### 2015-09-23 doc/committee-meeting/committee-meeting-1-summary.tex
Changes to R package

\documentclass{article}
\usepackage[T1]{fontenc}
\usepackage[osf]{garamondx}
\usepackage[garamondx,cmbraces]{newtxmath}
\usepackage{fullpage}
\usepackage{wrapfig}
\usepackage{tikz}
\usepackage[backend=biber]{biblatex}

\usetikzlibrary{arrows,shapes,positioning}

\author{Rosemary McCloskey}
\title{M.Sc. Thesis Proposal: Inference of Network Parameters with Kernel-ABC}
\addbibresource{refs.bib}
\begin{document}

\maketitle

\section{Motivation}

\begin{wrapfigure}{R}{0.6\textwidth}
    \centering
    \includegraphics[scale=0.8, trim=0 0 2in 0, clip=true]{transmission.pdf}
    \includegraphics[scale=0.8, trim=0.5in 0 2in 0, clip=true]{transmission_tree.pdf}
    \caption{Correspondence between transmission events (left) and viral
    phylogeny (right).}
    \label{fig:trans}
\end{wrapfigure}

RNA viruses, such as HIV, evolve rapidly within their hosts, accumulating
significant genetic diversity within months of a transmission event. This
process of transmission and subsequent divergent evolution is similar to the
speciation process observed in the evolution of macro-organisms, which often
occurs following the geographic isolation of a sub-population in a new
environment. For both viruses and macro-organisms, these isolation events
roughly correspond to branching points in their ``family tree'', or
\emph{phylogeny}. The relationship between transmissions and viral phylogenies
is illustrated in Figure \ref{fig:trans}, although several complicating factors
prevent the correspondence from being exact.

The practical consequence of this correspondence is the potential to infer
information about transmission patterns from the viral phylogeny. At least
theoretically, phylogenies are easy to construct, and are less susceptible to
biases associated with more conventional methods of investigating transmission
(such as contact tracing). This is the basis for the emerging area of research
known as \emph{phylodynamics}.

Recently, many phylodynamic investigations have been performed aiming to
uncover population-level parameters, such as basic reproductive
number~\cite{stadler2014insights}. Most often, these studies assume a
homogeneously mixed population where contact probabilities between all pairs of
individuals are the same, regardless of geographic or social structure.
However, it is also known that different contact network structures can have
strong effects on epidemic dynamics~\cite{volz2009epidemic}. A small number of
studies have investigated the potential for phylodynamic inference of contact
network structure~\cite{colijn2014phylogenetic,leventhal2012inferring} by
showing that extremely different networks result in differently shaped
phylogenies. I plan to carry this idea further, by identifying parameters of
contact networks which can be effectively estimated with phylodynamic methods
for real, ongoing epidemics.

\section{Methods}

\begin{wrapfigure}{R}{0.6\textwidth}
    \centering
    \begin{tikzpicture}
        [state/.style={rounded corners, rectangle, draw, inner sep=6pt, text width=2cm, align=center},
        decision/.style={diamond, aspect=2, draw},
        every path/.style={->, >=stealth, thick}]
        \node[state] (1) {guess parameters};
        \node[state, below=of 1] (2) {simulate network};
        \node[state, below=of 2] (3) {simulate tree};
        \node[state, right=of 3] (4) {compare to true tree};
        \node[decision, above=of 4] (5) {close enough?};
        \node[state, right=of 5] (6) {accept parameters};
        \node at (1.north west) [anchor=north west] {1};
        \node at (2.north west) [anchor=north west] {2};
        \node at (3.north west) [anchor=north west] {3};
        \node at (4.north west) [anchor=north west] {4};
        \node at (6.north west) [anchor=north west] {5};
        \draw (1) -- (2);
        \draw (2) -- (3);
        \draw (3) -- (4);
        \draw (4) -- (5);
        \draw (5.north) |- (1);
        \draw (5) -- (6);
    \end{tikzpicture}
    \caption{Schematic of phylodynamic ABC-rejection algorithm.}
    \label{fig:abc}
\end{wrapfigure}

I propose to use \emph{approximate Bayesian computation} (ABC) to investigate
the contact network structure underlying observed viral phylogenies. ABC
provides a method to estimate the parameters of models without explicit
likelihood calculations, when these are unavailable or computationally
prohibitive.

To illustrate, I describe the simplest ABC algorithm, called ABC-rejection, as
applied to this problem. A graphical schematic of the algorithm is shown in
Figure~\ref{fig:abc}. The objective of the algorithm is to fit a generative
model for contact networks with a small number of parameters. For example, the
Barabasi-Albert model~\cite{albert2002statistical}, in its simplest form, has
two parameters: the number of nodes in the network ($N$), and the mean degree
of each node in the network ($k$, corresponding to the mean number of contacts
per person). We assume, as is commonly done in phylodynamic studies, that the
true transmission tree is known with certainty.

We begin by guessing a set of parameters (Figure \ref{fig:abc}, box 1), for
example, an at-risk population size of $N = 5000$ with an average of $k = 8$
contacts per person.  A network is generated which satisfies these parameters
(box 2). We then simulate an epidemic over the network, and by keeping track of
who infects who, we derive a simulated transmission tree (box 3). This
simulated tree is compared to the true tree using a phylogenetic
kernel~\cite{poon2013mapping} (box 4). If the simulate tree is sufficiently
similar to the real tree, we accept the model parameters as possibly having
generated the true data (box 5). Otherwise, another set of parameters is
guessed, and we start at the beginning.

In practice, I intend to use either ABC-MCMC or ABC-SMC, rather than this
rejection method. These algorithms have the same basic premise as described
above, but use more sophisticated methods of ``guessing'' parameters, to reduce
the number of iterations needed to converge to the true values.

\section{Progress}

I have implemented a method to simulate transmission trees over networks, and
partially validated it by reconstructing Figure 1a
from~\cite{leventhal2012inferring}, which examined the effect of network type
(scale-free vs. small-world vs. random) on a particular tree summary statistic
(Sackin's index). As a proof of concept, I used the same parameters as in that
paper to develop a tree kernel-based classifier for tree shape, which was
94.6\% accurate at classifying the network type underlying simulated
phylogenies (Figure \ref{fig:kmat}).

\section{Goals}

The schematic in Figure \ref{fig:abc} outlines five major tasks required for
this project.

\begin{enumerate}
    \item Decide how to parameterize contact networks, and what sorts of
        parameters are both useful and feasible to infer.
    \item Simulate contact networks from a generative model. Several
        out-of-the-box graph generators exist (eg. for scale-free, small-world,
        and random graphs), but may need to be adapted depending on the
        generative model.
    \item Simulate phylogenies over contact networks. This has already been
        done.
    \item Compare simulated to observed phylogenies. I will use the tree
        kernel~\cite{poon2013mapping} for this task, which has an existing
        implementation.
    \item Implement the overall algorithm for proposing and accepting
        parameters, in other words, ABC-SMC or ABC-MCMC.
\end{enumerate}

\begin{figure}
    \centering
    \includegraphics[trim=0 0 0 0, clip=true, scale=0.5]{kmat.pdf}
    \label{fig:kmat}
    \caption{Principal components projection of kernel matrix of 300 simulated
    trees from three different network types.}
\end{figure}
\printbibliography

\end{document}

### 2015-09-23 doc/committee-meeting/presentation.tex
Changes to R package

\documentclass{beamer}
\usepackage{tikz}
\usepackage{tikz-qtree}
\usepackage{gillius}
\usepackage{mathdots}
\usepackage{abraces}
\usepackage{etex}
\usepackage{array}
\usepackage[backend=biber]{biblatex}

% http://tex.stackexchange.com/questions/12703/how-to-create-fixed-width-table-columns-with-text-raggedright-centered-raggedlef
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

\usetikzlibrary{external}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{decorations.pathreplacing}

\usetheme[everytitleformat=regular]{m}
\setbeamertemplate{navigation symbols}{}

\title{M.Sc. Thesis Proposal: Phylodynamic Inference of Contact Network
Structures with Approximate Bayesian Computation}
\author[RM \& AFYP]{Rosemary McCloskey and Art FY Poon}
\institute[UBC \& BCCfE]{BC Centre for Excellence in HIV/AIDS \\ University of British Columbia}
\date{September 22, 2015}

\newcommand{\dd}[2]{\frac{\text{d}\,#1}{\text{d}\,#2}}

\begin{document}
\definecolor{red}{RGB}{228,26,28}
\definecolor{blue}{RGB}{55,126,184}
\definecolor{green}{RGB}{77,175,74}
\definecolor{purple}{RGB}{152,78,163}

\maketitle
\begin{frame}{Outline}
    \begin{enumerate}
        \setlength{\itemsep}{12pt}
        \item Motivation
        \item Proposed methods
        \item Proof of concept
        \item Rotation project: phylogenetic clustering
    \end{enumerate}
\end{frame}

\section{Motivation}

\begin{frame}{Phylodynamics}
    \vspace{-1cm}
    \begin{center}
    \begin{tikzpicture}
        \uncover<1-2>{
        \node (a) {\includegraphics[width=1cm]{hiv}};
        \node at (a.north west) {\includegraphics[width=1cm]{person2}};
        }\uncover<2>{
        \node [right=1.5cm of a] (b) {\includegraphics[width=1cm]{hiv}};
        \node at (b.north west) {\includegraphics[width=1cm]{person1}};
        }

        \uncover<3->{
        \node [right=1.5cm of a, opacity=0.4] (b) {\includegraphics[width=1cm]{hiv}};
        \node at (b.north west) [opacity=0.4] {\includegraphics[width=1cm]{person1}};
        \node (a) [opacity=0.4] {\includegraphics[width=1cm]{hiv}};
        \node at (a.north west) [opacity=0.4] {\includegraphics[width=1cm]{person2}};
        }

        \uncover<2-3>{
        \draw[->, >=stealth, thick] (a) -- (b);
        }\uncover<3->{
        \node [below=2cm of a] (c) {\includegraphics[width=1cm]{hiv}};
        \node at (c.north west) {\includegraphics[width=1cm]{person2}};
        \node [below=2cm of b] (d) {\includegraphics[width=1cm]{hiv}};
        \node at (d.north west) {\includegraphics[width=1cm]{person1}};
        }\uncover<3->{
        \coordinate [left= of a] (timestart);
        \coordinate [left= of c] (timeend);
        \draw[->, >=stealth, thick] (timestart) -- node[auto, swap] {time} (timeend);
        }\uncover<4->{
        \coordinate [right=1.4 of a.center] (root);
        \draw [very thick] (a.center) +(1.4, 0) to +(1.4, 1);
        \draw [very thick] (c.center) -- (root);
        \draw [very thick] (d.center) -- (root);
        \draw [very thick] (root) to +(0, 1);
        }

        \uncover<2->{
        \node (note1) [right= of b, text width=4.5cm] {transmission to new host \\
        $\Rightarrow$ isolated sub-population};
        }
        \uncover<3->{
            \node (note2) [below=0.8cm of note1, text width=4.5 cm] {within-host evolution};
        }
        \uncover<4->{
            \node [below=of note2, text width=4.5 cm] {virus phylogeny \\ $\approx$ transmission tree};
        }
    \end{tikzpicture}
\end{center}
\end{frame}

\begin{frame}{State-of-the-art: population-level parameters}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            \includegraphics[width=\textwidth]{cumulative}
        \end{column}
        \begin{column}{0.5\textwidth}
            \includegraphics[width=\textwidth]{tmrca}
        \end{column}
    \end{columns}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            \tiny{Volz, Erik, and Sergei Pond. ``Phylodynamic analysis of Ebola virus in the 2014 Sierra Leone epidemic.'' PLoS currents 6 (2014).}
        \end{column}
        \begin{column}{0.5\textwidth}
            \tiny{Liao, Huanan, et al. ``Phylodynamic analysis of the dissemination of HIV-1 CRF01\_AE in Vietnam.'' Virology 391.1 (2009): 51-56.}
        \end{column}
    \end{columns}
\end{frame}

\begin{frame}{Network structure profoundly affects tree shape}

    \includegraphics[trim=0 -1cm 0 0]{odea-f2a}\includegraphics{odea-f2d}

    \begin{center}
    ``Contact heterogeneity is well known to have a strong effect on infectious
    disease dynamics. We have shown how the relationship between infectious
    disease dynamics and genealogies is similarly sensitive to the contact
    heterogeneity specified by a network.''

    \tiny{O'Dea, Eamon B., and Claus O. Wilke. ``Contact heterogeneity and
        phylodynamics: how contact networks shape parasite evolutionary
        trees.'' Interdisciplinary Perspectives on Infectious Diseases 2011
    (2010).}
    \end{center}
\end{frame}

\begin{frame}{Research question}
    \large
    \begin{enumerate}
        \setlength{\itemsep}{12pt}
        \item Are there contact network parameters which are possible to infer
            with phylogenetic methods?
        \pause
        \item What are the estimated values of those parameters for the HIV
            subtype B-infected population in British Columbia?
    \end{enumerate}
\end{frame}

\section{Proposed Methods}

\begin{frame}{Approximate Bayesian computation}
    \begin{tabular}{C{4cm}C{3cm}C{2cm}}
        goal & & data \\
        \hline\\
        $N$ = number of people, \newline $k$ = average number of partners & &
        \includegraphics[scale=0.5]{tree2.pdf} \\
        \\ \hline \\
        \uncover<2->{$N = 1000, k = 3$} & 
        \uncover<3->{\includegraphics[scale=0.5]{net1.pdf}} &
        \uncover<4->{\includegraphics[scale=0.5]{tree1.pdf}} \\
        \uncover<5->{$N = 1000, k = 12$} & 
        \uncover<6->{\includegraphics[scale=0.5]{net2.pdf}} &
        \uncover<7->{\includegraphics[scale=0.5]{tree3.pdf}} \\
    \end{tabular}
\end{frame}

\begin{frame}{Approximate Bayesian computation}
    \begin{tikzpicture}
        [state/.style={rounded corners, rectangle, draw, inner sep=8pt, text width=2cm, align=center},
        decision/.style={diamond, aspect=2, draw},
        every path/.style={->, >=stealth, thick}]
        \node[state] (1) {guess parameters};
        \node[state, below=of 1] (2) {simulate network};
        \node[state, below=of 2] (3) {simulate tree};
        \node[state, right=of 3] (4) {compare to true tree};
        \node[decision, above=of 4] (5) {close enough?};
        \node[state, right=of 5] (6) {accept parameters};
        \node[above=of 6, text width=4.5cm] {\textbf{Goal}: network parameters \newline (eg. $N$, $k$)};
        \node[below=of 6, text width=4.5cm] {\textbf{Data}: observed \newline transmission tree};
        \node at (1.north west) [anchor=north west] {1};
        \node at (2.north west) [anchor=north west] {2};
        \node at (3.north west) [anchor=north west] {3};
        \node at (4.north west) [anchor=north west] {4};
        \node at (6.north west) [anchor=north west] {5};
        \draw (1) -- (2);
        \draw (2) -- (3);
        \draw (3) -- (4);
        \draw (4) -- (5);
        \draw (5.north) |- (1);
        \draw (5) -- (6);
    \end{tikzpicture}
\end{frame}

\begin{frame}{Plan}
    Data: phylogeny of $\sim$8000 HIV infections in BC.
    \begin{enumerate}
        \item[0.] Literature review.
        \item Guess parameters: preliminary experiments to determine which
            parameters are identifiable.
        \pause
        \item Simulate network: out-of-the-box, possibly exponential random
            graph model (ERGM).
        \pause
        \item Simulate tree: next-reaction method (already done).
        \pause
        \item Compare to true tree: Art's tree kernel.
        \pause
        \item ABC algorithm: sequential Monte-Carlo or Markov chain Monte Carlo.
    \end{enumerate}
\end{frame}

\section{Proof of Concept}

\begin{frame}{Verifying tree simulations}
    Leventhal et al: Sackin's index for simulated phylogenies over networks of
    5000 nodes.
    \begin{center}
    \includegraphics[scale=1.5, trim=0 0 3.5cm 0, clip=true]{f1}
    \uncover<2->{\includegraphics[scale=0.07]{sackin}}
    \end{center}

    \tiny{Leventhal, Gabriel E., et al. ``Inferring epidemic contact structure from phylogenetic trees.'' PLoS Comput Biol 8.3 (2012)}
\end{frame}

\begin{frame}{Kernel matrices of simulated trees}
    \includegraphics[scale=0.3]{kmat5k}
    \includegraphics[scale=0.3]{kmat100}

    $\qquad$$\qquad$ 5000 nodes \hfill 100 nodes $\qquad$$\qquad$$\quad$
\end{frame}

\begin{frame}{Aside: clustering}
    \includegraphics[trim=1in 0 0 4in, clip=true]{f3a}
    \includegraphics[scale=0.7]{f3b}

    \tiny{Lewis, Fraser, et al. ``Episodic sexual transmission of HIV revealed by molecular phylodynamics.'' PLoS Med 5.3 (2008).}
\end{frame}

\begin{frame}{Aside: clustering}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            \includegraphics[width=\textwidth]{ba-net}
        \end{column}
        \begin{column}{0.5\textwidth}
            \includegraphics[width=\textwidth]{ba-tree}
        \end{column}
    \end{columns}
\end{frame}

\section{Rotation Project: Phylogenetic Clustering}

\begin{frame}{Clustering by Branching Rates}
    \only<1>{\includegraphics[height=\textwidth,angle=-90]{pcbr_example}}
    \only<2->{\includegraphics[height=\textwidth,angle=-90]{pcbr_example_color}}

    \hspace{1.2cm}
    \uncover<3->{
    \begin{tikzpicture}
        [node distance=0.5,
         every node/.style={circle, draw=red, fill=red, inner sep=4pt}]
        \node (a) { };
        \node [below=of a] (b) { };
        \node [below=of b] (c) { };
        \node [right=of b] (d) { };
        \node [left=of b] (e) { };
        \draw [thick] (a) -- (b) -- (d) -- (a);
        \draw [thick] (a) -- (b) -- (e) -- (a);
        \draw [thick] (c) -- (b) -- (d) -- (c);
        \draw [thick] (c) -- (b) -- (e) -- (c);
        \node [right=of a, draw=none, fill=none, text=red] {fast};
    \end{tikzpicture}
    }
    \hspace{0.2cm}
    \uncover<4->{
    \begin{tikzpicture}
        [node distance=0.5,
         every node/.style={circle, draw=blue, fill=blue, inner sep=4pt}]
        \node (a) { };
        \node [below=of a] (b) { };
        \node [below=of b] (c) { };
        \node [right=of b] (d) { };
        \node [left=of b] (e) { };
        \draw [thick] (a) -- (b) -- (d) -- (c) -- (e);
        \node [right=of a, draw=none, fill=none, text=blue] {slow};
    \end{tikzpicture}
    }
\end{frame}

\begin{frame}{Modelling branching rates}
    \begin{tikzpicture}
        [every node/.style={inner sep=2pt}]
        \coordinate (a);
        \coordinate [below right=3 of a] (b);
        \draw [thick] (a) -- ++ (-0.5, 0.5) node [anchor=south east] {$\ddots$};
        \draw [thick] (a) -- ++ (-0.5, -0.4) node [inner ysep=-4pt, inner xsep=2pt, anchor=north east] {$\iddots$};
        \draw [thick] (b) -- ++ (-0.5, -0.4) node [inner ysep=-4pt, inner xsep=2pt, anchor=north east] {$\iddots$};
        \draw [thick] (b) -- ++ (0.5, -0.5) node [anchor=north west, inner ysep=-4pt] {$\ddots$};
        \draw [thick] (a) -- (b);
        \uncover<2->{
            \node at (a) [circle, draw=blue, fill=blue] { };
            \node [right=0.2 of a, anchor=west, text=blue] {slow};
            \node (note1) [above right=0.5 and 6 of a, text width=4cm] {branching rates \\ attached to nodes};
        }
        \uncover<3->{
            \draw [thick] (a) -- node [auto] {
                $T \sim $ Exponential($\lambda$)
            } (b);
            \node at (a) [circle, draw=blue, fill=blue] { };
            \node (note2) [below=0.5 of note1, text width=4cm] {branching is \\ a Poisson process};
        }
        \uncover<4->{
            \node at (b) [circle, draw=red, fill=red] { };
            \node [right=0.2 of b, anchor=west, text=red] {fast};
            \node [below=0.5 of note2, text width=4cm] {rate switching is \\ a Poisson process};
        }
    \end{tikzpicture}

    \begin{center}
    \begin{tikzpicture}
    \uncover<5->{
        \node [circle, draw, text=blue, inner sep=2pt] (slow) {$\lambda_{\text{slow}}$};
        \node [right=of slow, circle, draw, text=red, inner sep=2pt] (fast) {$\lambda_{\text{fast}}$};
        \draw [->, >=stealth, thick] (slow) to [bend right] node[auto, swap] {$q_{10}$} (fast);
        \draw [->, >=stealth, thick] (fast) to [bend right] node[auto, swap] {$q_{01}$} (slow);
    }
    \uncover<6->{
        \node [right=of fast] {Markov-modulated Poisson process (MMPP)};
    }
    \end{tikzpicture}
    \end{center}
\end{frame}

\begin{frame}{Validation}
    Works fairly well for superspreader networks.
    \vspace{0.5cm}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            \includegraphics[width=\textwidth]{stadler2}
        \end{column}
        \begin{column}{0.5\textwidth}
            \includegraphics[width=\textwidth]{2_2_008}
        \end{column}
    \end{columns}
\end{frame}

\begin{frame}{Validation}
    Doesn't work for interconnected islands model.
    \vspace{0.5cm}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            \includegraphics[width=\textwidth]{stadler1}
        \end{column}
        \begin{column}{0.5\textwidth}
            \includegraphics[width=\textwidth]{1_2_197}
        \end{column}
    \end{columns}
\end{frame}

\begin{frame}{Validation}
    Doesn't work well for scale-free networks (hard to define ``true'' clusters).
    \vspace{0.5cm}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            \includegraphics[width=\textwidth]{stadler3}
        \end{column}
        \begin{column}{0.5\textwidth}
            \includegraphics[width=\textwidth]{3_3_072}
        \end{column}
    \end{columns}
\end{frame}

\begin{frame}{Validation}
    No clusters detected with differential risk trees.

    \vspace{0.5cm}
    \begin{center}
    \includegraphics[scale=0.44445]{diffrisk}
    \end{center}
\end{frame}

\begin{frame}
    \begin{center}
    \Huge{Thank you!}
    \end{center}
\end{frame}

\end{document}

### 2015-09-23 doc/committee-meeting/aside/make_net.R

@@ -1,0 +1,54 @@
+#!/usr/bin/env Rscript
+
+library(igraph)
+library(ape)
+g <- barabasi.game(100, m=2, directed=FALSE)
+g <- as.directed(g, "mutual")
+V(g)$remove <- 0
+E(g)$transmit <- 1
+write.graph(g, "ba.net", "gml")
+
+clus <- cluster_edge_betweenness(g, directed=FALSE)
+clus <- tapply(1:vcount(g), membership(clus), identity, simplify=FALSE)        
+subg <- mapply(induced_subgraph, list(g), clus, impl="create_from_scratch",       
+               SIMPLIFY=FALSE)                                                   
+                                                                                 
+# keep only clusters which have non-zero clustering coefficient                  
+keep.clus <- clus[head(order(-sapply(subg, transitivity)), 1)]                  
+
+rbw <- rainbow(length(keep.clus))                                                
+keep.memb <- mapply(match, list(1:vcount(g)), keep.clus)                       
+keep.memb <- unlist(ifelse(!is.na(apply(keep.memb, 1, any)),                     
+                           apply(keep.memb, 1, function (x) which(!is.na(x))),   
+                           NA))                                                  
+keep.edge <- do.call(c, lapply(keep.clus, function (x) E(g)[x %--% x]))        
+                                                                                 
+edge.col <- rep("gray", ecount(g))                                             
+edge.col[keep.edge] <- "black" 
+
+pdf("ba-net.pdf")
+par(mar=c(0, 0, 0, 0))
+plot(g, 
+     edge.arrow.size=0,
+     vertex.label=ifelse(is.na(keep.memb), NA, V(g)),
+     vertex.size=ifelse(is.na(keep.memb), 3, 10),                                
+     vertex.color=rbw[keep.memb],                                                
+     edge.color=edge.col)
+dev.off()
+
+system("nettree ba.net ba.nwk")
+
+t <- read.tree("ba.nwk")
+t$tip.label <- as.character(as.integer(t$tip.label)+1)
+
+tip.col <- rep("black", 100)
+tip.col[match(keep.clus[[1]], as.integer(t$tip.label))] <- "red"
+tip.cex <- rep(par("cex"), 100)
[…]

### 2015-09-23 programs/netabc/R/collect.R

@@ -1,11 +1,52 @@
+collect.metadata.mm <- function (mm.file)
+{
+    as.data.frame(yaml.load(substring(process.archive(mm.file, readLines, n=2)[2], 2)))
+}
+
+collect.metadata.newick <- function (nwk.file)
+{
+    as.data.frame(yaml.load(substring(process.archive(nwk.file, readLines, n=1)[1], 2)))
+}
+
+collect.metadata <- function (data.files)
+{
+    data.files <- sort(Sys.glob(data.files))
+    names <- basename(data.files)
+    names <- sapply(names, function (x) sub("\\.gz", "", x))
+    names <- sapply(names, function (x) sub("\\.tar", "", x))
+    names <- sapply(names, function (x) sub("\\.bz2?", "", x, perl=TRUE))
+    names <- sapply(names, function (x) sub("\\.xz", "", x, perl=TRUE))
+
+    extensions <- lapply(strsplit(names, ".", fixed=TRUE), tail, 1)
+    metadata <- do.call(rbind, mapply(function (f, ext) {
+        switch(ext, mtx=collect.metadata.mm(f),
+                    nwk=collect.metadata.newick(f))
+    }, data.files, extensions, SIMPLIFY=FALSE))
+
+    wcon <- textConnection("buf", open="w")
+    write.table(metadata, wcon)
+    close(wcon)
+
+    rcon <- textConnection(buf, open="r")
+    metadata <- read.table(rcon)
+    close(rcon)
+
+    metadata
+}
+
 # Collect data from several files whose names contain metadata in the format
 # key1-value1_key2-value2.foo
 collect.data <- function (data.files)
 {
-    names <- basename(data.files) 
-    names <- lapply(names, function (x) sub("(.*)\\..*?$", "\\1", x, perl=TRUE))
-    names <- sapply(names, strsplit, "_")
+    names <- basename(data.files)
 
+    names <- lapply(names, function (x) sub("\\.gz", "", x))
+    names <- lapply(names, function (x) sub("\\.tar", "", x))
+    names <- lapply(names, function (x) sub("\\.bz2?", "", x, perl=TRUE))
+    names <- lapply(names, function (x) sub("\\.xz", "", x, perl=TRUE))
+    names <- lapply(names, function (x) sub("(.*)\\..*?$", "\\1", x, perl=TRUE))
+
+    names <- sapply(names, strsplit, "_")
     names <- lapply(names, strsplit, "-")
[…]

### 2015-09-23 programs/netabc/R/ksvm.cv.R

@@ -38,6 +38,7 @@ ksvm.cv <- function (kmat, y, n.cv=1000, stats=c("accuracy"), show.progress=TRUE
             setTxtProgressBar(pb, getTxtProgressBar(pb)+1)
         sapply(stat.functions[stats], do.call, list(pred, y[holdout]))
     }, simplify=FALSE))
+    colnames(result) <- stats
 
     if (show.progress)
         close(pb)

### 2015-09-23 programs/netabc/R/plots.R

@@ -11,12 +11,19 @@ kpca.plot <- function (kmat, color=NULL, shape=NULL)
     ggplot(plot.data, plot.aes) + geom_point() + theme_bw()
 }
 
-summary.plot <- function (data, x, y, group=NULL, fun="mean")
+summary.plot <- function (data, x, y, group=NULL, fun="mean", x.factor=TRUE,
+                          y.factor=FALSE, group.factor=TRUE)
 {
-    plot.data <- aggregate(y~x+group, data, fun)
-    if (!is.null(group) & !is.factor(plot.data[,group]))
+    agg.formula <- as.formula(paste0(y, "~", x, "+", group))
+    plot.data <- aggregate(agg.formula, data, fun)
+
+    if (y.factor)
+        plot.data[,y] <- as.factor(plot.data[,y])
+    if (x.factor)
+        plot.data[,x] <- as.factor(plot.data[,x])
+    if (!is.null(group) & group.factor)
         plot.data[,group] <- as.factor(plot.data[,group])
 
-    ggplot(plot.data, aes_string(x=x, y=y, color=group)) + 
+    ggplot(plot.data, aes_string(x=x, y=y, color=group, group=group)) + 
         geom_point() + geom_line() + theme_bw()
 }

### 2015-09-23 programs/netabc/R/process.archive.R

@@ -1,0 +1,15 @@
+process.archive <- function (file.name, fun, ...)
+{
+    if (grepl("\\.bz2?$", file.name, perl=TRUE)) {
+        f <- bzfile(file.name)
+    } else if (grepl("\\.gz$", file.name)) {
+        f <- gzfile(file.name)
+    } else if (grepl("\\.xz$", file.name)) {
+        f <- xzfile(file.name)
+    } else {
+        f <- file(file.name)
+    }
+    result <- fun(f, ...)
+    close(f)
+    result
+}

### 2015-09-23 programs/netsim/src/nettree.c

@@ -81,7 +81,9 @@ struct nettree_options get_options(int argc, char **argv)
     {
         opts.net_file = fopen(argv[optind++], "r");
         if (optind < argc)
+        {
             opts.tree_file = fopen(argv[optind++], "w");
+        }
     }
 
     return opts;

### 2015-09-23 simulations/kernel-mean-degree.py

@@ -1,0 +1,86 @@
+#!/usr/bin/env python3
+
+from pprint import pprint
+import sys
+from makefile import *
+
+data_params = {"mean_degree": [2, 4, 8, 16],
+               "nnode": [5000],
+               "ntip": [100],
+               "seed": list(range(100))}
+
+data_targets = {
+    ("mean_degree", "nnode"): [
+        ("{file_basename}.gml",
+         "",
+         """
+         Rscript -e "g <- barabasi.game({nnode}, m={mean_degree}/2, directed=F)" \
+                 -e 'graph_attr(g, "comment") <- "{yaml}"' \
+                 -e "write.graph(SI.net(g), '{file_basename}.gml', format='gml')"
+         """)
+    ],
+    ("mean_degree", "nnode", "ntip", "seed"): [
+        ("{file_basename}.nwk",
+         "test/data/{mean_degree}_{nnode}.gml",
+         """
+         echo "#{yaml}" > $@
+         nettree --tree-tips {ntip} --seed {seed} $^ >> $@
+         """)
+    ]
+}
+
+data_makefile, data_files = create_makefile(data_params, data_targets, "test/data")
+
+results_params = {"branch_scaling": ["mean"],
+                  "decay_factor": [0.1, 0.2, 0.3],
+                  "rbf_variance": [0.0625, 0.125, 0.25, 0.5, 1, 2, 5],
+                  "sst_control": [0, 0.5, 1],
+                  "coal_power": [0, 1, 2],
+                  "ncrossv": [1000]}
+
+results_targets = {
+    ("branch_scaling", "decay_factor", "rbf_variance", "sst_control", "coal_power"): [
+        ("{file_basename}.mtx.bz2",
+         " ".join(f for f in data_files if f.endswith("nwk")),
+         """
+         echo "%%MatrixMarket matrix array real symmetric" > {file_basename}.mtx
+         echo "%{yaml}" >> {file_basename}.mtx
+         echo $(words $^) $(words $^) >> {file_basename}.mtx
+         for T1 in $(sort $^); do \
+             for T2 in $(sort $^); do \
+                 if [[ "$$T1" < "$$T2" || "$$T1" == "$$T2" ]]; then \
+                     treekernel -d -l {decay_factor} -g {rbf_variance} -b {branch_scaling} -s {sst_control} -c {coal_power} $$T1 $$T2 >> {file_basename}.mtx; \
+                 fi \
[…]

### 2015-09-23 simulations/makefile.py

@@ -1,0 +1,60 @@
+import itertools
+import re
+import yaml
+import os.path
+
+def clean_rule(rule):
+    rule = re.sub("(\s)\s+", "\\1", rule)
+    rule = re.sub("^\s", "", rule)
+    rule = re.sub("\s$", "", rule)
+    rule = re.sub("\n", "\n\t", rule)
+    return rule
+
+def create_makefile(params, targets, target_dir):
+
+    # targets is keyed by tuples of parameters, make them sorted
+    targets = {tuple(sorted(k)):v for k, v in targets.items()}
+
+    # clean up the rules so they can be directly printed to a Makefile
+    for k in targets:
+        for i, rule in enumerate(targets[k]):
+            target, depends, command = rule
+            targets[k][i] = (target, depends, clean_rule(command))
+
+    # if any of the parameters are all numeric, pad with zeroes so they are all
+    # the same width
+    for k, v in params.items():
+        if all(isinstance(x, (int)) for x in v):
+            v = [str(x) for x in v]
+            width = max(len(x) for x in v)
+            params[k] = ["0"*(width - len(x)) + x for x in v]
+
+        elif all(isinstance(x, (float, int)) for x in v):
+            s = [str(x) for x in v]
+            left_width = max(len(x.split(".")[0]) for x in s)
+            right_width = max(len(x.split(".")[1]) if "." in x else 0 for x in s)
+            width = left_width + 1 + right_width
+            fmt = "{{:0{width}.{right_width}f}}".format(width=width, right_width=right_width)
+            params[k] = [fmt.format(x) for x in v]
+
+    # keep track of a list of targets, and also the text for the Makefile
+    to_make = []
+    makefile = ""
+
+    # go through targets in order of fewest-to-most parameter dependencies
+    for num_params in range(len(params)+1, -1, -1):
+        for combo in itertools.combinations(params.keys(), num_params):
+            key = tuple(sorted(combo))
+            try:
+                for target, depends, command in targets[key]:
+                    for p in itertools.product(*[params[k] for k in key]):
+                        d = dict(zip(key, p))
+                        d["yaml"] = yaml.dump(d, width=1000)[:-1]
+                        d["file_basename"] = os.path.join(target_dir, "_".join(map(str, p)))
+                        makefile += "{}: {}\n\t{}\n\n".format(target.format(**d),
+                                depends.format(**d), command.format(**d))
+                        to_make.append(target.format(**d))
+            except KeyError as e:
+                pass
+    return makefile, to_make
+

### 2015-10-02 programs/pcbr/src/likelihood.c
Moved pcbr from separate repo

@@ -1,0 +1,465 @@
+#define NDEBUG
+
+#include <stdio.h>
+#include <stdlib.h>
+#include <string.h>
+#include <signal.h>
+#include <pll/pll.h>
+#include <math.h>
+#include <assert.h>
+#include <Rinternals.h>
+#include <Rembedded.h>
+#include <gsl/gsl_matrix.h>
+#include <gsl/gsl_linalg.h>
+#include <gsl/gsl_odeiv2.h>
+#include <gsl/gsl_cdf.h>
+#include <gsl/gsl_randist.h>
+#include "likelihood.h"
+#include "rwrapper.h"
+#include "util.h"
+
+// helpers for dynamic programming
+#define L(i, j) (w->L[ (i) * nrates + (j) ])
+#define C(i, j) (w->C[ (i) * nrates + (j) ])
+#define A(i) (w->A[ (i) ])
+#define pi(i) (w->pi[ (i) ])
+#define scale(i) (w->scale[ (i) ])
+#define left_child(i) (w->topology[ (i) * 2 ])
+#define right_child(i) (w->topology[ (i) * 2 + 1])
+#define t(i) (w->t[ (i) ])
+
+#define ROOT w->nnode - 1
+
+#define USE_TIPS TRANS_ONLY
+#define TRANS_PER_BRANCH ONE
+#define CHEATING 0
+#define TIP_PDF 0
+
+void calculate_P(pcbr_workspace *w, double *lambda, trans_per_branch tpb, 
+                 use_tips ut);
+
+pcbr_workspace *pcbr_create(const pllNewickTree *tree, int nrates)
+{
+    pcbr_workspace *w = malloc(sizeof(pcbr_workspace));
+
+    w->nnode = tree->nodes;
+    w->nrates = nrates;
+
+    w->topology = malloc(2 * tree->nodes * sizeof(int));
+    w->branch_lengths = malloc(tree->nodes * sizeof(double));
+    w->sorted_branch_lengths = malloc(tree->nodes * sizeof(double));
+    flatten(tree->tree, tree->nodes, w->topology, w->branch_lengths, 0);
+
+    w->t = malloc(tree->nodes * sizeof(double));
+    w->brlen_order_inv = malloc(tree->nodes*sizeof(int));
+    scale_branches(w);
+
+    // dynamic programming matrices and scratch space
+    w->L = malloc(nrates * sizeof(double) * tree->nodes);
[…]

### 2015-10-02 programs/pcbr/src/rwrapper.c

@@ -1,0 +1,50 @@
+#include <string.h>
+#include <Rinternals.h>
+#include <Rembedded.h>
+#include "rwrapper.h"
+
+SEXP construct_call(const SEXP *args, const char **tags, int narg);
+
+void start_R(void)
+{
+	char *argv[] = { "R", "--no-save", "--silent" };
+    Rf_initEmbeddedR(3, argv);
+}
+
+void stop_R(void)
+{
+    Rf_endEmbeddedR(0);
+}
+
+SEXP call_R(const char *func, const SEXP *args, const char **tags, int narg)
+{
+    int Rerror;
+    SEXP expr, result;
+    expr = PROTECT(LCONS(install(func), construct_call(args, tags, narg)));
+    result = R_tryEval(expr, R_GlobalEnv, &Rerror);
+    UNPROTECT(1);
+    return Rerror ? R_NilValue : result;
+}
+
+void R_library(const char *libname)
+{
+    SEXP expr;
+    int Rerror;
+    expr = PROTECT(lang2(install("library"), mkString(libname)));
+    R_tryEval(expr, R_GlobalEnv, &Rerror);
+    UNPROTECT(1);
+}
+
+// inspiration from Advanced R
+SEXP construct_call(const SEXP *args, const char **tags, int narg)
+{
+    SEXP call;
+    if (narg == 1) 
+        call = PROTECT(LCONS(args[0], R_NilValue));
+    else
+        call = PROTECT(LCONS(args[0], construct_call(&args[1], &tags[1], narg-1)));
+    if (tags[0])
+        SET_TAG(call, install(tags[0]));
+    UNPROTECT(1);
+    return call;
+}

### 2015-10-02 programs/pcbr/src/rwrapper.h

@@ -1,0 +1,24 @@
+#ifndef RWRAPPER_H
+#define RWRAPPER_H
+
+#include <Rinternals.h>
+
+/** Initialize an embedded R session. */
+void start_R(void);
+
+/** Terminate an embedded R session. */
+void stop_R(void);
+
+/** Call an R function.
+ *
+ * \param func the name of the function to call
+ * \param args arguments to supply to the function
+ * \param tags names for the arguments, if any
+ * \param narg number of arguments
+ */
+SEXP call_R(const char *func, const SEXP *args, const char **tags, int narg);
+
+/** Import an R library into the embedded session. */
+void R_library(const char *libname);
+
+#endif /* RWRAPPER_H */

### 2015-10-02 programs/pcbr/src/thread.h

@@ -1,0 +1,29 @@
+#ifndef THREAD_H
+#define THREAD_H
+
+#include <pll/pll.h>
+#include "config.h"
+#include "likelihood.h"
+
+typedef struct {
+    pcbr_workspace *w;
+    double **args;
+    double *arFunvals;
+    int neval;
+    use_tips ut;
+    trans_per_branch tpb;
+    int tip_pdf;
+    int cheating;
+} thread_data;
+
+thread_data *create_thread_data(const pllNewickTree *tree, int nrate, 
+        int nthread, int neval, use_tips ut, trans_per_branch tpb, int tip_pdf,
+        int cheating);
+
+void set_args(thread_data *tdata, double **args, int nthread, int narg);
+
+void *do_likelihood(void *tdata);
+
+void destroy_thread_data(thread_data *tdata, int nthread);
+
+#endif

### 2015-10-02 programs/pcbr/src/tree.h

@@ -1,0 +1,8 @@
+#ifndef TREE_H
+#define TREE_H
+
+#include "likelihood.h"
+
+int output_tree(const pcbr_workspace *tree, const double *rates, int nnode, char *buf);
+
+#endif

### 2015-10-02 programs/pcbr/src/util.c

@@ -1,0 +1,126 @@
+#include <math.h>
+#include <stdlib.h>
+#include "util.h"
+
+void swap(double *x, int i, int j)
+{
+    double tmp = x[i];
+
+    x[i] = x[j];
+    x[j] = tmp;
+}
+
+int which_max(double *x, int n)
+{
+    double max = x[0];
+    int which_max = 0;
+    int i;
+
+    for (i = 1; i < n; ++i) {
+        if (max < x[i]) {
+            max = x[i];
+            which_max = i;
+        }
+    }
+    return which_max;
+}
+
+double double_sum(double *x, int n)
+{
+    int i;
+    double sum = 0;
+
+    for (i = 0; i < n; ++i)
+        sum += x[i];
+    return sum;
+}
+
+double double_max(double *x, int n)
+{
+    int i;
+    double max = x[0];
+
+    for (i = 1; i < n; ++i)
+        max = fmax(max, x[i]);
+    return max;
+}
+
+double double_min(double *x, int n)
+{
+    int i;
+    double min = x[0];
+
+    for (i = 1; i < n; ++i)
+        min = fmin(min, x[i]);
+    return min;
+}
+
+void double_exp(double *x, double *result, int n)
+{
+    int i;
+
+    for (i = 0; i < n; ++i)
+        result[i] = exp(x[i]);
+}
+
+void double_log(double *x, double *result, int n)
+{
+    int i;
+
+    for (i = 0; i < n; ++i)
+        result[i] = log(x[i]);
+}
+
+int get_scale(double *x, int n)
+{
+    int i, nkeep = 0;
+    double ilog, logsum = 0, logmax = LOG_ZERO;
+
+    for (i = 0; i < n; ++i)
+        logmax = fmax(x[i] == 0 ? LOG_ZERO : log10(x[i]), logmax);
+
+    for (i = 0; i < n; ++i) {
+        if (x[i] > 0) {
+            ilog = log10(x[i]);
+            if (ilog - logmax > LOG_ZERO) {
+                logsum += ilog;
+                nkeep += 1;
+            }
+        }
+    }
+    if (nkeep == 0)
+        return 0;
+    return (int) ceil(logsum / nkeep);
+}
+
+int compare_doubles(const void *a, const void *b)
+{
+    return ( *(double*) a > *(double*) b ? 1 : -1);
+}
+
+void order(const double *x, double *sorted, int *order, int n)
+{
[…]

### 2015-10-02 programs/pcbr/tests/check_thread.c

@@ -1,0 +1,197 @@
+#include <check.h>
+#include "../src/thread.h"
+#include "../src/likelihood.h"
+#include "../C-Thread-Pool/thpool.h"
+
+Suite *thread_suite(void);
+
+const char *tree_6 = "(((t5:0.4119189046,t6:0.7903786616):0.8664339224,t4:0.03210153081):0.5706496239,((t2:0.805275284,t1:0.7685161901):0.4054109394,t3:0.4271961174):0.3987961309);";
+
+START_TEST (test_thread_data_create)
+{
+    int nthread = 3;
+    int nrate = 2;
+    int popsize = 100;
+
+    pllNewickTree *tree = pllNewickParseString(tree_6);
+    thread_data *tdata = create_thread_data(tree, nrate, nthread, popsize);
+
+    ck_assert_int_eq(tdata[0].neval, 34);
+    ck_assert_int_eq(tdata[1].neval, 33);
+    ck_assert_int_eq(tdata[2].neval, 33);
+
+    ck_assert_int_eq(memcmp(tdata[0].w->branch_lengths, tdata[1].w->branch_lengths, 11*sizeof(double)), 0);
+    ck_assert_int_eq(memcmp(tdata[1].w->branch_lengths, tdata[2].w->branch_lengths, 11*sizeof(double)), 0);
+
+    destroy_thread_data(tdata, nthread);
+    pllNewickParseDestroy(&tree);
+}
+END_TEST
+
+START_TEST (test_set_args)
+{
+    int nthread = 3;
+    int nrate = 2;
+    int popsize = 10;
+    int narg = nrate * nrate;
+    int i, j;
+    double **args = malloc(popsize * sizeof(double*));
+
+    pllNewickTree *tree = pllNewickParseString(tree_6);
+    thread_data *tdata = create_thread_data(tree, nrate, nthread, popsize);
+
+    srand(0);
+    for (i = 0; i < popsize; ++i) {
+        args[i] = malloc(narg * sizeof(double));
+        for (j = 0; j < narg; ++j)
+            args[i][j] = (double) rand() / (double) RAND_MAX;
+    }
+    set_args(tdata, args, nthread, narg);
+
+    for (i = 0; i < 3; ++i) {
+        for (j = 0; j < narg; ++j) {
+            ck_assert(tdata[0].args[i][j] == args[i][j]);
+            ck_assert(tdata[1].args[i][j] == args[i+4][j]);
[…]

### 2015-10-02 programs/pcbr/tests/check_tree.c

@@ -1,0 +1,98 @@
+#include <check.h>
+#include <stdlib.h>
+#include <pll/pll.h>
+#include <Rinternals.h>
+#include "../src/tree.h"
+#include "../src/likelihood.h"
+#include "config.h"
+
+#define NO_CHILD -1
+
+Suite *tree_suite(void);
+
+const char *tree_3 = "(v1:0.5,(v3:0.5,v4:0.25)v2:0.25)v0;";
+
+START_TEST (test_flatten)
+{
+    pllNewickTree *tree = pllNewickParseString(tree_3);
+    double *bl = malloc(5*sizeof(double));
+    int *order = malloc(10*sizeof(double));
+
+    flatten(tree->tree, tree->nodes, order, bl, 0);
+    ck_assert(order[0] == NO_CHILD);
+    ck_assert(order[1] == NO_CHILD);
+    ck_assert(order[2] == NO_CHILD);
+    ck_assert(order[3] == NO_CHILD);
+    ck_assert(order[4] == NO_CHILD);
+    ck_assert(order[5] == NO_CHILD);
+    ck_assert(order[6] == 1);
+    ck_assert(order[7] == 2);
+    ck_assert(order[8] == 0);
+    ck_assert(order[9] == 3);
+    ck_assert(bl[0] == 0.5);
+    ck_assert(bl[1] == 0.5);
+    ck_assert(bl[2] == 0.25);
+    ck_assert(bl[3] == 0.25);
+    ck_assert(bl[4] == 0);
+        
+    free(order);
+    free(bl);
+    pllNewickParseDestroy(&tree);
+}
+END_TEST
+
+START_TEST (test_output)
+{
+    pllNewickTree *tree = pllNewickParseString(tree_3);
+    pcbr_workspace *p = pcbr_create(tree, 2);
+    char buf[1000];
+    double rates[2] = {1, 2};
+    int i;
+
+    for (i = 0; i < 5; ++i) {
+        p->A[i] = i % 2;
+        p->L[2*i] = p->L[2*i+1] = 1;
+        p->scale[i] = 0;
+    }
+
+    get_clusters(p, rates);
+    output_tree(p, rates, p->nnode, buf);
+    fprintf(stderr, "%s\n", buf);
[…]

### 2015-10-02 programs/pcbr/tests/check_util.c

@@ -1,0 +1,103 @@
+#include <check.h>
+#include "../src/util.h"
+
+Suite *util_suite(void);
+
+START_TEST (test_swap)
+{
+    double test[4] = {1, 2, 3, 4};
+    swap(test, 0, 2);
+    ck_assert(test[0] == 3);
+    ck_assert(test[2] == 1);
+}
+END_TEST
+
+START_TEST (test_get_scale_nonzero)
+{
+    double x[2] = { 1e5, 1e-10 };
+    int scale = get_scale(x, 2);
+    ck_assert_int_eq(scale, -2);
+}
+END_TEST
+
+START_TEST (test_get_scale_zero)
+{
+    double x[2] = { 1e5, 0 };
+    int scale = get_scale((double *) &x, 2);
+    ck_assert_int_eq(scale, 5);
+    x[0] = 0; x[1] = 1e5;
+    scale = get_scale((double *) &x, 2);
+    ck_assert_int_eq(scale, 5);
+}
+END_TEST
+
+START_TEST (test_get_scale_tiny)
+{
+    double x[2] = { 1e5, 1e-100 };
+    int scale = get_scale(x, 2);
+    ck_assert_int_eq(scale, 5);
+    x[0] = 1e-100; x[1] = 1e5;
+    scale = get_scale(x, 2);
+    ck_assert_int_eq(scale, 5);
+}
+END_TEST
+
+START_TEST (test_order)
+{
+    double x[5] = {-1, 8, -4, 7, 1};
+    int result[5];
+    order(x, x, result, 5);
+    ck_assert_int_eq(result[0], 1);
+    ck_assert_int_eq(result[1], 4);
+    ck_assert_int_eq(result[2], 0);
+    ck_assert_int_eq(result[3], 3);
+    ck_assert_int_eq(result[4], 2);
+}
+END_TEST
+
+START_TEST (test_inverse_perm)
+{
+    int p[5] = {1, 4, 0, 3, 2};
+    int pinv[5];
+    inverse_perm(p, pinv, 5);
+    ck_assert_int_eq(pinv[0], 2);
+    ck_assert_int_eq(pinv[1], 0);
+    ck_assert_int_eq(pinv[2], 4);
+    ck_assert_int_eq(pinv[3], 3);
+    ck_assert_int_eq(pinv[4], 1);
+}
+END_TEST
+
+Suite *util_suite(void)
+{
+    Suite *s;
+    TCase *tc_core;
+
+    s = suite_create("util");
+
+    tc_core = tcase_create("Core");
[…]

### 2015-10-02 simulations/expt.py
Updated expt.py for python2, added more integrity checking

@@ -1,4 +1,4 @@
-#!/usr/bin/env python3
+#!/usr/bin/env python
 
 import shlex
 import subprocess
@@ -11,6 +11,7 @@ from pprint import pprint
 import random
 import sqlite3
 import hashlib
+import logging
 
 try:
     subprocess_run = subprocess.run
@@ -82,26 +83,36 @@ def get_dependencies(expt, step_name, parameters, cur):
 
 def needs_update(target, depends, cur):
     if not os.path.exists(target):
+        logging.info("File {} doesn't exist, remaking".format(target))
         return True
+
     for step in depends:
         for file_name in depends[step].split(" "):
             with open(file_name, "rb") as f:
                 checksum = hashlib.md5(f.read()).hexdigest()
-                cur.execute("SELECT checksum FROM md5 WHERE file = ?", (file_name,))
+                cur.execute("SELECT checksum FROM md5 WHERE path = ?", (file_name,))
                 row = cur.fetchone()
-                if row is None or row[0] != checksum:
+                if row is None:
+                    logging.error("Dependency {} doesn't exist".format(file_name))
+                    return False
+                if row[0] != checksum:
+                    logging.info("Checksum on dependency {} has changed, remaking file {}".format(file_name, target))
                     return True
+
+    logging.info("File {} doesn't need remaking".format(target))
     return False
 
 def run_step(expt, step_name, con, cur, nproc):
     step = expt["Steps"][step_name]
 
-    interpreter = shlex.split(step["Interpreter"])
+    #interpreter = shlex.split(step["Interpreter"])
+    interpreter = step["Interpreter"]
     try:
         startup = step["Startup"].rstrip() + "\n"
     except KeyError:
         startup = ""
-    startup = bytes(startup, "UTF-8")
+    if sys.version_info.major == 3:
+        startup = bytes(startup, "UTF-8")
 
     # steps are placed in expt_name/step_name
     expt_name = expt["Name"]
@@ -130,7 +141,10 @@ def run_step(expt, step_name, con, cur, nproc):
                 old_basename &= set([row[0] for row in cur.fetchall()])
 
         if len(old_basename) == 0:
-            basename = hashlib.md5(bytes(str(parameters), "UTF-8")).hexdigest()[:8]
+            if sys.version_info.major == 3:
+                basename = hashlib.md5(bytes(str(parameters), "UTF-8")).hexdigest()[:8]
+            else:
[…]

### 2015-10-02 doc/journal-club/worby2014within.tex
Mostly documentation

\documentclass{beamer}
\usepackage{tikz}
\usepackage{tikz-qtree}
\usepackage{gillius}
\usepackage{mathdots}
\usepackage{abraces}
\usepackage{etex}
\usepackage{array}
\usepackage[backend=biber]{biblatex}

% http://tex.stackexchange.com/questions/12703/how-to-create-fixed-width-table-columns-with-text-raggedright-centered-raggedlef
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

\usetikzlibrary{external}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{decorations.pathreplacing}

\usetheme[everytitleformat=regular]{m}
\setbeamertemplate{navigation symbols}{}

\title{Within-host bacterial diversity hinders accurate reconstruction of
transmission networks from genomic distance data}
\author[Worby, Lipsitch, and Hanage]{Colin J. Worby \and Marc Lipsitch \and William P. Hanage}

\newcommand{\dd}[2]{\frac{\text{d}\,#1}{\text{d}\,#2}}

\begin{document}
\definecolor{red}{RGB}{228,26,28}
\definecolor{blue}{RGB}{55,126,184}
\definecolor{green}{RGB}{77,175,74}
\definecolor{purple}{RGB}{152,78,163}

\maketitle
\begin{frame}{Motivation}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            \centering
            Assumption

            \hfill\\
            \begin{tikzpicture}
                [scale=2, sibling distance=0.8cm]
                \node{\includegraphics[width=1cm]{stick}\includegraphics[width=1cm]{bacteria}}
                    child {
                        node{\includegraphics[width=1cm]{stick}\includegraphics[width=1cm]{bacteria}}
                    }
                    child {
                        node[anchor=west]{\includegraphics[width=1cm]{stick2}\includegraphics[width=1cm]{bacteria}}
                    };
            \end{tikzpicture}
        \end{column}
        \begin{column}{0.5\textwidth}
            \centering
            Reality

            \hfill\\
            \begin{tikzpicture}
                [scale=2, sibling distance=0.8cm]
                \node{\includegraphics[width=1cm]{stick}\includegraphics[width=1cm]{petri}}
                    child {
                        node{\includegraphics[width=1cm]{stick}\includegraphics[width=1cm]{petri}}
                    }
                    child {
                        node[anchor=west]{\includegraphics[width=1cm]{stick2}\includegraphics[width=1cm]{petri}}
                    };
            \end{tikzpicture}
        \end{column}
    \end{columns}
    \pause
    \begin{center}
        How accurately can transmission networks be recovered from genetic data
        when within-host diversity is present?
    \end{center}
\end{frame}

\begin{frame}{Within-host diversity}
    \begin{columns}
        \begin{column}{0.4\textwidth}
            Under neutral evolution, maximal bottlenecking is insufficient to
            explain observed within-host diversity in acute \textit{S. aureus}
            infections.
        \end{column}
        \begin{column}{0.6\textwidth}
            \includegraphics[width=\textwidth]{f1}
        \end{column}
    \end{columns}
\end{frame}

\begin{frame}{Bottlenecks in transmission chains reduce diversity}
    \includegraphics[trim=0 2in 0 0, clip=true, width=\textwidth]{fs3}
\end{frame}

\begin{frame}{High mutation rate mitigates loss of diversity}
    \vspace{-0.5cm}
    \begin{center}
        \includegraphics[scale=0.3]{fs4}
    \end{center}
\end{frame}

\begin{frame}{Simulated transmission chains}
    \begin{itemize}
        \setlength{\itemsep}{10pt}
        \item Transmissions occur every 1000 bacterial generations.
        \item Upon infection, an inoculum of size $N_B$ is transferred from the
            source to the target.
        \item Order of infection is known.
        \item All infected individuals are observed and periodically sampled.
        \item Reconstructed transmission links: probability of transmission is
            inversely proportional to mean pairwise genetic distance.
    \end{itemize}
\end{frame}

\begin{frame}{More sampling = better reconstruction}
    \begin{center}
        \includegraphics[scale=1.5]{f3}
    \end{center}
\end{frame}

\begin{frame}{Simulated epidemics}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            \includegraphics[width=\textwidth, trim=0 1.3in 0 0, clip=true]{f4}
        \end{column}
        \begin{column}{0.5\textwidth}
            \includegraphics[width=\textwidth, trim=0 0 0 1.1in, clip=true]{f4}
        \end{column}
    \end{columns}

    \begin{center}
        Diversity increases with distance from epidemic source (obviously?).
    \end{center}
\end{frame}

\begin{frame}{Importance factors}
    \includegraphics[width=\textwidth]{f5}

    \begin{center}
        Assigning greater importance to genetic similarity increases precision
        at the expense of accuracy.
    \end{center}
\end{frame}

\end{document}

### 2015-10-02 simulations/kernel-mean-degree.py

@@ -1,111 +1,0 @@
-#!/usr/bin/env python3
-
-from pprint import pprint
-import sys
-from makefile import *
-
-data_params = {"mean_degree": [2, 4, 8, 16],
-               "nnode": [5000],
-               "nsimnode": [1000],
-               "ntip": [200],
-               "seed": list(range(100))}
-
-data_targets = {
-    ("mean_degree", "nnode"): [
-        ("{file_basename}.gml",
-         "",
-         """
-         Rscript -e "g <- barabasi.game({nnode}, m={mean_degree}/2, directed=F)" \
-                 -e 'graph_attr(g, "comment") <- "{yaml}"' \
-                 -e "write.graph(SI.net(g), '{file_basename}.gml', format='gml')"
-         """)
-    ],
-    ("mean_degree", "nnode", "ntip", "nsimnode", "seed"): [
-        ("{file_basename}.nwk",
-         "kernel-mean-degree/data/{mean_degree}_{nnode}.gml",
-         """
-         echo "#{yaml}" > $@
-         nettree --sim-nodes {nsimnode} --tree-tips {ntip} --seed {seed} $^ >> $@
-         """)
-    ]
-}
-
-data_makefile, data_files = create_makefile(data_params, data_targets, "kernel-mean-degree/data")
-
-results_params = {"branch_scaling": ["mean"],
-                  "decay_factor": [0.1, 0.2, 0.4, 0.5],
-                  "rbf_variance": [0.0625, 0.125, 1, 2],
-                  "sst_control": [0, 1],
-                  "coal_variance": ["INFINITY", 2],
-                  "ncrossv": [1000]}
-
-results_targets = {
-    ("branch_scaling", "decay_factor", "rbf_variance", "sst_control", "coal_variance"): [
-        ("{file_basename}.mtx.bz2",
-         " ".join(f for f in data_files if f.endswith("nwk")),
-         """
-         echo "%%MatrixMarket matrix array real symmetric" > {file_basename}.mtx
-         echo "%{yaml}" >> {file_basename}.mtx
-         echo $(words $^) $(words $^) >> {file_basename}.mtx
-         for T1 in kernel-mean-degree/data/*.nwk; do \
-             for T2 in kernel-mean-degree/data/*.nwk; do \
-                 if [[ "$$T1" < "$$T2" || "$$T1" == "$$T2" ]]; then \
[…]

### 2015-10-17 programs/netsim/src/netabc.c
Mostly implemented SMC

@@ -1,0 +1,201 @@
+#include <stdlib.h>
+#include <stdio.h>
+#include <getopt.h>
+#include <math.h>
+#include <float.h>
+#include <string.h>
+#include <igraph/igraph.h>
+#include <gsl/gsl_randist.h>
+#include <gsl/gsl_rng.h>
+#include <gsl/gsl_statistics_double.h>
+
+#include "tree.h"
+#include "treestats.h"
+#include "simulate.h"
+#include "smc.h"
+#include "util.h"
+
+#define NNODE 5000
+#define NSIMNODE 1000
+#define NTIP 200
+#define MEAN_DEGREE 8
+#define PA_POWER_MIN 0.1
+#define PA_POWER_MAX 1.0
+#define DECAY_FACTOR 0.4
+#define RBF_VARIANCE 2
+
+struct netabc_options {
+    FILE *tree_file;
+};
+
+struct option long_options[] =
+{
+    {"help", no_argument, 0, 'h'},
+    {0, 0, 0, 0}
+};
+
+void usage(void)
+{
+    fprintf(stderr, "Usage: netabc [options] [tree]\n\n");
+    fprintf(stderr, "Options:\n");
+    fprintf(stderr, "  -h, --help                display this message\n");
+}
+
+struct netabc_options get_options(int argc, char **argv)
+{
+    int i, c = 0;
+    struct netabc_options opts = {
+        .tree_file = stdin
+    };
+
+    while (c != -1)
+    {
+        c = getopt_long(argc, argv, "hl:g:s:c:ndb:", long_options, &i);
+        if (c == -1)
+            break;
+
+        switch (c)
+        {
+            case 0:
+                break;
+            case 'h':
+                usage();
+                exit(EXIT_SUCCESS);
+            case '?':
+                break;
+            default:
+                usage();
+                exit(EXIT_FAILURE);
+        }
+    }
+
+    // TODO: safety
+    if (optind < argc) {
+        opts.tree_file = fopen(argv[optind++], "r");
+    }
+
+    return opts;
+}
+
+void ba_sample_from_prior(gsl_rng *rng, double *theta)
+{
+    *theta = gsl_ran_flat(rng, 0.1, 1.0);
+}
+
+double ba_prior_density(const double *theta)
+{
+    if (*theta < PA_POWER_MIN || *theta > PA_POWER_MAX) {
[…]

### 2015-10-17 programs/netsim/src/nettree.c

@@ -24,6 +24,7 @@ struct nettree_options {
 
 struct option long_options[] =
 {
+    {"help", no_argument, 0, 'h'},
     {"sim-time", required_argument, 0, 's'},
     {"sim-nodes", required_argument, 0, 'n'},
     {"tree-height", required_argument, 0, 'r'},
@@ -32,6 +33,19 @@ struct option long_options[] =
     {"seed", required_argument, 0, 'd'},
     {0, 0, 0, 0}
 };
+
+void usage(void)
+{
+    fprintf(stderr, "Usage: nettree [options] [graph]\n\n");
+    fprintf(stderr, "Options:\n");
+    fprintf(stderr, "  -h, --help                display this message\n");
+    fprintf(stderr, "  -s, --sim-time            amount of epidemic time to simulate (default infinity)\n");
+    fprintf(stderr, "  -n, --sim-nodes           stop simulation after this many nodes are infected (default infinity)\n");
+    fprintf(stderr, "  -r, --tree-height         cut the tree at this height (default infinity)\n");
+    fprintf(stderr, "  -t, --tree-tips           sample this many tips from the tree (default infinity)\n");
+    fprintf(stderr, "  -e, --extant-only         sample only extant tips (default false)\n");
+    fprintf(stderr, "  -d, --seed                random seed\n");
+}
 
 struct nettree_options get_options(int argc, char **argv)
 {
@@ -49,10 +63,15 @@ struct nettree_options get_options(int argc, char **argv)
 
     while (c != -1)
     {
-        c = getopt_long(argc, argv, "s:n:r:t:ed:", long_options, &i);
+        c = getopt_long(argc, argv, "hs:n:r:t:ed:", long_options, &i);
+        if (c == -1)
+            break;
 
         switch (c)
         {
+            case 'h':
+                usage();
+                exit(EXIT_SUCCESS);
             case 's':
                 opts.sim_time = atof(optarg);
                 break;
@@ -71,8 +90,12 @@ struct nettree_options get_options(int argc, char **argv)
             case 'd':
                 opts.seed = atoi(optarg);
                 break;
-            default:
+            case '?':
+            case 0:
                 break;
+            default:
+                usage();
+                exit(EXIT_FAILURE);
         }
     }
 
@@ -115,7 +138,7 @@ int main (int argc, char **argv)
[…]

### 2015-10-17 programs/netsim/src/simulate.c

@@ -10,14 +10,13 @@
 
 void add_vertex_id(const igraph_t *net, igraph_strvector_t *v, int node, int numeric_ids);
 
-igraph_t *simulate_phylogeny(igraph_t *net, gsl_rng *rng, double stop_time, 
-        int stop_nodes, int numeric_ids)
+void simulate_phylogeny(igraph_t *tree, igraph_t *net, gsl_rng *rng,
+        double stop_time, int stop_nodes, int numeric_ids)
 {
     long i;
     int inode, snode, e, v, head, tail, nnode_tree = 0;
     double t, trans_rate = 0., remove_rate = 0., time = 0.;
     char buf[128];
-    igraph_t *tree = malloc(sizeof(igraph_t));
     igraph_vector_t discordant, incident, edges, branch_lengths, tip_map;
     igraph_vector_int_t infected, removed; 
     igraph_strvector_t node_map;
@@ -174,8 +173,6 @@ igraph_t *simulate_phylogeny(igraph_t *net, gsl_rng *rng, double stop_time,
     igraph_vector_destroy(&incident);
     igraph_vector_destroy(&edges);
     igraph_vector_destroy(&branch_lengths);
-
-    return tree;
 }
 
 /* Private */

### 2015-10-17 programs/netsim/src/simulate.h

@@ -40,9 +40,10 @@
  * \param[in] stop_time maximum amount of time to run the simulation for
  * \param[in] stop_nodes maximum number of nodes to infect
  * \param[in] numeric_ids 1 if the "id" attribute of the network is numeric
+ * \param[in] tree an uninitialized igraph_t object
  * \return a phylogeny simulated over the contact network
  */
-igraph_t *simulate_phylogeny(igraph_t *net, gsl_rng *rng, double stop_time, 
-        int stop_nodes, int numeric_ids);
+void simulate_phylogeny(igraph_t *tree, igraph_t *net, gsl_rng *rng, double
+        stop_time, int stop_nodes, int numeric_ids);
 
 #endif

### 2015-10-17 programs/netsim/src/smc.c

@@ -1,111 +1,334 @@
 #include <stdlib.h>
 #include <string.h>
 #include <math.h>
+#include <float.h>
+#include <gsl/gsl_roots.h>
 #include "smc.h"
+#include "util.h"
 
-#define NT 200
-#define M 15
-#define POPSIZE 100
-#define END_EPSILON 0.01
-#define alpha 0.9
+#define RESIZE_AMOUNT 100
+#define NTHREAD 1
 
-double ess(const double *W);
-double calc_new_epsilon(int nparam, double *W, void **X, metric d, 
-        double old_epsilon, const void *data);
+struct objfun_params {
+    double *X;
+    double *W;
+    double *new_W;
+    double epsilon;
+    const smc_config *config;
+};
+
+struct perturb_workspace {
+    gsl_rng *rng;
+    const smc_config *config;
+    const smc_functions *functions;
+    double *W;
+    double *theta;
+    double *X;
+    double epsilon;
+    const void *data;
+    char *fdbk;
+    double *new_theta;
+    double *new_X;
+    char *z;
+    int nparticle;
+};
+
+void resample(const smc_config config, double *W, double *theta);
+double next_epsilon(struct objfun_params *objpar);
+double epsilon_objfun(double epsilon, void *params);
+double ess(const double *W, int n);
+double perturb(struct perturb_workspace *ws);
 
 // see Del Moral et al. 2012: An adaptive sequential Monte Carlo method for
 // approximate Bayesian computation
-void abc_smc(int nparam, real_rv *prior_rvs, sampler data_sampler, metric d,
-             const void *data)
+smc_result *abc_smc(const smc_config config, const smc_functions functions,
+                    int seed, const void *data)
 {
     int i, j;
-    double *theta = malloc(nparam * POPSIZE * sizeof(double));
-    void **X = malloc(M * POPSIZE * sizeof(void*));
-    double *W = malloc(POPSIZE * sizeof(double));
-    double prev_ess = POPSIZE, prev_epsilon, cur_epsilon = INFINITY;
-    double num, denom;
+    double *X = malloc(config.nsample * config.nparticle * sizeof(double));
+    double *W = malloc(config.nparticle * sizeof(double));
+    char *z = malloc(config.dataset_size);
+    double epsilon = 1;
+    size_t new_size;
+    gsl_rng *rng = set_seed(seed);
 
[…]

### 2015-10-17 programs/netsim/src/smc.h

@@ -1,9 +1,57 @@
 #ifndef SMC_H
 #define SMC_H
 
-typedef double (*real_rv) (void);
-typedef void* (*sampler) (double*);
-typedef double (*pdf) (double*);
-typedef double (*metric) (const void *, const void *);
+#include <gsl/gsl_rng.h>
+
+typedef struct {
+    int nparam; /**< number of parameters in the model */
+    int nparticle; /**< number of particles used to approximate the posterior */
+    int nsample; /**< number of sampled data points per particle */
+    int ess_tolerance; /**< ESS below this value triggers resampling */
+    double final_epsilon; /**< tolerance level to end at */
+    double quality; /**< between 0 and 1, where 0 is fast and coarse, 1 is slow and accurate */
+    double step_tolerance; /**< tolerance for bisection solution of next epsilon */
+    size_t dataset_size; /**< size of data objects */
+    size_t feedback_size; /**< size of feedback objects */
+} smc_config;
+
+typedef struct {
+    void   (*sample_from_prior) (gsl_rng *rng, double *);
+    double (*prior_density)     (const double *);
+    void   (*propose)           (gsl_rng *rng, double *, const void *);
+    double (*proposal_density)  (const double *, const double *, const void *);
+    void   (*sample_dataset)    (gsl_rng *rng, const double *, void *);
+    double (*distance)          (const void *, const void *);
+    void   (*feedback)          (const double *, int, void *);
+} smc_functions;
+
+typedef struct {
+    int niter;
+    double *epsilon;
+    double *acceptance_rate;
+    double *theta;
+} smc_result;
+
+/** Perform ABC-SMC.
+ *
+ * This implements the adaptive tolerance algorithm from DelMoral et al. 2012.
+ * The object returned must be passed to smc_result_free() when you are done
+ * with it.
+ *
+ * \param[in] config control parameters for the algorithm
+ * \param[in] functions functions to use in the algorithm
+ * \param[in] seed random seed (or NULL)
+ * \param[in] data true data
+ * \return an smc_result object with the posterior distribution (theta) and
+ * other information about the run
+ */
+smc_result *abc_smc(const smc_config config, const smc_functions functions,
+                    int seed, const void *data);
+
+/** Free an smc_result object.
+ *
+ * \param[in] r object to free
+ */
+void smc_result_free(smc_result *r);
 
 #endif

### 2015-10-17 programs/netsim/src/treekernel.c

@@ -46,7 +46,6 @@ void usage(void)
     fprintf(stderr, "  -b, --scale-branches      type of branch scaling to apply (mean/median/max/none, default none)\n");
 }
 
-
 struct treekernel_options get_options(int argc, char **argv)
 {
     int i, c = 0;

### 2015-10-17 programs/netsim/tests/check_smc.c

@@ -1,0 +1,179 @@
+#include <check.h>
+#include <math.h>
+#include <string.h>
+#include <float.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <gsl/gsl_randist.h>
+#include <gsl/gsl_rng.h>
+#include <gsl/gsl_cdf.h>
+#include <gsl/gsl_statistics_double.h>
+#include "../src/smc.h"
+#include "../src/util.h"
+
+Suite *smc_suite(void);
+
+void toy_sample_from_prior (gsl_rng *rng, double *theta)
+{
+    *theta = gsl_ran_flat(rng, -10, 10);
+}
+
+double toy_prior_density(const double *theta)
+{
+    return 1./20.;
+}
+
+void toy_propose(gsl_rng *rng, double *theta, const void *params)
+{
+    double var = *((double *) params);
+    *theta += gsl_ran_gaussian(rng, sqrt(2*var));
+}
+
+double toy_proposal_density(const double *from, const double *to, const void *params)
+{
+    double var = * ((double *) params);
+    return gsl_ran_gaussian_pdf(*to - *from, sqrt(2*var));
+}
+
+void toy_sample_dataset(gsl_rng *rng, const double *theta, void *X)
+{
+    double x;
+    if (rand() < RAND_MAX / 2) {
+        x = (*theta + gsl_ran_gaussian(rng, 1)); 
+    }
+    else {
+        x = (*theta + gsl_ran_gaussian(rng, 0.1));
+    }
+    memcpy(X, &x, sizeof(double));
+}
+
+double toy_distance(const void *x, const void *y)
+{
+    return fabs(*(double *) x - *(double *) y);
+}
+
+void toy_feedback(const double *theta, int nparticle, void *params)
+{
+    double var = gsl_stats_variance(theta, 1, nparticle);
+    memcpy(params, &var, sizeof(double));
+}
+
+smc_functions toy_functions = {
+    .sample_from_prior = toy_sample_from_prior,
+    .prior_density = toy_prior_density,
+    .propose = toy_propose,
+    .proposal_density = toy_proposal_density,
+    .sample_dataset = toy_sample_dataset,
+    .distance = toy_distance,
+    .feedback = toy_feedback
+};
+
+void plot(const double *x, const double *y, int nx, const char *pdf, 
[…]

### 2015-10-17 doc/stats-seminar/presentation.tex
Most of presentation for stats seminar

\documentclass{beamer}
\usepackage{tikz}
\usepackage{tikz-qtree}
\usepackage{gillius}
\usepackage{mathdots}
\usepackage{abraces}
\usepackage{etex}
\usepackage{nicefrac}

\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{decorations.pathreplacing}
\usetikzlibrary{decorations.pathmorphing}

\usetheme[everytitleformat=regular]{m}
\setbeamertemplate{navigation symbols}{}

\title{Phylogenetic clustering with a Markov-modulated Poisson process}
\author[RM \& AFYP]{Rosemary McCloskey and Art FY Poon}
\institute[UBC \& BCCfE]{BC Centre for Excellence in HIV/AIDS \\ University of British Columbia}
\date{October 12, 2015}

\graphicspath{{figures/}}

\newcommand{\dd}[2]{\frac{\text{d}\,#1}{\text{d}\,#2}}

\setlength{\parindent}{0pt}

\begin{document}
\definecolor{red}{RGB}{228,26,28}
\definecolor{blue}{RGB}{55,126,184}
\definecolor{green}{RGB}{77,175,74}
\definecolor{purple}{RGB}{152,78,163}

\maketitle
\tableofcontents

\section{Introduction to phylodynamics}

\begin{frame}{Phylogenetic trees}
    \begin{tikzpicture}
        [sibling distance=2.5cm, level distance=2cm]
        \node (root) [circle, draw, very thick, minimum width=0.7cm] { \only<3->{?} };
        \node (fish) [below left=3cm and 1cm of root] {\includegraphics[width=1.8cm]{fish}};
        \node (cat) [below=3cm of root] {\includegraphics[width=1.8cm]{cat}};
        \node (dog) [below right=3cm and 1cm of root] {\includegraphics[width=1.8cm]{dog}};
        \node (cd) [below right=1.5cm and 0.5cm of root, circle, draw, very thick, minimum width=0.7cm] {\only<3->{?}};
        \draw [very thick] (root) -- (fish);
        \draw [very thick] (root) -- (cd) -- (cat);
        \draw [very thick] (cd) -- (dog);

        \uncover<4->{
        \draw [decorate, decoration={brace, amplitude=10pt, mirror}] (dog.north east) -- ++(0cm, 3cm);
        \node [above right=1.5cm and 0.5cm of dog.north east, text width=2.5cm, anchor=west] {evolutionary  relationships};
        }

        \uncover<2->{
        \node (known) [below=1cm of cat] {known species};
        \draw [->, >=stealth] (known) -- (fish);
        \draw [->, >=stealth] (known) -- (cat);
        \draw [->, >=stealth] (known) -- (dog);
        }

        \uncover<3->{
        \node (anc) [above=2cm of fish.west, anchor=south, text width=2cm] {common ancestors};
        \draw [->, >=stealth, shorten >=2mm] (anc) -- (root);
        \draw [->, >=stealth, shorten >=2mm] (anc) -- (cd);
        }

        \uncover<5->{
        \draw [decorate, decoration={brace, amplitude=10pt, raise=5pt}] (root.east) -- (cd.east);
        \node [below right=-0.5 and 1.2cm of root, text width=3cm] {genetic \\ distance \\ or time};
    }
    \end{tikzpicture}
\end{frame}

\begin{frame}{Phylogenetic splits correspond to speciation}
    \begin{tikzpicture}
        \node (root) [circle, draw, minimum width=0.7cm, fill=purple] { };
        \node (parent) [above right=of root] { };
        \node (lc) [below left=2.5cm and 1cm of root, circle, draw, fill=blue, minimum width=0.7cm] { };
        \node (rc) [below right=2.5cm and 1cm of root, circle, draw, fill=red, minimum width=0.7cm] { };
        \draw (root) -- (parent);
        \draw (root) -- (lc);
        \draw (root) -- (rc);
        \coordinate [left=4cm of parent] (top);
        \coordinate [below=5cm of top] (bot);
        \draw [->, >=stealth, very thick] (top) -- node [auto, swap] {time} (bot);

        \uncover<2->{
        \node (pop1) [right=3cm of root, circle, fill=purple, minimum width=1cm, draw] { };
        \coordinate [above right=of pop1.center] (rtop);
        \coordinate [below=2cm of rtop] (rbot);
        \draw [decorate, decoration={snake}, blue!50, very thick] (rtop) -- (rbot);
        }

        \uncover<3->{
        \node (pop2) [right=of pop1, circle, fill=purple, minimum width=0.5cm, draw] { };
        \draw (pop1.center) [->, >=stealth, very thick] (pop1.center) -- (pop2.center);
        }

        \uncover<4->{
        \node (pop1a) [below=2cm of pop1, circle, fill=blue, minimum width=1cm, draw] { };
        \node (pop2a) [right=of pop1a, circle, fill=red, minimum width=0.5cm, draw] { };
        \coordinate [above right=of pop1a.center] (rtop);
        \coordinate [below=2cm of rtop] (rbot);
        \draw [decorate, decoration={snake}, blue!50, very thick] (rtop) -- (rbot);
        }
    \end{tikzpicture}
\end{frame}

\begin{frame}{Phylogenetics and viral disease}
    \vspace{-1cm}
    \begin{center}
    \begin{tikzpicture}
        \uncover<1-2>{
        \node (a) {\includegraphics[width=1cm]{hiv}};
        \node at (a.north west) {\includegraphics[width=1cm]{person2}};
        }\uncover<2>{
        \node [right=1.5cm of a] (b) {\includegraphics[width=1cm]{hiv}};
        \node at (b.north west) {\includegraphics[width=1cm]{person1}};
        }

        \uncover<3->{
        \node [right=1.5cm of a, opacity=0.4] (b) {\includegraphics[width=1cm]{hiv}};
        \node at (b.north west) [opacity=0.4] {\includegraphics[width=1cm]{person1}};
        \node (a) [opacity=0.4] {\includegraphics[width=1cm]{hiv}};
        \node at (a.north west) [opacity=0.4] {\includegraphics[width=1cm]{person2}};
        }

        \uncover<2-3>{
        \draw[->, >=stealth, thick] (a) -- (b);
        }\uncover<3->{
        \node [below=2cm of a] (c) {\includegraphics[width=1cm]{hiv}};
        \node at (c.north west) {\includegraphics[width=1cm]{person2}};
        \node [below=2cm of b] (d) {\includegraphics[width=1cm]{hiv}};
        \node at (d.north west) {\includegraphics[width=1cm]{person1}};
        }\uncover<3->{
        \coordinate [left= of a] (timestart);
        \coordinate [left= of c] (timeend);
        \draw[->, >=stealth, thick] (timestart) -- node[auto, swap] {time} (timeend);
        }\uncover<4->{
        \coordinate [right=1.4 of a.center] (root);
        \draw [very thick] (a.center) +(1.4, 0) to +(1.4, 1);
        \draw [very thick] (c.center) -- (root);
        \draw [very thick] (d.center) -- (root);
        \draw [very thick] (root) to +(0, 1);
        }

        \uncover<2->{
        \node (note1) [right= of b, text width=4.5cm] {transmission to new host \\
        $\Rightarrow$ isolated sub-population};
        }
        \uncover<3->{
            \node (note2) [below=0.8cm of note1, text width=4.5 cm] {within-host evolution};
        }
        \uncover<4->{
            \node [below=of note2, text width=4.5 cm] {virus phylogeny \\ $\approx$ transmission tree};
        }

    \end{tikzpicture}

    \uncover<5->{\textbf{phylodynamics}}
    \end{center}
\end{frame}

\section{Motivation: more informed phylogenetic clustering}

\begin{frame}{Phylogenetic clustering}
    \begin{tikzpicture}
        [node distance=0.5, every node/.style={inner sep=-2pt}]
        \node (a) {\includegraphics[width=1cm]{person1}};
        \node [below left=0.5 and 0 of a] (b) {\includegraphics[width=1cm]{person2}};
        \node [right=of a] (c) {\includegraphics[width=1cm]{person3}};
        \node [below=of c] (d) {\includegraphics[width=1cm]{person4}};

        \node (e) [below left=2 and 0.5 of d] {\includegraphics[width=1cm]{person5}};
        \node (f) [right=of e] {\includegraphics[width=1cm]{person6}};

        \node (g) [above right=0.5 and 2 of f] {\includegraphics[width=1cm]{person7}};
        \node (h) [below=of g] {\includegraphics[width=1cm]{person8}};
        \node (i) [above right=0.5 and 0.5 of h] {\includegraphics[width=1cm]{person9}};

        \node (j) [above right=1 and 0.5 of g] {\includegraphics[width=1cm]{person10}};
        \node (k) [right=of j] {\includegraphics[width=1cm]{person11}};
        \node (l) [above right=of k] {\includegraphics[width=1cm]{person12}};

        \draw (a) -- (b) -- (c) -- (a) -- (d) -- (e) -- (f) -- (g) --
        (h) -- (i) -- (g) -- (j) -- (k) -- (l) -- (j) -- (d);
        \draw (b) -- (d);

        \uncover<2->{
        \node (groups) [inner sep=5pt, above right=0.5cm and 1cm of a, text width=4.5cm, align=center]
        {groups of epidemiologically \\ related individuals};
        \draw [->, >=stealth, thick] (groups) -- (c);
        \draw [->, >=stealth, thick, shorten >=0.5cm] (groups) -- (k.north);
        }
    \end{tikzpicture}
\end{frame}

\newcommand{\p}[1]{\includegraphics[width=0.7cm, angle=-90]{person#1} 
\includegraphics[width=0.7cm, angle=-90]{hiv}}
\begin{frame}{Clustering by genetic distance}
    \hspace{-0.5cm}
    % http://tex.stackexchange.com/questions/5078/tikz-skipping-levels-when-drawing-trees
    \begin{tikzpicture}[rotate=45, scale=0.95]
    \begin{scope}[rotate=45]
    \node[coordinate] (start) at (0,0) {};
    \xdef\pl{start}
    \foreach \n/\l in {
      \p{11}/p11,
      \p{10}/p10,
      \p{13}/p13,
      \p{7}/p7,
      \p{8}/p8,
      \p{9}/p9,
      \p{6}/p6,
      \p{5}/p5,
      \p{1}/p1,
      \p{2}/p2,
      \p{3}/p3,
      \p{4}/p4%
    } {
      \node[transform shape,below right] (\l) at (\pl.south west) {\n};
      \xdef\pl{\l}
    }
    \end{scope}

    \foreach \a/\b in {
      p4/p11,
      p10/p11,
      p13/p11,
      p9/p11,
      p9/p7,
      p9/p8,
      p5/p11,
      p5/p6,
      p2/p1,
      p4/p1,
      p4/p3%
    } {
      \draw (\a.east) |- (\b.east);
    }
    \uncover<2->{
        \draw [red, very thick] (p5.east) |- (p6.east);
        \node (gd) [above right=1cm and 0.5cm of p5, inner sep=0pt, text width=2cm] {genetic \\ distance};
        \draw [->, >=stealth, thick, shorten >=6pt, shorten <=6pt] (gd) -- (p5.east);
    }
    \uncover<3->{
        \draw [red, very thick] (p10.east) |- (p11.east);
        \draw [red, very thick] (p13.east) |- (p11.east);
        \draw [red, very thick] (p9.east) |- (p7.east);
        \draw [red, very thick] (p9.east) |- (p8.east);
        \draw [red, very thick] (p2.east) |- (p1.east);
        \draw [red, very thick] (p4.east) |- (p1.east);
        \draw [red, very thick] (p4.east) |- (p3.east);
        \draw (5, 5) to node [auto, swap] {distance threshold} +(3, -3);
    }
    \end{tikzpicture}
\end{frame}

\begin{frame}{Clustering by genetic distance}
    \begin{tikzpicture}
        [node distance=0.5, every node/.style={inner sep=-2pt}]
        \node (a) {\includegraphics[width=1cm]{person1}};
        \node [below left=0.5 and 0 of a] (b) {\includegraphics[width=1cm]{person2}};
        \node [right=of a] (c) {\includegraphics[width=1cm]{person3}};
        \node [below=of c] (d) {\includegraphics[width=1cm]{person4}};

        \node (e) [below left=2 and 0.5 of d] {\includegraphics[width=1cm]{person5}};
        \node (f) [right=of e] {\includegraphics[width=1cm]{person6}};

        \node (g) [above right=0.5 and 2 of f] {\includegraphics[width=1cm]{person7}};
        \node (h) [below=of g] {\includegraphics[width=1cm]{person8}};
        \node (i) [above right=0.5 and 0.5 of h] {\includegraphics[width=1cm]{person9}};

        \node (j) [above right=1 and 0.5 of g] {\includegraphics[width=1cm]{person10}};
        \node (k) [right=of j] {\includegraphics[width=1cm]{person11}};
        \node (l) [above right=of k] {\includegraphics[width=1cm]{person12}};

        \uncover<1-2> {
        \draw (a) -- (b) -- (c) -- (a) -- (d) -- (e) -- (f) -- (g) --
        (h) -- (i) -- (g) -- (j) -- (k) -- (l) -- (j) -- (d);
        \draw (b) -- (d);
        }

        \uncover<2->{
            \draw [red, very thick] (b) -- (a) -- (c);
            \draw [red, very thick] (a) -- (d);
            \draw [red, very thick] (e) -- (f); 
            \draw [red, very thick] (g) -- (i);
            \draw [red, very thick] (j) -- (l);
            \draw [red, very thick] (g) -- (h);
            \draw [red, very thick] (j) -- (k);
        }

        \uncover<4->{
            \draw (a) + (0.4, -0.8) circle (1.8);
            \draw (e) + (0.7, 0) ellipse(1.5 and 0.8);
            \draw [rotate=45] (g) + (-0.1, -0.6) ellipse(1.6 and 1.2);
            \draw [rotate=25] (k) + (0, 0.5) ellipse(2.2 and 1);
        }
    \end{tikzpicture}
\end{frame}

\begin{frame}{Popular problems}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            \includegraphics[width=\textwidth, trim=0 11in 0 0, clip=true]{hughes2009molecular-f1}
        \end{column}
        \begin{column}{0.5\textwidth}
            \includegraphics[width=\textwidth]{lewis2008episodic-f2}
        \end{column}
    \end{columns}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            \tiny
            Hughes, Gareth J., et al. ``Molecular phylodynamics of the heterosexual HIV epidemic in the United Kingdom.'' PLoS Pathog 5.9 (2009): e1000590.
        \end{column}
        \begin{column}{0.5\textwidth}
            \tiny
            Lewis, Fraser, et al. ``Episodic sexual transmission of HIV revealed by molecular phylodynamics.'' PLoS Med 5.3 (2008): e50.
        \end{column}
    \end{columns}
\end{frame}

\begin{frame}{Genetic distance can be misleading}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            \centering
            \includegraphics[height=0.8\textwidth, angle=-90]{fasttree}

            \uncover<2->{
            \begin{tikzpicture}
                [every node/.style={circle, draw, fill, inner sep=6pt}]
                \node (a) { };
                \node (b) [below=of a] { };
                \node (c) [below right=of a] { };
                \node (d) [below left=of a] { };
                \node (e) [left=of a] { };
                \draw (a) -- (b);
                \draw (a) -- (c);
                \draw (a) -- (d);
                \draw (a) -- (e);
            \end{tikzpicture}
            }
        \end{column}
        \begin{column}{0.5\textwidth}
            \centering
            \includegraphics[height=0.8\textwidth, angle=-90]{slowtree}

            \uncover<2->{
            \begin{tikzpicture}
                [every node/.style={circle, draw, fill, inner sep=6pt}]
                \node (a) { };
                \node (b) [below=of a] { };
                \node (c) [below right=of a] { };
                \node (d) [below left=of a] { };
                \node (e) [left=of a] { };
                \draw (a) -- (e);
                \draw (e) -- (d);
                \draw (d) -- (b);
                \draw (b) -- (c);
            \end{tikzpicture}
            }
        \end{column}
    \end{columns}
\end{frame}

\begin{frame}{Clustering by branching rates}
    \only<1>{\includegraphics[height=\textwidth,angle=-90]{pcbr_example}}
    \only<2->{\includegraphics[height=\textwidth,angle=-90]{pcbr_example_color}}

    \hspace{1.2cm}
    \uncover<3->{
    \begin{tikzpicture}
        [node distance=0.5,
         every node/.style={circle, draw=red, fill=red, inner sep=4pt}]
        \node (a) { };
        \node [below=of a] (b) { };
        \node [below=of b] (c) { };
        \node [right=of b] (d) { };
        \node [left=of b] (e) { };
        \draw [thick] (a) -- (b) -- (d) -- (a);
        \draw [thick] (a) -- (b) -- (e) -- (a);
        \draw [thick] (c) -- (b) -- (d) -- (c);
        \draw [thick] (c) -- (b) -- (e) -- (c);
        \node [right=of a, draw=none, fill=none, text=red] {fast};
    \end{tikzpicture}
    }
    \hspace{0.2cm}
    \uncover<4->{
    \begin{tikzpicture}
        [node distance=0.5,
         every node/.style={circle, draw=blue, fill=blue, inner sep=4pt}]
        \node (a) { };
        \node [below=of a] (b) { };
        \node [below=of b] (c) { };
        \node [right=of b] (d) { };
        \node [left=of b] (e) { };
        \draw [thick] (a) -- (b) -- (d) -- (c) -- (e);
        \node [right=of a, draw=none, fill=none, text=blue] {slow};
    \end{tikzpicture}
    }
\end{frame}

\section{A Markov-modulated Poisson process on phylogenies}

\begin{frame}{Modelling branching rates}
    \begin{tikzpicture}
        [every node/.style={inner sep=2pt}]
        \coordinate (a);
        \coordinate [below right=3 of a] (b);
        \draw [thick] (a) -- ++ (-0.5, 0.5) node [anchor=south east] {$\ddots$};
        \draw [thick] (a) -- ++ (-0.5, -0.4) node [inner ysep=-4pt, inner xsep=2pt, anchor=north east] {$\iddots$};
        \draw [thick] (b) -- ++ (-0.5, -0.4) node [inner ysep=-4pt, inner xsep=2pt, anchor=north east] {$\iddots$};
        \draw [thick] (b) -- ++ (0.5, -0.5) node [anchor=north west, inner ysep=-4pt] {$\ddots$};
        \draw [thick] (a) -- (b);
        \uncover<2->{
            \draw [blue, very thick] (a) -- node [auto, swap, text=blue] {slow} (b);
            \node (note1) [above right=0.5 and 6 of a, text width=4cm] {branching rates \\ along phylogeny};
        }
        \uncover<3>{
            \draw [blue] (a) -- node [auto, text=black] {
                $T \sim $ Exponential(\color{blue}$\lambda_\text{slow}$\color{black})
            } (b);
        }
        \uncover<4->{
            \draw [blue] (a) -- node [auto, text=black] {
                $T \sim $ Exponential(?)
            } (b);
        }
        \uncover<3->{
            \node (note2) [below=0.5 of note1, text width=4cm] {branching is \\ a Poisson process};
        }
        \uncover<4->{
            \node [below=0.5 of note2, text width=4cm] {rate switching is \\ a Poisson process};
            \draw [very thick, red] (b) -- node [auto] {fast} ++ (-0.5, 0.5);
        }
    \end{tikzpicture}

    \begin{center}
    \begin{tikzpicture}
    \uncover<5->{
        \node [circle, draw, text=blue, inner sep=2pt] (slow) {$\lambda_{\text{slow}}$};
        \node [right=of slow, circle, draw, text=red, inner sep=2pt] (fast) {$\lambda_{\text{fast}}$};
        \draw [->, >=stealth, thick] (slow) to [bend right] node[auto, swap] {$q_{sf}$} (fast);
        \draw [->, >=stealth, thick] (fast) to [bend right] node[auto, swap] {$q_{fs}$} (slow);
    }
    \uncover<6->{
        \node [right=of fast] {Markov-modulated Poisson process (MMPP)};
    }
    \end{tikzpicture}
    \end{center}
\end{frame}

\begin{frame}{Markov-modulated Poisson processes}
    \includegraphics[width=\textwidth]{muscariello2004mmpp}

    \includegraphics[width=\textwidth]{ching1997markov}

    \includegraphics[width=\textwidth]{lu2012markov}
\end{frame}

\newcommand{\ls}{\lambda_{\text{slow}}}
\newcommand{\lf}{\lambda_{\text{fast}}}

\begin{frame}[label=steps]{Using the MMPP for clustering}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            \begin{center}
                \textbf{Step 1}
            \end{center}
        \end{column}
        \begin{column}{0.5\textwidth}
            \only<2>{
            \begin{center}
                \textbf{Step 2}
            \end{center}
            }
        \end{column}
    \end{columns}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            \begin{center}
                \begin{tikzpicture}
                    \node [circle, draw, text=blue, inner sep=2pt] (slow) {$\lambda_{\text{slow}}$};
                    \node [right=of slow, circle, draw, text=red, inner sep=2pt] (fast) {$\lambda_{\text{fast}}$};
                    \draw [->, >=stealth, thick] (slow) to [bend right] node[auto, swap] {$q_{sf}$} (fast);
                    \draw [->, >=stealth, thick] (fast) to [bend right] node[auto, swap] {$q_{fs}$} (slow);
                \end{tikzpicture}
            \end{center}
        \end{column}
        \begin{column}{0.5\textwidth}
            \only<2->{\includegraphics[height=\textwidth,angle=-90]{pcbr_example_color}}
        \end{column}
    \end{columns}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            Given a phylogeny $T$ with branch lengths, find parameters $\theta$
            to optimize
            \[
                \mathcal{L}(\theta \mid T) = \Pr(T \mid \theta).
            \]
        \end{column}
        \begin{column}{0.5\textwidth}
            \only<2->{
            Given $T$ and $\theta$, find an assignment $\nu$ of states to nodes
            to optimize
            \[
                \Pr(T, \nu \mid \theta).
            \]
        }
        \end{column}
    \end{columns}
\end{frame}

\begin{frame}{MMPP likelihood}
    Recall the likelihood of a realization $D$ of an ordinary Poisson
    process with rate $\lambda$.

    \begin{center}
    \begin{tikzpicture}
        \node at (-0.5, 0) {$D = $};

        \draw (0, 0) -- (4.5, 0);
        \draw (1, -0.1) -- (1, 0.1);
        \draw (3, -0.1) -- (3, 0.1);
        \draw (4.5, -0.1) -- (4.5, 0.1);

        \node at (0.5, -0.3) {$t_1$};
        \node at (2, -0.3) {$t_2$};
        \node at (3.75, -0.3) {$t_3$};
    \end{tikzpicture}
    \end{center}
    \[
        \mathcal{L}(\lambda \mid D) = \color{blue}\lambda\color{black} \times 
        \color{red}\exp(-\lambda t_1)\color{black} \times
        \color{green}\lambda \exp(-\lambda t_2) \lambda \exp(-\lambda t_3)\color{black},
    \]
    \pause
    where
    \begin{align*}
        \color{blue}\lambda &= \text{ arrival rate at time } t_1 \\
        \color{red}\exp(-\lambda t_1) &= \text{ probability of no arrivals by time } t_1 \\
        \color{green}\lambda \exp(-\lambda t_2) \lambda \exp(-\lambda t_3) &= \text{ probability of rest of } D.
    \end{align*}
\end{frame}

\begin{frame}{MMPP likelihood}
    \begin{center}\begin{tikzpicture}
        \node [circle, draw, inner sep=2pt] (slow) {$\lambda_0$};
        \node [right=of slow, circle, draw, inner sep=2pt] (fast) {$\lambda_1$};
        \draw [->, >=stealth, thick] (slow) to [bend right] node[auto, swap] {$q_{01}$} (fast);
        \draw [->, >=stealth, thick] (fast) to [bend right] node[auto, swap] {$q_{10}$} (slow);
    \end{tikzpicture}\end{center}
    \vspace{-1cm}
    Let 
    \begin{itemize}
        \item $\theta = (\lambda_0, \lambda_1, q_{01}, q_{10})$,
            \pause
        \item $P_{ij}(t)$ be the probability that the MMPP is in state $j$
            after time $t$, and that no arrivals have occured by time $t$,
            \pause
        \item $L_i(D) = \mathcal{L}_i(\theta | D)$ be the likelihood of
            $\theta$ given that the Markov chain starts in state $i$.
            \pause
        \item $\pi$ be the equilibrium distribution of the Markov chain.
    \end{itemize}
\end{frame}

\begin{frame}{MMPP likelihood}
    \begin{center}
    \begin{tikzpicture}
        \node at (-0.5, 0) {$D = $};

        \draw (0, 0) -- (4.5, 0);
        \draw (1, -0.1) -- (1, 0.1);
        \draw (3, -0.1) -- (3, 0.1);
        \draw (4.5, -0.1) -- (4.5, 0.1);

        \draw [decorate, decoration={brace, amplitude=10pt}] (1, 0.3) -- (4.5, 0.3);
        \node at (2.75, 1) {$D'$};

        \node at (0.5, -0.3) {$t_1$};
        \node at (2, -0.3) {$t_2$};
        \node at (3.75, -0.3) {$t_3$};
    \end{tikzpicture}
    \end{center}
    \uncover<2->{
    For the ordinary Poisson process,
    \[
        \mathcal{L}(\lambda \mid D) = \exp(-\lambda t_1) \times \lambda \times 
        \mathcal{L}(\lambda \mid D').
    \]
    }
    \uncover<3->{
    For the MMPP,
    }
    \begin{align*}
        \uncover<3->{L_0(D) &= P_{00}(t_1)\,\lambda_0\,L_0(D')}
        \uncover<4->{ + P_{01}(t_1)\,\lambda_1\,L_1(D')} \\[1em]
        \uncover<5->{
        \mathcal{L}(\theta \mid D) &= \pi_0 L_0(D) + \pi_1 L_1(D)
        }
    \end{align*}
\end{frame}

\begin{frame}{MMPP likeilhood}
    \begin{center}
    \begin{tikzpicture}
        \coordinate (root);
        \node (lc) [below left=2cm and 1cm of root, regular polygon, regular polygon sides=3, draw] {$T_L$};
        \node (rc) [below right=2cm and 1cm of root, regular polygon, regular polygon sides=3, draw] {$T_R$};
        \draw (lc.north) -- node [auto] {$t_1$} (root) -- node [auto] {$t_2$} (rc.north);
        \node [above left=0.5cm and 1cm of lc.north] {$T = $};
    \end{tikzpicture}
    \end{center}

    For a tree, we take the product of the left and right children's likelihoods.
    \begin{align*}
        L_0(T) = &\left[P_{00}(t_1)\,\lambda_0\,L_0(T_L) + P_{01}(t_1)\,\lambda_1\,L_1(T_L)\right] \\
        \times &\left[P_{00}(t_1)\,\lambda_0\,L_0(T_R) + P_{01}(t_1)\,\lambda_1\,L_1(T_R)\right].
    \end{align*}
\end{frame}

\begin{frame}{Calculating $P_{ij}(t)$}
    Recall $P_{ij}(t)$ is the probability that, after time $t$, no branchings
    have occured \emph{and} the Markov chain is in state $j$, given that the
    Markov chain started in state $i$.
    \begin{align*}
    \uncover<2->{
        P_{ij}(0) &= 
        \begin{cases}
            1 & i = j \\
            0 & \text{otherwise}
        \end{cases} \\
    }\uncover<3->{
        \dd{P_{ij}}{t} &= 
        \begin{cases}
            \sum_{k} P_{ik}(t) q_{kj} & i \neq j \\
            \uncover<4->{P_{ij}(t) (-\sum_{k} q_{jk}) - \lambda_j & i = j.}
        \end{cases}
    }
    \end{align*}

    \tiny
    Fischer, Wolfgang, and Kathleen Meier-Hellstern. ``The Markov-modulated
    Poisson process (MMPP) cookbook.'' Performance evaluation 18.2 (1993):
    149-171.\par
\end{frame}

\begin{frame}{Finally: the likelihood}
    To calculate $\mathcal{L}(\theta \mid T)$:
    \begin{itemize}
        \uncover<2->{
        \item calculate $P_{ij}(t)$ for all states $i, j$ and all branch
            lengths $t$ by numerical integration
        }
        \uncover<3->{
        \item calculate $L_i(T')$ for all states $i$ and all subtrees $T'$ by
            Felsenstein's pruning algorithm (dynamic programming)
        }
    \end{itemize}

    \uncover<4->{
    Optimize $\mathcal{L}(\theta \mid T)$ for $\theta$ using the covariance
    matrix adaptation evolutionary strategy (CMA-ES).
    }

    \tiny
    \vfill
    Hansen, Nikolaus, Sibylle D. Müller, and Petros Koumoutsakos. ``Reducing
    the time complexity of the derandomized evolution strategy with covariance
    matrix adaptation (CMA-ES).'' Evolutionary Computation 11.1 (2003): 1-18.\par
\end{frame}

\againframe<2>{steps}

\begin{frame}{Assigning states to nodes}
    \begin{center}
    \begin{tikzpicture}
        \coordinate (root);
        \node (lc) [below left=2cm and 1cm of root, regular polygon, regular polygon sides=3, draw] {$T_L$};
        \node (rc) [below right=2cm and 1cm of root, regular polygon, regular polygon sides=3, draw] {$T_R$};
        \draw (lc.north) -- node [auto] {$t_1$} (root) -- node [auto] {$t_2$} (rc.north);
        \node [above left=0.5cm and 1cm of lc.north] {$T = $};
    \end{tikzpicture}
    \end{center}

    \begin{align*}
        L_0(T) = &\uncover<2->{\alert{\max}}\left(P_{00}(t_1)\,\lambda_0\,L_0(T_L) 
        \uncover<2->{,}\uncover<1>{\alert{+}}
        P_{01}(t_1)\,\lambda_1\,L_1(T_L)\right) \times \\
        &\uncover<2->{\alert{\max}}\left(P_{00}(t_1)\,\lambda_0\,L_0(T_R)
        \uncover<2->{,}\uncover<1>{\alert{+}} P_{01}(t_1)\,\lambda_1\,L_1(T_R)\right).
    \end{align*}
\end{frame}

\section{Application: HIV in British Columbia}

\begin{frame}{Human immunodeficiency virus (HIV)}
    \vspace{-0.5cm}
    \begin{center}
    \includegraphics[width=0.75\textwidth]{hiv-structure}
    \end{center}
    \tiny
    By Thomas Splettstoesser (www.scistyle.com) (Own work) [CC BY-SA 4.0
    (http://creativecommons.org/licenses/by-sa/4.0)], via Wikimedia Commons \par
\end{frame}

\begin{frame}{HIV genome}
    \vspace{1cm}

    \includegraphics[width=\textwidth]{hiv-genome}

    \vspace{2cm}
    \tiny
    By Thomas Splettstoesser (www.scistyle.com) (Own work) [CC BY-SA 3.0
    (http://creativecommons.org/licenses/by-sa/3.0) or GFDL
    (http://www.gnu.org/copyleft/fdl.html)], via Wikimedia Commons\par
\end{frame}

\begin{frame}{Making an HIV phylogeny}
    \begin{tikzpicture}

    \end{tikzpicture}
\end{frame}

\begin{frame}{HIV subtype B in British Columbia}
    \tikz[remember picture,overlay] \node [anchor=north east, inner sep=1.8 cm] at (current page.north east) {\includegraphics[width=0.3\textwidth]{timetree-riskgroup-legend}};

    \vspace{-1cm}
    \includegraphics[height=\textwidth,angle=-90]{timetree-riskgroup}
\end{frame}

\begin{frame}{Experimental parameters}
    \begin{description}
        \item[number of rates:] 2, 3, 4
        \item[upper bound on branching rates (branchings/day):] 1, \nicefrac17, \nicefrac{1}{30}
        \item[upper bound on transition rates (transitions/day):] 1, \nicefrac17, \nicefrac{1}{30}
        \item[use tip branches:] yes, no
        \item[branching rates change at nodes only:] yes, no
    \end{description}
\end{frame}

\begin{frame}{3 rates, upper bounds = \nicefrac17}
    \includegraphics[width=\textwidth, trim=0 0 0 0.8in, clip=true]{pcbr1}
\end{frame}

\begin{frame}{4 rates, upper bounds = \nicefrac17, rate changes at nodes}
    \includegraphics[width=\textwidth, trim=0 0 0 0.8in, clip=true]{pcbr2}
\end{frame}

\section{Challenges: validation}

\end{document}

\begin{frame}
    \frametitle{Genetic distance based clustering}
    \tikz[remember picture,overlay] \node [anchor=north east, inner sep=1.5 cm] at (current page.north east) {\includegraphics[width=0.3\textwidth]{timetree-riskgroup-legend}};

    \tikz[remember picture,overlay] \node [anchor=north west, inner ysep=1 cm] at (current page.north west) {\includegraphics[width=0.6\textwidth]{timetree-pd-polar}};

    \tikz[remember picture,overlay] \node [anchor=south east, inner ysep=0 cm] at (current page.south east) {\includegraphics[width=0.6\textwidth]{timetree-riskgroup-polar}};
\end{frame}

\begin{frame}
    \frametitle{Model-based clustering}
    \tikz[remember picture,overlay] \node [anchor=north east, inner sep=1.5 cm] at (current page.north east) {\includegraphics[width=0.3\textwidth]{timetree-riskgroup-legend}};

    \tikz[remember picture,overlay] \node [anchor=north west, inner ysep=1 cm] at (current page.north west) {\includegraphics[width=0.6\textwidth]{timetree_4}};

    \tikz[remember picture,overlay] \node [anchor=south east, inner ysep=0 cm] at (current page.south east) {\includegraphics[width=0.6\textwidth]{timetree-riskgroup-polar}};
\end{frame}

\begin{frame}
    \frametitle{Preliminary comparison}
    \tikz[remember picture,overlay] \node [anchor=north east, inner ysep=1.5 cm, inner xsep=0.5 cm, text width=6cm] at (current page.north east) {$\longleftarrow$ 79 clusters \\ median [IQR] 29 [16-56] individuals};

    \tikz[remember picture,overlay] \node [anchor=north west, inner ysep=1 cm] at (current page.north west) {\includegraphics[width=0.6\textwidth]{cluster4}};

    \tikz[remember picture,overlay] \node [anchor=south west, inner xsep=0.5 cm, inner ysep=1 cm, text width=6cm] at (current page.south west) {\hfill 221 clusters $\longrightarrow$ \\ \hfill median [IQR] 5 [8-15] individuals};

    \tikz[remember picture,overlay] \node [anchor=south east, inner ysep=0 cm] at (current page.south east) {\includegraphics[width=0.6\textwidth]{timetree-pd-polar}};
\end{frame}

\begin{frame}
    \frametitle{Challenges}
    \tikz[remember picture,overlay] \node [anchor=north east, inner ysep=1.5 cm, inner xsep=0.5 cm, text width=6.5cm] at (current page.north east) {$\longleftarrow$ likelihood ratio test doesn't work? \\ $\qquad$ how to define clusters?};

    \tikz[remember picture,overlay] \node [anchor=north west, inner ysep=1 cm] at (current page.north west) {\includegraphics[width=0.6\textwidth]{timetree_6}};

    \tikz[remember picture,overlay] \node [anchor=south west, inner xsep=1.5 cm, inner ysep=1 cm] at (current page.south west) {scaling tip branches $\longrightarrow$};

    \tikz[remember picture,overlay] \node [anchor=south east, inner ysep=0 cm] at (current page.south east) {\includegraphics[width=0.6\textwidth]{timetree-qnorm_3}};
\end{frame}

\begin{frame}
    \frametitle{Acknowledgements}

    \textbf{BC Centre for Excellence in HIV/AIDS} \\
    $\qquad$ Art Poon \\
    $\qquad$ Richard Liang \\
    $\qquad$ Jeffrey Joy \\
    $\qquad$ T. Nguyen \\

    \vspace{3cm}
    \includegraphics[width=0.2\textwidth]{logos/nserc}
    \includegraphics[width=0.35\textwidth]{logos/cfe}
    \includegraphics[width=0.35\textwidth]{logos/btp}
    \includegraphics[width=0.1\textwidth]{logos/ubc}
\end{frame}

\end{document}

### 2015-10-18 doc/stats-seminar/presentation.tex
Almost finished stats seminar talk...

@@
 
-\begin{frame}{Making an HIV phylogeny}
+\begin{frame}{Population-level HIV sequencing}
     \begin{tikzpicture}
+        \node (a) {\includegraphics[height=1.5cm]{person7}};
+        \node (b) [right=of a] {\includegraphics[height=1.5cm]{blood_testing}};
+        \draw [->, >=stealth, thick] (a) -- (b);
+        \uncover<2->{
+        \node (c) [below=of b] {\includegraphics[height=1.5cm]{hiv}\includegraphics[height=1.5cm]{alignment}};
+        \draw [->, >=stealth, thick] (b) -- (c);
+        }
+        \uncover<3->{
+        \node (d) [right=of c, rectangle, draw, inner ysep=12pt] {\includegraphics[height=1.5cm]{medical}};
+        \draw [->, >=stealth, thick] (c) -- (d);
+        }
+        \uncover<4->{
+        \node (e) [below=of c] {\includegraphics[height=0.25cm]{sequence1}};
+        \draw [->, >=stealth, thick] (c) -- (e);
+        }
+    \end{tikzpicture}
+\end{frame}
 
-    \end{tikzpicture}
+\begin{frame}{Maximum likelihood HIV phylogeny}
+    \begin{columns}
+        \begin{column}{0.5\textwidth}
+            \includegraphics[height=1cm]{person7}\includegraphics[height=0.25cm]{sequence1}
+
+            \includegraphics[height=1cm]{person8}\includegraphics[height=0.25cm]{sequence2}
+
+            \includegraphics[height=1cm]{person9}\includegraphics[height=0.25cm]{sequence4}
+
+            \includegraphics[height=1cm]{person10}\includegraphics[height=0.25cm]{sequence5}
+        \end{column}
+        \begin{column}{0.5\textwidth}
+            \centering
+            \uncover<2->{
+            \includegraphics[width=\textwidth]{unrooted}
+
+            FastTree 2
+            }
+        \end{column}
+    \end{columns}
+
+    \tiny
+    Price, Morgan N., Paramvir S. Dehal, and Adam P. Arkin. ``FastTree
+    2 -- approximately maximum-likelihood trees for large alignments.'' PloS one
+    5.3 (2010): e9490.\par
+\end{frame}
+
+\begin{frame}{Time scaling the phylogeny}
+
+    \begin{columns}
+        \begin{column}{0.5\textwidth}
+            \centering
+            \includegraphics[height=1cm]{unrooted}
+
+            \includegraphics[width=\textwidth]{rtt}
+        \end{column}
+        \begin{column}{0.5\textwidth}
+            \uncover<2->{
+            \includegraphics[width=\textwidth]{rooted}
+            }
+        \end{column}
+    \end{columns}
+
+    \tiny
+    Drummond, Alexei, G. Oliver, and Andrew Rambaut. ``Inference of viral
+    evolutionary rates from molecular sequences.'' Advances in parasitology 54
+    (2003): 331-358.\par
 \end{frame}
@@
 
+\begin{frame}{4 rates, upper bounds = \nicefrac17, rate changes at nodes}
+    \includegraphics[width=\textwidth, trim=0 0 0 0.8in, clip=true]{pcbr2}
+\end{frame}
+
 \begin{frame}{3 rates, upper bounds = \nicefrac17}
@@
 
-\begin{frame}{4 rates, upper bounds = \nicefrac17, rate changes at nodes}
-    \includegraphics[width=\textwidth, trim=0 0 0 0.8in, clip=true]{pcbr2}
+\section{Challenges}
+
+\section{Previous projects}
+
+\begin{frame}{Phylogenetic estimation of transmitted/founder viruses}
+    \begin{center}
+    \begin{tikzpicture}
+        [scale=1.7, every node/.style={inner sep=0pt}]
+        \node (root) at (1, 2) {\includegraphics[width=1cm]{hiv}};
+        \node (lc) at (0.5, 1) {\includegraphics[width=1cm]{hiv}};
+        \node (t1) at (0, 0) {\includegraphics[width=1cm]{hiv}};
+        \node (t2) at (1, 0) {\includegraphics[width=1cm]{hiv}};
+        \node (t3) at (2, 0) {\includegraphics[width=1cm]{hiv}};
+        \draw (root) -- (lc) -- (t1);
+        \draw (lc) -- (t2);
+        \draw (root) -- (t3);
+
+        \node at (3, 2) {$\longrightarrow$ transmitted/founder};
+        \node at (3, 0) {$\longrightarrow$ sampled};
+    \end{tikzpicture}
+\end{center}
+
+    Find patients with known early sequences; try to reconstruct with
+    phylogenetic methods.
+
+    \tiny
+    McCloskey, Rosemary M., et al. ``An Evaluation of Phylogenetic Methods for
+    Reconstructing Transmitted HIV Variants using Longitudinal Clonal HIV
+    Sequence Data.'' Journal of virology 88.11 (2014): 6181-6194.\par
 \end{frame}
 
-\section{Challenges: validation}
+\begin{frame}{Phylogenetic estimation of transmitted/founder viruses}
+    \begin{description}
+        \setlength{\itemsep}{16pt}
+        \item [{\Huge55}] publications with early and late sequences
+        \item [{\Huge232}] patients
+        \item [{\Huge335}] gene data sets
+        \item [{\Huge19,486}] total HIV sequences
+    \end{description}
+\end{frame}
 
-\end{document}
+\begin{frame}{Phylogenetic estimation of transmitted/founder viruses}
+    \includegraphics[width=\textwidth]{mccloskey2014evaluation_f1}
+\end{frame}
 
-\begin{frame}
-    \frametitle{Genetic distance based clustering}
-    \tikz[remember picture,overlay] \node [anchor=north east, inner sep=1.5 cm] at (current page.north east) {\includegraphics[width=0.3\textwidth]{timetree-riskgroup-legend}};
+\begin{frame}{Phylogenetic methods are informative}
+    \includegraphics[height=\textheight]{mccloskey2014evaluation_f2}
+\end{frame}
 
-    \tikz[remember picture,overlay] \node [anchor=north west, inner ysep=1 cm] at (current page.north west) {\includegraphics[width=0.6\textwidth]{timetree-pd-polar}};
+\begin{frame}{Single origin of Hepatitis C virus Q80K polymorphism}
+    \begin{itemize}
+        \setlength{\itemsep}{12pt}
+        \item The Q80K polymorphism occurs in 5-47\% of HCV cases
+        \item Confers mild resistance to some therapies 
+        \item We investigated the origin and transmissibility of Q80K using
+            publicly available sequence data
+    \end{itemize}
 
-    \tikz[remember picture,overlay] \node [anchor=south east, inner ysep=0 cm] at (current page.south east) {\includegraphics[width=0.6\textwidth]{timetree-riskgroup-polar}};
+    \vspace{1cm}
+    \tiny Lin, M. Valerie, and Raymond Chung. ``Recent FDA approval of sofosbuvir and simeprevir. Implications for current HCV treatment.'' Clinical Liver Disease 3.3 (2014): 65-68. \par
+    \tiny McCloskey, Rosemary M., et al. ``Global origin and transmission of hepatitis C virus NS3 Q80 K polymorphism.'' Journal of Infectious Diseases (2014): jiu613. \par
+\end{frame}
+
+\begin{frame}{Single origin of Hepatitis C virus Q80K polymorphism}
+    \begin{description}
+        \setlength{\itemsep}{16pt}
+        \item [{\Huge160 071}] unfiltered sequences
+        \item [{\Huge13}] countries of origin
+        \item [{\Huge4}] filtering steps
+        \item [{\Huge724}] filtered sequences
+    \end{description}
+\end{frame}
+
+\begin{frame}{Single origin of Hepatitis C virus Q80K polymorphism}
+    \includegraphics[width=\textwidth]{mccloskey2015global_f1}
+
+    \includegraphics[width=\textwidth]{mccloskey2015global_supp}
 \end{frame}
@@
 \begin{frame}
-    \frametitle{Model-based clustering}
-    \tikz[remember picture,overlay] \node [anchor=north east, inner sep=1.5 cm] at (current page.north east) {\includegraphics[width=0.3\textwidth]{timetree-riskgroup-legend}};
-
-    \tikz[remember picture,overlay] \node [anchor=north west, inner ysep=1 cm] at (current page.north west) {\includegraphics[width=0.6\textwidth]{timetree_4}};
-
-    \tikz[remember picture,overlay] \node [anchor=south east, inner ysep=0 cm] at (current page.south east) {\includegraphics[width=0.6\textwidth]{timetree-riskgroup-polar}};
-\end{frame}
-
-\begin{frame}
-    \frametitle{Preliminary comparison}
-    \tikz[remember picture,overlay] \node [anchor=north east, inner ysep=1.5 cm, inner xsep=0.5 cm, text width=6cm] at (current page.north east) {$\longleftarrow$ 79 clusters \\ median [IQR] 29 [16-56] individuals};
-
-    \tikz[remember picture,overlay] \node [anchor=north west, inner ysep=1 cm] at (current page.north west) {\includegraphics[width=0.6\textwidth]{cluster4}};
-
-    \tikz[remember picture,overlay] \node [anchor=south west, inner xsep=0.5 cm, inner ysep=1 cm, text width=6cm] at (current page.south west) {\hfill 221 clusters $\longrightarrow$ \\ \hfill median [IQR] 5 [8-15] individuals};
-
-    \tikz[remember picture,overlay] \node [anchor=south east, inner ysep=0 cm] at (current page.south east) {\includegraphics[width=0.6\textwidth]{timetree-pd-polar}};
-\end{frame}
-
-\begin{frame}
-    \frametitle{Challenges}
-    \tikz[remember picture,overlay] \node [anchor=north east, inner ysep=1.5 cm, inner xsep=0.5 cm, text width=6.5cm] at (current page.north east) {$\longleftarrow$ likelihood ratio test doesn't work?
-\\ $\qquad$ how to define clusters?};
-
-    \tikz[remember picture,overlay] \node [anchor=north west, inner ysep=1 cm] at (current page.north west) {\includegraphics[width=0.6\textwidth]{timetree_6}};
-
-    \tikz[remember picture,overlay] \node [anchor=south west, inner xsep=1.5 cm, inner ysep=1 cm] at (current page.south west) {scaling tip branches $\longrightarrow$};
-
-    \tikz[remember picture,overlay] \node [anchor=south east, inner ysep=0 cm] at (current page.south east) {\includegraphics[width=0.6\textwidth]{timetree-qnorm_3}};
-\end{frame}
-
-\begin{frame}
-    \frametitle{Acknowledgements}
-
+    \frametitle{Thanks}
+    \begin{columns}
+        \begin{column}{0.5\textwidth}
             \textbf{BC Centre for Excellence in HIV/AIDS} \\
@@
             $\qquad$ Richard Liang \\
+            $\qquad$ P. Richard Harrigan \\
             $\qquad$ Jeffrey Joy \\
-    $\qquad$ T. Nguyen \\
+            $\qquad$ Thuy Nguyen
+        \end{column}
+        \begin{column}{0.5\textwidth}
+            \textbf{UBC} \\
+            $\qquad$ Sarah Otto \\
+            $\qquad$ Alexandre Bouchard-C\^ot\'e \\
+            $\qquad$ Wayne Maddison \\
+            $\qquad$ Sara Mostafavi \\
+            \textbf{SFU} \\
+            $\qquad$ Ryan Morin \\
+            $\qquad$ Marni Mishna \\
+            $\qquad$ Cedric Chauve
+        \end{column}
+    \end{columns}
 
-    \vspace{3cm}
-    \includegraphics[width=0.2\textwidth]{logos/nserc}
-    \includegraphics[width=0.35\textwidth]{logos/cfe}
-    \includegraphics[width=0.35\textwidth]{logos/btp}
-    \includegraphics[width=0.1\textwidth]{logos/ubc}
+    \vspace{1cm}
+    \includegraphics[width=0.35\textwidth]{cfe}
+    \includegraphics[width=0.35\textwidth]{btp}
+    \includegraphics[width=0.1\textwidth]{ubc}
+    \hspace{0.5cm}
+    \includegraphics[width=0.1\textwidth]{ubcstats}
 \end{frame}

### 2015-10-19 doc/stats-seminar/presentation.tex
Updates to stats seminar talk

@@
 
-\section{Motivation: more informed phylogenetic clustering}
+\section{Motivation: phylogenetic clustering}
 
@@
 
-\begin{frame}{Popular problems}
-    \begin{columns}
-        \begin{column}{0.5\textwidth}
-            \includegraphics[width=\textwidth, trim=0 11in 0 0, clip=true]{hughes2009molecular-f1}
-        \end{column}
-        \begin{column}{0.5\textwidth}
-            \includegraphics[width=\textwidth]{lewis2008episodic-f2}
-        \end{column}
-    \end{columns}
-    \begin{columns}
-        \begin{column}{0.5\textwidth}
-            \tiny
-            Hughes, Gareth J., et al. ``Molecular phylodynamics of the heterosexual HIV epidemic in the United Kingdom.'' PLoS Pathog 5.9 (2009): e1000590.
-        \end{column}
-        \begin{column}{0.5\textwidth}
-            \tiny
-            Lewis, Fraser, et al. ``Episodic sexual transmission of HIV revealed by molecular phylodynamics.'' PLoS Med 5.3 (2008): e50.
-        \end{column}
-    \end{columns}
-\end{frame}
+%\begin{frame}{Popular problems}
+%    \begin{columns}
+%        \begin{column}{0.5\textwidth}
+%            \includegraphics[width=\textwidth, trim=0 11in 0 0, clip=true]{hughes2009molecular-f1}
+%        \end{column}
+%        \begin{column}{0.5\textwidth}
+%            \includegraphics[width=\textwidth]{lewis2008episodic-f2}
+%        \end{column}
+%    \end{columns}
+%    \begin{columns}
+%        \begin{column}{0.5\textwidth}
+%            \tiny
+%            Hughes, Gareth J., et al. ``Molecular phylodynamics of the heterosexual HIV epidemic in the United Kingdom.'' PLoS Pathog 5.9 (2009): e1000590.
+%        \end{column}
+%        \begin{column}{0.5\textwidth}
+%            \tiny
+%            Lewis, Fraser, et al. ``Episodic sexual transmission of HIV revealed by molecular phylodynamics.'' PLoS Med 5.3 (2008): e50.
+%        \end{column}
+%    \end{columns}
+%\end{frame}
 
@@
                 \draw (a) -- (e);
+                \draw (b) -- (e);
+                \draw (c) -- (e);
+                \draw (c) -- (d);
+                \draw (d) -- (e);
             \end{tikzpicture}
@@
 
-\section{A Markov-modulated Poisson process on phylogenies}
+\section{Methods: a Markov-modulated Poisson process on phylogenies}
 
@@
             \draw [blue] (a) -- node [auto, text=black] {
-                $T \sim $ Exponential(?)
+                $T \sim $ Exponential($\cdots$)
             } (b);
@@
 
-\begin{frame}{Markov-modulated Poisson processes}
-    \includegraphics[width=\textwidth]{muscariello2004mmpp}
-
-    \includegraphics[width=\textwidth]{ching1997markov}
-
-    \includegraphics[width=\textwidth]{lu2012markov}
-\end{frame}
+%\begin{frame}{Markov-modulated Poisson processes}
+%    \includegraphics[width=\textwidth]{muscariello2004mmpp}
+%
+%    \includegraphics[width=\textwidth]{ching1997markov}
+%
+%    \includegraphics[width=\textwidth]{lu2012markov}
+%\end{frame}
 
@@
     \end{align*}
+
+    \tiny
+    Felsenstein, Joseph. ``Evolutionary trees from DNA sequences: a maximum
+    likelihood approach.'' Journal of molecular evolution 17.6 (1981): 368-376.\par
 \end{frame}
@@
     \end{align*}
+
+    \tiny
+    Pupko, Tal, et al. ``A fast algorithm for joint reconstruction of ancestral
+    amino acid sequences.'' Molecular Biology and Evolution 17.6 (2000):
+    890-896.\par
 \end{frame}
@@
 
+\begin{frame}{Long terminal branches}
+    todo
+\end{frame}
+
+\begin{frame}{Phylogenetic clusters $\neq$ contact network clusters}
+    \begin{columns}
+        \begin{column}{0.5\textwidth}
+            \includegraphics[width=\textwidth]{ba-net}
+        \end{column}
+        \begin{column}{0.5\textwidth}
+            \includegraphics[width=\textwidth]{ba-tree}
+        \end{column}
+    \end{columns}
+\end{frame}
+
 \section{Previous projects}
@@
 \begin{frame}{Phylogenetic estimation of transmitted/founder viruses}
-    \includegraphics[width=\textwidth]{mccloskey2014evaluation_f1}
+    \begin{center}
+        \includegraphics[width=0.9\textwidth]{mccloskey2014evaluation_f1}
+    \end{center}
 \end{frame}
@@
 
-    \vspace{1cm}
+    \includegraphics[width=0.2\textwidth]{nserc}
+    \hspace{0.2cm}
+    \includegraphics[width=0.15\textwidth]{sfu}
+
     \includegraphics[width=0.35\textwidth]{cfe}

### 2015-10-20 programs/netsim/src/netabc.c
Fixed up SMC (it's not working, of course)

@@ -79,7 +79,7 @@ struct netabc_options get_options(int argc, char **argv)
 
 void ba_sample_from_prior(gsl_rng *rng, double *theta)
 {
-    *theta = gsl_ran_flat(rng, 0.1, 1.0);
+    *theta = gsl_ran_flat(rng, PA_POWER_MIN, PA_POWER_MAX);
 }
 
 double ba_prior_density(const double *theta)
@@ -99,10 +99,11 @@ void ba_propose(gsl_rng *rng, double *theta, const void *params)
 double ba_proposal_density(const double *from, const double *to, const void *params)
 {
     double var = * (double *) params;
+    if (*to < PA_POWER_MIN || *to > PA_POWER_MAX)
+        return 0.;
     return gsl_ran_gaussian_pdf(*to - *from, sqrt(2*var));
 }
 
-// TODO: I think this leaks memory
 void ba_sample_dataset(gsl_rng *rng, const double *theta, void *X)
 {
     int i;
@@ -111,8 +112,8 @@ void ba_sample_dataset(gsl_rng *rng, const double *theta, void *X)
     igraph_vector_t v;
 
     igraph_vector_init(&v, NNODE);
-    igraph_barabasi_game(&net, NNODE, *theta, MEAN_DEGREE/2, NULL, 0, 1, 0, 
-                         IGRAPH_BARABASI_PSUMTREE, NULL);
+    igraph_barabasi_game(&net, NNODE, *theta, MEAN_DEGREE/2, NULL, 0,
+            1, 0, IGRAPH_BARABASI_PSUMTREE, NULL);
     igraph_to_directed(&net, IGRAPH_TO_DIRECTED_MUTUAL);
 
     igraph_vector_fill(&v, 0);
@@ -143,13 +144,18 @@ double ba_distance(const void *x, const void *y)
     double kx = GAN(gx, "kernel");
     double ky = GAN(gy, "kernel");
     double kxy = kernel(gx, gy, DECAY_FACTOR, RBF_VARIANCE, 1, INFINITY);
-    return 1 - kxy / sqrt(kx * ky);
+    return 1. - kxy / sqrt(kx * ky);
 }
 
 void ba_feedback(const double *theta, int nparticle, void *params)
 {
     double var = gsl_stats_variance(theta, 1, nparticle);
     memcpy(params, &var, sizeof(double));
+}
+
+void ba_destroy_dataset(void *z)
+{
+    igraph_destroy((igraph_t *) z);
 }
 
[…]

### 2015-10-20 programs/netsim/src/smc.c

@@ -2,155 +2,219 @@
 #include <string.h>
 #include <math.h>
 #include <float.h>
+#include <pthread.h>
 #include <gsl/gsl_roots.h>
 #include "smc.h"
 #include "util.h"
 
 #define RESIZE_AMOUNT 100
+#define BISECTION_MAX_ITER 10000
 #define NTHREAD 1
 
-struct objfun_params {
-    double *X;
-    double *W;
-    double *new_W;
-    double epsilon;
-    const smc_config *config;
-};
+/** All the data used for SMC.
+ *
+ * Because the SMC algorithm is parallelizable, we store all the data in a
+ * global structure which is accessed by each thread.
+ */
+typedef struct {
+    const smc_config *config;       /**< configuration parameters */
+    const smc_functions *functions; /**< user-supplied functions */
 
-struct perturb_workspace {
-    gsl_rng *rng;
-    const smc_config *config;
-    const smc_functions *functions;
-    double *W;
-    double *theta;
-    double *X;
-    double epsilon;
-    const void *data;
-    char *fdbk;
-    double *new_theta;
-    double *new_X;
-    char *z;
-    int nparticle;
-};
+    const void *data;               /**< input data */
+    char *z;                        /**< simulated datasets */
+    char *fdbk;                     /**< feedback from current particles */
 
-void resample(const smc_config config, double *W, double *theta);
-double next_epsilon(struct objfun_params *objpar);
+    double *theta;                  /**< current particles */
+    double *new_theta;              /**< storage space for proposed particles */
+
+    double *W;                      /**< current weights */
+    double *new_W;                  /**< storage space for proposed new weights */
+
+    double *X;                      /**< distances of simulated datasets */
+    double *new_X;                  /**< storage space for proposed new dataset distances */
+
+    double epsilon;                 /**< current tolerance */
+
+    int accept;                     /**< number of accepted proposals */
+    int alive;                      /**< number of alive particles */
+} smc_workspace;
+
+/** Arguments to the perturb function.
+ *
+ * One of these objects is created for each thread.
+ */
+typedef struct {
[…]

### 2015-10-20 programs/netsim/src/smc.h

@@ -23,6 +23,7 @@ typedef struct {
     void   (*sample_dataset)    (gsl_rng *rng, const double *, void *);
     double (*distance)          (const void *, const void *);
     void   (*feedback)          (const double *, int, void *);
+    void   (*destroy_dataset)   (void *);
 } smc_functions;
 
 typedef struct {

### 2015-10-20 programs/netsim/src/tree.c

@@ -90,8 +90,8 @@ void ladderize(igraph_t *tree)
         igraph_vector_set(&permvec, perm[i], (double) i);
     igraph_permute_vertices(tree, &new_tree, &permvec);
 
-    // this maybe is questionable
-    memcpy(tree, &new_tree, sizeof(*tree));
+    igraph_destroy(tree);
+    memcpy(tree, &new_tree, sizeof(igraph_t));
 
     free(perm);
     igraph_vector_destroy(&work);
@@ -231,6 +231,7 @@ void subsample_tips(igraph_t *tree, int ntip, const gsl_rng *rng)
         for (j = 0; j < igraph_vector_size(elem); ++j) {
             igraph_vector_push_back(&keep_all, VECTOR(*elem)[j]);
         }
+        igraph_vector_destroy(elem);
     }
 
     igraph_vector_sort(&keep_all);

### 2015-10-20 programs/netsim/tests/check_simulate.c

@@ -34,7 +34,7 @@ igraph_t *tree_from_newick(const char *newick)
 
 START_TEST (test_simulate_chain)
 {
-    igraph_t net, *tree;
+    igraph_t net, tree;
     char buf[2];
     int i, from, to;
     double diff;
@@ -56,19 +56,19 @@ START_TEST (test_simulate_chain)
     {
         SETEAN(&net, "transmit", i, 1);
     }
-    tree = simulate_phylogeny(&net, rng, 100, 100, 0);
+    simulate_phylogeny(&tree, &net, rng, 100, 100, 0);
 
-    ck_assert_int_eq(igraph_vcount(tree), 9);
-    ck_assert_int_eq(igraph_ecount(tree), 8);
-    for (i = 0; i < igraph_ecount(tree); ++i)
+    ck_assert_int_eq(igraph_vcount(&tree), 9);
+    ck_assert_int_eq(igraph_ecount(&tree), 8);
+    for (i = 0; i < igraph_ecount(&tree); ++i)
     {
-        igraph_edge(tree, i, &from, &to);
-        diff = fabs(atoi(VAS(tree, "id", from)) - atoi(VAS(tree, "id", to)));
+        igraph_edge(&tree, i, &from, &to);
+        diff = fabs(atoi(VAS(&tree, "id", from)) - atoi(VAS(&tree, "id", to)));
         ck_assert(diff == 0.0 || diff == 1.0);
     }
 
     igraph_destroy(&net);
-    igraph_destroy(tree);
+    igraph_destroy(&tree);
 }
 END_TEST
 

### 2015-10-20 programs/netsim/tests/check_smc.c

@@ -20,6 +20,8 @@ void toy_sample_from_prior (gsl_rng *rng, double *theta)
 
 double toy_prior_density(const double *theta)
 {
+    if (*theta < -10 || *theta > 10)
+        return 0.;
     return 1./20.;
 }
 
@@ -58,6 +60,11 @@ void toy_feedback(const double *theta, int nparticle, void *params)
     memcpy(params, &var, sizeof(double));
 }
 
+void toy_destroy_dataset(void *z)
+{
+    return;
+}
+
 smc_functions toy_functions = {
     .sample_from_prior = toy_sample_from_prior,
     .prior_density = toy_prior_density,
@@ -65,7 +72,8 @@ smc_functions toy_functions = {
     .proposal_density = toy_proposal_density,
     .sample_dataset = toy_sample_dataset,
     .distance = toy_distance,
-    .feedback = toy_feedback
+    .feedback = toy_feedback,
+    .destroy_dataset = toy_destroy_dataset
 };
 
 void plot(const double *x, const double *y, int nx, const char *pdf, 
@@ -99,7 +107,6 @@ START_TEST (test_smc_toy_steps)
 {
     int i;
     double y = 0;
-    gsl_rng *rng = set_seed(0);
     smc_config toy_config = {
         .nparam = 1,
         .nparticle = 10000,
@@ -111,7 +118,8 @@ START_TEST (test_smc_toy_steps)
         .dataset_size = sizeof(double),
         .feedback_size = sizeof(double)
     };
-    smc_result *res = abc_smc(toy_config, toy_functions, rng, (void *) &y);
+
+    smc_result *res = abc_smc(toy_config, toy_functions, 0, (void *) &y);
 
     fprintf(stderr, "%d steps\n", res->niter);
     plot(&res->epsilon[1], NULL, res->niter-1, "check_smc_epsilon.pdf", 
@@ -125,7 +133,6 @@ END_TEST
 START_TEST (test_smc_toy)
 {
     double y = 0;
-    gsl_rng *rng = set_seed(0);
     int i, fd;
 
     smc_config config = {
@@ -140,10 +147,15 @@ START_TEST (test_smc_toy)
         .feedback_size = sizeof(double)
     };
 
-    smc_result *res = abc_smc(config, toy_functions, rng, (void *) &y);
+    smc_result *res = abc_smc(config, toy_functions, 0, (void *) &y);
+
[…]

### 2015-10-20 programs/netsim/tests/check_tree.c

@@ -39,6 +39,7 @@ START_TEST (test_parse_newick_topology)
 
     igraph_t *tree = parse_newick(f);
 
+    fprintf(stderr, "test_parse_newick_topology\n");
     ck_assert_int_eq(igraph_vcount(tree), 7);
     ck_assert_int_eq(igraph_ecount(tree), 6);
 
@@ -58,6 +59,7 @@ START_TEST (test_parse_newick_topology)
 
     igraph_vector_destroy(&vec);
     igraph_destroy(tree);
+    free(tree);
     fclose(f);
 }
 END_TEST
@@ -71,6 +73,7 @@ START_TEST (test_parse_newick_branch_lengths)
     igraph_vector_t size = new_vector();
     igraph_vector_t edge = new_vector();
 
+    fprintf(stderr, "test_parse_newick_branch_lengths\n");
     EANV(tree, "length", &bl);
     ck_assert(igraph_vector_sum(&bl) == 2.84);
 
@@ -93,6 +96,10 @@ START_TEST (test_parse_newick_branch_lengths)
 
     fclose(f);
     igraph_destroy(tree);
+    free(tree);
+    igraph_vector_destroy(&bl);
+    igraph_vector_destroy(&size);
+    igraph_vector_destroy(&edge);
 }
 END_TEST
 
@@ -102,6 +109,7 @@ START_TEST (test_write_newick)
     FILE *f = newick_file("(t1:0.4,((t3:0.05,t2:0.8):0.88,t4:0.25):0.46);");
     igraph_t *tree = parse_newick(f), *copy;
 
+    fprintf(stderr, "test_write_newick\n");
     fseek(f, 0, SEEK_SET);
     write_tree_newick(tree, f);
 
@@ -113,6 +121,9 @@ START_TEST (test_write_newick)
 
     fclose(f);
     igraph_destroy(tree);
+    igraph_destroy(copy);
+    free(tree);
+    free(copy);
 }
 END_TEST
 
@@ -122,6 +133,7 @@ START_TEST (test_root)
     int rt = root(tree);
     igraph_vector_t v = new_vector();
 
+    fprintf(stderr, "test_root\n");
     igraph_degree(tree, &v, igraph_vss_1(rt), IGRAPH_OUT, 1);
     ck_assert_int_eq((int) VECTOR(v)[0], 2);
     igraph_degree(tree, &v, igraph_vss_1(rt), IGRAPH_IN, 1);
[…]

### 2015-11-19 doc/journal-club/nov18/robinson2013dynamics.tex
Added Nov 18 JC presentation

\documentclass{beamer}
\usepackage{gillius}

\usetheme[everytitleformat=regular]{m}
\setbeamertemplate{navigation symbols}{}

\title{How the dynamics and structure of sexual contact networks shape pathogen
       phylogenies}
\author[Robinson et al.]{Katy Robinson \and Nick Fyson \and Ted Cohen \and
Christophe Fraser \and Caroline Colijn}

\begin{document}
\definecolor{red}{RGB}{228,26,28}
\definecolor{blue}{RGB}{55,126,184}
\definecolor{green}{RGB}{77,175,74}
\definecolor{purple}{RGB}{152,78,163}

\maketitle

\begin{frame}{Motivation}
    \begin{itemize}
        \item Information about contact networks is highly useful but
            challenging to gather.
        \item Contact networks affect the shape of pathogen phylogenies.
        \item Is there anything phylogenies can tell us that public health
            monitoring can't?
    \end{itemize}
    \begin{center}
        \itshape
        Here we ask to what extent dynamic host contact networks shape pathogen
        phylogenies, under conditions when the epidemic has similar
        trajectories on different networks.
        \normalfont
    \end{center}
\end{frame}

\begin{frame}{How phylogenies are shaped}
    \vspace{-0.5cm}
    \begin{center}
        \includegraphics[height=0.8\textheight]{f1}
    \end{center}
\end{frame}

\begin{frame}{Methods: simulations}
    \begin{itemize}
        \item Two types of dynamic network.
            \begin{itemize}
                \item Degree distribution from the British National Survey of
                    Sexual Attitudes and Lifestyles (NATSAL).
                \item Random.
            \end{itemize}
        \item Pathogens with 10- and 40-week durations of infectiousness.
        \item Simulated transmission and sequence evolution over the network.
            \begin{itemize}
                \item Transmission and mutation are both Poisson processes.
                \item One sequence per patient; no selection.
                \item When did they stop the simulation?
            \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}{Methods: phylogenetics}
    \begin{itemize}
        \item Three sampling schemes.
            \begin{itemize}
                \item 100 sequences at 2 time points.
                \item Uniform number of samples over time.
                \item Proportional sampling over time.
            \end{itemize}
        \item Maximum likelihood phylogenies with PHYLIP.
        \item Clustering by genetic distance cutoff (0.06 expected
            substitutions per site).
        \item Colless' index:
            \[
                \frac{2}{(n-1)(n-2)} \sum_b \lVert T_{iR} - T_{iL} \rVert.
            \]
    \end{itemize}
\end{frame}

\begin{frame}{Results: tree shapes}
    \includegraphics[width=\textwidth]{f2}
\end{frame}

\begin{frame}{Results: cluster numbers and sizes}
    \begin{center}
        \vspace{-0.5cm}
        \includegraphics[width=0.9\textwidth]{f3}
    \end{center}
\end{frame}

\begin{frame}{Results: network influences tree size}
    \begin{center}
        \vspace{-0.5cm}
        \includegraphics[height=0.8\textheight]{f4}
    \end{center}
\end{frame}

\begin{frame}{Results: degree distribution vs. cluster size distribution}
    \begin{center}
        \vspace{-0.5cm}
        \includegraphics[width=0.7\textwidth, trim=0 2.3cm 0 0, clip=true]{f5}
    \end{center}
\end{frame}

\begin{frame}{Results: branch lengths}
    \begin{center}
        \vspace{-0.5cm}
        \includegraphics[height=0.8\textheight]{f6}
    \end{center}
\end{frame}

\begin{frame}{Results: star-like-ness and imbalance}
    \begin{center}
        \vspace{-0.5cm}
        \includegraphics[width=0.8\textwidth]{f7}

        \includegraphics[width=0.7\textwidth]{f8}
    \end{center}
\end{frame}

\begin{frame}{Results: different trees from similar epidemics}
    \includegraphics[width=\textwidth]{f9}
\end{frame}

\begin{frame}{Results: sampling influences imbalance and clustering}
    \begin{center}
        \vspace{-0.5cm}
        \includegraphics[width=0.8\textwidth, trim=0 7.1cm 0 0, clip=true]{f11}

        \includegraphics[width=0.8\textwidth, trim=0 0 0 6.9cm, clip=true]{f11}
    \end{center}
\end{frame}

\end{document}

### 2015-11-19 programs/netabc/R/collect.R

@@ -85,11 +85,16 @@ collect.metadata <- function (data.files)
 
     # fill in the fields each file doesn't have with NA 
     metadata <- mapply(function (df, new) {
-        cbind(df, as.data.frame(setNames(as.list(rep(NA, length(new))), new)))
+        if (length(new) > 0) {
+            cbind(df, as.data.frame(setNames(as.list(rep(NA, length(new))), new)))
+        }
+        else {
+            df
+        }
     }, metadata, new.cols, SIMPLIFY=FALSE)
 
     # put all together and set row names to files
-    metadata <- do.call(rbind, lapply(metadata, "[", keep.cols))
+    metadata <- do.call(rbind, metadata)
     rownames(metadata) <- data.files
     metadata
 }

### 2015-11-19 programs/netabc/R/process.archive.R

@@ -2,6 +2,8 @@
 #'
 #' @param file.name name of the file to operate on
 #' @param fun function to apply to the opened file
+#' @param ... additional arguments for fun
+#' @export
 process.archive <- function (file.name, fun, ...)
 {
     if (grepl("\\.bz2?$", file.name, perl=TRUE)) {

### 2015-11-26 programs/netabc/R/read.mm.R
Added export tags to some functions

@@ -37,6 +37,7 @@ read.mm.general <- function (file)
 #'
 #' @param file.name MatrixMarket file to read
 #' @return a matrix
+#' @export
 read.mm <- function (file.name)
 {
     if (grepl("\\.bz2?$", file.name, perl=TRUE)) {

### 2015-11-26 programs/netabc/R/write.tsv.R

@@ -5,6 +5,7 @@
 #' 
 #' @param df data.frame to write
 #' @param ... other arguments for write.table, in particular the file name
+#' @export
 write.tsv <- function (df, ...)
 {
     write.table(df, ..., col.names=TRUE, row.names=FALSE, quote=FALSE, sep="\t")

### 2015-11-26 programs/netsim/src/netabc.c

@@ -674,7 +674,7 @@ int main (int argc, char **argv)
         fclose(opts.tree_file);
     }
 
-    ladderize(tree);
+    ladderize(tree); ladderize(tree);
     scale_branches(tree, MEAN);
     k = kernel(tree, tree, opts.decay_factor, opts.rbf_variance, 1);
     SETGAN(tree, "kernel", k);

### 2015-11-26 doc/thesis/intro.tex

\subsection{Overview of phylodynamics}

In evolutionary biology, a \defn{phylogeny}, or \defn{phylogenetic tree}, is a
graphical representation of the evolutionary relationships among a group of
organisms or species. Phylogenies consist of nodes, representing organisms or
species (generally, \defn{taxa}), and edges or branches, which link taxa to
their ancestors and descendants. The \defn{tips} of the phylogeny, that is, the
nodes without any descendants, correspond to \defn{extant}, or observed, taxa.
The \defn{internal nodes} correspond to their extinct common ancestors. The
phylogeny may have a \defn{root}, which is a node with no descendants
distinguished as the most recent common ancestor of all the extant taxa. When
such a root exists, the tree is referred to as being \defn{rooted}; otherwise,
it is \defn{unrooted}. The structural arrangement of nodes and edges in the
tree is referred to as its \defn{topology}. The branches of the phylogeny may
have lengths, which can represent either the genetic distance, or the calendar
time, between ancestors and their descendants. Phylogenies whose branch lengths
are in units of calendar time are referred to as \defn{time-scaled}.

In the case of a rooted tree with branch lengths in calendar time, the internal
nodes can be mapped onto a timeline by using the tips of the phylogeny, which
usually map to the present day, as a reference point. These \defn{branching
times} correspond to the times at which speciation events occurred. In
phylogenies relating animal species, speciation most often occurs by an
\defn{allopatric} process, where a sub-population of organisms is
geographically isolated from the original population. Over time, the two
populations diverge genetically, eventually resulting in two distinct species.

Certain viruses, particularly RNA viruses such as the human immuno-deficiency
virus (HIV), evolve rapidy enough that populations can accumulate measurable
genetic changes on time scales of months or years. Transmission of the virus
from one host to another closely mirrors the allopatric speciation process.
Upon transmission, a sub-population of virions from the donor host is
geographically isolated in the recpient, followed by genetic divergence of the
two hosts' infections. Therefore, the branching time relating the two hosts'
viral populations in a time-scaled phylogeny corresponds to the time of
transmission. This correspondence can be quite rough, and is subject to some
important limitations which are discussed below. However, even taking these
limitations into account, the relationship between branching times in the
phylogeny and transmission times in the population theoretically implies that
the evolutionary history of the virus is informative about its epidemiology. In
other words, information about transmission patterns can be inferred from the
viral phylogeny, complementing more traditional survey-based epidemiological
methods. The tools and techniques for making such inferences are referred to as
\defn{phylodynamic} methods.

\subsection{Sources of error in phylodynamics}

In general, the exact topology and branch lengths of a phylogeny relating
arbitrary taxa are not known and must be estimated. Such estimation is most
often carried out using some type of genetic data sampled from each of the
taxa, either amino acid sequences or gene orders. The resulting phylogeny, also
known as a \defn{gene tree}, differs from the time-scaled \defn{species tree}
which would contain complete information about past speciation events.  In
viral phylogenetics, the gene tree is the viral phylogeny, constructed using
the genetic sequences of the sampled viruses. On the other hand, the species
tree is the \defn{transmission tree}, wherein branching times correspond
exactly to transmission times. In general, phylodynamic methods perform
inference based on the transmission tree, which must be inferred from a viral
phylogeny based on the available sequence data. The inference of the viral
phylogeny itself is not an error-free process, as accurate phylogenetic
inference is still an active area of research. Moreover, the conversion of the
phylogeny into a transmission tree is a process which has the potential to
introduce at least three types of error, described below. We use the terms
``gene tree'', ``viral phylogeny'', and ``phylogeny'' interchangably, as well
as the terms ``species tree'' and ``transmission tree''.

Firstly, the branch lengths of the gene tree are measured in genetic distance,
and must be somehow converted to units of calendar time used in the species
tree. If we assume a constant rate of evolution along the whole tree (a
\defn{strict molecular clock}), then this conversion is straightforward - we
simply divide each of the branch lengths by the rate of evolution. However, it
is well known that rates of evolution are neither constant over time nor
constant between lineages, and the general problem of time-scaling a
phylogenetic tree is still an active area of research. Phylodynamic inference
relies on the assumption of a correctly time-scaled tree, and violations of
this assumption can mislead parameter estimates.

A second limitation which has perhaps greater potential to mislead phylodynamic
inference is the potential discordance between the gene tree and the species
tree. In an allopatric speciation, some of the genetic diversity separating the
two eventual species may already have been present in the population
\emph{before} the geographic isolation occured. In that case, the branching
time in the gene tree, which indicates the time when two genetically distinct
versions of the gene, may be earlier than the branching point in the species
tree. In evolutionary biology, this phenomenon is referred to as
\defn{incomplete lineage sorting}. Furthermore, if there was substantial
variation within a population which underwent two or more speciation events in
close succession, the order in which the species diverged may be different from
the order in which their distinct genes arose in the original heterogeneous
population. In this case, the topology of the species tree would be different
from that of the gene tree.

Finally, there is the problem of rooting the tree, that is, identifying the
node corresponding to the most recent common ancestor of all the sampled
viruses. The choice of root influences the temporal order of the internal
nodes in the phylogeny. If the tips are constrained to the present day, or some
other fixed sampling time, the branch lengths in the species tree must also be
altered to accomodate the choice of root.

To our knowledge, there has been no theoretical exploration of the quantitative
effect of any of these sources of error on the accuracy of phylodynamic
inference. Most phylodynamic studies assume, implicitly or explicitly, that the
viral phylogeny provides the same information as the transmission tree, and
errors introduced by this assumption are generally ignored. Studies employing
Bayesian methods eliminate some of the uncertainty associated with constructing
the viral phylogeny, but this does not account for the error introduced in the
phylogeny-to-transmission-tree conversion process. Our work does not take a
Bayesian approach to inferring the phylogeny, and therefore is susceptible to
all the sources of error discussed here.

\subsection{Contact networks}

Compartmental models of epidemic growth and spread often assume that human
populations are homogeneously mixed, that is, any person in the population is
equally likely to transmit the infection to any other. More sophisticated
models allow for the population to be divided into groups with higher or lower
contact rates, but these still do not account for individual variation or
population structure within compartments. In reality, and especially for
sexually-transmitted infections such as HIV, this assumption of homogeneity is
not realistic. There is substantial variation in the number of contacts per
individual, and potential transmissions are limited by geographic proximity.
The aim of using contact networks is to remove these unrealistic assumptions by 
explicitly simulating each individual in a population, along with the contacts
they have which could possibly lead to a transmission.

A \defn{contact network} is a graphical representation of a population over
which an epidemic can spread. Contact networks consist of nodes and edges,
where the nodes represent people, and edges represent contacts. An edge between
two nodes indicate that transmission is possible between the two corresponding
individuals. For example, if we are considering an HIV transmission network,
the edges may represent sexual contacts or individuals sharing injection drug
paraphernalia. For an airborne pathogen such as influenza, edges may simply
indicate that the two people were in the same room. The edges in a contact
network may be undirected, indicating that all contacts could possibly transmit
to each other, or directed, indicating that transmission risk is one-way. It is
possible to simulate an undirected network using directed edges by simply
including two directed edges, one in either direction, for each single directed
edge. Therefore, we exclusively consider directed contact networks here.

### 2015-11-26 programs/netabc/R/plots.R
Added function to plot marginals

@@ -145,3 +145,42 @@ edge.color.plot <- function (tree, labels, yaml="", palette="PuBu", parent=TRUE,
     plot.phylo(tree, edge.color=edge.col, ...)
     title(paste(strwrap(as.yaml(yaml.load(yaml)), 60), collapse="\n"))
 }
+
+#' Plot marginal kernel densities
+#'
+#' @param d data.frame to plot densities for
+#' @param truth true parameter values to indicate on plots
+#' @param limits axis limits for plots
+#' @export
+marginal.plot <- function (d, truth, limits) {                                   
+    vary.cols <- colnames(d)[apply(d, 2, function (x) length(unique(x)) > 1)]       
+                                                                                 
+    if (length(vary.cols) >= 2) {                                                
+    # 2D marginals                                                               
+        combos <- combn(vary.cols, 2)                                            
+        plot.data <- apply(combos, 2, function (c) d[, c, with=FALSE])           
+        plot.aes <- apply(combos, 2, function (c) aes_string(x=c[1], y=c[2]))       
+        plot.limits <- apply(combos, 2, function (c) limits[c])                  
+        plots <- mapply(ggplot, plot.data, plot.aes, SIMPLIFY=FALSE)             
+        plots <- mapply(function (p, lim) {                                      
+            p + stat_density2d(aes(fill=..level..), geom="polygon") + theme_bw() +
+                guides(fill=FALSE) +                                             
+                geom_point(data=truth, color="black", size=8) +                  
+                geom_point(data=truth, color="white", size=6) +                  
+                xlim(lim[[1]][1], lim[[1]][2]) +                                 
+                ylim(lim[[2]][1], lim[[2]][2])                                   
+        }, plots, plot.limits, SIMPLIFY=FALSE)                                   
+        do.call(grid.arrange, c(plots, ncol=ceiling(sqrt(ncol(combos))),         
+                                top="2D marginals"))                             
+    }                                                                            
+                                                                                 
+    # 1D marginals                                                               
+    plot.aes <- lapply(vary.cols, function (c) aes_string(x=c))                  
+    plots <- mapply(ggplot, list(d), plot.aes, SIMPLIFY=FALSE)                   
+    plots <- mapply(function (p, col) {                                          
+        p + geom_density() + theme_bw() + xlim(limits[[col]][1], limits[[col]][2]) +
+            geom_vline(xintercept=truth[,col], linetype=2)                       
[…]

### 2015-12-01 doc/thesis/intro.tex
Wrote some more

@@
 usually map to the present day, as a reference point. These \defn{branching
-times} correspond to the times at which speciation events occurred. In
-phylogenies relating animal species, speciation most often occurs by an
-\defn{allopatric} process, where a sub-population of organisms is
-geographically isolated from the original population. Over time, the two
-populations diverge genetically, eventually resulting in two distinct species.
+times} correspond to the times at which speciation events
+occurred~\autocite{nee1992tempo}. In phylogenies relating animal species,
+speciation most often occurs by an \defn{allopatric} process, where a
+sub-population of organisms is geographically isolated from the original
+population [a reference to Coyne \& Orr will go here once I skim the relevant
+bits of the book]. Over time, the two populations diverge genetically,
+eventually resulting in two distinct species.
 
-Certain viruses, particularly RNA viruses such as the human immuno-deficiency
-virus (HIV), evolve rapidy enough that populations can accumulate measurable
-genetic changes on time scales of months or years. Transmission of the virus
-from one host to another closely mirrors the allopatric speciation process.
-Upon transmission, a sub-population of virions from the donor host is
+Many RNA viruses, such as the human immuno-deficiency virus (HIV), evolve
+rapidy enough that populations can accumulate measurable genetic changes on
+time scales of months or years~\autocite{drummond2003measurably}. Transmission
+of the virus from one host to another closely mirrors the allopatric speciation
+process. Upon transmission, a sub-population of virions from the donor host is
 geographically isolated in the recpient, followed by genetic divergence of the
@@
 the edges may represent sexual contacts or individuals sharing injection drug
-paraphernalia. For an airborne pathogen such as influenza, edges may simply
+paraphernalia. For an airborne pathogen such as influenza, an edge may simply
 indicate that the two people were in the same room. The edges in a contact
@@
 edge. Therefore, we exclusively consider directed contact networks here.
+
+Contact networks have been studied extensively in a variety of social,
+economic, and public health contexts. However, investigation of the effects of
+contact network structures on viral phylogenies have been much more limited.
+Existing studies have to focused on a few broad classes of contact network, and
+how phylogenies deriving from these networks have
+differed~\autocite{leventhal2012inferring, colijn2014phylogenetic,
+robinson2013dynamics}. To demonstrate our method, we focus on the same three
+contact network types as studied in~\autocite{leventhal2012inferring}, namely
+Erdos-Renyi (ER) graphs, Barabasi-Albert (BA) graphs, and Watts-Strogatz (WS)
+graphs. ER graphs, also known as random graphs, are the simplest generative
+graph model~\autocite{erdos1960evolution}. They have a single parameter, $p$,
+which gives the probability of any edge occuring in the network. BA
+graphs~\autocite{barabasi1999emergence} model a preferential attachment process
+where nodes with high degree tend to attract more connections. They have two
+parameters, $m$ and $\alpha$, and are constructed by repeatedly adding new
+nodes with $m$ outgoing edges to the network. An existing node $v$ in the
+network is chosen to be the endpoint of these new edges with probability
+proportional to $d(v)^\alpha$, where $d(v)$ is the degree of $v$. WS
+graphs~\autocite{watts1998collective}, also known as small-world graphs, have
+two parameters, $k$ and $p$. They are constructed by creating a ring lattice
+with $k$ edges per node, and then randomly rewiring each edge with probability
+$p$.
+
+\subsection{Approximate Bayesian computation}
+
+Consider a model $M$, with parameters $\theta$, which we wish to fit to some
+observed data $D$. By ``fit'', we often mean that we want to find particular
+values $\hat{\theta}$ for the parameters which optimize the likelihood of our
+data given those parameters and the model,
+\[
+  \hat{\theta} = \argmax_\theta \Pr(D | M, \theta).
+\]
+This $\hat{\theta}$ is the \defn{maximum likelihood} parameter estimate. Note
+that we are employing a common abuse of notation here, where $\Pr(\cdots)$ is
+being unsed to refer to a probabilty \emph{density} rather than a true
+probability. Alternative to maximum likelihood, we may be interested less in a
+point estimate and more in the posterior distribution of possible values of
+$\theta$ given our data, $\Pr(\theta \mid M, D)$. This will be expanded upon
+below, but for the moment, for illustrative purposes, we restrict our attention
+to the maximum likelihood problem.
+
+If the model we are fitting is sufficiently simple, it may be possible to
+calculate $\hat{\theta}$ directly, using calculus. Most models do not admit
+analytic maximum likeilhood solutions, but if the likelihood any set of
+parameters can be calculated up to a normalizing constant, then
+${\Pr(D \mid M,\theta)}$ can be optimized numerically. A wide range of
+optimization strategies exist, the choice of which to use depending on the
+complexity of the model and whether or not we have access to the gradients of
+the likelihood function with respect to each of the parameters. The majority of
+modelling problems fit into this category, and numerical optimization is
+well-developed and extremely widely used.
+
+However, there are some cases, often when the observed data is of a complex
+type, that explicitly calculating the likelihood of some observed data is
+impossible, even up to a normalizing constant. For example, suppose that we
+want to model a chess player's behaviour. We will set up a simple one-parameter
+model which describes the chess playing process.  The parameter, $a \in [0, 1]$
+indicates the player's eagerness to remove his opponent's pieces from the
+board. We can write down an algorithm for the player's behaviour under such a
+model.
+
+% TODO: don't break this over the page
+
+\begin{algorithmic}
+  \While{the game is not over}
+    \If{I can capture an opponent's piece and $\Uniform(0, 1) < a$}
+      \State{capture the piece}
+    \Else
+      \State{make any other move at random}
+    \EndIf
+  \EndWhile
+\end{algorithmic}
+
+Suppose the observed data are the ending configurations of the board, after the
+player has concluded a game against an opponent with a known value of $a$. The
+model we have designed is very simple, but it is not obvious how to calculate
+the likelihood of a particular ending configuration. Indeed, it seems that the
+only way is to enumerate every possible path the game could have taken, and
+tabulate the ending configurations of each. Clearly, this is infeasible.
+Approximate Bayesian computation was designed for situations like these, where
+exact likelihoods are not available, perhaps due to the model involving an
+algorithm or generative process, but we still want to estimate the parameters.

### 2015-12-01 doc/thesis/methods.tex
Wrote some more

\subsection{Programs}

Most of the computer programs used for this work were written in the C
programming language. Contact networks were generated using the
\software{igraph} library~\autocite{csardi2006igraph}, which was also used to
store phylogenetic trees. Hash tables, and the dynamic programming matrices for
computing the tree kernel, were stored using Judy
arrays~\autocite{baskins2004judy}. The GNU Scientific Library
(GSL)~\autocite{gough2009gnu} was used to generate random draws from
probability distributions, and to perform the bisection step in the ABC-SMC
algorithm.

\subsubsection{Epidemic simulation over a network}

I implemented method of~\autocite{gillespie1977exact} for simulating epidemics,
and the corresponding transmission trees, over static contact networks. This
method has been applied previously for the same
purpose~\autocite{robinson2013dynamics, leventhal2012inferring}, however no
existing open source implementations were available.

Let $G = (V, E)$ be a directed contact network. Each directed edge $e = (u, v)$
in the network is associated with a transmission rate $\beta_e$, which indicates
that, once $u$ becomes infected, $u$ will infect $v$ with rate $\beta_e$. In
other words, the time between $u$ becoming infected and the infection passing
from $u$ to $v$ is a random variable with distribution $\Exponential(\beta_e)$.
Note that $v$ may become infected before this time has elapsed, if $v$ has
other incoming edges. Each vertex $v$ also has a removal rate $\gamma_v$, which
is the rate at which $v$ is removed from the population after becoming
infected. Removal may correspond to death or recovery with immunity, or a
combination of both, but in our implementation recovered nodes never re-enter
the susceptible population. Borrowing a term from HIV epidemiology, a
\defn{discordant edge} in $G$ is an edge $(u, v)$ where $u$ is infected and $v$
is uninfected.

To describe the algorithm, we introduce some notation and variables. Let
$\inc(v)$ be the set of incoming edges to $v$, and $\out(v)$ be the set of
outgoing edges from $v$. Let $I$ be the set of infected nodes in the network,
$R$ be the set of removed nodes, and $S$ be the remaining susceptible nodes.
Let $D$ be the set of discordant edges in the network, $\beta$ be the 
total transmission rate over all discordant edges, and $\gamma$ be the total
removal rate of all infected nodes. That is,
\[
  \beta = \sum_{e \in D} \beta_e, \quad
  \gamma = \sum_{v \in I} \gamma_v.
\]
The variables $S$, $I$, $R$, $D$, $\beta$, and $\gamma$ are all updated as the
simulation progresses, but since these updates are very straightforward we do
not write them explicitly in the algorithm. When a new node $v$ is infected, it
is deleted from $S$ and added to $I$, any formerly discordant edges in $\in(v)$
are deleted from $D$, and edges in $\out(v)$ to nodes in $S$ are added to $D$.
If $v$ is later removed, it is deleted from $I$ and added to $R$, and any
discordant edges in $\out(v)$ are deleted from $D$. In both cases, the
variables $\beta$ and $\gamma$ are updated to reflect the changes.

\newcommand{\tip}{\mathrm{tip}}

The full algorithm is shown in Algorithm \ref{alg:nettree}. The transmission
tree $T$ is simulated along with the epidemic. We keep a map called $\tip$,
which maps infected nodes in $I$ to the tips of $T$. The variable $n$ counts
the number of nodes, both extant and internal, in $T$. The simulation continues
until either there are no discordant edges left in the network, or we reach a
user-defined cutoff of time ($t_{\max}$) or number of infections ($I_{\max}$).

\begin{algorithm}
  \label{alg:nettree}
  \caption{Simulation of an epidemic and transmission tree over a contact network}
  \begin{algorithmic}
    \State infect one node $v$ at random
    \State let $T$ be a single node labeled $1$
    \State $\tip[v] \gets 1$
    \State $n \gets 1$
    \State $t \gets 0$
    \While{$D \neq \emptyset$ and $|I| + |R| < I_{\max}$ and $t < t_{\max}$}
      \State $s \gets \min(t_{\max} - t, \Exponential(\beta + \gamma))$
      \For{$v \in \tip$}
        \State{extend the branch length of $\tip[v]$ by $s$}
      \EndFor
      \State $t \gets t + s$
      \If{$t < t_{\max}$}
        \If{$\Uniform(0, 1) < \nicefrac{\beta}{\beta + \gamma}$}
          \State choose an edge $e = (u, v)$ from $D$ with probability $\beta_e / \beta$
                 and infect $v$
          \State add nodes with labels $n+1$ and $n+2$ to $T$
          \State connect $n+1$ and $n+2$ to $\tip[v]$ in $T$, with branch lengths $0$
          \State $\tip[v] \gets n + 1$
          \State $\tip[u] \gets n + 2$
          \State $n \gets n + 2$
        \Else
          \State choose a node $v$ from $I$ with probability $\gamma_v / \gamma$
                 and remove $v$
          \State remove $v$ from $\tip$
        \EndIf
        \State update $S$, $I$, $R$, $D$, $\beta$, and $\gamma$
      \EndIf
    \EndWhile
  \end{algorithmic}
\end{algorithm}

\subsubsection{Phylogenetic kernel and normalized lineages-through-time}

The tree kernel developed in~\autocite{poon2013mapping} provides a
comprehensive similarity score between two phylogenetic trees. The kernel
computes the dot-product of two feature vectors \ldots

In addition, we implemented a modified version of the normalized
lineages-through-time statistic developed in~\autocite{janzen2015approximate},
which uses piecewise linear functions instead of step functions for the
lineages-through-time plots.

\subsubsection{Adaptive sequential Monte-Carlo approximate Bayseian computation}

\subsection{Method validation}

\subsubsection{Identification of separable parameters in kernel space}

\subsubsection{Grid search}

\subsubsection{ABC-SMC}

### 2015-12-07 doc/thesis/intro.tex
More thesis writing

@@
 exact likelihoods are not available, perhaps due to the model involving an
-algorithm or generative process, but we still want to estimate the parameters.
+algorithm or generative process.

### 2015-12-07 doc/thesis/methods.tex
More thesis writing

@@
     \State infect one node $v$ at random
-    \State let $T$ be a single node labeled $1$
+    \State let $T$ be a single node with index $1$
     \State $\tip[v] \gets 1$
@@
 comprehensive similarity score between two phylogenetic trees. The kernel
-computes the dot-product of two feature vectors \ldots
+computes the dot-product of two feature vectors, corresponding to the two
+trees, in the infinite-dimensional feature space of all possible subset trees
+with branch lengths. I implemented the fast algorithm developed
+in~\autocite{moschitti2006making}, which first enumerates all pairs of subtrees
+with the same number of leaf children, and then computes the kernel by dynamic
+programming.
 
@@
 which uses piecewise linear functions instead of step functions for the
-lineages-through-time plots.
+lineages-through-time plots. This modification is to address a potential
+inconsistency for trees of different sizes, illustrated in
+Figure~\ref{fig:nltt}.
+
+\begin{figure}
+  \caption{Comparison of original formulation of normalized
+    lineages-through-time, developed in~\autocite{janzen2015approximate}, with
+    our modified version using linear interpolation. Here, the red and blue
+    trees both have uniformly spaced branching times. Using step functions
+    (left), the nLTT of the two trees is non-zero due to the differing numbers
+    of internal branches. Using linear interpolation, the nLTT is zero (right).
+  The lines on the right graph have been offset for visibility.}
+  \includegraphics[scale=0.75]{nltt}
+  \label{fig:nltt}
+\end{figure}
 
 \subsubsection{Adaptive sequential Monte-Carlo approximate Bayseian computation}
+
+I implemented the adaptive sequential Monte-Carlo algorithm
+developed in~\autocite{del2012adaptive} for approximate Bayesian computation.
+In contrast to most existing sequential Monte-Carlo methods, this algorithm
+does not require the user to specify a sequence of decreasing tolerances to
+approach the target posterior distribution. Rather, the tolerances are computed
+adaptively at each step, starting from infinity at the first iteration. The
+algorithm may be stopped when the tolerance reaches a user-defined final value,
+or when the rate of acceptance of the MCMC kernel reaches a user-defined
+threshold. Following a heuristic applied by the
+authors~\autocite{del2012adaptive}, we used the latter stopping criterion,
+accepting the SMC approximation to the posterior when the MCMC acceptance rate
+dropped below 1.5\%.
+
+\subsubsection{Management of computational experiments with combinatorially
+expanded parameters}
+
+This reasearch involved many computational experiments performed with all
+possible combinations of several parameters. For example, to identify the best
+combination of tree kernel parameters (see \ref{subsubsec:kernel} below), we
+considered at least three values each of the parameters $\lambda$ and $\sigma$,
+corresponding to at least nine possible combinations of these.
 
@@
 \subsubsection{Identification of separable parameters in kernel space}
+\label{subsubsec:kernel}
 

### 2015-12-07 doc/thesis/figures/nltt.R

@@ -1,0 +1,30 @@
+#!/usr/bin/env Rscript
+
+library(ape)
+
+t1 <- read.tree(text="(((:1,:1):1,:1):1,:1):0;")
+t2 <- read.tree(text="((((:1,:1):1,:1):1,:1):1,:1):0;")
+
+sf1 <- stepfun(c(0, 0.5), c(0, 0.5, 1))
+sf2 <- stepfun(c(0, 1/3, 2/3), c(0, 1/3, 2/3, 1))
+
+af1 <- approxfun(c(0, 0.5, 1), c(0, 0.5, 1))
+af2 <- approxfun(c(0, 1/3, 2/3, 1), c(0, 1/3, 2/3, 1))
+
+pdf("nltt.pdf")
+par(mfrow=c(2, 2), lwd=3)
+old.mar <- par("mar")
+par(mar=c(0, 2, 4, 2) + 0.1)
+plot(t1, edge.width=3, edge.color="blue", show.tip.label=FALSE, direction="down")
+plot(t2, edge.width=3, edge.color="red", show.tip.label=FALSE, direction="down")
+
+par(mar=old.mar)
+plot(sf1, col="blue", main="original nLTT", xlab="normalized lineages",
+     ylab="normalized time", do.points=FALSE, xlim=c(0, 1), xaxs="i")
+lines(sf2, col="red", do.points=FALSE)
+
+plot(af1, col="blue", main="updated nLTT", xlab="normalized lineages",
+     ylab="normalized time", xlim=c(0, 1), xaxs="i")
+x <- seq(0, 1, 0.1)
+lines(x, af2(x) + 0.05, col="red")
+dev.off()

### 2015-12-07 programs/netabc/R/plots.R

@@ -44,7 +44,7 @@ kpca.plot <- function (kmat, color=NULL, shape=NULL, yaml=NULL)
 #' @param yaml either NULL, or a YAML-formatted string to title the plot with
 #' @return a ggplot object
 #' @export
-summary.plot <- function (data, x, y, facet.x=".", facet.y=".", group=NULL,
+summarize.plot <- function (data, x, y, facet.x=".", facet.y=".", group=NULL,
                           fun="mean", x.factor=TRUE, y.factor=FALSE,
                           group.factor=TRUE, yaml=NULL)
 {
@@ -183,4 +183,4 @@ marginal.plot <- function (d, truth, limits) {
     }, plots, vary.cols, SIMPLIFY=FALSE)                                         
     do.call(grid.arrange, c(plots, ncol=ceiling(sqrt(length(vary.cols))),        
                             top="1D marginals"))                                 
-} 
+}

### 2015-12-07 programs/netsim/src/netabc.c
Redid some older changes to ABC simulations?

@@ -670,7 +670,7 @@ int main (int argc, char **argv)
         fclose(opts.tree_file);
     }
 
-    ladderize(tree);
+    ladderize(tree); 
     scale_branches(tree, MEAN);
     k = kernel(tree, tree, opts.decay_factor, opts.rbf_variance, 1);
     SETGAN(tree, "kernel", k);

### 2015-12-07 programs/netsim/src/simulate.c
Hopefully last changes to kernel-* simulations

@@ -10,6 +10,11 @@
 #include "simulate.h"
 #define NDEBUG
 
+#ifndef INFINITY
+#include <float.h>
+#define INFINITY DBL_MAX
+#endif
+
 void print_node(const igraph_t *net, char *buf, int node, int numeric_ids);
 
 void simulate_phylogeny(igraph_t *tree, igraph_t *net, gsl_rng *rng,

### 2015-12-07 programs/netsim/src/treestats.c

@@ -4,6 +4,7 @@
 #include <igraph/igraph.h>
 #include <Judy.h>
 #include <limits.h>
+#include <math.h>
 #include <gsl/gsl_math.h>
 #include <gsl/gsl_sf_psi.h>
 

### 2015-12-07 programs/expt/expt.py
Added check for correct hostname

@@ -9,6 +9,7 @@ import logging
 import os
 import pprint
 import random
+import re
 import sqlite3
 import subprocess
 import tempfile
@@ -512,5 +513,15 @@ if __name__ == "__main__":
     loglevels = [logging.WARNING, logging.INFO, logging.DEBUG]
     logging.basicConfig(level=loglevels[args.verbose])
 
-    expt = Experiment(yaml.load(args.yaml_file))
+    spec = yaml.load(args.yaml_file)
+
+    try:
+        hostname_re = re.compile(spec["Hostname"])
+    except KeyError:
+        hostname_re = re.compile(".*")
+    if hostname_re.match(socket.gethostname()) is None:
+        logging.error("Wrong hostname: expected to be running on {}, but we are on {}".format(hostname_re.pattern, socket.gethostname()))
+        sys.exit()
+
+    expt = Experiment(spec)
     expt.run(args.qsub, args.tmpdir, args.step)

### 2015-12-09 doc/thesis/intro.tex

@@
 time, between ancestors and their descendants. Phylogenies whose branch lengths
-are in units of calendar time are referred to as \defn{time-scaled}.
+are in units of calendar time are referred to as \defn{time-scaled}. In the
+case of a rooted tree with branch lengths in calendar time, the internal nodes
+can be mapped onto a timeline by using the tips of the phylogeny, which usually
+map to the present day, as a reference point. These \defn{branching times}
+correspond to the times at which speciation events
+occurred~\autocite{nee1992tempo}. These concepts are illustrated in
+Figure~\ref{fig:phylogeny}.
 
-In the case of a rooted tree with branch lengths in calendar time, the internal
-nodes can be mapped onto a timeline by using the tips of the phylogeny, which
-usually map to the present day, as a reference point. These \defn{branching
-times} correspond to the times at which speciation events
-occurred~\autocite{nee1992tempo}. In phylogenies relating animal species,
-speciation most often occurs by an \defn{allopatric} process, where a
-sub-population of organisms is geographically isolated from the original
-population [a reference to Coyne \& Orr will go here once I skim the relevant
-bits of the book]. Over time, the two populations diverge genetically,
-eventually resulting in two distinct species.
+\begin{figure}
+  \label{fig:phylogeny}
+  \includegraphics{phylogeny}
+  \caption{Cartoon illustration of a rooted, time-scaled phylogeny relating
+  several domestic animals. The tips of the phylogeny, which represent extant
+  taxa, are placed at the present day on the time axis. Internal nodes,
+  representing extinct common ancestors to the extant taxa, fall in the past,
+  at branching times corresponding to speciation events. The topology of the
+  tree indicates that cats and dogs are the most closely related pair of
+  species, whereas fish is most distantly related to any other node in the
+  tree.}
+\end{figure}
 
-Many RNA viruses, such as the human immuno-deficiency virus (HIV), evolve
-rapidy enough that populations can accumulate measurable genetic changes on
-time scales of months or years~\autocite{drummond2003measurably}. Transmission
-of the virus from one host to another closely mirrors the allopatric speciation
-process. Upon transmission, a sub-population of virions from the donor host is
-geographically isolated in the recpient, followed by genetic divergence of the
-two hosts' infections. Therefore, the branching time relating the two hosts'
-viral populations in a time-scaled phylogeny corresponds to the time of
-transmission. This correspondence can be quite rough, and is subject to some
-important limitations which are discussed below. However, even taking these
-limitations into account, the relationship between branching times in the
-phylogeny and transmission times in the population theoretically implies that
-the evolutionary history of the virus is informative about its epidemiology. In
-other words, information about transmission patterns can be inferred from the
-viral phylogeny, complementing more traditional survey-based epidemiological
-methods. The tools and techniques for making such inferences are referred to as
-\defn{phylodynamic} methods.
+Various population genetic processes cause the phylogenies that we observe to
+take the particular shapes that they do. Selection, due to environmental
+pressures, allows certain lineages to persist while driving others to
+extinction. Speciation, which causes branching, often occurs by an
+\defn{allopatric} process, where a sub-population of organisms is isolated from
+the original population by a geographic barrier~\autocite{coyne2004speciation}.
+Over time, the two populations diverge genetically, eventually resulting in two
+distinct species. Barriers are not necessary for speciation, however.
+Diversifying selection, which favours either of two ``extreme'' phenotypes, can
+cause \defn{sympatric} speciation without a geographic
+separation~\autocite{coyne2004speciation}.
+
+Now, consider a phylogeny not of macro-organisms, but of RNA viruses. Many of
+these, such as the human immuno-deficiency virus (HIV), evolve rapidy enough
+that populations can accumulate measurable genetic changes on time scales of
+months or years~\autocite{drummond2003measurably}. Because viruses cannot exist
+for long outside their hosts, the host replaces the outside world as the
+``environment'' in which evolution takes place. In this light, striking
+parallels can be drawn between immunological and epidemiological processes at
+the host level, and the evolutionary processes studied in population
+genetics~\autocite{grenfell2004unifying}. For example, both the host immune
+response~\autocite{??} and the presence of anti-retroviral drugs~\autocite{??}
+exert directional selection on the viral population within a single host.
+Transmission, where a sub-population of virions from the donor host is
+``geographically'' isolated in the recpient, closely mirrors the allopatric
+speciation process and causes a branching point in the viral phylogeny
+(although this may not correspond exactly to the time of transmission, as
+discussed in the next section). Consequently, any factor which influences
+transmission dynamics, such as the virus' mode of transmission or the
+population structure of its hosts, in turn affects the shape of the phylogeny.
+The study of this interaction between host-level factors and viral phylogies is
+known as \defn{phylodynamics}~\autocite{grenfell2004unifying}.
+
+In particular, if a time-scaled phylogeny is constructed where all viruses from
+the same host are collapsed into a single taxon, we might expect the branching
+time relating the two hosts to correspond to the time of transmission. Of
+course, this will only be the case if the two hosts' infections indeed diverged
+allopatrically at the time of transmission, and not sympatrically within the
+donor host at some previous time. This limitation, and others, in interpreting
+the branching times of the viral phylogeny as transmission times, are discussed
+in the next section. Nevertheless, even taking these limitations into account,
+the relationship between branching times in the phylogeny and transmission
+times in the population theoretically implies that the evolutionary history of
+the virus is informative about the biology and epidemiology of its hosts. In
+other words, information about, for example, transmission patterns, might be
+inferred from the viral phylogeny. Phylodynamic methods are intended to
+complement more traditional approaches to gathering epidemiological data, such
+as population surveys.
 

### 2015-12-17 doc/thesis/intro.tex
Reading and writing, plus cross-validation for ordinary lm

@@
 
-In evolutionary biology, a \defn{phylogeny}, or \defn{phylogenetic tree}, is a
-graphical representation of the evolutionary relationships among a group of
-organisms or species. Phylogenies consist of nodes, representing organisms or
-species (generally, \defn{taxa}), and edges or branches, which link taxa to
-their ancestors and descendants. The \defn{tips} of the phylogeny, that is, the
-nodes without any descendants, correspond to \defn{extant}, or observed, taxa.
-The \defn{internal nodes} correspond to their extinct common ancestors. The
-phylogeny may have a \defn{root}, which is a node with no descendants
-distinguished as the most recent common ancestor of all the extant taxa. When
-such a root exists, the tree is referred to as being \defn{rooted}; otherwise,
-it is \defn{unrooted}. The structural arrangement of nodes and edges in the
-tree is referred to as its \defn{topology}. The branches of the phylogeny may
-have lengths, which can represent either the genetic distance, or the calendar
-time, between ancestors and their descendants. Phylogenies whose branch lengths
-are in units of calendar time are referred to as \defn{time-scaled}. In the
-case of a rooted tree with branch lengths in calendar time, the internal nodes
-can be mapped onto a timeline by using the tips of the phylogeny, which usually
-map to the present day, as a reference point. These \defn{branching times}
-correspond to the times at which speciation events
-occurred~\autocite{nee1992tempo}. These concepts are illustrated in
-Figure~\ref{fig:phylogeny}.
+\subsubsection{Species trees}
 
-\begin{figure}
-  \label{fig:phylogeny}
-  \includegraphics{phylogeny}
-  \caption{Cartoon illustration of a rooted, time-scaled phylogeny relating
-  several domestic animals. The tips of the phylogeny, which represent extant
-  taxa, are placed at the present day on the time axis. Internal nodes,
-  representing extinct common ancestors to the extant taxa, fall in the past,
-  at branching times corresponding to speciation events. The topology of the
-  tree indicates that cats and dogs are the most closely related pair of
-  species, whereas fish is most distantly related to any other node in the
-  tree.}
+In evolutionary biology, the evolutionary relationships among a group of
+organisms or species (generally, \defn{taxa}) are often represented graphically
+in the form of a \defn{species tree}. These trees are often called
+\defn{phylogenies}, but to avoid ambiguity, we will reserve that term for gene
+trees as discussed below. Species trees consist of nodes, representing taxa,
+and edges or branches, which connect taxa to their evolutionary ancestors and
+descendants. The \defn{tips} of a species tree, that is, the nodes without any
+descendants, correspond to \defn{extant}, or observed, taxa, while the
+\defn{internal nodes} correspond to their extinct common ancestors. Species
+trees may have a \defn{root}, which is a node with no descendants distinguished
+as the most recent common ancestor of all the extant taxa. When such a root
+exists, the tree is referred to as being \defn{rooted}; otherwise, it is
+\defn{unrooted}. The structural arrangement of nodes and edges in the tree is
+referred to as its \defn{topology}. The branches of the tree may have
+associated lengths, which, represent the amount of calendar time elapsed
+between ancestors and their descendants. When these branch lengths are present,
+the internal nodes can be mapped onto a timeline by using the tips of the tree,
+which usually map to the present day, as a reference point. These
+\defn{branching times} correspond to the historical times at which speciation
+events occurred~\autocite{nee1992tempo}. These concepts are illustrated in
+Figure~\ref{fig:speciestree}.
+
+\begin{figure}[h]
+  \centering
+  \label{fig:speciestree}
+  \includegraphics{speciestree}
+  \caption{Illustration of a rooted species tree. The tips of the tree, which
+    represent extant taxa, are placed at the present day on the time axis.
+    Internal nodes, representing extinct common ancestors to the extant taxa,
+    fall in the past, at branching times corresponding to speciation events.
+    The topology of the tree indicates that cats and dogs are the most closely
+    related pair of species, whereas fish is most distantly related to any
+    other node in the tree.}
 \end{figure}
@@
 genetics~\autocite{grenfell2004unifying}. For example, both the host immune
-response~\autocite{??} and the presence of anti-retroviral drugs~\autocite{??}
-exert directional selection on the viral population within a single host.
-Transmission, where a sub-population of virions from the donor host is
-``geographically'' isolated in the recpient, closely mirrors the allopatric
-speciation process and causes a branching point in the viral phylogeny
-(although this may not correspond exactly to the time of transmission, as
-discussed in the next section). Consequently, any factor which influences
-transmission dynamics, such as the virus' mode of transmission or the
-population structure of its hosts, in turn affects the shape of the phylogeny.
-The study of this interaction between host-level factors and viral phylogies is
-known as \defn{phylodynamics}~\autocite{grenfell2004unifying}.
+response~\autocite{mcmichael2001cellular} and the presence of anti-retroviral
+drugs~\autocite{??} exert directional selection on the viral population within
+a single host. Transmission, where a sub-population of virions from the donor
+host is ``geographically'' isolated in the recpient, closely mirrors the
+allopatric speciation process and causes a branching point in the viral
+phylogeny (although this may not correspond exactly to the time of
+transmission, as discussed in the next section). Consequently, any factor which
+influences transmission dynamics, such as the virus' mode of transmission or
+the population structure of its hosts, in turn affects the shape of the
+phylogeny. The study of this interaction between host-level factors and viral
+phylogies is known as \defn{phylodynamics}~\autocite{grenfell2004unifying}.
 

### 2015-12-17 doc/thesis/methods.tex
Reading and writing, plus cross-validation for ordinary lm

@@
 
-\subsubsection{Management of computational experiments with combinatorially
-expanded parameters}
-
-This reasearch involved many computational experiments performed with all
-possible combinations of several parameters. For example, to identify the best
-combination of tree kernel parameters (see \ref{subsubsec:kernel} below), we
-considered at least three values each of the parameters $\lambda$ and $\sigma$,
-corresponding to at least nine possible combinations of these.
-
 \subsection{Method validation}
@@
 
-\subsubsection{ABC-SMC}
+\subsubsection{Approximate Bayesian computation}
+
+\subsection{Applications}
+
+\subsubsection{HIV in British Columubia}

### 2015-12-17 programs/netabc/R/cv.R

@@ -106,3 +106,29 @@ ksvm.cv <- function (kmat, y, n.cv=1000,
     do.cv(kmat, y, get.train, get.test, fit.model, predict, n.cv=n.cv,
           stats=stats, show.progress=show.progress, nthread=nthread)
 }
+
+#' Perform replicate 2-fold cross-validations on a linear model classifier.
+#'
+#' Returns a data.frame with the indicated statistics for each individual
+#' cross-validation. Currently the only supported statistic is "rsquared".
+#'
+#' @param x predictors
+#' @param y labels for data
+#' @param stats statistics to calculate
+#' @param show.progress whether to display a progress bar
+#' @param nthread number of threads to use
+#' @return A data.frame with the indicated statistics for each CV
+#' @seealso \url{http://stackoverflow.com/questions/1753299}
+#' @seealso \code{\link{kernlab::ksvm}}
+#' @export
+lm.cv <- function (x, y, n.cv=1000, stats="rsquared", show.progress=TRUE,
+                   nthread=1)
+{
+    get.train <- function(x, holdout) x[-holdout,,drop=FALSE]
+    get.test <- function(x, holdout, m) x[holdout,,drop=FALSE]
+    fit.model <- function(x, train.y) lm(train.y~pred, data=x)
+    predict <- function(m, x.test) predict.lm(m, newdata=x.test)
+    x.df <- data.frame(pred=x)
+    do.cv(x.df, y, get.train, get.test, fit.model, predict, n.cv=n.cv,
+          stats=stats, show.progress=show.progress, nthread=nthread)
+}

### 2015-12-17 programs/netsim/tests/check_smc.c

@@ -206,11 +206,9 @@ Suite *smc_suite(void)
     s = suite_create("smc");
 
     tc_smc = tcase_create("Core");
-    /*
     tcase_add_test(tc_smc, test_smc_toy);
     tcase_add_test(tc_smc, test_smc_toy_steps);
     tcase_add_test(tc_smc, test_smc_distance);
-    */
     tcase_set_timeout(tc_smc, 60);
     suite_add_tcase(s, tc_smc);
 

### 2016-01-01 doc/thesis/intro.tex
More thesis-ing

--- 2015-12-07 doc/thesis/intro.tex
@@
 
+\subsubsection{Phylogenetic trees}
+
 In evolutionary biology, a \defn{phylogeny}, or \defn{phylogenetic tree}, is a
-graphical representation of the evolutionary relationships among a group of
-organisms or species. Phylogenies consist of nodes, representing organisms or
-species (generally, \defn{taxa}), and edges or branches, which link taxa to
-their ancestors and descendants. The \defn{tips} of the phylogeny, that is, the
-nodes without any descendants, correspond to \defn{extant}, or observed, taxa.
-The \defn{internal nodes} correspond to their extinct common ancestors. The
-phylogeny may have a \defn{root}, which is a node with no descendants
-distinguished as the most recent common ancestor of all the extant taxa. When
-such a root exists, the tree is referred to as being \defn{rooted}; otherwise,
-it is \defn{unrooted}. The structural arrangement of nodes and edges in the
-tree is referred to as its \defn{topology}. The branches of the phylogeny may
-have lengths, which can represent either the genetic distance, or the calendar
-time, between ancestors and their descendants. Phylogenies whose branch lengths
-are in units of calendar time are referred to as \defn{time-scaled}.
+graphical representation of the the evolutionary relationships among a group of
+organisms or species (generally, \defn{taxa}). Phylogenies consist of nodes,
+representing taxa, and edges or branches, which connect taxa to their
+evolutionary ancestors and descendants. The \defn{tips} of a phylogeny, that
+is, the nodes without any descendants, correspond to \defn{extant}, or
+observed, taxa, while the \defn{internal nodes} correspond to their extinct
+common ancestors. Phylogenies may have a \defn{root}, which is a node with no
+descendants distinguished as the most recent common ancestor of all the extant
+taxa. When such a root exists, the tree is referred to as being \defn{rooted};
+otherwise, it is \defn{unrooted}. The structural arrangement of nodes and edges
+in the tree is referred to as its \defn{topology}. The branches of the tree may
+have associated lengths, representing either calendar time or genetic distance
+between ancestors and their descendants.
 
-In the case of a rooted tree with branch lengths in calendar time, the internal
-nodes can be mapped onto a timeline by using the tips of the phylogeny, which
-usually map to the present day, as a reference point. These \defn{branching
-times} correspond to the times at which speciation events
-occurred~\autocite{nee1992tempo}. In phylogenies relating animal species,
-speciation most often occurs by an \defn{allopatric} process, where a
-sub-population of organisms is geographically isolated from the original
-population [a reference to Coyne \& Orr will go here once I skim the relevant
-bits of the book]. Over time, the two populations diverge genetically,
-eventually resulting in two distinct species.
+A phylogeny with branch lengths in calendar time units is often referred to as
+\defn{time-scaled}. In a time-scaled phylogeny, the internal nodes can be
+mapped onto a timeline by using the tips of the tree, which usually map to the
+present day, as a reference point~\autocite{nee1992tempo}. The corresponding
+points on the timeline are called \defn{branching times}. These concepts are
+illustrated in Figure~\ref{fig:speciestree}.
 
-Many RNA viruses, such as the human immuno-deficiency virus (HIV), evolve
-rapidy enough that populations can accumulate measurable genetic changes on
-time scales of months or years~\autocite{drummond2003measurably}. Transmission
-of the virus from one host to another closely mirrors the allopatric speciation
-process. Upon transmission, a sub-population of virions from the donor host is
-geographically isolated in the recpient, followed by genetic divergence of the
-two hosts' infections. Therefore, the branching time relating the two hosts'
-viral populations in a time-scaled phylogeny corresponds to the time of
-transmission. This correspondence can be quite rough, and is subject to some
-important limitations which are discussed below. However, even taking these
-limitations into account, the relationship between branching times in the
-phylogeny and transmission times in the population theoretically implies that
-the evolutionary history of the virus is informative about its epidemiology. In
-other words, information about transmission patterns can be inferred from the
-viral phylogeny, complementing more traditional survey-based epidemiological
-methods. The tools and techniques for making such inferences are referred to as
-\defn{phylodynamic} methods.
+\begin{figure}[ht]
+  \centering
+  \label{fig:speciestree}
+  \includegraphics{speciestree}
+  \caption[Illustration of a rooted, time-scaled phylogeny]{Illustration of a
+    rooted, time-scaled phylogeny. The tips of the tree, which represent extant
+    taxa, are placed at the present day on the
+    time axis. Internal nodes, representing extinct common ancestors to the
+    extant taxa, fall in the past. The topology of the tree indicates that cats
+    and dogs are the most closely related pair of species, whereas fish is most
+    distantly related to any other node in the tree.}
+\end{figure}
 
-\subsection{Sources of error in phylodynamics}
+\subsubsection{Species trees and transmission trees}
 
-In general, the exact topology and branch lengths of a phylogeny relating
-arbitrary taxa are not known and must be estimated. Such estimation is most
-often carried out using some type of genetic data sampled from each of the
-taxa, either amino acid sequences or gene orders. The resulting phylogeny, also
-known as a \defn{gene tree}, differs from the time-scaled \defn{species tree}
-which would contain complete information about past speciation events. In
-viral phylogenetics, the gene tree is the viral phylogeny, constructed using
-the genetic sequences of the sampled viruses. On the other hand, the species
-tree is the \defn{transmission tree}, wherein branching times correspond
-exactly to transmission times. In general, phylodynamic methods perform
-inference based on the transmission tree, which must be inferred from a viral
-phylogeny based on the available sequence data. The inference of the viral
-phylogeny itself is not an error-free process, as accurate phylogenetic
-inference is still an active area of research. Moreover, the conversion of the
-phylogeny into a transmission tree is a process which has the potential to
-introduce at least three types of error, described below. We use the terms
-``gene tree'', ``viral phylogeny'', and ``phylogeny'' interchangably, as well
-as the terms ``species tree'' and ``transmission tree''.
+A \defn{species tree} is a particular type of phylogeny in which the taxa are
+species, and the branching times correspond to historical speciation events. In
+general, we do not have access to the true tree relating a particular group of
+extant species, as this would imply certain knowledge of speciation events in
+the distant past. Therefore, all species trees considered by researchers are
+estimates, based on the genetic or phenotypic similarity of the extant taxa,
+the fossil record, or both. We discuss some of the challenges associated with
+this estimation in subsection \ref{subsec:treeconv} below. However, for the
+moment, we ignore these complications and suppose that we know the true species
+tree.
 
-Firstly, the branch lengths of the gene tree are measured in genetic distance,
-and must be somehow converted to units of calendar time used in the species
-tree. If we assume a constant rate of evolution along the whole tree (a
-\defn{strict molecular clock}), then this conversion is straightforward - we
-simply divide each of the branch lengths by the rate of evolution. However, it
-is well known that rates of evolution are neither constant over time nor
-constant between lineages, and the general problem of time-scaling a
-phylogenetic tree is still an active area of research. Phylodynamic inference
-relies on the assumption of a correctly time-scaled tree, and violations of
-this assumption can mislead parameter estimates.
+The process of speciation, which causes branching in a species tree, often
+occurs by an \defn{allopatric} process, where a sub-population of organisms is
+isolated from the original population by a geographic
+barrier~\autocite{coyne2004speciation}. Over time, the two populations diverge
+genetically, eventually resulting in two distinct species.
 
-A second limitation which has perhaps greater potential to mislead phylodynamic
-inference is the potential discordance between the gene tree and the species
-tree. In an allopatric speciation, some of the genetic diversity separating the
-two eventual species may already have been present in the population
-\emph{before} the geographic isolation occured. In that case, the branching
-time in the gene tree, which indicates the time when two genetically distinct
-versions of the gene, may be earlier than the branching point in the species
-tree. In evolutionary biology, this phenomenon is referred to as
-\defn{incomplete lineage sorting}. Furthermore, if there was substantial
-variation within a population which underwent two or more speciation events in
-close succession, the order in which the species diverged may be different from
-the order in which their distinct genes arose in the original heterogeneous
-population. In this case, the topology of the species tree would be different
-from that of the gene tree.
+\subsubsection{Gene trees and viral phylogenies}
 
-Finally, there is the problem of rooting the tree, that is, identifying the
-node corresponding to the most recent common ancestor of all the sampled
-viruses. The choice of root influences the temporal order of the internal
-nodes in the phylogeny. If the tips are constrained to the present day, or some
-other fixed sampling time, the branch lengths in the species tree must also be
-altered to accomodate the choice of root.
+\subsubsection{Using gene trees to estimate species trees}
 
-To our knowledge, there has been no theoretical exploration of the quantitative
-effect of any of these sources of error on the accuracy of phylodynamic
-inference. Most phylodynamic studies assume, implicitly or explicitly, that the
-viral phylogeny provides the same information as the transmission tree, and
-errors introduced by this assumption are generally ignored. Studies employing
-Bayesian methods eliminate some of the uncertainty associated with constructing
-the viral phylogeny, but this does not account for the error introduced in the
-phylogeny-to-transmission-tree conversion process. Our work does not take a
-Bayesian approach to inferring the phylogeny, and therefore is susceptible to
-all the sources of error discussed here.
+\subsubsection{Tree shapes}
 
@@
 
-Compartmental models of epidemic growth and spread often assume that human
-populations are homogeneously mixed, that is, any person in the population is
-equally likely to transmit the infection to any other. More sophisticated
-models allow for the population to be divided into groups with higher or lower
-contact rates, but these still do not account for individual variation or
-population structure within compartments. In reality, and especially for
-sexually-transmitted infections such as HIV, this assumption of homogeneity is
-not realistic. There is substantial variation in the number of contacts per
-individual, and potential transmissions are limited by geographic proximity.
-The aim of using contact networks is to remove these unrealistic assumptions by 
-explicitly simulating each individual in a population, along with the contacts
-they have which could possibly lead to a transmission.
+\subsubsection{Overview and epidemiology}
 
-A \defn{contact network} is a graphical representation of a population over
-which an epidemic can spread. Contact networks consist of nodes and edges,
-where the nodes represent people, and edges represent contacts. An edge between
-two nodes indicate that transmission is possible between the two corresponding
-individuals. For example, if we are considering an HIV transmission network,
-the edges may represent sexual contacts or individuals sharing injection drug
-paraphernalia. For an airborne pathogen such as influenza, an edge may simply
-indicate that the two people were in the same room. The edges in a contact
-network may be undirected, indicating that all contacts could possibly transmit
-to each other, or directed, indicating that transmission risk is one-way. It is
-possible to simulate an undirected network using directed edges by simply
-including two directed edges, one in either direction, for each single directed
-edge. Therefore, we exclusively consider directed contact networks here.
+Epidemics spread through populations of hosts through \defn{contacts} between
+those hosts. The definition of contact depends on the mode of transmission of
+the pathogen in question. For an airborne pathogen like influenza, a contact
+may be simple physical proximity, while for a sexually transmitted pathogen
+like HIV, contacts would be sexual partnerships. A \defn{contact network} is a
+graphical representation of a host population and the contacts among its
+members. The \defn{nodes} in the network represent hosts, and \defn{edges}
+represent contacts between them.
 
-Contact networks have been studied extensively in a variety of social,
-economic, and public health contexts. However, investigation of the effects of
-contact network structures on viral phylogenies have been much more limited.
-Existing studies have to focused on a few broad classes of contact network, and
-how phylogenies deriving from these networks have
-differed~\autocite{leventhal2012inferring, colijn2014phylogenetic,
-robinson2013dynamics}. To demonstrate our method, we focus on the same three
-contact network types as studied in~\autocite{leventhal2012inferring}, namely
-Erdos-Renyi (ER) graphs, Barabasi-Albert (BA) graphs, and Watts-Strogatz (WS)
-graphs. ER graphs, also known as random graphs, are the simplest generative
-graph model~\autocite{erdos1960evolution}. They have a single parameter, $p$,
-which gives the probability of any edge occuring in the network. BA
-graphs~\autocite{barabasi1999emergence} model a preferential attachment process
-where nodes with high degree tend to attract more connections. They have two
-parameters, $m$ and $\alpha$, and are constructed by repeatedly adding new
-nodes with $m$ outgoing edges to the network. An existing node $v$ in the
-network is chosen to be the endpoint of these new edges with probability
-proportional to $d(v)^\alpha$, where $d(v)$ is the degree of $v$. WS
-graphs~\autocite{watts1998collective}, also known as small-world graphs, have
-two parameters, $k$ and $p$. They are constructed by creating a ring lattice
-with $k$ edges per node, and then randomly rewiring each edge with probability
-$p$.
+Edges in a contact networks may be \defn{directed}, representing one-way
+transmission risk, or \defn{undirected}, representing symmetric transmission
+risk. For example, a network for an airborne epidemic would use undirected
+edges, because the same physical proximity is required for a host to infect or
+to become infected. However, a blood-borne infection spread through
+transfusions would use undirected edges, since the donor has no chance of
+transmitting to the recipient. Directed edges are also useful when the
+transmission risk is not equal between the hosts, such as with HIV
+transmission, where acting as the receptive partner carries a higher risk of
+infection than acting as the insertive partner. In this case, a contact could
+be represented by two directed edges, one in each direction between the two
+hosts, with the edges annotated by what kind of risk they imply. In fact, it is
+possible to represent an undirected edge by two symmetric directed edges. For
+this reason, we consider only contact networks with directed edges in the
+sequel. A directed contact network is shown in Figure \ref{fig:contactnet}
+(left).
 
-\subsection{Approximate Bayesian computation}
+The path an epidemic takes through a contact network determines the topology of
+the transmission tree relating the infected hosts. The initially infected node
+who introduces the epidemic becomes the root of the tree. Each time a
+transmission occurs, the lineage corresponding to the donor host in the tree
+splits into two, representing the recipient lineage and the continuation of the
+donor lineage. This correspondence is illustrated in figure
+\ref{fig:contactnet}. It's important to note that, although the order and
+timing of transmissions determines the tree topology uniquely, the converse
+does not hold. That is, there are generally several orders of infection which
+could lead to the same topology, since the labels on the internal nodes of the
+tree are not available to the researcher.
 
-Consider a model $M$, with parameters $\theta$, which we wish to fit to some
-observed data $D$. By ``fit'', we often mean that we want to find particular
-values $\hat{\theta}$ for the parameters which optimize the likelihood of our
-data given those parameters and the model,
-\[
-  \hat{\theta} = \argmax_\theta \Pr(D | M, \theta).
-\]
-This $\hat{\theta}$ is the \defn{maximum likelihood} parameter estimate. Note
-that we are employing a common abuse of notation here, where $\Pr(\cdots)$ is
-being unsed to refer to a probabilty \emph{density} rather than a true
-probability. Alternative to maximum likelihood, we may be interested less in a
-point estimate and more in the posterior distribution of possible values of
-$\theta$ given our data, $\Pr(\theta \mid M, D)$. This will be expanded upon
-below, but for the moment, for illustrative purposes, we restrict our attention
-to the maximum likelihood problem.
+\begin{figure}[ht]
+  \centering
+  \label{fig:contactnet}
+  \includegraphics{contactnet}
+  \caption[Illustration of a epidemic spread over a contact network]
+  {Illustration of epidemic spread over a contact network. On the left, a
+   contact network with five hosts, labelled A through E. Directed edges
+   indicate six symmetric contacts among the hosts. Coloured, bolded edges
+   represent transmissions. The epidemic began at node A, who transmitted to
+   nodes B and C. Node C further transmitted to node D, and node E was not
+   infected. On the right, the transmission tree topology corresponding to this
+   scenario.}
+\end{figure}
 
-If the model we are fitting is sufficiently simple, it may be possible to
-calculate $\hat{\theta}$ directly, using calculus. Most models do not admit
-analytic maximum likeilhood solutions, but if the likelihood any set of
-parameters can be calculated up to a normalizing constant, then
-${\Pr(D \mid M,\theta)}$ can be optimized numerically. A wide range of
-optimization strategies exist, the choice of which to use depending on the
-complexity of the model and whether or not we have access to the gradients of
-the likelihood function with respect to each of the parameters. The majority of
-modelling problems fit into this category, and numerical optimization is
-well-developed and extremely widely used.
+\subsubsection{Generative models}
 
-However, there are some cases, often when the observed data is of a complex
-type, that explicitly calculating the likelihood of some observed data is
-impossible, even up to a normalizing constant. For example, suppose that we
-want to model a chess player's behaviour. We will set up a simple one-parameter
-model which describes the chess playing process.  The parameter, $a \in [0, 1]$
-indicates the player's eagerness to remove his opponent's pieces from the
-board. We can write down an algorithm for the player's behaviour under such a
-model.
+%Now, consider a phylogeny not of macro-organisms, but of RNA viruses. Many of
+%these, such as the human immuno-deficiency virus (HIV), evolve rapidy enough
+%that populations can accumulate measurable genetic changes on time scales of
+%months or years~\autocite{drummond2003measurably}. Because viruses cannot exist
+%for long outside their hosts, the host replaces the outside world as the
+%``environment'' in which evolution takes place. 
+%
+%In this light, striking parallels can be drawn between immunological and
+%epidemiological processes at the host level, and the evolutionary processes
+%studied in population genetics~\autocite{grenfell2004unifying}. For example,
+%both the host immune response~\autocite{mcmichael2001cellular} and the presence
+%of anti-retroviral drugs~\autocite{??} exert directional selection on the viral
+%population within a single host. Transmission, where a sub-population of
+%virions from the donor host is ``geographically'' isolated in the recpient,
+%closely mirrors the allopatric speciation process and causes a branching point
+%in the viral phylogeny (although this may not correspond exactly to the time of
+%transmission, as discussed in the next section). Consequently, any factor which
+%influences transmission dynamics, such as the virus' mode of transmission or
+%the population structure of its hosts, in turn affects the shape of the
+%phylogeny. The study of this interaction between host-level factors and viral
+%phylogies is known as \defn{phylodynamics}~\autocite{grenfell2004unifying}. 
+%
+%In particular, if a time-scaled phylogeny is constructed where all viruses from
+%the same host are collapsed into a single taxon, we might expect the branching
+%time relating the two hosts to correspond to the time of transmission. Of
+%course, this will only be the case if the two hosts' infections indeed diverged
+%allopatrically at the time of transmission, and not sympatrically within the
+%donor host at some previous time. This limitation, and others, in interpreting
+%the branching times of the viral phylogeny as transmission times, are discussed
+%in the next section. Nevertheless, even taking these limitations into account,
+%the relationship between branching times in the phylogeny and transmission
+%times in the population theoretically implies that the evolutionary history of
+%the virus is informative about the biology and epidemiology of its hosts. In
+%other words, information about, for example, transmission patterns, might be
+%inferred from the viral phylogeny. Phylodynamic methods are intended to
+%complement more traditional approaches to gathering epidemiological data, such
+%as population surveys.
+%
+%
+%\subsection{Sources of error in phylodynamics}
+%\label{subsec:treeconv}
+%
+%In general, the exact topology and branch lengths of a phylogeny relating
+%arbitrary taxa are not known and must be estimated. Such estimation is most
+%often carried out using some type of genetic data sampled from each of the
+%taxa, either amino acid sequences or gene orders. The resulting phylogeny, also
+%known as a \defn{gene tree}, differs from the time-scaled \defn{species tree}
+%which would contain complete information about past speciation events. In
+%viral phylogenetics, the gene tree is the viral phylogeny, constructed using
+%the genetic sequences of the sampled viruses. On the other hand, the species
+%tree is the \defn{transmission tree}, wherein branching times correspond
+%exactly to transmission times. In general, phylodynamic methods perform
+%inference based on the transmission tree, which must be inferred from a viral
+%phylogeny based on the available sequence data. The inference of the viral
+%phylogeny itself is not an error-free process, as accurate phylogenetic
+%inference is still an active area of research. Moreover, the conversion of the
+%phylogeny into a transmission tree is a process which has the potential to
+%introduce at least three types of error, described below. We use the terms
+%``gene tree'', ``viral phylogeny'', and ``phylogeny'' interchangably, as well
+%as the terms ``species tree'' and ``transmission tree''.
+%
+%Firstly, the branch lengths of the gene tree are measured in genetic distance,
+%and must be somehow converted to units of calendar time used in the species
+%tree. If we assume a constant rate of evolution along the whole tree (a
+%\defn{strict molecular clock}), then this conversion is straightforward - we
+%simply divide each of the branch lengths by the rate of evolution. However, it
+%is well known that rates of evolution are neither constant over time nor
+%constant between lineages, and the general problem of time-scaling a
+%phylogenetic tree is still an active area of research. Phylodynamic inference
+%relies on the assumption of a correctly time-scaled tree, and violations of
+%this assumption can mislead parameter estimates.
+%
+%A second limitation which has perhaps greater potential to mislead phylodynamic
+%inference is the potential discordance between the gene tree and the species
+%tree. In an allopatric speciation, some of the genetic diversity separating the
+%two eventual species may already have been present in the population
+%\emph{before} the geographic isolation occured. In that case, the branching
+%time in the gene tree, which indicates the time when two genetically distinct
+%versions of the gene, may be earlier than the branching point in the species
+%tree. In evolutionary biology, this phenomenon is referred to as
+%\defn{incomplete lineage sorting}. Furthermore, if there was substantial
+%variation within a population which underwent two or more speciation events in
+%close succession, the order in which the species diverged may be different from
+%the order in which their distinct genes arose in the original heterogeneous
+%population. In this case, the topology of the species tree would be different
+%from that of the gene tree. 
+%
+%Finally, there is the problem of rooting the tree, that is, identifying the
+%node corresponding to the most recent common ancestor of all the sampled
+%viruses. The choice of root influences the temporal order of the internal
+%nodes in the phylogeny. If the tips are constrained to the present day, or some
+%other fixed sampling time, the branch lengths in the species tree must also be
+%altered to accomodate the choice of root. 
+%
+%To our knowledge, there has been no theoretical exploration of the quantitative
+%effect of any of these sources of error on the accuracy of phylodynamic
+%inference. Most phylodynamic studies assume, implicitly or explicitly, that the
+%viral phylogeny provides the same information as the transmission tree, and
+%errors introduced by this assumption are generally ignored. Studies employing
+%Bayesian methods eliminate some of the uncertainty associated with constructing
+%the viral phylogeny, but this does not account for the error introduced in the
+%phylogeny-to-transmission-tree conversion process. Our work does not take a
+%Bayesian approach to inferring the phylogeny, and therefore is susceptible to
+%all the sources of error discussed here.
+%
+%\subsection{Contact networks}
+%
+%Compartmental models of epidemic growth and spread often assume that human
+%populations are homogeneously mixed, that is, any person in the population is
+%equally likely to transmit the infection to any other. More sophisticated
+%models allow for the population to be divided into groups with higher or lower
+%contact rates, but these still do not account for individual variation or
+%population structure within compartments. In reality, and especially for
+%sexually-transmitted infections such as HIV, this assumption of homogeneity is
+%not realistic. There is substantial variation in the number of contacts per
+%individual, and potential transmissions are limited by geographic proximity.
+%The aim of using contact networks is to remove these unrealistic assumptions by 
+%explicitly simulating each individual in a population, along with the contacts
+%they have which could possibly lead to a transmission.
+%
+%A \defn{contact network} is a graphical representation of a population over
+%which an epidemic can spread. Contact networks consist of nodes and edges,
+%where the nodes represent people, and edges represent contacts. An edge between
+%two nodes indicate that transmission is possible between the two corresponding
+%individuals. For example, if we are considering an HIV transmission network,
+%the edges may represent sexual contacts or individuals sharing injection drug
+%paraphernalia. For an airborne pathogen such as influenza, an edge may simply
+%indicate that the two people were in the same room. The edges in a contact
+%network may be undirected, indicating that all contacts could possibly transmit
+%to each other, or directed, indicating that transmission risk is one-way. It is
+%possible to simulate an undirected network using directed edges by simply
+%including two directed edges, one in either direction, for each single directed
+%edge. Therefore, we exclusively consider directed contact networks here.
+%
+%Contact networks have been studied extensively in a variety of social,
+%economic, and public health contexts. However, investigation of the effects of
+%contact network structures on viral phylogenies have been much more limited.
+%Existing studies have to focused on a few broad classes of contact network, and
+%how phylogenies deriving from these networks have
+%differed~\autocite{leventhal2012inferring, colijn2014phylogenetic,
+%robinson2013dynamics}. To demonstrate our method, we focus on the same three
+%contact network types as studied in~\autocite{leventhal2012inferring}, namely
+%Erdos-Renyi (ER) graphs, Barabasi-Albert (BA) graphs, and Watts-Strogatz (WS)
+%graphs. ER graphs, also known as random graphs, are the simplest generative
+%graph model~\autocite{erdos1960evolution}. They have a single parameter, $p$,
+%which gives the probability of any edge occuring in the network. BA
+%graphs~\autocite{barabasi1999emergence} model a preferential attachment process
+%where nodes with high degree tend to attract more connections. They have two
+%parameters, $m$ and $\alpha$, and are constructed by repeatedly adding new
+%nodes with $m$ outgoing edges to the network. An existing node $v$ in the
+%network is chosen to be the endpoint of these new edges with probability
+%proportional to $d(v)^\alpha$, where $d(v)$ is the degree of $v$. WS
+%graphs~\autocite{watts1998collective}, also known as small-world graphs, have
+%two parameters, $k$ and $p$. They are constructed by creating a ring lattice
+%with $k$ edges per node, and then randomly rewiring each edge with probability
+%$p$.
+%
+%\subsection{Approximate Bayesian computation}
 
-% TODO: don't break this over the page
-
-\begin{algorithmic}
-  \While{the game is not over}
-    \If{I can capture an opponent's piece and $\Uniform(0, 1) < a$}
-      \State{capture the piece}
-    \Else
-      \State{make any other move at random}
-    \EndIf
-  \EndWhile
-\end{algorithmic}
-
-Suppose the observed data are the ending configurations of the board, after the
-player has concluded a game against an opponent with a known value of $a$. The
-model we have designed is very simple, but it is not obvious how to calculate
-the likelihood of a particular ending configuration. Indeed, it seems that the
-only way is to enumerate every possible path the game could have taken, and
-tabulate the ending configurations of each. Clearly, this is infeasible.
-Approximate Bayesian computation was designed for situations like these, where
-exact likelihoods are not available, perhaps due to the model involving an
-algorithm or generative process.
+%Consider a model $M$, with parameters $\theta$, which we wish to fit to some
+%observed data $D$. By ``fit'', we often mean that we want to find particular
+%values $\hat{\theta}$ for the parameters which optimize the likelihood of our
+%data given those parameters and the model,
+%\[
+%  \hat{\theta} = \argmax_\theta \Pr(D | M, \theta).
+%\]
+%This $\hat{\theta}$ is the \defn{maximum likelihood} parameter estimate. Note
+%that we are employing a common abuse of notation here, where $\Pr(\cdots)$ is
+%being unsed to refer to a probabilty \emph{density} rather than a true
+%probability. Alternative to maximum likelihood, we may be interested less in a
+%point estimate and more in the posterior distribution of possible values of
+%$\theta$ given our data, $\Pr(\theta \mid M, D)$. This will be expanded upon
+%below, but for the moment, for illustrative purposes, we restrict our attention
+%to the maximum likelihood problem.
+%
+%If the model we are fitting is sufficiently simple, it may be possible to
+%calculate $\hat{\theta}$ directly, using calculus. Most models do not admit
+%analytic maximum likeilhood solutions, but if the likelihood any set of
+%parameters can be calculated up to a normalizing constant, then
+%${\Pr(D \mid M,\theta)}$ can be optimized numerically. A wide range of
+%optimization strategies exist, the choice of which to use depending on the
+%complexity of the model and whether or not we have access to the gradients of
+%the likelihood function with respect to each of the parameters. The majority of
+%modelling problems fit into this category, and numerical optimization is
+%well-developed and extremely widely used.
+%
+%However, there are some cases, often when the observed data is of a complex
+%type, that explicitly calculating the likelihood of some observed data is
+%impossible, even up to a normalizing constant. For example, suppose that we
+%want to model a chess player's behaviour. We will set up a simple one-parameter
+%model which describes the chess playing process.  The parameter, $a \in [0, 1]$
+%indicates the player's eagerness to remove his opponent's pieces from the
+%board. We can write down an algorithm for the player's behaviour under such a
+%model.
+%
+%% TODO: don't break this over the page
+%
+%\begin{algorithmic}
+%  \While{the game is not over}
+%    \If{I can capture an opponent's piece and $\Uniform(0, 1) < a$}
+%      \State{capture the piece}
+%    \Else
+%      \State{make any other move at random}
+%    \EndIf
+%  \EndWhile
+%\end{algorithmic}
+%
+%Suppose the observed data are the ending configurations of the board, after the
+%player has concluded a game against an opponent with a known value of $a$. The
+%model we have designed is very simple, but it is not obvious how to calculate
+%the likelihood of a particular ending configuration. Indeed, it seems that the
+%only way is to enumerate every possible path the game could have taken, and
+%tabulate the ending configurations of each. Clearly, this is infeasible.
+%Approximate Bayesian computation was designed for situations like these, where
+%exact likelihoods are not available, perhaps due to the model involving an
+%algorithm or generative process.

### 2016-01-11 doc/thesis/intro.tex
Reading and writing

@@
-\subsection{Overview of phylodynamics}
+\subsection{Phylogenetics and phylodynamics}
 
@@
 graphical representation of the the evolutionary relationships among a group of
-organisms or species (generally, \defn{taxa}). Phylogenies consist of nodes,
-representing taxa, and edges or branches, which connect taxa to their
-evolutionary ancestors and descendants. The \defn{tips} of a phylogeny, that
-is, the nodes without any descendants, correspond to \defn{extant}, or
-observed, taxa, while the \defn{internal nodes} correspond to their extinct
-common ancestors. Phylogenies may have a \defn{root}, which is a node with no
-descendants distinguished as the most recent common ancestor of all the extant
-taxa. When such a root exists, the tree is referred to as being \defn{rooted};
-otherwise, it is \defn{unrooted}. The structural arrangement of nodes and edges
-in the tree is referred to as its \defn{topology}. The branches of the tree may
-have associated lengths, representing either calendar time or genetic distance
-between ancestors and their descendants.
+organisms or species (generally, \defn{taxa})~\autocite{haeckel1866generelle}.
+The \defn{tips} of a phylogeny, that is, the nodes without any descendants,
+correspond to \defn{extant}, or observed, taxa, while the \defn{internal nodes}
+correspond to their common ancestors. The edges or \defn{branches} of the
+phylogeny connect ancestors to their descendants. Phylogenies may have a
+\defn{root}, which is a node with no descendants distinguished as the most
+recent common ancestor of all the extant
+taxa~\autocite{harding1971probabilities}. When such a root exists, the tree is
+referred to as being \defn{rooted}; otherwise, it is \defn{unrooted}. The
+structural arrangement of nodes and edges in the tree is referred to as its
+\defn{topology}~\autocite{cavalli1967phylogenetic}.
 
-A phylogeny with branch lengths in calendar time units is often referred to as
-\defn{time-scaled}. In a time-scaled phylogeny, the internal nodes can be
-mapped onto a timeline by using the tips of the tree, which usually map to the
-present day, as a reference point~\autocite{nee1992tempo}. The corresponding
-points on the timeline are called \defn{branching times}. These concepts are
-illustrated in Figure~\ref{fig:speciestree}.
+The branches of the tree may have associated lengths, representing either
+evolutionary distance or calendar time between ancestors and their descendants.
+The term ``evolutionary distance'' is used here imprecisely to mean any sort of
+quantitative measure of evolution, such as the number of differences between
+the DNA sequences of an ancestor its descendant, or the difference in average
+body mass or height. This will be clarified below in
+subsection~\ref{subsubsec:genetree}. A phylogeny with branch lengths in calendar
+time units is often referred to as \defn{time-scaled}. In a time-scaled
+phylogeny, the internal nodes can be mapped onto a timeline by using the tips
+of the tree, which usually map to the present day, as a reference
+point~\autocite{nee1992tempo}. The corresponding points on the timeline are
+called \defn{branching times}, and the rate of their accumulation is referred
+to as the \defn{branching rate}. Rooted trees whose tips are all the same
+distance from the root are called \defn{ultrametric}
+trees~\autocite{buneman1974note}. These concepts are illustrated in
+Figure~\ref{fig:speciestree}.
 
@@
   \includegraphics{speciestree}
-  \caption[Illustration of a rooted, time-scaled phylogeny]{Illustration of a
-    rooted, time-scaled phylogeny. The tips of the tree, which represent extant
-    taxa, are placed at the present day on the
-    time axis. Internal nodes, representing extinct common ancestors to the
-    extant taxa, fall in the past. The topology of the tree indicates that cats
-    and dogs are the most closely related pair of species, whereas fish is most
-    distantly related to any other node in the tree.}
+  \caption[Illustration of a rooted, ultrametric, time-scaled phylogeny]
+    {Illustration of a rooted, ultrametric, time-scaled phylogeny. The tips of
+      the tree, which represent extant taxa, are placed at the present day on
+      the time axis. Internal nodes, representing extinct common ancestors to
+      the extant taxa, fall in the past. The topology of the tree indicates
+      that cats and dogs are the most closely related pair of species, whereas
+      fish is most distantly related to any other node in the tree.}
 \end{figure}
@@
 \subsubsection{Species trees and transmission trees}
+\label{subsubsec:speciestree}
 
@@
 the fossil record, or both. We discuss some of the challenges associated with
-this estimation in subsection \ref{subsec:treeconv} below. However, for the
-moment, we ignore these complications and suppose that we know the true species
-tree.
+this estimation in subsection \ref{subsubsec:treeconv} below. However, for the
+moment, we ignore these complications and consider what can be said about the
+true species tree.
 
-The process of speciation, which causes branching in a species tree, often
-occurs by an \defn{allopatric} process, where a sub-population of organisms is
-isolated from the original population by a geographic
-barrier~\autocite{coyne2004speciation}. Over time, the two populations diverge
-genetically, eventually resulting in two distinct species.
+Speciation, which causes branching in a species tree, often occurs by an
+\defn{allopatric} process, where a sub-population of organisms is isolated from
+the original population by a geographic barrier~\autocite{coyne2004speciation}.
+Over time, differing selection pressures on either side of the barrier, coupled
+with genetic drift, cause the two populations to diverge genetically. This
+eventually results in two distinct species, which will appear as two subtrees
+or \defn{clades} in their species tree if descendants of both survive to the
+present. Of course, this is not necessarily the case, as selection may cause
+one or both of the new sub-populations or their descendants to become extinct.
+On the other hand, selection may favour one of the sub-populations, allowing it
+to expand into new regions or niches and possibly paving the way for subsequent
+speciation events. Therefore, \comment{Is this a fair statement?} we can point
+out two forces primarily responsible for the shapes of species trees:
+geographic barriers, which cause branching, and selection, which controls which
+lineages survive to the present and influences the branching rate.
+
+A different type of tree used in the study of pathogens is a \defn{transmission
+tree}. In such trees, tips represent infected hosts, while internal nodes
+correspond to transmissions from one host to another. Transmission trees are
+generally time-scaled \comment{Is this true?} and rooted, with the root
+corresponding to the index case and branching times indicating times of
+transmission. The internal nodes may be labeled with the donor of the
+transmission pair, if this is known. The tips of the tree, rather than being
+fixed at the present day, are placed at the time at which the individual was
+discovered or their pathogen population sampled. Consequently, the transmission
+tree may not be ultrametric, but may have tips located at varying distances
+from the root. Such trees are said to have \defn{heterochronous} taxa or
+samples~\autocite{drummond2003measurably}, in contrast to the
+\defn{isochronous} samples found in macro-organisms species trees. A
+transmission tree is illustrated in Figure~\ref{fig:contactnet} (right).
+
+When considering epidemics of RNA viruses, transmission trees and species trees
+are highly related. The viral population within a single host has often been
+referred to as a \defn{quasispecies}~\autocite{domingo2012viral}, and due to
+the fast evolutionary rate of these viruses, two hosts' viral populations are
+invariably genetically distinct. The transmission process is functionally
+identical to allopatric speciation, involving geographic isolation of a
+sub-population in a new host and subsequent genetic divergence. However, for
+the most part, it is not selection but rather host epidemiology which causes
+certain lineages to proliferate while driving others to extinction. A host who
+recovers, dies, or is isolated equates to an extinction in the transmission
+tree. On the other hand, dense populations in frequent contact often experience
+rapid epidemic growth, which causes similarly rapid accumulation of branching
+points in the tree.
+
+The fact that these evolutionary (ie. speciation) and epidemiological processes
+occur on the same time scale for RNA viruses~\autocite{drummond2003measurably}
+is the basis for the field of
+\defn{phylodynamics}~\autocite{grenfell2004unifying}. The similarity between
+transmission and speciation means that it is possible to apply many of the
+tools and techniques developed for species trees to viral transmission trees,
+and interpret the results in the context of transmission rates and host
+epidemiology. As a basic example, the lineages-through-time
+plot~\autocite{nee1992tempo}, which plots the historical speciation rate
+against time, can be used to quantify the incidence of new infections over the
+course of an epidemic. \comment{Reference?} In
+subsection~\ref{subsubsec:treeshape} below, we review some of these methods in
+more detail, as well as several which have been developed specifically for
+viral phylodynamics, and discuss what these methods can tell us about host
+epidemiology.
+
+Finally, we must note two important points. First, we have not yet considered
+the selection pressures operating within a single host, such as immune
+pressure, because these forces do not directly impact transmission trees unless
+they cause a host to recover or die. Second, we reiterate that both
+transmission trees and species trees are theoretical objects which are almost
+never known with certainty for real data. Indeed, the estimation of
+transmission trees is often problematic, and it is in such estimation that
+within-host forces play a role. Both of these issues are discussed further
+below in subsection~\ref{subsubsec:treeconv}, though we continue to ignore them
+in the next subsection.
 
 \subsubsection{Gene trees and viral phylogenies}
+\label{subsubsec:genetree}
 
-\subsubsection{Using gene trees to estimate species trees}
+\subsubsection{Estimating transmission trees}
+\label{subsubsec:treeconv}
+
+Most applications of viral phylodynamics, including this project, aim to answer
+epidemiological questions~\autocite{volz2013viral}. For example, phylodynamic
+methods have been used to investigate the degree of
+clustering~\autocite{hughes2009molecular} and the effect of elevated
+transmission risk in acute infection~\autocite{volz2012simple} (a more thorough
+review is given below in subsection~\ref{subsubsec:appphylo}). For these
+applications, we are ultimately interested in the transmission tree, which is a
+representation of an epidemiological process (transmission), as opposed to the
+viral phylogeny, which represents a biological process (viral evolution). As
+discussed above in subsection~\ref{subsubsec:speciestree}, transmission trees and
+are in practice unknown, and must be estimated from available data. Here, we
+review some of the ways to perform such estimation.
+
+In principle, transmission trees can be reconstructed by on-the-ground
+epidemiological methods, that is, by asking each person who they contacted.
+This is particularly relevant for sexually transmitted diseases such as HIV,
+where individuals are more likely to recall who they contacted and
+when~\autocite{klovdahl1985social}. This kind of data is challenging to collect
+in detail. Most commonly, when applying phylodynamics in practice, the
+transmission tree is estimated from the viral
+phylogeny~\autocite{volz2013viral}. As discussed above in
+subsection~\ref{subsubsec:genetree}, the viral phylogeny is itself also estimated;
+however, for the moment we ignore this complication and describe methods for
+estimating the transmission tree if the viral phylogeny were known.
+
+When we consider the fact that the viral phylogeny is not known with certainty,
+the methods just described follow a two-step procedure. The first step is to
+estimate the viral phylogeny, and the second step is to use it to estimate the
+transmission tree.
+
+\autocite{didelot2014bayesian} develop a Bayesian version of the two-step
+approach. They first estimate a time-scaled phylogeny from viral samples, then
+estimate the transmissions which took place along the phylogeny. Due to
+within-host evolution, the transmissions are allowed to occur anywhere along
+the branches, rather than being constrained to happen at branching points.
+However, the authors admit that their method requires sampling of every
+infected individual, although they indicate that it could be extended to relax
+this assumption.
+
+A different approach is undertaken by Jombart
+\etal~\autocite{jombart2011reconstructing}, who build transmission trees
+directly from sequence data without the intermediate step of constructing a
+phylogeny. Their method finds the most likely geneology among the sampled
+sequences, assuming that the that the common ancestors to sampled isolates are
+themselves sampled. Although this may be a realistic assumption in the early
+stages of an epidemic of a slower evolving pathogen, it is unlikely for
+ancestors to be present among sequences sampled from an ongoing HIV epidemic
+\comment{Needs a reference}.
+
+Several alternative approaches have been developed which aim to
+incorporate available epidemiological data as well. Cottam
+\etal~\autocite{cottam2008integrating} first built a viral phylogeny, then
+enumerated all transmission trees consistent with this phylogeny by labelling
+the internal nodes of the phylogeny in every possible way. Each of the putative
+transmission trees was assigned a likelihood based on infection time and
+geographic separation.
 
 \subsubsection{Tree shapes}
+\label{subsubsec:treeshape}
+
+The aim of viral phylodynamics is to glean some kind of knowledge, about the
+epidemic, the virus, or its hosts and their behaviour, by studying a phylogeny,
+most often a transmission tree. Phylogenies are complex objects, and it is not
+immediately obvious how to extract useful information from them with respect to
+fitting a parameter. Standard statistical methods, built for numeric data
+cannot be applied directly - for example, one cannot perform a regression of a
+parameter of interest against a phylogeny. Therefore, before we discuss exactly
+what phylodynamics can tell us about epidemics, and how it has been applied in
+the past, we will review some of the methods for quantifying the shapes of
+phylogenies and their similarity to each other.
+
+\subsubsection{Applications of phylodynamics}
+\label{subsubsec:appphylo}
 
 \subsection{Contact networks}
+\label{subsec:contactnet}
 
-\subsubsection{Overview and epidemiology}
+\subsubsection{Overview}
 
@@
    indicate six symmetric contacts among the hosts. Coloured, bolded edges
-   represent transmissions. The epidemic began at node A, who transmitted to
+   represent transmissions. The epidemic began with node A, who transmitted to
    nodes B and C. Node C further transmitted to node D, and node E was not
    infected. On the right, the transmission tree topology corresponding to this
-   scenario.}
+   scenario. Note that individual B was sampled before the other infected
+   individuals, resulting in a tree with heterochronous tips.}
 \end{figure}
@@
 \subsubsection{Generative models}
+\label{subsubsec:generative}
 
-%Now, consider a phylogeny not of macro-organisms, but of RNA viruses. Many of
-%these, such as the human immuno-deficiency virus (HIV), evolve rapidy enough
-%that populations can accumulate measurable genetic changes on time scales of
-%months or years~\autocite{drummond2003measurably}. Because viruses cannot exist
-%for long outside their hosts, the host replaces the outside world as the
-%``environment'' in which evolution takes place. 
-%
-%In this light, striking parallels can be drawn between immunological and
-%epidemiological processes at the host level, and the evolutionary processes
-%studied in population genetics~\autocite{grenfell2004unifying}. For example,
-%both the host immune response~\autocite{mcmichael2001cellular} and the presence
-%of anti-retroviral drugs~\autocite{??} exert directional selection on the viral
-%population within a single host. Transmission, where a sub-population of
-%virions from the donor host is ``geographically'' isolated in the recpient,
-%closely mirrors the allopatric speciation process and causes a branching point
-%in the viral phylogeny (although this may not correspond exactly to the time of
-%transmission, as discussed in the next section). Consequently, any factor which
-%influences transmission dynamics, such as the virus' mode of transmission or
-%the population structure of its hosts, in turn affects the shape of the
-%phylogeny. The study of this interaction between host-level factors and viral
-%phylogies is known as \defn{phylodynamics}~\autocite{grenfell2004unifying}. 
-%
-%In particular, if a time-scaled phylogeny is constructed where all viruses from
-%the same host are collapsed into a single taxon, we might expect the branching
-%time relating the two hosts to correspond to the time of transmission. Of
-%course, this will only be the case if the two hosts' infections indeed diverged
-%allopatrically at the time of transmission, and not sympatrically within the
-%donor host at some previous time. This limitation, and others, in interpreting
-%the branching times of the viral phylogeny as transmission times, are discussed
-%in the next section. Nevertheless, even taking these limitations into account,
-%the relationship between branching times in the phylogeny and transmission
-%times in the population theoretically implies that the evolutionary history of
-%the virus is informative about the biology and epidemiology of its hosts. In
-%other words, information about, for example, transmission patterns, might be
-%inferred from the viral phylogeny. Phylodynamic methods are intended to
-%complement more traditional approaches to gathering epidemiological data, such
-%as population surveys.
-%
-%
-%\subsection{Sources of error in phylodynamics}
-%\label{subsec:treeconv}
-%
-%In general, the exact topology and branch lengths of a phylogeny relating
-%arbitrary taxa are not known and must be estimated. Such estimation is most
-%often carried out using some type of genetic data sampled from each of the
-%taxa, either amino acid sequences or gene orders. The resulting phylogeny, also
-%known as a \defn{gene tree}, differs from the time-scaled \defn{species tree}
-%which would contain complete information about past speciation events. In
-%viral phylogenetics, the gene tree is the viral phylogeny, constructed using
-%the genetic sequences of the sampled viruses. On the other hand, the species
-%tree is the \defn{transmission tree}, wherein branching times correspond
-%exactly to transmission times. In general, phylodynamic methods perform
-%inference based on the transmission tree, which must be inferred from a viral
-%phylogeny based on the available sequence data. The inference of the viral
-%phylogeny itself is not an error-free process, as accurate phylogenetic
-%inference is still an active area of research. Moreover, the conversion of the
-%phylogeny into a transmission tree is a process which has the potential to
-%introduce at least three types of error, described below. We use the terms
-%``gene tree'', ``viral phylogeny'', and ``phylogeny'' interchangably, as well
-%as the terms ``species tree'' and ``transmission tree''.
-%
-%Firstly, the branch lengths of the gene tree are measured in genetic distance,
-%and must be somehow converted to units of calendar time used in the species
-%tree. If we assume a constant rate of evolution along the whole tree (a
-%\defn{strict molecular clock}), then this conversion is straightforward - we
-%simply divide each of the branch lengths by the rate of evolution. However, it
-%is well known that rates of evolution are neither constant over time nor
-%constant between lineages, and the general problem of time-scaling a
-%phylogenetic tree is still an active area of research. Phylodynamic inference
-%relies on the assumption of a correctly time-scaled tree, and violations of
-%this assumption can mislead parameter estimates.
-%
-%A second limitation which has perhaps greater potential to mislead phylodynamic
-%inference is the potential discordance between the gene tree and the species
-%tree. In an allopatric speciation, some of the genetic diversity separating the
-%two eventual species may already have been present in the population
-%\emph{before} the geographic isolation occured. In that case, the branching
-%time in the gene tree, which indicates the time when two genetically distinct
-%versions of the gene, may be earlier than the branching point in the species
-%tree. In evolutionary biology, this phenomenon is referred to as
-%\defn{incomplete lineage sorting}. Furthermore, if there was substantial
-%variation within a population which underwent two or more speciation events in
-%close succession, the order in which the species diverged may be different from
-%the order in which their distinct genes arose in the original heterogeneous
-%population. In this case, the topology of the species tree would be different
-%from that of the gene tree. 
-%
-%Finally, there is the problem of rooting the tree, that is, identifying the
-%node corresponding to the most recent common ancestor of all the sampled
-%viruses. The choice of root influences the temporal order of the internal
-%nodes in the phylogeny. If the tips are constrained to the present day, or some
-%other fixed sampling time, the branch lengths in the species tree must also be
-%altered to accomodate the choice of root. 
-%
-%To our knowledge, there has been no theoretical exploration of the quantitative
-%effect of any of these sources of error on the accuracy of phylodynamic
-%inference. Most phylodynamic studies assume, implicitly or explicitly, that the
-%viral phylogeny provides the same information as the transmission tree, and
-%errors introduced by this assumption are generally ignored. Studies employing
-%Bayesian methods eliminate some of the uncertainty associated with constructing
-%the viral phylogeny, but this does not account for the error introduced in the
-%phylogeny-to-transmission-tree conversion process. Our work does not take a
-%Bayesian approach to inferring the phylogeny, and therefore is susceptible to
-%all the sources of error discussed here.
-%
-%\subsection{Contact networks}
-%
-%Compartmental models of epidemic growth and spread often assume that human
-%populations are homogeneously mixed, that is, any person in the population is
-%equally likely to transmit the infection to any other. More sophisticated
-%models allow for the population to be divided into groups with higher or lower
-%contact rates, but these still do not account for individual variation or
-%population structure within compartments. In reality, and especially for
-%sexually-transmitted infections such as HIV, this assumption of homogeneity is
-%not realistic. There is substantial variation in the number of contacts per
-%individual, and potential transmissions are limited by geographic proximity.
-%The aim of using contact networks is to remove these unrealistic assumptions by 
-%explicitly simulating each individual in a population, along with the contacts
-%they have which could possibly lead to a transmission.
-%
-%A \defn{contact network} is a graphical representation of a population over
-%which an epidemic can spread. Contact networks consist of nodes and edges,
-%where the nodes represent people, and edges represent contacts. An edge between
-%two nodes indicate that transmission is possible between the two corresponding
-%individuals. For example, if we are considering an HIV transmission network,
-%the edges may represent sexual contacts or individuals sharing injection drug
-%paraphernalia. For an airborne pathogen such as influenza, an edge may simply
-%indicate that the two people were in the same room. The edges in a contact
-%network may be undirected, indicating that all contacts could possibly transmit
-%to each other, or directed, indicating that transmission risk is one-way. It is
-%possible to simulate an undirected network using directed edges by simply
-%including two directed edges, one in either direction, for each single directed
-%edge. Therefore, we exclusively consider directed contact networks here.
-%
-%Contact networks have been studied extensively in a variety of social,
-%economic, and public health contexts. However, investigation of the effects of
-%contact network structures on viral phylogenies have been much more limited.
-%Existing studies have to focused on a few broad classes of contact network, and
-%how phylogenies deriving from these networks have
-%differed~\autocite{leventhal2012inferring, colijn2014phylogenetic,
-%robinson2013dynamics}. To demonstrate our method, we focus on the same three
-%contact network types as studied in~\autocite{leventhal2012inferring}, namely
-%Erdos-Renyi (ER) graphs, Barabasi-Albert (BA) graphs, and Watts-Strogatz (WS)
-%graphs. ER graphs, also known as random graphs, are the simplest generative
-%graph model~\autocite{erdos1960evolution}. They have a single parameter, $p$,
-%which gives the probability of any edge occuring in the network. BA
-%graphs~\autocite{barabasi1999emergence} model a preferential attachment process
-%where nodes with high degree tend to attract more connections. They have two
-%parameters, $m$ and $\alpha$, and are constructed by repeatedly adding new
-%nodes with $m$ outgoing edges to the network. An existing node $v$ in the
-%network is chosen to be the endpoint of these new edges with probability
-%proportional to $d(v)^\alpha$, where $d(v)$ is the degree of $v$. WS
-%graphs~\autocite{watts1998collective}, also known as small-world graphs, have
-%two parameters, $k$ and $p$. They are constructed by creating a ring lattice
-%with $k$ edges per node, and then randomly rewiring each edge with probability
-%$p$.
-%
 %\subsection{Approximate Bayesian computation}
@@
 %algorithm or generative process.
+
+\subsection{Approximate Bayesian computation}

### 2016-01-11 doc/thesis/methods.tex
Reading and writing

@@
 
-\begin{figure}
-  \caption{Comparison of original formulation of normalized
-    lineages-through-time, developed in~\autocite{janzen2015approximate}, with
-    our modified version using linear interpolation. Here, the red and blue
-    trees both have uniformly spaced branching times. Using step functions
-    (left), the nLTT of the two trees is non-zero due to the differing numbers
-    of internal branches. Using linear interpolation, the nLTT is zero (right).
-  The lines on the right graph have been offset for visibility.}
+\begin{figure}[ht]
+  \centering
+  \caption[Comparison of original and modified normalised lineages-through-time]
+    {Comparison of original formulation of normalized lineages-through-time,
+     developed in~\autocite{janzen2015approximate}, with our modified version
+     using linear interpolation. Here, the red and blue trees both have
+     uniformly spaced branching times. Using step functions (left), the nLTT
+     of the two trees is non-zero due to the differing numbers of internal
+     branches. Using linear interpolation, the nLTT is zero (right). The lines
+     on the right graph have been offset for visibility.}
   \includegraphics[scale=0.75]{nltt}
@@
 
-\subsection{Method validation}
+\subsection{Simulation experiments}
 
@@
 \label{subsubsec:kernel}
+
+Recall that our approximate Bayesian computation approach to fitting contact
+network models involves simulating transmission trees under a wide variety of
+parameter values, and then comparing these simulated trees to the true
+transmission tree. Values which produce trees similar to the observed
+transmission tree are distinguished as more likely than values which produce
+trees very different from the truth. In order for this type of analysis to
+succeed, it is critical that different parameter values produce different
+looking trees. Otherwise, if many different values produce trees which are too
+similar to each other, it will be impossible to distinguish which value is most
+consistent with the real tree. Just as importantly, trees simulated with
+similar parameter values must be similar to each other. In mathematical terms,
+we require the trees simulated from distinct parameter values to be
+\defn{separable} in tree space. The concept of separability is illustrated in
+Figure~\ref{fig:separable}.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{separable}
+  \caption[Separable versus non-separable pararameters in kernel space]{
+    Separable versus non-separable parameters in kernel space. Trees have been
+    simulated under three sets of parameters, represented as blue, red, and
+    green. An observed tree is shown in black. Trees are layed out such that
+    the distance between two trees corresponds to their similarity. In panel A,
+    trees from the same parameter set are similar to each other, but different
+    from other trees. The true tree is most consistent with the red parameters.
+    In panels B and C, trees from the same parameter set are not similar to
+    each other (B), or trees from different parameter sets are similar to each
+    other (C). It's difficult to say which parameters the true tree is most
+    consistent with.
+  }
+  \label{fig:separable}
+\end{figure}
+
+Before undertaking a complete ABC analysis, I analysed four simple contact
+network models to determine whether their parameters could be separated in tree
+kernel space. The four models are described in detail in
+subsection~\ref{subsubsec:generative} of the introduction. Briefly, they are:
+random networks, where each possible contact has a fixed probability of
+occuring; preferential attachment networks, where highly connected nodes tend
+to attract more contacts; small world networks, where nodes are connected to
+their immediate neighbours and the occasional far-flung contact; and full
+networks, where every possible connection is present. I will describe here only
+the procedure and results for the preferential attachment networks. The details
+of the other three types of graph can be found in the supplemental materials.
+
+The method of testing for separability was described previously
+in~\autocite{poon2015phylodynamic}, but I will reiterate it here for
+completeness. As a concrete example, consider the attachment power parameter
+$\alpha$ of the preferential attachment networks. This parameter describes the
+strength of attraction to highly connected nodes and is bounded below by zero
+indicating no extra attraction. By qualitative observation, I determined that a
+power of 2.0, which produced networks with very few ``hub'' nodes with
+extremely high degree, was a suitable upper bound (see
+Figure~\ref{fig:pabounds}). Therefore, I chose to test the values 0.5, 1.0, and
+1.5 for separability. The other parameters were fixed: the number of nodes in
+the network was 5000, and the mean degree of each node in the network was four.
+As discussed in subsection~\ref{subsubsec:generative}, this is the smallest
+mean degree value for preferential attachment networks which produces networks
+which are more than trees.
+
+\begin{figure}[ht]
+  \centering
+  \label{fig:pabounds}
+  \includegraphics{pa_power_bounds}
+  \caption[Upper and lower bounds on preferential attachment power]{
+    Qualitative justification for choice of zero and two as lower and upper
+    bounds on preferential attachment power $\alpha$. (A) Preferential
+    attachment network on 50 nodes with $\alpha = 0$, the lower bound enforced
+    by the model. (B) Preferential attachment network on 50 nodes with $\alpha
+    = 2$, where a few nodes have very high degree but the majority have very
+    low degree. (C) Density plot of node degrees in a 5000-node network with
+    $\alpha = 0$. The maximum degree of a node in this network was 24. (D)
+    Density plot of node degrees in a 5000-node network with $\alpha = 2$. The
+    maximum degree of a node in this network was 4942. 
+  }
+\end{figure}
+
+For each of the values of $\alpha$, I generated 100 networks on 5000 nodes. An
+epidemic was simulated over each network (see subsection X) until 1000 nodes
+were infected, and 500 of those infected nodes were sampled to form a
+transmission tree. This resulted in 300 total simulated transmission trees -
+100 for each of the three values of $\alpha$. Next, I computed the tree
+kernel~\autocite{poon2013mapping} (see subsection~\ref{subsubsec:treeshape})
+for each pair of trees. These values were placed into a 300 $\times$ 300 kernel
+matrix, where the value at the $(i, j)$th position was the tree kernel of the 
+$i$th and $j$th trees.
+
+The tree kernel provides a pairwise similarity score between two trees. The
+higher the kernel score, the more similar the trees are to each other.
+Therefore, to have separability, we need trees simulated with the same value of
+$\alpha$ to have high kernel scores with each other, but low kernel scores with
+trees from different $\alpha$ values. We can visually check whether or not this
+is true by laying out the trees as points on a graph, in such a way that trees
+with high scores are close to each other, but trees with low scores are far
+apart. This is accomplished by performing a kernel principal components
+analysis (kPCA)~\autocite{scholkopf1998nonlinear} on the kernel matrix.
+Briefly, ordinary principal components analysis (PCA) finds a lower dimensional
+representation of points in a high dimensional space which preserves as much of
+the variation in the data as possible. kPCA performs the same task, but using
+the dot products of each pair of points as input, instead of the data points
+themselves. A two-dimensional kPCA projection of the simulated trees is shown
+in Figure~\ref{fig:pakpca}. The scenario just described - 500 samples from 1000
+infected nodes - is the central panel. To ensure that the method could be used
+in a variety of contexts, the same analysis was performed with 500 and 2000
+infected nodes, as well as 100 and 1000 sampled tips. The projections show that
+all three values are well separated from each other in many of the scenarios.
+With smaller trees, it becomes harder to disinguish $\alpha = 0.5$ from $\alpha
+= 1.0$ in two dimensions.
+
+\begin{figure}[ht]
+  \centering
+  \label{fig:pakpca}
+  \includegraphics{kernel_pa_kpca}
+  \caption[Projection of kernel matrix for different attachment power values
+  onto its first two principal components]{
+    Projection of the kernel matrix for different preferential attachment power
+    values onto its first two principal components, for eight simulation
+    scenarios. Each point corresponds to a simulated transmission tree, and is
+    coloured by preferential attachment power. Facets are number of infected
+    nodes (horizontal), and number of sampled tips (vertical). The parameters
+    to the tree kernel were $\lambda = 0.3$ and $\sigma = 5$, and the nLTT was
+    not used. Qualitatively, trees with a larger number of tips are easier to
+    separate in kernel space, regardless of what sampling proportion they
+    represent. In all cases, the highest attachment power can be separated from
+    the other two, but the two lowest values become hand to distinguish with in
+    the 100-tip trees.
+  }
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \label{fig:pacrossv}
+  \includegraphics{kernel_pa_crossv}
+  \caption[Cross-validation performance of kernel support vector machine
+  classifier for preferential attachment power]{
+    Cross-validation performance of kernel support vector machine classifier
+    for preferential attachment power.
+  }
+\end{figure}
 

### 2016-01-11 doc/thesis/figures/pa_power_bounds.R

@@ -1,0 +1,28 @@
+#!/usr/bin/env Rscript
+
+library(igraph)
+library(extrafont)
+
+set.seed(0)
+
+g0.small <- sample_pa(50, m=2, power=0, directed=FALSE)
+g2.small <- sample_pa(50, m=2, power=2, directed=FALSE)
+g0.large <- sample_pa(5000, m=2, power=0, directed=FALSE)
+g2.large <- sample_pa(5000, m=2, power=2, directed=FALSE)
+
+summary(degree(g0.large))
+summary(degree(g2.large))
+
+pdf("pa_power_bounds.pdf", family="Gillius ADF", width=6, height=6)
+par(mfrow=c(2, 2))
+par(mar=c(2, 2, 2, 2) + 0.1)
+plot(g0.small, vertex.label=NA, vertex.color="black", vertex.size=5)
+text(grconvertX(0.05, "ndc", "user"), grconvertY(0.95, "ndc", "user"), "A", cex=2)
+plot(g2.small, vertex.label=NA, vertex.color="black", vertex.size=5)
+text(grconvertX(0.55, "ndc", "user"), grconvertY(0.95, "ndc", "user"), "B", cex=2)
+par(mar=c(4, 4, 2, 2) + 0.1, xpd=NA)
+plot(density(degree(g0.large)), main=NA, xlab="degree", ylab="density", zero.line=F)
+text(grconvertX(0.05, "ndc", "user"), grconvertY(0.5, "ndc", "user"), "C", cex=2)
+plot(density(degree(g2.large)), main=NA, xlab="degree", ylab="density", zero.line=F)
+text(grconvertX(0.55, "ndc", "user"), grconvertY(0.5, "ndc", "user"), "D", cex=2)
+dev.off()

### 2016-01-11 doc/thesis/figures/separable.R

@@ -1,0 +1,46 @@
+#!/usr/bin/env Rscript
+
+library(ape)
+library(png)
+library(raster)
+library(extrafont)
+
+set.seed(4)
+
+tree.col <- c("black", rep(c("blue", "red", "forestgreen"), each=3))
+tmp <- tempfile()
+tinytree <- as.list(rep(0, 10))
+for (i in 1:10) {
+    png(tmp, width=50, height=50, bg="transparent")
+    plot(rtree(4), show.tip.label=F, direction="down", no.margin=TRUE, edge.width=5,
+         edge.color=tree.col[i])
+    dev.off()
+    tinytree[[i]] <- readPNG(tmp)
+}
+unlink(tmp)
+
+pdf("separable.pdf", width=6, height=2, family="Gillius ADF")
+par(mar=c(0, 0, 0, 0) + 0.1, mfrow=c(1, 3))
+
+plot(NA, xlim=c(0, 1), ylim=c(0, 1), axes=FALSE, xlab=NA, ylab=NA, frame.plot=TRUE)
+x <- c(rep(c(0.2, 0.4, 0.8), each=3) + rnorm(9, sd=0.1))
+y <- c(rep(c(0.7, 0.4, 0.6), each=3) + rnorm(9, sd=0.1))
+rasterImage(tinytree[[1]],0.5,0.5,0.6,0.6)
+mapply(rasterImage, tinytree[2:10], as.list(x), as.list(y), as.list(x+0.1), as.list(y+0.1))
+text(0.05, 0.95, "A", cex=2)
+
+plot(NA, xlim=c(0, 1), ylim=c(0, 1), axes=FALSE, xlab=NA, ylab=NA, frame.plot=TRUE)
+x <- c(rep(c(0.2, 0.4, 0.8), 3) + rnorm(9, sd=0.2))
+y <- c(rep(c(0.7, 0.4, 0.6), 3) + rnorm(9, sd=0.2))
+rasterImage(tinytree[[1]],0.5,0.5,0.6,0.6)
+mapply(rasterImage, tinytree[2:10], as.list(x), as.list(y), as.list(x+0.1), as.list(y+0.1))
+text(0.05, 0.95, "B", cex=2)
+
+plot(NA, xlim=c(0, 1), ylim=c(0, 1), axes=FALSE, xlab=NA, ylab=NA, frame.plot=TRUE)
[…]

### 2016-01-26 doc/thesis/figures/gridsearch_expt.tex
Writing yay

\documentclass{article}

\usepackage{tikz}
\usepackage{array}
\usepackage{nicefrac}
\usepackage{fullpage}
\usepackage[T1]{fontenc}
\usepackage[sfmath]{kpfonts}
\usepackage[default]{gillius}
\usepackage[normalem]{ulem}

\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{calc}

\begin{document}

\pagestyle{empty}

\newcommand{\boxw}{7cm}
\newcommand{\boxh}{2cm}

\begin{tikzpicture}
  \tikzstyle{box}=[minimum width=\boxw, minimum height=\boxh, draw]
  \tikzstyle{next}=[->, >=stealth, thick]

  \node (rect1) [box] { };
  \node at (rect1.north) [anchor=north] {\uline{$k$ testing values}};
  \node (t3) [above=0.5cm of rect1.south, inner sep=6pt, circle, draw, fill=blue!60!white] { };
  \node (t2) [left=0.5cm of t3, circle, inner sep=6pt, draw, fill=blue!80!white] { };
  \node (t4) [right=0.5cm of t3, circle, inner sep=6pt, draw, fill=blue!40!white] { };
  \node (t1) [left=0.5cm of t2, circle, inner sep=6pt, draw, fill=blue] { };
  \node (t5) [right=0.5cm of t4, circle, inner sep=6pt, draw, fill=blue!20!white] { };

  \node (rect2) [box, right=of rect1] { };
  \node at (rect2.north) [anchor=north] {\uline{$n$ training values}};
  \node (bar) [above=0.5cm of rect2.south, minimum width=6cm, minimum height=0.5cm, 
         rectangle, draw, left color=blue, right color=white] { };
  \foreach \x in {0.2, 0.4, ..., 6}{
    \draw ($(bar.north west) + (\x cm, 0)$) -- ($(bar.south west) + (\x cm, 0)$);
  }

  \node (rect3) [box, below=of rect1] { };
  \node at (rect3.north) [anchor=north] {\uline{$10 \times k$ testing trees}};
  \node [above=4pt of rect3.south] {
    \begin{tabular}{m{0.8cm} m{1cm} m{1cm} m{0.8cm} m{1cm}}
      \includegraphics[width=1cm]{testtree1} & $\times$ 10 &
      $\cdots$ &
      \includegraphics[width=1cm]{testtree3} & $\times$ 10
    \end{tabular}
  };

  \node (rect4) [box, below=of rect2] { };
  \node at (rect4.north) [anchor=north] {\uline{$15 \times n$ training trees}};
  \node [above=4pt of rect4.south] {
    \begin{tabular}{m{0.8cm} m{1cm} m{1cm} m{0.8cm} m{1cm}}
      \includegraphics[width=1cm]{testtree1} & $\times$ 15 &
      $\cdots$ &
      \includegraphics[width=1cm]{testtree3} & $\times$ 15
    \end{tabular}
  };

  \coordinate [left=0.5 of rect4.south west] (r4sw);
  \node (rect5) [box, below=of r4sw] { };
  \node at (rect5.north) [anchor=north] {\uline{$10 \times k$ kernel score vectors}};
  \node [above=0 of rect5.south, minimum width=\boxw, align=center] {
    \includegraphics[scale=0.6]{tinykscore}
  };

  \node (rect6) [box, below=of rect5] { };
  \node at (rect6.north) [anchor=north] {\uline{$10 \times k$ parameter estimates}};
  \node [above=6pt of rect6.south, minimum width=\boxw, text width=\boxw, align=center] {
    point estimate = training value with highest median kernel score
  };

  \coordinate [above=0.5 of rect5.north] (r5n);
  \draw [next] (rect1) -- (rect3);
  \draw [next] (rect2) -- (rect4);
  \draw [next] (rect3) |- (r5n) -- (rect5);
  \draw [next] (rect4) |- (r5n) -- (rect5);
  \draw [next] (rect5) -- (rect6);

\end{tikzpicture}

\end{document}

### 2016-01-26 doc/thesis/figures/kernel_expt.tex
Writing yay

\documentclass{article}

\usepackage{tikz}
\usepackage{array}
\usepackage{nicefrac}
\usepackage{fullpage}
\usepackage[T1]{fontenc}
\usepackage[sfmath]{kpfonts}
\usepackage[default]{gillius}
\usepackage[normalem]{ulem}

\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{calc}

\begin{document}

\pagestyle{empty}

\newcommand{\boxw}{7cm}
\newcommand{\boxh}{2cm}

\begin{tikzpicture}
  \tikzstyle{box}=[minimum width=\boxw, minimum height=\boxh, draw]
  \tikzstyle{next}=[->, >=stealth, thick]

  \node (rect1) [box] { };
  \node at (rect1.north) [anchor=north] {\uline{3 parameter values}};
  \node (b) [above=0.5cm of rect1.south, inner sep=6pt, circle, draw, fill=blue] { };
  \node (r) [left=1.5cm of b, circle, inner sep=6pt, draw, fill=red] { };
  \node (g) [right=1.5cm of b, circle, inner sep=6pt, draw, fill=green] { };

  \node (rect2) [below=of rect1, box] { };
  \node at (rect2.north) [anchor=north] {\uline{300 networks}};
  \node (b) [above=4pt of rect2.south] {
    \begin{tabular}{m{0.5cm} m{1cm} m{0.5cm} m{1cm} m{0.5cm} m{1cm}}
      \includegraphics[width=1cm]{tinynet1} & $\times$ 100 &
      \includegraphics[width=1cm]{tinynet2} & $\times$ 100 &
      \includegraphics[width=1cm]{tinynet3} & $\times$ 100
    \end{tabular}
  };

  \node (rect3) [below=of rect2, box] { };
  \node at (rect3.north) [anchor=north] {\uline{300 epidemics}};
  \node (b) [above=4pt of rect3.south] {
    \begin{tabular}{m{0.5cm} m{1cm} m{0.5cm} m{1cm} m{0.5cm} m{1cm}}
      \includegraphics[width=1cm]{tinyepi1} & $\times$ 100 &
      \includegraphics[width=1cm]{tinyepi2} & $\times$ 100 &
      \includegraphics[width=1cm]{tinyepi3} & $\times$ 100
    \end{tabular}
  };

  \node (rect4) [below=of rect3, box] { };
  \node at (rect4.north) [anchor=north] {\uline{300 transmission trees}};
  \node (b) [above=4pt of rect4.south] {
    \begin{tabular}{m{0.5cm} m{1cm} m{0.5cm} m{1cm} m{0.5cm} m{1cm}}
      \includegraphics[width=1cm]{tinytree1} & $\times$ 100 &
      \includegraphics[width=1cm]{tinytree2} & $\times$ 100 &
      \includegraphics[width=1cm]{tinytree3} & $\times$ 100
    \end{tabular}
  };

  \node (rect5) [right=of rect1, box] { };
  \node at (rect5.north) [anchor=north, text width=\boxw, align=center] {
    \uline{42 kernel matrices ($300 \times 300$)} \\ \vspace{4pt}
    $\lambda$ = \{ \nicefrac18, \nicefrac14, \nicefrac12, 1, 2, 4, 8 \} \\
    $\sigma$ = \{ 0.2, 0.3, 0.4 \} \\
    nLTT = \{ yes, no \} \\
  };

  \node (rect6) [below=of rect5, box] { };
  \node at (rect6.north) [anchor=north, text width=\boxw, align=center] {
    \uline{42 cross validations} \\ \hfill \\
    R$^2$ $\times$ 42
  };

  \node (rect7) [below=of rect6, box] { };
  \node at (rect7.north) [anchor=north, text width=\boxw, align=center] {
    \uline{optimal kernel meta-parameters} \\ \hfill \\
    $\lambda$ = 4,\; $\sigma$ = 0.3,\; nLTT = no
  };

  \node (rect8) [below=of rect7, minimum width=\boxw, minimum height=\boxh,
                 align=center, text width=\boxw] {
    repeat 8 times \\
    \vspace{4pt}
    infected nodes = \{ 500, 1000, 2000 \} \\
    sampled nodes = \{ 100, 500, 1000 \}
  };

  \coordinate [left=0.5 of rect5.west] (r5w);
  \coordinate [below=0.5cm of r] (r1ssw);
  \coordinate [below=of r1ssw] (r2nnw);
  \coordinate [below=0.5cm of g] (r1sse);
  \coordinate [below=of r1sse] (r2nne);
  \coordinate [below=3.5cm of r] (r2ssw);
  \coordinate [below=3.5cm of g] (r2sse);
  \coordinate [below=4.5cm of r] (r3nnw);
  \coordinate [below=4.5cm of g] (r3nne);
  \coordinate [below=6.5cm of r] (r3ssw);
  \coordinate [below=6.5cm of g] (r3sse);
  \coordinate [below=7.5cm of r] (r4nnw);
  \coordinate [below=7.5cm of g] (r4nne);
  \coordinate [below=0.5cm of rect4.north east] (r4ene);
  \coordinate [above=0.5cm of rect4.south east] (r4ese);

  \draw[next] (r1ssw) -- (r2nnw);
  \draw[next] (r1ssw) -- ($(r2nnw) + (-0.5cm, 0cm)$);
  \draw[next] (r1ssw) -- ($(r2nnw) + (0.5cm, 0cm)$);
  \draw[next] (r1sse) -- (r2nne);
  \draw[next] (r1sse) -- ($(r2nne) + (-0.5cm, 0cm)$);
  \draw[next] (r1sse) -- ($(r2nne) + (0.5cm, 0cm)$);
  \draw[next] (rect1.south) -- (rect2.north);
  \draw[next] (rect1.south) -- ($(rect2.north) + (-0.5cm, 0cm)$);
  \draw[next] (rect1.south) -- ($(rect2.north) + (0.5cm, 0cm)$);
  \draw [next] (rect2) -- (rect3);
  \draw [next] (r2ssw) -- (r3nnw);
  \draw [next] (r2sse) -- (r3nne);
  \draw [next] (rect3) -- (rect4);
  \draw [next] (r3ssw) -- (r4nnw);
  \draw [next] (r3sse) -- (r4nne);
  \draw [next] (rect4.east) -| (r5w) -- (rect5);
  \draw [thick] (r4ene) -- ++ (0.5cm, -0.5cm);
  \draw [thick] (r4ese) -- ++ (0.5cm, 0.5cm);
  \draw [next] (rect5) -- (rect6);
  \draw [next] (rect6) -- (rect7);
\end{tikzpicture}

\end{document}

### 2016-01-26 doc/thesis/intro.tex
Writing yay

@@
 A \defn{species tree} is a particular type of phylogeny in which the taxa are
-species, and the branching times correspond to historical speciation events. In
-general, we do not have access to the true tree relating a particular group of
-extant species, as this would imply certain knowledge of speciation events in
-the distant past. Therefore, all species trees considered by researchers are
-estimates, based on the genetic or phenotypic similarity of the extant taxa,
-the fossil record, or both. We discuss some of the challenges associated with
-this estimation in subsection \ref{subsubsec:treeconv} below. However, for the
-moment, we ignore these complications and consider what can be said about the
-true species tree.
+species, and the branching times correspond to historical speciation events. By
+speciation events, we mean times when two formerly interbreeding groups of
+organisms ceased to reproduce with each other. In general, we do not have
+access to the true tree relating a particular group of extant species, as this
+would imply certain knowledge of speciation events in the distant past.
+Therefore, all species trees considered by researchers are estimates, based on
+the genetic or phenotypic similarity of the extant taxa, the fossil record, or
+both. We discuss some of the challenges associated with this estimation in
+subsection \ref{subsubsec:treeconv} below. However, for the moment, we ignore
+these complications and consider what can be said about the true species tree.
 
@@
 to expand into new regions or niches and possibly paving the way for subsequent
-speciation events. Therefore, \comment{Is this a fair statement?} we can point
-out two forces primarily responsible for the shapes of species trees:
-geographic barriers, which cause branching, and selection, which controls which
-lineages survive to the present and influences the branching rate.
+speciation events. It must be noted that speciation is not necessarily
+allopatric. \defn{Sympatric} speciation refers to the emergence of a new
+species without any geographic barrier~\autocite{coyne2004speciation}. Indeed,
+there is a whole continuum of possible speciation
+processes~\autocite{fitzpatrick2008if}, ranging from completely allopatric to
+completely \defn{sympatric}. However, at least for eukaryotic organisms,
+allopatric speciation is thought to be by far the most common
+type~\autocite{coyne2004speciation}.
 
@@
 tree}. In such trees, tips represent infected hosts, while internal nodes
-correspond to transmissions from one host to another. Transmission trees are
-generally time-scaled \comment{Is this true?} and rooted, with the root
-corresponding to the index case and branching times indicating times of
+correspond to transmissions from one host to another. Transmission trees
+generally have branch lengths in units of calendar time, with a root
+corresponding to the index case, and branching times indicating times of
 transmission. The internal nodes may be labeled with the donor of the
@@
 against time, can be used to quantify the incidence of new infections over the
-course of an epidemic. \comment{Reference?} In
+course of an epidemic~\autocite{holmes1995revealing}. In
 subsection~\ref{subsubsec:treeshape} below, we review some of these methods in
@@
 \label{subsubsec:genetree}
+
+A \defn{gene tree} is another type of phylogeny which traces the evolutionary
+history of one or more sections of genetic material. Gene trees can be built
+from small parts of genes up to whole genomes, but for ease of exposition we
+will use only the word ``gene''. The tips of a gene tree represent the
+different \defn{alleles}, or versions, of the gene found in the studied taxa.
+The branching points in a gene tree represent genetic divergence events - that
+is, the times when two distinct alleles first came into existence through
+mutation.
+
+Gene trees are usually estimated using the nucleotide or amino acid sequences
+of the genes in question. Although there are methods for estimating gene trees
+using other types of molecular data, such as gene orders, we will focus on
+sequence-based methods here. As with species trees, it is not possible to know
+gene trees with certainty, only to estimate them. However, the problem of
+estimating gene trees directly from data is somewhat better defined than that
+of estimating species trees, because of the quantitative nature of the
+molecular data used to label the tips. In general, methods to estimate gene
+trees fall into three major categories: distance-based, maximum parsimony, and
+model-based~\autocite{nei2000molecular}. The first two categories are generally
+no longer used for phylogenetic inference, since model-based methods have been
+demonstrated to be almost universally more accurate. Therefore, we will review
+in detail only the last category.
+
+Model-based methods include both maximum likelihood and Bayesian approaches. To
+use them, one must first define a parametric model of sequence evolution, here
+denoted $M$. For all possible trees $T$ (with branch lengths) and choices of
+model parameters $\theta$, the model must define a probability density function
+on observed sequence data $D$, denoted $f(D \mid T,\,\theta)$. This is also
+sometimes written $\Pr(D \mid T,\,\theta)$, even though it is a probability
+\emph{density}, not a probability. Maximum likelihood aims to find a particular
+tree and values for the model parameters which optimize the \defn{likelihood
+function} $\L$ for the observed data,
+\[
+  \L(\theta,\, T \mid D) = f(D \mid T,\, \theta).
+\]
+Bayesian methods, rather than trying to find point estimates for $\theta$ and
+$T$, aim to approximate the entire \defn{posterior distribution} for the
+observed data,
+\[
+  f(T,\,\theta \mid D) = \frac{f(D \mid T,\,\theta)f(T,\,\theta)}{f(D)}.
+\]
+Here, $f(T, \theta)$ denotes the prior distribution on $T$ and $\theta$, and
+$f(D)$ is the marginal density of the data (that is, the integral of the
+numerator over all values of $T$ and $\theta$).
+
+A \defn{viral phylogeny} is a particular kind of gene tree which relates the
+genetic sequences of viruses carried within one or more hosts. In this work, we
+consider viral phylogenies constructed from one sequence per infected
+individual, so that the tips of the are associated with both the viral sequence
+and the corresponding host.
 
@@
 
+Many tree summary statistics have been developed to assign numerical values to
+phylogenies based on their properties. One of the most widely used is Sackin's
+index, which measures the imbalance or asymmetry in the tree.
+
 \subsubsection{Applications of phylodynamics}
@@
 
-\subsubsection{Generative models}
+\subsubsection{Models for contact networks}
 \label{subsubsec:generative}
+
+Throughout this section, the variable $n$ will be used to refer to the number
+of nodes in the network.
+
+An \defn{Erd\H{o}s-R\'enyi} (ER) network~\autocite{erdos1960evolution}, also
+known as a Bernoulli network, was the first network model described. It has a
+single parameter, $p$, which is the probability that any particular edge is
+present in the network. To construct an ER network, one simply chooses
+$p \times \binom{n}{2}$ distinct pairs of nodes and draws an edge between each
+pair. The average degree of a node in an ER network is equal to $p \times n$.
+
+A \defn{Barab\'asi-Albert} (BA) network~\autocite{barabasi1999emergence},
+incorproates the phenomenon of \defn{preferential attachment} seen in real life
+networks. This means that new connections are more likely to involve nodes
+which already have a high number of connections - popular nodes tend to become
+more popular. In the formulation we consider here, BA networks have two
+parameters, $m$ and $\alpha$, which are used to construct the network via the
+following algorithm. The network begins as a single node, then the remaining
+$n-1$ nodes are added one at a time. Each time a new node is added, it is
+connected to $m$ other nodes. The probability of one of these connections
+involving an existing node of degree $k$ is proportional to $k^\alpha$. Note
+that in the original formulation~\autocite{barabasi1999emergence}, there was an
+additional parameter $m_0$ which indicated the number of initial vertices in
+the network, and the only value of $\alpha$ considered was 1. For our purposes,
+we fix $m_0 = 1$ but allow $\alpha$ to vary. We refer to $\alpha$ throughout
+this work as the \defn{preferential attachment power} or just attachment power.
+
+Networks produced by the BA model are \defn{scale free}, which means that the
+probability of a node having $k$ connections is proportional to $k^\gamma$ for
+some constant $\gamma$. This is also called a \defn{power law} degree
+distribution. An important remark on notation must be made here: in network
+literature, the Greek letter $\alpha$ has often been used to refer to the power
+law exponent of the degree distribution, not the preferential attachment power.
+However, the paper defining the network model~\autocite{barabasi1999emergence}
+uses $\gamma$ for the power law exponent, and $\alpha$ for the attachment
+power, and we shall do likewise. This must be noted because there are reports
+in the literature of the power law exponent being as large as 4, which would
+seem to conflict with our choice (see subsection XX) to bound $\alpha$ above by
+2. The reader should rest assured that it is the attachment power we are
+bounding, not the power law exponent.
+
+WS and BA networks differ from ER networks in an important way: it is possible
+to explicitly calculate the likelihood of the ER model parameter $p$ given an
+observed network, but this cannot be done for the other two models. This is
+because the probability of a particular edge occuring is always equal to $p$ in
+an ER network, regardless of the presence or absence of any other edges. In
+other words, the edges can be viewed as a series of $\binom{n}{2}$ Bernoulli
+trials with success probability $p$, so that the likelihood of $p$ given an
+observed graph with $k$ edges is distributed as 
+Binomial$\left(\binom{n}{2}, p\right)$.
 
@@
 \subsection{Approximate Bayesian computation}
+
+\subsubsection{Overview and motivation}
+
+Mathematical models can be grouped into three categories in terms of our
+ability to fit them to observed data. First, there is a limited set of models
+which can be fit explicitly by using calculus to minimize a loss function. The
+canonical example of this type of model is a linear regression, with the
+ordinary least-squares loss function. A linear regression is a model of the
+relationship between some data points $x_1, \ldots, x_n$ and outcomes $y_1,
+\ldots, y_n$ which assumes that the outcomes are linearly related to the data,
+modulo some noise caused by measurement errors, environmental fluctuations, and
+so on. In other words, there is a constant $b$ such that 
+\begin{align}
+  \label{eq:regression}
+  y_i = bx_i + \varepsilon_i, 
+\end{align}
+where $\varepsilon_i$ is the error associated with the outcome $y_i$. Notice
+that, if we fix $b$, then since $y_i$ and $x_i$ are both given, $\varepsilon_i$
+is also fixed. The process of fitting a linear model means choosing the
+``best'' $b$. For least squares, this optimal $b$ is defined as the one which
+minimizes the sum of the squared $\varepsilon_i$.
+
+Now that we have defined what it means to fit a least-squares linear
+regression, there remains the question of how to go about finding this optimal
+$b$ value. Recall that the errors $\varepsilon_i$ are fixed for a particular
+$b$, and our objective is to minimize the loss function, which is the sum of
+the squared errors. Naively, we could simply try many different $b$ values,
+calculate the loss function for each, and choose the $b$ which gives the
+smallest value. This approach, though basic, falls under the umbrella of
+\defn{numerical optimization}, which is the most common way the majority of
+models are fit. Of course, there is a whole field devoted to more sophisticated
+methods for exploring the parameter space, but they all boil down to the basic
+method of trying different values and choosing the one which minimizes the
+loss.
+
+In the case of least-squares linear regression, there is another method we
+could use. Looking at equation~\ref{eq:regression} for the model, it is easy to
+rearrange terms to solve for $\varepsilon_i$ in terms of the other variables:
+we get $\varepsilon_i = y_i - bx_i$. In fact, we can write down the entire loss
+function, which is the sum of the squared $\varepsilon_i$ in terms of $b$:
+\[
+  \sum \varepsilon_i^2 = \sum (y_i - bx_i)^2,
+\]
+where the sums are over all $i$ from 1 to $n$, the number of points in the data
+set. We will not describe the procedure here, but it is possible to explicitly
+minimize the right-hand side by finding the point where the derivative with
+respect to $b$ is zero. However, the method of directly minimizing the loss
+function with calculus is only applicable to a very narrow class of simple
+models. For the majority, the loss function is too complicated to set the
+derivative with respect to the parameters to zero and solve.
+
+Because of the limited space of models which permit exact solutions, numerical
+optimization of a loss function is the most frequently used method to fit a
+model.
+
+An approximate Bayesian computation approach to least-squares regression might
+be as follows. We will need to make an additional assumption: that the errors
+$\varepsilon_i$ are normally distributed. That is, our linear model is now of
+the form
+\[
+  y_i = bx_i + \N(0, \sigma),
+\]
+where $\N(0, \sigma)$ indicates a normal distribution with mean zero and
+variance $\sigma$. We do not know $\sigma$ in advance, so this is another
+parameter we will need to estimate. There is nothing special about normal
+distributions - we could have chosen any other distribution we deemed
+appropriate.
+
+The main idea of approximate Bayesian computation is that the ``best''
+parameters will define a model which produces simulated data similar to the
+real data. Suppose we fix values of $b$ and $\sigma$. Then, using the given
+data points $x_i$, we can generate simulated observations $y_i^*$ by adding
+$bx_i$ and a randomly chosen error drawn from $\N(0, \sigma)$.
+
+\subsubsection{Algorithms for ABC}
+\label{subsubsec:abcalg}
+
+Approximate Bayesian computation does not refer to a particular procedure for
+model fitting. Rather, ABC refers to the general strategy of choosing model
+parameters based on the resulting model's propensity to generate data
+resembling the real data. There are three main classes of ABC algorithm which
+have been developed so far: rejection, \gls{MCMC}, and sequential Monte Carlo
+(SMC).
+
+All of these approaches require some common elements. First, as with all
+Bayesian methods, we are required to specify a \defn{prior distribution},
+denoted $\pi$, on the parameter space. The prior specifies what we already know
+or believe about the model parameters. Second, in order to compare simulated to
+observed data, we need to be able to summarize a data set in a numerical
+format. This is accomplished by a function, denoted $\eta$, which computes a
+vector of hopefully informative summary statistics on a data set. Third, we
+need a distance function $\rho$ which tells us how similar two data sets are to
+each other, based on their summary statistics.
+
+Continuing with the linear model example, we need to specify a prior $\pi(a, b,
+\sigma)$ on the three parameters. We do not have much certain information about
+these parameters except that $\sigma$ has to be at least zero, but it seems
+reasonable to assume that extreme relationships are fairly rare, and that
+positive and negative correlations are equiprobable. Therefore, we will let
+$a$, $b$, and $\sigma$ be independent, $a$ and $b$ be normally distributed,
+and $\sigma$ be log-normally distributed. That is,
+\[
+  \pi(a, b, \sigma) = 
+  \begin{cases}
+    \Pr[\N(0, 1) = a] \cdot \Pr[\N(0, 1) = b] \cdot \Pr[\N(0, 1) = \log\sigma]
+     & \sigma > 0 \\
+    0 & \sigma \le 0.
+  \end{cases}
+\]
+For the vector of summary statistics, we will use the mean and variance of the
+simulated data,
+\[
+  \eta(\vec{y}) = \left<\E[\vec{y}], \Var[\vec{y}]\right>.
+\]
+Finally, for the distance function $\rho$ we take the standard Euclidian
+distance.
+
+Rejection ABC is the simplest method, and also the one which was first
+proposed~\autocite{rubin1984bayesianly}. Effectively, it comes down to the
+approach described in the previous subsection of guessing parameter values
+until one is close enough to the truth. More specificially, a possible set of
+parameters $\theta$ is drawn from the prior distribution, and a simulated data
+set $z$ is generated from the model with those parameters. If the distance
+between the simulated data set and the real data, $\rho(\eta(y), \eta(z))$, is
+small enough, then we accept $\theta$ as a sample from the posterior. This can
+be repeated until as many samples as desired are obtained.
+
+% MCMC
+
+The most recently developed class of algorithms for ABC is sequential
+Monte-Carlo (SMC)~\autocite{sisson2007sequential}. As with the other two
+classes, we want to eventually obtain a sample from the posterior distribution
+$f(\theta \mid D)$. The idea of SMC is to begin with a sample from a
+distribution we know, most often the prior, and approach the posterior smoothly
+by progressing through a series of intermediate distributions.

### 2016-01-26 doc/thesis/methods.tex
Writing yay

@@
 
-I implemented the adaptive sequential Monte-Carlo algorithm
-developed in~\autocite{del2012adaptive} for approximate Bayesian computation.
-In contrast to most existing sequential Monte-Carlo methods, this algorithm
-does not require the user to specify a sequence of decreasing tolerances to
-approach the target posterior distribution. Rather, the tolerances are computed
-adaptively at each step, starting from infinity at the first iteration. The
-algorithm may be stopped when the tolerance reaches a user-defined final value,
-or when the rate of acceptance of the MCMC kernel reaches a user-defined
-threshold. Following a heuristic applied by the
-authors~\autocite{del2012adaptive}, we used the latter stopping criterion,
-accepting the SMC approximation to the posterior when the MCMC acceptance rate
-dropped below 1.5\%.
+I implemented the adaptive sequential Monte-Carlo (SMC) algorithm developed by
+Del Moral \etal~\autocite{del2012adaptive} for approximate Bayesian
+computation.
+
+For ease of exposition, we simplifiy the notation of Del Moral \etal by
+dropping the subscripts on the variables which indicate the current iteration
+of the algorithm. Instead, we will add a superscript $^*$ to indicate the value
+of the variable which will be used in the next iteration (this should become
+clear later on). Let $\set{\theta_i}_{i=1}^{N}$ be a set of $N$ particles,
+which are particular choices of the model parameters $\theta$. For each
+particle $\theta_i$, let $\set{X_{i,j}}_{j=1}^M$ be a set of $M$ simulated
+datasets, generated from the model with parameters $\theta_i$. Let
+$\set{W_i}_{i=1}^N$ be weights associated to each particle. Let $d$ be a
+distance measure on data sets, so that $d(x, y)$ is smaller the more similar
+$x$ and $y$ are to each other. Let $\varepsilon$ be the tolerance level which
+indicates whether a data set is ``close'' to the observed data. That is, if
+$d(x, y) < \varepsilon$, we will say that $x$ and $y$ are close, otherwise they
+are distant.
+
+% begin by sampling from the prior and setting all weights to equal
+
+Next, we calculate the next tolerance $\varepsilon^*$. Before we explain how
+this is done, we first need to define how we adjust the weights on the
+particles. As explained above (see subsection~\ref{subsubsec:abcalg}), the idea
+of sequential Monte-Carlo is to begin with the prior distribution $\pi$,
+progress smoothly through a series of intermediate distributions $\pi_1,
+\ldots, \pi_{n-1}$, and eventually arrive at the target posterior distribution
+$\pi_n$. In the $k$th iteration, the distribution $\pi_k$ is approximated by
+the particles and their weights.
+
+In contrast to most existing sequential Monte-Carlo methods, this
+algorithm does not require the user to specify a sequence of decreasing
+tolerances to approach the target posterior distribution. Rather, the
+tolerances are computed adaptively at each step, starting from infinity at the
+first iteration.
+
+The algorithm may be stopped when the tolerance reaches a
+user-defined final value, or when the rate of acceptance of the
+Metropolis-Hastings kernel reaches a user-defined threshold. Following a
+heuristic applied by the authors~\autocite{del2012adaptive}, we used the latter
+stopping criterion, accepting the SMC approximation to the posterior when the
+\gls{MCMC} acceptance rate dropped below 1.5\%.
 
@@
 the procedure and results for the preferential attachment networks. The details
-of the other three types of graph can be found in the supplemental materials.
+of the other three types of graph can be found in the supplemental materials. A
+graphical schematic of the analysis undertaken here is given in
+Figure~\ref{fig:kernelexpt}.
+
+\begin{figure}[ht]
+  \centering
+  \label{fig:kernelexpt}
+  \includegraphics{kernel_expt}
+  \caption[Schematic of first set of simulation experiments]{
+    Schematic of first set of simulation experiments to determine separable
+    parameters in kernel space, along with optimal tree kernel meta-parameters.
+  }
+\end{figure}
 
@@
 completeness. As a concrete example, consider the attachment power parameter
-$\alpha$ of the preferential attachment networks. This parameter describes the
+\gls{alpha} of the preferential attachment networks. This parameter describes the
 strength of attraction to highly connected nodes and is bounded below by zero
@@
 transmission tree. This resulted in 300 total simulated transmission trees -
-100 for each of the three values of $\alpha$. Next, I computed the tree
-kernel~\autocite{poon2013mapping} (see subsection~\ref{subsubsec:treeshape})
-for each pair of trees. These values were placed into a 300 $\times$ 300 kernel
-matrix, where the value at the $(i, j)$th position was the tree kernel of the 
-$i$th and $j$th trees.
+100 for each of the three values of $\alpha$. The data generation steps are
+shown on the left side of Figure~\ref{fig:kernelexpt}. Next, I computed the
+tree kernel~\autocite{poon2013mapping} (see
+subsection~\ref{subsubsec:treeshape}) for each pair of trees. These values were
+placed into a 300 $\times$ 300 kernel matrix, where the value at the $(i, j)$th
+position was the tree kernel of the $i$th and $j$th trees.
 
@@
 in a variety of contexts, the same analysis was performed with 500 and 2000
-infected nodes, as well as 100 and 1000 sampled tips. The projections show that
-all three values are well separated from each other in many of the scenarios.
-With smaller trees, it becomes harder to disinguish $\alpha = 0.5$ from $\alpha
-= 1.0$ in two dimensions.
+infected nodes, as well as 100 and 1000 sampled tips.
+
+The next step was to quantify how well trees simulated with different $\alpha$
+values could be distinguished from each other. As described
+previously~\autocite{poon2015phylodynamic}, this was done by assessing the
+accuracy of a support vector machine regression
+(SVR)~\autocite{smola1997support}. Briefly, an SVR operates by finding a
+hyperplane (in two dimensions, a line) such that the deviation of most of the
+data points from the line is less than some prescribed threshold. Points
+further away than this threshold are ignored in the model. I performed 1000
+replicate 2-fold cross-validations of an SVR predicting preferential attachment
+power on the simulated trees, using the \software{ksvm} function from
+\software{kernlab}~\autocite{karatzoglou2004kernlab}. That is, the SVR was
+trained on a random subset of 150 trees, and then used to predict $\alpha$ of
+the remaining 150 trees. The predictions were correlated against the true
+values of $\alpha$ to obtain an $R^2$, and this procedure was repeated 1000
+times with different subsets of trees.
+
+The cross-validation had the dual purpose of providing a means to selecting the
+optimal meta-parameters to the tree kernel (see
+subsection~\ref{subsubsec:treeshape}) - they are be those which provide the
+highest average $R^2$. To this end, the cross-validation was repeated for
+several values of $\lambda$ and $\sigma$ as shown in Figure~\ref{fig:kernelexpt}
+Each combination was evaluated with and without multiplying the tree kernel by
+the normalized lineages-through-time (nLTT) statistic, and was further repeated
+for the scenarios with differing numbers of infected and sampled nodes
+described above. To ensure that the tree kernel was the most appropriate
+similarity measure to use for ABC, we also computed the $R^2$ of $\alpha$
+against Sackin's index, a widely used tree balance statistic (see
+subsection~\ref{subsubsec:treeshape}), by the same cross-validation procedure.
+
+\subsubsection{Grid search}
+
+The previous set of simulations were intended to investigate which contact
+network parameters could and could not be inferred by examining transmission
+trees. However, they tell us nothing about the accuracy or precision we might
+expect when inferring those parameters numerically. As illustrated in
+Figure~\ref{fig:accprec}, the ideal situation is one where we are accurate and
+precise, and the worst situation is when we are precise but not accurate.
 
@@
   \centering
-  \label{fig:pakpca}
-  \includegraphics{kernel_pa_kpca}
-  \caption[Projection of kernel matrix for different attachment power values
-  onto its first two principal components]{
-    Projection of the kernel matrix for different preferential attachment power
-    values onto its first two principal components, for eight simulation
-    scenarios. Each point corresponds to a simulated transmission tree, and is
-    coloured by preferential attachment power. Facets are number of infected
-    nodes (horizontal), and number of sampled tips (vertical). The parameters
-    to the tree kernel were $\lambda = 0.3$ and $\sigma = 5$, and the nLTT was
-    not used. Qualitatively, trees with a larger number of tips are easier to
-    separate in kernel space, regardless of what sampling proportion they
-    represent. In all cases, the highest attachment power can be separated from
-    the other two, but the two lowest values become hand to distinguish with in
-    the 100-tip trees.
-  }
+  \includegraphics{acc_prec}
+  \caption{Illustration of accurate vs. precise kernel score estimates}
+  \label{fig:accprec}
 \end{figure}
+
+Figure~\ref{fig:gridsearch} shows a schematic of this experiment. A number of
+representative values, here denoted $k$, were chosen for the parameter of
+interest. In the case of preferential attachment power $\alpha$, I chose $k =
+8$ testing values, namely $0, 0.25, \ldots, 2.0$. For each of these values, ten
+networks on 5000 nodes were generated, an epidemic of 1000 nodes was simulated
+over each, and a transmission tree of 500 tips was sampled. The resulting $10
+\times k$ simulated transmission trees were referred to as the \defn{testing
+trees}. In addition, I chose a further $n \gg k$ training values spanning the
+range of the parameter. For $\alpha$, these were $0, 0.01, \ldots, 2.0$.
+Fifteen trees were simulated in the same manner for each of these values,
+referred to here as \defn{training trees}.
+
+For each of the testing trees, I computed the tree kernel with all of the
+training trees. This resulted in 15 kernel scores per training value. The
+training value with the highest median kernel score was used as a point
+estimate of the testing value. The kernel scores were then normalized to lie in
+[0, 1], and an interval was found which contained 95\% of the area under the
+curve and was of minimal width. This was used as a 95\% confidence interval for
+the parameter.
 
@@
   \centering
-  \label{fig:pacrossv}
-  \includegraphics{kernel_pa_crossv}
-  \caption[Cross-validation performance of kernel support vector machine
-  classifier for preferential attachment power]{
-    Cross-validation performance of kernel support vector machine classifier
-    for preferential attachment power.
-  }
+  \includegraphics{gridsearch_expt}
+  \caption{Schematic of grid search simulation experiments.}
+  \label{fig:gridsearch}
 \end{figure}
 
-\subsubsection{Grid search}
+\subsubsection{Approximate Bayesian computation}
 
-\subsubsection{Approximate Bayesian computation}
+Our final set of simulation experiments was to test the ABC-SMC algorithm on
+simulated data. For each replicate, we generated a network on 5000 nodes,
+simulated an epidemic over that network until 1000 nodes were infected, and
+sampled either 500 or all 1000 of those nodes to form a transmission tree.
+Priors were specified as follows: for the total number of nodes in the network,
+$\Uniform(1000, 10000)$; for the number of infected nodes, $\Uniform(500,
+2000)$; for the preferential attachment power, $\Uniform(0, 2)$. The mean
+degree of nodes in the network was fixed at either 4, 10, or 16. The
+algorithm was run with 1000 particles, and was stopped when the \gls{MCMC}
+acceptance probability dropped below 1.5\%.
 

### 2016-01-26 doc/thesis/objective.tex
Writing yay

Our objective is to develop a method to estimate structural parameters of the
contact network underlying an observed transmission tree.

\subsection{Prior work}

The present study is closely related to three groups of work within
computational epidemiology, which will be discussed in detail below. The first
and largest of these are phylodynamic investigations of epidemiological
parameters, such as transmission rate, recovery rate, and basic reproductive
number. Like our work, these studies make inferences from evolutionary
information in the form of a viral phylogeny, but their aims do not include
uncovering population structure and they typically assume a homogeneously mixed
population. A second, more recent group of studies has evaluated the effect of
network structure on transmission tree shape, by comparing trees simulated from
different network models. These have demonstrated that network structure does
profoundly influence tree shape, but did not attempt to quantitatively infer
the parameter from observed trees. The third group consists of studies which
infer a structural network parameter from namely observed infection and
recovery times using Bayesian methods. These are most similar in aims to our
work, but they rely on epidemiological observations rather than evolutionary
information from phylogenies.

In recent years, a large number of studies have been carried out using
phylodynamic methods to infer epidemiological quantities of interest. The
majority of these employ a Bayesian \gls{MCMC} approach to infer parameters of
an epidemiological model whose likelihood can be calculated, most often some
variation of the birth-death~\autocite{kendall1948generalized} or
coalescent~\autocite{kingman1982coalescent} models. Stadler
\etal~\autocite{stadler2011estimating} develop a formula for the likelihood of
a phylogeny with heterochronous tips under the birth-death model, which has
been used to estimate the basic reproductive number of several viral
epidemics~\autocite{stadler2011estimating}. However, the birth-death model is
cannot tell us anything about population structure, as it assumes that every
individual becomes infected at the same rate. Volz~\autocite{volz2012complex}
writes down the likelihood of a heterochronous phylogeny under a coalescent
model with arbitrarily complex population dynamics. This opens the door to more
complex inferences about population structure, as the population can be
partitioned into compartments with different transmission and recovery rates,
but still assumes that each compartment is homogeneously mixed. In other words, 
the coalescent model can tell us about the \emph{global} structure of a
population, such as whether there exists a high-risk subgroup, but not about
the \emph{local} structure, such as the average number of contacts each
individual has.

The use of models which assume a \defn{panmictic} (that is, homogeneously
mixed) population in phylodynamics has become widespread, so it is natural that
some researchers have investigated the effects of this assumption. In
particular, since phylodynamic methods use phylogenies as their data source,
several studies have examined the shapes of phylogenies arising from
non-panmictic populations. \citeauthor{o2010contact}~(\citeyear{o2010contact})
simulated epidemics over networks with four types of degree distribution. They
then estimated the Bayesian skyride~\autocite{minin2008smooth} population size
trajectory in two ways: from the phylogeny, using \gls{MCMC}; and from the
incidence and prevalence trajectories, using the method developed by Volz
\etal~\autocite{volz2009phylodynamics}. They found that the concordance between
the two skyrides, as well as the relationship between the skyride and
prevalence curve, was qualitatively different for each degree distribution.
\citeauthor{leventhal2012inferring} simulated transmission trees over \gls{ER},
\gls{WS}, \gls{BA}, and full networks with fixed number of nodes and mean degree.
They calculated Sackin's index of the simulated trees while varying the
epidemic, network, and sampling parameters, and found that the relationship
between these parameters and Sackin's index varies considerably among the
different network models.

The third group of studies was pioneered by Britton and
O'Niell~\autocite{britton2002bayesian}, who developed a Bayesian method to
infer the $p$ parameter of an \gls{ER} network, along with the $\beta$ and
$\gamma$ parameters of \gls{SI} model. Their method is able to use either
infection and removal times, or removal times only. However, it is designed for
only a small number of observations (15 and 42 cases in their applications),
and their estimates of $p$ for real outbreaks were mostly uninformative (95\%
confidence intervals [0.11-0.96] and 0.055-0.96]; $p$ is bounded in [0, 1]).

### 2016-01-26 doc/thesis/results.tex
Writing yay

\subsection{Estimable parameters}

\begin{figure}[ht]
  \centering
  \label{fig:pacrossv}
  \includegraphics{kernel_pa_crossv}
  \caption[Cross-validation performance of kernel support vector machine
  classifier for preferential attachment power]{
    Cross-validation performance of kernel support vector machine classifier
    for preferential attachment power.
  }
\end{figure}

The projections show that all three values are well separated from each other
in many of the scenarios. With smaller trees, it becomes harder to disinguish
$\alpha = 0.5$ from $\alpha = 1.0$ in two dimensions.

\begin{figure}[ht]
  \centering
  \label{fig:pakpca}
  \includegraphics{kernel_pa_kpca}
  \caption[Projection of kernel matrix for different attachment power values
  onto its first two principal components]{
    Projection of the kernel matrix for different preferential attachment power
    values onto its first two principal components, for eight simulation
    scenarios. Each point corresponds to a simulated transmission tree, and is
    coloured by preferential attachment power. Facets are number of infected
    nodes (horizontal), and number of sampled tips (vertical). The parameters
    to the tree kernel were $\lambda = 0.3$ and $\sigma = 5$, and the nLTT was
    not used. Qualitatively, trees with a larger number of tips are easier to
    separate in kernel space, regardless of what sampling proportion they
    represent. In all cases, the highest attachment power can be separated from
    the other two, but the two lowest values become hand to distinguish with in
    the 100-tip trees.
  }
\end{figure}

\subsection{Accuracy and precision}

\begin{figure}[ht]
  \centering
  %\includegraphics{gridsearch_pa_kernel}
  \caption{Grid search kernel scores.}
  \label{fig:gridkernel}
\end{figure}

### 2016-01-26 doc/thesis/figures/acc_prec.R

@@ -1,0 +1,36 @@
+#!/usr/bin/env Rscript
+
+library(ggplot2)
+library(data.table)
+library(extrafont)
+
+set.seed(0)
+
+dens1 <- density(rnorm(400))
+dens2 <- density(runif(400, -3, 3))
+dens3 <- density(rnorm(400))
+dens4 <- density(runif(400, -3, 3))
+
+d <- data.table(x=c(dens1$x, dens2$x, dens3$x, dens4$x), 
+                y=c(dens1$y, dens2$y, dens3$y, dens4$y), 
+                precise=rep(c("yes", "no", "yes", "no"), each=length(dens1$x)),
+                accurate=rep(c("yes", "no"), each=length(dens1$x)*2))
+d[,precise := factor(precise, levels=c("yes", "no"))]
+d[,accurate := factor(accurate, levels=c("yes", "no"))]
+d[,col := ifelse(precise == "yes", ifelse(accurate == "yes", "green", "red"), "orange")]
+
+lines <- data.frame(
+    v=c(0, dens2$x[which.max(dens2$y)], 2, 1),
+    precise=c("yes", "no", "yes", "no"),
+    accurate=c("yes", "yes", "no", "no")
+)
+
+pdf("acc_prec.pdf", height=4, family="Gillius ADF")
+ggplot(d, aes(x=x, y=y, color=col)) + geom_line() + theme_bw() + xlim(-4, 4) +
+    geom_vline(data=lines, aes(xintercept=v), lty=2) +
+    facet_grid(precise~accurate, scales="free", labeller="label_both") +
+    theme(axis.ticks=element_blank(), legend.position="none",
+          panel.grid=element_blank(), axis.text=element_blank()) +
+    scale_color_manual(values=c("green", "orange", "red")) +
+    labs(x="testing value", y="kernel score")
+dev.off()

### 2016-01-26 doc/thesis/figures/kernel_expt.R

@@ -1,0 +1,36 @@
+#!/usr/bin/env Rscript
+
+library(igraph)
+library(ape)
+
+nnode <- 7
+ninf <- 4
+set.seed(0)
+
+pow <- c(0.5, 1, 1.5)
+col <- c("red", "blue", "green")
+inf.col <- "black"
+
+for (i in 1:3) {
+    g <- sample_pa(nnode, m=2, power=pow[i], directed=FALSE)
+    png(sprintf("tinynet%d.png", i), bg="transparent", width=100, height=100)
+    par(mar=c(0, 0, 0, 0) + 0.1)
+    plot(g, vertex.size=10, vertex.label=NA, vertex.color=col[i],
+         vertex.frame.color=col[i], edge.color=col[i], edge.width=5)
+    dev.off()
+
+    png(sprintf("tinyepi%d.png", i), bg="transparent", width=100, height=100)
+    par(mar=c(0, 0, 0, 0) + 0.1)
+    ecol <- rep(col[i], ecount(g))
+    vcol <- rep(c(inf.col, col[i]), c(ninf, vcount(g)-ninf))
+    ecol[sample(unique(unlist(g[[1:ninf,1:ninf,edges=TRUE]])), ninf-1)] <- inf.col
+    plot(g, vertex.size=10, vertex.label=NA, vertex.color=vcol,
+         vertex.frame.color=vcol, edge.color=ecol, edge.width=5)
+    dev.off()
+
+    png(sprintf("tinytree%d.png", i), bg="transparent", width=100, height=100)
+    par(mar=c(0, 0, 0, 0) + 0.1)
+    t <- rcoal(ninf)
+    plot(t, direction="down", edge.width=5, edge.color=col[i], show.tip.label=FALSE)
+    dev.off()
+}

### 2016-01-26 doc/thesis/figures/nltt.R

@@ -1,6 +1,7 @@
 #!/usr/bin/env Rscript
 
 library(ape)
+library(extrafont)
 
 t1 <- read.tree(text="(((:1,:1):1,:1):1,:1):0;")
 t2 <- read.tree(text="((((:1,:1):1,:1):1,:1):1,:1):0;")
@@ -11,7 +12,7 @@ sf2 <- stepfun(c(0, 1/3, 2/3), c(0, 1/3, 2/3, 1))
 af1 <- approxfun(c(0, 0.5, 1), c(0, 0.5, 1))
 af2 <- approxfun(c(0, 1/3, 2/3, 1), c(0, 1/3, 2/3, 1))
 
-pdf("nltt.pdf")
+pdf("nltt.pdf", family="Gillius ADF")
 par(mfrow=c(2, 2), lwd=3)
 old.mar <- par("mar")
 par(mar=c(0, 2, 4, 2) + 0.1)

### 2016-01-26 doc/thesis/figures/separable.R

@@ -5,12 +5,12 @@ library(png)
 library(raster)
 library(extrafont)
 
-set.seed(4)
+set.seed(0)
 
-tree.col <- c("black", rep(c("blue", "red", "forestgreen"), each=3))
+tree.col <- c("black", rep(c("blue", "red"), each=3))
 tmp <- tempfile()
-tinytree <- as.list(rep(0, 10))
-for (i in 1:10) {
+tinytree <- as.list(rep(0, 7))
+for (i in 1:7) {
     png(tmp, width=50, height=50, bg="transparent")
     plot(rtree(4), show.tip.label=F, direction="down", no.margin=TRUE, edge.width=5,
          edge.color=tree.col[i])
@@ -23,24 +23,29 @@ pdf("separable.pdf", width=6, height=2, family="Gillius ADF")
 par(mar=c(0, 0, 0, 0) + 0.1, mfrow=c(1, 3))
 
 plot(NA, xlim=c(0, 1), ylim=c(0, 1), axes=FALSE, xlab=NA, ylab=NA, frame.plot=TRUE)
-x <- c(rep(c(0.2, 0.4, 0.8), each=3) + rnorm(9, sd=0.1))
-y <- c(rep(c(0.7, 0.4, 0.6), each=3) + rnorm(9, sd=0.1))
+x <- c(0.2, 0.3, 0.1, 0.55, 0.75, 0.65)
+y <- c(0.65, 0.75, 0.7, 0.4, 0.45, 0.35)
 rasterImage(tinytree[[1]],0.5,0.5,0.6,0.6)
-mapply(rasterImage, tinytree[2:10], as.list(x), as.list(y), as.list(x+0.1), as.list(y+0.1))
+mapply(rasterImage, tinytree[2:7], as.list(x), as.list(y), as.list(x+0.1), as.list(y+0.1))
+abline(a=0.1, b=1, lty=2)
 text(0.05, 0.95, "A", cex=2)
 
 plot(NA, xlim=c(0, 1), ylim=c(0, 1), axes=FALSE, xlab=NA, ylab=NA, frame.plot=TRUE)
-x <- c(rep(c(0.2, 0.4, 0.8), 3) + rnorm(9, sd=0.2))
[…]

### 2016-01-29 doc/thesis/intro.tex
More writing

@@
-\subsection{Phylogenetics and phylodynamics}
+\section{Phylogenetics and phylodynamics}
 
-\subsubsection{Phylogenetic trees}
+\subsection{Phylogenetic trees}
 
@@
 the DNA sequences of an ancestor its descendant, or the difference in average
-body mass or height. This will be clarified below in
-subsection~\ref{subsubsec:genetree}. A phylogeny with branch lengths in calendar
-time units is often referred to as \defn{time-scaled}. In a time-scaled
-phylogeny, the internal nodes can be mapped onto a timeline by using the tips
-of the tree, which usually map to the present day, as a reference
-point~\autocite{nee1992tempo}. The corresponding points on the timeline are
-called \defn{branching times}, and the rate of their accumulation is referred
-to as the \defn{branching rate}. Rooted trees whose tips are all the same
-distance from the root are called \defn{ultrametric}
+body mass or height. Particular examples of evolutionary distance, based on
+genetic data, will be discussed below in subsection~\ref{subsubsec:genetree}. A
+phylogeny with branch lengths in calendar time units is often referred to as
+\defn{time-scaled}. In a time-scaled phylogeny, the internal nodes can be
+mapped onto a timeline by using the tips of the tree, which usually map to the
+present day, as a reference point~\autocite{nee1992tempo}. The corresponding
+points on the timeline are called \defn{branching times}, and the rate of their
+accumulation is referred to as the \defn{branching rate}. Rooted trees whose
+tips are all the same distance from the root are called \defn{ultrametric}
 trees~\autocite{buneman1974note}. These concepts are illustrated in
@@
 
-\subsubsection{Species trees and transmission trees}
+\subsection{Species trees and transmission trees}
 \label{subsubsec:speciestree}
@@
 access to the true tree relating a particular group of extant species, as this
-would imply certain knowledge of speciation events in the distant past.
+would usually imply certain knowledge of speciation events in the distant past.
 Therefore, all species trees considered by researchers are estimates, based on
@@
 
-Speciation, which causes branching in a species tree, often occurs by an
-\defn{allopatric} process, where a sub-population of organisms is isolated from
-the original population by a geographic barrier~\autocite{coyne2004speciation}.
-Over time, differing selection pressures on either side of the barrier, coupled
-with genetic drift, cause the two populations to diverge genetically. This
-eventually results in two distinct species, which will appear as two subtrees
-or \defn{clades} in their species tree if descendants of both survive to the
-present. Of course, this is not necessarily the case, as selection may cause
-one or both of the new sub-populations or their descendants to become extinct.
-On the other hand, selection may favour one of the sub-populations, allowing it
-to expand into new regions or niches and possibly paving the way for subsequent
-speciation events. It must be noted that speciation is not necessarily
-allopatric. \defn{Sympatric} speciation refers to the emergence of a new
-species without any geographic barrier~\autocite{coyne2004speciation}. Indeed,
-there is a whole continuum of possible speciation
+Speciation often occurs by an \defn{allopatric} process, where a sub-population
+of organisms is isolated from the original population by a geographic
+barrier~\autocite{coyne2004speciation}. Over time, differing selection
+pressures on either side of the barrier, coupled with genetic drift, cause the
+two populations to diverge genetically. This eventually results in two distinct
+species, which will appear as two subtrees or \defn{clades} in their species
+tree if descendants of both survive to the present. Speciation is not
+necessarily allopatric: \defn{sympatric} speciation refers to the emergence of
+a new species without any geographic barrier~\autocite{coyne2004speciation}.
+Indeed, there is a whole continuum of possible speciation
 processes~\autocite{fitzpatrick2008if}, ranging from completely allopatric to
-completely \defn{sympatric}. However, at least for eukaryotic organisms,
-allopatric speciation is thought to be by far the most common
+completely sympatric. However, at least for eukaryotic organisms, allopatric
+speciation is thought to be by far the most common
 type~\autocite{coyne2004speciation}.
+
+%Of course, this is not
+%necessarily the case, as selection may cause one or both of the new
+%sub-populations or their descendants to become extinct. On the other hand,
+%selection may favour one of the sub-populations, allowing it to expand into new
+%regions or niches and possibly paving the way for subsequent speciation events.
 
@@
 transmission pair, if this is known. The tips of the tree, rather than being
-fixed at the present day, are placed at the time at which the individual was
-discovered or their pathogen population sampled. Consequently, the transmission
-tree may not be ultrametric, but may have tips located at varying distances
-from the root. Such trees are said to have \defn{heterochronous} taxa or
+fixed at the present day, are placed at the time at which the pathogen
+population of the individual was sampled. Consequently, the transmission tree
+may not be ultrametric, but may have tips located at varying distances from the
+root. Such trees are said to have \defn{heterochronous} taxa or
 samples~\autocite{drummond2003measurably}, in contrast to the
 \defn{isochronous} samples found in macro-organisms species trees. A
-transmission tree is illustrated in Figure~\ref{fig:contactnet} (right).
+transmission tree is illustrated in Figure~\ref{fig:contactnet} (A).
 
-When considering epidemics of RNA viruses, transmission trees and species trees
+\textit{A priori}, transmission trees and species trees have nothing to do with
+each other. One represents the evolutionary history of a group of organisms,
+the other traces the spread of a disease through a host population. However,
+when considering epidemics of RNA viruses, transmission trees and species trees
 are highly related. The viral population within a single host has often been
@@
 
-\subsubsection{Gene trees and viral phylogenies}
+\subsection{Gene trees and viral phylogenies}
 \label{subsubsec:genetree}
@@
 
-\subsubsection{Estimating transmission trees}
+\subsection{Estimating transmission trees}
 \label{subsubsec:treeconv}
@@
 Most applications of viral phylodynamics, including this project, aim to answer
-epidemiological questions~\autocite{volz2013viral}. For example, phylodynamic
-methods have been used to investigate the degree of
+epidemiological questions~\autocite{pybus2009evolutionary, volz2013viral}. For
+example, phylodynamic methods have been used to investigate the degree of
 clustering~\autocite{hughes2009molecular} and the effect of elevated
@@
 viral phylogeny, which represents a biological process (viral evolution). As
-discussed above in subsection~\ref{subsubsec:speciestree}, transmission trees and
-are in practice unknown, and must be estimated from available data. Here, we
-review some of the ways to perform such estimation.
+discussed above in subsection~\ref{subsubsec:speciestree}, transmission trees
+and are in practice unknown, and must be estimated from available data. Here,
+we review some of the ways to perform such estimation.
 
@@
 
-\subsubsection{Tree shapes}
+\subsection{Tree shapes}
 \label{subsubsec:treeshape}
@@
 phylogenies based on their properties. One of the most widely used is Sackin's
-index, which measures the imbalance or asymmetry in the tree.
+index~\cite{shao1990tree}, which measures the imbalance or asymmetry in the
+tree.
 
-\subsubsection{Applications of phylodynamics}
+\subsection{Applications of phylodynamics}
 \label{subsubsec:appphylo}
 
-\subsection{Contact networks}
+\section{Contact networks}
 \label{subsec:contactnet}
 
-\subsubsection{Overview}
+\subsection{Overview}
 
@@
 
-\subsubsection{Models for contact networks}
+\subsection{Models for contact networks}
 \label{subsubsec:generative}
@@
 
-%\subsection{Approximate Bayesian computation}
+%\section{Approximate Bayesian computation}
 
@@
 
-\subsection{Approximate Bayesian computation}
+\section{Approximate Bayesian computation}
 
-\subsubsection{Overview and motivation}
+\subsection{Overview and motivation}
 
-Mathematical models can be grouped into three categories in terms of our
-ability to fit them to observed data. First, there is a limited set of models
-which can be fit explicitly by using calculus to minimize a loss function. The
-canonical example of this type of model is a linear regression, with the
-ordinary least-squares loss function. A linear regression is a model of the
-relationship between some data points $x_1, \ldots, x_n$ and outcomes $y_1,
-\ldots, y_n$ which assumes that the outcomes are linearly related to the data,
-modulo some noise caused by measurement errors, environmental fluctuations, and
-so on. In other words, there is a constant $b$ such that 
+\gls{ABC} is the name for a family of techniques which can be used to fit
+complex models to observed data. We shall make this precise below, but to fully
+describe and motivate \gls{ABC}, it is necessary to first make explicit exactly
+what we mean by ``fitting'' a model.
+
+A linear regression is a model of the relationship between some data points
+$\vec{x} = (x_1, \ldots, x_n)$ and outcomes $\vec{y} = (y_1, \ldots, y_n)$.
+This model assumes that the outcomes are linearly related to the data, modulo
+some noise caused by measurement errors, environmental fluctuations, and so on.
+In other words, there is a constant $b$ such that $y_i = bx_i + \varepsilon_i$,
+where $\varepsilon_i$ is the error associated with the outcome $y_i$. If we
+make the additional assumption that the errors are normally distributed, then
+the model takes the form
 \begin{align}
+  y_i = bx_i + \N(0, \sigma^2).
   \label{eq:regression}
-  y_i = bx_i + \varepsilon_i, 
 \end{align}
-where $\varepsilon_i$ is the error associated with the outcome $y_i$. Notice
-that, if we fix $b$, then since $y_i$ and $x_i$ are both given, $\varepsilon_i$
-is also fixed. The process of fitting a linear model means choosing the
-``best'' $b$. For least squares, this optimal $b$ is defined as the one which
-minimizes the sum of the squared $\varepsilon_i$.
+When we say that we want to fit this model to observed data, we mean that we
+want to find particular values for $b$ and $\sigma$ which are in some way
+optimal given $\vec{x}$ and $\vec{y}$. We shall make this precise below. Notice
+that, if we fix $b$ and $\sigma$, then we can write down the probability density
+of observing $y_i$ given $x_i$. If $f_\N(\cdot \mid \mu, \sigma)$ is the
+\gls{pdf} of the normal distribution with mean $\mu$ and variance $\sigma^2$,
+and the $y_i$ are all independent, then
+\[
+  f(\vec{y} \mid \vec{x}, b, \sigma) = 
+  \prod_{i=1}^n f_\N(y_i - bx_i \mid 0, \sigma^2).
+\]
+This gives us a natural criterion for choosing the optimal parameters: we want
+to pick $b$ and $\sigma$ which define a model where the probability density of
+our observed data is as high as possible. When performing such an optimization,
+the above density function is also called the \defn{likelihood} of the
+parameters,
+\[
+  \L(b, \sigma \mid \vec{x}, \vec{y}) = f(\vec{y} \mid \vec{x}, b, \sigma).
+\]
+The particular $b$ and $\sigma$ which optimize $\L(b, \sigma)$ are called the
+\gls{ML} estimates.
+
+%Mathematical models can be grouped into three categories in terms of our
+%ability to fit them to observed data. First, there is a limited set of models
+%which can be fit explicitly by using calculus to minimize a loss function. The
+%canonical example of this type of model is a linear regression. 
+%Notice that, if we fix $b$, then since $y_i$ and $x_i$ are both given,
+%$\varepsilon_i$ is also fixed. The process of fitting a linear model means
+%choosing the ``best'' $b$. For least squares, this optimal $b$ is defined as
+%the one which minimizes the sum of the squared $\varepsilon_i$.
 
@@
 
-\subsubsection{Algorithms for ABC}
+\subsection{Algorithms for ABC}
 \label{subsubsec:abcalg}
@@
 % MCMC
+The second method is \gls{ABC}-\gls{MCMC}. This is similar to ordinary Bayesian
+\gls{MCMC}, except that a ratio of distances to the observed data replaces the
+likelihood ratio. The algorithm begins by sampling a single vector of parameter
+values $\theta$ from the prior distribution $\pi(\theta)$. It then proceeds
+iteratively: a new parameter vector $\theta^*$ is chosen according to a
+proposal distribution $q(\theta^* \mid \theta)$. The proposal distribution $q$
+is often taken to be a Gaussian centred at $\theta$. Then $\theta^*$ is
+accepted as the new $\theta$ with probability
+\[
+  \max\left(1, \text{TODO} \right),
+\]
+or discarded otherwise. This process is iterated until some stopping criterion
+is reached, typically a simple limit on the number of steps. After some initial
+number of iterations, known as \defn{burn-in}, parameters $\theta$ are
+routinely sampled. Since points in parameter space are visited in proportion to
+their posterior probability, these samples can be taken to approximate the
+posterior distribution on $\theta$, and can be used to calculate point
+estimates and confidence intervals.
 

### 2016-01-29 doc/thesis/objective.tex
More writing

@@
 
-The present study is closely related to three groups of work within
-computational epidemiology, which will be discussed in detail below. The first
-and largest of these are phylodynamic investigations of epidemiological
-parameters, such as transmission rate, recovery rate, and basic reproductive
-number. Like our work, these studies make inferences from evolutionary
-information in the form of a viral phylogeny, but their aims do not include
-uncovering population structure and they typically assume a homogeneously mixed
-population. A second, more recent group of studies has evaluated the effect of
-network structure on transmission tree shape, by comparing trees simulated from
-different network models. These have demonstrated that network structure does
-profoundly influence tree shape, but did not attempt to quantitatively infer
-the parameter from observed trees. The third group consists of studies which
-infer a structural network parameter from namely observed infection and
-recovery times using Bayesian methods. These are most similar in aims to our
-work, but they rely on epidemiological observations rather than evolutionary
-information from phylogenies.
+The present study is closely related to three groups of work with distinct
+objectives, which will be reviewed in detail below. 
+%, by
+%comparing trees simulated from different network models. These have
+%demonstrated that network structure does profoundly influence tree shape, but
+%did not attempt to quantitatively infer the parameter from observed trees. The
+%third group consists of studies which infer a structural network parameter from
+%observed infection and recovery times, using Bayesian methods. These are most
+%similar in aims to our work, but they rely on epidemiological observations
+%rather than evolutionary information from phylogenies.
 
-In recent years, a large number of studies have been carried out using
-phylodynamic methods to infer epidemiological quantities of interest. The
-majority of these employ a Bayesian \gls{MCMC} approach to infer parameters of
-an epidemiological model whose likelihood can be calculated, most often some
-variation of the birth-death~\autocite{kendall1948generalized} or
+The first and largest group of related studies are phylodynamic investigations
+of epidemiological parameters such as transmission rate, recovery rate, and
+basic reproductive number~\autocite{pybus2009evolutionary, volz2013viral}. Like
+our work, these studies make inferences about epidemiological processes from
+the genetic diversity of virus populations, which is usually represented in the
+form of a phylogeny. The majority of these employ a Bayesian \gls{MCMC}
+approach to infer parameters of an epidemiological model whose likelihood can
+be calculated, most often some variation of the
+birth-death~\autocite{kendall1948generalized} or
 coalescent~\autocite{kingman1982coalescent} models. Stadler
@@
 
-The use of models which assume a \defn{panmictic} (that is, homogeneously
-mixed) population in phylodynamics has become widespread, so it is natural that
-some researchers have investigated the effects of this assumption. In
-particular, since phylodynamic methods use phylogenies as their data source,
-several studies have examined the shapes of phylogenies arising from
-non-panmictic populations. \citeauthor{o2010contact}~(\citeyear{o2010contact})
-simulated epidemics over networks with four types of degree distribution. They
-then estimated the Bayesian skyride~\autocite{minin2008smooth} population size
-trajectory in two ways: from the phylogeny, using \gls{MCMC}; and from the
-incidence and prevalence trajectories, using the method developed by Volz
+A second, more recent group of studies has evaluated the effect of network
+structure on transmission tree shape. The use of models which assume a
+\defn{panmictic} (that is, homogeneously mixed) population in phylodynamics has
+become widespread, so it is natural that some researchers have investigated the
+effects of this assumption. In particular, since phylodynamic methods use
+phylogenies as their data source, several studies have examined the shapes of
+phylogenies arising from non-panmictic populations. 
+\citeauthor{o2010contact}~(\citeyear{o2010contact}) simulated epidemics over
+networks with four types of degree distribution. They then estimated the
+Bayesian skyride~\autocite{minin2008smooth} population size trajectory in two
+ways: from the phylogeny, using \gls{MCMC}; and from the incidence and
+prevalence trajectories, using the method developed by Volz
 \etal~\autocite{volz2009phylodynamics}. They found that the concordance between
@@
 \citeauthor{leventhal2012inferring} simulated transmission trees over \gls{ER},
-\gls{WS}, \gls{BA}, and full networks with fixed number of nodes and mean degree.
-They calculated Sackin's index of the simulated trees while varying the
+\gls{WS}, \gls{BA}, and full networks with fixed number of nodes and mean
+degree. They calculated Sackin's index of the simulated trees while varying the
 epidemic, network, and sampling parameters, and found that the relationship
 between these parameters and Sackin's index varies considerably among the
-different network models.
+different network models. In summary, studies in this group have demonstrated
+that network structure profoundly influences tree shape, but have not attempted
+to quantitatively infer network parameters from observed trees.
 
-The third group of studies was pioneered by Britton and
-O'Niell~\autocite{britton2002bayesian}, who developed a Bayesian method to
-infer the $p$ parameter of an \gls{ER} network, along with the $\beta$ and
-$\gamma$ parameters of \gls{SI} model. Their method is able to use either
-infection and removal times, or removal times only. However, it is designed for
-only a small number of observations (15 and 42 cases in their applications),
-and their estimates of $p$ for real outbreaks were mostly uninformative (95\%
-confidence intervals [0.11-0.96] and 0.055-0.96]; $p$ is bounded in [0, 1]).
+Finally, a third group of studies has used Bayesian methods to infer a
+structural network parameter from observed infection and recovery times. This
+third group is most similar in aims to our own work; rather than using
+epidemiological observations, we employ viral phylogenies. The pioneering study
+in this group was by
+\citeauthor{britton2002bayesian}~(\citeyear{britton2002bayesian}), who
+developed a Bayesian method to infer the $p$ parameter of an \gls{ER} network,
+along with the $\beta$ and $\gamma$ parameters of \gls{SI} model. Their method
+is able to use either infection and removal times, or removal times only.
+However, it is designed for only a small number of observations (15 and 42
+cases in their applications), and their estimates of $p$ for real outbreaks
+were mostly uninformative (95\% confidence intervals [0.11-0.96] and
+0.055-0.96]; $p$ is bounded in [0, 1]).
+\citeauthor{groendyke2011bayesian}~(\citeyear{groendyke2011bayesian})
+significantly updated and extended the methodology of
+\citeauthor{britton2002bayesian}, and applied it to a measles outbreak
+affecting 188 individuals. They were able to obtain a much more informative
+estimate of $p$, although this data set included both symptom onset and
+recovery times for all individuals. They also assumed that the entire network
+was infected, so that the network size was exactly 188. It is unclear how well
+their method would scale to networks with thousands of nodes.

### 2016-01-29 doc/thesis/results.tex
More writing

@@
-\subsection{Estimable parameters}
+\subsection{Separable parameters}
+
+Our first simulation-based experiment was designed to determine which network
+model parameters were potentially estimable by kernel-ABC. To do this, we
+simulated trees under three distinct values of each parameter of interest, and
+evaluated the accuracy of a kernel-\gls{SVM} classifier for the parameter. This
+is the same approach taken by \textcite{poon2015phylodynamic}, and has the dual
+purpose of allowing us to identify the optimal kernel meta-parameters
+\gls{sigma} and \gls{lambda}. The parameters of the \gls{ER}, \gls{WS}, and
+\gls{BA} network models were examined, but we will discuss only the results for
+the $\alpha$ and prevalence parameters of the \gls{BA} model in the main text.
+The results for the remaining models and their parameters can be found in the
+supplemental material.
+
+In Figure~\ref{fig:patrees}, we show that trees simulated under different
+values of \gls{alpha} are visibly quite distinctive. In particular, higher
+values of \gls{alpha} result in networks with a small number of highly
+connected nodes (see Figure~\ref{fig:pabounds}) which, once infected, are
+likely to transmit to many other nodes. This results in a more unbalanced,
+ladder-like structure in the phylogeny. Kernel-\gls{PCA} projections show that
+all three \gls{alpha} values are well separated from each other under several
+different sampling scenarios (Figure~\ref{fig:pakpca}). With smaller trees,
+it becomes harder to visually disinguish \gls{alpha} = 0.5 from \gls{alpha} =
+1.0 using the first two principal components.
+
+With suitably chosen meta-parameters, the accuracy of the kernel-\gls{SVM}
+classifier for \gls{alpha} was very high under a variety of prevalence and
+sampling scenarios (Figure~\ref{fig:pacrossv}). Accuracy was highest, with
+$R^2$ values above 0.95, for the largest trees and complete sampling (bottom
+center panel). However, even for trees of size 100 sampled from an epidemic on
+2000 nodes, the $R^2$ was above 0.8 (top right panel). In all cases, the
+accuracy of a Sackin's index-based classifier was also quite high, at about
+0.75. The fact that Sackin's index is informative of \gls{alpha} was
+unsurprising, given the clear differences in tree balance observed under
+different \gls{alpha} values. The \gls{nltt}
+statistic~\autocite{janzen2015approximate} did not improve the classifier, but
+rather slightly reduced the $R^2$ in most cases.
+
+\begin{figure}[ht]
+  \centering
+  \label{fig:patrees}
+  \includegraphics{kernel_pa_trees}
+  \caption[Visibly distinctive trees simulated under three values of \gls{alpha}]{
+    Epidemics of 1000 infected were simulated on \gls{BA} networks of 5000
+    nodes, with \gls{alpha} equal to 0.5, 1.0, or 1.5. Transmission trees were
+    created by sampling 500 infected nodes. Higher \gls{alpha} values produce
+    networks with a small number of highly-connected nodes, which results in a
+    highly unbalanced, ladder-like tree structure.
+  }
+\end{figure}
 
@@
 
-The projections show that all three values are well separated from each other
-in many of the scenarios. With smaller trees, it becomes harder to disinguish
-$\alpha = 0.5$ from $\alpha = 1.0$ in two dimensions.
-
 \begin{figure}[ht]
@@
     nodes (horizontal), and number of sampled tips (vertical). The parameters
-    to the tree kernel were $\lambda = 0.3$ and $\sigma = 5$, and the nLTT was
+    to the tree kernel were $\lambda = 0.3$ and $\sigma = 4$, and the nLTT was
     not used. Qualitatively, trees with a larger number of tips are easier to
@@
 
-\subsection{Accuracy and precision}
+We also considered the possibility of inferring the number of infected nodes,
+or \defn{prevalence}, under this model. All parameters except \gls{I} were
+fixed at the following values: \gls{N} = 5000, \gls{alpha} = 1.0, and \gls{m} =
+2. As shown in Figure~\ref{fig:prevtrees}, the prevalence had no obvious effect
+on the tree shape. However, a kernel-\gls{SVM} classifier was able to
+distinguish the number of infected nodes with high accuracy ($R^2 > 0.9$;
+Figure~\ref{fig:prevcrossv}). Moreover, the use of the \gls{nltt} statistic
+improved classification accuracy by a small amount, in contrast to the results
+for \gls{alpha}. In contrast, a Sackin's index-based classifier displayed
+extremely poor performance ($R^2$ < 0.1, not shown).
+
+It is important to note that, if we cut off the epidemic simulation when 500
+nodes are infected, the resulting tree will be shorter (in calendar time) than
+if we continue until 2000 nodes are infected. However, this information is not
+used when building the classifier, since the branch lengths in each tree are
+scaled by their mean. Therefore, the high performance of the classifier is due
+to structural differences captured by the tree kernel, rather than the trees
+simply having different heights.
 
@@
   \centering
-  %\includegraphics{gridsearch_pa_kernel}
-  \caption{Grid search kernel scores.}
+  \label{fig:prevtrees}
+  \includegraphics{kernel_nsimnode_trees}
+  \caption[Similarly shaped simulated under three values of \gls{I}]{
+    Epidemics were simulated on networks of size 5000 until \gls{I} = 500,
+    1000, or 2000 nodes were infected. When scaled to the same height, there is
+    no immediately visible distinction in shape between trees simulated under
+    different \gls{I} values.
+  }
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \label{fig:prevcrossv}
+  \includegraphics{kernel_nsimnode_crossv}
+  \caption[Cross-validation performance of kernel-\gls{SVM} classifier for
+           prevalence]{
+    For \gls{BA} networks of size \gls{N} = 5000, a kernel-\gls{SVM} classifier
+    is very accurate at predicting \gls{I}. Epidemics were simulated until
+    \gls{I} = 500, 1000, or 2000 nodes were infected, and either 100 or 500
+    nodes were sampled for inclusion in a transmission tree. The parameters of
+    the \gls{BA} network were \gls{alpha} = 1.0 and \gls{m} = 2. A Sackin's
+    index-based classifier performed very poorly on these data ($R^2 = $ TODO
+    for 100-tip trees and blah for 500-tip trees, not shown).
+  }
+\end{figure}
+
+\subsection{Accuracy of marginal estimates}
+
+We used grid search to obtain \defn{marginal} estimates for each network
+parameter while holding all other parameters fixed. We observed that kernel
+scores were highest at the values of \gls{alpha} on the grid closest to the
+true values, as shown in Figure~\ref{fig:gridkernel}. However, there was a much
+stronger spike in kernel scores near the true value for \gls{alpha} = 1.0 and
+1.25. This is recapitulated when we look at the accuracy of point estimates
+obtained by taking the grid value with the highest median kernel score. As
+shown in Figure~\ref{fig:gridest}, while the estimates are generally close to
+the true value, they are much closer for \gls{alpha} = 1.25 than for the other
+values.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{gridsearch_pa_kernel}
+  \caption[Grid search kernel scores]{
+    Grid search kernel scores for testing trees simulated under various
+    \gls{alpha} values. All epidemics had \gls{I} = 1000 infected nodes, on
+    \gls{BA} networks of size \gls{N} = 5000 with \gls{m} fixed at 2. Colours
+    indicate the number of sampled tips.
+  }
   \label{fig:gridkernel}
 \end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{gridsearch_pa_estimates}
+  \caption[Marginal estimates of \gls{alpha} obtained with grid search]{
+    Marginal estimates of \gls{alpha} obtained with grid search. Training trees
+    were simulated on a narrowly spaced grid of \gls{alpha} values, and
+    compared to testing trees using the tree kernel. The \gls{alpha} value in
+    the grid with the highest median kernel score was taken as the point
+    estimate for the testing tree. These point estimates are shown as black
+    dots. The dashed line is the identity.
+  }
+  \label{fig:gridest}
+\end{figure}
+
+\subsection{Accuracy of estimates with full ABC}

### 2016-01-29 doc/thesis/figures/abc_smc.R

@@ -1,0 +1,54 @@
+#!/usr/bin/env Rscript
+
+# This is really messy I need to fix it up
+
+suppressPackageStartupMessages(library(netabc))
+
+post <- rnorm(500)
+abc <- list(runif(500, -3, 3),
+            c(runif(250, -3, 3), rnorm(250)),
+            c(runif(100, -3, 3), rnorm(400)))
+dens <- lapply(abc, density)
+gradient <- colorRampPalette(c("blue", "white"))(length(dens[[1]]$x)*1.2)
+
+pdf("abc_smc.pdf", family="Gillius ADF", height=4, width=6)
+par(mfrow=c(2, 3), mgp=c(0, 0, 0), mar=c(1, 1, 2, 1) + 0.1, xpd=TRUE)
+for (i in 1:3) {
+    plot(density(post), main=NA, ylab=bquote(f[.(i)](theta*"|"*D)), 
+         xlab=expression(theta), tck=0, xaxt="n", yaxt="n", zero.line=FALSE,
+         xlim=c(-4, 4) + 0.1)
+    points(dens[[i]]$x, dens[[i]]$y, col=gradient, pch=16)
+    if (i == 1)
+        text(grconvertX(0, from="npc", to="user"),
+             grconvertY(1.1, from="npc", to="user"),
+             "A", cex=2)
+}
+
+true.tree <- tiny.tree(4, ultra=TRUE, color="black")
+tree.size <- 0.15
+offset <- tree.size/2
+epsilon <- c(0.45, 0.35, 0.25)
+
+for (i in 1:3) {
+    dist <- sample(dens[[i]]$x, 10, prob=dens[[i]]$y)
+    r <- (dist - min(dens[[i]]$x)) / (max(dens[[i]]$x) - min(dens[[i]]$x))
+    r <- abs(r - 0.5) * 1.2
+    r[r < 0.2] <- rnorm(sum(r < 0.2), mean=0.2, sd=0.05)
+    theta <- (1:length(r)) * 2 * pi / length(r)
+    coords <- polar2rect(r, theta) + 0.5
+    colors <- gradient[as.integer(r * length(gradient))]
+    trees <- lapply(colors, function (x) tiny.tree(ntip=4, color=x, ultra=TRUE))
+    plot.new()
+    for (j in 1:length(trees)) {
+            rasterImage(trees[[j]], coords[j,1]-offset, coords[j,2]-offset, 
[…]

### 2016-02-03 doc/thesis/intro.tex
Updates?

@@
 
+An alternative to summary statistics are \defn{distance measures} on trees.
+Rather than assigning a numerical value to each tree individually, a distance
+measure assocates each pair of trees with a number, indicating how different
+the trees are from each other. Distance measures allow us to identify groups of
+related phylogenies, for example, local epidemics which are undergoing a
+similar pattern of expansion.
+
 \subsection{Applications of phylodynamics}
@@
 
-\gls{ABC} is the name for a family of techniques which can be used to fit
-complex models to observed data. We shall make this precise below, but to fully
-describe and motivate \gls{ABC}, it is necessary to first make explicit exactly
-what we mean by ``fitting'' a model.
+\Gls{ABC} is a statistical method designed to fit complex models, which cannot
+be fit using more conventional methods, to observed
+\comment{these are review articles, Beaumont is specific to popgen, the others
+are general}
+data~\autocite{marin2012approximate, sunnaker2013approximate,
+beaumont2010approximate} We shall make this precise below, but to fully
+describe and motivate \gls{ABC}, it is necessary to first explain exactly what
+we mean by ``fitting'' a model. We will illustrate this concept with one of the
+simplest and most ubiquitous models: linear regression.
 
-A linear regression is a model of the relationship between some data points
-$\vec{x} = (x_1, \ldots, x_n)$ and outcomes $\vec{y} = (y_1, \ldots, y_n)$.
-This model assumes that the outcomes are linearly related to the data, modulo
-some noise caused by measurement errors, environmental fluctuations, and so on.
-In other words, there is a constant $b$ such that $y_i = bx_i + \varepsilon_i$,
+A linear regression is a model of the relationship between some data $\vec{x} =
+(x_1, \ldots, x_n)$ and outcomes $\vec{y} = (y_1, \ldots, y_n)$. This model
+assumes that the outcomes are linearly related to the data, modulo some noise
+introduced by, say, measurement errors and environmental fluctuations. It other
+words, there is a constant $\beta$ such that $y_i = \beta x_i + \varepsilon_i$,
 where $\varepsilon_i$ is the error associated with the outcome $y_i$. If we
@@
 the model takes the form
-\begin{align}
-  y_i = bx_i + \N(0, \sigma^2).
-  \label{eq:regression}
-\end{align}
-When we say that we want to fit this model to observed data, we mean that we
-want to find particular values for $b$ and $\sigma$ which are in some way
-optimal given $\vec{x}$ and $\vec{y}$. We shall make this precise below. Notice
-that, if we fix $b$ and $\sigma$, then we can write down the probability density
-of observing $y_i$ given $x_i$. If $f_\N(\cdot \mid \mu, \sigma)$ is the
-\gls{pdf} of the normal distribution with mean $\mu$ and variance $\sigma^2$,
-and the $y_i$ are all independent, then
 \[
-  f(\vec{y} \mid \vec{x}, b, \sigma) = 
-  \prod_{i=1}^n f_\N(y_i - bx_i \mid 0, \sigma^2).
+  y_i = \beta x_i + \N(0, \sigma^2),
 \]
-This gives us a natural criterion for choosing the optimal parameters: we want
-to pick $b$ and $\sigma$ which define a model where the probability density of
-our observed data is as high as possible. When performing such an optimization,
-the above density function is also called the \defn{likelihood} of the
+where $\N(0, \sigma^2)$ indicates a normally distributed random variable with
+mean 0 and variance $\sigma$. We will denote the \gls{pdf} of this random
+variable by $f_\N(\cdot \mid \mu, \sigma)$. In this formulation, the
+coefficient $\beta$ and the error variance $\sigma$ are the two parameters of
+the model. If the $y_i$ are all independent, then for particular choics of
+$\beta$ and $\sigma$, we can write down the probability density of observing
+$\vec{y}$ given $\vec{x}$.
+\[
+  f(\vec{y} \mid \vec{x}, \beta, \sigma) = 
+  \prod_{i=1}^n f_\N(y_i - \beta x_i \mid 0, \sigma^2).
+\]
+The higher the value of this probability density, the more likely the
+observations $\vec{y}$ are given $\vec{x}$ under the model. This gives us a
+natural criterion for choosing the parameters: we want to pick $\beta$ and
+$\sigma$ which define a model where the probability of our observed data is as
+high as possible. When performing such an optimization for fixed data, the
+density function just written is also called the \defn{likelihood} of the
 parameters,
 \[
-  \L(b, \sigma \mid \vec{x}, \vec{y}) = f(\vec{y} \mid \vec{x}, b, \sigma).
+  \L(\beta, \sigma \mid \vec{x}, \vec{y}) = f(\vec{y} \mid \vec{x}, \beta, \sigma).
 \]
-The particular $b$ and $\sigma$ which optimize $\L(b, \sigma)$ are called the
-\gls{ML} estimates.
+The particular $\beta$ and $\sigma$ which optimize $\L(\beta, \sigma \mid
+\vec{x}, \vec{y})$ are called the \gls{ML} estimates.
 
-%Mathematical models can be grouped into three categories in terms of our
-%ability to fit them to observed data. First, there is a limited set of models
-%which can be fit explicitly by using calculus to minimize a loss function. The
-%canonical example of this type of model is a linear regression. 
-%Notice that, if we fix $b$, then since $y_i$ and $x_i$ are both given,
-%$\varepsilon_i$ is also fixed. The process of fitting a linear model means
-%choosing the ``best'' $b$. For least squares, this optimal $b$ is defined as
-%the one which minimizes the sum of the squared $\varepsilon_i$.
+\Gls{ML} inference makes use only of the data at hand, in this case
+$\vec{x}$ and $\vec{y}$, to estimate the parameters $\beta$ and $\sigma$.
+However, it is frequently the case that the investigator has some additional
+\defn{prior} information or belief about what $\beta$ and $\sigma$ are likely
+to be. For example, the instrument used to measure the $\vec{y}$ may have a
+known margin of error, or the sign of the slope may be obvious from previous
+experiments. A Bayesian approach to fitting this model would make use of this
+information by codifying the investigator's beliefs as a \defn{prior
+distribution}, denoted $f(\beta, \sigma)$. The optimal parameters are then
+those which maximize the product of the prior and the likelihood,
+\[
+  f(\vec{y} \mid \vec{x}, \beta, \sigma) f(\beta, \sigma).
+\]
+Bayes' theorem tells us that this product is proportional to the
+\defn{posterior distribution} $f(\beta, \sigma \mid \vec{y}, \vec{x})$. The
+parameters which optimize the posterior, and hence also optimize the above
+product, are called the \gls{MAP} estimates.
 
-Now that we have defined what it means to fit a least-squares linear
-regression, there remains the question of how to go about finding this optimal
-$b$ value. Recall that the errors $\varepsilon_i$ are fixed for a particular
-$b$, and our objective is to minimize the loss function, which is the sum of
-the squared errors. Naively, we could simply try many different $b$ values,
-calculate the loss function for each, and choose the $b$ which gives the
-smallest value. This approach, though basic, falls under the umbrella of
-\defn{numerical optimization}, which is the most common way the majority of
-models are fit. Of course, there is a whole field devoted to more sophisticated
-methods for exploring the parameter space, but they all boil down to the basic
-method of trying different values and choosing the one which minimizes the
-loss.
+Now that we have defined what it means to fit a linear regression, there
+remains the question of how to go about finding this optimal $\beta$ value.
+Na\"{i}vely, we could simply try many different $\beta$ and $\sigma$ values,
+calculate the likelihood of each, and choose those which yield the highest
+value. This approach, though basic, falls under the umbrella of \defn{numerical
+optimization} of the likelihood function. Of course, there is a whole field
+devoted to more sophisticated methods for exploring the parameter space, but
+they all boil down to the basic idea of trying out different values and
+choosing those which give the highest likelihood.
 
-In the case of least-squares linear regression, there is another method we
-could use. Looking at equation~\ref{eq:regression} for the model, it is easy to
-rearrange terms to solve for $\varepsilon_i$ in terms of the other variables:
-we get $\varepsilon_i = y_i - bx_i$. In fact, we can write down the entire loss
-function, which is the sum of the squared $\varepsilon_i$ in terms of $b$:
-\[
-  \sum \varepsilon_i^2 = \sum (y_i - bx_i)^2,
-\]
-where the sums are over all $i$ from 1 to $n$, the number of points in the data
-set. We will not describe the procedure here, but it is possible to explicitly
-minimize the right-hand side by finding the point where the derivative with
-respect to $b$ is zero. However, the method of directly minimizing the loss
-function with calculus is only applicable to a very narrow class of simple
-models. For the majority, the loss function is too complicated to set the
-derivative with respect to the parameters to zero and solve.
-
-Because of the limited space of models which permit exact solutions, numerical
-optimization of a loss function is the most frequently used method to fit a
-model.
+In the case of linear regression, there is another method we could use. We know
+from calculus that the maximum of $\L$ occurs either on the boundary of its
+domain, or at a point where its partial derivatives with respect to $\beta$ and
+$\sigma$ are both zero. Though we will not show it here, it turns out to be
+straightforward to find the $\beta$ and $\sigma$ values where the partial
+derivatives are zero. In fact, these values co-incide with the least-squares
+estimates of $\beta$ and $\sigma$, which minimize the sum of the squared $x_i -
+\beta y_i$. However, the method of directly minimizing the loss function with
+calculus is only applicable to a narrow class of simple models. For the
+majority, the likelihood function is too complicated to set the partial
+derivatives with respect to the parameters to zero and solve.
 

### 2016-02-03 doc/thesis/figures/abc_smc.R

@@ -1,7 +1,5 @@
 #!/usr/bin/env Rscript
 
-# This is really messy I need to fix it up
-
 suppressPackageStartupMessages(library(netabc))
 
 post <- rnorm(500)
@@ -9,7 +7,7 @@ abc <- list(runif(500, -3, 3),
             c(runif(250, -3, 3), rnorm(250)),
             c(runif(100, -3, 3), rnorm(400)))
 dens <- lapply(abc, density)
-gradient <- colorRampPalette(c("blue", "white"))(length(dens[[1]]$x)*1.2)
+gradient <- colorRampPalette(c("steelblue1", "steelblue1"))(length(dens[[1]]$x)*1.2)
 
 pdf("abc_smc.pdf", family="Gillius ADF", height=4, width=6)
 par(mfrow=c(2, 3), mgp=c(0, 0, 0), mar=c(1, 1, 2, 1) + 0.1, xpd=TRUE)
@@ -40,13 +38,16 @@ for (i in 1:3) {
     trees <- lapply(colors, function (x) tiny.tree(ntip=4, color=x, ultra=TRUE))
     plot.new()
     for (j in 1:length(trees)) {
-            rasterImage(trees[[j]], coords[j,1]-offset, coords[j,2]-offset, 
-                        coords[j,1]+offset, coords[j,2]+offset)
+        rasterImage(trees[[j]], coords[j,1]-offset, coords[j,2]-offset, 
+                    coords[j,1]+offset, coords[j,2]+offset)
+        if (r[j] > epsilon[i]) {
+            points(coords[j,1], coords[j,2], pch=4, col="red", cex=4)
+        }
     }
     rasterImage(true.tree, 0.5-offset, 0.5-offset, 0.5+offset, 0.5+offset)
     symbols(0.5, 0.5, circles=epsilon[i], lty=2, add=T, inches=F)
-    arrows(0.55, -0.05, epsilon[i] + 0.5, -0.05, angle=90, length=0.05, code=3)
-    text(0.8 - i * 0.05, 0, bquote(epsilon[.(i)]))
+    #arrows(0.55, -0.05, epsilon[i] + 0.5, -0.05, angle=90, length=0.05, code=3)
+    #text(0.8 - i * 0.05, 0, bquote(epsilon[.(i)]))
     if (i == 1)
         text(0, 1, "B", cex=2)
 }

### 2016-02-03 programs/netabc/R/nltt.R
Updated install docs for netsim

@@ -1,0 +1,43 @@
+make.nltt.depths <- function (depths, stepfun = TRUE)                            
+{                                                                                
+    # normalize by maximum depth                                                 
+    depths <- depths / max(depths)                                               
+                                                                                 
+    # get number of lineages at each depth                                       
+    lin <- cumsum(as.integer(table(depths))) - 1                                 
+                                                                                 
+    # normalize lineages to be in [0, 1]                                         
+    lin <- lin / max(lin)                                                           
+                                                                                 
+    # make the nLTT functions                                                    
+    if (stepfun) {                                                               
+        stepfun(unique(depths), c(0, lin))                                       
+    }                                                                               
+    else {                                                                       
+        approxfun(unique(depths), lin)                                           
+    }                                                                            
+}                                                                                
+                                                                                    
+make.ltt <- function (t, stepfun = TRUE)                                         
+{                                                                                
+    # get node depths in sorted order; ignore tips                               
+    depths <- node.depth.edgelength(t)                                           
+    depths <- sort(depths[(t$Nnode+2):(2*t$Nnode+1)])                            
+    depths <- round(depths, 8)                                                      
+    make.nltt.depths(depths, stepfun)                                               
+}                                                                                   
+                                                                                    
+#' Calcluate the normalized lineages-through-time statistic.
+#'
+#' @param t1 first tree
+#' @param t2 second tree
+#' @param stepfun if FALSE, use a linear interpolation instead of a step function
+#' @param ... additional arguments passed to integrate
+#' @export
+nLTT <- function (t1, t2, stepfun = TRUE, ...)                        
+{                                                                                   
+    f1 <- make.ltt(t1, stepfun)                                                     
+    f2 <- make.ltt(t2, stepfun)                                                  
+    f <- function (x) abs(f1(x) - f2(x))                                            
+    integrate(f, 0, 1, ...)$value                                                   
+}

### 2016-02-03 doc/thesis/intro.tex
Wrote about tree shapes

@@
 epidemic, the virus, or its hosts and their behaviour, by studying a phylogeny,
-most often a transmission tree. Phylogenies are complex objects, and it is not
-immediately obvious how to extract useful information from them with respect to
-fitting a parameter. Standard statistical methods, built for numeric data
-cannot be applied directly - for example, one cannot perform a regression of a
-parameter of interest against a phylogeny. Therefore, before we discuss exactly
-what phylodynamics can tell us about epidemics, and how it has been applied in
-the past, we will review some of the methods for quantifying the shapes of
-phylogenies and their similarity to each other.
+most often a transmission tree~\autocite{pybus2009evolutionary, volz2013viral}.
+Phylogenies are complex objects, and it is not immediately obvious how to
+extract useful information from them with respect to fitting a parameter.
+Standard statistical methods built for numeric data cannot be applied directly
+- for example, one cannot perform a regression of a parameter of interest
+against a phylogeny. Therefore, before we discuss exactly what phylodynamics
+can tell us about epidemics, and how it has been applied in the past, we will
+review some of the methods for quantifying the shapes of phylogenies and their
+similarity to each other.
 
@@
 phylogenies based on their properties. One of the most widely used is Sackin's
-index~\cite{shao1990tree}, which measures the imbalance or asymmetry in the
-tree.
+index~\cite{shao1990tree}, which measures the imbalance or asymmetry in a
+rooted tree. For the $i$th tip of the tree, we define $N_i$ to be the number of
+branches between that tip and the root. The unnormalized Sackin's index is
+defined as the sum of all $N_i$. It is called unnormalized because it does not
+account for the number of tips in the tree. Among two trees having the same
+number of tips, the least-balanced tree will have the highest Sackin's index.
+However, among two equally balanced trees, the larger tree will have a higher
+Sackin's index. This makes it challenging to compare balances among trees of
+different sizes. To correct this, \textcite{kirkpatrick1993searching} derive
+the expected value of Sackin' index under the Yule
+model~\autocite{yule1925mathematical}. Dividing by this expected value
+normalizes Sackin's index, so that it can be used to compare trees of different
+sizes.
 
@@
 related phylogenies, for example, local epidemics which are undergoing a
-similar pattern of expansion.
+similar pattern of expansion. One such distance measure is the
+\gls{nltt}~\autocite{janzen2015approximate}, which compares the
+\gls{ltt}~\autocite{nee1992tempo} plots of two trees. Specifically, the two
+\gls{ltt} plots are normalized so that they begin at $(0, 0)$ and end at $(1,
+1)$, and the difference between the two plots is integrated between 0 and 1.
+In the context of infectious diseases, the \gls{nltt} is related to the
+prevalence~\autocite{holmes1995revealing}, so large values may indicate that
+the trees being compared are the products of different epidemic
+trajectories~\autocite{janzen2015approximate}.
+
+Another tree distance measure is the \defn{phylogenetic kernel}, or ``tree
+kernel'' developed by \textcite{poon2013mapping}. Strictly speaking, the tree
+kernel is a \defn{similarity measure}, rather than a distance measure, as it is
+maximized when the two trees being compared are the same. The basis of the tree
+kernel is the kernel trick originally developed for
+\glspl{SVM}~\autocite{burges1998tutorial}. The idea of the kernel trick is to
+compare objects by mapping them into a feature space of very high, possibly
+even infinite, dimension. The similarity between objects is taken to be their
+dot product in the feature space. It is called a ``trick'' because this dot
+product is computed using a \defn{kernel function} without explicitly mapping
+the objects to the feature space, which would be computationally prohibtive. In
+the case of the tree kernel, the feature space is the space of all possible
+\defn{subset trees}, which are subtrees that do not necessarily extend all the
+way to the tips. The subset-tree kernel was originally developed for comparing
+parse trees in natural language processing~\autocite{collins2002new} and did
+not incorporate branch length information. The version developed by
+\textcite{poon2013mapping} includes a radial basis function to compare the
+differences in branch lengths, thus incorporating both the trees' topologies
+and their branch lengths in a single similarity score. The tree kernel was
+later shown to be highly effective in differentiating trees simulated under a
+compartmental model with two risk groups of varying contact
+rates~\autocite{poon2015phylodynamic}. In that paper,
+\citeauthor{poon2015phylodynamic} used the tree kernel as the distance function
+in \gls{ABC} (see Section~\ref{sec:abc}), an approach dubbed kernel-\gls{ABC},
+to fit epidemiological models to observed trees.
 
@@
 \section{Approximate Bayesian computation}
+
+\label{sec:abc}
 

### 2016-02-12 doc/thesis/intro.tex
Updated reading and writing

@@
 species, which will appear as two subtrees or \defn{clades} in their species
-tree if descendants of both survive to the present. Speciation is not
-necessarily allopatric: \defn{sympatric} speciation refers to the emergence of
-a new species without any geographic barrier~\autocite{coyne2004speciation}.
-Indeed, there is a whole continuum of possible speciation
-processes~\autocite{fitzpatrick2008if}, ranging from completely allopatric to
-completely sympatric. However, at least for eukaryotic organisms, allopatric
-speciation is thought to be by far the most common
+tree if descendants of both survive to the present. Of course, this is not
+necessarily the case, as selection may cause one or both of the new
+sub-populations or their descendants to become extinct. On the other hand,
+selection may favour one of the sub-populations, allowing it to expand into new
+regions or niches and possibly paving the way for subsequent speciation events.
+Speciation is not necessarily allopatric: \defn{sympatric} speciation refers to
+the emergence of a new species without any geographic
+barrier~\autocite{coyne2004speciation}. Indeed, there is a whole continuum of
+possible speciation processes~\autocite{fitzpatrick2008if}, ranging from
+completely allopatric to completely sympatric. However, at least for eukaryotic
+organisms, allopatric speciation is thought to be by far the most common
 type~\autocite{coyne2004speciation}.
-
-%Of course, this is not
-%necessarily the case, as selection may cause one or both of the new
-%sub-populations or their descendants to become extinct. On the other hand,
-%selection may favour one of the sub-populations, allowing it to expand into new
-%regions or niches and possibly paving the way for subsequent speciation events.
 
@@
 corresponding to the index case, and branching times indicating times of
-transmission. The internal nodes may be labeled with the donor of the
+transmission. The internal nodes may be labelled with the donor of the
 transmission pair, if this is known. The tips of the tree, rather than being
@@
 
-\textit{A priori}, transmission trees and species trees have nothing to do with
-each other. One represents the evolutionary history of a group of organisms,
-the other traces the spread of a disease through a host population. However,
-when considering epidemics of RNA viruses, transmission trees and species trees
-are highly related. The viral population within a single host has often been
-referred to as a \defn{quasispecies}~\autocite{domingo2012viral}, and due to
-the fast evolutionary rate of these viruses, two hosts' viral populations are
-invariably genetically distinct. The transmission process is functionally
-identical to allopatric speciation, involving geographic isolation of a
-sub-population in a new host and subsequent genetic divergence. However, for
-the most part, it is not selection but rather host epidemiology which causes
-certain lineages to proliferate while driving others to extinction. A host who
-recovers, dies, or is isolated equates to an extinction in the transmission
-tree. On the other hand, dense populations in frequent contact often experience
-rapid epidemic growth, which causes similarly rapid accumulation of branching
-points in the tree.
+At first glance, transmission trees and species trees seem to have nothing to
+do with each other. One represents the evolutionary history of a group of
+organisms, the other traces the spread of a disease through a host population.
+However, when considering epidemics of RNA viruses, transmission trees and
+species trees are highly related. The viral population within a single host has
+often been referred to as a \defn{quasispecies}~\autocite{domingo2012viral},
+and due to the fast evolutionary rate of these viruses, two hosts' viral
+populations are invariably genetically distinct. The transmission process is
+functionally identical to allopatric speciation, involving geographic isolation
+of a sub-population in a new host and subsequent genetic divergence. However,
+for the most part, it is not selection but rather host epidemiology which
+causes certain lineages to proliferate while driving others to extinction. A
+host who recovers, dies, or is isolated equates to an extinction in the
+transmission tree. On the other hand, dense populations in frequent contact
+often experience rapid epidemic growth, which causes similarly rapid
+accumulation of branching points in the tree.
 
@@
 Finally, we must note two important points. First, we have not yet considered
-the selection pressures operating within a single host, such as immune
-pressure, because these forces do not directly impact transmission trees unless
-they cause a host to recover or die. Second, we reiterate that both
-transmission trees and species trees are theoretical objects which are almost
-never known with certainty for real data. Indeed, the estimation of
-transmission trees is often problematic, and it is in such estimation that
-within-host forces play a role. Both of these issues are discussed further
-below in subsection~\ref{subsubsec:treeconv}, though we continue to ignore them
-in the next subsection.
+the selection pressures operating within a single host, such as the immune
+response. Second, we reiterate that both transmission trees and species trees
+are theoretical objects which are almost never known with certainty for real
+data. Indeed, the estimation of transmission trees is often problematic, and it
+is in such estimation that within-host forces play a role. Both of these issues
+are discussed further below in subsection~\ref{subsubsec:treeconv}, though we
+continue to ignore them in the next subsection.
 
@@
 Rather than assigning a numerical value to each tree individually, a distance
-measure assocates each pair of trees with a number, indicating how different
+measure associates each pair of trees with a number, indicating how different
 the trees are from each other. Distance measures allow us to identify groups of
@@
 product is computed using a \defn{kernel function} without explicitly mapping
-the objects to the feature space, which would be computationally prohibtive. In
+the objects to the feature space, which would be computationally prohibitive. In
 the case of the tree kernel, the feature space is the space of all possible
@@
 A \defn{Barab\'asi-Albert} (BA) network~\autocite{barabasi1999emergence},
-incorproates the phenomenon of \defn{preferential attachment} seen in real life
+incorporates the phenomenon of \defn{preferential attachment} seen in real life
 networks. This means that new connections are more likely to involve nodes
@@
 observed network, but this cannot be done for the other two models. This is
-because the probability of a particular edge occuring is always equal to $p$ in
+because the probability of a particular edge occurring is always equal to $p$ in
 an ER network, regardless of the presence or absence of any other edges. In
@@
 
+\section{Sequential Monte Carlo}
+
+\Gls{SMC} is a statistical inference method which samples from a sequence of
+probability distributions in a fixed order~\autocite{del2006sequential}. This
+idea of \defn{sequential sampling} is useful several contexts, but we consider
+here its application to model fitting in a Bayesian setting.
+
+\texttt{TODO: move the basics of model fitting (likelihoods and priors) to here}
+
+Our objective is to obtain samples from, and summary statistics of, the
+posterior distribution on the model of interest's parameters. \gls{SMC} can be
+applied in this context by defining a sequence of distributions, starting from
+the posterior and ending at the prior, which constitute a ``smooth''
+trajectory. The main \gls{SMC} algorithm for this problem was developed by
+\textcite{del2006sequential}, based on an existing methodology called
+\gls{SIS}. \gls{SIS} is also designed to sample from a sequence of
+distributions, but rather than all being defined on the same space (such as
+model parameters), they are defined on nested spaces of increasing dimension.
+Following \textcite{del2006sequential} and other reviews of
+\gls{SMC}~\autocite{doucet2001introduction}, we will begin this section by
+describing \gls{SIS}, and then turn to the adaptation of the method to the case
+when the sequence of distributions are all defined on the same space. Note that
+\textcite{del2006sequential} use the variable $\pi$ for the target
+distributions, but \textcite{doucet2001introduction} use $\pi$ to indicate the
+importance distributions. We again follow the former and denote target
+distributions by $\pi$ and importance distributions by $\eta$.
+
+The basis of \gls{SIS} is \gls{IS}, which is a method of estimating summary
+statistics of distributions which are known only up to a normalizing constant,
+and therefore cannot be sampled from directly. That is, if $\pi$ is such a
+distribution and $f$ is any real-valued function, \gls{IS} is concerned with
+estimating
+\[
+  \pi(f) = \int f(x)\pi(x)\d x = \int f(x) \frac{\gamma(x)}{Z} \d x,
+\]
+where the integral is over the space on which $\pi$ is defined, $\gamma(x)$ is
+known, and $Z$ is the unknown normalizing constant, $Z = \int \gamma(x) \d x$.
+The posterior distributions of all but the simplest models' parameters fall
+into this category. Suppose we have at hand another distribution $\eta$, called
+the \defn{importance distribution}, from which we are able to sample. Define
+the \defn{importance weight} as the ratio ratio $w(x) = \gamma(x)/\eta(x)$. We
+can express the normalizing constant $Z$ in terms of the importance weight and
+distribution, $Z = \int w(x) \eta(x) \d x$, and in turn write the original
+integral of interest as
+\[
+  \int f(x) \pi(x) \d x = \frac{\int f(x) \gamma(x) \d x}
+                               {\int w(x) \eta(x) \d x}.
+\]
+If we sample a large number of points from $\eta$, then $\eta(x)$ can be
+approximated at each of them by an empirical distribution. Since the remaining
+quantities $f$, $\gamma$, and $w$ can all be evaluated pointwise, these are all
+the ingredients we need to obtain an estimate of $\pi(f)$. Although this is a
+simple and elegant approach, the drawback is that the variance of the estimate
+is proportional to the variance of the importance
+weights~\autocite{liu2008monte}, which may be quite large if $\eta$ and
+$\gamma$ are very different. Therefore, the practical use of \gls{IS} on its
+own is limited, since it depends on finding an importance distribution which is
+similar to $\pi$, which we usually know very little about \textit{a priori}.
+
+However, an ideal context for the use of \gls{IS} is when we want to sample
+from a sequence of nested probability distributions $\pi_1, \ldots, \pi_n$.
+By \defn{nested}, we mean that $\pi_{i+1}$ is defined on a space of dimension
+$i+1$ and admits $\pi_i$ as a marginal. That is,
+\[
+  \pi_{i+1}(x_1, \ldots, x_{i+1}) = \pi_i(x_1, \ldots, x_i) f_{i+1}(x_1, \ldots, x_{i+1}),
+\]
+where $f_{i+1}$ is a function which yields a distribution when multiplied with
+$\pi_i$. This situation may seem somewhat contrived, but it arises naturally
+when trying to infer the hidden sequence of parameters of a stateful model. For
+example, \textcite{doucet2001introduction} discuss the case when $\pi_i$ is the
+posterior distribution over the first $i-1$ states of a \gls{HMM}, conditional
+on the observed data up to time $i-i$. We assume that either $\pi_1$ is known
+explicitly, or we have enough information about it to find an adequate
+importance distribution. In the \gls{HMM} example, $\pi_1$ was taken to be the
+prior distribution on the starting state.
+
 \section{Approximate Bayesian computation}
@@
 coefficient $\beta$ and the error variance $\sigma$ are the two parameters of
-the model. If the $y_i$ are all independent, then for particular choics of
+the model. If the $y_i$ are all independent, then for particular choice of
 $\beta$ and $\sigma$, we can write down the probability density of observing
@@
 
-The main idea of approximate Bayesian computation is that the ``best''
-parameters will define a model which produces simulated data similar to the
-real data. Suppose we fix values of $b$ and $\sigma$. Then, using the given
-data points $x_i$, we can generate simulated observations $y_i^*$ by adding
-$bx_i$ and a randomly chosen error drawn from $\N(0, \sigma)$.
+The main idea of \gls{ABC} is to replace the traditional likelihood-based
+criteria for model ``goodness'' with one based on simulated data generated by
+the model. Good models should be able to simulate data which closely resemble
+reality. In \gls{ABC}, we are no longer interested in the posterior
+distribution, but rather in the distribution of model parameters which produce
+data sufficiently ``close'' to the real data. To formalize this, let $d(\cdot,
+\cdot)$ be a distance function on datasets, and $\varepsilon$ be a tolerance
+level. The objective of \gls{ABC} is to recover the distribution\ldots
 
@@
 approach described in the previous subsection of guessing parameter values
-until one is close enough to the truth. More specificially, a possible set of
+until one is close enough to the truth. More specifically, a possible set of
 parameters $\theta$ is drawn from the prior distribution, and a simulated data

### 2016-02-12 doc/thesis/methods.tex
Updated reading and writing

@@
-\subsection{Programs}
+\subsection{Computer program}
 
-Most of the computer programs used for this work were written in the C
-programming language. Contact networks were generated using the
-\software{igraph} library~\autocite{csardi2006igraph}, which was also used to
-store phylogenetic trees. Hash tables, and the dynamic programming matrices for
-computing the tree kernel, were stored using Judy
-arrays~\autocite{baskins2004judy}. The GNU Scientific Library
-(GSL)~\autocite{gough2009gnu} was used to generate random draws from
-probability distributions, and to perform the bisection step in the ABC-SMC
-algorithm.
+I implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
+\textcite{del2012adaptive}. The program was written in the
+\software{C} programming language. The \software{igraph}
+library~\autocite{csardi2006igraph} was used to generate and store contact
+networks and phylogenies. Judy arrays~\autocite{baskins2004judy} were used for
+hash tables and dynamic programming arrays. The
+\gls{GSL}~\autocite{gough2009gnu} was used to generate random draws from
+probability distributions, and to perform the bisection step in the
+\gls{ABC}-\gls{SMC} algorithm.
 
-\subsubsection{Epidemic simulation over a network}
+For ease of exposition, we simplifiy the notation of \textcite{del2012adaptive}
+by dropping the subscripts on the variables which indicate the current
+iteration number. Instead, we will add a prime $'$ to indicate a value which
+will be used in the next iteration (this should become clear later on).
 
-I implemented method of~\autocite{gillespie1977exact} for simulating epidemics,
-and the corresponding transmission trees, over static contact networks. This
-method has been applied previously for the same
-purpose~\autocite{robinson2013dynamics, leventhal2012inferring}, however no
-existing open source implementations were available.
+In the algorithm, we keep track of a population of $n$ sets of model
+parameters, called \defn{particles}, denoted $\set{X_i}_{i=1}^{n}$. For the
+\gls{BA} model, the particles would be 4-tuples (\gls{N}, \gls{I}, \gls{m},
+\gls{alpha}). Each particle $X_i$ is associated with a set of \gls{M} simulated
+datasets, denoted $\set{X_{i,j}}_{j=1}^M$, and a weight $W_i$.
 
-Let $G = (V, E)$ be a directed contact network. Each directed edge $e = (u, v)$
-in the network is associated with a transmission rate $\beta_e$, which indicates
-that, once $u$ becomes infected, $u$ will infect $v$ with rate $\beta_e$. In
-other words, the time between $u$ becoming infected and the infection passing
-from $u$ to $v$ is a random variable with distribution $\Exponential(\beta_e)$.
-Note that $v$ may become infected before this time has elapsed, if $v$ has
-other incoming edges. Each vertex $v$ also has a removal rate $\gamma_v$, which
-is the rate at which $v$ is removed from the population after becoming
-infected. Removal may correspond to death or recovery with immunity, or a
-combination of both, but in our implementation recovered nodes never re-enter
-the susceptible population. Borrowing a term from HIV epidemiology, a
-\defn{discordant edge} in $G$ is an edge $(u, v)$ where $u$ is infected and $v$
-is uninfected.
+The particles are initially drawn from the prior distribution.
+
+Let $d$ be a
+distance measure on data sets, so that $d(x, y)$ is smaller the more similar
+$x$ and $y$ are to each other. Let $\varepsilon$ be the tolerance level which
+indicates whether a data set is ``close'' to the observed data. That is, if
+$d(x, y) < \varepsilon$, we will say that $x$ and $y$ are close, otherwise they
+are distant.
+
+I implemented a Gillespie simulation algorithm~\autocite{gillespie1976general}
+for simulating epidemics, and the corresponding transmission trees, over static
+contact networks. This method has been independently implemented and applied by
+several authors~\autocite[\textit{e.g.}][]{o2010contact, robinson2013dynamics,
+leventhal2012inferring, groendyke2011bayesian}.
+\textcite{groendyke2011bayesian} published their implementation as an
+\software{R} package, but since the \gls{SMC} algorithm is quite
+computationally intensive, we chose to implement our own version in
+\software{C}.
+
+Let $G = (V, E)$ be a directed contact network. The individual nodes and edges
+of $G$ follow the dynamics of the \gls{SIR}
+model~\autocite{kermack1927contribution}. Each directed edge $e = (u, v)$ in
+the network is associated with a transmission rate $\beta_e$, which indicates
+that, once $u$ becomes infected, the waiting time until $u$ infects $v$ is
+distributed as $\Exponential(\beta_e)$. Note that $v$ may become infected
+before this time has elapsed, if $v$ has other incoming edges. $v$ also has a
+removal rate $\gamma_v$, so that the waiting time until removal of $v$ from the
+population is $\Exponential(\gamma_v)$. Removal may correspond to death or
+recovery with immunity, or a combination of both, but in our implementation
+recovered nodes never re-enter the susceptible population. We define a
+\defn{discordant edge} as an edge $(u, v)$ where $u$ is infected and $v$
+has never been infected.
 
@@
 outgoing edges from $v$. Let $I$ be the set of infected nodes in the network,
-$R$ be the set of removed nodes, and $S$ be the remaining susceptible nodes.
-Let $D$ be the set of discordant edges in the network, $\beta$ be the 
-total transmission rate over all discordant edges, and $\gamma$ be the total
-removal rate of all infected nodes. That is,
+$R$ be the set of removed nodes, and $S$ be the remaining susceptible nodes,
+and $D$ be the set of discordant edges in the network. Let $\beta$ be the total
+transmission rate over all discordant edges, and $\gamma$ be the total removal
+rate of all infected nodes,
 \[
@@
 The variables $S$, $I$, $R$, $D$, $\beta$, and $\gamma$ are all updated as the
-simulation progresses, but since these updates are very straightforward we do
-not write them explicitly in the algorithm. When a new node $v$ is infected, it
-is deleted from $S$ and added to $I$, any formerly discordant edges in $\in(v)$
-are deleted from $D$, and edges in $\out(v)$ to nodes in $S$ are added to $D$.
-If $v$ is later removed, it is deleted from $I$ and added to $R$, and any
-discordant edges in $\out(v)$ are deleted from $D$. In both cases, the
-variables $\beta$ and $\gamma$ are updated to reflect the changes.
+simulation progresses. When a node $v$ becomes infected, it is deleted from $S$
+and added to $I$, any formerly discordant edges in $\in(v)$ are deleted from
+$D$, and edges in $\out(v)$ to nodes in $S$ are added to $D$. If $v$ is later
+removed, it is deleted from $I$ and added to $R$, and any discordant edges in
+$\out(v)$ are deleted from $D$. In both cases, the variables $\beta$ and
+$\gamma$ are updated to reflect the changes. Since these updates are
+straightforward, we do not write them explicitly in the algorithm.
 
@@
 
-The full algorithm is shown in Algorithm \ref{alg:nettree}. The transmission
-tree $T$ is simulated along with the epidemic. We keep a map called $\tip$,
-which maps infected nodes in $I$ to the tips of $T$. The variable $n$ counts
-the number of nodes, both extant and internal, in $T$. The simulation continues
-until either there are no discordant edges left in the network, or we reach a
-user-defined cutoff of time ($t_{\max}$) or number of infections ($I_{\max}$).
+The Gillespie simulation algorithm is given as Algorithm~\ref{alg:nettree}. The
+transmission tree $T$ is simulated along with the epidemic. We keep a map
+called $\tip$, which maps infected nodes in $I$ to the tips of $T$. The
+simulation continues until either there are no discordant edges left in the
+network, or we reach a user-defined cutoff of time ($t_{\max}$) or number of
+infections ($I_{\max}$). We use the notation $\Uniform(0, 1)$ to indicate a
+number drawn from a uniform distribution on $(0, 1)$, and likewise for
+$\Exponential(\lambda)$. The combined number of internal nodes and tips in $T$
+is denoted $|T|$.
 
@@
   \begin{algorithmic}
-    \State infect one node $v$ at random
-    \State let $T$ be a single node with index $1$
+    \State infect a node $v$ at random, updating $S$, $I$, $D$, $\beta$ and $\gamma$
+    \State $T \gets$ a single node with label $1$
     \State $\tip[v] \gets 1$
-    \State $n \gets 1$
     \State $t \gets 0$
@@
       \If{$t < t_{\max}$}
-        \If{$\Uniform(0, 1) < \nicefrac{\beta}{\beta + \gamma}$}
+        \If{$\Uniform(0, \beta + \gamma) < \beta$}
           \State choose an edge $e = (u, v)$ from $D$ with probability $\beta_e / \beta$
                  and infect $v$
-          \State add nodes with labels $n+1$ and $n+2$ to $T$
-          \State connect $n+1$ and $n+2$ to $\tip[v]$ in $T$, with branch lengths $0$
-          \State $\tip[v] \gets n + 1$
-          \State $\tip[u] \gets n + 2$
-          \State $n \gets n + 2$
+          \State add tips with labels $(|T|+1)$ and $(|T|+2)$ to $T$
+          \Comment $|T|$ increased by 2
+          \State connect the new nodes to $\tip[v]$ in $T$, with branch lengths $0$
+          \State $\tip[v] \gets |T|-1$
+          \State $\tip[u] \gets |T|$
         \Else
@@
                  and remove $v$
-          \State remove $v$ from $\tip$
+          \State delete $v$ from $\tip$
         \EndIf
@@
 \end{figure}
-
-\subsubsection{Adaptive sequential Monte-Carlo approximate Bayseian computation}
-
-I implemented the adaptive sequential Monte-Carlo (SMC) algorithm developed by
-Del Moral \etal~\autocite{del2012adaptive} for approximate Bayesian
-computation.
-
-For ease of exposition, we simplifiy the notation of Del Moral \etal by
-dropping the subscripts on the variables which indicate the current iteration
-of the algorithm. Instead, we will add a superscript $^*$ to indicate the value
-of the variable which will be used in the next iteration (this should become
-clear later on). Let $\set{\theta_i}_{i=1}^{N}$ be a set of $N$ particles,
-which are particular choices of the model parameters $\theta$. For each
-particle $\theta_i$, let $\set{X_{i,j}}_{j=1}^M$ be a set of $M$ simulated
-datasets, generated from the model with parameters $\theta_i$. Let
-$\set{W_i}_{i=1}^N$ be weights associated to each particle. Let $d$ be a
-distance measure on data sets, so that $d(x, y)$ is smaller the more similar
-$x$ and $y$ are to each other. Let $\varepsilon$ be the tolerance level which
-indicates whether a data set is ``close'' to the observed data. That is, if
-$d(x, y) < \varepsilon$, we will say that $x$ and $y$ are close, otherwise they
-are distant.
 

### 2016-02-15 doc/thesis/intro.tex

@@
 the DNA sequences of an ancestor its descendant, or the difference in average
-body mass or height. Particular examples of evolutionary distance, based on
-genetic data, will be discussed below in subsection~\ref{subsubsec:genetree}. A
-phylogeny with branch lengths in calendar time units is often referred to as
-\defn{time-scaled}. In a time-scaled phylogeny, the internal nodes can be
-mapped onto a timeline by using the tips of the tree, which usually map to the
-present day, as a reference point~\autocite{nee1992tempo}. The corresponding
-points on the timeline are called \defn{branching times}, and the rate of their
-accumulation is referred to as the \defn{branching rate}. Rooted trees whose
-tips are all the same distance from the root are called \defn{ultrametric}
+body mass or height. A phylogeny with branch lengths in calendar time units is
+often referred to as \defn{time-scaled}. In a time-scaled phylogeny, the
+internal nodes can be mapped onto a timeline by using the tips of the tree,
+which usually map to the present day, as a reference
+point~\autocite{nee1992tempo}. The corresponding points on the timeline are
+called \defn{branching times}, and the rate of their accumulation is referred
+to as the \defn{branching rate}. Rooted trees whose tips are all the same
+distance from the root are called \defn{ultrametric}
 trees~\autocite{buneman1974note}. These concepts are illustrated in
@@
 
-\subsection{Species trees and transmission trees}
-\label{subsubsec:speciestree}
+\subsection{Transmission trees and epidemiology}
 
-A \defn{species tree} is a particular type of phylogeny in which the taxa are
-species, and the branching times correspond to historical speciation events. By
-speciation events, we mean times when two formerly interbreeding groups of
-organisms ceased to reproduce with each other. In general, we do not have
-access to the true tree relating a particular group of extant species, as this
-would usually imply certain knowledge of speciation events in the distant past.
-Therefore, all species trees considered by researchers are estimates, based on
-the genetic or phenotypic similarity of the extant taxa, the fossil record, or
-both. We discuss some of the challenges associated with this estimation in
-subsection \ref{subsubsec:treeconv} below. However, for the moment, we ignore
-these complications and consider what can be said about the true species tree.
+In epidemiology, a \defn{transmission tree} is a graphical representation of an
+epidemic's progress through a population. Like phylogenies, transmission trees
+have tips, nodes, edges, and branch lengths. However, rather than recording an
+evolutionary process (speciation), they record an epidemiolgical process
+(transmission). The tips of a transmission tree represent infected hosts, while
+internal nodes correspond to transmissions from one host to another.
+Transmission trees generally have branch lengths in units of calendar time,
+with a root corresponding to the index case, and branching times indicating
+times of transmission. The internal nodes may be labelled with
+the donor of the transmission pair, if this is known. The tips of the tree,
+rather than being fixed at the present day, are placed at the time at which the
+individual was removed from the epidemic, such as by death, recovery,
+isolation, behaviour change, or migration. Consequently, the transmission tree
+may not be ultrametric, but may have tips located at varying distances from the
+root. Such trees are said to have \defn{heterochronous}
+taxa~\autocite{drummond2003measurably}, in contrast to the
+\defn{isochronous} taxa found in the phylogenies of macro-organisms.  A
+transmission tree is illustrated in Figure~\ref{fig:virustrees}.
 
-Speciation often occurs by an \defn{allopatric} process, where a sub-population
-of organisms is isolated from the original population by a geographic
-barrier~\autocite{coyne2004speciation}. Over time, differing selection
-pressures on either side of the barrier, coupled with genetic drift, cause the
-two populations to diverge genetically. This eventually results in two distinct
-species, which will appear as two subtrees or \defn{clades} in their species
-tree if descendants of both survive to the present. Of course, this is not
-necessarily the case, as selection may cause one or both of the new
-sub-populations or their descendants to become extinct. On the other hand,
-selection may favour one of the sub-populations, allowing it to expand into new
-regions or niches and possibly paving the way for subsequent speciation events.
-Speciation is not necessarily allopatric: \defn{sympatric} speciation refers to
-the emergence of a new species without any geographic
-barrier~\autocite{coyne2004speciation}. Indeed, there is a whole continuum of
-possible speciation processes~\autocite{fitzpatrick2008if}, ranging from
-completely allopatric to completely sympatric. However, at least for eukaryotic
-organisms, allopatric speciation is thought to be by far the most common
-type~\autocite{coyne2004speciation}.
+\begin{figure}
+    \centering
+    \includegraphics{virustrees}
+    \label{fig:virustrees}
+    \caption[Illustration of a transmission tree]{
+        A transmission tree representing an epidemic's progress through
+        four individuals. The index case, patient $a$, transmitted the
+        virus to patients $b$ and $c$; patient $c$ went on to infect
+        patient $d$. 
+    }
+\end{figure}
 
-A different type of tree used in the study of pathogens is a \defn{transmission
-tree}. In such trees, tips represent infected hosts, while internal nodes
-correspond to transmissions from one host to another. Transmission trees
-generally have branch lengths in units of calendar time, with a root
-corresponding to the index case, and branching times indicating times of
-transmission. The internal nodes may be labelled with the donor of the
-transmission pair, if this is known. The tips of the tree, rather than being
-fixed at the present day, are placed at the time at which the pathogen
-population of the individual was sampled. Consequently, the transmission tree
-may not be ultrametric, but may have tips located at varying distances from the
-root. Such trees are said to have \defn{heterochronous} taxa or
-samples~\autocite{drummond2003measurably}, in contrast to the
-\defn{isochronous} samples found in macro-organisms species trees. A
-transmission tree is illustrated in Figure~\ref{fig:contactnet} (A).
+Since transmission trees are essentially a detailed record of an epidemic's
+progress, they contain substantial epidemiological information. As a basic
+example, the \gls{ltt} plot~\autocite{nee1992tempo}, which plots the number of
+lineages in a phylogeny against time, can be used to quantify the incidence of
+new infections over the course of an epidemic~\autocite{holmes1995revealing}.
+Many more diverse epidemiological parameters have been investigated using
+transmission trees, such as the degree of
+clustering~\autocite{hughes2009molecular} and the effect of elevated
+transmission risk in acute infection~\autocite{volz2012simple}. However, in all
+but the most well-studied of epidemics, this is not possible to obtain through
+traditional epidemiological methods. The time and effort to conduct detailed
+interviews and contact tracing of a sufficient number of infected individuals
+is usually prohibitive. Even when the resources for such methods are available,
+patients may not always recall whom they contacted and when, especially in the
+case of airborne transmission. Consequently, the transmission tree must be
+estimated using other methods. Most commonly, this is done by exploiting the
+relationship between transmission trees and viral phylogenies.
 
-At first glance, transmission trees and species trees seem to have nothing to
-do with each other. One represents the evolutionary history of a group of
-organisms, the other traces the spread of a disease through a host population.
-However, when considering epidemics of RNA viruses, transmission trees and
-species trees are highly related. The viral population within a single host has
-often been referred to as a \defn{quasispecies}~\autocite{domingo2012viral},
-and due to the fast evolutionary rate of these viruses, two hosts' viral
-populations are invariably genetically distinct. The transmission process is
-functionally identical to allopatric speciation, involving geographic isolation
-of a sub-population in a new host and subsequent genetic divergence. However,
-for the most part, it is not selection but rather host epidemiology which
-causes certain lineages to proliferate while driving others to extinction. A
-host who recovers, dies, or is isolated equates to an extinction in the
-transmission tree. On the other hand, dense populations in frequent contact
-often experience rapid epidemic growth, which causes similarly rapid
-accumulation of branching points in the tree.
+\subsection{Estimating transmission trees from viral phylogenies}
 
-The fact that these evolutionary (ie. speciation) and epidemiological processes
-occur on the same time scale for RNA viruses~\autocite{drummond2003measurably}
-is the basis for the field of
-\defn{phylodynamics}~\autocite{grenfell2004unifying}. The similarity between
-transmission and speciation means that it is possible to apply many of the
-tools and techniques developed for species trees to viral transmission trees,
-and interpret the results in the context of transmission rates and host
-epidemiology. As a basic example, the lineages-through-time
-plot~\autocite{nee1992tempo}, which plots the historical speciation rate
-against time, can be used to quantify the incidence of new infections over the
-course of an epidemic~\autocite{holmes1995revealing}. In
-subsection~\ref{subsubsec:treeshape} below, we review some of these methods in
-more detail, as well as several which have been developed specifically for
-viral phylodynamics, and discuss what these methods can tell us about host
-epidemiology.
+In general, \defn{viral phylogenies} are simply phylogenetic trees relating
+virus strains. \comment{Is ``strain'' the best word here?} In phylodynamics, we
+often consider \defn{inter-host} phylogenies, which relate one viral genotype
+from each host in a population. The tips of the tree are labelled with both the
+virus and its host. The crux of
+\defn{phylodynamics}~\autocite{grenfell2004unifying} is the fact that the
+processes generating these two types of tree - epidemiological processes, for
+transmission trees, and evolutionary processes, for viral phylogenies - occur
+on similar time scales for RNA viruses~\autocite{drummond2003measurably}. As a
+result, there is a close relationship between the two trees. In particular, the
+transmission process is quite similar to \defn{allopatric
+speciation}~\autocite{coyne2004speciation}, where genetic divergence follows
+the geographic isolation of a sub-population of organisms. Thus, transmission,
+which is represented as branching in the transmission tree, causes branching in
+the viral phylogeny as well. Similarly, the removal of an individual causes the
+extinction of their viral lineage. Due to these relationships, the topology of
+the viral phylogeny is often used as a proxy for the topology of the
+transmission tree. However, there are several complications and caveats which
+must be kept in mind when estimating the transmission tree in this manner.
 
-Finally, we must note two important points. First, we have not yet considered
-the selection pressures operating within a single host, such as the immune
-response. Second, we reiterate that both transmission trees and species trees
-are theoretical objects which are almost never known with certainty for real
-data. Indeed, the estimation of transmission trees is often problematic, and it
-is in such estimation that within-host forces play a role. Both of these issues
-are discussed further below in subsection~\ref{subsubsec:treeconv}, though we
-continue to ignore them in the next subsection.
+First are the issues of rooting and time-scaling. Modern likelihood-based
+methods of phylogenetic reconstruction produce unrooted trees whose branch
+lengths measure genetic distance in units of expected substitutions per site,
+whereas transmission trees are rooted with branches measuring calendar time.
+We generally assume that sampling a virus from the individual also corresponds
+to their removal from the transmission tree\comment{needs a reference}, so the
+positions of the tips in time are fixed. Therefore, to transform a viral
+phylogeny into an estimated transmission tree, we must find a root and an
+assignment of branch lengths such that the tips are placed at their proscribed
+times. Of course, there are an infinite number of possible combinations of root
+and branch lengths which result in the correct placement of the tips, so we
+would also like to find values which are the most ``reasonable'' given our
+data. By reasonable, we mean that the variation in \defn{evolutionary rate}, 
+which is the ratio of a branch's length in genetic distance to its length in
+calendar time, is low across the tree. While there is some variation among
+hosts, due to immunological and other factors, we generally expect to observe
+globally similar evolutionary rates. Methods for time-scaling a phylogeny
+include root-to-tip regression~\autocite{drummond2003inference}, which we apply
+in this work, and least-square dating~\autocite{to2015fast}. Both of these
+methods can be used to root the tree, by simply trying all possible root
+positions and choosing the one which minimizes a loss function
+($1-R^2$ or root-mean-square error for root-to-tip regression, sum of squared
+errors for least-square dating). Alternatively, the tree may be rooted with an
+outgroup before time-scaling.
 
-\subsection{Gene trees and viral phylogenies}
-\label{subsubsec:genetree}
+A second, perhaps more insidious problem is the fact that the correspondence
+between the topologies of the viral phylogeny and transmission tree is not
+necessarily exact. Due to intra-host diversity, the viral strain which is
+transmitted may have split from another lineage within the donor long before
+the transmission event occurred. Hence, the branching point in the viral
+phylogeny may be much earlier than that in the transmission tree. Another
+possibility is that one host transmitted to two or more recipients, but the
+lineages they each received originated within the donor host in a different
+order than that in which the transmissions occurred. In this case, the topology
+of the transmission tree and the viral phylogeny will be mismatched. Although
+phylodynamics is quite new, these phenomena have been studied in evolutionary
+biology for some time. Viral phylogenies are a specific version of a more
+general class of trees called \defn{gene trees}, which represent the
+evolutionary history of a section of genetic material. Transmission trees, on
+the other hand, are highly analogous to \defn{species trees}, whose tips are
+species and internal nodes are common ancestors. This analogy derives from the
+functional similarity between transmission and allopatric speciation. Hence,
+the potential discordance between transmission trees and viral phylogenies is
+the similar to that between gene and species trees, which is called
+\defn{incomplete lineage sorting}. In practice, this discordance has not proven
+an insurmountable problem: for example, \textcite{leitner1996accurate} were
+able to accurately recover a known transmission tree using a viral phylogeny.
 
-A \defn{gene tree} is another type of phylogeny which traces the evolutionary
-history of one or more sections of genetic material. Gene trees can be built
-from small parts of genes up to whole genomes, but for ease of exposition we
-will use only the word ``gene''. The tips of a gene tree represent the
-different \defn{alleles}, or versions, of the gene found in the studied taxa.
-The branching points in a gene tree represent genetic divergence events - that
-is, the times when two distinct alleles first came into existence through
-mutation.
+A final caveat is that the viral phylogeny itself is not known with certainty,
+so it must also be estimated from genetic data. Phylogenetic inference is a
+complex topic which we will not discuss in detail here (see e.g.
+\autocite{nei2000molecular} for a full review). Most modern analyses use
+model-based methods, which simultaneously estimate the phylogeny with branch
+lengths and the parameters of a model of evolution. Although they usually work
+well in practice \comment{needs a reference}, the estimated topology can vary
+based on the model used and, in the case of Bayesian analysis, the priors.  In
+addition, intra-host viral populations are genetically heterogeneous, so
+choosing a single representative genotype per host is necessarily imprecise.
+One can use either the genotype of a specific virion sampled from the host, or
+a synthetic genotype, such as a consensus or reconstructed ancestral sequence.
 
-Gene trees are usually estimated using the nucleotide or amino acid sequences
-of the genes in question. Although there are methods for estimating gene trees
-using other types of molecular data, such as gene orders, we will focus on
-sequence-based methods here. As with species trees, it is not possible to know
-gene trees with certainty, only to estimate them. However, the problem of
-estimating gene trees directly from data is somewhat better defined than that
-of estimating species trees, because of the quantitative nature of the
-molecular data used to label the tips. In general, methods to estimate gene
-trees fall into three major categories: distance-based, maximum parsimony, and
-model-based~\autocite{nei2000molecular}. The first two categories are generally
-no longer used for phylogenetic inference, since model-based methods have been
-demonstrated to be almost universally more accurate. Therefore, we will review
-in detail only the last category.
+What we have just discussed is a two-step procedure for estimating the
+transmission tree. First, a viral phylogeny is constructed from genetic
+sequence data, and then it is rooted and time-scaled into a transmission tree.
+This approach is straightforward, frequently used, and has the advantage of
+leveraging tried-and-true tools for phylogenetic inference. However, it also
+has drawbacks, perhaps the most obvious being the multiplication of errors
+produced by the separate steps. One commonly used alternative method is to
+directly estimate a time-scaled phylogeny by simultaneously inferring the tree
+topology, its root and branch lengths, and the parameters of a \defn{molecular
+clock} model. A molecular clock is a hypothesis about the evolutionary rates
+along the branches of the tree, such as that they are all equal (a \defn{strict
+clock}) or that they are \gls{iid} from a common distribution (a \defn{relaxed
+clock}). This inference is usually done in a Bayesian framework using
+\gls{MCMC}, so that prior information (including the tip dates) \comment{are
+tip dates part of the prior?} can be included in the analysis, and the
+so-called nuisance parameters of the molecular clock model can be marginalized
+out. Software packages for performing these analyses include
+BEAST~\autocite{bouckaert2014beast} and MrBayes~\autocite{ronquist2012mrbayes}.
 
-Model-based methods include both maximum likelihood and Bayesian approaches. To
-use them, one must first define a parametric model of sequence evolution, here
-denoted $M$. For all possible trees $T$ (with branch lengths) and choices of
-model parameters $\theta$, the model must define a probability density function
-on observed sequence data $D$, denoted $f(D \mid T,\,\theta)$. This is also
-sometimes written $\Pr(D \mid T,\,\theta)$, even though it is a probability
-\emph{density}, not a probability. Maximum likelihood aims to find a particular
-tree and values for the model parameters which optimize the \defn{likelihood
-function} $\L$ for the observed data,
-\[
-  \L(\theta,\, T \mid D) = f(D \mid T,\, \theta).
-\]
-Bayesian methods, rather than trying to find point estimates for $\theta$ and
-$T$, aim to approximate the entire \defn{posterior distribution} for the
-observed data,
-\[
-  f(T,\,\theta \mid D) = \frac{f(D \mid T,\,\theta)f(T,\,\theta)}{f(D)}.
-\]
-Here, $f(T, \theta)$ denotes the prior distribution on $T$ and $\theta$, and
-$f(D)$ is the marginal density of the data (that is, the integral of the
-numerator over all values of $T$ and $\theta$).
+Several other authors have developed methods tailor-made for inferring
+transmission trees. \textcite{didelot2014bayesian} develop a Bayesian version
+of the two-step approach which allows transmissions to occur anywhere along the
+branches of a transmission tree, rather than being constrained to the branching
+points in the viral phylogeny. The method requires sampling of every infected
+individual, although the authors indicate that it could be extended to relax
+this assumption. \textcite{cottam2008integrating} describe a likelihood-based
+method which enumerates all transmission trees consistent with an established
+phylogeny, assigning each a likelihood based on other epidemiological data. 
+This approach is novel in its integration of data from multiple sources,
+however because it enumerates a large portion of the tree space, it is unlikely
+to scale to larger epidemics. A different approach is undertaken by
+\textcite{jombart2011reconstructing}, who describe a method to build
+transmission trees directly from sequence data, contingent on the common
+ancestors also being sampled. This makes the method attractive for
+slow-evolving pathogens, but less practical for viral outbreaks where samples
+from common ancestors are unlikely to be available.
 
-A \defn{viral phylogeny} is a particular kind of gene tree which relates the
-genetic sequences of viruses carried within one or more hosts. In this work, we
-consider viral phylogenies constructed from one sequence per infected
-individual, so that the tips of the are associated with both the viral sequence
-and the corresponding host.
+%\subsection{Species trees and transmission trees}
+%\label{subsubsec:speciestree}
+%
+%A \defn{species tree} is a particular type of phylogeny in which the taxa are
+%species, and the branching times correspond to historical speciation events. By
+%speciation events, we mean times when two formerly interbreeding groups of
+%organisms ceased to reproduce with each other. In general, we do not have
+%access to the true tree relating a particular group of extant species, as this
+%would usually imply certain knowledge of speciation events in the distant past.
+%Therefore, all species trees considered by researchers are estimates, based on
+%the genetic or phenotypic similarity of the extant taxa, the fossil record, or
+%both. We discuss some of the challenges associated with this estimation in
+%subsection \ref{subsubsec:treeconv} below. However, for the moment, we ignore
+%these complications and consider what can be said about the true species tree.
+%
+%Speciation often occurs by an \defn{allopatric} process, where a sub-population
+%of organisms is isolated from the original population by a geographic
+%barrier~\autocite{coyne2004speciation}. Over time, differing selection
+%pressures on either side of the barrier, coupled with genetic drift, cause the
+%two populations to diverge genetically. This eventually results in two distinct
+%species, which will appear as two subtrees or \defn{clades} in their species
+%tree if descendants of both survive to the present. Of course, this is not
+%necessarily the case, as selection may cause one or both of the new
+%sub-populations or their descendants to become extinct. On the other hand,
+%selection may favour one of the sub-populations, allowing it to expand into new
+%regions or niches and possibly paving the way for subsequent speciation events.
+%Speciation is not necessarily allopatric: \defn{sympatric} speciation refers to
+%the emergence of a new species without any geographic
+%barrier~\autocite{coyne2004speciation}. Indeed, there is a whole continuum of
+%possible speciation processes~\autocite{fitzpatrick2008if}, ranging from
+%completely allopatric to completely sympatric. However, at least for eukaryotic
+%organisms, allopatric speciation is thought to be by far the most common
+%type~\autocite{coyne2004speciation}. 
+%
+%At first glance, transmission trees and species trees seem to have nothing to
+%do with each other. One represents the evolutionary history of a group of
+%organisms, the other traces the spread of a disease through a host population.
+%However, when considering epidemics of RNA viruses, transmission trees and
+%species trees are highly related. The viral population within a single host has
+%often been referred to as a \defn{quasispecies}~\autocite{domingo2012viral},
+%and due to the fast evolutionary rate of these viruses, two hosts' viral
+%populations are invariably genetically distinct. The transmission process is
+%functionally identical to allopatric speciation, involving geographic isolation
+%of a sub-population in a new host and subsequent genetic divergence. However,
+%for the most part, it is not selection but rather host epidemiology which
+%causes certain lineages to proliferate while driving others to extinction. A
+%host who recovers, dies, or is isolated equates to an extinction in the
+%transmission tree. On the other hand, dense populations in frequent contact
+%often experience rapid epidemic growth, which causes similarly rapid
+%accumulation of branching points in the tree. 
+%
+%The fact that these evolutionary (ie. speciation) and epidemiological processes
+%occur on the same time scale for RNA viruses~\autocite{drummond2003measurably}
+%is the basis for the field of
+%\defn{phylodynamics}~\autocite{grenfell2004unifying}. The similarity between
+%transmission and speciation means that it is possible to apply many of the
+%tools and techniques developed for species trees to viral transmission trees,
+%and interpret the results in the context of transmission rates and host
+%epidemiology. 
+%
+%%As a basic example, the lineages-through-time
+%%plot~\autocite{nee1992tempo}, which plots the historical speciation rate
+%%against time, can be used to quantify the incidence of new infections over the
+%%course of an epidemic~\autocite{holmes1995revealing}. 
+%In
+%subsection~\ref{subsubsec:treeshape} below, we review some of these methods in
+%more detail, as well as several which have been developed specifically for
+%viral phylodynamics, and discuss what these methods can tell us about host
+%epidemiology.
+%
+%Finally, we must note two important points. First, we have not yet considered
+%the selection pressures operating within a single host, such as the immune
+%response. Second, we reiterate that both transmission trees and species trees
+%are theoretical objects which are almost never known with certainty for real
+%data. Indeed, the estimation of transmission trees is often problematic, and it
+%is in such estimation that within-host forces play a role. Both of these issues
+%are discussed further below in subsection~\ref{subsubsec:treeconv}, though we
+%continue to ignore them in the next subsection.
+%
+%\subsection{Gene trees and viral phylogenies}
+%\label{subsubsec:genetree}
+%
+%A \defn{gene tree} is another type of phylogeny which traces the evolutionary
+%history of one or more sections of genetic material. Gene trees can be built
+%from small parts of genes up to whole genomes, but for ease of exposition we
+%will use only the word ``gene''. The tips of a gene tree represent the
+%different \defn{alleles}, or versions, of the gene found in the studied taxa.
+%The branching points in a gene tree represent genetic divergence events - that
+%is, the times when two distinct alleles first came into existence through
+%mutation.
+%
+%Gene trees are usually estimated using the nucleotide or amino acid sequences
+%of the genes in question. Although there are methods for estimating gene trees
+%using other types of molecular data, such as gene orders, we will focus on
+%sequence-based methods here. As with species trees, it is not possible to know
+%gene trees with certainty, only to estimate them. However, the problem of
+%estimating gene trees directly from data is somewhat better defined than that
+%of estimating species trees, because of the quantitative nature of the
+%molecular data used to label the tips. In general, methods to estimate gene
+%trees fall into three major categories: distance-based, maximum parsimony, and
+%model-based. The first two categories are generally
+%no longer used for phylogenetic inference, since model-based methods have been
+%demonstrated to be almost universally more accurate. Therefore, we will review
+%in detail only the last category.
+%
+%Model-based methods include both maximum likelihood and Bayesian approaches. To
+%use them, one must first define a parametric model of sequence evolution, here
+%denoted $M$. For all possible trees $T$ (with branch lengths) and choices of
+%model parameters $\theta$, the model must define a probability density function
+%on observed sequence data $D$, denoted $f(D \mid T,\,\theta)$. This is also
+%sometimes written $\Pr(D \mid T,\,\theta)$, even though it is a probability
+%\emph{density}, not a probability. Maximum likelihood aims to find a particular
+%tree and values for the model parameters which optimize the \defn{likelihood
+%function} $\L$ for the observed data,
+%\[
+%  \L(\theta,\, T \mid D) = f(D \mid T,\, \theta).
+%\]
+%Bayesian methods, rather than trying to find point estimates for $\theta$ and
+%$T$, aim to approximate the entire \defn{posterior distribution} for the
+%observed data,
+%\[
+%  f(T,\,\theta \mid D) = \frac{f(D \mid T,\,\theta)f(T,\,\theta)}{f(D)}.
+%\]
+%Here, $f(T, \theta)$ denotes the prior distribution on $T$ and $\theta$, and
+%$f(D)$ is the marginal density of the data (that is, the integral of the
+%numerator over all values of $T$ and $\theta$).
+%
+%\subsection{Estimating transmission trees}
+%\label{subsubsec:treeconv}
+%
+%Most applications of viral phylodynamics, including this project, aim to answer
+%epidemiological questions~\autocite{pybus2009evolutionary, volz2013viral}. 
+%For example, phylodynamic methods have been used to investigate the degree of
+%clustering~\autocite{hughes2009molecular} and the effect of elevated
+%transmission risk in acute infection~\autocite{volz2012simple} 
+%(a more thorough
+%review is given below in subsection~\ref{subsubsec:appphylo}). For these
+%applications, we are ultimately interested in the transmission tree, which is a
+%representation of an epidemiological process (transmission), as opposed to the
+%viral phylogeny, which represents a biological process (viral evolution). As
+%discussed above in subsection~\ref{subsubsec:speciestree}, transmission trees
+%and are in practice unknown, and must be estimated from available data. Here,
+%we review some of the ways to perform such estimation.
+%
+%In principle, transmission trees can be reconstructed by on-the-ground
+%epidemiological methods, that is, by asking each person who they contacted.
+%This is particularly relevant for sexually transmitted diseases such as HIV,
+%where individuals are more likely to recall who they contacted and
+%when~\autocite{klovdahl1985social}. This kind of data is challenging to collect
+%in detail. Most commonly, when applying phylodynamics in practice, the
+%transmission tree is estimated from the viral
+%phylogeny~\autocite{volz2013viral}. As discussed above in
+%subsection~\ref{subsubsec:genetree}, the viral phylogeny is itself also estimated;
+%however, for the moment we ignore this complication and describe methods for
+%estimating the transmission tree if the viral phylogeny were known.
+%
+%When we consider the fact that the viral phylogeny is not known with certainty,
+%the methods just described follow a two-step procedure. The first step is to
+%estimate the viral phylogeny, and the second step is to use it to estimate the
+%transmission tree.
+%
+%\autocite{didelot2014bayesian} develop a Bayesian version of the two-step
+%approach. They first estimate a time-scaled phylogeny from viral samples, then
+%estimate the transmissions which took place along the phylogeny. Due to
+%within-host evolution, the transmissions are allowed to occur anywhere along
+%the branches, rather than being constrained to happen at branching points.
+%However, the authors admit that their method requires sampling of every
+%infected individual, although they indicate that it could be extended to relax
+%this assumption.
+%
+%A different approach is undertaken by Jombart
+%\etal~\autocite{jombart2011reconstructing}, who build transmission trees
+%directly from sequence data without the intermediate step of constructing a
+%phylogeny. Their method finds the most likely geneology among the sampled
+%sequences, assuming that the that the common ancestors to sampled isolates are
+%themselves sampled. Although this may be a realistic assumption in the early
+%stages of an epidemic of a slower evolving pathogen, it is unlikely for
+%ancestors to be present among sequences sampled from an ongoing HIV epidemic
+%\comment{Needs a reference}.
+%
+%Several alternative approaches have been developed which aim to
+%incorporate available epidemiological data as well. Cottam
+%\etal~\autocite{cottam2008integrating} first built a viral phylogeny, then
+%enumerated all transmission trees consistent with this phylogeny by labelling
+%the internal nodes of the phylogeny in every possible way. Each of the putative
+%transmission trees was assigned a likelihood based on infection time and
+%geographic separation. 
+%
+%
 
-\subsection{Estimating transmission trees}
-\label{subsubsec:treeconv}
-
-Most applications of viral phylodynamics, including this project, aim to answer
-epidemiological questions~\autocite{pybus2009evolutionary, volz2013viral}. For
-example, phylodynamic methods have been used to investigate the degree of
-clustering~\autocite{hughes2009molecular} and the effect of elevated
-transmission risk in acute infection~\autocite{volz2012simple} (a more thorough
-review is given below in subsection~\ref{subsubsec:appphylo}). For these
-applications, we are ultimately interested in the transmission tree, which is a
-representation of an epidemiological process (transmission), as opposed to the
-viral phylogeny, which represents a biological process (viral evolution). As
-discussed above in subsection~\ref{subsubsec:speciestree}, transmission trees
-and are in practice unknown, and must be estimated from available data. Here,
-we review some of the ways to perform such estimation.
-
-In principle, transmission trees can be reconstructed by on-the-ground
-epidemiological methods, that is, by asking each person who they contacted.
-This is particularly relevant for sexually transmitted diseases such as HIV,
-where individuals are more likely to recall who they contacted and
-when~\autocite{klovdahl1985social}. This kind of data is challenging to collect
-in detail. Most commonly, when applying phylodynamics in practice, the
-transmission tree is estimated from the viral
-phylogeny~\autocite{volz2013viral}. As discussed above in
-subsection~\ref{subsubsec:genetree}, the viral phylogeny is itself also estimated;
-however, for the moment we ignore this complication and describe methods for
-estimating the transmission tree if the viral phylogeny were known.
-
-When we consider the fact that the viral phylogeny is not known with certainty,
-the methods just described follow a two-step procedure. The first step is to
-estimate the viral phylogeny, and the second step is to use it to estimate the
-transmission tree.
-
-\autocite{didelot2014bayesian} develop a Bayesian version of the two-step
-approach. They first estimate a time-scaled phylogeny from viral samples, then
-estimate the transmissions which took place along the phylogeny. Due to
-within-host evolution, the transmissions are allowed to occur anywhere along
-the branches, rather than being constrained to happen at branching points.
-However, the authors admit that their method requires sampling of every
-infected individual, although they indicate that it could be extended to relax
-this assumption.
-
-A different approach is undertaken by Jombart
-\etal~\autocite{jombart2011reconstructing}, who build transmission trees
-directly from sequence data without the intermediate step of constructing a
-phylogeny. Their method finds the most likely geneology among the sampled
-sequences, assuming that the that the common ancestors to sampled isolates are
-themselves sampled. Although this may be a realistic assumption in the early
-stages of an epidemic of a slower evolving pathogen, it is unlikely for
-ancestors to be present among sequences sampled from an ongoing HIV epidemic
-\comment{Needs a reference}.
-
-Several alternative approaches have been developed which aim to
-incorporate available epidemiological data as well. Cottam
-\etal~\autocite{cottam2008integrating} first built a viral phylogeny, then
-enumerated all transmission trees consistent with this phylogeny by labelling
-the internal nodes of the phylogeny in every possible way. Each of the putative
-transmission trees was assigned a likelihood based on infection time and
-geographic separation.
+\newpage
 

### 2016-02-23 doc/paper/paper.Rnw
Removed generated files; writing

\documentclass[12pt]{report}
\usepackage[osf]{garamondx}
\usepackage[garamondx,cmbraces]{newtxmath}
\usepackage{marginnote}
\usepackage[top=2cm, bottom=2cm, outer=3cm, inner=3cm, heightrounded, 
            marginparwidth=2cm, marginparsep=0.5cm]{geometry}
\usepackage{setspace}
\usepackage[backend=biber, style=authoryear]{biblatex}
\usepackage{nicefrac}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage[acronym,nonumberlist]{glossaries}

\newcommand{\comment}[1]{\marginnote{\textit{#1}}}

\newcommand{\defn}[1]{\textit{#1}}
\newcommand{\software}[1]{\textit{#1}}

\newcommand{\set}[1]{\left\lbrace#1\right\rbrace}
\newcommand{\sett}[1]{\{#1\}}
\renewcommand{\emptyset}{\varnothing}
\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\star}[1]{#1^*}
\renewcommand{\d}{\mathrm{d}\,}

\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

\DeclareMathOperator{\Exponential}{Exponential}
\DeclareMathOperator{\Uniform}{Uniform}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\Var}{Var}

\DeclareMathOperator{\inc}{in}
\DeclareMathOperator{\out}{out}

\newcommand{\N}{\mathcal{N}}
\renewcommand{\L}{\mathcal{L}}

\newacronym{BA}{BA}{Barab\'asi-Albert}
\newacronym{SI}{SI}{susceptible-infected}
\newacronym{kSVM}{kSVM}{kernel support vector machine}
\newacronym{nLTT}{nLTT}{normalized lineages-through-time}
\newacronym{SMC}{SMC}{sequential Monte-Carlo}
\newacronym{ABC}{ABC}{approximate Bayesian computation}
\newacronym{MSM}{MSM}{men who have sex with men}

\graphicspath{{figures/}}

\title{Phylodynamic inference of contact network parameters through approximate
Bayesian computation}
\author{Rosemary M. McCloskey \and Richard H. Liang \and Art F.Y. Poon}

\addbibresource{papers.bib}

\frenchspacing
\onehalfspacing

\begin{document}

\maketitle

\section*{Background}

From a theoretical perspective, network structure can profoundly influence the
incidence and prevalence trajectories of an epidemic, in turn affecting the
estimates of quantities such as effective population size. More practically,
contact networks have been explored as tools for curtailing epidemic spread, by
way of interventions targeted to well-connected
nodes~\autocite{little2014using}.

Survey-based studies of sexual networks~\autocite{liljeros2001web,
schneeberger2004scale} have found that sexual contact networks are best
described by a preferential attachment model \autocite[but not all,
see][]{jones2003assessment}. This result has been coroborated by more recent
investigations using phylogenetic methods~\autocite{brown2011transmission}.
Under these models, nodes with a high number of contacts attract new
connections at an elevated rate. Networks produced by preferential attachment
have a power-law degree distribution, meaning that the number of nodes of
degree $k$ is proportional to $k^\gamma$ for some constant $\gamma$. When $2 <
\gamma \leq 3$, the network is referred to as ``scale-free''. The first contact
network model incorporating preferential attachment was introduced by
\textcite{barabasi1999emergence}, and is now referred to as the \gls{BA} model.
Under this model, networks are formed by iteratively adding nodes with $m$ new
edges each. These new edges are joined to existing nodes of degree $k$ with
probability proportional to $k^\alpha$, so that nodes of high degree tend to
attract more connections (in the original paper, only $\alpha = 1$ was
investigated).

Due to their complexity, it is generally difficult to explicitly calculate the
likelihood of an observed transmission tree under a contact network model. To
calculate the normalizing constant, we would need to integrate over all
possible networks, and also over all possible labellings of the internal nodes
of the transmission tree. While it is not known (to us) whether such
integration is tractable, a simpler alternative is offered by likelihood-free
methods, namely \gls{ABC}. \gls{ABC} leverages the fact that, although
calculating the likelihood may be impossible, generating simulated datasets
according to a model is often straightforward. If our model fits the data well,
the simulated data it produces should be similar to the observed data. More
formally, if $D$ is the observed data, the posterior distribution $f(\theta
\mid D)$ on model parameters $\theta$ is replaced as the target of statistical
inference by $f(\theta \mid \rho(\hat{D}, D) < \varepsilon)$, where $\rho$ is a
distance function, $\hat{D}$ is a simulated dataset according to $\theta$, and
$\varepsilon$ is a small tolerance~\autocite{sunnaker2013approximate}. In the
specific case when $\rho$ is a kernel function, the approach is known as
kernel-\gls{ABC}.

Here, we apply kernel-\gls{ABC} to the problem of statistical inference of
contact network parameters from an estimated transmission tree. We then
estimate the parameters of the \gls{BA} model on a variety of simulated and
real data sets.

\section*{Methods}

We implemented a Gillespie simulation algorithm~\autocite{gillespie1976general}
for simulating epidemics and transmission trees over static contact networks,
as has been done previously \autocite[\textit{e.g.}][]{o2010contact,
robinson2013dynamics, leventhal2012inferring, groendyke2011bayesian}. To check
that our implementation was correct, we reproduced Figure 1 of
\textcite{leventhal2012inferring} (see Figure~\ref{fig:sf1}).

We chose to study the \gls{BA} network model~\autocite{barabasi1999emergence}.
In addition to $m$ and $\alpha$, we investigated the parameters $N$, which
denotes the total number of nodes in the network, and $I$, which is the number
of infected nodes at which to stop the simulation and sample the transmission
tree. Nodes in our networks followed simple \gls{SI} dynamics, meaning that
they became infected at a rate proportional to their numbers of infected
neighbours, and never recovered. For all analyses, the transmission trees'
branch lengths were scaled by dividing by their mean. We used the 
\software{igraph} library's implementation of the \gls{BA}
model~\autocite{csardi2006igraph} to generate the graphs.

\subsection*{Kernel classifiers}

We used the phylogenetic kernel developed by \textcite{poon2013mapping} to test
whether the parameters of the \gls{BA} model had an effect on tree shape. The
parameters not being tested were fixed at the values $N$ = 5000, $I$ = 1000,
$m$ = 2, and $\alpha$ = 1. The parameters were varied one at a time over the
following values: $N$ = \sett{3000, 5000, 8000}, $I$ = \sett{500, 1000, 2000},
$m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0.5, 1, 1.5}. For each parameter
set, 100 networks were generated, and a transmission tree was simulated over
each (300 trees per parameter set). The 300 trees were compared pairwise with
the tree kernel to form a $300 \times 300$ kernel matrix. We constructed a
\gls{kSVM} classifier for the parameter of interest using the
\software{kernlab} package~\autocite{karatzoglou2004kernlab}, and evaluated its
accuracy with 1000 two-fold cross-validations. The construction of the kernel
matrices and classifiers was repeated for several combinations of the kernel
meta-parameters $\lambda$ (the ``decay factor''), and $\sigma$ (the ``radial
basis function variance'')~\autocite[see][]{poon2013mapping}, and for
transmission trees of sizes 100, 500, and 1000. We also tested univariate
classifier based on Sackin's index~\autocite{shao1990tree} and an ordinary SVM
based on the \gls{nLTT} statistic~\autocite{janzen2015approximate}.

\subsection*{ABC simulations}

We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
\textcite{del2012adaptive}. To ensure our implementation was correct, we
applied it to the same mixture of Gaussians used by
\citeauthor{del2012adaptive} to demonstrate their method (originally used
by~\textcite{sisson2007sequential}). We were able to obtain a close
approximation to the function (see Figure~\ref{fig:smctest}), and attained the
stopping condition used by the authors in a comparable number of steps.

We simulated three transmission trees, each with 500 tips, under every element
of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
\sett{1000, 2000}, $m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0, 0.5, 1, 1.5,
2}. The adaptive \gls{ABC} algorithm was applied to each tree with these
priors: $N$ = Uniform(500, 15000), $I$ = Uniform(500, 5000), $m$ = Uniform(1,
5), and $\alpha$ = Uniform(0, 2). Following \textcite{del2012adaptive} and
\textcite{beaumont2009adaptive}, all proposals were Gaussian, with variance
equal to twice the empirical variance of the particles. Because there must be
at least as many nodes in the network as infected nodes, the proposal ratio was
set to zero whenever $I > N$. The algorithm was run with 1000 particles, 5
simulated datasets per particle, and the ``quality'' parameter set to 0.95. We
use the same stopping criterion as \textcite{del2012adaptive}, namely when the
MCMC acceptance rate dropped below 1.5\%. Point estimates for the parameters
were obtained by taking the highest point of an estimated kernel density on the
final set of particles.

Two further analyses were performed to address potential sources of error. To
evaluate the effect of model misspecification in the case of heterogeneity
among nodes, we generated a network where half the nodes were attached with
power $\alpha$ = 0.5, and the other half with power $\alpha$ = 1.5. The other
parameters for this network were $N$ = 5000, $I$ = 1000, and $m$ = 2. To
investigate the effects of potential sampling bias, we simulated a transmission
tree where the tips were sampled in a peer-driven fashion, rather than at
random. That is, the probability to sample a node was twice as high if any of
that node's network peers had already been sampled. The parameters of this
network were $N$ = 5000, $I$ = 2000, $m$ = 2, and $\alpha$ = 0.5.

\subsection*{Investigation of published data}

We applied our kernel-ABC method to several HIV datasets. Because the \gls{BA}
model generates networks with a single connected component, we specifically
searched for datasets which originated from existing clusters, either
phylogenetically or geographically defined. \textcite{wang2015targeting}
reported on a cohort of acutely infected \gls{MSM} in Beijing, China. Through
phylogenetic analysis, they found that they had ``deeply sampled'' the
local \gls{MSM} network, suggesting that many of the samples could be
epidemiologically related.

We downloaded all sequences associated with each paper from GenBank. Sequences
were multiply aligned using \software{MUSCLE} version 3.8.31, and alignments
were manually inspected with \software{Seaview} version 4.4.2. Phylogenies were
constructed from the nucleotide alignments by approximate maximum likelihood
using \software{FastTree2} version 2.1.7 with the GTR model. Transmission trees
were estimated by rooting and time-scaling the phylogenies by root-to-tip
regression, using a modified version of Path-o-Gen as described
previously~\autocite{poon2015phylodynamic}.

\section*{Results}

\subsection*{Kernel classifiers}

Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
tested. For appropriate choices of the kernel meta-parameters, the classifier
had an $R^2$ above 0.8 for all combinations of tree size and epidemic size. The
$I$ parameter was also classified accurately, with an average $R^2$ of 0.93 for
500-tip trees or 0.70 for 100-tip trees. The $m$ parameter was much harder to
classify, with $R^2$ varying between 0.01 and 0.36 depending on $I$ and the
tree size. Finally, the accuracy of the classifier for $N$ varied widely, from
0.08 with 100 tips and $I = 500$, to 0.82 with 1000 tips and $I = 2000$.
Figure~\ref{fig:crossv} shows the results for $\alpha$; similar figures for the
other parameters can be found in the supplemental materials. Based on
inspection of the cross-validation results, we chose to use the meta-parameters
$\lambda = 0.3$ and $\sigma = 4$ in further analyses, and not to use Sackin's
index or the \gls{nLTT}.

\begin{figure}
  \includegraphics{kernel-alpha-crossv}
  \caption{Cross-validation $R^2$ of kernel-SVM classifier for the $\alpha$
    parameter of the \gls{BA} network model.}
  \label{fig:crossv}
\end{figure}

\section*{Discussion}

The $m$ and $N$ parameters were much harder to recover than $\alpha$ and $I$,
suggesting differing influences of these parameters on tree shape. The $\alpha$
parameter in particular was accurately estimated in most cases, implying that
the strength of preferential attachment has a strong impact on the shape of the
transmission tree.

Our method has a number of caveats, perhaps the most significant being that it
takes a transmission tree as input. In reality, true transmission trees are not
available and must be estimated, often by way of a viral phylogeny. Although
this has been demonstrated to work in practice~\autocite{leitner1996accurate},
the topologies of a viral phylogeny and transmission tree can differ
significantly~\autocite{ypma2013relating}. Additionally, the \gls{BA} model
that we fitted here makes a number of unrealistic modelling assumptions, such
as that the network is connected and static, the nodes obey \gls{SI} dynamics,
and that the underlying behaviour of all nodes is the same. This last is
particularly problematic, as we showed by simulating a network where some nodes
exhibited a higher attachment power than others. The estimated attachment power
was simply the average of the two values, indicating that, although we could
characterize the network in aggregate, the estimated parameters could not be
said to apply to any individual node. Finally, the method is computationally
intensive, taking about a day when run on 20 cores in parallel.

\section*{Conclusions}

We developed a novel method which uses kernel-\gls{ABC} to fit contact network
models to transmission trees, and used it to fit the \gls{BA} preferential
attachment model to both simulated and real data.

\setcounter{figure}{0}
\renewcommand{\thefigure}{S\arabic{figure}}

\printbibliography

\section*{Supplemental Materials}

\begin{figure}
  \includegraphics{leventhal2012fig1}
  \caption{Reproduction of Figure 1A from \textcite{leventhal2012inferring}.}
  \label{fig:sf1}
\end{figure}

\begin{figure}
  \includegraphics{smc_test}
  \caption{Approximation of mixture of Gaussians used by
    \textcite{del2012adaptive} and \textcite{sisson2007sequential} to test
    \gls{SMC}.}
  \label{fig:smctest}
\end{figure}

\end{document}

### 2016-02-23 doc/thesis/intro.tex
Removed generated files; writing

@@
 lengths measure genetic distance in units of expected substitutions per site,
-whereas transmission trees are rooted with branches measuring calendar time.
-We generally assume that sampling a virus from the individual also corresponds
-to their removal from the transmission tree\comment{needs a reference}, so the
+whereas transmission trees are rooted with branches measuring calendar time. We
+generally assume that sampling a virus from the individual also corresponds to
+their removal from the transmission tree\comment{needs a reference}, so the
 positions of the tips in time are fixed. Therefore, to transform a viral
@@
 globally similar evolutionary rates. Methods for time-scaling a phylogeny
-include root-to-tip regression~\autocite{drummond2003inference}, which we apply
-in this work, and least-square dating~\autocite{to2015fast}. Both of these
-methods can be used to root the tree, by simply trying all possible root
-positions and choosing the one which minimizes a loss function
-($1-R^2$ or root-mean-square error for root-to-tip regression, sum of squared
-errors for least-square dating). Alternatively, the tree may be rooted with an
-outgroup before time-scaling.
+include root-to-tip regression~\autocite{shankarappa1999consistent,
+korber2000timing, drummond2003inference}, which we apply in this work, and
+least-square dating~\autocite{to2015fast}. Both of these methods can be used to
+root the tree, by simply trying all possible root positions and choosing the
+one which minimizes a loss function ($1-R^2$, or root-mean-square error, for
+root-to-tip regression; sum of squared errors for least-square dating).
+Alternatively, the tree may be rooted with an outgroup~\autocite{li1988rates}
+before time-scaling.
 
@@
 
-%\subsection{Species trees and transmission trees}
-%\label{subsubsec:speciestree}
-%
-%A \defn{species tree} is a particular type of phylogeny in which the taxa are
-%species, and the branching times correspond to historical speciation events. By
-%speciation events, we mean times when two formerly interbreeding groups of
-%organisms ceased to reproduce with each other. In general, we do not have
-%access to the true tree relating a particular group of extant species, as this
-%would usually imply certain knowledge of speciation events in the distant past.
-%Therefore, all species trees considered by researchers are estimates, based on
-%the genetic or phenotypic similarity of the extant taxa, the fossil record, or
-%both. We discuss some of the challenges associated with this estimation in
-%subsection \ref{subsubsec:treeconv} below. However, for the moment, we ignore
-%these complications and consider what can be said about the true species tree.
-%
-%Speciation often occurs by an \defn{allopatric} process, where a sub-population
-%of organisms is isolated from the original population by a geographic
-%barrier~\autocite{coyne2004speciation}. Over time, differing selection
-%pressures on either side of the barrier, coupled with genetic drift, cause the
-%two populations to diverge genetically. This eventually results in two distinct
-%species, which will appear as two subtrees or \defn{clades} in their species
-%tree if descendants of both survive to the present. Of course, this is not
-%necessarily the case, as selection may cause one or both of the new
-%sub-populations or their descendants to become extinct. On the other hand,
-%selection may favour one of the sub-populations, allowing it to expand into new
-%regions or niches and possibly paving the way for subsequent speciation events.
-%Speciation is not necessarily allopatric: \defn{sympatric} speciation refers to
-%the emergence of a new species without any geographic
-%barrier~\autocite{coyne2004speciation}. Indeed, there is a whole continuum of
-%possible speciation processes~\autocite{fitzpatrick2008if}, ranging from
-%completely allopatric to completely sympatric. However, at least for eukaryotic
-%organisms, allopatric speciation is thought to be by far the most common
-%type~\autocite{coyne2004speciation}. 
-%
-%At first glance, transmission trees and species trees seem to have nothing to
-%do with each other. One represents the evolutionary history of a group of
-%organisms, the other traces the spread of a disease through a host population.
-%However, when considering epidemics of RNA viruses, transmission trees and
-%species trees are highly related. The viral population within a single host has
-%often been referred to as a \defn{quasispecies}~\autocite{domingo2012viral},
-%and due to the fast evolutionary rate of these viruses, two hosts' viral
-%populations are invariably genetically distinct. The transmission process is
-%functionally identical to allopatric speciation, involving geographic isolation
-%of a sub-population in a new host and subsequent genetic divergence. However,
-%for the most part, it is not selection but rather host epidemiology which
-%causes certain lineages to proliferate while driving others to extinction. A
-%host who recovers, dies, or is isolated equates to an extinction in the
-%transmission tree. On the other hand, dense populations in frequent contact
-%often experience rapid epidemic growth, which causes similarly rapid
-%accumulation of branching points in the tree. 
-%
-%The fact that these evolutionary (ie. speciation) and epidemiological processes
-%occur on the same time scale for RNA viruses~\autocite{drummond2003measurably}
-%is the basis for the field of
-%\defn{phylodynamics}~\autocite{grenfell2004unifying}. The similarity between
-%transmission and speciation means that it is possible to apply many of the
-%tools and techniques developed for species trees to viral transmission trees,
-%and interpret the results in the context of transmission rates and host
-%epidemiology. 
-%
-%%As a basic example, the lineages-through-time
-%%plot~\autocite{nee1992tempo}, which plots the historical speciation rate
-%%against time, can be used to quantify the incidence of new infections over the
-%%course of an epidemic~\autocite{holmes1995revealing}. 
-%In
-%subsection~\ref{subsubsec:treeshape} below, we review some of these methods in
-%more detail, as well as several which have been developed specifically for
-%viral phylodynamics, and discuss what these methods can tell us about host
-%epidemiology.
-%
-%Finally, we must note two important points. First, we have not yet considered
-%the selection pressures operating within a single host, such as the immune
-%response. Second, we reiterate that both transmission trees and species trees
-%are theoretical objects which are almost never known with certainty for real
-%data. Indeed, the estimation of transmission trees is often problematic, and it
-%is in such estimation that within-host forces play a role. Both of these issues
-%are discussed further below in subsection~\ref{subsubsec:treeconv}, though we
-%continue to ignore them in the next subsection.
-%
-%\subsection{Gene trees and viral phylogenies}
-%\label{subsubsec:genetree}
-%
-%A \defn{gene tree} is another type of phylogeny which traces the evolutionary
-%history of one or more sections of genetic material. Gene trees can be built
-%from small parts of genes up to whole genomes, but for ease of exposition we
-%will use only the word ``gene''. The tips of a gene tree represent the
-%different \defn{alleles}, or versions, of the gene found in the studied taxa.
-%The branching points in a gene tree represent genetic divergence events - that
-%is, the times when two distinct alleles first came into existence through
-%mutation.
-%
-%Gene trees are usually estimated using the nucleotide or amino acid sequences
-%of the genes in question. Although there are methods for estimating gene trees
-%using other types of molecular data, such as gene orders, we will focus on
-%sequence-based methods here. As with species trees, it is not possible to know
-%gene trees with certainty, only to estimate them. However, the problem of
-%estimating gene trees directly from data is somewhat better defined than that
-%of estimating species trees, because of the quantitative nature of the
-%molecular data used to label the tips. In general, methods to estimate gene
-%trees fall into three major categories: distance-based, maximum parsimony, and
-%model-based. The first two categories are generally
-%no longer used for phylogenetic inference, since model-based methods have been
-%demonstrated to be almost universally more accurate. Therefore, we will review
-%in detail only the last category.
-%
-%Model-based methods include both maximum likelihood and Bayesian approaches. To
-%use them, one must first define a parametric model of sequence evolution, here
-%denoted $M$. For all possible trees $T$ (with branch lengths) and choices of
-%model parameters $\theta$, the model must define a probability density function
-%on observed sequence data $D$, denoted $f(D \mid T,\,\theta)$. This is also
-%sometimes written $\Pr(D \mid T,\,\theta)$, even though it is a probability
-%\emph{density}, not a probability. Maximum likelihood aims to find a particular
-%tree and values for the model parameters which optimize the \defn{likelihood
-%function} $\L$ for the observed data,
-%\[
-%  \L(\theta,\, T \mid D) = f(D \mid T,\, \theta).
-%\]
-%Bayesian methods, rather than trying to find point estimates for $\theta$ and
-%$T$, aim to approximate the entire \defn{posterior distribution} for the
-%observed data,
-%\[
-%  f(T,\,\theta \mid D) = \frac{f(D \mid T,\,\theta)f(T,\,\theta)}{f(D)}.
-%\]
-%Here, $f(T, \theta)$ denotes the prior distribution on $T$ and $\theta$, and
-%$f(D)$ is the marginal density of the data (that is, the integral of the
-%numerator over all values of $T$ and $\theta$).
-%
-%\subsection{Estimating transmission trees}
-%\label{subsubsec:treeconv}
-%
-%Most applications of viral phylodynamics, including this project, aim to answer
-%epidemiological questions~\autocite{pybus2009evolutionary, volz2013viral}. 
-%For example, phylodynamic methods have been used to investigate the degree of
-%clustering~\autocite{hughes2009molecular} and the effect of elevated
-%transmission risk in acute infection~\autocite{volz2012simple} 
-%(a more thorough
-%review is given below in subsection~\ref{subsubsec:appphylo}). For these
-%applications, we are ultimately interested in the transmission tree, which is a
-%representation of an epidemiological process (transmission), as opposed to the
-%viral phylogeny, which represents a biological process (viral evolution). As
-%discussed above in subsection~\ref{subsubsec:speciestree}, transmission trees
-%and are in practice unknown, and must be estimated from available data. Here,
-%we review some of the ways to perform such estimation.
-%
-%In principle, transmission trees can be reconstructed by on-the-ground
-%epidemiological methods, that is, by asking each person who they contacted.
-%This is particularly relevant for sexually transmitted diseases such as HIV,
-%where individuals are more likely to recall who they contacted and
-%when~\autocite{klovdahl1985social}. This kind of data is challenging to collect
-%in detail. Most commonly, when applying phylodynamics in practice, the
-%transmission tree is estimated from the viral
-%phylogeny~\autocite{volz2013viral}. As discussed above in
-%subsection~\ref{subsubsec:genetree}, the viral phylogeny is itself also estimated;
-%however, for the moment we ignore this complication and describe methods for
-%estimating the transmission tree if the viral phylogeny were known.
-%
-%When we consider the fact that the viral phylogeny is not known with certainty,
-%the methods just described follow a two-step procedure. The first step is to
-%estimate the viral phylogeny, and the second step is to use it to estimate the
-%transmission tree.
-%
-%\autocite{didelot2014bayesian} develop a Bayesian version of the two-step
-%approach. They first estimate a time-scaled phylogeny from viral samples, then
-%estimate the transmissions which took place along the phylogeny. Due to
-%within-host evolution, the transmissions are allowed to occur anywhere along
-%the branches, rather than being constrained to happen at branching points.
-%However, the authors admit that their method requires sampling of every
-%infected individual, although they indicate that it could be extended to relax
-%this assumption.
-%
-%A different approach is undertaken by Jombart
-%\etal~\autocite{jombart2011reconstructing}, who build transmission trees
-%directly from sequence data without the intermediate step of constructing a
-%phylogeny. Their method finds the most likely geneology among the sampled
-%sequences, assuming that the that the common ancestors to sampled isolates are
-%themselves sampled. Although this may be a realistic assumption in the early
-%stages of an epidemic of a slower evolving pathogen, it is unlikely for
-%ancestors to be present among sequences sampled from an ongoing HIV epidemic
-%\comment{Needs a reference}.
-%
-%Several alternative approaches have been developed which aim to
-%incorporate available epidemiological data as well. Cottam
-%\etal~\autocite{cottam2008integrating} first built a viral phylogeny, then
-%enumerated all transmission trees consistent with this phylogeny by labelling
-%the internal nodes of the phylogeny in every possible way. Each of the putative
-%transmission trees was assigned a likelihood based on infection time and
-%geographic separation. 
-%
-%
-
-\newpage
-
 \subsection{Tree shapes}
@@
 epidemic, the virus, or its hosts and their behaviour, by studying a phylogeny,
-most often a transmission tree~\autocite{pybus2009evolutionary, volz2013viral}.
-Phylogenies are complex objects, and it is not immediately obvious how to
-extract useful information from them with respect to fitting a parameter.
-Standard statistical methods built for numeric data cannot be applied directly
-- for example, one cannot perform a regression of a parameter of interest
-against a phylogeny. Therefore, before we discuss exactly what phylodynamics
-can tell us about epidemics, and how it has been applied in the past, we will
-review some of the methods for quantifying the shapes of phylogenies and their
-similarity to each other.
+most often an estimated transmission tree~\autocite{pybus2009evolutionary,
+volz2013viral}. Phylogenies are complex objects, and it is not immediately
+obvious how to extract useful information from them with respect to fitting a
+parameter. Standard statistical methods built for numeric data cannot be
+applied directly - for example, one cannot perform a regression of a parameter
+of interest against a phylogeny. Therefore, before we discuss exactly what
+phylodynamics can tell us about epidemics, and how it has been applied in the
+past, we will review some of the methods for quantifying the shapes of
+phylogenies and their similarity to each other.
 
-Many tree summary statistics have been developed to assign numerical values to
-phylogenies based on their properties. One of the most widely used is Sackin's
-index~\cite{shao1990tree}, which measures the imbalance or asymmetry in a
+The shape of a phylogeny consists of two components: the topology, and the
+distribution of branch lengths~\autocite{mooers1997inferring}. Many tree
+summary statistics have been developed to assign numerical values to
+phylogenies based on their shapes. One of the most widely used is Sackin's
+index~\autocite{shao1990tree}, which measures the imbalance or asymmetry in a
 rooted tree. For the $i$th tip of the tree, we define $N_i$ to be the number of
@@
 \gls{ltt} plots are normalized so that they begin at $(0, 0)$ and end at $(1,
-1)$, and the difference between the two plots is integrated between 0 and 1.
-In the context of infectious diseases, the \gls{nltt} is related to the
+1)$, and the difference between the two plots is integrated between 0 and 1. In
+the context of infectious diseases, the \gls{ltt} is related to the
 prevalence~\autocite{holmes1995revealing}, so large values may indicate that

### 2016-02-23 doc/thesis/objective.tex
Removed generated files; writing

@@
 epidemiological observations, we employ viral phylogenies. The pioneering study
-in this group was by
-\citeauthor{britton2002bayesian}~(\citeyear{britton2002bayesian}), who
-developed a Bayesian method to infer the $p$ parameter of an \gls{ER} network,
-along with the $\beta$ and $\gamma$ parameters of \gls{SI} model. Their method
-is able to use either infection and removal times, or removal times only.
-However, it is designed for only a small number of observations (15 and 42
-cases in their applications), and their estimates of $p$ for real outbreaks
-were mostly uninformative (95\% confidence intervals [0.11-0.96] and
-0.055-0.96]; $p$ is bounded in [0, 1]).
-\citeauthor{groendyke2011bayesian}~(\citeyear{groendyke2011bayesian})
-significantly updated and extended the methodology of
-\citeauthor{britton2002bayesian}, and applied it to a measles outbreak
-affecting 188 individuals. They were able to obtain a much more informative
-estimate of $p$, although this data set included both symptom onset and
-recovery times for all individuals. They also assumed that the entire network
-was infected, so that the network size was exactly 188. It is unclear how well
-their method would scale to networks with thousands of nodes.
+in this group was by \textcite{britton2002bayesian}, who developed a Bayesian
+method to infer the $p$ parameter of an \gls{ER} network, along with the
+$\beta$ and $\gamma$ parameters of \gls{SI} model. Their method is able to use
+either infection and removal times, or removal times only. However, it is
+designed for only a small number of observations (15 and 42 cases in their
+applications), and their estimates of $p$ for real outbreaks were mostly
+uninformative (95\% confidence intervals [0.11-0.96] and 0.055-0.96]; $p$ is
+bounded in [0, 1]).
+\textcite{groendyke2011bayesian} significantly updated and extended the
+methodology of \citeauthor{britton2002bayesian}, and applied it to a measles
+outbreak affecting 188 individuals. They were able to obtain a much more
+informative estimate of $p$, although this data set included both symptom onset
+and recovery times for all individuals. They also assumed that the entire
+network was infected, so that the network size was exactly 188. It is unclear
+how well their method would scale to networks with thousands of nodes.
+
+% TODO: I don't know where to put this so it's going here for now.
+Several studies~\autocite{little2014using, wang2015targeting,
+brown2011transmission, schneeberger2004scale} have found that the preferential
+attachment model provides a good fit to MSM networks.
+
+The phylogenetic network may show little to no agreement with a contact data
+obtained through epidemiological methods~\autocite{yirrell1998molecular,
+resik2007limitations}.

### 2016-02-23 doc/thesis/results.tex
Removed generated files; writing

@@
 Our first simulation-based experiment was designed to determine which network
-model parameters were potentially estimable by kernel-ABC. To do this, we
-simulated trees under three distinct values of each parameter of interest, and
-evaluated the accuracy of a kernel-\gls{SVM} classifier for the parameter. This
-is the same approach taken by \textcite{poon2015phylodynamic}, and has the dual
-purpose of allowing us to identify the optimal kernel meta-parameters
-\gls{sigma} and \gls{lambda}. The parameters of the \gls{ER}, \gls{WS}, and
-\gls{BA} network models were examined, but we will discuss only the results for
-the $\alpha$ and prevalence parameters of the \gls{BA} model in the main text.
-The results for the remaining models and their parameters can be found in the
-supplemental material.
+model parameters were potentially estimable by kernel-ABC, and to find the best
+set of kernel meta-parameters to carry through to future experiments. To do
+this, we simulated trees under three distinct values of each parameter of the
+\gls{BA} model, and evaluated the accuracy of a kernel-\gls{SVM} classifier for
+the parameter.
 

### 2016-03-01 doc/big-retreat/poster.tex
Added figures and started poster

\documentclass[14pt, a0paper, landscape]{tikzposter}
\usepackage[sfmath]{kpfonts}
\usepackage[default]{gillius}
\usepackage[backend=biber, style=authoryear, uniquelist=minyear, uniquename=false]{biblatex}
\usepackage{multirow}
\usepackage{capt-of}
\usepackage{adjustbox}
\graphicspath{{../figures/}}
\usetheme{Board}

\addbibresource{papers.bib}
\setlength{\bibhang}{0pt}

\title{Phylogenetic estimation of contact network parameters with kernel-ABC}
\author{Rosemary M. McCloskey$^1$ and Art F. Y. Poon $^{1,2}$}
\date{March 11, 2015}
\institute{$^1$ BC Centre for Excellence in HIV/AIDS 
           $^2$ Department of Medicine, University of British Columbia}
\begin{document}

\maketitle

\begin{columns}
  \column{0.25}
  \block{Contact networks shape viral transmission trees}{
    \centerline{\includegraphics[width=0.2\textwidth]{contactnet}}
  }
  \block{What can we tell about contact networks from viral sequence data?}{
  }

  \block{Barab\'asi-Albert preferential attachment networks}{

    \begin{itemize}
      \item Nodes are added one-at-a-time until there are $N$ total.
      \item Each time a node is added, $m$ new edges are connected to it.
      \item Each new edge is attached to an existing node of degree $d$ with
        probability proportional to $d^\alpha$.
    \end{itemize}

    \vspace{1cm}
    \includegraphics[width=0.2\textwidth, trim=0 22in 0 0, clip]{alpha-bounds}
    \begin{center}
      \begin{tabular}{cl}
        parameter & meaning \\
        \hline
        $\alpha$ & preferential attachment power: nodes of degree $d$ attract new edges with probability $\propto d^\alpha$ \\
        $m$ & number of new edges added for each new vertex \\
        $I$ & number of infected nodes when epidemic simulation is stopped \\
        $N$ & total number nodes in the network \\
        \hline
      \end{tabular}
    \end{center}
  }

  \block{Approximate Bayesian computation}{
  }

  \column{0.25}
  \block{Phylogenetic kernel lets us compare tree shapes}{
    \nocite{poon2013mapping}
  }

  \block{ABC with sequential Monte Carlo}{
    \centerline{\includegraphics[width=0.2\textwidth]{abc-smc}}
    \nocite{del2012adaptive}
  }

  \block{Varying $\alpha$ and $I$ affects tree shape, but not $m$ and $N$}{
    \centerline{\includegraphics[width=0.2\textwidth]{kernel-kpca}}
  }

  \column{0.25}
  \block{Simulations}{
    \begin{minipage}[t]{0.5\linewidth}
      \begin{tabular}{ccc}
        parameter & values & prior \\
        \hline
        $\alpha$ & 0.0, 0.5, 1.0, 1.5 & Uniform(0, 2) \\
        $m$ & 2, 3, 4 & Uniform(1, 6) \\
        $I$ & 1000, 2000 & Uniform(500, 5000) \\
        $N$ & 5000 & Uniform(500, 15000) \\
        \hline
      \end{tabular}
    \end{minipage}
    \hspace{0.05\linewidth}
    \begin{adjustbox}{valign=t}
      \begin{minipage}[t]{0.45\linewidth}
        Three networks/trees were simulated under each parameter combination.
        Kernel-ABC was used to re-estimate the parameters.
      \end{minipage}
    \end{adjustbox}
  }
  \block{Accurate estimates for $\alpha$, $I$, but not $m$, $N$}{
    \centerline{\includegraphics[width=0.2\textwidth]{abc-point-estimate}}
  }
  \block{Narrow dispersion of $\alpha$, $I$ compared to $m$, $N$}{
    \centerline{\includegraphics[width=0.2\textwidth]{abc-posterior-example}}
  }

  \column{0.25}

  \block{Real data: phylogenetic and geographic clusters}{
    \centering
    \begin{tabular}{ccccc}
      Reference & Risk group & Sequences ($n$) & Location & HIV gene \\
      \hline
      \textcite{niculescu2015recent} & IDU & 136 & Romania & \textit{pol} \\
      \textcite{li2015hiv} & MSM & 280$^1$ & Shanghai, China & \textit{pol} \\
      \textcite{shiino2014phylodynamic} & mixed & 211$^2$ & Japan & \textit{pol} \\
      \textcite{cuevas2009hiv} & mixed & 287 & Basque Country, Spain & \textit{pol} \\
      \textcite{novitsky2013phylogenetic} & \multirow{2}{*}{mixed} & 
      \multirow{2}{*}{218$^3$} &
      \multirow{2}{*}{Mochudi, Botswana} & 
      \multirow{2}{*}{\textit{env}} \\
      \textcite{novitsky2014impact} \\
      N/A & IDU & 399$^4$ & BC, Canada & \textit{pol} \\
      \hline
    \end{tabular}
    \captionof{table}{
      Characteristics of HIV datasets analysed with kernel-ABC. Abbreviations:
      MSM, men who have sex with men; IDU, injection drug users. $^1$subsampled
      from 1261 $^2$subsampled from 3531 $^3$subsampled from 1299 $^4$subsampled 
      from 7923.}
  }

  \block{Transmission trees were estimated from viral sequence data}{
    \centerline{\includegraphics[width=0.2\textwidth]{pipeline}}

    \nocite{edgar2004muscle}
    \nocite{price2010fasttree}
    \nocite{drummond2003inference}
  }

  \block{Parameter heterogeneity among real datasets}{
    \centerline{\includegraphics[width=0.2\textwidth]{realdata-hpd-bc}}
  }
\end{columns}

\begin{columns}
  \column{0.6}
  \block{Summary}{
    \Large
    We developed a kernel-ABC method for inference of structural contact
    network parameters from viral sequence data.

    The Barab\'asi-Albert model is one of the simplest preferential attachment
    models which makes a number of unrealistic modelling assumptions. In
    particular, the average degree of the network is equal to $2m$, which is
    constrained to a multiple of 2. Moreover, we assume that the network is
    static and connected, and that nodes are never removed and are homogeneous
    in their behaviour. These assumptions mean that the model is very likely
    misspecified for the real data in question. However, method can be applied
    to any network model from which simulations can be drawn, leaving open the
    possibility to investigate more realistic models in the future.
  }

  \column{0.4}
  \block{References} {
    \printbibliography[heading=none]
  }
  \block{Funding and acknowledgements} {
    some logos
  }
\end{columns}

\end{document}

### 2016-03-01 doc/committee-meeting-2/presentation.tex
Added figures and started poster

\documentclass{beamer}
\usepackage{tikz}
\usepackage{tikz-qtree}
\usepackage{gillius}
\usepackage{mathdots}
\usepackage{abraces}
\usepackage{etex}
\usepackage{array}
\usepackage[backend=biber]{biblatex}
\usepackage[export]{adjustbox}

% http://tex.stackexchange.com/questions/12703/how-to-create-fixed-width-table-columns-with-text-raggedright-centered-raggedlef
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

\usetikzlibrary{external}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{decorations.pathreplacing}

\definecolor{blue0}{HTML}{EFF3FF}
\definecolor{blue1}{HTML}{BDD7E7}
\definecolor{blue2}{HTML}{6BAED6}
\definecolor{blue3}{HTML}{2171B5}

\usetheme[everytitleformat=regular]{m}
\setbeamertemplate{navigation symbols}{}
\setbeamercolor{background canvas}{bg=white}

\renewcommand\footnoterule{}

\title{Progress update: Phylogenetic inference of contact network parameters with kernel-ABC}
\author[RM \& AFYP]{Rosemary McCloskey and Art FY Poon}
\institute[UBC \& BCCfE]{BC Centre for Excellence in HIV/AIDS \\ University of British Columbia}
\date{February 1, 2016}

\newcommand{\dd}[2]{\frac{\text{d}\,#1}{\text{d}\,#2}}

\begin{document}

\maketitle

\begin{frame}{Table of contents}
\tableofcontents
\end{frame}

\section{Project overview}

\begin{frame}{Objective: network parameters from phylogenies}
  \begin{center}
  \vspace{-0.5cm}
  \begin{tikzpicture}
    [every node/.style={inner sep=2pt}]
    \node (t) {\includegraphics{figures/phylo}};
    \node [below=0 of t, text width=3cm, align=center] {observed \\ phylogeny};
    \uncover<2->{
      \node [left=of t] (t2) {\includegraphics{figures/tree2}};
      \node [above=0 of t2] (t1) {\includegraphics{figures/tree1}};
      \node [below=0 of t2] (t3) {\includegraphics{figures/tree3}};
      \node [below=0 of t3, text width=3cm, align=center] {simulated \\ phylogenies};
    }
    \node [left=of t1] (n1) {\includegraphics{figures/net1}};
    \node [left=of t2] (n2) {\includegraphics{figures/net2}};
    \node [left=of t3] (n3) {\includegraphics{figures/net3}};
    \node [below=0 of n3, text width=3cm, align=center] {contact network \\ model};
    \uncover<2->{
      \draw [->, >=stealth, very thick, color=blue1] (n1) -- (t1);
      \draw [->, >=stealth, very thick, color=blue2] (n2) -- (t2);
      \draw [->, >=stealth, very thick, color=blue3] (n3) -- (t3);
    }
  \end{tikzpicture}
  \end{center}
\end{frame}

\begin{frame}{Our work in context}
  \vspace{-0.5cm}
  \colorbox{blue0}{\includegraphics[width=\textwidth]{figures/stadler2011estimating}}

  \pause
  \colorbox{blue0}{\includegraphics[width=\textwidth]{figures/leventhal2012inferring}}

  \pause
  \colorbox{blue0}{\includegraphics[width=0.75\textwidth]{figures/groendyke2011bayesian}}
\end{frame}

\begin{frame}{Parameter of interest: preferential attachment power $\alpha$}
  \vspace{-0.5cm}
  \includegraphics[height=0.9\textheight, center]{../thesis/figures/pa_power_bounds}
\end{frame}

\begin{frame}{Two similarity measures: nLTT and tree kernel}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=0.9\textwidth]{figures/treekernel}
    \end{column}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=0.9\textwidth]{figures/nltt}
    \end{column}
  \end{columns}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \tiny Poon, Art FY, et al. ``Mapping the shapes of phylogenetic trees
      from human and zoonotic RNA viruses.'' (2013): e78122.
    \end{column}
    \begin{column}{0.5\textwidth}
      \tiny Janzen, Thijs, Sebastian Höhna, and Rampal S. Etienne. ``Approximate
      Bayesian Computation of diversification rates from molecular phylogenies:
      introducing a new efficient summary statistic, the nLTT.'' Methods in
      Ecology and Evolution 6.5 (2015): 566-575.
    \end{column}
  \end{columns}
\end{frame}

\section{What's been done}

\subsection{Methods development}

\begin{frame}{Software}
  \textit{nettree:} simulate an epidemic over a contact network

  \begin{center}
    \includegraphics[scale=0.4, center]{figures/nettree}
  \end{center}
  \vspace{-0.5cm}
  \pause

  \textit{treekernel:} calculate the similarity of two trees
  \pause

  \textit{netabc:} infer contact network parameters from a phylogeny
\end{frame}

\begin{frame}{\textit{netabc} implements adaptive sequential Monte Carlo}
  \vspace{-0.5cm}
  \includegraphics[width=0.8\textwidth, center]{../thesis/figures/abc_smc.pdf}

  {\footnotesize Algorithm developed in: Del Moral, Pierre, Arnaud Doucet, and Ajay
  Jasra. ``An adaptive sequential Monte Carlo method for approximate Bayesian
  computation.'' Statistics and Computing 22.5 (2012): 1009-1020.}
\end{frame}

\subsection{Validation experiments}

\begin{frame}{Experiment 1: identify separable parameters}
  Trees simulated under different parameter values should segregate.

  \includegraphics[width=0.9\textwidth, center]{../thesis/figures/separable}
  \pause

  Use a kernel-SVM classifier to evaluate how well the values can be
  distinguished.
\end{frame}

\begin{frame}{Kernel-PCA to identify separable parameters}
  \includegraphics[width=0.9\textwidth, center]{../thesis/figures/kernel_expt}
\end{frame}

\begin{frame}{Trees with different attachment powers are distinctive}
  \vspace{-0.5cm}
  \includegraphics[width=1.1\textwidth, center]{figures/kernel-pa-tree}
\end{frame}

\begin{frame}{Higher attachment power $\Rightarrow$ more unbalanced trees}
  \vspace{-0.5cm}
  \includegraphics[height=0.9\textheight, center]{figures/kernel-pa-sackin}
\end{frame}

\begin{frame}{Attachment power is separable by the tree kernel}
  \vspace{-0.5cm}
  \includegraphics[height=0.9\textheight, center]{figures/kernel-pa-kpca}
\end{frame}

\begin{frame}{Tree kernel is more informative than Sackin's index}
  \vspace{-0.5cm}
  \includegraphics[height=0.9\textheight, center]{figures/kernel-pa-crossv}
\end{frame}

\begin{frame}{Prevalence parameter: sanity check on nLTT}
  Prevalence = how many nodes are infected before we stop the simulation.

  \includegraphics[width=\textwidth, center]{figures/prevalence}

  \begin{itemize}
      \pause
    \item Rate of new transmissions drops as network becomes saturated.
      \pause
    \item Prevalence should affect coalescence rate.
      \pause
    \item nLTT should be able to distinguish different coalescence rates.
  \end{itemize}
\end{frame}

\begin{frame}{Similar trees from different prevalence epidemics}
  \vspace{-0.5cm}
  \includegraphics[width=1.1\textwidth, center]{figures/kernel-nnode-tree}
\end{frame}

\begin{frame}{Tree balance is not associated with prevalence}
  \vspace{-0.5cm}
  \includegraphics[height=0.9\textheight, center]{figures/kernel-nnode-sackin}
\end{frame}

\begin{frame}{Prevalence is separable by the tree kernel}
  \vspace{-0.5cm}
  \includegraphics[height=0.9\textheight, center]{figures/kernel-nnode-kpca}
\end{frame}

\begin{frame}{Lineages-through-time is informative of prevalence}
  \vspace{-0.5cm}
  \includegraphics[height=0.9\textheight, center]{figures/kernel-nnode-crossv}
\end{frame}

\begin{frame}{Experiment 2: marginal parameter estimates}
  Grid search should be able to infer single parameters accurately when all
  others are fixed.

  \includegraphics[width=0.9\textwidth, center]{../thesis/figures/acc_prec}
\end{frame}

\begin{frame}{Grid search for marginal parameter estimates}
  \includegraphics[width=0.9\textwidth, center]{../thesis/figures/gridsearch_expt}
\end{frame}

\begin{frame}{Grid search precision depends on true $\alpha$ value}
  \vspace{-0.5cm}
  \includegraphics[height=0.9\textheight, center]{figures/gridsearch-pa-kernel}
\end{frame}

\begin{frame}{Accuracy varies for marginal estimates of $\alpha$}
  \vspace{-0.5cm}
  \includegraphics[height=0.9\textheight, center]{figures/gridsearch-pa-error}
\end{frame}

\begin{frame}{Experiment 3: full parameter estimates with ABC-SMC}
  Simulated trees under fixed parameter values; try to recover with ABC.

  \begin{center}
    \begin{tabular}{ccc}
      \textbf{parameter} & \textbf{value} & \textbf{prior} \\
      \hline
      $\alpha$ & 0.0, 0.25, \ldots, 2 & Uniform(0, 2) \\
      $m$ & 2, 5, 8 & fixed \\
      $N$ & 5000 & Uniform(1000, 10000) \\
      $I$ & 1000 & Uniform(500, 2000) \\
      \hline
    \end{tabular}
  \end{center}
\end{frame}

\begin{frame}{ABC point estimates are accurate for $\alpha$, not $N, I$}
  \vspace{-0.5cm}
  \includegraphics[height=0.9\textheight, center]{figures/abc-pa-point}
\end{frame}

\begin{frame}{Priors on $N, I$ do not influence estimates of $\alpha$}
  \vspace{-0.5cm}
  \includegraphics[height=0.9\textheight, page=1, center]{figures/abc-pa-accuracy}
\end{frame}

\begin{frame}{Dispersion of posterior on $N$ depends on true $\alpha$}
  \vspace{-0.5cm}
  \includegraphics[height=0.9\textheight, page=2, center]{figures/abc-pa-accuracy}
\end{frame}

\begin{frame}{Dispersion of posterior on $I$ depends on true $\alpha$}
  \vspace{-0.5cm}
  \includegraphics[height=0.9\textheight, page=3, center]{figures/abc-pa-accuracy}
\end{frame}

\subsection{Real data analysis}

\begin{frame}{HIV in people who inject drugs in BC and China}
  \begin{center}
    \vspace{-1cm}
    \includegraphics[width=\textwidth]{figures/realtrees}

    \begin{tabular}{ccc}
      & BC & China \\
      \hline
      source & CfE & \footnote{\tiny Tee, Kok Keng, et al. ``Temporal and spatial dynamics of human immunodeficiency virus type 1 circulating recombinant forms 08\_BC and 07\_BC in Asia.'' Journal of virology 82.18 (2008): 9206-9215.} \\
      number of sequences & 399 & 314 \\
      known cluster & yes & no \\
      \hline
    \end{tabular}
  \end{center}
\end{frame}

\begin{frame}{BC: informative estimates for $\alpha$, not $N, I$}
  \vspace{-0.5cm}
  \includegraphics[height=0.9\textheight, center]{figures/bctree-posterior}
\end{frame}

\begin{frame}{China: estimating $m$ does not improve estimates of $N, I$}
  \vspace{-0.5cm}
  \includegraphics[height=0.9\textheight, center]{figures/cn-posterior}
\end{frame}

\begin{frame}{Higher attachment power in BC than in China}
  \vspace{-0.5cm}
  \includegraphics[height=0.9\textheight, center]{figures/bc_vs_cn}
\end{frame}

\section{Conclusions and future work}

\begin{frame}{Future work: more parameters; more real data}
  \vspace{-0.25cm}
  \colorbox{blue0}{\includegraphics[height=2.7cm]{figures/li}}

  \colorbox{blue0}{\includegraphics[height=2cm]{figures/claiborne}}

  \colorbox{blue0}{\includegraphics[height=2cm]{figures/rumyantseva}}
\end{frame}

\begin{frame}{Logos!}
  \vspace{-0.5cm}
  \begin{center}
  \includegraphics[width=1in]{logos/btp}
  $\qquad$
  \includegraphics[width=1in]{logos/cfe}
  $\qquad$
  \includegraphics[width=1in]{logos/cihr}

  \includegraphics[width=1in]{logos/genomebc}
  $\qquad$
  \includegraphics[width=1in]{logos/msfhr}
  $\qquad$
  \includegraphics[width=1in]{logos/phc}

  \includegraphics[width=1in]{logos/phcri}
  $\qquad$
  \includegraphics[width=1in]{logos/sphf}
  $\qquad$
  \includegraphics[width=1in]{logos/westgrid}

  \includegraphics[width=0.5in]{logos/ubc}
  $\qquad$
  \includegraphics[width=1in]{logos/nserc}
  \end{center}
\end{frame}

\section{Appendix: clustering}

\begin{frame}{Appendix: replicating clustering in observed HIV tree}
  \vspace{-0.5cm}
  \includegraphics[width=\textwidth, trim=0 0 0 2cm, clip=true, center]{figures/pcbr}
\end{frame}

\begin{frame}{Cliques with baseline rate do not give similar trees}

  \begin{columns}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth, trim=0 0 0 1.5cm, clip=true]{figures/net20}
    \end{column}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth, trim=0 0 20cm 4cm, clip=true]{figures/tree20}
    \end{column}
  \end{columns}
\end{frame}

\end{document}

### 2016-03-01 doc/committee-meeting-2/summary.tex
Added figures and started poster

\documentclass{article}
\usepackage[T1]{fontenc}
\usepackage{garamondx}
\usepackage[garamondx,cmbraces]{newtxmath}
\usepackage{fullpage}
\usepackage[backend=biber, sorting=none]{biblatex}
\usepackage{setspace}

\newcommand{\sw}[1]{\textit{#1}}
\newcommand{\etal}{\textit{et al.}}

\title{Phylogenetic estimation of contact network parameters with kernel-ABC}
\author{Rosemary McCloskey \and Art FY Poon}
\addbibresource{papers.bib}

\begin{document}

\maketitle

\onehalfspacing

\section{Abstract}

Models of the spread of HIV in a population often make the simplifying
assumption that the population is homogeneously mixed, or is divided into
homogeneously mixed compartments. However, human populations are likely to have
complex structures formed by social contacts; this structure can have a
significant influence on the rate of epidemic spread. Contact network models
capture this structure by explicitly representing each possible transmission.

We developed a tree kernel-based approximate Bayesian computation (kernel-ABC)
method for estimating structural parameters of the contact network underlying
an observed HIV transmission tree, and used it to investigate the preferential
attachment (PA) power parameter of the Barab\'asi-Albert network model. To
validate the method, we constructed networks of 5000 nodes with mean degree 4
and PA power values \{0, 0.25, \ldots, 2\}. Epidemics were simulated over these
networks until 1000 nodes were infected, then transmission trees were obtained
by sub-sampling 500 tips each. We also applied the method to a viral phylogeny
relating a cluster of 399 HIV infections in British Columbia.

On simulated data, we were able to estimate the PA power: the median [IQR]
absolute deviation of our estimate from the true value was 0.08 [0.03-0.18],
with 95\% confidence intervals of width 0.57 [0.38-0.73]. However, our estimates
of the number of infected nodes in these networks, using a uniform prior on
[500, 2000], were mostly uninformative (1130, 95\% CI [645-1761], true value
1000). The estimated PA power of the contact network for the HIV phylogeny in
British Columbia was 1.01 (0.94-1.05).

Using a novel kernel-ABC method to estimate contact network parameters from
virus phylogenies, we found that structural parameters such as PA power can be
recovered. Our method offers the potential to quantitatively investigate social
network dynamics underlying the spread of viral epidemics.

\section{Background}

The domain of this work is the emerging field of
phylodynamics~\autocite{grenfell2004unifying}. Briefly, phylodynamics is the
study of the interaction between epidemiological and evolutionary processes, in
organisms such as RNA viruses where these occur on the same time
scale~\autocite{drummond2003measurably}. In particular, phylodynamic methods
provide a way to use evolutionary data, like phylogenies, to estimate
epidemiological quantities, like transmission rates.

One type of epidemiological data we might want to investigate with
phylodynamics is the structure of the host population. This includes questions
such as how many connections a person has on average, or whether there are
``superspreaders'' responsible for a disproportionately high number of
transmissions. A contact network is a graphical representation of a population
of individuals (nodes) and potential transmissions between them (edges).
Contact networks models remove the assumption of homogeneously mixed
populations used in many epidemiological investigations, and can potentially
offer much more detailed information about the structure of a population
experiencing an epidemic. However, contact network models can be challenging to
work with. Unlike standard compartmental models, one cannot generally write
down a set of equations describing the epidemic behaviour over time, nor
calculate the likelihood of observed data for a given contact network.

Approximate Bayesian computation (ABC) is a statistical method for estimating
parameters of models for which explicit likelihoods cannot be calculated. The
idea of ABC is to simulate data according to a broad range of possible model
parameters, and prefer parameters which produce simulated data similar to the
observed data. Contact network models are a good candidate for the use of ABC,
since simulating networks, and transmission trees over those networks, is
straightforward.

\section{Objective}

The goal of this work is to develop and validate a method to quantitatively
infer structural parameters of contact networks from observed transmission
trees. This idea fits in between three established bodies of work. First, there
are phylodynamic studies which estimate epidemiological parameters, such as the
transmission rate or basic reproductive number
(e.g.~\autocite{stadler2011estimating, volz2012simple}). These studies have
provided valuable insight into emerging epidemics, however they typically
either ignored any population structure or assumed the population was
panmictic. Second, there have been several \emph{qualitative} investigations of
the impact of contact network structure on phylogeny shape
(e.g.~\autocite{leventhal2012inferring, colijn2014phylogenetic, o2010contact}).
Third, there have been a couple of attempts to discern contact network
parameters from other epidemiological data, like infection and recovery
times~\autocite{britton2002bayesian, groendyke2011bayesian}, but these have
been limited to the edge density parameter $p$ in networks with randomly chosen
edges. So far as we are aware, there have been no attempts to
\emph{quantitatively} estimate contact network parameters other than $p$, and
no attempts at all to estimate these parameters directly from viral sequence
data.

\section{Methods}

\subsection{Programs}

Three software programs were written for this work: \sw{nettree},
\sw{treekernel}, and \sw{netabc}. \sw{Nettree} simulates a transmission tree
over a supplied contact network using Gillespie
simulation~\autocite{gillespie1976general}. \sw{Treekernel} is a fast
implementation of the phylogenetic kernel of Poon
\etal~\autocite{poon2013mapping} using the algorithm of
Moschitti~\autocite{moschitti2006making}. \sw{Netabc} is an implementation of
the adaptive sequential Monte-Carlo algorithm developed by Del Moral
\etal~\autocite{del2012adaptive}, tailored for approximate Bayesian computation
using the phylogenetic kernel (kernel-ABC).

\subsection{Simulation experiments}

I performed three sets of simulation experiments to validate the kernel-ABC
method. The first two sets have either already been performed, or will be in
the near future, for all the parameters of three contact network models:
Erd\H{o}s-R\`enyi or random networks~\autocite{erdos1960evolution},
Watts-Strogatz or small-world networks~\autocite{watts1998collective}, and
Barab\'asi-Albert or preferential attachment
networks~\autocite{barabasi1999emergence}. However, due to computational
constraints, I have so far only performed the third set of experiments for the
preferential attachment power parameter (henceforth $\alpha$) of the
Barab\'asi-Albert network model. Therefore, I will focus only on this parameter
for this meeting. $\alpha$ indicates the strength of attraction to popular
nodes: as $\alpha$ increases, so does the tendency of other nodes to attach
themselves to nodes of high degree.

The first experiment was similar to the validation performed previously by
Poon~\autocite{poon2015phylodynamic}. It is designed to ensure that the chosen
parameters are identifiable, in that they produce transmission trees which are
different enough from each other that the parameter can be recovered. This
analysis also serves the purpose of checking that the tree kernel is the most
effective distance measure to use for ABC, and of identifying the optimal
meta-parameters for tuning the kernel. I simulated 100 trees each under three
different values of $\alpha$ (0.5, 1.0, and 1.5), constructed a pairwise
distance matrix relating the 300 trees, and estimated the accuracy of a
kernel-SVM classifier trained with this matrix by cross-validation. This was
repeated for several different sets of simulation and tree kernel parameters.

The second experiment was a marginal grid search, designed to investigate our
ability to infer a particular model parameter when all others are fixed. For
each of nine different $\alpha$ values (0.0, 0.25, \ldots, 2.0), ten testing
trees were simulated. These were compared using the tree kernel to 6030
training trees, which had been simulated on a finely spaced grid of $\alpha$
values between 0 and 2. The $\alpha$ value with the highest median tree kernel
score was taken as the point estimate of $\alpha$ for the testing tree. These
point estimates were compared to the true values.

The third and final experiment tested the ability of the full ABC algorithm to
infer the $\alpha$ parameter. Ten transmission trees were simulated under each
of the same nine $\alpha$ values as in the previous experiment. The
simulations were performed on networks of 5000 nodes, 1000 of which eventually
became infected. We assumed a uniform prior on $[1000, 10000]$ for the number
of nodes in the network, on $[500, 2000]$ for the number of infected nodes, and
$[0, 2]$ for $\alpha$. The mean degree of the network was fixed at either 4,
10, or 16. In each case, we evaluated the accuracy and precision of the
posterior distributions on $\alpha$, as well as those on the number of nodes
and number of infected.

\subsection{Real world application}

I applied the ABC algorithm to a phylogeny relating 399 HIV infections in
British Columbia, which had previously been identified as being part of a
phylogenetic cluster. The priors used were the same as in the previous section.
Because this application involved only one replicate, I ran the ABC algorithm
with a ten-fold higher number of particles, to increase the resolution and
hopefully obtain a narrower confidence interval.

\section{Results}

\subsection{Simulation experiments}

The first validation experiment showed that trees simulated under different
values of the $\alpha$ parameter were sufficiently distinct to be identifiable
with the tree kernel. With suitable kernel meta-parameters, the $R^2$ of the
kernel-SVM classifier was above 0.95 for 1000- and 500- tip trees, dropping to
0.85 for 100-tip trees. Surprisingly, multiplication of the tree kernel by the
normalized lineages-through-time statistic
(see~\autocite{janzen2015approximate}) actually reduced accuracy somewhat. This
prompted us to perform a sanity check by doing the same type of experiment for
the number of infected nodes (henceforth $I$) under a full network model, where
every node is connected to every other node. The full graph satisfies the
assumption of panmixia required for Kingman's coalescent, so the
lineages-through-time statistic should be informative. We found that under the
full model, the $R^2$ did improve slightly with the addition of the
lineages-through-time statistic.

In the second experiment using grid search, we found that the point estimates
for $\alpha$ were quite accurate. The least accurate estimates were for the
true value of $\alpha = 0$, which had a mean of 0.17 and a standard error of
0.05 across replicates. The most accurate were for the true value of $\alpha =
1.25$, which had a mean of 1.25 with standard error 0.01.

Finally, in the third experiment using the full ABC algorithm, we were able to
recover the $\alpha$ parameter well in most cases, although the accuracy was
lower than with grid search. This was expected, since we were not fixing the
other model parameters. The least accurate point estimates were again for
$\alpha = 0$, with mean 0.31 and standard error 0.03. The most accurate were
for $\alpha = 1.25$, with mean 1.2 and standard error 0.01. The 95\% confidence
intervals on the parameters were fairly wide, averaging 0.64 across all
experiments. The estimates of the number of nodes and number of infected nodes
were mostly uninformative, with 95\% confidence intervals spanning almost the
entire prior region (average width 7313 and 1033 respectively, for prior
regions [1000, 10000] and [500, 2000]).

\subsection{Real world application}

We estimated the preferential attachment power of the real HIV network to be
1.01, with a 95\% confidence interval of [0.94, 1.05]. The reduced size of the
confidence interval compared to our simulations was likely due to the increased
number of particles in the ABC algorithm. As before, the estimates of numbers
of total nodes and infected nodes were uninformative, however the estimate of
$\alpha$ was robust to changing the mean degree of nodes in the network.

\section{Future work}

Two major tasks remain. First, I plan to perform the ABC validation experiments
for other parameters and network model types, namely random
networks~\autocite{erdos1960evolution} and small-world
networks~\autocite{watts1998collective}. This is a straightforward modification
of the existing experimental setup for $\alpha$, although the actual experiment
will take significant time. Second, I would like to find additional real-world
datasets to which to apply the method. Currently, I am running the algorithm on
a dataset from the CRF07 subtype of HIV in China, and am curating additional
datasets from Genbank.

\singlespacing
\printbibliography

\end{document}

### 2016-03-01 doc/journal-club/feb25/goodreau2006assessing.tex
Added figures and started poster

\documentclass{beamer}
\usepackage{tikz}
\usepackage{tikz-qtree}
\usepackage{gillius}
\usepackage{mathdots}
\usepackage{abraces}
\usepackage{etex}
\usepackage{array}
\usepackage[backend=biber]{biblatex}

% http://tex.stackexchange.com/questions/12703/how-to-create-fixed-width-table-columns-with-text-raggedright-centered-raggedlef
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

\usetikzlibrary{external}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{decorations.pathreplacing}

\usetheme[everytitleformat=regular]{m}
\setbeamertemplate{navigation symbols}{}

\title{Assessing the effects of human mixing patterns on human immunodeficiency
  virus-1 interhost phylogenetics through social network simulation}
\author[Goodreau]{Steven M. Goodeau}

\newcommand{\dd}[2]{\frac{\text{d}\,#1}{\text{d}\,#2}}

\begin{document}
\definecolor{red}{RGB}{228,26,28}
\definecolor{blue}{RGB}{55,126,184}
\definecolor{green}{RGB}{77,175,74}
\definecolor{purple}{RGB}{152,78,163}

\maketitle
\begin{frame}{Motivation}
  \begin{itemize}
    \setlength{\itemsep}{12pt}
    \item HIV evolution is affected by both within- and among-host processes.
    \item Host populations are highly structured, which violates assumptions of 
      many population genetic models.
    \item Aim: determine the effect of population structure on population
      size and epidemic growth rate, and evaluate accuracy of estimation
      methods for same.
  \end{itemize}
\end{frame}

\begin{frame}{ERGMs}
  \begin{itemize}
    \setlength{\itemsep}{12pt}
    \item Assigns a probability density to each possible network, as a function
      of several network statistics.
    \item $\Pr(\text{observe graph } x) \propto \exp(\theta' z(x))$, where
      $\theta$ are parameters and $z(x)$ is a vector of network statistics.
    \item Network statistics include number of edges, number of triangles,
      number of edges between nodes of the same ``type'', \&c.
  \end{itemize}
\end{frame}

\begin{frame}{Network types}
  \centerline{\includegraphics[height=0.8\textheight]{F1}}
\end{frame}

\begin{frame}{Simulations}
  \begin{enumerate}
    \setlength{\itemsep}{12pt}
    \item Network: use MCMC, adding and subtracting edges according to their
      log-likelihood.
    \item Epidemic: each sero-discordant partnership has a constant
      transmission rate.
    \item Viral evolution: coalescence rate for $n$ sequences of $n(n-1) /
      2N_eG$, where $N_e$ is the steady-state effective population size, and
      $G$ is viral the generation time. Also mutation rate $\mu_x$ for each
      site $x$.
  \end{enumerate}
\end{frame}

\begin{frame}{Prevalence varies across network types}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth, trim=0 23cm 0 0, clip]{F2}
    \end{column}
    \begin{column}{0.5\textwidth}
      \vspace{0.5cm}
      \includegraphics[width=\textwidth, trim=0 0 0 22cm, clip]{F2}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Simulations fall in expected range for random networks}
  \includegraphics[width=\textwidth]{F3}

  Points are simulations, lines are expected values under panmixis.
\end{frame}

\begin{frame}{Most simulations best fit by exponential growth}
  \centerline{\includegraphics[height=2.5cm, trim=0 15cm 0 0, clip]{F4}}

  \includegraphics[width=\textwidth]{T5}
\end{frame}

\begin{frame}{Predicted and actual pop. size discrepancies}
  \includegraphics[width=\textwidth]{F5}
\end{frame}

\begin{frame}{Growth rate, pop. size may be over- or under-estimated}
  \begin{columns}
    \begin{column}{0.2\textwidth}
      \includegraphics[width=\textwidth]{F6}
    \end{column}
    \begin{column}{0.7\textwidth}
      \vspace{0.5cm}
      \includegraphics[height=0.8\textheight]{F7}
    \end{column}
  \end{columns}
\end{frame}

\end{document}

### 2016-03-01 doc/paper/paper.tex
Added figures and started poster

--- 2016-02-23 doc/paper/paper.Rnw
@@
-\documentclass[12pt]{report}
+\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 \usepackage[osf]{garamondx}
 \usepackage[garamondx,cmbraces]{newtxmath}
-\usepackage{marginnote}
 \usepackage[top=2cm, bottom=2cm, outer=3cm, inner=3cm, heightrounded, 
@@
 \usepackage{setspace}
-\usepackage[backend=biber, style=authoryear]{biblatex}
-\usepackage{nicefrac}
-\usepackage{amsmath}
-\usepackage{algorithm}
-\usepackage{algorithmicx}
-\usepackage{algpseudocode}
+\usepackage[backend=biber, style=authoryear, uniquelist=minyear, uniquename=false]{biblatex}
 \usepackage[acronym,nonumberlist]{glossaries}
-
-\newcommand{\comment}[1]{\marginnote{\textit{#1}}}
+\usepackage{multirow}
 
@@
 \newacronym{MSM}{MSM}{men who have sex with men}
+\newacronym{MCMC}{MCMC}{Markov chain Monte Carlo}
+\newacronym{GTR}{GTR}{generalized time-reversible}
 
@@
 \onehalfspacing
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 
-From a theoretical perspective, network structure can profoundly influence the
-incidence and prevalence trajectories of an epidemic, in turn affecting the
-estimates of quantities such as effective population size. More practically,
+The structure of the contact network underlying a viral epidemic can profoundly
+impact the speed and pattern of the epidemic's spread. Network structure can
+influence the prevalence trajectory~\autocite{o2010contact} and epidemic
+threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates
+of quantities such as effective population
+size~\autocite{goodreau2006assessing}. From a public health perspective,
 contact networks have been explored as tools for curtailing epidemic spread, by
 way of interventions targeted to well-connected
-nodes~\autocite{little2014using}.
+nodes~\autocite{little2014using}. True contact networks are a challenging type
+of data to collect, requiring extensive epidemiological
+investigation~\autocite{welch2011statistical}.
+
+Viral sequence data, on the other hand, has become relatively inexpensive and
+straightforward to collect on a population level. Due to the high mutation rate
+of RNA viruses, epidemiological processes impact the course of viral evolution,
+thereby shaping the intra-host viral
+phylogeny~\autocite{drummond2003measurably}. The term ``phylodynamics'' was
+coined to describe this interaction, as well as the growing family of inference
+methods to estimate epidemiological parameters from viral
+phylogenies~\autocite{grenfell2004unifying}. These methods have revealed
+diverse properties of local viral outbreaks, from basic reproductive
+number~\autocite{stadler2011estimating}, to the degree of
+clustering~\autocite{hughes2009molecular}, to the elevated transmission risk
+during acute infection~\autocite{volz2012simple}. On the other hand, inference
+of structural network parameters has to date been limited. However, it has been
+shown that network structure has a tangible impact on phylogeny
+shape~\autocite{leventhal2012inferring, colijn2014phylogenetic,
+goodreau2006assessing, robinson2013dynamics}, suggesting that statistical
+inference might be possible~\autocite{welch2011statistical}.
 
@@
 schneeberger2004scale} have found that sexual contact networks are best
-described by a preferential attachment model \autocite[but not all,
-see][]{jones2003assessment}. This result has been coroborated by more recent
-investigations using phylogenetic methods~\autocite{brown2011transmission}.
-Under these models, nodes with a high number of contacts attract new
-connections at an elevated rate. Networks produced by preferential attachment
-have a power-law degree distribution, meaning that the number of nodes of
-degree $k$ is proportional to $k^\gamma$ for some constant $\gamma$. When $2 <
-\gamma \leq 3$, the network is referred to as ``scale-free''. The first contact
-network model incorporating preferential attachment was introduced by
-\textcite{barabasi1999emergence}, and is now referred to as the \gls{BA} model.
-Under this model, networks are formed by iteratively adding nodes with $m$ new
-edges each. These new edges are joined to existing nodes of degree $k$ with
-probability proportional to $k^\alpha$, so that nodes of high degree tend to
-attract more connections (in the original paper, only $\alpha = 1$ was
-investigated).
+described by a preferential attachment model \autocite[although there has been
+some disagreement, see][]{jones2003assessment}. Under these models, nodes with
+a high number of contacts attract new connections at an elevated rate. Networks
+produced by preferential attachment have a power-law degree distribution,
+meaning that the number of nodes of degree $k$ is proportional to $k^\gamma$
+for some constant $\gamma$. When $2 < \gamma \leq 3$, the network is referred
+to as ``scale-free''. The first contact network model incorporating
+preferential attachment was introduced by \textcite{barabasi1999emergence}, and
+is now referred to as the \gls{BA} model. Under this model, networks are formed
+by iteratively adding nodes with $m$ new edges each. These new edges are joined
+to existing nodes of degree $k$ with probability proportional to $k^\alpha$, so
+that nodes of high degree tend to attract more connections (in the original
+paper, only $\alpha = 1$ was investigated).
+
+Previous work offers precedent for the possibility of statistical inference of
+structural network parameters. \textcite{britton2002bayesian} develop a
+Bayesian approach to estimate the edge density in an Erd\H{o}s-R\'enyi
+network~\autocite{erdos1960evolution} given observed infection dates, and
+optionally recovery dates. Their approach was later extended by
+\textcite{groendyke2011bayesian} and applied to a much larger data set of 188
+individuals. \textcite{brown2011transmission} fit a preferential attachment
+model to a phylogenetically estimated transmission network relating 60\% of HIV
+infected \gls{MSM} in the United Kingdom. The transmission network is a
+subgraph of the contact network which includes only those edges which have
+already led to a new infection.
 
 Due to their complexity, it is generally difficult to explicitly calculate the
-likelihood of an observed transmission tree under a contact network model. To
-calculate the normalizing constant, we would need to integrate over all
-possible networks, and also over all possible labellings of the internal nodes
-of the transmission tree. While it is not known (to us) whether such
-integration is tractable, a simpler alternative is offered by likelihood-free
-methods, namely \gls{ABC}. \gls{ABC} leverages the fact that, although
-calculating the likelihood may be impossible, generating simulated datasets
-according to a model is often straightforward. If our model fits the data well,
-the simulated data it produces should be similar to the observed data. More
-formally, if $D$ is the observed data, the posterior distribution $f(\theta
-\mid D)$ on model parameters $\theta$ is replaced as the target of statistical
-inference by $f(\theta \mid \rho(\hat{D}, D) < \varepsilon)$, where $\rho$ is a
-distance function, $\hat{D}$ is a simulated dataset according to $\theta$, and
-$\varepsilon$ is a small tolerance~\autocite{sunnaker2013approximate}. In the
-specific case when $\rho$ is a kernel function, the approach is known as
-kernel-\gls{ABC}.
+likelihood of a transmission tree under a contact network model. To do so, we
+would need to integrate over all possible networks generated by that model, and
+also over all possible labellings of the internal nodes of the transmission
+tree. While it is not known (to us) whether such integration is tractable, a
+simpler alternative is offered by likelihood-free methods, namely
+\gls{ABC}~\autocite{rubin1984bayesianly, tavare1997inferring}. \gls{ABC}
+leverages the fact that, although calculating the likelihood may be impossible,
+generating simulated datasets according to a model is often straightforward. If
+our model fits the data well, the simulated data it produces should be similar
+to the observed data. More formally, if $D$ is the observed data, the posterior
+distribution $f(\theta \mid D)$ on model parameters $\theta$ is replaced as the
+target of statistical inference by $f(\theta \mid \rho(\hat{D}, D) <
+\varepsilon)$, where $\rho$ is a distance function, $\hat{D}$ is a simulated
+dataset according to $\theta$, and $\varepsilon$ is a small
+tolerance~\autocite{sunnaker2013approximate}. In the specific case when $\rho$
+is a kernel function, the approach is known as
+kernel-\gls{ABC}~\autocite{nakagome2013kernel}.
 
 Here, we apply kernel-\gls{ABC} to the problem of statistical inference of
-contact network parameters from an estimated transmission tree. We then
-estimate the parameters of the \gls{BA} model on a variety of simulated and
-real data sets.
+contact network parameters from an estimated transmission tree, using the tree
+kernel developed by \textcite{poon2013mapping}. We then estimate the parameters
+of the \gls{BA} model on a variety of simulated and real data sets.
 
@@
 robinson2013dynamics, leventhal2012inferring, groendyke2011bayesian}. To check
-that our implementation was correct, we reproduced Figure 1 of
+that our implementation was correct, we reproduced Figure 1A of
 \textcite{leventhal2012inferring} (see Figure~\ref{fig:sf1}).
@@
 \software{igraph} library's implementation of the \gls{BA}
-model~\autocite{csardi2006igraph} to generate the graphs.
+model~\autocite{csardi2006igraph} to generate the graphs. The analyses were run
+on Westgrid (\url{https://www.westgrid.ca/}) and a local computer cluster.
 
@@
 applied it to the same mixture of Gaussians used by
-\citeauthor{del2012adaptive} to demonstrate their method (originally used
-by~\textcite{sisson2007sequential}). We were able to obtain a close
-approximation to the function (see Figure~\ref{fig:smctest}), and attained the
-stopping condition used by the authors in a comparable number of steps.
+\citeauthor{del2012adaptive} to demonstrate their method \autocite[originally
+used by][]{sisson2007sequential}. We were able to obtain a close approximation
+to the function (see Figure~\ref{fig:smctest}), and attained the stopping
+condition used by the authors in a comparable number of steps.
 
@@
 of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
-\sett{1000, 2000}, $m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0, 0.5, 1, 1.5,
-2}. The adaptive \gls{ABC} algorithm was applied to each tree with these
-priors: $N$ = Uniform(500, 15000), $I$ = Uniform(500, 5000), $m$ = Uniform(1,
-5), and $\alpha$ = Uniform(0, 2). Following \textcite{del2012adaptive} and
+\sett{1000, 2000}, $m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0.0, 0.5, 1, 1.5}.
+The adaptive \gls{ABC} algorithm was applied to each tree with these priors:
+$m \sim$ Uniform(1, 5), $\alpha \sim$ Uniform(0, 2), and $(N, I)$ jointly
+uniform on the triangular region \{$500 \leq N \leq 15000$, $500 \leq I \leq
+ 5000$, $I \leq N$\}. Following \textcite{del2012adaptive} and
 \textcite{beaumont2009adaptive}, all proposals were Gaussian, with variance
-equal to twice the empirical variance of the particles. Because there must be
-at least as many nodes in the network as infected nodes, the proposal ratio was
-set to zero whenever $I > N$. The algorithm was run with 1000 particles, 5
-simulated datasets per particle, and the ``quality'' parameter set to 0.95. We
-use the same stopping criterion as \textcite{del2012adaptive}, namely when the
-MCMC acceptance rate dropped below 1.5\%. Point estimates for the parameters
-were obtained by taking the highest point of an estimated kernel density on the
-final set of particles.
+equal to twice the empirical variance of the particles. The algorithm was run
+with 1000 particles, 5 simulated datasets per particle, and the ``quality''
+parameter set to 0.95. We use the same stopping criterion as
+\textcite{del2012adaptive}, namely when the MCMC acceptance rate dropped below
+1.5\%. Point estimates for the parameters were obtained by taking the highest
+point of an estimated kernel density on the final set of particles.
 
@@
 searched for datasets which originated from existing clusters, either
-phylogenetically or geographically defined. \textcite{wang2015targeting}
-reported on a cohort of acutely infected \gls{MSM} in Beijing, China. Through
-phylogenetic analysis, they found that they had ``deeply sampled'' the
-local \gls{MSM} network, suggesting that many of the samples could be
-epidemiologically related.
+phylogenetically or geographically defined. Characteristics of the datasets we
+investigated are given in Table~\ref{tab:data}.
 
-We downloaded all sequences associated with each paper from GenBank. Sequences
-were multiply aligned using \software{MUSCLE} version 3.8.31, and alignments
-were manually inspected with \software{Seaview} version 4.4.2. Phylogenies were
+\begin{table}
+  \centering
+  \begin{tabular}{ccccc}
+    Reference & Sequences ($n$) & Location & Risk group & Gene \\
+    \hline
+    \autocite{wang2015targeting} & 173 & Beijing, China & MSM & \textit{pol} \\
+    \autocite{cuevas2009hiv} & 287 & Basque Country, Spain & mixed & \textit{pol} \\
+    \autocite{novitsky2013phylogenetic} & \multirow{2}{*}{1299} & \multirow{2}{*}{Mochudi, Botswana} & \multirow{2}{*}{mixed} & \multirow{2}{*}{\textit{env}} \\
+    \autocite{novitsky2014impact} \\
+    \hline
+  \end{tabular}
+  \caption{Characteristics of published datasets investigated with kernel-ABC.
+  Acronyms: MSM, men who have sex with men.}
+  \label{tab:data}
+\end{table}
+
+We downloaded all sequences associated with each study from GenBank. Sequences
+were multiply aligned using \software{MUSCLE} version 3.8.31
+\autocite{edgar2004muscle}, and alignments were manually inspected with
+\software{Seaview} version 4.4.2 \autocite{gouy2010seaview}. Phylogenies were
 constructed from the nucleotide alignments by approximate maximum likelihood
-using \software{FastTree2} version 2.1.7 with the GTR model. Transmission trees
-were estimated by rooting and time-scaling the phylogenies by root-to-tip
-regression, using a modified version of Path-o-Gen as described
+using \software{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
+\gls{GTR} model. Transmission trees were estimated by rooting and time-scaling
+the phylogenies by root-to-tip regression, using a modified version of
+Path-o-Gen \autocite[distributed as part of \software{BEAST},
+][]{drummond2007beast}, as described
 previously~\autocite{poon2015phylodynamic}.
+
+Two additional steps were undertaken for the
+\textcite{novitsky2013phylogenetic, novitsky2014impact} data. First, before
+performing the multiple sequence alignment, we aligned each
+sequence pairwise to the HXB2 reference sequence (GenBank accession number
+HIVHXB2CG) using \software{MAFFT} version 7.130b \autocite{katoh2013mafft} and
+clipped out the hypervariable regions with \software{Biopython}
+version 1.66+ \autocite{cock2009biopython}. Second, after multiple sequence
+alignment, the sequences were trimmed to an informative region with
+\software{trimAl} version 1.4.rev15 \autocite{capella2009trimal}.
+
+Kernel-ABC was used to fit the \gls{BA} model to each dataset. To make the
+analyses comparable, the same priors and configuration parameters were used for
+each dataset. The priors were as follows: $\alpha \sim$ Uniform(0, 2), $m \sim$
+Uniform(1, 5), and $(N, I) \sim$ Uniform(\sett{$n \leq N \leq 10000$, $n \leq I
+\leq 1000$, $I \leq N$}), where $n$ is the number of sequences in the dataset.
+The kernel meta-parameters and SMC configuration were the same as in the
+\gls{ABC} simulations.
 
@@
 tested. For appropriate choices of the kernel meta-parameters, the classifier
-had an $R^2$ above 0.8 for all combinations of tree size and epidemic size. The
-$I$ parameter was also classified accurately, with an average $R^2$ of 0.93 for
-500-tip trees or 0.70 for 100-tip trees. The $m$ parameter was much harder to
-classify, with $R^2$ varying between 0.01 and 0.36 depending on $I$ and the
-tree size. Finally, the accuracy of the classifier for $N$ varied widely, from
-0.08 with 100 tips and $I = 500$, to 0.82 with 1000 tips and $I = 2000$.
+had an $R^2$ above 
+    0.8 
+for all combinations of tree size and epidemic size. The $I$ parameter was also
+classified accurately, with an average $R^2$ of 
+    0.93 
+for 500-tip trees or 
+    0.7 
+for 100-tip trees. The $m$ parameter was much harder to classify, with $R^2$
+varying between 
+    0.004
+and
+    0.36
+depending on $I$ and the tree size. Finally, the accuracy of the classifier for
+$N$ varied widely, from 
+    0.08
+to
+    0.82.
 Figure~\ref{fig:crossv} shows the results for $\alpha$; similar figures for the
@@
 
+\subsection*{ABC simulations}
+
+Figure~\ref{fig:abcpt} shows point estimates for the simulations with $m = 2$.
+The results for the other values of $m$ are similar and can be found in the
+supplemental materials.
+
+The marginal approximated posterior distributions for one simulated dataset are
+shown in Figure~\ref{fig:abcex}.
+
+\begin{figure}
+    \includegraphics{abc-point-est}
+    \caption{Point estimates of \gls{BA} model parameters obtained by running
+    kernel-ABC on simulated phylogenies. Dotted lines indicate the true values,
+    and limits of the $y$-axes are the regions of uniform prior density.}
+    \label{fig:abcpt}
+\end{figure}
+
+\begin{figure}
+    \includegraphics{abc-example}
+    \caption{Marginal estimated posterior distributions obtained with ABC on a
+    single simulated dataset. Dotted lines and shaded polygon indicate true
+    values.}
+    \label{fig:abcex}
+\end{figure}
+
+\subsection*{Real data}
+
 \section*{Discussion}
+
+% kernel classifiers
 
@@
 transmission tree.
+
+% abc simulations
+
+The total number of nodes $N$ was almost always significantly over-estimated.
+This suggests that $N$ has very little effect on tree shape, and the estimates
+were influenced primarily by the prior. Since the prior on $N$ and $I$ is
+jointly uniform on a triangular region, there is more prior mass on high $N$
+values. In retrospect, this is reasonable because adding more nodes to a
+\gls{BA} network does not change the edge density or the network's overall
+shape. This can illustrated by imagining that we grow the network after the
+epidemic simulation has already been completed. It is possible (though perhaps
+unlikely) that none of the new nodes attains a connection to any infected node.
+Thus, running the simulation again on the new, larger network could produce the
+exact same transmission tree as before. This is an extreme example, but it
+serves to demonstrate the point that $N$ may be impossible to determine if the
+epidemic does not come close to saturating the network.
+
+% real data
+
+We estimated the parameters of the \gls{BA} model for several published
+datasets (see~\ref{tab:data}). \textcite{cuevas2009hiv} performed a
+phylogenetic analysis of 261 patients of mixed risk groups in the Basque
+Country, Spain, who were newly diagnosed between 2004 and 2007. Although their
+data did not comprise a single cluster, they did find that almost half (47\%)
+of the samples grouped in transmission clusters. \textcite{wang2015targeting}
+reported on a cohort of acutely infected \gls{MSM} in Beijing, China. Through
+phylogenetic analysis, they found that they had ``deeply sampled'' the local
+\gls{MSM} network, suggesting that many of the samples could be
+epidemiologically related. \textcite{novitsky2013phylogenetic} investigated a
+local HIV epidemic in Mochudi, Botswana. They later extended their
+analysis~\autocite{novitsky2014impact} to include an estimated 70\% of HIV
+infected persons in the town.
 
@@
 
+\newpage
+
 \section*{Supplemental Materials}
 
-\begin{figure}
-  \includegraphics{leventhal2012fig1}
+\begin{figure}[ht]
+  \includegraphics[scale=0.8]{leventhal2012fig1}
   \caption{Reproduction of Figure 1A from \textcite{leventhal2012inferring}.}
@@
     \textcite{del2012adaptive} and \textcite{sisson2007sequential} to test
-    \gls{SMC}.}
+    \gls{SMC}. Solid black line indicates true distribution. Grey shaded area
+    shows SMC approximation obtained with our implementation.}
   \label{fig:smctest}

### 2016-03-01 doc/thesis/results.tex
Added figures and started poster

@@
-\subsection{Separable parameters}
+\subsection{Kernel-SVM classifiers}
 
 Our first simulation-based experiment was designed to determine which network
-model parameters were potentially estimable by kernel-ABC, and to find the best
-set of kernel meta-parameters to carry through to future experiments. To do
-this, we simulated trees under three distinct values of each parameter of the
-\gls{BA} model, and evaluated the accuracy of a kernel-\gls{SVM} classifier for
-the parameter.
+model parameters were potentially estimable by kernel-\gls{ABC}, and to find
+the best set of kernel meta-parameters to carry through to future experiments.
+We simulated trees under three distinct values of each parameter of the
+\gls{BA} model, and calculated the cross-validation accuracy of a
+kernel-\gls{SVM} classifier. Many of the figures in this section pertain to the
+attachment power parameter $\alpha$ only, but equivalent figures for the other
+three parameters can be found in the supplemental materials.
 
-In Figure~\ref{fig:patrees}, we show that trees simulated under different
-values of \gls{alpha} are visibly quite distinctive. In particular, higher
-values of \gls{alpha} result in networks with a small number of highly
-connected nodes (see Figure~\ref{fig:pabounds}) which, once infected, are
-likely to transmit to many other nodes. This results in a more unbalanced,
-ladder-like structure in the phylogeny. Kernel-\gls{PCA} projections show that
-all three \gls{alpha} values are well separated from each other under several
-different sampling scenarios (Figure~\ref{fig:pakpca}). With smaller trees,
-it becomes harder to visually disinguish \gls{alpha} = 0.5 from \gls{alpha} =
-1.0 using the first two principal components.
+Trees simulated under different values of \gls{alpha} are visibly quite
+distinctive (Figure~\ref{fig:patrees}). In particular, higher values of
+\gls{alpha} result in networks with a small number of highly connected nodes
+(see Figure~\ref{fig:pabounds}) which, once infected, are likely to transmit to
+many other nodes. This results in a more unbalanced, ladder-like structure in
+the phylogeny.
+
+Kernel-\gls{PCA} projections show that all three \gls{alpha} values are well
+separated from each other under several different sampling scenarios
+(Figure~\ref{fig:pakpca}). For smaller trees, though, it becomes harder to
+visually disinguish \gls{alpha} = 0.5 from \gls{alpha} = 1.0 using the first
+two principal components.
 
@@
 unsurprising, given the clear differences in tree balance observed under
-different \gls{alpha} values. The \gls{nltt}
-statistic~\autocite{janzen2015approximate} did not improve the classifier, but
-rather slightly reduced the $R^2$ in most cases.
+different \gls{alpha} values.
 
@@
 
-We also considered the possibility of inferring the number of infected nodes,
-or \defn{prevalence}, under this model. All parameters except \gls{I} were
-fixed at the following values: \gls{N} = 5000, \gls{alpha} = 1.0, and \gls{m} =
-2. As shown in Figure~\ref{fig:prevtrees}, the prevalence had no obvious effect
-on the tree shape. However, a kernel-\gls{SVM} classifier was able to
-distinguish the number of infected nodes with high accuracy ($R^2 > 0.9$;
-Figure~\ref{fig:prevcrossv}). Moreover, the use of the \gls{nltt} statistic
-improved classification accuracy by a small amount, in contrast to the results
-for \gls{alpha}. In contrast, a Sackin's index-based classifier displayed
-extremely poor performance ($R^2$ < 0.1, not shown).
-
-It is important to note that, if we cut off the epidemic simulation when 500
-nodes are infected, the resulting tree will be shorter (in calendar time) than
-if we continue until 2000 nodes are infected. However, this information is not
-used when building the classifier, since the branch lengths in each tree are
-scaled by their mean. Therefore, the high performance of the classifier is due
-to structural differences captured by the tree kernel, rather than the trees
-simply having different heights.
+Figure~\ref{fig:rsquared} shows the cross-validation accuracy of classifiers
+for all four parameters of the \gls{BA} model. The classifier for \gls{I} also
+exhibited high performance, although not quite as high as that for \gls{alpha}.
+The average $R^2$ values were 0.70 for 100-tip trees, or 0.93 for 500-tip
+trees. On the other hand, $m$ and $N$ were much harder to classsify. For $m$,
+all $R^2$ values were below 0.5, falling to virtually zero in four scenarios.
+The $R^2$ values for $N$ were highly variable, ranging from 0.08 to 0.82
+depending on $I$ and the number of tips in the tree. In almost all cases, both
+the nLTT and Sackin's index classifiers were worse than the tree kernel, except
+for classifying $N$ with $I = 2000$ and $100$ tips, where the nLTT gave a small
+advantage (see the supplements). The greatest disadvantage for the Sackin's
+index-based classifier was on the $I$ parameter, where it had an $R^2$ of less
+than 0.1, despite the relatively high accuracy of the tree kernel (see the
+supplements).
 
@@
   \centering
-  \label{fig:prevtrees}
-  \includegraphics{kernel_nsimnode_trees}
-  \caption[Similarly shaped simulated under three values of \gls{I}]{
-    Epidemics were simulated on networks of size 5000 until \gls{I} = 500,
-    1000, or 2000 nodes were infected. When scaled to the same height, there is
-    no immediately visible distinction in shape between trees simulated under
-    different \gls{I} values.
-  }
+  \includegraphics{kernel_rsquared}
+  \caption[Cross-validation accuracy of kSVM classifiers for \gls{BA}
+  model parameters.]{Cross-validation $R^2$ values of kernel-SVM classifier for
+    \gls{BA} model parameters. Points and ranges are median and 95\% quantiles
+    over 1000 2-fold cross-validations.}
+  \label{fig:rsquared}
 \end{figure}
 
-\begin{figure}[ht]
-  \centering
-  \label{fig:prevcrossv}
-  \includegraphics{kernel_nsimnode_crossv}
-  \caption[Cross-validation performance of kernel-\gls{SVM} classifier for
-           prevalence]{
-    For \gls{BA} networks of size \gls{N} = 5000, a kernel-\gls{SVM} classifier
-    is very accurate at predicting \gls{I}. Epidemics were simulated until
-    \gls{I} = 500, 1000, or 2000 nodes were infected, and either 100 or 500
-    nodes were sampled for inclusion in a transmission tree. The parameters of
-    the \gls{BA} network were \gls{alpha} = 1.0 and \gls{m} = 2. A Sackin's
-    index-based classifier performed very poorly on these data ($R^2 = $ TODO
-    for 100-tip trees and blah for 500-tip trees, not shown).
-  }
-\end{figure}
+% needs to go in methods
+
+%It is important to note that, if we cut off the epidemic simulation when 500
+%nodes are infected, the resulting tree will be shorter (in calendar time) than
+%if we continue until 2000 nodes are infected. However, this information is not
+%used when building the classifier, since the branch lengths in each tree are
+%scaled by their mean. Therefore, the high performance of the classifier is due
+%to structural differences captured by the tree kernel, rather than the trees
+%simply having different heights.
 

### 2016-03-01 doc/committee-meeting-2/figures/bc-vs-cn.R

@@ -1,0 +1,43 @@
+#!/usr/bin/env Rscript
+
+suppressPackageStartupMessages(library(netabc))
+
+m <- collect.metadata("../../../simulations/abc-pa-bctree/abc/*.tsv")
+bc.file <- rownames(subset(m, m == 2 & rbf_variance == 5))
+cn.file <- Sys.glob("../../../simulations/abc-pa-cn/abc/*.tsv")
+stopifnot(length(bc.file) == 1)
+stopifnot(length(cn.file) == 1)
+
+get.posterior <- function (file) {
+    d <- fread(paste("head -n -1", file), select=c("iter", "weight", "attach_power"))
+    d <- subset(d, iter == d[,max(iter)])
+    idx <- sample(1:nrow(d), prob=d[,weight], replace=TRUE)
+    d[idx,]
+}
+
+bc.data <- get.posterior(bc.file)
+cn.data <- get.posterior(cn.file)
+
+bc.data <- bc.data[,"attach_power",with=FALSE][,data := "BC"]
+cn.data <- cn.data[,"attach_power",with=FALSE][,data := "China"]
+d <- rbind(bc.data, cn.data)
+
+bc.dens <- bc.data[,density(attach_power)]
+cn.dens <- cn.data[,density(attach_power)]
+
+bc.max <- bc.dens$x[which.max(bc.dens$y)]
+cn.max <- cn.dens$x[which.max(cn.dens$y)]
+
+lines <- data.frame(attach_power=c(bc.max, cn.max), data=c("BC", "China"),
+                    density=c(max(bc.dens$y), max(cn.dens$y)))
+lines$label <- round(lines$attach_power, 2)
+
+pdf("bc_vs_cn.pdf", family="Gillius ADF")
+ggplot(d, aes(x=attach_power)) + geom_density() + facet_grid(data~.) +
+    geom_vline(data=lines, lty=2, aes(xintercept=attach_power)) +
+    geom_text(data=lines, aes(x=attach_power - 0.05, y=density + 0.05, label=label), 
+              hjust=1, vjust=0) +
+    labs(x="preferential attachment power", y="posterior density") +
+    theme_bw() +
+    theme(text = element_text(size = 20))
+dev.off()

### 2016-03-01 doc/committee-meeting-2/figures/nettree.R

@@ -1,0 +1,21 @@
+#!/usr/bin/env Rscript
+
+suppressPackageStartupMessages(library(netabc))
+
+n <- 6
+g <- SIR.net(sample_pa(n, m=2, power=1, directed=FALSE))
+tmp.graph <- tempfile()
+tmp.tree <- tempfile()
+write.graph(g, tmp.graph, "gml")
+system2("nettree", args=tmp.graph, stdout=tmp.tree)
+t <- read.tree(tmp.tree)
+t$tip.label <- as.character(as.integer(t$tip.label)+1)
+t$node.label <- as.character(as.integer(t$node.label)+1)
+
+pdf("nettree.pdf", height=4)
+par(mar=c(0, 1, 0, 1) + 0.1, mfrow=c(1, 2))
+plot(g, vertex.color='#eff3ff', vertex.label.cex=1.5, vertex.size=20,
+     vertex.label.color="black", edge.width=2)
+plot(t, show.node.label=TRUE, edge.color="gray50", direction="down", srt=90, 
+     label.offset=0.05, cex=1.5, edge.width=2)
+dev.off()

### 2016-03-01 doc/committee-meeting-2/figures/phylo.R

@@ -1,0 +1,26 @@
+#!/usr/bin/env Rscript
+
+library(ape)
+library(igraph)
+
+n <- 15
+
+pdf("phylo.pdf", width=0.8, height=0.8)
+plot.phylo(rcoal(n), show.tip.label=FALSE, direction="down", edge.width=2,
+           no.margin=TRUE)
+dev.off()
+
+col <- c('#bdd7e7','#6baed6','#2171b5')
+for (i in 1:3) {
+    g <- sample_pa(n, m=2, power=(i-1)*0.8, directed=FALSE)
+    pdf(paste0("net", i, ".pdf"), width=0.8, height=0.8)
+    par(mar=c(0, 0, 0, 0) + 0.1)
+    plot(g, vertex.label=NA, vertex.size=8, vertex.color=col[i],
+         edge.width=2, vertex.frame.color=col[i], edge.color=col[i])
+    dev.off()
+    pdf(paste0("tree", i, ".pdf"), width=0.8, height=0.8)
+    par(mar=c(0, 0, 0, 0) + 0.1)
+    plot.phylo(rcoal(n), show.tip.label=FALSE, direction="down", edge.width=2,
+               edge.color=col[i], no.margin=TRUE)
+    dev.off()
+}

### 2016-03-01 doc/figures/abc-smc.R

@@ -1,0 +1,47 @@
+#!/usr/bin/env Rscript
+
+suppressPackageStartupMessages(library(netabc))
+
+post <- rnorm(500)
+abc <- list(runif(500, -3, 3),
+            c(runif(250, -3, 3), rnorm(250)),
+            c(runif(100, -3, 3), rnorm(400)))
+dens <- lapply(abc, density)
+gradient <- colorRampPalette(c("steelblue1", "steelblue1"))(length(dens[[1]]$x)*1.2)
+
+pdf("abc-smc.pdf", family="Gillius ADF", height=4, width=6)
+par(mfrow=c(2, 3), mgp=c(0, 0, 0), mar=c(1, 1, 2, 1) + 0.1, xpd=TRUE)
+for (i in 1:3) {
+    plot(density(post), main=NA, ylab=bquote(f[.(i)](theta*"|"*D)), 
+         xlab=expression(theta), tck=0, xaxt="n", yaxt="n", zero.line=FALSE,
+         xlim=c(-4, 4) + 0.1)
+    points(dens[[i]]$x, dens[[i]]$y, col=gradient, pch=16)
+}
+
+true.tree <- tiny.tree(4, ultra=TRUE, color="black")
+tree.size <- 0.15
+offset <- tree.size/2
+epsilon <- c(0.45, 0.35, 0.25)
+
+for (i in 1:3) {
+    dist <- sample(dens[[i]]$x, 10, prob=dens[[i]]$y)
+    r <- (dist - min(dens[[i]]$x)) / (max(dens[[i]]$x) - min(dens[[i]]$x))
+    r <- abs(r - 0.5) * 1.2
+    r[r < 0.2] <- rnorm(sum(r < 0.2), mean=0.2, sd=0.05)
+    theta <- (1:length(r)) * 2 * pi / length(r)
+    coords <- polar2rect(r, theta) + 0.5
+    colors <- gradient[as.integer(r * length(gradient))]
+    trees <- lapply(colors, function (x) tiny.tree(ntip=4, color=x, ultra=TRUE))
+    plot.new()
+    for (j in 1:length(trees)) {
+        rasterImage(trees[[j]], coords[j,1]-offset, coords[j,2]-offset, 
+                    coords[j,1]+offset, coords[j,2]+offset)
+        if (r[j] > epsilon[i]) {
+            points(coords[j,1], coords[j,2], pch=4, col="red", cex=4)
+        }
+    }
[…]

### 2016-03-01 doc/paper/paper.tex
Oh dear this is going to cause a merge conflict

--- 2016-02-23 doc/paper/paper.Rnw
@@
-\documentclass[12pt]{report}
+\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 \usepackage[osf]{garamondx}
@@
 
+\subsection*{ABC simulations}
+
+Point estimates for each parameter are shown in Figure~\ref{fig:abcpt}. Of the
+four parameters, $\alpha$ was the most accurately estimated, with a median
+[IQR] absolute error of 
+    0.11 
+    [0.04-0.18].
+The accuracy of the estimates was not significantly different between values of
+$m$ or $I$ (both one-way ANOVA,
+    $p$ = 0.18
+and 
+    0.47),
+although the errors when the true value of $\alpha$ was zero were
+significantly greater than the other values 
+    (Wilcoxon rank-sum test, $p$ = \ensuremath{1.02\times 10^{-3}}).
+The error in the estimated value of $I$ was
+    337.22 
+    [105.39-592.01].
+Errors were significantly higher for $\alpha \geq 1$
+    (Wilcoxon rank-sum test, $p$ = \ensuremath{1.03\times 10^{-3}})
+and for $I$ = 2000
+    ($p$ = \ensuremath{2.26\times 10^{-7}}),
+but not for any values of $m$
+    (one-way ANOVA, $p$ = 0.34).
+The $m$ parameter was estimated correctly in
+    33 \%
+of simulations, with an error of one in
+    43 \%
+and of two or more in 
+    23 \%
+(the only possible $m$ values were 2, 3, 4, or 5). The true values of
+$m$ and $I$ did not significantly affect the error
+    (one-way ANOVA, $p$ = 0.44 and
+                          0.68),
+but the accuracy was significantly lower for integral than non-integral
+values of $\alpha$
+    (Wilcoxon rank-sum test, $p$ = \ensuremath{8.4\times 10^{-4}}).
+Finally, the total number of nodes $N$ was consistently over-estimated by about a factor of two (error
+    \ensuremath{6.74\times 10^{3}} 
+    [\ensuremath{5.28\times 10^{3}}-\ensuremath{8.29\times 10^{3}}]).
+No other parameters influenced the accuracy of the $N$ estimates 
+    (one-way ANOVA, $p \geq$ NA).
+
+\begin{figure}
+    \includegraphics{missing}
+    \caption{Point estimates for \gls{BA} model parameters obtained with
+             kernel-ABC.}
+    \label{fig:abcpt}
+\end{figure}
+
 \section*{Discussion}

### 2016-03-01 doc/thesis/results.Rnw
Oh dear this is going to cause a merge conflict

<<setup, include=FALSE>>=
    library(netabc)
@

\subsection{Separable parameters}

Our first simulation-based experiment was designed to determine which network
model parameters were potentially estimable by kernel-ABC, and to find the best
set of kernel meta-parameters to carry through to future experiments. For each
parameter of the \gls{BA} model, we simulated trees under three distinct
values, and computed the cross-validation accuracy of a kernel-\gls{SVM}
classifier for the parameter.

In Figure~\ref{fig:patrees}, we show that trees simulated under different
values of \gls{alpha} are visibly quite distinctive. In particular, higher
values of \gls{alpha} result in networks with a small number of highly
connected nodes (see Figure~\ref{fig:pabounds}) which, once infected, are
likely to transmit to many other nodes. This results in a more unbalanced,
ladder-like structure in the phylogeny. Kernel-\gls{PCA} projections show that
all three \gls{alpha} values are well separated from each other under several
different sampling scenarios (Figure~\ref{fig:pakpca}). With smaller trees,
it becomes harder to visually disinguish \gls{alpha} = 0.5 from \gls{alpha} =
1.0 using the first two principal components.

With suitably chosen meta-parameters, the accuracy of the kernel-\gls{SVM}
classifier for \gls{alpha} was very high under a variety of prevalence and
sampling scenarios (Figure~\ref{fig:pacrossv}). Accuracy was highest, with
$R^2$ values above 0.95, for the largest trees and complete sampling (bottom
center panel). However, even for trees of size 100 sampled from an epidemic on
2000 nodes, the $R^2$ was above 0.8 (top right panel). In all cases, the
accuracy of a Sackin's index-based classifier was also quite high, at about
0.75. An \gls{SVM} classifier using only the \gls{nltt}
statistic~\autocite{janzen2015approximate} was slightly worse than Sackin's
index, with an $R^2$ of X.

The fact that Sackin's index is informative of \gls{alpha} was
unsurprising, given the clear differences in tree balance observed under
different \gls{alpha} values.

\begin{figure}[ht]
  \centering
  \label{fig:patrees}
  \includegraphics{kernel_pa_trees}
  \caption[Visibly distinctive trees simulated under three values of \gls{alpha}]{
    Epidemics of 1000 infected were simulated on \gls{BA} networks of 5000
    nodes, with \gls{alpha} equal to 0.5, 1.0, or 1.5. Transmission trees were
    created by sampling 500 infected nodes. Higher \gls{alpha} values produce
    networks with a small number of highly-connected nodes, which results in a
    highly unbalanced, ladder-like tree structure.
  }
\end{figure}

\begin{figure}[ht]
  \centering
  \label{fig:pacrossv}
  \includegraphics{kernel_pa_crossv}
  \caption[Cross-validation performance of kernel support vector machine
  classifier for preferential attachment power]{
    Cross-validation performance of kernel support vector machine classifier
    for preferential attachment power.
  }
\end{figure}

\begin{figure}[ht]
  \centering
  \label{fig:pakpca}
  \includegraphics{kernel_pa_kpca}
  \caption[Projection of kernel matrix for different attachment power values
  onto its first two principal components]{
    Projection of the kernel matrix for different preferential attachment power
    values onto its first two principal components, for eight simulation
    scenarios. Each point corresponds to a simulated transmission tree, and is
    coloured by preferential attachment power. Facets are number of infected
    nodes (horizontal), and number of sampled tips (vertical). The parameters
    to the tree kernel were $\lambda = 0.3$ and $\sigma = 4$, and the nLTT was
    not used. Qualitatively, trees with a larger number of tips are easier to
    separate in kernel space, regardless of what sampling proportion they
    represent. In all cases, the highest attachment power can be separated from
    the other two, but the two lowest values become hand to distinguish with in
    the 100-tip trees.
  }
\end{figure}

We also considered the possibility of inferring the number of infected nodes,
or \defn{prevalence}, under this model. All parameters except \gls{I} were
fixed at the following values: \gls{N} = 5000, \gls{alpha} = 1.0, and \gls{m} =
2. As shown in Figure~\ref{fig:prevtrees}, the prevalence had no obvious effect
on the tree shape. However, a kernel-\gls{SVM} classifier was able to
distinguish the number of infected nodes with high accuracy ($R^2 > 0.9$;
Figure~\ref{fig:prevcrossv}). Moreover, the use of the \gls{nltt} statistic
improved classification accuracy by a small amount, in contrast to the results
for \gls{alpha}. In contrast, a Sackin's index-based classifier displayed
extremely poor performance ($R^2$ < 0.1, not shown).

It is important to note that, if we cut off the epidemic simulation when 500
nodes are infected, the resulting tree will be shorter (in calendar time) than
if we continue until 2000 nodes are infected. However, this information is not
used when building the classifier, since the branch lengths in each tree are
scaled by their mean. Therefore, the high performance of the classifier is due
to structural differences captured by the tree kernel, rather than the trees
simply having different heights.

\begin{figure}[ht]
  \centering
  \label{fig:prevtrees}
  \includegraphics{kernel_nsimnode_trees}
  \caption[Similarly shaped simulated under three values of \gls{I}]{
    Epidemics were simulated on networks of size 5000 until \gls{I} = 500,
    1000, or 2000 nodes were infected. When scaled to the same height, there is
    no immediately visible distinction in shape between trees simulated under
    different \gls{I} values.
  }
\end{figure}

\begin{figure}[ht]
  \centering
  \label{fig:prevcrossv}
  \includegraphics{kernel_nsimnode_crossv}
  \caption[Cross-validation performance of kernel-\gls{SVM} classifier for
           prevalence]{
    For \gls{BA} networks of size \gls{N} = 5000, a kernel-\gls{SVM} classifier
    is very accurate at predicting \gls{I}. Epidemics were simulated until
    \gls{I} = 500, 1000, or 2000 nodes were infected, and either 100 or 500
    nodes were sampled for inclusion in a transmission tree. The parameters of
    the \gls{BA} network were \gls{alpha} = 1.0 and \gls{m} = 2. A Sackin's
    index-based classifier performed very poorly on these data ($R^2 = $ TODO
    for 100-tip trees and blah for 500-tip trees, not shown).
  }
\end{figure}

\subsection{Accuracy of marginal estimates}

We used grid search to obtain \defn{marginal} estimates for each network
parameter while holding all other parameters fixed. We observed that kernel
scores were highest at the values of \gls{alpha} on the grid closest to the
true values, as shown in Figure~\ref{fig:gridkernel}. However, there was a much
stronger spike in kernel scores near the true value for \gls{alpha} = 1.0 and
1.25. This is recapitulated when we look at the accuracy of point estimates
obtained by taking the grid value with the highest median kernel score. As
shown in Figure~\ref{fig:gridest}, while the estimates are generally close to
the true value, they are much closer for \gls{alpha} = 1.25 than for the other
values.

\begin{figure}[ht]
  \centering
  \includegraphics{gridsearch_pa_kernel}
  \caption[Grid search kernel scores]{
    Grid search kernel scores for testing trees simulated under various
    \gls{alpha} values. All epidemics had \gls{I} = 1000 infected nodes, on
    \gls{BA} networks of size \gls{N} = 5000 with \gls{m} fixed at 2. Colours
    indicate the number of sampled tips.
  }
  \label{fig:gridkernel}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics{gridsearch_pa_estimates}
  \caption[Marginal estimates of \gls{alpha} obtained with grid search]{
    Marginal estimates of \gls{alpha} obtained with grid search. Training trees
    were simulated on a narrowly spaced grid of \gls{alpha} values, and
    compared to testing trees using the tree kernel. The \gls{alpha} value in
    the grid with the highest median kernel score was taken as the point
    estimate for the testing tree. These point estimates are shown as black
    dots. The dashed line is the identity.
  }
  \label{fig:gridest}
\end{figure}

\subsection{Accuracy of estimates with full ABC}

<<point_est, include=FALSE>>=
    f <- Sys.glob("../../simulations/abc-pa-free-m/point-estimate/*")
    d <- fread(f)
    d[,m := floor(m)]
    d[,alpha_error := abs(true_alpha - alpha)]
    d[,N_error := abs(true_N - N)]
    d[,I_error := abs(true_I - I)]
    d[,m_error := abs(true_m - m)]

    alpha.av <- anova(lm(alpha_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
    m.av <- anova(lm(m_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
    N.av <- anova(lm(N_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
    I.av <- anova(lm(I_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))

    m.tbl <- d[,prop.table(table(m_error))]
    zero.alpha.test <- wilcox.test(d[true_alpha == 0, alpha_error], 
                                   d[true_alpha != 0, alpha_error])
    alpha.I.test <- wilcox.test(d[true_alpha < 1, I_error],
                                d[true_alpha >= 1, I_error])
    I.I.test <- wilcox.test(d[true_I == 1000, I_error],
                            d[true_I == 2000, I_error])
    alpha.m.test <- wilcox.test(d[true_alpha == 0 | true_alpha == 1, m_error],
                                d[true_alpha == 0.5 | true_alpha == 1.5, m_error])

    options(scipen=-1, digits=2)
@

We used kernel-\gls{ABC} to estimate the parameters of the \gls{BA} model 
on simulated trees where the true parameter values were known. Point estimates
for each parameter are shown in Figure~\ref{fig:abcpt}. Of the four parameters,
$\alpha$ was the most accurately estimated, with a median [IQR] absolute error
of 
    \Sexpr{d[,median(alpha_error)]} 
    [\Sexpr{d[,quantile(alpha_error, 0.25)]}-\Sexpr{d[,quantile(alpha_error, 0.75)]}].
The accuracy of the estimates was not significantly different between values of
$m$ or $I$ (both one-way ANOVA,
    $p$ = \Sexpr{alpha.av["factor(true_m)", "Pr(>F)"]}
and 
    \Sexpr{alpha.av["factor(true_I)", "Pr(>F)"]}),
although the errors when the true value of $\alpha$ was zero were
significantly greater than the other values 
    (Wilcoxon rank-sum test, $p$ = \Sexpr{zero.alpha.test$p.value}).
The error in the estimated value of $I$ was
    \Sexpr{d[,median(I_error)]} 
    [\Sexpr{d[,quantile(I_error, 0.25)]}-\Sexpr{d[,quantile(I_error, 0.75)]}].
Errors were significantly higher for $\alpha \geq 1$
    (Wilcoxon rank-sum test, $p$ = \Sexpr{alpha.I.test$p.value})
and for $I$ = 2000
    ($p$ = \Sexpr{I.I.test$p.value}),
but not for any values of $m$
    (one-way ANOVA, $p$ = \Sexpr{I.av["factor(true_m)", "Pr(>F)"]}).
The $m$ parameter was estimated correctly in
    \Sexpr{as.integer(m.tbl[1] * 100)} \%
of simulations, with an error of one in
    \Sexpr{as.integer(m.tbl[2] * 100)} \%
and of two or more in 
    \Sexpr{as.integer(sum(m.tbl[3:length(m.tbl)]) * 100)} \%
(the only possible $m$ values were 2, 3, 4, or 5). The true values of
$m$ and $I$ did not significantly affect the error
    (one-way ANOVA, $p$ = \Sexpr{m.av["factor(true_m)", "Pr(>F)"]} and
                          \Sexpr{m.av["factor(true_I)", "Pr(>F)"]}),
but the accuracy was significantly lower for integral than non-integral
values of $\alpha$
    (Wilcoxon rank-sum test, $p$ = \Sexpr{alpha.m.test$p.value}).
Finally, the total number of nodes $N$ was consistently over-estimated by about
a factor of two
    (error \Sexpr{d[,median(N_error)]} 
    [\Sexpr{d[,quantile(N_error, 0.25)]}-\Sexpr{d[,quantile(N_error, 0.75)]}]).
No other parameters influenced the accuracy of the $N$ estimates 
    (one-way ANOVA, $p \geq$ \Sexpr{min(N.av[,"Pr(>F)"])}).

Figure~\ref{fig:abcex} shows the \gls{ABC} approximation to the
posterior distribution on the \gls{BA} parameters for one simulation
(equivalent plots for all the simulations can be found in the supplemental
materials). \gls{HPD} intervals around $\alpha$ and $I$ were narrow relative
to the region of nonzero prior density, whereas the intervals for $m$ and $N$
were widely dispersed. Table~\ref{tab:abchpd} shows point estimates and 95\%
\gls{HPD} intervals averaged over all simulations.

### 2016-03-01 doc/committee-meeting-2/post.tex

\documentclass{article}
\usepackage[T1]{fontenc}
\usepackage{garamondx}
\usepackage[garamondx,cmbraces]{newtxmath}
\usepackage{fullpage}
\usepackage[backend=biber, sorting=none]{biblatex}
\usepackage{setspace}

\newcommand{\sw}[1]{\textit{#1}}
\newcommand{\etal}{\textit{et al.}}

\title{Committee meeting follow-up}
\author{Rosemary McCloskey}
\addbibresource{papers.bib}

\begin{document}

\maketitle

\onehalfspacing

Thanks everyone for your useful feedback at my committee meeting. I've come
away with a few new analyses to do and some tweaks to make, which I'll
summarize below. Please let me know if I have missed anything.

\subsection*{Bimodal $\alpha$}

Evaluate the effect of a ``bimodal'' distribution on $\alpha$, by simulating a
network where half the nodes use one value, and half with another. I tried this
with the two values being equal to 0.5 and 1.5, and the resulting posterior
distribution was centred at 1. This brings up an important caveat: what we are
able to estimate is the \emph{average} preferential attachment power, but not
the variance.

\includegraphics[scale=0.15]{mixed-alpha}

\subsection*{Realistic values of $\alpha$, as defined by $\gamma$}

The degree distributions of preferential attachment networks follow a power
law, meaning that the number of nodes of degree $k$ is proportional to
$k^\gamma$ for some $\gamma$ which is determined by $\alpha$. I looked at how
$\gamma$ varies with $\alpha$. Judging by this plot, we should upper bound
$\alpha$ at around 1.3 to 1.4. This is a nice result, since the posterior
density for values above 1.3 seems to always be zero (see eg. the above
$\alpha$ density), even though we put a uniform prior between 0 and 2.

\includegraphics[scale=0.15]{gamma-alpha}

Relatedly, I will plot the estimated vs. actual values of $\gamma$ in addition
to $\alpha$.

\subsection*{Lineages-through-time as a standalone classifier}

I had previously evaluated combining nLTT with the tree kernel and found that
it brought down the accuracy of the classifier. I also looked at nLTT by itself,
and it's even worse than Sackin's index (see below).

\includegraphics[scale=0.15]{crossv}

\subsection*{Expanded priors, free $m$}

We discussed expanding the prior regions so that the ``true'' values were less
in the middle, and also of allowing the $m$ parameter to be free in all
analyses. These updates are in progress. A bit of a preview is contained in the
first figure in this document, which shows that most of the posterior mass on
$I$ (the number of infected nodes) is concentrated around the true value on the
low end of the prior.

\subsection*{Biased sampling}

It was brought up that the assumption of random sampling from the population
might bias the results. I will evaluate the effect of non-random sampling based
on whether or not any of one's peers have been sampled. I have discussed an
implementation strategy with Art and will incorporate it into the program soon.

\subsection*{New real data}

For additional real-world use, as well as to circumvent possible privacy issues
with locally produced data, I am looking for more real-world data sets to which
I can apply the method. The best would be datasets which originate from known
phylogenetic clusters or epidemiologically linked patients. I have identified
the following one which describes their sample as a ``deeply sampled
subnetwork''.

\begin{quote}
Wang, Xicheng, et al. ``Targeting HIV Prevention Based on Molecular
Epidemiology Among Deeply Sampled Subnetworks of Men Who Have Sex With Men.''
Clinical Infectious Diseases (2015): civ526.
\end{quote}

I'm still on the lookout for others.

\subsection*{Literature support for the approach}

I need to put more effort into justifying why I'm doing this and why it's
important, such as the search for superspreader dynamics. Also, I would like to
emphasize a bit more that this is a general approach for fitting contact
network models to phylogenies, and is not restricted to this $\alpha$
parameter. I just happen to be using it because it is a simple and well-known
model, but is more realistic than the random graph model. Making this clearer
would (I hope) expand the potential utility of the method.

\end{document}

### 2016-03-01 doc/figures/pipeline.tex

\documentclass{article}
\usepackage[default]{gillius}
\usepackage{tikz}

\usetikzlibrary{positioning}
\usetikzlibrary{arrows}

\begin{document}
\pagestyle{empty}

\begin{tikzpicture}
  [every node/.style={rectangle, draw, text width=4cm, align=center, rounded corners},
   every path/.style={->, >=stealth, thick}]
  \node (dl) {download sequences \\ from GenBank};
  \node (align) [below=of dl] {multiple sequence alignment with MUSCLE};
  \node (clip) [below=of align, dashed] {remove hypervariable regions};
  \node (tree) [right=of dl] {estimate unrooted phylogeny with FastTree2};
  \node (cluster) [below=of tree, dashed] {select cluster by patristic distance};
  \node (root) [below=of cluster] {root and scale phylogeny with root-to-tip regression};

  \draw (dl) -- (align);
  \draw (align) -- (clip);
  \draw (clip.east) -- ++ (0.5cm, 0) |- (tree);
  \draw (tree) -- (cluster);
  \draw (cluster) -- (root);
\end{tikzpicture}

\end{document}

### 2016-03-01 doc/paper/paper.Rnw

@@
-\documentclass[12pt]{report}
+\documentclass[12pt]{article}
 \usepackage[osf]{garamondx}
@@
 \newacronym{MSM}{MSM}{men who have sex with men}
+\newacronym{HPD}{HPD}{highest posterior density}
 
@@
 
+<<setup, include=FALSE>>=
+    library(netabc)
+    library(xtable)
+@
+
 \maketitle
@@
 schneeberger2004scale} have found that sexual contact networks are best
-described by a preferential attachment model \autocite[but not all,
-see][]{jones2003assessment}. This result has been coroborated by more recent
-investigations using phylogenetic methods~\autocite{brown2011transmission}.
-Under these models, nodes with a high number of contacts attract new
-connections at an elevated rate. Networks produced by preferential attachment
-have a power-law degree distribution, meaning that the number of nodes of
-degree $k$ is proportional to $k^\gamma$ for some constant $\gamma$. When $2 <
-\gamma \leq 3$, the network is referred to as ``scale-free''. The first contact
-network model incorporating preferential attachment was introduced by
-\textcite{barabasi1999emergence}, and is now referred to as the \gls{BA} model.
-Under this model, networks are formed by iteratively adding nodes with $m$ new
-edges each. These new edges are joined to existing nodes of degree $k$ with
-probability proportional to $k^\alpha$, so that nodes of high degree tend to
-attract more connections (in the original paper, only $\alpha = 1$ was
-investigated).
+described by a preferential attachment model \autocite[although there has been
+some disagreement, see][]{jones2003assessment}. Under these models, nodes with
+a high number of contacts attract new connections at an elevated rate. Networks
+produced by preferential attachment have a power-law degree distribution,
+meaning that the number of nodes of degree $k$ is proportional to $k^\gamma$
+for some constant $\gamma$. When $2 < \gamma \leq 3$, the network is referred
+to as ``scale-free''. The first contact network model incorporating
+preferential attachment was introduced by \textcite{barabasi1999emergence}, and
+is now referred to as the \gls{BA} model. Under this model, networks are formed
+by iteratively adding nodes with $m$ new edges each. These new edges are joined
+to existing nodes of degree $k$ with probability proportional to $k^\alpha$, so
+that nodes of high degree tend to attract more connections (in the original
+paper, only $\alpha = 1$ was investigated).
+
+Previous work offers precedent for the possibility of statistical inference of
+structural network parameters. \textcite{britton2002bayesian} develop a
+Bayesian approach to estimate the edge density in an Erd\H{o}s-R\'enyi
+network~\autocite{erdos1960evolution} given observed infection dates, and
+optionally recovery dates. Their approach was later extended by
+\textcite{groendyke2011bayesian} and applied to a much larger data set of 188
+individuals. \textcite{brown2011transmission} fit a preferential attachment
+model to a phylogenetically estimated transmission network relating 60\% of HIV
+infected \gls{MSM} in the United Kingdom. The transmission network is a
+subgraph of the contact network which includes only those edges which have
+already led to a new infection.
 
@@
 Here, we apply kernel-\gls{ABC} to the problem of statistical inference of
-contact network parameters from an estimated transmission tree. We then
-estimate the parameters of the \gls{BA} model on a variety of simulated and
-real data sets.
+contact network parameters from an estimated transmission tree, using the tree
+kernel developed by \textcite{poon2013mapping}. We then estimate the parameters
+of the \gls{BA} model on a variety of simulated and real data sets.
 
@@
 searched for datasets which originated from existing clusters, either
-phylogenetically or geographically defined. \textcite{wang2015targeting}
-reported on a cohort of acutely infected \gls{MSM} in Beijing, China. Through
-phylogenetic analysis, they found that they had ``deeply sampled'' the
-local \gls{MSM} network, suggesting that many of the samples could be
-epidemiologically related.
+phylogenetically or geographically defined. Characteristics of the datasets we
+investigated are given in Table~\ref{tab:data}.
 
-We downloaded all sequences associated with each paper from GenBank. Sequences
-were multiply aligned using \software{MUSCLE} version 3.8.31, and alignments
-were manually inspected with \software{Seaview} version 4.4.2. Phylogenies were
+\begin{table}
+  \centering
+  \begin{tabular}{ccccc}
+    Reference & Sequences ($n$) & Location & Risk group & Gene \\
+    \hline
+    \autocite{wang2015targeting} & 173 & Beijing, China & MSM & \textit{pol} \\
+    \autocite{cuevas2009hiv} & 287 & Basque Country, Spain & mixed & \textit{pol} \\
+    \autocite{novitsky2013phylogenetic} & \multirow{2}{*}{201^*} &
+    \multirow{2}{*}{Mochudi, Botswana} & \multirow{2}{*}{mixed} &
+    \multirow{2}{*}{\textit{env}} \\ \autocite{novitsky2014impact} \\
+    \hline
+  \end{tabular}
+  \caption{Characteristics of published datasets investigated with kernel-ABC.
+  Acronyms: MSM, men who have sex with men. $^*$ Sub-sampled from 1299.}
+  \label{tab:data}
+\end{table}
+
+We downloaded all sequences associated with each study from GenBank. Sequences
+were multiply aligned using \software{MUSCLE} version 3.8.31
+\autocite{edgar2004muscle}, and alignments were manually inspected with
+\software{Seaview} version 4.4.2 \autocite{gouy2010seaview}. Phylogenies were
 constructed from the nucleotide alignments by approximate maximum likelihood
@@
 
+<<classifiers, include=FALSE>>=
+    get.kernel.data <- function (param) {
+        md <- collect.metadata(paste0('../../simulations/kernel-', param, '/classifier/*'))
+        md <- subset(md, rbf_variance == 4 & decay_factor == 0.3)
+        if (param == 'alpha') {
+            md <- subset(md, m == 2)
+        }
+        d <- setDT(collect.data(rownames(md)))
+        if (param %in% c('alpha', 'm')) {
+            d[,list(rsquared=mean(rsquared)), by=c('nsimnode', 'ntip')]
+        } else if (param == 'I') {
+            d[,list(rsquared=mean(rsquared)), by=ntip]
+        } else {
+            d[,list(rsquared=mean(rsquared)), by=c('I', 'ntip')]
+        }
+    }
+    kernel.alpha <- get.kernel.data('alpha')
+    kernel.m <- get.kernel.data('m')
+    kernel.I <- get.kernel.data('I')
+    kernel.N <- get.kernel.data('N')
+@
+
 Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
 tested. For appropriate choices of the kernel meta-parameters, the classifier
-had an $R^2$ above 0.8 for all combinations of tree size and epidemic size. The
-$I$ parameter was also classified accurately, with an average $R^2$ of 0.93 for
-500-tip trees or 0.70 for 100-tip trees. The $m$ parameter was much harder to
-classify, with $R^2$ varying between 0.01 and 0.36 depending on $I$ and the
-tree size. Finally, the accuracy of the classifier for $N$ varied widely, from
-0.08 with 100 tips and $I = 500$, to 0.82 with 1000 tips and $I = 2000$.
+had an $R^2$ above 
+    \Sexpr{kernel.alpha[,min(rsquared %/% 0.05 * 0.05)]} 
+for all combinations of tree size and epidemic size. The $I$ parameter was also
+classified accurately, with an average $R^2$ of 
+    \Sexpr{kernel.I[ntip == 500, round(mean(rsquared), 2)]} 
+for 500-tip trees or 
+    \Sexpr{kernel.I[ntip == 100, round(mean(rsquared), 2)]} 
+for 100-tip trees. The $m$ parameter was much harder to classify, with $R^2$
+varying between 
+    \Sexpr{kernel.m[,round(min(rsquared), 3)]}
+and
+    \Sexpr{kernel.m[,round(max(rsquared), 2)]}
+depending on $I$ and the tree size. Finally, the accuracy of the classifier for
+$N$ varied widely, from 
+    \Sexpr{kernel.N[,round(min(rsquared), 2)]}
+to
+    \Sexpr{kernel.N[,round(max(rsquared), 2)]}.
 Figure~\ref{fig:crossv} shows the results for $\alpha$; similar figures for the
@@
 
+\subsection*{ABC simulations}
+
+Across replicates and other model parameters, the mean absolute difference
+between the estimated and true values of $\alpha$ was X, with a standard error
+of X.
+
+Figure~\ref{fig:abcpt} shows point estimates for the simulations with $m = 2$.
+The results for the other values of $m$ are similar and can be found in the
+supplemental materials.
+
+The marginal approximated posterior distributions for one simulated dataset are
+shown in Figure~\ref{fig:abcex}.
+
+\begin{figure}
+    \includegraphics{abc-point-est}
+    \caption{Point estimates of \gls{BA} model parameters obtained by running
+    kernel-ABC on simulated phylogenies. Dotted lines indicate the true values,
+    and limits of the $y$-axes are the regions of uniform prior density.}
+    \label{fig:abcpt}
+\end{figure}
+
+\begin{figure}
+    \includegraphics{abc-example}
+    \caption{Marginal estimated posterior distributions obtained with ABC on a
+    single simulated dataset. Dotted lines and shaded polygon indicate true
+    values.}
+    \label{fig:abcex}
+\end{figure}
+
+\subsection*{Real data}
+
 \section*{Discussion}
+
+% kernel classifiers
 

### 2016-03-01 doc/thesis/results.tex

@@
-\subsection{Kernel-SVM classifiers}
+\subsection{Separable parameters}
 
 Our first simulation-based experiment was designed to determine which network
-model parameters were potentially estimable by kernel-\gls{ABC}, and to find
-the best set of kernel meta-parameters to carry through to future experiments.
-We simulated trees under three distinct values of each parameter of the
-\gls{BA} model, and calculated the cross-validation accuracy of a
-kernel-\gls{SVM} classifier. Many of the figures in this section pertain to the
-attachment power parameter $\alpha$ only, but equivalent figures for the other
-three parameters can be found in the supplemental materials.
+model parameters were potentially estimable by kernel-ABC, and to find the best
+set of kernel meta-parameters to carry through to future experiments. For each
+parameter of the \gls{BA} model, we simulated trees under three distinct
+values, and computed the cross-validation accuracy of a kernel-\gls{SVM}
+classifier for the parameter.
 
-Trees simulated under different values of \gls{alpha} are visibly quite
-distinctive (Figure~\ref{fig:patrees}). In particular, higher values of
-\gls{alpha} result in networks with a small number of highly connected nodes
-(see Figure~\ref{fig:pabounds}) which, once infected, are likely to transmit to
-many other nodes. This results in a more unbalanced, ladder-like structure in
-the phylogeny.
-
-Kernel-\gls{PCA} projections show that all three \gls{alpha} values are well
-separated from each other under several different sampling scenarios
-(Figure~\ref{fig:pakpca}). For smaller trees, though, it becomes harder to
-visually disinguish \gls{alpha} = 0.5 from \gls{alpha} = 1.0 using the first
-two principal components.
+In Figure~\ref{fig:patrees}, we show that trees simulated under different
+values of \gls{alpha} are visibly quite distinctive. In particular, higher
+values of \gls{alpha} result in networks with a small number of highly
+connected nodes (see Figure~\ref{fig:pabounds}) which, once infected, are
+likely to transmit to many other nodes. This results in a more unbalanced,
+ladder-like structure in the phylogeny. Kernel-\gls{PCA} projections show that
+all three \gls{alpha} values are well separated from each other under several
+different sampling scenarios (Figure~\ref{fig:pakpca}). With smaller trees,
+it becomes harder to visually disinguish \gls{alpha} = 0.5 from \gls{alpha} =
+1.0 using the first two principal components.
 
@@
 accuracy of a Sackin's index-based classifier was also quite high, at about
-0.75. The fact that Sackin's index is informative of \gls{alpha} was
+0.75. An \gls{SVM} classifier using only the \gls{nltt}
+statistic~\autocite{janzen2015approximate} was slightly worse than Sackin's
+index, with an $R^2$ of X.
+
+The fact that Sackin's index is informative of \gls{alpha} was
 unsurprising, given the clear differences in tree balance observed under

### 2016-03-03 doc/paper/paper.tex
First draft of paper! Yesgit add paper.pdf -f

--- 2016-03-01 doc/paper/paper.tex
@@
 \usepackage[garamondx,cmbraces]{newtxmath}
+\usepackage{marginnote}
 \usepackage[top=2cm, bottom=2cm, outer=3cm, inner=3cm, heightrounded, 
@@
 \usepackage{multirow}
+\usepackage{cleveref}
 
@@
 \newacronym{MSM}{MSM}{men who have sex with men}
+\newacronym{HPD}{HPD}{highest posterior density}
+\newacronym{GTR}{GTR}{general time-reversible}
 \newacronym{MCMC}{MCMC}{Markov chain Monte Carlo}
-\newacronym{GTR}{GTR}{generalized time-reversible}
+\newacronym{MAP}{MAP}{maximum \textit{a posteriori}}
+\newacronym{IDU}{IDU}{injection drug user}
 
-\graphicspath{{figures/}}
+\graphicspath{{../figures/}}
+\newcommand{\tablepath}{../tables}
+\def\figwidth{0.8\textwidth}
 
@@
 
-The structure of the contact network underlying a viral epidemic can profoundly
-impact the speed and pattern of the epidemic's spread. Network structure can
-influence the prevalence trajectory~\autocite{o2010contact} and epidemic
+When an infectious disease spreads through a population, transmissions cannot
+generally occur between any two individuals at any time. Rather, the
+individuals must have a particular mode of contact with one another, which
+varies with the disease's mode of transmission. For airborne pathogens,
+physical proximity may be sufficient, but for sexually transmitted diseases,
+either sexual or blood-to-blood contact is required. The set of links between
+individuals along which transmission can occur is called the contact
+network~\autocite{klovdahl1985social}. The structure of the contact network
+underlying an epidemic can profoundly impact the speed and pattern of the
+epidemic's expansion. Network structure can influence the prevalence
+trajectory~\autocite{o2010contact} and epidemic
 threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates
@@
 way of interventions targeted to well-connected
-nodes~\autocite{little2014using}. True contact networks are a challenging type
-of data to collect, requiring extensive epidemiological
+nodes~\autocite{wang2015targeting}. True contact networks are a challenging
+type of data to collect, requiring extensive epidemiological
 investigation~\autocite{welch2011statistical}.
@@
 shape~\autocite{leventhal2012inferring, colijn2014phylogenetic,
-goodreau2006assessing, robinson2013dynamics}, suggesting that statistical
+goodreau2006assessing, robinson2013dynamics}, suggesting that such statistical
 inference might be possible~\autocite{welch2011statistical}.
@@
 meaning that the number of nodes of degree $k$ is proportional to $k^\gamma$
-for some constant $\gamma$. When $2 < \gamma \leq 3$, the network is referred
-to as ``scale-free''. The first contact network model incorporating
-preferential attachment was introduced by \textcite{barabasi1999emergence}, and
-is now referred to as the \gls{BA} model. Under this model, networks are formed
-by iteratively adding nodes with $m$ new edges each. These new edges are joined
-to existing nodes of degree $k$ with probability proportional to $k^\alpha$, so
+for some constant $\gamma$. These networks are also referred to as
+``scale-free''. The first contact network model incorporating preferential
+attachment was introduced by \textcite{barabasi1999emergence}, and is now
+referred to as the \gls{BA} model. Under this model, networks are formed by
+iteratively adding nodes with $m$ new edges each. These new edges are joined to
+existing nodes of degree $k$ with probability proportional to $k^\alpha$, so
 that nodes of high degree tend to attract more connections (in the original
@@
 \textcite{groendyke2011bayesian} and applied to a much larger data set of 188
-individuals. \textcite{brown2011transmission} fit a preferential attachment
-model to a phylogenetically estimated transmission network relating 60\% of HIV
+individuals. \textcite{brown2011transmission} found that a Waring distribution,
+which is produced by a different preferential attachment model, was a good fit
+to a phylogenetically estimated transmission network relating 60\% of HIV
 infected \gls{MSM} in the United Kingdom. The transmission network is a
@@
 
-Due to their complexity, it is generally difficult to explicitly calculate the
-likelihood of a transmission tree under a contact network model. To do so, we
-would need to integrate over all possible networks generated by that model, and
-also over all possible labellings of the internal nodes of the transmission
-tree. While it is not known (to us) whether such integration is tractable, a
-simpler alternative is offered by likelihood-free methods, namely
-\gls{ABC}~\autocite{rubin1984bayesianly, tavare1997inferring}. \gls{ABC}
-leverages the fact that, although calculating the likelihood may be impossible,
+Standard methods of model fitting involve calculation of the likelihood of
+observed data under the model. In maximum likelihood estimation, a quantity
+proportional to the likelihood is optimized, often through a standard
+multi-dimensional numerical optimization procedure. Bayesian methods integrate
+prior information by optimizing the posterior probability instead. To avoid
+calculation of a normalizing constant, Bayesian inference is often performed
+using \gls{MCMC}, which requires the calculation of likelihood ratios.
+Unfortunately, it is generally difficult to explicitly calculate the likelihood
+of an observed transmission tree under a contact network model. To do so, it
+would be necessary to integrate over all possible networks, and also over all
+possible labellings of the internal nodes of the transmission tree. While it is
+not known (to us) whether such integration is tractable, a simpler alternative
+is offered by likelihood-free methods, namely \gls{ABC}. \gls{ABC} leverages
+the fact that, although calculating the likelihood may be impossible,
 generating simulated datasets according to a model is often straightforward. If
@@
 tolerance~\autocite{sunnaker2013approximate}. In the specific case when $\rho$
-is a kernel function, the approach is known as
-kernel-\gls{ABC}~\autocite{nakagome2013kernel}.
+is a kernel function, the approach is known as kernel-\gls{ABC}.
 
@@
 kernel developed by \textcite{poon2013mapping}. We then estimate the parameters
-of the \gls{BA} model on a variety of simulated and real data sets.
+of the \gls{BA} model on a variety of simulated and real data sets. Our results
+show that the attachment power parameter $\alpha$ can be inferred with
+reasonable accuracy, and can vary considerably between epidemics from different
+settings.
 
@@
 for simulating epidemics and transmission trees over static contact networks,
-as has been done previously \autocite[\textit{e.g.}][]{o2010contact,
-robinson2013dynamics, leventhal2012inferring, groendyke2011bayesian}. To check
+in the same fashion as several previous studies
+\autocite[\textit{e.g.}][]{o2010contact, robinson2013dynamics,
+leventhal2012inferring, groendyke2011bayesian, goodreau2006assessing}. To check
 that our implementation was correct, we reproduced Figure 1A of
-\textcite{leventhal2012inferring} (see Figure~\ref{fig:sf1}).
+\textcite{leventhal2012inferring} (our \cref{fig:sf1}), which plots the
+unbalancedness of transmission trees simulated over four network models at
+various levels of pathogen transmissibility.
 
@@
 whether the parameters of the \gls{BA} model had an effect on tree shape. The
-parameters not being tested were fixed at the values $N$ = 5000, $I$ = 1000,
-$m$ = 2, and $\alpha$ = 1. The parameters were varied one at a time over the
-following values: $N$ = \sett{3000, 5000, 8000}, $I$ = \sett{500, 1000, 2000},
-$m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0.5, 1, 1.5}. For each parameter
-set, 100 networks were generated, and a transmission tree was simulated over
-each (300 trees per parameter set). The 300 trees were compared pairwise with
-the tree kernel to form a $300 \times 300$ kernel matrix. We constructed a
-\gls{kSVM} classifier for the parameter of interest using the
+parameters were varied one at a time over the following values: $N$ =
+\sett{3000, 5000, 8000}, $I$ = \sett{500, 1000, 2000}, $m$ = \sett{2, 3, 4},
+and $\alpha$ = \sett{0.5, 1, 1.5}. The parameters not being tested were fixed
+at the values $N$ = 5000, $I$ = 1000, $m$ = 2, and $\alpha$ = 1. For each
+parameter set, 100 networks were generated, and a transmission tree was
+simulated over each (300 trees per parameter set). The 300 trees were compared
+pairwise with the tree kernel to form a $300 \times 300$ kernel matrix. We
+constructed a \gls{kSVM} classifier for the parameter of interest using the
 \software{kernlab} package~\autocite{karatzoglou2004kernlab}, and evaluated its
@@
 We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
-\textcite{del2012adaptive}. To ensure our implementation was correct, we
+\textcite{del2012adaptive}. To check that our implementation was correct, we
 applied it to the same mixture of Gaussians used by
-\citeauthor{del2012adaptive} to demonstrate their method \autocite[originally
-used by][]{sisson2007sequential}. We were able to obtain a close approximation
-to the function (see Figure~\ref{fig:smctest}), and attained the stopping
-condition used by the authors in a comparable number of steps.
+\citeauthor{del2012adaptive} to demonstrate their method (originally used
+by~\textcite{sisson2007sequential}). We were able to obtain a close
+approximation to the function (see \cref{fig:smctest}), and attained the
+stopping condition used by the authors in a comparable number of steps.
 
@@
 of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
-\sett{1000, 2000}, $m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0.0, 0.5, 1, 1.5}.
-The adaptive \gls{ABC} algorithm was applied to each tree with these priors:
-$m \sim$ Uniform(1, 5), $\alpha \sim$ Uniform(0, 2), and $(N, I)$ jointly
-uniform on the triangular region \{$500 \leq N \leq 15000$, $500 \leq I \leq
- 5000$, $I \leq N$\}. Following \textcite{del2012adaptive} and
-\textcite{beaumont2009adaptive}, all proposals were Gaussian, with variance
-equal to twice the empirical variance of the particles. The algorithm was run
-with 1000 particles, 5 simulated datasets per particle, and the ``quality''
-parameter set to 0.95. We use the same stopping criterion as
-\textcite{del2012adaptive}, namely when the MCMC acceptance rate dropped below
-1.5\%. Point estimates for the parameters were obtained by taking the highest
-point of an estimated kernel density on the final set of particles.
+\sett{1000, 2000}, $m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0.0, 0.5, 1,
+1.5}. This produced a total of 24 parameter combinations $\times$ three trees
+per combination = 72 trees total. The adaptive \gls{ABC} algorithm was applied
+to each tree with these priors: $m \sim$ Uniform(1, 5), $\alpha \sim$
+Uniform(0, 2), and $(N, I)$ jointly uniform on the triangular region \{$500
+\leq N \leq 15000$, $500 \leq I \leq 5000$, $I \leq N$\}. Following
+\textcite{del2012adaptive} and \textcite{beaumont2009adaptive}, all proposals
+were Gaussian, with variance equal to twice the empirical variance of the
+particles. The algorithm was run with 1000 particles, 5 simulated datasets per
+particle, and the ``quality'' parameter controlling the decay rate of the
+tolerance $\varepsilon$ set to 0.95. We used the same stopping criterion as
+\citeauthor{del2012adaptive}, namely when the \gls{MCMC} acceptance rate
+dropped below 1.5\%. Point estimates for the parameters were obtained by taking
+the highest point of an estimated kernel density on the final set of particles,
+using the \software{density} function with the default parameters in
+\software{R}. \gls{HPD} intervals were calculated with the
+\software{HPDinterval} function from the \software{R} package
+\software{coda}~\autocite{plummer2006coda}.
 
@@
 
-We applied our kernel-ABC method to several HIV datasets. Because the \gls{BA}
-model generates networks with a single connected component, we specifically
-searched for datasets which originated from existing clusters, either
-phylogenetically or geographically defined. Characteristics of the datasets we
-investigated are given in Table~\ref{tab:data}.
+We applied our kernel-ABC method to several published HIV datasets. Because the
+\gls{BA} model generates networks with a single connected component, we
+specifically searched for datasets which originated from existing clusters,
+either phylogenetically or geographically defined. Characteristics of the
+datasets we investigated are given in \cref{tab:data}.
 
-\begin{table}
+\begin{table}[ht]
   \centering
-  \begin{tabular}{ccccc}
-    Reference & Sequences ($n$) & Location & Risk group & Gene \\
-    \hline
-    \autocite{wang2015targeting} & 173 & Beijing, China & MSM & \textit{pol} \\
-    \autocite{cuevas2009hiv} & 287 & Basque Country, Spain & mixed & \textit{pol} \\
-    \autocite{novitsky2013phylogenetic} & \multirow{2}{*}{1299} & \multirow{2}{*}{Mochudi, Botswana} & \multirow{2}{*}{mixed} & \multirow{2}{*}{\textit{env}} \\
-    \autocite{novitsky2014impact} \\
-    \hline
-  \end{tabular}
+  \input{\tablepath/realdata}
   \caption{Characteristics of published datasets investigated with kernel-ABC.
-  Acronyms: MSM, men who have sex with men.}
+  Acronyms: MSM, men who have sex with men; IDU, injection drug users.}
   \label{tab:data}
@@
 
-We downloaded all sequences associated with each study from GenBank. Sequences
-were multiply aligned using \software{MUSCLE} version 3.8.31
-\autocite{edgar2004muscle}, and alignments were manually inspected with
-\software{Seaview} version 4.4.2 \autocite{gouy2010seaview}. Phylogenies were
-constructed from the nucleotide alignments by approximate maximum likelihood
-using \software{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
-\gls{GTR} model. Transmission trees were estimated by rooting and time-scaling
-the phylogenies by root-to-tip regression, using a modified version of
-Path-o-Gen \autocite[distributed as part of \software{BEAST},
-][]{drummond2007beast}, as described
+We downloaded all sequences associated with each study from GenBank. For the
+\textcite{novitsky2014impact} data, each sequence was aligned pairwise to the
+HXB2 reference sequence (Genbank accession number HIVHXB2CG) and the
+hypervariable regions were clipped out with BioPython version
+1.66+~\autocite{cock2009biopython}. Sequences were multiply aligned using
+\software{MUSCLE} version 3.8.31 \autocite{edgar2004muscle}, and alignments
+were manually inspected with \software{Seaview} version 4.4.2
+\autocite{gouy2010seaview}. Phylogenies were constructed from the nucleotide
+alignments by approximate maximum likelihood using \software{FastTree2} version
+2.1.7 with the \gls{GTR} model. Transmission trees were estimated by rooting
+and time-scaling the phylogenies by root-to-tip regression, using a modified
+version of Path-O-Gen (distributed as part of
+BEAST~\autocite{drummond2007beast}) as described
 previously~\autocite{poon2015phylodynamic}.
 
-Two additional steps were undertaken for the
-\textcite{novitsky2013phylogenetic, novitsky2014impact} data. First, before
-performing the multiple sequence alignment, we aligned each
-sequence pairwise to the HXB2 reference sequence (GenBank accession number
-HIVHXB2CG) using \software{MAFFT} version 7.130b \autocite{katoh2013mafft} and
-clipped out the hypervariable regions with \software{Biopython}
-version 1.66+ \autocite{cock2009biopython}. Second, after multiple sequence
-alignment, the sequences were trimmed to an informative region with
-\software{trimAl} version 1.4.rev15 \autocite{capella2009trimal}.
-
-Kernel-ABC was used to fit the \gls{BA} model to each dataset. To make the
-analyses comparable, the same priors and configuration parameters were used for
-each dataset. The priors were as follows: $\alpha \sim$ Uniform(0, 2), $m \sim$
-Uniform(1, 5), and $(N, I) \sim$ Uniform(\sett{$n \leq N \leq 10000$, $n \leq I
-\leq 1000$, $I \leq N$}), where $n$ is the number of sequences in the dataset.
-The kernel meta-parameters and SMC configuration were the same as in the
-\gls{ABC} simulations.
+Two of the datasets \autocite{li2015hiv,novitsky2014impact} were initially much
+larger than the others, containing 1265 and 1299 sequences respectively. To
+ensure that the analyses were comparable, we reduced these to a number of
+sequences similar to the smaller datasets. For the \textcite{li2015hiv} data,
+we detected a cluster of size 280 using a patristic distance cutoff of 0.02 as
+described previously~\autocite{poon2014impact}. Only sequences within this
+cluster were carried forward. For the \textcite{novitsky2014impact} data, no
+large clusters were detected using the same cutoff, so we analysed a subtree of
+size 180 chosen arbitrarily.
 
@@
 Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
-tested. For appropriate choices of the kernel meta-parameters, the classifier
-had an $R^2$ above 
+tested~\cref{fig:rsquared}. Classifiers based on two other tree statistics, the
+\gls{nLTT} and Sackin's index, generally exhibited worse performance than the
+tree kernel, although the magnitude of the disparity varied between the
+parameters (\cref{fig:crossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}). For
+appropriate choices of the kernel meta-parameters, the kSVM classifier for
+$\alpha$ had an $R^2$ above 
     0.8 
-for all combinations of tree size and epidemic size. The $I$ parameter was also
-classified accurately, with an average $R^2$ of 
+for all combinations of tree size and epidemic size (\cref{fig:crossv}). The
+other classifiers' $R^2$ averaged between
+  0.49
+and
+  0.61
+for the SVM based on the \gls{nLTT}, and between
+  0.73
+and
+  0.76
+for the linear regression against Sackin's index. The $I$ parameter was also
+classified accurately by the tree kernel (\cref{fig:Icrossv}), with an average
+$R^2$ of 
   0.93 
@@
   0.7 
-for 100-tip trees. The $m$ parameter was much harder to classify, with $R^2$
-varying between 
+for 100-tip trees. The \gls{nLTT} performed almost as well on this parameter as
+the tree kernel, with $R^2$ values of
+  0.83
+for 500-tip trees or
+  0.55
+for 100-tip trees. Sackin's index was an extremely poor classifier for this
+parameter, with both $R^2$ values below
+  0.08.
+
+The $m$ parameter was much harder to classify (\cref{fig:mcrossv}), with $R^2$
+for the kSVM varying between 
   0.004
@@
   0.36
-depending on $I$ and the tree size. Finally, the accuracy of the classifier for
-$N$ varied widely, from 
+depending on $I$ and the tree size. The \gls{nLTT} was also universally poor,
+with $R^2$ below
+  0.02
+for all scenarios. Sackin's index fared slightly better, especially on trees
+with a higher number of tips, with $R^2$ between
+  0
+and
+  0.18.
+Finally, the accuracy of the kSVM classifier for $N$ varied widely
+(\cref{fig:Ncrossv}), from 
   0.08
@@
   0.82.
-Figure~\ref{fig:crossv} shows the results for $\alpha$; similar figures for the
-other parameters can be found in the supplemental materials. Based on
-inspection of the cross-validation results, we chose to use the meta-parameters
-$\lambda = 0.3$ and $\sigma = 4$ in further analyses, and not to use Sackin's
-index or the \gls{nLTT}.
+The \gls{nLTT} performed well in classifying $N$ when the epidemic was large,
+improving upon the tree kernel in the scenarios with $I$ = 2000 and 100- or
+500-tip transmission trees ($R^2$ = 
+  0.31
+and
+  0.53),
+although it performed poorly when $I$ was 500 or 1000 (average $R^2$ =
+  0.07).
+Sackin's index could not classify $N$, with all $R^2$ below
+  0.11.
+Based on inspection of the cross-validation results, we chose to use the
+meta-parameters $\lambda = 0.3$ and $\sigma = 4$ in further analyses, and not
+to use Sackin's index or the \gls{nLTT}.
 
-\begin{figure}
-  \includegraphics{kernel-alpha-crossv}
-  \caption{Cross-validation $R^2$ of kernel-SVM classifier for the $\alpha$
-    parameter of the \gls{BA} network model.}
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\figwidth]{kernel-rsquared}
+  \caption{
+    Cross-validation accuracy of kernel-SVM classifiers for \gls{BA} model
+    parameters, with kernel meta-parameters $\lambda = 0.3$ and $\sigma = 4$.
+    Each point was calculated based on 300 simulated transmission trees over
+    networks with three different values of the parameter being tested.
+    Vertical lines are empirical 95\% confidence intervals based on 1000 2-fold
+    cross-validations.
+  }
+  \label{fig:rsquared}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\figwidth]{kernel-alpha-crossv}
+  \caption{
+    Cross-validation accuracy of kernel-SVM classifiers for $\alpha$ parameter
+    of \gls{BA} network model, for various tree kernel meta-parameters and
+    epidemic scenarios. Each point was calculated based on 300 simulated
+    transmission trees over networks with $\alpha$ = 0.5, 1.0, or 1.5. Dotted
+    and and dashed lines indicate, respectively, performance of SVM using the
+    \gls{nLTT} statistic, and linear regression using Sackin's index. Facets
+    are number of infected nodes before the simulation was stopped ($I$) and
+    number of tips in the sampled transmission tree.
+  }
   \label{fig:crossv}
@@
 
-Figure~\ref{fig:abcpt} shows point estimates for the simulations with $m = 2$.
-The results for the other values of $m$ are similar and can be found in the
-supplemental materials.
+Of the four parameters, $\alpha$ was the most accurately estimated, with a
+median [IQR] absolute error of 
+    0.11 
+    [0.05-0.18].
+(\cref{fig:abcpt,tab:abchpd}). The accuracy of the estimates was not
+significantly different between values of $m$ or $I$ (both one-way ANOVA,
+    $p$ = 0.1
+and 
+    0.25),
+although the errors when the true value of $\alpha$ was zero were
+significantly greater than the other values 
+    (Wilcoxon rank-sum test, $p$ = \ensuremath{6.41\times 10^{-4}}).
+The error in the estimated value of $I$ was
+    305.66 
+    [107.76-606.59].
+Errors were significantly higher for $\alpha \geq 1$
+    (Wilcoxon rank-sum test, $p$ = \ensuremath{6.12\times 10^{-4}})
+and for $I$ = 2000
+    ($p$ = \ensuremath{1.58\times 10^{-6}}),
+but not for any values of $m$
+    (one-way ANOVA, $p$ = 0.33).
+The $m$ parameter was estimated correctly in
+    37 \%
+of simulations, with an error of one in
+    40 \%
+and of two or more in 
+    22 \%
+(the only possible $m$ values were 1, 2, 3, 4, or 5). The true values of $m$
+and $I$ did not significantly affect the error
+    (one-way ANOVA, $p$ = 0.5 and
+                          0.68).
+%but the accuracy was significantly lower for integral than non-integral
+%values of $\alpha$
+%    (Wilcoxon rank-sum test, $p$ = alpha.m.test$p.value).
+Finally, the total number of nodes $N$ was consistently over-estimated by about
+a factor of two
+    (error \ensuremath{6.59\times 10^{3}} 
+    [\ensuremath{4.21\times 10^{3}}-\ensuremath{8.28\times 10^{3}}]).
+No other parameters influenced the accuracy of the $N$ estimates 
+    (one-way ANOVA, $p \geq$ 0.72).
 
-The marginal approximated posterior distributions for one simulated dataset are
-shown in Figure~\ref{fig:abcex}.
-
-\begin{figure}
-    \includegraphics{abc-point-est}
-    \caption{Point estimates of \gls{BA} model parameters obtained by running
-    kernel-ABC on simulated phylogenies. Dotted lines indicate the true values,
-    and limits of the $y$-axes are the regions of uniform prior density.}
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\figwidth]{abc-point-estimate}
+  \caption{
+    Point estimates of \gls{BA} model parameters obtained by running kernel-ABC
+    on simulated phylogenies. Dotted lines indicate the true values, and limits
+    of the $y$-axes are the regions of uniform prior density.
+  }
   \label{fig:abcpt}
@@
 
-\begin{figure}
-    \includegraphics{abc-example}
+\begin{table}[ht]
+  \centering
+  \input{\tablepath/abc-hpd}
+  \caption{Average \gls{HPD} interval widths for \gls{ABC} model parameter
+           estimates.}
+  \label{tab:abchpd}
+\end{table}
+
+The dispersion of the \gls{ABC} approximation to the posterior also varied
+between the parameters, with narrower \gls{HPD} intervals for those with the
+most accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows the
+distributions for for one simulation (equivalent plots for all the simulations
+can be found in the supplemental materials). \gls{HPD} intervals around
+$\alpha$ and $I$ were narrow relative to the region of nonzero prior density,
+whereas the intervals for $m$ and $N$ were more widely dispersed.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\figwidth]{abc-posterior-example}
   \caption{Marginal estimated posterior distributions obtained with ABC on a
@@
 
+To test the effect of model misspecification, we simulated one network where
+the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
+other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The \gls{MAP} [95\%
+HPD] estimates for each parameter were: 
+$\alpha$, 
+  1 [0.85-
+                           1.12];
+$I$,
+  5.23 [2.05-
+                       5.85];
+$m$,
+  \ensuremath{10^{4}} [\ensuremath{3.06\times 10^{3}}-
+                              \ensuremath{1.5\times 10^{4}}];
+$N$,
+  \ensuremath{1.17\times 10^{3}} [505.71-
+                       \ensuremath{1.75\times 10^{3}}].
+
+To test the effect of sampling bias, we sampled one transmission tree in a
+peer-driven fashion, where the probability to sample a node was twice as high
+if one of it's peers had already been sampled. The parameters for this
+experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and $I$ = 2000. The
+estimated values were
+$\alpha$, 
+  0.48 [0.03-
+                           0.83];
+$I$,
+  3.6 [2.18-
+                       5.9];
+$m$,
+  \ensuremath{1.08\times 10^{4}} [\ensuremath{3.11\times 10^{3}}-
+                              \ensuremath{1.48\times 10^{4}}];
+$N$,
+  \ensuremath{2.48\times 10^{3}} [\ensuremath{1.28\times 10^{3}}-
+                       \ensuremath{3.87\times 10^{3}}].
+
 \subsection*{Real data}
+
+There was substantial heterogeneity among the parameter estimates for the five
+published HIV datasets we analysed. Two of the
+datasets~\autocite{niculescu2015recent, wang2015targeting} had estimated
+$\alpha$ values near unity (\gls{MAP} estimate [95\% \gls{HPD}] 
+  1.06 
+  [0.63-1.27]
+and
+  1 
+  [0.41-1.16]). Another two
+datasets~\autocite{li2015hiv, cuevas2009hiv} had lower estimated values and wider
+\gls{HPD} intervals
+  (0.77 
+  [0.01-1.03]
+and
+  0.66 
+  [0.03-0.84]). 
+The \textcite{novitsky2014impact} data had an extremely low estimated $\alpha$
+and a very wide \gls{HPD} interval
+  0.17 
+  [0.04-1.39]). 
+For all the datasets except \textcite{novitsky2014impact}, estimated values of
+$I$ were below 2000, with narrow \gls{HPD} intervals around two of the
+datasets
+  (\textcite{cuevas2009hiv}, 880.52 
+  [290.7-\ensuremath{1.51\times 10^{3}}];
+   \textcite{niculescu2015recent}, 175.05
+  [138.74-454.51])
+and wider intervals around the other two
+  (\textcite{li2015hiv}, \ensuremath{1.59\times 10^{3}} 
+  [284.01-\ensuremath{3.81\times 10^{3}}];
+   \textcite{wang2015targeting}, 651.72
+  [268.94-\ensuremath{4.24\times 10^{3}}]).
+The \textcite{novitsky2014impact} data was again the outlier, with a very high
+estimated $I$, and \gls{HPD} interval spanning almost the entire prior region
+  (\ensuremath{7.55\times 10^{3}} 
+  [228.96-\ensuremath{8.92\times 10^{3}}]).
+Little information was gleaned about the $m$ parameter, with the \gls{HPD}
+interval occuping the entire prior region for all datasets. The estimates of
+$N$ were similarly uninformative, with the exception that the point estimate
+for the \textcite{wang2015targeting} data was smaller
+  (\ensuremath{5.84\times 10^{3}})
+than the estimates for other datasets.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\figwidth]{realdata-hpd}
+  \caption{
+    Point estimates and 95\% \gls{HPD} intervals for \gls{BA} network
+    parameters of published HIV datasets, obtained with kernel-ABC.
+  }
+\end{figure}
 
@@
 
-% kernel classifiers
+Contact networks can have a strong influence on epidemic progression, and are
+potentially useful as a public health tool. Despite this, few methods exist for
+investigating contact network parameters in a phylodynamic framework.
+Kernel-ABC is a model-agnostic method which can be used to investigate any
+quantity that affects tree shape. In this work, we developed a kernel-ABC-based
+method to infer the parameters of a contact network model. The method is
+general, and could be applied to any model from which contact networks can be
+simulated. We demonstrated the method on the \gls{BA} model, which is a simple 
+model incorprorating the preferential attachment feature commonly seen in
+real-world networks.
 
-The $m$ and $N$ parameters were much harder to recover than $\alpha$ and $I$,
-suggesting differing influences of these parameters on tree shape. The $\alpha$
-parameter in particular was accurately estimated in most cases, implying that
-the strength of preferential attachment has a strong impact on the shape of the
-transmission tree.
+By training a kernel-SVM classifier, we found that the $\alpha$ and $I$
+parameters, representing preferential attachment power and number of infected
+nodes, had a strong influence on tree shape. This was reflected in the relative
+accuracy of the kernel-ABC estimates of these parameters. On the other hand,
+the $m$ and $N$ parameters did not produce much variation in tree shape,
+resulting in both poorly performing classifiers and uninformative kernel-ABC
+estimates. The total number of nodes $N$ was almost always significantly
+over-estimated. Since the prior on $N$ and $I$ is jointly uniform on a
+triangular region ($I \leq N$), there is more prior mass on high $N$ values. In
+retrospect, it is unreasonable to expect good estimation of $N$, because adding
+more nodes to a \gls{BA} network does not change the edge density or overall
+shape. This can be illustrated by imaging that we add a small number of nodes
+to a network after the epidemic simulation has already been completed. It is
+possible that none of these new nodes attains a connection to any infected
+node. Thus, running the simulation again on the new, larger network could
+produce the exact same transmission tree as before.
 
-% abc simulations
+Our investigation of published HIV datasets indicated heterogeneity in the
+contact network structures underlying several distinct local epidemics.
+\textcite{cuevas2009hiv} studied a group of newly diagnosed individuals in the
+Basque Country, Spain. Although the individuals were of mixed risk group, and
+therefore unlikely to comprise a single contact network, a high proportion of
+them (47\%) grouped into local transmission clusters, allowing for the
+possibility of weak preferential attachment dynamics. We estimated a relatively
+low attachment power for these data, which is consistent with the sampled
+sequences comprising many distinct sub-networks. \textcite{li2015} sampled a
+large number of acutely infected \gls{MSM} in Shanghai, China. By using a
+patristic distance cutoff~\autocite{poon2014impact}, we identified and analysed
+a large cluster. Surprisingly, the estimated attachment power was fairly low
+with a large credible interval.
 
-The total number of nodes $N$ was almost always significantly over-estimated.
-This suggests that $N$ has very little effect on tree shape, and the estimates
-were influenced primarily by the prior. Since the prior on $N$ and $I$ is
-jointly uniform on a triangular region, there is more prior mass on high $N$
-values. In retrospect, this is reasonable because adding more nodes to a
-\gls{BA} network does not change the edge density or the network's overall
-shape. This can illustrated by imagining that we grow the network after the
-epidemic simulation has already been completed. It is possible (though perhaps
-unlikely) that none of the new nodes attains a connection to any infected node.
-Thus, running the simulation again on the new, larger network could produce the
-exact same transmission tree as before. This is an extreme example, but it
-serves to demonstrate the point that $N$ may be impossible to determine if the
-epidemic does not come close to saturating the network.
+\textcite{niculescu2015recent} studied a recent outbreak among Romainian
+\gls{IDU}. The estimated attachment power for this dataset was the highest
+among all we considered, at slightly above one. The dataset constituted a
+fairly high sampling prevalence of 33\% of newly diagnosed \gls{IDU} infections
+over the study period. \textcite{wang2015targeting} sampled acutely infected
+\gls{MSM} in Beijing, China, and discovered a high degree of clustering which
+indicated that the local \gls{MSM} subnetwork had been deeply sampled. In both
+these datasets, the estimated attachment power was close to one.
 
-% real data
+\textcite{novitsky2013phylogenetic} sampled approximately 44\% of the
+HIV-infected individuals in the northern area of Mochudi, Botswana. Additional
+sampling in a later study~\autocite{novitsky2014impact} brought the genotyping
+coverage up to 70\%. Even with such a high sampling coverage, we did not detect
+any large clusters using patristic distance, and therefore chose to analyze a
+subtree instead. The estimated attachment power was extremely low, with a very
+wide credible interval. The estimated number of infected nodes was also
+extremely high, much higher than the HIV-infected population of the town, with
+a similarly wide credible interval. Several factors may have contributed to
+this result. First, the authors note that the their sample was 75\% female. In
+a primarily heterosexual risk environmet, removal of a disproportionate nmuber
+of males from the network could obfuscate the true network structure. Second,
+the town in question was in close proximity to the country's capital, and the
+authors indicated that a high amount of migration takes place between the two
+locations. This suggests a reason for the high estimate of $I$: the contact
+network includes a much larger group based in the capital city.
 
-We estimated the parameters of the \gls{BA} model for several published
-datasets (see~\ref{tab:data}). \textcite{cuevas2009hiv} performed a
-phylogenetic analysis of 261 patients of mixed risk groups in the Basque
-Country, Spain, who were newly diagnosed between 2004 and 2007. Although their
-data did not comprise a single cluster, they did find that almost half (47\%)
-of the samples grouped in transmission clusters. \textcite{wang2015targeting}
-reported on a cohort of acutely infected \gls{MSM} in Beijing, China. Through
-phylogenetic analysis, they found that they had ``deeply sampled'' the local
-\gls{MSM} network, suggesting that many of the samples could be
-epidemiologically related. \textcite{novitsky2013phylogenetic} investigated a
-local HIV epidemic in Mochudi, Botswana. They later extended their
-analysis~\autocite{novitsky2014impact} to include an estimated 70\% of HIV
-infected persons in the town.
+When interpreting these results, we caution that the \gls{BA} model is quite
+simple and most likely misspecified for these data. In particular, the average
+degree of a node in the network is equal to $2m$, and therefore is constrained
+to be a multiple of 2. Furthermore, we considered the case $m = 1$, where the
+network has no cycles, to be implausable and assigned it zero prior
+probability. This forces the average degree to be at least four, which may be
+unrealistically high for sexual networks. Additional modelling assumptions,
+include the network being connected and static, the epidemic following simple
+\gls{SI} dynamics, and the underlying behaviour of all nodes is the same. This
+last is particularly problematic, as we showed by simulating a network where
+some nodes exhibited a higher attachment power than others. The estimated
+attachment power was simply the average of the two values, indicating that,
+although we could characterize the network in aggregate, the estimated
+parameters could not be said to apply to any individual node. However, we note
+that, given the model-agnostic nature of our method, it is possible to fit a
+model with heterogeneous node behaviour, and indeed this may prove a fruitful
+avenue for future investigations.
 
-Our method has a number of caveats, perhaps the most significant being that it
+Our method itself has of caveats, perhaps the most significant being that it
 takes a transmission tree as input. In reality, true transmission trees are not
-available and must be estimated, often by way of a viral phylogeny. Although
-this has been demonstrated to work in practice~\autocite{leitner1996accurate},
-the topologies of a viral phylogeny and transmission tree can differ
-significantly~\autocite{ypma2013relating}. Additionally, the \gls{BA} model
-that we fitted here makes a number of unrealistic modelling assumptions, such
-as that the network is connected and static, the nodes obey \gls{SI} dynamics,
-and that the underlying behaviour of all nodes is the same. This last is
-particularly problematic, as we showed by simulating a network where some nodes
-exhibited a higher attachment power than others. The estimated attachment power
-was simply the average of the two values, indicating that, although we could
-characterize the network in aggregate, the estimated parameters could not be
-said to apply to any individual node. Finally, the method is computationally
-intensive, taking about a day when run on 20 cores in parallel.
-
-\section*{Conclusions}
-
-We developed a novel method which uses kernel-\gls{ABC} to fit contact network
-models to transmission trees, and used it to fit the \gls{BA} preferential
-attachment model to both simulated and real data.
+available and must be approximated, often by way of a viral phylogeny. Although
+this has been demonstrated to be a fair approximation in
+practice~\autocite[e.g.][]{leitner1996accurate}, the topologies of a viral
+phylogeny and transmission tree can differ
+significantly~\autocite{ypma2013relating}. In addition, it is computationally
+intensive, taking about a day when run on 20 cores in parallel with the setings
+we described in the methods. However, there are no other methods available to
+perform the types of estimation we do here, and we therefore believe our method
+will be useful to epidemiological researchers interested in the network
+structure underlying disease outbreaks.
 
@@
 
+\end{document}
 \newpage
-
 \section*{Supplemental Materials}
@@
 \begin{figure}[ht]
-  \includegraphics[scale=0.8]{leventhal2012fig1}
+  \centering
+  %\includegraphics[scale=0.8]{leventhal2012fig1}
   \caption{Reproduction of Figure 1A from \textcite{leventhal2012inferring}.}
@@
 
-\begin{figure}
-  \includegraphics{smc_test}
+\begin{figure}[ht]
+  \centering
+  %\includegraphics{smc_test}
   \caption{Approximation of mixture of Gaussians used by
@@
 
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\figwidth]{kernel-m-crossv}
+  \caption{Cross-validation accuracy of kernel-SVM classifiers for $m$
+      parameter of \gls{BA} network model, for various tree kernel
+      meta-parameters and epidemic scenarios. Each point was calculated based
+      on 300 simulated transmission trees over networks with $m$ = 2, 3, or 4.
+      Dotted and and dashed lines indicate, respectively, performance of SVM
+      using the \gls{nLTT} statistic, and linear regression using Sackin's
+      index. Facets are number of infected nodes before the simulation was
+      stopped ($I$) and number of tips in the sampled transmission tree.
+  }
+  \label{fig:mcrossv}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\figwidth]{kernel-I-crossv}
+  \caption{Cross-validation accuracy of kernel-SVM classifiers for number of
+      infected nodes ($I$) under \gls{BA} network model, for various tree
+      kernel meta-parameters and two tree sizes. Each point was calculated
+      based on 300 simulated transmission trees over networks with $I$ = 500,
+      1000, or 2000. Dotted and and dashed lines indicate, respectively,
+      performance of SVM using the \gls{nLTT} statistic, and linear regression
+      using Sackin's index. Facets are the number of tips in the sampled
+      transmission tree.
+  }
+  \label{fig:Icrossv}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\figwidth]{kernel-N-crossv}
+  \caption{Cross-validation accuracy of kernel-SVM classifiers for total number
+      of nodes ($N$) under \gls{BA} network model, for various tree kernel
+      meta-parameters and epidemic scenarios sizes. Each point was calculated
+      based on 300 simulated transmission trees over networks with $N$ = 3000,
+      5000, or 8000. Dotted and and dashed lines indicate, respectively,
+      performance of SVM using the \gls{nLTT} statistic, and linear regression
+      using Sackin's index. Facets are the number of tips in the sampled
+      transmission tree.
+  }
+  \label{fig:Ncrossv}
+\end{figure}
+
 \end{document}

### 2016-03-03 doc/tables/realdata.tex
First draft of paper! Yesgit add paper.pdf -f

--- 2016-02-23 doc/paper/paper.Rnw
@@
-\documentclass[12pt]{report}
-\usepackage[osf]{garamondx}
-\usepackage[garamondx,cmbraces]{newtxmath}
-\usepackage{marginnote}
-\usepackage[top=2cm, bottom=2cm, outer=3cm, inner=3cm, heightrounded, 
-            marginparwidth=2cm, marginparsep=0.5cm]{geometry}
-\usepackage{setspace}
-\usepackage[backend=biber, style=authoryear]{biblatex}
-\usepackage{nicefrac}
-\usepackage{amsmath}
-\usepackage{algorithm}
-\usepackage{algorithmicx}
-\usepackage{algpseudocode}
-\usepackage[acronym,nonumberlist]{glossaries}
-
-\newcommand{\comment}[1]{\marginnote{\textit{#1}}}
-
-\newcommand{\defn}[1]{\textit{#1}}
-\newcommand{\software}[1]{\textit{#1}}
-
-\newcommand{\set}[1]{\left\lbrace#1\right\rbrace}
-\newcommand{\sett}[1]{\{#1\}}
-\renewcommand{\emptyset}{\varnothing}
-\renewcommand{\vec}[1]{\mathbf{#1}}
-\renewcommand{\star}[1]{#1^*}
-\renewcommand{\d}{\mathrm{d}\,}
-
-\DeclareMathOperator*{\argmin}{arg\,min}
-\DeclareMathOperator*{\argmax}{arg\,max}
-
-\DeclareMathOperator{\Exponential}{Exponential}
-\DeclareMathOperator{\Uniform}{Uniform}
-\DeclareMathOperator{\E}{E}
-\DeclareMathOperator{\Var}{Var}
-
-\DeclareMathOperator{\inc}{in}
-\DeclareMathOperator{\out}{out}
-
-\newcommand{\N}{\mathcal{N}}
-\renewcommand{\L}{\mathcal{L}}
-
-\newacronym{BA}{BA}{Barab\'asi-Albert}
-\newacronym{SI}{SI}{susceptible-infected}
-\newacronym{kSVM}{kSVM}{kernel support vector machine}
-\newacronym{nLTT}{nLTT}{normalized lineages-through-time}
-\newacronym{SMC}{SMC}{sequential Monte-Carlo}
-\newacronym{ABC}{ABC}{approximate Bayesian computation}
-\newacronym{MSM}{MSM}{men who have sex with men}
-
-\graphicspath{{figures/}}
-
-\title{Phylodynamic inference of contact network parameters through approximate
-Bayesian computation}
-\author{Rosemary M. McCloskey \and Richard H. Liang \and Art F.Y. Poon}
-
-\addbibresource{papers.bib}
-
-\frenchspacing
-\onehalfspacing
-
-\begin{document}
-
-\maketitle
-
-\section*{Background}
-
-From a theoretical perspective, network structure can profoundly influence the
-incidence and prevalence trajectories of an epidemic, in turn affecting the
-estimates of quantities such as effective population size. More practically,
-contact networks have been explored as tools for curtailing epidemic spread, by
-way of interventions targeted to well-connected
-nodes~\autocite{little2014using}.
-
-Survey-based studies of sexual networks~\autocite{liljeros2001web,
-schneeberger2004scale} have found that sexual contact networks are best
-described by a preferential attachment model \autocite[but not all,
-see][]{jones2003assessment}. This result has been coroborated by more recent
-investigations using phylogenetic methods~\autocite{brown2011transmission}.
-Under these models, nodes with a high number of contacts attract new
-connections at an elevated rate. Networks produced by preferential attachment
-have a power-law degree distribution, meaning that the number of nodes of
-degree $k$ is proportional to $k^\gamma$ for some constant $\gamma$. When $2 <
-\gamma \leq 3$, the network is referred to as ``scale-free''. The first contact
-network model incorporating preferential attachment was introduced by
-\textcite{barabasi1999emergence}, and is now referred to as the \gls{BA} model.
-Under this model, networks are formed by iteratively adding nodes with $m$ new
-edges each. These new edges are joined to existing nodes of degree $k$ with
-probability proportional to $k^\alpha$, so that nodes of high degree tend to
-attract more connections (in the original paper, only $\alpha = 1$ was
-investigated).
-
-Due to their complexity, it is generally difficult to explicitly calculate the
-likelihood of an observed transmission tree under a contact network model. To
-calculate the normalizing constant, we would need to integrate over all
-possible networks, and also over all possible labellings of the internal nodes
-of the transmission tree. While it is not known (to us) whether such
-integration is tractable, a simpler alternative is offered by likelihood-free
-methods, namely \gls{ABC}. \gls{ABC} leverages the fact that, although
-calculating the likelihood may be impossible, generating simulated datasets
-according to a model is often straightforward. If our model fits the data well,
-the simulated data it produces should be similar to the observed data. More
-formally, if $D$ is the observed data, the posterior distribution $f(\theta
-\mid D)$ on model parameters $\theta$ is replaced as the target of statistical
-inference by $f(\theta \mid \rho(\hat{D}, D) < \varepsilon)$, where $\rho$ is a
-distance function, $\hat{D}$ is a simulated dataset according to $\theta$, and
-$\varepsilon$ is a small tolerance~\autocite{sunnaker2013approximate}. In the
-specific case when $\rho$ is a kernel function, the approach is known as
-kernel-\gls{ABC}.
-
-Here, we apply kernel-\gls{ABC} to the problem of statistical inference of
-contact network parameters from an estimated transmission tree. We then
-estimate the parameters of the \gls{BA} model on a variety of simulated and
-real data sets.
-
-\section*{Methods}
-
-We implemented a Gillespie simulation algorithm~\autocite{gillespie1976general}
-for simulating epidemics and transmission trees over static contact networks,
-as has been done previously \autocite[\textit{e.g.}][]{o2010contact,
-robinson2013dynamics, leventhal2012inferring, groendyke2011bayesian}. To check
-that our implementation was correct, we reproduced Figure 1 of
-\textcite{leventhal2012inferring} (see Figure~\ref{fig:sf1}).
-
-We chose to study the \gls{BA} network model~\autocite{barabasi1999emergence}.
-In addition to $m$ and $\alpha$, we investigated the parameters $N$, which
-denotes the total number of nodes in the network, and $I$, which is the number
-of infected nodes at which to stop the simulation and sample the transmission
-tree. Nodes in our networks followed simple \gls{SI} dynamics, meaning that
-they became infected at a rate proportional to their numbers of infected
-neighbours, and never recovered. For all analyses, the transmission trees'
-branch lengths were scaled by dividing by their mean. We used the 
-\software{igraph} library's implementation of the \gls{BA}
-model~\autocite{csardi2006igraph} to generate the graphs.
-
-\subsection*{Kernel classifiers}
-
-We used the phylogenetic kernel developed by \textcite{poon2013mapping} to test
-whether the parameters of the \gls{BA} model had an effect on tree shape. The
-parameters not being tested were fixed at the values $N$ = 5000, $I$ = 1000,
-$m$ = 2, and $\alpha$ = 1. The parameters were varied one at a time over the
-following values: $N$ = \sett{3000, 5000, 8000}, $I$ = \sett{500, 1000, 2000},
-$m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0.5, 1, 1.5}. For each parameter
-set, 100 networks were generated, and a transmission tree was simulated over
-each (300 trees per parameter set). The 300 trees were compared pairwise with
-the tree kernel to form a $300 \times 300$ kernel matrix. We constructed a
-\gls{kSVM} classifier for the parameter of interest using the
-\software{kernlab} package~\autocite{karatzoglou2004kernlab}, and evaluated its
-accuracy with 1000 two-fold cross-validations. The construction of the kernel
-matrices and classifiers was repeated for several combinations of the kernel
-meta-parameters $\lambda$ (the ``decay factor''), and $\sigma$ (the ``radial
-basis function variance'')~\autocite[see][]{poon2013mapping}, and for
-transmission trees of sizes 100, 500, and 1000. We also tested univariate
-classifier based on Sackin's index~\autocite{shao1990tree} and an ordinary SVM
-based on the \gls{nLTT} statistic~\autocite{janzen2015approximate}.
-
-\subsection*{ABC simulations}
-
-We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
-\textcite{del2012adaptive}. To ensure our implementation was correct, we
-applied it to the same mixture of Gaussians used by
-\citeauthor{del2012adaptive} to demonstrate their method (originally used
-by~\textcite{sisson2007sequential}). We were able to obtain a close
-approximation to the function (see Figure~\ref{fig:smctest}), and attained the
-stopping condition used by the authors in a comparable number of steps.
-
-We simulated three transmission trees, each with 500 tips, under every element
-of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
-\sett{1000, 2000}, $m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0, 0.5, 1, 1.5,
-2}. The adaptive \gls{ABC} algorithm was applied to each tree with these
-priors: $N$ = Uniform(500, 15000), $I$ = Uniform(500, 5000), $m$ = Uniform(1,
-5), and $\alpha$ = Uniform(0, 2). Following \textcite{del2012adaptive} and
-\textcite{beaumont2009adaptive}, all proposals were Gaussian, with variance
-equal to twice the empirical variance of the particles. Because there must be
-at least as many nodes in the network as infected nodes, the proposal ratio was
-set to zero whenever $I > N$. The algorithm was run with 1000 particles, 5
-simulated datasets per particle, and the ``quality'' parameter set to 0.95. We
-use the same stopping criterion as \textcite{del2012adaptive}, namely when the
-MCMC acceptance rate dropped below 1.5\%. Point estimates for the parameters
-were obtained by taking the highest point of an estimated kernel density on the
-final set of particles.
-
-Two further analyses were performed to address potential sources of error. To
-evaluate the effect of model misspecification in the case of heterogeneity
-among nodes, we generated a network where half the nodes were attached with
-power $\alpha$ = 0.5, and the other half with power $\alpha$ = 1.5. The other
-parameters for this network were $N$ = 5000, $I$ = 1000, and $m$ = 2. To
-investigate the effects of potential sampling bias, we simulated a transmission
-tree where the tips were sampled in a peer-driven fashion, rather than at
-random. That is, the probability to sample a node was twice as high if any of
-that node's network peers had already been sampled. The parameters of this
-network were $N$ = 5000, $I$ = 2000, $m$ = 2, and $\alpha$ = 0.5.
-
-\subsection*{Investigation of published data}
-
-We applied our kernel-ABC method to several HIV datasets. Because the \gls{BA}
-model generates networks with a single connected component, we specifically
-searched for datasets which originated from existing clusters, either
-phylogenetically or geographically defined. \textcite{wang2015targeting}
-reported on a cohort of acutely infected \gls{MSM} in Beijing, China. Through
-phylogenetic analysis, they found that they had ``deeply sampled'' the
-local \gls{MSM} network, suggesting that many of the samples could be
-epidemiologically related.
-
-We downloaded all sequences associated with each paper from GenBank. Sequences
-were multiply aligned using \software{MUSCLE} version 3.8.31, and alignments
-were manually inspected with \software{Seaview} version 4.4.2. Phylogenies were
-constructed from the nucleotide alignments by approximate maximum likelihood
-using \software{FastTree2} version 2.1.7 with the GTR model. Transmission trees
-were estimated by rooting and time-scaling the phylogenies by root-to-tip
-regression, using a modified version of Path-o-Gen as described
-previously~\autocite{poon2015phylodynamic}.
-
-\section*{Results}
-
-\subsection*{Kernel classifiers}
-
-Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
-tested. For appropriate choices of the kernel meta-parameters, the classifier
-had an $R^2$ above 0.8 for all combinations of tree size and epidemic size. The
-$I$ parameter was also classified accurately, with an average $R^2$ of 0.93 for
-500-tip trees or 0.70 for 100-tip trees. The $m$ parameter was much harder to
-classify, with $R^2$ varying between 0.01 and 0.36 depending on $I$ and the
-tree size. Finally, the accuracy of the classifier for $N$ varied widely, from
-0.08 with 100 tips and $I = 500$, to 0.82 with 1000 tips and $I = 2000$.
-Figure~\ref{fig:crossv} shows the results for $\alpha$; similar figures for the
-other parameters can be found in the supplemental materials. Based on
-inspection of the cross-validation results, we chose to use the meta-parameters
-$\lambda = 0.3$ and $\sigma = 4$ in further analyses, and not to use Sackin's
-index or the \gls{nLTT}.
-
-\begin{figure}
-  \includegraphics{kernel-alpha-crossv}
-  \caption{Cross-validation $R^2$ of kernel-SVM classifier for the $\alpha$
-    parameter of the \gls{BA} network model.}
-  \label{fig:crossv}
-\end{figure}
-
-\section*{Discussion}
-
-The $m$ and $N$ parameters were much harder to recover than $\alpha$ and $I$,
-suggesting differing influences of these parameters on tree shape. The $\alpha$
-parameter in particular was accurately estimated in most cases, implying that
-the strength of preferential attachment has a strong impact on the shape of the
-transmission tree.
-
-Our method has a number of caveats, perhaps the most significant being that it
-takes a transmission tree as input. In reality, true transmission trees are not
-available and must be estimated, often by way of a viral phylogeny. Although
-this has been demonstrated to work in practice~\autocite{leitner1996accurate},
-the topologies of a viral phylogeny and transmission tree can differ
-significantly~\autocite{ypma2013relating}. Additionally, the \gls{BA} model
-that we fitted here makes a number of unrealistic modelling assumptions, such
-as that the network is connected and static, the nodes obey \gls{SI} dynamics,
-and that the underlying behaviour of all nodes is the same. This last is
-particularly problematic, as we showed by simulating a network where some nodes
-exhibited a higher attachment power than others. The estimated attachment power
-was simply the average of the two values, indicating that, although we could
-characterize the network in aggregate, the estimated parameters could not be
-said to apply to any individual node. Finally, the method is computationally
-intensive, taking about a day when run on 20 cores in parallel.
-
-\section*{Conclusions}
-
-We developed a novel method which uses kernel-\gls{ABC} to fit contact network
-models to transmission trees, and used it to fit the \gls{BA} preferential
-attachment model to both simulated and real data.
-
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{S\arabic{figure}}
-
-\printbibliography
-
-\section*{Supplemental Materials}
-
-\begin{figure}
-  \includegraphics{leventhal2012fig1}
-  \caption{Reproduction of Figure 1A from \textcite{leventhal2012inferring}.}
-  \label{fig:sf1}
-\end{figure}
-
-\begin{figure}
-  \includegraphics{smc_test}
-  \caption{Approximation of mixture of Gaussians used by
-    \textcite{del2012adaptive} and \textcite{sisson2007sequential} to test
-    \gls{SMC}.}
-  \label{fig:smctest}
-\end{figure}
-
-\end{document}
+\begin{tabular}{ccccc}
+  Reference & Sequences ($n$) & Location & Risk group & Gene \\
+  \hline
+  \autocite{wang2015targeting} & 173 & Beijing, China & MSM & \textit{pol} \\
+  \autocite{cuevas2009hiv} & 287 & Basque Country, Spain & mixed & \textit{pol} \\
+  \autocite{novitsky2013phylogenetic} & \multirow{2}{*}{180} &
+  \multirow{2}{*}{Mochudi, Botswana} & \multirow{2}{*}{mixed} &
+  \multirow{2}{*}{\textit{env}} \\ \autocite{novitsky2014impact} \\
+  \autocite{li2015hiv} & 280 & Shanghai, China & MSM & \textit{pol} \\
+  \autocite{niculescu2015recent} & 136 & Romaina & IDU & \textit{pol} \\
+  \hline
+\end{tabular}

### 2016-03-17 doc/paper/paper.Rnw
Mostly writing

@@
 \usepackage{multirow}
+\usepackage{cleveref}
+\usepackage{array}
 
@@
 \newacronym{HPD}{HPD}{highest posterior density}
+\newacronym{GTR}{GTR}{general time-reversible}
+\newacronym{MCMC}{MCMC}{Markov chain Monte Carlo}
+\newacronym{MAP}{MAP}{maximum \textit{a posteriori}}
+\newacronym{IDU}{IDU}{injection drug user}
 
-\graphicspath{{figures/}}
+\graphicspath{{../figures/}}
+\newcommand{\tablepath}{../tables}
+\def\figwidth{0.8\textwidth}
 
@@
     library(netabc)
-    library(xtable)
 @
@@
 
-The structure of the contact network underlying a viral epidemic can profoundly
-impact the speed and pattern of the epidemic's spread. Network structure can
-influence the prevalence trajectory~\autocite{o2010contact} and epidemic
+When an infectious disease spreads through a population, transmissions are
+generally more likely to occur between certain pairs of individuals. Such pairs
+must have a particular mode of contact with one another, which varies with the
+mode of transmission of the disease. For airborne pathogens, physical proximity
+may be sufficient, but for sexually transmitted diseases, either sexual or
+blood-to-blood contact is required. The set of links between individuals along
+which transmission can occur is called the contact
+network~\autocite{klovdahl1985social, morris1993epidemiology}. The structure of
+the contact network underlying an epidemic can profoundly impact the speed and
+pattern of the epidemic's expansion. Network structure can influence the
+prevalence trajectory~\autocite{o2010contact} and epidemic
 threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates
@@
 clustering~\autocite{hughes2009molecular}, to the elevated transmission risk
-during acute infection~\autocite{volz2012simple}. On the other hand, inference
+during acute infection~\autocite{volz2012simple}. On the other hand, although
+sophisticated methods have been developed for fitting complex population
+genetic models to phylogenies~\autocite{rasmussen2014phylodynamic}, inference
 of structural network parameters has to date been limited. However, it has been
@@
 \textcite{groendyke2011bayesian} and applied to a much larger data set of 188
-individuals. \textcite{brown2011transmission} fit a preferential attachment
-model to a phylogenetically estimated transmission network relating 60\% of HIV
-infected \gls{MSM} in the United Kingdom. The transmission network is a
-subgraph of the contact network which includes only those edges which have
-already led to a new infection.
+individuals. \textcite{brown2011transmission} analysed the degree distribution
+of an approximate transmission network, estimated based on genetic similarity
+and estimated times of infection, relating 60\% of HIV infected \gls{MSM} in
+the United Kingdom. The transmission network is a subgraph of the contact
+network which includes only those edges which have already led to a new
+infection. The authors found that a Waring distribution, which is produced by a
+different preferential attachment model, was a good fit to their estimated
+network.
 
-Due to their complexity, it is generally difficult to explicitly calculate the
-likelihood of an observed transmission tree under a contact network model. To
-calculate the normalizing constant, we would need to integrate over all
+Standard methods of model fitting involve calculation of the likelihood of
+observed data under the model. In maximum likelihood estimation, a quantity
+proportional to the likelihood is optimized, often through a standard
+multi-dimensional numerical optimization procedure. Bayesian methods integrate
+prior information by optimizing the posterior probability instead. To avoid
+calculation of a normalizing constant, Bayesian inference is often performed
+using \gls{MCMC}. Rather than calculating explicit likeilihoods, \gls{MCMC}
+uses likelihood \emph{ratios} in which the normalizing constant cancel out.
+Unfortunately, it is generally difficult to explicitly calculate the likelihood
+of an observed transmission tree under a contact network model, even up to a
+normalizing constant. To do so, it would be necessary to integrate over all
 possible networks, and also over all possible labellings of the internal nodes
@@
 specific case when $\rho$ is a kernel function, the approach is known as
-kernel-\gls{ABC}.
+kernel-\gls{ABC}~\autocite{nakagome2013kernel, poon2015phylodynamic}.
 
@@
 for simulating epidemics and transmission trees over static contact networks,
-as has been done previously \autocite[\textit{e.g.}][]{o2010contact,
-robinson2013dynamics, leventhal2012inferring, groendyke2011bayesian}. To check
+in the same fashion as several previous studies
+\autocite[\textit{e.g.}][]{o2010contact, robinson2013dynamics,
+leventhal2012inferring, groendyke2011bayesian, goodreau2006assessing}. To check
 that our implementation was correct, we reproduced Figure 1A of
-\textcite{leventhal2012inferring} (see Figure~\ref{fig:sf1}).
+\textcite{leventhal2012inferring} (our \cref{fig:sf1}), which plots the
+unbalancedness of transmission trees simulated over four network models at
+various levels of pathogen transmissibility. Our program is freely available at
+\url{https://github.com/rmcclosk/netabc}.
 
@@
 We used the phylogenetic kernel developed by \textcite{poon2013mapping} to test
-whether the parameters of the \gls{BA} model had an effect on tree shape. The
+whether the parameters of the \gls{BA} model had an effect on tree shape. We
+simulated 100 networks under each of three different values of $\alpha$: 0.5,
+1.0, and 1.5 (300 networks total). The other parameters were fixed to the
+following values: $N$ = 5000, $I$ = 1000, and $m$ = 2. A transmission tree with
+500 tips was simulated over each network (300 transmission trees total). The
+300 trees were compared pairwise with the tree kernel to form a $300 \times
+300$ kernel matrix. The kernel meta-parameters $\lambda$ (the ``decay
+factor''), and $\sigma$ (the ``radial basis function
+variance'')~\autocite[see][]{poon2013mapping}, were set to 0.3 and 4
+respectively. We constructed a \gls{kSVM} classifier for $\alpha$ using the
+\software{kernlab} package~\autocite{karatzoglou2004kernlab}, and evaluated its
+accuracy with 1000 two-fold cross-validations.
+
+Three similar experiments were performed for the other \gls{BA} model
+parameters (one experiment per parameter). $m$ was varied between 2, 3, and 4;
+$I$ between 500, 1000, and 2000; and $N$ between 3000, 5000, and 8000. The
 parameters not being tested were fixed at the values $N$ = 5000, $I$ = 1000,
-$m$ = 2, and $\alpha$ = 1. The parameters were varied one at a time over the
-following values: $N$ = \sett{3000, 5000, 8000}, $I$ = \sett{500, 1000, 2000},
-$m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0.5, 1, 1.5}. For each parameter
-set, 100 networks were generated, and a transmission tree was simulated over
-each (300 trees per parameter set). The 300 trees were compared pairwise with
-the tree kernel to form a $300 \times 300$ kernel matrix. We constructed a
-\gls{kSVM} classifier for the parameter of interest using the
-\software{kernlab} package~\autocite{karatzoglou2004kernlab}, and evaluated its
-accuracy with 1000 two-fold cross-validations. The construction of the kernel
-matrices and classifiers was repeated for several combinations of the kernel
-meta-parameters $\lambda$ (the ``decay factor''), and $\sigma$ (the ``radial
-basis function variance'')~\autocite[see][]{poon2013mapping}, and for
-transmission trees of sizes 100, 500, and 1000. We also tested univariate
-classifier based on Sackin's index~\autocite{shao1990tree} and an ordinary SVM
-based on the \gls{nLTT} statistic~\autocite{janzen2015approximate}.
+$m$ = 2, and $\alpha$ = 1. Thus, we performed a total of four kSVM
+cross-validations, one for each of the \gls{BA} model parameters $\alpha$, $I$,
+$m$, and $N$.
+
+We repeated these four cross-validations with different values of $\lambda$
+(0.2, 0.3, and 0.4) and $\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$), as well
+as on trees with differing numbers of tips (100, 500, and 1000) and in
+epidemics of differing size $I$ (500, 1000, and 2000). When evaluating the
+classifier for $I$, we did not consider trees with 1000 tips, because one of
+the tested $I$ values was 500, and the number of tips cannot be larger than
+$I$.
+
+For each of the four parameters, we also tested univariate classifier based on
+Sackin's index~\autocite{shao1990tree} and an ordinary SVM based on the
+\gls{nLTT} statistic~\autocite{janzen2015approximate}.
 
@@
 We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
-\textcite{del2012adaptive}. To ensure our implementation was correct, we
+\textcite{del2012adaptive}. To check that our implementation was correct, we
 applied it to the same mixture of Gaussians used by
@@
 by~\textcite{sisson2007sequential}). We were able to obtain a close
-approximation to the function (see Figure~\ref{fig:smctest}), and attained the
+approximation to the function (see \cref{fig:smctest}), and attained the
 stopping condition used by the authors in a comparable number of steps.
@@
 
-We applied our kernel-ABC method to several HIV datasets. Because the \gls{BA}
-model generates networks with a single connected component, we specifically
-searched for datasets which originated from existing clusters, either
-phylogenetically or geographically defined. Characteristics of the datasets we
-investigated are given in Table~\ref{tab:data}.
+We applied our kernel-ABC method to several published HIV datasets. Because the
+\gls{BA} model generates networks with a single connected component, we
+specifically searched for datasets which originated from existing clusters,
+either phylogenetically or geographically defined. Characteristics of the
+datasets we investigated are given in \cref{tab:data}.
 
-\begin{table}
+\begin{table}[ht]
   \centering
-  \begin{tabular}{ccccc}
-    Reference & Sequences ($n$) & Location & Risk group & Gene \\
-    \hline
-    \autocite{wang2015targeting} & 173 & Beijing, China & MSM & \textit{pol} \\
-    \autocite{cuevas2009hiv} & 287 & Basque Country, Spain & mixed & \textit{pol} \\
-    \autocite{novitsky2013phylogenetic} & \multirow{2}{*}{201^*} &
-    \multirow{2}{*}{Mochudi, Botswana} & \multirow{2}{*}{mixed} &
-    \multirow{2}{*}{\textit{env}} \\ \autocite{novitsky2014impact} \\
-    \hline
-  \end{tabular}
+  \input{\tablepath/realdata}
   \caption{Characteristics of published datasets investigated with kernel-ABC.
-  Acronyms: MSM, men who have sex with men. $^*$ Sub-sampled from 1299.}
+  Acronyms: MSM, men who have sex with men; IDU, injection drug users.}
   \label{tab:data}
@@
 
-We downloaded all sequences associated with each study from GenBank. Sequences
-were multiply aligned using \software{MUSCLE} version 3.8.31
-\autocite{edgar2004muscle}, and alignments were manually inspected with
-\software{Seaview} version 4.4.2 \autocite{gouy2010seaview}. Phylogenies were
-constructed from the nucleotide alignments by approximate maximum likelihood
-using \software{FastTree2} version 2.1.7 with the GTR model. Transmission trees
-were estimated by rooting and time-scaling the phylogenies by root-to-tip
-regression, using a modified version of Path-o-Gen as described
+We downloaded all sequences associated with each study from GenBank. For the
+\textcite{novitsky2014impact} data, each sequence was aligned pairwise to the
+HXB2 reference sequence (Genbank accession number HIVHXB2CG) and the
+hypervariable regions were clipped out with BioPython version
+1.66+~\autocite{cock2009biopython}. Sequences were multiply aligned using
+\software{MUSCLE} version 3.8.31 \autocite{edgar2004muscle}, and alignments
+were manually inspected with \software{Seaview} version 4.4.2
+\autocite{gouy2010seaview}. Phylogenies were constructed from the nucleotide
+alignments by approximate maximum likelihood using \software{FastTree2} version
+2.1.7 with the \gls{GTR} model. Transmission trees were estimated by rooting
+and time-scaling the phylogenies by root-to-tip regression, using a modified
+version of Path-O-Gen (distributed as part of
+BEAST~\autocite{drummond2007beast}) as described
 previously~\autocite{poon2015phylodynamic}.
+
+Two of the datasets \autocite{li2015hiv,novitsky2014impact} were initially much
+larger than the others, containing 1265 and 1299 sequences respectively. To
+ensure that the analyses were comparable, we reduced these to a number of
+sequences similar to the smaller datasets. For the \textcite{li2015hiv} data,
+we detected a cluster of size 280 using a patristic distance cutoff of 0.02 as
+described previously~\autocite{poon2014impact}. Only sequences within this
+cluster were carried forward. For the \textcite{novitsky2014impact} data, no
+large clusters were detected using the same cutoff, so we analysed a subtree of
+size 180 chosen arbitrarily.
 
@@
 <<classifiers, include=FALSE>>=
-    get.kernel.data <- function (param) {
-        md <- collect.metadata(paste0('../../simulations/kernel-', param, '/classifier/*'))
+    get.kernel.data <- function (param, step) {
+        md <- collect.metadata(paste0('../../simulations/kernel-', param, '/', step, '/*'))
+        if ('rbf_variance' %in% colnames(md)) {
           md <- subset(md, rbf_variance == 4 & decay_factor == 0.3)
+        }
         if (param == 'alpha') {
@@
     }
-    kernel.alpha <- get.kernel.data('alpha')
-    kernel.m <- get.kernel.data('m')
-    kernel.I <- get.kernel.data('I')
-    kernel.N <- get.kernel.data('N')
+    kernel.alpha <- get.kernel.data('alpha' ,'classifier')
+    kernel.m <- get.kernel.data('m', 'classifier')
+    kernel.I <- get.kernel.data('I', 'classifier')
+    kernel.N <- get.kernel.data('N', 'classifier')
+
+    sackin.alpha <- get.kernel.data('alpha', 'stats-classifier')
+    sackin.m <- get.kernel.data('m', 'stats-classifier')
+    sackin.I <- get.kernel.data('I', 'stats-classifier')
+    sackin.N <- get.kernel.data('N', 'stats-classifier')
+
+    nltt.alpha <- get.kernel.data('alpha', 'nltt-classifier')
+    nltt.m <- get.kernel.data('m', 'nltt-classifier')
+    nltt.I <- get.kernel.data('I', 'nltt-classifier')
+    nltt.N <- get.kernel.data('N', 'nltt-classifier')
 @
@@
 Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
-tested. For appropriate choices of the kernel meta-parameters, the classifier
-had an $R^2$ above 
-    \Sexpr{kernel.alpha[,min(rsquared %/% 0.05 * 0.05)]} 
-for all combinations of tree size and epidemic size. The $I$ parameter was also
-classified accurately, with an average $R^2$ of 
+tested~\cref{fig:rsquared}. Classifiers based on two other tree statistics, the
+\gls{nLTT} and Sackin's index, generally exhibited worse performance than the
+tree kernel, although the magnitude of the disparity varied between the
+parameters. \Cref{fig:rsquared} shows the cross-validation accuracy with
+$\lambda$ = 0.3, $\sigma$ = 4, $I$ = 1000, and trees of size 500. The results
+were largely robust to variations in $\lambda$ and $\sigma$, although accuracy
+varied between different epidemic and sampling scenarios
+(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}).
+
+The kSVM classifier for $\alpha$ had an average $R^2$ of
+    \Sexpr{kernel.alpha[,round(mean(rsquared), 2)]} 
+(\cref{fig:alphacrossv}). The $R^2$ of the for the SVM classifer for $\alpha$
+using the \gls{nLTT} varied between
+\Sexpr{nltt.alpha[,round(min(rsquared), 2)]}
+and
+  \Sexpr{nltt.alpha[,round(max(rsquared), 2)]}.
+The $R^2$ of the linear regression against Sackin's index varied between
+  \Sexpr{sackin.alpha[,round(min(rsquared), 2)]}
+and
+  \Sexpr{sackin.alpha[,round(max(rsquared), 2)]}.
+The $I$ parameter was also classified accurately by the tree kernel
+(\cref{fig:Icrossv}), with an average
+$R^2$ of 
   \Sexpr{kernel.I[ntip == 500, round(mean(rsquared), 2)]} 
@@
   \Sexpr{kernel.I[ntip == 100, round(mean(rsquared), 2)]} 
-for 100-tip trees. The $m$ parameter was much harder to classify, with $R^2$
-varying between 
+for 100-tip trees. The \gls{nLTT} performed almost as well on this parameter as
+the tree kernel, with $R^2$ values of
+  \Sexpr{nltt.I[ntip == 500, round(mean(rsquared), 2)]}
+for 500-tip trees or
+  \Sexpr{nltt.I[ntip == 100, round(mean(rsquared), 2)]}
+for 100-tip trees. Sackin's index was an extremely poor classifier for this
+parameter, with both $R^2$ values below
+  \Sexpr{sackin.I[,round(max(rsquared) + 0.01, 2)]}.
+
+The $m$ parameter was much harder to classify (\cref{fig:mcrossv}), with $R^2$
+for the kSVM varying between 
   \Sexpr{kernel.m[,round(min(rsquared), 3)]}
@@
   \Sexpr{kernel.m[,round(max(rsquared), 2)]}
-depending on $I$ and the tree size. Finally, the accuracy of the classifier for
-$N$ varied widely, from 
+depending on $I$ and the tree size. The \gls{nLTT} was also universally poor,
+with $R^2$ below
+  \Sexpr{nltt.m[,round(max(rsquared) + 0.01, 2)]}
+for all scenarios. Sackin's index fared slightly better than the \gls{nLTT},
+especially on trees with a higher number of tips, with $R^2$ between
+  \Sexpr{sackin.m[,round(min(rsquared), 2)]}
+and
+  \Sexpr{sackin.m[,round(max(rsquared), 2)]}.
+Finally, the accuracy of the kSVM classifier for $N$ varied widely
+(\cref{fig:Ncrossv}), from 
   \Sexpr{kernel.N[,round(min(rsquared), 2)]}
@@
   \Sexpr{kernel.N[,round(max(rsquared), 2)]}.
-Figure~\ref{fig:crossv} shows the results for $\alpha$; similar figures for the
-other parameters can be found in the supplemental materials. Based on
-inspection of the cross-validation results, we chose to use the meta-parameters
-$\lambda = 0.3$ and $\sigma = 4$ in further analyses, and not to use Sackin's
-index or the \gls{nLTT}.
+The \gls{nLTT} performed well in classifying $N$ when the epidemic was large,
+improving upon the tree kernel in the scenarios with $I$ = 2000 and 100- or
+500-tip transmission trees ($R^2$ = 
+  \Sexpr{nltt.N[which(I == 2000 & ntip == 100), round(mean(rsquared), 2)]}
+and
+  \Sexpr{nltt.N[which(I == 2000 & ntip == 500), round(mean(rsquared), 2)]}),
+although it performed poorly when $I$ was 500 or 1000 (average $R^2$ =
+  \Sexpr{nltt.N[which(I < 2000), round(mean(rsquared), 2)]}).
+Sackin's index could not classify $N$, with all $R^2$ below
+  \Sexpr{sackin.N[,round(max(rsquared) + 0.01, 2)]}.
+Based on inspection of the cross-validation results, we chose to use the
+meta-parameters $\lambda = 0.3$ and $\sigma = 4$ in further analyses, and not
+to use Sackin's index or the \gls{nLTT}.
 
-\begin{figure}
-  \includegraphics{kernel-alpha-crossv}
-  \caption{Cross-validation $R^2$ of kernel-SVM classifier for the $\alpha$
-    parameter of the \gls{BA} network model.}
-  \label{fig:crossv}
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
+  \caption{
+      Cross-validation accuracy of kernel-SVM classifier (left), SVM classifier
+      using \gls{nLTT} (centre), and linear regression using Sackin's index
+      (right) for \gls{BA} model parameters. Kernel meta-parameters were set to
+      $\lambda = 0.3$ and $\sigma = 4$. Each point was calculated based on 300
+      simulated transmission trees over networks with three different values of
+      the parameter being tested. Vertical lines are empirical 95\% confidence
+      intervals based on 1000 2-fold cross-validations.
+  }
+  \label{fig:rsquared}
 \end{figure}
@@
 
-Across replicates and other model parameters, the mean absolute difference
-between the estimated and true values of $\alpha$ was X, with a standard error
-of X.
+<<point_est, include=FALSE>>=
+    f <- Sys.glob("../../simulations/abc-pa-free-m/point-estimate/*")
+    d <- fread(f)
+    d[,m := floor(m)]
+    d[,alpha_error := abs(true_alpha - alpha)]
+    d[,N_error := abs(true_N - N)]
+    d[,I_error := abs(true_I - I)]
+    d[,m_error := abs(true_m - m)]
 
-Figure~\ref{fig:abcpt} shows point estimates for the simulations with $m = 2$.
-The results for the other values of $m$ are similar and can be found in the
-supplemental materials.
+    alpha.av <- anova(lm(alpha_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
+    m.av <- anova(lm(m_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
+    N.av <- anova(lm(N_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
+    I.av <- anova(lm(I_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
 
-The marginal approximated posterior distributions for one simulated dataset are
-shown in Figure~\ref{fig:abcex}.
+    m.tbl <- d[,prop.table(table(m_error))]
+    zero.alpha.test <- wilcox.test(d[true_alpha == 0, alpha_error], 
+                                   d[true_alpha != 0, alpha_error])
+    alpha.I.test <- wilcox.test(d[true_alpha < 1, I_error],
+                                d[true_alpha >= 1, I_error])
+    I.I.test <- wilcox.test(d[true_I == 1000, I_error],
+                            d[true_I == 2000, I_error])
+    alpha.m.test <- wilcox.test(d[true_alpha == 0 | true_alpha == 1, m_error],
+                                d[true_alpha == 0.5 | true_alpha == 1.5, m_error])
 
-\begin{figure}
-    \includegraphics{abc-point-est}
-    \caption{Point estimates of \gls{BA} model parameters obtained by running
-    kernel-ABC on simulated phylogenies. Dotted lines indicate the true values,
-    and limits of the $y$-axes are the regions of uniform prior density.}
+    options(scipen=-1, digits=2)
+@
+
+Of the four parameters, $\alpha$ was the most accurately estimated, with a
+median [IQR] absolute error of 
+    \Sexpr{d[,median(alpha_error)]} 
+    [\Sexpr{d[,quantile(alpha_error, 0.25)]}-\Sexpr{d[,quantile(alpha_error, 0.75)]}].
+\Cref{fig:abcpt} shows point estimates for all simulations with $m = 2$. The
+results $m = 3$ and $m = 4$ were similar (\cref{fig:abcptm3,fig:abcptm4}).
+Average boundaries of 95\% \gls{HPD} intervals are given in \cref{tab:abchpd}.
+The accuracy of the estimates was not significantly different between values of
+$m$ or $I$ (both one-way ANOVA,
+    $p$ = \Sexpr{alpha.av["factor(true_m)", "Pr(>F)"]}
+and 
+    \Sexpr{alpha.av["factor(true_I)", "Pr(>F)"]}),
+although the errors when the true value of $\alpha$ was zero were
+significantly greater than the other values 
+    (Wilcoxon rank-sum test, $p$ = \Sexpr{zero.alpha.test$p.value}).
+The error in the estimated value of $I$ was
+    \Sexpr{d[,median(I_error)]} 
+    [\Sexpr{d[,quantile(I_error, 0.25)]}-\Sexpr{d[,quantile(I_error, 0.75)]}].
+Errors were significantly higher for $\alpha \geq 1$
+    (Wilcoxon rank-sum test, $p$ = \Sexpr{alpha.I.test$p.value})
+and for $I$ = 2000
+    ($p$ = \Sexpr{I.I.test$p.value}),
+but not for any values of $m$
+    (one-way ANOVA, $p$ = \Sexpr{I.av["factor(true_m)", "Pr(>F)"]}).
+The $m$ parameter was estimated correctly in
+    \Sexpr{as.integer(m.tbl[1] * 100)} \%
+of simulations, with an error of one in
+    \Sexpr{as.integer(m.tbl[2] * 100)} \%
+and of two or more in 
+    \Sexpr{as.integer(sum(m.tbl[3:length(m.tbl)]) * 100)} \%
+(the only possible $m$ values were 1, 2, 3, 4, or 5). The true values of $m$
+and $I$ did not significantly affect the error
+    (one-way ANOVA, $p$ = \Sexpr{m.av["factor(true_m)", "Pr(>F)"]} and
+                          \Sexpr{m.av["factor(true_I)", "Pr(>F)"]}).
+Finally, the total number of nodes $N$ was consistently over-estimated by about
+a factor of two
+    (error \Sexpr{d[,median(N_error)]} 
+    [\Sexpr{d[,quantile(N_error, 0.25)]}-\Sexpr{d[,quantile(N_error, 0.75)]}]).
+No other parameters influenced the accuracy of the $N$ estimates 
+    (one-way ANOVA, $p \geq$ \Sexpr{min(N.av[1:3,"Pr(>F)"])}).
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth]{abc-point-estimate-m2.pdf}
+  \caption{
+    Point estimates of \gls{BA} model parameters obtained by running kernel-ABC
+    on simulated phylogenies without training, for simulations with $m = 2$.
+    Dotted lines indicate true values, and limits of the $y$-axes are regions
+    of uniform prior density.
+  }
   \label{fig:abcpt}
@@
 
-\begin{figure}
-    \includegraphics{abc-example}
+\begin{table}[ht]
+  \centering
+  \input{\tablepath/abc-hpd}
+  \caption{Average \gls{HPD} interval widths for \gls{ABC} model parameter
+           estimates.}
+  \label{tab:abchpd}
+\end{table}
+
+The dispersion of the \gls{ABC} approximation to the posterior also varied
+between the parameters, with narrower \gls{HPD} intervals for those with the
+most accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows the
+distributions for for one simulation (equivalent plots for all the simulations
+can be found in supplemental data). \Gls{HPD} intervals around $\alpha$ and $I$
+were narrow relative to the region of nonzero prior density, whereas the
+intervals for $m$ and $N$ were more widely dispersed.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth]{abc-posterior-example.pdf}
   \caption{Marginal estimated posterior distributions obtained with ABC on a
@@
 
+<<mixed_alpha, include=FALSE>>=
+  library(coda)
+  d <- fread("bzcat ../../simulations/abc-pa-mixed-alpha/abc/*")
+  d <- d[iter == max(iter)]
+  d <- d[sample(1:nrow(d), prob=weight, replace=TRUE)]
+  params <- c("alpha", "I", "m", "N")
+  setnames(d, c("attach_power", "edges_per_vertex", "nodes", "sim_nodes"),
+           params)
+  d <- melt(d, measure.vars=params, variable.name="parameter")
+  f <- function (x) {
+    dens <- density(x)
+    hpd <- HPDinterval(mcmc(x))
+    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
+         hpd.max=hpd[,"upper"])
+  }
+  d <- d[,f(value), by=parameter]
+  setkey(d, parameter)
+@
+
+To test the effect of model misspecification, we simulated one network where
+the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
+other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The \gls{MAP} [95\%
+HPD] estimates for each parameter were: 
+$\alpha$, 
+  \Sexpr{d["alpha", map]} [\Sexpr{d["alpha", hpd.min]}-
+                           \Sexpr{d["alpha", hpd.max]}];
+$I$,
+  \Sexpr{d["I", map]} [\Sexpr{d["I", hpd.min]}-
+                       \Sexpr{d["I", hpd.max]}];
+$m$,
+  \Sexpr{d["m", floor(map)]} [\Sexpr{d["m", floor(hpd.min)]}-
+                              \Sexpr{d["m", floor(hpd.max)]}];
+$N$,
+  \Sexpr{d["N", map]} [\Sexpr{d["N", hpd.min]}-
+                       \Sexpr{d["N", hpd.max]}].
+
+<<peerdriven, include=FALSE>>=
+  d <- fread("bzcat ../../simulations/abc-pa-peerdriven/abc/*")
+  d <- d[iter == max(iter)]
+  d <- d[sample(1:nrow(d), prob=weight, replace=TRUE)]
+  params <- c("alpha", "I", "m", "N")
+  setnames(d, c("attach_power", "edges_per_vertex", "nodes", "sim_nodes"),
+           params)
+  d <- melt(d, measure.vars=params, variable.name="parameter")
+  f <- function (x) {
+    dens <- density(x)
+    hpd <- HPDinterval(mcmc(x))
+    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
+         hpd.max=hpd[,"upper"])
+  }
+  d <- d[,f(value), by=parameter]
+  setkey(d, parameter)
+@
+
+To test the effect of sampling bias, we sampled one transmission tree in a
+peer-driven fashion, where the probability to sample a node was twice as high
+if one of it's peers had already been sampled. The parameters for this
+experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and $I$ = 2000. The
+estimated values were
+$\alpha$, 
+  \Sexpr{d["alpha", map]} [\Sexpr{d["alpha", hpd.min]}-
+                           \Sexpr{d["alpha", hpd.max]}];
+$I$,
+  \Sexpr{d["I", map]} [\Sexpr{d["I", hpd.min]}-
+                       \Sexpr{d["I", hpd.max]}];
+$m$,
+  \Sexpr{d["m", floor(map)]} [\Sexpr{d["m", floor(hpd.min)]}-
+                              \Sexpr{d["m", floor(hpd.max)]}];
+$N$,
+  \Sexpr{d["N", map]} [\Sexpr{d["N", hpd.min]}-
+                       \Sexpr{d["N", hpd.max]}].
+
 \subsection*{Real data}
+
+<<realdata, include=FALSE>>=
+  d <- fread(Sys.glob("../../simulations/aggregates/hpd/*"))
+  map <- dcast.data.table(d, dataset~parameter, value.var="map")
+  up <- dcast.data.table(d, dataset~parameter, value.var="hpd.max")
+  low <- dcast.data.table(d, dataset~parameter, value.var="hpd.min")
+  setkey(map, dataset)
+  setkey(up, dataset)
+  setkey(low, dataset)
+@
+
+There was substantial heterogeneity among the parameter estimates for the five
+published HIV datasets we analysed. Two of the
+datasets~\autocite{niculescu2015recent, wang2015targeting} had estimated
+$\alpha$ values near unity (\gls{MAP} estimate [95\% \gls{HPD}] 
+  \Sexpr{map["niculescu2015", alpha]} 
+  [\Sexpr{low["niculescu2015", alpha]}-\Sexpr{up["niculescu2015", alpha]}]
+and
+  \Sexpr{map["wang2015", alpha]} 
+  [\Sexpr{low["wang2015", alpha]}-\Sexpr{up["wang2015", alpha]}]). Another two
+datasets~\autocite{li2015hiv, cuevas2009hiv} had lower estimated values and wider
+\gls{HPD} intervals
+  (\Sexpr{map["li2015", alpha]} 
+  [\Sexpr{low["li2015", alpha]}-\Sexpr{up["li2015", alpha]}]
+and
+  \Sexpr{map["cuevas2009", alpha]} 
+  [\Sexpr{low["cuevas2009", alpha]}-\Sexpr{up["cuevas2009", alpha]}]). 
+The \textcite{novitsky2014impact} data had an extremely low estimated $\alpha$
+and a very wide \gls{HPD} interval
+  \Sexpr{map["novitsky2014", alpha]} 
+  [\Sexpr{low["novitsky2014", alpha]}-\Sexpr{up["novitsky2014", alpha]}]). 
+For all the datasets except \textcite{novitsky2014impact}, estimated values of
+$I$ were below 2000, with narrow \gls{HPD} intervals around two of the
+datasets
+  (\textcite{cuevas2009hiv}, \Sexpr{map["cuevas2009", I]} 
+  [\Sexpr{low["cuevas2009", I]}-\Sexpr{up["cuevas2009", I]}];
+   \textcite{niculescu2015recent}, \Sexpr{map["niculescu2015", I]}
+  [\Sexpr{low["niculescu2015", I]}-\Sexpr{up["niculescu2015", I]}])
+and wider intervals around the other two
+  (\textcite{li2015hiv}, \Sexpr{map["li2015", I]} 
+  [\Sexpr{low["li2015", I]}-\Sexpr{up["li2015", I]}];
+   \textcite{wang2015targeting}, \Sexpr{map["wang2015", I]}
+  [\Sexpr{low["wang2015", I]}-\Sexpr{up["wang2015", I]}]).
+The \textcite{novitsky2014impact} data was again the outlier, with a very high
+estimated $I$, and \gls{HPD} interval spanning almost the entire prior region
+  (\Sexpr{map["novitsky2014", I]} 
+  [\Sexpr{low["novitsky2014", I]}-\Sexpr{up["novitsky2014", I]}]).
+Little information was gleaned about the $m$ parameter, with the \gls{HPD}
+interval occuping the entire prior region for all datasets. The estimates of
+$N$ were similarly uninformative, with the exception that the point estimate
+for the \textcite{wang2015targeting} data was smaller
+  (\Sexpr{map["wang2015", N]})
+than the estimates for other datasets.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{realdata-hpd.pdf}
+  \caption{
+    Point estimates and 95\% \gls{HPD} intervals for \gls{BA} network
+    parameters of published HIV datasets, obtained with kernel-ABC.
+  }
+\end{figure}
 
@@
 
-% kernel classifiers
+Contact networks can have a strong influence on epidemic progression, and are
+potentially useful as a public health tool. Despite this, few methods exist for
+investigating contact network parameters in a phylodynamic framework.
+Kernel-ABC is a model-agnostic method which can be used to investigate any
+quantity that affects tree shape. In this work, we developed a kernel-ABC-based
+method to infer the parameters of a contact network model. The method is
+general, and could be applied to any model from which contact networks can be
+simulated. We demonstrated the method on the \gls{BA} model, which is a simple 
+model incorprorating the preferential attachment feature commonly seen in
+real-world networks.
 
-The $m$ and $N$ parameters were much harder to recover than $\alpha$ and $I$,
-suggesting differing influences of these parameters on tree shape. The $\alpha$
-parameter in particular was accurately estimated in most cases, implying that
-the strength of preferential attachment has a strong impact on the shape of the
-transmission tree.
+By training a kernel-SVM classifier, we found that the $\alpha$ and $I$
+parameters, representing preferential attachment power and number of infected
+nodes, had a strong influence on tree shape. This was reflected in the relative
+accuracy of the kernel-ABC estimates of these parameters. On the other hand,
+the $m$ and $N$ parameters did not produce much variation in tree shape,
+resulting in both poorly performing classifiers and uninformative kernel-ABC
+estimates. The total number of nodes $N$ was almost always significantly
+over-estimated. Since the prior on $N$ and $I$ is jointly uniform on a
+triangular region ($I \leq N$), there is more prior mass on high $N$ values. In
+retrospect, it is unreasonable to expect good estimation of $N$, because adding
+more nodes to a \gls{BA} network does not change the edge density or overall
+shape. This can be illustrated by imaging that we add a small number of nodes
+to a network after the epidemic simulation has already been completed. It is
+possible that none of these new nodes attains a connection to any infected
+node. Thus, running the simulation again on the new, larger network could
+produce the exact same transmission tree as before.
 
-% abc simulations
+As noted by \textcite{lintusaari2016identifiability}, uniform priors on model
+parameters may translate to highly informative priors on quantities of
+interest. We observed an exponential relationship between the preferential
+attachment power $\alpha$ and the power-law exponent $\gamma$
+(\cref{fig:gamma}). Therefore, placing a uniform prior on $\alpha$ between 0
+and 2 is equivalent to placing an informative prior that $\gamma$ is close to
+2. Therefore, if we were primarily interested in $\gamma$ rather than $\alpha$,
+a more sensible choice of prior might have a shape similar to \cref{fig:gamma}
+and be bounded above by approximately $\alpha$ = 1.5. This would uniformly
+bound $\gamma$ in the region $2 \leq \gamma \leq 4$ commonly reported in the
+network literature~\autocite{liljeros2001web, schneeberger2004scale,
+colgate1989risk, brown2011transmission}. We note however that
+\textcite{jones2003assessment} estimated $\gamma$ values greater than four, in
+one case as high as 17, for some datasets, indicating that a wider range of
+permitted $\gamma$ values may be warranted.
 
-The total number of nodes $N$ was almost always significantly over-estimated.
-This suggests that $N$ has very little effect on tree shape, and the estimates
-were influenced primarily by the prior. Since the prior on $N$ and $I$ is
-jointly uniform on a triangular region, there is more prior mass on high $N$
-values. In retrospect, this is reasonable because adding more nodes to a
-\gls{BA} network does not change the edge density or the network's overall
-shape. This can illustrated by imagining that we grow the network after the
-epidemic simulation has already been completed. It is possible (though perhaps
-unlikely) that none of the new nodes attains a connection to any infected node.
-Thus, running the simulation again on the new, larger network could produce the
-exact same transmission tree as before. This is an extreme example, but it
-serves to demonstrate the point that $N$ may be impossible to determine if the
-epidemic does not come close to saturating the network.
+Our investigation of published HIV datasets indicated heterogeneity in the
+contact network structures underlying several distinct local epidemics.
+\textcite{cuevas2009hiv} studied a group of newly diagnosed individuals in the
+Basque Country, Spain. Although the individuals were of mixed risk group, and
+therefore unlikely to comprise a single contact network, a high proportion of
+them (47\%) grouped into local transmission clusters, allowing for the
+possibility of weak preferential attachment dynamics. We estimated a relatively
+low attachment power for these data, which is consistent with the sampled
+sequences comprising many distinct sub-networks. \textcite{li2015hiv} sampled a
+large number of acutely infected \gls{MSM} in Shanghai, China. By using a
+patristic distance cutoff~\autocite{poon2014impact}, we identified and analysed
+a large cluster. Surprisingly, the estimated attachment power was fairly low
+with a large credible interval.
 
-% real data
+\textcite{niculescu2015recent} studied a recent outbreak among Romainian
+\gls{IDU}. The estimated attachment power for this dataset was the highest
+among all we considered, at slightly above one. The dataset constituted a
+fairly high sampling prevalence of 33\% of newly diagnosed \gls{IDU} infections
+over the study period. \textcite{wang2015targeting} sampled acutely infected
+\gls{MSM} in Beijing, China, and discovered a high degree of clustering which
+indicated that the local \gls{MSM} subnetwork had been deeply sampled. In both
+these datasets, the estimated attachment power was close to one.
 
-We estimated the parameters of the \gls{BA} model for several published
-datasets (see~\ref{tab:data}). \textcite{cuevas2009hiv} performed a
-phylogenetic analysis of 261 patients of mixed risk groups in the Basque
-Country, Spain, who were newly diagnosed between 2004 and 2007. Although their
-data did not comprise a single cluster, they did find that almost half (47\%)
-of the samples grouped in transmission clusters. \textcite{wang2015targeting}
-reported on a cohort of acutely infected \gls{MSM} in Beijing, China. Through
-phylogenetic analysis, they found that they had ``deeply sampled'' the local
-\gls{MSM} network, suggesting that many of the samples could be
-epidemiologically related. \textcite{novitsky2013phylogenetic} investigated a
-local HIV epidemic in Mochudi, Botswana. They later extended their
-analysis~\autocite{novitsky2014impact} to include an estimated 70\% of HIV
-infected persons in the town.
+\textcite{novitsky2013phylogenetic} sampled approximately 44\% of the
+HIV-infected individuals in the northern area of Mochudi, Botswana. Additional
+sampling in a later study~\autocite{novitsky2014impact} brought the genotyping
+coverage up to 70\%. Even with such a high sampling coverage, we did not detect
+any large clusters using patristic distance, and therefore chose to analyze a
+subtree instead. The estimated attachment power was extremely low, with a very
+wide credible interval. The estimated number of infected nodes was also
+extremely high, much higher than the HIV-infected population of the town, with
+a similarly wide credible interval. Several factors may have contributed to
+this result. First, the authors note that the their sample was 75\% female. In
+a primarily heterosexual risk environmet, removal of a disproportionate nmuber
+of males from the network could obfuscate the true network structure. Second,
+the town in question was in close proximity to the country's capital, and the
+authors indicated that a high amount of migration takes place between the two
+locations. This suggests a reason for the high estimate of $I$: the contact
+network includes a much larger group based in the capital city.
 
-Our method has a number of caveats, perhaps the most significant being that it
+When interpreting these results, we caution that the \gls{BA} model is quite
+simple and most likely misspecified for these data. In particular, the average
+degree of a node in the network is equal to $2m$, and therefore is constrained
+to be a multiple of 2. Furthermore, we considered the case $m = 1$, where the
+network has no cycles, to be implausable and assigned it zero prior
+probability. This forces the average degree to be at least four, which may be
+unrealistically high for sexual networks. Additional modelling assumptions,
+include the network being connected and static, the epidemic following simple
+\gls{SI} dynamics, and the underlying behaviour of all nodes is the same. This
+last is particularly problematic, as we showed by simulating a network where
+some nodes exhibited a higher attachment power than others. The estimated
+attachment power was simply the average of the two values, indicating that,
+although we could characterize the network in aggregate, the estimated
+parameters could not be said to apply to any individual node. However, we note
+that, given the model-agnostic nature of our method, it is possible to fit a
+model with heterogeneous node behaviour, and indeed this may prove a fruitful
+avenue for future investigations.
+
+Our method itself has of caveats, perhaps the most significant being that it
 takes a transmission tree as input. In reality, true transmission trees are not
-available and must be estimated, often by way of a viral phylogeny. Although
-this has been demonstrated to work in practice~\autocite{leitner1996accurate},
-the topologies of a viral phylogeny and transmission tree can differ
-significantly~\autocite{ypma2013relating}. Additionally, the \gls{BA} model
-that we fitted here makes a number of unrealistic modelling assumptions, such
-as that the network is connected and static, the nodes obey \gls{SI} dynamics,
-and that the underlying behaviour of all nodes is the same. This last is
-particularly problematic, as we showed by simulating a network where some nodes
-exhibited a higher attachment power than others. The estimated attachment power
-was simply the average of the two values, indicating that, although we could
-characterize the network in aggregate, the estimated parameters could not be
-said to apply to any individual node. Finally, the method is computationally
-intensive, taking about a day when run on 20 cores in parallel.
-
-\section*{Conclusions}
-
-We developed a novel method which uses kernel-\gls{ABC} to fit contact network
-models to transmission trees, and used it to fit the \gls{BA} preferential
-attachment model to both simulated and real data.
+available and must be approximated, often by way of a viral phylogeny. Although
+this has been demonstrated to be a fair
+approximation~\autocite[e.g.][]{leitner1996accurate}, and is frequently used in
+practice~\autocite[e.g.][]{stadler2013uncovering}, the topologies of a viral
+phylogeny and transmission tree can differ
+significantly~\autocite{ypma2013relating}. In addition, it is computationally
+intensive, taking about a day when run on 20 cores in parallel with the setings
+we described in the methods. However, there are no other methods available to
+perform the types of estimation we do here, and we therefore believe our method
+will be useful to epidemiological researchers interested in the network
+structure underlying disease outbreaks.
 
@@
 \newpage
-
 \section*{Supplemental Materials}
@@
 \begin{figure}[ht]
-  \includegraphics[scale=0.8]{leventhal2012fig1}
+  \centering
+  \includegraphics{leventhal2012fig1.pdf}
   \caption{Reproduction of Figure 1A from \textcite{leventhal2012inferring}.}
@@
 
+\begin{figure}[ht]
+  \centering
+  \includegraphics{kernel-alpha-crossv.pdf}
+  \caption{
+    Cross-validation accuracy of kernel-SVM classifiers for $\alpha$ parameter
+    of \gls{BA} network model, for various tree kernel meta-parameters and
+    epidemic scenarios. Each point was calculated based on 300 simulated
+    transmission trees over networks with $\alpha$ = 0.5, 1.0, or 1.5. Dotted
+    and and dashed lines indicate, respectively, performance of SVM using the
+    \gls{nLTT} statistic, and linear regression using Sackin's index. Facets
+    are number of infected nodes before the simulation was stopped ($I$) and
+    number of tips in the sampled transmission tree.
+  }
+  \label{fig:alphacrossv}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{kernel-m-crossv.pdf}
+  \caption{Cross-validation accuracy of kernel-SVM classifiers for $m$
+      parameter of \gls{BA} network model, for various tree kernel
+      meta-parameters and epidemic scenarios. Each point was calculated based
+      on 300 simulated transmission trees over networks with $m$ = 2, 3, or 4.
+      Dotted and and dashed lines indicate, respectively, performance of SVM
+      using the \gls{nLTT} statistic, and linear regression using Sackin's
+      index. Facets are number of infected nodes before the simulation was
+      stopped ($I$) and number of tips in the sampled transmission tree.
+  }
+  \label{fig:mcrossv}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{kernel-I-crossv.pdf}
+  \caption{Cross-validation accuracy of kernel-SVM classifiers for number of
+      infected nodes ($I$) under \gls{BA} network model, for various tree
+      kernel meta-parameters and two tree sizes. Each point was calculated
+      based on 300 simulated transmission trees over networks with $I$ = 500,
+      1000, or 2000. Dotted and and dashed lines indicate, respectively,
+      performance of SVM using the \gls{nLTT} statistic, and linear regression
+      using Sackin's index. Facets are the number of tips in the sampled
+      transmission tree.
+  }
+  \label{fig:Icrossv}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{kernel-N-crossv.pdf}
+  \caption{Cross-validation accuracy of kernel-SVM classifiers for total number
+      of nodes ($N$) under \gls{BA} network model, for various tree kernel
+      meta-parameters and epidemic scenarios sizes. Each point was calculated
+      based on 300 simulated transmission trees over networks with $N$ = 3000,
+      5000, or 8000. Dotted and and dashed lines indicate, respectively,
+      performance of SVM using the \gls{nLTT} statistic, and linear regression
+      using Sackin's index. Facets are the number of tips in the sampled
+      transmission tree.
+  }
+  \label{fig:Ncrossv}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{abc-point-estimate-m3.pdf}
+  \caption{
+    Point estimates of \gls{BA} model parameters obtained by running kernel-ABC
+    on simulated phylogenies without training, for simulations with $m = 3$.
+    Dotted lines indicate true values, and limits of the $y$-axes are regions
+    of uniform prior density.
+  }
+  \label{fig:abcptm3}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{abc-point-estimate-m4.pdf}
+  \caption{
+    Point estimates of \gls{BA} model parameters obtained by running kernel-ABC
+    on simulated phylogenies without training, for simulations with $m = 4$.
+    Dotted lines indicate true values, and limits of the $y$-axes are regions
+    of uniform prior density.
+  }
+  \label{fig:abcptm4}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{alpha-gamma.pdf}
+  \caption{
+      Relationship between preferential attachment power parameter $\alpha$
+      and power-law exponent $\gamma$ for networks simulated under the \gls{BA}
+      network model with $N$ = 5000 and $m$ = 2.
+  }
+  \label{fig:gamma}
+\end{figure}
+
 \end{document}

### 2016-03-17 doc/paper/paper.tex
Mostly writing

@@
 \usepackage{cleveref}
+\usepackage{array}
 
@@
 
-When an infectious disease spreads through a population, transmissions cannot
-generally occur between any two individuals at any time. Rather, the
-individuals must have a particular mode of contact with one another, which
-varies with the disease's mode of transmission. For airborne pathogens,
-physical proximity may be sufficient, but for sexually transmitted diseases,
-either sexual or blood-to-blood contact is required. The set of links between
-individuals along which transmission can occur is called the contact
-network~\autocite{klovdahl1985social}. The structure of the contact network
-underlying an epidemic can profoundly impact the speed and pattern of the
-epidemic's expansion. Network structure can influence the prevalence
-trajectory~\autocite{o2010contact} and epidemic
+When an infectious disease spreads through a population, transmissions are
+generally more likely to occur between certain pairs of individuals. Such pairs
+must have a particular mode of contact with one another, which varies with the
+mode of transmission of the disease. For airborne pathogens, physical proximity
+may be sufficient, but for sexually transmitted diseases, either sexual or
+blood-to-blood contact is required. The set of links between individuals along
+which transmission can occur is called the contact
+network~\autocite{klovdahl1985social, morris1993epidemiology}. The structure of
+the contact network underlying an epidemic can profoundly impact the speed and
+pattern of the epidemic's expansion. Network structure can influence the
+prevalence trajectory~\autocite{o2010contact} and epidemic
 threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates
@@
 \textcite{groendyke2011bayesian} and applied to a much larger data set of 188
-individuals. \textcite{brown2011transmission} found that a Waring distribution,
-which is produced by a different preferential attachment model, was a good fit
-to a phylogenetically estimated transmission network relating 60\% of HIV
-infected \gls{MSM} in the United Kingdom. The transmission network is a
-subgraph of the contact network which includes only those edges which have
-already led to a new infection.
+individuals. \textcite{brown2011transmission} analysed the degree distribution
+of an approximate transmission network, estimated based on genetic similarity
+and estimated times of infection, relating 60\% of HIV infected \gls{MSM} in
+the United Kingdom. The transmission network is a subgraph of the contact
+network which includes only those edges which have already led to a new
+infection. The authors found that a Waring distribution, which is produced by a
+different preferential attachment model, was a good fit to their estimated
+network.
 
@@
 calculation of a normalizing constant, Bayesian inference is often performed
-using \gls{MCMC}, which requires the calculation of likelihood ratios.
+using \gls{MCMC}. Rather than calculating explicit likeilihoods, \gls{MCMC}
+uses likelihood \emph{ratios} in which the normalizing constant cancel out.
 Unfortunately, it is generally difficult to explicitly calculate the likelihood
-of an observed transmission tree under a contact network model. To do so, it
-would be necessary to integrate over all possible networks, and also over all
-possible labellings of the internal nodes of the transmission tree. While it is
-not known (to us) whether such integration is tractable, a simpler alternative
-is offered by likelihood-free methods, namely \gls{ABC}. \gls{ABC} leverages
-the fact that, although calculating the likelihood may be impossible,
-generating simulated datasets according to a model is often straightforward. If
-our model fits the data well, the simulated data it produces should be similar
-to the observed data. More formally, if $D$ is the observed data, the posterior
-distribution $f(\theta \mid D)$ on model parameters $\theta$ is replaced as the
-target of statistical inference by $f(\theta \mid \rho(\hat{D}, D) <
-\varepsilon)$, where $\rho$ is a distance function, $\hat{D}$ is a simulated
-dataset according to $\theta$, and $\varepsilon$ is a small
-tolerance~\autocite{sunnaker2013approximate}. In the specific case when $\rho$
-is a kernel function, the approach is known as kernel-\gls{ABC}.
+of an observed transmission tree under a contact network model, even up to a
+normalizing constant. To do so, it would be necessary to integrate over all
+possible networks, and also over all possible labellings of the internal nodes
+of the transmission tree. While it is not known (to us) whether such
+integration is tractable, a simpler alternative is offered by likelihood-free
+methods, namely \gls{ABC}. \gls{ABC} leverages the fact that, although
+calculating the likelihood may be impossible, generating simulated datasets
+according to a model is often straightforward. If our model fits the data well,
+the simulated data it produces should be similar to the observed data. More
+formally, if $D$ is the observed data, the posterior distribution $f(\theta
+\mid D)$ on model parameters $\theta$ is replaced as the target of statistical
+inference by $f(\theta \mid \rho(\hat{D}, D) < \varepsilon)$, where $\rho$ is a
+distance function, $\hat{D}$ is a simulated dataset according to $\theta$, and
+$\varepsilon$ is a small tolerance~\autocite{sunnaker2013approximate}. In the
+specific case when $\rho$ is a kernel function, the approach is known as
+kernel-\gls{ABC}~\autocite{nakagome2013kernel, poon2015phylodynamic}.
 
@@
 unbalancedness of transmission trees simulated over four network models at
-various levels of pathogen transmissibility.
+various levels of pathogen transmissibility. Our program is freely available at
+\url{https://github.com/rmcclosk/netabc}.
 
@@
 We used the phylogenetic kernel developed by \textcite{poon2013mapping} to test
-whether the parameters of the \gls{BA} model had an effect on tree shape. The
-parameters were varied one at a time over the following values: $N$ =
-\sett{3000, 5000, 8000}, $I$ = \sett{500, 1000, 2000}, $m$ = \sett{2, 3, 4},
-and $\alpha$ = \sett{0.5, 1, 1.5}. The parameters not being tested were fixed
-at the values $N$ = 5000, $I$ = 1000, $m$ = 2, and $\alpha$ = 1. For each
-parameter set, 100 networks were generated, and a transmission tree was
-simulated over each (300 trees per parameter set). The 300 trees were compared
-pairwise with the tree kernel to form a $300 \times 300$ kernel matrix. We
-constructed a \gls{kSVM} classifier for the parameter of interest using the
+whether the parameters of the \gls{BA} model had an effect on tree shape. We
+simulated 100 networks under each of three different values of $\alpha$: 0.5,
+1.0, and 1.5 (300 networks total). The other parameters were fixed to the
+following values: $N$ = 5000, $I$ = 1000, and $m$ = 2. A transmission tree with
+500 tips was simulated over each network (300 transmission trees total). The
+300 trees were compared pairwise with the tree kernel to form a $300 \times
+300$ kernel matrix. The kernel meta-parameters $\lambda$ (the ``decay
+factor''), and $\sigma$ (the ``radial basis function
+variance'')~\autocite[see][]{poon2013mapping}, were set to 0.3 and 4
+respectively. We constructed a \gls{kSVM} classifier for $\alpha$ using the
 \software{kernlab} package~\autocite{karatzoglou2004kernlab}, and evaluated its
-accuracy with 1000 two-fold cross-validations. The construction of the kernel
-matrices and classifiers was repeated for several combinations of the kernel
-meta-parameters $\lambda$ (the ``decay factor''), and $\sigma$ (the ``radial
-basis function variance'')~\autocite[see][]{poon2013mapping}, and for
-transmission trees of sizes 100, 500, and 1000. We also tested univariate
-classifier based on Sackin's index~\autocite{shao1990tree} and an ordinary SVM
-based on the \gls{nLTT} statistic~\autocite{janzen2015approximate}.
+accuracy with 1000 two-fold cross-validations.
+
+Three similar experiments were performed for the other \gls{BA} model
+parameters (one experiment per parameter). $m$ was varied between 2, 3, and 4;
+$I$ between 500, 1000, and 2000; and $N$ between 3000, 5000, and 8000. The
+parameters not being tested were fixed at the values $N$ = 5000, $I$ = 1000,
+$m$ = 2, and $\alpha$ = 1. Thus, we performed a total of four kSVM
+cross-validations, one for each of the \gls{BA} model parameters $\alpha$, $I$,
+$m$, and $N$.
+
+We repeated these four cross-validations with different values of $\lambda$
+(0.2, 0.3, and 0.4) and $\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$), as well
+as on trees with differing numbers of tips (100, 500, and 1000) and in
+epidemics of differing size $I$ (500, 1000, and 2000). When evaluating the
+classifier for $I$, we did not consider trees with 1000 tips, because one of
+the tested $I$ values was 500, and the number of tips cannot be larger than
+$I$.
+
+For each of the four parameters, we also tested univariate classifier based on
+Sackin's index~\autocite{shao1990tree} and an ordinary SVM based on the
+\gls{nLTT} statistic~\autocite{janzen2015approximate}.
 
@@
 tree kernel, although the magnitude of the disparity varied between the
-parameters (\cref{fig:crossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}). For
-appropriate choices of the kernel meta-parameters, the kSVM classifier for
-$\alpha$ had an $R^2$ above 
-    0.8 
-for all combinations of tree size and epidemic size (\cref{fig:crossv}). The
-other classifiers' $R^2$ averaged between
+parameters. \Cref{fig:rsquared} shows the cross-validation accuracy with
+$\lambda$ = 0.3, $\sigma$ = 4, $I$ = 1000, and trees of size 500. The results
+were largely robust to variations in $\lambda$ and $\sigma$, although accuracy
+varied between different epidemic and sampling scenarios
+(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}).
+
+The kSVM classifier for $\alpha$ had an average $R^2$ of
+    0.92 
+(\cref{fig:alphacrossv}). The $R^2$ of the for the SVM classifer for $\alpha$
+using the \gls{nLTT} varied between
 0.49
 and
-  0.61
-for the SVM based on the \gls{nLTT}, and between
+  0.61.
+The $R^2$ of the linear regression against Sackin's index varied between
   0.73
 and
-  0.76
-for the linear regression against Sackin's index. The $I$ parameter was also
-classified accurately by the tree kernel (\cref{fig:Icrossv}), with an average
+  0.76.
+The $I$ parameter was also classified accurately by the tree kernel
+(\cref{fig:Icrossv}), with an average
 $R^2$ of 
@@
   0.02
-for all scenarios. Sackin's index fared slightly better, especially on trees
-with a higher number of tips, with $R^2$ between
+for all scenarios. Sackin's index fared slightly better than the \gls{nLTT},
+especially on trees with a higher number of tips, with $R^2$ between
   0
@@
   \centering
-  \includegraphics[width=\figwidth]{kernel-rsquared}
+  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
   \caption{
-    Cross-validation accuracy of kernel-SVM classifiers for \gls{BA} model
-    parameters, with kernel meta-parameters $\lambda = 0.3$ and $\sigma = 4$.
-    Each point was calculated based on 300 simulated transmission trees over
-    networks with three different values of the parameter being tested.
-    Vertical lines are empirical 95\% confidence intervals based on 1000 2-fold
-    cross-validations.
+      Cross-validation accuracy of kernel-SVM classifier (left), SVM classifier
+      using \gls{nLTT} (centre), and linear regression using Sackin's index
+      (right) for \gls{BA} model parameters. Kernel meta-parameters were set to
+      $\lambda = 0.3$ and $\sigma = 4$. Each point was calculated based on 300
+      simulated transmission trees over networks with three different values of
+      the parameter being tested. Vertical lines are empirical 95\% confidence
+      intervals based on 1000 2-fold cross-validations.
   }
   \label{fig:rsquared}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\figwidth]{kernel-alpha-crossv}
-  \caption{
-    Cross-validation accuracy of kernel-SVM classifiers for $\alpha$ parameter
-    of \gls{BA} network model, for various tree kernel meta-parameters and
-    epidemic scenarios. Each point was calculated based on 300 simulated
-    transmission trees over networks with $\alpha$ = 0.5, 1.0, or 1.5. Dotted
-    and and dashed lines indicate, respectively, performance of SVM using the
-    \gls{nLTT} statistic, and linear regression using Sackin's index. Facets
-    are number of infected nodes before the simulation was stopped ($I$) and
-    number of tips in the sampled transmission tree.
-  }
-  \label{fig:crossv}
 \end{figure}
@@
     [0.05-0.18].
-(\cref{fig:abcpt,tab:abchpd}). The accuracy of the estimates was not
-significantly different between values of $m$ or $I$ (both one-way ANOVA,
+\Cref{fig:abcpt} shows point estimates for all simulations with $m = 2$. The
+results $m = 3$ and $m = 4$ were similar (\cref{fig:abcptm3,fig:abcptm4}).
+Average boundaries of 95\% \gls{HPD} intervals are given in \cref{tab:abchpd}.
+The accuracy of the estimates was not significantly different between values of
+$m$ or $I$ (both one-way ANOVA,
     $p$ = 0.1
@@
                           0.68).
-%but the accuracy was significantly lower for integral than non-integral
-%values of $\alpha$
-%    (Wilcoxon rank-sum test, $p$ = alpha.m.test$p.value).
 Finally, the total number of nodes $N$ was consistently over-estimated by about
@@
   \centering
-  \includegraphics[width=\figwidth]{abc-point-estimate}
+  \includegraphics[width=\textwidth]{abc-point-estimate-m2.pdf}
   \caption{
     Point estimates of \gls{BA} model parameters obtained by running kernel-ABC
-    on simulated phylogenies. Dotted lines indicate the true values, and limits
-    of the $y$-axes are the regions of uniform prior density.
+    on simulated phylogenies without training, for simulations with $m = 2$.
+    Dotted lines indicate true values, and limits of the $y$-axes are regions
+    of uniform prior density.
   }
@@
 distributions for for one simulation (equivalent plots for all the simulations
-can be found in the supplemental materials). \gls{HPD} intervals around
-$\alpha$ and $I$ were narrow relative to the region of nonzero prior density,
-whereas the intervals for $m$ and $N$ were more widely dispersed.
+can be found in supplemental data). \Gls{HPD} intervals around $\alpha$ and $I$
+were narrow relative to the region of nonzero prior density, whereas the
+intervals for $m$ and $N$ were more widely dispersed.
 
@@
   \centering
-  \includegraphics[width=\figwidth]{abc-posterior-example}
+  \includegraphics[width=\textwidth]{abc-posterior-example.pdf}
   \caption{Marginal estimated posterior distributions obtained with ABC on a
@@
 $\alpha$, 
-  1 [0.85-
-                           1.12];
+  1 [0.07-
+                           1.07];
 $I$,
-  5.23 [2.05-
-                       5.85];
+  5.37 [2.05-
+                       5.82];
 $m$,
-  \ensuremath{10^{4}} [\ensuremath{3.06\times 10^{3}}-
+  \ensuremath{1.01\times 10^{4}} [\ensuremath{3.06\times 10^{3}}-
                               \ensuremath{1.5\times 10^{4}}];
 $N$,
-  \ensuremath{1.17\times 10^{3}} [505.71-
-                       \ensuremath{1.75\times 10^{3}}].
+  \ensuremath{1.18\times 10^{3}} [505.71-
+                       \ensuremath{1.62\times 10^{3}}].
 
@@
 $\alpha$, 
-  0.48 [0.03-
+  0.46 [0.03-
                            0.83];
 $I$,
-  3.6 [2.18-
+  3.53 [2.18-
                        5.9];
 $m$,
-  \ensuremath{1.08\times 10^{4}} [\ensuremath{3.11\times 10^{3}}-
-                              \ensuremath{1.48\times 10^{4}}];
+  \ensuremath{1.07\times 10^{4}} [\ensuremath{3.11\times 10^{3}}-
+                              \ensuremath{1.47\times 10^{4}}];
 $N$,
-  \ensuremath{2.48\times 10^{3}} [\ensuremath{1.28\times 10^{3}}-
-                       \ensuremath{3.87\times 10^{3}}].
+  \ensuremath{2.46\times 10^{3}} [\ensuremath{1.28\times 10^{3}}-
+                       \ensuremath{3.88\times 10^{3}}].
 
@@
   \centering
-  \includegraphics[width=\figwidth]{realdata-hpd}
+  \includegraphics{realdata-hpd.pdf}
   \caption{
@@
 
+As noted by \textcite{lintusaari2016identifiability}, uniform priors on model
+parameters may translate to highly informative priors on quantities of
+interest. We observed an exponential relationship between the preferential
+attachment power $\alpha$ and the power-law exponent $\gamma$
+(\cref{fig:gamma}). Therefore, placing a uniform prior on $\alpha$ between 0
+and 2 is equivalent to placing an informative prior that $\gamma$ is close to
+2. Therefore, if we were primarily interested in $\gamma$ rather than $\alpha$,
+a more sensible choice of prior might have a shape similar to \cref{fig:gamma}
+and be bounded above by approximately $\alpha$ = 1.5. This would uniformly
+bound $\gamma$ in the region $2 \leq \gamma \leq 4$ commonly reported in the
+network literature~\autocite{liljeros2001web, schneeberger2004scale,
+colgate1989risk, brown2011transmission}. We note however that
+\textcite{jones2003assessment} estimated $\gamma$ values greater than four, in
+one case as high as 17, for some datasets, indicating that a wider range of
+permitted $\gamma$ values may be warranted.
+
 Our investigation of published HIV datasets indicated heterogeneity in the
@@
 low attachment power for these data, which is consistent with the sampled
-sequences comprising many distinct sub-networks. \textcite{li2015} sampled a
+sequences comprising many distinct sub-networks. \textcite{li2015hiv} sampled a
 large number of acutely infected \gls{MSM} in Shanghai, China. By using a
@@
 available and must be approximated, often by way of a viral phylogeny. Although
-this has been demonstrated to be a fair approximation in
-practice~\autocite[e.g.][]{leitner1996accurate}, the topologies of a viral
+this has been demonstrated to be a fair
+approximation~\autocite[e.g.][]{leitner1996accurate}, and is frequently used in
+practice~\autocite[e.g.][]{stadler2013uncovering}, the topologies of a viral
 phylogeny and transmission tree can differ
@@
 
-\end{document}
 \newpage
@@
   \centering
-  %\includegraphics[scale=0.8]{leventhal2012fig1}
+  \includegraphics{leventhal2012fig1.pdf}
   \caption{Reproduction of Figure 1A from \textcite{leventhal2012inferring}.}
@@
   \centering
-  \includegraphics[width=\figwidth]{kernel-m-crossv}
+  \includegraphics{kernel-alpha-crossv.pdf}
+  \caption{
+    Cross-validation accuracy of kernel-SVM classifiers for $\alpha$ parameter
+    of \gls{BA} network model, for various tree kernel meta-parameters and
+    epidemic scenarios. Each point was calculated based on 300 simulated
+    transmission trees over networks with $\alpha$ = 0.5, 1.0, or 1.5. Dotted
+    and and dashed lines indicate, respectively, performance of SVM using the
+    \gls{nLTT} statistic, and linear regression using Sackin's index. Facets
+    are number of infected nodes before the simulation was stopped ($I$) and
+    number of tips in the sampled transmission tree.
+  }
+  \label{fig:alphacrossv}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{kernel-m-crossv.pdf}
   \caption{Cross-validation accuracy of kernel-SVM classifiers for $m$
@@
   \centering
-  \includegraphics[width=\figwidth]{kernel-I-crossv}
+  \includegraphics{kernel-I-crossv.pdf}
   \caption{Cross-validation accuracy of kernel-SVM classifiers for number of
@@
   \centering
-  \includegraphics[width=\figwidth]{kernel-N-crossv}
+  \includegraphics{kernel-N-crossv.pdf}
   \caption{Cross-validation accuracy of kernel-SVM classifiers for total number
@@
 
+\begin{figure}[ht]
+  \centering
+  \includegraphics{abc-point-estimate-m3.pdf}
+  \caption{
+    Point estimates of \gls{BA} model parameters obtained by running kernel-ABC
+    on simulated phylogenies without training, for simulations with $m = 3$.
+    Dotted lines indicate true values, and limits of the $y$-axes are regions
+    of uniform prior density.
+  }
+  \label{fig:abcptm3}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{abc-point-estimate-m4.pdf}
+  \caption{
+    Point estimates of \gls{BA} model parameters obtained by running kernel-ABC
+    on simulated phylogenies without training, for simulations with $m = 4$.
+    Dotted lines indicate true values, and limits of the $y$-axes are regions
+    of uniform prior density.
+  }
+  \label{fig:abcptm4}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{alpha-gamma.pdf}
+  \caption{
+      Relationship between preferential attachment power parameter $\alpha$
+      and power-law exponent $\gamma$ for networks simulated under the \gls{BA}
+      network model with $N$ = 5000 and $m$ = 2.
+  }
+  \label{fig:gamma}
+\end{figure}
+
 \end{document}

### 2016-03-17 doc/thesis/conclusion.tex
Mostly writing

\section{Extensions}

% ergm

An alternative approach is the deterministic framework outlined by
\textcite{morris1993epidemiology}, who proposes to apply the standard
compartmental modelling framework to contact networks by assigning each
individual their own compartment. Thus, each individual is associated with a
single \gls{ODE}, with the entire \gls{ODE} system parameterized by the
adjacency matrix of the contact network. \citeauthor{morris1993epidemiology}
proposes to use log-linear models to parameterize the matrix. This framework is
highly expressive, and allows straightforward incorporation of time-dependent
dynamics. However, simulating a transmission tree would require the numerical
solution of a very large system of \glspl{ODE}. Given the large number of
simulations required for kernel-\gls{ABC}, it is not clear if this method would
be computationally feasible in this context.

% spatial networks, see section 5.4 in keeling2005networks

% Caveats

It is important to note that our kernel-ABC method takes a transmission tree as
input, rather than a viral phylogeny. Thus, we have left the estimation of a
transmission tree up to the user. There were two reasons for this choice. First
and foremost, we wished to avoid the extra complexity and keep the number of
estimated parameters small. In theory, it is possible to incorporate the
process by which a viral phylogeny is generated along with a transmission tree
into our method, for example by simulating within-host dynamics. Although this
may be an avenue for future extension, we felt that it would obscure the
primary purpose of this work, which is to study contact network parameters.
Second, there are a number of different methods available for inferring
transmission trees~\autocite{didelot2014bayesian, yima2012unravelling,
jombart2011reconstructing, cottam2008integrating, poon2015phylodynamic}, some
of which incorporate geographic and/or epidemiological data not accomodated by
our method. We therefore felt it would be best to allow researchers to use
their own preferred tree building method.

### 2016-03-17 doc/thesis/intro.tex
Mostly writing

@@
 \section{Phylogenetics and phylodynamics}
+\label{sec:phylo}
 
@@
 trees~\autocite{buneman1974note}. These concepts are illustrated in
-Figure~\ref{fig:speciestree}.
+\cref{fig:speciestree}.
 
@@
   \centering
-  \label{fig:speciestree}
-  \includegraphics{speciestree}
+  \includegraphics{speciestree.pdf}
   \caption[Illustration of a rooted, ultrametric, time-scaled phylogeny]
@@
       fish is most distantly related to any other node in the tree.}
+  \label{fig:speciestree}
 \end{figure}
 
-\subsection{Transmission trees and epidemiology}
+\subsection{Transmission trees}
 
@@
 have tips, nodes, edges, and branch lengths. However, rather than recording an
-evolutionary process (speciation), they record an epidemiolgical process
+evolutionary process (speciation), they record an epidemiological process
 (transmission). The tips of a transmission tree represent infected hosts, while
@@
 \defn{isochronous} taxa found in the phylogenies of macro-organisms. A
-transmission tree is illustrated in Figure~\ref{fig:virustrees}.
+transmission tree is illustrated in \cref{fig:contactnet} (right).
+
+In a transmission tree, each sampled individual appears as a tip when they are
+removed from the infected population. In addition, they appear as an internal
+node each time they transmit to an individual who is either eventually sampled,
+or who transmits directly or indirectly to a sampled individual. This is not
+the same as the transmission \emph{network}, in which each infected individual
+appears exactly once, and edges are in one-to-one correspondence with
+transmissions~\autocite{welch2011statistical, keeling2005networks}.
+Transmission networks are discussed further in \cref{sec:contactnet}, and the
+distinction between the two objects is illustrated in \cref{fig:contactnet}.
+However, since transmission networks generally have no cycles (unless
+re-infection occurs), they are trees in the graph theoretical sense, and hence
+are sometimes also referred to as transmission
+trees~\autocite[\eg][]{kenah2015algorithms}. In this work, we reserve the term
+``transmission tree'' for the objects depicted on the right side of
+\cref{fig:contactnet}, following~\autocite[\eg][]{stadler2013uncovering}. The
+term ``transmission network'' is taken to mean the subgraph of the contact
+network along which transmissions occured,
+following~\autocite{welch2011statistical, keeling2005networks}.
 
@@
     \centering
-    \includegraphics{virustrees}
-    \label{fig:virustrees}
-    \caption[Illustration of a transmission tree]{
-        A transmission tree representing an epidemic's progress through
-        four individuals. The index case, patient $a$, transmitted the
-        virus to patients $b$ and $c$; patient $c$ went on to infect
-        patient $d$. 
+    \includegraphics[width=\textwidth]{contactnet.pdf}
+    \caption[Illustration of a contact network and transmission tree]{
+      Illustration of epidemic spread over a contact network, and the
+      corresponding transmission tree. (Left) A contact network with five
+      hosts, labelled $a$ through $e$. Thick shaded edges indicate symmetric
+      contacts among the hosts. The transmission network is indicated by
+      coloured arrows. The epidemic began with node $a$, who transmitted to
+      nodes $b$ and $c$. Node $c$ further transmitted to node $d$. Node $e$ was
+      not infected. (Right) The transmission tree corresponding to this
+      scenario, with a timeline of transmission and removal times.
     }
+    \label{fig:contactnet}
 \end{figure}
@@
 but the most well-studied of epidemics, this is not possible to obtain through
-traditional epidemiological methods. The time and effort to conduct detailed
-interviews and contact tracing of a sufficient number of infected individuals
-is usually prohibitive. Even when the resources for such methods are available,
-patients may not always recall whom they contacted and when, especially in the
-case of airborne transmission. Consequently, the transmission tree must be
-estimated using other methods. Most commonly, this is done by exploiting the
-relationship between transmission trees and viral phylogenies.
+traditional epidemiological methods~\autocite{welch2011statistical}. The time
+and effort to conduct detailed interviews and contact tracing of a sufficient
+number of infected individuals is usually prohibitive. Even when the resources
+for such methods are available, patients may not always recall whom they
+contacted and when, especially in the case of airborne transmission.
+Consequently, the transmission tree must be estimated using other methods. Most
+commonly, this is done by exploiting the relationship between transmission
+trees and viral phylogenies~\autocite{volz2013viral}.
 
-\subsection{Estimating transmission trees from viral phylogenies}
+\subsection{Relationship between transmission trees and viral phylogenies}
 
 In general, \defn{viral phylogenies} are simply phylogenetic trees relating
-virus strains. \comment{Is ``strain'' the best word here?} In phylodynamics, we
-often consider \defn{inter-host} phylogenies, which relate one viral genotype
-from each host in a population. The tips of the tree are labelled with both the
-virus and its host. The crux of
+virus strains. In phylodynamics, we often consider \defn{inter-host}
+phylogenies, which relate one viral genotype from each host in a population.
+The tips of the tree are labelled with both the virus and its host. The crux of
 \defn{phylodynamics}~\autocite{grenfell2004unifying} is the fact that the
@@
 generally assume that sampling a virus from the individual also corresponds to
-their removal from the transmission tree\comment{needs a reference}, so the
-positions of the tips in time are fixed. Therefore, to transform a viral
-phylogeny into an estimated transmission tree, we must find a root and an
-assignment of branch lengths such that the tips are placed at their proscribed
-times. Of course, there are an infinite number of possible combinations of root
-and branch lengths which result in the correct placement of the tips, so we
-would also like to find values which are the most ``reasonable'' given our
-data. By reasonable, we mean that the variation in \defn{evolutionary rate},
-which is the ratio of a branch's length in genetic distance to its length in
-calendar time, is low across the tree. While there is some variation among
-hosts, due to immunological and other factors, we generally expect to observe
-globally similar evolutionary rates. Methods for time-scaling a phylogeny
-include root-to-tip regression~\autocite{shankarappa1999consistent,
-korber2000timing, drummond2003inference}, which we apply in this work, and
-least-square dating~\autocite{to2015fast}. Both of these methods can be used to
-root the tree, by simply trying all possible root positions and choosing the
-one which minimizes a loss function ($1-R^2$, or root-mean-square error, for
-root-to-tip regression; sum of squared errors for least-square dating).
-Alternatively, the tree may be rooted with an outgroup~\autocite{li1988rates}
-before time-scaling.
+their removal from the transmission tree, so the positions of the tips in time
+are fixed. Therefore, to transform a viral phylogeny into an estimated
+transmission tree, we must find a root and an assignment of branch lengths such
+that the tips are placed at their proscribed times. Of course, there are an
+infinite number of possible combinations of root and branch lengths which
+result in the correct placement of the tips, so we would also like to find
+values which are the most ``reasonable'' given our data. By reasonable, we mean
+that the variation in \defn{evolutionary rate}, which is the ratio of a
+branch's length in genetic distance to its length in calendar time, is low
+across the tree. While there is some variation among hosts, due to
+immunological and other factors, we generally expect to observe globally
+similar evolutionary rates. Methods for time-scaling a phylogeny include
+root-to-tip regression~\autocite{shankarappa1999consistent, korber2000timing,
+drummond2003inference}, which we apply in this work, and least-square
+dating~\autocite{to2015fast}. Both of these methods can be used to root the
+tree, by simply trying all possible root positions and choosing the one which
+minimizes a loss function ($1-R^2$, or root-mean-square error, for root-to-tip
+regression; sum of squared errors for least-square dating). Alternatively, the
+tree may be rooted with an outgroup~\autocite{li1988rates} before time-scaling.
 
@@
 lengths and the parameters of a model of evolution. Although they usually work
-well in practice \comment{needs a reference}, the estimated topology can vary
-based on the model used and, in the case of Bayesian analysis, the priors.  In
-addition, intra-host viral populations are genetically heterogeneous, so
-choosing a single representative genotype per host is necessarily imprecise.
-One can use either the genotype of a specific virion sampled from the host, or
-a synthetic genotype, such as a consensus or reconstructed ancestral sequence.
+well in practice, the estimated topology can vary based on the model used and,
+in the case of Bayesian analysis, the priors.  In addition, intra-host viral
+populations are genetically heterogeneous, so choosing a single representative
+genotype per host is necessarily imprecise. One can use either the genotype of
+a specific virion sampled from the host, or a synthetic genotype, such as a
+consensus or reconstructed ancestral sequence.
 
@@
 clock}). This inference is usually done in a Bayesian framework using
-\gls{MCMC}, so that prior information (including the tip dates) \comment{are
-tip dates part of the prior?} can be included in the analysis, and the
-so-called nuisance parameters of the molecular clock model can be marginalized
-out. Software packages for performing these analyses include
-BEAST~\autocite{bouckaert2014beast} and MrBayes~\autocite{ronquist2012mrbayes}.
+\gls{MCMC}, so that prior information (including the tip dates) 
+%\comment{are tip dates part of the prior?} 
+can be included in the analysis, and the so-called nuisance parameters of the
+molecular clock model can be marginalized out. Software packages for performing
+these analyses include BEAST~\autocite{bouckaert2014beast} and
+MrBayes~\autocite{ronquist2012mrbayes}.
 
@@
 however because it enumerates a large portion of the tree space, it is unlikely
-to scale to larger epidemics. A different approach is undertaken by
-\textcite{jombart2011reconstructing}, who describe a method to build
-transmission trees directly from sequence data, contingent on the common
-ancestors also being sampled. This makes the method attractive for
-slow-evolving pathogens, but less practical for viral outbreaks where samples
-from common ancestors are unlikely to be available.
+to scale to larger epidemics. \textcite{ypma2012unravelling} develop a joint 
+likelihood function integrating temporal, geographic, and genetic observations, 
+and use Bayesian \gls{MCMC} to estimate both the tree and the parameters of the
+likelihood function. Their approach can handle missing data and produces high
+resolution transmission trees when multiple types of data are available. A
+different approach is undertaken by \textcite{jombart2011reconstructing}, who
+describe a method to build transmission trees directly from sequence data,
+contingent on the common ancestors also being sampled. This makes the method
+attractive for slow-evolving pathogens, but less practical for viral outbreaks
+where samples from common ancestors are unlikely to be available.
 
@@
 most often an estimated transmission tree~\autocite{pybus2009evolutionary,
-volz2013viral}. Phylogenies are complex objects, and it is not immediately
-obvious how to extract useful information from them with respect to fitting a
-parameter. Standard statistical methods built for numeric data cannot be
-applied directly - for example, one cannot perform a regression of a parameter
-of interest against a phylogeny. Therefore, before we discuss exactly what
-phylodynamics can tell us about epidemics, and how it has been applied in the
-past, we will review some of the methods for quantifying the shapes of
-phylogenies and their similarity to each other.
+volz2013viral}. What is informative about a phylogeny, beyond the demographic
+characteristics of the individuals it relates, is its \defn{shape}. The shape
+of a phylogeny has two components: the topology, and the distribution of branch
+lengths~\autocite{mooers1997inferring}. Methods of quantifying tree shape fall
+into two categories: summary statistics, and pairwise measures.
 
-The shape of a phylogeny consists of two components: the topology, and the
-distribution of branch lengths~\autocite{mooers1997inferring}. Many tree
-summary statistics have been developed to assign numerical values to
-phylogenies based on their shapes. One of the most widely used is Sackin's
-index~\autocite{shao1990tree}, which measures the imbalance or asymmetry in a
-rooted tree. For the $i$th tip of the tree, we define $N_i$ to be the number of
-branches between that tip and the root. The unnormalized Sackin's index is
-defined as the sum of all $N_i$. It is called unnormalized because it does not
-account for the number of tips in the tree. Among two trees having the same
-number of tips, the least-balanced tree will have the highest Sackin's index.
-However, among two equally balanced trees, the larger tree will have a higher
-Sackin's index. This makes it challenging to compare balances among trees of
-different sizes. To correct this, \textcite{kirkpatrick1993searching} derive
-the expected value of Sackin' index under the Yule
-model~\autocite{yule1925mathematical}. Dividing by this expected value
-normalizes Sackin's index, so that it can be used to compare trees of different
-sizes.
+Summary statistics assign a numeric value to each individual tree. One of the
+most widely used is Sackin's index~\autocite{shao1990tree}, which measures the
+imbalance or asymmetry in a rooted tree. For the $i$th tip of the tree, we
+define $N_i$ to be the number of branches between that tip and the root. The
+unnormalized Sackin's index is defined as the sum of all $N_i$. It is called
+unnormalized because it does not account for the number of tips in the tree.
+Among two trees having the same number of tips, the least-balanced tree will
+have the highest Sackin's index. However, among two equally balanced trees, the
+larger tree will have a higher Sackin's index. This makes it challenging to
+compare balances among trees of different sizes. To correct this,
+\textcite{kirkpatrick1993searching} derive the expected value of Sackin' index
+under the Yule model~\autocite{yule1925mathematical}. Dividing by this expected
+value normalizes Sackin's index, so that it can be used to compare trees of
+different sizes.
 
-An alternative to summary statistics are \defn{distance measures} on trees.
-Rather than assigning a numerical value to each tree individually, a distance
-measure associates each pair of trees with a number, indicating how different
-the trees are from each other. Distance measures allow us to identify groups of
+Rather than assigning numbers to individual trees, pairwise measures associate
+a numeric value to each \textit{pair} of trees, indicating how different the
+trees are from each other. Distance measures allow us to identify groups of
 related phylogenies, for example, local epidemics which are undergoing a
@@
 Another tree distance measure is the \defn{phylogenetic kernel}, or ``tree
-kernel'' developed by \textcite{poon2013mapping}. Strictly speaking, the tree
-kernel is a \defn{similarity measure}, rather than a distance measure, as it is
-maximized when the two trees being compared are the same. The basis of the tree
-kernel is the kernel trick originally developed for
+kernel'' developed by \textcite{poon2013mapping}. As opposed to the \gls{nltt},
+the tree kernel is maximized when the two trees being compared are the same.
+The basis of the tree kernel is the kernel trick originally developed for
 \glspl{SVM}~\autocite{burges1998tutorial}. The idea of the kernel trick is to
@@
 product is computed using a \defn{kernel function} without explicitly mapping
-the objects to the feature space, which would be computationally prohibitive. In
-the case of the tree kernel, the feature space is the space of all possible
+the objects to the feature space, which would be computationally prohibitive.
+In the case of the tree kernel, the feature space is the space of all possible
 \defn{subset trees}, which are subtrees that do not necessarily extend all the
@@
 \citeauthor{poon2015phylodynamic} used the tree kernel as the distance function
-in \gls{ABC} (see Section~\ref{sec:abc}), an approach dubbed kernel-\gls{ABC},
-to fit epidemiological models to observed trees.
-
-\subsection{Applications of phylodynamics}
-\label{subsubsec:appphylo}
+in \gls{ABC} (see \cref{sec:abc}), to fit epidemiological models to observed
+trees. This is an example of kernel-ABC~\autocite{nakagome2013kernel}, which
+will be discussed further in \cref{sec:abc}.
 
 \section{Contact networks}
-\label{subsec:contactnet}
+\label{sec:contactnet}
 
 \subsection{Overview}
+\label{subsec:netoverview}
 
@@
 the pathogen in question. For an airborne pathogen like influenza, a contact
-may be simple physical proximity, while for a sexually transmitted pathogen
-like HIV, contacts would be sexual partnerships. A \defn{contact network} is a
-graphical representation of a host population and the contacts among its
-members. The \defn{nodes} in the network represent hosts, and \defn{edges}
-represent contacts between them.
+may be simple physical proximity, while for \gls{HIV}, contact could be via
+unprotected sexual relations or blood-to-blood contact (such as through needle
+sharing). A \defn{contact network} is a graphical representation of a host
+population and the contacts among its members~\autocite{klovdahl1985social,
+morris1993epidemiology, keeling2005networks}. The \defn{nodes} in the network
+represent hosts, and \defn{edges} or \defn{links} represent contacts between
+them. A contact network is shown in \cref{fig:contactnet} (left).
 
@@
 edges, because the same physical proximity is required for a host to infect or
-to become infected. However, a blood-borne infection spread through
-transfusions would use undirected edges, since the donor has no chance of
-transmitting to the recipient. Directed edges are also useful when the
-transmission risk is not equal between the hosts, such as with HIV
-transmission, where acting as the receptive partner carries a higher risk of
-infection than acting as the insertive partner. In this case, a contact could
-be represented by two directed edges, one in each direction between the two
-hosts, with the edges annotated by what kind of risk they imply. In fact, it is
-possible to represent an undirected edge by two symmetric directed edges. For
-this reason, we consider only contact networks with directed edges in the
-sequel. A directed contact network is shown in Figure \ref{fig:contactnet}
-(left).
+to become infected. However, an infection which may be spread through
+blood-to-blood contact through transfusions transfusions would use undirected
+edges, since the donor has no chance of transmitting to the recipient. Directed
+edges are also useful when the transmission risk is not equal between the
+hosts, such as with HIV transmission, where acting as the receptive partner
+carries a higher risk of infection than acting as the insertive partner. In
+this case, a contact could be represented by two directed edges, one in each
+direction between the two hosts, with the edges annotated by what kind of risk
+they imply~\autocite{wasserman1994social}. An undirected contact network is
+equivalent to a directed network where each contact is represented by two
+symmetric directed edges. The \defn{degree} of a node in the network is how
+many contacts it has. In directed networks, we may make the distinction between
+\defn{out-degree} and \defn{in-degree}, which count respectively the number
+incoming and outgoing edges. The set of edges attached to a node are referred
+to as its \defn{incident} edges.
 
-The path an epidemic takes through a contact network determines the topology of
-the transmission tree relating the infected hosts. The initially infected node
-who introduces the epidemic becomes the root of the tree. Each time a
-transmission occurs, the lineage corresponding to the donor host in the tree
-splits into two, representing the recipient lineage and the continuation of the
-donor lineage. This correspondence is illustrated in figure
-\ref{fig:contactnet}. It's important to note that, although the order and
-timing of transmissions determines the tree topology uniquely, the converse
-does not hold. That is, there are generally several orders of infection which
-could lead to the same topology, since the labels on the internal nodes of the
-tree are not available to the researcher.
+The path an epidemic takes through a contact network defines the
+\defn{transmission network}, which is the subgraph of the contact network
+consisting of infected nodes and the edges which led to their
+infections~\autocite{welch2011statistical} (\cref{fig:contactnet}, left). The
+transmission network determines the topology of the transmission tree relating
+the infected hosts. The initially infected node who introduces the epidemic
+into the network, also known as the \defn{index case}, becomes the root of the
+tree. Each time a transmission occurs, the lineage corresponding to the donor
+host in the tree splits into two, representing the recipient lineage and the
+continuation of the donor lineage. \Cref{fig:contactnet} illustrates this
+correspondence. It's important to note that, although the order and timing of
+transmissions determines the tree topology uniquely, the converse does not
+hold. That is, for any given topology, there are in general many transmission
+networks which would lead to that topology. In other words, it impossible to
+distinguish who transmitted to whom from a transmission tree
+alone~\autocite{bernard2007hiv}.
 
-\begin{figure}[ht]
-  \centering
-  \label{fig:contactnet}
-  \includegraphics{contactnet}
-  \caption[Illustration of a epidemic spread over a contact network]
-  {Illustration of epidemic spread over a contact network. On the left, a
-   contact network with five hosts, labelled A through E. Directed edges
-   indicate six symmetric contacts among the hosts. Coloured, bolded edges
-   represent transmissions. The epidemic began with node A, who transmitted to
-   nodes B and C. Node C further transmitted to node D, and node E was not
-   infected. On the right, the transmission tree topology corresponding to this
-   scenario. Note that individual B was sampled before the other infected
-   individuals, resulting in a tree with heterochronous tips.}
-\end{figure}
+From a public health perspective, knowledge of contact networks has the
+potential to be extremely useful. On a population level, network structure can
+dramatically affect the speed and pattern of epidemic
+spread~\autocite[\eg][]{barthelemy2005dynamical}. For example, epidemics are
+expected to spread more rapidly in networks having the ``small world''
+property, where the average path length between two nodes in the network is
+relatively low~\autocite{watts1998collective}. Some sexually transmitted
+infections would not be expected to survive in a homogeneously mixed
+population, but their long-term persistance can be explained by contact
+heterogeneity~\autocite{anderson1992infectious, pastor2001epidemic}. Hence, the
+contact network can provide an idea of what to expect as an epidemic unfolds.
+In terms of actionable information, vaccination strategies which would
+irradicate an epidemic in a random network might not work if the network is
+scale-free~\autocite[][see \cref{subsec:pa}]{keeling2005networks}. On a local
+level, contact networks can be informative about the groups or individuals who
+are at highest risk of acquiring or transmitting infection, and would therefore
+benefit most from public health interventions~\autocite{wang2015targeting,
+little2014using}.
 
-\subsection{Models for contact networks}
-\label{subsubsec:generative}
+Contact networks are a challenging type of data to collect, requiring extensive
+epidemiological investigation in the form of contact
+tracing~\autocite{morris1993epidemiology, welch2011statistical,
+keeling2005networks}. Several less resource-intensive alternatives have been
+explored. Some work focuses on the phylogenetic network, detecting patterns
+such as clustering. However, the phylogenetic network may show little to no
+agreement with a contact data obtained through epidemiological
+methods~\autocite{yirrell1998molecular, resik2007limitations}. Some
+studies~\autocite{brown2011transmission} have focused on parameters of the
+transmission network, rather than the more complete contact network, because it
+is possible to derive an estimate of the transmission network phylogenetically.
 
-Throughout this section, the variable $n$ will be used to refer to the number
-of nodes in the network.
+Rather than considering entire social networks, some epidemiological research
+has focused on properties which can be investigated without extensive contact
+tracing. The most common of these properties is the \defn{degree distribution}
+of the network, which denotes the probability that a node has any given number
+of links. Degree distributions for sexual networks can be estimated by simply
+asking each person how many contacts they had in, say, the last
+year~\autocite{liljeros2001web, schneeberger2004scale, colgate1989risk}.
 
-An \defn{Erd\H{o}s-R\'enyi} (ER) network~\autocite{erdos1960evolution}, also
-known as a Bernoulli network, was the first network model described. It has a
-single parameter, $p$, which is the probability that any particular edge is
-present in the network. To construct an ER network, one simply chooses
-$p \times \binom{n}{2}$ distinct pairs of nodes and draws an edge between each
-pair. The average degree of a node in an ER network is equal to $p \times n$.
+There have been a few statistical methods developed to infer contact network
+parameters without extensive contact tracing. The pioneering study in this
+group was by \textcite{britton2002bayesian}, who developed a Bayesian method to
+infer the $p$ parameter of an \gls{ER} network, along with the transmission and
+removal rate parameters of the \gls{SI} model. Their method is able to use
+either observed infection and removal times, or removal times only. However, it
+is designed for only a small number of observations (fifteen and 42 cases in
+their applications), and their estimates of $p$ for real outbreaks were mostly
+uninformative (95\% confidence intervals [0.11-0.96] and [0.055-0.96]; $p$ is
+bounded in [0, 1]). \textcite{groendyke2011bayesian} significantly updated and
+extended the methodology of \citeauthor{britton2002bayesian}, and applied it to
+a measles outbreak affecting 188 individuals. They were able to obtain a much
+more informative estimate of $p$, although this data set included both symptom
+onset and recovery times for all individuals. They also assumed that the entire
+network was infected, so that the network size was exactly 188. It is unclear
+how well their method would scale to networks with thousands of nodes.
 
-A \defn{Barab\'asi-Albert} (BA) network~\autocite{barabasi1999emergence},
-incorporates the phenomenon of \defn{preferential attachment} seen in real life
-networks. This means that new connections are more likely to involve nodes
-which already have a high number of connections - popular nodes tend to become
-more popular. In the formulation we consider here, BA networks have two
-parameters, $m$ and $\alpha$, which are used to construct the network via the
-following algorithm. The network begins as a single node, then the remaining
-$n-1$ nodes are added one at a time. Each time a new node is added, it is
-connected to $m$ other nodes. The probability of one of these connections
-involving an existing node of degree $k$ is proportional to $k^\alpha$. Note
-that in the original formulation~\autocite{barabasi1999emergence}, there was an
-additional parameter $m_0$ which indicated the number of initial vertices in
-the network, and the only value of $\alpha$ considered was 1. For our purposes,
-we fix $m_0 = 1$ but allow $\alpha$ to vary. We refer to $\alpha$ throughout
-this work as the \defn{preferential attachment power} or just attachment power.
+\subsection{Scale-free networks and preferential attachment}
+\label{subsec:pa}
 
-Networks produced by the BA model are \defn{scale free}, which means that the
-probability of a node having $k$ connections is proportional to $k^\gamma$ for
-some constant $\gamma$. This is also called a \defn{power law} degree
-distribution. An important remark on notation must be made here: in network
-literature, the Greek letter $\alpha$ has often been used to refer to the power
-law exponent of the degree distribution, not the preferential attachment power.
-However, the paper defining the network model~\autocite{barabasi1999emergence}
-uses $\gamma$ for the power law exponent, and $\alpha$ for the attachment
-power, and we shall do likewise. This must be noted because there are reports
-in the literature of the power law exponent being as large as 4, which would
-seem to conflict with our choice (see subsection XX) to bound $\alpha$ above by
-2. The reader should rest assured that it is the attachment power we are
-bounding, not the power law exponent.
+A \defn{scale-free} network is one whose degree distribution follows a power
+law, meaning that the number of nodes in the network with degree $k$ is
+proportional to $k^{-\gamma}$ for some constant
+$\gamma$~\autocite{barabasi1999emergence}. Scale-free networks are
+characterized by a large number of nodes of low degree, with relatively few
+``hub'' nodes of very high degree.
 
-WS and BA networks differ from ER networks in an important way: it is possible
-to explicitly calculate the likelihood of the ER model parameter $p$ given an
-observed network, but this cannot be done for the other two models. This is
-because the probability of a particular edge occurring is always equal to $p$ in
-an ER network, regardless of the presence or absence of any other edges. In
-other words, the edges can be viewed as a series of $\binom{n}{2}$ Bernoulli
-trials with success probability $p$, so that the likelihood of $p$ given an
-observed graph with $k$ edges is distributed as 
-Binomial$\left(\binom{n}{2}, p\right)$.
+Epidemiological surveys have indicated that human sexual networks tend to be
+scale-free~\autocite{liljeros2001web, schneeberger2004scale, colgate1989risk}.
+In this context, the high-degree hub nodes are known as
+\defn{superspreaders}~\autocite{kemper1980identification}.
+
+One mechanism which has been shown to lead to scale-free networks is
+\defn{preferential attachment}~\autocite{simon1955class, barabasi1999emergence}.
+Under this process, networks are formed by starting with a small number $m_0$
+of nodes. New nodes are added one at a time until there are a total of $N$ in
+the network. Each time a new node is added, $m \geq 1$ edges are added from it
+to other nodes in the graph. In the original
+formulation~\autocite{barabasi1999emergence}, the partners of the new node are
+chosen with probability linearly proportional to their degree. However,
+\citeauthor{barabasi1999emergence} suggest extending the model such that the
+probability of choosing a partner of degree $d$ is proportional to $d^\alpha$
+for some constant $\alpha$, and we use this extension here.
+
+There has been some contention of the idea that sexual networks are scale-free.
+\textcite{handcock2004likelihood} fit several stochastic models of partner
+formation to empirical degree distributions derived from population surveys of
+sexual behaviour. They found that a negative binomial distribution, rather than
+a power law, was the best fit to five out of six datasets, although the
+difference in goodness of fit was extremely small in four out of these five.
+
+\subsection{Relationship between network structure and transmission trees}
+
+Clearly, the shape of a transmission tree is shaped by the underlying contact
+network (\cref{fig:contactnet}). A number of studies have made progress in
+quantifying this relationship. \textcite{o2010contact} simulated epidemics over
+networks with four types of degree distribution. They then estimated the
+Bayesian skyride~\autocite{minin2008smooth} population size trajectory in two
+ways: from the phylogeny, using \gls{MCMC}; and from the incidence and
+prevalence trajectories, using the method developed by Volz
+\etal~\autocite{volz2009phylodynamics}. They found that the concordance between
+the two skyrides, as well as the relationship between the skyride and
+prevalence curve, was qualitatively different for each degree distribution.
+\citeauthor{leventhal2012inferring} simulated transmission trees over \gls{ER},
+\gls{WS}, \gls{BA}, and full networks with fixed number of nodes and mean
+degree. They calculated Sackin's index of the simulated trees while varying the
+epidemic, network, and sampling parameters, and found that the relationship
+between these parameters and Sackin's index varies considerably among the
+different network models. In summary, studies in this group have demonstrated
+that network structure profoundly influences tree shape, but have not attempted
+to quantitatively infer network parameters from observed trees.
 
@@
 
-\section{Sequential Monte Carlo}
-
-\Gls{SMC} is a statistical inference method which samples from a sequence of
-probability distributions in a fixed order~\autocite{del2006sequential}. This
-idea of \defn{sequential sampling} is useful several contexts, but we consider
-here its application to model fitting in a Bayesian setting.
-
-\texttt{TODO: move the basics of model fitting (likelihoods and priors) to here}
-
-Our objective is to obtain samples from, and summary statistics of, the
-posterior distribution on the model of interest's parameters. \gls{SMC} can be
-applied in this context by defining a sequence of distributions, starting from
-the posterior and ending at the prior, which constitute a ``smooth''
-trajectory. The main \gls{SMC} algorithm for this problem was developed by
-\textcite{del2006sequential}, based on an existing methodology called
-\gls{SIS}. \gls{SIS} is also designed to sample from a sequence of
-distributions, but rather than all being defined on the same space (such as
-model parameters), they are defined on nested spaces of increasing dimension.
-Following \textcite{del2006sequential} and other reviews of
-\gls{SMC}~\autocite{doucet2001introduction}, we will begin this section by
-describing \gls{SIS}, and then turn to the adaptation of the method to the case
-when the sequence of distributions are all defined on the same space. Note that
-\textcite{del2006sequential} use the variable $\pi$ for the target
-distributions, but \textcite{doucet2001introduction} use $\pi$ to indicate the
-importance distributions. We again follow the former and denote target
-distributions by $\pi$ and importance distributions by $\eta$.
-
-The basis of \gls{SIS} is \gls{IS}, which is a method of estimating summary
-statistics of distributions which are known only up to a normalizing constant,
-and therefore cannot be sampled from directly. That is, if $\pi$ is such a
-distribution and $f$ is any real-valued function, \gls{IS} is concerned with
-estimating
-\[
-  \pi(f) = \int f(x)\pi(x)\d x = \int f(x) \frac{\gamma(x)}{Z} \d x,
-\]
-where the integral is over the space on which $\pi$ is defined, $\gamma(x)$ is
-known, and $Z$ is the unknown normalizing constant, $Z = \int \gamma(x) \d x$.
-The posterior distributions of all but the simplest models' parameters fall
-into this category. Suppose we have at hand another distribution $\eta$, called
-the \defn{importance distribution}, from which we are able to sample. Define
-the \defn{importance weight} as the ratio ratio $w(x) = \gamma(x)/\eta(x)$. We
-can express the normalizing constant $Z$ in terms of the importance weight and
-distribution, $Z = \int w(x) \eta(x) \d x$, and in turn write the original
-integral of interest as
-\[
-  \int f(x) \pi(x) \d x = \frac{\int f(x) \gamma(x) \d x}
-                               {\int w(x) \eta(x) \d x}.
-\]
-If we sample a large number of points from $\eta$, then $\eta(x)$ can be
-approximated at each of them by an empirical distribution. Since the remaining
-quantities $f$, $\gamma$, and $w$ can all be evaluated pointwise, these are all
-the ingredients we need to obtain an estimate of $\pi(f)$. Although this is a
-simple and elegant approach, the drawback is that the variance of the estimate
-is proportional to the variance of the importance
-weights~\autocite{liu2008monte}, which may be quite large if $\eta$ and
-$\gamma$ are very different. Therefore, the practical use of \gls{IS} on its
-own is limited, since it depends on finding an importance distribution which is
-similar to $\pi$, which we usually know very little about \textit{a priori}.
-
-However, an ideal context for the use of \gls{IS} is when we want to sample
-from a sequence of nested probability distributions $\pi_1, \ldots, \pi_n$.
-By \defn{nested}, we mean that $\pi_{i+1}$ is defined on a space of dimension
-$i+1$ and admits $\pi_i$ as a marginal. That is,
-\[
-  \pi_{i+1}(x_1, \ldots, x_{i+1}) = \pi_i(x_1, \ldots, x_i) f_{i+1}(x_1, \ldots, x_{i+1}),
-\]
-where $f_{i+1}$ is a function which yields a distribution when multiplied with
-$\pi_i$. This situation may seem somewhat contrived, but it arises naturally
-when trying to infer the hidden sequence of parameters of a stateful model. For
-example, \textcite{doucet2001introduction} discuss the case when $\pi_i$ is the
-posterior distribution over the first $i-1$ states of a \gls{HMM}, conditional
-on the observed data up to time $i-i$. We assume that either $\pi_1$ is known
-explicitly, or we have enough information about it to find an adequate
-importance distribution. In the \gls{HMM} example, $\pi_1$ was taken to be the
-prior distribution on the starting state.
-
 \section{Approximate Bayesian computation}
-
 \label{sec:abc}
@@
 be fit using more conventional methods, to observed
-\comment{these are review articles, Beaumont is specific to popgen, the others
-are general}
 data~\autocite{marin2012approximate, sunnaker2013approximate,
@@
 by progressing through a series of intermediate distributions.
+
+\section{Sequential Monte Carlo}
+
+\Gls{SMC} is a statistical inference method which samples from a sequence of
+probability distributions in a fixed order~\autocite{del2006sequential}. This
+idea of \defn{sequential sampling} is useful several contexts, but we consider
+here its application to model fitting in a Bayesian setting.
+
+\texttt{TODO: move the basics of model fitting (likelihoods and priors) to here}
+
+Our objective is to obtain samples from, and summary statistics of, the
+posterior distribution on the model of interest's parameters. \gls{SMC} can be
+applied in this context by defining a sequence of distributions, starting from
+the posterior and ending at the prior, which constitute a ``smooth''
+trajectory. The main \gls{SMC} algorithm for this problem was developed by
+\textcite{del2006sequential}, based on an existing methodology called
+\gls{SIS}. \gls{SIS} is also designed to sample from a sequence of
+distributions, but rather than all being defined on the same space (such as
+model parameters), they are defined on nested spaces of increasing dimension.
+Following \textcite{del2006sequential} and other reviews of
+\gls{SMC}~\autocite{doucet2001introduction}, we will begin this section by
+describing \gls{SIS}, and then turn to the adaptation of the method to the case
+when the sequence of distributions are all defined on the same space. Note that
+\textcite{del2006sequential} use the variable $\pi$ for the target
+distributions, but \textcite{doucet2001introduction} use $\pi$ to indicate the
+importance distributions. We follow the former and denote target distributions
+by $\pi$ and importance distributions by $\eta$.
+
+\subsection{Sequential importance sampling}
+
+The basis of \gls{SIS} is \gls{IS}, which is a method of estimating summary
+statistics of distributions which are known only up to a normalizing constant,
+and therefore cannot be sampled from directly. That is, if $\pi$ is such a
+distribution and $f$ is any real-valued function, \gls{IS} is concerned with
+estimating
+\[
+  \pi(f) = \int f(x)\pi(x)\d x = \int f(x) \frac{\gamma(x)}{Z} \d x,
+\]
+where the integral is over the space on which $\pi$ is defined, $\gamma(x)$ is
+known, and $Z$ is the unknown normalizing constant, $Z = \int \gamma(x) \d x$.
+The posterior distributions of all but the simplest models' parameters fall
+into this category. Suppose we have at hand another distribution $\eta$, called
+the \defn{importance distribution}, from which we are able to sample. Define
+the \defn{importance weight} as the ratio ratio $w(x) = \gamma(x)/\eta(x)$. We
+can express the normalizing constant $Z$ in terms of the importance weight and
+distribution, $Z = \int w(x) \eta(x) \d x$, and in turn write the original
+integral of interest as
+\[
+  \int f(x) \pi(x) \d x = \frac{\int f(x) \gamma(x) \d x}
+                               {\int w(x) \eta(x) \d x}.
+\]
+If we sample a large number of points from $\eta$, then $\eta(x)$ can be
+approximated at each of them by an empirical distribution. Since the remaining
+quantities $f$, $\gamma$, and $w$ can all be evaluated pointwise, these are all
+the ingredients we need to obtain an estimate of $\pi(f)$. Although this is a
+simple and elegant approach, the drawback is that the variance of the estimate
+is proportional to the variance of the importance
+weights~\autocite{liu2008monte}, which may be quite large if $\eta$ and
+$\gamma$ are very different. Therefore, the practical use of \gls{IS} on its
+own is limited, since it depends on finding an importance distribution which is
+similar to $\pi$, which we usually know very little about \textit{a priori}.
+
+However, an ideal context for the use of \gls{IS} is when we want to sample
+from a sequence of nested probability distributions $\pi_1, \ldots, \pi_n$.
+By \defn{nested}, we mean that $\pi_{i+1}$ is defined on a space of dimension
+$i+1$ and admits $\pi_i$ as a marginal. That is,
+\[
+  \pi_{i+1}(x_1, \ldots, x_{i+1}) = \pi_i(x_1, \ldots, x_i) f_{i+1}(x_1, \ldots, x_{i+1}),
+\]
+where $f_{i+1}$ is a function which yields a distribution when multiplied with
+$\pi_i$. This situation may seem somewhat contrived, but it arises naturally
+when trying to infer the hidden sequence of parameters of a stateful model. For
+example, \textcite{doucet2001introduction} discuss the case when $\pi_i$ is the
+posterior distribution over the first $i-1$ states of a \gls{HMM}, conditional
+on the observed data up to time $i-i$. We assume that either $\pi_1$ is known
+explicitly, or we have enough information about it to find an adequate
+importance distribution. In the \gls{HMM} example, $\pi_1$ was taken to be the
+prior distribution on the starting state.

### 2016-03-17 doc/thesis/methods.tex
Mostly writing

@@
 are distant.
+
+\subsubsection*{Epidemic simulation}
+\label{subsubsec:nettree}
 
@@
      on the right graph have been offset for visibility.}
-  \includegraphics[scale=0.75]{nltt}
+  \includegraphics{nltt.pdf}
   \label{fig:nltt}
@@
   \centering
-  \includegraphics{separable}
+  \includegraphics{separable.pdf}
   \caption[Separable versus non-separable pararameters in kernel space]{
@@
 
-Before undertaking a complete ABC analysis, I analysed four simple contact
-network models to determine whether their parameters could be separated in tree
-kernel space. The four models are described in detail in
-subsection~\ref{subsubsec:generative} of the introduction. Briefly, they are:
-random networks, where each possible contact has a fixed probability of
-occuring; preferential attachment networks, where highly connected nodes tend
-to attract more contacts; small world networks, where nodes are connected to
-their immediate neighbours and the occasional far-flung contact; and full
-networks, where every possible connection is present. I will describe here only
-the procedure and results for the preferential attachment networks. The details
-of the other three types of graph can be found in the supplemental materials. A
-graphical schematic of the analysis undertaken here is given in
+Before undertaking a complete ABC analysis, I analysed the parameters of the
+\gls{BA} network model to to determine whether they could be separated in tree
+kernel space. A graphical schematic of the analysis undertaken here is given in
 Figure~\ref{fig:kernelexpt}.
@@
   \label{fig:kernelexpt}
-  \includegraphics{kernel_expt}
+  \includegraphics{kernel-expt.pdf}
   \caption[Schematic of first set of simulation experiments]{
@@
 The method of testing for separability was described previously
-in~\autocite{poon2015phylodynamic}, but I will reiterate it here for
-completeness. As a concrete example, consider the attachment power parameter
-\gls{alpha} of the preferential attachment networks. This parameter describes the
-strength of attraction to highly connected nodes and is bounded below by zero
-indicating no extra attraction. By qualitative observation, I determined that a
-power of 2.0, which produced networks with very few ``hub'' nodes with
-extremely high degree, was a suitable upper bound (see
-Figure~\ref{fig:pabounds}). Therefore, I chose to test the values 0.5, 1.0, and
-1.5 for separability. The other parameters were fixed: the number of nodes in
-the network was 5000, and the mean degree of each node in the network was four.
-As discussed in subsection~\ref{subsubsec:generative}, this is the smallest
-mean degree value for preferential attachment networks which produces networks
-which are more than trees.
+in~\autocite{poon2015phylodynamic}. As a concrete example, consider the
+attachment power parameter \gls{alpha}. This parameter describes the strength
+of attraction to highly connected nodes and is bounded below by zero indicating
+no extra attraction. By qualitative observation, I determined that a power of
+2.0, which produced networks with very few ``hub'' nodes with extremely high
+degree, was a suitable upper bound (see Figure~\ref{fig:alphabds}). Therefore,
+I chose to test the values 0.5, 1.0, and 1.5 for separability. The other
+parameters were fixed: the number of nodes in the network was 5000, and the
+mean degree of each node in the network was four. As discussed in
+subsection~\ref{subsubsec:generative}, this is the smallest mean degree value
+for preferential attachment networks which produces networks which are more
+than trees.
 
@@
   \centering
-  \label{fig:pabounds}
-  \includegraphics{pa_power_bounds}
+  \label{fig:alphabds}
+  \includegraphics{alpha-bounds.pdf}
   \caption[Upper and lower bounds on preferential attachment power]{
@@
 
-For each of the values of $\alpha$, I generated 100 networks on 5000 nodes. An
-epidemic was simulated over each network (see subsection X) until 1000 nodes
-were infected, and 500 of those infected nodes were sampled to form a
-transmission tree. This resulted in 300 total simulated transmission trees -
-100 for each of the three values of $\alpha$. The data generation steps are
-shown on the left side of Figure~\ref{fig:kernelexpt}. Next, I computed the
-tree kernel~\autocite{poon2013mapping} (see
-subsection~\ref{subsubsec:treeshape}) for each pair of trees. These values were
-placed into a 300 $\times$ 300 kernel matrix, where the value at the $(i, j)$th
-position was the tree kernel of the $i$th and $j$th trees.
+For each of the values of $\alpha$, I generated 100 networks, each with 5000
+nodes. An epidemic was simulated over each network (see
+\cref{subsubsec:nettree}) until 1000 nodes were infected, and 500 of those
+infected nodes were sampled to form a transmission tree. This resulted in 300
+total simulated transmission trees - 100 for each of the three values of
+$\alpha$. The data generation steps are shown on the left side of
+Figure~\ref{fig:kernelexpt}. Next, I computed the tree
+kernel~\autocite{poon2013mapping} (see subsection~\ref{subsubsec:treeshape})
+for each pair of trees. These values were placed into a 300 $\times$ 300 kernel
+matrix, where the value at the $(i, j)$th position was the tree kernel of the
+$i$th and $j$th trees.
 
@@
   \centering
-  \includegraphics{acc_prec}
+  \includegraphics{acc-prec.pdf}
   \caption{Illustration of accurate vs. precise kernel score estimates}
@@
   \centering
-  \includegraphics{gridsearch_expt}
+  \includegraphics{gridsearch-expt.pdf}
   \caption{Schematic of grid search simulation experiments.}
@@
 
-\subsection{Applications}
+\subsubsection{Characterization of power-law exponent in Barab\'asi-Albert networks}
 
-\subsubsection{HIV in British Columubia}
+Most studies of social network or transmission network
+parameters~\autocite[e.g.][]{liljeros2001web, jones2003assessment,
+schneeberger2004scale, brown2011transmission} report the coefficient
+\gls{gamma} of the power law degree distribution. To make our results
+comparable to previous work, we used simulated networks to investigate the
+relationship between the \gls{BA} model parameters and \gls{gamma}. A network
+was simulated for each element of the Cartesian product of these parameters:
+\gls{N} = \sett{500, 600, \ldots, 15000}, \gls{m} = \sett{1, 2, \ldots, 8}, and
+\gls{alpha} = \sett{0.00, 0.01, \ldots, 2.00}. A power law distribution was
+fitted to the degree distribution of each simulated network using the
+\software{fit\_power\_law} function in \software{igraph} with the `R.mle'
+implementation. We fitted a \gls{GLM} with Gamma-distributed errors and a log
+link function to the observed distribution of \gls{gamma} values, with
+\gls{alpha}, \gls{m}, \gls{N}, and all possible interaction terms as
+predictors.
+
+\subsection{Real data experiments}
+
+We applied kernel-ABC to six published HIV datasets, summarized in
+\cref{tab:realdata}.
+
+\begin{table}
+  \centering
+  \input{\tablepath/realdata-bc}
+  \caption{Characteristics of published HIV datasets analysed with kernel-ABC.}
+  \label{tab:realdata}
+\end{table}

### 2016-03-17 doc/thesis/objective.tex
Mostly writing

--- 2016-01-29 doc/thesis/objective.tex
@@
 
-A second, more recent group of studies has evaluated the effect of network
-structure on transmission tree shape. The use of models which assume a
-\defn{panmictic} (that is, homogeneously mixed) population in phylodynamics has
-become widespread, so it is natural that some researchers have investigated the
-effects of this assumption. In particular, since phylodynamic methods use
-phylogenies as their data source, several studies have examined the shapes of
-phylogenies arising from non-panmictic populations. 
-\citeauthor{o2010contact}~(\citeyear{o2010contact}) simulated epidemics over
-networks with four types of degree distribution. They then estimated the
-Bayesian skyride~\autocite{minin2008smooth} population size trajectory in two
-ways: from the phylogeny, using \gls{MCMC}; and from the incidence and
-prevalence trajectories, using the method developed by Volz
-\etal~\autocite{volz2009phylodynamics}. They found that the concordance between
-the two skyrides, as well as the relationship between the skyride and
-prevalence curve, was qualitatively different for each degree distribution.
-\citeauthor{leventhal2012inferring} simulated transmission trees over \gls{ER},
-\gls{WS}, \gls{BA}, and full networks with fixed number of nodes and mean
-degree. They calculated Sackin's index of the simulated trees while varying the
-epidemic, network, and sampling parameters, and found that the relationship
-between these parameters and Sackin's index varies considerably among the
-different network models. In summary, studies in this group have demonstrated
-that network structure profoundly influences tree shape, but have not attempted
-to quantitatively infer network parameters from observed trees.
+The use of models which assume a \defn{panmictic} (that is, homogeneously
+mixed) population in phylodynamics has become widespread, so it is natural that
+some researchers have investigated the effects of this assumption. In
+particular, since phylodynamic methods use phylogenies as their data source,
+several studies have examined the shapes of phylogenies arising from
+non-panmictic populations.
 
@@
 third group is most similar in aims to our own work; rather than using
-epidemiological observations, we employ viral phylogenies. The pioneering study
-in this group was by
-\citeauthor{britton2002bayesian}~(\citeyear{britton2002bayesian}), who
-developed a Bayesian method to infer the $p$ parameter of an \gls{ER} network,
-along with the $\beta$ and $\gamma$ parameters of \gls{SI} model. Their method
-is able to use either infection and removal times, or removal times only.
-However, it is designed for only a small number of observations (15 and 42
-cases in their applications), and their estimates of $p$ for real outbreaks
-were mostly uninformative (95\% confidence intervals [0.11-0.96] and
-0.055-0.96]; $p$ is bounded in [0, 1]).
-\citeauthor{groendyke2011bayesian}~(\citeyear{groendyke2011bayesian})
-significantly updated and extended the methodology of
-\citeauthor{britton2002bayesian}, and applied it to a measles outbreak
-affecting 188 individuals. They were able to obtain a much more informative
-estimate of $p$, although this data set included both symptom onset and
-recovery times for all individuals. They also assumed that the entire network
-was infected, so that the network size was exactly 188. It is unclear how well
-their method would scale to networks with thousands of nodes.
+epidemiological observations, we employ viral phylogenies. These were discussed
+in \cref{subsec:netoverview}.

### 2016-03-17 doc/thesis/results.Rnw
Mostly writing

@@
   \label{fig:patrees}
-  \includegraphics{kernel_pa_trees}
+  \includegraphics{kernel-alpha-tree.pdf}
   \caption[Visibly distinctive trees simulated under three values of \gls{alpha}]{
@@
   \label{fig:pacrossv}
-  \includegraphics{kernel_pa_crossv}
+  \includegraphics{kernel-alpha-crossv.pdf}
   \caption[Cross-validation performance of kernel support vector machine
@@
   \label{fig:pakpca}
-  \includegraphics{kernel_pa_kpca}
+  \includegraphics{kernel-kpca.pdf}
   \caption[Projection of kernel matrix for different attachment power values
@@
 
-\begin{figure}[ht]
-  \centering
-  \label{fig:prevtrees}
-  \includegraphics{kernel_nsimnode_trees}
-  \caption[Similarly shaped simulated under three values of \gls{I}]{
-    Epidemics were simulated on networks of size 5000 until \gls{I} = 500,
-    1000, or 2000 nodes were infected. When scaled to the same height, there is
-    no immediately visible distinction in shape between trees simulated under
-    different \gls{I} values.
-  }
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \label{fig:prevcrossv}
-  \includegraphics{kernel_nsimnode_crossv}
-  \caption[Cross-validation performance of kernel-\gls{SVM} classifier for
-           prevalence]{
-    For \gls{BA} networks of size \gls{N} = 5000, a kernel-\gls{SVM} classifier
-    is very accurate at predicting \gls{I}. Epidemics were simulated until
-    \gls{I} = 500, 1000, or 2000 nodes were infected, and either 100 or 500
-    nodes were sampled for inclusion in a transmission tree. The parameters of
-    the \gls{BA} network were \gls{alpha} = 1.0 and \gls{m} = 2. A Sackin's
-    index-based classifier performed very poorly on these data ($R^2 = $ TODO
-    for 100-tip trees and blah for 500-tip trees, not shown).
-  }
-\end{figure}
-
 \subsection{Accuracy of marginal estimates}
@@
   \centering
-  \includegraphics{gridsearch_pa_kernel}
+  \includegraphics{gridsearch-alpha-kernel.pdf}
   \caption[Grid search kernel scores]{
@@
   \centering
-  \includegraphics{gridsearch_pa_estimates}
+  \includegraphics{gridsearch-alpha-point-estimate.pdf}
   \caption[Marginal estimates of \gls{alpha} obtained with grid search]{
@@
 for each parameter are shown in Figure~\ref{fig:abcpt}. Of the four parameters,
-$\alpha$ was the most accurately estimated, with a median [IQR] absolute error
+\gls{alpha} was the most accurately estimated, with a median [IQR] absolute error
 of 
@@
     \Sexpr{alpha.av["factor(true_I)", "Pr(>F)"]}),
-although the errors when the true value of $\alpha$ was zero were
+although the errors when the true value of \gls{alpha} was zero were
 significantly greater than the other values 
@@
 but the accuracy was significantly lower for integral than non-integral
-values of $\alpha$
+values of \gls{alpha}
     (Wilcoxon rank-sum test, $p$ = \Sexpr{alpha.m.test$p.value}).
-Finally, the total number of nodes $N$ was consistently over-estimated by about
-a factor of two
+Finally, the total number of nodes \gls{N} was consistently over-estimated by
+about a factor of two
     (error \Sexpr{d[,median(N_error)]} 
@@
 (equivalent plots for all the simulations can be found in the supplemental
-materials). \gls{HPD} intervals around $\alpha$ and $I$ were narrow relative
-to the region of nonzero prior density, whereas the intervals for $m$ and $N$
-were widely dispersed. Table~\ref{tab:abchpd} shows point estimates and 95\%
-\gls{HPD} intervals averaged over all simulations.
+materials). \Gls{HPD} intervals around \gls{alpha} and \gls{I} were narrow
+relative to the region of nonzero prior density, whereas the intervals for $m$
+and \gls{N} were widely dispersed. Table~\ref{tab:abchpd} shows point estimates and
+95\% \gls{HPD} intervals averaged over all simulations.
+
+\begin{table}
+    \centering
+    \input{\tablepath/abc-hpd}
+    \caption{Average widths of 95\% confidence intervals for \gls{BA} model
+    parameters estimated with kernel-\gls{ABC}.}
+    \label{tab:glm}
+\end{table}
+
+\subsection{Characterization of power-law exponent in Barab\'asi-Albert networks}
+
+\Cref{tab:glm} shows the estimated parameters for a log-link \gls{GLM} fitted
+to the observed distribution of \gls{gamma} values. The coefficients are
+interpretable as multiplicative effects.
+
+\begin{table}
+    \centering
+    \input{\tablepath/pa-gamma-glm}
+    \caption{Estimated \gls{GLM} parameters for relationship between power-law
+    exponent \gls{gamma} and \gls{BA} model parameters.}
+    \label{tab:glm}
+\end{table}

### 2016-03-17 doc/figures/alpha-bounds.R

@@ -16,13 +16,25 @@ summary(degree(g2.large))
 pdf("alpha-bounds.pdf", family="Gillius ADF", width=6, height=6)
 par(mfrow=c(2, 2))
 par(mar=c(2, 2, 2, 2) + 0.1)
+
 plot(g0.small, vertex.label=NA, vertex.color="black", vertex.size=5)
-#text(grconvertX(0.05, "ndc", "user"), grconvertY(0.95, "ndc", "user"), "A", cex=2)
+text(grconvertX(0.05, "ndc", "user"), grconvertY(0.95, "ndc", "user"), "A", cex=2)
+
 plot(g2.small, vertex.label=NA, vertex.color="black", vertex.size=5)
-#text(grconvertX(0.55, "ndc", "user"), grconvertY(0.95, "ndc", "user"), "B", cex=2)
+text(grconvertX(0.55, "ndc", "user"), grconvertY(0.95, "ndc", "user"), "B", cex=2)
+
 par(mar=c(4, 4, 2, 2) + 0.1, xpd=NA)
-plot(density(degree(g0.large)), main=NA, xlab="degree", ylab="density", zero.line=F)
-#text(grconvertX(0.05, "ndc", "user"), grconvertY(0.5, "ndc", "user"), "C", cex=2)
-plot(density(degree(g2.large)), main=NA, xlab="degree", ylab="density", zero.line=F)
-#text(grconvertX(0.55, "ndc", "user"), grconvertY(0.5, "ndc", "user"), "D", cex=2)
+
+d0 <- density(degree(g0.large))
+y0 <- d0$y[d0$x > 0]
+x0 <- log(d0$x[d0$x > 0])
+plot(x0, y0, type="l", main=NA, xlab="log(degree)", ylab="density")
+text(grconvertX(0.05, "ndc", "user"), grconvertY(0.5, "ndc", "user"), "C", cex=2)
+
+d2 <- density(degree(g2.large))
+y2 <- d2$y[d2$x > 0]
+x2 <- log(d2$x[d2$x > 0])
+plot(x2, y2, type="l", main=NA, xlab="log(degree)", ylab="density")
+text(grconvertX(0.55, "ndc", "user"), grconvertY(0.5, "ndc", "user"), "D", cex=2)
+
 dev.off()

### 2016-03-17 programs/netabc/R/collect.R

@@ -6,7 +6,8 @@
 #' @param mm.file MatrixMarket formatted file
 collect.metadata.mm <- function (mm.file)
 {
-    as.data.frame(yaml.load(substring(process.archive(mm.file, readLines, n=2)[2], 2)))
+    as.data.frame(yaml.load(substring(process.archive(mm.file, readLines, n=2)[2], 2)),
+                  stringsAsFactors=FALSE)
 }
 
 #' Collect YAML-formatted metadata from a Newick file
@@ -17,7 +18,8 @@ collect.metadata.mm <- function (mm.file)
 #' @param nwk.file Newick formatted file
 collect.metadata.newick <- function (nwk.file)
 {
-    as.data.frame(yaml.load(substring(process.archive(nwk.file, readLines, n=1)[1], 2)))
+    as.data.frame(yaml.load(substring(process.archive(nwk.file, readLines, n=1)[1], 2)),
+                  stringsAsFactors=FALSE)
 }
 
 #' Collect YAML-formatted metadata from a TSV file
@@ -29,7 +31,8 @@ collect.metadata.newick <- function (nwk.file)
 #' @param tsv.file tab-separated value file
 collect.metadata.tsv <- function (tsv.file)
 {
-    as.data.frame(yaml.load(substring(process.archive(tsv.file, readLines, n=1)[1], 2)))
+    as.data.frame(yaml.load(substring(process.archive(tsv.file, readLines, n=1)[1], 2)),
+                  stringsAsFactors=FALSE)
 }
 
 #' Collect YAML-formatted metadata from a GML file
@@ -44,7 +47,8 @@ collect.metadata.gml <- function (gml.file)
     getter <- function (f) { 
         strsplit(grep("comment", readLines(f, n=20), value=TRUE), '"')[[1]][2]
     }
-    as.data.frame(yaml.load(process.archive(gml.file, getter)))
+    as.data.frame(yaml.load(process.archive(gml.file, getter)),
+                  stringsAsFactors=FALSE)
 }
 
 #' Collect YAML-formatted metadata from several files.
@@ -86,7 +90,8 @@ collect.metadata <- function (data.files)
     # fill in the fields each file doesn't have with NA 
     metadata <- mapply(function (df, new) {
         if (length(new) > 0) {
-            cbind(df, as.data.frame(setNames(as.list(rep(NA, length(new))), new)))
+            cbind(df, as.data.frame(setNames(as.list(rep(NA, length(new))), new),
[…]

### 2016-03-19 doc/thesis/conclusion.tex
Draft of contact network section done

@@
 
+The two-step process of simulating a contact network and subsequently allowing
+an epidemic to spread over that network carries with it the assumption that the
+contact network is static over the duration of the epidemic. Clearly this
+assumption is invalid, as people make and break partnerships on a regular
+basis. Our work has not addressed this assumption, primarily due to our desire
+to avoid the additional complexity required to address the dynamic nature of
+networks. This simplifying assumption is made by most studies using contact
+network models in an epidemiological context~\autocite{welch2011statistical,
+bansal2007individual}. However, in principle, kernel-\gls{ABC} could be
+adapted to dynamic contact networks by using a method such as that developed by
+\textcite{robinson2012dynamics} to simulate a dynamic contact network, while
+concurrently simulating the spread of an epidemic.
+
 % Caveats
@@
 transmission tree up to the user. There were two reasons for this choice. First
-and foremost, we wished to avoid the extra complexity and keep the number of
+and foremost, we wished again to avoid extra complexity and keep the number of
 estimated parameters small. In theory, it is possible to incorporate the

### 2016-03-19 doc/thesis/intro.tex
Draft of contact network section done

@@
 represent hosts, and \defn{edges} or \defn{links} represent contacts between
-them. A contact network is shown in \cref{fig:contactnet} (left).
+them. A contact network is shown in \cref{fig:contactnet} (left). Contact
+networks are a particular type of \defn{social
+network}~\autocite{moreno1953shall, barnes1954class}, which is a network in
+which edges may represent any kind of social or economic relationship. Social
+networks are frequently used in the social sciences to study phenomena where
+relationships between people or entities are important \autocite[for a review
+see][]{wasserman1994social}.
 
@@
 to become infected. However, an infection which may be spread through
-blood-to-blood contact through transfusions transfusions would use undirected
+blood-to-blood contact through transfusions transfusions would use directed
 edges, since the donor has no chance of transmitting to the recipient. Directed
@@
 \defn{out-degree} and \defn{in-degree}, which count respectively the number
-incoming and outgoing edges. The set of edges attached to a node are referred
-to as its \defn{incident} edges.
+incoming and outgoing edges. The \defn{degree distribution} of a network
+denotes the probability that a node has any given number of links. The set of
+edges attached to a node are referred to as its \defn{incident} edges.
 
-The path an epidemic takes through a contact network defines the
-\defn{transmission network}, which is the subgraph of the contact network
-consisting of infected nodes and the edges which led to their
-infections~\autocite{welch2011statistical} (\cref{fig:contactnet}, left). The
-transmission network determines the topology of the transmission tree relating
-the infected hosts. The initially infected node who introduces the epidemic
-into the network, also known as the \defn{index case}, becomes the root of the
-tree. Each time a transmission occurs, the lineage corresponding to the donor
-host in the tree splits into two, representing the recipient lineage and the
-continuation of the donor lineage. \Cref{fig:contactnet} illustrates this
-correspondence. It's important to note that, although the order and timing of
-transmissions determines the tree topology uniquely, the converse does not
-hold. That is, for any given topology, there are in general many transmission
-networks which would lead to that topology. In other words, it impossible to
-distinguish who transmitted to whom from a transmission tree
-alone~\autocite{bernard2007hiv}.
+Epidemiological models most often assume some form of contact homogeneity. The
+simplest models, such as the \gls{SIR} model, assume a completely homogeneously
+mixed population, where every pair of contacts is equally likely. More
+sophisticated models partition the population into groups with different
+contact rates between and among each group. However, these models still assume
+that every possible contact between a member of group $i$ and a member of group
+$j$ is equally likely. This assumption is clearly unrealistic for the majority
+of human communities, and can lead to significant errors in predicted epidemic
+trajectories when there is substantial heterogeneity
+present~\autocite{bansal2007individual, volz2007susceptible}. Contact networks
+provide a way to relax this assumption by representing individuals and their
+contacts explicitly. It is important to note that, although panmixia is an
+unrealistic modelling assumption, it has not proven a substantial hurdle to
+epidemic modelling in practice~\autocite{anderson1992infectious}. Using this
+assumption, researchers have been able to derive estimates of the transmission
+rate and the basic reproductive number of various outbreaks, which have agreed
+with values obtained by on-the-ground data collection. Therefore, if one is
+interested only in these population-level variables, the additional complexity
+of contact network models may not be warranted. Rather, these models are most
+useful when we are interested in properties of the network itself, such as
+centrality, structural balance, and
+transitivity~\autocite{wasserman1994social}.
 
@@
 dramatically affect the speed and pattern of epidemic
-spread~\autocite[\eg][]{barthelemy2005dynamical}. For example, epidemics are
-expected to spread more rapidly in networks having the ``small world''
-property, where the average path length between two nodes in the network is
-relatively low~\autocite{watts1998collective}. Some sexually transmitted
-infections would not be expected to survive in a homogeneously mixed
-population, but their long-term persistance can be explained by contact
+spread~\autocite[\eg][]{barthelemy2005dynamical, volz2008sir}. For example,
+epidemics are expected to spread more rapidly in networks having the ``small
+world'' property, where the average path length between two nodes in the
+network is relatively low~\autocite{watts1998collective}. Some sexually
+transmitted infections would not be expected to survive in a homogeneously
+mixed population, but their long-term persistance can be explained by contact
 heterogeneity~\autocite{anderson1992infectious, pastor2001epidemic}. Hence, the
@@
 tracing~\autocite{morris1993epidemiology, welch2011statistical,
-keeling2005networks}. Several less resource-intensive alternatives have been
-explored. Some work focuses on the phylogenetic network, detecting patterns
-such as clustering. However, the phylogenetic network may show little to no
-agreement with a contact data obtained through epidemiological
-methods~\autocite{yirrell1998molecular, resik2007limitations}. Some
-studies~\autocite{brown2011transmission} have focused on parameters of the
-transmission network, rather than the more complete contact network, because it
-is possible to derive an estimate of the transmission network phylogenetically.
+keeling2005networks}. Therefore, it has been necessary to explore less
+resource-intensive alternatives which still contain information about
+population structure. For instance, it is possible to obtain limited
+information about the contact network by individual interviews without contact
+tracing. Variables which can be estimated in this fashion are referred to as
+\defn{node-level} measures~\autocite{wasserman1994social}. One of the most
+well-studied of these is the degree distribution, which can be estimated by
+simply asking each person how many contacts they had in some interval of
+time~\autocite{liljeros2001web, schneeberger2004scale, colgate1989risk}.
 
-Rather than considering entire social networks, some epidemiological research
-has focused on properties which can be investigated without extensive contact
-tracing. The most common of these properties is the \defn{degree distribution}
-of the network, which denotes the probability that a node has any given number
-of links. Degree distributions for sexual networks can be estimated by simply
-asking each person how many contacts they had in, say, the last
-year~\autocite{liljeros2001web, schneeberger2004scale, colgate1989risk}.
+An alternative approach has been the analysis of other networks, which can be
+estimated with phylogenetic methods from viral sequence data. Some work focuses
+on the \defn{phylogenetic network}, in which two nodes are connected if the
+genetic distance between their viral sequences is below some threshold.
+Primarily, this work has focused on the detection of \defn{phylogenetic
+clusters}, which are groups of individuals whose viral sequences are
+significantly more similar to each other's than to the general population's.
+The phylogenetic network is informative about ``hotspots'' of transmission and
+can be used to identify demographic groups to whom targeted intervetions are
+likely to have the greatest effect~\autocite{poon2014impact}.  However, this
+network may show little to no agreement with a contact data obtained through
+epidemiological methods~\autocite{yirrell1998molecular, resik2007limitations,
+robinson2013dynamics}, and therefore may be a poor proxy for the contact
+network. Other studies~\autocite{brown2011transmission} have investigated the
+\defn{transmission network}, which is the subgraph of the contact network
+consisting of infected nodes and the edges which led to their
+infections~\autocite{welch2011statistical} (\cref{fig:contactnet}, left). It is
+possible to estimate the transmission network phylogenetically, although the
+methods required for doing so are more sophisticated than for estimating the
+phylogenetic network~\autocite{brown2011transmission}. These studies again
+mostly focusing on clustering, and also on degree distributions.
 
-There have been a few statistical methods developed to infer contact network
-parameters without extensive contact tracing. The pioneering study in this
-group was by \textcite{britton2002bayesian}, who developed a Bayesian method to
-infer the $p$ parameter of an \gls{ER} network, along with the transmission and
-removal rate parameters of the \gls{SI} model. Their method is able to use
-either observed infection and removal times, or removal times only. However, it
-is designed for only a small number of observations (fifteen and 42 cases in
-their applications), and their estimates of $p$ for real outbreaks were mostly
-uninformative (95\% confidence intervals [0.11-0.96] and [0.055-0.96]; $p$ is
-bounded in [0, 1]). \textcite{groendyke2011bayesian} significantly updated and
-extended the methodology of \citeauthor{britton2002bayesian}, and applied it to
-a measles outbreak affecting 188 individuals. They were able to obtain a much
-more informative estimate of $p$, although this data set included both symptom
-onset and recovery times for all individuals. They also assumed that the entire
-network was infected, so that the network size was exactly 188. It is unclear
-how well their method would scale to networks with thousands of nodes.
+Other statistical methods have been developed to infer contact network
+parameters strictly from the timeline of an epidemic, using neither genetic
+data nor reported contacts. \textcite{britton2002bayesian} developed a Bayesian
+method to infer the $p$ parameter of an \gls{ER} network, along with the
+transmission and removal rate parameters of the \gls{SI} model, using observed
+infection and optionally removal times. However, it was designed for only a
+small number of observations, and was unable to estimate $p$ independently from
+the transmission rate.  \textcite{groendyke2011bayesian} significantly updated
+and extended the methodology of \citeauthor{britton2002bayesian}, and applied
+it to a measles outbreak affecting 188 individuals. They were able to obtain a
+much more informative estimate of $p$, although this data set included both
+symptom onset and recovery times for all individuals, and was unusual in that
+the entire contact network was presumed to be infected. \textcite{volz2008sir}
+developed differential equations describing the dynamics of the \gls{SIR} model
+on a wide variety of random networks defined by their degree distributions.
+Although the topic of estimation was not addressed in the original paper,
+\citeauthor{volz2008sir}'s method could in principle be used to fit such models
+to observed epidemic trajectories, similar to what is done with the ordinary
+\gls{SIR} model. \textcite{volz2007susceptible} later extended the method to
+dynamic contact networks and applied it to a sexual network relating 99
+individuals investigated during a syphilis outbreak.
 
@@
 characterized by a large number of nodes of low degree, with relatively few
-``hub'' nodes of very high degree.
+``hub'' nodes of very high degree. Epidemiological surveys have indicated that
+human sexual networks tend to be scale-free~\autocite{liljeros2001web,
+schneeberger2004scale, colgate1989risk}. Interestingly, many other types of
+network, including computer networks, biological neural networks, metabolic
+networks~\autocite{jeong2000large}, and academic co-author networks, also have
+the scale-free property.
 
-Epidemiological surveys have indicated that human sexual networks tend to be
-scale-free~\autocite{liljeros2001web, schneeberger2004scale, colgate1989risk}.
-In this context, the high-degree hub nodes are known as
-\defn{superspreaders}~\autocite{kemper1980identification}.
+Several properties of scale-free networks are relevant in epidemiology.  The
+high-degree hub nodes are known as
+\defn{superspreaders}~\autocite{kemper1980identification}, which have been
+postulated to contribute in varying degree to the spread of diseases such as
+\gls{HIV}~\autocite{stadler2013uncovering} and
+\gls{SARS}~\autocite{shen2004superspreading}. Scale-free networks have no
+epidemic threshold~\autocite{pastor2001epidemic}, meaning that diseases with
+arbitrarily low transmissibility can persist at low levels indefinitely. This
+is in contrast with homogeneously mixed populations, in which transmissibility
+below the epidemic threshold would result in exponential decay in the number of
+infected individuals and eventual extinction of the pathogen (Anderson \& May,
+I think).
 
@@
 
-There has been some contention of the idea that sexual networks are scale-free.
-\textcite{handcock2004likelihood} fit several stochastic models of partner
-formation to empirical degree distributions derived from population surveys of
-sexual behaviour. They found that a negative binomial distribution, rather than
-a power law, was the best fit to five out of six datasets, although the
-difference in goodness of fit was extremely small in four out of these five.
+There has been some contention of the idea that contact networks are
+scale-free. \textcite{handcock2004likelihood} fit several stochastic models of
+partner formation to empirical degree distributions derived from population
+surveys of sexual behaviour. They found that a negative binomial distribution,
+rather than a power law, was the best fit to five out of six datasets, although
+the difference in goodness of fit was extremely small in four out of these
+five. \textcite{bansal2007individual} found that an exponential distribution,
+rather than a power law, was the best fit to degree distributions of six social
+or sexual networks.
 
@@
 
-Clearly, the shape of a transmission tree is shaped by the underlying contact
-network (\cref{fig:contactnet}). A number of studies have made progress in
-quantifying this relationship. \textcite{o2010contact} simulated epidemics over
-networks with four types of degree distribution. They then estimated the
-Bayesian skyride~\autocite{minin2008smooth} population size trajectory in two
-ways: from the phylogeny, using \gls{MCMC}; and from the incidence and
-prevalence trajectories, using the method developed by Volz
-\etal~\autocite{volz2009phylodynamics}. They found that the concordance between
+The contact network underlying an epidemic constrains the shape of the
+transmission network, which in turn determines the topology of the transmission
+tree relating the infected hosts (\cref{fig:contactnet}). The initially
+infected node who introduces the epidemic into the network, also known as the
+\defn{index case}, becomes the root of the tree. Each time a transmission
+occurs, the lineage corresponding to the donor host in the tree splits into
+two, representing the recipient lineage and the continuation of the donor
+lineage. \Cref{fig:contactnet} illustrates this correspondence. It's important
+to note that, although the order and timing of transmissions determines the
+tree topology uniquely, the converse does not hold. That is, for any given
+topology, there are in general many transmission networks which would lead to
+that topology. In other words, it impossible to distinguish who transmitted to
+whom from a transmission tree alone~\autocite{bernard2007hiv}.
+
+A number of studies have made progress in quantifying the relationship between
+contact networks and transmission trees. \textcite{o2010contact} simulated
+epidemics over networks with four types of degree distribution. They then
+estimated the Bayesian skyride~\autocite{minin2008smooth} population size
+trajectory in two ways: from the phylogeny, using \gls{MCMC}; and from the
+incidence and prevalence trajectories, using the method developed by 
+\textcite{volz2009phylodynamics}. They found that the concordance between
 the two skyrides, as well as the relationship between the skyride and

### 2016-03-22 doc/paper/paper.Rnw
changes from meeting today

@@
 Little information was gleaned about the $m$ parameter, with the \gls{HPD}
-interval occuping the entire prior region for all datasets. The estimates of
+interval occupying the entire prior region for all datasets. The estimates of
 $N$ were similarly uninformative, with the exception that the point estimate
@@
   }
+  \label{fig:abchpd}
 \end{figure}
@@
 Contact networks can have a strong influence on epidemic progression, and are
-potentially useful as a public health tool. Despite this, few methods exist for
-investigating contact network parameters in a phylodynamic framework.
-Kernel-ABC is a model-agnostic method which can be used to investigate any
-quantity that affects tree shape. In this work, we developed a kernel-ABC-based
-method to infer the parameters of a contact network model. The method is
-general, and could be applied to any model from which contact networks can be
-simulated. We demonstrated the method on the \gls{BA} model, which is a simple 
-model incorprorating the preferential attachment feature commonly seen in
-real-world networks.
+potentially useful as a public health tool~\autocite{wang2015targeting,
+little2014using}. Despite this, few methods exist for investigating contact
+network parameters in a phylodynamic framework~\autocite{groendyke2011bayesian,
+brown2011transmission}. Kernel-ABC is a model-agnostic method which can be used
+to investigate any quantity that affects tree shape. In this work, we developed
+a kernel-ABC-based method to infer the parameters of a contact network model.
+The method is general, and could be applied to any model from which contact
+networks can be simulated. We demonstrated the method on the \gls{BA} model,
+which is a simple model incorporating the preferential attachment feature
+commonly seen in real-world networks.
 
@@
 nodes, had a strong influence on tree shape. This was reflected in the relative
-accuracy of the kernel-ABC estimates of these parameters. On the other hand,
-the $m$ and $N$ parameters did not produce much variation in tree shape,
-resulting in both poorly performing classifiers and uninformative kernel-ABC
-estimates. The total number of nodes $N$ was almost always significantly
-over-estimated. Since the prior on $N$ and $I$ is jointly uniform on a
-triangular region ($I \leq N$), there is more prior mass on high $N$ values. In
-retrospect, it is unreasonable to expect good estimation of $N$, because adding
-more nodes to a \gls{BA} network does not change the edge density or overall
-shape. This can be illustrated by imaging that we add a small number of nodes
-to a network after the epidemic simulation has already been completed. It is
-possible that none of these new nodes attains a connection to any infected
-node. Thus, running the simulation again on the new, larger network could
-produce the exact same transmission tree as before.
+accuracy of the kernel-ABC estimates of these parameters. The total number of
+nodes $N$ had a weak influence on tree shape, which was most prominent when the
+epidemic size $I$ and number of sampled tips were both large. The $m$
+parameter, representing the number of edges created in the network per vertex,
+did not produce much variation in tree shape, resulting in in both poorly
+performing classifiers and uninformative kernel-ABC estimates.
+
+$N$ was almost always significantly over-estimated using kernel-ABC. Since the
+prior on $N$ and $I$ is jointly uniform on a triangular region ($I \leq N$),
+there is more prior mass on high $N$ values. In retrospect, it is unreasonable
+to expect good estimation of $N$, because adding more nodes to a \gls{BA}
+network does not change the edge density or overall shape. This can be
+illustrated by imagining that we add a small number of nodes to a network after
+the epidemic simulation has already been completed. It is possible that none of
+these new nodes attains a connection to any infected node. Thus, running the
+simulation again on the new, larger network could produce the exact same
+transmission tree as before.
 
@@
 Our investigation of published HIV datasets indicated heterogeneity in the
-contact network structures underlying several distinct local epidemics.
-\textcite{cuevas2009hiv} studied a group of newly diagnosed individuals in the
-Basque Country, Spain. Although the individuals were of mixed risk group, and
-therefore unlikely to comprise a single contact network, a high proportion of
-them (47\%) grouped into local transmission clusters, allowing for the
-possibility of weak preferential attachment dynamics. We estimated a relatively
-low attachment power for these data, which is consistent with the sampled
-sequences comprising many distinct sub-networks. \textcite{li2015hiv} sampled a
-large number of acutely infected \gls{MSM} in Shanghai, China. By using a
-patristic distance cutoff~\autocite{poon2014impact}, we identified and analysed
-a large cluster. Surprisingly, the estimated attachment power was fairly low
-with a large credible interval.
+contact network structures underlying several distinct local epidemics. The
+five datasets analysed fell into three categories (\cref{fig:abchpd}). First,
+we estimated a preferential attachment power between 0.5 and 1 for the
+epidemics studied by \textcite{cuevas2009hiv} and \textcite{li2015hiv}, with
+credible intervals occupying nearly the entire region from 0 to 1.
+\citeauthor{cuevas2009hiv} studied a group of newly diagnosed individuals in
+the Basque Country, Spain. Although the individuals were of mixed risk groups,
+and therefore unlikely to comprise a single contact network, a high proportion
+of them (47\%) grouped into local clusters based on genetic distance. The low
+estimated attachment power for these data is consistent with the sampled
+sequences comprising many distinct sub-networks rather than a single connected
+network. \citeauthor{li2015hiv} sampled a large number of acutely infected
+\gls{MSM} in Shanghai, China, in which we identified a large cluster from the
+phylogeny using a patristic distance cutoff~\autocite{poon2014impact}. The low
+attachment power estimated for this dataset was surprising given the high
+phylogenetic relatedness of the sequences. It is possible that the number and
+diversity of circulating recombinant forms introduced errors into the estimated
+viral phylogeny.
 
-\textcite{niculescu2015recent} studied a recent outbreak among Romainian
-\gls{IDU}. The estimated attachment power for this dataset was the highest
-among all we considered, at slightly above one. The dataset constituted a
-fairly high sampling prevalence of 33\% of newly diagnosed \gls{IDU} infections
-over the study period. \textcite{wang2015targeting} sampled acutely infected
-\gls{MSM} in Beijing, China, and discovered a high degree of clustering which
-indicated that the local \gls{MSM} subnetwork had been deeply sampled. In both
-these datasets, the estimated attachment power was close to one.
+For the outbreaks studied by \textcite{niculescu2015recent} and
+\textcite{wang2015targeting}, the estimated $\alpha$ was close to one, with a
+narrower credible interval than for the other studies.
+\citeauthor{niculescu2015recent} studied a recent outbreak among Romanian
+\gls{IDU}, while \citeauthor{wang2015targeting} sampled acutely infected
+\gls{MSM} in Beijing, China. Both studies discovered a high degree of
+phylogenetic relatedness owing to the recent infection times and homogeneous
+risk groups of the studied populations. The estimated number of infections for
+these datasets were also quite low, although the \gls{HPD} interval for
+\citeauthor{wang2015targeting} was much wider than that for
+\citeauthor{niculescu2015recent}.
 
+The final studied dataset was an outlier in terms of estimated parameters.
 \textcite{novitsky2013phylogenetic} sampled approximately 44\% of the
@@
 any large clusters using patristic distance, and therefore chose to analyze a
-subtree instead. The estimated attachment power was extremely low, with a very
-wide credible interval. The estimated number of infected nodes was also
-extremely high, much higher than the HIV-infected population of the town, with
-a similarly wide credible interval. Several factors may have contributed to
-this result. First, the authors note that the their sample was 75\% female. In
-a primarily heterosexual risk environmet, removal of a disproportionate nmuber
-of males from the network could obfuscate the true network structure. Second,
-the town in question was in close proximity to the country's capital, and the
-authors indicated that a high amount of migration takes place between the two
-locations. This suggests a reason for the high estimate of $I$: the contact
-network includes a much larger group based in the capital city.
+subtree instead. Estimates of $\alpha$ and $N$ both had very wide \gls{HPD}
+intervals and were markedly different from the other datasets. The estimated
+number of infected nodes was also extremely high, much higher than the
+estimated HIV prevalence of the town. Several factors may have contributed to
+these results. First, the authors note that the their sample was 75\% female.
+In a primarily heterosexual risk environment, removal of a disproportionate
+number of males from the network could obfuscate the true network structure.
+Second, the town in question was in close proximity to the country's capital,
+and the authors indicated that a high amount of migration takes place between
+the two locations. This suggests that the contact network may include a much
+larger group based in the capital city, which would explain the high estimate
+of $I$.
 
@@
 to be a multiple of 2. Furthermore, we considered the case $m = 1$, where the
-network has no cycles, to be implausable and assigned it zero prior
+network has no cycles, to be implausible and assigned it zero prior
 probability. This forces the average degree to be at least four, which may be
-unrealistically high for sexual networks. Additional modelling assumptions,
-include the network being connected and static, the epidemic following simple
-\gls{SI} dynamics, and the underlying behaviour of all nodes is the same. This
+unrealistically high for sexual networks. Additional modelling assumptions
+include the network being connected and static, all transmission rates being
+equal, no removal after infection, and identical behaviour of all nodes. This
 last is particularly problematic, as we showed by simulating a network where
@@
 although we could characterize the network in aggregate, the estimated
-parameters could not be said to apply to any individual node. However, we note
-that, given the model-agnostic nature of our method, it is possible to fit a
-model with heterogeneous node behaviour, and indeed this may prove a fruitful
-avenue for future investigations.
+parameters could not be said to apply to any individual node. Despite these
+issues, we felt it was best to demonstrate the method first on a simple model.
+It is possible to fit more complex models which address some of these issues,
+such as one incorporating heterogeneous node behaviour. This may prove a
+fruitful avenue for future investigations.
 
-Our method itself has of caveats, perhaps the most significant being that it
+Our method has a number of caveats, perhaps the most significant being that it
 takes a transmission tree as input. In reality, true transmission trees are not
@@
 phylogeny and transmission tree can differ
-significantly~\autocite{ypma2013relating}. In addition, it is computationally
-intensive, taking about a day when run on 20 cores in parallel with the setings
-we described in the methods. However, there are no other methods available to
-perform the types of estimation we do here, and we therefore believe our method
-will be useful to epidemiological researchers interested in the network
-structure underlying disease outbreaks.
+significantly~\autocite{ypma2013relating}. In addition, the \gls{ABC}-\gls{SMC}
+algorithm is computationally intensive, taking about a day when run on 20 cores
+in parallel with the settings we described in the methods. Nevertheless, our
+method is potentially useful to epidemiological researchers interested in the
+general characteristics of the network structure underlying disease outbreaks.
 
@@
   \centering
-  %\includegraphics{smc_test}
+  \includegraphics{smc-test.pdf}
   \caption{Approximation of mixture of Gaussians used by

### 2016-03-22 doc/paper/paper.tex
changes from meeting today

@@
 $\alpha$, 
-  1 [0.07-
-                           1.07];
+  1 [0.75-
+                           1.12];
 $I$,
-  5.37 [2.05-
-                       5.82];
+  5.24 [2.05-
+                       5.81];
 $m$,
-  \ensuremath{1.01\times 10^{4}} [\ensuremath{3.06\times 10^{3}}-
+  \ensuremath{1.04\times 10^{4}} [\ensuremath{2.27\times 10^{3}}-
                               \ensuremath{1.5\times 10^{4}}];
 $N$,
-  \ensuremath{1.18\times 10^{3}} [505.71-
-                       \ensuremath{1.62\times 10^{3}}].
+  \ensuremath{1.17\times 10^{3}} [502.34-
+                       \ensuremath{1.58\times 10^{3}}].
 
@@
 $I$,
-  3.53 [2.18-
+  3.47 [2.18-
                        5.9];
 $m$,
-  \ensuremath{1.07\times 10^{4}} [\ensuremath{3.11\times 10^{3}}-
-                              \ensuremath{1.47\times 10^{4}}];
+  \ensuremath{1.08\times 10^{4}} [\ensuremath{3.11\times 10^{3}}-
+                              \ensuremath{1.48\times 10^{4}}];
 $N$,
-  \ensuremath{2.46\times 10^{3}} [\ensuremath{1.28\times 10^{3}}-
-                       \ensuremath{3.88\times 10^{3}}].
+  \ensuremath{2.94\times 10^{3}} [\ensuremath{1.28\times 10^{3}}-
+                       \ensuremath{3.94\times 10^{3}}].
 

### 2016-03-23 doc/paper/paper.Rnw
Rewrote results section

@@
 \usepackage{array}
+\usepackage{float}
+
+\newfloat{data}{hb}{lod}
+\floatname{data}{Data}
+\crefname{data}{data}{data}
 
@@
 \newcommand{\sett}[1]{\{#1\}}
+\newcommand{\ie}{\textit{i.e.}}
 \renewcommand{\emptyset}{\varnothing}
@@
 \newacronym{IDU}{IDU}{injection drug user}
+\newacronym{HIV}{HIV}{human immuno-deficiency virus}
 
@@
 mode of transmission of the disease. For airborne pathogens, physical proximity
-may be sufficient, but for sexually transmitted diseases, either sexual or
-blood-to-blood contact is required. The set of links between individuals along
-which transmission can occur is called the contact
-network~\autocite{klovdahl1985social, morris1993epidemiology}. The structure of
-the contact network underlying an epidemic can profoundly impact the speed and
-pattern of the epidemic's expansion. Network structure can influence the
-prevalence trajectory~\autocite{o2010contact} and epidemic
+may be sufficient, while for sexually transmitted diseases, sexual or in some
+cases blood-to-blood contact is required. The population together with the set
+of links between individuals along which transmission can occur is called the
+contact network~\autocite{klovdahl1985social, morris1993epidemiology}. The
+structure of the contact network underlying an epidemic can profoundly impact
+the speed and pattern of the epidemic's expansion. Network structure can
+influence the prevalence trajectory~\autocite{o2010contact} and epidemic
 threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates
@@
 Survey-based studies of sexual networks~\autocite{liljeros2001web,
-schneeberger2004scale} have found that sexual contact networks are best
-described by a preferential attachment model \autocite[although there has been
-some disagreement, see][]{jones2003assessment}. Under these models, nodes with
-a high number of contacts attract new connections at an elevated rate. Networks
-produced by preferential attachment have a power-law degree distribution,
-meaning that the number of nodes of degree $k$ is proportional to $k^\gamma$
-for some constant $\gamma$. These networks are also referred to as
-``scale-free''. The first contact network model incorporating preferential
-attachment was introduced by \textcite{barabasi1999emergence}, and is now
-referred to as the \gls{BA} model. Under this model, networks are formed by
-iteratively adding nodes with $m$ new edges each. These new edges are joined to
-existing nodes of degree $k$ with probability proportional to $k^\alpha$, so
-that nodes of high degree tend to attract more connections (in the original
-paper, only $\alpha = 1$ was investigated).
+schneeberger2004scale, colgate1989risk} have found that these networks have a
+degree distribution which follows a power law \autocite[although there has been
+some disagreement, see][]{handcock2004likelihood}. That is, number of nodes of
+degree $k$ is proportional to $k^-\gamma$ for some constant $\gamma$. These
+networks are also referred to as ``scale-free''. One process by which
+scale-free networks can be generated is preferential attachment, where nodes
+with a high number of contacts attract new connections at an elevated rate. The
+first contact network model incorporating preferential attachment was
+introduced by \textcite{barabasi1999emergence}, and is now referred to as the
+\gls{BA} model. Under this model, networks are formed by iteratively adding
+nodes with $m$ new edges each. In the most commonly studied formulation, these
+new edges are joined to existing nodes of degree $k$ with probability
+proportional to $k$, so that nodes of high degree tend to attract more
+connections. \citeauthor{barabasi1999emergence} suggested an extension where
+the probability of attaching to a node of degree $k$ is $k^\alpha$ for some
+non-negative constant $\alpha$, and we use this extension here.
 
@@
 \textcite{groendyke2011bayesian} and applied to a much larger data set of 188
-individuals. \textcite{brown2011transmission} analysed the degree distribution
-of an approximate transmission network, estimated based on genetic similarity
-and estimated times of infection, relating 60\% of HIV infected \gls{MSM} in
+individuals. \textcite{volz2007susceptible, volz2008sir} developed differential
+equations describing the spread of a \gls{SI} epidemic on static and dynamic
+contact networks with several degree distributions, which could in principle be
+used for inference if observed incidence trajectories were available.
+\textcite{brown2011transmission} analysed the degree distribution of an
+approximate transmission network, estimated based on genetic similarity and
+estimated times of infection, relating 60\% of \gls{HIV}-infected \gls{MSM} in
 the United Kingdom. The transmission network is a subgraph of the contact
@@
 infection. The authors found that a Waring distribution, which is produced by a
-different preferential attachment model, was a good fit to their estimated
-network.
+more sophisticated preferential attachment model, was a good fit to their
+estimated network.
 
@@
 calculation of a normalizing constant, Bayesian inference is often performed
-using \gls{MCMC}. Rather than calculating explicit likeilihoods, \gls{MCMC}
-uses likelihood \emph{ratios} in which the normalizing constant cancel out.
-Unfortunately, it is generally difficult to explicitly calculate the likelihood
-of an observed transmission tree under a contact network model, even up to a
-normalizing constant. To do so, it would be necessary to integrate over all
-possible networks, and also over all possible labellings of the internal nodes
-of the transmission tree. While it is not known (to us) whether such
-integration is tractable, a simpler alternative is offered by likelihood-free
-methods, namely \gls{ABC}. \gls{ABC} leverages the fact that, although
-calculating the likelihood may be impossible, generating simulated datasets
-according to a model is often straightforward. If our model fits the data well,
-the simulated data it produces should be similar to the observed data. More
-formally, if $D$ is the observed data, the posterior distribution $f(\theta
-\mid D)$ on model parameters $\theta$ is replaced as the target of statistical
-inference by $f(\theta \mid \rho(\hat{D}, D) < \varepsilon)$, where $\rho$ is a
-distance function, $\hat{D}$ is a simulated dataset according to $\theta$, and
-$\varepsilon$ is a small tolerance~\autocite{sunnaker2013approximate}. In the
-specific case when $\rho$ is a kernel function, the approach is known as
+using \gls{MCMC}, which uses likelihood \emph{ratios} in which the normalizing
+constants cancel out. Unfortunately, it is generally difficult to explicitly
+calculate the likelihood of an observed transmission tree under a contact
+network model, even up to a normalizing constant. To do so, it would be
+necessary to integrate over all possible networks, and also over all possible
+labellings of the internal nodes of the transmission tree. While it is not
+known (to us) whether such integration is tractable, a simpler alternative is
+offered by likelihood-free methods, namely \gls{ABC}. \gls{ABC} leverages the
+fact that, although calculating the likelihood may be impractical, generating
+simulated datasets according to a model is often straightforward. If our model
+fits the data well, the simulated data it produces should be similar to the
+observed data. More formally, if $D$ is the observed data, the posterior
+distribution $f(\theta \mid D)$ on model parameters $\theta$ is replaced as the
+target of statistical inference by $f(\theta \mid \rho(\hat{D}, D) <
+\varepsilon)$, where $\rho$ is a distance function, $\hat{D}$ is a simulated
+dataset according to $\theta$, and $\varepsilon$ is a small
+tolerance~\autocite{sunnaker2013approximate}. In the specific case when $\rho$
+is a kernel function, the approach is known as
 kernel-\gls{ABC}~\autocite{nakagome2013kernel, poon2015phylodynamic}.
@@
 tree. Nodes in our networks followed simple \gls{SI} dynamics, meaning that
-they became infected at a rate proportional to their numbers of infected
+they became infected at a rate proportional to their number of infected
 neighbours, and never recovered. For all analyses, the transmission trees'
@@
 cross-validations, one for each of the \gls{BA} model parameters $\alpha$, $I$,
-$m$, and $N$.
-
-We repeated these four cross-validations with different values of $\lambda$
-(0.2, 0.3, and 0.4) and $\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$), as well
-as on trees with differing numbers of tips (100, 500, and 1000) and in
-epidemics of differing size $I$ (500, 1000, and 2000). When evaluating the
+$m$, and $N$. We repeated these four cross-validations with different values of
+$\lambda$ (0.2, 0.3, and 0.4) and $\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$),
+as well as on trees with differing numbers of tips (100, 500, and 1000) and in
+epidemics of differing size (500, 1000, and 2000). When evaluating the
 classifier for $I$, we did not consider trees with 1000 tips, because one of
@@
 to each tree with these priors: $m \sim$ Uniform(1, 5), $\alpha \sim$
-Uniform(0, 2), and $(N, I)$ jointly uniform on the triangular region \{$500
-\leq N \leq 15000$, $500 \leq I \leq 5000$, $I \leq N$\}. Following
+Uniform(0, 2), and $(N, I)$ jointly uniform on the region \{$500 \leq N \leq
+15000$, $500 \leq I \leq 5000$, $I \leq N$\}. Following
 \textcite{del2012adaptive} and \textcite{beaumont2009adaptive}, all proposals
@@
 using the \software{density} function with the default parameters in
-\software{R}. \gls{HPD} intervals were calculated with the
+\software{R}. \Gls{HPD} intervals were calculated with the
 \software{HPDinterval} function from the \software{R} package
@@
 HXB2 reference sequence (Genbank accession number HIVHXB2CG) and the
-hypervariable regions were clipped out with BioPython version
+hypervariable regions were clipped out with \software{BioPython} version
 1.66+~\autocite{cock2009biopython}. Sequences were multiply aligned using
@@
 Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
-tested~\cref{fig:rsquared}. Classifiers based on two other tree statistics, the
-\gls{nLTT} and Sackin's index, generally exhibited worse performance than the
-tree kernel, although the magnitude of the disparity varied between the
-parameters. \Cref{fig:rsquared} shows the cross-validation accuracy with
-$\lambda$ = 0.3, $\sigma$ = 4, $I$ = 1000, and trees of size 500. The results
-were largely robust to variations in $\lambda$ and $\sigma$, although accuracy
-varied between different epidemic and sampling scenarios
+tested (\cref{fig:rsquared}, left). Classifiers based on two other tree
+statistics, the \gls{nLTT} and Sackin's index, generally exhibited worse
+performance than the tree kernel, although the magnitude of the disparity
+varied between the parameters (\cref{fig:rsquared}, centre and right).
+\Cref{fig:rsquared} shows the cross-validation accuracy of each of the tree
+classifiers. The results were largely robust to variations in the tree kernel
+meta-parameters $\lambda$ and $\sigma$, although accuracy varied between
+different epidemic and sampling scenarios
 (\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}).
 
-The kSVM classifier for $\alpha$ had an average $R^2$ of
-    \Sexpr{kernel.alpha[,round(mean(rsquared), 2)]} 
-(\cref{fig:alphacrossv}). The $R^2$ of the for the SVM classifer for $\alpha$
-using the \gls{nLTT} varied between
-\Sexpr{nltt.alpha[,round(min(rsquared), 2)]}
-and
-  \Sexpr{nltt.alpha[,round(max(rsquared), 2)]}.
-The $R^2$ of the linear regression against Sackin's index varied between
-  \Sexpr{sackin.alpha[,round(min(rsquared), 2)]}
-and
-  \Sexpr{sackin.alpha[,round(max(rsquared), 2)]}.
-The $I$ parameter was also classified accurately by the tree kernel
-(\cref{fig:Icrossv}), with an average
+When classifying $\alpha$, the kernel-SVM classifier had an average $R^2$ of 
+    \Sexpr{kernel.alpha[,round(mean(rsquared), 2)]},
+compared to 
+    \Sexpr{nltt.alpha[,round(mean(rsquared), 2)]}
+for the \gls{nLTT}-based SVM, and
+    \Sexpr{sackin.alpha[,round(mean(rsquared), 2)]}
+for the linear regression against Sackin's index. There was little variation
+about the mean for different tree and epidemic sizes. No classifier could
+accurately identify the $m$ parameter in any epidemic scenario, with average
+$R^2$ values of 
+  \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for \gls{kSVM},
+  \Sexpr{nltt.m[,round(mean(rsquared), 2)]} for the \gls{nLTT}, and
+  \Sexpr{sackin.m[,round(mean(rsquared), 2)]}
+for Sackin's index. Again, there was little variation in accuracy between
+epidemic scenarios, although the accuracy of the \gls{kSVM} was slightly higher
+on 1000-tip trees.
+
+The accuracy of classifiers $I$ varied significantly with the number of tips in
+the tree. For 100-tip trees, the average $R^2$ values were
+  \Sexpr{kernel.I[ntip == 100, round(mean(rsquared), 2)]},
+  \Sexpr{nltt.I[ntip == 100, round(mean(rsquared), 2)]}, and
+  \Sexpr{sackin.I[ntip == 100, round(mean(rsquared), 2)]}
+for the tree kernel, \gls{nLTT}, and Sackin's index respectively. For 500-tip
+trees, the values increased to
+  \Sexpr{kernel.I[ntip == 500, round(mean(rsquared), 2)]},
+  \Sexpr{nltt.I[ntip == 500, round(mean(rsquared), 2)]}, and
+  \Sexpr{sackin.I[ntip == 500, round(mean(rsquared), 2)]}.
+Finally, the performance of classifiers for $N$ depended heavily on the
+epidemic scenario. The $R^2$ of the \gls{kSVM} classifier ranged from
+  \Sexpr{kernel.N[,round(min(rsquared), 2)]}
+for the smallest epidemic and smallest sample size, to
+  \Sexpr{kernel.N[,round(max(rsquared), 2)]}
+for the largest. Likewise, $R^2$ for the \gls{nLTT}-based SVM ranged from 
+  \Sexpr{nltt.N[,round(min(rsquared), 2)]}
+to
+  \Sexpr{nltt.N[,round(max(rsquared), 2)]}.
+Sackin's index did not accurately classify $N$ in any scenario, with an average
 $R^2$ of
-  \Sexpr{kernel.I[ntip == 500, round(mean(rsquared), 2)]} 
-for 500-tip trees or 
-  \Sexpr{kernel.I[ntip == 100, round(mean(rsquared), 2)]} 
-for 100-tip trees. The \gls{nLTT} performed almost as well on this parameter as
-the tree kernel, with $R^2$ values of
-  \Sexpr{nltt.I[ntip == 500, round(mean(rsquared), 2)]}
-for 500-tip trees or
-  \Sexpr{nltt.I[ntip == 100, round(mean(rsquared), 2)]}
-for 100-tip trees. Sackin's index was an extremely poor classifier for this
-parameter, with both $R^2$ values below
-  \Sexpr{sackin.I[,round(max(rsquared) + 0.01, 2)]}.
-
-The $m$ parameter was much harder to classify (\cref{fig:mcrossv}), with $R^2$
-for the kSVM varying between 
-  \Sexpr{kernel.m[,round(min(rsquared), 3)]}
-and
-  \Sexpr{kernel.m[,round(max(rsquared), 2)]}
-depending on $I$ and the tree size. The \gls{nLTT} was also universally poor,
-with $R^2$ below
-  \Sexpr{nltt.m[,round(max(rsquared) + 0.01, 2)]}
-for all scenarios. Sackin's index fared slightly better than the \gls{nLTT},
-especially on trees with a higher number of tips, with $R^2$ between
-  \Sexpr{sackin.m[,round(min(rsquared), 2)]}
-and
-  \Sexpr{sackin.m[,round(max(rsquared), 2)]}.
-Finally, the accuracy of the kSVM classifier for $N$ varied widely
-(\cref{fig:Ncrossv}), from 
-  \Sexpr{kernel.N[,round(min(rsquared), 2)]}
-to
-  \Sexpr{kernel.N[,round(max(rsquared), 2)]}.
-The \gls{nLTT} performed well in classifying $N$ when the epidemic was large,
-improving upon the tree kernel in the scenarios with $I$ = 2000 and 100- or
-500-tip transmission trees ($R^2$ = 
-  \Sexpr{nltt.N[which(I == 2000 & ntip == 100), round(mean(rsquared), 2)]}
-and
-  \Sexpr{nltt.N[which(I == 2000 & ntip == 500), round(mean(rsquared), 2)]}),
-although it performed poorly when $I$ was 500 or 1000 (average $R^2$ =
-  \Sexpr{nltt.N[which(I < 2000), round(mean(rsquared), 2)]}).
-Sackin's index could not classify $N$, with all $R^2$ below
-  \Sexpr{sackin.N[,round(max(rsquared) + 0.01, 2)]}.
-Based on inspection of the cross-validation results, we chose to use the
-meta-parameters $\lambda = 0.3$ and $\sigma = 4$ in further analyses, and not
-to use Sackin's index or the \gls{nLTT}.
+  \Sexpr{sackin.N[,round(mean(rsquared), 2)]}
+and little variation between scenarios.
 
@@
       the parameter being tested. Vertical lines are empirical 95\% confidence
-      intervals based on 1000 2-fold cross-validations.
+      intervals based on 1000 two-fold cross-validations.
   }
@@
                                 d[true_alpha == 0.5 | true_alpha == 1.5, m_error])
-
-    options(scipen=-1, digits=2)
+    options(scipen=4)
+    stopifnot(alpha.av["factor(true_m)", "Pr(>F)"] > 0.05)
+    stopifnot(alpha.av["factor(true_I)", "Pr(>F)"] > 0.05)
+    stopifnot(I.I.test$p.value < 1e-5)
+    stopifnot(I.av["factor(true_m)", "Pr(>F)"] > 0.05)
+    stopifnot(m.av["factor(true_m)", "Pr(>F)"] > 0.05)
+    stopifnot(m.av["factor(true_I)", "Pr(>F)"] > 0.05)
+    stopifnot(min(N.av[1:3,"Pr(>F)"]) > 0.05)
 @
 
-Of the four parameters, $\alpha$ was the most accurately estimated, with a
+\Cref{fig:abcpt} shows \gls{MAP} point estimates of the \gls{BA} model
+parameters obtained with kernel-\gls{ABC} on simulated data. The estimates
+shown correspond only to the simulations where the $m$ parameter was set to 2,
+however the results for $m = 3$ and $m = 4$ were similar 
+(\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% \gls{HPD}
+intervals are given in \cref{tab:abchpd}.
+
+The accuracy of the parameter estimates obtained with kernel-\gls{ABC}
+paralleled the results from the \gls{kSVM} classifier. Of the four parameters,
+$\alpha$ was the most accurately estimated, with point estimates having a
 median [IQR] absolute error of 
-    \Sexpr{d[,median(alpha_error)]} 
-    [\Sexpr{d[,quantile(alpha_error, 0.25)]}-\Sexpr{d[,quantile(alpha_error, 0.75)]}].
-\Cref{fig:abcpt} shows point estimates for all simulations with $m = 2$. The
-results $m = 3$ and $m = 4$ were similar (\cref{fig:abcptm3,fig:abcptm4}).
-Average boundaries of 95\% \gls{HPD} intervals are given in \cref{tab:abchpd}.
-The accuracy of the estimates was not significantly different between values of
-$m$ or $I$ (both one-way ANOVA,
-    $p$ = \Sexpr{alpha.av["factor(true_m)", "Pr(>F)"]}
-and 
-    \Sexpr{alpha.av["factor(true_I)", "Pr(>F)"]}),
-although the errors when the true value of $\alpha$ was zero were
-significantly greater than the other values 
-    (Wilcoxon rank-sum test, $p$ = \Sexpr{zero.alpha.test$p.value}).
-The error in the estimated value of $I$ was
-    \Sexpr{d[,median(I_error)]} 
-    [\Sexpr{d[,quantile(I_error, 0.25)]}-\Sexpr{d[,quantile(I_error, 0.75)]}].
-Errors were significantly higher for $\alpha \geq 1$
-    (Wilcoxon rank-sum test, $p$ = \Sexpr{alpha.I.test$p.value})
-and for $I$ = 2000
-    ($p$ = \Sexpr{I.I.test$p.value}),
-but not for any values of $m$
-    (one-way ANOVA, $p$ = \Sexpr{I.av["factor(true_m)", "Pr(>F)"]}).
-The $m$ parameter was estimated correctly in
+    \Sexpr{d[,round(median(alpha_error), 2)]} 
+    [\Sexpr{d[,round(quantile(alpha_error, 0.25), 2)]} - 
+    \Sexpr{d[,round(quantile(alpha_error, 0.75), 2)]}].
+The errors when the true value of $\alpha$ was zero were significantly greater
+than those for the other values 
+    (Wilcoxon rank-sum test, $p$ = \Sexpr{round(zero.alpha.test$p.value, 4)}),
+but did not vary across the true values of the other parameters (one-way
+ANOVA). Estimates for $I$ were also relatively accurate, with point estimate
+errors of
+    \Sexpr{d[,round(median(I_error))]} 
+    [\Sexpr{d[,round(quantile(I_error, 0.25))]} - 
+    \Sexpr{d[,round(quantile(I_error, 0.75))]}].
+These errors were significantly higher when the true value of $\alpha$ was
+at least 1
+    (Wilcoxon rank-sum test, $p$ = \Sexpr{round(alpha.I.test$p.value, 4)})
+and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
+did not affect the estimates of $I$ (one-way ANOVA).
+
+The $m$ parameter was estimated correctly in only
     \Sexpr{as.integer(m.tbl[1] * 100)} \%
-of simulations, with an error of one in
-    \Sexpr{as.integer(m.tbl[2] * 100)} \%
-and of two or more in 
-    \Sexpr{as.integer(sum(m.tbl[3:length(m.tbl)]) * 100)} \%
-(the only possible $m$ values were 1, 2, 3, 4, or 5). The true values of $m$
-and $I$ did not significantly affect the error
-    (one-way ANOVA, $p$ = \Sexpr{m.av["factor(true_m)", "Pr(>F)"]} and
-                          \Sexpr{m.av["factor(true_I)", "Pr(>F)"]}).
-Finally, the total number of nodes $N$ was consistently over-estimated by about
-a factor of two
-    (error \Sexpr{d[,median(N_error)]} 
-    [\Sexpr{d[,quantile(N_error, 0.25)]}-\Sexpr{d[,quantile(N_error, 0.75)]}]).
-No other parameters influenced the accuracy of the $N$ estimates 
-    (one-way ANOVA, $p \geq$ \Sexpr{min(N.av[1:3,"Pr(>F)"])}).
+of simulations. Oddly, the error in the estimated $m$ was higher for integer
+values of $\alpha$ (\ie~0 and 1) than non-integer values 
+    (Wilcoxon rask-sum test, $p$ = \Sexpr{round(alpha.m.test$p.value, 3)}).
+The true values of the other parameters did not significantly affect the
+estimates of $m$ (both one-way ANOVA). Finally, the total number of nodes $N$
+was consistently over-estimated by about a factor of two
+    (error \Sexpr{d[,round(median(N_error))]} 
+    [\Sexpr{d[,round(quantile(N_error, 0.25))]} - 
+     \Sexpr{d[,round(quantile(N_error, 0.75))]}]).
+No parameters influenced the accuracy of the $N$ estimates (all one-way ANOVA).
 
@@
   \input{\tablepath/abc-hpd}
-  \caption{Average \gls{HPD} interval widths for \gls{ABC} model parameter
-           estimates.}
+  \caption{
+    \glsreset{HPD}
+    Average \acrlong{MAP} point estimates and 95\% \gls{HPD} interval widths
+    for \gls{BA} model parameter estimates obtained with kernel-\gls{ABC}.
+    Three transmission trees were simulated under each combination of the
+    listed parameter values, and the parameters were estimated with
+    kernel-\gls{ABC} without training.
+  }
   \label{tab:abchpd}
@@
 The dispersion of the \gls{ABC} approximation to the posterior also varied
-between the parameters, with narrower \gls{HPD} intervals for those with the
-most accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows the
-distributions for for one simulation (equivalent plots for all the simulations
-can be found in supplemental data). \Gls{HPD} intervals around $\alpha$ and $I$
-were narrow relative to the region of nonzero prior density, whereas the
-intervals for $m$ and $N$ were more widely dispersed.
+between the parameters, with narrower \gls{HPD} intervals for the parameters
+with more accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows
+the distributions for for one simulation (equivalent plots for all the
+simulations can be found in \cref{dat:abcpost}). \Gls{HPD} intervals around
+$\alpha$ and $I$ were narrow relative to the region of nonzero prior density,
+whereas the intervals for $m$ and $N$ were more widely dispersed.
 
@@
   \includegraphics[width=\textwidth]{abc-posterior-example.pdf}
-  \caption{Marginal estimated posterior distributions obtained with ABC on a
-  single simulated dataset. Dotted lines and shaded polygon indicate true
-  values.}
+  \caption{
+    Marginal posterior distributions of \gls{BA} model parameters estimated
+    with kernel-\gls{ABC} for a single simulated transmission tree. Dotted
+    lines and shaded polygon indicate true values.
+  }
   \label{fig:abcex}
@@
 
-<<mixed_alpha, include=FALSE>>=
+<<mixed_peerdriven, include=FALSE>>=
   library(coda)
@@
   d <- d[sample(1:nrow(d), prob=weight, replace=TRUE)]
-  params <- c("alpha", "I", "m", "N")
+  params <- c("alpha", "m", "N", "I")
   setnames(d, c("attach_power", "edges_per_vertex", "nodes", "sim_nodes"),
@@
   setkey(d, parameter)
+
+  p <- fread("bzcat ../../simulations/abc-pa-peerdriven/abc/*")
+  p <- p[iter == max(iter)]
+  p <- p[sample(1:nrow(p), prob=weight, replace=TRUE)]
+  params <- c("alpha", "m", "N", "I")
+  setnames(p, c("attach_power", "edges_per_vertex", "nodes", "sim_nodes"),
+           params)
+  p <- melt(p, measure.vars=params, variable.name="parameter")
+  f <- function (x) {
+    dens <- density(x)
+    hpd <- HPDinterval(mcmc(x))
+    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
+         hpd.max=hpd[,"upper"])
+  }
+  p <- p[,f(value), by=parameter]
+  setkey(p, parameter)
 @
@@
 $\alpha$, 
-  \Sexpr{d["alpha", map]} [\Sexpr{d["alpha", hpd.min]}-
-                           \Sexpr{d["alpha", hpd.max]}];
+  \Sexpr{round(d["alpha", map], 2)} 
+  [\Sexpr{round(d["alpha", hpd.min], 2)}-
+   \Sexpr{round(d["alpha", hpd.max], 2)}];
 $I$,
-  \Sexpr{d["I", map]} [\Sexpr{d["I", hpd.min]}-
-                       \Sexpr{d["I", hpd.max]}];
+  \Sexpr{d["I", round(map)]} 
+  [\Sexpr{d["I", round(hpd.min)]} -
+   \Sexpr{d["I", round(hpd.max)]}];
 $m$,
-  \Sexpr{d["m", floor(map)]} [\Sexpr{d["m", floor(hpd.min)]}-
+  \Sexpr{d["m", floor(map)]} 
+  [\Sexpr{d["m", floor(hpd.min)]} -
    \Sexpr{d["m", floor(hpd.max)]}];
 $N$,
-  \Sexpr{d["N", map]} [\Sexpr{d["N", hpd.min]}-
-                       \Sexpr{d["N", hpd.max]}].
-
-<<peerdriven, include=FALSE>>=
-  d <- fread("bzcat ../../simulations/abc-pa-peerdriven/abc/*")
-  d <- d[iter == max(iter)]
-  d <- d[sample(1:nrow(d), prob=weight, replace=TRUE)]
-  params <- c("alpha", "I", "m", "N")
-  setnames(d, c("attach_power", "edges_per_vertex", "nodes", "sim_nodes"),
-           params)
-  d <- melt(d, measure.vars=params, variable.name="parameter")
-  f <- function (x) {
-    dens <- density(x)
-    hpd <- HPDinterval(mcmc(x))
-    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
-         hpd.max=hpd[,"upper"])
-  }
-  d <- d[,f(value), by=parameter]
-  setkey(d, parameter)
-@
-
+  \Sexpr{d["N", round(map)]} 
+  [\Sexpr{d["N", round(hpd.min)]}-
+   \Sexpr{d["N", round(hpd.max)]}].
 To test the effect of sampling bias, we sampled one transmission tree in a
@@
 $\alpha$, 
-  \Sexpr{d["alpha", map]} [\Sexpr{d["alpha", hpd.min]}-
-                           \Sexpr{d["alpha", hpd.max]}];
+  \Sexpr{p["alpha", round(map, 2)]} 
+  [\Sexpr{p["alpha", round(hpd.min, 2)]} -
+   \Sexpr{p["alpha", round(hpd.max, 2)]}];
 $I$,
-  \Sexpr{d["I", map]} [\Sexpr{d["I", hpd.min]}-
-                       \Sexpr{d["I", hpd.max]}];
+  \Sexpr{p["I", floor(map)]} 
+  [\Sexpr{p["I", floor(hpd.min)]} -
+   \Sexpr{p["I", floor(hpd.max)]}];
 $m$,
-  \Sexpr{d["m", floor(map)]} [\Sexpr{d["m", floor(hpd.min)]}-
-                              \Sexpr{d["m", floor(hpd.max)]}];
+  \Sexpr{p["m", floor(map)]} 
+  [\Sexpr{p["m", floor(hpd.min)]} -
+   \Sexpr{p["m", floor(hpd.max)]}];
 $N$,
-  \Sexpr{d["N", map]} [\Sexpr{d["N", hpd.min]}-
-                       \Sexpr{d["N", hpd.max]}].
+  \Sexpr{p["N", floor(map)]} 
+  [\Sexpr{p["N", floor(hpd.min)]} -
+   \Sexpr{p["N", floor(hpd.max)]}].
+Both of these results were in line with estimates obtained on other simulated
+datasets (\cref{tab:abchpd}).
 
@@
 $\alpha$ values near unity (\gls{MAP} estimate [95\% \gls{HPD}] 
-  \Sexpr{map["niculescu2015", alpha]} 
-  [\Sexpr{low["niculescu2015", alpha]}-\Sexpr{up["niculescu2015", alpha]}]
+  \Sexpr{map["niculescu2015", round(alpha, 2)]} 
+  [\Sexpr{low["niculescu2015", round(alpha, 2)]} - 
+   \Sexpr{up["niculescu2015", round(alpha, 2)]}]
 and
-  \Sexpr{map["wang2015", alpha]} 
-  [\Sexpr{low["wang2015", alpha]}-\Sexpr{up["wang2015", alpha]}]). Another two
-datasets~\autocite{li2015hiv, cuevas2009hiv} had lower estimated values and wider
-\gls{HPD} intervals
-  (\Sexpr{map["li2015", alpha]} 
-  [\Sexpr{low["li2015", alpha]}-\Sexpr{up["li2015", alpha]}]
+  \Sexpr{map["wang2015", round(alpha, 2)]} 
+  [\Sexpr{low["wang2015", round(alpha, 2)]} -
+   \Sexpr{up["wang2015", round(alpha, 2)]}]). 
+Another two datasets~\autocite{li2015hiv, cuevas2009hiv} had lower estimated
+values and wider \gls{HPD} intervals
+  (\Sexpr{map["li2015", round(alpha, 2)]} 
+  [\Sexpr{low["li2015", round(alpha, 2)]} - 
+  \Sexpr{up["li2015", round(alpha, 2)]}]
 and
-  \Sexpr{map["cuevas2009", alpha]} 
-  [\Sexpr{low["cuevas2009", alpha]}-\Sexpr{up["cuevas2009", alpha]}]). 
+  \Sexpr{map["cuevas2009", round(alpha, 2)]} 
+  [\Sexpr{low["cuevas2009", round(alpha, 2)]} -
+   \Sexpr{up["cuevas2009", round(alpha, 2)]}]). 
 The \textcite{novitsky2014impact} data had an extremely low estimated $\alpha$
 and a very wide \gls{HPD} interval
-  \Sexpr{map["novitsky2014", alpha]} 
-  [\Sexpr{low["novitsky2014", alpha]}-\Sexpr{up["novitsky2014", alpha]}]). 
-For all the datasets except \textcite{novitsky2014impact}, estimated values of
+  (\Sexpr{map["novitsky2014", round(alpha, 2)]} 
+  [\Sexpr{low["novitsky2014", round(alpha, 2)]} -
+   \Sexpr{up["novitsky2014", round(alpha, 2)]}]). 
+For all the datasets except \citeauthor{novitsky2014impact}, estimated values of
 $I$ were below 2000, with narrow \gls{HPD} intervals around two of the
 datasets
-  (\textcite{cuevas2009hiv}, \Sexpr{map["cuevas2009", I]} 
-  [\Sexpr{low["cuevas2009", I]}-\Sexpr{up["cuevas2009", I]}];
-   \textcite{niculescu2015recent}, \Sexpr{map["niculescu2015", I]}
-  [\Sexpr{low["niculescu2015", I]}-\Sexpr{up["niculescu2015", I]}])
+  (\citeauthor{cuevas2009hiv}, \Sexpr{map["cuevas2009", floor(I)]} 
+  [\Sexpr{low["cuevas2009", floor(I)]} -
+   \Sexpr{up["cuevas2009", floor(I)]}];
+   \citeauthor{niculescu2015recent}, \Sexpr{map["niculescu2015", floor(I)]}
+  [\Sexpr{low["niculescu2015", floor(I)]} - 
+   \Sexpr{up["niculescu2015", floor(I)]}])
 and wider intervals around the other two
-  (\textcite{li2015hiv}, \Sexpr{map["li2015", I]} 
-  [\Sexpr{low["li2015", I]}-\Sexpr{up["li2015", I]}];
-   \textcite{wang2015targeting}, \Sexpr{map["wang2015", I]}
-  [\Sexpr{low["wang2015", I]}-\Sexpr{up["wang2015", I]}]).
-The \textcite{novitsky2014impact} data was again the outlier, with a very high
+  (\citeauthor{li2015hiv}, \Sexpr{map["li2015", floor(I)]} 
+  [\Sexpr{low["li2015", floor(I)]} -
+   \Sexpr{up["li2015", floor(I)]}];
+   \citeauthor{wang2015targeting}, \Sexpr{map["wang2015", floor(I)]}
+  [\Sexpr{low["wang2015", floor(I)]} - 
+   \Sexpr{up["wang2015", floor(I)]}]).
+The \citeauthor{novitsky2014impact} data was again the outlier, with a very high
 estimated $I$, and \gls{HPD} interval spanning almost the entire prior region
-  (\Sexpr{map["novitsky2014", I]} 
-  [\Sexpr{low["novitsky2014", I]}-\Sexpr{up["novitsky2014", I]}]).
-Little information was gleaned about the $m$ parameter, with the \gls{HPD}
-interval occupying the entire prior region for all datasets. The estimates of
-$N$ were similarly uninformative, with the exception that the point estimate
-for the \textcite{wang2015targeting} data was smaller
-  (\Sexpr{map["wang2015", N]})
-than the estimates for other datasets.
+  (\Sexpr{map["novitsky2014", floor(I)]} 
+  [\Sexpr{low["novitsky2014", floor(I)]} -
+   \Sexpr{up["novitsky2014", floor(I)]}]).
+No information was gleaned about the $m$ parameter, with the \gls{HPD} interval
+occupying the entire prior region for all datasets. The estimates of $N$ were
+similarly uninformative, with the exception that the point estimate for the
+\citeauthor{wang2015targeting} data was smaller
+  (\Sexpr{map["wang2015", floor(N)]})
+than the estimates for other datasets
+ (average \Sexpr{map[dataset != "wang2015", floor(mean(N))]}).
 
@@
   \caption{
-    Point estimates and 95\% \gls{HPD} intervals for \gls{BA} network
-    parameters of published HIV datasets, obtained with kernel-ABC.
+    \acrlong{MAP} point estimates and 95\% \gls{HPD} intervals for parameters
+    of the \gls{BA} network model, fitted to five published \gls{HIV} datasets
+    with kernel-\gls{ABC}.
   }
@@
 network parameters in a phylodynamic framework~\autocite{groendyke2011bayesian,
-brown2011transmission}. Kernel-ABC is a model-agnostic method which can be used
-to investigate any quantity that affects tree shape. In this work, we developed
-a kernel-ABC-based method to infer the parameters of a contact network model.
-The method is general, and could be applied to any model from which contact
-networks can be simulated. We demonstrated the method on the \gls{BA} model,
-which is a simple model incorporating the preferential attachment feature
-commonly seen in real-world networks.
+volz2008sir, brown2011transmission}. Kernel-ABC is a model-agnostic method
+which can be used to investigate any quantity that affects tree shape. In this
+work, we developed a kernel-ABC-based method to infer the parameters of a
+contact network model. The method is general, and could be applied to any model
+from which contact networks can be simulated. We demonstrated the method on the
+\gls{BA} model, which is a simple model giving rise to the power law degree
+distributions commonly observed in real-world networks.
 
@@
 parameters may translate to highly informative priors on quantities of
-interest. We observed an exponential relationship between the preferential
-attachment power $\alpha$ and the power-law exponent $\gamma$
+interest. We observed a non-linear relationship between the preferential
+attachment power $\alpha$ and the power law exponent $\gamma$
 (\cref{fig:gamma}). Therefore, placing a uniform prior on $\alpha$ between 0
@@
 In a primarily heterosexual risk environment, removal of a disproportionate
-number of males from the network could obfuscate the true network structure.
-Second, the town in question was in close proximity to the country's capital,
-and the authors indicated that a high amount of migration takes place between
-the two locations. This suggests that the contact network may include a much
-larger group based in the capital city, which would explain the high estimate
-of $I$.
+number of males from the network could obfuscate the true network structure,
+for example if there were a disproportionate number of highly connected nodes
+of one gender. Second, the town in question was in close proximity to the
+country's capital, and the authors indicated that a high amount of migration
+takes place between the two locations. This suggests that the contact network
+may include a much larger group based in the capital city, which would explain
+the high estimate of $I$.
 
@@
 phylogeny and transmission tree can differ
-significantly~\autocite{ypma2013relating}. In addition, the \gls{ABC}-\gls{SMC}
-algorithm is computationally intensive, taking about a day when run on 20 cores
-in parallel with the settings we described in the methods. Nevertheless, our
-method is potentially useful to epidemiological researchers interested in the
-general characteristics of the network structure underlying disease outbreaks.
+significantly~\autocite{ypma2013relating} due to within-host evolution and the
+sampling process. In addition, the \gls{ABC}-\gls{SMC} algorithm is
+computationally intensive, taking about a day when run on 20 cores in parallel
+with the settings we described in the methods. Nevertheless, our method is
+potentially useful to epidemiological researchers interested in the general
+characteristics of the network structure underlying disease outbreaks. This
+work, and previous work by our group~\autocite{poon2015phylodynamic}, has
+demonstrated that kernel-ABC is a broadly applicable and effective framework in
+which to perform phylodynamic inference.
 
@@
 \renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{data}{0}
+\renewcommand{\thedata}{S\arabic{data}}
 
@@
 
-\begin{figure}[ht]
+\begin{figure}[H]
   \centering
   \includegraphics{leventhal2012fig1.pdf}
-  \caption{Reproduction of Figure 1A from \textcite{leventhal2012inferring}.}
+  \caption{
+      Reproduction of Figure 1A from \textcite{leventhal2012inferring} used to
+      check the accuracy of our implementation of Gillespie simulation.
+      Transmission trees were simulated over three types of network, with
+      pathogen transmissimility varying from 0 to 1. Sackin's index was
+      calculated for each simulated transmission tree.
+  }
   \label{fig:sf1}
@@
 
-\begin{figure}[ht]
+\begin{figure}[H]
   \centering
@@
     \gls{SMC}. Solid black line indicates true distribution. Grey shaded area
-    shows SMC approximation obtained with our implementation.}
+    shows SMC approximation obtained with our implementation, using 10000
+    particles with one simulated data point per particle.
+  }
   \label{fig:smctest}
@@
       Relationship between preferential attachment power parameter $\alpha$
-      and power-law exponent $\gamma$ for networks simulated under the \gls{BA}
+      and power law exponent $\gamma$ for networks simulated under the \gls{BA}
       network model with $N$ = 5000 and $m$ = 2.
@@
 
+\begin{data}[H]
+  \centering
+  \includegraphics{stock/missing}
+  \caption{
+    Plots of marginal posterior distributions estimated with kernel-\gls{ABC}
+    for all simulated transmission trees.
+  }
+  \label{dat:abcpost}
+\end{data}
+
 \end{document}

### 2016-03-23 doc/figures/pa.R

@@ -1,0 +1,21 @@
+#!/usr/bin/env Rscript
+
+library(igraph)
+
+nv <- c(3, 4, 5, 6, 12)
+
+g <- list(NULL)
+for (n in nv) {
+    g[[length(g)+1]] <- sample_pa(n, m=2, power=1, directed=FALSE, start.graph=tail(g, 1)[[1]])
+}
+g <- tail(g, length(g)-1)
+
+vcol <- mapply(rep, list(c("black", "white")), 
+               as.list(as.data.frame(rbind(nv, max(nv)-nv))), SIMPLIFY=FALSE)
+
+lay <- layout_in_circle(tail(g, 1)[[1]])
+pdf("pa.pdf")
+mapply(plot, g, vertex.color=vcol, vertex.frame.color=vcol,
+       vertex.label=list(NA), vertex.size=list(5), 
+       layout=list(lay))
+dev.off()

### 2016-03-23 doc/figures/smc-test.R

@@ -1,0 +1,20 @@
+#!/usr/bin/env Rscript
+
+library(data.table)
+library(extrafont)
+
+trace.file <- system2(c("locate", "check_smc_trace.tsv"), stdout=TRUE)
+trace <- fread(trace.file)
+
+trace <- trace[iter == max(iter)]
+trace <- trace[sample(1:nrow(trace), prob=weight, replace=TRUE)]
+dens <- density(trace$theta)
+
+x <- seq(-4, 4, 0.01)
+target <- 0.5 * dnorm(x) + 0.5 * dnorm(x, sd=0.1)
+
+pdf("smc-test.pdf", family="Gillius ADF", width=5, height=5)
+plot(x, target, type="n", xlab="theta", ylab="density")
+polygon(dens$x, dens$y, col="gray80")
+lines(x, target)
+dev.off()

### 2016-03-23 doc/paper/paper.Rnw
Fixed tiny typo

@@
 \usepackage[garamondx,cmbraces]{newtxmath}
-\usepackage{marginnote}
-\usepackage[top=2cm, bottom=2cm, outer=3cm, inner=3cm, heightrounded, 
-            marginparwidth=2cm, marginparsep=0.5cm]{geometry}
+\usepackage{fullpage}
 \usepackage{setspace}
@@
 $N$ was almost always significantly over-estimated using kernel-ABC. Since the
-prior on $N$ and $I$ is jointly uniform on a triangular region ($I \leq N$),
-there is more prior mass on high $N$ values. In retrospect, it is unreasonable
-to expect good estimation of $N$, because adding more nodes to a \gls{BA}
-network does not change the edge density or overall shape. This can be
+prior on $N$ and $I$ is jointly uniform on a non-rectangular region ($I \leq
+N$), there is more prior mass on high $N$ values. In retrospect, it is
+unreasonable to expect good estimation of $N$, because adding more nodes to a
+\gls{BA} network does not change the edge density or overall shape. This can be
 illustrated by imagining that we add a small number of nodes to a network after
@@
 It is possible to fit more complex models which address some of these issues,
-such as one incorporating heterogeneous node behaviour. This may prove a
+such as one incorporating heterogeneous node behaviour, which may prove a
 fruitful avenue for future investigations.

### 2016-03-24 doc/thesis/intro.tex
Added stuff about model fitting

@@
 internal nodes can be mapped onto a timeline by using the tips of the tree,
-which usually map to the present day, as a reference
+which usually correspond to the present day, as a reference
 point~\autocite{nee1992tempo}. The corresponding points on the timeline are
@@
 Transmission trees generally have branch lengths in units of calendar time,
-with a root corresponding to the index case, and branching times indicating
-times of transmission. The internal nodes may be labelled with
-the donor of the transmission pair, if this is known. The tips of the tree,
-rather than being fixed at the present day, are placed at the time at which the
-individual was removed from the epidemic, such as by death, recovery,
-isolation, behaviour change, or migration. Consequently, the transmission tree
-may not be ultrametric, but may have tips located at varying distances from the
-root. Such trees are said to have \defn{heterochronous}
-taxa~\autocite{drummond2003measurably}, in contrast to the
-\defn{isochronous} taxa found in the phylogenies of macro-organisms. A
+with branching times indicating times of transmission. The root of a
+transmission tree corresponds to the initially infected patient who introduced
+the epidemic into the network, also known as the \defn{index case}. The
+internal nodes may be labelled with the donor of the transmission pair, if this
+is known. The tips of the tree, rather than being fixed at the present day, are
+placed at the time at which the individual was removed from the epidemic, such
+as by death, recovery, isolation, behaviour change, or migration. Consequently,
+the transmission tree may not be ultrametric, but may have tips located at
+varying distances from the root. Such trees are said to have
+\defn{heterochronous} taxa~\autocite{drummond2003measurably}, in contrast to
+the \defn{isochronous} taxa found in most phylogenies of macro-organisms. A
 transmission tree is illustrated in \cref{fig:contactnet} (right).
 
-In a transmission tree, each sampled individual appears as a tip when they are
-removed from the infected population. In addition, they appear as an internal
-node each time they transmit to an individual who is either eventually sampled,
-or who transmits directly or indirectly to a sampled individual. This is not
-the same as the transmission \emph{network}, in which each infected individual
-appears exactly once, and edges are in one-to-one correspondence with
+Due to the internal nodes, each infected individual in an epidemic may appear
+in the transmission tree more than once. This is different from the
+transmission \emph{network}, in which each infected individual appears exactly
+once, and edges are in one-to-one correspondence with
 transmissions~\autocite{welch2011statistical, keeling2005networks}.
@@
 phylogenies, which relate one viral genotype from each host in a population.
-The tips of the tree are labelled with both the virus and its host. The crux of
-\defn{phylodynamics}~\autocite{grenfell2004unifying} is the fact that the
-processes generating these two types of tree - epidemiological processes, for
-transmission trees, and evolutionary processes, for viral phylogenies - occur
-on similar time scales for RNA viruses~\autocite{drummond2003measurably}. As a
-result, there is a close relationship between the two trees. In particular, the
+The crux of \defn{phylodynamics}~\autocite{grenfell2004unifying} is the fact
+that the epidemiological processes recorded in transmission trees, and the
+evolutionary processes recorded in viral phylogenies, occur on similar time
+scales for RNA viruses~\autocite{drummond2003measurably}. As a result, there is
+a close relationship between the two types of tree. In particular, the
 transmission process is quite similar to \defn{allopatric
@@
 which is represented as branching in the transmission tree, causes branching in
-the viral phylogeny as well. Similarly, the removal of an individual causes the
-extinction of their viral lineage. Due to these relationships, the topology of
-the viral phylogeny is often used as a proxy for the topology of the
-transmission tree. However, there are several complications and caveats which
-must be kept in mind when estimating the transmission tree in this manner.
+the viral phylogeny as well. Similarly, the removal of an individual from the
+transmission tree causes the extinction of their viral lineage in the
+phylogeny. Due to these relationships, the topology of the viral phylogeny is
+often used as a proxy for the topology of the transmission tree. However, there
+are several complications and caveats which must be kept in mind when
+estimating the transmission tree in this manner.
 
 First are the issues of rooting and time-scaling. Modern likelihood-based
-methods of phylogenetic reconstruction produce unrooted trees whose branch
-lengths measure genetic distance in units of expected substitutions per site,
-whereas transmission trees are rooted with branches measuring calendar time. We
-generally assume that sampling a virus from the individual also corresponds to
-their removal from the transmission tree, so the positions of the tips in time
-are fixed. Therefore, to transform a viral phylogeny into an estimated
-transmission tree, we must find a root and an assignment of branch lengths such
-that the tips are placed at their proscribed times. Of course, there are an
-infinite number of possible combinations of root and branch lengths which
-result in the correct placement of the tips, so we would also like to find
-values which are the most ``reasonable'' given our data. By reasonable, we mean
-that the variation in \defn{evolutionary rate}, which is the ratio of a
+methods of phylogenetic reconstruction~\autocite[\eg][]{price2010fasttree,
+stamatakis2014raxml} produce unrooted trees whose branch lengths measure
+genetic distance in units of expected substitutions per site. On the other
+hand, transmission trees are rooted, and have branches measuring calendar
+time~\autocite{pybus2009evolutionary}. It is generally assumed that sampling a
+virus from the individual also corresponds to their removal from the
+transmission tree, so the positions of the tips in time are fixed. Therefore,
+to transform a viral phylogeny into an estimated transmission tree, we must
+find a root and an assignment of branch lengths such that the tips are placed
+at their proscribed times. In addition, the branch lengths should be chosen
+such that the variation in \defn{evolutionary rate}, which is the ratio of a
 branch's length in genetic distance to its length in calendar time, is low
@@
 regression; sum of squared errors for least-square dating). Alternatively, the
-tree may be rooted with an outgroup~\autocite{li1988rates} before time-scaling.
+tree may be rooted separately with an outgroup~\autocite{li1988rates} before
+time-scaling.
 
@@
 clock}). This inference is usually done in a Bayesian framework using
-\gls{MCMC}, so that prior information (including the tip dates) 
-%\comment{are tip dates part of the prior?} 
-can be included in the analysis, and the so-called nuisance parameters of the
-molecular clock model can be marginalized out. Software packages for performing
-these analyses include BEAST~\autocite{bouckaert2014beast} and
+\gls{MCMC}, so that prior information (including the tip dates) can be included
+in the analysis, and the so-called nuisance parameters of the molecular clock
+model can be marginalized out. Software packages for performing these analyses
+include BEAST~\autocite{bouckaert2014beast} and
 MrBayes~\autocite{ronquist2012mrbayes}.
@@
 transmission network, which in turn determines the topology of the transmission
-tree relating the infected hosts (\cref{fig:contactnet}). The initially
-infected node who introduces the epidemic into the network, also known as the
-\defn{index case}, becomes the root of the tree. Each time a transmission
-occurs, the lineage corresponding to the donor host in the tree splits into
-two, representing the recipient lineage and the continuation of the donor
-lineage. \Cref{fig:contactnet} illustrates this correspondence. It's important
-to note that, although the order and timing of transmissions determines the
-tree topology uniquely, the converse does not hold. That is, for any given
-topology, there are in general many transmission networks which would lead to
-that topology. In other words, it impossible to distinguish who transmitted to
-whom from a transmission tree alone~\autocite{bernard2007hiv}.
+tree relating the infected hosts (\cref{fig:contactnet}). The index case who
+introduces the epidemic into the network becomes the root of the tree. Each
+time a transmission occurs, the lineage corresponding to the donor host in the
+tree splits into two, representing the recipient lineage and the continuation
+of the donor lineage. \Cref{fig:contactnet} illustrates this correspondence.
+It's important to note that, although the order and timing of transmissions
+determines the tree topology uniquely, the converse does not hold. That is, for
+any given topology, there are in general many transmission networks which would
+lead to that topology. In other words, it impossible to distinguish who
+transmitted to whom from a transmission tree alone~\autocite{bernard2007hiv}.
 
@@
 \label{sec:abc}
+\glsreset{ABC}
 
-\subsection{Overview and motivation}
+\subsection{Model fitting}
 
-\Gls{ABC} is a statistical method designed to fit complex models, which cannot
-be fit using more conventional methods, to observed
-data~\autocite{marin2012approximate, sunnaker2013approximate,
-beaumont2010approximate} We shall make this precise below, but to fully
-describe and motivate \gls{ABC}, it is necessary to first explain exactly what
-we mean by ``fitting'' a model. We will illustrate this concept with one of the
-simplest and most ubiquitous models: linear regression.
+A \defn{mathematical model} is a formal description of a hypothesized
+relationship between some observed data, $\vec{x} = \set{x_1, \ldots, x_n}$,
+and outcomes, $\vec{y} = \set{y_1, \ldots, y_n}$. A \defn{parametric} model
+defines a family of possible relationships between data and outcomes, indexed
+by one or more numeric parameters $\theta$. A \defn{statistical} model
+describes the relationship between data and outcomes in terms of probabilities.
+Statistical models define, either explicitly or implicitly, the probability of
+observing $\vec{y}$ given $\vec{x}$ and, if the model is parameteric, $\theta$.
+In this context, the observed outcomes are taken to be realizations of random
+variables $\vec{Y} = \set{Y_1, \ldots, Y_n}$. Note that it is entirely possible
+to have no data $\vec{x}$, only observed outcomes $\vec{y}$. In this case, a
+model would describe the process by which $\vec{y}$ is generated.
 
-A linear regression is a model of the relationship between some data $\vec{x} =
-(x_1, \ldots, x_n)$ and outcomes $\vec{y} = (y_1, \ldots, y_n)$. This model
-assumes that the outcomes are linearly related to the data, modulo some noise
-introduced by, say, measurement errors and environmental fluctuations. It other
-words, there is a constant $\beta$ such that $y_i = \beta x_i + \varepsilon_i$,
-where $\varepsilon_i$ is the error associated with the outcome $y_i$. If we
-make the additional assumption that the errors are normally distributed, then
-the model takes the form
-\[
-  y_i = \beta x_i + \N(0, \sigma^2),
-\]
-where $\N(0, \sigma^2)$ indicates a normally distributed random variable with
-mean 0 and variance $\sigma$. We will denote the \gls{pdf} of this random
-variable by $f_\N(\cdot \mid \mu, \sigma)$. In this formulation, the
-coefficient $\beta$ and the error variance $\sigma$ are the two parameters of
-the model. If the $y_i$ are all independent, then for particular choice of
-$\beta$ and $\sigma$, we can write down the probability density of observing
-$\vec{y}$ given $\vec{x}$.
-\[
-  f(\vec{y} \mid \vec{x}, \beta, \sigma) = 
-  \prod_{i=1}^n f_\N(y_i - \beta x_i \mid 0, \sigma^2).
-\]
-The higher the value of this probability density, the more likely the
-observations $\vec{y}$ are given $\vec{x}$ under the model. This gives us a
-natural criterion for choosing the parameters: we want to pick $\beta$ and
-$\sigma$ which define a model where the probability of our observed data is as
-high as possible. When performing such an optimization for fixed data, the
-density function just written is also called the \defn{likelihood} of the
+To illustrate these concepts, consider the well-known linear model. For
+clarity, we will restrict our attention to the case of one-dimensional data and
+outcomes where each $x_i$ and $y_i$ is a real number. The linear model
+postulates that the outcomes are linearly related to the data, modulo some
+noise introduced by measurement error, environmental fluctuations, and other
+external factors. Formally, $y_i = \beta x_i + \varepsilon_i$, where $\beta$ is
+the slope of the linear relationship, and $\varepsilon_i$ is the error
+associated with measurement $i$. We can make this model a statistical one by
+hypothesizing a distribution for the error terms $\varepsilon_i$; most
+commonly, it is assumed that they are normally distributed with variance
+$\sigma$. In mathematical terms, $Y_i \sim \beta x_i + \N(0, \sigma^2)$, where
+``$\sim$'' means ``is distributed as''. We can see from this formulation that
+the model is parametric, with parameters $\beta$ and $\sigma$. Moreover, we can
+write down the probability density of observing outcome $y_i$ given the
 parameters,
 \[
-  \L(\beta, \sigma \mid \vec{x}, \vec{y}) = f(\vec{y} \mid \vec{x}, \beta, \sigma).
+  \Pr(Y_i = y_i \mid \beta, \sigma) = 
+  \Pr\left(\N(0, \sigma^2) = y_i - \beta x_i\right).
 \]
-The particular $\beta$ and $\sigma$ which optimize $\L(\beta, \sigma \mid
-\vec{x}, \vec{y})$ are called the \gls{ML} estimates.
+Note that we have followed the standard statistical abuse of notation and used
+``$\Pr$'' to refer to a probability \emph{density}, rather than a probability.
+Assuming all the $y_i$ are independent, the probability density of the entire
+observed set of outcomes is the product of the probability density of each
+individual $y_i$,
+\[
+  \Pr(\vec{Y} = \vec{y} \mid \beta, \sigma) = 
+  \prod_{i=1}^N \Pr(Y_i = y_i \mid \beta, \sigma).
+\]
 
-\Gls{ML} inference makes use only of the data at hand, in this case
-$\vec{x}$ and $\vec{y}$, to estimate the parameters $\beta$ and $\sigma$.
-However, it is frequently the case that the investigator has some additional
-\defn{prior} information or belief about what $\beta$ and $\sigma$ are likely
-to be. For example, the instrument used to measure the $\vec{y}$ may have a
-known margin of error, or the sign of the slope may be obvious from previous
-experiments. A Bayesian approach to fitting this model would make use of this
-information by codifying the investigator's beliefs as a \defn{prior
-distribution}, denoted $f(\beta, \sigma)$. The optimal parameters are then
-those which maximize the product of the prior and the likelihood,
+For a general model, the probability density of $\vec{y}$ given the parameters
+$\theta$ is also known as the \defn{likelihood}, written $\L$, of $\theta$.
+That is, $\L(\theta \mid \vec{y}) = \Pr(\vec{y} \mid \theta)$. The higher the
+value of the likelihood, the more likely the observations $\vec{y}$ are under
+the model. Thus, the likelihood provides a natural criterion for fitting the
+model parameters: we want to pick $\theta$ such that the probability density of
+our observed outcomes $\vec{y}$ is as high as possible. The parameters which
+optimize the likelihood are known as the \textit{\gls{ML}} estimates, denoted
+$\hat{\theta}$. \Gls{ML} estimation is usually performed with numerical
+optimization. In the simplest terms, many possible values for $\theta$ are
+examined, $\L(\theta \mid \vec{y})$ is calculated for each, and the parameters
+which produce the highest value are accepted. Many sophisticated numerical
+optimization methods exist, although they may not be guaranteed to find the
+true \gls{ML} estimates if the likelihood function is complex. Occasionally, as
+in the case of least squares, the \gls{ML} estimates can be found explicitly by
+setting the likelihood function's derivatives to zero.
+
+\Gls{ML} estimation makes use only of the data and outcomes to estimate the
+model parameters $\theta$. However, it is frequently the case that the
+investigator has some additional information or \defn{prior belief} about what
+$\theta$ are likely to be. For example, in the linear regression case, the
+intstrument used to measure the outcomes may have a well-known margin of error,
+or the sign of the slope may be obvious from previous experiments. The Bayesian
+approach to model fitting makes use of this information by codifying the
+investigator's beliefs as a \defn{prior distribution} on the parameters,
+denoted $\Pr(\theta)$. Instead of considering only the likeilhood, Bayesian
+inference focuses on the product of the likelihood and the prior, $\Pr(\vec{y}
+\mid \theta) \Pr(\theta)$. Bayes' theorem tells us that this product is related
+to the \textit{posterior distribution} on $\theta$,
 \[
-  f(\vec{y} \mid \vec{x}, \beta, \sigma) f(\beta, \sigma).
+  \Pr(\theta \mid \vec{y}) = \frac{\Pr(\vec{y} \mid \theta) \Pr(\theta)}
+                                  {\Pr(\vec{y})}.
 \]
-Bayes' theorem tells us that this product is proportional to the
-\defn{posterior distribution} $f(\beta, \sigma \mid \vec{y}, \vec{x})$. The
-parameters which optimize the posterior, and hence also optimize the above
-product, are called the \gls{MAP} estimates.
 
-Now that we have defined what it means to fit a linear regression, there
-remains the question of how to go about finding this optimal $\beta$ value.
-Na\"{i}vely, we could simply try many different $\beta$ and $\sigma$ values,
-calculate the likelihood of each, and choose those which yield the highest
-value. This approach, though basic, falls under the umbrella of \defn{numerical
-optimization} of the likelihood function. Of course, there is a whole field
-devoted to more sophisticated methods for exploring the parameter space, but
-they all boil down to the basic idea of trying out different values and
-choosing those which give the highest likelihood.
+In principle, $\Pr(\vec{y} \mid \theta) \Pr(\theta)$ can be optimized
+numerically just like $\L(\theta \mid \vec{y})$, which would also optimize the
+posterior distribution. The resulting optimal parameters are called the
+\gls{MAP} estimates. However, from a Bayesian perspective, $\theta$ is not a
+fixed quantity to be estimated, but rather a random variable with an associated
+distribution (the posterior). Therefore, the \gls{MAP} estimate by itself is of
+limited value without associated statistics about the posterior distribution,
+such as the mean or credible intervals. Unfortunately, to calculate such
+statistics, it is necessary to evaluate the normalizing constant
+$\Pr(\vec{y})$, which is almost always an intractible integral.
 
-In the case of linear regression, there is another method we could use. We know
-from calculus that the maximum of $\L$ occurs either on the boundary of its
-domain, or at a point where its partial derivatives with respect to $\beta$ and
-$\sigma$ are both zero. Though we will not show it here, it turns out to be
-straightforward to find the $\beta$ and $\sigma$ values where the partial
-derivatives are zero. In fact, these values co-incide with the least-squares
-estimates of $\beta$ and $\sigma$, which minimize the sum of the squared $x_i -
-\beta y_i$. However, the method of directly minimizing the loss function with
-calculus is only applicable to a narrow class of simple models. For the
-majority, the likelihood function is too complicated to set the partial
-derivatives with respect to the parameters to zero and solve.
+A popular method for circumventing the normalizing constant is the use of
+\gls{MCMC} to obtain a sample from the posterior distribution. \Gls{MCMC} works
+by defining a Markov chain whose states are indexed by possible model
+parameters. The transition probability from state $\theta_1$ to state
+$\theta_2$ is taken to be
+\[
+  \max\left(1, \frac{\Pr(\vec{y} \mid \theta_2) \Pr(\theta_2) q(\theta_2, \theta_1)}
+                    {\Pr(\vec{y} \mid \theta_1) \Pr(\theta_2) q(\theta_1, \theta_2)} \right),
+\]
+where $q(\theta, \theta')$ is a symmetric \defn{proposal distribution} used in
+the algorithm to generate the chain. The stationary distribution of this Markov
+chain is equal to the posterior distribution on $\theta$. Therefore, if a long
+enough random walk is performed on the chain, the distribution of states
+visited will be a Monte Carlo approximation of $\Pr(\theta \mid \vec{y})$, from
+which we can calculate statistics of interest. Actually performing this random
+walk is straightforward and is accomplished via the Metropolis-Hastings
+algorithm.
 
-An approximate Bayesian computation approach to least-squares regression might
-be as follows. We will need to make an additional assumption: that the errors
-$\varepsilon_i$ are normally distributed. That is, our linear model is now of
-the form
+\begin{algorithm}
+  \begin{algorithmic}
+    \State Draw $\theta$ according to $\Pr(\theta)$
+    \Loop
+      \State Propose $\theta'$ according to $q(\theta, \theta')$
+      \State Accept $\theta \gets \theta'$ with probability
+      $\max \left( 1, 
+       \dfrac{\Pr(\vec{y} \mid \theta') \Pr(\theta') q(\theta', \theta)}
+             {\Pr(\vec{y} \mid \theta\phantom{'}) \Pr(\theta\phantom{'}) q(\theta, \theta')}
+       \right)$
+    \EndLoop
+  \end{algorithmic}
+  \label{alg:mh}
+\end{algorithm}
+
+\subsection{Overview and motivation for ABC}
+
+Most mathematical models are amenable to fitting via one or both of the
+approaches, \gls{ML} or Bayesian inference, discussed above. However, there are
+some, particularly in the domain of population genetics, for which calculation
+of either the likelihood or the product of the likelihood and the prior may be
+infeasible. For example, one or both of these quantities may be expressible
+only as an intractible integral. \Gls{ABC} is designed for such cases, where
+standard likelihood-based techniques for model fitting cannot be applied.
+
+The first practical application of \gls{ABC} was by
+\textcite{tavare1997inferring}, who wanted to calculate the posterior
+distribution on the coalescence time $t$ of $n$ sequences with $k$ observed
+segregating sites under the coalescent model~\autocite{kingman1982coalescent}.
+That is,
 \[
-  y_i = bx_i + \N(0, \sigma),
+  \Pr(t \mid k) \propto \Pr(t) \Pr(k \mid t),
 \]
-where $\N(0, \sigma)$ indicates a normal distribution with mean zero and
-variance $\sigma$. We do not know $\sigma$ in advance, so this is another
-parameter we will need to estimate. There is nothing special about normal
-distributions - we could have chosen any other distribution we deemed
-appropriate.
+where again we abuse $\Pr$ to refer to a probability density.
+\citeauthor{tavare1997inferring} observed that the expected number of
+segregating sites depends on the total branch length of the tree, rather than
+its height. This enabled them to write an alternative expression for the
+product of the likelihood and prior in terms of the total branch length $l$,
+\[
+  \int_0^\infty \Pr(t, l) \Pr(k \mid t, l) \d l.
+\]
+While it is unclear how to evaluate this integral, the power of this
+formulation is the fact that it is straightforward to \emph{simulate} points
+according to $\Pr(t, l)$, the joint probability of $t$ and $l$, by simply
+running the (stochastic) coalescent process and observing the values $\bar{t}$
+and $\bar{l}$ which result.
 
-The main idea of \gls{ABC} is to replace the traditional likelihood-based
-criteria for model ``goodness'' with one based on simulated data generated by
-the model. Good models should be able to simulate data which closely resemble
-reality. In \gls{ABC}, we are no longer interested in the posterior
-distribution, but rather in the distribution of model parameters which produce
-data sufficiently ``close'' to the real data. To formalize this, let $d(\cdot,
-\cdot)$ be a distance function on datasets, and $\varepsilon$ be a tolerance
-level. The objective of \gls{ABC} is to recover the distribution\ldots
+%\Gls{ABC} is a statistical method designed to fit complex models, which cannot
+%be fit using more conventional methods, to observed
+%data~\autocite{marin2012approximate, sunnaker2013approximate,
+%beaumont2010approximate}. We shall make this precise below, but to fully
+%describe and motivate \gls{ABC}, it is necessary to first explain exactly what
+%we mean by ``fitting'' a model. We will illustrate this concept with one of the
+%simplest and most ubiquitous models: linear regression.
+%
+%An approximate Bayesian computation approach to least-squares regression might
+%be as follows. We will need to make an additional assumption: that the errors
+%$\varepsilon_i$ are normally distributed. That is, our linear model is now of
+%the form
+%\[
+%  y_i = bx_i + \N(0, \sigma),
+%\]
+%where $\N(0, \sigma)$ indicates a normal distribution with mean zero and
+%variance $\sigma$. We do not know $\sigma$ in advance, so this is another
+%parameter we will need to estimate. There is nothing special about normal
+%distributions - we could have chosen any other distribution we deemed
+%appropriate. 
+%
+%The main idea of \gls{ABC} is to replace the traditional likelihood-based
+%criteria for model ``goodness'' with one based on simulated data generated by
+%the model. Good models should be able to simulate data which closely resemble
+%reality. In \gls{ABC}, we are no longer interested in the posterior
+%distribution, but rather in the distribution of model parameters which produce
+%data sufficiently ``close'' to the real data. To formalize this, let $d(\cdot,
+%\cdot)$ be a distance function on datasets, and $\varepsilon$ be a tolerance
+%level. The objective of \gls{ABC} is to recover the distribution\ldots
 
@@
 
-Approximate Bayesian computation does not refer to a particular procedure for
-model fitting. Rather, ABC refers to the general strategy of choosing model
-parameters based on the resulting model's propensity to generate data
-resembling the real data. There are three main classes of ABC algorithm which
-have been developed so far: rejection, \gls{MCMC}, and sequential Monte Carlo
-(SMC).
+\Gls{ABC} does not refer to a particular procedure for model fitting. Rather,
+\gls{ABC} refers to the general strategy of choosing model parameters based on
+the resulting model's propensity to generate data resembling the real data.
+There are three main classes of \gls{ABC} algorithm which have been developed
+so far: rejection, \gls{MCMC}, and \gls{SMC}.
 

### 2016-03-28 doc/thesis/methods.tex
Wrote some stuff about netabc

@@
 
-I implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
-\textcite{del2012adaptive}. The program was written in the
-\software{C} programming language. The \software{igraph}
-library~\autocite{csardi2006igraph} was used to generate and store contact
-networks and phylogenies. Judy arrays~\autocite{baskins2004judy} were used for
-hash tables and dynamic programming arrays. The
-\gls{GSL}~\autocite{gough2009gnu} was used to generate random draws from
-probability distributions, and to perform the bisection step in the
-\gls{ABC}-\gls{SMC} algorithm.
+\software{Netabc} is a computer program designed to perform statistical
+inference of contact network parameters from an estimated transmission tree
+using kernel-\gls{ABC}. The program combines three major components: Gillespie
+simulation, to simulate transmission trees on contact networks; the
+phylogenetic kernel, to compare simulated to observed transmission trees; and
+adaptive \gls{ABC}-\gls{SMC}, to maintain a population of particles and
+advance it toward the \gls{ABC} target distribution. We give a high-level
+overview of the program here, before describing these three components in
+detail.
 
-For ease of exposition, we simplifiy the notation of \textcite{del2012adaptive}
-by dropping the subscripts on the variables which indicate the current
-iteration number. Instead, we will add a prime $'$ to indicate a value which
-will be used in the next iteration (this should become clear later on).
+As described in \cref{sec:smc}, \software{netabc} keeps track of a population
+of particles, each of which contains particular parameter values for the model
+we are trying to fit. A small number of contact networks are generated for each
+particle, in accordance with that particle's parameters. An epidemic is
+simulated over each of these networks using Gillespie simulation, and by
+keeping track of its progress, a transmission tree is obtained. Thus, each
+particle becomes associated with several simulated transmission trees.  These
+trees are compared to the observed tree using the phyolgenetic kernel.
+Particles are weighted according to the similarity of their associated
+simulated trees with the true tree, with more similar trees recieving higher
+weights. The particles are iteratively perturbed to explore the parameter
+space, and particles with simulated trees too distant from the true tree are
+periodically dropped and resampled. Once a convergence criterion is attained,
+the final set of particles is used as a Monte Carlo approximation to the target
+distribution of \gls{ABC}, which is assumed to resemble the posterior
+distribution on model parameters (see \cref{sec:abc}). A graphical schematic of
+this algorithm is given in \cref{fig:abcsmc}.
 
-In the algorithm, we keep track of a population of $n$ sets of model
-parameters, called \defn{particles}, denoted $\set{X_i}_{i=1}^{n}$. For the
-\gls{BA} model, the particles would be 4-tuples (\gls{N}, \gls{I}, \gls{m},
-\gls{alpha}). Each particle $X_i$ is associated with a set of \gls{M} simulated
-datasets, denoted $\set{X_{i,j}}_{j=1}^M$, and a weight $W_i$.
+The generation of contact networks from the model of interest, though critical
+to the program's function, is not included here as a component of
+\software{netabc}.  Generation of networks is a model-specific task, whereas
+\software{netabc} is intended to be generic and work with any network model.
+For ease of use, generators for three basic network models, namely the
+\gls{BA}, \gls{ER}, and \gls{WS} models, are included in the program through
+the \software{igraph} library~\autocite{csardi2006igraph}. To investigate other
+models, additional graph generators will need to be added into the program.
 
-The particles are initially drawn from the prior distribution.
+\begin{figure}
+    \includegraphics{abc-smc.pdf}
+    \caption[Graphical schematic of \gls{ABC}-\gls{SMC} algorithm.]{
+      Graphical schematic of the \gls{ABC}-\gls{SMC} algorithm implemented in
+      \software{netabc}. Particles are initially drawn from their prior
+      distributions, making the initial population a Monte Carlo approximation
+      to the prior. At each iteration, particles are perturbed, and a distance
+      threshold around the true tree contracts. Particles are rejected, and
+      eventually resampled, when all their associated simulated trees lie
+      outside the threshold. As the algorithm progresses, the population
+      smoothly approaches a Monte Carlo approximation of the \gls{ABC} target
+      distribution, which is assumed to resemble the posterior.
+    }
+    \label{fig:abcsmc}
+\end{figure}
 
-Let $d$ be a
-distance measure on data sets, so that $d(x, y)$ is smaller the more similar
-$x$ and $y$ are to each other. Let $\varepsilon$ be the tolerance level which
-indicates whether a data set is ``close'' to the observed data. That is, if
-$d(x, y) < \varepsilon$, we will say that $x$ and $y$ are close, otherwise they
-are distant.
+\software{Netabc} is written in the \software{C} programming language. The
+\software{igraph} library~\autocite{csardi2006igraph} is used to generate and
+store contact networks and phylogenies. Judy arrays~\autocite{baskins2004judy}
+are used for hash tables and dynamic programming matrices. The
+\gls{GSL}~\autocite{gough2009gnu} is used to generate random draws from
+probability distributions, and to perform the bisection step in the adaptive
+\gls{ABC}-\gls{SMC} algorithm. The program is freely available at
+\url{github.com/rmcclosk/netabc}.
 
@@
 
-I implemented a Gillespie simulation algorithm~\autocite{gillespie1976general}
-for simulating epidemics, and the corresponding transmission trees, over static
-contact networks. This method has been independently implemented and applied by
-several authors~\autocite[\textit{e.g.}][]{o2010contact, robinson2013dynamics,
+The simulation of epidemics, and the corresponding transmission, trees over
+contact networks is performed in \software{netabc} using the Gillespie
+simulation algorithm~\autocite{gillespie1976general}. This method has been
+independently implemented and applied by several
+authors~\autocite[\textit{e.g.}][]{o2010contact, robinson2013dynamics,
 leventhal2012inferring, groendyke2011bayesian}.
@@
 
-Let $G = (V, E)$ be a directed contact network. The individual nodes and edges
-of $G$ follow the dynamics of the \gls{SIR}
+Let $G = (V, E)$ be a directed contact network. We assume the individual nodes
+and edges of $G$ follow the dynamics of the \gls{SIR}
 model~\autocite{kermack1927contribution}. Each directed edge $e = (u, v)$ in
@@
 recovered nodes never re-enter the susceptible population. We define a
-\defn{discordant edge} as an edge $(u, v)$ where $u$ is infected and $v$
-has never been infected.
+\defn{discordant edge} as an edge $(u, v)$ where $u$ is infected and $v$ has
+never been infected.
 
@@
 simulation progresses. When a node $v$ becomes infected, it is deleted from $S$
-and added to $I$, any formerly discordant edges in $\in(v)$ are deleted from
+and added to $I$. Any formerly discordant edges in $\in(v)$ are deleted from
 $D$, and edges in $\out(v)$ to nodes in $S$ are added to $D$. If $v$ is later
 removed, it is deleted from $I$ and added to $R$, and any discordant edges in
-$\out(v)$ are deleted from $D$. In both cases, the variables $\beta$ and
-$\gamma$ are updated to reflect the changes. Since these updates are
-straightforward, we do not write them explicitly in the algorithm.
+$\out(v)$ are deleted from $D$. At the time of either infection or removal, the
+variables $\beta$ and $\gamma$ are updated to reflect the changes in the
+network. Since these updates are straightforward, we do not write them
+explicitly in the algorithm.
 
-\newcommand{\tip}{\mathrm{tip}}
+\newcommand{\tip}{\mathit{tip}}
 
@@
           \State add tips with labels $(|T|+1)$ and $(|T|+2)$ to $T$
-          \Comment $|T|$ increased by 2
           \State connect the new nodes to $\tip[v]$ in $T$, with branch lengths $0$
@@
 
-\subsubsection{Phylogenetic kernel and normalized lineages-through-time}
+\subsubsection{Phylogenetic kernel}
 
-The tree kernel developed in~\autocite{poon2013mapping} provides a
+The tree kernel developed by \textcite{poon2013mapping} provides a
 comprehensive similarity score between two phylogenetic trees. The kernel
@@
 trees, in the infinite-dimensional feature space of all possible subset trees
-with branch lengths. I implemented the fast algorithm developed
-in~\autocite{moschitti2006making}, which first enumerates all pairs of subtrees
-with the same number of leaf children, and then computes the kernel by dynamic
-programming.
+with branch lengths (see \cref{subsec:treeshape}). The kernel was implemented
+using the fast algorithm developed by \textcite{moschitti2006making}. First,
+the number of leaf children of each node, also known as its \defn{production
+rule}, is recorded. The nodes of both trees are ordered by production rule, and
+a list of pairs of nodes sharing the same production rule is created. These are
+the nodes for which the value of the tree kernel must be computed - all other
+pairs have a value of zero. The pairs to be compared are then re-ordered so
+that the child nodes are always evaluated before their parents. Due to its
+recursive definition, ordering the pairs in this way allows the tree kernel to
+be computed by dynamic programming. The complexity of this implementation is
+$O(|T_1||T_2|)$, where $|T|$ counts the number of nodes in the tree $T$.
 
-In addition, we implemented a modified version of the normalized
-lineages-through-time statistic developed in~\autocite{janzen2015approximate},
-which uses piecewise linear functions instead of step functions for the
-lineages-through-time plots. This modification is to address a potential
-inconsistency for trees of different sizes, illustrated in
-Figure~\ref{fig:nltt}.
+\subsubsection*{Adaptive sequential Monte Carlo for Approximate Bayesian computation}
 
-\begin{figure}[ht]
-  \centering
-  \caption[Comparison of original and modified normalised lineages-through-time]
-    {Comparison of original formulation of normalized lineages-through-time,
-     developed in~\autocite{janzen2015approximate}, with our modified version
-     using linear interpolation. Here, the red and blue trees both have
-     uniformly spaced branching times. Using step functions (left), the nLTT
-     of the two trees is non-zero due to the differing numbers of internal
-     branches. Using linear interpolation, the nLTT is zero (right). The lines
-     on the right graph have been offset for visibility.}
-  \includegraphics{nltt.pdf}
-  \label{fig:nltt}
-\end{figure}
+I implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
+\textcite{del2012adaptive}.
+
+For ease of exposition, we simplifiy the notation of \textcite{del2012adaptive}
+by dropping the subscripts on the variables which indicate the current
+iteration number. Instead, we will add a prime $'$ to indicate a value which
+will be used in the next iteration (this should become clear later on).
+
+In the algorithm, we keep track of a population of $n$ sets of model
+parameters, called \defn{particles}, denoted $\set{X_i}_{i=1}^{n}$. For the
+\gls{BA} model, the particles would be 4-tuples (\gls{N}, \gls{I}, \gls{m},
+\gls{alpha}). Each particle $X_i$ is associated with a set of \gls{M} simulated
+datasets, denoted $\set{X_{i,j}}_{j=1}^M$, and a weight $W_i$.
+
+The particles are initially drawn from the prior distribution.
+
+Let $d$ be a
+distance measure on data sets, so that $d(x, y)$ is smaller the more similar
+$x$ and $y$ are to each other. Let $\varepsilon$ be the tolerance level which
+indicates whether a data set is ``close'' to the observed data. That is, if
+$d(x, y) < \varepsilon$, we will say that $x$ and $y$ are close, otherwise they
+are distant.
 
@@
 
-We applied kernel-ABC to six published HIV datasets, summarized in
+We applied kernel-\gls{ABC} to six \gls{HIV} datasets, summarized in
 \cref{tab:realdata}.

### 2016-03-29 doc/thesis/appendix.tex
Writing stuff

\begin{figure}
  \centering
  \includegraphics{kernel-I-tree.pdf}
  \caption{
    Transmission trees simulated over \gls{BA} networks with varying values of
    $I$, the number of infected nodes when the epidemic simulation was stopped.
  }
  \label{fig:Itrees}
\end{figure}

\begin{figure}
  \centering
  \includegraphics{kernel-m-tree.pdf}
  \caption{
    Transmission trees simulated over \gls{BA} networks with varying values of
    $m$, the number of edges added per vertex.
  }
  \label{fig:mtrees}
\end{figure}

\begin{figure}
  \centering
  \includegraphics{kernel-N-tree.pdf}
  \caption{
    Transmission trees simulated over \gls{BA} networks with varying values of
    $N$, the number of nodes in the network.
  }
  \label{fig:Ntrees}
\end{figure}

\begin{figure}
  \centering
  \includegraphics{kernel-alpha-crossv.pdf}
  \label{fig:alphacrossv}
\end{figure}

\begin{figure}
  \centering
  \includegraphics{kernel-I-crossv.pdf}
  \label{fig:Icrossv}
\end{figure}

\begin{figure}
  \centering
  \includegraphics{kernel-m-crossv.pdf}
  \label{fig:mcrossv}
\end{figure}

\begin{figure}
  \centering
  \includegraphics{kernel-N-crossv.pdf}
  \label{fig:Ncrossv}
\end{figure}

\begin{figure}
  \centering
  \includegraphics{kernel-alpha-kpca.pdf}
  \label{fig:alphakpca}
\end{figure}

\begin{figure}
  \centering
  \includegraphics{kernel-I-kpca.pdf}
  \label{fig:Ikpca}
\end{figure}

\begin{figure}
  \centering
  \includegraphics{kernel-m-kpca.pdf}
  \label{fig:mkpca}
\end{figure}

\begin{figure}
  \centering
  \includegraphics{kernel-N-kpca.pdf}
  \label{fig:Nkpca}
\end{figure}

### 2016-03-29 doc/thesis/conclusion.tex
Writing stuff

@@
-\section{Extensions}
+Contact network structure has a substantial impact on epidemic trajectory
+\autocite{barthelemy2005dynamical, volz2007susceptible, volz2008sir}, and a few
+methods have been developed to estimate network parameters from epidemiological
+data \autocite{volz2007susceptible, britton2002bayesian,
+groendyke2011bayesian}. It is known that contact network structure can have a
+substantial impact on transmission tree shape \autocite{o2010contact,
+robinson2013dynamics, leventhal2012inferring, colijn2014phylogenetic,
+villandre2016assessment}.
+
+Our work had three main aims. First, we developed a method to estimate contact
+network model parameters from phylogenetic data. This method widens the field
+of models which may be investigated in a phylodynamic context. Second, we
+investigated the parameters of the \gls{BA} model. We determined through
+simulation studies that the preferential attachment power $\alpha$ and number
+of infected nodes $I$ had a substantial impact on transmission tree shape, and
+could be estimated using our method.
 
@@
 Second, there are a number of different methods available for inferring
-transmission trees~\autocite{didelot2014bayesian, yima2012unravelling,
+transmission trees~\autocite{didelot2014bayesian, ypma2012unravelling,
 jombart2011reconstructing, cottam2008integrating, poon2015phylodynamic}, some
-of which incorporate geographic and/or epidemiological data not accomodated by
+of which incorporate geographic and/or epidemiological data not accommodated by
 our method. We therefore felt it would be best to allow researchers to use

### 2016-03-29 doc/thesis/intro.tex
Writing stuff

@@
 term ``transmission network'' is taken to mean the subgraph of the contact
-network along which transmissions occured,
+network along which transmissions occurred,
 following~\autocite{welch2011statistical, keeling2005networks}.
@@
 
+\subsection{Applications of phylodynamics}
+\label{subsec:phyloapp}
+
+Phylodynamic methods have been used to investigate epidemiological parameters
+such as transmission rate, recovery rate, and basic reproductive
+number~\autocite{pybus2009evolutionary, volz2013viral}. These studies make
+inferences about epidemiological processes from the genetic diversity of virus
+populations, which is usually represented in the form of a phylogeny. The
+majority of these employ a Bayesian \gls{MCMC} approach to infer parameters of
+an epidemiological model whose likelihood can be calculated, most often some
+variation of the birth-death~\autocite{kendall1948generalized} or
+coalescent~\autocite{kingman1982coalescent} models.
+\textcite{stadler2011estimating} develop a formula for the likelihood of a
+phylogeny with heterochronous tips under the birth-death model, which has been
+used to estimate the basic reproductive number of several viral
+epidemics~\autocite{stadler2011estimating}. However, the birth-death model is
+cannot tell us anything about population structure, as it assumes that every
+individual becomes infected at the same rate. \textcite{volz2012complex} writes
+down the likelihood of a heterochronous phylogeny under a coalescent model with
+arbitrarily complex population dynamics. This opens the door to more complex
+inferences about population structure, as the population can be partitioned
+into compartments with different transmission and recovery rates, but still
+assumes that each compartment is homogeneously mixed. In other words, the
+coalescent model can tell us about the \emph{global} structure of a population,
+such as whether there exists a high-risk subgroup, but not about the
+\emph{local} structure, such as the average number of contacts each individual
+has.
+
 \section{Contact networks}
@@
 transmitted infections would not be expected to survive in a homogeneously
-mixed population, but their long-term persistance can be explained by contact
+mixed population, but their long-term persistence can be explained by contact
 heterogeneity~\autocite{anderson1992infectious, pastor2001epidemic}. Hence, the
@@
 In terms of actionable information, vaccination strategies which would
-irradicate an epidemic in a random network might not work if the network is
+eradicate an epidemic in a random network might not work if the network is
 scale-free~\autocite[][see \cref{subsec:pa}]{keeling2005networks}. On a local
@@
 The phylogenetic network is informative about ``hotspots'' of transmission and
-can be used to identify demographic groups to whom targeted intervetions are
+can be used to identify demographic groups to whom targeted interventions are
 likely to have the greatest effect~\autocite{poon2014impact}.  However, this
@@
 prevalence curve, was qualitatively different for each degree distribution.
-\citeauthor{leventhal2012inferring} simulated transmission trees over \gls{ER},
-\gls{WS}, \gls{BA}, and full networks with fixed number of nodes and mean
-degree. They calculated Sackin's index of the simulated trees while varying the
-epidemic, network, and sampling parameters, and found that the relationship
-between these parameters and Sackin's index varies considerably among the
-different network models. In summary, studies in this group have demonstrated
-that network structure profoundly influences tree shape, but have not attempted
-to quantitatively infer network parameters from observed trees.
+\textcite{leventhal2012inferring} investigated the relationship between
+transmission tree imbalance and several epidemic parameters under four contact
+network models, and found that these relationships varied considerably
+depending on which model was being considered. \textcite{welch2011network}
+simulated transmission trees over networks with varying degrees of community
+structure. They found that transmission trees simulated under networks with low
+clustering could not generally be distinguished from those simulated under
+highly clustered networks, and concluded that contact network clusters do not
+affect transmission tree shape. However, more recently,
+\textcite{villandre2016assessment} investigated the correspondence between
+contact network clusters and transmission tree clusters, and found a moderate
+correspondence between the two.
+
+In summary, studies in this group have demonstrated that network structure
+profoundly influences tree shape, but have not attempted to quantitatively
+infer network parameters from observed trees.
 
@@
 \subsection{Model fitting}
+\label{subsec:mfit}
 
@@
 Statistical models define, either explicitly or implicitly, the probability of
-observing $\vec{y}$ given $\vec{x}$ and, if the model is parameteric, $\theta$.
+observing $\vec{y}$ given $\vec{x}$ and, if the model is parametric, $\theta$.
 In this context, the observed outcomes are taken to be realizations of random
@@
 $\theta$ are likely to be. For example, in the linear regression case, the
-intstrument used to measure the outcomes may have a well-known margin of error,
+instrument used to measure the outcomes may have a well-known margin of error,
 or the sign of the slope may be obvious from previous experiments. The Bayesian
@@
 investigator's beliefs as a \defn{prior distribution} on the parameters,
-denoted $\Pr(\theta)$. Instead of considering only the likeilhood, Bayesian
+denoted $\Pr(\theta)$. Instead of considering only the likelihood, Bayesian
 inference focuses on the product of the likelihood and the prior, $\Pr(\vec{y}
@@
 statistics, it is necessary to evaluate the normalizing constant
-$\Pr(\vec{y})$, which is almost always an intractible integral.
+$\Pr(\vec{y})$, which is almost always an intractable integral.
 
@@
 walk is straightforward and is accomplished via the Metropolis-Hastings
-algorithm.
+algorithm (\cref{alg:mh}).
 
 \begin{algorithm}
+  \label{alg:mh}
+  \caption{Metropolis-Hastings algorithm for Markov chain Monte Carlo.}
   \begin{algorithmic}
@@
 \subsection{Overview and motivation for ABC}
+\label{subsec:abcoverview}
 
@@
 infeasible. For example, one or both of these quantities may be expressible
-only as an intractible integral. \Gls{ABC} is designed for such cases, where
-standard likelihood-based techniques for model fitting cannot be applied.
+only as an intractable integral. \Gls{ABC} is designed for such cases, where
+standard likelihood-based techniques for model fitting cannot be applied. Such
+models are particularly prevalent in population
+genetics~\autocite{beaumont2002approximate, beaumont2010approximate}. We will
+begin this subsection by describing \gls{ABC} in general terms. Then, we will
+use linear regression as a toy problem to demonstrate how \gls{ABC} could be
+applied. Finally, we will give an example of a model which is impractical to
+fit using likelihood-based methods, and show how its parameters were estimated
+using \gls{ABC}.
 
-The first practical application of \gls{ABC} was by
-\textcite{tavare1997inferring}, who wanted to calculate the posterior
-distribution on the coalescence time $t$ of $n$ sequences with $k$ observed
-segregating sites under the coalescent model~\autocite{kingman1982coalescent}.
-That is,
+Ordinarily, Bayesian inference targets the posterior distribution.
+
+%traditional likelihood criterion
+%for model ``goodness'' with one based on simulated data generated by the model.
+%Good models should be able to simulate data which closely resemble reality. In
+%\gls{ABC}, we are no longer interested in the posterior distribution, but
+%rather in the distribution of model parameters which produce data sufficiently
+%``close'' to the real data. To formalize this, let $d(\cdot, \cdot)$ be a
+%distance function on datasets, and $\varepsilon$ be a tolerance level.
+
+Of course, linear models can be fit more easily using likelihood-based methods, 
+since the likelihood $\L(\theta \mid \vec{y})$ can be easily calculated (see
+\cref{subsec:fit}). One model where this is not the case is Kingman's
+coalescent~\autocite{kingman1982coalescent}.
+
+As an example of such a model, we will review the first practical application
+of \gls{ABC} by \textcite{tavare1997inferring}. The aim of the investigation
+was to calculate the posterior distribution on the coalescence time $t$ of some
+observed sequences $D$ under the coalescent model
+\autocite{kingman1982coalescent} with the infinite-sites assumption
+\autocite{watterson1975number}. Their estimation method relied on two key
+observations. First, the number $k$ of \defn{segregating sites} in the data is
+informative about the coalescence time while being much easier to work with
+than the data itself. A segregating site is a position which is polymorphic in
+the observed sequences. Using this observation, the authors replaced their
+posterior distribution of interest, $\Pr(t \mid D)$, by
 \[
@@
 \]
-where again we abuse $\Pr$ to refer to a probability density.
-\citeauthor{tavare1997inferring} observed that the expected number of
-segregating sites depends on the total branch length of the tree, rather than
-its height. This enabled them to write an alternative expression for the
-product of the likelihood and prior in terms of the total branch length $l$,
+where again we abuse $\Pr$ to refer to a probability density. Second,
+the authors observed that the expected number of segregating sites depends on
+the total branch length $l$ of the tree, rather than its height. This enabled
+them to write an alternative expression for the product of the likelihood and
+prior in terms of $l$,
 \[
-  \int_0^\infty \Pr(t, l) \Pr(k \mid t, l) \d l.
+  \Pr(t) \Pr(k \mid t) = \int_0^\infty \Pr(t, l) \Pr(k \mid l) \d l.
 \]
-While it is unclear how to evaluate this integral, the power of this
-formulation is the fact that it is straightforward to \emph{simulate} points
-according to $\Pr(t, l)$, the joint probability of $t$ and $l$, by simply
-running the (stochastic) coalescent process and observing the values $\bar{t}$
-and $\bar{l}$ which result.
+While this integral may be difficult to evaluate analytically, the power of
+this formulation is the fact that it is straightforward to \emph{simulate}
+values of $t$ and $l$ according to $\Pr(t, l)$, their joint probability under
+the coalescent model. One simply runs the (stochastic) coalescent process and
+observing the values $\hat{t}$ and $\hat{l}$ which result. It is
+straightforward to calculate $\Pr(k \mid \hat{l})$, the probability
+that $k$ segregating sites were observed in a tree with total branch length
+$\hat{l}$, using the Poisson distribution. Hence if $\hat{l}$ and $\hat{t}$,
+having been generated according to $\Pr(t, l)$, are sampled with probability
+proportional to $\Pr(k \mid \hat{l})$, they represent an unbiased sample from
+the above integral. Taking a large number of such samples allowed the authors
+to obtain a Monte Carlo estimate of $\Pr(t \mid k)$.
 
@@
 %
-%The main idea of \gls{ABC} is to replace the traditional likelihood-based
-%criteria for model ``goodness'' with one based on simulated data generated by
-%the model. Good models should be able to simulate data which closely resemble
-%reality. In \gls{ABC}, we are no longer interested in the posterior
-%distribution, but rather in the distribution of model parameters which produce
-%data sufficiently ``close'' to the real data. To formalize this, let $d(\cdot,
-%\cdot)$ be a distance function on datasets, and $\varepsilon$ be a tolerance
-%level. The objective of \gls{ABC} is to recover the distribution\ldots
 
@@
 \]
-Finally, for the distance function $\rho$ we take the standard Euclidian
+Finally, for the distance function $\rho$ we take the standard Euclidean
 distance.
@@
 own is limited, since it depends on finding an importance distribution which is
-similar to $\pi$, which we usually know very little about \textit{a priori}.
+similar to $\pi$, which we usually know very little about \textit{a priory}.
 

### 2016-03-29 doc/thesis/methods.tex
Writing stuff

@@
 particle becomes associated with several simulated transmission trees.  These
-trees are compared to the observed tree using the phyolgenetic kernel.
+trees are compared to the observed tree using the phylogenetic kernel.
 Particles are weighted according to the similarity of their associated
-simulated trees with the true tree, with more similar trees recieving higher
+simulated trees with the true tree, with more similar trees receiving higher
 weights. The particles are iteratively perturbed to explore the parameter
@@
 probability distributions, and to perform the bisection step in the adaptive
-\gls{ABC}-\gls{SMC} algorithm. The program is freely available at
-\url{github.com/rmcclosk/netabc}.
+\gls{ABC}-\gls{SMC} algorithm. Parallelization is implemented with POSIX
+threads~\autocite{barney2009posix}. In addition to the \software{netabc} binary
+to perform kernel-\gls{ABC}, we provide three additional stand-alone utilities:
+\software{treekernel}, to calculate the phylogenetic kernel;
+\software{nettree}, to simulate a transmission tree over a contact network; and
+\software{treestat}, to compute various summary statistics of phylogenies. The
+programs are freely available at \url{github.com/rmcclosk/netabc}.
 
@@
 authors~\autocite[\textit{e.g.}][]{o2010contact, robinson2013dynamics,
-leventhal2012inferring, groendyke2011bayesian}.
+leventhal2012inferring, groendyke2011bayesian, villandre2016assessment}.
 \textcite{groendyke2011bayesian} published their implementation as an
@@
 
-For ease of exposition, we simplifiy the notation of \textcite{del2012adaptive}
+For ease of exposition, we simplify the notation of \textcite{del2012adaptive}
 by dropping the subscripts on the variables which indicate the current
@@
 
-\subsection{Simulation experiments}
+\subsection{Kernel classifiers}
+\label{subsec:kernel}
 
-\subsubsection{Identification of separable parameters in kernel space}
-\label{subsubsec:kernel}
-
-Recall that our approximate Bayesian computation approach to fitting contact
-network models involves simulating transmission trees under a wide variety of
-parameter values, and then comparing these simulated trees to the true
-transmission tree. Values which produce trees similar to the observed
-transmission tree are distinguished as more likely than values which produce
-trees very different from the truth. In order for this type of analysis to
-succeed, it is critical that different parameter values produce different
-looking trees. Otherwise, if many different values produce trees which are too
-similar to each other, it will be impossible to distinguish which value is most
-consistent with the real tree. Just as importantly, trees simulated with
-similar parameter values must be similar to each other. In mathematical terms,
-we require the trees simulated from distinct parameter values to be
-\defn{separable} in tree space. The concept of separability is illustrated in
-Figure~\ref{fig:separable}.
+Recall that our \gls{ABC} approach to fitting contact network models involves
+simulating transmission trees under a wide variety of parameter values, and
+then comparing these simulated trees to the true transmission tree. Values
+which produce trees similar to the observed transmission tree are distinguished
+as more likely than values which produce trees very different from the truth.
+In order for this type of analysis to succeed, it is critical that different
+parameter values produce different looking trees. Otherwise, if many different
+values produce trees which are too similar to each other, it will be impossible
+to distinguish which value is most consistent with the real tree. Just as
+importantly, trees simulated with similar parameter values must be similar to
+each other. In mathematical terms, we require the trees simulated from distinct
+parameter values to be \defn{separable} in tree space. The concept of
+separability is illustrated in Figure~\ref{fig:separable}.
 
@@
   \includegraphics{separable.pdf}
-  \caption[Separable versus non-separable pararameters in kernel space]{
+  \caption[Separable versus non-separable parameters in kernel space]{
     Separable versus non-separable parameters in kernel space. Trees have been
     simulated under three sets of parameters, represented as blue, red, and
-    green. An observed tree is shown in black. Trees are layed out such that
+    green. An observed tree is shown in black. Trees are laid out such that
     the distance between two trees corresponds to their similarity. In panel A,
@@
 
-Before undertaking a complete ABC analysis, I analysed the parameters of the
+Before undertaking a complete ABC analysis, I analyzed the parameters of the
 \gls{BA} network model to to determine whether they could be separated in tree
@@
 2.0, which produced networks with very few ``hub'' nodes with extremely high
-degree, was a suitable upper bound (see Figure~\ref{fig:alphabds}). Therefore,
+degree, was a suitable upper bound (\cref{fig:alphabds}). Therefore,
 I chose to test the values 0.5, 1.0, and 1.5 for separability. The other
@@
   \centering
-  \label{fig:alphabds}
   \includegraphics{alpha-bounds.pdf}
@@
   }
+  \label{fig:alphabds}
 \end{figure}
@@
 
-\subsubsection{Grid search}
+\subsection{Grid search}
 
@@
 
-\subsubsection{Approximate Bayesian computation}
+\subsection{Approximate Bayesian computation}
 
@@
 
-\subsubsection{Characterization of power-law exponent in Barab\'asi-Albert networks}
+\subsection{Characterization of power-law exponent in Barab\'asi-Albert networks}
 
@@
   \input{\tablepath/realdata-bc}
-  \caption{Characteristics of published HIV datasets analysed with kernel-ABC.}
+  \caption{Characteristics of published HIV datasets analyzed with kernel-ABC.}
   \label{tab:realdata}

### 2016-03-29 doc/thesis/objective.tex
Writing stuff

@@
-Our objective is to develop a method to estimate structural parameters of the
-contact network underlying an observed transmission tree.
+\itshape
+This section is intended to be a synthesis of all the background material
+together into some research objectives.
+\normalfont
 
-\subsection{Prior work}
+The spread of a disease through a susceptible population is most often modelled
+by assuming either a homogeneously mixed population, or a population divided
+into a small number of homogeneously mixed compartments. This assumption, also
+called the ``law of mass action'', implies that any two individuals in the same
+compartment are equally likely to come into contact causing transmission.
+Although this is clearly unrealistic for most populations, the predictions they
+make have proven to be fairly accurate in practice. However, when there is
+substantial contact heterogeneity, such as when considering a sexually
+transmitted infection, the error introduced by assuming a panmictic population
+may be substantial.
 
-The present study is closely related to three groups of work with distinct
-objectives, which will be reviewed in detail below. 
-%, by
-%comparing trees simulated from different network models. These have
-%demonstrated that network structure does profoundly influence tree shape, but
-%did not attempt to quantitatively infer the parameter from observed trees. The
-%third group consists of studies which infer a structural network parameter from
-%observed infection and recovery times, using Bayesian methods. These are most
-%similar in aims to our work, but they rely on epidemiological observations
-%rather than evolutionary information from phylogenies.
+Contact network models provide an alternative to compartmental models which do
+not require the assumption of panmixia. In addition to more accurate
+prevalence predictions, the parameters of the networks themselves may be of
+interest from a public health perspective. For example, certain vaccination
+strategies may be more or less effective in curtailing an epidemic depending on
+the underlying network's degree distribution.
 
-The first and largest group of related studies are phylodynamic investigations
-of epidemiological parameters such as transmission rate, recovery rate, and
-basic reproductive number~\autocite{pybus2009evolutionary, volz2013viral}. Like
-our work, these studies make inferences about epidemiological processes from
-the genetic diversity of virus populations, which is usually represented in the
-form of a phylogeny. The majority of these employ a Bayesian \gls{MCMC}
-approach to infer parameters of an epidemiological model whose likelihood can
-be calculated, most often some variation of the
-birth-death~\autocite{kendall1948generalized} or
-coalescent~\autocite{kingman1982coalescent} models. Stadler
-\etal~\autocite{stadler2011estimating} develop a formula for the likelihood of
-a phylogeny with heterochronous tips under the birth-death model, which has
-been used to estimate the basic reproductive number of several viral
-epidemics~\autocite{stadler2011estimating}. However, the birth-death model is
-cannot tell us anything about population structure, as it assumes that every
-individual becomes infected at the same rate. Volz~\autocite{volz2012complex}
-writes down the likelihood of a heterochronous phylogeny under a coalescent
-model with arbitrarily complex population dynamics. This opens the door to more
-complex inferences about population structure, as the population can be
-partitioned into compartments with different transmission and recovery rates,
-but still assumes that each compartment is homogeneously mixed. In other words,
-the coalescent model can tell us about the \emph{global} structure of a
-population, such as whether there exists a high-risk subgroup, but not about
-the \emph{local} structure, such as the average number of contacts each
-individual has.
+Phylodynamic methods have been used to investigate a great diversity of
+epidemiological parameters. However, as far as we are aware, no phylodynamic
+methods have been developed to fit contact network models. The primary
+objective of this work is to develop such a method.
 
-The use of models which assume a \defn{panmictic} (that is, homogeneously
-mixed) population in phylodynamics has become widespread, so it is natural that
-some researchers have investigated the effects of this assumption. In
-particular, since phylodynamic methods use phylogenies as their data source,
-several studies have examined the shapes of phylogenies arising from
-non-panmictic populations.
+Network structure has a substantial impact on the speed and pattern of epidemic
+spread. Indeed, a small number of methods have been developed which exploit
+this fact to perform statistical inference. Contact network parameters, such as
+edge density and degree distribution, have been estimated from epidemiological
+data, such as incidence reports and individual recovery times.
 
-Finally, a third group of studies has used Bayesian methods to infer a
-structural network parameter from observed infection and recovery times. This
-third group is most similar in aims to our own work; rather than using
-epidemiological observations, we employ viral phylogenies. These were discussed
-in \cref{subsec:netoverview}.
+Contact networks' parameters can have a strong effect on the shape of
+transmission trees, and therefore on viral phylogenies, resulting from
+epidemics on those networks. Yet, this relationship has not yet been
+``inverted'' to quantitatively estimate network parameters from transmission
+trees.
+
+\newcommand{\G}{\mathcal{G}}
+\newcommand{\Nu}{\mathcal{N}}
+
+Although we have not proven it analytically, calculating the likelihood of the
+parameters of a contact network models seems likely to be an intractable
+problem. This would imply that these models are amenable to neither \gls{ML}
+nor Bayesian inference. To illustrate this, consider a contact network model
+with parameters $\theta$, and an observed transmission tree $T$ with $n$ tips.
+In general, we do not know the labels of the internal nodes of $T$, only the
+labels of its tips. To fit this model using likelihood-based methods, we must
+calculate the likelihood of $\theta$, that is, $\Pr(T \mid \theta)$. Let $\G$
+be the set of all possible contact networks, and $\Nu$ be the set of all
+possible labellings of the internal nodes of $T$. We can write the likelihood as
+\begin{align}
+\begin{split}
+  \label{eq:netlik}
+  \Pr(T \mid \theta)
+    &= \sum_{\nu \in \Nu} \Pr(T, \nu \mid \theta) \\
+    &= \sum_{G \in \G} \sum_{\nu \in \Nu} \Pr(T, \nu \mid G, \theta) \Pr(G \mid \theta) \\
+    &= \sum_{G \in \G} \sum_{\nu \in \Nu} \Pr(T, \nu \mid G) \Pr(G \mid \theta),
+\end{split}
+\end{align}
+the last equality following from the fact that $T$ and $\nu$ depend only on
+$G$, not on $\theta$. Although $\Pr(T, \nu \mid G)$ and $\Pr(G \mid \theta)$
+may individually be straightforward to calculate, the number of possible
+directed graphs on $N$ nodes is $2^{N(N-1)}$, larger if the nodes and edges in
+the graph may have different labels or attributes. Hence, the number of terms
+in the sum is at least exponential in $n$, more if we do not know \textit{a
+priori} how many nodes are in the network (as is likely). In addition,
+\cref{eq:netlik} assumes that $T$ is complete, meaning that all infected
+individuals were sampled. This is rarely the case in practice - most often, the
+observed tree is a subsampled version of the true tree. In this case, the
+likelihood calculation becomes even more complex, because we must also sum over
+all possible complete trees.
+
+Depending on the network model studied, it is possible that \cref{eq:netlik}
+could be simplified into a tractable expression. However, a simpler alternative 
+to likelihood-based methods, which would apply to any network model, is
+provided by \gls{ABC}. All of the ingredients required to apply \gls{ABC} are
+readily available. Simulating networks is straightforward under a variety of
+models. Epidemics on those networks, and the corresponding transmission trees,
+can also be easily simulated. As mentioned above, contact networks can
+profoundly affect transmission tree shape, and those shapes can be compared
+using a highly informative similarity measure. \Gls{SMC} has several advantages
+over other algorithms for \gls{ABC}~\autocite{mckinley2009inference}, including
+a recently-developed adaptive algorithm requiring minimal tuning on the part of
+the user~\autocite{del2012adaptive}. In summary, our method to infer contact
+network parameters will combine the following: stochastic simulation of
+epidemics on networks, the tree kernel, and adaptive \gls{ABC}-\gls{SMC}.

### 2016-03-29 doc/thesis/preamble.tex
Writing stuff

\documentclass[12pt]{report}
\usepackage[osf]{garamondx}
\usepackage[garamondx,cmbraces]{newtxmath}
\usepackage{marginnote}
\usepackage{fullpage}
\usepackage{setspace}
\usepackage[backend=biber, style=numeric-comp, sorting=none]{biblatex}
\usepackage{nicefrac}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{multicol}
\usepackage{multirow}
\usepackage[hidelinks]{hyperref}
\usepackage[acronym,nonumberlist]{glossaries}
\usepackage{glossary-mcols}
\usepackage{cleveref}
\usepackage{array}
\usepackage{textcomp}
\usepackage{float}

\newcommand{\comment}[1]{(\textbf{#1})}

\newcommand{\defn}[1]{\textit{#1}}
\newcommand{\software}[1]{\textit{#1}}
\newcommand{\etal}{\textit{et al.}}
\newcommand{\eg}{\textit{e.g.}}

\newcommand{\set}[1]{\left\lbrace#1\right\rbrace}
\newcommand{\sett}[1]{\{#1\}}
\renewcommand{\emptyset}{\varnothing}
\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\star}[1]{#1^*}
\renewcommand{\d}{\mathrm{d}\,}

\newcommand{\N}{\mathcal{N}}
\renewcommand{\L}{\mathcal{L}}

\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

\DeclareMathOperator{\Exponential}{Exponential}
\DeclareMathOperator{\Uniform}{Uniform}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\Var}{Var}

\DeclareMathOperator{\inc}{in}
\DeclareMathOperator{\out}{out}

\newfloat{data}{hb}{lod}
\floatname{data}{Data}
\crefname{data}{data}{data}

\newacronym{MCMC}{MCMC}{Markov chain Monte Carlo}
\newacronym{ER}{ER}{Erd\H{o}s-R\'enyi}
\newacronym{BA}{BA}{Barab\'asi-Albert}
\newacronym{WS}{WS}{Watts-Strogatz}
\newacronym{SI}{SI}{susceptible-infected}
\newacronym{SVM}{SVM}{support vector machine}
\newacronym{PCA}{PCA}{principal components analysis}
\newacronym{SMC}{SMC}{sequential Monte Carlo}
\newacronym{ABC}{ABC}{approximate Bayesian computation}
\newacronym{nltt}{nLTT}{normalized lineages-through-time}
\newacronym{ltt}{LTT}{lineages-through-time}
\newacronym{pdf}{pdf}{probability density function}
\newacronym{rv}{rv}{random variable}
\newacronym{ML}{ML}{maximum likelihood}
\newacronym{MAP}{MAP}{maximum \textit{a posteriori}}
\newacronym{GSL}{GSL}{GNU scientific library}
\newacronym{SIR}{SIR}{susceptible-infected-recovered}
\newacronym{IS}{IS}{importance sampling}
\newacronym{SIS}{SIS}{sequential importance sampling}
\newacronym{HMM}{HMM}{hidden Markov model}
\newacronym{iid}{i.i.d.}{independent and identically distributed}
\newacronym{HPD}{HPD}{highest posterior density}
\newacronym{HIV}{HIV}{human immunodeficiency virus}
\newacronym{SARS}{SARS}{severe acute respiratory syndrome}
\newacronym{ODE}{ODE}{ordinary differential equation}
\newacronym{GLM}{GLM}{generalized linear model}

\newglossaryentry{alpha}{
  name=$\alpha$,
  description=preferential attachment power parameter in Barab\'asi-Albert networks
}
\newglossaryentry{m}{
  name=$m$,
  description=number of edges added per vertex when constructing a Barab\'asi-Albert network
}
\newglossaryentry{N}{
  name=$N$,
  description=number of nodes in the network
}
\newglossaryentry{I}{
  name=$I$,
  description=number of nodes which are eventually infected
}
\newglossaryentry{gamma}{
  name=$\gamma$,
  description=exponent of power-law degree distribution in scale-free networks
}
\newglossaryentry{p}{
  name=$p$,
  description=edge density parameter in Erd\H{o}s-R\'enyi networks
}
\newglossaryentry{lambda}{
  name=$\lambda$,
  description=decay factor meta-parameter for tree kernel
}
\newglossaryentry{sigma}{
  name=$\sigma$,
  description=radial basis function variance meta-parameter for tree kernel
}
\newglossaryentry{n}{
  name=$n$,
  description=number of particles used for ABC-SMC
}
\newglossaryentry{M}{
  name=$M$,
  description=number of simulated datasets per particle in ABC-SMC
}
\newglossaryentry{tmrca}{
  name=$t_{\mathrm{mrca}}$,
  description=time to most recent common ancestor of a group of extant taxa
}

\graphicspath{{../figures/}}
\newcommand{\tablepath}{../tables}
\frenchspacing
\onehalfspacing

\title{Phylodynamic inference of contact network parameters through approximate
Bayesian computation}
\author{Rosemary M. McCloskey and Art F.Y. Poon}

\addbibresource{papers.bib}
\makeglossaries

### 2016-03-29 doc/thesis/preface.tex
Writing stuff

This work was conducted at the BC Centre for Excellence in HIV/AIDS under the
supervision of Dr. Art Poon. I wrote the code, performed the experiments, and
wrote the thesis. Dr. Richard Liang assisted with the development of the
Gillespie simulation algorithm.

A version of chapters 2 and 3 has been submitted for publication in Molecular
Biology and Evolution, with the title ``Phylogenetic estimation of contact
network parameters with kernel-ABC''. A presentation with the same title was
given at the 23rd HIV Dynamics and Evolution meeting on April 25, in Woods
Hole, Massachussets, USA.

### 2016-03-29 doc/thesis/results.Rnw
Writing stuff

@@
 
-\subsection{Separable parameters}
+\subsection{Kernel classifiers}
 
+% overview of experiment
 Our first simulation-based experiment was designed to determine which network
@@
 set of kernel meta-parameters to carry through to future experiments. For each
-parameter of the \gls{BA} model, we simulated trees under three distinct
-values, and computed the cross-validation accuracy of a kernel-\gls{SVM}
-classifier for the parameter.
+of the four \gls{BA} model parameters, we simulated trees under three distinct
+values of that parameter while holding the others fixed, then cross-validated a
+kernel-\gls{SVM} classifier for that parameter.
 
-In Figure~\ref{fig:patrees}, we show that trees simulated under different
-values of \gls{alpha} are visibly quite distinctive. In particular, higher
-values of \gls{alpha} result in networks with a small number of highly
-connected nodes (see Figure~\ref{fig:pabounds}) which, once infected, are
-likely to transmit to many other nodes. This results in a more unbalanced,
-ladder-like structure in the phylogeny. Kernel-\gls{PCA} projections show that
-all three \gls{alpha} values are well separated from each other under several
-different sampling scenarios (Figure~\ref{fig:pakpca}). With smaller trees,
-it becomes harder to visually disinguish \gls{alpha} = 0.5 from \gls{alpha} =
-1.0 using the first two principal components.
+<<treestats, include=FALSE>>=
+    params <- c("alpha", "I", "m", "N")
+    ptn <- paste0("../../simulations/kernel-", params, "/statistics/*")
+    files <- lapply(ptn, Sys.glob)
+    d <- lapply(files, collect.data)
+    d <- mapply(cbind, d, parameter=as.list(params), SIMPLIFY=FALSE)
+    lapply(d, setDT)
+    d[[1]] <- subset(d[[1]], m == 2)
+    setnames(d[[1]], "nsimnode", "I")
+@
+
+% trees look different
+Trees simulated under different values of \gls{alpha} are visibly quite
+distinctive (\cref{fig:patrees}). In particular, higher values of \gls{alpha}
+produce networks with a small number of highly connected nodes which, once
+infected, are likely to transmit to many other nodes (\cref{fig:alphabds}).
+This results in a more unbalanced, ladder-like structure in the phylogeny,
+compared to networks with lower \gls{alpha} values. None of the other three
+parameters produced trees which were as easily distinguished from each other
+(\cref{fig:Itrees,fig:mtrees,fig:Ntrees,fig:Itrees}). However, higher $I$
+values produce trees which appear to have longer terminal branches,
+
+Kernel-\gls{PCA} projections show that all three \gls{alpha} values are well
+separated from each other in feature space under several different sampling
+scenarios (Figure~\ref{fig:pakpca}). With smaller trees, it becomes harder to
+visually distinguish \gls{alpha} = 0.5 from \gls{alpha} = 1.0 using the first
+two principal components.
 
@@
 classifier for \gls{alpha} was very high under a variety of prevalence and
-sampling scenarios (Figure~\ref{fig:pacrossv}). Accuracy was highest, with
+sampling scenarios. Accuracy was highest, with
 $R^2$ values above 0.95, for the largest trees and complete sampling (bottom
@@
   \centering
-  \label{fig:patrees}
   \includegraphics{kernel-alpha-tree.pdf}
   \caption[Visibly distinctive trees simulated under three values of \gls{alpha}]{
-    Epidemics of 1000 infected were simulated on \gls{BA} networks of 5000
-    nodes, with \gls{alpha} equal to 0.5, 1.0, or 1.5. Transmission trees were
-    created by sampling 500 infected nodes. Higher \gls{alpha} values produce
-    networks with a small number of highly-connected nodes, which results in a
-    highly unbalanced, ladder-like tree structure.
+    Epidemics simulated on \gls{BA} networks of 5000 nodes, with \gls{alpha}
+    equal to 0.5, 1.0, or 1.5, until 1000 individuals were infected.
+    Transmission trees were created by sampling 500 infected nodes. Higher
+    \gls{alpha} values produced networks with a small number of
+    highly-connected nodes, resulting in highly unbalanced, ladder-like trees.
   }
+  \label{fig:patrees}
 \end{figure}
@@
   \centering
-  \label{fig:pacrossv}
-  \includegraphics{kernel-alpha-crossv.pdf}
-  \caption[Cross-validation performance of kernel support vector machine
-  classifier for preferential attachment power]{
-    Cross-validation performance of kernel support vector machine classifier
-    for preferential attachment power.
-  }
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \label{fig:pakpca}
   \includegraphics{kernel-kpca.pdf}
@@
   }
+  \label{fig:pakpca}
 \end{figure}

### 2016-03-30 doc/thesis/objective.tex
Small changes to objective

@@
-\itshape
-This section is intended to be a synthesis of all the background material
-together into some research objectives.
-\normalfont
-
-The spread of a disease through a susceptible population is most often modelled
-by assuming either a homogeneously mixed population, or a population divided
-into a small number of homogeneously mixed compartments. This assumption, also
-called the ``law of mass action'', implies that any two individuals in the same
-compartment are equally likely to come into contact causing transmission.
-Although this is clearly unrealistic for most populations, the predictions they
-make have proven to be fairly accurate in practice. However, when there is
-substantial contact heterogeneity, such as when considering a sexually
-transmitted infection, the error introduced by assuming a panmictic population
-may be substantial.
-
-Contact network models provide an alternative to compartmental models which do
-not require the assumption of panmixia. In addition to more accurate
-prevalence predictions, the parameters of the networks themselves may be of
-interest from a public health perspective. For example, certain vaccination
-strategies may be more or less effective in curtailing an epidemic depending on
-the underlying network's degree distribution.
-
-Phylodynamic methods have been used to investigate a great diversity of
-epidemiological parameters. However, as far as we are aware, no phylodynamic
-methods have been developed to fit contact network models. The primary
-objective of this work is to develop such a method.
-
-Network structure has a substantial impact on the speed and pattern of epidemic
-spread. Indeed, a small number of methods have been developed which exploit
-this fact to perform statistical inference. Contact network parameters, such as
-edge density and degree distribution, have been estimated from epidemiological
-data, such as incidence reports and individual recovery times.
-
-Contact networks' parameters can have a strong effect on the shape of
-transmission trees, and therefore on viral phylogenies, resulting from
-epidemics on those networks. Yet, this relationship has not yet been
-``inverted'' to quantitatively estimate network parameters from transmission
-trees.
+The spread of a disease is most often modelled by assuming either a
+homogeneously mixed population, or a population divided into a small number of
+homogeneously mixed compartments. This assumption, also called the ``law of
+mass action'', implies that any two individuals in the same compartment are
+equally likely to come into contact causing transmission. Although this
+provides a reasonable approximation in many cases, the error introduced by
+assuming a panmictic population can be substantial when significant contact
+heterogeneity exists in the underlying population. Contact network models
+provide an alternative to compartmental models which do not require the
+assumption of panmixia. In addition to more accurate predictions, the
+parameters of the networks themselves may be of interest from a public health
+perspective. For example, certain vaccination strategies may be more or less
+effective in curtailing an epidemic depending on the underlying network's
+degree distribution. Phylodynamic methods have been used to fit many different
+types of model to phylogenetic data, but as far as we know, no methods have yet
+been developed to fit contact network models. The primary objective of this
+work is to develop such a method.
 
@@
 
-Although we have not proven it analytically, calculating the likelihood of the
-parameters of a contact network models seems likely to be an intractable
-problem. This would imply that these models are amenable to neither \gls{ML}
-nor Bayesian inference. To illustrate this, consider a contact network model
-with parameters $\theta$, and an observed transmission tree $T$ with $n$ tips.
-In general, we do not know the labels of the internal nodes of $T$, only the
-labels of its tips. To fit this model using likelihood-based methods, we must
-calculate the likelihood of $\theta$, that is, $\Pr(T \mid \theta)$. Let $\G$
-be the set of all possible contact networks, and $\Nu$ be the set of all
-possible labellings of the internal nodes of $T$. We can write the likelihood as
+Calculating the likelihood of the parameters of a contact network models seems
+likely to be an intractable problem, which would imply that these models are
+amenable to neither \gls{ML} nor Bayesian inference. We not proven this is the
+case, but some intuition can be provided by examining the process involved in
+the likelihood calculation. Consider a contact network model with parameters
+$\theta$, and an observed transmission tree $T$ with $n$ tips.  In general, we
+do not know the labels of the internal nodes of $T$, only the labels of its
+tips. To fit this model using likelihood-based methods, we must calculate the
+likelihood of $\theta$, that is, $\Pr(T \mid \theta)$. Let $\G$ be the set of
+all possible contact networks, and $\Nu$ be the set of all possible labellings
+of the internal nodes of $T$. We can write the likelihood as
 \begin{align}
@@
 to likelihood-based methods, which would apply to any network model, is
-provided by \gls{ABC}. All of the ingredients required to apply \gls{ABC} are
-readily available. Simulating networks is straightforward under a variety of
-models. Epidemics on those networks, and the corresponding transmission trees,
-can also be easily simulated. As mentioned above, contact networks can
-profoundly affect transmission tree shape, and those shapes can be compared
-using a highly informative similarity measure. \Gls{SMC} has several advantages
-over other algorithms for \gls{ABC}~\autocite{mckinley2009inference}, including
-a recently-developed adaptive algorithm requiring minimal tuning on the part of
-the user~\autocite{del2012adaptive}. In summary, our method to infer contact
-network parameters will combine the following: stochastic simulation of
-epidemics on networks, the tree kernel, and adaptive \gls{ABC}-\gls{SMC}.
+provided by \gls{ABC}. All of the ingredients required to apply \gls{ABC} to
+this problem are readily available. Simulating networks is straightforward
+under a variety of models. Epidemics on those networks, and the corresponding
+transmission trees, can also be easily simulated. As mentioned above, contact
+networks can profoundly affect transmission tree shape, and those shapes can be
+compared using a highly informative similarity measure. \Gls{SMC} has several
+advantages over other algorithms for \gls{ABC}~\autocite{mckinley2009inference},
+including a recently-developed adaptive algorithm requiring minimal tuning on
+the part of the user~\autocite{del2012adaptive}. In summary, our method to
+infer contact network parameters will combine the following: stochastic
+simulation of epidemics on networks, the tree kernel, and adaptive
+\gls{ABC}-\gls{SMC}. Since our distance measure is a kernel function, our
+method is a type of kernel-\gls{ABC}. For ease of exposition, we will often use
+the term ``kernel-\gls{ABC}'' to refer to our method specifically.
+
+Empirical studies of sexual contact networks have found that these networks
+tend to be scale-free, meaning that their degree distributions follow a power
+law (although there has been some disagreement). Preferential attachment has
+been postulated as a mechanism by which scale-free networks could be generated.
+This makes the \gls{BA} model, one of the simplest preferential attachment
+models, a natural choice to explore with our method. The second aim of this
+work is to use simulations to investigate the parameters of the \gls{BA} model,
+including whether they have a detectable impact on tree shape, and whether they
+can be accurately recovered using kernel-\gls{ABC}.
+
+Due to its high global prevalence and fast mutation rate, \gls{HIV} is one of
+the most commonly-studied viruses in a phylodynamic context. Consequently, a
+large volume of \gls{HIV} sequence data is publicly available, more than for
+any other pathogen, and including sequences sampled from diverse geographic and
+demographic contexts. Since \gls{HIV} is almost always spread through either
+sexual contact or sharing of injection drug supplies, the contact networks
+underlying \gls{HIV} epidemics are highly structured. Moreover, since no cure
+yet exists, efforts to curtail the progression of an epidemic have relied on
+preventing further transmissions through measures such as \gls{tasp} and
+education leading to behaviour change. The effectiveness of this type of
+intervention can vary significantly based on the underlying structure of the
+network and the particular nodes to whom the intervention is targeted. Due to
+this combination of data availability and potential public health impact,
+\gls{HIV} is an obvious context in which our method could be applied.
+Therefore, the third and final aim of this work is to apply kernel-\gls{ABC} to
+fit the \gls{BA} model to existing \gls{HIV} outbreaks.
+
+To summarize, this work has three objectives. First, we will develop a method
+which uses kernel-\gls{ABC} to infer parameters of contact network models from
+observed transmission trees. Second, we will use simulations to characterize
+the parameters of the \gls{BA} network model in terms of their effect on tree
+shape and how accurately they can be recovered with kernel-\gls{ABC}. Finally,
+we will apply the method fit the \gls{BA} model to several real-world \gls{HIV}
+datasets.

### 2016-03-30 doc/thesis/results.Rnw
Small changes to objective

@@
 <<treestats, include=FALSE>>=
-    params <- c("alpha", "I", "m", "N")
-    ptn <- paste0("../../simulations/kernel-", params, "/statistics/*")
-    files <- lapply(ptn, Sys.glob)
-    d <- lapply(files, collect.data)
-    d <- mapply(cbind, d, parameter=as.list(params), SIMPLIFY=FALSE)
-    lapply(d, setDT)
-    d[[1]] <- subset(d[[1]], m == 2)
-    setnames(d[[1]], "nsimnode", "I")
+    #params <- c("alpha", "I", "m", "N")
+    #ptn <- paste0("../../simulations/kernel-", params, "/statistics/*")
+    #files <- lapply(ptn, Sys.glob)
+    #d <- lapply(files, collect.data)
+    #d <- mapply(cbind, d, parameter=as.list(params), SIMPLIFY=FALSE)
+    #lapply(d, setDT)
+    #d[[1]] <- subset(d[[1]], m == 2)
+    #setnames(d[[1]], "nsimnode", "I")
 @

### 2016-03-31 doc/thesis/conclusion.tex
Finished results! Yes!

@@
 their own preferred tree building method.
+
+Our use of the \gls{BA} model makes several simplifying assumptions. First, we
+assume homogeneity across the network with respect to node behaviour and
+transmission risk. In reality, the attraction to high-degree nodes seems likely
+to vary among individuals, as does their risk of transmitting or contracting
+the virus. We have also assumed that all transmission risks are symmetric,
+which is clearly false for all known modes of \gls{HIV} transmission, and that
+infected individuals never recover but remain infectious indefinitely. These
+assumptions were made for the purpose of keeping the model as simple as
+possible, since this is the very first attempt to fit a contact network model
+in a phylodynamic context. However, the Gillespie simulation algorithm built
+into \software{netabc} can handle arbitrary transmission and removal rates
+which need not be homogeneous across the network. Moreover, it is possible to
+use kernel-\gls{ABC} to fit a model which relaxes some or all of these
+assumptions, which may be a fruitful avenue for future investigation.

### 2016-03-31 doc/thesis/methods.tex
Finished results! Yes!

@@
-\subsection{Computer program}
+\subsection{Kernel-ABC method}
 
-\software{Netabc} is a computer program designed to perform statistical
+I wrote a computer program called \software{netabc} to perform statistical
 inference of contact network parameters from an estimated transmission tree
@@
 phylogenetic kernel, to compare simulated to observed transmission trees; and
-adaptive \gls{ABC}-\gls{SMC}, to maintain a population of particles and
-advance it toward the \gls{ABC} target distribution. We give a high-level
-overview of the program here, before describing these three components in
-detail.
+adaptive \gls{ABC}-\gls{SMC}, to maintain a population of particles and advance
+it toward the \gls{ABC} target distribution. We give a high-level overview of
+the program here, before describing these three components in detail.
 
@@
 \software{treestat}, to compute various summary statistics of phylogenies. The
-programs are freely available at \url{github.com/rmcclosk/netabc}.
+programs are freely available at \url{https://github.com/rmcclosk/netabc}.
 
@@
 
-The algorithm may be stopped when the tolerance reaches a
-user-defined final value, or when the rate of acceptance of the
-Metropolis-Hastings kernel reaches a user-defined threshold. Following a
-heuristic applied by the authors~\autocite{del2012adaptive}, we used the latter
-stopping criterion, accepting the SMC approximation to the posterior when the
-\gls{MCMC} acceptance rate dropped below 1.5\%.
+The algorithm may be stopped when the tolerance reaches a user-defined final
+value, or when the rate of acceptance of the \gls{MH} kernel reaches a
+user-defined threshold.
 
-\subsection{Kernel classifiers}
+\subsection{Analysis of \acrlong{BA} model}
+
+We investigated four parameters related to the \gls{BA} model, denoted
+\gls{N}, \gls{m}, \gls{alpha}, \gls{I}. The first three of these are parameters
+of the model itself, while \gls{I} is related to the simulation of transmission
+trees over the network. However, we will refer to all four as \gls{BA}
+parameters. \gls{N} denotes the total number of nodes in the network, or
+equivalently, susceptible individuals in the population. When a node is added
+to the network, \gls{m} new undirected edges are added incident to it, and are
+attached to existing nodes of degree $k$ with probability proportional to
+$k^\alpha$ (\cref{subsec:pa}). To simulate transmission trees over a
+\gls{BA} network, we allowed an epidemic to spread until \gls{I} nodes were
+infected, and sampled a transmission tree at that time. We assumed that all
+contacts had symmetric transmission risk, which was implemented by replacing
+each undirected edge in the network with two directed edges (one in each
+direction).
+
+We did not consider the time scale of the transmission trees in these
+simulations, only their shape. Therefore, the transmission rate along each edge
+in the network was set to 1, and all transmission trees' branch lengths were
+scaled by their mean. The removal rate of each node was set to 0, implying no
+recovery or death in the population. These assumptions are similar to those
+made by \textcite{leventhal2012inferring}.
+
+\subsubsection*{Kernel classifiers}
 \label{subsec:kernel}
 
-Recall that our \gls{ABC} approach to fitting contact network models involves
-simulating transmission trees under a wide variety of parameter values, and
-then comparing these simulated trees to the true transmission tree. Values
-which produce trees similar to the observed transmission tree are distinguished
-as more likely than values which produce trees very different from the truth.
-In order for this type of analysis to succeed, it is critical that different
-parameter values produce different looking trees. Otherwise, if many different
-values produce trees which are too similar to each other, it will be impossible
-to distinguish which value is most consistent with the real tree. Just as
-importantly, trees simulated with similar parameter values must be similar to
-each other. In mathematical terms, we require the trees simulated from distinct
-parameter values to be \defn{separable} in tree space. The concept of
-separability is illustrated in Figure~\ref{fig:separable}.
+The experiments presented here involved a large number of variables which were
+varied combinatorially. For ease of exposition, we will describe a single
+experiment first, then enumerate the values of all variables for which the
+experiment was repeated. The parameters of the tree kernel, $\lambda$ and
+$\sigma$ (\cref{subsubsec:treeshape}) will be referred to as
+\defn{meta-parameters} to distinguish them from the parameters of the \gls{BA}
+model. With the exception of our own binaries, all analyses were done in
+\software{R}, and all packages listed below are \software{R} packages.
+
+The attachment power parameter \gls{alpha} was varied among three values: 0.5,
+1.0, and 1.5. For each value, the \software{sample\_pa} function in the
+\software{igraph} package was used to simulate 100 networks, with the other
+parameters set to \gls{N} = 5000 and \gls{m} = 2. This step yielded a total of
+300 networks. An epidemic was simulated on each network using our
+\software{nettree} binary until \gls{I} = 1000 nodes were infected, at which
+point 500 of them were sampled to form a transmission tree. A total of 300
+transmission trees were thus obtained, comprised of 100 trees for each of the
+three values of \gls{alpha}. The trees were ``ladderized'' so that the subtree
+descending from the left child of each node was not smaller than that
+descending from the right child. Summary statistics, such as Sackin's index and
+the ratio of internal to terminal branch lengths, were computed for each
+simulated tree using our \software{treestat} binary. The trees were visualized
+using the \software{ape} package. Our \software{treekernel} binary was used to
+calculate the value of the kernel for each pair of trees, with the
+meta-parameters set to $\lambda = 0.3$ and $\sigma = 4$. These values were
+stored in a symmetric 300 $\times$ 300 kernel matrix. Similarly, we computed
+the \gls{nltt} statistic between each pair of trees using our
+\software{treestat} binary, and stored them in a second $300 \times 300$
+matrix.
+
+To investigate the effect of \gls{alpha} on tree shape, we constructed
+classifiers for \gls{alpha} based on three statistics. First, we used the
+\software{kernlab} package~\autocite{karatzoglou2004kernlab} to create a
+\gls{kSVM} classifier using the computed kernel matrix. Second, we used the
+\software{e1071} package to create an ordinary \gls{SVM} classifier using the
+pairwise \gls{nltt} matrix. Finally, we performed an ordinary linear regression
+of \gls{alpha} against Sackin's index. Each of these classifiers was evaluated
+with 1000 two-fold cross-validations. We also performed a \gls{kPCA} projection
+of the kernel matrix, and used it to visualize the separation of the different
+\gls{alpha} values in the tree kernel's feature space. A schematic of this
+experiment is presented in \cref{fig:kernelexpt}.
+
+Similar experiments were performed with the values shown in
+\cref{tab:kernelexpt}. The other three \gls{BA} parameters, namely \gls{N},
+\gls{m}, and \gls{I}, were each varied while holding the others fixed. The
+experiments for \gls{alpha}, \gls{m}, and \gls{N} were repeated with three
+different values of \gls{I}. All experiments were repeated with trees having
+three different numbers of tips. Kernel matrices were computed for all pairs of
+the meta-parameters \gls{lambda} = \sett{0.2, 0.3, 0.4} and \gls{sigma} =
+\sett{\nicefrac18, \nicefrac14, \nicefrac12, 1, 2, 4, 8}.
+
+\begin{landscape}
+\begin{table}[ht]
+  \centering
+  \input{\tablepath/kernel-expt}
+  \caption[Variables used in tree kernel simulation experiments]
+  {
+    Values of parameters and other variables used in tree kernel simulation
+    experiments. Each row corresponds to one of the \gls{BA} model parameters.
+    One kernel matrix was created for every combination of values except the
+    one indicated in the ``varied parameter'' column, which was varied when
+    producing simulated trees.
+  }
+  \label{tab:kernelexpt}
+\end{table}
+
+\begin{table}[ht]
+  \centering
+  \input{\tablepath/gridsearch-expt}
+  \caption[Variables used in grid search experiments]
+  {
+    Variables and \gls{BA} parameter values used for grid search experiments. 
+    Trees were simulated under the test values, and compared to a grid of trees
+    simulated under the grid values. Kernel scores were used to calculate point
+    estimates and credible intervals for the test values.
+  }
+  \label{tab:gridexpt}
+\end{table}
+\end{landscape}
 
@@
   \centering
-  \includegraphics{separable.pdf}
-  \caption[Separable versus non-separable parameters in kernel space]{
-    Separable versus non-separable parameters in kernel space. Trees have been
-    simulated under three sets of parameters, represented as blue, red, and
-    green. An observed tree is shown in black. Trees are laid out such that
-    the distance between two trees corresponds to their similarity. In panel A,
-    trees from the same parameter set are similar to each other, but different
-    from other trees. The true tree is most consistent with the red parameters.
-    In panels B and C, trees from the same parameter set are not similar to
-    each other (B), or trees from different parameter sets are similar to each
-    other (C). It's difficult to say which parameters the true tree is most
-    consistent with.
+  \includegraphics{kernel-expt.pdf}
+  \caption[Schematic of kernel classifier experiment]{
+    Schematic of investigation of \gls{BA} model parameters using the tree
+    kernel.
   }
-  \label{fig:separable}
+  \label{fig:kernelexpt}
 \end{figure}
 
-Before undertaking a complete ABC analysis, I analyzed the parameters of the
-\gls{BA} network model to to determine whether they could be separated in tree
-kernel space. A graphical schematic of the analysis undertaken here is given in
-Figure~\ref{fig:kernelexpt}.
+\subsubsection*{Grid search}
 
-\begin{figure}[ht]
+As in the previous section, we will begin by describing a single experiment,
+and then list the variables for which similar experiments were performed. We
+varied \gls{alpha} along a narrowly spaced grid of values: 0, 0.01, \ldots, 2.
+For each value, fifteen networks were generated with \software{igraph}, and
+transmission trees were simulated over each using \software{nettree}. These
+trees will be referred to as ``grid trees'', and their associated values ``grid
+values''. Next, one further test tree was simulated with the test value
+\gls{alpha} = 0. Both the grid trees and the test tree had 500 tips, and were
+simulated with the other \gls{BA} parameters set to $N$ = 5000, $m$ = 2, and
+$I$ = 1000. The test tree was compared to each of the grid trees using the tree
+kernel, with the meta-parameters set to $\lambda = 0.3$ and $\sigma = 4$, using
+the \software{treekernel} binary. The median kernel score was calculated for
+each grid value. Grid values were resampled with probability proportional to
+their median kernel scores. A point estimate for the test value was obtained by
+taking the highest point of an estimated kernel density of the resampled grid
+values, and a 95\% credible interval was obtained using the
+\software{HPDinterval} function in the \software{coda} package.
+
+Each experiment of the type just described was repeated ten times with the same
+test value. Similar experiments were performed for each of the four \gls{BA}
+parameters, with several test values and trees of varying sizes. The variables
+are listed in \cref{tab:gridexpt}.
+
+\subsubsection*{Approximate Bayesian computation}
+
+To test the full kernel-\gls{ABC} algorithm, we simulated four trees each under
+a variety of parameter values, and ran the \software{netabc} program to
+estimate posterior distributions for the parameters. The parameter values and
+priors used are listed in \ref{tab:abcexpt}. The tree kernel meta-parameters
+were set to $\lambda = 0.3$ and $\sigma = 4$. The \gls{SMC} algorithm was run
+with 1000 particles, five sampled datasets per particle, and the parameter
+controlling the tradeoff between speed and approximation quality
+\textcite[called ``$\alpha$'' in][]{del2012adaptive} set to 0.95. The algorithm
+was stopped when the acceptance rate of the \gls{MH} kernel dropped below
+1.5\%, the same criterion used by \citeauthor{del2012adaptive}. Approximate
+marginal posterior densities for each parameter were calculated using the
+\software{density} function in \software{R} applied to the final weighted
+population of particles. Credible intervals were obtained for each parameter
+using the \software{HPDinterval} function in the \software{coda} package.
+
+\begin{table}[ht]
   \centering
-  \label{fig:kernelexpt}
-  \includegraphics{kernel-expt.pdf}
-  \caption[Schematic of first set of simulation experiments]{
-    Schematic of first set of simulation experiments to determine separable
-    parameters in kernel space, along with optimal tree kernel meta-parameters.
+  \input{\tablepath/abc-expt}
+  \caption[Variables used in grid search experiments]
+  {
+    Variables and \gls{BA} parameter values used for \gls{ABC} validation
+    experiments. Trees were simulated under the test values, and
+    kernel-\gls{ABC} was used to re-estimate posterior distributions for the
+    \gls{BA} parameters without training.
   }
-\end{figure}
+  \label{tab:abcexpt}
+\end{table}
 
-The method of testing for separability was described previously
-in~\autocite{poon2015phylodynamic}. As a concrete example, consider the
-attachment power parameter \gls{alpha}. This parameter describes the strength
-of attraction to highly connected nodes and is bounded below by zero indicating
-no extra attraction. By qualitative observation, I determined that a power of
-2.0, which produced networks with very few ``hub'' nodes with extremely high
-degree, was a suitable upper bound (\cref{fig:alphabds}). Therefore,
-I chose to test the values 0.5, 1.0, and 1.5 for separability. The other
-parameters were fixed: the number of nodes in the network was 5000, and the
-mean degree of each node in the network was four. As discussed in
-subsection~\ref{subsubsec:generative}, this is the smallest mean degree value
-for preferential attachment networks which produces networks which are more
-than trees.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{alpha-bounds.pdf}
-  \caption[Upper and lower bounds on preferential attachment power]{
-    Qualitative justification for choice of zero and two as lower and upper
-    bounds on preferential attachment power $\alpha$. (A) Preferential
-    attachment network on 50 nodes with $\alpha = 0$, the lower bound enforced
-    by the model. (B) Preferential attachment network on 50 nodes with $\alpha
-    = 2$, where a few nodes have very high degree but the majority have very
-    low degree. (C) Density plot of node degrees in a 5000-node network with
-    $\alpha = 0$. The maximum degree of a node in this network was 24. (D)
-    Density plot of node degrees in a 5000-node network with $\alpha = 2$. The
-    maximum degree of a node in this network was 4942. 
-  }
-  \label{fig:alphabds}
-\end{figure}
-
-For each of the values of $\alpha$, I generated 100 networks, each with 5000
-nodes. An epidemic was simulated over each network (see
-\cref{subsubsec:nettree}) until 1000 nodes were infected, and 500 of those
-infected nodes were sampled to form a transmission tree. This resulted in 300
-total simulated transmission trees - 100 for each of the three values of
-$\alpha$. The data generation steps are shown on the left side of
-Figure~\ref{fig:kernelexpt}. Next, I computed the tree
-kernel~\autocite{poon2013mapping} (see subsection~\ref{subsubsec:treeshape})
-for each pair of trees. These values were placed into a 300 $\times$ 300 kernel
-matrix, where the value at the $(i, j)$th position was the tree kernel of the
-$i$th and $j$th trees.
-
-The tree kernel provides a pairwise similarity score between two trees. The
-higher the kernel score, the more similar the trees are to each other.
-Therefore, to have separability, we need trees simulated with the same value of
-$\alpha$ to have high kernel scores with each other, but low kernel scores with
-trees from different $\alpha$ values. We can visually check whether or not this
-is true by laying out the trees as points on a graph, in such a way that trees
-with high scores are close to each other, but trees with low scores are far
-apart. This is accomplished by performing a kernel principal components
-analysis (kPCA)~\autocite{scholkopf1998nonlinear} on the kernel matrix.
-Briefly, ordinary principal components analysis (PCA) finds a lower dimensional
-representation of points in a high dimensional space which preserves as much of
-the variation in the data as possible. kPCA performs the same task, but using
-the dot products of each pair of points as input, instead of the data points
-themselves. A two-dimensional kPCA projection of the simulated trees is shown
-in Figure~\ref{fig:pakpca}. The scenario just described - 500 samples from 1000
-infected nodes - is the central panel. To ensure that the method could be used
-in a variety of contexts, the same analysis was performed with 500 and 2000
-infected nodes, as well as 100 and 1000 sampled tips.
-
-The next step was to quantify how well trees simulated with different $\alpha$
-values could be distinguished from each other. As described
-previously~\autocite{poon2015phylodynamic}, this was done by assessing the
-accuracy of a support vector machine regression
-(SVR)~\autocite{smola1997support}. Briefly, an SVR operates by finding a
-hyperplane (in two dimensions, a line) such that the deviation of most of the
-data points from the line is less than some prescribed threshold. Points
-further away than this threshold are ignored in the model. I performed 1000
-replicate 2-fold cross-validations of an SVR predicting preferential attachment
-power on the simulated trees, using the \software{ksvm} function from
-\software{kernlab}~\autocite{karatzoglou2004kernlab}. That is, the SVR was
-trained on a random subset of 150 trees, and then used to predict $\alpha$ of
-the remaining 150 trees. The predictions were correlated against the true
-values of $\alpha$ to obtain an $R^2$, and this procedure was repeated 1000
-times with different subsets of trees.
-
-The cross-validation had the dual purpose of providing a means to selecting the
-optimal meta-parameters to the tree kernel (see
-subsection~\ref{subsubsec:treeshape}) - they are be those which provide the
-highest average $R^2$. To this end, the cross-validation was repeated for
-several values of $\lambda$ and $\sigma$ as shown in Figure~\ref{fig:kernelexpt}
-Each combination was evaluated with and without multiplying the tree kernel by
-the normalized lineages-through-time (nLTT) statistic, and was further repeated
-for the scenarios with differing numbers of infected and sampled nodes
-described above. To ensure that the tree kernel was the most appropriate
-similarity measure to use for ABC, we also computed the $R^2$ of $\alpha$
-against Sackin's index, a widely used tree balance statistic (see
-subsection~\ref{subsubsec:treeshape}), by the same cross-validation procedure.
-
-\subsection{Grid search}
-
-The previous set of simulations were intended to investigate which contact
-network parameters could and could not be inferred by examining transmission
-trees. However, they tell us nothing about the accuracy or precision we might
-expect when inferring those parameters numerically. As illustrated in
-Figure~\ref{fig:accprec}, the ideal situation is one where we are accurate and
-precise, and the worst situation is when we are precise but not accurate.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{acc-prec.pdf}
-  \caption{Illustration of accurate vs. precise kernel score estimates}
-  \label{fig:accprec}
-\end{figure}
-
-Figure~\ref{fig:gridsearch} shows a schematic of this experiment. A number of
-representative values, here denoted $k$, were chosen for the parameter of
-interest. In the case of preferential attachment power $\alpha$, I chose $k =
-8$ testing values, namely $0, 0.25, \ldots, 2.0$. For each of these values, ten
-networks on 5000 nodes were generated, an epidemic of 1000 nodes was simulated
-over each, and a transmission tree of 500 tips was sampled. The resulting $10
-\times k$ simulated transmission trees were referred to as the \defn{testing
-trees}. In addition, I chose a further $n \gg k$ training values spanning the
-range of the parameter. For $\alpha$, these were $0, 0.01, \ldots, 2.0$.
-Fifteen trees were simulated in the same manner for each of these values,
-referred to here as \defn{training trees}.
-
-For each of the testing trees, I computed the tree kernel with all of the
-training trees. This resulted in 15 kernel scores per training value. The
-training value with the highest median kernel score was used as a point
-estimate of the testing value. The kernel scores were then normalized to lie in
-[0, 1], and an interval was found which contained 95\% of the area under the
-curve and was of minimal width. This was used as a 95\% confidence interval for
-the parameter.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{gridsearch-expt.pdf}
-  \caption{Schematic of grid search simulation experiments.}
-  \label{fig:gridsearch}
-\end{figure}
-
-\subsection{Approximate Bayesian computation}
-
-Our final set of simulation experiments was to test the ABC-SMC algorithm on
-simulated data. For each replicate, we generated a network on 5000 nodes,
-simulated an epidemic over that network until 1000 nodes were infected, and
-sampled either 500 or all 1000 of those nodes to form a transmission tree.
-Priors were specified as follows: for the total number of nodes in the network,
-$\Uniform(1000, 10000)$; for the number of infected nodes, $\Uniform(500,
-2000)$; for the preferential attachment power, $\Uniform(0, 2)$. The mean
-degree of nodes in the network was fixed at either 4, 10, or 16. The
-algorithm was run with 1000 particles, and was stopped when the \gls{MCMC}
-acceptance probability dropped below 1.5\%.
-
-\subsection{Characterization of power-law exponent in Barab\'asi-Albert networks}
+\subsubsection*{Characterization of power-law exponent in Barab\'asi-Albert networks}
 
@@
 relationship between the \gls{BA} model parameters and \gls{gamma}. A network
-was simulated for each element of the Cartesian product of these parameters:
-\gls{N} = \sett{500, 600, \ldots, 15000}, \gls{m} = \sett{1, 2, \ldots, 8}, and
-\gls{alpha} = \sett{0.00, 0.01, \ldots, 2.00}. A power law distribution was
-fitted to the degree distribution of each simulated network using the
-\software{fit\_power\_law} function in \software{igraph} with the `R.mle'
-implementation. We fitted a \gls{GLM} with Gamma-distributed errors and a log
-link function to the observed distribution of \gls{gamma} values, with
-\gls{alpha}, \gls{m}, \gls{N}, and all possible interaction terms as
-predictors.
+was simulated for each combination of parameters listed in
+\cref{tab:gammaexpt}. A power law distribution was fitted to the degree
+distribution of each simulated network using the \software{fit\_power\_law}
+function in \software{igraph} with the `R.mle' implementation. We fitted a
+\gls{GLM} with Gamma-distributed errors and a log link function to the observed
+distribution of \gls{gamma} values, with \gls{alpha}, \gls{m}, \gls{N}, and all
+possible interaction terms as predictors.
+
+\begin{table}
+  \centering
+  \input{\tablepath/gamma-expt}
+  \caption[\gls{BA} parameters used as input \gls{GLM} predicting $\gamma$]
+  {
+    \gls{BA} model parameters used as input to \gls{GLM} predicting power law
+    exponent $\gamma$. One network was simulated with each combination of
+    parameters, and $\gamma$ was calculated for each network. A \gls{GLM} with
+    Gamma-distributed errors and a log link function was fit to the $\gamma$
+    values with all parameters and interaction terms as predictors.
+  }
+  \label{tab:gammaexpt}
+\end{table}
 
@@
 
-We applied kernel-\gls{ABC} to six \gls{HIV} datasets, summarized in
-\cref{tab:realdata}.
+Because the \gls{BA} model assumes a single connected contact network, it is
+most appropriate to apply to groups of individuals who are epidemiologically
+related. Therefore, we searched for published \gls{HIV} datasets which
+originated from existing clusters, either phylogenetically or geographically
+defined. In addition, we analysed an in-house dataset sampled from
+\gls{HIV}-positive individuals in British Columbia, Canada (the ``BC data'').
+The datasets are summarized in \cref{tab:realdata}.
 
-\begin{table}
+We downloaded all sequences associated with each published study from GenBank.
+For the \textcite{novitsky2014impact} data, each sequence was aligned pairwise
+to the HXB2 reference sequence (GenBank accession number HIVHXB2CG) and the
+hypervariable regions were clipped out with \software{BioPython} version
+1.66+~\autocite{cock2009biopython}. Sequences were multiply aligned using
+\software{MUSCLE} version 3.8.31 \autocite{edgar2004muscle}, and alignments
+were manually inspected with \software{Seaview} version 4.4.2
+\autocite{gouy2010seaview}. Phylogenies were constructed from the nucleotide
+alignments by approximate maximum likelihood using \software{FastTree2} version
+2.1.7 with the \gls{GTR} model. Transmission trees were estimated by rooting
+and time-scaling the phylogenies by root-to-tip regression, using a modified
+version of Path-O-Gen (distributed as part of
+BEAST~\autocite{drummond2007beast}) as described
+previously~\autocite{poon2015phylodynamic}.
+
+Three of the datasets \autocite[][and the BC data]{li2015hiv,novitsky2014impact}
+were initially much larger than the others, containing 1265, 1299, and 7923
+sequences respectively. To ensure that the analyses were comparable, we reduced
+these to a number of sequences similar to the smaller datasets. For the
+\citeauthor{li2015hiv} and BC datasets, we detected clusters of size 280 and
+399 respectively using a patristic distance cutoff of 0.02 as described
+previously~\autocite{poon2014impact}. Only sequences within these clusters were
+carried forward. For the \textcite{novitsky2014impact} data, no large clusters
+were detected using the same cutoff, so we analysed a subtree of size 180
+chosen arbitrarily.
+
+\begin{table}[ht]
   \centering

### 2016-03-31 doc/thesis/results.Rnw
Finished results! Yes!

--- 2016-03-29 doc/thesis/results.Rnw
@@
 
-% overview of experiment
-Our first simulation-based experiment was designed to determine which network
-model parameters were potentially estimable by kernel-ABC, and to find the best
-set of kernel meta-parameters to carry through to future experiments. For each
-of the four \gls{BA} model parameters, we simulated trees under three distinct
-values of that parameter while holding the others fixed, then cross-validated a
-kernel-\gls{SVM} classifier for that parameter.
-
 <<treestats, include=FALSE>>=
     params <- c("alpha", "I", "m", "N")
-    ptn <- paste0("../../simulations/kernel-", params, "/statistics/*")
-    files <- lapply(ptn, Sys.glob)
-    d <- lapply(files, collect.data)
-    d <- mapply(cbind, d, parameter=as.list(params), SIMPLIFY=FALSE)
-    lapply(d, setDT)
-    d[[1]] <- subset(d[[1]], m == 2)
-    setnames(d[[1]], "nsimnode", "I")
+
+    get.treestats <- function (param) {
+        stats <- paste0("../../simulations/kernel-", param, "/statistics/*")
+        d <- collect.data(stats)
+        trees <- paste0("../../simulations/kernel-", param, "/tree/*")
+        m <- collect.metadata(trees)
+        rownames(m) <- sub("../../simulations/", "", rownames(m))
+        d <- merge(d, m, by=0, suffixes=c("", ".1"))
+        if (param == "alpha") {
+            d <- subset(d, m == 2)
+            colnames(d)[colnames(d) == "nsimnode"] <- "I"
+        }
+        setDT(d)
+    }
+
+    alpha.stats <- get.treestats("alpha")
+    I.stats <- get.treestats("I")
+    m.stats <- get.treestats("m")
+    N.stats <- get.treestats("N")
+
+    alpha.sackin.cor <- alpha.stats[,cor.test(alpha, sackin, method="spearman")]
+    stopifnot(alpha.sackin.cor$p.value < 1e-5)
+
+    N.ratio.cor <- N.stats[,cor.test(int.tip.ratio, N)]
+    I.ratio.cor <- I.stats[,cor.test(int.tip.ratio, I)]
+    stopifnot(N.ratio.cor$p.value < 1e-5)
+    stopifnot(I.ratio.cor$p.value < 1e-5)
 @
 
-% trees look different
 Trees simulated under different values of \gls{alpha} are visibly quite
-distinctive (\cref{fig:patrees}). In particular, higher values of \gls{alpha}
+distinct (\cref{fig:alphatrees}). In particular, higher values of \gls{alpha}
 produce networks with a small number of highly connected nodes which, once
@@
 This results in a more unbalanced, ladder-like structure in the phylogeny,
-compared to networks with lower \gls{alpha} values. None of the other three
-parameters produced trees which were as easily distinguished from each other
-(\cref{fig:Itrees,fig:mtrees,fig:Ntrees,fig:Itrees}). However, higher $I$
-values produce trees which appear to have longer terminal branches,
+compared to networks with lower \gls{alpha} values. Sackin's index was
+significantly correlated with \gls{alpha} 
+    (Spearman's rho \Sexpr{round(alpha.sackin.cor$estimate, 2)},
+     $p < 10^{-5}$).
+None of the other three parameters produced trees which were as easily
+distinguished from each other
+(\cref{fig:Itrees,fig:mtrees,fig:Ntrees,fig:Itrees}). However, the ratio of
+internal to terminal branch lengths was positively correlated with $N$ and
+negatively correlated with $I$ (Spearman's rho
+    \Sexpr{round(N.ratio.cor$estimate, 2)} and
+    \Sexpr{round(I.ratio.cor$estimate, 2)},
+    both $p < 1e-5$).
 
@@
 
-The fact that Sackin's index is informative of \gls{alpha} was
-unsurprising, given the clear differences in tree balance observed under
-different \gls{alpha} values.
-
 \begin{figure}[ht]
@@
   }
-  \label{fig:patrees}
+  \label{fig:alphatrees}
 \end{figure}

### 2016-04-05 doc/paper/abstract.tex
MBE Submission

Models of the spread of disease in a population often make the simplifying
assumption that the population is homogeneously mixed, or is divided into
homogeneously mixed compartments. However, human populations have complex
structures formed by social contacts, which can have a significant influence on
the rate of epidemic spread. Contact network models capture this structure by
explicitly representing each contact which could possibly lead to a
transmission. We developed a method based on kernel approximate Bayesian
computation (kernel-ABC) for estimating structural parameters of the contact
network underlying an observed viral phylogeny. The method combines adaptive
sequential Monte Carlo for ABC, Gillespie simulation for propagating epidemics
though networks, and a kernel-based tree similarity score. We used the method
to fit the Barab\'{a}si-Albert network model to simulated transmission trees,
and also applied it to viral phylogenies estimated from five published HIV
sequence datasets. On simulated data, we found that the preferential attachment
power and the number of infected nodes in the network can often be accurately
estimated. On the other hand, the mean degree of the network, as well as the
total number of nodes, were not estimable with kernel-ABC. We observed
substantial heterogeneity in the parameter estimates on real datasets, with
point estimates for the preferential attachment power ranging from 0.2 to 1.1.
These results underscore the importance of considering contact structures when
performing phylodynamic inference. Our method offers the potential to
quantitatively investigate the network structure underlying viral epidemics,
using models appropriate to the context being studied.

### 2016-04-05 doc/paper/paper.tex
MBE Submission

@@
-\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
+\documentclass[nogrid]{MBE}
+\usepackage{url}
 
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+% Mol Biol Evol stuff
+\jshort{mst}
+\volname{}
+\jvolume{0}
+\jvol{}
+\jissue{0}
+\pubyear{2016}
+\mstype{Article}
+\artid{???}
+\access{???}
+\begin{document}
 
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
-\usepackage[osf]{garamondx}
-\usepackage[garamondx,cmbraces]{newtxmath}
-\usepackage{marginnote}
-\usepackage[top=2cm, bottom=2cm, outer=3cm, inner=3cm, heightrounded, 
-            marginparwidth=2cm, marginparsep=0.5cm]{geometry}
-\usepackage{setspace}
-\usepackage[backend=biber, style=authoryear, uniquelist=minyear, uniquename=false]{biblatex}
-\usepackage[acronym,nonumberlist]{glossaries}
-\usepackage{multirow}
-\usepackage{cleveref}
-\usepackage{array}
-
-\newcommand{\defn}[1]{\textit{#1}}
-\newcommand{\software}[1]{\textit{#1}}
-
-\newcommand{\set}[1]{\left\lbrace#1\right\rbrace}
-\newcommand{\sett}[1]{\{#1\}}
-\renewcommand{\emptyset}{\varnothing}
-\renewcommand{\vec}[1]{\mathbf{#1}}
-\renewcommand{\star}[1]{#1^*}
-\renewcommand{\d}{\mathrm{d}\,}
-
-\DeclareMathOperator*{\argmin}{arg\,min}
-\DeclareMathOperator*{\argmax}{arg\,max}
-
-\DeclareMathOperator{\Exponential}{Exponential}
-\DeclareMathOperator{\Uniform}{Uniform}
-\DeclareMathOperator{\E}{E}
-\DeclareMathOperator{\Var}{Var}
-
-\DeclareMathOperator{\inc}{in}
-\DeclareMathOperator{\out}{out}
-
-\newcommand{\N}{\mathcal{N}}
-\renewcommand{\L}{\mathcal{L}}
-
-\newacronym{BA}{BA}{Barab\'asi-Albert}
-\newacronym{SI}{SI}{susceptible-infected}
-\newacronym{kSVM}{kSVM}{kernel support vector machine}
-\newacronym{nLTT}{nLTT}{normalized lineages-through-time}
-\newacronym{SMC}{SMC}{sequential Monte-Carlo}
-\newacronym{ABC}{ABC}{approximate Bayesian computation}
-\newacronym{MSM}{MSM}{men who have sex with men}
-\newacronym{HPD}{HPD}{highest posterior density}
-\newacronym{GTR}{GTR}{general time-reversible}
-\newacronym{MCMC}{MCMC}{Markov chain Monte Carlo}
-\newacronym{MAP}{MAP}{maximum \textit{a posteriori}}
-\newacronym{IDU}{IDU}{injection drug user}
-
-\graphicspath{{../figures/}}
-\newcommand{\tablepath}{../tables}
-\def\figwidth{0.8\textwidth}
-
-\title{Phylodynamic inference of contact network parameters through approximate
-Bayesian computation}
-\author{Rosemary M. McCloskey \and Richard H. Liang \and Art F.Y. Poon}
-
-\addbibresource{papers.bib}
-
-\frenchspacing
-\onehalfspacing
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\begin{document}
+\title[Contact networks and ABC]{Phylodynamic inference of contact network
+parameters through approximate Bayesian computation}
+\author[McCloskey, Liang, and Poon]{Rosemary M. McCloskey$^{*,1}$, 
+    Richard H. Liang$^1$, and Art F.Y. Poon$^{1,2}$}
+\address{$^1$BC Centre for Excellence in HIV/AIDS, 608-1081 Burrard Street, Vancouver, Canada \\
+$^2$Department of Medicine, University of British Columbia, Vancouver, Canada}
+\history{Submitted 5 April 2016}
+\coresp{E-mail: rmccloskey@cfenet.ubc.ca.}
+\editor{}
+\abstract{\input{abstract}}
+\keyword{Phylogenetics, phylodynamics, approximate Bayesian computation,
+contact network, transmission tree, human immunodeficiency virus.}
 
@@
 
-\section*{Background}
+\section{Introduction}
 
@@
 mode of transmission of the disease. For airborne pathogens, physical proximity
-may be sufficient, but for sexually transmitted diseases, either sexual or
-blood-to-blood contact is required. The set of links between individuals along
-which transmission can occur is called the contact
-network~\autocite{klovdahl1985social, morris1993epidemiology}. The structure of
-the contact network underlying an epidemic can profoundly impact the speed and
-pattern of the epidemic's expansion. Network structure can influence the
-prevalence trajectory~\autocite{o2010contact} and epidemic
-threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates
-of quantities such as effective population
-size~\autocite{goodreau2006assessing}. From a public health perspective,
-contact networks have been explored as tools for curtailing epidemic spread, by
-way of interventions targeted to well-connected
-nodes~\autocite{wang2015targeting}. True contact networks are a challenging
-type of data to collect, requiring extensive epidemiological
-investigation~\autocite{welch2011statistical}.
+may be sufficient, while for sexually transmitted diseases, \newpage\noindent %hack
+sexual or in some cases blood-to-blood contact is required. The population
+together with the set of links between individuals along which transmission can
+occur is called the contact network~\citep{klovdahl1985social,
+morris1993epidemiology}. The structure of the contact network underlying an
+epidemic can profoundly impact the speed and pattern of the epidemic's
+expansion. Network structure can influence the prevalence
+trajectory~\citep{o2011contact} and epidemic
+threshold~\citep{barthelemy2005dynamical}, in turn affecting the estimates of
+quantities such as effective population size~\citep{goodreau2006assessing}.
+From a public health perspective, contact networks have been explored as tools
+for curtailing epidemic spread, by way of interventions targeted to
+well-connected nodes~\citep{wang2015targeting}. True contact networks are a
+challenging type of data to collect, requiring extensive epidemiological
+investigation~\citep{welch2011statistical}.
 
@@
 of RNA viruses, epidemiological processes impact the course of viral evolution,
-thereby shaping the intra-host viral
-phylogeny~\autocite{drummond2003measurably}. The term ``phylodynamics'' was
+thereby shaping the inter-host viral
+phylogeny~\citep{drummond2003measurably}. The term ``phylodynamics'' was
 coined to describe this interaction, as well as the growing family of inference
 methods to estimate epidemiological parameters from viral
-phylogenies~\autocite{grenfell2004unifying}. These methods have revealed
+phylogenies~\citep{grenfell2004unifying}. These methods have revealed
 diverse properties of local viral outbreaks, from basic reproductive
-number~\autocite{stadler2011estimating}, to the degree of
-clustering~\autocite{hughes2009molecular}, to the elevated transmission risk
-during acute infection~\autocite{volz2012simple}. On the other hand, although
+number~\citep{stadler2012estimating}, to the degree of
+clustering~\citep{hughes2009molecular}, to the elevated transmission risk
+during acute infection~\citep{volz2012simple}. On the other hand, although
 sophisticated methods have been developed for fitting complex population
-genetic models to phylogenies~\autocite{rasmussen2014phylodynamic}, inference
+genetic models to phylogenies~\citep{rasmussen2014phylodynamic}, inference
 of structural network parameters has to date been limited. However, it has been
 shown that network structure has a tangible impact on phylogeny
-shape~\autocite{leventhal2012inferring, colijn2014phylogenetic,
-goodreau2006assessing, robinson2013dynamics}, suggesting that such statistical
-inference might be possible~\autocite{welch2011statistical}.
+shape~\citep{leventhal2012inferring, colijn2014phylogenetic,
+goodreau2006assessing, robinson2013dynamics, villandre2016assessment},
+suggesting that such statistical inference might be
+possible~\citep{welch2011statistical}.
 
-Survey-based studies of sexual networks~\autocite{liljeros2001web,
-schneeberger2004scale} have found that sexual contact networks are best
-described by a preferential attachment model \autocite[although there has been
-some disagreement, see][]{jones2003assessment}. Under these models, nodes with
-a high number of contacts attract new connections at an elevated rate. Networks
-produced by preferential attachment have a power-law degree distribution,
-meaning that the number of nodes of degree $k$ is proportional to $k^\gamma$
-for some constant $\gamma$. These networks are also referred to as
-``scale-free''. The first contact network model incorporating preferential
-attachment was introduced by \textcite{barabasi1999emergence}, and is now
-referred to as the \gls{BA} model. Under this model, networks are formed by
-iteratively adding nodes with $m$ new edges each. These new edges are joined to
-existing nodes of degree $k$ with probability proportional to $k^\alpha$, so
-that nodes of high degree tend to attract more connections (in the original
-paper, only $\alpha = 1$ was investigated).
+Survey-based studies of sexual networks~\citep{liljeros2001web,
+schneeberger2004scale, colgate1989risk} have found that these networks tend to
+have a degree distribution which follows a power law \citep[although there has
+been some disagreement, see][]{handcock2004likelihood}. That is, the number of
+nodes of degree $k$ is proportional to $k^{-\gamma}$ for some constant
+$\gamma$. These networks are also referred to as ``scale-free''. One process by
+which scale-free networks can be generated is preferential attachment, where
+nodes with a high number of contacts attract new connections at an elevated
+rate. The first contact network model incorporating preferential attachment was
+introduced by \citet{barabasi1999emergence}, and is now referred to as the
+Barab\'asi-Albert (BA) model. Under this model, networks are formed by
+iteratively adding nodes with $m$ new edges each. In the most commonly studied
+formulation, these new edges are joined to existing nodes of degree $k$ with
+probability proportional to $k$, so that nodes of high degree tend to attract
+more connections. \citeauthor{barabasi1999emergence} suggested an extension
+where the probability of attaching to a node of degree $k$ is $k^\alpha$ for
+some non-negative constant $\alpha$, and we use this extension here.
 
 Previous work offers precedent for the possibility of statistical inference of
-structural network parameters. \textcite{britton2002bayesian} develop a
-Bayesian approach to estimate the edge density in an Erd\H{o}s-R\'enyi
-network~\autocite{erdos1960evolution} given observed infection dates, and
+structural network parameters. \citet{britton2002bayesian} develop a Bayesian
+approach to estimate the edge density in an Erd\H{o}s-R\'enyi
+network~\citep{erdos1960evolution} given observed infection dates, and
 optionally recovery dates. Their approach was later extended by
-\textcite{groendyke2011bayesian} and applied to a much larger data set of 188
-individuals. \textcite{brown2011transmission} analysed the degree distribution
+\citet{groendyke2011bayesian} and applied to a much larger data set of 188
+individuals. \citet{volz2007susceptible, volz2008sir} developed differential
+equations describing the spread of a susceptible-infected (SI) epidemic on
+static and dynamic contact networks with several degree distributions, which
+could in principle be used for inference if observed incidence trajectories
+were available. \citet{brown2011transmission} analysed the degree distribution
 of an approximate transmission network, estimated based on genetic similarity
-and estimated times of infection, relating 60\% of HIV infected \gls{MSM} in
-the United Kingdom. The transmission network is a subgraph of the contact
-network which includes only those edges which have already led to a new
-infection. The authors found that a Waring distribution, which is produced by a
-different preferential attachment model, was a good fit to their estimated
-network.
+and estimated times of infection, relating 60\% of HIV-infected men who have
+sex with men (MSM) in the United Kingdom. The transmission network is a
+subgraph of the contact network which includes only those edges which have
+already led to a new infection. The authors found that a Waring distribution,
+which is produced by a more sophisticated preferential attachment model, was a
+good fit to their estimated network.
 
@@
 calculation of a normalizing constant, Bayesian inference is often performed
-using \gls{MCMC}. Rather than calculating explicit likeilihoods, \gls{MCMC}
-uses likelihood \emph{ratios} in which the normalizing constant cancel out.
-Unfortunately, it is generally difficult to explicitly calculate the likelihood
-of an observed transmission tree under a contact network model, even up to a
-normalizing constant. To do so, it would be necessary to integrate over all
-possible networks, and also over all possible labellings of the internal nodes
-of the transmission tree. While it is not known (to us) whether such
-integration is tractable, a simpler alternative is offered by likelihood-free
-methods, namely \gls{ABC}. \gls{ABC} leverages the fact that, although
-calculating the likelihood may be impossible, generating simulated datasets
-according to a model is often straightforward. If our model fits the data well,
-the simulated data it produces should be similar to the observed data. More
-formally, if $D$ is the observed data, the posterior distribution $f(\theta
-\mid D)$ on model parameters $\theta$ is replaced as the target of statistical
-inference by $f(\theta \mid \rho(\hat{D}, D) < \varepsilon)$, where $\rho$ is a
-distance function, $\hat{D}$ is a simulated dataset according to $\theta$, and
-$\varepsilon$ is a small tolerance~\autocite{sunnaker2013approximate}. In the
+using Markov chain Monte Carlo (MCMC), which uses likelihood \emph{ratios} in
+which the normalizing constants cancel out. Unfortunately, it is generally
+difficult to explicitly calculate the likelihood of an observed transmission
+tree under a contact network model, even up to a normalizing constant. To do
+so, it would be necessary to integrate over all possible networks, and also
+over all possible labellings of the internal nodes of the transmission tree.
+While it is not known (to us) whether such integration is tractable, a simpler
+alternative is offered by likelihood-free methods, namely approximate Bayesian
+computation (ABC). ABC leverages the fact that, although calculating the
+likelihood may be impractical, generating simulated datasets according to a
+model is often straightforward. If our model fits the data well, the simulated
+data it produces should be similar to the observed data. More formally, if $D$
+is the observed data, the posterior distribution $f(\theta \mid D)$ on model
+parameters $\theta$ is replaced as the target of statistical inference by
+$f(\theta \mid \rho(\hat{D}, D) < \varepsilon)$, where $\rho$ is a distance
+function, $\hat{D}$ is a simulated dataset according to $\theta$, and
+$\varepsilon$ is a small tolerance~\citep{sunnaaker2013approximate}. In the
 specific case when $\rho$ is a kernel function, the approach is known as
-kernel-\gls{ABC}~\autocite{nakagome2013kernel, poon2015phylodynamic}.
+kernel-ABC~\citep{nakagome2013kernel, poon2015phylodynamic}.
 
-Here, we apply kernel-\gls{ABC} to the problem of statistical inference of
-contact network parameters from an estimated transmission tree, using the tree
-kernel developed by \textcite{poon2013mapping}. We then estimate the parameters
-of the \gls{BA} model on a variety of simulated and real data sets. Our results
-show that the attachment power parameter $\alpha$ can be inferred with
-reasonable accuracy, and can vary considerably between epidemics from different
-settings.
+Here, we develop a method using kernel-ABC to estimate the parameters of
+contact network models from observed phylogenetic data. The distance function
+we use is the tree kernel developed by \citet{poon2013mapping}, which computes
+a weighted dot product of the trees' representations in the space of all
+possible subset trees. We apply the method to investigate the parameters of the
+BA network model on a variety of simulated and real datasets. Our results show
+that some network parameters can be inferred with reasonable accuracy, while
+others have a minimal detectable impact on tree shape and therefore cannot be
+estimated accurately. We also find that these parameters can vary considerably
+between real epidemics from different settings.
 
-\section*{Methods}
+\section{New Approaches}
 
-We implemented a Gillespie simulation algorithm~\autocite{gillespie1976general}
-for simulating epidemics and transmission trees over static contact networks,
-in the same fashion as several previous studies
-\autocite[\textit{e.g.}][]{o2010contact, robinson2013dynamics,
-leventhal2012inferring, groendyke2011bayesian, goodreau2006assessing}. To check
-that our implementation was correct, we reproduced Figure 1A of
-\textcite{leventhal2012inferring} (our \cref{fig:sf1}), which plots the
-unbalancedness of transmission trees simulated over four network models at
-various levels of pathogen transmissibility. Our program is freely available at
-\url{https://github.com/rmcclosk/netabc}.
+We have developed a kernel-ABC-based method to perform statistical inference of
+contact network parameters from a transmission tree estimated from an observed
+viral phylogeny. We implemented the adaptive sequential Monte Carlo (SMC)
+algorithm for ABC developed by \citet{del2012adaptive}. The SMC algorithm keeps
+track of a population of parameter ``particles'', which are initially sampled
+from the parameters' joint prior distribution. Several datasets are simulated
+under the model of interest for each of the particles. In this case, the
+datasets are transmission trees, which are generated by a two-step process.
+First, a contact network is simulated according to the network model being fit.
+Second, a transmission tree is simulated over that network with a Gillespie
+simulation algorithm~\citep{gillespie1976general}, in the same fashion as
+several previous studies \citep[\textit{e.g.}][]{robinson2013dynamics,
+leventhal2012inferring}. The particles are weighted according to the similarity
+between their associated simulated trees and the observed tree. To quantify
+this similarity, we used the tree kernel developed by \citet{poon2013mapping}.
+Particles are iteratively perturbed by applying a Metropolis-Hastings kernel
+and, if the move is accepted, simulating new datasets under the new parameters.
+When a particle's weight drops to zero, because its simulated trees are too
+dissimilar to the observed tree, the particle is dropped from the population,
+and eventually replaced by a resampled particle with a higher weight. As the
+algorithm progresses, the population converges to a Monte Carlo approximation
+of the ABC target distribution, which is assumed to approximate the desired
+posterior~\citep{del2012adaptive, sunnaaker2013approximate}.
 
-We chose to study the \gls{BA} network model~\autocite{barabasi1999emergence}.
-In addition to $m$ and $\alpha$, we investigated the parameters $N$, which
-denotes the total number of nodes in the network, and $I$, which is the number
-of infected nodes at which to stop the simulation and sample the transmission
-tree. Nodes in our networks followed simple \gls{SI} dynamics, meaning that
-they became infected at a rate proportional to their numbers of infected
-neighbours, and never recovered. For all analyses, the transmission trees'
-branch lengths were scaled by dividing by their mean. We used the 
-\software{igraph} library's implementation of the \gls{BA}
-model~\autocite{csardi2006igraph} to generate the graphs. The analyses were run
-on Westgrid (\url{https://www.westgrid.ca/}) and a local computer cluster.
+A computer program implementing our method is freely available at
+\url{https://github.com/rmcclosk/netabc} (last accessed April 3, 2016).
 
-\subsection*{Kernel classifiers}
+\section{Results}
 
-We used the phylogenetic kernel developed by \textcite{poon2013mapping} to test
-whether the parameters of the \gls{BA} model had an effect on tree shape. We
-simulated 100 networks under each of three different values of $\alpha$: 0.5,
-1.0, and 1.5 (300 networks total). The other parameters were fixed to the
-following values: $N$ = 5000, $I$ = 1000, and $m$ = 2. A transmission tree with
-500 tips was simulated over each network (300 transmission trees total). The
-300 trees were compared pairwise with the tree kernel to form a $300 \times
-300$ kernel matrix. The kernel meta-parameters $\lambda$ (the ``decay
-factor''), and $\sigma$ (the ``radial basis function
-variance'')~\autocite[see][]{poon2013mapping}, were set to 0.3 and 4
-respectively. We constructed a \gls{kSVM} classifier for $\alpha$ using the
-\software{kernlab} package~\autocite{karatzoglou2004kernlab}, and evaluated its
-accuracy with 1000 two-fold cross-validations.
+\subsection{Kernel classifiers}
 
-Three similar experiments were performed for the other \gls{BA} model
-parameters (one experiment per parameter). $m$ was varied between 2, 3, and 4;
-$I$ between 500, 1000, and 2000; and $N$ between 3000, 5000, and 8000. The
-parameters not being tested were fixed at the values $N$ = 5000, $I$ = 1000,
-$m$ = 2, and $\alpha$ = 1. Thus, we performed a total of four kSVM
-cross-validations, one for each of the \gls{BA} model parameters $\alpha$, $I$,
-$m$, and $N$.
+We investigated the parameters of the BA network
+model~\citep{barabasi1999emergence}. In addition to $m$ and $\alpha$ (see
+Introduction), we considered $N$, which denotes the total number of nodes in
+the network, and $I$, which is the number of infected nodes at which to stop
+the simulation and sample the transmission tree. To examine the effect of these
+parameters on tree shape, we simulated transmission trees under different
+parameter values, calculated pairwise tree kernel scores between them, and
+attempted to classify the trees using a kernel support vector machine (kSVM).
+We also tested classifiers based on Sackin's index~\citep{shao1990tree} and the
+normalized lineages-through-time (nLTT)
+statistic~\citep{janzen2015approximate}. Accuracy of the kSVMs varied based on
+the parameter being tested (Figure~\ref{fig:rsquared}, left). Classifiers based
+on two other tree statistics, the nLTT and Sackin's index, generally exhibited
+worse performance than the tree kernel, although the magnitude of the disparity
+varied between the parameters (Figure~\ref{fig:rsquared}, centre and right).
+The results were largely robust to variations in the tree kernel
+meta-parameters $\lambda$ and $\sigma$ (Figures~S3-S6).
 
-We repeated these four cross-validations with different values of $\lambda$
-(0.2, 0.3, and 0.4) and $\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$), as well
-as on trees with differing numbers of tips (100, 500, and 1000) and in
-epidemics of differing size $I$ (500, 1000, and 2000). When evaluating the
-classifier for $I$, we did not consider trees with 1000 tips, because one of
-the tested $I$ values was 500, and the number of tips cannot be larger than
-$I$.
+When classifying $\alpha$, the kernel-SVM classifier had an average $R^2$ of 
+    0.92,
+compared to 
+    0.56
+for the nLTT-based SVM, and
+    0.75
+for the linear regression against Sackin's index. There was little variation
+about the mean for different tree and epidemic sizes. No classifier could
+accurately identify the $m$ parameter in any epidemic scenario, with average
+$R^2$ values of 
+  0.12 for kSVM,
+  0.01 for the nLTT, and
+  0.06
+for Sackin's index. Again, there was little variation in accuracy between
+epidemic scenarios, although the accuracy of the kSVM was slightly higher
+on 1000-tip trees.
 
-For each of the four parameters, we also tested univariate classifier based on
-Sackin's index~\autocite{shao1990tree} and an ordinary SVM based on the
-\gls{nLTT} statistic~\autocite{janzen2015approximate}.
+The accuracy of classifiers for $I$ varied significantly with the number of
+tips in the tree. For 100-tip trees, the average $R^2$ values were
+  0.7,
+  0.55, and
+  0.02
+for the tree kernel, nLTT, and Sackin's index respectively. For 500-tip
+trees, the values increased to
+  0.93,
+  0.83, and
+  0.07.
+Finally, the performance of classifiers for $N$ depended heavily on the
+epidemic scenario. The $R^2$ of the kSVM classifier ranged from
+  0.08
+for the smallest epidemic and smallest sample size, to
+  0.82
+for the largest. Likewise, $R^2$ for the nLTT-based SVM ranged from 
+  0.01
+to
+  0.54.
+Sackin's index did not accurately classify $N$ in any scenario, with an average
+$R^2$ of
+  0.03
+and little variation between scenarios.
 
-\subsection*{ABC simulations}
-
-We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
-\textcite{del2012adaptive}. To check that our implementation was correct, we
-applied it to the same mixture of Gaussians used by
-\citeauthor{del2012adaptive} to demonstrate their method (originally used
-by~\textcite{sisson2007sequential}). We were able to obtain a close
-approximation to the function (see \cref{fig:smctest}), and attained the
-stopping condition used by the authors in a comparable number of steps.
-
-We simulated three transmission trees, each with 500 tips, under every element
-of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
-\sett{1000, 2000}, $m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0.0, 0.5, 1,
-1.5}. This produced a total of 24 parameter combinations $\times$ three trees
-per combination = 72 trees total. The adaptive \gls{ABC} algorithm was applied
-to each tree with these priors: $m \sim$ Uniform(1, 5), $\alpha \sim$
-Uniform(0, 2), and $(N, I)$ jointly uniform on the triangular region \{$500
-\leq N \leq 15000$, $500 \leq I \leq 5000$, $I \leq N$\}. Following
-\textcite{del2012adaptive} and \textcite{beaumont2009adaptive}, all proposals
-were Gaussian, with variance equal to twice the empirical variance of the
-particles. The algorithm was run with 1000 particles, 5 simulated datasets per
-particle, and the ``quality'' parameter controlling the decay rate of the
-tolerance $\varepsilon$ set to 0.95. We used the same stopping criterion as
-\citeauthor{del2012adaptive}, namely when the \gls{MCMC} acceptance rate
-dropped below 1.5\%. Point estimates for the parameters were obtained by taking
-the highest point of an estimated kernel density on the final set of particles,
-using the \software{density} function with the default parameters in
-\software{R}. \gls{HPD} intervals were calculated with the
-\software{HPDinterval} function from the \software{R} package
-\software{coda}~\autocite{plummer2006coda}.
-
-Two further analyses were performed to address potential sources of error. To
-evaluate the effect of model misspecification in the case of heterogeneity
-among nodes, we generated a network where half the nodes were attached with
-power $\alpha$ = 0.5, and the other half with power $\alpha$ = 1.5. The other
-parameters for this network were $N$ = 5000, $I$ = 1000, and $m$ = 2. To
-investigate the effects of potential sampling bias, we simulated a transmission
-tree where the tips were sampled in a peer-driven fashion, rather than at
-random. That is, the probability to sample a node was twice as high if any of
-that node's network peers had already been sampled. The parameters of this
-network were $N$ = 5000, $I$ = 2000, $m$ = 2, and $\alpha$ = 0.5.
-
-\subsection*{Investigation of published data}
-
-We applied our kernel-ABC method to several published HIV datasets. Because the
-\gls{BA} model generates networks with a single connected component, we
-specifically searched for datasets which originated from existing clusters,
-either phylogenetically or geographically defined. Characteristics of the
-datasets we investigated are given in \cref{tab:data}.
-
-\begin{table}[ht]
+\begin{figure*}[ht]
   \centering
-  \input{\tablepath/realdata}
-  \caption{Characteristics of published datasets investigated with kernel-ABC.
-  Acronyms: MSM, men who have sex with men; IDU, injection drug users.}
-  \label{tab:data}
-\end{table}
-
-We downloaded all sequences associated with each study from GenBank. For the
-\textcite{novitsky2014impact} data, each sequence was aligned pairwise to the
-HXB2 reference sequence (Genbank accession number HIVHXB2CG) and the
-hypervariable regions were clipped out with BioPython version
-1.66+~\autocite{cock2009biopython}. Sequences were multiply aligned using
-\software{MUSCLE} version 3.8.31 \autocite{edgar2004muscle}, and alignments
-were manually inspected with \software{Seaview} version 4.4.2
-\autocite{gouy2010seaview}. Phylogenies were constructed from the nucleotide
-alignments by approximate maximum likelihood using \software{FastTree2} version
-2.1.7 with the \gls{GTR} model. Transmission trees were estimated by rooting
-and time-scaling the phylogenies by root-to-tip regression, using a modified
-version of Path-O-Gen (distributed as part of
-BEAST~\autocite{drummond2007beast}) as described
-previously~\autocite{poon2015phylodynamic}.
-
-Two of the datasets \autocite{li2015hiv,novitsky2014impact} were initially much
-larger than the others, containing 1265 and 1299 sequences respectively. To
-ensure that the analyses were comparable, we reduced these to a number of
-sequences similar to the smaller datasets. For the \textcite{li2015hiv} data,
-we detected a cluster of size 280 using a patristic distance cutoff of 0.02 as
-described previously~\autocite{poon2014impact}. Only sequences within this
-cluster were carried forward. For the \textcite{novitsky2014impact} data, no
-large clusters were detected using the same cutoff, so we analysed a subtree of
-size 180 chosen arbitrarily.
-
-\section*{Results}
-
-\subsection*{Kernel classifiers}
-
-Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
-tested~\cref{fig:rsquared}. Classifiers based on two other tree statistics, the
-\gls{nLTT} and Sackin's index, generally exhibited worse performance than the
-tree kernel, although the magnitude of the disparity varied between the
-parameters. \Cref{fig:rsquared} shows the cross-validation accuracy with
-$\lambda$ = 0.3, $\sigma$ = 4, $I$ = 1000, and trees of size 500. The results
-were largely robust to variations in $\lambda$ and $\sigma$, although accuracy
-varied between different epidemic and sampling scenarios
-(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}).
-
-The kSVM classifier for $\alpha$ had an average $R^2$ of
-    0.92 
-(\cref{fig:alphacrossv}). The $R^2$ of the for the SVM classifer for $\alpha$
-using the \gls{nLTT} varied between
-0.49
-and
-  0.61.
-The $R^2$ of the linear regression against Sackin's index varied between
-  0.73
-and
-  0.76.
-The $I$ parameter was also classified accurately by the tree kernel
-(\cref{fig:Icrossv}), with an average
-$R^2$ of 
-  0.93 
-for 500-tip trees or 
-  0.7 
-for 100-tip trees. The \gls{nLTT} performed almost as well on this parameter as
-the tree kernel, with $R^2$ values of
-  0.83
-for 500-tip trees or
-  0.55
-for 100-tip trees. Sackin's index was an extremely poor classifier for this
-parameter, with both $R^2$ values below
-  0.08.
-
-The $m$ parameter was much harder to classify (\cref{fig:mcrossv}), with $R^2$
-for the kSVM varying between 
-  0.004
-and
-  0.36
-depending on $I$ and the tree size. The \gls{nLTT} was also universally poor,
-with $R^2$ below
-  0.02
-for all scenarios. Sackin's index fared slightly better than the \gls{nLTT},
-especially on trees with a higher number of tips, with $R^2$ between
-  0
-and
-  0.18.
-Finally, the accuracy of the kSVM classifier for $N$ varied widely
-(\cref{fig:Ncrossv}), from 
-  0.08
-to
-  0.82.
-The \gls{nLTT} performed well in classifying $N$ when the epidemic was large,
-improving upon the tree kernel in the scenarios with $I$ = 2000 and 100- or
-500-tip transmission trees ($R^2$ = 
-  0.31
-and
-  0.53),
-although it performed poorly when $I$ was 500 or 1000 (average $R^2$ =
-  0.07).
-Sackin's index could not classify $N$, with all $R^2$ below
-  0.11.
-Based on inspection of the cross-validation results, we chose to use the
-meta-parameters $\lambda = 0.3$ and $\sigma = 4$ in further analyses, and not
-to use Sackin's index or the \gls{nLTT}.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
+  \includegraphics[width=\textwidth]{kernel-rsquared.eps}
+  \vspace{6pt}
   \caption{
       Cross-validation accuracy of kernel-SVM classifier (left), SVM classifier
-      using \gls{nLTT} (centre), and linear regression using Sackin's index
-      (right) for \gls{BA} model parameters. Kernel meta-parameters were set to
+      using nLTT (centre), and linear regression using Sackin's index
+      (right) for BA model parameters. Kernel meta-parameters were set to
       $\lambda = 0.3$ and $\sigma = 4$. Each point was calculated based on 300
@@
       the parameter being tested. Vertical lines are empirical 95\% confidence
-      intervals based on 1000 2-fold cross-validations.
+      intervals based on 1000 two-fold cross-validations.
   }
   \label{fig:rsquared}
-\end{figure}
+\end{figure*}
 
-\subsection*{ABC simulations}
+\subsection{ABC simulations}
 
-Of the four parameters, $\alpha$ was the most accurately estimated, with a
+Figure~\ref{fig:abcpt} shows maximum \textit{a posteriori} (MAP) point
+estimates of the BA model parameters obtained with kernel-ABC on simulated
+data. The estimates shown correspond only to the simulations where the $m$
+parameter was set to 2, however the results for $m = 3$ and $m = 4$ were
+similar (Figures S7 and S8). Average boundaries of 95\% HPD intervals are given
+in Table~\ref{tab:abchpd}.
+
+The accuracy of the parameter estimates obtained with kernel-ABC
+paralleled the results from the kSVM classifier. Of the four parameters,
+$\alpha$ was the most accurately estimated, with point estimates having a
 median [IQR] absolute error of 
     0.11 
-    [0.05-0.18].
-\Cref{fig:abcpt} shows point estimates for all simulations with $m = 2$. The
-results $m = 3$ and $m = 4$ were similar (\cref{fig:abcptm3,fig:abcptm4}).
-Average boundaries of 95\% \gls{HPD} intervals are given in \cref{tab:abchpd}.
-The accuracy of the estimates was not significantly different between values of
-$m$ or $I$ (both one-way ANOVA,
-    $p$ = 0.1
-and 
-    0.25),
-although the errors when the true value of $\alpha$ was zero were
-significantly greater than the other values 
-    (Wilcoxon rank-sum test, $p$ = \ensuremath{6.41\times 10^{-4}}).
-The error in the estimated value of $I$ was
-    305.66 
-    [107.76-606.59].
-Errors were significantly higher for $\alpha \geq 1$
-    (Wilcoxon rank-sum test, $p$ = \ensuremath{6.12\times 10^{-4}})
-and for $I$ = 2000
-    ($p$ = \ensuremath{1.58\times 10^{-6}}),
-but not for any values of $m$
-    (one-way ANOVA, $p$ = 0.33).
-The $m$ parameter was estimated correctly in
+    [0.05 - 
+    0.18].
+The errors when the true value of $\alpha$ was zero were significantly greater
+than those for the other values 
+    (Wilcoxon rank-sum test, $p$ = $6\!\times\!10^{-4}$),
+but did not vary across the true values of the other parameters (one-way
+ANOVA). Estimates for $I$ were also relatively accurate, with point estimate
+errors of
+    306 
+    [108 - 
+    607].
+These errors were significantly higher when the true value of $\alpha$ was
+at least 1
+    (Wilcoxon rank-sum test, $p$ = $6\!\times\!10^{-4}$)
+and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
+did not affect the estimates of $I$ (one-way ANOVA).
+
+The $m$ parameter was estimated correctly in only
     37 \%
-of simulations, with an error of one in
-    40 \%
-and of two or more in 
-    22 \%
-(the only possible $m$ values were 1, 2, 3, 4, or 5). The true values of $m$
-and $I$ did not significantly affect the error
-    (one-way ANOVA, $p$ = 0.5 and
-                          0.68).
-Finally, the total number of nodes $N$ was consistently over-estimated by about
-a factor of two
-    (error \ensuremath{6.59\times 10^{3}} 
-    [\ensuremath{4.21\times 10^{3}}-\ensuremath{8.28\times 10^{3}}]).
-No other parameters influenced the accuracy of the $N$ estimates 
-    (one-way ANOVA, $p \geq$ 0.72).
+of simulations. Oddly, the error in the estimated $m$ was higher for integer
+values of $\alpha$ (\text{i.e.}~0 and 1) than non-integer values 
+    (Wilcoxon rank-sum test, $p$ = 0.007).
+The true values of the other parameters did not significantly affect the
+estimates of $m$ (both one-way ANOVA). Finally, the total number of nodes $N$
+was consistently over-estimated by about a factor of two
+    (error 6588 
+    [4214 - 
+     8284]).
+No parameters influenced the accuracy of the $N$ estimates (all one-way ANOVA).
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
   \centering
-  \includegraphics[width=\textwidth]{abc-point-estimate-m2.pdf}
+  \includegraphics[width=\textwidth]{abc-point-estimate-m2.eps}
+  \vspace{6pt}
   \caption{
-    Point estimates of \gls{BA} model parameters obtained by running kernel-ABC
+    Point estimates of BA model parameters obtained by running kernel-ABC
     on simulated phylogenies without training, for simulations with $m = 2$.
     Dotted lines indicate true values, and limits of the $y$-axes are regions
-    of uniform prior density.
+    of uniform prior density. (A) Estimates for $\alpha$ and $I$ against their
+    true values in simulations. (B) Estimates for $m$ and $N$, which were held
+    fixed in these simulations, against true values of $\alpha$.
   }
   \label{fig:abcpt}
-\end{figure}
+\end{figure*}
 
-\begin{table}[ht]
+\begin{table*}[ht]
   \centering
-  \input{\tablepath/abc-hpd}
-  \caption{Average \gls{HPD} interval widths for \gls{ABC} model parameter
-           estimates.}
+    \begin{tabular}{lr>{\raggedleft\arraybackslash}p{2.5cm}>{\raggedleft\arraybackslash}p{2.5cm}>{\raggedleft\arraybackslash}p{2.5cm}}
+      \hline
+    Parameter & True value & Mean point estimate & Mean HPD lower bound & Mean HPD upper bound \\ 
+      \hline
+    $\alpha$ & 0.0 & 0.24 & 0.02 & 0.73 \\ 
+       & 0.5 & 0.42 & 0.02 & 0.81 \\ 
+       & 1.0 & 0.97 & 0.61 & 1.11 \\ 
+       & 1.5 & 1.48 & 1.26 & 1.83 \\ 
+      $I$ & 1000 & 1155.68 & 598.68 & 2402.84 \\ 
+       & 2000 & 2646.07 & 1182.31 & 4058.13 \\ 
+      $m$ & 2 & 2.92 & 1.75 & 4.92 \\ 
+       & 3 & 3.33 & 1.96 & 4.92 \\ 
+       & 4 & 3.62 & 1.88 & 5.00 \\ 
+      $N$ & 5000 & 10962.61 & 2732.55 & 14701.87 \\ 
+       \hline
+    \end{tabular}
+  \caption{
+      Average maximum \textit{a posteriori} point estimates and 95\% highest
+      posterior density (HPD) interval widths for BA model parameter estimates
+      obtained with kernel-ABC. Three transmission trees were simulated under
+      each combination of the listed parameter values, and the parameters were
+      estimated with kernel-ABC without training.
+  }
   \label{tab:abchpd}
-\end{table}
+\end{table*}
 
-The dispersion of the \gls{ABC} approximation to the posterior also varied
-between the parameters, with narrower \gls{HPD} intervals for those with the
-most accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows the
-distributions for for one simulation (equivalent plots for all the simulations
-can be found in supplemental data). \Gls{HPD} intervals around $\alpha$ and $I$
-were narrow relative to the region of nonzero prior density, whereas the
-intervals for $m$ and $N$ were more widely dispersed.
+The dispersion of the ABC approximation to the posterior also varied between
+the parameters, with narrower HPD intervals for the parameters with more
+accurate point estimates (Table~\ref{tab:abchpd}). Figure~\ref{fig:abcex} shows
+the distributions for for one simulation (equivalent plots for all the
+simulations can be found in Supplementary Figure S10). HPD intervals around $\alpha$
+and $I$ were narrow relative to the region of nonzero prior density, whereas
+the intervals for $m$ and $N$ were more widely dispersed.
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
     \centering
-  \includegraphics[width=\textwidth]{abc-posterior-example.pdf}
-  \caption{Marginal estimated posterior distributions obtained with ABC on a
-  single simulated dataset. Dotted lines and shaded polygon indicate true
-  values.}
+  \includegraphics[width=\textwidth]{abc-posterior-example.eps}
+  \vspace{6pt}
+  \caption{
+    Marginal posterior distributions of BA model parameters estimated
+    with kernel-ABC for a single simulated transmission tree. Dotted
+    lines and shaded polygon indicate true values.
+  }
   \label{fig:abcex}
-\end{figure}
+\end{figure*}
 
@@
 the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
-other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The \gls{MAP} [95\%
+other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
 HPD] estimates for each parameter were: 
 $\alpha$, 
-  1 [0.75-
-                           1.12];
+  1 
+  [0.07-
+   1.07];
 $I$,
-  5.24 [2.05-
-                       5.81];
+  1170 
+  [506 -
+   1616];
 $m$,
-  \ensuremath{1.04\times 10^{4}} [\ensuremath{2.27\times 10^{3}}-
-                              \ensuremath{1.5\times 10^{4}}];
+  5 
+  [2 -
+   5];
 $N$,
-  \ensuremath{1.17\times 10^{3}} [502.34-
-                       \ensuremath{1.58\times 10^{3}}].
-
+  \ensuremath{1.033\times 10^{4}} 
+  [2408-
+   \ensuremath{1.4974\times 10^{4}}].
 To test the effect of sampling bias, we sampled one transmission tree in a
 peer-driven fashion, where the probability to sample a node was twice as high
-if one of it's peers had already been sampled. The parameters for this
+if one of its peers had already been sampled. The parameters for this
 experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and $I$ = 2000. The
@@
 $\alpha$, 
-  0.46 [0.03-
-                           0.83];
+  0.48 
+  [0.03 -
+   0.82];
 $I$,
-  3.47 [2.18-
-                       5.9];
+  2468 
+  [1281 -
+   3938];
 $m$,
-  \ensuremath{1.08\times 10^{4}} [\ensuremath{3.11\times 10^{3}}-
-                              \ensuremath{1.48\times 10^{4}}];
+  3 
+  [2 -
+   5];
 $N$,
-  \ensuremath{2.94\times 10^{3}} [\ensuremath{1.28\times 10^{3}}-
-                       \ensuremath{3.94\times 10^{3}}].
+  \ensuremath{1.0738\times 10^{4}} 
+  [3106 -
+   \ensuremath{1.4783\times 10^{4}}].
+Both of these results were in line with estimates obtained on other simulated
+datasets (Table~\ref{tab:abchpd}).
 
-\subsection*{Real data}
+\subsection{Real data}
 
-There was substantial heterogeneity among the parameter estimates for the five
-published HIV datasets we analysed. Two of the
-datasets~\autocite{niculescu2015recent, wang2015targeting} had estimated
-$\alpha$ values near unity (\gls{MAP} estimate [95\% \gls{HPD}] 
+We applied kernel-ABC to five published HIV datasets (Table~\ref{tab:data}),
+and found substantial heterogeneity among the parameter estimates
+(Figure~\ref{fig:abchpd}). Two of the datasets~\citep{niculescu2015recent,
+wang2015targeting} had estimated $\alpha$ values near unity (MAP estimates
+[95\% HPD] 
   1.06 
-  [0.63-1.27]
+  [0.63 - 
+   1.27]
 and
   1 
-  [0.41-1.16]). Another two
-datasets~\autocite{li2015hiv, cuevas2009hiv} had lower estimated values and wider
-\gls{HPD} intervals
+  [0.41 -
+   1.16] respectively). 
+Another two datasets~\citep{li2015hiv, cuevas2009hiv} had lower estimated
+values and wider HPD intervals
   (0.77 
-  [0.01-1.03]
+  [0.01 - 
+  1.03]
 and
   0.66 
-  [0.03-0.84]). 
-The \textcite{novitsky2014impact} data had an extremely low estimated $\alpha$
-and a very wide \gls{HPD} interval
-  0.17 
-  [0.04-1.39]). 
-For all the datasets except \textcite{novitsky2014impact}, estimated values of
-$I$ were below 2000, with narrow \gls{HPD} intervals around two of the
+  [0.03 -
+   0.84]). 
+The \citet{novitsky2014impact} data had an extremely low estimated $\alpha$
+and a very wide HPD interval
+  (0.17 
+  [0.04 -
+   1.39]). 
+For all the datasets except \citeauthor{novitsky2014impact}, estimated values of
+$I$ were below 2000, with narrow HPD intervals around two of the
 datasets
-  (\textcite{cuevas2009hiv}, 880.52 
-  [290.7-\ensuremath{1.51\times 10^{3}}];
-   \textcite{niculescu2015recent}, 175.05
-  [138.74-454.51])
+  (\citeauthor{cuevas2009hiv}, 880 
+  [290 -
+   1511];
+   \citeauthor{niculescu2015recent}, 175
+  [138 - 
+   454])
 and wider intervals around the other two
-  (\textcite{li2015hiv}, \ensuremath{1.59\times 10^{3}} 
-  [284.01-\ensuremath{3.81\times 10^{3}}];
-   \textcite{wang2015targeting}, 651.72
-  [268.94-\ensuremath{4.24\times 10^{3}}]).
-The \textcite{novitsky2014impact} data was again the outlier, with a very high
-estimated $I$, and \gls{HPD} interval spanning almost the entire prior region
-  (\ensuremath{7.55\times 10^{3}} 
-  [228.96-\ensuremath{8.92\times 10^{3}}]).
-Little information was gleaned about the $m$ parameter, with the \gls{HPD}
-interval occupying the entire prior region for all datasets. The estimates of
-$N$ were similarly uninformative, with the exception that the point estimate
-for the \textcite{wang2015targeting} data was smaller
-  (\ensuremath{5.84\times 10^{3}})
-than the estimates for other datasets.
+  (\citeauthor{li2015hiv}, 1590 
+  [284 -
+   3807];
+   \citeauthor{wang2015targeting}, 651
+  [268 - 
+   4235]).
+The \citeauthor{novitsky2014impact} data was again the outlier, with a very high
+estimated $I$, and HPD interval spanning almost the entire prior region
+  (7547 
+  [228 -
+   8921]).
+No information was gleaned about the $m$ parameter, with the HPD interval
+occupying the entire prior region for all datasets. The estimates of $N$ were
+similarly uninformative, with the exception that the point estimate for the
+\citeauthor{wang2015targeting} data was smaller
+  (5839)
+than the estimates for other datasets
+ (average 8927).
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
   \centering
-  \includegraphics{realdata-hpd.pdf}
+  \includegraphics{realdata-hpd.eps}
+  \vspace{8pt}
   \caption{
-    Point estimates and 95\% \gls{HPD} intervals for \gls{BA} network
-    parameters of published HIV datasets, obtained with kernel-ABC.
+      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
+      parameters of the BA network model, fitted to five published HIV datasets
+      with kernel-ABC.
   }
   \label{fig:abchpd}
-\end{figure}
+\end{figure*}
 
-\section*{Discussion}
+\section{Discussion}
 
 Contact networks can have a strong influence on epidemic progression, and are
-potentially useful as a public health tool~\autocite{wang2015targeting,
+potentially useful as a public health tool~\citep{wang2015targeting,
 little2014using}. Despite this, few methods exist for investigating contact
-network parameters in a phylodynamic framework~\autocite{groendyke2011bayesian,
-brown2011transmission}. Kernel-ABC is a model-agnostic method which can be used
-to investigate any quantity that affects tree shape. In this work, we developed
-a kernel-ABC-based method to infer the parameters of a contact network model.
-The method is general, and could be applied to any model from which contact
-networks can be simulated. We demonstrated the method on the \gls{BA} model,
-which is a simple model incorporating the preferential attachment feature
-commonly seen in real-world networks.
+network parameters in a phylodynamic framework~\citep{groendyke2011bayesian,
+volz2008sir, brown2011transmission, leventhal2012inferring}. Kernel-ABC is a
+model-agnostic method which can be used to investigate any quantity that
+affects tree shape~\citep{poon2015phylodynamic}. In this work, we developed a
+kernel-ABC-based method to infer the parameters of a contact network model. The
+method is general, and could be applied to any model from which contact
+networks can be simulated. We demonstrated the method on the BA model,
+which is a simple preferential attachment model giving rise to the power law
+degree distributions commonly observed in real-world networks.
 
@@
 $N$ was almost always significantly over-estimated using kernel-ABC. Since the
-prior on $N$ and $I$ is jointly uniform on a triangular region ($I \leq N$),
-there is more prior mass on high $N$ values. In retrospect, it is unreasonable
-to expect good estimation of $N$, because adding more nodes to a \gls{BA}
-network does not change the edge density or overall shape. This can be
+prior on $N$ and $I$ is jointly uniform on a non-rectangular region ($I \leq
+N$), there is more prior mass on high $N$ values. In retrospect, it is
+unreasonable to expect good estimation of $N$, because adding more nodes to a
+BA network does not change the edge density or overall shape. This can be
 illustrated by imagining that we add a small number of nodes to a network after
@@
 
-As noted by \textcite{lintusaari2016identifiability}, uniform priors on model
+As noted by \citet{lintusaari2016identifiability}, uniform priors on model
 parameters may translate to highly informative priors on quantities of
-interest. We observed an exponential relationship between the preferential
-attachment power $\alpha$ and the power-law exponent $\gamma$
-(\cref{fig:gamma}). Therefore, placing a uniform prior on $\alpha$ between 0
-and 2 is equivalent to placing an informative prior that $\gamma$ is close to
-2. Therefore, if we were primarily interested in $\gamma$ rather than $\alpha$,
-a more sensible choice of prior might have a shape similar to \cref{fig:gamma}
-and be bounded above by approximately $\alpha$ = 1.5. This would uniformly
-bound $\gamma$ in the region $2 \leq \gamma \leq 4$ commonly reported in the
-network literature~\autocite{liljeros2001web, schneeberger2004scale,
-colgate1989risk, brown2011transmission}. We note however that
-\textcite{jones2003assessment} estimated $\gamma$ values greater than four, in
-one case as high as 17, for some datasets, indicating that a wider range of
-permitted $\gamma$ values may be warranted.
+interest. We observed a non-linear relationship between the preferential
+attachment power $\alpha$ and the power law exponent $\gamma$ (Figure~S9).
+Therefore, placing a uniform prior on $\alpha$ between 0 and 2 is equivalent to
+placing an informative prior that $\gamma$ is close to 2. Therefore, if we were
+primarily interested in $\gamma$ rather than $\alpha$, a more sensible choice
+of prior might have a shape similar to the inverse of Figure~S9 and be bounded
+above by approximately $\alpha$ = 1.5. This would uniformly bound $\gamma$ in
+the region $2 \leq \gamma \leq 4$ commonly reported in the network
+literature~\citep{liljeros2001web, schneeberger2004scale, colgate1989risk,
+brown2011transmission}. We note however that \citet{jones2003assessment}
+estimated $\gamma$ values greater than four, in one case as high as 17, for
+some datasets, indicating that a wider range of permitted $\gamma$ values may
+be warranted.
 
@@
 contact network structures underlying several distinct local epidemics. The
-five datasets analysed fell into three categories (\cref{fig:abchpd}). First,
-we estimated a preferential attachment power between 0.5 and 1 for the
-epidemics studied by \textcite{cuevas2009hiv} and \textcite{li2015hiv}, with
-credible intervals occupying nearly the entire region from 0 to 1.
+five datasets analysed fell into three categories (Figure~\ref{fig:abchpd}).
+First, we estimated a preferential attachment power between 0.5 and 1 for the
+epidemics studied by \citet{cuevas2009hiv} and \citet{li2015hiv}, with credible
+intervals occupying nearly the entire region from 0 to 1.
 \citeauthor{cuevas2009hiv} studied a group of newly diagnosed individuals in
@@
 network. \citeauthor{li2015hiv} sampled a large number of acutely infected
-\gls{MSM} in Shanghai, China, in which we identified a large cluster from the
-phylogeny using a patristic distance cutoff~\autocite{poon2014impact}. The low
+MSM in Shanghai, China, in which we identified a large cluster from the
+phylogeny using a patristic distance cutoff~\citep{poon2015impact}. The low
 attachment power estimated for this dataset was surprising given the high
 phylogenetic relatedness of the sequences. It is possible that the number and
-diversity of circulating recombinant forms introduced errors into the estimated
-viral phylogeny.
+diversity of circulating recombinant forms in the data introduced errors into
+our estimated viral phylogeny.
 
-For the outbreaks studied by \textcite{niculescu2015recent} and
-\textcite{wang2015targeting}, the estimated $\alpha$ was close to one, with a
+For the outbreaks studied by \citet{niculescu2015recent} and
+\citet{wang2015targeting}, the estimated $\alpha$ was close to one, with a
 narrower credible interval than for the other studies.
 \citeauthor{niculescu2015recent} studied a recent outbreak among Romanian
-\gls{IDU}, while \citeauthor{wang2015targeting} sampled acutely infected
-\gls{MSM} in Beijing, China. Both studies discovered a high degree of
-phylogenetic relatedness owing to the recent infection times and homogeneous
+injection drug users (IDU), while \citeauthor{wang2015targeting} sampled
+acutely infected MSM in Beijing, China. Both studies discovered a high degree
+of phylogenetic relatedness owing to the recent infection times and homogeneous
 risk groups of the studied populations. The estimated number of infections for
-these datasets were also quite low, although the \gls{HPD} interval for
+these datasets were also quite low, although the HPD interval for
 \citeauthor{wang2015targeting} was much wider than that for
@@
 The final studied dataset was an outlier in terms of estimated parameters.
-\textcite{novitsky2013phylogenetic} sampled approximately 44\% of the
+\citet{novitsky2013phylogenetic} sampled approximately 44\% of the
 HIV-infected individuals in the northern area of Mochudi, Botswana. Additional
-sampling in a later study~\autocite{novitsky2014impact} brought the genotyping
+sampling in a later study~\citep{novitsky2014impact} brought the genotyping
 coverage up to 70\%. Even with such a high sampling coverage, we did not detect
 any large clusters using patristic distance, and therefore chose to analyze a
-subtree instead. Estimates of $\alpha$ and $N$ both had very wide \gls{HPD}
+subtree instead. Estimates of $\alpha$ and $N$ both had very wide HPD
 intervals and were markedly different from the other datasets. The estimated
@@
 In a primarily heterosexual risk environment, removal of a disproportionate
-number of males from the network could obfuscate the true network structure.
+number of males from the network could obfuscate the true network structure,
+for example if the majority of highly connected nodes were of one gender.
 Second, the town in question was in close proximity to the country's capital,
@@
 
-When interpreting these results, we caution that the \gls{BA} model is quite
+When interpreting these results, we caution that the BA model is quite
 simple and most likely misspecified for these data. In particular, the average
@@
 issues, we felt it was best to demonstrate the method first on a simple model.
-It is possible to fit more complex models which address some of these issues,
-such as one incorporating heterogeneous node behaviour. This may prove a
-fruitful avenue for future investigations.
+It is possible to use this framework to fit more complex models which address
+some of these issues, such as one incorporating heterogeneous node behaviour,
+which may prove a fruitful avenue for future investigations.
 
@@
 this has been demonstrated to be a fair
-approximation~\autocite[e.g.][]{leitner1996accurate}, and is frequently used in
-practice~\autocite[e.g.][]{stadler2013uncovering}, the topologies of a viral
+approximation~\citep[e.g.][]{leitner1996accurate}, and is frequently used in
+practice~\citep[e.g.][]{stadler2013uncovering}, the topologies of a viral
 phylogeny and transmission tree can differ
-significantly~\autocite{ypma2013relating}. In addition, the \gls{ABC}-\gls{SMC}
-algorithm is computationally intensive, taking about a day when run on 20 cores
-in parallel with the settings we described in the methods. Nevertheless, our
-method is potentially useful to epidemiological researchers interested in the
-general characteristics of the network structure underlying disease outbreaks.
+significantly~\citep{ypma2013relating} due to within-host evolution and the
+sampling process. In addition, the ABC-SMC algorithm is
+computationally intensive, taking about a day when run on 20 cores in parallel
+with the settings we described in the methods. Nevertheless, our method is
+potentially useful to epidemiological researchers interested in the general
+characteristics of the network structure underlying disease outbreaks. This
+work, and previous work by our group~\citep{poon2015phylodynamic}, has
+demonstrated that kernel-ABC is a broadly applicable and effective framework in
+which to perform phylodynamic inference.
 
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{S\arabic{figure}}
+\section{Materials and Methods}
 
-\printbibliography
+Nodes in our networks followed simple SI dynamics, meaning that they
+became infected at a rate proportional to their number of infected neighbours,
+and never recovered. For all analyses, the transmission trees' branch lengths
+were scaled by dividing by their mean. We used the \textit{igraph} library's
+implementation of the BA model~\citep{csardi2006igraph} to generate the graphs.
+The analyses were run on Westgrid (\url{https://www.westgrid.ca/}) and a local
+computer cluster.
 
-\newpage
-\section*{Supplemental Materials}
+\subsection{Computer program validation}
 
-\begin{figure}[ht]
+To check that our implementation of Gillespie simulation was correct, we
+reproduced Figure 1A of \citet{leventhal2012inferring} (our Figure~S1), which
+plots the unbalancedness of transmission trees simulated over four network
+models at various levels of pathogen transmissibility. Our implementation of
+adaptive ABC-SMC was tested by applying it to the same mixture of Gaussians
+used by \citeauthor{del2012adaptive} to demonstrate their method (originally
+used by~\citet{sisson2007sequential}). We were able to obtain a close
+approximation to the function (see Figure~S2), and attained the stopping
+condition used by the authors in a comparable number of steps.
+
+\subsection{Kernel classifiers}
+
+We used the phylogenetic kernel developed by \citet{poon2013mapping} to test
+whether the parameters of the BA model had an effect on tree shape. 100
+networks were simulated under each of three different values of $\alpha$: 0.5,
+1.0, and 1.5 (300 networks total). The other parameters were fixed to the
+following values: $N$ = 5000, $I$ = 1000, and $m$ = 2. A transmission tree with
+500 tips was simulated over each network (300 transmission trees total). The
+300 trees were compared pairwise with the tree kernel to form a $300 \times
+300$ kernel matrix. The kernel meta-parameters $\lambda$ (the ``decay
+factor''), and $\sigma$ (the ``radial basis function
+variance'')~\citep[see][]{poon2013mapping}, were set to 0.3 and 4 respectively.
+We constructed a kSVM classifier for $\alpha$ using the \textit{kernlab}
+package~\citep{zeileis2004kernlab}, and evaluated its accuracy with 1000
+two-fold cross-validations.
+
+Three similar experiments were performed for the other BA model parameters (one
+experiment per parameter). $m$ was varied between 2, 3, and 4; $I$ between 500,
+1000, and 2000; and $N$ between 3000, 5000, and 8000. The parameters not being
+tested were fixed at the values $N$ = 5000, $I$ = 1000, $m$ = 2, and $\alpha$ =
+1. Thus, we performed a total of four kSVM cross-validations, one for each of
+the BA model parameters $\alpha$, $I$, $m$, and $N$. We repeated these four
+cross-validations with different values of $\lambda$ (0.2, 0.3, and 0.4) and
+$\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$), as well as on trees with
+differing numbers of tips (100, 500, and 1000) and in epidemics of differing
+size (500, 1000, and 2000). The combination of the number of sampled
+individuals (\textit{i.e.} the number of tips) and the epidemic size
+(\textit{i.e.} $I$) will be referred to as an ``epidemic scenario''. When
+evaluating the classifier for $I$, we did not consider trees with 1000 tips,
+because one of the tested $I$ values was 500, and the number of tips cannot be
+larger than $I$.
+
+For each of the four parameters, we also tested a linear regression against
+Sackin's index~\citep{shao1990tree} and an ordinary SVM based on the normalized
+lineages-through-time (nLTT) statistic~\citep{janzen2015approximate}.
+
+\subsection{ABC simulations}
+
+We simulated three transmission trees, each with 500 tips, under every element
+of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
+\{1000, 2000\}, $m$ = \{2, 3, 4\}, and $\alpha$ = \{0.0, 0.5, 1,
+1.5\}. This produced a total of 24 parameter combinations $\times$ three trees
+per combination = 72 trees total. The adaptive ABC algorithm was applied to
+each tree with these priors: $m \sim$ Uniform(1, 5), $\alpha \sim$ Uniform(0,
+2), and $(N, I)$ jointly uniform on the region \{$500 \leq N \leq 15000$, $500
+\leq I \leq 5000$, $I \leq N$\}. Following \citet{del2012adaptive} and
+\citet{beaumont2009adaptive}, all proposals were Gaussian, with variance equal
+to twice the empirical variance of the particles. The algorithm was run with
+1000 particles, 5 simulated datasets per particle, and the ``quality''
+parameter controlling the decay rate of the tolerance $\varepsilon$ set to
+0.95. We used the same stopping criterion as \citeauthor{del2012adaptive},
+namely when the MCMC acceptance rate dropped below 1.5\%. Point estimates for
+the parameters were obtained by taking the highest point of an estimated kernel
+density on the final set of particles, calculated using the \textit{density}
+function with the default parameters in \textit{R}. Highest posterior density
+(HPD) intervals were calculated with the \textit{HPDinterval} function from
+the \textit{R} package \textit{coda}~\citep{plummer2006coda}.
+
+Two further analyses were performed to address potential sources of error. To
+evaluate the effect of model misspecification in the case of heterogeneity
+among nodes, we generated a network where half the nodes were attached with
+power $\alpha$ = 0.5, and the other half with power $\alpha$ = 1.5. The other
+parameters for this network were $N$ = 5000, $I$ = 1000, and $m$ = 2. To
+investigate the effects of potential sampling bias, we simulated a transmission
+tree where the tips were sampled in a peer-driven fashion, rather than at
+random. That is, the probability to sample a node was twice as high if any of
+that node's network peers had already been sampled. The parameters of this
+network were $N$ = 5000, $I$ = 2000, $m$ = 2, and $\alpha$ = 0.5.
+
+\subsection{Investigation of published data}
+
+We applied our kernel-ABC method to several published HIV datasets. Because the
+BA model generates networks with a single connected component, we
+specifically searched for datasets which originated from existing clusters,
+either phylogenetically or geographically defined. Characteristics of the
+datasets we investigated are given in Table~\ref{tab:data}.
+
+\begin{table*}[!t]
   \centering
-  \includegraphics{leventhal2012fig1.pdf}
-  \caption{Reproduction of Figure 1A from \textcite{leventhal2012inferring}.}
-  \label{fig:sf1}
-\end{figure}
+  \begin{tabular}{ccccc}
+    Reference & Sequences ($n$) & Location & Risk group & Gene \\
+    \hline
+    \citet{wang2015targeting} & 173 & Beijing, China & MSM & \textit{pol} \\
+    \citet{cuevas2009hiv} & 287 & Basque Country, Spain & mixed & \textit{pol} \\
+    \citet{novitsky2013phylogenetic} & \multirow{2}{*}{180} &
+    \multirow{2}{*}{Mochudi, Botswana} & \multirow{2}{*}{mixed} &
+    \multirow{2}{*}{\textit{env}} \\ \citet{novitsky2014impact} \\
+    \citet{li2015hiv} & 280 & Shanghai, China & MSM & \textit{pol} \\
+    \citet{niculescu2015recent} & 136 & Romaina & IDU & \textit{pol} \\
+    \hline
+  \end{tabular}
+  \caption{Characteristics of published datasets investigated with kernel-ABC.
+  Acronyms: MSM, men who have sex with men; IDU, injection drug users.}
+  \label{tab:data}
+\end{table*}
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics{smc-test.pdf}
-  \caption{Approximation of mixture of Gaussians used by
-    \textcite{del2012adaptive} and \textcite{sisson2007sequential} to test
-    \gls{SMC}. Solid black line indicates true distribution. Grey shaded area
-    shows SMC approximation obtained with our implementation.}
-  \label{fig:smctest}
-\end{figure}
+We downloaded all sequences associated with each study from GenBank. For the
+\citet{novitsky2014impact} data, each sequence was aligned pairwise to the HXB2
+reference sequence (Genbank accession number K03455) and the hypervariable
+regions were clipped out with \textit{BioPython} version
+1.66+~\citep{cock2009biopython}. Sequences were multiply aligned using
+\textit{MUSCLE} version 3.8.31 \citep{edgar2004muscle}, and alignments were
+manually inspected with \textit{Seaview} version 4.4.2
+\citep{gouy2010seaview}. Phylogenies were constructed from the nucleotide
+alignments by approximate maximum likelihood using \textit{FastTree2} version
+2.1.7 with the generalized time-reversible model. Transmission trees were
+estimated by rooting and time-scaling the phylogenies by root-to-tip
+regression, using a modified version of Path-O-Gen (distributed as part of
+BEAST~\citep{drummond2007beast}) as described
+previously~\citep{poon2015phylodynamic}.
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics{kernel-alpha-crossv.pdf}
-  \caption{
-    Cross-validation accuracy of kernel-SVM classifiers for $\alpha$ parameter
-    of \gls{BA} network model, for various tree kernel meta-parameters and
-    epidemic scenarios. Each point was calculated based on 300 simulated
-    transmission trees over networks with $\alpha$ = 0.5, 1.0, or 1.5. Dotted
-    and and dashed lines indicate, respectively, performance of SVM using the
-    \gls{nLTT} statistic, and linear regression using Sackin's index. Facets
-    are number of infected nodes before the simulation was stopped ($I$) and
-    number of tips in the sampled transmission tree.
-  }
-  \label{fig:alphacrossv}
-\end{figure}
+Two of the datasets \citep{li2015hiv,novitsky2014impact} were initially much
+larger than the others, containing 1265 and 1299 sequences respectively. To
+ensure that the analyses were comparable, we reduced these to a number of
+sequences similar to the smaller datasets. For the \citet{li2015hiv} data,
+we detected a cluster of size 280 using a patristic distance cutoff of 0.02 as
+described previously~\citep{poon2015impact}. Only sequences within this
+cluster were carried forward. For the \citet{novitsky2014impact} data, no
+large clusters were detected using the same cutoff, so we analysed a subtree of
+size 180 chosen arbitrarily.
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics{kernel-m-crossv.pdf}
-  \caption{Cross-validation accuracy of kernel-SVM classifiers for $m$
-      parameter of \gls{BA} network model, for various tree kernel
-      meta-parameters and epidemic scenarios. Each point was calculated based
-      on 300 simulated transmission trees over networks with $m$ = 2, 3, or 4.
-      Dotted and and dashed lines indicate, respectively, performance of SVM
-      using the \gls{nLTT} statistic, and linear regression using Sackin's
-      index. Facets are number of infected nodes before the simulation was
-      stopped ($I$) and number of tips in the sampled transmission tree.
-  }
-  \label{fig:mcrossv}
-\end{figure}
+For all datasets, we used the priors $\alpha$ $\sim$ Uniform(0, 2), $m$ $\sim$
+Uniform(2, 5), and $N$ and $I$ jointly uniform on the region 
+\{$n \leq N \leq 10000$, $n \leq I \leq 10000$, $I \leq N$\}, where $n$ is the
+number of tips in the tree. The other parameters to the SMC algorithm
+were the same as used for the simulation experiments.
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics{kernel-I-crossv.pdf}
-  \caption{Cross-validation accuracy of kernel-SVM classifiers for number of
-      infected nodes ($I$) under \gls{BA} network model, for various tree
-      kernel meta-parameters and two tree sizes. Each point was calculated
-      based on 300 simulated transmission trees over networks with $I$ = 500,
-      1000, or 2000. Dotted and and dashed lines indicate, respectively,
-      performance of SVM using the \gls{nLTT} statistic, and linear regression
-      using Sackin's index. Facets are the number of tips in the sampled
-      transmission tree.
-  }
-  \label{fig:Icrossv}
-\end{figure}
+\section{Supplementary data}
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics{kernel-N-crossv.pdf}
-  \caption{Cross-validation accuracy of kernel-SVM classifiers for total number
-      of nodes ($N$) under \gls{BA} network model, for various tree kernel
-      meta-parameters and epidemic scenarios sizes. Each point was calculated
-      based on 300 simulated transmission trees over networks with $N$ = 3000,
-      5000, or 8000. Dotted and and dashed lines indicate, respectively,
-      performance of SVM using the \gls{nLTT} statistic, and linear regression
-      using Sackin's index. Facets are the number of tips in the sampled
-      transmission tree.
-  }
-  \label{fig:Ncrossv}
-\end{figure}
+Supplementary figures S1-S10 are available at Molecular Biology and Evolution
+online (\url{http://www.mbe.oxfordjournals.org/}).
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics{abc-point-estimate-m3.pdf}
-  \caption{
-    Point estimates of \gls{BA} model parameters obtained by running kernel-ABC
-    on simulated phylogenies without training, for simulations with $m = 3$.
-    Dotted lines indicate true values, and limits of the $y$-axes are regions
-    of uniform prior density.
-  }
-  \label{fig:abcptm3}
-\end{figure}
+\section{Acknowledgements}
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics{abc-point-estimate-m4.pdf}
-  \caption{
-    Point estimates of \gls{BA} model parameters obtained by running kernel-ABC
-    on simulated phylogenies without training, for simulations with $m = 4$.
-    Dotted lines indicate true values, and limits of the $y$-axes are regions
-    of uniform prior density.
-  }
-  \label{fig:abcptm4}
-\end{figure}
+This work was supported by grants from the Canadian Institutes of Health
+Research (CIHR, operating grant HOP-111406), and the Bill \& Melinda Gates
+Foundation (award number OPP1110049). R.M.M. was supported by a scholarship
+from the CIHR Strategic Training Program in Bioinformatics. A.F.Y.P. was
+supported by a CIHR New Investigator Award (Canadian HIV Vaccine Initiative,
+Vaccine Discovery and Social Research) and by a Career Investigator Scholar
+Award from the Michael Smith Foundation for Health Research, in partnership
+with the Providence Health Care Research Institute and St. Paul's Hospital
+Foundation.
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics{alpha-gamma.pdf}
-  \caption{
-      Relationship between preferential attachment power parameter $\alpha$
-      and power-law exponent $\gamma$ for networks simulated under the \gls{BA}
-      network model with $N$ = 5000 and $m$ = 2.
-  }
-  \label{fig:gamma}
-\end{figure}
+\bibliographystyle{natbib}
+\bibliography{papers}
 

### 2016-04-06 doc/thesis/objective.tex
Finished results! Yes!
More writing

@@
+\section{Objective}
+
 The spread of a disease is most often modelled by assuming either a
-homogeneously mixed population, or a population divided into a small number of
-homogeneously mixed compartments. This assumption, also called the ``law of
-mass action'', implies that any two individuals in the same compartment are
-equally likely to come into contact causing transmission. Although this
-provides a reasonable approximation in many cases, the error introduced by
-assuming a panmictic population can be substantial when significant contact
-heterogeneity exists in the underlying population. Contact network models
-provide an alternative to compartmental models which do not require the
-assumption of panmixia. In addition to more accurate predictions, the
-parameters of the networks themselves may be of interest from a public health
-perspective. For example, certain vaccination strategies may be more or less
-effective in curtailing an epidemic depending on the underlying network's
-degree distribution. Phylodynamic methods have been used to fit many different
-types of model to phylogenetic data, but as far as we know, no methods have yet
-been developed to fit contact network models. The primary objective of this
-work is to develop such a method.
+homogeneously mixed population~\autocite{hamer1906milroy,
+kermack1927contribution}, or a population divided into a small number of
+homogeneously mixed groups~\autocite{rushton1955deterministic}. This
+assumption, also called the ``law of mass action'', implies that any two
+individuals in the same compartment are equally likely to come into contact
+causing transmission. Although this provides a reasonable approximation in many
+cases, the error introduced by assuming a panmictic population can be
+substantial when significant contact heterogeneity exists in the underlying
+population. Contact network models provide an alternative to compartmental
+models which do not require the assumption of panmixia. In addition to more
+accurate predictions, the parameters of the networks themselves may be of
+interest from a public health perspective. For example, certain vaccination
+strategies may be more or less effective in curtailing an epidemic depending on
+the underlying network's degree distribution. Phylodynamic methods have been
+used to fit many different types of model to phylogenetic data, but as far as
+we know, no methods have yet been developed to fit contact network models. The
+primary objective of this work is to develop such a method.
 
@@
 the graph may have different labels or attributes. Hence, the number of terms
-in the sum is at least exponential in $n$, more if we do not know \textit{a
-priori} how many nodes are in the network (as is likely). In addition,
-\cref{eq:netlik} assumes that $T$ is complete, meaning that all infected
-individuals were sampled. This is rarely the case in practice - most often, the
-observed tree is a subsampled version of the true tree. In this case, the
-likelihood calculation becomes even more complex, because we must also sum over
-all possible complete trees.
+in the sum is at least exponential in $n$, as there must be at least $n$ nodes
+in the network. In addition, \cref{eq:netlik} assumes that $T$ is complete,
+meaning that all infected individuals were sampled. This is rarely the case in
+practice - most often, the observed tree is a subsampled version of the true
+tree. In this case, the likelihood calculation becomes even more complex,
+because we must also sum over all possible complete trees.
 

### 2016-04-06 doc/thesis/results.Rnw
More writing

@@
 produce networks with a small number of highly connected nodes which, once
-infected, are likely to transmit to many other nodes (\cref{fig:alphabds}).
-This results in a more unbalanced, ladder-like structure in the phylogeny,
-compared to networks with lower \gls{alpha} values. Sackin's index was
-significantly correlated with \gls{alpha} 
+infected, are likely to transmit to many other nodes. This results in a more
+unbalanced, ladder-like structure in the phylogeny, compared to networks with
+lower \gls{alpha} values. Sackin's index was significantly correlated with
+\gls{alpha} 
     (Spearman's rho \Sexpr{round(alpha.sackin.cor$estimate, 2)},
@@
     \Sexpr{round(I.ratio.cor$estimate, 2)},
-    both $p < 1e-5$).
-
-Kernel-\gls{PCA} projections show that all three \gls{alpha} values are well
-separated from each other in feature space under several different sampling
-scenarios (Figure~\ref{fig:pakpca}). With smaller trees, it becomes harder to
-visually distinguish \gls{alpha} = 0.5 from \gls{alpha} = 1.0 using the first
-two principal components.
-
-With suitably chosen meta-parameters, the accuracy of the kernel-\gls{SVM}
-classifier for \gls{alpha} was very high under a variety of prevalence and
-sampling scenarios. Accuracy was highest, with
-$R^2$ values above 0.95, for the largest trees and complete sampling (bottom
-center panel). However, even for trees of size 100 sampled from an epidemic on
-2000 nodes, the $R^2$ was above 0.8 (top right panel). In all cases, the
-accuracy of a Sackin's index-based classifier was also quite high, at about
-0.75. An \gls{SVM} classifier using only the \gls{nltt}
-statistic~\autocite{janzen2015approximate} was slightly worse than Sackin's
-index, with an $R^2$ of X.
+    both $p < 10^{-5}$).
 
@@
 
+\Cref{fig:kpca} shows \gls{kPCA} projections of the simulated trees onto the
+first two principal components of the kernel matrix. The figure shows only the
+simulations with 500-tip trees and, for all parameters except $I$, 1000
+infected nodes. The three \gls{alpha} and \gls{I} values considered are well
+separated from each other in feature space (\cref{fig:pakpca}). On the other
+hand, the three \gls{N} values overlap significantly, and the three \gls{m}
+values are virtually indistinguishable. Similar observations can be made for
+other values of \gls{I} and the number of tips 
+(\cref{fig:alphakpca,fig:Nkpca,fig:Ikpca,fig:mkpca}). The values of \gls{I} and
+\gls{N} separate more clearly with larger numbers of tips, and in the case of
+\gls{N} larger epidemic sizes.
+
 \begin{figure}[ht]
@@
   \includegraphics{kernel-kpca.pdf}
-  \caption[Projection of kernel matrix for different attachment power values
-  onto its first two principal components]{
-    Projection of the kernel matrix for different preferential attachment power
-    values onto its first two principal components, for eight simulation
-    scenarios. Each point corresponds to a simulated transmission tree, and is
-    coloured by preferential attachment power. Facets are number of infected
-    nodes (horizontal), and number of sampled tips (vertical). The parameters
-    to the tree kernel were $\lambda = 0.3$ and $\sigma = 4$, and the nLTT was
-    not used. Qualitatively, trees with a larger number of tips are easier to
-    separate in kernel space, regardless of what sampling proportion they
-    represent. In all cases, the highest attachment power can be separated from
-    the other two, but the two lowest values become hand to distinguish with in
-    the 100-tip trees.
+  \caption[\gls{kPCA} projections of simulated trees under varying \gls{BA}
+           parameter values]{
+    Each parameter of the \gls{BA} model was individually varied to produce 300
+    simulated trees. Kernel matrices were formed from all pairwise kernel
+    scores among each set of 300 trees. The trees were projected onto the first
+    two principal components of the kernel matrix calculated using \gls{kPCA}.
+    All trees had 500 tips. The parameters not being varied were set to
+    \gls{alpha} = 1, \gls{I} = 1000, \gls{m} = 2, and \gls{N} = 5000. The tree
+    kernel meta-parameters were $\lambda = 0.3$ and $\sigma = 4$.
   }
@@
 
-We also considered the possibility of inferring the number of infected nodes,
-or \defn{prevalence}, under this model. All parameters except \gls{I} were
-fixed at the following values: \gls{N} = 5000, \gls{alpha} = 1.0, and \gls{m} =
-2. As shown in Figure~\ref{fig:prevtrees}, the prevalence had no obvious effect
-on the tree shape. However, a kernel-\gls{SVM} classifier was able to
-distinguish the number of infected nodes with high accuracy ($R^2 > 0.9$;
-Figure~\ref{fig:prevcrossv}). Moreover, the use of the \gls{nltt} statistic
-improved classification accuracy by a small amount, in contrast to the results
-for \gls{alpha}. In contrast, a Sackin's index-based classifier displayed
-extremely poor performance ($R^2$ < 0.1, not shown).
+<<classifiers, include=FALSE>>=
+    get.kernel.data <- function (param, step) {
+        md <- collect.metadata(paste0('../../simulations/kernel-', param, '/', step, '/*'))
+        if ('rbf_variance' %in% colnames(md)) {
+          md <- subset(md, rbf_variance == 4 & decay_factor == 0.3)
+        }
+        if (param == 'alpha') {
+            md <- subset(md, m == 2)
+        }
+        d <- setDT(collect.data(rownames(md)))
+        if (param %in% c('alpha', 'm')) {
+            d[,list(rsquared=mean(rsquared)), by=c('nsimnode', 'ntip')]
+        } else if (param == 'I') {
+            d[,list(rsquared=mean(rsquared)), by=ntip]
+        } else {
+            d[,list(rsquared=mean(rsquared)), by=c('I', 'ntip')]
+        }
+    }
+    kernel.alpha <- get.kernel.data('alpha' ,'classifier')
+    kernel.m <- get.kernel.data('m', 'classifier')
+    kernel.I <- get.kernel.data('I', 'classifier')
+    kernel.N <- get.kernel.data('N', 'classifier')
 
-It is important to note that, if we cut off the epidemic simulation when 500
-nodes are infected, the resulting tree will be shorter (in calendar time) than
-if we continue until 2000 nodes are infected. However, this information is not
-used when building the classifier, since the branch lengths in each tree are
-scaled by their mean. Therefore, the high performance of the classifier is due
-to structural differences captured by the tree kernel, rather than the trees
-simply having different heights.
+    sackin.alpha <- get.kernel.data('alpha', 'stats-classifier')
+    sackin.m <- get.kernel.data('m', 'stats-classifier')
+    sackin.I <- get.kernel.data('I', 'stats-classifier')
+    sackin.N <- get.kernel.data('N', 'stats-classifier')
 
-\subsection{Accuracy of marginal estimates}
+    nltt.alpha <- get.kernel.data('alpha', 'nltt-classifier')
+    nltt.m <- get.kernel.data('m', 'nltt-classifier')
+    nltt.I <- get.kernel.data('I', 'nltt-classifier')
+    nltt.N <- get.kernel.data('N', 'nltt-classifier')
+@
+
+Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
+tested (\cref{fig:rsquared}, left). Classifiers based on two other tree
+statistics, the \gls{nltt} and Sackin's index, generally exhibited worse
+performance than the tree kernel, although the magnitude of the disparity
+varied between the parameters (\cref{fig:rsquared}, centre and right).
+\Cref{fig:rsquared} shows the cross-validation accuracy of each of the tree
+classifiers. The results were largely robust to variations in the tree kernel
+meta-parameters $\lambda$ and $\sigma$, although accuracy varied between
+different epidemic and sampling scenarios
+(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}).
+
+When classifying $\alpha$, the kernel-SVM classifier had an average $R^2$ of 
+    \Sexpr{kernel.alpha[,round(mean(rsquared), 2)]},
+compared to 
+    \Sexpr{nltt.alpha[,round(mean(rsquared), 2)]}
+for the \gls{nltt}-based SVM, and
+    \Sexpr{sackin.alpha[,round(mean(rsquared), 2)]}
+for the linear regression against Sackin's index. There was little variation
+about the mean for different tree and epidemic sizes. No classifier could
+accurately identify the $m$ parameter in any epidemic scenario, with average
+$R^2$ values of 
+  \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for \gls{kSVM},
+  \Sexpr{nltt.m[,round(mean(rsquared), 2)]} for the \gls{nltt}, and
+  \Sexpr{sackin.m[,round(mean(rsquared), 2)]}
+for Sackin's index. Again, there was little variation in accuracy between
+epidemic scenarios, although the accuracy of the \gls{kSVM} was slightly higher
+on 1000-tip trees.
+
+The accuracy of classifiers $I$ varied significantly with the number of tips in
+the tree. For 100-tip trees, the average $R^2$ values were
+  \Sexpr{kernel.I[ntip == 100, round(mean(rsquared), 2)]},
+  \Sexpr{nltt.I[ntip == 100, round(mean(rsquared), 2)]}, and
+  \Sexpr{sackin.I[ntip == 100, round(mean(rsquared), 2)]}
+for the tree kernel, \gls{nltt}, and Sackin's index respectively. For 500-tip
+trees, the values increased to
+  \Sexpr{kernel.I[ntip == 500, round(mean(rsquared), 2)]},
+  \Sexpr{nltt.I[ntip == 500, round(mean(rsquared), 2)]}, and
+  \Sexpr{sackin.I[ntip == 500, round(mean(rsquared), 2)]}.
+Finally, the performance of classifiers for $N$ depended heavily on the
+epidemic scenario. The $R^2$ of the \gls{kSVM} classifier ranged from
+  \Sexpr{kernel.N[,round(min(rsquared), 2)]}
+for the smallest epidemic and smallest sample size, to
+  \Sexpr{kernel.N[,round(max(rsquared), 2)]}
+for the largest. Likewise, $R^2$ for the \gls{nltt}-based SVM ranged from 
+  \Sexpr{nltt.N[,round(min(rsquared), 2)]}
+to
+  \Sexpr{nltt.N[,round(max(rsquared), 2)]}.
+Sackin's index did not accurately classify $N$ in any scenario, with an average
+$R^2$ of
+  \Sexpr{sackin.N[,round(mean(rsquared), 2)]}
+and little variation between scenarios.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
+  \caption{
+      Cross-validation accuracy of kernel-SVM classifier (left), SVM classifier
+      using \gls{nltt} (centre), and linear regression using Sackin's index
+      (right) for \gls{BA} model parameters. Kernel meta-parameters were set to
+      $\lambda = 0.3$ and $\sigma = 4$. Each point was calculated based on 300
+      simulated transmission trees over networks with three different values of
+      the parameter being tested. Vertical lines are empirical 95\% confidence
+      intervals based on 1000 two-fold cross-validations.
+  }
+  \label{fig:rsquared}
+\end{figure}
+
+\subsection{Grid search}
 
@@
 The error in the estimated value of $I$ was
-    \Sexpr{d[,median(I_error)]} 
-    [\Sexpr{d[,quantile(I_error, 0.25)]}-\Sexpr{d[,quantile(I_error, 0.75)]}].
+    \Sexpr{d[,round(median(I_error))]} 
+    [\Sexpr{d[,round(quantile(I_error, 0.25))]}-\Sexpr{d[,round(quantile(I_error, 0.75))]}].
 Errors were significantly higher for $\alpha \geq 1$
@@
 
-Figure~\ref{fig:abcex} shows the \gls{ABC} approximation to the
-posterior distribution on the \gls{BA} parameters for one simulation
-(equivalent plots for all the simulations can be found in the supplemental
-materials). \Gls{HPD} intervals around \gls{alpha} and \gls{I} were narrow
-relative to the region of nonzero prior density, whereas the intervals for $m$
-and \gls{N} were widely dispersed. Table~\ref{tab:abchpd} shows point estimates and
-95\% \gls{HPD} intervals averaged over all simulations.
+\Cref{fig:abcex} shows the \gls{ABC} approximation to the posterior
+distribution on the \gls{BA} parameters for one simulation (equivalent plots
+for all the simulations can be found in the supplemental materials). \Gls{HPD}
+intervals around \gls{alpha} and \gls{I} were narrow relative to the region of
+nonzero prior density, whereas the intervals for $m$ and \gls{N} were widely
+dispersed. Table~\ref{tab:abchpd} shows point estimates and 95\% \gls{HPD}
+intervals averaged over all simulations.
 

### 2016-04-11 programs/netsim/src/simulate.c
Removed netsim program (has its own repo now)

@@ -1,281 +1,0 @@
-#include <stdio.h>
-#include <stdlib.h>
-#include <string.h>
-#include <assert.h>
-#include <math.h>
-#include <Judy.h>
-#include <igraph/igraph.h>
-#include <gsl/gsl_randist.h>
-
-#include "simulate.h"
-#define NDEBUG
-
-#ifndef INFINITY
-#include <float.h>
-#define INFINITY DBL_MAX
-#endif
-
-void print_node(const igraph_t *net, char *buf, int node, int numeric_ids);
-
-void simulate_phylogeny(igraph_t *tree, igraph_t *net, gsl_rng *rng,
-        double stop_time, int stop_nodes, int numeric_ids)
-{
-    long i;
-    int inode, snode, e, v, n, head, tail, nnode_tree = 0;
-    int Rc_int, ndiscordant, ninfected = 1;
-    double t, r, sum, trans_rate = 0., remove_rate = 0., time = 0.;
-    char buf[128];
-    igraph_vector_int_t *incident;
-    igraph_vector_t edges, branch_lengths;
-    igraph_inclist_t inclist_in, inclist_out;
-    Word_t Bytes, Index, *PValue;
-
-    if (stop_nodes <= 0) {
-        stop_nodes = igraph_vcount(net);
-    }
-    if (stop_time <= 0) {
-        stop_time = INFINITY;
-    }
-
-    // set of infected nodes
-    Pvoid_t infected = (Pvoid_t) NULL;
-
-    // set of removed nodes
-    Pvoid_t removed = (Pvoid_t) NULL;
-
-    // set of discordant edges
-    Pvoid_t discordant = (Pvoid_t) NULL;
-
-    // map from nodes in the network to extant tips in the tree
-    Pvoid_t tip_map = (Pvoid_t) NULL;
-
-    // map from nodes (internal + terminal) in the tree to nodes in the network
-    Pvoid_t node_map = (Pvoid_t) NULL;
-
-    igraph_inclist_init(net, &inclist_in, IGRAPH_IN);
-    igraph_inclist_init(net, &inclist_out, IGRAPH_OUT);
-
-    igraph_vector_init(&edges, 0);
-    igraph_vector_init(&branch_lengths, 0);
-
-    // start the epidemic
-    inode = gsl_rng_get(rng) % igraph_vcount(net);
-    J1S(Rc_int, infected, inode);
-#ifndef NDEBUG
[…]

### 2016-04-11 programs/netsim/tests/check_util.c

@@ -1,326 +1,0 @@
-#include <check.h>
-#include <limits.h>
-#include <igraph/igraph.h>
-#include <gsl/gsl_rng.h>
-
-#include "../src/util.h"
-
-Suite *util_suite(void);
-
-START_TEST (test_get_scale_nonzero)
-{
-    double x[2] = { 1e5, 1e-10 };
-    int scale = get_scale(x, 2);
-    ck_assert_int_eq(scale, -2);
-}
-END_TEST
-
-START_TEST (test_get_scale_zero)
-{
-    double x[2] = { 1e5, 0 };
-    int scale = get_scale((double *) &x, 2);
-    ck_assert_int_eq(scale, 5);
-    x[0] = 0; x[1] = 1e5;
-    scale = get_scale((double *) &x, 2);
-    ck_assert_int_eq(scale, 5);
-}
-END_TEST
-
-START_TEST (test_get_scale_tiny)
-{
-    double x[2] = { 1e5, 1e-200 };
-    int scale = get_scale(x, 2);
-    ck_assert_int_eq(scale, 5);
-    x[0] = 1e-200; x[1] = 1e5;
-    scale = get_scale(x, 2);
-    ck_assert_int_eq(scale, 5);
-}
-END_TEST
-
-START_TEST (test_set_seed)
-{
-    gsl_rng *rng = set_seed(0);
-    ck_assert(rand() == 1804289383);
-    ck_assert(gsl_rng_get(rng) == 4293858116);
-    ck_assert(igraph_rng_get_integer(igraph_rng_default(), 0, INT_MAX) == 2146929058);
-}
-END_TEST
-
-START_TEST (test_order_ints)
-{
-    int x[5] = {4, 2, 1, 5, 3};
-    int ord[5];
-    order(x, ord, sizeof(int), 5, compare_ints);
-
-    ck_assert_int_eq(ord[0], 2);
-    ck_assert_int_eq(ord[1], 1);
-    ck_assert_int_eq(ord[2], 4);
-    ck_assert_int_eq(ord[3], 0);
-    ck_assert_int_eq(ord[4], 3);
-}
-END_TEST
-
-START_TEST (test_order_doubles)
-{
-    double x[5] = {4, 2, 1, 5, 3};
-    int ord[5];
-    order(x, ord, sizeof(double), 5, compare_doubles);
-
-    ck_assert_int_eq(ord[0], 2);
-    ck_assert_int_eq(ord[1], 1);
[…]

### 2016-04-12 doc/thesis/intro.tex
More writing.

@@
 \subsection{Tree shapes}
-\label{subsubsec:treeshape}
+\label{subsec:treeshape}
 
@@
 coalescent~\autocite{kingman1982coalescent} models.
-\textcite{stadler2011estimating} develop a formula for the likelihood of a
+\textcite{stadler2012estimating} develop a formula for the likelihood of a
 phylogeny with heterochronous tips under the birth-death model, which has been
 used to estimate the basic reproductive number of several viral
-epidemics~\autocite{stadler2011estimating}. However, the birth-death model is
+epidemics~\autocite{stadler2012estimating}. However, the birth-death model is
 cannot tell us anything about population structure, as it assumes that every
@@
 
+\section{Sequential Monte Carlo}
+\label{sec:smc}
+
+\Gls{SMC} is a statistical inference method which samples from a sequence of
+probability distributions in a fixed order~\autocite{del2006sequential}, based
+on a technique called \gls{SIS}.
+
+\subsection{Sequential importance sampling}
+\label{subsec:sis}
+
+The basis of \gls{SIS} is \gls{IS}, which is a method of estimating summary
+statistics of distributions which are known only up to a normalizing constant,
+and therefore cannot be sampled from directly. That is, if $\pi$ is such a
+distribution and $f$ is any real-valued function, \gls{IS} is concerned with
+estimating
+\[
+  \pi(f) = \int f(x)\pi(x)\d x = \int f(x) \frac{\gamma(x)}{Z} \d x,
+\]
+where the integral is over the space on which $\pi$ is defined, $\gamma(x)$ is
+known pointwise, and $Z = \int \gamma(x) \d x$ is the unknown normalizing
+constant. Suppose we have at hand another distribution $\eta$, called the
+\defn{importance distribution}, from which we are able to sample. Define the
+\defn{importance weight} as the ratio ratio $w(x) = \gamma(x)/\eta(x)$. We can
+express the normalizing constant $Z$ in terms of the importance weight and
+distribution, $Z = \int w(x) \eta(x) \d x$, and in turn write the expectation
+of interest as
+\[
+  \int f(x) \pi(x) \d x = \frac{\int f(x) \gamma(x) \d x}
+                               {\int w(x) \eta(x) \d x}.
+\]
+If we sample a large number of points from $\eta$, then $\eta(x)$ can be
+approximated by a Monte Carlo estimate. Since the remaining quantities $f$,
+$\gamma$, and $w$ can all be evaluated pointwise, these are all the ingredients
+we need to obtain an estimate of $\pi(f)$. Although this is a simple and
+elegant approach, the drawback is that the variance of the estimate is
+proportional to the variance of the importance weights~\autocite{liu2008monte},
+which may be quite large if $\eta$ and $\gamma$ are very different. Therefore,
+the practical use of \gls{IS} on its own is limited, since it depends on
+finding an importance distribution similar to $\pi$, which we usually know very
+little about \textit{a priori}.
+
+\gls{SIS} applies \gls{IS} to the problem of sampling from a distribution $\pi$ 
+on an high-dimensional space, say $\pi(\vec{x}) = \pi(x_1, \ldots, x_d)$. The
+objective is to build up an importance distribution $\eta$ for $\pi$
+sequentially. Let $\vec{x_i} = (x_1, \ldots, x_i)$, for $1 \leq i \leq d$. By
+the general product rule,
+\[
+  \pi(\vec{x}) 
+  = \pi(x_1) \pi(x_2 \mid x_1) \cdots
+    \pi(x_{d-1} \mid \vec{x_{d-2}}) \pi(x_d \mid \vec{x_{d-1}}).
+\]
+This decomposition is natural in many contexts, particularly for on-line
+estimation. For example, in a stateful model like an \gls{HMM}, $x_i$ may
+represent the state at time $i$, with $\pi(\vec{x})$ being the posterior
+distribution over possible paths. The importance distribution $\eta$ for $\pi$
+will be constructed using a similar decomposition,
+\[
+  \eta(\vec{x}) 
+  = \eta(x_1) \eta(x_2 \mid x_1) \cdots
+    \eta(x_{d-1} \mid \vec{x_{d-2}}) \eta(x_d \mid \vec{x_{d-1}}).
+\]
+The importance weights for $\eta$ with respect to $\pi$ are then defined
+recursively as
+\begin{align}
+  \label{eq:sisw}
+  w(\vec{x_i}) = \frac{\pi(\vec{x_i})}{\eta(\vec{x_i})}
+  = \frac{\pi(x_i \mid \vec{x_{i-1}})\pi(\vec{x_{i-1}})}
+         {\eta(x_i \mid \vec{x_{i-1}})\eta(\vec{x_{i-1}})}
+  = \frac{\pi(x_i \mid \vec{x_{i-1}})}
+         {\eta(x_i \mid \vec{x_{i-1}})}\cdot w(\vec{x_{i-1}}).
+\end{align}
+Thus, we can choose $\eta(x_i \mid \vec{x_{i-1}})$ such that the variance of
+the importance weights is as small as possible at every step, eventually
+arriving at a full importance distribution. The process for choosing $\eta$
+will differ according to the problem at hand. One potential choice for
+$\eta(x_i \mid \vec{x_{i-1}})$ is simply $\pi(x_i \mid \vec{x_{i-1}})$, if it
+is possible to compute. For example, in the case of an \gls{HMM}, $\pi(x_i \mid
+\vec{x_{i-1}})$ is simply the transition probability from $x_{i-1}$ to $x_i$.
+In a Bayesian setting, $\eta$ may be taken to be the prior distribution. In
+some applications, it is possible to compute or approximate the form of $\eta$
+which minimizes the variance of the weights. The resulting importance
+distribution is known as the \defn{optimal
+kernel}~\autocite{cappe2007overview}. Choosing each element of the
+decomposition of $\eta$ is a problem-specific task (see \textit{e.g.}
+\autocite{smith2013sequential,liu2008monte} for many examples).
+
+The recursive definition \cref{eq:sisw} suggests an algorithm for obtaining a
+sample from $\pi$. We begin with a sample of $n$ ``particles'' which have been
+sampled from the importance distribution $\eta(x_0)$ for $\pi(x_0)$. The
+particles are updated and reweighted $d$ times, corresponding to the $d$
+elements of the decomposition of $\pi$. At the $i$th step, each particle is
+extended to include $x_i$ drawn from the chosen $\eta(x_i \mid \vec{x_{i-1}})$,
+and the importance weights are recalculated and normalized.
+
+Of course, $\eta$ is merely an approximation to $\pi$, and may be a fairly poor
+one depending on the application. Try as we might to keep the variances of the
+weights low, the cumulative errors at each sequential step tend to push many of
+the weights to very low values. This results in a poor approximation to $\pi$,
+since only a few particles retain high importance weights after all $d$
+sequential steps. To mitigate this problem, we periodically apply a resampling
+step when the variance in the importance weights becomes too high. Several
+different criteria have been proposed for when to resample, but we focus here
+on the one described by \textcite{liu2008monte}, namely the decay of the
+\gls{ESS} below a prescribed threshold. The \gls{ESS} of the population of
+particles is defined as
+\[
+  \ESS(w) = \frac{n}{1 + \Var(w)},
+\]
+where $n$ is the number of particles~\autocite{liu2008monte}. When the
+\gls{ESS} drops below the threshold, we resample the particles according to
+their weights. This results in the removal of low-weight particles from the
+population, and also equalizes all the weights. Various resampling strategies
+beyond the basic sampling with replacement have been proposed, but we will not
+discuss those here. The full \gls{SIS} algorithm with resampling is given as
+\cref{alg:sis}. We use a superscript $^{(k)}$ to refer to the $k$th particle,
+and for brevity write $w$ for $w(\vec{x_i})$. The variable $T$ is the threshold
+for the \gls{ESS}, below which the population is resampled. It is conventional
+to take $T = n/2$.
+
+\begin{algorithm}
+  \caption{Sequential importance sampling.}
+  \begin{algorithmic}
+    \For {$k = 1$ to $n$}
+      \State Sample $x_0^{(k)}$ from $\eta(x_0)$
+      \Comment{Initialize the $k$th particle}
+      \State $w^{(k)} \gets \dfrac{\pi\left(x_0^{(k)}\right)}{\eta\left(x_0^{(k)}\right)}$
+    \EndFor
+    \For {$i = 1$ to $d$}
+      \For {$k = 1$ to $n$}
+        \State Sample $x_i^{(k)}$ from $\eta\left(x_i \mid \vec{x_{i-1}^{(k)}}\right)$
+        \Comment Extend the $k$th particle
+        \State $w^{(k)} \gets \dfrac{\pi\left(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}\right)}{\eta\left(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}\right)} \cdot w^{(k)}$
+      \EndFor
+      \State Normalize the weights so that $\sum w = 1$
+      \If {$\ESS(w) < T$}
+        \State Resample the particles according to $w$
+        \For {$k = 1$ to $n$}
+          \State $w^{(k)} \gets 1/n$
+        \EndFor
+      \EndIf
+    \EndFor
+    \State Sample $n$ particles with probabilities $w$
+  \end{algorithmic}
+  \label{alg:sis}
+\end{algorithm}
+
+\subsection{Sequential Monte Carlo samplers}
+
+The \gls{SIS} algorithm described above aims to sample from a high-dimensional
+distribution $\pi(x)$, by sequentially sampling from $d$ distributions of lower
+but increasing dimension. \textcite{del2006sequential} developed an
+\defn{\gls{SMC} sampler} with an alternative objective: to sample sequentially
+from $d$ distributions $\pi_1, \ldots, \pi_d$, all of the same dimension and
+defined on the same space. The $\pi_i$ are assumed to form a related sequence,
+such as posterior distributions attained by sequentially considering new
+evidence. As with \gls{SIS}, we assume that $\pi_i = \gamma_i / Z_i$, where
+$\gamma_i$ is known pointwise and the normalizing constant $Z_i$ is unknown.
+
+Both algorithms follow the same general pattern: the target distribution is
+attained by progressing through a sequence of intermediate distributions. For
+\gls{SIS}, the intermediate distributions are lower-dimensional marginals of
+the target, while for the \gls{SMC} sampler, they are of the same dimension and
+constitute a smooth progression from an initial distribution to the target. In
+both cases, the neighbouring distributions in the sequence are related to each
+other in some way, and we can take advantage of that relationship to create a
+sequence of importance distributions alongside the sequence of targets. In
+\gls{SIS}, the neighbouring marginals $\pi(\vec{x_i})$ and
+$\pi(\vec{x_{i+1}})$ were related by the conditional density $\pi(x_i \mid
+\vec{x_{i-1}})$, which we used to inform the importance distribution.
+
+The sequence of targets considered by the \gls{SMC} sampler are assumed to be
+closely enough related, that we can shift an importance distribution for
+$\pi_i$ into one for $\pi_{i+1}$ by perturbing the particles in a systematic
+way.
+
 \section{Approximate Bayesian computation}
 \label{sec:abc}
+
 \glsreset{ABC}
@@
 clarity, we will restrict our attention to the case of one-dimensional data and
-outcomes where each $x_i$ and $y_i$ is a real number. The linear model
+outcomes where each $x_i$ and $y_i$ is a single real number. The linear model
 postulates that the outcomes are linearly related to the data, modulo some
@@
 ``$\sim$'' means ``is distributed as''. We can see from this formulation that
-the model is parametric, with parameters $\beta$ and $\sigma$. Moreover, we can
-write down the probability density of observing outcome $y_i$ given the
-parameters,
+the model is parametric, with parameters $\theta$ = ($\beta$, $\sigma$).
+Moreover, we can write down the probability density of observing outcome $y_i$
+given the parameters,
 \[
@@
 ``$\Pr$'' to refer to a probability \emph{density}, rather than a probability.
-Assuming all the $y_i$ are independent, the probability density of the entire
-observed set of outcomes is the product of the probability density of each
-individual $y_i$,
+Also, we are treating the $x_i$ as fixed quantities, and therefore have not
+conditioned the probability density on $\vec{x}$. Assuming all the $y_i$ are
+independent, the probability density of the entire observed set of outcomes is
+the product of the probability density of each individual $y_i$,
 \[
@@
 optimize the likelihood are known as the \textit{\gls{ML}} estimates, denoted
-$\hat{\theta}$. \Gls{ML} estimation is usually performed with numerical
-optimization. In the simplest terms, many possible values for $\theta$ are
-examined, $\L(\theta \mid \vec{y})$ is calculated for each, and the parameters
-which produce the highest value are accepted. Many sophisticated numerical
-optimization methods exist, although they may not be guaranteed to find the
-true \gls{ML} estimates if the likelihood function is complex. Occasionally, as
-in the case of least squares, the \gls{ML} estimates can be found explicitly by
-setting the likelihood function's derivatives to zero.
+$\hat{\theta}$. That is,
+\[
+  \hat{\theta} = \argmax_\theta\; \Pr(\vec{y} \mid \theta).
+\]
+\Gls{ML} estimation is usually performed with numerical optimization. In the
+simplest terms, many possible values for $\theta$ are examined, $\L(\theta \mid
+\vec{y})$ is calculated for each, and the parameters which produce the highest
+value are accepted. Many sophisticated numerical optimization methods exist,
+although they may not be guaranteed to find the true \gls{ML} estimates if the
+likelihood function is complex. Occasionally, as is the case with linear
+regression, the \gls{ML} estimates can be found explicitly by setting the
+likelihood function's derivatives to zero.
 
@@
 model parameters $\theta$. However, it is frequently the case that the
-investigator has some additional information or \defn{prior belief} about what
-$\theta$ are likely to be. For example, in the linear regression case, the
-instrument used to measure the outcomes may have a well-known margin of error,
-or the sign of the slope may be obvious from previous experiments. The Bayesian
-approach to model fitting makes use of this information by codifying the
-investigator's beliefs as a \defn{prior distribution} on the parameters,
-denoted $\Pr(\theta)$. Instead of considering only the likelihood, Bayesian
-inference focuses on the product of the likelihood and the prior, $\Pr(\vec{y}
-\mid \theta) \Pr(\theta)$. Bayes' theorem tells us that this product is related
-to the \textit{posterior distribution} on $\theta$,
+investigator has some additional information or belief about what $\theta$ are
+likely to be. For example, in the linear regression case, the instrument used
+to measure the outcomes may have a well-known margin of error, or the sign of
+the slope may be obvious from previous experiments. The Bayesian approach to
+model fitting makes use of this information by codifying the investigator's
+beliefs as a \defn{prior distribution} on the parameters, denoted
+$\Pr(\theta)$. Instead of considering only the likelihood, Bayesian inference
+focuses on the product of the likelihood and the prior, $\Pr(\vec{y} \mid
+\theta) \Pr(\theta)$. Bayes' theorem tells us that this product is related to
+the \textit{posterior distribution} on $\theta$,
 \[
@@
 which we can calculate statistics of interest. Actually performing this random
-walk is straightforward and is accomplished via the Metropolis-Hastings
+walk is straightforward and can be accomplished via the Metropolis-Hastings
 algorithm (\cref{alg:mh}).
@@
 \begin{algorithm}
-  \label{alg:mh}
   \caption{Metropolis-Hastings algorithm for Markov chain Monte Carlo.}
@@
 
-\subsection{Overview and motivation for ABC}
+\subsection{Overview of ABC}
 \label{subsec:abcoverview}
@@
 models are particularly prevalent in population
-genetics~\autocite{beaumont2002approximate, beaumont2010approximate}. We will
-begin this subsection by describing \gls{ABC} in general terms. Then, we will
-use linear regression as a toy problem to demonstrate how \gls{ABC} could be
-applied. Finally, we will give an example of a model which is impractical to
-fit using likelihood-based methods, and show how its parameters were estimated
-using \gls{ABC}.
-
-Ordinarily, Bayesian inference targets the posterior distribution.
-
-%traditional likelihood criterion
-%for model ``goodness'' with one based on simulated data generated by the model.
-%Good models should be able to simulate data which closely resemble reality. In
-%\gls{ABC}, we are no longer interested in the posterior distribution, but
-%rather in the distribution of model parameters which produce data sufficiently
-%``close'' to the real data. To formalize this, let $d(\cdot, \cdot)$ be a
-%distance function on datasets, and $\varepsilon$ be a tolerance level.
-
-Of course, linear models can be fit more easily using likelihood-based methods, 
-since the likelihood $\L(\theta \mid \vec{y})$ can be easily calculated (see
-\cref{subsec:fit}). One model where this is not the case is Kingman's
-coalescent~\autocite{kingman1982coalescent}.
-
-As an example of such a model, we will review the first practical application
-of \gls{ABC} by \textcite{tavare1997inferring}. The aim of the investigation
-was to calculate the posterior distribution on the coalescence time $t$ of some
-observed sequences $D$ under the coalescent model
-\autocite{kingman1982coalescent} with the infinite-sites assumption
-\autocite{watterson1975number}. Their estimation method relied on two key
-observations. First, the number $k$ of \defn{segregating sites} in the data is
-informative about the coalescence time while being much easier to work with
-than the data itself. A segregating site is a position which is polymorphic in
-the observed sequences. Using this observation, the authors replaced their
-posterior distribution of interest, $\Pr(t \mid D)$, by
-\[
-  \Pr(t \mid k) \propto \Pr(t) \Pr(k \mid t),
-\]
-where again we abuse $\Pr$ to refer to a probability density. Second,
-the authors observed that the expected number of segregating sites depends on
-the total branch length $l$ of the tree, rather than its height. This enabled
-them to write an alternative expression for the product of the likelihood and
-prior in terms of $l$,
-\[
-  \Pr(t) \Pr(k \mid t) = \int_0^\infty \Pr(t, l) \Pr(k \mid l) \d l.
-\]
-While this integral may be difficult to evaluate analytically, the power of
-this formulation is the fact that it is straightforward to \emph{simulate}
-values of $t$ and $l$ according to $\Pr(t, l)$, their joint probability under
-the coalescent model. One simply runs the (stochastic) coalescent process and
-observing the values $\hat{t}$ and $\hat{l}$ which result. It is
-straightforward to calculate $\Pr(k \mid \hat{l})$, the probability
-that $k$ segregating sites were observed in a tree with total branch length
-$\hat{l}$, using the Poisson distribution. Hence if $\hat{l}$ and $\hat{t}$,
-having been generated according to $\Pr(t, l)$, are sampled with probability
-proportional to $\Pr(k \mid \hat{l})$, they represent an unbiased sample from
-the above integral. Taking a large number of such samples allowed the authors
-to obtain a Monte Carlo estimate of $\Pr(t \mid k)$.
-
-%\Gls{ABC} is a statistical method designed to fit complex models, which cannot
-%be fit using more conventional methods, to observed
-%data~\autocite{marin2012approximate, sunnaker2013approximate,
-%beaumont2010approximate}. We shall make this precise below, but to fully
-%describe and motivate \gls{ABC}, it is necessary to first explain exactly what
-%we mean by ``fitting'' a model. We will illustrate this concept with one of the
-%simplest and most ubiquitous models: linear regression.
-%
-%An approximate Bayesian computation approach to least-squares regression might
-%be as follows. We will need to make an additional assumption: that the errors
-%$\varepsilon_i$ are normally distributed. That is, our linear model is now of
-%the form
+genetics~\autocite{beaumont2002approximate, beaumont2010approximate}.
+
+Ordinarily, Bayesian inference targets the posterior distribution $\Pr(\theta
+\mid \vec{y})$. That is, in the Bayesian framework, model parameters with
+higher posterior density are ``better'' in the sense that they offer a more
+credible explanation for the observed data. Approximate Bayesian computation
+offers an alternative metric for parameter credibility, namely the similarity
+simulated datasets to the observed data. If datasets simulated under the model
+closely resemble the real data, it follows that the model is a reasonable
+approximation to the real-world process generating the observed data. More
+formally, suppose we have a distance measure $\rho$ defined on the space of all
+possible data our model could generate. \gls{ABC} aims to sample from the joint
+posterior distribution of model parameters and simulated datasets $\vec{z}$
+which are within some small distance $\varepsilon$ of the observed data
+$\vec{y}$,
+\[
+  \Pr(\theta, \vec{z} \mid \vec{y}, \varepsilon) =
+  \frac{\Pr(\theta) \Pr(\vec{z} \mid \theta) \I_{A_{\varepsilon, \vec{y}}} (\vec{z})}
+  {\int_{A_{\varepsilon, \vec{y}} \times \Theta} \Pr(\theta) \Pr(\vec{z} \mid \theta) \d \theta}.
+\]
+Here, $A_{\varepsilon, \vec{y}}$ is an $\varepsilon$-ball around $\vec{y}$ with
+respect to $\rho$, $\Theta$ is the space of all possible model parameters, and
+$\I$ is the indicator function~\autocite{marin2012approximate}. As we shall
+see in the next section, this distribution can be sampled from exactly. The
+word ``approximate'' derives from the assumption that, for a suitably chosen
+distance $\rho$ and a small enough $\varepsilon$, the marginal in $\vec{z}$ of
+this distribution approximates the posterior of
+interest~\autocite{marin2012approximate}. That is,
+\[
+  \int \Pr(\theta, \vec{z} \mid \vec{y}, \varepsilon) \d\vec{z}
+  \approx \Pr(\theta \mid \vec{y}).
+\]
+To emphasize that the distribution on the left only approximates the posterior,
+we will refer to it as the \textit{\gls{ABC} target distribution}. Note that in
+many formulations, the distance function $\rho$ is defined as $\rho(S(\cdot),
+S(\cdot))$ where $S$ is a function which maps data points into a vector of
+summary statistics. This can be useful if the data are high-dimensional or of a
+complex type, but it is not strictly necessary. For instance, if the data are
+numeric and of low dimension, the distance function may simply be the Euclidian
+distance~\autocite{sisson2007sequential}. For more complex data,
+\textcite{nakagome2013kernel} proposed the use of a kernel function (defined in
+\cref{subsec:treeshape}), an approach they dubbed \defn{kernel-\gls{ABC}}.
+
+\subsection{Algorithms for ABC}
+\label{subsec:abcalg}
+
+Algorithms for performing \gls{ABC} fall into one of three categories:
+rejection, \gls{MCMC}, and \gls{SMC}. Rejection ABC is the simplest method, and
+also the one which was first proposed~\autocite{rubin1984bayesianly,
+tavare1997inferring}. The algorithm, outlined in \cref{alg:abcrej}, repeats the
+following steps until a desired number of samples from the target distribution
+are obtained. Parameter values $\theta$ are sampled according to the prior
+distribution $\Pr(\theta)$. Then, a simulated dataset $\vec{z}$ is generated
+from the model with the sampled parameter values. By definition, the
+probability density of obtaining the particular dataset $\vec{z}$ is
+$\Pr(\vec{z} \mid \theta)$. Finally, the parameters are sampled if the distance
+of $\vec{z}$ from the observed data $\vec{y}$ is less than $\varepsilon$ - that
+is, with probability $\I_{A_{\varepsilon, \vec{y}}}(\vec{z})$. Putting this all
+together, the parameters $\theta$ are sampled with probability proportional to
+\[
+  \Pr(\theta) \Pr(\vec{z} \mid \theta) \I_{A_{\varepsilon, \vec{y}}}(\vec{z}),
+\]
+which is exactly the numerator of the \gls{ABC} target distribution. Thus,
+$\theta$ represents an unbiased sample from the approximate posterior.
+
+\begin{algorithm}
+  \caption{Rejection \gls{ABC}.}
+  \begin{algorithmic}
+    \Loop
+      \State Draw $\theta$ according to $\Pr(\theta)$
+      \State Simulate a dataset $\vec{z}$ from the model with parameters $\theta$
+      \If{$\rho(\vec{y}, \vec{z}) < \varepsilon$}
+        \State Sample $\theta$
+      \EndIf
+    \EndLoop
+  \end{algorithmic}
+  \label{alg:abcrej}
+\end{algorithm}
+
+Rejection \gls{ABC} is easy to understand and implement, but it is not
+generally computationally feasible. If the posterior is very different from the
+prior, a very large number of samples may need to be taken in order to find a
+simulated dataset which is close to $\vec{z}$. The inefficiency is compounded
+by the curse of dimensionality - the measure of the $\varepsilon$-ball around
+$\vec{y}$ decreases exponentially with the number of dimensions.
+\gls{ABC}-\gls{MCMC} (\cref{alg:abcmcmc}) was designed to overcome these
+hurdles~\autocite{marjoram2003markov}. The approach is similar to ordinary
+Bayesian \gls{MCMC} (\cref{subsec:mfit}), except that a distance cutoff
+replaces the likelihood ratio. That is, the transition probability between
+states $\theta_1$ and $\theta_2$, with simulated datasets $\vec{z_1}$ and
+$\vec{z_2}$, is defined as
+\[
+  \max\left(1, \frac{\Pr(\vec{z_2} \mid \theta_2) q(\theta_2, \theta_1)}
+                    {\Pr(\vec{z_1} \mid \theta_1) q(\theta_1, \theta_2)} 
+    \cdot \I_{A_{\varepsilon, \vec{y}}}(\vec{z_2}) \right).
+\]
+
+\begin{algorithm}
+  \caption{ABC-MCMC.}
+  \begin{algorithmic}
+    \State Draw $\theta$ according to $\Pr(\theta)$
+    \Loop
+      \State Propose $\theta'$ according to $q(\theta, \theta')$
+      \State Simulate a dataset $\vec{z'}$ according to the model with
+             parameters $\theta$
+      \State Accept $\theta \gets \theta'$ with probability
+      $\max \left( 1, 
+       \dfrac{\Pr(\theta') q(\theta', \theta)}
+             {\Pr(\theta\phantom{'}) q(\theta, \theta')} 
+       \cdot \I_{A_{\varepsilon, \vec{y}}}(\vec{z'}) \right)$
+    \EndLoop
+  \end{algorithmic}
+  \label{alg:abcmcmc}
+\end{algorithm}
+
+Some of the same computational inefficiencies arise with \gls{ABC}-\gls{MCMC}
+as with rejection. For example, in regions of low posterior density, the
+probability to simulate a dataset proximal to the observed data is low. Various
+strategies have been developed to mitigate this, including reducing the
+tolerance level $\varepsilon$ as the chain
+progresses~\autocite{ratmann2007using}.
+
+The most recently developed class of algorithm for \gls{ABC} is
+\gls{ABC}-\gls{SMC}~\autocite{sisson2007sequential, 2009adaptive}. As with
+\gls{ABC}-\gls{MCMC}, the algorithm is a straightforward modification of an
+existing Bayesian inference method, in this case \gls{SMC} (\cref{sec:smc}).
+
+%As an example of such a model, we will review the first practical application
+%of \gls{ABC} by \textcite{tavare1997inferring}. The aim of the investigation
+%was to calculate the posterior distribution on the coalescence time $t$ of some
+%observed sequences $D$ under the coalescent model
+%\autocite{kingman1982coalescent} with the infinite-sites assumption
+%\autocite{watterson1975number}. Their estimation method relied on two key
+%observations. First, the number $k$ of \defn{segregating sites} in the data is
+%informative about the coalescence time while being much easier to work with
+%than the data itself. A segregating site is a position which is polymorphic in
+%the observed sequences. Using this observation, the authors replaced their
+%posterior distribution of interest, $\Pr(t \mid D)$, by
 %\[
-%  y_i = bx_i + \N(0, \sigma),
+%  \Pr(t \mid k) \propto \Pr(t) \Pr(k \mid t),
 %\]
-%where $\N(0, \sigma)$ indicates a normal distribution with mean zero and
-%variance $\sigma$. We do not know $\sigma$ in advance, so this is another
-%parameter we will need to estimate. There is nothing special about normal
-%distributions - we could have chosen any other distribution we deemed
-%appropriate. 
-%
-
-\subsection{Algorithms for ABC}
-\label{subsubsec:abcalg}
-
-\Gls{ABC} does not refer to a particular procedure for model fitting. Rather,
-\gls{ABC} refers to the general strategy of choosing model parameters based on
-the resulting model's propensity to generate data resembling the real data.
-There are three main classes of \gls{ABC} algorithm which have been developed
-so far: rejection, \gls{MCMC}, and \gls{SMC}.
-
-All of these approaches require some common elements. First, as with all
-Bayesian methods, we are required to specify a \defn{prior distribution},
-denoted $\pi$, on the parameter space. The prior specifies what we already know
-or believe about the model parameters. Second, in order to compare simulated to
-observed data, we need to be able to summarize a data set in a numerical
-format. This is accomplished by a function, denoted $\eta$, which computes a
-vector of hopefully informative summary statistics on a data set. Third, we
-need a distance function $\rho$ which tells us how similar two data sets are to
-each other, based on their summary statistics.
-
-Continuing with the linear model example, we need to specify a prior $\pi(a, b,
-\sigma)$ on the three parameters. We do not have much certain information about
-these parameters except that $\sigma$ has to be at least zero, but it seems
-reasonable to assume that extreme relationships are fairly rare, and that
-positive and negative correlations are equiprobable. Therefore, we will let
-$a$, $b$, and $\sigma$ be independent, $a$ and $b$ be normally distributed,
-and $\sigma$ be log-normally distributed. That is,
-\[
-  \pi(a, b, \sigma) = 
-  \begin{cases}
-    \Pr[\N(0, 1) = a] \cdot \Pr[\N(0, 1) = b] \cdot \Pr[\N(0, 1) = \log\sigma]
-     & \sigma > 0 \\
-    0 & \sigma \le 0.
-  \end{cases}
-\]
-For the vector of summary statistics, we will use the mean and variance of the
-simulated data,
-\[
-  \eta(\vec{y}) = \left<\E[\vec{y}], \Var[\vec{y}]\right>.
-\]
-Finally, for the distance function $\rho$ we take the standard Euclidean
-distance.
-
-Rejection ABC is the simplest method, and also the one which was first
-proposed~\autocite{rubin1984bayesianly}. Effectively, it comes down to the
-approach described in the previous subsection of guessing parameter values
-until one is close enough to the truth. More specifically, a possible set of
-parameters $\theta$ is drawn from the prior distribution, and a simulated data
-set $z$ is generated from the model with those parameters. If the distance
-between the simulated data set and the real data, $\rho(\eta(y), \eta(z))$, is
-small enough, then we accept $\theta$ as a sample from the posterior. This can
-be repeated until as many samples as desired are obtained.
-
-% MCMC
-The second method is \gls{ABC}-\gls{MCMC}. This is similar to ordinary Bayesian
-\gls{MCMC}, except that a ratio of distances to the observed data replaces the
-likelihood ratio. The algorithm begins by sampling a single vector of parameter
-values $\theta$ from the prior distribution $\pi(\theta)$. It then proceeds
-iteratively: a new parameter vector $\theta^*$ is chosen according to a
-proposal distribution $q(\theta^* \mid \theta)$. The proposal distribution $q$
-is often taken to be a Gaussian centred at $\theta$. Then $\theta^*$ is
-accepted as the new $\theta$ with probability
-\[
-  \max\left(1, \text{TODO} \right),
-\]
-or discarded otherwise. This process is iterated until some stopping criterion
-is reached, typically a simple limit on the number of steps. After some initial
-number of iterations, known as \defn{burn-in}, parameters $\theta$ are
-routinely sampled. Since points in parameter space are visited in proportion to
-their posterior probability, these samples can be taken to approximate the
-posterior distribution on $\theta$, and can be used to calculate point
-estimates and confidence intervals.
-
-The most recently developed class of algorithms for ABC is sequential
-Monte-Carlo (SMC)~\autocite{sisson2007sequential}. As with the other two
-classes, we want to eventually obtain a sample from the posterior distribution
-$f(\theta \mid D)$. The idea of SMC is to begin with a sample from a
-distribution we know, most often the prior, and approach the posterior smoothly
-by progressing through a series of intermediate distributions.
-
-\section{Sequential Monte Carlo}
-
-\Gls{SMC} is a statistical inference method which samples from a sequence of
-probability distributions in a fixed order~\autocite{del2006sequential}. This
-idea of \defn{sequential sampling} is useful several contexts, but we consider
-here its application to model fitting in a Bayesian setting.
-
-\texttt{TODO: move the basics of model fitting (likelihoods and priors) to here}
-
-Our objective is to obtain samples from, and summary statistics of, the
-posterior distribution on the model of interest's parameters. \gls{SMC} can be
-applied in this context by defining a sequence of distributions, starting from
-the posterior and ending at the prior, which constitute a ``smooth''
-trajectory. The main \gls{SMC} algorithm for this problem was developed by
-\textcite{del2006sequential}, based on an existing methodology called
-\gls{SIS}. \gls{SIS} is also designed to sample from a sequence of
-distributions, but rather than all being defined on the same space (such as
-model parameters), they are defined on nested spaces of increasing dimension.
-Following \textcite{del2006sequential} and other reviews of
-\gls{SMC}~\autocite{doucet2001introduction}, we will begin this section by
-describing \gls{SIS}, and then turn to the adaptation of the method to the case
-when the sequence of distributions are all defined on the same space. Note that
-\textcite{del2006sequential} use the variable $\pi$ for the target
-distributions, but \textcite{doucet2001introduction} use $\pi$ to indicate the
-importance distributions. We follow the former and denote target distributions
-by $\pi$ and importance distributions by $\eta$.
-
-\subsection{Sequential importance sampling}
-
-The basis of \gls{SIS} is \gls{IS}, which is a method of estimating summary
-statistics of distributions which are known only up to a normalizing constant,
-and therefore cannot be sampled from directly. That is, if $\pi$ is such a
-distribution and $f$ is any real-valued function, \gls{IS} is concerned with
-estimating
-\[
-  \pi(f) = \int f(x)\pi(x)\d x = \int f(x) \frac{\gamma(x)}{Z} \d x,
-\]
-where the integral is over the space on which $\pi$ is defined, $\gamma(x)$ is
-known, and $Z$ is the unknown normalizing constant, $Z = \int \gamma(x) \d x$.
-The posterior distributions of all but the simplest models' parameters fall
-into this category. Suppose we have at hand another distribution $\eta$, called
-the \defn{importance distribution}, from which we are able to sample. Define
-the \defn{importance weight} as the ratio ratio $w(x) = \gamma(x)/\eta(x)$. We
-can express the normalizing constant $Z$ in terms of the importance weight and
-distribution, $Z = \int w(x) \eta(x) \d x$, and in turn write the original
-integral of interest as
-\[
-  \int f(x) \pi(x) \d x = \frac{\int f(x) \gamma(x) \d x}
-                               {\int w(x) \eta(x) \d x}.
-\]
-If we sample a large number of points from $\eta$, then $\eta(x)$ can be
-approximated at each of them by an empirical distribution. Since the remaining
-quantities $f$, $\gamma$, and $w$ can all be evaluated pointwise, these are all
-the ingredients we need to obtain an estimate of $\pi(f)$. Although this is a
-simple and elegant approach, the drawback is that the variance of the estimate
-is proportional to the variance of the importance
-weights~\autocite{liu2008monte}, which may be quite large if $\eta$ and
-$\gamma$ are very different. Therefore, the practical use of \gls{IS} on its
-own is limited, since it depends on finding an importance distribution which is
-similar to $\pi$, which we usually know very little about \textit{a priory}.
-
-However, an ideal context for the use of \gls{IS} is when we want to sample
-from a sequence of nested probability distributions $\pi_1, \ldots, \pi_n$.
-By \defn{nested}, we mean that $\pi_{i+1}$ is defined on a space of dimension
-$i+1$ and admits $\pi_i$ as a marginal. That is,
-\[
-  \pi_{i+1}(x_1, \ldots, x_{i+1}) = \pi_i(x_1, \ldots, x_i) f_{i+1}(x_1, \ldots, x_{i+1}),
-\]
-where $f_{i+1}$ is a function which yields a distribution when multiplied with
-$\pi_i$. This situation may seem somewhat contrived, but it arises naturally
-when trying to infer the hidden sequence of parameters of a stateful model. For
-example, \textcite{doucet2001introduction} discuss the case when $\pi_i$ is the
-posterior distribution over the first $i-1$ states of a \gls{HMM}, conditional
-on the observed data up to time $i-i$. We assume that either $\pi_1$ is known
-explicitly, or we have enough information about it to find an adequate
-importance distribution. In the \gls{HMM} example, $\pi_1$ was taken to be the
-prior distribution on the starting state.
+%where again we abuse $\Pr$ to refer to a probability density. Second,
+%the authors observed that the expected number of segregating sites depends on
+%the total branch length $l$ of the tree, rather than its height. This enabled
+%them to write an alternative expression for the product of the likelihood and
+%prior in terms of $l$,
+%\[
+%  \Pr(t) \Pr(k \mid t) = \int_0^\infty \Pr(t, l) \Pr(k \mid l) \d l.
+%\]
+%While this integral may be difficult to evaluate analytically, the power of
+%this formulation is the fact that it is straightforward to \emph{simulate}
+%values of $t$ and $l$ according to $\Pr(t, l)$, their joint probability under
+%the coalescent model. One simply runs the (stochastic) coalescent process and
+%observing the values $\hat{t}$ and $\hat{l}$ which result. It is
+%straightforward to calculate $\Pr(k \mid \hat{l})$, the probability
+%that $k$ segregating sites were observed in a tree with total branch length
+%$\hat{l}$, using the Poisson distribution. Hence if $\hat{l}$ and $\hat{t}$,
+%having been generated according to $\Pr(t, l)$, are sampled with probability
+%proportional to $\Pr(k \mid \hat{l})$, they represent an unbiased sample from
+%the above integral. Taking a large number of such samples allowed the authors
+%to obtain a Monte Carlo estimate of $\Pr(t \mid k)$.

### 2016-04-12 doc/thesis/methods.tex
More writing.

@@
 
-I wrote a computer program called \software{netabc} to perform statistical
-inference of contact network parameters from an estimated transmission tree
-using kernel-\gls{ABC}. The program combines three major components: Gillespie
+\software{Netabc} is a computer program to perform statistical inference of
+contact network parameters from an estimated transmission tree using
+kernel-\gls{ABC}. The program combines three major components: Gillespie
 simulation, to simulate transmission trees on contact networks; the
@@
 this algorithm is given in \cref{fig:abcsmc}.
-
-The generation of contact networks from the model of interest, though critical
-to the program's function, is not included here as a component of
-\software{netabc}.  Generation of networks is a model-specific task, whereas
-\software{netabc} is intended to be generic and work with any network model.
-For ease of use, generators for three basic network models, namely the
-\gls{BA}, \gls{ER}, and \gls{WS} models, are included in the program through
-the \software{igraph} library~\autocite{csardi2006igraph}. To investigate other
-models, additional graph generators will need to be added into the program.
 
@@
 independently implemented and applied by several
-authors~\autocite[\textit{e.g.}][]{o2010contact, robinson2013dynamics,
+authors~\autocite[\textit{e.g.}][]{o2011contact, robinson2013dynamics,
 leventhal2012inferring, groendyke2011bayesian, villandre2016assessment}.
@@
 The tree kernel developed by \textcite{poon2013mapping} provides a
-comprehensive similarity score between two phylogenetic trees. The kernel
-computes the dot-product of two feature vectors, corresponding to the two
-trees, in the infinite-dimensional feature space of all possible subset trees
-with branch lengths (see \cref{subsec:treeshape}). The kernel was implemented
-using the fast algorithm developed by \textcite{moschitti2006making}. First,
-the number of leaf children of each node, also known as its \defn{production
-rule}, is recorded. The nodes of both trees are ordered by production rule, and
-a list of pairs of nodes sharing the same production rule is created. These are
-the nodes for which the value of the tree kernel must be computed - all other
-pairs have a value of zero. The pairs to be compared are then re-ordered so
-that the child nodes are always evaluated before their parents. Due to its
-recursive definition, ordering the pairs in this way allows the tree kernel to
-be computed by dynamic programming. The complexity of this implementation is
-$O(|T_1||T_2|)$, where $|T|$ counts the number of nodes in the tree $T$.
+comprehensive similarity score between two phylogenetic trees, via the
+dot-product of the two trees' feature vectors in the infinite-dimensional space
+of all possible subset trees with branch lengths (see \cref{subsec:treeshape}).
+The kernel was implemented using the fast algorithm developed by
+\textcite{moschitti2006making}. First, the number of leaf children of each
+node, also known as its \defn{production rule}, is recorded. The nodes of both
+trees are ordered by production rule, and a list of pairs of nodes sharing the
+same production rule is created. These are the nodes for which the value of the
+tree kernel must be computed - all other pairs have a value of zero. The pairs
+to be compared are then re-ordered so that the child nodes are always evaluated
+before their parents. Due to its recursive definition, ordering the pairs in
+this way allows the tree kernel to be computed by dynamic programming. The
+complexity of this implementation is $O(|T_1||T_2|)$, where $|T|$ counts the
+number of nodes in the tree $T$.
 
 \subsubsection*{Adaptive sequential Monte Carlo for Approximate Bayesian computation}
+\label{subsubsec:adaptsmc}
 
 I implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
-\textcite{del2012adaptive}.
+\textcite{del2012adaptive}. This algorithm is similar to the reference
+\gls{ABC}-\gls{SMC} algorithm described in \cref{subsec:abcalg}, except that
+the sequence of tolerances $\varepsilon_i$ is automatically determined rather
+than specified by the user. The tolerances are chosen such that the \gls{ESS}
+of the particle population, which indicates the quality of the Monte Carlo
+approximation (see \cref{subsec:sis}), decays at a controlled rate. This
+prevents the Monte Carlo approxmiation from collapsing by the variance in
+weights growing too large. A single parameter $\alpha$ (not to be confused with
+the \gls{BA} model parameter) controls the decay rate, with $\varepsilon_i$
+being chosen to satisfy
+\[
+  \ESS(w_i) = \alpha \ESS(w_{i-1}).
+\]
+Here, $w_i$ is the vector of weights at the $i$th step. Note that, since $w_i$
+depends on $\varepsilon_i$, this equation solves for the updated weights and
+the updated tolerance simultaneously. As pointed out by
+\textcite{del2012adaptive}, the equation has no analytic solution, but can be
+solved by bisection. As in the reference version, the particles are perturbed
+by applying an \gls{MCMC} kernel, and resampling occurs when the \gls{ESS}
+drops below a prescribed threshold. In my implementation, following all
+proposals of the \gls{MCMC} kernel are Gaussian with variance equal to twice
+the empirical variance of the
+particles~\autocite{beaumont2009adaptive,del2012adaptive}.
 
-For ease of exposition, we simplify the notation of \textcite{del2012adaptive}
-by dropping the subscripts on the variables which indicate the current
-iteration number. Instead, we will add a prime $'$ to indicate a value which
-will be used in the next iteration (this should become clear later on).
-
-In the algorithm, we keep track of a population of $n$ sets of model
-parameters, called \defn{particles}, denoted $\set{X_i}_{i=1}^{n}$. For the
-\gls{BA} model, the particles would be 4-tuples (\gls{N}, \gls{I}, \gls{m},
-\gls{alpha}). Each particle $X_i$ is associated with a set of \gls{M} simulated
-datasets, denoted $\set{X_{i,j}}_{j=1}^M$, and a weight $W_i$.
-
-The particles are initially drawn from the prior distribution.
-
-Let $d$ be a
-distance measure on data sets, so that $d(x, y)$ is smaller the more similar
-$x$ and $y$ are to each other. Let $\varepsilon$ be the tolerance level which
-indicates whether a data set is ``close'' to the observed data. That is, if
-$d(x, y) < \varepsilon$, we will say that $x$ and $y$ are close, otherwise they
-are distant.
-
-% begin by sampling from the prior and setting all weights to equal
-
-Next, we calculate the next tolerance $\varepsilon^*$. Before we explain how
-this is done, we first need to define how we adjust the weights on the
-particles. As explained above (see subsection~\ref{subsubsec:abcalg}), the idea
-of sequential Monte-Carlo is to begin with the prior distribution $\pi$,
-progress smoothly through a series of intermediate distributions $\pi_1,
-\ldots, \pi_{n-1}$, and eventually arrive at the target posterior distribution
-$\pi_n$. In the $k$th iteration, the distribution $\pi_k$ is approximated by
-the particles and their weights.
-
-In contrast to most existing sequential Monte-Carlo methods, this
-algorithm does not require the user to specify a sequence of decreasing
-tolerances to approach the target posterior distribution. Rather, the
-tolerances are computed adaptively at each step, starting from infinity at the
-first iteration.
-
-The algorithm may be stopped when the tolerance reaches a user-defined final
-value, or when the rate of acceptance of the \gls{MH} kernel reaches a
-user-defined threshold.
+The algorithm may be stopped when one of two termination conditions is reached.
+The user may specify a final tolerance $\varepsilon$, or a final acceptance
+rate of the \gls{MCMC} kernel. The latter stops the algorithm when the
+particles are not moving around very much, implying little change in the
+estimated target.
 
@@
 experiment was repeated. The parameters of the tree kernel, $\lambda$ and
-$\sigma$ (\cref{subsubsec:treeshape}) will be referred to as
+$\sigma$ (\cref{subsec:treeshape}) will be referred to as
 \defn{meta-parameters} to distinguish them from the parameters of the \gls{BA}
-model. With the exception of our own binaries, all analyses were done in
+model. With the exception of our own programs, all analyses were done in
 \software{R}, and all packages listed below are \software{R} packages.
@@
 simulated tree using our \software{treestat} binary. The trees were visualized
-using the \software{ape} package. Our \software{treekernel} binary was used to
-calculate the value of the kernel for each pair of trees, with the
-meta-parameters set to $\lambda = 0.3$ and $\sigma = 4$. These values were
-stored in a symmetric 300 $\times$ 300 kernel matrix. Similarly, we computed
-the \gls{nltt} statistic between each pair of trees using our
-\software{treestat} binary, and stored them in a second $300 \times 300$
-matrix.
+using the \software{ape} package~\autocite{paradis2004ape}. Our
+\software{treekernel} binary was used to calculate the value of the kernel for
+each pair of trees, with the meta-parameters set to $\lambda = 0.3$ and $\sigma
+= 4$. These values were stored in a symmetric 300 $\times$ 300 kernel matrix.
+Similarly, we computed the \gls{nltt} statistic between each pair of trees
+using our \software{treestat} binary, and stored them in a second $300 \times
+300$ matrix.
 
@@
 classifiers for \gls{alpha} based on three statistics. First, we used the
-\software{kernlab} package~\autocite{karatzoglou2004kernlab} to create a
-\gls{kSVM} classifier using the computed kernel matrix. Second, we used the
-\software{e1071} package to create an ordinary \gls{SVM} classifier using the
-pairwise \gls{nltt} matrix. Finally, we performed an ordinary linear regression
-of \gls{alpha} against Sackin's index. Each of these classifiers was evaluated
-with 1000 two-fold cross-validations. We also performed a \gls{kPCA} projection
-of the kernel matrix, and used it to visualize the separation of the different
-\gls{alpha} values in the tree kernel's feature space. A schematic of this
-experiment is presented in \cref{fig:kernelexpt}.
+\software{kernlab} package~\autocite{zeileis2004kernlab} to create a \gls{kSVM}
+classifier using the computed kernel matrix. Second, we used the
+\software{e1071} package~\autocite{meyer2015e1071} to create an ordinary
+\gls{SVM} classifier using the pairwise \gls{nltt} matrix. Finally, we
+performed an ordinary linear regression of \gls{alpha} against Sackin's index.
+Each of these classifiers was evaluated with 1000 two-fold cross-validations.
+We also performed a \gls{kPCA} projection of the kernel matrix, and used it to
+visualize the separation of the different \gls{alpha} values in the tree
+kernel's feature space. A schematic of this experiment is presented in
+\cref{fig:kernelexpt}.
 
@@
 parameters, with several test values and trees of varying sizes. The variables
-are listed in \cref{tab:gridexpt}.
+are listed in \cref{tab:gridexpt}. A graphical schematic of the grid search
+experiments is shown in \cref{fig:gridexpt}.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth]{gridsearch-expt}
+  \caption{Graphical schematic of grid search experiments used to investigate 
+    \gls{BA} model parameters.}
+  \label{fig:gridexpt}
+\end{figure}
 
@@
 were set to $\lambda = 0.3$ and $\sigma = 4$. The \gls{SMC} algorithm was run
-with 1000 particles, five sampled datasets per particle, and the parameter
-controlling the tradeoff between speed and approximation quality
-\textcite[called ``$\alpha$'' in][]{del2012adaptive} set to 0.95. The algorithm
-was stopped when the acceptance rate of the \gls{MH} kernel dropped below
-1.5\%, the same criterion used by \citeauthor{del2012adaptive}. Approximate
-marginal posterior densities for each parameter were calculated using the
+with 1000 particles, five sampled datasets per particle, and the $\alpha$
+parameter (not to be confused with the \gls{BA} preferential attachment
+parameter, see \cref{subsubsec:adaptsmc}) set to 0.95. The algorithm was
+stopped when the acceptance rate of the \gls{MH} kernel dropped below 1.5\%,
+the same criterion used by \citeauthor{del2012adaptive}. Approximate marginal
+posterior densities for each parameter were calculated using the
 \software{density} function in \software{R} applied to the final weighted
 population of particles. Credible intervals were obtained for each parameter
-using the \software{HPDinterval} function in the \software{coda} package.
+using the \software{HPDinterval} function in the \software{coda}
+package~\autocite{plummer2006coda}.
 
@@
 We downloaded all sequences associated with each published study from GenBank.
-For the \textcite{novitsky2014impact} data, each sequence was aligned pairwise
-to the HXB2 reference sequence (GenBank accession number HIVHXB2CG) and the
-hypervariable regions were clipped out with \software{BioPython} version
-1.66+~\autocite{cock2009biopython}. Sequences were multiply aligned using
-\software{MUSCLE} version 3.8.31 \autocite{edgar2004muscle}, and alignments
-were manually inspected with \software{Seaview} version 4.4.2
-\autocite{gouy2010seaview}. Phylogenies were constructed from the nucleotide
-alignments by approximate maximum likelihood using \software{FastTree2} version
-2.1.7 with the \gls{GTR} model. Transmission trees were estimated by rooting
-and time-scaling the phylogenies by root-to-tip regression, using a modified
-version of Path-O-Gen (distributed as part of
+For the \textcite{novitsky2014impact} data, each \textit{env} sequence was
+aligned pairwise to the HXB2 reference sequence (GenBank accession number
+HIVHXB2CG) and the hypervariable regions were clipped out with
+\software{BioPython} version 1.66+~\autocite{cock2009biopython}. Sequences were
+multiply aligned using \software{MUSCLE} version 3.8.31
+\autocite{edgar2004muscle}, and alignments were manually inspected with
+\software{Seaview} version 4.4.2 \autocite{gouy2010seaview}. Phylogenies were
+constructed from the nucleotide alignments by approximate maximum likelihood
+using \software{FastTree2} version 2.1.7 with the \gls{GTR} model. Transmission
+trees were estimated by rooting and time-scaling the phylogenies by root-to-tip
+regression, using a modified version of Path-O-Gen (distributed as part of
 BEAST~\autocite{drummond2007beast}) as described
@@
 399 respectively using a patristic distance cutoff of 0.02 as described
-previously~\autocite{poon2014impact}. Only sequences within these clusters were
+previously~\autocite{poon2015impact}. Only sequences within these clusters were
 carried forward. For the \textcite{novitsky2014impact} data, no large clusters

### 2016-04-15 doc/hiv-dynamics/presentation.tex
Added HIV Dynamics & Evolution talk

\documentclass{beamer}
\usepackage{tikz}
\usepackage{tikz-qtree}
\usepackage{gillius}
\usepackage{abraces}
\usepackage{etex}
\usepackage{array}
\usepackage{multirow}
\usepackage{qrcode}
\usepackage[backend=biber]{biblatex}

% http://tex.stackexchange.com/questions/12703/how-to-create-fixed-width-table-columns-with-text-raggedright-centered-raggedlef
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

\usetikzlibrary{external}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{positioning}
\usetikzlibrary{decorations.pathreplacing}

\usetheme[everytitleformat=regular]{m}
\setbeamertemplate{navigation symbols}{}

\graphicspath{{../figures/}}
\newcommand{\tablepath}{../tables}

\title{Phylogenetic inference of contact network parameters with kernel approximate Bayesian computation}
\author[RMM, RHL \& AFYP]{Rosemary M McCloskey$^1$ \and Richard H Liang$^1$ \and Art FY Poon$^{1,2}$}
\institute[UBC \& BCCfE]{$^1$BC Centre for Excellence in HIV/AIDS, Vancouver, Canada \\ $^2$Department of Medicine, University of British Columbia, Vancouver, Canada}
\date{HIV Dynamics \& Evolution, Woods Hole, USA, April 25, 2016}

\newcommand{\dd}[2]{\frac{\text{d}\,#1}{\text{d}\,#2}}

\addbibresource{papers.bib}

\begin{document}
\setbeamercolor{background canvas}{bg=white}
\renewcommand{\footnotesize}{\tiny}
\definecolor{red}{RGB}{228,26,28}
\definecolor{blue}{RGB}{55,126,184}
\definecolor{green}{RGB}{77,175,74}
\definecolor{purple}{RGB}{152,78,163}

\maketitle

\begin{frame}{Most epidemiological models assume homogeneous mixing}
  \begin{columns}
    \begin{column}{0.6\textwidth}
      \centering
      \includegraphics{compartments}
      \vspace{1cm}

      \uncover<2->{\includegraphics[width=\textwidth]{sir-trajectories}}
    \end{column}
    \begin{column}{0.4\textwidth}
      \vspace{-2cm}
      \begin{itemize}
        \setlength{\itemsep}{24pt}
        \item Often provides reasonable approximation in practice.
        \uncover<2->{
        \item Can be inaccurate when substantial contact heterogeneity exists.
        }
      \end{itemize}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Network models capture contact heterogeneity}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \includegraphics[width=\textwidth]{contactnet-empty}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{itemize}
        \setlength{\itemsep}{12pt}
          \pause
        \item May offer more accurate predictions for highly structured
          populations.
          \pause
        \item Network parameters may be of interest for their own sake,
          \textit{e.g.} are there superspreaders?
          \pause
        \item Extremely difficult to estimate in practice.
      \end{itemize}
    \end{column}
  \end{columns}

\end{frame}

\begin{frame}{Contact networks shape transmission trees}
  \includegraphics[width=\textwidth]{contactnet}

  \begin{itemize}
      \pause
    \item Transmission trees in turn shape viral phylogenies.
      \pause
    \item \textbf{Aim:} estimate contact network parameters from viral
      phylogenies.
  \end{itemize}
\end{frame}

\begin{frame}{Scale-free networks and preferential attachment}
  \includegraphics[width=\textwidth]{pa-example}
  \begin{itemize}
    \item Sexual networks are often scale-free: degree distributions appear
      follow a power law ($\Pr(\text{degree} = d) \propto d^{\gamma}$).
      \pause
    \item Preferential attachment models generate scale-free networks.
  \end{itemize}
\end{frame}

\begin{frame}{Barab\'asi-Albert (BA) network model}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{minipage}[p][\textheight][t]{\textwidth}
        \centering
        \only<1>{\includegraphics[page=1, width=\textwidth]{pa.pdf}}
        \only<2>{\includegraphics[page=2, width=\textwidth]{pa.pdf}}
        \only<3>{\includegraphics[page=3, width=\textwidth]{pa.pdf}}
        \only<4>{\includegraphics[page=4, width=\textwidth]{pa.pdf}}
        \only<5>{\includegraphics[page=5, width=\textwidth]{pa.pdf}}
        \only<6>{\includegraphics[page=6, width=\textwidth]{pa.pdf}}
      \end{minipage}
    \end{column}
    \begin{column}{0.5\textwidth}
      \begin{minipage}[t][\textheight][t]{\textwidth}
      \begin{itemize}
        \uncover<1->{ \item Start with a small number of connected nodes. }
        \uncover<2->{ \item Attach new nodes with $m$ edges. }
        \uncover<3->{ \item Other endpoints of degree $d$ are chosen with probability $\propto d^{\alpha} + 1$. }
        \uncover<5->{ \item Continue until network has $N$ nodes. }
        \uncover<6->{ \item Also consider the prevalence $I$ when the transmission
          tree is sampled.}
      \end{itemize}
      \end{minipage}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Do network parameters measurably affect tree shape?}
  \only<1>{
    \includegraphics[height=1in, trim=0 2in 4in 0, clip]{kernel-idea}
  }
  \only<2>{
    \includegraphics[height=1in, trim=0 2in 2in 0, clip]{kernel-idea}
  }
  \only<3->{
    \includegraphics[height=1in, trim=0 2in 0 0, clip]{kernel-idea}
  }
  \begin{itemize}
    \item Generate networks under different parameter values ($N$, $m$, $\alpha$).
    \uncover<2->{
    \item Simulate epidemic over each network with Gillespie algorithm ($I$).
    }
    \uncover<3->{
    \item Randomly subsample to form transmission trees.
    }
    \uncover<4->{
    \item Compare pairwise using tree kernel\footfullcite{poon2013mapping}.
    }
  \end{itemize}
\end{frame}

\begin{frame}{$\alpha$ and $I$ separate well in feature space}
  \begin{tikzpicture}
    \node {\includegraphics[height=0.8\textheight]{kernel-kpca.pdf}};
    \fill<1>[white] (-4.6,-3.2) rectangle (5,0.2);
  \end{tikzpicture}
  \pause
\end{frame}

\begin{frame}{Classifiers for network parameters}
  \vspace{-1cm}
  \begin{itemize}
    \item kSVM using tree kernel
    \item SVM using normalized
      lineages-through-time\footfullcite{janzen2015approximate}
    \item linear regression using Sackin's index~\footfullcite{shao1990tree}
  \end{itemize}

  \pause
  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
\end{frame}

\begin{frame}{Approximate Bayesian computation (ABC) for networks}
  \begin{minipage}[p][\textheight][t]{\textwidth}
    \only<1>{\includegraphics[width=\textwidth, trim=0 3in 0 0, clip]{abc-idea.pdf}}
    \only<2>{\includegraphics[width=\textwidth, trim=0 1.8in 0 0, clip]{abc-idea.pdf}}
    \only<3>{\includegraphics[width=\textwidth, trim=0 1in 0 0, clip]{abc-idea.pdf}}
    \only<4>{\includegraphics[width=\textwidth, trim=0 0in 0 0, clip]{abc-idea.pdf}}
  \end{minipage}
\end{frame}

\begin{frame}{Sequential Monte Carlo for ABC}
  \begin{tikzpicture}
    \only<1-2>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 4in 0, clip]{abc-smc.pdf}}};
    \only<1>{\fill[white] (0.2,-6.5) rectangle (4,-3.5)};
    \only<3>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 2in 0, clip]{abc-smc.pdf}}};
    \only<4>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 0 0, clip]{abc-smc.pdf}}};
  \end{tikzpicture}
\end{frame}

\begin{frame}{Marginal posterior distributions for one simulated tree}
  \begin{minipage}[p][\textheight][t]{\textwidth}
    \only<1>{\centerline{\includegraphics[width=0.9\textwidth, trim=0 2.5in 0 0, clip]{abc-posterior-example.pdf}}}
    \only<2>{\centerline{\includegraphics[width=0.9\textwidth]{abc-posterior-example.pdf}}}
  \end{minipage}
\end{frame}

\begin{frame}{Accuracy of maximum \textit{a posteriori} estimates for $m = 2$}
  \begin{minipage}[p][\textheight][t]{\textwidth}
    \vspace{-0.5cm}
    \only<1>{\includegraphics[height=0.9\textheight, trim=0 0 3.4in 0, clip]{abc-point-estimate-m2.pdf}}
    \only<2>{\includegraphics[height=0.9\textheight]{abc-point-estimate-m2.pdf}}
  \end{minipage}
\end{frame}

\begin{frame}{Application to published HIV datasets}
  \uncover<2->{
  \includegraphics[width=\textwidth]{realdata-hpd-bc-horiz}
  }
  \vspace{0.5cm}

  \scriptsize
  \input{\tablepath/realdata-bc}
  \normalsize
\end{frame}

\begin{frame}{Conclusions}
  \begin{itemize}
    \item Kernel-ABC is a viable method for fitting contact network models to
      phylogenetic data.
      \pause
    \item Some parameters of the Barab\'asi-Albert preferential attachment data
      have a measurable impact on tree shape and can be re-estimated using
      kernel-ABC.
      \pause
    \item The networks underlying real epidemics may be heterogeneous,
      underscoring the importance of considering network structure in
      phylodynamic analyses.
      \pause
  \end{itemize}
  \centerline{\qrcode{github.com/rmcclosk/netabc} github.com/rmcclosk/netabc}
\end{frame}

\begin{frame}{Acknowledgements}
  \begin{columns}
    \begin{column}{0.6\textwidth}

      \textbf{BC Centre for Excellence in HIV/AIDS}

      Art Poon

      Jeff Joy

      Richard Liang

      Thuy Nguyen

      P. Richard Harrigan

      \hfill\\
      \textbf{University of British Columbia}

      Sarah Otto

      Alexandre Bouchard-C\^ot\'e

      \vfill
    \end{column}
    \begin{column}{0.4\textwidth}
      \centering

      \includegraphics[width=3cm]{logos/cihr}
      \vspace{0.5cm}

      \includegraphics[width=3cm]{logos/bmgf}
      \vspace{0.5cm}

      \includegraphics[width=1cm]{logos/ubc}
      \vspace{0.5cm}

      \includegraphics[width=3cm]{logos/btp}
    \end{column}
  \end{columns}
\end{frame}
\end{document}

### 2016-04-15 doc/thesis/intro.tex
Mostly SMC background

@@
 can be used to identify demographic groups to whom targeted interventions are
-likely to have the greatest effect~\autocite{poon2014impact}.  However, this
+likely to have the greatest effect~\autocite{poon2015impact}.  However, this
 network may show little to no agreement with a contact data obtained through
@@
 such as posterior distributions attained by sequentially considering new
-evidence. As with \gls{SIS}, we assume that $\pi_i = \gamma_i / Z_i$, where
-$\gamma_i$ is known pointwise and the normalizing constant $Z_i$ is unknown.
-
-Both algorithms follow the same general pattern: the target distribution is
-attained by progressing through a sequence of intermediate distributions. For
-\gls{SIS}, the intermediate distributions are lower-dimensional marginals of
-the target, while for the \gls{SMC} sampler, they are of the same dimension and
-constitute a smooth progression from an initial distribution to the target. In
-both cases, the neighbouring distributions in the sequence are related to each
-other in some way, and we can take advantage of that relationship to create a
-sequence of importance distributions alongside the sequence of targets. In
-\gls{SIS}, the neighbouring marginals $\pi(\vec{x_i})$ and
-$\pi(\vec{x_{i+1}})$ were related by the conditional density $\pi(x_i \mid
-\vec{x_{i-1}})$, which we used to inform the importance distribution.
-
-The sequence of targets considered by the \gls{SMC} sampler are assumed to be
-closely enough related, that we can shift an importance distribution for
-$\pi_i$ into one for $\pi_{i+1}$ by perturbing the particles in a systematic
-way.
+evidence. As with \gls{SIS}, we assume that $\pi_i(x) = \gamma_i(x) / Z_i$,
+where $\gamma_i$ is known pointwise and the normalizing constant $Z_i$ is
+unknown.
+
+Both algorithms involve progression through a sequence of related
+distributions. For \gls{SIS}, these distributions are lower-dimensional
+marginals of the target distribution, while for the \gls{SMC} sampler, they are
+of the same dimension and constitute a smooth progression from an initial to a
+final distribution. In both cases, the neighbouring distributions in the
+sequence are related to each other in some way, and we can take advantage of
+that relationship to create a sequence of importance distributions alongside
+the sequence of targets. In \gls{SIS}, the neighbouring marginals
+$\pi(\vec{x_i})$ and $\pi(\vec{x_{i+1}})$ were related by the conditional
+density $\pi(x_i \mid \vec{x_{i-1}})$, which we used to inform the importance
+distribution. In \gls{SMC}, the relationship between subsequent distributions
+is less explicit, but it is assumed that they are related closely enough that
+an importance distribution for $\pi_i$ can be easily transformed into one for
+$\pi_{i+1}$. In particular, the sequence of importance distributions $\eta_i$
+is constructed as
+\begin{align}
+  \label{eq:impint}
+  \eta_i(x) = \int \eta_{i-1}(\xi) K_i(\xi, x) \d\xi,
+\end{align}
+where $K_i$ is any Markov kernel and the integral is over the space on which the
+$\pi_i$ are defined. The choice of $K_i$ should be based on the percieved
+relationship between $\pi_{i-1}$ and $\pi_i$. \textcite{del2006sequential}
+propose the use of a \gls{MCMC} kernel with equilibrium distribution $\pi_i$.
+That is,
+\[
+  K_i(\xi, x) = \max\left(1, \frac{q(x, \xi)\pi_i(\xi)}{q(\xi, x)\pi_i(x)}\right),
+\]
+where $q(\xi, x)$ is a proposal function such as a Gaussian distribution
+centered at $\xi$ (see \cref{subsec:mfit}).
+
+Although this method of building up $\eta$ appears straightforward, the
+drawback is that the importance distribution itself becomes intractible. In
+particular, evaluating $\eta_d(x)$ involves a $d$-dimensional integral of the
+type \cref{eq:impint}. As it is necessary to evaluate $\eta(x)$ pointwise to
+perform \gls{IS}, this construction appears to have defeated the purpose of
+providing an importance distribution for each $\pi_i$.
+\textcite{del2006sequential} overcome this problem with two ``artificial''
+objects. First, they propose the existence of \textit{backward} kernels
+$L_{i-1}(x_i, x_{i-1})$ which perform the inverse operation as $K_i$, namely
+transforming $\eta_i$ back into $\eta_{i-1}$. As we shall see, these backward
+kernels do not need to be calculated analytically in terms of $K_i$. Second,
+they define a sequence of $i$-dimensional target distributions
+\[
+  \tilde{\pi}_i(\vec{x_i}) = \frac{\gamma_i(x_i)}{Z_i} \prod_{k=1}^{i-1} L_k(x_{k+1}, x_k)
+\]
+which admit $\pi_i(\vec{x_i})$ as marginals.
 

### 2016-04-15 doc/thesis/results.Rnw
Mostly SMC background

@@
 infected nodes. The three \gls{alpha} and \gls{I} values considered are well
-separated from each other in feature space (\cref{fig:pakpca}). On the other
-hand, the three \gls{N} values overlap significantly, and the three \gls{m}
-values are virtually indistinguishable. Similar observations can be made for
-other values of \gls{I} and the number of tips 
+separated from each other in feature space. On the other hand, the three
+\gls{N} values overlap significantly, and the three \gls{m} values are
+virtually indistinguishable. Similar observations can be made for other values
+of \gls{I} and the number of tips
 (\cref{fig:alphakpca,fig:Nkpca,fig:Ikpca,fig:mkpca}). The values of \gls{I} and
@@
   }
-  \label{fig:pakpca}
+  \label{fig:kpca}
 \end{figure}
@@
 performance than the tree kernel, although the magnitude of the disparity
-varied between the parameters (\cref{fig:rsquared}, centre and right).
-\Cref{fig:rsquared} shows the cross-validation accuracy of each of the tree
-classifiers. The results were largely robust to variations in the tree kernel
-meta-parameters $\lambda$ and $\sigma$, although accuracy varied between
-different epidemic and sampling scenarios
+varied between the parameters (\cref{fig:rsquared}, centre and right). The
+results were largely robust to variations in the tree kernel meta-parameters
+$\lambda$ and $\sigma$, although accuracy varied between different epidemic and
+sampling scenarios
 (\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}).
@@
 
-We used grid search to obtain \defn{marginal} estimates for each network
-parameter while holding all other parameters fixed. We observed that kernel
-scores were highest at the values of \gls{alpha} on the grid closest to the
-true values, as shown in Figure~\ref{fig:gridkernel}. However, there was a much
-stronger spike in kernel scores near the true value for \gls{alpha} = 1.0 and
-1.25. This is recapitulated when we look at the accuracy of point estimates
-obtained by taking the grid value with the highest median kernel score. As
-shown in Figure~\ref{fig:gridest}, while the estimates are generally close to
-the true value, they are much closer for \gls{alpha} = 1.25 than for the other
-values.
+The accuracy of grid search estimates largely paralleled that of the \gls{kSVM}
+classifiers.
+
+We used grid search to obtain marginal estimates for each network parameter
+while holding all other parameters fixed. We observed that kernel scores were
+highest at the values of \gls{alpha} on the grid closest to the true values, as
+shown in Figure~\ref{fig:gridkernel}. However, there was a much stronger spike
+in kernel scores near the true value for \gls{alpha} = 1.0 and 1.25. This is
+recapitulated when we look at the accuracy of point estimates obtained by
+taking the grid value with the highest median kernel score. As shown in
+Figure~\ref{fig:gridest}, while the estimates are generally close to the true
+value, they are much closer for \gls{alpha} = 1.25 than for the other values.
 

### 2016-04-16 doc/thesis/appendix.tex
More writing yay

@@
 \end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics{gridsearch-alpha-kernel.pdf}
+  \caption{                                              
+    Grid search kernel scores for testing trees simulated under various             
+    \gls{alpha} values. The other \gls{BA} parameters were fixed at \gls{I} =
+    1000, \gls{N} = 5000, and \gls{m} = 2. 
+  }        
+  \label{fig:gridalpha}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics{gridsearch-I-kernel.pdf}
+  \caption{                                              
+    Grid search kernel scores for testing trees simulated under various             
+    \gls{I} values. The other \gls{BA} parameters were fixed at \gls{alpha} =
+    1.0, \gls{N} = 5000, and \gls{m} = 2. 
+  }        
+  \label{fig:gridI}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics{gridsearch-m-kernel.pdf}
+  \caption{                                              
+    Grid search kernel scores for testing trees simulated under various             
+    \gls{m} values. The other \gls{BA} parameters were fixed at \gls{alpha} =
+    1.0, \gls{I} = 1000, and \gls{N} = 5000.
+  }        
+  \label{fig:gridm}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics{gridsearch-N-kernel.pdf}
+  \caption{                                              
+    Grid search kernel scores for testing trees simulated under various             
+    \gls{N} values. The other \gls{BA} parameters were fixed at \gls{alpha} =
+    1.0, \gls{I} = 1000, and \gls{m} = 2. 
+  }        
+  \label{fig:gridN}
+\end{figure}

### 2016-04-16 doc/thesis/conclusion.tex
More writing yay

@@
 Our work had three main aims. First, we developed a method to estimate contact
-network model parameters from phylogenetic data. This method widens the field
-of models which may be investigated in a phylodynamic context. Second, we
-investigated the parameters of the \gls{BA} model. We determined through
-simulation studies that the preferential attachment power $\alpha$ and number
-of infected nodes $I$ had a substantial impact on transmission tree shape, and
-could be estimated using our method.
+network model parameters from phylogenetic data, based on kernel approximate
+Bayesian computation. This method widens the field of models which may be
+investigated in a phylodynamic context.
+
+Our second aim was to investigate the parameters of the \gls{BA} model, in
+terms of their impact on tree shape and our ability to reconstruct them with
+kernel-\gls{ABC}. We determined through simulation studies that the
+preferential attachment power $\alpha$ and number of infected nodes $I$ had a
+substantial impact on transmission tree shape, and could be estimated using our
+method.
+
+The final aim of our work was to apply our method to real world \gls{HIV}
+datasets. This analysis indicated heterogeneity in the contact network
+structures underlying several distinct local epidemics. The five datasets
+analysed fell into three categories (Figure~\ref{fig:abchpd}). First, we
+estimated a preferential attachment power between 0.5 and 1 for the epidemics
+studied by \textcite{cuevas2009hiv} and \textcite{li2015hiv}, with credible
+intervals occupying nearly the entire region from 0 to 1.
+\citeauthor{cuevas2009hiv} studied a group of newly diagnosed individuals in
+the Basque Country, Spain. Although the individuals were of mixed risk groups,
+and therefore unlikely to comprise a single contact network, a high proportion
+of them (47\%) grouped into local clusters based on genetic distance. The low
+estimated attachment power for these data is consistent with the sampled
+sequences comprising many distinct sub-networks rather than a single connected
+network. \citeauthor{li2015hiv} sampled a large number of acutely infected MSM
+in Shanghai, China, in which we identified a large cluster from the phylogeny
+using a patristic distance cutoff~\autocite{poon2015impact}. The low attachment
+power estimated for this dataset was surprising given the high phylogenetic
+relatedness of the sequences. It is possible that the number and diversity of
+circulating recombinant forms in the data introduced errors into our estimated
+viral phylogeny.
+
+For the outbreaks studied by \textcite{niculescu2015recent} and
+\textcite{wang2015targeting}, the estimated $\alpha$ was close to one, with a
+narrower credible interval than for the other studies.
+\citeauthor{niculescu2015recent} studied a recent outbreak among Romanian
+injection drug users (IDU), while \citeauthor{wang2015targeting} sampled
+acutely infected MSM in Beijing, China. Both studies discovered a high degree
+of phylogenetic relatedness owing to the recent infection times and homogeneous
+risk groups of the studied populations. The estimated number of infections for
+these datasets were also quite low, although the HPD interval for
+\citeauthor{wang2015targeting} was much wider than that for
+\citeauthor{niculescu2015recent}.
+
+The final studied dataset was an outlier in terms of estimated parameters.
+\textcite{novitsky2013phylogenetic} sampled approximately 44\% of the
+HIV-infected individuals in the northern area of Mochudi, Botswana. Additional
+sampling in a later study~\autocite{novitsky2014impact} brought the genotyping
+coverage up to 70\%. Even with such a high sampling coverage, we did not detect
+any large clusters using patristic distance, and therefore chose to analyze a
+subtree instead. Estimates of $\alpha$ and $N$ both had very wide HPD
+intervals and were markedly different from the other datasets. The estimated
+number of infected nodes was also extremely high, much higher than the
+estimated HIV prevalence of the town. Several factors may have contributed to
+these results. First, the authors note that the their sample was 75\% female.
+In a primarily heterosexual risk environment, removal of a disproportionate
+number of males from the network could obfuscate the true network structure,
+for example if the majority of highly connected nodes were of one gender.
+Second, the town in question was in close proximity to the country's capital,
+and the authors indicated that a high amount of migration takes place between
+the two locations. This suggests that the contact network may include a much
+larger group based in the capital city, which would explain the high estimate
+of $I$.
 

### 2016-04-16 doc/thesis/intro.tex
More writing yay

@@
 One mechanism which has been shown to lead to scale-free networks is
-\defn{preferential attachment}~\autocite{simon1955class, barabasi1999emergence}.
-Under this process, networks are formed by starting with a small number $m_0$
-of nodes. New nodes are added one at a time until there are a total of $N$ in
-the network. Each time a new node is added, $m \geq 1$ edges are added from it
-to other nodes in the graph. In the original
-formulation~\autocite{barabasi1999emergence}, the partners of the new node are
-chosen with probability linearly proportional to their degree. However,
+\defn{preferential attachment}~\autocite{simon1955class,
+barabasi1999emergence}. The simplest preferential attachment model is known as
+the \acrlong{BA} (BA) model after its
+inventors~\autocite{barabasi1999emergence}. Under this model, networks are
+formed by starting with a small number $m_0$ of nodes. New nodes are added one
+at a time until there are a total of $N$ in the network. Each time a new node
+is added, $m \geq 1$ edges are added from it to other nodes in the graph. In
+the original formulation, the partners of the new node are chosen with
+probability linearly proportional to their degree. However,
 \citeauthor{barabasi1999emergence} suggest extending the model such that the
 probability of choosing a partner of degree $d$ is proportional to $d^\alpha$
-for some constant $\alpha$, and we use this extension here.
+for some constant $\alpha$, and we use this extension here. When $m = 1$, the
+network takes on the distinctive shape of a tree, that is, it does not contain
+any cycles. Cycles are present in the network for all all other $m$ values.
 

### 2016-04-16 doc/thesis/methods.tex
More writing yay

@@
 the \software{treekernel} binary. The median kernel score was calculated for
-each grid value. Grid values were resampled with probability proportional to
-their median kernel scores. A point estimate for the test value was obtained by
-taking the highest point of an estimated kernel density of the resampled grid
-values, and a 95\% credible interval was obtained using the
-\software{HPDinterval} function in the \software{coda} package.
+each grid value, and the scores were normalized such that the area under the
+curve was equal to 1. The grid value with the highest median kernel score was
+taken as the point estimate for the test value, and a 95\% credible interval
+was obtained using the \software{hpd} function in the \software{TeachingDemos}
+package.
 

### 2016-04-16 doc/thesis/results.tex
More writing yay

@@
-\subsection{Separable parameters}
+\subsection{Kernel classifiers}
 
-Our first simulation-based experiment was designed to determine which network
-model parameters were potentially estimable by kernel-ABC, and to find the best
-set of kernel meta-parameters to carry through to future experiments. For each
-parameter of the \gls{BA} model, we simulated trees under three distinct
-values, and computed the cross-validation accuracy of a kernel-\gls{SVM}
-classifier for the parameter.
-
-In Figure~\ref{fig:patrees}, we show that trees simulated under different
-values of \gls{alpha} are visibly quite distinctive. In particular, higher
-values of \gls{alpha} result in networks with a small number of highly
-connected nodes (see Figure~\ref{fig:pabounds}) which, once infected, are
-likely to transmit to many other nodes. This results in a more unbalanced,
-ladder-like structure in the phylogeny. Kernel-\gls{PCA} projections show that
-all three \gls{alpha} values are well separated from each other under several
-different sampling scenarios (Figure~\ref{fig:pakpca}). With smaller trees,
-it becomes harder to visually disinguish \gls{alpha} = 0.5 from \gls{alpha} =
-1.0 using the first two principal components.
-
-With suitably chosen meta-parameters, the accuracy of the kernel-\gls{SVM}
-classifier for \gls{alpha} was very high under a variety of prevalence and
-sampling scenarios (Figure~\ref{fig:pacrossv}). Accuracy was highest, with
-$R^2$ values above 0.95, for the largest trees and complete sampling (bottom
-center panel). However, even for trees of size 100 sampled from an epidemic on
-2000 nodes, the $R^2$ was above 0.8 (top right panel). In all cases, the
-accuracy of a Sackin's index-based classifier was also quite high, at about
-0.75. An \gls{SVM} classifier using only the \gls{nltt}
-statistic~\autocite{janzen2015approximate} was slightly worse than Sackin's
-index, with an $R^2$ of X.
-
-The fact that Sackin's index is informative of \gls{alpha} was
-unsurprising, given the clear differences in tree balance observed under
-different \gls{alpha} values.
+Trees simulated under different values of \gls{alpha} were visibly quite
+distinct (\cref{fig:alphatrees}). In particular, higher values of \gls{alpha}
+produce networks with a small number of highly connected nodes which, once
+infected, are likely to transmit to many other nodes. This results in a more
+unbalanced, ladder-like structure in the phylogeny, compared to networks with
+lower \gls{alpha} values. None of the other three parameters produced trees
+which were as easily distinguished from each other
+(\cref{fig:Itrees,fig:mtrees,fig:Ntrees,fig:Itrees}).
+Sackin's index, which measures tree imbalance, was significantly correlated with
+all four parameters
+    (for $\alpha$, $I$, $m$, and $N$ respectively: Spearman's rho =
+     0.85,
+     \ensuremath{-0.12},
+     \ensuremath{-0.13},
+     0.09;
+     $p$-values
+     $<10^{-5}$,
+     $0.003$,
+     $<10^{-5}$,
+     $<10^{-5}$)
+The ratio of internal to terminal branch lengths was negatively correlated with
+\gls{alpha} and \gls{I}, and positively corelated with \gls{N} and \gls{m}
+  (Spearman's rho
+    \ensuremath{-0.84},
+    \ensuremath{-0.69},
+    0.1,
+    0.18;
+  $p$-values
+    $0$,
+    $2\!\times\!10^{-85}$,
+    $4\!\times\!10^{-7}$,
+    $0$).
 
@@
   \centering
-  \label{fig:patrees}
-  \includegraphics{kernel_pa_trees}
+  \includegraphics{kernel-alpha-tree.pdf}
   \caption[Visibly distinctive trees simulated under three values of \gls{alpha}]{
-    Epidemics of 1000 infected were simulated on \gls{BA} networks of 5000
-    nodes, with \gls{alpha} equal to 0.5, 1.0, or 1.5. Transmission trees were
-    created by sampling 500 infected nodes. Higher \gls{alpha} values produce
-    networks with a small number of highly-connected nodes, which results in a
-    highly unbalanced, ladder-like tree structure.
+    Epidemics simulated on \gls{BA} networks of 5000 nodes, with \gls{alpha}
+    equal to 0.5, 1.0, or 1.5, until 1000 individuals were infected.
+    Transmission trees were created by sampling 500 infected nodes. Higher
+    \gls{alpha} values produced networks with a small number of
+    highly-connected nodes, resulting in highly unbalanced, ladder-like trees.
   }
+  \label{fig:alphatrees}
 \end{figure}
+
+\Cref{fig:kpca} shows \gls{kPCA} projections of the simulated trees onto the
+first two principal components of the kernel matrix. The figure shows only the
+simulations with 500-tip trees and, for all parameters except $I$, 1000
+infected nodes. The three \gls{alpha} and \gls{I} values considered are well
+separated from each other in feature space. On the other hand, the three
+\gls{N} values overlap significantly, and the three \gls{m} values are
+virtually indistinguishable. Similar observations can be made for other values
+of \gls{I} and the number of tips
+(\cref{fig:alphakpca,fig:Nkpca,fig:Ikpca,fig:mkpca}). The values of \gls{I} and
+\gls{N} separate more clearly with larger numbers of tips, and in the case of
+\gls{N} larger epidemic sizes.
 
@@
   \centering
-  \label{fig:pacrossv}
-  \includegraphics{kernel_pa_crossv}
-  \caption[Cross-validation performance of kernel support vector machine
-  classifier for preferential attachment power]{
-    Cross-validation performance of kernel support vector machine classifier
-    for preferential attachment power.
+  \includegraphics{kernel-kpca.pdf}
+  \caption[\gls{kPCA} projections of simulated trees under varying \gls{BA}
+           parameter values]{
+    Each parameter of the \gls{BA} model was individually varied to produce 300
+    simulated trees. Kernel matrices were formed from all pairwise kernel
+    scores among each set of 300 trees. The trees were projected onto the first
+    two principal components of the kernel matrix calculated using \gls{kPCA}.
+    All trees had 500 tips. The parameters not being varied were set to
+    \gls{alpha} = 1, \gls{I} = 1000, \gls{m} = 2, and \gls{N} = 5000. The tree
+    kernel meta-parameters were $\lambda = 0.3$ and $\sigma = 4$.
   }
+  \label{fig:kpca}
 \end{figure}
+
+Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
+tested (\cref{fig:rsquared}, left). Classifiers based on two other tree
+statistics, the \gls{nltt} and Sackin's index, generally exhibited worse
+performance than the tree kernel, although the magnitude of the disparity
+varied between the parameters (\cref{fig:rsquared}, centre and right). The
+results were largely robust to variations in the tree kernel meta-parameters
+$\lambda$ and $\sigma$, although accuracy varied between different epidemic and
+sampling scenarios
+(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}).
+
+When classifying $\alpha$, the kernel-SVM classifier had an average $R^2$ of 
+    0.92,
+compared to 
+    0.56
+for the \gls{nltt}-based SVM, and
+    0.75
+for the linear regression against Sackin's index. There was little variation
+about the mean for different tree and epidemic sizes. No classifier could
+accurately identify the $m$ parameter in any epidemic scenario, with average
+$R^2$ values of 
+  0.12 for \gls{kSVM},
+  0.01 for the \gls{nltt}, and
+  0.06
+for Sackin's index. Again, there was little variation in accuracy between
+epidemic scenarios, although the accuracy of the \gls{kSVM} was slightly higher
+on 1000-tip trees.
+
+The accuracy of classifiers $I$ varied significantly with the number of tips in
+the tree. For 100-tip trees, the average $R^2$ values were
+  0.7,
+  0.55, and
+  0.02
+for the tree kernel, \gls{nltt}, and Sackin's index respectively. For 500-tip
+trees, the values increased to
+  0.93,
+  0.83, and
+  0.07.
+Finally, the performance of classifiers for $N$ depended heavily on the
+epidemic scenario. The $R^2$ of the \gls{kSVM} classifier ranged from
+  0.08
+for the smallest epidemic and smallest sample size, to
+  0.82
+for the largest. Likewise, $R^2$ for the \gls{nltt}-based SVM ranged from 
+  0.01
+to
+  0.54.
+Sackin's index did not accurately classify $N$ in any scenario, with an average
+$R^2$ of
+  0.03
+and little variation between scenarios.
 
@@
   \centering
-  \label{fig:pakpca}
-  \includegraphics{kernel_pa_kpca}
-  \caption[Projection of kernel matrix for different attachment power values
-  onto its first two principal components]{
-    Projection of the kernel matrix for different preferential attachment power
-    values onto its first two principal components, for eight simulation
-    scenarios. Each point corresponds to a simulated transmission tree, and is
-    coloured by preferential attachment power. Facets are number of infected
-    nodes (horizontal), and number of sampled tips (vertical). The parameters
-    to the tree kernel were $\lambda = 0.3$ and $\sigma = 4$, and the nLTT was
-    not used. Qualitatively, trees with a larger number of tips are easier to
-    separate in kernel space, regardless of what sampling proportion they
-    represent. In all cases, the highest attachment power can be separated from
-    the other two, but the two lowest values become hand to distinguish with in
-    the 100-tip trees.
+  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
+  \caption{
+      Cross-validation accuracy of kernel-SVM classifier (left), SVM classifier
+      using \gls{nltt} (centre), and linear regression using Sackin's index
+      (right) for \gls{BA} model parameters. Kernel meta-parameters were set to
+      $\lambda = 0.3$ and $\sigma = 4$. Each point was calculated based on 300
+      simulated transmission trees over networks with three different values of
+      the parameter being tested. Vertical lines are empirical 95\% confidence
+      intervals based on 1000 two-fold cross-validations.
   }
+  \label{fig:rsquared}
 \end{figure}
 
-Figure~\ref{fig:rsquared} shows the cross-validation accuracy of classifiers
-for all four parameters of the \gls{BA} model. The classifier for \gls{I} also
-exhibited high performance, although not quite as high as that for \gls{alpha}.
-The average $R^2$ values were 0.70 for 100-tip trees, or 0.93 for 500-tip
-trees. On the other hand, $m$ and $N$ were much harder to classsify. For $m$,
-all $R^2$ values were below 0.5, falling to virtually zero in four scenarios.
-The $R^2$ values for $N$ were highly variable, ranging from 0.08 to 0.82
-depending on $I$ and the number of tips in the tree. In almost all cases, both
-the nLTT and Sackin's index classifiers were worse than the tree kernel, except
-for classifying $N$ with $I = 2000$ and $100$ tips, where the nLTT gave a small
-advantage (see the supplements). The greatest disadvantage for the Sackin's
-index-based classifier was on the $I$ parameter, where it had an $R^2$ of less
-than 0.1, despite the relatively high accuracy of the tree kernel (see the
-supplements).
+\subsection{Grid search}
+
+The accuracy of grid search estimates largely paralleled that of the \gls{kSVM}
+classifiers. \Cref{fig:gridest} shows point estimates and 95\% highest density
+intervals for each of the \gls{BA} parameters, for one replicate experiment
+with 500-tip trees. For all parameters except $m$, the error of point estimates
+was negatively correlated with the number of sampled tips in the tree (for
+\gls{alpha}, \gls{I}, and \gls{N} respectively: Spearman's $\rho$ = 
+    \ensuremath{-0.22},
+    \ensuremath{-0.41},
+    \ensuremath{-0.16};
+$p$-value = 
+    $4\!\times\!10^{-4}$,
+    $2\!\times\!10^{-4}$,
+    $0.01)$.
+The highest density intervals obtained for all parameters were extremely wide,
+occcupying $>$90\% of the grid in all cases.
+
+The \gls{alpha} parameter was the most accurately estimated, with point
+estimates having an average deviation of 
+    0.14
+from the true value, on a grid from 0 to 2. The error was negatively correlated
+with the true value of \gls{alpha} 
+    (Spearman's $\rho$ = \ensuremath{-0.26},
+     $p = 1\!\times\!10^{-5}$),
+although the relationship was clearly nonlinear (\cref{fig:gridest,fig:gridalpha}).
+The accuracy was highest for the test value \gls{alpha} = 1.25
+    (mean error 0.02)
+which exhibited markedly different behaviour than the other values in terms of
+the distribution of kernel scores along the grid (\cref{fig:gridalpha}). In
+particular, there was a very pronounced peak in scores around the true value,
+in contrast to most other values where the scores were flat around the true
+value. The peak was also observed to a lesser degree for \gls{alpha} = 1. The
+average absolute error of the point estimates for \gls{I} was 
+    298,
+on a grid of 500 to 5000, and the errors were not significantly correlated with
+the true value of \gls{I}. Kernel score distributions for all test values
+exhibited the same rounded shape (\cref{fig:gridI}).
+
+For the other two parameters, \gls{m} and \gls{N}, accuracy was much poorer
+(\cref{fig:gridest,fig:gridm,fig:gridN}). The average error for \gls{m} was
+    1.31,
+on a grid from 1 to 6; this error was positively correlated with increasing
+\gls{m}
+    (Spearman's $\rho$ = 0.25,
+     $p = 6\!\times\!10^{-4}$).
+This positive correlation was apparently due to the much lower error for $m =
+1$ than for the other $m$ values (\cref{fig:gridm})
+    (mean errors 0.1 for
+     $m = 1$ vs. 1.5466667 for
+     $m > 1$).
+The value $m = 1$ causes the network to take on a distinct shape relative to
+higher $m$ values, namely a tree (\textit{i.e.} there are no cycles,
+\cref{subsec:treeshape}). The average error for \gls{N} was 
+    2419,
+on a grid from 1000 to 15000, and was positively correlated with the true value
+of \gls{N}
+    (Spearman's $\rho$ = 0.43,
+     $p = 2\!\times\!10^{-12}$).
 
@@
   \centering
-  \includegraphics{kernel_rsquared}
-  \caption[Cross-validation accuracy of kSVM classifiers for \gls{BA}
-  model parameters.]{Cross-validation $R^2$ values of kernel-SVM classifier for
-    \gls{BA} model parameters. Points and ranges are median and 95\% quantiles
-    over 1000 2-fold cross-validations.}
-  \label{fig:rsquared}
-\end{figure}
-
-% needs to go in methods
-
-%It is important to note that, if we cut off the epidemic simulation when 500
-%nodes are infected, the resulting tree will be shorter (in calendar time) than
-%if we continue until 2000 nodes are infected. However, this information is not
-%used when building the classifier, since the branch lengths in each tree are
-%scaled by their mean. Therefore, the high performance of the classifier is due
-%to structural differences captured by the tree kernel, rather than the trees
-%simply having different heights.
-
-\subsection{Accuracy of marginal estimates}
-
-We used grid search to obtain \defn{marginal} estimates for each network
-parameter while holding all other parameters fixed. We observed that kernel
-scores were highest at the values of \gls{alpha} on the grid closest to the
-true values, as shown in Figure~\ref{fig:gridkernel}. However, there was a much
-stronger spike in kernel scores near the true value for \gls{alpha} = 1.0 and
-1.25. This is recapitulated when we look at the accuracy of point estimates
-obtained by taking the grid value with the highest median kernel score. As
-shown in Figure~\ref{fig:gridest}, while the estimates are generally close to
-the true value, they are much closer for \gls{alpha} = 1.25 than for the other
-values.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{gridsearch_pa_kernel}
-  \caption[Grid search kernel scores]{
-    Grid search kernel scores for testing trees simulated under various
-    \gls{alpha} values. All epidemics had \gls{I} = 1000 infected nodes, on
-    \gls{BA} networks of size \gls{N} = 5000 with \gls{m} fixed at 2. Colours
-    indicate the number of sampled tips.
-  }
-  \label{fig:gridkernel}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{gridsearch_pa_estimates}
-  \caption[Marginal estimates of \gls{alpha} obtained with grid search]{
-    Marginal estimates of \gls{alpha} obtained with grid search. Training trees
-    were simulated on a narrowly spaced grid of \gls{alpha} values, and
-    compared to testing trees using the tree kernel. The \gls{alpha} value in
-    the grid with the highest median kernel score was taken as the point
-    estimate for the testing tree. These point estimates are shown as black
-    dots. The dashed line is the identity.
+  \includegraphics[width=\textwidth]{gridsearch-example}
+  \caption[Grid search estimates of \gls{BA} model parameters]{Point estimates
+      and 95\% highest density intervals for each \gls{BA} model parameter,
+      obtained using grid search. Networks and transmission trees were
+      simulated over a grid of values for each parameter while holding the
+      others fixed. For a subset of the grid values ($x$-axis), test networks
+      and trees were created and compared to each tree on the grid using the
+      tree kernel. The kernel scores along the grid were normalized to resemble
+      a probability distribution, from which the mode and highest density
+      interval were calculated. Shown values correspond to one replicate
+      experiment, with trees of size 500.
   } 
@@
 \subsection{Accuracy of estimates with full ABC}
+
+We used kernel-\gls{ABC} to estimate the parameters of the \gls{BA} model 
+on simulated trees where the true parameter values were known. Point estimates
+for each parameter are shown in Figure~\ref{fig:abcpt}. Of the four parameters,
+\gls{alpha} was the most accurately estimated, with a median [IQR] absolute error
+of 
+    0.11 
+    [0.05-0.18].
+The accuracy of the estimates was not significantly different between values of
+$m$ or $I$ (both one-way ANOVA,
+    $p$ = 0.1
+and 
+    0.25),
+although the errors when the true value of \gls{alpha} was zero were
+significantly greater than the other values 
+    (Wilcoxon rank-sum test, $p$ = \ensuremath{6.41\times 10^{-4}}).
+The error in the estimated value of $I$ was
+    306 
+    [108-607].
+Errors were significantly higher for $\alpha \geq 1$
+    (Wilcoxon rank-sum test, $p$ = \ensuremath{6.12\times 10^{-4}})
+and for $I$ = 2000
+    ($p$ = \ensuremath{1.58\times 10^{-6}}),
+but not for any values of $m$
+    (one-way ANOVA, $p$ = 0.33).
+The $m$ parameter was estimated correctly in
+    37 \%
+of simulations, with an error of one in
+    40 \%
+and of two or more in 
+    22 \%
+(the only possible $m$ values were 2, 3, 4, or 5). The true values of
+$m$ and $I$ did not significantly affect the error
+    (one-way ANOVA, $p$ = 0.5 and
+                          0.68),
+but the accuracy was significantly lower for integral than non-integral
+values of \gls{alpha}
+    (Wilcoxon rank-sum test, $p$ = \ensuremath{7.2\times 10^{-3}}).
+Finally, the total number of nodes \gls{N} was consistently over-estimated by
+about a factor of two
+    (error \ensuremath{6.59\times 10^{3}} 
+    [\ensuremath{4.21\times 10^{3}}-\ensuremath{8.28\times 10^{3}}]).
+No other parameters influenced the accuracy of the $N$ estimates 
+    (one-way ANOVA, $p \geq$ NA).
+
+\Cref{fig:abcex} shows the \gls{ABC} approximation to the posterior
+distribution on the \gls{BA} parameters for one simulation (equivalent plots
+for all the simulations can be found in the supplemental materials). \Gls{HPD}
+intervals around \gls{alpha} and \gls{I} were narrow relative to the region of
+nonzero prior density, whereas the intervals for $m$ and \gls{N} were widely
+dispersed. \Cref{tab:abchpd} shows point estimates and 95\% \gls{HPD}
+intervals averaged over all simulations.
+
+\begin{table}
+    \centering
+    \input{\tablepath/abc-hpd}
+    \caption{Average widths of 95\% confidence intervals for \gls{BA} model
+    parameters estimated with kernel-\gls{ABC}.}
+    \label{tab:glm}
+\end{table}
+
+\subsection{Characterization of power-law exponent in Barab\'asi-Albert networks}
+
+\Cref{tab:glm} shows the estimated parameters for a log-link \gls{GLM} fitted
+to the observed distribution of \gls{gamma} values. The coefficients are
+interpretable as multiplicative effects.
+
+\begin{table}
+    \centering
+    \input{\tablepath/pa-gamma-glm}
+    \caption{Estimated \gls{GLM} parameters for relationship between power-law
+    exponent \gls{gamma} and \gls{BA} model parameters.}
+    \label{tab:glm}
+\end{table}
+
+\subsection{Real data}
+
+We applied kernel-ABC to five published HIV datasets (\cref{tab:data}),
+and found substantial heterogeneity among the parameter estimates
+(\cref{fig:abchpd}). Two of the datasets~\autocite{niculescu2015recent,
+wang2015targeting} had estimated $\alpha$ values near unity (MAP estimates
+[95\% HPD] 
+  1.06 
+  [0.63 - 
+   1.27]
+and
+  1 
+  [0.41 -
+   1.16] respectively). 
+Another two datasets~\autocite{li2015hiv, cuevas2009hiv} had lower estimated
+values and wider HPD intervals
+  (0.77 
+  [0.01 - 
+  1.03]
+and
+  0.66 
+  [0.03 -
+   0.84]). 
+The \textcite{novitsky2014impact} data had an extremely low estimated $\alpha$
+and a very wide HPD interval
+  (0.17 
+  [0.04 -
+   1.39]). 
+For all the datasets except \citeauthor{novitsky2014impact}, estimated values of
+$I$ were below 2000, with narrow HPD intervals around two of the
+datasets
+  (\citeauthor{cuevas2009hiv}, 880 
+  [290 -
+   \ensuremath{1.51\times 10^{3}}];
+   \citeauthor{niculescu2015recent}, 175
+  [138 - 
+   454])
+and wider intervals around the other two
+  (\citeauthor{li2015hiv}, \ensuremath{1.59\times 10^{3}} 
+  [284 -
+   \ensuremath{3.81\times 10^{3}}];
+   \citeauthor{wang2015targeting}, 651
+  [268 - 
+   \ensuremath{4.24\times 10^{3}}]).
+The \citeauthor{novitsky2014impact} data was again the outlier, with a very high
+estimated $I$, and HPD interval spanning almost the entire prior region
+  (\ensuremath{7.55\times 10^{3}} 
+  [228 -
+   \ensuremath{8.92\times 10^{3}}]).
+No information was gleaned about the $m$ parameter, with the HPD interval
+occupying the entire prior region for all datasets. The estimates of $N$ were
+similarly uninformative, with the exception that the point estimate for the
+\citeauthor{wang2015targeting} data was smaller
+  (\ensuremath{5.84\times 10^{3}})
+than the estimates for other datasets
+ (average \ensuremath{8.93\times 10^{3}}).
+
+\begin{figure*}[ht]
+  \centering
+  \includegraphics{realdata-hpd-bc.pdf}
+  \vspace{8pt}
+  \caption{
+      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
+      parameters of the BA network model, fitted to five published HIV datasets
+      with kernel-ABC.
+  }
+  \label{fig:abchpd}
+\end{figure*}

### 2016-04-16 doc/thesis/conclusion.tex
Wrote discussion mostly

--- 2016-03-29 doc/thesis/conclusion.tex
@@
+Due to the rapid advancement of nucleotide sequencing technology, viral
+sequence data data have become increasingly feasible to collect on a population
+level. Through phylodynamic methods, these data offer a window into
+epidemiological processes which would otherwise be virtually impossible to
+study on a realistic scale. Although phylodynamic methods have been developed
+to fit a wide variety of epidemiological models to phylogenetic
+data~\autocite{volz2012complex, rasmussen2014phylodynamic}, our method is the
+first (to our knowledge) which can fit models that do not assume panmixia. We
+believe this capability will be of broad interest to the molecular evolution
+and epidemiology community, as it widens the field of epidemiological
+parameters which may be investigated through viral sequence data. In addition,
+the characterization of local contact networks could be valuable from a public
+health perspective, such as for investigating optimal vaccination strategies.
+This information could assist in curtailing current epimemics, as well as
+preventing future epidemics of different diseases over the same contact
+network.
+
 Contact network structure has a substantial impact on epidemic trajectory
@@
 
-Our work had three main aims. First, we developed a method to estimate contact
-network model parameters from phylogenetic data. This method widens the field
-of models which may be investigated in a phylodynamic context. Second, we
-investigated the parameters of the \gls{BA} model. We determined through
-simulation studies that the preferential attachment power $\alpha$ and number
-of infected nodes $I$ had a substantial impact on transmission tree shape, and
-could be estimated using our method.
+Our work had three main aims. First, we developed phylodynamic a method to
+estimate contact network model parameters.
+
+Our second aim was to investigate the parameters of the \gls{BA} model, in
+terms of their impact on tree shape and our ability to reconstruct them with
+kernel-\gls{ABC}. We determined through simulation studies that the
+preferential attachment power $\alpha$ and number of infected nodes $I$ had a
+substantial impact on transmission tree shape, and could be estimated using our
+method.
+
+The final aim of our work was to apply our method to real world
+\gls{HIV} datasets.
 
 % ergm
-
-An alternative approach is the deterministic framework outlined by
-\textcite{morris1993epidemiology}, who proposes to apply the standard
-compartmental modelling framework to contact networks by assigning each
-individual their own compartment. Thus, each individual is associated with a
-single \gls{ODE}, with the entire \gls{ODE} system parameterized by the
-adjacency matrix of the contact network. \citeauthor{morris1993epidemiology}
-proposes to use log-linear models to parameterize the matrix. This framework is
-highly expressive, and allows straightforward incorporation of time-dependent
-dynamics. However, simulating a transmission tree would require the numerical
-solution of a very large system of \glspl{ODE}. Given the large number of
-simulations required for kernel-\gls{ABC}, it is not clear if this method would
-be computationally feasible in this context.
-
-% spatial networks, see section 5.4 in keeling2005networks
-
-The two-step process of simulating a contact network and subsequently allowing
-an epidemic to spread over that network carries with it the assumption that the
-contact network is static over the duration of the epidemic. Clearly this
-assumption is invalid, as people make and break partnerships on a regular
-basis. Our work has not addressed this assumption, primarily due to our desire
-to avoid the additional complexity required to address the dynamic nature of
-networks. This simplifying assumption is made by most studies using contact
-network models in an epidemiological context~\autocite{welch2011statistical,
-bansal2007individual}. However, in principle, kernel-\gls{ABC} could be
-adapted to dynamic contact networks by using a method such as that developed by
-\textcite{robinson2012dynamics} to simulate a dynamic contact network, while
-concurrently simulating the spread of an epidemic.
-
-% Caveats
-
-It is important to note that our kernel-ABC method takes a transmission tree as
-input, rather than a viral phylogeny. Thus, we have left the estimation of a
-transmission tree up to the user. There were two reasons for this choice. First
-and foremost, we wished again to avoid extra complexity and keep the number of
-estimated parameters small. In theory, it is possible to incorporate the
-process by which a viral phylogeny is generated along with a transmission tree
-into our method, for example by simulating within-host dynamics. Although this
-may be an avenue for future extension, we felt that it would obscure the
-primary purpose of this work, which is to study contact network parameters.
-Second, there are a number of different methods available for inferring
-transmission trees~\autocite{didelot2014bayesian, ypma2012unravelling,
-jombart2011reconstructing, cottam2008integrating, poon2015phylodynamic}, some
-of which incorporate geographic and/or epidemiological data not accommodated by
-our method. We therefore felt it would be best to allow researchers to use
-their own preferred tree building method.

### 2016-04-16 doc/thesis/discussion.tex
Wrote discussion mostly

\subsection{\software{Netabc}: uses, limitations, and possible extensions}

The method behind \software{netabc} is model-agnostic, meaning that it can be
used to infer parameters of any network model, as long as it allows simulated
networks can be easily generated. We have included generators for the \gls{BA}
model discussed here, as well as the \gls{ER} and \gls{SW} network models.
Instructions for adding additional models are available in the project's online
documentation. We have made \software{netabc} publicly available at
\url{github.com/rmcclosk/netabc} under a permissive open source license, to
encourage other researchers to apply and extend our method.

% spatial networks, see section 5.4 in keeling2005networks

The two-step process of simulating a contact network and subsequently allowing
an epidemic to spread over that network carries with it the assumption that the
contact network is static over the duration of the epidemic. Clearly this
assumption is invalid, as people make and break partnerships on a regular
basis. Our work has not addressed this assumption, primarily due to our desire
to avoid the additional complexity required to address the dynamic nature of
networks. This simplifying assumption is made by most studies using contact
network models in an epidemiological context~\autocite{welch2011statistical,
bansal2007individual}. However, in principle, kernel-\gls{ABC} could be
adapted to dynamic contact networks by using a method such as that developed by
\textcite{robinson2012dynamics} to simulate a dynamic contact network, while
concurrently simulating the spread of an epidemic.

An alternative approach is the deterministic framework outlined by
\textcite{morris1993epidemiology}, who proposes to apply the standard
compartmental modelling framework to contact networks by assigning each
individual their own compartment. Thus, each individual is associated with a
single \gls{ODE}, with the entire \gls{ODE} system parameterized by the
adjacency matrix of the contact network. \citeauthor{morris1993epidemiology}
proposes to use log-linear models to parameterize the matrix. This framework is
highly expressive, and allows straightforward incorporation of time-dependent
dynamics. However, simulating a transmission tree would require the numerical
solution of a very large system of \glspl{ODE}. Given the large number of
simulations required for kernel-\gls{ABC}, it is not clear if this method would
be computationally feasible in this context.

It is important to note that \software{netabc} takes a transmission tree as
input, rather than a viral phylogeny. Thus, we have left the estimation of a
transmission tree up to the user. There were two reasons for this choice. First
and foremost, we wished again to avoid extra complexity and keep the number of
estimated parameters small. In theory, it is possible to incorporate the
process by which a viral phylogeny is generated along with a transmission tree
into our method, for example by simulating within-host dynamics. Although this
may be an avenue for future extension, we felt that it would obscure the
primary purpose of this work, which is to study contact network parameters.
Second, there are a number of different methods available for inferring
transmission trees~\autocite{didelot2014bayesian, ypma2012unravelling,
jombart2011reconstructing, cottam2008integrating, poon2015phylodynamic}, some
of which incorporate geographic and/or epidemiological data not accommodated by
our method. We therefore felt it would be best to allow researchers to use
their own preferred tree building method.

\subsection{Analysis of \acrlong{BA} model}

The preferential attachment power \gls{alpha} had a very strong influence on
tree shape in the range of values we considered. Although the tree kernel was
the most effective classifier for \gls{alpha}, a tree balance statistic
performed nearly as well. This result was intuitive: high \gls{alpha} values
produce networks with few well-connected superspreader nodes which are involved
in a large number of transmissions, resulting in a highly unbalanced
ladder-like tree structure. The \gls{I} parameter, representing the prevalence
at the time of sampling, was also generally estimable. The dynamics of the
\gls{SI} model, and the coalescent process, offer a potential explanation for
this result. In the initial phase of the epidemic, when \gls{I} is small, each
new transmission results in potentially many new discordant edges, thus
decreasing the waiting time until the next transmission. Hence, there is an
early exponential growth phase, producing many short branches near the root of
the tree. As the epidemic gets closer to saturating the network, the number of
discordant edges decays, causing longer waiting times. The distribution of
coalescent times in the tree should therefore be informative about \gls{I}.
This information is captured by the tree kernel, and also by the \gls{nltt}
statistic, which both performed quite will in classifying \gls{I}
(\cref{fig:rsquared}).

The number of nodes in the network, \gls{N}, exhibited the most variation in
terms of being estimable. There was almost no difference between trees
simulated under different \gls{N} values when the number of infected nodes
\gls{I} was very small. There is an intuitive explanation for this result,
namely that adding additional nodes does not change the edge density or overall
shape of a \gls{BA} network. This can be illustrated by imagining that we add a
small number of nodes to a network after the epidemic simulation has already
been completed. It is possible that none of these new nodes attains a
connection to any infected node. Thus, running the simulation again on the new,
larger network could produce the exact same transmission tree as before. On the
other hand, when \gls{I} is large, the coalescent dynamics discussed above for
\gls{I} also apply, as evidenced by the relative accuracy of the \gls{nltt}.
The \gls{m} parameter, which controls the number of connections added to the
network per vertex, did not have a measurable impact on tree shape and was not
estimable with kernel-ABC. The exception to this was the value \gls{m} = 1,
which produces networks without cycles whose associated trees were more easily
distinguished. However, all the analyses presented here did not take the
absolute size of the transmission trees into account, as the branch lengths
were rescaled by their mean. Because higher \gls{m} values imply higher edge
density, an epidemic should spread more quickly for higher \gls{m} than lower
\gls{m} with the same per-edge transmission probability.  Hence, considering
the absolute height of the trees may improve our method's ability to
reconstruct \gls{m}.

In addition to the tree height, many summary statistics have been developed to
capture particular details of tree shape. Two of these, Sackin's index and the
ratio of internal to terminal branch lengths, were correlated with every
\gls{BA} parameter. Classifiers based on Sackin's index and the \gls{nltt}
similarity measure performed well in some cases, though poorly in others.
\gls{ABC} is often applied using a vector of summary statistics, rather than a
kernel-based similarity score as we have done here, and methods have been
developed to select an optimal combination of summary statistics for a given
inference task~\autocite{fearnhead2012constructing}. Hence, an avenue for
future improvement of our method may be the inclusion of additional summary
statistics to supplement the tree kernel. In addition, all four parameters were
more accurately classified when the number of tips in the transmission trees
was larger, underscoring the importance of adequate sampling for accurate
phylodynamic inference.

For the more estimable parameters, the credible intervals attained from the
marginal \gls{ABC} target distributions were much narrower than those obtained
through grid search, while point estimates were of comparable accuracy. This
was likely due to the fact that \gls{SMC} employs importance sampling to
approximate the posterior distribution, while grid search simply calculates a
distance metric which may not have any resemblance to the posterior.
Admittedly, our method of finding credible intervals from kernel scores along
the grid, namely by normalizing the scores to resemble a probability
distribution, was somewhat ad hoc, which may also have played a role.
Regardless, this result indicates that there is benefit to applying the more
sophisticated method, even if values for some of the parameters are known
\textit{a priori}, and especially if credible intervals are desired on the
parameters of interest.

\subsection{Application to HIV data}

The analysis of \gls{HIV} datasets indicated heterogeneity in the contact
network structures underlying several distinct local epidemics. The five
datasets analysed fell into three categories (Figure~\ref{fig:abchpd}). First,
we estimated a preferential attachment power between 0.5 and 1 for the
epidemics studied by \textcite{cuevas2009hiv} and \textcite{li2015hiv}, with
credible intervals occupying nearly the entire region from 0 to 1.
\citeauthor{cuevas2009hiv} studied a group of newly diagnosed individuals in
the Basque Country, Spain. Although the individuals were of mixed risk groups,
and therefore unlikely to comprise a single contact network, a high proportion
of them (47\%) grouped into local clusters based on genetic distance. The low
estimated attachment power for these data is consistent with the sampled
sequences comprising many distinct sub-networks rather than a single connected
network. \citeauthor{li2015hiv} sampled a large number of acutely infected MSM
in Shanghai, China, in which we identified a large cluster from the phylogeny
using a patristic distance cutoff~\autocite{poon2015impact}. The low attachment
power estimated for this dataset was surprising given the high phylogenetic
relatedness of the sequences. It is possible that the number and diversity of
circulating recombinant forms in the data introduced errors into our estimated
viral phylogeny.

For the outbreaks studied by \textcite{niculescu2015recent} and
\textcite{wang2015targeting}, the estimated $\alpha$ was close to one, with a
narrower credible interval than for the other studies.
\citeauthor{niculescu2015recent} studied a recent outbreak among Romanian
injection drug users (IDU), while \citeauthor{wang2015targeting} sampled
acutely infected MSM in Beijing, China. Both studies discovered a high degree
of phylogenetic relatedness owing to the recent infection times and homogeneous
risk groups of the studied populations. The estimated number of infections for
these datasets were also quite low, although the HPD interval for
\citeauthor{wang2015targeting} was much wider than that for
\citeauthor{niculescu2015recent}.

The final studied dataset was an outlier in terms of estimated parameters.
\textcite{novitsky2013phylogenetic} sampled approximately 44\% of the
HIV-infected individuals in the northern area of Mochudi, Botswana. Additional
sampling in a later study~\autocite{novitsky2014impact} brought the genotyping
coverage up to 70\%. Even with such a high sampling coverage, we did not detect
any large clusters using patristic distance, and therefore chose to analyze a
subtree instead. Estimates of $\alpha$ and $N$ both had very wide HPD intervals
and were markedly different from the other datasets. The estimated number of
infected nodes was also extremely high, much higher than the estimated HIV
prevalence of the town. Several factors may have contributed to these results.
First, the authors note that the their sample was 75\% female.  In a primarily
heterosexual risk environment, removal of a disproportionate number of males
from the network could obfuscate the true network structure, for example if the
majority of highly connected nodes were of one gender.  Second, the town in
question was in close proximity to the country's capital, and the authors
indicated that a high amount of migration takes place between the two
locations. This suggests that the contact network may include a much larger
group based in the capital city, which would explain the high estimate of $I$.

Our use of the \gls{BA} model makes several simplifying assumptions. First, we
assume homogeneity across the network with respect to node behaviour and
transmission risk. In reality, the attraction to high-degree nodes seems likely
to vary among individuals, as does their risk of transmitting or contracting
the virus. We have also assumed that all transmission risks are symmetric,
which is clearly false for all known modes of \gls{HIV} transmission, and that
infected individuals never recover but remain infectious indefinitely. These
assumptions were made for the purpose of keeping the model as simple as
possible, since this is the very first attempt to fit a contact network model
in a phylodynamic context. However, the Gillespie simulation algorithm built
into \software{netabc} can handle arbitrary transmission and removal rates
which need not be homogeneous across the network. Moreover, it is possible to
use kernel-\gls{ABC} to fit a model which relaxes some or all of these
assumptions, which may be a fruitful avenue for future investigation.

### 2016-04-16 doc/thesis/results.Rnw
Wrote discussion mostly

@@
+\subsection{Analysis of \acrlong{BA} model}
+
 <<setup, include=FALSE>>=
     library(netabc)
+    library(Hmisc)
 @
 
-\subsection{Kernel classifiers}
+\subsubsection*{Classifiers for parameters based on tree shape}
 
@@
     params <- c("alpha", "I", "m", "N")
+
+    pp <- function (p) {
+        if (p < 1e-5) {
+            "<10^{-5}"
+        } else {
+            latexSN(round(p, -floor(log10(p))))
+        }
+    }
 
@@
     alpha.sackin.cor <- alpha.stats[,cor.test(alpha, sackin, method="spearman")]
+    I.sackin.cor <- I.stats[,cor.test(I, sackin, method="spearman")]
+    m.sackin.cor <- m.stats[,cor.test(m, sackin, method="spearman")]
+    N.sackin.cor <- N.stats[,cor.test(N, sackin, method="spearman")]
     stopifnot(alpha.sackin.cor$p.value < 1e-5)
 
+    alpha.ratio.cor <- alpha.stats[,cor.test(int.tip.ratio, alpha)]
+    I.ratio.cor <- I.stats[,cor.test(int.tip.ratio, I)]
+    m.ratio.cor <- m.stats[,cor.test(int.tip.ratio, m)]
     N.ratio.cor <- N.stats[,cor.test(int.tip.ratio, N)]
-    I.ratio.cor <- I.stats[,cor.test(int.tip.ratio, I)]
+    stopifnot(alpha.ratio.cor$p.value < 1e-5)
+    stopifnot(I.ratio.cor$p.value < 1e-5)
+    stopifnot(m.ratio.cor$p.value < 1e-5)
     stopifnot(N.ratio.cor$p.value < 1e-5)
-    stopifnot(I.ratio.cor$p.value < 1e-5)
 @
 
-Trees simulated under different values of \gls{alpha} are visibly quite
+Trees simulated under different values of \gls{alpha} were visibly quite
 distinct (\cref{fig:alphatrees}). In particular, higher values of \gls{alpha}
@@
 unbalanced, ladder-like structure in the phylogeny, compared to networks with
-lower \gls{alpha} values. Sackin's index was significantly correlated with
-\gls{alpha} 
-    (Spearman's rho \Sexpr{round(alpha.sackin.cor$estimate, 2)},
-     $p < 10^{-5}$).
-None of the other three parameters produced trees which were as easily
-distinguished from each other
-(\cref{fig:Itrees,fig:mtrees,fig:Ntrees,fig:Itrees}). However, the ratio of
-internal to terminal branch lengths was positively correlated with $N$ and
-negatively correlated with $I$ (Spearman's rho
-    \Sexpr{round(N.ratio.cor$estimate, 2)} and
+lower \gls{alpha} values. None of the other three parameters produced trees
+which were as easily distinguished from each other
+(\cref{fig:Itrees,fig:mtrees,fig:Ntrees,fig:Itrees}).
+Sackin's index, which measures tree imbalance, was significantly correlated with
+all four parameters
+    (for $\alpha$, $I$, $m$, and $N$ respectively: Spearman's rho =
+     \Sexpr{round(alpha.sackin.cor$estimate, 2)},
+     \Sexpr{round(I.sackin.cor$estimate, 2)},
+     \Sexpr{round(m.sackin.cor$estimate, 2)},
+     \Sexpr{round(N.sackin.cor$estimate, 2)};
+     $p$-values
+     $\Sexpr{pp(alpha.sackin.cor$p.value)}$,
+     $\Sexpr{pp(I.sackin.cor$p.value)}$,
+     $\Sexpr{pp(m.sackin.cor$p.value)}$,
+     $\Sexpr{pp(N.sackin.cor$p.value)}$)
+The ratio of internal to terminal branch lengths was negatively correlated with
+\gls{alpha} and \gls{I}, and positively corelated with \gls{N} and \gls{m}
+  (Spearman's rho
+    \Sexpr{round(alpha.ratio.cor$estimate, 2)},
     \Sexpr{round(I.ratio.cor$estimate, 2)},
-    both $p < 10^{-5}$).
+    \Sexpr{round(m.ratio.cor$estimate, 2)},
+    \Sexpr{round(N.ratio.cor$estimate, 2)};
+  $p$-values
+    $\Sexpr{latexSN(round(alpha.ratio.cor$p.value, -floor(log10(alpha.ratio.cor$p.value))))}$,
+    $\Sexpr{latexSN(round(I.ratio.cor$p.value, -floor(log10(I.ratio.cor$p.value))))}$,
+    $\Sexpr{latexSN(round(m.ratio.cor$p.value, -floor(log10(m.ratio.cor$p.value))))}$,
+    $\Sexpr{latexSN(round(N.ratio.cor$p.value, -floor(log10(N.ratio.cor$p.value))))}$).
 
@@
 
-\subsection{Grid search}
+\subsubsection*{Marginal parameter estimates with grid search}
+
+<<gridsearch, include=FALSE>>=
+    d <- setDT(collect.data("../../simulations/aggregates/gridsearch/*"))
+    d[,error := abs(true_value - point.est)]
+
+    # correlation between number of tips and error
+    tip.cor <- d[,cor.test(error, tips, method="spearman")[c("p.value", "estimate")], by=parameter]
+    setkey(tip.cor, parameter)
+
+    # CI takes up >90% of grid
+    d[parameter == "alpha", all(upper - lower >= 1.8)]
+    d[parameter == "I", all(upper - lower >= 4050)]
+    d[parameter == "N", all(upper - lower >= 12555)]
+    d[parameter == "m", all(upper - lower >= 5)]
+
+    # correlation between value and error
+    value.cor <- d[,cor.test(error, true_value)[c("p.value", "estimate")], by=parameter]
+    setkey(value.cor, parameter)
+    stopifnot(value.cor["I", p.value] >= 0.05)
+@
 
 The accuracy of grid search estimates largely paralleled that of the \gls{kSVM}
-classifiers.
+classifiers. \Cref{fig:gridest} shows point estimates and 95\% highest density
+intervals for each of the \gls{BA} parameters, for one replicate experiment
+with 500-tip trees. For all parameters except $m$, the error of point estimates
+was negatively correlated with the number of sampled tips in the tree (for
+\gls{alpha}, \gls{I}, and \gls{N} respectively: Spearman's $\rho$ = 
+    \Sexpr{tip.cor["alpha", round(estimate, 2)]},
+    \Sexpr{tip.cor["I", round(estimate, 2)]},
+    \Sexpr{tip.cor["N", round(estimate, 2)]};
+$p$-value = 
+    $\Sexpr{tip.cor["alpha", latexSN(round(p.value, -floor(log10(p.value))))]}$,
+    $\Sexpr{tip.cor["I", latexSN(round(p.value, -floor(log10(p.value))))]}$,
+    $\Sexpr{tip.cor["N", latexSN(round(p.value, -floor(log10(p.value))))]})$.
+The highest density intervals obtained for all parameters were extremely wide,
+occcupying $>$90\% of the grid in all cases.
 
-We used grid search to obtain marginal estimates for each network parameter
-while holding all other parameters fixed. We observed that kernel scores were
-highest at the values of \gls{alpha} on the grid closest to the true values, as
-shown in Figure~\ref{fig:gridkernel}. However, there was a much stronger spike
-in kernel scores near the true value for \gls{alpha} = 1.0 and 1.25. This is
-recapitulated when we look at the accuracy of point estimates obtained by
-taking the grid value with the highest median kernel score. As shown in
-Figure~\ref{fig:gridest}, while the estimates are generally close to the true
-value, they are much closer for \gls{alpha} = 1.25 than for the other values.
+The \gls{alpha} parameter was the most accurately estimated, with point
+estimates having an average deviation of 
+    \Sexpr{d[parameter == "alpha", round(mean(error), 2)]}
+from the true value, on a grid from 0 to 2. The error was negatively correlated
+with the true value of \gls{alpha} 
+    (Spearman's $\rho$ = \Sexpr{value.cor["alpha", round(estimate, 2)]},
+     $p = \Sexpr{value.cor["alpha", latexSN(round(p.value, -floor(log10(p.value))))]}$),
+although the relationship was clearly nonlinear (\cref{fig:gridest,fig:gridalpha}).
+The accuracy was highest for the test value \gls{alpha} = 1.25
+    (mean error \Sexpr{d[parameter == "alpha" & true_value == 1.25, round(mean(error), 2)]})
+which exhibited markedly different behaviour than the other values in terms of
+the distribution of kernel scores along the grid (\cref{fig:gridalpha}). In
+particular, there was a very pronounced peak in scores around the true value,
+in contrast to most other values where the scores were flat around the true
+value. The peak was also observed to a lesser degree for \gls{alpha} = 1. The
+average absolute error of the point estimates for \gls{I} was 
+    \Sexpr{d[parameter == "I", round(mean(error))]},
+on a grid of 500 to 5000, and the errors were not significantly correlated with
+the true value of \gls{I}. Kernel score distributions for all test values
+exhibited the same rounded shape (\cref{fig:gridI}).
+
+For the other two parameters, \gls{m} and \gls{N}, accuracy was much poorer
+(\cref{fig:gridest,fig:gridm,fig:gridN}). The average error for \gls{m} was
+    \Sexpr{d[parameter == "m", round(mean(error), 2)]},
+on a grid from 1 to 6; this error was positively correlated with increasing
+\gls{m}
+    (Spearman's $\rho$ = \Sexpr{value.cor["m", round(estimate, 2)]},
+     $p = \Sexpr{value.cor["m", latexSN(round(p.value, -floor(log10(p.value))))]}$).
+This positive correlation was apparently due to the much lower error for $m =
+1$ than for the other $m$ values (\cref{fig:gridm})
+    (mean errors \Sexpr{d[parameter == "m" & true_value == 1, mean(error)]} for
+     $m = 1$ vs. \Sexpr{d[parameter == "m" & true_value > 1, mean(error)]} for
+     $m > 1$).
+The value $m = 1$ causes the network to take on a distinct shape relative to
+higher $m$ values, namely a tree (\textit{i.e.} there are no cycles,
+\cref{subsec:treeshape}). The average error for \gls{N} was 
+    \Sexpr{d[parameter == "N", round(mean(error))]},
+on a grid from 1000 to 15000, and was positively correlated with the true value
+of \gls{N}
+    (Spearman's $\rho$ = \Sexpr{value.cor["N", round(estimate, 2)]},
+     $p = \Sexpr{value.cor["N", latexSN(round(p.value, -floor(log10(p.value))))]}$).
 
@@
   \centering
-  \includegraphics{gridsearch-alpha-kernel.pdf}
-  \caption[Grid search kernel scores]{
-    Grid search kernel scores for testing trees simulated under various
-    \gls{alpha} values. All epidemics had \gls{I} = 1000 infected nodes, on
-    \gls{BA} networks of size \gls{N} = 5000 with \gls{m} fixed at 2. Colours
-    indicate the number of sampled tips.
-  }
-  \label{fig:gridkernel}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{gridsearch-alpha-point-estimate.pdf}
-  \caption[Marginal estimates of \gls{alpha} obtained with grid search]{
-    Marginal estimates of \gls{alpha} obtained with grid search. Training trees
-    were simulated on a narrowly spaced grid of \gls{alpha} values, and
-    compared to testing trees using the tree kernel. The \gls{alpha} value in
-    the grid with the highest median kernel score was taken as the point
-    estimate for the testing tree. These point estimates are shown as black
-    dots. The dashed line is the identity.
+  \includegraphics[width=\textwidth]{gridsearch-example}
+  \caption[Grid search estimates of \gls{BA} model parameters]{Point estimates
+      and 95\% highest density intervals for each \gls{BA} model parameter,
+      obtained using grid search. Networks and transmission trees were
+      simulated over a grid of values for each parameter while holding the
+      others fixed. For a subset of the grid values ($x$-axis), test networks
+      and trees were created and compared to each tree on the grid using the
+      tree kernel. The kernel scores along the grid were normalized to resemble
+      a probability distribution, from which the mode and highest density
+      interval were calculated. Shown values correspond to one replicate
+      experiment, with trees of size 500.
   } 
@@
 
-\subsection{Accuracy of estimates with full ABC}
+\subsubsection*{Joint parameter estimates with kernel-ABC}
 
@@
 nonzero prior density, whereas the intervals for $m$ and \gls{N} were widely
-dispersed. Table~\ref{tab:abchpd} shows point estimates and 95\% \gls{HPD}
+dispersed. \Cref{tab:abchpd} shows point estimates and 95\% \gls{HPD}
 intervals averaged over all simulations.
@@
 
-\subsection{Characterization of power-law exponent in Barab\'asi-Albert networks}
+\subsubsection*{Effect of parameters on power-law exponent}
 
@@
 \end{table}
+
+\subsection{Application to HIV data}
+
+<<realdata, include=FALSE>>=
+  d <- fread(Sys.glob("../../simulations/aggregates/hpd/*"))
+  map <- dcast.data.table(d, dataset~parameter, value.var="map")
+  up <- dcast.data.table(d, dataset~parameter, value.var="hpd.max")
+  low <- dcast.data.table(d, dataset~parameter, value.var="hpd.min")
+  setkey(map, dataset)
+  setkey(up, dataset)
+  setkey(low, dataset)
+@
+
+We applied kernel-ABC to five published HIV datasets (\cref{tab:data}),
+and found substantial heterogeneity among the parameter estimates
+(\cref{fig:abchpd}). Two of the datasets~\autocite{niculescu2015recent,
+wang2015targeting} had estimated $\alpha$ values near unity (MAP estimates
+[95\% HPD] 
+  \Sexpr{map["niculescu2015", round(alpha, 2)]} 
+  [\Sexpr{low["niculescu2015", round(alpha, 2)]} - 
+   \Sexpr{up["niculescu2015", round(alpha, 2)]}]
+and
+  \Sexpr{map["wang2015", round(alpha, 2)]} 
+  [\Sexpr{low["wang2015", round(alpha, 2)]} -
+   \Sexpr{up["wang2015", round(alpha, 2)]}] respectively). 
+Another two datasets~\autocite{li2015hiv, cuevas2009hiv} had lower estimated
+values and wider HPD intervals
+  (\Sexpr{map["li2015", round(alpha, 2)]} 
+  [\Sexpr{low["li2015", round(alpha, 2)]} - 
+  \Sexpr{up["li2015", round(alpha, 2)]}]
+and
+  \Sexpr{map["cuevas2009", round(alpha, 2)]} 
+  [\Sexpr{low["cuevas2009", round(alpha, 2)]} -
+   \Sexpr{up["cuevas2009", round(alpha, 2)]}]). 
+The \textcite{novitsky2014impact} data had an extremely low estimated $\alpha$
+and a very wide HPD interval
+  (\Sexpr{map["novitsky2014", round(alpha, 2)]} 
+  [\Sexpr{low["novitsky2014", round(alpha, 2)]} -
+   \Sexpr{up["novitsky2014", round(alpha, 2)]}]). 
+For all the datasets except \citeauthor{novitsky2014impact}, estimated values of
+$I$ were below 2000, with narrow HPD intervals around two of the
+datasets
+  (\citeauthor{cuevas2009hiv}, \Sexpr{map["cuevas2009", floor(I)]} 
+  [\Sexpr{low["cuevas2009", floor(I)]} -
+   \Sexpr{up["cuevas2009", floor(I)]}];
+   \citeauthor{niculescu2015recent}, \Sexpr{map["niculescu2015", floor(I)]}
+  [\Sexpr{low["niculescu2015", floor(I)]} - 
+   \Sexpr{up["niculescu2015", floor(I)]}])
+and wider intervals around the other two
+  (\citeauthor{li2015hiv}, \Sexpr{map["li2015", floor(I)]} 
+  [\Sexpr{low["li2015", floor(I)]} -
+   \Sexpr{up["li2015", floor(I)]}];
+   \citeauthor{wang2015targeting}, \Sexpr{map["wang2015", floor(I)]}
+  [\Sexpr{low["wang2015", floor(I)]} - 
+   \Sexpr{up["wang2015", floor(I)]}]).
+The \citeauthor{novitsky2014impact} data was again the outlier, with a very high
+estimated $I$, and HPD interval spanning almost the entire prior region
+  (\Sexpr{map["novitsky2014", floor(I)]} 
+  [\Sexpr{low["novitsky2014", floor(I)]} -
+   \Sexpr{up["novitsky2014", floor(I)]}]).
+No information was gleaned about the $m$ parameter, with the HPD interval
+occupying the entire prior region for all datasets. The estimates of $N$ were
+similarly uninformative, with the exception that the point estimate for the
+\citeauthor{wang2015targeting} data was smaller
+  (\Sexpr{map["wang2015", floor(N)]})
+than the estimates for other datasets
+ (average \Sexpr{map[dataset != "wang2015", floor(mean(N))]}).
+
+\begin{figure*}[ht]
+  \centering
+  \includegraphics{realdata-hpd-bc.pdf}
+  \vspace{8pt}
+  \caption{
+      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
+      parameters of the BA network model, fitted to five published HIV datasets
+      with kernel-ABC.
+  }
+  \label{fig:abchpd}
+\end{figure*}

### 2016-04-18 doc/thesis/discussion.tex
More writing

@@
 encourage other researchers to apply and extend our method.
+
+In retrospect, there are a few algorithmic improvements which could have been
+made to \software{netabc}. Firstly, our implementation uses simple multinomial
+sampling for the resampling step, as well as for generating the final sample
+from the target distribution.  Several other sampling strategies have been
+developed~\autocite{douc2005comparison}, and it is possible that the use of a
+more sophisticated scheme might increase the algorithm's accuracy. Secondly,
+the model parameters \gls{I}, \gls{N}, and \gls{m} are all integer-valued, but
+we treat them as continuous-valued when performing \gls{SMC}. This is not
+inherently wrong, in that it simply assigns equal probability density to the
+interval $[i, i+1)$ for all integers $i$. However, computational effort could
+be wasted if particles are moved around within these intervals. To improve
+efficiency, alternative proposals could be used for discrete-valued particles.
 

### 2016-04-18 doc/thesis/intro.tex
More writing

@@
 The \defn{tips} of a phylogeny, that is, the nodes without any descendants,
-correspond to \defn{extant}, or observed, taxa, while the \defn{internal nodes}
-correspond to their common ancestors. The edges or \defn{branches} of the
-phylogeny connect ancestors to their descendants. Phylogenies may have a
-\defn{root}, which is a node with no descendants distinguished as the most
-recent common ancestor of all the extant
+correspond to \defn{extant}, or observed, taxa. The \defn{internal nodes}
+correspond to their (usually extinct) common ancestors. The edges or
+\defn{branches} of the phylogeny connect ancestors to their descendants.
+Phylogenies may have a \defn{root}, which is a node with no descendants
+distinguished as the most recent common ancestor of all the extant
 taxa~\autocite{harding1971probabilities}. When such a root exists, the tree is
@@
 evolutionary process (speciation), they record an epidemiological process
-(transmission). The tips of a transmission tree represent infected hosts, while
-internal nodes correspond to transmissions from one host to another.
-Transmission trees generally have branch lengths in units of calendar time,
-with branching times indicating times of transmission. The root of a
-transmission tree corresponds to the initially infected patient who introduced
-the epidemic into the network, also known as the \defn{index case}. The
-internal nodes may be labelled with the donor of the transmission pair, if this
-is known. The tips of the tree, rather than being fixed at the present day, are
-placed at the time at which the individual was removed from the epidemic, such
-as by death, recovery, isolation, behaviour change, or migration. Consequently,
-the transmission tree may not be ultrametric, but may have tips located at
-varying distances from the root. Such trees are said to have
-\defn{heterochronous} taxa~\autocite{drummond2003measurably}, in contrast to
-the \defn{isochronous} taxa found in most phylogenies of macro-organisms. A
+(transmission). The tips of a transmission tree represent the removal of
+infected hosts, while internal nodes correspond to transmissions from one host
+to another.Transmission trees generally have branch lengths in units of
+calendar time, with branching times indicating times of transmission. The root
+of a transmission tree corresponds to the initially infected patient who
+introduced the epidemic into the network, also known as the \defn{index case}.
+The internal nodes may be labelled with the donor of the transmission pair, if
+this is known. The tips of the tree, rather than being fixed at the present
+day, are placed at the time at which the individual was removed from the
+epidemic, such as by death, recovery, isolation, behaviour change, or
+migration. Consequently, the transmission tree may not be ultrametric, but may
+have tips located at varying distances from the root. Such trees are said to
+have \defn{heterochronous} taxa~\autocite{drummond2003measurably}, in contrast
+to the \defn{isochronous} taxa found in most phylogenies of macro-organisms. A
 transmission tree is illustrated in \cref{fig:contactnet} (right).
 
-Due to the internal nodes, each infected individual in an epidemic may appear
-in the transmission tree more than once. This is different from the
-transmission \emph{network}, in which each infected individual appears exactly
-once, and edges are in one-to-one correspondence with
-transmissions~\autocite{welch2011statistical, keeling2005networks}.
-Transmission networks are discussed further in \cref{sec:contactnet}, and the
-distinction between the two objects is illustrated in \cref{fig:contactnet}.
-However, since transmission networks generally have no cycles (unless
-re-infection occurs), they are trees in the graph theoretical sense, and hence
-are sometimes also referred to as transmission
-trees~\autocite[\eg][]{kenah2015algorithms}. In this work, we reserve the term
-``transmission tree'' for the objects depicted on the right side of
-\cref{fig:contactnet}, following~\autocite[\eg][]{stadler2013uncovering}. The
-term ``transmission network'' is taken to mean the subgraph of the contact
-network along which transmissions occurred,
-following~\autocite{welch2011statistical, keeling2005networks}.
+Each infected individual in an epidemic may appear in the transmission tree
+more than once. This is different from the transmission \emph{network}, in
+which each infected individual appears exactly once, and edges are in
+one-to-one correspondence with transmissions~\autocite{welch2011statistical,
+keeling2005networks}. Transmission networks are discussed further in
+\cref{sec:contactnet}, and the distinction between the two objects is
+illustrated in \cref{fig:contactnet}. However, since transmission networks
+generally have no cycles (unless re-infection occurs), they are trees in the
+graph theoretical sense, and hence are sometimes also referred to as
+transmission trees~\autocite[\eg][]{kenah2015algorithms}. In this work, we
+reserve the term ``transmission tree'' for the objects depicted on the right
+side of \cref{fig:contactnet}, following \eg~\autocite{stadler2013uncovering}.
+The term ``transmission network'' is taken to mean the subgraph of the contact
+network along which transmissions occurred, following
+\eg~\autocite{welch2011statistical, keeling2005networks}.
 
@@
 new infections over the course of an epidemic~\autocite{holmes1995revealing}.
-Many more diverse epidemiological parameters have been investigated using
-transmission trees, such as the degree of
-clustering~\autocite{hughes2009molecular} and the effect of elevated
-transmission risk in acute infection~\autocite{volz2012simple}. However, in all
-but the most well-studied of epidemics, this is not possible to obtain through
-traditional epidemiological methods~\autocite{welch2011statistical}. The time
-and effort to conduct detailed interviews and contact tracing of a sufficient
-number of infected individuals is usually prohibitive. Even when the resources
-for such methods are available, patients may not always recall whom they
-contacted and when, especially in the case of airborne transmission.
-Consequently, the transmission tree must be estimated using other methods. Most
-commonly, this is done by exploiting the relationship between transmission
-trees and viral phylogenies~\autocite{volz2013viral}.
-
-\subsection{Relationship between transmission trees and viral phylogenies}
+However, in all but the most well-studied of epidemics, transmission trees are
+not possible to assemble through traditional epidemiological
+methods~\autocite{welch2011statistical}. The time and effort to conduct
+detailed interviews and contact tracing of a sufficient number of infected
+individuals is usually prohibitive, and may be additionally be confounded by
+misreporting~\autocite{eames2015six}. However, it turns out that for viral
+epidemics, some of the epidemiological information contained in the
+transmission tree leaves a mark on the viral genetic material circulating in
+the population. A family of methods called
+\defn{phylodynamics}~\autocite{grenfell2004unifying} addresses the challenge of
+estimating epidemiological parameters from viral sequence
+data~\autocite{volz2013viral}.
+
+\subsection{Phylodynamics: linking evolution and epidemiology}
+
+%Most modern analyses use model-based methods, which simultaneously
+%estimate the phylogeny with branch lengths and the parameters of a model of
+%evolution. Although they usually work well in practice, the estimated topology
+%can vary based on the model used and, in the case of Bayesian analysis, the
+%priors. In addition, intra-host viral populations are genetically
+%heterogeneous, so choosing a single representative genotype per host is
+%necessarily imprecise. One can use either the genotype of a specific virion
+%sampled from the host, or a synthetic genotype, such as a consensus or
+%reconstructed ancestral sequence.
 
 In general, \defn{viral phylogenies} are simply phylogenetic trees relating
-virus strains. In phylodynamics, we often consider \defn{inter-host}
-phylogenies, which relate one viral genotype from each host in a population.
-The crux of \defn{phylodynamics}~\autocite{grenfell2004unifying} is the fact
-that the epidemiological processes recorded in transmission trees, and the
-evolutionary processes recorded in viral phylogenies, occur on similar time
-scales for RNA viruses~\autocite{drummond2003measurably}. As a result, there is
-a close relationship between the two types of tree. In particular, the
-transmission process is quite similar to \defn{allopatric
+virus strains. \defn{Inter-host} phylogenies relate one representative viral
+genotype from each host in a population. In this work, when we use the phrase
+``viral phylogeny'', we are referring to an inter-host phylogeny unless
+otherwise stated. Phylogenetic inference is a complex topic which we will not
+discuss here (see \eg \autocite{nei2000molecular} for a full review). We assume
+here that the viral phylogeny is known, a common assumption in the development
+and application of phylodynamic methods.
+
+The crux of phylodynamics is the fact that the epidemiological processes
+recorded in transmission trees, and the evolutionary processes recorded in
+viral phylogenies, occur on similar time scales for RNA
+viruses~\autocite{drummond2003measurably}. Phylodynamic methods have been used
+to investigate epidemiological parameters such as transmission rate, recovery
+rate, and basic reproductive number~\autocite{pybus2009evolutionary,
+volz2013viral}. The majority of phylodynamic studies attempt to infer the
+parameters of an epidemiological model for which the likelihood of an observed
+phylogeny can be calculated. Most often, this is some variation of the
+birth-death~\autocite{kendall1948generalized, stadler2012estimating} or
+coalescent~\autocite{kingman1982coalescent, volz2012complex} models.
+
+Due to the relationship between the aforementioned processes, there is a close
+relationship between viral phylogenies and transmission trees. In particular,
+the transmission process is quite similar to \defn{allopatric
 speciation}~\autocite{coyne2004speciation}, where genetic divergence follows
@@
 
-A final caveat is that the viral phylogeny itself is not known with certainty,
-so it must also be estimated from genetic data. Phylogenetic inference is a
-complex topic which we will not discuss in detail here (see e.g.
-\autocite{nei2000molecular} for a full review). Most modern analyses use
-model-based methods, which simultaneously estimate the phylogeny with branch
-lengths and the parameters of a model of evolution. Although they usually work
-well in practice, the estimated topology can vary based on the model used and,
-in the case of Bayesian analysis, the priors.  In addition, intra-host viral
-populations are genetically heterogeneous, so choosing a single representative
-genotype per host is necessarily imprecise. One can use either the genotype of
-a specific virion sampled from the host, or a synthetic genotype, such as a
-consensus or reconstructed ancestral sequence.
-
 What we have just discussed is a two-step procedure for estimating the
@@
 trajectories~\autocite{janzen2015approximate}.
+
+\textcite{poon2013mapping} developed an alternative pairwise measure which
+applies the concept of a \defn{kernel function} to phylogenies. Kernel
+functions, originally developed for \glspl{SVM}~\autocite{burges1998tutorial}, 
+compare objects by mapping them into a feature space of very high, possibly
+even infinite, dimension.
 
@@
 will be discussed further in \cref{sec:abc}.
-
-\subsection{Applications of phylodynamics}
-\label{subsec:phyloapp}
-
-Phylodynamic methods have been used to investigate epidemiological parameters
-such as transmission rate, recovery rate, and basic reproductive
-number~\autocite{pybus2009evolutionary, volz2013viral}. These studies make
-inferences about epidemiological processes from the genetic diversity of virus
-populations, which is usually represented in the form of a phylogeny. The
-majority of these employ a Bayesian \gls{MCMC} approach to infer parameters of
-an epidemiological model whose likelihood can be calculated, most often some
-variation of the birth-death~\autocite{kendall1948generalized} or
-coalescent~\autocite{kingman1982coalescent} models.
-\textcite{stadler2012estimating} develop a formula for the likelihood of a
-phylogeny with heterochronous tips under the birth-death model, which has been
-used to estimate the basic reproductive number of several viral
-epidemics~\autocite{stadler2012estimating}. However, the birth-death model is
-cannot tell us anything about population structure, as it assumes that every
-individual becomes infected at the same rate. \textcite{volz2012complex} writes
-down the likelihood of a heterochronous phylogeny under a coalescent model with
-arbitrarily complex population dynamics. This opens the door to more complex
-inferences about population structure, as the population can be partitioned
-into compartments with different transmission and recovery rates, but still
-assumes that each compartment is homogeneously mixed. In other words, the
-coalescent model can tell us about the \emph{global} structure of a population,
-such as whether there exists a high-risk subgroup, but not about the
-\emph{local} structure, such as the average number of contacts each individual
-has.
 
@@
 
+To fully describe \gls{SMC}, we will introduce some notation and terminology.
+For a sequence $x_1, \ldots, x_d$, we will write $\vec{x_i}$ to mean the
+partial sequence $x_1, \ldots, x_i$.
+
+A \defn{Markov kernel} is the continuous analogue of the transition matrix in a
+finite-state Markov model. For some spaces $X$ and $Y$, $K \maps X \times Y \to
+[0, 1]$ such that
+\begin{align}
+    \label{eq:mk}
+    \int_Y K(x, y) \d y = 1
+\end{align}
+for all $x \in X$. This is an ``operational'' definition of Markov kernel which
+will be suitable for our purposes. A more precise definition can be found in
+\textbf{TODO}. Note that a Markov kernels have nothing to do with the kernel
+functions defined in \cref{subsec:treeshape}, other than sharing a name.
+Unfortunately, the word ``kernel'' is ubiquitous in mathematics and refers to
+many unrelated concepts.
+
 \subsection{Sequential importance sampling}
@@
 \subsection{Sequential Monte Carlo samplers}
+\label{subsec:smcsamp}
 
@@
 \end{align}
-where $K_i$ is any Markov kernel and the integral is over the space on which the
+where $K_i$ is a Markov kernel and the integral is over the space on which the
 $\pi_i$ are defined. The choice of $K_i$ should be based on the percieved
@@
 drawback is that the importance distribution itself becomes intractible. In
-particular, evaluating $\eta_d(x)$ involves a $d$-dimensional integral of the
-type \cref{eq:impint}. As it is necessary to evaluate $\eta(x)$ pointwise to
+particular, evaluating $\eta_i(x)$ involves a $i$-dimensional integral of the
+type in \cref{eq:impint}. As it is necessary to evaluate $\eta(x)$ pointwise to
 perform \gls{IS}, this construction appears to have defeated the purpose of
@@
 \textcite{del2006sequential} overcome this problem with two ``artificial''
-objects. First, they propose the existence of \textit{backward} kernels
-$L_{i-1}(x_i, x_{i-1})$ which perform the inverse operation as $K_i$, namely
-transforming $\eta_i$ back into $\eta_{i-1}$. As we shall see, these backward
-kernels do not need to be calculated analytically in terms of $K_i$. Second,
-they define a sequence of $i$-dimensional target distributions
-\[
-  \tilde{\pi}_i(\vec{x_i}) = \frac{\gamma_i(x_i)}{Z_i} \prod_{k=1}^{i-1} L_k(x_{k+1}, x_k)
-\]
-which admit $\pi_i(\vec{x_i})$ as marginals.
+objects. First, they propose the existence of \textit{backward} Markov kernels
+$L_{i-1}(x_i, x_{i-1})$. For now, these kernels are arbitrary, and will be
+precisely defined on a problem-specific basis. Second, they define an
+alternative sequence of target distributions
+\[
+  \tilde{\pi}_i(\vec{x_i}) = \pi_i(x_i) \prod_{k=1}^{i-1} L_k(x_{k+1}, x_k)
+\]
+of increasing dimension. This brings us back to the setting described above in
+\cref{subsec:sis}, namely of building up an importance distribution of
+dimension $d$ sequentially through lower-dimensional distributions. We can
+write $\tilde{\pi}_i$ in terms of $\tilde{\pi}_{i-1}$ by noticing that
+\begin{align*}
+  \frac{\tilde{\pi}_i(\vec{x_i})}{\tilde{\pi}_{i-1}(\vec{x_{i-1}})} 
+  = \frac{\pi_i(x_i) \prod_{k=1}^{i-1} L(x_{k+1}, x_k)}
+  {\pi_{i-1}(x_{i-1}) \prod_{k=1}^{i-2} L(x_{k+1}, x_k)}
+  = \frac{\pi_i(x_i) L(x_i, x_{i-1})}{\pi_{i-1}(x_{i-1})},
+\end{align*}
+and hence
+\[
+  \tilde{\pi}_i = \frac{\pi_i(x_i) L(x_i, x_{i-1})}{\pi_{i-1}(x_{i-1})} \cdot \tilde{\pi}_{i-1}.
+\]
+Therefore, the importance weights for these new targets are defined recursively as
+\begin{align}
+  w(\vec{x_i}) 
+    &= \frac{\tilde{\pi}_i(\vec{x_i})}{\eta_i(\vec{x_i})} \\
+    &= \frac{\tilde{\pi}_{i-1}(\vec{x_{i-1}}) \pi_i(x_i) L(x_i, x_{i-1})}
+           {\eta_{i-1}(\vec{x_{i-1}}) \pi_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)} \\
+    &= w(\vec{x_{i-1}}) \cdot
+      \frac{\pi_i(x_i) L_{i-1}(x_i, x_{i-1})}
+           {\pi_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)} \\
+    &\propto w(\vec{x_{i-1}}) \cdot
+      \frac{\gamma_i(x_i) L_{i-1}(x_i, x_{i-1})}
+           {\gamma_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}.
+    \label{eq:smcwt}
+\end{align}
+The final key piece of information is to notice that, because the $L_i$ are
+Markov kernels, $\pi_i$ is simply the marginal in $\vec{x_{i-1}}$ of
+$\tilde{\pi}$. Therefore, a sample from $\tilde{\pi}_i$ automatically gets us a
+sample from $\pi_i$, by considering only the $i$th component of $\vec{x_i}$.
+These are all the ingredients we need to apply \gls{SMC}. The sequences of
+kernels $L$ and $K$ should be chosen based on the problem at hand to inimize
+the variance in the importance weights as well as possible. For a fixed choice
+of $K_i$, the backward kernels $L_i$ which minimize this variance are called
+the \defn{optimal} backward kernels. The full \gls{SMC} sampler algorithm is
+presented as \cref{alg:smcsamp}.
+
+\begin{algorithm}
+  \caption{Sequential Monte Carlo sampler of \textcite{del2006sequential}.}
+  \begin{algorithmic}
+    \For {$k = 1$ to $n$}
+      \State Sample $x_1^{(k)}$ from $\eta_1(x_1)$
+      \Comment{Initialize the $k$th particle}
+      \State $w^{(k)} \gets \dfrac{\gamma_1\left(x_1^{(k)}\right)}{\eta_1\left(x_1^{(k)}\right)}$
+      \State Normalize the weights so that $\sum w = 1$
+    \EndFor
+    \For {$i = 2$ to $d$}
+      \For {$k = 1$ to $n$}
+        \State Sample $x_i^{(k)}$ from $K(x_{i-1}^{(k)}, x_i)$
+        \Comment Extend the $k$th particle
+        \State $w^{(k)} \gets w^{(k)} \cdot \dfrac{\gamma_i(x_i) L_{i-1}(x_i, x_{i-1})}{\gamma_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}$
+      \EndFor
+      \State Normalize the weights so that $\sum w = 1$
+      \If {$\ESS(w) < T$}
+        \State Resample the particles according to $w$
+        \For {$k = 1$ to $n$}
+          \State $w^{(k)} \gets 1/n$
+        \EndFor
+      \EndIf
+      \State Sample the $i$th component of $n$ particles with probabilities $w$
+    \EndFor
+  \end{algorithmic}
+  \label{alg:smcsamp}
+\end{algorithm}
 
@@
 offers an alternative metric for parameter credibility, namely the similarity
-simulated datasets to the observed data. If datasets simulated under the model
-closely resemble the real data, it follows that the model is a reasonable
+of simulated datasets to the observed data. If datasets simulated under the
+model closely resemble the real data, it follows that the model is a reasonable
 approximation to the real-world process generating the observed data. More
@@
 Algorithms for performing \gls{ABC} fall into one of three categories:
-rejection, \gls{MCMC}, and \gls{SMC}. Rejection ABC is the simplest method, and
-also the one which was first proposed~\autocite{rubin1984bayesianly,
-tavare1997inferring}. The algorithm, outlined in \cref{alg:abcrej}, repeats the
-following steps until a desired number of samples from the target distribution
-are obtained. Parameter values $\theta$ are sampled according to the prior
-distribution $\Pr(\theta)$. Then, a simulated dataset $\vec{z}$ is generated
-from the model with the sampled parameter values. By definition, the
-probability density of obtaining the particular dataset $\vec{z}$ is
-$\Pr(\vec{z} \mid \theta)$. Finally, the parameters are sampled if the distance
-of $\vec{z}$ from the observed data $\vec{y}$ is less than $\varepsilon$ - that
-is, with probability $\I_{A_{\varepsilon, \vec{y}}}(\vec{z})$. Putting this all
-together, the parameters $\theta$ are sampled with probability proportional to
+rejection, \gls{MCMC}, and \gls{SMC}. To simplify the math, we shall restrict
+the descriptions of these algorithms to the case of one simulated dataset per
+parameter particle (the meaning of this will become clear shortly). The
+extension to multiple datasets per particle will be given at the end.
+
+Rejection ABC is the simplest method, and also the one which was first
+proposed~\autocite{rubin1984bayesianly, tavare1997inferring}. The algorithm,
+outlined in \cref{alg:abcrej}, repeats the following steps until a desired
+number of samples from the target distribution are obtained. Parameter values
+$\theta$ are sampled according to the prior distribution $\Pr(\theta)$. Then, a
+simulated dataset $\vec{z}$ is generated from the model with the sampled
+parameter values. By definition, the probability density of obtaining the
+particular dataset $\vec{z}$ is $\Pr(\vec{z} \mid \theta)$. Finally, the
+parameters are sampled if the distance of $\vec{z}$ from the observed data
+$\vec{y}$ is less than $\varepsilon$ - that is, with probability
+$\I_{A_{\varepsilon, \vec{y}}}(\vec{z})$. Putting this all together, the
+parameters $\theta$ are sampled with probability proportional to
 \[
@@
 The most recently developed class of algorithm for \gls{ABC} is
-\gls{ABC}-\gls{SMC}~\autocite{sisson2007sequential, 2009adaptive}. As with
-\gls{ABC}-\gls{MCMC}, the algorithm is a straightforward modification of an
-existing Bayesian inference method, in this case \gls{SMC} (\cref{sec:smc}).
-
-%As an example of such a model, we will review the first practical application
-%of \gls{ABC} by \textcite{tavare1997inferring}. The aim of the investigation
-%was to calculate the posterior distribution on the coalescence time $t$ of some
-%observed sequences $D$ under the coalescent model
-%\autocite{kingman1982coalescent} with the infinite-sites assumption
-%\autocite{watterson1975number}. Their estimation method relied on two key
-%observations. First, the number $k$ of \defn{segregating sites} in the data is
-%informative about the coalescence time while being much easier to work with
-%than the data itself. A segregating site is a position which is polymorphic in
-%the observed sequences. Using this observation, the authors replaced their
-%posterior distribution of interest, $\Pr(t \mid D)$, by
-%\[
-%  \Pr(t \mid k) \propto \Pr(t) \Pr(k \mid t),
-%\]
-%where again we abuse $\Pr$ to refer to a probability density. Second,
-%the authors observed that the expected number of segregating sites depends on
-%the total branch length $l$ of the tree, rather than its height. This enabled
-%them to write an alternative expression for the product of the likelihood and
-%prior in terms of $l$,
-%\[
-%  \Pr(t) \Pr(k \mid t) = \int_0^\infty \Pr(t, l) \Pr(k \mid l) \d l.
-%\]
-%While this integral may be difficult to evaluate analytically, the power of
-%this formulation is the fact that it is straightforward to \emph{simulate}
-%values of $t$ and $l$ according to $\Pr(t, l)$, their joint probability under
-%the coalescent model. One simply runs the (stochastic) coalescent process and
-%observing the values $\hat{t}$ and $\hat{l}$ which result. It is
-%straightforward to calculate $\Pr(k \mid \hat{l})$, the probability
-%that $k$ segregating sites were observed in a tree with total branch length
-%$\hat{l}$, using the Poisson distribution. Hence if $\hat{l}$ and $\hat{t}$,
-%having been generated according to $\Pr(t, l)$, are sampled with probability
-%proportional to $\Pr(k \mid \hat{l})$, they represent an unbiased sample from
-%the above integral. Taking a large number of such samples allowed the authors
-%to obtain a Monte Carlo estimate of $\Pr(t \mid k)$.
+\gls{ABC}-\gls{SMC}~\autocite{sisson2007sequential, beaumont2009adaptive}. As
+with \gls{ABC}-\gls{MCMC}, the algorithm is a straightforward modification of
+an existing Bayesian inference method, in this case the \gls{SMC} sampler
+(\cref{subsec:smcsamp}). The sequence of target distributions is defined as
+$\pi_i = \Pr(\theta, \vec{z} \mid \vec{y}, \varepsilon_i)$ for a decreasing
+sequence of tolerances $\varepsilon_i$. The intention is for the algorithm to
+progress smoothly through a sequence of target distributions which ends at the
+\gls{ABC} approximation to the posterior. As discussed in
+\cref{subsec:smcsamp}, the choices of the kernels $K$ and $L$ is
+problem-specific, and so appropriate kernels must be chosen for \gls{ABC}.
+Several options have been proposed~\autocite{beaumont2009adaptive,
+sisson2007sequential, del2012adaptive}.
+
+All the algorithms discussed in this section can be straightforwardly extended
+to sample from the joint distribution
+\[
+  \pi_\varepsilon(\theta, z_1, \ldots, z_M \mid y),
+\]
+which is equivalent to associating $M$ simulated datasets to each parameter
+particle instead of just one. The simulated dataset $z$ is replaced by
+$\vec{z} = z_1, \ldots, z_M$, and the indicator function for the
+$\varepsilon$-ball around $y$ is replaced by
+\[
+  \sum_{k=1}^M \I_{A_{\varepsilon, y}} (z_i).
+\]
+For \gls{ABC}-\gls{MCMC} and \gls{ABC}-\gls{SMC}, the proposal distribution
+$q(\theta, \theta') \Pr(z \mid \theta')$ is replaced by
+\[
+  q_i(\theta, \theta') \prod_{k=1}^M \Pr(z_i \mid \theta').
+\]

### 2016-04-18 doc/thesis/methods.tex
More writing

@@
 the sequence of tolerances $\varepsilon_i$ is automatically determined rather
-than specified by the user. The tolerances are chosen such that the \gls{ESS}
-of the particle population, which indicates the quality of the Monte Carlo
+than specified in advance. The tolerances are chosen such that the \gls{ESS} of
+the particle population, which indicates the quality of the Monte Carlo
 approximation (see \cref{subsec:sis}), decays at a controlled rate. This
@@
 \textcite{del2012adaptive}, the equation has no analytic solution, but can be
-solved by bisection. As in the reference version, the particles are perturbed
-by applying an \gls{MCMC} kernel, and resampling occurs when the \gls{ESS}
-drops below a prescribed threshold. In my implementation, following all
-proposals of the \gls{MCMC} kernel are Gaussian with variance equal to twice
-the empirical variance of the
-particles~\autocite{beaumont2009adaptive,del2012adaptive}.
-
+solved numerically by bisection. The forward kernels $K_i$ are taken to be
+\gls{MCMC} kernels with stationary distributions $\pi_{\varepsilon_i}$ and
+proposal distributions
+\[
+  q_i(\theta, \theta') \prod_{k=1}^M \Pr(X_i^{(k)}' \mid \theta'),
+\]
+where $\theta$ is the vector of model parameters and $X_k$ are $M$ datasets
+simulated according to $\theta$. In our implementation, $q$ is a Gaussian
+proposal with variance equal to twice the empirical variance of the particles,
+following~\autocite{beaumont2009adaptive, del2012adaptive}. The backwards
+kernels are
+\[
+  L_{i-1}(z', z) = \frac{\pi_n(z)K(z, z')}{\pi_n(z')}.
+\]
+When substituted into \cref{eq:smcwt}, the forward kernels $K(z, z')$ and
+densities $\pi_n(z')$ cancel out, and we are left with the weight update
+\begin{align*}
+  w(z') 
+    &\propto w(z) \frac{\pi_n(z \mid y)}{\pi_{n-1}(z \mid y)} \\
+    &= w(z) \frac{\pi(z) \pi_n(y \mid z)}{\pi(z) \pi_{n-1}(y \mid z)} \\
+    &= w(z) \frac{\sum_{k=i}^M \I_{A_{\varepsilon_n, y}}(x_k)}
+            {\sum_{k=i}^M \I_{A_{\varepsilon_{n-1}, y}}(x_k)}.
+\end{align*}
 The algorithm may be stopped when one of two termination conditions is reached.

### 2016-04-18 doc/thesis/intro.tex
Panicked 1AM writing

@@
 functions, originally developed for \glspl{SVM}~\autocite{burges1998tutorial}, 
-compare objects by mapping them into a feature space of very high, possibly
-even infinite, dimension.
-
-Another tree distance measure is the \defn{phylogenetic kernel}, or ``tree
-kernel'' developed by \textcite{poon2013mapping}. As opposed to the \gls{nltt},
-the tree kernel is maximized when the two trees being compared are the same.
-The basis of the tree kernel is the kernel trick originally developed for
-\glspl{SVM}~\autocite{burges1998tutorial}. The idea of the kernel trick is to
-compare objects by mapping them into a feature space of very high, possibly
-even infinite, dimension. The similarity between objects is taken to be their
-dot product in the feature space. It is called a ``trick'' because this dot
-product is computed using a \defn{kernel function} without explicitly mapping
-the objects to the feature space, which would be computationally prohibitive.
-In the case of the tree kernel, the feature space is the space of all possible
-\defn{subset trees}, which are subtrees that do not necessarily extend all the
-way to the tips. The subset-tree kernel was originally developed for comparing
-parse trees in natural language processing~\autocite{collins2002new} and did
-not incorporate branch length information. The version developed by
+compare objects in a space $\X$ by mapping them into a feature space $\F$ of
+high or infinite dimension via a function $\varphi$. The similarity between the
+objects is defined as
+\[  
+  K(x, x') = <\varphi(x), \varphi(x')>,
+\]
+that is, the inner product of the objects' representations in the feature
+space. Computing $\varphi(x)$ may be computationally prohibitive due to the
+dimension of $\F$. The utility of a kernel function $K$ is that it is
+constructed in such a way that it can compute the inner product without
+explicitly computing $\varphi(x)$. The kernel function developed in
+\autocite{poon2013mapping} will henceforth be referred to as the \defn{tree
+kernel}. The feature space into which trees are mapped is the space of all
+possible \defn{subset trees}, which are subtrees that do not necessarily extend
+all the way to the tips. The subset-tree kernel was originally developed for
+comparing parse trees in natural language processing~\autocite{collins2002new}
+and did not incorporate branch length information. The version developed by
 \textcite{poon2013mapping} includes a radial basis function to compare the
 differences in branch lengths, thus incorporating both the trees' topologies
-and their branch lengths in a single similarity score. The tree kernel was
-later shown to be highly effective in differentiating trees simulated under a
-compartmental model with two risk groups of varying contact
-rates~\autocite{poon2015phylodynamic}. In that paper,
-\citeauthor{poon2015phylodynamic} used the tree kernel as the distance function
-in \gls{ABC} (see \cref{sec:abc}), to fit epidemiological models to observed
-trees. This is an example of kernel-ABC~\autocite{nakagome2013kernel}, which
-will be discussed further in \cref{sec:abc}.
+and their branch lengths in a single similarity score.
+
+%The tree kernel was later shown to be highly effective in differentiating trees
+%simulated under a compartmental model with two risk groups of varying contact
+%rates~\autocite{poon2015phylodynamic}. In that paper,
+%\citeauthor{poon2015phylodynamic} used the tree kernel as the distance function
+%in \gls{ABC} (see \cref{sec:abc}), to fit epidemiological models to observed
+%trees. This is an example of kernel-ABC~\autocite{nakagome2013kernel}, which
+%will be discussed further in \cref{sec:abc}.
 
@@
 tracing~\autocite{morris1993epidemiology, welch2011statistical,
-keeling2005networks}. Therefore, it has been necessary to explore less
-resource-intensive alternatives which still contain information about
+keeling2005networks, eames2015six}. Therefore, it has been necessary to explore
+less resource-intensive alternatives which still contain information about
 population structure. For instance, it is possible to obtain limited
@@
 rather than a power law, was the best fit to degree distributions of six social
-or sexual networks.
+and sexual networks.
 
@@
 of the donor lineage. \Cref{fig:contactnet} illustrates this correspondence.
-It's important to note that, although the order and timing of transmissions
+It must be emphasized that, although the order and timing of transmissions
 determines the tree topology uniquely, the converse does not hold. That is, for
@@
 incidence and prevalence trajectories, using the method developed by
-\textcite{volz2009phylodynamics}. They found that the concordance between
-the two skyrides, as well as the relationship between the skyride and
-prevalence curve, was qualitatively different for each degree distribution.
+\textcite{volz2009phylodynamics}. The concordance between the two skyrides, as
+well as the relationship between the skyride and prevalence curve, was
+qualitatively different for each degree distribution.
 \textcite{leventhal2012inferring} investigated the relationship between
@@
 
-In summary, studies in this group have demonstrated that network structure
-profoundly influences tree shape, but have not attempted to quantitatively
-infer network parameters from observed trees.
-
-%\section{Approximate Bayesian computation}
-
-%Consider a model $M$, with parameters $\theta$, which we wish to fit to some
-%observed data $D$. By ``fit'', we often mean that we want to find particular
-%values $\hat{\theta}$ for the parameters which optimize the likelihood of our
-%data given those parameters and the model,
-%\[
-%  \hat{\theta} = \argmax_\theta \Pr(D | M, \theta).
-%\]
-%This $\hat{\theta}$ is the \defn{maximum likelihood} parameter estimate. Note
-%that we are employing a common abuse of notation here, where $\Pr(\cdots)$ is
-%being unsed to refer to a probabilty \emph{density} rather than a true
-%probability. Alternative to maximum likelihood, we may be interested less in a
-%point estimate and more in the posterior distribution of possible values of
-%$\theta$ given our data, $\Pr(\theta \mid M, D)$. This will be expanded upon
-%below, but for the moment, for illustrative purposes, we restrict our attention
-%to the maximum likelihood problem.
-%
-%If the model we are fitting is sufficiently simple, it may be possible to
-%calculate $\hat{\theta}$ directly, using calculus. Most models do not admit
-%analytic maximum likeilhood solutions, but if the likelihood any set of
-%parameters can be calculated up to a normalizing constant, then
-%${\Pr(D \mid M,\theta)}$ can be optimized numerically. A wide range of
-%optimization strategies exist, the choice of which to use depending on the
-%complexity of the model and whether or not we have access to the gradients of
-%the likelihood function with respect to each of the parameters. The majority of
-%modelling problems fit into this category, and numerical optimization is
-%well-developed and extremely widely used.
-%
-%However, there are some cases, often when the observed data is of a complex
-%type, that explicitly calculating the likelihood of some observed data is
-%impossible, even up to a normalizing constant. For example, suppose that we
-%want to model a chess player's behaviour. We will set up a simple one-parameter
-%model which describes the chess playing process.  The parameter, $a \in [0, 1]$
-%indicates the player's eagerness to remove his opponent's pieces from the
-%board. We can write down an algorithm for the player's behaviour under such a
-%model.
-%
-%% TODO: don't break this over the page
-%
-%\begin{algorithmic}
-%  \While{the game is not over}
-%    \If{I can capture an opponent's piece and $\Uniform(0, 1) < a$}
-%      \State{capture the piece}
-%    \Else
-%      \State{make any other move at random}
-%    \EndIf
-%  \EndWhile
-%\end{algorithmic}
-%
-%Suppose the observed data are the ending configurations of the board, after the
-%player has concluded a game against an opponent with a known value of $a$. The
-%model we have designed is very simple, but it is not obvious how to calculate
-%the likelihood of a particular ending configuration. Indeed, it seems that the
-%only way is to enumerate every possible path the game could have taken, and
-%tabulate the ending configurations of each. Clearly, this is infeasible.
-%Approximate Bayesian computation was designed for situations like these, where
-%exact likelihoods are not available, perhaps due to the model involving an
-%algorithm or generative process.
-
 \section{Sequential Monte Carlo}
@@
 To fully describe \gls{SMC}, we will introduce some notation and terminology.
-For a sequence $x_1, \ldots, x_d$, we will write $\vec{x_i}$ to mean the
-partial sequence $x_1, \ldots, x_i$.
-
-A \defn{Markov kernel} is the continuous analogue of the transition matrix in a
-finite-state Markov model. For some spaces $X$ and $Y$, $K \maps X \times Y \to
-[0, 1]$ such that
+The definitions of these terms will become clearer as they are used. For a
+sequence $x_1, \ldots, x_d$, we will write $\vec{x_i}$ to mean the partial
+sequence $x_1, \ldots, x_i$. The subscript $^{(k)}$ will be used to indicate
+the $k$th particle in a population. To ease the notational burden we will omit
+the superscripts and subscripts on the weight functions $w$.
+
+We define a \defn{Markov kernel} as the continuous analogue of the transition
+matrix in a finite-state Markov model.  For some spaces $X$ and $Y$, $K \maps X
+\times Y \to [0, 1]$ such that
 \begin{align}
@@
 \textbf{TODO}. Note that a Markov kernels have nothing to do with the kernel
-functions defined in \cref{subsec:treeshape}, other than sharing a name.
-Unfortunately, the word ``kernel'' is ubiquitous in mathematics and refers to
-many unrelated concepts.
+functions defined in \cref{subsec:treeshape}, other than sharing a name (the
+word ``kernel'' is ubiquitous in mathematics).
 
@@
 objective is to build up an importance distribution $\eta$ for $\pi$
-sequentially. Let $\vec{x_i} = (x_1, \ldots, x_i)$, for $1 \leq i \leq d$. By
-the general product rule,
+sequentially. By the general product rule, $\pi(\vec{x})$ can be decomposed as
 \[
@@
 \]
-The importance weights for $\eta$ with respect to $\pi$ are then defined
-recursively as
+The importance weights for $\eta$ can be written recursively as
 \begin{align}
@@
 the importance weights is as small as possible at every step, eventually
-arriving at a full importance distribution. The process for choosing $\eta$
-will differ according to the problem at hand. One potential choice for
-$\eta(x_i \mid \vec{x_{i-1}})$ is simply $\pi(x_i \mid \vec{x_{i-1}})$, if it
-is possible to compute. For example, in the case of an \gls{HMM}, $\pi(x_i \mid
-\vec{x_{i-1}})$ is simply the transition probability from $x_{i-1}$ to $x_i$.
-In a Bayesian setting, $\eta$ may be taken to be the prior distribution. In
-some applications, it is possible to compute or approximate the form of $\eta$
-which minimizes the variance of the weights. The resulting importance
-distribution is known as the \defn{optimal
-kernel}~\autocite{cappe2007overview}. Choosing each element of the
-decomposition of $\eta$ is a problem-specific task (see \textit{e.g.}
-\autocite{smith2013sequential,liu2008monte} for many examples).
+arriving at a full importance distribution. This choice is made on a
+problem-specific basis, taking any available information about $\pi(x_i \mid
+\vec{x_{i-1}})$ into account (see \textit{e.g.}
+\autocite{smith2013sequential,liu2008monte} for many examples).  One potential
+choice for $\eta(x_i \mid \vec{x_{i-1}})$ is simply $\pi(x_i \mid
+\vec{x_{i-1}})$, if it is possible to compute. In a Bayesian setting, the
+prior distribution may be used. The exact form of $\eta(x_i, \vec{x_{i-1}})$
+which minimizes the variance of the weights is called the \defn{optimal
+kernel}~\autocite{cappe2007overview}, the name deriving from the fact that
+$k(x_i, \vec{x_{i-1}}) = \eta(x_i, \vec{x_{i-1}})$ is a Markov kernel. In some
+applications, it is possible to approximate the optimal kernel or even compute
+it explicitly.
 
 The recursive definition \cref{eq:sisw} suggests an algorithm for obtaining a
-sample from $\pi$. We begin with a sample of $n$ ``particles'' which have been
-sampled from the importance distribution $\eta(x_0)$ for $\pi(x_0)$. The
+sample from $\pi$ (\cref{alg:sis}).  We begin with $n$ ``particles'' which have
+been sampled from the importance distribution $\eta(x_0)$ for $\pi(x_0)$. The
 particles are updated and reweighted $d$ times, corresponding to the $d$
 elements of the decomposition of $\pi$. At the $i$th step, each particle is
-extended to include $x_i$ drawn from the chosen $\eta(x_i \mid \vec{x_{i-1}})$,
-and the importance weights are recalculated and normalized.
+extended to include $x_i$ drawn according to the chosen $\eta(x_i \mid
+\vec{x_{i-1}})$, and the importance weights are recalculated and normalized.
+
+\begin{algorithm}
+  \caption{Sequential importance sampling.}
+  \begin{algorithmic}
+    \For {$k = 1$ to $n$}
+      \State Sample $x_1^{(k)}$ from $\eta(x_1)$
+      \Comment{Initialize the $k$th particle}
+      \State $w^{(k)} \gets \dfrac{\pi\left(x_1^{(k)}\right)}{\eta\left(x_1^{(k)}\right)}$
+    \EndFor
+    \For {$i = 2$ to $d$}
+      \For {$k = 1$ to $n$}
+        \State Sample $x_i^{(k)}$ from $\eta\left(x_i \mid \vec{x_{i-1}^{(k)}}\right)$
+        \Comment Extend the $k$th particle
+        \State $w(\vec{x_i}^{(k)}) \gets \dfrac{\pi\left(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}\right)}{\eta\left(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}\right)} \cdot w(\vec{x_{i-1}}^{(k)})$
+      \EndFor
+      \State Normalize the weights so that $\sum w = 1$
+    \EndFor
+    \State Sample $n$ particles with probabilities $w$
+  \end{algorithmic}
+  \label{alg:sis}
+\end{algorithm}
 
@@
 on the one described by \textcite{liu2008monte}, namely the decay of the
-\gls{ESS} below a prescribed threshold. The \gls{ESS} of the population of
-particles is defined as
+\gls{ESS} below a prescribed threshold, conventionally $n/2$. The \gls{ESS} of
+the population of particles is defined as
 \[
@@
 beyond the basic sampling with replacement have been proposed, but we will not
-discuss those here. The full \gls{SIS} algorithm with resampling is given as
-\cref{alg:sis}. We use a superscript $^{(k)}$ to refer to the $k$th particle,
-and for brevity write $w$ for $w(\vec{x_i})$. The variable $T$ is the threshold
-for the \gls{ESS}, below which the population is resampled. It is conventional
-to take $T = n/2$.
-
-\begin{algorithm}
-  \caption{Sequential importance sampling.}
-  \begin{algorithmic}
-    \For {$k = 1$ to $n$}
-      \State Sample $x_0^{(k)}$ from $\eta(x_0)$
-      \Comment{Initialize the $k$th particle}
-      \State $w^{(k)} \gets \dfrac{\pi\left(x_0^{(k)}\right)}{\eta\left(x_0^{(k)}\right)}$
-    \EndFor
-    \For {$i = 1$ to $d$}
-      \For {$k = 1$ to $n$}
-        \State Sample $x_i^{(k)}$ from $\eta\left(x_i \mid \vec{x_{i-1}^{(k)}}\right)$
-        \Comment Extend the $k$th particle
-        \State $w^{(k)} \gets \dfrac{\pi\left(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}\right)}{\eta\left(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}\right)} \cdot w^{(k)}$
-      \EndFor
-      \State Normalize the weights so that $\sum w = 1$
-      \If {$\ESS(w) < T$}
-        \State Resample the particles according to $w$
-        \For {$k = 1$ to $n$}
-          \State $w^{(k)} \gets 1/n$
-        \EndFor
-      \EndIf
-    \EndFor
-    \State Sample $n$ particles with probabilities $w$
-  \end{algorithmic}
-  \label{alg:sis}
-\end{algorithm}
+discuss those here.
 

### 2016-04-19 doc/thesis/appendix.tex
Finished draft intro!

@@
   \includegraphics{kernel-alpha-crossv.pdf}
+  \caption{
+    Cross validation accuracy of classifiers for \gls{BA} model parameter
+    \gls{alpha} for eight epidemic scenarios. Solid lines and points are
+    $R^2$ of tree kernel \gls{kSVR} under various kernel meta-parameters.
+    Dashed and dotted lines are $R^2$ of linear regression against Sackin's
+    index, and \gls{SVR} using \gls{nltt}.
+  }
   \label{fig:alphacrossv}
@@
   \includegraphics{kernel-I-crossv.pdf}
+  \caption{
+    Cross validation accuracy of classifiers for \gls{BA} model parameter
+    \gls{I} for eight epidemic scenarios. Solid lines and points are $R^2$ of
+    tree kernel \gls{kSVR} under various kernel meta-parameters. Dashed and
+    dotted lines are $R^2$ of linear regression against Sackin's index, and
+    \gls{SVR} using \gls{nltt}.
+  }
   \label{fig:Icrossv}
@@
   \includegraphics{kernel-m-crossv.pdf}
+  \caption{
+    Cross validation accuracy of classifiers for \gls{BA} model parameter
+    \gls{m} for eight epidemic scenarios. Solid lines and points are $R^2$ of
+    tree kernel \gls{kSVR} under various kernel meta-parameters. Dashed and
+    dotted lines are $R^2$ of linear regression against Sackin's index, and
+    \gls{SVR} using \gls{nltt}.
+  }
   \label{fig:mcrossv}
@@
   \includegraphics{kernel-N-crossv.pdf}
+  \caption{
+    Cross validation accuracy of classifiers for \gls{BA} model parameter
+    \gls{N} for eight epidemic scenarios. Solid lines and points are $R^2$ of
+    tree kernel \gls{kSVR} under various kernel meta-parameters. Dashed and
+    dotted lines are $R^2$ of linear regression against Sackin's index, and
+    \gls{SVR} using \gls{nltt}.
+  }
   \label{fig:Ncrossv}
@@
   \includegraphics{kernel-alpha-kpca.pdf}
+  \caption{
+    Kernel principal components projection of trees simulated under three
+    different values of \gls{BA} parameter \gls{alpha}, for eight epidemic
+    scenarios.
+  }
   \label{fig:alphakpca}
@@
   \includegraphics{kernel-I-kpca.pdf}
+  \caption{
+    Kernel principal components projection of trees simulated under three
+    different values of \gls{BA} parameter \gls{I}, for eight epidemic
+    scenarios.
+  }
   \label{fig:Ikpca}
@@
   \includegraphics{kernel-m-kpca.pdf}
+  \caption{
+    Kernel principal components projection of trees simulated under three
+    different values of \gls{BA} parameter \gls{m}, for eight epidemic
+    scenarios.
+  }
   \label{fig:mkpca}
@@
   \includegraphics{kernel-N-kpca.pdf}
+  \caption{
+    Kernel principal components projection of trees simulated under three
+    different values of \gls{BA} parameter \gls{N}, for eight epidemic
+    scenarios.
+  }
   \label{fig:Nkpca}
@@
 \end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics{abc-point-estimate-m3.pdf}
+  \caption{                                              
+    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
+    running \software{netabc} on simulated data. Values shown are for
+    simulations with \gls{m} = 3. Dashed lines indicate true values. (A)
+    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations
+    against known values. (B) Estimates of \gls{m} and \gls{N} which were held
+    fixed in these simulations at the values \gls{m} = 3 and \gls{N} = 5000.
+  }        
+  \label{fig:abcm3}
+\end{figure}
+
+\begin{figure}
+  \centering
+  \includegraphics{abc-point-estimate-m4.pdf}
+  \caption{                                              
+    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
+    running \software{netabc} on simulated data. Values shown are for
+    simulations with \gls{m} = 4. Dashed lines indicate true values. (A)
+    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations
+    against known values. (B) Estimates of \gls{m} and \gls{N} which were held
+    fixed in these simulations at the values \gls{m} = 4 and \gls{N} = 5000.
+  }        
+  \label{fig:abcm4}
+\end{figure}

### 2016-04-19 doc/thesis/conclusion.tex
Finished draft intro!

@@
-Due to the rapid advancement of nucleotide sequencing technology, viral
-sequence data data have become increasingly feasible to collect on a population
-level. Through phylodynamic methods, these data offer a window into
-epidemiological processes which would otherwise be virtually impossible to
-study on a realistic scale. Although phylodynamic methods have been developed
-to fit a wide variety of epidemiological models to phylogenetic
+This thesis developed \software{netabc}, a computer program implementing a
+statistical inference method for contact network parameters from viral
+phylogenetic data. Due to the rapid advancement of nucleotide sequencing
+technology, viral sequence data data have become increasingly feasible to
+collect on a population level. Through phylodynamic methods, these data offer a
+window into epidemiological processes which would otherwise be virtually
+impossible to study on a realistic scale. Although phylodynamic methods have
+been developed to fit a wide variety of epidemiological models to phylogenetic
 data~\autocite{volz2012complex, rasmussen2014phylodynamic}, our method is the
@@
 villandre2016assessment}.
-
-Our work had three main aims. First, we developed phylodynamic a method to
-estimate contact network model parameters.
-
-Our second aim was to investigate the parameters of the \gls{BA} model, in
-terms of their impact on tree shape and our ability to reconstruct them with
-kernel-\gls{ABC}. We determined through simulation studies that the
-preferential attachment power $\alpha$ and number of infected nodes $I$ had a
-substantial impact on transmission tree shape, and could be estimated using our
-method.
-
-The final aim of our work was to apply our method to real world
-\gls{HIV} datasets.
-
-% ergm

### 2016-04-19 doc/thesis/intro.tex
Finished draft intro!

@@
 
-In general, \defn{viral phylogenies} are simply phylogenetic trees relating
-virus strains. \defn{Inter-host} phylogenies relate one representative viral
-genotype from each host in a population. In this work, when we use the phrase
-``viral phylogeny'', we are referring to an inter-host phylogeny unless
-otherwise stated. Phylogenetic inference is a complex topic which we will not
-discuss here (see \eg \autocite{nei2000molecular} for a full review). We assume
-here that the viral phylogeny is known, a common assumption in the development
-and application of phylodynamic methods.
-
-The crux of phylodynamics is the fact that the epidemiological processes
-recorded in transmission trees, and the evolutionary processes recorded in
-viral phylogenies, occur on similar time scales for RNA
-viruses~\autocite{drummond2003measurably}. Phylodynamic methods have been used
-to investigate epidemiological parameters such as transmission rate, recovery
-rate, and basic reproductive number~\autocite{pybus2009evolutionary,
+The basis of phylodynamics is the fact that, for RNA viruses, epidemiological
+and evolutionary processes occur on similar time
+scales~\autocite{drummond2003measurably}. In fact, these two processes
+interact, such that it is possible to detect the influence of host epidemiology
+on the evolutionary history of the virus as recorded in an \defn{inter-host
+viral phylogeny}. Phylodynamic methods aim to detect and quantify the
+signatures of epidemiological processes in these
+phylogenies~\autocite{pybus2009evolutionary, volz2013viral}, which relate one
+representative viral genotype from each host in an infected population. These
+methods have been used to investigate parameters such as transmission rate,
+recovery rate, and basic reproductive number~\autocite{pybus2009evolutionary,
 volz2013viral}. The majority of phylodynamic studies attempt to infer the
@@
 birth-death~\autocite{kendall1948generalized, stadler2012estimating} or
-coalescent~\autocite{kingman1982coalescent, volz2012complex} models.
-
-Due to the relationship between the aforementioned processes, there is a close
-relationship between viral phylogenies and transmission trees. In particular,
-the transmission process is quite similar to \defn{allopatric
+coalescent~\autocite{kingman1982coalescent, volz2012complex} models. These
+methods generally assume the viral phylogeny is known, and we assume the same
+in this work (see \eg \autocite{nei2000molecular} for a review of phylogenetic
+inference).
+
+Due to the relationship between the aforementioned processes, there is a degree
+of correspondence between viral phylogenies and transmission
+trees~\autocite{ypma2013relating, kenah2015algorithms, kenah2016molecular}. In
+particular, the transmission process is quite similar to \defn{allopatric
 speciation}~\autocite{coyne2004speciation}, where genetic divergence follows
@@
 transmission tree causes the extinction of their viral lineage in the
-phylogeny. Due to these relationships, the topology of the viral phylogeny is
-often used as a proxy for the topology of the transmission tree. However, there
-are several complications and caveats which must be kept in mind when
-estimating the transmission tree in this manner.
-
-First are the issues of rooting and time-scaling. Modern likelihood-based
+phylogeny. Consequently, the topology of the viral phylogeny is sometimes used
+as a proxy for the topology of the transmission tree. Modern likelihood-based
 methods of phylogenetic reconstruction~\autocite[\eg][]{price2010fasttree,
@@
 hand, transmission trees are rooted, and have branches measuring calendar
-time~\autocite{pybus2009evolutionary}. It is generally assumed that sampling a
-virus from the individual also corresponds to their removal from the
-transmission tree, so the positions of the tips in time are fixed. Therefore,
-to transform a viral phylogeny into an estimated transmission tree, we must
-find a root and an assignment of branch lengths such that the tips are placed
-at their proscribed times. In addition, the branch lengths should be chosen
-such that the variation in \defn{evolutionary rate}, which is the ratio of a
-branch's length in genetic distance to its length in calendar time, is low
-across the tree. While there is some variation among hosts, due to
-immunological and other factors, we generally expect to observe globally
-similar evolutionary rates. Methods for time-scaling a phylogeny include
-root-to-tip regression~\autocite{shankarappa1999consistent, korber2000timing,
+time~\autocite{pybus2009evolutionary}. Therefore, estimating a transmission
+tree from a viral phylogeny requires the phylogeny to be rooted and
+time-scaled. Methods for performing this process include root-to-tip
+regression~\autocite{shankarappa1999consistent, korber2000timing,
 drummond2003inference}, which we apply in this work, and least-square
-dating~\autocite{to2015fast}. Both of these methods can be used to root the
-tree, by simply trying all possible root positions and choosing the one which
-minimizes a loss function ($1-R^2$, or root-mean-square error, for root-to-tip
-regression; sum of squared errors for least-square dating). Alternatively, the
-tree may be rooted separately with an outgroup~\autocite{li1988rates} before
-time-scaling.
-
-A second, perhaps more insidious problem is the fact that the correspondence
-between the topologies of the viral phylogeny and transmission tree is not
-necessarily exact. Due to intra-host diversity, the viral strain which is
-transmitted may have split from another lineage within the donor long before
-the transmission event occurred. Hence, the branching point in the viral
-phylogeny may be much earlier than that in the transmission tree. Another
-possibility is that one host transmitted to two or more recipients, but the
-lineages they each received originated within the donor host in a different
-order than that in which the transmissions occurred. In this case, the topology
-of the transmission tree and the viral phylogeny will be mismatched. Although
-phylodynamics is quite new, these phenomena have been studied in evolutionary
-biology for some time. Viral phylogenies are a specific version of a more
-general class of trees called \defn{gene trees}, which represent the
-evolutionary history of a section of genetic material. Transmission trees, on
-the other hand, are highly analogous to \defn{species trees}, whose tips are
-species and internal nodes are common ancestors. This analogy derives from the
-functional similarity between transmission and allopatric speciation. Hence,
-the potential discordance between transmission trees and viral phylogenies is
-the similar to that between gene and species trees, which is called
-\defn{incomplete lineage sorting}. In practice, this discordance has not proven
-an insurmountable problem: for example, \textcite{leitner1996accurate} were
-able to accurately recover a known transmission tree using a viral phylogeny.
-
-What we have just discussed is a two-step procedure for estimating the
-transmission tree. First, a viral phylogeny is constructed from genetic
-sequence data, and then it is rooted and time-scaled into a transmission tree.
-This approach is straightforward, frequently used, and has the advantage of
-leveraging tried-and-true tools for phylogenetic inference. However, it also
-has drawbacks, perhaps the most obvious being the multiplication of errors
-produced by the separate steps. One commonly used alternative method is to
-directly estimate a time-scaled phylogeny by simultaneously inferring the tree
-topology, its root and branch lengths, and the parameters of a \defn{molecular
-clock} model. A molecular clock is a hypothesis about the evolutionary rates
-along the branches of the tree, such as that they are all equal (a \defn{strict
-clock}) or that they are \gls{iid} from a common distribution (a \defn{relaxed
-clock}). This inference is usually done in a Bayesian framework using
-\gls{MCMC}, so that prior information (including the tip dates) can be included
-in the analysis, and the so-called nuisance parameters of the molecular clock
-model can be marginalized out. Software packages for performing these analyses
-include BEAST~\autocite{bouckaert2014beast} and
-MrBayes~\autocite{ronquist2012mrbayes}.
-
-Several other authors have developed methods tailor-made for inferring
-transmission trees. \textcite{didelot2014bayesian} develop a Bayesian version
-of the two-step approach which allows transmissions to occur anywhere along the
-branches of a transmission tree, rather than being constrained to the branching
-points in the viral phylogeny. The method requires sampling of every infected
-individual, although the authors indicate that it could be extended to relax
-this assumption. \textcite{cottam2008integrating} describe a likelihood-based
-method which enumerates all transmission trees consistent with an established
-phylogeny, assigning each a likelihood based on other epidemiological data. 
-This approach is novel in its integration of data from multiple sources,
-however because it enumerates a large portion of the tree space, it is unlikely
-to scale to larger epidemics. \textcite{ypma2012unravelling} develop a joint 
-likelihood function integrating temporal, geographic, and genetic observations, 
-and use Bayesian \gls{MCMC} to estimate both the tree and the parameters of the
-likelihood function. Their approach can handle missing data and produces high
-resolution transmission trees when multiple types of data are available. A
-different approach is undertaken by \textcite{jombart2011reconstructing}, who
-describe a method to build transmission trees directly from sequence data,
-contingent on the common ancestors also being sampled. This makes the method
-attractive for slow-evolving pathogens, but less practical for viral outbreaks
-where samples from common ancestors are unlikely to be available.
+dating~\autocite{to2015fast}. Alternatively, the tree may be rooted separately
+with an outgroup~\autocite{li1988rates} before time-scaling.
+
+A caveat of estimating transmission trees in this manner is that the
+correspondence between the topologies of the viral phylogeny and transmission
+tree is not necessarily exact~\autocite{ypma2013relating}. Due to intra-host
+diversity, the viral strain which is transmitted may have split from another
+lineage within the donor long before the transmission event occurred. Hence,
+the branching point in the viral phylogeny may be much earlier than that in the
+transmission tree. Another possibility is that one host transmitted to two or
+more recipients, but the lineages they each received originated within the
+donor host in a different order than that in which the transmissions occurred.
+In this case, the topology of the transmission tree and the viral phylogeny
+will be mismatched. In practice, this discordance has not proven an
+insurmountable problem: for example, \textcite{leitner1996accurate} were able
+to accurately recover a known transmission tree using a viral phylogeny. The
+problem of accurately estimating transmission trees is an ongoing area of
+research~\autocite{cottam2008integrating, jombart2011reconstructing,
+ypma2012unravelling, didelot2014bayesian}.
+
+%Although phylodynamics is quite new, these phenomena have been studied in
+%evolutionary biology for some time. Viral phylogenies are a specific version of
+%a more general class of trees called \defn{gene trees}, which represent the
+%evolutionary history of a section of genetic material. Transmission trees, on
+%the other hand, are highly analogous to \defn{species trees}, whose tips are
+%species and internal nodes are common ancestors. This analogy derives from the
+%functional similarity between transmission and allopatric speciation. Hence,
+%the potential discordance between transmission trees and viral phylogenies is
+%the similar to that between gene and species trees, which is called
+%\defn{incomplete lineage sorting}.
+
+%What we have just discussed is a two-step procedure for estimating the
+%transmission tree. First, a viral phylogeny is constructed from genetic
+%sequence data, and then it is rooted and time-scaled into a transmission tree.
+%This approach is straightforward, frequently used, and has the advantage of
+%leveraging tried-and-true tools for phylogenetic inference. However, it also
+%has drawbacks, perhaps the most obvious being the multiplication of errors
+%produced by the separate steps. One commonly used alternative method is to
+%directly estimate a time-scaled phylogeny by simultaneously inferring the tree
+%topology, its root and branch lengths, and the parameters of a \defn{molecular
+%clock} model. A molecular clock is a hypothesis about the evolutionary rates
+%along the branches of the tree, such as that they are all equal (a \defn{strict
+%clock}) or that they are \gls{iid} from a common distribution (a \defn{relaxed
+%clock}). This inference is usually done in a Bayesian framework using
+%\gls{MCMC}, so that prior information (including the tip dates) can be included
+%in the analysis, and the so-called nuisance parameters of the molecular clock
+%model can be marginalized out. Software packages for performing these analyses
+%include BEAST~\autocite{bouckaert2014beast} and
+%MrBayes~\autocite{ronquist2012mrbayes}.
+
+%Other methods are tailor-made for inferring transmission trees.
+%\textcite{didelot2014bayesian} develop a Bayesian approach which allows
+%transmissions to occur anywhere along the
+%branches of a transmission tree, rather than being constrained to the branching
+%points in the viral phylogeny. The method requires sampling of every infected
+%individual, although the authors indicate that it could be extended to relax
+%this assumption. \textcite{cottam2008integrating} describe a likelihood-based
+%method which enumerates all transmission trees consistent with an established
+%phylogeny, assigning each a likelihood based on other epidemiological data.
+%This approach is novel in its integration of data from multiple sources,
+%however because it enumerates a large portion of the tree space, it is unlikely
+%to scale to larger epidemics. \textcite{ypma2012unravelling} develop a joint
+%likelihood function integrating temporal, geographic, and genetic observations,
+%and use Bayesian \gls{MCMC} to estimate both the tree and the parameters of the
+%likelihood function. Their approach can handle missing data and produces high
+%resolution transmission trees when multiple types of data are available. A
+%different approach is undertaken by \textcite{jombart2011reconstructing}, who
+%describe a method to build transmission trees directly from sequence data,
+%contingent on the common ancestors also being sampled. This makes the method
+%attractive for slow-evolving pathogens, but less practical for viral outbreaks
+%where samples from common ancestors are unlikely to be available.
 
@@
 
-The aim of viral phylodynamics is to glean some kind of knowledge, about the
-epidemic, the virus, or its hosts and their behaviour, by studying a phylogeny,
-most often an estimated transmission tree~\autocite{pybus2009evolutionary,
-volz2013viral}. What is informative about a phylogeny, beyond the demographic
-characteristics of the individuals it relates, is its \defn{shape}. The shape
-of a phylogeny has two components: the topology, and the distribution of branch
-lengths~\autocite{mooers1997inferring}. Methods of quantifying tree shape fall
-into two categories: summary statistics, and pairwise measures.
-
-Summary statistics assign a numeric value to each individual tree. One of the
-most widely used is Sackin's index~\autocite{shao1990tree}, which measures the
-imbalance or asymmetry in a rooted tree. For the $i$th tip of the tree, we
-define $N_i$ to be the number of branches between that tip and the root. The
-unnormalized Sackin's index is defined as the sum of all $N_i$. It is called
-unnormalized because it does not account for the number of tips in the tree.
-Among two trees having the same number of tips, the least-balanced tree will
-have the highest Sackin's index. However, among two equally balanced trees, the
-larger tree will have a higher Sackin's index. This makes it challenging to
-compare balances among trees of different sizes. To correct this,
-\textcite{kirkpatrick1993searching} derive the expected value of Sackin' index
-under the Yule model~\autocite{yule1925mathematical}. Dividing by this expected
-value normalizes Sackin's index, so that it can be used to compare trees of
-different sizes.
-
-Rather than assigning numbers to individual trees, pairwise measures associate
-a numeric value to each \textit{pair} of trees, indicating how different the
-trees are from each other. Distance measures allow us to identify groups of
-related phylogenies, for example, local epidemics which are undergoing a
-similar pattern of expansion. One such distance measure is the
+To perform phylodynamic inference, we must be able to extract quantitative
+information from viral phylogenies. What is informative about a phylogeny,
+beyond the demographic characteristics of the individuals it relates, is its
+\defn{shape}. The shape of a phylogeny has two components: the topology, and
+the distribution of branch lengths~\autocite{mooers1997inferring}. Methods of
+quantifying tree shape fall into two categories: summary statistics, and
+pairwise measures. Summary statistics assign a numeric value to each individual
+tree, while pairwise measures quantify the similarity between pairs of trees.
+
+One of the most widely used tree summary statistics is Sackin's
+index~\autocite{shao1990tree}, which measures the imbalance or asymmetry in a
+rooted tree. For the $i$th tip of the tree, we define $N_i$ to be the number of
+branches between that tip and the root. The unnormalized Sackin's index is
+defined as the sum of all $N_i$. It is called unnormalized because it does not
+account for the number of tips in the tree. Among two trees having the same
+number of tips, the least-balanced tree will have the highest Sackin's index.
+However, among two equally balanced trees, the larger tree will have a higher
+Sackin's index. This makes it challenging to compare balances among trees of
+different sizes. To correct this, \textcite{kirkpatrick1993searching} derive
+the expected value of Sackin' index under the Yule
+model~\autocite{yule1925mathematical}. Dividing by this expected value
+normalizes Sackin's index, so that it can be used to compare trees of different
+sizes. An example of a pairwise measure is the
 \gls{nltt}~\autocite{janzen2015approximate}, which compares the
@@
 \gls{ltt} plots are normalized so that they begin at $(0, 0)$ and end at $(1,
-1)$, and the difference between the two plots is integrated between 0 and 1. In
-the context of infectious diseases, the \gls{ltt} is related to the
+1)$, and the absolute difference between the two plots is integrated between 0
+and 1. In the context of infectious diseases, the \gls{ltt} is related to the
 prevalence~\autocite{holmes1995revealing}, so large values may indicate that
@@
 \[  
-  K(x, x') = <\varphi(x), \varphi(x')>,
+  K(x, x') = \langle \varphi(x), \varphi(x')\rangle,
 \]
@@
 below the epidemic threshold would result in exponential decay in the number of
-infected individuals and eventual extinction of the pathogen (Anderson \& May,
-I think).
+infected individuals and eventual extinction of the
+pathogen~\autocite{anderson1992infectious}.
 
@@
 A number of studies have made progress in quantifying the relationship between
-contact networks and transmission trees. \textcite{o2010contact} simulated
+contact networks and transmission trees. \textcite{o2011contact} simulated
 epidemics over networks with four types of degree distribution. They then
@@
 \textcite{villandre2016assessment} investigated the correspondence between
-contact network clusters and transmission tree clusters, and found a moderate
-correspondence between the two.
+contact network clusters and transmission tree clusters, and did find a
+moderate correspondence between the two in some cases.
 
@@
 
-\Gls{SMC} is a statistical inference method which samples from a sequence of
-probability distributions in a fixed order~\autocite{del2006sequential}, based
-on a technique called \gls{SIS}.
+\glsreset{SMC}
+
+\subsection{Overview and notation}
+
+\Gls{SMC} is the name for a family of statistical inference methods which rely
+on approximating probability distributions of interest with large collections
+of \defn{particles}, here denoted
+$\set{x^{(k)}}$~\autocite{doucet2001introduction, liu2008monte}. These
+collections or \defn{populations} are constructed to form a \defn{Monte Carlo
+approximation} to some distribution of interest $\pi$, meaning that the
+empirical distribution of the particles converges in distribution to $\pi$ as
+the population size gets large~\autocite{liu2001theoretical}. The word
+\defn{sequential} is used because the particle populations are modified in an
+iterative fashion over time, for example, to incorporate new evidence.
 
@@
 for all $x \in X$. This is an ``operational'' definition of Markov kernel which
-will be suitable for our purposes. A more precise definition can be found in
-\textbf{TODO}. Note that a Markov kernels have nothing to do with the kernel
-functions defined in \cref{subsec:treeshape}, other than sharing a name (the
-word ``kernel'' is ubiquitous in mathematics).
-
-\subsection{Sequential importance sampling}
+will be suitable for our purposes. Note that a Markov kernels have nothing to
+do with the kernel functions defined in \cref{subsec:treeshape}, other than
+sharing a name (the word ``kernel'' is ubiquitous in mathematics).
+
+\subsection{Sequential importance sampling for sequential Monte Carlo}
 \label{subsec:sis}
 
-The basis of \gls{SIS} is \gls{IS}, which is a method of estimating summary
-statistics of distributions which are known only up to a normalizing constant,
-and therefore cannot be sampled from directly. That is, if $\pi$ is such a
-distribution and $f$ is any real-valued function, \gls{IS} is concerned with
-estimating
+\Gls{SIS} is one type of \gls{SMC} method, whose aim is to sample from a
+distribution $\pi$ on an high-dimensional space, say $\pi(\vec{x}) = \pi(x_1,
+\ldots, x_d)$. The basis of \gls{SIS} is \gls{IS}, which is a method of
+estimating summary statistics of distributions which are known only up to a
+normalizing constant, and therefore cannot be sampled from directly. That is,
+if $\pi$ is such a distribution and $f$ is any real-valued function, \gls{IS}
+is concerned with estimating
 \[
@@
 
-\gls{SIS} applies \gls{IS} to the problem of sampling from a distribution $\pi$ 
-on an high-dimensional space, say $\pi(\vec{x}) = \pi(x_1, \ldots, x_d)$. The
-objective is to build up an importance distribution $\eta$ for $\pi$
-sequentially. By the general product rule, $\pi(\vec{x})$ can be decomposed as
+The objective of \gls{SIS} is to build up an importance distribution $\eta$ for
+$\pi$ sequentially. By the general product rule, $\pi(\vec{x})$ can be
+decomposed as
 \[
@@
 
-\subsection{Sequential Monte Carlo samplers}
+\subsection{The sequential Monte Carlo sampler}
 \label{subsec:smcsamp}
@@
   \label{eq:impint}
-  \eta_i(x) = \int \eta_{i-1}(\xi) K_i(\xi, x) \d\xi,
+  \eta_i(x') = \int \eta_{i-1}(x) K_i(x, x') \d x,
 \end{align}
@@
 \[
-  K_i(\xi, x) = \max\left(1, \frac{q(x, \xi)\pi_i(\xi)}{q(\xi, x)\pi_i(x)}\right),
+  K_i(x, x') = \max\left(1, \frac{q(x', x)\pi_i(x)}{q(x, x')\pi_i(x')}\right),
 \]
@@
 sample from $\pi_i$, by considering only the $i$th component of $\vec{x_i}$.
-These are all the ingredients we need to apply \gls{SMC}. The sequences of
+These are all the ingredients we need to apply \gls{SIS}. The sequences of
 kernels $L$ and $K$ should be chosen based on the problem at hand to inimize
@@
 the \defn{optimal} backward kernels. The full \gls{SMC} sampler algorithm is
-presented as \cref{alg:smcsamp}.
+presented as \cref{alg:smcsamp}. A resampling step is applied whenever the
+\gls{ESS} of the population drops too low, as discussed in the previous
+section.
 
@@
 A \defn{mathematical model} is a formal description of a hypothesized
-relationship between some observed data, $\vec{x} = \set{x_1, \ldots, x_n}$,
-and outcomes, $\vec{y} = \set{y_1, \ldots, y_n}$. A \defn{parametric} model
-defines a family of possible relationships between data and outcomes, indexed
-by one or more numeric parameters $\theta$. A \defn{statistical} model
-describes the relationship between data and outcomes in terms of probabilities.
-Statistical models define, either explicitly or implicitly, the probability of
-observing $\vec{y}$ given $\vec{x}$ and, if the model is parametric, $\theta$.
-In this context, the observed outcomes are taken to be realizations of random
-variables $\vec{Y} = \set{Y_1, \ldots, Y_n}$. Note that it is entirely possible
-to have no data $\vec{x}$, only observed outcomes $\vec{y}$. In this case, a
-model would describe the process by which $\vec{y}$ is generated.
+relationship between some observed data, $x$ and outcomes $y$. A
+\defn{parametric} model defines a family of possible relationships between data
+and outcomes, indexed by one or more numeric parameters $\theta$. A
+\defn{statistical} model describes the relationship between data and outcomes
+in terms of probabilities. Statistical models define, either explicitly or
+implicitly, the probability of observing $y$ given $\vec{x}$ and, if the
+model is parametric, $\theta$. Note that it is entirely possible to have no
+data $\vec{x}$, only observed outcomes $y$. In this case, a model would
+describe the process by which $y$ is generated.
 
@@
 clarity, we will restrict our attention to the case of one-dimensional data and
-outcomes where each $x_i$ and $y_i$ is a single real number. The linear model
-postulates that the outcomes are linearly related to the data, modulo some
-noise introduced by measurement error, environmental fluctuations, and other
-external factors. Formally, $y_i = \beta x_i + \varepsilon_i$, where $\beta$ is
-the slope of the linear relationship, and $\varepsilon_i$ is the error
-associated with measurement $i$. We can make this model a statistical one by
-hypothesizing a distribution for the error terms $\varepsilon_i$; most
-commonly, it is assumed that they are normally distributed with variance
-$\sigma$. In mathematical terms, $Y_i \sim \beta x_i + \N(0, \sigma^2)$, where
-``$\sim$'' means ``is distributed as''. We can see from this formulation that
-the model is parametric, with parameters $\theta$ = ($\beta$, $\sigma$).
-Moreover, we can write down the probability density of observing outcome $y_i$
-given the parameters,
-\[
-  \Pr(Y_i = y_i \mid \beta, \sigma) = 
-  \Pr\left(\N(0, \sigma^2) = y_i - \beta x_i\right).
-\]
-Note that we have followed the standard statistical abuse of notation and used
-``$\Pr$'' to refer to a probability \emph{density}, rather than a probability.
-Also, we are treating the $x_i$ as fixed quantities, and therefore have not
-conditioned the probability density on $\vec{x}$. Assuming all the $y_i$ are
-independent, the probability density of the entire observed set of outcomes is
-the product of the probability density of each individual $y_i$,
-\[
-  \Pr(\vec{Y} = \vec{y} \mid \beta, \sigma) = 
-  \prod_{i=1}^N \Pr(Y_i = y_i \mid \beta, \sigma).
-\]
-
-For a general model, the probability density of $\vec{y}$ given the parameters
+outcomes where $x = \set{x_1, \ldots, x_n}$ and $y = \set{y_1, \ldots, y_n}$
+are vectors of real numbers. The linear model postulates that the outcomes are
+linearly related to the data, modulo some noise introduced by measurement
+error, environmental fluctuations, and other external factors. Formally, $y_i =
+\beta x_i + \varepsilon_i$, where $\beta$ is the slope of the linear
+relationship, and $\varepsilon_i$ is the error associated with measurement $i$.
+We can make this model a statistical one by hypothesizing a distribution for
+the error terms $\varepsilon_i$; most commonly, it is assumed that they are
+normally distributed with variance $\sigma$. In mathematical terms, $Y_i \sim
+\beta x_i + \N(0, \sigma^2)$, where ``$\sim$'' means ``is distributed as''. We
+can see from this formulation that the model is parametric, with parameters
+$\theta$ = ($\beta$, $\sigma$). Moreover, we can write down the probability
+density $\pi$ of observing outcome $y_i$ given the parameters,
+\[
+  \pi(y \mid \beta, \sigma) = 
+  \prod_{i=1}^n f_{\N(0, \sigma^2)} (y_i - \beta x_i),
+\]
+where $f_{\N(0, \sigma^2)}$ is the probability density of the normal
+distribution with mean zero and variance $\sigma^2$. Note that we are treating
+the $x_i$ as fixed quantities, and therefore have not conditioned the
+probability density on $\vec{x}$. Also, we have assumed that all the $y_i$ are
+independent.
+
+TODO
+
+For a general model, the probability density of $y$ given the parameters
 $\theta$ is also known as the \defn{likelihood}, written $\L$, of $\theta$.
-That is, $\L(\theta \mid \vec{y}) = \Pr(\vec{y} \mid \theta)$. The higher the
-value of the likelihood, the more likely the observations $\vec{y}$ are under
-the model. Thus, the likelihood provides a natural criterion for fitting the
-model parameters: we want to pick $\theta$ such that the probability density of
-our observed outcomes $\vec{y}$ is as high as possible. The parameters which
-optimize the likelihood are known as the \textit{\gls{ML}} estimates, denoted
-$\hat{\theta}$. That is,
-\[
-  \hat{\theta} = \argmax_\theta\; \Pr(\vec{y} \mid \theta).
+That is, $\L(\theta \mid y) = f(y \mid \theta)$ for the model's \gls{pdf} $f$.
+The higher the value of the likelihood, the more likely the observations $y$
+are under the model. Thus, the likelihood provides a natural criterion for
+fitting the model parameters: we want to pick $\theta$ such that the
+probability density of our observed outcomes $y$ is as high as possible. The
+parameters which optimize the likelihood are known as the \textit{\gls{ML}}
+estimates, denoted $\hat{\theta}$. That is,
+\[
+  \hat{\theta} = \argmax_\theta\; \L(\theta \mid y).
 \]
@@
 simplest terms, many possible values for $\theta$ are examined, $\L(\theta \mid
-\vec{y})$ is calculated for each, and the parameters which produce the highest
-value are accepted. Many sophisticated numerical optimization methods exist,
-although they may not be guaranteed to find the true \gls{ML} estimates if the
-likelihood function is complex. Occasionally, as is the case with linear
-regression, the \gls{ML} estimates can be found explicitly by setting the
-likelihood function's derivatives to zero.
+y)$ is calculated for each, and the parameters which produce the highest value
+are accepted. Many sophisticated numerical optimization methods exist, although
+they may not be guaranteed to find the true \gls{ML} estimates if the
+likelihood function is complex.
 
@@
 beliefs as a \defn{prior distribution} on the parameters, denoted
-$\Pr(\theta)$. Instead of considering only the likelihood, Bayesian inference
-focuses on the product of the likelihood and the prior, $\Pr(\vec{y} \mid
-\theta) \Pr(\theta)$. Bayes' theorem tells us that this product is related to
-the \textit{posterior distribution} on $\theta$,
-\[
-  \Pr(\theta \mid \vec{y}) = \frac{\Pr(\vec{y} \mid \theta) \Pr(\theta)}
-                                  {\Pr(\vec{y})}.
-\]
-
-In principle, $\Pr(\vec{y} \mid \theta) \Pr(\theta)$ can be optimized
-numerically just like $\L(\theta \mid \vec{y})$, which would also optimize the
-posterior distribution. The resulting optimal parameters are called the
-\gls{MAP} estimates. However, from a Bayesian perspective, $\theta$ is not a
-fixed quantity to be estimated, but rather a random variable with an associated
-distribution (the posterior). Therefore, the \gls{MAP} estimate by itself is of
-limited value without associated statistics about the posterior distribution,
-such as the mean or credible intervals. Unfortunately, to calculate such
-statistics, it is necessary to evaluate the normalizing constant
-$\Pr(\vec{y})$, which is almost always an intractable integral.
+$\pi(\theta)$. Instead of considering only the likelihood, Bayesian inference
+focuses on the product of the likelihood and the prior, $f(y \mid \theta)
+\pi(\theta)$. Bayes' theorem tells us that this product is related to the
+\textit{posterior distribution} on $\theta$,
+\begin{align}
+  f(\theta \mid y) 
+    = \frac{f(y \mid \theta) \pi(\theta)}
+           {\int f(y \mid \theta) \pi(\theta) \d \theta}.
+  \label{eq:bayes}
+\end{align}
+In principle, $f(y \mid \theta) \pi(\theta)$ can be optimized numerically just
+like $\L(\theta \mid y)$, which would also optimize the posterior distribution.
+The resulting optimal parameters are called the \gls{MAP} estimates. However,
+from a Bayesian perspective, $\theta$ is not a fixed quantity to be estimated,
+but rather a random variable with an associated distribution (the posterior).
+Therefore, the \gls{MAP} estimate by itself is of limited value without
+associated statistics about the posterior distribution, such as the mean or
+credible intervals. Unfortunately, to calculate such statistics, it is
+necessary to evaluate the normalizing constant in the denominator of
+\cref{eq:bayes}, which is almost always an intractable integral.
 
@@
 \[
-  \max\left(1, \frac{\Pr(\vec{y} \mid \theta_2) \Pr(\theta_2) q(\theta_2, \theta_1)}
-                    {\Pr(\vec{y} \mid \theta_1) \Pr(\theta_2) q(\theta_1, \theta_2)} \right),
+  \max\left(1, \frac{f(y \mid \theta_2) \pi(\theta_2) q(\theta_2, \theta_1)}
+                    {f(y \mid \theta_1) \pi(\theta_2) q(\theta_1, \theta_2)} \right),
 \]
@@
 enough random walk is performed on the chain, the distribution of states
-visited will be a Monte Carlo approximation of $\Pr(\theta \mid \vec{y})$, from
+visited will be a Monte Carlo approximation of $\Pr(\theta \mid y)$, from
 which we can calculate statistics of interest. Actually performing this random
@@
   \begin{algorithmic}
-    \State Draw $\theta$ according to $\Pr(\theta)$
+    \State Draw $\theta$ according to the prior $\pi(\theta)$
     \Loop
@@
       $\max \left( 1, 
-       \dfrac{\Pr(\vec{y} \mid \theta') \Pr(\theta') q(\theta', \theta)}
-             {\Pr(\vec{y} \mid \theta\phantom{'}) \Pr(\theta\phantom{'}) q(\theta, \theta')}
+       \dfrac{f(y \mid \theta') \pi(\theta') q(\theta', \theta)}
+             {f(y \mid \theta\phantom{'}) \pi(\theta\phantom{'}) q(\theta, \theta')}
        \right)$
@@
 approaches, \gls{ML} or Bayesian inference, discussed above. However, there are
-some, particularly in the domain of population genetics, for which calculation
-of either the likelihood or the product of the likelihood and the prior may be
-infeasible. For example, one or both of these quantities may be expressible
-only as an intractable integral. \Gls{ABC} is designed for such cases, where
-standard likelihood-based techniques for model fitting cannot be applied. Such
-models are particularly prevalent in population
-genetics~\autocite{beaumont2002approximate, beaumont2010approximate}.
-
-Ordinarily, Bayesian inference targets the posterior distribution $\Pr(\theta
-\mid \vec{y})$. That is, in the Bayesian framework, model parameters with
-higher posterior density are ``better'' in the sense that they offer a more
-credible explanation for the observed data. Approximate Bayesian computation
-offers an alternative metric for parameter credibility, namely the similarity
-of simulated datasets to the observed data. If datasets simulated under the
-model closely resemble the real data, it follows that the model is a reasonable
+some, particularly in the domain of population
+genetics~\autocite{beaumont2002approximate, beaumont2010approximate}, for which
+calculation of either the likelihood or the product of the likelihood and the
+prior may be infeasible. For example, one or both of these quantities may be
+expressible only as an intractable integral. \Gls{ABC} is designed for such
+cases, where standard likelihood-based techniques for model fitting cannot be
+applied.
+
+Ordinarily, Bayesian inference targets the posterior distribution $f(\theta
+\mid y)$. That is, in the Bayesian framework, model parameters with higher
+posterior density are ``better'' in the sense that they offer a more credible
+explanation for the observed data. Approximate Bayesian computation offers an
+alternative metric for parameter credibility, namely the similarity of
+simulated datasets to the observed data. If datasets simulated under the model
+closely resemble the real data, it follows that the model is a reasonable
 approximation to the real-world process generating the observed data. More
@@
 possible data our model could generate. \gls{ABC} aims to sample from the joint
-posterior distribution of model parameters and simulated datasets $\vec{z}$
-which are within some small distance $\varepsilon$ of the observed data
-$\vec{y}$,
-\[
-  \Pr(\theta, \vec{z} \mid \vec{y}, \varepsilon) =
-  \frac{\Pr(\theta) \Pr(\vec{z} \mid \theta) \I_{A_{\varepsilon, \vec{y}}} (\vec{z})}
-  {\int_{A_{\varepsilon, \vec{y}} \times \Theta} \Pr(\theta) \Pr(\vec{z} \mid \theta) \d \theta}.
-\]
-Here, $A_{\varepsilon, \vec{y}}$ is an $\varepsilon$-ball around $\vec{y}$ with
+posterior distribution of model parameters and simulated datasets $z$ which are
+within some small distance $\varepsilon$ of the observed data $y$,
+\[
+  \pi_{\varepsilon}(\theta, z \mid y) =
+  \frac{\pi(\theta) f(z \mid \theta) \I_{A_{\varepsilon, y}} (z)}
+  {\int_{A_{\varepsilon, y} \times \Theta} \pi(\theta) f(z \mid \theta) \d \theta}.
+\]
+Here, $A_{\varepsilon, y}$ is an $\varepsilon$-ball around $y$ with
 respect to $\rho$, $\Theta$ is the space of all possible model parameters, and
@@
 word ``approximate'' derives from the assumption that, for a suitably chosen
-distance $\rho$ and a small enough $\varepsilon$, the marginal in $\vec{z}$ of
+distance $\rho$ and a small enough $\varepsilon$, the marginal in $z$ of
 this distribution approximates the posterior of
@@
 \[
-  \int \Pr(\theta, \vec{z} \mid \vec{y}, \varepsilon) \d\vec{z}
-  \approx \Pr(\theta \mid \vec{y}).
-\]
-To emphasize that the distribution on the left only approximates the posterior,
-we will refer to it as the \textit{\gls{ABC} target distribution}. Note that in
+  \int \pi_\varepsilon(\theta, z \mid y) \d z \approx f(\theta \mid y).
+\]
+This distribution is variously referred to as the  \textit{\gls{ABC} target
+distribution} or the \gls{ABC} approximation to the posterior. Note that in
 many formulations, the distance function $\rho$ is defined as $\rho(S(\cdot),
@@
 parameter particle (the meaning of this will become clear shortly). The
-extension to multiple datasets per particle will be given at the end.
+extension to multiple datasets per particle is straightforward and will be
+given at the end of the section. We use the variable $x$ to refer to the pair
+$(\theta, z)$, so that the \gls{ABC} target distribution can be written
+$\pi_\varepsilon(x \mid y)$.
 
@@
 number of samples from the target distribution are obtained. Parameter values
-$\theta$ are sampled according to the prior distribution $\Pr(\theta)$. Then, a
-simulated dataset $\vec{z}$ is generated from the model with the sampled
-parameter values. By definition, the probability density of obtaining the
-particular dataset $\vec{z}$ is $\Pr(\vec{z} \mid \theta)$. Finally, the
-parameters are sampled if the distance of $\vec{z}$ from the observed data
-$\vec{y}$ is less than $\varepsilon$ - that is, with probability
-$\I_{A_{\varepsilon, \vec{y}}}(\vec{z})$. Putting this all together, the
+$\theta$ are sampled according to the prior distribution $\pi(\theta)$. Then, a
+simulated dataset $z$ is generated from the model with the sampled parameter
+values. By definition, the probability density of obtaining the particular
+dataset $z$ is $f(z \mid \theta)$. Finally, the parameters are sampled if the
+distance of $z$ from the observed data $y$ is less than $\varepsilon$, that is,
+with probability $\I_{A_{\varepsilon, y}}(z)$. Putting this all together, the
 parameters $\theta$ are sampled with probability proportional to
 \[
-  \Pr(\theta) \Pr(\vec{z} \mid \theta) \I_{A_{\varepsilon, \vec{y}}}(\vec{z}),
+  \pi(\theta) f(z \mid \theta) \I_{A_{\varepsilon, y}}(z),
 \]
@@
     \Loop
-      \State Draw $\theta$ according to $\Pr(\theta)$
-      \State Simulate a dataset $\vec{z}$ from the model with parameters $\theta$
-      \If{$\rho(\vec{y}, \vec{z}) < \varepsilon$}
+      \State Draw $\theta$ according to $\pi(\theta)$
+      \State Simulate a dataset $z$ from the model with parameters $\theta$
+      \If{$\rho(y, z) < \varepsilon$}
         \State Sample $\theta$
@@
 prior, a very large number of samples may need to be taken in order to find a
-simulated dataset which is close to $\vec{z}$. The inefficiency is compounded
+simulated dataset which is close to $z$. The inefficiency is compounded
 by the curse of dimensionality - the measure of the $\varepsilon$-ball around
-$\vec{y}$ decreases exponentially with the number of dimensions.
+$y$ decreases exponentially with the number of dimensions.
 \gls{ABC}-\gls{MCMC} (\cref{alg:abcmcmc}) was designed to overcome these
@@
 replaces the likelihood ratio. That is, the transition probability between
-states $\theta_1$ and $\theta_2$, with simulated datasets $\vec{z_1}$ and
-$\vec{z_2}$, is defined as
-\[
-  \max\left(1, \frac{\Pr(\vec{z_2} \mid \theta_2) q(\theta_2, \theta_1)}
-                    {\Pr(\vec{z_1} \mid \theta_1) q(\theta_1, \theta_2)} 
-    \cdot \I_{A_{\varepsilon, \vec{y}}}(\vec{z_2}) \right).
+states $x$ and $x'$ is defined as
+\[
+  \max\left(1, \frac{f(z' \mid \theta') q(\theta', \theta)}
+                    {f(z \mid \theta) q(\theta, \theta')} 
+    \cdot \I_{A_{\varepsilon, y}}(z') \right).
 \]
@@
 \begin{algorithm}
-  \caption{ABC-MCMC.}
+  \caption{\gls{ABC}-\gls{MCMC}.}
   \begin{algorithmic}
@@
       $\max \left( 1, 
-       \dfrac{\Pr(\theta') q(\theta', \theta)}
-             {\Pr(\theta\phantom{'}) q(\theta, \theta')} 
-       \cdot \I_{A_{\varepsilon, \vec{y}}}(\vec{z'}) \right)$
+       \dfrac{\pi(\theta') q(\theta', \theta)}
+             {\pi(\theta\phantom{'}) q(\theta, \theta')} 
+       \cdot \I_{A_{\varepsilon, y}}(z') \right)$
     \EndLoop
@@
 (\cref{subsec:smcsamp}). The sequence of target distributions is defined as
-$\pi_i = \Pr(\theta, \vec{z} \mid \vec{y}, \varepsilon_i)$ for a decreasing
-sequence of tolerances $\varepsilon_i$. The intention is for the algorithm to
-progress smoothly through a sequence of target distributions which ends at the
-\gls{ABC} approximation to the posterior. As discussed in
-\cref{subsec:smcsamp}, the choices of the kernels $K$ and $L$ is
-problem-specific, and so appropriate kernels must be chosen for \gls{ABC}.
-Several options have been proposed~\autocite{beaumont2009adaptive,
-sisson2007sequential, del2012adaptive}.
+$\pi_i = \pi_{\varepsilon_i}(x \mid y)$ for a decreasing sequence of tolerances
+$\varepsilon_i$. The intention is for the algorithm to progress smoothly
+through a sequence of target distributions which ends at the \gls{ABC}
+approximation to the posterior. As discussed in \cref{subsec:smcsamp}, the
+choices of the kernels $K$ and $L$ is problem-specific, and so appropriate
+kernels must be chosen for \gls{ABC}. Several options have been
+proposed~\autocite{beaumont2009adaptive, sisson2007sequential,
+del2012adaptive}.
 
@@
 particle instead of just one. The simulated dataset $z$ is replaced by
-$\vec{z} = z_1, \ldots, z_M$, and the indicator function for the
+$z = z_1, \ldots, z_M$, and the indicator function for the
 $\varepsilon$-ball around $y$ is replaced by

### 2016-04-19 doc/thesis/methods.tex
Finished draft intro!

@@
 \[
-  q_i(\theta, \theta') \prod_{k=1}^M \Pr(X_i^{(k)}' \mid \theta'),
+  q_i(\theta, \theta') \prod_{k=1}^M \Pr(X_i^{(k)'} \mid \theta'),
 \]
@@
 classifiers for \gls{alpha} based on three statistics. First, we used the
-\software{kernlab} package~\autocite{zeileis2004kernlab} to create a \gls{kSVM}
+\software{kernlab} package~\autocite{zeileis2004kernlab} to create a \gls{kSVR}
 classifier using the computed kernel matrix. Second, we used the
 \software{e1071} package~\autocite{meyer2015e1071} to create an ordinary
-\gls{SVM} classifier using the pairwise \gls{nltt} matrix. Finally, we
+\gls{SVR} classifier using the pairwise \gls{nltt} matrix. Finally, we
 performed an ordinary linear regression of \gls{alpha} against Sackin's index.

### 2016-04-19 doc/thesis/results.Rnw
Finished draft intro!

@@
     library(Hmisc)
+
+    pp <- function (p) {
+        if (p < 1e-5) {
+            "{<}10^{-5}"
+        } else {
+            latexSN(round(p, -floor(log10(p))))
+        }
+    }
 @
@@
     params <- c("alpha", "I", "m", "N")
-
-    pp <- function (p) {
-        if (p < 1e-5) {
-            "<10^{-5}"
-        } else {
-            latexSN(round(p, -floor(log10(p))))
-        }
-    }
 
@@
     \Sexpr{round(N.ratio.cor$estimate, 2)};
-  $p$-values
-    $\Sexpr{latexSN(round(alpha.ratio.cor$p.value, -floor(log10(alpha.ratio.cor$p.value))))}$,
-    $\Sexpr{latexSN(round(I.ratio.cor$p.value, -floor(log10(I.ratio.cor$p.value))))}$,
-    $\Sexpr{latexSN(round(m.ratio.cor$p.value, -floor(log10(m.ratio.cor$p.value))))}$,
-    $\Sexpr{latexSN(round(N.ratio.cor$p.value, -floor(log10(N.ratio.cor$p.value))))}$).
+  all $p < 10^{-5}$).
 
@@
 first two principal components of the kernel matrix. The figure shows only the
-simulations with 500-tip trees and, for all parameters except $I$, 1000
-infected nodes. The three \gls{alpha} and \gls{I} values considered are well
-separated from each other in feature space. On the other hand, the three
-\gls{N} values overlap significantly, and the three \gls{m} values are
-virtually indistinguishable. Similar observations can be made for other values
-of \gls{I} and the number of tips
+simulations with 500-tip trees and 1000 infected nodes. The three \gls{alpha}
+and \gls{I} values considered are well separated from each other in feature
+space. On the other hand, the three \gls{N} values overlap significantly, and
+the three \gls{m} values are virtually indistinguishable. Similar observations
+can be made for other values of \gls{I} and the number of tips
 (\cref{fig:alphakpca,fig:Nkpca,fig:Ikpca,fig:mkpca}). The values of \gls{I} and
-\gls{N} separate more clearly with larger numbers of tips, and in the case of
-\gls{N} larger epidemic sizes.
+\gls{N} separated more clearly with larger numbers of tips, and in the case of
+\gls{N}, larger epidemic sizes.
 
@@
 
-Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
+Accuracy of the \gls{kSVR} classifiers varied based on the parameter being
 tested (\cref{fig:rsquared}, left). Classifiers based on two other tree
@@
 
-When classifying $\alpha$, the kernel-SVM classifier had an average $R^2$ of 
+When classifying $\alpha$, the \gls{kSVR} classifier had an average $R^2$ of 
     \Sexpr{kernel.alpha[,round(mean(rsquared), 2)]},
@@
     \Sexpr{nltt.alpha[,round(mean(rsquared), 2)]}
-for the \gls{nltt}-based SVM, and
+for the \gls{nltt}-based SVR, and
     \Sexpr{sackin.alpha[,round(mean(rsquared), 2)]}
@@
 $R^2$ values of 
-  \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for \gls{kSVM},
+  \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for \gls{kSVR},
   \Sexpr{nltt.m[,round(mean(rsquared), 2)]} for the \gls{nltt}, and
@@
 for Sackin's index. Again, there was little variation in accuracy between
-epidemic scenarios, although the accuracy of the \gls{kSVM} was slightly higher
-on 1000-tip trees.
+epidemic scenarios, although the accuracy of the \gls{kSVR} was slightly higher
+on 1000-tip trees 
+    (average $R^2$ 
+     \Sexpr{kernel.m[ntip == 100,round(mean(rsquared, 2))]},
+     \Sexpr{kernel.m[ntip == 500,round(mean(rsquared, 2))]},
+     \Sexpr{kernel.m[ntip == 1000,round(mean(rsquared, 2))]}
+     for 100, 500, and 1000 tips respectively).
 
@@
 Finally, the performance of classifiers for $N$ depended heavily on the
-epidemic scenario. The $R^2$ of the \gls{kSVM} classifier ranged from
+epidemic scenario. The $R^2$ of the \gls{kSVR} classifier ranged from
   \Sexpr{kernel.N[,round(min(rsquared), 2)]}
@@
   \Sexpr{kernel.N[,round(max(rsquared), 2)]}
-for the largest. Likewise, $R^2$ for the \gls{nltt}-based SVM ranged from 
+for the largest. Likewise, $R^2$ for the \gls{nltt}-based SVR ranged from 
   \Sexpr{nltt.N[,round(min(rsquared), 2)]}
@@
   \caption{
-      Cross-validation accuracy of kernel-SVM classifier (left), SVM classifier
+      Cross-validation accuracy of kernel-SVR classifier (left), SVR classifier
       using \gls{nltt} (centre), and linear regression using Sackin's index
@@
 
-The accuracy of grid search estimates largely paralleled that of the \gls{kSVM}
+The accuracy of grid search estimates largely paralleled that of the \gls{kSVR}
 classifiers. \Cref{fig:gridest} shows point estimates and 95\% highest density
@@
     \Sexpr{tip.cor["N", round(estimate, 2)]};
-$p$-value = 
-    $\Sexpr{tip.cor["alpha", latexSN(round(p.value, -floor(log10(p.value))))]}$,
-    $\Sexpr{tip.cor["I", latexSN(round(p.value, -floor(log10(p.value))))]}$,
-    $\Sexpr{tip.cor["N", latexSN(round(p.value, -floor(log10(p.value))))]})$.
+$p$-values
+    $\Sexpr{tip.cor["alpha", pp(p.value)]}$,
+    $\Sexpr{tip.cor["I", pp(p.value)]}$,
+    $\Sexpr{tip.cor["N", pp(p.value)]}$).
 The highest density intervals obtained for all parameters were extremely wide,
-occcupying $>$90\% of the grid in all cases.
+occcupying $>$90\% of the grid in all cases (\cref{fig:gridest}).
 
@@
     (Spearman's $\rho$ = \Sexpr{value.cor["alpha", round(estimate, 2)]},
-     $p = \Sexpr{value.cor["alpha", latexSN(round(p.value, -floor(log10(p.value))))]}$),
+     $p = \Sexpr{value.cor["alpha", pp(p.value)]}$),
 although the relationship was clearly nonlinear (\cref{fig:gridest,fig:gridalpha}).
@@
 the true value of \gls{I}. Kernel score distributions for all test values
-exhibited the same rounded shape (\cref{fig:gridI}).
+exhibited a similar rounded shape (\cref{fig:gridI}).
 
-For the other two parameters, \gls{m} and \gls{N}, accuracy was much poorer
-(\cref{fig:gridest,fig:gridm,fig:gridN}). The average error for \gls{m} was
+The average error for \gls{m} was
     \Sexpr{d[parameter == "m", round(mean(error), 2)]},
@@
     (Spearman's $\rho$ = \Sexpr{value.cor["m", round(estimate, 2)]},
-     $p = \Sexpr{value.cor["m", latexSN(round(p.value, -floor(log10(p.value))))]}$).
-This positive correlation was apparently due to the much lower error for $m =
-1$ than for the other $m$ values (\cref{fig:gridm})
-    (mean errors \Sexpr{d[parameter == "m" & true_value == 1, mean(error)]} for
-     $m = 1$ vs. \Sexpr{d[parameter == "m" & true_value > 1, mean(error)]} for
-     $m > 1$).
+     $p = \Sexpr{value.cor["m", pp(p.value)]}$).
+This positive correlation was apparently due to the much lower error for
+\gls{m} = 1 and 2 than for the other $m$ values 
+    (mean errors 
+    \Sexpr{d[parameter == "m" & true_value <= 2, round(mean(error), 2)]} 
+    for $m \leq 2$ vs. 
+    \Sexpr{d[parameter == "m" & true_value > 2, round(mean(error), 2)]} 
+    for $m > 2$, \cref{fig:gridm}).
 The value $m = 1$ causes the network to take on a distinct shape relative to
-higher $m$ values, namely a tree (\textit{i.e.} there are no cycles,
-\cref{subsec:treeshape}). The average error for \gls{N} was 
+higher \gls{m} values, namely a tree (\textit{i.e.} there are no cycles,
+see \cref{subsec:treeshape}). The average error for \gls{N} was 
     \Sexpr{d[parameter == "N", round(mean(error))]},
@@
     (Spearman's $\rho$ = \Sexpr{value.cor["N", round(estimate, 2)]},
-     $p = \Sexpr{value.cor["N", latexSN(round(p.value, -floor(log10(p.value))))]}$).
+    $p \Sexpr{value.cor["N", pp(p.value)]}$).
 
@@
     I.av <- anova(lm(I_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
+    stopifnot(I.av["factor(true_m)", "Pr(>F)"] >= 0.05)
+    stopifnot(m.av["factor(true_m)", "Pr(>F)"] >= 0.05)
+    stopifnot(m.av["factor(true_I)", "Pr(>F)"] >= 0.05)
+    stopifnot(min(na.omit(N.av[,"Pr(>F)"])) >= 0.05)
 
@@
                                 d[true_alpha == 0.5 | true_alpha == 1.5, m_error])
-
     options(scipen=-1, digits=2)
@@
 
-We used kernel-\gls{ABC} to estimate the parameters of the \gls{BA} model 
-on simulated trees where the true parameter values were known. Point estimates
-for each parameter are shown in Figure~\ref{fig:abcpt}. Of the four parameters,
-\gls{alpha} was the most accurately estimated, with a median [IQR] absolute error
-of 
+We used \software{netabc} to estimate the parameters of the \gls{BA} model on
+simulated trees where the true parameter values were known. Point estimates for
+each parameter are shown in \cref{fig:abcpt} for the simulations with \gls{m} =
+2. The results for the other values of \gls{m} were similar
+(\cref{fig:abcm3,fig:abcm4}). The median [IQR] absolute error of estimates of
+\gls{alpha} across all simulations was
     \Sexpr{d[,median(alpha_error)]} 
@@
 significantly greater than the other values 
-    (Wilcoxon rank-sum test, $p$ = \Sexpr{zero.alpha.test$p.value}).
+    (Wilcoxon rank-sum test, $p = \Sexpr{pp(zero.alpha.test$p.value)}$).
 The error in the estimated value of $I$ was
@@
 Errors were significantly higher for $\alpha \geq 1$
-    (Wilcoxon rank-sum test, $p$ = \Sexpr{alpha.I.test$p.value})
+    (Wilcoxon rank-sum test, $p = \Sexpr{pp(alpha.I.test$p.value)}$)
 and for $I$ = 2000
-    ($p$ = \Sexpr{I.I.test$p.value}),
-but not for any values of $m$
-    (one-way ANOVA, $p$ = \Sexpr{I.av["factor(true_m)", "Pr(>F)"]}).
-The $m$ parameter was estimated correctly in
+    ($p = \Sexpr{pp(I.I.test$p.value)}$),
+but not for any values of $m$ (one-way ANOVA). The $m$ parameter was estimated
+correctly in only
     \Sexpr{as.integer(m.tbl[1] * 100)} \%
-of simulations, with an error of one in
-    \Sexpr{as.integer(m.tbl[2] * 100)} \%
-and of two or more in 
-    \Sexpr{as.integer(sum(m.tbl[3:length(m.tbl)]) * 100)} \%
-(the only possible $m$ values were 2, 3, 4, or 5). The true values of
-$m$ and $I$ did not significantly affect the error
-    (one-way ANOVA, $p$ = \Sexpr{m.av["factor(true_m)", "Pr(>F)"]} and
-                          \Sexpr{m.av["factor(true_I)", "Pr(>F)"]}),
-but the accuracy was significantly lower for integral than non-integral
-values of \gls{alpha}
-    (Wilcoxon rank-sum test, $p$ = \Sexpr{alpha.m.test$p.value}).
+of simulations. The true values of $m$ and $I$ did not significantly affect the
+error (one-way ANOVA), but the accuracy was significantly lower for integral
+than non-integral values of \gls{alpha}
+    (Wilcoxon rank-sum test, $p = \Sexpr{pp(alpha.m.test$p.value)}$).
 Finally, the total number of nodes \gls{N} was consistently over-estimated by
 about a factor of two
-    (error \Sexpr{d[,median(N_error)]} 
-    [\Sexpr{d[,quantile(N_error, 0.25)]}-\Sexpr{d[,quantile(N_error, 0.75)]}]).
-No other parameters influenced the accuracy of the $N$ estimates 
-    (one-way ANOVA, $p \geq$ \Sexpr{min(N.av[,"Pr(>F)"])}).
+    (error \Sexpr{d[,format(median(N_error), scientific=FALSE)]} 
+    [\Sexpr{d[,format(quantile(N_error, 0.25), scientific=FALSE)]} - 
+     \Sexpr{d[,format(quantile(N_error, 0.75), scientific=FALSE)]}]).
+No other parameters influenced the accuracy of the $N$ estimates (one-way
+ANOVA)
+
+\begin{figure}
+  \includegraphics{abc-point-estimate-m2}
+  \caption{
+    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by         
+    running \software{netabc} on simulated data. Values shown are for               
+    simulations with \gls{m} = 3. Dashed lines indicate true values. (A)            
+    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations  
+    against known values. (B) Estimates of \gls{m} and \gls{N} which were held   
+    fixed in these simulations at the values \gls{m} = 3 and \gls{N} = 5000. 
+  }
+  \label{fig:abcm2}
+\end{figure}
 

### 2016-04-19 doc/thesis/results.tex
Finished draft intro!

@@
-\subsection{Kernel classifiers}
+\subsection{Analysis of \acrlong{BA} model}
+
+\subsubsection*{Classifiers for parameters based on tree shape}
 
@@
      $p$-values
-     $<10^{-5}$,
+     ${<}10^{-5}$,
      $0.003$,
-     $<10^{-5}$,
-     $<10^{-5}$)
+     ${<}10^{-5}$,
+     ${<}10^{-5}$)
 The ratio of internal to terminal branch lengths was negatively correlated with
@@
     0.18;
-  $p$-values
-    $0$,
-    $2\!\times\!10^{-85}$,
-    $4\!\times\!10^{-7}$,
-    $0$).
+  all $p < 10^{-5}$).
 
@@
 
-When classifying $\alpha$, the kernel-SVM classifier had an average $R^2$ of 
+When classifying $\alpha$, the \gls{kSVR} classifier had an average $R^2$ of 
     0.92,
@@
     0.56
-for the \gls{nltt}-based SVM, and
+for the \gls{nltt}-based SVR, and
     0.75
@@
 $R^2$ values of 
-  0.12 for \gls{kSVM},
+  0.12 for \gls{kSVR},
   0.01 for the \gls{nltt}, and
@@
 for Sackin's index. Again, there was little variation in accuracy between
-epidemic scenarios, although the accuracy of the \gls{kSVM} was slightly higher
-on 1000-tip trees.
+epidemic scenarios, although the accuracy of the \gls{kSVR} was slightly higher
+on 1000-tip trees 
+    (average $R^2$ 
+     0,
+     0,
+     0
+     for 100, 500, and 1000 tips respectively).
 
@@
 Finally, the performance of classifiers for $N$ depended heavily on the
-epidemic scenario. The $R^2$ of the \gls{kSVM} classifier ranged from
+epidemic scenario. The $R^2$ of the \gls{kSVR} classifier ranged from
   0.08
@@
   0.82
-for the largest. Likewise, $R^2$ for the \gls{nltt}-based SVM ranged from 
+for the largest. Likewise, $R^2$ for the \gls{nltt}-based SVR ranged from 
   0.01
@@
 
-\subsection{Grid search}
+\subsubsection*{Marginal parameter estimates with grid search}
 
-The accuracy of grid search estimates largely paralleled that of the \gls{kSVM}
+The accuracy of grid search estimates largely paralleled that of the \gls{kSVR}
 classifiers. \Cref{fig:gridest} shows point estimates and 95\% highest density
@@
     \ensuremath{-0.22},
-    \ensuremath{-0.41},
+    \ensuremath{-0.51},
     \ensuremath{-0.16};
-$p$-value = 
+$p$-values
     $4\!\times\!10^{-4}$,
-    $2\!\times\!10^{-4}$,
-    $0.01)$.
+    ${<}10^{-5}$,
+    $0.01$).
 The highest density intervals obtained for all parameters were extremely wide,
-occcupying $>$90\% of the grid in all cases.
+occcupying $>$90\% of the grid in all cases (\cref{fig:gridest}).
 
@@
 average absolute error of the point estimates for \gls{I} was 
-    298,
+    310,
 on a grid of 500 to 5000, and the errors were not significantly correlated with
 the true value of \gls{I}. Kernel score distributions for all test values
-exhibited the same rounded shape (\cref{fig:gridI}).
+exhibited a similar rounded shape (\cref{fig:gridI}).
 
-For the other two parameters, \gls{m} and \gls{N}, accuracy was much poorer
-(\cref{fig:gridest,fig:gridm,fig:gridN}). The average error for \gls{m} was
+The average error for \gls{m} was
     1.31,
@@
      $p = 6\!\times\!10^{-4}$).
-This positive correlation was apparently due to the much lower error for $m =
-1$ than for the other $m$ values (\cref{fig:gridm})
-    (mean errors 0.1 for
-     $m = 1$ vs. 1.5466667 for
-     $m > 1$).
+This positive correlation was apparently due to the much lower error for
+\gls{m} = 1 and 2 than for the other $m$ values 
+    (mean errors 
+    0.93 
+    for $m \leq 2$ vs. 
+    1.49 
+    for $m > 2$, \cref{fig:gridm}).
 The value $m = 1$ causes the network to take on a distinct shape relative to
-higher $m$ values, namely a tree (\textit{i.e.} there are no cycles,
-\cref{subsec:treeshape}). The average error for \gls{N} was 
+higher \gls{m} values, namely a tree (\textit{i.e.} there are no cycles,
+see \cref{subsec:treeshape}). The average error for \gls{N} was 
     2419,
@@
     (Spearman's $\rho$ = 0.43,
-     $p = 2\!\times\!10^{-12}$).
+    $p {<}10^{-5}$).
 
@@
 
-\subsection{Accuracy of estimates with full ABC}
+\subsubsection*{Joint parameter estimates with kernel-ABC}
 
-We used kernel-\gls{ABC} to estimate the parameters of the \gls{BA} model 
-on simulated trees where the true parameter values were known. Point estimates
-for each parameter are shown in Figure~\ref{fig:abcpt}. Of the four parameters,
-\gls{alpha} was the most accurately estimated, with a median [IQR] absolute error
-of 
+We used \software{netabc} to estimate the parameters of the \gls{BA} model on
+simulated trees where the true parameter values were known. Point estimates for
+each parameter are shown in \cref{fig:abcpt} for the simulations with \gls{m} =
+2. The results for the other values of \gls{m} were similar
+(\cref{fig:abcm3,fig:abcm4}). The median [IQR] absolute error of estimates of
+\gls{alpha} across all simulations was
     0.11 
@@
 significantly greater than the other values 
-    (Wilcoxon rank-sum test, $p$ = \ensuremath{6.41\times 10^{-4}}).
+    (Wilcoxon rank-sum test, $p = 6\!\times\!10^{-4}$).
 The error in the estimated value of $I$ was
@@
 Errors were significantly higher for $\alpha \geq 1$
-    (Wilcoxon rank-sum test, $p$ = \ensuremath{6.12\times 10^{-4}})
+    (Wilcoxon rank-sum test, $p = 6\!\times\!10^{-4}$)
 and for $I$ = 2000
-    ($p$ = \ensuremath{1.58\times 10^{-6}}),
-but not for any values of $m$
-    (one-way ANOVA, $p$ = 0.33).
-The $m$ parameter was estimated correctly in
+    ($p = {<}10^{-5}$),
+but not for any values of $m$ (one-way ANOVA). The $m$ parameter was estimated
+correctly in only
     37 \%
-of simulations, with an error of one in
-    40 \%
-and of two or more in 
-    22 \%
-(the only possible $m$ values were 2, 3, 4, or 5). The true values of
-$m$ and $I$ did not significantly affect the error
-    (one-way ANOVA, $p$ = 0.5 and
-                          0.68),
-but the accuracy was significantly lower for integral than non-integral
-values of \gls{alpha}
-    (Wilcoxon rank-sum test, $p$ = \ensuremath{7.2\times 10^{-3}}).
+of simulations. The true values of $m$ and $I$ did not significantly affect the
+error (one-way ANOVA), but the accuracy was significantly lower for integral
+than non-integral values of \gls{alpha}
+    (Wilcoxon rank-sum test, $p = 7\!\times\!10^{-3}$).
 Finally, the total number of nodes \gls{N} was consistently over-estimated by
 about a factor of two
-    (error \ensuremath{6.59\times 10^{3}} 
-    [\ensuremath{4.21\times 10^{3}}-\ensuremath{8.28\times 10^{3}}]).
-No other parameters influenced the accuracy of the $N$ estimates 
-    (one-way ANOVA, $p \geq$ NA).
+    (error 6588 
+    [4214 - 
+     8284]).
+No other parameters influenced the accuracy of the $N$ estimates (one-way
+ANOVA)
+
+\begin{figure}
+  \includegraphics{abc-point-estimate-m2}
+  \caption{
+    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by         
+    running \software{netabc} on simulated data. Values shown are for               
+    simulations with \gls{m} = 3. Dashed lines indicate true values. (A)            
+    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations  
+    against known values. (B) Estimates of \gls{m} and \gls{N} which were held   
+    fixed in these simulations at the values \gls{m} = 3 and \gls{N} = 5000. 
+  }
+  \label{fig:abcm2}
+\end{figure}
 
@@
 
-\subsection{Real data}
+\subsection{Application to HIV data}
 

### 2016-04-20 doc/thesis/discussion.tex
Finished draft intro!
I don't know, more writing

--- 2016-04-16 doc/thesis/discussion.tex
@@
 networks can be easily generated. We have included generators for the \gls{BA}
-model discussed here, as well as the \gls{ER} and \gls{SW} network models.
+model discussed here, as well as the \gls{ER} and \gls{WS} network models.
 Instructions for adding additional models are available in the project's online
@@
 encourage other researchers to apply and extend our method.
+
+In retrospect, there are a few algorithmic improvements which could have been
+made to \software{netabc}. Firstly, our implementation uses simple multinomial
+sampling for the resampling step, as well as for generating the final sample
+from the target distribution.  Several other sampling strategies have been
+developed~\autocite{douc2005comparison}, and it is possible that the use of a
+more sophisticated scheme might increase the algorithm's accuracy.
 

### 2016-04-20 doc/thesis/intro.tex
I don't know, more writing

@@
 and their branch lengths in a single similarity score.
+
+The kernel score of a pair of trees, denoted $K(T_1, T_2)$, is defined as a sum
+over all pairs of nodes $(n_1, n_2)$, where $n_1$ is a node in $T_1$ and $n_2$
+is a node in $T_2$. Folloting \textcite{poon2013mapping}, let $N(T)$ denote the
+set of all nodes in $T$, $\nc(n)$ be the number of children of node $n$,
+$c_n^{j}$ be the $j$th child of node $n$, and $l_n$ be the vector of branch
+lengths connecting node $n$ to its descendants. The \defn{production rule} of
+$n$ is its total number of children, and its number of leaf children. That is,
+if two nodes have the same number of children and among these, the same number
+of leaves, then they have the same production rule. Let $k_G(x, y)$ be a
+Gaussian radial basis function of the vectors $x$ and $y$,
+\[
+  k_G(x, y) = \exp\left(-\frac{1}{2\sigma} \norm{x - y}_2^2\right),
+\]
+where $\norm{\cdot}_2$ is the Euclidian norm and $\sigma$ is a variance
+parameter. The tree kernel is defined as 
+\[
+  K(T_1, T_2) = \sum_{n_1 \in N(T_1)} \sum{n_2 \in N(T_2)} \Delta (n_1, n_2),
+\]
+where
+\[
+  \Delta(n_1, n_2) =
+  \begin{cases}
+    \lambda & n_1 \text{ and } n_2 \text{ are leaves} \\
+    \lambda k_G(l_{n_1}, l_{n_2}) \displaystyle\prod_{j=1}^{\nc(n_1)} \left(1 +
+    \Delta(c_{n_1}^j, c_{n_2}^j) \right) & \begin{aligned} n_1 \text{ and } n_2 \text{ have the same} \\ \text{production rule} \end{aligned} \\
+    0 & \text{otherwise}.
+
+  \end{cases}
+\]
+Here $\lambda$ is a decay factor parameter, which penalizes large matches that
+tend to dominate the kernel score. In this work, we refer to the parameters
+$\lambda$ and $\sigma$ as \defn{meta-parameters}, to avoid confusing them with
+model parameters we are trying to estimate.
 
@@
 enough random walk is performed on the chain, the distribution of states
-visited will be a Monte Carlo approximation of $\Pr(\theta \mid y)$, from
+visited will be a Monte Carlo approximation of $f(\theta \mid y)$, from
 which we can calculate statistics of interest. Actually performing this random
@@
   \begin{algorithmic}
-    \State Draw $\theta$ according to $\Pr(\theta)$
+    \State Draw $\theta$ according to $\pi(\theta)$
     \Loop
@@
 For \gls{ABC}-\gls{MCMC} and \gls{ABC}-\gls{SMC}, the proposal distribution
-$q(\theta, \theta') \Pr(z \mid \theta')$ is replaced by
-\[
-  q_i(\theta, \theta') \prod_{k=1}^M \Pr(z_i \mid \theta').
-\]
+$q(\theta, \theta') f(z \mid \theta')$ is replaced by
+\[
+  q_i(\theta, \theta') \prod_{k=1}^M f(z_i \mid \theta').
+\]

### 2016-04-20 doc/thesis/methods.tex
I don't know, more writing

@@
 As described in \cref{sec:smc}, \software{netabc} keeps track of a population
-of particles, each of which contains particular parameter values for the model
-we are trying to fit. A small number of contact networks are generated for each
-particle, in accordance with that particle's parameters. An epidemic is
-simulated over each of these networks using Gillespie simulation, and by
-keeping track of its progress, a transmission tree is obtained. Thus, each
-particle becomes associated with several simulated transmission trees.  These
-trees are compared to the observed tree using the phylogenetic kernel.
-Particles are weighted according to the similarity of their associated
-simulated trees with the true tree, with more similar trees receiving higher
-weights. The particles are iteratively perturbed to explore the parameter
-space, and particles with simulated trees too distant from the true tree are
-periodically dropped and resampled. Once a convergence criterion is attained,
-the final set of particles is used as a Monte Carlo approximation to the target
-distribution of \gls{ABC}, which is assumed to resemble the posterior
-distribution on model parameters (see \cref{sec:abc}). A graphical schematic of
-this algorithm is given in \cref{fig:abcsmc}.
+of particles $x^{(k)}$, each of which contains particular parameter values
+$\theta^{(k)}$ for the model we are trying to fit. A small number of contact
+networks $z^{(k)}$ are generated for each particle, in accordance with that
+particle's parameters. An epidemic is simulated over each of these networks
+using Gillespie simulation, and by keeping track of its progress, a
+transmission tree is obtained. Thus, each particle becomes associated with
+several simulated transmission trees. These trees are compared to the observed
+tree using the phylogenetic kernel. Particles are weighted according to the
+similarity of their associated simulated trees with the true tree, with more
+similar trees receiving higher weights. The particles are iteratively perturbed
+to explore the parameter space, and particles with simulated trees too distant
+from the true tree are periodically dropped and resampled. Once a convergence
+criterion is attained, the final set of particles is used as a Monte Carlo
+approximation to the target distribution of \gls{ABC}, which is assumed to
+resemble the posterior distribution on model parameters (see \cref{sec:abc}). A
+graphical schematic of this algorithm is given in \cref{fig:abcsmc}.
 
@@
 The kernel was implemented using the fast algorithm developed by
-\textcite{moschitti2006making}. First, the number of leaf children of each
-node, also known as its \defn{production rule}, is recorded. The nodes of both
-trees are ordered by production rule, and a list of pairs of nodes sharing the
-same production rule is created. These are the nodes for which the value of the
-tree kernel must be computed - all other pairs have a value of zero. The pairs
-to be compared are then re-ordered so that the child nodes are always evaluated
-before their parents. Due to its recursive definition, ordering the pairs in
-this way allows the tree kernel to be computed by dynamic programming. The
-complexity of this implementation is $O(|T_1||T_2|)$, where $|T|$ counts the
-number of nodes in the tree $T$.
+\textcite{moschitti2006making}. First, the production rule of each node, which
+is the total number of children and the number of leaf children, is recorded.
+The nodes of both trees are ordered by production rule, and a list of pairs of
+nodes sharing the same production rule is created. These are the nodes for
+which the value of the tree kernel must be computed - all other pairs have a
+value of zero. The pairs to be compared are then re-ordered so that the child
+nodes are always evaluated before their parents. Due to its recursive
+definition, ordering the pairs in this way allows the tree kernel to be
+computed by dynamic programming. The complexity of this implementation is
+$O(|T_1||T_2|)$, where $|T|$ counts the number of nodes in the tree $T$.
+
+The tree kernel cannot be used directly as a distance measure for \gls{ABC},
+since it is maximized, not minimized, when the two trees being compared are the
+same. Therefore, we defined the distance between two trees as
+\[
+  \rho(T_1, T_2) = 1 - \frac{K(T_1, T_2)}{\sqrt{K(T_1, T_1) K(T_2, T_2)}},
+\]
+which is a number between 0 and 1 minimized when $T_1 = T_2$. This is similar
+to the normalization used by \textcite{collins2002new, poon2013mapping}.
 
@@
 
-I implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
+We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
 \textcite{del2012adaptive}. This algorithm is similar to the reference
@@
 \[
-  q_i(\theta, \theta') \prod_{k=1}^M \Pr(X_i^{(k)'} \mid \theta'),
+  q_i(\theta, \theta') \prod_{k=1}^M \Pr(z_i^{(k)'} \mid \theta'),
 \]
-where $\theta$ is the vector of model parameters and $X_k$ are $M$ datasets
-simulated according to $\theta$. In our implementation, $q$ is a Gaussian
-proposal with variance equal to twice the empirical variance of the particles,
+where $\theta$ is the vector of model parameters and $z_k$ are $M$ datasets
+simulated according to $\theta'$. In our implementation, $q$ is either a
+Gaussian proposal for continuous parameters, or a Poisson proposal for discrete
+parameters. For the Poisson proposals, the number of steps to move the particle
+is drawn from a Poisson distribution, and the direction in which to move the
+particle is chosen uniformly at random. For both proposals, the variance was
+set equal to twice the empirical variance of the particles,
 following~\autocite{beaumont2009adaptive, del2012adaptive}. The backwards
@@
 \[
-  L_{i-1}(z', z) = \frac{\pi_n(z)K(z, z')}{\pi_n(z')}.
+  L_{i-1}(x', x) = \frac{\pi_n(x)K(x, x')}{\pi_n(x')}.
 \]
-When substituted into \cref{eq:smcwt}, the forward kernels $K(z, z')$ and
-densities $\pi_n(z')$ cancel out, and we are left with the weight update
+When substituted into \cref{eq:smcwt}, the forward kernels $K(x, x')$ and
+densities $\pi_n(x') = \pi_{\varepsilon_n}(x')$ cancel out, and we are left
+with the weight update 
 \begin{align*}
-  w(z') 
-    &\propto w(z) \frac{\pi_n(z \mid y)}{\pi_{n-1}(z \mid y)} \\
-    &= w(z) \frac{\pi(z) \pi_n(y \mid z)}{\pi(z) \pi_{n-1}(y \mid z)} \\
-    &= w(z) \frac{\sum_{k=i}^M \I_{A_{\varepsilon_n, y}}(x_k)}
-            {\sum_{k=i}^M \I_{A_{\varepsilon_{n-1}, y}}(x_k)}.
+  w_i(x) 
+    &\propto w_{i-1}(x) \frac{\pi_n(x \mid y)}{\pi_{i-1}(x \mid y)} \\
+    &= w_{i-1}(x) \frac{\pi(x) \pi_i(y \mid x)}{\pi(x) \pi_{i-1}(y \mid x)} \\
+    &= w_{i-1}(x) \frac{\sum_{k=i}^M \I_{A_{\varepsilon_i, y}}(z_k)}
+            {\sum_{k=i}^M \I_{A_{\varepsilon_{i-1}, y}}(z_k)}.
 \end{align*}
-The algorithm may be stopped when one of two termination conditions is reached.
-The user may specify a final tolerance $\varepsilon$, or a final acceptance
-rate of the \gls{MCMC} kernel. The latter stops the algorithm when the
-particles are not moving around very much, implying little change in the
-estimated target.
+In other words, when the distance threshold $\varepsilon_{i-1}$ is contracted
+to $\varepsilon_i$, the particles' weights are multiplied by the proportion of
+simulated datasets which are still inside the new threshold. The algorithm may
+be stopped when one of two termination conditions is reached. The user may
+specify a final tolerance $\varepsilon$, or a final acceptance rate of the
+\gls{MCMC} kernel. The latter condition stops the algorithm when the particles
+are not moving around very much, implying little change in the estimated
+target.
 
@@
 
-\subsubsection*{Characterization of power-law exponent in Barab\'asi-Albert networks}
-
-Most studies of social network or transmission network
-parameters~\autocite[e.g.][]{liljeros2001web, jones2003assessment,
-schneeberger2004scale, brown2011transmission} report the coefficient
-\gls{gamma} of the power law degree distribution. To make our results
-comparable to previous work, we used simulated networks to investigate the
-relationship between the \gls{BA} model parameters and \gls{gamma}. A network
-was simulated for each combination of parameters listed in
-\cref{tab:gammaexpt}. A power law distribution was fitted to the degree
-distribution of each simulated network using the \software{fit\_power\_law}
-function in \software{igraph} with the `R.mle' implementation. We fitted a
-\gls{GLM} with Gamma-distributed errors and a log link function to the observed
-distribution of \gls{gamma} values, with \gls{alpha}, \gls{m}, \gls{N}, and all
-possible interaction terms as predictors.
+%\subsubsection*{Characterization of power-law exponent in Barab\'asi-Albert networks}
+%
+%Most studies of social network or transmission network
+%parameters~\autocite[e.g.][]{liljeros2001web, jones2003assessment,
+%schneeberger2004scale, brown2011transmission} report the coefficient
+%\gls{gamma} of the power law degree distribution. To make our results
+%comparable to previous work, we used simulated networks to investigate the
+%relationship between the \gls{BA} model parameters and \gls{gamma}. A network
+%was simulated for each combination of parameters listed in
+%\cref{tab:gammaexpt}. A power law distribution was fitted to the degree
+%distribution of each simulated network using the \software{fit\_power\_law}
+%function in \software{igraph} with the `R.mle' implementation. We fitted a
+%\gls{GLM} with Gamma-distributed errors and a log link function to the observed
+%distribution of \gls{gamma} values, with \gls{alpha}, \gls{m}, \gls{N}, and all
+%possible interaction terms as predictors.
 
@@
 aligned pairwise to the HXB2 reference sequence (GenBank accession number
-HIVHXB2CG) and the hypervariable regions were clipped out with
+K03455) and the hypervariable regions were clipped out with
 \software{BioPython} version 1.66+~\autocite{cock2009biopython}. Sequences were

### 2016-04-20 doc/thesis/preface.tex
Wrote discussion mostly
I don't know, more writing

@@
-This work was conducted at the BC Centre for Excellence in HIV/AIDS under the
-supervision of Dr. Art Poon. I wrote the code, performed the experiments, and
-wrote the thesis. Dr. Richard Liang assisted with the development of the
-Gillespie simulation algorithm.
+The initial idea to use kernel approximate Bayesian computation to infer
+contact network model parameters was Dr. Poon's. The tree kernel was originally
+developed by Dr. Poon, but the version used here was implemented by me to
+improve computational efficiency. The idea to apply sequential Monte Carlo is
+credited to Dr. Alexandre Bouchard-C\^ot\'e. Dr. Sarah Otto suggested the
+experiments involving a network with a heterogeneous $\alpha$ parameter and
+peer-driven sampling. Dr. Richard H. Liang provided guidance in the development
+of the Gillespie simulation algorithm. The \software{netabc} program, and all
+supplementary analysis programs, were written by me.
 
-A version of chapters 2 and 3 has been submitted for publication in Molecular
-Biology and Evolution, with the title ``Phylogenetic estimation of contact
-network parameters with kernel-ABC''. A presentation with the same title was
-given at the 23rd HIV Dynamics and Evolution meeting on April 25, in Woods
-Hole, Massachussets, USA.
+A version of chapter 2 has been submitted for publication in Virus Evolution
+with the title ``Reconstructing network parameters from viral phylogenies.'' A
+presentation entitled ``Phylodynamic inference of contact network parameters
+with kernel-ABC'' was given based on chapter 2 at the 23rd HIV Dynamics and
+Evolution meeting on April 25, in Woods Hole, Massachussets, USA.
+
+Throughout this work, the pronouns \textit{we} and \textit{our} refer to
+Rosemary M. McCloskey unless otherwise stated.

### 2016-04-20 doc/thesis/results.Rnw
I don't know, more writing

@@
     (Spearman's $\rho$ = \Sexpr{value.cor["N", round(estimate, 2)]},
-    $p \Sexpr{value.cor["N", pp(p.value)]}$).
+    $p\ \Sexpr{value.cor["N", pp(p.value)]}$).
 

### 2016-04-20 doc/thesis/intro.tex
Few more references for phylodynamics

@@
 phylogeny. Consequently, the topology of the viral phylogeny is sometimes used
-as a proxy for the topology of the transmission tree. Modern likelihood-based
-methods of phylogenetic reconstruction~\autocite[\eg][]{price2010fasttree,
+as a proxy for the topology of the transmission
+tree~\autocite{hall2015epidemic}. Modern likelihood-based methods of
+phylogenetic reconstruction~\autocite[\eg][]{price2010fasttree,
 stamatakis2014raxml} produce unrooted trees whose branch lengths measure
@@
 correspondence between the topologies of the viral phylogeny and transmission
-tree is not necessarily exact~\autocite{ypma2013relating}. Due to intra-host
-diversity, the viral strain which is transmitted may have split from another
-lineage within the donor long before the transmission event occurred. Hence,
-the branching point in the viral phylogeny may be much earlier than that in the
-transmission tree. Another possibility is that one host transmitted to two or
-more recipients, but the lineages they each received originated within the
-donor host in a different order than that in which the transmissions occurred.
-In this case, the topology of the transmission tree and the viral phylogeny
-will be mismatched. In practice, this discordance has not proven an
-insurmountable problem: for example, \textcite{leitner1996accurate} were able
-to accurately recover a known transmission tree using a viral phylogeny. The
+tree is not necessarily exact~\autocite{ypma2013relating, romero2014timing}.
+Due to intra-host diversity, the viral strain which is transmitted may have
+split from another lineage within the donor long before the transmission event
+occurred. Hence, the branching point in the viral phylogeny may be much earlier
+than that in the transmission tree. Another possibility is that one host
+transmitted to two or more recipients, but the lineages they each received
+originated within the donor host in a different order than that in which the
+transmissions occurred.  In this case, the topology of the transmission tree
+and the viral phylogeny will be mismatched. In practice, this discordance has
+not proven an insurmountable problem: for example,
+\textcite{leitner1996accurate, paraskevis2004phylogenetic} were able to
+accurately recover a known transmission tree using a viral phylogeny. The
 problem of accurately estimating transmission trees is an ongoing area of

### 2016-04-20 doc/big-retreat/poster.tex

@@
-\documentclass[14pt, a0paper, landscape]{tikzposter}
+\documentclass[20pt, a0paper, landscape, colspace=10mm, blockverticalspace=10mm]{tikzposter}
 \usepackage[sfmath]{kpfonts}
@@
 \usepackage{adjustbox}
+\usepackage{qrcode}
+\usepackage{multicol}
 \graphicspath{{../figures/}}
+\newcommand{\tablepath}{../tables}
+
 \usetheme{Board}
+\usetikzlibrary{calc}
 
@@
 \setlength{\bibhang}{0pt}
+\tikzposterlatexaffectionproofoff
+\settitle{ \vbox{ \centering \color{titlefgcolor}
+{\fontsize{2.5cm}{2cm}\selectfont \@title}
+\\ --- \huge \@author~--- \\ \@institute } }
 
@@
 
-\maketitle
+\node at (current page.west) 
+  [anchor=north] {
+  \includegraphics[width=\paperwidth, trim=0 0 0 1in]{logos/cfe_banner}
+};
+
+\maketitle[width=1.2\linewidth]
+
+% hacky things
+\node [text width=6in] at ($ (current page.west) + (-0.42\paperwidth, -1.5in) $)
+  { \LARGE B.I.G. Research Day \\ March 11, 2016 \\ Vancouver, Canada};
+\node [text width=6in, align=right] at ($ (current page.west) + (0.42\paperwidth, -1.5in) $)
+  { \LARGE rmccloskey@cfenet.ubc.ca \\ (604) 682-6344 ext. 63266 };
 
@@
     \centerline{\includegraphics[width=0.2\textwidth]{contactnet}}
+      Network structure can change the speed and pattern of epidemic spread,
+      potentially informing public health interventions.
   }
-  \block{What can we tell about contact networks from viral sequence data?}{
-  }
+  \block{}{\vspace{1cm}\color{blocktitlefgcolor}\bfseries\LARGE{
+    \centering
+    \centerline{What can we tell about contact}
+
+    \centerline{networks from viral sequences?}
+  }}
 
   \block{Barab\'asi-Albert preferential attachment networks}{
-
-    \begin{itemize}
-      \item Nodes are added one-at-a-time until there are $N$ total.
-      \item Each time a node is added, $m$ new edges are connected to it.
-      \item Each new edge is attached to an existing node of degree $d$ with
-        probability proportional to $d^\alpha$.
-    \end{itemize}
+    Nodes are added one-at-a-time until there are $N$ total. Each time a node
+    is added, $m$ new edges are connected to it. Each new edge is attached to
+    an existing node of degree $d$ with probability proportional to
+    $d^\alpha$.
 
     \vspace{1cm}
-    \includegraphics[width=0.2\textwidth, trim=0 22in 0 0, clip]{alpha-bounds}
+    \centerline{\includegraphics[width=0.2\textwidth, trim=0 22in 0 0, clip]{alpha-bounds}}
+
     \begin{center}
@@
         \hline
-        $\alpha$ & preferential attachment power: nodes of degree $d$ attract new edges with probability $\propto d^\alpha$ \\
+        $\alpha$ & preferential attachment power \\
         $m$ & number of new edges added for each new vertex \\
-        $I$ & number of infected nodes when epidemic simulation is stopped \\
+        $I$ & number of infected nodes when epidemic is stopped \\
         $N$ & total number nodes in the network \\
@@
     \end{center}
+    \node at (1, 10) {low $\alpha$};
+    \node at (14, 10) {high $\alpha$};
   }
 
-  \block{Approximate Bayesian computation}{
+  \block{}{
+    \vspace{1.5cm}
+    \centerline{\qrcode{https://github.com/rmcclosk/netabc}
+    \LARGE{github.com/rmcclosk/netabc}}
   }
 
-  \column{0.25}
-  \block{Phylogenetic kernel lets us compare tree shapes}{
+  \column{0.225}
+  \block{\textit{netabc}: tree kernel + approximate Bayesian computation +
+         sequential Monte Carlo}{
+    \centerline{\includegraphics[width=0.18\textwidth]{abc-idea}}
+    \nocite{del2012adaptive}
     \nocite{poon2013mapping}
-  }
 
-  \block{ABC with sequential Monte Carlo}{
-    \centerline{\includegraphics[width=0.2\textwidth]{abc-smc}}
-    \nocite{del2012adaptive}
+    \vspace{1cm}
+      ``Particles'' are sampled from prior distributions on model parameters.
+      Simulated trees are generated for each particle and compared to the
+      observed tree with a phylogenetic kernel~\autocite{poon2013mapping}.
+      Particles whose simulated trees are distant from the observed tree are
+      removed. Surviving particles are perturbed, and new simulated datasets
+      are generated. Iteration leads to an approximation of the posterior
+      distribution on the model parameters.
   }
@@
   \block{Varying $\alpha$ and $I$ affects tree shape, but not $m$ and $N$}{
-    \centerline{\includegraphics[width=0.2\textwidth]{kernel-kpca}}
+    \centerline{\includegraphics[width=0.18\textwidth]{kernel-kpca}}
+
+      We simulated 100 trees of size 500 under each parameter value. Parameters
+      not being tested were fixed at $\alpha$ = 1.0, $m$ = 2, $I$ = 1000, $N$ =
+      5000. A $300 \times 300$ kernel matrix of all pairwise similarities was 
+      projected onto its first two principal components.
   }
 
-  \column{0.25}
+  \column{0.225}
   \block{Simulations}{
-    \begin{minipage}[t]{0.5\linewidth}
+    \centering
     \begin{tabular}{ccc}
@@
     \end{tabular}
-    \end{minipage}
-    \hspace{0.05\linewidth}
-    \begin{adjustbox}{valign=t}
-      \begin{minipage}[t]{0.45\linewidth}
+
+    \vspace{0.5cm}
+    \captionof{table}{
     Three networks/trees were simulated under each parameter combination.
     Kernel-ABC was used to re-estimate the parameters.
-      \end{minipage}
-    \end{adjustbox}
     }
-  \block{Accurate estimates for $\alpha$, $I$, but not $m$, $N$}{
-    \centerline{\includegraphics[width=0.2\textwidth]{abc-point-estimate}}
+  }
+  \block{Accurate estimates for $\alpha$, $I$; not $m$, $N$}{
+    \centerline{\includegraphics[width=0.16\textwidth]{abc-point-estimate}}
+    Dotted lines are true values. $\alpha$ and $I$, which had the most impact
+    on tree shape, were possible to estimate accurately. Posterior
+    distributions of $m$ and $N$ heavily resemble the priors, indicating
+    little information about these parameters in the trees.
   }
   \block{Narrow dispersion of $\alpha$, $I$ compared to $m$, $N$}{
-    \centerline{\includegraphics[width=0.2\textwidth]{abc-posterior-example}}
+    \centerline{\includegraphics[width=0.16\textwidth]{abc-posterior-example}}
   }
 
-  \column{0.25}
+  \column{0.3}
 
@@
     \centering
-    \begin{tabular}{ccccc}
-      Reference & Risk group & Sequences ($n$) & Location & HIV gene \\
-      \hline
-      \textcite{niculescu2015recent} & IDU & 136 & Romania & \textit{pol} \\
-      \textcite{li2015hiv} & MSM & 280$^1$ & Shanghai, China & \textit{pol} \\
-      \textcite{shiino2014phylodynamic} & mixed & 211$^2$ & Japan & \textit{pol} \\
-      \textcite{cuevas2009hiv} & mixed & 287 & Basque Country, Spain & \textit{pol} \\
-      \textcite{novitsky2013phylogenetic} & \multirow{2}{*}{mixed} & 
-      \multirow{2}{*}{218$^3$} &
-      \multirow{2}{*}{Mochudi, Botswana} & 
-      \multirow{2}{*}{\textit{env}} \\
-      \textcite{novitsky2014impact} \\
-      N/A & IDU & 399$^4$ & BC, Canada & \textit{pol} \\
-      \hline
-    \end{tabular}
+    \input{\tablepath/realdata}
     \captionof{table}{
       Characteristics of HIV datasets analysed with kernel-ABC. Abbreviations:
-      MSM, men who have sex with men; IDU, injection drug users. $^1$subsampled
-      from 1261 $^2$subsampled from 3531 $^3$subsampled from 1299 $^4$subsampled 
-      from 7923.}
+      MSM, men who have sex with men; IDU, injection drug users.
+    }
   }
 
-  \block{Transmission trees were estimated from viral sequence data}{
+  \block{Transmission trees were estimated from viral sequences}{
     \centerline{\includegraphics[width=0.2\textwidth]{pipeline}}
@@
   \block{Parameter heterogeneity among real datasets}{
-    \centerline{\includegraphics[width=0.2\textwidth]{realdata-hpd-bc}}
+    \centerline{\includegraphics[width=0.2\textwidth]{realdata-hpd}}
   }
@@
 \begin{columns}
-  \column{0.6}
+  \column{0.65}
   \block{Summary}{
-    \Large
+    \setlength{\columnsep}{1cm}
+    \begin{multicols}{3}
       We developed a kernel-ABC method for inference of structural contact
-    network parameters from viral sequence data.
+      network parameters from viral sequence data. The method is applicable to
+      any network model from which networks can be simulated. We demonstrated
+      that two of the parameters of the Barab\'asi-Albert preferential
+      attachment model can be estimated with kernel-ABC. When applied to real
+      data, we observed heterogeneity among model parameters, although
+      estimated values of $\alpha$ were consistent with those observed in real
+      social networks.
 
-    The Barab\'asi-Albert model is one of the simplest preferential attachment
-    models which makes a number of unrealistic modelling assumptions. In
-    particular, the average degree of the network is equal to $2m$, which is
-    constrained to a multiple of 2. Moreover, we assume that the network is
-    static and connected, and that nodes are never removed and are homogeneous
-    in their behaviour. These assumptions mean that the model is very likely
-    misspecified for the real data in question. However, method can be applied
-    to any network model from which simulations can be drawn, leaving open the
-    possibility to investigate more realistic models in the future.
+      Due to the simple nature of the Barab\'asi-Albert model, it is likely
+      misspecified for the real data we analyzed, and therefore caution must be
+      exercised in interpreting the estimated values. However, the possibility
+      to investigate more realistic models in the future makes our approach a
+      useful tool for phylodynamic research on contact networks.
+    \end{multicols}
+  }
+  \block{}{
+    \begin{minipage}{0.55\linewidth}
+      \vspace{1cm}
+      \includegraphics[width=3in]{logos/btp}
+      \includegraphics[width=3in]{logos/cfe}
+      \includegraphics[width=1.5in]{logos/cihr}
+      \includegraphics[width=3in]{logos/msfhr}
+      \includegraphics[width=2in]{logos/phc}
+      \includegraphics[width=0.9in]{logos/ubc}
+      \vspace{0.5cm}
+    \end{minipage}
+    \begin{adjustbox}{valign=t}
+      \begin{minipage}[t]{0.45\linewidth}
+        \small
+        This work was supported by a grant from the Canadian Institutes for
+        Health Research (CIHR HOP-111406 to AFYP) and a scholarship from the 
+        Canadian Institutes for Health Research  Strategic Training Program in
+        Bioinformatics (to RMM). AFYP is the recipient of a Scholar Award from
+        the MSFHR/St. Paul’s Hospital Foundation - Providence Health Care
+        Research Institute Career Investigator program, and a CIHR New
+        Investigator Award.
+      \end{minipage}
+    \end{adjustbox}
   }
 
-  \column{0.4}
+  \column{0.35}
   \block{References} {
+    \renewcommand*{\bibfont}{\footnotesize}
     \printbibliography[heading=none]
+    \small
+    \includegraphics[height=1cm]{logos/westgrid} Analyses were run in the
+    Westgrid high performance computing environment.
   }
-  \block{Funding and acknowledgements} {
-    some logos
-  }
+
 \end{columns}

### 2016-04-20 doc/paper/mbe_submission.tex

\documentclass[12pt]{article}
\usepackage{fullpage}
\usepackage{garamondx}
\usepackage{parskip}
\usepackage{setspace}

% http://tex.stackexchange.com/questions/44618/dynamically-count-and-return-number-of-words-in-a-section
\newcommand\wordcount{\immediate\write18{texcount -sub=section \jobname.tex  | grep "Section" | sed -e 's/+.*//' | sed -n \thesection p > '/tmp/count.txt'}\ignorespaces\input{/tmp/count.txt}}

\title{Phylodynamic inference of contact network parameters through approximate Bayesian computation}
\author{Rosemary M. McCloskey, Richard H. Liang, and Art F.Y. Poon}

\doublespacing
\pagestyle{empty}
\begin{document}
\maketitle

\section{Abstract}

(\wordcount / 250 words)

Models of the spread of disease in a population often make the simplifying
assumption that the population is homogeneously mixed, or is divided into
homogeneously mixed compartments. However, human populations have complex
structures formed by social contacts, which can have a significant influence on
the rate of epidemic spread. Contact network models capture this structure by
explicitly representing each contact which could possibly lead to a
transmission. We developed a method based on kernel approximate Bayesian
computation (kernel-ABC) for estimating structural parameters of the contact
network underlying an observed viral phylogeny. The method combines adaptive
sequential Monte Carlo for ABC, Gillespie simulation for propagating epidemics
though networks, and a kernel-based tree similarity score. We used the method
to fit the Barab\'{a}si-Albert network model to simulated transmission trees,
and also applied it to viral phylogenies estimated from five published HIV
sequence datasets. On simulated data, we found that the preferential attachment
power and the number of infected nodes in the network can often be accurately
estimated. On the other hand, the mean degree of the network, as well as the
total number of nodes, were not estimable with kernel-ABC. We observed
substantial heterogeneity in the parameter estimates on real datasets, with
point estimates for the preferential attachment power ranging from 0.2 to 1.1.
These results underscore the importance of considering contact structures when
performing phylodynamic inference. Our method offers the potential to
quantitatively investigate the network structure underlying viral epidemics,
using models appropriate to the context being studied.

\section{Keywords}

Phylogenetics, phylodynamics, approximate Bayesian computation, contact network,
transmission tree, human immunodeficiency virus.

\section{Cover Letter}

Dear editors of \textit{Molecular Biology and Evolution},

On behalf of my coauthors and myself, I am submitting a manuscript entitled
``\textit{Phylodynamic inference of contact network parameters through
approximate Bayesian computation}'' for your consideration.

In this work, we have developed a method to fit contact network models to viral
phylogenetic data. These models capture the contact heterogeneity and
population structure often observed in real host communities. It has previously
been demonstrated analytically and through simulation studies that network
structure can affect the speed and pattern of epidemic spread. Consequently,
the often-made assumption of a homogeneously mixed population may introduce
errors into epidemiological predictions. Furthermore, the network itself may be
of interest from a public health perspective, such as for investigating the
presence of highly-connected ``superspreader'' nodes.

Despite their potential usefulness, most network parameters are extraordinarily
difficult to estimate with existing methods. Direct epidemiological
investigation of contact networks requires extensive contact tracing and may be
hampered by misreporting. Viral genetic data are much easier to collect, but as
far as we are aware, no other phylodynamic methods have been developed to make
use of this data to fit network models. Therefore, our work opens new avenues
for phylodynamic investigation, and we believe it will be of broad interest to
the molecular evolution and epidemiology community.

All authors fulfill the authorship criteria, and have agreed to this
submission. \textit{Molecular Biology and Evolution} is our first choice for
this manuscript, which has not been submitted to any other journal.

Thank you for your time,

Rosemary M. McCloskey

\section{Reviewers}

\begin{enumerate}
  \item David Welch
  \item Gabriel Leventhal
  \item Erik Volz
  \item Tanja Stadler
  \item Oliver Ratmann
  \item Andrew Rambaut
\end{enumerate}

\section{Editors}

\begin{enumerate}
  \item Thomas Leitner
  \item Beth Shapiro
  \item Kieth Crandall
  \item John Novembre
\end{enumerate}

\section{Funding}

\begin{enumerate}
  \item Bill \& Melinda Gates Foundation (OPP1110049)
  \item Canadian Institutes for Health Research (HOP-111406)
\end{enumerate}

\section{Significant discoveries}

(\wordcount / 150 words)

Through a simulation study, we have found that certain parameters of the
contact network underlying a viral epidemic can be quantitatively estimated
from phylogenetic data. In particular, we investigated a preferential
attachment network model and were able to accurately infer the attachment power
and the number of infected nodes in the network, in most simulation scenarios
we considered. On the other hand, the average degree and total number of nodes
in the network had an insufficient effect on tree shape and could not be
quantitatively estimated. When applied five real-world HIV epidemics, we found
substantial heterogeneity of these parameters, which are known to have a
significant impact on the speed and pattern of epidemic spread.

\section{Significant methodological or theoretical advances}

(\wordcount / 150 words)

We have developed a phylodynamic method to quantitatively estimate the
parameters of contact network models from phylogenetic data. Our method is
based on approximate Bayesian computation, and combines adaptive sequential
Monte Carlo, Gillespie simulation, and a kernel-based tree similarity score.
Due to the use of the tree kernel, the method falls under the heading of
"kernel-ABC", which has been previously demonstrated to be a robust and
effective means of estimating epidemiological quantities from observed
phylogenies. To our knowledge, no other methods have been developed to estimate
network parameters from this type of data. Therefore, our method expands the
domain of phylodynamic inference to include network structure, offering
molecular epidemiologists a new perspective into ongoing epidemics.

\section{Significant new tools or resources}

(\wordcount / 150 words)

We provide a computer program "netabc" implementing our kernel-ABC method. The
C source code for the tool is freely available under a permissive license at
github.com/rmcclosk/netabc. This allows users to both apply the method as-is to
their data, and to incorporate new network models tailored to particular
research questions. The program is multithreaded, facilitating faster analysis
when more computing power is available. In addition, we provide two
supplemental programs to assist in phylodynamic analyses involving contact
networks: "nettree", which simulates a transmission tree over a contact network
by Gillespie simulation, and "treekernel", which quickly computes the kernel
similarity score between two trees. These tools will enable researchers to
quickly test hypotheses about the effect of network structure on transmission
tree shape.

\section{Broad impact}

(\wordcount / 150 words)

Phylodynamics and molecular epidemiology are relatively new areas of study, yet
they have undergone rapid development and have already seen public health
applications for real epidemics. Recent advances in sequencing technology have
allowed the collection of viral genetic data on a population scale in diverse
geographic and demographic settings. These data, together with phylodynamic
methods, provide an opportunity to gain insight into the real-world dynamics of
disease spread. Our work expands the domain of phylodynamic inference to
include structural parameters of the contact network underlying an epidemic.
Because our method is broadly applicable to any model from which networks can
be sampled, it will enable researchers to quantitatively investigate population
structure and contact heterogeneity in a variety of contexts.

\end{document}

### 2016-04-20 doc/paper/paper.Rnw

@@
-\documentclass[12pt]{article}
-\usepackage[osf]{garamondx}
-\usepackage[garamondx,cmbraces]{newtxmath}
-\usepackage{fullpage}
-\usepackage{setspace}
-\usepackage[backend=biber, style=authoryear, uniquelist=minyear, uniquename=false]{biblatex}
-\usepackage[acronym,nonumberlist]{glossaries}
-\usepackage{multirow}
-\usepackage{cleveref}
-\usepackage{array}
-\usepackage{float}
+\documentclass[nogrid]{MBE}
+\usepackage{url}
 
-\newfloat{data}{hb}{lod}
-\floatname{data}{Data}
-\crefname{data}{data}{data}
-
-\newcommand{\defn}[1]{\textit{#1}}
-\newcommand{\software}[1]{\textit{#1}}
-
-\newcommand{\set}[1]{\left\lbrace#1\right\rbrace}
-\newcommand{\sett}[1]{\{#1\}}
-\newcommand{\ie}{\textit{i.e.}}
-\renewcommand{\emptyset}{\varnothing}
-\renewcommand{\vec}[1]{\mathbf{#1}}
-\renewcommand{\star}[1]{#1^*}
-\renewcommand{\d}{\mathrm{d}\,}
-
-\DeclareMathOperator*{\argmin}{arg\,min}
-\DeclareMathOperator*{\argmax}{arg\,max}
-
-\DeclareMathOperator{\Exponential}{Exponential}
-\DeclareMathOperator{\Uniform}{Uniform}
-\DeclareMathOperator{\E}{E}
-\DeclareMathOperator{\Var}{Var}
-
-\DeclareMathOperator{\inc}{in}
-\DeclareMathOperator{\out}{out}
-
-\newcommand{\N}{\mathcal{N}}
-\renewcommand{\L}{\mathcal{L}}
-
-\newacronym{BA}{BA}{Barab\'asi-Albert}
-\newacronym{SI}{SI}{susceptible-infected}
-\newacronym{kSVM}{kSVM}{kernel support vector machine}
-\newacronym{nLTT}{nLTT}{normalized lineages-through-time}
-\newacronym{SMC}{SMC}{sequential Monte-Carlo}
-\newacronym{ABC}{ABC}{approximate Bayesian computation}
-\newacronym{MSM}{MSM}{men who have sex with men}
-\newacronym{HPD}{HPD}{highest posterior density}
-\newacronym{GTR}{GTR}{general time-reversible}
-\newacronym{MCMC}{MCMC}{Markov chain Monte Carlo}
-\newacronym{MAP}{MAP}{maximum \textit{a posteriori}}
-\newacronym{IDU}{IDU}{injection drug user}
-\newacronym{HIV}{HIV}{human immuno-deficiency virus}
-
-\graphicspath{{../figures/}}
-\newcommand{\tablepath}{../tables}
-\def\figwidth{0.8\textwidth}
-
-\title{Phylodynamic inference of contact network parameters through approximate
-Bayesian computation}
-\author{Rosemary M. McCloskey \and Richard H. Liang \and Art F.Y. Poon}
-
-\addbibresource{papers.bib}
-
-\frenchspacing
-\onehalfspacing
+% Mol Biol Evol stuff
+\jshort{mst}
+\volname{}
+\jvolume{0}
+\jvol{}
+\jissue{0}
+\pubyear{2016}
+\mstype{Article}
+\artid{???}
+\access{???}
 
@@
 
+\title[Contact networks and ABC]{Phylodynamic inference of contact network
+parameters through approximate Bayesian computation}
+\author[McCloskey, Liang, and Poon]{Rosemary M. McCloskey$^{*,1}$, 
+    Richard H. Liang$^1$, and Art F.Y. Poon$^{1,2}$}
+\address{$^1$BC Centre for Excellence in HIV/AIDS, 608-1081 Burrard Street, Vancouver, Canada \\
+$^2$Department of Medicine, University of British Columbia, Vancouver, Canada}
+\history{Submitted 5 April 2016}
+\coresp{E-mail: rmccloskey@cfenet.ubc.ca.}
+\editor{}
+\abstract{\input{abstract}}
+\keyword{Phylogenetics, phylodynamics, approximate Bayesian computation,
+contact network, transmission tree, human immunodeficiency virus.}
+
 <<setup, include=FALSE>>=
     library(netabc)
+    library(Hmisc)
 @
@@
 mode of transmission of the disease. For airborne pathogens, physical proximity
-may be sufficient, while for sexually transmitted diseases, sexual or in some
-cases blood-to-blood contact is required. The population together with the set
-of links between individuals along which transmission can occur is called the
-contact network~\autocite{klovdahl1985social, morris1993epidemiology}. The
-structure of the contact network underlying an epidemic can profoundly impact
-the speed and pattern of the epidemic's expansion. Network structure can
-influence the prevalence trajectory~\autocite{o2010contact} and epidemic
-threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates
-of quantities such as effective population
-size~\autocite{goodreau2006assessing}. From a public health perspective,
-contact networks have been explored as tools for curtailing epidemic spread, by
-way of interventions targeted to well-connected
-nodes~\autocite{wang2015targeting}. True contact networks are a challenging
-type of data to collect, requiring extensive epidemiological
-investigation~\autocite{welch2011statistical}.
+may be sufficient, while for sexually transmitted diseases, \newpage\noindent %hack
+sexual or in some cases blood-to-blood contact is required. The population
+together with the set of links between individuals along which transmission can
+occur is called the contact network~\citep{klovdahl1985social,
+morris1993epidemiology}. The structure of the contact network underlying an
+epidemic can profoundly impact the speed and pattern of the epidemic's
+expansion. Network structure can influence the prevalence
+trajectory~\citep{o2011contact} and epidemic
+threshold~\citep{barthelemy2005dynamical}, in turn affecting the estimates of
+quantities such as effective population size~\citep{goodreau2006assessing}.
+From a public health perspective, contact networks have been explored as tools
+for curtailing epidemic spread, by way of interventions targeted to
+well-connected nodes~\citep{wang2015targeting}. True contact networks are a
+challenging type of data to collect, requiring extensive epidemiological
+investigation~\citep{welch2011statistical}.
 
@@
 of RNA viruses, epidemiological processes impact the course of viral evolution,
-thereby shaping the intra-host viral
-phylogeny~\autocite{drummond2003measurably}. The term ``phylodynamics'' was
+thereby shaping the inter-host viral
+phylogeny~\citep{drummond2003measurably}. The term ``phylodynamics'' was
 coined to describe this interaction, as well as the growing family of inference
 methods to estimate epidemiological parameters from viral
-phylogenies~\autocite{grenfell2004unifying}. These methods have revealed
+phylogenies~\citep{grenfell2004unifying}. These methods have revealed
 diverse properties of local viral outbreaks, from basic reproductive
-number~\autocite{stadler2011estimating}, to the degree of
-clustering~\autocite{hughes2009molecular}, to the elevated transmission risk
-during acute infection~\autocite{volz2012simple}. On the other hand, although
+number~\citep{stadler2012estimating}, to the degree of
+clustering~\citep{hughes2009molecular}, to the elevated transmission risk
+during acute infection~\citep{volz2012simple}. On the other hand, although
 sophisticated methods have been developed for fitting complex population
-genetic models to phylogenies~\autocite{rasmussen2014phylodynamic}, inference
+genetic models to phylogenies~\citep{rasmussen2014phylodynamic}, inference
 of structural network parameters has to date been limited. However, it has been
 shown that network structure has a tangible impact on phylogeny
-shape~\autocite{leventhal2012inferring, colijn2014phylogenetic,
-goodreau2006assessing, robinson2013dynamics}, suggesting that such statistical
-inference might be possible~\autocite{welch2011statistical}.
+shape~\citep{leventhal2012inferring, colijn2014phylogenetic,
+goodreau2006assessing, robinson2013dynamics, villandre2016assessment},
+suggesting that such statistical inference might be
+possible~\citep{welch2011statistical}.
 
-Survey-based studies of sexual networks~\autocite{liljeros2001web,
-schneeberger2004scale, colgate1989risk} have found that these networks have a
-degree distribution which follows a power law \autocite[although there has been
-some disagreement, see][]{handcock2004likelihood}. That is, number of nodes of
-degree $k$ is proportional to $k^-\gamma$ for some constant $\gamma$. These
-networks are also referred to as ``scale-free''. One process by which
-scale-free networks can be generated is preferential attachment, where nodes
-with a high number of contacts attract new connections at an elevated rate. The
-first contact network model incorporating preferential attachment was
-introduced by \textcite{barabasi1999emergence}, and is now referred to as the
-\gls{BA} model. Under this model, networks are formed by iteratively adding
-nodes with $m$ new edges each. In the most commonly studied formulation, these
-new edges are joined to existing nodes of degree $k$ with probability
-proportional to $k$, so that nodes of high degree tend to attract more
-connections. \citeauthor{barabasi1999emergence} suggested an extension where
-the probability of attaching to a node of degree $k$ is $k^\alpha$ for some
-non-negative constant $\alpha$, and we use this extension here.
+Survey-based studies of sexual networks~\citep{liljeros2001web,
+schneeberger2004scale, colgate1989risk} have found that these networks tend to
+have a degree distribution which follows a power law \citep[although there has
+been some disagreement, see][]{handcock2004likelihood}. That is, the number of
+nodes of degree $k$ is proportional to $k^{-\gamma}$ for some constant
+$\gamma$. These networks are also referred to as ``scale-free''. One process by
+which scale-free networks can be generated is preferential attachment, where
+nodes with a high number of contacts attract new connections at an elevated
+rate. The first contact network model incorporating preferential attachment was
+introduced by \citet{barabasi1999emergence}, and is now referred to as the
+Barab\'asi-Albert (BA) model. Under this model, networks are formed by
+iteratively adding nodes with $m$ new edges each. In the most commonly studied
+formulation, these new edges are joined to existing nodes of degree $k$ with
+probability proportional to $k$, so that nodes of high degree tend to attract
+more connections. \citeauthor{barabasi1999emergence} suggested an extension
+where the probability of attaching to a node of degree $k$ is $k^\alpha$ for
+some non-negative constant $\alpha$, and we use this extension here.
 
 Previous work offers precedent for the possibility of statistical inference of
-structural network parameters. \textcite{britton2002bayesian} develop a
-Bayesian approach to estimate the edge density in an Erd\H{o}s-R\'enyi
-network~\autocite{erdos1960evolution} given observed infection dates, and
+structural network parameters. \citet{britton2002bayesian} develop a Bayesian
+approach to estimate the edge density in an Erd\H{o}s-R\'enyi
+network~\citep{erdos1960evolution} given observed infection dates, and
 optionally recovery dates. Their approach was later extended by
-\textcite{groendyke2011bayesian} and applied to a much larger data set of 188
-individuals. \textcite{volz2007susceptible, volz2008sir} developed differential
-equations describing the spread of a \gls{SI} epidemic on static and dynamic
-contact networks with several degree distributions, which could in principle be
-used for inference if observed incidence trajectories were available.
-\textcite{brown2011transmission} analysed the degree distribution of an
-approximate transmission network, estimated based on genetic similarity and
-estimated times of infection, relating 60\% of \gls{HIV}-infected \gls{MSM} in
-the United Kingdom. The transmission network is a subgraph of the contact
-network which includes only those edges which have already led to a new
-infection. The authors found that a Waring distribution, which is produced by a
-more sophisticated preferential attachment model, was a good fit to their
-estimated network.
+\citet{groendyke2011bayesian} and applied to a much larger data set of 188
+individuals. \citet{volz2007susceptible, volz2008sir} developed differential
+equations describing the spread of a susceptible-infected (SI) epidemic on
+static and dynamic contact networks with several degree distributions, which
+could in principle be used for inference if observed incidence trajectories
+were available. \citet{brown2011transmission} analysed the degree distribution
+of an approximate transmission network, estimated based on genetic similarity
+and estimated times of infection, relating 60\% of HIV-infected men who have
+sex with men (MSM) in the United Kingdom. The transmission network is a
+subgraph of the contact network which includes only those edges which have
+already led to a new infection. The authors found that a Waring distribution,
+which is produced by a more sophisticated preferential attachment model, was a
+good fit to their estimated network.
 
@@
 calculation of a normalizing constant, Bayesian inference is often performed
-using \gls{MCMC}, which uses likelihood \emph{ratios} in which the normalizing
-constants cancel out. Unfortunately, it is generally difficult to explicitly
-calculate the likelihood of an observed transmission tree under a contact
-network model, even up to a normalizing constant. To do so, it would be
-necessary to integrate over all possible networks, and also over all possible
-labellings of the internal nodes of the transmission tree. While it is not
-known (to us) whether such integration is tractable, a simpler alternative is
-offered by likelihood-free methods, namely \gls{ABC}. \gls{ABC} leverages the
-fact that, although calculating the likelihood may be impractical, generating
-simulated datasets according to a model is often straightforward. If our model
-fits the data well, the simulated data it produces should be similar to the
-observed data. More formally, if $D$ is the observed data, the posterior
-distribution $f(\theta \mid D)$ on model parameters $\theta$ is replaced as the
-target of statistical inference by $f(\theta \mid \rho(\hat{D}, D) <
-\varepsilon)$, where $\rho$ is a distance function, $\hat{D}$ is a simulated
-dataset according to $\theta$, and $\varepsilon$ is a small
-tolerance~\autocite{sunnaker2013approximate}. In the specific case when $\rho$
-is a kernel function, the approach is known as
-kernel-\gls{ABC}~\autocite{nakagome2013kernel, poon2015phylodynamic}.
+using Markov chain Monte Carlo (MCMC), which uses likelihood \emph{ratios} in
+which the normalizing constants cancel out. Unfortunately, it is generally
+difficult to explicitly calculate the likelihood of an observed transmission
+tree under a contact network model, even up to a normalizing constant. To do
+so, it would be necessary to integrate over all possible networks, and also
+over all possible labellings of the internal nodes of the transmission tree.
+While it is not known (to us) whether such integration is tractable, a simpler
+alternative is offered by likelihood-free methods, namely approximate Bayesian
+computation (ABC). ABC leverages the fact that, although calculating the
+likelihood may be impractical, generating simulated datasets according to a
+model is often straightforward. If our model fits the data well, the simulated
+data it produces should be similar to the observed data. More formally, if $D$
+is the observed data, the posterior distribution $f(\theta \mid D)$ on model
+parameters $\theta$ is replaced as the target of statistical inference by
+$f(\theta \mid \rho(\hat{D}, D) < \varepsilon)$, where $\rho$ is a distance
+function, $\hat{D}$ is a simulated dataset according to $\theta$, and
+$\varepsilon$ is a small tolerance~\citep{sunnaaker2013approximate}. In the
+specific case when $\rho$ is a kernel function, the approach is known as
+kernel-ABC~\citep{nakagome2013kernel, poon2015phylodynamic}.
 
-Here, we apply kernel-\gls{ABC} to the problem of statistical inference of
-contact network parameters from an estimated transmission tree, using the tree
-kernel developed by \textcite{poon2013mapping}. We then estimate the parameters
-of the \gls{BA} model on a variety of simulated and real data sets. Our results
-show that the attachment power parameter $\alpha$ can be inferred with
-reasonable accuracy, and can vary considerably between epidemics from different
-settings.
+Here, we develop a method using kernel-ABC to estimate the parameters of
+contact network models from observed phylogenetic data. The distance function
+we use is the tree kernel developed by \citet{poon2013mapping}, which computes
+a weighted dot product of the trees' representations in the space of all
+possible subset trees. We apply the method to investigate the parameters of the
+BA network model on a variety of simulated and real datasets. Our results show
+that some network parameters can be inferred with reasonable accuracy, while
+others have a minimal detectable impact on tree shape and therefore cannot be
+estimated accurately. We also find that these parameters can vary considerably
+between real epidemics from different settings.
 
-\section*{Methods}
+\section{New Approaches}
 
-We implemented a Gillespie simulation algorithm~\autocite{gillespie1976general}
-for simulating epidemics and transmission trees over static contact networks,
-in the same fashion as several previous studies
-\autocite[\textit{e.g.}][]{o2010contact, robinson2013dynamics,
-leventhal2012inferring, groendyke2011bayesian, goodreau2006assessing}. To check
-that our implementation was correct, we reproduced Figure 1A of
-\textcite{leventhal2012inferring} (our \cref{fig:sf1}), which plots the
-unbalancedness of transmission trees simulated over four network models at
-various levels of pathogen transmissibility. Our program is freely available at
-\url{https://github.com/rmcclosk/netabc}.
+We have developed a kernel-ABC-based method to perform statistical inference of
+contact network parameters from a transmission tree estimated from an observed
+viral phylogeny. We implemented the adaptive sequential Monte Carlo (SMC)
+algorithm for ABC developed by \citet{del2012adaptive}. The SMC algorithm keeps
+track of a population of parameter ``particles'', which are initially sampled
+from the parameters' joint prior distribution. Several datasets are simulated
+under the model of interest for each of the particles. In this case, the
+datasets are transmission trees, which are generated by a two-step process.
+First, a contact network is simulated according to the network model being fit.
+Second, a transmission tree is simulated over that network with a Gillespie
+simulation algorithm~\citep{gillespie1976general}, in the same fashion as
+several previous studies \citep[\textit{e.g.}][]{robinson2013dynamics,
+leventhal2012inferring}. The particles are weighted according to the similarity
+between their associated simulated trees and the observed tree. To quantify
+this similarity, we used the tree kernel developed by \citet{poon2013mapping}.
+Particles are iteratively perturbed by applying a Metropolis-Hastings kernel
+and, if the move is accepted, simulating new datasets under the new parameters.
+When a particle's weight drops to zero, because its simulated trees are too
+dissimilar to the observed tree, the particle is dropped from the population,
+and eventually replaced by a resampled particle with a higher weight. As the
+algorithm progresses, the population converges to a Monte Carlo approximation
+of the ABC target distribution, which is assumed to approximate the desired
+posterior~\citep{del2012adaptive, sunnaaker2013approximate}.
 
-We chose to study the \gls{BA} network model~\autocite{barabasi1999emergence}.
-In addition to $m$ and $\alpha$, we investigated the parameters $N$, which
-denotes the total number of nodes in the network, and $I$, which is the number
-of infected nodes at which to stop the simulation and sample the transmission
-tree. Nodes in our networks followed simple \gls{SI} dynamics, meaning that
-they became infected at a rate proportional to their number of infected
-neighbours, and never recovered. For all analyses, the transmission trees'
-branch lengths were scaled by dividing by their mean. We used the 
-\software{igraph} library's implementation of the \gls{BA}
-model~\autocite{csardi2006igraph} to generate the graphs. The analyses were run
-on Westgrid (\url{https://www.westgrid.ca/}) and a local computer cluster.
+A computer program implementing our method is freely available at
+\url{https://github.com/rmcclosk/netabc} (last accessed April 3, 2016).
 
-\subsection*{Kernel classifiers}
+\section{Results}
 
-We used the phylogenetic kernel developed by \textcite{poon2013mapping} to test
-whether the parameters of the \gls{BA} model had an effect on tree shape. We
-simulated 100 networks under each of three different values of $\alpha$: 0.5,
-1.0, and 1.5 (300 networks total). The other parameters were fixed to the
-following values: $N$ = 5000, $I$ = 1000, and $m$ = 2. A transmission tree with
-500 tips was simulated over each network (300 transmission trees total). The
-300 trees were compared pairwise with the tree kernel to form a $300 \times
-300$ kernel matrix. The kernel meta-parameters $\lambda$ (the ``decay
-factor''), and $\sigma$ (the ``radial basis function
-variance'')~\autocite[see][]{poon2013mapping}, were set to 0.3 and 4
-respectively. We constructed a \gls{kSVM} classifier for $\alpha$ using the
-\software{kernlab} package~\autocite{karatzoglou2004kernlab}, and evaluated its
-accuracy with 1000 two-fold cross-validations.
-
-Three similar experiments were performed for the other \gls{BA} model
-parameters (one experiment per parameter). $m$ was varied between 2, 3, and 4;
-$I$ between 500, 1000, and 2000; and $N$ between 3000, 5000, and 8000. The
-parameters not being tested were fixed at the values $N$ = 5000, $I$ = 1000,
-$m$ = 2, and $\alpha$ = 1. Thus, we performed a total of four kSVM
-cross-validations, one for each of the \gls{BA} model parameters $\alpha$, $I$,
-$m$, and $N$. We repeated these four cross-validations with different values of
-$\lambda$ (0.2, 0.3, and 0.4) and $\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$),
-as well as on trees with differing numbers of tips (100, 500, and 1000) and in
-epidemics of differing size (500, 1000, and 2000). When evaluating the
-classifier for $I$, we did not consider trees with 1000 tips, because one of
-the tested $I$ values was 500, and the number of tips cannot be larger than
-$I$.
-
-For each of the four parameters, we also tested univariate classifier based on
-Sackin's index~\autocite{shao1990tree} and an ordinary SVM based on the
-\gls{nLTT} statistic~\autocite{janzen2015approximate}.
-
-\subsection*{ABC simulations}
-
-We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
-\textcite{del2012adaptive}. To check that our implementation was correct, we
-applied it to the same mixture of Gaussians used by
-\citeauthor{del2012adaptive} to demonstrate their method (originally used
-by~\textcite{sisson2007sequential}). We were able to obtain a close
-approximation to the function (see \cref{fig:smctest}), and attained the
-stopping condition used by the authors in a comparable number of steps.
-
-We simulated three transmission trees, each with 500 tips, under every element
-of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
-\sett{1000, 2000}, $m$ = \sett{2, 3, 4}, and $\alpha$ = \sett{0.0, 0.5, 1,
-1.5}. This produced a total of 24 parameter combinations $\times$ three trees
-per combination = 72 trees total. The adaptive \gls{ABC} algorithm was applied
-to each tree with these priors: $m \sim$ Uniform(1, 5), $\alpha \sim$
-Uniform(0, 2), and $(N, I)$ jointly uniform on the region \{$500 \leq N \leq
-15000$, $500 \leq I \leq 5000$, $I \leq N$\}. Following
-\textcite{del2012adaptive} and \textcite{beaumont2009adaptive}, all proposals
-were Gaussian, with variance equal to twice the empirical variance of the
-particles. The algorithm was run with 1000 particles, 5 simulated datasets per
-particle, and the ``quality'' parameter controlling the decay rate of the
-tolerance $\varepsilon$ set to 0.95. We used the same stopping criterion as
-\citeauthor{del2012adaptive}, namely when the \gls{MCMC} acceptance rate
-dropped below 1.5\%. Point estimates for the parameters were obtained by taking
-the highest point of an estimated kernel density on the final set of particles,
-using the \software{density} function with the default parameters in
-\software{R}. \Gls{HPD} intervals were calculated with the
-\software{HPDinterval} function from the \software{R} package
-\software{coda}~\autocite{plummer2006coda}.
-
-Two further analyses were performed to address potential sources of error. To
-evaluate the effect of model misspecification in the case of heterogeneity
-among nodes, we generated a network where half the nodes were attached with
-power $\alpha$ = 0.5, and the other half with power $\alpha$ = 1.5. The other
-parameters for this network were $N$ = 5000, $I$ = 1000, and $m$ = 2. To
-investigate the effects of potential sampling bias, we simulated a transmission
-tree where the tips were sampled in a peer-driven fashion, rather than at
-random. That is, the probability to sample a node was twice as high if any of
-that node's network peers had already been sampled. The parameters of this
-network were $N$ = 5000, $I$ = 2000, $m$ = 2, and $\alpha$ = 0.5.
-
-\subsection*{Investigation of published data}
-
-We applied our kernel-ABC method to several published HIV datasets. Because the
-\gls{BA} model generates networks with a single connected component, we
-specifically searched for datasets which originated from existing clusters,
-either phylogenetically or geographically defined. Characteristics of the
-datasets we investigated are given in \cref{tab:data}.
-
-\begin{table}[ht]
-  \centering
-  \input{\tablepath/realdata}
-  \caption{Characteristics of published datasets investigated with kernel-ABC.
-  Acronyms: MSM, men who have sex with men; IDU, injection drug users.}
-  \label{tab:data}
-\end{table}
-
-We downloaded all sequences associated with each study from GenBank. For the
-\textcite{novitsky2014impact} data, each sequence was aligned pairwise to the
-HXB2 reference sequence (Genbank accession number HIVHXB2CG) and the
-hypervariable regions were clipped out with \software{BioPython} version
-1.66+~\autocite{cock2009biopython}. Sequences were multiply aligned using
-\software{MUSCLE} version 3.8.31 \autocite{edgar2004muscle}, and alignments
-were manually inspected with \software{Seaview} version 4.4.2
-\autocite{gouy2010seaview}. Phylogenies were constructed from the nucleotide
-alignments by approximate maximum likelihood using \software{FastTree2} version
-2.1.7 with the \gls{GTR} model. Transmission trees were estimated by rooting
-and time-scaling the phylogenies by root-to-tip regression, using a modified
-version of Path-O-Gen (distributed as part of
-BEAST~\autocite{drummond2007beast}) as described
-previously~\autocite{poon2015phylodynamic}.
-
-Two of the datasets \autocite{li2015hiv,novitsky2014impact} were initially much
-larger than the others, containing 1265 and 1299 sequences respectively. To
-ensure that the analyses were comparable, we reduced these to a number of
-sequences similar to the smaller datasets. For the \textcite{li2015hiv} data,
-we detected a cluster of size 280 using a patristic distance cutoff of 0.02 as
-described previously~\autocite{poon2014impact}. Only sequences within this
-cluster were carried forward. For the \textcite{novitsky2014impact} data, no
-large clusters were detected using the same cutoff, so we analysed a subtree of
-size 180 chosen arbitrarily.
-
-\section*{Results}
-
-\subsection*{Kernel classifiers}
+\subsection{Kernel classifiers}
 
@@
 
-Accuracy of the \gls{kSVM} classifiers varied based on the parameter being
-tested (\cref{fig:rsquared}, left). Classifiers based on two other tree
-statistics, the \gls{nLTT} and Sackin's index, generally exhibited worse
-performance than the tree kernel, although the magnitude of the disparity
-varied between the parameters (\cref{fig:rsquared}, centre and right).
-\Cref{fig:rsquared} shows the cross-validation accuracy of each of the tree
-classifiers. The results were largely robust to variations in the tree kernel
-meta-parameters $\lambda$ and $\sigma$, although accuracy varied between
-different epidemic and sampling scenarios
-(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}).
+We investigated the parameters of the BA network
+model~\citep{barabasi1999emergence}. In addition to $m$ and $\alpha$ (see
+Introduction), we considered $N$, which denotes the total number of nodes in
+the network, and $I$, which is the number of infected nodes at which to stop
+the simulation and sample the transmission tree. To examine the effect of these
+parameters on tree shape, we simulated transmission trees under different
+parameter values, calculated pairwise tree kernel scores between them, and
+attempted to classify the trees using a kernel support vector machine (kSVM).
+We also tested classifiers based on Sackin's index~\citep{shao1990tree} and the
+normalized lineages-through-time (nLTT)
+statistic~\citep{janzen2015approximate}. Accuracy of the kSVMs varied based on
+the parameter being tested (Figure~\ref{fig:rsquared}, left). Classifiers based
+on two other tree statistics, the nLTT and Sackin's index, generally exhibited
+worse performance than the tree kernel, although the magnitude of the disparity
+varied between the parameters (Figure~\ref{fig:rsquared}, centre and right).
+The results were largely robust to variations in the tree kernel
+meta-parameters $\lambda$ and $\sigma$ (Figures~S3-S6).
 
@@
     \Sexpr{nltt.alpha[,round(mean(rsquared), 2)]}
-for the \gls{nLTT}-based SVM, and
+for the nLTT-based SVM, and
     \Sexpr{sackin.alpha[,round(mean(rsquared), 2)]}
@@
 $R^2$ values of 
-  \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for \gls{kSVM},
-  \Sexpr{nltt.m[,round(mean(rsquared), 2)]} for the \gls{nLTT}, and
+  \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for kSVM,
+  \Sexpr{nltt.m[,round(mean(rsquared), 2)]} for the nLTT, and
   \Sexpr{sackin.m[,round(mean(rsquared), 2)]}
 for Sackin's index. Again, there was little variation in accuracy between
-epidemic scenarios, although the accuracy of the \gls{kSVM} was slightly higher
+epidemic scenarios, although the accuracy of the kSVM was slightly higher
 on 1000-tip trees.
 
-The accuracy of classifiers $I$ varied significantly with the number of tips in
-the tree. For 100-tip trees, the average $R^2$ values were
+The accuracy of classifiers for $I$ varied significantly with the number of
+tips in the tree. For 100-tip trees, the average $R^2$ values were
   \Sexpr{kernel.I[ntip == 100, round(mean(rsquared), 2)]},
@@
   \Sexpr{sackin.I[ntip == 100, round(mean(rsquared), 2)]}
-for the tree kernel, \gls{nLTT}, and Sackin's index respectively. For 500-tip
+for the tree kernel, nLTT, and Sackin's index respectively. For 500-tip
 trees, the values increased to
@@
 Finally, the performance of classifiers for $N$ depended heavily on the
-epidemic scenario. The $R^2$ of the \gls{kSVM} classifier ranged from
+epidemic scenario. The $R^2$ of the kSVM classifier ranged from
   \Sexpr{kernel.N[,round(min(rsquared), 2)]}
@@
   \Sexpr{kernel.N[,round(max(rsquared), 2)]}
-for the largest. Likewise, $R^2$ for the \gls{nLTT}-based SVM ranged from 
+for the largest. Likewise, $R^2$ for the nLTT-based SVM ranged from 
   \Sexpr{nltt.N[,round(min(rsquared), 2)]}
@@
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
   \centering
-  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
+  \includegraphics[width=\textwidth]{kernel-rsquared.eps}
+  \vspace{6pt}
   \caption{
       Cross-validation accuracy of kernel-SVM classifier (left), SVM classifier
-      using \gls{nLTT} (centre), and linear regression using Sackin's index
-      (right) for \gls{BA} model parameters. Kernel meta-parameters were set to
+      using nLTT (centre), and linear regression using Sackin's index
+      (right) for BA model parameters. Kernel meta-parameters were set to
       $\lambda = 0.3$ and $\sigma = 4$. Each point was calculated based on 300
@@
   \label{fig:rsquared}
-\end{figure}
+\end{figure*}
 
-\subsection*{ABC simulations}
+\subsection{ABC simulations}
 
@@
                                 d[true_alpha == 0.5 | true_alpha == 1.5, m_error])
-    options(scipen=4)
     stopifnot(alpha.av["factor(true_m)", "Pr(>F)"] > 0.05)
@@
 
-\Cref{fig:abcpt} shows \gls{MAP} point estimates of the \gls{BA} model
-parameters obtained with kernel-\gls{ABC} on simulated data. The estimates
-shown correspond only to the simulations where the $m$ parameter was set to 2,
-however the results for $m = 3$ and $m = 4$ were similar 
-(\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% \gls{HPD}
-intervals are given in \cref{tab:abchpd}.
+Figure~\ref{fig:abcpt} shows maximum \textit{a posteriori} (MAP) point
+estimates of the BA model parameters obtained with kernel-ABC on simulated
+data. The estimates shown correspond only to the simulations where the $m$
+parameter was set to 2, however the results for $m = 3$ and $m = 4$ were
+similar (Figures S7 and S8). Average boundaries of 95\% HPD intervals are given
+in Table~\ref{tab:abchpd}.
 
-The accuracy of the parameter estimates obtained with kernel-\gls{ABC}
-paralleled the results from the \gls{kSVM} classifier. Of the four parameters,
+The accuracy of the parameter estimates obtained with kernel-ABC
+paralleled the results from the kSVM classifier. Of the four parameters,
 $\alpha$ was the most accurately estimated, with point estimates having a
@@
 than those for the other values 
-    (Wilcoxon rank-sum test, $p$ = \Sexpr{round(zero.alpha.test$p.value, 4)}),
+    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(zero.alpha.test$p.value, 4))}$),
 but did not vary across the true values of the other parameters (one-way
@@
 at least 1
-    (Wilcoxon rank-sum test, $p$ = \Sexpr{round(alpha.I.test$p.value, 4)})
+    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(alpha.I.test$p.value, 4))}$)
 and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
@@
 of simulations. Oddly, the error in the estimated $m$ was higher for integer
-values of $\alpha$ (\ie~0 and 1) than non-integer values 
-    (Wilcoxon rask-sum test, $p$ = \Sexpr{round(alpha.m.test$p.value, 3)}).
+values of $\alpha$ (\text{i.e.}~0 and 1) than non-integer values 
+    (Wilcoxon rank-sum test, $p$ = \Sexpr{round(alpha.m.test$p.value, 3)}).
 The true values of the other parameters did not significantly affect the
@@
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
   \centering
-  \includegraphics[width=\textwidth]{abc-point-estimate-m2.pdf}
+  \includegraphics[width=\textwidth]{abc-point-estimate-m2.eps}
+  \vspace{6pt}
   \caption{
-    Point estimates of \gls{BA} model parameters obtained by running kernel-ABC
+    Point estimates of BA model parameters obtained by running kernel-ABC
     on simulated phylogenies without training, for simulations with $m = 2$.
     Dotted lines indicate true values, and limits of the $y$-axes are regions
-    of uniform prior density.
+    of uniform prior density. (A) Estimates for $\alpha$ and $I$ against their
+    true values in simulations. (B) Estimates for $m$ and $N$, which were held
+    fixed in these simulations, against true values of $\alpha$.
   }
   \label{fig:abcpt}
-\end{figure}
+\end{figure*}
 
-\begin{table}[ht]
+\begin{table*}[ht]
   \centering
-  \input{\tablepath/abc-hpd}
+    \begin{tabular}{lr>{\raggedleft\arraybackslash}p{2.5cm}>{\raggedleft\arraybackslash}p{2.5cm}>{\raggedleft\arraybackslash}p{2.5cm}}
+      \hline
+    Parameter & True value & Mean point estimate & Mean HPD lower bound & Mean HPD upper bound \\ 
+      \hline
+    $\alpha$ & 0.0 & 0.24 & 0.02 & 0.73 \\ 
+       & 0.5 & 0.42 & 0.02 & 0.81 \\ 
+       & 1.0 & 0.97 & 0.61 & 1.11 \\ 
+       & 1.5 & 1.48 & 1.26 & 1.83 \\ 
+      $I$ & 1000 & 1155.68 & 598.68 & 2402.84 \\ 
+       & 2000 & 2646.07 & 1182.31 & 4058.13 \\ 
+      $m$ & 2 & 2.92 & 1.75 & 4.92 \\ 
+       & 3 & 3.33 & 1.96 & 4.92 \\ 
+       & 4 & 3.62 & 1.88 & 5.00 \\ 
+      $N$ & 5000 & 10962.61 & 2732.55 & 14701.87 \\ 
+       \hline
+    \end{tabular}
   \caption{
-    \glsreset{HPD}
-    Average \acrlong{MAP} point estimates and 95\% \gls{HPD} interval widths
-    for \gls{BA} model parameter estimates obtained with kernel-\gls{ABC}.
-    Three transmission trees were simulated under each combination of the
-    listed parameter values, and the parameters were estimated with
-    kernel-\gls{ABC} without training.
+      Average maximum \textit{a posteriori} point estimates and 95\% highest
+      posterior density (HPD) interval widths for BA model parameter estimates
+      obtained with kernel-ABC. Three transmission trees were simulated under
+      each combination of the listed parameter values, and the parameters were
+      estimated with kernel-ABC without training.
   }
   \label{tab:abchpd}
-\end{table}
+\end{table*}
 
-The dispersion of the \gls{ABC} approximation to the posterior also varied
-between the parameters, with narrower \gls{HPD} intervals for the parameters
-with more accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows
+The dispersion of the ABC approximation to the posterior also varied between
+the parameters, with narrower HPD intervals for the parameters with more
+accurate point estimates (Table~\ref{tab:abchpd}). Figure~\ref{fig:abcex} shows
 the distributions for for one simulation (equivalent plots for all the
-simulations can be found in \cref{dat:abcpost}). \Gls{HPD} intervals around
-$\alpha$ and $I$ were narrow relative to the region of nonzero prior density,
-whereas the intervals for $m$ and $N$ were more widely dispersed.
+simulations can be found in Supplementary Figure S10). HPD intervals around $\alpha$
+and $I$ were narrow relative to the region of nonzero prior density, whereas
+the intervals for $m$ and $N$ were more widely dispersed.
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
     \centering
-  \includegraphics[width=\textwidth]{abc-posterior-example.pdf}
+  \includegraphics[width=\textwidth]{abc-posterior-example.eps}
+  \vspace{6pt}
   \caption{
-    Marginal posterior distributions of \gls{BA} model parameters estimated
-    with kernel-\gls{ABC} for a single simulated transmission tree. Dotted
+    Marginal posterior distributions of BA model parameters estimated
+    with kernel-ABC for a single simulated transmission tree. Dotted
     lines and shaded polygon indicate true values.
@@
   \label{fig:abcex}
-\end{figure}
+\end{figure*}
 
@@
 the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
-other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The \gls{MAP} [95\%
+other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
 HPD] estimates for each parameter were: 
@@
 peer-driven fashion, where the probability to sample a node was twice as high
-if one of it's peers had already been sampled. The parameters for this
+if one of its peers had already been sampled. The parameters for this
 experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and $I$ = 2000. The
@@
 Both of these results were in line with estimates obtained on other simulated
-datasets (\cref{tab:abchpd}).
+datasets (Table~\ref{tab:abchpd}).
 
-\subsection*{Real data}
+\subsection{Real data}
 
@@
 
-There was substantial heterogeneity among the parameter estimates for the five
-published HIV datasets we analysed. Two of the
-datasets~\autocite{niculescu2015recent, wang2015targeting} had estimated
-$\alpha$ values near unity (\gls{MAP} estimate [95\% \gls{HPD}] 
+We applied kernel-ABC to five published HIV datasets (Table~\ref{tab:data}),
+and found substantial heterogeneity among the parameter estimates
+(Figure~\ref{fig:abchpd}). Two of the datasets~\citep{niculescu2015recent,
+wang2015targeting} had estimated $\alpha$ values near unity (MAP estimates
+[95\% HPD] 
   \Sexpr{map["niculescu2015", round(alpha, 2)]} 
@@
   [\Sexpr{low["wang2015", round(alpha, 2)]} -
-   \Sexpr{up["wang2015", round(alpha, 2)]}]). 
-Another two datasets~\autocite{li2015hiv, cuevas2009hiv} had lower estimated
-values and wider \gls{HPD} intervals
+   \Sexpr{up["wang2015", round(alpha, 2)]}] respectively). 
+Another two datasets~\citep{li2015hiv, cuevas2009hiv} had lower estimated
+values and wider HPD intervals
   (\Sexpr{map["li2015", round(alpha, 2)]} 
@@
    \Sexpr{up["cuevas2009", round(alpha, 2)]}]). 
-The \textcite{novitsky2014impact} data had an extremely low estimated $\alpha$
-and a very wide \gls{HPD} interval
+The \citet{novitsky2014impact} data had an extremely low estimated $\alpha$
+and a very wide HPD interval
   (\Sexpr{map["novitsky2014", round(alpha, 2)]} 
@@
 For all the datasets except \citeauthor{novitsky2014impact}, estimated values of
-$I$ were below 2000, with narrow \gls{HPD} intervals around two of the
+$I$ were below 2000, with narrow HPD intervals around two of the
 datasets
@@
 The \citeauthor{novitsky2014impact} data was again the outlier, with a very high
-estimated $I$, and \gls{HPD} interval spanning almost the entire prior region
+estimated $I$, and HPD interval spanning almost the entire prior region
   (\Sexpr{map["novitsky2014", floor(I)]} 
@@
    \Sexpr{up["novitsky2014", floor(I)]}]).
-No information was gleaned about the $m$ parameter, with the \gls{HPD} interval
+No information was gleaned about the $m$ parameter, with the HPD interval
 occupying the entire prior region for all datasets. The estimates of $N$ were
@@
 
-\begin{figure}[ht]
+\begin{figure*}[ht]
   \centering
-  \includegraphics{realdata-hpd.pdf}
+  \includegraphics{realdata-hpd.eps}
+  \vspace{8pt}
   \caption{
-    \acrlong{MAP} point estimates and 95\% \gls{HPD} intervals for parameters
-    of the \gls{BA} network model, fitted to five published \gls{HIV} datasets
-    with kernel-\gls{ABC}.
+      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
+      parameters of the BA network model, fitted to five published HIV datasets
+      with kernel-ABC.
   }
   \label{fig:abchpd}
-\end{figure}
+\end{figure*}
 
-\section*{Discussion}
+\section{Discussion}
 
 Contact networks can have a strong influence on epidemic progression, and are
-potentially useful as a public health tool~\autocite{wang2015targeting,
+potentially useful as a public health tool~\citep{wang2015targeting,
 little2014using}. Despite this, few methods exist for investigating contact
-network parameters in a phylodynamic framework~\autocite{groendyke2011bayesian,
-volz2008sir, brown2011transmission}. Kernel-ABC is a model-agnostic method
-which can be used to investigate any quantity that affects tree shape. In this
-work, we developed a kernel-ABC-based method to infer the parameters of a
-contact network model. The method is general, and could be applied to any model
-from which contact networks can be simulated. We demonstrated the method on the
-\gls{BA} model, which is a simple model giving rise to the power law degree
-distributions commonly observed in real-world networks.
+network parameters in a phylodynamic framework~\citep{groendyke2011bayesian,
+volz2008sir, brown2011transmission, leventhal2012inferring}. Kernel-ABC is a
+model-agnostic method which can be used to investigate any quantity that
+affects tree shape~\citep{poon2015phylodynamic}. In this work, we developed a
+kernel-ABC-based method to infer the parameters of a contact network model. The
+method is general, and could be applied to any model from which contact
+networks can be simulated. We demonstrated the method on the BA model,
+which is a simple preferential attachment model giving rise to the power law
+degree distributions commonly observed in real-world networks.
 
@@
 unreasonable to expect good estimation of $N$, because adding more nodes to a
-\gls{BA} network does not change the edge density or overall shape. This can be
+BA network does not change the edge density or overall shape. This can be
 illustrated by imagining that we add a small number of nodes to a network after
@@
 
-As noted by \textcite{lintusaari2016identifiability}, uniform priors on model
+As noted by \citet{lintusaari2016identifiability}, uniform priors on model
 parameters may translate to highly informative priors on quantities of
 interest. We observed a non-linear relationship between the preferential
-attachment power $\alpha$ and the power law exponent $\gamma$
-(\cref{fig:gamma}). Therefore, placing a uniform prior on $\alpha$ between 0
-and 2 is equivalent to placing an informative prior that $\gamma$ is close to
-2. Therefore, if we were primarily interested in $\gamma$ rather than $\alpha$,
-a more sensible choice of prior might have a shape similar to \cref{fig:gamma}
-and be bounded above by approximately $\alpha$ = 1.5. This would uniformly
-bound $\gamma$ in the region $2 \leq \gamma \leq 4$ commonly reported in the
-network literature~\autocite{liljeros2001web, schneeberger2004scale,
-colgate1989risk, brown2011transmission}. We note however that
-\textcite{jones2003assessment} estimated $\gamma$ values greater than four, in
-one case as high as 17, for some datasets, indicating that a wider range of
-permitted $\gamma$ values may be warranted.
+attachment power $\alpha$ and the power law exponent $\gamma$ (Figure~S9).
+Therefore, placing a uniform prior on $\alpha$ between 0 and 2 is equivalent to
+placing an informative prior that $\gamma$ is close to 2. Therefore, if we were
+primarily interested in $\gamma$ rather than $\alpha$, a more sensible choice
+of prior might have a shape similar to the inverse of Figure~S9 and be bounded
+above by approximately $\alpha$ = 1.5. This would uniformly bound $\gamma$ in
+the region $2 \leq \gamma \leq 4$ commonly reported in the network
+literature~\citep{liljeros2001web, schneeberger2004scale, colgate1989risk,
+brown2011transmission}. We note however that \citet{jones2003assessment}
+estimated $\gamma$ values greater than four, in one case as high as 17, for
+some datasets, indicating that a wider range of permitted $\gamma$ values may
+be warranted.
 
@@
 number of males from the network could obfuscate the true network structure,
-for example if there were a disproportionate number of highly connected nodes
-of one gender. Second, the town in question was in close proximity to the
-country's capital, and the authors indicated that a high amount of migration
-takes place between the two locations. This suggests that the contact network
-may include a much larger group based in the capital city, which would explain
-the high estimate of $I$.
+for example if the majority of highly connected nodes were of one gender.
+Second, the town in question was in close proximity to the country's capital,
+and the authors indicated that a high amount of migration takes place between
+the two locations. This suggests that the contact network may include a much
+larger group based in the capital city, which would explain the high estimate
+of $I$.
 
-When interpreting these results, we caution that the \gls{BA} model is quite
+When interpreting these results, we caution that the BA model is quite
 simple and most likely misspecified for these data. In particular, the average
@@
 issues, we felt it was best to demonstrate the method first on a simple model.
-It is possible to fit more complex models which address some of these issues,
-such as one incorporating heterogeneous node behaviour, which may prove a
-fruitful avenue for future investigations.
+It is possible to use this framework to fit more complex models which address
+some of these issues, such as one incorporating heterogeneous node behaviour,
+which may prove a fruitful avenue for future investigations.
 
@@
 this has been demonstrated to be a fair
-approximation~\autocite[e.g.][]{leitner1996accurate}, and is frequently used in
-practice~\autocite[e.g.][]{stadler2013uncovering}, the topologies of a viral
+approximation~\citep[e.g.][]{leitner1996accurate}, and is frequently used in
+practice~\citep[e.g.][]{stadler2013uncovering}, the topologies of a viral
 phylogeny and transmission tree can differ
-significantly~\autocite{ypma2013relating} due to within-host evolution and the
-sampling process. In addition, the \gls{ABC}-\gls{SMC} algorithm is
+significantly~\citep{ypma2013relating} due to within-host evolution and the
+sampling process. In addition, the ABC-SMC algorithm is
 computationally intensive, taking about a day when run on 20 cores in parallel
@@
 characteristics of the network structure underlying disease outbreaks. This
-work, and previous work by our group~\autocite{poon2015phylodynamic}, has
+work, and previous work by our group~\citep{poon2015phylodynamic}, has
 demonstrated that kernel-ABC is a broadly applicable and effective framework in
@@
 
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{data}{0}
-\renewcommand{\thedata}{S\arabic{data}}
+\section{Materials and Methods}
 
-\printbibliography
+Nodes in our networks followed simple SI dynamics, meaning that they
+became infected at a rate proportional to their number of infected neighbours,
+and never recovered. For all analyses, the transmission trees' branch lengths
+were scaled by dividing by their mean. We used the \textit{igraph} library's
+implementation of the BA model~\citep{csardi2006igraph} to generate the graphs.
+The analyses were run on Westgrid (\url{https://www.westgrid.ca/}) and a local
+computer cluster.
 
-\newpage
-\section*{Supplemental Materials}
+\subsection{Computer program validation}
 
-\begin{figure}[H]
+To check that our implementation of Gillespie simulation was correct, we
+reproduced Figure 1A of \citet{leventhal2012inferring} (our Figure~S1), which
+plots the unbalancedness of transmission trees simulated over four network
+models at various levels of pathogen transmissibility. Our implementation of
+adaptive ABC-SMC was tested by applying it to the same mixture of Gaussians
+used by \citeauthor{del2012adaptive} to demonstrate their method (originally
+used by~\citet{sisson2007sequential}). We were able to obtain a close
+approximation to the function (see Figure~S2), and attained the stopping
+condition used by the authors in a comparable number of steps.
+
+\subsection{Kernel classifiers}
+
+We used the phylogenetic kernel developed by \citet{poon2013mapping} to test
+whether the parameters of the BA model had an effect on tree shape. 100
+networks were simulated under each of three different values of $\alpha$: 0.5,
+1.0, and 1.5 (300 networks total). The other parameters were fixed to the
+following values: $N$ = 5000, $I$ = 1000, and $m$ = 2. A transmission tree with
+500 tips was simulated over each network (300 transmission trees total). The
+300 trees were compared pairwise with the tree kernel to form a $300 \times
+300$ kernel matrix. The kernel meta-parameters $\lambda$ (the ``decay
+factor''), and $\sigma$ (the ``radial basis function
+variance'')~\citep[see][]{poon2013mapping}, were set to 0.3 and 4 respectively.
+We constructed a kSVM classifier for $\alpha$ using the \textit{kernlab}
+package~\citep{zeileis2004kernlab}, and evaluated its accuracy with 1000
+two-fold cross-validations.
+
+Three similar experiments were performed for the other BA model parameters (one
+experiment per parameter). $m$ was varied between 2, 3, and 4; $I$ between 500,
+1000, and 2000; and $N$ between 3000, 5000, and 8000. The parameters not being
+tested were fixed at the values $N$ = 5000, $I$ = 1000, $m$ = 2, and $\alpha$ =
+1. Thus, we performed a total of four kSVM cross-validations, one for each of
+the BA model parameters $\alpha$, $I$, $m$, and $N$. We repeated these four
+cross-validations with different values of $\lambda$ (0.2, 0.3, and 0.4) and
+$\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$), as well as on trees with
+differing numbers of tips (100, 500, and 1000) and in epidemics of differing
+size (500, 1000, and 2000). The combination of the number of sampled
+individuals (\textit{i.e.} the number of tips) and the epidemic size
+(\textit{i.e.} $I$) will be referred to as an ``epidemic scenario''. When
+evaluating the classifier for $I$, we did not consider trees with 1000 tips,
+because one of the tested $I$ values was 500, and the number of tips cannot be
+larger than $I$.
+
+For each of the four parameters, we also tested a linear regression against
+Sackin's index~\citep{shao1990tree} and an ordinary SVM based on the normalized
+lineages-through-time (nLTT) statistic~\citep{janzen2015approximate}.
+
+\subsection{ABC simulations}
+
+We simulated three transmission trees, each with 500 tips, under every element
+of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
+\{1000, 2000\}, $m$ = \{2, 3, 4\}, and $\alpha$ = \{0.0, 0.5, 1,
+1.5\}. This produced a total of 24 parameter combinations $\times$ three trees
+per combination = 72 trees total. The adaptive ABC algorithm was applied to
+each tree with these priors: $m \sim$ Uniform(1, 5), $\alpha \sim$ Uniform(0,
+2), and $(N, I)$ jointly uniform on the region \{$500 \leq N \leq 15000$, $500
+\leq I \leq 5000$, $I \leq N$\}. Following \citet{del2012adaptive} and
+\citet{beaumont2009adaptive}, all proposals were Gaussian, with variance equal
+to twice the empirical variance of the particles. The algorithm was run with
+1000 particles, 5 simulated datasets per particle, and the ``quality''
+parameter controlling the decay rate of the tolerance $\varepsilon$ set to
+0.95. We used the same stopping criterion as \citeauthor{del2012adaptive},
+namely when the MCMC acceptance rate dropped below 1.5\%. Point estimates for
+the parameters were obtained by taking the highest point of an estimated kernel
+density on the final set of particles, calculated using the \textit{density}
+function with the default parameters in \textit{R}. Highest posterior density
+(HPD) intervals were calculated with the \textit{HPDinterval} function from
+the \textit{R} package \textit{coda}~\citep{plummer2006coda}.
+
+Two further analyses were performed to address potential sources of error. To
+evaluate the effect of model misspecification in the case of heterogeneity
+among nodes, we generated a network where half the nodes were attached with
+power $\alpha$ = 0.5, and the other half with power $\alpha$ = 1.5. The other
+parameters for this network were $N$ = 5000, $I$ = 1000, and $m$ = 2. To
+investigate the effects of potential sampling bias, we simulated a transmission
+tree where the tips were sampled in a peer-driven fashion, rather than at
+random. That is, the probability to sample a node was twice as high if any of
+that node's network peers had already been sampled. The parameters of this
+network were $N$ = 5000, $I$ = 2000, $m$ = 2, and $\alpha$ = 0.5.
+
+\subsection{Investigation of published data}
+
+We applied our kernel-ABC method to several published HIV datasets. Because the
+BA model generates networks with a single connected component, we
+specifically searched for datasets which originated from existing clusters,
+either phylogenetically or geographically defined. Characteristics of the
+datasets we investigated are given in Table~\ref{tab:data}.
+
+\begin{table*}[!t]
   \centering
-  \includegraphics{leventhal2012fig1.pdf}
-  \caption{
-      Reproduction of Figure 1A from \textcite{leventhal2012inferring} used to
-      check the accuracy of our implementation of Gillespie simulation.
-      Transmission trees were simulated over three types of network, with
-      pathogen transmissimility varying from 0 to 1. Sackin's index was
-      calculated for each simulated transmission tree.
-  }
-  \label{fig:sf1}
-\end{figure}
+  \begin{tabular}{ccccc}
+    Reference & Sequences ($n$) & Location & Risk group & Gene \\
+    \hline
+    \citet{wang2015targeting} & 173 & Beijing, China & MSM & \textit{pol} \\
+    \citet{cuevas2009hiv} & 287 & Basque Country, Spain & mixed & \textit{pol} \\
+    \citet{novitsky2013phylogenetic} & \multirow{2}{*}{180} &
+    \multirow{2}{*}{Mochudi, Botswana} & \multirow{2}{*}{mixed} &
+    \multirow{2}{*}{\textit{env}} \\ \citet{novitsky2014impact} \\
+    \citet{li2015hiv} & 280 & Shanghai, China & MSM & \textit{pol} \\
+    \citet{niculescu2015recent} & 136 & Romaina & IDU & \textit{pol} \\
+    \hline
+  \end{tabular}
+  \caption{Characteristics of published datasets investigated with kernel-ABC.
+  Acronyms: MSM, men who have sex with men; IDU, injection drug users.}
+  \label{tab:data}
+\end{table*}
 
-\begin{figure}[H]
-  \centering
-  \includegraphics{smc-test.pdf}
-  \caption{Approximation of mixture of Gaussians used by
-    \textcite{del2012adaptive} and \textcite{sisson2007sequential} to test
-    \gls{SMC}. Solid black line indicates true distribution. Grey shaded area
-    shows SMC approximation obtained with our implementation, using 10000
-    particles with one simulated data point per particle.
-  }
-  \label{fig:smctest}
-\end{figure}
+We downloaded all sequences associated with each study from GenBank. For the
+\citet{novitsky2014impact} data, each sequence was aligned pairwise to the HXB2
+reference sequence (Genbank accession number K03455) and the hypervariable
+regions were clipped out with \textit{BioPython} version
+1.66+~\citep{cock2009biopython}. Sequences were multiply aligned using
+\textit{MUSCLE} version 3.8.31 \citep{edgar2004muscle}, and alignments were
+manually inspected with \textit{Seaview} version 4.4.2
+\citep{gouy2010seaview}. Phylogenies were constructed from the nucleotide
+alignments by approximate maximum likelihood using \textit{FastTree2} version
+2.1.7 with the generalized time-reversible model. Transmission trees were
+estimated by rooting and time-scaling the phylogenies by root-to-tip
+regression, using a modified version of Path-O-Gen (distributed as part of
+BEAST~\citep{drummond2007beast}) as described
+previously~\citep{poon2015phylodynamic}.
 
-\begin{figure}[H]
-  \centering
-  \includegraphics{kernel-alpha-crossv.pdf}
-  \caption{
-    Cross-validation accuracy of kernel-SVM classifiers for $\alpha$ parameter
-    of \gls{BA} network model, for various tree kernel meta-parameters and
-    epidemic scenarios. Each point was calculated based on 300 simulated
-    transmission trees over networks with $\alpha$ = 0.5, 1.0, or 1.5. Dotted
-    and and dashed lines indicate, respectively, performance of SVM using the
-    \gls{nLTT} statistic, and linear regression using Sackin's index. Facets
-    are number of infected nodes before the simulation was stopped ($I$) and
-    number of tips in the sampled transmission tree.
-  }
-  \label{fig:alphacrossv}
-\end{figure}
+Two of the datasets \citep{li2015hiv,novitsky2014impact} were initially much
+larger than the others, containing 1265 and 1299 sequences respectively. To
+ensure that the analyses were comparable, we reduced these to a number of
+sequences similar to the smaller datasets. For the \citet{li2015hiv} data,
+we detected a cluster of size 280 using a patristic distance cutoff of 0.02 as
+described previously~\citep{poon2015impact}. Only sequences within this
+cluster were carried forward. For the \citet{novitsky2014impact} data, no
+large clusters were detected using the same cutoff, so we analysed a subtree of
+size 180 chosen arbitrarily.
 
-\begin{figure}[H]
-  \centering
-  \includegraphics{kernel-m-crossv.pdf}
-  \caption{Cross-validation accuracy of kernel-SVM classifiers for $m$
-      parameter of \gls{BA} network model, for various tree kernel
-      meta-parameters and epidemic scenarios. Each point was calculated based
-      on 300 simulated transmission trees over networks with $m$ = 2, 3, or 4.
-      Dotted and and dashed lines indicate, respectively, performance of SVM
-      using the \gls{nLTT} statistic, and linear regression using Sackin's
-      index. Facets are number of infected nodes before the simulation was
-      stopped ($I$) and number of tips in the sampled transmission tree.
-  }
-  \label{fig:mcrossv}
-\end{figure}
+For all datasets, we used the priors $\alpha$ $\sim$ Uniform(0, 2), $m$ $\sim$
+Uniform(2, 5), and $N$ and $I$ jointly uniform on the region 
+\{$n \leq N \leq 10000$, $n \leq I \leq 10000$, $I \leq N$\}, where $n$ is the
+number of tips in the tree. The other parameters to the SMC algorithm
+were the same as used for the simulation experiments.
 
-\begin{figure}[H]
-  \centering
-  \includegraphics{kernel-I-crossv.pdf}
-  \caption{Cross-validation accuracy of kernel-SVM classifiers for number of
-      infected nodes ($I$) under \gls{BA} network model, for various tree
-      kernel meta-parameters and two tree sizes. Each point was calculated
-      based on 300 simulated transmission trees over networks with $I$ = 500,
-      1000, or 2000. Dotted and and dashed lines indicate, respectively,
-      performance of SVM using the \gls{nLTT} statistic, and linear regression
-      using Sackin's index. Facets are the number of tips in the sampled
-      transmission tree.
-  }
-  \label{fig:Icrossv}
-\end{figure}
+\section{Supplementary data}
 
-\begin{figure}[H]
-  \centering
-  \includegraphics{kernel-N-crossv.pdf}
-  \caption{Cross-validation accuracy of kernel-SVM classifiers for total number
-      of nodes ($N$) under \gls{BA} network model, for various tree kernel
-      meta-parameters and epidemic scenarios sizes. Each point was calculated
-      based on 300 simulated transmission trees over networks with $N$ = 3000,
-      5000, or 8000. Dotted and and dashed lines indicate, respectively,
-      performance of SVM using the \gls{nLTT} statistic, and linear regression
-      using Sackin's index. Facets are the number of tips in the sampled
-      transmission tree.
-  }
-  \label{fig:Ncrossv}
-\end{figure}
+Supplementary figures S1-S10 are available at Molecular Biology and Evolution
+online (\url{http://www.mbe.oxfordjournals.org/}).
 
-\begin{figure}[H]
-  \centering
-  \includegraphics{abc-point-estimate-m3.pdf}
-  \caption{
-    Point estimates of \gls{BA} model parameters obtained by running kernel-ABC
-    on simulated phylogenies without training, for simulations with $m = 3$.
-    Dotted lines indicate true values, and limits of the $y$-axes are regions
-    of uniform prior density.
-  }
-  \label{fig:abcptm3}
-\end{figure}
+\section{Acknowledgements}
 
-\begin{figure}[H]
-  \centering
-  \includegraphics{abc-point-estimate-m4.pdf}
-  \caption{
-    Point estimates of \gls{BA} model parameters obtained by running kernel-ABC
-    on simulated phylogenies without training, for simulations with $m = 4$.
-    Dotted lines indicate true values, and limits of the $y$-axes are regions
-    of uniform prior density.
-  }
-  \label{fig:abcptm4}
-\end{figure}
+This work was supported by grants from the Canadian Institutes of Health
+Research (CIHR, operating grant HOP-111406), and the Bill \& Melinda Gates
+Foundation (award number OPP1110049). R.M.M. was supported by a scholarship
+from the CIHR Strategic Training Program in Bioinformatics. A.F.Y.P. was
+supported by a CIHR New Investigator Award (Canadian HIV Vaccine Initiative,
+Vaccine Discovery and Social Research) and by a Career Investigator Scholar
+Award from the Michael Smith Foundation for Health Research, in partnership
+with the Providence Health Care Research Institute and St. Paul's Hospital
+Foundation.
 
-\begin{figure}[H]
-  \centering
-  \includegraphics{alpha-gamma.pdf}
-  \caption{
-      Relationship between preferential attachment power parameter $\alpha$
-      and power law exponent $\gamma$ for networks simulated under the \gls{BA}
-      network model with $N$ = 5000 and $m$ = 2.
-  }
-  \label{fig:gamma}
-\end{figure}
-
-\begin{data}[H]
-  \centering
-  \includegraphics{stock/missing}
-  \caption{
-    Plots of marginal posterior distributions estimated with kernel-\gls{ABC}
-    for all simulated transmission trees.
-  }
-  \label{dat:abcpost}
-\end{data}
+\bibliographystyle{natbib}
+\bibliography{papers}
 

### 2016-04-20 doc/paper/supp-f1-9.tex

\documentclass[12pt]{article}
\usepackage{graphicx}
\usepackage{fullpage}

\begin{document}

\section*{Supplementary Figures S1-S9}

\renewcommand{\thefigure}{S\arabic{figure}}

\begin{figure}[ht]
  \centering
  \includegraphics{leventhal2012fig1.pdf}
  \caption{
    Reproduction of Figure 1A from Leventhal \textit{et al.} (2012) used to
    check the accuracy of our implementation of Gillespie simulation.
    Transmission trees were simulated over three types of network, with
    pathogen transmissibility varying from 0 to 1. Sackin's index was
    calculated for each simulated transmission tree.
  }
  \label{fig:sf1}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics{smc-test.pdf}
  \caption{Approximation of mixture of Gaussians used by
    Del Moral \textit{et al.} (2012) and Sisson \textit{et al.} (2009) to test
    SMC. Solid black line indicates true distribution. Grey shaded area shows
    SMC approximation obtained with our implementation, using 10000
    particles with one simulated data point per particle.
  }
  \label{fig:smctest}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics{kernel-alpha-crossv.pdf}
  \caption{
    Cross-validation accuracy of kernel-SVM classifiers for $\alpha$ parameter
    of BA network model, for various tree kernel meta-parameters and
    epidemic scenarios. Each point was calculated based on 300 simulated
    transmission trees over networks with $\alpha$ = 0.5, 1.0, or 1.5. Dotted
    and and dashed lines indicate, respectively, performance of SVM using the
    nLTT statistic, and linear regression using Sackin's index. Facets
    are number of infected nodes before the simulation was stopped ($I$) and
    number of tips in the sampled transmission tree.
  }
  \label{fig:alphacrossv}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics{kernel-m-crossv.pdf}
  \caption{Cross-validation accuracy of kernel-SVM classifiers for $m$
      parameter of BA network model, for various tree kernel
      meta-parameters and epidemic scenarios. Each point was calculated based
      on 300 simulated transmission trees over networks with $m$ = 2, 3, or 4.
      Dotted and and dashed lines indicate, respectively, performance of SVM
      using the nLTT statistic, and linear regression using Sackin's
      index. Facets are number of infected nodes before the simulation was
      stopped ($I$) and number of tips in the sampled transmission tree.
  }
  \label{fig:mcrossv}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics{kernel-I-crossv.pdf}
  \caption{Cross-validation accuracy of kernel-SVM classifiers for number of
      infected nodes ($I$) under BA network model, for various tree
      kernel meta-parameters and two tree sizes. Each point was calculated
      based on 300 simulated transmission trees over networks with $I$ = 500,
      1000, or 2000. Dotted and and dashed lines indicate, respectively,
      performance of SVM using the nLTT statistic, and linear regression
      using Sackin's index. Facets are the number of tips in the sampled
      transmission tree.
  }
  \label{fig:Icrossv}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics{kernel-N-crossv.pdf}
  \caption{Cross-validation accuracy of kernel-SVM classifiers for total number
      of nodes ($N$) under BA network model, for various tree kernel
      meta-parameters and epidemic scenarios sizes. Each point was calculated
      based on 300 simulated transmission trees over networks with $N$ = 3000,
      5000, or 8000. Dotted and and dashed lines indicate, respectively,
      performance of SVM using the nLTT statistic, and linear regression
      using Sackin's index. Facets are the number of tips in the sampled
      transmission tree.
  }
  \label{fig:Ncrossv}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics{abc-point-estimate-m3.pdf}
  \caption{
    Point estimates of BA model parameters obtained by running kernel-ABC
    on simulated phylogenies without training, for simulations with $m = 3$.
    Dotted lines indicate true values, and limits of the $y$-axes are regions
    of uniform prior density. (A) Estimates for $\alpha$ and $I$ against their
    true values in simulations. (B) Estimates for $m$ and $N$, which were held
    fixed in these simulations, against true values of $\alpha$.
  }
  \label{fig:abcptm3}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics{abc-point-estimate-m4.pdf}
  \caption{
    Point estimates of BA model parameters obtained by running kernel-ABC
    on simulated phylogenies without training, for simulations with $m = 4$.
    Dotted lines indicate true values, and limits of the $y$-axes are regions
    of uniform prior density. (A) Estimates for $\alpha$ and $I$ against their
    true values in simulations. (B) Estimates for $m$ and $N$, which were held
    fixed in these simulations, against true values of $\alpha$.
  }
  \label{fig:abcptm4}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics{alpha-gamma.pdf}
  \caption{
      Relationship between preferential attachment power parameter $\alpha$
      and power law exponent $\gamma$ for networks simulated under the BA
      network model with $N$ = 5000 and $m$ = 2.
  }
  \label{fig:gamma}
\end{figure}

\end{document}

### 2016-04-20 doc/thesis/conclusion.tex

@@
+Due to the rapid advancement of nucleotide sequencing technology, viral
+sequence data data have become increasingly feasible to collect on a population
+level. Through phylodynamic methods, these data offer a window into
+epidemiological processes which would otherwise be virtually impossible to
+study on a realistic scale.
+
 This thesis developed \software{netabc}, a computer program implementing a
 statistical inference method for contact network parameters from viral
-phylogenetic data. Due to the rapid advancement of nucleotide sequencing
-technology, viral sequence data data have become increasingly feasible to
-collect on a population level. Through phylodynamic methods, these data offer a
-window into epidemiological processes which would otherwise be virtually
-impossible to study on a realistic scale. Although phylodynamic methods have
-been developed to fit a wide variety of epidemiological models to phylogenetic
-data~\autocite{volz2012complex, rasmussen2014phylodynamic}, our method is the
-first (to our knowledge) which can fit models that do not assume panmixia. We
-believe this capability will be of broad interest to the molecular evolution
-and epidemiology community, as it widens the field of epidemiological
-parameters which may be investigated through viral sequence data. In addition,
-the characterization of local contact networks could be valuable from a public
-health perspective, such as for investigating optimal vaccination strategies.
-This information could assist in curtailing current epimemics, as well as
-preventing future epidemics of different diseases over the same contact
-network.
+phylogenetic data. \software{Netabc} brings together the areas of viral
+phylodynamics and network epidemiology, which have only intersected in a very
+limited fashion thus far~\autocite{welch2011statistical}. The use of
+kernel-\gls{ABC}, a likelihood-free method, makes it possible to fit network
+models to phylogenies without calculating intractible likelihoods.
 
-Contact network structure has a substantial impact on epidemic trajectory
-\autocite{barthelemy2005dynamical, volz2007susceptible, volz2008sir}, and a few
-methods have been developed to estimate network parameters from epidemiological
-data \autocite{volz2007susceptible, britton2002bayesian,
-groendyke2011bayesian}. It is known that contact network structure can have a
-substantial impact on transmission tree shape \autocite{o2010contact,
-robinson2013dynamics, leventhal2012inferring, colijn2014phylogenetic,
-villandre2016assessment}.
+Although phylodynamic methods have been developed to fit a wide variety of
+epidemiological models to phylogenetic data~\autocite{volz2012complex,
+rasmussen2014phylodynamic}, our method is the first (to our knowledge) which
+can fit models that do not assume panmixia. We believe this capability will be
+of broad interest to the molecular evolution and epidemiology community, as it
+widens the field of epidemiological parameters which may be investigated
+through viral sequence data. In addition, the characterization of local contact
+networks could be valuable from a public health perspective, such as for
+investigating optimal vaccination strategies. This information could assist in
+curtailing current epimemics, as well as preventing future epidemics of
+different diseases over the same contact network.

### 2016-04-20 doc/thesis/objective.tex

@@
-\section{Objective}
-
 The spread of a disease is most often modelled by assuming either a
@@
 homogeneously mixed groups~\autocite{rushton1955deterministic}. This
-assumption, also called the ``law of mass action'', implies that any two
-individuals in the same compartment are equally likely to come into contact
-causing transmission. Although this provides a reasonable approximation in many
-cases, the error introduced by assuming a panmictic population can be
-substantial when significant contact heterogeneity exists in the underlying
-population. Contact network models provide an alternative to compartmental
-models which do not require the assumption of panmixia. In addition to more
-accurate predictions, the parameters of the networks themselves may be of
-interest from a public health perspective. For example, certain vaccination
-strategies may be more or less effective in curtailing an epidemic depending on
-the underlying network's degree distribution. Phylodynamic methods have been
-used to fit many different types of model to phylogenetic data, but as far as
-we know, no methods have yet been developed to fit contact network models. The
-primary objective of this work is to develop such a method.
+assumption, also called the ``law of mass action'' or \defn{panmixia}, implies
+that any two individuals in the same compartment are equally likely to come
+into contact causing transmission. Although this provides a reasonable
+approximation in many cases, the error introduced by assuming a panmictic
+population can be substantial when significant contact heterogeneity exists in
+the underlying population. Contact network models provide an alternative to
+compartmental models which do not require the assumption of panmixia. In
+addition to more accurate predictions, the parameters of the networks
+themselves may be of interest from a public health perspective. For example,
+certain vaccination strategies may be more or less effective in curtailing an
+epidemic depending on the underlying network's degree distribution.
+Phylodynamic methods have been used to fit many different types of model to
+phylogenetic data, but as far as we know, no methods have yet been developed to
+fit contact network models. The primary objective of this work is to develop
+such a method.
 

### 2016-04-20 programs/netabc/R/plots.R
Added posterior plot function

@@ -1,4 +1,4 @@
-#' Plot a 2-dimensional PCA projection of a kernel-matrix 
+#' Plot a 2-dimensional PCA projection of a kernel-matrix
 #'
 #' @param kmat kernel matrix
 #' @param color either NULL, or a list of one element to display with color
@@ -101,7 +101,7 @@ cluster.plot <- function (net, tree, yaml="", palette="Set1", status.only=FALSE,
     }
     d$cluster[bg] <- 1
     cluster.colors <- c("#bdbdbd", brewer.pal(max(d$cluster)-1, palette))
-    
+
     labels <- c(tree$tip.label, tree$node.label)
     parents <- labels[tree$edge[,1]]
     children <- labels[tree$edge[,2]]
@@ -152,37 +152,37 @@ edge.color.plot <- function (tree, labels, yaml="", palette="PuBu", parent=TRUE,
 #' @param truth true parameter values to indicate on plots
 #' @param limits axis limits for plots
 #' @export
-marginal.plot <- function (d, truth, limits) {                                   
-    vary.cols <- colnames(d)[apply(d, 2, function (x) length(unique(x)) > 1)]       
-                                                                                 
-    if (length(vary.cols) >= 2) {                                                
-    # 2D marginals                                                               
-        combos <- combn(vary.cols, 2)                                            
-        plot.data <- apply(combos, 2, function (c) d[, c, with=FALSE])           
-        plot.aes <- apply(combos, 2, function (c) aes_string(x=c[1], y=c[2]))       
-        plot.limits <- apply(combos, 2, function (c) limits[c])                  
-        plots <- mapply(ggplot, plot.data, plot.aes, SIMPLIFY=FALSE)             
-        plots <- mapply(function (p, lim) {                                      
+marginal.plot <- function (d, truth, limits) {
+    vary.cols <- colnames(d)[apply(d, 2, function (x) length(unique(x)) > 1)]
+
+    if (length(vary.cols) >= 2) {
+    # 2D marginals
+        combos <- combn(vary.cols, 2)
+        plot.data <- apply(combos, 2, function (c) d[, c, with=FALSE])
+        plot.aes <- apply(combos, 2, function (c) aes_string(x=c[1], y=c[2]))
+        plot.limits <- apply(combos, 2, function (c) limits[c])
[…]

### 2016-04-22 doc/tables/realdata-bc.tex
Updated analyses...

--- 2016-03-03 doc/tables/realdata.tex
@@
   \hline
-  \autocite{wang2015targeting} & 173 & Beijing, China & MSM & \textit{pol} \\
-  \autocite{cuevas2009hiv} & 287 & Basque Country, Spain & mixed & \textit{pol} \\
-  \autocite{novitsky2013phylogenetic} & \multirow{2}{*}{180} &
-  \multirow{2}{*}{Mochudi, Botswana} & \multirow{2}{*}{mixed} &
-  \multirow{2}{*}{\textit{env}} \\ \autocite{novitsky2014impact} \\
-  \autocite{li2015hiv} & 280 & Shanghai, China & MSM & \textit{pol} \\
-  \autocite{niculescu2015recent} & 136 & Romaina & IDU & \textit{pol} \\
+  \textcite{wang2015targeting} & 173 & Beijing, China & MSM & \textit{pol} \\
+  \textcite{cuevas2009hiv} & 287 & Basque Country, Spain & mixed & \textit{pol} \\
+  \textcite{novitsky2013phylogenetic} & \multirow{2}{*}{180} &
+  \multirow{2}{*}{Mochudi, Botswana} & \multirow{2}{*}{HET} &
+  \multirow{2}{*}{\textit{env}} \\ \textcite{novitsky2014impact} \\
+  \textcite{li2015hiv} & 280 & Shanghai, China & MSM & \textit{pol} \\
+  \textcite{niculescu2015recent} & 136 & Romania & IDU & \textit{pol} \\
+  unpublished & 399 & British Columbia, Canada & IDU & \textit{pol} \\
   \hline

### 2016-04-22 doc/paper/paper.Rnw
Stuff to work on for dynamics

@@
-\documentclass[nogrid]{MBE}
-\usepackage{url}
-
-% Mol Biol Evol stuff
-\jshort{mst}
-\volname{}
-\jvolume{0}
-\jvol{}
-\jissue{0}
-\pubyear{2016}
-\mstype{Article}
-\artid{???}
-\access{???}
-
+\documentclass[12pt]{article}
+\usepackage[T1]{fontenc}
+\usepackage{garamondx}
+\usepackage[garamondx,cmbraces]{newtxmath}
+\usepackage[backend=biber]{biblatex}
+\usepackage{fullpage}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{setspace}
+\usepackage{cleveref}
+
+\addbibresource{papers.bib}
 \begin{document}
-
-\title[Contact networks and ABC]{Phylodynamic inference of contact network
-parameters through approximate Bayesian computation}
-\author[McCloskey, Liang, and Poon]{Rosemary M. McCloskey$^{*,1}$, 
-    Richard H. Liang$^1$, and Art F.Y. Poon$^{1,2}$}
-\address{$^1$BC Centre for Excellence in HIV/AIDS, 608-1081 Burrard Street, Vancouver, Canada \\
-$^2$Department of Medicine, University of British Columbia, Vancouver, Canada}
-\history{Submitted 5 April 2016}
-\coresp{E-mail: rmccloskey@cfenet.ubc.ca.}
-\editor{}
+\onehalfspacing
+
+\newcommand{\tablepath}{../tables}
+\graphicspath{{../figures/}}
+
+\title{Reconstructing contact network parameters from viral phylogenies}
+\author{Rosemary M. McCloskey, Richard H. Liang, and Art F.Y. Poon}
 \abstract{\input{abstract}}
-\keyword{Phylogenetics, phylodynamics, approximate Bayesian computation,
-contact network, transmission tree, human immunodeficiency virus.}
 
@@
 mode of transmission of the disease. For airborne pathogens, physical proximity
-may be sufficient, while for sexually transmitted diseases, \newpage\noindent %hack
-sexual or in some cases blood-to-blood contact is required. The population
-together with the set of links between individuals along which transmission can
-occur is called the contact network~\citep{klovdahl1985social,
-morris1993epidemiology}. The structure of the contact network underlying an
-epidemic can profoundly impact the speed and pattern of the epidemic's
-expansion. Network structure can influence the prevalence
-trajectory~\citep{o2011contact} and epidemic
-threshold~\citep{barthelemy2005dynamical}, in turn affecting the estimates of
-quantities such as effective population size~\citep{goodreau2006assessing}.
+may be sufficient, while for sexually transmitted diseases, sexual or in some
+cases blood-to-blood contact is required. The population together with the set
+of links between individuals along which transmission can occur is called the
+contact network~\autocite{klovdahl1985social, morris1993epidemiology}. The
+structure of the contact network underlying an epidemic can profoundly impact
+the speed and pattern of the epidemic's expansion. Network structure can
+influence the prevalence trajectory~\autocite{o2011contact} and epidemic
+threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates of
+quantities such as effective population size~\autocite{goodreau2006assessing}.
 From a public health perspective, contact networks have been explored as tools
 for curtailing epidemic spread, by way of interventions targeted to
-well-connected nodes~\citep{wang2015targeting}. True contact networks are a
+well-connected nodes~\autocite{wang2015targeting}. True contact networks are a
 challenging type of data to collect, requiring extensive epidemiological
-investigation~\citep{welch2011statistical}.
+investigation~\autocite{welch2011statistical}.
 
@@
 thereby shaping the inter-host viral
-phylogeny~\citep{drummond2003measurably}. The term ``phylodynamics'' was
+phylogeny~\autocite{drummond2003measurably}. The term ``phylodynamics'' was
 coined to describe this interaction, as well as the growing family of inference
 methods to estimate epidemiological parameters from viral
-phylogenies~\citep{grenfell2004unifying}. These methods have revealed
+phylogenies~\autocite{grenfell2004unifying}. These methods have revealed
 diverse properties of local viral outbreaks, from basic reproductive
-number~\citep{stadler2012estimating}, to the degree of
-clustering~\citep{hughes2009molecular}, to the elevated transmission risk
-during acute infection~\citep{volz2012simple}. On the other hand, although
+number~\autocite{stadler2012estimating}, to the degree of
+clustering~\autocite{hughes2009molecular}, to the elevated transmission risk
+during acute infection~\autocite{volz2012simple}. On the other hand, although
 sophisticated methods have been developed for fitting complex population
-genetic models to phylogenies~\citep{rasmussen2014phylodynamic}, inference
+genetic models to phylogenies~\autocite{rasmussen2014phylodynamic}, inference
 of structural network parameters has to date been limited. However, it has been
 shown that network structure has a tangible impact on phylogeny
-shape~\citep{leventhal2012inferring, colijn2014phylogenetic,
+shape~\autocite{leventhal2012inferring, colijn2014phylogenetic,
 goodreau2006assessing, robinson2013dynamics, villandre2016assessment},
 suggesting that such statistical inference might be
-possible~\citep{welch2011statistical}.
-
-Survey-based studies of sexual networks~\citep{liljeros2001web,
+possible~\autocite{welch2011statistical}.
+
+Survey-based studies of sexual networks~\autocite{liljeros2001web,
 schneeberger2004scale, colgate1989risk} have found that these networks tend to
-have a degree distribution which follows a power law \citep[although there has
+have a degree distribution which follows a power law \autocite[although there has
 been some disagreement, see][]{handcock2004likelihood}. That is, the number of
 nodes of degree $k$ is proportional to $k^{-\gamma}$ for some constant
-$\gamma$. These networks are also referred to as ``scale-free''. One process by
-which scale-free networks can be generated is preferential attachment, where
+$\gamma$. These networks are also creferred to as ``scale-free''. One process by
+which scale-free networks can be generated is pcreferential attachment, where
 nodes with a high number of contacts attract new connections at an elevated
-rate. The first contact network model incorporating preferential attachment was
-introduced by \citet{barabasi1999emergence}, and is now referred to as the
+rate. The first contact network model incorporating pcreferential attachment was
+introduced by \textcite{barabasi1999emergence}, and is now creferred to as the
 Barab\'asi-Albert (BA) model. Under this model, networks are formed by
@@
 Previous work offers precedent for the possibility of statistical inference of
-structural network parameters. \citet{britton2002bayesian} develop a Bayesian
+structural network parameters. \textcite{britton2002bayesian} develop a Bayesian
 approach to estimate the edge density in an Erd\H{o}s-R\'enyi
-network~\citep{erdos1960evolution} given observed infection dates, and
+network~\autocite{erdos1960evolution} given observed infection dates, and
 optionally recovery dates. Their approach was later extended by
-\citet{groendyke2011bayesian} and applied to a much larger data set of 188
-individuals. \citet{volz2007susceptible, volz2008sir} developed differential
+\textcite{groendyke2011bayesian} and applied to a much larger data set of 188
+individuals. \textcite{volz2007susceptible, volz2008sir} developed differential
 equations describing the spread of a susceptible-infected (SI) epidemic on
@@
 could in principle be used for inference if observed incidence trajectories
-were available. \citet{brown2011transmission} analysed the degree distribution
+were available. \textcite{brown2011transmission} analysed the degree distribution
 of an approximate transmission network, estimated based on genetic similarity
@@
 already led to a new infection. The authors found that a Waring distribution,
-which is produced by a more sophisticated preferential attachment model, was a
+which is produced by a more sophisticated pcreferential attachment model, was a
 good fit to their estimated network.
@@
 function, $\hat{D}$ is a simulated dataset according to $\theta$, and
-$\varepsilon$ is a small tolerance~\citep{sunnaaker2013approximate}. In the
+$\varepsilon$ is a small tolerance~\autocite{sunnaaker2013approximate}. In the
 specific case when $\rho$ is a kernel function, the approach is known as
-kernel-ABC~\citep{nakagome2013kernel, poon2015phylodynamic}.
+kernel-ABC~\autocite{nakagome2013kernel, poon2015phylodynamic}.
 
@@
 contact network models from observed phylogenetic data. The distance function
-we use is the tree kernel developed by \citet{poon2013mapping}, which computes
+we use is the tree kernel developed by \textcite{poon2013mapping}, which computes
 a weighted dot product of the trees' representations in the space of all
@@
 that some network parameters can be inferred with reasonable accuracy, while
-others have a minimal detectable impact on tree shape and therefore cannot be
+others have a minimal detectable impact on tree shape and thecrefore cannot be
 estimated accurately. We also find that these parameters can vary considerably
@@
 
-\section{New Approaches}
+\section{Methods}
+
+\subsection{\textit{Netabc}: phylogenetic inference of contact network
+parameters with kernel-ABC}
 
@@
 viral phylogeny. We implemented the adaptive sequential Monte Carlo (SMC)
-algorithm for ABC developed by \citet{del2012adaptive}. The SMC algorithm keeps
-track of a population of parameter ``particles'', which are initially sampled
-from the parameters' joint prior distribution. Several datasets are simulated
-under the model of interest for each of the particles. In this case, the
-datasets are transmission trees, which are generated by a two-step process.
+algorithm for ABC developed by \textcite{del2012adaptive}. The SMC algorithm
+keeps track of a population of parameter ``particles'', which are initially
+sampled from the parameters' joint prior distribution. Several datasets are
+simulated under the model of interest for each of the particles. In this case,
+the datasets are transmission trees, which are generated by a two-step process.
 First, a contact network is simulated according to the network model being fit.
 Second, a transmission tree is simulated over that network with a Gillespie
-simulation algorithm~\citep{gillespie1976general}, in the same fashion as
-several previous studies \citep[\textit{e.g.}][]{robinson2013dynamics,
+simulation algorithm~\autocite{gillespie1976general}, in the same fashion as
+several previous studies \autocite[\textit{e.g.}][]{robinson2013dynamics,
 leventhal2012inferring}. The particles are weighted according to the similarity
 between their associated simulated trees and the observed tree. To quantify
-this similarity, we used the tree kernel developed by \citet{poon2013mapping}.
-Particles are iteratively perturbed by applying a Metropolis-Hastings kernel
-and, if the move is accepted, simulating new datasets under the new parameters.
-When a particle's weight drops to zero, because its simulated trees are too
-dissimilar to the observed tree, the particle is dropped from the population,
-and eventually replaced by a resampled particle with a higher weight. As the
-algorithm progresses, the population converges to a Monte Carlo approximation
-of the ABC target distribution, which is assumed to approximate the desired
-posterior~\citep{del2012adaptive, sunnaaker2013approximate}.
-
-A computer program implementing our method is freely available at
-\url{https://github.com/rmcclosk/netabc} (last accessed April 3, 2016).
+this similarity, we used the tree kernel developed by
+\textcite{poon2013mapping}. Particles are iteratively perturbed by applying a
+Metropolis-Hastings kernel and, if the move is accepted, simulating new
+datasets under the new parameters. When a particle's weight drops to zero,
+because its simulated trees are too dissimilar to the observed tree, the
+particle is dropped from the population, and eventually replaced by a resampled
+particle with a higher weight. As the algorithm progresses, the population
+converges to a Monte Carlo approximation of the ABC target distribution, which
+is assumed to approximate the desired posterior~\autocite{del2012adaptive,
+sunnaaker2013approximate}. A computer program implementing our method is freely
+available at \url{https://github.com/rmcclosk/netabc} (last accessed April 3,
+2016).
+
+To check that our implementation of Gillespie simulation was correct, we
+reproduced Figure 1A of \textcite{leventhal2012inferring} (our
+\cref{fig:leventhal}), which plots the unbalancedness of transmission trees
+simulated over four network models at various levels of pathogen
+transmissibility. Our implementation of adaptive ABC-SMC was tested by applying
+it to the same mixture of Gaussians used by \citeauthor{del2012adaptive} to
+demonstrate their method (originally used by~\textcite{sisson2007sequential}).
+We were able to obtain a close approximation to the function (see
+\cref{fig:smctest}), and attained the stopping condition used by the authors in
+a comparable number of steps.
+
+Nodes in our networks followed simple SI dynamics, meaning that they became
+infected at a rate proportional to their number of infected neighbours, and
+never recovered. For all analyses, the transmission trees' branch lengths were
+scaled by dividing by their mean. We used the \textit{igraph} library's
+implementation of the BA model~\autocite{csardi2006igraph} to generate the
+graphs. The analyses were run on Westgrid (\url{https://www.westgrid.ca/}) and
+a local computer cluster.
+
+\subsection{Kernel classifiers}
+
+We used the phylogenetic kernel developed by \textcite{poon2013mapping} to test
+whether the parameters of the BA model had an effect on tree shape. 100
+networks were simulated under each of three different values of $\alpha$: 0.5,
+1.0, and 1.5 (300 networks total). The other parameters were fixed to the
+following values: $N$ = 5000, $I$ = 1000, and $m$ = 2. A transmission tree with
+500 tips was simulated over each network (300 transmission trees total). The
+300 trees were compared pairwise with the tree kernel to form a $300 \times
+300$ kernel matrix. The kernel meta-parameters $\lambda$ (the ``decay
+factor''), and $\sigma$ (the ``radial basis function
+variance'')~\autocite[see][]{poon2013mapping}, were set to 0.3 and 4 respectively.
+We constructed a kSVR classifier for $\alpha$ using the \textit{kernlab}
+package~\autocite{zeileis2004kernlab}, and evaluated its accuracy with 1000
+two-fold cross-validations.
+
+Three similar experiments were performed for the other BA model parameters (one
+experiment per parameter). $m$ was varied between 2, 3, and 4; $I$ between 500,
+1000, and 2000; and $N$ between 3000, 5000, and 8000. The parameters not being
+tested were fixed at the values $N$ = 5000, $I$ = 1000, $m$ = 2, and $\alpha$ =
+1. Thus, we performed a total of four kSVR cross-validations, one for each of
+the BA model parameters $\alpha$, $I$, $m$, and $N$. We repeated these four
+cross-validations with different values of $\lambda$ (0.2, 0.3, and 0.4) and
+$\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$), as well as on trees with
+differing numbers of tips (100, 500, and 1000) and in epidemics of differing
+size (500, 1000, and 2000). The combination of the number of sampled
+individuals (\textit{i.e.} the number of tips) and the epidemic size
+(\textit{i.e.} $I$) will be creferred to as an ``epidemic scenario''. When
+evaluating the classifier for $I$, we did not consider trees with 1000 tips,
+because one of the tested $I$ values was 500, and the number of tips cannot be
+larger than $I$.
+
+For each of the four parameters, we also tested a linear regression against
+Sackin's index~\autocite{shao1990tree} and an ordinary SVR based on the normalized
+lineages-through-time (nLTT) statistic~\autocite{janzen2015approximate}.
+
+\subsection{ABC simulations}
+
+We simulated three transmission trees, each with 500 tips, under every element
+of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
+\{1000, 2000\}, $m$ = \{2, 3, 4\}, and $\alpha$ = \{0.0, 0.5, 1,
+1.5\}. This produced a total of 24 parameter combinations $\times$ three trees
+per combination = 72 trees total. The adaptive ABC algorithm was applied to
+each tree with these priors: $m \sim$ DiscreteUniform(1, 5), $\alpha \sim$
+Uniform(0, 2), and $(N, I)$ jointly uniform on the region \{$500 \leq N \leq
+15000$, $500 \leq I \leq 5000$, $I \leq N$\}. Proposals for $\alpha$, $N$, and
+$I$ were Gaussian, while proposals for $m$ were Poisson. Following
+\textcite{del2012adaptive} and \textcite{beaumont2009adaptive}, the variance of
+all proposals was equal to the empirical variance of the particles.
+
+The algorithm was run with 1000 particles, 5 simulated datasets per particle,
+and the ``quality'' parameter controlling the decay rate of the tolerance
+$\varepsilon$ set to 0.95. We used the same stopping criterion as
+\citeauthor{del2012adaptive}, namely when the MCMC acceptance rate dropped
+below 1.5\%. Point estimates for the parameters were obtained by taking the
+highest point of an estimated kernel density on the final set of particles,
+calculated using the \textit{density} function with the default parameters in
+\textit{R}. Highest posterior density (HPD) intervals were calculated with the
+\textit{HPDinterval} function from the \textit{R} package
+\textit{coda}~\autocite{plummer2006coda}.
+
+Two further analyses were performed to address potential sources of error. To
+evaluate the effect of model misspecification in the case of heterogeneity
+among nodes, we generated a network where half the nodes were attached with
+power $\alpha$ = 0.5, and the other half with power $\alpha$ = 1.5. The other
+parameters for this network were $N$ = 5000, $I$ = 1000, and $m$ = 2. To
+investigate the effects of potential sampling bias, we simulated a transmission
+tree where the tips were sampled in a peer-driven fashion, rather than at
+random. That is, the probability to sample a node was twice as high if any of
+that node's network peers had already been sampled. The parameters of this
+network were $N$ = 5000, $I$ = 2000, $m$ = 2, and $\alpha$ = 0.5.
+
+\subsection{Investigation of published data}
+
+We applied our kernel-ABC method to several published HIV datasets. Because the
+BA model generates networks with a single connected component, we
+specifically searched for datasets which originated from existing clusters,
+either phylogenetically or geographically defined. Characteristics of the
+datasets we investigated are given in \cref{tab:data}.
+
+\begin{table*}[!t]
+  \centering
+  \input{\tablepath/realdata}
+  \caption{Characteristics of published datasets investigated with kernel-ABC.
+  Acronyms: MSM, men who have sex with men; IDU, injection drug users.}
+  \label{tab:data}
+\end{table*}
+
+We downloaded all sequences associated with each study from GenBank. For the
+\textcite{novitsky2014impact} data, each sequence was aligned pairwise to the HXB2
+creference sequence (Genbank accession number K03455) and the hypervariable
+regions were clipped out with \textit{BioPython} version
+1.66+~\autocite{cock2009biopython}. Sequences were multiply aligned using
+\textit{MUSCLE} version 3.8.31 \autocite{edgar2004muscle}, and alignments were
+manually inspected with \textit{Seaview} version 4.4.2
+\autocite{gouy2010seaview}. Phylogenies were constructed from the nucleotide
+alignments by approximate maximum likelihood using \textit{FastTree2} version
+2.1.7 with the generalized time-reversible model. Transmission trees were
+estimated by rooting and time-scaling the phylogenies by root-to-tip
+regression, using a modified version of Path-O-Gen (distributed as part of
+BEAST~\autocite{drummond2007beast}) as described
+previously~\autocite{poon2015phylodynamic}.
+
+Two of the datasets \autocite{li2015hiv,novitsky2014impact} were initially much
+larger than the others, containing 1265 and 1299 sequences respectively. To
+ensure that the analyses were comparable, we reduced these to a number of
+sequences similar to the smaller datasets. For the \textcite{li2015hiv} data,
+we detected a cluster of size 280 using a patristic distance cutoff of 0.02 as
+described previously~\autocite{poon2015impact}. Only sequences within this
+cluster were carried forward. For the \textcite{novitsky2014impact} data, no
+large clusters were detected using the same cutoff, so we analysed a subtree of
+size 180 chosen arbitrarily.
+
+For all datasets, we used the priors $\alpha$ $\sim$ Uniform(0, 2) and $N$ and
+$I$ jointly uniform on the region \{$n \leq N \leq 10000$, $n \leq I \leq
+10000$, $I \leq N$\}, where $n$ is the number of tips in the tree. Since the
+value $m = 1$ produces networks with no cycles, which we considered fairly
+implausible, we ran one analysis with the prior $m \sim$ DiscreteUniform(1, 5),
+and one with the prior $m \sim$ DiscreteUniform(2, 5). The other parameters to
+the SMC algorithm were the same as used for the simulation experiments.
 
@@
 We investigated the parameters of the BA network
-model~\citep{barabasi1999emergence}. In addition to $m$ and $\alpha$ (see
+model~\autocite{barabasi1999emergence}. In addition to $m$ and $\alpha$ (see
 Introduction), we considered $N$, which denotes the total number of nodes in
@@
 parameter values, calculated pairwise tree kernel scores between them, and
-attempted to classify the trees using a kernel support vector machine (kSVM).
-We also tested classifiers based on Sackin's index~\citep{shao1990tree} and the
-normalized lineages-through-time (nLTT)
-statistic~\citep{janzen2015approximate}. Accuracy of the kSVMs varied based on
-the parameter being tested (Figure~\ref{fig:rsquared}, left). Classifiers based
-on two other tree statistics, the nLTT and Sackin's index, generally exhibited
+attempted to classify the trees using a kernel support vector machine (kSVR).
+We also tested classifiers based on Sackin's index~\autocite{shao1990tree} and
+the normalized lineages-through-time (nLTT)
+statistic~\autocite{janzen2015approximate}. Accuracy of the kSVRs varied based
+on the parameter being tested (\cref{fig:rsquared}, left). Classifiers based on
+two other tree statistics, the nLTT and Sackin's index, generally exhibited
 worse performance than the tree kernel, although the magnitude of the disparity
-varied between the parameters (Figure~\ref{fig:rsquared}, centre and right).
-The results were largely robust to variations in the tree kernel
-meta-parameters $\lambda$ and $\sigma$ (Figures~S3-S6).
-
-When classifying $\alpha$, the kernel-SVM classifier had an average $R^2$ of 
+varied between the parameters (\cref{fig:rsquared}, centre and right). The
+results were largely robust to variations in the tree kernel meta-parameters
+$\lambda$ and $\sigma$ (\cref{fig:alphacrossv,fig:mcrossv,fig:Ncrossv,fig:Icrossv}).
+
+When classifying $\alpha$, the kernel-SVR classifier had an average $R^2$ of 
     \Sexpr{kernel.alpha[,round(mean(rsquared), 2)]},
@@
     \Sexpr{nltt.alpha[,round(mean(rsquared), 2)]}
-for the nLTT-based SVM, and
+for the nLTT-based SVR, and
     \Sexpr{sackin.alpha[,round(mean(rsquared), 2)]}
@@
 $R^2$ values of 
-  \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for kSVM,
+  \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for kSVR,
   \Sexpr{nltt.m[,round(mean(rsquared), 2)]} for the nLTT, and
@@
 for Sackin's index. Again, there was little variation in accuracy between
-epidemic scenarios, although the accuracy of the kSVM was slightly higher
-on 1000-tip trees.
+epidemic scenarios, although the accuracy of the kSVR was slightly higher
+on 1000-tip trees (\cref{fig:rsquared}, left).
 
@@
 Finally, the performance of classifiers for $N$ depended heavily on the
-epidemic scenario. The $R^2$ of the kSVM classifier ranged from
+epidemic scenario. The $R^2$ of the kSVR classifier ranged from
   \Sexpr{kernel.N[,round(min(rsquared), 2)]}
@@
   \Sexpr{kernel.N[,round(max(rsquared), 2)]}
-for the largest. Likewise, $R^2$ for the nLTT-based SVM ranged from 
+for the largest. Likewise, $R^2$ for the nLTT-based SVR ranged from 
   \Sexpr{nltt.N[,round(min(rsquared), 2)]}
@@
   \caption{
-      Cross-validation accuracy of kernel-SVM classifier (left), SVM classifier
+      Cross-validation accuracy of kernel-SVR classifier (left), SVR classifier
       using nLTT (centre), and linear regression using Sackin's index
@@
 
-<<point_est, include=FALSE>>=
-    f <- Sys.glob("../../simulations/abc-pa-free-m/point-estimate/*")
-    d <- fread(f)
-    d[,m := floor(m)]
-    d[,alpha_error := abs(true_alpha - alpha)]
-    d[,N_error := abs(true_N - N)]
-    d[,I_error := abs(true_I - I)]
-    d[,m_error := abs(true_m - m)]
-
-    alpha.av <- anova(lm(alpha_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-    m.av <- anova(lm(m_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-    N.av <- anova(lm(N_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-    I.av <- anova(lm(I_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-
-    m.tbl <- d[,prop.table(table(m_error))]
-    zero.alpha.test <- wilcox.test(d[true_alpha == 0, alpha_error], 
-                                   d[true_alpha != 0, alpha_error])
-    alpha.I.test <- wilcox.test(d[true_alpha < 1, I_error],
-                                d[true_alpha >= 1, I_error])
-    I.I.test <- wilcox.test(d[true_I == 1000, I_error],
-                            d[true_I == 2000, I_error])
-    alpha.m.test <- wilcox.test(d[true_alpha == 0 | true_alpha == 1, m_error],
-                                d[true_alpha == 0.5 | true_alpha == 1.5, m_error])
-    stopifnot(alpha.av["factor(true_m)", "Pr(>F)"] > 0.05)
-    stopifnot(alpha.av["factor(true_I)", "Pr(>F)"] > 0.05)
-    stopifnot(I.I.test$p.value < 1e-5)
-    stopifnot(I.av["factor(true_m)", "Pr(>F)"] > 0.05)
-    stopifnot(m.av["factor(true_m)", "Pr(>F)"] > 0.05)
-    stopifnot(m.av["factor(true_I)", "Pr(>F)"] > 0.05)
-    stopifnot(min(N.av[1:3,"Pr(>F)"]) > 0.05)
-@
-
-Figure~\ref{fig:abcpt} shows maximum \textit{a posteriori} (MAP) point
+%<<point_est, include=FALSE>>=
+%    f <- Sys.glob("../../simulations/abc-pa-free-m/point-estimate/*")
+%    d <- fread(f)
+%    d[,m := floor(m)]
+%    d[,alpha_error := abs(true_alpha - alpha)]
+%    d[,N_error := abs(true_N - N)]
+%    d[,I_error := abs(true_I - I)]
+%    d[,m_error := abs(true_m - m)]
+%
+%    alpha.av <- anova(lm(alpha_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
+%    m.av <- anova(lm(m_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
+%    N.av <- anova(lm(N_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
+%    I.av <- anova(lm(I_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
+%
+%    m.tbl <- d[,prop.table(table(m_error))]
+%    zero.alpha.test <- wilcox.test(d[true_alpha == 0, alpha_error], 
+%                                   d[true_alpha != 0, alpha_error])
+%    alpha.I.test <- wilcox.test(d[true_alpha < 1, I_error],
+%                                d[true_alpha >= 1, I_error])
+%    I.I.test <- wilcox.test(d[true_I == 1000, I_error],
+%                            d[true_I == 2000, I_error])
+%    alpha.m.test <- wilcox.test(d[true_alpha == 0 | true_alpha == 1, m_error],
+%                                d[true_alpha == 0.5 | true_alpha == 1.5, m_error])
+%    stopifnot(alpha.av["factor(true_m)", "Pr(>F)"] > 0.05)
+%    stopifnot(alpha.av["factor(true_I)", "Pr(>F)"] > 0.05)
+%    stopifnot(I.I.test$p.value < 1e-5)
+%    stopifnot(I.av["factor(true_m)", "Pr(>F)"] > 0.05)
+%    stopifnot(m.av["factor(true_m)", "Pr(>F)"] > 0.05)
+%    stopifnot(m.av["factor(true_I)", "Pr(>F)"] > 0.05)
+%    stopifnot(min(N.av[1:3,"Pr(>F)"]) > 0.05)
+%@
+
+\Cref{fig:abcpt} shows maximum \textit{a posteriori} (MAP) point
 estimates of the BA model parameters obtained with kernel-ABC on simulated
@@
 parameter was set to 2, however the results for $m = 3$ and $m = 4$ were
-similar (Figures S7 and S8). Average boundaries of 95\% HPD intervals are given
-in Table~\ref{tab:abchpd}.
+similar (\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% HPD
+intervals are given in \cref{tab:abchpd}.
 
 The accuracy of the parameter estimates obtained with kernel-ABC
-paralleled the results from the kSVM classifier. Of the four parameters,
+paralleled the results from the kSVR classifier. Of the four parameters,
 $\alpha$ was the most accurately estimated, with point estimates having a
 median [IQR] absolute error of 
-    \Sexpr{d[,round(median(alpha_error), 2)]} 
-    [\Sexpr{d[,round(quantile(alpha_error, 0.25), 2)]} - 
-    \Sexpr{d[,round(quantile(alpha_error, 0.75), 2)]}].
+%    \Sexpr{d[,round(median(alpha_error), 2)]} 
+%    [\Sexpr{d[,round(quantile(alpha_error, 0.25), 2)]} - 
+%    \Sexpr{d[,round(quantile(alpha_error, 0.75), 2)]}].
 The errors when the true value of $\alpha$ was zero were significantly greater
 than those for the other values 
-    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(zero.alpha.test$p.value, 4))}$),
+%    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(zero.alpha.test$p.value, 4))}$),
 but did not vary across the true values of the other parameters (one-way
@@
 errors of
-    \Sexpr{d[,round(median(I_error))]} 
-    [\Sexpr{d[,round(quantile(I_error, 0.25))]} - 
-    \Sexpr{d[,round(quantile(I_error, 0.75))]}].
+%    \Sexpr{d[,round(median(I_error))]} 
+%    [\Sexpr{d[,round(quantile(I_error, 0.25))]} - 
+%    \Sexpr{d[,round(quantile(I_error, 0.75))]}].
 These errors were significantly higher when the true value of $\alpha$ was
 at least 1
-    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(alpha.I.test$p.value, 4))}$)
+%    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(alpha.I.test$p.value, 4))}$)
 and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
@@
 The $m$ parameter was estimated correctly in only
-    \Sexpr{as.integer(m.tbl[1] * 100)} \%
+%    \Sexpr{as.integer(m.tbl[1] * 100)} \%
 of simulations. Oddly, the error in the estimated $m$ was higher for integer
-values of $\alpha$ (\text{i.e.}~0 and 1) than non-integer values 
-    (Wilcoxon rank-sum test, $p$ = \Sexpr{round(alpha.m.test$p.value, 3)}).
+values of $\alpha$ (\textit{i.e.}~0 and 1) than non-integer values 
+%    (Wilcoxon rank-sum test, $p$ = \Sexpr{round(alpha.m.test$p.value, 3)}).
 The true values of the other parameters did not significantly affect the
@@
 was consistently over-estimated by about a factor of two
-    (error \Sexpr{d[,round(median(N_error))]} 
-    [\Sexpr{d[,round(quantile(N_error, 0.25))]} - 
-     \Sexpr{d[,round(quantile(N_error, 0.75))]}]).
+%    (error \Sexpr{d[,round(median(N_error))]} 
+%    [\Sexpr{d[,round(quantile(N_error, 0.25))]} - 
+%     \Sexpr{d[,round(quantile(N_error, 0.75))]}]).
 No parameters influenced the accuracy of the $N$ estimates (all one-way ANOVA).
@@
   \centering
-    \begin{tabular}{lr>{\raggedleft\arraybackslash}p{2.5cm}>{\raggedleft\arraybackslash}p{2.5cm}>{\raggedleft\arraybackslash}p{2.5cm}}
-      \hline
-    Parameter & True value & Mean point estimate & Mean HPD lower bound & Mean HPD upper bound \\ 
-      \hline
-    $\alpha$ & 0.0 & 0.24 & 0.02 & 0.73 \\ 
-       & 0.5 & 0.42 & 0.02 & 0.81 \\ 
-       & 1.0 & 0.97 & 0.61 & 1.11 \\ 
-       & 1.5 & 1.48 & 1.26 & 1.83 \\ 
-      $I$ & 1000 & 1155.68 & 598.68 & 2402.84 \\ 
-       & 2000 & 2646.07 & 1182.31 & 4058.13 \\ 
-      $m$ & 2 & 2.92 & 1.75 & 4.92 \\ 
-       & 3 & 3.33 & 1.96 & 4.92 \\ 
-       & 4 & 3.62 & 1.88 & 5.00 \\ 
-      $N$ & 5000 & 10962.61 & 2732.55 & 14701.87 \\ 
-       \hline
-    \end{tabular}
+  \input{\tablepath/abc-hpd.tex}
   \caption{
@@
 the parameters, with narrower HPD intervals for the parameters with more
-accurate point estimates (Table~\ref{tab:abchpd}). Figure~\ref{fig:abcex} shows
+accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows
 the distributions for for one simulation (equivalent plots for all the
-simulations can be found in Supplementary Figure S10). HPD intervals around $\alpha$
-and $I$ were narrow relative to the region of nonzero prior density, whereas
-the intervals for $m$ and $N$ were more widely dispersed.
+simulations can be found in Supplementary Figures TODO). HPD intervals around
+$\alpha$ and $I$ were narrow relative to the region of nonzero prior density,
+whereas the intervals for $m$ and $N$ were more widely dispersed.
 
@@
   params <- c("alpha", "m", "N", "I")
-  setnames(d, c("attach_power", "edges_per_vertex", "nodes", "sim_nodes"),
-           params)
   d <- melt(d, measure.vars=params, variable.name="parameter")
@@
 
-  p <- fread("bzcat ../../simulations/abc-pa-peerdriven/abc/*")
+  m <- collect.metadata("../../simulations/abc-pa-peerdriven/abc/*")
+  p <- fread(paste("bzcat", rownames(m)[m$sample_peer == 1]))
   p <- p[iter == max(iter)]
@@
   params <- c("alpha", "m", "N", "I")
-  setnames(p, c("attach_power", "edges_per_vertex", "nodes", "sim_nodes"),
-           params)
   p <- melt(p, measure.vars=params, variable.name="parameter")
@@
   setkey(p, parameter)
+  options(scipen=5)
 @
@@
 To test the effect of model misspecification, we simulated one network where
-the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
+the nodes exhibited heterogeneous pcreferential attachment power (half 0.5, the
 other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
@@
   \Sexpr{round(d["alpha", map], 2)} 
-  [\Sexpr{round(d["alpha", hpd.min], 2)}-
+  [\Sexpr{round(d["alpha", hpd.min], 2)} -
    \Sexpr{round(d["alpha", hpd.max], 2)}];
@@
 Both of these results were in line with estimates obtained on other simulated
-datasets (Table~\ref{tab:abchpd}).
+datasets (\cref{tab:abchpd}).
 
@@
 <<realdata, include=FALSE>>=
-  d <- fread(Sys.glob("../../simulations/aggregates/hpd/*"))
-  map <- dcast.data.table(d, dataset~parameter, value.var="map")
-  up <- dcast.data.table(d, dataset~parameter, value.var="hpd.max")
-  low <- dcast.data.table(d, dataset~parameter, value.var="hpd.min")
-  setkey(map, dataset)
-  setkey(up, dataset)
-  setkey(low, dataset)
+  m <- collect.metadata("../../simulations/aggregates/hpd/*")
+  d1 <- fread(rownames(m)[m$m_min == 1])
+  d2 <- fread(rownames(m)[m$m_min == 2])
+
+  map1 <- dcast.data.table(d1, dataset~parameter, value.var="map")
+  up1 <- dcast.data.table(d1, dataset~parameter, value.var="hpd.max")
+  low1 <- dcast.data.table(d1, dataset~parameter, value.var="hpd.min")
+  setkey(map1, dataset)
+  setkey(up1, dataset)
+  setkey(low1, dataset)
+
+  map2 <- dcast.data.table(d2, dataset~parameter, value.var="map")
+  up2 <- dcast.data.table(d2, dataset~parameter, value.var="hpd.max")
+  low2 <- dcast.data.table(d2, dataset~parameter, value.var="hpd.min")
+  setkey(map2, dataset)
+  setkey(up2, dataset)
+  setkey(low2, dataset)
 @
 
-We applied kernel-ABC to five published HIV datasets (Table~\ref{tab:data}),
+We applied kernel-ABC to five published HIV datasets (\cref{tab:data}),
 and found substantial heterogeneity among the parameter estimates
-(Figure~\ref{fig:abchpd}). Two of the datasets~\citep{niculescu2015recent,
-wang2015targeting} had estimated $\alpha$ values near unity (MAP estimates
+(\cref{fig:abchpd,fig:abchpdm2}). Two of the
+datasets~\autocite{niculescu2015recent, wang2015targeting} had estimated
+$\alpha$ values near unity for the prior allowing $m = 1$ (MAP estimates
 [95\% HPD] 
-  \Sexpr{map["niculescu2015", round(alpha, 2)]} 
-  [\Sexpr{low["niculescu2015", round(alpha, 2)]} - 
-   \Sexpr{up["niculescu2015", round(alpha, 2)]}]
+  \Sexpr{map1["niculescu2015", round(alpha, 2)]} 
+  [\Sexpr{low1["niculescu2015", round(alpha, 2)]} - 
+   \Sexpr{up1["niculescu2015", round(alpha, 2)]}]
 and
-  \Sexpr{map["wang2015", round(alpha, 2)]} 
-  [\Sexpr{low["wang2015", round(alpha, 2)]} -
-   \Sexpr{up["wang2015", round(alpha, 2)]}] respectively). 
-Another two datasets~\citep{li2015hiv, cuevas2009hiv} had lower estimated
-values and wider HPD intervals
-  (\Sexpr{map["li2015", round(alpha, 2)]} 
-  [\Sexpr{low["li2015", round(alpha, 2)]} - 
-  \Sexpr{up["li2015", round(alpha, 2)]}]
+  \Sexpr{map1["wang2015", round(alpha, 2)]} 
+  [\Sexpr{low1["wang2015", round(alpha, 2)]} -
+   \Sexpr{up1["wang2015", round(alpha, 2)]}] respectively).
+The MAP estimates did not change appreciably when $m = 1$ was disallowed by the
+prior, although the credible interval of the \textcite{niculescu2015recent}
+data was narrower
+  (\Sexpr{low1["niculescu2015", round(alpha, 2)]} - 
+   \Sexpr{up1["niculescu2015", round(alpha, 2)]}).
+When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
+low estimated $\alpha$ values
+  (\Sexpr{map1["li2015", round(alpha, 2)]} 
+  [\Sexpr{low1["li2015", round(alpha, 2)]} - 
+  \Sexpr{up1["li2015", round(alpha, 2)]}]
 and
-  \Sexpr{map["cuevas2009", round(alpha, 2)]} 
-  [\Sexpr{low["cuevas2009", round(alpha, 2)]} -
-   \Sexpr{up["cuevas2009", round(alpha, 2)]}]). 
-The \citet{novitsky2014impact} data had an extremely low estimated $\alpha$
-and a very wide HPD interval
-  (\Sexpr{map["novitsky2014", round(alpha, 2)]} 
-  [\Sexpr{low["novitsky2014", round(alpha, 2)]} -
-   \Sexpr{up["novitsky2014", round(alpha, 2)]}]). 
-For all the datasets except \citeauthor{novitsky2014impact}, estimated values of
-$I$ were below 2000, with narrow HPD intervals around two of the
-datasets
-  (\citeauthor{cuevas2009hiv}, \Sexpr{map["cuevas2009", floor(I)]} 
-  [\Sexpr{low["cuevas2009", floor(I)]} -
-   \Sexpr{up["cuevas2009", floor(I)]}];
-   \citeauthor{niculescu2015recent}, \Sexpr{map["niculescu2015", floor(I)]}
-  [\Sexpr{low["niculescu2015", floor(I)]} - 
-   \Sexpr{up["niculescu2015", floor(I)]}])
-and wider intervals around the other two
-  (\citeauthor{li2015hiv}, \Sexpr{map["li2015", floor(I)]} 
-  [\Sexpr{low["li2015", floor(I)]} -
-   \Sexpr{up["li2015", floor(I)]}];
-   \citeauthor{wang2015targeting}, \Sexpr{map["wang2015", floor(I)]}
-  [\Sexpr{low["wang2015", floor(I)]} - 
-   \Sexpr{up["wang2015", floor(I)]}]).
-The \citeauthor{novitsky2014impact} data was again the outlier, with a very high
+  \Sexpr{map1["cuevas2009", round(alpha, 2)]} 
+  [\Sexpr{low1["cuevas2009", round(alpha, 2)]} -
+   \Sexpr{up1["cuevas2009", round(alpha, 2)]}]). 
+However, the MAP estimates increased when $m = 1$ was not permitted, although
+the HPD intervals remained roughly the same
+  (\Sexpr{map2["li2015", round(alpha, 2)]} 
+  [\Sexpr{low2["li2015", round(alpha, 2)]} - 
+  \Sexpr{up2["li2015", round(alpha, 2)]}]
+and
+  \Sexpr{map2["cuevas2009", round(alpha, 2)]} 
+  [\Sexpr{low2["cuevas2009", round(alpha, 2)]} -
+   \Sexpr{up2["cuevas2009", round(alpha, 2)]}]).
+The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
+for both priors on $m$
+  (\Sexpr{map1["novitsky2014", round(alpha, 2)]} for $m \geq 1$;
+   \Sexpr{map2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).
+However, the confidence interval was much wider when $m = 1$ was allowed
+  ([\Sexpr{low1["novitsky2014", round(alpha, 2)]} -
+    \Sexpr{up1["novitsky2014", round(alpha, 2)]}] for $m \geq 1$ vs.
+    \Sexpr{low2["novitsky2014", round(alpha, 2)]} -
+    \Sexpr{up2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).
+
+For all the datasets except \citeauthor{novitsky2014impact}, estimated values
+of $I$ were below 2000 when $m = 1$ was allowed, with relatively narrow HPD
+intervals compared to the nonzero prior density region
+  (\citeauthor{cuevas2009hiv}, \Sexpr{map1["cuevas2009", floor(I)]} 
+  [\Sexpr{low1["cuevas2009", floor(I)]} -
+   \Sexpr{up1["cuevas2009", floor(I)]}];
+   \citeauthor{niculescu2015recent}, \Sexpr{map1["niculescu2015", floor(I)]}
+  [\Sexpr{low1["niculescu2015", floor(I)]} - 
+   \Sexpr{up1["niculescu2015", floor(I)]}];
+  (\citeauthor{li2015hiv}, \Sexpr{map1["li2015", floor(I)]} 
+  [\Sexpr{low1["li2015", floor(I)]} -
+   \Sexpr{up1["li2015", floor(I)]}];
+   \citeauthor{wang2015targeting}, \Sexpr{map1["wang2015", floor(I)]}
+  [\Sexpr{low1["wang2015", floor(I)]} - 
+   \Sexpr{up1["wang2015", floor(I)]}]).
+The \citeauthor{novitsky2014impact} data was the outlier, with a very high
 estimated $I$, and HPD interval spanning almost the entire prior region
-  (\Sexpr{map["novitsky2014", floor(I)]} 
-  [\Sexpr{low["novitsky2014", floor(I)]} -
-   \Sexpr{up["novitsky2014", floor(I)]}]).
-No information was gleaned about the $m$ parameter, with the HPD interval
-occupying the entire prior region for all datasets. The estimates of $N$ were
-similarly uninformative, with the exception that the point estimate for the
-\citeauthor{wang2015targeting} data was smaller
-  (\Sexpr{map["wang2015", floor(N)]})
-than the estimates for other datasets
- (average \Sexpr{map[dataset != "wang2015", floor(mean(N))]}).
+  (\Sexpr{map1["novitsky2014", floor(I)]} 
+  [\Sexpr{low1["novitsky2014", floor(I)]} -
+   \Sexpr{up1["novitsky2014", floor(I)]}]).
+The $I$ estimates and HPD intervals were generally robust to the choice of
+prior on $m$, with slightly narrower HPD intervals (compare
+\cref{fig:abchpd,fig:abchpdm2}).
+
+The MAP estimate of $m$ was equal to 1 for all but the
+\citeauthor{novitsky2014impact} data, when this value was allowed. However, the
+upper bound of the HPD interval was different for each dataset
+  (\citeauthor{niculescu2015recent}, \Sexpr{up1["niculescu2015", round(m)]};
+   \citeauthor{wang2015targeting}, \Sexpr{up1["wang2015", round(m)]};
+   \citeauthor{li2015hiv}, \Sexpr{up1["li2015", round(m)]};
+   \citeauthor{cuevas2009hiv}, \Sexpr{up1["cuevas2009", round(m)]}).
+When $m = 1$ was disallowed, the MAP for all datasets was either 2 or 3, with
+HPD intervals spanning the entire prior region. The estimates for the total
+number of nodes $N$ were largely uninformative for all samples, with almost all
+MAP estimates greater than 5000 and HPD intervals spanning almost the entire
+nonzero prior density region. The only exception was the \citeauthor{li2015hiv}
+data, for which the MAP estimate was lower 
+  (\Sexpr{map1["li2015", floor(N)]})
+when $m = 1$ was allowed.
 
@@
   \centering
-  \includegraphics{realdata-hpd.eps}
+  \includegraphics{realdata-hpd}
   \vspace{8pt}
@@
 Contact networks can have a strong influence on epidemic progression, and are
-potentially useful as a public health tool~\citep{wang2015targeting,
+potentially useful as a public health tool~\autocite{wang2015targeting,
 little2014using}. Despite this, few methods exist for investigating contact
-network parameters in a phylodynamic framework~\citep{groendyke2011bayesian,
+network parameters in a phylodynamic framework~\autocite{groendyke2011bayesian,
 volz2008sir, brown2011transmission, leventhal2012inferring}. Kernel-ABC is a
 model-agnostic method which can be used to investigate any quantity that
-affects tree shape~\citep{poon2015phylodynamic}. In this work, we developed a
+affects tree shape~\autocite{poon2015phylodynamic}. In this work, we developed a
 kernel-ABC-based method to infer the parameters of a contact network model. The
@@
 networks can be simulated. We demonstrated the method on the BA model,
-which is a simple preferential attachment model giving rise to the power law
+which is a simple pcreferential attachment model giving rise to the power law
 degree distributions commonly observed in real-world networks.
 
-By training a kernel-SVM classifier, we found that the $\alpha$ and $I$
-parameters, representing preferential attachment power and number of infected
-nodes, had a strong influence on tree shape. This was reflected in the relative
+By training a kernel-SVR classifier, we found that the $\alpha$ and $I$
+parameters, representing pcreferential attachment power and number of infected
+nodes, had a strong influence on tree shape. This was creflected in the relative
 accuracy of the kernel-ABC estimates of these parameters. The total number of
@@
 simulation again on the new, larger network could produce the exact same
-transmission tree as before.
-
-As noted by \citet{lintusaari2016identifiability}, uniform priors on model
+transmission tree as before. We note also that our accurate estimates of $I$
+may have been influenced by this prior, which places more mass on low $I$
+values. However, the MAP estimate of $I$ was very high for the
+\textcite{novitsky2013phylogenetic,novitsky2014impact} data, suggesting that a
+strong enough signal in the data can overcome the prior.
+
+As noted by \textcite{lintusaari2016identifiability}, uniform priors on model
 parameters may translate to highly informative priors on quantities of
-interest. We observed a non-linear relationship between the preferential
-attachment power $\alpha$ and the power law exponent $\gamma$ (Figure~S9).
-Therefore, placing a uniform prior on $\alpha$ between 0 and 2 is equivalent to
-placing an informative prior that $\gamma$ is close to 2. Therefore, if we were
-primarily interested in $\gamma$ rather than $\alpha$, a more sensible choice
-of prior might have a shape similar to the inverse of Figure~S9 and be bounded
-above by approximately $\alpha$ = 1.5. This would uniformly bound $\gamma$ in
-the region $2 \leq \gamma \leq 4$ commonly reported in the network
-literature~\citep{liljeros2001web, schneeberger2004scale, colgate1989risk,
-brown2011transmission}. We note however that \citet{jones2003assessment}
-estimated $\gamma$ values greater than four, in one case as high as 17, for
-some datasets, indicating that a wider range of permitted $\gamma$ values may
-be warranted.
+interest. We observed a non-linear relationship between the pcreferential
+attachment power $\alpha$ and the power law exponent $\gamma$
+(\cref{fig:gamma}). Thecrefore, placing a uniform prior on $\alpha$ between 0
+and 2 is equivalent to placing an informative prior that $\gamma$ is close to
+2. Thecrefore, if we were primarily interested in $\gamma$ rather than
+$\alpha$, a more sensible choice of prior might have a shape similar to the
+inverse of \cref{fig:gamma} and be bounded above by approximately $\alpha$ =
+1.5. This would uniformly bound $\gamma$ in the region $2 \leq \gamma \leq 4$
+commonly reported in the network literature~\autocite{liljeros2001web,
+schneeberger2004scale, colgate1989risk, brown2011transmission}. We note however
+that \textcite{jones2003assessment} estimated $\gamma$ values greater than
+four, in one case as high as 17, for some datasets, indicating that a wider
+range of permitted $\gamma$ values may be warranted.
 
 Our investigation of published HIV datasets indicated heterogeneity in the
-contact network structures underlying several distinct local epidemics. The
-five datasets analysed fell into three categories (Figure~\ref{fig:abchpd}).
-First, we estimated a preferential attachment power between 0.5 and 1 for the
-epidemics studied by \citet{cuevas2009hiv} and \citet{li2015hiv}, with credible
-intervals occupying nearly the entire region from 0 to 1.
-\citeauthor{cuevas2009hiv} studied a group of newly diagnosed individuals in
-the Basque Country, Spain. Although the individuals were of mixed risk groups,
-and therefore unlikely to comprise a single contact network, a high proportion
-of them (47\%) grouped into local clusters based on genetic distance. The low
-estimated attachment power for these data is consistent with the sampled
-sequences comprising many distinct sub-networks rather than a single connected
-network. \citeauthor{li2015hiv} sampled a large number of acutely infected
-MSM in Shanghai, China, in which we identified a large cluster from the
-phylogeny using a patristic distance cutoff~\citep{poon2015impact}. The low
-attachment power estimated for this dataset was surprising given the high
-phylogenetic relatedness of the sequences. It is possible that the number and
-diversity of circulating recombinant forms in the data introduced errors into
-our estimated viral phylogeny.
-
-For the outbreaks studied by \citet{niculescu2015recent} and
-\citet{wang2015targeting}, the estimated $\alpha$ was close to one, with a
-narrower credible interval than for the other studies.
-\citeauthor{niculescu2015recent} studied a recent outbreak among Romanian
-injection drug users (IDU), while \citeauthor{wang2015targeting} sampled
-acutely infected MSM in Beijing, China. Both studies discovered a high degree
-of phylogenetic relatedness owing to the recent infection times and homogeneous
-risk groups of the studied populations. The estimated number of infections for
-these datasets were also quite low, although the HPD interval for
-\citeauthor{wang2015targeting} was much wider than that for
-\citeauthor{niculescu2015recent}.
-
-The final studied dataset was an outlier in terms of estimated parameters.
-\citet{novitsky2013phylogenetic} sampled approximately 44\% of the
-HIV-infected individuals in the northern area of Mochudi, Botswana. Additional
-sampling in a later study~\citep{novitsky2014impact} brought the genotyping
-coverage up to 70\%. Even with such a high sampling coverage, we did not detect
-any large clusters using patristic distance, and therefore chose to analyze a
-subtree instead. Estimates of $\alpha$ and $N$ both had very wide HPD
-intervals and were markedly different from the other datasets. The estimated
-number of infected nodes was also extremely high, much higher than the
-estimated HIV prevalence of the town. Several factors may have contributed to
-these results. First, the authors note that the their sample was 75\% female.
-In a primarily heterosexual risk environment, removal of a disproportionate
-number of males from the network could obfuscate the true network structure,
-for example if the majority of highly connected nodes were of one gender.
-Second, the town in question was in close proximity to the country's capital,
-and the authors indicated that a high amount of migration takes place between
-the two locations. This suggests that the contact network may include a much
-larger group based in the capital city, which would explain the high estimate
-of $I$.
-
-When interpreting these results, we caution that the BA model is quite
-simple and most likely misspecified for these data. In particular, the average
-degree of a node in the network is equal to $2m$, and therefore is constrained
-to be a multiple of 2. Furthermore, we considered the case $m = 1$, where the
-network has no cycles, to be implausible and assigned it zero prior
-probability. This forces the average degree to be at least four, which may be
-unrealistically high for sexual networks. Additional modelling assumptions
-include the network being connected and static, all transmission rates being
-equal, no removal after infection, and identical behaviour of all nodes. This
-last is particularly problematic, as we showed by simulating a network where
-some nodes exhibited a higher attachment power than others. The estimated
-attachment power was simply the average of the two values, indicating that,
-although we could characterize the network in aggregate, the estimated
-parameters could not be said to apply to any individual node. Despite these
-issues, we felt it was best to demonstrate the method first on a simple model.
-It is possible to use this framework to fit more complex models which address
-some of these issues, such as one incorporating heterogeneous node behaviour,
-which may prove a fruitful avenue for future investigations.
+contact network structures underlying several distinct local epidemics. When
+interpreting these results, we caution that the BA model is quite simple and
+most likely misspecified for these data. In particular, the average degree of a
+node in the network is equal to $2m$, and thecrefore is constrained to be a
+multiple of 2. Furthermore, we considered the case $m = 1$, where the network
+has no cycles, to be implausible and thecrefore assigned it zero prior
+probability in one set of analyses. This forced the average degree to be at
+least four, which may be unrealistically high for sexual networks. The fact
+that the estimated values of $\alpha$ differed substatially for three datasets
+depending on whether or not $m = 1$ was allowed by the prior is futher evidence
+of this potential misspecification. However, we note that for two of the
+datasets, the estimated values of $\alpha$ did not change much between priors,
+and the estimates of $I$ were robust to the choice of prior for all datasets
+studied. More sophisticated models, for example models incorporating
+heterogeneity in node behaviour, are likely to provide a better fit to these
+data.
+
+With respect to the pcreferential attachment power $\alpha$, the five datasets
+analysed fell into three categories (\cref{fig:abchpd}). First, we
+estimated a pcreferential attachment power close to 1, indicating linear
+pcreferential attachment, for the outbreaks studied by
+\textcite{niculescu2015recent} and \textcite{wang2015targeting}. These values
+were robust to specifying different priors for $m$. Both studies were of
+populations in which we would expect a high degree of epidemiological
+relatedness: \textcite{niculescu2015recent} studied a recent outbreak among
+Romanian injection drug users (IDU), while \citeauthor{wang2015targeting}
+sampled acutely infected MSM in Beijing, China. Both these are contexts in
+which we would expect some of the assumptions of the BA model, such as a
+connected network, relatively high mean degree, and pcreferential attachment
+dynamics, to hold.
+
+The remaining three datasets (\textcite{cuevas2009hiv, novitsky2014impact,
+li2015hiv}) had estimated values of $\alpha$ below 0.5 when $m = 1$ was
+included in the prior, but these were not robust to changing the prior to
+exclude $m = 1$. For the \citeauthor{cuevas2009hiv} data, model
+misspecification is likely partially responsible. While the authors found that
+a large proportion of the samples were epidemiologically linked, these were
+mainly in small local clusters rather than the single large component
+postulated by the BA model. In addition, the mixed risk groups in the dataset
+would be unlikely to significantly interact, further weakening any global
+pcreferential attachment dynamics. The dataset studied by
+\textcite{novitsky2014impact} originated from a densely sampled population
+where the predominant risk factor was believed to be heterosexual exposure.
+Although the MAP estimate of $\alpha$ was almost unchanged when the value $m =
+1$ was excluded from the prior, the confidence interval shrank significantly.
+For both priors, the estimated prevalence was extremely high, in fact higher
+than the estimated HIV prevalence in the sampled region. The authors indicated
+that the source of the samples was a town in close proximity to the country's
+capital city, and suggested that there may have been a high degree of migration
+and partner interchange between the two locations. It is possible that the
+contact network underlying the subtree we investigated includes a much larger
+group based in the capital city, which would explain the high estimate of $I$.
+There is no clear explanation for the discrepancy between the two priors for
+the \textcite{li2015hiv} data, as the subset we analyzed formed a phylogenetic
+cluster and thecrefore was a good candidate for the BA model. However, nearly
+all the posterior density was assigned to $m = 1$ when this value was allowed,
+indicating that the network was more likely to have an acyclic tree structure.
+
+In addition to the aforementioned possibility of misspecification, additional
+modelling assumptions include the network being connected and static, all
+transmission rates being equal, no removal after infection, and identical
+behaviour of all nodes. This last is particularly problematic, as we showed by
+simulating a network where some nodes exhibited a higher attachment power than
+others. The estimated attachment power was simply the average of the two
+values, indicating that, although we could characterize the network in
+aggregate, the estimated parameters could not be said to apply to any
+individual node. Despite these issues, we felt it was best to demonstrate the
+method first on a simple model. It is possible to use this framework to fit
+more complex models which address some of these issues, such as one
+incorporating heterogeneous node behaviour, which may prove a fruitful avenue
+for future investigations.
 
@@
 this has been demonstrated to be a fair
-approximation~\citep[e.g.][]{leitner1996accurate}, and is frequently used in
-practice~\citep[e.g.][]{stadler2013uncovering}, the topologies of a viral
+approximation~\autocite[e.g.][]{leitner1996accurate}, and is frequently used in
+practice~\autocite[e.g.][]{stadler2013uncovering}, the topologies of a viral
 phylogeny and transmission tree can differ
-significantly~\citep{ypma2013relating} due to within-host evolution and the
+significantly~\autocite{ypma2013relating} due to within-host evolution and the
 sampling process. In addition, the ABC-SMC algorithm is
@@
 characteristics of the network structure underlying disease outbreaks. This
-work, and previous work by our group~\citep{poon2015phylodynamic}, has
+work, and previous work by our group~\autocite{poon2015phylodynamic}, has
 demonstrated that kernel-ABC is a broadly applicable and effective framework in
 which to perform phylodynamic inference.
-
-\section{Materials and Methods}
-
-Nodes in our networks followed simple SI dynamics, meaning that they
-became infected at a rate proportional to their number of infected neighbours,
-and never recovered. For all analyses, the transmission trees' branch lengths
-were scaled by dividing by their mean. We used the \textit{igraph} library's
-implementation of the BA model~\citep{csardi2006igraph} to generate the graphs.
-The analyses were run on Westgrid (\url{https://www.westgrid.ca/}) and a local
-computer cluster.
-
-\subsection{Computer program validation}
-
-To check that our implementation of Gillespie simulation was correct, we
-reproduced Figure 1A of \citet{leventhal2012inferring} (our Figure~S1), which
-plots the unbalancedness of transmission trees simulated over four network
-models at various levels of pathogen transmissibility. Our implementation of
-adaptive ABC-SMC was tested by applying it to the same mixture of Gaussians
-used by \citeauthor{del2012adaptive} to demonstrate their method (originally
-used by~\citet{sisson2007sequential}). We were able to obtain a close
-approximation to the function (see Figure~S2), and attained the stopping
-condition used by the authors in a comparable number of steps.
-
-\subsection{Kernel classifiers}
-
-We used the phylogenetic kernel developed by \citet{poon2013mapping} to test
-whether the parameters of the BA model had an effect on tree shape. 100
-networks were simulated under each of three different values of $\alpha$: 0.5,
-1.0, and 1.5 (300 networks total). The other parameters were fixed to the
-following values: $N$ = 5000, $I$ = 1000, and $m$ = 2. A transmission tree with
-500 tips was simulated over each network (300 transmission trees total). The
-300 trees were compared pairwise with the tree kernel to form a $300 \times
-300$ kernel matrix. The kernel meta-parameters $\lambda$ (the ``decay
-factor''), and $\sigma$ (the ``radial basis function
-variance'')~\citep[see][]{poon2013mapping}, were set to 0.3 and 4 respectively.
-We constructed a kSVM classifier for $\alpha$ using the \textit{kernlab}
-package~\citep{zeileis2004kernlab}, and evaluated its accuracy with 1000
-two-fold cross-validations.
-
-Three similar experiments were performed for the other BA model parameters (one
-experiment per parameter). $m$ was varied between 2, 3, and 4; $I$ between 500,
-1000, and 2000; and $N$ between 3000, 5000, and 8000. The parameters not being
-tested were fixed at the values $N$ = 5000, $I$ = 1000, $m$ = 2, and $\alpha$ =
-1. Thus, we performed a total of four kSVM cross-validations, one for each of
-the BA model parameters $\alpha$, $I$, $m$, and $N$. We repeated these four
-cross-validations with different values of $\lambda$ (0.2, 0.3, and 0.4) and
-$\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$), as well as on trees with
-differing numbers of tips (100, 500, and 1000) and in epidemics of differing
-size (500, 1000, and 2000). The combination of the number of sampled
-individuals (\textit{i.e.} the number of tips) and the epidemic size
-(\textit{i.e.} $I$) will be referred to as an ``epidemic scenario''. When
-evaluating the classifier for $I$, we did not consider trees with 1000 tips,
-because one of the tested $I$ values was 500, and the number of tips cannot be
-larger than $I$.
-
-For each of the four parameters, we also tested a linear regression against
-Sackin's index~\citep{shao1990tree} and an ordinary SVM based on the normalized
-lineages-through-time (nLTT) statistic~\citep{janzen2015approximate}.
-
-\subsection{ABC simulations}
-
-We simulated three transmission trees, each with 500 tips, under every element
-of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
-\{1000, 2000\}, $m$ = \{2, 3, 4\}, and $\alpha$ = \{0.0, 0.5, 1,
-1.5\}. This produced a total of 24 parameter combinations $\times$ three trees
-per combination = 72 trees total. The adaptive ABC algorithm was applied to
-each tree with these priors: $m \sim$ Uniform(1, 5), $\alpha \sim$ Uniform(0,
-2), and $(N, I)$ jointly uniform on the region \{$500 \leq N \leq 15000$, $500
-\leq I \leq 5000$, $I \leq N$\}. Following \citet{del2012adaptive} and
-\citet{beaumont2009adaptive}, all proposals were Gaussian, with variance equal
-to twice the empirical variance of the particles. The algorithm was run with
-1000 particles, 5 simulated datasets per particle, and the ``quality''
-parameter controlling the decay rate of the tolerance $\varepsilon$ set to
-0.95. We used the same stopping criterion as \citeauthor{del2012adaptive},
-namely when the MCMC acceptance rate dropped below 1.5\%. Point estimates for
-the parameters were obtained by taking the highest point of an estimated kernel
-density on the final set of particles, calculated using the \textit{density}
-function with the default parameters in \textit{R}. Highest posterior density
-(HPD) intervals were calculated with the \textit{HPDinterval} function from
-the \textit{R} package \textit{coda}~\citep{plummer2006coda}.
-
-Two further analyses were performed to address potential sources of error. To
-evaluate the effect of model misspecification in the case of heterogeneity
-among nodes, we generated a network where half the nodes were attached with
-power $\alpha$ = 0.5, and the other half with power $\alpha$ = 1.5. The other
-parameters for this network were $N$ = 5000, $I$ = 1000, and $m$ = 2. To
-investigate the effects of potential sampling bias, we simulated a transmission
-tree where the tips were sampled in a peer-driven fashion, rather than at
-random. That is, the probability to sample a node was twice as high if any of
-that node's network peers had already been sampled. The parameters of this
-network were $N$ = 5000, $I$ = 2000, $m$ = 2, and $\alpha$ = 0.5.
-
-\subsection{Investigation of published data}
-
-We applied our kernel-ABC method to several published HIV datasets. Because the
-BA model generates networks with a single connected component, we
-specifically searched for datasets which originated from existing clusters,
-either phylogenetically or geographically defined. Characteristics of the
-datasets we investigated are given in Table~\ref{tab:data}.
-
-\begin{table*}[!t]
-  \centering
-  \begin{tabular}{ccccc}
-    Reference & Sequences ($n$) & Location & Risk group & Gene \\
-    \hline
-    \citet{wang2015targeting} & 173 & Beijing, China & MSM & \textit{pol} \\
-    \citet{cuevas2009hiv} & 287 & Basque Country, Spain & mixed & \textit{pol} \\
-    \citet{novitsky2013phylogenetic} & \multirow{2}{*}{180} &
-    \multirow{2}{*}{Mochudi, Botswana} & \multirow{2}{*}{mixed} &
-    \multirow{2}{*}{\textit{env}} \\ \citet{novitsky2014impact} \\
-    \citet{li2015hiv} & 280 & Shanghai, China & MSM & \textit{pol} \\
-    \citet{niculescu2015recent} & 136 & Romaina & IDU & \textit{pol} \\
-    \hline
-  \end{tabular}
-  \caption{Characteristics of published datasets investigated with kernel-ABC.
-  Acronyms: MSM, men who have sex with men; IDU, injection drug users.}
-  \label{tab:data}
-\end{table*}
-
-We downloaded all sequences associated with each study from GenBank. For the
-\citet{novitsky2014impact} data, each sequence was aligned pairwise to the HXB2
-reference sequence (Genbank accession number K03455) and the hypervariable
-regions were clipped out with \textit{BioPython} version
-1.66+~\citep{cock2009biopython}. Sequences were multiply aligned using
-\textit{MUSCLE} version 3.8.31 \citep{edgar2004muscle}, and alignments were
-manually inspected with \textit{Seaview} version 4.4.2
-\citep{gouy2010seaview}. Phylogenies were constructed from the nucleotide
-alignments by approximate maximum likelihood using \textit{FastTree2} version
-2.1.7 with the generalized time-reversible model. Transmission trees were
-estimated by rooting and time-scaling the phylogenies by root-to-tip
-regression, using a modified version of Path-O-Gen (distributed as part of
-BEAST~\citep{drummond2007beast}) as described
-previously~\citep{poon2015phylodynamic}.
-
-Two of the datasets \citep{li2015hiv,novitsky2014impact} were initially much
-larger than the others, containing 1265 and 1299 sequences respectively. To
-ensure that the analyses were comparable, we reduced these to a number of
-sequences similar to the smaller datasets. For the \citet{li2015hiv} data,
-we detected a cluster of size 280 using a patristic distance cutoff of 0.02 as
-described previously~\citep{poon2015impact}. Only sequences within this
-cluster were carried forward. For the \citet{novitsky2014impact} data, no
-large clusters were detected using the same cutoff, so we analysed a subtree of
-size 180 chosen arbitrarily.
-
-For all datasets, we used the priors $\alpha$ $\sim$ Uniform(0, 2), $m$ $\sim$
-Uniform(2, 5), and $N$ and $I$ jointly uniform on the region 
-\{$n \leq N \leq 10000$, $n \leq I \leq 10000$, $I \leq N$\}, where $n$ is the
-number of tips in the tree. The other parameters to the SMC algorithm
-were the same as used for the simulation experiments.
-
-\section{Supplementary data}
-
-Supplementary figures S1-S10 are available at Molecular Biology and Evolution
-online (\url{http://www.mbe.oxfordjournals.org/}).
 
@@
 
-\bibliographystyle{natbib}
-\bibliography{papers}
+\printbibliography
+
+\newpage
+
+\section*{Supplementary Figures}
+
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{leventhal2012fig1.pdf}
+  \caption{
+    Reproduction of Figure 1A from Leventhal \textit{et al.} (2012) used to
+    check the accuracy of our implementation of Gillespie simulation.
+    Transmission trees were simulated over three types of network, with
+    pathogen transmissibility varying from 0 to 1. Sackin's index was
+    calculated for each simulated transmission tree.
+  }
+  \label{fig:leventhal}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{smc-test.pdf}
+  \caption{Approximation of mixture of Gaussians used by
+    Del Moral \textit{et al.} (2012) and Sisson \textit{et al.} (2009) to test
+    SMC. Solid black line indicates true distribution. Grey shaded area shows
+    SMC approximation obtained with our implementation, using 10000
+    particles with one simulated data point per particle.
+  }
+  \label{fig:smctest}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{kernel-alpha-crossv.pdf}
+  \caption{
+    Cross-validation accuracy of kernel-SVM classifiers for $\alpha$ parameter
+    of BA network model, for various tree kernel meta-parameters and
+    epidemic scenarios. Each point was calculated based on 300 simulated
+    transmission trees over networks with $\alpha$ = 0.5, 1.0, or 1.5. Dotted
+    and and dashed lines indicate, respectively, performance of SVM using the
+    nLTT statistic, and linear regression using Sackin's index. Facets
+    are number of infected nodes before the simulation was stopped ($I$) and
+    number of tips in the sampled transmission tree.
+  }
+  \label{fig:alphacrossv}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{kernel-m-crossv.pdf}
+  \caption{Cross-validation accuracy of kernel-SVM classifiers for $m$
+      parameter of BA network model, for various tree kernel
+      meta-parameters and epidemic scenarios. Each point was calculated based
+      on 300 simulated transmission trees over networks with $m$ = 2, 3, or 4.
+      Dotted and and dashed lines indicate, respectively, performance of SVM
+      using the nLTT statistic, and linear regression using Sackin's
+      index. Facets are number of infected nodes before the simulation was
+      stopped ($I$) and number of tips in the sampled transmission tree.
+  }
+  \label{fig:mcrossv}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{kernel-I-crossv.pdf}
+  \caption{Cross-validation accuracy of kernel-SVM classifiers for number of
+      infected nodes ($I$) under BA network model, for various tree
+      kernel meta-parameters and two tree sizes. Each point was calculated
+      based on 300 simulated transmission trees over networks with $I$ = 500,
+      1000, or 2000. Dotted and and dashed lines indicate, respectively,
+      performance of SVM using the nLTT statistic, and linear regression
+      using Sackin's index. Facets are the number of tips in the sampled
+      transmission tree.
+  }
+  \label{fig:Icrossv}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{kernel-N-crossv.pdf}
+  \caption{Cross-validation accuracy of kernel-SVM classifiers for total number
+      of nodes ($N$) under BA network model, for various tree kernel
+      meta-parameters and epidemic scenarios sizes. Each point was calculated
+      based on 300 simulated transmission trees over networks with $N$ = 3000,
+      5000, or 8000. Dotted and and dashed lines indicate, respectively,
+      performance of SVM using the nLTT statistic, and linear regression
+      using Sackin's index. Facets are the number of tips in the sampled
+      transmission tree.
+  }
+  \label{fig:Ncrossv}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{abc-point-estimate-m3.pdf}
+  \caption{
+    Point estimates of BA model parameters obtained by running kernel-ABC
+    on simulated phylogenies without training, for simulations with $m = 3$.
+    Dotted lines indicate true values, and limits of the $y$-axes are regions
+    of uniform prior density. (A) Estimates for $\alpha$ and $I$ against their
+    true values in simulations. (B) Estimates for $m$ and $N$, which were held
+    fixed in these simulations, against true values of $\alpha$.
+  }
+  \label{fig:abcptm3}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{abc-point-estimate-m4.pdf}
+  \caption{
+    Point estimates of BA model parameters obtained by running kernel-ABC
+    on simulated phylogenies without training, for simulations with $m = 4$.
+    Dotted lines indicate true values, and limits of the $y$-axes are regions
+    of uniform prior density. (A) Estimates for $\alpha$ and $I$ against their
+    true values in simulations. (B) Estimates for $m$ and $N$, which were held
+    fixed in these simulations, against true values of $\alpha$.
+  }
+  \label{fig:abcptm4}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{alpha-gamma.pdf}
+  \caption{
+      Relationship between pcreferential attachment power parameter $\alpha$
+      and power law exponent $\gamma$ for networks simulated under the BA
+      network model with $N$ = 5000 and $m$ = 2.
+  }
+  \label{fig:gamma}
+\end{figure}
+
+\begin{figure}[ht]
+  \includegraphics{realdata-hpd-m2}
+  \vspace{8pt}
+  \caption{
+      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
+      parameters of the BA network model, fitted to five published HIV datasets
+      with kernel-ABC. Regions of nonzero prior density are shown on $x$-axes.
+  }
+  \label{fig:abchpdm2}
+\end{figure}
+
+\begin{figure}[ht]
+  \includegraphics{cuevas2009-posterior}
+  \caption{
+      Posterior distribution of BA model parameters for
+      \textcite{cuevas2009hiv} data. Vertical lines indicate maximum \textit{a
+      posteriori} estimates, and shaded areas are 95\% highest posterior
+      density intervals. $x$-axis indicates regions of nonzero prior density.
+  }
+  \label{fig:cuevas}
+\end{figure}
+
+\begin{figure}[ht]
+  \includegraphics{li2015-posterior}
+  \caption{
+      Posterior distribution of BA model parameters for \textcite{li2015hiv}
+      data. Vertical lines indicate maximum \textit{a posteriori} estimates,
+      and shaded areas are 95\% highest posterior density intervals. $x$-axis
+      indicates regions of nonzero prior density.
+  }
+  \label{fig:li}
+\end{figure}
+
+\begin{figure}[ht]
+  \includegraphics{niculescu2015-posterior}
+  \caption{
+      Posterior distribution of BA model parameters for
+      \textcite{niculescu2015recent} data. Vertical lines indicate maximum
+      \textit{a posteriori} estimates, and shaded areas are 95\% highest
+      posterior density intervals. $x$-axis indicates regions of nonzero prior
+      density.
+  }
+  \label{fig:niculescu}
+\end{figure}
+
+\begin{figure}[ht]
+  \includegraphics{novitsky2014-posterior}
+  \caption{
+      Posterior distribution of BA model parameters for
+      \textcite{novitsky2013phylogenetic, novitsky2014impact} data. Vertical
+      lines indicate maximum \textit{a posteriori} estimates, and shaded areas
+      are 95\% highest posterior density intervals. $x$-axis indicates regions
+      of nonzero prior density.
+  }
+  \label{fig:novitsky}
+\end{figure}
+
+\begin{figure}[ht]
+  \includegraphics{wang2015-posterior}
+  \caption{
+      Posterior distribution of BA model parameters for
+      \textcite{wang2015targeting} data. Vertical lines indicate maximum
+      \textit{a posteriori} estimates, and shaded areas are 95\% highest
+      posterior density intervals. $x$-axis indicates regions of nonzero prior
+      density.
+  }
+  \label{fig:wang}
+\end{figure}
 

### 2016-04-22 doc/paper/paper.tex
Stuff to work on for dynamics

@@
-\documentclass[nogrid]{MBE}
-\usepackage{url}
+\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
 
-% Mol Biol Evol stuff
-\jshort{mst}
-\volname{}
-\jvolume{0}
-\jvol{}
-\jissue{0}
-\pubyear{2016}
-\mstype{Article}
-\artid{???}
-\access{???}
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
+\usepackage[T1]{fontenc}
+\usepackage{garamondx}
+\usepackage[garamondx,cmbraces]{newtxmath}
+\usepackage[backend=biber]{biblatex}
+\usepackage{fullpage}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{setspace}
+\usepackage{cleveref}
+
+\addbibresource{papers.bib}
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
+\onehalfspacing
 
-\title[Contact networks and ABC]{Phylodynamic inference of contact network
-parameters through approximate Bayesian computation}
-\author[McCloskey, Liang, and Poon]{Rosemary M. McCloskey$^{*,1}$, 
-    Richard H. Liang$^1$, and Art F.Y. Poon$^{1,2}$}
-\address{$^1$BC Centre for Excellence in HIV/AIDS, 608-1081 Burrard Street, Vancouver, Canada \\
-$^2$Department of Medicine, University of British Columbia, Vancouver, Canada}
-\history{Submitted 5 April 2016}
-\coresp{E-mail: rmccloskey@cfenet.ubc.ca.}
-\editor{}
+\newcommand{\tablepath}{../tables}
+\graphicspath{{../figures/}}
+
+\title{Reconstructing contact network parameters from viral phylogenies}
+\author{Rosemary M. McCloskey, Richard H. Liang, and Art F.Y. Poon}
 \abstract{\input{abstract}}
-\keyword{Phylogenetics, phylodynamics, approximate Bayesian computation,
-contact network, transmission tree, human immunodeficiency virus.}
 
@@
 thereby shaping the inter-host viral
-phylogeny~\citep{drummond2003measurably}. The term ``phylodynamics'' was
+phylogeny~\autocite{drummond2003measurably}. The term ``phylodynamics'' was
 coined to describe this interaction, as well as the growing family of inference
 methods to estimate epidemiological parameters from viral
-phylogenies~\citep{grenfell2004unifying}. These methods have revealed
+phylogenies~\autocite{grenfell2004unifying}. These methods have revealed
 diverse properties of local viral outbreaks, from basic reproductive
-number~\citep{stadler2012estimating}, to the degree of
-clustering~\citep{hughes2009molecular}, to the elevated transmission risk
-during acute infection~\citep{volz2012simple}. On the other hand, although
+number~\autocite{stadler2012estimating}, to the degree of
+clustering~\autocite{hughes2009molecular}, to the elevated transmission risk
+during acute infection~\autocite{volz2012simple}. On the other hand, although
 sophisticated methods have been developed for fitting complex population
-genetic models to phylogenies~\citep{rasmussen2014phylodynamic}, inference
+genetic models to phylogenies~\autocite{rasmussen2014phylodynamic}, inference
 of structural network parameters has to date been limited. However, it has been
 shown that network structure has a tangible impact on phylogeny
-shape~\citep{leventhal2012inferring, colijn2014phylogenetic,
+shape~\autocite{leventhal2012inferring, colijn2014phylogenetic,
 goodreau2006assessing, robinson2013dynamics, villandre2016assessment},
 suggesting that such statistical inference might be
-possible~\citep{welch2011statistical}.
+possible~\autocite{welch2011statistical}.
 
-Survey-based studies of sexual networks~\citep{liljeros2001web,
+Survey-based studies of sexual networks~\autocite{liljeros2001web,
 schneeberger2004scale, colgate1989risk} have found that these networks tend to
-have a degree distribution which follows a power law \citep[although there has
+have a degree distribution which follows a power law \autocite[although there has
 been some disagreement, see][]{handcock2004likelihood}. That is, the number of
 nodes of degree $k$ is proportional to $k^{-\gamma}$ for some constant
-$\gamma$. These networks are also referred to as ``scale-free''. One process by
-which scale-free networks can be generated is preferential attachment, where
+$\gamma$. These networks are also creferred to as ``scale-free''. One process by
+which scale-free networks can be generated is pcreferential attachment, where
 nodes with a high number of contacts attract new connections at an elevated
-rate. The first contact network model incorporating preferential attachment was
-introduced by \citet{barabasi1999emergence}, and is now referred to as the
+rate. The first contact network model incorporating pcreferential attachment was
+introduced by \textcite{barabasi1999emergence}, and is now creferred to as the
 Barab\'asi-Albert (BA) model. Under this model, networks are formed by
@@
 viral phylogeny. We implemented the adaptive sequential Monte Carlo (SMC)
-algorithm for ABC developed by \citet{del2012adaptive}. The SMC algorithm keeps
-track of a population of parameter ``particles'', which are initially sampled
-from the parameters' joint prior distribution. Several datasets are simulated
-under the model of interest for each of the particles. In this case, the
-datasets are transmission trees, which are generated by a two-step process.
+algorithm for ABC developed by \textcite{del2012adaptive}. The SMC algorithm
+keeps track of a population of parameter ``particles'', which are initially
+sampled from the parameters' joint prior distribution. Several datasets are
+simulated under the model of interest for each of the particles. In this case,
+the datasets are transmission trees, which are generated by a two-step process.
 First, a contact network is simulated according to the network model being fit.
 Second, a transmission tree is simulated over that network with a Gillespie
-simulation algorithm~\citep{gillespie1976general}, in the same fashion as
-several previous studies \citep[\textit{e.g.}][]{robinson2013dynamics,
+simulation algorithm~\autocite{gillespie1976general}, in the same fashion as
+several previous studies \autocite[\textit{e.g.}][]{robinson2013dynamics,
 leventhal2012inferring}. The particles are weighted according to the similarity
 between their associated simulated trees and the observed tree. To quantify
-this similarity, we used the tree kernel developed by \citet{poon2013mapping}.
-Particles are iteratively perturbed by applying a Metropolis-Hastings kernel
-and, if the move is accepted, simulating new datasets under the new parameters.
-When a particle's weight drops to zero, because its simulated trees are too
-dissimilar to the observed tree, the particle is dropped from the population,
-and eventually replaced by a resampled particle with a higher weight. As the
-algorithm progresses, the population converges to a Monte Carlo approximation
-of the ABC target distribution, which is assumed to approximate the desired
-posterior~\citep{del2012adaptive, sunnaaker2013approximate}.
+this similarity, we used the tree kernel developed by
+\textcite{poon2013mapping}. Particles are iteratively perturbed by applying a
+Metropolis-Hastings kernel and, if the move is accepted, simulating new
+datasets under the new parameters. When a particle's weight drops to zero,
+because its simulated trees are too dissimilar to the observed tree, the
+particle is dropped from the population, and eventually replaced by a resampled
+particle with a higher weight. As the algorithm progresses, the population
+converges to a Monte Carlo approximation of the ABC target distribution, which
+is assumed to approximate the desired posterior~\autocite{del2012adaptive,
+sunnaaker2013approximate}. A computer program implementing our method is freely
+available at \url{https://github.com/rmcclosk/netabc} (last accessed April 3,
+2016).
 
-A computer program implementing our method is freely available at
-\url{https://github.com/rmcclosk/netabc} (last accessed April 3, 2016).
+To check that our implementation of Gillespie simulation was correct, we
+reproduced Figure 1A of \textcite{leventhal2012inferring} (our
+\cref{fig:leventhal}), which plots the unbalancedness of transmission trees
+simulated over four network models at various levels of pathogen
+transmissibility. Our implementation of adaptive ABC-SMC was tested by applying
+it to the same mixture of Gaussians used by \citeauthor{del2012adaptive} to
+demonstrate their method (originally used by~\textcite{sisson2007sequential}).
+We were able to obtain a close approximation to the function (see
+\cref{fig:smctest}), and attained the stopping condition used by the authors in
+a comparable number of steps.
+
+Nodes in our networks followed simple SI dynamics, meaning that they became
+infected at a rate proportional to their number of infected neighbours, and
+never recovered. For all analyses, the transmission trees' branch lengths were
+scaled by dividing by their mean. We used the \textit{igraph} library's
+implementation of the BA model~\autocite{csardi2006igraph} to generate the
+graphs. The analyses were run on Westgrid (\url{https://www.westgrid.ca/}) and
+a local computer cluster.
+
+\subsection{Kernel classifiers}
+
+We used the phylogenetic kernel developed by \textcite{poon2013mapping} to test
+whether the parameters of the BA model had an effect on tree shape. 100
+networks were simulated under each of three different values of $\alpha$: 0.5,
+1.0, and 1.5 (300 networks total). The other parameters were fixed to the
+following values: $N$ = 5000, $I$ = 1000, and $m$ = 2. A transmission tree with
+500 tips was simulated over each network (300 transmission trees total). The
+300 trees were compared pairwise with the tree kernel to form a $300 \times
+300$ kernel matrix. The kernel meta-parameters $\lambda$ (the ``decay
+factor''), and $\sigma$ (the ``radial basis function
+variance'')~\autocite[see][]{poon2013mapping}, were set to 0.3 and 4 respectively.
+We constructed a kSVR classifier for $\alpha$ using the \textit{kernlab}
+package~\autocite{zeileis2004kernlab}, and evaluated its accuracy with 1000
+two-fold cross-validations.
+
+Three similar experiments were performed for the other BA model parameters (one
+experiment per parameter). $m$ was varied between 2, 3, and 4; $I$ between 500,
+1000, and 2000; and $N$ between 3000, 5000, and 8000. The parameters not being
+tested were fixed at the values $N$ = 5000, $I$ = 1000, $m$ = 2, and $\alpha$ =
+1. Thus, we performed a total of four kSVR cross-validations, one for each of
+the BA model parameters $\alpha$, $I$, $m$, and $N$. We repeated these four
+cross-validations with different values of $\lambda$ (0.2, 0.3, and 0.4) and
+$\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$), as well as on trees with
+differing numbers of tips (100, 500, and 1000) and in epidemics of differing
+size (500, 1000, and 2000). The combination of the number of sampled
+individuals (\textit{i.e.} the number of tips) and the epidemic size
+(\textit{i.e.} $I$) will be creferred to as an ``epidemic scenario''. When
+evaluating the classifier for $I$, we did not consider trees with 1000 tips,
+because one of the tested $I$ values was 500, and the number of tips cannot be
+larger than $I$.
+
+For each of the four parameters, we also tested a linear regression against
+Sackin's index~\autocite{shao1990tree} and an ordinary SVR based on the normalized
+lineages-through-time (nLTT) statistic~\autocite{janzen2015approximate}.
+
+\subsection{ABC simulations}
+
+We simulated three transmission trees, each with 500 tips, under every element
+of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
+\{1000, 2000\}, $m$ = \{2, 3, 4\}, and $\alpha$ = \{0.0, 0.5, 1,
+1.5\}. This produced a total of 24 parameter combinations $\times$ three trees
+per combination = 72 trees total. The adaptive ABC algorithm was applied to
+each tree with these priors: $m \sim$ DiscreteUniform(1, 5), $\alpha \sim$
+Uniform(0, 2), and $(N, I)$ jointly uniform on the region \{$500 \leq N \leq
+15000$, $500 \leq I \leq 5000$, $I \leq N$\}. Proposals for $\alpha$, $N$, and
+$I$ were Gaussian, while proposals for $m$ were Poisson. Following
+\textcite{del2012adaptive} and \textcite{beaumont2009adaptive}, the variance of
+all proposals was equal to the empirical variance of the particles.
+
+The algorithm was run with 1000 particles, 5 simulated datasets per particle,
+and the ``quality'' parameter controlling the decay rate of the tolerance
+$\varepsilon$ set to 0.95. We used the same stopping criterion as
+\citeauthor{del2012adaptive}, namely when the MCMC acceptance rate dropped
+below 1.5\%. Point estimates for the parameters were obtained by taking the
+highest point of an estimated kernel density on the final set of particles,
+calculated using the \textit{density} function with the default parameters in
+\textit{R}. Highest posterior density (HPD) intervals were calculated with the
+\textit{HPDinterval} function from the \textit{R} package
+\textit{coda}~\autocite{plummer2006coda}.
+
+Two further analyses were performed to address potential sources of error. To
+evaluate the effect of model misspecification in the case of heterogeneity
+among nodes, we generated a network where half the nodes were attached with
+power $\alpha$ = 0.5, and the other half with power $\alpha$ = 1.5. The other
+parameters for this network were $N$ = 5000, $I$ = 1000, and $m$ = 2. To
+investigate the effects of potential sampling bias, we simulated a transmission
+tree where the tips were sampled in a peer-driven fashion, rather than at
+random. That is, the probability to sample a node was twice as high if any of
+that node's network peers had already been sampled. The parameters of this
+network were $N$ = 5000, $I$ = 2000, $m$ = 2, and $\alpha$ = 0.5.
+
+\subsection{Investigation of published data}
+
+We applied our kernel-ABC method to several published HIV datasets. Because the
+BA model generates networks with a single connected component, we
+specifically searched for datasets which originated from existing clusters,
+either phylogenetically or geographically defined. Characteristics of the
+datasets we investigated are given in \cref{tab:data}.
+
+\begin{table*}[!t]
+  \centering
+  \input{\tablepath/realdata}
+  \caption{Characteristics of published datasets investigated with kernel-ABC.
+  Acronyms: MSM, men who have sex with men; IDU, injection drug users.}
+  \label{tab:data}
+\end{table*}
+
+We downloaded all sequences associated with each study from GenBank. For the
+\textcite{novitsky2014impact} data, each sequence was aligned pairwise to the HXB2
+creference sequence (Genbank accession number K03455) and the hypervariable
+regions were clipped out with \textit{BioPython} version
+1.66+~\autocite{cock2009biopython}. Sequences were multiply aligned using
+\textit{MUSCLE} version 3.8.31 \autocite{edgar2004muscle}, and alignments were
+manually inspected with \textit{Seaview} version 4.4.2
+\autocite{gouy2010seaview}. Phylogenies were constructed from the nucleotide
+alignments by approximate maximum likelihood using \textit{FastTree2} version
+2.1.7 with the generalized time-reversible model. Transmission trees were
+estimated by rooting and time-scaling the phylogenies by root-to-tip
+regression, using a modified version of Path-O-Gen (distributed as part of
+BEAST~\autocite{drummond2007beast}) as described
+previously~\autocite{poon2015phylodynamic}.
+
+Two of the datasets \autocite{li2015hiv,novitsky2014impact} were initially much
+larger than the others, containing 1265 and 1299 sequences respectively. To
+ensure that the analyses were comparable, we reduced these to a number of
+sequences similar to the smaller datasets. For the \textcite{li2015hiv} data,
+we detected a cluster of size 280 using a patristic distance cutoff of 0.02 as
+described previously~\autocite{poon2015impact}. Only sequences within this
+cluster were carried forward. For the \textcite{novitsky2014impact} data, no
+large clusters were detected using the same cutoff, so we analysed a subtree of
+size 180 chosen arbitrarily.
+
+For all datasets, we used the priors $\alpha$ $\sim$ Uniform(0, 2) and $N$ and
+$I$ jointly uniform on the region \{$n \leq N \leq 10000$, $n \leq I \leq
+10000$, $I \leq N$\}, where $n$ is the number of tips in the tree. Since the
+value $m = 1$ produces networks with no cycles, which we considered fairly
+implausible, we ran one analysis with the prior $m \sim$ DiscreteUniform(1, 5),
+and one with the prior $m \sim$ DiscreteUniform(2, 5). The other parameters to
+the SMC algorithm were the same as used for the simulation experiments.
 
@@
 parameter values, calculated pairwise tree kernel scores between them, and
-attempted to classify the trees using a kernel support vector machine (kSVM).
-We also tested classifiers based on Sackin's index~\citep{shao1990tree} and the
-normalized lineages-through-time (nLTT)
-statistic~\citep{janzen2015approximate}. Accuracy of the kSVMs varied based on
-the parameter being tested (Figure~\ref{fig:rsquared}, left). Classifiers based
-on two other tree statistics, the nLTT and Sackin's index, generally exhibited
+attempted to classify the trees using a kernel support vector machine (kSVR).
+We also tested classifiers based on Sackin's index~\autocite{shao1990tree} and
+the normalized lineages-through-time (nLTT)
+statistic~\autocite{janzen2015approximate}. Accuracy of the kSVRs varied based
+on the parameter being tested (\cref{fig:rsquared}, left). Classifiers based on
+two other tree statistics, the nLTT and Sackin's index, generally exhibited
 worse performance than the tree kernel, although the magnitude of the disparity
-varied between the parameters (Figure~\ref{fig:rsquared}, centre and right).
-The results were largely robust to variations in the tree kernel
-meta-parameters $\lambda$ and $\sigma$ (Figures~S3-S6).
+varied between the parameters (\cref{fig:rsquared}, centre and right). The
+results were largely robust to variations in the tree kernel meta-parameters
+$\lambda$ and $\sigma$ (\cref{fig:alphacrossv,fig:mcrossv,fig:Ncrossv,fig:Icrossv}).
 
-When classifying $\alpha$, the kernel-SVM classifier had an average $R^2$ of 
+When classifying $\alpha$, the kernel-SVR classifier had an average $R^2$ of 
     0.92,
@@
     0.56
-for the nLTT-based SVM, and
+for the nLTT-based SVR, and
     0.75
@@
 $R^2$ values of 
-  0.12 for kSVM,
+  0.12 for kSVR,
   0.01 for the nLTT, and
@@
 Finally, the performance of classifiers for $N$ depended heavily on the
-epidemic scenario. The $R^2$ of the kSVM classifier ranged from
+epidemic scenario. The $R^2$ of the kSVR classifier ranged from
   0.08
@@
   0.82
-for the largest. Likewise, $R^2$ for the nLTT-based SVM ranged from 
+for the largest. Likewise, $R^2$ for the nLTT-based SVR ranged from 
   0.01
@@
 
-Figure~\ref{fig:abcpt} shows maximum \textit{a posteriori} (MAP) point
+%<<point_est, include=FALSE>>=
+%    f <- Sys.glob("../../simulations/abc-pa-free-m/point-estimate/*")
+%    d <- fread(f)
+%    d[,m := floor(m)]
+%    d[,alpha_error := abs(true_alpha - alpha)]
+%    d[,N_error := abs(true_N - N)]
+%    d[,I_error := abs(true_I - I)]
+%    d[,m_error := abs(true_m - m)]
+%
+%    alpha.av <- anova(lm(alpha_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
+%    m.av <- anova(lm(m_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
+%    N.av <- anova(lm(N_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
+%    I.av <- anova(lm(I_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
+%
+%    m.tbl <- d[,prop.table(table(m_error))]
+%    zero.alpha.test <- wilcox.test(d[true_alpha == 0, alpha_error], 
+%                                   d[true_alpha != 0, alpha_error])
+%    alpha.I.test <- wilcox.test(d[true_alpha < 1, I_error],
+%                                d[true_alpha >= 1, I_error])
+%    I.I.test <- wilcox.test(d[true_I == 1000, I_error],
+%                            d[true_I == 2000, I_error])
+%    alpha.m.test <- wilcox.test(d[true_alpha == 0 | true_alpha == 1, m_error],
+%                                d[true_alpha == 0.5 | true_alpha == 1.5, m_error])
+%    stopifnot(alpha.av["factor(true_m)", "Pr(>F)"] > 0.05)
+%    stopifnot(alpha.av["factor(true_I)", "Pr(>F)"] > 0.05)
+%    stopifnot(I.I.test$p.value < 1e-5)
+%    stopifnot(I.av["factor(true_m)", "Pr(>F)"] > 0.05)
+%    stopifnot(m.av["factor(true_m)", "Pr(>F)"] > 0.05)
+%    stopifnot(m.av["factor(true_I)", "Pr(>F)"] > 0.05)
+%    stopifnot(min(N.av[1:3,"Pr(>F)"]) > 0.05)
+%@
+
+\Cref{fig:abcpt} shows maximum \textit{a posteriori} (MAP) point
 estimates of the BA model parameters obtained with kernel-ABC on simulated
@@
 parameter was set to 2, however the results for $m = 3$ and $m = 4$ were
-similar (Figures S7 and S8). Average boundaries of 95\% HPD intervals are given
-in Table~\ref{tab:abchpd}.
+similar (\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% HPD
+intervals are given in \cref{tab:abchpd}.
 
 The accuracy of the parameter estimates obtained with kernel-ABC
-paralleled the results from the kSVM classifier. Of the four parameters,
+paralleled the results from the kSVR classifier. Of the four parameters,
 $\alpha$ was the most accurately estimated, with point estimates having a
 median [IQR] absolute error of 
-    0.11 
-    [0.05 - 
-    0.18].
+%    d[,round(median(alpha_error), 2)] 
+%    [d[,round(quantile(alpha_error, 0.25), 2)] - 
+%    d[,round(quantile(alpha_error, 0.75), 2)]].
 The errors when the true value of $\alpha$ was zero were significantly greater
 than those for the other values 
-    (Wilcoxon rank-sum test, $p$ = $6\!\times\!10^{-4}$),
+%    (Wilcoxon rank-sum test, $p$ = $latexSN(round(zero.alpha.test$p.value, 4))$),
 but did not vary across the true values of the other parameters (one-way
@@
 errors of
-    306 
-    [108 - 
-    607].
+%    d[,round(median(I_error))] 
+%    [d[,round(quantile(I_error, 0.25))] - 
+%    d[,round(quantile(I_error, 0.75))]].
 These errors were significantly higher when the true value of $\alpha$ was
 at least 1
-    (Wilcoxon rank-sum test, $p$ = $6\!\times\!10^{-4}$)
+%    (Wilcoxon rank-sum test, $p$ = $latexSN(round(alpha.I.test$p.value, 4))$)
 and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
@@
 The $m$ parameter was estimated correctly in only
-    37 \%
+%    as.integer(m.tbl[1] * 100) \%
 of simulations. Oddly, the error in the estimated $m$ was higher for integer
-values of $\alpha$ (\text{i.e.}~0 and 1) than non-integer values 
-    (Wilcoxon rank-sum test, $p$ = 0.007).
+values of $\alpha$ (\textit{i.e.}~0 and 1) than non-integer values 
+%    (Wilcoxon rank-sum test, $p$ = round(alpha.m.test$p.value, 3)).
 The true values of the other parameters did not significantly affect the
@@
 was consistently over-estimated by about a factor of two
-    (error 6588 
-    [4214 - 
-     8284]).
+%    (error d[,round(median(N_error))] 
+%    [d[,round(quantile(N_error, 0.25))] - 
+%     d[,round(quantile(N_error, 0.75))]]).
 No parameters influenced the accuracy of the $N$ estimates (all one-way ANOVA).
@@
 $\alpha$, 
-  1 
-  [0.07-
-   1.07];
+  1.1 
+  [0.6 -
+   1.15];
 $I$,
-  1170 
-  [506 -
-   1616];
+  1141 
+  [662 -
+   4455];
 $m$,
-  5 
+  3 
   [2 -
@@
 $N$,
-  \ensuremath{1.033\times 10^{4}} 
-  [2408-
-   \ensuremath{1.4974\times 10^{4}}].
+  12689 
+  [3733-
+   14977].
 To test the effect of sampling bias, we sampled one transmission tree in a
@@
 $\alpha$, 
-  0.48 
-  [0.03 -
-   0.82];
+  0.36 
+  [0.01 -
+   0.63];
 $I$,
-  2468 
-  [1281 -
-   3938];
+  2392 
+  [1419 -
+   3767];
 $m$,
@@
 $N$,
-  \ensuremath{1.0738\times 10^{4}} 
-  [3106 -
-   \ensuremath{1.4783\times 10^{4}}].
+  8760 
+  [2881 -
+   14788].
 Both of these results were in line with estimates obtained on other simulated
-datasets (Table~\ref{tab:abchpd}).
+datasets (\cref{tab:abchpd}).
 
@@
 
-We applied kernel-ABC to five published HIV datasets (Table~\ref{tab:data}),
+We applied kernel-ABC to five published HIV datasets (\cref{tab:data}),
 and found substantial heterogeneity among the parameter estimates
-(Figure~\ref{fig:abchpd}). Two of the datasets~\citep{niculescu2015recent,
-wang2015targeting} had estimated $\alpha$ values near unity (MAP estimates
+(\cref{fig:abchpd,fig:abchpdm2}). Two of the
+datasets~\autocite{niculescu2015recent, wang2015targeting} had estimated
+$\alpha$ values near unity for the prior allowing $m = 1$ (MAP estimates
 [95\% HPD] 
-  1.06 
-  [0.63 - 
+  1.05 
+  [0.04 - 
    1.27]
 and
-  1 
-  [0.41 -
-   1.16] respectively). 
-Another two datasets~\citep{li2015hiv, cuevas2009hiv} had lower estimated
-values and wider HPD intervals
-  (0.77 
+  0.84 
   [0.01 -
-  1.03]
+   1.02] respectively).
+The MAP estimates did not change appreciably when $m = 1$ was disallowed by the
+prior, although the credible interval of the \textcite{niculescu2015recent}
+data was narrower
+  (0.04 - 
+   1.27).
+When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
+low estimated $\alpha$ values
+  (0.06 
+  [0.01 - 
+  0.73]
 and
-  0.66 
-  [0.03 -
-   0.84]). 
-The \citet{novitsky2014impact} data had an extremely low estimated $\alpha$
-and a very wide HPD interval
-  (0.17 
-  [0.04 -
-   1.39]). 
-For all the datasets except \citeauthor{novitsky2014impact}, estimated values of
-$I$ were below 2000, with narrow HPD intervals around two of the
-datasets
-  (\citeauthor{cuevas2009hiv}, 880 
-  [290 -
-   1511];
-   \citeauthor{niculescu2015recent}, 175
-  [138 - 
-   454])
-and wider intervals around the other two
-  (\citeauthor{li2015hiv}, 1590 
-  [284 -
-   3807];
-   \citeauthor{wang2015targeting}, 651
-  [268 - 
-   4235]).
-The \citeauthor{novitsky2014impact} data was again the outlier, with a very high
+  0.19 
+  [0.01 -
+   0.8]). 
+However, the MAP estimates increased when $m = 1$ was not permitted, although
+the HPD intervals remained roughly the same
+  (0.78 
+  [0.02 - 
+  0.94]
+and
+  0.59 
+  [0.07 -
+   0.95]).
+The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
+for both priors on $m$
+  (0.32 for $m \geq 1$;
+   0.39 for $m \geq 2$).
+However, the confidence interval was much wider when $m = 1$ was allowed
+  ([0.04 -
+    1.62] for $m \geq 1$ vs.
+    0 -
+    0.73 for $m \geq 2$).
+
+For all the datasets except \citeauthor{novitsky2014impact}, estimated values
+of $I$ were below 2000 when $m = 1$ was allowed, with relatively narrow HPD
+intervals compared to the nonzero prior density region
+  (\citeauthor{cuevas2009hiv}, 482 
+  [293 -
+   2111];
+   \citeauthor{niculescu2015recent}, 307
+  [136 - 
+   2822];
+  (\citeauthor{li2015hiv}, 1183 
+  [413 -
+   2897];
+   \citeauthor{wang2015targeting}, 719
+  [176 - 
+   2114]).
+The \citeauthor{novitsky2014impact} data was the outlier, with a very high
 estimated $I$, and HPD interval spanning almost the entire prior region
-  (7547 
-  [228 -
-   8921]).
-No information was gleaned about the $m$ parameter, with the HPD interval
-occupying the entire prior region for all datasets. The estimates of $N$ were
-similarly uninformative, with the exception that the point estimate for the
-\citeauthor{wang2015targeting} data was smaller
-  (5839)
-than the estimates for other datasets
- (average 8927).
+  (7409 
+  [187 -
+   8819]).
+The $I$ estimates and HPD intervals were generally robust to the choice of
+prior on $m$, with slightly narrower HPD intervals (compare
+\cref{fig:abchpd,fig:abchpdm2}).
+
+The MAP estimate of $m$ was equal to 1 for all but the
+\citeauthor{novitsky2014impact} data, when this value was allowed. However, the
+upper bound of the HPD interval was different for each dataset
+  (\citeauthor{niculescu2015recent}, 5;
+   \citeauthor{wang2015targeting}, 4;
+   \citeauthor{li2015hiv}, 1;
+   \citeauthor{cuevas2009hiv}, 2).
+When $m = 1$ was disallowed, the MAP for all datasets was either 2 or 3, with
+HPD intervals spanning the entire prior region. The estimates for the total
+number of nodes $N$ were largely uninformative for all samples, with almost all
+MAP estimates greater than 5000 and HPD intervals spanning almost the entire
+nonzero prior density region. The only exception was the \citeauthor{li2015hiv}
+data, for which the MAP estimate was lower 
+  (6973)
+when $m = 1$ was allowed.
 
@@
 simulation again on the new, larger network could produce the exact same
-transmission tree as before.
+transmission tree as before. We note also that our accurate estimates of $I$
+may have been influenced by this prior, which places more mass on low $I$
+values. However, the MAP estimate of $I$ was very high for the
+\textcite{novitsky2013phylogenetic,novitsky2014impact} data, suggesting that a
+strong enough signal in the data can overcome the prior.
 
-As noted by \citet{lintusaari2016identifiability}, uniform priors on model
+As noted by \textcite{lintusaari2016identifiability}, uniform priors on model
 parameters may translate to highly informative priors on quantities of
-interest. We observed a non-linear relationship between the preferential
-attachment power $\alpha$ and the power law exponent $\gamma$ (Figure~S9).
-Therefore, placing a uniform prior on $\alpha$ between 0 and 2 is equivalent to
-placing an informative prior that $\gamma$ is close to 2. Therefore, if we were
-primarily interested in $\gamma$ rather than $\alpha$, a more sensible choice
-of prior might have a shape similar to the inverse of Figure~S9 and be bounded
-above by approximately $\alpha$ = 1.5. This would uniformly bound $\gamma$ in
-the region $2 \leq \gamma \leq 4$ commonly reported in the network
-literature~\citep{liljeros2001web, schneeberger2004scale, colgate1989risk,
-brown2011transmission}. We note however that \citet{jones2003assessment}
-estimated $\gamma$ values greater than four, in one case as high as 17, for
-some datasets, indicating that a wider range of permitted $\gamma$ values may
-be warranted.
+interest. We observed a non-linear relationship between the pcreferential
+attachment power $\alpha$ and the power law exponent $\gamma$
+(\cref{fig:gamma}). Thecrefore, placing a uniform prior on $\alpha$ between 0
+and 2 is equivalent to placing an informative prior that $\gamma$ is close to
+2. Thecrefore, if we were primarily interested in $\gamma$ rather than
+$\alpha$, a more sensible choice of prior might have a shape similar to the
+inverse of \cref{fig:gamma} and be bounded above by approximately $\alpha$ =
+1.5. This would uniformly bound $\gamma$ in the region $2 \leq \gamma \leq 4$
+commonly reported in the network literature~\autocite{liljeros2001web,
+schneeberger2004scale, colgate1989risk, brown2011transmission}. We note however
+that \textcite{jones2003assessment} estimated $\gamma$ values greater than
+four, in one case as high as 17, for some datasets, indicating that a wider
+range of permitted $\gamma$ values may be warranted.
 
 Our investigation of published HIV datasets indicated heterogeneity in the
-contact network structures underlying several distinct local epidemics. The
-five datasets analysed fell into three categories (Figure~\ref{fig:abchpd}).
-First, we estimated a preferential attachment power between 0.5 and 1 for the
-epidemics studied by \citet{cuevas2009hiv} and \citet{li2015hiv}, with credible
-intervals occupying nearly the entire region from 0 to 1.
-\citeauthor{cuevas2009hiv} studied a group of newly diagnosed individuals in
-the Basque Country, Spain. Although the individuals were of mixed risk groups,
-and therefore unlikely to comprise a single contact network, a high proportion
-of them (47\%) grouped into local clusters based on genetic distance. The low
-estimated attachment power for these data is consistent with the sampled
-sequences comprising many distinct sub-networks rather than a single connected
-network. \citeauthor{li2015hiv} sampled a large number of acutely infected
-MSM in Shanghai, China, in which we identified a large cluster from the
-phylogeny using a patristic distance cutoff~\citep{poon2015impact}. The low
-attachment power estimated for this dataset was surprising given the high
-phylogenetic relatedness of the sequences. It is possible that the number and
-diversity of circulating recombinant forms in the data introduced errors into
-our estimated viral phylogeny.
+contact network structures underlying several distinct local epidemics. When
+interpreting these results, we caution that the BA model is quite simple and
+most likely misspecified for these data. In particular, the average degree of a
+node in the network is equal to $2m$, and thecrefore is constrained to be a
+multiple of 2. Furthermore, we considered the case $m = 1$, where the network
+has no cycles, to be implausible and thecrefore assigned it zero prior
+probability in one set of analyses. This forced the average degree to be at
+least four, which may be unrealistically high for sexual networks. The fact
+that the estimated values of $\alpha$ differed substatially for three datasets
+depending on whether or not $m = 1$ was allowed by the prior is futher evidence
+of this potential misspecification. However, we note that for two of the
+datasets, the estimated values of $\alpha$ did not change much between priors,
+and the estimates of $I$ were robust to the choice of prior for all datasets
+studied. More sophisticated models, for example models incorporating
+heterogeneity in node behaviour, are likely to provide a better fit to these
+data.
 
-For the outbreaks studied by \citet{niculescu2015recent} and
-\citet{wang2015targeting}, the estimated $\alpha$ was close to one, with a
-narrower credible interval than for the other studies.
-\citeauthor{niculescu2015recent} studied a recent outbreak among Romanian
-injection drug users (IDU), while \citeauthor{wang2015targeting} sampled
-acutely infected MSM in Beijing, China. Both studies discovered a high degree
-of phylogenetic relatedness owing to the recent infection times and homogeneous
-risk groups of the studied populations. The estimated number of infections for
-these datasets were also quite low, although the HPD interval for
-\citeauthor{wang2015targeting} was much wider than that for
-\citeauthor{niculescu2015recent}.
+With respect to the pcreferential attachment power $\alpha$, the five datasets
+analysed fell into three categories (\cref{fig:abchpd}). First, we
+estimated a pcreferential attachment power close to 1, indicating linear
+pcreferential attachment, for the outbreaks studied by
+\textcite{niculescu2015recent} and \textcite{wang2015targeting}. These values
+were robust to specifying different priors for $m$. Both studies were of
+populations in which we would expect a high degree of epidemiological
+relatedness: \textcite{niculescu2015recent} studied a recent outbreak among
+Romanian injection drug users (IDU), while \citeauthor{wang2015targeting}
+sampled acutely infected MSM in Beijing, China. Both these are contexts in
+which we would expect some of the assumptions of the BA model, such as a
+connected network, relatively high mean degree, and pcreferential attachment
+dynamics, to hold.
 
-The final studied dataset was an outlier in terms of estimated parameters.
-\citet{novitsky2013phylogenetic} sampled approximately 44\% of the
-HIV-infected individuals in the northern area of Mochudi, Botswana. Additional
-sampling in a later study~\citep{novitsky2014impact} brought the genotyping
-coverage up to 70\%. Even with such a high sampling coverage, we did not detect
-any large clusters using patristic distance, and therefore chose to analyze a
-subtree instead. Estimates of $\alpha$ and $N$ both had very wide HPD
-intervals and were markedly different from the other datasets. The estimated
-number of infected nodes was also extremely high, much higher than the
-estimated HIV prevalence of the town. Several factors may have contributed to
-these results. First, the authors note that the their sample was 75\% female.
-In a primarily heterosexual risk environment, removal of a disproportionate
-number of males from the network could obfuscate the true network structure,
-for example if the majority of highly connected nodes were of one gender.
-Second, the town in question was in close proximity to the country's capital,
-and the authors indicated that a high amount of migration takes place between
-the two locations. This suggests that the contact network may include a much
-larger group based in the capital city, which would explain the high estimate
-of $I$.
+The remaining three datasets (\textcite{cuevas2009hiv, novitsky2014impact,
+li2015hiv}) had estimated values of $\alpha$ below 0.5 when $m = 1$ was
+included in the prior, but these were not robust to changing the prior to
+exclude $m = 1$. For the \citeauthor{cuevas2009hiv} data, model
+misspecification is likely partially responsible. While the authors found that
+a large proportion of the samples were epidemiologically linked, these were
+mainly in small local clusters rather than the single large component
+postulated by the BA model. In addition, the mixed risk groups in the dataset
+would be unlikely to significantly interact, further weakening any global
+pcreferential attachment dynamics. The dataset studied by
+\textcite{novitsky2014impact} originated from a densely sampled population
+where the predominant risk factor was believed to be heterosexual exposure.
+Although the MAP estimate of $\alpha$ was almost unchanged when the value $m =
+1$ was excluded from the prior, the confidence interval shrank significantly.
+For both priors, the estimated prevalence was extremely high, in fact higher
+than the estimated HIV prevalence in the sampled region. The authors indicated
+that the source of the samples was a town in close proximity to the country's
+capital city, and suggested that there may have been a high degree of migration
+and partner interchange between the two locations. It is possible that the
+contact network underlying the subtree we investigated includes a much larger
+group based in the capital city, which would explain the high estimate of $I$.
+There is no clear explanation for the discrepancy between the two priors for
+the \textcite{li2015hiv} data, as the subset we analyzed formed a phylogenetic
+cluster and thecrefore was a good candidate for the BA model. However, nearly
+all the posterior density was assigned to $m = 1$ when this value was allowed,
+indicating that the network was more likely to have an acyclic tree structure.
 
-When interpreting these results, we caution that the BA model is quite
-simple and most likely misspecified for these data. In particular, the average
-degree of a node in the network is equal to $2m$, and therefore is constrained
-to be a multiple of 2. Furthermore, we considered the case $m = 1$, where the
-network has no cycles, to be implausible and assigned it zero prior
-probability. This forces the average degree to be at least four, which may be
-unrealistically high for sexual networks. Additional modelling assumptions
-include the network being connected and static, all transmission rates being
-equal, no removal after infection, and identical behaviour of all nodes. This
-last is particularly problematic, as we showed by simulating a network where
-some nodes exhibited a higher attachment power than others. The estimated
-attachment power was simply the average of the two values, indicating that,
-although we could characterize the network in aggregate, the estimated
-parameters could not be said to apply to any individual node. Despite these
-issues, we felt it was best to demonstrate the method first on a simple model.
-It is possible to use this framework to fit more complex models which address
-some of these issues, such as one incorporating heterogeneous node behaviour,
-which may prove a fruitful avenue for future investigations.
+In addition to the aforementioned possibility of misspecification, additional
+modelling assumptions include the network being connected and static, all
+transmission rates being equal, no removal after infection, and identical
+behaviour of all nodes. This last is particularly problematic, as we showed by
+simulating a network where some nodes exhibited a higher attachment power than
+others. The estimated attachment power was simply the average of the two
+values, indicating that, although we could characterize the network in
+aggregate, the estimated parameters could not be said to apply to any
+individual node. Despite these issues, we felt it was best to demonstrate the
+method first on a simple model. It is possible to use this framework to fit
+more complex models which address some of these issues, such as one
+incorporating heterogeneous node behaviour, which may prove a fruitful avenue
+for future investigations.
 

### 2016-04-22 doc/thesis/intro.tex
Stuff to work on for dynamics

@@
 research~\autocite{cottam2008integrating, jombart2011reconstructing,
-ypma2012unravelling, didelot2014bayesian}.
+ypma2012unravelling, morelli2012bayesian, didelot2014bayesian}.
 
@@
 trajectories when there is substantial heterogeneity
-present~\autocite{bansal2007individual, volz2007susceptible}. Contact networks
-provide a way to relax this assumption by representing individuals and their
-contacts explicitly. It is important to note that, although panmixia is an
-unrealistic modelling assumption, it has not proven a substantial hurdle to
-epidemic modelling in practice~\autocite{anderson1992infectious}. Using this
-assumption, researchers have been able to derive estimates of the transmission
-rate and the basic reproductive number of various outbreaks, which have agreed
-with values obtained by on-the-ground data collection. Therefore, if one is
-interested only in these population-level variables, the additional complexity
-of contact network models may not be warranted. Rather, these models are most
-useful when we are interested in properties of the network itself, such as
-centrality, structural balance, and
-transitivity~\autocite{wasserman1994social}.
+present~\autocite{bansal2007individual, volz2007susceptible,
+rolls2015simulation}. Contact networks provide a way to relax this assumption
+by representing individuals and their contacts explicitly. It is important to
+note that, although panmixia is an unrealistic modelling assumption, it has not
+proven a substantial hurdle to epidemic modelling in
+practice~\autocite{anderson1992infectious}. Using this assumption, researchers
+have been able to derive estimates of the transmission rate and the basic
+reproductive number of various outbreaks, which have agreed with values
+obtained by on-the-ground data collection. Therefore, if one is interested only
+in these population-level variables, the additional complexity of contact
+network models may not be warranted. Rather, these models are most useful when
+we are interested in properties of the network itself, such as centrality,
+structural balance, and transitivity~\autocite{wasserman1994social}.
 
@@
 for all $x \in X$. This is an ``operational'' definition of Markov kernel which
-will be suitable for our purposes. Note that a Markov kernels have nothing to
-do with the kernel functions defined in \cref{subsec:treeshape}, other than
-sharing a name (the word ``kernel'' is ubiquitous in mathematics).
+will be suitable for our purposes. A more rigorous definition can be found in
+\eg \autocite{kallenberg2006foundations}. Note that a Markov kernels have
+nothing to do with the kernel functions defined in \cref{subsec:treeshape},
+other than sharing a name (the word ``kernel'' is ubiquitous in mathematics).
 
@@
 independent.
-
-TODO
 

### 2016-04-22 doc/thesis/methods.tex
Stuff to work on for dynamics

@@
 
+Two further experiments were performed to address potential sources of error.
+To evaluate the effect of model misspecification in the case of heterogeneity
+among nodes, we generated a network where half the nodes were attached with
+power $\alpha$ = 0.5, and the other half with power $\alpha$ = 1.5. The other
+parameters for this network were $N$ = 5000, $I$ = 1000, and $m$ = 2. To
+investigate the effects of potential sampling
+bias~\autocite{karcher2016quantifying}, we simulated a transmission tree where
+the tips were sampled in a peer-driven fashion, rather than at random. That is,
+the probability to sample a node was twice as high if any of that node's
+network peers had already been sampled. The parameters of this network were $N$
+= 5000, $I$ = 2000, $m$ = 2, and $\alpha$ = 0.5.
+
 \subsection{Real data experiments}

### 2016-04-22 doc/thesis/objective.tex
Stuff to work on for dynamics

@@
+\glsreset{ABC}
+
 The spread of a disease is most often modelled by assuming either a
@@
 into contact causing transmission. Although this provides a reasonable
-approximation in many cases, the error introduced by assuming a panmictic
-population can be substantial when significant contact heterogeneity exists in
-the underlying population. Contact network models provide an alternative to
-compartmental models which do not require the assumption of panmixia. In
-addition to more accurate predictions, the parameters of the networks
-themselves may be of interest from a public health perspective. For example,
-certain vaccination strategies may be more or less effective in curtailing an
-epidemic depending on the underlying network's degree distribution.
-Phylodynamic methods have been used to fit many different types of model to
-phylogenetic data, but as far as we know, no methods have yet been developed to
-fit contact network models. The primary objective of this work is to develop
-such a method.
+approximation in many cases~\autocite{anderson1992infectious}, the error
+introduced by assuming a panmictic population can be substantial when
+significant contact heterogeneity exists in the underlying population. Contact
+network models provide an alternative to compartmental models which do not
+require the assumption of panmixia. In addition to more accurate predictions,
+the parameters of the networks themselves may be of interest from a public
+health perspective. For example, certain vaccination strategies may be more or
+less effective in curtailing an epidemic depending on the underlying network's
+degree distribution~\autocite{peng2013vaccination}. Phylodynamic methods have
+been used to fit many different types of model to phylogenetic
+data~\autocite{pybus2009evolutionary,volz2013viral}, but as far as we know, no
+methods have yet been developed to fit contact network models. The primary
+objective of this work is to develop such a method.
 
@@
 likely to be an intractable problem, which would imply that these models are
-amenable to neither \gls{ML} nor Bayesian inference. We not proven this is the
-case, but some intuition can be provided by examining the process involved in
-the likelihood calculation. Consider a contact network model with parameters
-$\theta$, and an observed transmission tree $T$ with $n$ tips.  In general, we
+amenable to neither \gls{ML} nor Bayesian inference. We have not proven this is
+the case, but some intuition can be provided by examining the process involved
+in the likelihood calculation. Consider a contact network model with parameters
+$\theta$, and an estimated transmission tree $T$ with $n$ tips. In general, we
 do not know the labels of the internal nodes of $T$, only the labels of its
@@
 meaning that all infected individuals were sampled. This is rarely the case in
-practice - most often, the observed tree is a subsampled version of the true
-tree. In this case, the likelihood calculation becomes even more complex,
-because we must also sum over all possible complete trees.
+practice - most often, we only have access to a subset of the infected
+individuals. In this case, the likelihood calculation becomes even more
+complex, because we must also sum over all possible complete trees.
 
@@
 networks can profoundly affect transmission tree shape, and those shapes can be
-compared using a highly informative similarity measure. \Gls{SMC} has several
-advantages over other algorithms for \gls{ABC}~\autocite{mckinley2009inference},
-including a recently-developed adaptive algorithm requiring minimal tuning on
-the part of the user~\autocite{del2012adaptive}. In summary, our method to
-infer contact network parameters will combine the following: stochastic
-simulation of epidemics on networks, the tree kernel, and adaptive
+compared using a highly informative similarity measure called the ``tree
+kernel''~\autocite{poon2013mapping}. \Gls{ABC} can be implemented with
+\gls{SMC}, which has several advantages over other
+algorithms~\autocite{mckinley2009inference}. A recently-developed adaptive
+algorithm requiring minimal tuning on the part of the user makes \gls{SMC} an
+even more attractive approach~\autocite{del2012adaptive}. In summary, our
+method to infer contact network parameters will combine the following:
+stochastic simulation of epidemics on networks, the tree kernel, and adaptive
 \gls{ABC}-\gls{SMC}. Since our distance measure is a kernel function, our
@@
 Empirical studies of sexual contact networks have found that these networks
-tend to be scale-free, meaning that their degree distributions follow a power
-law (although there has been some disagreement). Preferential attachment has
-been postulated as a mechanism by which scale-free networks could be generated.
-This makes the \gls{BA} model, one of the simplest preferential attachment
-models, a natural choice to explore with our method. The second aim of this
-work is to use simulations to investigate the parameters of the \gls{BA} model,
-including whether they have a detectable impact on tree shape, and whether they
-can be accurately recovered using kernel-\gls{ABC}.
+tend to be scale-free~\autocite{colgate1989risk, liljeros2001web,
+schneeberger2004scale}, meaning that their degree distributions follow a power
+law (although there has been some disagreement, see
+\autocite{handcock2004likelihood, bansal2007individual}). Preferential
+attachment has been postulated as a mechanism by which scale-free networks
+could be generated~\autocite{barabasi1999emergence}. This makes the \gls{BA}
+model, one of the simplest preferential attachment models, a natural choice to
+explore with our method. The second aim of this work is to use simulations to
+investigate the parameters of the \gls{BA} model, including whether they have a
+detectable impact on tree shape, and whether they can be accurately recovered
+using kernel-\gls{ABC}.
 
@@
 any other pathogen, and including sequences sampled from diverse geographic and
-demographic contexts. Since \gls{HIV} is almost always spread through either
-sexual contact or sharing of injection drug supplies, the contact networks
-underlying \gls{HIV} epidemics are highly structured. Moreover, since no cure
-yet exists, efforts to curtail the progression of an epidemic have relied on
-preventing further transmissions through measures such as \gls{tasp} and
-education leading to behaviour change. The effectiveness of this type of
-intervention can vary significantly based on the underlying structure of the
-network and the particular nodes to whom the intervention is targeted. Due to
-this combination of data availability and potential public health impact,
-\gls{HIV} is an obvious context in which our method could be applied.
-Therefore, the third and final aim of this work is to apply kernel-\gls{ABC} to
-fit the \gls{BA} model to existing \gls{HIV} outbreaks.
+demographic contexts. At the time of this writing, there were $635400$ HIV
+sequences publicly available in GenBank, annotated with 172 distinct countries
+of origin. Since \gls{HIV} is almost always spread through either sexual
+contact or sharing of injection drug supplies, the contact networks underlying
+\gls{HIV} epidemics are driven by social dynamics and are therefore likely to
+be highly nonrandom. Moreover, since no cure yet exists, efforts to curtail the
+progression of an epidemic have relied on preventing further transmissions
+through measures such as \gls{tasp} and education leading to behaviour change.
+The effectiveness of this type of intervention can vary significantly based on
+the underlying structure of the network and the particular nodes to whom the
+intervention is targeted. Due to this combination of data availability and
+potential public health impact, \gls{HIV} is an obvious context in which our
+method could be applied. Therefore, the third and final aim of this work is to
+apply kernel-\gls{ABC} to fit the \gls{BA} model to existing \gls{HIV}
+outbreaks.
 

### 2016-04-22 doc/thesis/preface.tex
Stuff to work on for dynamics

@@
 A version of chapter 2 has been submitted for publication in Virus Evolution
-with the title ``Reconstructing network parameters from viral phylogenies.'' A
-presentation entitled ``Phylodynamic inference of contact network parameters
-with kernel-ABC'' was given based on chapter 2 at the 23rd HIV Dynamics and
-Evolution meeting on April 25, in Woods Hole, Massachussets, USA.
+with the title ``Reconstructing network parameters from viral phylogenies.'' An
+oral presentation entitled ``Phylodynamic inference of contact network
+parameters with kernel-ABC'' was given based on chapter 2 at the 23rd HIV
+Dynamics and Evolution meeting on April 25, in Woods Hole, Massachussets, USA.
+A poster entitled ``Likelihood-free estimation of contact network parameters
+from viral phylogenies'' was presented at the Intelligent Systems for Molecular
+Biology meeting in Orlando, Florida, USA.
 

### 2016-04-22 doc/thesis/results.tex
Stuff to work on for dynamics

@@
     (Spearman's $\rho$ = 0.43,
-    $p {<}10^{-5}$).
+    $p\ {<}10^{-5}$).
 
@@
 \subsection{Application to HIV data}
-
-We applied kernel-ABC to five published HIV datasets (\cref{tab:data}),
-and found substantial heterogeneity among the parameter estimates
-(\cref{fig:abchpd}). Two of the datasets~\autocite{niculescu2015recent,
-wang2015targeting} had estimated $\alpha$ values near unity (MAP estimates
-[95\% HPD] 
-  1.06 
-  [0.63 - 
-   1.27]
-and
-  1 
-  [0.41 -
-   1.16] respectively). 
-Another two datasets~\autocite{li2015hiv, cuevas2009hiv} had lower estimated
-values and wider HPD intervals
-  (0.77 
-  [0.01 - 
-  1.03]
-and
-  0.66 
-  [0.03 -
-   0.84]). 
-The \textcite{novitsky2014impact} data had an extremely low estimated $\alpha$
-and a very wide HPD interval
-  (0.17 
-  [0.04 -
-   1.39]). 
-For all the datasets except \citeauthor{novitsky2014impact}, estimated values of
-$I$ were below 2000, with narrow HPD intervals around two of the
-datasets
-  (\citeauthor{cuevas2009hiv}, 880 
-  [290 -
-   \ensuremath{1.51\times 10^{3}}];
-   \citeauthor{niculescu2015recent}, 175
-  [138 - 
-   454])
-and wider intervals around the other two
-  (\citeauthor{li2015hiv}, \ensuremath{1.59\times 10^{3}} 
-  [284 -
-   \ensuremath{3.81\times 10^{3}}];
-   \citeauthor{wang2015targeting}, 651
-  [268 - 
-   \ensuremath{4.24\times 10^{3}}]).
-The \citeauthor{novitsky2014impact} data was again the outlier, with a very high
-estimated $I$, and HPD interval spanning almost the entire prior region
-  (\ensuremath{7.55\times 10^{3}} 
-  [228 -
-   \ensuremath{8.92\times 10^{3}}]).
-No information was gleaned about the $m$ parameter, with the HPD interval
-occupying the entire prior region for all datasets. The estimates of $N$ were
-similarly uninformative, with the exception that the point estimate for the
-\citeauthor{wang2015targeting} data was smaller
-  (\ensuremath{5.84\times 10^{3}})
-than the estimates for other datasets
- (average \ensuremath{8.93\times 10^{3}}).
-
-\begin{figure*}[ht]
-  \centering
-  \includegraphics{realdata-hpd-bc.pdf}
-  \vspace{8pt}
-  \caption{
-      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
-      parameters of the BA network model, fitted to five published HIV datasets
-      with kernel-ABC.
-  }
-  \label{fig:abchpd}
-\end{figure*}

### 2016-04-25 doc/thesis/discussion.tex
Writing

--- 2016-04-16 doc/thesis/discussion.tex
@@
 
-% spatial networks, see section 5.4 in keeling2005networks
+Several alternative network models and modeling frameworks have been developed
+which may provide useful future targets for kernel-\gls{ABC}. Waring
+models~\autocite{irwin1963place,handcock2004likelihood} are a more flexible
+type of preferential attachment model which permit a subset of attachments to
+be formed non-prefentially. These models were used by
+\textcite{brown2011transmission} to characterize the transmission network in
+the United Kingdom. \Glspl{ERGM}~\autocite{robins2007introduction} are a
+flexible and expressive parameterization of contact networks in terms of
+statistics of network features such as pairs and triads.
+\textcite{goodreau2007assessing} evaluated the effect of several different
+\gls{ERGM} parameterizations on transmission tree shape and effective
+population size. The author suggested the use of \glspl{ERGM} as a general
+framework for estimation of epidemiological quantities related to \gls{HIV}
+transmission. Except for a few special cases, simulating a network according to
+an \gls{ERGM} generally requires \gls{MCMC}, which would be too computationally
+intensive to integrate into \software{netabc} as it currently stands. To fit
+\gls{ERGM} with kernel-\gls{ABC}, one possibility would be to consider the
+network itself as a parameter to be modified by the \gls{MCMC} kernel.
+Other network modelling frameworks include the partnership-centric formulation
+developed by~\textcite{eames2002modeling} and the log-linear adjacency matrix
+parameterization developed by~\textcite{morriz1993epidemiology}.
+
+%\Textcite{morris1993epidemiology} proposes to apply the standard compartmental
+%modelling framework to contact networks by assigning each individual their own
+%compartment. Thus, each individual is associated with a single \gls{ODE}, with
+%the entire \gls{ODE} system parameterized by the adjacency matrix of the
+%contact network. 
+%The author proposes to use log-linear models to parameterize the adjacency
+%5matrix of a contact network. This framework is highly expressive, and allows
+%straightforward incorporation of time-dependent dynamics. 
+%However, simulating a transmission tree would require the numerical solution of
+%a very large system of \glspl{ODE}, which may be computationally prohibitive
+%given the number of simulations required for kernel-\gls{ABC}.
 
@@
 assumption is invalid, as people make and break partnerships on a regular
-basis. Our work has not addressed this assumption, primarily due to our desire
-to avoid the additional complexity required to address the dynamic nature of
-networks. This simplifying assumption is made by most studies using contact
-network models in an epidemiological context~\autocite{welch2011statistical,
-bansal2007individual}. However, in principle, kernel-\gls{ABC} could be
-adapted to dynamic contact networks by using a method such as that developed by
-\textcite{robinson2012dynamics} to simulate a dynamic contact network, while
-concurrently simulating the spread of an epidemic.
-
-An alternative approach is the deterministic framework outlined by
-\textcite{morris1993epidemiology}, who proposes to apply the standard
-compartmental modelling framework to contact networks by assigning each
-individual their own compartment. Thus, each individual is associated with a
-single \gls{ODE}, with the entire \gls{ODE} system parameterized by the
-adjacency matrix of the contact network. \citeauthor{morris1993epidemiology}
-proposes to use log-linear models to parameterize the matrix. This framework is
-highly expressive, and allows straightforward incorporation of time-dependent
-dynamics. However, simulating a transmission tree would require the numerical
-solution of a very large system of \glspl{ODE}. Given the large number of
-simulations required for kernel-\gls{ABC}, it is not clear if this method would
-be computationally feasible in this context.
+basis. Addressing the impact of this simplifying assumption is outside the
+scope of this work. However, the same assumption is made by most studies using
+contact network models in an epidemiological
+context~\autocite{welch2011statistical, bansal2007individual}. In principle,
+kernel-\gls{ABC} could be adapted to dynamic contact networks by using a method
+such as that developed by \textcite{robinson2012dynamics} to simulate such a
+network, while concurrently simulating the spread of an epidemic.
 
@@
 input, rather than a viral phylogeny. Thus, we have left the estimation of a
-transmission tree up to the user. There were two reasons for this choice. First
-and foremost, we wished again to avoid extra complexity and keep the number of
-estimated parameters small. In theory, it is possible to incorporate the
+transmission tree up to the user. In theory, it is possible to incorporate the
 process by which a viral phylogeny is generated along with a transmission tree
@@
 primary purpose of this work, which is to study contact network parameters.
-Second, there are a number of different methods available for inferring
+In addition, there are a number of different methods available for inferring
 transmission trees~\autocite{didelot2014bayesian, ypma2012unravelling,
@@
 their own preferred tree building method.
+
+Our implementation of \gls{SMC} uses a simple multinomial scheme to sample
+particles from the population according to their weights. Several other
+sampling strategies have been developed~\autocite{douc2005comparison}, and it
+is possible that the use of a more sophisticated technique might increase the
+algorithm's accuracy.
 
@@
 
+As noted by \textcite{lintusaari2016identifiability}, uniform priors on model
+parameters may translate to highly informative priors on quantities of
+interest. We observed a non-linear relationship between the pcreferential
+attachment power $\alpha$ and the power law exponent $\gamma$
+(\cref{fig:gamma}). Thecrefore, placing a uniform prior on $\alpha$ between 0
+and 2 is equivalent to placing an informative prior that $\gamma$ is close to
+2. Thecrefore, if we were primarily interested in $\gamma$ rather than
+$\alpha$, a more sensible choice of prior might have a shape informed by
+\cref{fig:gamma} and be bounded above by approximately $\alpha$ = 1.5. This
+would uniformly bound $\gamma$ in the region $2 \leq \gamma \leq 4$ commonly
+reported in the network literature~\autocite{liljeros2001web,
+schneeberger2004scale, colgate1989risk, brown2011transmission}. We note however
+that \textcite{jones2003assessment} estimated $\gamma$ values greater than
+four for some datasets, in one case as high as 17, indicating that a wider
+range of permitted $\gamma$ values may be warranted.
+
 \subsection{Application to HIV data}
 
-The analysis of \gls{HIV} datasets indicated heterogeneity in the contact
-network structures underlying several distinct local epidemics. The five
-datasets analysed fell into three categories (Figure~\ref{fig:abchpd}). First,
-we estimated a preferential attachment power between 0.5 and 1 for the
-epidemics studied by \textcite{cuevas2009hiv} and \textcite{li2015hiv}, with
-credible intervals occupying nearly the entire region from 0 to 1.
-\citeauthor{cuevas2009hiv} studied a group of newly diagnosed individuals in
-the Basque Country, Spain. Although the individuals were of mixed risk groups,
-and therefore unlikely to comprise a single contact network, a high proportion
-of them (47\%) grouped into local clusters based on genetic distance. The low
-estimated attachment power for these data is consistent with the sampled
-sequences comprising many distinct sub-networks rather than a single connected
-network. \citeauthor{li2015hiv} sampled a large number of acutely infected MSM
-in Shanghai, China, in which we identified a large cluster from the phylogeny
-using a patristic distance cutoff~\autocite{poon2015impact}. The low attachment
-power estimated for this dataset was surprising given the high phylogenetic
-relatedness of the sequences. It is possible that the number and diversity of
-circulating recombinant forms in the data introduced errors into our estimated
-viral phylogeny.
+Our investigation of published HIV datasets indicated heterogeneity in the
+contact network structures underlying several distinct local epidemics. When
+interpreting these results, we caution that the BA model is quite simple and
+most likely misspecified for these data. In particular, the average degree of a
+node in the network is equal to $2m$, and thecrefore is constrained to be a
+multiple of 2. Furthermore, we considered the case $m = 1$, where the network
+has no cycles, to be implausible and thecrefore assigned it zero prior
+probability in one set of analyses. This forced the average degree to be at
+least four, which may be unrealistically high for sexual networks. The fact
+that the estimated values of $\alpha$ differed substatially for three datasets
+depending on whether or not $m = 1$ was allowed by the prior is futher evidence
+of this potential misspecification. However, we note that for two of the
+datasets, the estimated values of $\alpha$ did not change much between priors,
+and the estimates of $I$ were robust to the choice of prior for all datasets
+studied. More sophisticated models, for example models incorporating
+heterogeneity in node behaviour, are likely to provide a better fit to these
+data.
 
-For the outbreaks studied by \textcite{niculescu2015recent} and
-\textcite{wang2015targeting}, the estimated $\alpha$ was close to one, with a
-narrower credible interval than for the other studies.
-\citeauthor{niculescu2015recent} studied a recent outbreak among Romanian
-injection drug users (IDU), while \citeauthor{wang2015targeting} sampled
-acutely infected MSM in Beijing, China. Both studies discovered a high degree
-of phylogenetic relatedness owing to the recent infection times and homogeneous
-risk groups of the studied populations. The estimated number of infections for
-these datasets were also quite low, although the HPD interval for
-\citeauthor{wang2015targeting} was much wider than that for
-\citeauthor{niculescu2015recent}.
+With respect to the pcreferential attachment power $\alpha$, the five datasets
+analysed fell into three categories (\cref{fig:abchpd}). First, we
+estimated a pcreferential attachment power close to 1, indicating linear
+pcreferential attachment, for the outbreaks studied by
+\textcite{niculescu2015recent} and \textcite{wang2015targeting}. These values
+were robust to specifying different priors for $m$. Both studies were of
+populations in which we would expect a high degree of epidemiological
+relatedness: \textcite{niculescu2015recent} studied a recent outbreak among
+Romanian injection drug users (IDU), while \citeauthor{wang2015targeting}
+sampled acutely infected MSM in Beijing, China. Both these are contexts in
+which we would expect some of the assumptions of the BA model, such as a
+connected network, relatively high mean degree, and pcreferential attachment
+dynamics, to hold.
 
-The final studied dataset was an outlier in terms of estimated parameters.
-\textcite{novitsky2013phylogenetic} sampled approximately 44\% of the
-HIV-infected individuals in the northern area of Mochudi, Botswana. Additional
-sampling in a later study~\autocite{novitsky2014impact} brought the genotyping
-coverage up to 70\%. Even with such a high sampling coverage, we did not detect
-any large clusters using patristic distance, and therefore chose to analyze a
-subtree instead. Estimates of $\alpha$ and $N$ both had very wide HPD intervals
-and were markedly different from the other datasets. The estimated number of
-infected nodes was also extremely high, much higher than the estimated HIV
-prevalence of the town. Several factors may have contributed to these results.
-First, the authors note that the their sample was 75\% female.  In a primarily
-heterosexual risk environment, removal of a disproportionate number of males
-from the network could obfuscate the true network structure, for example if the
-majority of highly connected nodes were of one gender.  Second, the town in
-question was in close proximity to the country's capital, and the authors
-indicated that a high amount of migration takes place between the two
-locations. This suggests that the contact network may include a much larger
+The remaining three datasets (\textcite{cuevas2009hiv, novitsky2014impact,
+li2015hiv}) had estimated values of $\alpha$ below 0.5 when $m = 1$ was
+included in the prior, but these were not robust to changing the prior to
+exclude $m = 1$. For the \citeauthor{cuevas2009hiv} data, model
+misspecification is likely partially responsible. While the authors found that
+a large proportion of the samples were epidemiologically linked, these were
+mainly in small local clusters rather than the single large component
+postulated by the BA model. In addition, the mixed risk groups in the dataset
+would be unlikely to significantly interact, further weakening any global
+pcreferential attachment dynamics. The dataset studied by
+\textcite{novitsky2014impact} originated from a densely sampled population
+where the predominant risk factor was believed to be heterosexual exposure.
+Although the MAP estimate of $\alpha$ was almost unchanged when the value $m =
+1$ was excluded from the prior, the confidence interval shrank substantially.
+For both priors, the estimated prevalence was extremely high, in fact higher
+than the estimated HIV prevalence in the sampled region. The authors indicated
+that the source of the samples was a town in close proximity to the country's
+capital city, and suggested that there may have been a high degree of migration
+and partner interchange between the two locations. It is possible that the
+contact network underlying the subtree we investigated includes a much larger
 group based in the capital city, which would explain the high estimate of $I$.
+There is no clear explanation for the discrepancy between the two priors for
+the \textcite{li2015hiv} data, as the subset we analyzed formed a phylogenetic
+cluster and thecrefore was a good candidate for the BA model. However, nearly
+all the posterior density was assigned to $m = 1$ when this value was allowed,
+indicating that the network was more likely to have an acyclic tree structure.
 

### 2016-04-25 doc/thesis/intro.tex
Writing

@@
 infected hosts, while internal nodes correspond to transmissions from one host
-to another.Transmission trees generally have branch lengths in units of
+to another. Transmission trees generally have branch lengths in units of
 calendar time, with branching times indicating times of transmission. The root
@@
 coalescent~\autocite{kingman1982coalescent, volz2012complex} models. These
-methods generally assume the viral phylogeny is known, and we assume the same
-in this work (see \eg \autocite{nei2000molecular} for a review of phylogenetic
-inference).
+methods either assume the viral phylogeny is known, as we do in this work, or
+(more commonly) integrate over phylogenetic uncertainty using Bayesian methods.
+Phylogenetic inference is a complex topic which we shall not discuss here; see
+\eg \autocite{nei2000molecular} for a full review.
 
@@
 of correspondence between viral phylogenies and transmission
-trees~\autocite{ypma2013relating, kenah2015algorithms, kenah2016molecular}. In
-particular, the transmission process is quite similar to \defn{allopatric
-speciation}~\autocite{coyne2004speciation}, where genetic divergence follows
-the geographic isolation of a sub-population of organisms. Thus, transmission,
-which is represented as branching in the transmission tree, causes branching in
-the viral phylogeny as well. Similarly, the removal of an individual from the
-transmission tree causes the extinction of their viral lineage in the
-phylogeny. Consequently, the topology of the viral phylogeny is sometimes used
-as a proxy for the topology of the transmission
+trees~\autocite{leitner2002molecular, ypma2013relating, kenah2015algorithms,
+kenah2016molecular}. In particular, the transmission process is quite similar
+to \defn{allopatric speciation}~\autocite{coyne2004speciation}, where genetic
+divergence follows the geographic isolation of a sub-population of organisms.
+Thus, transmission, which is represented as branching in the transmission tree,
+causes branching in the viral phylogeny as well. Similarly, the removal of an
+individual from the transmission tree causes the extinction of their viral
+lineage in the phylogeny. Consequently, the topology of the viral phylogeny is
+sometimes used as a proxy for the topology of the transmission
 tree~\autocite{hall2015epidemic}. Modern likelihood-based methods of
@@
 correspondence between the topologies of the viral phylogeny and transmission
-tree is not necessarily exact~\autocite{ypma2013relating, romero2014timing}.
-Due to intra-host diversity, the viral strain which is transmitted may have
-split from another lineage within the donor long before the transmission event
-occurred. Hence, the branching point in the viral phylogeny may be much earlier
-than that in the transmission tree. Another possibility is that one host
-transmitted to two or more recipients, but the lineages they each received
-originated within the donor host in a different order than that in which the
-transmissions occurred.  In this case, the topology of the transmission tree
-and the viral phylogeny will be mismatched. In practice, this discordance has
-not proven an insurmountable problem: for example,
-\textcite{leitner1996accurate, paraskevis2004phylogenetic} were able to
-accurately recover a known transmission tree using a viral phylogeny. The
-problem of accurately estimating transmission trees is an ongoing area of
+tree is far from exact~\autocite{ypma2013relating, romero2014timing}.  Due to
+intra-host diversity, the viral strain which is transmitted may have split from
+another lineage within the donor long before the transmission event occurred.
+Hence, the branching point in the viral phylogeny may be much earlier than that
+in the transmission tree. Another possibility is that one host transmitted to
+two or more recipients, but the lineages they each received originated within
+the donor host in a different order than that in which the transmissions
+occurred.  In this case, the topology of the transmission tree and the viral
+phylogeny will be mismatched. In practice, this discordance has not proven an
+insurmountable problem: for example, \textcite{leitner1996accurate,
+paraskevis2004phylogenetic} were able to accurately recover a known
+transmission tree using a viral phylogeny. The problem of accurately estimating
+transmission trees is an ongoing area of
 research~\autocite{cottam2008integrating, jombart2011reconstructing,
@@
 \autocite{poon2013mapping} will henceforth be referred to as the \defn{tree
-kernel}. The feature space into which trees are mapped is the space of all
-possible \defn{subset trees}, which are subtrees that do not necessarily extend
-all the way to the tips. The subset-tree kernel was originally developed for
-comparing parse trees in natural language processing~\autocite{collins2002new}
-and did not incorporate branch length information. The version developed by
+kernel}. This kernel maps trees into the space of all possible possible
+\defn{subset trees}, which are subtrees that do not necessarily extend all the
+way to the tips. The subset-tree kernel was originally developed for comparing
+parse trees in natural language processing~\autocite{collins2002new} and did
+not incorporate branch length information. The version developed by
 \textcite{poon2013mapping} includes a radial basis function to compare the
@@
 edges are also useful when the transmission risk is not equal between the
-hosts, such as with HIV transmission, where acting as the receptive partner
-carries a higher risk of infection than acting as the insertive partner. In
-this case, a contact could be represented by two directed edges, one in each
-direction between the two hosts, with the edges annotated by what kind of risk
-they imply~\autocite{wasserman1994social}. An undirected contact network is
+hosts, such as with \gls{HIV} transmission among \gls{MSM}, where the receptive
+partner carries a higher risk of infection than the insertive
+partner~\autocite{baggaley2010hiv}. In this case, a contact could be
+represented by two directed edges, one in each direction between the two hosts,
+with the edges annotated by what kind of risk they
+imply~\autocite{wasserman1994social}. An undirected contact network is
 equivalent to a directed network where each contact is represented by two
@@
 Epidemiological models most often assume some form of contact homogeneity. The
-simplest models, such as the \gls{SIR} model, assume a completely homogeneously
-mixed population, where every pair of contacts is equally likely. More
-sophisticated models partition the population into groups with different
-contact rates between and among each group. However, these models still assume
-that every possible contact between a member of group $i$ and a member of group
-$j$ is equally likely. This assumption is clearly unrealistic for the majority
-of human communities, and can lead to significant errors in predicted epidemic
+simplest models, such as the \gls{SIR} model~\autocite{anderson1992infectious},
+assume a completely homogeneously mixed population, where every pair of
+contacts is equally likely. More sophisticated models partition the population
+into groups with different contact rates between and among each
+group~\autocite{jacquez1988modeling}. However, these models still assume that
+every possible contact between a member of group $i$ and a member of group $j$
+is equally likely. This assumption is clearly unrealistic for the majority of
+human communities, and can lead to significant errors in predicted epidemic
 trajectories when there is substantial heterogeneity
@@
 reproductive number of various outbreaks, which have agreed with values
-obtained by on-the-ground data collection. Therefore, if one is interested only
-in these population-level variables, the additional complexity of contact
-network models may not be warranted. Rather, these models are most useful when
-we are interested in properties of the network itself, such as centrality,
-structural balance, and transitivity~\autocite{wasserman1994social}.
+obtained by on-the-ground data collection~\autocite{stadler2014insights}.
+Therefore, if one is interested only in these population-level variables, the
+additional complexity of contact network models may not be warranted. Rather,
+these models are most useful when we are interested in properties of the
+network itself, such as centrality, structural balance, and
+transitivity~\autocite{wasserman1994social}.
 
@@
 contact network can provide an idea of what to expect as an epidemic unfolds.
-In terms of actionable information, vaccination strategies which would
-eradicate an epidemic in a random network might not work if the network is
-scale-free~\autocite[][see \cref{subsec:pa}]{keeling2005networks}. On a local
-level, contact networks can be informative about the groups or individuals who
-are at highest risk of acquiring or transmitting infection, and would therefore
-benefit most from public health interventions~\autocite{wang2015targeting,
-little2014using}.
+In terms of actionable information, the efficacy of different vaccination
+strategies may depend on the topology of the
+network~\autocite{keeling2005networks,peng2013vaccination, ma2013importance,
+rushmore2014network}. On a local level, contact networks can be informative
+about the groups or individuals who are at highest risk of acquiring or
+transmitting infection, and would therefore benefit most from public health
+interventions~\autocite{wang2015targeting, little2014using}.
 
@@
 \defn{node-level} measures~\autocite{wasserman1994social}. One of the most
-well-studied of these is the degree distribution, which can be estimated by
-simply asking each person how many contacts they had in some interval of
-time~\autocite{liljeros2001web, schneeberger2004scale, colgate1989risk}.
-
-An alternative approach has been the analysis of other networks, which can be
-estimated with phylogenetic methods from viral sequence data. Some work focuses
-on the \defn{phylogenetic network}, in which two nodes are connected if the
-genetic distance between their viral sequences is below some threshold.
-Primarily, this work has focused on the detection of \defn{phylogenetic
-clusters}, which are groups of individuals whose viral sequences are
-significantly more similar to each other's than to the general population's.
-The phylogenetic network is informative about ``hotspots'' of transmission and
-can be used to identify demographic groups to whom targeted interventions are
-likely to have the greatest effect~\autocite{poon2015impact}.  However, this
-network may show little to no agreement with a contact data obtained through
-epidemiological methods~\autocite{yirrell1998molecular, resik2007limitations,
-robinson2013dynamics}, and therefore may be a poor proxy for the contact
-network. Other studies~\autocite{brown2011transmission} have investigated the
-\defn{transmission network}, which is the subgraph of the contact network
-consisting of infected nodes and the edges which led to their
+well-studied of these is the degree distribution mentioned above, which can 
+theoretically be estimated by simply asking each person how many contacts they
+had in some interval of time. However, the degree distributions often observed
+in real-world sexual networks are heavy-tailed~\autocite{liljeros2001web,
+schneeberger2004scale, colgate1989risk}, so dense or peer-driven sampling would
+be needed to capture the high-degree nodes characterizing the tail of the
+distribution.
+
+An alternative approach has been the analysis of other types of network, which
+can be directly estimated with phylogenetic methods from viral sequence data.
+Some work focuses on the \defn{phylogenetic network}, in which two nodes are
+connected if the genetic distance between their viral sequences is below some
+threshold.  Primarily, this work has focused on the detection of
+\defn{phylogenetic clusters}, which are groups of individuals whose viral
+sequences are significantly more similar to each other's than to the general
+population's.  The phylogenetic network is informative about ``hotspots'' of
+transmission and can be used to identify demographic groups to whom targeted
+interventions are likely to have the greatest effect~\autocite{poon2015impact}.
+However, this network may show little to no agreement with a contact data
+obtained through epidemiological methods~\autocite{yirrell1998molecular,
+resik2007limitations, robinson2013dynamics}, and therefore may be a poor proxy
+for the contact network. Other studies~\autocite{brown2011transmission} have
+investigated the \defn{transmission network}, which is the subgraph of the
+contact network consisting of infected nodes and the edges which led to their
 infections~\autocite{welch2011statistical} (\cref{fig:contactnet}, left). It is
@@
 network models, and found that these relationships varied considerably
-depending on which model was being considered. \textcite{welch2011network}
+depending on which model was being considered. The authors also investigated a
+real-world \gls{HIV} phylogeny and found a level of unbalancedeness
+inconsistent with a randomly mixing population. \textcite{welch2011network}
 simulated transmission trees over networks with varying degrees of community
@@
 will be suitable for our purposes. A more rigorous definition can be found in
-\eg \autocite{kallenberg2006foundations}. Note that a Markov kernels have
+\eg~\autocite{kallenberg2006foundations}. Note that a Markov kernels have
 nothing to do with the kernel functions defined in \cref{subsec:treeshape},

### 2016-04-25 doc/thesis/objective.tex
Writing

@@
 less effective in curtailing an epidemic depending on the underlying network's
-degree distribution~\autocite{peng2013vaccination}. Phylodynamic methods have
-been used to fit many different types of model to phylogenetic
-data~\autocite{pybus2009evolutionary,volz2013viral}, but as far as we know, no
-methods have yet been developed to fit contact network models. The primary
-objective of this work is to develop such a method.
+degree distribution~\autocite{peng2013vaccination, ma2013importance}.
+Phylodynamic methods have been used to fit many different types of model to
+phylogenetic data~\autocite{pybus2009evolutionary,volz2013viral}, but as far as
+we know, no methods have yet been developed to fit contact network models. The
+primary objective of this work is to develop such a method.
 
@@
 datasets.
+
+\glsreset{ABC}
+\glsreset{BA}
+\glsreset{HIV}
+\glsreset{ML}
+\glsreset{SMC}

### 2016-04-25 programs/netabc/R/plots.R
Added y-axis label to posterior densities

@@ -324,5 +324,7 @@ posterior.plot.pa <- function (trace, I_min=500, I_max=10000, N_min=500, N_max=1
         pm <- pm + geom_vline(xintercept=true_m, linetype="dashed")
     }
 
-    arrangeGrob(palpha, pI, pm, pN, ncol=2)
+    arrangeGrob(palpha, pI, pm, pN, ncol=2, 
+                left=textGrob("approximate posterior density", rot=90, 
+                              gp=gpar(fontsize=14, fontfamily="Gillius ADF")))
 }

### 2016-04-25 doc/hiv-dynamics/presentation.tex

@@
 \usepackage{tikz}
-\usepackage{tikz-qtree}
 \usepackage{gillius}
@@
 \usepackage{qrcode}
-\usepackage[backend=biber]{biblatex}
+\usepackage[backend=biber, style=authoryear, uniquelist=minyear, uniquename=false]{biblatex}
 
@@
       \centering
-      \includegraphics{compartments}
+      \includegraphics[scale=0.6]{compartments}
       \vspace{1cm}
 
-      \uncover<2->{\includegraphics[width=\textwidth]{sir-trajectories}}
+      \uncover<3->{\includegraphics[width=\textwidth]{sir-trajectories}}
     \end{column}
@@
         \setlength{\itemsep}{24pt}
-        \item Often provides reasonable approximation in practice.
         \uncover<2->{
+        \item Often provide a reasonable approximation in practice.
+        }
+        \uncover<3->{
         \item Can be inaccurate when substantial contact heterogeneity exists.
@@
 
-\begin{frame}{Scale-free networks and preferential attachment}
-  \includegraphics[width=\textwidth]{pa-example}
-  \begin{itemize}
-    \item Sexual networks are often scale-free: degree distributions appear
-      follow a power law ($\Pr(\text{degree} = d) \propto d^{\gamma}$).
-      \pause
-    \item Preferential attachment models generate scale-free networks.
-  \end{itemize}
-\end{frame}
+%\begin{frame}{Preferential attachment generates scale-free networks}
+%  \includegraphics[width=\textwidth]{pa-example}
+%  \begin{itemize}
+%    \item Sexual networks are often scale-free: degree distributions appear
+%      follow a power law ($\Pr(\text{degree} = d) \propto d^{-\gamma}$).
+%      \pause
+%    \item Preferential attachment models generate scale-free networks.
+%  \end{itemize}
+%\end{frame}
 
-\begin{frame}{Barab\'asi-Albert (BA) network model}
+\begin{frame}{Barab\'asi-Albert model incorporates preferential attachment}
   \begin{columns}
@@
         \uncover<5->{ \item Continue until network has $N$ nodes. }
-        \uncover<6->{ \item Also consider the prevalence $I$ when the transmission
-          tree is sampled.}
+        \uncover<6->{ \item Also consider the prevalence $I$. }
       \end{itemize}
@@
   \begin{itemize}
-    \item Generate networks under different parameter values ($N$, $m$, $\alpha$).
+    \setlength{\itemsep}{0pt}
+    \item Generate networks under different parameter values (number of nodes
+      $N$, number of edges per vertex $m$, pereferential attachment power
+      $\alpha$).
     \uncover<2->{
-    \item Simulate epidemic over each network with Gillespie algorithm ($I$).
+    \item Simulate epidemic over each network until $I$ nodes are infected.
     }
@@
     \uncover<4->{
-    \item Compare pairwise using tree kernel\footfullcite{poon2013mapping}.
+    \item Compare trees pairwise using tree kernel\footfullcite{poon2013mapping}.
     }
@@
 
-\begin{frame}{$\alpha$ and $I$ separate well in feature space}
-  \begin{tikzpicture}
-    \node {\includegraphics[height=0.8\textheight]{kernel-kpca.pdf}};
-    \fill<1>[white] (-4.6,-3.2) rectangle (5,0.2);
+\begin{frame}{Preferential attachment power $\alpha$ affects tree shape}
+  \begin{tikzpicture}[remember picture, overlay]
+    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=0.2in 2.7in 3.3in 0, clip]{kernel-kpca.pdf}};
+    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-alpha-tree.pdf}};
   \end{tikzpicture}
-  \pause
 \end{frame}
 
-\begin{frame}{Classifiers for network parameters}
-  \vspace{-1cm}
-  \begin{itemize}
-    \item kSVM using tree kernel
-    \item SVM using normalized
-      lineages-through-time\footfullcite{janzen2015approximate}
-    \item linear regression using Sackin's index~\footfullcite{shao1990tree}
-  \end{itemize}
-
-  \pause
-  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
+\begin{frame}{Prevalence $I$ affects tree shape}
+  \begin{tikzpicture}[remember picture, overlay]
+    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=3.5in 2.7in 0in 0, clip]{kernel-kpca.pdf}};
+    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-I-tree.pdf}};
+  \end{tikzpicture}
 \end{frame}
 
-\begin{frame}{Approximate Bayesian computation (ABC) for networks}
+\begin{frame}{Number of edges per vertex $m$ does not affect tree shape}
+  \begin{tikzpicture}[remember picture, overlay]
+    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=0.2in 0.3in 3.3in 2.2in, clip]{kernel-kpca.pdf}};
+    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-m-tree.pdf}};
+  \end{tikzpicture}
+\end{frame}
+
+\begin{frame}{Number nodes $N$ modestly affects tree shape}
+  \begin{tikzpicture}[remember picture, overlay]
+    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=3.5in 0.3in 0 2.2in, clip]{kernel-kpca.pdf}};
+    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-N-tree.pdf}};
+  \end{tikzpicture}
+\end{frame}
+
+\begin{frame}{Simulated trees can be used to fit a network model}
   \begin{minipage}[p][\textheight][t]{\textwidth}
-    \only<1>{\includegraphics[width=\textwidth, trim=0 3in 0 0, clip]{abc-idea.pdf}}
-    \only<2>{\includegraphics[width=\textwidth, trim=0 1.8in 0 0, clip]{abc-idea.pdf}}
-    \only<3>{\includegraphics[width=\textwidth, trim=0 1in 0 0, clip]{abc-idea.pdf}}
-    \only<4>{\includegraphics[width=\textwidth, trim=0 0in 0 0, clip]{abc-idea.pdf}}
+    \only<1>{\includegraphics[width=0.9\textwidth, trim=0 3in 0 0, clip]{abc-idea.pdf}}
+    \only<2>{\includegraphics[width=0.9\textwidth, trim=0 1.8in 0 0, clip]{abc-idea.pdf}}
+    \only<3>{\includegraphics[width=0.9\textwidth, trim=0 1in 0 0, clip]{abc-idea.pdf}}
+    \uncover<4->{\includegraphics[width=0.9\textwidth, trim=0 0in 0 0, clip]{abc-idea.pdf}}
+    \uncover<5->{\vspace{0.5cm}\centerline{\textbf{kernel approximate Bayesian computation (kernel-ABC)}}}
   \end{minipage}
@@
 
-\begin{frame}{Sequential Monte Carlo for ABC}
-  \begin{tikzpicture}
-    \only<1-2>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 4in 0, clip]{abc-smc.pdf}}};
-    \only<1>{\fill[white] (0.2,-6.5) rectangle (4,-3.5)};
-    \only<3>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 2in 0, clip]{abc-smc.pdf}}};
-    \only<4>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 0 0, clip]{abc-smc.pdf}}};
-  \end{tikzpicture}
-\end{frame}
+\begin{frame}{Example simulation shows $\alpha$ and $I$ can be reconstructed}
+  \begin{minipage}[p][\textheight][t]{\textwidth}
+    \begin{center}
+    preferential attachment power \hfill prevalence
+    \only<1>{
+      \includegraphics[width=0.8\textwidth, trim=0 2.5in 0 0, clip]{abc-posterior-example.pdf}
+    }
+    \only<2>{
+      \includegraphics[width=0.8\textwidth]{abc-posterior-example.pdf}
 
-\begin{frame}{Marginal posterior distributions for one simulated tree}
-  \begin{minipage}[p][\textheight][t]{\textwidth}
-    \only<1>{\centerline{\includegraphics[width=0.9\textwidth, trim=0 2.5in 0 0, clip]{abc-posterior-example.pdf}}}
-    \only<2>{\centerline{\includegraphics[width=0.9\textwidth]{abc-posterior-example.pdf}}}
+      \vspace{-0.25cm}
+      edges per vertex \hfill total nodes
+    }
+    \end{center}
   \end{minipage}
@@
 
-\begin{frame}{Accuracy of maximum \textit{a posteriori} estimates for $m = 2$}
+\begin{frame}{Maximum \textit{a posteriori} estimates are accurate for $\alpha$ and $I$}
   \begin{minipage}[p][\textheight][t]{\textwidth}
     \vspace{-0.5cm}
-    \only<1>{\includegraphics[height=0.9\textheight, trim=0 0 3.4in 0, clip]{abc-point-estimate-m2.pdf}}
-    \only<2>{\includegraphics[height=0.9\textheight]{abc-point-estimate-m2.pdf}}
+    preferential attachment power \hfill \uncover<2>{edges per vertex}
+
+    \hspace{0.75cm}\only<1>{\includegraphics[height=0.8\textheight, trim=0 0 3.4in 0, clip]{abc-point-estimate-m2.pdf}}
+    \only<2>{\includegraphics[height=0.8\textheight]{abc-point-estimate-m2.pdf}}
+
+    prevalence \hfill \uncover<2>{total nodes}
   \end{minipage}
@@
 
-\begin{frame}{Application to published HIV datasets}
+\begin{frame}{Real world HIV datasets exhibit network heterogeneity}
   \uncover<2->{
-  \includegraphics[width=\textwidth]{realdata-hpd-bc-horiz}
+    \centerline{\includegraphics[height=0.8\textheight]{realdata-hpd-bc}}
   }
-  \vspace{0.5cm}
 
-  \scriptsize
-  \input{\tablepath/realdata-bc}
-  \normalsize
+%  \scriptsize
+%    \begin{tabular}{ccccc}
+%      Reference & Sequences ($n$) & Location & Risk group & Gene \\
+%      \hline
+%      \color{magenta} \textcite{novitsky2013phylogenetic} &
+%        \color{magenta} \multirow{2}{*}{180} &
+%        \color{magenta} \multirow{2}{*}{Mochudi, Botswana} & 
+%        \color{magenta} \multirow{2}{*}{HET} &
+%        \color{magenta} \multirow{2}{*}{\textit{env}} \\ 
+%      \color{magenta} \textcite{novitsky2014impact} \\
+%      \color{blue} \textcite{cuevas2009hiv} & 
+%        \color{blue} 287 & 
+%        \color{blue} Basque Country, Spain & 
+%        \color{blue} mixed & 
+%        \color{blue} \textit{pol} \\
+%      \color{cyan} \textcite{li2015hiv} & 
+%        \color{cyan} 280 & 
+%        \color{cyan} Shanghai, China & 
+%        \color{cyan} MSM & 
+%        \color{cyan} \textit{pol} \\
+%      \color{green} unpublished & 
+%        \color{green} 399 & 
+%        \color{green} British Columbia, Canada & 
+%        \color{green} IDU & 
+%        \color{green} \textit{pol} \\
+%      \color{orange} \textcite{wang2015targeting} & 
+%        \color{orange} 173 & 
+%        \color{orange} Beijing, China & 
+%        \color{orange} MSM & 
+%        \color{orange} \textit{pol} \\
+%      \color{red} \textcite{niculescu2015recent} & 
+%        \color{red} 136 & 
+%        \color{red} Romaina & 
+%        \color{red} IDU & 
+%        \color{red} \textit{pol} \\
+%      \hline
+%    \end{tabular}
+%  \normalsize
 \end{frame}
@@
   \begin{itemize}
-    \item Kernel-ABC is a viable method for fitting contact network models to
-      phylogenetic data.
+    \item Kernel-ABC is the first phylodynamic method to fit contact network
+      models to phylogenetic data.
       \pause
-    \item Some parameters of the Barab\'asi-Albert preferential attachment data
-      have a measurable impact on tree shape and can be re-estimated using
-      kernel-ABC.
+    \item The preferential attachment power of the Barab\'asi-Albert network
+      model, which is challenging to estimate by traditional epidemiological
+      methods, can be estimated with kernel-ABC.
       \pause
-    \item The networks underlying real epidemics may be heterogeneous,
+    \item The networks underlying real epidemics are heterogeneous,
       underscoring the importance of considering network structure in
@@
       \vfill
+      \vspace{0.5cm}
+      $\qquad\qquad$\includegraphics[width=2cm]{logos/genomecanada}
+      $\qquad$
+      \includegraphics[width=2cm]{logos/genomequebec}
     \end{column}
@@
 
-      \includegraphics[width=3cm]{logos/cihr}
+      \includegraphics[width=3cm]{logos/cfe}
       \vspace{0.5cm}
 
+      \includegraphics[width=3cm]{logos/cihr}
+
+      \includegraphics[width=3cm]{logos/genomebc}
+
       \includegraphics[width=3cm]{logos/bmgf}
-      \vspace{0.5cm}
-
-      \includegraphics[width=1cm]{logos/ubc}
       \vspace{0.5cm}
@@
 \end{frame}
+
+\begin{frame}{Classifiers for network parameters}
+  \vspace{-1cm}
+  \begin{itemize}
+    \item kSVR using tree kernel
+    \item SVR using normalized
+      lineages-through-time\footfullcite{janzen2015approximate}
+    \item linear regression using Sackin's index~\footfullcite{shao1990tree}
+  \end{itemize}
+
+  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
+\end{frame}
+
+\begin{frame}{Sequential Monte Carlo for ABC}
+  \begin{tikzpicture}
+    \only<1-2>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 4in 0, clip]{abc-smc.pdf}}};
+    \only<1>{\fill[white] (0.2,-6.5) rectangle (4,-3.5)};
+    \only<3>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 2in 0, clip]{abc-smc.pdf}}};
+    \only<4>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 0 0, clip]{abc-smc.pdf}}};
+  \end{tikzpicture}
+\end{frame}
+
 \end{document}

### 2016-04-25 doc/tables/realdata.tex

@@
   \hline
-  \autocite{wang2015targeting} & 173 & Beijing, China & MSM & \textit{pol} \\
-  \autocite{cuevas2009hiv} & 287 & Basque Country, Spain & mixed & \textit{pol} \\
-  \autocite{novitsky2013phylogenetic} & \multirow{2}{*}{180} &
-  \multirow{2}{*}{Mochudi, Botswana} & \multirow{2}{*}{mixed} &
-  \multirow{2}{*}{\textit{env}} \\ \autocite{novitsky2014impact} \\
-  \autocite{li2015hiv} & 280 & Shanghai, China & MSM & \textit{pol} \\
-  \autocite{niculescu2015recent} & 136 & Romaina & IDU & \textit{pol} \\
+  \textcite{wang2015targeting} & 173 & Beijing, China & MSM & \textit{pol} \\
+  \textcite{cuevas2009hiv} & 287 & Basque Country, Spain & mixed & \textit{pol} \\
+  \textcite{novitsky2013phylogenetic} & \multirow{2}{*}{180} &
+  \multirow{2}{*}{Mochudi, Botswana} & \multirow{2}{*}{HET} &
+  \multirow{2}{*}{\textit{env}} \\ \textcite{novitsky2014impact} \\
+  \textcite{li2015hiv} & 280 & Shanghai, China & MSM & \textit{pol} \\
+  \textcite{niculescu2015recent} & 136 & Romania & IDU & \textit{pol} \\
   \hline

### 2016-04-25 doc/thesis/appendix.Rnw

--- 2016-04-19 doc/thesis/appendix.tex
@@
 \end{figure}
+
+\begin{figure}
+  \includegraphics{cuevas2009-posterior}
+  \caption{
+      Posterior distribution of BA model parameters for
+      \textcite{cuevas2009hiv} data. Vertical lines indicate maximum \textit{a
+      posteriori} estimates, and shaded areas are 95\% highest posterior
+      density intervals. $x$-axis indicates regions of nonzero prior density.
+  }
+  \label{fig:cuevas}
+\end{figure}
+
+\begin{figure}
+  \includegraphics{li2015-posterior}
+  \caption{
+      Posterior distribution of BA model parameters for \textcite{li2015hiv}
+      data. Vertical lines indicate maximum \textit{a posteriori} estimates,
+      and shaded areas are 95\% highest posterior density intervals. $x$-axis
+      indicates regions of nonzero prior density.
+  }
+  \label{fig:li}
+\end{figure}
+
+\begin{figure}
+  \includegraphics{niculescu2015-posterior}
+  \caption{
+      Posterior distribution of BA model parameters for
+      \textcite{niculescu2015recent} data. Vertical lines indicate maximum
+      \textit{a posteriori} estimates, and shaded areas are 95\% highest
+      posterior density intervals. $x$-axis indicates regions of nonzero prior
+      density.
+  }
+  \label{fig:niculescu}
+\end{figure}
+
+\begin{figure}
+  \includegraphics{novitsky2014-posterior}
+  \caption{
+      Posterior distribution of BA model parameters for
+      \textcite{novitsky2013phylogenetic, novitsky2014impact} data. Vertical
+      lines indicate maximum \textit{a posteriori} estimates, and shaded areas
+      are 95\% highest posterior density intervals. $x$-axis indicates regions
+      of nonzero prior density.
+  }
+  \label{fig:novitsky}
+\end{figure}
+
+\begin{figure}
+  \includegraphics{wang2015-posterior}
+  \caption{
+      Posterior distribution of BA model parameters for
+      \textcite{wang2015targeting} data. Vertical lines indicate maximum
+      \textit{a posteriori} estimates, and shaded areas are 95\% highest
+      posterior density intervals. $x$-axis indicates regions of nonzero prior
+      density.
+  }
+  \label{fig:wang}
+\end{figure}
+
+<<abc_posterior, echo=FALSE, results="asis">>=
+    N <- 5000
+    replicate <- 0
+    for (m in c(2, 3, 4)) {
+    for (I in c(1000, 2000)) {
+    for (alpha in c(0, 0.5, 1, 1.5, 2)) {
+        fn <- sprintf("abc-posterior/%2f_%d_%d_%d_%d.pdf", alpha, I, m, N, replicate)
+        lab <- sprintf("fig:%2f-%d-%d-%d-%d", alpha, I, m, N, replicate)
+        cat("\\begin{figure}\n")
+        cat("\\centering\n")
+        cat("\\includegraphics{", fn, "}\n")
+        cat("\\caption{\n")
+        cat("Approximate marginal posterior distributions of Barab\\'asi-Albert model parameters obtained using kernel-ABC for one simulated dataset.
+Dashed lines indicate true values, and shaded areas shown 95\\% highest posterior density intervals.\n")
+        cat("}\n")
+        cat("\\label{", lab, "}\n")
+    }
+    }
+    }
+@

### 2016-04-25 doc/thesis/methods.tex

@@
   \includegraphics[width=\textwidth]{gridsearch-expt}
-  \caption{Graphical schematic of grid search experiments used to investigate 
-    \gls{BA} model parameters.}
+  \caption{
+    Graphical schematic of grid search experiments used to investigate \gls{BA}
+    model parameters. Trees were simulated along a narrowly spaced grid of
+    parameter values (``grid trees''). Separate trees were simulated for a
+    small subset of the grid values (``test trees''). Each test tree was
+    compared to every grid tree using the tree kernel, and the resulting kernel
+    scores were normalized to resemble a probability density from which the
+    mode and 95\% highest density interval were calculated.
+  }
   \label{fig:gridexpt}
@@
 
-To test the full kernel-\gls{ABC} algorithm, we simulated four trees each under
+To test the full kernel-\gls{ABC} algorithm, we simulated three trees each under
 a variety of parameter values, and ran the \software{netabc} program to
 estimate posterior distributions for the parameters. The parameter values and
-priors used are listed in \ref{tab:abcexpt}. The tree kernel meta-parameters
+priors used are listed in \cref{tab:abcexpt}. The tree kernel meta-parameters
 were set to $\lambda = 0.3$ and $\sigma = 4$. The \gls{SMC} algorithm was run
@@
 constructed from the nucleotide alignments by approximate maximum likelihood
-using \software{FastTree2} version 2.1.7 with the \gls{GTR} model. Transmission
-trees were estimated by rooting and time-scaling the phylogenies by root-to-tip
-regression, using a modified version of Path-O-Gen (distributed as part of
+using \software{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
+\gls{GTR} model~\autocite{tavare1986some}. Transmission trees were estimated by
+rooting and time-scaling the phylogenies by root-to-tip regression, using a
+modified version of Path-O-Gen (distributed as part of
 BEAST~\autocite{drummond2007beast}) as described

### 2016-04-25 doc/thesis/preamble.tex
Small changes to objective
Finished results! Yes!
More writing.
Wrote discussion mostly
More writing
Panicked 1AM writing
Finished draft intro!

@@
-\documentclass[12pt]{report}
+\documentclass[11pt]{report}
 \usepackage[osf]{garamondx}
 \usepackage[garamondx,cmbraces]{newtxmath}
-\usepackage{marginnote}
-\usepackage{fullpage}
+\usepackage[top=1in, bottom=1in, inner=1.25in, outer=1in]{geometry}
 \usepackage{setspace}
@@
 \usepackage{float}
+\usepackage{pdflscape}
+\usepackage{fancyhdr}
+\usepackage[dvipsnames]{xcolor}
 
@@
 \renewcommand{\d}{\mathrm{d}\,}
+\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
+\newcommand{\maps}{:}
 
 \newcommand{\N}{\mathcal{N}}
+\newcommand{\I}{\mathbb{I}}
+\newcommand{\F}{\mathcal{F}}
+\newcommand{\X}{\mathcal{X}}
 \renewcommand{\L}{\mathcal{L}}
@@
 \DeclareMathOperator{\Var}{Var}
-
 \DeclareMathOperator{\inc}{in}
 \DeclareMathOperator{\out}{out}
+\DeclareMathOperator{\ESS}{ESS}
+\DeclareMathOperator{\nc}{nc}
 
@@
 \newacronym{SVM}{SVM}{support vector machine}
+\newacronym{SVR}{SVR}{support vector regression}
+\newacronym{kSVM}{kSVM}{kernel support vector machine}
+\newacronym{kSVR}{kSVR}{kernel support vector regression}
 \newacronym{PCA}{PCA}{principal components analysis}
+\newacronym{kPCA}{kPCA}{kernel principal components analysis}
 \newacronym{SMC}{SMC}{sequential Monte Carlo}
@@
 \newacronym{HPD}{HPD}{highest posterior density}
+\newacronym{HDI}{HDI}{highest density interval}
 \newacronym{HIV}{HIV}{human immunodeficiency virus}
@@
 \newacronym{GLM}{GLM}{generalized linear model}
+\newacronym{tasp}{TasP}{treatment as prevention}
+\newacronym{MH}{MH}{Metropolis-Hastings}
+\newacronym{GTR}{GTR}{generalized time-reversible}
+\newacronym{ESS}{ESS}{expected sample size}
+\newacronym{MSM}{MSM}{men who have sex with men}
+\newacronym{ERGM}{ERGM}{exponential random graph model}
 
@@
 
+\pagestyle{fancyplain}
+\fancyhf{}
+\renewcommand{\headrulewidth}{0pt}
+\rfoot{\thepage}
+
+\renewcommand\UrlFont{\color{Blue}\rmfamily}
+
 \title{Phylodynamic inference of contact network parameters through approximate

### 2016-04-25 doc/thesis/results.Rnw

@@
 simulated trees where the true parameter values were known. Point estimates for
-each parameter are shown in \cref{fig:abcpt} for the simulations with \gls{m} =
+each parameter are shown in \cref{fig:abcptm2} for the simulations with \gls{m} =
 2. The results for the other values of \gls{m} were similar
@@
 No other parameters influenced the accuracy of the $N$ estimates (one-way
-ANOVA)
+ANOVA).
 
@@
   \includegraphics{abc-point-estimate-m2}
+  \caption[\Acrlong{MAP} point estimates for \gls{BA} model parameters obtained
+    by running \software{netabc} on simulated data, for simulations with $m = 2$.] 
+  {
+    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by         
+    running \software{netabc} on simulated data. Values shown are for               
+    simulations with \gls{m} = 2. Dashed lines indicate true values. (A)            
+    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations  
+    against known values. (B) Estimates of \gls{m} and \gls{N} which were held   
+    fixed in these simulations at the values \gls{m} = 2 and \gls{N} = 5000. 
+  }
+  \label{fig:abcptm2}
+\end{figure}
+
+\Cref{fig:abcex} shows the \gls{ABC} approximation to the posterior
+distribution on the \gls{BA} parameters for one simulation. Equivalent plots
+for one replicate simulation with each combination of parameters can be found
+in
+<<posterior_sims, echo=FALSE, results="asis">>=
+    N <- 5000
+    replicate <- 0
+    lab <- NULL
+    for (m in c(2, 3, 4)) {
+    for (I in c(1000, 2000)) {
+    for (alpha in c(0, 0.5, 1, 1.5, 2)) {
+        lab <- c(lab, sprintf("fig:%2f-%d-%d-%d-%d", alpha, I, m, N, replicate))
+    }
+    }
+    }
+    cat(paste0("\\cref{", paste(lab, sep=","), "}."))
+@
+\Gls{HPD} intervals around \gls{alpha} and \gls{I} were narrow relative to the
+region of nonzero prior density, whereas the intervals for $m$ and \gls{N} were
+widely dispersed. \Cref{tab:abchpd} shows point estimates and 95\% \gls{HPD}
+intervals averaged over all simulations.
+
+\begin{figure}
+  \includegraphics{{abc-posterior/1.0_1000_2_5000_0}.pdf}
   \caption{
@@
   }
-  \label{fig:abcm2}
+  \label{fig:abcptm2}
 \end{figure}
-
-\Cref{fig:abcex} shows the \gls{ABC} approximation to the posterior
-distribution on the \gls{BA} parameters for one simulation (equivalent plots
-for all the simulations can be found in the supplemental materials). \Gls{HPD}
-intervals around \gls{alpha} and \gls{I} were narrow relative to the region of
-nonzero prior density, whereas the intervals for $m$ and \gls{N} were widely
-dispersed. \Cref{tab:abchpd} shows point estimates and 95\% \gls{HPD}
-intervals averaged over all simulations.
 
@@
 <<realdata, include=FALSE>>=
-  d <- fread(Sys.glob("../../simulations/aggregates/hpd/*"))
-  map <- dcast.data.table(d, dataset~parameter, value.var="map")
-  up <- dcast.data.table(d, dataset~parameter, value.var="hpd.max")
-  low <- dcast.data.table(d, dataset~parameter, value.var="hpd.min")
-  setkey(map, dataset)
-  setkey(up, dataset)
-  setkey(low, dataset)
+  m <- collect.metadata("../../simulations/aggregates/hpd/*")
+  d1 <- fread(rownames(m)[m$m_min == 1])
+  d2 <- fread(rownames(m)[m$m_min == 2])
+
+  map1 <- dcast.data.table(d1, dataset~parameter, value.var="map")
+  up1 <- dcast.data.table(d1, dataset~parameter, value.var="hpd.max")
+  low1 <- dcast.data.table(d1, dataset~parameter, value.var="hpd.min")
+  setkey(map1, dataset)
+  setkey(up1, dataset)
+  setkey(low1, dataset)
+
+  map2 <- dcast.data.table(d2, dataset~parameter, value.var="map")
+  up2 <- dcast.data.table(d2, dataset~parameter, value.var="hpd.max")
+  low2 <- dcast.data.table(d2, dataset~parameter, value.var="hpd.min")
+  setkey(map2, dataset)
+  setkey(up2, dataset)
+  setkey(low2, dataset)
 @
@@
 and found substantial heterogeneity among the parameter estimates
-(\cref{fig:abchpd}). Two of the datasets~\autocite{niculescu2015recent,
-wang2015targeting} had estimated $\alpha$ values near unity (MAP estimates
-[95\% HPD] 
-  \Sexpr{map["niculescu2015", round(alpha, 2)]} 
-  [\Sexpr{low["niculescu2015", round(alpha, 2)]} - 
-   \Sexpr{up["niculescu2015", round(alpha, 2)]}]
+(\cref{fig:abchpd,fig:abchpdm2}). Plots of the marginal posterior distributions
+for each dataset are shown in
+\cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.
+Two of the datasets~\autocite{niculescu2015recent, wang2015targeting} had
+estimated $\alpha$ values near unity for the prior allowing $m = 1$ (MAP
+estimates [95\% HPD] 
+  \Sexpr{map1["niculescu2015", round(alpha, 2)]} 
+  [\Sexpr{low1["niculescu2015", round(alpha, 2)]} - 
+   \Sexpr{up1["niculescu2015", round(alpha, 2)]}]
 and
-  \Sexpr{map["wang2015", round(alpha, 2)]} 
-  [\Sexpr{low["wang2015", round(alpha, 2)]} -
-   \Sexpr{up["wang2015", round(alpha, 2)]}] respectively). 
-Another two datasets~\autocite{li2015hiv, cuevas2009hiv} had lower estimated
-values and wider HPD intervals
-  (\Sexpr{map["li2015", round(alpha, 2)]} 
-  [\Sexpr{low["li2015", round(alpha, 2)]} - 
-  \Sexpr{up["li2015", round(alpha, 2)]}]
+  \Sexpr{map1["wang2015", round(alpha, 2)]} 
+  [\Sexpr{low1["wang2015", round(alpha, 2)]} -
+   \Sexpr{up1["wang2015", round(alpha, 2)]}] respectively).
+The MAP estimates did not change appreciably when $m = 1$ was disallowed by the
+prior, although the credible interval of the \textcite{niculescu2015recent}
+data was narrower
+  (\Sexpr{low1["niculescu2015", round(alpha, 2)]} - 
+   \Sexpr{up1["niculescu2015", round(alpha, 2)]}).
+When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
+low estimated $\alpha$ values
+  (\Sexpr{map1["li2015", round(alpha, 2)]} 
+  [\Sexpr{low1["li2015", round(alpha, 2)]} - 
+  \Sexpr{up1["li2015", round(alpha, 2)]}]
 and
-  \Sexpr{map["cuevas2009", round(alpha, 2)]} 
-  [\Sexpr{low["cuevas2009", round(alpha, 2)]} -
-   \Sexpr{up["cuevas2009", round(alpha, 2)]}]). 
-The \textcite{novitsky2014impact} data had an extremely low estimated $\alpha$
-and a very wide HPD interval
-  (\Sexpr{map["novitsky2014", round(alpha, 2)]} 
-  [\Sexpr{low["novitsky2014", round(alpha, 2)]} -
-   \Sexpr{up["novitsky2014", round(alpha, 2)]}]). 
-For all the datasets except \citeauthor{novitsky2014impact}, estimated values of
-$I$ were below 2000, with narrow HPD intervals around two of the
-datasets
-  (\citeauthor{cuevas2009hiv}, \Sexpr{map["cuevas2009", floor(I)]} 
-  [\Sexpr{low["cuevas2009", floor(I)]} -
-   \Sexpr{up["cuevas2009", floor(I)]}];
-   \citeauthor{niculescu2015recent}, \Sexpr{map["niculescu2015", floor(I)]}
-  [\Sexpr{low["niculescu2015", floor(I)]} - 
-   \Sexpr{up["niculescu2015", floor(I)]}])
-and wider intervals around the other two
-  (\citeauthor{li2015hiv}, \Sexpr{map["li2015", floor(I)]} 
-  [\Sexpr{low["li2015", floor(I)]} -
-   \Sexpr{up["li2015", floor(I)]}];
-   \citeauthor{wang2015targeting}, \Sexpr{map["wang2015", floor(I)]}
-  [\Sexpr{low["wang2015", floor(I)]} - 
-   \Sexpr{up["wang2015", floor(I)]}]).
-The \citeauthor{novitsky2014impact} data was again the outlier, with a very high
+  \Sexpr{map1["cuevas2009", round(alpha, 2)]} 
+  [\Sexpr{low1["cuevas2009", round(alpha, 2)]} -
+   \Sexpr{up1["cuevas2009", round(alpha, 2)]}]). 
+However, the MAP estimates increased when $m = 1$ was not permitted, although
+the HPD intervals remained roughly the same
+  (\Sexpr{map2["li2015", round(alpha, 2)]} 
+  [\Sexpr{low2["li2015", round(alpha, 2)]} - 
+  \Sexpr{up2["li2015", round(alpha, 2)]}]
+and
+  \Sexpr{map2["cuevas2009", round(alpha, 2)]} 
+  [\Sexpr{low2["cuevas2009", round(alpha, 2)]} -
+   \Sexpr{up2["cuevas2009", round(alpha, 2)]}]).
+The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
+for both priors on $m$
+  (\Sexpr{map1["novitsky2014", round(alpha, 2)]} for $m \geq 1$;
+   \Sexpr{map2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).
+However, the confidence interval was much wider when $m = 1$ was allowed
+  ([\Sexpr{low1["novitsky2014", round(alpha, 2)]} -
+    \Sexpr{up1["novitsky2014", round(alpha, 2)]}] for $m \geq 1$ vs.
+    \Sexpr{low2["novitsky2014", round(alpha, 2)]} -
+    \Sexpr{up2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).
+
+For all the datasets except \citeauthor{novitsky2014impact}, estimated values
+of $I$ were below 2000 when $m = 1$ was allowed, with relatively narrow HPD
+intervals compared to the nonzero prior density region
+  (\citeauthor{cuevas2009hiv}, \Sexpr{map1["cuevas2009", floor(I)]} 
+  [\Sexpr{low1["cuevas2009", floor(I)]} -
+   \Sexpr{up1["cuevas2009", floor(I)]}];
+   \citeauthor{niculescu2015recent}, \Sexpr{map1["niculescu2015", floor(I)]}
+  [\Sexpr{low1["niculescu2015", floor(I)]} - 
+   \Sexpr{up1["niculescu2015", floor(I)]}];
+  \citeauthor{li2015hiv}, \Sexpr{map1["li2015", floor(I)]} 
+  [\Sexpr{low1["li2015", floor(I)]} -
+   \Sexpr{up1["li2015", floor(I)]}];
+   \citeauthor{wang2015targeting}, \Sexpr{map1["wang2015", floor(I)]}
+  [\Sexpr{low1["wang2015", floor(I)]} - 
+   \Sexpr{up1["wang2015", floor(I)]}]).
+The \citeauthor{novitsky2014impact} data was the outlier, with a very high
 estimated $I$, and HPD interval spanning almost the entire prior region
-  (\Sexpr{map["novitsky2014", floor(I)]} 
-  [\Sexpr{low["novitsky2014", floor(I)]} -
-   \Sexpr{up["novitsky2014", floor(I)]}]).
-No information was gleaned about the $m$ parameter, with the HPD interval
-occupying the entire prior region for all datasets. The estimates of $N$ were
-similarly uninformative, with the exception that the point estimate for the
-\citeauthor{wang2015targeting} data was smaller
-  (\Sexpr{map["wang2015", floor(N)]})
-than the estimates for other datasets
- (average \Sexpr{map[dataset != "wang2015", floor(mean(N))]}).
+  (\Sexpr{map1["novitsky2014", floor(I)]} 
+  [\Sexpr{low1["novitsky2014", floor(I)]} -
+   \Sexpr{up1["novitsky2014", floor(I)]}]).
+The $I$ estimates and HPD intervals were generally robust to the choice of
+prior on $m$, with slightly narrower HPD intervals (compare
+\cref{fig:abchpd,fig:abchpdm2}).
+
+The MAP estimate of $m$ was equal to 1 for all but the
+\citeauthor{novitsky2014impact} data, when this value was allowed. However, the
+upper bound of the HPD interval was different for each dataset
+  (\citeauthor{niculescu2015recent}, \Sexpr{up1["niculescu2015", round(m)]};
+   \citeauthor{wang2015targeting}, \Sexpr{up1["wang2015", round(m)]};
+   \citeauthor{li2015hiv}, \Sexpr{up1["li2015", round(m)]};
+   \citeauthor{cuevas2009hiv}, \Sexpr{up1["cuevas2009", round(m)]}).
+When $m = 1$ was disallowed, the MAP for all datasets was either 2 or 3, with
+HPD intervals spanning the entire prior region. The estimates for the total
+number of nodes $N$ were largely uninformative for all samples, with almost all
+MAP estimates greater than 7500 and HPD intervals spanning almost the entire
+nonzero prior density region. The only exception was the \citeauthor{li2015hiv}
+data, for which the MAP estimate was lower 
+  (\Sexpr{map1["li2015", floor(N)]})
+when $m = 1$ was allowed.
 
@@
   \centering
-  \includegraphics{realdata-hpd-bc.pdf}
+  %\includegraphics{realdata-hpd}
   \vspace{8pt}

### 2016-04-25 doc/thesis/appendix.Rnw
Appendix

@@
-\begin{figure}
+\begin{figure}[H]
   \centering
@@
 
-\begin{figure}
+\begin{figure}[H]
   \centering
@@
     Cross validation accuracy of classifiers for \gls{BA} model parameter
-    \gls{alpha} for eight epidemic scenarios. Solid lines and points are
+    \gls{alpha} for eigH epidemic scenarios. Solid lines and points are
     $R^2$ of tree kernel \gls{kSVR} under various kernel meta-parameters.
@@
     Cross validation accuracy of classifiers for \gls{BA} model parameter
-    \gls{I} for eight epidemic scenarios. Solid lines and points are $R^2$ of
+    \gls{I} for eigH epidemic scenarios. Solid lines and points are $R^2$ of
     tree kernel \gls{kSVR} under various kernel meta-parameters. Dashed and
@@
     Cross validation accuracy of classifiers for \gls{BA} model parameter
-    \gls{m} for eight epidemic scenarios. Solid lines and points are $R^2$ of
+    \gls{m} for eigH epidemic scenarios. Solid lines and points are $R^2$ of
     tree kernel \gls{kSVR} under various kernel meta-parameters. Dashed and
@@
     Cross validation accuracy of classifiers for \gls{BA} model parameter
-    \gls{N} for eight epidemic scenarios. Solid lines and points are $R^2$ of
+    \gls{N} for eigH epidemic scenarios. Solid lines and points are $R^2$ of
     tree kernel \gls{kSVR} under various kernel meta-parameters. Dashed and
@@
     Kernel principal components projection of trees simulated under three
-    different values of \gls{BA} parameter \gls{alpha}, for eight epidemic
+    different values of \gls{BA} parameter \gls{alpha}, for eigH epidemic
     scenarios.
@@
     Kernel principal components projection of trees simulated under three
-    different values of \gls{BA} parameter \gls{I}, for eight epidemic
+    different values of \gls{BA} parameter \gls{I}, for eigH epidemic
     scenarios.
@@
     Kernel principal components projection of trees simulated under three
-    different values of \gls{BA} parameter \gls{m}, for eight epidemic
+    different values of \gls{BA} parameter \gls{m}, for eigH epidemic
     scenarios.
@@
     Kernel principal components projection of trees simulated under three
-    different values of \gls{BA} parameter \gls{N}, for eight epidemic
+    different values of \gls{BA} parameter \gls{N}, for eigH epidemic
     scenarios.
@@
 
-\begin{figure}
+\begin{figure}[H]
   \includegraphics{cuevas2009-posterior}
@@
 
-\begin{figure}
+\begin{figure}[H]
   \includegraphics{li2015-posterior}
@@
 
-\begin{figure}
+\begin{figure}[H]
   \includegraphics{niculescu2015-posterior}
@@
 
-\begin{figure}
+\begin{figure}[H]
   \includegraphics{novitsky2014-posterior}
@@
 
-\begin{figure}
+\begin{figure}[H]
   \includegraphics{wang2015-posterior}
@@
     for (I in c(1000, 2000)) {
-    for (alpha in c(0, 0.5, 1, 1.5, 2)) {
-        fn <- sprintf("abc-posterior/%2f_%d_%d_%d_%d.pdf", alpha, I, m, N, replicate)
-        lab <- sprintf("fig:%2f-%d-%d-%d-%d", alpha, I, m, N, replicate)
-        cat("\\begin{figure}\n")
+    for (alpha in c(0, 0.5, 1, 1.5)) {
+        fn <- sprintf("{abc-posterior/%.1f_%d_%d_%d_%d}.pdf", alpha, I, m, N, replicate)
+        lab <- sprintf("{fig:%.1f-%d-%d-%d-%d}", alpha, I, m, N, replicate)
+        cat("\\begin{figure}[H]\n")
         cat("\\centering\n")
-        cat("\\includegraphics{", fn, "}\n")
+        cat(paste0("\\includegraphics{", fn, "}\n"))
         cat("\\caption{\n")
@@
         cat("\\label{", lab, "}\n")
+        cat("\\end{figure}\n")
     }

### 2016-04-25 doc/paper/paper.Rnw
Paper updates

@@
 the speed and pattern of the epidemic's expansion. Network structure can
-influence the prevalence trajectory~\autocite{o2011contact} and epidemic
-threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates of
-quantities such as effective population size~\autocite{goodreau2006assessing}.
-From a public health perspective, contact networks have been explored as tools
-for curtailing epidemic spread, by way of interventions targeted to
-well-connected nodes~\autocite{wang2015targeting}. True contact networks are a
-challenging type of data to collect, requiring extensive epidemiological
+influence the prevalence trajectory~\autocite{o2011contact, ma2013importance}
+and epidemic threshold~\autocite{barthelemy2005dynamical}, in turn affecting
+the estimates of quantities such as effective population
+size~\autocite{goodreau2006assessing}.  From a public health perspective,
+contact networks have been explored as tools for curtailing epidemic spread, by
+way of interventions targeted to well-connected
+nodes~\autocite{wang2015targeting}. True contact networks are a challenging
+type of data to collect, requiring extensive epidemiological
 investigation~\autocite{welch2011statistical}.
@@
 sophisticated methods have been developed for fitting complex population
-genetic models to phylogenies~\autocite{rasmussen2014phylodynamic}, inference
-of structural network parameters has to date been limited. However, it has been
-shown that network structure has a tangible impact on phylogeny
-shape~\autocite{leventhal2012inferring, colijn2014phylogenetic,
-goodreau2006assessing, robinson2013dynamics, villandre2016assessment},
-suggesting that such statistical inference might be
+genetic models to phylogenies~\autocite{volz2012complex,
+rasmussen2014phylodynamic}, inference of structural network parameters has to
+date been limited. However, it has been shown that network structure has a
+tangible impact on phylogeny shape~\autocite{leventhal2012inferring,
+colijn2014phylogenetic, goodreau2006assessing, robinson2013dynamics,
+villandre2016assessment}, suggesting that such statistical inference might be
 possible~\autocite{welch2011statistical}.
 
-Survey-based studies of sexual networks~\autocite{liljeros2001web,
-schneeberger2004scale, colgate1989risk} have found that these networks tend to
-have a degree distribution which follows a power law \autocite[although there has
-been some disagreement, see][]{handcock2004likelihood}. That is, the number of
-nodes of degree $k$ is proportional to $k^{-\gamma}$ for some constant
-$\gamma$. These networks are also creferred to as ``scale-free''. One process by
-which scale-free networks can be generated is pcreferential attachment, where
-nodes with a high number of contacts attract new connections at an elevated
-rate. The first contact network model incorporating pcreferential attachment was
+Survey-based studies of sexual networks~\autocite{colgate1989risk,
+liljeros2001web, schneeberger2004scale} have found that these networks tend to
+have a degree distribution which follows a power law \autocite[although there
+has been some disagreement, see][]{handcock2004likelihood}. That is, the number
+of nodes of degree $k$ is proportional to $k^{-\gamma}$ for some constant
+$\gamma$. These networks are also referred to as
+``scale-free''~\autocite{barabasi1999emergence}. One process by which
+scale-free networks can be generated is preferential attachment, where nodes
+with a high number of contacts attract new connections at an elevated rate. The
+first contact network model incorporating pcreferential attachment was
 introduced by \textcite{barabasi1999emergence}, and is now creferred to as the
@@
 where the probability of attaching to a node of degree $k$ is $k^\alpha$ for
-some non-negative constant $\alpha$, and we use this extension here.
+some non-negative constant $\alpha$, and we use this extension in this work.
 
@@
 alternative is offered by likelihood-free methods, namely approximate Bayesian
-computation (ABC). ABC leverages the fact that, although calculating the
-likelihood may be impractical, generating simulated datasets according to a
-model is often straightforward. If our model fits the data well, the simulated
-data it produces should be similar to the observed data. More formally, if $D$
-is the observed data, the posterior distribution $f(\theta \mid D)$ on model
+computation (ABC)~\autocite{tavare1997inferring, beaumont2002approximate}. ABC
+leverages the fact that, although calculating the likelihood may be
+impractical, generating simulated datasets according to a model is often
+straightforward. If our model fits the data well, the simulated data it
+produces should be similar to the observed data. More formally, if $D$ is the
+observed data, the posterior distribution $f(\theta \mid D)$ on model
 parameters $\theta$ is replaced as the target of statistical inference by
@@
 sunnaaker2013approximate}. A computer program implementing our method is freely
-available at \url{https://github.com/rmcclosk/netabc} (last accessed April 3,
-2016).
+available at \url{https://github.com/rmcclosk/netabc} (last accessed \today).
 
@@
 We constructed a kSVR classifier for $\alpha$ using the \textit{kernlab}
-package~\autocite{zeileis2004kernlab}, and evaluated its accuracy with 1000
+package~\autocite{karatzoglou2004kernlab}, and evaluated its accuracy with 1000
 two-fold cross-validations.
@@
   \input{\tablepath/realdata}
-  \caption{Characteristics of published datasets investigated with kernel-ABC.
-  Acronyms: MSM, men who have sex with men; IDU, injection drug users.}
+  \caption{
+    Characteristics of published datasets investigated with kernel-ABC.
+    Acronyms: MSM, men who have sex with men; IDU, injection drug users; HET,
+    heterosexual. The \textcite{novitsky2013phylogenetic, novitsky2014impact}
+    data was sampled from a primarily heterosexual risk environment, but did
+    not explicitly exclude other risk groups.
+  }
   \label{tab:data}
@@
 alignments by approximate maximum likelihood using \textit{FastTree2} version
-2.1.7 with the generalized time-reversible model. Transmission trees were
-estimated by rooting and time-scaling the phylogenies by root-to-tip
-regression, using a modified version of Path-O-Gen (distributed as part of
-BEAST~\autocite{drummond2007beast}) as described
+2.1.7 \autocite{price2010fasttree} with the generalized time-reversible model.
+Transmission trees were estimated by rooting and time-scaling the phylogenies
+by root-to-tip regression, using a modified version of Path-O-Gen (distributed
+as part of BEAST~\autocite{drummond2007beast}) as described
 previously~\autocite{poon2015phylodynamic}.
@@
 and found substantial heterogeneity among the parameter estimates
-(\cref{fig:abchpd,fig:abchpdm2}). Two of the
-datasets~\autocite{niculescu2015recent, wang2015targeting} had estimated
-$\alpha$ values near unity for the prior allowing $m = 1$ (MAP estimates
-[95\% HPD] 
+(\cref{fig:abchpd,fig:abchpdm2}). Plots of the marginal posterior distributions
+for each dataset are shown in
+\cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.
+Two of the datasets~\autocite{niculescu2015recent, wang2015targeting} had
+estimated $\alpha$ values near unity for the prior allowing $m = 1$ (MAP
+estimates [95\% HPD] 
   \Sexpr{map1["niculescu2015", round(alpha, 2)]} 
@@
    \Sexpr{up1["niculescu2015", floor(I)]}];
-  (\citeauthor{li2015hiv}, \Sexpr{map1["li2015", floor(I)]} 
+  \citeauthor{li2015hiv}, \Sexpr{map1["li2015", floor(I)]} 
   [\Sexpr{low1["li2015", floor(I)]} -
@@
 number of nodes $N$ were largely uninformative for all samples, with almost all
-MAP estimates greater than 5000 and HPD intervals spanning almost the entire
+MAP estimates greater than 7500 and HPD intervals spanning almost the entire
 nonzero prior density region. The only exception was the \citeauthor{li2015hiv}
@@
 little2014using}. Despite this, few methods exist for investigating contact
-network parameters in a phylodynamic framework~\autocite{groendyke2011bayesian,
-volz2008sir, brown2011transmission, leventhal2012inferring}. Kernel-ABC is a
-model-agnostic method which can be used to investigate any quantity that
-affects tree shape~\autocite{poon2015phylodynamic}. In this work, we developed a
+network parameters in a phylodynamic framework \autocite[although see][for
+related work]{groendyke2011bayesian, volz2008sir, brown2011transmission,
+leventhal2012inferring}. Kernel-ABC is a model-agnostic method which can be
+used to investigate any quantity that affects tree
+shape~\autocite{poon2015phylodynamic}. In this work, we developed a
 kernel-ABC-based method to infer the parameters of a contact network model. The
 method is general, and could be applied to any model from which contact
-networks can be simulated. We demonstrated the method on the BA model,
-which is a simple pcreferential attachment model giving rise to the power law
-degree distributions commonly observed in real-world networks.
+networks can be simulated. We demonstrated the method on the BA model, which is
+a simple pcreferential attachment model giving rise to the power law degree
+distributions commonly observed in real-world networks.
 
@@
 2. Thecrefore, if we were primarily interested in $\gamma$ rather than
-$\alpha$, a more sensible choice of prior might have a shape similar to the
-inverse of \cref{fig:gamma} and be bounded above by approximately $\alpha$ =
-1.5. This would uniformly bound $\gamma$ in the region $2 \leq \gamma \leq 4$
-commonly reported in the network literature~\autocite{liljeros2001web,
+$\alpha$, a more sensible choice of prior might have a shape informed by
+\cref{fig:gamma} and be bounded above by approximately $\alpha$ = 1.5. This
+would uniformly bound $\gamma$ in the region $2 \leq \gamma \leq 4$ commonly
+reported in the network literature~\autocite{liljeros2001web,
 schneeberger2004scale, colgate1989risk, brown2011transmission}. We note however
 that \textcite{jones2003assessment} estimated $\gamma$ values greater than
-four, in one case as high as 17, for some datasets, indicating that a wider
+four for some datasets, in one case as high as 17, indicating that a wider
 range of permitted $\gamma$ values may be warranted.

### 2016-04-27 doc/thesis/acknowledgements.tex
Thesis almost done

First and foremost I am deeply indebted to my supervisor, Dr. Art Poon, 
for mentoring me through the majority of my academic career.

I also thank my labmates, especially Jeff, Richard, and Thuy. I'll miss our
Thursday lunches.

I am grateful for the mentorship of Dr. Sara Mostafavi and Dr. Ryan Morin, who
accepted me into their labs and always had time for my many questions.  Thanks
also to Dr. Marni Mishna and Dr. Cedric Chauve, for starting me on this path.

I owe thanks to my fellow graduate students of the bioinformatics training
program. Special thanks go to Ogan for keeping me on an even keel when times
got tough.

Finally, I dedicate this thesis to my parents, who have supported me morally
and financially, endured my frustrations, and celebrated my successes.

### 2016-04-27 doc/thesis/appendix.Rnw
Thesis almost done

--- 2016-04-25 doc/thesis/appendix.Rnw
@@
-\begin{figure}
+\begin{figure}[H]
   \centering
-  \includegraphics{kernel-I-tree.pdf}
-  \caption{
-    Transmission trees simulated over \gls{BA} networks with varying values of
-    $I$, the number of infected nodes when the epidemic simulation was stopped.
+  \includegraphics[width=\textwidth]{kernel-I-tree.pdf}
+  \caption[Simulated transmission trees under three different values of BA parameter $I$]{
+    Simulated transmission trees under three different values of BA parameter
+    $I$. Epidemics were simulated on \gls{BA} networks with parameters $\alpha$
+    = 1.0, $m = 2$, and $N = 5000$. Epidemics were simulated until $I = 500$,
+    1000, or 2000 nodes were infected. Transmission trees were created by
+    sampling 500 infected nodes. For higher \gls{I} values, the network was
+    closer to saturation at the time of sampling, resulting in longer terminal
+    branches as the waiting time until the next transmission increased.
   }
@@
 
-\begin{figure}
+\begin{figure}[H]
   \centering
-  \includegraphics{kernel-m-tree.pdf}
-  \caption{
-    Transmission trees simulated over \gls{BA} networks with varying values of
-    $m$, the number of edges added per vertex.
+  \includegraphics[width=\textwidth]{kernel-m-tree.pdf}
+  \caption[Simulated transmission trees under three different values of BA parameter $m$]{
+    Simulated transmission trees under three different values of BA parameter
+    $m$. Epidemics were simulated on \gls{BA} networks with parameters $\alpha$
+    = 1.0, $N = 5000$, and $m$ = 2, 3, or 4. Epidemics were simulated until $I
+    = 1000$ nodes were infected. Transmission trees were created by sampling
+    500 infected nodes.
   }
@@
 
-\begin{figure}
+\begin{figure}[H]
   \centering
-  \includegraphics{kernel-N-tree.pdf}
-  \caption{
-    Transmission trees simulated over \gls{BA} networks with varying values of
-    $N$, the number of nodes in the network.
+  \includegraphics[width=\textwidth]{kernel-N-tree.pdf}
+  \caption[Simulated transmission trees under three different values of BA parameter $N$]{
+    Simulated transmission trees under three different values of BA parameter
+    $N$. Epidemics were simulated on \gls{BA} networks with parameters $\alpha$
+    = 1.0, $m = 2$, and $N$ = 3000, 5000, or 8000. Epidemics were simulated
+    until $I = 1000$ nodes were infected. Transmission trees were created by
+    sampling 500 infected nodes. For lower $N$ values, the network was closer
+    to saturation at the time of sampling, resulting in longer waiting times
+    until the next transmission and longer terminal branch lengths.
   }
@@
 
-\begin{figure}
+\begin{figure}[H]
+  \centering
+  \includegraphics{gridsearch-alpha-point-estimate.pdf}
+  \caption{
+      Point estimates of preferential attachment power \gls{alpha} of
+      \acrlong{BA} network model, obtained on simulated trees with
+      kernel-score-based grid search. Test trees were simulated according to
+      several values of \gls{alpha} ($x$-axis) with other model parameters
+      fixed at \gls{m} = 2, \gls{N} = 5000, and \gls{I} = 1000. The test trees
+      were compared to trees simulated along a narrowly spaced grid of
+      \gls{alpha} values using the tree kernel, with the same values of the
+      other parameters. The grid value with the highest median kernel score was
+      taken as a point estimate for \gls{alpha} ($y$-axis).
+  }
+  \label{fig:gridptalpha}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics{gridsearch-I-point-estimate.pdf}
+  \caption{
+      Point estimates of prevalence at time of sampling \gls{I} of \acrlong{BA}
+      network model, obtained on simulated trees with kernel-score-based grid
+      search. Test trees were simulated according to several values of \gls{I}
+      ($x$-axis) with other model parameters fixed at \gls{alpha} = 1, \gls{m}
+      = 2, and \gls{N} = 5000. The test trees were compared to trees simulated
+      along a narrowly spaced grid of \gls{I} values using the tree kernel,
+      with the same values of the other parameters. The grid value with the
+      highest median kernel score was taken as a point estimate for \gls{I}
+      ($y$-axis).
+  }
+  \label{fig:gridptI}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics{gridsearch-m-point-estimate.pdf}
+  \caption{
+      Point estimates of number of edges per vertex \gls{m} of \acrlong{BA}
+      network model, obtained on simulated trees with kernel-score-based grid
+      search. Test trees were simulated according to several values of \gls{m}
+      ($x$-axis) with other model parameters fixed at \gls{alpha} = 1, \gls{I}
+      = 1000, and \gls{N} = 5000. The test trees were compared to trees
+      simulated along a narrowly spaced grid of \gls{m} values using the tree
+      kernel, with the same values of the other parameters. The grid value with
+      the highest median kernel score was taken as a point estimate for \gls{m}
+      ($y$-axis).
+  }
+  \label{fig:gridptm}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics{gridsearch-N-point-estimate.pdf}
+  \caption{
+      Point estimates of number of edges per vertex \gls{N} of \acrlong{BA}
+      network model, obtained on simulated trees with kernel-score-based grid
+      search. Test trees were simulated according to several values of \gls{N}
+      ($x$-axis) with other model parameters fixed at \gls{alpha} = 1, \gls{m}
+      = 2, and \gls{I} = 1000. The test trees were compared to trees simulated
+      along a narrowly spaced grid of \gls{N} values using the tree kernel,
+      with the same values of the other parameters. The grid value with the
+      highest median kernel score was taken as a point estimate for \gls{m}
+      ($y$-axis).
+  }
+  \label{fig:gridptN}
+\end{figure}
+
+\begin{figure}[H]
   \centering
@@
   }        
-  \label{fig:abcm3}
+  \label{fig:abcptm3}
 \end{figure}
 
-\begin{figure}
+\begin{figure}[H]
   \centering
@@
   }        
-  \label{fig:abcm4}
+  \label{fig:abcptm4}
 \end{figure}
 
-\begin{figure}
+\begin{figure}[ht]
+  \centering
+  \includegraphics{alpha-gamma.pdf}
+  \caption{
+      Relationship between preferential attachment power parameter $\alpha$
+      and power law exponent $\gamma$ for networks simulated under the BA
+      network model with $N$ = 5000 and $m$ = 2.
+  }
+  \label{fig:gamma}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{mixed-posterior}
+    \caption[Approximate marginal posterior distributions of BA model
+        parameters obtained using kernel-ABC for a network with heterogeneous
+        node behaviour.]
+    {
+        Approximate marginal posterior distributions of Barab\'asi-Albert
+        model parameters obtained using kernel-ABC for a network with
+        heterogeneous node behaviour. Half of the nodes were attached with
+        $\alpha = 0.5$, and the other half with $\alpha = 1.5$ (vertical
+        dashed lines, top left). Other parameter values were $m = 2$, $I =
+        1000$, and $N = 5000$ (vertical dashed lines, other than top left).
+        Shaded areas indicate 95\% highest posterior density intervals.
+    }
+    \label{fig:mixed}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{peerdriven-posterior}
+    \caption[Approximate marginal posterior distributions of Barab\'asi-Albert
+        model parameters obtained using kernel-ABC for a network with
+        peer-driven sampling.]
+    {
+        Approximate marginal posterior distributions of Barab\'asi-Albert
+        model parameters obtained using kernel-ABC for a network with
+        peer-driven sampling. An epidemic was simulated in the usual fashion,
+        but rather than being sampled at random, infected nodes were sampled
+        with a probability two times higher if they had any sampled neighbours
+        in the contact network. Vertical dashed lines indicate true parameter
+        values, and shaded areas indicate 95\% highest posterior density
+        intervals.
+    }
+    \label{fig:peerdriven}
+\end{figure}
+
+\begin{figure}[ht]
+  \includegraphics{realdata-hpd-bc-m2}
+  \vspace{8pt}
+  \caption{
+      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
+      parameters of the BA network model, fitted to five published HIV datasets
+      with kernel-ABC. $x$-axes indicate regions of nonzero prior density. In
+      particular, the prior on $m$ was DiscreteUniform(2, 5).
+  }
+  \label{fig:abchpdm2}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics{bctree-posterior}
+  \caption{
+      Posterior distribution of BA model parameters for BC data. Vertical lines
+      indicate maximum \textit{a posteriori} estimates, and shaded areas are
+      95\% highest posterior density intervals. $x$-axis indicates regions of
+      nonzero prior density.
+  }
+  \label{fig:bctree}
+\end{figure}
+
+\begin{figure}[H]
   \includegraphics{cuevas2009-posterior}
@@
     for (I in c(1000, 2000)) {
-    for (alpha in c(0, 0.5, 1, 1.5, 2)) {
-        fn <- sprintf("abc-posterior/%2f_%d_%d_%d_%d.pdf", alpha, I, m, N, replicate)
-        lab <- sprintf("fig:%2f-%d-%d-%d-%d", alpha, I, m, N, replicate)
-        cat("\\begin{figure}\n")
+    for (alpha in c(0, 0.5, 1, 1.5)) {
+        fn <- sprintf("{abc-posterior/%.1f_%d_%d_%d_%d}.pdf", alpha, I, m, N, replicate)
+        lab <- sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate)
+        cat("\\begin{figure}[H]\n")
         cat("\\centering\n")
-        cat("\\includegraphics{", fn, "}\n")
-        cat("\\caption{\n")
-        cat("Approximate marginal posterior distributions of Barab\\'asi-Albert model parameters obtained using kernel-ABC for one simulated dataset.
-Dashed lines indicate true values, and shaded areas shown 95\\% highest posterior density intervals.\n")
+        cat(paste0("\\includegraphics{", fn, "}\n"))
+        cat(sprintf("\\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\\alpha$ = %.1f, $I$ = %d, $m$ = %d, and $N$ = %d]{\n", alpha, I, m, N))
+        cat(sprintf("Approximate marginal posterior distributions of BA model parameters obtained by applying \textit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = %.1f, $I$ = %d, $m$ = %d, and $N$ = %d.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\\%% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.\n", alpha, I, m, N))
         cat("}\n")
-        cat("\\label{", lab, "}\n")
+        cat(paste0("\\label{", lab, "}\n"))
+        cat("\\end{figure}\n")
+        cat("\\clearpage\n\n")
     }

### 2016-04-27 doc/thesis/appendix.tex
Thesis almost done

@@
   }        
-  \label{fig:abcm4}
+  \label{fig:abcptm4}
 \end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{alpha-gamma.pdf}
+  \caption{
+      Relationship between preferential attachment power parameter $\alpha$
+      and power law exponent $\gamma$ for networks simulated under the BA
+      network model with $N$ = 5000 and $m$ = 2.
+  }
+  \label{fig:gamma}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{mixed-posterior}
+    \caption[Approximate marginal posterior distributions of BA model
+        parameters obtained using kernel-ABC for a network with heterogeneous
+        node behaviour.]
+    {
+        Approximate marginal posterior distributions of Barab\'asi-Albert
+        model parameters obtained using kernel-ABC for a network with
+        heterogeneous node behaviour. Half of the nodes were attached with
+        $\alpha = 0.5$, and the other half with $\alpha = 1.5$ (vertical
+        dashed lines, top left). Other parameter values were $m = 2$, $I =
+        1000$, and $N = 5000$ (vertical dashed lines, other than top left).
+        Shaded areas indicate 95\% highest posterior density intervals.
+    }
+    \label{fig:mixed}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{peerdriven-posterior}
+    \caption[Approximate marginal posterior distributions of Barab\'asi-Albert
+        model parameters obtained using kernel-ABC for a network with
+        peer-driven sampling.]
+    {
+        Approximate marginal posterior distributions of Barab\'asi-Albert
+        model parameters obtained using kernel-ABC for a network with
+        peer-driven sampling. An epidemic was simulated in the usual fashion,
+        but rather than being sampled at random, infected nodes were sampled
+        with a probability two times higher if they had any sampled neighbours
+        in the contact network. Vertical dashed lines indicate true parameter
+        values, and shaded areas indicate 95\% highest posterior density
+        intervals.
+    }
+    \label{fig:peerdriven}
+\end{figure}
+
+\begin{figure}[ht]
+  \includegraphics{realdata-hpd-bc-m2}
+  \vspace{8pt}
+  \caption{
+      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
+      parameters of the BA network model, fitted to five published HIV datasets
+      with kernel-ABC. $x$-axes indicate regions of nonzero prior density. In
+      particular, the prior on $m$ was DiscreteUniform(2, 5).
+  }
+  \label{fig:abchpdm2}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics{bctree-posterior}
+  \caption{
+      Posterior distribution of BA model parameters for BC data. Vertical lines
+      indicate maximum \textit{a posteriori} estimates, and shaded areas are
+      95\% highest posterior density intervals. $x$-axis indicates regions of
+      nonzero prior density.
+  }
+  \label{fig:bctree}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics{cuevas2009-posterior}
+  \caption{
+      Posterior distribution of BA model parameters for
+      \textcite{cuevas2009hiv} data. Vertical lines indicate maximum \textit{a
+      posteriori} estimates, and shaded areas are 95\% highest posterior
+      density intervals. $x$-axis indicates regions of nonzero prior density.
+  }
+  \label{fig:cuevas}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics{li2015-posterior}
+  \caption{
+      Posterior distribution of BA model parameters for \textcite{li2015hiv}
+      data. Vertical lines indicate maximum \textit{a posteriori} estimates,
+      and shaded areas are 95\% highest posterior density intervals. $x$-axis
+      indicates regions of nonzero prior density.
+  }
+  \label{fig:li}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics{niculescu2015-posterior}
+  \caption{
+      Posterior distribution of BA model parameters for
+      \textcite{niculescu2015recent} data. Vertical lines indicate maximum
+      \textit{a posteriori} estimates, and shaded areas are 95\% highest
+      posterior density intervals. $x$-axis indicates regions of nonzero prior
+      density.
+  }
+  \label{fig:niculescu}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics{novitsky2014-posterior}
+  \caption{
+      Posterior distribution of BA model parameters for
+      \textcite{novitsky2013phylogenetic, novitsky2014impact} data. Vertical
+      lines indicate maximum \textit{a posteriori} estimates, and shaded areas
+      are 95\% highest posterior density intervals. $x$-axis indicates regions
+      of nonzero prior density.
+  }
+  \label{fig:novitsky}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics{wang2015-posterior}
+  \caption{
+      Posterior distribution of BA model parameters for
+      \textcite{wang2015targeting} data. Vertical lines indicate maximum
+      \textit{a posteriori} estimates, and shaded areas are 95\% highest
+      posterior density intervals. $x$-axis indicates regions of nonzero prior
+      density.
+  }
+  \label{fig:wang}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.0_1000_2_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 1000, $m$ = 2, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.0-1000-2-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.5_1000_2_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 1000, $m$ = 2, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.5, $I$ = 1000, $m$ = 2, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.5-1000-2-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.0_1000_2_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.0-1000-2-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.5_1000_2_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 1000, $m$ = 2, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.5, $I$ = 1000, $m$ = 2, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.5-1000-2-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.0_2000_2_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 2000, $m$ = 2, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.0, $I$ = 2000, $m$ = 2, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.0-2000-2-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.5_2000_2_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 2000, $m$ = 2, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.5, $I$ = 2000, $m$ = 2, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.5-2000-2-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.0_2000_2_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 2000, $m$ = 2, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.0, $I$ = 2000, $m$ = 2, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.0-2000-2-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.5_2000_2_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 2000, $m$ = 2, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.5, $I$ = 2000, $m$ = 2, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.5-2000-2-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.0_1000_3_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 1000, $m$ = 3, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.0, $I$ = 1000, $m$ = 3, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.0-1000-3-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.5_1000_3_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 1000, $m$ = 3, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.5, $I$ = 1000, $m$ = 3, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.5-1000-3-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.0_1000_3_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 1000, $m$ = 3, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.0, $I$ = 1000, $m$ = 3, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.0-1000-3-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.5_1000_3_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 1000, $m$ = 3, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.5, $I$ = 1000, $m$ = 3, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.5-1000-3-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.0_2000_3_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 2000, $m$ = 3, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.0, $I$ = 2000, $m$ = 3, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.0-2000-3-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.5_2000_3_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 2000, $m$ = 3, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.5, $I$ = 2000, $m$ = 3, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.5-2000-3-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.0_2000_3_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 2000, $m$ = 3, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.0, $I$ = 2000, $m$ = 3, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.0-2000-3-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.5_2000_3_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 2000, $m$ = 3, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.5, $I$ = 2000, $m$ = 3, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.5-2000-3-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.0_1000_4_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 1000, $m$ = 4, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.0, $I$ = 1000, $m$ = 4, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.0-1000-4-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.5_1000_4_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 1000, $m$ = 4, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.5, $I$ = 1000, $m$ = 4, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.5-1000-4-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.0_1000_4_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 1000, $m$ = 4, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.0, $I$ = 1000, $m$ = 4, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.0-1000-4-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.5_1000_4_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 1000, $m$ = 4, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.5, $I$ = 1000, $m$ = 4, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.5-1000-4-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.0_2000_4_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 2000, $m$ = 4, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.0, $I$ = 2000, $m$ = 4, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.0-2000-4-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.5_2000_4_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 2000, $m$ = 4, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.5, $I$ = 2000, $m$ = 4, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.5-2000-4-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.0_2000_4_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 2000, $m$ = 4, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.0, $I$ = 2000, $m$ = 4, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.0-2000-4-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.5_2000_4_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 2000, $m$ = 4, and $N$ = 5000]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.5, $I$ = 2000, $m$ = 4, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.5-2000-4-5000-0}
+\end{figure}
+\clearpage

### 2016-04-27 doc/thesis/discussion.tex
Thesis almost done

@@
 
-The method behind \software{netabc} is model-agnostic, meaning that it can be
-used to infer parameters of any network model, as long as it allows simulated
-networks can be easily generated. We have included generators for the \gls{BA}
-model discussed here, as well as the \gls{ER} and \gls{WS} network models.
+Contact networks can have a strong influence on epidemic progression, and are
+potentially useful as a public health tool~\autocite{wang2015targeting,
+little2014using}. Despite this, few methods exist for investigating contact
+network parameters in a phylodynamic framework \autocite[although see][for
+related work]{groendyke2011bayesian, volz2008sir, brown2011transmission,
+leventhal2012inferring, greenbaum2016inference}. Kernel-ABC is a model-agnostic
+method which can be used to investigate any quantity that affects tree
+shape~\autocite{poon2015phylodynamic}. In this work, we developed
+\software{netabc}, a method based on kernel-ABC to infer the parameters of a
+contact network model. The method is general, meaning that it can be used to
+infer parameters of any network model, as long as it allows simulated networks
+can be easily generated. We have included generators for the \gls{BA} model
+discussed here, as well as the \gls{ER} and \gls{WS} network models.
 Instructions for adding additional models are available in the project's online
@@
 statistics of network features such as pairs and triads.
-\textcite{goodreau2007assessing} evaluated the effect of several different
+\textcite{goodreau2006assessing} evaluated the effect of several different
 \gls{ERGM} parameterizations on transmission tree shape and effective
@@
 developed by~\textcite{eames2002modeling} and the log-linear adjacency matrix
-parameterization developed by~\textcite{morriz1993epidemiology}.
+parameterization applied by~\textcite{morris1993epidemiology}.
 
@@
 It is important to note that \software{netabc} takes a transmission tree as
-input, rather than a viral phylogeny. Thus, we have left the estimation of a
-transmission tree up to the user. In theory, it is possible to incorporate the
-process by which a viral phylogeny is generated along with a transmission tree
-into our method, for example by simulating within-host dynamics. Although this
-may be an avenue for future extension, we felt that it would obscure the
-primary purpose of this work, which is to study contact network parameters.
-In addition, there are a number of different methods available for inferring
-transmission trees~\autocite{didelot2014bayesian, ypma2012unravelling,
-jombart2011reconstructing, cottam2008integrating, poon2015phylodynamic}, some
-of which incorporate geographic and/or epidemiological data not accommodated by
-our method. We therefore felt it would be best to allow researchers to use
-their own preferred tree building method.
+input, rather than a viral phylogeny. In reality, true transmission trees are
+not available and must be approximated, often by way of a viral phylogeny.
+Although this has been demonstrated to be a fair
+approximation~\autocite[e.g.][]{leitner1996accurate}, and is frequently used in
+practice~\autocite[e.g.][]{stadler2013uncovering}, the topologies of a viral
+phylogeny and transmission tree can differ
+significantly~\autocite{ypma2013relating} due to within-host evolution and the
+sampling process. We have left the estimation of a transmission tree up to the
+user. In theory, it is possible to incorporate the process by which a viral
+phylogeny is generated along with a transmission tree into our method, for
+example by simulating within-host dynamics. Although this may be an avenue for
+future extension, we felt that it would obscure the primary purpose of this
+work, which is to study contact network parameters. In addition, there are a
+number of different methods available for inferring transmission
+trees~\autocite{didelot2014bayesian, ypma2012unravelling,
+jombart2011reconstructing, cottam2008integrating}, some of which incorporate
+geographic and/or epidemiological data not accommodated by our method. We
+therefore felt it would be best to allow researchers to use their own preferred
+method of constructing a transmission tree.
 
@@
 is possible that the use of a more sophisticated technique might increase the
-algorithm's accuracy.
+algorithm's accuracy. Finally, the \gls{ABC}-\gls{SMC} algorithm is
+computationally intensive, taking about a day when run on 20 cores in parallel
+with the settings described in the methods section.
 
@@
 parameters may translate to highly informative priors on quantities of
-interest. We observed a non-linear relationship between the pcreferential
+interest. We observed a non-linear relationship between the preferential
 attachment power $\alpha$ and the power law exponent $\gamma$
@@
 
-With respect to the pcreferential attachment power $\alpha$, the five datasets
+With respect to the preferential attachment power $\alpha$, the five datasets
 analysed fell into three categories (\cref{fig:abchpd}). First, we
-estimated a pcreferential attachment power close to 1, indicating linear
-pcreferential attachment, for the outbreaks studied by
+estimated a preferential attachment power close to 1, indicating linear
+preferential attachment, for the outbreaks studied by
 \textcite{niculescu2015recent} and \textcite{wang2015targeting}. These values
@@
 which we would expect some of the assumptions of the BA model, such as a
-connected network, relatively high mean degree, and pcreferential attachment
+connected network, relatively high mean degree, and preferential attachment
 dynamics, to hold.
@@
 would be unlikely to significantly interact, further weakening any global
-pcreferential attachment dynamics. The dataset studied by
+preferential attachment dynamics. The dataset studied by
 \textcite{novitsky2014impact} originated from a densely sampled population

### 2016-04-27 doc/thesis/intro.tex
Thesis almost done

@@
   \includegraphics{speciestree.pdf}
-  \caption[Illustration of a rooted, ultrametric, time-scaled phylogeny]
+  \caption[Illustration of a rooted, ultrametric, time-scaled phylogeny.]
     {Illustration of a rooted, ultrametric, time-scaled phylogeny. The tips of
@@
     \includegraphics[width=\textwidth]{contactnet.pdf}
-    \caption[Illustration of a contact network and transmission tree]{
+    \caption[Illustration of a contact network and transmission tree.]{
       Illustration of epidemic spread over a contact network, and the
@@
 \citeauthor{barabasi1999emergence} suggest extending the model such that the
-probability of choosing a partner of degree $d$ is proportional to $d^\alpha$
-for some constant $\alpha$, and we use this extension here. When $m = 1$, the
-network takes on the distinctive shape of a tree, that is, it does not contain
-any cycles. Cycles are present in the network for all all other $m$ values.
+probability of choosing a partner of degree $d$ is proportional to $d^\alpha +
+1$ for some constant $\alpha$, and we use this extension here. When $m = 1$,
+the network takes on the distinctive shape of a tree, that is, it does not
+contain any cycles. Cycles are present in the network for all all other $m$
+values.
 

### 2016-04-27 doc/thesis/methods.tex
Thesis almost done

@@
     \includegraphics{abc-smc.pdf}
-    \caption[Graphical schematic of \gls{ABC}-\gls{SMC} algorithm.]{
+    \caption[Graphical schematic of the ABC-SMC algorithm implemented in \software{netabc}.]{
       Graphical schematic of the \gls{ABC}-\gls{SMC} algorithm implemented in
@@
 the particle population, which indicates the quality of the Monte Carlo
-approximation (see \cref{subsec:sis}), decays at a controlled rate. This
-prevents the Monte Carlo approxmiation from collapsing by the variance in
-weights growing too large. A single parameter $\alpha$ (not to be confused with
-the \gls{BA} model parameter) controls the decay rate, with $\varepsilon_i$
-being chosen to satisfy
+approximation (see \cref{subsec:sis}), decays at a controlled rate. A
+sudden precipitous drop in \gls{ESS} would indicate that only a small number of
+particles had non-zero importance weights, which would result in a very poor
+Monte Carlo approximation to the target distribution. This situation is
+referred to as the ``collapse'' of the approximation, and is mitigated by the
+adaptive approach. A single parameter $\alpha$ (not to be confused with the
+\gls{BA} model parameter) controls the decay rate, with $\varepsilon_i$ being
+chosen to satisfy
 \[
@@
 attached to existing nodes of degree $k$ with probability proportional to
-$k^\alpha$ (\cref{subsec:pa}). To simulate transmission trees over a
+$k^\alpha + 1$ (\cref{subsec:pa}). To simulate transmission trees over a
 \gls{BA} network, we allowed an epidemic to spread until \gls{I} nodes were
@@
   \includegraphics{kernel-expt.pdf}
-  \caption[Schematic of kernel classifier experiment]{
-    Schematic of investigation of \gls{BA} model parameters using the tree
-    kernel.
+  \caption[Schematic of experiments investigating impact of BA model parameters
+           on tree shape.]{
+    Schematic of experiments designed to investigate the impact of variations
+    in BA model parameters on transmission tree shapes. The parameters of the
+    BA model were varied one at a time. Transmission trees were simulated under
+    three different values of each parameter, then compared pairwise using the
+    tree kernel. Classifiers were constructed for each parameter, and their
+    accuracy was evaluated by cross-validation. Kernel-PCA projections were
+    used to visually examine the separation of the trees in the feature space
+    defined by the tree kernel.
   }
@@
   \includegraphics[width=\textwidth]{gridsearch-expt}
-  \caption{
+  \caption[Schematic of grid search experiment.]{
     Graphical schematic of grid search experiments used to investigate \gls{BA}
@@
 \gls{HIV}-positive individuals in British Columbia, Canada (the ``BC data'').
-The datasets are summarized in \cref{tab:realdata}.
+The datasets are summarized in \cref{tab:data}.
 
@@
   \input{\tablepath/realdata-bc}
-  \caption{Characteristics of published HIV datasets analyzed with kernel-ABC.}
-  \label{tab:realdata}
+  \caption[Characteristics of published HIV datasets analyzed with \software{netabc}.]
+  {
+    Characteristics of published HIV datasets analyzed with \software{netabc}. 
+    Abbreviations: MSM, men who have sex with men; HET, heterosexual; IDU,
+    injection drug users. The \textcite{novitsky2013phylogenetic,novitsky2014impact} data
+    were sampled from a primarily heterosexual risk environment, but did not
+    explicitly exclude other risk factors.
+  }
+  \label{tab:data}
 \end{table}

### 2016-04-27 doc/thesis/objective.tex
Thesis almost done

@@
 \glsreset{ABC}
+\glsreset{BA}
 
@@
 homogeneously mixed groups~\autocite{rushton1955deterministic}. This
-assumption, also called the ``law of mass action'' or \defn{panmixia}, implies
+assumption, also called \defn{mass
+action}~\autocite{heesterbeek2000mathematical}, or \defn{panmixia}, implies
 that any two individuals in the same compartment are equally likely to come
@@
 introduced by assuming a panmictic population can be substantial when
-significant contact heterogeneity exists in the underlying population. Contact
-network models provide an alternative to compartmental models which do not
-require the assumption of panmixia. In addition to more accurate predictions,
-the parameters of the networks themselves may be of interest from a public
-health perspective. For example, certain vaccination strategies may be more or
-less effective in curtailing an epidemic depending on the underlying network's
-degree distribution~\autocite{peng2013vaccination, ma2013importance}.
+significant contact heterogeneity exists in the underlying
+population~\autocite{bansal2007individual,barthelemy2005dynamical,keeling2005networks}.
+Contact network models provide an alternative to compartmental models which do
+not require the assumption of panmixia. In addition to more accurate
+predictions, the parameters of the networks themselves may be of interest from
+a public health perspective. For example, certain vaccination strategies may be
+more or less effective in curtailing an epidemic depending on the underlying
+network's degree distribution~\autocite{peng2013vaccination, ma2013importance}.
 Phylodynamic methods have been used to fit many different types of model to
@@
 may individually be straightforward to calculate, the number of possible
-directed graphs on $N$ nodes is $2^{N(N-1)}$, larger if the nodes and edges in
-the graph may have different labels or attributes. Hence, the number of terms
-in the sum is at least exponential in $n$, as there must be at least $n$ nodes
-in the network. In addition, \cref{eq:netlik} assumes that $T$ is complete,
-meaning that all infected individuals were sampled. This is rarely the case in
-practice - most often, we only have access to a subset of the infected
-individuals. In this case, the likelihood calculation becomes even more
-complex, because we must also sum over all possible complete trees.
+directed graphs on $N$ nodes is $2^{N(N-1)}$~\autocite{harary2014graphical},
+larger if the nodes and edges in the graph may have different labels or
+attributes. Hence, the number of terms in the sum is at least exponential in
+$n$, as there must be at least $n$ nodes in the network. In addition,
+\cref{eq:netlik} assumes that $T$ is complete, meaning that all infected
+individuals were sampled. This is rarely the case in practice - most often, we
+only have access to a subset of the infected individuals. In this case, the
+likelihood calculation becomes even more complex, because we must also sum over
+all possible complete trees.
 
@@
 to likelihood-based methods, which would apply to any network model, is
-provided by \gls{ABC}. All of the ingredients required to apply \gls{ABC} to
-this problem are readily available. Simulating networks is straightforward
-under a variety of models. Epidemics on those networks, and the corresponding
-transmission trees, can also be easily simulated. As mentioned above, contact
-networks can profoundly affect transmission tree shape, and those shapes can be
-compared using a highly informative similarity measure called the ``tree
+provided by
+\gls{ABC}~\autocite{rubin1984bayesianly,tavare1997inferring,fu1997estimating,beaumont2002approximate}.
+All of the ingredients required to apply \gls{ABC} to this problem are readily
+available. Simulating networks is straightforward under a variety of models.
+Epidemics on those networks, and the corresponding transmission trees, can also
+be easily simulated. As mentioned above, contact networks can profoundly affect
+transmission tree shape, and those shapes can be compared using a highly
+informative similarity measure called the ``tree
 kernel''~\autocite{poon2013mapping}. \Gls{ABC} can be implemented with
@@
 attachment has been postulated as a mechanism by which scale-free networks
-could be generated~\autocite{barabasi1999emergence}. This makes the \gls{BA}
-model, one of the simplest preferential attachment models, a natural choice to
-explore with our method. The second aim of this work is to use simulations to
-investigate the parameters of the \gls{BA} model, including whether they have a
-detectable impact on tree shape, and whether they can be accurately recovered
-using kernel-\gls{ABC}.
+could be generated~\autocite{barabasi1999emergence}. The \gls{BA}
+model~\autocite{barabasi1999emergence} is one of the simplest preferential
+attachment models, which makes it a natural choice to explore with our method.
+The second aim of this work is to use simulations to investigate the parameters
+of the \gls{BA} model, including whether they have a detectable impact on tree
+shape, and whether they can be accurately recovered using kernel-\gls{ABC}.
 
@@
 the underlying structure of the network and the particular nodes to whom the
-intervention is targeted. Due to this combination of data availability and
-potential public health impact, \gls{HIV} is an obvious context in which our
-method could be applied. Therefore, the third and final aim of this work is to
-apply kernel-\gls{ABC} to fit the \gls{BA} model to existing \gls{HIV}
-outbreaks.
+intervention is targeted~\autocite{little2014using,wang2015targeting}. Due to
+this combination of data availability and potential public health impact,
+\gls{HIV} is an obvious context in which our method could be applied.
+Therefore, the third and final aim of this work is to apply kernel-\gls{ABC} to
+fit the \gls{BA} model to existing \gls{HIV} outbreaks.
 
@@
 shape and how accurately they can be recovered with kernel-\gls{ABC}. Finally,
-we will apply the method fit the \gls{BA} model to several real-world \gls{HIV}
-datasets.
+we will apply the method to fit the \gls{BA} model to several real-world
+\gls{HIV} datasets.
 

### 2016-04-27 doc/thesis/results.tex
Thesis almost done

@@
 which were as easily distinguished from each other
-(\cref{fig:Itrees,fig:mtrees,fig:Ntrees,fig:Itrees}).
-Sackin's index, which measures tree imbalance, was significantly correlated with
-all four parameters
+(\cref{fig:Itrees,fig:mtrees,fig:Ntrees,fig:Itrees}).  Sackin's index, which
+measures tree imbalance, was significantly correlated with all four parameters
     (for $\alpha$, $I$, $m$, and $N$ respectively: Spearman's rho =
@@
   \centering
-  \includegraphics{kernel-alpha-tree.pdf}
-  \caption[Visibly distinctive trees simulated under three values of \gls{alpha}]{
-    Epidemics simulated on \gls{BA} networks of 5000 nodes, with \gls{alpha}
-    equal to 0.5, 1.0, or 1.5, until 1000 individuals were infected.
-    Transmission trees were created by sampling 500 infected nodes. Higher
-    \gls{alpha} values produced networks with a small number of
+  \includegraphics[width=\textwidth]{kernel-alpha-tree.pdf}
+  \caption[Simulated transmission trees under three different values of BA parameter $\alpha$]{
+    Simulated transmission trees under three different values of BA parameter
+    $\alpha$. Epidemics were simulated on \gls{BA} networks of 5000 nodes, with
+    \gls{alpha} equal to 0.5, 1.0, or 1.5, until 1000 individuals were
+    infected. Transmission trees were created by sampling 500 infected nodes.
+    Higher \gls{alpha} values produced networks with a small number of
     highly-connected nodes, resulting in highly unbalanced, ladder-like trees.
@@
   \includegraphics{kernel-kpca.pdf}
-  \caption[\gls{kPCA} projections of simulated trees under varying \gls{BA}
-           parameter values]{
+  \caption[Kernel-PCA projections of simulated trees under varying BA
+           parameter values.]{
     Each parameter of the \gls{BA} model was individually varied to produce 300
@@
     (average $R^2$ 
-     0,
-     0,
-     0
+     0.01,
+     0.11,
+     0.32
      for 100, 500, and 1000 tips respectively).
@@
   \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
-  \caption{
+  \caption[Cross-validation accuracy of kernel-SVR, nLTT-based SVR, and
+  Sackin's index regression classifiers for BA model parameters.]{
       Cross-validation accuracy of kernel-SVR classifier (left), SVR classifier
@@
 intervals for each of the \gls{BA} parameters, for one replicate experiment
-with 500-tip trees. For all parameters except $m$, the error of point estimates
-was negatively correlated with the number of sampled tips in the tree (for
+with 500-tip trees. Plots showing the point estimates for all replicates can be
+found in \cref{fig:gridptalpha,fig:gridptI,fig:gridptm,fig:gridptN}. For all
+parameters except $m$, the error of point estimates was negatively correlated
+with the number of sampled tips in the tree (for
 \gls{alpha}, \gls{I}, and \gls{N} respectively: Spearman's $\rho$ = 
@@
     $p\ {<}10^{-5}$). 
+Most of the kernel score distributions were extremely flat, except for the
+value $N = 1000$ which exhibited a pronounced peak around the true value
+(\cref{fig:gridN}).
 
@@
   \includegraphics[width=\textwidth]{gridsearch-example}
-  \caption[Grid search estimates of \gls{BA} model parameters]{Point estimates
+  \caption[Grid search estimates of \gls{BA} model parameters.]{Point estimates
       and 95\% highest density intervals for each \gls{BA} model parameter,
@@
 simulated trees where the true parameter values were known. Point estimates for
-each parameter are shown in \cref{fig:abcpt} for the simulations with \gls{m} =
+each parameter are shown in \cref{fig:abcptm2} for the simulations with \gls{m} =
 2. The results for the other values of \gls{m} were similar
-(\cref{fig:abcm3,fig:abcm4}). The median [IQR] absolute error of estimates of
+(\cref{fig:abcptm3,fig:abcptm4}). The median [IQR] absolute error of estimates of
 \gls{alpha} across all simulations was
-    0.11 
-    [0.05-0.18].
+    0.08 
+    [0.05-0.17].
 The accuracy of the estimates was not significantly different between values of
 $m$ or $I$ (both one-way ANOVA,
-    $p$ = 0.1
+    $p$ = 0.05
 and 
-    0.25),
+    0.07),
 although the errors when the true value of \gls{alpha} was zero were
 significantly greater than the other values 
-    (Wilcoxon rank-sum test, $p = 6\!\times\!10^{-4}$).
+    (Wilcoxon rank-sum test, $p = 8\!\times\!10^{-3}$).
 The error in the estimated value of $I$ was
-    306 
-    [108-607].
+    395 
+    [207-683].
 Errors were significantly higher for $\alpha \geq 1$
-    (Wilcoxon rank-sum test, $p = 6\!\times\!10^{-4}$)
+    (Wilcoxon rank-sum test, $p = 8\!\times\!10^{-3}$)
 and for $I$ = 2000
@@
 correctly in only
-    37 \%
+    27 \%
 of simulations. The true values of $m$ and $I$ did not significantly affect the
@@
 than non-integral values of \gls{alpha}
-    (Wilcoxon rank-sum test, $p = 7\!\times\!10^{-3}$).
+    (Wilcoxon rank-sum test, $p = 0.3$).
 Finally, the total number of nodes \gls{N} was consistently over-estimated by
 about a factor of two
-    (error 6588 
-    [4214 - 
-     8284]).
+    (error 5987 
+    [2060 - 
+     7999]).
 No other parameters influenced the accuracy of the $N$ estimates (one-way
-ANOVA)
+ANOVA).
 
@@
   \includegraphics{abc-point-estimate-m2}
-  \caption{
+  \caption[\Acrlong{MAP} point estimates for \gls{BA} model parameters obtained
+    by running \software{netabc} on simulated data, for simulations with $m = 2$.] 
+  {
     \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by         
     running \software{netabc} on simulated data. Values shown are for               
-    simulations with \gls{m} = 3. Dashed lines indicate true values. (A)            
+    simulations with \gls{m} = 2. Dashed lines indicate true values. (A)            
     Estimates of \gls{alpha} and \gls{I} which were varied in these simulations  
     against known values. (B) Estimates of \gls{m} and \gls{N} which were held   
-    fixed in these simulations at the values \gls{m} = 3 and \gls{N} = 5000. 
+    fixed in these simulations at the values \gls{m} = 2 and \gls{N} = 5000. 
   }
-  \label{fig:abcm2}
+  \label{fig:abcptm2}
 \end{figure}
@@
 \Cref{fig:abcex} shows the \gls{ABC} approximation to the posterior
-distribution on the \gls{BA} parameters for one simulation (equivalent plots
-for all the simulations can be found in the supplemental materials). \Gls{HPD}
-intervals around \gls{alpha} and \gls{I} were narrow relative to the region of
-nonzero prior density, whereas the intervals for $m$ and \gls{N} were widely
-dispersed. \Cref{tab:abchpd} shows point estimates and 95\% \gls{HPD}
-intervals averaged over all simulations.
+distribution on the \gls{BA} parameters for one simulation. Equivalent plots
+for one replicate simulation with each combination of parameters can be found
+in \cref{fig:0.0-1000-2-5000-0,fig:0.5-1000-2-5000-0,fig:1.0-1000-2-5000-0,fig:1.5-1000-2-5000-0,fig:0.0-2000-2-5000-0,fig:0.5-2000-2-5000-0,fig:1.0-2000-2-5000-0,fig:1.5-2000-2-5000-0,fig:0.0-1000-3-5000-0,fig:0.5-1000-3-5000-0,fig:1.0-1000-3-5000-0,fig:1.5-1000-3-5000-0,fig:0.0-2000-3-5000-0,fig:0.5-2000-3-5000-0,fig:1.0-2000-3-5000-0,fig:1.5-2000-3-5000-0,fig:0.0-1000-4-5000-0,fig:0.5-1000-4-5000-0,fig:1.0-1000-4-5000-0,fig:1.5-1000-4-5000-0,fig:0.0-2000-4-5000-0,fig:0.5-2000-4-5000-0,fig:1.0-2000-4-5000-0,fig:1.5-2000-4-5000-0}.
+\Gls{HPD} intervals around \gls{alpha} and \gls{I} were
+narrow relative to the region of nonzero prior density, whereas the intervals
+for $m$ and \gls{N} were widely dispersed. \Cref{tab:abchpd} shows point
+estimates and 95\% \gls{HPD} intervals averaged over all simulations.
+
+\begin{figure}
+  \includegraphics{{abc-posterior/1.0_1000_2_5000_0}.pdf}
+  \caption[Approximate marginal posterior distributions of BA model parameters
+      obtained by applying \textit{netabc} to a simulated transmission tree
+      with values $alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]
+    {
+        Approximate marginal posterior distributions of BA model parameters
+        obtained by applying \textit{netabc} to a simulated transmission tree
+        with BA parameter values $alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ =
+        5000. Vertical dashed lines indicate true values. Shaded areas are 95\%
+        highest posterior density intervals. $x$-axes indicate regions of
+        nonzero prior density.
+    }
+  \label{fig:abcex}
+\end{figure}
 
@@
     parameters estimated with kernel-\gls{ABC}.}
-    \label{tab:glm}
+    \label{tab:abchpd}
 \end{table}
+
+To test the effect of model misspecification, we simulated one network where
+the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
+other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
+HPD] estimates for each parameter were: 
+$\alpha$, 
+  1.09 
+  [0.6 -
+   1.16];
+$I$,
+  1103 
+  [662 -
+   4455];
+$m$,
+  3 
+  [2 -
+   5];
+$N$,
+  12582 
+  [3590-
+   14977].
+The approximate posterior distributions for this simulation are shown in
+\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
+transmission tree in a peer-driven fashion, where the probability to sample a
+node was twice as high if one of its peers had already been sampled. The
+parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
+$I$ = 2000. The estimated values were
+$\alpha$, 
+  0.39 
+  [0 -
+   0.62];
+$I$,
+  2360 
+  [1379 -
+   3811];
+$m$,
+  2 
+  [2 -
+   5];
+$N$,
+  13537 
+  [2855 -
+   14649].
+The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
+of these results were in line with estimates obtained on other simulated
+datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
+$\alpha$ was somewhat lower than typical.
 
@@
 \subsection{Application to HIV data}
+
+We applied kernel-ABC to five published HIV datasets (\cref{tab:data}),
+and found substantial heterogeneity among the parameter estimates
+(\cref{fig:abchpd,fig:abchpdm2}). Plots of the marginal posterior distributions
+for each dataset are shown in
+\cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.
+Two of the datasets~\autocite{niculescu2015recent, wang2015targeting} had
+estimated $\alpha$ values near unity for the prior allowing $m = 1$ (MAP
+estimates [95\% HPD] 
+  1.05 
+  [0.04 - 
+   1.27]
+and
+  0.84 
+  [0.01 -
+   1.02] respectively).
+The MAP estimates did not change appreciably when $m = 1$ was disallowed by the
+prior, although the credible interval of the \textcite{niculescu2015recent}
+data was narrower
+  (0.04 - 
+   1.27).
+When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
+low estimated $\alpha$ values
+  (0.06 
+  [0.01 - 
+  0.73]
+and
+  0.19 
+  [0.01 -
+   0.8]). 
+However, the MAP estimates increased when $m = 1$ was not permitted, although
+the HPD intervals remained roughly the same
+  (0.78 
+  [0.02 - 
+  0.94]
+and
+  0.59 
+  [0.07 -
+   0.95]).
+The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
+for both priors on $m$
+  (0.32 for $m \geq 1$;
+   0.39 for $m \geq 2$).
+However, the confidence interval was much wider when $m = 1$ was allowed
+  ([0.04 -
+    1.62] for $m \geq 1$ vs.
+    0 -
+    0.73 for $m \geq 2$).
+
+For all the datasets except \citeauthor{novitsky2014impact}, estimated values
+of $I$ were below 2000 when $m = 1$ was allowed, with relatively narrow HPD
+intervals compared to the nonzero prior density region
+  (\citeauthor{cuevas2009hiv}, 482 
+  [293 -
+   2111];
+   \citeauthor{niculescu2015recent}, 307
+  [136 - 
+   2822];
+  \citeauthor{li2015hiv}, 1183 
+  [413 -
+   2897];
+   \citeauthor{wang2015targeting}, 719
+  [176 - 
+   2114]).
+The \citeauthor{novitsky2014impact} data was the outlier, with a very high
+estimated $I$, and HPD interval spanning almost the entire prior region
+  (7409 
+  [187 -
+   8819]).
+The $I$ estimates and HPD intervals were generally robust to the choice of
+prior on $m$, with slightly narrower HPD intervals (compare
+\cref{fig:abchpd,fig:abchpdm2}).
+
+The MAP estimate of $m$ was equal to 1 for all but the
+\citeauthor{novitsky2014impact} data, when this value was allowed. However, the
+upper bound of the HPD interval was different for each dataset
+  (\citeauthor{niculescu2015recent}, 5;
+   \citeauthor{wang2015targeting}, 4;
+   \citeauthor{li2015hiv}, 1;
+   \citeauthor{cuevas2009hiv}, 2).
+When $m = 1$ was disallowed, the MAP for all datasets was either 2 or 3, with
+HPD intervals spanning the entire prior region. The estimates for the total
+number of nodes $N$ were largely uninformative for all samples, with almost all
+MAP estimates greater than 7500 and HPD intervals spanning almost the entire
+nonzero prior density region. The only exception was the \citeauthor{li2015hiv}
+data, for which the MAP estimate was lower 
+  (6973)
+when $m = 1$ was allowed.
+
+\begin{figure*}[ht]
+  \centering
+  \includegraphics{realdata-hpd-bc}
+  \caption{
+      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
+      parameters of the BA network model, fitted to five published HIV datasets
+      with kernel-ABC.
+  }
+  \label{fig:abchpd}
+\end{figure*}

### 2016-04-28 doc/thesis/appendix.Rnw
Draft???

@@
   \includegraphics{kernel-alpha-crossv.pdf}
-  \caption{
+  \caption[Cross validation accuracy of classifiers for \gls{BA} model parameter
+    \gls{alpha} for eight epidemic scenarios.]
+  {
     Cross validation accuracy of classifiers for \gls{BA} model parameter
@@
   \includegraphics{kernel-I-crossv.pdf}
-  \caption{
+  \caption[Cross validation accuracy of classifiers for \gls{BA} model parameter
+    \gls{I} for eight epidemic scenarios.]
+  {
     Cross validation accuracy of classifiers for \gls{BA} model parameter
@@
   \includegraphics{kernel-m-crossv.pdf}
-  \caption{
+  \caption[Cross validation accuracy of classifiers for \gls{BA} model parameter
+    \gls{m} for eight epidemic scenarios.]
+  {
     Cross validation accuracy of classifiers for \gls{BA} model parameter
@@
   \includegraphics{kernel-N-crossv.pdf}
-  \caption{
+  \caption[Cross validation accuracy of classifiers for \gls{BA} model parameter
+    \gls{N} for eight epidemic scenarios.]
+  {
     Cross validation accuracy of classifiers for \gls{BA} model parameter
@@
   \includegraphics{gridsearch-alpha-point-estimate.pdf}
-  \caption{
+  \caption[
+      Point estimates of preferential attachment power \gls{alpha} of
+      \acrlong{BA} network model, obtained on simulated trees with
+      kernel-score-based grid search.]
+  {
       Point estimates of preferential attachment power \gls{alpha} of
@@
   \includegraphics{gridsearch-I-point-estimate.pdf}
-  \caption{
+  \caption[ Point estimates of prevalence at time of sampling \gls{I} of
+      \acrlong{BA} network model, obtained on simulated trees with
+      kernel-score-based grid search. ]
+  {
       Point estimates of prevalence at time of sampling \gls{I} of \acrlong{BA}
@@
   \includegraphics{gridsearch-m-point-estimate.pdf}
-  \caption{
+  \caption[
+      Point estimates of number of edges per vertex \gls{m} of \acrlong{BA}
+      network model, obtained on simulated trees with kernel-score-based grid
+      search.
+  ]
+  {
       Point estimates of number of edges per vertex \gls{m} of \acrlong{BA}
@@
   \includegraphics{gridsearch-N-point-estimate.pdf}
-  \caption{
+  \caption[
+      Point estimates of number of edges per vertex \gls{N} of \acrlong{BA}
+      network model, obtained on simulated trees with kernel-score-based grid
+      search.
+  ]{
       Point estimates of number of edges per vertex \gls{N} of \acrlong{BA}
@@
   \includegraphics{abc-point-estimate-m3.pdf}
-  \caption{                                              
+  \caption[
     \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
-    running \software{netabc} on simulated data. Values shown are for
-    simulations with \gls{m} = 3. Dashed lines indicate true values. (A)
-    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations
-    against known values. (B) Estimates of \gls{m} and \gls{N} which were held
-    fixed in these simulations at the values \gls{m} = 3 and \gls{N} = 5000.
+    running \software{netabc} on simulated data, for simulations with \gls{m} = 3.
+  ]{
+    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
+    running \software{netabc} on simulated data, for simulations with \gls{m} = 3.
+    Dashed lines indicate true values. (A) Estimates of \gls{alpha} and \gls{I}
+    which were varied in these simulations against known values. (B) Estimates
+    of \gls{m} and \gls{N} which were held fixed in these simulations at the
+    values \gls{m} = 3 and \gls{N} = 5000.
   }        
@@
   \includegraphics{abc-point-estimate-m4.pdf}
-  \caption{                                              
+  \caption[
     \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
-    running \software{netabc} on simulated data. Values shown are for
-    simulations with \gls{m} = 4. Dashed lines indicate true values. (A)
-    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations
-    against known values. (B) Estimates of \gls{m} and \gls{N} which were held
-    fixed in these simulations at the values \gls{m} = 4 and \gls{N} = 5000.
+    running \software{netabc} on simulated data, for simulations with \gls{m} = 3.
+  ]{
+    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
+    running \software{netabc} on simulated data, for simulations with \gls{m} = 4.
+    Dashed lines indicate true values. (A) Estimates of \gls{alpha} and \gls{I}
+    which were varied in these simulations against known values. (B) Estimates
+    of \gls{m} and \gls{N} which were held fixed in these simulations at the
+    values \gls{m} = 4 and \gls{N} = 5000.
   }        
@@
   \label{fig:gamma}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{sir-trajectories.pdf}
+  \caption[Average trajectories of epidemics simulated according to the
+  \gls{SIR} model over full and \gls{BA} networks.]
+  {
+      Average trajectories of epidemics simulated according to the \gls{SIR}
+      model over full and \gls{BA} networks. 100 epidemics each were simulated
+      over completely connected networks and networks generated from the
+      \gls{BA} model. The network parameters were $\alpha = 1$ and $m = 2$.
+      The transmission rate along each edge of a network $G$ was taken as
+      $2/|E(G)|$, and the removal rate of each vertex was taken as $1/|V(G)|$.
+      These choices ensured that both networks had the same waiting time until
+      the next transmission at the beginning of the epidemic.
+  }
+  \label{fig:sir}
 \end{figure}
@@
   \vspace{8pt}
-  \caption{
+  \caption[
       Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
       parameters of the BA network model, fitted to five published HIV datasets
-      with kernel-ABC. $x$-axes indicate regions of nonzero prior density. In
-      particular, the prior on $m$ was DiscreteUniform(2, 5).
+      with \software{netabc} using the prior $m \sim DiscreteUniform(2, 5)$.
+  ]{
+      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
+      parameters of the BA network model, fitted to five published HIV datasets
+      with \software{netabc} using the prior $m \sim$ DiscreteUniform(2, 5).
+      $x$-axes indicate regions of nonzero prior density. In particular, the
+      prior on $m$ was DiscreteUniform(2, 5).
   }
@@
   \includegraphics{bctree-posterior}
-  \caption{
-      Posterior distribution of BA model parameters for BC data. Vertical lines
-      indicate maximum \textit{a posteriori} estimates, and shaded areas are
-      95\% highest posterior density intervals. $x$-axis indicates regions of
-      nonzero prior density.
+  \caption[
+      Approximate marginal posterior distributions of BA model parameters for
+      BC data.
+  ]
+  {
+      Approximate marginal posterior distributions of BA model parameters for
+      BC data. Vertical lines indicate maximum \textit{a posteriori} estimates,
+      and shaded areas are 95\% highest posterior density intervals. $x$-axis
+      indicates regions of nonzero prior density.
   }
@@
   \includegraphics{cuevas2009-posterior}
-  \caption{
-      Posterior distribution of BA model parameters for
+  \caption[
+      Approximate marginal poosterior distributions of BA model parameters for
+      \textcite{cuevas2009hiv} data. 
+  ]{
+      Approximate marginal poosterior distributions of BA model parameters for
       \textcite{cuevas2009hiv} data. Vertical lines indicate maximum \textit{a
@@
   \includegraphics{li2015-posterior}
-  \caption{
-      Posterior distribution of BA model parameters for \textcite{li2015hiv}
-      data. Vertical lines indicate maximum \textit{a posteriori} estimates,
-      and shaded areas are 95\% highest posterior density intervals. $x$-axis
-      indicates regions of nonzero prior density.
+  \caption[
+      Approximate marginal posterior distributions of BA model parameters for
+      \textcite{li2015hiv} data. 
+  ]{
+      Approximate marginal posterior distributions of BA model parameters for
+      \textcite{li2015hiv} data. Vertical lines indicate maximum \textit{a
+      posteriori} estimates, and shaded areas are 95\% highest posterior
+      density intervals. $x$-axis indicates regions of nonzero prior density.
   }
@@
   \includegraphics{niculescu2015-posterior}
-  \caption{
-      Posterior distribution of BA model parameters for
+  \caption[
+      Approximate marginal posterior distributions of BA model parameters for
+      \textcite{niculescu2015recent} data.
+  ]{
+      Approximate marginal posterior distributions of BA model parameters for
       \textcite{niculescu2015recent} data. Vertical lines indicate maximum
@@
   \includegraphics{novitsky2014-posterior}
-  \caption{
-      Posterior distribution of BA model parameters for
+  \caption[
+      Approximate marginal posterior distributions of BA model parameters for
+      \textcite{novitsky2013phylogenetic, novitsky2014impact} data.
+  ]{
+      Approximate marginal posterior distributions of BA model parameters for
       \textcite{novitsky2013phylogenetic, novitsky2014impact} data. Vertical
@@
   \includegraphics{wang2015-posterior}
-  \caption{
-      Posterior distribution of BA model parameters for
+  \caption[
+      Approximate marginal posterior distributions of BA model parameters for
+      \textcite{wang2015targeting} data.
+  ]{
+      Approximate marginal posterior distributions of BA model parameters for
       \textcite{wang2015targeting} data. Vertical lines indicate maximum
@@
         cat(paste0("\\includegraphics{", fn, "}\n"))
-        cat(sprintf("\\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\\alpha$ = %.1f, $I$ = %d, $m$ = %d, and $N$ = %d]{\n", alpha, I, m, N))
+        cat(sprintf("\\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\\alpha$ = %.1f, $I$ = %d, $m$ = %d, and $N$ = %d.]{\n", alpha, I, m, N))
         cat(sprintf("Approximate marginal posterior distributions of BA model parameters obtained by applying \textit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = %.1f, $I$ = %d, $m$ = %d, and $N$ = %d.

### 2016-04-28 doc/thesis/discussion.tex
Draft???

@@
 computationally intensive, taking about a day when run on 20 cores in parallel
-with the settings described in the methods section.
+with the settings described in the methods section. Implementing parallization
+using MPI, rather than POSIX threads as we have done here, would allow the
+program to be run over a larger number of cores on multiple CPUs in parallel.
 
@@
 performed nearly as well. This result was intuitive: high \gls{alpha} values
-produce networks with few well-connected superspreader nodes which are involved
-in a large number of transmissions, resulting in a highly unbalanced
-ladder-like tree structure. The \gls{I} parameter, representing the prevalence
-at the time of sampling, was also generally estimable. The dynamics of the
-\gls{SI} model, and the coalescent process, offer a potential explanation for
-this result. In the initial phase of the epidemic, when \gls{I} is small, each
-new transmission results in potentially many new discordant edges, thus
-decreasing the waiting time until the next transmission. Hence, there is an
+produce networks with few well-connected ``superspreader'' nodes which are
+involved in a large number of transmissions, resulting in a highly unbalanced
+ladder-like tree structure (\cref{fig:alphatrees}). The \gls{I} parameter,
+representing the prevalence at the time of sampling, was also generally
+estimable. The dynamics of the \gls{SI} model, and the coalescent
+process~\autocite{kingman1982coalescent}, offer a potential explanation for
+this result. \Cref{fig:sir} shows simulated trajectories of the \gls{SIR}
+model. In the initial phase of the epidemic, when \gls{I} is small, each new
+transmission results in potentially many new discordant edges, thus decreasing
+the waiting time until the next transmission. Hence, there is an
 early exponential growth phase, producing many short branches near the root of
@@
 discordant edges decays, causing longer waiting times. The distribution of
-coalescent times in the tree should therefore be informative about \gls{I}.
-This information is captured by the tree kernel, and also by the \gls{nltt}
-statistic, which both performed quite will in classifying \gls{I}
-(\cref{fig:rsquared}).
+coalescence times in the tree should therefore be informative about
+\gls{I}~\autocite{volz2009phylodynamics}. This information is captured by the
+tree kernel, and also by the \gls{nltt} statistic, which both performed quite
+will in classifying \gls{I} (\cref{fig:rsquared}).
 
@@
 attachment power $\alpha$ and the power law exponent $\gamma$
-(\cref{fig:gamma}). Thecrefore, placing a uniform prior on $\alpha$ between 0
+(\cref{fig:gamma}). Therefore, placing a uniform prior on $\alpha$ between 0
 and 2 is equivalent to placing an informative prior that $\gamma$ is close to
-2. Thecrefore, if we were primarily interested in $\gamma$ rather than
+2. Therefore, if we were primarily interested in $\gamma$ rather than
 $\alpha$, a more sensible choice of prior might have a shape informed by
@@
 most likely misspecified for these data. In particular, the average degree of a
-node in the network is equal to $2m$, and thecrefore is constrained to be a
+node in the network is equal to $2m$, and therefore is constrained to be a
 multiple of 2. Furthermore, we considered the case $m = 1$, where the network
-has no cycles, to be implausible and thecrefore assigned it zero prior
+has no cycles, to be implausible and therefore assigned it zero prior
 probability in one set of analyses. This forced the average degree to be at
@@
 
-With respect to the preferential attachment power $\alpha$, the five datasets
-analysed fell into three categories (\cref{fig:abchpd}). First, we
-estimated a preferential attachment power close to 1, indicating linear
-preferential attachment, for the outbreaks studied by
+With respect to the preferential attachment power $\alpha$, the six datasets
+analysed fell into two categories (\cref{fig:abchpd}). First, we estimated a
+preferential attachment power close to 1, indicating linear preferential
+attachment, for the BC data and the outbreaks studied by
 \textcite{niculescu2015recent} and \textcite{wang2015targeting}. These values
-were robust to specifying different priors for $m$. Both studies were of
-populations in which we would expect a high degree of epidemiological
-relatedness: \textcite{niculescu2015recent} studied a recent outbreak among
-Romanian injection drug users (IDU), while \citeauthor{wang2015targeting}
-sampled acutely infected MSM in Beijing, China. Both these are contexts in
-which we would expect some of the assumptions of the BA model, such as a
-connected network, relatively high mean degree, and preferential attachment
-dynamics, to hold.
+were robust to specifying different priors for $m$. All three datasets were
+sampled from populations in which we would expect a high degree of
+epidemiological relatedness: \textcite{niculescu2015recent} studied a recent
+outbreak among Romanian \gls{IDU}, \citeauthor{wang2015targeting} sampled
+acutely infected MSM in Beijing, China, and the BC data constituted a
+phylogenetic \gls{IDU} cluster. These are all contexts in which we would expect
+some of the assumptions of the BA model, such as a connected network,
+relatively high mean degree, and preferential attachment dynamics, to hold.
 
@@
 the \textcite{li2015hiv} data, as the subset we analyzed formed a phylogenetic
-cluster and thecrefore was a good candidate for the BA model. However, nearly
+cluster and therefore was a good candidate for the BA model. However, nearly
 all the posterior density was assigned to $m = 1$ when this value was allowed,

### 2016-04-28 doc/thesis/intro.tex
Draft???

@@
 Thus, transmission, which is represented as branching in the transmission tree,
-causes branching in the viral phylogeny as well. Similarly, the removal of an
-individual from the transmission tree causes the extinction of their viral
-lineage in the phylogeny. Consequently, the topology of the viral phylogeny is
-sometimes used as a proxy for the topology of the transmission
+causes branching in the viral phylogeny as
+well~\autocite{volz2009phylodynamics}. Similarly, the removal of an individual
+from the transmission tree causes the extinction of their viral lineage in the
+phylogeny. Consequently, the topology of the viral phylogeny is sometimes used
+as a proxy for the topology of the transmission
 tree~\autocite{hall2015epidemic}. Modern likelihood-based methods of
@@
 schneeberger2004scale, colgate1989risk}. Interestingly, many other types of
-network, including computer networks, biological neural networks, metabolic
-networks~\autocite{jeong2000large}, and academic co-author networks, also have
-the scale-free property.
+network, including computer networks~\autocite{pastor2001epidemic}, biological
+metabolic networks~\autocite{jeong2000large}, and academic co-author
+networks~\autocite{barabasi2002evolution}, also have the scale-free property.
 

### 2016-04-28 doc/thesis/methods.tex
Draft???

@@
-\subsection{Kernel-ABC method}
+\subsection{\software{Netabc}: a computer program for estimation of contact
+network parameters with kernel-ABC}
 
@@
 kernel-\gls{ABC}. The program combines three major components: Gillespie
-simulation, to simulate transmission trees on contact networks; the
-phylogenetic kernel, to compare simulated to observed transmission trees; and
-adaptive \gls{ABC}-\gls{SMC}, to maintain a population of particles and advance
-it toward the \gls{ABC} target distribution. We give a high-level overview of
-the program here, before describing these three components in detail.
+simulation, to simulate transmission trees on contact networks; the tree
+kernel, to compare simulated to observed transmission trees; and adaptive
+\gls{ABC}-\gls{SMC}, to maintain a population of particles and advance it
+toward the \gls{ABC} target distribution. We give a high-level overview of the
+program here, before describing these three components in detail.
 
@@
 several simulated transmission trees. These trees are compared to the observed
-tree using the phylogenetic kernel. Particles are weighted according to the
-similarity of their associated simulated trees with the true tree, with more
-similar trees receiving higher weights. The particles are iteratively perturbed
-to explore the parameter space, and particles with simulated trees too distant
-from the true tree are periodically dropped and resampled. Once a convergence
-criterion is attained, the final set of particles is used as a Monte Carlo
-approximation to the target distribution of \gls{ABC}, which is assumed to
-resemble the posterior distribution on model parameters (see \cref{sec:abc}). A
-graphical schematic of this algorithm is given in \cref{fig:abcsmc}.
+tree using the tree kernel. Particles are weighted according to the similarity
+of their associated simulated trees with the true tree, with more similar trees
+receiving higher weights. The particles are iteratively perturbed to explore
+the parameter space, and particles with simulated trees too distant from the
+true tree are periodically dropped and resampled. Once a convergence criterion
+is attained, the final set of particles is used as a Monte Carlo approximation
+to the target distribution of \gls{ABC}, which is assumed to resemble the
+posterior distribution on model parameters (see \cref{sec:abc}). A graphical
+schematic of this algorithm is given in \cref{fig:abcsmc}.
 
@@
 to perform kernel-\gls{ABC}, we provide three additional stand-alone utilities:
-\software{treekernel}, to calculate the phylogenetic kernel;
-\software{nettree}, to simulate a transmission tree over a contact network; and
-\software{treestat}, to compute various summary statistics of phylogenies. The
-programs are freely available at \url{https://github.com/rmcclosk/netabc}.
+\software{treekernel}, to calculate the tree kernel; \software{nettree}, to
+simulate a transmission tree over a contact network; and \software{treestat},
+to compute various summary statistics of phylogenies. The programs are freely
+available at \url{https://github.com/rmcclosk/netabc}.
 
@@
 
-\subsection{Real data experiments}
+\subsection{Application to HIV data}
 

### 2016-04-28 doc/thesis/results.Rnw
Appendix
Draft???

@@
 
-    pp <- function (p) {
+    pp <- function (p, eq=FALSE) {
         if (p < 1e-5) {
@@
         } else {
-            latexSN(round(p, -floor(log10(p))))
+            n <- latexSN(round(p, -floor(log10(p))))
+            n <- sub("^1\\\\!\\\\times\\\\!", "", n)
+            if (eq) {
+                paste0("=", n)
+            } else {
+                n
+            }
         }
@@
     (average $R^2$ 
-     \Sexpr{kernel.m[ntip == 100,round(mean(rsquared, 2))]},
-     \Sexpr{kernel.m[ntip == 500,round(mean(rsquared, 2))]},
-     \Sexpr{kernel.m[ntip == 1000,round(mean(rsquared, 2))]}
+     \Sexpr{kernel.m[ntip == 100,round(mean(rsquared), 2)]},
+     \Sexpr{kernel.m[ntip == 500,round(mean(rsquared), 2)]},
+     \Sexpr{kernel.m[ntip == 1000,round(mean(rsquared), 2)]}
      for 100, 500, and 1000 tips respectively).
@@
     $p\ \Sexpr{value.cor["N", pp(p.value)]}$). 
+Most of the kernel score distributions were extremely flat, except for the
+value $N = 1000$ which exhibited a pronounced peak around the true value
+(\cref{fig:gridN}).
 
@@
 2. The results for the other values of \gls{m} were similar
-(\cref{fig:abcm3,fig:abcm4}). The median [IQR] absolute error of estimates of
+(\cref{fig:abcptm3,fig:abcptm4}). The median [IQR] absolute error of estimates of
 \gls{alpha} across all simulations was
@@
     \Sexpr{as.integer(m.tbl[1] * 100)} \%
-of simulations. The true values of $m$ and $I$ did not significantly affect the
-error (one-way ANOVA), but the accuracy was significantly lower for integral
-than non-integral values of \gls{alpha}
-    (Wilcoxon rank-sum test, $p = \Sexpr{pp(alpha.m.test$p.value)}$).
-Finally, the total number of nodes \gls{N} was consistently over-estimated by
-about a factor of two
+of simulations, little better than random guessing. The true values of $m$ and
+$I$ did not significantly affect the error (one-way ANOVA). Finally, the total
+number of nodes \gls{N} was consistently over-estimated by about a factor of
+two
     (error \Sexpr{d[,format(median(N_error), scientific=FALSE)]} 
@@
 
-\Cref{fig:abcex} shows the \gls{ABC} approximation to the posterior
-distribution on the \gls{BA} parameters for one simulation. Equivalent plots
-for one replicate simulation with each combination of parameters can be found
-in
-<<posterior_sims, echo=FALSE, results="asis">>=
+<<posterior_sims, echo=FALSE>>=
     N <- 5000
@@
     for (I in c(1000, 2000)) {
-    for (alpha in c(0, 0.5, 1, 1.5, 2)) {
-        lab <- c(lab, sprintf("fig:%2f-%d-%d-%d-%d", alpha, I, m, N, replicate))
+    for (alpha in c(0, 0.5, 1, 1.5)) {
+        lab <- c(lab, sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate))
     }
@@
     }
-    cat(paste0("\\cref{", paste(lab, sep=","), "}."))
+    lab <- paste(lab, collapse=",")
 @
-\Gls{HPD} intervals around \gls{alpha} and \gls{I} were narrow relative to the
-region of nonzero prior density, whereas the intervals for $m$ and \gls{N} were
-widely dispersed. \Cref{tab:abchpd} shows point estimates and 95\% \gls{HPD}
-intervals averaged over all simulations.
+
+\Cref{fig:abcex} shows the \gls{ABC} approximation to the posterior
+distribution on the \gls{BA} parameters for one simulation. Equivalent plots
+for one replicate simulation with each combination of parameters can be found
+in \cref{\Sexpr{lab}}. \Gls{HPD} intervals around \gls{alpha} and \gls{I} were
+narrow relative to the region of nonzero prior density, whereas the intervals
+for $m$ and \gls{N} were widely dispersed. \Cref{tab:abchpd} shows point
+estimates and 95\% \gls{HPD} intervals averaged over all simulations.
 
@@
   \includegraphics{{abc-posterior/1.0_1000_2_5000_0}.pdf}
-  \caption{
-    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by         
-    running \software{netabc} on simulated data. Values shown are for               
-    simulations with \gls{m} = 3. Dashed lines indicate true values. (A)            
-    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations  
-    against known values. (B) Estimates of \gls{m} and \gls{N} which were held   
-    fixed in these simulations at the values \gls{m} = 3 and \gls{N} = 5000. 
+  \caption[Approximate marginal posterior distributions of BA model parameters
+      obtained by applying \textit{netabc} to a simulated transmission tree
+      with values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]
+    {
+        Approximate marginal posterior distributions of BA model parameters
+        obtained by applying \textit{netabc} to a simulated transmission tree
+        with BA parameter values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ =
+        5000. Vertical dashed lines indicate true values. Shaded areas are 95\%
+        highest posterior density intervals. $x$-axes indicate regions of
+        nonzero prior density.
     }
-  \label{fig:abcptm2}
+  \label{fig:abcex}
 \end{figure}
@@
     \input{\tablepath/abc-hpd}
-    \caption{Average widths of 95\% confidence intervals for \gls{BA} model
-    parameters estimated with kernel-\gls{ABC}.}
-    \label{tab:glm}
+    \caption[
+        Maximum \textit{a priori} estimates and 95\% highest posterior density
+        (HPD) interval boundaries for \gls{BA} model parameters estimated with
+        \software{netabc}, averaged over simulated transmission trees.
+    ]
+    {
+        Maximum \textit{a priori} estimates and 95\% highest posterior density
+        (HPD) interval boundaries for \gls{BA} model parameters estimated with
+        \software{netabc}, averaged over simulated transmission trees.
+    }
+    \label{tab:abchpd}
 \end{table}
 
-\subsubsection*{Effect of parameters on power-law exponent}
+<<mixed_peerdriven, include=FALSE>>=
+  library(coda)
+  d <- fread("bzcat ../../simulations/abc-pa-mixed-alpha/abc/*")
+  d <- d[iter == max(iter)]
+  d <- d[sample(1:nrow(d), prob=weight, replace=TRUE)]
+  params <- c("alpha", "m", "N", "I")
+  d <- melt(d, measure.vars=params, variable.name="parameter")
+  f <- function (x) {
+    dens <- density(x)
+    hpd <- HPDinterval(mcmc(x))
+    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
+         hpd.max=hpd[,"upper"])
+  }
+  d <- d[,f(value), by=parameter]
+  setkey(d, parameter)
 
-\Cref{tab:glm} shows the estimated parameters for a log-link \gls{GLM} fitted
-to the observed distribution of \gls{gamma} values. The coefficients are
-interpretable as multiplicative effects.
+  m <- collect.metadata("../../simulations/abc-pa-peerdriven/abc/*")
+  p <- fread(paste("bzcat", rownames(m)[m$sample_peer == 1]))
+  p <- p[iter == max(iter)]
+  p <- p[sample(1:nrow(p), prob=weight, replace=TRUE)]
+  params <- c("alpha", "m", "N", "I")
+  p <- melt(p, measure.vars=params, variable.name="parameter")
+  f <- function (x) {
+    dens <- density(x)
+    hpd <- HPDinterval(mcmc(x))
+    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
+         hpd.max=hpd[,"upper"])
+  }
+  p <- p[,f(value), by=parameter]
+  setkey(p, parameter)
+  options(scipen=5)
+@
 
-\begin{table}
-    \centering
-    \input{\tablepath/pa-gamma-glm}
-    \caption{Estimated \gls{GLM} parameters for relationship between power-law
-    exponent \gls{gamma} and \gls{BA} model parameters.}
-    \label{tab:glm}
-\end{table}
+To test the effect of model misspecification, we simulated one network where
+the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
+other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
+HPD] estimates for each parameter were: 
+$\alpha$, 
+  \Sexpr{round(d["alpha", map], 2)} 
+  [\Sexpr{round(d["alpha", hpd.min], 2)} -
+   \Sexpr{round(d["alpha", hpd.max], 2)}];
+$I$,
+  \Sexpr{d["I", round(map)]} 
+  [\Sexpr{d["I", round(hpd.min)]} -
+   \Sexpr{d["I", round(hpd.max)]}];
+$m$,
+  \Sexpr{d["m", floor(map)]} 
+  [\Sexpr{d["m", floor(hpd.min)]} -
+   \Sexpr{d["m", floor(hpd.max)]}];
+$N$,
+  \Sexpr{d["N", round(map)]} 
+  [\Sexpr{d["N", round(hpd.min)]} -
+   \Sexpr{d["N", round(hpd.max)]}].
+The approximate posterior distributions for this simulation are shown in
+\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
+transmission tree in a peer-driven fashion, where the probability to sample a
+node was twice as high if one of its peers had already been sampled. The
+parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
+$I$ = 2000. The estimated values were
+$\alpha$, 
+  \Sexpr{p["alpha", round(map, 2)]} 
+  [\Sexpr{p["alpha", round(hpd.min, 2)]} -
+   \Sexpr{p["alpha", round(hpd.max, 2)]}];
+$I$,
+  \Sexpr{p["I", floor(map)]} 
+  [\Sexpr{p["I", floor(hpd.min)]} -
+   \Sexpr{p["I", floor(hpd.max)]}];
+$m$,
+  \Sexpr{p["m", floor(map)]} 
+  [\Sexpr{p["m", floor(hpd.min)]} -
+   \Sexpr{p["m", floor(hpd.max)]}];
+$N$,
+  \Sexpr{p["N", floor(map)]} 
+  [\Sexpr{p["N", floor(hpd.min)]} -
+   \Sexpr{p["N", floor(hpd.max)]}].
+The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
+of these results were in line with estimates obtained on other simulated
+datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
+$\alpha$ was somewhat lower than typical.
+
+%\subsubsection*{Effect of parameters on power-law exponent}
+%
+%\Cref{tab:glm} shows the estimated parameters for a log-link \gls{GLM} fitted
+%to the observed distribution of \gls{gamma} values. The coefficients are
+%interpretable as multiplicative effects.
+%
+%\begin{table}
+%    \centering
+%    \input{\tablepath/pa-gamma-glm}
+%    \caption{Estimated \gls{GLM} parameters for relationship between power-law
+%    exponent \gls{gamma} and \gls{BA} model parameters.}
+%    \label{tab:glm}
+%\end{table}
 
@@
 \cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.
-Two of the datasets~\autocite{niculescu2015recent, wang2015targeting} had
-estimated $\alpha$ values near unity for the prior allowing $m = 1$ (MAP
-estimates [95\% HPD] 
+Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had
+estimated $\alpha$ values near unity for the prior allowing $m = 1$ (\gls{MAP}
+estimates [95\% \gls{HPD}] 
   \Sexpr{map1["niculescu2015", round(alpha, 2)]} 
@@
     \Sexpr{up1["novitsky2014", round(alpha, 2)]}] for $m \geq 1$ vs.
-    \Sexpr{low2["novitsky2014", round(alpha, 2)]} -
-    \Sexpr{up2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).
+   [\Sexpr{low2["novitsky2014", round(alpha, 2)]} -
+    \Sexpr{up2["novitsky2014", round(alpha, 2)]}] for $m \geq 2$).
 
@@
   \centering
-  %\includegraphics{realdata-hpd}
-  \vspace{8pt}
-  \caption{
+  \includegraphics{realdata-hpd-bc}
+  \caption[
       Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
-      parameters of the BA network model, fitted to five published HIV datasets
-      with kernel-ABC.
+      parameters of the BA network model, fitted to six HIV datasets
+      with \software{netabc}.]
+  {
+      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
+      parameters of the BA network model, fitted to six HIV datasets with
+      \software{netabc}. Legend labels indicate risk group and country of
+      origin. Abbreviations: IDU, injection drug users; MSM, men who have sex
+      with men; HET, heterosexual.
   }

### 2016-04-28 doc/thesis/results.tex
Draft???

@@
     (Spearman's $\rho$ = \ensuremath{-0.26},
-     $p = 1\!\times\!10^{-5}$),
+     $p = 10^{-5}$),
 although the relationship was clearly nonlinear (\cref{fig:gridest,fig:gridalpha}).
@@
     27 \%
-of simulations. The true values of $m$ and $I$ did not significantly affect the
-error (one-way ANOVA), but the accuracy was significantly lower for integral
-than non-integral values of \gls{alpha}
-    (Wilcoxon rank-sum test, $p = 0.3$).
-Finally, the total number of nodes \gls{N} was consistently over-estimated by
-about a factor of two
+of simulations, little better than random guessing. The true values of $m$ and
+$I$ did not significantly affect the error (one-way ANOVA). Finally, the total
+number of nodes \gls{N} was consistently over-estimated by about a factor of
+two
     (error 5987 
@@
       obtained by applying \textit{netabc} to a simulated transmission tree
-      with values $alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]
+      with values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]
     {
@@
         obtained by applying \textit{netabc} to a simulated transmission tree
-        with BA parameter values $alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ =
+        with BA parameter values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ =
         5000. Vertical dashed lines indicate true values. Shaded areas are 95\%
@@
     \input{\tablepath/abc-hpd}
-    \caption{Average widths of 95\% confidence intervals for \gls{BA} model
-    parameters estimated with kernel-\gls{ABC}.}
+    \caption[
+        Maximum \textit{a priori} estimates and 95\% highest posterior density
+        (HPD) interval boundaries for \gls{BA} model parameters estimated with
+        \software{netabc}, averaged over simulated transmission trees.
+    ]
+    {
+        Maximum \textit{a priori} estimates and 95\% highest posterior density
+        (HPD) interval boundaries for \gls{BA} model parameters estimated with
+        \software{netabc}, averaged over simulated transmission trees.
+    }
     \label{tab:abchpd}
@@
 $\alpha$, 
-  1.09 
+  1.1 
   [0.6 -
@@
 $I$,
-  1103 
+  1091 
   [662 -
@@
   3 
-  [2 -
+  [1 -
    5];
 $N$,
-  12582 
-  [3590-
-   14977].
+  12566 
+  [3733 -
+   14945].
 The approximate posterior distributions for this simulation are shown in
@@
 $\alpha$, 
-  0.39 
+  0.37 
   [0 -
@@
 $I$,
-  2360 
-  [1379 -
+  2407 
+  [1423 -
    3811];
 $m$,
-  2 
+  3 
   [2 -
@@
 $N$,
-  13537 
-  [2855 -
-   14649].
+  13811 
+  [2624 -
+   14647].
 The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
@@
 
-\subsubsection*{Effect of parameters on power-law exponent}
-
-\Cref{tab:glm} shows the estimated parameters for a log-link \gls{GLM} fitted
-to the observed distribution of \gls{gamma} values. The coefficients are
-interpretable as multiplicative effects.
-
-\begin{table}
-    \centering
-    \input{\tablepath/pa-gamma-glm}
-    \caption{Estimated \gls{GLM} parameters for relationship between power-law
-    exponent \gls{gamma} and \gls{BA} model parameters.}
-    \label{tab:glm}
-\end{table}
+%\subsubsection*{Effect of parameters on power-law exponent}
+%
+%\Cref{tab:glm} shows the estimated parameters for a log-link \gls{GLM} fitted
+%to the observed distribution of \gls{gamma} values. The coefficients are
+%interpretable as multiplicative effects.
+%
+%\begin{table}
+%    \centering
+%    \input{\tablepath/pa-gamma-glm}
+%    \caption{Estimated \gls{GLM} parameters for relationship between power-law
+%    exponent \gls{gamma} and \gls{BA} model parameters.}
+%    \label{tab:glm}
+%\end{table}
 
@@
 \cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.
-Two of the datasets~\autocite{niculescu2015recent, wang2015targeting} had
-estimated $\alpha$ values near unity for the prior allowing $m = 1$ (MAP
-estimates [95\% HPD] 
+Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had
+estimated $\alpha$ values near unity for the prior allowing $m = 1$ (\gls{MAP}
+estimates [95\% \gls{HPD}] 
   1.05 
@@
     1.62] for $m \geq 1$ vs.
-    0 -
-    0.73 for $m \geq 2$).
+   [0 -
+    0.73] for $m \geq 2$).
 
@@
   \includegraphics{realdata-hpd-bc}
-  \caption{
+  \caption[
       Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
-      parameters of the BA network model, fitted to five published HIV datasets
-      with kernel-ABC.
+      parameters of the BA network model, fitted to six HIV datasets
+      with \software{netabc}.]
+  {
+      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
+      parameters of the BA network model, fitted to six HIV datasets with
+      \software{netabc}. Legend labels indicate risk group and country of
+      origin. Abbreviations: IDU, injection drug users; MSM, men who have sex
+      with men; HET, heterosexual.
   }

### 2016-04-28 doc/thesis/figures/acc_prec.R

@@ -1,36 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(ggplot2)
-library(data.table)
-library(extrafont)
-
-set.seed(0)
-
-dens1 <- density(rnorm(400))
-dens2 <- density(runif(400, -3, 3))
-dens3 <- density(rnorm(400))
-dens4 <- density(runif(400, -3, 3))
-
-d <- data.table(x=c(dens1$x, dens2$x, dens3$x, dens4$x), 
-                y=c(dens1$y, dens2$y, dens3$y, dens4$y), 
-                precise=rep(c("yes", "no", "yes", "no"), each=length(dens1$x)),
-                accurate=rep(c("yes", "no"), each=length(dens1$x)*2))
-d[,precise := factor(precise, levels=c("yes", "no"))]
-d[,accurate := factor(accurate, levels=c("yes", "no"))]
-d[,col := ifelse(precise == "yes", ifelse(accurate == "yes", "green", "red"), "orange")]
-
-lines <- data.frame(
-    v=c(0, dens2$x[which.max(dens2$y)], 2, 1),
-    precise=c("yes", "no", "yes", "no"),
-    accurate=c("yes", "yes", "no", "no")
-)
-
-pdf("acc_prec.pdf", height=4, family="Gillius ADF")
-ggplot(d, aes(x=x, y=y, color=col)) + geom_line() + theme_bw() + xlim(-4, 4) +
-    geom_vline(data=lines, aes(xintercept=v), lty=2) +
-    facet_grid(precise~accurate, scales="free", labeller="label_both") +
-    theme(axis.ticks=element_blank(), legend.position="none",
-          panel.grid=element_blank(), axis.text=element_blank()) +
-    scale_color_manual(values=c("green", "orange", "red")) +
-    labs(x="testing value", y="kernel score")
-dev.off()

### 2016-04-28 doc/thesis/figures/gridsearch_expt.R

@@ -1,30 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(igraph)
-library(ape)
-library(extrafont)
-
-set.seed(0)
-ninf <- 4
-col <- c(rgb(0, 0, 1), rgb(0.4, 0.4, 1), rgb(0.8, 0.8, 1))
-
-for (i in 1:3) {
-    png(sprintf("testtree%d.png", i), bg="transparent", width=100, height=100)
-    par(mar=c(0, 0, 0, 0) + 0.1)
-    t <- rcoal(ninf)
-    plot(t, direction="down", edge.width=5, edge.color=col[i], show.tip.label=FALSE)
-    dev.off()
-}
-
-dens <- density(rnorm(200))
-x <- dens$x
-y <- dens$y
-gradient <- colorRampPalette(c("blue", "white"))(length(x))
-true.x <- x[length(x)*0.4]
-true.x.col <- gradient[length(x)*0.4]
-pdf("tinykscore.pdf", family="Gillius ADF", width=3, height=0.8)
-par(mar=c(1, 1, 0, 0) + 0.1, mgp=c(0, 1, 0), cex=1.25)
-plot(x, y, type="p", xlim=c(-4, 4), main=NA, pch=16,
-     xlab="training value", ylab="score", col=gradient, xaxt="n", yaxt="n")
-abline(v=true.x, col=true.x.col)
-dev.off()

### 2016-04-28 doc/thesis/figures/kernel_expt.R

@@ -1,36 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(igraph)
-library(ape)
-
-nnode <- 7
-ninf <- 4
-set.seed(0)
-
-pow <- c(0.5, 1, 1.5)
-col <- c("red", "blue", "green")
-inf.col <- "black"
-
-for (i in 1:3) {
-    g <- sample_pa(nnode, m=2, power=pow[i], directed=FALSE)
-    png(sprintf("tinynet%d.png", i), bg="transparent", width=100, height=100)
-    par(mar=c(0, 0, 0, 0) + 0.1)
-    plot(g, vertex.size=10, vertex.label=NA, vertex.color=col[i],
-         vertex.frame.color=col[i], edge.color=col[i], edge.width=5)
-    dev.off()
-
-    png(sprintf("tinyepi%d.png", i), bg="transparent", width=100, height=100)
-    par(mar=c(0, 0, 0, 0) + 0.1)
-    ecol <- rep(col[i], ecount(g))
-    vcol <- rep(c(inf.col, col[i]), c(ninf, vcount(g)-ninf))
-    ecol[sample(unique(unlist(g[[1:ninf,1:ninf,edges=TRUE]])), ninf-1)] <- inf.col
-    plot(g, vertex.size=10, vertex.label=NA, vertex.color=vcol,
-         vertex.frame.color=vcol, edge.color=ecol, edge.width=5)
-    dev.off()
-
-    png(sprintf("tinytree%d.png", i), bg="transparent", width=100, height=100)
-    par(mar=c(0, 0, 0, 0) + 0.1)
-    t <- rcoal(ninf)
-    plot(t, direction="down", edge.width=5, edge.color=col[i], show.tip.label=FALSE)
-    dev.off()
-}

### 2016-04-28 doc/thesis/figures/nltt.R

@@ -1,31 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(ape)
-library(extrafont)
-
-t1 <- read.tree(text="(((:1,:1):1,:1):1,:1):0;")
-t2 <- read.tree(text="((((:1,:1):1,:1):1,:1):1,:1):0;")
-
-sf1 <- stepfun(c(0, 0.5), c(0, 0.5, 1))
-sf2 <- stepfun(c(0, 1/3, 2/3), c(0, 1/3, 2/3, 1))
-
-af1 <- approxfun(c(0, 0.5, 1), c(0, 0.5, 1))
-af2 <- approxfun(c(0, 1/3, 2/3, 1), c(0, 1/3, 2/3, 1))
-
-pdf("nltt.pdf", family="Gillius ADF")
-par(mfrow=c(2, 2), lwd=3)
-old.mar <- par("mar")
-par(mar=c(0, 2, 4, 2) + 0.1)
-plot(t1, edge.width=3, edge.color="blue", show.tip.label=FALSE, direction="down")
-plot(t2, edge.width=3, edge.color="red", show.tip.label=FALSE, direction="down")
-
-par(mar=old.mar)
-plot(sf1, col="blue", main="original nLTT", xlab="normalized lineages",
-     ylab="normalized time", do.points=FALSE, xlim=c(0, 1), xaxs="i")
-lines(sf2, col="red", do.points=FALSE)
-
-plot(af1, col="blue", main="updated nLTT", xlab="normalized lineages",
-     ylab="normalized time", xlim=c(0, 1), xaxs="i")
-x <- seq(0, 1, 0.1)
-lines(x, af2(x) + 0.05, col="red")
-dev.off()

### 2016-04-28 doc/thesis/figures/separable.R

@@ -1,51 +1,0 @@
-#!/usr/bin/env Rscript
-
-library(ape)
-library(png)
-library(raster)
-library(extrafont)
-
-set.seed(0)
-
-tree.col <- c("black", rep(c("blue", "red"), each=3))
-tmp <- tempfile()
-tinytree <- as.list(rep(0, 7))
-for (i in 1:7) {
-    png(tmp, width=50, height=50, bg="transparent")
-    plot(rtree(4), show.tip.label=F, direction="down", no.margin=TRUE, edge.width=5,
-         edge.color=tree.col[i])
-    dev.off()
-    tinytree[[i]] <- readPNG(tmp)
-}
-unlink(tmp)
-
-pdf("separable.pdf", width=6, height=2, family="Gillius ADF")
-par(mar=c(0, 0, 0, 0) + 0.1, mfrow=c(1, 3))
-
-plot(NA, xlim=c(0, 1), ylim=c(0, 1), axes=FALSE, xlab=NA, ylab=NA, frame.plot=TRUE)
-x <- c(0.2, 0.3, 0.1, 0.55, 0.75, 0.65)
-y <- c(0.65, 0.75, 0.7, 0.4, 0.45, 0.35)
-rasterImage(tinytree[[1]],0.5,0.5,0.6,0.6)
-mapply(rasterImage, tinytree[2:7], as.list(x), as.list(y), as.list(x+0.1), as.list(y+0.1))
-abline(a=0.1, b=1, lty=2)
-text(0.05, 0.95, "A", cex=2)
-
-plot(NA, xlim=c(0, 1), ylim=c(0, 1), axes=FALSE, xlab=NA, ylab=NA, frame.plot=TRUE)
-x <- (c(1, 0.5, -0.5, -1, -0.5, 0.5) + 1) * 0.45
-y <- (c(0, sqrt(3)/2, sqrt(3)/2, 0, -sqrt(3)/2, -sqrt(3)/2, 0) + 1) * 0.45
-x <- x[c(1, 3, 5, 2, 4, 6)]
-y <- y[c(1, 3, 5, 2, 4, 6)]
-rasterImage(tinytree[[1]],0.5,0.5,0.6,0.6)
-mapply(rasterImage, tinytree[2:7], as.list(x), as.list(y), as.list(x+0.1), as.list(y+0.1))
[…]

### 2016-04-28 doc/thesis/conclusion.tex
Added gamma things

@@
 networks could be valuable from a public health perspective, such as for
-investigating optimal vaccination strategies. This information could assist in
-curtailing current epimemics, as well as preventing future epidemics of
-different diseases over the same contact network.
+investigating optimal vaccination strategies~\autocite{keeling2005networks,
+peng2013vaccination, ma2013importance, rushmore2014network}. This information
+could assist in curtailing current epimemics, as well as preventing future
+epidemics of different diseases over the same contact network.
+
+The paricular model we have investigated uses a preferential attachment
+mechanism to generate scale-free networks resembling real-world social and
+sexual networks~\autocite{liljeros2001web, schneeberger2004scale,
+colgate1989risk}. Of the four parameters we considered, the pereferential
+attachment power $\alpha$ was the most readily estimable. Estimating $\alpha$
+with traditional epidemiological methods is challenging due to the requirement
+of sampling the high-degree nodes making up the tail of the power law
+distribution, although approaches such as respondent-driven
+sampling~\autocite{heckathorn1997respondent} may be effective.
+
+In closing, \software{netabc} combines phylodynamics, contact network
+epidemiology, approximate Bayesian computation, and sequentiial Monte Carlo 
+to provide a source of insight into network structures complementary to
+traditional epidemiology.

### 2016-04-28 doc/thesis/results.Rnw
Added gamma things

@@
 \end{figure*}
+
+To make our analyses comparable to the existing network literature, we
+estimated values of the power law exponent $\gamma$ for each of the datasets 
+investigated. The results are summarized in \cref{tab:gamma}. All of the
+estimated exponents were in the range $2 \leq \gamma \leq 4$ commonly reported
+in the literature~\autocite{liljeros2001web, schneeberger2004scale,
+colgate1989risk, brown2011transmission}. Higher estimated values of $\alpha$
+translated into higher estimated values of $\gamma$, which is consistent with
+the observed relationship between these two quantities (\cref{fig:gamma}).
+
+\begin{table}
+    \caption[
+        Estimated power law exponents for six HIV datasets based on maximum
+        \textit{a priori} estimates of BA model parameters.
+    ]{
+        Estimated power law exponents for six HIV datasets based on maximum
+        \textit{a priori} estimates of BA model parameters. 100 networks were
+        simulated using \textit{MAP} parameter estimates obtained with
+        \software{netabc}. The power law exponent $\gamma$ was estimated for
+        each, and the median of those estimates was used as a point estimate
+        for the corresponding dataset.
+    }
+    \label{tab:gamma}
+\end{table}

### 2016-05-02 doc/thesis/abstract.tex

--- 2016-04-05 doc/paper/abstract.tex
@@
 substantial heterogeneity in the parameter estimates on real datasets, with
-point estimates for the preferential attachment power ranging from 0.2 to 1.1.
+point estimates for the preferential attachment power ranging from 0.06 to 1.05.
 These results underscore the importance of considering contact structures when
 performing phylodynamic inference. Our method offers the potential to
-quantitatively investigate the network structure underlying viral epidemics,
-using models appropriate to the context being studied.
+quantitatively investigate the contact network structure underlying viral
+epidemics.

### 2016-05-02 doc/thesis/appendix.tex

--- 2016-04-19 doc/thesis/appendix.tex
@@
 
-\begin{figure}
+\begin{figure}[H]
   \centering
   \includegraphics{kernel-alpha-crossv.pdf}
-  \caption{
+  \caption[Cross validation accuracy of classifiers for \gls{BA} model parameter
+    \gls{alpha} for eight epidemic scenarios.]
+  {
     Cross validation accuracy of classifiers for \gls{BA} model parameter
@@
 
-\begin{figure}
+\begin{figure}[H]
   \centering
   \includegraphics{kernel-I-crossv.pdf}
-  \caption{
+  \caption[Cross validation accuracy of classifiers for \gls{BA} model parameter
+    \gls{I} for eight epidemic scenarios.]
+  {
     Cross validation accuracy of classifiers for \gls{BA} model parameter
@@
 
-\begin{figure}
+\begin{figure}[H]
   \centering
   \includegraphics{kernel-m-crossv.pdf}
-  \caption{
+  \caption[Cross validation accuracy of classifiers for \gls{BA} model parameter
+    \gls{m} for eight epidemic scenarios.]
+  {
     Cross validation accuracy of classifiers for \gls{BA} model parameter
@@
 
-\begin{figure}
+\begin{figure}[H]
   \centering
   \includegraphics{kernel-N-crossv.pdf}
-  \caption{
+  \caption[Cross validation accuracy of classifiers for \gls{BA} model parameter
+    \gls{N} for eight epidemic scenarios.]
+  {
     Cross validation accuracy of classifiers for \gls{BA} model parameter
@@
 
-\begin{figure}
+\begin{figure}[H]
+  \centering
+  \includegraphics{gridsearch-alpha-point-estimate.pdf}
+  \caption[
+      Point estimates of preferential attachment power \gls{alpha} of
+      \acrlong{BA} network model, obtained on simulated trees with
+      kernel-score-based grid search.]
+  {
+      Point estimates of preferential attachment power \gls{alpha} of
+      \acrlong{BA} network model, obtained on simulated trees with
+      kernel-score-based grid search. Test trees were simulated according to
+      several values of \gls{alpha} ($x$-axis) with other model parameters
+      fixed at \gls{m} = 2, \gls{N} = 5000, and \gls{I} = 1000. The test trees
+      were compared to trees simulated along a narrowly spaced grid of
+      \gls{alpha} values using the tree kernel, with the same values of the
+      other parameters. The grid value with the highest median kernel score was
+      taken as a point estimate for \gls{alpha} ($y$-axis).
+  }
+  \label{fig:gridptalpha}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics{gridsearch-I-point-estimate.pdf}
+  \caption[ Point estimates of prevalence at time of sampling \gls{I} of
+      \acrlong{BA} network model, obtained on simulated trees with
+      kernel-score-based grid search. ]
+  {
+      Point estimates of prevalence at time of sampling \gls{I} of \acrlong{BA}
+      network model, obtained on simulated trees with kernel-score-based grid
+      search. Test trees were simulated according to several values of \gls{I}
+      ($x$-axis) with other model parameters fixed at \gls{alpha} = 1, \gls{m}
+      = 2, and \gls{N} = 5000. The test trees were compared to trees simulated
+      along a narrowly spaced grid of \gls{I} values using the tree kernel,
+      with the same values of the other parameters. The grid value with the
+      highest median kernel score was taken as a point estimate for \gls{I}
+      ($y$-axis).
+  }
+  \label{fig:gridptI}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics{gridsearch-m-point-estimate.pdf}
+  \caption[
+      Point estimates of number of edges per vertex \gls{m} of \acrlong{BA}
+      network model, obtained on simulated trees with kernel-score-based grid
+      search.
+  ]
+  {
+      Point estimates of number of edges per vertex \gls{m} of \acrlong{BA}
+      network model, obtained on simulated trees with kernel-score-based grid
+      search. Test trees were simulated according to several values of \gls{m}
+      ($x$-axis) with other model parameters fixed at \gls{alpha} = 1, \gls{I}
+      = 1000, and \gls{N} = 5000. The test trees were compared to trees
+      simulated along a narrowly spaced grid of \gls{m} values using the tree
+      kernel, with the same values of the other parameters. The grid value with
+      the highest median kernel score was taken as a point estimate for \gls{m}
+      ($y$-axis).
+  }
+  \label{fig:gridptm}
+\end{figure}
+
+\begin{figure}[H]
+  \centering
+  \includegraphics{gridsearch-N-point-estimate.pdf}
+  \caption[
+      Point estimates of number of edges per vertex \gls{N} of \acrlong{BA}
+      network model, obtained on simulated trees with kernel-score-based grid
+      search.
+  ]{
+      Point estimates of number of edges per vertex \gls{N} of \acrlong{BA}
+      network model, obtained on simulated trees with kernel-score-based grid
+      search. Test trees were simulated according to several values of \gls{N}
+      ($x$-axis) with other model parameters fixed at \gls{alpha} = 1, \gls{m}
+      = 2, and \gls{I} = 1000. The test trees were compared to trees simulated
+      along a narrowly spaced grid of \gls{N} values using the tree kernel,
+      with the same values of the other parameters. The grid value with the
+      highest median kernel score was taken as a point estimate for \gls{m}
+      ($y$-axis).
+  }
+  \label{fig:gridptN}
+\end{figure}
+
+\begin{figure}[H]
   \centering
   \includegraphics{abc-point-estimate-m3.pdf}
-  \caption{                                              
+  \caption[
     \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
-    running \software{netabc} on simulated data. Values shown are for
-    simulations with \gls{m} = 3. Dashed lines indicate true values. (A)
-    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations
-    against known values. (B) Estimates of \gls{m} and \gls{N} which were held
-    fixed in these simulations at the values \gls{m} = 3 and \gls{N} = 5000.
+    running \software{netabc} on simulated data, for simulations with \gls{m} = 3.
+  ]{
+    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
+    running \software{netabc} on simulated data, for simulations with \gls{m} = 3.
+    Dashed lines indicate true values. (A) Estimates of \gls{alpha} and \gls{I}
+    which were varied in these simulations against known values. (B) Estimates
+    of \gls{m} and \gls{N} which were held fixed in these simulations at the
+    values \gls{m} = 3 and \gls{N} = 5000.
   }        
-  \label{fig:abcm3}
+  \label{fig:abcptm3}
 \end{figure}
 
-\begin{figure}
+\begin{figure}[H]
   \centering
   \includegraphics{abc-point-estimate-m4.pdf}
+  \caption[
+    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
+    running \software{netabc} on simulated data, for simulations with \gls{m} = 3.
+  ]{
+    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
+    running \software{netabc} on simulated data, for simulations with \gls{m} = 4.
+    Dashed lines indicate true values. (A) Estimates of \gls{alpha} and \gls{I}
+    which were varied in these simulations against known values. (B) Estimates
+    of \gls{m} and \gls{N} which were held fixed in these simulations at the
+    values \gls{m} = 4 and \gls{N} = 5000.
+  }        
+  \label{fig:abcptm4}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{alpha-gamma.pdf}
   \caption{
-    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
-    running \software{netabc} on simulated data. Values shown are for
-    simulations with \gls{m} = 4. Dashed lines indicate true values. (A)
-    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations
-    against known values. (B) Estimates of \gls{m} and \gls{N} which were held
-    fixed in these simulations at the values \gls{m} = 4 and \gls{N} = 5000.
+      Relationship between preferential attachment power parameter $\alpha$
+      and power law exponent $\gamma$ for networks simulated under the BA
+      network model with $N$ = 5000 and $m$ = 2.
   }
-  \label{fig:abcm4}
+  \label{fig:gamma}
 \end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{sir-trajectories.pdf}
+  \caption[Average trajectories of epidemics simulated according to the
+  \gls{SIR} model over full and \gls{BA} networks.]
+  {
+      Average trajectories of epidemics simulated according to the \gls{SIR}
+      model over full and \gls{BA} networks. 100 epidemics each were simulated
+      over completely connected networks and networks generated from the
+      \gls{BA} model. The network parameters were $\alpha = 1$ and $m = 2$.
+      The transmission rate along each edge of a network $G$ was taken as
+      $2/|E(G)|$, and the removal rate of each vertex was taken as $1/|V(G)|$.
+      These choices ensured that both networks had the same waiting time until
+      the next transmission at the beginning of the epidemic.
+  }
+  \label{fig:sir}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{mixed-posterior}
+    \caption[Approximate marginal posterior distributions of BA model
+        parameters obtained using kernel-ABC for a network with heterogeneous
+        node behaviour.]
+    {
+        Approximate marginal posterior distributions of Barab\'asi-Albert
+        model parameters obtained using kernel-ABC for a network with
+        heterogeneous node behaviour. Half of the nodes were attached with
+        $\alpha = 0.5$, and the other half with $\alpha = 1.5$ (vertical
+        dashed lines, top left). Other parameter values were $m = 2$, $I =
+        1000$, and $N = 5000$ (vertical dashed lines, other than top left).
+        Shaded areas indicate 95\% highest posterior density intervals.
+    }
+    \label{fig:mixed}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{peerdriven-posterior}
+    \caption[Approximate marginal posterior distributions of Barab\'asi-Albert
+        model parameters obtained using kernel-ABC for a network with
+        peer-driven sampling.]
+    {
+        Approximate marginal posterior distributions of Barab\'asi-Albert
+        model parameters obtained using kernel-ABC for a network with
+        peer-driven sampling. An epidemic was simulated in the usual fashion,
+        but rather than being sampled at random, infected nodes were sampled
+        with a probability two times higher if they had any sampled neighbours
+        in the contact network. Vertical dashed lines indicate true parameter
+        values, and shaded areas indicate 95\% highest posterior density
+        intervals.
+    }
+    \label{fig:peerdriven}
+\end{figure}
+
+\begin{figure}[ht]
+  \includegraphics{realdata-hpd-bc-m2}
+  \vspace{8pt}
+  \caption[
+      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
+      parameters of the BA network model, fitted to five published HIV datasets
+      with \software{netabc} using the prior $m \sim DiscreteUniform(2, 5)$.
+  ]{
+      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
+      parameters of the BA network model, fitted to five published HIV datasets
+      with \software{netabc} using the prior $m \sim$ DiscreteUniform(2, 5).
+      $x$-axes indicate regions of nonzero prior density. In particular, the
+      prior on $m$ was DiscreteUniform(2, 5).
+  }
+  \label{fig:abchpdm2}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics{bctree-posterior}
+  \caption[
+      Approximate marginal posterior distributions of BA model parameters for
+      BC data.
+  ]
+  {
+      Approximate marginal posterior distributions of BA model parameters for
+      BC data. Vertical lines indicate maximum \textit{a posteriori} estimates,
+      and shaded areas are 95\% highest posterior density intervals. $x$-axis
+      indicates regions of nonzero prior density.
+  }
+  \label{fig:bctree}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics{cuevas2009-posterior}
+  \caption[
+      Approximate marginal poosterior distributions of BA model parameters for
+      \textcite{cuevas2009hiv} data. 
+  ]{
+      Approximate marginal poosterior distributions of BA model parameters for
+      \textcite{cuevas2009hiv} data. Vertical lines indicate maximum \textit{a
+      posteriori} estimates, and shaded areas are 95\% highest posterior
+      density intervals. $x$-axis indicates regions of nonzero prior density.
+  }
+  \label{fig:cuevas}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics{li2015-posterior}
+  \caption[
+      Approximate marginal posterior distributions of BA model parameters for
+      \textcite{li2015hiv} data. 
+  ]{
+      Approximate marginal posterior distributions of BA model parameters for
+      \textcite{li2015hiv} data. Vertical lines indicate maximum \textit{a
+      posteriori} estimates, and shaded areas are 95\% highest posterior
+      density intervals. $x$-axis indicates regions of nonzero prior density.
+  }
+  \label{fig:li}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics{niculescu2015-posterior}
+  \caption[
+      Approximate marginal posterior distributions of BA model parameters for
+      \textcite{niculescu2015recent} data.
+  ]{
+      Approximate marginal posterior distributions of BA model parameters for
+      \textcite{niculescu2015recent} data. Vertical lines indicate maximum
+      \textit{a posteriori} estimates, and shaded areas are 95\% highest
+      posterior density intervals. $x$-axis indicates regions of nonzero prior
+      density.
+  }
+  \label{fig:niculescu}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics{novitsky2014-posterior}
+  \caption[
+      Approximate marginal posterior distributions of BA model parameters for
+      \textcite{novitsky2013phylogenetic, novitsky2014impact} data.
+  ]{
+      Approximate marginal posterior distributions of BA model parameters for
+      \textcite{novitsky2013phylogenetic, novitsky2014impact} data. Vertical
+      lines indicate maximum \textit{a posteriori} estimates, and shaded areas
+      are 95\% highest posterior density intervals. $x$-axis indicates regions
+      of nonzero prior density.
+  }
+  \label{fig:novitsky}
+\end{figure}
+
+\begin{figure}[H]
+  \includegraphics{wang2015-posterior}
+  \caption[
+      Approximate marginal posterior distributions of BA model parameters for
+      \textcite{wang2015targeting} data.
+  ]{
+      Approximate marginal posterior distributions of BA model parameters for
+      \textcite{wang2015targeting} data. Vertical lines indicate maximum
+      \textit{a posteriori} estimates, and shaded areas are 95\% highest
+      posterior density intervals. $x$-axis indicates regions of nonzero prior
+      density.
+  }
+  \label{fig:wang}
+\end{figure}
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.0_1000_2_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.0-1000-2-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.5_1000_2_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 1000, $m$ = 2, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.5, $I$ = 1000, $m$ = 2, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.5-1000-2-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.0_1000_2_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.0-1000-2-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.5_1000_2_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 1000, $m$ = 2, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.5, $I$ = 1000, $m$ = 2, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.5-1000-2-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.0_2000_2_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 2000, $m$ = 2, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.0, $I$ = 2000, $m$ = 2, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.0-2000-2-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.5_2000_2_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 2000, $m$ = 2, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.5, $I$ = 2000, $m$ = 2, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.5-2000-2-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.0_2000_2_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 2000, $m$ = 2, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.0, $I$ = 2000, $m$ = 2, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.0-2000-2-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.5_2000_2_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 2000, $m$ = 2, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.5, $I$ = 2000, $m$ = 2, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.5-2000-2-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.0_1000_3_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 1000, $m$ = 3, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.0, $I$ = 1000, $m$ = 3, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.0-1000-3-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.5_1000_3_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 1000, $m$ = 3, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.5, $I$ = 1000, $m$ = 3, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.5-1000-3-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.0_1000_3_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 1000, $m$ = 3, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.0, $I$ = 1000, $m$ = 3, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.0-1000-3-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.5_1000_3_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 1000, $m$ = 3, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.5, $I$ = 1000, $m$ = 3, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.5-1000-3-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.0_2000_3_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 2000, $m$ = 3, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.0, $I$ = 2000, $m$ = 3, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.0-2000-3-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.5_2000_3_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 2000, $m$ = 3, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.5, $I$ = 2000, $m$ = 3, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.5-2000-3-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.0_2000_3_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 2000, $m$ = 3, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.0, $I$ = 2000, $m$ = 3, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.0-2000-3-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.5_2000_3_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 2000, $m$ = 3, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.5, $I$ = 2000, $m$ = 3, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.5-2000-3-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.0_1000_4_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 1000, $m$ = 4, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.0, $I$ = 1000, $m$ = 4, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.0-1000-4-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.5_1000_4_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 1000, $m$ = 4, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.5, $I$ = 1000, $m$ = 4, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.5-1000-4-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.0_1000_4_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 1000, $m$ = 4, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.0, $I$ = 1000, $m$ = 4, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.0-1000-4-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.5_1000_4_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 1000, $m$ = 4, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.5, $I$ = 1000, $m$ = 4, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.5-1000-4-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.0_2000_4_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 2000, $m$ = 4, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.0, $I$ = 2000, $m$ = 4, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.0-2000-4-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/0.5_2000_4_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 2000, $m$ = 4, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.5, $I$ = 2000, $m$ = 4, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:0.5-2000-4-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.0_2000_4_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 2000, $m$ = 4, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.0, $I$ = 2000, $m$ = 4, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.0-2000-4-5000-0}
+\end{figure}
+\clearpage
+
+\begin{figure}[H]
+\centering
+\includegraphics{{abc-posterior/1.5_2000_4_5000_0}.pdf}
+\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 2000, $m$ = 4, and $N$ = 5000.]{
+Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.5, $I$ = 2000, $m$ = 4, and $N$ = 5000.
+Vertical dashed lines indicate true values.
+Shaded areas are 95\% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.
+}
+\label{fig:1.5-2000-4-5000-0}
+\end{figure}
+\clearpage

### 2016-05-02 doc/thesis/discussion.tex

@@
 larger network could produce the exact same transmission tree as before. On the
-other hand, when \gls{I} is large, the coalescent dynamics discussed above for
-\gls{I} also apply, as evidenced by the relative accuracy of the \gls{nltt}.
+other hand, when \gls{I} is large relative to \gls{N}, the coalescent dynamics
+discussed above also apply. That is, the waiting times until the next infection
+increase, resulting in longer coalescence times toward the tips. The relative
+accuracy of the \gls{nltt} in these situations coroborates this hypothesis.
 The \gls{m} parameter, which controls the number of connections added to the
@@
 similarity measure performed well in some cases, though poorly in others.
-\gls{ABC} is often applied using a vector of summary statistics, rather than a
-kernel-based similarity score as we have done here, and methods have been
+\gls{ABC} is often applied using a vector of summary
+statistics~\autocite{marin2012approximate, sunnaaker2013approximate}, rather
+than a kernel-based similarity score as we have done here. Methods have been
 developed to select an optimal combination of summary statistics for a given
@@
 use kernel-\gls{ABC} to fit a model which relaxes some or all of these
-assumptions, which may be a fruitful avenue for future investigation.
+assumptions, which may be a fruitful avenue for future investigation. Despite
+the possible misspecification, our estimates of the power law exponent $\gamma$
+were in line with those reported in the literature (\cref{tab:gamma}).

### 2016-05-02 doc/thesis/methods.tex

@@
 
-We did not consider the time scale of the transmission trees in these
-simulations, only their shape. Therefore, the transmission rate along each edge
-in the network was set to 1, and all transmission trees' branch lengths were
-scaled by their mean. The removal rate of each node was set to 0, implying no
-recovery or death in the population. These assumptions are similar to those
-made by \textcite{leventhal2012inferring}.
+Nodes in our networks followed simple \gls{SI} dynamics, meaning that they
+became infected at a rate proportional to their number of infected neighbours,
+and never recovered.  We did not consider the time scale of the transmission
+trees in these simulations, only their shape. Therefore, the transmission rate
+along each edge in the network was set to 1, the removal rate of each node was
+set to 0, and all transmission trees' branch lengths were scaled by their mean.
+For all analyses, the transmission trees' branch lengths were scaled by
+dividing by their mean. We used the \textit{igraph} library's implementation of
+the BA model~\autocite{csardi2006igraph} to generate the graphs. The analyses
+were run on Westgrid (\url{https://www.westgrid.ca/}) and a local computer
+cluster.
 
@@
 
+\begin{table}[ht]
+  \centering
+  \input{\tablepath/realdata-bc}
+  \caption[Characteristics of published HIV datasets analyzed with \software{netabc}.]
+  {
+    Characteristics of published HIV datasets analyzed with \software{netabc}. 
+    Abbreviations: MSM, men who have sex with men; HET, heterosexual; IDU,
+    injection drug users. The \textcite{novitsky2013phylogenetic,novitsky2014impact} data
+    were sampled from a primarily heterosexual risk environment, but did not
+    explicitly exclude other risk factors.
+  }
+  \label{tab:data}
+\end{table}
+
 We downloaded all sequences associated with each published study from GenBank.
@@
 
-\begin{table}[ht]
-  \centering
-  \input{\tablepath/realdata-bc}
-  \caption[Characteristics of published HIV datasets analyzed with \software{netabc}.]
-  {
-    Characteristics of published HIV datasets analyzed with \software{netabc}. 
-    Abbreviations: MSM, men who have sex with men; HET, heterosexual; IDU,
-    injection drug users. The \textcite{novitsky2013phylogenetic,novitsky2014impact} data
-    were sampled from a primarily heterosexual risk environment, but did not
-    explicitly exclude other risk factors.
-  }
-  \label{tab:data}
-\end{table}
+Empirical studies of contact networks often report the exponent $\gamma$ of the
+power law degree distribution. To compare our results to the literature, we
+simulated 100 networks each according to the \gls{MAP} parameter estimates
+obtained for each investigated dataset. The power-law exponent $\gamma$ was
+calculated for each network using the \software{fit\_power\_law} function in
+\software{igraph}, with the `R.mle' implementation. The median of the 100
+$\gamma$ values was taken as a point estimate for the associated dataset.
+
+For all datasets, we used the priors $\alpha$ $\sim$ Uniform(0, 2) and $N$ and
+$I$ jointly uniform on the region \{$n \leq N \leq 10000$, $n \leq I \leq
+10000$, $I \leq N$\}, where $n$ is the number of tips in the tree (see
+\cref{tab:data}). Since the value $m = 1$ produces networks with no cycles,
+which we considered fairly implausible, we ran one analysis with the prior $m
+\sim$ DiscreteUniform(1, 5), and one with the prior $m \sim$ DiscreteUniform(2,
+5). The other parameters to the SMC algorithm were the same as used for the
+simulation experiments, except that we used 10000 particles instead of 1000 to
+increase the accuracy of the estimated posterior. This was computationally
+feasible due to the small number of runs required for this analysis.

### 2016-05-02 doc/thesis/preamble.tex
Thesis almost done

@@
 \documentclass[11pt]{report}
-\usepackage[osf]{garamondx}
-\usepackage[garamondx,cmbraces]{newtxmath}
+\usepackage{mathptmx}
+\usepackage{graphicx}
 \usepackage[top=1in, bottom=1in, inner=1.25in, outer=1in]{geometry}
@@
 \usepackage{nicefrac}
-\usepackage{amsmath}
+\usepackage{amsmath, amssymb}
 \usepackage{algorithm}
@@
 \newacronym{ERGM}{ERGM}{exponential random graph model}
+\newacronym{IDU}{IDU}{injection drug users}
 
@@
   name=$N$,
-  description=number of nodes in the network
+  description=total number of nodes in a contact network
 }
@@
   name=$I$,
-  description=number of nodes which are eventually infected
+  description=number of infected nodes in a contact network at the time of
+  transmission tree sampling
 }

### 2016-05-02 doc/thesis/preface.tex

@@
 Dynamics and Evolution meeting on April 25, in Woods Hole, Massachussets, USA.
-A poster entitled ``Likelihood-free estimation of contact network parameters
-from viral phylogenies'' was presented at the Intelligent Systems for Molecular
-Biology meeting in Orlando, Florida, USA.
+A poster based on chapter 2 entitled ``Likelihood-free estimation of contact
+network parameters from viral phylogenies'' was presented at the Intelligent
+Systems for Molecular Biology meeting in Orlando, Florida, USA.
 

### 2016-05-03 doc/figures/contactnet-empty.tex
GLM thingies

\documentclass[landscape]{article}
\usepackage{tikz}
\usepackage[T1]{fontenc}
\usepackage{mathptmx}
\usepackage{fullpage}

\usetikzlibrary{positioning}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}
\usetikzlibrary{automata}

\graphicspath{{stock/}}

\begin{document}
\pagestyle{empty}

\begin{tikzpicture}
  [every node/.style = {inner sep=0pt},
   every path/.style = {->, >=stealth, very thick},
   label/.style = {anchor=south, color=white, inner sep=6pt},
   time/.style = {anchor=north west, inner sep=2pt}]
  \def\pht{1.5cm}

  \node at (-6, 5.5) (a) {\includegraphics[height=\pht]{person}};
  \node at (a.south) [label] {\huge a};
  \node [below left=of a] (b) {\includegraphics[height=\pht]{person}};
  \node at (b.south) [label] {\huge b};
  \node [below=of b] (e) {\includegraphics[height=\pht]{person}};
  \node at (e.south) [label] {\huge e};
  \node [below right=of a] (c) {\includegraphics[height=\pht]{person}};
  \node at (c.south) [label] {\huge c};
  \node [below=of c] (d) {\includegraphics[height=\pht]{person}};
  \node at (d.south) [label] {\huge d};

  \fill [shade, opacity=0.2] (a.east) -- (c.north) -- (c.north west) -- (a.south east) -- cycle;
  \fill [shade, opacity=0.2] (a.west) -- (b.north) -- (b.north east) -- (a.south west) -- cycle;
  \fill [shade, opacity=0.2] (a.south) -- (a.south west) -- (e.north) -- (e.north east) -- cycle;
  \fill [shade, opacity=0.2] (b.south west) -- (b.south) -- (e.north) -- (e.north west) -- cycle;
  \fill [shade, opacity=0.2] ($(c.south) + (-8pt, 0)$) -- ++(16pt, 0) -- ($(d.north) + (8pt, 0)$) -- ++(-16pt, 0) -- cycle;
  \fill [shade, opacity=0.2] (b.east) -- (b.south east) -- (d.west) -- (d.north west) -- cycle;
\end{tikzpicture}

\end{document}

### 2016-05-03 doc/figures/gridsearch-expt.tex
Mostly writing
GLM thingies

--- 2016-01-26 doc/thesis/figures/gridsearch_expt.tex
@@
 \usepackage[T1]{fontenc}
-\usepackage[sfmath]{kpfonts}
-\usepackage[default]{gillius}
+\usepackage{mathptmx}
 \usepackage[normalem]{ulem}
@@
 
+\immediate\write18{./gridsearch-expt-extra.R}
 \pagestyle{empty}
@@
   \node (rect1) [box] { };
-  \node at (rect1.north) [anchor=north] {\uline{$k$ testing values}};
+  \node at (rect1.north) [anchor=north] {\uline{$k$ test values}};
   \node (t3) [above=0.5cm of rect1.south, inner sep=6pt, circle, draw, fill=blue!60!white] { };
@@
   \node (rect2) [box, right=of rect1] { };
-  \node at (rect2.north) [anchor=north] {\uline{$n$ training values}};
+  \node at (rect2.north) [anchor=north] {\uline{$n$ grid values}};
   \node (bar) [above=0.5cm of rect2.south, minimum width=6cm, minimum height=0.5cm, 
@@
   \node (rect3) [box, below=of rect1] { };
-  \node at (rect3.north) [anchor=north] {\uline{$10 \times k$ testing trees}};
+  \node at (rect3.north) [anchor=north] {\uline{$10 \times k$ test trees}};
   \node [above=4pt of rect3.south] {
@@
   \node (rect4) [box, below=of rect2] { };
-  \node at (rect4.north) [anchor=north] {\uline{$15 \times n$ training trees}};
+  \node at (rect4.north) [anchor=north] {\uline{$15 \times n$ grid trees}};
   \node [above=4pt of rect4.south] {
@@
   \node [above=6pt of rect6.south, minimum width=\boxw, text width=\boxw, align=center] {
-    point estimate = training value with highest median kernel score
+    point estimate = grid value with highest median kernel score
   };

### 2016-05-03 doc/figures/kernel-expt-extra.R

@@ -34,3 +34,43 @@ for (i in 1:3) {
     plot(t, direction="down", edge.width=5, edge.color=col[i], show.tip.label=FALSE)
     dev.off()
 }
+
+npt <- 5
+x <- c(rnorm(npt, sd=0.2), rnorm(npt, mean=2, sd=0.2), rnorm(npt, mean=4, sd=0.2))
+y <- rnorm(npt*3)
+png("tinypca.png", bg="transparent", width=500, height=100)
+par(mar=c(0, 0, 0, 0) + 0.1)
+plot(x, y, pch=16, cex=3, col=rep(col, each=npt))
+box(lwd=4)
+dev.off()
+
+x <- c(rnorm(npt, sd=0.2), rnorm(npt, mean=2, sd=0.2), rnorm(npt, mean=4, sd=0.2))
+y <- c(rnorm(npt, mean=1, sd=0.2), rnorm(npt, mean=0, sd=0.2), rnorm(npt, mean=1, sd=0.2))
+png("tinyksvr.png", bg="transparent", width=100, height=100, type="cairo")
+par(mar=c(0, 0, 0, 0) + 0.1)
+plot(splinefun(x=c(0, 2, 4), y=c(1, 0, 1)), lwd=4, xlim=c(-0.5, 4.5))
+plot(splinefun(x=c(0, 2, 4), y=c(1, 0, 1)), lwd=32, add=TRUE, xlim=c(-0.5, 4.5), 
+     col=rgb(0, 0, 0, alpha=0.5))
+points(x, y, pch=16, cex=2, col=rep(col, each=npt))
+box(lwd=4)
+dev.off()
+
+x <- c(rnorm(npt, sd=0.2), rnorm(npt, mean=2, sd=0.2), rnorm(npt, mean=4, sd=0.2))
+y <- c(rnorm(npt, mean=0, sd=0.2), rnorm(npt, mean=0.33, sd=0.2), rnorm(npt, mean=1, sd=0.2))
+png("tinysvr.png", bg="transparent", width=100, height=100, type="cairo")
+par(mar=c(0, 0, 0, 0) + 0.1)
+plot(x, y, pch=16, cex=2, col=rep(col, each=npt), type="n")
+abline(a=0, b=0.2, lwd=4)
+abline(a=0, b=0.2, lwd=32, col=rgb(0, 0, 0, alpha=0.5))
[…]

### 2016-05-03 doc/figures/kernel-idea.R

@@ -1,0 +1,74 @@
+#!/usr/bin/env Rscript
+
+suppressPackageStartupMessages(library(netabc))
+library(RColorBrewer)
+library(grid)
+
+epidemic.mst <- function (g, nI) {
+    t <- mst(g)
+    I <- vcount(g)
+
+    while (I > nI) {
+        v <- which(degree(t) == 1)[1]
+        t <- delete_edges(t, E(t)[inc(v)])
+        I <- I - 1
+    }
+
+    inodes <- which(degree(t) > 0)
+    iedges <- get.edge.ids(g, c(t(as_edgelist(t))))
+    list(nodes=inodes, edges=iedges)
+}
+
+n <- 40
+t <- 12
+alpha <- c(0, 2)
+m <- 2
+
+br <- list(rep(1, t - 1), c(1, rep(0.1, t-1)))
+tree <- lapply(mapply(rcoal, list(t), br=br, SIMPLIFY=FALSE), ladderize)
+net <- mapply(sample_pa, list(n), power=as.list(alpha), m=m, directed=FALSE, SIMPLIFY=FALSE)
+
+epi <- lapply(net, epidemic.mst, t)
+inodes <- lapply(epi, "[[", "nodes")
+iedges <- lapply(epi, "[[", "edges")
+ecol <- mapply(rep, list(NA), lapply(net, ecount), SIMPLIFY=FALSE)
+ecol <- mapply(function (x, i) { x[i] <- "red"; x}, ecol, iedges, SIMPLIFY=FALSE)
+vcol <- mapply(rep, list("black"), lapply(net, ecount), SIMPLIFY=FALSE)
+vcol <- mapply(function (x, i) { x[i] <- "red"; x}, vcol, inodes, SIMPLIFY=FALSE)
+
+ewidth <- mapply(rep, list(0), lapply(net, ecount), SIMPLIFY=FALSE)
+ewidth <- mapply(function (x, i) { x[i] <- 3; x}, ewidth, iedges, SIMPLIFY=FALSE)
+
+lay <- lapply(net, layout_nicely)
+
+pdf("kernel-idea.pdf", width=6, height=4, family="Gillius ADF")
+par(mfcol=c(2, 3), mar=c(0, 2, 0, 2))
+
+# plot networks
+mapply(plot, net, edge.color="gray", vertex.color="black", vertex.label=NA,
+       vertex.size=6, edge.width=3, layout=lay)
+
+# plot epidemics
[…]

### 2016-05-03 doc/figures/pa-example.R

@@ -1,0 +1,15 @@
+#!/usr/bin/env Rscript
+
+library(igraph)
+library(extrafont)
+
+set.seed(0)
+g0.small <- sample_pa(50, m=2, power=0, directed=FALSE)
+g2.small <- sample_pa(50, m=2, power=2, directed=FALSE)
+
+pdf("pa-example.pdf", family="Gillius ADF", width=6, height=3)
+par(mfrow=c(1, 2))
+par(mar=c(2, 2, 2, 2) + 0.1)
+plot(g0.small, vertex.label=NA, vertex.color="black", vertex.size=5)
+plot(g2.small, vertex.label=NA, vertex.color="black", vertex.size=5)
+dev.off()

### 2016-05-03 doc/figures/pa.R

@@ -1,21 +1,42 @@
 #!/usr/bin/env Rscript
 
 library(igraph)
+set.seed(4)
 
-nv <- c(3, 4, 5, 6, 12)
+nv <- c(3, 4, 5, 6, 12, 12)
+nI <- 7
 
 g <- list(NULL)
 for (n in nv) {
-    g[[length(g)+1]] <- sample_pa(n, m=2, power=1, directed=FALSE, start.graph=tail(g, 1)[[1]])
+    g[[length(g)+1]] <- sample_pa(n, m=2, power=1, directed=FALSE, 
+                                  start.graph=tail(g, 1)[[1]])
 }
 g <- tail(g, length(g)-1)
+fullg <- tail(g, 1)[[1]]
+
+t <- mst(fullg)
+I <- vcount(fullg)
+while (I > nI) {
+    v <- which(degree(t) == 1)[1]
+    t <- delete_edges(t, E(t)[inc(v)])
+    I <- I - 1
+}
+
+inodes <- which(degree(t) > 0)
+iedges <- get.edge.ids(fullg, c(t(as_edgelist(t))))
 
 vcol <- mapply(rep, list(c("black", "white")), 
                as.list(as.data.frame(rbind(nv, max(nv)-nv))), SIMPLIFY=FALSE)
+vcol[[length(vcol)]][inodes] <- "red"
+ecol <- mapply(rep, list("black"), lapply(g, ecount), SIMPLIFY=FALSE)
+ecol[[length(ecol)]][iedges] <- "red"
+vsize <- mapply(rep, list(5), as.list(nv), SIMPLIFY=FALSE)
+vsize[[length(vsize)]][inodes] <- 10
 
 lay <- layout_in_circle(tail(g, 1)[[1]])
 pdf("pa.pdf")
+par(mar=c(0, 0, 0, 0) + 0.1)
 mapply(plot, g, vertex.color=vcol, vertex.frame.color=vcol,
-       vertex.label=list(NA), vertex.size=list(5), 
-       layout=list(lay))
+       vertex.label=list(NA), vertex.size=vsize, edge.color=ecol,
+       edge.width=2, layout=list(lay))
 dev.off()

### 2016-05-03 doc/figures/smc-test.R

@@ -4,7 +4,7 @@ library(data.table)
 library(extrafont)
 
 trace.file <- system2(c("locate", "check_smc_trace.tsv"), stdout=TRUE)
-trace <- fread(trace.file)
+trace <- fread(trace.file[1])
 
 trace <- trace[iter == max(iter)]
 trace <- trace[sample(1:nrow(trace), prob=weight, replace=TRUE)]

### 2016-05-03 doc/thesis/global.R

@@ -1,0 +1,54 @@
+#!/usr/bin/env Rscript
+
+library(netabc)
+library(Hmisc)
+library(xtable)
+
+pp <- function (p, eq=FALSE) {
+    if (p < 1e-5) {
+        "{<}10^{-5}"
+    } else {
+        n <- latexSN(round(p, -floor(log10(p))))
+        n <- sub("^1\\\\!\\\\times\\\\!", "", n)
+        if (eq) {
+            paste0("=", n)
+        } else {
+            n
+        }
+    }
+}
+
+f <- Sys.glob("../../simulations/abc-pa-free-m/point-estimate/*")
+d <- fread(f)
+d <- d[,m := round(m)]
+d <- d[,alpha_error := abs(true_alpha - alpha)]
+d <- d[,N_error := abs(true_N - N)]
+d <- d[,I_error := abs(true_I - I)]
+d <- d[,m_error := abs(true_m - m)]
+
+make.glm <- function (param, family, link) {
+    frm <- paste0(param, "_error~factor(true_alpha)+factor(true_I)+factor(true_m)")
+    m <- glm(as.formula(frm), family(link=link), d)
+    df <- as.data.frame(coef(summary(m)))
+    df$Parameter <- sub(").*", "", sub("factor(true_", "", rownames(df), fixed=TRUE))
+    df$Parameter[1] <- "(Intercept)"
+    df$Value <- sub(".*)", "", rownames(df))
+    df$Value[1] <- "-"
+    setDT(df)
+    setnames(df, "Std. Error", "Standard error")
+    if ("Pr(>|t|)" %in% colnames(df)) {
+        df$`t value` <- NULL
+        setnames(df, "Pr(>|t|)", "p")
+    } else {
+        df$`z value` <- NULL
+        setnames(df, "Pr(>|z|)", "p")
+    }
+    setcolorder(df, c("Parameter", "Value", "Estimate", "Standard error", "p"))
+    df
+}
+ 
+# make GLMs
+alpha.glm <- make.glm("alpha", Gamma, "inverse")
+I.glm <- make.glm("I", Gamma, "inverse")
+N.glm <- make.glm("N", Gamma, "inverse")
+m.glm <- make.glm("m", poisson, "log")

### 2016-05-04 doc/thesis/appendix.Rnw
Moar

@@
-\begin{figure}[H]
+<<setup, include=FALSE>>=
+    source("global.R")
+@
+
+\chapter{Additional plots}
+
+\begin{figure}[ht]
   \centering
@@
 
-\begin{figure}[H]
+\begin{figure}[ht]
   \centering
@@
 
-\begin{figure}[H]
+\clearpage
+
+\begin{figure}[ht]
   \centering
@@
 
-\begin{figure}[H]
+\clearpage
+
+%\begin{figure}[ht]
+%  \centering
+%  \includegraphics{abc-point-estimate-m1.pdf}
+%  \caption[
+%    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
+%    running \software{netabc} on simulated data, for simulations with \gls{m} = 1.
+%  ]{
+%    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
+%    running \software{netabc} on simulated data, for simulations with \gls{m} = 3.
+%    Dashed lines indicate true values. (A) Estimates of \gls{alpha} and \gls{I}
+%    which were varied in these simulations against known values. (B) Estimates
+%    of \gls{m} and \gls{N} which were held fixed in these simulations at the
+%    values \gls{m} = 1 and \gls{N} = 5000.
+%  }        
+%  \label{fig:abcptm3}
+%\end{figure}
+
+\begin{figure}[ht]
   \centering
@@
   \centering
-  \includegraphics{sir-trajectories.pdf}
+  \includegraphics[width=\textwidth]{sir-trajectories.pdf}
   \caption[Average trajectories of epidemics simulated according to the
@@
 
-\begin{figure}[H]
-  \includegraphics{bctree-posterior}
+\clearpage
+
+\begin{figure}[ht]
+  %\includegraphics{bctree-posterior}
   \caption[
@@
 
-\begin{figure}[H]
+\begin{figure}[ht]
   \includegraphics{cuevas2009-posterior}
@@
 
-\begin{figure}[H]
+\begin{figure}[ht]
   \includegraphics{li2015-posterior}
@@
 
-\begin{figure}[H]
+\begin{figure}[ht]
   \includegraphics{niculescu2015-posterior}
@@
 
-\begin{figure}[H]
+\begin{figure}[ht]
   \includegraphics{novitsky2014-posterior}
@@
 
-\begin{figure}[H]
+\begin{figure}[ht]
   \includegraphics{wang2015-posterior}
@@
 
+\clearpage
+
 <<abc_posterior, echo=FALSE, results="asis">>=
@@
         lab <- sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate)
-        cat("\\begin{figure}[H]\n")
+        cat("\\begin{figure}[ht]\n")
         cat("\\centering\n")
@@
         cat(sprintf("\\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\\alpha$ = %.1f, $I$ = %d, $m$ = %d, and $N$ = %d.]{\n", alpha, I, m, N))
-        cat(sprintf("Approximate marginal posterior distributions of BA model parameters obtained by applying \textit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = %.1f, $I$ = %d, $m$ = %d, and $N$ = %d.
+        cat(sprintf("Approximate marginal posterior distributions of BA model parameters obtained by applying \\software{netabc} to a simulated transmission tree with BA parameter values $\\alpha$ = %.1f, $I$ = %d, $m$ = %d, and $N$ = %d.
 Vertical dashed lines indicate true values.
@@
 @
+
+%\chapter{GLM results}
+%
+%<<glm_tables, echo=FALSE, warning=FALSE, error=FALSE, results="asis">>=
+%    format.glm <- function (df) {
+%        df[Parameter == "alpha", Parameter := "$\\alpha$"]
+%        df[,p := paste0("$", sapply(p, pp, eq=FALSE), "$")]
+%        setnames(df, "p", "$p$-value")
+%        df
+%    }
+%
+%    alpha.glm <- format.glm(alpha.glm)
+%    I.glm <- format.glm(I.glm)
+%    m.glm <- format.glm(m.glm)
+%    N.glm <- format.glm(N.glm)
+%
+%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+%                        "$\\alpha$ to true values of BA parameters.")
+%    capt <- paste(capt.short, "GLM was fitted with a Gamma distribution and",
+%                  "inverse link function. Coefficients are interpretable as",
+%                  "additive effects on the inverse of the mean error.")
+%    print(xtable(alpha.glm, label="tab:glmalpha", caption=c(capt, capt.short)),
+%          include.rownames=FALSE, sanitize.text.function=identity,
+%          sanitize.colnames.function=identity)
+%
+%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+%                        "$I$ to true values of BA parameters.")
+%    capt <- paste(capt.short, "GLM was fitted with a Gamma distribution and",
+%                  "inverse link function. Coefficients are interpretable as",
+%                  "additive effects on the inverse of the mean error.")
+%    print(xtable(I.glm, label="tab:glmI", caption=c(capt, capt.short)),
+%          include.rownames=FALSE, sanitize.text.function=identity,
+%          sanitize.colnames.function=identity)
+%
+%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+%                        "$m$ to true values of BA parameters.")
+%    capt <- paste(capt.short, "GLM was fitted with a Poisson distribution and",
+%                  "identity link function. Coefficients are interpretable as",
+%                  "additive effects on the exponential of the mean error.")
+%    print(xtable(m.glm, label="tab:glmm", caption=c(capt, capt.short)),
+%          include.rownames=FALSE, sanitize.text.function=identity,
+%          sanitize.colnames.function=identity)
+%
+%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+%                        "$N$ to true values of BA parameters.")
+%    capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
+%                  "identity link function. Coefficients are interpretable as",
+%                  "additive effects on the inverse of the mean error.")
+%    print(xtable(N.glm, label="tab:glmN", caption=c(capt, capt.short)),
+%          include.rownames=FALSE, sanitize.text.function=identity,
+%          sanitize.colnames.function=identity)
+%@

### 2016-05-04 doc/thesis/appendix.tex
Moar

@@
-\begin{figure}[H]
+\chapter{Additional plots}
+
+\begin{figure}[ht]
   \centering
@@
 
-\begin{figure}[H]
+\clearpage
+
+\begin{figure}[ht]
 \centering
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 
-\begin{figure}[H]
+\begin{figure}[ht]
 \centering
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 1000, $m$ = 2, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.5, $I$ = 1000, $m$ = 2, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.5, $I$ = 1000, $m$ = 2, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 1000, $m$ = 2, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.5, $I$ = 1000, $m$ = 2, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.5, $I$ = 1000, $m$ = 2, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 2000, $m$ = 2, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.0, $I$ = 2000, $m$ = 2, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.0, $I$ = 2000, $m$ = 2, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 2000, $m$ = 2, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.5, $I$ = 2000, $m$ = 2, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.5, $I$ = 2000, $m$ = 2, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 2000, $m$ = 2, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.0, $I$ = 2000, $m$ = 2, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.0, $I$ = 2000, $m$ = 2, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 2000, $m$ = 2, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.5, $I$ = 2000, $m$ = 2, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.5, $I$ = 2000, $m$ = 2, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 1000, $m$ = 3, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.0, $I$ = 1000, $m$ = 3, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.0, $I$ = 1000, $m$ = 3, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 1000, $m$ = 3, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.5, $I$ = 1000, $m$ = 3, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.5, $I$ = 1000, $m$ = 3, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 1000, $m$ = 3, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.0, $I$ = 1000, $m$ = 3, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.0, $I$ = 1000, $m$ = 3, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 1000, $m$ = 3, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.5, $I$ = 1000, $m$ = 3, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.5, $I$ = 1000, $m$ = 3, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 2000, $m$ = 3, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.0, $I$ = 2000, $m$ = 3, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.0, $I$ = 2000, $m$ = 3, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 2000, $m$ = 3, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.5, $I$ = 2000, $m$ = 3, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.5, $I$ = 2000, $m$ = 3, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 2000, $m$ = 3, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.0, $I$ = 2000, $m$ = 3, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.0, $I$ = 2000, $m$ = 3, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 2000, $m$ = 3, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.5, $I$ = 2000, $m$ = 3, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.5, $I$ = 2000, $m$ = 3, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 1000, $m$ = 4, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.0, $I$ = 1000, $m$ = 4, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.0, $I$ = 1000, $m$ = 4, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 1000, $m$ = 4, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.5, $I$ = 1000, $m$ = 4, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.5, $I$ = 1000, $m$ = 4, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 1000, $m$ = 4, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.0, $I$ = 1000, $m$ = 4, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.0, $I$ = 1000, $m$ = 4, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 1000, $m$ = 4, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.5, $I$ = 1000, $m$ = 4, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.5, $I$ = 1000, $m$ = 4, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 2000, $m$ = 4, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.0, $I$ = 2000, $m$ = 4, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.0, $I$ = 2000, $m$ = 4, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 2000, $m$ = 4, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 0.5, $I$ = 2000, $m$ = 4, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.5, $I$ = 2000, $m$ = 4, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 2000, $m$ = 4, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.0, $I$ = 2000, $m$ = 4, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.0, $I$ = 2000, $m$ = 4, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 2000, $m$ = 4, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying 	extit{netabc} to a simulated transmission tree with BA parameter values $alpha$ = 1.5, $I$ = 2000, $m$ = 4, and $N$ = 5000.
+Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.5, $I$ = 2000, $m$ = 4, and $N$ = 5000.
 Vertical dashed lines indicate true values.
@@
 \clearpage
+
+%\chapter{GLM results}
+%
+%<<glm_tables, echo=FALSE, warning=FALSE, error=FALSE, results="asis">>=
+%    format.glm <- function (df) {
+%        df[Parameter == "alpha", Parameter := "$\\alpha$"]
+%        df[,p := paste0("$", sapply(p, pp, eq=FALSE), "$")]
+%        setnames(df, "p", "$p$-value")
+%        df
+%    }
+%
+%    alpha.glm <- format.glm(alpha.glm)
+%    I.glm <- format.glm(I.glm)
+%    m.glm <- format.glm(m.glm)
+%    N.glm <- format.glm(N.glm)
+%
+%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+%                        "$\\alpha$ to true values of BA parameters.")
+%    capt <- paste(capt.short, "GLM was fitted with a Gamma distribution and",
+%                  "inverse link function. Coefficients are interpretable as",
+%                  "additive effects on the inverse of the mean error.")
+%    print(xtable(alpha.glm, label="tab:glmalpha", caption=c(capt, capt.short)),
+%          include.rownames=FALSE, sanitize.text.function=identity,
+%          sanitize.colnames.function=identity)
+%
+%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+%                        "$I$ to true values of BA parameters.")
+%    capt <- paste(capt.short, "GLM was fitted with a Gamma distribution and",
+%                  "inverse link function. Coefficients are interpretable as",
+%                  "additive effects on the inverse of the mean error.")
+%    print(xtable(I.glm, label="tab:glmI", caption=c(capt, capt.short)),
+%          include.rownames=FALSE, sanitize.text.function=identity,
+%          sanitize.colnames.function=identity)
+%
+%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+%                        "$m$ to true values of BA parameters.")
+%    capt <- paste(capt.short, "GLM was fitted with a Poisson distribution and",
+%                  "identity link function. Coefficients are interpretable as",
+%                  "additive effects on the exponential of the mean error.")
+%    print(xtable(m.glm, label="tab:glmm", caption=c(capt, capt.short)),
+%          include.rownames=FALSE, sanitize.text.function=identity,
+%          sanitize.colnames.function=identity)
+%
+%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+%                        "$N$ to true values of BA parameters.")
+%    capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
+%                  "identity link function. Coefficients are interpretable as",
+%                  "additive effects on the inverse of the mean error.")
+%    print(xtable(N.glm, label="tab:glmN", caption=c(capt, capt.short)),
+%          include.rownames=FALSE, sanitize.text.function=identity,
+%          sanitize.colnames.function=identity)
+%@

### 2016-05-04 doc/thesis/discussion.tex
Moar

@@
 The preferential attachment power \gls{alpha} had a very strong influence on
-tree shape in the range of values we considered. Although the tree kernel was
-the most effective classifier for \gls{alpha}, a tree balance statistic
-performed nearly as well. This result was intuitive: high \gls{alpha} values
-produce networks with few well-connected ``superspreader'' nodes which are
-involved in a large number of transmissions, resulting in a highly unbalanced
-ladder-like tree structure (\cref{fig:alphatrees}). The \gls{I} parameter,
-representing the prevalence at the time of sampling, was also generally
-estimable. The dynamics of the \gls{SI} model, and the coalescent
-process~\autocite{kingman1982coalescent}, offer a potential explanation for
-this result. \Cref{fig:sir} shows simulated trajectories of the \gls{SIR}
-model. In the initial phase of the epidemic, when \gls{I} is small, each new
-transmission results in potentially many new discordant edges, thus decreasing
-the waiting time until the next transmission. Hence, there is an
-early exponential growth phase, producing many short branches near the root of
-the tree. As the epidemic gets closer to saturating the network, the number of
-discordant edges decays, causing longer waiting times. The distribution of
-coalescence times in the tree should therefore be informative about
-\gls{I}~\autocite{volz2009phylodynamics}. This information is captured by the
-tree kernel, and also by the \gls{nltt} statistic, which both performed quite
-will in classifying \gls{I} (\cref{fig:rsquared}).
+tree shape in the range of values we considered
+(\cref{fig:alphatrees,fig:kpca}). Although the tree kernel was the most
+effective classifier for \gls{alpha}, a Sackin's index of tree imbalance
+performed nearly as well (\cref{fig:rsquared}). This result was intuitive: high
+\gls{alpha} values produce networks with few well-connected ``superspreader''
+nodes which are involved in a large number of transmissions, resulting in a
+highly unbalanced ladder-like tree structure (\cref{fig:alphatrees}). There was
+no observeable bias in the estimates of \gls{alpha} obtained with
+\software{netabc}, however the variation in these estimates was higher for
+$\alpha < 1$ than for $\alpha \geq 1$ (\cref{fig:abcptm2,tab:abchpd}). The
+relationship between \gls{alpha} and the power law exponent \gls{gamma} may
+explain this result (\cref{fig:gamma}). The \gls{gamma} values associated with
+\gls{alpha} = 0 and \gls{alpha} = 0.5 are nearly identical (about 2.28 for
+\gls{alpha} = 0 and 2.33 for \gls{alpha} = 0.5 with \gls{N} = 5000 and \gls{m}
+= 2). In other words, the degree distributions of networks with \gls{alpha} $<
+1$ are similar to each other, which may result in similarity of corresponding
+transmission trees as well. 
+%The accuracy of \gls{alpha} estimates also increased when the value of the
+%\gls{m} parameter was higher (compare
+%\cref{fig:abcptm2,fig:abcptm3,fig:abcptm4}).
+
+\textbf{ART: THIS SECTION IS IN PROGRESS PLEASE DO NOT READ IT YET}
+
+The \gls{I} parameter, representing the prevalence at the time of sampling, was
+accurately classified by the tree kernel and the \gls{nltt}, but not by
+Sackin's index (\cref{fig:rsquared}). When \gls{N} was held fixed, \gls{I} was
+estimated accurately and without bias by grid search
+\cref{fig:gridest,fig:gridI}. When the \gls{BA} model was fit with
+\software{netabc}, the 95\% \gls{HPD} intervals for \gls{I} usually contained
+the true value, but the \gls{MAP} estimates were biased slightly upward
+(\cref{tab:abchpd}).
+
+also generally estimable, although it was slightly over-estimated for both
+cases we considered with kernel-\gls{ABC}.
+
+The dynamics of the \gls{SI} model,
+and the coalescent process~\autocite{kingman1982coalescent}, offer a potential
+explanation for this result. In our simulations, we assumed that all discordant
+edges shared the same transmission rate, so that the waiting time until the
+next transmission in the entire network was always inversely proportional to
+the number of discordant edges. Therefore, when there are a high number of
+discordant edges,
+
+In the initial phase of the epidemic, when \gls{I}
+is small, each new transmission results in many new discordant edges.
+
+thus
+decreasing the waiting time until the next transmission. Hence,
+there is an early exponential growth phase, producing many short branches near
+the root of the tree. As the epidemic gets closer to saturating the network,
+the number of discordant edges decays, causing longer waiting times. The
+distribution of coalescence times in the tree should therefore be informative
+about \gls{I}~\autocite{volz2009phylodynamics}. This information is captured by
+the tree kernel, and also by the \gls{nltt} statistic, which both performed
+quite well in classifying \gls{I} (\cref{fig:rsquared}).
+
+\textbf{END OF IN PROGRESS}
 
 The number of nodes in the network, \gls{N}, exhibited the most variation in
-terms of being estimable. There was almost no difference between trees
+terms of its effect on tree shape. There was almost no difference between trees
 simulated under different \gls{N} values when the number of infected nodes
-\gls{I} was very small. There is an intuitive explanation for this result,
-namely that adding additional nodes does not change the edge density or overall
-shape of a \gls{BA} network. This can be illustrated by imagining that we add a
-small number of nodes to a network after the epidemic simulation has already
-been completed. It is possible that none of these new nodes attains a
-connection to any infected node. Thus, running the simulation again on the new,
-larger network could produce the exact same transmission tree as before. On the
-other hand, when \gls{I} is large relative to \gls{N}, the coalescent dynamics
+\gls{I} was small. There is an intuitive explanation for this result, namely
+that adding additional nodes does not change the edge density or overall shape
+of a \gls{BA} network. This can be illustrated by imagining that we add a small
+number of nodes to a network after the epidemic simulation has already been
+completed. It is possible that none of these new nodes attains a connection to
+any infected node. Thus, running the simulation again on the new, larger
+network could produce the exact same transmission tree as before. On the other
+hand, when \gls{I} is large relative to \gls{N}, the coalescent dynamics
 discussed above also apply. That is, the waiting times until the next infection
@@
 accuracy of the \gls{nltt} in these situations coroborates this hypothesis.
+
+%As mentioned above, \gls{I} and \gls{N} were both systematically over-estimated
+%by our method, although the bias was more severe for \gls{N} than for \gls{I}.
+%
+%When \gls{N} was held fixed and \gls{I} was varied, grid search was able to
+%recover \gls{I} without bias (\cref{fig:est}).
+
 The \gls{m} parameter, which controls the number of connections added to the
@@
 range of permitted $\gamma$ values may be warranted.
+
+The combination of method, model, and priors we employed did not produce
+perfect estimates of any of the parameters. The estimates of \gls{alpha} were
+the most accurate, although the variance of the estimates was high and the
+confidence intervals were wide for \gls{alpha} $< 1$
+(\cref{tab:abchpd,fig:abcptm2,fig:abcptm3,fig:abcptm4}). The estimates of
+\gls{N} and \gls{I} were both biased, and the estimates of \gls{m} were largely
+uninformative. Despite these issues, a major result of our our investigation is
+that some contact network parameters have a measurable impact on tree shape
+which can be used to perform statistical inference. Further refinements to
+\software{netabc}, as well as the use of more sophisticated network models, may
+improve the accuracy and precision of these estimates.
 

### 2016-05-04 doc/thesis/intro.tex
Moar

@@
 over all pairs of nodes $(n_1, n_2)$, where $n_1$ is a node in $T_1$ and $n_2$
-is a node in $T_2$. Folloting \textcite{poon2013mapping}, let $N(T)$ denote the
+is a node in $T_2$. Following \textcite{poon2013mapping}, let $N(T)$ denote the
 set of all nodes in $T$, $\nc(n)$ be the number of children of node $n$,
@@
 Algorithms for performing \gls{ABC} fall into one of three categories:
-rejection, \gls{MCMC}, and \gls{SMC}. To simplify the math, we shall restrict
-the descriptions of these algorithms to the case of one simulated dataset per
-parameter particle (the meaning of this will become clear shortly). The
-extension to multiple datasets per particle is straightforward and will be
-given at the end of the section. We use the variable $x$ to refer to the pair
-$(\theta, z)$, so that the \gls{ABC} target distribution can be written
-$\pi_\varepsilon(x \mid y)$.
+rejection, \gls{MCMC}, and \gls{SMC}~\autocite{marin2012approximate}. To
+simplify the math, we shall restrict the descriptions of these algorithms to
+the case of one simulated dataset per parameter particle (the meaning of this
+will become clear shortly). The extension to multiple datasets per particle is
+straightforward and will be given at the end of the section. We use the
+variable $x$ to refer to the pair $(\theta, z)$, so that the \gls{ABC} target
+distribution can be written $\pi_\varepsilon(x \mid y)$.
 

### 2016-05-04 doc/thesis/methods.tex
Moar

@@
 \gls{BA} network, we allowed an epidemic to spread until \gls{I} nodes were
-infected, and sampled a transmission tree at that time. We assumed that all
-contacts had symmetric transmission risk, which was implemented by replacing
-each undirected edge in the network with two directed edges (one in each
-direction).
+infected, and sampled a transmission tree at that time. The \gls{alpha}
+parameter is unitless, while \gls{m} has units of edges or connections per
+vertex, and \gls{N} and \gls{I} both have units of nodes or individuals.
 
-Nodes in our networks followed simple \gls{SI} dynamics, meaning that they
-became infected at a rate proportional to their number of infected neighbours,
-and never recovered.  We did not consider the time scale of the transmission
-trees in these simulations, only their shape. Therefore, the transmission rate
-along each edge in the network was set to 1, the removal rate of each node was
-set to 0, and all transmission trees' branch lengths were scaled by their mean.
-For all analyses, the transmission trees' branch lengths were scaled by
-dividing by their mean. We used the \textit{igraph} library's implementation of
-the BA model~\autocite{csardi2006igraph} to generate the graphs. The analyses
-were run on Westgrid (\url{https://www.westgrid.ca/}) and a local computer
-cluster.
+We assumed that all contacts had symmetric transmission risk, which was
+implemented by replacing each undirected edge in the network with two directed
+edges (one in each direction). Nodes in our networks followed simple \gls{SI}
+dynamics, meaning that they became infected at a rate proportional to their
+number of infected neighbours, and never recovered. We did not consider the
+time scale of the transmission trees in these simulations, only their shape.
+Therefore, the transmission rate along each edge in the network was set to 1,
+the removal rate of each node was set to 0, and all transmission trees' branch
+lengths were scaled by their mean. For all analyses, the transmission trees'
+branch lengths were scaled by dividing by their mean. We used the
+\textit{igraph} library's implementation of the BA
+model~\autocite{csardi2006igraph} to generate the graphs. The analyses were run
+on Westgrid (\url{https://www.westgrid.ca/}) and a local computer cluster. With
+the exception of our own programs, all analyses were done in \software{R}, and
+all packages listed below are \software{R} packages.
 
-\subsubsection*{Kernel classifiers}
+\subsubsection*{Classifiers for BA model parameters based on tree shape}
 \label{subsec:kernel}
@@
 \defn{meta-parameters} to distinguish them from the parameters of the \gls{BA}
-model. With the exception of our own programs, all analyses were done in
-\software{R}, and all packages listed below are \software{R} packages.
+model.
 
@@
 was obtained using the \software{hpd} function in the \software{TeachingDemos}
-package.
+package~\autocite{snow2013teachingdemos}.
 
@@
 experiments is shown in \cref{fig:gridexpt}. 
+%Spearman's correlation was used to test whether the number of tips in the tree
+%was correlated with the acccuracy of the point estimates. One-way \gls{ANOVA}
+%was used to test whether the accuracy of the point estimates were significantly
+%higher for any parameter values. If so, the distribution of errors was examined
+%to choose a suitable post hoc test. If there appeared to be a correlation
+%between the parameter value and the accuracy of the estimate, Spearman's
+%correlation was calculated. If there were one or more particular values which
+%appeared to have lower error, we used the Wilcoxon rank-sum test.
 
@@
 %possible interaction terms as predictors.
+%
+%\begin{table}
+%  \centering
+%  \input{\tablepath/gamma-expt}
+%  \caption[\gls{BA} parameters used as input \gls{GLM} predicting $\gamma$]
+%  {
+%    \gls{BA} model parameters used as input to \gls{GLM} predicting power law
+%    exponent $\gamma$. One network was simulated with each combination of
+%    parameters, and $\gamma$ was calculated for each network. A \gls{GLM} with
+%    Gamma-distributed errors and a log link function was fit to the $\gamma$
+%    values with all parameters and interaction terms as predictors.
+%  }
+%  \label{tab:gammaexpt}
+%\end{table}
 
-\begin{table}
-  \centering
-  \input{\tablepath/gamma-expt}
-  \caption[\gls{BA} parameters used as input \gls{GLM} predicting $\gamma$]
-  {
-    \gls{BA} model parameters used as input to \gls{GLM} predicting power law
-    exponent $\gamma$. One network was simulated with each combination of
-    parameters, and $\gamma$ was calculated for each network. A \gls{GLM} with
-    Gamma-distributed errors and a log link function was fit to the $\gamma$
-    values with all parameters and interaction terms as predictors.
-  }
-  \label{tab:gammaexpt}
-\end{table}
+%To evaluate the effects of the true parameter values on the accuracy of the
+%\gls{MAP} estimates, we analysed each parameter individually using a \gls{GLM}
+%(four \glspl{GLM} total). The response variable was the error of the point
+%estimate, and the predictor variables were the true values of \gls{alpha},
+%\gls{I}, and \gls{m}. We did not test for differences across true values of
+%\gls{N}, because \gls{N} was not varied in these simulations. The distribution
+%family and link function for each \gls{GLM} were chosen by examination of
+%residual plots and \gls{AIC} of the model fits. For \gls{alpha}, \gls{I}, and
+%\gls{N}, we used the Gamma distribution with the inverse link function. For
+%\gls{m}, we used the Poisson distribution with the log link function. To
+%allow for the possibility that the relationships between parameters and error
+%rates were nonlinear, the true parameter values were treated as discrete rather
+%than continuous predictors. When significant associations were found by the
+%\glspl{GLM}, additional post hoc tests were performed based on the observed
+%relationship between the error rate and true parameter value in question. If
+%the relationship appeared to be monotonic, Spearman's correlation was
+%calculated. Otherwise, if there were particular values which appeared to have a
+%higher error rate than the others, the Wilcoxon rank-sum test was used to check
+%for a significant difference between the two groups.
 

### 2016-05-04 doc/thesis/preface.tex
Moar

@@
 developed by Dr. Poon, but the version used here was implemented by me to
-improve computational efficiency. The idea to apply sequential Monte Carlo is
-credited to Dr. Alexandre Bouchard-C\^ot\'e. Dr. Sarah Otto suggested the
-experiments involving a network with a heterogeneous $\alpha$ parameter and
-peer-driven sampling. Dr. Richard H. Liang provided guidance in the development
-of the Gillespie simulation algorithm. The \software{netabc} program, and all
+improve computational efficiency. The idea to apply sequential Monte Carlo was
+mine, but Dr. Alexandre Bouchard-C\^ot\'e made me aware of the adaptive version
+used in this work. Dr. Sarah Otto suggested the experiments involving a network
+with a heterogeneous $\alpha$ parameter and peer-driven sampling. Dr. Richard
+H. Liang provided guidance in the development of the Gillespie simulation
+algorithm and statistical advice. The \software{netabc} program, and all
 supplementary analysis programs, were written by me.
@@
 oral presentation entitled ``Phylodynamic inference of contact network
-parameters with kernel-ABC'' was given based on chapter 2 at the 23rd HIV
-Dynamics and Evolution meeting on April 25, in Woods Hole, Massachussets, USA.
-A poster based on chapter 2 entitled ``Likelihood-free estimation of contact
-network parameters from viral phylogenies'' was presented at the Intelligent
-Systems for Molecular Biology meeting in Orlando, Florida, USA.
+parameters with kernel-ABC'' was given based on chapter 2 to the 23rd HIV
+Dynamics and Evolution meeting on April 25, in Woods Hole, Massachussets, USA
+(the presentation was delivered remotely). A poster based on chapter 2 entitled
+``Likelihood-free estimation of contact network parameters from viral
+phylogenies'' was presented at the Intelligent Systems for Molecular Biology
+meeting in Orlando, Florida, USA.
+
+Use of the BC data is in accordance with an ethics application that was
+reviewed and approved by the UBC/Providence Health Care Research Ethics Board
+(H07-02559).
 

### 2016-05-04 doc/thesis/results.Rnw
Moar

@@
 <<setup, include=FALSE>>=
-    library(netabc)
-    library(Hmisc)
-
-    pp <- function (p, eq=FALSE) {
-        if (p < 1e-5) {
-            "{<}10^{-5}"
-        } else {
-            n <- latexSN(round(p, -floor(log10(p))))
-            n <- sub("^1\\\\!\\\\times\\\\!", "", n)
-            if (eq) {
-                paste0("=", n)
-            } else {
-                n
-            }
-        }
-    }
+    source("global.R")
 @
 
-\subsubsection*{Classifiers for parameters based on tree shape}
+\subsubsection*{Classifiers for BA model parameters based on tree shape}
 
@@
     d[,error := abs(true_value - point.est)]
+    setkey(d, parameter)
 
@@
     setkey(tip.cor, parameter)
+    stopifnot(tip.cor["alpha", p.value] < 0.05)
+    stopifnot(tip.cor["I", p.value] < 0.05)
+    stopifnot(tip.cor["N", p.value] < 0.05)
+    stopifnot(tip.cor["m", p.value] > 0.05)
 
-    # CI takes up >90% of grid
-    d[parameter == "alpha", all(upper - lower >= 1.8)]
-    d[parameter == "I", all(upper - lower >= 4050)]
-    d[parameter == "N", all(upper - lower >= 12555)]
-    d[parameter == "m", all(upper - lower >= 5)]
+    # CI takes up >75% of grid
+    stopifnot(d["alpha", all(upper - lower >= 0.75*2)])
+    stopifnot(d["I", all(upper - lower >= 0.75*5000)])
+    stopifnot(d["N", all(upper - lower >= 0.75*15000)])
+    stopifnot(d["m", all(upper - lower >= 0.75*6)])
 
     # correlation between value and error
+    av <- d[,list(p=summary(aov(error~factor(true_value)))[[1]][["Pr(>F)"]][1]),
+            by=parameter]
+    stopifnot(av["alpha", p] < 0.05)
+    stopifnot(av["I", p] < 0.05)
+    stopifnot(av["m", p] < 0.05)
+    stopifnot(av["N", p] < 0.05)
+
+    alpha.test <- d["alpha", wilcox.test(.SD[true_value %in% c(1.0, 1.25),error], 
+                                         .SD[!true_value %in% c(1.0, 1.25),error])]
+    stopifnot(alpha.test$p.value < 0.05)
+
+    I.test <- d["I", wilcox.test(.SD[true_value >= 2000 & true_value <= 3000, error], 
+                                 .SD[true_value < 2000 | true_value > 3000, error])]
+    stopifnot(I.test$p.value < 0.05)
+
+    m.test <- d["m", wilcox.test(.SD[true_value == 1, error], 
+                                 .SD[true_value > 1, error])]
+    stopifnot(m.test$p.value < 0.05)
+
+    N.test <- d["N", wilcox.test(.SD[true_value <= 3000, error], 
+                                 .SD[true_value > 3000, error])]
+    stopifnot(N.test$p.value < 0.05)
+
     value.cor <- d[,cor.test(error, true_value)[c("p.value", "estimate")], by=parameter]
@@
 The highest density intervals obtained for all parameters were extremely wide,
-occcupying $>$90\% of the grid in all cases (\cref{fig:gridest}).
+occcupying $>$75\% of the grid in all cases (\cref{fig:gridest}).
 
@@
 estimates having an average deviation of 
-    \Sexpr{d[parameter == "alpha", round(mean(error), 2)]}
-from the true value, on a grid from 0 to 2. The error was negatively correlated
-with the true value of \gls{alpha} 
-    (Spearman's $\rho$ = \Sexpr{value.cor["alpha", round(estimate, 2)]},
-     $p = \Sexpr{value.cor["alpha", pp(p.value)]}$),
-although the relationship was clearly nonlinear (\cref{fig:gridest,fig:gridalpha}).
-The accuracy was highest for the test value \gls{alpha} = 1.25
-    (mean error \Sexpr{d[parameter == "alpha" & true_value == 1.25, round(mean(error), 2)]})
-which exhibited markedly different behaviour than the other values in terms of
-the distribution of kernel scores along the grid (\cref{fig:gridalpha}). In
-particular, there was a very pronounced peak in scores around the true value,
-in contrast to most other values where the scores were flat around the true
-value. The peak was also observed to a lesser degree for \gls{alpha} = 1. The
-average absolute error of the point estimates for \gls{I} was 
-    \Sexpr{d[parameter == "I", round(mean(error))]},
-on a grid of 500 to 5000, and the errors were not significantly correlated with
-the true value of \gls{I}. Kernel score distributions for all test values
-exhibited a similar rounded shape (\cref{fig:gridI}).
+    \Sexpr{d["alpha", round(mean(error), 2)]}
+from the true value, on a grid from 0 to 2. The error of point estimates varied
+significantly between true values of \gls{alpha}
+    (one-way \gls{ANOVA}, $p \Sexpr{pp(av["alpha", p], eq=TRUE)}$). In
+particular, errors were lower for the values \gls{alpha} = 1.0 and 1.25 than
+for the other values
+    (average errors 
+    \Sexpr{d["alpha", .SD[true_value %in% c(1.0, 1.25), round(mean(error), 2)]]}
+    for \gls{alpha} = 1.0 or 1.5 vs.
+    \Sexpr{d["alpha", .SD[!true_value %in% c(1.0, 1.25), round(mean(error), 2)]]}
+    for \gls{alpha} $\neq$ 1.0 or 1.5),
+and this difference was significant
+    (Wilcoxon rank-sum test, $p \Sexpr{pp(alpha.test$p.value, eq=TRUE)}$,
+     \cref{fig:gridptalpha}).
+These two values exhibited different qualitative behaviour than the other
+values in terms of the distribution of kernel scores along the grid
+(\cref{fig:gridalpha}). In particular, there was a pronounced peak in scores
+around the true value, in contrast to the other values where the scores were
+flat around the true value. The effect was most obvious for the value
+\gls{alpha} = 1.25.
+
+The average absolute error of the point estimates for \gls{I} was 
+    \Sexpr{d["I", round(mean(error))]} individuals,
+on a grid of 500 to 5000, and these errors differed between true values of
+\gls{I}
+    (one-way \gls{ANOVA}, $p \Sexpr{pp(av["I", p], eq=TRUE)}$).
+The errors for $2000 \leq I \leq 3000$ were higher than those for the other
+values
+    (average errors
+     \Sexpr{d["I", .SD[true_value >= 2000 & true_value <= 3000, round(mean(error))]]}
+     for $2000 \leq I \leq 3000$ vs.
+     \Sexpr{d["I", .SD[true_value < 2000 | true_value > 3000, round(mean(error))]]}
+     for $I < 2000$ or $I > 3000$),
+and this difference was significant
+    (Wilcoxon rank-sum test, $p \Sexpr{pp(I.test$p.value, eq=TRUE)}$,
+     \cref{fig:gridptI}).
+Kernel score distributions for all test values exhibited a similar rounded
+shape (\cref{fig:gridI}).
 
 The average error for \gls{m} was
-    \Sexpr{d[parameter == "m", round(mean(error), 2)]},
-on a grid from 1 to 6; this error was positively correlated with increasing
-\gls{m}
-    (Spearman's $\rho$ = \Sexpr{value.cor["m", round(estimate, 2)]},
-     $p = \Sexpr{value.cor["m", pp(p.value)]}$).
-This positive correlation was apparently due to the much lower error for
-\gls{m} = 1 and 2 than for the other $m$ values 
-    (mean errors 
-    \Sexpr{d[parameter == "m" & true_value <= 2, round(mean(error), 2)]} 
-    for $m \leq 2$ vs. 
-    \Sexpr{d[parameter == "m" & true_value > 2, round(mean(error), 2)]} 
-    for $m > 2$, \cref{fig:gridm}).
+    \Sexpr{d["m", round(mean(error), 2)]} edges per vertex,
+on a grid from 1 to 6. The error varied significantly between the true values
+of \gls{m} 
+    (one-way \gls{ANOVA}, $p \Sexpr{pp(av["m", p], eq=TRUE)}$).
+Errors for the value \gls{m} = 1 were lower than the other values
+    (average errors
+    \Sexpr{d["m", .SD[true_value == 1, round(mean(error), 2)]]}
+    for $m = 1$ vs.
+    \Sexpr{d["m", .SD[true_value > 1, round(mean(error), 2)]]}
+    for $m > 1$),
+and this difference was significant
+    (Wilcoxon rank-sum test, $p \Sexpr{pp(m.test$p.value, eq=TRUE)}$,
+     \cref{fig:gridptm}).
 The value $m = 1$ causes the network to take on a distinct shape relative to
-higher \gls{m} values, namely a tree (\textit{i.e.} there are no cycles,
-see \cref{subsec:treeshape}). The average error for \gls{N} was 
-    \Sexpr{d[parameter == "N", round(mean(error))]},
-on a grid from 1000 to 15000, and was positively correlated with the true value
+higher \gls{m} values, namely a tree (\ie there are no cycles, see
+\cref{subsec:treeshape}). The kernel score distribution had a peak at $m = 1$
+when this was the true value, and a valley at $m = 1$ when the true value of
+$m$ was greater that 1 (\cref{fig:gridm}).
+
+The average error for \gls{N} was 
+    \Sexpr{d[parameter == "N", round(mean(error))]} individuals,
+on a grid from 1000 to 15000, and was varied significantly with the true value
 of \gls{N}
-    (Spearman's $\rho$ = \Sexpr{value.cor["N", round(estimate, 2)]},
-    $p\ \Sexpr{value.cor["N", pp(p.value)]}$). 
-Most of the kernel score distributions were extremely flat, except for the
-value $N = 1000$ which exhibited a pronounced peak around the true value
-(\cref{fig:gridN}).
+    (one-way \gls{ANOVA}, $p \Sexpr{pp(av["N", p], eq=TRUE)}$).
+The errors were lower for $N \leq 3000$
+    (average errors
+     \Sexpr{d["N", .SD[true_value <= 3000, round(mean(error))]]}
+     for $N \leq 3000$ vs.
+     \Sexpr{d["N", .SD[true_value > 3000, round(mean(error))]]}
+     for $N > 3000$),
+and this diffrence was significant
+    (Wilcoxon rank-sum test, $p \Sexpr{pp(N.test$p.value, eq=TRUE)}$,
+     \cref{fig:gridptN}).
+The kernel score distribution had a peak at $N = 1000$ when this was the true
+value, and a valley there otherwise (\cref{fig:gridN}). Except for this valley,
+the distributions were flat for $N > 3000$.
 
@@
     I.av <- anova(lm(I_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-    stopifnot(I.av["factor(true_m)", "Pr(>F)"] >= 0.05)
-    stopifnot(m.av["factor(true_m)", "Pr(>F)"] >= 0.05)
-    stopifnot(m.av["factor(true_I)", "Pr(>F)"] >= 0.05)
-    stopifnot(min(na.omit(N.av[,"Pr(>F)"])) >= 0.05)
 
@@
                                 d[true_alpha == 0.5 | true_alpha == 1.5, m_error])
-    options(scipen=-1, digits=2)
+    stopifnot(alpha.av["factor(true_m)", "Pr(>F)"] <= 0.05)
+    stopifnot(alpha.av["factor(true_I)", "Pr(>F)"] > 0.05)
+    stopifnot(I.I.test$p.value < 1e-5)
+    stopifnot(I.av["factor(true_m)", "Pr(>F)"] > 0.05)
+    stopifnot(m.av["factor(true_m)", "Pr(>F)"] > 0.05)
+    stopifnot(m.av["factor(true_I)", "Pr(>F)"] > 0.05)
+    stopifnot(m.av["factor(true_alpha)", "Pr(>F)"] > 0.05)
+    stopifnot(min(N.av[1:3,"Pr(>F)"]) > 0.05)
+
+    alpha.m.cor <- cor.test(d[,alpha_error], d[,true_m], method="spearman")
 @
 
-We used \software{netabc} to estimate the parameters of the \gls{BA} model on
-simulated trees where the true parameter values were known. Point estimates for
-each parameter are shown in \cref{fig:abcptm2} for the simulations with \gls{m} =
-2. The results for the other values of \gls{m} were similar
-(\cref{fig:abcptm3,fig:abcptm4}). The median [IQR] absolute error of estimates of
-\gls{alpha} across all simulations was
-    \Sexpr{d[,median(alpha_error)]} 
-    [\Sexpr{d[,quantile(alpha_error, 0.25)]}-\Sexpr{d[,quantile(alpha_error, 0.75)]}].
-The accuracy of the estimates was not significantly different between values of
-$m$ or $I$ (both one-way ANOVA,
-    $p$ = \Sexpr{alpha.av["factor(true_m)", "Pr(>F)"]}
-and 
-    \Sexpr{alpha.av["factor(true_I)", "Pr(>F)"]}),
-although the errors when the true value of \gls{alpha} was zero were
-significantly greater than the other values 
-    (Wilcoxon rank-sum test, $p = \Sexpr{pp(zero.alpha.test$p.value)}$).
-The error in the estimated value of $I$ was
+\Cref{fig:abcpt} shows \gls{MAP} point estimates of the BA model parameters
+obtained with kernel-ABC on simulated data. The estimates shown correspond only
+to the simulations where the $m$ parameter was set to 2, however the results
+for $m = 3$ and $m = 4$ were similar (\cref{fig:abcptm3,fig:abcptm4}). Average
+boundaries of 95\% HPD intervals are given in \cref{tab:abchpd}.
+
+The accuracy of the parameter estimates obtained with kernel-ABC
+paralleled the results from the \gls{kSVR} classifier. Of the four parameters,
+$\alpha$ was the most accurately estimated, with point estimates having a
+median [IQR] absolute error of 
+    \Sexpr{d[,round(median(alpha_error), 2)]} 
+    [\Sexpr{d[,round(quantile(alpha_error, 0.25), 2)]} - 
+    \Sexpr{d[,round(quantile(alpha_error, 0.75), 2)]}].
+The errors when the true value of $\alpha$ was zero were significantly greater
+than those for the other values 
+    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(zero.alpha.test$p.value, 4))}$).
+Errors in estimating $\alpha$ also varied with the true value of $m$ just at
+the threshold of statistical significance
+    ($p \Sexpr{pp(alpha.av["factor(true_m)", "Pr(>F)"])}$),
+but did not vary across the true values of $N$ or $I$ (both one-way ANOVA).
+Estimates for $I$ were relatively accurate, with point estimate errors of
     \Sexpr{d[,round(median(I_error))]} 
-    [\Sexpr{d[,round(quantile(I_error, 0.25))]}-\Sexpr{d[,round(quantile(I_error, 0.75))]}].
-Errors were significantly higher for $\alpha \geq 1$
-    (Wilcoxon rank-sum test, $p = \Sexpr{pp(alpha.I.test$p.value)}$)
-and for $I$ = 2000
-    ($p = \Sexpr{pp(I.I.test$p.value)}$),
-but not for any values of $m$ (one-way ANOVA). The $m$ parameter was estimated
-correctly in only
+    [\Sexpr{d[,round(quantile(I_error, 0.25))]} - 
+    \Sexpr{d[,round(quantile(I_error, 0.75))]}] individuals.
+These errors were significantly higher when the true value of $\alpha$ was
+at least 1
+    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(alpha.I.test$p.value, 4))}$)
+and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
+did not affect the estimates of $I$ (one-way ANOVA).
+
+The $m$ parameter was estimated correctly in only
     \Sexpr{as.integer(m.tbl[1] * 100)} \%
-of simulations, little better than random guessing. The true values of $m$ and
-$I$ did not significantly affect the error (one-way ANOVA). Finally, the total
-number of nodes \gls{N} was consistently over-estimated by about a factor of
-two
-    (error \Sexpr{d[,format(median(N_error), scientific=FALSE)]} 
-    [\Sexpr{d[,format(quantile(N_error, 0.25), scientific=FALSE)]} - 
-     \Sexpr{d[,format(quantile(N_error, 0.75), scientific=FALSE)]}]).
-No other parameters influenced the accuracy of the $N$ estimates (one-way
-ANOVA).
+of simulations, barely better than random guessing. The true values of the
+other parameters did not significantly affect the estimates of $m$ (both
+one-way ANOVA). Finally, the total number of nodes $N$ was consistently
+over-estimated by about a factor of two
+    (error \Sexpr{d[,round(median(N_error))]} 
+    [\Sexpr{d[,round(quantile(N_error, 0.25))]} - 
+     \Sexpr{d[,round(quantile(N_error, 0.75))]}] individuals).
+No parameters influenced the accuracy of the $N$ estimates (all one-way ANOVA).
 
-\begin{figure}
-  \includegraphics{abc-point-estimate-m2}
-  \caption[\Acrlong{MAP} point estimates for \gls{BA} model parameters obtained
-    by running \software{netabc} on simulated data, for simulations with $m = 2$.] 
-  {
-    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by         
-    running \software{netabc} on simulated data. Values shown are for               
-    simulations with \gls{m} = 2. Dashed lines indicate true values. (A)            
-    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations  
-    against known values. (B) Estimates of \gls{m} and \gls{N} which were held   
-    fixed in these simulations at the values \gls{m} = 2 and \gls{N} = 5000. 
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth]{abc-point-estimate-m2}
+  \vspace{6pt}
+  \caption{
+    Point estimates of BA model parameters obtained by running kernel-ABC
+    on simulated phylogenies without training, for simulations with $m = 2$.
+    Dotted lines indicate true values, and limits of the $y$-axes are regions
+    of uniform prior density. (A) Estimates for $\alpha$ and $I$ against their
+    true values in simulations. (B) Estimates for $m$ and $N$, which were held
+    fixed in these simulations, against true values of $\alpha$.
   }
-  \label{fig:abcptm2}
+  \label{fig:abcpt}
 \end{figure}
 
-<<posterior_sims, echo=FALSE>>=
+\begin{table*}[ht]
+  \centering
+  \input{\tablepath/abc-hpd.tex}
+  \caption{
+      Average maximum \textit{a posteriori} point estimates and 95\% highest
+      posterior density (HPD) interval widths for BA model parameter estimates
+      obtained with kernel-ABC. Three transmission trees were simulated under
+      each combination of the listed parameter values, and the parameters were
+      estimated with kernel-ABC without training.
+  }
+  \label{tab:abchpd}
+\end{table*}
+
+<<supp_labels, echo=FALSE, results="asis">>=
     N <- 5000
@@
     }
-    lab <- paste(lab, collapse=",")
+    lab <- paste(lab, collapse = ",")
 @
 
-\Cref{fig:abcex} shows the \gls{ABC} approximation to the posterior
-distribution on the \gls{BA} parameters for one simulation. Equivalent plots
-for one replicate simulation with each combination of parameters can be found
-in \cref{\Sexpr{lab}}. \Gls{HPD} intervals around \gls{alpha} and \gls{I} were
-narrow relative to the region of nonzero prior density, whereas the intervals
-for $m$ and \gls{N} were widely dispersed. \Cref{tab:abchpd} shows point
-estimates and 95\% \gls{HPD} intervals averaged over all simulations.
+The dispersion of the ABC approximation to the posterior also varied between
+the parameters, with narrower HPD intervals for the parameters with more
+accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows
+the distributions for for one simulation. Equivalent plots for one replicate
+simulation with each studied parameter combination can be found in
+\cref{\Sexpr{lab}}. HPD intervals around $\alpha$ and $I$ were often narrow
+relative to the region of nonzero prior density, whereas the intervals for $m$
+and $N$ were more widely dispersed.
 
-\begin{figure}
-  \includegraphics{{abc-posterior/1.0_1000_2_5000_0}.pdf}
-  \caption[Approximate marginal posterior distributions of BA model parameters
-      obtained by applying \textit{netabc} to a simulated transmission tree
-      with values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]
-    {
-        Approximate marginal posterior distributions of BA model parameters
-        obtained by applying \textit{netabc} to a simulated transmission tree
-        with BA parameter values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ =
-        5000. Vertical dashed lines indicate true values. Shaded areas are 95\%
-        highest posterior density intervals. $x$-axes indicate regions of
-        nonzero prior density.
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{abc-posterior-example}
+  \vspace{6pt}
+  \caption{
+    Marginal posterior distributions of BA model parameters estimated
+    with kernel-ABC for a single simulated transmission tree. Dotted
+    lines and shaded polygon indicate true values.
   }
@@
 \end{figure}
-
-\begin{table}
-    \centering
-    \input{\tablepath/abc-hpd}
-    \caption[
-        Maximum \textit{a priori} estimates and 95\% highest posterior density
-        (HPD) interval boundaries for \gls{BA} model parameters estimated with
-        \software{netabc}, averaged over simulated transmission trees.
-    ]
-    {
-        Maximum \textit{a priori} estimates and 95\% highest posterior density
-        (HPD) interval boundaries for \gls{BA} model parameters estimated with
-        \software{netabc}, averaged over simulated transmission trees.
-    }
-    \label{tab:abchpd}
-\end{table}
 
@@
   \Sexpr{d["N", round(map)]} 
-  [\Sexpr{d["N", round(hpd.min)]} -
+  [\Sexpr{d["N", round(hpd.min)]}-
    \Sexpr{d["N", round(hpd.max)]}].
@@
 $\alpha$ was somewhat lower than typical.
+
+%<<abc_glm, include=FALSE>>=
+%    source("global.R")
+%    options(scipen=-1, digits=2)
+%@
+%
+%<<alpha_glm, include=FALSE>>=
+%    # alpha_error is influenced by alpha and m but not I
+%    stopifnot(alpha.glm[Parameter == "alpha", min(p) < 0.05])
+%    stopifnot(alpha.glm[Parameter == "m", min(p) < 0.05])
+%    stopifnot(alpha.glm[Parameter == "I", min(p) > 0.05])
+%
+%    # alpha_error is correlated with true_alpha
+%    alpha.test <- d[,cor.test(alpha_error, true_alpha, method="spearman")]
+%    stopifnot(alpha.test$p.value < 0.05)
+%
+%    # alpha_error is not correlated with true_m
+%    alpha.m.cor <- d[,cor.test(alpha_error, true_m)]
+%    stopifnot(alpha.m.cor$p.value < 0.05)
+%@
+%
+%We used \software{netabc} to estimate the parameters of the \gls{BA} model on
+%simulated trees where the true parameter values were known. Point estimates for
+%each parameter are shown in \cref{fig:abcptm2} for the simulations with \gls{m} =
+%2. The results for the other values of \gls{m} were similar
+%(\cref{fig:abcptm3,fig:abcptm4}). The median [IQR] absolute error of estimates
+%of \gls{alpha} across all simulations was
+%    \Sexpr{d[,median(alpha_error)]} 
+%    [\Sexpr{d[,quantile(alpha_error, 0.25)]}-\Sexpr{d[,quantile(alpha_error, 0.75)]}].
+%\Gls{GLM} analysis indicated that the true values of both \gls{alpha} and
+%\gls{m} had significant effects on the error in estimated \gls{alpha}
+%    ($p$ values $\Sexpr{pp(alpha.glm[Parameter == "alpha", min(p)], eq=FALSE)}$
+%     $\Sexpr{pp(alpha.glm[Parameter == "m", min(p)], eq=FALSE)}$,
+%     \cref{tab:glmalpha}),
+%but the true value of \gls{I} did not. There was a significant negative
+%correlation between the true value of \gls{alpha} and the error
+%    (Spearman's $\rho$ = \Sexpr{round(alpha.test$estimate, 2)},
+%     $p \Sexpr{pp(alpha.test$p.value, eq=TRUE)}$).
+%There was a negative correlation between \gls{m} and the error in \gls{alpha}
+%    (Spearman's $\rho$ = \Sexpr{round(alpha.m.cor$estimate, 2)},
+%    $p \Sexpr{pp(alpha.m.cor$p.value, eq=FALSE)}$).
+%
+%<<I_glm, include=FALSE>>=
+%    # I_error is influenced by alpha and I but not by m
+%    stopifnot(I.glm[Parameter == "alpha", min(p) < 0.05])
+%    stopifnot(I.glm[Parameter == "m", min(p) > 0.05])
+%    stopifnot(I.glm[Parameter == "I", min(p) < 0.05])
+%
+%    # correlation between I_error and true_alpha
+%    I.alpha.cor <- d[,cor.test(I_error, true_alpha, method="spearman")]
+%    stopifnot(I.alpha.cor$p.value < 0.05)
+%
+%    # correlation between I_error and true_I
+%    I.I.cor <- d[,cor.test(I_error, true_I, method="spearman")]
+%    stopifnot(I.alpha.cor$p.value < 0.05)
+%@
+%
+%The mean error in the estimated value of \gls{I} was
+%    \Sexpr{d[,round(median(I_error))]} 
+%    [\Sexpr{d[,round(quantile(I_error, 0.25))]}-\Sexpr{d[,round(quantile(I_error, 0.75))]}],
+%an over-estimate of roughly a factor of 1.5
+%(\cref{fig:abcptm2,fig:abcptm1,fig:abcptm3,fig:abcptm4}). \Gls{GLM} analysis
+%indicated a relationship between the error in estimated \gls{I}, and the true
+%values of \gls{alpha} and \gls{I}
+%    ($p$ values $\Sexpr{pp(I.glm[Parameter == "alpha", min(p)], eq=FALSE)}$
+%     $\Sexpr{pp(alpha.glm[Parameter == "I", min(p)], eq=FALSE)}$,
+%     \cref{tab:glmalpha}),
+%but not the true value of \gls{m}. There was a significant correlation between 
+%the true value of \gls{alpha} and the error in estimated \gls{I}
+%    (Spearman's $\rho$ = \Sexpr{I.alpha.cor$estimate},
+%     $p \Sexpr{pp(I.alpha.cor$p.value, eq=TRUE)}$),
+%and between the true value of \gls{I} and the error in estimated \gls{I}
+%    (Spearman's $\rho$ = \Sexpr{I.I.cor$estimate},
+%     $p \Sexpr{pp(I.I.cor$p.value, eq=TRUE)}$).
+%
+%<<m_glm, include=FALSE>>=
+%    # effects of parameters on m_error
+%    stopifnot(m.glm[Parameter == "m", min(p)] < 0.05)
+%    stopifnot(m.glm[Parameter == "alpha", min(p)] > 0.05)
+%    stopifnot(m.glm[Parameter == "I", min(p)] > 0.05)
+%
+%    # m was only estimated correctly in just over 20\% of simulations with m > 2
+%    m.tbl <- prop.table(table(d[true_m > 1,m_error]))
+%    stopifnot(m.tbl[1] > 0.2 & m.tbl[1] < 0.3)
+%
+%    # with m = 1, over 95% of simulations are right
+%    m1.est <- d[true_m == 1, m_error]
+%    stopifnot(sum(m1.est == 0) / length(m1.est) > 0.95)
+%    stopifnot(sum(m1.est == 0) / length(m1.est) < 1)
+%@
+%
+%\gls{GLM} analysis showed an effect of the true value of \gls{m} on the error
+%in the estimated \gls{m}
+%    ($p \Sexpr{pp(m.glm[Parameter == "m", min(p)])}$,
+%     \cref{tab:glmm}).
+%When the true value of \gls{m} was 1, the correct value was recovered by
+%\software{netabc} in virtually every case
+%    (\Sexpr{sum(m1.est == 0)} out of \Sexpr{length(m1.est)} simulations).
+%However, when the true value of \gls{m} was 2 or higher, the correct value was
+%recovered in only 
+%    \Sexpr{as.integer(m.tbl[1] * 100)}\%
+%of simulations, little better than random guessing. The \gls{GLM} analysis did
+%not indicate any effects of the true parameter values on the error in estimated
+%\gls{m} (\cref{tab:glmm}).
+%
+%<<N_glm, include=FALSE>>=
+%    # alpha_error is influenced by alpha and m but not I
+%    stopifnot(N.glm[Parameter == "alpha", min(p) < 0.05])
+%    stopifnot(N.glm[Parameter == "I", min(p) < 0.05])
+%    stopifnot(N.glm[Parameter == "m", min(p) > 0.05])
+%
+%    N.alpha.test <- d[,wilcox.test(.SD[true_alpha == 1.5, N_error], 
+%                                   .SD[true_alpha < 1.5, N_error])]
+%    stopifnot(N.alpha.test$p.value < 0.05)
+%    N.I.test <- d[,wilcox.test(.SD[true_I == 3000, N_error], 
+%                               .SD[true_I < 3000, N_error])]
+%    stopifnot(N.I.test$p.value < 0.05)
+%@
+%
+%Finally, the total number of nodes \gls{N} was consistently over-estimated by
+%about a factor of two
+%    (error \Sexpr{d[,format(median(N_error), scientific=FALSE)]} 
+%    [\Sexpr{d[,format(quantile(N_error, 0.25), scientific=FALSE)]} - 
+%     \Sexpr{d[,format(quantile(N_error, 0.75), scientific=FALSE)]}]).
+%The fitted \gls{GLM} indicated that the true values of both \gls{alpha} and
+%\gls{I} had an effect on the error in the estimated \gls{N}
+%    ($p$-values $\Sexpr{pp(N.glm[Parameter == "alpha", min(p)], eq=FALSE)}$ and 
+%     $\Sexpr{pp(N.glm[Parameter == "I", min(p)], eq=FALSE)}$,
+%     \cref{tab:glmN}),
+%but that the true value of \gls{m} did not. The error in the estimated \gls{N}
+%when \gls{alpha} was equal to 1.5 was slightly lower than for other values of
+%\gls{alpha}
+%    (median [IQR] error rates 
+%     \Sexpr{d[true_alpha == 1.5, format(median(N_error), scientific=FALSE)]} 
+%    [\Sexpr{d[true_alpha == 1.5, format(quantile(N_error, 0.25), scientific=FALSE)]} - 
+%     \Sexpr{d[true_alpha == 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]}]
+%     for \gls{alpha} = 1.5 vs. 
+%     \Sexpr{d[true_alpha < 1.5, format(median(N_error), scientific=FALSE)]} 
+%    [\Sexpr{d[true_alpha < 1.5, format(quantile(N_error, 0.25), scientific=FALSE)]} - 
+%     \Sexpr{d[true_alpha < 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]}]
+%     for \gls{alpha} < 1.5),
+%and this difference was statistically significant
+%    (Wilcoxon rank-sum test, $p \Sexpr{pp(N.alpha.test$p.value, eq=TRUE)}$).
+%Similarly, the error in the estimated \gls{N} when \gls{I} was equal to 3000 was
+%lower than for other values of \gls{I}
+%    (median [IQR] error rates 
+%     \Sexpr{d[true_I == 3000, format(median(N_error), scientific=FALSE)]} 
+%    [\Sexpr{d[true_I == 3000, format(quantile(N_error, 0.25), scientific=FALSE)]} - 
+%     \Sexpr{d[true_I == 3000, format(quantile(N_error, 0.75), scientific=FALSE)]}]
+%     for \gls{I} = 3000 vs. 
+%     \Sexpr{d[true_alpha < 3000, format(median(N_error), scientific=FALSE)]} 
+%    [\Sexpr{d[true_alpha < 3000, format(quantile(N_error, 0.25), scientific=FALSE)]} - 
+%     \Sexpr{d[true_alpha < 3000, format(quantile(N_error, 0.75), scientific=FALSE)]}]
+%     for \gls{I} < 3000),
+%and this difference was statistically significant
+%    (Wilcoxon rank-sum test, $p \Sexpr{pp(N.I.test$p.value, eq=TRUE)}$).
+%
+%\begin{figure}
+%  \includegraphics{abc-point-estimate-m2}
+%  \caption[\Acrlong{MAP} point estimates for \gls{BA} model parameters obtained
+%    by running \software{netabc} on simulated data, for simulations with $m = 2$.] 
+%  {
+%    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by         
+%    running \software{netabc} on simulated data. Values shown are for               
+%    simulations with \gls{m} = 2. Dashed lines indicate true values. (A)            
+%    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations  
+%    against known values. (B) Estimates of \gls{m} and \gls{N} which were held   
+%    fixed in these simulations at the values \gls{m} = 2 and \gls{N} = 5000. 
+%  }
+%  \label{fig:abcptm2}
+%\end{figure}
+%
+%<<posterior_sims, echo=FALSE>>=
+%    N <- 5000
+%    replicate <- 0
+%    lab <- NULL
+%    for (m in c(2, 3, 4)) {
+%    for (I in c(1000, 2000)) {
+%    for (alpha in c(0, 0.5, 1, 1.5)) {
+%        lab <- c(lab, sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate))
+%    }
+%    }
+%    }
+%    lab <- paste(lab, collapse=",")
+%@
+%
+%\Cref{fig:abcex} shows the \gls{ABC} approximation to the posterior
+%distribution on the \gls{BA} parameters for one simulation. Equivalent plots
+%for one replicate simulation with each combination of parameters can be found
+%in \cref{\Sexpr{lab}}. \Gls{HPD} intervals around \gls{alpha} and \gls{I} were
+%narrow relative to the region of nonzero prior density, whereas the intervals
+%for $m$ and \gls{N} were widely dispersed. \Cref{tab:abchpd} shows point
+%estimates and 95\% \gls{HPD} intervals averaged over all simulations.
+%
+%\begin{figure}
+%  \includegraphics{{abc-posterior/1.0_1000_2_5000_0}.pdf}
+%  \caption[Approximate marginal posterior distributions of BA model parameters
+%      obtained by applying \textit{netabc} to a simulated transmission tree
+%      with values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]
+%    {
+%        Approximate marginal posterior distributions of BA model parameters
+%        obtained by applying \textit{netabc} to a simulated transmission tree
+%        with BA parameter values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ =
+%        5000. Vertical dashed lines indicate true values. Shaded areas are 95\%
+%        highest posterior density intervals. $x$-axes indicate regions of
+%        nonzero prior density.
+%    }
+%  \label{fig:abcex}
+%\end{figure}
+%
+%
+%\begin{table}
+%    \centering
+%    \input{\tablepath/abc-hpd}
+%    \caption[
+%        Maximum \textit{a priori} estimates and 95\% highest posterior density
+%        (HPD) interval boundaries for \gls{BA} model parameters estimated with
+%        \software{netabc}, averaged over simulated transmission trees.
+%    ]
+%    {
+%        Maximum \textit{a priori} estimates and 95\% highest posterior density
+%        (HPD) interval boundaries for \gls{BA} model parameters estimated with
+%        \software{netabc}, averaged over simulated transmission trees.
+%    }
+%    \label{tab:abchpd}
+%\end{table}
+%
+%<<mixed_peerdriven, include=FALSE>>=
+%  library(coda)
+%  d <- fread("bzcat ../../simulations/abc-pa-mixed-alpha/abc/*")
+%  d <- d[iter == max(iter)]
+%  d <- d[sample(1:nrow(d), prob=weight, replace=TRUE)]
+%  params <- c("alpha", "m", "N", "I")
+%  d <- melt(d, measure.vars=params, variable.name="parameter")
+%  f <- function (x) {
+%    dens <- density(x)
+%    hpd <- HPDinterval(mcmc(x))
+%    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
+%         hpd.max=hpd[,"upper"])
+%  }
+%  d <- d[,f(value), by=parameter]
+%  setkey(d, parameter)
+%
+%  m <- collect.metadata("../../simulations/abc-pa-peerdriven/abc/*")
+%  p <- fread(paste("bzcat", rownames(m)[m$sample_peer == 1]))
+%  p <- p[iter == max(iter)]
+%  p <- p[sample(1:nrow(p), prob=weight, replace=TRUE)]
+%  params <- c("alpha", "m", "N", "I")
+%  p <- melt(p, measure.vars=params, variable.name="parameter")
+%  f <- function (x) {
+%    dens <- density(x)
+%    hpd <- HPDinterval(mcmc(x))
+%    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
+%         hpd.max=hpd[,"upper"])
+%  }
+%  p <- p[,f(value), by=parameter]
+%  setkey(p, parameter)
+%  options(scipen=5)
+%@
+%
+%To test the effect of model misspecification, we simulated one network where
+%the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
+%other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
+%HPD] estimates for each parameter were: 
+%$\alpha$, 
+%  \Sexpr{round(d["alpha", map], 2)} 
+%  [\Sexpr{round(d["alpha", hpd.min], 2)} -
+%   \Sexpr{round(d["alpha", hpd.max], 2)}];
+%$I$,
+%  \Sexpr{d["I", round(map)]} 
+%  [\Sexpr{d["I", round(hpd.min)]} -
+%   \Sexpr{d["I", round(hpd.max)]}];
+%$m$,
+%  \Sexpr{d["m", floor(map)]} 
+%  [\Sexpr{d["m", floor(hpd.min)]} -
+%   \Sexpr{d["m", floor(hpd.max)]}];
+%$N$,
+%  \Sexpr{d["N", round(map)]} 
+%  [\Sexpr{d["N", round(hpd.min)]} -
+%   \Sexpr{d["N", round(hpd.max)]}].
+%The approximate posterior distributions for this simulation are shown in
+%\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
+%transmission tree in a peer-driven fashion, where the probability to sample a
+%node was twice as high if one of its peers had already been sampled. The
+%parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
+%$I$ = 2000. The estimated values were
+%$\alpha$, 
+%  \Sexpr{p["alpha", round(map, 2)]} 
+%  [\Sexpr{p["alpha", round(hpd.min, 2)]} -
+%   \Sexpr{p["alpha", round(hpd.max, 2)]}];
+%$I$,
+%  \Sexpr{p["I", floor(map)]} 
+%  [\Sexpr{p["I", floor(hpd.min)]} -
+%   \Sexpr{p["I", floor(hpd.max)]}];
+%$m$,
+%  \Sexpr{p["m", floor(map)]} 
+%  [\Sexpr{p["m", floor(hpd.min)]} -
+%   \Sexpr{p["m", floor(hpd.max)]}];
+%$N$,
+%  \Sexpr{p["N", floor(map)]} 
+%  [\Sexpr{p["N", floor(hpd.min)]} -
+%   \Sexpr{p["N", floor(hpd.max)]}].
+%The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
+%of these results were in line with estimates obtained on other simulated
+%datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
+%$\alpha$ was somewhat lower than typical.
 
@@
 in the literature~\autocite{liljeros2001web, schneeberger2004scale,
-colgate1989risk, brown2011transmission}. Higher estimated values of $\alpha$
-translated into higher estimated values of $\gamma$, which is consistent with
-the observed relationship between these two quantities (\cref{fig:gamma}).
+colgate1989risk, brown2011transmission}.
 
 \begin{table}
+    \centering
+    \input{\tablepath/realdata-gamma.tex}
     \caption[

### 2016-05-04 doc/thesis/results.tex
Moar

@@
 
-\subsubsection*{Classifiers for parameters based on tree shape}
+\subsubsection*{Classifiers for BA model parameters based on tree shape}
 
@@
     0.14
-from the true value, on a grid from 0 to 2. The error was negatively correlated
-with the true value of \gls{alpha} 
-    (Spearman's $\rho$ = \ensuremath{-0.26},
-     $p = 10^{-5}$),
-although the relationship was clearly nonlinear (\cref{fig:gridest,fig:gridalpha}).
-The accuracy was highest for the test value \gls{alpha} = 1.25
-    (mean error 0.02)
-which exhibited markedly different behaviour than the other values in terms of
-the distribution of kernel scores along the grid (\cref{fig:gridalpha}). In
-particular, there was a very pronounced peak in scores around the true value,
-in contrast to most other values where the scores were flat around the true
-value. The peak was also observed to a lesser degree for \gls{alpha} = 1. The
-average absolute error of the point estimates for \gls{I} was 
-    310,
-on a grid of 500 to 5000, and the errors were not significantly correlated with
-the true value of \gls{I}. Kernel score distributions for all test values
-exhibited a similar rounded shape (\cref{fig:gridI}).
+from the true value, on a grid from 0 to 2. The error of point estimates varied
+significantly between true values of \gls{alpha}
+    (one-way \gls{ANOVA}, $p {<}10^{-5}$). In
+particular, errors were lower for the values \gls{alpha} = 1.0 and 1.25 than
+for the other values
+    (average errors 
+    0.03
+    for \gls{alpha} = 1.0 or 1.5 vs.
+    0.17
+    for \gls{alpha} $\neq$ 1.0 or 1.5),
+and this difference was significant
+    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
+     \cref{fig:gridptalpha}).
+These two values exhibited different qualitative behaviour than the other
+values in terms of the distribution of kernel scores along the grid
+(\cref{fig:gridalpha}). In particular, there was a pronounced peak in scores
+around the true value, in contrast to the other values where the scores were
+flat around the true value. The effect was most obvious for the value
+\gls{alpha} = 1.25.
+
+The average absolute error of the point estimates for \gls{I} was 
+    310 individuals,
+on a grid of 500 to 5000, and these errors differed between true values of
+\gls{I}
+    (one-way \gls{ANOVA}, $p =0.001$).
+The errors for $2000 \leq I \leq 3000$ were higher than those for the other
+values
+    (average errors
+     430
+     for $2000 \leq I \leq 3000$ vs.
+     250
+     for $I < 2000$ or $I > 3000$),
+and this difference was significant
+    (Wilcoxon rank-sum test, $p =6\!\times\!10^{-4}$,
+     \cref{fig:gridptI}).
+Kernel score distributions for all test values exhibited a similar rounded
+shape (\cref{fig:gridI}).
 
 The average error for \gls{m} was
-    1.31,
-on a grid from 1 to 6; this error was positively correlated with increasing
-\gls{m}
-    (Spearman's $\rho$ = 0.25,
-     $p = 6\!\times\!10^{-4}$).
-This positive correlation was apparently due to the much lower error for
-\gls{m} = 1 and 2 than for the other $m$ values 
-    (mean errors 
-    0.93 
-    for $m \leq 2$ vs. 
-    1.49 
-    for $m > 2$, \cref{fig:gridm}).
+    1.31 edges per vertex,
+on a grid from 1 to 6. The error varied significantly between the true values
+of \gls{m} 
+    (one-way \gls{ANOVA}, $p {<}10^{-5}$).
+Errors for the value \gls{m} = 1 were lower than the other values
+    (average errors
+    0.1
+    for $m = 1$ vs.
+    1.55
+    for $m > 1$),
+and this difference was significant
+    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
+     \cref{fig:gridptm}).
 The value $m = 1$ causes the network to take on a distinct shape relative to
-higher \gls{m} values, namely a tree (\textit{i.e.} there are no cycles,
-see \cref{subsec:treeshape}). The average error for \gls{N} was 
-    2419,
-on a grid from 1000 to 15000, and was positively correlated with the true value
+higher \gls{m} values, namely a tree (\ie there are no cycles, see
+\cref{subsec:treeshape}). The kernel score distribution had a peak at $m = 1$
+when this was the true value, and a valley at $m = 1$ when the true value of
+$m$ was greater that 1 (\cref{fig:gridm}).
+
+The average error for \gls{N} was 
+    2419 individuals,
+on a grid from 1000 to 15000, and was varied significantly with the true value
 of \gls{N}
-    (Spearman's $\rho$ = 0.43,
-    $p\ {<}10^{-5}$). 
-Most of the kernel score distributions were extremely flat, except for the
-value $N = 1000$ which exhibited a pronounced peak around the true value
-(\cref{fig:gridN}).
+    (one-way \gls{ANOVA}, $p {<}10^{-5}$).
+The errors were lower for $N \leq 3000$
+    (average errors
+     740
+     for $N \leq 3000$ vs.
+     2979
+     for $N > 3000$),
+and this diffrence was significant
+    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
+     \cref{fig:gridptN}).
+The kernel score distribution had a peak at $N = 1000$ when this was the true
+value, and a valley there otherwise (\cref{fig:gridN}). Except for this valley,
+the distributions were flat for $N > 3000$.
 
@@
 
-We used \software{netabc} to estimate the parameters of the \gls{BA} model on
-simulated trees where the true parameter values were known. Point estimates for
-each parameter are shown in \cref{fig:abcptm2} for the simulations with \gls{m} =
-2. The results for the other values of \gls{m} were similar
-(\cref{fig:abcptm3,fig:abcptm4}). The median [IQR] absolute error of estimates of
-\gls{alpha} across all simulations was
+\Cref{fig:abcpt} shows \gls{MAP} point estimates of the BA model parameters
+obtained with kernel-ABC on simulated data. The estimates shown correspond only
+to the simulations where the $m$ parameter was set to 2, however the results
+for $m = 3$ and $m = 4$ were similar (\cref{fig:abcptm3,fig:abcptm4}). Average
+boundaries of 95\% HPD intervals are given in \cref{tab:abchpd}.
+
+The accuracy of the parameter estimates obtained with kernel-ABC
+paralleled the results from the \gls{kSVR} classifier. Of the four parameters,
+$\alpha$ was the most accurately estimated, with point estimates having a
+median [IQR] absolute error of 
     0.08 
-    [0.05-0.17].
-The accuracy of the estimates was not significantly different between values of
-$m$ or $I$ (both one-way ANOVA,
-    $p$ = 0.05
-and 
-    0.07),
-although the errors when the true value of \gls{alpha} was zero were
-significantly greater than the other values 
-    (Wilcoxon rank-sum test, $p = 8\!\times\!10^{-3}$).
-The error in the estimated value of $I$ was
+    [0.05 - 
+    0.17].
+The errors when the true value of $\alpha$ was zero were significantly greater
+than those for the other values 
+    (Wilcoxon rank-sum test, $p$ = $0.0078$).
+Errors in estimating $\alpha$ also varied with the true value of $m$ just at
+the threshold of statistical significance
+    ($p 0.05$),
+but did not vary across the true values of $N$ or $I$ (both one-way ANOVA).
+Estimates for $I$ were relatively accurate, with point estimate errors of
     395 
-    [207-683].
-Errors were significantly higher for $\alpha \geq 1$
-    (Wilcoxon rank-sum test, $p = 8\!\times\!10^{-3}$)
-and for $I$ = 2000
-    ($p = {<}10^{-5}$),
-but not for any values of $m$ (one-way ANOVA). The $m$ parameter was estimated
-correctly in only
+    [207 - 
+    683] individuals.
+These errors were significantly higher when the true value of $\alpha$ was
+at least 1
+    (Wilcoxon rank-sum test, $p$ = $0.0077$)
+and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
+did not affect the estimates of $I$ (one-way ANOVA).
+
+The $m$ parameter was estimated correctly in only
     27 \%
-of simulations, little better than random guessing. The true values of $m$ and
-$I$ did not significantly affect the error (one-way ANOVA). Finally, the total
-number of nodes \gls{N} was consistently over-estimated by about a factor of
-two
+of simulations, barely better than random guessing. The true values of the
+other parameters did not significantly affect the estimates of $m$ (both
+one-way ANOVA). Finally, the total number of nodes $N$ was consistently
+over-estimated by about a factor of two
     (error 5987 
     [2060 - 
-     7999]).
-No other parameters influenced the accuracy of the $N$ estimates (one-way
-ANOVA).
+     7999] individuals).
+No parameters influenced the accuracy of the $N$ estimates (all one-way ANOVA).
 
-\begin{figure}
-  \includegraphics{abc-point-estimate-m2}
-  \caption[\Acrlong{MAP} point estimates for \gls{BA} model parameters obtained
-    by running \software{netabc} on simulated data, for simulations with $m = 2$.] 
-  {
-    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by         
-    running \software{netabc} on simulated data. Values shown are for               
-    simulations with \gls{m} = 2. Dashed lines indicate true values. (A)            
-    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations  
-    against known values. (B) Estimates of \gls{m} and \gls{N} which were held   
-    fixed in these simulations at the values \gls{m} = 2 and \gls{N} = 5000. 
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth]{abc-point-estimate-m2}
+  \vspace{6pt}
+  \caption{
+    Point estimates of BA model parameters obtained by running kernel-ABC
+    on simulated phylogenies without training, for simulations with $m = 2$.
+    Dotted lines indicate true values, and limits of the $y$-axes are regions
+    of uniform prior density. (A) Estimates for $\alpha$ and $I$ against their
+    true values in simulations. (B) Estimates for $m$ and $N$, which were held
+    fixed in these simulations, against true values of $\alpha$.
   }
-  \label{fig:abcptm2}
+  \label{fig:abcpt}
 \end{figure}
 
-\Cref{fig:abcex} shows the \gls{ABC} approximation to the posterior
-distribution on the \gls{BA} parameters for one simulation. Equivalent plots
-for one replicate simulation with each combination of parameters can be found
-in \cref{fig:0.0-1000-2-5000-0,fig:0.5-1000-2-5000-0,fig:1.0-1000-2-5000-0,fig:1.5-1000-2-5000-0,fig:0.0-2000-2-5000-0,fig:0.5-2000-2-5000-0,fig:1.0-2000-2-5000-0,fig:1.5-2000-2-5000-0,fig:0.0-1000-3-5000-0,fig:0.5-1000-3-5000-0,fig:1.0-1000-3-5000-0,fig:1.5-1000-3-5000-0,fig:0.0-2000-3-5000-0,fig:0.5-2000-3-5000-0,fig:1.0-2000-3-5000-0,fig:1.5-2000-3-5000-0,fig:0.0-1000-4-5000-0,fig:0.5-1000-4-5000-0,fig:1.0-1000-4-5000-0,fig:1.5-1000-4-5000-0,fig:0.0-2000-4-5000-0,fig:0.5-2000-4-5000-0,fig:1.0-2000-4-5000-0,fig:1.5-2000-4-5000-0}.
-\Gls{HPD} intervals around \gls{alpha} and \gls{I} were
-narrow relative to the region of nonzero prior density, whereas the intervals
-for $m$ and \gls{N} were widely dispersed. \Cref{tab:abchpd} shows point
-estimates and 95\% \gls{HPD} intervals averaged over all simulations.
+\begin{table*}[ht]
+  \centering
+  \input{\tablepath/abc-hpd.tex}
+  \caption{
+      Average maximum \textit{a posteriori} point estimates and 95\% highest
+      posterior density (HPD) interval widths for BA model parameter estimates
+      obtained with kernel-ABC. Three transmission trees were simulated under
+      each combination of the listed parameter values, and the parameters were
+      estimated with kernel-ABC without training.
+  }
+  \label{tab:abchpd}
+\end{table*}
 
-\begin{figure}
-  \includegraphics{{abc-posterior/1.0_1000_2_5000_0}.pdf}
-  \caption[Approximate marginal posterior distributions of BA model parameters
-      obtained by applying \textit{netabc} to a simulated transmission tree
-      with values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]
-    {
-        Approximate marginal posterior distributions of BA model parameters
-        obtained by applying \textit{netabc} to a simulated transmission tree
-        with BA parameter values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ =
-        5000. Vertical dashed lines indicate true values. Shaded areas are 95\%
-        highest posterior density intervals. $x$-axes indicate regions of
-        nonzero prior density.
+The dispersion of the ABC approximation to the posterior also varied between
+the parameters, with narrower HPD intervals for the parameters with more
+accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows
+the distributions for for one simulation. Equivalent plots for one replicate
+simulation with each studied parameter combination can be found in
+\cref{fig:0.0-1000-2-5000-0,fig:0.5-1000-2-5000-0,fig:1.0-1000-2-5000-0,fig:1.5-1000-2-5000-0,fig:0.0-2000-2-5000-0,fig:0.5-2000-2-5000-0,fig:1.0-2000-2-5000-0,fig:1.5-2000-2-5000-0,fig:0.0-1000-3-5000-0,fig:0.5-1000-3-5000-0,fig:1.0-1000-3-5000-0,fig:1.5-1000-3-5000-0,fig:0.0-2000-3-5000-0,fig:0.5-2000-3-5000-0,fig:1.0-2000-3-5000-0,fig:1.5-2000-3-5000-0,fig:0.0-1000-4-5000-0,fig:0.5-1000-4-5000-0,fig:1.0-1000-4-5000-0,fig:1.5-1000-4-5000-0,fig:0.0-2000-4-5000-0,fig:0.5-2000-4-5000-0,fig:1.0-2000-4-5000-0,fig:1.5-2000-4-5000-0}.
+HPD intervals around $\alpha$ and $I$ were often narrow
+relative to the region of nonzero prior density, whereas the intervals for $m$
+and $N$ were more widely dispersed.
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{abc-posterior-example}
+  \vspace{6pt}
+  \caption{
+    Marginal posterior distributions of BA model parameters estimated
+    with kernel-ABC for a single simulated transmission tree. Dotted
+    lines and shaded polygon indicate true values.
   }
@@
 $\alpha$, 
-  1.1 
+  1.09 
   [0.6 -
@@
 $I$,
-  1091 
+  1097 
   [662 -
@@
 $N$,
-  12566 
-  [3733 -
+  12766 
+  [3367-
    14945].
@@
 $\alpha$, 
-  0.37 
-  [0 -
-   0.62];
+  0.1 
+  [0.02 -
+   0.63];
 $I$,
-  2407 
-  [1423 -
-   3811];
+  2399 
+  [1419 -
+   3742];
 $m$,
@@
 $N$,
-  13811 
-  [2624 -
+  9338 
+  [2881 -
    14647].
@@
 $\alpha$ was somewhat lower than typical.
+
+%<<abc_glm, include=FALSE>>=
+%    source("global.R")
+%    options(scipen=-1, digits=2)
+%@
+%
+%<<alpha_glm, include=FALSE>>=
+%    # alpha_error is influenced by alpha and m but not I
+%    stopifnot(alpha.glm[Parameter == "alpha", min(p) < 0.05])
+%    stopifnot(alpha.glm[Parameter == "m", min(p) < 0.05])
+%    stopifnot(alpha.glm[Parameter == "I", min(p) > 0.05])
+%
+%    # alpha_error is correlated with true_alpha
+%    alpha.test <- d[,cor.test(alpha_error, true_alpha, method="spearman")]
+%    stopifnot(alpha.test$p.value < 0.05)
+%
+%    # alpha_error is not correlated with true_m
+%    alpha.m.cor <- d[,cor.test(alpha_error, true_m)]
+%    stopifnot(alpha.m.cor$p.value < 0.05)
+%@
+%
+%We used \software{netabc} to estimate the parameters of the \gls{BA} model on
+%simulated trees where the true parameter values were known. Point estimates for
+%each parameter are shown in \cref{fig:abcptm2} for the simulations with \gls{m} =
+%2. The results for the other values of \gls{m} were similar
+%(\cref{fig:abcptm3,fig:abcptm4}). The median [IQR] absolute error of estimates
+%of \gls{alpha} across all simulations was
+%    d[,median(alpha_error)] 
+%    [d[,quantile(alpha_error, 0.25)]-d[,quantile(alpha_error, 0.75)]].
+%\Gls{GLM} analysis indicated that the true values of both \gls{alpha} and
+%\gls{m} had significant effects on the error in estimated \gls{alpha}
+%    ($p$ values $pp(alpha.glm[Parameter == "alpha", min(p)], eq=FALSE)$
+%     $pp(alpha.glm[Parameter == "m", min(p)], eq=FALSE)$,
+%     \cref{tab:glmalpha}),
+%but the true value of \gls{I} did not. There was a significant negative
+%correlation between the true value of \gls{alpha} and the error
+%    (Spearman's $\rho$ = round(alpha.test$estimate, 2),
+%     $p pp(alpha.test$p.value, eq=TRUE)$).
+%There was a negative correlation between \gls{m} and the error in \gls{alpha}
+%    (Spearman's $\rho$ = round(alpha.m.cor$estimate, 2),
+%    $p pp(alpha.m.cor$p.value, eq=FALSE)$).
+%
+%<<I_glm, include=FALSE>>=
+%    # I_error is influenced by alpha and I but not by m
+%    stopifnot(I.glm[Parameter == "alpha", min(p) < 0.05])
+%    stopifnot(I.glm[Parameter == "m", min(p) > 0.05])
+%    stopifnot(I.glm[Parameter == "I", min(p) < 0.05])
+%
+%    # correlation between I_error and true_alpha
+%    I.alpha.cor <- d[,cor.test(I_error, true_alpha, method="spearman")]
+%    stopifnot(I.alpha.cor$p.value < 0.05)
+%
+%    # correlation between I_error and true_I
+%    I.I.cor <- d[,cor.test(I_error, true_I, method="spearman")]
+%    stopifnot(I.alpha.cor$p.value < 0.05)
+%@
+%
+%The mean error in the estimated value of \gls{I} was
+%    d[,round(median(I_error))] 
+%    [d[,round(quantile(I_error, 0.25))]-d[,round(quantile(I_error, 0.75))]],
+%an over-estimate of roughly a factor of 1.5
+%(\cref{fig:abcptm2,fig:abcptm1,fig:abcptm3,fig:abcptm4}). \Gls{GLM} analysis
+%indicated a relationship between the error in estimated \gls{I}, and the true
+%values of \gls{alpha} and \gls{I}
+%    ($p$ values $pp(I.glm[Parameter == "alpha", min(p)], eq=FALSE)$
+%     $pp(alpha.glm[Parameter == "I", min(p)], eq=FALSE)$,
+%     \cref{tab:glmalpha}),
+%but not the true value of \gls{m}. There was a significant correlation between 
+%the true value of \gls{alpha} and the error in estimated \gls{I}
+%    (Spearman's $\rho$ = I.alpha.cor$estimate,
+%     $p pp(I.alpha.cor$p.value, eq=TRUE)$),
+%and between the true value of \gls{I} and the error in estimated \gls{I}
+%    (Spearman's $\rho$ = I.I.cor$estimate,
+%     $p pp(I.I.cor$p.value, eq=TRUE)$).
+%
+%<<m_glm, include=FALSE>>=
+%    # effects of parameters on m_error
+%    stopifnot(m.glm[Parameter == "m", min(p)] < 0.05)
+%    stopifnot(m.glm[Parameter == "alpha", min(p)] > 0.05)
+%    stopifnot(m.glm[Parameter == "I", min(p)] > 0.05)
+%
+%    # m was only estimated correctly in just over 20\% of simulations with m > 2
+%    m.tbl <- prop.table(table(d[true_m > 1,m_error]))
+%    stopifnot(m.tbl[1] > 0.2 & m.tbl[1] < 0.3)
+%
+%    # with m = 1, over 95% of simulations are right
+%    m1.est <- d[true_m == 1, m_error]
+%    stopifnot(sum(m1.est == 0) / length(m1.est) > 0.95)
+%    stopifnot(sum(m1.est == 0) / length(m1.est) < 1)
+%@
+%
+%\gls{GLM} analysis showed an effect of the true value of \gls{m} on the error
+%in the estimated \gls{m}
+%    ($p pp(m.glm[Parameter == "m", min(p)])$,
+%     \cref{tab:glmm}).
+%When the true value of \gls{m} was 1, the correct value was recovered by
+%\software{netabc} in virtually every case
+%    (sum(m1.est == 0) out of length(m1.est) simulations).
+%However, when the true value of \gls{m} was 2 or higher, the correct value was
+%recovered in only 
+%    as.integer(m.tbl[1] * 100)\%
+%of simulations, little better than random guessing. The \gls{GLM} analysis did
+%not indicate any effects of the true parameter values on the error in estimated
+%\gls{m} (\cref{tab:glmm}).
+%
+%<<N_glm, include=FALSE>>=
+%    # alpha_error is influenced by alpha and m but not I
+%    stopifnot(N.glm[Parameter == "alpha", min(p) < 0.05])
+%    stopifnot(N.glm[Parameter == "I", min(p) < 0.05])
+%    stopifnot(N.glm[Parameter == "m", min(p) > 0.05])
+%
+%    N.alpha.test <- d[,wilcox.test(.SD[true_alpha == 1.5, N_error], 
+%                                   .SD[true_alpha < 1.5, N_error])]
+%    stopifnot(N.alpha.test$p.value < 0.05)
+%    N.I.test <- d[,wilcox.test(.SD[true_I == 3000, N_error], 
+%                               .SD[true_I < 3000, N_error])]
+%    stopifnot(N.I.test$p.value < 0.05)
+%@
+%
+%Finally, the total number of nodes \gls{N} was consistently over-estimated by
+%about a factor of two
+%    (error d[,format(median(N_error), scientific=FALSE)] 
+%    [d[,format(quantile(N_error, 0.25), scientific=FALSE)] - 
+%     d[,format(quantile(N_error, 0.75), scientific=FALSE)]]).
+%The fitted \gls{GLM} indicated that the true values of both \gls{alpha} and
+%\gls{I} had an effect on the error in the estimated \gls{N}
+%    ($p$-values $pp(N.glm[Parameter == "alpha", min(p)], eq=FALSE)$ and 
+%     $pp(N.glm[Parameter == "I", min(p)], eq=FALSE)$,
+%     \cref{tab:glmN}),
+%but that the true value of \gls{m} did not. The error in the estimated \gls{N}
+%when \gls{alpha} was equal to 1.5 was slightly lower than for other values of
+%\gls{alpha}
+%    (median [IQR] error rates 
+%     d[true_alpha == 1.5, format(median(N_error), scientific=FALSE)] 
+%    [d[true_alpha == 1.5, format(quantile(N_error, 0.25), scientific=FALSE)] - 
+%     d[true_alpha == 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]]
+%     for \gls{alpha} = 1.5 vs. 
+%     d[true_alpha < 1.5, format(median(N_error), scientific=FALSE)] 
+%    [d[true_alpha < 1.5, format(quantile(N_error, 0.25), scientific=FALSE)] - 
+%     d[true_alpha < 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]]
+%     for \gls{alpha} < 1.5),
+%and this difference was statistically significant
+%    (Wilcoxon rank-sum test, $p pp(N.alpha.test$p.value, eq=TRUE)$).
+%Similarly, the error in the estimated \gls{N} when \gls{I} was equal to 3000 was
+%lower than for other values of \gls{I}
+%    (median [IQR] error rates 
+%     d[true_I == 3000, format(median(N_error), scientific=FALSE)] 
+%    [d[true_I == 3000, format(quantile(N_error, 0.25), scientific=FALSE)] - 
+%     d[true_I == 3000, format(quantile(N_error, 0.75), scientific=FALSE)]]
+%     for \gls{I} = 3000 vs. 
+%     d[true_alpha < 3000, format(median(N_error), scientific=FALSE)] 
+%    [d[true_alpha < 3000, format(quantile(N_error, 0.25), scientific=FALSE)] - 
+%     d[true_alpha < 3000, format(quantile(N_error, 0.75), scientific=FALSE)]]
+%     for \gls{I} < 3000),
+%and this difference was statistically significant
+%    (Wilcoxon rank-sum test, $p pp(N.I.test$p.value, eq=TRUE)$).
+%
+%\begin{figure}
+%  \includegraphics{abc-point-estimate-m2}
+%  \caption[\Acrlong{MAP} point estimates for \gls{BA} model parameters obtained
+%    by running \software{netabc} on simulated data, for simulations with $m = 2$.] 
+%  {
+%    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by         
+%    running \software{netabc} on simulated data. Values shown are for               
+%    simulations with \gls{m} = 2. Dashed lines indicate true values. (A)            
+%    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations  
+%    against known values. (B) Estimates of \gls{m} and \gls{N} which were held   
+%    fixed in these simulations at the values \gls{m} = 2 and \gls{N} = 5000. 
+%  }
+%  \label{fig:abcptm2}
+%\end{figure}
+%
+%<<posterior_sims, echo=FALSE>>=
+%    N <- 5000
+%    replicate <- 0
+%    lab <- NULL
+%    for (m in c(2, 3, 4)) {
+%    for (I in c(1000, 2000)) {
+%    for (alpha in c(0, 0.5, 1, 1.5)) {
+%        lab <- c(lab, sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate))
+%    }
+%    }
+%    }
+%    lab <- paste(lab, collapse=",")
+%@
+%
+%\Cref{fig:abcex} shows the \gls{ABC} approximation to the posterior
+%distribution on the \gls{BA} parameters for one simulation. Equivalent plots
+%for one replicate simulation with each combination of parameters can be found
+%in \cref{lab}. \Gls{HPD} intervals around \gls{alpha} and \gls{I} were
+%narrow relative to the region of nonzero prior density, whereas the intervals
+%for $m$ and \gls{N} were widely dispersed. \Cref{tab:abchpd} shows point
+%estimates and 95\% \gls{HPD} intervals averaged over all simulations.
+%
+%\begin{figure}
+%  \includegraphics{{abc-posterior/1.0_1000_2_5000_0}.pdf}
+%  \caption[Approximate marginal posterior distributions of BA model parameters
+%      obtained by applying \textit{netabc} to a simulated transmission tree
+%      with values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]
+%    {
+%        Approximate marginal posterior distributions of BA model parameters
+%        obtained by applying \textit{netabc} to a simulated transmission tree
+%        with BA parameter values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ =
+%        5000. Vertical dashed lines indicate true values. Shaded areas are 95\%
+%        highest posterior density intervals. $x$-axes indicate regions of
+%        nonzero prior density.
+%    }
+%  \label{fig:abcex}
+%\end{figure}
+%
+%
+%\begin{table}
+%    \centering
+%    \input{\tablepath/abc-hpd}
+%    \caption[
+%        Maximum \textit{a priori} estimates and 95\% highest posterior density
+%        (HPD) interval boundaries for \gls{BA} model parameters estimated with
+%        \software{netabc}, averaged over simulated transmission trees.
+%    ]
+%    {
+%        Maximum \textit{a priori} estimates and 95\% highest posterior density
+%        (HPD) interval boundaries for \gls{BA} model parameters estimated with
+%        \software{netabc}, averaged over simulated transmission trees.
+%    }
+%    \label{tab:abchpd}
+%\end{table}
+%
+%<<mixed_peerdriven, include=FALSE>>=
+%  library(coda)
+%  d <- fread("bzcat ../../simulations/abc-pa-mixed-alpha/abc/*")
+%  d <- d[iter == max(iter)]
+%  d <- d[sample(1:nrow(d), prob=weight, replace=TRUE)]
+%  params <- c("alpha", "m", "N", "I")
+%  d <- melt(d, measure.vars=params, variable.name="parameter")
+%  f <- function (x) {
+%    dens <- density(x)
+%    hpd <- HPDinterval(mcmc(x))
+%    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
+%         hpd.max=hpd[,"upper"])
+%  }
+%  d <- d[,f(value), by=parameter]
+%  setkey(d, parameter)
+%
+%  m <- collect.metadata("../../simulations/abc-pa-peerdriven/abc/*")
+%  p <- fread(paste("bzcat", rownames(m)[m$sample_peer == 1]))
+%  p <- p[iter == max(iter)]
+%  p <- p[sample(1:nrow(p), prob=weight, replace=TRUE)]
+%  params <- c("alpha", "m", "N", "I")
+%  p <- melt(p, measure.vars=params, variable.name="parameter")
+%  f <- function (x) {
+%    dens <- density(x)
+%    hpd <- HPDinterval(mcmc(x))
+%    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
+%         hpd.max=hpd[,"upper"])
+%  }
+%  p <- p[,f(value), by=parameter]
+%  setkey(p, parameter)
+%  options(scipen=5)
+%@
+%
+%To test the effect of model misspecification, we simulated one network where
+%the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
+%other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
+%HPD] estimates for each parameter were: 
+%$\alpha$, 
+%  round(d["alpha", map], 2) 
+%  [round(d["alpha", hpd.min], 2) -
+%   round(d["alpha", hpd.max], 2)];
+%$I$,
+%  d["I", round(map)] 
+%  [d["I", round(hpd.min)] -
+%   d["I", round(hpd.max)]];
+%$m$,
+%  d["m", floor(map)] 
+%  [d["m", floor(hpd.min)] -
+%   d["m", floor(hpd.max)]];
+%$N$,
+%  d["N", round(map)] 
+%  [d["N", round(hpd.min)] -
+%   d["N", round(hpd.max)]].
+%The approximate posterior distributions for this simulation are shown in
+%\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
+%transmission tree in a peer-driven fashion, where the probability to sample a
+%node was twice as high if one of its peers had already been sampled. The
+%parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
+%$I$ = 2000. The estimated values were
+%$\alpha$, 
+%  p["alpha", round(map, 2)] 
+%  [p["alpha", round(hpd.min, 2)] -
+%   p["alpha", round(hpd.max, 2)]];
+%$I$,
+%  p["I", floor(map)] 
+%  [p["I", floor(hpd.min)] -
+%   p["I", floor(hpd.max)]];
+%$m$,
+%  p["m", floor(map)] 
+%  [p["m", floor(hpd.min)] -
+%   p["m", floor(hpd.max)]];
+%$N$,
+%  p["N", floor(map)] 
+%  [p["N", floor(hpd.min)] -
+%   p["N", floor(hpd.max)]].
+%The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
+%of these results were in line with estimates obtained on other simulated
+%datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
+%$\alpha$ was somewhat lower than typical.
 
@@
 estimates [95\% \gls{HPD}] 
-  1.05 
+  1.04 
   [0.04 - 
-   1.27]
+   1.25]
 and
-  0.84 
+  0.82 
   [0.01 -
-   1.02] respectively).
+   1.03] respectively).
 The MAP estimates did not change appreciably when $m = 1$ was disallowed by the
@@
   (0.04 - 
-   1.27).
+   1.25).
 When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
 low estimated $\alpha$ values
-  (0.06 
+  (0.3 
+  [0 - 
+  0.75]
+and
+  0.3 
   [0.01 -
-  0.73]
-and
-  0.19 
-  [0.01 -
-   0.8]). 
+   0.79]). 
 However, the MAP estimates increased when $m = 1$ was not permitted, although
@@
   [0.02 - 
-  0.94]
+  0.93]
 and
-  0.59 
-  [0.07 -
+  0.58 
+  [0.08 -
    0.95]).
@@
 for both priors on $m$
-  (0.32 for $m \geq 1$;
-   0.39 for $m \geq 2$).
+  (0.29 for $m \geq 1$;
+   0.23 for $m \geq 2$).
 However, the confidence interval was much wider when $m = 1$ was allowed
-  ([0.04 -
-    1.62] for $m \geq 1$ vs.
+  ([0.03 -
+    1.6] for $m \geq 1$ vs.
    [0 -
@@
 intervals compared to the nonzero prior density region
-  (\citeauthor{cuevas2009hiv}, 482 
-  [293 -
-   2111];
+  (\citeauthor{cuevas2009hiv}, 497 
+  [287 -
+   2430];
    \citeauthor{niculescu2015recent}, 307
-  [136 - 
+  [138 - 
    2822];
-  \citeauthor{li2015hiv}, 1183 
-  [413 -
+  \citeauthor{li2015hiv}, 1217 
+  [383 -
    2897];
-   \citeauthor{wang2015targeting}, 719
-  [176 - 
-   2114]).
+   \citeauthor{wang2015targeting}, 621
+  [182 - 
+   2139]).
 The \citeauthor{novitsky2014impact} data was the outlier, with a very high
 estimated $I$, and HPD interval spanning almost the entire prior region
-  (7409 
+  (7642 
   [187 -
-   8819]).
+   8836]).
 The $I$ estimates and HPD intervals were generally robust to the choice of
@@
 data, for which the MAP estimate was lower 
-  (6973)
+  (7428)
 when $m = 1$ was allowed.
@@
 \end{figure*}
+
+To make our analyses comparable to the existing network literature, we
+estimated values of the power law exponent $\gamma$ for each of the datasets 
+investigated. The results are summarized in \cref{tab:gamma}. All of the
+estimated exponents were in the range $2 \leq \gamma \leq 4$ commonly reported
+in the literature~\autocite{liljeros2001web, schneeberger2004scale,
+colgate1989risk, brown2011transmission}.
+
+\begin{table}
+    \centering
+    \input{\tablepath/realdata-gamma.tex}
+    \caption[
+        Estimated power law exponents for six HIV datasets based on maximum
+        \textit{a priori} estimates of BA model parameters.
+    ]{
+        Estimated power law exponents for six HIV datasets based on maximum
+        \textit{a priori} estimates of BA model parameters. 100 networks were
+        simulated using \textit{MAP} parameter estimates obtained with
+        \software{netabc}. The power law exponent $\gamma$ was estimated for
+        each, and the median of those estimates was used as a point estimate
+        for the corresponding dataset.
+    }
+    \label{tab:gamma}
+\end{table}

### 2016-05-04 doc/thesis/discussion.tex
Made plots of derivatives to explain N/I bias

@@
 \software{netabc}, however the variation in these estimates was higher for
-$\alpha < 1$ than for $\alpha \geq 1$ (\cref{fig:abcptm2,tab:abchpd}). The
+$\alpha < 1$ than for $\alpha \geq 1$ (\cref{fig:abcpt,tab:abchpd}). The
 relationship between \gls{alpha} and the power law exponent \gls{gamma} may
@@
 
-\textbf{ART: THIS SECTION IS IN PROGRESS PLEASE DO NOT READ IT YET}
-
 The \gls{I} parameter, representing the prevalence at the time of sampling, was
-accurately classified by the tree kernel and the \gls{nltt}, but not by
-Sackin's index (\cref{fig:rsquared}). When \gls{N} was held fixed, \gls{I} was
-estimated accurately and without bias by grid search
-\cref{fig:gridest,fig:gridI}. When the \gls{BA} model was fit with
-\software{netabc}, the 95\% \gls{HPD} intervals for \gls{I} usually contained
-the true value, but the \gls{MAP} estimates were biased slightly upward
-(\cref{tab:abchpd}).
-
 also generally estimable, although it was slightly over-estimated for both
-cases we considered with kernel-\gls{ABC}.
-
-The dynamics of the \gls{SI} model,
+cases we considered with kernel-\gls{ABC}. The dynamics of the \gls{SI} model,
 and the coalescent process~\autocite{kingman1982coalescent}, offer a potential
-explanation for this result. In our simulations, we assumed that all discordant
-edges shared the same transmission rate, so that the waiting time until the
-next transmission in the entire network was always inversely proportional to
-the number of discordant edges. Therefore, when there are a high number of
-discordant edges,
-
-In the initial phase of the epidemic, when \gls{I}
-is small, each new transmission results in many new discordant edges.
-
-thus
-decreasing the waiting time until the next transmission. Hence,
-there is an early exponential growth phase, producing many short branches near
-the root of the tree. As the epidemic gets closer to saturating the network,
-the number of discordant edges decays, causing longer waiting times. The
-distribution of coalescence times in the tree should therefore be informative
-about \gls{I}~\autocite{volz2009phylodynamics}. This information is captured by
-the tree kernel, and also by the \gls{nltt} statistic, which both performed
-quite well in classifying \gls{I} (\cref{fig:rsquared}).
-
-\textbf{END OF IN PROGRESS}
+explanation for the identifiability of \gls{I}. In our simulations, we assumed
+that all discordant edges shared the same transmission rate, so that the
+waiting time until the next transmission in the entire network was always
+inversely proportional to the number of discordant edges. In the initial phase
+of the epidemic, when \gls{I} is small, each new transmission results in many
+new discordant edges. Hence, there is an early exponential growth phase,
+producing many short branches near the root of the tree. As the epidemic gets
+closer to saturating the network, the number of discordant edges decays,
+causing longer waiting times. The distribution of coalescence times in the tree
+should therefore be informative about \gls{I}~\autocite{volz2009phylodynamics}.
+This information is captured by the tree kernel, and also by the \gls{nltt}
+statistic, which both performed quite well in classifying \gls{I}
+(\cref{fig:rsquared}).
 
@@
 increase, resulting in longer coalescence times toward the tips. The relative
-accuracy of the \gls{nltt} in these situations coroborates this hypothesis.
+accuracy of the \gls{nltt} in these situations
+(\cref{fig:rsquared,fig:Ncrossv}) coroborates this hypothesis, as the
+\gls{nltt} uses only information about the coalescence times. When all \gls{BA}
+parameters were simultaneously estimated with kernel-\gls{ABC}, \gls{N} was
+nearly always over-estimated by approximately a factor of two
+(\cref{fig:abcpt,tab:abchpd}). One factor which may have contributed to this
+bias was our choice of prior distribution. Since the prior for \gls{I} and
+\gls{N} was jointly uniform on a region where $I \leq N$, more prior weight was
+assigned to higher \gls{N} values. Another contributing factor relates to the
+dynamics of the \gls{SI} model and the coalescent process.
 
-%As mentioned above, \gls{I} and \gls{N} were both systematically over-estimated
-%by our method, although the bias was more severe for \gls{N} than for \gls{I}.
-%
-%When \gls{N} was held fixed and \gls{I} was varied, grid search was able to
-%recover \gls{I} without bias (\cref{fig:est}).
+\gls{I} and \gls{N} were both systematically over-estimated by
+\software{netabc}, although the bias was more severe for \gls{N} than for
+\gls{I}. The number of infected individuals follows a logistic growth curve
+under the \gls{SI} model. This kind of growth curve has three qualitative
+phases: a slow ramp-up, an exponential growth phase, and a slow final phase
+when the susceptible population is almost depleted. The waiting times until the
+next transmission, which determine the coalescence times in the tree, are
+dependent on the growth phase of the epidemic. Therefore, we hypothesize that
+it is the growth phase at the time of sampling which most affects tree shape,
+rather than the specific values of \gls{I} or \gls{N}. To investigate this
+hypothesis, we simulated transmission trees over networks on a grid of \gls{I}
+and \gls{N} values in the region of uniform prior density. We fit logistic
+growth curves to the proportion of infected individuals over time, and
+calculated the first and second derivatives of these curves at the time of
+transmission tree sampling. These derivatives give us an indication of the
+growth rates of the epidemics at the time of sampling. As shown in
+\cref{fig:derivs}, there are bands along which both derivatives are similar
+which contain the values we tested. These bands span mostly higher values of
+\gls{N} and \gls{I} than the true values. Therefore, if \gls{N} and \gls{I} are
+free to vary (as is the case in kernel-\gls{ABC}), and our hypothesis is true,
+both parameters will tend to be overestimated due to being less identifiable
+within their own band. However, when \gls{N} is fixed at 5000, the derivatives
+vary substantially along the \gls{I}-axis, which explains why the grid search
+estimates of \gls{I} were accurate and unbiased (\cref{fig:gridptI}).
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{derivatives.pdf}
+  \caption[
+    First and second derivatives of epidemic growth curves at time of sampling
+    for various values of $I$ and $N$.
+  ]{
+    First and second derivatives of epidemic growth curves at time of sampling
+    for various values of \gls{I} and \gls{N}. Networks were simulated under
+    the \gls{BA} model with \gls{alpha} = 1.0, \gls{m} = 2, and \gls{N} varied 
+    along the values shown on the $x$-axis. Transmission trees were sampled at
+    the time when \gls{I} nodes were infected ($y$-axis). Logistic growth
+    curves were fit to epidemic trajectories derived from the transmission
+    trees, and their first and second derivatives were calculated at the time
+    of sampling. Contours show first derivatives, while colours indicate second
+    derivatives. Values of \gls{I} and \gls{N} used in simulation experiments
+    with kernel-\gls{ABC} are indicated by diamonds.
+  }
+  \label{fig:derivs}
+\end{figure}
 

### 2016-05-04 doc/thesis/results.tex
Made plots of derivatives to explain N/I bias

@@
 
-\Cref{fig:abcpt} shows \gls{MAP} point estimates of the BA model parameters
+\Cref{fig:abcptm2} shows \gls{MAP} point estimates of the BA model parameters
 obtained with kernel-ABC on simulated data. The estimates shown correspond only
@@
   }
-  \label{fig:abcpt}
+  \label{fig:abcptm2}
 \end{figure}
@@
 $I$,
-  1097 
+  1109 
   [662 -
@@
   3 
-  [1 -
+  [2 -
    5];
 $N$,
-  12766 
-  [3367-
-   14945].
+  12673 
+  [3733-
+   14977].
 The approximate posterior distributions for this simulation are shown in
@@
 $\alpha$, 
-  0.1 
-  [0.02 -
+  0.36 
+  [0.01 -
    0.63];
@@
   2399 
-  [1419 -
-   3742];
+  [1379 -
+   3838];
 $m$,
@@
 $N$,
-  9338 
-  [2881 -
-   14647].
+  8830 
+  [2836 -
+   14720].
 The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both

### 2016-05-04 doc/thesis/discussion.tex
Fix reference

@@
 \software{netabc}, however the variation in these estimates was higher for
-$\alpha < 1$ than for $\alpha \geq 1$ (\cref{fig:abcpt,tab:abchpd}). The
+$\alpha < 1$ than for $\alpha \geq 1$ (\cref{fig:abcptm2,tab:abchpd}). The
 relationship between \gls{alpha} and the power law exponent \gls{gamma} may
@@
 nearly always over-estimated by approximately a factor of two
-(\cref{fig:abcpt,tab:abchpd}). One factor which may have contributed to this
+(\cref{fig:abcptm2,tab:abchpd}). One factor which may have contributed to this
 bias was our choice of prior distribution. Since the prior for \gls{I} and

### 2016-05-05 doc/thesis/discussion.tex
Updated references

@@
 the possible misspecification, our estimates of the power law exponent $\gamma$
-were in line with those reported in the literature (\cref{tab:gamma}).
+were within the range of values reported in the literature (\cref{tab:gamma}).

### 2016-05-05 doc/thesis/objective.tex
Updated references

@@
 tend to be scale-free~\autocite{colgate1989risk, liljeros2001web,
-schneeberger2004scale}, meaning that their degree distributions follow a power
-law (although there has been some disagreement, see
-\autocite{handcock2004likelihood, bansal2007individual}). Preferential
+schneeberger2004scale, clemencon2015statistical}, meaning that their degree
+distributions follow a power law (although there has been some disagreement,
+see \autocite{handcock2004likelihood, bansal2007individual}). Preferential
 attachment has been postulated as a mechanism by which scale-free networks
@@
 \gls{HIV} epidemics are driven by social dynamics and are therefore likely to
-be highly nonrandom. Moreover, since no cure yet exists, efforts to curtail the
-progression of an epidemic have relied on preventing further transmissions
-through measures such as \gls{tasp} and education leading to behaviour change.
-The effectiveness of this type of intervention can vary significantly based on
-the underlying structure of the network and the particular nodes to whom the
-intervention is targeted~\autocite{little2014using,wang2015targeting}. Due to
-this combination of data availability and potential public health impact,
-\gls{HIV} is an obvious context in which our method could be applied.
-Therefore, the third and final aim of this work is to apply kernel-\gls{ABC} to
-fit the \gls{BA} model to existing \gls{HIV} outbreaks.
+be highly nonrandom~\autocite{clemencon2015statistical}. Moreover, since no
+cure yet exists, efforts to curtail the progression of an epidemic have relied
+on preventing further transmissions through measures such as \gls{tasp} and
+education leading to behaviour change. The effectiveness of this type of
+intervention can vary significantly based on the underlying structure of the
+network and the particular nodes to whom the intervention is
+targeted~\autocite{little2014using,wang2015targeting}. Due to this combination
+of data availability and potential public health impact, \gls{HIV} is an
+obvious context in which our method could be applied. Therefore, the third and
+final aim of this work is to apply kernel-\gls{ABC} to fit the \gls{BA} model
+to existing \gls{HIV} outbreaks.
 

### 2016-05-05 doc/thesis/results.Rnw
Made plots of derivatives to explain N/I bias
Updated references

@@
 investigated. The results are summarized in \cref{tab:gamma}. All of the
-estimated exponents were in the range $2 \leq \gamma \leq 4$ commonly reported
-in the literature~\autocite{liljeros2001web, schneeberger2004scale,
-colgate1989risk, brown2011transmission}.
+estimated exponents were in the range $2 \leq \gamma \leq 2.5$.
 

### 2016-05-05 doc/thesis/results.tex
Fix reference
Updated references

@@
 $I$,
-  1109 
+  1137 
   [662 -
@@
 $N$,
-  12673 
-  [3733-
-   14977].
+  12785 
+  [3861-
+   14945].
 The approximate posterior distributions for this simulation are shown in
@@
 $\alpha$, 
-  0.36 
-  [0.01 -
+  0.37 
+  [0.02 -
    0.63];
 $I$,
-  2399 
-  [1379 -
-   3838];
+  2352 
+  [1423 -
+   3811];
 $m$,
@@
 $N$,
-  8830 
-  [2836 -
-   14720].
+  8508 
+  [2636 -
+   14615].
 The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both

### 2016-05-05 doc/thesis/intro.tex
Formatting for GPS

@@
 In epidemiology, a \defn{transmission tree} is a graphical representation of an
-epidemic's progress through a population. Like phylogenies, transmission trees
-have tips, nodes, edges, and branch lengths. However, rather than recording an
-evolutionary process (speciation), they record an epidemiological process
-(transmission). The tips of a transmission tree represent the removal of
-infected hosts, while internal nodes correspond to transmissions from one host
-to another. Transmission trees generally have branch lengths in units of
-calendar time, with branching times indicating times of transmission. The root
-of a transmission tree corresponds to the initially infected patient who
-introduced the epidemic into the network, also known as the \defn{index case}.
-The internal nodes may be labelled with the donor of the transmission pair, if
-this is known. The tips of the tree, rather than being fixed at the present
-day, are placed at the time at which the individual was removed from the
-epidemic, such as by death, recovery, isolation, behaviour change, or
-migration. Consequently, the transmission tree may not be ultrametric, but may
-have tips located at varying distances from the root. Such trees are said to
-have \defn{heterochronous} taxa~\autocite{drummond2003measurably}, in contrast
-to the \defn{isochronous} taxa found in most phylogenies of macro-organisms. A
-transmission tree is illustrated in \cref{fig:contactnet} (right).
+epidemic's progress through a population~\autocite{ypma2013relating}. Like
+phylogenies, transmission trees have tips, nodes, edges, and branch lengths.
+However, rather than recording an evolutionary process (speciation), they
+record an epidemiological process (transmission). The tips of a transmission
+tree represent the removal of infected hosts, while internal nodes correspond
+to transmissions from one host to another. Transmission trees generally have
+branch lengths in units of calendar time, with branching times indicating times
+of transmission. The root of a transmission tree corresponds to the initially
+infected patient who introduced the epidemic into the network, also known as
+the \defn{index case}. The internal nodes may be labelled with the donor of the
+transmission pair, if this is known. The tips of the tree, rather than being
+fixed at the present day, are placed at the time at which the individual was
+removed from the epidemic, such as by death, recovery, isolation, behaviour
+change, or migration~\autocite{stadler2013uncovering}. Consequently, the
+transmission tree may not be ultrametric, but may have tips located at varying
+distances from the root. Such trees are said to have \defn{heterochronous}
+taxa~\autocite{drummond2003measurably}, in contrast to the \defn{isochronous}
+taxa found in most phylogenies of macro-organisms. A transmission tree is
+illustrated in \cref{fig:contactnet} (right).
 
@@
 one-to-one correspondence with transmissions~\autocite{welch2011statistical,
-keeling2005networks}. Transmission networks are discussed further in
-\cref{sec:contactnet}, and the distinction between the two objects is
-illustrated in \cref{fig:contactnet}. However, since transmission networks
-generally have no cycles (unless re-infection occurs), they are trees in the
-graph theoretical sense, and hence are sometimes also referred to as
-transmission trees~\autocite[\eg][]{kenah2015algorithms}. In this work, we
-reserve the term ``transmission tree'' for the objects depicted on the right
-side of \cref{fig:contactnet}, following \eg~\autocite{stadler2013uncovering}.
-The term ``transmission network'' is taken to mean the subgraph of the contact
-network along which transmissions occurred, following
+keeling2005networks}. The distinction between the two objects is illustrated in
+\cref{fig:contactnet}. However, since transmission networks generally have no
+cycles (unless re-infection occurs), they are trees in the graph theoretical
+sense, and hence are sometimes also referred to as transmission
+trees~\autocite[\eg][]{kenah2015algorithms}. In this work, we reserve the term
+``transmission tree'' for the objects depicted on the right side of
+\cref{fig:contactnet}, following \eg~\autocite{stadler2013uncovering}. The term
+``transmission network'' is taken to mean the subgraph of the contact network
+along which transmissions occurred, following
 \eg~\autocite{welch2011statistical, keeling2005networks}.
@@
 individuals is usually prohibitive, and may be additionally be confounded by
-misreporting~\autocite{eames2015six}. However, it turns out that for viral
-epidemics, some of the epidemiological information contained in the
-transmission tree leaves a mark on the viral genetic material circulating in
-the population. A family of methods called
+misreporting and other challenges~\autocite{eames2015six}. However, it turns
+out that for viral epidemics, some of the epidemiological information contained
+in the transmission tree leaves a mark on the viral genetic material
+circulating in the population. A family of methods called
 \defn{phylodynamics}~\autocite{grenfell2004unifying} addresses the challenge of
@@
 methods either assume the viral phylogeny is known, as we do in this work, or
-(more commonly) integrate over phylogenetic uncertainty using Bayesian methods.
-Phylogenetic inference is a complex topic which we shall not discuss here; see
-\eg \autocite{nei2000molecular} for a full review.
+(more commonly) integrate over phylogenetic uncertainty in a Bayesian
+framework. Phylogenetic inference is a complex topic which we shall not discuss
+here; see \eg~\autocite{nei2000molecular} for a full review.
 
@@
 in the transmission tree. Another possibility is that one host transmitted to
-two or more recipients, but the lineages they each received originated within
-the donor host in a different order than that in which the transmissions
-occurred.  In this case, the topology of the transmission tree and the viral
-phylogeny will be mismatched. In practice, this discordance has not proven an
-insurmountable problem: for example, \textcite{leitner1996accurate,
-paraskevis2004phylogenetic} were able to accurately recover a known
-transmission tree using a viral phylogeny. The problem of accurately estimating
-transmission trees is an ongoing area of
+two or more recipients in one order, but the transmitted lineages originated
+within the donor in a different order. In this case, the topology of the
+transmission tree and the viral phylogeny will be mismatched. In practice, this
+discordance has not proven an insurmountable problem: for example,
+\textcite{leitner1996accurate, paraskevis2004phylogenetic} were able to
+accurately recover a known transmission tree using a viral phylogeny. The
+problem of accurately estimating transmission trees is an ongoing area of
 research~\autocite{cottam2008integrating, jombart2011reconstructing,
@@
 prevalence~\autocite{holmes1995revealing}, so large values may indicate that
-the trees being compared are the products of different epidemic
+the trees being compared were produced by different epidemic
 trajectories~\autocite{janzen2015approximate}.
@@
 is equally likely. This assumption is clearly unrealistic for the majority of
-human communities, and can lead to significant errors in predicted epidemic
-trajectories when there is substantial heterogeneity
-present~\autocite{bansal2007individual, volz2007susceptible,
-rolls2015simulation}. Contact networks provide a way to relax this assumption
-by representing individuals and their contacts explicitly. It is important to
-note that, although panmixia is an unrealistic modelling assumption, it has not
-proven a substantial hurdle to epidemic modelling in
-practice~\autocite{anderson1992infectious}. Using this assumption, researchers
-have been able to derive estimates of the transmission rate and the basic
-reproductive number of various outbreaks, which have agreed with values
+human communities, and can lead to errors in predicted epidemic trajectories
+when there is substantial heterogeneity present~\autocite{bansal2007individual,
+volz2007susceptible, rolls2015simulation}. Contact networks provide a way to
+relax this assumption by representing individuals and their contacts
+explicitly. It is important to note that, although panmixia is an unrealistic
+modelling assumption, it has not proven a substantial hurdle to epidemic
+modelling in practice~\autocite{anderson1992infectious}. Using this assumption,
+researchers have been able to derive estimates of the transmission rate and the
+basic reproductive number of various outbreaks, which have agreed with values
 obtained by on-the-ground data collection~\autocite{stadler2014insights}.
@@
 in real-world sexual networks are heavy-tailed~\autocite{liljeros2001web,
-schneeberger2004scale, colgate1989risk}, so dense or peer-driven sampling would
-be needed to capture the high-degree nodes characterizing the tail of the
-distribution.
+schneeberger2004scale, colgate1989risk}, so dense or respondent-driven
+sampling~\autocite{heckathorn1997respondent} would be needed to capture the
+high-degree nodes characterizing the tail of the distribution.
 
@@
 human sexual networks tend to be scale-free~\autocite{liljeros2001web,
-schneeberger2004scale, colgate1989risk}. Interestingly, many other types of
-network, including computer networks~\autocite{pastor2001epidemic}, biological
-metabolic networks~\autocite{jeong2000large}, and academic co-author
+schneeberger2004scale, colgate1989risk, clemenccon2015statistical}.
+Interestingly, many other types of network, including computer
+networks~\autocite{pastor2001epidemic}, biological metabolic
+networks~\autocite{jeong2000large}, and academic co-author
 networks~\autocite{barabasi2002evolution}, also have the scale-free property.
@@
 
-\subsection{Sequential importance sampling for sequential Monte Carlo}
+\subsection{Sequential importance sampling}
 \label{subsec:sis}
@@
 \defn{importance distribution}, from which we are able to sample. Define the
-\defn{importance weight} as the ratio ratio $w(x) = \gamma(x)/\eta(x)$. We can
+\defn{importance weight} as the ratio $w(x) = \gamma(x)/\eta(x)$. We can
 express the normalizing constant $Z$ in terms of the importance weight and
@@
 \vec{x_{i-1}})$, if it is possible to compute. In a Bayesian setting, the
-prior distribution may be used. The exact form of $\eta(x_i, \vec{x_{i-1}})$
-which minimizes the variance of the weights is called the \defn{optimal
-kernel}~\autocite{cappe2007overview}, the name deriving from the fact that
-$k(x_i, \vec{x_{i-1}}) = \eta(x_i, \vec{x_{i-1}})$ is a Markov kernel. In some
-applications, it is possible to approximate the optimal kernel or even compute
-it explicitly.
+prior distribution may be used. The exact form of $\eta(x_i \mid
+\vec{x_{i-1}})$ which minimizes the variance of the weights is called the
+\defn{optimal kernel}~\autocite{cappe2007overview}, the name deriving from the
+fact that $k(x_i, \vec{x_{i-1}}) = \eta(x_i \mid \vec{x_{i-1}})$ is a Markov
+kernel. In some applications, it is possible to approximate the optimal kernel
+or even compute it explicitly.
 
@@
 since only a few particles retain high importance weights after all $d$
-sequential steps. To mitigate this problem, we periodically apply a resampling
-step when the variance in the importance weights becomes too high. Several
+sequential steps. To mitigate this problem, a resampling step is periodically
+applied when the variance in the importance weights becomes too high. Several
 different criteria have been proposed for when to resample, but we focus here
@@
 \gls{ESS} below a prescribed threshold, conventionally $n/2$. The \gls{ESS} of
-the population of particles is defined as
-\[
-  \ESS(w) = \frac{n}{1 + \Var(w)},
-\]
-where $n$ is the number of particles~\autocite{liu2008monte}. When the
-\gls{ESS} drops below the threshold, we resample the particles according to
-their weights. This results in the removal of low-weight particles from the
+the population of particles is defined as \[ \ESS(w) = \frac{n}{1 + \Var(w)},
+\] where $n$ is the number of particles~\autocite{liu2008monte}. When the
+\gls{ESS} drops below the threshold, particles are resampled according to their
+weights. This results in the removal of low-weight particles from the
 population, and also equalizes all the weights. Various resampling strategies
-beyond the basic sampling with replacement have been proposed, but we will not
-discuss those here.
+beyond the basic sampling with replacement have been
+proposed~\autocite{douc2005comparison}, but we will not discuss those here.
 
@@
 \]
-where $q(\xi, x)$ is a proposal function such as a Gaussian distribution
-centered at $\xi$ (see \cref{subsec:mfit}).
+where $q(x, x')$ is a proposal function such as a Gaussian distribution
+centered at $x$ (see \cref{subsec:mfit}).
 
@@
       \If {$\ESS(w) < T$}
+        \Comment{$T$ is a user-defined threshold}
         \State Resample the particles according to $w$
@@
 in terms of probabilities. Statistical models define, either explicitly or
-implicitly, the probability of observing $y$ given $\vec{x}$ and, if the
-model is parametric, $\theta$. Note that it is entirely possible to have no
-data $\vec{x}$, only observed outcomes $y$. In this case, a model would
-describe the process by which $y$ is generated.
+implicitly, the probability of observing $y$ given $x$ and, if the model is
+parametric, $\theta$. Note that it is entirely possible to have no data $x$,
+only observed outcomes $y$. In this case, a model would describe the process by
+which $y$ is generated.
 
@@
 the $x_i$ as fixed quantities, and therefore have not conditioned the
-probability density on $\vec{x}$. Also, we have assumed that all the $y_i$ are
+probability density on $x$. Also, we have assumed that all the $y_i$ are
 independent.
@@
 walk is straightforward and can be accomplished via the Metropolis-Hastings
-algorithm (\cref{alg:mh}).
+algorithm~\autocite{metropolis1953equation,hastings1970monte} (\cref{alg:mh}).
 

### 2016-05-05 doc/thesis/objective.tex
Formatting for GPS

@@
 \gls{ABC}-\gls{SMC}. Since our distance measure is a kernel function, our
-method is a type of kernel-\gls{ABC}. For ease of exposition, we will often use
-the term ``kernel-\gls{ABC}'' to refer to our method specifically.
+method is a type of kernel-\gls{ABC}~\autocite{nakagome2013kernel}. For ease of
+exposition, we will often use the term ``kernel-\gls{ABC}'' to refer to our
+method specifically.
 
@@
 tend to be scale-free~\autocite{colgate1989risk, liljeros2001web,
-schneeberger2004scale, clemencon2015statistical}, meaning that their degree
+schneeberger2004scale,clemenccon2015statistical}, meaning that their degree
 distributions follow a power law (although there has been some disagreement,
@@
 \gls{HIV} epidemics are driven by social dynamics and are therefore likely to
-be highly nonrandom~\autocite{clemencon2015statistical}. Moreover, since no
+be highly nonrandom~\autocite{clemenccon2015statistical}. Moreover, since no
 cure yet exists, efforts to curtail the progression of an epidemic have relied

### 2016-05-05 doc/thesis/preface.tex
Formatting for GPS

@@
 
-A version of chapter 2 has been submitted for publication in Virus Evolution
-with the title ``Reconstructing network parameters from viral phylogenies.'' An
-oral presentation entitled ``Phylodynamic inference of contact network
-parameters with kernel-ABC'' was given based on chapter 2 to the 23rd HIV
-Dynamics and Evolution meeting on April 25, in Woods Hole, Massachussets, USA
-(the presentation was delivered remotely). A poster based on chapter 2 entitled
+A version of chapter 2 has been submitted for publication with the title
+``Reconstructing network parameters from viral phylogenies.'' An oral
+presentation entitled ``Phylodynamic inference of contact network parameters
+with kernel-ABC'' was given based on chapter 2 to the 23rd HIV Dynamics and
+Evolution meeting on April 25, in Woods Hole, Massachussets, USA (the
+presentation was delivered remotely). A poster based on chapter 2 entitled
 ``Likelihood-free estimation of contact network parameters from viral

### 2016-05-06 doc/thesis/appendix.Rnw
Put test figures back in

@@
 \chapter{Additional plots}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{leventhal2012fig1.pdf}
+  \caption[
+    Reproduction of Figure 1A from Leventhal \textit{et al.} (2012) used to
+    check the accuracy of our implementation of Gillespie simulation.
+  ]{
+    Reproduction of Figure 1A from Leventhal \textit{et al.} (2012) used to
+    check the accuracy of our implementation of Gillespie simulation.
+    Transmission trees were simulated over three types of network, with
+    pathogen transmissibility varying from 0 to 1. Sackin's index was
+    calculated for each simulated transmission tree.
+  }
+  \label{fig:leventhal}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{smc-test.pdf}
+  \caption[
+      Approximation of mixture of Gaussians used by Del Moral \textit{et al.}
+      (2012) and Sisson \textit{et al.} (2009) to test adaptive
+      \gls{ABC}-\gls{SMC}.
+    ]{
+      Approximation of mixture of Gaussians used by Del Moral \textit{et al.}
+      (2012) and Sisson \textit{et al.} (2009) to test SMC. Solid black line
+      indicates true distribution. Grey shaded area shows \gls{ABC}
+      approximation obtained with our implementation of adaptive
+      \gls{ABC}-\gls{SMC}, using 10000 particles with one simulated data point
+      per particle.
+    }
+  \label{fig:smctest}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{smc-test-bimodal.pdf}
+  \caption[
+    Approximation of mixture of two Gaussians used to test convergence of
+    adaptive \gls{ABC}-\gls{SMC} algorithm to a bimodal distribution.
+  ]{
+    Approximation of mixture of two Gaussians used to test convergence of SMC
+    algorithm to a bimodal distribution. Solid black line indicates true
+    distribution. Grey shaded area shows \gls{ABC}-\gls{SMC} approximation
+    obtained with our implementation, using 10000 particles with one simulated
+    data point per particle.
+  }
+  \label{fig:smctest2}
+\end{figure}
 
@@
 %  }        
-%  \label{fig:abcptm3}
+%  \label{fig:abcptm1}
 %\end{figure}
@@
   \label{fig:gamma}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth]{sir-trajectories.pdf}
-  \caption[Average trajectories of epidemics simulated according to the
-  \gls{SIR} model over full and \gls{BA} networks.]
-  {
-      Average trajectories of epidemics simulated according to the \gls{SIR}
-      model over full and \gls{BA} networks. 100 epidemics each were simulated
-      over completely connected networks and networks generated from the
-      \gls{BA} model. The network parameters were $\alpha = 1$ and $m = 2$.
-      The transmission rate along each edge of a network $G$ was taken as
-      $2/|E(G)|$, and the removal rate of each vertex was taken as $1/|V(G)|$.
-      These choices ensured that both networks had the same waiting time until
-      the next transmission at the beginning of the epidemic.
-  }
-  \label{fig:sir}
 \end{figure}

### 2016-05-06 doc/thesis/intro.tex
Put test figures back in

@@
 \subsection{Phylodynamics: linking evolution and epidemiology}
+\label{subsec:phylodynamics}
 
@@
 barabasi1999emergence}. The simplest preferential attachment model is known as
-the \acrlong{BA} (BA) model after its
-inventors~\autocite{barabasi1999emergence}. Under this model, networks are
-formed by starting with a small number $m_0$ of nodes. New nodes are added one
-at a time until there are a total of $N$ in the network. Each time a new node
-is added, $m \geq 1$ edges are added from it to other nodes in the graph. In
-the original formulation, the partners of the new node are chosen with
-probability linearly proportional to their degree. However,
-\citeauthor{barabasi1999emergence} suggest extending the model such that the
-probability of choosing a partner of degree $d$ is proportional to $d^\alpha +
-1$ for some constant $\alpha$, and we use this extension here. When $m = 1$,
-the network takes on the distinctive shape of a tree, that is, it does not
-contain any cycles. Cycles are present in the network for all all other $m$
+the \gls{BA} model after its inventors~\autocite{barabasi1999emergence}. Under
+this model, networks are formed by starting with a small number $m_0$ of nodes.
+New nodes are added one at a time until there are a total of $N$ in the
+network. Each time a new node is added, $m \geq 1$ edges are added from it to
+other nodes in the graph. In the original formulation, the partners of the new
+node are chosen with probability linearly proportional to their degree.
+However, \citeauthor{barabasi1999emergence} suggest extending the model such
+that the probability of choosing a partner of degree $d$ is proportional to
+$d^\alpha + 1$ for some constant $\alpha$, and we use this extension here. When
+$m = 1$, the network takes on the distinctive shape of a tree, that is, it does
+not contain any cycles. Cycles are present in the network for all all other $m$
 values.
@@
 \]
-This distribution is variously referred to as the  \textit{\gls{ABC} target
-distribution} or the \gls{ABC} approximation to the posterior. Note that in
-many formulations, the distance function $\rho$ is defined as $\rho(S(\cdot),
-S(\cdot))$ where $S$ is a function which maps data points into a vector of
-summary statistics. This can be useful if the data are high-dimensional or of a
-complex type, but it is not strictly necessary. For instance, if the data are
-numeric and of low dimension, the distance function may simply be the Euclidian
-distance~\autocite{sisson2007sequential}. For more complex data,
-\textcite{nakagome2013kernel} proposed the use of a kernel function (defined in
-\cref{subsec:treeshape}), an approach they dubbed \defn{kernel-\gls{ABC}}.
+The intuition for why this approximation might hold comes from the fact that, 
+when $\varepsilon = 0$, the intergral on the left is exactly equal to the
+posterior. Thus, by taking $\varepsilon$ small, we should attain something
+close to the posterior.
+
+The distribution $\pi_\varepsilon(\theta, z \mid y)$ is variously referred to
+as the  \textit{\gls{ABC} target distribution} or the \gls{ABC} approximation
+to the posterior. Note that in many formulations, the distance function $\rho$
+is defined as $\rho(S(\cdot), S(\cdot))$ where $S$ is a function which maps
+data points into a vector of summary statistics. This can be useful if the data
+are high-dimensional or of a complex type, but it is not strictly necessary.
+For instance, if the data are numeric and of low dimension, the distance
+function may simply be the Euclidian distance~\autocite{sisson2007sequential}.
+For more complex data, \textcite{nakagome2013kernel} proposed the use of a
+kernel function (defined in \cref{subsec:treeshape}), an approach they dubbed
+\defn{kernel-\gls{ABC}}.
 
@@
 \]
+%\subsection{Convergence properties}
+%
+%The convergence of \gls{ABC} to the true posterior distribution of interest
+%depends on two components: the convergence of the underlying algorithm to the
+%\gls{ABC} target distribution, and the resemblance of this target distribution
+%to the true posterior~\autocite{marin2012approximate}. We shall address these 
+%components one at a time. Rejection \gls{ABC} and \gls{ABC}-\gls{MCMC} use,
+%respectively, rejection sampling and \gls{MCMC} to sample from the target
+%distributions. Both of these methods are well-established and their convergence
+%properties have been extensively studied. Since we do not apply rejection or
+%\gls{MCMC} in this work, we shall restrict our attention to the convergence
+%properties of \gls{SMC}.
+
+%Many results about the rate of convergence of \gls{MCMC} are available, as well
+%as convergence diagnostics. \Gls{SMC} is a much newer method than \gls{MCMC}\ldots

### 2016-05-06 doc/thesis/methods.tex
Put test figures back in

@@
 simulation, to simulate transmission trees on contact networks; the tree
-kernel, to compare simulated to observed transmission trees; and adaptive
-\gls{ABC}-\gls{SMC}, to maintain a population of particles and advance it
-toward the \gls{ABC} target distribution. We give a high-level overview of the
-program here, before describing these three components in detail.
+kernel, to compare transmission trees; and adaptive \gls{ABC}-\gls{SMC}, to
+maintain a population of particles and advance it toward the \gls{ABC} target
+distribution. We give a high-level overview of the program here, before
+describing these three components in detail. \software{Netabc} takes as input
+an estimated transmission tree, which can be derived from a viral phylogeny by
+rooting and time-scaling as described in \cref{subsec:phylodynamics} or
+estimated by other methods~\autocite{cottam2008integrating,
+jombart2011reconstructing, ypma2012unravelling, morelli2012bayesian,
+didelot2014bayesian}. We variously refer to this estimated transmission tree as
+the observed tree, input tree, or true tree.
 
@@
 transmission tree is obtained. Thus, each particle becomes associated with
-several simulated transmission trees. These trees are compared to the observed
+several simulated transmission trees. These trees are compared to the input
 tree using the tree kernel. Particles are weighted according to the similarity
@@
 available at \url{https://github.com/rmcclosk/netabc}.
+
+To check that our implementation of Gillespie simulation was correct, we
+reproduced Figure 1A of \textcite{leventhal2012inferring} (our
+\cref{fig:leventhal}), which plots the unbalancedness of transmission trees
+simulated over four network models at various levels of pathogen
+transmissibility. Our implementation of adaptive ABC-SMC was tested by applying
+it to the same mixture of Gaussians used by \citeauthor{del2012adaptive} to
+demonstrate their method (originally used by~\textcite{sisson2007sequential}).
+We were able to obtain a close approximation to the function (see
+\cref{fig:smctest}), and attained the stopping condition used by the authors in
+a comparable number of steps. To check that the algorithm would converge to a
+bimodal distribution, we also applied it to a mixture of two Gaussians with
+means $\pm$4 and variances 1. The algorithm was able to recover both peaks
+(\cref{fig:smctest2}).
 
@@
 the removal rate of each node was set to 0, and all transmission trees' branch
-lengths were scaled by their mean. For all analyses, the transmission trees'
-branch lengths were scaled by dividing by their mean. We used the
-\textit{igraph} library's implementation of the BA
-model~\autocite{csardi2006igraph} to generate the graphs. The analyses were run
-on Westgrid (\url{https://www.westgrid.ca/}) and a local computer cluster. With
-the exception of our own programs, all analyses were done in \software{R}, and
-all packages listed below are \software{R} packages.
+lengths were scaled by their mean. \textit{igraph} library's implementation of
+the BA model~\autocite{csardi2006igraph} to generate the graphs. The analyses
+were run on Westgrid (\url{https://www.westgrid.ca/}) and a local computer
+cluster. With the exception of our own programs, all analyses were done in
+\software{R}, and all packages listed below are \software{R} packages.
 
@@
 transmission trees were simulated over each using \software{nettree}. These
-trees will be referred to as ``grid trees'', and their associated values ``grid
-values''. Next, one further test tree was simulated with the test value
-\gls{alpha} = 0. Both the grid trees and the test tree had 500 tips, and were
-simulated with the other \gls{BA} parameters set to $N$ = 5000, $m$ = 2, and
-$I$ = 1000. The test tree was compared to each of the grid trees using the tree
-kernel, with the meta-parameters set to $\lambda = 0.3$ and $\sigma = 4$, using
-the \software{treekernel} binary. The median kernel score was calculated for
-each grid value, and the scores were normalized such that the area under the
-curve was equal to 1. The grid value with the highest median kernel score was
-taken as the point estimate for the test value, and a 95\% credible interval
-was obtained using the \software{hpd} function in the \software{TeachingDemos}
-package~\autocite{snow2013teachingdemos}.
+trees will be referred to as ``grid trees''. Next, one further test tree was
+simulated with the test value \gls{alpha} = 0. Both the grid trees and the test
+tree had 500 tips, and were simulated with the other \gls{BA} parameters set to
+$N$ = 5000, $m$ = 2, and $I$ = 1000. The test tree was compared to each of the
+grid trees using the tree kernel, with the meta-parameters set to $\lambda =
+0.3$ and $\sigma = 4$, using the \software{treekernel} binary. The median
+kernel score was calculated for each grid value, and the scores were normalized
+such that the area under the curve was equal to 1. The grid value with the
+highest median kernel score was taken as the point estimate for the test value,
+and a 95\% credible interval was obtained using the \software{hpd} function in
+the \software{TeachingDemos} package~\autocite{snow2013teachingdemos}.
 

### 2016-05-06 doc/thesis/results.Rnw
Put test figures back in

@@
     N.sackin.cor <- N.stats[,cor.test(N, sackin, method="spearman")]
-    stopifnot(alpha.sackin.cor$p.value < 1e-5)
+    stopifnot(alpha.sackin.cor$p.value < 0.05)
+    stopifnot(I.sackin.cor$p.value < 0.05)
+    stopifnot(m.sackin.cor$p.value < 0.05)
+    stopifnot(N.sackin.cor$p.value < 0.05)
 
@@
      $\Sexpr{pp(m.sackin.cor$p.value)}$,
-     $\Sexpr{pp(N.sackin.cor$p.value)}$)
+     $\Sexpr{pp(N.sackin.cor$p.value)}$).
 The ratio of internal to terminal branch lengths was negatively correlated with
-\gls{alpha} and \gls{I}, and positively corelated with \gls{N} and \gls{m}
+\gls{alpha} and \gls{I}, and positively corelated with \gls{m} and \gls{N}
   (Spearman's rho
@@
 significantly between true values of \gls{alpha}
-    (one-way \gls{ANOVA}, $p \Sexpr{pp(av["alpha", p], eq=TRUE)}$). In
+    (one-way \gls{ANOVA}, $p$ $\Sexpr{pp(av["alpha", p], eq=TRUE)}$). In
 particular, errors were lower for the values \gls{alpha} = 1.0 and 1.25 than
@@
   \vspace{6pt}
-  \caption{
-    Point estimates of BA model parameters obtained by running kernel-ABC
-    on simulated phylogenies without training, for simulations with $m = 2$.
-    Dotted lines indicate true values, and limits of the $y$-axes are regions
-    of uniform prior density. (A) Estimates for $\alpha$ and $I$ against their
-    true values in simulations. (B) Estimates for $m$ and $N$, which were held
-    fixed in these simulations, against true values of $\alpha$.
+  \caption[
+    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
+    running \software{netabc} on simulated data, for simulations with \gls{m} =
+    2.
+  ]{
+    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
+    running \software{netabc} on simulated data, for simulations with \gls{m} =
+    2. Dashed lines indicate true values. (A) Estimates of \gls{alpha} and
+    \gls{I} which were varied in these simulations against known values. (B)
+    Estimates of \gls{m} and \gls{N} which were held fixed in these simulations
+    at the values \gls{m} = 2 and \gls{N} = 5000.
   }        

### 2016-05-06 doc/thesis/results.tex
Put test figures back in

@@
      ${<}10^{-5}$,
-     ${<}10^{-5}$)
+     ${<}10^{-5}$).
 The ratio of internal to terminal branch lengths was negatively correlated with
-\gls{alpha} and \gls{I}, and positively corelated with \gls{N} and \gls{m}
+\gls{alpha} and \gls{I}, and positively corelated with \gls{m} and \gls{N}
   (Spearman's rho
@@
 significantly between true values of \gls{alpha}
-    (one-way \gls{ANOVA}, $p {<}10^{-5}$). In
+    (one-way \gls{ANOVA}, $p$ ${<}10^{-5}$). In
 particular, errors were lower for the values \gls{alpha} = 1.0 and 1.25 than
@@
 $I$,
-  1137 
+  1119 
   [662 -
@@
   3 
-  [2 -
+  [1 -
    5];
 $N$,
-  12785 
-  [3861-
-   14945].
+  12644 
+  [3258-
+   14967].
 The approximate posterior distributions for this simulation are shown in
@@
 $\alpha$, 
-  0.37 
-  [0.02 -
+  0.21 
+  [0 -
    0.63];
 $I$,
-  2352 
-  [1423 -
+  2390 
+  [1419 -
    3811];
@@
 $N$,
-  8508 
-  [2636 -
-   14615].
+  8944 
+  [2836 -
+   14647].
 The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both

### 2016-05-08 doc/thesis/conclusion.tex
Fix final spelling and grammar mistakes

@@
 kernel-\gls{ABC}, a likelihood-free method, makes it possible to fit network
-models to phylogenies without calculating intractible likelihoods.
+models to phylogenies without calculating intractable likelihoods.
 
 Although phylodynamic methods have been developed to fit a wide variety of
-epidemiological models to phylogenetic data~\autocite{volz2012complex,
-rasmussen2014phylodynamic}, our method is the first (to our knowledge) which
-can fit models that do not assume panmixia. We believe this capability will be
-of broad interest to the molecular evolution and epidemiology community, as it
-widens the field of epidemiological parameters which may be investigated
-through viral sequence data. In addition, the characterization of local contact
-networks could be valuable from a public health perspective, such as for
-investigating optimal vaccination strategies~\autocite{keeling2005networks,
-peng2013vaccination, ma2013importance, rushmore2014network}. This information
-could assist in curtailing current epimemics, as well as preventing future
-epidemics of different diseases over the same contact network.
+epidemiological models to phylogenetic data assuming homogeneous
+mixing~\autocite{volz2012complex, rasmussen2014phylodynamic}, our method is
+able to fit models not requiring this assumption. We believe this capability
+will be of broad interest to the molecular evolution and epidemiology
+community, as it widens the field of epidemiological parameters which may be
+investigated through viral sequence data. In addition, the characterization of
+local contact networks could be valuable from a public health perspective, such
+as for investigating optimal vaccination
+strategies~\autocite{keeling2005networks, peng2013vaccination,
+ma2013importance, rushmore2014network}. This information could assist in
+curtailing current epidemics, as well as preventing future epidemics of
+different diseases over the same contact network.
 
-The paricular model we have investigated uses a preferential attachment
+{\color{blue}
+The particular model we have investigated uses a preferential attachment
 mechanism to generate scale-free networks resembling real-world social and
 sexual networks~\autocite{liljeros2001web, schneeberger2004scale,
-colgate1989risk}. Of the four parameters we considered, the pereferential
+colgate1989risk}. Of the four parameters we considered, the preferential
 attachment power $\alpha$ was the most readily estimable. Estimating $\alpha$
@@
 In closing, \software{netabc} combines phylodynamics, contact network
-epidemiology, approximate Bayesian computation, and sequentiial Monte Carlo 
+epidemiology, approximate Bayesian computation, and sequential Monte Carlo 
 to provide a source of insight into network structures complementary to
 traditional epidemiology.
+}

### 2016-05-08 doc/thesis/discussion.tex
Fix final spelling and grammar mistakes

@@
 
-Several alternative network models and modeling frameworks have been developed
+Several alternative network models and modelling frameworks have been developed
 which may provide useful future targets for kernel-\gls{ABC}. Waring
@@
 type of preferential attachment model which permit a subset of attachments to
-be formed non-prefentially. These models were used by
+be formed non-preferentially. These models were used by
 \textcite{brown2011transmission} to characterize the transmission network in
@@
 nodes which are involved in a large number of transmissions, resulting in a
-highly unbalanced ladder-like tree structure (\cref{fig:alphatrees}). There was
-no observeable bias in the estimates of \gls{alpha} obtained with
+highly unbalanced ladder-like tree structure (\cref{fig:alphatrees}). 
+{\color{blue}
+There was no observable bias in the estimates of \gls{alpha} obtained with
 \software{netabc}, however the variation in these estimates was higher for
@@
 transmission trees as well. 
+}
 %The accuracy of \gls{alpha} estimates also increased when the value of the
@@
 accuracy of the \gls{nltt} in these situations
-(\cref{fig:rsquared,fig:Ncrossv}) coroborates this hypothesis, as the
-\gls{nltt} uses only information about the coalescence times. When all \gls{BA}
+(\cref{fig:rsquared,fig:Ncrossv}) corroborates this hypothesis, as the
+\gls{nltt} uses only information about the coalescence times. 
+{\color{blue}When all \gls{BA}
 parameters were simultaneously estimated with kernel-\gls{ABC}, \gls{N} was
@@
 estimates of \gls{I} were accurate and unbiased (\cref{fig:gridptI}).
+}
 
@@
 least four, which may be unrealistically high for sexual networks. The fact
-that the estimated values of $\alpha$ differed substatially for three datasets
-depending on whether or not $m = 1$ was allowed by the prior is futher evidence
+that the estimated values of $\alpha$ differed substantially for three datasets
+depending on whether or not $m = 1$ was allowed by the prior is further evidence
 of this potential misspecification. However, we note that for two of the

### 2016-05-08 doc/thesis/intro.tex
Fix final spelling and grammar mistakes

@@
 quantitative measure of evolution, such as the number of differences between
-the DNA sequences of an ancestor its descendant, or the difference in average
-body mass or height. A phylogeny with branch lengths in calendar time units is
-often referred to as \defn{time-scaled}. In a time-scaled phylogeny, the
-internal nodes can be mapped onto a timeline by using the tips of the tree,
-which usually correspond to the present day, as a reference
-point~\autocite{nee1992tempo}. The corresponding points on the timeline are
+the DNA sequences of an ancestor and its descendant, or the difference in
+average body mass or height. A phylogeny with branch lengths in calendar time
+units is often referred to as \defn{time-scaled}. In a time-scaled phylogeny,
+the internal nodes can be mapped onto a timeline by using the tips of the tree,
+which usually correspond to the present day, reference
+points~\autocite{nee1992tempo}. The corresponding points on the timeline are
 called \defn{branching times}, and the rate of their accumulation is referred
@@
 record an epidemiological process (transmission). The tips of a transmission
-tree represent the removal of infected hosts, while internal nodes correspond
-to transmissions from one host to another. Transmission trees generally have
-branch lengths in units of calendar time, with branching times indicating times
-of transmission. The root of a transmission tree corresponds to the initially
-infected patient who introduced the epidemic into the network, also known as
-the \defn{index case}. The internal nodes may be labelled with the donor of the
-transmission pair, if this is known. The tips of the tree, rather than being
-fixed at the present day, are placed at the time at which the individual was
-removed from the epidemic, such as by death, recovery, isolation, behaviour
-change, or migration~\autocite{stadler2013uncovering}. Consequently, the
-transmission tree may not be ultrametric, but may have tips located at varying
-distances from the root. Such trees are said to have \defn{heterochronous}
-taxa~\autocite{drummond2003measurably}, in contrast to the \defn{isochronous}
-taxa found in most phylogenies of macro-organisms. A transmission tree is
-illustrated in \cref{fig:contactnet} (right).
-
-Each infected individual in an epidemic may appear in the transmission tree
-more than once. This is different from the transmission \emph{network}, in
+tree represent the removal by sampling of infected hosts, while internal nodes
+correspond to transmissions from one host to another. Transmission trees
+generally have branch lengths in units of calendar time, with branching times
+indicating times of transmission. The root of a transmission tree corresponds
+to the initially infected patient who introduced the epidemic into the network,
+also known as the \defn{index case}. The internal nodes may be labelled with
+the donor of the transmission pair, if this is known. The tips of the tree,
+rather than being fixed at the present day, are placed at the time at which the
+individual was removed from the epidemic, such as by death, recovery,
+isolation, behaviour change, or migration~\autocite{stadler2013uncovering}.
+Consequently, the transmission tree may not be ultrametric, but may have tips
+located at varying distances from the root. Such trees are said to have
+\defn{heterochronous} taxa~\autocite{drummond2003measurably}, in contrast to
+the \defn{isochronous} taxa found in most phylogenies of macro-organisms. A
+transmission tree is illustrated in \cref{fig:contactnet} (right). 
+{\color{blue} The object on the right of the figure is called a \defn{contact
+network}, which depicts the entire susceptible population along with all
+possible routes of disease transmission. Contact networks, and their
+relationships to transmission trees, will be discussed further in
+\cref{sec:contactnet}}.
+
+Each infected individual in an epidemic may appear at nodes of the transmission
+tree more than once. This is different from the transmission \emph{network}, in
 which each infected individual appears exactly once, and edges are in
@@
 \]
-where $\norm{\cdot}_2$ is the Euclidian norm and $\sigma$ is a variance
+where $\norm{\cdot}_2$ is the Euclidean norm and $\sigma$ is a variance
 parameter. The tree kernel is defined as 
@@
 
-\Gls{SIS} is one type of \gls{SMC} method, whose aim is to sample from a
-distribution $\pi$ on an high-dimensional space, say $\pi(\vec{x}) = \pi(x_1,
-\ldots, x_d)$. The basis of \gls{SIS} is \gls{IS}, which is a method of
-estimating summary statistics of distributions which are known only up to a
-normalizing constant, and therefore cannot be sampled from directly. That is,
-if $\pi$ is such a distribution and $f$ is any real-valued function, \gls{IS}
-is concerned with estimating
+\Gls{SIS}~\autocite{gordon1993novel} is one type of \gls{SMC} method, whose aim
+is to sample from a distribution $\pi$ on an high-dimensional space, say
+$\pi(\vec{x}) = \pi(x_1, \ldots, x_d)$. The basis of \gls{SIS} is \gls{IS},
+which is a method of estimating summary statistics of distributions which are
+known only up to a normalizing constant, and therefore cannot be sampled from
+directly. That is, if $\pi$ is such a distribution and $f$ is any real-valued
+function, \gls{IS} is concerned with estimating
 \[
@@
 where $K_i$ is a Markov kernel and the integral is over the space on which the
-$\pi_i$ are defined. The choice of $K_i$ should be based on the percieved
+$\pi_i$ are defined. The choice of $K_i$ should be based on the perceived
 relationship between $\pi_{i-1}$ and $\pi_i$. \textcite{del2006sequential}
@@
 where $q(x, x')$ is a proposal function such as a Gaussian distribution
-centered at $x$ (see \cref{subsec:mfit}).
+centred at $x$ (see \cref{subsec:mfit}).
 
 Although this method of building up $\eta$ appears straightforward, the
-drawback is that the importance distribution itself becomes intractible. In
+drawback is that the importance distribution itself becomes intractable. In
 particular, evaluating $\eta_i(x)$ involves a $i$-dimensional integral of the
@@
 These are all the ingredients we need to apply \gls{SIS}. The sequences of
-kernels $L$ and $K$ should be chosen based on the problem at hand to inimize
+kernels $L$ and $K$ should be chosen based on the problem at hand to minimize
 the variance in the importance weights as well as possible. For a fixed choice
@@
 The intuition for why this approximation might hold comes from the fact that, 
-when $\varepsilon = 0$, the intergral on the left is exactly equal to the
+when $\varepsilon = 0$, the integral on the left is exactly equal to the
 posterior. Thus, by taking $\varepsilon$ small, we should attain something
@@
 For instance, if the data are numeric and of low dimension, the distance
-function may simply be the Euclidian distance~\autocite{sisson2007sequential}.
+function may simply be the Euclidean distance~\autocite{sisson2007sequential}.
 For more complex data, \textcite{nakagome2013kernel} proposed the use of a
@@
 \]
+
 %\subsection{Convergence properties}
@@
 %\gls{ABC} target distribution, and the resemblance of this target distribution
-%to the true posterior~\autocite{marin2012approximate}. We shall address these 
-%components one at a time. Rejection \gls{ABC} and \gls{ABC}-\gls{MCMC} use,
-%respectively, rejection sampling and \gls{MCMC} to sample from the target
-%distributions. Both of these methods are well-established and their convergence
-%properties have been extensively studied. Since we do not apply rejection or
-%\gls{MCMC} in this work, we shall restrict our attention to the convergence
-%properties of \gls{SMC}.
-
-%Many results about the rate of convergence of \gls{MCMC} are available, as well
-%as convergence diagnostics. \Gls{SMC} is a much newer method than \gls{MCMC}\ldots
+%to the true
+%posterior~\autocite{marin2012approximate,fearnhead2012constructing}. We shall
+%address these components one at a time. Rejection \gls{ABC} and
+%\gls{ABC}-\gls{MCMC} use, respectively, rejection sampling and \gls{MCMC} to
+%sample from the target distributions. Both of these methods are
+%well-established and their convergence properties have been extensively
+%studied. Since we do not apply rejection or \gls{MCMC} in this work, we shall
+%restrict our attention to the convergence properties of \gls{SMC}.

### 2016-05-08 doc/thesis/methods.tex
Fix final spelling and grammar mistakes

@@
 
-The simulation of epidemics, and the corresponding transmission, trees over
+The simulation of epidemics, and the corresponding transmission trees, over
 contact networks is performed in \software{netabc} using the Gillespie
@@
 simulation progresses. When a node $v$ becomes infected, it is deleted from $S$
-and added to $I$. Any formerly discordant edges in $\in(v)$ are deleted from
+and added to $I$. Any formerly discordant edges in $\inc(v)$ are deleted from
 $D$, and edges in $\out(v)$ to nodes in $S$ are added to $D$. If $v$ is later

### 2016-05-08 doc/thesis/objective.tex
Fix final spelling and grammar mistakes

@@
 that any two individuals in the same compartment are equally likely to come
-into contact causing transmission. Although this provides a reasonable
-approximation in many cases~\autocite{anderson1992infectious}, the error
-introduced by assuming a panmictic population can be substantial when
-significant contact heterogeneity exists in the underlying
+into contact making transmission possible at some predefined rate. Although
+this provides a reasonable approximation in many
+cases~\autocite{anderson1992infectious}, the error introduced by assuming a
+panmictic population can be substantial when significant contact heterogeneity
+exists in the underlying
 population~\autocite{bansal2007individual,barthelemy2005dynamical,keeling2005networks}.
@@
 network's degree distribution~\autocite{peng2013vaccination, ma2013importance}.
-Phylodynamic methods have been used to fit many different types of model to
-phylogenetic data~\autocite{pybus2009evolutionary,volz2013viral}, but as far as
-we know, no methods have yet been developed to fit contact network models. The
-primary objective of this work is to develop such a method.
+Phylodynamic methods have been used to fit many different types of models to
+phylogenetic data~\autocite{pybus2009evolutionary,volz2013viral}, but these
+models generally assume a panmictic population. The primary objective of this
+work is to develop a method to fit contact network models in a phylodynamic
+framework.
 

### 2016-05-08 doc/thesis/preface.tex
Fix final spelling and grammar mistakes

@@
 with kernel-ABC'' was given based on chapter 2 to the 23rd HIV Dynamics and
-Evolution meeting on April 25, in Woods Hole, Massachussets, USA (the
+Evolution meeting on April 25, in Woods Hole, Massachusetts, USA (the
 presentation was delivered remotely). A poster based on chapter 2 entitled
@@
 (H07-02559).
-
-Throughout this work, the pronouns \textit{we} and \textit{our} refer to
-Rosemary M. McCloskey unless otherwise stated.

### 2016-05-08 doc/thesis/results.Rnw
Fix final spelling and grammar mistakes

@@
 The ratio of internal to terminal branch lengths was negatively correlated with
-\gls{alpha} and \gls{I}, and positively corelated with \gls{m} and \gls{N}
+\gls{alpha} and \gls{I}, and positively correlated with \gls{m} and \gls{N}
   (Spearman's rho
@@
 The highest density intervals obtained for all parameters were extremely wide,
-occcupying $>$75\% of the grid in all cases (\cref{fig:gridest}).
+occupying $>$75\% of the grid in all cases (\cref{fig:gridest}).
 
@@
      for $N > 3000$),
-and this diffrence was significant
+and this difference was significant
     (Wilcoxon rank-sum test, $p \Sexpr{pp(N.test$p.value, eq=TRUE)}$,
@@
 
+    # one-way ANOVAs
+    stopifnot(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"] < 0.06)
+    stopifnot(anova(lm(alpha_error~factor(true_I), d))[1, "Pr(>F)"] > 0.05)
+    stopifnot(anova(lm(I_error~true_m, d))[1, "Pr(>F)"] > 0.05)
+    stopifnot(anova(lm(m_error~factor(true_alpha), d))[1, "Pr(>F)"] > 0.05)
+    stopifnot(anova(lm(m_error~factor(true_I), d))[1, "Pr(>F)"] > 0.05)
+
+    stopifnot(anova(lm(N_error~factor(true_alpha), d))[1, "Pr(>F)"] > 0.05)
+    stopifnot(anova(lm(N_error~factor(true_I), d))[1, "Pr(>F)"] > 0.05)
+    stopifnot(anova(lm(N_error~true_m, d))[1, "Pr(>F)"] > 0.05)
+
     alpha.m.cor <- cor.test(d[,alpha_error], d[,true_m], method="spearman")
@@
 the threshold of statistical significance
-    ($p \Sexpr{pp(alpha.av["factor(true_m)", "Pr(>F)"])}$),
-but did not vary across the true values of $N$ or $I$ (both one-way ANOVA).
-Estimates for $I$ were relatively accurate, with point estimate errors of
+    (one-way ANOVA, $p 
+    \Sexpr{pp(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"], eq=TRUE)}$),
+but did not vary across the true value of $I$ (one-way ANOVA). Estimates for
+$I$ were relatively accurate, with point estimate errors of
     \Sexpr{d[,round(median(I_error))]} 
@@
   \input{\tablepath/abc-hpd.tex}
-  \caption{
+  \caption[
+      Average maximum \textit{a posteriori} point estimates and 95\% highest
+      posterior density (HPD) interval widths for BA model parameter estimates
+      obtained with kernel-ABC.
+  ]{
       Average maximum \textit{a posteriori} point estimates and 95\% highest

### 2016-05-08 doc/thesis/results.tex
Fix final spelling and grammar mistakes

@@
      for $N > 3000$),
-and this diffrence was significant
+and this difference was significant
     (Wilcoxon rank-sum test, $p {<}10^{-5}$,
@@
 the threshold of statistical significance
-    ($p 0.05$),
-but did not vary across the true values of $N$ or $I$ (both one-way ANOVA).
-Estimates for $I$ were relatively accurate, with point estimate errors of
+    (one-way ANOVA, $p 
+    =0.05$),
+but did not vary across the true value of $I$ (one-way ANOVA). Estimates for
+$I$ were relatively accurate, with point estimate errors of
     395 
@@
 $I$,
-  1119 
+  1115 
   [662 -
@@
 $N$,
-  12644 
-  [3258-
-   14967].
+  12680 
+  [3377-
+   14933].
 The approximate posterior distributions for this simulation are shown in
@@
 $\alpha$, 
-  0.21 
+  0.08 
   [0 -
-   0.63];
+   0.62];
 $I$,
-  2390 
-  [1419 -
-   3811];
+  2391 
+  [1379 -
+   3860];
 $m$,
@@
 $N$,
-  8944 
-  [2836 -
-   14647].
+  10675 
+  [2881 -
+   14788].
 The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both

### 2016-05-08 doc/thesis/discussion.tex
Highlight new things

@@
 use kernel-\gls{ABC} to fit a model which relaxes some or all of these
-assumptions, which may be a fruitful avenue for future investigation. Despite
-the possible misspecification, our estimates of the power law exponent $\gamma$
-were within the range of values reported in the literature (\cref{tab:gamma}).
+assumptions, which may be a fruitful avenue for future investigation. 
+{\color{blue}Despite the possible misspecification, our estimates of the power
+law exponent $\gamma$ were within the range of values reported in the
+literature (\cref{tab:gamma}).}

### 2016-05-08 doc/thesis/preface.tex
Highlight new things

@@
 with kernel-ABC'' was given based on chapter 2 to the 23rd HIV Dynamics and
-Evolution meeting on April 25, in Woods Hole, Massachusetts, USA (the
-presentation was delivered remotely). A poster based on chapter 2 entitled
-``Likelihood-free estimation of contact network parameters from viral
-phylogenies'' was presented at the Intelligent Systems for Molecular Biology
-meeting in Orlando, Florida, USA.
+Evolution meeting on April 25, 2016, in Woods Hole, Massachusetts, USA (the
+presentation was delivered remotely). 
+%A poster based on chapter 2 entitled
+%``Likelihood-free estimation of contact network parameters from viral
+%phylogenies'' was presented at the Intelligent Systems for Molecular Biology
+%meeting in Orlando, Florida, USA.
 

### 2016-05-08 doc/thesis/results.Rnw
Highlight new things

@@
 
+{\color{blue}
 To make our analyses comparable to the existing network literature, we
@@
 \end{table}
+}

### 2016-05-08 doc/thesis/results.tex
Highlight new things

@@
 $I$,
-  1115 
+  1142 
   [662 -
@@
 $N$,
-  12680 
-  [3377-
-   14933].
+  12602 
+  [3861-
+   14977].
 The approximate posterior distributions for this simulation are shown in
@@
 $\alpha$, 
-  0.08 
+  0.22 
   [0 -
-   0.62];
+   0.61];
 $I$,
   2391 
-  [1379 -
-   3860];
+  [1464 -
+   3838];
 $m$,
@@
 $N$,
-  10675 
+  8725 
   [2881 -
-   14788].
+   14780].
 The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both

### 2016-05-10 doc/thesis/acknowledgements.tex
Underline changes for printing

@@
 
-I am grateful for the mentorship of Dr. Sara Mostafavi and Dr. Ryan Morin, who
-accepted me into their labs and always had time for my many questions.  Thanks
-also to Dr. Marni Mishna and Dr. Cedric Chauve, for starting me on this path.
+I am grateful for the mentorship of my committee members Dr. Alexandre
+Bouchard-C\^ot\'e, Dr. Richard Harrigan, and Dr. Sarah Otto. Thanks also to my
+rotation supervisors Dr. Sara Mostafavi and Dr. Ryan Morin, who accepted me
+into their labs and always had time for my many questions.
 
 I owe thanks to my fellow graduate students of the bioinformatics training
-program. Special thanks go to Ogan for keeping me on an even keel when times
-got tough.
+program, especially to Ogan.
 

### 2016-05-10 doc/thesis/conclusion.tex
Underline changes for printing

@@
 
-{\color{blue}
+\uline{
 The particular model we have investigated uses a preferential attachment
 mechanism to generate scale-free networks resembling real-world social and
-sexual networks~\autocite{liljeros2001web, schneeberger2004scale,
-colgate1989risk}. Of the four parameters we considered, the preferential
+sexual networks}~\autocite{liljeros2001web, schneeberger2004scale,
+colgate1989risk}. \uline{Of the four parameters we considered, the preferential
 attachment power $\alpha$ was the most readily estimable. Estimating $\alpha$
@@
 distribution, although approaches such as respondent-driven
-sampling~\autocite{heckathorn1997respondent} may be effective.
+sampling~\autocite{heckathorn1997respondent} may be effective. }
 
+\uline{
 In closing, \software{netabc} combines phylodynamics, contact network

### 2016-05-10 doc/thesis/discussion.tex
Underline changes for printing

--- 2016-05-08 doc/thesis/discussion.tex
@@
 input, rather than a viral phylogeny. In reality, true transmission trees are
-not available and must be approximated, often by way of a viral phylogeny.
-Although this has been demonstrated to be a fair
+not available and must be estimated; these estimates are often based on the
+viral phylogeny. Although this has been demonstrated to be a fair
 approximation~\autocite[e.g.][]{leitner1996accurate}, and is frequently used in
@@
 phylogeny and transmission tree can differ
-significantly~\autocite{ypma2013relating} due to within-host evolution and the
-sampling process. We have left the estimation of a transmission tree up to the
-user. In theory, it is possible to incorporate the process by which a viral
-phylogeny is generated along with a transmission tree into our method, for
-example by simulating within-host dynamics. Although this may be an avenue for
-future extension, we felt that it would obscure the primary purpose of this
-work, which is to study contact network parameters. In addition, there are a
-number of different methods available for inferring transmission
-trees~\autocite{didelot2014bayesian, ypma2012unravelling,
-jombart2011reconstructing, cottam2008integrating}, some of which incorporate
-geographic and/or epidemiological data not accommodated by our method. We
-therefore felt it would be best to allow researchers to use their own preferred
-method of constructing a transmission tree.
+significantly~\autocite{ypma2013relating, hall2015epidemic} due to within-host
+evolution and the sampling process. We have left the estimation of a
+transmission tree up to the user. In theory, it is possible to incorporate the
+process by which a viral phylogeny is generated along with a transmission tree
+into our method, for example by simulating within-host dynamics. Although this
+may be an avenue for future extension, we felt that it would obscure the
+primary purpose of this work, which is to study contact network parameters. In
+addition, there are a number of different methods available for inferring
+transmission trees~\autocite{didelot2014bayesian, ypma2012unravelling,
+jombart2011reconstructing, cottam2008integrating, hall2015epidemic}, some of
+which incorporate geographic and/or epidemiological data not accommodated by
+our method. We therefore felt it would be best to allow researchers to use
+their own preferred method of constructing a transmission tree.
 
@@
 highly unbalanced ladder-like tree structure (\cref{fig:alphatrees}). 
-{\color{blue}
+\uline{
 There was no observable bias in the estimates of \gls{alpha} obtained with
 \software{netabc}, however the variation in these estimates was higher for
-$\alpha < 1$ than for $\alpha \geq 1$ (\cref{fig:abcptm2,tab:abchpd}). The
+$\alpha < 1$ than for $\alpha \geq 1$} (\cref{fig:abcptm2,tab:abchpd}). \uline{The
 relationship between \gls{alpha} and the power law exponent \gls{gamma} may
-explain this result (\cref{fig:gamma}). The \gls{gamma} values associated with
+explain this result} (\cref{fig:gamma}).\uline{ The \gls{gamma} values associated with
 \gls{alpha} = 0 and \gls{alpha} = 0.5 are nearly identical (about 2.28 for
@@
 \gls{nltt} uses only information about the coalescence times. 
-{\color{blue}When all \gls{BA}
+\uline{When all \gls{BA}
 parameters were simultaneously estimated with kernel-\gls{ABC}, \gls{N} was
 nearly always over-estimated by approximately a factor of two
-(\cref{fig:abcptm2,tab:abchpd}). One factor which may have contributed to this
+}(\cref{fig:abcptm2,tab:abchpd}). \uline{One factor which may have contributed to this
 bias was our choice of prior distribution. Since the prior for \gls{I} and
@@
 dynamics of the \gls{SI} model and the coalescent process.
+}
 
+\uline{
 \gls{I} and \gls{N} were both systematically over-estimated by
@@
 growth rates of the epidemics at the time of sampling. As shown in
-\cref{fig:derivs}, there are bands along which both derivatives are similar
+}\cref{fig:derivs},\uline{there are bands along which both derivatives are similar
 which contain the values we tested. These bands span mostly higher values of
@@
 vary substantially along the \gls{I}-axis, which explains why the grid search
-estimates of \gls{I} were accurate and unbiased (\cref{fig:gridptI}).
-}
+estimates of \gls{I} were accurate and unbiased} (\cref{fig:gridptI}).
 
@@
 use kernel-\gls{ABC} to fit a model which relaxes some or all of these
-assumptions, which may be a fruitful avenue for future investigation. Despite
-the possible misspecification, our estimates of the power law exponent $\gamma$
-were within the range of values reported in the literature (\cref{tab:gamma}).
+assumptions, which may be a fruitful avenue for future investigation. 
+\uline{Despite the possible misspecification, our estimates of the power
+law exponent $\gamma$ were within the range of values reported in the
+literature} (\cref{tab:gamma}).

### 2016-05-10 doc/thesis/intro.tex
Underline changes for printing

@@
 transmission tree is illustrated in \cref{fig:contactnet} (right). 
-{\color{blue} The object on the right of the figure is called a \defn{contact
+\uline{The object on the right of the figure is called a \defn{contact
 network}, which depicts the entire susceptible population along with all
 possible routes of disease transmission. Contact networks, and their
-relationships to transmission trees, will be discussed further in
-\cref{sec:contactnet}}.
+relationships to transmission trees, will be discussed further in}
+\cref{sec:contactnet}.
 
@@
 drummond2003inference}, which we apply in this work, and least-square
-dating~\autocite{to2015fast}. Alternatively, the tree may be rooted separately
+dating~\autocite{to2016fast}. Alternatively, the tree may be rooted separately
 with an outgroup~\autocite{li1988rates} before time-scaling.
@@
 research~\autocite{cottam2008integrating, jombart2011reconstructing,
-ypma2012unravelling, morelli2012bayesian, didelot2014bayesian}.
+ypma2012unravelling, morelli2012bayesian, didelot2014bayesian, hall2015epidemic}.
+\uline{For example, \textcite{hall2015epidemic} developed a Bayesian
+method to jointly estimate a transmission tree and viral phylogeny by combining
+models of agent-based transmission, within-host population dynamics, and
+sequence evolution.}
 
@@
 networks are a particular type of \defn{social
-network}~\autocite{moreno1953shall, barnes1954class}, which is a network in
+network}~\autocite{moreno1934shall, barnes1954class}, which is a network in
 which edges may represent any kind of social or economic relationship. Social
@@
 not contain any cycles. Cycles are present in the network for all all other $m$
-values.
+values. \uline{Examples of \gls{BA} networks with three different values of the
+preferential attachment power \gls{alpha} are shown in} \cref{fig:baeg}.
+
+\begin{figure}
+  \includegraphics[width=\textwidth]{pa-example}
+  \caption[
+    Examples of Barab\'asi-Albert networks with preferential attachment power
+    $\alpha$ = 0, 1, and 2.
+  ]{
+    Examples of Barab\'asi-Albert networks with preferential attachment power
+    $\alpha$ = 0 (left), 1 (centre), and 2 (right). All networks have $N = 50$
+    nodes and were constructed with $m = 2$ edges per vertex. When $\alpha =
+    0$, attachments are formed at random and most nodes have low degree. When
+    $\alpha = 1$, preferential attachment is linear and several higher-degree
+    nodes are observable. When $\alpha = 2$, preferential attachment is
+    quadratic and nearly every vertex is attached to a small number of hub
+    nodes.
+  }
+  \label{fig:baeg}
+\end{figure}
 
@@
 moderate correspondence between the two in some cases.
+\uline{\textcite{goodreau2006assessing} combined a dynamic contact
+network model with a model of within-host viral evolution to simulate viral
+phylogenies over eight types of contact network. Estimates of prevalence and 
+effective population size were calculated for each simulated phylogeny under
+three models of epidemic growth. The author found that estimates for networks
+with a small high-risk subgroup and networks involving commercial sex workers
+were substantially different than estimates for random networks or networks
+with segregated equal-risk groups.}
 

### 2016-05-10 doc/thesis/methods.tex
Highlight new things
Underline changes for printing

@@
 
+\uline{
 Empirical studies of contact networks often report the exponent $\gamma$ of the
@@
 $\gamma$ values was taken as a point estimate for the associated dataset.
+}
 
+\uline{
 For all datasets, we used the priors $\alpha$ $\sim$ Uniform(0, 2) and $N$ and
 $I$ jointly uniform on the region \{$n \leq N \leq 10000$, $n \leq I \leq
-10000$, $I \leq N$\}, where $n$ is the number of tips in the tree (see
-\cref{tab:data}). Since the value $m = 1$ produces networks with no cycles,
+10000$, $I \leq N$\}, where $n$ is the number of tips in the tree} (see
+\cref{tab:data}). \uline{Since the value $m = 1$ produces networks with no cycles,
 which we considered fairly implausible, we ran one analysis with the prior $m
@@
 increase the accuracy of the estimated posterior. This was computationally
-feasible due to the small number of runs required for this analysis.
+feasible due to the small number of runs required for this analysis.}

### 2016-05-10 doc/thesis/results.Rnw
Underline changes for printing

--- 2016-05-08 doc/thesis/results.Rnw
@@
 
+\uline{
 To make our analyses comparable to the existing network literature, we
 estimated values of the power law exponent $\gamma$ for each of the datasets 
-investigated. The results are summarized in \cref{tab:gamma}. All of the
-estimated exponents were in the range $2 \leq \gamma \leq 2.5$.
+investigated. The results are summarized in} \cref{tab:gamma}. \uline{All of the
+estimated exponents were in the range $2 \leq \gamma \leq 2.5$, which is on the
+lower end of the range $2 \leq \gamma \leq 4$ reported in the literature.
+}
 

### 2016-05-10 doc/thesis/results.tex
Underline changes for printing

--- 2016-05-08 doc/thesis/results.tex
@@
   [0.6 -
-   1.16];
+   1.14];
 $I$,
-  1115 
-  [662 -
+  1120 
+  [666 -
    4455];
@@
 $N$,
-  12680 
-  [3377-
-   14933].
+  12768 
+  [4021-
+   14967].
 The approximate posterior distributions for this simulation are shown in
@@
 $\alpha$, 
-  0.08 
+  0.21 
   [0 -
-   0.62];
+   0.63];
 $I$,
-  2391 
-  [1379 -
-   3860];
+  2399 
+  [1464 -
+   3811];
 $m$,
@@
 $N$,
-  10675 
-  [2881 -
-   14788].
+  13817 
+  [3080 -
+   14970].
 The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both

### 2016-05-17 doc/thesis/abstract.tex
Change kernel-ABC to kernel-assisted ABC

@@
 explicitly representing each contact which could possibly lead to a
-transmission. We developed a method based on kernel approximate Bayesian
-computation (kernel-ABC) for estimating structural parameters of the contact
+transmission. We developed a method based on kernel-assisted approximate
+Bayesian computation (ABC) for estimating structural parameters of the contact
 network underlying an observed viral phylogeny. The method combines adaptive
@@
 estimated. On the other hand, the mean degree of the network, as well as the
-total number of nodes, were not estimable with kernel-ABC. We observed
-substantial heterogeneity in the parameter estimates on real datasets, with
-point estimates for the preferential attachment power ranging from 0.06 to 1.05.
-These results underscore the importance of considering contact structures when
-performing phylodynamic inference. Our method offers the potential to
-quantitatively investigate the contact network structure underlying viral
-epidemics.
+total number of nodes, appeared to be weakly or non-identifiable with
+kernel-assisted ABC. We observed substantial heterogeneity in the parameter
+estimates on real datasets, with point estimates for the preferential
+attachment power ranging from 0.06 to 1.05. These results underscore the
+importance of considering contact structures when performing phylodynamic
+inference. Our method offers the potential to quantitatively investigate the
+contact network structure underlying viral epidemics.

### 2016-05-17 doc/thesis/appendix.Rnw
Fix final spelling and grammar mistakes
Change kernel-ABC to kernel-assisted ABC

@@
     pathogen transmissibility varying from 0 to 1. Sackin's index was
-    calculated for each simulated transmission tree.
+    calculated for each simulated transmission tree. Lines indicate median
+    Sackin's index values, and shaded areas are interquartile ranges.
   }
@@
     \caption[Approximate marginal posterior distributions of BA model
-        parameters obtained using kernel-ABC for a network with heterogeneous
+        parameters obtained using kernel-assisted ABC for a network with heterogeneous
         node behaviour.]
@@
         Approximate marginal posterior distributions of Barab\'asi-Albert
-        model parameters obtained using kernel-ABC for a network with
+        model parameters obtained using kernel-assisted ABC for a network with
         heterogeneous node behaviour. Half of the nodes were attached with
@@
     \caption[Approximate marginal posterior distributions of Barab\'asi-Albert
-        model parameters obtained using kernel-ABC for a network with
+        model parameters obtained using kernel-assisted ABC for a network with
         peer-driven sampling.]
@@
         Approximate marginal posterior distributions of Barab\'asi-Albert
-        model parameters obtained using kernel-ABC for a network with
+        model parameters obtained using kernel-assisted ABC for a network with
         peer-driven sampling. An epidemic was simulated in the usual fashion,
@@
   \caption[
-      Approximate marginal poosterior distributions of BA model parameters for
+      Approximate marginal posterior distributions of BA model parameters for
       \textcite{cuevas2009hiv} data. 
   ]{
-      Approximate marginal poosterior distributions of BA model parameters for
+      Approximate marginal posterior distributions of BA model parameters for
       \textcite{cuevas2009hiv} data. Vertical lines indicate maximum \textit{a

### 2016-05-17 doc/thesis/conclusion.tex
Change kernel-ABC to kernel-assisted ABC

--- 2016-05-08 doc/thesis/conclusion.tex
@@
 
-{\color{blue}
 The particular model we have investigated uses a preferential attachment
@@
 traditional epidemiology.
-}

### 2016-05-17 doc/thesis/discussion.tex
Change kernel-ABC to kernel-assisted ABC

--- 2016-05-08 doc/thesis/discussion.tex
@@
 Several alternative network models and modelling frameworks have been developed
-which may provide useful future targets for kernel-\gls{ABC}. Waring
+which may provide useful future targets for kernel-assisted \gls{ABC}. Waring
 models~\autocite{irwin1963place,handcock2004likelihood} are a more flexible
@@
 intensive to integrate into \software{netabc} as it currently stands. To fit
-\gls{ERGM} with kernel-\gls{ABC}, one possibility would be to consider the
+\gls{ERGM} with kernel-assisted \gls{ABC}, one possibility would be to consider the
 network itself as a parameter to be modified by the \gls{MCMC} kernel.
@@
 %a very large system of \glspl{ODE}, which may be computationally prohibitive
-%given the number of simulations required for kernel-\gls{ABC}.
+%given the number of simulations required for kernel-assisted \gls{ABC}.
 
@@
 context~\autocite{welch2011statistical, bansal2007individual}. In principle,
-kernel-\gls{ABC} could be adapted to dynamic contact networks by using a method
+kernel-assisted \gls{ABC} could be adapted to dynamic contact networks by using a method
 such as that developed by \textcite{robinson2012dynamics} to simulate such a
@@
 highly unbalanced ladder-like tree structure (\cref{fig:alphatrees}). 
-{\color{blue}
 There was no observable bias in the estimates of \gls{alpha} obtained with
@@
 transmission trees as well. 
-}
 %The accuracy of \gls{alpha} estimates also increased when the value of the
@@
 The \gls{I} parameter, representing the prevalence at the time of sampling, was
-also generally estimable, although it was slightly over-estimated for both
-cases we considered with kernel-\gls{ABC}. The dynamics of the \gls{SI} model,
+also generally identifiable, although it was slightly over-estimated for both
+cases we considered with kernel-assisted \gls{ABC}. The dynamics of the \gls{SI} model,
 and the coalescent process~\autocite{kingman1982coalescent}, offer a potential
@@
 (\cref{fig:rsquared,fig:Ncrossv}) corroborates this hypothesis, as the
-\gls{nltt} uses only information about the coalescence times. 
-{\color{blue}When all \gls{BA}
-parameters were simultaneously estimated with kernel-\gls{ABC}, \gls{N} was
+\gls{nltt} uses only information about the coalescence times. When all \gls{BA}
+parameters were simultaneously estimated with kernel-assisted \gls{ABC}, \gls{N} was
 nearly always over-estimated by approximately a factor of two
@@
 \gls{N} and \gls{I} than the true values. Therefore, if \gls{N} and \gls{I} are
-free to vary (as is the case in kernel-\gls{ABC}), and our hypothesis is true,
+free to vary (as is the case in kernel-assisted \gls{ABC}), and our hypothesis is true,
 both parameters will tend to be overestimated due to being less identifiable
@@
 estimates of \gls{I} were accurate and unbiased (\cref{fig:gridptI}).
-}
 
@@
     derivatives. Values of \gls{I} and \gls{N} used in simulation experiments
-    with kernel-\gls{ABC} are indicated by diamonds.
+    with kernel-assisted \gls{ABC} are indicated by diamonds.
   }
@@
 network per vertex, did not have a measurable impact on tree shape and was not
-estimable with kernel-ABC. The exception to this was the value \gls{m} = 1,
+identifiable with kernel-ABC. The exception to this was the value \gls{m} = 1,
 which produces networks without cycles whose associated trees were more easily
@@
 
-For the more estimable parameters, the credible intervals attained from the
+For the more identifiable parameters, the credible intervals attained from the
 marginal \gls{ABC} target distributions were much narrower than those obtained
@@
 which need not be homogeneous across the network. Moreover, it is possible to
-use kernel-\gls{ABC} to fit a model which relaxes some or all of these
-assumptions, which may be a fruitful avenue for future investigation. Despite
-the possible misspecification, our estimates of the power law exponent $\gamma$
-were within the range of values reported in the literature (\cref{tab:gamma}).
+use kernel-assisted \gls{ABC} to fit a model which relaxes some or all of these
+assumptions, which may be a fruitful avenue for future investigation. 
+Despite the possible misspecification, our estimates of the power
+law exponent $\gamma$ were within the range of values reported in the
+literature (\cref{tab:gamma}).

### 2016-05-17 doc/thesis/intro.tex
Change kernel-ABC to kernel-assisted ABC

@@
 transmission tree is illustrated in \cref{fig:contactnet} (right). 
-\uline{The object on the right of the figure is called a \defn{contact
+The object on the right of the figure is called a \defn{contact
 network}, which depicts the entire susceptible population along with all
 possible routes of disease transmission. Contact networks, and their
-relationships to transmission trees, will be discussed further in}
+relationships to transmission trees, will be discussed further in
 \cref{sec:contactnet}.
@@
 research~\autocite{cottam2008integrating, jombart2011reconstructing,
-ypma2012unravelling, morelli2012bayesian, didelot2014bayesian, hall2015epidemic}.
-\uline{For example, \textcite{hall2015epidemic} developed a Bayesian
-method to jointly estimate a transmission tree and viral phylogeny by combining
-models of agent-based transmission, within-host population dynamics, and
-sequence evolution.}
+ypma2012unravelling, morelli2012bayesian, didelot2014bayesian,
+hall2015epidemic}. For example, \textcite{hall2015epidemic} developed a
+Bayesian method to jointly estimate a transmission tree and viral phylogeny by
+combining models of agent-based transmission, within-host population dynamics,
+and sequence evolution.
 
@@
 not contain any cycles. Cycles are present in the network for all all other $m$
-values. \uline{Examples of \gls{BA} networks with three different values of the
-preferential attachment power \gls{alpha} are shown in} \cref{fig:baeg}.
+values. Examples of \gls{BA} networks with three different values of the
+preferential attachment power \gls{alpha} are shown in \cref{fig:baeg}.
 
@@
 moderate correspondence between the two in some cases.
-\uline{\textcite{goodreau2006assessing} combined a dynamic contact
-network model with a model of within-host viral evolution to simulate viral
-phylogenies over eight types of contact network. Estimates of prevalence and 
-effective population size were calculated for each simulated phylogeny under
-three models of epidemic growth. The author found that estimates for networks
-with a small high-risk subgroup and networks involving commercial sex workers
-were substantially different than estimates for random networks or networks
-with segregated equal-risk groups.}
+\textcite{goodreau2006assessing} combined a dynamic contact network model with
+a model of within-host viral evolution to simulate viral phylogenies over eight
+types of contact network. Estimates of prevalence and effective population size
+were calculated for each simulated phylogeny under three models of epidemic
+growth. The author found that estimates for networks with a small high-risk
+subgroup and networks involving commercial sex workers were substantially
+different than estimates for random networks or networks with segregated
+equal-risk groups.
 

### 2016-05-17 doc/thesis/methods.tex
Change kernel-ABC to kernel-assisted ABC

--- 2016-05-08 doc/thesis/methods.tex
@@
 \subsection{\software{Netabc}: a computer program for estimation of contact
-network parameters with kernel-ABC}
+network parameters with kernel-assisted ABC}
 
@@
 contact network parameters from an estimated transmission tree using
-kernel-\gls{ABC}. The program combines three major components: Gillespie
+kernel-assisted \gls{ABC}. The program combines three major components: Gillespie
 simulation, to simulate transmission trees on contact networks; the tree
@@
 threads~\autocite{barney2009posix}. In addition to the \software{netabc} binary
-to perform kernel-\gls{ABC}, we provide three additional stand-alone utilities:
-\software{treekernel}, to calculate the tree kernel; \software{nettree}, to
-simulate a transmission tree over a contact network; and \software{treestat},
-to compute various summary statistics of phylogenies. The programs are freely
-available at \url{https://github.com/rmcclosk/netabc}.
+to perform kernel-assisted \gls{ABC}, we provide three additional stand-alone
+utilities: \software{treekernel}, to calculate the tree kernel;
+\software{nettree}, to simulate a transmission tree over a contact network; and
+\software{treestat}, to compute various summary statistics of phylogenies. The
+programs are freely available at \url{https://github.com/rmcclosk/netabc}.
 
@@
 
-To test the full kernel-\gls{ABC} algorithm, we simulated three trees each under
+To test the full kernel-assisted \gls{ABC} algorithm, we simulated three trees each under
 a variety of parameter values, and ran the \software{netabc} program to
@@
     experiments. Trees were simulated under the test values, and
-    kernel-\gls{ABC} was used to re-estimate posterior distributions for the
+    kernel-assisted \gls{ABC} was used to re-estimate posterior distributions for the
     \gls{BA} parameters without training.

### 2016-05-17 doc/thesis/objective.tex
Change kernel-ABC to kernel-assisted ABC

@@
 to likelihood-based methods, which would apply to any network model, is
-provided by
-\gls{ABC}~\autocite{rubin1984bayesianly,tavare1997inferring,fu1997estimating,beaumont2002approximate}.
-All of the ingredients required to apply \gls{ABC} to this problem are readily
-available. Simulating networks is straightforward under a variety of models.
-Epidemics on those networks, and the corresponding transmission trees, can also
-be easily simulated. As mentioned above, contact networks can profoundly affect
-transmission tree shape, and those shapes can be compared using a highly
-informative similarity measure called the ``tree
-kernel''~\autocite{poon2013mapping}. \Gls{ABC} can be implemented with
-\gls{SMC}, which has several advantages over other
+provided by \gls{ABC}~\autocite{rubin1984bayesianly, tavare1997inferring,
+fu1997estimating, beaumont2002approximate}. All of the ingredients required to
+apply \gls{ABC} to this problem are readily available. Simulating networks is
+straightforward under a variety of models. Epidemics on those networks, and the
+corresponding transmission trees, can also be easily simulated. As mentioned
+above, contact networks can profoundly affect transmission tree shape, and
+those shapes can be compared using a highly informative similarity measure
+called the ``tree kernel''~\autocite{poon2013mapping}. \Gls{ABC} can be
+implemented with \gls{SMC}, which has several advantages over other
 algorithms~\autocite{mckinley2009inference}. A recently-developed adaptive
@@
 stochastic simulation of epidemics on networks, the tree kernel, and adaptive
-\gls{ABC}-\gls{SMC}. Since our distance measure is a kernel function, our
-method is a type of kernel-\gls{ABC}~\autocite{nakagome2013kernel}. For ease of
-exposition, we will often use the term ``kernel-\gls{ABC}'' to refer to our
-method specifically.
+\gls{ABC}-\gls{SMC}.
+
+Our method will expand on the framework developed
+by~\autocite{poon2015phylodynamic}, who combined \gls{ABC}-\gls{MCMC} with the
+tree kernel to infer parameters of population genetic models from viral
+phylogenies. The term ``kernel-assisted \gls{ABC}'' refers to any application
+of \gls{ABC} using a kernel function as a distance metric, including that
+developed by \citeauthor{poon2015phylodynamic}. However, for ease of
+exposition, we will often use the term ``kernel-assisted \gls{ABC}'' to refer
+to our method specifically. Note that this is distinct from the version of
+\gls{ABC} based on kernel Bayes rule introduced by
+\textcite{fukumizu2011kernel}.
 
@@
 of the \gls{BA} model, including whether they have a detectable impact on tree
-shape, and whether they can be accurately recovered using kernel-\gls{ABC}.
+shape, and whether they can be accurately recovered using kernel-assisted
+\gls{ABC}.
 
@@
 obvious context in which our method could be applied. Therefore, the third and
-final aim of this work is to apply kernel-\gls{ABC} to fit the \gls{BA} model
-to existing \gls{HIV} outbreaks.
+final aim of this work is to apply kernel-assisted \gls{ABC} to fit the
+\gls{BA} model to existing \gls{HIV} outbreaks.
 
 To summarize, this work has three objectives. First, we will develop a method
-which uses kernel-\gls{ABC} to infer parameters of contact network models from
-observed transmission trees. Second, we will use simulations to characterize
-the parameters of the \gls{BA} network model in terms of their effect on tree
-shape and how accurately they can be recovered with kernel-\gls{ABC}. Finally,
-we will apply the method to fit the \gls{BA} model to several real-world
-\gls{HIV} datasets.
+which uses kernel-assisted \gls{ABC} to infer parameters of contact network
+models from observed transmission trees. Second, we will use simulations to
+characterize the parameters of the \gls{BA} network model in terms of their
+effect on tree shape and how accurately they can be recovered with
+kernel-assisted \gls{ABC}. Finally, we will apply the method to fit the
+\gls{BA} model to several real-world \gls{HIV} datasets.
 

### 2016-05-17 doc/thesis/preface.tex
Underline changes for printing
Change kernel-ABC to kernel-assisted ABC

@@
-The initial idea to use kernel approximate Bayesian computation to infer
-contact network model parameters was Dr. Poon's. The tree kernel was originally
-developed by Dr. Poon, but the version used here was implemented by me to
-improve computational efficiency. The idea to apply sequential Monte Carlo was
-mine, but Dr. Alexandre Bouchard-C\^ot\'e made me aware of the adaptive version
-used in this work. Dr. Sarah Otto suggested the experiments involving a network
-with a heterogeneous $\alpha$ parameter and peer-driven sampling. Dr. Richard
-H. Liang provided guidance in the development of the Gillespie simulation
-algorithm and statistical advice. The \software{netabc} program, and all
-supplementary analysis programs, were written by me.
+The initial idea to use approximate Bayesian computation (ABC) to infer contact
+network model parameters was Dr. Poon's, based on his previous work using ABC
+to infer parameters of population genetic models. The tree kernel was
+originally developed by Dr. Poon, but the version used here was implemented by
+me to improve computational efficiency. The idea to apply sequential Monte
+Carlo was mine, but Dr. Alexandre Bouchard-C\^ot\'e made me aware of the
+adaptive version used in this work. Dr. Sarah Otto suggested the experiments
+involving a network with a heterogeneous $\alpha$ parameter and peer-driven
+sampling. Dr. Richard Liang provided guidance in the development of the
+Gillespie simulation algorithm and statistical advice. The \software{netabc}
+program, and all supplementary analysis programs, were written by me.
 

### 2016-05-17 doc/thesis/results.Rnw
Change kernel-ABC to kernel-assisted ABC

--- 2016-05-08 doc/thesis/results.Rnw
@@
 
-\subsubsection*{Joint parameter estimates with kernel-ABC}
+\subsubsection*{Joint parameter estimates with kernel-assisted ABC}
 
@@
 \Cref{fig:abcptm2} shows \gls{MAP} point estimates of the BA model parameters
-obtained with kernel-ABC on simulated data. The estimates shown correspond only
+obtained with kernel-assisted ABC on simulated data. The estimates shown correspond only
 to the simulations where the $m$ parameter was set to 2, however the results
@@
 
-The accuracy of the parameter estimates obtained with kernel-ABC
+The accuracy of the parameter estimates obtained with kernel-assisted ABC
 paralleled the results from the \gls{kSVR} classifier. Of the four parameters,
@@
       posterior density (HPD) interval widths for BA model parameter estimates
-      obtained with kernel-ABC.
+      obtained with kernel-assisted ABC.
   ]{
@@
       posterior density (HPD) interval widths for BA model parameter estimates
-      obtained with kernel-ABC. Three transmission trees were simulated under
+      obtained with kernel-assisted ABC. Three transmission trees were simulated under
       each combination of the listed parameter values, and the parameters were
-      estimated with kernel-ABC without training.
+      estimated with kernel-assisted ABC without training.
   }
@@
     Marginal posterior distributions of BA model parameters estimated
-    with kernel-ABC for a single simulated transmission tree. Dotted
+    with kernel-assisted ABC for a single simulated transmission tree. Dotted
     lines and shaded polygon indicate true values.
@@
 
-We applied kernel-ABC to five published HIV datasets (\cref{tab:data}),
+We applied kernel-assisted ABC to five published HIV datasets (\cref{tab:data}),
 and found substantial heterogeneity among the parameter estimates
@@
 investigated. The results are summarized in \cref{tab:gamma}. All of the
-estimated exponents were in the range $2 \leq \gamma \leq 2.5$.
+estimated exponents were in the range $2 \leq \gamma \leq 2.5$, which is on the
+lower end of the range $2 \leq \gamma \leq 4$ reported in the literature.
 

### 2016-05-17 doc/thesis/results.tex
Change kernel-ABC to kernel-assisted ABC

--- 2016-05-08 doc/thesis/results.tex
@@
 
-\subsubsection*{Joint parameter estimates with kernel-ABC}
+\subsubsection*{Joint parameter estimates with kernel-assisted ABC}
 
 \Cref{fig:abcptm2} shows \gls{MAP} point estimates of the BA model parameters
-obtained with kernel-ABC on simulated data. The estimates shown correspond only
+obtained with kernel-assisted ABC on simulated data. The estimates shown correspond only
 to the simulations where the $m$ parameter was set to 2, however the results
@@
 $I$,
-  1115 
+  1119 
   [662 -
@@
 $N$,
-  12680 
-  [3377-
-   14933].
+  12678 
+  [3367-
+   14977].
 The approximate posterior distributions for this simulation are shown in
@@
 $\alpha$, 
-  0.08 
-  [0 -
-   0.62];
+  0.23 
+  [0.01 -
+   0.63];
 $I$,
-  2391 
-  [1379 -
-   3860];
+  2398 
+  [1419 -
+   3767];
 $m$,
@@
 $N$,
-  10675 
-  [2881 -
+  9761 
+  [2836 -
    14788].

### 2016-05-21 doc/thesis/discussion.tex
Most of Alex's changes done

@@
 computationally intensive, taking about a day when run on 20 cores in parallel
-with the settings described in the methods section. Implementing parallization
-using MPI, rather than POSIX threads as we have done here, would allow the
-program to be run over a larger number of cores on multiple CPUs in parallel.
+with the settings described in the methods section. Implementing
+parallelization using MPI, rather than POSIX threads as we have done here,
+would allow the program to be run over a larger number of cores on multiple
+CPUs in parallel.
 
-\subsection{Analysis of \acrlong{BA} model}
+\subsection{Analysis of \acrlong{BA} model with synthetic data}
 
@@
 
-\subsection{Application to HIV data}
+\subsection{Application to real world HIV data}
 

### 2016-05-21 doc/thesis/intro.tex
Most of Alex's changes done

@@
 correspond to \defn{extant}, or observed, taxa. The \defn{internal nodes}
-correspond to their (usually extinct) common ancestors. The edges or
-\defn{branches} of the phylogeny connect ancestors to their descendants.
-Phylogenies may have a \defn{root}, which is a node with no descendants
-distinguished as the most recent common ancestor of all the extant
+correspond to their common ancestors, {\color{blue}\uline{usually extinct
+(although occasionally the internal nodes may be observed as well, eg.
+\autocite{jombart2011reconstructing}).}} The edges or \defn{branches} of the
+phylogeny connect ancestors to their descendants. Phylogenies may have a
+\defn{root}, which is a node with no descendants distinguished as the most
+recent common ancestor of all the extant
 taxa~\autocite{harding1971probabilities}. When such a root exists, the tree is
@@
 parameter. The tree kernel is defined as 
-\[
-  K(T_1, T_2) = \sum_{n_1 \in N(T_1)} \sum{n_2 \in N(T_2)} \Delta (n_1, n_2),
-\]
+\begin{align}
+  K(T_1, T_2) = \sum_{n_1 \in N(T_1)} \sum_{n_2 \in N(T_2)} \Delta (n_1, n_2),
+  \label{eq:kernel}
+\end{align}
 where
@@
     0 & \text{otherwise}.
-
   \end{cases}
 \]
-Here $\lambda$ is a decay factor parameter, which penalizes large matches that
-tend to dominate the kernel score. In this work, we refer to the parameters
-$\lambda$ and $\sigma$ as \defn{meta-parameters}, to avoid confusing them with
-model parameters we are trying to estimate.
-
-%The tree kernel was later shown to be highly effective in differentiating trees
-%simulated under a compartmental model with two risk groups of varying contact
-%rates~\autocite{poon2015phylodynamic}. In that paper,
-%\citeauthor{poon2015phylodynamic} used the tree kernel as the distance function
-%in \gls{ABC} (see \cref{sec:abc}), to fit epidemiological models to observed
-%trees. This is an example of kernel-ABC~\autocite{nakagome2013kernel}, which
-%will be discussed further in \cref{sec:abc}.
+{\color{blue}\uline{
+The parameter $\lambda$ in the above expression, is called the \emph{decay
+factor}~\autocite{moschitti2006making}, and takes a value between 0 and 1.
+Without this parameter, terms in the sum} \ref{eq:kernel} \uline{corresponding
+to large subset trees with the same topology would be similarly large and tend
+to dominate the kernel score. $\lambda$ penalizes $\Delta$ more strongly as the
+number of recursive calls increases, which downweights the largest matching
+substructures and allows smaller matches to contribute more to the kernel
+score. }} In this work, we refer to the parameters $\lambda$ and $\sigma$ as
+\defn{meta-parameters}, to avoid confusing them with model parameters we are
+trying to estimate. {\color{blue}\uline{ When evaluating the tree kernel, it is
+helpful to reorder the children of each internal node such that the larger of
+the two subtrees is on the right-hand side. If the two subtrees have equal
+sizes, then the child with the longer branch length can be put on the
+right-hand side. This operation is referred to as \defn{ladderizing}. Since the
+ordering of children is arbitrary in phylogenies, this operation ensures that a
+maximal number of matching subset trees are counted by the tree kernel without 
+making meaningful changes to the trees. }}
+
+{\color{blue}\uline{
+The tree kernel was later shown to be highly effective in differentiating trees
+simulated under a compartmental model with two risk groups of varying contact
+rates~\autocite{poon2015phylodynamic}. In that paper,
+\citeauthor{poon2015phylodynamic} used the tree kernel as the distance function
+in \gls{ABC}} (see \cref{sec:abc}), \uline{to fit epidemiological models to
+observed trees. }}
 
@@
 of \defn{particles}, here denoted
-$\set{x^{(k)}}$~\autocite{doucet2001introduction, liu2008monte}. These
-collections or \defn{populations} are constructed to form a \defn{Monte Carlo
-approximation} to some distribution of interest $\pi$, meaning that the
-empirical distribution of the particles converges in distribution to $\pi$ as
-the population size gets large~\autocite{liu2001theoretical}. The word
-\defn{sequential} is used because the particle populations are modified in an
-iterative fashion over time, for example, to incorporate new evidence.
+$\set{x^{(k)}}$~\autocite{doucet2000sequential, doucet2001introduction,
+liu2008monte}. These collections or \defn{populations} are constructed to form
+a \defn{Monte Carlo approximation} to some distribution of interest $\pi$,
+meaning that the empirical distribution of the particles converges in
+distribution to $\pi$ as the population size gets
+large~\autocite{liu2001theoretical}. The word \defn{sequential} is used because
+the particle populations are modified in an iterative fashion over time, for
+example, to incorporate new evidence.
 
@@
 
+{\color{blue}\uline{
+The goal of this section of the introduction is to describe an algorithm called
+the \gls{SMC} sampler~\autocite{del2006sequential}, which forms the basis of
+the adaptive \gls{ABC}-\gls{SMC} algorithm we apply toward the main objective
+of this work. We begin by describing \gls{SIS}, which is a precursor to
+\gls{SMC} that samples from a sequence of distributions defined on spaces of
+increasing dimension. We then describe \gls{SMC} itself, which extends
+\gls{SIS} with a resampling step to fight particle degeneracy. Finally, we
+outline the \gls{SMC} sampler, which allows \gls{SMC} to be applied to
+sequences of distributions all defined on the same space. This terminology will
+become clear as the methods are described.
+}}
+
 \subsection{Sequential importance sampling}
 \label{subsec:sis}
-
-\Gls{SIS}~\autocite{gordon1993novel} is one type of \gls{SMC} method, whose aim
-is to sample from a distribution $\pi$ on an high-dimensional space, say
-$\pi(\vec{x}) = \pi(x_1, \ldots, x_d)$. The basis of \gls{SIS} is \gls{IS},
-which is a method of estimating summary statistics of distributions which are
-known only up to a normalizing constant, and therefore cannot be sampled from
-directly. That is, if $\pi$ is such a distribution and $f$ is any real-valued
-function, \gls{IS} is concerned with estimating
+\glsreset{SIS}
+
+\Gls{SIS}~\autocite{gordon1993novel} {\color{blue}\uline{is a particle-based
+method}} whose aim is to sample from a distribution $\pi$ on an
+high-dimensional space, say $\pi(\vec{x}) = \pi(x_1, \ldots, x_d)$. The basis
+of \gls{SIS} is \gls{IS}, which is a method of estimating summary statistics of
+distributions which are known only up to a normalizing constant, and therefore
+cannot be sampled from directly. That is, if $\pi$ is such a distribution and
+$f$ is any real-valued function, \gls{IS} is concerned with estimating
 \[
@@
 \defn{importance distribution}, from which we are able to sample. Define the
-\defn{importance weight} as the ratio $w(x) = \gamma(x)/\eta(x)$. We can
-express the normalizing constant $Z$ in terms of the importance weight and
-distribution, $Z = \int w(x) \eta(x) \d x$, and in turn write the expectation
-of interest as
-\[
-  \int f(x) \pi(x) \d x = \frac{\int f(x) \gamma(x) \d x}
-                               {\int w(x) \eta(x) \d x}.
-\]
-If we sample a large number of points from $\eta$, then $\eta(x)$ can be
-approximated by a Monte Carlo estimate. Since the remaining quantities $f$,
-$\gamma$, and $w$ can all be evaluated pointwise, these are all the ingredients
-we need to obtain an estimate of $\pi(f)$. Although this is a simple and
-elegant approach, the drawback is that the variance of the estimate is
-proportional to the variance of the importance weights~\autocite{liu2008monte},
-which may be quite large if $\eta$ and $\gamma$ are very different. Therefore,
-the practical use of \gls{IS} on its own is limited, since it depends on
-finding an importance distribution similar to $\pi$, which we usually know very
-little about \textit{a priori}.
+\defn{importance weight} as the ratio $w(x) = \gamma(x)/\eta(x)$. We can write
+the expectation of interest as
+{\color{blue}
+\begin{align}
+  \int f(x) \pi(x) \d x = \frac{1}{Z} \int w(x) f(x) \d x.
+  \label{eq:is}
+\end{align}
+}
+{\color{blue}\uline{
+Since $\eta$ can be sampled from exactly, and $\gamma$ and $f$ can both be
+evaluated pointwise, the integral $\int w(x) f(x) \d x$ can be approximated by
+a Monte Carlo estimate. Moreover, the normalizing constant $Z$ can be expressed
+in terms of the importance weight and distribution, $Z = \int w(x) \eta(x) \d
+x$. Therefore, we have all the ingredients we need to obtain an estimate of
+$\pi(f)$ using} \cref{eq:is}.} Although this is a simple and elegant approach,
+the drawback is that the variance of the estimate is proportional to the
+variance of the importance weights~\autocite{liu2008monte}, which may be quite
+large if $\eta$ and $\gamma$ are very different. Therefore, the practical use
+of \gls{IS} on its own is limited, since it depends on finding an importance
+distribution similar to $\pi$, which we usually know very little about
+\textit{a priori}.
 
@@
 
-The recursive definition \cref{eq:sisw} suggests an algorithm for obtaining a
-sample from $\pi$ (\cref{alg:sis}).  We begin with $n$ ``particles'' which have
-been sampled from the importance distribution $\eta(x_0)$ for $\pi(x_0)$. The
-particles are updated and reweighted $d$ times, corresponding to the $d$
-elements of the decomposition of $\pi$. At the $i$th step, each particle is
-extended to include $x_i$ drawn according to the chosen $\eta(x_i \mid
-\vec{x_{i-1}})$, and the importance weights are recalculated and normalized.
+The recursive definition \ref{eq:sisw} suggests an algorithm for obtaining a
+sample from {\color{blue}\uline{$\eta$ and using it to obtain an approximate
+sample from $\pi$ by \gls{IS}}} (\cref{alg:sis}).  We begin with $n$
+``particles'' which have been sampled from the importance distribution
+$\eta(x_0)$ for $\pi(x_0)$. The particles are updated and reweighted $d$ times,
+corresponding to the $d$ elements of the decomposition of $\pi$. At the $i$th
+step, each particle is extended to include $x_i$ drawn according to the chosen
+$\eta(x_i \mid \vec{x_{i-1}})$, and the importance weights are recalculated and
+normalized.
 
@@
       \Comment{Initialize the $k$th particle}
-      \State $w^{(k)} \gets \dfrac{\pi\left(x_1^{(k)}\right)}{\eta\left(x_1^{(k)}\right)}$
+      %\State $w^{(k)} \gets \dfrac{\pi\left(x_1^{(k)}\right)}{\eta\left(x_1^{(k)}\right)}$
+      \State $w^{(k)} \gets \pi(x_1^{(k)}) / \eta(x_1^{(k)})$
     \EndFor
@@
       \For {$k = 1$ to $n$}
-        \State Sample $x_i^{(k)}$ from $\eta\left(x_i \mid \vec{x_{i-1}^{(k)}}\right)$
+        \State Sample $x_i^{(k)}$ from $\eta(x_i \mid \vec{x_{i-1}^{(k)}})$
         \Comment Extend the $k$th particle
-        \State $w(\vec{x_i}^{(k)}) \gets \dfrac{\pi\left(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}\right)}{\eta\left(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}\right)} \cdot w(\vec{x_{i-1}}^{(k)})$
+        \State $w(\vec{x_i}^{(k)}) \gets [\pi(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}) \,/\, \eta(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}})] \cdot w(\vec{x_{i-1}}^{(k)})$
       \EndFor
@@
     \EndFor
-    \State Sample $n$ particles with probabilities $w$
   \end{algorithmic}
@@
 
-Of course, $\eta$ is merely an approximation to $\pi$, and may be a fairly poor
-one depending on the application. Try as we might to keep the variances of the
-weights low, the cumulative errors at each sequential step tend to push many of
-the weights to very low values. This results in a poor approximation to $\pi$,
-since only a few particles retain high importance weights after all $d$
-sequential steps. To mitigate this problem, a resampling step is periodically
-applied when the variance in the importance weights becomes too high. Several
-different criteria have been proposed for when to resample, but we focus here
-on the one described by \textcite{liu2008monte}, namely the decay of the
-\gls{ESS} below a prescribed threshold, conventionally $n/2$. The \gls{ESS} of
-the population of particles is defined as \[ \ESS(w) = \frac{n}{1 + \Var(w)},
-\] where $n$ is the number of particles~\autocite{liu2008monte}. When the
-\gls{ESS} drops below the threshold, particles are resampled according to their
-weights. This results in the removal of low-weight particles from the
-population, and also equalizes all the weights. Various resampling strategies
-beyond the basic sampling with replacement have been
-proposed~\autocite{douc2005comparison}, but we will not discuss those here.
+\subsection{Sequential Monte Carlo}
+
+The importance distribution $\eta$ constructed with \gls{SIS} is merely an
+approximation to $\pi$, and may be a fairly poor one in practice depending on
+the application. Try as we might to keep the variances of the weights low, the
+cumulative errors at each sequential step tend to push many of the weights to
+very low values~\autocite{doucet2000sequential}. This results in a poor
+approximation to $\pi$, since only a few particles retain high importance
+weights after all $d$ sequential steps, 
+{\color{blue}\uline{ a problem known as \defn{particle degeneracy}. To mitigate
+this problem, \textcite{gordon1993novel} introduced technique they called the
+\defn{bootstrap filter}, which involves a resampling of the population of
+particles after each sequential step in accordance with their importance
+weights. A similar idea, termed \defn{particle rejuvination}, was proposed by
+\textcite{liu1995blind}. These approaches cause particles with high importance
+weights to be replicated in the population, while particles with low weights
+may be removed. After each resampling step, the importance weights for all
+particles are set equal.}}
+
+\begin{algorithm}
+  \caption{Sequential Monte Carlo \autocite{doucet2000sequential}.}
+  \begin{algorithmic}
+    \For {$k = 1$ to $n$}
+      \State Sample $x_1^{(k)}$ from $\eta(x_1)$
+      \Comment{Initialize the $k$th particle}
+      \State $w^{(k)} \gets \pi(x_1^{(k)}) / \eta(x_1^{(k)})$
+    \EndFor
+    \For {$i = 2$ to $d$}
+      \For {$k = 1$ to $n$}
+        \State Sample $x_i^{(k)}$ from $\eta(x_i \mid \vec{x_{i-1}^{(k)}})$
+        \Comment Extend the $k$th particle
+        \State $w(\vec{x_i}^{(k)}) \gets [\pi(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}) \,/\, \eta(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}})] \cdot w(\vec{x_{i-1}}^{(k)})$
+      \EndFor
+      \If {$\ESS(w) < T$}
+        \Comment{$T$ is a user-defined threshold}
+        \State Resample the particles according to $w$
+        \For {$k = 1$ to $n$}
+          \State $w^{(k)} \gets 1/n$
+        \EndFor
+      \EndIf
+    \EndFor
+  \end{algorithmic}
+  \label{alg:smc}
+\end{algorithm}
+
+{\color{blue}\uline{The resampling step was formally integrated with
+\gls{SIS} by \citeauthor{doucet2000sequential} to form the first \gls{SMC}
+algorithm }(\cref{alg:smc}). \uline{Rather than resample at every step as the
+bootstrap filter proposed, the authors use a criterion based on the \gls{ESS}
+the particle population to determine when resampling is necessary. }} The
+\gls{ESS} of the population of particles is defined as 
+\[ 
+  \ESS(w) = \frac{n}{1 + \Var(w)}, 
+\]
+where $n$ is the number of particles in the population.
+When the \gls{ESS} drops below the threshold
+{\color{blue}\uline{(conventionally $n/2$~\autocite{liu2008monte})}}, particles
+are resampled according to their importance weights. This results in the
+removal of low-weight particles from the population, and also equalizes all the
+weights. Various resampling strategies beyond the basic sampling with
+replacement have been proposed~\autocite{douc2005comparison}, but we will not
+discuss those here.
 
@@
 
-The \gls{SIS} algorithm described above aims to sample from a high-dimensional
-distribution $\pi(x)$, by sequentially sampling from $d$ distributions of lower
-but increasing dimension. \textcite{del2006sequential} developed an
-\defn{\gls{SMC} sampler} with an alternative objective: to sample sequentially
-from $d$ distributions $\pi_1, \ldots, \pi_d$, all of the same dimension and
-defined on the same space. The $\pi_i$ are assumed to form a related sequence,
-such as posterior distributions attained by sequentially considering new
-evidence. As with \gls{SIS}, we assume that $\pi_i(x) = \gamma_i(x) / Z_i$,
-where $\gamma_i$ is known pointwise and the normalizing constant $Z_i$ is
-unknown.
+The \gls{SIS} and \gls{SMC} algorithms described above aim to sample from a
+high-dimensional distribution $\pi(x)$, by sequentially sampling from $d$
+distributions of lower but increasing dimension. \textcite{del2006sequential}
+developed an \defn{\gls{SMC} sampler} with an alternative objective: to sample
+sequentially from $d$ distributions $\pi_1, \ldots, \pi_d$, all of the same
+dimension and defined on the same space. The $\pi_i$ are assumed to form a
+related sequence, such as posterior distributions attained by sequentially
+considering new evidence. As with \gls{SIS}, we assume that $\pi_i(x) =
+\gamma_i(x) / Z_i$, where $\gamma_i$ is known pointwise and the normalizing
+constant $Z_i$ is unknown.
 
@@
       \Comment{Initialize the $k$th particle}
-      \State $w^{(k)} \gets \dfrac{\gamma_1\left(x_1^{(k)}\right)}{\eta_1\left(x_1^{(k)}\right)}$
+      \State $w^{(k)} \gets \gamma_1(x_1^{(k)}) / \eta_1(x_1^{(k)})$
       \State Normalize the weights so that $\sum w = 1$
@@
       \EndIf
-      \State Sample the $i$th component of $n$ particles with probabilities $w$
     \EndFor
@@
 they may not be guaranteed to find the true \gls{ML} estimates if the
-likelihood function is complex.
+likelihood function is multi-modal.
 
@@
 \]
-The intuition for why this approximation might hold comes from the fact that, 
-when $\varepsilon = 0$, the integral on the left is exactly equal to the
-posterior. Thus, by taking $\varepsilon$ small, we should attain something
-close to the posterior.
-
 The distribution $\pi_\varepsilon(\theta, z \mid y)$ is variously referred to
 as the \textit{\gls{ABC} target distribution} or the \gls{ABC} approximation
-to the posterior. Note that in many formulations, the distance function $\rho$
-is defined as $\rho(S(\cdot), S(\cdot))$ where $S$ is a function which maps
-data points into a vector of summary statistics. This can be useful if the data
-are high-dimensional or of a complex type, but it is not strictly necessary.
-For instance, if the data are numeric and of low dimension, the distance
-function may simply be the Euclidean distance~\autocite{sisson2007sequential}.
-For more complex data, \textcite{nakagome2013kernel} proposed the use of a
-kernel function (defined in \cref{subsec:treeshape}), an approach they dubbed
-\defn{kernel-\gls{ABC}}.
+to the posterior. The intuition for why this approximation might be reasonable
+comes from the fact that, when $\varepsilon = 0$ and $\rho$ satisfies $\rho(z,
+y) = 0 \Rightarrow x = y$, the integral on the left is exactly equal to the
+posterior. Thus, by taking $\varepsilon$ small, we should attain something
+close to the posterior {\color{blue}\uline{if $\rho$ captures the similarity
+between datasets reasonably well. However, the accuracy of the \gls{ABC}
+approximation depends heavily on the choice of distance function. }}
+
+In many applications {\color{purple}\uline{(eg. \autocite{fu1997estimating,
+tanaka2006using})}}, $\rho$ is defined as $\rho(S(\cdot), S(\cdot))$ where $S$
+is a function which maps data points into a vector of summary statistics.
+{\color{blue}\uline{In the context of \gls{ABC}, a summary statistic $S$ is
+called \defn{sufficient} if} 
+\[
+  f(\theta \mid y) = f(\theta \mid S(y)).
+\]
+\uline{That is, sufficiency implies that the data can be replaced with the
+summary statistic without losing any information about the posterior
+distribution~\autocite{marjoram2006modern}. For most problems, it is not
+possible to find sufficient summary statistics~\autocite{marjoram2006modern}.
+TODO
+}}
+
+This can be useful if the data are high-dimensional or of a complex type, but
+it is not strictly necessary. For instance, if the data are numeric and of low
+dimension, the distance function may simply be the Euclidean
+distance~\autocite{sisson2007sequential}. \textcite{poon2015phylodynamic}
+proposed the use of a kernel function (defined in \cref{subsec:treeshape}),
+which allows
 
@@
 rejection, \gls{MCMC}, and \gls{SMC}~\autocite{marin2012approximate}. To
-simplify the math, we shall restrict the descriptions of these algorithms to
-the case of one simulated dataset per parameter particle (the meaning of this
-will become clear shortly). The extension to multiple datasets per particle is
-straightforward and will be given at the end of the section. We use the
-variable $x$ to refer to the pair $(\theta, z)$, so that the \gls{ABC} target
-distribution can be written $\pi_\varepsilon(x \mid y)$.
+simplify the notation, we shall restrict the descriptions of these algorithms
+to the case of one simulated dataset per parameter particle (the meaning of
+this will become clear shortly). The extension to multiple datasets per
+particle is straightforward and will be given at the end of the section. We use
+the variable $x$ to refer to the pair $(\theta, z)$, so that the \gls{ABC}
+target distribution can be written $\pi_\varepsilon(x \mid y)$.
 

### 2016-05-21 doc/thesis/methods.tex
Most of Alex's changes done

@@
 lengths were scaled by their mean. \textit{igraph} library's implementation of
-the BA model~\autocite{csardi2006igraph} to generate the graphs. The analyses
-were run on Westgrid (\url{https://www.westgrid.ca/}) and a local computer
-cluster. With the exception of our own programs, all analyses were done in
-\software{R}, and all packages listed below are \software{R} packages.
+the BA model~\autocite{csardi2006igraph} was used to generate the graphs. The
+analyses were run on Westgrid (\url{https://www.westgrid.ca/}) and a local
+computer cluster. With the exception of our own programs, all analyses were
+done in \software{R}, and all packages listed below are \software{R} packages.
 

### 2016-05-21 doc/thesis/methods_results.Rnw
Most of Alex's changes done

--- 2016-05-17 doc/thesis/results.Rnw
@@
-\subsection{Analysis of \acrlong{BA} model}
-
 <<setup, include=FALSE>>=
@@
 @
+
+\section{\software{Netabc}: a computer program for estimation of contact
+network parameters with kernel-assisted ABC}
+
+\software{Netabc} is a computer program to perform statistical inference of
+contact network parameters from an estimated transmission tree using
+kernel-assisted \gls{ABC}. The program combines three major components: Gillespie
+simulation, to simulate transmission trees on contact networks; the tree
+kernel, to compare transmission trees; and adaptive \gls{ABC}-\gls{SMC}, to
+maintain a population of particles and advance it toward the \gls{ABC} target
+distribution. We give a high-level overview of the program here, before
+describing these three components in detail. \software{Netabc} takes as input
+an estimated transmission tree, which can be derived from a viral phylogeny by
+rooting and time-scaling as described in \cref{subsec:phylodynamics} or
+estimated by other methods~\autocite{cottam2008integrating,
+jombart2011reconstructing, ypma2012unravelling, morelli2012bayesian,
+didelot2014bayesian}. We variously refer to this estimated transmission tree as
+the observed tree, input tree, or true tree.
+
+As described in \cref{sec:smc}, \software{netabc} keeps track of a population
+of particles $x^{(k)}$, each of which contains particular parameter values
+$\theta^{(k)}$ for the model we are trying to fit. A small number of contact
+networks $z^{(k)}$ are generated for each particle, in accordance with that
+particle's parameters. An epidemic is simulated over each of these networks
+using Gillespie simulation, and by keeping track of its progress, a
+transmission tree is obtained. Thus, each particle becomes associated with
+several simulated transmission trees. These trees are compared to the input
+tree using the tree kernel. Particles are weighted according to the similarity
+of their associated simulated trees with the true tree, with more similar trees
+receiving higher weights. The particles are iteratively perturbed to explore
+the parameter space, and particles with simulated trees too distant from the
+true tree are periodically dropped and resampled. Once a convergence criterion
+is attained, the final set of particles is used as a Monte Carlo approximation
+to the target distribution of \gls{ABC}, which is assumed to resemble the
+posterior distribution on model parameters (see \cref{sec:abc}). A graphical
+schematic of this algorithm is given in \cref{fig:abcsmc}.
+
+\begin{figure}
+    \includegraphics{abc-smc.pdf}
+    \caption[Graphical schematic of the ABC-SMC algorithm implemented in \software{netabc}.]{
+      Graphical schematic of the \gls{ABC}-\gls{SMC} algorithm implemented in
+      \software{netabc}. Particles are initially drawn from their prior
+      distributions, making the initial population a Monte Carlo approximation
+      to the prior. At each iteration, particles are perturbed, and a distance
+      threshold around the true tree contracts. Particles are rejected, and
+      eventually resampled, when all their associated simulated trees lie
+      outside the threshold. As the algorithm progresses, the population
+      smoothly approaches a Monte Carlo approximation of the \gls{ABC} target
+      distribution, which is assumed to resemble the posterior.
+    }
+    \label{fig:abcsmc}
+\end{figure}
+
+\software{Netabc} is written in the \software{C} programming language. The
+\software{igraph} library~\autocite{csardi2006igraph} is used to generate and
+store contact networks and phylogenies. Judy arrays~\autocite{baskins2004judy}
+are used for hash tables and dynamic programming matrices. The
+\gls{GSL}~\autocite{gough2009gnu} is used to generate random draws from
+probability distributions, and to perform the bisection step in the adaptive
+\gls{ABC}-\gls{SMC} algorithm. Parallelization is implemented with POSIX
+threads~\autocite{barney2009posix}. In addition to the \software{netabc} binary
+to perform kernel-assisted \gls{ABC}, we provide three additional stand-alone
+utilities: \software{treekernel}, to calculate the tree kernel;
+\software{nettree}, to simulate a transmission tree over a contact network; and
+\software{treestat}, to compute various summary statistics of phylogenies. The
+programs are freely available at \url{https://github.com/rmcclosk/netabc}.
+
+To check that our implementation of Gillespie simulation was correct, we
+reproduced Figure 1A of \textcite{leventhal2012inferring} (our
+\cref{fig:leventhal}), which plots the unbalancedness of transmission trees
+simulated over four network models at various levels of pathogen
+transmissibility. Our implementation of adaptive ABC-SMC was tested by applying
+it to the same mixture of Gaussians used by \citeauthor{del2012adaptive} to
+demonstrate their method (originally used by~\textcite{sisson2007sequential}).
+We were able to obtain a close approximation to the function (see
+\cref{fig:smctest}), and attained the stopping condition used by the authors in
+a comparable number of steps. To check that the algorithm would converge to a
+bimodal distribution, we also applied it to a mixture of two Gaussians with
+means $\pm$4 and variances 1. The algorithm was able to recover both peaks
+(\cref{fig:smctest2}).
+
+\subsection{Epidemic simulation}
+\label{subsec:nettree}
+
+The simulation of epidemics, and the corresponding transmission trees, over
+contact networks is performed in \software{netabc} using the Gillespie
+simulation algorithm~\autocite{gillespie1976general}. This method has been
+independently implemented and applied by several
+authors~\autocite[\textit{e.g.}][]{o2011contact, robinson2013dynamics,
+leventhal2012inferring, groendyke2011bayesian, villandre2016assessment}.
+\textcite{groendyke2011bayesian} published their implementation as an
+\software{R} package, but since the \gls{SMC} algorithm is quite
+computationally intensive, we chose to implement our own version in
+\software{C}.
+
+Let $G = (V, E)$ be a directed contact network. We assume the individual nodes
+and edges of $G$ follow the dynamics of the \gls{SIR}
+model~\autocite{kermack1927contribution}. Each directed edge $e = (u, v)$ in
+the network is associated with a transmission rate $\beta_e$, which indicates
+that, once $u$ becomes infected, the waiting time until $u$ infects $v$ is
+distributed as $\Exponential(\beta_e)$. Note that $v$ may become infected
+before this time has elapsed, if $v$ has other incoming edges. $v$ also has a
+removal rate $\gamma_v$, so that the waiting time until removal of $v$ from the
+population is $\Exponential(\gamma_v)$. Removal may correspond to death or
+recovery with immunity, or a combination of both, but in our implementation
+recovered nodes never re-enter the susceptible population. We define a
+\defn{discordant edge} as an edge $(u, v)$ where $u$ is infected and $v$ has
+never been infected.
+
+To describe the algorithm, we introduce some notation and variables. Let
+$\inc(v)$ be the set of incoming edges to $v$, and $\out(v)$ be the set of
+outgoing edges from $v$. Let $I$ be the set of infected nodes in the network,
+$R$ be the set of removed nodes, and $S$ be the remaining susceptible nodes,
+and $D$ be the set of discordant edges in the network. Let $\beta$ be the total
+transmission rate over all discordant edges, and $\gamma$ be the total removal
+rate of all infected nodes,
+\[
+  \beta = \sum_{e \in D} \beta_e, \quad
+  \gamma = \sum_{v \in I} \gamma_v.
+\]
+The variables $S$, $I$, $R$, $D$, $\beta$, and $\gamma$ are all updated as the
+simulation progresses. When a node $v$ becomes infected, it is deleted from $S$
+and added to $I$. Any formerly discordant edges in $\inc(v)$ are deleted from
+$D$, and edges in $\out(v)$ to nodes in $S$ are added to $D$. If $v$ is later
+removed, it is deleted from $I$ and added to $R$, and any discordant edges in
+$\out(v)$ are deleted from $D$. At the time of either infection or removal, the
+variables $\beta$ and $\gamma$ are updated to reflect the changes in the
+network. Since these updates are straightforward, we do not write them
+explicitly in the algorithm.
+
+\newcommand{\tip}{\mathit{tip}}
+
+The Gillespie simulation algorithm is given as Algorithm~\ref{alg:nettree}. The
+transmission tree $T$ is simulated along with the epidemic. We keep a map
+called $\tip$, which maps infected nodes in $I$ to the tips of $T$. The
+simulation continues until either there are no discordant edges left in the
+network, or we reach a user-defined cutoff of time ($t_{\max}$) or number of
+infections ($I_{\max}$). We use the notation $\Uniform(0, 1)$ to indicate a
+number drawn from a uniform distribution on $(0, 1)$, and likewise for
+$\Exponential(\lambda)$. The combined number of internal nodes and tips in $T$
+is denoted $|T|$.
+
+\begin{algorithm}
+  \label{alg:nettree}
+  \caption{Simulation of an epidemic and transmission tree over a contact network}
+  \begin{algorithmic}
+    \State infect a node $v$ at random, updating $S$, $I$, $D$, $\beta$ and $\gamma$
+    \State $T \gets$ a single node with label $1$
+    \State $\tip[v] \gets 1$
+    \State $t \gets 0$
+    \While{$D \neq \emptyset$ and $|I| + |R| < I_{\max}$ and $t < t_{\max}$}
+      \State $s \gets \min(t_{\max} - t, \Exponential(\beta + \gamma))$
+      \For{$v \in \tip$}
+        \State{extend the branch length of $\tip[v]$ by $s$}
+      \EndFor
+      \State $t \gets t + s$
+      \If{$t < t_{\max}$}
+        \If{$\Uniform(0, \beta + \gamma) < \beta$}
+          \State choose an edge $e = (u, v)$ from $D$ with probability $\beta_e / \beta$
+                 and infect $v$
+          \State add tips with labels $(|T|+1)$ and $(|T|+2)$ to $T$
+          \State connect the new nodes to $\tip[v]$ in $T$, with branch lengths $0$
+          \State $\tip[v] \gets |T|-1$
+          \State $\tip[u] \gets |T|$
+        \Else
+          \State choose a node $v$ from $I$ with probability $\gamma_v / \gamma$
+                 and remove $v$
+          \State delete $v$ from $\tip$
+        \EndIf
+        \State update $S$, $I$, $R$, $D$, $\beta$, and $\gamma$
+      \EndIf
+    \EndWhile
+  \end{algorithmic}
+\end{algorithm}
+
+\subsection{Phylogenetic kernel}
+
+The tree kernel developed by \textcite{poon2013mapping} provides a
+comprehensive similarity score between two phylogenetic trees, via the
+dot-product of the two trees' feature vectors in the infinite-dimensional space
+of all possible subset trees with branch lengths (see \cref{subsec:treeshape}).
+The kernel was implemented using the fast algorithm developed by
+\textcite{moschitti2006making}. First, the production rule of each node, which
+is the total number of children and the number of leaf children, is recorded.
+The nodes of both trees are ordered by production rule, and a list of pairs of
+nodes sharing the same production rule is created. These are the nodes for
+which the value of the tree kernel must be computed - all other pairs have a
+value of zero. The pairs to be compared are then re-ordered so that the child
+nodes are always evaluated before their parents. Due to its recursive
+definition, ordering the pairs in this way allows the tree kernel to be
+computed by dynamic programming. The complexity of this implementation is
+$O(|T_1||T_2|)$, where $|T|$ counts the number of nodes in the tree $T$.
+
+The tree kernel cannot be used directly as a distance measure for \gls{ABC},
+since it is maximized, not minimized, when the two trees being compared are the
+same. Therefore, we defined the distance between two trees as
+\[
+  \rho(T_1, T_2) = 1 - \frac{K(T_1, T_2)}{\sqrt{K(T_1, T_1) K(T_2, T_2)}},
+\]
+which is a number between 0 and 1 minimized when $T_1 = T_2$. This is similar
+to the normalization used by \textcite{collins2002new, poon2013mapping}.
+
+\subsection{Adaptive sequential Monte Carlo for Approximate Bayesian computation}
+\label{subsec:adaptsmc}
+
+We implemented the adaptive \gls{SMC} algorithm for \gls{ABC} developed by
+\textcite{del2012adaptive}. This algorithm is similar to the reference
+\gls{ABC}-\gls{SMC} algorithm described in \cref{subsec:abcalg}, except that
+the sequence of tolerances $\varepsilon_i$ is automatically determined rather
+than specified in advance. The tolerances are chosen such that the \gls{ESS} of
+the particle population, which indicates the quality of the Monte Carlo
+approximation (see \cref{subsec:sis}), decays at a controlled rate. A
+sudden precipitous drop in \gls{ESS} would indicate that only a small number of
+particles had non-zero importance weights, which would result in a very poor
+Monte Carlo approximation to the target distribution. This situation is
+referred to as the ``collapse'' of the approximation, and is mitigated by the
+adaptive approach. A single parameter $\alpha$ (not to be confused with the
+\gls{BA} model parameter) controls the decay rate, with $\varepsilon_i$ being
+chosen to satisfy
+\[
+  \ESS(w_i) = \alpha \ESS(w_{i-1}).
+\]
+Here, $w_i$ is the vector of weights at the $i$th step. Note that, since $w_i$
+depends on $\varepsilon_i$, this equation solves for the updated weights and
+the updated tolerance simultaneously. As pointed out by
+\textcite{del2012adaptive}, the equation has no analytic solution, but can be
+solved numerically by bisection. The forward kernels $K_i$ are taken to be
+\gls{MCMC} kernels with stationary distributions $\pi_{\varepsilon_i}$ and
+proposal distributions
+\[
+  q_i(\theta, \theta') \prod_{k=1}^M \Pr(z_i^{(k)'} \mid \theta'),
+\]
+where $\theta$ is the vector of model parameters and $z_k$ are $M$ datasets
+simulated according to $\theta'$. In our implementation, $q$ is either a
+Gaussian proposal for continuous parameters, or a Poisson proposal for discrete
+parameters. For the Poisson proposals, the number of steps to move the particle
+is drawn from a Poisson distribution, and the direction in which to move the
+particle is chosen uniformly at random. For both proposals, the variance was
+set equal to twice the empirical variance of the particles,
+following~\autocite{beaumont2009adaptive, del2012adaptive}. The backwards
+kernels are
+\[
+  L_{i-1}(x', x) = \frac{\pi_n(x)K(x, x')}{\pi_n(x')}.
+\]
+When substituted into \cref{eq:smcwt}, the forward kernels $K(x, x')$ and
+densities $\pi_n(x') = \pi_{\varepsilon_n}(x')$ cancel out, and we are left
+with the weight update 
+\begin{align*}
+  w_i(x) 
+    &\propto w_{i-1}(x) \frac{\pi_n(x \mid y)}{\pi_{i-1}(x \mid y)} \\
+    &= w_{i-1}(x) \frac{\pi(x) \pi_i(y \mid x)}{\pi(x) \pi_{i-1}(y \mid x)} \\
+    &= w_{i-1}(x) \frac{\sum_{k=i}^M \I_{A_{\varepsilon_i, y}}(z_k)}
+            {\sum_{k=i}^M \I_{A_{\varepsilon_{i-1}, y}}(z_k)}.
+\end{align*}
+In other words, when the distance threshold $\varepsilon_{i-1}$ is contracted
+to $\varepsilon_i$, the particles' weights are multiplied by the proportion of
+simulated datasets which are still inside the new threshold. The algorithm may
+be stopped when one of two termination conditions is reached. The user may
+specify a final tolerance $\varepsilon$, or a final acceptance rate of the
+\gls{MCMC} kernel. The latter condition stops the algorithm when the particles
+are not moving around very much, implying little change in the estimated
+target.
+
+\section{Analysis of \acrlong{BA} model with synthetic data}
+
+\subsection{Methods}
+\glsreset{BA}
+
+{\color{blue}\uline{Using synthetic data,}} we investigated four parameters
+related to the \gls{BA} contact network model, denoted \gls{N}, \gls{m},
+\gls{alpha}, \gls{I} (see \cref{subsec:pa}). The first three of these are
+parameters of the model itself, while \gls{I} is related to the simulation of
+transmission trees over the network. However, we will refer to all four as
+\gls{BA} parameters. \gls{N} denotes the total number of nodes in the network,
+or equivalently, susceptible individuals in the population. When a node is
+added to the network, \gls{m} new undirected edges are added incident to it,
+and are attached to existing nodes of degree $k$ with probability proportional
+to $k^\alpha + 1$ (\cref{subsec:pa}). To simulate transmission trees over a
+\gls{BA} network, we allowed an epidemic to spread until \gls{I} nodes were
+infected, and sampled a transmission tree at that time. The \gls{alpha}
+parameter is unitless, while \gls{m} has units of edges or connections per
+vertex, and \gls{N} and \gls{I} both have units of nodes or individuals.
+
+{\color{blue}\uline{Our investigation comprised three sets of computational
+experiments aimed at determining which of these parameters could be estimated
+from phylogenetic data, and with what degree of accuracy. First, we performed
+an exploratory analysis using kernel-\gls{PCA} to determine which of the four
+\gls{BA} parameters might be identifiable from tree shapes. This analysis was
+also used to validate our choice of the tree kernel as a distance function for
+\gls{ABC}. Second, we used grid search to investigate the accuracy and
+precision of marginal parameter estimates when all but one parameters of the
+model were held fixed. Finally, we applied the full \gls{ABC}-\gls{SMC}
+algorithm to jointly estimate the four parameters of the \gls{BA} model
+simultaneously.}}
+
+We assumed that all contacts had symmetric transmission risk, which was
+implemented by replacing each undirected edge in the network with two directed
+edges (one in each direction). Nodes in our networks followed simple \gls{SI}
+dynamics, meaning that they became infected at a rate proportional to their
+number of infected neighbours, and never recovered. We did not consider the
+time scale of the transmission trees in these simulations, only their shape.
+Therefore, the transmission rate along each edge in the network was set to 1,
+the removal rate of each node was set to 0, and all transmission trees' branch
+lengths were scaled by their mean. \textit{igraph} library's implementation of
+the BA model~\autocite{csardi2006igraph} was used to generate the graphs. The
+analyses were run on Westgrid (\url{https://www.westgrid.ca/}) and a local
+computer cluster. With the exception of our own programs, all analyses were
+done in \software{R}, and all packages listed below are \software{R} packages.
+
+\subsubsection*{Classifiers for BA model parameters based on tree shape}
+\label{subsec:kernel}
+
+The experiments presented here involved a large number of variables which were
+varied combinatorially. For ease of exposition, we will describe a single
+experiment first, then enumerate the values of all variables for which the
+experiment was repeated. The parameters of the tree kernel, $\lambda$ and
+$\sigma$ (\cref{subsec:treeshape}) will be referred to as
+\defn{meta-parameters} to distinguish them from the parameters of the \gls{BA}
+model.
+
+The attachment power parameter \gls{alpha} was varied among three values: 0.5,
+1.0, and 1.5. For each value, the \software{sample\_pa} function in the
+\software{igraph} package was used to simulate 100 networks, with the other
+parameters set to \gls{N} = 5000 and \gls{m} = 2. This step yielded a total of
+300 networks. An epidemic was simulated on each network using our
+\software{nettree} binary until \gls{I} = 1000 nodes were infected, at which
+point 500 of them were sampled to form a transmission tree. A total of 300
+transmission trees were thus obtained, comprised of 100 trees for each of the
+three values of \gls{alpha}. The trees were ``ladderized'' so that the subtree
+descending from the left child of each node was not smaller than that
+descending from the right child. Summary statistics, such as Sackin's index and
+the ratio of internal to terminal branch lengths, were computed for each
+simulated tree using our \software{treestat} binary. The trees were visualized
+using the \software{ape} package~\autocite{paradis2004ape}. Our
+\software{treekernel} binary was used to calculate the value of the kernel for
+each pair of trees, with the meta-parameters set to $\lambda = 0.3$ and $\sigma
+= 4$. These values were stored in a symmetric 300 $\times$ 300 kernel matrix.
+Similarly, we computed the \gls{nltt} statistic between each pair of trees
+using our \software{treestat} binary, and stored them in a second $300 \times
+300$ matrix.
+
+To investigate the effect of \gls{alpha} on tree shape, we constructed
+classifiers for \gls{alpha} based on three statistics. First, we used the
+\software{kernlab} package~\autocite{zeileis2004kernlab} to create a \gls{kSVR}
+classifier using the computed kernel matrix. Second, we used the
+\software{e1071} package~\autocite{meyer2015e1071} to create an ordinary
+\gls{SVR} classifier using the pairwise \gls{nltt} matrix. Finally, we
+performed an ordinary linear regression of \gls{alpha} against Sackin's index.
+Each of these classifiers was evaluated with 1000 two-fold cross-validations.
+We also performed a \gls{kPCA} projection of the kernel matrix, and used it to
+visualize the separation of the different \gls{alpha} values in the tree
+kernel's feature space. A schematic of this experiment is presented in
+\cref{fig:kernelexpt}.
+
+Similar experiments were performed with the values shown in
+\cref{tab:kernelexpt}. The other three \gls{BA} parameters, namely \gls{N},
+\gls{m}, and \gls{I}, were each varied while holding the others fixed. The
+experiments for \gls{alpha}, \gls{m}, and \gls{N} were repeated with three
+different values of \gls{I}. All experiments were repeated with trees having
+three different numbers of tips. Kernel matrices were computed for all pairs of
+the meta-parameters \gls{lambda} = \sett{0.2, 0.3, 0.4} and \gls{sigma} =
+\sett{\nicefrac18, \nicefrac14, \nicefrac12, 1, 2, 4, 8}.
+
+\begin{landscape}
+\begin{table}[ht]
+  \centering
+  \input{\tablepath/kernel-expt}
+  \caption[Variables used in tree kernel simulation experiments]
+  {
+    Values of parameters and other variables used in tree kernel simulation
+    experiments. Each row corresponds to one of the \gls{BA} model parameters.
+    One kernel matrix was created for every combination of values except the
+    one indicated in the ``varied parameter'' column, which was varied when
+    producing simulated trees.
+  }
+  \label{tab:kernelexpt}
+\end{table}
+
+\begin{table}[ht]
+  \centering
+  \input{\tablepath/gridsearch-expt}
+  \caption[Variables used in grid search experiments]
+  {
+    Variables and \gls{BA} parameter values used for grid search experiments. 
+    Trees were simulated under the test values, and compared to a grid of trees
+    simulated under the grid values. Kernel scores were used to calculate point
+    estimates and credible intervals for the test values.
+  }
+  \label{tab:gridexpt}
+\end{table}
+\end{landscape}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics{kernel-expt.pdf}
+  \caption[Schematic of experiments investigating impact of BA model parameters
+           on tree shape.]{
+    Schematic of experiments designed to investigate the impact of variations
+    in BA model parameters on transmission tree shapes. The parameters of the
+    BA model were varied one at a time. Transmission trees were simulated under
+    three different values of each parameter, then compared pairwise using the
+    tree kernel. Classifiers were constructed for each parameter, and their
+    accuracy was evaluated by cross-validation. Kernel-PCA projections were
+    used to visually examine the separation of the trees in the feature space
+    defined by the tree kernel.
+  }
+  \label{fig:kernelexpt}
+\end{figure}
+
+\subsubsection*{Grid search}
+
+As in the previous section, we will begin by describing a single experiment,
+and then list the variables for which similar experiments were performed. We
+varied \gls{alpha} along a narrowly spaced grid of values: 0, 0.01, \ldots, 2.
+For each value, fifteen networks were generated with \software{igraph}, and
+transmission trees were simulated over each using \software{nettree}. These
+trees will be referred to as ``grid trees''. Next, one further test tree was
+simulated with the test value \gls{alpha} = 0. Both the grid trees and the test
+tree had 500 tips, and were simulated with the other \gls{BA} parameters set to
+$N$ = 5000, $m$ = 2, and $I$ = 1000. The test tree was compared to each of the
+grid trees using the tree kernel, with the meta-parameters set to $\lambda =
+0.3$ and $\sigma = 4$, using the \software{treekernel} binary. The median
+kernel score was calculated for each grid value, and the scores were normalized
+such that the area under the curve was equal to 1. The grid value with the
+highest median kernel score was taken as the point estimate for the test value,
+and a 95\% credible interval was obtained using the \software{hpd} function in
+the \software{TeachingDemos} package~\autocite{snow2013teachingdemos}.
+
+Each experiment of the type just described was repeated ten times with the same
+test value. Similar experiments were performed for each of the four \gls{BA}
+parameters, with several test values and trees of varying sizes. The variables
+are listed in \cref{tab:gridexpt}. A graphical schematic of the grid search
+experiments is shown in \cref{fig:gridexpt}. 
+%Spearman's correlation was used to test whether the number of tips in the tree
+%was correlated with the acccuracy of the point estimates. One-way \gls{ANOVA}
+%was used to test whether the accuracy of the point estimates were significantly
+%higher for any parameter values. If so, the distribution of errors was examined
+%to choose a suitable post hoc test. If there appeared to be a correlation
+%between the parameter value and the accuracy of the estimate, Spearman's
+%correlation was calculated. If there were one or more particular values which
+%appeared to have lower error, we used the Wilcoxon rank-sum test.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth]{gridsearch-expt}
+  \caption[Schematic of grid search experiment.]{
+    Graphical schematic of grid search experiments used to investigate \gls{BA}
+    model parameters. Trees were simulated along a narrowly spaced grid of
+    parameter values (``grid trees''). Separate trees were simulated for a
+    small subset of the grid values (``test trees''). Each test tree was
+    compared to every grid tree using the tree kernel, and the resulting kernel
+    scores were normalized to resemble a probability density from which the
+    mode and 95\% highest density interval were calculated.
+  }
+  \label{fig:gridexpt}
+\end{figure}
+
+\subsubsection*{Approximate Bayesian computation}
+
+To test the full kernel-assisted \gls{ABC} algorithm, we simulated three trees each under
+a variety of parameter values, and ran the \software{netabc} program to
+estimate posterior distributions for the parameters. The parameter values and
+priors used are listed in \cref{tab:abcexpt}. The tree kernel meta-parameters
+were set to $\lambda = 0.3$ and $\sigma = 4$. The \gls{SMC} algorithm was run
+with 1000 particles, five sampled datasets per particle, and the $\alpha$
+parameter (not to be confused with the \gls{BA} preferential attachment
+parameter, see \cref{subsec:adaptsmc}) set to 0.95. The algorithm was
+stopped when the acceptance rate of the \gls{MH} kernel dropped below 1.5\%,
+the same criterion used by \citeauthor{del2012adaptive}. Approximate marginal
+posterior densities for each parameter were calculated using the
+\software{density} function in \software{R} applied to the final weighted
+population of particles. {\color{red}\sout{Credible intervals were obtained for
+each parameter using the \software{HPDinterval} function in the \software{coda}
+package~\autocite{plummer2006coda}.}} {\color{blue}\uline{Posterior means
+obtained for each parameter using the \software{wtd.mean} function in the
+\software{Hmisc} package~\autocite{harrell2016hmisc}. Credible intervals were
+obtained using the \software{hpd} function in the \software{TeachingDemos}
+package~\autocite{snow2013teachingdemos}.}}
+
+\begin{table}[ht]
+  \centering
+  \input{\tablepath/abc-expt}
+  \caption[Variables used in grid search experiments]
+  {
+    Variables and \gls{BA} parameter values used for \gls{ABC} validation
+    experiments. Trees were simulated under the test values, and
+    kernel-assisted \gls{ABC} was used to re-estimate posterior distributions for the
+    \gls{BA} parameters without training.
+  }
+  \label{tab:abcexpt}
+\end{table}
+
+%\subsubsection*{Characterization of power-law exponent in Barab\'asi-Albert networks}
+%
+%Most studies of social network or transmission network
+%parameters~\autocite[e.g.][]{liljeros2001web, jones2003assessment,
+%schneeberger2004scale, brown2011transmission} report the coefficient
+%\gls{gamma} of the power law degree distribution. To make our results
+%comparable to previous work, we used simulated networks to investigate the
+%relationship between the \gls{BA} model parameters and \gls{gamma}. A network
+%was simulated for each combination of parameters listed in
+%\cref{tab:gammaexpt}. A power law distribution was fitted to the degree
+%distribution of each simulated network using the \software{fit\_power\_law}
+%function in \software{igraph} with the `R.mle' implementation. We fitted a
+%\gls{GLM} with Gamma-distributed errors and a log link function to the observed
+%distribution of \gls{gamma} values, with \gls{alpha}, \gls{m}, \gls{N}, and all
+%possible interaction terms as predictors.
+%
+%\begin{table}
+%  \centering
+%  \input{\tablepath/gamma-expt}
+%  \caption[\gls{BA} parameters used as input \gls{GLM} predicting $\gamma$]
+%  {
+%    \gls{BA} model parameters used as input to \gls{GLM} predicting power law
+%    exponent $\gamma$. One network was simulated with each combination of
+%    parameters, and $\gamma$ was calculated for each network. A \gls{GLM} with
+%    Gamma-distributed errors and a log link function was fit to the $\gamma$
+%    values with all parameters and interaction terms as predictors.
+%  }
+%  \label{tab:gammaexpt}
+%\end{table}
+
+%To evaluate the effects of the true parameter values on the accuracy of the
+%\gls{MAP} estimates, we analysed each parameter individually using a \gls{GLM}
+%(four \glspl{GLM} total). The response variable was the error of the point
+%estimate, and the predictor variables were the true values of \gls{alpha},
+%\gls{I}, and \gls{m}. We did not test for differences across true values of
+%\gls{N}, because \gls{N} was not varied in these simulations. The distribution
+%family and link function for each \gls{GLM} were chosen by examination of
+%residual plots and \gls{AIC} of the model fits. For \gls{alpha}, \gls{I}, and
+%\gls{N}, we used the Gamma distribution with the inverse link function. For
+%\gls{m}, we used the Poisson distribution with the log link function. To
+%allow for the possibility that the relationships between parameters and error
+%rates were nonlinear, the true parameter values were treated as discrete rather
+%than continuous predictors. When significant associations were found by the
+%\glspl{GLM}, additional post hoc tests were performed based on the observed
+%relationship between the error rate and true parameter value in question. If
+%the relationship appeared to be monotonic, Spearman's correlation was
+%calculated. Otherwise, if there were particular values which appeared to have a
+%higher error rate than the others, the Wilcoxon rank-sum test was used to check
+%for a significant difference between the two groups.
+
+Two further experiments were performed to address potential sources of error.
+To evaluate the effect of model misspecification in the case of heterogeneity
+among nodes, we generated a network where half the nodes were attached with
+power $\alpha$ = 0.5, and the other half with power $\alpha$ = 1.5. The other
+parameters for this network were $N$ = 5000, $I$ = 1000, and $m$ = 2. To
+investigate the effects of potential sampling
+bias~\autocite{karcher2016quantifying}, we simulated a transmission tree where
+the tips were sampled in a peer-driven fashion, rather than at random. That is,
+the probability to sample a node was twice as high if any of that node's
+network peers had already been sampled. The parameters of this network were $N$
+= 5000, $I$ = 2000, $m$ = 2, and $\alpha$ = 0.5.
+
+\subsection{Results}
 
@@
                                 d[true_alpha == 0.5 | true_alpha == 1.5, m_error])
-    stopifnot(alpha.av["factor(true_m)", "Pr(>F)"] <= 0.05)
-    stopifnot(alpha.av["factor(true_I)", "Pr(>F)"] > 0.05)
-    stopifnot(I.I.test$p.value < 1e-5)
-    stopifnot(I.av["factor(true_m)", "Pr(>F)"] > 0.05)
-    stopifnot(m.av["factor(true_m)", "Pr(>F)"] > 0.05)
-    stopifnot(m.av["factor(true_I)", "Pr(>F)"] > 0.05)
-    stopifnot(m.av["factor(true_alpha)", "Pr(>F)"] > 0.05)
-    stopifnot(min(N.av[1:3,"Pr(>F)"]) > 0.05)
+    #stopifnot(alpha.av["factor(true_m)", "Pr(>F)"] <= 0.05)
+    #stopifnot(alpha.av["factor(true_I)", "Pr(>F)"] > 0.05)
+    #stopifnot(I.I.test$p.value < 1e-5)
+    #stopifnot(I.av["factor(true_m)", "Pr(>F)"] > 0.05)
+    #stopifnot(m.av["factor(true_m)", "Pr(>F)"] > 0.05)
+    #stopifnot(m.av["factor(true_I)", "Pr(>F)"] > 0.05)
+    #stopifnot(m.av["factor(true_alpha)", "Pr(>F)"] > 0.05)
+    #stopifnot(min(N.av[1:3,"Pr(>F)"]) > 0.05)
 
     # one-way ANOVAs
-    stopifnot(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"] < 0.06)
+    stopifnot(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"] > 0.05)
     stopifnot(anova(lm(alpha_error~factor(true_I), d))[1, "Pr(>F)"] > 0.05)
@@
 
-\Cref{fig:abcptm2} shows \gls{MAP} point estimates of the BA model parameters
-obtained with kernel-assisted ABC on simulated data. The estimates shown correspond only
-to the simulations where the $m$ parameter was set to 2, however the results
-for $m = 3$ and $m = 4$ were similar (\cref{fig:abcptm3,fig:abcptm4}). Average
-boundaries of 95\% HPD intervals are given in \cref{tab:abchpd}.
+\Cref{fig:abcptm2} shows {\color{red}\cout{\gls{MAP}}}
+{\color{blue}\uline{posterior mean}} point estimates of the BA model parameters
+obtained with kernel-assisted ABC on simulated data. The estimates shown
+correspond only to the simulations where the $m$ parameter was set to 2,
+however the results for $m = 3$ and $m = 4$ were similar
+(\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% HPD intervals are
+given in \cref{tab:abchpd}.
 
@@
     (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(zero.alpha.test$p.value, 4))}$).
-Errors in estimating $\alpha$ also varied with the true value of $m$ just at
-the threshold of statistical significance
-    (one-way ANOVA, $p 
-    \Sexpr{pp(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"], eq=TRUE)}$),
-but did not vary across the true value of $I$ (one-way ANOVA). Estimates for
-$I$ were relatively accurate, with point estimate errors of
+Errors in estimating $\alpha$ 
+%also varied with the true value of $m$ just at
+%the threshold of statistical significance
+%    (one-way ANOVA, $p 
+%    \Sexpr{pp(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"], eq=TRUE)}$),
+{\color{blue}\uline{did not vary across the true values of $m$ or $I$ (both one-way ANOVA).}} 
+Estimates for $I$ were relatively accurate, with point estimate errors of
     \Sexpr{d[,round(median(I_error))]} 
@@
     \Sexpr{as.integer(m.tbl[1] * 100)} \%
-of simulations, barely better than random guessing. The true values of the
-other parameters did not significantly affect the estimates of $m$ (both
-one-way ANOVA). Finally, the total number of nodes $N$ was consistently
-over-estimated by about a factor of two
+of simulations {\color{red}\sout{barely better than random guessing}}. The true
+values of the other parameters did not significantly affect the estimates of
+$m$ (both one-way ANOVA). Finally, the total number of nodes $N$ was
+consistently over-estimated by about a factor of two
     (error \Sexpr{d[,round(median(N_error))]} 
@@
   \caption[
-    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
+    Posterior mean point estimates for \gls{BA} model parameters obtained by
     running \software{netabc} on simulated data, for simulations with \gls{m} =
@@
   ]{
-    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
+    Posterior mean point estimates for \gls{BA} model parameters obtained by
     running \software{netabc} on simulated data, for simulations with \gls{m} =
@@
 accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows
-the distributions for for one simulation. Equivalent plots for one replicate
+the distributions for one simulation. Equivalent plots for one replicate
 simulation with each studied parameter combination can be found in
@@
 
-\subsection{Application to HIV data}
+\section{Application to real world HIV data}
+
+\subsection{Methods}
+
+{\color{blue}\uline{
+The synthetic data experiments presented in the previous section demonstrated 
+that the \gls{alpha} and \gls{I} parameters, and to a lesser extent the \gls{N}
+parameter, had measurable impacts on tree shape and were identifiable with
+kernel-assisted \gls{ABC}. Therefore, our next step was to use
+\software{netabc} to estimate the parameters of the \gls{BA} model for real
+world \gls{HIV} epidemics. }} Because the \gls{BA} model assumes a single
+connected contact network, it is most appropriate to apply to groups of
+individuals who are epidemiologically related. Therefore, we searched for
+published \gls{HIV} datasets which originated from existing clusters, either
+phylogenetically or geographically defined. In addition, we analysed an
+in-house dataset sampled from \gls{HIV}-positive individuals in British
+Columbia, Canada (the ``BC data''). The datasets are summarized in
+\cref{tab:data}.
+
+\begin{table}[ht]
+  \centering
+  \input{\tablepath/realdata-bc}
+  \caption[Characteristics of published HIV datasets analyzed with \software{netabc}.]
+  {
+    Characteristics of published HIV datasets analyzed with \software{netabc}. 
+    Abbreviations: MSM, men who have sex with men; HET, heterosexual; IDU,
+    injection drug users. The \textcite{novitsky2013phylogenetic,novitsky2014impact} data
+    were sampled from a primarily heterosexual risk environment, but did not
+    explicitly exclude other risk factors.
+  }
+  \label{tab:data}
+\end{table}
+
+We downloaded all sequences associated with each published study from GenBank.
+For the \textcite{novitsky2014impact} data, each \textit{env} sequence was
+aligned pairwise to the HXB2 reference sequence (GenBank accession number
+K03455) and the hypervariable regions were clipped out with
+\software{BioPython} version 1.66+~\autocite{cock2009biopython}. Sequences were
+multiply aligned using \software{MUSCLE} version 3.8.31
+\autocite{edgar2004muscle}, and alignments were manually inspected with
+\software{Seaview} version 4.4.2 \autocite{gouy2010seaview}. Phylogenies were
+constructed from the nucleotide alignments by approximate maximum likelihood
+using \software{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
+\gls{GTR} model~\autocite{tavare1986some}. Transmission trees were estimated by
+rooting and time-scaling the phylogenies by root-to-tip regression, using a
+modified version of Path-O-Gen (distributed as part of
+BEAST~\autocite{drummond2007beast}) as described
+previously~\autocite{poon2015phylodynamic}.
+
+Three of the datasets \autocite[][and the BC data]{li2015hiv,novitsky2014impact}
+were initially much larger than the others, containing 1265, 1299, and 7923
+sequences respectively. To ensure that the analyses were comparable, we reduced
+these to a number of sequences similar to the smaller datasets. For the
+\citeauthor{li2015hiv} and BC datasets, we detected clusters of size 280 and
+399 respectively using a patristic distance cutoff of 0.02 as described
+previously~\autocite{poon2015impact}. Only sequences within these clusters were
+carried forward. For the \textcite{novitsky2014impact} data, no large clusters
+were detected using the same cutoff, so we analysed a subtree of size 180
+chosen arbitrarily.
+
+Empirical studies of contact networks often report the exponent $\gamma$ of the
+power law degree distribution. To compare our results to the literature, we
+simulated 100 networks each according to the \gls{MAP} parameter estimates
+obtained for each investigated dataset. The power-law exponent $\gamma$ was
+calculated for each network using the \software{fit\_power\_law} function in
+\software{igraph}, with the `R.mle' implementation. The median of the 100
+$\gamma$ values was taken as a point estimate for the associated dataset.
+
+For all datasets, we used the priors $\alpha$ $\sim$ Uniform(0, 2) and $N$ and
+$I$ jointly uniform on the region \{$n \leq N \leq 10000$, $n \leq I \leq
+10000$, $I \leq N$\}, where $n$ is the number of tips in the tree (see
+\cref{tab:data}). Since the value $m = 1$ produces networks with no cycles,
+which we considered fairly implausible, we ran one analysis with the prior $m
+\sim$ DiscreteUniform(1, 5), and one with the prior $m \sim$ DiscreteUniform(2,
+5). The other parameters to the SMC algorithm were the same as used for the
+simulation experiments, except that we used 10000 particles instead of 1000 to
+increase the accuracy of the estimated posterior. This was computationally
+feasible due to the small number of runs required for this analysis.
+
+\subsection{Results}
 

### 2016-05-21 doc/thesis/methods_results.tex
Most of Alex's changes done

--- 2016-05-21 doc/thesis/methods_results.Rnw
@@
-<<setup, include=FALSE>>=
-    source("global.R")
-@
-
 \section{\software{Netabc}: a computer program for estimation of contact
@@
 each parameter using the \software{HPDinterval} function in the \software{coda}
-package~\autocite{plummer2006coda}.}} {\color{blue}\uline{Posterior means
-obtained for each parameter using the \software{wtd.mean} function in the
-\software{Hmisc} package~\autocite{harrell2016hmisc}. Credible intervals were
-obtained using the \software{hpd} function in the \software{TeachingDemos}
-package~\autocite{snow2013teachingdemos}.}}
+package~\autocite{plummer2006coda}.}} {\color{blue}\uline{Posterior means and
+credible intervals were obtained for each parameter using the
+\software{wtd.mean} and \software{wtd.quantile} functions in the
+\software{Hmisc} package~\autocite{harrell2016hmisc}. }}
 
@@
 
-<<treestats, include=FALSE>>=
-    params <- c("alpha", "I", "m", "N")
-
-    get.treestats <- function (param) {
-        stats <- paste0("../../simulations/kernel-", param, "/statistics/*")
-        d <- collect.data(stats)
-        trees <- paste0("../../simulations/kernel-", param, "/tree/*")
-        m <- collect.metadata(trees)
-        rownames(m) <- sub("../../simulations/", "", rownames(m))
-        d <- merge(d, m, by=0, suffixes=c("", ".1"))
-        if (param == "alpha") {
-            d <- subset(d, m == 2)
-            colnames(d)[colnames(d) == "nsimnode"] <- "I"
-        }
-        setDT(d)
-    }
-
-    alpha.stats <- get.treestats("alpha")
-    I.stats <- get.treestats("I")
-    m.stats <- get.treestats("m")
-    N.stats <- get.treestats("N")
-
-    alpha.sackin.cor <- alpha.stats[,cor.test(alpha, sackin, method="spearman")]
-    I.sackin.cor <- I.stats[,cor.test(I, sackin, method="spearman")]
-    m.sackin.cor <- m.stats[,cor.test(m, sackin, method="spearman")]
-    N.sackin.cor <- N.stats[,cor.test(N, sackin, method="spearman")]
-    stopifnot(alpha.sackin.cor$p.value < 0.05)
-    stopifnot(I.sackin.cor$p.value < 0.05)
-    stopifnot(m.sackin.cor$p.value < 0.05)
-    stopifnot(N.sackin.cor$p.value < 0.05)
-
-    alpha.ratio.cor <- alpha.stats[,cor.test(int.tip.ratio, alpha)]
-    I.ratio.cor <- I.stats[,cor.test(int.tip.ratio, I)]
-    m.ratio.cor <- m.stats[,cor.test(int.tip.ratio, m)]
-    N.ratio.cor <- N.stats[,cor.test(int.tip.ratio, N)]
-    stopifnot(alpha.ratio.cor$p.value < 1e-5)
-    stopifnot(I.ratio.cor$p.value < 1e-5)
-    stopifnot(m.ratio.cor$p.value < 1e-5)
-    stopifnot(N.ratio.cor$p.value < 1e-5)
-@
-
 Trees simulated under different values of \gls{alpha} were visibly quite
@@
     (for $\alpha$, $I$, $m$, and $N$ respectively: Spearman's rho =
-     \Sexpr{round(alpha.sackin.cor$estimate, 2)},
-     \Sexpr{round(I.sackin.cor$estimate, 2)},
-     \Sexpr{round(m.sackin.cor$estimate, 2)},
-     \Sexpr{round(N.sackin.cor$estimate, 2)};
+     0.85,
+     \ensuremath{-0.12},
+     \ensuremath{-0.13},
+     0.09;
      $p$-values
-     $\Sexpr{pp(alpha.sackin.cor$p.value)}$,
-     $\Sexpr{pp(I.sackin.cor$p.value)}$,
-     $\Sexpr{pp(m.sackin.cor$p.value)}$,
-     $\Sexpr{pp(N.sackin.cor$p.value)}$).
+     ${<}10^{-5}$,
+     $0.003$,
+     ${<}10^{-5}$,
+     ${<}10^{-5}$).
 The ratio of internal to terminal branch lengths was negatively correlated with
@@
   (Spearman's rho
-    \Sexpr{round(alpha.ratio.cor$estimate, 2)},
-    \Sexpr{round(I.ratio.cor$estimate, 2)},
-    \Sexpr{round(m.ratio.cor$estimate, 2)},
-    \Sexpr{round(N.ratio.cor$estimate, 2)};
+    \ensuremath{-0.84},
+    \ensuremath{-0.69},
+    0.1,
+    0.18;
   all $p < 10^{-5}$).
@@
 
-<<classifiers, include=FALSE>>=
-    get.kernel.data <- function (param, step) {
-        md <- collect.metadata(paste0('../../simulations/kernel-', param, '/', step, '/*'))
-        if ('rbf_variance' %in% colnames(md)) {
-          md <- subset(md, rbf_variance == 4 & decay_factor == 0.3)
-        }
-        if (param == 'alpha') {
-            md <- subset(md, m == 2)
-        }
-        d <- setDT(collect.data(rownames(md)))
-        if (param %in% c('alpha', 'm')) {
-            d[,list(rsquared=mean(rsquared)), by=c('nsimnode', 'ntip')]
-        } else if (param == 'I') {
-            d[,list(rsquared=mean(rsquared)), by=ntip]
-        } else {
-            d[,list(rsquared=mean(rsquared)), by=c('I', 'ntip')]
-        }
-    }
-    kernel.alpha <- get.kernel.data('alpha' ,'classifier')
-    kernel.m <- get.kernel.data('m', 'classifier')
-    kernel.I <- get.kernel.data('I', 'classifier')
-    kernel.N <- get.kernel.data('N', 'classifier')
-
-    sackin.alpha <- get.kernel.data('alpha', 'stats-classifier')
-    sackin.m <- get.kernel.data('m', 'stats-classifier')
-    sackin.I <- get.kernel.data('I', 'stats-classifier')
-    sackin.N <- get.kernel.data('N', 'stats-classifier')
-
-    nltt.alpha <- get.kernel.data('alpha', 'nltt-classifier')
-    nltt.m <- get.kernel.data('m', 'nltt-classifier')
-    nltt.I <- get.kernel.data('I', 'nltt-classifier')
-    nltt.N <- get.kernel.data('N', 'nltt-classifier')
-@
-
 Accuracy of the \gls{kSVR} classifiers varied based on the parameter being
@@
 When classifying $\alpha$, the \gls{kSVR} classifier had an average $R^2$ of 
-    \Sexpr{kernel.alpha[,round(mean(rsquared), 2)]},
+    0.92,
 compared to 
-    \Sexpr{nltt.alpha[,round(mean(rsquared), 2)]}
+    0.56
 for the \gls{nltt}-based SVR, and
-    \Sexpr{sackin.alpha[,round(mean(rsquared), 2)]}
+    0.75
 for the linear regression against Sackin's index. There was little variation
@@
 $R^2$ values of 
-  \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for \gls{kSVR},
-  \Sexpr{nltt.m[,round(mean(rsquared), 2)]} for the \gls{nltt}, and
-  \Sexpr{sackin.m[,round(mean(rsquared), 2)]}
+  0.12 for \gls{kSVR},
+  0.01 for the \gls{nltt}, and
+  0.06
 for Sackin's index. Again, there was little variation in accuracy between
@@
     (average $R^2$ 
-     \Sexpr{kernel.m[ntip == 100,round(mean(rsquared), 2)]},
-     \Sexpr{kernel.m[ntip == 500,round(mean(rsquared), 2)]},
-     \Sexpr{kernel.m[ntip == 1000,round(mean(rsquared), 2)]}
+     0.01,
+     0.11,
+     0.32
      for 100, 500, and 1000 tips respectively).
@@
 the tree. For 100-tip trees, the average $R^2$ values were
-  \Sexpr{kernel.I[ntip == 100, round(mean(rsquared), 2)]},
-  \Sexpr{nltt.I[ntip == 100, round(mean(rsquared), 2)]}, and
-  \Sexpr{sackin.I[ntip == 100, round(mean(rsquared), 2)]}
+  0.7,
+  0.55, and
+  0.02
 for the tree kernel, \gls{nltt}, and Sackin's index respectively. For 500-tip
 trees, the values increased to
-  \Sexpr{kernel.I[ntip == 500, round(mean(rsquared), 2)]},
-  \Sexpr{nltt.I[ntip == 500, round(mean(rsquared), 2)]}, and
-  \Sexpr{sackin.I[ntip == 500, round(mean(rsquared), 2)]}.
+  0.93,
+  0.83, and
+  0.07.
 Finally, the performance of classifiers for $N$ depended heavily on the
 epidemic scenario. The $R^2$ of the \gls{kSVR} classifier ranged from
-  \Sexpr{kernel.N[,round(min(rsquared), 2)]}
+  0.08
 for the smallest epidemic and smallest sample size, to
-  \Sexpr{kernel.N[,round(max(rsquared), 2)]}
+  0.82
 for the largest. Likewise, $R^2$ for the \gls{nltt}-based SVR ranged from 
-  \Sexpr{nltt.N[,round(min(rsquared), 2)]}
+  0.01
 to
-  \Sexpr{nltt.N[,round(max(rsquared), 2)]}.
+  0.54.
 Sackin's index did not accurately classify $N$ in any scenario, with an average
 $R^2$ of
-  \Sexpr{sackin.N[,round(mean(rsquared), 2)]}
+  0.03
 and little variation between scenarios.
@@
 
-<<gridsearch, include=FALSE>>=
-    d <- setDT(collect.data("../../simulations/aggregates/gridsearch/*"))
-    d[,error := abs(true_value - point.est)]
-    setkey(d, parameter)
-
-    # correlation between number of tips and error
-    tip.cor <- d[,cor.test(error, tips, method="spearman")[c("p.value", "estimate")], by=parameter]
-    setkey(tip.cor, parameter)
-    stopifnot(tip.cor["alpha", p.value] < 0.05)
-    stopifnot(tip.cor["I", p.value] < 0.05)
-    stopifnot(tip.cor["N", p.value] < 0.05)
-    stopifnot(tip.cor["m", p.value] > 0.05)
-
-    # CI takes up >75% of grid
-    stopifnot(d["alpha", all(upper - lower >= 0.75*2)])
-    stopifnot(d["I", all(upper - lower >= 0.75*5000)])
-    stopifnot(d["N", all(upper - lower >= 0.75*15000)])
-    stopifnot(d["m", all(upper - lower >= 0.75*6)])
-
-    # correlation between value and error
-    av <- d[,list(p=summary(aov(error~factor(true_value)))[[1]][["Pr(>F)"]][1]),
-            by=parameter]
-    stopifnot(av["alpha", p] < 0.05)
-    stopifnot(av["I", p] < 0.05)
-    stopifnot(av["m", p] < 0.05)
-    stopifnot(av["N", p] < 0.05)
-
-    alpha.test <- d["alpha", wilcox.test(.SD[true_value %in% c(1.0, 1.25),error], 
-                                         .SD[!true_value %in% c(1.0, 1.25),error])]
-    stopifnot(alpha.test$p.value < 0.05)
-
-    I.test <- d["I", wilcox.test(.SD[true_value >= 2000 & true_value <= 3000, error], 
-                                 .SD[true_value < 2000 | true_value > 3000, error])]
-    stopifnot(I.test$p.value < 0.05)
-
-    m.test <- d["m", wilcox.test(.SD[true_value == 1, error], 
-                                 .SD[true_value > 1, error])]
-    stopifnot(m.test$p.value < 0.05)
-
-    N.test <- d["N", wilcox.test(.SD[true_value <= 3000, error], 
-                                 .SD[true_value > 3000, error])]
-    stopifnot(N.test$p.value < 0.05)
-
-    value.cor <- d[,cor.test(error, true_value)[c("p.value", "estimate")], by=parameter]
-    setkey(value.cor, parameter)
-    stopifnot(value.cor["I", p.value] >= 0.05)
-@
-
 The accuracy of grid search estimates largely paralleled that of the \gls{kSVR}
@@
 \gls{alpha}, \gls{I}, and \gls{N} respectively: Spearman's $\rho$ = 
-    \Sexpr{tip.cor["alpha", round(estimate, 2)]},
-    \Sexpr{tip.cor["I", round(estimate, 2)]},
-    \Sexpr{tip.cor["N", round(estimate, 2)]};
+    \ensuremath{-0.22},
+    \ensuremath{-0.51},
+    \ensuremath{-0.16};
 $p$-values
-    $\Sexpr{tip.cor["alpha", pp(p.value)]}$,
-    $\Sexpr{tip.cor["I", pp(p.value)]}$,
-    $\Sexpr{tip.cor["N", pp(p.value)]}$).
+    $4\!\times\!10^{-4}$,
+    ${<}10^{-5}$,
+    $0.01$).
 The highest density intervals obtained for all parameters were extremely wide,
@@
 estimates having an average deviation of 
-    \Sexpr{d["alpha", round(mean(error), 2)]}
+    0.14
 from the true value, on a grid from 0 to 2. The error of point estimates varied
 significantly between true values of \gls{alpha}
-    (one-way \gls{ANOVA}, $p$ $\Sexpr{pp(av["alpha", p], eq=TRUE)}$). In
+    (one-way \gls{ANOVA}, $p$ ${<}10^{-5}$). In
 particular, errors were lower for the values \gls{alpha} = 1.0 and 1.25 than
@@
     (average errors 
-    \Sexpr{d["alpha", .SD[true_value %in% c(1.0, 1.25), round(mean(error), 2)]]}
+    0.03
     for \gls{alpha} = 1.0 or 1.5 vs.
-    \Sexpr{d["alpha", .SD[!true_value %in% c(1.0, 1.25), round(mean(error), 2)]]}
+    0.17
     for \gls{alpha} $\neq$ 1.0 or 1.5),
 and this difference was significant
-    (Wilcoxon rank-sum test, $p \Sexpr{pp(alpha.test$p.value, eq=TRUE)}$,
+    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
      \cref{fig:gridptalpha}).
@@
 The average absolute error of the point estimates for \gls{I} was 
-    \Sexpr{d["I", round(mean(error))]} individuals,
+    310 individuals,
 on a grid of 500 to 5000, and these errors differed between true values of
 \gls{I}
-    (one-way \gls{ANOVA}, $p \Sexpr{pp(av["I", p], eq=TRUE)}$).
+    (one-way \gls{ANOVA}, $p =0.001$).
 The errors for $2000 \leq I \leq 3000$ were higher than those for the other
@@
     (average errors
-     \Sexpr{d["I", .SD[true_value >= 2000 & true_value <= 3000, round(mean(error))]]}
+     430
      for $2000 \leq I \leq 3000$ vs.
-     \Sexpr{d["I", .SD[true_value < 2000 | true_value > 3000, round(mean(error))]]}
+     250
      for $I < 2000$ or $I > 3000$),
 and this difference was significant
-    (Wilcoxon rank-sum test, $p \Sexpr{pp(I.test$p.value, eq=TRUE)}$,
+    (Wilcoxon rank-sum test, $p =6\!\times\!10^{-4}$,
      \cref{fig:gridptI}).
@@
 The average error for \gls{m} was
-    \Sexpr{d["m", round(mean(error), 2)]} edges per vertex,
+    1.31 edges per vertex,
 on a grid from 1 to 6. The error varied significantly between the true values
 of \gls{m} 
-    (one-way \gls{ANOVA}, $p \Sexpr{pp(av["m", p], eq=TRUE)}$).
+    (one-way \gls{ANOVA}, $p {<}10^{-5}$).
 Errors for the value \gls{m} = 1 were lower than the other values
     (average errors
-    \Sexpr{d["m", .SD[true_value == 1, round(mean(error), 2)]]}
+    0.1
     for $m = 1$ vs.
-    \Sexpr{d["m", .SD[true_value > 1, round(mean(error), 2)]]}
+    1.55
     for $m > 1$),
 and this difference was significant
-    (Wilcoxon rank-sum test, $p \Sexpr{pp(m.test$p.value, eq=TRUE)}$,
+    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
      \cref{fig:gridptm}).
@@
 The average error for \gls{N} was 
-    \Sexpr{d[parameter == "N", round(mean(error))]} individuals,
+    2419 individuals,
 on a grid from 1000 to 15000, and was varied significantly with the true value
 of \gls{N}
-    (one-way \gls{ANOVA}, $p \Sexpr{pp(av["N", p], eq=TRUE)}$).
+    (one-way \gls{ANOVA}, $p {<}10^{-5}$).
 The errors were lower for $N \leq 3000$
     (average errors
-     \Sexpr{d["N", .SD[true_value <= 3000, round(mean(error))]]}
+     740
      for $N \leq 3000$ vs.
-     \Sexpr{d["N", .SD[true_value > 3000, round(mean(error))]]}
+     2979
      for $N > 3000$),
 and this difference was significant
-    (Wilcoxon rank-sum test, $p \Sexpr{pp(N.test$p.value, eq=TRUE)}$,
+    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
      \cref{fig:gridptN}).
@@
 
-<<point_est, include=FALSE>>=
-    f <- Sys.glob("../../simulations/abc-pa-free-m/point-estimate/*")
-    d <- fread(f)
-    d[,m := floor(m)]
-    d[,alpha_error := abs(true_alpha - alpha)]
-    d[,N_error := abs(true_N - N)]
-    d[,I_error := abs(true_I - I)]
-    d[,m_error := abs(true_m - m)]
-
-    alpha.av <- anova(lm(alpha_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-    m.av <- anova(lm(m_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-    N.av <- anova(lm(N_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-    I.av <- anova(lm(I_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-
-    m.tbl <- d[,prop.table(table(m_error))]
-    zero.alpha.test <- wilcox.test(d[true_alpha == 0, alpha_error], 
-                                   d[true_alpha != 0, alpha_error])
-    alpha.I.test <- wilcox.test(d[true_alpha < 1, I_error],
-                                d[true_alpha >= 1, I_error])
-    I.I.test <- wilcox.test(d[true_I == 1000, I_error],
-                            d[true_I == 2000, I_error])
-    alpha.m.test <- wilcox.test(d[true_alpha == 0 | true_alpha == 1, m_error],
-                                d[true_alpha == 0.5 | true_alpha == 1.5, m_error])
-    #stopifnot(alpha.av["factor(true_m)", "Pr(>F)"] <= 0.05)
-    #stopifnot(alpha.av["factor(true_I)", "Pr(>F)"] > 0.05)
-    #stopifnot(I.I.test$p.value < 1e-5)
-    #stopifnot(I.av["factor(true_m)", "Pr(>F)"] > 0.05)
-    #stopifnot(m.av["factor(true_m)", "Pr(>F)"] > 0.05)
-    #stopifnot(m.av["factor(true_I)", "Pr(>F)"] > 0.05)
-    #stopifnot(m.av["factor(true_alpha)", "Pr(>F)"] > 0.05)
-    #stopifnot(min(N.av[1:3,"Pr(>F)"]) > 0.05)
-
-    # one-way ANOVAs
-    stopifnot(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"] > 0.05)
-    stopifnot(anova(lm(alpha_error~factor(true_I), d))[1, "Pr(>F)"] > 0.05)
-    stopifnot(anova(lm(I_error~true_m, d))[1, "Pr(>F)"] > 0.05)
-    stopifnot(anova(lm(m_error~factor(true_alpha), d))[1, "Pr(>F)"] > 0.05)
-    stopifnot(anova(lm(m_error~factor(true_I), d))[1, "Pr(>F)"] > 0.05)
-
-    stopifnot(anova(lm(N_error~factor(true_alpha), d))[1, "Pr(>F)"] > 0.05)
-    stopifnot(anova(lm(N_error~factor(true_I), d))[1, "Pr(>F)"] > 0.05)
-    stopifnot(anova(lm(N_error~true_m, d))[1, "Pr(>F)"] > 0.05)
-
-    alpha.m.cor <- cor.test(d[,alpha_error], d[,true_m], method="spearman")
-@
-
-\Cref{fig:abcptm2} shows {\color{red}\cout{\gls{MAP}}}
-{\color{blue}\uline{posterior mean}} point estimates of the BA model parameters
-obtained with kernel-assisted ABC on simulated data. The estimates shown
-correspond only to the simulations where the $m$ parameter was set to 2,
-however the results for $m = 3$ and $m = 4$ were similar
-(\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% HPD intervals are
-given in \cref{tab:abchpd}.
+\Cref{fig:abcptm2} shows \gls{MAP} point estimates of the BA model parameters
+obtained with kernel-assisted ABC on simulated data. The estimates shown correspond only
+to the simulations where the $m$ parameter was set to 2, however the results
+for $m = 3$ and $m = 4$ were similar (\cref{fig:abcptm3,fig:abcptm4}). Average
+boundaries of 95\% HPD intervals are given in \cref{tab:abchpd}.
 
@@
 median [IQR] absolute error of 
-    \Sexpr{d[,round(median(alpha_error), 2)]} 
-    [\Sexpr{d[,round(quantile(alpha_error, 0.25), 2)]} - 
-    \Sexpr{d[,round(quantile(alpha_error, 0.75), 2)]}].
+    0.11 
+    [0.03 - 
+    0.25].
 The errors when the true value of $\alpha$ was zero were significantly greater
 than those for the other values 
-    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(zero.alpha.test$p.value, 4))}$).
+    (Wilcoxon rank-sum test, $p$ = $0$).
 Errors in estimating $\alpha$ 
@@
 %    (one-way ANOVA, $p 
-%    \Sexpr{pp(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"], eq=TRUE)}$),
+%    pp(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"], eq=TRUE)$),
 {\color{blue}\uline{did not vary across the true values of $m$ or $I$ (both one-way ANOVA).}} 
 Estimates for $I$ were relatively accurate, with point estimate errors of
-    \Sexpr{d[,round(median(I_error))]} 
-    [\Sexpr{d[,round(quantile(I_error, 0.25))]} - 
-    \Sexpr{d[,round(quantile(I_error, 0.75))]}] individuals.
+    492 
+    [294 - 
+    782] individuals.
 These errors were significantly higher when the true value of $\alpha$ was
 at least 1
-    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(alpha.I.test$p.value, 4))}$)
+    (Wilcoxon rank-sum test, $p$ = $0$)
 and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
@@
 The $m$ parameter was estimated correctly in only
-    \Sexpr{as.integer(m.tbl[1] * 100)} \%
-of simulations {\color{red}\sout{barely better than random guessing}}. The true
-values of the other parameters did not significantly affect the estimates of
-$m$ (both one-way ANOVA). Finally, the total number of nodes $N$ was
-consistently over-estimated by about a factor of two
-    (error \Sexpr{d[,round(median(N_error))]} 
-    [\Sexpr{d[,round(quantile(N_error, 0.25))]} - 
-     \Sexpr{d[,round(quantile(N_error, 0.75))]}] individuals).
+    37 \%
+of simulations, barely better than random guessing. The true values of the
+other parameters did not significantly affect the estimates of $m$ (both
+one-way ANOVA). Finally, the total number of nodes $N$ was consistently
+over-estimated by about a factor of two
+    (error 4153 
+    [3660 - 
+     4489] individuals).
 No parameters influenced the accuracy of the $N$ estimates (all one-way ANOVA).
@@
   \caption[
-    Posterior mean point estimates for \gls{BA} model parameters obtained by
+    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
     running \software{netabc} on simulated data, for simulations with \gls{m} =
@@
   ]{
-    Posterior mean point estimates for \gls{BA} model parameters obtained by
+    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
     running \software{netabc} on simulated data, for simulations with \gls{m} =
@@
 
-<<supp_labels, echo=FALSE, results="asis">>=
-    N <- 5000
-    replicate <- 0
-    lab <- NULL
-    for (m in c(2, 3, 4)) {
-    for (I in c(1000, 2000)) {
-    for (alpha in c(0, 0.5, 1, 1.5)) {
-        lab <- c(lab, sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate))
-    }
-    }
-    }
-    lab <- paste(lab, collapse = ",")
-@
-
 The dispersion of the ABC approximation to the posterior also varied between
@@
 simulation with each studied parameter combination can be found in
-\cref{\Sexpr{lab}}. HPD intervals around $\alpha$ and $I$ were often narrow
+\cref{fig:0.0-1000-2-5000-0,fig:0.5-1000-2-5000-0,fig:1.0-1000-2-5000-0,fig:1.5-1000-2-5000-0,fig:0.0-2000-2-5000-0,fig:0.5-2000-2-5000-0,fig:1.0-2000-2-5000-0,fig:1.5-2000-2-5000-0,fig:0.0-1000-3-5000-0,fig:0.5-1000-3-5000-0,fig:1.0-1000-3-5000-0,fig:1.5-1000-3-5000-0,fig:0.0-2000-3-5000-0,fig:0.5-2000-3-5000-0,fig:1.0-2000-3-5000-0,fig:1.5-2000-3-5000-0,fig:0.0-1000-4-5000-0,fig:0.5-1000-4-5000-0,fig:1.0-1000-4-5000-0,fig:1.5-1000-4-5000-0,fig:0.0-2000-4-5000-0,fig:0.5-2000-4-5000-0,fig:1.0-2000-4-5000-0,fig:1.5-2000-4-5000-0}.
+HPD intervals around $\alpha$ and $I$ were often narrow
 relative to the region of nonzero prior density, whereas the intervals for $m$
@@
 
-<<mixed_peerdriven, include=FALSE>>=
-  library(coda)
-  d <- fread("bzcat ../../simulations/abc-pa-mixed-alpha/abc/*")
-  d <- d[iter == max(iter)]
-  d <- d[sample(1:nrow(d), prob=weight, replace=TRUE)]
-  params <- c("alpha", "m", "N", "I")
-  d <- melt(d, measure.vars=params, variable.name="parameter")
-  f <- function (x) {
-    dens <- density(x)
-    hpd <- HPDinterval(mcmc(x))
-    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
-         hpd.max=hpd[,"upper"])
-  }
-  d <- d[,f(value), by=parameter]
-  setkey(d, parameter)
-
-  m <- collect.metadata("../../simulations/abc-pa-peerdriven/abc/*")
-  p <- fread(paste("bzcat", rownames(m)[m$sample_peer == 1]))
-  p <- p[iter == max(iter)]
-  p <- p[sample(1:nrow(p), prob=weight, replace=TRUE)]
-  params <- c("alpha", "m", "N", "I")
-  p <- melt(p, measure.vars=params, variable.name="parameter")
-  f <- function (x) {
-    dens <- density(x)
-    hpd <- HPDinterval(mcmc(x))
-    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
-         hpd.max=hpd[,"upper"])
-  }
-  p <- p[,f(value), by=parameter]
-  setkey(p, parameter)
-  options(scipen=5)
-@
-
 To test the effect of model misspecification, we simulated one network where
@@
 $\alpha$, 
-  \Sexpr{round(d["alpha", map], 2)} 
-  [\Sexpr{round(d["alpha", hpd.min], 2)} -
-   \Sexpr{round(d["alpha", hpd.max], 2)}];
+  1.1 
+  [0.6 -
+   1.16];
 $I$,
-  \Sexpr{d["I", round(map)]} 
-  [\Sexpr{d["I", round(hpd.min)]} -
-   \Sexpr{d["I", round(hpd.max)]}];
+  1109 
+  [662 -
+   4455];
 $m$,
-  \Sexpr{d["m", floor(map)]} 
-  [\Sexpr{d["m", floor(hpd.min)]} -
-   \Sexpr{d["m", floor(hpd.max)]}];
+  3 
+  [1 -
+   5];
 $N$,
-  \Sexpr{d["N", round(map)]} 
-  [\Sexpr{d["N", round(hpd.min)]}-
-   \Sexpr{d["N", round(hpd.max)]}].
+  12717 
+  [4021-
+   14977].
 The approximate posterior distributions for this simulation are shown in
@@
 $\alpha$, 
-  \Sexpr{p["alpha", round(map, 2)]} 
-  [\Sexpr{p["alpha", round(hpd.min, 2)]} -
-   \Sexpr{p["alpha", round(hpd.max, 2)]}];
+  0.1 
+  [0.01 -
+   0.63];
 $I$,
-  \Sexpr{p["I", floor(map)]} 
-  [\Sexpr{p["I", floor(hpd.min)]} -
-   \Sexpr{p["I", floor(hpd.max)]}];
+  2399 
+  [1423 -
+   3860];
 $m$,
-  \Sexpr{p["m", floor(map)]} 
-  [\Sexpr{p["m", floor(hpd.min)]} -
-   \Sexpr{p["m", floor(hpd.max)]}];
+  3 
+  [2 -
+   5];
 $N$,
-  \Sexpr{p["N", floor(map)]} 
-  [\Sexpr{p["N", floor(hpd.min)]} -
-   \Sexpr{p["N", floor(hpd.max)]}].
+  8762 
+  [2836 -
+   14788].
 The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
@@
 %of \gls{alpha} across all simulations was
-%    \Sexpr{d[,median(alpha_error)]} 
-%    [\Sexpr{d[,quantile(alpha_error, 0.25)]}-\Sexpr{d[,quantile(alpha_error, 0.75)]}].
+%    d[,median(alpha_error)] 
+%    [d[,quantile(alpha_error, 0.25)]-d[,quantile(alpha_error, 0.75)]].
 %\Gls{GLM} analysis indicated that the true values of both \gls{alpha} and
 %\gls{m} had significant effects on the error in estimated \gls{alpha}
-%    ($p$ values $\Sexpr{pp(alpha.glm[Parameter == "alpha", min(p)], eq=FALSE)}$
-%     $\Sexpr{pp(alpha.glm[Parameter == "m", min(p)], eq=FALSE)}$,
+%    ($p$ values $pp(alpha.glm[Parameter == "alpha", min(p)], eq=FALSE)$
+%     $pp(alpha.glm[Parameter == "m", min(p)], eq=FALSE)$,
 %     \cref{tab:glmalpha}),
@@
 %correlation between the true value of \gls{alpha} and the error
-%    (Spearman's $\rho$ = \Sexpr{round(alpha.test$estimate, 2)},
-%     $p \Sexpr{pp(alpha.test$p.value, eq=TRUE)}$).
+%    (Spearman's $\rho$ = round(alpha.test$estimate, 2),
+%     $p pp(alpha.test$p.value, eq=TRUE)$).
 %There was a negative correlation between \gls{m} and the error in \gls{alpha}
-%    (Spearman's $\rho$ = \Sexpr{round(alpha.m.cor$estimate, 2)},
-%    $p \Sexpr{pp(alpha.m.cor$p.value, eq=FALSE)}$).
+%    (Spearman's $\rho$ = round(alpha.m.cor$estimate, 2),
+%    $p pp(alpha.m.cor$p.value, eq=FALSE)$).
 %
@@
 %The mean error in the estimated value of \gls{I} was
-%    \Sexpr{d[,round(median(I_error))]} 
-%    [\Sexpr{d[,round(quantile(I_error, 0.25))]}-\Sexpr{d[,round(quantile(I_error, 0.75))]}],
+%    d[,round(median(I_error))] 
+%    [d[,round(quantile(I_error, 0.25))]-d[,round(quantile(I_error, 0.75))]],
 %an over-estimate of roughly a factor of 1.5
@@
 %values of \gls{alpha} and \gls{I}
-%    ($p$ values $\Sexpr{pp(I.glm[Parameter == "alpha", min(p)], eq=FALSE)}$
-%     $\Sexpr{pp(alpha.glm[Parameter == "I", min(p)], eq=FALSE)}$,
+%    ($p$ values $pp(I.glm[Parameter == "alpha", min(p)], eq=FALSE)$
+%     $pp(alpha.glm[Parameter == "I", min(p)], eq=FALSE)$,
 %     \cref{tab:glmalpha}),
@@
 %the true value of \gls{alpha} and the error in estimated \gls{I}
-%    (Spearman's $\rho$ = \Sexpr{I.alpha.cor$estimate},
-%     $p \Sexpr{pp(I.alpha.cor$p.value, eq=TRUE)}$),
+%    (Spearman's $\rho$ = I.alpha.cor$estimate,
+%     $p pp(I.alpha.cor$p.value, eq=TRUE)$),
 %and between the true value of \gls{I} and the error in estimated \gls{I}
-%    (Spearman's $\rho$ = \Sexpr{I.I.cor$estimate},
-%     $p \Sexpr{pp(I.I.cor$p.value, eq=TRUE)}$).
+%    (Spearman's $\rho$ = I.I.cor$estimate,
+%     $p pp(I.I.cor$p.value, eq=TRUE)$).
 %
@@
 %in the estimated \gls{m}
-%    ($p \Sexpr{pp(m.glm[Parameter == "m", min(p)])}$,
+%    ($p pp(m.glm[Parameter == "m", min(p)])$,
 %     \cref{tab:glmm}).
@@
 %\software{netabc} in virtually every case
-%    (\Sexpr{sum(m1.est == 0)} out of \Sexpr{length(m1.est)} simulations).
+%    (sum(m1.est == 0) out of length(m1.est) simulations).
 %However, when the true value of \gls{m} was 2 or higher, the correct value was
 %recovered in only 
-%    \Sexpr{as.integer(m.tbl[1] * 100)}\%
+%    as.integer(m.tbl[1] * 100)\%
 %of simulations, little better than random guessing. The \gls{GLM} analysis did
@@
 %about a factor of two
-%    (error \Sexpr{d[,format(median(N_error), scientific=FALSE)]} 
-%    [\Sexpr{d[,format(quantile(N_error, 0.25), scientific=FALSE)]} - 
-%     \Sexpr{d[,format(quantile(N_error, 0.75), scientific=FALSE)]}]).
+%    (error d[,format(median(N_error), scientific=FALSE)] 
+%    [d[,format(quantile(N_error, 0.25), scientific=FALSE)] - 
+%     d[,format(quantile(N_error, 0.75), scientific=FALSE)]]).
 %The fitted \gls{GLM} indicated that the true values of both \gls{alpha} and
 %\gls{I} had an effect on the error in the estimated \gls{N}
-%    ($p$-values $\Sexpr{pp(N.glm[Parameter == "alpha", min(p)], eq=FALSE)}$ and 
-%     $\Sexpr{pp(N.glm[Parameter == "I", min(p)], eq=FALSE)}$,
+%    ($p$-values $pp(N.glm[Parameter == "alpha", min(p)], eq=FALSE)$ and 
+%     $pp(N.glm[Parameter == "I", min(p)], eq=FALSE)$,
 %     \cref{tab:glmN}),
@@
 %    (median [IQR] error rates 
-%     \Sexpr{d[true_alpha == 1.5, format(median(N_error), scientific=FALSE)]} 
-%    [\Sexpr{d[true_alpha == 1.5, format(quantile(N_error, 0.25), scientific=FALSE)]} - 
-%     \Sexpr{d[true_alpha == 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]}]
+%     d[true_alpha == 1.5, format(median(N_error), scientific=FALSE)] 
+%    [d[true_alpha == 1.5, format(quantile(N_error, 0.25), scientific=FALSE)] - 
+%     d[true_alpha == 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]]
 %     for \gls{alpha} = 1.5 vs. 
-%     \Sexpr{d[true_alpha < 1.5, format(median(N_error), scientific=FALSE)]} 
-%    [\Sexpr{d[true_alpha < 1.5, format(quantile(N_error, 0.25), scientific=FALSE)]} - 
-%     \Sexpr{d[true_alpha < 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]}]
+%     d[true_alpha < 1.5, format(median(N_error), scientific=FALSE)] 
+%    [d[true_alpha < 1.5, format(quantile(N_error, 0.25), scientific=FALSE)] - 
+%     d[true_alpha < 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]]
 %     for \gls{alpha} < 1.5),
 %and this difference was statistically significant
-%    (Wilcoxon rank-sum test, $p \Sexpr{pp(N.alpha.test$p.value, eq=TRUE)}$).
+%    (Wilcoxon rank-sum test, $p pp(N.alpha.test$p.value, eq=TRUE)$).
 %Similarly, the error in the estimated \gls{N} when \gls{I} was equal to 3000 was
@@
 %    (median [IQR] error rates 
-%     \Sexpr{d[true_I == 3000, format(median(N_error), scientific=FALSE)]} 
-%    [\Sexpr{d[true_I == 3000, format(quantile(N_error, 0.25), scientific=FALSE)]} - 
-%     \Sexpr{d[true_I == 3000, format(quantile(N_error, 0.75), scientific=FALSE)]}]
+%     d[true_I == 3000, format(median(N_error), scientific=FALSE)] 
+%    [d[true_I == 3000, format(quantile(N_error, 0.25), scientific=FALSE)] - 
+%     d[true_I == 3000, format(quantile(N_error, 0.75), scientific=FALSE)]]
 %     for \gls{I} = 3000 vs. 
-%     \Sexpr{d[true_alpha < 3000, format(median(N_error), scientific=FALSE)]} 
-%    [\Sexpr{d[true_alpha < 3000, format(quantile(N_error, 0.25), scientific=FALSE)]} - 
-%     \Sexpr{d[true_alpha < 3000, format(quantile(N_error, 0.75), scientific=FALSE)]}]
+%     d[true_alpha < 3000, format(median(N_error), scientific=FALSE)] 
+%    [d[true_alpha < 3000, format(quantile(N_error, 0.25), scientific=FALSE)] - 
+%     d[true_alpha < 3000, format(quantile(N_error, 0.75), scientific=FALSE)]]
 %     for \gls{I} < 3000),
 %and this difference was statistically significant
-%    (Wilcoxon rank-sum test, $p \Sexpr{pp(N.I.test$p.value, eq=TRUE)}$).
+%    (Wilcoxon rank-sum test, $p pp(N.I.test$p.value, eq=TRUE)$).
 %
@@
 %for one replicate simulation with each combination of parameters can be found
-%in \cref{\Sexpr{lab}}. \Gls{HPD} intervals around \gls{alpha} and \gls{I} were
+%in \cref{lab}. \Gls{HPD} intervals around \gls{alpha} and \gls{I} were
 %narrow relative to the region of nonzero prior density, whereas the intervals
@@
 %$\alpha$, 
-%  \Sexpr{round(d["alpha", map], 2)} 
-%  [\Sexpr{round(d["alpha", hpd.min], 2)} -
-%   \Sexpr{round(d["alpha", hpd.max], 2)}];
+%  round(d["alpha", map], 2) 
+%  [round(d["alpha", hpd.min], 2) -
+%   round(d["alpha", hpd.max], 2)];
 %$I$,
-%  \Sexpr{d["I", round(map)]} 
-%  [\Sexpr{d["I", round(hpd.min)]} -
-%   \Sexpr{d["I", round(hpd.max)]}];
+%  d["I", round(map)] 
+%  [d["I", round(hpd.min)] -
+%   d["I", round(hpd.max)]];
 %$m$,
-%  \Sexpr{d["m", floor(map)]} 
-%  [\Sexpr{d["m", floor(hpd.min)]} -
-%   \Sexpr{d["m", floor(hpd.max)]}];
+%  d["m", floor(map)] 
+%  [d["m", floor(hpd.min)] -
+%   d["m", floor(hpd.max)]];
 %$N$,
-%  \Sexpr{d["N", round(map)]} 
-%  [\Sexpr{d["N", round(hpd.min)]} -
-%   \Sexpr{d["N", round(hpd.max)]}].
+%  d["N", round(map)] 
+%  [d["N", round(hpd.min)] -
+%   d["N", round(hpd.max)]].
 %The approximate posterior distributions for this simulation are shown in
@@
 %$\alpha$, 
-%  \Sexpr{p["alpha", round(map, 2)]} 
-%  [\Sexpr{p["alpha", round(hpd.min, 2)]} -
-%   \Sexpr{p["alpha", round(hpd.max, 2)]}];
+%  p["alpha", round(map, 2)] 
+%  [p["alpha", round(hpd.min, 2)] -
+%   p["alpha", round(hpd.max, 2)]];
 %$I$,
-%  \Sexpr{p["I", floor(map)]} 
-%  [\Sexpr{p["I", floor(hpd.min)]} -
-%   \Sexpr{p["I", floor(hpd.max)]}];
+%  p["I", floor(map)] 
+%  [p["I", floor(hpd.min)] -
+%   p["I", floor(hpd.max)]];
 %$m$,
-%  \Sexpr{p["m", floor(map)]} 
-%  [\Sexpr{p["m", floor(hpd.min)]} -
-%   \Sexpr{p["m", floor(hpd.max)]}];
+%  p["m", floor(map)] 
+%  [p["m", floor(hpd.min)] -
+%   p["m", floor(hpd.max)]];
 %$N$,
-%  \Sexpr{p["N", floor(map)]} 
-%  [\Sexpr{p["N", floor(hpd.min)]} -
-%   \Sexpr{p["N", floor(hpd.max)]}].
+%  p["N", floor(map)] 
+%  [p["N", floor(hpd.min)] -
+%   p["N", floor(hpd.max)]].
 %The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
@@
 
-<<realdata, include=FALSE>>=
-  m <- collect.metadata("../../simulations/aggregates/hpd/*")
-  d1 <- fread(rownames(m)[m$m_min == 1])
-  d2 <- fread(rownames(m)[m$m_min == 2])
-
-  map1 <- dcast.data.table(d1, dataset~parameter, value.var="map")
-  up1 <- dcast.data.table(d1, dataset~parameter, value.var="hpd.max")
-  low1 <- dcast.data.table(d1, dataset~parameter, value.var="hpd.min")
-  setkey(map1, dataset)
-  setkey(up1, dataset)
-  setkey(low1, dataset)
-
-  map2 <- dcast.data.table(d2, dataset~parameter, value.var="map")
-  up2 <- dcast.data.table(d2, dataset~parameter, value.var="hpd.max")
-  low2 <- dcast.data.table(d2, dataset~parameter, value.var="hpd.min")
-  setkey(map2, dataset)
-  setkey(up2, dataset)
-  setkey(low2, dataset)
-@
-
 We applied kernel-assisted ABC to five published HIV datasets (\cref{tab:data}),
@@
 estimates [95\% \gls{HPD}] 
-  \Sexpr{map1["niculescu2015", round(alpha, 2)]} 
-  [\Sexpr{low1["niculescu2015", round(alpha, 2)]} - 
-   \Sexpr{up1["niculescu2015", round(alpha, 2)]}]
+  1.04 
+  [0.04 - 
+   1.25]
 and
-  \Sexpr{map1["wang2015", round(alpha, 2)]} 
-  [\Sexpr{low1["wang2015", round(alpha, 2)]} -
-   \Sexpr{up1["wang2015", round(alpha, 2)]}] respectively).
+  0.82 
+  [0.01 -
+   1.03] respectively).
 The MAP estimates did not change appreciably when $m = 1$ was disallowed by the
@@
 data was narrower
-  (\Sexpr{low1["niculescu2015", round(alpha, 2)]} - 
-   \Sexpr{up1["niculescu2015", round(alpha, 2)]}).
+  (0.04 - 
+   1.25).
 When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
 low estimated $\alpha$ values
-  (\Sexpr{map1["li2015", round(alpha, 2)]} 
-  [\Sexpr{low1["li2015", round(alpha, 2)]} - 
-  \Sexpr{up1["li2015", round(alpha, 2)]}]
+  (0.3 
+  [0 - 
+  0.75]
 and
-  \Sexpr{map1["cuevas2009", round(alpha, 2)]} 
-  [\Sexpr{low1["cuevas2009", round(alpha, 2)]} -
-   \Sexpr{up1["cuevas2009", round(alpha, 2)]}]). 
+  0.3 
+  [0.01 -
+   0.79]). 
 However, the MAP estimates increased when $m = 1$ was not permitted, although
 the HPD intervals remained roughly the same
-  (\Sexpr{map2["li2015", round(alpha, 2)]} 
-  [\Sexpr{low2["li2015", round(alpha, 2)]} - 
-  \Sexpr{up2["li2015", round(alpha, 2)]}]
+  (0.78 
+  [0.02 - 
+  0.93]
 and
-  \Sexpr{map2["cuevas2009", round(alpha, 2)]} 
-  [\Sexpr{low2["cuevas2009", round(alpha, 2)]} -
-   \Sexpr{up2["cuevas2009", round(alpha, 2)]}]).
+  0.58 
+  [0.08 -
+   0.95]).
 The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
 for both priors on $m$
-  (\Sexpr{map1["novitsky2014", round(alpha, 2)]} for $m \geq 1$;
-   \Sexpr{map2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).
+  (0.29 for $m \geq 1$;
+   0.23 for $m \geq 2$).
 However, the confidence interval was much wider when $m = 1$ was allowed
-  ([\Sexpr{low1["novitsky2014", round(alpha, 2)]} -
-    \Sexpr{up1["novitsky2014", round(alpha, 2)]}] for $m \geq 1$ vs.
-   [\Sexpr{low2["novitsky2014", round(alpha, 2)]} -
-    \Sexpr{up2["novitsky2014", round(alpha, 2)]}] for $m \geq 2$).
+  ([0.03 -
+    1.6] for $m \geq 1$ vs.
+   [0 -
+    0.73] for $m \geq 2$).
 
@@
 intervals compared to the nonzero prior density region
-  (\citeauthor{cuevas2009hiv}, \Sexpr{map1["cuevas2009", floor(I)]} 
-  [\Sexpr{low1["cuevas2009", floor(I)]} -
-   \Sexpr{up1["cuevas2009", floor(I)]}];
-   \citeauthor{niculescu2015recent}, \Sexpr{map1["niculescu2015", floor(I)]}
-  [\Sexpr{low1["niculescu2015", floor(I)]} - 
-   \Sexpr{up1["niculescu2015", floor(I)]}];
-  \citeauthor{li2015hiv}, \Sexpr{map1["li2015", floor(I)]} 
-  [\Sexpr{low1["li2015", floor(I)]} -
-   \Sexpr{up1["li2015", floor(I)]}];
-   \citeauthor{wang2015targeting}, \Sexpr{map1["wang2015", floor(I)]}
-  [\Sexpr{low1["wang2015", floor(I)]} - 
-   \Sexpr{up1["wang2015", floor(I)]}]).
+  (\citeauthor{cuevas2009hiv}, 497 
+  [287 -
+   2430];
+   \citeauthor{niculescu2015recent}, 307
+  [138 - 
+   2822];
+  \citeauthor{li2015hiv}, 1217 
+  [383 -
+   2897];
+   \citeauthor{wang2015targeting}, 621
+  [182 - 
+   2139]).
 The \citeauthor{novitsky2014impact} data was the outlier, with a very high
 estimated $I$, and HPD interval spanning almost the entire prior region
-  (\Sexpr{map1["novitsky2014", floor(I)]} 
-  [\Sexpr{low1["novitsky2014", floor(I)]} -
-   \Sexpr{up1["novitsky2014", floor(I)]}]).
+  (7642 
+  [187 -
+   8836]).
 The $I$ estimates and HPD intervals were generally robust to the choice of
@@
 upper bound of the HPD interval was different for each dataset
-  (\citeauthor{niculescu2015recent}, \Sexpr{up1["niculescu2015", round(m)]};
-   \citeauthor{wang2015targeting}, \Sexpr{up1["wang2015", round(m)]};
-   \citeauthor{li2015hiv}, \Sexpr{up1["li2015", round(m)]};
-   \citeauthor{cuevas2009hiv}, \Sexpr{up1["cuevas2009", round(m)]}).
+  (\citeauthor{niculescu2015recent}, 5;
+   \citeauthor{wang2015targeting}, 4;
+   \citeauthor{li2015hiv}, 1;
+   \citeauthor{cuevas2009hiv}, 2).
 When $m = 1$ was disallowed, the MAP for all datasets was either 2 or 3, with
@@
 data, for which the MAP estimate was lower 
-  (\Sexpr{map1["li2015", floor(N)]})
+  (7428)
 when $m = 1$ was allowed.

### 2016-05-21 doc/thesis/objective.tex
Most of Alex's changes done

@@
 \glsreset{BA}
+\glsreset{SMC}
 
@@
 models generally assume a panmictic population. The primary objective of this
-work is to develop a method to fit contact network models in a phylodynamic
-framework.
+work is \emph{to develop a method to fit contact network models in a
+phylodynamic framework.}
 
@@
 Calculating the likelihood of the parameters of a contact network models seems
-likely to be an intractable problem, which would imply that these models are
-amenable to neither \gls{ML} nor Bayesian inference. We have not proven this is
-the case, but some intuition can be provided by examining the process involved
-in the likelihood calculation. Consider a contact network model with parameters
-$\theta$, and an estimated transmission tree $T$ with $n$ tips. In general, we
-do not know the labels of the internal nodes of $T$, only the labels of its
-tips. To fit this model using likelihood-based methods, we must calculate the
-likelihood of $\theta$, that is, $\Pr(T \mid \theta)$. Let $\G$ be the set of
-all possible contact networks, and $\Nu$ be the set of all possible labellings
-of the internal nodes of $T$. We can write the likelihood as
+likely to be an intractable problem. We have not proven this is the case, but
+some intuition can be provided by examining the process involved in the
+likelihood calculation. Consider a contact network model with parameters
+$\theta$, and an estimated transmission tree $T$ with $n$ tips.
+{\color{blue}\uline{The transmission tree is a record of an epidemic's progress
+through a network. Tips of the transmission tree correspond to sampled
+individuals, while internal nodes correspond to transmissions and may be
+labelled with the transmitting individual. Transmission trees will be described
+in more detail in the sequel, but for now they can be thought of as
+representing the exact path an epidemic has taken. }} In general, we do not
+know the labels of the internal nodes of $T$, only the labels of its tips. To
+fit this model using likelihood-based methods, we must calculate the likelihood
+of $\theta$, that is, $\Pr(T \mid \theta)$. Let $\G$ be the set of all possible
+contact networks, and $\Nu$ be the set of all possible labellings of the
+internal nodes of $T$. We can write the likelihood as
 \begin{align}
@@
 
+{\color{blue}\uline{
+In a Bayesian setting, our goal is to estimate the posterior distribution on
+model parameters given our data, $\Pr(\theta \mid T)$. Bayes' theorem relates
+the posterior distribution to the likelihood,}
+\[
+  \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)}.
+\]
+\uline{For all but the simplest mathematical models, the calculation of the
+normalizing constant $\Pr(T)$ is an intractable problem. What we have argued in
+the previous paragraph is that calculating $\Pr(T \mid \theta)$ for a single
+parameter candidate $\theta$ also seems likely to be computationally hard.
+Therefore, the problem of fitting contact network models to phylogenies seems
+to be of the \emph{doubly-intractible} type~\autocite{murray2012mcmc}, which
+would imply that these models are not amenable to neither \gls{ML} nor Bayesian
+inference techniques. Although both methods are able to cope with an
+intractible normalizing constant (for example, by local search for \gls{ML} or
+Bayesian \gls{MCMC}), neither can avoid the intractable likelihood
+calculations. }}
+
 Depending on the network model studied, it is possible that \cref{eq:netlik}
@@
 
+{\color{blue}\uline{
 Our method will expand on the framework developed
-by~\autocite{poon2015phylodynamic}, who combined \gls{ABC}-\gls{MCMC} with the
-tree kernel to infer parameters of population genetic models from viral
-phylogenies. The term ``kernel-assisted \gls{ABC}'' refers to any application
-of \gls{ABC} using a kernel function as a distance metric, including that
+by~\autocite{poon2015phylodynamic}, who combined \gls{ABC} with the tree kernel
+to infer parameters of population genetic models from viral phylogenies using
+\gls{MCMC}. The term ``kernel-assisted \gls{ABC}'' refers to any application of
+\gls{ABC} using a kernel function as a distance metric, including that
 developed by \citeauthor{poon2015phylodynamic}. However, for ease of
@@
 \textcite{fukumizu2011kernel}.
+}}
 
@@
 attachment models, which makes it a natural choice to explore with our method.
-The second aim of this work is to use simulations to investigate the parameters
-of the \gls{BA} model, including whether they have a detectable impact on tree
-shape, and whether they can be accurately recovered using kernel-assisted
-\gls{ABC}.
+The second aim of this work is \emph{to use simulations to investigate the
+parameters of the \gls{BA} model, including whether they have a detectable
+impact on tree shape, and whether they can be accurately recovered using
+kernel-assisted \gls{ABC}.}
 
@@
 any other pathogen, and including sequences sampled from diverse geographic and
-demographic contexts. At the time of this writing, there were $635400$ HIV
+demographic contexts. At the time of this writing, there were $635,400$ HIV
 sequences publicly available in GenBank, annotated with 172 distinct countries
@@
 \gls{HIV} epidemics are driven by social dynamics and are therefore likely to
-be highly nonrandom~\autocite{clemenccon2015statistical}. Moreover, since no
+be highly structured~\autocite{clemenccon2015statistical}. Moreover, since no
 cure yet exists, efforts to curtail the progression of an epidemic have relied
@@
 obvious context in which our method could be applied. Therefore, the third and
-final aim of this work is to apply kernel-assisted \gls{ABC} to fit the
-\gls{BA} model to existing \gls{HIV} outbreaks.
+final aim of this work is \emph{to apply kernel-assisted \gls{ABC} to fit the
+\gls{BA} model to existing \gls{HIV} outbreaks}.
 

### 2016-05-21 doc/thesis/results.tex
Most of Alex's changes done

@@
   [0.6 -
-   1.16];
+   1.15];
 $I$,
-  1119 
+  1133 
   [662 -
@@
 $N$,
-  12678 
-  [3367-
-   14977].
+  12655 
+  [3733-
+   14933].
 The approximate posterior distributions for this simulation are shown in
@@
 $\alpha$, 
-  0.23 
+  0.37 
   [0.01 -
@@
 $I$,
-  2398 
+  2415 
   [1419 -
-   3767];
+   3811];
 $m$,
@@
 $N$,
-  9761 
-  [2836 -
+  8973 
+  [2881 -
    14788].

### 2016-05-26 doc/thesis/discussion.tex
Started working on Sally's revisions

@@
+\label{sec:disc}
+
 \subsection{\software{Netabc}: uses, limitations, and possible extensions}
@@
 context~\autocite{welch2011statistical, bansal2007individual}. In principle,
-kernel-assisted \gls{ABC} could be adapted to dynamic contact networks by using a method
-such as that developed by \textcite{robinson2012dynamics} to simulate such a
-network, while concurrently simulating the spread of an epidemic.
+kernel-assisted \gls{ABC} could be adapted to dynamic contact networks by using
+a method such as that developed by \textcite{robinson2012dynamics} to simulate
+such a network, while concurrently simulating the spread of an epidemic.
 
@@
 process by which a viral phylogeny is generated along with a transmission tree
-into our method, for example by simulating within-host dynamics. Although this
-may be an avenue for future extension, we felt that it would obscure the
+into our method, for example by simulating within-host dynamics. {\color{blue}
+\uline{Indeed, progress has very recently been demonstrated on this front in a
+talk by \citeauthor{giardina2016inference}, who have independently developed in
+a method similar to ours to fit contact network models to phylogenetic data
+that additionally incorporates a within-host evolutionary model.}} Although
+this may be an avenue for future extension, we felt that it would obscure the
 primary purpose of this work, which is to study contact network parameters. In

### 2016-05-26 doc/thesis/intro.tex
Started working on Sally's revisions

@@
 \textcite{leitner1996accurate, paraskevis2004phylogenetic} were able to
-accurately recover a known transmission tree using a viral phylogeny. The
+accurately recover known transmission trees using viral phylogenies. The
 problem of accurately estimating transmission trees is an ongoing area of
@@
 
+{\color{blue}\uline{
+Recall that the primary objective of our work is to develop a statistical
+inference method for estimating contact network parameters from transmission
+trees. In particular, for a network model with parameters $\theta$ and an input
+transmission tree $T$, we are interested in the posterior distribution}
+\[
+  \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)}.
+\]
+\uline{As outlined in} \cref{sec:obj}, \uline{both the likelihood $\Pr(T \mid
+\theta)$ and the normalizing constant $\Pr(T)$ are likely computationally
+intractable. Hence, rather than computing the posterior distribution
+analytically, we will attempt to approximate it using a \defn{Monte Carlo}
+approach.
+TODO
+}}
+
 \glsreset{SMC}
@@
 the error terms $\varepsilon_i$; most commonly, it is assumed that they are
-normally distributed with variance $\sigma$. In mathematical terms, $Y_i \sim
+normally distributed with variance $\sigma$. In mathematical terms, $y_i \sim
 \beta x_i + \N(0, \sigma^2)$, where ``$\sim$'' means ``is distributed as''. We
@@
 to the posterior. The intuition for why this approximation might be reasonable
-comes from the fact that, when $\varepsilon = 0$ and $\rho$ satisfies $\rho(z,
-y) = 0 \Rightarrow x = y$, the integral on the left is exactly equal to the
-posterior. Thus, by taking $\varepsilon$ small, we should attain something
-close to the posterior {\color{blue}\uline{if $\rho$ captures the similarity
-between datasets reasonably well. However, the accuracy of the \gls{ABC}
-approximation depends heavily on the choice of distance function. }}
-
-In many applications {\color{purple}\uline{(eg. \autocite{fu1997estimating,
+comes from the fact that, when $\varepsilon = 0$, the $\varepsilon$-ball around
+$y$ should contain only $y$ itself, hence the integral on the left is exactly
+equal to the posterior. Thus, by taking $\varepsilon$ small, we should attain
+something close to the posterior {\color{blue}\uline{if $\rho$ captures the
+similarity between datasets reasonably well. However, the accuracy of the
+\gls{ABC} approximation depends heavily on the choice of distance function
+\autocite{aeschbacher2012novel,blum2013comparative}.}}
+
+In many applications {\color{blue}\uline{(eg. \autocite{fu1997estimating,
 tanaka2006using})}}, $\rho$ is defined as $\rho(S(\cdot), S(\cdot))$ where $S$
@@
 possible to find sufficient summary statistics~\autocite{marjoram2006modern}.
-TODO
-}}
-
-This can be useful if the data are high-dimensional or of a complex type, but
-it is not strictly necessary. For instance, if the data are numeric and of low
-dimension, the distance function may simply be the Euclidean
-distance~\autocite{sisson2007sequential}. \textcite{poon2015phylodynamic}
-proposed the use of a kernel function (defined in \cref{subsec:treeshape}),
-which allows
+A number of sophisticated methods have been developed for selecting and
+weighting summary statistics based on various optimality criteria}
+\autocite[][and references therein]{aeschbacher2012novel,
+blum2013comparative}}.
+
+Summary statistics can be useful if the data are high-dimensional or of a
+complex type, but they are not strictly necessary. For instance, if the data
+are numeric and of low dimension, the distance function may simply be the
+Euclidean distance~\autocite{sisson2007sequential}.
+{\color{blue}\textcite{park2015k2} \uline{proposed the use of a kernel function}
+(defined in \cref{subsec:treeshape}) \uline{in place of a distance function.
+The authors referred to their approach as ``double-kernel \gls{ABC}'' due to
+the use of a second kernel function to compute the weights of the particles.
+The work by \textcite{poon2015phylodynamic}, upon which ours is based, employed
+a similar approach, replacing the likelihood ratio in Bayesian \gls{MCMC} with
+a ratio of kernel scores.}}
 
@@
 %studied. Since we do not apply rejection or \gls{MCMC} in this work, we shall
-%restrict our attention to the convergence properties of \gls{SMC}.
+%}restrict our attention to the convergence properties of \gls{SMC}.
+
+\section{Summary}
+
+{\color{blue}\uline{ Our method integrates the four distinct research areas
+  just described: phylogenetics, contact networks, \acrlong{SMC}, and
+  \acrlong{ABC}. The first two topics together form the problem domain.
+  Phylogenetic data is the input to our method, while estimates of the
+  parameters of contact network models are the desired output. The latter two
+  topics define the algorithm and statistical framework that our inference
+method will use. }}

### 2016-05-26 doc/thesis/methods_results.Rnw
Started working on Sally's revisions

@@
 
+{\color{blue}\uline{
+The chapter is organized as follows. One section is devoted to each of our
+three research aims: the development of a method for inference of contact
+network parameters from transmission trees} (\cref{sec:netabc}), \uline{the
+investigation of the \acrlong{BA} network model using simulated data
+}(\cref{sec:ba}), \uline{ and the application of the method to real-world
+\gls{HIV} epidemics }(\cref{sec:hiv}). \uline{Each of these sections contains a
+description of the methods used and, in the case of the latter two objectives,
+the results of the experiments. The fourth and final section of the chapter
+}(\cref{sec:disc}) \uline{contains a unified discussion of all three
+objectives. }}
+
 \section{\software{Netabc}: a computer program for estimation of contact
 network parameters with kernel-assisted ABC}
+\label{sec:netabc}
 
@@
 contact network parameters from an estimated transmission tree using
-kernel-assisted \gls{ABC}. The program combines three major components: Gillespie
-simulation, to simulate transmission trees on contact networks; the tree
-kernel, to compare transmission trees; and adaptive \gls{ABC}-\gls{SMC}, to
-maintain a population of particles and advance it toward the \gls{ABC} target
-distribution. We give a high-level overview of the program here, before
-describing these three components in detail. \software{Netabc} takes as input
-an estimated transmission tree, which can be derived from a viral phylogeny by
+\gls{ABC}. {\color{blue}\uline{As discussed in }\cref{sec:obj}, \uline{the principal
+statistical algorithm used by \software{netabc} is adaptive
+\gls{ABC}-\gls{SMC}~\autocite{del2012adaptive}. In addition, there are two
+supplementary components which are specific to the domain of phylogenetics and
+contact networks: Gillespie simulation~\autocite{gillespie1976general}, to
+simulate transmission trees on contact networks; and the tree
+kernel~\autocite{poon2013mapping}, which is used as the distance function in
+\gls{ABC} to compare transmission trees~\autocite{poon2015phylodynamic}} (see
+\cref{sec:abc}).} We give a high-level overview of the program here, before
+describing these components in detail. \software{Netabc} takes as input an
+estimated transmission tree, which can be derived from a viral phylogeny by
 rooting and time-scaling as described in \cref{subsec:phylodynamics} or
@@
 jombart2011reconstructing, ypma2012unravelling, morelli2012bayesian,
-didelot2014bayesian}. We variously refer to this estimated transmission tree as
-the observed tree, input tree, or true tree.
+didelot2014bayesian, hall2015epidemic}. We variously refer to this estimated
+transmission tree as the observed tree, input tree, or true tree.
 
@@
 of particles $x^{(k)}$, each of which contains particular parameter values
-$\theta^{(k)}$ for the model we are trying to fit. A small number of contact
-networks $z^{(k)}$ are generated for each particle, in accordance with that
-particle's parameters. An epidemic is simulated over each of these networks
-using Gillespie simulation, and by keeping track of its progress, a
-transmission tree is obtained. Thus, each particle becomes associated with
-several simulated transmission trees. These trees are compared to the input
-tree using the tree kernel. Particles are weighted according to the similarity
-of their associated simulated trees with the true tree, with more similar trees
-receiving higher weights. The particles are iteratively perturbed to explore
-the parameter space, and particles with simulated trees too distant from the
-true tree are periodically dropped and resampled. Once a convergence criterion
-is attained, the final set of particles is used as a Monte Carlo approximation
-to the target distribution of \gls{ABC}, which is assumed to resemble the
-posterior distribution on model parameters (see \cref{sec:abc}). A graphical
-schematic of this algorithm is given in \cref{fig:abcsmc}.
+$\theta^{(k)}$ for the {\color{blue}\uline{contact network}} model we are
+trying to fit {\color{blue}\uline{to the input tree}}. A small number of
+contact networks $z^{(k)}$ are generated under the model for each particle, in
+accordance with that particle's parameters. An epidemic is simulated over each
+of these networks using Gillespie simulation, and by keeping track of its
+progress, a transmission tree is obtained. Thus, each particle becomes
+associated with several simulated transmission trees. These trees are compared
+to the input tree using the tree kernel. Particles are weighted according to
+the similarity of their associated simulated trees with the true tree, with
+more similar trees receiving higher weights. The particles are iteratively
+perturbed to explore the parameter space, and particles with simulated trees
+too distant from the true tree are periodically dropped and resampled. Once a
+convergence criterion is attained, the final set of particles is used as a
+Monte Carlo approximation to the target distribution of \gls{ABC}, which is
+assumed to resemble the posterior distribution on model parameters (see
+\cref{sec:abc}). A graphical schematic of this algorithm is given in
+\cref{fig:abcsmc}.
 
@@
 probability distributions, and to perform the bisection step in the adaptive
-\gls{ABC}-\gls{SMC} algorithm. Parallelization is implemented with POSIX
+\gls{ABC}-\gls{SMC} algorithm. Parallelization is implemented with \gls{POSIX}
 threads~\autocite{barney2009posix}. In addition to the \software{netabc} binary
@@
 simulated over four network models at various levels of pathogen
-transmissibility. Our implementation of adaptive ABC-SMC was tested by applying
-it to the same mixture of Gaussians used by \citeauthor{del2012adaptive} to
-demonstrate their method (originally used by~\textcite{sisson2007sequential}).
-We were able to obtain a close approximation to the function (see
-\cref{fig:smctest}), and attained the stopping condition used by the authors in
-a comparable number of steps. To check that the algorithm would converge to a
-bimodal distribution, we also applied it to a mixture of two Gaussians with
-means $\pm$4 and variances 1. The algorithm was able to recover both peaks
-(\cref{fig:smctest2}).
-
-\subsection{Epidemic simulation}
+transmissibility. Our implementation of adaptive \gls{ABC}-\gls{SMC} was tested
+by applying it to the same mixture of Gaussians used by
+\textcite{del2012adaptive} to demonstrate their method (originally used
+by~\textcite{sisson2007sequential}). We were able to obtain a close
+approximation to the function (see \cref{fig:smctest}), and attained the
+stopping condition used by the authors in a comparable number of steps. To
+check that the algorithm would converge to a bimodal distribution, we also
+applied it to a mixture of two Gaussians with means $\pm$4 and variances 1. The
+algorithm was able to recover both peaks (\cref{fig:smctest2}).
+
+\subsection{Simulation of transmission trees over contact networks}
 \label{subsec:nettree}
@@
 \section{Analysis of \acrlong{BA} model with synthetic data}
+\label{sec:ba}
 
@@
 from phylogenetic data, and with what degree of accuracy. First, we performed
-an exploratory analysis using kernel-\gls{PCA} to determine which of the four
-\gls{BA} parameters might be identifiable from tree shapes. This analysis was
-also used to validate our choice of the tree kernel as a distance function for
-\gls{ABC}. Second, we used grid search to investigate the accuracy and
-precision of marginal parameter estimates when all but one parameters of the
-model were held fixed. Finally, we applied the full \gls{ABC}-\gls{SMC}
-algorithm to jointly estimate the four parameters of the \gls{BA} model
-simultaneously.}}
+an exploratory analysis using a tree kernel-based classifier to determine which
+of the four \gls{BA} parameters might be identifiable from tree shapes when all
+others were held fixed. This analysis was also used to validate our choice of
+the tree kernel as a distance function for \gls{ABC}, by evaluating two further
+classifiers based on other tree summary statistics from the literature. Second,
+we used grid search to investigate the accuracy and precision of marginal
+parameter estimates, again holding all but one parameters of the model fixed.
+Finally, we applied the full \gls{ABC}-\gls{SMC} algorithm to jointly estimate
+the four parameters of the \gls{BA} model simultaneously.}}
 
@@
 done in \software{R}, and all packages listed below are \software{R} packages.
+{\color{blue}\uline{Code to run all simulation-based experiments is freely
+available at \url{https://github.com/rmcclosk/thesis}.}}
 
@@
 
-The experiments presented here involved a large number of variables which were
-varied combinatorially. For ease of exposition, we will describe a single
-experiment first, then enumerate the values of all variables for which the
-experiment was repeated. The parameters of the tree kernel, $\lambda$ and
-$\sigma$ (\cref{subsec:treeshape}) will be referred to as
-\defn{meta-parameters} to distinguish them from the parameters of the \gls{BA}
-model.
+\glsreset{nltt}
+
+{\color{blue}\uline{
+Our first computational experiment was designed as an exploratory analysis of
+the four \gls{BA} model parameters defined above: \gls{alpha}, \gls{I},
+\gls{m}, and \gls{N}. The objective of this experiment was to determine whether
+any of the four parameters might be identifiable from the shape of the
+transmission tree. TODO }
+
+\uline{In addition, a secondary objective of this section was to validate the
+use of the tree kernel as a distance measure for \gls{ABC} in our context. As
+discussed in} \cref{sec:abc}, \uline{the choice of distance function is
+extremely important for the accuracy of the \gls{ABC} approximation to the
+posterior. Therefore, we evaluated two additional tree statistics in the same
+manner as we evaluated the tree kernel (that is, by constructing and testing a
+classifier). First, we considered Sackin's index~\autocite{shao1990tree}, which
+measures the degree of unbalancedeness or asymmetry in a phylogeny} (see
+\cref{subsec:treeshape}). \uline{Sackin's index is widely used for
+characterizing phylogenies~\autocite{frost2013modelling} and has been
+demonstrated to vary between transmission trees simulated under different
+contact network types~\autocite{leventhal2012inferring}. Sackin's index does
+not take branch lengths into account, considering only the tree's topology. The
+other statistic we considered was the
+\gls{nltt}~\autocite{janzen2015approximate}, which compares two trees based on
+normalized distributions of their branching times} (see \cref{sec:treeshape}).
+\uline{In contrast with Sackin's index, the \gls{nltt} does not explicity
+consider the trees' topologies, but it does use their normalized branch
+lengths. While the \gls{nltt} is a newly developed statistic not yet in
+widespread use, the unnormalized \gls{ltt}~\autocite{nee1992tempo} was the
+basis of seminal early work extracting epidemiological information from
+phylodynamics~\autocite{holmes1995revealing}.}
+
+\uline{We expect the tree kernel to classify the \gls{BA} parameters more
+accurately than either Sackin's index or the \gls{nltt}, since the tree kernel
+takes both topology and branch lengths into account. }}
+
+This experiment involved a large number of variables which were varied
+combinatorially. For ease of exposition, we will describe a single experiment
+first, then enumerate the values of all variables for which the experiment was
+repeated. The parameters of the tree kernel, $\lambda$ and $\sigma$
+(\cref{subsec:treeshape}) will be referred to as \defn{meta-parameters} to
+distinguish them from the parameters of the \gls{BA} model.
 
@@
 
-To investigate the effect of \gls{alpha} on tree shape, we constructed
-classifiers for \gls{alpha} based on three statistics. First, we used the
-\software{kernlab} package~\autocite{zeileis2004kernlab} to create a \gls{kSVR}
-classifier using the computed kernel matrix. Second, we used the
-\software{e1071} package~\autocite{meyer2015e1071} to create an ordinary
-\gls{SVR} classifier using the pairwise \gls{nltt} matrix. Finally, we
-performed an ordinary linear regression of \gls{alpha} against Sackin's index.
-Each of these classifiers was evaluated with 1000 two-fold cross-validations.
-We also performed a \gls{kPCA} projection of the kernel matrix, and used it to
-visualize the separation of the different \gls{alpha} values in the tree
-kernel's feature space. A schematic of this experiment is presented in
-\cref{fig:kernelexpt}.
+To investigate the identifiability of \gls{alpha} from tree shape, we
+constructed classifiers for \gls{alpha} based on the three tree shape
+statistics discussed above. First, we used the \software{kernlab}
+package~\autocite{zeileis2004kernlab} to create a \gls{kSVR} classifier using
+the computed kernel matrix. Second, we used the \software{e1071}
+package~\autocite{meyer2015e1071} to create an ordinary \gls{SVR} classifier
+using the pairwise \gls{nltt} matrix. Finally, we performed an ordinary linear
+regression of \gls{alpha} against Sackin's index. Each of these classifiers was
+evaluated with 1000 two-fold cross-validations. We also performed a \gls{kPCA}
+projection of the kernel matrix, and used it to visualize the separation of the
+different \gls{alpha} values in the tree kernel's feature space. A schematic of
+this experiment is presented in \cref{fig:kernelexpt}.
 
@@
     in BA model parameters on transmission tree shapes. The parameters of the
-    BA model were varied one at a time. Transmission trees were simulated under
-    three different values of each parameter, then compared pairwise using the
-    tree kernel. Classifiers were constructed for each parameter, and their
-    accuracy was evaluated by cross-validation. Kernel-PCA projections were
-    used to visually examine the separation of the trees in the feature space
-    defined by the tree kernel.
+    BA model were varied one at a time {\color{blue}\uline{while holding all
+    others fixed}}. Transmission trees were simulated under three different
+    values of each parameter, then compared pairwise using the tree kernel.
+    Classifiers were constructed for each parameter, and their accuracy was
+    evaluated by cross-validation. Kernel-PCA projections were used to visually
+    examine the separation of the trees in the feature space defined by the
+    tree kernel.
   }
@@
 \subsubsection*{Grid search}
+
+{\color{blue}\uline{The previous experiment was an exploratory analysis
+intended to determine which of the \gls{BA} parameters were identifiable, and
+whether the tree kernel could potentially be used to distinguish different
+parameter values when all others were held fixed. In this experiment, which was
+still of an exploratory nature, we continued to consider one parameter at a
+time while fixing the other three. However, rather than checking for
+identifiability, we were now interested in quantifying the accuracy and
+precision of kernel score-based estimates. This was done by examining the
+distribution of kernel scores on a grid of parameter values, when trees
+simulated according to those values were compared with a single simulated test
+tree. This experiment did not involve the full \gls{ABC}-\gls{SMC} algorithm,
+nor did we continue to use Sackin's index or the \gls{nltt}.}}
 
@@
 tree had 500 tips, and were simulated with the other \gls{BA} parameters set to
-$N$ = 5000, $m$ = 2, and $I$ = 1000. The test tree was compared to each of the
-grid trees using the tree kernel, with the meta-parameters set to $\lambda =
-0.3$ and $\sigma = 4$, using the \software{treekernel} binary. The median
-kernel score was calculated for each grid value, and the scores were normalized
-such that the area under the curve was equal to 1. The grid value with the
-highest median kernel score was taken as the point estimate for the test value,
-and a 95\% credible interval was obtained using the \software{hpd} function in
-the \software{TeachingDemos} package~\autocite{snow2013teachingdemos}.
+{\color{blue}\uline{the known values}} $N$ = 5000, $m$ = 2, and $I$ = 1000. The
+test tree was compared to each of the grid trees using the tree kernel, with
+the meta-parameters set to $\lambda = 0.3$ and $\sigma = 4$, using the
+\software{treekernel} binary. The median kernel score was calculated for each
+grid value, and the scores were normalized such that the area under the curve
+was equal to 1. The grid value with the highest median kernel score was taken
+as the point estimate for the test value, and a 95\% credible interval was
+obtained using the \software{hpd} function in the \software{TeachingDemos}
+package~\autocite{snow2013teachingdemos}.
 
@@
 are listed in \cref{tab:gridexpt}. A graphical schematic of the grid search
-experiments is shown in \cref{fig:gridexpt}. 
+experiments is shown in \cref{fig:gridexpt}. {\color{blue}\uline{We emphasize
+that these experiments were marginal in nature; that is, the each parameter was
+varied and estimated individually while holding the others at known, fixed
+values.}}
+
 %Spearman's correlation was used to test whether the number of tips in the tree
@@
     model parameters. Trees were simulated along a narrowly spaced grid of
-    parameter values (``grid trees''). Separate trees were simulated for a
-    small subset of the grid values (``test trees''). Each test tree was
+    values of one parameter (``grid trees'') {\color{blue}\uline{with all other
+    parameters fixed to known values}}. Separate trees were simulated for a
+    small subset of the grid values (``test trees''), {\color{blue}\uline{also
+    holding the other parameters fixed}}. Each test tree was
     compared to every grid tree using the tree kernel, and the resulting kernel
@@
 
-To test the full kernel-assisted \gls{ABC} algorithm, we simulated three trees each under
-a variety of parameter values, and ran the \software{netabc} program to
-estimate posterior distributions for the parameters. The parameter values and
-priors used are listed in \cref{tab:abcexpt}. The tree kernel meta-parameters
-were set to $\lambda = 0.3$ and $\sigma = 4$. The \gls{SMC} algorithm was run
-with 1000 particles, five sampled datasets per particle, and the $\alpha$
-parameter (not to be confused with the \gls{BA} preferential attachment
-parameter, see \cref{subsec:adaptsmc}) set to 0.95. The algorithm was
-stopped when the acceptance rate of the \gls{MH} kernel dropped below 1.5\%,
-the same criterion used by \citeauthor{del2012adaptive}. Approximate marginal
-posterior densities for each parameter were calculated using the
-\software{density} function in \software{R} applied to the final weighted
-population of particles. {\color{red}\sout{Credible intervals were obtained for
-each parameter using the \software{HPDinterval} function in the \software{coda}
+{\color{blue}\uline{
+Our final simulation-based experiment was designed to test the full
+\gls{ABC}-\gls{SMC} algorithm by jointly estimating the four parameters of the
+\gls{BA} model. In contrast to the previous two experiments, we did not inform
+the algorithm of any of the true parameter values.
+}}
+
+We simulated three trees each under a variety of parameter values, and ran the
+\software{netabc} program to estimate posterior distributions for the
+parameters. The parameter values and priors used are listed in
+\cref{tab:abcexpt}. The tree kernel meta-parameters were set to $\lambda = 0.3$
+and $\sigma = 4$. The \gls{SMC} algorithm was run with 1000 particles, five
+sampled datasets per particle, and the $\alpha$ parameter (not to be confused
+with the \gls{BA} preferential attachment parameter, see
+\cref{subsec:adaptsmc}) set to 0.95. The algorithm was stopped when the
+acceptance rate of the \gls{MH} kernel dropped below 1.5\%, the same criterion
+used by \citeauthor{del2012adaptive}. Approximate marginal posterior densities
+for each parameter were calculated using the \software{density} function in
+\software{R} applied to the final weighted population of particles.
+{\color{red}\sout{Credible intervals were obtained for each parameter using the
+\software{HPDinterval} function in the \software{coda}
 package~\autocite{plummer2006coda}.}} {\color{blue}\uline{Posterior means
@@
 
-\Cref{fig:abcptm2} shows {\color{red}\cout{\gls{MAP}}}
+\Cref{fig:abcptm2} shows {\color{red}\sout{\gls{MAP}}}
 {\color{blue}\uline{posterior mean}} point estimates of the BA model parameters
@@
 %    \Sexpr{pp(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"], eq=TRUE)}$),
-{\color{blue}\uline{did not vary across the true values of $m$ or $I$ (both one-way ANOVA).}} 
+{\color{blue}\uline{did not vary across the true values of $m$ or $I$ (both
+one-way ANOVA).}}
+
 Estimates for $I$ were relatively accurate, with point estimate errors of
@@
 
-The $m$ parameter was estimated correctly in only
+The $m$ parameter was estimated correctly in
     \Sexpr{as.integer(m.tbl[1] * 100)} \%
-of simulations {\color{red}\sout{barely better than random guessing}}. The true
-values of the other parameters did not significantly affect the estimates of
-$m$ (both one-way ANOVA). Finally, the total number of nodes $N$ was
-consistently over-estimated by about a factor of two
-    (error \Sexpr{d[,round(median(N_error))]} 
-    [\Sexpr{d[,round(quantile(N_error, 0.25))]} - 
-     \Sexpr{d[,round(quantile(N_error, 0.75))]}] individuals).
-No parameters influenced the accuracy of the $N$ estimates (all one-way ANOVA).
+of simulations {\color{red}\sout{barely better than random guessing}}.
+{\color{blue}\uline{However this was due to the estimated value being $m = 3$
+in the large majority of simulations (\Sexpr{d[,sum(m == 3)]} out of
+\Sexpr{nrow(d)}), with one-third of the simulations having the true value $m =
+3$.}} The true values of the other parameters did not significantly affect the
+estimates of $m$ (both one-way ANOVA).
+
+Finally, the total number of nodes $N$ was consistently over-estimated by about
+a factor of two (error \Sexpr{d[,round(median(N_error))]}
+[\Sexpr{d[,round(quantile(N_error, 0.25))]} - \Sexpr{d[,round(quantile(N_error,
+0.75))]}] individuals). No parameters influenced the accuracy of the $N$
+estimates (all one-way ANOVA).
 
@@
   \caption[
-      Average maximum \textit{a posteriori} point estimates and 95\% highest
-      posterior density (HPD) interval widths for BA model parameter estimates
-      obtained with kernel-assisted ABC.
+      Average posterior mean point estimates and 95\% highest posterior density
+      (HPD) interval widths for BA model parameter estimates obtained with
+      kernel-assisted ABC.
   ]{
-      Average maximum \textit{a posteriori} point estimates and 95\% highest
-      posterior density (HPD) interval widths for BA model parameter estimates
-      obtained with kernel-assisted ABC. Three transmission trees were simulated under
-      each combination of the listed parameter values, and the parameters were
+      Average posterior mean point estimates and 95\% highest posterior density
+      (HPD) interval widths for BA model parameter estimates obtained with
+      kernel-assisted ABC. Three transmission trees were simulated under each
+      combination of the listed parameter values, and the parameters were
       estimated with kernel-assisted ABC without training.
@@
 the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
-other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
-HPD] estimates for each parameter were: 
+other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The posterior mean
+[95\% HPD] estimates for each parameter were: 
 $\alpha$, 
@@
 \section{Application to real world HIV data}
+\label{sec:hiv}
 
@@
 
-  map1 <- dcast.data.table(d1, dataset~parameter, value.var="map")
+  map1 <- dcast.data.table(d1, dataset~parameter, value.var="mean")
   up1 <- dcast.data.table(d1, dataset~parameter, value.var="hpd.max")
@@
 
-  map2 <- dcast.data.table(d2, dataset~parameter, value.var="map")
+  map2 <- dcast.data.table(d2, dataset~parameter, value.var="mean")
   up2 <- dcast.data.table(d2, dataset~parameter, value.var="hpd.max")

### 2016-05-26 doc/thesis/methods_results.tex
Started working on Sally's revisions

@@
+{\color{blue}\uline{
+The chapter is organized as follows. One section is devoted to each of our
+three research aims: the development of a method for inference of contact
+network parameters from transmission trees} (\cref{sec:netabc}), \uline{the
+investigation of the \acrlong{BA} network model using simulated data
+}(\cref{sec:ba}), \uline{ and the application of the method to real-world
+\gls{HIV} epidemics }(\cref{sec:hiv}). \uline{Each of these sections contains a
+description of the methods used and, in the case of the latter two objectives,
+the results of the experiments. The fourth and final section of the chapter
+}(\cref{sec:disc}) \uline{contains a unified discussion of all three
+objectives. }}
+
 \section{\software{Netabc}: a computer program for estimation of contact
 network parameters with kernel-assisted ABC}
+\label{sec:netabc}
 
@@
 
-To test the full kernel-assisted \gls{ABC} algorithm, we simulated three trees each under
-a variety of parameter values, and ran the \software{netabc} program to
-estimate posterior distributions for the parameters. The parameter values and
-priors used are listed in \cref{tab:abcexpt}. The tree kernel meta-parameters
-were set to $\lambda = 0.3$ and $\sigma = 4$. The \gls{SMC} algorithm was run
-with 1000 particles, five sampled datasets per particle, and the $\alpha$
-parameter (not to be confused with the \gls{BA} preferential attachment
-parameter, see \cref{subsec:adaptsmc}) set to 0.95. The algorithm was
-stopped when the acceptance rate of the \gls{MH} kernel dropped below 1.5\%,
-the same criterion used by \citeauthor{del2012adaptive}. Approximate marginal
-posterior densities for each parameter were calculated using the
-\software{density} function in \software{R} applied to the final weighted
-population of particles. {\color{red}\sout{Credible intervals were obtained for
-each parameter using the \software{HPDinterval} function in the \software{coda}
-package~\autocite{plummer2006coda}.}} {\color{blue}\uline{Posterior means and
-credible intervals were obtained for each parameter using the
-\software{wtd.mean} and \software{wtd.quantile} functions in the
-\software{Hmisc} package~\autocite{harrell2016hmisc}. }}
+{\color{blue}\uline{
+Our final simulation-based experiment was designed to test the full
+\gls{ABC}-\gls{SMC} algorithm by jointly estimating the four parameters of the
+\gls{BA} model. In contrast to the previous two experiments, we did not inform
+the algorithm of any of the true parameter values.
+}}
+
+We simulated three trees each under a variety of parameter values, and ran the
+\software{netabc} program to estimate posterior distributions for the
+parameters. The parameter values and priors used are listed in
+\cref{tab:abcexpt}. The tree kernel meta-parameters were set to $\lambda = 0.3$
+and $\sigma = 4$. The \gls{SMC} algorithm was run with 1000 particles, five
+sampled datasets per particle, and the $\alpha$ parameter (not to be confused
+with the \gls{BA} preferential attachment parameter, see
+\cref{subsec:adaptsmc}) set to 0.95. The algorithm was stopped when the
+acceptance rate of the \gls{MH} kernel dropped below 1.5\%, the same criterion
+used by \citeauthor{del2012adaptive}. Approximate marginal posterior densities
+for each parameter were calculated using the \software{density} function in
+\software{R} applied to the final weighted population of particles.
+{\color{red}\sout{Credible intervals were obtained for each parameter using the
+\software{HPDinterval} function in the \software{coda}
+package~\autocite{plummer2006coda}.}} {\color{blue}\uline{Posterior means
+obtained for each parameter using the \software{wtd.mean} function in the
+\software{Hmisc} package~\autocite{harrell2016hmisc}. Credible intervals were
+obtained using the \software{hpd} function in the \software{TeachingDemos}
+package~\autocite{snow2013teachingdemos}.}}
 
@@
 
-\Cref{fig:abcptm2} shows \gls{MAP} point estimates of the BA model parameters
-obtained with kernel-assisted ABC on simulated data. The estimates shown correspond only
-to the simulations where the $m$ parameter was set to 2, however the results
-for $m = 3$ and $m = 4$ were similar (\cref{fig:abcptm3,fig:abcptm4}). Average
-boundaries of 95\% HPD intervals are given in \cref{tab:abchpd}.
+\Cref{fig:abcptm2} shows {\color{red}\sout{\gls{MAP}}}
+{\color{blue}\uline{posterior mean}} point estimates of the BA model parameters
+obtained with kernel-assisted ABC on simulated data. The estimates shown
+correspond only to the simulations where the $m$ parameter was set to 2,
+however the results for $m = 3$ and $m = 4$ were similar
+(\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% HPD intervals are
+given in \cref{tab:abchpd}.
 
@@
 %    pp(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"], eq=TRUE)$),
-{\color{blue}\uline{did not vary across the true values of $m$ or $I$ (both one-way ANOVA).}} 
+{\color{blue}\uline{did not vary across the true values of $m$ or $I$ (both
+one-way ANOVA).}}
+
 Estimates for $I$ were relatively accurate, with point estimate errors of
@@
 
-The $m$ parameter was estimated correctly in only
+The $m$ parameter was estimated correctly in
     37 \%
-of simulations, barely better than random guessing. The true values of the
-other parameters did not significantly affect the estimates of $m$ (both
-one-way ANOVA). Finally, the total number of nodes $N$ was consistently
-over-estimated by about a factor of two
-    (error 4153 
-    [3660 - 
-     4489] individuals).
-No parameters influenced the accuracy of the $N$ estimates (all one-way ANOVA).
+of simulations {\color{red}\sout{barely better than random guessing}}.
+{\color{blue}\uline{However this was due to the estimated value being $m = 3$
+in the large majority of simulations (62 out of
+72), with one-third of the simulations having the true value $m =
+3$.}} The true values of the other parameters did not significantly affect the
+estimates of $m$ (both one-way ANOVA).
+
+Finally, the total number of nodes $N$ was consistently over-estimated by about
+a factor of two (error 4153
+[3660 - 4489] individuals). No parameters influenced the accuracy of the $N$
+estimates (all one-way ANOVA).
 
@@
 the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
-other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
-HPD] estimates for each parameter were: 
+other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The posterior mean
+[95\% HPD] estimates for each parameter were: 
 $\alpha$, 
-  1.1 
+  1.09 
   [0.6 -
-   1.16];
+   1.14];
 $I$,
-  1109 
-  [662 -
+  1143 
+  [666 -
    4455];
@@
   3 
-  [1 -
+  [2 -
    5];
 $N$,
-  12717 
-  [4021-
+  12635 
+  [4282-
    14977].
@@
 $\alpha$, 
-  0.1 
-  [0.01 -
+  0.23 
+  [0 -
    0.63];
 $I$,
-  2399 
-  [1423 -
-   3860];
+  2383 
+  [1419 -
+   3811];
 $m$,
-  3 
+  4 
   [2 -
@@
 $N$,
-  8762 
+  13517 
   [2836 -
-   14788].
+   14720].
 The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
@@
 estimates [95\% \gls{HPD}] 
-  1.04 
-  [0.04 - 
-   1.25]
+  0.76 
+  [0.07 - 
+   1.31]
 and
-  0.82 
+  0.61 
   [0.01 -
@@
 data was narrower
-  (0.04 - 
-   1.25).
+  (0.07 - 
+   1.31).
 When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
 low estimated $\alpha$ values
-  (0.3 
+  (0.31 
   [0 - 
-  0.75]
+  0.73]
 and
-  0.3 
-  [0.01 -
-   0.79]). 
+  0.38 
+  [0.02 -
+   0.82]). 
 However, the MAP estimates increased when $m = 1$ was not permitted, although
 the HPD intervals remained roughly the same
-  (0.78 
-  [0.02 - 
+  (0.55 
+  [0.01 - 
   0.93]
 and
-  0.58 
-  [0.08 -
-   0.95]).
+  0.5 
+  [0.07 -
+   0.94]).
 The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
 for both priors on $m$
-  (0.29 for $m \geq 1$;
-   0.23 for $m \geq 2$).
+  (0.5 for $m \geq 1$;
+   0.34 for $m \geq 2$).
 However, the confidence interval was much wider when $m = 1$ was allowed
   ([0.03 -
-    1.6] for $m \geq 1$ vs.
+    1.61] for $m \geq 1$ vs.
    [0 -
@@
 intervals compared to the nonzero prior density region
-  (\citeauthor{cuevas2009hiv}, 497 
-  [287 -
-   2430];
-   \citeauthor{niculescu2015recent}, 307
-  [138 - 
-   2822];
-  \citeauthor{li2015hiv}, 1217 
-  [383 -
-   2897];
-   \citeauthor{wang2015targeting}, 621
-  [182 - 
-   2139]).
+  (\citeauthor{cuevas2009hiv}, 943 
+  [285 -
+   2111];
+   \citeauthor{niculescu2015recent}, 1032
+  [136 - 
+   2847];
+  \citeauthor{li2015hiv}, 1437 
+  [364 -
+   2659];
+   \citeauthor{wang2015targeting}, 872
+  [173 - 
+   2119]).
 The \citeauthor{novitsky2014impact} data was the outlier, with a very high
 estimated $I$, and HPD interval spanning almost the entire prior region
-  (7642 
+  (5586 
   [187 -
-   8836]).
+   8812]).
 The $I$ estimates and HPD intervals were generally robust to the choice of
@@
 upper bound of the HPD interval was different for each dataset
-  (\citeauthor{niculescu2015recent}, 5;
-   \citeauthor{wang2015targeting}, 4;
+  (\citeauthor{niculescu2015recent}, 4;
+   \citeauthor{wang2015targeting}, 3;
    \citeauthor{li2015hiv}, 1;
@@
 data, for which the MAP estimate was lower 
-  (7428)
+  (6049)
 when $m = 1$ was allowed.

### 2016-05-26 doc/thesis/objective.tex
Started working on Sally's revisions

@@
+\label{sec:obj}
 \glsreset{ABC}
@@
 network's degree distribution~\autocite{peng2013vaccination, ma2013importance}.
-Phylodynamic methods have been used to fit many different types of models to
-phylogenetic data~\autocite{pybus2009evolutionary,volz2013viral}, but these
-models generally assume a panmictic population. The primary objective of this
-work is \emph{to develop a method to fit contact network models in a
-phylodynamic framework.}
+Phylodynamic methods, {\color{blue}\uline{which link viruses' evolutionary and
+epidemiological dynamics}}, have been used to fit many different types of models
+to phylogenetic data~\autocite{pybus2009evolutionary,volz2013viral}. However,
+these models generally assume a panmictic population. The primary objective of
+this work is \emph{to develop a method to fit contact network models,
+{\color{blue}\uline{and thereby relax the assumption of homogeneous mixing}},
+in a phylodynamic framework.}
 
@@
 corresponding transmission trees, can also be easily simulated. As mentioned
-above, contact networks can profoundly affect transmission tree shape, and
-those shapes can be compared using a highly informative similarity measure
-called the ``tree kernel''~\autocite{poon2013mapping}. \Gls{ABC} can be
-implemented with \gls{SMC}, which has several advantages over other
+above, contact networks can profoundly affect transmission tree shape. Those
+shapes can be compared using a highly informative similarity measure called the
+``tree kernel''~\autocite{poon2013mapping}; {\color{blue}\uline{similar kernel
+functions have been demonstrated to work well as distance functions in
+\gls{ABC}~\autocite{park2015k2}}}. \Gls{ABC} can be implemented with \gls{SMC},
+which has several advantages over other
 algorithms~\autocite{mckinley2009inference}. A recently-developed adaptive
@@
 stochastic simulation of epidemics on networks, the tree kernel, and adaptive
-\gls{ABC}-\gls{SMC}.
-
-{\color{blue}\uline{
-Our method will expand on the framework developed
-by~\autocite{poon2015phylodynamic}, who combined \gls{ABC} with the tree kernel
-to infer parameters of population genetic models from viral phylogenies using
-\gls{MCMC}. The term ``kernel-assisted \gls{ABC}'' refers to any application of
-\gls{ABC} using a kernel function as a distance metric, including that
-developed by \citeauthor{poon2015phylodynamic}. However, for ease of
-exposition, we will often use the term ``kernel-assisted \gls{ABC}'' to refer
-to our method specifically. Note that this is distinct from the version of
-\gls{ABC} based on kernel Bayes rule introduced by
-\textcite{fukumizu2011kernel}.
-}}
+\gls{ABC}-\gls{SMC}. {\color{blue}\uline{ Our method will expand on the
+framework developed by~\autocite{poon2015phylodynamic}, who combined \gls{ABC}
+with the tree kernel to infer parameters of population genetic models from
+viral phylogenies using \gls{MCMC}. }}
 
@@
 impact on tree shape, and whether they can be accurately recovered using
-kernel-assisted \gls{ABC}.}
+\gls{ABC}.}
 
@@
 any other pathogen, and including sequences sampled from diverse geographic and
-demographic contexts. At the time of this writing, there were $635,400$ HIV
+demographic settings. At the time of this writing, there were $635,400$ HIV
 sequences publicly available in GenBank, annotated with 172 distinct countries
@@
 obvious context in which our method could be applied. Therefore, the third and
-final aim of this work is \emph{to apply kernel-assisted \gls{ABC} to fit the
-\gls{BA} model to existing \gls{HIV} outbreaks}.
+final aim of this work is \emph{to apply \gls{ABC} to fit the \gls{BA} model to
+existing \gls{HIV} outbreaks}.
 
 To summarize, this work has three objectives. First, we will develop a method
-which uses kernel-assisted \gls{ABC} to infer parameters of contact network
-models from observed transmission trees. Second, we will use simulations to
-characterize the parameters of the \gls{BA} network model in terms of their
-effect on tree shape and how accurately they can be recovered with
-kernel-assisted \gls{ABC}. Finally, we will apply the method to fit the
-\gls{BA} model to several real-world \gls{HIV} datasets.
+which uses \gls{ABC} to infer parameters of contact network models from
+observed transmission trees. Second, we will use simulations to characterize
+the parameters of the \gls{BA} network model in terms of their effect on tree
+shape and how accurately they can be recovered with \gls{ABC}. Finally, we will
+apply the method to fit the \gls{BA} model to several real-world \gls{HIV}
+datasets.
+
+{\color{blue}\uline{
+The remainder of this background chapter is organized in four sections. The
+first section introduces phylogenies and transmission trees, which are the
+input data from which our method aims to make statistical inferences. This
+section also introduces phylodynamics, a family of methods that, like ours, aim
+to infer epidemiological parameters from evolutionary data. The second section
+focuses on contact networks and network models, whose parameters we are
+attempting to infer. The relationship between contact networks and transmission
+trees is also discussed. The third and fourth sections introduce \gls{SMC}
+and \gls{ABC} respectively, which are the two algorithmic components of the
+method we will implement. In particular, \gls{ABC} refers to the general
+approach of using simulations to replace likelihood calculations in a Bayesian
+setting, while \gls{SMC} is a particular algorithm which can be used to
+implement \gls{ABC}. }}
 

### 2016-05-26 doc/thesis/preface.tex
Started working on Sally's revisions

@@
 (H07-02559).
+
+{\color{blue}\uline{Source code for the \software{netabc} program is freely
+available at \url{https://github.com/rmcclosk/netabc} under the GPL-3 license.
+Scripts to run all computational experiments, as well as the source code for
+this thesis, are available at \url{https://github.com/rmcclosk/thesis}.}}

### 2016-05-26 doc/thesis/discussion.Rnw
Started working on HIV studies prevalence stuff

--- 2016-05-26 doc/thesis/discussion.tex
@@
 1$ was excluded from the prior, the confidence interval shrank substantially.
-For both priors, the estimated prevalence was extremely high, in fact higher
-than the estimated HIV prevalence in the sampled region. The authors indicated
-that the source of the samples was a town in close proximity to the country's
-capital city, and suggested that there may have been a high degree of migration
-and partner interchange between the two locations. It is possible that the
-contact network underlying the subtree we investigated includes a much larger
-group based in the capital city, which would explain the high estimate of $I$.
-There is no clear explanation for the discrepancy between the two priors for
-the \textcite{li2015hiv} data, as the subset we analyzed formed a phylogenetic
-cluster and therefore was a good candidate for the BA model. However, nearly
-all the posterior density was assigned to $m = 1$ when this value was allowed,
-indicating that the network was more likely to have an acyclic tree structure.
+
+%For both priors, the estimated prevalence was extremely high, in fact higher
+%than the estimated HIV prevalence in the sampled region. The authors indicated
+%that the source of the samples was a town in close proximity to the country's
+%capital city, and suggested that there may have been a high degree of migration
+%and partner interchange between the two locations. It is possible that the
+%contact network underlying the subtree we investigated includes a much larger
+%group based in the capital city, which would explain the high estimate of $I$.
+%There is no clear explanation for the discrepancy between the two priors for
+%the \textcite{li2015hiv} data, as the subset we analyzed formed a phylogenetic
+%cluster and therefore was a good candidate for the BA model. However, nearly
+%all the posterior density was assigned to $m = 1$ when this value was allowed,
+%indicating that the network was more likely to have an acyclic tree structure.
+
+{\color{blue}\uline{
+The estimates of the prevalence \gls{I} were largely robust to the choice of
+prior on \gls{m}} (\cref{fig:abchpd,fig:abchpdm2}).
+\uline{\textcite{niculescu2015recent} reported that 494 new \gls{HIV}
+infections were diagnosed among the studied population (\glspl{IDU} in
+Bucharest) during the study period. Our estimate of the prevalence was higher
+(X), although the 95\% \gls{HPD} interval did contain the value 494. In
+addition, it is not unreasonable to imagine that a substantial fraction of the
+\gls{HIV}-positive individuals in an \gls{IDU} population could be undiagnosed,
+given that these populations are often marginalized.
+
+The individuals sampled by \autocite{wang2015targeting} were recruited by
+following a prospective cohort of 2000 \gls{MSM} in Beijing. Of the 2000, 179
+new infections were identified during the study period. This number is much
+lower than the estimated prevalence of X obtained with \software{netabc}.
+However, although \citeauthor{wang2015targeting} ... TODO
+
+\textcite{cuevas2009hiv} reported that 620 newly infected and 1500
+chronically infected patients had been sampled during the study period, for a
+total of 2120. This is substantially greater than the point estimate of X we
+obtained with \gls{ABC}, even without considering the likely possibility that
+there were more infected than sampled individuals. Again, this error may in
+part be due to model misspecification, which highlights the importance of
+selecting a model appropriate to the dataset at hand.
+
+}}
 
@@
 which need not be homogeneous across the network. Moreover, it is possible to
-use kernel-assisted \gls{ABC} to fit a model which relaxes some or all of these
-assumptions, which may be a fruitful avenue for future investigation. 
-Despite the possible misspecification, our estimates of the power
-law exponent $\gamma$ were within the range of values reported in the
-literature (\cref{tab:gamma}).
+use \gls{ABC} to fit a model which relaxes some or all of these assumptions,
+which may be a fruitful avenue for future investigation.  Despite the possible
+misspecification, our estimates of the power law exponent $\gamma$ were within
+the range of values reported in the literature (\cref{tab:gamma}).

### 2016-05-27 doc/thesis/discussion.tex
More of Sally's revisions

@@
 1$ was excluded from the prior, the confidence interval shrank substantially.
-For both priors, the estimated prevalence was extremely high, in fact higher
-than the estimated HIV prevalence in the sampled region. The authors indicated
-that the source of the samples was a town in close proximity to the country's
-capital city, and suggested that there may have been a high degree of migration
-and partner interchange between the two locations. It is possible that the
-contact network underlying the subtree we investigated includes a much larger
-group based in the capital city, which would explain the high estimate of $I$.
-There is no clear explanation for the discrepancy between the two priors for
-the \textcite{li2015hiv} data, as the subset we analyzed formed a phylogenetic
-cluster and therefore was a good candidate for the BA model. However, nearly
-all the posterior density was assigned to $m = 1$ when this value was allowed,
-indicating that the network was more likely to have an acyclic tree structure.
+
+%For both priors, the estimated prevalence was extremely high, in fact higher
+%than the estimated HIV prevalence in the sampled region. The authors indicated
+%that the source of the samples was a town in close proximity to the country's
+%capital city, and suggested that there may have been a high degree of migration
+%and partner interchange between the two locations. It is possible that the
+%contact network underlying the subtree we investigated includes a much larger
+%group based in the capital city, which would explain the high estimate of $I$.
+%There is no clear explanation for the discrepancy between the two priors for
+%the \textcite{li2015hiv} data, as the subset we analyzed formed a phylogenetic
+%cluster and therefore was a good candidate for the BA model. However, nearly
+%all the posterior density was assigned to $m = 1$ when this value was allowed,
+%indicating that the network was more likely to have an acyclic tree structure.
+
+{\color{blue}\uline{
+The estimates of the prevalence \gls{I} were largely robust to the choice of
+prior on \gls{m}} (\cref{fig:abchpd,fig:abchpdm2}).
+\textcite{niculescu2015recent} \uline{reported that 494 new HIV infections were
+diagnosed among the studied population (\glspl{IDU} in Bucharest) during the
+study period. Our estimate of the prevalence was higher (X), although the 95\%
+\gls{HPD} interval did contain the value 494. In addition, it is not
+unreasonable to imagine that a substantial fraction of the \gls{HIV}-positive
+individuals in an \gls{IDU} population could be undiagnosed, given that these
+populations are often marginalized.}
+
+\uline{The individuals sampled by \autocite{wang2015targeting} were recruited by
+following a prospective cohort of 2000 \gls{MSM} in Beijing. Of the 2000, 179
+new infections were identified during the study period. This number is much
+lower than the estimated prevalence of X obtained with \software{netabc}.
+However, although} \citeauthor{wang2015targeting} TODO
+
+\textcite{cuevas2009hiv} \uline{reported that 620 newly infected and 1500
+chronically infected patients had been sampled during the study period, for a
+total of 2120. This is substantially greater than the point estimate of X we
+obtained with \gls{ABC}, even without considering the likely possibility that
+there were more infected than sampled individuals. Again, this error may in
+part be due to model misspecification, which highlights the importance of
+selecting a model appropriate to the dataset at hand.}}
 

### 2016-05-27 doc/thesis/methods_results.Rnw
More of Sally's revisions

@@
 grid value, and the scores were normalized such that the area under the curve
-was equal to 1. The grid value with the highest median kernel score was taken
-as the point estimate for the test value, and a 95\% credible interval was
-obtained using the \software{hpd} function in the \software{TeachingDemos}
-package~\autocite{snow2013teachingdemos}.
+was equal to 1. {\color{red}\sout{The grid value with the highest median kernel
+score was taken as the point estimate for the test value}}{\color{blue}\uline{
+A point estimate for the test value was obtained by calculating the weighted
+mean of the grid values with the kernel scores as weights}}, and a 95\%
+credible interval was obtained using the \software{hpd} function in the
+\software{TeachingDemos} package~\autocite{snow2013teachingdemos}.
 
@@
 
+{\color{blue}\uline{
+To evaluate the effects of the true parameter values on the accuracy of the
+posterior mean estimates, we analysed each parameter individually using a
+\gls{GLM} (four \glspl{GLM} total). The response variable was the error of the
+point estimate, and the predictor variables were the true values of
+\gls{alpha}, \gls{I}, and \gls{m}. We did not test for differences across true
+values of \gls{N}, because \gls{N} was not varied in these simulations. The
+distribution family and link function for each \gls{GLM} were chosen by
+examination of residual plots and \gls{AIC}. For \gls{alpha}, \gls{I}, and
+\gls{N}, we used the Gaussian distribution with the inverse link function. For
+\gls{m}, we used the Poisson distribution with the log link function. The
+$p$-values of the estimated \glspl{GLM} coefficients were corrected using
+Holm-Bonferroni correction~\autocite{holm1979simple} with $n = 12$ (four
+\glspl{GLM} with three predictors each). }}
+
 %\subsubsection*{Characterization of power-law exponent in Barab\'asi-Albert networks}
@@
 %\end{table}
-
-%To evaluate the effects of the true parameter values on the accuracy of the
-%\gls{MAP} estimates, we analysed each parameter individually using a \gls{GLM}
-%(four \glspl{GLM} total). The response variable was the error of the point
-%estimate, and the predictor variables were the true values of \gls{alpha},
-%\gls{I}, and \gls{m}. We did not test for differences across true values of
-%\gls{N}, because \gls{N} was not varied in these simulations. The distribution
-%family and link function for each \gls{GLM} were chosen by examination of
-%residual plots and \gls{AIC} of the model fits. For \gls{alpha}, \gls{I}, and
-%\gls{N}, we used the Gamma distribution with the inverse link function. For
-%\gls{m}, we used the Poisson distribution with the log link function. To
-%allow for the possibility that the relationships between parameters and error
-%rates were nonlinear, the true parameter values were treated as discrete rather
-%than continuous predictors. When significant associations were found by the
-%\glspl{GLM}, additional post hoc tests were performed based on the observed
-%relationship between the error rate and true parameter value in question. If
-%the relationship appeared to be monotonic, Spearman's correlation was
-%calculated. Otherwise, if there were particular values which appeared to have a
-%higher error rate than the others, the Wilcoxon rank-sum test was used to check
-%for a significant difference between the two groups.
 
@@
       the parameter being tested. Vertical lines are empirical 95\% confidence
-      intervals based on 1000 two-fold cross-validations.
+      intervals based on 1000 two-fold cross-validations. {\color{blue}\uline{
+      The classifiers for \gls{I} were not evaluated with 1000-tip trees,
+      because one of the tested \gls{I} values was 500, and it is not possible
+      to sample a tree of size 1000 from 500 infected individuals.}}
   }
@@
     # CI takes up >75% of grid
-    stopifnot(d["alpha", all(upper - lower >= 0.75*2)])
-    stopifnot(d["I", all(upper - lower >= 0.75*5000)])
-    stopifnot(d["N", all(upper - lower >= 0.75*15000)])
-    stopifnot(d["m", all(upper - lower >= 0.75*6)])
+    stopifnot(d["alpha", all(upper.95 - lower.95 >= 0.75*2)])
+    stopifnot(d["I", all(upper.95 - lower.95 >= 0.75*5000)])
+    stopifnot(d["N", all(upper.95 - lower.95 >= 0.75*15000)])
+    stopifnot(d["m", all(upper.95 - lower.95 >= 0.75*6)])
 
@@
 
+    tests <- d[,list(rsquared=cor.test(true_value, point.est)$estimate^2,
+                     rel.err=mean(abs(true_value - point.est) * 100 / diff(range(true_value))),
+                     abs.err=mean(abs(true_value - point.est))),
+                by=parameter]
+    setkey(tests, parameter)
+
     alpha.test <- d["alpha", wilcox.test(.SD[true_value %in% c(1.0, 1.25),error], 
@@
 
-The accuracy of grid search estimates largely paralleled that of the \gls{kSVR}
-classifiers. \Cref{fig:gridest} shows point estimates and 95\% highest density
-intervals for each of the \gls{BA} parameters, for one replicate experiment
-with 500-tip trees. Plots showing the point estimates for all replicates can be
-found in \cref{fig:gridptalpha,fig:gridptI,fig:gridptm,fig:gridptN}. For all
-parameters except $m$, the error of point estimates was negatively correlated
-with the number of sampled tips in the tree (for
-\gls{alpha}, \gls{I}, and \gls{N} respectively: Spearman's $\rho$ = 
-    \Sexpr{tip.cor["alpha", round(estimate, 2)]},
-    \Sexpr{tip.cor["I", round(estimate, 2)]},
-    \Sexpr{tip.cor["N", round(estimate, 2)]};
-$p$-values
+{\color{red}\sout{The accuracy of grid search estimates largely paralleled that
+of the \gls{kSVR} classifiers.}}\Cref{fig:gridest} shows point estimates and
+95\% highest density intervals for each of the \gls{BA} parameters, for one
+replicate experiment with 500-tip trees. Plots showing the point estimates for
+all replicates can be found in
+\cref{fig:gridptalpha,fig:gridptI,fig:gridptm,fig:gridptN}. For all parameters
+except $m$, the error of point estimates was negatively correlated with the
+number of sampled tips in the tree (for \gls{alpha}, \gls{I}, and \gls{N}
+respectively: Spearman's $\rho$ = \Sexpr{tip.cor["alpha", round(estimate, 2)]},
+\Sexpr{tip.cor["I", round(estimate, 2)]}, \Sexpr{tip.cor["N", round(estimate,
+2)]}; $p$-values
     $\Sexpr{tip.cor["alpha", pp(p.value)]}$,
@@
     $\Sexpr{tip.cor["N", pp(p.value)]}$).
-The highest density intervals obtained for all parameters were extremely wide,
-occupying $>$75\% of the grid in all cases (\cref{fig:gridest}).
-
-The \gls{alpha} parameter was the most accurately estimated, with point
-estimates having an average deviation of 
+The 95\% highest density intervals obtained for all parameters were extremely
+wide, occupying $>$75\% of the grid in all cases (\cref{fig:gridest}).
+
+{\color{blue}\uline{Similar to the results obtained in the previous experiment,
+the \gls{alpha} and \gls{I} parameters were estimated with comparably high
+accuracy, while the other two parameters were less identifiable. Across all
+replicates, $R^2$ values for the correlations between the estimated and true
+values were
+    $\Sexpr{tests["alpha", round(rsquared, 2)]}$,
+    $\Sexpr{tests["I", round(rsquared, 2)]}$,
+    $\Sexpr{tests["m", round(rsquared, 2)]}$, and
+    $\Sexpr{tests["N", round(rsquared, 2)]}$
+for \gls{alpha}, \gls{I}, \gls{m}, and \gls{N} respectively. The mean absolute
+errors of the point estimates were
+    $\Sexpr{tests["alpha", round(abs.err, 2)]}$,
+    $\Sexpr{tests["I", round(abs.err)]}$,
+    $\Sexpr{tests["m", round(abs.err, 2)]}$, and
+    $\Sexpr{tests["N", round(abs.err)]}$,
+representing
+    $\Sexpr{tests["alpha", round(rel.err)]}$\%,
+    $\Sexpr{tests["I", round(rel.err)]}$\%,
+    $\Sexpr{tests["m", round(rel.err)]}$\%, and
+    $\Sexpr{tests["N", round(rel.err)]}$\%
+of the respective grids.}
+
+\uline{Qualitatively, three of the parameters (excluding \gls{N}) exhibited
+weak identifiability within particular sections of the grid. The \glspl{IQR}
+for \gls{alpha} were nearly identical for the values \gls{alpha} $\leq 1$,
+spanning approximately [0-1.3]. Similarly, the \glspl{IQR} for \gls{alpha} $>
+1$ were all approximately [0.7-2]} (\cref{fig:gridest}). \uline{The
+distributions of kernel scores for these groups of values were also
+qualitatively similar} (\cref{fig:gridalpha}), \uline{although the value
+\gls{alpha} = 1.25 exhibited a more pronounced peak in kernel scores around the
+true value.}
+
+\uline{For \gls{I}, the \glspl{IQR} were nearly identical for the values
+\gls{I} $\leq 2000$ (approximately [500,3800]), and for the values \gls{I} $>
+3000$ (approximately [1600-5000]). Kernel score distributions for all \gls{I}
+values exibited a similar rounded shape} (\cref{fig:gridI}), \uline{although
+this was most pronounced for values near the upper and lower extremes of the
+grid.}
+
+\uline{The \glspl{IQR} for \gls{m} were distinct for \gls{m} = 1 and 2, but
+were similar for all values of \gls{m} $> 2$ (approximately [2-6]). The kernel
+score distribution for \gls{m} = 1 was distinct from the others, having a large
+peak at \gls{m} = 1, while the others exhibited a valley at \gls{m} = 1 and
+were mostly flat everywhere else} (\cref{fig:gridm}).
+
+\uline{The \glspl{IQR} for \gls{N} did not form any obvious groups like the
+other parameters' \glspl{IQR}} (\cref{fig:gridest}). \uline{The kernel score
+distribution for \gls{N} = 1000 was distinct, having a peak at the true value}
+(\cref{fig:gridN}). \uline{The distributions for \gls{N} $> 1000$ were all
+relatively similar, having a valley at the low end of the grid and being mostly 
+flat everywhere else.}}
+
+{\color{red}\sout{The \gls{alpha} parameter was the most accurately estimated,
+with point estimates having an average deviation of 
     \Sexpr{d["alpha", round(mean(error), 2)]}
@@
     (Wilcoxon rank-sum test, $p \Sexpr{pp(alpha.test$p.value, eq=TRUE)}$,
-     \cref{fig:gridptalpha}).
-These two values exhibited different qualitative behaviour than the other
+}\cref{fig:gridptalpha}).
+\sout{These two values exhibited different qualitative behaviour than the other
 values in terms of the distribution of kernel scores along the grid
-(\cref{fig:gridalpha}). In particular, there was a pronounced peak in scores
+}(\cref{fig:gridalpha}). \sout{In particular, there was a pronounced peak in scores
 around the true value, in contrast to the other values where the scores were
 flat around the true value. The effect was most obvious for the value
-\gls{alpha} = 1.25.
-
-The average absolute error of the point estimates for \gls{I} was 
-    \Sexpr{d["I", round(mean(error))]} individuals,
-on a grid of 500 to 5000, and these errors differed between true values of
-\gls{I}
-    (one-way \gls{ANOVA}, $p \Sexpr{pp(av["I", p], eq=TRUE)}$).
-The errors for $2000 \leq I \leq 3000$ were higher than those for the other
-values
-    (average errors
-     \Sexpr{d["I", .SD[true_value >= 2000 & true_value <= 3000, round(mean(error))]]}
-     for $2000 \leq I \leq 3000$ vs.
-     \Sexpr{d["I", .SD[true_value < 2000 | true_value > 3000, round(mean(error))]]}
-     for $I < 2000$ or $I > 3000$),
-and this difference was significant
-    (Wilcoxon rank-sum test, $p \Sexpr{pp(I.test$p.value, eq=TRUE)}$,
-     \cref{fig:gridptI}).
-Kernel score distributions for all test values exhibited a similar rounded
-shape (\cref{fig:gridI}).
-
-The average error for \gls{m} was
-    \Sexpr{d["m", round(mean(error), 2)]} edges per vertex,
-on a grid from 1 to 6. The error varied significantly between the true values
-of \gls{m} 
-    (one-way \gls{ANOVA}, $p \Sexpr{pp(av["m", p], eq=TRUE)}$).
-Errors for the value \gls{m} = 1 were lower than the other values
-    (average errors
-    \Sexpr{d["m", .SD[true_value == 1, round(mean(error), 2)]]}
-    for $m = 1$ vs.
-    \Sexpr{d["m", .SD[true_value > 1, round(mean(error), 2)]]}
-    for $m > 1$),
-and this difference was significant
-    (Wilcoxon rank-sum test, $p \Sexpr{pp(m.test$p.value, eq=TRUE)}$,
-     \cref{fig:gridptm}).
-The value $m = 1$ causes the network to take on a distinct shape relative to
-higher \gls{m} values, namely a tree (\ie there are no cycles, see
-\cref{subsec:treeshape}). The kernel score distribution had a peak at $m = 1$
-when this was the true value, and a valley at $m = 1$ when the true value of
-$m$ was greater that 1 (\cref{fig:gridm}).
-
-The average error for \gls{N} was 
-    \Sexpr{d[parameter == "N", round(mean(error))]} individuals,
-on a grid from 1000 to 15000, and was varied significantly with the true value
-of \gls{N}
-    (one-way \gls{ANOVA}, $p \Sexpr{pp(av["N", p], eq=TRUE)}$).
-The errors were lower for $N \leq 3000$
-    (average errors
-     \Sexpr{d["N", .SD[true_value <= 3000, round(mean(error))]]}
-     for $N \leq 3000$ vs.
-     \Sexpr{d["N", .SD[true_value > 3000, round(mean(error))]]}
-     for $N > 3000$),
-and this difference was significant
-    (Wilcoxon rank-sum test, $p \Sexpr{pp(N.test$p.value, eq=TRUE)}$,
-     \cref{fig:gridptN}).
-The kernel score distribution had a peak at $N = 1000$ when this was the true
-value, and a valley there otherwise (\cref{fig:gridN}). Except for this valley,
-the distributions were flat for $N > 3000$.
+\gls{alpha} = 1.25.}}
+
+%The average absolute error of the point estimates for \gls{I} was 
+%    \Sexpr{d["I", round(mean(error))]} individuals,
+%on a grid of 500 to 5000, and these errors differed between true values of
+%\gls{I}
+%    (one-way \gls{ANOVA}, $p \Sexpr{pp(av["I", p], eq=TRUE)}$).
+%The errors for $2000 \leq I \leq 3000$ were higher than those for the other
+%values
+%    (average errors
+%     \Sexpr{d["I", .SD[true_value >= 2000 & true_value <= 3000, round(mean(error))]]}
+%     for $2000 \leq I \leq 3000$ vs.
+%     \Sexpr{d["I", .SD[true_value < 2000 | true_value > 3000, round(mean(error))]]}
+%     for $I < 2000$ or $I > 3000$),
+%and this difference was significant
+%    (Wilcoxon rank-sum test, $p \Sexpr{pp(I.test$p.value, eq=TRUE)}$,
+%     \cref{fig:gridptI}).
+%Kernel score distributions for all test values exhibited a similar rounded
+%shape (\cref{fig:gridI}). 
+%
+%The average error for \gls{m} was
+%    \Sexpr{d["m", round(mean(error), 2)]} edges per vertex,
+%on a grid from 1 to 6. The error varied significantly between the true values
+%of \gls{m} 
+%    (one-way \gls{ANOVA}, $p \Sexpr{pp(av["m", p], eq=TRUE)}$).
+%Errors for the value \gls{m} = 1 were lower than the other values
+%    (average errors
+%    \Sexpr{d["m", .SD[true_value == 1, round(mean(error), 2)]]}
+%    for $m = 1$ vs.
+%    \Sexpr{d["m", .SD[true_value > 1, round(mean(error), 2)]]}
+%    for $m > 1$),
+%and this difference was significant
+%    (Wilcoxon rank-sum test, $p \Sexpr{pp(m.test$p.value, eq=TRUE)}$,
+%     \cref{fig:gridptm}).
+%The value $m = 1$ causes the network to take on a distinct shape relative to
+%higher \gls{m} values, namely a tree (\ie there are no cycles, see
+%\cref{subsec:treeshape}). The kernel score distribution had a peak at $m = 1$
+%when this was the true value, and a valley at $m = 1$ when the true value of
+%$m$ was greater that 1 (\cref{fig:gridm}).
+%
+%The average error for \gls{N} was 
+%    \Sexpr{d[parameter == "N", round(mean(error))]} individuals,
+%on a grid from 1000 to 15000, and was varied significantly with the true value
+%of \gls{N}
+%    (one-way \gls{ANOVA}, $p \Sexpr{pp(av["N", p], eq=TRUE)}$).
+%The errors were lower for $N \leq 3000$
+%    (average errors
+%     \Sexpr{d["N", .SD[true_value <= 3000, round(mean(error))]]}
+%     for $N \leq 3000$ vs.
+%     \Sexpr{d["N", .SD[true_value > 3000, round(mean(error))]]}
+%     for $N > 3000$),
+%and this difference was significant
+%    (Wilcoxon rank-sum test, $p \Sexpr{pp(N.test$p.value, eq=TRUE)}$,
+%     \cref{fig:gridptN}).
+%The kernel score distribution had a peak at $N = 1000$ when this was the true
+%value, and a valley there otherwise (\cref{fig:gridN}). Except for this valley,
+%the distributions were flat for $N > 3000$.
 
@@
   \caption[Grid search estimates of \gls{BA} model parameters.]{Point estimates
-      and 95\% highest density intervals for each \gls{BA} model parameter,
-      obtained using grid search. Networks and transmission trees were
-      simulated over a grid of values for each parameter while holding the
-      others fixed. For a subset of the grid values ($x$-axis), test networks
-      and trees were created and compared to each tree on the grid using the
-      tree kernel. The kernel scores along the grid were normalized to resemble
-      a probability distribution, from which the mode and highest density
-      interval were calculated. Shown values correspond to one replicate
-      experiment, with trees of size 500.
+      (center lines) and 75\% (boxes) and 95\% (whiskers) highest density
+      intervals for each \gls{BA} model parameter, obtained using grid search.
+      Networks and transmission trees were simulated over a grid of values for
+      each parameter while holding the others fixed. For a subset of the grid
+      values ($x$-axis), test networks and trees were created and compared to
+      each tree on the grid using the tree kernel. The kernel scores along the
+      grid were normalized to resemble a probability distribution, from which
+      the mode and highest density interval were calculated. Shown values
+      correspond to one replicate experiment, with trees of size 500.
   } 

### 2016-05-27 doc/thesis/methods_results.tex
More of Sally's revisions

@@
 
-The accuracy of grid search estimates largely paralleled that of the \gls{kSVR}
-classifiers. \Cref{fig:gridest} shows point estimates and 95\% highest density
-intervals for each of the \gls{BA} parameters, for one replicate experiment
-with 500-tip trees. Plots showing the point estimates for all replicates can be
-found in \cref{fig:gridptalpha,fig:gridptI,fig:gridptm,fig:gridptN}. For all
-parameters except $m$, the error of point estimates was negatively correlated
-with the number of sampled tips in the tree (for
-\gls{alpha}, \gls{I}, and \gls{N} respectively: Spearman's $\rho$ = 
-    \ensuremath{-0.22},
-    \ensuremath{-0.51},
-    \ensuremath{-0.16};
-$p$-values
+{\color{red}\sout{The accuracy of grid search estimates largely paralleled that
+of the \gls{kSVR} classifiers.}}\Cref{fig:gridest} shows point estimates and
+95\% highest density intervals for each of the \gls{BA} parameters, for one
+replicate experiment with 500-tip trees. Plots showing the point estimates for
+all replicates can be found in
+\cref{fig:gridptalpha,fig:gridptI,fig:gridptm,fig:gridptN}. For all parameters
+except $m$, the error of point estimates was negatively correlated with the
+number of sampled tips in the tree (for \gls{alpha}, \gls{I}, and \gls{N}
+respectively: Spearman's $\rho$ = \ensuremath{-0.22},
+\ensuremath{-0.51}, \ensuremath{-0.16}; $p$-values
     $4\!\times\!10^{-4}$,
@@
     $0.01$).
-The highest density intervals obtained for all parameters were extremely wide,
-occupying $>$75\% of the grid in all cases (\cref{fig:gridest}).
-
-The \gls{alpha} parameter was the most accurately estimated, with point
-estimates having an average deviation of 
+The 95\% highest density intervals obtained for all parameters were extremely
+wide, occupying $>$75\% of the grid in all cases (\cref{fig:gridest}).
+
+{\color{blue}\uline{Similar to the results obtained in the previous experiment,
+the \gls{alpha} and \gls{I} parameters were estimated with comparably high
+accuracy, while the other two parameters were less identifiable. Across all
+replicates, $R^2$ values for the correlations between the estimated and true
+values were
+    $0.91$,
+    $0.91$,
+    $0.25$, and
+    $0.54$
+for \gls{alpha}, \gls{I}, \gls{m}, and \gls{N} respectively. The mean absolute
+errors of the point estimates were
+    $0.14$,
+    $310$,
+    $1.31$, and
+    $2419$,
+representing
+    $7$\%,
+    $8$\%,
+    $26$\%, and
+    $17$\%
+of the respective grids.}
+
+\uline{Qualitatively, three of the parameters (excluding \gls{N}) exhibited
+weak identifiability within particular sections of the grid. The \glspl{IQR}
+for \gls{alpha} were nearly identical for the values \gls{alpha} $\leq 1$,
+spanning approximately [0-1.3]. Similarly, the \glspl{IQR} for \gls{alpha} $>
+1$ were all approximately [0.7-2]} (\cref{fig:gridest}). \uline{The
+distributions of kernel scores for these groups of values were also
+qualitatively similar} (\cref{fig:gridalpha}), \uline{although the value
+\gls{alpha} = 1.25 exhibited a more pronounced peak in kernel scores around the
+true value.}
+
+\uline{For \gls{I}, the \glspl{IQR} were nearly identical for the values
+\gls{I} $\leq 2000$ (approximately [500,3800]), and for the values \gls{I} $>
+3000$ (approximately [1600-5000]). Kernel score distributions for all \gls{I}
+values exibited a similar rounded shape} (\cref{fig:gridI}), \uline{although
+this was most pronounced for values near the upper and lower extremes of the
+grid.}
+
+\uline{The \glspl{IQR} for \gls{m} were distinct for \gls{m} = 1 and 2, but
+were similar for all values of \gls{m} $> 2$ (approximately [2-6]). The kernel
+score distribution for \gls{m} = 1 was distinct from the others, having a large
+peak at \gls{m} = 1, while the others exhibited a valley at \gls{m} = 1 and
+were mostly flat everywhere else} (\cref{fig:gridm}).
+
+\uline{The \glspl{IQR} for \gls{N} did not form any obvious groups like the
+other parameters' \glspl{IQR}} (\cref{fig:gridest}). \uline{The kernel score
+distribution for \gls{N} = 1000 was distinct, having a peak at the true value}
+(\cref{fig:gridN}). \uline{The distributions for \gls{N} $> 1000$ were all
+relatively similar, having a valley at the low end of the grid and being mostly 
+flat everywhere else.}}
+
+{\color{red}\sout{The \gls{alpha} parameter was the most accurately estimated,
+with point estimates having an average deviation of 
     0.14
@@
     (Wilcoxon rank-sum test, $p {<}10^{-5}$,
-     \cref{fig:gridptalpha}).
-These two values exhibited different qualitative behaviour than the other
+}\cref{fig:gridptalpha}).
+\sout{These two values exhibited different qualitative behaviour than the other
 values in terms of the distribution of kernel scores along the grid
-(\cref{fig:gridalpha}). In particular, there was a pronounced peak in scores
+}(\cref{fig:gridalpha}). \sout{In particular, there was a pronounced peak in scores
 around the true value, in contrast to the other values where the scores were
 flat around the true value. The effect was most obvious for the value
-\gls{alpha} = 1.25.
-
-The average absolute error of the point estimates for \gls{I} was 
-    310 individuals,
-on a grid of 500 to 5000, and these errors differed between true values of
-\gls{I}
-    (one-way \gls{ANOVA}, $p =0.001$).
-The errors for $2000 \leq I \leq 3000$ were higher than those for the other
-values
-    (average errors
-     430
-     for $2000 \leq I \leq 3000$ vs.
-     250
-     for $I < 2000$ or $I > 3000$),
-and this difference was significant
-    (Wilcoxon rank-sum test, $p =6\!\times\!10^{-4}$,
-     \cref{fig:gridptI}).
-Kernel score distributions for all test values exhibited a similar rounded
-shape (\cref{fig:gridI}).
-
-The average error for \gls{m} was
-    1.31 edges per vertex,
-on a grid from 1 to 6. The error varied significantly between the true values
-of \gls{m} 
-    (one-way \gls{ANOVA}, $p {<}10^{-5}$).
-Errors for the value \gls{m} = 1 were lower than the other values
-    (average errors
-    0.1
-    for $m = 1$ vs.
-    1.55
-    for $m > 1$),
-and this difference was significant
-    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
-     \cref{fig:gridptm}).
-The value $m = 1$ causes the network to take on a distinct shape relative to
-higher \gls{m} values, namely a tree (\ie there are no cycles, see
-\cref{subsec:treeshape}). The kernel score distribution had a peak at $m = 1$
-when this was the true value, and a valley at $m = 1$ when the true value of
-$m$ was greater that 1 (\cref{fig:gridm}).
-
-The average error for \gls{N} was 
-    2419 individuals,
-on a grid from 1000 to 15000, and was varied significantly with the true value
-of \gls{N}
-    (one-way \gls{ANOVA}, $p {<}10^{-5}$).
-The errors were lower for $N \leq 3000$
-    (average errors
-     740
-     for $N \leq 3000$ vs.
-     2979
-     for $N > 3000$),
-and this difference was significant
-    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
-     \cref{fig:gridptN}).
-The kernel score distribution had a peak at $N = 1000$ when this was the true
-value, and a valley there otherwise (\cref{fig:gridN}). Except for this valley,
-the distributions were flat for $N > 3000$.
+\gls{alpha} = 1.25.}}
+
+%The average absolute error of the point estimates for \gls{I} was 
+%    d["I", round(mean(error))] individuals,
+%on a grid of 500 to 5000, and these errors differed between true values of
+%\gls{I}
+%    (one-way \gls{ANOVA}, $p pp(av["I", p], eq=TRUE)$).
+%The errors for $2000 \leq I \leq 3000$ were higher than those for the other
+%values
+%    (average errors
+%     d["I", .SD[true_value >= 2000 & true_value <= 3000, round(mean(error))]]
+%     for $2000 \leq I \leq 3000$ vs.
+%     d["I", .SD[true_value < 2000 | true_value > 3000, round(mean(error))]]
+%     for $I < 2000$ or $I > 3000$),
+%and this difference was significant
+%    (Wilcoxon rank-sum test, $p pp(I.test$p.value, eq=TRUE)$,
+%     \cref{fig:gridptI}).
+%Kernel score distributions for all test values exhibited a similar rounded
+%shape (\cref{fig:gridI}). 
+%
+%The average error for \gls{m} was
+%    d["m", round(mean(error), 2)] edges per vertex,
+%on a grid from 1 to 6. The error varied significantly between the true values
+%of \gls{m} 
+%    (one-way \gls{ANOVA}, $p pp(av["m", p], eq=TRUE)$).
+%Errors for the value \gls{m} = 1 were lower than the other values
+%    (average errors
+%    d["m", .SD[true_value == 1, round(mean(error), 2)]]
+%    for $m = 1$ vs.
+%    d["m", .SD[true_value > 1, round(mean(error), 2)]]
+%    for $m > 1$),
+%and this difference was significant
+%    (Wilcoxon rank-sum test, $p pp(m.test$p.value, eq=TRUE)$,
+%     \cref{fig:gridptm}).
+%The value $m = 1$ causes the network to take on a distinct shape relative to
+%higher \gls{m} values, namely a tree (\ie there are no cycles, see
+%\cref{subsec:treeshape}). The kernel score distribution had a peak at $m = 1$
+%when this was the true value, and a valley at $m = 1$ when the true value of
+%$m$ was greater that 1 (\cref{fig:gridm}).
+%
+%The average error for \gls{N} was 
+%    d[parameter == "N", round(mean(error))] individuals,
+%on a grid from 1000 to 15000, and was varied significantly with the true value
+%of \gls{N}
+%    (one-way \gls{ANOVA}, $p pp(av["N", p], eq=TRUE)$).
+%The errors were lower for $N \leq 3000$
+%    (average errors
+%     d["N", .SD[true_value <= 3000, round(mean(error))]]
+%     for $N \leq 3000$ vs.
+%     d["N", .SD[true_value > 3000, round(mean(error))]]
+%     for $N > 3000$),
+%and this difference was significant
+%    (Wilcoxon rank-sum test, $p pp(N.test$p.value, eq=TRUE)$,
+%     \cref{fig:gridptN}).
+%The kernel score distribution had a peak at $N = 1000$ when this was the true
+%value, and a valley there otherwise (\cref{fig:gridN}). Except for this valley,
+%the distributions were flat for $N > 3000$.
 
@@
 $\alpha$, 
-  1.09 
+  1.1 
   [0.6 -
-   1.14];
+   1.15];
 $I$,
-  1143 
-  [666 -
+  1099 
+  [662 -
    4455];
@@
   3 
-  [2 -
+  [1 -
    5];
 $N$,
-  12635 
-  [4282-
-   14977].
+  12791 
+  [3861-
+   14967].
 The approximate posterior distributions for this simulation are shown in
@@
 $\alpha$, 
-  0.23 
+  0.24 
   [0 -
-   0.63];
+   0.62];
 $I$,
-  2383 
+  2384 
   [1419 -
@@
 $m$,
-  4 
+  3 
   [2 -
@@
 $N$,
-  13517 
-  [2836 -
-   14720].
+  11565 
+  [2741 -
+   14780].
 The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both

### 2016-05-27 doc/thesis/global.R

@@ -27,13 +27,11 @@ d <- d[,I_error := abs(true_I - I)]
 d <- d[,m_error := abs(true_m - m)]
 
 make.glm <- function (param, family, link) {
-    frm <- paste0(param, "_error~factor(true_alpha)+factor(true_I)+factor(true_m)")
+    frm <- paste0(param, "_error~true_alpha+true_I+true_m")
     m <- glm(as.formula(frm), family(link=link), d)
     df <- as.data.frame(coef(summary(m)))
-    df$Parameter <- sub(").*", "", sub("factor(true_", "", rownames(df), fixed=TRUE))
+    df$Parameter <- sub(").*", "", sub("true_", "", rownames(df), fixed=TRUE))
     df$Parameter[1] <- "(Intercept)"
-    df$Value <- sub(".*)", "", rownames(df))
-    df$Value[1] <- "-"
     setDT(df)
     setnames(df, "Std. Error", "Standard error")
     if ("Pr(>|t|)" %in% colnames(df)) {
@@ -43,12 +41,13 @@ make.glm <- function (param, family, link) {
         df$`z value` <- NULL
         setnames(df, "Pr(>|z|)", "p")
     }
-    setcolorder(df, c("Parameter", "Value", "Estimate", "Standard error", "p"))
+    setcolorder(df, c("Parameter", "Estimate", "Standard error", "p"))
+    df[,p := p.adjust(p, method="holm", n=12)]
     df
 }
  
 # make GLMs
-alpha.glm <- make.glm("alpha", Gamma, "inverse")
-I.glm <- make.glm("I", Gamma, "inverse")
-N.glm <- make.glm("N", Gamma, "inverse")
+alpha.glm <- make.glm("alpha", gaussian, "inverse")
+I.glm <- make.glm("I", gaussian, "inverse")
+N.glm <- make.glm("N", gaussian, "inverse")
 m.glm <- make.glm("m", poisson, "log")

### 2016-05-27 doc/thesis/discussion.Rnw
Added some MSM/China stuff

@@
 prior on \gls{m}} (\cref{fig:abchpd,fig:abchpdm2}).
-\uline{\textcite{niculescu2015recent} reported that 494 new \gls{HIV}
-infections were diagnosed among the studied population (\glspl{IDU} in
-Bucharest) during the study period. Our estimate of the prevalence was higher
-(X), although the 95\% \gls{HPD} interval did contain the value 494. In
-addition, it is not unreasonable to imagine that a substantial fraction of the
-\gls{HIV}-positive individuals in an \gls{IDU} population could be undiagnosed,
-given that these populations are often marginalized.
+\textcite{niculescu2015recent} \uline{reported that 494 new HIV infections were
+diagnosed among the studied population (\glspl{IDU} in Bucharest) during the
+study period. Our estimate of the prevalence was higher (X), although the 95\%
+\gls{HPD} interval did contain the value 494. In addition, it is not
+unreasonable that a substantial fraction of the \gls{HIV}-positive individuals
+in an \gls{IDU} population could be undiagnosed, given that these populations
+are often marginalized.}
 
-The individuals sampled by \autocite{wang2015targeting} were recruited by
-following a prospective cohort of 2000 \gls{MSM} in Beijing. Of the 2000, 179
-new infections were identified during the study period. This number is much
-lower than the estimated prevalence of X obtained with \software{netabc}.
-However, although \citeauthor{wang2015targeting} ... TODO
+\uline{The individuals sampled by \autocite{wang2015targeting} were recruited by
+following a prospective cohort of 2000 \gls{MSM} in Beijing, China. Of the
+2000, 179 new infections were identified during the study period. This number
+is much lower than the estimated prevalence of X obtained with
+\software{netabc}, however the authors did not claim to have sampled the entire
+sexual network. In a nationwide survey, \textcite{wu2013hiv} estimated that
+there were 24,198 \gls{MSM} living in Beijing, of whom 5.7\%, or 1379, were
+\gls{HIV}-infected, which is within the 95\% confidence interval of our
+estimate of \gls{I}.}
 
-\textcite{cuevas2009hiv} reported that 620 newly infected and 1500
+\uline{\textcite{li2015hiv} studied an \gls{MSM} population in Shanghai, China.
+Despite the apparent similarity to the context of the
+\textcite{wang2015targeting} study, the estimated \gls{HIV} prevalence of the
+\citeauthor{li2015hiv} data was higher (X). The aforementioned
+survey~\autocite{wu2013hiv} estimated that there were 14,511 \gls{MSM} living
+in Shanghai, with an \gls{HIV} prevalence of 6.8\% or approximately 987
+individuals. This was within the 95\% confidence interval of the estimate
+obtained with \software{netabc}. It is worth noting that \citeauthor{li2015hiv}
+claimed that there were 80,000 \gls{MSM} living in Shanghai, which is a
+significantly different estimate than that obtained by \textcite{wu2013hiv}.}
+
+\textcite{cuevas2009hiv} \uline{reported that 620 newly infected and 1500
 chronically infected patients had been sampled during the study period, for a
@@
 part be due to model misspecification, which highlights the importance of
-selecting a model appropriate to the dataset at hand.
-
-}}
+selecting a model appropriate to the dataset at hand.}
+}
 

### 2016-05-28 doc/thesis/abstract.tex
More of Sally's changes

--- 2016-05-02 doc/thesis/abstract.tex
@@
 the rate of epidemic spread. Contact network models capture this structure by
-explicitly representing each contact which could possibly lead to a
-transmission. We developed a method based on kernel approximate Bayesian
-computation (kernel-ABC) for estimating structural parameters of the contact
-network underlying an observed viral phylogeny. The method combines adaptive
-sequential Monte Carlo for ABC, Gillespie simulation for propagating epidemics
-though networks, and a kernel-based tree similarity score. We used the method
-to fit the Barab\'{a}si-Albert network model to simulated transmission trees,
-and also applied it to viral phylogenies estimated from five published HIV
-sequence datasets. On simulated data, we found that the preferential attachment
-power and the number of infected nodes in the network can often be accurately
-estimated. On the other hand, the mean degree of the network, as well as the
-total number of nodes, were not estimable with kernel-ABC. We observed
+explicitly representing each contact that could possibly lead to a
+transmission. We developed a method based on approximate Bayesian computation
+(ABC) for estimating structural parameters of the contact network underlying an
+observed viral phylogeny. The method combines adaptive sequential Monte Carlo
+for ABC, Gillespie simulation for propagating epidemics though networks, and a
+kernel-based tree similarity score. We used the method to fit the
+Barab\'{a}si-Albert network model to simulated transmission trees and applied
+it to viral phylogenies estimated from six real-world HIV sequence datasets. On
+simulated data, we found that the preferential attachment power and the number
+of infected nodes in the network can often be accurately estimated. On the
+other hand, the mean degree of the network, as well as the total number of
+nodes, appeared to be weakly or non-identifiable with ABC. We observed
 substantial heterogeneity in the parameter estimates on real datasets, with
-point estimates for the preferential attachment power ranging from 0.06 to 1.05.
-These results underscore the importance of considering contact structures when
-performing phylodynamic inference. Our method offers the potential to
+point estimates for the preferential attachment power ranging from 0.06 to
+1.05. These results underscore the importance of considering contact structures
+when performing phylodynamic inference. Our method offers the potential to
 quantitatively investigate the contact network structure underlying viral

### 2016-05-28 doc/thesis/appendix.Rnw
More of Sally's changes

@@
 
-%\chapter{GLM results}
-%
-%<<glm_tables, echo=FALSE, warning=FALSE, error=FALSE, results="asis">>=
-%    format.glm <- function (df) {
-%        df[Parameter == "alpha", Parameter := "$\\alpha$"]
-%        df[,p := paste0("$", sapply(p, pp, eq=FALSE), "$")]
-%        setnames(df, "p", "$p$-value")
-%        df
-%    }
-%
-%    alpha.glm <- format.glm(alpha.glm)
-%    I.glm <- format.glm(I.glm)
-%    m.glm <- format.glm(m.glm)
-%    N.glm <- format.glm(N.glm)
-%
-%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-%                        "$\\alpha$ to true values of BA parameters.")
-%    capt <- paste(capt.short, "GLM was fitted with a Gamma distribution and",
-%                  "inverse link function. Coefficients are interpretable as",
-%                  "additive effects on the inverse of the mean error.")
-%    print(xtable(alpha.glm, label="tab:glmalpha", caption=c(capt, capt.short)),
-%          include.rownames=FALSE, sanitize.text.function=identity,
-%          sanitize.colnames.function=identity)
-%
-%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-%                        "$I$ to true values of BA parameters.")
-%    capt <- paste(capt.short, "GLM was fitted with a Gamma distribution and",
-%                  "inverse link function. Coefficients are interpretable as",
-%                  "additive effects on the inverse of the mean error.")
-%    print(xtable(I.glm, label="tab:glmI", caption=c(capt, capt.short)),
-%          include.rownames=FALSE, sanitize.text.function=identity,
-%          sanitize.colnames.function=identity)
-%
-%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-%                        "$m$ to true values of BA parameters.")
-%    capt <- paste(capt.short, "GLM was fitted with a Poisson distribution and",
-%                  "identity link function. Coefficients are interpretable as",
-%                  "additive effects on the exponential of the mean error.")
-%    print(xtable(m.glm, label="tab:glmm", caption=c(capt, capt.short)),
-%          include.rownames=FALSE, sanitize.text.function=identity,
-%          sanitize.colnames.function=identity)
-%
-%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-%                        "$N$ to true values of BA parameters.")
-%    capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
-%                  "identity link function. Coefficients are interpretable as",
-%                  "additive effects on the inverse of the mean error.")
-%    print(xtable(N.glm, label="tab:glmN", caption=c(capt, capt.short)),
-%          include.rownames=FALSE, sanitize.text.function=identity,
-%          sanitize.colnames.function=identity)
-%@
+\chapter{GLM results}
+
+<<glm_tables, echo=FALSE, warning=FALSE, error=FALSE, results="asis">>=
+    format.glm <- function (df) {
+        df[Parameter == "alpha", Parameter := "$\\alpha$"]
+        df[,p := paste0("$", sapply(p, pp, eq=FALSE), "$")]
+        df[,Estimate := paste0("$", sapply(Estimate, pp, eq=FALSE, thres=NA), "$")]
+        df[,`Standard error` := paste0("$", sapply(`Standard error`, pp, eq=FALSE, thres=NA), "$")]
+        setnames(df, "p", "$p$-value")
+        df
+    }
+
+    alpha.glm <- format.glm(alpha.glm)
+    I.glm <- format.glm(I.glm)
+    #m.glm <- format.glm(m.glm)
+    #N.glm <- format.glm(N.glm)
+
+    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+                        "$\\alpha$ to true values of BA parameters.")
+    capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
+                  "inverse link function. Coefficients are interpretable as",
+                  "additive effects on the inverse of the mean error.")
+    print(xtable(alpha.glm, label="tab:glmalpha", caption=c(capt, capt.short)),
+          include.rownames=FALSE, sanitize.text.function=identity,
+          sanitize.colnames.function=identity)
+
+    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+                        "$I$ to true values of BA parameters.")
+    capt <- paste(capt.short, "GLM was fitted with a gaussian distribution and",
+                  "inverse link function. Coefficients are interpretable as",
+                  "additive effects on the inverse of the mean error.")
+    print(xtable(I.glm, label="tab:glmI", caption=c(capt, capt.short)),
+          include.rownames=FALSE, sanitize.text.function=identity,
+          sanitize.colnames.function=identity)
+
+    #capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+    #                    "$m$ to true values of BA parameters.")
+    #capt <- paste(capt.short, "GLM was fitted with a Poisson distribution and",
+    #              "log link function. Coefficients are interpretable as",
+    #              "additive effects on the exponential of the mean error.")
+    #print(xtable(m.glm, label="tab:glmm", caption=c(capt, capt.short)),
+    #      include.rownames=FALSE, sanitize.text.function=identity,
+    #      sanitize.colnames.function=identity)
+
+    #capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+    #                    "$N$ to true values of BA parameters.")
+    #capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
+    #              "inverse link function. Coefficients are interpretable as",
+    #              "additive effects on the inverse of the mean error.")
+    #print(xtable(N.glm, label="tab:glmN", caption=c(capt, capt.short)),
+    #      include.rownames=FALSE, sanitize.text.function=identity,
+    #      sanitize.colnames.function=identity)
+@

### 2016-05-28 doc/thesis/appendix.tex
More of Sally's changes

@@
 
-%\chapter{GLM results}
-%
-%<<glm_tables, echo=FALSE, warning=FALSE, error=FALSE, results="asis">>=
-%    format.glm <- function (df) {
-%        df[Parameter == "alpha", Parameter := "$\\alpha$"]
-%        df[,p := paste0("$", sapply(p, pp, eq=FALSE), "$")]
-%        setnames(df, "p", "$p$-value")
-%        df
-%    }
-%
-%    alpha.glm <- format.glm(alpha.glm)
-%    I.glm <- format.glm(I.glm)
-%    m.glm <- format.glm(m.glm)
-%    N.glm <- format.glm(N.glm)
-%
-%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-%                        "$\\alpha$ to true values of BA parameters.")
-%    capt <- paste(capt.short, "GLM was fitted with a Gamma distribution and",
-%                  "inverse link function. Coefficients are interpretable as",
-%                  "additive effects on the inverse of the mean error.")
-%    print(xtable(alpha.glm, label="tab:glmalpha", caption=c(capt, capt.short)),
-%          include.rownames=FALSE, sanitize.text.function=identity,
-%          sanitize.colnames.function=identity)
-%
-%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-%                        "$I$ to true values of BA parameters.")
-%    capt <- paste(capt.short, "GLM was fitted with a Gamma distribution and",
-%                  "inverse link function. Coefficients are interpretable as",
-%                  "additive effects on the inverse of the mean error.")
-%    print(xtable(I.glm, label="tab:glmI", caption=c(capt, capt.short)),
-%          include.rownames=FALSE, sanitize.text.function=identity,
-%          sanitize.colnames.function=identity)
-%
-%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-%                        "$m$ to true values of BA parameters.")
-%    capt <- paste(capt.short, "GLM was fitted with a Poisson distribution and",
-%                  "identity link function. Coefficients are interpretable as",
-%                  "additive effects on the exponential of the mean error.")
-%    print(xtable(m.glm, label="tab:glmm", caption=c(capt, capt.short)),
-%          include.rownames=FALSE, sanitize.text.function=identity,
-%          sanitize.colnames.function=identity)
-%
-%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-%                        "$N$ to true values of BA parameters.")
-%    capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
-%                  "identity link function. Coefficients are interpretable as",
-%                  "additive effects on the inverse of the mean error.")
-%    print(xtable(N.glm, label="tab:glmN", caption=c(capt, capt.short)),
-%          include.rownames=FALSE, sanitize.text.function=identity,
-%          sanitize.colnames.function=identity)
-%@
+\chapter{GLM results}
+
+% latex table generated in R 3.2.3 by xtable 1.8-2 package
+% Fri May 27 14:04:29 2016
+\begin{table}[ht]
+\centering
+\begin{tabular}{llll}
+  \hline
+Parameter & Estimate & Standard error & $p$-value \\ 
+  \hline
+(Intercept) & $2$ & $0.6$ & $0.01$ \\ 
+  $\alpha$ & $10$ & $2$ & ${<}10^{-5}$ \\ 
+  I & $-3\!\times\!10^{-4}$ & $2\!\times\!10^{-4}$ & $0.7$ \\ 
+  m & $0.5$ & $0.2$ & $0.01$ \\ 
+   \hline
+\end{tabular}
+\caption[Parameters of a fitted GLM relating error in estimated $\alpha$ to true values of BA parameters.]{Parameters of a fitted GLM relating error in estimated $\alpha$ to true values of BA parameters.
+GLM was fitted with a Gaussian distribution and inverse link function.
+Coefficients are interpretable as additive effects on the inverse of the mean error.}
+\label{tab:glmalpha}
+\end{table}
+% latex table generated in R 3.2.3 by xtable 1.8-2 package
+% Fri May 27 14:04:29 2016
+\begin{table}[ht]
+\centering
+\begin{tabular}{llll}
+  \hline
+Parameter & Estimate & Standard error & $p$-value \\ 
+  \hline
+(Intercept) & $0.004$ & $5\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
+  $\alpha$ & $-0.001$ & $2\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
+  I & $-4\!\times\!10^{-7}$ & $2\!\times\!10^{-7}$ & $0.05$ \\ 
+  m & $-7\!\times\!10^{-5}$ & $8\!\times\!10^{-5}$ & $1$ \\ 
+   \hline
+\end{tabular}
+\caption[Parameters of a fitted GLM relating error in estimated $I$ to true values of BA parameters.]{Parameters of a fitted GLM relating error in estimated $I$ to true values of BA parameters.
+GLM was fitted with a gaussian distribution and inverse link function.
+Coefficients are interpretable as additive effects on the inverse of the mean error.}
+\label{tab:glmI}
+\end{table}

### 2016-05-28 doc/thesis/discussion.Rnw
More of Sally's changes

@@
 
-\uline{The individuals sampled by \autocite{wang2015targeting} were recruited by
-following a prospective cohort of 2000 \gls{MSM} in Beijing, China. Of the
-2000, 179 new infections were identified during the study period. This number
-is much lower than the estimated prevalence of X obtained with
+\uline{The individuals sampled by} \textcite{wang2015targeting} \uline{were
+recruited by following a prospective cohort of 2000 \gls{MSM} in Beijing,
+China. Of the 2000, 179 new infections were identified during the study period.
+This number is much lower than the estimated prevalence of X obtained with
 \software{netabc}, however the authors did not claim to have sampled the entire
-sexual network. In a nationwide survey, \textcite{wu2013hiv} estimated that
-there were 24,198 \gls{MSM} living in Beijing, of whom 5.7\%, or 1379, were
-\gls{HIV}-infected, which is within the 95\% confidence interval of our
+sexual network. In a nationwide survey,} \textcite{wu2013hiv} \uline{estimated
+that there were 24,198 \gls{MSM} living in Beijing, of whom 5.7\%, or 1379,
+were \gls{HIV}-infected, which is within the 95\% confidence interval of our
 estimate of \gls{I}.}
 
-\uline{\textcite{li2015hiv} studied an \gls{MSM} population in Shanghai, China.
+\textcite{li2015hiv} \uline{studied an \gls{MSM} population in Shanghai, China.
 Despite the apparent similarity to the context of the
-\textcite{wang2015targeting} study, the estimated \gls{HIV} prevalence of the
-\citeauthor{li2015hiv} data was higher (X). The aforementioned
+}\textcite{wang2015targeting} \uline{study, the estimated \gls{HIV} prevalence
+of the } \citeauthor{li2015hiv} \uline{data was higher (X). The aforementioned
 survey~\autocite{wu2013hiv} estimated that there were 14,511 \gls{MSM} living
@@
 individuals. This was within the 95\% confidence interval of the estimate
-obtained with \software{netabc}. It is worth noting that \citeauthor{li2015hiv}
-claimed that there were 80,000 \gls{MSM} living in Shanghai, which is a
-significantly different estimate than that obtained by \textcite{wu2013hiv}.}
+obtained with \software{netabc}. It is worth noting that} \citeauthor{li2015hiv}
+\uline{claimed that there were 80,000 \gls{MSM} living in Shanghai, which is a
+significantly different estimate than that obtained by} \textcite{wu2013hiv}.
 
@@
 part be due to model misspecification, which highlights the importance of
-selecting a model appropriate to the dataset at hand.}
-}
+selecting a model appropriate to the dataset at hand.} }
 

### 2016-05-28 doc/thesis/intro.tex
More of Sally's changes

@@
 In evolutionary biology, a \defn{phylogeny}, or \defn{phylogenetic tree}, is a
-graphical representation of the the evolutionary relationships among a group of
+graphical representation of the evolutionary relationships among a group of
 organisms or species (generally, \defn{taxa})~\autocite{haeckel1866generelle}.
@@
       that cats and dogs are the most closely related pair of species, whereas
-      fish is most distantly related to any other node in the tree.}
+      fish is most distantly related to any other taxon in the tree.}
   \label{fig:speciestree}
@@
 detailed interviews and contact tracing of a sufficient number of infected
-individuals is usually prohibitive, and may be additionally be confounded by
+individuals is usually prohibitive, and may additionally be confounded by
 misreporting and other challenges~\autocite{eames2015six}. However, it turns
@@
 $c_n^{j}$ be the $j$th child of node $n$, and $l_n$ be the vector of branch
-lengths connecting node $n$ to its descendants. The \defn{production rule} of
-$n$ is its total number of children, and its number of leaf children. That is,
-if two nodes have the same number of children and among these, the same number
-of leaves, then they have the same production rule. Let $k_G(x, y)$ be a
-Gaussian radial basis function of the vectors $x$ and $y$,
+lengths connecting node $n$ to its $\nc(n)$ children. {\color{blue}\uline{
+Furthermore, let $\nl(n)$ be the number of children of $n$ which are leaves (we
+always have $\nl(n) \leq \nc(n)$). The \defn{production rule} of $n$ is the
+pair $(\nc(n), \nl(n))$.}} That is, if two nodes have the same number of
+children and among these, the same number of leaves, then they have the same
+production rule. Let $k_G(x, y)$ be a Gaussian radial basis function of the
+vectors $x$ and $y$,
 \[
@@
 to become infected. However, an infection which may be spread through
-blood-to-blood contact through transfusions transfusions would use directed
-edges, since the donor has no chance of transmitting to the recipient. Directed
-edges are also useful when the transmission risk is not equal between the
-hosts, such as with \gls{HIV} transmission among \gls{MSM}, where the receptive
-partner carries a higher risk of infection than the insertive
-partner~\autocite{baggaley2010hiv}. In this case, a contact could be
-represented by two directed edges, one in each direction between the two hosts,
-with the edges annotated by what kind of risk they
-imply~\autocite{wasserman1994social}. An undirected contact network is
+blood-to-blood contact through transfusions would use directed edges, since the
+recipient has no chance of transmitting to the donor. Directed edges are also
+useful when the transmission risk is not equal between the hosts, such as with
+\gls{HIV} transmission among \gls{MSM}, where the receptive partner carries a
+higher risk of infection than the insertive partner~\autocite{baggaley2010hiv}.
+In this case, a contact could be represented by two directed edges, one in each
+direction between the two hosts, with the edges annotated by what kind of risk
+they imply~\autocite{wasserman1994social}. An undirected contact network is
 equivalent to a directed network where each contact is represented by two
@@
 many contacts it has. In directed networks, we may make the distinction between
-\defn{out-degree} and \defn{in-degree}, which count respectively the number
+\defn{in-degree} and \defn{out-degree}, which count respectively the number
 incoming and outgoing edges. The \defn{degree distribution} of a network
@@
 is equally likely. This assumption is clearly unrealistic for the majority of
-human communities, and can lead to errors in predicted epidemic trajectories
+human communities and can lead to errors in predicted epidemic trajectories
 when there is substantial heterogeneity present~\autocite{bansal2007individual,
@@
 about the groups or individuals who are at highest risk of acquiring or
-transmitting infection, and would therefore benefit most from public health
+transmitting infection who would therefore benefit most from public health
 interventions~\autocite{wang2015targeting, little2014using}.
@@
 interventions are likely to have the greatest effect~\autocite{poon2015impact}.
-However, this network may show little to no agreement with a contact data
+However, this network may show little to no agreement with contact data
 obtained through epidemiological methods~\autocite{yirrell1998molecular,
-resik2007limitations, robinson2013dynamics}, and therefore may be a poor proxy
+resik2007limitations, robinson2013dynamics} and therefore may be a poor proxy
 for the contact network. Other studies~\autocite{brown2011transmission} have
 investigated the \defn{transmission network}, which is the subgraph of the
-contact network consisting of infected nodes and the edges which led to their
+contact network consisting of infected nodes and the edges that led to their
 infections~\autocite{welch2011statistical} (\cref{fig:contactnet}, left). It is
@@
 phylogenetic network~\autocite{brown2011transmission}. These studies again
-mostly focusing on clustering, and also on degree distributions.
+mostly focus on clustering and also on degree distributions.
 
@@
 the transmission rate.  \textcite{groendyke2011bayesian} significantly updated
-and extended the methodology of \citeauthor{britton2002bayesian}, and applied
+and extended the methodology of \citeauthor{britton2002bayesian} and applied
 it to a measles outbreak affecting 188 individuals. They were able to obtain a
 much more informative estimate of $p$, although this data set included both
-symptom onset and recovery times for all individuals, and was unusual in that
+symptom onset and recovery times for all individuals and was unusual in that
 the entire contact network was presumed to be infected. \textcite{volz2008sir}
@@
 epidemic threshold~\autocite{pastor2001epidemic}, meaning that diseases with
-arbitrarily low transmissibility can persist at low levels indefinitely. This
-is in contrast with homogeneously mixed populations, in which transmissibility
-below the epidemic threshold would result in exponential decay in the number of
-infected individuals and eventual extinction of the
+arbitrarily low transmissibility {\color{red}\sout{can persist}} 
+{\color{blue}\uline{have a chance, however small, of persisting}} at low levels
+indefinitely. This is in contrast with homogeneously mixed populations, in
+which transmissibility below the epidemic threshold would result in exponential
+decay in the number of infected individuals and eventual extinction of the
 pathogen~\autocite{anderson1992infectious}.
@@
 $m = 1$, the network takes on the distinctive shape of a tree, that is, it does
-not contain any cycles. Cycles are present in the network for all all other $m$
+not contain any cycles. Cycles are present in the network for all other $m$
 values. Examples of \gls{BA} networks with three different values of the
@@
 
-There has been some contention of the idea that contact networks are
+There has been some contention over the idea that contact networks are
 scale-free. \textcite{handcock2004likelihood} fit several stochastic models of
@@
 transmission tree imbalance and several epidemic parameters under four contact
-network models, and found that these relationships varied considerably
+network models and found that these relationships varied considerably
 depending on which model was being considered. The authors also investigated a
-real-world \gls{HIV} phylogeny and found a level of unbalancedeness
-inconsistent with a randomly mixing population. \textcite{welch2011network}
-simulated transmission trees over networks with varying degrees of community
-structure. They found that transmission trees simulated under networks with low
-clustering could not generally be distinguished from those simulated under
-highly clustered networks, and concluded that contact network clusters do not
-affect transmission tree shape. However, more recently,
+real-world \gls{HIV} phylogeny and found a level of imbalance inconsistent with
+a randomly mixing population. \textcite{welch2011network} simulated
+transmission trees over networks with varying degrees of community structure.
+They found that transmission trees simulated under networks with low clustering
+could not generally be distinguished from those simulated under highly
+clustered networks and concluded that contact network clusters do not affect
+transmission tree shape. However, more recently,
 \textcite{villandre2016assessment} investigated the correspondence between
-contact network clusters and transmission tree clusters, and did find a
+contact network clusters and transmission tree clusters and did find a
 moderate correspondence between the two in some cases.
@@
 
-\Gls{SMC} is the name for a family of statistical inference methods which rely
+\Gls{SMC} is the name for a family of statistical inference methods that rely
 on approximating probability distributions of interest with large collections
@@
 will be suitable for our purposes. A more rigorous definition can be found in
-\eg~\autocite{kallenberg2006foundations}. Note that a Markov kernels have
-nothing to do with the kernel functions defined in \cref{subsec:treeshape},
-other than sharing a name (the word ``kernel'' is ubiquitous in mathematics).
+\eg~\autocite{kallenberg2006foundations}. Note that Markov kernels have nothing
+to do with the kernel functions defined in \cref{subsec:treeshape}, other than
+sharing a name (the word ``kernel'' is ubiquitous in mathematics).
 
@@
 sample from $\pi$ by \gls{IS}}} (\cref{alg:sis}).  We begin with $n$
-``particles'' which have been sampled from the importance distribution
+``particles'', which have been sampled from the importance distribution
 $\eta(x_0)$ for $\pi(x_0)$. The particles are updated and reweighted $d$ times,
@@
 where $q(x, x')$ is a proposal function such as a Gaussian distribution
-centred at $x$ (see \cref{subsec:mfit}).
+centred at $x$ {\color{blue}\uline{from which $x'$ is drawn}} (see
+\cref{subsec:mfit}).
 
@@
 objects. First, they propose the existence of \textit{backward} Markov kernels
-$L_{i-1}(x_i, x_{i-1})$. For now, these kernels are arbitrary, and will be
-precisely defined on a problem-specific basis. Second, they define an
+$L_{i-1}(x_i, x_{i-1})$. For now, these kernels are arbitrary; they will later
+be precisely defined on a problem-specific basis. Second, the authors define an
 alternative sequence of target distributions
@@
 distribution with mean zero and variance $\sigma^2$. Note that we are treating
-the $x_i$ as fixed quantities, and therefore have not conditioned the
+the $x_i$ as fixed quantities and therefore have not conditioned the
 probability density on $x$. Also, we have assumed that all the $y_i$ are
@@
 probability density of our observed outcomes $y$ is as high as possible. The
-parameters which optimize the likelihood are known as the \textit{\gls{ML}}
+parameters that optimize the likelihood are known as the \textit{\gls{ML}}
 estimates, denoted $\hat{\theta}$. That is,
@@
 simplest terms, many possible values for $\theta$ are examined, $\L(\theta \mid
-y)$ is calculated for each, and the parameters which produce the highest value
+y)$ is calculated for each, and the parameters that produce the highest value
 are accepted. Many sophisticated numerical optimization methods exist, although
@@
 \[
-  \max\left(1, \frac{f(y \mid \theta_2) \pi(\theta_2) q(\theta_2, \theta_1)}
+  \min\left(1, \frac{f(y \mid \theta_2) \pi(\theta_2) q(\theta_2, \theta_1)}
                     {f(y \mid \theta_1) \pi(\theta_2) q(\theta_1, \theta_2)} \right),
@@
       \State Accept $\theta \gets \theta'$ with probability
-      $\max \left( 1, 
+      $\min \left( 1, 
        \dfrac{f(y \mid \theta') \pi(\theta') q(\theta', \theta)}
@@
 \[
-  \max\left(1, \frac{f(z' \mid \theta') q(\theta', \theta)}
-                    {f(z \mid \theta) q(\theta, \theta')} 
+  \min\left(1, \frac{\pi(\theta') q(\theta', \theta)}
+                    {\pi(\theta) q(\theta, \theta')} 
     \cdot \I_{A_{\varepsilon, y}}(z') \right).
@@
       \State Accept $\theta \gets \theta'$ with probability
-      $\max \left( 1, 
+      $\min \left( 1, 
        \dfrac{\pi(\theta') q(\theta', \theta)}

### 2016-05-28 doc/thesis/methods_results.Rnw
More of Sally's changes

@@
 reproduced Figure 1A of \textcite{leventhal2012inferring} (our
-\cref{fig:leventhal}), which plots the unbalancedness of transmission trees
+\cref{fig:leventhal}), which plots the imbalance of transmission trees
 simulated over four network models at various levels of pathogen
@@
 computationally intensive, we chose to implement our own version in
-\software{C}.
+\software{C} {\color{blue}\uline{as part of \software{netabc}}}.
 
@@
 classifier). First, we considered Sackin's index~\autocite{shao1990tree}, which
-measures the degree of unbalancedeness or asymmetry in a phylogeny} (see
+measures the degree of imbalance or asymmetry in a phylogeny} (see
 \cref{subsec:treeshape}). \uline{Sackin's index is widely used for
@@
 To evaluate the effects of the true parameter values on the accuracy of the
-posterior mean estimates, we analysed each parameter individually using a
-\gls{GLM} (four \glspl{GLM} total). The response variable was the error of the
+posterior mean estimates, we analysed the \gls{alpha} and \gls{I} parameters
+individually using \glspl{GLM}. The response variable was the error of the
 point estimate, and the predictor variables were the true values of
@@
 values of \gls{N}, because \gls{N} was not varied in these simulations. The
-distribution family and link function for each \gls{GLM} were chosen by
-examination of residual plots and \gls{AIC}. For \gls{alpha}, \gls{I}, and
-\gls{N}, we used the Gaussian distribution with the inverse link function. For
-\gls{m}, we used the Poisson distribution with the log link function. The
-$p$-values of the estimated \glspl{GLM} coefficients were corrected using
-Holm-Bonferroni correction~\autocite{holm1979simple} with $n = 12$ (four
-\glspl{GLM} with three predictors each). }}
+distribution family and link function for the \glspl{GLM} were chosen as
+Gaussian and inverse, respectively, by examination of residual plots and
+\gls{AIC}. The $p$-values of the estimated \glspl{GLM} coefficients were
+corrected using Holm-Bonferroni correction~\autocite{holm1979simple} with $n =
+6$ (two \glspl{GLM} with three predictors each). Because there was clearly
+little to no identifiability of the \gls{N} and \gls{m} parameters with
+\gls{ABC} (see Results in next section), we did not construct \glspl{GLM} for
+those parameters. }}
 
@@
 however the results for $m = 3$ and $m = 4$ were similar
-(\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% HPD intervals are
-given in \cref{tab:abchpd}.
-
+(\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% \gls{HPD}
+intervals are given in \cref{tab:abchpd}.
+
+{\color{blue}
+\uline{Across all simulations, the median [\gls{IQR}] absolute errors of the parameter 
+estimates obtained with \software{netabc} were
+    \Sexpr{d[,round(median(alpha_error), 2)]} 
+    [\Sexpr{d[,round(quantile(alpha_error, 0.25), 2)]} - 
+    \Sexpr{d[,round(quantile(alpha_error, 0.75), 2)]}]
+for \gls{alpha},
+    \Sexpr{d[,round(median(I_error))]} 
+    [\Sexpr{d[,round(quantile(I_error, 0.25))]} - 
+    \Sexpr{d[,round(quantile(I_error, 0.75))]}]
+for \gls{I},
+    \Sexpr{d[,round(median(m_error), 2)]} 
+    [\Sexpr{d[,round(quantile(m_error, 0.25), 2)]} - 
+    \Sexpr{d[,round(quantile(m_error, 0.75), 2)]}]
+for \gls{m}, and
+    \Sexpr{d[,round(median(N_error))]} 
+    [\Sexpr{d[,round(quantile(N_error, 0.25))]} - 
+    \Sexpr{d[,round(quantile(N_error, 0.75))]}]
+for \gls{N}. These errors comprised, respectively,
+    \Sexpr{d[,round(median(alpha_error) * 100 / 2)]}\%,
+    \Sexpr{d[,round(median(I_error) * 100 / 4500)]}\%,
+    \Sexpr{d[,round(median(m_error) * 100 / 6)]}\%, and
+    \Sexpr{d[,round(median(N_error) * 100 / 14500)]}\%
+of the regions of nonzero prior density. Average \gls{HPD} interval widths were
+    \Sexpr{d[,round(mean(alpha.upper - alpha.lower), 2)]},
+    \Sexpr{d[,round(mean(I.upper - I.lower))]},
+    \Sexpr{d[,round(mean(m.upper - m.lower), 2)]}, and
+    \Sexpr{d[,format(round(mean(N.upper - N.lower)), digits=5)]},
+representing 
+    \Sexpr{d[,round(mean(alpha.upper - alpha.lower) * 100 / 2)]}\%,
+    \Sexpr{d[,round(mean(I.upper - I.lower) * 100 / 4500)]}\%,
+    \Sexpr{d[,round(mean(m.upper - m.lower) * 100 / 6)]}\%, and
+    \Sexpr{d[,round(mean(N.upper - N.lower) * 100 / 14500)]}\%
+of the nonzero prior density regions. Point estimates of \gls{I} were upwardly
+biased: \gls{I} was overestimated in
+    \Sexpr{d[,sum(I > true_I)]}
+out of 
+    \Sexpr{nrow(d)}
+simulations
+    (\Sexpr{round(d[,sum(I < true_I)] * 100 / nrow(d))}\%).
+The estimates for \gls{m} and \gls{N} were similar across all simulations
+(median [IQR] point estimates
+    \Sexpr{d[,round(median(m), 2)]}
+    [\Sexpr{d[,round(quantile(m, 0.25), 2)]} - 
+    \Sexpr{d[,round(quantile(m, 0.75), 2)]}]
+and 
+    \Sexpr{d[,round(median(N))]}
+    [\Sexpr{d[,round(quantile(N, 0.25))]} - 
+    \Sexpr{d[,round(quantile(N, 0.75))]}])
+regardless of the true values of any of the \gls{BA} parameters.}
+
+\uline{To analyse the effects of the true parameter values on the
+accuracy our estimates of \gls{alpha} and \gls{I}, we fitted one \glspl{GLM}
+for each of these two parameters, with error rate as the dependent variable and
+the true parameter values as independent variables. Since the estimates of
+\gls{m} and \gls{N} were roughly equal across all simulations}
+(\cref{fig:abcptm2,fig:abcptm3,fig:abcptm4}), \uline{\glspl{GLM} were not
+fitted for these parameters. The estimated coefficients are shown in
+}\cref{tab:glmalpha,tab:glmI}. \uline{For the parameters \gls{alpha}, \gls{I},
+and \gls{N}, the \gls{GLM} were fitted using the inverse link function. That
+is, if $p$ is the true value of the parameter and $\hat{p}$ is a random
+variable representing our estimate of the parameter, the \gls{GLM} posits a
+relationship of the form }
+\[
+    \E(|p-\hat{p}|) = (\beta_0 + \beta_\alpha \alpha + \beta_I I + \beta_m m)^{-1},
+\]
+\uline{where the $\beta$'s are coefficients to be fitted. If the true value
+\gls{alpha}, say, is increased by one, the \emph{inverse} of the expected
+absolute error will increase by $\beta_\alpha$. If $\beta_\alpha$ is positive,
+it means that the absolute error decreases as the true value of \gls{alpha}
+increases.}
+
+<<abc_glm, include=FALSE>>=
+    stopifnot(alpha.glm[Parameter == "alpha",p] < 0.05)
+    stopifnot(alpha.glm[Parameter == "I",p] > 0.05)
+    stopifnot(alpha.glm[Parameter == "m",p] < 0.05)
+
+    stopifnot(I.glm[Parameter == "alpha",p] < 0.05)
+    stopifnot(I.glm[Parameter == "I",p] < 0.05)
+    stopifnot(I.glm[Parameter == "m",p] > 0.05)
+@
+
+\uline{The \gls{GLM} analysis indicated that the error in estimates of
+\gls{alpha} decreased with larger true values of \gls{alpha} 
+    ($p \Sexpr{pp(alpha.glm[Parameter == "alpha", p], eq=TRUE)}$)
+and \gls{m}
+    ($p \Sexpr{pp(alpha.glm[Parameter == "m", p], eq=TRUE)}$)
+but was not significatly affected by \gls{I}} (\cref{tab:glmalpha}). 
+\uline{Qualitatively, \gls{alpha} seemed to be only weakly identifiable between
+the values of 0 and 0.5} (\cref{fig:abcptm2}). \uline{The error in the
+estimated prevalence \gls{I} was slightly lower for smaller values of
+\gls{alpha}
+    ($p \Sexpr{pp(I.glm[Parameter == "alpha", p], eq=TRUE)}$)
+and \gls{I}
+    ($p \Sexpr{pp(I.glm[Parameter == "I", p], eq=TRUE)}$),
+but was not significantly affected by the true value of \gls{m}} 
+(\cref{tab:glmI}).}
+
+{\color{red}\sout{
 The accuracy of the parameter estimates obtained with kernel-assisted ABC
@@
 %    \Sexpr{pp(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"], eq=TRUE)}$),
-{\color{blue}\uline{did not vary across the true values of $m$ or $I$ (both
-one-way ANOVA).}}
-
-Estimates for $I$ were relatively accurate, with point estimate errors of
+did not vary across the true values of $m$ or $I$ (both
+one-way ANOVA).}
+
+\sout{Estimates for $I$ were relatively accurate, with point estimate errors of
     \Sexpr{d[,round(median(I_error))]} 
@@
 and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
-did not affect the estimates of $I$ (one-way ANOVA).
-
-The $m$ parameter was estimated correctly in
+did not affect the estimates of $I$ (one-way ANOVA).}
+
+\sout{The $m$ parameter was estimated correctly in
     \Sexpr{as.integer(m.tbl[1] * 100)} \%
-of simulations {\color{red}\sout{barely better than random guessing}}.
-{\color{blue}\uline{However this was due to the estimated value being $m = 3$
-in the large majority of simulations (\Sexpr{d[,sum(m == 3)]} out of
-\Sexpr{nrow(d)}), with one-third of the simulations having the true value $m =
-3$.}} The true values of the other parameters did not significantly affect the
-estimates of $m$ (both one-way ANOVA).
-
-Finally, the total number of nodes $N$ was consistently over-estimated by about
+of simulations barely better than random guessing. The true values of the other
+parameters did not significantly affect the
+estimates of $m$ (both one-way ANOVA).}
+
+\sout{Finally, the total number of nodes $N$ was consistently over-estimated by about
 a factor of two (error \Sexpr{d[,round(median(N_error))]}
@@
 0.75))]}] individuals). No parameters influenced the accuracy of the $N$
-estimates (all one-way ANOVA).
+estimates (all one-way ANOVA).}}
 

### 2016-05-28 doc/thesis/objective.tex
More of Sally's changes

@@
 
-Calculating the likelihood of the parameters of a contact network models seems
+Calculating the likelihood of the parameters of a contact network model seems
 likely to be an intractable problem. We have not proven this is the case, but
@@
 likelihood calculation. Consider a contact network model with parameters
-$\theta$, and an estimated transmission tree $T$ with $n$ tips.
+$\theta$ and an estimated transmission tree $T$ with $n$ tips.
 {\color{blue}\uline{The transmission tree is a record of an epidemic's progress

### 2016-05-28 doc/thesis/global.R

@@ -4,11 +4,11 @@ library(netabc)
 library(Hmisc)
 library(xtable)
 
-pp <- function (p, eq=FALSE) {
-    if (p < 1e-5) {
-        "{<}10^{-5}"
+pp <- function (p, eq=FALSE, thres=1e-5) {
+    if (!is.na(thres) & p < thres) {
+        paste0("{<}", pp(thres, thres=0))
     } else {
-        n <- latexSN(round(p, -floor(log10(p))))
+        n <- latexSN(round(p, -floor(log10(abs(p)))))
         n <- sub("^1\\\\!\\\\times\\\\!", "", n)
         if (eq) {
             paste0("=", n)
@@ -42,12 +42,12 @@ make.glm <- function (param, family, link) {
         setnames(df, "Pr(>|z|)", "p")
     }
     setcolorder(df, c("Parameter", "Estimate", "Standard error", "p"))
-    df[,p := p.adjust(p, method="holm", n=12)]
+    df[,p := p.adjust(p, method="holm", n=6)]
     df
 }
  
 # make GLMs
 alpha.glm <- make.glm("alpha", gaussian, "inverse")
 I.glm <- make.glm("I", gaussian, "inverse")
-N.glm <- make.glm("N", gaussian, "inverse")
-m.glm <- make.glm("m", poisson, "log")
+#N.glm <- make.glm("N", gaussian, "inverse")
+#m.glm <- make.glm("m", poisson, "log")

### 2016-05-30 doc/thesis/intro.tex
A bit more writing

@@
 intractable. Hence, rather than computing the posterior distribution
-analytically, we will attempt to approximate it using a \defn{Monte Carlo}
-approach.
-TODO
+analytically, we will approximate it using a \defn{Monte Carlo} approach.
+The fundamental idea behind Monte Carlo methods is succinctly expressed by
+\textcite{liu2001theoretical}:
+\begin{quote}
+    Monte Carlo's view of the world is that any probability distribution $\pi$, 
+    regardless of its complexity, can always be \emph{represented} by a
+    discrete sample from it. By ``represented'', we mean that any computation
+    of expectations using $\pi$ can be replaced to an acceptable degree of
+    accuracy by using the empirical distribution resulting from the discrete
+    sample.
+\end{quote}
+In other words, if we are able to sample enough points from a distribution of
+interest, we will be able to make reasonably accurate statements about the
+distribution itself. For example, the expected value of the distribution can be
+estimated by the sample's population mean. Several algorithms exist for
+obtaining such a sample, some of which will be discussed further in
+\cref{subsec:mfit}. The particular method we apply in this work is \gls{SMC},
+which is also known as the \emph{particle filter}. The reason for this name is
+that the desired number of samples from the target distribution is decided at
+the outset, and that number of ``particles'' \ldots TODO
+
+This is in contrast 
+}
+
+\uline{In the
+context of this work, the distribution of interest is the posterior
+distribution on the parameters of a contact network model, given observed
+phylogenetic data in the form of a transmission tree.}
+
+\uline{A typical Bayesian Monte Carlo approach to this problem would be to
+calculate the product of the likelihood and the prior for each of the parameter
+particles.
+
 }}
@@
 \subsection{Sequential Monte Carlo}
+\label{subsec:smc}
 

### 2016-05-30 doc/thesis/methods_results.Rnw
A bit more writing

@@
                  and infect $v$
+          \State $\tip[v] \gets |T|+1$
+          \Comment{add new tips to tree and tip array}
+          \State $\tip[u] \gets |T|+2$
+          \Comment{corresponding to $u$ and $v$}
           \State add tips with labels $(|T|+1)$ and $(|T|+2)$ to $T$
           \State connect the new nodes to $\tip[v]$ in $T$, with branch lengths $0$
-          \State $\tip[v] \gets |T|-1$
-          \State $\tip[u] \gets |T|$
         \Else
@@
 comprehensive similarity score between two phylogenetic trees, via the
-dot-product of the two trees' feature vectors in the infinite-dimensional space
-of all possible subset trees with branch lengths (see \cref{subsec:treeshape}).
+dot-product of the two trees' feature vectors in the space of all possible
+subset trees with branch lengths (see \cref{subsec:treeshape}). {\color{blue}
+\uline{Because the branch lengths are continuous, there are infinitely many
+possible subset trees; hence, the feature space is infinite-dimensional.}}
 The kernel was implemented using the fast algorithm developed by
@@
 computed by dynamic programming. The complexity of this implementation is
-$O(|T_1||T_2|)$, where $|T|$ counts the number of nodes in the tree $T$.
+$O(|T_1||T_2|)$ for the two trees $T_1$ and $T_2$ being compared.
 
@@
 \]
-which is a number between 0 and 1 minimized when $T_1 = T_2$. This is similar
-to the normalization used by \textcite{collins2002new, poon2013mapping}.
+which is a number between 0 and 1 that is minimized when $T_1 = T_2$. This is
+similar to the normalization used by \textcite{collins2002new,
+poon2013mapping}.
 
@@
 Monte Carlo approximation to the target distribution. This situation is
-referred to as the ``collapse'' of the approximation, and is mitigated by the
-adaptive approach. A single parameter $\alpha$ (not to be confused with the
-\gls{BA} model parameter) controls the decay rate, with $\varepsilon_i$ being
-chosen to satisfy
+referred to as the ``collapse'' of the approximation, {\color{blue}\uline{also
+known as particle degeneracy} (see \cref{subsec:smc})} and is mitigated by the
+adaptive approach. A single parameter {\color{red}\sout{$\alpha$ (not to be
+confused with the \gls{BA} model parameter)}} controls the decay rate.
+{\color{blue}\uline{In the original paper of \textcite{del2012adaptive}, the
+parameter is called $\alpha$, but to avoid confusion with the \gls{BA}
+parameter of the same name, we will refer to it here as $\alpha_{\ESS}$.}}
+The tolerance $\varepsilon_i$ is chosen to satisfy
 \[
-  \ESS(w_i) = \alpha \ESS(w_{i-1}).
+  \ESS(w_i) = \alpha_{\ESS} \ESS(w_{i-1}),
 \]
-Here, $w_i$ is the vector of weights at the $i$th step. Note that, since $w_i$
+where, $w_i$ is the vector of weights at the $i$th step. Note that, since $w_i$
 depends on $\varepsilon_i$, this equation solves for the updated weights and
@@
 to $\varepsilon_i$, the particles' weights are multiplied by the proportion of
-simulated datasets which are still inside the new threshold. The algorithm may
-be stopped when one of two termination conditions is reached. The user may
-specify a final tolerance $\varepsilon$, or a final acceptance rate of the
-\gls{MCMC} kernel. The latter condition stops the algorithm when the particles
-are not moving around very much, implying little change in the estimated
-target.
+simulated datasets that are still inside the new threshold. 
+The user may specify a final tolerance $\varepsilon$, or a final acceptance
+rate of the \gls{MCMC} kernel, and the algorithm will be stopped when either of
+these termination conditions is reached. The latter condition stops the
+algorithm when the particles are not moving around very much, implying little
+change in the estimated target.
 
@@
 the removal rate of each node was set to 0, and all transmission trees' branch
-lengths were scaled by their mean. \textit{igraph} library's implementation of
-the BA model~\autocite{csardi2006igraph} was used to generate the graphs. The
-analyses were run on Westgrid (\url{https://www.westgrid.ca/}) and a local
+lengths were scaled by their mean. The \textit{igraph} library's implementation
+of the BA model~\autocite{csardi2006igraph} was used to generate the graphs.
+The analyses were run on Westgrid (\url{https://www.westgrid.ca/}) and a local
 computer cluster. With the exception of our own programs, all analyses were
@@
 \gls{m}, and \gls{N}. The objective of this experiment was to determine whether
-any of the four parameters might be identifiable from the shape of the
-transmission tree. TODO }
-
-\uline{In addition, a secondary objective of this section was to validate the
-use of the tree kernel as a distance measure for \gls{ABC} in our context. As
-discussed in} \cref{sec:abc}, \uline{the choice of distance function is
-extremely important for the accuracy of the \gls{ABC} approximation to the
-posterior. Therefore, we evaluated two additional tree statistics in the same
-manner as we evaluated the tree kernel (that is, by constructing and testing a
-classifier). First, we considered Sackin's index~\autocite{shao1990tree}, which
-measures the degree of imbalance or asymmetry in a phylogeny} (see
-\cref{subsec:treeshape}). \uline{Sackin's index is widely used for
-characterizing phylogenies~\autocite{frost2013modelling} and has been
-demonstrated to vary between transmission trees simulated under different
-contact network types~\autocite{leventhal2012inferring}. Sackin's index does
-not take branch lengths into account, considering only the tree's topology. The
-other statistic we considered was the
-\gls{nltt}~\autocite{janzen2015approximate}, which compares two trees based on
-normalized distributions of their branching times} (see \cref{sec:treeshape}).
-\uline{In contrast with Sackin's index, the \gls{nltt} does not explicity
-consider the trees' topologies, but it does use their normalized branch
-lengths. While the \gls{nltt} is a newly developed statistic not yet in
-widespread use, the unnormalized \gls{ltt}~\autocite{nee1992tempo} was the
-basis of seminal early work extracting epidemiological information from
-phylodynamics~\autocite{holmes1995revealing}.}
-
-\uline{We expect the tree kernel to classify the \gls{BA} parameters more
-accurately than either Sackin's index or the \gls{nltt}, since the tree kernel
-takes both topology and branch lengths into account. }}
-
-This experiment involved a large number of variables which were varied
+any of the four parameters were identifiable from the shape of the transmission
+tree, as quantified by the tree kernel. Each of the \gls{BA} model parameters
+was varied one at a time, while holding the other parameters at fixed, known
+values. Contact networks were generated according to each set of parameter
+values, and transmission trees were simulated over the networks. We then
+evaluated how well a classifier based on the tree kernel could differentiate
+the trees simulated under distinct parameter values. If the classifier's
+cross-validation accuracy was high, this could be taken as an indication that
+the parameter in question was identifiable in the range of values considered. 
+A caveat of this preliminary analysis is that, since all parameters but one
+were held at known values, nothing could be said about the identifiability of
+\emph{combinations} of parameters; this issue will be explored later by jointly
+estimating all parameters with \gls{ABC}.}
+
+\uline{In addition to testing for identifiability, a secondary objective of
+this analysis was to validate the use of the tree kernel as a distance measure
+for \gls{ABC} in our context. As discussed in} \cref{sec:abc}, \uline{the
+choice of distance function is extremely important for the accuracy of the
+\gls{ABC} approximation to the posterior. Therefore, we evaluated two
+additional tree statistics in the same manner as we evaluated the tree kernel
+(that is, by constructing and testing a classifier). First, we considered
+Sackin's index~\autocite{shao1990tree}, which measures the degree of imbalance
+or asymmetry in a phylogeny} (see \cref{subsec:treeshape}). \uline{Sackin's
+index is widely used for characterizing
+phylogenies~\autocite{frost2013modelling} and has been demonstrated to vary
+between transmission trees simulated under different contact network
+types~\autocite{leventhal2012inferring}. Sackin's index does not take branch
+lengths into account, considering only the tree's topology. The other statistic
+we considered was the \gls{nltt}~\autocite{janzen2015approximate}, which
+compares two trees based on normalized distributions of their branching times}
+(see \cref{sec:treeshape}).  \uline{In contrast with Sackin's index, the
+\gls{nltt} does not explicity consider the trees' topologies, but it does use
+their normalized branch lengths. While the \gls{nltt} is a newly developed
+statistic not yet in widespread use, the unnormalized
+\gls{ltt}~\autocite{nee1992tempo} was the basis of seminal early work
+extracting epidemiological information from
+phylodynamics~\autocite{holmes1995revealing}. We expected the tree kernel to
+classify the \gls{BA} parameters more accurately than either Sackin's index or
+the \gls{nltt}, since the tree kernel takes both topology and branch lengths
+into account. }}
+
+This experiment involved a large number of variables that were varied
 combinatorially. For ease of exposition, we will describe a single experiment
@@
 repeated. The parameters of the tree kernel, $\lambda$ and $\sigma$
-(\cref{subsec:treeshape}) will be referred to as \defn{meta-parameters} to
+(\cref{subsec:treeshape}), will be referred to as \defn{meta-parameters} to
 distinguish them from the parameters of the \gls{BA} model.
@@
 simulated tree using our \software{treestat} binary. The trees were visualized
-using the \software{ape} package~\autocite{paradis2004ape}. Our
-\software{treekernel} binary was used to calculate the value of the kernel for
-each pair of trees, with the meta-parameters set to $\lambda = 0.3$ and $\sigma
-= 4$. These values were stored in a symmetric 300 $\times$ 300 kernel matrix.
-Similarly, we computed the \gls{nltt} statistic between each pair of trees
-using our \software{treestat} binary, and stored them in a second $300 \times
-300$ matrix.
+using the \software{ape} package~\autocite{paradis2004ape}. {\color{blue}\uline{
+Both the tree kernel and the \gls{nltt} are pairwise statistics, and the
+\glspl{SVM} classifiers we used to investigate them operate on pairwise
+distance matrices.}} Our \software{treekernel} binary was used to calculate the
+value of the kernel for each pair of trees, with the meta-parameters set to
+$\lambda = 0.3$ and $\sigma = 4$. These values were stored in a symmetric 300
+$\times$ 300 kernel matrix.  Similarly, we computed the \gls{nltt} statistic
+between each pair of trees using our \software{treestat} binary, and stored
+them in a second $300 \times 300$ matrix.
 
@@
 \gls{BA} model. In contrast to the previous two experiments, we did not inform
-the algorithm of any of the true parameter values.
-}}
-
-We simulated three trees each under a variety of parameter values, and ran the
+the algorithm of any of the true parameter values.  }}
+
+We simulated three trees each under a variety of parameter values and ran the
 \software{netabc} program to estimate posterior distributions for the
@@
 and $\sigma = 4$. The \gls{SMC} algorithm was run with 1000 particles, five
-sampled datasets per particle, and the $\alpha$ parameter (not to be confused
-with the \gls{BA} preferential attachment parameter, see
+sampled datasets per particle, and the $\alpha_{\ESS}$ parameter (not to be
+confused with the \gls{BA} preferential attachment parameter, see
 \cref{subsec:adaptsmc}) set to 0.95. The algorithm was stopped when the
@@
 among nodes, we generated a network where half the nodes were attached with
-power $\alpha$ = 0.5, and the other half with power $\alpha$ = 1.5. The other
+power $\alpha$ = 0.5 and the other half with power $\alpha$ = 1.5. The other
 parameters for this network were $N$ = 5000, $I$ = 1000, and $m$ = 2. To
@@
 distinct (\cref{fig:alphatrees}). In particular, higher values of \gls{alpha}
-produce networks with a small number of highly connected nodes which, once
+produce networks with a small number of highly connected nodes, which, once
 infected, are likely to transmit to many other nodes. This results in a more
@@
 lower \gls{alpha} values. None of the other three parameters produced trees
-which were as easily distinguished from each other
+that were as easily distinguished from each other
 (\cref{fig:Itrees,fig:mtrees,fig:Ntrees,fig:Itrees}).  Sackin's index, which
@@
 \gls{N} separated more clearly with larger numbers of tips, and in the case of
-\gls{N}, larger epidemic sizes.
+\gls{N}, with larger epidemic sizes {\color{blue}(\cref{fig:Ikpca,fig:Nkpca})}.
 
@@
 
-The accuracy of classifiers $I$ varied significantly with the number of tips in
-the tree. For 100-tip trees, the average $R^2$ values were
+The accuracy of classifiers for $I$ varied significantly with the number of
+tips in the tree. For 100-tip trees, the average $R^2$ values were
   \Sexpr{kernel.I[ntip == 100, round(mean(rsquared), 2)]},
@@
       simulated transmission trees over networks with three different values of
-      the parameter being tested. Vertical lines are empirical 95\% confidence
-      intervals based on 1000 two-fold cross-validations. {\color{blue}\uline{
-      The classifiers for \gls{I} were not evaluated with 1000-tip trees,
-      because one of the tested \gls{I} values was 500, and it is not possible
-      to sample a tree of size 1000 from 500 infected individuals.}}
+      the parameter being tested, assuming perfect knowledge of the other
+      parameters. Vertical lines are empirical 95\% confidence intervals based
+      on 1000 two-fold cross-validations. {\color{blue}\uline{ The classifiers
+      for \gls{I} were not evaluated with 1000-tip trees, because one of the
+      tested \gls{I} values was 500, and it is not possible to sample a tree of
+      size 1000 from 500 infected individuals.}}
   }
@@
   \vspace{6pt}
-  \caption{
+  \caption[
+    Marginal posterior distributions of BA model parameters estimated
+    with kernel-assisted ABC for a single simulated transmission tree.
+  ]{
     Marginal posterior distributions of BA model parameters estimated
     with kernel-assisted ABC for a single simulated transmission tree. Dotted
-    lines and shaded polygon indicate true values.
+    lines indicate true values, solid lines indicate posterior mean point
+    estimates, and shaded areas show 95\% highest posterior density intervals.
   }
@@
     injection drug users. The \textcite{novitsky2013phylogenetic,novitsky2014impact} data
-    were sampled from a primarily heterosexual risk environment, but did not
+    were sampled from a primarily heterosexual risk environment but did not
     explicitly exclude other risk factors.
@@
 aligned pairwise to the HXB2 reference sequence (GenBank accession number
-K03455) and the hypervariable regions were clipped out with
+K03455), and the hypervariable regions were clipped out with
 \software{BioPython} version 1.66+~\autocite{cock2009biopython}. Sequences were

### 2016-05-30 doc/thesis/methods_results.tex
A bit more writing

--- 2016-05-26 doc/thesis/methods_results.tex
@@
 \end{table}
+
+{\color{blue}\uline{
+To evaluate the effects of the true parameter values on the accuracy of the
+posterior mean estimates, we analysed the \gls{alpha} and \gls{I} parameters
+individually using \glspl{GLM}. The response variable was the error of the
+point estimate, and the predictor variables were the true values of
+\gls{alpha}, \gls{I}, and \gls{m}. We did not test for differences across true
+values of \gls{N}, because \gls{N} was not varied in these simulations. The
+distribution family and link function for the \glspl{GLM} were chosen as
+Gaussian and inverse, respectively, by examination of residual plots and
+\gls{AIC}. The $p$-values of the estimated \glspl{GLM} coefficients were
+corrected using Holm-Bonferroni correction~\autocite{holm1979simple} with $n =
+6$ (two \glspl{GLM} with three predictors each). Because there was clearly
+little to no identifiability of the \gls{N} and \gls{m} parameters with
+\gls{ABC} (see Results in next section), we did not construct \glspl{GLM} for
+those parameters. }}
 
@@
 
-%To evaluate the effects of the true parameter values on the accuracy of the
-%\gls{MAP} estimates, we analysed each parameter individually using a \gls{GLM}
-%(four \glspl{GLM} total). The response variable was the error of the point
-%estimate, and the predictor variables were the true values of \gls{alpha},
-%\gls{I}, and \gls{m}. We did not test for differences across true values of
-%\gls{N}, because \gls{N} was not varied in these simulations. The distribution
-%family and link function for each \gls{GLM} were chosen by examination of
-%residual plots and \gls{AIC} of the model fits. For \gls{alpha}, \gls{I}, and
-%\gls{N}, we used the Gamma distribution with the inverse link function. For
-%\gls{m}, we used the Poisson distribution with the log link function. To
-%allow for the possibility that the relationships between parameters and error
-%rates were nonlinear, the true parameter values were treated as discrete rather
-%than continuous predictors. When significant associations were found by the
-%\glspl{GLM}, additional post hoc tests were performed based on the observed
-%relationship between the error rate and true parameter value in question. If
-%the relationship appeared to be monotonic, Spearman's correlation was
-%calculated. Otherwise, if there were particular values which appeared to have a
-%higher error rate than the others, the Wilcoxon rank-sum test was used to check
-%for a significant difference between the two groups.
-
 Two further experiments were performed to address potential sources of error.
@@
 \subsubsection*{Marginal parameter estimates with grid search}
-
-The accuracy of grid search estimates largely paralleled that of the \gls{kSVR}
-classifiers. \Cref{fig:gridest} shows point estimates and 95\% highest density
-intervals for each of the \gls{BA} parameters, for one replicate experiment
-with 500-tip trees. Plots showing the point estimates for all replicates can be
-found in \cref{fig:gridptalpha,fig:gridptI,fig:gridptm,fig:gridptN}. For all
-parameters except $m$, the error of point estimates was negatively correlated
-with the number of sampled tips in the tree (for
-\gls{alpha}, \gls{I}, and \gls{N} respectively: Spearman's $\rho$ = 
-    \ensuremath{-0.22},
-    \ensuremath{-0.51},
-    \ensuremath{-0.16};
-$p$-values
-    $4\!\times\!10^{-4}$,
-    ${<}10^{-5}$,
-    $0.01$).
-The highest density intervals obtained for all parameters were extremely wide,
-occupying $>$75\% of the grid in all cases (\cref{fig:gridest}).
-
-The \gls{alpha} parameter was the most accurately estimated, with point
-estimates having an average deviation of 
-    0.14
-from the true value, on a grid from 0 to 2. The error of point estimates varied
-significantly between true values of \gls{alpha}
-    (one-way \gls{ANOVA}, $p$ ${<}10^{-5}$). In
-particular, errors were lower for the values \gls{alpha} = 1.0 and 1.25 than
-for the other values
-    (average errors 
-    0.03
-    for \gls{alpha} = 1.0 or 1.5 vs.
-    0.17
-    for \gls{alpha} $\neq$ 1.0 or 1.5),
-and this difference was significant
-    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
-     \cref{fig:gridptalpha}).
-These two values exhibited different qualitative behaviour than the other
-values in terms of the distribution of kernel scores along the grid
-(\cref{fig:gridalpha}). In particular, there was a pronounced peak in scores
-around the true value, in contrast to the other values where the scores were
-flat around the true value. The effect was most obvious for the value
-\gls{alpha} = 1.25.
-
-The average absolute error of the point estimates for \gls{I} was 
-    310 individuals,
-on a grid of 500 to 5000, and these errors differed between true values of
-\gls{I}
-    (one-way \gls{ANOVA}, $p =0.001$).
-The errors for $2000 \leq I \leq 3000$ were higher than those for the other
-values
-    (average errors
-     430
-     for $2000 \leq I \leq 3000$ vs.
-     250
-     for $I < 2000$ or $I > 3000$),
-and this difference was significant
-    (Wilcoxon rank-sum test, $p =6\!\times\!10^{-4}$,
-     \cref{fig:gridptI}).
-Kernel score distributions for all test values exhibited a similar rounded
-shape (\cref{fig:gridI}).
-
-The average error for \gls{m} was
-    1.31 edges per vertex,
-on a grid from 1 to 6. The error varied significantly between the true values
-of \gls{m} 
-    (one-way \gls{ANOVA}, $p {<}10^{-5}$).
-Errors for the value \gls{m} = 1 were lower than the other values
-    (average errors
-    0.1
-    for $m = 1$ vs.
-    1.55
-    for $m > 1$),
-and this difference was significant
-    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
-     \cref{fig:gridptm}).
-The value $m = 1$ causes the network to take on a distinct shape relative to
-higher \gls{m} values, namely a tree (\ie there are no cycles, see
-\cref{subsec:treeshape}). The kernel score distribution had a peak at $m = 1$
-when this was the true value, and a valley at $m = 1$ when the true value of
-$m$ was greater that 1 (\cref{fig:gridm}).
-
-The average error for \gls{N} was 
-    2419 individuals,
-on a grid from 1000 to 15000, and was varied significantly with the true value
-of \gls{N}
-    (one-way \gls{ANOVA}, $p {<}10^{-5}$).
-The errors were lower for $N \leq 3000$
-    (average errors
-     740
-     for $N \leq 3000$ vs.
-     2979
-     for $N > 3000$),
-and this difference was significant
-    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
-     \cref{fig:gridptN}).
-The kernel score distribution had a peak at $N = 1000$ when this was the true
-value, and a valley there otherwise (\cref{fig:gridN}). Except for this valley,
-the distributions were flat for $N > 3000$.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth]{gridsearch-example}
-  \caption[Grid search estimates of \gls{BA} model parameters.]{Point estimates
-      and 95\% highest density intervals for each \gls{BA} model parameter,
-      obtained using grid search. Networks and transmission trees were
-      simulated over a grid of values for each parameter while holding the
-      others fixed. For a subset of the grid values ($x$-axis), test networks
-      and trees were created and compared to each tree on the grid using the
-      tree kernel. The kernel scores along the grid were normalized to resemble
-      a probability distribution, from which the mode and highest density
-      interval were calculated. Shown values correspond to one replicate
-      experiment, with trees of size 500.
-  } 
-  \label{fig:gridest}
-\end{figure}
-
-\subsubsection*{Joint parameter estimates with kernel-assisted ABC}
-
-\Cref{fig:abcptm2} shows {\color{red}\sout{\gls{MAP}}}
-{\color{blue}\uline{posterior mean}} point estimates of the BA model parameters
-obtained with kernel-assisted ABC on simulated data. The estimates shown
-correspond only to the simulations where the $m$ parameter was set to 2,
-however the results for $m = 3$ and $m = 4$ were similar
-(\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% HPD intervals are
-given in \cref{tab:abchpd}.
-
-The accuracy of the parameter estimates obtained with kernel-assisted ABC
-paralleled the results from the \gls{kSVR} classifier. Of the four parameters,
-$\alpha$ was the most accurately estimated, with point estimates having a
-median [IQR] absolute error of 
-    0.11 
-    [0.03 - 
-    0.25].
-The errors when the true value of $\alpha$ was zero were significantly greater
-than those for the other values 
-    (Wilcoxon rank-sum test, $p$ = $0$).
-Errors in estimating $\alpha$ 
-%also varied with the true value of $m$ just at
-%the threshold of statistical significance
-%    (one-way ANOVA, $p 
-%    pp(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"], eq=TRUE)$),
-{\color{blue}\uline{did not vary across the true values of $m$ or $I$ (both
-one-way ANOVA).}}
-
-Estimates for $I$ were relatively accurate, with point estimate errors of
-    492 
-    [294 - 
-    782] individuals.
-These errors were significantly higher when the true value of $\alpha$ was
-at least 1
-    (Wilcoxon rank-sum test, $p$ = $0$)
-and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
-did not affect the estimates of $I$ (one-way ANOVA).
-
-The $m$ parameter was estimated correctly in
-    37 \%
-of simulations {\color{red}\sout{barely better than random guessing}}.
-{\color{blue}\uline{However this was due to the estimated value being $m = 3$
-in the large majority of simulations (62 out of
-72), with one-third of the simulations having the true value $m =
-3$.}} The true values of the other parameters did not significantly affect the
-estimates of $m$ (both one-way ANOVA).
-
-Finally, the total number of nodes $N$ was consistently over-estimated by about
-a factor of two (error 4153
-[3660 - 4489] individuals). No parameters influenced the accuracy of the $N$
-estimates (all one-way ANOVA).
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth]{abc-point-estimate-m2}
-  \vspace{6pt}
-  \caption[
-    Posterior mean point estimates for \gls{BA} model parameters obtained by
-    running \software{netabc} on simulated data, for simulations with \gls{m} =
-    2.
-  ]{
-    Posterior mean point estimates for \gls{BA} model parameters obtained by
-    running \software{netabc} on simulated data, for simulations with \gls{m} =
-    2. Dashed lines indicate true values. (A) Estimates of \gls{alpha} and
-    \gls{I} which were varied in these simulations against known values. (B)
-    Estimates of \gls{m} and \gls{N} which were held fixed in these simulations
-    at the values \gls{m} = 2 and \gls{N} = 5000.
-  }
-  \label{fig:abcptm2}
-\end{figure}
-
-\begin{table*}[ht]
-  \centering
-  \input{\tablepath/abc-hpd.tex}
-  \caption[
-      Average posterior mean point estimates and 95\% highest posterior density
-      (HPD) interval widths for BA model parameter estimates obtained with
-      kernel-assisted ABC.
-  ]{
-      Average posterior mean point estimates and 95\% highest posterior density
-      (HPD) interval widths for BA model parameter estimates obtained with
-      kernel-assisted ABC. Three transmission trees were simulated under each
-      combination of the listed parameter values, and the parameters were
-      estimated with kernel-assisted ABC without training.
-  }
-  \label{tab:abchpd}
-\end{table*}
-
-The dispersion of the ABC approximation to the posterior also varied between
-the parameters, with narrower HPD intervals for the parameters with more
-accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows
-the distributions for one simulation. Equivalent plots for one replicate
-simulation with each studied parameter combination can be found in
-\cref{fig:0.0-1000-2-5000-0,fig:0.5-1000-2-5000-0,fig:1.0-1000-2-5000-0,fig:1.5-1000-2-5000-0,fig:0.0-2000-2-5000-0,fig:0.5-2000-2-5000-0,fig:1.0-2000-2-5000-0,fig:1.5-2000-2-5000-0,fig:0.0-1000-3-5000-0,fig:0.5-1000-3-5000-0,fig:1.0-1000-3-5000-0,fig:1.5-1000-3-5000-0,fig:0.0-2000-3-5000-0,fig:0.5-2000-3-5000-0,fig:1.0-2000-3-5000-0,fig:1.5-2000-3-5000-0,fig:0.0-1000-4-5000-0,fig:0.5-1000-4-5000-0,fig:1.0-1000-4-5000-0,fig:1.5-1000-4-5000-0,fig:0.0-2000-4-5000-0,fig:0.5-2000-4-5000-0,fig:1.0-2000-4-5000-0,fig:1.5-2000-4-5000-0}.
-HPD intervals around $\alpha$ and $I$ were often narrow
-relative to the region of nonzero prior density, whereas the intervals for $m$
-and $N$ were more widely dispersed.
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{abc-posterior-example}
-  \vspace{6pt}
-  \caption{
-    Marginal posterior distributions of BA model parameters estimated
-    with kernel-assisted ABC for a single simulated transmission tree. Dotted
-    lines and shaded polygon indicate true values.
-  }
-  \label{fig:abcex}
-\end{figure}
-
-To test the effect of model misspecification, we simulated one network where
-the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
-other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The posterior mean
-[95\% HPD] estimates for each parameter were: 
-$\alpha$, 
-  1.09 
-  [0.6 -
-   1.14];
-$I$,
-  1143 
-  [666 -
-   4455];
-$m$,
-  3 
-  [2 -
-   5];
-$N$,
-  12635 
-  [4282-
-   14977].
-The approximate posterior distributions for this simulation are shown in
-\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
-transmission tree in a peer-driven fashion, where the probability to sample a
-node was twice as high if one of its peers had already been sampled. The
-parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
-$I$ = 2000. The estimated values were
-$\alpha$, 
-  0.23 
-  [0 -
-   0.63];
-$I$,
-  2383 
-  [1419 -
-   3811];
-$m$,
-  4 
-  [2 -
-   5];
-$N$,
-  13517 
-  [2836 -
-   14720].
-The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
-of these results were in line with estimates obtained on other simulated
-datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
-$\alpha$ was somewhat lower than typical.
-
-%<<abc_glm, include=FALSE>>=
-%    source("global.R")
-%    options(scipen=-1, digits=2)
-%@
-%
-%<<alpha_glm, include=FALSE>>=
-%    # alpha_error is influenced by alpha and m but not I
-%    stopifnot(alpha.glm[Parameter == "alpha", min(p) < 0.05])
-%    stopifnot(alpha.glm[Parameter == "m", min(p) < 0.05])
-%    stopifnot(alpha.glm[Parameter == "I", min(p) > 0.05])
-%
-%    # alpha_error is correlated with true_alpha
-%    alpha.test <- d[,cor.test(alpha_error, true_alpha, method="spearman")]
-%    stopifnot(alpha.test$p.value < 0.05)
-%
-%    # alpha_error is not correlated with true_m
-%    alpha.m.cor <- d[,cor.test(alpha_error, true_m)]
-%    stopifnot(alpha.m.cor$p.value < 0.05)
-%@
-%
-%We used \software{netabc} to estimate the parameters of the \gls{BA} model on
-%simulated trees where the true parameter values were known. Point estimates for
-%each parameter are shown in \cref{fig:abcptm2} for the simulations with \gls{m} =
-%2. The results for the other values of \gls{m} were similar
-%(\cref{fig:abcptm3,fig:abcptm4}). The median [IQR] absolute error of estimates
-%of \gls{alpha} across all simulations was
-%    d[,median(alpha_error)] 
-%    [d[,quantile(alpha_error, 0.25)]-d[,quantile(alpha_error, 0.75)]].
-%\Gls{GLM} analysis indicated that the true values of both \gls{alpha} and
-%\gls{m} had significant effects on the error in estimated \gls{alpha}
-%    ($p$ values $pp(alpha.glm[Parameter == "alpha", min(p)], eq=FALSE)$
-%     $pp(alpha.glm[Parameter == "m", min(p)], eq=FALSE)$,
-%     \cref{tab:glmalpha}),
-%but the true value of \gls{I} did not. There was a significant negative
-%correlation between the true value of \gls{alpha} and the error
-%    (Spearman's $\rho$ = round(alpha.test$estimate, 2),
-%     $p pp(alpha.test$p.value, eq=TRUE)$).
-%There was a negative correlation between \gls{m} and the error in \gls{alpha}
-%    (Spearman's $\rho$ = round(alpha.m.cor$estimate, 2),
-%    $p pp(alpha.m.cor$p.value, eq=FALSE)$).
-%
-%<<I_glm, include=FALSE>>=
-%    # I_error is influenced by alpha and I but not by m
-%    stopifnot(I.glm[Parameter == "alpha", min(p) < 0.05])
-%    stopifnot(I.glm[Parameter == "m", min(p) > 0.05])
-%    stopifnot(I.glm[Parameter == "I", min(p) < 0.05])
-%
-%    # correlation between I_error and true_alpha
-%    I.alpha.cor <- d[,cor.test(I_error, true_alpha, method="spearman")]
-%    stopifnot(I.alpha.cor$p.value < 0.05)
-%
-%    # correlation between I_error and true_I
-%    I.I.cor <- d[,cor.test(I_error, true_I, method="spearman")]
-%    stopifnot(I.alpha.cor$p.value < 0.05)
-%@
-%
-%The mean error in the estimated value of \gls{I} was
-%    d[,round(median(I_error))] 
-%    [d[,round(quantile(I_error, 0.25))]-d[,round(quantile(I_error, 0.75))]],
-%an over-estimate of roughly a factor of 1.5
-%(\cref{fig:abcptm2,fig:abcptm1,fig:abcptm3,fig:abcptm4}). \Gls{GLM} analysis
-%indicated a relationship between the error in estimated \gls{I}, and the true
-%values of \gls{alpha} and \gls{I}
-%    ($p$ values $pp(I.glm[Parameter == "alpha", min(p)], eq=FALSE)$
-%     $pp(alpha.glm[Parameter == "I", min(p)], eq=FALSE)$,
-%     \cref{tab:glmalpha}),
-%but not the true value of \gls{m}. There was a significant correlation between 
-%the true value of \gls{alpha} and the error in estimated \gls{I}
-%    (Spearman's $\rho$ = I.alpha.cor$estimate,
-%     $p pp(I.alpha.cor$p.value, eq=TRUE)$),
-%and between the true value of \gls{I} and the error in estimated \gls{I}
-%    (Spearman's $\rho$ = I.I.cor$estimate,
-%     $p pp(I.I.cor$p.value, eq=TRUE)$).
-%
-%<<m_glm, include=FALSE>>=
-%    # effects of parameters on m_error
-%    stopifnot(m.glm[Parameter == "m", min(p)] < 0.05)
-%    stopifnot(m.glm[Parameter == "alpha", min(p)] > 0.05)
-%    stopifnot(m.glm[Parameter == "I", min(p)] > 0.05)
-%
-%    # m was only estimated correctly in just over 20\% of simulations with m > 2
-%    m.tbl <- prop.table(table(d[true_m > 1,m_error]))
-%    stopifnot(m.tbl[1] > 0.2 & m.tbl[1] < 0.3)
-%
-%    # with m = 1, over 95% of simulations are right
-%    m1.est <- d[true_m == 1, m_error]
-%    stopifnot(sum(m1.est == 0) / length(m1.est) > 0.95)
-%    stopifnot(sum(m1.est == 0) / length(m1.est) < 1)
-%@
-%
-%\gls{GLM} analysis showed an effect of the true value of \gls{m} on the error
-%in the estimated \gls{m}
-%    ($p pp(m.glm[Parameter == "m", min(p)])$,
-%     \cref{tab:glmm}).
-%When the true value of \gls{m} was 1, the correct value was recovered by
-%\software{netabc} in virtually every case
-%    (sum(m1.est == 0) out of length(m1.est) simulations).
-%However, when the true value of \gls{m} was 2 or higher, the correct value was
-%recovered in only 
-%    as.integer(m.tbl[1] * 100)\%
-%of simulations, little better than random guessing. The \gls{GLM} analysis did
-%not indicate any effects of the true parameter values on the error in estimated
-%\gls{m} (\cref{tab:glmm}).
-%
-%<<N_glm, include=FALSE>>=
-%    # alpha_error is influenced by alpha and m but not I
-%    stopifnot(N.glm[Parameter == "alpha", min(p) < 0.05])
-%    stopifnot(N.glm[Parameter == "I", min(p) < 0.05])
-%    stopifnot(N.glm[Parameter == "m", min(p) > 0.05])
-%
-%    N.alpha.test <- d[,wilcox.test(.SD[true_alpha == 1.5, N_error], 
-%                                   .SD[true_alpha < 1.5, N_error])]
-%    stopifnot(N.alpha.test$p.value < 0.05)
-%    N.I.test <- d[,wilcox.test(.SD[true_I == 3000, N_error], 
-%                               .SD[true_I < 3000, N_error])]
-%    stopifnot(N.I.test$p.value < 0.05)
-%@
-%
-%Finally, the total number of nodes \gls{N} was consistently over-estimated by
-%about a factor of two
-%    (error d[,format(median(N_error), scientific=FALSE)] 
-%    [d[,format(quantile(N_error, 0.25), scientific=FALSE)] - 
-%     d[,format(quantile(N_error, 0.75), scientific=FALSE)]]).
-%The fitted \gls{GLM} indicated that the true values of both \gls{alpha} and
-%\gls{I} had an effect on the error in the estimated \gls{N}
-%    ($p$-values $pp(N.glm[Parameter == "alpha", min(p)], eq=FALSE)$ and 
-%     $pp(N.glm[Parameter == "I", min(p)], eq=FALSE)$,
-%     \cref{tab:glmN}),
-%but that the true value of \gls{m} did not. The error in the estimated \gls{N}
-%when \gls{alpha} was equal to 1.5 was slightly lower than for other values of
-%\gls{alpha}
-%    (median [IQR] error rates 
-%     d[true_alpha == 1.5, format(median(N_error), scientific=FALSE)] 
-%    [d[true_alpha == 1.5, format(quantile(N_error, 0.25), scientific=FALSE)] - 
-%     d[true_alpha == 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]]
-%     for \gls{alpha} = 1.5 vs. 
-%     d[true_alpha < 1.5, format(median(N_error), scientific=FALSE)] 
-%    [d[true_alpha < 1.5, format(quantile(N_error, 0.25), scientific=FALSE)] - 
-%     d[true_alpha < 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]]
-%     for \gls{alpha} < 1.5),
-%and this difference was statistically significant
-%    (Wilcoxon rank-sum test, $p pp(N.alpha.test$p.value, eq=TRUE)$).
-%Similarly, the error in the estimated \gls{N} when \gls{I} was equal to 3000 was
-%lower than for other values of \gls{I}
-%    (median [IQR] error rates 
-%     d[true_I == 3000, format(median(N_error), scientific=FALSE)] 
-%    [d[true_I == 3000, format(quantile(N_error, 0.25), scientific=FALSE)] - 
-%     d[true_I == 3000, format(quantile(N_error, 0.75), scientific=FALSE)]]
-%     for \gls{I} = 3000 vs. 
-%     d[true_alpha < 3000, format(median(N_error), scientific=FALSE)] 
-%    [d[true_alpha < 3000, format(quantile(N_error, 0.25), scientific=FALSE)] - 
-%     d[true_alpha < 3000, format(quantile(N_error, 0.75), scientific=FALSE)]]
-%     for \gls{I} < 3000),
-%and this difference was statistically significant
-%    (Wilcoxon rank-sum test, $p pp(N.I.test$p.value, eq=TRUE)$).
-%
-%\begin{figure}
-%  \includegraphics{abc-point-estimate-m2}
-%  \caption[\Acrlong{MAP} point estimates for \gls{BA} model parameters obtained
-%    by running \software{netabc} on simulated data, for simulations with $m = 2$.] 
-%  {
-%    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by         
-%    running \software{netabc} on simulated data. Values shown are for               
-%    simulations with \gls{m} = 2. Dashed lines indicate true values. (A)            
-%    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations  
-%    against known values. (B) Estimates of \gls{m} and \gls{N} which were held   
-%    fixed in these simulations at the values \gls{m} = 2 and \gls{N} = 5000. 
-%  }
-%  \label{fig:abcptm2}
-%\end{figure}
-%
-%<<posterior_sims, echo=FALSE>>=
-%    N <- 5000
-%    replicate <- 0
-%    lab <- NULL
-%    for (m in c(2, 3, 4)) {
-%    for (I in c(1000, 2000)) {
-%    for (alpha in c(0, 0.5, 1, 1.5)) {
-%        lab <- c(lab, sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate))
-%    }
-%    }
-%    }
-%    lab <- paste(lab, collapse=",")
-%@
-%
-%\Cref{fig:abcex} shows the \gls{ABC} approximation to the posterior
-%distribution on the \gls{BA} parameters for one simulation. Equivalent plots
-%for one replicate simulation with each combination of parameters can be found
-%in \cref{lab}. \Gls{HPD} intervals around \gls{alpha} and \gls{I} were
-%narrow relative to the region of nonzero prior density, whereas the intervals
-%for $m$ and \gls{N} were widely dispersed. \Cref{tab:abchpd} shows point
-%estimates and 95\% \gls{HPD} intervals averaged over all simulations.
-%
-%\begin{figure}
-%  \includegraphics{{abc-posterior/1.0_1000_2_5000_0}.pdf}
-%  \caption[Approximate marginal posterior distributions of BA model parameters
-%      obtained by applying \textit{netabc} to a simulated transmission tree
-%      with values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]
-%    {
-%        Approximate marginal posterior distributions of BA model parameters
-%        obtained by applying \textit{netabc} to a simulated transmission tree
-%        with BA parameter values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ =
-%        5000. Vertical dashed lines indicate true values. Shaded areas are 95\%
-%        highest posterior density intervals. $x$-axes indicate regions of
-%        nonzero prior density.
-%    }
-%  \label{fig:abcex}
-%\end{figure}
-%
-%
-%\begin{table}
-%    \centering
-%    \input{\tablepath/abc-hpd}
-%    \caption[
-%        Maximum \textit{a priori} estimates and 95\% highest posterior density
-%        (HPD) interval boundaries for \gls{BA} model parameters estimated with
-%        \software{netabc}, averaged over simulated transmission trees.
-%    ]
-%    {
-%        Maximum \textit{a priori} estimates and 95\% highest posterior density
-%        (HPD) interval boundaries for \gls{BA} model parameters estimated with
-%        \software{netabc}, averaged over simulated transmission trees.
-%    }
-%    \label{tab:abchpd}
-%\end{table}
-%
-%<<mixed_peerdriven, include=FALSE>>=
-%  library(coda)
-%  d <- fread("bzcat ../../simulations/abc-pa-mixed-alpha/abc/*")
-%  d <- d[iter == max(iter)]
-%  d <- d[sample(1:nrow(d), prob=weight, replace=TRUE)]
-%  params <- c("alpha", "m", "N", "I")
-%  d <- melt(d, measure.vars=params, variable.name="parameter")
-%  f <- function (x) {
-%    dens <- density(x)
-%    hpd <- HPDinterval(mcmc(x))
-%    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
-%         hpd.max=hpd[,"upper"])
-%  }
-%  d <- d[,f(value), by=parameter]
-%  setkey(d, parameter)
-%
-%  m <- collect.metadata("../../simulations/abc-pa-peerdriven/abc/*")
-%  p <- fread(paste("bzcat", rownames(m)[m$sample_peer == 1]))
-%  p <- p[iter == max(iter)]
-%  p <- p[sample(1:nrow(p), prob=weight, replace=TRUE)]
-%  params <- c("alpha", "m", "N", "I")
-%  p <- melt(p, measure.vars=params, variable.name="parameter")
-%  f <- function (x) {
-%    dens <- density(x)
-%    hpd <- HPDinterval(mcmc(x))
-%    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
-%         hpd.max=hpd[,"upper"])
-%  }
-%  p <- p[,f(value), by=parameter]
-%  setkey(p, parameter)
-%  options(scipen=5)
-%@
-%
-%To test the effect of model misspecification, we simulated one network where
-%the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
-%other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
-%HPD] estimates for each parameter were: 
-%$\alpha$, 
-%  round(d["alpha", map], 2) 
-%  [round(d["alpha", hpd.min], 2) -
-%   round(d["alpha", hpd.max], 2)];
-%$I$,
-%  d["I", round(map)] 
-%  [d["I", round(hpd.min)] -
-%   d["I", round(hpd.max)]];
-%$m$,
-%  d["m", floor(map)] 
-%  [d["m", floor(hpd.min)] -
-%   d["m", floor(hpd.max)]];
-%$N$,
-%  d["N", round(map)] 
-%  [d["N", round(hpd.min)] -
-%   d["N", round(hpd.max)]].
-%The approximate posterior distributions for this simulation are shown in
-%\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
-%transmission tree in a peer-driven fashion, where the probability to sample a
-%node was twice as high if one of its peers had already been sampled. The
-%parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
-%$I$ = 2000. The estimated values were
-%$\alpha$, 
-%  p["alpha", round(map, 2)] 
-%  [p["alpha", round(hpd.min, 2)] -
-%   p["alpha", round(hpd.max, 2)]];
-%$I$,
-%  p["I", floor(map)] 
-%  [p["I", floor(hpd.min)] -
-%   p["I", floor(hpd.max)]];
-%$m$,
-%  p["m", floor(map)] 
-%  [p["m", floor(hpd.min)] -
-%   p["m", floor(hpd.max)]];
-%$N$,
-%  p["N", floor(map)] 
-%  [p["N", floor(hpd.min)] -
-%   p["N", floor(hpd.max)]].
-%The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
-%of these results were in line with estimates obtained on other simulated
-%datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
-%$\alpha$ was somewhat lower than typical.
-
-%\subsubsection*{Effect of parameters on power-law exponent}
-%
-%\Cref{tab:glm} shows the estimated parameters for a log-link \gls{GLM} fitted
-%to the observed distribution of \gls{gamma} values. The coefficients are
-%interpretable as multiplicative effects.
-%
-%\begin{table}
-%    \centering
-%    \input{\tablepath/pa-gamma-glm}
-%    \caption{Estimated \gls{GLM} parameters for relationship between power-law
-%    exponent \gls{gamma} and \gls{BA} model parameters.}
-%    \label{tab:glm}
-%\end{table}
-
-\section{Application to real world HIV data}
-\label{sec:hiv}
-
-\subsection{Methods}
-
-{\color{blue}\uline{
-The synthetic data experiments presented in the previous section demonstrated 
-that the \gls{alpha} and \gls{I} parameters, and to a lesser extent the \gls{N}
-parameter, had measurable impacts on tree shape and were identifiable with
-kernel-assisted \gls{ABC}. Therefore, our next step was to use
-\software{netabc} to estimate the parameters of the \gls{BA} model for real
-world \gls{HIV} epidemics. }} Because the \gls{BA} model assumes a single
-connected contact network, it is most appropriate to apply to groups of
-individuals who are epidemiologically related. Therefore, we searched for
-published \gls{HIV} datasets which originated from existing clusters, either
-phylogenetically or geographically defined. In addition, we analysed an
-in-house dataset sampled from \gls{HIV}-positive individuals in British
-Columbia, Canada (the ``BC data''). The datasets are summarized in
-\cref{tab:data}.
-
-\begin{table}[ht]
-  \centering
-  \input{\tablepath/realdata-bc}
-  \caption[Characteristics of published HIV datasets analyzed with \software{netabc}.]
-  {
-    Characteristics of published HIV datasets analyzed with \software{netabc}. 
-    Abbreviations: MSM, men who have sex with men; HET, heterosexual; IDU,
-    injection drug users. The \textcite{novitsky2013phylogenetic,novitsky2014impact} data
-    were sampled from a primarily heterosexual risk environment, but did not
-    explicitly exclude other risk factors.
-  }
-  \label{tab:data}
-\end{table}
-
-We downloaded all sequences associated with each published study from GenBank.
-For the \textcite{novitsky2014impact} data, each \textit{env} sequence was
-aligned pairwise to the HXB2 reference sequence (GenBank accession number
-K03455) and the hypervariable regions were clipped out with
-\software{BioPython} version 1.66+~\autocite{cock2009biopython}. Sequences were
-multiply aligned using \software{MUSCLE} version 3.8.31
-\autocite{edgar2004muscle}, and alignments were manually inspected with
-\software{Seaview} version 4.4.2 \autocite{gouy2010seaview}. Phylogenies were
-constructed from the nucleotide alignments by approximate maximum likelihood
-using \software{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
-\gls{GTR} model~\autocite{tavare1986some}. Transmission trees were estimated by
-rooting and time-scaling the phylogenies by root-to-tip regression, using a
-modified version of Path-O-Gen (distributed as part of
-BEAST~\autocite{drummond2007beast}) as described
-previously~\autocite{poon2015phylodynamic}.
-
-Three of the datasets \autocite[][and the BC data]{li2015hiv,novitsky2014impact}
-were initially much larger than the others, containing 1265, 1299, and 7923
-sequences respectively. To ensure that the analyses were comparable, we reduced
-these to a number of sequences similar to the smaller datasets. For the
-\citeauthor{li2015hiv} and BC datasets, we detected clusters of size 280 and
-399 respectively using a patristic distance cutoff of 0.02 as described
-previously~\autocite{poon2015impact}. Only sequences within these clusters were
-carried forward. For the \textcite{novitsky2014impact} data, no large clusters
-were detected using the same cutoff, so we analysed a subtree of size 180
-chosen arbitrarily.
-
-Empirical studies of contact networks often report the exponent $\gamma$ of the
-power law degree distribution. To compare our results to the literature, we
-simulated 100 networks each according to the \gls{MAP} parameter estimates
-obtained for each investigated dataset. The power-law exponent $\gamma$ was
-calculated for each network using the \software{fit\_power\_law} function in
-\software{igraph}, with the `R.mle' implementation. The median of the 100
-$\gamma$ values was taken as a point estimate for the associated dataset.
-
-For all datasets, we used the priors $\alpha$ $\sim$ Uniform(0, 2) and $N$ and
-$I$ jointly uniform on the region \{$n \leq N \leq 10000$, $n \leq I \leq
-10000$, $I \leq N$\}, where $n$ is the number of tips in the tree (see
-\cref{tab:data}). Since the value $m = 1$ produces networks with no cycles,
-which we considered fairly implausible, we ran one analysis with the prior $m
-\sim$ DiscreteUniform(1, 5), and one with the prior $m \sim$ DiscreteUniform(2,
-5). The other parameters to the SMC algorithm were the same as used for the
-simulation experiments, except that we used 10000 particles instead of 1000 to
-increase the accuracy of the estimated posterior. This was computationally
-feasible due to the small number of runs required for this analysis.
-
-\subsection{Results}
-
-We applied kernel-assisted ABC to five published HIV datasets (\cref{tab:data}),
-and found substantial heterogeneity among the parameter estimates
-(\cref{fig:abchpd,fig:abchpdm2}). Plots of the marginal posterior distributions
-for each dataset are shown in
-\cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.
-Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had
-estimated $\alpha$ values near unity for the prior allowing $m = 1$ (\gls{MAP}
-estimates [95\% \gls{HPD}] 
-  0.76 
-  [0.07 - 
-   1.31]
-and
-  0.61 
-  [0.01 -
-   1.03] respectively).
-The MAP estimates did not change appreciably when $m = 1$ was disallowed by the
-prior, although the credible interval of the \textcite{niculescu2015recent}
-data was narrower
-  (0.07 - 
-   1.31).
-When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
-low estimated $\alpha$ values
-  (0.31 
-  [0 - 
-  0.73]
-and
-  0.38 
-  [0.02 -
-   0.82]). 
-However, the MAP estimates increased when $m = 1$ was not permitted, although
-the HPD intervals remained roughly the same
-  (0.55 
-  [0.01 - 
-  0.93]
-and
-  0.5 
-  [0.07 -
-   0.94]).
-The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
-for both priors on $m$
-  (0.5 for $m \geq 1$;
-   0.34 for $m \geq 2$).
-However, the confidence interval was much wider when $m = 1$ was allowed
-  ([0.03 -
-    1.61] for $m \geq 1$ vs.
-   [0 -
-    0.73] for $m \geq 2$).
-
-For all the datasets except \citeauthor{novitsky2014impact}, estimated values
-of $I$ were below 2000 when $m = 1$ was allowed, with relatively narrow HPD
-intervals compared to the nonzero prior density region
-  (\citeauthor{cuevas2009hiv}, 943 
-  [285 -
-   2111];
-   \citeauthor{niculescu2015recent}, 1032
-  [136 - 
-   2847];
-  \citeauthor{li2015hiv}, 1437 
-  [364 -
-   2659];
-   \citeauthor{wang2015targeting}, 872
-  [173 - 
-   2119]).
-The \citeauthor{novitsky2014impact} data was the outlier, with a very high
-estimated $I$, and HPD interval spanning almost the entire prior region
-  (5586 
-  [187 -
-   8812]).
-The $I$ estimates and HPD intervals were generally robust to the choice of
-prior on $m$, with slightly narrower HPD intervals (compare
-\cref{fig:abchpd,fig:abchpdm2}).
-
-The MAP estimate of $m$ was equal to 1 for all but the
-\citeauthor{novitsky2014impact} data, when this value was allowed. However, the
-upper bound of the HPD interval was different for each dataset
-  (\citeauthor{niculescu2015recent}, 4;
-   \citeauthor{wang2015targeting}, 3;
-   \citeauthor{li2015hiv}, 1;
-   \citeauthor{cuevas2009hiv}, 2).
-When $m = 1$ was disallowed, the MAP for all datasets was either 2 or 3, with
-HPD intervals spanning the entire prior region. The estimates for the total
-number of nodes $N$ were largely uninformative for all samples, with almost all
-MAP estimates greater than 7500 and HPD intervals spanning almost the entire
-nonzero prior density region. The only exception was the \citeauthor{li2015hiv}
-data, for which the MAP estimate was lower 
-  (6049)
-when $m = 1$ was allowed.
-
-\begin{figure*}[ht]
-  \centering
-  \includegraphics{realdata-hpd-bc}
-  \caption[
-      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
-      parameters of the BA network model, fitted to six HIV datasets
-      with \software{netabc}.]
-  {
-      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
-      parameters of the BA network model, fitted to six HIV datasets with
-      \software{netabc}. Legend labels indicate risk group and country of
-      origin. Abbreviations: IDU, injection drug users; MSM, men who have sex
-      with men; HET, heterosexual.
-  }
-  \label{fig:abchpd}
-\end{figure*}
-
-To make our analyses comparable to the existing network literature, we
-estimated values of the power law exponent $\gamma$ for each of the datasets 
-investigated. The results are summarized in \cref{tab:gamma}. All of the
-estimated exponents were in the range $2 \leq \gamma \leq 2.5$, which is on the
-lower end of the range $2 \leq \gamma \leq 4$ reported in the literature.
-
-\begin{table}
-    \centering
-    \input{\tablepath/realdata-gamma.tex}
-    \caption[
-        Estimated power law exponents for six HIV datasets based on maximum
-        \textit{a priori} estimates of BA model parameters.
-    ]{
-        Estimated power law exponents for six HIV datasets based on maximum
-        \textit{a priori} estimates of BA model parameters. 100 networks were
-        simulated using \textit{MAP} parameter estimates obtained with
-        \software{netabc}. The power law exponent $\gamma$ was estimated for
-        each, and the median of those estimates was used as a point estimate
-        for the corresponding dataset.
-    }
-    \label{tab:gamma}
-\end{table}

### 2016-05-31 doc/thesis/intro.tex
More changes... i'm so unproductive

--- 2016-05-28 doc/thesis/intro.tex
@@
 
+\glsreset{SMC}
+
+\subsection{Overview and notation}
+
 {\color{blue}\uline{
@@
 transmission tree $T$, we are interested in the posterior distribution}
-\[
+\begin{align}
   \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)}.
-\]
+  \label{eq:post}
+\end{align}
 \uline{As outlined in} \cref{sec:obj}, \uline{both the likelihood $\Pr(T \mid
@@
 intractable. Hence, rather than computing the posterior distribution
-analytically, we will attempt to approximate it using a \defn{Monte Carlo}
-approach.
-TODO
+analytically, we will approximate it using a \defn{Monte Carlo} approach.
+The fundamental idea behind Monte Carlo methods is succinctly expressed by
+\textcite{liu2001theoretical}}:
+\begin{quote}
+    \uline{Monte Carlo's view of the world is that any probability distribution $\pi$, 
+    regardless of its complexity, can always be \emph{represented} by a
+    discrete sample from it. By ``represented'', we mean that any computation
+    of expectations using $\pi$ can be replaced to an acceptable degree of
+    accuracy by using the empirical distribution resulting from the discrete
+    sample.}
+\end{quote}
+\uline{In other words, if we are able to sample enough points from a
+distribution of interest, we will be able to make reasonably accurate
+statements about the distribution itself. For example, the expected value of
+the distribution can be estimated by the sample's population mean. The reason
+Monte Carlo methods will be useful in this work is that algorithms exist for
+obtaining samples from distributions, including as the aforementioned
+posterior, that are analytically intractable and from which direct sampling is
+not possible } (for a review see \autocite{robert2004monte}).
+\uline{\Gls{SMC}}~\autocite{doucet2000sequential, doucet2001introduction,
+liu2008monte} \uline{is one such algorithm.}
+
+\uline{ \Gls{SMC} is also known as the \emph{particle filter}. Rather than
+sampling points one at a time from the target distribution, \gls{SMC} considers
+a population of points or ``particles'', here denoted $\set{x^{(k)}}$. The
+particles are associated with weights, $\set{w(x^{(k)})}$. These weighted
+particles are a representation for the target distribution. For example, the 
+expected value of the distribution is approximated by}
+\[
+  \frac{1}{n} \sum_{k=1}^n x^{(k)} w(x^{(k)}),
+\]
+\uline{where $n$ is the number of particles. Initially, the particles do not
+represent the target distribution but rather a more tractable distribution from
+which direct sampling is straightforward. The word ``sequential'' is used to
+describe the iterative process of perturbation, resampling, and reweighting
+applied to the particles in such a way that they converge, collectively, to a
+sample from the target. }
+
+\uline{In this work, the distribution of interest is the posterior distribution
+\ref{eq:post}. The particles are particular values of the parameters $\theta$
+of the contact network model being studied. If we were taking a typical
+Bayesian Monte Carlo approach to this problem, the particles would end up
+weighted by their posterior probability and distributed in such a way that the
+weighted population was a reasonable representation of $\Pr(\theta \mid T)$.
+This turns out to be not straightforward in our case, as we will address in
+}\cref{sec:abc}. \uline{For now, nothing is lost by \ldots}}
+
+{\color{blue}\uline{
+The goal of this section of the introduction is to describe an algorithm called
+the \gls{SMC} sampler~\autocite{del2006sequential}, which forms the basis of
+the adaptive \gls{ABC}-\gls{SMC} algorithm we apply toward the main objective
+of this work. We begin by describing \gls{SIS}, which is a precursor to
+\gls{SMC} that samples from a sequence of distributions defined on spaces of
+increasing dimension. We then describe \gls{SMC} itself, which extends
+\gls{SIS} with a resampling step to fight particle degeneracy. Finally, we
+outline the \gls{SMC} sampler, which allows \gls{SMC} to be applied to
+sequences of distributions all defined on the same space. This terminology will
+become clear as the methods are described.
 }}
-
-\glsreset{SMC}
-
-\subsection{Overview and notation}
 
@@
 sharing a name (the word ``kernel'' is ubiquitous in mathematics).
-
-{\color{blue}\uline{
-The goal of this section of the introduction is to describe an algorithm called
-the \gls{SMC} sampler~\autocite{del2006sequential}, which forms the basis of
-the adaptive \gls{ABC}-\gls{SMC} algorithm we apply toward the main objective
-of this work. We begin by describing \gls{SIS}, which is a precursor to
-\gls{SMC} that samples from a sequence of distributions defined on spaces of
-increasing dimension. We then describe \gls{SMC} itself, which extends
-\gls{SIS} with a resampling step to fight particle degeneracy. Finally, we
-outline the \gls{SMC} sampler, which allows \gls{SMC} to be applied to
-sequences of distributions all defined on the same space. This terminology will
-become clear as the methods are described.
-}}
 
@@
 {\color{blue}\uline{The resampling step was formally integrated with
-\gls{SIS} by \citeauthor{doucet2000sequential} to form the first \gls{SMC}
+\gls{SIS} by \textcite{doucet2000sequential} to form the first \gls{SMC}
 algorithm }(\cref{alg:smc}). \uline{Rather than resample at every step as the
@@
 \defn{parametric} model defines a family of possible relationships between data
-and outcomes, indexed by one or more numeric parameters $\theta$. A
+and outcomes, parameterized by one or more numeric parameters $\theta$. A
 \defn{statistical} model describes the relationship between data and outcomes
@@
 \gls{MCMC} to obtain a sample from the posterior distribution. \Gls{MCMC} works
-by defining a Markov chain whose states are indexed by possible model
-parameters. The transition probability from state $\theta_1$ to state
-$\theta_2$ is taken to be
+by defining a Markov chain {\color{red}\sout{whose states are indexed by
+possible model parameters. The transition probability from state $\theta_1$ to
+state $\theta_2$ is taken to be}} {\color{blue}\uline{on the space of possible
+model parameters. The transition density from parameters $\theta_1$ to
+$\theta_2$ is taken to be}}
 \[

### 2016-05-31 doc/thesis/methods_results.Rnw
More changes... i'm so unproductive

@@
 confused with the \gls{BA} model parameter)}} controls the decay rate.
-{\color{blue}\uline{In the original paper of \textcite{del2012adaptive}, the
+{\color{blue}\uline{In the original paper of }\textcite{del2012adaptive}, \uline{the
 parameter is called $\alpha$, but to avoid confusion with the \gls{BA}
@@
 \emph{combinations} of parameters; this issue will be explored later by jointly
-estimating all parameters with \gls{ABC}.}
+estimating all parameters with \gls{ABC}} (\cref{subsec:abc}).
 
@@
 <<mixed_peerdriven, include=FALSE>>=
-  library(coda)
   d <- fread("bzcat ../../simulations/abc-pa-mixed-alpha/abc/*")
   d <- d[iter == max(iter)]
-  d <- d[sample(1:nrow(d), prob=weight, replace=TRUE)]
+  #d <- d[sample(1:nrow(d), prob=weight, replace=TRUE)]
   params <- c("alpha", "m", "N", "I")
   d <- melt(d, measure.vars=params, variable.name="parameter")
-  f <- function (x) {
-    dens <- density(x)
-    hpd <- HPDinterval(mcmc(x))
-    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
-         hpd.max=hpd[,"upper"])
-  }
-  d <- d[,f(value), by=parameter]
+  f <- function (x, wt) {
+    hpd <- wtd.hpd(x, wt/sum(wt))
+    list(pmean=wtd.mean(x, wt, normwt=TRUE), 
+         hpd.min=hpd[1], 
+         hpd.max=hpd[2])
+  }
+  d <- d[,f(value, weight), by=parameter]
   setkey(d, parameter)
@@
   p <- p[iter == max(iter)]
-  p <- p[sample(1:nrow(p), prob=weight, replace=TRUE)]
+  #p <- p[sample(1:nrow(p), prob=weight, replace=TRUE)]
   params <- c("alpha", "m", "N", "I")
   p <- melt(p, measure.vars=params, variable.name="parameter")
-  f <- function (x) {
-    dens <- density(x)
-    hpd <- HPDinterval(mcmc(x))
-    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
-         hpd.max=hpd[,"upper"])
-  }
-  p <- p[,f(value), by=parameter]
+  p <- p[,f(value, weight), by=parameter]
   setkey(p, parameter)
@@
 $\alpha$, 
-  \Sexpr{round(d["alpha", map], 2)} 
+  \Sexpr{round(d["alpha", pmean], 2)} 
   [\Sexpr{round(d["alpha", hpd.min], 2)} -
@@
 $I$,
-  \Sexpr{d["I", round(map)]} 
+  \Sexpr{d["I", round(pmean)]} 
   [\Sexpr{d["I", round(hpd.min)]} -
@@
 $m$,
-  \Sexpr{d["m", floor(map)]} 
+  \Sexpr{d["m", floor(pmean)]} 
   [\Sexpr{d["m", floor(hpd.min)]} -
@@
 $N$,
-  \Sexpr{d["N", round(map)]} 
+  \Sexpr{d["N", round(pmean)]} 
   [\Sexpr{d["N", round(hpd.min)]}-
@@
 $\alpha$, 
-  \Sexpr{p["alpha", round(map, 2)]} 
+  \Sexpr{p["alpha", round(pmean, 2)]} 
   [\Sexpr{p["alpha", round(hpd.min, 2)]} -
@@
 $I$,
-  \Sexpr{p["I", floor(map)]} 
+  \Sexpr{p["I", floor(pmean)]} 
   [\Sexpr{p["I", floor(hpd.min)]} -
@@
 $m$,
-  \Sexpr{p["m", floor(map)]} 
+  \Sexpr{p["m", floor(pmean)]} 
   [\Sexpr{p["m", floor(hpd.min)]} -
@@
 $N$,
-  \Sexpr{p["N", floor(map)]} 
+  \Sexpr{p["N", floor(pmean)]} 
   [\Sexpr{p["N", floor(hpd.min)]} -
@@
 %
-%<<mixed_peerdriven, include=FALSE>>=
-%  library(coda)
-%  d <- fread("bzcat ../../simulations/abc-pa-mixed-alpha/abc/*")
-%  d <- d[iter == max(iter)]
-%  d <- d[sample(1:nrow(d), prob=weight, replace=TRUE)]
-%  params <- c("alpha", "m", "N", "I")
-%  d <- melt(d, measure.vars=params, variable.name="parameter")
-%  f <- function (x) {
-%    dens <- density(x)
-%    hpd <- HPDinterval(mcmc(x))
-%    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
-%         hpd.max=hpd[,"upper"])
-%  }
-%  d <- d[,f(value), by=parameter]
-%  setkey(d, parameter)
-%
-%  m <- collect.metadata("../../simulations/abc-pa-peerdriven/abc/*")
-%  p <- fread(paste("bzcat", rownames(m)[m$sample_peer == 1]))
-%  p <- p[iter == max(iter)]
-%  p <- p[sample(1:nrow(p), prob=weight, replace=TRUE)]
-%  params <- c("alpha", "m", "N", "I")
-%  p <- melt(p, measure.vars=params, variable.name="parameter")
-%  f <- function (x) {
-%    dens <- density(x)
-%    hpd <- HPDinterval(mcmc(x))
-%    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
-%         hpd.max=hpd[,"upper"])
-%  }
-%  p <- p[,f(value), by=parameter]
-%  setkey(p, parameter)
-%  options(scipen=5)
-%@
-%
-%To test the effect of model misspecification, we simulated one network where
-%the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
-%other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
-%HPD] estimates for each parameter were: 
-%$\alpha$, 
-%  \Sexpr{round(d["alpha", map], 2)} 
-%  [\Sexpr{round(d["alpha", hpd.min], 2)} -
-%   \Sexpr{round(d["alpha", hpd.max], 2)}];
-%$I$,
-%  \Sexpr{d["I", round(map)]} 
-%  [\Sexpr{d["I", round(hpd.min)]} -
-%   \Sexpr{d["I", round(hpd.max)]}];
-%$m$,
-%  \Sexpr{d["m", floor(map)]} 
-%  [\Sexpr{d["m", floor(hpd.min)]} -
-%   \Sexpr{d["m", floor(hpd.max)]}];
-%$N$,
-%  \Sexpr{d["N", round(map)]} 
-%  [\Sexpr{d["N", round(hpd.min)]} -
-%   \Sexpr{d["N", round(hpd.max)]}].
-%The approximate posterior distributions for this simulation are shown in
-%\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
-%transmission tree in a peer-driven fashion, where the probability to sample a
-%node was twice as high if one of its peers had already been sampled. The
-%parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
-%$I$ = 2000. The estimated values were
-%$\alpha$, 
-%  \Sexpr{p["alpha", round(map, 2)]} 
-%  [\Sexpr{p["alpha", round(hpd.min, 2)]} -
-%   \Sexpr{p["alpha", round(hpd.max, 2)]}];
-%$I$,
-%  \Sexpr{p["I", floor(map)]} 
-%  [\Sexpr{p["I", floor(hpd.min)]} -
-%   \Sexpr{p["I", floor(hpd.max)]}];
-%$m$,
-%  \Sexpr{p["m", floor(map)]} 
-%  [\Sexpr{p["m", floor(hpd.min)]} -
-%   \Sexpr{p["m", floor(hpd.max)]}];
-%$N$,
-%  \Sexpr{p["N", floor(map)]} 
-%  [\Sexpr{p["N", floor(hpd.min)]} -
-%   \Sexpr{p["N", floor(hpd.max)]}].
-%The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
-%of these results were in line with estimates obtained on other simulated
-%datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
-%$\alpha$ was somewhat lower than typical.
-
 %\subsubsection*{Effect of parameters on power-law exponent}

### 2016-05-31 doc/thesis/methods_results.tex
More changes... i'm so unproductive

--- 2016-05-27 doc/thesis/methods_results.tex
@@
 Monte Carlo approximation to the target distribution. This situation is
-referred to as the ``collapse'' of the approximation, and is mitigated by the
-adaptive approach. A single parameter $\alpha$ (not to be confused with the
-\gls{BA} model parameter) controls the decay rate, with $\varepsilon_i$ being
-chosen to satisfy
+referred to as the ``collapse'' of the approximation, {\color{blue}\uline{also
+known as particle degeneracy} (see \cref{subsec:smc})} and is mitigated by the
+adaptive approach. A single parameter {\color{red}\sout{$\alpha$ (not to be
+confused with the \gls{BA} model parameter)}} controls the decay rate.
+{\color{blue}\uline{In the original paper of }\textcite{del2012adaptive}, \uline{the
+parameter is called $\alpha$, but to avoid confusion with the \gls{BA}
+parameter of the same name, we will refer to it here as $\alpha_{\ESS}$.}}
+The tolerance $\varepsilon_i$ is chosen to satisfy
 \[
-  \ESS(w_i) = \alpha \ESS(w_{i-1}).
+  \ESS(w_i) = \alpha_{\ESS} \ESS(w_{i-1}),
 \]
-Here, $w_i$ is the vector of weights at the $i$th step. Note that, since $w_i$
+where, $w_i$ is the vector of weights at the $i$th step. Note that, since $w_i$
 depends on $\varepsilon_i$, this equation solves for the updated weights and
@@
 \gls{m}, and \gls{N}. The objective of this experiment was to determine whether
-any of the four parameters might be identifiable from the shape of the
-transmission tree. TODO }
-
-\uline{In addition, a secondary objective of this section was to validate the
-use of the tree kernel as a distance measure for \gls{ABC} in our context. As
-discussed in} \cref{sec:abc}, \uline{the choice of distance function is
-extremely important for the accuracy of the \gls{ABC} approximation to the
-posterior. Therefore, we evaluated two additional tree statistics in the same
-manner as we evaluated the tree kernel (that is, by constructing and testing a
-classifier). First, we considered Sackin's index~\autocite{shao1990tree}, which
-measures the degree of unbalancedeness or asymmetry in a phylogeny} (see
-\cref{subsec:treeshape}). \uline{Sackin's index is widely used for
-characterizing phylogenies~\autocite{frost2013modelling} and has been
-demonstrated to vary between transmission trees simulated under different
-contact network types~\autocite{leventhal2012inferring}. Sackin's index does
-not take branch lengths into account, considering only the tree's topology. The
-other statistic we considered was the
-\gls{nltt}~\autocite{janzen2015approximate}, which compares two trees based on
-normalized distributions of their branching times} (see \cref{sec:treeshape}).
-\uline{In contrast with Sackin's index, the \gls{nltt} does not explicity
-consider the trees' topologies, but it does use their normalized branch
-lengths. While the \gls{nltt} is a newly developed statistic not yet in
-widespread use, the unnormalized \gls{ltt}~\autocite{nee1992tempo} was the
-basis of seminal early work extracting epidemiological information from
-phylodynamics~\autocite{holmes1995revealing}.}
-
-\uline{We expect the tree kernel to classify the \gls{BA} parameters more
-accurately than either Sackin's index or the \gls{nltt}, since the tree kernel
-takes both topology and branch lengths into account. }}
-
-This experiment involved a large number of variables which were varied
+any of the four parameters were identifiable from the shape of the transmission
+tree, as quantified by the tree kernel. Each of the \gls{BA} model parameters
+was varied one at a time, while holding the other parameters at fixed, known
+values. Contact networks were generated according to each set of parameter
+values, and transmission trees were simulated over the networks. We then
+evaluated how well a classifier based on the tree kernel could differentiate
+the trees simulated under distinct parameter values. If the classifier's
+cross-validation accuracy was high, this could be taken as an indication that
+the parameter in question was identifiable in the range of values considered. 
+A caveat of this preliminary analysis is that, since all parameters but one
+were held at known values, nothing could be said about the identifiability of
+\emph{combinations} of parameters; this issue will be explored later by jointly
+estimating all parameters with \gls{ABC}.}
+
+\uline{In addition to testing for identifiability, a secondary objective of
+this analysis was to validate the use of the tree kernel as a distance measure
+for \gls{ABC} in our context. As discussed in} \cref{sec:abc}, \uline{the
+choice of distance function is extremely important for the accuracy of the
+\gls{ABC} approximation to the posterior. Therefore, we evaluated two
+additional tree statistics in the same manner as we evaluated the tree kernel
+(that is, by constructing and testing a classifier). First, we considered
+Sackin's index~\autocite{shao1990tree}, which measures the degree of imbalance
+or asymmetry in a phylogeny} (see \cref{subsec:treeshape}). \uline{Sackin's
+index is widely used for characterizing
+phylogenies~\autocite{frost2013modelling} and has been demonstrated to vary
+between transmission trees simulated under different contact network
+types~\autocite{leventhal2012inferring}. Sackin's index does not take branch
+lengths into account, considering only the tree's topology. The other statistic
+we considered was the \gls{nltt}~\autocite{janzen2015approximate}, which
+compares two trees based on normalized distributions of their branching times}
+(see \cref{sec:treeshape}).  \uline{In contrast with Sackin's index, the
+\gls{nltt} does not explicity consider the trees' topologies, but it does use
+their normalized branch lengths. While the \gls{nltt} is a newly developed
+statistic not yet in widespread use, the unnormalized
+\gls{ltt}~\autocite{nee1992tempo} was the basis of seminal early work
+extracting epidemiological information from
+phylodynamics~\autocite{holmes1995revealing}. We expected the tree kernel to
+classify the \gls{BA} parameters more accurately than either Sackin's index or
+the \gls{nltt}, since the tree kernel takes both topology and branch lengths
+into account. }}
+
+This experiment involved a large number of variables that were varied
 combinatorially. For ease of exposition, we will describe a single experiment
@@
 
-The accuracy of classifiers $I$ varied significantly with the number of tips in
-the tree. For 100-tip trees, the average $R^2$ values were
+The accuracy of classifiers for $I$ varied significantly with the number of
+tips in the tree. For 100-tip trees, the average $R^2$ values were
   0.7,
@@
 however the results for $m = 3$ and $m = 4$ were similar
-(\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% HPD intervals are
-given in \cref{tab:abchpd}.
-
+(\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% \gls{HPD}
+intervals are given in \cref{tab:abchpd}.
+
+{\color{blue}
+\uline{Across all simulations, the median [\gls{IQR}] absolute errors of the parameter 
+estimates obtained with \software{netabc} were
+    0.11 
+    [0.03 - 
+    0.25]
+for \gls{alpha},
+    492 
+    [294 - 
+    782]
+for \gls{I},
+    1 
+    [0 - 
+    1]
+for \gls{m}, and
+    4153 
+    [3660 - 
+    4489]
+for \gls{N}. These errors comprised, respectively,
+    6\%,
+    11\%,
+    17\%, and
+    29\%
+of the regions of nonzero prior density. Average \gls{HPD} interval widths were
+    0.56,
+    1737,
+    3.06, and
+    12078,
+representing 
+    28\%,
+    39\%,
+    51\%, and
+    83\%
+of the nonzero prior density regions. Point estimates of \gls{I} were upwardly
+biased: \gls{I} was overestimated in
+    69
+out of 
+    72
+simulations
+    (4\%).
+The estimates for \gls{m} and \gls{N} were similar across all simulations
+(median [IQR] point estimates
+    3
+    [3 - 
+    3]
+and 
+    9153
+    [8660 - 
+    9489])
+regardless of the true values of any of the \gls{BA} parameters.}
+
+\uline{To analyse the effects of the true parameter values on the
+accuracy our estimates of \gls{alpha} and \gls{I}, we fitted one \glspl{GLM}
+for each of these two parameters, with error rate as the dependent variable and
+the true parameter values as independent variables. Since the estimates of
+\gls{m} and \gls{N} were roughly equal across all simulations}
+(\cref{fig:abcptm2,fig:abcptm3,fig:abcptm4}), \uline{\glspl{GLM} were not
+fitted for these parameters. The estimated coefficients are shown in
+}\cref{tab:glmalpha,tab:glmI}. \uline{For the parameters \gls{alpha}, \gls{I},
+and \gls{N}, the \gls{GLM} were fitted using the inverse link function. That
+is, if $p$ is the true value of the parameter and $\hat{p}$ is a random
+variable representing our estimate of the parameter, the \gls{GLM} posits a
+relationship of the form }
+\[
+    \E(|p-\hat{p}|) = (\beta_0 + \beta_\alpha \alpha + \beta_I I + \beta_m m)^{-1},
+\]
+\uline{where the $\beta$'s are coefficients to be fitted. If the true value
+\gls{alpha}, say, is increased by one, the \emph{inverse} of the expected
+absolute error will increase by $\beta_\alpha$. If $\beta_\alpha$ is positive,
+it means that the absolute error decreases as the true value of \gls{alpha}
+increases.}
+
+\uline{The \gls{GLM} analysis indicated that the error in estimates of
+\gls{alpha} decreased with larger true values of \gls{alpha} 
+    ($p {<}10^{-5}$)
+and \gls{m}
+    ($p =0.01$)
+but was not significatly affected by \gls{I}} (\cref{tab:glmalpha}). 
+\uline{Qualitatively, \gls{alpha} seemed to be only weakly identifiable between
+the values of 0 and 0.5} (\cref{fig:abcptm2}). \uline{The error in the
+estimated prevalence \gls{I} was slightly lower for smaller values of
+\gls{alpha}
+    ($p {<}10^{-5}$)
+and \gls{I}
+    ($p =0.05$),
+but was not significantly affected by the true value of \gls{m}} 
+(\cref{tab:glmI}).}
+
+{\color{red}\sout{
 The accuracy of the parameter estimates obtained with kernel-assisted ABC
@@
 %    pp(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"], eq=TRUE)$),
-{\color{blue}\uline{did not vary across the true values of $m$ or $I$ (both
-one-way ANOVA).}}
-
-Estimates for $I$ were relatively accurate, with point estimate errors of
+did not vary across the true values of $m$ or $I$ (both
+one-way ANOVA).}
+
+\sout{Estimates for $I$ were relatively accurate, with point estimate errors of
     492 
@@
 and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
-did not affect the estimates of $I$ (one-way ANOVA).
-
-The $m$ parameter was estimated correctly in
+did not affect the estimates of $I$ (one-way ANOVA).}
+
+\sout{The $m$ parameter was estimated correctly in
     37 \%
-of simulations {\color{red}\sout{barely better than random guessing}}.
-{\color{blue}\uline{However this was due to the estimated value being $m = 3$
-in the large majority of simulations (62 out of
-72), with one-third of the simulations having the true value $m =
-3$.}} The true values of the other parameters did not significantly affect the
-estimates of $m$ (both one-way ANOVA).
-
-Finally, the total number of nodes $N$ was consistently over-estimated by about
+of simulations barely better than random guessing. The true values of the other
+parameters did not significantly affect the
+estimates of $m$ (both one-way ANOVA).}
+
+\sout{Finally, the total number of nodes $N$ was consistently over-estimated by about
 a factor of two (error 4153
 [3660 - 4489] individuals). No parameters influenced the accuracy of the $N$
-estimates (all one-way ANOVA).
+estimates (all one-way ANOVA).}}
 
@@
 $\alpha$, 
-  1.1 
-  [0.6 -
-   1.15];
+  1.03 
+  [0.67 -
+   1.18];
 $I$,
-  1099 
-  [662 -
-   4455];
+  1474 
+  [511 -
+   2990];
 $m$,
@@
   [1 -
-   5];
+   4];
 $N$,
-  12791 
-  [3861-
-   14967].
+  9861 
+  [3710-
+   14977].
 The approximate posterior distributions for this simulation are shown in
@@
 $\alpha$, 
-  0.24 
+  0.3 
   [0 -
-   0.62];
+   0.63];
 $I$,
-  2384 
-  [1419 -
+  2449 
+  [1417 -
    3811];
@@
   [2 -
-   5];
+   4];
 $N$,
-  11565 
-  [2741 -
+  9132 
+  [2852 -
    14780].

### 2016-05-31 doc/thesis/appendix.Rnw
Reorganizing

@@
 @
+
+\chapter{Mathematical models, likelihood, and Bayesian inference}
+\label{chp:prelim}
+
+A \defn{mathematical model} is a formal description of a hypothesized
+relationship between some observed data, $x$ and outcomes $y$. A
+\defn{parametric} model defines a family of possible relationships between data
+and outcomes, parameterized by one or more numeric parameters $\theta$. A
+\defn{statistical} model describes the relationship between data and outcomes
+in terms of probabilities. Statistical models define, either explicitly or
+implicitly, the probability of observing $y$ given $x$ and, if the model is
+parametric, $\theta$. Note that it is entirely possible to have no data $x$,
+only observed outcomes $y$. In this case, a model would describe the process by
+which $y$ is generated.
+
+To illustrate these concepts, consider the well-known linear model. For
+clarity, we will restrict our attention to the case of one-dimensional data and
+outcomes where $x = \set{x_1, \ldots, x_n}$ and $y = \set{y_1, \ldots, y_n}$
+are vectors of real numbers. The linear model postulates that the outcomes are
+linearly related to the data, modulo some noise introduced by measurement
+error, environmental fluctuations, and other external factors. Formally, $y_i =
+\beta x_i + \varepsilon_i$, where $\beta$ is the slope of the linear
+relationship, and $\varepsilon_i$ is the error associated with measurement $i$.
+We can make this model a statistical one by hypothesizing a distribution for
+the error terms $\varepsilon_i$; most commonly, it is assumed that they are
+normally distributed with variance $\sigma$. In mathematical terms, $y_i \sim
+\beta x_i + \N(0, \sigma^2)$, where ``$\sim$'' means ``is distributed as''. We
+can see from this formulation that the model is parametric, with parameters
+$\theta$ = ($\beta$, $\sigma$). Moreover, we can write down the probability
+density $\pi$ of observing outcome $y_i$ given the parameters,
+\[
+  \pi(y \mid \beta, \sigma) = 
+  \prod_{i=1}^n f_{\N(0, \sigma^2)} (y_i - \beta x_i),
+\]
+where $f_{\N(0, \sigma^2)}$ is the probability density of the normal
+distribution with mean zero and variance $\sigma^2$. Note that we are treating
+the $x_i$ as fixed quantities and therefore have not conditioned the
+probability density on $x$. Also, we have assumed that all the $y_i$ are
+independent.
+
+For a general model, the probability density of $y$ given the parameters
+$\theta$ is also known as the \defn{likelihood}, written $\L$, of $\theta$.
+That is, $\L(\theta \mid y) = f(y \mid \theta)$ for the model's \gls{pdf} $f$.
+The higher the value of the likelihood, the more likely the observations $y$
+are under the model. Thus, the likelihood provides a natural criterion for
+fitting the model parameters: we want to pick $\theta$ such that the
+probability density of our observed outcomes $y$ is as high as possible. The
+parameters that optimize the likelihood are known as the \textit{\gls{ML}}
+estimates, denoted $\hat{\theta}$. That is,
+\[
+  \hat{\theta} = \argmax_\theta\; \L(\theta \mid y).
+\]
+\Gls{ML} estimation is usually performed with numerical optimization. In the
+simplest terms, many possible values for $\theta$ are examined, $\L(\theta \mid
+y)$ is calculated for each, and the parameters that produce the highest value
+are accepted. Many sophisticated numerical optimization methods exist, although
+they may not be guaranteed to find the true \gls{ML} estimates if the
+likelihood function is multi-modal.
+
+\Gls{ML} estimation makes use only of the data and outcomes to estimate the
+model parameters $\theta$. However, it is frequently the case that the
+investigator has some additional information or belief about what $\theta$ are
+likely to be. For example, in the linear regression case, the instrument used
+to measure the outcomes may have a well-known margin of error, or the sign of
+the slope may be obvious from previous experiments. The Bayesian approach to
+model fitting makes use of this information by codifying the investigator's
+beliefs as a \defn{prior distribution} on the parameters, denoted
+$\pi(\theta)$. Instead of considering only the likelihood, Bayesian inference
+focuses on the product of the likelihood and the prior, $f(y \mid \theta)
+\pi(\theta)$. Bayes' theorem tells us that this product is related to the
+\textit{posterior distribution} on $\theta$,
+\begin{align}
+  f(\theta \mid y) 
+    = \frac{f(y \mid \theta) \pi(\theta)}
+           {\int f(y \mid \theta) \pi(\theta) \d \theta}.
+  \label{eq:bayes}
+\end{align}
+In principle, $f(y \mid \theta) \pi(\theta)$ can be optimized numerically just
+like $\L(\theta \mid y)$, which would also optimize the posterior distribution.
+The resulting optimal parameters are called the \gls{MAP} estimates. However,
+from a Bayesian perspective, $\theta$ is not a fixed quantity to be estimated,
+but rather a random variable with an associated distribution (the posterior).
+Therefore, the \gls{MAP} estimate by itself is of limited value without
+associated statistics about the posterior distribution, such as the mean or
+credible intervals. Unfortunately, to calculate such statistics, it is
+necessary to evaluate the normalizing constant in the denominator of
+\cref{eq:bayes}, which is almost always an intractable integral.
+
+A popular method for circumventing the normalizing constant is the use of
+\gls{MCMC} to obtain a sample from the posterior distribution. \Gls{MCMC} works
+by defining a Markov chain {\color{red}\sout{whose states are indexed by
+possible model parameters. The transition probability from state $\theta_1$ to
+state $\theta_2$ is taken to be}} {\color{blue}\uline{on the space of possible
+model parameters. The transition density from parameters $\theta_1$ to
+$\theta_2$ is taken to be}}
+\[
+  \min\left(1, \frac{f(y \mid \theta_2) \pi(\theta_2) q(\theta_2, \theta_1)}
+                    {f(y \mid \theta_1) \pi(\theta_2) q(\theta_1, \theta_2)} \right),
+\]
+where $q(\theta, \theta')$ is a symmetric \defn{proposal distribution} used in
+the algorithm to generate the chain. The stationary distribution of this Markov
+chain is equal to the posterior distribution on $\theta$. Therefore, if a long
+enough random walk is performed on the chain, the distribution of states
+visited will be a Monte Carlo approximation of $f(\theta \mid y)$, from
+which we can calculate statistics of interest. Actually performing this random
+walk is straightforward and can be accomplished via the Metropolis-Hastings
+algorithm~\autocite{metropolis1953equation,hastings1970monte} (\cref{alg:mh}).
+
+\begin{algorithm}
+  \caption{Metropolis-Hastings algorithm for Markov chain Monte Carlo.}
+  \begin{algorithmic}
+    \State Draw $\theta$ according to the prior $\pi(\theta)$
+    \Loop
+      \State Propose $\theta'$ according to $q(\theta, \theta')$
+      \State Accept $\theta \gets \theta'$ with probability
+      $\min \left( 1, 
+       \dfrac{f(y \mid \theta') \pi(\theta') q(\theta', \theta)}
+             {f(y \mid \theta\phantom{'}) \pi(\theta\phantom{'}) q(\theta, \theta')}
+       \right)$
+    \EndLoop
+  \end{algorithmic}
+  \label{alg:mh}
+\end{algorithm}
 

### 2016-05-31 doc/thesis/intro.tex
Reorganizing

@@
 trees. In particular, for a network model with parameters $\theta$ and an input
-transmission tree $T$, we are interested in the posterior distribution}
+transmission tree $T$, we are interested in the \emph{posterior distribution}
 \begin{align}
@@
 \glsreset{ABC}
-
-\subsection{Model fitting}
-\label{subsec:mfit}
-
-A \defn{mathematical model} is a formal description of a hypothesized
-relationship between some observed data, $x$ and outcomes $y$. A
-\defn{parametric} model defines a family of possible relationships between data
-and outcomes, parameterized by one or more numeric parameters $\theta$. A
-\defn{statistical} model describes the relationship between data and outcomes
-in terms of probabilities. Statistical models define, either explicitly or
-implicitly, the probability of observing $y$ given $x$ and, if the model is
-parametric, $\theta$. Note that it is entirely possible to have no data $x$,
-only observed outcomes $y$. In this case, a model would describe the process by
-which $y$ is generated.
-
-To illustrate these concepts, consider the well-known linear model. For
-clarity, we will restrict our attention to the case of one-dimensional data and
-outcomes where $x = \set{x_1, \ldots, x_n}$ and $y = \set{y_1, \ldots, y_n}$
-are vectors of real numbers. The linear model postulates that the outcomes are
-linearly related to the data, modulo some noise introduced by measurement
-error, environmental fluctuations, and other external factors. Formally, $y_i =
-\beta x_i + \varepsilon_i$, where $\beta$ is the slope of the linear
-relationship, and $\varepsilon_i$ is the error associated with measurement $i$.
-We can make this model a statistical one by hypothesizing a distribution for
-the error terms $\varepsilon_i$; most commonly, it is assumed that they are
-normally distributed with variance $\sigma$. In mathematical terms, $y_i \sim
-\beta x_i + \N(0, \sigma^2)$, where ``$\sim$'' means ``is distributed as''. We
-can see from this formulation that the model is parametric, with parameters
-$\theta$ = ($\beta$, $\sigma$). Moreover, we can write down the probability
-density $\pi$ of observing outcome $y_i$ given the parameters,
-\[
-  \pi(y \mid \beta, \sigma) = 
-  \prod_{i=1}^n f_{\N(0, \sigma^2)} (y_i - \beta x_i),
-\]
-where $f_{\N(0, \sigma^2)}$ is the probability density of the normal
-distribution with mean zero and variance $\sigma^2$. Note that we are treating
-the $x_i$ as fixed quantities and therefore have not conditioned the
-probability density on $x$. Also, we have assumed that all the $y_i$ are
-independent.
-
-For a general model, the probability density of $y$ given the parameters
-$\theta$ is also known as the \defn{likelihood}, written $\L$, of $\theta$.
-That is, $\L(\theta \mid y) = f(y \mid \theta)$ for the model's \gls{pdf} $f$.
-The higher the value of the likelihood, the more likely the observations $y$
-are under the model. Thus, the likelihood provides a natural criterion for
-fitting the model parameters: we want to pick $\theta$ such that the
-probability density of our observed outcomes $y$ is as high as possible. The
-parameters that optimize the likelihood are known as the \textit{\gls{ML}}
-estimates, denoted $\hat{\theta}$. That is,
-\[
-  \hat{\theta} = \argmax_\theta\; \L(\theta \mid y).
-\]
-\Gls{ML} estimation is usually performed with numerical optimization. In the
-simplest terms, many possible values for $\theta$ are examined, $\L(\theta \mid
-y)$ is calculated for each, and the parameters that produce the highest value
-are accepted. Many sophisticated numerical optimization methods exist, although
-they may not be guaranteed to find the true \gls{ML} estimates if the
-likelihood function is multi-modal.
-
-\Gls{ML} estimation makes use only of the data and outcomes to estimate the
-model parameters $\theta$. However, it is frequently the case that the
-investigator has some additional information or belief about what $\theta$ are
-likely to be. For example, in the linear regression case, the instrument used
-to measure the outcomes may have a well-known margin of error, or the sign of
-the slope may be obvious from previous experiments. The Bayesian approach to
-model fitting makes use of this information by codifying the investigator's
-beliefs as a \defn{prior distribution} on the parameters, denoted
-$\pi(\theta)$. Instead of considering only the likelihood, Bayesian inference
-focuses on the product of the likelihood and the prior, $f(y \mid \theta)
-\pi(\theta)$. Bayes' theorem tells us that this product is related to the
-\textit{posterior distribution} on $\theta$,
-\begin{align}
-  f(\theta \mid y) 
-    = \frac{f(y \mid \theta) \pi(\theta)}
-           {\int f(y \mid \theta) \pi(\theta) \d \theta}.
-  \label{eq:bayes}
-\end{align}
-In principle, $f(y \mid \theta) \pi(\theta)$ can be optimized numerically just
-like $\L(\theta \mid y)$, which would also optimize the posterior distribution.
-The resulting optimal parameters are called the \gls{MAP} estimates. However,
-from a Bayesian perspective, $\theta$ is not a fixed quantity to be estimated,
-but rather a random variable with an associated distribution (the posterior).
-Therefore, the \gls{MAP} estimate by itself is of limited value without
-associated statistics about the posterior distribution, such as the mean or
-credible intervals. Unfortunately, to calculate such statistics, it is
-necessary to evaluate the normalizing constant in the denominator of
-\cref{eq:bayes}, which is almost always an intractable integral.
-
-A popular method for circumventing the normalizing constant is the use of
-\gls{MCMC} to obtain a sample from the posterior distribution. \Gls{MCMC} works
-by defining a Markov chain {\color{red}\sout{whose states are indexed by
-possible model parameters. The transition probability from state $\theta_1$ to
-state $\theta_2$ is taken to be}} {\color{blue}\uline{on the space of possible
-model parameters. The transition density from parameters $\theta_1$ to
-$\theta_2$ is taken to be}}
-\[
-  \min\left(1, \frac{f(y \mid \theta_2) \pi(\theta_2) q(\theta_2, \theta_1)}
-                    {f(y \mid \theta_1) \pi(\theta_2) q(\theta_1, \theta_2)} \right),
-\]
-where $q(\theta, \theta')$ is a symmetric \defn{proposal distribution} used in
-the algorithm to generate the chain. The stationary distribution of this Markov
-chain is equal to the posterior distribution on $\theta$. Therefore, if a long
-enough random walk is performed on the chain, the distribution of states
-visited will be a Monte Carlo approximation of $f(\theta \mid y)$, from
-which we can calculate statistics of interest. Actually performing this random
-walk is straightforward and can be accomplished via the Metropolis-Hastings
-algorithm~\autocite{metropolis1953equation,hastings1970monte} (\cref{alg:mh}).
-
-\begin{algorithm}
-  \caption{Metropolis-Hastings algorithm for Markov chain Monte Carlo.}
-  \begin{algorithmic}
-    \State Draw $\theta$ according to the prior $\pi(\theta)$
-    \Loop
-      \State Propose $\theta'$ according to $q(\theta, \theta')$
-      \State Accept $\theta \gets \theta'$ with probability
-      $\min \left( 1, 
-       \dfrac{f(y \mid \theta') \pi(\theta') q(\theta', \theta)}
-             {f(y \mid \theta\phantom{'}) \pi(\theta\phantom{'}) q(\theta, \theta')}
-       \right)$
-    \EndLoop
-  \end{algorithmic}
-  \label{alg:mh}
-\end{algorithm}
+{\color{blue}\uline{\acrlong{SMC}, and the \gls{SMC} sampler, were developed
+for sampling from distributions which can be evaluated up to a normalizing
+constant. TODO }
+
+\uline{This section is organized as follows. First, we formalize the concept of a
+mathematical model and discuss what it means to ``fit'' such models to data 
+and how this is typically accomplished. Second, we introduce \gls{ABC}, which
+was designed for situations like the one we face in this work where the most
+commonly used means of fitting models cannot be applied. Finally, we describe
+algorithms which can be used to implement \gls{ABC}, including a variant of
+\gls{SMC}.}}
 
@@
 
-%\subsection{Convergence properties}
-%
-%The convergence of \gls{ABC} to the true posterior distribution of interest
-%depends on two components: the convergence of the underlying algorithm to the
-%\gls{ABC} target distribution, and the resemblance of this target distribution
-%to the true
-%posterior~\autocite{marin2012approximate,fearnhead2012constructing}. We shall
-%address these components one at a time. Rejection \gls{ABC} and
-%\gls{ABC}-\gls{MCMC} use, respectively, rejection sampling and \gls{MCMC} to
-%sample from the target distributions. Both of these methods are
-%well-established and their convergence properties have been extensively
-%studied. Since we do not apply rejection or \gls{MCMC} in this work, we shall
-%}restrict our attention to the convergence properties of \gls{SMC}.
-
 \section{Summary}
@@
   topics define the algorithm and statistical framework that our inference
-method will use. }}
+method will use. }
+
+\uline{Having introduced transmission trees and contact network models, we are
+now in a position to argue that approximate Bayesian computation is an
+appropriate tool to apply towards our main objective: fitting network models to
+phylogenetic data. Let $M$ be a contact network model having parameters
+$\theta$.
+
+}}

### 2016-05-31 doc/thesis/methods_results.Rnw
Reorganizing

@@
 {\color{blue}\uline{
-The chapter is organized as follows. One section is devoted to each of our
+This chapter is organized as follows. One section is devoted to each of our
 three research aims: the development of a method for inference of contact
@@
 
+\subsection{Justification for approach}
+
+We present here a mathematical justification for the use of \gls{ABC} for the
+problem at hand, as opposed to more frequently-used approaches for fitting
+mathematical models. Consider a contact network model with parameters $\theta$,
+and an estimated transmission tree $T$. TODO
+
 \section{Analysis of \acrlong{BA} model with synthetic data}

### 2016-05-31 doc/thesis/objective.tex
Reorganizing

@@
 \newcommand{\Nu}{\mathcal{N}}
+
+{\color{blue}\uline{In this work, we take a Bayesian approach: our goal is to
+estimate the posterior distribution on model parameters given our data,}
+\[
+    \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)},
+\]
+\uline{where $\Pr(T \mid \theta)$ is the likelihood, $\Pr(\theta)$ is the prior,
+and $\Pr(T)$ is the marginal probability of $T$ which acts as a normalizing
+constant on the posterior (see \cref{chp:prelim} \uline{for a review of
+mathematical modeling and Bayesian inference, including definitions of these
+concepts).
+}}
 
@@
 {\color{blue}\uline{
-In a Bayesian setting, our goal is to estimate the posterior distribution on
-model parameters given our data, $\Pr(\theta \mid T)$. Bayes' theorem relates
-the posterior distribution to the likelihood,}
-\[
-  \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)}.
-\]
-\uline{For all but the simplest mathematical models, the calculation of the
-normalizing constant $\Pr(T)$ is an intractable problem. What we have argued in
+\uline{
+For all but the simplest mathematical models, the calculation of the
+normalizing constant $\Pr(T)$ is an intractable problem. 
+What we have argued in
 the previous paragraph is that calculating $\Pr(T \mid \theta)$ for a single
@@
 functions have been demonstrated to work well as distance functions in
-\gls{ABC}~\autocite{park2015k2}}}. \Gls{ABC} can be implemented with \gls{SMC},
-which has several advantages over other
-algorithms~\autocite{mckinley2009inference}. A recently-developed adaptive
-algorithm requiring minimal tuning on the part of the user makes \gls{SMC} an
-even more attractive approach~\autocite{del2012adaptive}. In summary, our
-method to infer contact network parameters will combine the following:
-stochastic simulation of epidemics on networks, the tree kernel, and adaptive
-\gls{ABC}-\gls{SMC}. {\color{blue}\uline{ Our method will expand on the
-framework developed by~\autocite{poon2015phylodynamic}, who combined \gls{ABC}
-with the tree kernel to infer parameters of population genetic models from
-viral phylogenies using \gls{MCMC}. }}
+\gls{ABC}~\autocite{park2015k2}}}. \Gls{ABC} can be implemented with several
+algorithms, but \gls{SMC} has advantages over others,
+{\color{blue}\uline{including improved accuracy in low-density regions and
+parallelizability}}~\autocite{mckinley2009inference}. A recently-developed
+adaptive algorithm requiring minimal tuning on the part of the user makes
+\gls{SMC} an even more attractive approach~\autocite{del2012adaptive}. In
+summary, our method to infer contact network parameters will combine the
+following: stochastic simulation of epidemics on networks, the tree kernel, and
+adaptive \gls{ABC}-\gls{SMC}. {\color{blue}\uline{ Our method will expand on
+the framework developed by~\autocite{poon2015phylodynamic}, who combined
+\gls{ABC} with the tree kernel to infer parameters of population genetic models
+from viral phylogenies using \gls{MCMC}. }}
 

### 2016-06-01 doc/thesis/appendix.Rnw
moved objective stuff to methods and expanded

@@
 \chapter{Mathematical models, likelihood, and Bayesian inference}
-\label{chp:prelim}
+\label[appendix]{chp:prelim}
 
@@
 
-\chapter{GLM results}
-
-<<glm_tables, echo=FALSE, warning=FALSE, error=FALSE, results="asis">>=
-    format.glm <- function (df) {
-        df[Parameter == "alpha", Parameter := "$\\alpha$"]
-        df[,p := paste0("$", sapply(p, pp, eq=FALSE), "$")]
-        df[,Estimate := paste0("$", sapply(Estimate, pp, eq=FALSE, thres=NA), "$")]
-        df[,`Standard error` := paste0("$", sapply(`Standard error`, pp, eq=FALSE, thres=NA), "$")]
-        setnames(df, "p", "$p$-value")
-        df
-    }
-
-    alpha.glm <- format.glm(alpha.glm)
-    I.glm <- format.glm(I.glm)
-    #m.glm <- format.glm(m.glm)
-    #N.glm <- format.glm(N.glm)
-
-    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-                        "$\\alpha$ to true values of BA parameters.")
-    capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
-                  "inverse link function. Coefficients are interpretable as",
-                  "additive effects on the inverse of the mean error.")
-    print(xtable(alpha.glm, label="tab:glmalpha", caption=c(capt, capt.short)),
-          include.rownames=FALSE, sanitize.text.function=identity,
-          sanitize.colnames.function=identity)
-
-    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-                        "$I$ to true values of BA parameters.")
-    capt <- paste(capt.short, "GLM was fitted with a gaussian distribution and",
-                  "inverse link function. Coefficients are interpretable as",
-                  "additive effects on the inverse of the mean error.")
-    print(xtable(I.glm, label="tab:glmI", caption=c(capt, capt.short)),
-          include.rownames=FALSE, sanitize.text.function=identity,
-          sanitize.colnames.function=identity)
-
-    #capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-    #                    "$m$ to true values of BA parameters.")
-    #capt <- paste(capt.short, "GLM was fitted with a Poisson distribution and",
-    #              "log link function. Coefficients are interpretable as",
-    #              "additive effects on the exponential of the mean error.")
-    #print(xtable(m.glm, label="tab:glmm", caption=c(capt, capt.short)),
-    #      include.rownames=FALSE, sanitize.text.function=identity,
-    #      sanitize.colnames.function=identity)
-
-    #capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-    #                    "$N$ to true values of BA parameters.")
-    #capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
-    #              "inverse link function. Coefficients are interpretable as",
-    #              "additive effects on the inverse of the mean error.")
-    #print(xtable(N.glm, label="tab:glmN", caption=c(capt, capt.short)),
-    #      include.rownames=FALSE, sanitize.text.function=identity,
-    #      sanitize.colnames.function=identity)
-@
+%\chapter{GLM results}
+%<<glm_tables, echo=FALSE, warning=FALSE, error=FALSE, results="asis">>=
+%    format.glm <- function (df) {
+%        df[Parameter == "alpha", Parameter := "$\\alpha$"]
+%        df[,p := paste0("$", sapply(p, pp, eq=FALSE), "$")]
+%        df[,Estimate := paste0("$", sapply(Estimate, pp, eq=FALSE, thres=NA), "$")]
+%        df[,`Standard error` := paste0("$", sapply(`Standard error`, pp, eq=FALSE, thres=NA), "$")]
+%        setnames(df, "p", "$p$-value")
+%        df
+%    }
+%
+%    alpha.glm <- format.glm(alpha.glm)
+%    I.glm <- format.glm(I.glm)
+%    #m.glm <- format.glm(m.glm)
+%    #N.glm <- format.glm(N.glm)
+%
+%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+%                        "$\\alpha$ to true values of BA parameters.")
+%    capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
+%                  "inverse link function. Coefficients are interpretable as",
+%                  "additive effects on the inverse of the mean error.")
+%    print(xtable(alpha.glm, label="tab:glmalpha", caption=c(capt, capt.short)),
+%          include.rownames=FALSE, sanitize.text.function=identity,
+%          sanitize.colnames.function=identity)
+%
+%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+%                        "$I$ to true values of BA parameters.")
+%    capt <- paste(capt.short, "GLM was fitted with a gaussian distribution and",
+%                  "inverse link function. Coefficients are interpretable as",
+%                  "additive effects on the inverse of the mean error.")
+%    print(xtable(I.glm, label="tab:glmI", caption=c(capt, capt.short)),
+%          include.rownames=FALSE, sanitize.text.function=identity,
+%          sanitize.colnames.function=identity)
+%
+%    #capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+%    #                    "$m$ to true values of BA parameters.")
+%    #capt <- paste(capt.short, "GLM was fitted with a Poisson distribution and",
+%    #              "log link function. Coefficients are interpretable as",
+%    #              "additive effects on the exponential of the mean error.")
+%    #print(xtable(m.glm, label="tab:glmm", caption=c(capt, capt.short)),
+%    #      include.rownames=FALSE, sanitize.text.function=identity,
+%    #      sanitize.colnames.function=identity)
+%
+%    #capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+%    #                    "$N$ to true values of BA parameters.")
+%    #capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
+%    #              "inverse link function. Coefficients are interpretable as",
+%    #              "additive effects on the inverse of the mean error.")
+%    #print(xtable(N.glm, label="tab:glmN", caption=c(capt, capt.short)),
+%    #      include.rownames=FALSE, sanitize.text.function=identity,
+%    #      sanitize.colnames.function=identity)
+%@

### 2016-06-01 doc/thesis/appendix.tex
moved objective stuff to methods and expanded

--- 2016-05-17 doc/thesis/appendix.tex
@@
+\chapter{Mathematical models, likelihood, and Bayesian inference}
+\label[appendix]{chp:prelim}
+
+A \defn{mathematical model} is a formal description of a hypothesized
+relationship between some observed data, $x$ and outcomes $y$. A
+\defn{parametric} model defines a family of possible relationships between data
+and outcomes, parameterized by one or more numeric parameters $\theta$. A
+\defn{statistical} model describes the relationship between data and outcomes
+in terms of probabilities. Statistical models define, either explicitly or
+implicitly, the probability of observing $y$ given $x$ and, if the model is
+parametric, $\theta$. Note that it is entirely possible to have no data $x$,
+only observed outcomes $y$. In this case, a model would describe the process by
+which $y$ is generated.
+
+To illustrate these concepts, consider the well-known linear model. For
+clarity, we will restrict our attention to the case of one-dimensional data and
+outcomes where $x = \set{x_1, \ldots, x_n}$ and $y = \set{y_1, \ldots, y_n}$
+are vectors of real numbers. The linear model postulates that the outcomes are
+linearly related to the data, modulo some noise introduced by measurement
+error, environmental fluctuations, and other external factors. Formally, $y_i =
+\beta x_i + \varepsilon_i$, where $\beta$ is the slope of the linear
+relationship, and $\varepsilon_i$ is the error associated with measurement $i$.
+We can make this model a statistical one by hypothesizing a distribution for
+the error terms $\varepsilon_i$; most commonly, it is assumed that they are
+normally distributed with variance $\sigma$. In mathematical terms, $y_i \sim
+\beta x_i + \N(0, \sigma^2)$, where ``$\sim$'' means ``is distributed as''. We
+can see from this formulation that the model is parametric, with parameters
+$\theta$ = ($\beta$, $\sigma$). Moreover, we can write down the probability
+density $\pi$ of observing outcome $y_i$ given the parameters,
+\[
+  \pi(y \mid \beta, \sigma) = 
+  \prod_{i=1}^n f_{\N(0, \sigma^2)} (y_i - \beta x_i),
+\]
+where $f_{\N(0, \sigma^2)}$ is the probability density of the normal
+distribution with mean zero and variance $\sigma^2$. Note that we are treating
+the $x_i$ as fixed quantities and therefore have not conditioned the
+probability density on $x$. Also, we have assumed that all the $y_i$ are
+independent.
+
+For a general model, the probability density of $y$ given the parameters
+$\theta$ is also known as the \defn{likelihood}, written $\L$, of $\theta$.
+That is, $\L(\theta \mid y) = f(y \mid \theta)$ for the model's \gls{pdf} $f$.
+The higher the value of the likelihood, the more likely the observations $y$
+are under the model. Thus, the likelihood provides a natural criterion for
+fitting the model parameters: we want to pick $\theta$ such that the
+probability density of our observed outcomes $y$ is as high as possible. The
+parameters that optimize the likelihood are known as the \textit{\gls{ML}}
+estimates, denoted $\hat{\theta}$. That is,
+\[
+  \hat{\theta} = \argmax_\theta\; \L(\theta \mid y).
+\]
+\Gls{ML} estimation is usually performed with numerical optimization. In the
+simplest terms, many possible values for $\theta$ are examined, $\L(\theta \mid
+y)$ is calculated for each, and the parameters that produce the highest value
+are accepted. Many sophisticated numerical optimization methods exist, although
+they may not be guaranteed to find the true \gls{ML} estimates if the
+likelihood function is multi-modal.
+
+\Gls{ML} estimation makes use only of the data and outcomes to estimate the
+model parameters $\theta$. However, it is frequently the case that the
+investigator has some additional information or belief about what $\theta$ are
+likely to be. For example, in the linear regression case, the instrument used
+to measure the outcomes may have a well-known margin of error, or the sign of
+the slope may be obvious from previous experiments. The Bayesian approach to
+model fitting makes use of this information by codifying the investigator's
+beliefs as a \defn{prior distribution} on the parameters, denoted
+$\pi(\theta)$. Instead of considering only the likelihood, Bayesian inference
+focuses on the product of the likelihood and the prior, $f(y \mid \theta)
+\pi(\theta)$. Bayes' theorem tells us that this product is related to the
+\textit{posterior distribution} on $\theta$,
+\begin{align}
+  f(\theta \mid y) 
+    = \frac{f(y \mid \theta) \pi(\theta)}
+           {\int f(y \mid \theta) \pi(\theta) \d \theta}.
+  \label{eq:bayes}
+\end{align}
+In principle, $f(y \mid \theta) \pi(\theta)$ can be optimized numerically just
+like $\L(\theta \mid y)$, which would also optimize the posterior distribution.
+The resulting optimal parameters are called the \gls{MAP} estimates. However,
+from a Bayesian perspective, $\theta$ is not a fixed quantity to be estimated,
+but rather a random variable with an associated distribution (the posterior).
+Therefore, the \gls{MAP} estimate by itself is of limited value without
+associated statistics about the posterior distribution, such as the mean or
+credible intervals. Unfortunately, to calculate such statistics, it is
+necessary to evaluate the normalizing constant in the denominator of
+\cref{eq:bayes}, which is almost always an intractable integral.
+
+A popular method for circumventing the normalizing constant is the use of
+\gls{MCMC} to obtain a sample from the posterior distribution. \Gls{MCMC} works
+by defining a Markov chain {\color{red}\sout{whose states are indexed by
+possible model parameters. The transition probability from state $\theta_1$ to
+state $\theta_2$ is taken to be}} {\color{blue}\uline{on the space of possible
+model parameters. The transition density from parameters $\theta_1$ to
+$\theta_2$ is taken to be}}
+\[
+  \min\left(1, \frac{f(y \mid \theta_2) \pi(\theta_2) q(\theta_2, \theta_1)}
+                    {f(y \mid \theta_1) \pi(\theta_2) q(\theta_1, \theta_2)} \right),
+\]
+where $q(\theta, \theta')$ is a symmetric \defn{proposal distribution} used in
+the algorithm to generate the chain. The stationary distribution of this Markov
+chain is equal to the posterior distribution on $\theta$. Therefore, if a long
+enough random walk is performed on the chain, the distribution of states
+visited will be a Monte Carlo approximation of $f(\theta \mid y)$, from
+which we can calculate statistics of interest. Actually performing this random
+walk is straightforward and can be accomplished via the Metropolis-Hastings
+algorithm~\autocite{metropolis1953equation,hastings1970monte} (\cref{alg:mh}).
+
+\begin{algorithm}
+  \caption{Metropolis-Hastings algorithm for Markov chain Monte Carlo.}
+  \begin{algorithmic}
+    \State Draw $\theta$ according to the prior $\pi(\theta)$
+    \Loop
+      \State Propose $\theta'$ according to $q(\theta, \theta')$
+      \State Accept $\theta \gets \theta'$ with probability
+      $\min \left( 1, 
+       \dfrac{f(y \mid \theta') \pi(\theta') q(\theta', \theta)}
+             {f(y \mid \theta\phantom{'}) \pi(\theta\phantom{'}) q(\theta, \theta')}
+       \right)$
+    \EndLoop
+  \end{algorithmic}
+  \label{alg:mh}
+\end{algorithm}
+
 \chapter{Additional plots}
@@
 %\chapter{GLM results}
-%
 %<<glm_tables, echo=FALSE, warning=FALSE, error=FALSE, results="asis">>=
@@
 %        df[,p := paste0("$", sapply(p, pp, eq=FALSE), "$")]
+%        df[,Estimate := paste0("$", sapply(Estimate, pp, eq=FALSE, thres=NA), "$")]
+%        df[,`Standard error` := paste0("$", sapply(`Standard error`, pp, eq=FALSE, thres=NA), "$")]
 %        setnames(df, "p", "$p$-value")
@@
 %    I.glm <- format.glm(I.glm)
-%    m.glm <- format.glm(m.glm)
-%    N.glm <- format.glm(N.glm)
+%    #m.glm <- format.glm(m.glm)
+%    #N.glm <- format.glm(N.glm)
 %
@@
 %                        "$\\alpha$ to true values of BA parameters.")
-%    capt <- paste(capt.short, "GLM was fitted with a Gamma distribution and",
+%    capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
 %                  "inverse link function. Coefficients are interpretable as",
@@
 %                        "$I$ to true values of BA parameters.")
-%    capt <- paste(capt.short, "GLM was fitted with a Gamma distribution and",
+%    capt <- paste(capt.short, "GLM was fitted with a gaussian distribution and",
 %                  "inverse link function. Coefficients are interpretable as",
@@
 %
-%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-%                        "$m$ to true values of BA parameters.")
-%    capt <- paste(capt.short, "GLM was fitted with a Poisson distribution and",
-%                  "identity link function. Coefficients are interpretable as",
-%                  "additive effects on the exponential of the mean error.")
-%    print(xtable(m.glm, label="tab:glmm", caption=c(capt, capt.short)),
-%          include.rownames=FALSE, sanitize.text.function=identity,
-%          sanitize.colnames.function=identity)
+%    #capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+%    #                    "$m$ to true values of BA parameters.")
+%    #capt <- paste(capt.short, "GLM was fitted with a Poisson distribution and",
+%    #              "log link function. Coefficients are interpretable as",
+%    #              "additive effects on the exponential of the mean error.")
+%    #print(xtable(m.glm, label="tab:glmm", caption=c(capt, capt.short)),
+%    #      include.rownames=FALSE, sanitize.text.function=identity,
+%    #      sanitize.colnames.function=identity)
 %
-%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-%                        "$N$ to true values of BA parameters.")
-%    capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
-%                  "identity link function. Coefficients are interpretable as",
-%                  "additive effects on the inverse of the mean error.")
-%    print(xtable(N.glm, label="tab:glmN", caption=c(capt, capt.short)),
-%          include.rownames=FALSE, sanitize.text.function=identity,
-%          sanitize.colnames.function=identity)
+%    #capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+%    #                    "$N$ to true values of BA parameters.")
+%    #capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
+%    #              "inverse link function. Coefficients are interpretable as",
+%    #              "additive effects on the inverse of the mean error.")
+%    #print(xtable(N.glm, label="tab:glmN", caption=c(capt, capt.short)),
+%    #      include.rownames=FALSE, sanitize.text.function=identity,
+%    #      sanitize.colnames.function=identity)
 %@

### 2016-06-01 doc/thesis/intro.tex
moved objective stuff to methods and expanded

@@
 \subsection{Transmission trees}
+\label{subsec:tt}
 
@@
 trees. In particular, for a network model with parameters $\theta$ and an input
-transmission tree $T$, we are interested in the \emph{posterior distribution}
+transmission tree $T$, we are interested in the \emph{posterior distribution}}
 \begin{align}
@@
 
-\uline{ \Gls{SMC} is also known as the \emph{particle filter}. Rather than
+\uline{\Gls{SMC} is also known as the \emph{particle filter}. Rather than
 sampling points one at a time from the target distribution, \gls{SMC} considers
@@
 applied to the particles in such a way that they converge, collectively, to a
-sample from the target. }
-
-\uline{In this work, the distribution of interest is the posterior distribution
-\ref{eq:post}. The particles are particular values of the parameters $\theta$
-of the contact network model being studied. If we were taking a typical
-Bayesian Monte Carlo approach to this problem, the particles would end up
-weighted by their posterior probability and distributed in such a way that the
-weighted population was a reasonable representation of $\Pr(\theta \mid T)$.
-This turns out to be not straightforward in our case, as we will address in
-}\cref{sec:abc}. \uline{For now, nothing is lost by \ldots}}
+sample from the target.}
+
+\uline{In this work, the distribution of interest is the posterior distribution}
+\ref{eq:post}. \uline{The particles are particular values of the parameters
+$\theta$ of the contact network model being studied. If we were taking a
+typical Bayesian Monte Carlo approach to this problem, the particles would end
+up weighted by their posterior probability and distributed in such a way that
+the weighted population was a reasonable representation of $\Pr(\theta \mid
+T)$. This turns out to be not straightforward in our case, as we will address
+in} \cref{sec:abc}. \uline{For now, nothing is lost by \ldots}}
 
@@
   topics define the algorithm and statistical framework that our inference
-method will use. }
-
-\uline{Having introduced transmission trees and contact network models, we are
-now in a position to argue that approximate Bayesian computation is an
-appropriate tool to apply towards our main objective: fitting network models to
-phylogenetic data. Let $M$ be a contact network model having parameters
-$\theta$.
-
-}}
+method will use. }}

### 2016-06-01 doc/thesis/methods_results.Rnw
moved objective stuff to methods and expanded

--- 2016-05-31 doc/thesis/methods_results.Rnw
@@
 
+\subsection{Justification for approach}
+\label{subsec:just}
+
+{\color{blue}\uline{We present here a non-rigorous justification for the use of
+\gls{ABC} for the problem at hand, as opposed to more frequently-used
+approaches for fitting mathematical models} (see \cref{chp:prelim}).
+\uline{Consider a contact network model with parameters $\theta$, and an
+estimated transmission tree $T$. Taking a Bayesian approach, our aim is to
+obtain a sample from the posterior distribution on the model's parameters given
+our data,}
+\[
+    \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)}.
+\]
+\uline{For all but the simplest models, the normalizing constant $\Pr(T)$ is an
+intractable integral. What we shall argue here is that, in contrast to most
+commonly studied mathematical models, the likelihood $\Pr(T \mid \theta)$ is
+also likely to be intractible in our case.}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{transtree}
+    \caption[
+        Illustration of an estimated transmission tree without labels and two
+        possible underlying complete transmission trees with labels.
+    ]{
+        Illustration of an estimated transmission tree without labels (left)
+        and two possible underlying complete transmission trees with labels
+        (right). In the top right scenario, the epidemic began with node $a$
+        who transmitted first to $b$ and then to $c$. In the bottom right
+        scenario, $b$ was the index case; $b$ infected $c$, who went on to
+        infect $a$. A transmission tree estimated from a viral phylogeny would
+        have the same topology and tip labels in both cases.
+    }
+    \label{fig:tt}
+\end{figure}
+
+\uline{As discussed in} \cref{subsec:tt}, \uline{the internal nodes of
+transmission trees represent transmission events, and are labelled with the
+donor in the associated transmission pair. However, when we estimate a
+transmission tree from viral sequence data, we generally only know the labels
+of the tips of the tree, not the labels of the internal nodes. In viral
+phylogenies, the transmissions are at least partially preserved through the
+evolutionary relationships among the viruses, but the directionality of those
+transmissions is unknown. Thus, a single estimated transmission tree can
+correspond to many possible pathways of the epidemic through the network.
+}\Cref{fig:tt} \uline{illustrates this concept for a simple transmission tree
+with three tips. When calculating a likelihood given a transmission tree, we
+must sum over all possible labellings of the internal nodes. Let $\Nu$ be the
+set of such labellings. Then}
+\begin{align}
+    \Pr(T \mid \theta) = \sum{\nu \in \Nu} \Pr(T, \nu \mid \theta).
+    \label{eq:lik1}
+\end{align}
+\uline{A contact network model assigns a probability to each possible contact
+network. Transmission trees are realized over particular contact networks, not
+over the model itself. Therefore, we must also sum over all contact networks
+which could be generated by the model. Let $\G$ be the set of all possible
+contact networks. Summing } \cref{eq:lik1} \uline{over $\G$ gives}
+\begin{align}
+    \Pr(T \mid \theta) = \sum_{G \in \G} \sum{\nu \in \Nu} 
+        \Pr(T, \nu \mid G, \theta) \Pr(G \mid \theta).
+    \label{eq:lik2}
+\end{align}
+\uline{This can be simplified somewhat by noticing that, given a specific
+contact network, the labelled transmission tree depends only on that network
+and not on the model that generated it. That is, $\Pr(T, \nu \mid G, \theta) =
+\Pr(T, \nu \mid G)$, and}
+\begin{align}
+    \Pr(T \mid \theta) = \sum_{G \in \G} \Pr(G \mid \theta) \sum{\nu \in \Nu} 
+        \Pr(T, \nu \mid G)
+    \label{eq:lik3}
+\end{align}
+\uline{Under the assumption that both transmission and removal are Poisson
+processes, calculating $\Pr(T, \nu \mid G)$ can be accomplished by a
+straightforward modification of the Gillespie simulation algorithm} 
+(\cref{alg:nettree}). \uline{Rather than choosing transmission or removal
+events according to their probabilities, the events would be deterministically
+chosen based on the transmission tree and the probabilities of each event
+would be multiplied together. Assuming efficient data structures for storing
+lists of nodes and edges, the complexity of this calculation would be $O(|T|)$.
+The number of possible labellings of internal nodes of $T$ is easily seen to be
+$2^{(|T|-1)/2}$ by noticing that each of the $(|T|-1)/2$ internal nodes must be
+labelled with the same label as either its right child or its left child.
+Although exponential calculations of this nature can are often be simplified on 
+trees using dynamic programming} (\eg \autocite{pupko2000fast}), \uline{it
+cannot be straightforwardly applied in this case because the subtrees'
+probabilities depend on the existing epidemic progress (their parents and
+siblings). Hence, calculating the inner sum over labels may take time
+$O(2^{(|T|-1)/2})$.}
+
+\uline{The outer sum, over all contact networks, is also difficult to evaluate
+in general. There are $2^{N(N-1)}$ directed graphs on $N$
+nodes~\autocite{harary2014graphical}. There must be at least as many nodes in
+the contact network as the number of tips in the tree, which is $(|T|+1)/2$. Of
+course, it is very likely that there are more nodes in the network than
+observed tips because some individuals are never infected and/or some infected
+individuals are never sampled. The complexity of calculating $\Pr(G \mid
+\theta)$ is obviously dependent on the particular model being investigated. For
+the \gls{BA} model, we might have to sum over all possible orders in which the
+nodes could be added, and all assignments of edges to the nodes which generated
+them. However, even in the unlikely case that calculating $\Pr(G \mid \theta)$
+can be done in constant time, the sum} \ref{eq:lik3} \uline{still has at least
+$O(2^{|T|^2})$ terms.}
+
+\uline{We have shown that both the normalizing constant $\Pr(T)$ and the
+likelihood $\Pr(T \mid \theta)$ are likely computationally prohibitive to
+calculate. If this is the case, the problem of fitting contact network models
+to phylogenies seems to be of the \emph{doubly-intractible}
+type~\autocite{murray2012mcmc}, which would imply that these models are not
+amenable to neither \gls{ML} nor Bayesian inference techniques. Although both
+methods are able to cope with an intractible normalizing constant (for example,
+by local search for \gls{ML} or Bayesian \gls{MCMC}), neither can avoid the
+intractable likelihood calculations. This justifies the use of \gls{ABC}, which
+is a likelihood-free method.}
+
+\uline{We have not proven here that} \cref{eq:lik3} \uline{is impossible to
+calculate in polynomial time - it could be possible to algebraically simplify
+the sum into a tractable expression. Furthermore, under certain models, a large
+proportion of $\G$ may have zero probability, which would enable the
+simplification of the outer sum on a model-specific basis. It should also be
+noted that extensions of Bayesian \gls{MCMC} have been developed for
+doubly-intractable problems~\autocite{liang2010double, murray2012mcmc}, which
+might be adaptable to the problem at hand. These have not been as widely used
+as \gls{ABC}, nor are they as easily parallelizable as \gls{SMC}. }}
+
 \section{Analysis of \acrlong{BA} model with synthetic data}
@@
 it means that the absolute error decreases as the true value of \gls{alpha}
-increases.}
+increases.}}
 
@@
 
-\uline{The \gls{GLM} analysis indicated that the error in estimates of
+<<glm_tables, echo=FALSE, warning=FALSE, error=FALSE, results="asis">>=
+    source("global.R")
+    format.glm <- function (df) {
+        df.copy <- copy(df)
+        df.copy[Parameter == "alpha", Parameter := "$\\alpha$"]
+        df.copy[,p := paste0("$", sapply(p, pp, eq=FALSE), "$")]
+        df.copy[,Estimate := paste0("$", sapply(Estimate, pp, eq=FALSE, thres=NA), "$")]
+        df.copy[,`Standard error` := paste0("$", sapply(`Standard error`, pp, eq=FALSE, thres=NA), "$")]
+        setnames(df.copy, "p", "$p$-value")
+        df.copy
+    }
+
+    alpha.glm.tbl <- format.glm(alpha.glm)
+    I.glm.tbl <- format.glm(I.glm)
+
+    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+                        "$\\alpha$ to true values of BA parameters.")
+    capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
+                  "inverse link function. Coefficients are interpretable as",
+                  "additive effects on the inverse of the mean error.")
+    print(xtable(alpha.glm.tbl, label="tab:glmalpha", caption=c(capt, capt.short)),
+          include.rownames=FALSE, sanitize.text.function=identity,
+          sanitize.colnames.function=identity)
+
+    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
+                        "$I$ to true values of BA parameters.")
+    capt <- paste(capt.short, "GLM was fitted with a gaussian distribution and",
+                  "inverse link function. Coefficients are interpretable as",
+                  "additive effects on the inverse of the mean error.")
+    print(xtable(I.glm.tbl, label="tab:glmI", caption=c(capt, capt.short)),
+          include.rownames=FALSE, sanitize.text.function=identity,
+          sanitize.colnames.function=identity)
+@
+
+{\color{blue}\uline{The \gls{GLM} analysis indicated that the error in estimates of
 \gls{alpha} decreased with larger true values of \gls{alpha} 

### 2016-06-01 doc/thesis/methods_results.tex
moved objective stuff to methods and expanded

@@
 
+\subsection{Justification for approach}
+\label{subsec:just}
+
+{\color{blue}\uline{We present here a non-rigorous justification for the use of
+\gls{ABC} for the problem at hand, as opposed to more frequently-used
+approaches for fitting mathematical models} (see \cref{chp:prelim}).
+\uline{Consider a contact network model with parameters $\theta$, and an
+estimated transmission tree $T$. Taking a Bayesian approach, our aim is to
+obtain a sample from the posterior distribution on the model's parameters given
+our data,}
+\[
+    \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)}.
+\]
+\uline{For all but the simplest models, the normalizing constant $\Pr(T)$ is an
+intractable integral. What we shall argue here is that, in contrast to most
+commonly studied mathematical models, the likelihood $\Pr(T \mid \theta)$ is
+also likely to be intractible in our case.}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{transtree}
+    \caption[
+        Illustration of an estimated transmission tree without labels and two
+        possible underlying complete transmission trees with labels.
+    ]{
+        Illustration of an estimated transmission tree without labels (left)
+        and two possible underlying complete transmission trees with labels
+        (right). In the top right scenario, the epidemic began with node $a$
+        who transmitted first to $b$ and then to $c$. In the bottom right
+        scenario, $b$ was the index case; $b$ infected $c$, who went on to
+        infect $a$. A transmission tree estimated from a viral phylogeny would
+        have the same topology and tip labels in both cases.
+    }
+    \label{fig:tt}
+\end{figure}
+
+\uline{As discussed in} \cref{subsec:tt}, \uline{the internal nodes of
+transmission trees represent transmission events, and are labelled with the
+donor in the associated transmission pair. However, when we estimate a
+transmission tree from viral sequence data, we generally only know the labels
+of the tips of the tree, not the labels of the internal nodes. In viral
+phylogenies, the transmissions are at least partially preserved through the
+evolutionary relationships among the viruses, but the directionality of those
+transmissions is unknown. Thus, a single estimated transmission tree can
+correspond to many possible pathways of the epidemic through the network.
+}\Cref{fig:tt} \uline{illustrates this concept for a simple transmission tree
+with three tips. When calculating a likelihood given a transmission tree, we
+must sum over all possible labellings of the internal nodes. Let $\Nu$ be the
+set of such labellings. Then}
+\begin{align}
+    \Pr(T \mid \theta) = \sum{\nu \in \Nu} \Pr(T, \nu \mid \theta).
+    \label{eq:lik1}
+\end{align}
+\uline{A contact network model assigns a probability to each possible contact
+network. Transmission trees are realized over particular contact networks, not
+over the model itself. Therefore, we must also sum over all contact networks
+which could be generated by the model. Let $\G$ be the set of all possible
+contact networks. Summing } \cref{eq:lik1} \uline{over $\G$ gives}
+\begin{align}
+    \Pr(T \mid \theta) = \sum_{G \in \G} \sum{\nu \in \Nu} 
+        \Pr(T, \nu \mid G, \theta) \Pr(G \mid \theta).
+    \label{eq:lik2}
+\end{align}
+\uline{This can be simplified somewhat by noticing that, given a specific
+contact network, the labelled transmission tree depends only on that network
+and not on the model that generated it. That is, $\Pr(T, \nu \mid G, \theta) =
+\Pr(T, \nu \mid G)$, and}
+\begin{align}
+    \Pr(T \mid \theta) = \sum_{G \in \G} \Pr(G \mid \theta) \sum{\nu \in \Nu} 
+        \Pr(T, \nu \mid G)
+    \label{eq:lik3}
+\end{align}
+\uline{Under the assumption that both transmission and removal are Poisson
+processes, calculating $\Pr(T, \nu \mid G)$ can be accomplished by a
+straightforward modification of the Gillespie simulation algorithm} 
+(\cref{alg:nettree}). \uline{Rather than choosing transmission or removal
+events according to their probabilities, the events would be deterministically
+chosen based on the transmission tree and the probabilities of each event
+would be multiplied together. Assuming efficient data structures for storing
+lists of nodes and edges, the complexity of this calculation would be $O(|T|)$.
+The number of possible labellings of internal nodes of $T$ is easily seen to be
+$2^{(|T|-1)/2}$ by noticing that each of the $(|T|-1)/2$ internal nodes must be
+labelled with the same label as either its right child or its left child.
+Although exponential calculations of this nature can are often be simplified on 
+trees using dynamic programming} (\eg \autocite{pupko2000fast}), \uline{it
+cannot be straightforwardly applied in this case because the subtrees'
+probabilities depend on the existing epidemic progress (their parents and
+siblings). Hence, calculating the inner sum over labels may take time
+$O(2^{(|T|-1)/2})$.}
+
+\uline{The outer sum, over all contact networks, is also difficult to evaluate
+in general. There are $2^{N(N-1)}$ directed graphs on $N$
+nodes~\autocite{harary2014graphical}. There must be at least as many nodes in
+the contact network as the number of tips in the tree, which is $(|T|+1)/2$. Of
+course, it is very likely that there are more nodes in the network than
+observed tips because some individuals are never infected and/or some infected
+individuals are never sampled. The complexity of calculating $\Pr(G \mid
+\theta)$ is obviously dependent on the particular model being investigated. For
+the \gls{BA} model, we might have to sum over all possible orders in which the
+nodes could be added, and all assignments of edges to the nodes which generated
+them. However, even in the unlikely case that calculating $\Pr(G \mid \theta)$
+can be done in constant time, the sum} \ref{eq:lik3} \uline{still has at least
+$O(2^{|T|^2})$ terms.}
+
+\uline{We have shown that both the normalizing constant $\Pr(T)$ and the
+likelihood $\Pr(T \mid \theta)$ are likely computationally prohibitive to
+calculate. If this is the case, the problem of fitting contact network models
+to phylogenies seems to be of the \emph{doubly-intractible}
+type~\autocite{murray2012mcmc}, which would imply that these models are not
+amenable to neither \gls{ML} nor Bayesian inference techniques. Although both
+methods are able to cope with an intractible normalizing constant (for example,
+by local search for \gls{ML} or Bayesian \gls{MCMC}), neither can avoid the
+intractable likelihood calculations. This justifies the use of \gls{ABC}, which
+is a likelihood-free method.}
+
+\uline{We have not proven here that} \cref{eq:lik3} \uline{is impossible to
+calculate in polynomial time - it could be possible to algebraically simplify
+the sum into a tractable expression. Furthermore, under certain models, a large
+proportion of $\G$ may have zero probability, which would enable the
+simplification of the outer sum on a model-specific basis. It should also be
+noted that extensions of Bayesian \gls{MCMC} have been developed for
+doubly-intractable problems~\autocite{lang2010double, murray2012mcmc}, which
+might be adaptable to the problem at hand. These have not been as widely used
+as \gls{ABC}, nor as easily parallelizable as \gls{SMC}. }}
+
 \section{Analysis of \acrlong{BA} model with synthetic data}
@@
 it means that the absolute error decreases as the true value of \gls{alpha}
-increases.}
-
-\uline{The \gls{GLM} analysis indicated that the error in estimates of
+increases.}}
+
+% latex table generated in R 3.2.3 by xtable 1.8-2 package
+% Tue May 31 17:03:28 2016
+\begin{table}[ht]
+\centering
+\begin{tabular}{llll}
+  \hline
+Parameter & Estimate & Standard error & $p$-value \\ 
+  \hline
+(Intercept) & $2$ & $0.6$ & $0.01$ \\ 
+  $\alpha$ & $10$ & $2$ & ${<}10^{-5}$ \\ 
+  I & $-3\!\times\!10^{-4}$ & $2\!\times\!10^{-4}$ & $0.7$ \\ 
+  m & $0.5$ & $0.2$ & $0.01$ \\ 
+   \hline
+\end{tabular}
+\caption[Parameters of a fitted GLM relating error in estimated $\alpha$ to true values of BA parameters.]{Parameters of a fitted GLM relating error in estimated $\alpha$ to true values of BA parameters.
+GLM was fitted with a Gaussian distribution and inverse link function.
+Coefficients are interpretable as additive effects on the inverse of the mean error.}
+\label{tab:glmalpha}
+\end{table}
+% latex table generated in R 3.2.3 by xtable 1.8-2 package
+% Tue May 31 17:03:28 2016
+\begin{table}[ht]
+\centering
+\begin{tabular}{llll}
+  \hline
+Parameter & Estimate & Standard error & $p$-value \\ 
+  \hline
+(Intercept) & $0.004$ & $5\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
+  $\alpha$ & $-0.001$ & $2\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
+  I & $-4\!\times\!10^{-7}$ & $2\!\times\!10^{-7}$ & $0.05$ \\ 
+  m & $-7\!\times\!10^{-5}$ & $8\!\times\!10^{-5}$ & $1$ \\ 
+   \hline
+\end{tabular}
+\caption[Parameters of a fitted GLM relating error in estimated $I$ to true values of BA parameters.]{Parameters of a fitted GLM relating error in estimated $I$ to true values of BA parameters.
+GLM was fitted with a gaussian distribution and inverse link function.
+Coefficients are interpretable as additive effects on the inverse of the mean error.}
+\label{tab:glmI}
+\end{table}
+
+{\color{blue}\uline{The \gls{GLM} analysis indicated that the error in estimates of
 \gls{alpha} decreased with larger true values of \gls{alpha} 
@@
 %
-%<<mixed_peerdriven, include=FALSE>>=
-%  library(coda)
-%  d <- fread("bzcat ../../simulations/abc-pa-mixed-alpha/abc/*")
-%  d <- d[iter == max(iter)]
-%  d <- d[sample(1:nrow(d), prob=weight, replace=TRUE)]
-%  params <- c("alpha", "m", "N", "I")
-%  d <- melt(d, measure.vars=params, variable.name="parameter")
-%  f <- function (x) {
-%    dens <- density(x)
-%    hpd <- HPDinterval(mcmc(x))
-%    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
-%         hpd.max=hpd[,"upper"])
-%  }
-%  d <- d[,f(value), by=parameter]
-%  setkey(d, parameter)
-%
-%  m <- collect.metadata("../../simulations/abc-pa-peerdriven/abc/*")
-%  p <- fread(paste("bzcat", rownames(m)[m$sample_peer == 1]))
-%  p <- p[iter == max(iter)]
-%  p <- p[sample(1:nrow(p), prob=weight, replace=TRUE)]
-%  params <- c("alpha", "m", "N", "I")
-%  p <- melt(p, measure.vars=params, variable.name="parameter")
-%  f <- function (x) {
-%    dens <- density(x)
-%    hpd <- HPDinterval(mcmc(x))
-%    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
-%         hpd.max=hpd[,"upper"])
-%  }
-%  p <- p[,f(value), by=parameter]
-%  setkey(p, parameter)
-%  options(scipen=5)
-%@
-%
-%To test the effect of model misspecification, we simulated one network where
-%the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
-%other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
-%HPD] estimates for each parameter were: 
-%$\alpha$, 
-%  round(d["alpha", map], 2) 
-%  [round(d["alpha", hpd.min], 2) -
-%   round(d["alpha", hpd.max], 2)];
-%$I$,
-%  d["I", round(map)] 
-%  [d["I", round(hpd.min)] -
-%   d["I", round(hpd.max)]];
-%$m$,
-%  d["m", floor(map)] 
-%  [d["m", floor(hpd.min)] -
-%   d["m", floor(hpd.max)]];
-%$N$,
-%  d["N", round(map)] 
-%  [d["N", round(hpd.min)] -
-%   d["N", round(hpd.max)]].
-%The approximate posterior distributions for this simulation are shown in
-%\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
-%transmission tree in a peer-driven fashion, where the probability to sample a
-%node was twice as high if one of its peers had already been sampled. The
-%parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
-%$I$ = 2000. The estimated values were
-%$\alpha$, 
-%  p["alpha", round(map, 2)] 
-%  [p["alpha", round(hpd.min, 2)] -
-%   p["alpha", round(hpd.max, 2)]];
-%$I$,
-%  p["I", floor(map)] 
-%  [p["I", floor(hpd.min)] -
-%   p["I", floor(hpd.max)]];
-%$m$,
-%  p["m", floor(map)] 
-%  [p["m", floor(hpd.min)] -
-%   p["m", floor(hpd.max)]];
-%$N$,
-%  p["N", floor(map)] 
-%  [p["N", floor(hpd.min)] -
-%   p["N", floor(hpd.max)]].
-%The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
-%of these results were in line with estimates obtained on other simulated
-%datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
-%$\alpha$ was somewhat lower than typical.
-
 %\subsubsection*{Effect of parameters on power-law exponent}

### 2016-06-01 doc/thesis/objective.tex
moved objective stuff to methods and expanded

@@
 and $\Pr(T)$ is the marginal probability of $T$ which acts as a normalizing
-constant on the posterior (see \cref{chp:prelim} \uline{for a review of
+constant on the posterior (see} \cref{chp:prelim} \uline{for a review of
 mathematical modeling and Bayesian inference, including definitions of these
-concepts).
-}}
+concepts). As we shall show} (see \cref{subsec:just}), \uline{estimating this
+distribution presents computational challenges beyond those usually encountered
+in Bayesian inference. Both the likelihood $\Pr(T \mid \theta)$ and the
+normalizing constant $\Pr(T)$ seem to be intractable, which rules out the use
+of most common \acrlong{ML} and Bayesian methods. }}
 
-Calculating the likelihood of the parameters of a contact network model seems
-likely to be an intractable problem. We have not proven this is the case, but
-some intuition can be provided by examining the process involved in the
-likelihood calculation. Consider a contact network model with parameters
-$\theta$ and an estimated transmission tree $T$ with $n$ tips.
-{\color{blue}\uline{The transmission tree is a record of an epidemic's progress
-through a network. Tips of the transmission tree correspond to sampled
-individuals, while internal nodes correspond to transmissions and may be
-labelled with the transmitting individual. Transmission trees will be described
-in more detail in the sequel, but for now they can be thought of as
-representing the exact path an epidemic has taken. }} In general, we do not
-know the labels of the internal nodes of $T$, only the labels of its tips. To
-fit this model using likelihood-based methods, we must calculate the likelihood
-of $\theta$, that is, $\Pr(T \mid \theta)$. Let $\G$ be the set of all possible
-contact networks, and $\Nu$ be the set of all possible labellings of the
-internal nodes of $T$. We can write the likelihood as
+{\color{red}\sout{Calculating the likelihood of the parameters of a contact
+network model seems likely to be an intractable problem. We have not proven
+this is the case, but some intuition can be provided by examining the process
+involved in the likelihood calculation. Consider a contact network model with
+parameters $\theta$ and an estimated transmission tree $T$ with $n$ tips. In
+general, we do not know the labels of the internal nodes of $T$, only the
+labels of its tips. To fit this model using likelihood-based methods, we must
+calculate the likelihood of $\theta$, that is, $\Pr(T \mid \theta)$. Let $\G$
+be the set of all possible contact networks, and $\Nu$ be the set of all
+possible labellings of the internal nodes of $T$. We can write the likelihood
+as}
 \begin{align}
@@
 \end{align}
-the last equality following from the fact that $T$ and $\nu$ depend only on
-$G$, not on $\theta$. Although $\Pr(T, \nu \mid G)$ and $\Pr(G \mid \theta)$
+\sout{the last equality following from the fact that $T$ and $\nu$ depend only
+on $G$, not on $\theta$. Although $\Pr(T, \nu \mid G)$ and $\Pr(G \mid \theta)$
 may individually be straightforward to calculate, the number of possible
@@
 $n$, as there must be at least $n$ nodes in the network. In addition,
-\cref{eq:netlik} assumes that $T$ is complete, meaning that all infected
+}\cref{eq:netlik} \sout{assumes that $T$ is complete, meaning that all infected
 individuals were sampled. This is rarely the case in practice - most often, we
@@
 likelihood calculation becomes even more complex, because we must also sum over
-all possible complete trees.
+all possible complete trees.}}
 
-{\color{blue}\uline{
-\uline{
-For all but the simplest mathematical models, the calculation of the
-normalizing constant $\Pr(T)$ is an intractable problem. 
-What we have argued in
-the previous paragraph is that calculating $\Pr(T \mid \theta)$ for a single
-parameter candidate $\theta$ also seems likely to be computationally hard.
-Therefore, the problem of fitting contact network models to phylogenies seems
-to be of the \emph{doubly-intractible} type~\autocite{murray2012mcmc}, which
-would imply that these models are not amenable to neither \gls{ML} nor Bayesian
-inference techniques. Although both methods are able to cope with an
-intractible normalizing constant (for example, by local search for \gls{ML} or
-Bayesian \gls{MCMC}), neither can avoid the intractable likelihood
-calculations. }}
-
-Depending on the network model studied, it is possible that \cref{eq:netlik}
-could be simplified into a tractable expression. However, a simpler alternative
-to likelihood-based methods, which would apply to any network model, is
-provided by \gls{ABC}~\autocite{rubin1984bayesianly, tavare1997inferring,
-fu1997estimating, beaumont2002approximate}. All of the ingredients required to
-apply \gls{ABC} to this problem are readily available. Simulating networks is
-straightforward under a variety of models. Epidemics on those networks, and the
-corresponding transmission trees, can also be easily simulated. As mentioned
-above, contact networks can profoundly affect transmission tree shape. Those
-shapes can be compared using a highly informative similarity measure called the
-``tree kernel''~\autocite{poon2013mapping}; {\color{blue}\uline{similar kernel
+{\color{red}\sout{Depending on the network model studied, it is possible that}
+\cref{eq:netlik} \sout{could be simplified into a tractable expression.}}
+An alternative to likelihood-based methods, which could be applied to any
+network model, is provided by \gls{ABC}~\autocite{rubin1984bayesianly,
+tavare1997inferring, fu1997estimating, beaumont2002approximate}. All of the
+ingredients required to apply \gls{ABC} to this problem are readily available.
+Simulating networks is straightforward under a variety of models. Epidemics on
+those networks, and the corresponding transmission trees, can also be easily
+simulated. As mentioned above, contact networks can profoundly affect
+transmission tree shape. Those shapes can be compared using a highly
+informative similarity measure called the ``tree
+kernel''~\autocite{poon2013mapping}; {\color{blue}\uline{similar kernel
 functions have been demonstrated to work well as distance functions in
@@
 The second aim of this work is \emph{to use simulations to investigate the
-parameters of the \gls{BA} model, including whether they have a detectable
+parameters of the \acrlong{BA} model, including whether they have a detectable
 impact on tree shape, and whether they can be accurately recovered using
@@
 obvious context in which our method could be applied. Therefore, the third and
-final aim of this work is \emph{to apply \gls{ABC} to fit the \gls{BA} model to
-existing \gls{HIV} outbreaks}.
+final aim of this work is \emph{to apply \gls{ABC} to fit the \acrlong{BA}
+model to existing \gls{HIV} outbreaks}.
 

### 2016-06-02 programs/netabc/R/hpd.R
Redid weighted HPD for discrete values

@@ -6,37 +6,61 @@
 #' @return a two-element vector containing upper and lower HPD interval bounds
 #' @export
 wtd.hpd <- function (x, wt, conf=0.95) {
+    if (is.integer(x)) {
+        message("Integer values passed, calling wtd.hpd.discrete")
+        return(wtd.hpd.discrete(x, wt, conf))
+    }
     cdf <- setDT(wtd.Ecdf(x, wt))
     setkey(cdf, x)
-    cdf <- cdf[,tail(.SD, 1), by=x]
-
-    if (is.integer(x)) {
-        icdf <- function (p)
-            sapply(p, function (pi) cdf[ecdf >= pi, head(x, 1)])
-        find.hpd.discrete(icdf, x[which.max(wt)], conf=conf)
-    } else {
-        icdf <- cdf[,approxfun(ecdf, x)]
-        hpd(icdf, conf=conf)
-    }
+    cdf <- cdf[,head(.SD, 1), by=x]
+    icdf <- cdf[,approxfun(ecdf, x)]
+    hpd(icdf, conf=conf)
 }
 
 #' Find a highest density interval for a discrete-valued distribution.
 #'
-#' Helper function for wtd.hpd().
-#'
-#' @param icdf data.table containing inverse CDF values
-#' @param conf how much mass the HPD interval should contain
-#' @param tol narrowness of steps to search ICDF
-find.hpd.discrete <- function (icdf, mode, conf=0.95, tol=1e-3) {
-    # interval widths for [0, conf], [tol, conf+tol], ...
-    p <- seq(0, 1-conf, tol)
-    w <- na.omit(sapply(p, function (pi) icdf(pi + conf) - icdf(pi)))
-
-    # intervals with minimum width 
-    mins <- which(w == min(w))
-    all.ints <- cbind(icdf(p[mins]), icdf(p[mins]+conf))
-
-    # interval which is closest to being symmetric about the mode    
-    keep.int <- which.min(abs((mode - all.ints[,1]) - (all.ints[,2] - mode)))
-    c(lower=all.ints[keep.int,1], upper=all.ints[keep.int,2])
+#' @param x sample from distribution
+#' @param wt weights of each element in the sample
+#' @param conf proportion of probability density contained in interval
+#' @return a two-element vector containing upper and lower HPD interval bounds
+#' @export
+wtd.hpd.discrete <- function (x, wt, conf=0.95) {
+    if (any(diff(x) != 1)) {
[…]

### 2016-06-02 programs/netabc/R/hpd.R
Fix error in consecutive integer checking

@@ -6,7 +6,7 @@
 #' @return a two-element vector containing upper and lower HPD interval bounds
 #' @export
 wtd.hpd <- function (x, wt, conf=0.95) {
-    if (is.integer(x)) {
+    if (all(x == as.integer(x))) {
         message("Integer values passed, calling wtd.hpd.discrete")
         return(wtd.hpd.discrete(x, wt, conf))
     }
@@ -25,7 +25,7 @@ wtd.hpd <- function (x, wt, conf=0.95) {
 #' @return a two-element vector containing upper and lower HPD interval bounds
 #' @export
 wtd.hpd.discrete <- function (x, wt, conf=0.95) {
-    if (any(diff(x) != 1)) {
+    if (any(diff(sort(unique(x))) != 1)) {
         stop("Not implemented yet for non-consecutive integers.")
     }
     

### 2016-06-02 doc/thesis/methods_results.Rnw
Added intro section to chapter 2

@@
 
-{\color{blue}\uline{
-This chapter is organized as follows. One section is devoted to each of our
-three research aims: the development of a method for inference of contact
-network parameters from transmission trees} (\cref{sec:netabc}), \uline{the
-investigation of the \acrlong{BA} network model using simulated data
-}(\cref{sec:ba}), \uline{ and the application of the method to real-world
-\gls{HIV} epidemics }(\cref{sec:hiv}). \uline{Each of these sections contains a
-description of the methods used and, in the case of the latter two objectives,
-the results of the experiments. The fourth and final section of the chapter
-}(\cref{sec:disc}) \uline{contains a unified discussion of all three
-objectives. }}
+{\color{blue}\uline{ In this chapter, we will address the three research aims
+of this thesis introduced in} \cref{sec:obj}. \uline{First, in}
+\cref{sec:netabc}, \uline{we describe \software{netabc}, a computer program
+that implements an \acrlong{ABC}-based algorithm to fit contact network models
+to phylogenetic data. We also provide a justification for the use of \gls{ABC}
+for this problem by arguing that the likelihood functions required to fit these
+models by more conventional means are likely to be computationally intractable.
+Second, in} \cref{sec:ba}, \uline{we perform a simulation study to investigate
+the \acrlong{BA} network model, which uses a preferential attachment mechanism
+to generate networks with the power law degree distributions observed in real
+world social and sexual networks. We progress through two exploratory analyses
+testing the identifiability of the model's parameters, and conclude by testing
+\software{netabc}'s ability to recover the parameters from simulated
+transmission trees. Third, in} \cref{sec:hiv}, \uline{we apply
+\software{netabc} to fit the \gls{BA} model to six real world \gls{HIV}
+datasets, with the understanding of the model parameters' identifiability
+gained through the simulation experiments. We conclude the chapter with a
+unified discussion of the three research aims, including interpretation of the
+results of both the simulated and real data experiments, as well as an
+examination of the limitations of our approach and opportunities for future
+investigation. }}
 

### 2016-06-02 programs/netabc/R/hpd.R
Final (?) fix to discrete HPD...

@@ -35,19 +35,19 @@ wtd.hpd.discrete <- function (x, wt, conf=0.95) {
     
     # record the cdf excluding and including each value
     setkey(cdf, x)
-    cdf <- cdf[,list(start=head(ecdf, 1), end=tail(ecdf, 1)), by=x]
-    cdf[,start := c(head(start, 1), head(end, -1))]
+    cdf <- cdf[,list(dstart=head(ecdf, 1), dend=tail(ecdf, 1)), by=x]
+    cdf[,dstart := c(head(dstart, 1), head(dend, -1))]
     
     # find every possible interval (yeah it's a bit hacky, ideally we would
     # stream them or something but I don't want to bother right now because all
     # my use cases are small)
-    intervals <- setnames(as.data.table(t(cdf[,combn(x, 2)])), c("start", "end"))
-    intervals <- rbind(intervals, cdf[,list(start=x, end=x)])
-    setkey(intervals, start, end)
+    intervals <- setnames(as.data.table(t(cdf[,combn(x, 2)])), c("istart", "iend"))
+    intervals <- rbind(intervals, cdf[,list(istart=x, iend=x)])
+    setkey(intervals, istart, iend)
     
     # record the width of each interval and the density they contain
-    intervals[,width := end - start + 1]
-    intervals[,density := cdf[end,end] - cdf[start,start]]
+    intervals[,width := iend - istart + 1]
+    intervals[,density := cdf[J(iend),dend] - cdf[J(istart),dstart]]
     
     # pick the minimum width interval containing at least conf of the density
     intervals <- intervals[density >= conf]
@@ -55,12 +55,12 @@ wtd.hpd.discrete <- function (x, wt, conf=0.95) {
     
     # if there are multiple intervals left, pick the one which contains the
     # mode
-    intervals <- intervals[mode >= start & mode <= end]
+    intervals <- intervals[mode >= istart & mode <= iend]
     
     # if there are still multiple left, pick the one with the highest density
     intervals <- intervals[density == min(density)]
     
     # finally, just pick one at random
     intervals <- intervals[sample(1:nrow(intervals), 1)]
-    intervals[,c(lower=start, upper=end)]
+    intervals[,c(lower=istart, upper=iend)]
 }

### 2016-06-02 programs/netabc/R/hpd.R
Fix again

@@ -31,6 +31,9 @@ wtd.hpd.discrete <- function (x, wt, conf=0.95) {
     
     # make an empirical CDF and find the mode
     cdf <- setDT(wtd.Ecdf(x, wt))
+    if (cdf[1,x] != cdf[2,x]) {
+        cdf <- rbind(data.table(x=cdf[1,x], ecdf=0), cdf)
+    }
     mode <- cdf[,unique(x)][cdf[,which.max(diff(ecdf))]]
     
     # record the cdf excluding and including each value
@@ -55,7 +58,9 @@ wtd.hpd.discrete <- function (x, wt, conf=0.95) {
     
     # if there are multiple intervals left, pick the one which contains the
     # mode
-    intervals <- intervals[mode >= istart & mode <= iend]
+    if (nrow(intervals) > 1) {
+        intervals <- intervals[mode >= istart & mode <= iend]
+    }
     
     # if there are still multiple left, pick the one with the highest density
     intervals <- intervals[density == min(density)]

### 2016-06-03 doc/thesis/abstract.tex
Updated abstract and a bit more

@@
 structures formed by social contacts, which can have a significant influence on
-the rate of epidemic spread. Contact network models capture this structure by
-explicitly representing each contact that could possibly lead to a
-transmission. We developed a method based on approximate Bayesian computation
-(ABC) for estimating structural parameters of the contact network underlying an
+the rate {\color{blue}\uline{and pattern}} of epidemic spread. Contact
+{\color{red}\sout{network models}} {\color{blue}\uline{networks}} capture this
+structure by explicitly representing each contact that could possibly lead to a
+transmission. {\color{blue}\uline{Contact network models parameterize the
+structure of these networks, but \ldots }}
+
+We developed a method based on approximate Bayesian computation (ABC) for
+estimating structural parameters of the contact network underlying an
 observed viral phylogeny. The method combines adaptive sequential Monte Carlo
 for ABC, Gillespie simulation for propagating epidemics though networks, and a
-kernel-based tree similarity score. We used the method to fit the
-Barab\'{a}si-Albert network model to simulated transmission trees and applied
-it to viral phylogenies estimated from six real-world HIV sequence datasets. On
-simulated data, we found that the preferential attachment power and the number
-of infected nodes in the network can often be accurately estimated. On the
-other hand, the mean degree of the network, as well as the total number of
-nodes, appeared to be weakly or non-identifiable with ABC. We observed
-substantial heterogeneity in the parameter estimates on real datasets, with
-point estimates for the preferential attachment power ranging from 0.06 to
-1.05. These results underscore the importance of considering contact structures
-when performing phylodynamic inference. Our method offers the potential to
-quantitatively investigate the contact network structure underlying viral
-epidemics.
+{\color{blue}\uline{previously developed}} kernel-based tree similarity score.
+{\color{blue}\uline{Our method offers the potential to quantitatively
+investigate contact network structure from phylogenies derived from viral
+sequence data, as a complementary approach to traditional epidemiological
+investigation.}}
+
+We used the method to {\color{red}\sout{fit}}
+{\color{blue}\uline{investigate}} the Barab\'{a}si-Albert network model.
+{\color{blue}\uline{This model incorporates the mechanism of preferential
+attachment observed in real world social and sexual networks, whereby
+individuals with a higher number of connections attract new contacts at an
+elevated rate (``the rich get richer'').}}
+{\color{red}\sout{to simulated transmission trees and applied it to viral
+phylogenies estimated from six real-world HIV sequence datasets.}} Using
+simulated data, we found that the strength of preferential attachment and the
+number of infected nodes in the network could often be accurately estimated. On
+the other hand, the mean degree of the network, as well as the total number of
+nodes, appeared to be weakly or non-identifiable with ABC.
+
+Finally, the Barab\'{a}si-Albert model was fit to six real-world HIV datasets.
+We observed substantial heterogeneity in the parameter estimates on real
+datasets, with point estimates for the preferential attachment power ranging
+from 0.06 to 1.05. 
+{\color{blue}\uline{Point estimates of the strength of preferential attachment 
+were higher in injection drug user populations, potentially indicating\ldots}}
+{\color{red}\sout{These results underscore the importance of considering
+contact structures when performing phylodynamic inference.}}

### 2016-06-03 doc/thesis/discussion.Rnw
A bit more writing
Updated abstract and a bit more

@@
 There was no observable bias in the estimates of \gls{alpha} obtained with
-\software{netabc}, however the variation in these estimates was higher for
-$\alpha < 1$ than for $\alpha \geq 1$ (\cref{fig:abcptm2,tab:abchpd}). The
-relationship between \gls{alpha} and the power law exponent \gls{gamma} may
-explain this result (\cref{fig:gamma}). The \gls{gamma} values associated with
-\gls{alpha} = 0 and \gls{alpha} = 0.5 are nearly identical (about 2.28 for
-\gls{alpha} = 0 and 2.33 for \gls{alpha} = 0.5 with \gls{N} = 5000 and \gls{m}
-= 2). In other words, the degree distributions of networks with \gls{alpha} $<
-1$ are similar to each other, which may result in similarity of corresponding
-transmission trees as well. 
+\software{netabc}, however {\color{red}\sout{the variation in these estimates
+was higher for $\alpha < 1$ than for $\alpha \geq 1$}} {\color{blue}\uline{there
+appeared to be weaker identifiability for \gls{alpha} $<$ 1 than for \gls{alpha}
+$\geq$ 1}} (\cref{fig:abcptm2,tab:abchpd}). The relationship between
+\gls{alpha} and the power law exponent \gls{gamma} may explain this result
+(\cref{fig:gamma}). The \gls{gamma} values associated with \gls{alpha} = 0 and
+\gls{alpha} = 0.5 are nearly identical (about 2.28 for \gls{alpha} = 0 and 2.33
+for \gls{alpha} = 0.5 with \gls{N} = 5000 and \gls{m} = 2). In other words, the
+degree distributions of networks with \gls{alpha} $< 1$ are similar to each
+other, which may result in similarity of corresponding transmission trees as
+well. 
 %The accuracy of \gls{alpha} estimates also increased when the value of the
@@
 
+{\color{blue}\uline{
+The posterior mean estimates for the real world \gls{HIV} datasets corresponded
+to power law exponents $\gamma$ between 2.09 and 2.49.  TODO
+}}
+
 %For both priors, the estimated prevalence was extremely high, in fact higher

### 2016-06-03 doc/thesis/methods_results.Rnw
Updated abstract and a bit more

@@
 
-\subsubsection*{Joint parameter estimates with kernel-assisted ABC}
+\subsubsection*{Joint parameter estimates with \software{netabc}}
 
@@
 of the regions of nonzero prior density. Average \gls{HPD} interval widths were
-    \Sexpr{d[,round(mean(alpha.upper - alpha.lower), 2)]},
-    \Sexpr{d[,round(mean(I.upper - I.lower))]},
-    \Sexpr{d[,round(mean(m.upper - m.lower), 2)]}, and
-    \Sexpr{d[,format(round(mean(N.upper - N.lower)), digits=5)]},
+    \Sexpr{d[,round(mean(alpha.upper.95 - alpha.lower.95), 2)]},
+    \Sexpr{d[,round(mean(I.upper.95 - I.lower.95))]},
+    \Sexpr{d[,round(mean(m.upper.95 - m.lower.95), 2)]}, and
+    \Sexpr{d[,format(round(mean(N.upper.95 - N.lower.95)), digits=5)]},
 representing 
-    \Sexpr{d[,round(mean(alpha.upper - alpha.lower) * 100 / 2)]}\%,
-    \Sexpr{d[,round(mean(I.upper - I.lower) * 100 / 4500)]}\%,
-    \Sexpr{d[,round(mean(m.upper - m.lower) * 100 / 6)]}\%, and
-    \Sexpr{d[,round(mean(N.upper - N.lower) * 100 / 14500)]}\%
+    \Sexpr{d[,round(mean(alpha.upper.95 - alpha.lower.95) * 100 / 2)]}\%,
+    \Sexpr{d[,round(mean(I.upper.95 - I.lower.95) * 100 / 4500)]}\%,
+    \Sexpr{d[,round(mean(m.upper.95 - m.lower.95) * 100 / 6)]}\%, and
+    \Sexpr{d[,round(mean(N.upper.95 - N.lower.95) * 100 / 14500)]}\%
 of the nonzero prior density regions. Point estimates of \gls{I} were upwardly
@@
 
-  map1 <- dcast.data.table(d1, dataset~parameter, value.var="mean")
-  up1 <- dcast.data.table(d1, dataset~parameter, value.var="hpd.max")
-  low1 <- dcast.data.table(d1, dataset~parameter, value.var="hpd.min")
-  setkey(map1, dataset)
+  pmean1 <- dcast.data.table(d1, dataset~parameter, value.var="mean")
+  up1 <- dcast.data.table(d1, dataset~parameter, value.var="hpd.max.95")
+  low1 <- dcast.data.table(d1, dataset~parameter, value.var="hpd.min.95")
+  setkey(pmean1, dataset)
   setkey(up1, dataset)
@@
 
-  map2 <- dcast.data.table(d2, dataset~parameter, value.var="mean")
-  up2 <- dcast.data.table(d2, dataset~parameter, value.var="hpd.max")
-  low2 <- dcast.data.table(d2, dataset~parameter, value.var="hpd.min")
-  setkey(map2, dataset)
+  pmean2 <- dcast.data.table(d2, dataset~parameter, value.var="mean")
+  up2 <- dcast.data.table(d2, dataset~parameter, value.var="hpd.max.95")
+  low2 <- dcast.data.table(d2, dataset~parameter, value.var="hpd.min.95")
+  setkey(pmean2, dataset)
   setkey(up2, dataset)
@@
 estimates [95\% \gls{HPD}] 
-  \Sexpr{map1["niculescu2015", round(alpha, 2)]} 
+  \Sexpr{pmean1["niculescu2015", round(alpha, 2)]} 
   [\Sexpr{low1["niculescu2015", round(alpha, 2)]} - 
@@
 and
-  \Sexpr{map1["wang2015", round(alpha, 2)]} 
+  \Sexpr{pmean1["wang2015", round(alpha, 2)]} 
   [\Sexpr{low1["wang2015", round(alpha, 2)]} -
@@
 low estimated $\alpha$ values
-  (\Sexpr{map1["li2015", round(alpha, 2)]} 
+  (\Sexpr{pmean1["li2015", round(alpha, 2)]} 
   [\Sexpr{low1["li2015", round(alpha, 2)]} - 
@@
 and
-  \Sexpr{map1["cuevas2009", round(alpha, 2)]} 
+  \Sexpr{pmean1["cuevas2009", round(alpha, 2)]} 
   [\Sexpr{low1["cuevas2009", round(alpha, 2)]} -
@@
 the HPD intervals remained roughly the same
-  (\Sexpr{map2["li2015", round(alpha, 2)]} 
+  (\Sexpr{pmean2["li2015", round(alpha, 2)]} 
   [\Sexpr{low2["li2015", round(alpha, 2)]} - 
@@
 and
-  \Sexpr{map2["cuevas2009", round(alpha, 2)]} 
+  \Sexpr{pmean2["cuevas2009", round(alpha, 2)]} 
   [\Sexpr{low2["cuevas2009", round(alpha, 2)]} -
@@
 for both priors on $m$
-  (\Sexpr{map1["novitsky2014", round(alpha, 2)]} for $m \geq 1$;
-   \Sexpr{map2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).
+  (\Sexpr{pmean1["novitsky2014", round(alpha, 2)]} for $m \geq 1$;
+   \Sexpr{pmean2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).
 However, the confidence interval was much wider when $m = 1$ was allowed
@@
 intervals compared to the nonzero prior density region
-  (\citeauthor{cuevas2009hiv}, \Sexpr{map1["cuevas2009", floor(I)]} 
+  (\citeauthor{cuevas2009hiv}, \Sexpr{pmean1["cuevas2009", floor(I)]} 
   [\Sexpr{low1["cuevas2009", floor(I)]} -
    \Sexpr{up1["cuevas2009", floor(I)]}];
-   \citeauthor{niculescu2015recent}, \Sexpr{map1["niculescu2015", floor(I)]}
+   \citeauthor{niculescu2015recent}, \Sexpr{pmean1["niculescu2015", floor(I)]}
   [\Sexpr{low1["niculescu2015", floor(I)]} - 
    \Sexpr{up1["niculescu2015", floor(I)]}];
-  \citeauthor{li2015hiv}, \Sexpr{map1["li2015", floor(I)]} 
+  \citeauthor{li2015hiv}, \Sexpr{pmean1["li2015", floor(I)]} 
   [\Sexpr{low1["li2015", floor(I)]} -
    \Sexpr{up1["li2015", floor(I)]}];
-   \citeauthor{wang2015targeting}, \Sexpr{map1["wang2015", floor(I)]}
+   \citeauthor{wang2015targeting}, \Sexpr{pmean1["wang2015", floor(I)]}
   [\Sexpr{low1["wang2015", floor(I)]} - 
@@
 estimated $I$, and HPD interval spanning almost the entire prior region
-  (\Sexpr{map1["novitsky2014", floor(I)]} 
+  (\Sexpr{pmean1["novitsky2014", floor(I)]} 
   [\Sexpr{low1["novitsky2014", floor(I)]} -
@@
 data, for which the MAP estimate was lower 
-  (\Sexpr{map1["li2015", floor(N)]})
+  (\Sexpr{pmean1["li2015", floor(N)]})
 when $m = 1$ was allowed.

### 2016-06-03 doc/thesis/methods_results.tex
Updated abstract and a bit more

@@
-{\color{blue}\uline{
-This chapter is organized as follows. One section is devoted to each of our
-three research aims: the development of a method for inference of contact
-network parameters from transmission trees} (\cref{sec:netabc}), \uline{the
-investigation of the \acrlong{BA} network model using simulated data
-}(\cref{sec:ba}), \uline{ and the application of the method to real-world
-\gls{HIV} epidemics }(\cref{sec:hiv}). \uline{Each of these sections contains a
-description of the methods used and, in the case of the latter two objectives,
-the results of the experiments. The fourth and final section of the chapter
-}(\cref{sec:disc}) \uline{contains a unified discussion of all three
-objectives. }}
+{\color{blue}\uline{ In this chapter, we will address the three research aims
+of this thesis introduced in} \cref{sec:obj}. \uline{First, in}
+\cref{sec:netabc}, \uline{we describe \software{netabc}, a computer program
+that implements an \acrlong{ABC}-based algorithm to fit contact network models
+to phylogenetic data. We also provide a justification for the use of \gls{ABC}
+for this problem by arguing that the likelihood functions required to fit these
+models by more conventional means are likely to be computationally intractable.
+Second, in} \cref{sec:ba}, \uline{we perform a simulation study to investigate
+the \acrlong{BA} network model, which uses a preferential attachment mechanism
+to generate networks with the power law degree distributions observed in real
+world social and sexual networks. We progress through two exploratory analyses
+testing the identifiability of the model's parameters, and conclude by testing
+\software{netabc}'s ability to recover the parameters from simulated
+transmission trees. Third, in} \cref{sec:hiv}, \uline{we apply
+\software{netabc} to fit the \gls{BA} model to six real world \gls{HIV}
+datasets, with the understanding of the model parameters' identifiability
+gained through the simulation experiments. We conclude the chapter with a
+unified discussion of the three research aims, including interpretation of the
+results of both the simulated and real data experiments, as well as an
+examination of the limitations of our approach and opportunities for future
+investigation. }}
 
@@
 noted that extensions of Bayesian \gls{MCMC} have been developed for
-doubly-intractable problems~\autocite{lang2010double, murray2012mcmc}, which
+doubly-intractable problems~\autocite{liang2010double, murray2012mcmc}, which
 might be adaptable to the problem at hand. These have not been as widely used
-as \gls{ABC}, nor as easily parallelizable as \gls{SMC}. }}
+as \gls{ABC}, nor are they as easily parallelizable as \gls{SMC}. }}
 
@@
 of the regions of nonzero prior density. Average \gls{HPD} interval widths were
-    0.56,
-    1737,
-    3.06, and
-    12078,
+    0.68,
+    2454,
+    3.01, and
+    12046,
 representing 
-    28\%,
-    39\%,
-    51\%, and
+    34\%,
+    55\%,
+    50\%, and
     83\%
@@
 % latex table generated in R 3.2.3 by xtable 1.8-2 package
-% Tue May 31 17:03:28 2016
+% Thu Jun  2 15:02:59 2016
 \begin{table}[ht]
@@
   [1 -
-   4];
+   5];
 $N$,
@@
   [2 -
-   4];
+   5];
 $N$,
@@
   [0.07 - 
-   1.31]
+   1.34]
 and
@@
   [0.01 -
-   1.03] respectively).
+   1.02] respectively).
 The MAP estimates did not change appreciably when $m = 1$ was disallowed by the
@@
   (0.07 - 
-   1.31).
+   1.34).
 When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
@@
   0.38 
-  [0.02 -
-   0.82]). 
+  [0.01 -
+   0.8]). 
 However, the MAP estimates increased when $m = 1$ was not permitted, although
@@
   (\citeauthor{cuevas2009hiv}, 943 
-  [285 -
-   2111];
+  [287 -
+   2093];
    \citeauthor{niculescu2015recent}, 1032
   [136 - 
-   2847];
+   2851];
   \citeauthor{li2015hiv}, 1437 
-  [364 -
-   2659];
+  [365 -
+   2897];
    \citeauthor{wang2015targeting}, 872
@@
   [187 -
-   8812]).
+   8806]).
 The $I$ estimates and HPD intervals were generally robust to the choice of
@@
 upper bound of the HPD interval was different for each dataset
-  (\citeauthor{niculescu2015recent}, 4;
-   \citeauthor{wang2015targeting}, 3;
+  (\citeauthor{niculescu2015recent}, 5;
+   \citeauthor{wang2015targeting}, 4;
    \citeauthor{li2015hiv}, 1;

### 2016-06-03 doc/thesis/abstract.tex
Tiny abstract changes... zzz

@@
 transmission. {\color{blue}\uline{Contact network models parameterize the
-structure of these networks, but \ldots }}
+structure of these networks, but estimating their parameters from contact data
+requires extensive, often prohibitive, epidemiological investigation. }}
 
@@
 sequence data, as a complementary approach to traditional epidemiological
-investigation.}}
+methods.}}
 
-We used the method to {\color{red}\sout{fit}}
+We applied our method to {\color{red}\sout{fit}}
 {\color{blue}\uline{investigate}} the Barab\'{a}si-Albert network model.
@@
 
-Finally, the Barab\'{a}si-Albert model was fit to six real-world HIV datasets.
+Finally, the Barab\'{a}si-Albert model was fit to six real world HIV datasets.
 We observed substantial heterogeneity in the parameter estimates on real
@@
 {\color{blue}\uline{Point estimates of the strength of preferential attachment 
-were higher in injection drug user populations, potentially indicating\ldots}}
-{\color{red}\sout{These results underscore the importance of considering
-contact structures when performing phylodynamic inference.}}
+were higher in injection drug user populations, potentially indicating that
+high-degree ''hub'' nodes may play a role in epidemics among this risk group.}}
+Our results underscore the importance of considering contact structures when
+{\color{red}\sout{performing phylodynamic inference}}{\color{blue}\uline{
+investigating viral outbreaks.}}

### 2016-06-06 doc/thesis/new.tex
Wrote some things

% for intro about power law
Strictly speaking, when $\alpha \neq 1$, the degree distribution no longer
obeys a power law, but we do not believe this affects the applicability of the
model to real world networks. When $\alpha < 1$, the distribution follows a
stretched exponential~\autocite{krapivsky2000connectivity}.
\textcite{laherrere1998stretched} argue that stretched exponentials offer a
better explanation for diverse natural phenomena than power laws, and also
point out that the two distributions are often similar except in the far right
tail. When $\alpha > 1$, the degree distribution exhibits behaviour referred to
as \defn{gelation}, namely a small number of hub nodes, possibly only one,
which collectively are connected to every other node in the graph. Despite
these deviations from the power law distribution, the idea that the strength of
preferential attachment may be more or less than exactly linear, and that it
may vary between geographic and demographic contexts, seems a natural one.
In fact, \textcite{de2007preferential} were able to estimate the \acrlong{PA}
power from partner count data collected from the same individuals for
consecutive time intervals, and found a value less than one in all cases.  It
is also worth noting that, in addition to the \gls{BA} model, other
investigations of the interaction between contact networks and transmission
trees have studied the \gls{ER} and \gls{WS} models, whose degree distributions 
do not generally follow a power law under any parameter settings.

% intro or objective - it's hard to get real contact networks
True contact networks have been estimated in several communities experiencing 
sexually transmitted infection outbreaks (\autocite{clemencon2015statistical},
also see \autocite{morris2011hiv, rothenberg2007large} and references therein).
Though valuable, this kind of investigation is a substantial undertaking,
especially in the time-sensitive situation of an emerging epidemic. The
in-depth surveying, contact tracing, and data analysis required are simply
infeasible in most contexts; even if the resources exist, there are often more
pressing concerns.

% methods: fitting gamma to real data
As discussed in \cref{subsec:pa}, the degree distributions generated by values
of \gls{alpha} not equal to one do not follow a power law, but rather a
stretched exponential (for \gls{alpha} $<$ 1) or a gelation-type distribution
(for \gls{alpha} $>$ 1). However, for many

% results: fitting gamma to real data

% discussion about PA power estimation paper
For all datasets we examined, the posterior mean estimates for \gls{alpha} were
sub-linear, ranging from X to X. The sub-linearity is consistent with the
results of \textcite{de2007preferential}, who developed a statistical inference
method to estimate the parameters of a more sophisticated \acrlong{PA} model
incorporating heterogeneous node behaviour. They found \gls{alpha} values
ranging from 0.26 to 0.62, depending on the gender and time period considered. 
Our estimates of \gls{alpha} for the \textcite{niculescu2015recent} was above
this range under both priors, as were the estimates for the
\textcite{wang2015targeting} data and the BC data when \gls{m} = 1 was
disallowed by the prior. The dataset investigated by
\textcite{de2007preferential} was derived from a survey of a random sample of
the Norwegian population, whereas our investigation focused on datasets from
known phylogenetic or geographic clusters of \gls{HIV} infected persons. It is
therefore unsurprising that we detected stronger \acrlong{PA} dynamics in some
cases. For instance, random sampling is much less likely to discover the
high-degree nodes characterizing the tail of the degree distribution, simply
because those individuals are rare in the general population. In addition, it
is plausible that \gls{HIV}-positive individuals are more likely to be highly
connected in their sexual networks, as the odds of acquiring \gls{HIV} increase
with the number of unprotected sexual contacts.

Both \textcite{de2007preferential} and \textcite{novitsky2014impact} studied
populations whose primary risk factor for \gls{HIV} infection was heterosexual
contact. \citeauthor{de2007preferential} explicitly excluded reported
homosexual contacts; \citeauthor{novitsky2014impact} did not, but noted that
heterosexual contact is the primary mode of transmission in Botswana where the
study was done. In the first of the two papers describing the Botswana
study~\autocite{novitsky2013phylogenetic}, the authors noted that their sample
was gender-biased, being composed of approximately 75\% women. Our estimate of
\gls{alpha} for these data was X or X, depending on the prior on \gls{m};
\citeauthor{de2007preferential} estimated 0.54, 0.57, and 0.29 for 3-year,
5-year, and lifetime partnership networks respectively for the female portion
of their sample.

For both choices of prior on \gls{m}, the datasets derived from \gls{IDU}
populations had a higher estimated \acrlong{PA} power than the other datasets
(\cref{fig:abchpd,fig:abchpdm2}). This finding is in line with
\textcite{dombrowski2013topological}, who reanalyzed a network of \glspl{IDU}
in Brooklyn, USA, collected between 1991 and
1993~\autocite{friedman2006social}. They found that the the \gls{IDU} network 
resembled a \gls{BA} network much more closely than other social and sexual 
networks, and offered sociological explanations for the apparent \acrlong{PA}
dynamics in this population. Importantly, from a public health perspective,
the authors asserted that the removal of \emph{random} individuals from
\gls{IDU} networks may have the paradoxical effect of increasing the network's
epidemic susceptibility. When low-degree nodes are removed, as would occur
during a police crackdown, their network neighbours may turn to well-known
community members for advice or supplies, thus increasing the connectivity of
these high-degree nodes.

One somewhat surprising result was the difference between parameter estimates 
for the \textcite{li2015hiv} and \textcite{wang2015targeting} datasets. Both
groups studied cohorts of acutely infected \gls{MSM} in major Chinese cities
(Shanghai and Beijing respectively); yet, the \citeauthor{li2015hiv} data was 
estimated to have a lower \acrlong{PA} power and larger infected population
than the \textcite{wang2015targeting} data.

% for discussion
In order to compare our results to existing literature on networks and
distributions of partner counts, we have reported estimated values for the
power law exponent \gls{gamma} of the real data sets we evaluated. However, the
posterior means for \gls{alpha} for all six datasets were less than one; the
degree distributions in this parameter range are stretched exponential, not
power law~\autocite{krapivsky2000connectivity}. As we show in
\cref{fig:powerlaw}, the power law fit does capture the slope of the degree
distribution fairly well, but the results should still be interpreted
cautiously. \textcite{krapivsky2000connectivity} showed that the power law
distribution can be maintained, with \gls{gamma} tuned to any desired value, by
a straightforward modification of the \gls{BA} model. The authors define the
``connection kernel'' $A_k$ the probability of a new connection to a node of
degree $k$, up to a normalizing constant. In the \gls{BA} model as we have
presented it here, $A_k = k^{\alpha} + 1$. Taking $A_k$ as any asymptotically
linear function will result in a power law distribution, with the exponent
\gls{gamma} determined by the properties of $A_k$. Implementing such a model
would be straightforward and seems a natural next step toward improving the
realism of the \gls{BA} model.

\textcite{rothenberg2007large} estimated the power law exponents for 15 major
network studies of sexual and injection drug use networks carried out between
1981 and 2000. The estimated \gls{gamma} values for ``all contacts'' (that is,
including both interviewed and non-interviewed individuals) were 
between 2.08 and 2.87 for all but one of the networks.

% misc
romano2010social (Social Networks Shape the Transmission Dynamics of
Hepatitis C Virus) found power law exponent 2.15 with xmin = 10

hughes2009molecular (Molecular Phylodynamics of the Heterosexual HIV Epidemic
in the United Kingdom) found a power law was a good fit to UK sexual networks
with slope = 2.1

see morris2011hiv (http://www.icpsr.umich.edu/icpsrweb/NAHDAP/studies/22140) and
rothenberg2007large (Large-Network Concepts and Small-Network Characteristics:
Fixed and Variable Factors) for empirical HIV networks

clemencon2015statistical estimated xmin = 7, gamma = 3.02 for MSM (see supplements)

otoh schneeberger has gamma estimates for MSM of 1.75, 1.57, 1.43

zarabi2013combining may have some data for mixed risk group studies, useful for cuevas

### 2016-06-07 doc/thesis/abstract.tex
More exposition for ABC; rewrote discussion of real data

@@
 investigate contact network structure from phylogenies derived from viral
-sequence data, as a complementary approach to traditional epidemiological
-methods.}}
+sequence data, complementing traditional epidemiological methods.}}
 
@@
 {\color{blue}\uline{investigate}} the Barab\'{a}si-Albert network model.
-{\color{blue}\uline{This model incorporates the mechanism of preferential
-attachment observed in real world social and sexual networks, whereby
-individuals with a higher number of connections attract new contacts at an
-elevated rate (``the rich get richer'').}}
+{\color{blue}\uline{This model incorporates the preferential attachment
+mechanism observed in real world social and sexual networks, whereby
+individuals with more connections attract new contacts at an elevated rate
+(``the rich get richer'').}}
 {\color{red}\sout{to simulated transmission trees and applied it to viral
@@
 simulated data, we found that the strength of preferential attachment and the
-number of infected nodes in the network could often be accurately estimated. On
-the other hand, the mean degree of the network, as well as the total number of
-nodes, appeared to be weakly or non-identifiable with ABC.
+number of infected nodes could often be accurately estimated. However, the mean
+degree of the network and the total number of nodes appeared to be weakly- or
+non-identifiable with ABC.
 
-Finally, the Barab\'{a}si-Albert model was fit to six real world HIV datasets.
-We observed substantial heterogeneity in the parameter estimates on real
-datasets, with point estimates for the preferential attachment power ranging
-from 0.06 to 1.05. 
-{\color{blue}\uline{Point estimates of the strength of preferential attachment 
-were higher in injection drug user populations, potentially indicating that
-high-degree ''hub'' nodes may play a role in epidemics among this risk group.}}
-Our results underscore the importance of considering contact structures when
-{\color{red}\sout{performing phylodynamic inference}}{\color{blue}\uline{
-investigating viral outbreaks.}}
+Finally, the Barab\'{a}si-Albert model was fit to six real world HIV datasets,
+and substantial heterogeneity in the parameter estimates was observed.
+{\color{red}\sout{Point estimates}}{\color{blue}\uline{Depending on the choice
+of prior, posterior means}} for the preferential attachment power ranged from
+{\color{red}\sout{0.06 to 1.05}} {\color{blue}\uline{0.26, slightly less than
+logarithmic, to 1.00, exactly linear.}} {\color{blue}\uline{Point estimates of
+the strength of preferential attachment were higher in two injection drug user
+populations, potentially indicating that high-degree ''hub'' nodes may play a
+role in epidemics among this risk group.}} Our results underscore the
+importance of considering contact structures when {\color{red}\sout{performing
+phylodynamic inference}}{\color{blue}\uline{investigating viral outbreaks.}}

### 2016-06-07 doc/thesis/discussion.tex
More exposition for ABC; rewrote discussion of real data

@@
 
-With respect to the preferential attachment power $\alpha$, the six datasets
-analysed fell into two categories (\cref{fig:abchpd}). First, we estimated a
-preferential attachment power close to 1, indicating linear preferential
-attachment, for the BC data and the outbreaks studied by
-\textcite{niculescu2015recent} and \textcite{wang2015targeting}. These values
-were robust to specifying different priors for $m$. All three datasets were
-sampled from populations in which we would expect a high degree of
-epidemiological relatedness: \textcite{niculescu2015recent} studied a recent
-outbreak among Romanian \gls{IDU}, \citeauthor{wang2015targeting} sampled
-acutely infected MSM in Beijing, China, and the BC data constituted a
-phylogenetic \gls{IDU} cluster. These are all contexts in which we would expect
-some of the assumptions of the BA model, such as a connected network,
-relatively high mean degree, and preferential attachment dynamics, to hold.
+{\color{red}\sout{With respect to the preferential attachment power $\alpha$,
+the six datasets analysed fell into two categories} (\cref{fig:abchpd}).
+\sout{First, we estimated a preferential attachment power close to 1,
+indicating linear preferential attachment, for the BC data and the outbreaks
+studied by} \textcite{niculescu2015recent} and \textcite{wang2015targeting}.
+\sout{These values were robust to specifying different priors for $m$. All three
+datasets were sampled from populations in which we would expect a high degree
+of epidemiological relatedness:} \textcite{niculescu2015recent} \sout{studied a
+recent outbreak among Romanian \gls{IDU},} \citeauthor{wang2015targeting}
+\sout{sampled acutely infected MSM in Beijing, China, and the BC data
+constituted a phylogenetic \gls{IDU} cluster. These are all contexts in which
+we would expect some of the assumptions of the BA model, such as a connected
+network, relatively high mean degree, and preferential attachment dynamics, to
+hold.}
 
-The remaining three datasets (\textcite{cuevas2009hiv, novitsky2014impact,
-li2015hiv}) had estimated values of $\alpha$ below 0.5 when $m = 1$ was
-included in the prior, but these were not robust to changing the prior to
-exclude $m = 1$. For the \citeauthor{cuevas2009hiv} data, model
-misspecification is likely partially responsible. While the authors found that
-a large proportion of the samples were epidemiologically linked, these were
-mainly in small local clusters rather than the single large component
-postulated by the BA model. In addition, the mixed risk groups in the dataset
-would be unlikely to significantly interact, further weakening any global
-preferential attachment dynamics. The dataset studied by
-\textcite{novitsky2014impact} originated from a densely sampled population
-where the predominant risk factor was believed to be heterosexual exposure.
-Although the MAP estimate of $\alpha$ was almost unchanged when the value $m =
-1$ was excluded from the prior, the confidence interval shrank substantially.
+\sout{The remaining three datasets} (\textcite{cuevas2009hiv,
+novitsky2014impact, li2015hiv}) \sout{had estimated values of $\alpha$ below
+0.5 when $m = 1$ was included in the prior, but these were not robust to
+changing the prior to exclude $m = 1$. For the} \citeauthor{cuevas2009hiv} 
+\sout{data, model misspecification is likely partially responsible. While the
+authors found that a large proportion of the samples were epidemiologically
+linked, these were mainly in small local clusters rather than the single large
+component postulated by the BA model. In addition, the mixed risk groups in the
+dataset would be unlikely to significantly interact, further weakening any
+global preferential attachment dynamics. The dataset studied by}
+\textcite{novitsky2014impact} \sout{originated from a densely sampled
+population where the predominant risk factor was believed to be heterosexual
+exposure. Although the MAP estimate of $\alpha$ was almost unchanged when the
+value $m = 1$ was excluded from the prior, the confidence interval shrank
+substantially.}}
+
+For all datasets we examined, the posterior mean estimates for \gls{alpha} were
+sub-linear, ranging from 
+    0.27
+to 
+    0.73.
+The sub-linearity is consistent with the results of
+\textcite{de2007preferential}, who developed a statistical inference method to
+estimate the parameters of a more sophisticated \acrlong{PA} model
+incorporating heterogeneous node behaviour. They found \gls{alpha} values
+ranging from 0.26 to 0.62, depending on the gender and time period considered. 
+Our estimates of \gls{alpha} for the \textcite{niculescu2015recent} was above
+this range under both priors, as were the estimates for the
+\textcite{wang2015targeting} data and the BC data when \gls{m} = 1 was
+disallowed by the prior. The dataset investigated by
+\textcite{de2007preferential} was derived from a survey of a random sample of
+the Norwegian population, whereas our investigation focused on datasets from
+known phylogenetic or geographic clusters of \gls{HIV} infected persons. It is
+therefore unsurprising that we detected stronger \acrlong{PA} dynamics in some
+cases. For instance, random sampling is much less likely to discover the
+high-degree nodes characterizing the tail of the degree distribution, simply
+because those individuals are rare in the general population. In addition, it
+is plausible that \gls{HIV}-positive individuals are more likely to be highly
+connected in their sexual networks, as the odds of acquiring \gls{HIV} increase
+with the number of unprotected sexual contacts.
+
+Both \textcite{de2007preferential} and \textcite{novitsky2014impact} studied
+populations whose primary risk factor for \gls{HIV} infection was heterosexual
+contact. \citeauthor{de2007preferential} explicitly excluded reported
+homosexual contacts; \citeauthor{novitsky2014impact} did not, but noted that
+heterosexual contact is the primary mode of transmission in Botswana where the
+study was done. In the first of the two papers describing the Botswana
+study~\autocite{novitsky2013phylogenetic}, the authors noted that their sample
+was gender-biased, being composed of approximately 75\% women. Our estimate of
+\gls{alpha} for these data was X or X, depending on the prior on \gls{m};
+\citeauthor{de2007preferential} estimated 0.54, 0.57, and 0.29 for 3-year,
+5-year, and lifetime partnership networks respectively for the female portion
+of their sample.
+
+For both choices of prior on \gls{m}, the datasets derived from \gls{IDU}
+populations had a higher estimated \acrlong{PA} power than the other datasets
+(\cref{fig:abchpd,fig:abchpdm2}). This finding is in line with
+\textcite{dombrowski2013topological}, who reanalyzed a network of \glspl{IDU}
+in Brooklyn, USA, collected between 1991 and
+1993~\autocite{friedman2006social}. They found that the the \gls{IDU} network 
+resembled a \gls{BA} network much more closely than other social and sexual 
+networks, and offered sociological explanations for the apparent \acrlong{PA}
+dynamics in this population. Importantly, from a public health perspective,
+the authors asserted that the removal of \emph{random} individuals from
+\gls{IDU} networks may have the paradoxical effect of increasing the network's
+epidemic susceptibility. When low-degree nodes are removed, as would occur
+during a police crackdown, their network neighbours may turn to well-known
+community members for advice or supplies, thus increasing the connectivity of
+these high-degree nodes.
 
@@
 \gls{HPD} interval did contain the value 494. In addition, it is not
-unreasonable to imagine that a substantial fraction of the \gls{HIV}-positive
-individuals in an \gls{IDU} population could be undiagnosed, given that these
-populations are often marginalized.}
+unreasonable that a substantial fraction of the \gls{HIV}-positive individuals
+in an \gls{IDU} population could be undiagnosed, given that these populations
+are often marginalized.}
 
-\uline{The individuals sampled by \autocite{wang2015targeting} were recruited by
-following a prospective cohort of 2000 \gls{MSM} in Beijing. Of the 2000, 179
-new infections were identified during the study period. This number is much
-lower than the estimated prevalence of X obtained with \software{netabc}.
-However, although} \citeauthor{wang2015targeting} TODO
+\uline{The individuals sampled by} \textcite{wang2015targeting} \uline{were
+recruited by following a prospective cohort of 2000 \gls{MSM} in Beijing,
+China. Of the 2000, 179 new infections were identified during the study period.
+This number is much lower than the estimated prevalence of X obtained with
+\software{netabc}, however the authors did not claim to have sampled the entire
+sexual network. In a nationwide survey,} \textcite{wu2013hiv} \uline{estimated
+that there were 24,198 \gls{MSM} living in Beijing, of whom 5.7\%, or 1379,
+were \gls{HIV}-infected, which is within the 95\% confidence interval of our
+estimate of \gls{I}.}
+
+\textcite{li2015hiv} \uline{studied an \gls{MSM} population in Shanghai, China.
+Despite the apparent similarity to the context of the
+}\textcite{wang2015targeting} \uline{study, the estimated \gls{HIV} prevalence
+of the } \citeauthor{li2015hiv} \uline{data was higher (X). The aforementioned
+survey~\autocite{wu2013hiv} estimated that there were 14,511 \gls{MSM} living
+in Shanghai, with an \gls{HIV} prevalence of 6.8\% or approximately 987
+individuals. This was within the 95\% confidence interval of the estimate
+obtained with \software{netabc}. It is worth noting that} \citeauthor{li2015hiv}
+\uline{claimed that there were 80,000 \gls{MSM} living in Shanghai, which is a
+significantly different estimate than that obtained by} \textcite{wu2013hiv}.
 
@@
 part be due to model misspecification, which highlights the importance of
-selecting a model appropriate to the dataset at hand.}}
+selecting a model appropriate to the dataset at hand.} }
 

### 2016-06-07 doc/thesis/methods_results.tex
More exposition for ABC; rewrote discussion of real data

@@
 confused with the \gls{BA} model parameter)}} controls the decay rate.
-{\color{blue}\uline{In the original paper of }\textcite{del2012adaptive}, \uline{the
-parameter is called $\alpha$, but to avoid confusion with the \gls{BA}
-parameter of the same name, we will refer to it here as $\alpha_{\ESS}$.}}
+{\color{blue}\uline{In the original paper of }\textcite{del2012adaptive},
+\uline{the parameter is called $\alpha$, but to avoid confusion with the
+\gls{BA} parameter of the same name, we will refer to it here as
+\gls{alphaess}.}}
 The tolerance $\varepsilon_i$ is chosen to satisfy
@@
 to phylogenies seems to be of the \emph{doubly-intractible}
-type~\autocite{murray2012mcmc}, which would imply that these models are not
+ype~\autocite{murray2012mcmc}, which would imply that these models are not
 amenable to neither \gls{ML} nor Bayesian inference techniques. Although both
@@
 \glsreset{BA}
+
+%{\color{blue}\uline{The \gls{BA} contact network model} (see \cref{subsec:pa})
+%\uline{generates networks with a power law degree distribution. Power law
+%distributions have been observed by several empirical studies of sexual
+%behaviour}~\autocite{colgate1989risk, liljeros2001web, schneeberger2004scale,
+%clemencon2015statistical} (although not all, see
+%\autocite{handcock2004likelihood, bansal2007individual}).}
 
@@
 \emph{combinations} of parameters; this issue will be explored later by jointly
-estimating all parameters with \gls{ABC}} (\cref{subsec:abc}).
+estimating all parameters with \gls{ABC}}.
 
@@
 compares two trees based on normalized distributions of their branching times}
-(see \cref{sec:treeshape}).  \uline{In contrast with Sackin's index, the
+(see \cref{subsec:treeshape}).  \uline{In contrast with Sackin's index, the
 \gls{nltt} does not explicity consider the trees' topologies, but it does use
@@
   \includegraphics[width=\textwidth]{gridsearch-example}
-  \caption[Grid search estimates of \gls{BA} model parameters.]{Point estimates
-      (center lines) and 75\% (boxes) and 95\% (whiskers) highest density
-      intervals for each \gls{BA} model parameter, obtained using grid search.
-      Networks and transmission trees were simulated over a grid of values for
-       each parameter while holding the others fixed. For a subset of the grid
-      values ($x$-axis), test networks and trees were created and compared to
-      each tree on the grid using the tree kernel. The kernel scores along the
-      grid were normalized to resemble a probability distribution, from which
-      the mode and highest density interval were calculated. Shown values
-      correspond to one replicate experiment, with trees of size 500.
+  \caption[
+    Grid search estimates of \gls{BA} model parameters for one replicate
+    experiment with trees of size 500.
+  ]{
+    Grid search estimates of \gls{BA} model parameters for one replicate
+    experiment with trees of size 500. Point estimates (dots), 95\% highest
+    density intervals (lines), and 50\% highest density intervals (notches) for
+    each \gls{BA} model parameter, obtained using grid search. Networks and
+    transmission trees were simulated over a grid of values for each parameter
+    while holding the others fixed. For a subset of the grid values ($x$-axis),
+    test networks and trees were created and compared to each tree on the grid
+    using the tree kernel. The kernel scores along the grid were normalized to
+    resemble a probability distribution, from which the mode and highest
+    density interval were calculated. 
   } 
@@
 simulations
-    (4\%).
+    (96\%).
 The estimates for \gls{m} and \gls{N} were similar across all simulations
@@
 % latex table generated in R 3.2.3 by xtable 1.8-2 package
-% Thu Jun  2 15:02:59 2016
+% Mon Jun  6 15:27:33 2016
 \begin{table}[ht]
@@
 
-Empirical studies of contact networks often report the exponent $\gamma$ of the
-power law degree distribution. To compare our results to the literature, we
-simulated 100 networks each according to the \gls{MAP} parameter estimates
-obtained for each investigated dataset. The power-law exponent $\gamma$ was
-calculated for each network using the \software{fit\_power\_law} function in
-\software{igraph}, with the `R.mle' implementation. The median of the 100
-$\gamma$ values was taken as a point estimate for the associated dataset.
-
 For all datasets, we used the priors $\alpha$ $\sim$ Uniform(0, 2) and $N$ and
@@
 
+Empirical studies of contact networks often report the exponent $\gamma$ of the
+power law degree distribution. To compare our results to the literature, we
+simulated 100 networks each according to the posterior mean parameter estimates
+obtained for each investigated dataset. The power-law exponent $\gamma$ was
+calculated for each network using the \software{fit\_power\_law} function in
+\software{igraph}, with the `R.mle' implementation. The median of the 100
+$\gamma$ values was taken as a point estimate for the associated dataset.
+
 \subsection{Results}
@@
 Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had
-estimated $\alpha$ values near unity for the prior allowing $m = 1$ (\gls{MAP}
-estimates [95\% \gls{HPD}] 
-  0.76 
-  [0.07 - 
-   1.34]
+estimated $\alpha$ values near unity for the prior allowing $m = 1$ (posterior
+mean [95\% \gls{HPD}] 
+  0.73 
+  [0.05 - 
+   1.18]
 and
-  0.61 
+  0.55 
   [0.01 -
-   1.02] respectively).
-The MAP estimates did not change appreciably when $m = 1$ was disallowed by the
+   0.99] respectively).
+The estimates did not change appreciably when $m = 1$ was disallowed by the
 prior, although the credible interval of the \textcite{niculescu2015recent}
 data was narrower
-  (0.07 - 
-   1.34).
+  (0.05 - 
+   1.18).
 When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
 low estimated $\alpha$ values
-  (0.31 
+  (0.33 
   [0 - 
-  0.73]
+  0.76]
 and
-  0.38 
-  [0.01 -
-   0.8]). 
-However, the MAP estimates increased when $m = 1$ was not permitted, although
-the HPD intervals remained roughly the same
-  (0.55 
-  [0.01 - 
-  0.93]
+  0.27 
+  [0 -
+   0.59]). 
+However, the estimates increased when $m = 1$ was not permitted, although the
+HPD intervals remained roughly the same
+  (0.58 
+  [0.06 - 
+  0.99]
 and
-  0.5 
-  [0.07 -
-   0.94]).
+  0.48 
+  [0.02 -
+   0.87]).
 The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
 for both priors on $m$
-  (0.5 for $m \geq 1$;
-   0.34 for $m \geq 2$).
+  (0.55 for $m \geq 1$;
+   0.53 for $m \geq 2$).
 However, the confidence interval was much wider when $m = 1$ was allowed
-  ([0.03 -
-    1.61] for $m \geq 1$ vs.
+  ([0 -
+    1.75] for $m \geq 1$ vs.
    [0 -
-    0.73] for $m \geq 2$).
+    1.75] for $m \geq 2$).
 
@@
 intervals compared to the nonzero prior density region
-  (\citeauthor{cuevas2009hiv}, 943 
-  [287 -
-   2093];
-   \citeauthor{niculescu2015recent}, 1032
+  (\citeauthor{cuevas2009hiv}, 701 
+  [289 -
+   1279];
+   \citeauthor{niculescu2015recent}, 747
   [136 - 
-   2851];
-  \citeauthor{li2015hiv}, 1437 
-  [365 -
-   2897];
-   \citeauthor{wang2015targeting}, 872
-  [173 - 
-   2119]).
+   2378];
+  \citeauthor{li2015hiv}, 1390 
+  [310 -
+   2821];
+   \citeauthor{wang2015targeting}, 675
+  [175 - 
+   1400]).
 The \citeauthor{novitsky2014impact} data was the outlier, with a very high
 estimated $I$, and HPD interval spanning almost the entire prior region
-  (5586 
-  [187 -
-   8806]).
+  (5431 
+  [183 -
+   8739]).
 The $I$ estimates and HPD intervals were generally robust to the choice of
@@
 
-The MAP estimate of $m$ was equal to 1 for all but the
+The posterior mean of $m$ was equal to 1 for all but the
 \citeauthor{novitsky2014impact} data, when this value was allowed. However, the
 upper bound of the HPD interval was different for each dataset
-  (\citeauthor{niculescu2015recent}, 5;
-   \citeauthor{wang2015targeting}, 4;
+  (\citeauthor{niculescu2015recent}, 4;
+   \citeauthor{wang2015targeting}, 1;
    \citeauthor{li2015hiv}, 1;
-   \citeauthor{cuevas2009hiv}, 2).
-When $m = 1$ was disallowed, the MAP for all datasets was either 2 or 3, with
-HPD intervals spanning the entire prior region. The estimates for the total
-number of nodes $N$ were largely uninformative for all samples, with almost all
-MAP estimates greater than 7500 and HPD intervals spanning almost the entire
-nonzero prior density region. The only exception was the \citeauthor{li2015hiv}
-data, for which the MAP estimate was lower 
-  (6049)
+   \citeauthor{cuevas2009hiv}, 1).
+When $m = 1$ was disallowed, the estimate for all datasets was either 2 or 3,
+with HPD intervals spanning the entire prior region. The estimates for the
+total number of nodes $N$ were largely uninformative for all samples, with
+almost all posterior mean estimates greater than 7500 and HPD intervals
+spanning almost the entire nonzero prior density region. The only exception was
+the \citeauthor{li2015hiv} data, for which the estimate was lower 
+  (5916)
 when $m = 1$ was allowed.
@@
 estimated values of the power law exponent $\gamma$ for each of the datasets 
-investigated. The results are summarized in \cref{tab:gamma}. All of the
+investigated, using the posterior median estimates with the prior allowing
+\gls{m} = 1. The results are summarized in \cref{tab:gamma}. All of the
 estimated exponents were in the range $2 \leq \gamma \leq 2.5$, which is on the
@@
         \textit{a priori} estimates of BA model parameters. 100 networks were
-        simulated using \textit{MAP} parameter estimates obtained with
+        simulated using posterior mean parameter estimates obtained with
         \software{netabc}. The power law exponent $\gamma$ was estimated for

### 2016-06-07 doc/thesis/discussion.Rnw

@@
+<<setup, include=FALSE>>=
+    source("global.R")
+@
+
 \label{sec:disc}
@@
 
-With respect to the preferential attachment power $\alpha$, the six datasets
-analysed fell into two categories (\cref{fig:abchpd}). First, we estimated a
-preferential attachment power close to 1, indicating linear preferential
-attachment, for the BC data and the outbreaks studied by
-\textcite{niculescu2015recent} and \textcite{wang2015targeting}. These values
-were robust to specifying different priors for $m$. All three datasets were
-sampled from populations in which we would expect a high degree of
-epidemiological relatedness: \textcite{niculescu2015recent} studied a recent
-outbreak among Romanian \gls{IDU}, \citeauthor{wang2015targeting} sampled
-acutely infected MSM in Beijing, China, and the BC data constituted a
-phylogenetic \gls{IDU} cluster. These are all contexts in which we would expect
-some of the assumptions of the BA model, such as a connected network,
-relatively high mean degree, and preferential attachment dynamics, to hold.
+{\color{red}\sout{With respect to the preferential attachment power $\alpha$,
+the six datasets analysed fell into two categories} (\cref{fig:abchpd}).
+\sout{First, we estimated a preferential attachment power close to 1,
+indicating linear preferential attachment, for the BC data and the outbreaks
+studied by} \textcite{niculescu2015recent} and \textcite{wang2015targeting}.
+\sout{These values were robust to specifying different priors for $m$. All three
+datasets were sampled from populations in which we would expect a high degree
+of epidemiological relatedness:} \textcite{niculescu2015recent} \sout{studied a
+recent outbreak among Romanian \gls{IDU},} \citeauthor{wang2015targeting}
+\sout{sampled acutely infected MSM in Beijing, China, and the BC data
+constituted a phylogenetic \gls{IDU} cluster. These are all contexts in which
+we would expect some of the assumptions of the BA model, such as a connected
+network, relatively high mean degree, and preferential attachment dynamics, to
+hold.}
 
-The remaining three datasets (\textcite{cuevas2009hiv, novitsky2014impact,
-li2015hiv}) had estimated values of $\alpha$ below 0.5 when $m = 1$ was
-included in the prior, but these were not robust to changing the prior to
-exclude $m = 1$. For the \citeauthor{cuevas2009hiv} data, model
-misspecification is likely partially responsible. While the authors found that
-a large proportion of the samples were epidemiologically linked, these were
-mainly in small local clusters rather than the single large component
-postulated by the BA model. In addition, the mixed risk groups in the dataset
-would be unlikely to significantly interact, further weakening any global
-preferential attachment dynamics. The dataset studied by
-\textcite{novitsky2014impact} originated from a densely sampled population
-where the predominant risk factor was believed to be heterosexual exposure.
-Although the MAP estimate of $\alpha$ was almost unchanged when the value $m =
-1$ was excluded from the prior, the confidence interval shrank substantially.
+\sout{The remaining three datasets} (\textcite{cuevas2009hiv,
+novitsky2014impact, li2015hiv}) \sout{had estimated values of $\alpha$ below
+0.5 when $m = 1$ was included in the prior, but these were not robust to
+changing the prior to exclude $m = 1$. For the} \citeauthor{cuevas2009hiv} 
+\sout{data, model misspecification is likely partially responsible. While the
+authors found that a large proportion of the samples were epidemiologically
+linked, these were mainly in small local clusters rather than the single large
+component postulated by the BA model. In addition, the mixed risk groups in the
+dataset would be unlikely to significantly interact, further weakening any
+global preferential attachment dynamics. The dataset studied by}
+\textcite{novitsky2014impact} \sout{originated from a densely sampled
+population where the predominant risk factor was believed to be heterosexual
+exposure. Although the MAP estimate of $\alpha$ was almost unchanged when the
+value $m = 1$ was excluded from the prior, the confidence interval shrank
+substantially.}}
 
-{\color{blue}\uline{
-The posterior mean estimates for the real world \gls{HIV} datasets corresponded
-to power law exponents $\gamma$ between 2.09 and 2.49.  TODO
-}}
+<<disc, include=FALSE>>=
+    d <- setDT(collect.data("../../simulations/aggregates/hpd/*"))
+@
+
+For all datasets we examined, the posterior mean estimates for \gls{alpha} were
+sub-linear, ranging from 
+    \Sexpr{d[m_min == 1 & parameter == "alpha", round(min(mean), 2)]}
+to 
+    \Sexpr{d[m_min == 1 & parameter == "alpha", round(max(mean), 2)]}.
+The sub-linearity is consistent with the results of
+\textcite{de2007preferential}, who developed a statistical inference method to
+estimate the parameters of a more sophisticated \acrlong{PA} model
+incorporating heterogeneous node behaviour. They found \gls{alpha} values
+ranging from 0.26 to 0.62, depending on the gender and time period considered. 
+Our estimates of \gls{alpha} for the \textcite{niculescu2015recent} was above
+this range under both priors, as were the estimates for the
+\textcite{wang2015targeting} data and the BC data when \gls{m} = 1 was
+disallowed by the prior. The dataset investigated by
+\textcite{de2007preferential} was derived from a survey of a random sample of
+the Norwegian population, whereas our investigation focused on datasets from
+known phylogenetic or geographic clusters of \gls{HIV} infected persons. It is
+therefore unsurprising that we detected stronger \acrlong{PA} dynamics in some
+cases. For instance, random sampling is much less likely to discover the
+high-degree nodes characterizing the tail of the degree distribution, simply
+because those individuals are rare in the general population. In addition, it
+is plausible that \gls{HIV}-positive individuals are more likely to be highly
+connected in their sexual networks, as the odds of acquiring \gls{HIV} increase
+with the number of unprotected sexual contacts.
+
+Both \textcite{de2007preferential} and \textcite{novitsky2014impact} studied
+populations whose primary risk factor for \gls{HIV} infection was heterosexual
+contact. \citeauthor{de2007preferential} explicitly excluded reported
+homosexual contacts; \citeauthor{novitsky2014impact} did not, but noted that
+heterosexual contact is the primary mode of transmission in Botswana where the
+study was done. In the first of the two papers describing the Botswana
+study~\autocite{novitsky2013phylogenetic}, the authors noted that their sample
+was gender-biased, being composed of approximately 75\% women. Our estimate of
+\gls{alpha} for these data was X or X, depending on the prior on \gls{m};
+\citeauthor{de2007preferential} estimated 0.54, 0.57, and 0.29 for 3-year,
+5-year, and lifetime partnership networks respectively for the female portion
+of their sample.
+
+For both choices of prior on \gls{m}, the datasets derived from \gls{IDU}
+populations had a higher estimated \acrlong{PA} power than the other datasets
+(\cref{fig:abchpd,fig:abchpdm2}). This finding is in line with
+\textcite{dombrowski2013topological}, who reanalyzed a network of \glspl{IDU}
+in Brooklyn, USA, collected between 1991 and
+1993~\autocite{friedman2006social}. They found that the the \gls{IDU} network 
+resembled a \gls{BA} network much more closely than other social and sexual 
+networks, and offered sociological explanations for the apparent \acrlong{PA}
+dynamics in this population. Importantly, from a public health perspective,
+the authors asserted that the removal of \emph{random} individuals from
+\gls{IDU} networks may have the paradoxical effect of increasing the network's
+epidemic susceptibility. When low-degree nodes are removed, as would occur
+during a police crackdown, their network neighbours may turn to well-known
+community members for advice or supplies, thus increasing the connectivity of
+these high-degree nodes.
 

### 2016-06-07 doc/thesis/intro.tex

@@
 trees. In particular, for a network model with parameters $\theta$ and an input
-transmission tree $T$, we are interested in the \emph{posterior distribution}}
+transmission tree $T$, we are interested in the posterior distribution}
 \begin{align}
-  \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)}.
+  \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)},
   \label{eq:post}
 \end{align}
-\uline{As outlined in} \cref{sec:obj}, \uline{both the likelihood $\Pr(T \mid
-\theta)$ and the normalizing constant $\Pr(T)$ are likely computationally
-intractable. Hence, rather than computing the posterior distribution
-analytically, we will approximate it using a \defn{Monte Carlo} approach.
-The fundamental idea behind Monte Carlo methods is succinctly expressed by
-\textcite{liu2001theoretical}}:
+where we have adopted the common abuse of the symbol $\Pr$ to refer to a
+probability density. \uline{As alluded to in} \cref{sec:obj}, \uline{both the
+likelihood $\Pr(T \mid \theta)$ and the normalizing constant $\Pr(T)$ are
+likely computationally intractable (this will be explained further in}
+\cref{subsec:just}\uline{). Hence, rather than computing the posterior
+distribution analytically, we will approximate it using a \defn{Monte Carlo}
+approach. The fundamental idea behind Monte Carlo methods is succinctly
+expressed by \textcite{liu2001theoretical}}:
 \begin{quote}
@@
 Monte Carlo methods will be useful in this work is that algorithms exist for
-obtaining samples from distributions, including as the aforementioned
-posterior, that are analytically intractable and from which direct sampling is
-not possible } (for a review see \autocite{robert2004monte}).
-\uline{\Gls{SMC}}~\autocite{doucet2000sequential, doucet2001introduction,
-liu2008monte} \uline{is one such algorithm.}
+obtaining samples from distributions that are analytically intractable and from
+which direct sampling is not possible } (for a review see
+\autocite{robert2004monte}). \uline{\Gls{SMC}}~\autocite{doucet2000sequential,
+doucet2001introduction, liu2008monte} \uline{is one such algorithm.}
 
@@
 the weighted population was a reasonable representation of $\Pr(\theta \mid
-T)$. This turns out to be not straightforward in our case, as we will address
-in} \cref{sec:abc}. \uline{For now, nothing is lost by \ldots}}
-
-{\color{blue}\uline{
-The goal of this section of the introduction is to describe an algorithm called
-the \gls{SMC} sampler~\autocite{del2006sequential}, which forms the basis of
-the adaptive \gls{ABC}-\gls{SMC} algorithm we apply toward the main objective
-of this work. We begin by describing \gls{SIS}, which is a precursor to
-\gls{SMC} that samples from a sequence of distributions defined on spaces of
+T)$. In our case, due to the intractable likelihood, we will need to consider
+an approximation to the posterior } (see \cref{sec:abc}). \uline{However, for
+now, nothing is lost by assuming that our target distribution is the posterior
+itself.}
+
+\uline{The goal of this section of the introduction is to describe an algorithm
+called the \gls{SMC} sampler~\autocite{del2006sequential}, which forms the
+basis of the adaptive \gls{ABC}-\gls{SMC} algorithm we apply toward the main
+objective of this work. We begin by describing \gls{SIS}, which is a precursor
+to \gls{SMC} that samples from a sequence of distributions defined on spaces of
 increasing dimension. We then describe \gls{SMC} itself, which extends
@@
 sequences of distributions all defined on the same space. This terminology will
-become clear as the methods are described.
-}}
-
-\Gls{SMC} is the name for a family of statistical inference methods that rely
-on approximating probability distributions of interest with large collections
-of \defn{particles}, here denoted
-$\set{x^{(k)}}$~\autocite{doucet2000sequential, doucet2001introduction,
-liu2008monte}. These collections or \defn{populations} are constructed to form
+become clear as the methods are described. }}
+
+{\color{red}\sout{Sequential Monte Carlo (SMC) is the name for a family of
+statistical inference methods that rely on approximating probability
+distributions of interest with large collections of \defn{particles}, here
+denoted $\set{x^{(k)}}$}\autocite{doucet2000sequential, doucet2001introduction,
+liu2008monte}. \sout{These collections or \defn{populations} are constructed to form
 a \defn{Monte Carlo approximation} to some distribution of interest $\pi$,
@@
 distribution to $\pi$ as the population size gets
-large~\autocite{liu2001theoretical}. The word \defn{sequential} is used because
+large}\autocite{liu2001theoretical}. \sout{The word \defn{sequential} is used because
 the particle populations are modified in an iterative fashion over time, for
-example, to incorporate new evidence.
+example, to incorporate new evidence.}}
 
@@
 centred at $x$ {\color{blue}\uline{from which $x'$ is drawn}} (see
-\cref{subsec:mfit}).
+\cref{chp:prelim}).
 
@@
 sample from $\pi_i$, by considering only the $i$th component of $\vec{x_i}$.
-These are all the ingredients we need to apply \gls{SIS}. The sequences of
-kernels $L$ and $K$ should be chosen based on the problem at hand to minimize
-the variance in the importance weights as well as possible. For a fixed choice
-of $K_i$, the backward kernels $L_i$ which minimize this variance are called
-the \defn{optimal} backward kernels. The full \gls{SMC} sampler algorithm is
-presented as \cref{alg:smcsamp}. A resampling step is applied whenever the
-\gls{ESS} of the population drops too low, as discussed in the previous
-section.
+{\color{blue}\uline{In fact, since the weight update} \cref{eq:smcwt} 
+\uline{depends only on the $i$th and $i-1$st components of each particle, we do
+not even need to keep track of the complete particles if we are only interested
+in the final distribution.}} These are all the ingredients we need to apply
+\gls{SIS}. The sequences of kernels $L$ and $K$ should be chosen based on the
+problem at hand to minimize the variance in the importance weights as well
+as possible. For a fixed choice of $K_i$, the backward kernels $L_i$ which
+minimize this variance are called the \defn{optimal} backward kernels. The full
+\gls{SMC} sampler algorithm is presented as \cref{alg:smcsamp}. A resampling
+step is applied whenever the \gls{ESS} of the population drops too low, as
+discussed in the previous section.
 
@@
 \glsreset{ABC}
-{\color{blue}\uline{\acrlong{SMC}, and the \gls{SMC} sampler, were developed
+
+\subsection{Overview and motivation}
+\label{subsec:abcoverview}
+
+{\color{blue}\uline{\Acrlong{SMC}, and the \gls{SMC} sampler, were developed
 for sampling from distributions which can be evaluated up to a normalizing
-constant. TODO }
-
-\uline{This section is organized as follows. First, we formalize the concept of a
-mathematical model and discuss what it means to ``fit'' such models to data 
-and how this is typically accomplished. Second, we introduce \gls{ABC}, which
-was designed for situations like the one we face in this work where the most
-commonly used means of fitting models cannot be applied. Finally, we describe
-algorithms which can be used to implement \gls{ABC}, including a variant of
-\gls{SMC}.}}
-
-\subsection{Overview of ABC}
-\label{subsec:abcoverview}
-
-Most mathematical models are amenable to fitting via one or both of the
-approaches, \gls{ML} or Bayesian inference, discussed above. However, there are
-some, particularly in the domain of population
-genetics~\autocite{beaumont2002approximate, beaumont2010approximate}, for which
-calculation of either the likelihood or the product of the likelihood and the
-prior may be infeasible. For example, one or both of these quantities may be
-expressible only as an intractable integral. \Gls{ABC} is designed for such
-cases, where standard likelihood-based techniques for model fitting cannot be
-applied.
-
-Ordinarily, Bayesian inference targets the posterior distribution $f(\theta
-\mid y)$. That is, in the Bayesian framework, model parameters with higher
-posterior density are ``better'' in the sense that they offer a more credible
-explanation for the observed data. Approximate Bayesian computation offers an
+constant. We claim, and shall argue more thoroughly below}
+(\cref{subsec:just}), \uline{that the posterior distribution}
+\[
+  \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)}
+\]
+\uline{for a contact network model with parameters $\theta$ and an input
+transmission tree $T$ does not fall in this category (note that, as in the
+previous section, we are using $\Pr$ to denote a probability density).
+Therefore, \gls{SMC}, and other Bayesian and \gls{ML} techniques for fitting
+mathematical models} (see \cref{chp:prelim}), \uline{cannot be directly
+applied to our problem. In particular, \gls{MCMC} and the \gls{SMC} sampler are
+designed for distributions $\pi$ which can be evaluated up to a normalizing
+constant $Z$, that is, $\pi(x) = \gamma(x) / Z$. Both algorithms calculate the
+ratio $\pi(x) / \pi(x') = \gamma(x) / \gamma(x')$ for a current value $x$ and
+proposed updated value $x'$ - for \gls{MCMC}, this is part of the \acrlong{MH}
+ratio, while for the \gls{SMC} sampler, it is required to calculate the
+importance weights. In the context of Bayesian inference, this ratio is a
+likelihood ratio, which must be calculated by computing the individual
+likelihoods and dividing them. If the likeilhood is intractible, this is
+clearly not a viable approach.}
+
+\uline{\Gls{ABC}}~\autocite{rubin1984bayesianly, tavare1997inferring,
+fu1997estimating}\uline{ was developed to estimate posterior distributions with
+intractable likelihoods, which have arisen frequently in the domain of
+population genetics} \autocite{beaumont2002approximate,
+beaumont2010approximate}. \uline{\Gls{ABC} navigates around the intractable
+likelihood by replacing the posterior as the target of inference by an
+\emph{approximate} posterior. This distribution is constructed in such a way
+that the ratios required for \gls{MCMC} and the \gls{SMC} sampler can be
+computed. This conveniently allows us to apply the existing \gls{MCMC} and
+\gls{SMC} algorithms with minimal changes. In the next section, we shall
+demonstrate how this is done, but first we give the definition of the
+approximate posterior. }}
+
+{\color{red}\sout{Most mathematical models are amenable to fitting via one or
+both of the approaches, \gls{ML} or Bayesian inference, discussed above.
+However, there are some, particularly in the domain of population
+genetics}~\autocite{beaumont2002approximate, beaumont2010approximate},
+\sout{for which calculation of either the likelihood or the product of the
+likelihood and the prior may be infeasible. For example, one or both of these
+quantities may be expressible only as an intractable integral. \Gls{ABC} is
+designed for such cases, where standard likelihood-based techniques for model
+fitting cannot be applied.}}
+
+{\color{red}\sout{Ordinarily, Bayesian inference targets the posterior
+distribution $\Pr(\theta \mid y)$. That is, in the Bayesian framework,}} 
+{\color{blue}\uline{By targeting the posterior distribution, Bayesian inference
+makes the assertion that}} model parameters with higher posterior density are
+``better'' in the sense that they offer a more credible explanation for the
+observed data. The approximate posterior targeted by \gls{ABC} uses an
 alternative metric for parameter credibility, namely the similarity of
@@
 approximation to the real-world process generating the observed data. More
-formally, suppose we have a distance measure $\rho$ defined on the space of all
-possible data our model could generate. \gls{ABC} aims to sample from the joint
-posterior distribution of model parameters and simulated datasets $z$ which are
-within some small distance $\varepsilon$ of the observed data $y$,
+formally, {\color{blue}\uline{let $y$ be the observed data to which we are
+trying to fit a model with parameters $\theta$. In the case of this work, the
+data is a transmission tree $T$, but we shall stick with the generic variable
+$y$ for now.}} Suppose we have a distance measure $\rho$ defined on the space
+of all possible data our model could generate. \gls{ABC} aims to sample from
+the joint posterior distribution of model parameters and simulated datasets $z$
+which are within some small distance $\varepsilon$ of the observed data $y$,
 \[
   \pi_{\varepsilon}(\theta, z \mid y) =
-  \frac{\pi(\theta) f(z \mid \theta) \I_{A_{\varepsilon, y}} (z)}
-  {\int_{A_{\varepsilon, y} \times \Theta} \pi(\theta) f(z \mid \theta) \d \theta}.
+  \frac{\pi(\theta) \Pr(z \mid \theta) \I_{A_{\varepsilon, y}} (z)}
+  {\int_{A_{\varepsilon, y} \times \Theta} \pi(\theta) \Pr(z \mid \theta) \d \theta}.
 \]
@@
 respect to $\rho$, $\Theta$ is the space of all possible model parameters, and
-$\I$ is the indicator function~\autocite{marin2012approximate}. As we shall
-see in the next section, this distribution can be sampled from exactly. The
-word ``approximate'' derives from the assumption that, for a suitably chosen
-distance $\rho$ and a small enough $\varepsilon$, the marginal in $z$ of
-this distribution approximates the posterior of
+$\I$ is the indicator function~\autocite{marin2012approximate}.
+{\color{blue}\uline{The term $\Pr(z \mid \theta)$ appears to be the bothersome
+likelihood again, but this will turn out not to be a problem because we are
+simulating $z$ ourselves. In fact, it is possible to sample from this
+distribution exactly (see next section).}}
+
+{\color{blue}\uline{To return to the context of this thesis, the observed data
+$y$ is an estimated transmission tree for a viral epidemic under investigation.
+The model in question is a contact network model with parameters $\theta$. $z$
+is a simulated transmission tree, obtained by first generating a contact
+network under the model, and then simulating the spread of an epidemic over
+that network. A transmission tree can be constructed by keeping track of who
+infected whom during the simulated epidemic (further details will be given in
+} \cref{subsec:nettree} \uline{). The distance function $\rho$ must compare two
+transmission trees - the observed tree $y$, and the simulated tree $z$. We will
+define this distance function using the tree kernel discussed in}
+\cref{subsec:treeshape}. \uline{In words, the approximate posterior we consider
+here is a distribution which assigns a joint probability density to model
+parameters and simulated transmission trees under those parameters. The
+probability density is proportional to the product of the prior on the
+parameters, and the likelihood of the simulated transmission tree under those
+parameters, but only if the simulated transmission tree is sufficiently close
+to the true tree. Otherwise, the probability density is zero. }}
+
+{\color{red}\sout{As we shall see in the next section, this distribution can
+be sampled from exactly. The word ``approximate'' derives from the assumption
+that,}}
+for a suitably chosen distance $\rho$ and a small enough $\varepsilon$, the
+marginal in $z$ of this distribution approximates the posterior of
 interest~\autocite{marin2012approximate}. That is,
@@
 hurdles~\autocite{marjoram2003markov}. The approach is similar to ordinary
-Bayesian \gls{MCMC} (\cref{subsec:mfit}), except that a distance cutoff
+Bayesian \gls{MCMC} (\cref{chp:prelim}), except that a distance cutoff
 replaces the likelihood ratio. That is, the transition probability between
@@
       \State Propose $\theta'$ according to $q(\theta, \theta')$
-      \State Simulate a dataset $\vec{z'}$ according to the model with
-             parameters $\theta$
+      \State Simulate a dataset $z'$ according to the model with parameters
+             $\theta$
       \State Accept $\theta \gets \theta'$ with probability
@@
 through a sequence of target distributions which ends at the \gls{ABC}
-approximation to the posterior. As discussed in \cref{subsec:smcsamp}, the
+approximation to the posterior. {\color{blue}\uline{The initial importance
+distribution, denoted as $\eta_0$ in} \cref{alg:smcsamp}, \uline{is taken to be}
+$\Pr(\theta) \Pr(z \mid \theta)$. TODO}
+
+As discussed in \cref{subsec:smcsamp}, the
 choices of the kernels $K$ and $L$ is problem-specific, and so appropriate
@@
 del2012adaptive}.
+
+\begin{algorithm}
+  \caption{\gls{ABC}-\gls{SMC}.}
+  \begin{algorithmic}
+    \For {$k = 1$ to $n$}
+      \State Sample $x_1^{(k)}$ from $\eta_1(x_1)$
+      \Comment{Initialize the $k$th particle}
+      \State $w^{(k)} \gets \gamma_1(x_1^{(k)}) / \eta_1(x_1^{(k)})$
+      \State Normalize the weights so that $\sum w = 1$
+    \EndFor
+    \For {$i = 2$ to $d$}
+      \For {$k = 1$ to $n$}
+        \State Sample $x_i^{(k)}$ from $K(x_{i-1}^{(k)}, x_i)$
+        \Comment Extend the $k$th particle
+        \State $w^{(k)} \gets w^{(k)} \cdot \dfrac{\gamma_i(x_i) L_{i-1}(x_i, x_{i-1})}{\gamma_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}$
+      \EndFor
+      \State Normalize the weights so that $\sum w = 1$
+      \If {$\ESS(w) < T$}
+        \Comment{$T$ is a user-defined threshold}
+        \State Resample the particles according to $w$
+        \For {$k = 1$ to $n$}
+          \State $w^{(k)} \gets 1/n$
+        \EndFor
+      \EndIf
+    \EndFor
+  \end{algorithmic}
+  \label{alg:abcsmc}
+\end{algorithm}
 

### 2016-06-07 doc/thesis/methods_results.Rnw

@@
 simulations
-    (\Sexpr{round(d[,sum(I < true_I)] * 100 / nrow(d))}\%).
+    (\Sexpr{round(d[,sum(I > true_I)] * 100 / nrow(d))}\%).
 The estimates for \gls{m} and \gls{N} were similar across all simulations
@@
 Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had
-estimated $\alpha$ values near unity for the prior allowing $m = 1$ (\gls{MAP}
-estimates [95\% \gls{HPD}] 
+estimated $\alpha$ values near unity for the prior allowing $m = 1$ (posterior
+mean [95\% \gls{HPD}] 
   \Sexpr{pmean1["niculescu2015", round(alpha, 2)]} 
@@
    \Sexpr{up1["wang2015", round(alpha, 2)]}] respectively).
-The MAP estimates did not change appreciably when $m = 1$ was disallowed by the
+The estimates did not change appreciably when $m = 1$ was disallowed by the
 prior, although the credible interval of the \textcite{niculescu2015recent}
@@
    \Sexpr{up1["cuevas2009", round(alpha, 2)]}]). 
-However, the MAP estimates increased when $m = 1$ was not permitted, although
-the HPD intervals remained roughly the same
+However, the estimates increased when $m = 1$ was not permitted, although the
+HPD intervals remained roughly the same
   (\Sexpr{pmean2["li2015", round(alpha, 2)]} 
@@
 
-The MAP estimate of $m$ was equal to 1 for all but the
+The posterior mean of $m$ was equal to 1 for all but the
 \citeauthor{novitsky2014impact} data, when this value was allowed. However, the
@@
    \citeauthor{cuevas2009hiv}, \Sexpr{up1["cuevas2009", round(m)]}).
-When $m = 1$ was disallowed, the MAP for all datasets was either 2 or 3, with
-HPD intervals spanning the entire prior region. The estimates for the total
-number of nodes $N$ were largely uninformative for all samples, with almost all
-MAP estimates greater than 7500 and HPD intervals spanning almost the entire
-nonzero prior density region. The only exception was the \citeauthor{li2015hiv}
-data, for which the MAP estimate was lower 
+When $m = 1$ was disallowed, the estimate for all datasets was either 2 or 3,
+with HPD intervals spanning the entire prior region. The estimates for the
+total number of nodes $N$ were largely uninformative for all samples, with
+almost all posterior mean estimates greater than 7500 and HPD intervals
+spanning almost the entire nonzero prior density region. The only exception was
+the \citeauthor{li2015hiv} data, for which the estimate was lower 
   (\Sexpr{pmean1["li2015", floor(N)]})

### 2016-06-08 doc/thesis/discussion.Rnw
Up to ABC in intro now

@@
 with the settings described in the methods section. Implementing
-parallelization using MPI, rather than POSIX threads as we have done here,
-would allow the program to be run over a larger number of cores on multiple
-CPUs in parallel.
+parallelization using \gls{MPI}, rather than \gls{POSIX} threads as we have
+done here, would allow the program to be run over a larger number of cores on
+multiple CPUs in parallel.
 
@@
 was gender-biased, being composed of approximately 75\% women. Our estimate of
-\gls{alpha} for these data was X or X, depending on the prior on \gls{m};
-\citeauthor{de2007preferential} estimated 0.54, 0.57, and 0.29 for 3-year,
-5-year, and lifetime partnership networks respectively for the female portion
-of their sample.
+\gls{alpha} for these data was 
+    \Sexpr{d[m_min == 1 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]}
+        or 
+    \Sexpr{d[m_min == 2 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]},
+depending on the prior on \gls{m}; \citeauthor{de2007preferential} estimated
+0.54, 0.57, and 0.29 for 3-year, 5-year, and lifetime partnership networks
+respectively for the female portion of their sample.
 
@@
 these high-degree nodes.
+
+One somewhat surprising result was the difference between parameter estimates 
+for the \textcite{li2015hiv} and \textcite{wang2015targeting} datasets. Both
+groups studied cohorts of acutely infected \gls{MSM} in major Chinese cities
+(Shanghai and Beijing respectively); yet, the \citeauthor{li2015hiv} data was 
+estimated to have a lower \acrlong{PA} power and larger infected population
+than the \textcite{wang2015targeting} data\ldots
+
+% for discussion
+In order to compare our results to existing literature on networks and
+distributions of partner counts, we have reported estimated values for the
+power law exponent \gls{gamma} of the real data sets we evaluated. However, the
+posterior means for \gls{alpha} for all six datasets were less than one; the
+degree distributions in this parameter range are stretched exponential, not
+power law~\autocite{krapivsky2000connectivity}. As we show in
+\cref{fig:powerlaw}, the power law fit does capture the slope of the degree
+distribution fairly well, but the results should still be interpreted
+cautiously. \textcite{krapivsky2000connectivity} showed that the power law
+distribution can be maintained, with \gls{gamma} tuned to any desired value, by
+a straightforward modification of the \gls{BA} model. The authors define the
+``connection kernel'' $A_k$ the probability of a new connection to a node of
+degree $k$, up to a normalizing constant. In the \gls{BA} model as we have
+presented it here, $A_k = k^{\alpha} + 1$. Taking $A_k$ as any asymptotically
+linear function will result in a power law distribution, with the exponent
+\gls{gamma} determined by the properties of $A_k$. Implementing such a model
+would be straightforward and seems a natural next step toward improving the
+realism of the \gls{BA} model.
+
+\textcite{rothenberg2007large} estimated the power law exponents for 15 major
+network studies of sexual and injection drug use networks carried out between
+1981 and 2000. The estimated \gls{gamma} values for ``all contacts'' (that is,
+including both interviewed and non-interviewed individuals) were 
+between 2.08 and 2.87 for all but one of the networks\ldots
 

### 2016-06-08 doc/thesis/discussion.tex
Up to ABC in intro now

@@
 was gender-biased, being composed of approximately 75\% women. Our estimate of
-\gls{alpha} for these data was X or X, depending on the prior on \gls{m};
-\citeauthor{de2007preferential} estimated 0.54, 0.57, and 0.29 for 3-year,
-5-year, and lifetime partnership networks respectively for the female portion
-of their sample.
+\gls{alpha} for these data was 
+    0.55
+        or 
+    0.53,
+depending on the prior on \gls{m}; \citeauthor{de2007preferential} estimated
+0.54, 0.57, and 0.29 for 3-year, 5-year, and lifetime partnership networks
+respectively for the female portion of their sample.
 

### 2016-06-08 doc/thesis/intro.tex
Up to ABC in intro now

@@
 other nodes in the graph. In the original formulation, the partners of the new
-node are chosen with probability linearly proportional to their degree.
-However, \citeauthor{barabasi1999emergence} suggest extending the model such
-that the probability of choosing a partner of degree $d$ is proportional to
-$d^\alpha + 1$ for some constant $\alpha$, and we use this extension here. When
-$m = 1$, the network takes on the distinctive shape of a tree, that is, it does
-not contain any cycles. Cycles are present in the network for all other $m$
-values. Examples of \gls{BA} networks with three different values of the
+node are chosen with probability linearly proportional to their degree plus
+one.
+
+There has been some contention over the idea that contact networks are
+scale-free. \textcite{handcock2004likelihood} fit several stochastic models of
+partner formation to empirical degree distributions derived from population
+surveys of sexual behaviour. They found that a negative binomial distribution,
+rather than a power law, was the best fit to five out of six datasets, although
+the difference in goodness of fit was extremely small in four out of these
+five. \textcite{bansal2007individual} found that an exponential distribution,
+rather than a power law, was the best fit to degree distributions of six social
+and sexual networks. \textcite{drombowski2013} contend that sexual networks are
+shaped more by homophily (``like attract like'') than by preferential
+attachment, but find that \gls{IDU} network do demonstrate a scale-free
+structure.
+
+In the paper describing the \gls{BA} model, \citeauthor{barabasi1999emergence}
+suggest an extension where the probability of choosing a partner of degree $d$
+is proportional to $d^\alpha + 1$ for some constant \gls{alpha}.
+{\color{blue}\uline{When $\gls{alpha} \neq 1$, the degree distribution no
+longer follows a power law~\autocite{krapivsky2000connectivity}. For
+$\gls{alpha} < 1$, the distribution is a stretched exponential, meaning that
+the number of nodes of degree $k$ is proportional to $\exp(-k^\beta)$ for some
+constant $\beta$. For $\gls{alpha} > 1$, the distribution takes on a
+characteristic called \emph{gelation}, where a one or a few high-degree hub
+nodes are connected to nearly every other node in the graph. We do not believe
+these departures from the power law affect the applicability of the model to
+real world networks. In fact,} \textcite{de2007preferential} \uline{were able to
+estimate the \acrlong{PA} power from partner count data collected from the same
+individuals for consecutive time intervals, and found a value less than one in
+all cases. It is also worth noting that, in addition to the \gls{BA} model,
+other investigations of the interaction between contact networks and
+transmission trees have studied the \acrlong{ER} and \acrlong{WS}
+models~\autocite{leventhal2012inferring}, whose degree distributions do not
+generally follow a power law under any parameter settings.}}
+
+When $m = 1$, the network takes on the distinctive shape of a tree, that is, it
+does not contain any cycles. Cycles are present in the network for all other
+$m$ values. Examples of \gls{BA} networks with three different values of the
 preferential attachment power \gls{alpha} are shown in \cref{fig:baeg}.
@@
 \end{figure}
-
-There has been some contention over the idea that contact networks are
-scale-free. \textcite{handcock2004likelihood} fit several stochastic models of
-partner formation to empirical degree distributions derived from population
-surveys of sexual behaviour. They found that a negative binomial distribution,
-rather than a power law, was the best fit to five out of six datasets, although
-the difference in goodness of fit was extremely small in four out of these
-five. \textcite{bansal2007individual} found that an exponential distribution,
-rather than a power law, was the best fit to degree distributions of six social
-and sexual networks.
 
@@
 \end{align}
-where we have adopted the common abuse of the symbol $\Pr$ to refer to a
-probability density. \uline{As alluded to in} \cref{sec:obj}, \uline{both the
+\uline{where we have adopted the common abuse of the symbol $\Pr$ to refer to a
+probability density. As alluded to in} \cref{sec:obj}, \uline{both the
 likelihood $\Pr(T \mid \theta)$ and the normalizing constant $\Pr(T)$ are
@@
 sampling points one at a time from the target distribution, \gls{SMC} considers
-a population of points or ``particles'', here denoted $\set{x^{(k)}}$. The
-particles are associated with weights, $\set{w(x^{(k)})}$. These weighted
-particles are a representation for the target distribution. For example, the 
-expected value of the distribution is approximated by}
+a population of points or ``particles'', here denoted $\set{x^{(k)}}$ and
+indexed by an integer $k$. The particles are associated with weights,
+$\set{w(x^{(k)})}$. These weighted particles are a representation for the
+target distribution. For example, the expected value of the distribution is
+approximated by}
 \[
@@
   \pi_{\varepsilon}(\theta, z \mid y) =
-  \frac{\pi(\theta) \Pr(z \mid \theta) \I_{A_{\varepsilon, y}} (z)}
-  {\int_{A_{\varepsilon, y} \times \Theta} \pi(\theta) \Pr(z \mid \theta) \d \theta}.
+  \frac{\Pr(\theta) \Pr(z \mid \theta) \I_{A_{\varepsilon, y}} (z)}
+  {\int_{A_{\varepsilon, y} \times \Theta} \Pr(\theta) \Pr(z \mid \theta) \d \theta}.
 \]
@@
 parameters, but only if the simulated transmission tree is sufficiently close
-to the true tree. Otherwise, the probability density is zero. }}
+to the true tree. Otherwise, the probability density is zero.}}
 
@@
 that,}}
+{\color{blue}\uline{It is not yet obvious that the \gls{ABC} target
+distribution just presented can in fact be used to approximate the true
+posterior. TODO }}
+
 for a suitably chosen distance $\rho$ and a small enough $\varepsilon$, the

### 2016-06-08 doc/thesis/preamble.tex
Moar
Formatting for GPS
Underline changes for printing
Started working on Sally's revisions
More of Sally's revisions
More of Sally's changes
More exposition for ABC; rewrote discussion of real data
Up to ABC in intro now

@@
 \usepackage[dvipsnames]{xcolor}
+\usepackage[titletoc]{appendix}
+\usepackage{tocloft}
+\usepackage[normalem]{ulem}
 
@@
 \newcommand{\eg}{\textit{e.g.}}
+\newcommand{\ie}{\textit{i.e.}}
 
@@
 \newcommand{\X}{\mathcal{X}}
+\newcommand{\E}{\mathbb{E}}
 \renewcommand{\L}{\mathcal{L}}
@@
 \DeclareMathOperator{\Uniform}{Uniform}
-\DeclareMathOperator{\E}{E}
 \DeclareMathOperator{\Var}{Var}
@@
 \DeclareMathOperator{\nc}{nc}
+\DeclareMathOperator{\nl}{nl}
 
@@
 \newacronym{IDU}{IDU}{injection drug users}
+\newacronym{ANOVA}{ANOVA}{analysis of variance}
+\newacronym{MANOVA}{MANOVA}{multivariate analysis of variance}
+\newacronym{AIC}{AIC}{Akaike information criterion}
+\newacronym{POSIX}{POSIX}{Portable Operating System Interface}
+\newacronym{IQR}{IQR}{interquartile range}
+\newacronym{MPI}{MPI}{message passing interface}
+\newacronym{PA}{PA}{preferential attachment}
 
 \newglossaryentry{alpha}{
-  name=$\alpha$,
+  name=\ensuremath{\alpha},
   description=preferential attachment power parameter in Barab\'asi-Albert networks
@@
 }
+\newglossaryentry{alphaess}{
+  name=$\alpha_{\ESS}$,
+  description=per-iteration rate of expected sample size decay in adaptive ABC-SMC
+}
 

### 2016-06-09 doc/thesis/methods_results.Rnw
Updated grid search results

@@
 of the \gls{kSVR} classifiers.}}\Cref{fig:gridest} shows point estimates and
-95\% highest density intervals for each of the \gls{BA} parameters, for one
-replicate experiment with 500-tip trees. Plots showing the point estimates for
-all replicates can be found in
+{\color{blue}\uline{50\%}} and 95\% highest density intervals for each of the
+\gls{BA} parameters, for one replicate experiment with 500-tip trees. Plots
+showing the point estimates for all replicates can be found in
 \cref{fig:gridptalpha,fig:gridptI,fig:gridptm,fig:gridptN}. For all parameters
@@
 
-{\color{blue}\uline{Similar to the results obtained in the previous experiment,
-the \gls{alpha} and \gls{I} parameters were estimated with comparably high
-accuracy, while the other two parameters were less identifiable. Across all
-replicates, $R^2$ values for the correlations between the estimated and true
-values were
+{\color{blue}\uline{Across all replicates, $R^2$ values for the correlations
+between the estimated and true values were
     $\Sexpr{tests["alpha", round(rsquared, 2)]}$,
@@
 
-\uline{Qualitatively, three of the parameters (excluding \gls{N}) exhibited
-weak identifiability within particular sections of the grid. The \glspl{IQR}
-for \gls{alpha} were nearly identical for the values \gls{alpha} $\leq 1$,
-spanning approximately [0-1.3]. Similarly, the \glspl{IQR} for \gls{alpha} $>
-1$ were all approximately [0.7-2]} (\cref{fig:gridest}). \uline{The
-distributions of kernel scores for these groups of values were also
-qualitatively similar} (\cref{fig:gridalpha}), \uline{although the value
-\gls{alpha} = 1.25 exhibited a more pronounced peak in kernel scores around the
-true value.}
-
-\uline{For \gls{I}, the \glspl{IQR} were nearly identical for the values
-\gls{I} $\leq 2000$ (approximately [500,3800]), and for the values \gls{I} $>
-3000$ (approximately [1600-5000]). Kernel score distributions for all \gls{I}
-values exibited a similar rounded shape} (\cref{fig:gridI}), \uline{although
-this was most pronounced for values near the upper and lower extremes of the
-grid.}
-
-\uline{The \glspl{IQR} for \gls{m} were distinct for \gls{m} = 1 and 2, but
-were similar for all values of \gls{m} $> 2$ (approximately [2-6]). The kernel
-score distribution for \gls{m} = 1 was distinct from the others, having a large
-peak at \gls{m} = 1, while the others exhibited a valley at \gls{m} = 1 and
-were mostly flat everywhere else} (\cref{fig:gridm}).
-
-\uline{The \glspl{IQR} for \gls{N} did not form any obvious groups like the
-other parameters' \glspl{IQR}} (\cref{fig:gridest}). \uline{The kernel score
-distribution for \gls{N} = 1000 was distinct, having a peak at the true value}
-(\cref{fig:gridN}). \uline{The distributions for \gls{N} $> 1000$ were all
-relatively similar, having a valley at the low end of the grid and being mostly 
-flat everywhere else.}}
+\uline{Qualitatively, each of the parameters exhibited weak identifiability
+within particular sections of the grid} (\cref{fig:gridest}). \uline{The 50\%
+\glspl{HDI} for \gls{alpha} were nearly identical for the values $\gls{alpha}
+\leq 0.5$ (on average [X-X]) and for $\gls{alpha} \geq 1.5$ ([X-X]).
+For \gls{I}, similar 50\% \gls{HPD} were observed for $\gls{I} \leq 1500$
+(average [X-X]) and for $\gls{I} > 3000$ ([X-X]). For \gls{m}, both the 50\%
+and 95\% \glspl{HDI} were generally identical for $\gls{m} > 2$ (average [X-X]
+for 50\% and [X-X] for 95\% \gls{HDI}). We did not observe a similar pattern
+for \gls{N}, however the boundaries of the \gls{HDI} did not always
+monotonically increase with the actual value.}
+
+\Cref{fig:gridalpha,fig:gridI,fig:gridm,fig:gridN} \uline{show kernel score
+distributions of kernel scores along the grid for each parameter. The
+distributions for some values, such as $\gls{alpha} = 1.25$, $\gls{I} = 500$
+and $4500$, $\gls{m} = 1$, and $\gls{N} = 1000$, exhibited distinct peaks
+around the true value. This indicates that these values produce distinctively
+shaped trees that can be identified with the tree kernel, when the other
+parameter values are fixed and known. However, for the majority of values of
+each parameter, the score distributions were fairly flat around the true value. 
+This means there is a range of values which produce similarly shaped trees, and
+the parameter is less identifiable within that range.}}
 

### 2016-06-09 doc/thesis/abstract.tex
Edits and edits

@@
 logarithmic, to 1.00, exactly linear.}} {\color{blue}\uline{Point estimates of
-the strength of preferential attachment were higher in two injection drug user
-populations, potentially indicating that high-degree ''hub'' nodes may play a
+the strength of preferential attachment were higher in injection drug user
+populations, potentially indicating that high-degree ``hub'' nodes may play a
 role in epidemics among this risk group.}} Our results underscore the

### 2016-06-09 doc/thesis/intro.tex
Edits and edits

@@
 which are within some small distance $\varepsilon$ of the observed data $y$,
-\[
+\begin{align}
   \pi_{\varepsilon}(\theta, z \mid y) =
@@
   {\int_{A_{\varepsilon, y} \times \Theta} \Pr(\theta) \Pr(z \mid \theta) \d \theta}.
-\]
+  \label{eq:abctarget}
+\end{align}
 Here, $A_{\varepsilon, y}$ is an $\varepsilon$-ball around $y$ with
 respect to $\rho$, $\Theta$ is the space of all possible model parameters, and
-$\I$ is the indicator function~\autocite{marin2012approximate}.
-{\color{blue}\uline{The term $\Pr(z \mid \theta)$ appears to be the bothersome
-likelihood again, but this will turn out not to be a problem because we are
-simulating $z$ ourselves. In fact, it is possible to sample from this
-distribution exactly (see next section).}}
+$\I$ is the indicator function~\autocite{marin2012approximate}. The
+distribution $\pi_\varepsilon(\theta, z \mid y)$ will be referred to as
+the \textit{\gls{ABC} target distribution}. {\color{blue}\uline{The term $\Pr(z
+\mid \theta)$ appears to be the bothersome likelihood again, but this will turn
+out not to be a problem because we are simulating $z$ ourselves. In fact, it is
+possible to sample from this distribution exactly (see next section).}}
 
@@
 {\color{red}\sout{As we shall see in the next section, this distribution can
-be sampled from exactly. The word ``approximate'' derives from the assumption
-that,}}
-{\color{blue}\uline{It is not yet obvious that the \gls{ABC} target
-distribution just presented can in fact be used to approximate the true
-posterior. TODO }}
-
+be sampled from exactly. The word ``approximate'' derives from assumption that,
 for a suitably chosen distance $\rho$ and a small enough $\varepsilon$, the
 marginal in $z$ of this distribution approximates the posterior of
-interest~\autocite{marin2012approximate}. That is,
-\[
-  \int \pi_\varepsilon(\theta, z \mid y) \d z \approx f(\theta \mid y).
-\]
-The distribution $\pi_\varepsilon(\theta, z \mid y)$ is variously referred to
-as the \textit{\gls{ABC} target distribution} or the \gls{ABC} approximation
-to the posterior. The intuition for why this approximation might be reasonable
+interest~\autocite{marin2012approximate}. That is, }}
+{\color{blue}\uline{In fact, it is not the \gls{ABC} target distribution
+itself, but rather its marginal in $z$, which approximates the posterior
+distribution. In other words, we claim that }} 
+\[
+  \int \pi_\varepsilon(\theta, z \mid y) \d z \approx \Pr(\theta \mid y).
+\]
+The intuition for why this approximation might be reasonable
 comes from the fact that, when $\varepsilon = 0$, the $\varepsilon$-ball around
@@
 
+\subsubsection{Distance functions and summary statistics in ABC}
+
 In many applications {\color{blue}\uline{(eg. \autocite{fu1997estimating,
@@
 \[
-  f(\theta \mid y) = f(\theta \mid S(y)).
+  \Pr(\theta \mid y) = \Pr(\theta \mid S(y)).
 \]
@@
 (\cref{subsec:smcsamp}). The sequence of target distributions is defined as
-$\pi_i = \pi_{\varepsilon_i}(x \mid y)$ for a decreasing sequence of tolerances
-$\varepsilon_i$. The intention is for the algorithm to progress smoothly
-through a sequence of target distributions which ends at the \gls{ABC}
-approximation to the posterior. {\color{blue}\uline{The initial importance
-distribution, denoted as $\eta_0$ in} \cref{alg:smcsamp}, \uline{is taken to be}
-$\Pr(\theta) \Pr(z \mid \theta)$. TODO}
+$\pi_i(x) = \pi_{\varepsilon_i}(x \mid y)$ for a decreasing sequence of
+tolerances $\varepsilon_i$. The intention is for the algorithm to progress
+smoothly through a sequence of target distributions which ends at the \gls{ABC}
+approximation to the posterior. {\color{blue}\uline{The initial value
+$\varepsilon_0$ is set to $\infty$, which makes the first distribution in the
+sequence}
+\[
+  \pi_0(\theta, z) = \frac{\Pr(\theta) \Pr(z \mid \theta)}
+               {\int_{\R \times \Theta} \Pr(\theta) \Pr(z \mid \theta) \d\theta \d z}.
+\]
+\uline{For brevity, we have ommitted the dependence on the observed data $y$.
+In the terminology of} \cref{alg:smc}, \uline{the numerator is the first of the
+$\gamma$'s, that is, $\gamma_0 = \Pr(\theta) \Pr(z \mid \theta)$. Sampling in
+proportion to $\gamma_0$ is straightforward and was already demonstrated for
+rejection \gls{ABC} above. Because the sampling is exact, the initial
+importance weights are all set equal to 1 and normalized to $1/n$ where $n$ is
+the number of particles.}}
 
@@
     \For {$k = 1$ to $n$}
-      \State Sample $x_1^{(k)}$ from $\eta_1(x_1)$
-      \Comment{Initialize the $k$th particle}
-      \State $w^{(k)} \gets \gamma_1(x_1^{(k)}) / \eta_1(x_1^{(k)})$
-      \State Normalize the weights so that $\sum w = 1$
+      \State Sample $x_1^{(k)}$ from $\Pr(\theta) \Pr(z \mid \theta)$
+      \Comment{Draw $\theta$ from the prior and simulate $z$ from $\theta$}
+      \State $w^{(k)} \gets 1/n$
     \EndFor
-    \For {$i = 2$ to $d$}
+    \For {$i = 2$ to $i_{\max}$}
+      \Comment{$i_{\max}$ is a user-specified number of iterations}
       \For {$k = 1$ to $n$}

### 2016-06-09 doc/thesis/methods_results.Rnw
Edits and edits

@@
 examination of the limitations of our approach and opportunities for future
-investigation. }}
+investigation.}}
 
@@
 This means there is a range of values which produce similarly shaped trees, and
-the parameter is less identifiable within that range.}}
+the parameter is less identifiable within that range. The exception was
+\gls{I}, whose score distributions exhibited a more or less rounded shape with
+the highest point near the true value. }}
 
@@
 We downloaded all sequences associated with each published study from GenBank.
-For the \textcite{novitsky2014impact} data, each \textit{env} sequence was
-aligned pairwise to the HXB2 reference sequence (GenBank accession number
-K03455), and the hypervariable regions were clipped out with
-\software{BioPython} version 1.66+~\autocite{cock2009biopython}. Sequences were
-multiply aligned using \software{MUSCLE} version 3.8.31
+{\color{blue}\uline{For the \textcite{niculescu2015recent} data, only sequences
+from \gls{IDU} (whose sequence identifiers included the letters ``DU'') were
+included in the analysis. \textcite{kao2011surveillance} found a very strong
+association in their study population between subtype and risk group - subtype
+B was almost always associated with \gls{MSM}, whereas \gls{IDU} were usually
+infected with a circulating recombinant form. Since there were many more
+subtype B sequences in their data than sequences of other subtypes, we
+restricted our analysis to the subtype B sequences and labelled this dataset as
+\gls{MSM}. Two datasets (\textcite{zetterberg2004two} and
+\textcite{grabowski2014role}) included both \textit{env} and \textit{gag}
+sequences. Each gene was analyzed separately to assess the robustness of
+\software{netabc} to the particular \gls{HIV} gene sequence used to estimate a
+transmission tree.}}
+
+{\color{red}\sout{For the \textcite{novitsky2014impact} data,}} Each
+\textit{env} sequence was aligned pairwise to the HXB2 reference sequence
+(GenBank accession number K03455), and the hypervariable regions were clipped
+out with \software{BioPython} version 1.66+~\autocite{cock2009biopython}.
+Sequences were multiply aligned using \software{MUSCLE} version 3.8.31
 \autocite{edgar2004muscle}, and alignments were manually inspected with
@@
 
-We applied kernel-assisted ABC to five published HIV datasets (\cref{tab:data}),
-and found substantial heterogeneity among the parameter estimates
-(\cref{fig:abchpd,fig:abchpdm2}). Plots of the marginal posterior distributions
-for each dataset are shown in
+We applied kernel-assisted ABC to {\color{red}\sout{five}}
+{\color{blue}\uline{seven}} published HIV datasets, {\color{blue}\uline{plus
+one in-house, unpublished dataset}} (\cref{tab:data}).
+{\color{red}\sout{and found substantial heterogeneity among the parameter
+estimates}} {\color{blue}\uline{Posterior mean point estimates and 50\% and
+95\% \glspl{HPD} for each parameter are shown in} \cref{fig:abchpd}, \uline{and
+marginal posterior distributions in}
 \cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.
+\Cref{fig:abchpdm2} \uline{shows point estimates and \glspl{HPD} obtained when
+the value $\gls{m} = 1$ was disallowed by the prior. Since the results
+indicated that $\gls{m} = 1$ was the most credible value for several datasets, 
+all results discussed henceforth are for the prior $\gls{m} \sim
+\DiscreteUniform(1, 5)$ unless otherwise stated.}}
+
 Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had

### 2016-06-09 doc/thesis/methods_results.tex
Edits and edits

@@
 
-{\color{blue}\uline{Similar to the results obtained in the previous experiment,
-the \gls{alpha} and \gls{I} parameters were estimated with comparably high
-accuracy, while the other two parameters were less identifiable. Across all
-replicates, $R^2$ values for the correlations between the estimated and true
-values were
+{\color{blue}\uline{Across all replicates, $R^2$ values for the correlations
+between the estimated and true values were
     $0.91$,
@@
 
-\uline{Qualitatively, three of the parameters (excluding \gls{N}) exhibited
-weak identifiability within particular sections of the grid. The \glspl{IQR}
-for \gls{alpha} were nearly identical for the values \gls{alpha} $\leq 1$,
-spanning approximately [0-1.3]. Similarly, the \glspl{IQR} for \gls{alpha} $>
-1$ were all approximately [0.7-2]} (\cref{fig:gridest}). \uline{The
-distributions of kernel scores for these groups of values were also
-qualitatively similar} (\cref{fig:gridalpha}), \uline{although the value
-\gls{alpha} = 1.25 exhibited a more pronounced peak in kernel scores around the
-true value.}
-
-\uline{For \gls{I}, the \glspl{IQR} were nearly identical for the values
-\gls{I} $\leq 2000$ (approximately [500,3800]), and for the values \gls{I} $>
-3000$ (approximately [1600-5000]). Kernel score distributions for all \gls{I}
-values exibited a similar rounded shape} (\cref{fig:gridI}), \uline{although
-this was most pronounced for values near the upper and lower extremes of the
-grid.}
-
-\uline{The \glspl{IQR} for \gls{m} were distinct for \gls{m} = 1 and 2, but
-were similar for all values of \gls{m} $> 2$ (approximately [2-6]). The kernel
-score distribution for \gls{m} = 1 was distinct from the others, having a large
-peak at \gls{m} = 1, while the others exhibited a valley at \gls{m} = 1 and
-were mostly flat everywhere else} (\cref{fig:gridm}).
-
-\uline{The \glspl{IQR} for \gls{N} did not form any obvious groups like the
-other parameters' \glspl{IQR}} (\cref{fig:gridest}). \uline{The kernel score
-distribution for \gls{N} = 1000 was distinct, having a peak at the true value}
-(\cref{fig:gridN}). \uline{The distributions for \gls{N} $> 1000$ were all
-relatively similar, having a valley at the low end of the grid and being mostly 
-flat everywhere else.}}
+\uline{Qualitatively, each of the parameters exhibited weak identifiability
+within particular sections of the grid} (\cref{fig:gridest}). \uline{The 50\%
+\glspl{HDI} for \gls{alpha} were nearly identical for the values $\gls{alpha}
+\leq 0.5$ (on average [X-X]) and for $\gls{alpha} \geq 1.5$ ([X-X]).
+For \gls{I}, similar 50\% \gls{HPD} were observed for $\gls{I} \leq 1500$
+(average [X-X]) and for $\gls{I} > 3000$ ([X-X]). For \gls{m}, both the 50\%
+and 95\% \glspl{HDI} were generally identical for $\gls{m} > 2$ (average [X-X]
+for 50\% and [X-X] for 95\% \gls{HDI}). We did not observe a similar pattern
+for \gls{N}, however the boundaries of the \gls{HDI} did not always
+monotonically increase with the actual value.}
+
+\Cref{fig:gridalpha,fig:gridI,fig:gridm,fig:gridN} \uline{show kernel score
+distributions of kernel scores along the grid for each parameter. The
+distributions for some values, such as $\gls{alpha} = 1.25$, $\gls{I} = 500$
+and $4500$, $\gls{m} = 1$, and $\gls{N} = 1000$, exhibited distinct peaks
+around the true value. This indicates that these values produce distinctively
+shaped trees that can be identified with the tree kernel, when the other
+parameter values are fixed and known. However, for the majority of values of
+each parameter, the score distributions were fairly flat around the true value. 
+This means there is a range of values which produce similarly shaped trees, and
+the parameter is less identifiable within that range. The exception was
+\gls{I}, whose score distributions exhibited a more or less rounded shape with
+the highest point near the true value. }}
 
@@
 % latex table generated in R 3.2.3 by xtable 1.8-2 package
-% Mon Jun  6 15:27:33 2016
+% Thu Jun  9 16:06:29 2016
 \begin{table}[ht]
@@
 \subsection{Results}
-
-We applied kernel-assisted ABC to five published HIV datasets (\cref{tab:data}),
-and found substantial heterogeneity among the parameter estimates
-(\cref{fig:abchpd,fig:abchpdm2}). Plots of the marginal posterior distributions
-for each dataset are shown in
-\cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.
-Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had
-estimated $\alpha$ values near unity for the prior allowing $m = 1$ (posterior
-mean [95\% \gls{HPD}] 
-  0.73 
-  [0.05 - 
-   1.18]
-and
-  0.55 
-  [0.01 -
-   0.99] respectively).
-The estimates did not change appreciably when $m = 1$ was disallowed by the
-prior, although the credible interval of the \textcite{niculescu2015recent}
-data was narrower
-  (0.05 - 
-   1.18).
-When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
-low estimated $\alpha$ values
-  (0.33 
-  [0 - 
-  0.76]
-and
-  0.27 
-  [0 -
-   0.59]). 
-However, the estimates increased when $m = 1$ was not permitted, although the
-HPD intervals remained roughly the same
-  (0.58 
-  [0.06 - 
-  0.99]
-and
-  0.48 
-  [0.02 -
-   0.87]).
-The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
-for both priors on $m$
-  (0.55 for $m \geq 1$;
-   0.53 for $m \geq 2$).
-However, the confidence interval was much wider when $m = 1$ was allowed
-  ([0 -
-    1.75] for $m \geq 1$ vs.
-   [0 -
-    1.75] for $m \geq 2$).
-
-For all the datasets except \citeauthor{novitsky2014impact}, estimated values
-of $I$ were below 2000 when $m = 1$ was allowed, with relatively narrow HPD
-intervals compared to the nonzero prior density region
-  (\citeauthor{cuevas2009hiv}, 701 
-  [289 -
-   1279];
-   \citeauthor{niculescu2015recent}, 747
-  [136 - 
-   2378];
-  \citeauthor{li2015hiv}, 1390 
-  [310 -
-   2821];
-   \citeauthor{wang2015targeting}, 675
-  [175 - 
-   1400]).
-The \citeauthor{novitsky2014impact} data was the outlier, with a very high
-estimated $I$, and HPD interval spanning almost the entire prior region
-  (5431 
-  [183 -
-   8739]).
-The $I$ estimates and HPD intervals were generally robust to the choice of
-prior on $m$, with slightly narrower HPD intervals (compare
-\cref{fig:abchpd,fig:abchpdm2}).
-
-The posterior mean of $m$ was equal to 1 for all but the
-\citeauthor{novitsky2014impact} data, when this value was allowed. However, the
-upper bound of the HPD interval was different for each dataset
-  (\citeauthor{niculescu2015recent}, 4;
-   \citeauthor{wang2015targeting}, 1;
-   \citeauthor{li2015hiv}, 1;
-   \citeauthor{cuevas2009hiv}, 1).
-When $m = 1$ was disallowed, the estimate for all datasets was either 2 or 3,
-with HPD intervals spanning the entire prior region. The estimates for the
-total number of nodes $N$ were largely uninformative for all samples, with
-almost all posterior mean estimates greater than 7500 and HPD intervals
-spanning almost the entire nonzero prior density region. The only exception was
-the \citeauthor{li2015hiv} data, for which the estimate was lower 
-  (5916)
-when $m = 1$ was allowed.
-
-\begin{figure*}[ht]
-  \centering
-  \includegraphics{realdata-hpd-bc}
-  \caption[
-      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
-      parameters of the BA network model, fitted to six HIV datasets
-      with \software{netabc}.]
-  {
-      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
-      parameters of the BA network model, fitted to six HIV datasets with
-      \software{netabc}. Legend labels indicate risk group and country of
-      origin. Abbreviations: IDU, injection drug users; MSM, men who have sex
-      with men; HET, heterosexual.
-  }
-  \label{fig:abchpd}
-\end{figure*}
-
-To make our analyses comparable to the existing network literature, we
-estimated values of the power law exponent $\gamma$ for each of the datasets 
-investigated, using the posterior median estimates with the prior allowing
-\gls{m} = 1. The results are summarized in \cref{tab:gamma}. All of the
-estimated exponents were in the range $2 \leq \gamma \leq 2.5$, which is on the
-lower end of the range $2 \leq \gamma \leq 4$ reported in the literature.
-
-\begin{table}
-    \centering
-    \input{\tablepath/realdata-gamma.tex}
-    \caption[
-        Estimated power law exponents for six HIV datasets based on maximum
-        \textit{a priori} estimates of BA model parameters.
-    ]{
-        Estimated power law exponents for six HIV datasets based on maximum
-        \textit{a priori} estimates of BA model parameters. 100 networks were
-        simulated using posterior mean parameter estimates obtained with
-        \software{netabc}. The power law exponent $\gamma$ was estimated for
-        each, and the median of those estimates was used as a point estimate
-        for the corresponding dataset.
-    }
-    \label{tab:gamma}
-\end{table}

### 2016-06-09 doc/thesis/preamble.tex
Edits and edits

@@
 \newcommand{\E}{\mathbb{E}}
+\newcommand{\R}{\mathbb{R}}
 \renewcommand{\L}{\mathcal{L}}
@@
 \DeclareMathOperator{\Uniform}{Uniform}
+\DeclareMathOperator{\DiscreteUniform}{DiscreteUniform}
 \DeclareMathOperator{\Var}{Var}
@@
 \newglossaryentry{m}{
-  name=$m$,
+  name=\ensuremath{m},
   description=number of edges added per vertex when constructing a Barab\'asi-Albert network
@@
 \newglossaryentry{N}{
-  name=$N$,
+  name=\ensuremath{N},
   description=total number of nodes in a contact network
@@
 \newglossaryentry{I}{
-  name=$I$,
+  name=\ensuremath{I},
   description=number of infected nodes in a contact network at the time of
@@
 \newglossaryentry{p}{
-  name=$p$,
+  name=\ensuremath{p},
   description=edge density parameter in Erd\H{o}s-R\'enyi networks
@@
 \newglossaryentry{lambda}{
-  name=$\lambda$,
+  name=\ensuremath{\lambda},
   description=decay factor meta-parameter for tree kernel
@@
 \newglossaryentry{sigma}{
-  name=$\sigma$,
+  name=\ensuremath{\sigma},
   description=radial basis function variance meta-parameter for tree kernel
@@
 \newglossaryentry{n}{
-  name=$n$,
+  name=\ensuremath{n},
   description=number of particles used for ABC-SMC
@@
 \newglossaryentry{M}{
-  name=$M$,
+  name=\ensuremath{M},
   description=number of simulated datasets per particle in ABC-SMC
 }
-\newglossaryentry{tmrca}{
-  name=$t_{\mathrm{mrca}}$,
-  description=time to most recent common ancestor of a group of extant taxa
-}
 \newglossaryentry{alphaess}{
-  name=$\alpha_{\ESS}$,
+  name=\ensuremath{\alpha_{\ESS}},
   description=per-iteration rate of expected sample size decay in adaptive ABC-SMC

### 2016-06-10 doc/thesis/discussion.Rnw
Some results

@@
 
+\subsubsection{Preferential attachment power}
+
 For all datasets we examined, the posterior mean estimates for \gls{alpha} were
@@
 
+\subsubsection{Power law exponent}
+
 % for discussion
@@
 
-{\color{blue}\uline{
+\subsubsection{HIV prevalence}
+
+{\color{blue}\uline{The true \gls{HIV} prevalence in a population can be
+difficult to estimate for several reasons. \gls{HIV}-infected individuals may
+be asymptomatic for months or years, possibly delaying their awareness of their
+status. In many contexts, the risk factors for acquisition of \gls{HIV} are
+illegal or stigmatized, which may represent a barrier to testing, treatment,
+and/or disclosure of status. Several of the studies whose data we analysed
+report the number of new infections over the study period. We expect these
+numbers to be lower, perhaps substantially, than the true prevalence. We also
+note that our simulation study showed that the estimates of \gls{I} obtained
+with \software{netabc} were upwardly biased, the severity increasing with the
+true value of \gls{I}. In addition, our initial exploratory analysis showed
+that the identifiability of \gls{I} decreases with the number of sampled tips 
+(\cref{fig:Icrossv,fig:Ikpca}). In real world studies, the proportion of
+infected individuals sampled is usually low, whic may imede our ability to
+estimate \gls{I}.
+
 The estimates of the prevalence \gls{I} were largely robust to the choice of
 prior on \gls{m}} (\cref{fig:abchpd,fig:abchpdm2}).
-\textcite{niculescu2015recent} \uline{reported that 494 new HIV infections were
-diagnosed among the studied population (\glspl{IDU} in Bucharest) during the
-study period. Our estimate of the prevalence was higher (X), although the 95\%
-\gls{HPD} interval did contain the value 494. In addition, it is not
-unreasonable that a substantial fraction of the \gls{HIV}-positive individuals
-in an \gls{IDU} population could be undiagnosed, given that these populations
-are often marginalized.}
+
+\textcite{niculescu2015recent} and \textcite{zetterberg2004two} both reported 
+on epidemics in \gls{IDU} populations in Eastern Europe (Romania and Estonia,
+respectively). The point estimates of \gls{I} for these two datasets were
+similar (X and X), although the confidence interval around the
+\textcite{niculescu2015recent} data was wider. \textcite{niculescu2015recent}
+\uline{reported that 494 new \gls{HIV} infections were diagnosed among the
+studied population (\glspl{IDU} in Bucharest) during the study period, and this
+value was contained in our 95\% \gls{HPD} (X-X). On the other hand,
+\textcite{zetterberg2004two} collected samples from several locations
+throughout Estonia; using their reported incidence estimates, there were
+approximately 2000 new infections in the country during the study period. This
+value was outside our 95\% \gls{HPD} (X-X). The authors indicated that their
+sample was ``convenience based''. It is possible that other, harder-to-reach
+individuals had no contacts with those sampled, which would reduce the apparent
+network size. Alternatively, the low estimated prevalence may have been due to
+the low sampling fraction.}
+
+\uline{The cluster we analysed within the BC data consists primarily of
+\gls{IDU} in Vancouver's downtown east side. This population has been
+intensively targeted by public health interventions. Combined with the routine 
+clinical genotyping performed in BC for all new \gls{HIV} infections, it is not
+unreasonable to suppose that the sampled individuals (399) comprise the
+majority of \gls{HIV}-positive network members.
+
+\textcite{wang2015targeting} and \textcite{li2015hiv} studied \gls{MSM}
+populations in major Chinese cities (Shanghai and Beijing).
 
@@
 part be due to model misspecification, which highlights the importance of
-selecting a model appropriate to the dataset at hand.} }
+selecting a model appropriate to the dataset at hand.}}
+
+\subsubsection{Caveats? TODO}
 

### 2016-06-11 doc/thesis/discussion.tex
Writing...

@@
 to 
-    0.73.
+    0.95.
 The sub-linearity is consistent with the results of
@@
 \gls{alpha} for these data was 
-    0.55
+    0.55, 0.55
         or 
-    0.53,
+    0.53, 0.53,
 depending on the prior on \gls{m}; \citeauthor{de2007preferential} estimated
@@
 than the \textcite{wang2015targeting} data\ldots
+
+\subsubsection{Power law exponent}
 

### 2016-06-11 doc/thesis/methods_results.Rnw
Writing...

@@
 restricted our analysis to the subtype B sequences and labelled this dataset as
-\gls{MSM}. Two datasets (\textcite{zetterberg2004two} and
-\textcite{grabowski2014role}) included both \textit{env} and \textit{gag}
-sequences. Each gene was analyzed separately to assess the robustness of
-\software{netabc} to the particular \gls{HIV} gene sequence used to estimate a
-transmission tree.}}
+\gls{MSM}. Three datasets (\textcite{zetterberg2004two},
+\textcite{grabowski2014role}, and \textcite{mccormack2002early}) included both
+\textit{env} and \textit{gag} sequences. Each gene was analyzed separately to
+assess the robustness of \software{netabc} to the particular \gls{HIV} gene
+sequence used to estimate a transmission tree.}}
 
@@
 
-Three of the datasets \autocite[][and the BC data]{li2015hiv,novitsky2014impact}
-were initially much larger than the others, containing 1265, 1299, and 7923
-sequences respectively. To ensure that the analyses were comparable, we reduced
-these to a number of sequences similar to the smaller datasets. For the
-\citeauthor{li2015hiv} and BC datasets, we detected clusters of size 280 and
-399 respectively using a patristic distance cutoff of 0.02 as described
+{\color{red}\sout{Three}} 
+{\color{blue}\uline{Six}} of the datasets 
+\autocite[][and the BC data]{li2015hiv,novitsky2014impact,grabowski2014role,little2014using}
+were initially larger than the others, containing 1265, 1299, 1026/915
+(\textit{env}/\textit{gag}), 648, and 7923 sequences respectively. To ensure
+that the analyses were comparable, we reduced these to a number of sequences
+similar to the smaller datasets. For the \citeauthor{li2015hiv} and BC
+datasets, we detected clusters of size 280 and 399 respectively using a
+patristic distance cutoff of 0.02 as described
 previously~\autocite{poon2015impact}. Only sequences within these clusters were
-carried forward. For the \textcite{novitsky2014impact} data, no large clusters
-were detected using the same cutoff, so we analysed a subtree of size 180
-chosen arbitrarily.
+carried forward. For the
+\textcite{grabowski2014role,novitsky2014impact,little2014using} datasets, no
+large clusters were detected using the same cutoff, so we analysed subsets of
+size 225, 180, and 180 respectively. {\color{blue}\uline{The subset of the
+\citeauthor{grabowski2014role} data was chosen by eye such that the individuals
+were monophyletic in both the \textit{gag} and \textit{env} trees. The other
+subsets were arbitrarily chosen subtrees from phylogenies of the complete
+datasets.}}
 
@@
   m <- collect.metadata("../../simulations/aggregates/hpd/*")
-  d1 <- fread(rownames(m)[m$m_min == 1])
-  d2 <- fread(rownames(m)[m$m_min == 2])
+  d1 <- fread(rownames(m)[m$m_min == 1 & m$gene == "gag"])
+  d2 <- fread(rownames(m)[m$m_min == 2 & m$gene == "gag"])
+
+  d1.env <- fread(rownames(m)[m$m_min == 1 & m$gene == "env"])
+  d2.env <- fread(rownames(m)[m$m_min == 2 & m$gene == "env"])
 
@@
   setkey(low2, dataset)
+
+  pmean1.env <- dcast.data.table(d1.env, dataset~parameter, value.var="mean")
+  up1.env <- dcast.data.table(d1.env, dataset~parameter, value.var="hpd.max.95")
+  low1.env <- dcast.data.table(d1.env, dataset~parameter, value.var="hpd.min.95")
+  setkey(pmean1.env, dataset)
+  setkey(up1.env, dataset)
+  setkey(low1.env, dataset)
+
+  pmean2.env <- dcast.data.table(d2.env, dataset~parameter, value.var="mean")
+  up2.env <- dcast.data.table(d2.env, dataset~parameter, value.var="hpd.max.95")
+  low2.env <- dcast.data.table(d2.env, dataset~parameter, value.var="hpd.min.95")
+  setkey(pmean2.env, dataset)
+  setkey(up2.env, dataset)
+  setkey(low2.env, dataset)
+
+  stopifnot(pmean1[,dataset[which.max(alpha)]] == "zetterberg2004")
+  stopifnot(pmean1[,dataset[which.min(alpha)]] == "cuevas2009")
 @
 
-We applied kernel-assisted ABC to {\color{red}\sout{five}}
-{\color{blue}\uline{seven}} published HIV datasets, {\color{blue}\uline{plus
+We applied \software{netabc} to {\color{red}\sout{five}}
+{\color{blue}\uline{ten}} published HIV datasets, {\color{blue}\uline{plus
 one in-house, unpublished dataset}} (\cref{tab:data}).
@@
 all results discussed henceforth are for the prior $\gls{m} \sim
-\DiscreteUniform(1, 5)$ unless otherwise stated.}}
-
+\DiscreteUniform(1, 5)$ unless otherwise stated. For clarity, datasets will be
+referred to as their risk group and location. For example, \gls{IDU}/Estonia
+refers to the} \textcite{zetterberg2004two} \uline{dataset.}}
+
+{\color{red}\sout{
 Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had
@@
    [\Sexpr{low2["novitsky2014", round(alpha, 2)]} -
-    \Sexpr{up2["novitsky2014", round(alpha, 2)]}] for $m \geq 2$).
-
-For all the datasets except \citeauthor{novitsky2014impact}, estimated values
-of $I$ were below 2000 when $m = 1$ was allowed, with relatively narrow HPD
-intervals compared to the nonzero prior density region
+    \Sexpr{up2["novitsky2014", round(alpha, 2)]}] for $m \geq 2$). }}
+
+{\color{blue}\uline{Posterior mean point estimates for the \acrlong{PA} power
+\gls{alpha} ranged from 
+    \Sexpr{pmean1["zetterberg2004", round(alpha, 2)]}
+for the \gls{IDU}/Estonia data to
+    \Sexpr{pmean1["cuevas2009", round(alpha, 2)]}
+for the mixed/Spain data. When aggregated by risk group, the average estimates
+were
+    \Sexpr{pmean1[c("zetterberg2004", "niculescu2015", "bctree"), round(mean(alpha), 2)]}
+for \gls{IDU},
+    \Sexpr{pmean1[c("novitsky2014", "mccormack2002", "grabowski2014"), round(mean(alpha), 2)]}
+for primarily heterosexual risk, and
+    \Sexpr{pmean1[c("wang2015", "kao2011", "little2014", "li2015"), round(mean(alpha), 2)]}
+for \gls{MSM}. These values were obtained with \textit{gag} for the datasets
+where both \textit{gag} and \textit{env} were sequenced, but the estimates for
+\gls{alpha} did not change appreciably between the two genes} (TODO figure).
+\uline{There was a large amount of uncertainty associated with these
+estimates. 95\% \gls{HPD} were very wide for most datasets, often encompassing
+almost the entire range from 0 to 1} (\cref{fig:abchpd}). \uline{For all the
+datasets except HET/Malawi and MSM/Beijing, the posterior mean was either
+outside, or very close to the border of, the 50\% \gls{HPD}. This indicates
+that the posterior distributions were diffuse with heavy tails, rather than
+having most of their mass around the mode.}}
+
+{\color{red}\sout{For all the datasets except \citeauthor{novitsky2014impact},
+estimated values of $I$ were below 2000 when $m = 1$ was allowed, with
+relatively narrow HPD intervals compared to the nonzero prior density region
   (\citeauthor{cuevas2009hiv}, \Sexpr{pmean1["cuevas2009", floor(I)]} 
@@
 prior on $m$, with slightly narrower HPD intervals (compare
-\cref{fig:abchpd,fig:abchpdm2}).
+\cref{fig:abchpd,fig:abchpdm2}).}}
 

### 2016-06-11 doc/thesis/methods_results.tex
Writing...

@@
 % latex table generated in R 3.2.3 by xtable 1.8-2 package
-% Thu Jun  9 16:06:29 2016
+% Fri Jun 10 18:17:33 2016
 \begin{table}[ht]
@@
 \subsection{Results}
+
+We applied \software{netabc} to {\color{red}\sout{five}}
+{\color{blue}\uline{ten}} published HIV datasets, {\color{blue}\uline{plus
+one in-house, unpublished dataset}} (\cref{tab:data}).
+{\color{red}\sout{and found substantial heterogeneity among the parameter
+estimates}} {\color{blue}\uline{Posterior mean point estimates and 50\% and
+95\% \glspl{HPD} for each parameter are shown in} \cref{fig:abchpd}, \uline{and
+marginal posterior distributions in}
+\cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.
+\Cref{fig:abchpdm2} \uline{shows point estimates and \glspl{HPD} obtained when
+the value $\gls{m} = 1$ was disallowed by the prior. Since the results
+indicated that $\gls{m} = 1$ was the most credible value for several datasets, 
+all results discussed henceforth are for the prior $\gls{m} \sim
+\DiscreteUniform(1, 5)$ unless otherwise stated. For clarity, datasets will be
+referred to as their risk group and location. For example, \gls{IDU}/Estonia
+refers to the} \textcite{zetterberg2004two} \uline{dataset.}}
+
+{\color{red}\sout{
+Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had
+estimated $\alpha$ values near unity for the prior allowing $m = 1$ (posterior
+mean [95\% \gls{HPD}] 
+  0.73 
+  [0.05 - 
+   1.18]
+and
+  0.55 
+  [0.01 -
+   0.99] respectively).
+The estimates did not change appreciably when $m = 1$ was disallowed by the
+prior, although the credible interval of the \textcite{niculescu2015recent}
+data was narrower
+  (0.05 - 
+   1.18).
+When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
+low estimated $\alpha$ values
+  (0.33 
+  [0 - 
+  0.76]
+and
+  0.27 
+  [0 -
+   0.59]). 
+However, the estimates increased when $m = 1$ was not permitted, although the
+HPD intervals remained roughly the same
+  (0.58 
+  [0.06 - 
+  0.99]
+and
+  0.48 
+  [0.02 -
+   0.87]).
+The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
+for both priors on $m$
+  (0.55 for $m \geq 1$;
+   0.53 for $m \geq 2$).
+However, the confidence interval was much wider when $m = 1$ was allowed
+  ([0 -
+    1.75] for $m \geq 1$ vs.
+   [0 -
+    1.75] for $m \geq 2$). }}
+
+{\color{blue}\uline{Posterior mean point estimates for the \acrlong{PA} power
+\gls{alpha} ranged from 
+    0.9
+for the \gls{IDU}/Estonia data to
+    0.27
+for the mixed/Spain data. When aggregated by risk group, the average estimates
+were
+    0.73
+for \gls{IDU},
+    0.43
+for primarily heterosexual risk, and
+    0.39
+for \gls{MSM}. These values were obtained with \textit{gag} for the datasets
+where both \textit{gag} and \textit{env} were sequenced, but the estimates for
+\gls{alpha} did not change appreciably between the two genes} (TODO figure).
+\uline{There was a large amount of uncertainty associated with these
+estimates. 95\% \gls{HPD} were very wide for most datasets, often encompassing
+almost the entire range from 0 to 1} (\cref{fig:abchpd}). \uline{For all the
+datasets except HET/Malawi and MSM/Beijing, the posterior mean was either
+outside, or very close to the border of, the 50\% \gls{HPD}. This indicates
+that the posterior distributions were diffuse with heavy tails, rather than
+having most of their mass around the mode.}}
+
+{\color{red}\sout{For all the datasets except \citeauthor{novitsky2014impact},
+estimated values of $I$ were below 2000 when $m = 1$ was allowed, with
+relatively narrow HPD intervals compared to the nonzero prior density region
+  (\citeauthor{cuevas2009hiv}, 701 
+  [289 -
+   1279];
+   \citeauthor{niculescu2015recent}, 747
+  [136 - 
+   2378];
+  \citeauthor{li2015hiv}, 1390 
+  [310 -
+   2821];
+   \citeauthor{wang2015targeting}, 675
+  [175 - 
+   1400]).
+The \citeauthor{novitsky2014impact} data was the outlier, with a very high
+estimated $I$, and HPD interval spanning almost the entire prior region
+  (5431 
+  [183 -
+   8739]).
+The $I$ estimates and HPD intervals were generally robust to the choice of
+prior on $m$, with slightly narrower HPD intervals (compare
+\cref{fig:abchpd,fig:abchpdm2}).}}
+
+The posterior mean of $m$ was equal to 1 for all but the
+\citeauthor{novitsky2014impact} data, when this value was allowed. However, the
+upper bound of the HPD interval was different for each dataset
+  (\citeauthor{niculescu2015recent}, 4;
+   \citeauthor{wang2015targeting}, 1;
+   \citeauthor{li2015hiv}, 1;
+   \citeauthor{cuevas2009hiv}, 1).
+When $m = 1$ was disallowed, the estimate for all datasets was either 2 or 3,
+with HPD intervals spanning the entire prior region. The estimates for the
+total number of nodes $N$ were largely uninformative for all samples, with
+almost all posterior mean estimates greater than 7500 and HPD intervals
+spanning almost the entire nonzero prior density region. The only exception was
+the \citeauthor{li2015hiv} data, for which the estimate was lower 
+  (5916)
+when $m = 1$ was allowed.
+
+\begin{figure*}[ht]
+  \centering
+  \includegraphics{realdata-hpd-bc}
+  \caption[
+      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
+      parameters of the BA network model, fitted to six HIV datasets
+      with \software{netabc}.]
+  {
+      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
+      parameters of the BA network model, fitted to six HIV datasets with
+      \software{netabc}. Legend labels indicate risk group and country of
+      origin. Abbreviations: IDU, injection drug users; MSM, men who have sex
+      with men; HET, heterosexual.
+  }
+  \label{fig:abchpd}
+\end{figure*}
+
+To make our analyses comparable to the existing network literature, we
+estimated values of the power law exponent $\gamma$ for each of the datasets 
+investigated, using the posterior median estimates with the prior allowing
+\gls{m} = 1. The results are summarized in \cref{tab:gamma}. All of the
+estimated exponents were in the range $2 \leq \gamma \leq 2.5$, which is on the
+lower end of the range $2 \leq \gamma \leq 4$ reported in the literature.
+
+\begin{table}
+    \centering
+    \input{\tablepath/realdata-gamma.tex}
+    \caption[
+        Estimated power law exponents for six HIV datasets based on maximum
+        \textit{a priori} estimates of BA model parameters.
+    ]{
+        Estimated power law exponents for six HIV datasets based on maximum
+        \textit{a priori} estimates of BA model parameters. 100 networks were
+        simulated using posterior mean parameter estimates obtained with
+        \software{netabc}. The power law exponent $\gamma$ was estimated for
+        each, and the median of those estimates was used as a point estimate
+        for the corresponding dataset.
+    }
+    \label{tab:gamma}
+\end{table}

### 2016-06-13 doc/thesis/discussion.tex
Intro summary; finished ABC

--- 2016-06-07 doc/thesis/discussion.tex
@@
   \centering
-  \includegraphics{derivatives.pdf}
+  %\includegraphics{derivatives.pdf}
   \caption[
@@
 substantially.}}
-
-For all datasets we examined, the posterior mean estimates for \gls{alpha} were
-sub-linear, ranging from 
-    0.27
-to 
-    0.73.
-The sub-linearity is consistent with the results of
-\textcite{de2007preferential}, who developed a statistical inference method to
-estimate the parameters of a more sophisticated \acrlong{PA} model
-incorporating heterogeneous node behaviour. They found \gls{alpha} values
-ranging from 0.26 to 0.62, depending on the gender and time period considered. 
-Our estimates of \gls{alpha} for the \textcite{niculescu2015recent} was above
-this range under both priors, as were the estimates for the
-\textcite{wang2015targeting} data and the BC data when \gls{m} = 1 was
-disallowed by the prior. The dataset investigated by
-\textcite{de2007preferential} was derived from a survey of a random sample of
-the Norwegian population, whereas our investigation focused on datasets from
-known phylogenetic or geographic clusters of \gls{HIV} infected persons. It is
-therefore unsurprising that we detected stronger \acrlong{PA} dynamics in some
-cases. For instance, random sampling is much less likely to discover the
-high-degree nodes characterizing the tail of the degree distribution, simply
-because those individuals are rare in the general population. In addition, it
-is plausible that \gls{HIV}-positive individuals are more likely to be highly
-connected in their sexual networks, as the odds of acquiring \gls{HIV} increase
-with the number of unprotected sexual contacts.
-
-Both \textcite{de2007preferential} and \textcite{novitsky2014impact} studied
-populations whose primary risk factor for \gls{HIV} infection was heterosexual
-contact. \citeauthor{de2007preferential} explicitly excluded reported
-homosexual contacts; \citeauthor{novitsky2014impact} did not, but noted that
-heterosexual contact is the primary mode of transmission in Botswana where the
-study was done. In the first of the two papers describing the Botswana
-study~\autocite{novitsky2013phylogenetic}, the authors noted that their sample
-was gender-biased, being composed of approximately 75\% women. Our estimate of
-\gls{alpha} for these data was X or X, depending on the prior on \gls{m};
-\citeauthor{de2007preferential} estimated 0.54, 0.57, and 0.29 for 3-year,
-5-year, and lifetime partnership networks respectively for the female portion
-of their sample.
-
-For both choices of prior on \gls{m}, the datasets derived from \gls{IDU}
-populations had a higher estimated \acrlong{PA} power than the other datasets
-(\cref{fig:abchpd,fig:abchpdm2}). This finding is in line with
-\textcite{dombrowski2013topological}, who reanalyzed a network of \glspl{IDU}
-in Brooklyn, USA, collected between 1991 and
-1993~\autocite{friedman2006social}. They found that the the \gls{IDU} network 
-resembled a \gls{BA} network much more closely than other social and sexual 
-networks, and offered sociological explanations for the apparent \acrlong{PA}
-dynamics in this population. Importantly, from a public health perspective,
-the authors asserted that the removal of \emph{random} individuals from
-\gls{IDU} networks may have the paradoxical effect of increasing the network's
-epidemic susceptibility. When low-degree nodes are removed, as would occur
-during a police crackdown, their network neighbours may turn to well-known
-community members for advice or supplies, thus increasing the connectivity of
-these high-degree nodes.
-
-%For both priors, the estimated prevalence was extremely high, in fact higher
-%than the estimated HIV prevalence in the sampled region. The authors indicated
-%that the source of the samples was a town in close proximity to the country's
-%capital city, and suggested that there may have been a high degree of migration
-%and partner interchange between the two locations. It is possible that the
-%contact network underlying the subtree we investigated includes a much larger
-%group based in the capital city, which would explain the high estimate of $I$.
-%There is no clear explanation for the discrepancy between the two priors for
-%the \textcite{li2015hiv} data, as the subset we analyzed formed a phylogenetic
-%cluster and therefore was a good candidate for the BA model. However, nearly
-%all the posterior density was assigned to $m = 1$ when this value was allowed,
-%indicating that the network was more likely to have an acyclic tree structure.
-
-{\color{blue}\uline{
-The estimates of the prevalence \gls{I} were largely robust to the choice of
-prior on \gls{m}} (\cref{fig:abchpd,fig:abchpdm2}).
-\textcite{niculescu2015recent} \uline{reported that 494 new HIV infections were
-diagnosed among the studied population (\glspl{IDU} in Bucharest) during the
-study period. Our estimate of the prevalence was higher (X), although the 95\%
-\gls{HPD} interval did contain the value 494. In addition, it is not
-unreasonable that a substantial fraction of the \gls{HIV}-positive individuals
-in an \gls{IDU} population could be undiagnosed, given that these populations
-are often marginalized.}
-
-\uline{The individuals sampled by} \textcite{wang2015targeting} \uline{were
-recruited by following a prospective cohort of 2000 \gls{MSM} in Beijing,
-China. Of the 2000, 179 new infections were identified during the study period.
-This number is much lower than the estimated prevalence of X obtained with
-\software{netabc}, however the authors did not claim to have sampled the entire
-sexual network. In a nationwide survey,} \textcite{wu2013hiv} \uline{estimated
-that there were 24,198 \gls{MSM} living in Beijing, of whom 5.7\%, or 1379,
-were \gls{HIV}-infected, which is within the 95\% confidence interval of our
-estimate of \gls{I}.}
-
-\textcite{li2015hiv} \uline{studied an \gls{MSM} population in Shanghai, China.
-Despite the apparent similarity to the context of the
-}\textcite{wang2015targeting} \uline{study, the estimated \gls{HIV} prevalence
-of the } \citeauthor{li2015hiv} \uline{data was higher (X). The aforementioned
-survey~\autocite{wu2013hiv} estimated that there were 14,511 \gls{MSM} living
-in Shanghai, with an \gls{HIV} prevalence of 6.8\% or approximately 987
-individuals. This was within the 95\% confidence interval of the estimate
-obtained with \software{netabc}. It is worth noting that} \citeauthor{li2015hiv}
-\uline{claimed that there were 80,000 \gls{MSM} living in Shanghai, which is a
-significantly different estimate than that obtained by} \textcite{wu2013hiv}.
-
-\textcite{cuevas2009hiv} \uline{reported that 620 newly infected and 1500
-chronically infected patients had been sampled during the study period, for a
-total of 2120. This is substantially greater than the point estimate of X we
-obtained with \gls{ABC}, even without considering the likely possibility that
-there were more infected than sampled individuals. Again, this error may in
-part be due to model misspecification, which highlights the importance of
-selecting a model appropriate to the dataset at hand.} }
-
-Our use of the \gls{BA} model makes several simplifying assumptions. First, we
-assume homogeneity across the network with respect to node behaviour and
-transmission risk. In reality, the attraction to high-degree nodes seems likely
-to vary among individuals, as does their risk of transmitting or contracting
-the virus. We have also assumed that all transmission risks are symmetric,
-which is clearly false for all known modes of \gls{HIV} transmission, and that
-infected individuals never recover but remain infectious indefinitely. These
-assumptions were made for the purpose of keeping the model as simple as
-possible, since this is the very first attempt to fit a contact network model
-in a phylodynamic context. However, the Gillespie simulation algorithm built
-into \software{netabc} can handle arbitrary transmission and removal rates
-which need not be homogeneous across the network. Moreover, it is possible to
-use \gls{ABC} to fit a model which relaxes some or all of these assumptions,
-which may be a fruitful avenue for future investigation.  Despite the possible
-misspecification, our estimates of the power law exponent $\gamma$ were within
-the range of values reported in the literature (\cref{tab:gamma}).

### 2016-06-13 doc/thesis/intro.tex
Intro summary; finished ABC

@@
 
-As discussed in \cref{subsec:smcsamp}, the
-choices of the kernels $K$ and $L$ is problem-specific, and so appropriate
-kernels must be chosen for \gls{ABC}. Several options have been
-proposed~\autocite{beaumont2009adaptive, sisson2007sequential,
-del2012adaptive}.
-
-\begin{algorithm}
-  \caption{\gls{ABC}-\gls{SMC}.}
-  \begin{algorithmic}
-    \For {$k = 1$ to $n$}
-      \State Sample $x_1^{(k)}$ from $\Pr(\theta) \Pr(z \mid \theta)$
-      \Comment{Draw $\theta$ from the prior and simulate $z$ from $\theta$}
-      \State $w^{(k)} \gets 1/n$
-    \EndFor
-    \For {$i = 2$ to $i_{\max}$}
-      \Comment{$i_{\max}$ is a user-specified number of iterations}
-      \For {$k = 1$ to $n$}
-        \State Sample $x_i^{(k)}$ from $K(x_{i-1}^{(k)}, x_i)$
-        \Comment Extend the $k$th particle
-        \State $w^{(k)} \gets w^{(k)} \cdot \dfrac{\gamma_i(x_i) L_{i-1}(x_i, x_{i-1})}{\gamma_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}$
-      \EndFor
-      \State Normalize the weights so that $\sum w = 1$
-      \If {$\ESS(w) < T$}
-        \Comment{$T$ is a user-defined threshold}
-        \State Resample the particles according to $w$
-        \For {$k = 1$ to $n$}
-          \State $w^{(k)} \gets 1/n$
-        \EndFor
-      \EndIf
-    \EndFor
-  \end{algorithmic}
-  \label{alg:abcsmc}
-\end{algorithm}
+As discussed in \cref{subsec:smcsamp}, the choices of the kernels $K$ and $L$
+is problem-specific, and so appropriate kernels must be chosen for \gls{ABC}.
+Several options have been proposed~\autocite{beaumont2009adaptive,
+sisson2007sequential, del2012adaptive}. {\color{blue}\uline{With an
+appropriately chosen kernel, the weight update} (\cref{eq:smcwt}) \uline{will
+simplify into a computable expression.} \textcite{sisson2007sequential} and
+\textcite{beaumont2009adaptive} \uline{both suggest random walk kernels for
+$K$, where each particle is
+perturbed according to a Gaussian distribution. The backwards kernels $L$ are
+chosen to approximately minimize the variance in importance weights. In this
+thesis, we use an \gls{MCMC} kernel, as proposed by} \textcite{del2012adaptive}.
+\uline{The associated backwards kernels and weight updates are discussed in the
+next chapter} (\cref{subsec:adaptsmc}). \uline{With generic kernels $K$ and
+$L$, the \gls{ABC}-\gls{SMC} algorithm is almost identical to the \gls{SMC}
+sampler} (\cref{alg:smcsamp}), \uline{with $\gamma_i$ replaced with $\pi_i$.}
+Therefore we will not repeat the full algorithm here.}
 
@@
 
-{\color{blue}\uline{ Our method integrates the four distinct research areas
-  just described: phylogenetics, contact networks, \acrlong{SMC}, and
-  \acrlong{ABC}. The first two topics together form the problem domain.
-  Phylogenetic data is the input to our method, while estimates of the
-  parameters of contact network models are the desired output. The latter two
-  topics define the algorithm and statistical framework that our inference
-method will use. }}
+\glsreset{ABC}
+\glsreset{SMC}
+\glsreset{BA}
+
+{\color{blue}\uline{In} \cref{sec:obj}, \uline{we outlined three research
+objectives that will be addressed in this thesis. First, we aim to develop a
+method for fitting contact network models to estimated transmission trees.
+Although transmission trees can be estimated for any epidemic by thorough
+contact tracing, viral diseases are the most useful context for this method.
+Transmission and sequence evolution occur on similar time scales for RNA
+viruses, resulting in viral phylogenies whose shapes are heavily constrained by
+the transmission process. The study of this interaction between evolution and
+epidemiology is called \emph{phylodynamics}. Phylodynamic methods make it
+possible to estimate transmission trees from viral sequence data; these
+estimated trees form the input data for our method.}
+
+\uline{From these estimated transmission trees, our method's purpose is to
+estimate the parameters of a contact network model. Rather than assuming a
+homogeneously mixed population, as most epidemiological models do, network
+models take the more realistic view that human populations are structured. That
+is, contacts which allow for transmission may occur in a nonrandom way, rather
+than between every pair of individuals in the population. A network model
+parameterizes this structure. In particular, our second research objective is
+to characterize our ability to fit the \gls{BA} model, which incorporates
+preferential attachment to generate networks with realistic degree
+distributions.}
+
+\uline{Our method will apply \gls{ABC}, a simulation-based approach, for
+fitting these models. Simulating a transmission tree according to a network
+model is straightforward: a network can be generated according to the model,
+and the spread of an epidemic can be simulated over the network and recorded in
+a transmission tree. \gls{ABC} uses the concordance between these simulated
+transmission trees and the ``true'' estimated tree as an indicator of parameter
+credibility. The closer the simulated transmission trees appear to the true
+tree, the more weight is assigned to the associated network parameters.}
+
+\uline{\gls{ABC} can be implemented by at least three algorithms, but the one
+we choose to apply in this work is \gls{SMC}. \Gls{SMC} uses a population of
+parameter ``particles'' to approximate a distribution of interest, in this case
+the approximate posterior distribution targeted by \gls{ABC}. After running the
+\gls{ABC}-\gls{SMC} algorithm, statistics on the model parameters can be
+approximated by the weighted population of particles. For example, a weighted
+average would give an approximate expected value for each parameter.}
+
+\uline{Thus, our method integrates four research topics: phylogenetics, contact
+networks, \acrlong{SMC}, and \acrlong{ABC}. The first two topics together form
+the problem domain. Phylogenetic data is the input to our method, while
+estimates of the parameters of contact network models are the desired output.
+The latter two topics define the algorithm and statistical framework that our
+inference method will use. }}

### 2016-06-14 doc/thesis/abstract.Rnw
Editing and proofreading intro

%TC:macro \del [ignore]

Models of the spread of disease in a population often make the simplifying
assumption that the population is homogeneously mixed, or is divided into
homogeneously mixed compartments. However, human populations have complex
structures formed by social contacts, which can have a significant influence on
the rate \add{and pattern} of epidemic spread. Contact \del{network models}
\add{networks} capture this structure by explicitly representing each contact
that could possibly lead to a transmission. \add{Contact network models
parameterize the structure of these networks, but estimating their parameters
from contact data requires extensive, often prohibitive, epidemiological
investigation.}

We developed a method based on approximate Bayesian computation (ABC) for
estimating structural parameters of the contact network underlying an
observed viral phylogeny. The method combines adaptive sequential Monte Carlo
for ABC, Gillespie simulation for propagating epidemics though networks, and a
\add{previously developed} kernel-based tree similarity score. \add{Our method
offers the potential to quantitatively investigate contact network structure
from phylogenies derived from viral sequence data, complementing traditional
epidemiological methods.}

We applied our method to \del{fit} the Barab\'{a}si-Albert network model.
\add{This model incorporates the preferential attachment mechanism observed in
real world social and sexual networks, whereby individuals with more
connections attract new contacts at an elevated rate (``the rich get
richer'').} \del{to simulated transmission trees and applied it to viral
phylogenies estimated from six real-world HIV sequence datasets.} Using
simulated data, we found that the strength of preferential attachment and the
number of infected nodes could often be accurately estimated. However, the mean
degree of the network and the total number of nodes appeared to be weakly- or
non-identifiable with ABC.

<<alpha, include=FALSE>>=
    source("global.R") 
    d <- setDT(collect.data("../../simulations/aggregates/hpd/*"))
    d <- d[m_min == 1 & agg_gene == "gag",]

    stopifnot(d[parameter == "alpha", dataset[which.min(mean)]] == "cuevas2009")
    stopifnot(d[parameter == "alpha", dataset[which.max(mean)]] == "zetterberg2004")

    stopifnot(d[parameter == "alpha", max(mean)] >= 0.9)
    stopifnot(d[parameter == "alpha", max(mean)] < 1.0)
@

Finally, the Barab\'{a}si-Albert model was fit to \del{six}\add{eleven} real
world HIV datasets, and substantial heterogeneity in the parameter estimates
was observed. \del{Point estimates} \add{Posterior means} for the preferential
attachment power \add{were all sub-linear} and ranged from \del{0.06 to 1.05}
\add{ 
    \Sexpr{d[parameter == "alpha", round(min(mean), 2)]}
to
    \Sexpr{d[parameter == "alpha", round(max(mean), 2)]}.
Point estimates of the strength of preferential attachment were higher in
injection drug user populations, potentially indicating that high-degree
``hub'' nodes may play a role in epidemics among this risk group.} Our results
underscore the importance of considering contact structures when
\del{performing phylodynamic inference}\add{investigating viral outbreaks.}

### 2016-06-14 doc/thesis/intro.tex
Editing and proofreading intro

@@
 correspond to \defn{extant}, or observed, taxa. The \defn{internal nodes}
-correspond to their common ancestors, {\color{blue}\uline{usually extinct
-(although occasionally the internal nodes may be observed as well, eg.
-\autocite{jombart2011reconstructing}).}} The edges or \defn{branches} of the
+correspond to their common ancestors, \add{usually extinct (although
+occasionally the internal nodes may be observed as well, eg.
+\autocite{jombart2011reconstructing}).} The edges or \defn{branches} of the
 phylogeny connect ancestors to their descendants. Phylogenies may have a
@@
 $c_n^{j}$ be the $j$th child of node $n$, and $l_n$ be the vector of branch
-lengths connecting node $n$ to its $\nc(n)$ children. {\color{blue}\uline{
-Furthermore, let $\nl(n)$ be the number of children of $n$ which are leaves (we
-always have $\nl(n) \leq \nc(n)$). The \defn{production rule} of $n$ is the
-pair $(\nc(n), \nl(n))$.}} That is, if two nodes have the same number of
-children and among these, the same number of leaves, then they have the same
-production rule. Let $k_G(x, y)$ be a Gaussian radial basis function of the
-vectors $x$ and $y$,
+lengths connecting node $n$ to its $\nc(n)$ children. \add{Furthermore, let
+$\nl(n)$ be the number of children of $n$ which are leaves (we always have
+$\nl(n) \leq \nc(n)$). The \defn{production rule} of $n$ is the pair $(\nc(n),
+\nl(n))$.} That is, if two nodes have the same number of children and among
+these, the same number of leaves, then they have the same production rule. Let
+$k_G(x, y)$ be a Gaussian radial basis function of the vectors $x$ and $y$,
 \[
@@
 \]
-{\color{blue}\uline{
-The parameter $\lambda$ in the above expression, is called the \emph{decay
+\add{The parameter $\lambda$ in the above expression, is called the \emph{decay
 factor}~\autocite{moschitti2006making}, and takes a value between 0 and 1.
-Without this parameter, terms in the sum} \ref{eq:kernel} \uline{corresponding
-to large subset trees with the same topology would be similarly large and tend
-to dominate the kernel score. $\lambda$ penalizes $\Delta$ more strongly as the
+Without this parameter, terms in the sum \ref{eq:kernel} corresponding to large
+subset trees with the same topology would be similarly large and tend to
+dominate the kernel score. $\lambda$ penalizes $\Delta$ more strongly as the
 number of recursive calls increases, which downweights the largest matching
 substructures and allows smaller matches to contribute more to the kernel
-score. }} In this work, we refer to the parameters $\lambda$ and $\sigma$ as
+score.} In this work, we refer to the parameters $\lambda$ and $\sigma$ as
 \defn{meta-parameters}, to avoid confusing them with model parameters we are
-trying to estimate. {\color{blue}\uline{ When evaluating the tree kernel, it is
-helpful to reorder the children of each internal node such that the larger of
-the two subtrees is on the right-hand side. If the two subtrees have equal
-sizes, then the child with the longer branch length can be put on the
-right-hand side. This operation is referred to as \defn{ladderizing}. Since the
-ordering of children is arbitrary in phylogenies, this operation ensures that a
-maximal number of matching subset trees are counted by the tree kernel without 
-making meaningful changes to the trees. }}
-
-{\color{blue}\uline{
-The tree kernel was later shown to be highly effective in differentiating trees
-simulated under a compartmental model with two risk groups of varying contact
-rates~\autocite{poon2015phylodynamic}. In that paper,
+trying to estimate. \add{When evaluating the tree kernel, it is helpful to
+reorder the children of each internal node such that the larger of the two
+subtrees is on the right-hand side. If the two subtrees have equal sizes, then
+the child with the longer branch length can be put on the right-hand side. This
+operation is referred to as \defn{ladderizing}. Since the ordering of children
+is arbitrary in phylogenies, this operation ensures that a maximal number of
+matching subset trees are counted by the tree kernel without making meaningful
+changes to the trees.}
+
+\add{The tree kernel was later shown to be highly effective in differentiating
+trees simulated under a compartmental model with two risk groups of varying
+contact rates~\autocite{poon2015phylodynamic}. In that paper,
 \citeauthor{poon2015phylodynamic} used the tree kernel as the distance function
-in \gls{ABC}} (see \cref{sec:abc}), \uline{to fit epidemiological models to
-observed trees. }}
+in \gls{ABC} (see \cref{sec:abc}), to fit epidemiological models to observed
+trees.}
 
@@
 epidemic threshold~\autocite{pastor2001epidemic}, meaning that diseases with
-arbitrarily low transmissibility {\color{red}\sout{can persist}} 
-{\color{blue}\uline{have a chance, however small, of persisting}} at low levels
-indefinitely. This is in contrast with homogeneously mixed populations, in
-which transmissibility below the epidemic threshold would result in exponential
-decay in the number of infected individuals and eventual extinction of the
+arbitrarily low transmissibility \del{can persist} \add{have a chance, however
+small, of persisting} at low levels indefinitely. This is in contrast with
+homogeneously mixed populations, in which transmissibility below the epidemic
+threshold would result in exponential decay in the number of infected
+individuals and eventual extinction of the
 pathogen~\autocite{anderson1992infectious}.
@@
 rather than a power law, was the best fit to degree distributions of six social
-and sexual networks. \textcite{drombowski2013} contend that sexual networks are
-shaped more by homophily (``like attract like'') than by preferential
-attachment, but find that \gls{IDU} network do demonstrate a scale-free
-structure.
+and sexual networks. \add{\textcite{dombrowski2013topological} contend that
+sexual networks are shaped more by homophily (``like attract like'') than by
+preferential attachment, but find that \gls{IDU} network do demonstrate a
+scale-free structure.}
 
@@
 is proportional to $d^\alpha + 1$ for some constant \gls{alpha}.
-{\color{blue}\uline{When $\gls{alpha} \neq 1$, the degree distribution no
-longer follows a power law~\autocite{krapivsky2000connectivity}. For
-$\gls{alpha} < 1$, the distribution is a stretched exponential, meaning that
-the number of nodes of degree $k$ is proportional to $\exp(-k^\beta)$ for some
-constant $\beta$. For $\gls{alpha} > 1$, the distribution takes on a
-characteristic called \emph{gelation}, where a one or a few high-degree hub
-nodes are connected to nearly every other node in the graph. We do not believe
-these departures from the power law affect the applicability of the model to
-real world networks. In fact,} \textcite{de2007preferential} \uline{were able to
-estimate the \acrlong{PA} power from partner count data collected from the same
-individuals for consecutive time intervals, and found a value less than one in
-all cases. It is also worth noting that, in addition to the \gls{BA} model,
-other investigations of the interaction between contact networks and
-transmission trees have studied the \acrlong{ER} and \acrlong{WS}
-models~\autocite{leventhal2012inferring}, whose degree distributions do not
-generally follow a power law under any parameter settings.}}
+\add{When $\gls{alpha} \neq 1$, the degree distribution no longer follows
+a power law~\autocite{krapivsky2000connectivity}. For $\gls{alpha} < 1$, the
+distribution is a stretched exponential, meaning that the number of nodes of
+degree $k$ is proportional to $\exp(-k^\beta)$ for some constant $\beta$. For
+$\gls{alpha} > 1$, the distribution takes on a characteristic called
+\emph{gelation}, where a one or a few high-degree hub nodes are connected to
+nearly every other node in the graph. We do not believe these departures from
+the power law affect the applicability of the model to real world networks. In
+fact, \textcite{de2007preferential} were able to estimate the \acrlong{PA}
+power from partner count data collected from the same individuals for
+consecutive time intervals, and found a value less than one in all cases. It is
+also worth noting that, in addition to the \gls{BA} model, other investigations
+of the interaction between contact networks and transmission trees have studied
+the \acrlong{ER} and \acrlong{WS} models~\autocite{leventhal2012inferring},
+whose degree distributions do not generally follow a power law under any
+parameter settings.}
 
@@
 
-{\color{blue}\uline{
-Recall that the primary objective of our work is to develop a statistical
+\add{Recall that the primary objective of our work is to develop a statistical
 inference method for estimating contact network parameters from transmission
-trees. In particular, for a network model with parameters $\theta$ and an input
-transmission tree $T$, we are interested in the posterior distribution}
+trees. For a network model with parameters $\theta$ and an input transmission
+tree $T$, our goal is to estimate statistics, such as means and credible
+intervals, of the posterior distribution}
 \begin{align}
-  \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)},
+  \madd{\post(\theta \mid T) = \frac{\lik(T \mid \theta) \prior(\theta)}
+    {\int_\Theta\lik(T \mid \theta) \prior(\theta)}.}
   \label{eq:post}
 \end{align}
-\uline{where we have adopted the common abuse of the symbol $\Pr$ to refer to a
-probability density. As alluded to in} \cref{sec:obj}, \uline{both the
-likelihood $\Pr(T \mid \theta)$ and the normalizing constant $\Pr(T)$ are
-likely computationally intractable (this will be explained further in}
-\cref{subsec:just}\uline{). Hence, rather than computing the posterior
-distribution analytically, we will approximate it using a \defn{Monte Carlo}
-approach. The fundamental idea behind Monte Carlo methods is succinctly
-expressed by \textcite{liu2001theoretical}}:
+\add{As alluded to in \cref{sec:obj}, both the likelihood $\lik(T \mid \theta)$
+and the normalizing constant are likely computationally intractable
+(this will be discussed further in \cref{subsec:just}). Hence, rather than
+computing the posterior distribution analytically, we will approximate it using
+a \defn{Monte Carlo} approach. The fundamental idea behind Monte Carlo methods
+is succinctly expressed by \textcite{liu2001theoretical}:}
 \begin{quote}
-    \uline{Monte Carlo's view of the world is that any probability distribution $\pi$, 
-    regardless of its complexity, can always be \emph{represented} by a
-    discrete sample from it. By ``represented'', we mean that any computation
-    of expectations using $\pi$ can be replaced to an acceptable degree of
-    accuracy by using the empirical distribution resulting from the discrete
-    sample.}
+    \add{Monte Carlo's view of the world is that any probability distribution
+    $\pi$, regardless of its complexity, can always be \emph{represented} by a
+    discrete sample from it. By ``represented'', we mean that any
+    computation of expectations using $\pi$ can be replaced to an acceptable
+    degree of accuracy by using the empirical distribution resulting from the
+    discrete sample.}
 \end{quote}
-\uline{In other words, if we are able to sample enough points from a
-distribution of interest, we will be able to make reasonably accurate
-statements about the distribution itself. For example, the expected value of
-the distribution can be estimated by the sample's population mean. The reason
-Monte Carlo methods will be useful in this work is that algorithms exist for
-obtaining samples from distributions that are analytically intractable and from
-which direct sampling is not possible } (for a review see
-\autocite{robert2004monte}). \uline{\Gls{SMC}}~\autocite{doucet2000sequential,
-doucet2001introduction, liu2008monte} \uline{is one such algorithm.}
-
-\uline{\Gls{SMC} is also known as the \emph{particle filter}. Rather than
-sampling points one at a time from the target distribution, \gls{SMC} considers
-a population of points or ``particles'', here denoted $\set{x^{(k)}}$ and
-indexed by an integer $k$. The particles are associated with weights,
-$\set{w(x^{(k)})}$. These weighted particles are a representation for the
-target distribution. For example, the expected value of the distribution is
-approximated by}
-\[
-  \frac{1}{n} \sum_{k=1}^n x^{(k)} w(x^{(k)}),
-\]
-\uline{where $n$ is the number of particles. Initially, the particles do not
+\add{In other words, if we are able to sample enough points from a distribution
+of interest, we will be able to make reasonably accurate statements about the
+distribution itself. For example, the expected value of the distribution can be
+estimated by the sample's population mean. The reason Monte Carlo methods will
+be useful in this work is that algorithms exist for obtaining samples from
+distributions that are analytically intractable and from which direct sampling
+is not possible (for a review see \autocite{robert2004monte}).
+\Gls{SMC}~\autocite{doucet2000sequential, doucet2001introduction, liu2008monte}
+is one such algorithm.}
+
+\add{\Gls{SMC} considers a population of points or ``particles'', here denoted
+$\set{x^{(k)}}$ and indexed by an integer $k$. The particles are associated
+with weights, $\set{w(x^{(k)})}$. After the algorithm has been run, the
+weighted particles are a Monte Carlo representation for the target
+distribution. For example, the expected value is approximated by}
+\[
+  \madd{\frac{1}{n} \sum_{k=1}^n x^{(k)} w(x^{(k)}),}
+\]
+\add{where $n$ is the number of particles. Initially, the particles do not
 represent the target distribution but rather a more tractable distribution from
@@
 applied to the particles in such a way that they converge, collectively, to a
-sample from the target.}
-
-\uline{In this work, the distribution of interest is the posterior distribution}
-\ref{eq:post}. \uline{The particles are particular values of the parameters
-$\theta$ of the contact network model being studied. If we were taking a
-typical Bayesian Monte Carlo approach to this problem, the particles would end
-up weighted by their posterior probability and distributed in such a way that
-the weighted population was a reasonable representation of $\Pr(\theta \mid
-T)$. In our case, due to the intractable likelihood, we will need to consider
-an approximation to the posterior } (see \cref{sec:abc}). \uline{However, for
-now, nothing is lost by assuming that our target distribution is the posterior
-itself.}
-
-\uline{The goal of this section of the introduction is to describe an algorithm
-called the \gls{SMC} sampler~\autocite{del2006sequential}, which forms the
-basis of the adaptive \gls{ABC}-\gls{SMC} algorithm we apply toward the main
-objective of this work. We begin by describing \gls{SIS}, which is a precursor
-to \gls{SMC} that samples from a sequence of distributions defined on spaces of
-increasing dimension. We then describe \gls{SMC} itself, which extends
-\gls{SIS} with a resampling step to fight particle degeneracy. Finally, we
-outline the \gls{SMC} sampler, which allows \gls{SMC} to be applied to
-sequences of distributions all defined on the same space. This terminology will
-become clear as the methods are described. }}
-
-{\color{red}\sout{Sequential Monte Carlo (SMC) is the name for a family of
-statistical inference methods that rely on approximating probability
-distributions of interest with large collections of \defn{particles}, here
-denoted $\set{x^{(k)}}$}\autocite{doucet2000sequential, doucet2001introduction,
-liu2008monte}. \sout{These collections or \defn{populations} are constructed to form
+representation of the target. \Gls{SMC} is also known as the \defn{particle
+filter}.}
+
+\add{In this work, the distribution of interest is the posterior distribution
+\ref{eq:post}. The particles are particular values of the parameters $\theta$
+of the contact network model being studied. If we were taking a typical
+Bayesian Monte Carlo approach to this problem, the particles would end up
+weighted by their posterior probability and distributed in such a way that the
+weighted population was a reasonable representation of $\post(\theta \mid T)$.
+In our case, due to the intractable likelihood, we will need to consider an
+approximation to the posterior (see \cref{sec:abc}). However, for now, nothing
+is lost by assuming that our target distribution is the posterior itself.}
+
+\add{Here we describe an algorithm called the \gls{SMC} sampler, developed by
+\textcite{del2006sequential}, which forms the basis of the adaptive
+\gls{ABC}-\gls{SMC} algorithm we apply toward the main objective of this work.
+We begin by describing \gls{SIS}, which is a precursor to \gls{SMC} that
+samples from a sequence of distributions defined on spaces of increasing
+dimension. We then describe \gls{SMC} itself, which extends \gls{SIS} with a
+resampling step to fight particle degeneracy. Finally, we outline the \gls{SMC}
+sampler, which allows \gls{SMC} to be applied to sequences of distributions all
+defined on the same space. This terminology will become clear as the methods
+are described.}
+
+\del{\Gls{SMC} is the name for a family of statistical inference methods that
+rely on approximating probability distributions of interest with large
+collections of \defn{particles}, here denoted
+$\set{x^{(k)}}$~\autocite{doucet2000sequential, doucet2001introduction,
+liu2008monte}. These collections or \defn{populations} are constructed to form
 a \defn{Monte Carlo approximation} to some distribution of interest $\pi$,
@@
 distribution to $\pi$ as the population size gets
-large}\autocite{liu2001theoretical}. \sout{The word \defn{sequential} is used because
+large~\autocite{liu2001theoretical}. The word \defn{sequential} is used because
 the particle populations are modified in an iterative fashion over time, for
-example, to incorporate new evidence.}}
+example, to incorporate new evidence.}
 
@@
 the $k$th particle in a population. To ease the notational burden we will omit
-the superscripts and subscripts on the weight functions $w$.
-
-We define a \defn{Markov kernel} as the continuous analogue of the transition
-matrix in a finite-state Markov model.  For some spaces $X$ and $Y$, $K \maps X
-\times Y \to [0, 1]$ such that
+the superscripts and subscripts on the weight functions $w$. We define a
+\defn{Markov kernel} as the continuous analogue of the transition matrix in a
+finite-state Markov model. For some spaces $X$ and $Y$, $K \maps X \times Y
+\to [0, 1]$ such that
 \begin{align}
@@
 to do with the kernel functions defined in \cref{subsec:treeshape}, other than
-sharing a name (the word ``kernel'' is ubiquitous in mathematics).
+sharing a name (the word ``kernel'' is ubiquitous in mathematics). \add{Also, 
+the variable $\pi$ is used here for a generic target distribution to describe
+the algorithm, and does not refer to the prior or posterior distributions we
+will eventually want to work with.}
 
@@
 
-\Gls{SIS}~\autocite{gordon1993novel} {\color{blue}\uline{is a particle-based
-method}} whose aim is to sample from a distribution $\pi$ on an
-high-dimensional space, say $\pi(\vec{x}) = \pi(x_1, \ldots, x_d)$. The basis
-of \gls{SIS} is \gls{IS}, which is a method of estimating summary statistics of
-distributions which are known only up to a normalizing constant, and therefore
-cannot be sampled from directly. That is, if $\pi$ is such a distribution and
-$f$ is any real-valued function, \gls{IS} is concerned with estimating
+\Gls{SIS}~\autocite{gordon1993novel} \add{is a particle-based method} whose aim
+is to sample from a distribution $\pi$ on an high-dimensional space, say
+$\pi(\vec{x}) = \pi(x_1, \ldots, x_d)$. The basis of \gls{SIS} is \gls{IS},
+which is a method of estimating summary statistics of distributions which are
+known only up to a normalizing constant, and therefore cannot be sampled from
+directly. That is, if $\pi$ is such a distribution and $f$ is any real-valued
+function, \gls{IS} is concerned with estimating
 \[
@@
 the expectation of interest as
-{\color{blue}
 \begin{align}
-  \int f(x) \pi(x) \d x = \frac{1}{Z} \int w(x) f(x) \d x.
+  \int f(x) \pi(x) \d x = 
+  \mdel{\dfrac{\int f(x) \gamma(x) \d x}{\int w(x) \eta(x) \d x}}\quad
+  \madd{\frac{1}{Z} \int w(x) \eta(x) f(x) \d x}.
   \label{eq:is}
 \end{align}
-}
-{\color{blue}\uline{
-Since $\eta$ can be sampled from exactly, and $\gamma$ and $f$ can both be
-evaluated pointwise, the integral $\int w(x) f(x) \d x$ can be approximated by
-a Monte Carlo estimate. Moreover, the normalizing constant $Z$ can be expressed
-in terms of the importance weight and distribution, $Z = \int w(x) \eta(x) \d
-x$. Therefore, we have all the ingredients we need to obtain an estimate of
-$\pi(f)$ using} \cref{eq:is}.} Although this is a simple and elegant approach,
+\add{Since $\eta$ can be sampled from exactly, and $\gamma$ and $f$ can both be
+evaluated pointwise, the integral $\int w(x) \eta(x) f(x) \d x$ can be
+approximated by a Monte Carlo estimate. We simply take a sample from $\eta$,
+multiply each point in the sample by $f$ and $\gamma$ evaluated at that point,
+and sum the results. Moreover, the normalizing constant $Z$ can be expressed in
+terms of the importance weight and distribution, $Z = \int w(x) \eta(x) \d x$.
+Therefore, we have all the ingredients we need to obtain an estimate of
+$\pi(f)$ using \cref{eq:is}.} Although this is a simple and elegant approach,
 the drawback is that the variance of the estimate is proportional to the
@@
 The importance weights for $\eta$ can be written recursively as
-\begin{align}
+\begin{equation}
   \label{eq:sisw}
-  w(\vec{x_i}) = \frac{\pi(\vec{x_i})}{\eta(\vec{x_i})}
-  = \frac{\pi(x_i \mid \vec{x_{i-1}})\pi(\vec{x_{i-1}})}
+  \begin{alignedat}{3}
+  w(\vec{x_i}) 
+    &= \frac{\pi(\vec{x_i})}{\eta(\vec{x_i})}
+     & \text{definition of importance weight} \\
+    &= \frac{\pi(x_i \mid \vec{x_{i-1}})\pi(\vec{x_{i-1}})}
          {\eta(x_i \mid \vec{x_{i-1}})\eta(\vec{x_{i-1}})}
-  = \frac{\pi(x_i \mid \vec{x_{i-1}})}
-         {\eta(x_i \mid \vec{x_{i-1}})}\cdot w(\vec{x_{i-1}}).
-\end{align}
+      & \qquad\qquad\text{definition of conditional probability} \\
+    &= \frac{\pi(x_i \mid \vec{x_{i-1}})}
+            {\eta(x_i \mid \vec{x_{i-1}})}\cdot w(\vec{x_{i-1}})
+      & \text{definition of importance weight}.
+  \end{alignedat}
+\end{equation}
 Thus, we can choose $\eta(x_i \mid \vec{x_{i-1}})$ such that the variance of
@@
 The recursive definition \ref{eq:sisw} suggests an algorithm for obtaining a
-sample from {\color{blue}\uline{$\eta$ and using it to obtain an approximate
-sample from $\pi$ by \gls{IS}}} (\cref{alg:sis}).  We begin with $n$
-``particles'', which have been sampled from the importance distribution
-$\eta(x_0)$ for $\pi(x_0)$. The particles are updated and reweighted $d$ times,
-corresponding to the $d$ elements of the decomposition of $\pi$. At the $i$th
-step, each particle is extended to include $x_i$ drawn according to the chosen
-$\eta(x_i \mid \vec{x_{i-1}})$, and the importance weights are recalculated and
-normalized.
+sample from \add{$\eta$ and using it to obtain an approximate sample from $\pi$
+by \gls{IS}} (\cref{alg:sis}). We begin with $n$ particles, which have been
+sampled from the importance distribution $\eta(x_1)$ for $\pi(x_1)$. \del{The
+particles are updated and reweighted $d$ times, corresponding to the $d$
+elements of the decomposition of $\pi$. At the $i$th step, each particle is
+extended to include $x_i$ drawn according to the chosen $\eta(x_i \mid
+\vec{x_{i-1}})$, and the importance weights are recalculated and normalized.}
+\add{Each particle is extended by drawing $x_2$ from $\eta(x_2 \mid x_1)$, and
+the importance weights are updated by \cref{eq:sisw}. This procedure is
+repeated $d$ times until the particles have dimension equal to that of $\pi$.}
 
@@
       \State Sample $x_1^{(k)}$ from $\eta(x_1)$
-      \Comment{Initialize the $k$th particle}
-      %\State $w^{(k)} \gets \dfrac{\pi\left(x_1^{(k)}\right)}{\eta\left(x_1^{(k)}\right)}$
+      \Comment Initialize the $k$th particle
       \State $w^{(k)} \gets \pi(x_1^{(k)}) / \eta(x_1^{(k)})$
+      \Comment Initialize importance weight
     \EndFor
@@
         \Comment Extend the $k$th particle
-        \State $w(\vec{x_i}^{(k)}) \gets [\pi(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}) \,/\, \eta(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}})] \cdot w(\vec{x_{i-1}}^{(k)})$
+        \State $w^{(k)} \gets [\pi(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}}) \,/\, \eta(x_i^{(k)} \mid \vec{x_{i-1}^{(k)}})] \cdot w^{(k)}$
+        \Comment Update importance weight
       \EndFor
@@
 approximation to $\pi$, since only a few particles retain high importance
-weights after all $d$ sequential steps, 
-{\color{blue}\uline{ a problem known as \defn{particle degeneracy}. To mitigate
-this problem, \textcite{gordon1993novel} introduced technique they called the
-\defn{bootstrap filter}, which involves a resampling of the population of
-particles after each sequential step in accordance with their importance
-weights. A similar idea, termed \defn{particle rejuvination}, was proposed by
-\textcite{liu1995blind}. These approaches cause particles with high importance
-weights to be replicated in the population, while particles with low weights
-may be removed. After each resampling step, the importance weights for all
-particles are set equal.}}
+weights after all $d$ sequential steps, \add{a problem known as \defn{particle
+degeneracy}. To mitigate this problem, \textcite{gordon1993novel} introduced
+technique they called the \defn{bootstrap filter}, which involves a resampling
+of the population of particles after each sequential step in accordance with
+their importance weights. A similar idea, termed \defn{particle rejuvination},
+was proposed by \textcite{liu1995blind}. These approaches cause particles with
+high importance weights to be replicated in the population, while particles
+with low weights may be removed. After each resampling step, the importance
+weights for all particles are set equal.}
+
+\add{The resampling step was formally integrated with \gls{SIS} by
+\textcite{doucet2000sequential} to form the first \gls{SMC} algorithm
+(\cref{alg:smc}). Rather than resample at every step as the bootstrap filter
+proposed, the authors use a criterion based on the \gls{ESS} the particle
+population to determine when resampling is necessary.} The \gls{ESS} of the
+population of particles is defined as 
+\[ 
+  \ESS(w) = \frac{n}{1 + \Var(w)}, 
+\]
+where \gls{n} is the number of particles in the population. Resampling is
+triggered when the \gls{ESS} drops below the threshold \add{(conventionally
+$n/2$~\autocite{liu2008monte})}. \del{This results in the removal of low-weight
+particles from the population, and also equalizes all the weights.} Various
+resampling strategies beyond basic sampling with replacement have been
+proposed~\autocite{douc2005comparison}, but we will not discuss those here.
 
@@
       \State $w^{(k)} \gets \pi(x_1^{(k)}) / \eta(x_1^{(k)})$
+      \Comment{Initialize importance weight}
     \EndFor
@@
 
-{\color{blue}\uline{The resampling step was formally integrated with
-\gls{SIS} by \textcite{doucet2000sequential} to form the first \gls{SMC}
-algorithm }(\cref{alg:smc}). \uline{Rather than resample at every step as the
-bootstrap filter proposed, the authors use a criterion based on the \gls{ESS}
-the particle population to determine when resampling is necessary. }} The
-\gls{ESS} of the population of particles is defined as 
-\[ 
-  \ESS(w) = \frac{n}{1 + \Var(w)}, 
-\]
-where $n$ is the number of particles in the population.
-When the \gls{ESS} drops below the threshold
-{\color{blue}\uline{(conventionally $n/2$~\autocite{liu2008monte})}}, particles
-are resampled according to their importance weights. This results in the
-removal of low-weight particles from the population, and also equalizes all the
-weights. Various resampling strategies beyond the basic sampling with
-replacement have been proposed~\autocite{douc2005comparison}, but we will not
-discuss those here.
-
 \subsection{The sequential Monte Carlo sampler}
@@
 The \gls{SIS} and \gls{SMC} algorithms described above aim to sample from a
-high-dimensional distribution $\pi(x)$, by sequentially sampling from $d$
+high-dimensional distribution $\pi(\vec{x})$, by sequentially sampling from $d$
 distributions of lower but increasing dimension. \textcite{del2006sequential}
-developed an \defn{\gls{SMC} sampler} with an alternative objective: to sample
-sequentially from $d$ distributions $\pi_1, \ldots, \pi_d$, all of the same
-dimension and defined on the same space. The $\pi_i$ are assumed to form a
-related sequence, such as posterior distributions attained by sequentially
-considering new evidence. As with \gls{SIS}, we assume that $\pi_i(x) =
-\gamma_i(x) / Z_i$, where $\gamma_i$ is known pointwise and the normalizing
-constant $Z_i$ is unknown.
+developed the \defn{\gls{SMC} sampler} with an alternative objective: to sample
+sequentially from $d$ distributions $\pi_1, \ldots, \pi_d$, all of the
+\emph{same} dimension and defined on the same space. The $\pi_i$ are assumed to
+form a related sequence, such as posterior distributions attained by
+sequentially considering new evidence. As with \gls{SIS}, we assume that
+$\pi_i(x) = \gamma_i(x) / Z_i$, where each $\gamma_i$ is known pointwise and
+the normalizing constants $Z_i$ are unknown.
 
 Both algorithms involve progression through a sequence of related
-distributions. For \gls{SIS}, these distributions are lower-dimensional
-marginals of the target distribution, while for the \gls{SMC} sampler, they are
-of the same dimension and constitute a smooth progression from an initial to a
-final distribution. In both cases, the neighbouring distributions in the
-sequence are related to each other in some way, and we can take advantage of
-that relationship to create a sequence of importance distributions alongside
-the sequence of targets. In \gls{SIS}, the neighbouring marginals
-$\pi(\vec{x_i})$ and $\pi(\vec{x_{i+1}})$ were related by the conditional
-density $\pi(x_i \mid \vec{x_{i-1}})$, which we used to inform the importance
-distribution. In \gls{SMC}, the relationship between subsequent distributions
-is less explicit, but it is assumed that they are related closely enough that
-an importance distribution for $\pi_i$ can be easily transformed into one for
-$\pi_{i+1}$. In particular, the sequence of importance distributions $\eta_i$
-is constructed as
+distributions. For \gls{SIS} and \gls{SMC}, these distributions are
+lower-dimensional marginals of the target distribution, while for the \gls{SMC}
+sampler, they are of the same dimension and constitute a smooth progression
+from an initial to a final distribution. In both cases, the neighbouring
+distributions in the sequence are related to each other in some way, and we can
+take advantage of that relationship to create a sequence of importance
+distributions alongside the sequence of targets. In \gls{SIS}, the neighbouring
+marginals $\pi(\vec{x_i})$ and $\pi(\vec{x_{i+1}})$ were related by the
+conditional density $\pi(x_i \mid \vec{x_{i-1}})$, which we used to inform the
+importance distribution. In \gls{SMC}, the relationship between subsequent
+distributions is less explicit, but it is assumed that they are related closely
+enough that an importance distribution for $\pi_i$ can be easily transformed
+into one for $\pi_{i+1}$. In particular, the sequence of importance
+distributions $\eta_i$ is constructed as
 \begin{align}
@@
 \]
-of increasing dimension. This brings us back to the setting described above in
-\cref{subsec:sis}, namely of building up an importance distribution of
-dimension $d$ sequentially through lower-dimensional distributions. We can
-write $\tilde{\pi}_i$ in terms of $\tilde{\pi}_{i-1}$ by noticing that
+of increasing dimension. \add{That is, $\tilde{\pi}_i$ has dimension equal to
+the original dimension of $\pi_i$ raised to the power of $i$.} This brings us
+back to the \gls{SIS} setting described above (\cref{subsec:sis}), namely of
+building up an importance distribution sequentially through lower-dimensional
+distributions. \add{The dimension of the importance distributions $\eta$ is
+similarly augmented with the forward kernels,}
+\begin{align}
+  \eta_i(\vec{x_i}) = \eta_1(x_1) \prod_{k=1}^{i-1} K_k(x_k, x_{k+1})
+  \label{eq:smcimp}
+\end{align}
+\add{Thanks to the backwards kernels, we can write $\tilde{\pi}_i$ in terms of
+$\tilde{\pi}_{i-1}$ as follows.}
 \begin{align*}
-  \frac{\tilde{\pi}_i(\vec{x_i})}{\tilde{\pi}_{i-1}(\vec{x_{i-1}})} 
-  = \frac{\pi_i(x_i) \prod_{k=1}^{i-1} L(x_{k+1}, x_k)}
+  \mdel{\dfrac{\tilde{\pi}_i(\vec{x_i})}{\tilde{\pi}_{i-1}(\vec{x_{i-1}})} 
+  = \dfrac{\pi_i(x_i) \prod_{k=1}^{i-1} L(x_{k+1}, x_k)}
   {\pi_{i-1}(x_{i-1}) \prod_{k=1}^{i-2} L(x_{k+1}, x_k)}
-  = \frac{\pi_i(x_i) L(x_i, x_{i-1})}{\pi_{i-1}(x_{i-1})},
+  = \dfrac{\pi_i(x_i) L(x_i, x_{i-1})}{\pi_{i-1}(x_{i-1})},}
 \end{align*}
-and hence
-\[
-  \tilde{\pi}_i = \frac{\pi_i(x_i) L(x_i, x_{i-1})}{\pi_{i-1}(x_{i-1})} \cdot \tilde{\pi}_{i-1}.
+\del{and hence}
+\[
+  \mdel{\tilde{\pi}_i = \dfrac{\pi_i(x_i) L(x_i, x_{i-1})}{\pi_{i-1}(x_{i-1})} \cdot \tilde{\pi}_{i-1}.}
+\]
+\begin{align*}
+  \madd{\tilde{\pi}_i(\vx_i)} &\madd{= \pi_i(x_i) \prod_{k=1}^{i-1} L_k(x_{k+1}, x_k)}
+    & \madd{\text{definition of } \tilde{\pi_i}} \\
+  &\madd{= \pi_i(x_i) L_{i-1}(x_{i-1}, x_i) \prod_{k=1}^{i-2} L_k(x_{k+1}, x_k)}
+  & \madd{\text{pull out } k = i-1 \text{ term from product}} \\
+  &\madd{= \frac{\pi_i(x_i)L_{i-1}(x_{i-1}, x_i)}{\pi_{i-1}(x_{i-1})} \cdot \pi_{i-1}(x_{i-1}) \prod_{k=1}^{i-2} L_k(x_{k+1}, x_k)}
+  & \madd{\text{multiply and divide by } \pi_{i-1}(x_{i-1})} \\
+  &\madd{= \frac{\pi_i(x_i)L_{i-1}(x_{i-1}, x_i)}{\pi_{i-1}(x_{i-1})} \cdot \tilde{\pi}_{i-1}(\vec{x_{i-1}})}
+  & \madd{\text{definition of } \tilde{\pi}_{i-1}}.
+\end{align*}
+\add{The importance distributions can also be expressed recursively using the
+forward kernels, which follows directly from \cref{eq:smcimp},}
+\[
+  \madd{\eta_i(\vec{x_i}) = \eta_{i-1}(\vec{x_{i-1}})K_i(x_{i-1}, x_i).}
 \]
@@
   w(\vec{x_i}) 
-    &= \frac{\tilde{\pi}_i(\vec{x_i})}{\eta_i(\vec{x_i})} \\
-    &= \frac{\tilde{\pi}_{i-1}(\vec{x_{i-1}}) \pi_i(x_i) L(x_i, x_{i-1})}
-           {\eta_{i-1}(\vec{x_{i-1}}) \pi_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)} \\
+    &= \frac{\tilde{\pi}_i(\vec{x_i})}{\eta_i(\vec{x_i})} 
+      & \madd{\text{definition of importance weight}} \nonumber \\
+    &= \madd{\frac{\tilde{\pi}_i(\vec{x_i})}{\eta_{i-1}(\vec{x_{i-1}})K_i(x_{i-1}, x_i)}}
+      & \madd{\text{recursive definition of } \eta_i} \nonumber \\
+    &= \frac{\tilde{\pi}_{i-1}(\vec{x_{i-1}}) \pi_i(x_i) L_{i-1}(x_i, x_{i-1})}
+            {\eta_{i-1}(\vec{x_{i-1}}) \pi_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}
+      & \madd{\text{recursive definition of } \pi_i} \nonumber \\
     &= w(\vec{x_{i-1}}) \cdot
       \frac{\pi_i(x_i) L_{i-1}(x_i, x_{i-1})}
-           {\pi_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)} \\
+           {\pi_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)} 
+      & \madd{\text{definition of importance weight }} \nonumber \\
     &\propto w(\vec{x_{i-1}}) \cdot
       \frac{\gamma_i(x_i) L_{i-1}(x_i, x_{i-1})}
-           {\gamma_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}.
+           {\gamma_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}
+      & \madd{\text{remove normalizing constant $Z_i/Z_{i-1}$}} 
     \label{eq:smcwt}
@@
 sample from $\pi_i$, by considering only the $i$th component of $\vec{x_i}$.
-{\color{blue}\uline{In fact, since the weight update} \cref{eq:smcwt} 
-\uline{depends only on the $i$th and $i-1$st components of each particle, we do
-not even need to keep track of the complete particles if we are only interested
-in the final distribution.}} These are all the ingredients we need to apply
-\gls{SIS}. The sequences of kernels $L$ and $K$ should be chosen based on the
-problem at hand to minimize the variance in the importance weights as well
-as possible. For a fixed choice of $K_i$, the backward kernels $L_i$ which
-minimize this variance are called the \defn{optimal} backward kernels. The full
-\gls{SMC} sampler algorithm is presented as \cref{alg:smcsamp}. A resampling
-step is applied whenever the \gls{ESS} of the population drops too low, as
-discussed in the previous section.
+\add{In fact, since the weight update \cref{eq:smcwt} depends only on the $i$th
+and $i-1$st components of each particle, we do not even need to keep track of
+the complete particles if we are only interested in the final distribution.}
+\del{These are all the ingredients we need to apply \gls{SIS}.} The sequences
+of kernels $L$ and $K$ should be chosen based on the problem at hand to
+minimize the variance in the importance weights as well as possible. For a
+fixed choice of $K$, the backward kernels which minimize this variance are
+called the \defn{optimal} backward kernels. The full \gls{SMC} sampler
+algorithm is presented as \cref{alg:smcsamp}. A resampling step is applied
+whenever the \gls{ESS} of the population drops too low, as discussed in the
+previous section.
 
@@
       \State $w^{(k)} \gets \gamma_1(x_1^{(k)}) / \eta_1(x_1^{(k)})$
+      \Comment{Initialize the importance weights}
       \State Normalize the weights so that $\sum w = 1$
@@
         \State $w^{(k)} \gets w^{(k)} \cdot \dfrac{\gamma_i(x_i) L_{i-1}(x_i, x_{i-1})}{\gamma_{i-1}(x_{i-1}) K_i(x_{i-1}, x_i)}$
+        \Comment{Update the importance weights}
       \EndFor
@@
 
-{\color{blue}\uline{\Acrlong{SMC}, and the \gls{SMC} sampler, were developed
-for sampling from distributions which can be evaluated up to a normalizing
-constant. We claim, and shall argue more thoroughly below}
-(\cref{subsec:just}), \uline{that the posterior distribution}
-\[
-  \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)}
-\]
-\uline{for a contact network model with parameters $\theta$ and an input
-transmission tree $T$ does not fall in this category (note that, as in the
-previous section, we are using $\Pr$ to denote a probability density).
-Therefore, \gls{SMC}, and other Bayesian and \gls{ML} techniques for fitting
-mathematical models} (see \cref{chp:prelim}), \uline{cannot be directly
-applied to our problem. In particular, \gls{MCMC} and the \gls{SMC} sampler are
-designed for distributions $\pi$ which can be evaluated up to a normalizing
-constant $Z$, that is, $\pi(x) = \gamma(x) / Z$. Both algorithms calculate the
-ratio $\pi(x) / \pi(x') = \gamma(x) / \gamma(x')$ for a current value $x$ and
-proposed updated value $x'$ - for \gls{MCMC}, this is part of the \acrlong{MH}
-ratio, while for the \gls{SMC} sampler, it is required to calculate the
-importance weights. In the context of Bayesian inference, this ratio is a
-likelihood ratio, which must be calculated by computing the individual
-likelihoods and dividing them. If the likeilhood is intractible, this is
-clearly not a viable approach.}
-
-\uline{\Gls{ABC}}~\autocite{rubin1984bayesianly, tavare1997inferring,
-fu1997estimating}\uline{ was developed to estimate posterior distributions with
+\add{\Acrlong{SMC}, and the \gls{SMC} sampler, were developed for sampling from
+distributions which can be evaluated up to a normalizing constant. We claim,
+and shall argue more thoroughly below (\cref{subsec:just}) that the posterior
+distribution}
+\[
+  \madd{\post(\theta \mid T) \propto \lik(T \mid \theta) \prior(\theta)}
+\]
+\add{for a contact network model with parameters $\theta$ and an input
+transmission tree $T$ does not fall in this category. Therefore, \gls{SMC}, and
+other Bayesian and \gls{ML} techniques for fitting mathematical models (see
+\cref{chp:prelim}), cannot be directly applied to our problem. In particular,
+\gls{MCMC} and the \gls{SMC} sampler are designed for distributions $\pi$ which
+can be evaluated up to a normalizing constant $Z$, that is, $\pi(x) = \gamma(x)
+/ Z$. Both algorithms calculate a ratio of the form $\pi(x) / \pi(x') \propto
+\gamma(x) / \gamma(x')$ for a current value $x$ and proposed updated value $x'$
+- for \gls{MCMC}, this is part of the \acrlong{MH} ratio, while for the
+\gls{SMC} sampler, a similar ratio is required to calculate the importance
+weights. In the context of Bayesian inference, this ratio contains a likelihood
+ratio, which must be calculated by computing the individual likelihoods and
+dividing them. If the likeilhood is intractible, this is clearly not a viable
+approach.}
+
+\add{\Gls{ABC}~\autocite{rubin1984bayesianly, tavare1997inferring,
+fu1997estimating} was developed to estimate posterior distributions with
 intractable likelihoods, which have arisen frequently in the domain of
-population genetics} \autocite{beaumont2002approximate,
-beaumont2010approximate}. \uline{\Gls{ABC} navigates around the intractable
-likelihood by replacing the posterior as the target of inference by an
-\emph{approximate} posterior. This distribution is constructed in such a way
-that the ratios required for \gls{MCMC} and the \gls{SMC} sampler can be
-computed. This conveniently allows us to apply the existing \gls{MCMC} and
-\gls{SMC} algorithms with minimal changes. In the next section, we shall
-demonstrate how this is done, but first we give the definition of the
-approximate posterior. }}
-
-{\color{red}\sout{Most mathematical models are amenable to fitting via one or
-both of the approaches, \gls{ML} or Bayesian inference, discussed above.
-However, there are some, particularly in the domain of population
-genetics}~\autocite{beaumont2002approximate, beaumont2010approximate},
-\sout{for which calculation of either the likelihood or the product of the
-likelihood and the prior may be infeasible. For example, one or both of these
-quantities may be expressible only as an intractable integral. \Gls{ABC} is
-designed for such cases, where standard likelihood-based techniques for model
-fitting cannot be applied.}}
-
-{\color{red}\sout{Ordinarily, Bayesian inference targets the posterior
-distribution $\Pr(\theta \mid y)$. That is, in the Bayesian framework,}} 
-{\color{blue}\uline{By targeting the posterior distribution, Bayesian inference
-makes the assertion that}} model parameters with higher posterior density are
-``better'' in the sense that they offer a more credible explanation for the
-observed data. The approximate posterior targeted by \gls{ABC} uses an
-alternative metric for parameter credibility, namely the similarity of
-simulated datasets to the observed data. If datasets simulated under the model
-closely resemble the real data, it follows that the model is a reasonable
-approximation to the real-world process generating the observed data. More
-formally, {\color{blue}\uline{let $y$ be the observed data to which we are
-trying to fit a model with parameters $\theta$. In the case of this work, the
-data is a transmission tree $T$, but we shall stick with the generic variable
-$y$ for now.}} Suppose we have a distance measure $\rho$ defined on the space
-of all possible data our model could generate. \gls{ABC} aims to sample from
-the joint posterior distribution of model parameters and simulated datasets $z$
-which are within some small distance $\varepsilon$ of the observed data $y$,
+population genetics~\autocite{beaumont2002approximate,
+beaumont2010approximate}. \Gls{ABC} navigates around the intractable likelihood
+by replacing the posterior as the target of inference by an \emph{approximate}
+posterior. This distribution is constructed in such a way that the ratios
+required for \gls{MCMC} and the \gls{SMC} sampler can be computed, conveniently
+allowing us to apply those algorithms with minimal changes. In the next
+section, we shall demonstrate how this is done, but first we give the
+definition of the approximate posterior.}
+
+\del{Most mathematical models are amenable to fitting via one or both of the
+approaches, \gls{ML} or Bayesian inference, discussed above. However, there are
+some, particularly in the domain of population
+genetics~\autocite{beaumont2002approximate, beaumont2010approximate}, for which
+calculation of either the likelihood or the product of the likelihood and the
+prior may be infeasible. For example, one or both of these quantities may be
+expressible only as an intractable integral. \Gls{ABC} is designed for such
+cases, where standard likelihood-based techniques for model fitting cannot be
+applied.}
+
+\del{Ordinarily, Bayesian inference targets the posterior distribution
+$\post(\theta \mid y)$. That is, in the Bayesian framework,}
+\add{By targeting the posterior distribution, Bayesian inference makes the
+assertion that} model parameters with higher posterior density are ``better''
+in the sense that they offer a more credible explanation for the observed data.
+The approximate posterior targeted by \gls{ABC} uses an alternative metric for
+parameter credibility: the similarity of simulated datasets to the observed
+data. If datasets simulated under the model closely resemble the real data, it
+follows that the model is a reasonable approximation to the real-world process
+generating the observed data. More formally, \add{let $y$ be the observed data
+to which we are trying to fit a model with parameters $\theta$. In the case of
+this work, the data is a transmission tree $T$, but we shall stick with the
+generic variable $y$ for now.} Suppose we have a distance measure $\rho$
+defined on the space of all possible data our model could generate. \gls{ABC}
+aims to sample from the joint posterior distribution of model parameters and
+simulated datasets $z$ which are within some small distance $\varepsilon$ of
+the observed data $y$,
 \begin{align}
   \pi_{\varepsilon}(\theta, z \mid y) =
-  \frac{\Pr(\theta) \Pr(z \mid \theta) \I_{A_{\varepsilon, y}} (z)}
-  {\int_{A_{\varepsilon, y} \times \Theta} \Pr(\theta) \Pr(z \mid \theta) \d \theta}.
+  \frac{\prior(\theta) \lik(z \mid \theta) \I_{A_{\varepsilon, y}} (z)}
+  {\int_{A_{\varepsilon, y} \times \gls{Theta}} \prior(\theta) \lik(z \mid \theta) \d \theta}.
   \label{eq:abctarget}
 \end{align}
-Here, $A_{\varepsilon, y}$ is an $\varepsilon$-ball around $y$ with
-respect to $\rho$, $\Theta$ is the space of all possible model parameters, and
-$\I$ is the indicator function~\autocite{marin2012approximate}. The
-distribution $\pi_\varepsilon(\theta, z \mid y)$ will be referred to as
-the \textit{\gls{ABC} target distribution}. {\color{blue}\uline{The term $\Pr(z
-\mid \theta)$ appears to be the bothersome likelihood again, but this will turn
-out not to be a problem because we are simulating $z$ ourselves. In fact, it is
-possible to sample from this distribution exactly (see next section).}}
-
-{\color{blue}\uline{To return to the context of this thesis, the observed data
-$y$ is an estimated transmission tree for a viral epidemic under investigation.
-The model in question is a contact network model with parameters $\theta$. $z$
-is a simulated transmission tree, obtained by first generating a contact
-network under the model, and then simulating the spread of an epidemic over
-that network. A transmission tree can be constructed by keeping track of who
+Here, $A_{\varepsilon, y}$ is an $\varepsilon$-ball around $y$ with respect to
+$\rho$, \gls{Theta} is the space of all possible model parameters, and $\I$ is
+the indicator function~\autocite{marin2012approximate}. The distribution
+$\pi_\varepsilon(\theta, z \mid y)$ will be referred to as the
+\textit{\gls{ABC} target distribution}. \add{The term $\lik(z \mid \theta)$
+appears to be the bothersome likelihood again, but this will turn out not to
+be a problem because we are simulating $z$ ourselves.}
+
+\end{document}
+
+\add{To return to the context of this thesis, the observed data $y$ is an
+estimated transmission tree for a viral epidemic under investigation. The model
+in question is a contact network model with parameters $\theta$. $z$ is a
+simulated transmission tree, obtained by first generating a contact network
+under the model, and then simulating the spread of an epidemic over that
+network. A transmission tree can be constructed by keeping track of who
 infected whom during the simulated epidemic (further details will be given in
-} \cref{subsec:nettree} \uline{). The distance function $\rho$ must compare two
+\cref{subsec:nettree}). The distance function $\rho$ must compare two
 transmission trees - the observed tree $y$, and the simulated tree $z$. We will
-define this distance function using the tree kernel discussed in}
-\cref{subsec:treeshape}. \uline{In words, the approximate posterior we consider
-here is a distribution which assigns a joint probability density to model
-parameters and simulated transmission trees under those parameters. The
-probability density is proportional to the product of the prior on the
-parameters, and the likelihood of the simulated transmission tree under those
-parameters, but only if the simulated transmission tree is sufficiently close
-to the true tree. Otherwise, the probability density is zero.}}
+define this distance function using the tree kernel discussed in
+\cref{subsec:treeshape}. In words, the approximate posterior we consider here
+is a distribution which assigns a joint probability density to model parameters
+and simulated transmission trees under those parameters. The probability
+density is proportional to the product of the prior on the parameters, and the
+likelihood of the simulated transmission tree under those parameters, but only
+if the simulated transmission tree is sufficiently close to the true tree.
+Otherwise, the probability density is zero.}
 

### 2016-06-14 doc/thesis/objective.tex
Editing and proofreading intro

@@
 network's degree distribution~\autocite{peng2013vaccination, ma2013importance}.
-Phylodynamic methods, {\color{blue}\uline{which link viruses' evolutionary and
-epidemiological dynamics}}, have been used to fit many different types of models
-to phylogenetic data~\autocite{pybus2009evolutionary,volz2013viral}. However,
-these models generally assume a panmictic population. The primary objective of
-this work is \emph{to develop a method to fit contact network models,
-{\color{blue}\uline{and thereby relax the assumption of homogeneous mixing}},
-in a phylodynamic framework.}
+Phylodynamic methods, \add{which link viruses' evolutionary and epidemiological
+dynamics}, have been used to fit many different types of models to phylogenetic
+data~\autocite{pybus2009evolutionary,volz2013viral}. However, these models
+generally assume a panmictic population. The primary objective of this work is
+\emph{to develop a method to fit contact network models, \add{and thereby relax
+the assumption of homogeneous mixing,} in a phylodynamic framework.}
 
@@
 
-{\color{blue}\uline{In this work, we take a Bayesian approach: our goal is to
+\add{In this work, we take a Bayesian approach: our goal is to
 estimate the posterior distribution on model parameters given our data,}
 \[
-    \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)},
+  \madd{\post(\theta \mid T) = \frac{\lik(T \mid \theta) \prior(\theta)}
+    {\int_\Theta \lik(T \mid \theta) \prior(\theta) \d \theta}},
 \]
-\uline{where $\Pr(T \mid \theta)$ is the likelihood, $\Pr(\theta)$ is the prior,
-and $\Pr(T)$ is the marginal probability of $T$ which acts as a normalizing
-constant on the posterior (see} \cref{chp:prelim} \uline{for a review of
-mathematical modeling and Bayesian inference, including definitions of these
-concepts). As we shall show} (see \cref{subsec:just}), \uline{estimating this
-distribution presents computational challenges beyond those usually encountered
-in Bayesian inference. Both the likelihood $\Pr(T \mid \theta)$ and the
-normalizing constant $\Pr(T)$ seem to be intractable, which rules out the use
-of most common \acrlong{ML} and Bayesian methods. }}
+\add{where $\lik(T \mid \theta)$ is the likelihood of the parameters given $T$,
+$\prior(\theta)$ is the prior on $\theta$, and $\Theta$ is the space of
+possible model parameters. The denominator on the right-hand side is the
+marginal probability of $T$ which acts as a normalizing constant on the
+posterior (see \cref{chp:prelim} for a review of mathematical modeling and
+Bayesian inference, including definitions of these concepts). As we shall show
+(\cref{subsec:just}), estimating this distribution presents computational
+challenges beyond those usually encountered in Bayesian inference. Both the
+likelihood $\lik(T \mid \theta)$ and the normalizing constant seem to be
+intractable, which rules out the use of most common \acrlong{ML} and Bayesian
+methods.}
 
-{\color{red}\sout{Calculating the likelihood of the parameters of a contact
-network model seems likely to be an intractable problem. We have not proven
-this is the case, but some intuition can be provided by examining the process
-involved in the likelihood calculation. Consider a contact network model with
-parameters $\theta$ and an estimated transmission tree $T$ with $n$ tips. In
-general, we do not know the labels of the internal nodes of $T$, only the
-labels of its tips. To fit this model using likelihood-based methods, we must
-calculate the likelihood of $\theta$, that is, $\Pr(T \mid \theta)$. Let $\G$
-be the set of all possible contact networks, and $\Nu$ be the set of all
-possible labellings of the internal nodes of $T$. We can write the likelihood
-as}
+\del{Calculating the likelihood of the parameters of a contact network model
+seems likely to be an intractable problem. We have not proven this is the case,
+but some intuition can be provided by examining the process involved in the
+likelihood calculation. Consider a contact network model with parameters
+$\theta$ and an estimated transmission tree $T$ with $n$ tips. In general, we
+do not know the labels of the internal nodes of $T$, only the labels of its
+tips. To fit this model using likelihood-based methods, we must calculate the
+likelihood of $\theta$, that is, $\Pr(T \mid \theta)$. Let $\G$ be the set of
+all possible contact networks, and $\Nu$ be the set of all possible labellings
+of the internal nodes of $T$. We can write the likelihood as}
 \begin{align}
@@
   \label{eq:netlik}
-  \Pr(T \mid \theta)
-    &= \sum_{\nu \in \Nu} \Pr(T, \nu \mid \theta) \\
-    &= \sum_{G \in \G} \sum_{\nu \in \Nu} \Pr(T, \nu \mid G, \theta) \Pr(G \mid \theta) \\
-    &= \sum_{G \in \G} \sum_{\nu \in \Nu} \Pr(T, \nu \mid G) \Pr(G \mid \theta),
+  \mdel{\Pr(T \mid \theta)}
+    &\mdel{=\sum_{\nu \in \Nu} \Pr(T, \nu \mid \theta)} \\
+    &\mdel{=\sum_{G \in \G} \sum_{\nu \in \Nu} \Pr(T, \nu \mid G, \theta) \Pr(G \mid \theta)} \\
+    &\mdel{=\sum_{G \in \G} \sum_{\nu \in \Nu} \Pr(T, \nu \mid G) \Pr(G \mid \theta),}
 \end{split}
 \end{align}
-\sout{the last equality following from the fact that $T$ and $\nu$ depend only
+\del{the last equality following from the fact that $T$ and $\nu$ depend only
 on $G$, not on $\theta$. Although $\Pr(T, \nu \mid G)$ and $\Pr(G \mid \theta)$
@@
 $n$, as there must be at least $n$ nodes in the network. In addition,
-}\cref{eq:netlik} \sout{assumes that $T$ is complete, meaning that all infected
+\cref{eq:netlik} assumes that $T$ is complete, meaning that all infected
 individuals were sampled. This is rarely the case in practice - most often, we
@@
 likelihood calculation becomes even more complex, because we must also sum over
-all possible complete trees.}}
+all possible complete trees.}
 
-{\color{red}\sout{Depending on the network model studied, it is possible that}
-\cref{eq:netlik} \sout{could be simplified into a tractable expression.}}
-An alternative to likelihood-based methods, which could be applied to any
-network model, is provided by \gls{ABC}~\autocite{rubin1984bayesianly,
+\del{Depending on the network model studied, it is possible that
+\cref{eq:netlik} could be simplified into a tractable expression.} An
+alternative to likelihood-based methods, which could be applied to any network
+model, is provided by \gls{ABC}~\autocite{rubin1984bayesianly,
 tavare1997inferring, fu1997estimating, beaumont2002approximate}. All of the
@@
 informative similarity measure called the ``tree
-kernel''~\autocite{poon2013mapping}; {\color{blue}\uline{similar kernel
-functions have been demonstrated to work well as distance functions in
-\gls{ABC}~\autocite{park2015k2}}}. \Gls{ABC} can be implemented with several
+kernel''~\autocite{poon2013mapping}; \add{similar kernel functions have been
+demonstrated to work well as distance functions in
+\gls{ABC}~\autocite{park2015k2}}. \Gls{ABC} can be implemented with several
 algorithms, but \gls{SMC} has advantages over others,
-{\color{blue}\uline{including improved accuracy in low-density regions and
-parallelizability}}~\autocite{mckinley2009inference}. A recently-developed
+\add{including improved accuracy in low-density regions and
+parallelizability}~\autocite{mckinley2009inference}. A recently-developed
 adaptive algorithm requiring minimal tuning on the part of the user makes
@@
 following: stochastic simulation of epidemics on networks, the tree kernel, and
-adaptive \gls{ABC}-\gls{SMC}. {\color{blue}\uline{ Our method will expand on
-the framework developed by~\autocite{poon2015phylodynamic}, who combined
-\gls{ABC} with the tree kernel to infer parameters of population genetic models
-from viral phylogenies using \gls{MCMC}. }}
+adaptive \gls{ABC}-\gls{SMC}. \add{Our method will expand on the framework
+developed by~\autocite{poon2015phylodynamic}, who combined \gls{ABC} with the
+tree kernel to infer parameters of population genetic models from viral
+phylogenies using \gls{MCMC}.}
 
@@
 
-{\color{blue}\uline{
-The remainder of this background chapter is organized in four sections. The
-first section introduces phylogenies and transmission trees, which are the
+\add{The remainder of this background chapter is organized in four sections.
+The first section introduces phylogenies and transmission trees, which are the
 input data from which our method aims to make statistical inferences. This
@@
 setting, while \gls{SMC} is a particular algorithm which can be used to
-implement \gls{ABC}. }}
+implement \gls{ABC}.}
 

### 2016-06-14 doc/thesis/preamble.tex
Editing and proofreading intro

@@
 \usepackage[normalem]{ulem}
+\usepackage{calc}
 
@@
 \newcommand{\maps}{:}
+\newcommand{\vx}{\vec{x}}
+
+\newcommand{\prior}{\pi}
+\newcommand{\lik}{f}
+\newcommand{\post}{\pi}
+
+\let\mref\cref
+\let\Mref\Cref
+\let\mtextcite\textcite
+\let\mautocite\autocite
+\renewcommand{\cref}[1]{\mbox{\mref{#1}}}
+\renewcommand{\Cref}[1]{\mbox{\Mref{#1}}}
+\renewcommand{\textcite}[1]{\mbox{\mtextcite{#1}}}
+\renewcommand{\autocite}[1]{\mbox{\mautocite{#1}}}
+\newcommand{\add}[1]{\color{blue}\uline{#1} \color{black}}
+\newcommand{\madd}[1]{\color{blue}\underline{#1} \color{black}}
+\newcommand{\del}[1]{\color{red}\sout{#1} \color{black}}
+% http://tex.stackexchange.com/a/20612/60841
+\newsavebox\CBox
+\newcommand\mdel[1]{%
+  \color{red}%
+  \ifmmode\sbox\CBox{$#1$}\else\sbox\CBox{#1}\fi%
+  \makebox[0pt][l]{\usebox\CBox}%
+  \rule[0.5\ht\CBox-0.25pt]{\wd\CBox}{0.5pt}
+  \color{black}%
+}
+%TC:macro \del [ignore]
+%TC:macro \mdel [ignore]
 
@@
 \DeclareMathOperator{\nl}{nl}
-
-\newfloat{data}{hb}{lod}
-\floatname{data}{Data}
-\crefname{data}{data}{data}
 
@@
   name=\ensuremath{n},
-  description=number of particles used for ABC-SMC
+  description=number of particles used for sequential Monte Carlo
 }
@@
   description=per-iteration rate of expected sample size decay in adaptive ABC-SMC
+}
+\newglossaryentry{Theta}{
+  name=\ensuremath{\Theta},
+  description=parameter space for a given mathematical model
 }

### 2016-06-14 doc/thesis/preface.tex
Editing and proofreading intro

--- 2016-05-17 doc/thesis/preface.tex
@@
 Evolution meeting on April 25, 2016, in Woods Hole, Massachusetts, USA (the
-presentation was delivered remotely). 
-%A poster based on chapter 2 entitled
-%``Likelihood-free estimation of contact network parameters from viral
-%phylogenies'' was presented at the Intelligent Systems for Molecular Biology
-%meeting in Orlando, Florida, USA.
+presentation was delivered remotely). A poster based on chapter 2 entitled
+``Likelihood-free estimation of contact network parameters from viral
+phylogenies'' is scheduled for presentation at the Intelligent Systems for
+Molecular Biology meeting on July 8, 2016, in Orlando, Florida, USA.
 
@@
 reviewed and approved by the UBC/Providence Health Care Research Ethics Board
-(H07-02559).
+(H07-02559). Rosemary M. McCloskey completed the Tri-Council Policy Statement:
+Ethical Conduct for Research Involving Humans Course on Research Ethics (TCPS
+2: CORE) tutorial on June 8, 2016.
+
+\add{Source code for the \software{netabc} program is freely available at
+\url{https://github.com/rmcclosk/netabc} under the GPL-3 license. Scripts to
+run all computational experiments, as well as the source code for this thesis,
+are available at \url{https://github.com/rmcclosk/thesis}.}

### 2016-06-14 doc/thesis/discussion.tex
Writing from last night

--- 2016-06-11 doc/thesis/discussion.tex
@@
+%<<setup, include=FALSE>>=
+%    source("global.R")
+%@
+
 \label{sec:disc}
@@
 \subsection{\software{Netabc}: uses, limitations, and possible extensions}
+
+\glsreset{ER}
+\glsreset{WS}
 
@@
 \software{netabc}, a method based on kernel-ABC to infer the parameters of a
-contact network model. The method is general, meaning that it can be used to
-infer parameters of any network model, as long as it allows simulated networks
-can be easily generated. We have included generators for the \gls{BA} model
-discussed here, as well as the \gls{ER} and \gls{WS} network models.
-Instructions for adding additional models are available in the project's online
-documentation. We have made \software{netabc} publicly available at
+contact network model. The method \del{is general, meaning that it} can be used
+to infer parameters of any network model, as long as it allows simulated
+networks to be easily generated. We have included generators for the \gls{BA}
+model discussed here, as well as the \gls{ER}~\autocite{erdos1960evolution} and
+\gls{WS}~\autocite{watts1998collective} network models.  Instructions for
+adding additional models are available in the project's online documentation.
+We have made \software{netabc} publicly available at
 \url{github.com/rmcclosk/netabc} under a permissive open source license, to
@@
 Several alternative network models and modelling frameworks have been developed
-which may provide useful future targets for kernel-assisted \gls{ABC}. Waring
+which may provide useful future targets for \gls{ABC}. Waring
 models~\autocite{irwin1963place,handcock2004likelihood} are a more flexible
@@
 \gls{ERGM} parameterizations on transmission tree shape and effective
-population size. The author suggested the use of \glspl{ERGM} as a general
-framework for estimation of epidemiological quantities related to \gls{HIV}
-transmission. Except for a few special cases, simulating a network according to
-an \gls{ERGM} generally requires \gls{MCMC}, which would be too computationally
-intensive to integrate into \software{netabc} as it currently stands. To fit
-\gls{ERGM} with kernel-assisted \gls{ABC}, one possibility would be to consider the
-network itself as a parameter to be modified by the \gls{MCMC} kernel.
-Other network modelling frameworks include the partnership-centric formulation
-developed by~\textcite{eames2002modeling} and the log-linear adjacency matrix
+population size, and suggested the use of \glspl{ERGM} as a general framework
+for estimation of epidemiological quantities related to \gls{HIV} transmission.
+Except for a few special cases, simulating a network according to an \gls{ERGM}
+generally requires \gls{MCMC}, which would be too computationally intensive to
+integrate into \software{netabc} as it currently stands. To fit \gls{ERGM} with
+\gls{ABC}, one possibility would be to consider the network
+itself as a parameter to be modified by the \gls{MCMC} kernel.  Other network
+modelling frameworks include the partnership-centric formulation developed
+by~\textcite{eames2002modeling} and the log-linear adjacency matrix
 parameterization applied by~\textcite{morris1993epidemiology}.
-
-%\Textcite{morris1993epidemiology} proposes to apply the standard compartmental
-%modelling framework to contact networks by assigning each individual their own
-%compartment. Thus, each individual is associated with a single \gls{ODE}, with
-%the entire \gls{ODE} system parameterized by the adjacency matrix of the
-%contact network. 
-%The author proposes to use log-linear models to parameterize the adjacency
-%5matrix of a contact network. This framework is highly expressive, and allows
-%straightforward incorporation of time-dependent dynamics. 
-%However, simulating a transmission tree would require the numerical solution of
-%a very large system of \glspl{ODE}, which may be computationally prohibitive
-%given the number of simulations required for kernel-assisted \gls{ABC}.
 
@@
 contact network is static over the duration of the epidemic. Clearly this
-assumption is invalid, as people make and break partnerships on a regular
-basis. Addressing the impact of this simplifying assumption is outside the
-scope of this work. However, the same assumption is made by most studies using
-contact network models in an epidemiological
-context~\autocite{welch2011statistical, bansal2007individual}. In principle,
-kernel-assisted \gls{ABC} could be adapted to dynamic contact networks by using
-a method such as that developed by \textcite{robinson2012dynamics} to simulate
-such a network, while concurrently simulating the spread of an epidemic.
+assumption is false, as people make and break partnerships on a regular basis.
+Addressing the impact of this simplifying assumption is outside the scope of
+this work. However, the same assumption is made by most studies using contact
+network models in an epidemiological context~\autocite{welch2011statistical,
+bansal2007individual}. In principle, \gls{ABC} could be adapted
+to dynamic contact networks by using a method such as that developed by
+\textcite{robinson2012dynamics} to simulate such a network, while concurrently
+simulating the spread of an epidemic.
 
@@
 process by which a viral phylogeny is generated along with a transmission tree
-into our method, for example by simulating within-host dynamics. {\color{blue}
-\uline{Indeed, progress has very recently been demonstrated on this front in a
-talk by \citeauthor{giardina2016inference}, who have independently developed in
-a method similar to ours to fit contact network models to phylogenetic data
-that additionally incorporates a within-host evolutionary model.}} Although
-this may be an avenue for future extension, we felt that it would obscure the
-primary purpose of this work, which is to study contact network parameters. In
+into our method, for example by simulating within-host dynamics. \add{Indeed,
+progress has very recently been demonstrated on this front in a talk by
+\textcite{giardina2016inference}, who have independently developed in a method
+similar to ours to fit contact network models to phylogenetic data that
+additionally incorporates a within-host evolutionary model.} Although this may
+be an avenue for future extension, we felt that it would obscure the primary
+purpose of this work, which is to study contact network parameters. In
 addition, there are a number of different methods available for inferring
@@
 highly unbalanced ladder-like tree structure (\cref{fig:alphatrees}). 
-There was no observable bias in the estimates of \gls{alpha} obtained with
-\software{netabc}, however {\color{red}\sout{the variation in these estimates
-was higher for $\alpha < 1$ than for $\alpha \geq 1$}} {\color{blue}\uline{there
-appeared to be weaker identifiability for \gls{alpha} $<$ 1 than for \gls{alpha}
-$\geq$ 1}} (\cref{fig:abcptm2,tab:abchpd}). The relationship between
-\gls{alpha} and the power law exponent \gls{gamma} may explain this result
-(\cref{fig:gamma}). The \gls{gamma} values associated with \gls{alpha} = 0 and
-\gls{alpha} = 0.5 are nearly identical (about 2.28 for \gls{alpha} = 0 and 2.33
-for \gls{alpha} = 0.5 with \gls{N} = 5000 and \gls{m} = 2). In other words, the
-degree distributions of networks with \gls{alpha} $< 1$ are similar to each
-other, which may result in similarity of corresponding transmission trees as
-well. 
-%The accuracy of \gls{alpha} estimates also increased when the value of the
-%\gls{m} parameter was higher (compare
-%\cref{fig:abcptm2,fig:abcptm3,fig:abcptm4}).
+\del{There was no observable bias in the estimates of \gls{alpha} obtained with
+\software{netabc}, however the variation in these estimates
+was higher for $\alpha < 1$ than for $\alpha \geq 1$} \add{There appeared to be
+weaker identifiability for \gls{alpha} $<$ 1 than for \gls{alpha} $\geq$ 1}
+(\cref{fig:abcptm2,tab:abchpd}), which may be partially explained by the
+relationship between \gls{alpha} and the power law exponent \gls{gamma}
+(\cref{fig:gamma}). \add{Although the degree distributions do not truly follow
+a power law for $\gls{alpha} \neq 1$, the fitted exponent still captures the
+slope of the degree distribution reasonably well (\cref{fig:powerlaw}), and 
+distributions with similar fitted \gls{gamma} values appear qualitatively
+similar in other respects.} The \gls{gamma} values fitted to \gls{alpha} = 0
+and \gls{alpha} = 0.5 are nearly identical (about 2.28 for \gls{alpha} = 0 and
+2.33 for \gls{alpha} = 0.5 with \gls{N} = 5000 and \gls{m} = 2). In other
+words, the degree distributions of networks with \gls{alpha} $< 1$ are similar
+to each other, which may result in similarity of corresponding transmission
+trees as well.
 
@@
 also generally identifiable, although it was slightly over-estimated for both
-cases we considered with kernel-assisted \gls{ABC}. The dynamics of the \gls{SI} model,
-and the coalescent process~\autocite{kingman1982coalescent}, offer a potential
-explanation for the identifiability of \gls{I}. In our simulations, we assumed
-that all discordant edges shared the same transmission rate, so that the
-waiting time until the next transmission in the entire network was always
-inversely proportional to the number of discordant edges. In the initial phase
-of the epidemic, when \gls{I} is small, each new transmission results in many
-new discordant edges. Hence, there is an early exponential growth phase,
-producing many short branches near the root of the tree. As the epidemic gets
-closer to saturating the network, the number of discordant edges decays,
-causing longer waiting times. The distribution of coalescence times in the tree
-should therefore be informative about \gls{I}~\autocite{volz2009phylodynamics}.
+cases we considered with \gls{ABC}. \add{Sackin's index was better able to
+capture the effect of \gls{I} on tree shape than the \gls{nltt}
+(\cref{fig:rsquared,fig:Icrossv}), suggesting that this parameter impacts the
+distribution of branching times in the tree more than the topology. In a
+homogeneously-mixed population, branching times can be modelled by the
+coalescent process~\autocite{kingman1982coalescent}, in our case under the
+\gls{SI} model~\autocite{volz2009phylodynamics}. Although our populations are
+not homogeneously mixed, the forces which affect the distributions of branching
+times still apply.} \del{The dynamics of the \gls{SI} model, and the coalescent
+process, offer a potential explanation for the identifiability of \gls{I}.} In
+our simulations, we assumed that all discordant edges shared the same
+transmission rate, so that the waiting time until the next transmission in the
+entire network was always inversely proportional to the number of discordant
+edges. In the initial phase of the epidemic, when \gls{I} is small, each new
+transmission results in many new discordant edges. Hence, there is an early
+exponential growth phase, producing many short branches near the root of the
+tree. As the epidemic gets closer to saturating the network, the number of
+discordant edges decays, causing longer waiting times. \del{The distribution of
+coalescence times in the tree should therefore be informative about \gls{I}.
 This information is captured by the tree kernel, and also by the \gls{nltt}
 statistic, which both performed quite well in classifying \gls{I}
-(\cref{fig:rsquared}).
+(\cref{fig:rsquared}).}
 
@@
 simulated under different \gls{N} values when the number of infected nodes
-\gls{I} was small. There is an intuitive explanation for this result, namely
-that adding additional nodes does not change the edge density or overall shape
-of a \gls{BA} network. This can be illustrated by imagining that we add a small
+\gls{I} was small (\cref{fig:Nkpca,fig:Ncrossv}). There is an intuitive
+explanation for this result, namely that adding additional nodes does not
+change the edge density or overall shape of a \gls{BA} network. \add{In fact,
+this is why the degree distributions of these networks are called
+``scale-free.''} This can be illustrated by imagining that we add a small
 number of nodes to a network after the epidemic simulation has already been
@@
 hand, when \gls{I} is large relative to \gls{N}, the coalescent dynamics
-discussed above also apply. That is, the waiting times until the next infection
+discussed above also apply. The waiting times until the next infection
 increase, resulting in longer coalescence times toward the tips. The relative
@@
 \gls{nltt} uses only information about the coalescence times. When all \gls{BA}
-parameters were simultaneously estimated with kernel-assisted \gls{ABC}, \gls{N} was
-nearly always over-estimated by approximately a factor of two
+parameters were simultaneously estimated with \gls{ABC}, \gls{N} was nearly
+always over-estimated by approximately a factor of two
 (\cref{fig:abcptm2,tab:abchpd}). One factor which may have contributed to this
@@
 assigned to higher \gls{N} values. Another contributing factor relates to the
-dynamics of the \gls{SI} model and the coalescent process.
+dynamics of the \gls{SI} model and the coalescent process and is discussed
+below.
 
@@
 growth curves to the proportion of infected individuals over time, and
-calculated the first and second derivatives of these curves at the time of
-transmission tree sampling. These derivatives give us an indication of the
-growth rates of the epidemics at the time of sampling. As shown in
-\cref{fig:derivs}, there are bands along which both derivatives are similar
-which contain the values we tested. These bands span mostly higher values of
-\gls{N} and \gls{I} than the true values. Therefore, if \gls{N} and \gls{I} are
-free to vary (as is the case in kernel-assisted \gls{ABC}), and our hypothesis is true,
-both parameters will tend to be overestimated due to being less identifiable
-within their own band. However, when \gls{N} is fixed at 5000, the derivatives
-vary substantially along the \gls{I}-axis, which explains why the grid search
-estimates of \gls{I} were accurate and unbiased (\cref{fig:gridptI}).
+calculated the first and second derivatives of these curves (with respect to
+time) at the time of transmission tree sampling. These derivatives give us an
+indication of the growth rates of the epidemics at the time of sampling. As
+shown in \cref{fig:derivs}, there are bands along which both derivatives are
+similar which contain the values we tested. These bands span mostly higher
+values of \gls{N} and \gls{I} than the true values. Therefore, if \gls{N} and
+\gls{I} are free to vary (as is the case in \gls{ABC}), and our hypothesis is
+true, both parameters will tend to be overestimated due to being less
+identifiable within their own band. However, when \gls{N} is fixed at 5000, the
+derivatives vary substantially along the \gls{I}-axis, which explains why the
+grid search estimates of \gls{I} were accurate and unbiased
+(\cref{fig:gridptI}).
 
@@
   \centering
-  \includegraphics{derivatives.pdf}
+  %\includegraphics{derivatives.pdf}
   \caption[
-    First and second derivatives of epidemic growth curves at time of sampling
-    for various values of $I$ and $N$.
+    First and second time derivatives of epidemic growth curves at time of
+    sampling for various values of $I$ and $N$.
   ]{
-    First and second derivatives of epidemic growth curves at time of sampling
-    for various values of \gls{I} and \gls{N}. Networks were simulated under
-    the \gls{BA} model with \gls{alpha} = 1.0, \gls{m} = 2, and \gls{N} varied 
-    along the values shown on the $x$-axis. Transmission trees were sampled at
-    the time when \gls{I} nodes were infected ($y$-axis). Logistic growth
-    curves were fit to epidemic trajectories derived from the transmission
-    trees, and their first and second derivatives were calculated at the time
-    of sampling. Contours show first derivatives, while colours indicate second
-    derivatives. Values of \gls{I} and \gls{N} used in simulation experiments
-    with kernel-assisted \gls{ABC} are indicated by diamonds.
+    First and second \add{time} derivatives of epidemic growth curves at time
+    of sampling for various values of \gls{I} and \gls{N}. Networks were
+    simulated under the \gls{BA} model with \gls{alpha} = 1.0, \gls{m} = 2, and
+    \gls{N} varied along the values shown on the $x$-axis. Transmission trees
+    were sampled at the time when \gls{I} nodes were infected ($y$-axis).
+    Logistic growth curves were fit to epidemic trajectories derived from the
+    transmission trees, and their first and second derivatives were calculated
+    at the time of sampling. Contours show first derivatives, while colours
+    indicate second derivatives. Values of \gls{I} and \gls{N} used in
+    simulation experiments with \gls{ABC} are indicated by diamonds.
   }
@@
 network per vertex, did not have a measurable impact on tree shape and was not
-identifiable with kernel-ABC. The exception to this was the value \gls{m} = 1,
+identifiable with \gls{ABC}, \add{nor with grid search when the other
+parameters were held fixed.} The exception to this was the value \gls{m} = 1,
 which produces networks without cycles whose associated trees were more easily
-distinguished. However, all the analyses presented here did not take the
-absolute size of the transmission trees into account, as the branch lengths
-were rescaled by their mean. Because higher \gls{m} values imply higher edge
-density, an epidemic should spread more quickly for higher \gls{m} than lower
-\gls{m} with the same per-edge transmission probability.  Hence, considering
-the absolute height of the trees may improve our method's ability to
-reconstruct \gls{m}.
+distinguished (\cref{fig:gridest,fig:gridm}). However, all the analyses
+presented here did not take the absolute size of the transmission trees into
+account, as the branch lengths were rescaled by their mean. Because higher
+\gls{m} values imply higher edge density, an epidemic should spread more
+quickly for higher \gls{m} than lower \gls{m} with the same per-edge
+transmission probability.  Hence, considering the absolute height of the trees
+may improve our method's ability to reconstruct \gls{m}. \add{It was also
+pointed out to us by an anonymous reviewer that for a fixed \gls{I}, an
+infected node may only end up transmitting along a fraction of its outgoing
+edges, which could mask the presence of the extra edges associated with higher
+\gls{m}}.
+
+\end{document}
 
@@
 
+%<<disc, include=FALSE>>=
+%    d <- setDT(collect.data("../../simulations/aggregates/hpd/*"))
+%@
+
 \subsubsection{Preferential attachment power}
@@
 sub-linear, ranging from 
-    0.27
+    X%\S%expr{d[m_min == 1 & parameter == "alpha", round(min(mean), 2)]}
 to 
-    0.95.
+    X%\S%expr{d[m_min == 1 & parameter == "alpha", round(max(mean), 2)]}.
 The sub-linearity is consistent with the results of
@@
 \gls{alpha} for these data was 
-    0.55, 0.55
+    X%\S%expr{d[m_min == 1 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]}
         or 
-    0.53, 0.53,
+    X%\S%expr{d[m_min == 2 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]},
 depending on the prior on \gls{m}; \citeauthor{de2007preferential} estimated

### 2016-06-14 doc/thesis/intro.tex
Writing from last night

@@
 
-\end{document}
-
 \add{To return to the context of this thesis, the observed data $y$ is an
 estimated transmission tree for a viral epidemic under investigation. The model
-in question is a contact network model with parameters $\theta$. $z$ is a
-simulated transmission tree, obtained by first generating a contact network
-under the model, and then simulating the spread of an epidemic over that
-network. A transmission tree can be constructed by keeping track of who
+in question is a contact network model with parameters $\theta$. The simulated
+dataset $z$ is a transmission tree, obtained by first generating a contact
+network under the model, and then simulating the spread of an epidemic over
+that network. A transmission tree can be constructed by keeping track of who
 infected whom during the simulated epidemic (further details will be given in
 \cref{subsec:nettree}). The distance function $\rho$ must compare two
-transmission trees - the observed tree $y$, and the simulated tree $z$. We will
-define this distance function using the tree kernel discussed in
-\cref{subsec:treeshape}. In words, the approximate posterior we consider here
-is a distribution which assigns a joint probability density to model parameters
-and simulated transmission trees under those parameters. The probability
-density is proportional to the product of the prior on the parameters, and the
-likelihood of the simulated transmission tree under those parameters, but only
-if the simulated transmission tree is sufficiently close to the true tree.
-Otherwise, the probability density is zero.}
-
-{\color{red}\sout{As we shall see in the next section, this distribution can
-be sampled from exactly. The word ``approximate'' derives from assumption that,
-for a suitably chosen distance $\rho$ and a small enough $\varepsilon$, the
-marginal in $z$ of this distribution approximates the posterior of
-interest~\autocite{marin2012approximate}. That is, }}
-{\color{blue}\uline{In fact, it is not the \gls{ABC} target distribution
+transmission trees - the observed tree, which takes the place of $y$, and the
+simulated tree $z$. We will define this distance function using the tree kernel
+discussed in \cref{subsec:treeshape}. In words, the approximate posterior we
+consider here is a distribution which assigns a joint probability density to
+model parameters and simulated transmission trees under those parameters. The
+probability density is proportional to the product of the prior on the
+parameters, and the likelihood of the simulated transmission tree under those
+parameters, but only if the simulated transmission tree is sufficiently close
+to the true tree.  Otherwise, the probability density is zero.}
+
+\del{As we shall see in the next section, this distribution can be sampled from
+exactly. The word ``approximate'' derives from assumption that, for a suitably
+chosen distance $\rho$ and a small enough $\varepsilon$, the marginal in $z$ of
+this distribution approximates the posterior of
+interest~\autocite{marin2012approximate}. That is,}
+\add{In fact, it is not the \gls{ABC} target distribution
 itself, but rather its marginal in $z$, which approximates the posterior
-distribution. In other words, we claim that }} 
-\[
-  \int \pi_\varepsilon(\theta, z \mid y) \d z \approx \Pr(\theta \mid y).
-\]
-The intuition for why this approximation might be reasonable
-comes from the fact that, when $\varepsilon = 0$, the $\varepsilon$-ball around
-$y$ should contain only $y$ itself, hence the integral on the left is exactly
-equal to the posterior. Thus, by taking $\varepsilon$ small, we should attain
-something close to the posterior {\color{blue}\uline{if $\rho$ captures the
-similarity between datasets reasonably well. However, the accuracy of the
-\gls{ABC} approximation depends heavily on the choice of distance function
-\autocite{aeschbacher2012novel,blum2013comparative}.}}
+distribution. In other words, we claim that}
+\[
+  \int \pi_\varepsilon(\theta, z \mid y) \d z \approx \post(\theta \mid y).
+\]
+The intuition for why this approximation might be reasonable comes from
+\add{considering the case when $\varepsilon$ = 0 and $\rho$ has the property
+that $\rho(z, y)$ if and only if $x = y$. In that case,} \del{the fact that,
+when $\varepsilon = 0$,} the $\varepsilon$-ball around $y$ should contain only
+$y$ itself, hence the integral on the left is exactly equal to the posterior.
+Thus, by taking $\varepsilon$ small, we should attain something close to the
+posterior \add{if $\rho$ captures the similarity between datasets reasonably
+well. However, the accuracy of the \gls{ABC} approximation depends heavily on
+the choice of distance function
+\autocite{aeschbacher2012novel,blum2013comparative}.}
 
@@
 
-In many applications {\color{blue}\uline{(eg. \autocite{fu1997estimating,
-tanaka2006using})}}, $\rho$ is defined as $\rho(S(\cdot), S(\cdot))$ where $S$
-is a function which maps data points into a vector of summary statistics.
-{\color{blue}\uline{In the context of \gls{ABC}, a summary statistic $S$ is
-called \defn{sufficient} if} 
-\[
-  \Pr(\theta \mid y) = \Pr(\theta \mid S(y)).
-\]
-\uline{That is, sufficiency implies that the data can be replaced with the
+In many applications \add{(eg. \autocite{fu1997estimating,tanaka2006using})},
+$\rho$ is defined as $\rho(S(\cdot), S(\cdot))$ where $S$ is a function which
+maps data points into a vector of summary statistics.
+\add{In the context of \gls{ABC}, a summary statistic $S$ is called
+\defn{sufficient} if}
+\[
+  \post(\theta \mid y) = \post(\theta \mid S(y)).
+\]
+\add{That is, sufficiency implies that the data can be replaced with the
 summary statistic without losing any information about the posterior
@@
 A number of sophisticated methods have been developed for selecting and
-weighting summary statistics based on various optimality criteria}
-\autocite[][and references therein]{aeschbacher2012novel,
-blum2013comparative}}.
-
-Summary statistics can be useful if the data are high-dimensional or of a
-complex type, but they are not strictly necessary. For instance, if the data
-are numeric and of low dimension, the distance function may simply be the
-Euclidean distance~\autocite{sisson2007sequential}.
-{\color{blue}\textcite{park2015k2} \uline{proposed the use of a kernel function}
-(defined in \cref{subsec:treeshape}) \uline{in place of a distance function.
-The authors referred to their approach as ``double-kernel \gls{ABC}'' due to
-the use of a second kernel function to compute the weights of the particles.
-The work by \textcite{poon2015phylodynamic}, upon which ours is based, employed
-a similar approach, replacing the likelihood ratio in Bayesian \gls{MCMC} with
-a ratio of kernel scores.}}
+weighting summary statistics based on various optimality criteria
+\autocite[][and~references~therein]{aeschbacher2012novel,blum2013comparative}.
+We do not apply these methods in this work, instead focusing on a distance
+function which is not based on summary statistics.}
+
+\del{Summary statistics can be useful if the data are high-dimensional or of a
+complex type, but}\add{Although summary statistics are often presented as a
+fundamental part of \gls{ABC},} they are not strictly necessary. For instance,
+if the data are numeric and of low dimension, the distance function may simply
+be the Euclidean distance~\autocite{sisson2007sequential}.
+\add{\textcite{park2015k2} proposed the use of a kernel function
+(as defined in \cref{subsec:treeshape}) in place of a distance function.  The
+authors referred to their approach as ``double-kernel \gls{ABC}'' due to the
+use of a second (unrelated) kernel function to compute the weights of the
+particles. The work by \textcite{poon2015phylodynamic}, upon which ours is
+based, employed a similar approach, replacing the likelihood ratio in Bayesian
+\gls{MCMC} with a ratio of kernel scores.}
 
@@
 
-Algorithms for performing \gls{ABC} fall into one of three categories:
+Algorithms for performing \gls{ABC} can be grouped into three categories:
 rejection, \gls{MCMC}, and \gls{SMC}~\autocite{marin2012approximate}. To
@@
 the variable $x$ to refer to the pair $(\theta, z)$, so that the \gls{ABC}
-target distribution can be written $\pi_\varepsilon(x \mid y)$.
+target distribution can be written $\pi_\varepsilon(x \mid y)$. \add{This makes
+our notation consistent with \cref{sec:smc}.}
 
@@
 \[
-  \pi(\theta) f(z \mid \theta) \I_{A_{\varepsilon, y}}(z),
+  \prior(\theta) \lik(z \mid \theta) \I_{A_{\varepsilon, y}}(z),
 \]
@@
 smoothly through a sequence of target distributions which ends at the \gls{ABC}
-approximation to the posterior. {\color{blue}\uline{The initial value
-$\varepsilon_0$ is set to $\infty$, which makes the first distribution in the
-sequence}
-\[
-  \pi_0(\theta, z) = \frac{\Pr(\theta) \Pr(z \mid \theta)}
-               {\int_{\R \times \Theta} \Pr(\theta) \Pr(z \mid \theta) \d\theta \d z}.
-\]
-\uline{For brevity, we have ommitted the dependence on the observed data $y$.
-In the terminology of} \cref{alg:smc}, \uline{the numerator is the first of the
-$\gamma$'s, that is, $\gamma_0 = \Pr(\theta) \Pr(z \mid \theta)$. Sampling in
-proportion to $\gamma_0$ is straightforward and was already demonstrated for
-rejection \gls{ABC} above. Because the sampling is exact, the initial
-importance weights are all set equal to 1 and normalized to $1/n$ where $n$ is
-the number of particles.}}
+approximation to the posterior. \add{The initial value $\varepsilon_1$ is set
+to $\infty$, which makes the first distribution in the sequence}
+\[
+  \madd{\pi_1(\theta, z) = \frac{\prior(\theta) \lik(z \mid \theta)}
+  {\int_{\R \times \Theta} \prior(\theta) \lik(z \mid \theta) \d\theta \d z}}.
+\]
+\add{This initial distribution does not depend on the observed data $y$.  In
+the terminology of the \gls{SMC} sampler (\cref{alg:smc}), the numerator is the
+first of the $\gamma$'s, that is, $\gamma_1 = \prior(\theta) \lik(z \mid
+\theta)$. Sampling in proportion to $\gamma_1$ is straightforward and was
+already demonstrated for rejection \gls{ABC} above. Because the sampling is
+exact, the initial importance weights are all set equal to 1 and normalized to
+$1/\gls{n}$ where \gls{n} is the number of particles.}
 
@@
 Several options have been proposed~\autocite{beaumont2009adaptive,
-sisson2007sequential, del2012adaptive}. {\color{blue}\uline{With an
-appropriately chosen kernel, the weight update} (\cref{eq:smcwt}) \uline{will
-simplify into a computable expression.} \textcite{sisson2007sequential} and
-\textcite{beaumont2009adaptive} \uline{both suggest random walk kernels for
-$K$, where each particle is
-perturbed according to a Gaussian distribution. The backwards kernels $L$ are
-chosen to approximately minimize the variance in importance weights. In this
-thesis, we use an \gls{MCMC} kernel, as proposed by} \textcite{del2012adaptive}.
-\uline{The associated backwards kernels and weight updates are discussed in the
-next chapter} (\cref{subsec:adaptsmc}). \uline{With generic kernels $K$ and
-$L$, the \gls{ABC}-\gls{SMC} algorithm is almost identical to the \gls{SMC}
-sampler} (\cref{alg:smcsamp}), \uline{with $\gamma_i$ replaced with $\pi_i$.}
-Therefore we will not repeat the full algorithm here.}
+sisson2007sequential, del2012adaptive}. \add{With an appropriately chosen
+kernel, the weight update (\cref{eq:smcwt}) will simplify into a computable
+expression. \textcite{sisson2007sequential} and \textcite{beaumont2009adaptive}
+both suggest random walk kernels for $K$, where each particle is perturbed
+according to a Gaussian distribution. The backwards kernels $L$ are chosen to
+approximately minimize the variance in importance weights. In this thesis, we
+use an \gls{MCMC} kernel, as proposed by \textcite{del2012adaptive}.  The
+associated backwards kernels and weight updates are discussed in the next
+chapter (\cref{subsec:adaptsmc}). With generic kernels $K$ and $L$, the
+\gls{ABC}-\gls{SMC} algorithm is almost identical to the \gls{SMC} sampler
+(\cref{alg:smcsamp}), with $\gamma_i$ replaced with $\pi_i$.  Therefore we will
+not repeat the full algorithm here.}
 
@@
 
-{\color{blue}\uline{In} \cref{sec:obj}, \uline{we outlined three research
-objectives that will be addressed in this thesis. First, we aim to develop a
-method for fitting contact network models to estimated transmission trees.
-Although transmission trees can be estimated for any epidemic by thorough
-contact tracing, viral diseases are the most useful context for this method.
-Transmission and sequence evolution occur on similar time scales for RNA
-viruses, resulting in viral phylogenies whose shapes are heavily constrained by
-the transmission process. The study of this interaction between evolution and
-epidemiology is called \emph{phylodynamics}. Phylodynamic methods make it
-possible to estimate transmission trees from viral sequence data; these
-estimated trees form the input data for our method.}
-
-\uline{From these estimated transmission trees, our method's purpose is to
-estimate the parameters of a contact network model. Rather than assuming a
-homogeneously mixed population, as most epidemiological models do, network
-models take the more realistic view that human populations are structured. That
-is, contacts which allow for transmission may occur in a nonrandom way, rather
-than between every pair of individuals in the population. A network model
-parameterizes this structure. In particular, our second research objective is
-to characterize our ability to fit the \gls{BA} model, which incorporates
-preferential attachment to generate networks with realistic degree
-distributions.}
-
-\uline{Our method will apply \gls{ABC}, a simulation-based approach, for
-fitting these models. Simulating a transmission tree according to a network
-model is straightforward: a network can be generated according to the model,
-and the spread of an epidemic can be simulated over the network and recorded in
-a transmission tree. \gls{ABC} uses the concordance between these simulated
+\add{In \cref{sec:obj}, we outlined three research objectives that will be
+addressed in this thesis. First, we aim to develop a method for fitting contact
+network models to estimated transmission trees. Although transmission trees
+can be estimated for any epidemic by thorough contact tracing, viral diseases
+are the most useful context for this method due to the possibility of inferring
+the trees from sequence data. Transmission and sequence evolution occur on
+similar time scales for RNA viruses, resulting in viral phylogenies whose
+shapes are heavily constrained by the transmission process. The study of this
+interaction between evolution and epidemiology is called \emph{phylodynamics};
+phylodynamic methods make it possible to estimate transmission trees from viral
+sequence data. These estimated trees form the input data for our method.}
+
+\add{The desired output of our method is a posterior distribution of the
+parameters of a contact network model. Rather than assuming a homogeneously
+mixed population, as most epidemiological models do, network models take the
+more realistic view that human populations are structured. That is, contacts
+which allow for transmission may occur in a nonrandom way, rather than between
+every pair of individuals in the population. A network model parameterizes this
+structure. In particular, our second research objective is to characterize our
+ability to fit the \gls{BA} model, which incorporates preferential attachment
+to generate networks with realistic degree distributions.}
+
+\add{To fit these models, our method will apply \gls{ABC}, a simulation-based
+approach. Simulating a transmission tree according to a network model is
+straightforward: a network can be generated according to the model, and the
+spread of an epidemic can be simulated over the network and recorded in a
+transmission tree. \Gls{ABC} uses the concordance between these simulated
 transmission trees and the ``true'' estimated tree as an indicator of parameter
@@
 
-\uline{\gls{ABC} can be implemented by at least three algorithms, but the one
-we choose to apply in this work is \gls{SMC}. \Gls{SMC} uses a population of
-parameter ``particles'' to approximate a distribution of interest, in this case
-the approximate posterior distribution targeted by \gls{ABC}. After running the
-\gls{ABC}-\gls{SMC} algorithm, statistics on the model parameters can be
-approximated by the weighted population of particles. For example, a weighted
-average would give an approximate expected value for each parameter.}
-
-\uline{Thus, our method integrates four research topics: phylogenetics, contact
+\add{\Gls{ABC} can be implemented by at least three classes of algorithm, but
+the one we choose to apply in this work is \gls{SMC}. \Gls{SMC} uses a
+population of parameter ``particles'' to approximate a distribution of
+interest, in this case the approximate posterior distribution targeted by
+\gls{ABC}. After running the \gls{ABC}-\gls{SMC} algorithm, statistics on the
+model parameters can be approximated by the weighted population of particles.
+For example, a weighted average would give an approximate expected value for
+each parameter.}
+
+\add{Thus, our method integrates four research topics: phylogenetics, contact
 networks, \acrlong{SMC}, and \acrlong{ABC}. The first two topics together form
@@
 The latter two topics define the algorithm and statistical framework that our
-inference method will use. }}
+inference method will use.}

### 2016-06-14 doc/thesis/methods_results.Rnw
Writing from last night

@@
 
-{\color{blue}\uline{ In this chapter, we will address the three research aims
-of this thesis introduced in} \cref{sec:obj}. \uline{First, in}
-\cref{sec:netabc}, \uline{we describe \software{netabc}, a computer program
-that implements an \acrlong{ABC}-based algorithm to fit contact network models
-to phylogenetic data. We also provide a justification for the use of \gls{ABC}
-for this problem by arguing that the likelihood functions required to fit these
-models by more conventional means are likely to be computationally intractable.
-Second, in} \cref{sec:ba}, \uline{we perform a simulation study to investigate
-the \acrlong{BA} network model, which uses a preferential attachment mechanism
-to generate networks with the power law degree distributions observed in real
-world social and sexual networks. We progress through two exploratory analyses
-testing the identifiability of the model's parameters, and conclude by testing
+\add{In this chapter, we will address the three research aims
+of this thesis introduced in \cref{sec:obj}. First, in \cref{sec:netabc}, we
+describe \software{netabc}, a computer program that implements an
+\acrlong{ABC}-based algorithm to fit contact network models to phylogenetic
+data. We also provide a justification for the use of \gls{ABC} for this problem
+by arguing that the likelihood functions required to fit these models by more
+conventional means are likely to be computationally intractable.  Second, in
+\cref{sec:ba}, we perform a simulation study to investigate the \acrlong{BA}
+network model, which uses a preferential attachment mechanism to generate
+networks with the power law degree distributions observed in real world social
+and sexual networks. We progress through two exploratory analyses testing the
+identifiability of the model's parameters, and conclude by testing
 \software{netabc}'s ability to recover the parameters from simulated
-transmission trees. Third, in} \cref{sec:hiv}, \uline{we apply
-\software{netabc} to fit the \gls{BA} model to six real world \gls{HIV}
-datasets, with the understanding of the model parameters' identifiability
-gained through the simulation experiments. We conclude the chapter with a
-unified discussion of the three research aims, including interpretation of the
-results of both the simulated and real data experiments, as well as an
-examination of the limitations of our approach and opportunities for future
-investigation.}}
+transmission trees. Third, in \cref{sec:hiv}, we apply \software{netabc} to fit
+the \gls{BA} model to six real world \gls{HIV} datasets, with the understanding
+of the model parameters' identifiability gained through the simulation
+experiments. We conclude the chapter with a unified discussion of the three
+research aims, including interpretation of the results of both the simulated
+and real data experiments, as well as an examination of the limitations of our
+approach and opportunities for future investigation.}
 
@@
 
-<<treestats, include=FALSE>>=
+<<treestats, include=FALSE, eval=FALSE>>=
     params <- c("alpha", "I", "m", "N")
@@
 
-<<classifiers, include=FALSE>>=
+<<classifiers, include=FALSE, eval=FALSE>>=
     get.kernel.data <- function (param, step) {
@@
 
-<<gridsearch, include=FALSE>>=
+<<gridsearch, include=FALSE, eval=FALSE>>=
     d <- setDT(collect.data("../../simulations/aggregates/gridsearch/*"))
@@
 
-<<point_est, include=FALSE>>=
+<<point_est, include=FALSE, eval=FALSE>>=
     f <- Sys.glob("../../simulations/abc-pa-free-m/point-estimate/*")
@@
 
-<<abc_glm, include=FALSE>>=
+<<abc_glm, include=FALSE, eval=FALSE>>=
     stopifnot(alpha.glm[Parameter == "alpha",p] < 0.05)
@@
 
-<<glm_tables, echo=FALSE, warning=FALSE, error=FALSE, results="asis">>=
+<<glm_tables, echo=FALSE, warning=FALSE, error=FALSE, results="asis", eval=FALSE>>=
     source("global.R")
@@
 
-<<supp_labels, echo=FALSE, results="asis">>=
+<<supp_labels, echo=FALSE, results="asis", eval=FALSE>>=
     N <- 5000
@@
 
-<<mixed_peerdriven, include=FALSE>>=
+<<mixed_peerdriven, include=FALSE, eval=FALSE>>=
   d <- fread("bzcat ../../simulations/abc-pa-mixed-alpha/abc/*")
@@
 
-%<<abc_glm, include=FALSE>>=
-%    source("global.R")
-%    options(scipen=-1, digits=2)
-%@
-%
-%<<alpha_glm, include=FALSE>>=
-%    # alpha_error is influenced by alpha and m but not I
-%    stopifnot(alpha.glm[Parameter == "alpha", min(p) < 0.05])
-%    stopifnot(alpha.glm[Parameter == "m", min(p) < 0.05])
-%    stopifnot(alpha.glm[Parameter == "I", min(p) > 0.05])
-%
-%    # alpha_error is correlated with true_alpha
-%    alpha.test <- d[,cor.test(alpha_error, true_alpha, method="spearman")]
-%    stopifnot(alpha.test$p.value < 0.05)
-%
-%    # alpha_error is not correlated with true_m
-%    alpha.m.cor <- d[,cor.test(alpha_error, true_m)]
-%    stopifnot(alpha.m.cor$p.value < 0.05)
-%@
-%
-%We used \software{netabc} to estimate the parameters of the \gls{BA} model on
-%simulated trees where the true parameter values were known. Point estimates for
-%each parameter are shown in \cref{fig:abcptm2} for the simulations with \gls{m} =
-%2. The results for the other values of \gls{m} were similar
-%(\cref{fig:abcptm3,fig:abcptm4}). The median [IQR] absolute error of estimates
-%of \gls{alpha} across all simulations was
-%    \Sexpr{d[,median(alpha_error)]} 
-%    [\Sexpr{d[,quantile(alpha_error, 0.25)]}-\Sexpr{d[,quantile(alpha_error, 0.75)]}].
-%\Gls{GLM} analysis indicated that the true values of both \gls{alpha} and
-%\gls{m} had significant effects on the error in estimated \gls{alpha}
-%    ($p$ values $\Sexpr{pp(alpha.glm[Parameter == "alpha", min(p)], eq=FALSE)}$
-%     $\Sexpr{pp(alpha.glm[Parameter == "m", min(p)], eq=FALSE)}$,
-%     \cref{tab:glmalpha}),
-%but the true value of \gls{I} did not. There was a significant negative
-%correlation between the true value of \gls{alpha} and the error
-%    (Spearman's $\rho$ = \Sexpr{round(alpha.test$estimate, 2)},
-%     $p \Sexpr{pp(alpha.test$p.value, eq=TRUE)}$).
-%There was a negative correlation between \gls{m} and the error in \gls{alpha}
-%    (Spearman's $\rho$ = \Sexpr{round(alpha.m.cor$estimate, 2)},
-%    $p \Sexpr{pp(alpha.m.cor$p.value, eq=FALSE)}$).
-%
-%<<I_glm, include=FALSE>>=
-%    # I_error is influenced by alpha and I but not by m
-%    stopifnot(I.glm[Parameter == "alpha", min(p) < 0.05])
-%    stopifnot(I.glm[Parameter == "m", min(p) > 0.05])
-%    stopifnot(I.glm[Parameter == "I", min(p) < 0.05])
-%
-%    # correlation between I_error and true_alpha
-%    I.alpha.cor <- d[,cor.test(I_error, true_alpha, method="spearman")]
-%    stopifnot(I.alpha.cor$p.value < 0.05)
-%
-%    # correlation between I_error and true_I
-%    I.I.cor <- d[,cor.test(I_error, true_I, method="spearman")]
-%    stopifnot(I.alpha.cor$p.value < 0.05)
-%@
-%
-%The mean error in the estimated value of \gls{I} was
-%    \Sexpr{d[,round(median(I_error))]} 
-%    [\Sexpr{d[,round(quantile(I_error, 0.25))]}-\Sexpr{d[,round(quantile(I_error, 0.75))]}],
-%an over-estimate of roughly a factor of 1.5
-%(\cref{fig:abcptm2,fig:abcptm1,fig:abcptm3,fig:abcptm4}). \Gls{GLM} analysis
-%indicated a relationship between the error in estimated \gls{I}, and the true
-%values of \gls{alpha} and \gls{I}
-%    ($p$ values $\Sexpr{pp(I.glm[Parameter == "alpha", min(p)], eq=FALSE)}$
-%     $\Sexpr{pp(alpha.glm[Parameter == "I", min(p)], eq=FALSE)}$,
-%     \cref{tab:glmalpha}),
-%but not the true value of \gls{m}. There was a significant correlation between 
-%the true value of \gls{alpha} and the error in estimated \gls{I}
-%    (Spearman's $\rho$ = \Sexpr{I.alpha.cor$estimate},
-%     $p \Sexpr{pp(I.alpha.cor$p.value, eq=TRUE)}$),
-%and between the true value of \gls{I} and the error in estimated \gls{I}
-%    (Spearman's $\rho$ = \Sexpr{I.I.cor$estimate},
-%     $p \Sexpr{pp(I.I.cor$p.value, eq=TRUE)}$).
-%
-%<<m_glm, include=FALSE>>=
-%    # effects of parameters on m_error
-%    stopifnot(m.glm[Parameter == "m", min(p)] < 0.05)
-%    stopifnot(m.glm[Parameter == "alpha", min(p)] > 0.05)
-%    stopifnot(m.glm[Parameter == "I", min(p)] > 0.05)
-%
-%    # m was only estimated correctly in just over 20\% of simulations with m > 2
-%    m.tbl <- prop.table(table(d[true_m > 1,m_error]))
-%    stopifnot(m.tbl[1] > 0.2 & m.tbl[1] < 0.3)
-%
-%    # with m = 1, over 95% of simulations are right
-%    m1.est <- d[true_m == 1, m_error]
-%    stopifnot(sum(m1.est == 0) / length(m1.est) > 0.95)
-%    stopifnot(sum(m1.est == 0) / length(m1.est) < 1)
-%@
-%
-%\gls{GLM} analysis showed an effect of the true value of \gls{m} on the error
-%in the estimated \gls{m}
-%    ($p \Sexpr{pp(m.glm[Parameter == "m", min(p)])}$,
-%     \cref{tab:glmm}).
-%When the true value of \gls{m} was 1, the correct value was recovered by
-%\software{netabc} in virtually every case
-%    (\Sexpr{sum(m1.est == 0)} out of \Sexpr{length(m1.est)} simulations).
-%However, when the true value of \gls{m} was 2 or higher, the correct value was
-%recovered in only 
-%    \Sexpr{as.integer(m.tbl[1] * 100)}\%
-%of simulations, little better than random guessing. The \gls{GLM} analysis did
-%not indicate any effects of the true parameter values on the error in estimated
-%\gls{m} (\cref{tab:glmm}).
-%
-%<<N_glm, include=FALSE>>=
-%    # alpha_error is influenced by alpha and m but not I
-%    stopifnot(N.glm[Parameter == "alpha", min(p) < 0.05])
-%    stopifnot(N.glm[Parameter == "I", min(p) < 0.05])
-%    stopifnot(N.glm[Parameter == "m", min(p) > 0.05])
-%
-%    N.alpha.test <- d[,wilcox.test(.SD[true_alpha == 1.5, N_error], 
-%                                   .SD[true_alpha < 1.5, N_error])]
-%    stopifnot(N.alpha.test$p.value < 0.05)
-%    N.I.test <- d[,wilcox.test(.SD[true_I == 3000, N_error], 
-%                               .SD[true_I < 3000, N_error])]
-%    stopifnot(N.I.test$p.value < 0.05)
-%@
-%
-%Finally, the total number of nodes \gls{N} was consistently over-estimated by
-%about a factor of two
-%    (error \Sexpr{d[,format(median(N_error), scientific=FALSE)]} 
-%    [\Sexpr{d[,format(quantile(N_error, 0.25), scientific=FALSE)]} - 
-%     \Sexpr{d[,format(quantile(N_error, 0.75), scientific=FALSE)]}]).
-%The fitted \gls{GLM} indicated that the true values of both \gls{alpha} and
-%\gls{I} had an effect on the error in the estimated \gls{N}
-%    ($p$-values $\Sexpr{pp(N.glm[Parameter == "alpha", min(p)], eq=FALSE)}$ and 
-%     $\Sexpr{pp(N.glm[Parameter == "I", min(p)], eq=FALSE)}$,
-%     \cref{tab:glmN}),
-%but that the true value of \gls{m} did not. The error in the estimated \gls{N}
-%when \gls{alpha} was equal to 1.5 was slightly lower than for other values of
-%\gls{alpha}
-%    (median [IQR] error rates 
-%     \Sexpr{d[true_alpha == 1.5, format(median(N_error), scientific=FALSE)]} 
-%    [\Sexpr{d[true_alpha == 1.5, format(quantile(N_error, 0.25), scientific=FALSE)]} - 
-%     \Sexpr{d[true_alpha == 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]}]
-%     for \gls{alpha} = 1.5 vs. 
-%     \Sexpr{d[true_alpha < 1.5, format(median(N_error), scientific=FALSE)]} 
-%    [\Sexpr{d[true_alpha < 1.5, format(quantile(N_error, 0.25), scientific=FALSE)]} - 
-%     \Sexpr{d[true_alpha < 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]}]
-%     for \gls{alpha} < 1.5),
-%and this difference was statistically significant
-%    (Wilcoxon rank-sum test, $p \Sexpr{pp(N.alpha.test$p.value, eq=TRUE)}$).
-%Similarly, the error in the estimated \gls{N} when \gls{I} was equal to 3000 was
-%lower than for other values of \gls{I}
-%    (median [IQR] error rates 
-%     \Sexpr{d[true_I == 3000, format(median(N_error), scientific=FALSE)]} 
-%    [\Sexpr{d[true_I == 3000, format(quantile(N_error, 0.25), scientific=FALSE)]} - 
-%     \Sexpr{d[true_I == 3000, format(quantile(N_error, 0.75), scientific=FALSE)]}]
-%     for \gls{I} = 3000 vs. 
-%     \Sexpr{d[true_alpha < 3000, format(median(N_error), scientific=FALSE)]} 
-%    [\Sexpr{d[true_alpha < 3000, format(quantile(N_error, 0.25), scientific=FALSE)]} - 
-%     \Sexpr{d[true_alpha < 3000, format(quantile(N_error, 0.75), scientific=FALSE)]}]
-%     for \gls{I} < 3000),
-%and this difference was statistically significant
-%    (Wilcoxon rank-sum test, $p \Sexpr{pp(N.I.test$p.value, eq=TRUE)}$).
-%
-%\begin{figure}
-%  \includegraphics{abc-point-estimate-m2}
-%  \caption[\Acrlong{MAP} point estimates for \gls{BA} model parameters obtained
-%    by running \software{netabc} on simulated data, for simulations with $m = 2$.] 
-%  {
-%    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by         
-%    running \software{netabc} on simulated data. Values shown are for               
-%    simulations with \gls{m} = 2. Dashed lines indicate true values. (A)            
-%    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations  
-%    against known values. (B) Estimates of \gls{m} and \gls{N} which were held   
-%    fixed in these simulations at the values \gls{m} = 2 and \gls{N} = 5000. 
-%  }
-%  \label{fig:abcptm2}
-%\end{figure}
-%
-%<<posterior_sims, echo=FALSE>>=
-%    N <- 5000
-%    replicate <- 0
-%    lab <- NULL
-%    for (m in c(2, 3, 4)) {
-%    for (I in c(1000, 2000)) {
-%    for (alpha in c(0, 0.5, 1, 1.5)) {
-%        lab <- c(lab, sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate))
-%    }
-%    }
-%    }
-%    lab <- paste(lab, collapse=",")
-%@
-%
-%\Cref{fig:abcex} shows the \gls{ABC} approximation to the posterior
-%distribution on the \gls{BA} parameters for one simulation. Equivalent plots
-%for one replicate simulation with each combination of parameters can be found
-%in \cref{\Sexpr{lab}}. \Gls{HPD} intervals around \gls{alpha} and \gls{I} were
-%narrow relative to the region of nonzero prior density, whereas the intervals
-%for $m$ and \gls{N} were widely dispersed. \Cref{tab:abchpd} shows point
-%estimates and 95\% \gls{HPD} intervals averaged over all simulations.
-%
-%\begin{figure}
-%  \includegraphics{{abc-posterior/1.0_1000_2_5000_0}.pdf}
-%  \caption[Approximate marginal posterior distributions of BA model parameters
-%      obtained by applying \textit{netabc} to a simulated transmission tree
-%      with values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]
-%    {
-%        Approximate marginal posterior distributions of BA model parameters
-%        obtained by applying \textit{netabc} to a simulated transmission tree
-%        with BA parameter values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ =
-%        5000. Vertical dashed lines indicate true values. Shaded areas are 95\%
-%        highest posterior density intervals. $x$-axes indicate regions of
-%        nonzero prior density.
-%    }
-%  \label{fig:abcex}
-%\end{figure}
-%
-%
-%\begin{table}
-%    \centering
-%    \input{\tablepath/abc-hpd}
-%    \caption[
-%        Maximum \textit{a priori} estimates and 95\% highest posterior density
-%        (HPD) interval boundaries for \gls{BA} model parameters estimated with
-%        \software{netabc}, averaged over simulated transmission trees.
-%    ]
-%    {
-%        Maximum \textit{a priori} estimates and 95\% highest posterior density
-%        (HPD) interval boundaries for \gls{BA} model parameters estimated with
-%        \software{netabc}, averaged over simulated transmission trees.
-%    }
-%    \label{tab:abchpd}
-%\end{table}
-%
-%\subsubsection*{Effect of parameters on power-law exponent}
-%
-%\Cref{tab:glm} shows the estimated parameters for a log-link \gls{GLM} fitted
-%to the observed distribution of \gls{gamma} values. The coefficients are
-%interpretable as multiplicative effects.
-%
-%\begin{table}
-%    \centering
-%    \input{\tablepath/pa-gamma-glm}
-%    \caption{Estimated \gls{GLM} parameters for relationship between power-law
-%    exponent \gls{gamma} and \gls{BA} model parameters.}
-%    \label{tab:glm}
-%\end{table}
-
 \section{Application to real world HIV data}
@@
 
-<<realdata, include=FALSE>>=
+<<realdata, include=FALSE, eval=FALSE>>=
   m <- collect.metadata("../../simulations/aggregates/hpd/*")

### 2016-06-14 doc/thesis/methods_results.tex
Editing and proofreading intro
Writing from last night

--- 2016-06-09 doc/thesis/methods_results.tex
@@
-{\color{blue}\uline{ In this chapter, we will address the three research aims
-of this thesis introduced in} \cref{sec:obj}. \uline{First, in}
-\cref{sec:netabc}, \uline{we describe \software{netabc}, a computer program
-that implements an \acrlong{ABC}-based algorithm to fit contact network models
-to phylogenetic data. We also provide a justification for the use of \gls{ABC}
-for this problem by arguing that the likelihood functions required to fit these
-models by more conventional means are likely to be computationally intractable.
-Second, in} \cref{sec:ba}, \uline{we perform a simulation study to investigate
-the \acrlong{BA} network model, which uses a preferential attachment mechanism
-to generate networks with the power law degree distributions observed in real
-world social and sexual networks. We progress through two exploratory analyses
-testing the identifiability of the model's parameters, and conclude by testing
+\add{In this chapter, we will address the three research aims
+of this thesis introduced in \cref{sec:obj}. First, in \cref{sec:netabc}, we
+describe \software{netabc}, a computer program that implements an
+\acrlong{ABC}-based algorithm to fit contact network models to phylogenetic
+data. We also provide a justification for the use of \gls{ABC} for this problem
+by arguing that the likelihood functions required to fit these models by more
+conventional means are likely to be computationally intractable.  Second, in
+\cref{sec:ba}, we perform a simulation study to investigate the \acrlong{BA}
+network model, which uses a preferential attachment mechanism to generate
+networks with the power law degree distributions observed in real world social
+and sexual networks. We progress through two exploratory analyses testing the
+identifiability of the model's parameters, and conclude by testing
 \software{netabc}'s ability to recover the parameters from simulated
-transmission trees. Third, in} \cref{sec:hiv}, \uline{we apply
-\software{netabc} to fit the \gls{BA} model to six real world \gls{HIV}
-datasets, with the understanding of the model parameters' identifiability
-gained through the simulation experiments. We conclude the chapter with a
-unified discussion of the three research aims, including interpretation of the
-results of both the simulated and real data experiments, as well as an
-examination of the limitations of our approach and opportunities for future
-investigation.}}
+transmission trees. Third, in \cref{sec:hiv}, we apply \software{netabc} to fit
+the \gls{BA} model to six real world \gls{HIV} datasets, with the understanding
+of the model parameters' identifiability gained through the simulation
+experiments. We conclude the chapter with a unified discussion of the three
+research aims, including interpretation of the results of both the simulated
+and real data experiments, as well as an examination of the limitations of our
+approach and opportunities for future investigation.}
 
@@
 \subsubsection*{Classifiers for BA model parameters based on tree shape}
-
-Trees simulated under different values of \gls{alpha} were visibly quite
-distinct (\cref{fig:alphatrees}). In particular, higher values of \gls{alpha}
-produce networks with a small number of highly connected nodes, which, once
-infected, are likely to transmit to many other nodes. This results in a more
-unbalanced, ladder-like structure in the phylogeny, compared to networks with
-lower \gls{alpha} values. None of the other three parameters produced trees
-that were as easily distinguished from each other
-(\cref{fig:Itrees,fig:mtrees,fig:Ntrees,fig:Itrees}).  Sackin's index, which
-measures tree imbalance, was significantly correlated with all four parameters
-    (for $\alpha$, $I$, $m$, and $N$ respectively: Spearman's rho =
-     0.85,
-     \ensuremath{-0.12},
-     \ensuremath{-0.13},
-     0.09;
-     $p$-values
-     ${<}10^{-5}$,
-     $0.003$,
-     ${<}10^{-5}$,
-     ${<}10^{-5}$).
-The ratio of internal to terminal branch lengths was negatively correlated with
-\gls{alpha} and \gls{I}, and positively correlated with \gls{m} and \gls{N}
-  (Spearman's rho
-    \ensuremath{-0.84},
-    \ensuremath{-0.69},
-    0.1,
-    0.18;
-  all $p < 10^{-5}$).
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth]{kernel-alpha-tree.pdf}
-  \caption[Simulated transmission trees under three different values of BA parameter $\alpha$]{
-    Simulated transmission trees under three different values of BA parameter
-    $\alpha$. Epidemics were simulated on \gls{BA} networks of 5000 nodes, with
-    \gls{alpha} equal to 0.5, 1.0, or 1.5, until 1000 individuals were
-    infected. Transmission trees were created by sampling 500 infected nodes.
-    Higher \gls{alpha} values produced networks with a small number of
-    highly-connected nodes, resulting in highly unbalanced, ladder-like trees.
-  }
-  \label{fig:alphatrees}
-\end{figure}
-
-\Cref{fig:kpca} shows \gls{kPCA} projections of the simulated trees onto the
-first two principal components of the kernel matrix. The figure shows only the
-simulations with 500-tip trees and 1000 infected nodes. The three \gls{alpha}
-and \gls{I} values considered are well separated from each other in feature
-space. On the other hand, the three \gls{N} values overlap significantly, and
-the three \gls{m} values are virtually indistinguishable. Similar observations
-can be made for other values of \gls{I} and the number of tips
-(\cref{fig:alphakpca,fig:Nkpca,fig:Ikpca,fig:mkpca}). The values of \gls{I} and
-\gls{N} separated more clearly with larger numbers of tips, and in the case of
-\gls{N}, with larger epidemic sizes {\color{blue}(\cref{fig:Ikpca,fig:Nkpca})}.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{kernel-kpca.pdf}
-  \caption[Kernel-PCA projections of simulated trees under varying BA
-           parameter values.]{
-    Each parameter of the \gls{BA} model was individually varied to produce 300
-    simulated trees. Kernel matrices were formed from all pairwise kernel
-    scores among each set of 300 trees. The trees were projected onto the first
-    two principal components of the kernel matrix calculated using \gls{kPCA}.
-    All trees had 500 tips. The parameters not being varied were set to
-    \gls{alpha} = 1, \gls{I} = 1000, \gls{m} = 2, and \gls{N} = 5000. The tree
-    kernel meta-parameters were $\lambda = 0.3$ and $\sigma = 4$.
-  }
-  \label{fig:kpca}
-\end{figure}
-
-Accuracy of the \gls{kSVR} classifiers varied based on the parameter being
-tested (\cref{fig:rsquared}, left). Classifiers based on two other tree
-statistics, the \gls{nltt} and Sackin's index, generally exhibited worse
-performance than the tree kernel, although the magnitude of the disparity
-varied between the parameters (\cref{fig:rsquared}, centre and right). The
-results were largely robust to variations in the tree kernel meta-parameters
-$\lambda$ and $\sigma$, although accuracy varied between different epidemic and
-sampling scenarios
-(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}).
-
-When classifying $\alpha$, the \gls{kSVR} classifier had an average $R^2$ of 
-    0.92,
-compared to 
-    0.56
-for the \gls{nltt}-based SVR, and
-    0.75
-for the linear regression against Sackin's index. There was little variation
-about the mean for different tree and epidemic sizes. No classifier could
-accurately identify the $m$ parameter in any epidemic scenario, with average
-$R^2$ values of 
-  0.12 for \gls{kSVR},
-  0.01 for the \gls{nltt}, and
-  0.06
-for Sackin's index. Again, there was little variation in accuracy between
-epidemic scenarios, although the accuracy of the \gls{kSVR} was slightly higher
-on 1000-tip trees 
-    (average $R^2$ 
-     0.01,
-     0.11,
-     0.32
-     for 100, 500, and 1000 tips respectively).
-
-The accuracy of classifiers for $I$ varied significantly with the number of
-tips in the tree. For 100-tip trees, the average $R^2$ values were
-  0.7,
-  0.55, and
-  0.02
-for the tree kernel, \gls{nltt}, and Sackin's index respectively. For 500-tip
-trees, the values increased to
-  0.93,
-  0.83, and
-  0.07.
-Finally, the performance of classifiers for $N$ depended heavily on the
-epidemic scenario. The $R^2$ of the \gls{kSVR} classifier ranged from
-  0.08
-for the smallest epidemic and smallest sample size, to
-  0.82
-for the largest. Likewise, $R^2$ for the \gls{nltt}-based SVR ranged from 
-  0.01
-to
-  0.54.
-Sackin's index did not accurately classify $N$ in any scenario, with an average
-$R^2$ of
-  0.03
-and little variation between scenarios.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
-  \caption[Cross-validation accuracy of kernel-SVR, nLTT-based SVR, and
-  Sackin's index regression classifiers for BA model parameters.]{
-      Cross-validation accuracy of kernel-SVR classifier (left), SVR classifier
-      using \gls{nltt} (centre), and linear regression using Sackin's index
-      (right) for \gls{BA} model parameters. Kernel meta-parameters were set to
-      $\lambda = 0.3$ and $\sigma = 4$. Each point was calculated based on 300
-      simulated transmission trees over networks with three different values of
-      the parameter being tested, assuming perfect knowledge of the other
-      parameters. Vertical lines are empirical 95\% confidence intervals based
-      on 1000 two-fold cross-validations. {\color{blue}\uline{ The classifiers
-      for \gls{I} were not evaluated with 1000-tip trees, because one of the
-      tested \gls{I} values was 500, and it is not possible to sample a tree of
-      size 1000 from 500 infected individuals.}}
-  }
-  \label{fig:rsquared}
-\end{figure}
-
-\subsubsection*{Marginal parameter estimates with grid search}
-
-{\color{red}\sout{The accuracy of grid search estimates largely paralleled that
-of the \gls{kSVR} classifiers.}}\Cref{fig:gridest} shows point estimates and
-{\color{blue}\uline{50\%}} and 95\% highest density intervals for each of the
-\gls{BA} parameters, for one replicate experiment with 500-tip trees. Plots
-showing the point estimates for all replicates can be found in
-\cref{fig:gridptalpha,fig:gridptI,fig:gridptm,fig:gridptN}. For all parameters
-except $m$, the error of point estimates was negatively correlated with the
-number of sampled tips in the tree (for \gls{alpha}, \gls{I}, and \gls{N}
-respectively: Spearman's $\rho$ = \ensuremath{-0.22},
-\ensuremath{-0.51}, \ensuremath{-0.16}; $p$-values
-    $4\!\times\!10^{-4}$,
-    ${<}10^{-5}$,
-    $0.01$).
-The 95\% highest density intervals obtained for all parameters were extremely
-wide, occupying $>$75\% of the grid in all cases (\cref{fig:gridest}).
-
-{\color{blue}\uline{Across all replicates, $R^2$ values for the correlations
-between the estimated and true values were
-    $0.91$,
-    $0.91$,
-    $0.25$, and
-    $0.54$
-for \gls{alpha}, \gls{I}, \gls{m}, and \gls{N} respectively. The mean absolute
-errors of the point estimates were
-    $0.14$,
-    $310$,
-    $1.31$, and
-    $2419$,
-representing
-    $7$\%,
-    $8$\%,
-    $26$\%, and
-    $17$\%
-of the respective grids.}
-
-\uline{Qualitatively, each of the parameters exhibited weak identifiability
-within particular sections of the grid} (\cref{fig:gridest}). \uline{The 50\%
-\glspl{HDI} for \gls{alpha} were nearly identical for the values $\gls{alpha}
-\leq 0.5$ (on average [X-X]) and for $\gls{alpha} \geq 1.5$ ([X-X]).
-For \gls{I}, similar 50\% \gls{HPD} were observed for $\gls{I} \leq 1500$
-(average [X-X]) and for $\gls{I} > 3000$ ([X-X]). For \gls{m}, both the 50\%
-and 95\% \glspl{HDI} were generally identical for $\gls{m} > 2$ (average [X-X]
-for 50\% and [X-X] for 95\% \gls{HDI}). We did not observe a similar pattern
-for \gls{N}, however the boundaries of the \gls{HDI} did not always
-monotonically increase with the actual value.}
-
-\Cref{fig:gridalpha,fig:gridI,fig:gridm,fig:gridN} \uline{show kernel score
-distributions of kernel scores along the grid for each parameter. The
-distributions for some values, such as $\gls{alpha} = 1.25$, $\gls{I} = 500$
-and $4500$, $\gls{m} = 1$, and $\gls{N} = 1000$, exhibited distinct peaks
-around the true value. This indicates that these values produce distinctively
-shaped trees that can be identified with the tree kernel, when the other
-parameter values are fixed and known. However, for the majority of values of
-each parameter, the score distributions were fairly flat around the true value. 
-This means there is a range of values which produce similarly shaped trees, and
-the parameter is less identifiable within that range. The exception was
-\gls{I}, whose score distributions exhibited a more or less rounded shape with
-the highest point near the true value. }}
-
-{\color{red}\sout{The \gls{alpha} parameter was the most accurately estimated,
-with point estimates having an average deviation of 
-    0.14
-from the true value, on a grid from 0 to 2. The error of point estimates varied
-significantly between true values of \gls{alpha}
-    (one-way \gls{ANOVA}, $p$ ${<}10^{-5}$). In
-particular, errors were lower for the values \gls{alpha} = 1.0 and 1.25 than
-for the other values
-    (average errors 
-    0.03
-    for \gls{alpha} = 1.0 or 1.5 vs.
-    0.17
-    for \gls{alpha} $\neq$ 1.0 or 1.5),
-and this difference was significant
-    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
-}\cref{fig:gridptalpha}).
-\sout{These two values exhibited different qualitative behaviour than the other
-values in terms of the distribution of kernel scores along the grid
-}(\cref{fig:gridalpha}). \sout{In particular, there was a pronounced peak in scores
-around the true value, in contrast to the other values where the scores were
-flat around the true value. The effect was most obvious for the value
-\gls{alpha} = 1.25.}}
-
-%The average absolute error of the point estimates for \gls{I} was 
-%    d["I", round(mean(error))] individuals,
-%on a grid of 500 to 5000, and these errors differed between true values of
-%\gls{I}
-%    (one-way \gls{ANOVA}, $p pp(av["I", p], eq=TRUE)$).
-%The errors for $2000 \leq I \leq 3000$ were higher than those for the other
-%values
-%    (average errors
-%     d["I", .SD[true_value >= 2000 & true_value <= 3000, round(mean(error))]]
-%     for $2000 \leq I \leq 3000$ vs.
-%     d["I", .SD[true_value < 2000 | true_value > 3000, round(mean(error))]]
-%     for $I < 2000$ or $I > 3000$),
-%and this difference was significant
-%    (Wilcoxon rank-sum test, $p pp(I.test$p.value, eq=TRUE)$,
-%     \cref{fig:gridptI}).
-%Kernel score distributions for all test values exhibited a similar rounded
-%shape (\cref{fig:gridI}). 
-%
-%The average error for \gls{m} was
-%    d["m", round(mean(error), 2)] edges per vertex,
-%on a grid from 1 to 6. The error varied significantly between the true values
-%of \gls{m} 
-%    (one-way \gls{ANOVA}, $p pp(av["m", p], eq=TRUE)$).
-%Errors for the value \gls{m} = 1 were lower than the other values
-%    (average errors
-%    d["m", .SD[true_value == 1, round(mean(error), 2)]]
-%    for $m = 1$ vs.
-%    d["m", .SD[true_value > 1, round(mean(error), 2)]]
-%    for $m > 1$),
-%and this difference was significant
-%    (Wilcoxon rank-sum test, $p pp(m.test$p.value, eq=TRUE)$,
-%     \cref{fig:gridptm}).
-%The value $m = 1$ causes the network to take on a distinct shape relative to
-%higher \gls{m} values, namely a tree (\ie there are no cycles, see
-%\cref{subsec:treeshape}). The kernel score distribution had a peak at $m = 1$
-%when this was the true value, and a valley at $m = 1$ when the true value of
-%$m$ was greater that 1 (\cref{fig:gridm}).
-%
-%The average error for \gls{N} was 
-%    d[parameter == "N", round(mean(error))] individuals,
-%on a grid from 1000 to 15000, and was varied significantly with the true value
-%of \gls{N}
-%    (one-way \gls{ANOVA}, $p pp(av["N", p], eq=TRUE)$).
-%The errors were lower for $N \leq 3000$
-%    (average errors
-%     d["N", .SD[true_value <= 3000, round(mean(error))]]
-%     for $N \leq 3000$ vs.
-%     d["N", .SD[true_value > 3000, round(mean(error))]]
-%     for $N > 3000$),
-%and this difference was significant
-%    (Wilcoxon rank-sum test, $p pp(N.test$p.value, eq=TRUE)$,
-%     \cref{fig:gridptN}).
-%The kernel score distribution had a peak at $N = 1000$ when this was the true
-%value, and a valley there otherwise (\cref{fig:gridN}). Except for this valley,
-%the distributions were flat for $N > 3000$.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth]{gridsearch-example}
-  \caption[
-    Grid search estimates of \gls{BA} model parameters for one replicate
-    experiment with trees of size 500.
-  ]{
-    Grid search estimates of \gls{BA} model parameters for one replicate
-    experiment with trees of size 500. Point estimates (dots), 95\% highest
-    density intervals (lines), and 50\% highest density intervals (notches) for
-    each \gls{BA} model parameter, obtained using grid search. Networks and
-    transmission trees were simulated over a grid of values for each parameter
-    while holding the others fixed. For a subset of the grid values ($x$-axis),
-    test networks and trees were created and compared to each tree on the grid
-    using the tree kernel. The kernel scores along the grid were normalized to
-    resemble a probability distribution, from which the mode and highest
-    density interval were calculated. 
-  } 
-  \label{fig:gridest}
-\end{figure}
-
-\subsubsection*{Joint parameter estimates with \software{netabc}}
-
-\Cref{fig:abcptm2} shows {\color{red}\sout{\gls{MAP}}}
-{\color{blue}\uline{posterior mean}} point estimates of the BA model parameters
-obtained with kernel-assisted ABC on simulated data. The estimates shown
-correspond only to the simulations where the $m$ parameter was set to 2,
-however the results for $m = 3$ and $m = 4$ were similar
-(\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% \gls{HPD}
-intervals are given in \cref{tab:abchpd}.
-
-{\color{blue}
-\uline{Across all simulations, the median [\gls{IQR}] absolute errors of the parameter 
-estimates obtained with \software{netabc} were
-    0.11 
-    [0.03 - 
-    0.25]
-for \gls{alpha},
-    492 
-    [294 - 
-    782]
-for \gls{I},
-    1 
-    [0 - 
-    1]
-for \gls{m}, and
-    4153 
-    [3660 - 
-    4489]
-for \gls{N}. These errors comprised, respectively,
-    6\%,
-    11\%,
-    17\%, and
-    29\%
-of the regions of nonzero prior density. Average \gls{HPD} interval widths were
-    0.68,
-    2454,
-    3.01, and
-    12046,
-representing 
-    34\%,
-    55\%,
-    50\%, and
-    83\%
-of the nonzero prior density regions. Point estimates of \gls{I} were upwardly
-biased: \gls{I} was overestimated in
-    69
-out of 
-    72
-simulations
-    (96\%).
-The estimates for \gls{m} and \gls{N} were similar across all simulations
-(median [IQR] point estimates
-    3
-    [3 - 
-    3]
-and 
-    9153
-    [8660 - 
-    9489])
-regardless of the true values of any of the \gls{BA} parameters.}
-
-\uline{To analyse the effects of the true parameter values on the
-accuracy our estimates of \gls{alpha} and \gls{I}, we fitted one \glspl{GLM}
-for each of these two parameters, with error rate as the dependent variable and
-the true parameter values as independent variables. Since the estimates of
-\gls{m} and \gls{N} were roughly equal across all simulations}
-(\cref{fig:abcptm2,fig:abcptm3,fig:abcptm4}), \uline{\glspl{GLM} were not
-fitted for these parameters. The estimated coefficients are shown in
-}\cref{tab:glmalpha,tab:glmI}. \uline{For the parameters \gls{alpha}, \gls{I},
-and \gls{N}, the \gls{GLM} were fitted using the inverse link function. That
-is, if $p$ is the true value of the parameter and $\hat{p}$ is a random
-variable representing our estimate of the parameter, the \gls{GLM} posits a
-relationship of the form }
-\[
-    \E(|p-\hat{p}|) = (\beta_0 + \beta_\alpha \alpha + \beta_I I + \beta_m m)^{-1},
-\]
-\uline{where the $\beta$'s are coefficients to be fitted. If the true value
-\gls{alpha}, say, is increased by one, the \emph{inverse} of the expected
-absolute error will increase by $\beta_\alpha$. If $\beta_\alpha$ is positive,
-it means that the absolute error decreases as the true value of \gls{alpha}
-increases.}}
-
-% latex table generated in R 3.2.3 by xtable 1.8-2 package
-% Thu Jun  9 16:06:29 2016
-\begin{table}[ht]
-\centering
-\begin{tabular}{llll}
-  \hline
-Parameter & Estimate & Standard error & $p$-value \\ 
-  \hline
-(Intercept) & $2$ & $0.6$ & $0.01$ \\ 
-  $\alpha$ & $10$ & $2$ & ${<}10^{-5}$ \\ 
-  I & $-3\!\times\!10^{-4}$ & $2\!\times\!10^{-4}$ & $0.7$ \\ 
-  m & $0.5$ & $0.2$ & $0.01$ \\ 
-   \hline
-\end{tabular}
-\caption[Parameters of a fitted GLM relating error in estimated $\alpha$ to true values of BA parameters.]{Parameters of a fitted GLM relating error in estimated $\alpha$ to true values of BA parameters.
-GLM was fitted with a Gaussian distribution and inverse link function.
-Coefficients are interpretable as additive effects on the inverse of the mean error.}
-\label{tab:glmalpha}
-\end{table}
-% latex table generated in R 3.2.3 by xtable 1.8-2 package
-% Thu Jun  9 16:06:29 2016
-\begin{table}[ht]
-\centering
-\begin{tabular}{llll}
-  \hline
-Parameter & Estimate & Standard error & $p$-value \\ 
-  \hline
-(Intercept) & $0.004$ & $5\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
-  $\alpha$ & $-0.001$ & $2\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
-  I & $-4\!\times\!10^{-7}$ & $2\!\times\!10^{-7}$ & $0.05$ \\ 
-  m & $-7\!\times\!10^{-5}$ & $8\!\times\!10^{-5}$ & $1$ \\ 
-   \hline
-\end{tabular}
-\caption[Parameters of a fitted GLM relating error in estimated $I$ to true values of BA parameters.]{Parameters of a fitted GLM relating error in estimated $I$ to true values of BA parameters.
-GLM was fitted with a gaussian distribution and inverse link function.
-Coefficients are interpretable as additive effects on the inverse of the mean error.}
-\label{tab:glmI}
-\end{table}
-
-{\color{blue}\uline{The \gls{GLM} analysis indicated that the error in estimates of
-\gls{alpha} decreased with larger true values of \gls{alpha} 
-    ($p {<}10^{-5}$)
-and \gls{m}
-    ($p =0.01$)
-but was not significatly affected by \gls{I}} (\cref{tab:glmalpha}). 
-\uline{Qualitatively, \gls{alpha} seemed to be only weakly identifiable between
-the values of 0 and 0.5} (\cref{fig:abcptm2}). \uline{The error in the
-estimated prevalence \gls{I} was slightly lower for smaller values of
-\gls{alpha}
-    ($p {<}10^{-5}$)
-and \gls{I}
-    ($p =0.05$),
-but was not significantly affected by the true value of \gls{m}} 
-(\cref{tab:glmI}).}
-
-{\color{red}\sout{
-The accuracy of the parameter estimates obtained with kernel-assisted ABC
-paralleled the results from the \gls{kSVR} classifier. Of the four parameters,
-$\alpha$ was the most accurately estimated, with point estimates having a
-median [IQR] absolute error of 
-    0.11 
-    [0.03 - 
-    0.25].
-The errors when the true value of $\alpha$ was zero were significantly greater
-than those for the other values 
-    (Wilcoxon rank-sum test, $p$ = $0$).
-Errors in estimating $\alpha$ 
-%also varied with the true value of $m$ just at
-%the threshold of statistical significance
-%    (one-way ANOVA, $p 
-%    pp(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"], eq=TRUE)$),
-did not vary across the true values of $m$ or $I$ (both
-one-way ANOVA).}
-
-\sout{Estimates for $I$ were relatively accurate, with point estimate errors of
-    492 
-    [294 - 
-    782] individuals.
-These errors were significantly higher when the true value of $\alpha$ was
-at least 1
-    (Wilcoxon rank-sum test, $p$ = $0$)
-and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
-did not affect the estimates of $I$ (one-way ANOVA).}
-
-\sout{The $m$ parameter was estimated correctly in
-    37 \%
-of simulations barely better than random guessing. The true values of the other
-parameters did not significantly affect the
-estimates of $m$ (both one-way ANOVA).}
-
-\sout{Finally, the total number of nodes $N$ was consistently over-estimated by about
-a factor of two (error 4153
-[3660 - 4489] individuals). No parameters influenced the accuracy of the $N$
-estimates (all one-way ANOVA).}}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth]{abc-point-estimate-m2}
-  \vspace{6pt}
-  \caption[
-    Posterior mean point estimates for \gls{BA} model parameters obtained by
-    running \software{netabc} on simulated data, for simulations with \gls{m} =
-    2.
-  ]{
-    Posterior mean point estimates for \gls{BA} model parameters obtained by
-    running \software{netabc} on simulated data, for simulations with \gls{m} =
-    2. Dashed lines indicate true values. (A) Estimates of \gls{alpha} and
-    \gls{I} which were varied in these simulations against known values. (B)
-    Estimates of \gls{m} and \gls{N} which were held fixed in these simulations
-    at the values \gls{m} = 2 and \gls{N} = 5000.
-  }
-  \label{fig:abcptm2}
-\end{figure}
-
-\begin{table*}[ht]
-  \centering
-  \input{\tablepath/abc-hpd.tex}
-  \caption[
-      Average posterior mean point estimates and 95\% highest posterior density
-      (HPD) interval widths for BA model parameter estimates obtained with
-      kernel-assisted ABC.
-  ]{
-      Average posterior mean point estimates and 95\% highest posterior density
-      (HPD) interval widths for BA model parameter estimates obtained with
-      kernel-assisted ABC. Three transmission trees were simulated under each
-      combination of the listed parameter values, and the parameters were
-      estimated with kernel-assisted ABC without training.
-  }
-  \label{tab:abchpd}
-\end{table*}
-
-The dispersion of the ABC approximation to the posterior also varied between
-the parameters, with narrower HPD intervals for the parameters with more
-accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows
-the distributions for one simulation. Equivalent plots for one replicate
-simulation with each studied parameter combination can be found in
-\cref{fig:0.0-1000-2-5000-0,fig:0.5-1000-2-5000-0,fig:1.0-1000-2-5000-0,fig:1.5-1000-2-5000-0,fig:0.0-2000-2-5000-0,fig:0.5-2000-2-5000-0,fig:1.0-2000-2-5000-0,fig:1.5-2000-2-5000-0,fig:0.0-1000-3-5000-0,fig:0.5-1000-3-5000-0,fig:1.0-1000-3-5000-0,fig:1.5-1000-3-5000-0,fig:0.0-2000-3-5000-0,fig:0.5-2000-3-5000-0,fig:1.0-2000-3-5000-0,fig:1.5-2000-3-5000-0,fig:0.0-1000-4-5000-0,fig:0.5-1000-4-5000-0,fig:1.0-1000-4-5000-0,fig:1.5-1000-4-5000-0,fig:0.0-2000-4-5000-0,fig:0.5-2000-4-5000-0,fig:1.0-2000-4-5000-0,fig:1.5-2000-4-5000-0}.
-HPD intervals around $\alpha$ and $I$ were often narrow
-relative to the region of nonzero prior density, whereas the intervals for $m$
-and $N$ were more widely dispersed.
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{abc-posterior-example}
-  \vspace{6pt}
-  \caption[
-    Marginal posterior distributions of BA model parameters estimated
-    with kernel-assisted ABC for a single simulated transmission tree.
-  ]{
-    Marginal posterior distributions of BA model parameters estimated
-    with kernel-assisted ABC for a single simulated transmission tree. Dotted
-    lines indicate true values, solid lines indicate posterior mean point
-    estimates, and shaded areas show 95\% highest posterior density intervals.
-  }
-  \label{fig:abcex}
-\end{figure}
-
-To test the effect of model misspecification, we simulated one network where
-the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
-other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The posterior mean
-[95\% HPD] estimates for each parameter were: 
-$\alpha$, 
-  1.03 
-  [0.67 -
-   1.18];
-$I$,
-  1474 
-  [511 -
-   2990];
-$m$,
-  3 
-  [1 -
-   5];
-$N$,
-  9861 
-  [3710-
-   14977].
-The approximate posterior distributions for this simulation are shown in
-\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
-transmission tree in a peer-driven fashion, where the probability to sample a
-node was twice as high if one of its peers had already been sampled. The
-parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
-$I$ = 2000. The estimated values were
-$\alpha$, 
-  0.3 
-  [0 -
-   0.63];
-$I$,
-  2449 
-  [1417 -
-   3811];
-$m$,
-  3 
-  [2 -
-   5];
-$N$,
-  9132 
-  [2852 -
-   14780].
-The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
-of these results were in line with estimates obtained on other simulated
-datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
-$\alpha$ was somewhat lower than typical.
-
-%<<abc_glm, include=FALSE>>=
-%    source("global.R")
-%    options(scipen=-1, digits=2)
-%@
-%
-%<<alpha_glm, include=FALSE>>=
-%    # alpha_error is influenced by alpha and m but not I
-%    stopifnot(alpha.glm[Parameter == "alpha", min(p) < 0.05])
-%    stopifnot(alpha.glm[Parameter == "m", min(p) < 0.05])
-%    stopifnot(alpha.glm[Parameter == "I", min(p) > 0.05])
-%
-%    # alpha_error is correlated with true_alpha
-%    alpha.test <- d[,cor.test(alpha_error, true_alpha, method="spearman")]
-%    stopifnot(alpha.test$p.value < 0.05)
-%
-%    # alpha_error is not correlated with true_m
-%    alpha.m.cor <- d[,cor.test(alpha_error, true_m)]
-%    stopifnot(alpha.m.cor$p.value < 0.05)
-%@
-%
-%We used \software{netabc} to estimate the parameters of the \gls{BA} model on
-%simulated trees where the true parameter values were known. Point estimates for
-%each parameter are shown in \cref{fig:abcptm2} for the simulations with \gls{m} =
-%2. The results for the other values of \gls{m} were similar
-%(\cref{fig:abcptm3,fig:abcptm4}). The median [IQR] absolute error of estimates
-%of \gls{alpha} across all simulations was
-%    d[,median(alpha_error)] 
-%    [d[,quantile(alpha_error, 0.25)]-d[,quantile(alpha_error, 0.75)]].
-%\Gls{GLM} analysis indicated that the true values of both \gls{alpha} and
-%\gls{m} had significant effects on the error in estimated \gls{alpha}
-%    ($p$ values $pp(alpha.glm[Parameter == "alpha", min(p)], eq=FALSE)$
-%     $pp(alpha.glm[Parameter == "m", min(p)], eq=FALSE)$,
-%     \cref{tab:glmalpha}),
-%but the true value of \gls{I} did not. There was a significant negative
-%correlation between the true value of \gls{alpha} and the error
-%    (Spearman's $\rho$ = round(alpha.test$estimate, 2),
-%     $p pp(alpha.test$p.value, eq=TRUE)$).
-%There was a negative correlation between \gls{m} and the error in \gls{alpha}
-%    (Spearman's $\rho$ = round(alpha.m.cor$estimate, 2),
-%    $p pp(alpha.m.cor$p.value, eq=FALSE)$).
-%
-%<<I_glm, include=FALSE>>=
-%    # I_error is influenced by alpha and I but not by m
-%    stopifnot(I.glm[Parameter == "alpha", min(p) < 0.05])
-%    stopifnot(I.glm[Parameter == "m", min(p) > 0.05])
-%    stopifnot(I.glm[Parameter == "I", min(p) < 0.05])
-%
-%    # correlation between I_error and true_alpha
-%    I.alpha.cor <- d[,cor.test(I_error, true_alpha, method="spearman")]
-%    stopifnot(I.alpha.cor$p.value < 0.05)
-%
-%    # correlation between I_error and true_I
-%    I.I.cor <- d[,cor.test(I_error, true_I, method="spearman")]
-%    stopifnot(I.alpha.cor$p.value < 0.05)
-%@
-%
-%The mean error in the estimated value of \gls{I} was
-%    d[,round(median(I_error))] 
-%    [d[,round(quantile(I_error, 0.25))]-d[,round(quantile(I_error, 0.75))]],
-%an over-estimate of roughly a factor of 1.5
-%(\cref{fig:abcptm2,fig:abcptm1,fig:abcptm3,fig:abcptm4}). \Gls{GLM} analysis
-%indicated a relationship between the error in estimated \gls{I}, and the true
-%values of \gls{alpha} and \gls{I}
-%    ($p$ values $pp(I.glm[Parameter == "alpha", min(p)], eq=FALSE)$
-%     $pp(alpha.glm[Parameter == "I", min(p)], eq=FALSE)$,
-%     \cref{tab:glmalpha}),
-%but not the true value of \gls{m}. There was a significant correlation between 
-%the true value of \gls{alpha} and the error in estimated \gls{I}
-%    (Spearman's $\rho$ = I.alpha.cor$estimate,
-%     $p pp(I.alpha.cor$p.value, eq=TRUE)$),
-%and between the true value of \gls{I} and the error in estimated \gls{I}
-%    (Spearman's $\rho$ = I.I.cor$estimate,
-%     $p pp(I.I.cor$p.value, eq=TRUE)$).
-%
-%<<m_glm, include=FALSE>>=
-%    # effects of parameters on m_error
-%    stopifnot(m.glm[Parameter == "m", min(p)] < 0.05)
-%    stopifnot(m.glm[Parameter == "alpha", min(p)] > 0.05)
-%    stopifnot(m.glm[Parameter == "I", min(p)] > 0.05)
-%
-%    # m was only estimated correctly in just over 20\% of simulations with m > 2
-%    m.tbl <- prop.table(table(d[true_m > 1,m_error]))
-%    stopifnot(m.tbl[1] > 0.2 & m.tbl[1] < 0.3)
-%
-%    # with m = 1, over 95% of simulations are right
-%    m1.est <- d[true_m == 1, m_error]
-%    stopifnot(sum(m1.est == 0) / length(m1.est) > 0.95)
-%    stopifnot(sum(m1.est == 0) / length(m1.est) < 1)
-%@
-%
-%\gls{GLM} analysis showed an effect of the true value of \gls{m} on the error
-%in the estimated \gls{m}
-%    ($p pp(m.glm[Parameter == "m", min(p)])$,
-%     \cref{tab:glmm}).
-%When the true value of \gls{m} was 1, the correct value was recovered by
-%\software{netabc} in virtually every case
-%    (sum(m1.est == 0) out of length(m1.est) simulations).
-%However, when the true value of \gls{m} was 2 or higher, the correct value was
-%recovered in only 
-%    as.integer(m.tbl[1] * 100)\%
-%of simulations, little better than random guessing. The \gls{GLM} analysis did
-%not indicate any effects of the true parameter values on the error in estimated
-%\gls{m} (\cref{tab:glmm}).
-%
-%<<N_glm, include=FALSE>>=
-%    # alpha_error is influenced by alpha and m but not I
-%    stopifnot(N.glm[Parameter == "alpha", min(p) < 0.05])
-%    stopifnot(N.glm[Parameter == "I", min(p) < 0.05])
-%    stopifnot(N.glm[Parameter == "m", min(p) > 0.05])
-%
-%    N.alpha.test <- d[,wilcox.test(.SD[true_alpha == 1.5, N_error], 
-%                                   .SD[true_alpha < 1.5, N_error])]
-%    stopifnot(N.alpha.test$p.value < 0.05)
-%    N.I.test <- d[,wilcox.test(.SD[true_I == 3000, N_error], 
-%                               .SD[true_I < 3000, N_error])]
-%    stopifnot(N.I.test$p.value < 0.05)
-%@
-%
-%Finally, the total number of nodes \gls{N} was consistently over-estimated by
-%about a factor of two
-%    (error d[,format(median(N_error), scientific=FALSE)] 
-%    [d[,format(quantile(N_error, 0.25), scientific=FALSE)] - 
-%     d[,format(quantile(N_error, 0.75), scientific=FALSE)]]).
-%The fitted \gls{GLM} indicated that the true values of both \gls{alpha} and
-%\gls{I} had an effect on the error in the estimated \gls{N}
-%    ($p$-values $pp(N.glm[Parameter == "alpha", min(p)], eq=FALSE)$ and 
-%     $pp(N.glm[Parameter == "I", min(p)], eq=FALSE)$,
-%     \cref{tab:glmN}),
-%but that the true value of \gls{m} did not. The error in the estimated \gls{N}
-%when \gls{alpha} was equal to 1.5 was slightly lower than for other values of
-%\gls{alpha}
-%    (median [IQR] error rates 
-%     d[true_alpha == 1.5, format(median(N_error), scientific=FALSE)] 
-%    [d[true_alpha == 1.5, format(quantile(N_error, 0.25), scientific=FALSE)] - 
-%     d[true_alpha == 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]]
-%     for \gls{alpha} = 1.5 vs. 
-%     d[true_alpha < 1.5, format(median(N_error), scientific=FALSE)] 
-%    [d[true_alpha < 1.5, format(quantile(N_error, 0.25), scientific=FALSE)] - 
-%     d[true_alpha < 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]]
-%     for \gls{alpha} < 1.5),
-%and this difference was statistically significant
-%    (Wilcoxon rank-sum test, $p pp(N.alpha.test$p.value, eq=TRUE)$).
-%Similarly, the error in the estimated \gls{N} when \gls{I} was equal to 3000 was
-%lower than for other values of \gls{I}
-%    (median [IQR] error rates 
-%     d[true_I == 3000, format(median(N_error), scientific=FALSE)] 
-%    [d[true_I == 3000, format(quantile(N_error, 0.25), scientific=FALSE)] - 
-%     d[true_I == 3000, format(quantile(N_error, 0.75), scientific=FALSE)]]
-%     for \gls{I} = 3000 vs. 
-%     d[true_alpha < 3000, format(median(N_error), scientific=FALSE)] 
-%    [d[true_alpha < 3000, format(quantile(N_error, 0.25), scientific=FALSE)] - 
-%     d[true_alpha < 3000, format(quantile(N_error, 0.75), scientific=FALSE)]]
-%     for \gls{I} < 3000),
-%and this difference was statistically significant
-%    (Wilcoxon rank-sum test, $p pp(N.I.test$p.value, eq=TRUE)$).
-%
-%\begin{figure}
-%  \includegraphics{abc-point-estimate-m2}
-%  \caption[\Acrlong{MAP} point estimates for \gls{BA} model parameters obtained
-%    by running \software{netabc} on simulated data, for simulations with $m = 2$.] 
-%  {
-%    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by         
-%    running \software{netabc} on simulated data. Values shown are for               
-%    simulations with \gls{m} = 2. Dashed lines indicate true values. (A)            
-%    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations  
-%    against known values. (B) Estimates of \gls{m} and \gls{N} which were held   
-%    fixed in these simulations at the values \gls{m} = 2 and \gls{N} = 5000. 
-%  }
-%  \label{fig:abcptm2}
-%\end{figure}
-%
-%<<posterior_sims, echo=FALSE>>=
-%    N <- 5000
-%    replicate <- 0
-%    lab <- NULL
-%    for (m in c(2, 3, 4)) {
-%    for (I in c(1000, 2000)) {
-%    for (alpha in c(0, 0.5, 1, 1.5)) {
-%        lab <- c(lab, sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate))
-%    }
-%    }
-%    }
-%    lab <- paste(lab, collapse=",")
-%@
-%
-%\Cref{fig:abcex} shows the \gls{ABC} approximation to the posterior
-%distribution on the \gls{BA} parameters for one simulation. Equivalent plots
-%for one replicate simulation with each combination of parameters can be found
-%in \cref{lab}. \Gls{HPD} intervals around \gls{alpha} and \gls{I} were
-%narrow relative to the region of nonzero prior density, whereas the intervals
-%for $m$ and \gls{N} were widely dispersed. \Cref{tab:abchpd} shows point
-%estimates and 95\% \gls{HPD} intervals averaged over all simulations.
-%
-%\begin{figure}
-%  \includegraphics{{abc-posterior/1.0_1000_2_5000_0}.pdf}
-%  \caption[Approximate marginal posterior distributions of BA model parameters
-%      obtained by applying \textit{netabc} to a simulated transmission tree
-%      with values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]
-%    {
-%        Approximate marginal posterior distributions of BA model parameters
-%        obtained by applying \textit{netabc} to a simulated transmission tree
-%        with BA parameter values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ =
-%        5000. Vertical dashed lines indicate true values. Shaded areas are 95\%
-%        highest posterior density intervals. $x$-axes indicate regions of
-%        nonzero prior density.
-%    }
-%  \label{fig:abcex}
-%\end{figure}
-%
-%
-%\begin{table}
-%    \centering
-%    \input{\tablepath/abc-hpd}
-%    \caption[
-%        Maximum \textit{a priori} estimates and 95\% highest posterior density
-%        (HPD) interval boundaries for \gls{BA} model parameters estimated with
-%        \software{netabc}, averaged over simulated transmission trees.
-%    ]
-%    {
-%        Maximum \textit{a priori} estimates and 95\% highest posterior density
-%        (HPD) interval boundaries for \gls{BA} model parameters estimated with
-%        \software{netabc}, averaged over simulated transmission trees.
-%    }
-%    \label{tab:abchpd}
-%\end{table}
-%
-%\subsubsection*{Effect of parameters on power-law exponent}
-%
-%\Cref{tab:glm} shows the estimated parameters for a log-link \gls{GLM} fitted
-%to the observed distribution of \gls{gamma} values. The coefficients are
-%interpretable as multiplicative effects.
-%
-%\begin{table}
-%    \centering
-%    \input{\tablepath/pa-gamma-glm}
-%    \caption{Estimated \gls{GLM} parameters for relationship between power-law
-%    exponent \gls{gamma} and \gls{BA} model parameters.}
-%    \label{tab:glm}
-%\end{table}
-
-\section{Application to real world HIV data}
-\label{sec:hiv}
-
-\subsection{Methods}
-
-{\color{blue}\uline{
-The synthetic data experiments presented in the previous section demonstrated 
-that the \gls{alpha} and \gls{I} parameters, and to a lesser extent the \gls{N}
-parameter, had measurable impacts on tree shape and were identifiable with
-kernel-assisted \gls{ABC}. Therefore, our next step was to use
-\software{netabc} to estimate the parameters of the \gls{BA} model for real
-world \gls{HIV} epidemics. }} Because the \gls{BA} model assumes a single
-connected contact network, it is most appropriate to apply to groups of
-individuals who are epidemiologically related. Therefore, we searched for
-published \gls{HIV} datasets which originated from existing clusters, either
-phylogenetically or geographically defined. In addition, we analysed an
-in-house dataset sampled from \gls{HIV}-positive individuals in British
-Columbia, Canada (the ``BC data''). The datasets are summarized in
-\cref{tab:data}.
-
-\begin{table}[ht]
-  \centering
-  \input{\tablepath/realdata-bc}
-  \caption[Characteristics of published HIV datasets analyzed with \software{netabc}.]
-  {
-    Characteristics of published HIV datasets analyzed with \software{netabc}. 
-    Abbreviations: MSM, men who have sex with men; HET, heterosexual; IDU,
-    injection drug users. The \textcite{novitsky2013phylogenetic,novitsky2014impact} data
-    were sampled from a primarily heterosexual risk environment but did not
-    explicitly exclude other risk factors.
-  }
-  \label{tab:data}
-\end{table}
-
-We downloaded all sequences associated with each published study from GenBank.
-{\color{blue}\uline{For the \textcite{niculescu2015recent} data, only sequences
-from \gls{IDU} (whose sequence identifiers included the letters ``DU'') were
-included in the analysis. \textcite{kao2011surveillance} found a very strong
-association in their study population between subtype and risk group - subtype
-B was almost always associated with \gls{MSM}, whereas \gls{IDU} were usually
-infected with a circulating recombinant form. Since there were many more
-subtype B sequences in their data than sequences of other subtypes, we
-restricted our analysis to the subtype B sequences and labelled this dataset as
-\gls{MSM}. Two datasets (\textcite{zetterberg2004two} and
-\textcite{grabowski2014role}) included both \textit{env} and \textit{gag}
-sequences. Each gene was analyzed separately to assess the robustness of
-\software{netabc} to the particular \gls{HIV} gene sequence used to estimate a
-transmission tree.}}
-
-{\color{red}\sout{For the \textcite{novitsky2014impact} data,}} Each
-\textit{env} sequence was aligned pairwise to the HXB2 reference sequence
-(GenBank accession number K03455), and the hypervariable regions were clipped
-out with \software{BioPython} version 1.66+~\autocite{cock2009biopython}.
-Sequences were multiply aligned using \software{MUSCLE} version 3.8.31
-\autocite{edgar2004muscle}, and alignments were manually inspected with
-\software{Seaview} version 4.4.2 \autocite{gouy2010seaview}. Phylogenies were
-constructed from the nucleotide alignments by approximate maximum likelihood
-using \software{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
-\gls{GTR} model~\autocite{tavare1986some}. Transmission trees were estimated by
-rooting and time-scaling the phylogenies by root-to-tip regression, using a
-modified version of Path-O-Gen (distributed as part of
-BEAST~\autocite{drummond2007beast}) as described
-previously~\autocite{poon2015phylodynamic}.
-
-Three of the datasets \autocite[][and the BC data]{li2015hiv,novitsky2014impact}
-were initially much larger than the others, containing 1265, 1299, and 7923
-sequences respectively. To ensure that the analyses were comparable, we reduced
-these to a number of sequences similar to the smaller datasets. For the
-\citeauthor{li2015hiv} and BC datasets, we detected clusters of size 280 and
-399 respectively using a patristic distance cutoff of 0.02 as described
-previously~\autocite{poon2015impact}. Only sequences within these clusters were
-carried forward. For the \textcite{novitsky2014impact} data, no large clusters
-were detected using the same cutoff, so we analysed a subtree of size 180
-chosen arbitrarily.
-
-For all datasets, we used the priors $\alpha$ $\sim$ Uniform(0, 2) and $N$ and
-$I$ jointly uniform on the region \{$n \leq N \leq 10000$, $n \leq I \leq
-10000$, $I \leq N$\}, where $n$ is the number of tips in the tree (see
-\cref{tab:data}). Since the value $m = 1$ produces networks with no cycles,
-which we considered fairly implausible, we ran one analysis with the prior $m
-\sim$ DiscreteUniform(1, 5), and one with the prior $m \sim$ DiscreteUniform(2,
-5). The other parameters to the SMC algorithm were the same as used for the
-simulation experiments, except that we used 10000 particles instead of 1000 to
-increase the accuracy of the estimated posterior. This was computationally
-feasible due to the small number of runs required for this analysis.
-
-Empirical studies of contact networks often report the exponent $\gamma$ of the
-power law degree distribution. To compare our results to the literature, we
-simulated 100 networks each according to the posterior mean parameter estimates
-obtained for each investigated dataset. The power-law exponent $\gamma$ was
-calculated for each network using the \software{fit\_power\_law} function in
-\software{igraph}, with the `R.mle' implementation. The median of the 100
-$\gamma$ values was taken as a point estimate for the associated dataset.
-
-\subsection{Results}

### 2016-06-14 doc/thesis/discussion.Rnw
Intro summary; finished ABC
Started writing intro for BA methods section

@@
-<<setup, include=FALSE>>=
-    source("global.R")
-@
+%<<setup, include=FALSE>>=
+%    source("global.R")
+%@
 
@@
 network per vertex, did not have a measurable impact on tree shape and was not
-identifiable with kernel-ABC. The exception to this was the value \gls{m} = 1,
+identifiable with \gls{ABC}, \add{nor with grid search when the other
+parameters were held fixed.} The exception to this was the value \gls{m} = 1,
 which produces networks without cycles whose associated trees were more easily
-distinguished. However, all the analyses presented here did not take the
-absolute size of the transmission trees into account, as the branch lengths
-were rescaled by their mean. Because higher \gls{m} values imply higher edge
-density, an epidemic should spread more quickly for higher \gls{m} than lower
-\gls{m} with the same per-edge transmission probability.  Hence, considering
-the absolute height of the trees may improve our method's ability to
-reconstruct \gls{m}.
+distinguished (\cref{fig:gridest,fig:gridm}). However, all the analyses
+presented here did not take the absolute size of the transmission trees into
+account, as the branch lengths were rescaled by their mean. Because higher
+\gls{m} values imply higher edge density, an epidemic should spread more
+quickly for higher \gls{m} than lower \gls{m} with the same per-edge
+transmission probability.  Hence, considering the absolute height of the trees
+may improve our method's ability to reconstruct \gls{m}. \add{It was also
+pointed out to us by an anonymous reviewer that for a fixed \gls{I}, an
+infected node may only end up transmitting along a fraction of its outgoing
+edges, which could mask the presence of the extra edges associated with higher
+\gls{m}}.
 
 In addition to the tree height, many summary statistics have been developed to
-capture particular details of tree shape. Two of these, Sackin's index and the
-ratio of internal to terminal branch lengths, were correlated with every
-\gls{BA} parameter. Classifiers based on Sackin's index and the \gls{nltt}
-similarity measure performed well in some cases, though poorly in others.
-\gls{ABC} is often applied using a vector of summary
-statistics~\autocite{marin2012approximate, sunnaaker2013approximate}, rather
-than a kernel-based similarity score as we have done here. Methods have been
-developed to select an optimal combination of summary statistics for a given
-inference task~\autocite{fearnhead2012constructing}. Hence, an avenue for
+capture particular details of tree shape~\autocite{bortolussi2006aptreeshape}.
+Two of these, Sackin's index and the ratio of internal to terminal branch
+lengths, were correlated with every \gls{BA} parameter. Classifiers based on
+Sackin's index and the \gls{nltt} similarity measure performed well in some
+cases, though poorly in others. \Gls{ABC} is often applied using a vector of
+summary statistics~\autocite{marin2012approximate, sunnaaker2013approximate},
+rather than a kernel-based similarity score as we have done here. Methods have
+been developed to select an optimal combination of summary statistics for a
+given inference task~\autocite{fearnhead2012constructing}. Hence, an avenue for
 future improvement of our method may be the inclusion of additional summary
-statistics to supplement the tree kernel. In addition, all four parameters were
-more accurately classified when the number of tips in the transmission trees
-was larger, underscoring the importance of adequate sampling for accurate
-phylodynamic inference.
+statistics to supplement the tree kernel. In addition, all four \gls{BA}
+parameters were more accurately classified when the number of tips in the
+transmission trees was larger, underscoring the importance of adequate sampling
+for accurate phylodynamic inference~\autocite{novitsky2014impact}.
 
-For the more identifiable parameters, the credible intervals attained from the
-marginal \gls{ABC} target distributions were much narrower than those obtained
-through grid search, while point estimates were of comparable accuracy. This
-was likely due to the fact that \gls{SMC} employs importance sampling to
-approximate the posterior distribution, while grid search simply calculates a
-distance metric which may not have any resemblance to the posterior.
-Admittedly, our method of finding credible intervals from kernel scores along
-the grid, namely by normalizing the scores to resemble a probability
-distribution, was somewhat ad hoc, which may also have played a role.
-Regardless, this result indicates that there is benefit to applying the more
-sophisticated method, even if values for some of the parameters are known
-\textit{a priori}, and especially if credible intervals are desired on the
-parameters of interest.
+For \del{the more identifiable parameters}\add{\gls{alpha} and \gls{I}}, the
+credible intervals attained from the marginal \gls{ABC} target distributions
+were \del{much} narrower than those obtained through grid search\del{, while
+point estimates were of comparable accuracy}. This was likely due to the fact
+that \gls{SMC} employs importance sampling to approximate the posterior
+distribution, while grid search simply calculates a distance metric which may
+not have any resemblance to the posterior. Our method of finding credible
+intervals from kernel scores along the grid, by normalizing the scores to
+resemble a probability distribution, was somewhat ad hoc, which may also have
+played a role. Regardless, this result indicates that there is benefit to
+applying the more sophisticated method, even if values for some of the
+parameters are known \textit{a priori}, and especially if credible intervals
+are desired on the parameters of interest.
 
@@
 interest. We observed a non-linear relationship between the preferential
-attachment power $\alpha$ and the power law exponent $\gamma$
-(\cref{fig:gamma}). Therefore, placing a uniform prior on $\alpha$ between 0
-and 2 is equivalent to placing an informative prior that $\gamma$ is close to
-2. Therefore, if we were primarily interested in $\gamma$ rather than
-$\alpha$, a more sensible choice of prior might have a shape informed by
-\cref{fig:gamma} and be bounded above by approximately $\alpha$ = 1.5. This
-would uniformly bound $\gamma$ in the region $2 \leq \gamma \leq 4$ commonly
-reported in the network literature~\autocite{liljeros2001web,
-schneeberger2004scale, colgate1989risk, brown2011transmission}. We note however
-that \textcite{jones2003assessment} estimated $\gamma$ values greater than
-four for some datasets, in one case as high as 17, indicating that a wider
-range of permitted $\gamma$ values may be warranted.
+attachment power \gls{alpha} and the fitted power law exponent \gls{gamma}
+(\cref{fig:gamma}, although see the discussion in \cref{TODO}). Therefore,
+placing a uniform prior on \gls{alpha} between 0 and 2 is equivalent to placing an
+informative prior that \gls{gamma} is close to 2. If we were primarily
+interested in \gls{gamma} rather than \gls{alpha}, a more sensible choice of
+prior might have a shape informed by \cref{fig:gamma} and be bounded above by
+approximately $\gls{alpha} = 1.5$. This would bound \gls{gamma} in the region
+$2 \leq \gls{gamma} \leq 4$ commonly reported in the network
+literature~\autocite{liljeros2001web, schneeberger2004scale, colgate1989risk,
+brown2011transmission}. We note however that \textcite{jones2003assessment}
+estimated \gls{gamma} values greater than four for some datasets, in one case
+as high as 17, indicating that consideration of a wider range of permitted
+\gls{gamma} values may be warranted.
 
@@
 the most accurate, although the variance of the estimates was high and the
-confidence intervals were wide for \gls{alpha} $< 1$
+confidence intervals were wide for $\gls{alpha} < 1$
 (\cref{tab:abchpd,fig:abcptm2,fig:abcptm3,fig:abcptm4}). The estimates of
-\gls{N} and \gls{I} were both biased, and the estimates of \gls{m} were largely
-uninformative. Despite these issues, a major result of our our investigation is
-that some contact network parameters have a measurable impact on tree shape
-which can be used to perform statistical inference. Further refinements to
-\software{netabc}, as well as the use of more sophisticated network models, may
-improve the accuracy and precision of these estimates.
+\gls{N} and \gls{I} were both upwardly biased, and the estimates of \gls{m}
+were largely uninformative. Despite these limitations, a major result of our
+investigation is that some contact network parameters have a measurable impact
+on tree shape which can be used to perform statistical inference. Further
+refinements to \software{netabc}, as well as the use of more sophisticated
+network models, may improve the accuracy and precision of these estimates.
 
@@
 contact network structures underlying several distinct local epidemics. When
-interpreting these results, we caution that the BA model is quite simple and
-most likely misspecified for these data. In particular, the average degree of a
-node in the network is equal to $2m$, and therefore is constrained to be a
-multiple of 2. Furthermore, we considered the case $m = 1$, where the network
-has no cycles, to be implausible and therefore assigned it zero prior
-probability in one set of analyses. This forced the average degree to be at
-least four, which may be unrealistically high for sexual networks. The fact
-that the estimated values of $\alpha$ differed substantially for three datasets
-depending on whether or not $m = 1$ was allowed by the prior is further evidence
-of this potential misspecification. However, we note that for two of the
-datasets, the estimated values of $\alpha$ did not change much between priors,
-and the estimates of $I$ were robust to the choice of prior for all datasets
-studied. More sophisticated models, for example models incorporating
-heterogeneity in node behaviour, are likely to provide a better fit to these
-data.
+interpreting these results, we caution that the \gls{BA} model is quite simple
+and most likely misspecified for these data. In particular, the average degree
+of a node in the network is equal to $2\gls{m}$, and therefore is constrained
+to be a multiple of 2. Furthermore, we considered the case $\gls{m} = 1$, where
+the network has no cycles, to be implausible and therefore assigned it zero
+prior probability in one set of analyses. This forced the average degree to be
+at least four, which may be unrealistically high for sexual networks. The fact
+that the estimated values of \gls{alpha} differed substantially for several
+datasets depending on whether or not $\gls{m} = 1$ was allowed by the prior is
+further evidence of this potential misspecification. However, we note that for
+two of the datasets, the estimated values of \gls{alpha} did not change much
+between priors, and the estimates of $\gls{I}$ were robust to the choice of
+prior for all datasets studied. More sophisticated models, for example models
+incorporating heterogeneity in node behaviour, are likely to provide a better
+fit to these data.
 
-{\color{red}\sout{With respect to the preferential attachment power $\alpha$,
-the six datasets analysed fell into two categories} (\cref{fig:abchpd}).
-\sout{First, we estimated a preferential attachment power close to 1,
-indicating linear preferential attachment, for the BC data and the outbreaks
-studied by} \textcite{niculescu2015recent} and \textcite{wang2015targeting}.
-\sout{These values were robust to specifying different priors for $m$. All three
-datasets were sampled from populations in which we would expect a high degree
-of epidemiological relatedness:} \textcite{niculescu2015recent} \sout{studied a
-recent outbreak among Romanian \gls{IDU},} \citeauthor{wang2015targeting}
-\sout{sampled acutely infected MSM in Beijing, China, and the BC data
+\del{With respect to the preferential attachment power $\alpha$, the six
+datasets analysed fell into two categories (\cref{fig:abchpd}). First, we
+estimated a preferential attachment power close to 1, indicating linear
+preferential attachment, for the BC data and the outbreaks studied by
+\textcite{niculescu2015recent} and \textcite{wang2015targeting}. These
+values were robust to specifying different priors for $m$. All three datasets
+were sampled from populations in which we would expect a high degree of
+epidemiological relatedness: \textcite{niculescu2015recent} studied a
+recent outbreak among Romanian \gls{IDU}, \citeauthor{wang2015targeting}
+sampled acutely infected MSM in Beijing, China, and the BC data
 constituted a phylogenetic \gls{IDU} cluster. These are all contexts in which
@@
 
-\sout{The remaining three datasets} (\textcite{cuevas2009hiv,
-novitsky2014impact, li2015hiv}) \sout{had estimated values of $\alpha$ below
-0.5 when $m = 1$ was included in the prior, but these were not robust to
-changing the prior to exclude $m = 1$. For the} \citeauthor{cuevas2009hiv} 
-\sout{data, model misspecification is likely partially responsible. While the
-authors found that a large proportion of the samples were epidemiologically
-linked, these were mainly in small local clusters rather than the single large
-component postulated by the BA model. In addition, the mixed risk groups in the
-dataset would be unlikely to significantly interact, further weakening any
-global preferential attachment dynamics. The dataset studied by}
-\textcite{novitsky2014impact} \sout{originated from a densely sampled
-population where the predominant risk factor was believed to be heterosexual
-exposure. Although the MAP estimate of $\alpha$ was almost unchanged when the
-value $m = 1$ was excluded from the prior, the confidence interval shrank
-substantially.}}
+\del{The remaining three datasets (\textcite{cuevas2009hiv, novitsky2014impact,
+li2015hiv}) had estimated values of $\alpha$ below 0.5 when $m = 1$ was
+included in the prior, but these were not robust to changing the prior to
+exclude $m = 1$. For the \citeauthor{cuevas2009hiv} data, model
+misspecification is likely partially responsible. While the authors found that
+a large proportion of the samples were epidemiologically linked, these were
+mainly in small local clusters rather than the single large component
+postulated by the BA model. In addition, the mixed risk groups in the dataset
+would be unlikely to significantly interact, further weakening any global
+preferential attachment dynamics. The dataset studied by
+\textcite{novitsky2014impact} originated from a densely sampled population
+where the predominant risk factor was believed to be heterosexual exposure.
+Although the MAP estimate of $\alpha$ was almost unchanged when the value $m =
+1$ was excluded from the prior, the confidence interval shrank substantially.}
 
 <<disc, include=FALSE>>=
+    source("global.R")
     d <- setDT(collect.data("../../simulations/aggregates/hpd/*"))
+    d <- d[agg_gene == "gag"]
+    d1 <- d[m_min == 1]
+    d2 <- d[m_min == 2]
 @
 
-\subsubsection{Preferential attachment power}
+\end{document}
 
-For all datasets we examined, the posterior mean estimates for \gls{alpha} were
-sub-linear, ranging from 
+\subsubsection{Preferential attachment power is sub-linear and higher for \gls{IDU}}
+
+\add{For all datasets we examined, the posterior mean estimates for \gls{alpha}
+were sub-linear, ranging from 
     \Sexpr{d[m_min == 1 & parameter == "alpha", round(min(mean), 2)]}
@@
 connected in their sexual networks, as the odds of acquiring \gls{HIV} increase
-with the number of unprotected sexual contacts.
+with the number of unprotected sexual contacts.}
 
@@
 \gls{alpha} for these data was 
-    \Sexpr{d[m_min == 1 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]}
+    X%\S%expr{d[m_min == 1 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]}
         or 
-    \Sexpr{d[m_min == 2 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]},
+    X%\S%expr{d[m_min == 2 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]},
 depending on the prior on \gls{m}; \citeauthor{de2007preferential} estimated

### 2016-06-14 doc/thesis/discussion.tex
Started writing intro for BA methods section

@@
 
-\end{document}
+In addition to the tree height, many summary statistics have been developed to
+capture particular details of tree shape~\autocite{bortolussi2006aptreeshape}.
+Two of these, Sackin's index and the ratio of internal to terminal branch
+lengths, were correlated with every \gls{BA} parameter. Classifiers based on
+Sackin's index and the \gls{nltt} similarity measure performed well in some
+cases, though poorly in others. \Gls{ABC} is often applied using a vector of
+summary statistics~\autocite{marin2012approximate, sunnaaker2013approximate},
+rather than a kernel-based similarity score as we have done here. Methods have
+been developed to select an optimal combination of summary statistics for a
+given inference task~\autocite{fearnhead2012constructing}. Hence, an avenue for
+future improvement of our method may be the inclusion of additional summary
+statistics to supplement the tree kernel. In addition, all four \gls{BA}
+parameters were more accurately classified when the number of tips in the
+transmission trees was larger, underscoring the importance of adequate sampling
+for accurate phylodynamic inference~\autocite{novitsky2014impact}.
 
-In addition to the tree height, many summary statistics have been developed to
-capture particular details of tree shape. Two of these, Sackin's index and the
-ratio of internal to terminal branch lengths, were correlated with every
-\gls{BA} parameter. Classifiers based on Sackin's index and the \gls{nltt}
-similarity measure performed well in some cases, though poorly in others.
-\gls{ABC} is often applied using a vector of summary
-statistics~\autocite{marin2012approximate, sunnaaker2013approximate}, rather
-than a kernel-based similarity score as we have done here. Methods have been
-developed to select an optimal combination of summary statistics for a given
-inference task~\autocite{fearnhead2012constructing}. Hence, an avenue for
-future improvement of our method may be the inclusion of additional summary
-statistics to supplement the tree kernel. In addition, all four parameters were
-more accurately classified when the number of tips in the transmission trees
-was larger, underscoring the importance of adequate sampling for accurate
-phylodynamic inference.
-
-For the more identifiable parameters, the credible intervals attained from the
-marginal \gls{ABC} target distributions were much narrower than those obtained
-through grid search, while point estimates were of comparable accuracy. This
-was likely due to the fact that \gls{SMC} employs importance sampling to
-approximate the posterior distribution, while grid search simply calculates a
-distance metric which may not have any resemblance to the posterior.
-Admittedly, our method of finding credible intervals from kernel scores along
-the grid, namely by normalizing the scores to resemble a probability
-distribution, was somewhat ad hoc, which may also have played a role.
-Regardless, this result indicates that there is benefit to applying the more
-sophisticated method, even if values for some of the parameters are known
-\textit{a priori}, and especially if credible intervals are desired on the
-parameters of interest.
+For \del{the more identifiable parameters}\add{\gls{alpha} and \gls{I}}, the
+credible intervals attained from the marginal \gls{ABC} target distributions
+were \del{much} narrower than those obtained through grid search\del{, while
+point estimates were of comparable accuracy}. This was likely due to the fact
+that \gls{SMC} employs importance sampling to approximate the posterior
+distribution, while grid search simply calculates a distance metric which may
+not have any resemblance to the posterior. Our method of finding credible
+intervals from kernel scores along the grid, by normalizing the scores to
+resemble a probability distribution, was somewhat ad hoc, which may also have
+played a role. Regardless, this result indicates that there is benefit to
+applying the more sophisticated method, even if values for some of the
+parameters are known \textit{a priori}, and especially if credible intervals
+are desired on the parameters of interest.
 
@@
 
-\sout{The remaining three datasets} (\textcite{cuevas2009hiv,
-novitsky2014impact, li2015hiv}) \sout{had estimated values of $\alpha$ below
-0.5 when $m = 1$ was included in the prior, but these were not robust to
-changing the prior to exclude $m = 1$. For the} \citeauthor{cuevas2009hiv} 
-\sout{data, model misspecification is likely partially responsible. While the
-authors found that a large proportion of the samples were epidemiologically
-linked, these were mainly in small local clusters rather than the single large
-component postulated by the BA model. In addition, the mixed risk groups in the
-dataset would be unlikely to significantly interact, further weakening any
-global preferential attachment dynamics. The dataset studied by}
-\textcite{novitsky2014impact} \sout{originated from a densely sampled
-population where the predominant risk factor was believed to be heterosexual
-exposure. Although the MAP estimate of $\alpha$ was almost unchanged when the
-value $m = 1$ was excluded from the prior, the confidence interval shrank
-substantially.}}
+\del{The remaining three datasets (\textcite{cuevas2009hiv, novitsky2014impact,
+li2015hiv}) had estimated values of $\alpha$ below 0.5 when $m = 1$ was
+included in the prior, but these were not robust to changing the prior to
+exclude $m = 1$. For the \citeauthor{cuevas2009hiv} data, model
+misspecification is likely partially responsible. While the authors found that
+a large proportion of the samples were epidemiologically linked, these were
+mainly in small local clusters rather than the single large component
+postulated by the BA model. In addition, the mixed risk groups in the dataset
+would be unlikely to significantly interact, further weakening any global
+preferential attachment dynamics. The dataset studied by
+\textcite{novitsky2014impact} originated from a densely sampled population
+where the predominant risk factor was believed to be heterosexual exposure.
+Although the MAP estimate of $\alpha$ was almost unchanged when the value $m =
+1$ was excluded from the prior, the confidence interval shrank substantially.}
 
-%<<disc, include=FALSE>>=
-%    d <- setDT(collect.data("../../simulations/aggregates/hpd/*"))
-%@
+\end{document}
 
-\subsubsection{Preferential attachment power}
+\subsubsection{Preferential attachment power is sub-linear and higher for \gls{IDU}}
 
-For all datasets we examined, the posterior mean estimates for \gls{alpha} were
-sub-linear, ranging from 
-    X%\S%expr{d[m_min == 1 & parameter == "alpha", round(min(mean), 2)]}
+\add{For all datasets we examined, the posterior mean estimates for \gls{alpha}
+were sub-linear, ranging from 
+    0.27
 to 
-    X%\S%expr{d[m_min == 1 & parameter == "alpha", round(max(mean), 2)]}.
+    0.9.
 The sub-linearity is consistent with the results of

### 2016-06-14 doc/thesis/methods_results.Rnw
Started writing intro for BA methods section

@@
 \section{\software{Netabc}: a computer program for estimation of contact
-network parameters with kernel-assisted ABC}
+network parameters with ABC}
 \label{sec:netabc}
@@
 contact network parameters from an estimated transmission tree using
-\gls{ABC}. {\color{blue}\uline{As discussed in }\cref{sec:obj}, \uline{the principal
-statistical algorithm used by \software{netabc} is adaptive
+\gls{ABC}. \add{As discussed in \cref{sec:obj}, the principal statistical
+algorithm used by \software{netabc} is adaptive
 \gls{ABC}-\gls{SMC}~\autocite{del2012adaptive}. In addition, there are two
@@
 kernel~\autocite{poon2013mapping}, which is used as the distance function in
-\gls{ABC} to compare transmission trees~\autocite{poon2015phylodynamic}} (see
+\gls{ABC} to compare transmission trees~\autocite{poon2015phylodynamic} (see
 \cref{sec:abc}).} We give a high-level overview of the program here, before
@@
 As described in \cref{sec:smc}, \software{netabc} keeps track of a population
-of particles $x^{(k)}$, each of which contains particular parameter values
-$\theta^{(k)}$ for the {\color{blue}\uline{contact network}} model we are
-trying to fit {\color{blue}\uline{to the input tree}}. A small number of
-contact networks $z^{(k)}$ are generated under the model for each particle, in
-accordance with that particle's parameters. An epidemic is simulated over each
-of these networks using Gillespie simulation, and by keeping track of its
-progress, a transmission tree is obtained. Thus, each particle becomes
-associated with several simulated transmission trees. These trees are compared
-to the input tree using the tree kernel. Particles are weighted according to
-the similarity of their associated simulated trees with the true tree, with
-more similar trees receiving higher weights. The particles are iteratively
-perturbed to explore the parameter space, and particles with simulated trees
-too distant from the true tree are periodically dropped and resampled. Once a
-convergence criterion is attained, the final set of particles is used as a
-Monte Carlo approximation to the target distribution of \gls{ABC}, which is
-assumed to resemble the posterior distribution on model parameters (see
-\cref{sec:abc}). A graphical schematic of this algorithm is given in
+of particles $x^{(k)}$ indexed by an integere $k$, each of which contains
+particular parameter values $\theta^{(k)}$ for the \add{contact network} model
+we are trying to fit \add{to the input tree}. A small number \gls{M} of contact
+networks $z^{(k,i)}$, $1 \leq i \leq \gls{M}$, are generated under the model
+for each particle in accordance with that particle's parameters. An epidemic is
+simulated over each of these networks using Gillespie simulation, and by
+keeping track of its progress, a transmission tree is obtained. Thus, each
+particle becomes associated with several simulated transmission trees. These
+trees are compared to the input tree using the tree kernel. Particles are
+weighted according to the similarity of their associated simulated trees with
+the true tree, with more similar trees receiving higher weights. The particles
+are iteratively perturbed to explore the parameter space, and particles with
+simulated trees too distant from the true tree are periodically dropped and
+resampled. Once a convergence criterion is attained, the final set of particles
+is used as a Monte Carlo approximation to the target distribution of \gls{ABC},
+which is assumed to resemble the posterior distribution on model parameters
+(see \cref{sec:abc}). A graphical schematic of this algorithm is shown in
 \cref{fig:abcsmc}.
@@
     \includegraphics{abc-smc.pdf}
-    \caption[Graphical schematic of the ABC-SMC algorithm implemented in \software{netabc}.]{
+    \caption[
+        Graphical schematic of the ABC-SMC algorithm implemented in \software{netabc}.
+    ]{
         Graphical schematic of the \gls{ABC}-\gls{SMC} algorithm implemented in
         \software{netabc}. Particles are initially drawn from their prior
-      distributions, making the initial population a Monte Carlo approximation
-      to the prior. At each iteration, particles are perturbed, and a distance
-      threshold around the true tree contracts. Particles are rejected, and
-      eventually resampled, when all their associated simulated trees lie
-      outside the threshold. As the algorithm progresses, the population
-      smoothly approaches a Monte Carlo approximation of the \gls{ABC} target
-      distribution, which is assumed to resemble the posterior.
+        distributions, making the initial population a Monte Carlo
+        approximation to the prior. At each iteration, particles are perturbed,
+        and a distance threshold around the true tree contracts. Particles are
+        rejected, and eventually resampled, when all their associated simulated
+        trees lie outside the threshold. As the algorithm progresses, the
+        population smoothly approaches a Monte Carlo approximation of the
+        \gls{ABC} target distribution, which is assumed to resemble the
+        posterior.
     }
@@
 \gls{GSL}~\autocite{gough2009gnu} is used to generate random draws from
-probability distributions, and to perform the bisection step in the adaptive
-\gls{ABC}-\gls{SMC} algorithm. Parallelization is implemented with \gls{POSIX}
+probability distributions, and to \del{perform the bisection step} \add{solve
+for the next $\varepsilon$ by bisection} in the adaptive \gls{ABC}-\gls{SMC}
+algorithm. Parallelization is implemented with \gls{POSIX}
 threads~\autocite{barney2009posix}. In addition to the \software{netabc} binary
-to perform kernel-assisted \gls{ABC}, we provide three additional stand-alone
-utilities: \software{treekernel}, to calculate the tree kernel;
-\software{nettree}, to simulate a transmission tree over a contact network; and
-\software{treestat}, to compute various summary statistics of phylogenies. The
-programs are freely available at \url{https://github.com/rmcclosk/netabc}.
+to perform \gls{ABC}, we provide three additional stand-alone utilities:
+\software{treekernel}, to calculate the tree kernel; \software{nettree}, to
+simulate a transmission tree over a contact network; and \software{treestat},
+to compute various summary statistics of phylogenies. The programs are freely
+available at \url{https://github.com/rmcclosk/netabc}.
 
@@
 check that the algorithm would converge to a bimodal distribution, we also
-applied it to a mixture of two Gaussians with means $\pm$4 and variances 1. The
+applied it to a mixture of two Gaussians with means $\pm4$ and variances 1. The
 algorithm was able to recover both peaks (\cref{fig:smctest2}).
@@
 computationally intensive, we chose to implement our own version in
-\software{C} {\color{blue}\uline{as part of \software{netabc}}}.
+\software{C} \add{as part of \software{netabc}}.
 
@@
 that, once $u$ becomes infected, the waiting time until $u$ infects $v$ is
-distributed as $\Exponential(\beta_e)$. Note that $v$ may become infected
+distributed as $\Exponential(\gls{beta}_e)$. Note that $v$ may become infected
 before this time has elapsed, if $v$ has other incoming edges. $v$ also has a
-removal rate $\gamma_v$, so that the waiting time until removal of $v$ from the
-population is $\Exponential(\gamma_v)$. Removal may correspond to death or
-recovery with immunity, or a combination of both, but in our implementation
+removal rate $\gls{nu}_v$, so that the waiting time until removal of $v$ from
+the population is $\Exponential(\gls{nu}_v)$. Removal may correspond to death
+or recovery with immunity, or a combination of both, but in our implementation
 recovered nodes never re-enter the susceptible population. We define a
 \defn{discordant edge} as an edge $(u, v)$ where $u$ is infected and $v$ has
-never been infected.
+never been infected. \add{In the epidemiology literature, the symbol $\gamma$
+is usually used in place of \gls{nu}; we use \gls{nu} here to distinguish the
+recovery rate from the power law exponent of scale free networks (see
+\cref{subsec:pa}).}
 
@@
 $\inc(v)$ be the set of incoming edges to $v$, and $\out(v)$ be the set of
-outgoing edges from $v$. Let $I$ be the set of infected nodes in the network,
-$R$ be the set of removed nodes, and $S$ be the remaining susceptible nodes,
-and $D$ be the set of discordant edges in the network. Let $\beta$ be the total
-transmission rate over all discordant edges, and $\gamma$ be the total removal
-rate of all infected nodes,
+outgoing edges from $v$. Let \gls{gI} be the set of infected nodes in the
+network, \gls{gR} be the set of removed nodes, \gls{gS} be the set of
+susceptible nodes, and \gls{gD} be the set of discordant edges in the network.
+Let \gls{beta} be the total transmission rate over all discordant edges, and
+\gls{nu} be the total removal rate of all infected nodes,
 \[
-  \beta = \sum_{e \in D} \beta_e, \quad
-  \gamma = \sum_{v \in I} \gamma_v.
+    \gls{beta} = \sum_{e \in \gls{gD}} \gls{beta}_e, \quad
+    \gls{nu} = \sum_{v \in \gls{gI}} \gls{nu}_v.
 \]
-The variables $S$, $I$, $R$, $D$, $\beta$, and $\gamma$ are all updated as the
-simulation progresses. When a node $v$ becomes infected, it is deleted from $S$
-and added to $I$. Any formerly discordant edges in $\inc(v)$ are deleted from
-$D$, and edges in $\out(v)$ to nodes in $S$ are added to $D$. If $v$ is later
-removed, it is deleted from $I$ and added to $R$, and any discordant edges in
-$\out(v)$ are deleted from $D$. At the time of either infection or removal, the
-variables $\beta$ and $\gamma$ are updated to reflect the changes in the
-network. Since these updates are straightforward, we do not write them
-explicitly in the algorithm.
+The variables \gls{gS}, \gls{gI}, \gls{gR}, \gls{gD}, \gls{beta}, and \gls{nu}
+are all updated as the simulation progresses. When a node $v$ becomes infected,
+it is deleted from \gls{gS} and added to \gls{gI}. Any formerly discordant
+edges in $\inc(v)$ are deleted from \gls{gD}, and edges in $\out(v)$ to nodes
+in \gls{gS} are added to \gls{gD}. If $v$ is later removed, it is deleted from
+\gls{gI} and added to \gls{gR}, and any discordant edges in $\out(v)$ are
+deleted from \gls{gD}. At the time of either infection or removal, the
+variables \gls{beta} and \gls{nu} are updated to reflect the changes in the
+network. \del{The updates to \gls{gS}, \gls{gI}, \gls{gR}, \gls{gD},
+\gls{beta}, and \gls{nu} are straightforward and are not written explicitly in
+the algorithm.}
 
@@
 
-The Gillespie simulation algorithm is given as Algorithm~\ref{alg:nettree}. The
-transmission tree $T$ is simulated along with the epidemic. We keep a map
-called $\tip$, which maps infected nodes in $I$ to the tips of $T$. The
-simulation continues until either there are no discordant edges left in the
-network, or we reach a user-defined cutoff of time ($t_{\max}$) or number of
-infections ($I_{\max}$). We use the notation $\Uniform(0, 1)$ to indicate a
-number drawn from a uniform distribution on $(0, 1)$, and likewise for
-$\Exponential(\lambda)$. The combined number of internal nodes and tips in $T$
-is denoted $|T|$.
+The Gillespie simulation algorithm is given as \cref{alg:nettree}. The
+transmission tree \gls{T} is simulated along with the epidemic. We keep a map
+called ``$\tip$'', which maps infected nodes in \gls{gI} to the tips of
+\gls{T}. The simulation continues until either there are no discordant edges
+left in the network, or we reach a user-defined cutoff of time (\gls{tmax}) or
+number of infections (\gls{I}). We use the notation $\Uniform(0, 1)$ to
+indicate a number drawn from a uniform distribution on $(0, 1)$, and
+$\Exponential(\lambda)$ to indicate a number drawn from an exponential
+distribution with rate $\lambda$. The combined number of internal nodes and
+tips in \gls{T} is denoted $|T|$. \add{The updates to \gls{gS}, \gls{gI},
+\gls{gR}, \gls{gD}, \gls{beta}, and \gls{nu} described in the previous
+paragraph are not written explicitly in \cref{alg:nettree}, as they are quite
+straightforward and would only obfuscate the pseudocode.}
 
@@
     \begin{algorithmic}
-    \State infect a node $v$ at random, updating $S$, $I$, $D$, $\beta$ and $\gamma$
-    \State $T \gets$ a single node with label $1$
+        \State infect a node $v$ at random, updating \gls{gS}, \gls{gI},
+                 \gls{gD}, \gls{beta}, and \gls{gamma}
+        \State $\gls{T} \gets$ a single node with label $1$
         \State $\tip[v] \gets 1$
-    \State $t \gets 0$
-    \While{$D \neq \emptyset$ and $|I| + |R| < I_{\max}$ and $t < t_{\max}$}
-      \State $s \gets \min(t_{\max} - t, \Exponential(\beta + \gamma))$
+        \State $\gls{t} \gets 0$
+        \While{$\gls{gD} \neq \emptyset$ and $|\gls{gI}| + |\gls{gR}| < \gls{I}$
+                and $\gls{t} < \gls{tmax}$}
+            \State $s \gets \min(\gls{tmax} - \gls{t},\,
+                \Exponential(\gls{beta} + \gls{nu}))$
             \For{$v \in \tip$}
@@
             \EndFor
-      \State $t \gets t + s$
-      \If{$t < t_{\max}$}
-        \If{$\Uniform(0, \beta + \gamma) < \beta$}
-          \State choose an edge $e = (u, v)$ from $D$ with probability $\beta_e / \beta$
-                 and infect $v$
-          \State $\tip[v] \gets |T|+1$
-          \Comment{add new tips to tree and tip array}
-          \State $\tip[u] \gets |T|+2$
-          \Comment{corresponding to $u$ and $v$}
-          \State add tips with labels $(|T|+1)$ and $(|T|+2)$ to $T$
-          \State connect the new nodes to $\tip[v]$ in $T$, with branch lengths $0$
+            \State $\gls{t} \gets \gls{t} + s$
+            \If{$\gls{t} < \gls{tmax}$}
+                \If{$\Uniform(0, \gls{beta} + \gls{nu}) < \gls{beta}$}
+                    \State choose an edge $e = (u, v)$ from \gls{gD} with
+                        probability $\gls{beta}_e / \gls{beta}$ and infect $v$
+                    \State $\tip[v] \gets |\gls{T}|+1$
+                    \Comment add new tips to tree and tip array
+                    \State $\tip[u] \gets |\gls{T}|+2$
+                    \Comment corresponding to $u$ and $v$ 
+                    \State add tips with labels $(|\gls{T}|+1)$ and
+                        $(|\gls{T}|+2)$ to \gls{T}
+                    \State connect the new nodes to $\tip[v]$ in $\gls{T}$,
+                        with branch lengths $0$ 
                 \Else
-          \State choose a node $v$ from $I$ with probability $\gamma_v / \gamma$
-                 and remove $v$
+                    \State choose a node $v$ from \gls{gI} with probability
+                        $\gls{nu}_v / \gls{nu}$ and remove $v$
                     \State delete $v$ from $\tip$
                 \EndIf
-        \State update $S$, $I$, $R$, $D$, $\beta$, and $\gamma$
+                \State update \gls{gS}, \gls{gI}, \gls{gR}, \gls{gD},
+                    \gls{beta}, and \gls{nu}
             \EndIf
@@
 dot-product of the two trees' feature vectors in the space of all possible
-subset trees with branch lengths (see \cref{subsec:treeshape}). {\color{blue}
-\uline{Because the branch lengths are continuous, there are infinitely many
-possible subset trees; hence, the feature space is infinite-dimensional.}}
-The kernel was implemented using the fast algorithm developed by
+subset trees with branch lengths (see \cref{subsec:treeshape}). \add{Because
+the branch lengths are continuous, there are infinitely many possible subset
+trees; hence, the feature space is infinite-dimensional.} The kernel was
+implemented using the fast algorithm developed by
 \textcite{moschitti2006making}. First, the production rule of each node, which
@@
 computed by dynamic programming. The complexity of this implementation is
-$O(|T_1||T_2|)$ for the two trees $T_1$ and $T_2$ being compared.
+$O(|\gls{T}_1||\gls{T}_2|)$ for the two trees $\gls{T}_1$ and $\gls{T}_2$ being
+compared.
 
@@
 \[
-  \rho(T_1, T_2) = 1 - \frac{K(T_1, T_2)}{\sqrt{K(T_1, T_1) K(T_2, T_2)}},
+    \gls{rho}(\gls{T}_1, \gls{T}_2) = 1 - \frac{K(\gls{T}_1, \gls{T}_2)}{\sqrt{K(\gls{T}_1, \gls{T}_1) K(\gls{T}_2, \gls{T}_2)}},
 \]
-which is a number between 0 and 1 that is minimized when $T_1 = T_2$. This is
+which is a number between 0 and 1 that is minimized when $\gls{T}_1 = \gls{T}_2$. This is
 similar to the normalization used by \textcite{collins2002new,
@@
 \gls{ABC}-\gls{SMC} algorithm described in \cref{subsec:abcalg}, except that
-the sequence of tolerances $\varepsilon_i$ is automatically determined rather
+the sequence of tolerances $\gls{epsilon}_i$ is automatically determined rather
 than specified in advance. The tolerances are chosen such that the \gls{ESS} of
@@
 Monte Carlo approximation to the target distribution. This situation is
-referred to as the ``collapse'' of the approximation, {\color{blue}\uline{also
-known as particle degeneracy} (see \cref{subsec:smc})} and is mitigated by the
-adaptive approach. A single parameter {\color{red}\sout{$\alpha$ (not to be
-confused with the \gls{BA} model parameter)}} controls the decay rate.
-{\color{blue}\uline{In the original paper of }\textcite{del2012adaptive},
-\uline{the parameter is called $\alpha$, but to avoid confusion with the
-\gls{BA} parameter of the same name, we will refer to it here as
-\gls{alphaess}.}}
-The tolerance $\varepsilon_i$ is chosen to satisfy
+referred to as the collapse of the approximation \add{or particle degeneracy
+(see \cref{subsec:smc})} and is mitigated by the adaptive approach. A single
+parameter \del{$\alpha$ (not to be confused with the \gls{BA} model parameter)}
+controls the decay rate. \add{In the original paper of
+\textcite{del2012adaptive}, the parameter is called $\alpha$, but to avoid
+confusion with the \gls{BA} parameter of the same name we will refer to it
+here as \gls{alphaess}.} The tolerance $\gls{epsilon}_i$ is chosen to satisfy
 \[
-  \ESS(w_i) = \alpha_{\ESS} \ESS(w_{i-1}),
+    \ESS(w_i) = \gls{alphaess} \ESS(w_{i-1}),
 \]
 where, $w_i$ is the vector of weights at the $i$th step. Note that, since $w_i$
-depends on $\varepsilon_i$, this equation solves for the updated weights and
+depends on $\gls{epsilon}_i$, this equation solves for the updated weights and
 the updated tolerance simultaneously. As pointed out by
@@
 solved numerically by bisection. The forward kernels $K_i$ are taken to be
-\gls{MCMC} kernels with stationary distributions $\pi_{\varepsilon_i}$ and
+\gls{MCMC} kernels with stationary distributions $\pi_{\gls{epsilon}_i}$ and
 proposal distributions
 \[
-  q_i(\theta, \theta') \prod_{k=1}^M \Pr(z_i^{(k)'} \mid \theta'),
+    \gls{q}_i\left(\theta^{(k)}, \theta^{(k)\prime}\right) 
+    \prod_{j=1}^{\gls{M}} \lik\left(z^{(j,k)\prime} \middle| \theta^{(k)\prime} \right),
 \]
-where $\theta$ is the vector of model parameters and $z_k$ are $M$ datasets
-simulated according to $\theta'$. In our implementation, $q$ is either a
-Gaussian proposal for continuous parameters, or a Poisson proposal for discrete
-parameters. For the Poisson proposals, the number of steps to move the particle
-is drawn from a Poisson distribution, and the direction in which to move the
-particle is chosen uniformly at random. For both proposals, the variance was
-set equal to twice the empirical variance of the particles,
-following~\autocite{beaumont2009adaptive, del2012adaptive}. The backwards
-kernels are
+where $\theta^{(k)}$ is the vector of model parameters \add{associated with
+particle $x^{(k)}$} and $z^{(j,k)\prime}$, $1 \leq j \leq \gls{M}$, are \gls{M}
+datasets simulated according to $\theta^{(k)\prime}$. In our implementation,
+the $\gls{q}_i$ \add{are constructed component-wise for $\theta$ out of}
+Gaussian proposals for continuous parameters and Poisson proposals for discrete
+parameters. For the Poisson proposals, the number of \add{discrete} steps to
+move the particle is drawn from a Poisson distribution, and the direction in
+which to move the particle is chosen uniformly at random. The variance of each
+proposal distribution was set equal to twice the empirical variance of the
+particles, following~\autocite{beaumont2009adaptive, del2012adaptive}. The
+backwards kernels are
 \[
-  L_{i-1}(x', x) = \frac{\pi_n(x)K(x, x')}{\pi_n(x')}.
+    L_{i-1}(x', x) = \frac{\pi_i(x)K_i(x, x')}{\pi_i(x')}.
 \]
-When substituted into \cref{eq:smcwt}, the forward kernels $K(x, x')$ and
-densities $\pi_n(x') = \pi_{\varepsilon_n}(x')$ cancel out, and we are left
-with the weight update 
+Here we have written $x'$ for $x_i$ and $x$ for $x_{i-1}$ to emphasize that
+$x_{i-1}$ is the current value of the particle and $x_i$ is the proposed value.
+When substituted into \cref{eq:smcwt}, the forward kernels $K_i(x, x')$ and
+densities $\pi_i(x') = \pi_{\varepsilon_i}(x')$ cancel out, and we are left
+with the following weight update.
 \begin{align*}
     w_i(x) 
-    &\propto w_{i-1}(x) \frac{\pi_n(x \mid y)}{\pi_{i-1}(x \mid y)} \\
-    &= w_{i-1}(x) \frac{\pi(x) \pi_i(y \mid x)}{\pi(x) \pi_{i-1}(y \mid x)} \\
-    &= w_{i-1}(x) \frac{\sum_{k=i}^M \I_{A_{\varepsilon_i, y}}(z_k)}
-            {\sum_{k=i}^M \I_{A_{\varepsilon_{i-1}, y}}(z_k)}.
+    & \madd{\propto w_{i-1}(x) \frac{\pi_i(x') L_{i-1}(x', x)}{\pi_{i-1}(x)K_i(x, x')}}
+        & \madd{\text{importance weight update from SMC}} \\
+    &\madd{=w_{i-1}(x) \frac{\pi_i(x') \pi_i(x) K_i(x, x')}{\pi_{i-1}(x)K_i(x, x')\pi_i(x')}}
+        & \madd{\text{substitute } L_{i-1}} \\
+    &= w_{i-1}(x) \frac{\pi_i(x)}{\pi_{i-1}(x)}
+        & \madd{\text{cancel } K_i(x, x') \text{ and } \pi_i(x')} \\
+    &= \madd{w_{i-1}(x) \frac{\pi(\theta) \prod_{j=1}^M f(z^{(j)\prime} \mid \theta) 
+                        \sum_{j=i}^M \I_{A_{\varepsilon_i, y}}(z^{(j)})}
+                       {\pi(\theta) \prod_{j=1}^M f(z^{(j)\prime} \mid \theta) 
+                       \sum_{j=i}^M \I_{A_{\varepsilon_{i-1}, y}}(z^{(j)})}}
+        & \madd{\text{definition of } \pi_i(x)} \\
+    &= w_{i-1}(x) \frac{\sum_{j=i}^M \I_{A_{\varepsilon_i, y}}(z^{(j)})}
+        {\sum_{j=i}^M \I_{A_{\varepsilon_{i-1}, y}}(z^{(j)})}
+        & \madd{\text{cancel prior and likelihood.}}
 \end{align*}
-In other words, when the distance threshold $\varepsilon_{i-1}$ is contracted
-to $\varepsilon_i$, the particles' weights are multiplied by the proportion of
-simulated datasets that are still inside the new threshold. 
-The user may specify a final tolerance $\varepsilon$, or a final acceptance
-rate of the \gls{MCMC} kernel, and the algorithm will be stopped when either of
-these termination conditions is reached. The latter condition stops the
-algorithm when the particles are not moving around very much, implying little
-change in the estimated target.
+In other words, when the distance threshold $\gls{epsilon}_{i-1}$ is contracted
+to $\gls{epsilon}_i$, the particles' weights are multiplied by the proportion of
+simulated datasets that are still inside the new threshold. The user may
+specify a final tolerance $\gls{epsilon}$, or a final acceptance rate of the
+\gls{MCMC} kernel, and the algorithm will be stopped when either of these
+termination conditions is reached. The latter condition stops the algorithm
+when the particles are not moving around very much, implying little change in
+the estimated target.
 
@@
 
-{\color{blue}\uline{We present here a non-rigorous justification for the use of
+\add{We present here a non-rigorous justification for the use of
 \gls{ABC} for the problem at hand, as opposed to more frequently-used
-approaches for fitting mathematical models} (see \cref{chp:prelim}).
-\uline{Consider a contact network model with parameters $\theta$, and an
-estimated transmission tree $T$. Taking a Bayesian approach, our aim is to
-obtain a sample from the posterior distribution on the model's parameters given
-our data,}
+approaches for fitting mathematical models (see \cref{chp:prelim}). Consider a
+contact network model with parameters $\theta$, and an estimated transmission
+tree \gls{T}. Taking a Bayesian approach, our aim is to obtain a sample from
+the posterior distribution on the model's parameters given our data,}
 \[
-    \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)}.
+    \madd{\post(\theta \mid \gls{T}) = \frac{\lik(\gls{T} \mid \theta) \prior(\theta)}
+    {\int_{\Theta}\lik(\gls{T} \mid \theta) \prior(\theta) \d\theta}.}
 \]
-\uline{For all but the simplest models, the normalizing constant $\Pr(T)$ is an
-intractable integral. What we shall argue here is that, in contrast to most
-commonly studied mathematical models, the likelihood $\Pr(T \mid \theta)$ is
-also likely to be intractible in our case.}
+\add{For all but the simplest models, the normalizing constant in the
+denominator is an intractable integral. What we shall argue here is that, in
+contrast to most commonly studied mathematical models, the likelihood
+$\lik(\gls{T} \mid \theta)$ is also likely to be intractible in our case.}
 
@@
     \centering
-    \includegraphics[width=\textwidth]{transtree}
+    \includegraphics[width=0.6\textwidth]{transtree}
     \caption[
@@
 
-\uline{As discussed in} \cref{subsec:tt}, \uline{the internal nodes of
-transmission trees represent transmission events, and are labelled with the
-donor in the associated transmission pair. However, when we estimate a
-transmission tree from viral sequence data, we generally only know the labels
-of the tips of the tree, not the labels of the internal nodes. In viral
-phylogenies, the transmissions are at least partially preserved through the
-evolutionary relationships among the viruses, but the directionality of those
-transmissions is unknown. Thus, a single estimated transmission tree can
-correspond to many possible pathways of the epidemic through the network.
-}\Cref{fig:tt} \uline{illustrates this concept for a simple transmission tree
-with three tips. When calculating a likelihood given a transmission tree, we
-must sum over all possible labellings of the internal nodes. Let $\Nu$ be the
-set of such labellings. Then}
+\add{As discussed in \cref{subsec:tt}, the internal nodes of transmission trees
+represent transmission events, and are labelled with the donor in the
+associated transmission pair. However, when we estimate a transmission tree
+from viral sequence data, we generally only know the labels of the tips of the
+tree, not the labels of the internal nodes. In viral phylogenies, the
+transmissions are at least partially preserved through the evolutionary
+relationships among the viruses, but the directionality of those transmissions
+is unknown. Thus, a single estimated transmission tree can correspond to many
+possible pathways of the epidemic through the network. \Cref{fig:tt}
+illustrates this concept for a simple transmission tree with three tips. When
+calculating a likelihood given a transmission tree, we must sum over all
+possible labellings of the internal nodes. Let $\L$ be the set of such
+labellings. Then}
 \begin{align}
-    \Pr(T \mid \theta) = \sum{\nu \in \Nu} \Pr(T, \nu \mid \theta).
+    \madd{\lik(\gls{T} \mid \theta) = \sum_{l \in \L} \lik(\gls{T},\, l \mid \theta).}
     \label{eq:lik1}
 \end{align}
-\uline{A contact network model assigns a probability to each possible contact
-network. Transmission trees are realized over particular contact networks, not
-over the model itself. Therefore, we must also sum over all contact networks
-which could be generated by the model. Let $\G$ be the set of all possible
-contact networks. Summing } \cref{eq:lik1} \uline{over $\G$ gives}
+\add{A contact network model assigns a probability density to each possible
+contact network. Transmission trees are realized over particular contact
+networks, not over the model itself. Therefore, we must also sum over all
+contact networks which could be generated by the model. Let $\G$ be the set of
+all possible contact networks. Summing \cref{eq:lik1} over $\G$ gives}
 \begin{align}
-    \Pr(T \mid \theta) = \sum_{G \in \G} \sum{\nu \in \Nu} 
-        \Pr(T, \nu \mid G, \theta) \Pr(G \mid \theta).
+    \madd{\lik(\gls{T} \mid \theta) = \sum_{G \in \G} \sum_{l \in \L} 
+        \lik(\gls{T},\, l \mid G, \theta) \lik(G \mid \theta).}
     \label{eq:lik2}
 \end{align}
-\uline{This can be simplified somewhat by noticing that, given a specific
+\add{This can be simplified somewhat by noticing that, given a specific
 contact network, the labelled transmission tree depends only on that network
-and not on the model that generated it. That is, $\Pr(T, \nu \mid G, \theta) =
-\Pr(T, \nu \mid G)$, and}
+and not on the model that generated it. That is, $\lik(\gls{T}, \nu \mid G,
+\theta) = \lik(\gls{T},\, l \mid G)$, and}
 \begin{align}
-    \Pr(T \mid \theta) = \sum_{G \in \G} \Pr(G \mid \theta) \sum{\nu \in \Nu} 
-        \Pr(T, \nu \mid G)
+    \madd{\lik(\gls{T} \mid \theta) = \sum_{G \in \G} \lik(G \mid \theta) \sum_{l \in \L} 
+    \lik(\gls{T}, \nu \mid G)}
     \label{eq:lik3}
 \end{align}
-\uline{Under the assumption that both transmission and removal are Poisson
-processes, calculating $\Pr(T, \nu \mid G)$ can be accomplished by a
-straightforward modification of the Gillespie simulation algorithm} 
-(\cref{alg:nettree}). \uline{Rather than choosing transmission or removal
-events according to their probabilities, the events would be deterministically
-chosen based on the transmission tree and the probabilities of each event
-would be multiplied together. Assuming efficient data structures for storing
-lists of nodes and edges, the complexity of this calculation would be $O(|T|)$.
-The number of possible labellings of internal nodes of $T$ is easily seen to be
-$2^{(|T|-1)/2}$ by noticing that each of the $(|T|-1)/2$ internal nodes must be
-labelled with the same label as either its right child or its left child.
-Although exponential calculations of this nature can are often be simplified on 
-trees using dynamic programming} (\eg \autocite{pupko2000fast}), \uline{it
-cannot be straightforwardly applied in this case because the subtrees'
-probabilities depend on the existing epidemic progress (their parents and
-siblings). Hence, calculating the inner sum over labels may take time
-$O(2^{(|T|-1)/2})$.}
-
-\uline{The outer sum, over all contact networks, is also difficult to evaluate
-in general. There are $2^{N(N-1)}$ directed graphs on $N$
+\add{Under the assumption that both transmission and removal are Poisson
+processes, calculating $\lik(\gls{T},\, l \mid G)$ can be accomplished by a
+straightforward modification of the Gillespie simulation algorithm
+(\cref{alg:nettree}). Rather than choosing transmission or removal events
+according to their probabilities, the events would be deterministically chosen
+based on the transmission tree and the probabilities of each event would be
+multiplied together. Assuming efficient data structures for storing lists of
+nodes and edges, the complexity of this calculation would be $O(|\gls{T}|)$.
+The number of possible labellings of internal nodes of $\gls{T}$ is easily seen
+to be $2^{(|\gls{T}|-1)/2}$ by noticing that each of the $(|\gls{T}|-1)/2$
+internal nodes must be labelled with the same label as either its right child
+or its left child. Although exponential calculations of this nature can are
+often be simplified on trees using dynamic programming (\eg
+\autocite{pupko2000fast}), it cannot be straightforwardly applied in this case
+because the subtrees' probabilities depend on the existing epidemic progress
+(their parents and siblings). Hence, calculating the inner sum over labels may
+take time $O(2^{(|\gls{T}|-1)/2})$.}
+
+\add{The outer sum, over all contact networks, is also difficult to evaluate
+in general. There are $2^{\gls{N}(\gls{N}-1)}$ directed graphs on \gls{N}
 nodes~\autocite{harary2014graphical}. There must be at least as many nodes in
-the contact network as the number of tips in the tree, which is $(|T|+1)/2$. Of
-course, it is very likely that there are more nodes in the network than
-observed tips because some individuals are never infected and/or some infected
-individuals are never sampled. The complexity of calculating $\Pr(G \mid
-\theta)$ is obviously dependent on the particular model being investigated. For
-the \gls{BA} model, we might have to sum over all possible orders in which the
-nodes could be added, and all assignments of edges to the nodes which generated
-them. However, even in the unlikely case that calculating $\Pr(G \mid \theta)$
-can be done in constant time, the sum} \ref{eq:lik3} \uline{still has at least
-$O(2^{|T|^2})$ terms.}
-
-\uline{We have shown that both the normalizing constant $\Pr(T)$ and the
-likelihood $\Pr(T \mid \theta)$ are likely computationally prohibitive to
-calculate. If this is the case, the problem of fitting contact network models
-to phylogenies seems to be of the \emph{doubly-intractible}
-ype~\autocite{murray2012mcmc}, which would imply that these models are not
-amenable to neither \gls{ML} nor Bayesian inference techniques. Although both
-methods are able to cope with an intractible normalizing constant (for example,
-by local search for \gls{ML} or Bayesian \gls{MCMC}), neither can avoid the
-intractable likelihood calculations. This justifies the use of \gls{ABC}, which
-is a likelihood-free method.}
-
-\uline{We have not proven here that} \cref{eq:lik3} \uline{is impossible to
-calculate in polynomial time - it could be possible to algebraically simplify
-the sum into a tractable expression. Furthermore, under certain models, a large
-proportion of $\G$ may have zero probability, which would enable the
-simplification of the outer sum on a model-specific basis. It should also be
-noted that extensions of Bayesian \gls{MCMC} have been developed for
-doubly-intractable problems~\autocite{liang2010double, murray2012mcmc}, which
-might be adaptable to the problem at hand. These have not been as widely used
-as \gls{ABC}, nor are they as easily parallelizable as \gls{SMC}. }}
+the contact network as the number of tips in the tree, which is
+$(|\gls{T}|+1)/2$. Of course, it is very likely that there are more nodes in
+the network than observed tips because some individuals are never infected
+and/or some infected individuals are never sampled. The complexity of
+calculating $\lik(G \mid \theta)$ is obviously dependent on the particular
+model being investigated. For the \gls{BA} model, we might have to sum over all
+possible orders in which the nodes could be added, and all assignments of edges
+to the nodes which generated them. However, even in the case that calculating
+$\lik(G \mid \theta)$ can be done in constant time, the sum (\ref{eq:lik3})
+still has at least $O(2^{|\gls{T}|^2})$ terms.}
+
+\add{We have shown that both the normalizing constant $\int_\Theta \lik(\gls{T}
+\mid \theta) \prior(\theta) \d\theta$ and the likelihood $\lik(T \mid \theta)$
+are likely computationally prohibitive to calculate. If this is the case, the
+problem of fitting contact network models to phylogenies seems to be of the
+\emph{doubly-intractible} type~\autocite{murray2012mcmc}, which would imply that
+these models are not amenable to neither \gls{ML} nor Bayesian inference
+techniques. Although both methods are able to cope with an intractible
+normalizing constant (for example, by local search for \gls{ML} or Bayesian
+\gls{MCMC}), neither can avoid the intractable likelihood calculations. This
+justifies the use of \gls{ABC}, which is a likelihood-free method.}
+
+\add{We have not proven here that \cref{eq:lik3} is impossible to calculate in
+polynomial time - it could be possible to algebraically simplify the sum into a
+tractable expression. Furthermore, under certain models, a large proportion of
+$\G$ may have zero probability, which would enable the simplification of the
+outer sum on a model-specific basis. It should also be noted that extensions of
+Bayesian \gls{MCMC} have been developed for doubly-intractable
+problems~\autocite{liang2010double, murray2012mcmc}, which might be adaptable
+to the problem at hand. These have not been as widely used as \gls{ABC}, nor
+are they as easily parallelizable as \gls{SMC}.}
 
@@
 \label{sec:ba}
+\glsreset{BA}
+
+\subsection{Motivation}
+
+\add{We developed \software{netabc} with the objective of extracting useful,
+quantitative information about network structures from viral phylogenies.}
+
+\add{There are three aspects to ``usefulness''. The first is model
+specification and the biological or epidemiological interpretation of the
+parameters. We want the model to be somewhat realistic, so that \ldots, but no
+too complicated that it becomes difficult to interpret. At least some of the
+parameters should be of interest from a theoretical or practical perspective.
+As this is a phylodynamic method, intended to be used with viral sequence data, 
+we would also like to choose parameters which may be difficult to estimate with
+more standard methods; otherwise, our method provides no advantage. The
+\gls{BA} model (\cref{subsec:pa}) satisfies these criteria, albeit some better
+than others. The purported realism of the model stems from the fat-tailed
+degree distributions it produces, which are similar to those observed in real
+world sexual networks~\autocite{colgate1989risk, liljeros2001web,
+schneeberger2004scale, clemenccon2015statistical, rothenberg2007large}.
+Moreover, the ``rich get richer'' phenomenon, where popular individuals attract
+new connections at an elevated rate, is intuitively reasonable for both
+sexual~\autocite{de2007preferential} and
+\gls{IDU}~\autocite{dombrowski2013topological} networks. However, the model is 
+very simple, assuming that all nodes form the same number of links when added
+to the network and share the same preference for popular individuals.}
+
+\add{In this thesis, we consider four parameters related to the \gls{BA} model,
+denoted \gls{N}, \gls{m}, \gls{alpha}, and \gls{I} (see \cref{subsec:pa}). The
+first three of these are parameters of the model itself, while \gls{I} is
+related to the simulation of transmission trees over the network. However, we
+will refer to all four as \gls{BA} parameters. \gls{N} denotes the total number
+of nodes in the network, or equivalently, susceptible individuals in the
+population. \gls{m} is the number of new undirected edges added for each new
+vertex, or equivalently one-half of the average degree. \gls{alpha} is the
+power of preferential attachment - new nodes are attached to existing nodes of
+degree $d$ with probability proportional to $d^{\gls{alpha}} + 1$. Finally,
+\gls{I} is the number of infected individuals at the time when sampling occurs.}
+
+\add{From a public health standpoint, all four parameters are of some interest.
+The prevalence \gls{I} can be used to estimate the resources required to combat
+an ongoing epidemic, while total susceptible population size \gls{N} provides a
+similar metric for preventative measures. The average degree, controlled by
+\gls{m}, may have some value in predicting how fast an epidemic will spread.
+When there are more connections in the network, the time intervals between
+transmission events decrease, as does the number of transmissions required for
+the epidemic to travel from one part of the network to another. High average
+degrees may be reduced by interventions causing behaviour change, such as using
+condoms or caughing into one's sleeve; these types of interventions will be
+less helpful when the average number of contacts per person is already low. The
+preferential attachment power \gls{alpha} is perhaps the least obviously
+useful, but it can offer some insight into the dynamics of the network and how
+it would react to the removal of nodes. \textcite{dombrowski2013topological} 
+found evidence of preferential attachment in \gls{IDU} networks, and suggested
+that as a consequence of this characteristic, the removal of random nodes (such
+as through a police crackdown) might inadventently make it easier for epidemics
+to spread. When individuals with only one or two connections lose them, they
+might tend to seek out well-known (that is, high-degree) members of the
+community, thus increasing those individuals' connectivity even further.}
+
+\add{All four \gls{BA} parameters can be estimated using more conventional
+approaches, but the difficulty of doing so varies between the parameters and
+based on other characteristics of the epidemic under consideration. All network
+parameters can theoretically be estimated by explicitly constructing the
+contact network, although this is highly resource intensive and is hampered by 
+misreporting and other challenges~\autocite{eames2015six}. In any event, if we
+are in a position to know the true contact network then the point of estimating
+network parameters is moot. Depending on the context, \gls{N} may be estimable
+with a survey, but if the population in question is ``hidden'' (such as
+\gls{IDU} in regions where drug use is illegal) estimation of numbers may be
+more challenging. Respondent-driven
+sampling~\autocite{heckathorn1997respondent} was developed to investigate such
+populations. The prevalence \gls{I} may be estimated from reports of hospital
+admissions or engagement in care, but this may be very difficult for diseases
+such as \gls{HCV} which may remain asymptomatic for decades. Routine
+population-level testing is another possibility, but is resource intensive and
+the barriers associated with hidden populations still apply. The average degree
+of the network ($2\gls{m}$) can be ascertained by simply asking a sample of
+people how many contacts they had, but they might be unwilling to disclose this
+information or, in the case of airborne transmissions, they might not know. The
+estimation of \gls{alpha} is even more complex, as this parameter is most
+strongly reflected in the connectivity of very high degree nodes, who are rare
+in the population. Even if full degree distributions are available, there are
+models other than preferential attachment which can produce scale-free 
+networks~\autocite[\eg][]{kumar2000stochastic}. \textcite{de2007preferential}
+were able to estimate \gls{alpha} by maximum likelihood using partner count
+data from several sequential time intervals, but they admit such detailed data
+are not usually available. Moreover, their dataset was constructed via a random
+survey, which would likely miss the few high-degree nodes characterizing a
+power law degree distribution. In short, each of these parameters may be
+estimated without phylodynamics, but there are sufficient difficulties that we
+believe an alternative method based on sequence data is warranted.}
+
+\end{document}
 
 \subsection{Methods}
-\glsreset{BA}
 
@@
 
-{\color{blue}\uline{Using synthetic data,}} we investigated four parameters
-related to the \gls{BA} contact network model, denoted \gls{N}, \gls{m},
-\gls{alpha}, \gls{I} (see \cref{subsec:pa}). The first three of these are
-parameters of the model itself, while \gls{I} is related to the simulation of
-transmission trees over the network. However, we will refer to all four as
-\gls{BA} parameters. \gls{N} denotes the total number of nodes in the network,
-or equivalently, susceptible individuals in the population. When a node is
-added to the network, \gls{m} new undirected edges are added incident to it,
-and are attached to existing nodes of degree $k$ with probability proportional
-to $k^\alpha + 1$ (\cref{subsec:pa}). To simulate transmission trees over a
-\gls{BA} network, we allowed an epidemic to spread until \gls{I} nodes were
-infected, and sampled a transmission tree at that time. The \gls{alpha}
-parameter is unitless, while \gls{m} has units of edges or connections per
-vertex, and \gls{N} and \gls{I} both have units of nodes or individuals.
+\add{Using synthetic data,} we investigated four parameters related to the
+\gls{BA} contact network model, 
+denoted \gls{N}, \gls{m}, \gls{alpha}, \gls{I}
+(see \cref{subsec:pa}). The first three of these are parameters of the model
+itself, while \gls{I} is related to the simulation of transmission trees over
+the network. However, we will refer to all four as \gls{BA} parameters. \gls{N}
+denotes the total number of nodes in the network, or equivalently, susceptible
+individuals in the population. When a node is added to the network, \gls{m} new
+undirected edges are added incident to it, and are attached to existing nodes
+of degree $d$ with probability proportional to $d^{\gls{alpha}} + 1$
+(\cref{subsec:pa}). To simulate transmission trees over a \gls{BA} network, we
+allowed an epidemic to spread until \gls{I} nodes were infected, and sampled a
+transmission tree at that time. The \gls{alpha} parameter is unitless, while
+\gls{m} has units of edges or connections per vertex, and \gls{N} and \gls{I}
+both have units of nodes or individuals.
+
+\end{document}
 
@@
     experiments. Trees were simulated under the test values, and
-    kernel-assisted \gls{ABC} was used to re-estimate posterior distributions for the
+    \gls{ABC} was used to re-estimate posterior distributions for the
     \gls{BA} parameters without training.
@@
 
-<<treestats, include=FALSE, eval=FALSE>>=
+<<treestats, include=FALSE>>=
     params <- c("alpha", "I", "m", "N")
@@
 
-<<classifiers, include=FALSE, eval=FALSE>>=
+<<classifiers, include=FALSE>>=
     get.kernel.data <- function (param, step) {
@@
 
-<<gridsearch, include=FALSE, eval=FALSE>>=
+<<gridsearch, include=FALSE>>=
     d <- setDT(collect.data("../../simulations/aggregates/gridsearch/*"))
@@
 
-<<point_est, include=FALSE, eval=FALSE>>=
+<<point_est, include=FALSE>>=
     f <- Sys.glob("../../simulations/abc-pa-free-m/point-estimate/*")
@@
 {\color{blue}\uline{posterior mean}} point estimates of the BA model parameters
-obtained with kernel-assisted ABC on simulated data. The estimates shown
+obtained with ABC on simulated data. The estimates shown
 correspond only to the simulations where the $m$ parameter was set to 2,
@@
 
-<<abc_glm, include=FALSE, eval=FALSE>>=
+<<abc_glm, include=FALSE>>=
     stopifnot(alpha.glm[Parameter == "alpha",p] < 0.05)
@@
 
-<<glm_tables, echo=FALSE, warning=FALSE, error=FALSE, results="asis", eval=FALSE>>=
+<<glm_tables, echo=FALSE, warning=FALSE, error=FALSE, results="asis">>=
     source("global.R")
@@
 {\color{red}\sout{
-The accuracy of the parameter estimates obtained with kernel-assisted ABC
+The accuracy of the parameter estimates obtained with ABC
 paralleled the results from the \gls{kSVR} classifier. Of the four parameters,
@@
       (HPD) interval widths for BA model parameter estimates obtained with
-      kernel-assisted ABC.
+      ABC.
   ]{
@@
       (HPD) interval widths for BA model parameter estimates obtained with
-      kernel-assisted ABC. Three transmission trees were simulated under each
+      ABC. Three transmission trees were simulated under each
       combination of the listed parameter values, and the parameters were
-      estimated with kernel-assisted ABC without training.
+      estimated with ABC without training.
   }
@@
 
-<<supp_labels, echo=FALSE, results="asis", eval=FALSE>>=
+<<supp_labels, echo=FALSE, results="asis">>=
     N <- 5000
@@
     Marginal posterior distributions of BA model parameters estimated
-    with kernel-assisted ABC for a single simulated transmission tree.
+    with ABC for a single simulated transmission tree.
   ]{
     Marginal posterior distributions of BA model parameters estimated
-    with kernel-assisted ABC for a single simulated transmission tree. Dotted
+    with ABC for a single simulated transmission tree. Dotted
     lines indicate true values, solid lines indicate posterior mean point
@@
 
-<<mixed_peerdriven, include=FALSE, eval=FALSE>>=
+<<mixed_peerdriven, include=FALSE>>=
   d <- fread("bzcat ../../simulations/abc-pa-mixed-alpha/abc/*")
@@
 parameter, had measurable impacts on tree shape and were identifiable with
-kernel-assisted \gls{ABC}. Therefore, our next step was to use
+\gls{ABC}. Therefore, our next step was to use
 \software{netabc} to estimate the parameters of the \gls{BA} model for real
@@
 
-<<realdata, include=FALSE, eval=FALSE>>=
+<<realdata, include=FALSE>>=
   m <- collect.metadata("../../simulations/aggregates/hpd/*")
-  d1 <- fread(rownames(m)[m$m_min == 1 & m$gene == "gag"])
-  d2 <- fread(rownames(m)[m$m_min == 2 & m$gene == "gag"])
-
-  d1.env <- fread(rownames(m)[m$m_min == 1 & m$gene == "env"])
-  d2.env <- fread(rownames(m)[m$m_min == 2 & m$gene == "env"])
+  d1 <- fread(rownames(m)[m$m_min == 1 & m$agg_gene == "gag"])
+  d2 <- fread(rownames(m)[m$m_min == 2 & m$agg_gene == "gag"])
+
+  d1.env <- fread(rownames(m)[m$m_min == 1 & m$agg_gene == "env"])
+  d2.env <- fread(rownames(m)[m$m_min == 2 & m$agg_gene == "env"])
 

### 2016-06-14 doc/thesis/methods_results.tex
Started writing intro for BA methods section

--- 2016-06-11 doc/thesis/methods_results.tex
@@
-{\color{blue}\uline{ In this chapter, we will address the three research aims
-of this thesis introduced in} \cref{sec:obj}. \uline{First, in}
-\cref{sec:netabc}, \uline{we describe \software{netabc}, a computer program
-that implements an \acrlong{ABC}-based algorithm to fit contact network models
-to phylogenetic data. We also provide a justification for the use of \gls{ABC}
-for this problem by arguing that the likelihood functions required to fit these
-models by more conventional means are likely to be computationally intractable.
-Second, in} \cref{sec:ba}, \uline{we perform a simulation study to investigate
-the \acrlong{BA} network model, which uses a preferential attachment mechanism
-to generate networks with the power law degree distributions observed in real
-world social and sexual networks. We progress through two exploratory analyses
-testing the identifiability of the model's parameters, and conclude by testing
+\add{In this chapter, we will address the three research aims
+of this thesis introduced in \cref{sec:obj}. First, in \cref{sec:netabc}, we
+describe \software{netabc}, a computer program that implements an
+\acrlong{ABC}-based algorithm to fit contact network models to phylogenetic
+data. We also provide a justification for the use of \gls{ABC} for this problem
+by arguing that the likelihood functions required to fit these models by more
+conventional means are likely to be computationally intractable.  Second, in
+\cref{sec:ba}, we perform a simulation study to investigate the \acrlong{BA}
+network model, which uses a preferential attachment mechanism to generate
+networks with the power law degree distributions observed in real world social
+and sexual networks. We progress through two exploratory analyses testing the
+identifiability of the model's parameters, and conclude by testing
 \software{netabc}'s ability to recover the parameters from simulated
-transmission trees. Third, in} \cref{sec:hiv}, \uline{we apply
-\software{netabc} to fit the \gls{BA} model to six real world \gls{HIV}
-datasets, with the understanding of the model parameters' identifiability
-gained through the simulation experiments. We conclude the chapter with a
-unified discussion of the three research aims, including interpretation of the
-results of both the simulated and real data experiments, as well as an
-examination of the limitations of our approach and opportunities for future
-investigation.}}
+transmission trees. Third, in \cref{sec:hiv}, we apply \software{netabc} to fit
+the \gls{BA} model to six real world \gls{HIV} datasets, with the understanding
+of the model parameters' identifiability gained through the simulation
+experiments. We conclude the chapter with a unified discussion of the three
+research aims, including interpretation of the results of both the simulated
+and real data experiments, as well as an examination of the limitations of our
+approach and opportunities for future investigation.}
 
 \section{\software{Netabc}: a computer program for estimation of contact
-network parameters with kernel-assisted ABC}
+network parameters with ABC}
 \label{sec:netabc}
@@
 % latex table generated in R 3.2.3 by xtable 1.8-2 package
-% Fri Jun 10 18:17:33 2016
+% Tue Jun 14 16:56:16 2016
 \begin{table}[ht]
@@
 
-%<<abc_glm, include=FALSE>>=
-%    source("global.R")
-%    options(scipen=-1, digits=2)
-%@
-%
-%<<alpha_glm, include=FALSE>>=
-%    # alpha_error is influenced by alpha and m but not I
-%    stopifnot(alpha.glm[Parameter == "alpha", min(p) < 0.05])
-%    stopifnot(alpha.glm[Parameter == "m", min(p) < 0.05])
-%    stopifnot(alpha.glm[Parameter == "I", min(p) > 0.05])
-%
-%    # alpha_error is correlated with true_alpha
-%    alpha.test <- d[,cor.test(alpha_error, true_alpha, method="spearman")]
-%    stopifnot(alpha.test$p.value < 0.05)
-%
-%    # alpha_error is not correlated with true_m
-%    alpha.m.cor <- d[,cor.test(alpha_error, true_m)]
-%    stopifnot(alpha.m.cor$p.value < 0.05)
-%@
-%
-%We used \software{netabc} to estimate the parameters of the \gls{BA} model on
-%simulated trees where the true parameter values were known. Point estimates for
-%each parameter are shown in \cref{fig:abcptm2} for the simulations with \gls{m} =
-%2. The results for the other values of \gls{m} were similar
-%(\cref{fig:abcptm3,fig:abcptm4}). The median [IQR] absolute error of estimates
-%of \gls{alpha} across all simulations was
-%    d[,median(alpha_error)] 
-%    [d[,quantile(alpha_error, 0.25)]-d[,quantile(alpha_error, 0.75)]].
-%\Gls{GLM} analysis indicated that the true values of both \gls{alpha} and
-%\gls{m} had significant effects on the error in estimated \gls{alpha}
-%    ($p$ values $pp(alpha.glm[Parameter == "alpha", min(p)], eq=FALSE)$
-%     $pp(alpha.glm[Parameter == "m", min(p)], eq=FALSE)$,
-%     \cref{tab:glmalpha}),
-%but the true value of \gls{I} did not. There was a significant negative
-%correlation between the true value of \gls{alpha} and the error
-%    (Spearman's $\rho$ = round(alpha.test$estimate, 2),
-%     $p pp(alpha.test$p.value, eq=TRUE)$).
-%There was a negative correlation between \gls{m} and the error in \gls{alpha}
-%    (Spearman's $\rho$ = round(alpha.m.cor$estimate, 2),
-%    $p pp(alpha.m.cor$p.value, eq=FALSE)$).
-%
-%<<I_glm, include=FALSE>>=
-%    # I_error is influenced by alpha and I but not by m
-%    stopifnot(I.glm[Parameter == "alpha", min(p) < 0.05])
-%    stopifnot(I.glm[Parameter == "m", min(p) > 0.05])
-%    stopifnot(I.glm[Parameter == "I", min(p) < 0.05])
-%
-%    # correlation between I_error and true_alpha
-%    I.alpha.cor <- d[,cor.test(I_error, true_alpha, method="spearman")]
-%    stopifnot(I.alpha.cor$p.value < 0.05)
-%
-%    # correlation between I_error and true_I
-%    I.I.cor <- d[,cor.test(I_error, true_I, method="spearman")]
-%    stopifnot(I.alpha.cor$p.value < 0.05)
-%@
-%
-%The mean error in the estimated value of \gls{I} was
-%    d[,round(median(I_error))] 
-%    [d[,round(quantile(I_error, 0.25))]-d[,round(quantile(I_error, 0.75))]],
-%an over-estimate of roughly a factor of 1.5
-%(\cref{fig:abcptm2,fig:abcptm1,fig:abcptm3,fig:abcptm4}). \Gls{GLM} analysis
-%indicated a relationship between the error in estimated \gls{I}, and the true
-%values of \gls{alpha} and \gls{I}
-%    ($p$ values $pp(I.glm[Parameter == "alpha", min(p)], eq=FALSE)$
-%     $pp(alpha.glm[Parameter == "I", min(p)], eq=FALSE)$,
-%     \cref{tab:glmalpha}),
-%but not the true value of \gls{m}. There was a significant correlation between 
-%the true value of \gls{alpha} and the error in estimated \gls{I}
-%    (Spearman's $\rho$ = I.alpha.cor$estimate,
-%     $p pp(I.alpha.cor$p.value, eq=TRUE)$),
-%and between the true value of \gls{I} and the error in estimated \gls{I}
-%    (Spearman's $\rho$ = I.I.cor$estimate,
-%     $p pp(I.I.cor$p.value, eq=TRUE)$).
-%
-%<<m_glm, include=FALSE>>=
-%    # effects of parameters on m_error
-%    stopifnot(m.glm[Parameter == "m", min(p)] < 0.05)
-%    stopifnot(m.glm[Parameter == "alpha", min(p)] > 0.05)
-%    stopifnot(m.glm[Parameter == "I", min(p)] > 0.05)
-%
-%    # m was only estimated correctly in just over 20\% of simulations with m > 2
-%    m.tbl <- prop.table(table(d[true_m > 1,m_error]))
-%    stopifnot(m.tbl[1] > 0.2 & m.tbl[1] < 0.3)
-%
-%    # with m = 1, over 95% of simulations are right
-%    m1.est <- d[true_m == 1, m_error]
-%    stopifnot(sum(m1.est == 0) / length(m1.est) > 0.95)
-%    stopifnot(sum(m1.est == 0) / length(m1.est) < 1)
-%@
-%
-%\gls{GLM} analysis showed an effect of the true value of \gls{m} on the error
-%in the estimated \gls{m}
-%    ($p pp(m.glm[Parameter == "m", min(p)])$,
-%     \cref{tab:glmm}).
-%When the true value of \gls{m} was 1, the correct value was recovered by
-%\software{netabc} in virtually every case
-%    (sum(m1.est == 0) out of length(m1.est) simulations).
-%However, when the true value of \gls{m} was 2 or higher, the correct value was
-%recovered in only 
-%    as.integer(m.tbl[1] * 100)\%
-%of simulations, little better than random guessing. The \gls{GLM} analysis did
-%not indicate any effects of the true parameter values on the error in estimated
-%\gls{m} (\cref{tab:glmm}).
-%
-%<<N_glm, include=FALSE>>=
-%    # alpha_error is influenced by alpha and m but not I
-%    stopifnot(N.glm[Parameter == "alpha", min(p) < 0.05])
-%    stopifnot(N.glm[Parameter == "I", min(p) < 0.05])
-%    stopifnot(N.glm[Parameter == "m", min(p) > 0.05])
-%
-%    N.alpha.test <- d[,wilcox.test(.SD[true_alpha == 1.5, N_error], 
-%                                   .SD[true_alpha < 1.5, N_error])]
-%    stopifnot(N.alpha.test$p.value < 0.05)
-%    N.I.test <- d[,wilcox.test(.SD[true_I == 3000, N_error], 
-%                               .SD[true_I < 3000, N_error])]
-%    stopifnot(N.I.test$p.value < 0.05)
-%@
-%
-%Finally, the total number of nodes \gls{N} was consistently over-estimated by
-%about a factor of two
-%    (error d[,format(median(N_error), scientific=FALSE)] 
-%    [d[,format(quantile(N_error, 0.25), scientific=FALSE)] - 
-%     d[,format(quantile(N_error, 0.75), scientific=FALSE)]]).
-%The fitted \gls{GLM} indicated that the true values of both \gls{alpha} and
-%\gls{I} had an effect on the error in the estimated \gls{N}
-%    ($p$-values $pp(N.glm[Parameter == "alpha", min(p)], eq=FALSE)$ and 
-%     $pp(N.glm[Parameter == "I", min(p)], eq=FALSE)$,
-%     \cref{tab:glmN}),
-%but that the true value of \gls{m} did not. The error in the estimated \gls{N}
-%when \gls{alpha} was equal to 1.5 was slightly lower than for other values of
-%\gls{alpha}
-%    (median [IQR] error rates 
-%     d[true_alpha == 1.5, format(median(N_error), scientific=FALSE)] 
-%    [d[true_alpha == 1.5, format(quantile(N_error, 0.25), scientific=FALSE)] - 
-%     d[true_alpha == 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]]
-%     for \gls{alpha} = 1.5 vs. 
-%     d[true_alpha < 1.5, format(median(N_error), scientific=FALSE)] 
-%    [d[true_alpha < 1.5, format(quantile(N_error, 0.25), scientific=FALSE)] - 
-%     d[true_alpha < 1.5, format(quantile(N_error, 0.75), scientific=FALSE)]]
-%     for \gls{alpha} < 1.5),
-%and this difference was statistically significant
-%    (Wilcoxon rank-sum test, $p pp(N.alpha.test$p.value, eq=TRUE)$).
-%Similarly, the error in the estimated \gls{N} when \gls{I} was equal to 3000 was
-%lower than for other values of \gls{I}
-%    (median [IQR] error rates 
-%     d[true_I == 3000, format(median(N_error), scientific=FALSE)] 
-%    [d[true_I == 3000, format(quantile(N_error, 0.25), scientific=FALSE)] - 
-%     d[true_I == 3000, format(quantile(N_error, 0.75), scientific=FALSE)]]
-%     for \gls{I} = 3000 vs. 
-%     d[true_alpha < 3000, format(median(N_error), scientific=FALSE)] 
-%    [d[true_alpha < 3000, format(quantile(N_error, 0.25), scientific=FALSE)] - 
-%     d[true_alpha < 3000, format(quantile(N_error, 0.75), scientific=FALSE)]]
-%     for \gls{I} < 3000),
-%and this difference was statistically significant
-%    (Wilcoxon rank-sum test, $p pp(N.I.test$p.value, eq=TRUE)$).
-%
-%\begin{figure}
-%  \includegraphics{abc-point-estimate-m2}
-%  \caption[\Acrlong{MAP} point estimates for \gls{BA} model parameters obtained
-%    by running \software{netabc} on simulated data, for simulations with $m = 2$.] 
-%  {
-%    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by         
-%    running \software{netabc} on simulated data. Values shown are for               
-%    simulations with \gls{m} = 2. Dashed lines indicate true values. (A)            
-%    Estimates of \gls{alpha} and \gls{I} which were varied in these simulations  
-%    against known values. (B) Estimates of \gls{m} and \gls{N} which were held   
-%    fixed in these simulations at the values \gls{m} = 2 and \gls{N} = 5000. 
-%  }
-%  \label{fig:abcptm2}
-%\end{figure}
-%
-%<<posterior_sims, echo=FALSE>>=
-%    N <- 5000
-%    replicate <- 0
-%    lab <- NULL
-%    for (m in c(2, 3, 4)) {
-%    for (I in c(1000, 2000)) {
-%    for (alpha in c(0, 0.5, 1, 1.5)) {
-%        lab <- c(lab, sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate))
-%    }
-%    }
-%    }
-%    lab <- paste(lab, collapse=",")
-%@
-%
-%\Cref{fig:abcex} shows the \gls{ABC} approximation to the posterior
-%distribution on the \gls{BA} parameters for one simulation. Equivalent plots
-%for one replicate simulation with each combination of parameters can be found
-%in \cref{lab}. \Gls{HPD} intervals around \gls{alpha} and \gls{I} were
-%narrow relative to the region of nonzero prior density, whereas the intervals
-%for $m$ and \gls{N} were widely dispersed. \Cref{tab:abchpd} shows point
-%estimates and 95\% \gls{HPD} intervals averaged over all simulations.
-%
-%\begin{figure}
-%  \includegraphics{{abc-posterior/1.0_1000_2_5000_0}.pdf}
-%  \caption[Approximate marginal posterior distributions of BA model parameters
-%      obtained by applying \textit{netabc} to a simulated transmission tree
-%      with values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]
-%    {
-%        Approximate marginal posterior distributions of BA model parameters
-%        obtained by applying \textit{netabc} to a simulated transmission tree
-%        with BA parameter values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ =
-%        5000. Vertical dashed lines indicate true values. Shaded areas are 95\%
-%        highest posterior density intervals. $x$-axes indicate regions of
-%        nonzero prior density.
-%    }
-%  \label{fig:abcex}
-%\end{figure}
-%
-%
-%\begin{table}
-%    \centering
-%    \input{\tablepath/abc-hpd}
-%    \caption[
-%        Maximum \textit{a priori} estimates and 95\% highest posterior density
-%        (HPD) interval boundaries for \gls{BA} model parameters estimated with
-%        \software{netabc}, averaged over simulated transmission trees.
-%    ]
-%    {
-%        Maximum \textit{a priori} estimates and 95\% highest posterior density
-%        (HPD) interval boundaries for \gls{BA} model parameters estimated with
-%        \software{netabc}, averaged over simulated transmission trees.
-%    }
-%    \label{tab:abchpd}
-%\end{table}
-%
-%\subsubsection*{Effect of parameters on power-law exponent}
-%
-%\Cref{tab:glm} shows the estimated parameters for a log-link \gls{GLM} fitted
-%to the observed distribution of \gls{gamma} values. The coefficients are
-%interpretable as multiplicative effects.
-%
-%\begin{table}
-%    \centering
-%    \input{\tablepath/pa-gamma-glm}
-%    \caption{Estimated \gls{GLM} parameters for relationship between power-law
-%    exponent \gls{gamma} and \gls{BA} model parameters.}
-%    \label{tab:glm}
-%\end{table}
-
 \section{Application to real world HIV data}

### 2016-06-14 doc/thesis/preamble.tex
Writing from last night
Started writing intro for BA methods section

@@
 \usepackage{calc}
+\usepackage{twoopt}
 
@@
 \let\mautocite\autocite
+\let\mciteauthor\citeauthor
 \renewcommand{\cref}[1]{\mbox{\mref{#1}}}
 \renewcommand{\Cref}[1]{\mbox{\Mref{#1}}}
-\renewcommand{\textcite}[1]{\mbox{\mtextcite{#1}}}
-\renewcommand{\autocite}[1]{\mbox{\mautocite{#1}}}
+\renewcommandtwoopt{\textcite}[3][][]{\mbox{\mtextcite[#1][#2]{#3}}}
+\renewcommandtwoopt{\autocite}[3][][]{\mbox{\mautocite[#1][#2]{#3}}}
+\renewcommand{\citeauthor}[1]{\mbox{\mciteauthor{#1}}}
 \newcommand{\add}[1]{\color{blue}\uline{#1} \color{black}}
@@
 
+\DeclareMathAlphabet{\mathcal}{OMS}{cmsy}{m}{n}
 \newcommand{\N}{\mathcal{N}}
@@
 \DeclareMathOperator{\Var}{Var}
-\DeclareMathOperator{\inc}{in}
+\DeclareMathOperator{\inc}{inc}
 \DeclareMathOperator{\out}{out}
@@
 \newacronym{PA}{PA}{preferential attachment}
+\newacronym{HCV}{HCV}{hepatitis C virus}
 
@@
   description=preferential attachment power parameter in Barab\'asi-Albert networks
+}
+\newglossaryentry{beta}{
+  name=\ensuremath{\beta},
+  description=transmission rate in susceptible-infected and susceptible-infected-removed epidemiological models
+}
+\newglossaryentry{nu}{
+  name=\ensuremath{\nu},
+  description=recovery rate in the susceptible-infected-removed epidemiological model
 }
@@
 }
+\newglossaryentry{gI}{
+  name=\ensuremath{\mathcal{I}},
+  description=set of infected nodes in Gillespie simulation
+}
+\newglossaryentry{gR}{
+  name=\ensuremath{\mathcal{R}},
+  description=set of recovered nodes in Gillespie simulation
+}
+\newglossaryentry{gS}{
+  name=\ensuremath{\mathcal{S}},
+  description=set of recovered nodes in Gillespie simulation
+}
+\newglossaryentry{gD}{
+  name=\ensuremath{\mathcal{D}},
+  description=set of discordant edges in Gillespie simulation
+}
 \newglossaryentry{gamma}{
-  name=$\gamma$,
+  name=\ensuremath{\gamma},
   description=exponent of power-law degree distribution in scale-free networks
@@
 }
+\newglossaryentry{T}{
+  name=\ensuremath{T},
+  description=a transmission tree
+}
+\newglossaryentry{tmax}{
+  name=\ensuremath{t_{\max}},
+  description=user-defined cutoff time at which to stop Gillespie simulation
+}
+\newglossaryentry{t}{
+  name=\ensuremath{t},
+  description=time since initial infection of index case in an epidemic
+}
+\newglossaryentry{rho}{
+  name=\ensuremath{\rho},
+  description=distance function for approximate Bayesian computation
+}
+\newglossaryentry{epsilon}{
+  name=\ensuremath{\varepsilon},
+  description=distance function for approximate Bayesian computation
+}
+\newglossaryentry{q}{
+  name=\ensuremath{q},
+  description=proposal function for Metropolis-Hastings kernel
+}
 

### 2016-06-15 doc/thesis/methods_results.Rnw
So much exposition I hate it

@@
 
+\add{Regardless of how useful or interesting the \gls{BA} model parameters are, 
+there is another aspect to the ``usefulness'' of our method - identifiability. 
+Roughly speaking, the identifiability of a parameter says how much information
+about that parameter can possibly be obtained from the observed data (see
+\cref{chp:prelim}). If the parameters of the \gls{BA} model do not influence
+tree shape at all, then the posterior distribution will exactly resemble the
+prior. In that case, no matter how accurate a representation of the posterior
+we are able to produce, no useful information can be learned about the network.
+Hence, before proceeding with a full validation of \software{netabc} on
+simulated data, we undertook two experiments designed to assess the
+identifiability of each parameter separately.}
+
+\add{First, we simulated trees under three different values of each parameter,
+and asked how well we could tell the different trees apart. The better we are
+able to distinguish the trees, the more identifiability we might expect for the
+corresponding parameter when we attempt to estimate it with \gls{ABC}. In this
+experiment, we fix all but one of the parameters at a time, so it is possible
+that identifiability will be lost when all parameters are free to vary. This
+experiment also has the secondary purpose of validating our choice of the tree
+kernel as a distance measure in \gls{ABC}. To tell the trees apart, we use a
+classifier based on the tree kernel, but we also test two other tree shape
+statistics: Sackin's index, which considers only the topology, and the
+\gls{nltt}, which considers only branching times. Since the tree kernel
+incorporates both of these sources of information, we expect it to outperform 
+the other two statistics. Finally, the tree kernel can be ``tuned'' by
+adjusting the values of the meta-parameters \gls{lambda} and \gls{sigma}. We
+used this experiment to choose values for these meta-parameters which were best
+able to distinguish the different trees. }
+
+\add{A second experiment was designed to test whether we could actually
+estimate the parameters numerically, rather than just telling three different
+values apart, and also to assess how identifiability varied in different
+regions of the parameter space. An individual tree was compared to simulated
+trees on a one-dimensional grid of values of one \gls{BA} parameter, to obtain
+a ``distribution'' of kernel score values. From this distribution, estimates
+and credible intervals of the parameter in question could be calculated. We
+still held all but one parameter fixed, as the curse of dimensionality made it
+infeasible to perform grid search on the full parameter space. Repeating this
+experiment with trees located throughout the parameter space allowed us to
+beter quantify the identifiability. Furthermore, doing marginal estimation
+(that is, estimating one parameter with all others fixed) can provide insight
+into any biases observed when doing joint estimation with \gls{ABC}. If grid
+search is inaccurate, it indicates a lack of parameter identifiability.
+However, if the marginal grid search estimates are accurate but the estimates
+obtained with \gls{ABC} are biased, this points to confounding between the
+parameters which could only be observed when they are all estimated jointly.}
+
+\add{After these preliminary experiments, the strategy we used for testing
+\software{netabc} was a standard simulation-based validation. Transmission
+trees were simulated under several combinations of parameter values; we tried
+to recover these values by running \software{netabc} on the simulated trees.}
+
+\add{We have argued that the \gls{BA} parameters are worth estimating, and
+described three simulation-based experiments to assess our ability to do so. 
+There is a final aspect of the usefulness of these estimates: the accuracy of
+the \gls{ABC} approximation to the posterior.}
+
+... There are two factors which can affect the accuracy of the \gls{ABC}
+approximation: the choice of distance function, and the settings used for
+\gls{SMC} such as the number of particles and \gls{ESS} decay rate
+\gls{alphaess}. Fortunately, these are both easy to test. If the \gls{ABC}
+estimates are wrong, we can check if the distance function is responsible by
+simulating several trees under both the estimated and true values. ...
+For the \gls{SMC} parameters, we can simply rerun one simulation with better
+settings (more particles, more samples per particle, and higher
+\gls{alphaess}), and see if the approximation improves.
+
 \end{document}

### 2016-06-16 doc/thesis/abstract.Rnw
Editing, midway through chp 2

@@
 was observed. \del{Point estimates} \add{Posterior means} for the preferential
-attachment power \add{were all sub-linear} and ranged from \del{0.06 to 1.05}
-\add{ 
-    \Sexpr{d[parameter == "alpha", round(min(mean), 2)]}
-to
-    \Sexpr{d[parameter == "alpha", round(max(mean), 2)]}.
-Point estimates of the strength of preferential attachment were higher in
-injection drug user populations, potentially indicating that high-degree
-``hub'' nodes may play a role in epidemics among this risk group.} Our results
-underscore the importance of considering contact structures when
+attachment power \add{were all sub-linear, consistent with literature results.
+We found that the strength of preferential attachment was higher in injection
+drug user populations, potentially indicating that high-degree
+``superspreader'' nodes may play a role in epidemics among this risk group.}
+Our results underscore the importance of considering contact structures when
 \del{performing phylodynamic inference}\add{investigating viral outbreaks.}

### 2016-06-16 doc/thesis/abstract.tex
Editing, midway through chp 2

@@
+%TC:macro \del [ignore]
+
 Models of the spread of disease in a population often make the simplifying
@@
 structures formed by social contacts, which can have a significant influence on
-the rate {\color{blue}\uline{and pattern}} of epidemic spread. Contact
-{\color{red}\sout{network models}} {\color{blue}\uline{networks}} capture this
-structure by explicitly representing each contact that could possibly lead to a
-transmission. {\color{blue}\uline{Contact network models parameterize the
-structure of these networks, but estimating their parameters from contact data
-requires extensive, often prohibitive, epidemiological investigation. }}
+the rate \add{and pattern} of epidemic spread. Contact \del{network models}
+\add{networks} capture this structure by explicitly representing each contact
+that could possibly lead to a transmission. \add{Contact network models
+parameterize the structure of these networks, but estimating their parameters
+from contact data requires extensive, often prohibitive, epidemiological
+investigation.}
 
@@
 for ABC, Gillespie simulation for propagating epidemics though networks, and a
-{\color{blue}\uline{previously developed}} kernel-based tree similarity score.
-{\color{blue}\uline{Our method offers the potential to quantitatively
-investigate contact network structure from phylogenies derived from viral
-sequence data, complementing traditional epidemiological methods.}}
+\add{previously developed} kernel-based tree similarity score. \add{Our method
+offers the potential to quantitatively investigate contact network structure
+from phylogenies derived from viral sequence data, complementing traditional
+epidemiological methods.}
 
-We applied our method to {\color{red}\sout{fit}}
-{\color{blue}\uline{investigate}} the Barab\'{a}si-Albert network model.
-{\color{blue}\uline{This model incorporates the preferential attachment
-mechanism observed in real world social and sexual networks, whereby
-individuals with more connections attract new contacts at an elevated rate
-(``the rich get richer'').}}
-{\color{red}\sout{to simulated transmission trees and applied it to viral
-phylogenies estimated from six real-world HIV sequence datasets.}} Using
+We applied our method to \del{fit} the Barab\'{a}si-Albert network model.
+\add{This model incorporates the preferential attachment mechanism observed in
+real world social and sexual networks, whereby individuals with more
+connections attract new contacts at an elevated rate (``the rich get
+richer'').} \del{to simulated transmission trees and applied it to viral
+phylogenies estimated from six real-world HIV sequence datasets.} Using
 simulated data, we found that the strength of preferential attachment and the
@@
 
-Finally, the Barab\'{a}si-Albert model was fit to six real world HIV datasets,
-and substantial heterogeneity in the parameter estimates was observed.
-{\color{red}\sout{Point estimates}}{\color{blue}\uline{Depending on the choice
-of prior, posterior means}} for the preferential attachment power ranged from
-{\color{red}\sout{0.06 to 1.05}} {\color{blue}\uline{0.26, slightly less than
-logarithmic, to 1.00, exactly linear.}} {\color{blue}\uline{Point estimates of
-the strength of preferential attachment were higher in injection drug user
-populations, potentially indicating that high-degree ``hub'' nodes may play a
-role in epidemics among this risk group.}} Our results underscore the
-importance of considering contact structures when {\color{red}\sout{performing
-phylodynamic inference}}{\color{blue}\uline{investigating viral outbreaks.}}
+Finally, the Barab\'{a}si-Albert model was fit to \del{six}\add{eleven} real
+world HIV datasets, and substantial heterogeneity in the parameter estimates
+was observed. \del{Point estimates} \add{Posterior means} for the preferential
+attachment power \add{were all sub-linear, consistent with literature results.
+We found that the strength of preferential attachment was higher in injection
+drug user populations, potentially indicating that high-degree
+``superspreader'' nodes may play a role in epidemics among this risk group.}
+Our results underscore the importance of considering contact structures when
+\del{performing phylodynamic inference}\add{investigating viral outbreaks.}

### 2016-06-16 doc/thesis/abstract_exit.tex
Editing, midway through chp 2

Models of the spread of disease in a population often make the simplifying assumption that the population is homogeneously mixed, or is divided into homogeneously mixed compartments. However, human populations have complex structures formed by social contacts, which can have a significant influence on the rate and pattern of epidemic spread. Contact networks capture this structure by explicitly representing each contact that could possibly lead to a transmission. Contact network models parameterize the structure of these networks, but estimating their parameters from contact data requires extensive, often prohibitive, epidemiological investigation.

We developed a method based on approximate Bayesian computation (ABC) for estimating structural parameters of the contact network underlying an observed viral phylogeny. The method combines adaptive sequential Monte Carlo for ABC, Gillespie simulation for propagating epidemics though networks, and a previously developed kernel-based tree similarity score. Our method offers the potential to quantitatively investigate contact network structure from phylogenies derived from viral sequence data, complementing traditional epidemiological methods.

We applied our method to the Barab\'{a}si-Albert network model. This model incorporates the preferential attachment mechanism observed in real world social and sexual networks, whereby individuals with more connections attract new contacts at an elevated rate (``the rich get richer''). Using simulated data, we found that the strength of preferential attachment and the number of infected nodes could often be accurately estimated. However, the mean degree of the network and the total number of nodes appeared to be weakly- or non-identifiable with ABC.

 Finally, the Barab\'{a}si-Albert model was fit to eleven real world HIV datasets, and substantial heterogeneity in the parameter estimates was observed. Posterior means for the preferential attachment power were all sub-linear, consistent with literature results. We found that the strength of preferential attachment was higher in injection drug user populations, potentially indicating that high-degree ``superspreader'' nodes may play a role in epidemics among this risk group. Our results underscore the importance of considering contact structures when investigating viral outbreaks.

### 2016-06-16 doc/thesis/methods_results.Rnw
Editing, midway through chp 2

--- 2016-06-14 doc/thesis/methods_results.Rnw
@@
 
-\subsection{Motivation}
+\subsection{Why study the \acrlong{BA} model?}
 
 \add{We developed \software{netabc} with the objective of extracting useful,
-quantitative information about network structures from viral phylogenies.}
-
-\add{There are three aspects to ``usefulness''. The first is model
-specification and the biological or epidemiological interpretation of the
-parameters. We want the model to be somewhat realistic, so that \ldots, but no
-too complicated that it becomes difficult to interpret. At least some of the
-parameters should be of interest from a theoretical or practical perspective.
-As this is a phylodynamic method, intended to be used with viral sequence data, 
-we would also like to choose parameters which may be difficult to estimate with
-more standard methods; otherwise, our method provides no advantage. The
-\gls{BA} model (\cref{subsec:pa}) satisfies these criteria, albeit some better
-than others. The purported realism of the model stems from the fat-tailed
-degree distributions it produces, which are similar to those observed in real
-world sexual networks~\autocite{colgate1989risk, liljeros2001web,
-schneeberger2004scale, clemenccon2015statistical, rothenberg2007large}.
-Moreover, the ``rich get richer'' phenomenon, where popular individuals attract
-new connections at an elevated rate, is intuitively reasonable for both
-sexual~\autocite{de2007preferential} and
+quantitative information about network structures from viral phylogenies. An
+important aspect of ``usefulness'' is model specification and the biological or
+epidemiological interpretation of the parameters. We want the model to be
+realistic, but no so complicated that it becomes difficult to interpret. At
+least some of the parameters should be of interest from a theoretical or
+practical perspective, or there would be no point in estimating them. Since
+\software{netabc} is a phylodynamic method, intended to be used with viral
+sequence data, we would also like to choose parameters which may be difficult
+to estimate with more standard methods. Otherwise, our method provides no
+advantage. The \gls{BA} model (\cref{subsec:pa}) satisfies these criteria,
+albeit some better than others. The purported realism of the model stems from
+the fat-tailed degree distributions it produces, which are similar to those
+observed in real world sexual networks~\autocite{colgate1989risk,
+liljeros2001web, schneeberger2004scale, clemenccon2015statistical,
+rothenberg2007large}. Moreover, the ``rich get richer'' phenomenon, where
+popular individuals attract new connections at an elevated rate, is intuitively
+reasonable for both sexual~\autocite{de2007preferential} and
 \gls{IDU}~\autocite{dombrowski2013topological} networks. However, the model is
@@
 denoted \gls{N}, \gls{m}, \gls{alpha}, and \gls{I} (see \cref{subsec:pa}). The
-first three of these are parameters of the model itself, while \gls{I} is
+first three of these parameterize the network structure, while \gls{I} is
 related to the simulation of transmission trees over the network. However, we
@@
 vertex, or equivalently one-half of the average degree. \gls{alpha} is the
-power of preferential attachment - new nodes are attached to existing nodes of
+power of preferential attachment -- new nodes are attached to existing nodes of
 degree $d$ with probability proportional to $d^{\gls{alpha}} + 1$. Finally,
-\gls{I} is the number of infected individuals at the time when sampling occurs.}
+\gls{I} is the number of infected individuals at the time when sampling occurs.
+The \gls{alpha} parameter is unitless, while \gls{m} has units of edges or
+connections per vertex, and \gls{N} and \gls{I} both have units of nodes or
+individuals.}
 
@@
 an ongoing epidemic, while total susceptible population size \gls{N} provides a
-similar metric for preventative measures. The average degree, controlled by
-\gls{m}, may have some value in predicting how fast an epidemic will spread.
-When there are more connections in the network, the time intervals between
-transmission events decrease, as does the number of transmissions required for
-the epidemic to travel from one part of the network to another. High average
-degrees may be reduced by interventions causing behaviour change, such as using
-condoms or caughing into one's sleeve; these types of interventions will be
-less helpful when the average number of contacts per person is already low. The
-preferential attachment power \gls{alpha} is perhaps the least obviously
-useful, but it can offer some insight into the dynamics of the network and how
-it would react to the removal of nodes. \textcite{dombrowski2013topological} 
-found evidence of preferential attachment in \gls{IDU} networks, and suggested
-that as a consequence of this characteristic, the removal of random nodes (such
-as through a police crackdown) might inadventently make it easier for epidemics
-to spread. When individuals with only one or two connections lose them, they
-might tend to seek out well-known (that is, high-degree) members of the
-community, thus increasing those individuals' connectivity even further.}
+similar metric for preventative measures. The average degree of the network, in
+this case $2\gls{m}$, is directly related to \gls{R0}, the basic reproductive
+number~\autocite{britton2002bayesian}. \gls{R0} quantifies the number of
+secondary infections ultimately caused by one infected individual; higher
+\gls{R0} generally indicates faster epidemic growth and/or larger eventual
+epidemic size~\autocite{anderson1992infectious}. In a homogeneously mixed
+population, the proportion of the population which must be vaccinated to
+control an epidemic can be expressed in terms of
+\gls{R0}~\autocite{anderson1982directly}. Although optimal vaccination
+strategies may differ in heterogeneous contact
+structures~\autocite{ma2013importance}, it is reasonable to suppose that there
+would still be a relationship between \gls{m}, \gls{R0}, and the vaccination
+threshold. The preferential attachment power \gls{alpha} quantifies the degree
+to which high-degree nodes, also called
+superspreaders~\autocite{kemper1980identification}, characterize the network
+structure. Superspreaders have been hypothesized to play a greater-than-average
+role in the spread of several diseases~\autocite{stadler2013uncovering,
+shen2004superspreading}. If so, network-based
+interventions~\autocite{little2014using, wang2015targeting} may be worth
+considering as part of an epidemic control strategy. \gls{alpha} can also offer
+some insight into how the network would react to the removal of nodes.
+\textcite{dombrowski2013topological} found evidence of preferential attachment
+in \gls{IDU} networks, and suggested that as a consequence of this
+characteristic, the removal of random nodes (such as through a police
+crackdown) might inadventently make it easier for epidemics to spread. When
+individuals with only one or two connections lose them, they might tend to seek
+out well-known (that is, high-degree) members of the community, thus increasing
+those individuals' connectivity even further.}
 
 \add{All four \gls{BA} parameters can be estimated using more conventional
-approaches, but the difficulty of doing so varies between the parameters and
-based on other characteristics of the epidemic under consideration. All network
-parameters can theoretically be estimated by explicitly constructing the
-contact network, although this is highly resource intensive and is hampered by 
-misreporting and other challenges~\autocite{eames2015six}. In any event, if we
-are in a position to know the true contact network then the point of estimating
-network parameters is moot. Depending on the context, \gls{N} may be estimable
-with a survey, but if the population in question is ``hidden'' (such as
-\gls{IDU} in regions where drug use is illegal) estimation of numbers may be
-more challenging. Respondent-driven
-sampling~\autocite{heckathorn1997respondent} was developed to investigate such
-populations. The prevalence \gls{I} may be estimated from reports of hospital
-admissions or engagement in care, but this may be very difficult for diseases
-such as \gls{HCV} which may remain asymptomatic for decades. Routine
-population-level testing is another possibility, but is resource intensive and
-the barriers associated with hidden populations still apply. The average degree
-of the network ($2\gls{m}$) can be ascertained by simply asking a sample of
-people how many contacts they had, but they might be unwilling to disclose this
-information or, in the case of airborne transmissions, they might not know. The
-estimation of \gls{alpha} is even more complex, as this parameter is most
-strongly reflected in the connectivity of very high degree nodes, who are rare
-in the population. Even if full degree distributions are available, there are
-models other than preferential attachment which can produce scale-free 
+approaches, but this estimation may be challenging depending on characteristics
+of the epidemic under consideration. In theory, any network parameter can be
+estimated by explicitly constructing the contact network, although this is
+highly resource intensive and is hampered by misreporting and other
+challenges~\autocite{eames2015six}. All parameters become more difficult to
+estimate when the infected population is ``hidden'' due to illegal or
+stigmatized behaviour, as is sometimes the case with \gls{HIV} outbreaks among
+\gls{IDU}, \gls{MSM}, or sex workers. Of course, phylodynamic methods are
+equally ineffective when the population is completely hidden, since we must at
+least be able to sample viral sequences from its members. If a population is
+accessible, \gls{N} and \gls{I} may be estimable by surveying and/or testing
+each member of the population for the disease. However, this will not tell us
+if there are large compartments of the population that we simply have not
+sampled. Our hope is that estimating \gls{N} and \gls{I} phylogenetically might
+provide this additional information. The average degree of the network,
+$2\gls{m}$, is also estimable by a survey, although individuals may be
+unwilling or unable to disclose how many contacts they have had. The estimation
+of \gls{alpha} is more complex, as this parameter is most strongly reflected in
+the connectivity of very high degree nodes, who are rare in the population.
+Locating them might require contact tracing, or respondent-driven
+sampling~\autocite{heckathorn1997respondent}. Even if the full degree
+distribution of a network is available, there are models other than
+preferential attachment which can produce scale-free
 networks~\autocite[\eg][]{kumar2000stochastic}. \textcite{de2007preferential}
@@
 survey, which would likely miss the few high-degree nodes characterizing a
-power law degree distribution. In short, each of these parameters may be
-estimated without phylodynamics, but there are sufficient difficulties that we
-believe an alternative method based on sequence data is warranted.}
-
-\end{document}
+power law degree distribution. In summary, each of the \gls{BA} parameters may
+be estimated without phylodynamics, but there are sufficient difficulties that
+we believe an alternative method using sequence data is warranted.}
+
+\subsection{Using synthetic data to investigate identifiability and sources of
+estimation error and bias}
+
+\add{We have argued that the parameters of the \gls{BA} model are interesting
+and worth estimating with phylodynamic methods. However, these estimates will
+only constitute ``useful'' information about the network if the parameters are
+identifiable from phylogenetic data. Roughly speaking, the identifiability of a
+parameter says how much information about that parameter can possibly be
+obtained from the observed data. If the parameters of the \gls{BA} model do not
+influence tree shape at all, then we cannot possibly estimate them -- the
+posterior distribution will exactly resemble the prior, no matter how accurate
+a representation of the posterior we are able to produce. Hence, before
+proceeding with a full validation of \software{netabc} on simulated data, we
+undertook two experiments designed to assess the identifiability of the
+\gls{BA} parameters. These experiments only investigated one parameter of the
+\gls{BA} model at a time while holding all others fixed, a strategy commonly
+used when performing sensitivity analyses of mathematical models. This allowed
+us to perform a fast preliminary analysis without dealing with the ``curse of
+dimensionality'' of the full parameter space. The experiments are motivated and
+described on a high level here, with more detail provided in the next section.}
+
+\add{First, we simulated trees under three different values of each parameter,
+and asked how well we could tell the different trees apart. The better we are
+able to distinguish the trees, the more identifiability we might expect for the
+corresponding parameter when we attempt to estimate it with \gls{ABC}. This
+experiment also had the secondary purpose of validating our choice of the tree
+kernel as a distance measure in \gls{ABC}. To tell the trees apart, we used a
+classifier based on the tree kernel, but we also tested two other tree shape
+statistics: one which considers only the topology, and another which considers
+only branching times. Since the tree kernel incorporates both of these sources
+of information, we expected it to outperform the other two statistics. Finally,
+the tree kernel can be ``tuned'' by adjusting the values of the meta-parameters
+\gls{lambda} and \gls{sigma}. The results of this experiment were used to
+select values for these meta-parameters to carry forward to the rest of the
+thesis, based on their accuracy in distinguishing the different trees. }
+
+\add{A second experiment was designed to test whether we could actually
+estimate the parameters numerically, rather than just telling three different
+values apart, and also to assess how identifiability varied in different
+regions of the parameter space. An individual tree was compared to simulated
+trees on a one-dimensional grid of values of one \gls{BA} parameter, to obtain
+a ``distribution'' of kernel score values. From this distribution, estimates
+and credible intervals of the parameter could be calculated. Repeating this
+experiment with trees located throughout the parameter space allowed us to
+beter quantify the identifiability. Furthermore, doing marginal estimation
+(that is, estimating one parameter with all others fixed) can provide insight
+into any biases observed when doing joint estimation with \gls{ABC}. If grid
+search is inaccurate, it indicates a lack of parameter identifiability.
+However, if the marginal grid search estimates are accurate but the estimates
+obtained with \gls{ABC} are biased, this points to confounding between the
+parameters which could only be observed when they are all estimated jointly.}
+
+\add{After these preliminary experiments, the strategy we used for testing
+\software{netabc} was a standard simulation-based validation. Transmission
+trees were simulated under several combinations of parameter values, and we
+tried to recover these values with \software{netabc}. We then used a
+multivariable analysis to investigate how accuracy of these estimates was
+influenced by the true parameter values.}
+
+\add{In the previous section, we argued that the \gls{BA} model parameters were
+worth investigating, and here we have presented several computational
+experiments designed to assess their identifiability. There is a final, more
+technical aspect of our method's ``usefulness'' to consider, which is the
+accuracy of the \gls{ABC} approximation to the posterior. As discussed in
+\cref{sec:abc}, \gls{ABC} does not target the posterior distribution directly,
+but rather an approximate posterior derived from simulated data and a distance
+function. \gls{ABC} \emph{assumes} that this approximate posterior resembles
+the true posterior, and it is critical for our estimates' relevance that this
+assumption holds. There are two potential causes of an inaccurate \gls{ABC}
+approximation~\autocite{fearnhead2012constructing}. First, the Monte Carlo
+approximation to the \gls{ABC} target distribution may be poor, due to the
+settings used for \gls{ABC}-\gls{SMC}. Second, the \gls{ABC} target
+distribution may not resemble the posterior, due to a poor choice of distance
+function. We designed two experiments to investigate the impact of these
+sources of error.}
+
+\add{The Monte Carlo approximation error is fairly easily quantified by simply
+increasing the computing power used for \gls{SMC}. We ran one simulation using
+a larger number of pakticles, more simulated datasets per particle, and a
+higher value for \gls{alphaess}. A substantial improvement in accuracy
+resulting from these changes would likely indicate a high Monte Carlo error
+with the lower settings. The second issue, the resemblance of the \gls{ABC}
+target distribution to the true posterior, is somewhat more difficult to
+investigate. We do not have access to the true posterior, even for simulations
+where the true parameter values are known. To address this source of error, we
+performed marginal parameter estimation with \gls{ABC} by informing
+\software{netabc} of some of the true parameter values. Any inaccuracy or bias
+observed only in the joint estimation results, but not the marginal estimates, 
+is most easily explained by interdependence between parameters in the true
+posterior. However, errors observed in both marginal and joint estimates could
+be due to either the shape of the true posterior or an inaccurate \gls{ABC}
+approximation, and we have no way to distinguish one from the other. In other
+words, this experiment provided only an upper bound on the error due to an
+inaccurate \gls{ABC} approximation.}
 
@@
 
-%{\color{blue}\uline{The \gls{BA} contact network model} (see \cref{subsec:pa})
-%\uline{generates networks with a power law degree distribution. Power law
-%distributions have been observed by several empirical studies of sexual
-%behaviour}~\autocite{colgate1989risk, liljeros2001web, schneeberger2004scale,
-%clemencon2015statistical} (although not all, see
-%\autocite{handcock2004likelihood, bansal2007individual}).}
-
-\add{Using synthetic data,} we investigated four parameters related to the
-\gls{BA} contact network model, 
-denoted \gls{N}, \gls{m}, \gls{alpha}, \gls{I}
-(see \cref{subsec:pa}). The first three of these are parameters of the model
-itself, while \gls{I} is related to the simulation of transmission trees over
-the network. However, we will refer to all four as \gls{BA} parameters. \gls{N}
-denotes the total number of nodes in the network, or equivalently, susceptible
-individuals in the population. When a node is added to the network, \gls{m} new
-undirected edges are added incident to it, and are attached to existing nodes
-of degree $d$ with probability proportional to $d^{\gls{alpha}} + 1$
-(\cref{subsec:pa}). To simulate transmission trees over a \gls{BA} network, we
-allowed an epidemic to spread until \gls{I} nodes were infected, and sampled a
-transmission tree at that time. The \gls{alpha} parameter is unitless, while
-\gls{m} has units of edges or connections per vertex, and \gls{N} and \gls{I}
-both have units of nodes or individuals.
-
-\end{document}
-
-{\color{blue}\uline{Our investigation comprised three sets of computational
-experiments aimed at determining which of these parameters could be estimated
-from phylogenetic data, and with what degree of accuracy. First, we performed
-an exploratory analysis using a tree kernel-based classifier to determine which
-of the four \gls{BA} parameters might be identifiable from tree shapes when all
-others were held fixed. This analysis was also used to validate our choice of
-the tree kernel as a distance function for \gls{ABC}, by evaluating two further
-classifiers based on other tree summary statistics from the literature. Second,
-we used grid search to investigate the accuracy and precision of marginal
-parameter estimates, again holding all but one parameters of the model fixed.
-Finally, we applied the full \gls{ABC}-\gls{SMC} algorithm to jointly estimate
-the four parameters of the \gls{BA} model simultaneously.}}
-
-We assumed that all contacts had symmetric transmission risk, which was
-implemented by replacing each undirected edge in the network with two directed
-edges (one in each direction). Nodes in our networks followed simple \gls{SI}
-dynamics, meaning that they became infected at a rate proportional to their
-number of infected neighbours, and never recovered. We did not consider the
-time scale of the transmission trees in these simulations, only their shape.
-Therefore, the transmission rate along each edge in the network was set to 1,
-the removal rate of each node was set to 0, and all transmission trees' branch
-lengths were scaled by their mean. The \textit{igraph} library's implementation
-of the BA model~\autocite{csardi2006igraph} was used to generate the graphs.
-The analyses were run on Westgrid (\url{https://www.westgrid.ca/}) and a local
-computer cluster. With the exception of our own programs, all analyses were
-done in \software{R}, and all packages listed below are \software{R} packages.
-{\color{blue}\uline{Code to run all simulation-based experiments is freely
-available at \url{https://github.com/rmcclosk/thesis}.}}
+\del{We investigated four parameters related to the \gls{BA} contact network
+model, denoted \gls{N}, \gls{m}, \gls{alpha}, \gls{I} (see \cref{subsec:pa}).
+The first three of these are parameters of the model itself, while \gls{I} is
+related to the simulation of transmission trees over the network. However, we
+will refer to all four as \gls{BA} parameters. \gls{N} denotes the total number
+of nodes in the network, or equivalently, susceptible individuals in the
+population. When a node is added to the network, \gls{m} new undirected edges
+are added incident to it, and are attached to existing nodes of degree $d$ with
+probability proportional to $d^{\gls{alpha}} + 1$ (\cref{subsec:pa}). To
+simulate transmission trees over a \gls{BA} network, we allowed an epidemic to
+spread until \gls{I} nodes were infected, and sampled a transmission tree at
+that time.}
+
+\add{For all simulations,} we assumed that all contacts had symmetric
+transmission risk, which was implemented by replacing each undirected edge in
+the network with two directed edges (one in each direction). Nodes in our
+networks followed simple \gls{SI} dynamics, meaning that they became infected
+at a rate proportional to their number of infected neighbours, and never
+recovered. We did not consider the time scale of the transmission trees in
+these simulations, only their shape. Therefore, the transmission rate along
+each edge in the network was set to 1, the removal rate of each node was set to
+0, and all transmission trees' branch lengths were scaled by their mean. The
+\textit{igraph} library's implementation of the BA
+model~\autocite{csardi2006igraph} was used to generate the graphs. The analyses
+were run on Westgrid (\url{https://www.westgrid.ca/}) and a local computer
+cluster. With the exception of our own C programs, all analyses were done in
+\software{R}, and all packages listed below are \software{R} packages.
+\add{Code to run all experiments is freely available at
+    \url{https://github.com/rmcclosk/thesis}.}
 
@@
 
-{\color{blue}\uline{
-Our first computational experiment was designed as an exploratory analysis of
-the four \gls{BA} model parameters defined above: \gls{alpha}, \gls{I},
+\add{Our first computational experiment was designed as an exploratory analysis
+of the four \gls{BA} model parameters defined above: \gls{alpha}, \gls{I},
 \gls{m}, and \gls{N}. The objective of this experiment was to determine whether
@@
 
-\uline{In addition to testing for identifiability, a secondary objective of
+\add{In addition to testing for identifiability, a secondary objective of
 this analysis was to validate the use of the tree kernel as a distance measure
-for \gls{ABC} in our context. As discussed in} \cref{sec:abc}, \uline{the
-choice of distance function is extremely important for the accuracy of the
-\gls{ABC} approximation to the posterior. Therefore, we evaluated two
-additional tree statistics in the same manner as we evaluated the tree kernel
-(that is, by constructing and testing a classifier). First, we considered
-Sackin's index~\autocite{shao1990tree}, which measures the degree of imbalance
-or asymmetry in a phylogeny} (see \cref{subsec:treeshape}). \uline{Sackin's
-index is widely used for characterizing
-phylogenies~\autocite{frost2013modelling} and has been demonstrated to vary
-between transmission trees simulated under different contact network
-types~\autocite{leventhal2012inferring}. Sackin's index does not take branch
-lengths into account, considering only the tree's topology. The other statistic
-we considered was the \gls{nltt}~\autocite{janzen2015approximate}, which
-compares two trees based on normalized distributions of their branching times}
-(see \cref{subsec:treeshape}).  \uline{In contrast with Sackin's index, the
-\gls{nltt} does not explicity consider the trees' topologies, but it does use
-their normalized branch lengths. While the \gls{nltt} is a newly developed
-statistic not yet in widespread use, the unnormalized
-\gls{ltt}~\autocite{nee1992tempo} was the basis of seminal early work
-extracting epidemiological information from
-phylodynamics~\autocite{holmes1995revealing}. We expected the tree kernel to
+for \gls{ABC} in our context. As discussed in \cref{sec:abc}, the choice of
+distance function is extremely important for the accuracy of the \gls{ABC}
+approximation to the posterior. Therefore, we evaluated two additional tree
+statistics in the same manner as we evaluated the tree kernel (that is, by
+constructing and testing a classifier). First, we considered Sackin's
+index~\autocite{shao1990tree}, which measures the degree of imbalance or
+asymmetry in a phylogeny (see \cref{subsec:treeshape}). Sackin's index is
+widely used for characterizing phylogenies~\autocite{frost2013modelling} and
+has been demonstrated to vary between transmission trees simulated under
+different contact network types~\autocite{leventhal2012inferring}. Sackin's
+index does not take branch lengths into account, considering only the tree's
+topology. The other statistic we considered was the
+\gls{nltt}~\autocite{janzen2015approximate}, which compares two trees based on
+normalized distributions of their branching times (see
+\cref{subsec:treeshape}). In contrast with Sackin's index, the \gls{nltt} does
+not explicity consider the trees' topologies, but it does use their normalized
+branch lengths. While the \gls{nltt} is a newly developed statistic not yet in
+widespread use, the unnormalized \gls{ltt}~\autocite{nee1992tempo} was the
+basis of seminal early work extracting epidemiological information from
+phylogenies~\autocite{holmes1995revealing}. We expected the tree kernel to
 classify the \gls{BA} parameters more accurately than either Sackin's index or
 the \gls{nltt}, since the tree kernel takes both topology and branch lengths
-into account. }}
+into account.}
 
@@
 
+\begin{figure}[ht]
+    \centering
+    \includegraphics{kernel-expt.pdf}
+    \caption[
+        Schematic of classifier experiments investigating identifiability of BA
+        model parameters from tree shapes.
+    ]{
+        Schematic of classifier experiments investigating identifiability of BA
+        model parameters from tree shapes. The parameters of the BA model were
+        varied one at a time \add{while holding all others fixed}. Transmission
+        trees were simulated under three different values of each
+        parameter\del{, then compared pairwise using the tree kernel.}
+        Classifiers were constructed for each parameter \add{based on three
+        tree shape statistics}, and their accuracy was evaluated by
+        cross-validation. Kernel-\gls{PCA} projections were used to visually
+        examine the separation of the trees in the feature space defined by the
+        tree kernel.
+    }
+    \label{fig:kernelexpt}
+\end{figure}
+
 Similar experiments were performed with the values shown in
-\cref{tab:kernelexpt}. The other three \gls{BA} parameters, namely \gls{N},
-\gls{m}, and \gls{I}, were each varied while holding the others fixed. The
-experiments for \gls{alpha}, \gls{m}, and \gls{N} were repeated with three
-different values of \gls{I}. All experiments were repeated with trees having
-three different numbers of tips. Kernel matrices were computed for all pairs of
-the meta-parameters \gls{lambda} = \sett{0.2, 0.3, 0.4} and \gls{sigma} =
+\cref{tab:kernelexpt}. The other three \gls{BA} parameters, \gls{N}, \gls{m},
+and \gls{I}, were each varied while holding the others fixed. The experiments
+for \gls{alpha}, \gls{m}, and \gls{N} were repeated with three different values
+of \gls{I}. All experiments were repeated with trees having three different
+numbers of tips. Kernel matrices were computed for all pairs of the
+meta-parameters \gls{lambda} = \sett{0.2, 0.3, 0.4} and \gls{sigma} =
 \sett{\nicefrac18, \nicefrac14, \nicefrac12, 1, 2, 4, 8}.
@@
     \input{\tablepath/kernel-expt}
-  \caption[Variables used in tree kernel simulation experiments]
-  {
-    Values of parameters and other variables used in tree kernel simulation
-    experiments. Each row corresponds to one of the \gls{BA} model parameters.
-    One kernel matrix was created for every combination of values except the
-    one indicated in the ``varied parameter'' column, which was varied when
+    \caption[
+        Values of parameters and meta-parameters used in classifier experiments
+        to investigate identifiability of BA model parameters from tree shapes.
+    ]{
+        Values of parameters and meta-parameters used in classifier experiments
+        to investigate identifiability of BA model parameters from tree shapes.
+        Each row corresponds to one of the BA model parameters. One kernel
+        matrix was created for every combination of values except the one
+        indicated in the ``varied parameter'' column, which was varied when
         producing simulated trees.
@@
     \input{\tablepath/gridsearch-expt}
-  \caption[Variables used in grid search experiments]
-  {
-    Variables and \gls{BA} parameter values used for grid search experiments. 
-    Trees were simulated under the test values, and compared to a grid of trees
-    simulated under the grid values. Kernel scores were used to calculate point
-    estimates and credible intervals for the test values.
+    \caption[
+        Values of parameters and meta-parameters used in grid search
+        experiments to further investigate identifiability of BA model
+        parameters.
+    ]{
+        Values of parameters and meta-parameters used in grid search
+        experiments to further investigate identifiability of BA model
+        parameters. Trees were simulated under the test values, and compared to
+        a grid of trees simulated under the grid values. Kernel scores were
+        used to calculate point estimates and credible intervals for each
+        parameter, which were compared to the test values.
     }
@@
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics{kernel-expt.pdf}
-  \caption[Schematic of experiments investigating impact of BA model parameters
-           on tree shape.]{
-    Schematic of experiments designed to investigate the impact of variations
-    in BA model parameters on transmission tree shapes. The parameters of the
-    BA model were varied one at a time {\color{blue}\uline{while holding all
-    others fixed}}. Transmission trees were simulated under three different
-    values of each parameter, then compared pairwise using the tree kernel.
-    Classifiers were constructed for each parameter, and their accuracy was
-    evaluated by cross-validation. Kernel-PCA projections were used to visually
-    examine the separation of the trees in the feature space defined by the
-    tree kernel.
-  }
-  \label{fig:kernelexpt}
-\end{figure}
-
 \subsubsection*{Grid search}
 
-{\color{blue}\uline{The previous experiment was an exploratory analysis
-intended to determine which of the \gls{BA} parameters were identifiable, and
-whether the tree kernel could potentially be used to distinguish different
-parameter values when all others were held fixed. In this experiment, which was
-still of an exploratory nature, we continued to consider one parameter at a
-time while fixing the other three. However, rather than checking for
-identifiability, we were now interested in quantifying the accuracy and
-precision of kernel score-based estimates. This was done by examining the
-distribution of kernel scores on a grid of parameter values, when trees
-simulated according to those values were compared with a single simulated test
-tree. This experiment did not involve the full \gls{ABC}-\gls{SMC} algorithm,
-nor did we continue to use Sackin's index or the \gls{nltt}.}}
+\add{The previous experiment was an exploratory analysis intended to determine
+which of the \gls{BA} parameters were identifiable, and whether the tree kernel
+could potentially be used to distinguish different parameter values when all
+others were held fixed. In this experiment, which was still of an exploratory
+nature, we continued to consider one parameter at a time while fixing the other
+three. However, rather than checking for identifiability, we were now
+interested in quantifying the accuracy and precision of kernel score-based
+estimates. This was done by examining the distribution of kernel scores on a
+grid of parameter values, when trees simulated according to those values were
+compared with a single simulated test tree.}
 
@@
 tree had 500 tips, and were simulated with the other \gls{BA} parameters set to
-{\color{blue}\uline{the known values}} $N$ = 5000, $m$ = 2, and $I$ = 1000. The
-test tree was compared to each of the grid trees using the tree kernel, with
-the meta-parameters set to $\lambda = 0.3$ and $\sigma = 4$, using the
-\software{treekernel} binary. The median kernel score was calculated for each
-grid value, and the scores were normalized such that the area under the curve
-was equal to 1. {\color{red}\sout{The grid value with the highest median kernel
-score was taken as the point estimate for the test value}}{\color{blue}\uline{
-A point estimate for the test value was obtained by calculating the weighted
-mean of the grid values with the kernel scores as weights}}, and a 95\%
-credible interval was obtained using the \software{hpd} function in the
-\software{TeachingDemos} package~\autocite{snow2013teachingdemos}.
+\add{the known values} $\gls{N} = 5000$, $\gls{m} = 2$, and $\gls{I} = 1000$.
+The test tree was compared to each of the grid trees using the tree kernel,
+with the meta-parameters set to $\gls{lambda} = 0.3$ and $\gls{sigma} = 4$,
+using the \software{treekernel} binary. The median kernel score was calculated
+for each grid value, and the scores were normalized such that the area under
+the curve was equal to 1. \del{The grid value with the highest median kernel
+score was taken as the point estimate for the test value.} \add{For all
+parameters except \gls{m}}, 50\% and 95\% highest denstity intervals wer
+obtained using the \software{hpd} function in the \software{TeachingDemos}
+package~\autocite{snow2013teachingdemos}. \add{Since the \software{hpd}
+function assumes a continuous distribution, we implemented our own version for
+discrete distributions to use for \gls{m}.}
 
@@
 are listed in \cref{tab:gridexpt}. A graphical schematic of the grid search
-experiments is shown in \cref{fig:gridexpt}. {\color{blue}\uline{We emphasize
-that these experiments were marginal in nature; that is, the each parameter was
-varied and estimated individually while holding the others at known, fixed
-values.}}
-
-%Spearman's correlation was used to test whether the number of tips in the tree
-%was correlated with the acccuracy of the point estimates. One-way \gls{ANOVA}
-%was used to test whether the accuracy of the point estimates were significantly
-%higher for any parameter values. If so, the distribution of errors was examined
-%to choose a suitable post hoc test. If there appeared to be a correlation
-%between the parameter value and the accuracy of the estimate, Spearman's
-%correlation was calculated. If there were one or more particular values which
-%appeared to have lower error, we used the Wilcoxon rank-sum test.
+experiments is shown in \cref{fig:gridexpt}.
 
@@
     \includegraphics[width=\textwidth]{gridsearch-expt}
-  \caption[Schematic of grid search experiment.]{
-    Graphical schematic of grid search experiments used to investigate \gls{BA}
-    model parameters. Trees were simulated along a narrowly spaced grid of
-    values of one parameter (``grid trees'') {\color{blue}\uline{with all other
-    parameters fixed to known values}}. Separate trees were simulated for a
-    small subset of the grid values (``test trees''), {\color{blue}\uline{also
-    holding the other parameters fixed}}. Each test tree was
-    compared to every grid tree using the tree kernel, and the resulting kernel
-    scores were normalized to resemble a probability density from which the
-    mode and 95\% highest density interval were calculated.
+    \caption[
+        Schematic of grid search experiment to further investigate
+        identifiability of BA model parameters from tree shapes.
+    ]{
+        Schematic of grid search experiment to further investigate
+        identifiability of BA model parameters from tree shapes. Trees were
+        simulated along a narrowly spaced grid of values of one parameter
+        (``grid trees'') \add{with all other parameters fixed to known values}.
+        Separate trees were simulated for a small subset of the grid values
+        (``test trees''), \add{also holding the other parameters fixed}. Each
+        test tree was compared to every grid tree using the tree kernel, and
+        the resulting kernel scores were normalized to resemble a probability
+        density from which the mode and 95\% highest density interval were
+        calculated.
     }
@@
 
-{\color{blue}\uline{
-Our final simulation-based experiment was designed to test the full
+\add{Our final synthetic data experiment was designed to test the full
 \gls{ABC}-\gls{SMC} algorithm by jointly estimating the four parameters of the
-\gls{BA} model. In contrast to the previous two experiments, we did not inform
-the algorithm of any of the true parameter values.  }}
-
-We simulated three trees each under a variety of parameter values and ran the
-\software{netabc} program to estimate posterior distributions for the
-parameters. The parameter values and priors used are listed in
-\cref{tab:abcexpt}. The tree kernel meta-parameters were set to $\lambda = 0.3$
-and $\sigma = 4$. The \gls{SMC} algorithm was run with 1000 particles, five
-sampled datasets per particle, and the $\alpha_{\ESS}$ parameter (not to be
-confused with the \gls{BA} preferential attachment parameter, see
-\cref{subsec:adaptsmc}) set to 0.95. The algorithm was stopped when the
+\gls{BA} model. We used the standard validation approach of simulating
+transmission trees under the model with known parameter values and attempting
+to recover those values with \software{netabc}. The algorithm was not informed
+of any of the true parameter values for the main set of simulations. Despite
+the fact that the parameter values used to generate the simulated transmission
+trees were known, the true posterior distributions on the \gls{BA} parameters
+were unknown. Therefore, any apparent errors or biases in the estimates could
+be due to either poor performance of our method, or to real features of the
+posterior distribution. The latter type of error reflects on the suitability of
+the model, but does not invalidate the use of our method in cases where the
+parameters are more identifiable. Two retrospective experiments were peformed
+to disambiguate some of the observed errors: one where we ran a simulation with
+increased computational power to test for an increase in accuracy, and a second
+where we estimated parameters marginally to remove confounding from the other
+parameters in the joint posterior.}
+
+\del{We simulated three trees each under a variety of parameter values and ran
+the \software{netabc} program to estimate posterior distributions for the
+parameters.} The parameter values and priors used are listed in
+\cref{tab:abcexpt}. The tree kernel meta-parameters were set to $\gls{lambda} =
+0.3$ and $\gls{sigma} = 4$. The \gls{SMC} algorithm was run with 1000
+particles, five sampled datasets per particle, and \gls{alphaess} \del{(not to
+be confused with the \gls{BA} preferential attachment parameter, see
+\cref{subsec:adaptsmc})} set to 0.95. The algorithm was stopped when the
 acceptance rate of the \gls{MH} kernel dropped below 1.5\%, the same criterion
-used by \citeauthor{del2012adaptive}. Approximate marginal posterior densities
-for each parameter were calculated using the \software{density} function in
-\software{R} applied to the final weighted population of particles.
-{\color{red}\sout{Credible intervals were obtained for each parameter using the
-\software{HPDinterval} function in the \software{coda}
-package~\autocite{plummer2006coda}.}} {\color{blue}\uline{Posterior means
-obtained for each parameter using the \software{wtd.mean} function in the
-\software{Hmisc} package~\autocite{harrell2016hmisc}. Credible intervals were
-obtained using the \software{hpd} function in the \software{TeachingDemos}
-package~\autocite{snow2013teachingdemos}.}}
+used by \textcite{del2012adaptive}. \add{For visualization,} approximate
+marginal posterior densities for each parameter were calculated using the
+\software{density} function in \software{R} applied to the final weighted
+population of particles. \del{Credible intervals were obtained for each
+parameter using the \software{HPDinterval} function in the \software{coda}
+package~\autocite{plummer2006coda}.} \add{Posterior means obtained for each
+parameter using the \software{wtd.mean} function in the \software{Hmisc}
+package~\autocite{harrell2016hmisc}. Credible intervals were obtained using the
+\software{hpd} function in the \software{TeachingDemos}
+package~\autocite{snow2013teachingdemos} for \gls{alpha} \gls{I}, and \gls{N}, 
+and using our own implementation for discrete distributions for \gls{m}.}
 
@@
     \input{\tablepath/abc-expt}
-  \caption[Variables used in grid search experiments]
-  {
-    Variables and \gls{BA} parameter values used for \gls{ABC} validation
-    experiments. Trees were simulated under the test values, and
-    \gls{ABC} was used to re-estimate posterior distributions for the
-    \gls{BA} parameters without training.
+    \caption[
+        Parameter values used in simulation experiments to test accuracy of
+        \gls{BA} model fitting with \software{netabc}.
+    ]{
+        Parameter values used in simulation experiments to test accuracy of
+        \gls{BA} model fitting with \software{netabc}. Trees were simulated
+        under the test values, and \software{netabc} was used to estimate
+        posterior distributions on the \gls{BA} parameters for each simulated
+        tree. \software{Netabc} was na\"ive to the true parameter values.
     }
@@
 
-{\color{blue}\uline{
-To evaluate the effects of the true parameter values on the accuracy of the
-posterior mean estimates, we analysed the \gls{alpha} and \gls{I} parameters
-individually using \glspl{GLM}. The response variable was the error of the
-point estimate, and the predictor variables were the true values of
+\add{To evaluate the effects of the true parameter values on the accuracy of
+the posterior mean estimates, we analysed the \gls{alpha} and \gls{I}
+parameters individually using \glspl{GLM}. The response variable was the error
+of the point estimate, and the predictor variables were the true values of
 \gls{alpha}, \gls{I}, and \gls{m}. We did not test for differences across true
@@
 6$ (two \glspl{GLM} with three predictors each). Because there was clearly
-little to no identifiability of the \gls{N} and \gls{m} parameters with
-\gls{ABC} (see Results in next section), we did not construct \glspl{GLM} for
-those parameters. }}
-
-%\subsubsection*{Characterization of power-law exponent in Barab\'asi-Albert networks}
-%
-%Most studies of social network or transmission network
-%parameters~\autocite[e.g.][]{liljeros2001web, jones2003assessment,
-%schneeberger2004scale, brown2011transmission} report the coefficient
-%\gls{gamma} of the power law degree distribution. To make our results
-%comparable to previous work, we used simulated networks to investigate the
-%relationship between the \gls{BA} model parameters and \gls{gamma}. A network
-%was simulated for each combination of parameters listed in
-%\cref{tab:gammaexpt}. A power law distribution was fitted to the degree
-%distribution of each simulated network using the \software{fit\_power\_law}
-%function in \software{igraph} with the `R.mle' implementation. We fitted a
-%\gls{GLM} with Gamma-distributed errors and a log link function to the observed
-%distribution of \gls{gamma} values, with \gls{alpha}, \gls{m}, \gls{N}, and all
-%possible interaction terms as predictors.
-%
-%\begin{table}
-%  \centering
-%  \input{\tablepath/gamma-expt}
-%  \caption[\gls{BA} parameters used as input \gls{GLM} predicting $\gamma$]
-%  {
-%    \gls{BA} model parameters used as input to \gls{GLM} predicting power law
-%    exponent $\gamma$. One network was simulated with each combination of
-%    parameters, and $\gamma$ was calculated for each network. A \gls{GLM} with
-%    Gamma-distributed errors and a log link function was fit to the $\gamma$
-%    values with all parameters and interaction terms as predictors.
-%  }
-%  \label{tab:gammaexpt}
-%\end{table}
-
-Two further experiments were performed to address potential sources of error.
-To evaluate the effect of model misspecification in the case of heterogeneity
-among nodes, we generated a network where half the nodes were attached with
-power $\alpha$ = 0.5 and the other half with power $\alpha$ = 1.5. The other
-parameters for this network were $N$ = 5000, $I$ = 1000, and $m$ = 2. To
-investigate the effects of potential sampling
-bias~\autocite{karcher2016quantifying}, we simulated a transmission tree where
-the tips were sampled in a peer-driven fashion, rather than at random. That is,
-the probability to sample a node was twice as high if any of that node's
-network peers had already been sampled. The parameters of this network were $N$
-= 5000, $I$ = 2000, $m$ = 2, and $\alpha$ = 0.5.
+little to no identifiability of \gls{N} and \gls{m} with \gls{ABC} (see results
+in next section), we did not construct \glspl{GLM} for those parameters.}
+
+Two additional simulations were performed to address \del{potential sources of error}
+\add{the possible impact of two types of model misspecification}. To consider
+the effect of heterogeneity among nodes, we generated a network where half the
+nodes were attached with power $\gls{alpha} = 0.5$ and the other half with power
+$\gls{alpha} = 1.5$. The other parameters for this network were $\gls{N} = 5000$,
+$\gls{I} = 1000$, and $\gls{m} = 2$. To investigate the effects of potential
+sampling bias~\autocite{karcher2016quantifying}, we simulated a transmission
+tree where the tips were sampled in a peer-driven fashion, rather than at
+random. That is, the probability to sample a node was twice as high if any of
+that node's network peers had already been sampled. The parameters of this
+network were $\gls{N} = 5000$, $\gls{I} = 2000$, $\gls{m} = 2$, and
+$\gls{alpha} = 0.5$.
+
+\end{document}
 

### 2016-06-16 doc/thesis/methods_results.tex
Editing, midway through chp 2

@@
 
-Similar experiments were performed with the values shown in
-\cref{tab:kernelexpt}. The other three \gls{BA} parameters, namely \gls{N},
-\gls{m}, and \gls{I}, were each varied while holding the others fixed. The
-experiments for \gls{alpha}, \gls{m}, and \gls{N} were repeated with three
-different values of \gls{I}. All experiments were repeated with trees having
-three different numbers of tips. Kernel matrices were computed for all pairs of
-the meta-parameters \gls{lambda} = \sett{0.2, 0.3, 0.4} and \gls{sigma} =
-\sett{\nicefrac18, \nicefrac14, \nicefrac12, 1, 2, 4, 8}.
-
-\begin{landscape}
-\begin{table}[ht]
-  \centering
-  \input{\tablepath/kernel-expt}
-  \caption[Variables used in tree kernel simulation experiments]
-  {
-    Values of parameters and other variables used in tree kernel simulation
-    experiments. Each row corresponds to one of the \gls{BA} model parameters.
-    One kernel matrix was created for every combination of values except the
-    one indicated in the ``varied parameter'' column, which was varied when
-    producing simulated trees.
-  }
-  \label{tab:kernelexpt}
-\end{table}
-
-\begin{table}[ht]
-  \centering
-  \input{\tablepath/gridsearch-expt}
-  \caption[Variables used in grid search experiments]
-  {
-    Variables and \gls{BA} parameter values used for grid search experiments. 
-    Trees were simulated under the test values, and compared to a grid of trees
-    simulated under the grid values. Kernel scores were used to calculate point
-    estimates and credible intervals for the test values.
-  }
-  \label{tab:gridexpt}
-\end{table}
-\end{landscape}
-
 \begin{figure}[ht]
@@
     \includegraphics{kernel-expt.pdf}
-  \caption[Schematic of experiments investigating impact of BA model parameters
-           on tree shape.]{
-    Schematic of experiments designed to investigate the impact of variations
-    in BA model parameters on transmission tree shapes. The parameters of the
-    BA model were varied one at a time {\color{blue}\uline{while holding all
-    others fixed}}. Transmission trees were simulated under three different
-    values of each parameter, then compared pairwise using the tree kernel.
-    Classifiers were constructed for each parameter, and their accuracy was
-    evaluated by cross-validation. Kernel-PCA projections were used to visually
+    \caption[
+        Schematic of classifier experiments investigating identifiability of BA
+        model parameters from tree shapes.
+    ]{
+        Schematic of classifier experiments investigating identifiability of BA
+        model parameters from tree shapes. The parameters of the BA model were
+        varied one at a time \add{while holding all others fixed}. Transmission
+        trees were simulated under three different values of each
+        parameter\del{, then compared pairwise using the tree kernel.}
+        Classifiers were constructed for each parameter \add{based on three
+        tree shape statistics}, and their accuracy was evaluated by
+        cross-validation. Kernel-\gls{PCA} projections were used to visually
         examine the separation of the trees in the feature space defined by the
@@
 
+Similar experiments were performed with the values shown in
+\cref{tab:kernelexpt}. The other three \gls{BA} parameters, \gls{N}, \gls{m},
+and \gls{I}, were each varied while holding the others fixed. The experiments
+for \gls{alpha}, \gls{m}, and \gls{N} were repeated with three different values
+of \gls{I}. All experiments were repeated with trees having three different
+numbers of tips. Kernel matrices were computed for all pairs of the
+meta-parameters \gls{lambda} = \sett{0.2, 0.3, 0.4} and \gls{sigma} =
+\sett{\nicefrac18, \nicefrac14, \nicefrac12, 1, 2, 4, 8}.
+
+\begin{landscape}
+\begin{table}[ht]
+    \centering
+    \input{\tablepath/kernel-expt}
+    \caption[
+        Values of parameters and meta-parameters used in classifier experiments
+        to investigate identifiability of BA model parameters from tree shapes.
+    ]{
+        Values of parameters and meta-parameters used in classifier experiments
+        to investigate identifiability of BA model parameters from tree shapes.
+        Each row corresponds to one of the BA model parameters. One kernel
+        matrix was created for every combination of values except the one
+        indicated in the ``varied parameter'' column, which was varied when
+        producing simulated trees.
+    }
+    \label{tab:kernelexpt}
+\end{table}
+
+\begin{table}[ht]
+    \centering
+    \input{\tablepath/gridsearch-expt}
+    \caption[
+        Values of parameters and meta-parameters used in grid search
+        experiments to further investigate identifiability of BA model
+        parameters.
+    ]{
+        Values of parameters and meta-parameters used in grid search
+        experiments to further investigate identifiability of BA model
+        parameters. Trees were simulated under the test values, and compared to
+        a grid of trees simulated under the grid values. Kernel scores were
+        used to calculate point estimates and credible intervals for each
+        parameter, which were compared to the test values.
+    }
+  \label{tab:gridexpt}
+\end{table}
+\end{landscape}
+
 \subsubsection*{Grid search}
 
-{\color{blue}\uline{The previous experiment was an exploratory analysis
-intended to determine which of the \gls{BA} parameters were identifiable, and
-whether the tree kernel could potentially be used to distinguish different
-parameter values when all others were held fixed. In this experiment, which was
-still of an exploratory nature, we continued to consider one parameter at a
-time while fixing the other three. However, rather than checking for
-identifiability, we were now interested in quantifying the accuracy and
-precision of kernel score-based estimates. This was done by examining the
-distribution of kernel scores on a grid of parameter values, when trees
-simulated according to those values were compared with a single simulated test
-tree. This experiment did not involve the full \gls{ABC}-\gls{SMC} algorithm,
-nor did we continue to use Sackin's index or the \gls{nltt}.}}
+\add{The previous experiment was an exploratory analysis intended to determine
+which of the \gls{BA} parameters were identifiable, and whether the tree kernel
+could potentially be used to distinguish different parameter values when all
+others were held fixed. In this experiment, which was still of an exploratory
+nature, we continued to consider one parameter at a time while fixing the other
+three. However, rather than checking for identifiability, we were now
+interested in quantifying the accuracy and precision of kernel score-based
+estimates. This was done by examining the distribution of kernel scores on a
+grid of parameter values, when trees simulated according to those values were
+compared with a single simulated test tree.}
 
@@
 % latex table generated in R 3.2.3 by xtable 1.8-2 package
-% Tue Jun 14 16:56:16 2016
+% Thu Jun 16 16:24:45 2016
 \begin{table}[ht]

### 2016-06-18 doc/thesis/methods_results.Rnw
Most of the way through methods/results

@@
 simulated tree using our \software{treestat} binary. The trees were visualized
-using the \software{ape} package~\autocite{paradis2004ape}. {\color{blue}\uline{
-Both the tree kernel and the \gls{nltt} are pairwise statistics, and the
-\glspl{SVM} classifiers we used to investigate them operate on pairwise
-distance matrices.}} Our \software{treekernel} binary was used to calculate the
-value of the kernel for each pair of trees, with the meta-parameters set to
-$\lambda = 0.3$ and $\sigma = 4$. These values were stored in a symmetric 300
-$\times$ 300 kernel matrix.  Similarly, we computed the \gls{nltt} statistic
-between each pair of trees using our \software{treestat} binary, and stored
-them in a second $300 \times 300$ matrix.
+using the \software{ape} package~\autocite{paradis2004ape}. \add{Both the tree
+kernel and the \gls{nltt} are pairwise statistics, and the \glspl{SVR}
+classifiers we used to investigate them operate on pairwise distance matrices.}
+Our \software{treekernel} binary was used to calculate the value of the kernel
+for each pair of trees, with the meta-parameters set to $\lambda = 0.3$ and
+$\sigma = 4$. These values were stored in a symmetric 300 $\times$ 300 kernel
+matrix.  Similarly, we computed the \gls{nltt} statistic between each pair of
+trees using our \software{treestat} binary, and stored them in a second $300
+\times 300$ matrix.
 
@@
 
-Two additional simulations were performed to address \del{potential sources of error}
-\add{the possible impact of two types of model misspecification}. To consider
-the effect of heterogeneity among nodes, we generated a network where half the
-nodes were attached with power $\gls{alpha} = 0.5$ and the other half with power
-$\gls{alpha} = 1.5$. The other parameters for this network were $\gls{N} = 5000$,
-$\gls{I} = 1000$, and $\gls{m} = 2$. To investigate the effects of potential
-sampling bias~\autocite{karcher2016quantifying}, we simulated a transmission
-tree where the tips were sampled in a peer-driven fashion, rather than at
-random. That is, the probability to sample a node was twice as high if any of
-that node's network peers had already been sampled. The parameters of this
-network were $\gls{N} = 5000$, $\gls{I} = 2000$, $\gls{m} = 2$, and
-$\gls{alpha} = 0.5$.
-
-\end{document}
+Two further simulations were performed to address \del{potential sources of
+error} \add{the possible impact of two types of model misspecification}. To
+consider the effect of heterogeneity among nodes, we generated a network where
+half the nodes were attached with power $\gls{alpha} = 0.5$ and the other half
+with power $\gls{alpha} = 1.5$. The other parameters for this network were
+$\gls{N} = 5000$, $\gls{I} = 1000$, and $\gls{m} = 2$. To investigate the
+effects of potential sampling bias~\autocite{karcher2016quantifying}, we
+simulated a transmission tree where the tips were sampled in a peer-driven
+fashion, rather than at random. That is, the probability to sample a node was
+twice as high if any of that node's network peers had already been sampled. The
+parameters of this network were $\gls{N} = 5000$, $\gls{I} = 2000$, $\gls{m} =
+2$, and $\gls{alpha} = 0.5$.
+
+\add{To assess the impact of the \gls{SMC} settings on \software{netabc}'s
+accuracy, we ran \software{netabc} twice on the same simulated transmission
+tree. For the first run, the \gls{SMC} settings were the same as in the other
+simulations: 1000 particles, 5 simulated transmission trees per particle, and
+\gls{alphaess} = 0.95. The second run was performed with 2000 particles, 10
+simulated transmission trees per particle, and \gls{alphaess} = 0.99. To
+investigate the extent to which errors in the estimated \gls{BA} parameters
+were due to true features of the posterior, rather than an inaccurate \gls{ABC}
+approximation, we performed marginal estimation for one set of parameter
+values. Each combination of 1, 2, or 3 model parameters (14 combinations total)
+was fixed to their known values, and the remaining parameters were estimated
+with \software{netabc}. The parameter values were $\gls{alpha} = 0.0$, $\gls{m}
+= 2$, $\gls{I} = 2000$, and $\gls{N} = 5000$. These values were chosen because
+they had the highest error rate of all combinations tested when estimated
+jointly (see results in next section).}
 
@@
 
-    alpha.ratio.cor <- alpha.stats[,cor.test(int.tip.ratio, alpha)]
-    I.ratio.cor <- I.stats[,cor.test(int.tip.ratio, I)]
-    m.ratio.cor <- m.stats[,cor.test(int.tip.ratio, m)]
-    N.ratio.cor <- N.stats[,cor.test(int.tip.ratio, N)]
+    alpha.ratio.cor <- alpha.stats[,cor.test(int.tip.ratio, alpha, method="spearman")]
+    I.ratio.cor <- I.stats[,cor.test(int.tip.ratio, I, method="spearman")]
+    m.ratio.cor <- m.stats[,cor.test(int.tip.ratio, m, method="spearman")]
+    N.ratio.cor <- N.stats[,cor.test(int.tip.ratio, N, method="spearman")]
     stopifnot(alpha.ratio.cor$p.value < 1e-5)
@@
 measures tree imbalance, was significantly correlated with all four parameters
-    (for $\alpha$, $I$, $m$, and $N$ respectively: Spearman's rho =
-     \Sexpr{round(alpha.sackin.cor$estimate, 2)},
-     \Sexpr{round(I.sackin.cor$estimate, 2)},
-     \Sexpr{round(m.sackin.cor$estimate, 2)},
-     \Sexpr{round(N.sackin.cor$estimate, 2)};
+    (for \gls{alpha}, \gls{I}, \gls{m}, and \gls{N} respectively: Spearman's rho =
+    \Sexpr{round(alpha.sackin.cor[["estimate"]], 2)},
+     \Sexpr{round(I.sackin.cor[["estimate"]], 2)},
+     \Sexpr{round(m.sackin.cor[["estimate"]], 2)},
+     \Sexpr{round(N.sackin.cor[["estimate"]], 2)};
      $p$-values
-     $\Sexpr{pp(alpha.sackin.cor$p.value)}$,
-     $\Sexpr{pp(I.sackin.cor$p.value)}$,
-     $\Sexpr{pp(m.sackin.cor$p.value)}$,
-     $\Sexpr{pp(N.sackin.cor$p.value)}$).
+     $\Sexpr{pp(alpha.sackin.cor[["p.value"]])}$,
+     $\Sexpr{pp(I.sackin.cor[["p.value"]])}$,
+     $\Sexpr{pp(m.sackin.cor[["p.value"]])}$,
+     $\Sexpr{pp(N.sackin.cor[["p.value"]])}$).
 The ratio of internal to terminal branch lengths was negatively correlated with
@@
     \includegraphics[width=\textwidth]{kernel-alpha-tree.pdf}
-  \caption[Simulated transmission trees under three different values of BA parameter $\alpha$]{
-    Simulated transmission trees under three different values of BA parameter
-    $\alpha$. Epidemics were simulated on \gls{BA} networks of 5000 nodes, with
-    \gls{alpha} equal to 0.5, 1.0, or 1.5, until 1000 individuals were
-    infected. Transmission trees were created by sampling 500 infected nodes.
-    Higher \gls{alpha} values produced networks with a small number of
-    highly-connected nodes, resulting in highly unbalanced, ladder-like trees.
+    \caption[
+        Simulated transmission trees under three different values of
+        preferential attachment power (\gls{alpha}) parameter of \acrshort{BA}
+        model.
+    ]{
+        Simulated transmission trees under three different values of
+        preferential attachment power (\gls{alpha}) parameter of \acrshort{BA}
+        model. Epidemics were simulated on \gls{BA} networks of 5000 nodes,
+        with \gls{alpha} equal to 0.5, 1.0, or 1.5, until 1000 individuals were
+        infected. Transmission trees were created by randomly sampling 500
+        infected nodes. Higher \gls{alpha} values produced networks with a
+        small number of highly-connected nodes, resulting in highly unbalanced,
+        ladder-like trees.
     }
@@
 simulations with 500-tip trees and 1000 infected nodes. The three \gls{alpha}
-and \gls{I} values considered are well separated from each other in feature
-space. On the other hand, the three \gls{N} values overlap significantly, and
-the three \gls{m} values are virtually indistinguishable. Similar observations
-can be made for other values of \gls{I} and the number of tips
-(\cref{fig:alphakpca,fig:Nkpca,fig:Ikpca,fig:mkpca}). The values of \gls{I} and
-\gls{N} separated more clearly with larger numbers of tips, and in the case of
-\gls{N}, with larger epidemic sizes {\color{blue}(\cref{fig:Ikpca,fig:Nkpca})}.
+and \gls{I} values considered are well separated from each other in the feature
+space \add{mapped to by the tree kernel}. On the other hand, the three \gls{N}
+values overlap significantly, and the three \gls{m} values are virtually
+indistinguishable. Similar observations can be made for other values of \gls{I}
+and the number of tips (\cref{fig:alphakpca,fig:Nkpca,fig:Ikpca,fig:mkpca}).
+The values of \gls{I} and \gls{N} separated more clearly with larger numbers of
+tips, and in the case of \gls{N}, with larger epidemic sizes
+\add{(\cref{fig:Ikpca,fig:Nkpca})}.
 
@@
     \includegraphics{kernel-kpca.pdf}
-  \caption[Kernel-PCA projections of simulated trees under varying BA
-           parameter values.]{
-    Each parameter of the \gls{BA} model was individually varied to produce 300
-    simulated trees. Kernel matrices were formed from all pairwise kernel
-    scores among each set of 300 trees. The trees were projected onto the first
-    two principal components of the kernel matrix calculated using \gls{kPCA}.
-    All trees had 500 tips. The parameters not being varied were set to
-    \gls{alpha} = 1, \gls{I} = 1000, \gls{m} = 2, and \gls{N} = 5000. The tree
-    kernel meta-parameters were $\lambda = 0.3$ and $\sigma = 4$.
+    \caption[
+        Kernel-\gls{PCA} projections of simulated trees under varying \gls{BA}
+        parameter values.
+    ]{
+        Each parameter of the \gls{BA} model was individually varied to produce
+        300 simulated trees with 500 tips each. Kernel matrices were formed
+        from all pairwise kernel scores among each set of 300 trees. The trees
+        were projected onto the first two principal components of the kernel
+        matrix calculated using \gls{kPCA}. The other parameters, which were
+        not varied, were set to $\gls{alpha} = 1$, $\gls{I} = 1000$, $\gls{m} =
+        2$, and $\gls{N} = 5000$. The tree kernel meta-parameters were
+        $\gls{lambda} = 0.3$ and $\gls{sigma} = 4$.
   }
@@
 
-Accuracy of the \gls{kSVR} classifiers varied based on the parameter being
-tested (\cref{fig:rsquared}, left). Classifiers based on two other tree
-statistics, the \gls{nltt} and Sackin's index, generally exhibited worse
-performance than the tree kernel, although the magnitude of the disparity
-varied between the parameters (\cref{fig:rsquared}, centre and right). The
-results were largely robust to variations in the tree kernel meta-parameters
-$\lambda$ and $\sigma$, although accuracy varied between different epidemic and
-sampling scenarios
-(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}).
-
-When classifying $\alpha$, the \gls{kSVR} classifier had an average $R^2$ of 
+The accuracy of each classifier \add{on all \gls{BA} parameters is shown in}
+\del{varied based on the parameter being tested} \cref{fig:rsquared}.
+Classifiers based on \del{two other tree statistics,} the \gls{nltt} and
+Sackin's index generally exhibited worse performance than the tree kernel,
+although the magnitude of the disparity varied between the parameters
+(\cref{fig:rsquared}, centre and right). The results were largely robust to
+variations in the tree kernel meta-parameters \gls{lambda} and \gls{sigma},
+although accuracy varied between different epidemic and sampling scenarios
+(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}). \add{For all
+parameters except \gls{m}, the absolute number of tips in the tree had a much
+greater impact on accuracy than the proportion of infected individuals these
+tips represented. However, \gls{m}, both the number and proportion of sampled
+tips had a strong impact or the accuracy of the \gls{kSVR}
+(\cref{fig:mcrossv}).}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
+    \caption[
+        Cross-validation accuracy of kernel-SVR, nLTT-based SVR, and Sackin's
+        index regression classifiers for BA model parameters.
+    ]{
+        Cross-validation accuracy of kernel-SVR classifier (left), SVR
+        classifier using \gls{nltt} (centre), and linear regression using
+        Sackin's index (right) for \gls{BA} model parameters. Kernel
+        meta-parameters were set to $\gls{lambda} = 0.3$ and $\gls{sigma} = 4$.
+        Each point was calculated based on 300 simulated transmission trees
+        over networks with three different values of the parameter being
+        tested, assuming perfect knowledge of the other parameters. Vertical
+        lines are empirical 95\% confidence intervals based on 1000 two-fold
+        cross-validations. \add{The classifiers for \gls{I} were not evaluated
+        with 1000-tip trees, because one of the tested \gls{I} values was 500,
+        and it is not possible to sample a tree of size 1000 from 500 infected
+        individuals.}
+    }
+    \label{fig:rsquared}
+\end{figure}
+
+The \gls{kSVR} classifier for \gls{alpha} had an average $R^2$ of 
     \Sexpr{kernel.alpha[,round(mean(rsquared), 2)]},
@@
 about the mean for different tree and epidemic sizes. No classifier could
-accurately identify the $m$ parameter in any epidemic scenario, with average
-$R^2$ values of 
+accurately identify \gls{m} in any epidemic scenario, with average $R^2$ values
+of 
   \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for \gls{kSVR},
@@
 
-The accuracy of classifiers for $I$ varied significantly with the number of
-tips in the tree. For 100-tip trees, the average $R^2$ values were
+The accuracy of classifiers for $I$ \del{varied significantly
+with}\add{appeared to be strongly influenced by} the number of tips in the
+tree. For 100-tip trees, the average $R^2$ values were
   \Sexpr{kernel.I[ntip == 100, round(mean(rsquared), 2)]},
@@
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
-  \caption[Cross-validation accuracy of kernel-SVR, nLTT-based SVR, and
-  Sackin's index regression classifiers for BA model parameters.]{
-      Cross-validation accuracy of kernel-SVR classifier (left), SVR classifier
-      using \gls{nltt} (centre), and linear regression using Sackin's index
-      (right) for \gls{BA} model parameters. Kernel meta-parameters were set to
-      $\lambda = 0.3$ and $\sigma = 4$. Each point was calculated based on 300
-      simulated transmission trees over networks with three different values of
-      the parameter being tested, assuming perfect knowledge of the other
-      parameters. Vertical lines are empirical 95\% confidence intervals based
-      on 1000 two-fold cross-validations. {\color{blue}\uline{ The classifiers
-      for \gls{I} were not evaluated with 1000-tip trees, because one of the
-      tested \gls{I} values was 500, and it is not possible to sample a tree of
-      size 1000 from 500 infected individuals.}}
-  }
-  \label{fig:rsquared}
-\end{figure}
-
 \subsubsection*{Marginal parameter estimates with grid search}
@@
     tests <- d[,list(rsquared=cor.test(true_value, point.est)$estimate^2,
-                     rel.err=mean(abs(true_value - point.est) * 100 / diff(range(true_value))),
-                     abs.err=mean(abs(true_value - point.est))),
+                     grid.err=mean(abs(true_value - point.est) * 100 / diff(range(true_value))),
+                     abs.err=mean(abs(true_value - point.est)),
+                     rel.err=mean(abs(true_value - point.est) * 100 / true_value)),
                 by=parameter]
@@
 
-{\color{red}\sout{The accuracy of grid search estimates largely paralleled that
-of the \gls{kSVR} classifiers.}}\Cref{fig:gridest} shows point estimates and
-{\color{blue}\uline{50\%}} and 95\% highest density intervals for each of the
-\gls{BA} parameters, for one replicate experiment with 500-tip trees. Plots
-showing the point estimates for all replicates can be found in
+\del{The accuracy of grid search estimates largely paralleled that of the
+\gls{kSVR} classifiers.} \Cref{fig:gridest} shows point estimates and
+\add{50\%} and 95\% highest density intervals for each of the \gls{BA}
+parameters, for one replicate experiment with 500-tip trees. Plots showing the
+point estimates for all replicates can be found in
 \cref{fig:gridptalpha,fig:gridptI,fig:gridptm,fig:gridptN}. For all parameters
-except $m$, the error of point estimates was negatively correlated with the
+except \gls{m}, the error of point estimates was negatively correlated with the
 number of sampled tips in the tree (for \gls{alpha}, \gls{I}, and \gls{N}
-respectively: Spearman's $\rho$ = \Sexpr{tip.cor["alpha", round(estimate, 2)]},
-\Sexpr{tip.cor["I", round(estimate, 2)]}, \Sexpr{tip.cor["N", round(estimate,
-2)]}; $p$-values
+respectively: Spearman's $\rho$ = 
+    \Sexpr{tip.cor["alpha", round(estimate, 2)]},
+    \Sexpr{tip.cor["I", round(estimate, 2)]}, 
+    \Sexpr{tip.cor["N", round(estimate, 2)]}; 
+$p$-values
     $\Sexpr{tip.cor["alpha", pp(p.value)]}$,
@@
 
-{\color{blue}\uline{Across all replicates, $R^2$ values for the correlations
-between the estimated and true values were
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{gridsearch-example}
+    \caption[
+        Grid search estimates of \gls{BA} model parameters for one replicate
+        experiment with trees of size 500.
+    ]{
+        Grid search estimates of \gls{BA} model parameters for one replicate
+        experiment with trees of size 500. Point estimates (dots), 95\%
+        \glspl{HDI} (lines), and 50\% \glspl{HDI} (notches) for each \gls{BA}
+        model parameter, obtained using grid search. Networks and transmission
+        trees were simulated over a grid of values for each parameter while
+        holding the others fixed \add{to known values}. For a subset of the
+        grid values ($x$-axis), test networks and trees were created and
+        compared to each tree on the grid using the tree kernel. The kernel
+        scores along the grid were normalized to resemble a probability
+        distribution, from which the mode and highest density interval were
+        calculated.
+    } 
+    \label{fig:gridest}
+\end{figure}
+
+\add{Across all replicates, $R^2$ values for the correlations between the
+estimated and true values were
     $\Sexpr{tests["alpha", round(rsquared, 2)]}$,
@@
 representing
-    $\Sexpr{tests["alpha", round(rel.err)]}$\%,
-    $\Sexpr{tests["I", round(rel.err)]}$\%,
-    $\Sexpr{tests["m", round(rel.err)]}$\%, and
-    $\Sexpr{tests["N", round(rel.err)]}$\%
-of the respective grids.}
-
-\uline{Qualitatively, each of the parameters exhibited weak identifiability
-within particular sections of the grid} (\cref{fig:gridest}). \uline{The 50\%
-\glspl{HDI} for \gls{alpha} were nearly identical for the values $\gls{alpha}
-\leq 0.5$ (on average [X-X]) and for $\gls{alpha} \geq 1.5$ ([X-X]).
+    $\Sexpr{tests["alpha", round(grid.err)]}$\%,
+    $\Sexpr{tests["I", round(grid.err)]}$\%,
+    $\Sexpr{tests["m", round(grid.err)]}$\%, and
+    $\Sexpr{tests["N", round(grid.err)]}$\%
+of the respective grids. For \gls{I} and \gls{N}, relative errors may be more
+appropriate to consider. An overestimate of 100 individuals would be very
+misleading if the true population was only of size 100, but almost negligable
+in a population of size 5000. The mean relative errors for \gls{I} and \gls{N}
+respectively were
+    $\Sexpr{tests["I", round(rel.err)]}$\% and 
+    $\Sexpr{tests["N", round(rel.err)]}$\%.}
+
+\add{Qualitatively, \gls{alpha} and \gls{I} exhibited weak identifiability
+within particular sections of the grid (\cref{fig:gridest}). The 50\%
+\glspl{HDI} for \gls{alpha} were similar for the values $\gls{alpha}
+\leq 0.5$ (on average 
+    \Sexpr{d[parameter == "alpha" & true_value <= 0.5, round(mean(lower.50), 2)]} -
+    \Sexpr{d[parameter == "alpha" & true_value <= 0.5, round(mean(upper.50), 2)]})
+and for $\gls{alpha} \geq 1.5$ 
+    (\Sexpr{d[parameter == "alpha" & true_value >= 1.5, round(mean(lower.50), 2)]} -
+    \Sexpr{d[parameter == "alpha" & true_value >= 1.5, round(mean(upper.50), 2)]}).
 For \gls{I}, similar 50\% \gls{HPD} were observed for $\gls{I} \leq 1500$
-(average [X-X]) and for $\gls{I} > 3000$ ([X-X]). For \gls{m}, both the 50\%
-and 95\% \glspl{HDI} were generally identical for $\gls{m} > 2$ (average [X-X]
-for 50\% and [X-X] for 95\% \gls{HDI}). We did not observe a similar pattern
-for \gls{N}, however the boundaries of the \gls{HDI} did not always
-monotonically increase with the actual value.}
-
-\Cref{fig:gridalpha,fig:gridI,fig:gridm,fig:gridN} \uline{show kernel score
+(average 
+    [\Sexpr{d[parameter == "I" & true_value <= 1500, round(mean(lower.50))]} -
+     \Sexpr{d[parameter == "I" & true_value <= 1500, round(mean(upper.50))]}])
+and for $\gls{I} > 3000$ 
+    ([\Sexpr{d[parameter == "I" & true_value > 3000, round(mean(lower.50))]} -
+      \Sexpr{d[parameter == "I" & true_value > 3000, round(mean(upper.50))]}]).
+No similar patterns were observed for \gls{m} or \gls{N} (although the 50\%
+\glspl{HDI} appear to be identical for $\gls{m} > 2$ in \cref{fig:gridexpt},
+this was not consistent across replicates).}
+
+\add{\Cref{fig:gridalpha,fig:gridI,fig:gridm,fig:gridN} show kernel score
 distributions of kernel scores along the grid for each parameter. The
@@
 \gls{I}, whose score distributions exhibited a more or less rounded shape with
-the highest point near the true value. }}
-
-{\color{red}\sout{The \gls{alpha} parameter was the most accurately estimated,
+the highest point near the true value.}
+
+\del{The \gls{alpha} parameter was the most accurately estimated,
 with point estimates having an average deviation of 
@@
 and this difference was significant
-    (Wilcoxon rank-sum test, $p \Sexpr{pp(alpha.test$p.value, eq=TRUE)}$,
-}\cref{fig:gridptalpha}).
-\sout{These two values exhibited different qualitative behaviour than the other
-values in terms of the distribution of kernel scores along the grid
-}(\cref{fig:gridalpha}). \sout{In particular, there was a pronounced peak in scores
-around the true value, in contrast to the other values where the scores were
-flat around the true value. The effect was most obvious for the value
-\gls{alpha} = 1.25.}}
-
-%The average absolute error of the point estimates for \gls{I} was 
-%    \Sexpr{d["I", round(mean(error))]} individuals,
-%on a grid of 500 to 5000, and these errors differed between true values of
-%\gls{I}
-%    (one-way \gls{ANOVA}, $p \Sexpr{pp(av["I", p], eq=TRUE)}$).
-%The errors for $2000 \leq I \leq 3000$ were higher than those for the other
-%values
-%    (average errors
-%     \Sexpr{d["I", .SD[true_value >= 2000 & true_value <= 3000, round(mean(error))]]}
-%     for $2000 \leq I \leq 3000$ vs.
-%     \Sexpr{d["I", .SD[true_value < 2000 | true_value > 3000, round(mean(error))]]}
-%     for $I < 2000$ or $I > 3000$),
-%and this difference was significant
-%    (Wilcoxon rank-sum test, $p \Sexpr{pp(I.test$p.value, eq=TRUE)}$,
-%     \cref{fig:gridptI}).
-%Kernel score distributions for all test values exhibited a similar rounded
-%shape (\cref{fig:gridI}). 
-%
-%The average error for \gls{m} was
-%    \Sexpr{d["m", round(mean(error), 2)]} edges per vertex,
-%on a grid from 1 to 6. The error varied significantly between the true values
-%of \gls{m} 
-%    (one-way \gls{ANOVA}, $p \Sexpr{pp(av["m", p], eq=TRUE)}$).
-%Errors for the value \gls{m} = 1 were lower than the other values
-%    (average errors
-%    \Sexpr{d["m", .SD[true_value == 1, round(mean(error), 2)]]}
-%    for $m = 1$ vs.
-%    \Sexpr{d["m", .SD[true_value > 1, round(mean(error), 2)]]}
-%    for $m > 1$),
-%and this difference was significant
-%    (Wilcoxon rank-sum test, $p \Sexpr{pp(m.test$p.value, eq=TRUE)}$,
-%     \cref{fig:gridptm}).
-%The value $m = 1$ causes the network to take on a distinct shape relative to
-%higher \gls{m} values, namely a tree (\ie there are no cycles, see
-%\cref{subsec:treeshape}). The kernel score distribution had a peak at $m = 1$
-%when this was the true value, and a valley at $m = 1$ when the true value of
-%$m$ was greater that 1 (\cref{fig:gridm}).
-%
-%The average error for \gls{N} was 
-%    \Sexpr{d[parameter == "N", round(mean(error))]} individuals,
-%on a grid from 1000 to 15000, and was varied significantly with the true value
-%of \gls{N}
-%    (one-way \gls{ANOVA}, $p \Sexpr{pp(av["N", p], eq=TRUE)}$).
-%The errors were lower for $N \leq 3000$
-%    (average errors
-%     \Sexpr{d["N", .SD[true_value <= 3000, round(mean(error))]]}
-%     for $N \leq 3000$ vs.
-%     \Sexpr{d["N", .SD[true_value > 3000, round(mean(error))]]}
-%     for $N > 3000$),
-%and this difference was significant
-%    (Wilcoxon rank-sum test, $p \Sexpr{pp(N.test$p.value, eq=TRUE)}$,
-%     \cref{fig:gridptN}).
-%The kernel score distribution had a peak at $N = 1000$ when this was the true
-%value, and a valley there otherwise (\cref{fig:gridN}). Except for this valley,
-%the distributions were flat for $N > 3000$.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth]{gridsearch-example}
-  \caption[
-    Grid search estimates of \gls{BA} model parameters for one replicate
-    experiment with trees of size 500.
-  ]{
-    Grid search estimates of \gls{BA} model parameters for one replicate
-    experiment with trees of size 500. Point estimates (dots), 95\% highest
-    density intervals (lines), and 50\% highest density intervals (notches) for
-    each \gls{BA} model parameter, obtained using grid search. Networks and
-    transmission trees were simulated over a grid of values for each parameter
-    while holding the others fixed. For a subset of the grid values ($x$-axis),
-    test networks and trees were created and compared to each tree on the grid
-    using the tree kernel. The kernel scores along the grid were normalized to
-    resemble a probability distribution, from which the mode and highest
-    density interval were calculated. 
-  } 
-  \label{fig:gridest}
-\end{figure}
+    (Wilcoxon rank-sum test, $p \Sexpr{pp(alpha.test[["p.value"]], eq=TRUE)}$,
+\cref{fig:gridptalpha}). These two values exhibited different qualitative
+behaviour than the other values in terms of the distribution of kernel scores
+along the grid (\cref{fig:gridalpha}). In particular, there was a pronounced
+peak in scores around the true value, in contrast to the other values where the
+scores were flat around the true value. The effect was most obvious for the
+value \gls{alpha} = 1.25.}
 
@@
 
-\Cref{fig:abcptm2} shows {\color{red}\sout{\gls{MAP}}}
-{\color{blue}\uline{posterior mean}} point estimates of the BA model parameters
-obtained with ABC on simulated data. The estimates shown
-correspond only to the simulations where the $m$ parameter was set to 2,
-however the results for $m = 3$ and $m = 4$ were similar
+\Cref{fig:abcptm2} shows \del{MAP} \add{posterior mean} point estimates of the
+\gls{BA} model parameters obtained with \gls{ABC} on simulated data. The
+estimates shown correspond only to the simulations where \gls{m} was set to 2,
+however the results for $\gls{m} = 3$ and $\gls{m} = 4$ were similar
 (\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% \gls{HPD}
@@
 
-{\color{blue}
-\uline{Across all simulations, the median [\gls{IQR}] absolute errors of the parameter 
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{abc-point-estimate-m2}
+    \vspace{6pt}
+    \caption[
+        Posterior mean point estimates for \gls{BA} model parameters obtained
+        by running \software{netabc} on simulated data, for simulations with
+        $\gls{m} = 2$.
+    ]{
+        Posterior mean point estimates for \gls{BA} model parameters obtained
+        by running \software{netabc} on simulated data, for simulations with
+        $\gls{m} = 2$. Dashed lines indicate true values. (A) Estimates of
+        \gls{alpha} and \gls{I} which were varied in these simulations against
+        known values. (B) Estimates of \gls{m} and \gls{N} which were held
+        fixed in these simulations at the values $\gls{m} = 2$ and 
+        $\gls{N} = 5000$.
+    }
+    \label{fig:abcptm2}
+\end{figure}
+
+\begin{table*}[ht]
+    \centering
+    \input{\tablepath/abc-hpd.tex}
+    \caption[
+        Average posterior mean point estimates and 95\% \glspl{HDI} interval
+        widths for \gls{BA} model parameter estimates obtained with
+        \software{netabc} on simulated data.
+    ]{
+        Average posterior mean point estimates and 95\% \glspl{HDI} interval
+        widths for \gls{BA} model parameter estimates obtained with
+        \software{netabc} on simulated data. Three transmission trees were
+        simulated under each combination of the listed parameter values, and
+        the parameters were estimated with \gls{ABC} without training.
+    }
+    \label{tab:abchpd}
+\end{table*}
+
+\add{Across all simulations, the median [\gls{IQR}] absolute errors of the parameter 
 estimates obtained with \software{netabc} were
@@
     \Sexpr{d[,round(median(N_error) * 100 / 14500)]}\%
-of the regions of nonzero prior density. Average \gls{HPD} interval widths were
+of the regions of nonzero prior density. For \gls{I} and \gls{N}, relative
+errors were
+    \Sexpr{d[,round(median(I_error * 100 / true_I))]}\%
+    [\Sexpr{d[,round(quantile(I_error * 100 / true_I, 0.25))]} - 
+    \Sexpr{d[,round(quantile(I_error * 100 / true_I, 0.75))]}\%]
+and
+    \Sexpr{d[,round(median(N_error * 100 / true_N))]}\%
+    [\Sexpr{d[,round(quantile(N_error * 100 / true_N, 0.25))]} - 
+    \Sexpr{d[,round(quantile(N_error * 100 / true_N, 0.75))]}\%].
+Average \gls{HPD} interval widths were
     \Sexpr{d[,round(mean(alpha.upper.95 - alpha.lower.95), 2)]},
@@
 
-\uline{To analyse the effects of the true parameter values on the
+\add{To analyse the effects of the true parameter values on the
 accuracy our estimates of \gls{alpha} and \gls{I}, we fitted one \glspl{GLM}
@@
 the true parameter values as independent variables. Since the estimates of
-\gls{m} and \gls{N} were roughly equal across all simulations}
-(\cref{fig:abcptm2,fig:abcptm3,fig:abcptm4}), \uline{\glspl{GLM} were not
-fitted for these parameters. The estimated coefficients are shown in
-}\cref{tab:glmalpha,tab:glmI}. \uline{For the parameters \gls{alpha}, \gls{I},
-and \gls{N}, the \gls{GLM} were fitted using the inverse link function. That
-is, if $p$ is the true value of the parameter and $\hat{p}$ is a random
-variable representing our estimate of the parameter, the \gls{GLM} posits a
-relationship of the form }
+\gls{m} and \gls{N} were roughly equal across all simulations
+(\cref{fig:abcptm2,fig:abcptm3,fig:abcptm4}), \glspl{GLM} were not fitted for
+these parameters. The estimated coefficients are shown in
+\cref{tab:glmalpha,tab:glmI}. For the parameters \gls{alpha}, \gls{I}, and
+\gls{N}, the \gls{GLM} were fitted using the inverse link function. That is, if
+$p$ is the true value of the parameter and $\hat{p}$ is a random variable
+representing our estimate of the parameter, the \gls{GLM} posits a relationship
+of the form}
 \[
-    \E(|p-\hat{p}|) = (\beta_0 + \beta_\alpha \alpha + \beta_I I + \beta_m m)^{-1},
+    \madd{\E(|p-\hat{p}|) = (\beta_0 + \beta_\alpha \alpha + \beta_I I + \beta_m m)^{-1}},
 \]
-\uline{where the $\beta$'s are coefficients to be fitted. If the true value
+\add{where the $\beta$'s are coefficients to be fitted. If the true value
 \gls{alpha}, say, is increased by one, the \emph{inverse} of the expected
@@
 it means that the absolute error decreases as the true value of \gls{alpha}
-increases.}}
+increases.}
 
@@
 
-    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-                        "$\\alpha$ to true values of BA parameters.")
-    capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
+    capt.short <- paste("Parameters of a fitted \\gls{GLM} relating error in estimated",
+                        "\\gls{alpha} to true values of \\gls{BA} parameters.")
+    capt <- paste(capt.short, "\\gls{GLM} was fitted with a Gaussian distribution and",
                   "inverse link function. Coefficients are interpretable as",
@@
 
-    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-                        "$I$ to true values of BA parameters.")
-    capt <- paste(capt.short, "GLM was fitted with a gaussian distribution and",
+    capt.short <- paste("Parameters of a fitted \\gls{GLM} relating error in estimated",
+                        "\\gls{I} to true values of \\gls{BA} parameters.")
+    capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
                   "inverse link function. Coefficients are interpretable as",
@@
 
-{\color{blue}\uline{The \gls{GLM} analysis indicated that the error in estimates of
+\add{The \gls{GLM} analysis indicated that the error in estimates of
 \gls{alpha} decreased with larger true values of \gls{alpha} 
@@
     ($p \Sexpr{pp(alpha.glm[Parameter == "m", p], eq=TRUE)}$)
-but was not significatly affected by \gls{I}} (\cref{tab:glmalpha}). 
-\uline{Qualitatively, \gls{alpha} seemed to be only weakly identifiable between
-the values of 0 and 0.5} (\cref{fig:abcptm2}). \uline{The error in the
-estimated prevalence \gls{I} was slightly lower for smaller values of
-\gls{alpha}
+but was not significatly affected by \gls{I} (\cref{tab:glmalpha}).
+Qualitatively, \gls{alpha} seemed to be only weakly identifiable between the
+values of 0 and 0.5 (\cref{fig:abcptm2}). The error in the estimated prevalence
+\gls{I} was slightly lower for smaller values of \gls{alpha}
     ($p \Sexpr{pp(I.glm[Parameter == "alpha", p], eq=TRUE)}$)
@@
     ($p \Sexpr{pp(I.glm[Parameter == "I", p], eq=TRUE)}$),
-but was not significantly affected by the true value of \gls{m}} 
+but was not significantly affected by the true value of \gls{m}
 (\cref{tab:glmI}).}
 
-{\color{red}\sout{
-The accuracy of the parameter estimates obtained with ABC
-paralleled the results from the \gls{kSVR} classifier. Of the four parameters,
-$\alpha$ was the most accurately estimated, with point estimates having a
-median [IQR] absolute error of 
+\del{The accuracy of the parameter estimates obtained with ABC paralleled the
+results from the \gls{kSVR} classifier. Of the four parameters, $\alpha$ was
+the most accurately estimated, with point estimates having a median [IQR]
+absolute error of 
     \Sexpr{d[,round(median(alpha_error), 2)]} 
@@
 than those for the other values 
-    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(zero.alpha.test$p.value, 4))}$).
-Errors in estimating $\alpha$ 
-%also varied with the true value of $m$ just at
-%the threshold of statistical significance
-%    (one-way ANOVA, $p 
-%    \Sexpr{pp(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"], eq=TRUE)}$),
-did not vary across the true values of $m$ or $I$ (both
-one-way ANOVA).}
-
-\sout{Estimates for $I$ were relatively accurate, with point estimate errors of
+    (Wilcoxon rank-sum test, $p$ = 
+    $\Sexpr{latexSN(round(zero.alpha.test[["p.value"]], 4))}$).
+Errors in estimating $\alpha$ did not vary across the true values of $m$ or $I$
+(both one-way ANOVA).}
+
+\del{Estimates for $I$ were relatively accurate, with point estimate errors of
     \Sexpr{d[,round(median(I_error))]} 
@@
 at least 1
-    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(alpha.I.test$p.value, 4))}$)
+    (Wilcoxon rank-sum test, $p$ = 
+    $\Sexpr{latexSN(round(alpha.I.test[["p.value"]], 4))}$)
 and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
@@
 
-\sout{The $m$ parameter was estimated correctly in
+\del{The $m$ parameter was estimated correctly in
     \Sexpr{as.integer(m.tbl[1] * 100)} \%
@@
 
-\sout{Finally, the total number of nodes $N$ was consistently over-estimated by about
+\del{Finally, the total number of nodes $N$ was consistently over-estimated by about
 a factor of two (error \Sexpr{d[,round(median(N_error))]}
@@
 0.75))]}] individuals). No parameters influenced the accuracy of the $N$
-estimates (all one-way ANOVA).}}
+estimates (all one-way ANOVA).}
+
+<<supp_labels, echo=FALSE, results="asis">>=
+    #N <- 5000
+    #replicate <- 0
+    #lab <- NULL
+    #for (m in c(2, 3, 4)) {
+    #for (I in c(1000, 2000)) {
+    #for (alpha in c(0, 0.5, 1, 1.5)) {
+    #    lab <- c(lab, sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate))
+    #}
+    #}
+    #}
+    #lab <- paste(lab, collapse = ",")
+    worst <- d[alpha_error > quantile(alpha_error, 0.75) & I_error > quantile(I_error, 0.75),]
+    best <- d[alpha_error < quantile(alpha_error, 0.25) & I_error < quantile(I_error, 0.25),]
+    stopifnot(nrow(worst) == 1)
+    stopifnot(nrow(best) == 1)
+
+    worst.plot <- worst[,sprintf("%.1f_%.0f_%.0f_%.0f_%.0f", true_alpha, true_I, true_m, true_N, replicate)]
+    best.plot <- best[,sprintf("%.1f_%.0f_%.0f_%.0f_%.0f", true_alpha, true_I, true_m, true_N, replicate)]
+@
+
+The dispersion of the \gls{ABC} approximation to the posterior also varied
+between the parameters (\cref{tab:abchpd}). \gls{HPD} intervals around
+\gls{alpha} and \gls{I} were often narrow relative to the region of nonzero
+prior density, whereas the intervals for \gls{m} and \gls{N} were more widely
+dispersed. \Cref{fig:abclow1,fig:abclow2} \del{shows the distributions for one
+simulation.} \add{show one- and two-dimensional marginal distributions for a
+simulation with relatively low error, where the errors in \gls{alpha} and
+\gls{I} were each below the 25\% quantiles. The parameters for this simulation
+were 
+    $\gls{alpha} = \Sexpr{best[,true_alpha]}$,
+    $\gls{I} = \Sexpr{best[,true_I]}$,
+    $\gls{m} = \Sexpr{best[,true_m]}$, and
+    $\gls{N} = \Sexpr{best[,true_N]}$.
+\Cref{fig:abchigh1,abchigh2} show the equivalent marginals for a different
+simulation with relatively high error rates above the 75\% quantiles. The
+parameters were
+    $\gls{alpha} = \Sexpr{worst[,true_alpha]}$,
+    $\gls{I} = \Sexpr{worst[,true_I]}$,
+    $\gls{m} = \Sexpr{worst[,true_m]}$, and
+    $\gls{N} = \Sexpr{worst[,true_N]}$.
+The two-dimensional marginals some dependence between pairs of parameters,
+particularly \gls{I} and \gls{N} which show a diagonally shaped region of high
+posterior density.}
 
@@
     \centering
-  \includegraphics[width=\textwidth]{abc-point-estimate-m2}
-  \vspace{6pt}
+    \includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior", best.plot)}}.pdf}
     \caption[
-    Posterior mean point estimates for \gls{BA} model parameters obtained by
-    running \software{netabc} on simulated data, for simulations with \gls{m} =
-    2.
+        One-dimensional marginal posterior distributions of \gls{BA} model
+        parameters estimated with low error by \software{netabc} from a
+        simulated transmission tree.
     ]{
-    Posterior mean point estimates for \gls{BA} model parameters obtained by
-    running \software{netabc} on simulated data, for simulations with \gls{m} =
-    2. Dashed lines indicate true values. (A) Estimates of \gls{alpha} and
-    \gls{I} which were varied in these simulations against known values. (B)
-    Estimates of \gls{m} and \gls{N} which were held fixed in these simulations
-    at the values \gls{m} = 2 and \gls{N} = 5000.
-  }
-  \label{fig:abcptm2}
+        One-dimensional marginal posterior distributions of \gls{BA} model
+        parameters estimated with low error by \software{netabc} from a
+        simulated transmission tree. Dashed lines indicate true values, solid
+        lines indicate posterior means, and shaded areas show 95\%
+        \acrlong{HPD} intervals.
+    }
+    \label{fig:abclow1}
 \end{figure}
-
-\begin{table*}[ht]
-  \centering
-  \input{\tablepath/abc-hpd.tex}
-  \caption[
-      Average posterior mean point estimates and 95\% highest posterior density
-      (HPD) interval widths for BA model parameter estimates obtained with
-      ABC.
-  ]{
-      Average posterior mean point estimates and 95\% highest posterior density
-      (HPD) interval widths for BA model parameter estimates obtained with
-      ABC. Three transmission trees were simulated under each
-      combination of the listed parameter values, and the parameters were
-      estimated with ABC without training.
-  }
-  \label{tab:abchpd}
-\end{table*}
-
-<<supp_labels, echo=FALSE, results="asis">>=
-    N <- 5000
-    replicate <- 0
-    lab <- NULL
-    for (m in c(2, 3, 4)) {
-    for (I in c(1000, 2000)) {
-    for (alpha in c(0, 0.5, 1, 1.5)) {
-        lab <- c(lab, sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate))
-    }
-    }
-    }
-    lab <- paste(lab, collapse = ",")
-@
-
-The dispersion of the ABC approximation to the posterior also varied between
-the parameters, with narrower HPD intervals for the parameters with more
-accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows
-the distributions for one simulation. Equivalent plots for one replicate
-simulation with each studied parameter combination can be found in
-\cref{\Sexpr{lab}}. HPD intervals around $\alpha$ and $I$ were often narrow
-relative to the region of nonzero prior density, whereas the intervals for $m$
-and $N$ were more widely dispersed.
 
@@
     \centering
-    \includegraphics[width=\textwidth]{abc-posterior-example}
-  \vspace{6pt}
+    \includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior-2d", best.plot)}}.pdf}
     \caption[
-    Marginal posterior distributions of BA model parameters estimated
-    with ABC for a single simulated transmission tree.
+        Two-dimensional marginal posterior distributions of \gls{BA} model
+        parameters estimated with low error by \software{netabc} from a
+        simulated transmission tree.
     ]{
-    Marginal posterior distributions of BA model parameters estimated
-    with ABC for a single simulated transmission tree. Dotted
-    lines indicate true values, solid lines indicate posterior mean point
-    estimates, and shaded areas show 95\% highest posterior density intervals.
-  }
-  \label{fig:abcex}
+        Two-dimensional marginal posterior distributions of \gls{BA} model
+        parameters estimated with low error by \software{netabc} from a
+        simulated transmission tree. White circles indicate true values,
+        magenta diamonds indicate posterior means.
+    }
+    \label{fig:abclow2}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior", worst.plot)}}.pdf}
+    \caption[
+        One-dimensional marginal posterior distributions of \gls{BA} model
+        parameters estimated with high error by \software{netabc} from a
+        simulated transmission tree.
+    ]{
+        One-dimensional marginal posterior distributions of \gls{BA} model
+        parameters estimated with high error by \software{netabc} from a
+        simulated transmission tree. Dashed lines indicate true values, solid
+        lines indicate posterior means, and shaded areas show 95\%
+        \acrlong{HPD} intervals.
+    }
+    \label{fig:abchigh1}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior-2d", worst.plot)}}.pdf}
+    \caption[
+        Two-dimensional marginal posterior distributions of \gls{BA} model
+        parameters estimated with high error by \software{netabc} from a
+        simulated transmission tree.
+    ]{
+        Two-dimensional marginal posterior distributions of \gls{BA} model
+        parameters estimated with high error by \software{netabc} from a
+        simulated transmission tree. White circles indicate true values,
+        magenta diamonds indicate posterior means.
+    }
+    \label{fig:abchigh2}
 \end{figure}
@@
   d <- d[iter == max(iter)]
-  #d <- d[sample(1:nrow(d), prob=weight, replace=TRUE)]
   params <- c("alpha", "m", "N", "I")
@@
   p <- p[iter == max(iter)]
-  #p <- p[sample(1:nrow(p), prob=weight, replace=TRUE)]
   params <- c("alpha", "m", "N", "I")
@@
 the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
-other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The posterior mean
-[95\% HPD] estimates for each parameter were: 
-$\alpha$, 
+other half 1.5), with $\gls{m} = 2$, $\gls{N} = 5000$, and $\gls{I} = 1000$.
+The posterior mean [95\% \gls{HPD}] estimates for each parameter were: 
+\gls{alpha},
   \Sexpr{round(d["alpha", pmean], 2)} 
@@
    \Sexpr{round(d["alpha", hpd.max], 2)}];
-$I$,
+\gls{I},
   \Sexpr{d["I", round(pmean)]} 
@@
    \Sexpr{d["I", round(hpd.max)]}];
-$m$,
+\gls{m},
   \Sexpr{d["m", floor(pmean)]} 
@@
    \Sexpr{d["m", floor(hpd.max)]}];
-$N$,
+\gls{N},
   \Sexpr{d["N", round(pmean)]} 
@@
    \Sexpr{d["N", round(hpd.max)]}].
-The approximate posterior distributions for this simulation are shown in
-\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
-transmission tree in a peer-driven fashion, where the probability to sample a
-node was twice as high if one of its peers had already been sampled. The
-parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
-$I$ = 2000. The estimated values were
-$\alpha$, 
+The one-dimensional marginal approximate posterior distributions for this
+simulation are shown in \cref{fig:mixed}. To test the effect of sampling bias,
+we sampled one transmission tree in a peer-driven fashion, where the
+probability to sample a node was twice as high if one of its peers had already
+been sampled. The parameters for this experiment were $\gls{N} = 5000$,
+$\gls{m} = 2$, $\gls{alpha} = 0.5$, and $\gls{I} = 2000$. The estimated values
+were: \gls{alpha},
   \Sexpr{p["alpha", round(pmean, 2)]} 
@@
    \Sexpr{p["alpha", round(hpd.max, 2)]}];
-$I$,
+\gls{I},
   \Sexpr{p["I", floor(pmean)]} 
@@
    \Sexpr{p["I", floor(hpd.max)]}];
-$m$,
+\gls{m},
   \Sexpr{p["m", floor(pmean)]} 
@@
    \Sexpr{p["m", floor(hpd.max)]}];
-$N$,
+\gls{N},
   \Sexpr{p["N", floor(pmean)]} 
@@
    \Sexpr{p["N", floor(hpd.max)]}].
-The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
-of these results were in line with estimates obtained on other simulated
+The approximate posterior distributions are shown in \cref{fig:peerdriven}.
+Both of these results were in line with estimates obtained on other simulated
 datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
-$\alpha$ was somewhat lower than typical.
+\gls{alpha} was somewhat lower than typical.
 
@@
 
-\subsection{Methods}
-
-{\color{blue}\uline{
-The synthetic data experiments presented in the previous section demonstrated 
-that the \gls{alpha} and \gls{I} parameters, and to a lesser extent the \gls{N}
-parameter, had measurable impacts on tree shape and were identifiable with
-\gls{ABC}. Therefore, our next step was to use
-\software{netabc} to estimate the parameters of the \gls{BA} model for real
-world \gls{HIV} epidemics. }} Because the \gls{BA} model assumes a single
-connected contact network, it is most appropriate to apply to groups of
-individuals who are epidemiologically related. Therefore, we searched for
-published \gls{HIV} datasets which originated from existing clusters, either
-phylogenetically or geographically defined. In addition, we analysed an
+\subsection{Identification of suitable datasets}
+
+Because the \gls{BA} model assumes a single connected contact network, it is
+most appropriate to apply to groups of individuals who are epidemiologically
+related. Therefore, we searched for published \gls{HIV} datasets which
+originated from existing clusters, either phylogenetically or geographically
+defined. \add{To identify datasets fitting these criteria, we used the
+\software{Entrez} module in the \software{BioPython}
+library~\autocite{cock2009biopython} to identify all studies which were linked
+to at least 150 sequences of the same \gls{HIV} gene in GenBank (635 at the
+time when the data was collected). We manually curated a subset of these
+articles which, based on their title and abstract, appeared to have sampled one
+sequence per individual in a group likely to be epidemiologically related. For
+example, studies were excluded if they were investigating response to a
+particular drug, pregnant women or pediatric \gls{HIV} patients, intra-host
+evolution, or a multi-region or multi-country cohort. We acknowledge that our
+perusal of the complete set of articles was rather cursory, often limited to
+reading the title, and it is quite possible that we failed to identify other
+studies which would have been suitable to include. Each potential dataset was
+revisited, and those without sampling time annotation in GenBank were excluded.
+We also excluded studies where all sequences were sampled at the same
+timepoint, which was necessary because the method we used to time scale the
+tree requires non-contemporaneous tips.} The datasets are summarized in
+\cref{tab:data}. In addition \add{to the published data}, we analysed an
 in-house dataset sampled from \gls{HIV}-positive individuals in British
-Columbia, Canada (the ``BC data''). The datasets are summarized in
-\cref{tab:data}.
+Columbia, Canada. \add{For clarity, we will refer to each dataset by its risk
+group and location of origin in the text. For example, the
+\textcite{zetterberg2004two} data will be referred to as IDU/Estonia.}
 
@@
     \input{\tablepath/realdata-bc}
-  \caption[Characteristics of published HIV datasets analyzed with \software{netabc}.]
-  {
-    Characteristics of published HIV datasets analyzed with \software{netabc}. 
-    Abbreviations: MSM, men who have sex with men; HET, heterosexual; IDU,
-    injection drug users. The \textcite{novitsky2013phylogenetic,novitsky2014impact} data
-    were sampled from a primarily heterosexual risk environment but did not
-    explicitly exclude other risk factors.
+    \caption[
+        Characteristics of published \gls{HIV} datasets analyzed with
+        \software{netabc}.
+    ]{
+        Characteristics of published \gls{HIV} datasets analyzed with
+        \software{netabc}. Abbreviations: \acrshort{MSM}, \acrlong{MSM}; HET,
+        heterosexual; \acrshort{IDU}, \acrlong{IDU}. The HET data were sampled
+        from a primarily heterosexual risk environment but did not explicitly
+        exclude other risk factors. The number of sequences column indicates
+        how many sequences were included in our analysis; there may have been
+        additional sequences linked to the study which we excluded for various
+        reasons (see methods).
     }
@@
 
+\subsection{Methods}
+
 We downloaded all sequences associated with each published study from GenBank.
-{\color{blue}\uline{For the \textcite{niculescu2015recent} data, only sequences
-from \gls{IDU} (whose sequence identifiers included the letters ``DU'') were
-included in the analysis. \textcite{kao2011surveillance} found a very strong
-association in their study population between subtype and risk group - subtype
-B was almost always associated with \gls{MSM}, whereas \gls{IDU} were usually
-infected with a circulating recombinant form. Since there were many more
-subtype B sequences in their data than sequences of other subtypes, we
-restricted our analysis to the subtype B sequences and labelled this dataset as
-\gls{MSM}. Three datasets (\textcite{zetterberg2004two},
-\textcite{grabowski2014role}, and \textcite{mccormack2002early}) included both
-\textit{env} and \textit{gag} sequences. Each gene was analyzed separately to
-assess the robustness of \software{netabc} to the particular \gls{HIV} gene
-sequence used to estimate a transmission tree.}}
-
-{\color{red}\sout{For the \textcite{novitsky2014impact} data,}} Each
-\textit{env} sequence was aligned pairwise to the HXB2 reference sequence
-(GenBank accession number K03455), and the hypervariable regions were clipped
-out with \software{BioPython} version 1.66+~\autocite{cock2009biopython}.
-Sequences were multiply aligned using \software{MUSCLE} version 3.8.31
+\add{For the IDU/Romania data, only sequences from \gls{IDU} (whose sequence
+identifiers included the letters ``DU'') were included in the analysis.
+\textcite{kao2011surveillance} (MSM/Taiwan) found a very strong association in
+their study population between subtype and risk group - subtype B was almost
+always associated with \gls{MSM}, whereas \gls{IDU} were usually infected with
+a circulating recombinant form. Since there were many more subtype B sequences
+in their data than sequences of other subtypes, we restricted our analysis to
+the subtype B sequences and labelled this dataset as \gls{MSM}. Three datasets
+(IDU/Estonia, HET/Uganda, and HET/Malawi) included both \textit{env} and
+\textit{gag} sequences. Each gene was analyzed separately to assess the
+robustness of \software{netabc} to the particular \gls{HIV} gene sequence used
+to estimate a transmission tree.}
+
+\del{For the \textcite{novitsky2014impact} data,} Each \textit{env} sequence
+was aligned pairwise to the HXB2 reference sequence (GenBank accession number
+K03455), and the hypervariable regions were clipped out with
+\software{BioPython} version 1.66+~\autocite{cock2009biopython}. Sequences were
+multiply aligned using \software{MUSCLE} version 3.8.31
 \autocite{edgar2004muscle}, and alignments were manually inspected with
@@
 
-{\color{red}\sout{Three}} 
-{\color{blue}\uline{Six}} of the datasets 
-\autocite[][and the BC data]{li2015hiv,novitsky2014impact,grabowski2014role,little2014using}
-were initially larger than the others, containing 1265, 1299, 1026/915
-(\textit{env}/\textit{gag}), 648, and 7923 sequences respectively. To ensure
+\del{Three} \add{Six} of the datasets (MSM/Shanghai, HET/Botswana, HET/Uganda, 
+and MSM/USA) were initially larger than the others, containing 1265, 1299,
+1026/915 (\textit{env}/\textit{gag}), and 648 sequences respectively. To ensure
 that the analyses were comparable, we reduced these to a number of sequences
-similar to the smaller datasets. For the \citeauthor{li2015hiv} and BC
-datasets, we detected clusters of size 280 and 399 respectively using a
-patristic distance cutoff of 0.02 as described
-previously~\autocite{poon2015impact}. Only sequences within these clusters were
-carried forward. For the
-\textcite{grabowski2014role,novitsky2014impact,little2014using} datasets, no
+similar to the smaller datasets. For the MSM/Shanghai dataset, we detected a
+cluster of size 280 using a patristic distance cutoff of 0.02 as described
+previously~\autocite{poon2015impact}. Only sequences within this cluster were
+carried forward. For the HET/Uganda, HET/Botswana, and MSM/USA datasets, no
 large clusters were detected using the same cutoff, so we analysed subsets of
-size 225, 180, and 180 respectively. {\color{blue}\uline{The subset of the
-\citeauthor{grabowski2014role} data was chosen by eye such that the individuals
-were monophyletic in both the \textit{gag} and \textit{env} trees. The other
-subsets were arbitrarily chosen subtrees from phylogenies of the complete
-datasets.}}
-
-For all datasets, we used the priors $\alpha$ $\sim$ Uniform(0, 2) and $N$ and
-$I$ jointly uniform on the region \{$n \leq N \leq 10000$, $n \leq I \leq
-10000$, $I \leq N$\}, where $n$ is the number of tips in the tree (see
-\cref{tab:data}). Since the value $m = 1$ produces networks with no cycles,
-which we considered fairly implausible, we ran one analysis with the prior $m
-\sim$ DiscreteUniform(1, 5), and one with the prior $m \sim$ DiscreteUniform(2,
-5). The other parameters to the SMC algorithm were the same as used for the
-simulation experiments, except that we used 10000 particles instead of 1000 to
-increase the accuracy of the estimated posterior. This was computationally
-feasible due to the small number of runs required for this analysis.
-
-Empirical studies of contact networks often report the exponent $\gamma$ of the
-power law degree distribution. To compare our results to the literature, we
+size 225, 180, and 180 respectively. \add{The subset of the HET/Uganda data was
+chosen by eye such that the individuals were monophyletic in both the
+\textit{gag} and \textit{env} trees. The other subsets were arbitrarily chosen
+subtrees from phylogenies of the complete datasets.}
+
+For all datasets, we used the priors $\gls{alpha} \sim \Uniform(0, 2)$ and
+\gls{N} and \gls{I} jointly uniform on the region $\{n \leq \gls{N} \leq
+10000$, $n \leq \gls{I} \leq 10000$, $\gls{I} \leq \gls{N}\}$, where $n$ is the
+number of tips in the tree (see \cref{tab:data}). Since the value $\gls{m} = 1$
+produces networks with no cycles, which we considered fairly implausible, we
+ran one analysis with the prior $\gls{m} \sim \DiscreteUniform(1, 5)$, and one
+with the prior $\gls{m} \sim \DiscreteUniform(2, 5)$. The other parameters to
+the SMC algorithm were the same as used for the simulation experiments, except
+that we used 10000 particles instead of 1000 to increase the accuracy of the
+estimated posterior. This was computationally feasible due to the small number
+of runs required for this analysis.
+
+Empirical studies of contact networks often report the exponent \gls{gamma} of
+the power law degree distribution. To compare our results to the literature, we
 simulated 100 networks each according to the posterior mean parameter estimates
-obtained for each investigated dataset. The power-law exponent $\gamma$ was
-calculated for each network using the \software{fit\_power\_law} function in
-\software{igraph}, with the `R.mle' implementation. The median of the 100
-$\gamma$ values was taken as a point estimate for the associated dataset.
+obtained for each investigated dataset. \add{Although the \gls{BA} model does
+not produce power law networks except when $\gls{alpha} = 1$, simulations show
+that the power law fit still captures the slope of the degree distribution
+reasonably well (\cref{fig:powerlaw}).} \gls{gamma} was calculated for each
+network using the \software{fit\_power\_law} function in \software{igraph},
+with the `R.mle' implementation. The median of the 100 \gls{gamma} values was
+taken as a point estimate for the associated dataset.
 
@@
 
-We applied \software{netabc} to {\color{red}\sout{five}}
-{\color{blue}\uline{ten}} published HIV datasets, {\color{blue}\uline{plus
-one in-house, unpublished dataset}} (\cref{tab:data}).
-{\color{red}\sout{and found substantial heterogeneity among the parameter
-estimates}} {\color{blue}\uline{Posterior mean point estimates and 50\% and
-95\% \glspl{HPD} for each parameter are shown in} \cref{fig:abchpd}, \uline{and
-marginal posterior distributions in}
-\cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.
-\Cref{fig:abchpdm2} \uline{shows point estimates and \glspl{HPD} obtained when
+\del{We applied \software{netabc} to five published HIV datasets (\cref{tab:data}).
+and found substantial heterogeneity among the parameter
+estimates.} \add{Posterior mean point estimates and 50\% and 95\% \glspl{HPD}
+for each parameter are shown in \cref{fig:abchpd}.} \del{and marginal posterior
+distributions in \cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}}.
+\add{\Cref{fig:abchpdm2} shows point estimates and \glspl{HPD} obtained when
 the value $\gls{m} = 1$ was disallowed by the prior. Since the results
@@
 all results discussed henceforth are for the prior $\gls{m} \sim
-\DiscreteUniform(1, 5)$ unless otherwise stated. For clarity, datasets will be
-referred to as their risk group and location. For example, \gls{IDU}/Estonia
-refers to the} \textcite{zetterberg2004two} \uline{dataset.}}
-
-{\color{red}\sout{
-Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had
+\DiscreteUniform(1, 5)$ unless otherwise stated.}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{realdata-hpd-bc}
+    \caption[
+        Posterior means and 50\%/95\% \gls{HPD} intervals for parameters of the
+        \gls{BA} network model, fitted to eleven
+        \gls{HIV} datasets with \software{netabc}.
+    ]{
+        Posterior means (points), 50\% (notches), and 95\% (lines) \gls{HPD}
+        intervals for parameters of the \gls{BA} network model, fitted to
+        eleven \gls{HIV} datasets with \software{netabc}. Legend labels
+        indicate risk group and country of origin. Abbreviations:
+        \acrshort{IDU}, \acrlong{IDU}; \acrshort{MSM}, \acrlong{MSM}; HET,
+        heterosexual. Note that posterior means can fall outside of \glspl{HPD} 
+        if the distribution is diffuse.
+    }
+    \label{fig:abchpd}
+\end{figure}
+
+\del{Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had
 estimated $\alpha$ values near unity for the prior allowing $m = 1$ (posterior
@@
    [\Sexpr{low2["novitsky2014", round(alpha, 2)]} -
-    \Sexpr{up2["novitsky2014", round(alpha, 2)]}] for $m \geq 2$). }}
-
-{\color{blue}\uline{Posterior mean point estimates for the \acrlong{PA} power
-\gls{alpha} ranged from 
+    \Sexpr{up2["novitsky2014", round(alpha, 2)]}] for $m \geq 2$).}
+
+\add{Posterior mean point estimates for the \acrlong{PA} power \gls{alpha}
+ranged from 
     \Sexpr{pmean1["zetterberg2004", round(alpha, 2)]}
@@
 where both \textit{gag} and \textit{env} were sequenced, but the estimates for
-\gls{alpha} did not change appreciably between the two genes} (TODO figure).
-\uline{There was a large amount of uncertainty associated with these
-estimates. 95\% \gls{HPD} were very wide for most datasets, often encompassing
-almost the entire range from 0 to 1} (\cref{fig:abchpd}). \uline{For all the
-datasets except HET/Malawi and MSM/Beijing, the posterior mean was either
-outside, or very close to the border of, the 50\% \gls{HPD}. This indicates
-that the posterior distributions were diffuse with heavy tails, rather than
-having most of their mass around the mode.}}
+\gls{alpha} did not change appreciably between the two genes (\cref{fig:genes}).
+There was a large amount of uncertainty associated with these estimates. 95\%
+\gls{HPD} were very wide for most datasets, often encompassing almost the
+entire range from 0 to 1 (\cref{fig:abchpd}). For all the datasets except
+HET/Malawi and MSM/Beijing, the posterior mean was either outside, or very
+close to the border of, the 50\% \gls{HPD}. This indicates that the posterior
+distributions were diffuse with heavy tails, rather than having most of their
+mass around the mode.}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{realdata-genes}
+    \caption[
+        TODO
+    ]{
+        TODO
+    }
+    \label{fig:genes}
+\end{figure}
+
+\end{document}
 
@@
 
-\begin{figure*}[ht]
-  \centering
-  \includegraphics{realdata-hpd-bc}
-  \caption[
-      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
-      parameters of the BA network model, fitted to six HIV datasets
-      with \software{netabc}.]
-  {
-      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
-      parameters of the BA network model, fitted to six HIV datasets with
-      \software{netabc}. Legend labels indicate risk group and country of
-      origin. Abbreviations: IDU, injection drug users; MSM, men who have sex
-      with men; HET, heterosexual.
-  }
-  \label{fig:abchpd}
-\end{figure*}
-
 To make our analyses comparable to the existing network literature, we

### 2016-06-20 doc/thesis/methods_results.Rnw
Finished I stuff finally omg

@@
         \State infect a node $v$ at random, updating \gls{gS}, \gls{gI},
-                 \gls{gD}, \gls{beta}, and \gls{gamma}
+                 \gls{gD}, \gls{beta}, and \gls{nu}
         \State $\gls{T} \gets$ a single node with label $1$
@@
 \subsection{Why study the \acrlong{BA} model?}
+\label{subsec:whyba}
 
@@
 with \software{netabc}. The parameter values were $\gls{alpha} = 0.0$, $\gls{m}
-= 2$, $\gls{I} = 2000$, and $\gls{N} = 5000$. These values were chosen because
-they had the highest error rate of all combinations tested when estimated
-jointly (see results in next section).}
+= 2$, $\gls{I} = 2000$, and $\gls{N} = 5000$.}
 
@@
 
+\add{\Cref{fig:marginal} shows the effect of performing marginal \gls{ABC}
+estimation of each of the \gls{BA} parameters on the same simulated
+transmission tree. The estimates of \gls{m} are apparently unaffected by
+marginalizing out the other parameters, coroborating the prevous experiments'
+findings that \gls{m} is not an identifiable parameter from scaled tree shapes.
+The estimates of \gls{alpha} and \gls{N} were weakly improved in marginal
+estimation, which is consistent with the result from grid search that all
+\gls{N} values and \gls{alpha} values below 0.5 are weakly identifiable. On the
+other hand, both the accuracy and precision (\ie width of the \gls{HPD}
+intervals) of the estimates of \gls{I} were substantially improved when \gls{N}
+and at least one other variable were marginalized out. Neither the accuracy nor
+the precision of any parameter estimates appeared to improve when more
+computing power was used for \gls{SMC} (\cref{fig:moresmc}).}
+
 \section{Application to real world HIV data}
@@
 
-\subsection{Identification of suitable datasets}
+\subsection{Methods}
 
@@
 
-\subsection{Methods}
-
 We downloaded all sequences associated with each published study from GenBank.
@@
 identifiers included the letters ``DU'') were included in the analysis.
-\textcite{kao2011surveillance} (MSM/Taiwan) found a very strong association in
-their study population between subtype and risk group - subtype B was almost
-always associated with \gls{MSM}, whereas \gls{IDU} were usually infected with
-a circulating recombinant form. Since there were many more subtype B sequences
-in their data than sequences of other subtypes, we restricted our analysis to
-the subtype B sequences and labelled this dataset as \gls{MSM}. Three datasets
+\textcite{kao2011surveillance} (MSM/Taiwan) found a strong association in their
+study population between subtype and risk group - subtype B was most often
+associated with \gls{MSM}, whereas \gls{IDU} were usually infected with a
+circulating recombinant form. Since there were many more subtype B sequences in
+their data than sequences of other subtypes, we restricted our analysis to the
+subtype B sequences and labelled this dataset as \gls{MSM}. Three datasets
 (IDU/Estonia, HET/Uganda, and HET/Malawi) included both \textit{env} and
@@
 Empirical studies of contact networks often report the exponent \gls{gamma} of
-the power law degree distribution. To compare our results to the literature, we
-simulated 100 networks each according to the posterior mean parameter estimates
-obtained for each investigated dataset. \add{Although the \gls{BA} model does
-not produce power law networks except when $\gls{alpha} = 1$, simulations show
-that the power law fit still captures the slope of the degree distribution
-reasonably well (\cref{fig:powerlaw}).} \gls{gamma} was calculated for each
-network using the \software{fit\_power\_law} function in \software{igraph},
-with the `R.mle' implementation. The median of the 100 \gls{gamma} values was
-taken as a point estimate for the associated dataset.
+the power law degree distribution~\autocite{colgate1989risk, liljeros2001web,
+schneeberger2004scale, clemenccon2015statistical, rothenberg2007large,
+brown2011transmission}. Although the \gls{BA} model does not produce networks
+with true power law degree distributions except when $\gls{alpha} = 1$, the
+power law still provides a reasonable approximation to the slope when
+$\gls{alpha} \neq 1$ (\cref{fig:powerlaw}).  To compare our results to
+\add{existing network} literature we simulated 100 networks each according to
+the posterior mean parameter estimates obtained for each investigated dataset.
+\add{Although the \gls{BA} model does not produce power law networks except
+when $\gls{alpha} = 1$, simulations show that the power law fit still captures
+the slope of the degree distribution reasonably well (\cref{fig:powerlaw}).}
+\gls{gamma} was calculated for each network using the
+\software{fit\_power\_law} function in \software{igraph}, with the `R.mle'
+implementation. The median of the 100 \gls{gamma} values was taken as a point
+estimate for the associated dataset.
 
@@
   stopifnot(pmean1[,dataset[which.min(alpha)]] == "cuevas2009")
+
+  stopifnot(pmean1[,dataset[which.max(I)]] == "zetterberg2004")
+  stopifnot(pmean1[,dataset[which.min(I)]] == "novitsky2014")
+  stopifnot(pmean1[dataset != "novitsky2014", dataset[which.min(I)]] == "bctree")
+
+  stopifnot(pmean1[,cor.test(ntip, I)$p.value, method="spearman"] > 0.05)
 @
@@
     ]{
-        Posterior means (points), 50\% (notches), and 95\% (lines) \gls{HPD}
-        intervals for parameters of the \gls{BA} network model, fitted to
-        eleven \gls{HIV} datasets with \software{netabc}. Legend labels
-        indicate risk group and country of origin. Abbreviations:
+        Posterior means (points), 50\% \gls{HPD} intervals (notches), and 95\%
+        (lines) \gls{HPD} intervals for parameters of the \gls{BA} network
+        model, fitted to eleven \gls{HIV} datasets with \software{netabc}.
+        Legend labels indicate risk group and country of origin. Abbreviations:
         \acrshort{IDU}, \acrlong{IDU}; \acrshort{MSM}, \acrlong{MSM}; HET,
@@
 \add{Posterior mean point estimates for the \acrlong{PA} power \gls{alpha}
-ranged from 
+were all sub-linear, ranging from 
     \Sexpr{pmean1["zetterberg2004", round(alpha, 2)]}
@@
 distributions were diffuse with heavy tails, rather than having most of their
-mass around the mode.}
+mass around the mode. When the value $\gls{m} = 1$ was disallowed by the prior, 
+TODO}
 
@@
     \caption[
-        TODO
+        Posterior means and 50\%/95\% \gls{HPD} intervals for parameters of the
+        \gls{BA} network model, fitted to three \gls{HIV} datasets where both
+        \textit{gag} and \textit{env} genes were sequenced.
     ]{
-        TODO
+        Posterior means (points) 50\% \gls{HPD} intervals (notches), and 95\%
+        \gls{HPD} intervals (lines) for parameters of the \gls{BA} network model,
+        fitted to three \gls{HIV} datasets where both \textit{gag} and
+        \textit{env} genes were sequenced.
     }
@@
 
-\end{document}
-
-{\color{red}\sout{For all the datasets except \citeauthor{novitsky2014impact},
+\del{For all the datasets except \citeauthor{novitsky2014impact},
 estimated values of $I$ were below 2000 when $m = 1$ was allowed, with
@@
 prior on $m$, with slightly narrower HPD intervals (compare
-\cref{fig:abchpd,fig:abchpdm2}).}}
-
-The posterior mean of $m$ was equal to 1 for all but the
+\cref{fig:abchpd,fig:abchpdm2}).}
+
+\add{For all but the HET/Botswana data, the posterior mean estimates for
+the prevalence \gls{I} were between
+    \Sexpr{pmean1["zetterberg2004", round(I)]}
+(IDU/Estonia) and
+    \Sexpr{pmean1["bctree", round(I)]}
+(IDU/Canada). The HET/Botswana data had a much higher estimated \gls{I} value
+    (\Sexpr{pmean1["novitsky2014", round(I)]})
+than the other datasets, with a very wide 95\% \gls{HPD} interval covering
+almost the entire prior region (\cref{fig:abchpd}). There was no significant
+correlation between the number of sequences in the tree and the estimated
+prevalence (Spearman correlation, 
+    $p \Sexpr{pp(pmean1[,cor.test(ntip, I)[["p.value"]], method="spearman"], eq=TRUE)}$).
+This indicates that the higher prevalence estimates were not simply due to
+increased sampling density. When both \textit{gag} and \textit{env} sequences
+were analyzed, the estimates from the \textit{env} data were substantially
+higher (\cref{fig:genes}). The ratios of terminal to internal branch lengths
+were higher for the \textit{env} trees than the \textit{gag} trees (average X
+for \textit{env} and X for \textit{gag}). As with \gls{alpha}, many of the
+posterior means were outside the 50\% \gls{HPD} intervals, suggesting diffuse
+posterior distributions. The values of \gls{I} were largely robust to
+restricting the prior to $\gls{m} > 1$ (compare
+\cref{fig:abchpd,fig:abchdpm2}), although TODO}
+
+\del{The posterior mean of $m$ was equal to 1 for all but the
 \citeauthor{novitsky2014impact} data, when this value was allowed. However, the
@@
   (\Sexpr{pmean1["li2015", floor(N)]})
-when $m = 1$ was allowed.
-
-To make our analyses comparable to the existing network literature, we
-estimated values of the power law exponent $\gamma$ for each of the datasets 
-investigated, using the posterior median estimates with the prior allowing
-\gls{m} = 1. The results are summarized in \cref{tab:gamma}. All of the
-estimated exponents were in the range $2 \leq \gamma \leq 2.5$, which is on the
-lower end of the range $2 \leq \gamma \leq 4$ reported in the literature.
+when $m = 1$ was allowed.}
+
+\add{The posterior means of \gls{m} were equal to one for 
+  \Sexpr{replace_number(pmean1[,sum(m == 1)])}
+of the datasets analysed. The widths of the 95\% \gls{HPD} varied from 0 (all
+the mass on the estimated value) to 5 (the entire prior region). When the value
+$\gls{m} = 1$ was disallowed by the prior, TODO.}
+
+\add{Estimates of \gls{N} were  mostly uninformative, with very similar estimates
+for all datasets (mean
+    \Sexpr{pmean1[,round(mean(N))]},
+range
+    \Sexpr{pmean1[,round(min(N))]} - \Sexpr{pmean1[,round(max(N))]}).
+This was similar to the pattern observed for the synthetic data, where the
+posterior mean always fell around the upper two-thirds mark of the region
+allowed by the prior. The results did not change appreciably (TODO) when the
+value $\gls{m} = 1$ was assigned zero prior probability.}
+
+To compare our results to the existing network literature, we estimated
+values of the power law exponent \gls{gamma} for each of the datasets
+investigated, using the posterior median estimates \del{with the prior allowing
+$\gls{m} = 1$}. Because the estimates of \gls{N} were uninformative, we
+simulated another set of networks with the estimated \gls{I} value, rather than
+\gls{N}, as the number of nodes. The results are summarized in
+\cref{tab:gamma}. All of the estimated exponents were in the range $2 \leq
+\gls{gamma} \leq 2.5$, which is on the lower end of the range $2 \leq
+\gls{gamma} \leq 4$ reported in the literature.
 

### 2016-06-20 doc/thesis/discussion.Rnw

@@
 identifiable within their own band. However, when \gls{N} is fixed at 5000, the
-derivatives vary substantially along the \gls{I}-axis, which explains why the
-grid search estimates of \gls{I} were accurate and unbiased
-(\cref{fig:gridptI}).
+derivatives vary substantially along the \gls{I}-axis, which explains why
+the grid search estimates of \gls{I} were accurate and unbiased
+(\cref{fig:gridptI}). \add{We also note the resemblance of the contour surface
+of \cref{fig:derivs} to the two-dimensional marginal posterior distributions on
+\gls{I} and \gls{N} obtained with simulated data
+(\cref{fig:abclow2,fig:abchigh2}). However, while fixing \gls{N} and any other
+parameter resulted in much more accurate estimates of \gls{I}, fixing \gls{N}
+alone did not (\cref{fig:marginal}). Therefore, there may be more complex
+parameter dependence in the \gls{BA} model which we have not considered here.}
 
@@
 literature~\autocite{liljeros2001web, schneeberger2004scale, colgate1989risk,
-brown2011transmission}. We note however that \textcite{jones2003assessment}
-estimated \gls{gamma} values greater than four for some datasets, in one case
-as high as 17, indicating that consideration of a wider range of permitted
-\gls{gamma} values may be warranted.
+clemenccon2015statistical, brown2011transmission}. We note however that
+\textcite{jones2003assessment} estimated \gls{gamma} values greater than four
+for some datasets, in one case as high as 17, indicating that consideration of
+a wider range of permitted \gls{gamma} values may be warranted.
 
@@
 this range under both priors, as were the estimates for the
-\textcite{wang2015targeting} data and the BC data when \gls{m} = 1 was
+\textcite{wang2015targeting} data and the BC data when $\gls{m} = 1$ was
 disallowed by the prior. The dataset investigated by
@@
     X%\S%expr{d[m_min == 2 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]},
-depending on the prior on \gls{m}; \citeauthor{de2007preferential} estimated
+depending on the prior on \gls{m}; \textcite{de2007preferential} estimated
 0.54, 0.57, and 0.29 for 3-year, 5-year, and lifetime partnership networks
@@
 
-One somewhat surprising result was the difference between parameter estimates 
-for the \textcite{li2015hiv} and \textcite{wang2015targeting} datasets. Both
-groups studied cohorts of acutely infected \gls{MSM} in major Chinese cities
-(Shanghai and Beijing respectively); yet, the \citeauthor{li2015hiv} data was 
-estimated to have a lower \acrlong{PA} power and larger infected population
-than the \textcite{wang2015targeting} data\ldots
+\add{Unfortunately, the sub-linear region for \gls{alpha} identified by both
+\textcite{de2007preferential} and \software{netabc} is also the region of
+poorest identifiability (\cref{fig:gridest,fig:abcptm2}). This was reflected in
+the high level of uncertainty in the estimates, with most 95\% \gls{HPD}
+intervals covering the majority of the range [0, 1]. The value $\gls{alpha} =
+0.5$ was contained in the 95\% \gls{HPD} interval for every dataset;
+consequently, it is not possible to say with high confidence that any of the
+\gls{alpha} values are different from each other. In synthetic data, the
+confidence intervals around \gls{alpha} narrowed when other parameters were
+marginalized out (\cref{fig:maginal}). Thus, it is possible that estimates of
+\gls{alpha} could be made more precise by specifying either exact values or
+informative priors on the other \gls{BA} parameters. We argued informally in
+\cref{subsec: whyba} that \gls{I}, \gls{m}, and \gls{N} can sometimes be
+estimated by standard epidemiological methods, although the difficulty may
+depend on the specific epidemic under consideration.}
+
+%One somewhat surprising result was the difference between parameter estimates 
+%for the \textcite{li2015hiv} and \textcite{wang2015targeting} datasets. Both
+%groups studied cohorts of acutely infected \gls{MSM} in major Chinese cities
+%%(Shanghai and Beijing respectively); yet, the \citeauthor{li2015hiv} data was 
+%estimated to have a lower \acrlong{PA} power and larger infected population
+%than the \textcite{wang2015targeting} data\ldots
 
@@
 
-\textcite{rothenberg2007large} estimated the power law exponents for 15 major
-network studies of sexual and injection drug use networks carried out between
-1981 and 2000. The estimated \gls{gamma} values for ``all contacts'' (that is,
-including both interviewed and non-interviewed individuals) were 
-between 2.08 and 2.87 for all but one of the networks\ldots
+\cref{tab:litgamma} shows some estimated \gls{gamma} values reported in the
+network literature.  The values mainly fall between 2 and 4, although
+substantial variation within this range exists for networks with apparently
+similar demographics. Some of these differences are likely due to the differing
+time intervals considered in each study -- some investigate the distribution of
+lifetime partners, while others consider only recent partnerships, such as in
+the past year. TODO
+
+\begin{table}[ht]
+    \centering
+    \input{\tablepath/lit-gamma}
+    \caption[
+        Values of power law exponent \gls{gamma} previously reported in the
+        literature.
+    ]{
+        Values of power law exponent \gls{gamma} previously reported in the
+        literature. In the risk group column, ``mixed sexual'' indicates
+        participants were not selected on the basis of sexual orientation;
+        these values can be assumed to reflect primarily heterosexual dynamics.
+        \textcite{brown2011transmission} studied a phylogenetically estimated
+        transmission network, rather than a contact network.
+        The values presented by \textcite{jones2003assessment} were in the
+        context of arguing that the power law distribution is not suitable for
+        use with network data.
+    }
+    \label{tab:kernelexpt}
+\end{table}
 
@@
 
-{\color{blue}\uline{The true \gls{HIV} prevalence in a population can be
+\add{The true \gls{HIV} prevalence in a population can be
 difficult to estimate for several reasons. \gls{HIV}-infected individuals may
@@
 (\cref{fig:Icrossv,fig:Ikpca}). In real world studies, the proportion of
-infected individuals sampled is usually low, whic may imede our ability to
-estimate \gls{I}.
+infected individuals sampled is usually low, which may impede our ability to
+estimate \gls{I}. In the following, we discuss any information available about
+the \gls{HIV} prevalence in the analysed study populations, and relate this to
+our estimates. However, due to the aforementioned challenges, this information
+cannot be taken as absolute ``ground truth''. Discrepancies between our results
+and the available data may indicate that the estimates produced by
+\software{netabc} are inaccurate, but they may also indicate that the available
+epidemiological data was not sufficient to estimate the true prevalence. Since
+the estimates of the prevalence \gls{I} were largely robust to the choice of
+prior on \gls{m} (\cref{fig:abchpd,fig:abchpdm2}), we discuss here only the
+results obtained with the prior $\gls{m} \sim \DiscreteUniform(1, 5)$.}
 
-The estimates of the prevalence \gls{I} were largely robust to the choice of
-prior on \gls{m}} (\cref{fig:abchpd,fig:abchpdm2}).
+\add{The prevalence estimates for two \gls{IDU} epidemics in Eastern Europe,
+IDU/Estonia and IDU/Romania, were similar 
+    (X and X), 
+although the confidence interval around the \textcite{niculescu2015recent} data
+was wider. \textcite{niculescu2015recent} reported that 494 new \gls{HIV}
+infections were diagnosed among the studied population (\glspl{IDU} in
+Bucharest) during the study period, and this value was contained in our 95\%
+\gls{HPD} 
+    (X-X). 
+On the other hand, \textcite{zetterberg2004two} collected samples from several
+locations throughout Estonia; using their reported incidence estimates, there
+were approximately 2000 new infections in the country during the study period.
+This value was outside our 95\% \gls{HPD} 
+    (X-X). 
+The authors indicated that their sample was ``convenience based''; it is
+possible that other, harder-to-reach individuals had no contacts with those
+sampled, reducing the apparent network size. In contrast, the IDU/Canada data
+was collected primarily from \gls{IDU} in Vancouver's downtown east side, a
+population that has been intensively targeted by public health interventions.
+Combined with the routine clinical genotyping performed in British Columbia for
+all new \gls{HIV} infections, it is not unreasonable to suppose that the
+sampled individuals (399) comprise the majority of \gls{HIV}-positive network
+members. TODO}
 
-\textcite{niculescu2015recent} and \textcite{zetterberg2004two} both reported 
-on epidemics in \gls{IDU} populations in Eastern Europe (Romania and Estonia,
-respectively). The point estimates of \gls{I} for these two datasets were
-similar (X and X), although the confidence interval around the
-\textcite{niculescu2015recent} data was wider. \textcite{niculescu2015recent}
-\uline{reported that 494 new \gls{HIV} infections were diagnosed among the
-studied population (\glspl{IDU} in Bucharest) during the study period, and this
-value was contained in our 95\% \gls{HPD} (X-X). On the other hand,
-\textcite{zetterberg2004two} collected samples from several locations
-throughout Estonia; using their reported incidence estimates, there were
-approximately 2000 new infections in the country during the study period. This
-value was outside our 95\% \gls{HPD} (X-X). The authors indicated that their
-sample was ``convenience based''. It is possible that other, harder-to-reach
-individuals had no contacts with those sampled, which would reduce the apparent
-network size. Alternatively, the low estimated prevalence may have been due to
-the low sampling fraction.}
+\add{Surprisingly, the prevalence estimates for two \gls{MSM} populations major
+Chinese cities (MSM/Shanghai and MSM/Beijing) were quite different
+    (X and X).
+The individuals sampled for the MSM/Beijing study were recruited by following a
+prospective cohort of size 2000. 179 new infections were identified in the
+cohort during the study period, which is much lower than the estimated
+prevalence obtained with \software{netabc}
+    (X).
+However, the authors did not claim to have sampled the entire sexual network.
+In a nationwide survey, \textcite{wu2013hiv} estimated that there were 24,198
+\gls{MSM} living in Beijing, of whom 5.7\%, or 1379, were \gls{HIV}-infected,
+which is within the 95\% confidence interval of our estimate of \gls{I}
+    (X-X).
+The same survey estimated the size of the \gls{MSM} population in Shanghai
+14,511, with an \gls{HIV} prevalence of 6.8\% or approximately 987 individuals.
+This was within the 95\% confidence interval of the estimate obtained with
+\software{netabc} for the MSM/Shanghai data
+    (X-X). 
+It is worth noting that the authors of the IDU/Shanghai
+study~\autocite{li2015hiv} claimed that there were 80,000 \gls{MSM} living in
+Shanghai, which is a significantly different estimate than that obtained by
+\textcite{wu2013hiv}. The authors of the MSM/Taiwan
+study~\autocite{kao2011surveillance} stated that there were 18,378
+\gls{HIV}-infected individuals in Taiwan, of which approximately 44\% (8086
+individuals) were \gls{MSM}. This value is well outside of our 95\% \gls{HPD} 
+    (X-X),
+suggesting that the entire \gls{MSM} population does not comprise a single
+network. In fact, our estimate of \gls{I} for these data
+    (X)
+was closer to the number of \gls{MSM} enrolled in the study, 301. The study
+focused exclusively on newly diagnosed individuals; perhaps chronically
+infected \gls{MSM} were no longer part of the same network. On the other hand,
+our estimated prevalence for the MSM/USA data
+    (X)
+was clearly an underestimate. The study enrolled almost 648 \gls{HIV} infected 
+individuals, which should have been a lower bound on \gls{I}, although the
+value was contained in the 95\% \gls{HPD} interval
+    (X-X).}
 
-\uline{The cluster we analysed within the BC data consists primarily of
-\gls{IDU} in Vancouver's downtown east side. This population has been
-intensively targeted by public health interventions. Combined with the routine 
-clinical genotyping performed in BC for all new \gls{HIV} infections, it is not
-unreasonable to suppose that the sampled individuals (399) comprise the
-majority of \gls{HIV}-positive network members.
+\add{The HET/Uganda, HET/Malawi, and HET/Botswana data all originated from
+populations in sub-Saharan Africa, where the dominant mode of transmission is
+believed to be heterosexual. The HET/Uganda study population included
+1434 seropositive individuals. Since the majority of inhabitants of the study
+district were enrolled in the study, it is likely that this number represents
+the majority of \gls{HIV} infected persons. Our estimate was slightly (TODO:
+lower/higher)
+    (X),
+but the value 1434 was contained ith the 95\% \gls{HPD} interval
+    (X-X).
+The HET/Malawi data represents possibly the oldest population-level \gls{HIV}
+sequencing effort, carried out on samples collected for other purposes between
+1981 and 1989. Every individual in the study district was sampled; among these,
+a total of 200 were \gls{HIV}-positive. Although the HET/Malawi data had the
+lowest estimated \gls{I} value among the heterosexual datasets, and 200 was
+contained within the 95\% \gls{HPD} interval
+    (X-X),
+the estimated prevalence was much higher than this value (X). Due to the
+thorough population-level screening, it is most likely that the posterior mean
+represents an overestimate, especially considering the observed upward bias of
+\gls{ABC} estimates of \gls{I} observed on simulated data (\cref{fig:abcptm2}).
+However, it is also possible that the true sexual network extended outside the
+study's geographic area to include a larger number of infected individuals.
+The authors of the HET/Botswana data~\autocite{novitsky2013phylogenetic, 
+novitsky2014impact} estimated that there were 1731 \gls{HIV}-positive
+individuals in the study area. \gls{HIV} sequences were obtained from
+approximately 70\% of these individuals. The estimated prevalence we obtained
+was much higher 
+    (X), 
+with a 95\% \gls{HPD} interval spanning nearly the entire prior region. Since
+the total population of the study area was approximately 15,000, this value of
+\gls{I} seems unlikely to be accurate. However, the authors noted that the town
+where the study took place was located proximally to the country's capital
+city, and suggested that frequent travel between the two locations may have
+facilitated linking of their sexual networks. Thus, the high estimate of
+\gls{I} we obtained for these data may include a larger network component based
+in the capital city.}
 
-\textcite{wang2015targeting} and \textcite{li2015hiv} studied \gls{MSM}
-populations in major Chinese cities (Shanghai and Beijing).
+\add{\textcite{cuevas2009hiv} reported that 620 newly infected and 1500
+chronically infected patients had been sampled during the study period, for a
+total of 2120. This is substantially greater than the point estimate of 
+    X 
+we obtained with \gls{ABC}. This error may in part be due to model
+misspecification -- it is highly unlikely that a mixed risk group population 
+would form a single, highly-connected contact network of the type generated by
+the \gls{BA} model.}
 
-\uline{The individuals sampled by} \textcite{wang2015targeting} \uline{were
-recruited by following a prospective cohort of 2000 \gls{MSM} in Beijing,
-China. Of the 2000, 179 new infections were identified during the study period.
-This number is much lower than the estimated prevalence of X obtained with
-\software{netabc}, however the authors did not claim to have sampled the entire
-sexual network. In a nationwide survey,} \textcite{wu2013hiv} \uline{estimated
-that there were 24,198 \gls{MSM} living in Beijing, of whom 5.7\%, or 1379,
-were \gls{HIV}-infected, which is within the 95\% confidence interval of our
-estimate of \gls{I}.}
-
-\textcite{li2015hiv} \uline{studied an \gls{MSM} population in Shanghai, China.
-Despite the apparent similarity to the context of the
-}\textcite{wang2015targeting} \uline{study, the estimated \gls{HIV} prevalence
-of the } \citeauthor{li2015hiv} \uline{data was higher (X). The aforementioned
-survey~\autocite{wu2013hiv} estimated that there were 14,511 \gls{MSM} living
-in Shanghai, with an \gls{HIV} prevalence of 6.8\% or approximately 987
-individuals. This was within the 95\% confidence interval of the estimate
-obtained with \software{netabc}. It is worth noting that} \citeauthor{li2015hiv}
-\uline{claimed that there were 80,000 \gls{MSM} living in Shanghai, which is a
-significantly different estimate than that obtained by} \textcite{wu2013hiv}.
-
-\textcite{cuevas2009hiv} \uline{reported that 620 newly infected and 1500
-chronically infected patients had been sampled during the study period, for a
-total of 2120. This is substantially greater than the point estimate of X we
-obtained with \gls{ABC}, even without considering the likely possibility that
-there were more infected than sampled individuals. Again, this error may in
-part be due to model misspecification, which highlights the importance of
-selecting a model appropriate to the dataset at hand.}}
-
-\subsubsection{Caveats? TODO}
+\subsubsection{Modelling assumptions}
 
@@
 misspecification, our estimates of the power law exponent $\gamma$ were within
-the range of values reported in the literature (\cref{tab:gamma}).
+the range of values reported in the literature (\cref{tab:gamma}), \add{and our
+estimates of \gls{alpha} were in rough agreement with a previous study using
+survey-based epidemiology~\autocite{de2007preferential}.}

### 2016-06-21 doc/thesis/appendix.Rnw
Updated abstract and a bit more
Up to ABC in intro now
Almooooooost!

@@
 
-A \defn{mathematical model} is a formal description of a hypothesized
-relationship between some observed data, $x$ and outcomes $y$. A
-\defn{parametric} model defines a family of possible relationships between data
-and outcomes, parameterized by one or more numeric parameters $\theta$. A
-\defn{statistical} model describes the relationship between data and outcomes
-in terms of probabilities. Statistical models define, either explicitly or
-implicitly, the probability of observing $y$ given $x$ and, if the model is
-parametric, $\theta$. Note that it is entirely possible to have no data $x$,
-only observed outcomes $y$. In this case, a model would describe the process by
-which $y$ is generated.
+\add{This appendix reviews some fundamental statistical concepts that are
+referred to throughout the body of this thesis.} A \defn{mathematical model} is
+a formal description of a hypothesized relationship between some observed data,
+\del{$x$ and outcomes $y$} \add{$y$ and covariates $x$}. A \defn{parametric}
+model defines a family of possible relationships between data and outcomes,
+parameterized by one or more numeric parameters $\theta$. A \defn{statistical}
+model describes the relationship between data and outcomes in terms of
+probabilities. Statistical models define, either explicitly or implicitly, the
+probability of observing $y$ given $x$ and, if the model is parametric,
+$\theta$. Note that it is entirely possible to have no data $x$, only observed
+outcomes $y$. In this case, a model would describe the process by which $y$ is
+generated.
 
@@
 $\theta$ = ($\beta$, $\sigma$). Moreover, we can write down the probability
-density $\pi$ of observing outcome $y_i$ given the parameters,
+density \lik of observing outcome $y_i$ given the parameters,
 \[
-  \pi(y \mid \beta, \sigma) = 
-  \prod_{i=1}^n f_{\N(0, \sigma^2)} (y_i - \beta x_i),
+  \lik(y \mid \beta, \sigma) = 
+  \prod_{i=1}^n \lik_{\N(0, \sigma^2)} (y_i - \beta x_i),
 \]
-where $f_{\N(0, \sigma^2)}$ is the probability density of the normal
+where $\lik_{\N(0, \sigma^2)}$ is the probability density of the normal
 distribution with mean zero and variance $\sigma^2$. Note that we are treating
@@
 $\theta$ is also known as the \defn{likelihood}, written $\L$, of $\theta$.
-That is, $\L(\theta \mid y) = f(y \mid \theta)$ for the model's \gls{pdf} $f$.
+That is, $\L(\theta \mid y) = \lik(y \mid \theta)$ for the model's \gls{pdf} $f$.
 The higher the value of the likelihood, the more likely the observations $y$
@@
 beliefs as a \defn{prior distribution} on the parameters, denoted
-$\pi(\theta)$. Instead of considering only the likelihood, Bayesian inference
-focuses on the product of the likelihood and the prior, $f(y \mid \theta)
-\pi(\theta)$. Bayes' theorem tells us that this product is related to the
+$\prior(\theta)$. Instead of considering only the likelihood, Bayesian inference
+focuses on the product of the likelihood and the prior, $\lik(y \mid \theta)
+\prior(\theta)$. Bayes' theorem tells us that this product is related to the
 \textit{posterior distribution} on $\theta$,
 \begin{align}
-  f(\theta \mid y) 
-    = \frac{f(y \mid \theta) \pi(\theta)}
-           {\int f(y \mid \theta) \pi(\theta) \d \theta}.
+  \post(\theta \mid y) 
+    = \frac{\lik(y \mid \theta) \prior(\theta)}
+           {\int \lik(y \mid \theta) \prior(\theta) \d \theta}.
   \label{eq:bayes}
 \end{align}
-In principle, $f(y \mid \theta) \pi(\theta)$ can be optimized numerically just
-like $\L(\theta \mid y)$, which would also optimize the posterior distribution.
-The resulting optimal parameters are called the \gls{MAP} estimates. However,
-from a Bayesian perspective, $\theta$ is not a fixed quantity to be estimated,
-but rather a random variable with an associated distribution (the posterior).
-Therefore, the \gls{MAP} estimate by itself is of limited value without
-associated statistics about the posterior distribution, such as the mean or
-credible intervals. Unfortunately, to calculate such statistics, it is
-necessary to evaluate the normalizing constant in the denominator of
-\cref{eq:bayes}, which is almost always an intractable integral.
+In principle, $\post(y \mid \theta) \prior(\theta)$ can be optimized
+numerically just like $\L(\theta \mid y)$, which would also optimize the
+posterior distribution. The resulting optimal parameters are called the
+\gls{MAP} estimates. However, from a Bayesian perspective, $\theta$ is not a
+fixed quantity to be estimated, but rather a random variable with an associated
+distribution (the posterior). Therefore, the \gls{MAP} estimate by itself is of
+limited value without associated statistics about the posterior distribution,
+such as the mean or credible intervals. Unfortunately, to calculate such
+statistics, it is necessary to evaluate the normalizing constant in the
+denominator of \cref{eq:bayes}, which is almost always an intractable integral.
 
@@
 \gls{MCMC} to obtain a sample from the posterior distribution. \Gls{MCMC} works
-by defining a Markov chain {\color{red}\sout{whose states are indexed by
-possible model parameters. The transition probability from state $\theta_1$ to
-state $\theta_2$ is taken to be}} {\color{blue}\uline{on the space of possible
-model parameters. The transition density from parameters $\theta_1$ to
-$\theta_2$ is taken to be}}
+by defining a Markov chain \del{whose states are indexed by possible model
+parameters. The transition probability from state $\theta_1$ to state
+$\theta_2$ is taken to be} \add{on the space of possible model parameters. The
+transition density from parameters $\theta_1$ to $\theta_2$ is taken to be}
 \[
-  \min\left(1, \frac{f(y \mid \theta_2) \pi(\theta_2) q(\theta_2, \theta_1)}
-                    {f(y \mid \theta_1) \pi(\theta_2) q(\theta_1, \theta_2)} \right),
+  \min\left(1, \frac{\lik(y \mid \theta_2) \prior(\theta_2) q(\theta_2, \theta_1)}
+                    {\lik(y \mid \theta_1) \prior(\theta_2) q(\theta_1, \theta_2)} \right),
 \]
@@
 enough random walk is performed on the chain, the distribution of states
-visited will be a Monte Carlo approximation of $f(\theta \mid y)$, from
+visited will be a Monte Carlo approximation of $\post(\theta \mid y)$, from
 which we can calculate statistics of interest. Actually performing this random
@@
   \begin{algorithmic}
-    \State Draw $\theta$ according to the prior $\pi(\theta)$
+    \State Draw $\theta$ according to the prior $\prior(\theta)$
     \Loop
@@
       $\min \left( 1, 
-       \dfrac{f(y \mid \theta') \pi(\theta') q(\theta', \theta)}
-             {f(y \mid \theta\phantom{'}) \pi(\theta\phantom{'}) q(\theta, \theta')}
+       \dfrac{\lik(y \mid \theta') \prior(\theta') q(\theta', \theta)}
+             {\lik(y \mid \theta\phantom{'}) \prior(\theta\phantom{'}) q(\theta, \theta')}
        \right)$
@@
     \includegraphics[width=\textwidth]{kernel-I-tree.pdf}
-  \caption[Simulated transmission trees under three different values of BA parameter $I$]{
-    Simulated transmission trees under three different values of BA parameter
-    $I$. Epidemics were simulated on \gls{BA} networks with parameters $\alpha$
-    = 1.0, $m = 2$, and $N = 5000$. Epidemics were simulated until $I = 500$,
-    1000, or 2000 nodes were infected. Transmission trees were created by
-    sampling 500 infected nodes. For higher \gls{I} values, the network was
-    closer to saturation at the time of sampling, resulting in longer terminal
-    branches as the waiting time until the next transmission increased.
+    \caption[
+        Simulated transmission trees under three different values of \gls{BA}
+        parameter \gls{I}
+    ]{
+        Simulated transmission trees under three different values of \gls{BA}
+        parameter \gls{I}. Epidemics were simulated on \gls{BA} networks with
+        parameters $\gls{alpha} = 1.0$, $\gls{m} = 2$, and $\gls{N} = 5000$.
+        Epidemics were simulated until $\gls{I} = 500$, 1000, or 2000 nodes
+        were infected. Transmission trees were created by sampling 500 infected
+        nodes. For higher \gls{I} values, the network was closer to saturation
+        at the time of sampling, resulting in longer terminal branches as the
+        waiting time until the next transmission increased.
     }
@@
     \includegraphics[width=\textwidth]{kernel-m-tree.pdf}
-  \caption[Simulated transmission trees under three different values of BA parameter $m$]{
-    Simulated transmission trees under three different values of BA parameter
-    $m$. Epidemics were simulated on \gls{BA} networks with parameters $\alpha$
-    = 1.0, $N = 5000$, and $m$ = 2, 3, or 4. Epidemics were simulated until $I
-    = 1000$ nodes were infected. Transmission trees were created by sampling
-    500 infected nodes.
+    \caption[
+        Simulated transmission trees under three different values of \gls{BA}
+        parameter \gls{m}
+    ]{
+        Simulated transmission trees under three different values of \gls{BA}
+        parameter \gls{m}. Epidemics were simulated on \gls{BA} networks with
+        parameters $\gls{alpha} = 1.0$, $\gls{n} = 5000$, and $\gls{m} = 2$, 3,
+        or 4. Epidemics were simulated until $\gls{I} = 1000$ nodes were
+        infected. Transmission trees were created by sampling 500 infected
+        nodes.
     }
@@
     \includegraphics[width=\textwidth]{kernel-N-tree.pdf}
-  \caption[Simulated transmission trees under three different values of BA parameter $N$]{
-    Simulated transmission trees under three different values of BA parameter
-    $N$. Epidemics were simulated on \gls{BA} networks with parameters $\alpha$
-    = 1.0, $m = 2$, and $N$ = 3000, 5000, or 8000. Epidemics were simulated
-    until $I = 1000$ nodes were infected. Transmission trees were created by
-    sampling 500 infected nodes. For lower $N$ values, the network was closer
-    to saturation at the time of sampling, resulting in longer waiting times
-    until the next transmission and longer terminal branch lengths.
+    \caption[
+        Simulated transmission trees under three different values of \gls{BA}
+        parameter \gls{N}
+    ]{
+        Simulated transmission trees under three different values of BA
+        parameter \gls{N}. Epidemics were simulated on \gls{BA} networks with
+        parameters $\gls{alpha} = 1.0$, $\gls{m} = 2$, and $\gls{N}$ = 3000,
+        5000, or 8000. Epidemics were simulated until $I = 1000$ nodes were
+        infected. Transmission trees were created by sampling 500 infected
+        nodes. For lower $N$ values, the network was closer to saturation at
+        the time of sampling, resulting in longer waiting times until the next
+        transmission and longer terminal branch lengths.
     }
@@
     \includegraphics{kernel-alpha-crossv.pdf}
-  \caption[Cross validation accuracy of classifiers for \gls{BA} model parameter
-    \gls{alpha} for eight epidemic scenarios.]
-  {
+    \caption[
+        Cross validation accuracy of classifiers for \gls{BA} model parameter
+        \gls{alpha} for eight epidemic scenarios.
+    ]{
         Cross validation accuracy of classifiers for \gls{BA} model parameter
@@
     \includegraphics{kernel-I-crossv.pdf}
-  \caption[Cross validation accuracy of classifiers for \gls{BA} model parameter
-    \gls{I} for eight epidemic scenarios.]
-  {
+    \caption[
         Cross validation accuracy of classifiers for \gls{BA} model parameter
-    \gls{I} for eight epidemic scenarios. Solid lines and points are $R^2$ of
-    tree kernel \gls{kSVR} under various kernel meta-parameters. Dashed and
-    dotted lines are $R^2$ of linear regression against Sackin's index, and
-    \gls{SVR} using \gls{nltt}.
+        \gls{I} for eight epidemic scenarios.
+    ]{
+        Cross validation accuracy of classifiers for \gls{BA} model parameter
+        \gls{I} for eight epidemic scenarios. Solid lines and points are $R^2$
+        of tree kernel \gls{kSVR} under various kernel meta-parameters. Dashed
+        and dotted lines are $R^2$ of linear regression against Sackin's index,
+        and \gls{SVR} using \gls{nltt}.
     }
@@
     \includegraphics{kernel-m-crossv.pdf}
-  \caption[Cross validation accuracy of classifiers for \gls{BA} model parameter
-    \gls{m} for eight epidemic scenarios.]
-  {
+    \caption[
         Cross validation accuracy of classifiers for \gls{BA} model parameter
-    \gls{m} for eight epidemic scenarios. Solid lines and points are $R^2$ of
-    tree kernel \gls{kSVR} under various kernel meta-parameters. Dashed and
-    dotted lines are $R^2$ of linear regression against Sackin's index, and
-    \gls{SVR} using \gls{nltt}.
+        \gls{m} for eight epidemic scenarios.
+    ]{
+        Cross validation accuracy of classifiers for \gls{BA} model parameter
+        \gls{m} for eight epidemic scenarios. Solid lines and points are $R^2$
+        of tree kernel \gls{kSVR} under various kernel meta-parameters. Dashed
+        and dotted lines are $R^2$ of linear regression against Sackin's index,
+        and \gls{SVR} using \gls{nltt}.
     }
@@
     \includegraphics{kernel-N-crossv.pdf}
-  \caption[Cross validation accuracy of classifiers for \gls{BA} model parameter
-    \gls{N} for eight epidemic scenarios.]
-  {
+    \caption[
         Cross validation accuracy of classifiers for \gls{BA} model parameter
-    \gls{N} for eight epidemic scenarios. Solid lines and points are $R^2$ of
-    tree kernel \gls{kSVR} under various kernel meta-parameters. Dashed and
-    dotted lines are $R^2$ of linear regression against Sackin's index, and
-    \gls{SVR} using \gls{nltt}.
+        \gls{N} for eight epidemic scenarios.
+    ]{
+        Cross validation accuracy of classifiers for \gls{BA} model parameter
+        \gls{N} for eight epidemic scenarios. Solid lines and points are $R^2$
+        of tree kernel \gls{kSVR} under various kernel meta-parameters. Dashed
+        and dotted lines are $R^2$ of linear regression against Sackin's index,
+        and \gls{SVR} using \gls{nltt}.
     }
@@
         Grid search kernel scores for testing trees simulated under various
-    \gls{alpha} values. The other \gls{BA} parameters were fixed at \gls{I} =
-    1000, \gls{N} = 5000, and \gls{m} = 2. 
+        \gls{alpha} values. The other \gls{BA} parameters were fixed at \gls{I}
+        = 1000, \gls{N} = 5000, and \gls{m} = 2. 
     }        
@@
         Grid search kernel scores for testing trees simulated under various
-    \gls{I} values. The other \gls{BA} parameters were fixed at \gls{alpha} =
-    1.0, \gls{N} = 5000, and \gls{m} = 2. 
+        \gls{I} values. The other \gls{BA} parameters were fixed at \gls{alpha}
+        = 1.0, \gls{N} = 5000, and \gls{m} = 2. 
     }        
@@
         Grid search kernel scores for testing trees simulated under various
-    \gls{m} values. The other \gls{BA} parameters were fixed at \gls{alpha} =
-    1.0, \gls{I} = 1000, and \gls{N} = 5000.
+        \gls{m} values. The other \gls{BA} parameters were fixed at \gls{alpha}
+        = 1.0, \gls{I} = 1000, and \gls{N} = 5000.
     }        
@@
         Grid search kernel scores for testing trees simulated under various
-    \gls{N} values. The other \gls{BA} parameters were fixed at \gls{alpha} =
-    1.0, \gls{I} = 1000, and \gls{m} = 2.
+        \gls{N} values. The other \gls{BA} parameters were fixed at \gls{alpha}
+        = 1.0, \gls{I} = 1000, and \gls{m} = 2.
     }        
@@
         Point estimates of preferential attachment power \gls{alpha} of
-      \acrlong{BA} network model, obtained on simulated trees with
-      kernel-score-based grid search.]
-  {
+        \gls{BA} network model, obtained on simulated trees with
+        kernel-score-based grid search.
+    ]{
         Point estimates of preferential attachment power \gls{alpha} of
@@
         several values of \gls{alpha} ($x$-axis) with other model parameters
-      fixed at \gls{m} = 2, \gls{N} = 5000, and \gls{I} = 1000. The test trees
-      were compared to trees simulated along a narrowly spaced grid of
+        fixed at \gls{m} = 2, \gls{N} = 5000, and \gls{I} = 1000. The test
+        trees were compared to trees simulated along a narrowly spaced grid of
         \gls{alpha} values using the tree kernel, with the same values of the
-      other parameters. The grid value with the highest median kernel score was
-      taken as a point estimate for \gls{alpha} ($y$-axis).
+        other parameters. The grid value with the highest median kernel score
+        was taken as a point estimate for \gls{alpha} ($y$-axis).
     }
@@
     \includegraphics{gridsearch-I-point-estimate.pdf}
-  \caption[ Point estimates of prevalence at time of sampling \gls{I} of
-      \acrlong{BA} network model, obtained on simulated trees with
-      kernel-score-based grid search. ]
-  {
-      Point estimates of prevalence at time of sampling \gls{I} of \acrlong{BA}
-      network model, obtained on simulated trees with kernel-score-based grid
-      search. Test trees were simulated according to several values of \gls{I}
-      ($x$-axis) with other model parameters fixed at \gls{alpha} = 1, \gls{m}
-      = 2, and \gls{N} = 5000. The test trees were compared to trees simulated
-      along a narrowly spaced grid of \gls{I} values using the tree kernel,
-      with the same values of the other parameters. The grid value with the
-      highest median kernel score was taken as a point estimate for \gls{I}
-      ($y$-axis).
+    \caption[ 
+        Point estimates of prevalence at time of sampling \gls{I} of
+        \gls{BA} network model, obtained on simulated trees with
+        kernel-score-based grid search.
+    ]{
+        Point estimates of prevalence at time of sampling \gls{I} of
+        \gls{BA} network model, obtained on simulated trees with
+        kernel-score-based grid search. Test trees were simulated according to
+        several values of \gls{I} ($x$-axis) with other model parameters fixed
+        at \gls{alpha} = 1, \gls{m} = 2, and \gls{N} = 5000. The test trees
+        were compared to trees simulated along a narrowly spaced grid of
+        \gls{I} values using the tree kernel, with the same values of the other
+        parameters. The grid value with the highest median kernel score was
+        taken as a point estimate for \gls{I} ($y$-axis).
     }
@@
     \caption[
-      Point estimates of number of edges per vertex \gls{m} of \acrlong{BA}
+        Point estimates of number of edges per vertex \gls{m} of \gls{BA}
         network model, obtained on simulated trees with kernel-score-based grid
         search.
-  ]
-  {
-      Point estimates of number of edges per vertex \gls{m} of \acrlong{BA}
+    ]{
+        Point estimates of number of edges per vertex \gls{m} of \gls{BA}
         network model, obtained on simulated trees with kernel-score-based grid
-      search. Test trees were simulated according to several values of \gls{m}
-      ($x$-axis) with other model parameters fixed at \gls{alpha} = 1, \gls{I}
-      = 1000, and \gls{N} = 5000. The test trees were compared to trees
-      simulated along a narrowly spaced grid of \gls{m} values using the tree
-      kernel, with the same values of the other parameters. The grid value with
-      the highest median kernel score was taken as a point estimate for \gls{m}
-      ($y$-axis).
+        search. Test trees were simulated according to several values of
+        \gls{m} ($x$-axis) with other model parameters fixed at \gls{alpha} =
+        1, \gls{I} = 1000, and \gls{N} = 5000. The test trees were compared to
+        trees simulated along a narrowly spaced grid of \gls{m} values using
+        the tree kernel, with the same values of the other parameters. The grid
+        value with the highest median kernel score was taken as a point
+        estimate for \gls{m} ($y$-axis).
     }
@@
     \caption[
-      Point estimates of number of edges per vertex \gls{N} of \acrlong{BA}
+        Point estimates of number of edges per vertex \gls{N} of \gls{BA}
         network model, obtained on simulated trees with kernel-score-based grid
@@
     ]{
-      Point estimates of number of edges per vertex \gls{N} of \acrlong{BA}
+        Point estimates of number of edges per vertex \gls{N} of \gls{BA}
         network model, obtained on simulated trees with kernel-score-based grid
-      search. Test trees were simulated according to several values of \gls{N}
-      ($x$-axis) with other model parameters fixed at \gls{alpha} = 1, \gls{m}
-      = 2, and \gls{I} = 1000. The test trees were compared to trees simulated
-      along a narrowly spaced grid of \gls{N} values using the tree kernel,
-      with the same values of the other parameters. The grid value with the
-      highest median kernel score was taken as a point estimate for \gls{m}
-      ($y$-axis).
+        search. Test trees were simulated according to several values of
+        \gls{N} ($x$-axis) with other model parameters fixed at \gls{alpha} =
+        1, \gls{m} = 2, and \gls{I} = 1000. The test trees were compared to
+        trees simulated along a narrowly spaced grid of \gls{N} values using
+        the tree kernel, with the same values of the other parameters. The grid
+        value with the highest median kernel score was taken as a point
+        estimate for \gls{m} ($y$-axis).
     }
@@
 
-%\begin{figure}[ht]
-%  \centering
-%  \includegraphics{abc-point-estimate-m1.pdf}
-%  \caption[
-%    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
-%    running \software{netabc} on simulated data, for simulations with \gls{m} = 1.
-%  ]{
-%    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
-%    running \software{netabc} on simulated data, for simulations with \gls{m} = 3.
-%    Dashed lines indicate true values. (A) Estimates of \gls{alpha} and \gls{I}
-%    which were varied in these simulations against known values. (B) Estimates
-%    of \gls{m} and \gls{N} which were held fixed in these simulations at the
-%    values \gls{m} = 1 and \gls{N} = 5000.
-%  }        
-%  \label{fig:abcptm1}
-%\end{figure}
-
 \begin{figure}[ht]
@@
     \caption[
-    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
-    running \software{netabc} on simulated data, for simulations with \gls{m} = 3.
+        Posterior mean point estimates for \gls{BA} model parameters obtained
+        by running \software{netabc} on simulated data, for simulations with
+        \gls{m} = 3. 
     ]{
-    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
-    running \software{netabc} on simulated data, for simulations with \gls{m} = 3.
-    Dashed lines indicate true values. (A) Estimates of \gls{alpha} and \gls{I}
-    which were varied in these simulations against known values. (B) Estimates
-    of \gls{m} and \gls{N} which were held fixed in these simulations at the
-    values \gls{m} = 3 and \gls{N} = 5000.
+        Posterior mean point estimates for \gls{BA} model parameters obtained
+        by running \software{netabc} on simulated data, for simulations with
+        \gls{m} = 3. Dashed lines indicate true values. (A) Estimates of
+        \gls{alpha} and \gls{I} which were varied in these simulations against
+        known values. (B) Estimates of \gls{m} and \gls{N} which were held
+        fixed in these simulations at the values \gls{m} = 3 and \gls{N} =
+        5000.
     }        
@@
     \caption[
-    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
-    running \software{netabc} on simulated data, for simulations with \gls{m} = 3.
+        Posterior mean point estimates for \gls{BA} model parameters obtained
+        by running \software{netabc} on simulated data, for simulations with
+        \gls{m} = 3.
     ]{
-    \Acrlong{MAP} point estimates for \gls{BA} model parameters obtained by
-    running \software{netabc} on simulated data, for simulations with \gls{m} = 4.
-    Dashed lines indicate true values. (A) Estimates of \gls{alpha} and \gls{I}
-    which were varied in these simulations against known values. (B) Estimates
-    of \gls{m} and \gls{N} which were held fixed in these simulations at the
-    values \gls{m} = 4 and \gls{N} = 5000.
+        Posterior mean point estimates for \gls{BA} model parameters obtained
+        by running \software{netabc} on simulated data, for simulations with
+        \gls{m} = 4. Dashed lines indicate true values. (A) Estimates of
+        \gls{alpha} and \gls{I} which were varied in these simulations against
+        known values. (B) Estimates of \gls{m} and \gls{N} which were held
+        fixed in these simulations at the values \gls{m} = 4 and \gls{N} =
+        5000.
     }        
@@
     \caption{
-      Relationship between preferential attachment power parameter $\alpha$
-      and power law exponent $\gamma$ for networks simulated under the BA
-      network model with $N$ = 5000 and $m$ = 2.
+        Relationship between preferential attachment power parameter
+        \gls{alpha} and fitted power law exponent \gls{gamma} for networks
+        simulated under the \gls{BA} network model with $\gls{N} = 5000$ and
+        $\gls{m} = 2$.
     }
@@
     \centering
+    \includegraphics{powerlaw.pdf}
+    \caption{
+        Best fit power law and stretched exponential curves for degree
+        distributions of simulated \acrlong{BA} networks for several values of
+        \gls{alpha} and \gls{m}.
+    }
+    \label{fig:powerlaw}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
     \includegraphics{mixed-posterior}
-    \caption[Approximate marginal posterior distributions of BA model
-        parameters obtained using kernel-assisted ABC for a network with heterogeneous
-        node behaviour.]
-    {
-        Approximate marginal posterior distributions of Barab\'asi-Albert
-        model parameters obtained using kernel-assisted ABC for a network with
+    \caption[
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters obtained with \software{netabc} for a network with
+        heterogeneous node behaviour.
+    ]{
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters obtained with \software{netabc} for a network with
         heterogeneous node behaviour. Half of the nodes were attached with
-        $\alpha = 0.5$, and the other half with $\alpha = 1.5$ (vertical
-        dashed lines, top left). Other parameter values were $m = 2$, $I =
-        1000$, and $N = 5000$ (vertical dashed lines, other than top left).
-        Shaded areas indicate 95\% highest posterior density intervals.
+        $\gls{alpha} = 0.5$, and the other half with $\gls{alpha} = 1.5$
+        (vertical dashed lines, top left). Other parameter values were $m = 2$,
+        $\gls{I} = 1000$, and $\gls{N} = 5000$ (vertical dashed lines, other
+        than top left). Shaded areas indicate 95\% highest posterior density
+        intervals.
     }
@@
     \includegraphics{peerdriven-posterior}
-    \caption[Approximate marginal posterior distributions of Barab\'asi-Albert
-        model parameters obtained using kernel-assisted ABC for a network with
-        peer-driven sampling.]
-    {
-        Approximate marginal posterior distributions of Barab\'asi-Albert
-        model parameters obtained using kernel-assisted ABC for a network with
+    \caption[
+        Approximate marginal posterior distributions of \gls{BA}
+        model parameters obtained with \software{netabc} for a network with
+        peer-driven sampling.
+    ]{
+        Approximate marginal posterior distributions of \gls{BA}
+        model parameters obtained with \software{netabc} for a network with
         peer-driven sampling. An epidemic was simulated in the usual fashion,
@@
     \includegraphics{realdata-hpd-bc-m2}
-  \vspace{8pt}
     \caption[
-      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
-      parameters of the BA network model, fitted to five published HIV datasets
-      with \software{netabc} using the prior $m \sim DiscreteUniform(2, 5)$.
+        Posterior mean point estimates and 95\% \gls{HPD} intervals for
+        parameters of the \gls{BA} network model, fitted to eleven \gls{HIV}
+        datasets with \software{netabc} using the prior $\gls{m} \sim
+        \DiscreteUniform(2, 5)$.
     ]{
-      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
-      parameters of the BA network model, fitted to five published HIV datasets
-      with \software{netabc} using the prior $m \sim$ DiscreteUniform(2, 5).
-      $x$-axes indicate regions of nonzero prior density. In particular, the
-      prior on $m$ was DiscreteUniform(2, 5).
+        Posterior mean point estimates and 95\% \gls{HPD} intervals for
+        parameters of the \gls{BA} network model, fitted to five published
+        \gls{HIV} datasets with \software{netabc} using the prior $\gls{m}
+        \sim \DiscreteUniform(2, 5)$. $x$-axes indicate regions of nonzero
+        prior density. 
     }
@@
 \begin{figure}[ht]
-  %\includegraphics{bctree-posterior}
-  \caption[
-      Approximate marginal posterior distributions of BA model parameters for
-      BC data.
-  ]
-  {
-      Approximate marginal posterior distributions of BA model parameters for
-      BC data. Vertical lines indicate maximum \textit{a posteriori} estimates,
-      and shaded areas are 95\% highest posterior density intervals. $x$-axis
-      indicates regions of nonzero prior density.
+    \includegraphics{bctree-posterior}
+    \caption
+    [
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to IDU/Canada data with \software{netabc}.
+    ]{
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to IDU/Canada data with \software{netabc}. Vertical
+        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
+        intervals. $x$-axis indicates regions of nonzero prior density.
     }
@@
     \caption[
-      Approximate marginal posterior distributions of BA model parameters for
-      \textcite{cuevas2009hiv} data. 
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to mixed/Spain data with \software{netabc}.
     ]{
-      Approximate marginal posterior distributions of BA model parameters for
-      \textcite{cuevas2009hiv} data. Vertical lines indicate maximum \textit{a
-      posteriori} estimates, and shaded areas are 95\% highest posterior
-      density intervals. $x$-axis indicates regions of nonzero prior density.
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to mixed/Spain data with \software{netabc}. Vertical
+        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
+        intervals. $x$-axis indicates regions of nonzero prior density.
     }
@@
 \begin{figure}[ht]
+    \includegraphics{grabowski2014-posterior}
+    \caption[
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to HET/Uganda data with \software{netabc}.
+    ]{
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to HET/Uganda data with \software{netabc}. Vertical
+        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
+        intervals. $x$-axis indicates regions of nonzero prior density.
+    }
+    \label{fig:grabowski}
+\end{figure}
+
+\begin{figure}[ht]
+    \includegraphics{kao2011-posterior}
+    \caption[
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to MSM/Taiwan data with \software{netabc}.
+    ]{
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to MSM/Taiwan data with \software{netabc}. Vertical
+        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
+        intervals. $x$-axis indicates regions of nonzero prior density.
+    }
+    \label{fig:kao}
+\end{figure}
+
+\begin{figure}[ht]
     \includegraphics{li2015-posterior}
     \caption[
-      Approximate marginal posterior distributions of BA model parameters for
-      \textcite{li2015hiv} data. 
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to MSM/Shanghai data with \software{netabc}.
     ]{
-      Approximate marginal posterior distributions of BA model parameters for
-      \textcite{li2015hiv} data. Vertical lines indicate maximum \textit{a
-      posteriori} estimates, and shaded areas are 95\% highest posterior
-      density intervals. $x$-axis indicates regions of nonzero prior density.
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to MSM/Shanghai data with \software{netabc}. Vertical
+        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
+        intervals. $x$-axis indicates regions of nonzero prior density. 
     }
@@
 \begin{figure}[ht]
+    \includegraphics{little2014-posterior}
+    \caption[
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to MSM/USA data with \software{netabc}.
+    ]{
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to MSM/USA data with \software{netabc}. Vertical
+        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
+        intervals. $x$-axis indicates regions of nonzero prior density. 
+    }
+    \label{fig:little}
+\end{figure}
+
+\begin{figure}[ht]
+    \includegraphics{mccormack2002-posterior}
+    \caption[
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to HET/Malawi data with \software{netabc}.
+    ]{
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to HET/Malawi data with \software{netabc}. Vertical
+        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
+        intervals. $x$-axis indicates regions of nonzero prior density. 
+    }
+    \label{fig:mccormack}
+\end{figure}
+
+\begin{figure}[ht]
     \includegraphics{niculescu2015-posterior}
     \caption[
-      Approximate marginal posterior distributions of BA model parameters for
-      \textcite{niculescu2015recent} data.
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to IDU/Romania data with \software{netabc}.
     ]{
-      Approximate marginal posterior distributions of BA model parameters for
-      \textcite{niculescu2015recent} data. Vertical lines indicate maximum
-      \textit{a posteriori} estimates, and shaded areas are 95\% highest
-      posterior density intervals. $x$-axis indicates regions of nonzero prior
-      density.
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to IDU/Romania data with \software{netabc}. Vertical
+        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
+        intervals. $x$-axis indicates regions of nonzero prior density.
     }
@@
     \caption[
-      Approximate marginal posterior distributions of BA model parameters for
-      \textcite{novitsky2013phylogenetic, novitsky2014impact} data.
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to HET/Botswana data with \software{netabc}.
     ]{
-      Approximate marginal posterior distributions of BA model parameters for
-      \textcite{novitsky2013phylogenetic, novitsky2014impact} data. Vertical
-      lines indicate maximum \textit{a posteriori} estimates, and shaded areas
-      are 95\% highest posterior density intervals. $x$-axis indicates regions
-      of nonzero prior density.
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to HET/Botswana data with \software{netabc}. Vertical
+        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
+        intervals. $x$-axis indicates regions of nonzero prior density.
     }
@@
     \caption[
-      Approximate marginal posterior distributions of BA model parameters for
-      \textcite{wang2015targeting} data.
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to MSM/Beijing data with \software{netabc}.
     ]{
-      Approximate marginal posterior distributions of BA model parameters for
-      \textcite{wang2015targeting} data. Vertical lines indicate maximum
-      \textit{a posteriori} estimates, and shaded areas are 95\% highest
-      posterior density intervals. $x$-axis indicates regions of nonzero prior
-      density.
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to MSM/Beijing data with \software{netabc}. Vertical
+        lines indicate posterior means, and shaded areas are 95\% highest
+        posterior density intervals. $x$-axis indicates regions of nonzero
+        prior density.
     }
@@
 
-\clearpage
+\begin{figure}[ht]
+    \includegraphics{zetterberg2004-posterior}
+    \caption[
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to IDU/Estonia data with \software{netabc}.
+    ]{
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to IDU/Estonia data with \software{netabc}. Vertical
+        lines indicate posterior means, and shaded areas are 95\% highest
+        posterior density intervals. $x$-axis indicates regions of nonzero
+        prior density.
+    }
+    \label{fig:zetterberg}
+\end{figure}
 
-<<abc_posterior, echo=FALSE, results="asis">>=
-    N <- 5000
-    replicate <- 0
-    for (m in c(2, 3, 4)) {
-    for (I in c(1000, 2000)) {
-    for (alpha in c(0, 0.5, 1, 1.5)) {
-        fn <- sprintf("{abc-posterior/%.1f_%d_%d_%d_%d}.pdf", alpha, I, m, N, replicate)
-        lab <- sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate)
-        cat("\\begin{figure}[ht]\n")
-        cat("\\centering\n")
-        cat(paste0("\\includegraphics{", fn, "}\n"))
-        cat(sprintf("\\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\\alpha$ = %.1f, $I$ = %d, $m$ = %d, and $N$ = %d.]{\n", alpha, I, m, N))
-        cat(sprintf("Approximate marginal posterior distributions of BA model parameters obtained by applying \\software{netabc} to a simulated transmission tree with BA parameter values $\\alpha$ = %.1f, $I$ = %d, $m$ = %d, and $N$ = %d.
+%<<abc_posterior, echo=FALSE, results="asis">>=
+%    N <- 5000
+%    replicate <- 0
+%    for (m in c(2, 3, 4)) {
+%    for (I in c(1000, 2000)) {
+%    for (alpha in c(0, 0.5, 1, 1.5)) {
+%        fn <- sprintf("{abc-posterior/%.1f_%d_%d_%d_%d}.pdf", alpha, I, m, N, replicate)
+%        lab <- sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate)
+%        cat("\\begin{figure}[ht]\n")
+%        cat("\\centering\n")
+%        cat(paste0("\\includegraphics{", fn, "}\n"))
+%        cat(sprintf("\\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\\alpha$ = %.1f, $I$ = %d, $m$ = %d, and $N$ = %d.]{\n", alpha, I, m, N))
+%        cat(sprintf("Approximate marginal posterior distributions of BA model parameters obtained by applying \\software{netabc} to a simulated transmission tree with BA parameter values $\\alpha$ = %.1f, $I$ = %d, $m$ = %d, and $N$ = %d.
 Vertical dashed lines indicate true values.
@@
 $x$-axes indicate regions of nonzero prior density.\n", alpha, I, m, N))
-        cat("}\n")
-        cat(paste0("\\label{", lab, "}\n"))
-        cat("\\end{figure}\n")
-        cat("\\clearpage\n\n")
-    }
-    }
-    }
-@
-
-%\chapter{GLM results}
-%<<glm_tables, echo=FALSE, warning=FALSE, error=FALSE, results="asis">>=
-%    format.glm <- function (df) {
-%        df[Parameter == "alpha", Parameter := "$\\alpha$"]
-%        df[,p := paste0("$", sapply(p, pp, eq=FALSE), "$")]
-%        df[,Estimate := paste0("$", sapply(Estimate, pp, eq=FALSE, thres=NA), "$")]
-%        df[,`Standard error` := paste0("$", sapply(`Standard error`, pp, eq=FALSE, thres=NA), "$")]
-%        setnames(df, "p", "$p$-value")
-%        df
+%        cat("}\n")
+%        cat(paste0("\\label{", lab, "}\n"))
+%        cat("\\end{figure}\n")
+%        cat("\\clearpage\n\n")
 %    }
-%
-%    alpha.glm <- format.glm(alpha.glm)
-%    I.glm <- format.glm(I.glm)
-%    #m.glm <- format.glm(m.glm)
-%    #N.glm <- format.glm(N.glm)
-%
-%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-%                        "$\\alpha$ to true values of BA parameters.")
-%    capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
-%                  "inverse link function. Coefficients are interpretable as",
-%                  "additive effects on the inverse of the mean error.")
-%    print(xtable(alpha.glm, label="tab:glmalpha", caption=c(capt, capt.short)),
-%          include.rownames=FALSE, sanitize.text.function=identity,
-%          sanitize.colnames.function=identity)
-%
-%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-%                        "$I$ to true values of BA parameters.")
-%    capt <- paste(capt.short, "GLM was fitted with a gaussian distribution and",
-%                  "inverse link function. Coefficients are interpretable as",
-%                  "additive effects on the inverse of the mean error.")
-%    print(xtable(I.glm, label="tab:glmI", caption=c(capt, capt.short)),
-%          include.rownames=FALSE, sanitize.text.function=identity,
-%          sanitize.colnames.function=identity)
-%
-%    #capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-%    #                    "$m$ to true values of BA parameters.")
-%    #capt <- paste(capt.short, "GLM was fitted with a Poisson distribution and",
-%    #              "log link function. Coefficients are interpretable as",
-%    #              "additive effects on the exponential of the mean error.")
-%    #print(xtable(m.glm, label="tab:glmm", caption=c(capt, capt.short)),
-%    #      include.rownames=FALSE, sanitize.text.function=identity,
-%    #      sanitize.colnames.function=identity)
-%
-%    #capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-%    #                    "$N$ to true values of BA parameters.")
-%    #capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
-%    #              "inverse link function. Coefficients are interpretable as",
-%    #              "additive effects on the inverse of the mean error.")
-%    #print(xtable(N.glm, label="tab:glmN", caption=c(capt, capt.short)),
-%    #      include.rownames=FALSE, sanitize.text.function=identity,
-%    #      sanitize.colnames.function=identity)
+%    }
+%    }
 %@

### 2016-06-21 doc/thesis/appendix.tex
Updated abstract and a bit more
Up to ABC in intro now
Almooooooost!

@@
 
-\clearpage
+\begin{figure}[ht]
+    \includegraphics{zetterberg2004-posterior}
+    \caption[
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to IDU/Estonia data with \software{netabc}.
+    ]{
+        Approximate marginal posterior distributions of \gls{BA} model
+        parameters fitted to IDU/Estonia data with \software{netabc}. Vertical
+        lines indicate posterior means, and shaded areas are 95\% highest
+        posterior density intervals. $x$-axis indicates regions of nonzero
+        prior density.
+    }
+    \label{fig:zetterberg}
+\end{figure}
 
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/0.0_1000_2_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.
+%<<abc_posterior, echo=FALSE, results="asis">>=
+%    N <- 5000
+%    replicate <- 0
+%    for (m in c(2, 3, 4)) {
+%    for (I in c(1000, 2000)) {
+%    for (alpha in c(0, 0.5, 1, 1.5)) {
+%        fn <- sprintf("{abc-posterior/%.1f_%d_%d_%d_%d}.pdf", alpha, I, m, N, replicate)
+%        lab <- sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate)
+%        cat("\\begin{figure}[ht]\n")
+%        cat("\\centering\n")
+%        cat(paste0("\\includegraphics{", fn, "}\n"))
+%        cat(sprintf("\\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\\alpha$ = %.1f, $I$ = %d, $m$ = %d, and $N$ = %d.]{\n", alpha, I, m, N))
+%        cat(sprintf("Approximate marginal posterior distributions of BA model parameters obtained by applying \\software{netabc} to a simulated transmission tree with BA parameter values $\\alpha$ = %.1f, $I$ = %d, $m$ = %d, and $N$ = %d.
 Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:0.0-1000-2-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/0.5_1000_2_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 1000, $m$ = 2, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.5, $I$ = 1000, $m$ = 2, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:0.5-1000-2-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/1.0_1000_2_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.0, $I$ = 1000, $m$ = 2, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:1.0-1000-2-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/1.5_1000_2_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 1000, $m$ = 2, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.5, $I$ = 1000, $m$ = 2, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:1.5-1000-2-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/0.0_2000_2_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 2000, $m$ = 2, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.0, $I$ = 2000, $m$ = 2, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:0.0-2000-2-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/0.5_2000_2_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 2000, $m$ = 2, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.5, $I$ = 2000, $m$ = 2, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:0.5-2000-2-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/1.0_2000_2_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 2000, $m$ = 2, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.0, $I$ = 2000, $m$ = 2, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:1.0-2000-2-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/1.5_2000_2_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 2000, $m$ = 2, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.5, $I$ = 2000, $m$ = 2, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:1.5-2000-2-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/0.0_1000_3_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 1000, $m$ = 3, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.0, $I$ = 1000, $m$ = 3, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:0.0-1000-3-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/0.5_1000_3_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 1000, $m$ = 3, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.5, $I$ = 1000, $m$ = 3, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:0.5-1000-3-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/1.0_1000_3_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 1000, $m$ = 3, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.0, $I$ = 1000, $m$ = 3, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:1.0-1000-3-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/1.5_1000_3_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 1000, $m$ = 3, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.5, $I$ = 1000, $m$ = 3, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:1.5-1000-3-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/0.0_2000_3_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 2000, $m$ = 3, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.0, $I$ = 2000, $m$ = 3, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:0.0-2000-3-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/0.5_2000_3_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 2000, $m$ = 3, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.5, $I$ = 2000, $m$ = 3, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:0.5-2000-3-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/1.0_2000_3_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 2000, $m$ = 3, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.0, $I$ = 2000, $m$ = 3, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:1.0-2000-3-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/1.5_2000_3_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 2000, $m$ = 3, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.5, $I$ = 2000, $m$ = 3, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:1.5-2000-3-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/0.0_1000_4_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 1000, $m$ = 4, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.0, $I$ = 1000, $m$ = 4, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:0.0-1000-4-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/0.5_1000_4_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 1000, $m$ = 4, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.5, $I$ = 1000, $m$ = 4, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:0.5-1000-4-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/1.0_1000_4_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 1000, $m$ = 4, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.0, $I$ = 1000, $m$ = 4, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:1.0-1000-4-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/1.5_1000_4_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 1000, $m$ = 4, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.5, $I$ = 1000, $m$ = 4, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:1.5-1000-4-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/0.0_2000_4_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.0, $I$ = 2000, $m$ = 4, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.0, $I$ = 2000, $m$ = 4, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:0.0-2000-4-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/0.5_2000_4_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 0.5, $I$ = 2000, $m$ = 4, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 0.5, $I$ = 2000, $m$ = 4, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:0.5-2000-4-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/1.0_2000_4_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.0, $I$ = 2000, $m$ = 4, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.0, $I$ = 2000, $m$ = 4, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:1.0-2000-4-5000-0}
-\end{figure}
-\clearpage
-
-\begin{figure}[ht]
-\centering
-\includegraphics{{abc-posterior/1.5_2000_4_5000_0}.pdf}
-\caption[Approximate marginal posterior distributions of BA model parameters for a simulated transmission tree with $\alpha$ = 1.5, $I$ = 2000, $m$ = 4, and $N$ = 5000.]{
-Approximate marginal posterior distributions of BA model parameters obtained by applying \software{netabc} to a simulated transmission tree with BA parameter values $\alpha$ = 1.5, $I$ = 2000, $m$ = 4, and $N$ = 5000.
-Vertical dashed lines indicate true values.
-Shaded areas are 95\% highest posterior density intervals.
-$x$-axes indicate regions of nonzero prior density.
-}
-\label{fig:1.5-2000-4-5000-0}
-\end{figure}
-\clearpage
-
-%\chapter{GLM results}
-%<<glm_tables, echo=FALSE, warning=FALSE, error=FALSE, results="asis">>=
-%    format.glm <- function (df) {
-%        df[Parameter == "alpha", Parameter := "$\\alpha$"]
-%        df[,p := paste0("$", sapply(p, pp, eq=FALSE), "$")]
-%        df[,Estimate := paste0("$", sapply(Estimate, pp, eq=FALSE, thres=NA), "$")]
-%        df[,`Standard error` := paste0("$", sapply(`Standard error`, pp, eq=FALSE, thres=NA), "$")]
-%        setnames(df, "p", "$p$-value")
-%        df
+Shaded areas are 95\\%% highest posterior density intervals.
+$x$-axes indicate regions of nonzero prior density.\n", alpha, I, m, N))
+%        cat("}\n")
+%        cat(paste0("\\label{", lab, "}\n"))
+%        cat("\\end{figure}\n")
+%        cat("\\clearpage\n\n")
 %    }
-%
-%    alpha.glm <- format.glm(alpha.glm)
-%    I.glm <- format.glm(I.glm)
-%    #m.glm <- format.glm(m.glm)
-%    #N.glm <- format.glm(N.glm)
-%
-%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-%                        "$\\alpha$ to true values of BA parameters.")
-%    capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
-%                  "inverse link function. Coefficients are interpretable as",
-%                  "additive effects on the inverse of the mean error.")
-%    print(xtable(alpha.glm, label="tab:glmalpha", caption=c(capt, capt.short)),
-%          include.rownames=FALSE, sanitize.text.function=identity,
-%          sanitize.colnames.function=identity)
-%
-%    capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-%                        "$I$ to true values of BA parameters.")
-%    capt <- paste(capt.short, "GLM was fitted with a gaussian distribution and",
-%                  "inverse link function. Coefficients are interpretable as",
-%                  "additive effects on the inverse of the mean error.")
-%    print(xtable(I.glm, label="tab:glmI", caption=c(capt, capt.short)),
-%          include.rownames=FALSE, sanitize.text.function=identity,
-%          sanitize.colnames.function=identity)
-%
-%    #capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-%    #                    "$m$ to true values of BA parameters.")
-%    #capt <- paste(capt.short, "GLM was fitted with a Poisson distribution and",
-%    #              "log link function. Coefficients are interpretable as",
-%    #              "additive effects on the exponential of the mean error.")
-%    #print(xtable(m.glm, label="tab:glmm", caption=c(capt, capt.short)),
-%    #      include.rownames=FALSE, sanitize.text.function=identity,
-%    #      sanitize.colnames.function=identity)
-%
-%    #capt.short <- paste("Parameters of a fitted GLM relating error in estimated",
-%    #                    "$N$ to true values of BA parameters.")
-%    #capt <- paste(capt.short, "GLM was fitted with a Gaussian distribution and",
-%    #              "inverse link function. Coefficients are interpretable as",
-%    #              "additive effects on the inverse of the mean error.")
-%    #print(xtable(N.glm, label="tab:glmN", caption=c(capt, capt.short)),
-%    #      include.rownames=FALSE, sanitize.text.function=identity,
-%    #      sanitize.colnames.function=identity)
+%    }
+%    }
 %@

### 2016-06-21 doc/thesis/conclusion.tex
Almooooooost!

@@
 colgate1989risk}. Of the four parameters we considered, the preferential
-attachment power $\alpha$ was the most readily estimable. Estimating $\alpha$
-with traditional epidemiological methods is challenging due to the requirement
-of sampling the high-degree nodes making up the tail of the power law
-distribution, although approaches such as respondent-driven
-sampling~\autocite{heckathorn1997respondent} may be effective.
+attachment power \gls{alpha} was the most readily estimable. Estimating
+\gls{alpha} with traditional epidemiological methods is challenging due to the
+requirement of sampling the high-degree nodes making up the tail of the power
+law distribution, although approaches such as respondent-driven
+sampling~\autocite{heckathorn1997respondent} \add{or collection of longitudinal
+partner count data~\autocite{de2007preferential}} may be effective.
 

### 2016-06-21 doc/thesis/discussion.Rnw
Almooooooost!

@@
-%<<setup, include=FALSE>>=
-%    source("global.R")
-%@
+<<setup, include=FALSE>>=
+    source("global.R")
+@
 
@@
 distributions with similar fitted \gls{gamma} values appear qualitatively
-similar in other respects.} The \gls{gamma} values fitted to \gls{alpha} = 0
-and \gls{alpha} = 0.5 are nearly identical (about 2.28 for \gls{alpha} = 0 and
-2.33 for \gls{alpha} = 0.5 with \gls{N} = 5000 and \gls{m} = 2). In other
-words, the degree distributions of networks with \gls{alpha} $< 1$ are similar
-to each other, which may result in similarity of corresponding transmission
-trees as well.
+similar in other respects.} The \gls{gamma} values fitted to $\gls{alpha} = 0$
+and $\gls{alpha} = 0.5$ are nearly identical (about $2.28 for \gls{alpha} = 0$
+and $2.33 for \gls{alpha} = 0.5$ with $\gls{N} = 5000$ and $\gls{m} = 2$). In
+other words, the degree distributions of networks with \gls{alpha} $< 1$ are
+similar to each other, which may result in similarity of corresponding
+transmission trees as well.
 
@@
 the grid search estimates of \gls{I} were accurate and unbiased
-(\cref{fig:gridptI}). \add{We also note the resemblance of the contour surface
-of \cref{fig:derivs} to the two-dimensional marginal posterior distributions on
-\gls{I} and \gls{N} obtained with simulated data
-(\cref{fig:abclow2,fig:abchigh2}). However, while fixing \gls{N} and any other
-parameter resulted in much more accurate estimates of \gls{I}, fixing \gls{N}
-alone did not (\cref{fig:marginal}). Therefore, there may be more complex
-parameter dependence in the \gls{BA} model which we have not considered here.}
+(\cref{fig:gridest,fig:gridptI}). \add{We also note the resemblance of the
+contour surface of \cref{fig:derivs} to the two-dimensional marginal posterior
+distributions on \gls{I} and \gls{N} obtained with simulated data
+(\cref{fig:abclow2,fig:abchigh2}). In addition, both the accuracy and precision
+of the \gls{ABC} estimates of \gls{I} improved substantially when \gls{N} was
+marginalized out (\cref{fig:marginal}).}
 
@@
   \centering
-  %\includegraphics{derivatives.pdf}
+  \includegraphics{derivatives.pdf}
   \caption[
@@
 
-\end{document}
-
 \subsubsection{Preferential attachment power is sub-linear and higher for \gls{IDU}}
@@
 estimate the parameters of a more sophisticated \acrlong{PA} model
-incorporating heterogeneous node behaviour. They found \gls{alpha} values
-ranging from 0.26 to 0.62, depending on the gender and time period considered. 
-Our estimates of \gls{alpha} for the \textcite{niculescu2015recent} was above
-this range under both priors, as were the estimates for the
-\textcite{wang2015targeting} data and the BC data when $\gls{m} = 1$ was
-disallowed by the prior. The dataset investigated by
+incorporating heterogeneous node behaviour. When used to analyse
+population-level longitudinal partner count data, they found \gls{alpha} values
+ranging from 0.26 to 0.62 depending on the gender and time period considered.
+Our estimates of \gls{alpha} for the IDU/Romania was above this range under
+both priors, as were the estimates for the MSM/Beijing data and the BC data
+when $\gls{m} = 1$ was disallowed by the prior. The dataset investigated by
 \textcite{de2007preferential} was derived from a survey of a random sample of
@@
 
-Both \textcite{de2007preferential} and \textcite{novitsky2014impact} studied
+\add{Both \textcite{de2007preferential} and the HET/Botswana data studied
 populations whose primary risk factor for \gls{HIV} infection was heterosexual
@@
 \gls{alpha} for these data was 
-    X%\S%expr{d[m_min == 1 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]}
+    \Sexpr{d[m_min == 1 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]}
         or 
-    X%\S%expr{d[m_min == 2 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]},
-depending on the prior on \gls{m}; \textcite{de2007preferential} estimated
-0.54, 0.57, and 0.29 for 3-year, 5-year, and lifetime partnership networks
-respectively for the female portion of their sample.
+    \Sexpr{d[m_min == 2 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]},
+depending on the prior on \gls{m}. Similarly, \textcite{de2007preferential}
+estimated 0.54, 0.57, and 0.29 for 3-year, 5-year, and lifetime partnership
+networks respectively for the female portion of their sample.}
 
-For both choices of prior on \gls{m}, the datasets derived from \gls{IDU}
+\add{For both choices of prior on \gls{m}, the datasets derived from \gls{IDU}
 populations had a higher estimated \acrlong{PA} power than the other datasets
@@
 in Brooklyn, USA, collected between 1991 and
-1993~\autocite{friedman2006social}. They found that the the \gls{IDU} network 
+1993~\autocite{friedman2006social}. They found that the \gls{IDU} network
 resembled a \gls{BA} network much more closely than other social and sexual 
@@
 community members for advice or supplies, thus increasing the connectivity of
-these high-degree nodes.
+these high-degree nodes.}
 
@@
 
-%One somewhat surprising result was the difference between parameter estimates 
-%for the \textcite{li2015hiv} and \textcite{wang2015targeting} datasets. Both
-%groups studied cohorts of acutely infected \gls{MSM} in major Chinese cities
-%%(Shanghai and Beijing respectively); yet, the \citeauthor{li2015hiv} data was 
-%estimated to have a lower \acrlong{PA} power and larger infected population
-%than the \textcite{wang2015targeting} data\ldots
+\subsubsection{Power law exponent is within reported range, but hard to interpret}
 
-\subsubsection{Power law exponent}
-
-% for discussion
-In order to compare our results to existing literature on networks and
+\add{In order to compare our results to existing literature on networks and
 distributions of partner counts, we have reported estimated values for the
@@
 would be straightforward and seems a natural next step toward improving the
-realism of the \gls{BA} model.
+realism of the \gls{BA} model.}
 
-\cref{tab:litgamma} shows some estimated \gls{gamma} values reported in the
-network literature.  The values mainly fall between 2 and 4, although
-substantial variation within this range exists for networks with apparently
-similar demographics. Some of these differences are likely due to the differing
-time intervals considered in each study -- some investigate the distribution of
-lifetime partners, while others consider only recent partnerships, such as in
-the past year. TODO
+\add{\Cref{tab:litgamma} shows some estimated \gls{gamma} values reported in the
+network literature.  Our estimates are most similar to those of
+\textcite{liljeros2001web, rothenberg2007large}, but interpretation of the
+exact values is challenging. Estimates from the literature mainly fall between
+2 and 4, although substantial variation within this range exists, even for
+networks with apparently similar demographics. For example,
+\textcite{schneeberger2004scale} estimate higher \gls{gamma} values for
+heterosexual females than heterosexual males, while
+\textcite{clemenccon2015statistical} find the opposite. Some of these
+differences are likely due to the differing time intervals considered in each
+study -- some investigate the distribution of lifetime partners, while others
+consider only recent partnerships, such as in the past year. However, the
+effect of the time period on \gls{gamma} is not readily obvious from the
+available estimates.}
 
@@
         participants were not selected on the basis of sexual orientation;
-        these values can be assumed to reflect primarily heterosexual dynamics.
+        these values likely reflect primarily heterosexual dynamics.
         \textcite{brown2011transmission} studied a phylogenetically estimated
@@
     }
-    \label{tab:kernelexpt}
+    \label{tab:litgamma}
 \end{table}
 
-%For both priors, the estimated prevalence was extremely high, in fact higher
-%than the estimated HIV prevalence in the sampled region. The authors indicated
-%that the source of the samples was a town in close proximity to the country's
-%capital city, and suggested that there may have been a high degree of migration
-%and partner interchange between the two locations. It is possible that the
-%contact network underlying the subtree we investigated includes a much larger
-%group based in the capital city, which would explain the high estimate of $I$.
-%There is no clear explanation for the discrepancy between the two priors for
-%the \textcite{li2015hiv} data, as the subset we analyzed formed a phylogenetic
-%cluster and therefore was a good candidate for the BA model. However, nearly
-%all the posterior density was assigned to $m = 1$ when this value was allowed,
-%indicating that the network was more likely to have an acyclic tree structure.
-
-\subsubsection{HIV prevalence}
+\subsubsection{HIV prevalence estimates are often discordant with epidemiological information}
 
@@
 epidemiological data was not sufficient to estimate the true prevalence. Since
-the estimates of the prevalence \gls{I} were largely robust to the choice of
-prior on \gls{m} (\cref{fig:abchpd,fig:abchpdm2}), we discuss here only the
-results obtained with the prior $\gls{m} \sim \DiscreteUniform(1, 5)$.}
+the estimates of \gls{I} were largely robust to the choice of prior on \gls{m}
+(\cref{fig:abchpd,fig:abchpdm2}), we discuss here only the results obtained
+with the prior $\gls{m} \sim \DiscreteUniform(1, 5)$.}
 
 \add{The prevalence estimates for two \gls{IDU} epidemics in Eastern Europe,
-IDU/Estonia and IDU/Romania, were similar 
-    (X and X), 
-although the confidence interval around the \textcite{niculescu2015recent} data
-was wider. \textcite{niculescu2015recent} reported that 494 new \gls{HIV}
-infections were diagnosed among the studied population (\glspl{IDU} in
-Bucharest) during the study period, and this value was contained in our 95\%
-\gls{HPD} 
-    (X-X). 
-On the other hand, \textcite{zetterberg2004two} collected samples from several
-locations throughout Estonia; using their reported incidence estimates, there
-were approximately 2000 new infections in the country during the study period.
-This value was outside our 95\% \gls{HPD} 
-    (X-X). 
+IDU/Romania and IDU/Estonia, were 
+    \Sexpr{d1[dataset == "niculescu2015" & parameter == "I", round(mean)]} and 
+    \Sexpr{d1[dataset == "zetterberg2004" & parameter == "I", round(mean)]}.
+The authors of the IDU/Romania study~\autocite{niculescu2015recent} reported
+that 494 new \gls{HIV} infections were diagnosed among the studied population
+(\glspl{IDU} in Bucharest) during the study period, and this value was
+contained in our 95\% \gls{HPD} interval
+    (\Sexpr{d1[dataset == "niculescu2015" & parameter == "I", round(hpd.min.95)]} -
+    \Sexpr{d1[dataset == "niculescu2015" & parameter == "I", round(hpd.max.95)]}).
+On the other hand, the IDU/Estonia data were collected from several locations
+in the country. Using their reported incidence estimates, there were
+approximately 2000 new infections in the country during the study period, a
+value outside our 95\% \gls{HPD} interval
+    (\Sexpr{d1[dataset == "zetterberg2004" & parameter == "I", round(hpd.min.95)]} -
+    \Sexpr{d1[dataset == "zetterberg2004" & parameter == "I", round(hpd.max.95)]}).
 The authors indicated that their sample was ``convenience based''; it is
@@
 sampled individuals (399) comprise the majority of \gls{HIV}-positive network
-members. TODO}
+members. This value was the lower bound of our estimated 95\% \gls{HPD}
+interval
+    (\Sexpr{d1[dataset == "bctree" & parameter == "I", round(hpd.min.95)]} -
+    \Sexpr{d1[dataset == "bctree" & parameter == "I", round(hpd.max.95)]}),
+but it seems likely that the posterior mean of
+    \Sexpr{d1[dataset == "bctree" & parameter == "I", round(mean)]}
+is an overestimate.}
 
-\add{Surprisingly, the prevalence estimates for two \gls{MSM} populations major
-Chinese cities (MSM/Shanghai and MSM/Beijing) were quite different
-    (X and X).
+\add{The prevalence estimates for two \gls{MSM} populations in major Chinese
+cities (MSM/Beijing and MSM/Shanghai) were 
+    \Sexpr{d1[dataset == "wang2015" & parameter == "I", round(mean)]} and
+    \Sexpr{d1[dataset == "li2015" & parameter == "I", round(mean)]}.
 The individuals sampled for the MSM/Beijing study were recruited by following a
@@
 cohort during the study period, which is much lower than the estimated
-prevalence obtained with \software{netabc}
-    (X).
-However, the authors did not claim to have sampled the entire sexual network.
-In a nationwide survey, \textcite{wu2013hiv} estimated that there were 24,198
-\gls{MSM} living in Beijing, of whom 5.7\%, or 1379, were \gls{HIV}-infected,
-which is within the 95\% confidence interval of our estimate of \gls{I}
-    (X-X).
+prevalence obtained with \software{netabc}. However, the authors did not claim
+to have sampled the entire sexual network. In a nationwide survey,
+\textcite{wu2013hiv} estimated that there were 24,198 \gls{MSM} living in
+Beijing, of whom 5.7\%, or 1379, were \gls{HIV}-infected, which is close to the
+upper limit of the 95\% confidence interval of our estimate of \gls{I}
+    (\Sexpr{d1[dataset == "wang2015" & parameter == "I", round(hpd.min.95)]} -
+    \Sexpr{d1[dataset == "wang2015" & parameter == "I", round(hpd.max.95)]}).
 The same survey estimated the size of the \gls{MSM} population in Shanghai
@@
 \software{netabc} for the MSM/Shanghai data
-    (X-X). 
+    (\Sexpr{d1[dataset == "li2015" & parameter == "I", round(hpd.min.95)]} -
+    \Sexpr{d1[dataset == "li2015" & parameter == "I", round(hpd.max.95)]}).
 It is worth noting that the authors of the IDU/Shanghai
@@
 individuals) were \gls{MSM}. This value is well outside of our 95\% \gls{HPD} 
-    (X-X),
+    (\Sexpr{d1[dataset == "kao2011" & parameter == "I", round(hpd.min.95)]} -
+    \Sexpr{d1[dataset == "kao2011" & parameter == "I", round(hpd.max.95)]}),
 suggesting that the entire \gls{MSM} population does not comprise a single
 network. In fact, our estimate of \gls{I} for these data
-    (X)
-was closer to the number of \gls{MSM} enrolled in the study, 301. The study
-focused exclusively on newly diagnosed individuals; perhaps chronically
-infected \gls{MSM} were no longer part of the same network. On the other hand,
-our estimated prevalence for the MSM/USA data
-    (X)
+    (\Sexpr{d1[dataset == "kao2011" & parameter == "I", round(mean)]}),
+was closer to the number of \gls{MSM} enrolled in the study (301) than to the
+epidemiologically estimated prevalence. The study focused exclusively on newly
+diagnosed individuals. On the other hand, our estimated prevalence for the
+MSM/USA data
+    (\Sexpr{d1[dataset == "little2014" & parameter == "I", round(mean)]})
 was clearly an underestimate. The study enrolled almost 648 \gls{HIV} infected 
@@
 value was contained in the 95\% \gls{HPD} interval
-    (X-X).}
+    (\Sexpr{d1[dataset == "little2014" & parameter == "I", round(hpd.min.95)]} -
+    \Sexpr{d1[dataset == "little2014" & parameter == "I", round(hpd.max.95)]}).}
 
@@
 district were enrolled in the study, it is likely that this number represents
-the majority of \gls{HIV} infected persons. Our estimate was slightly (TODO:
-lower/higher)
-    (X),
+the majority of \gls{HIV} infected persons. Our estimate was slightly lower
+    (\Sexpr{d1[dataset == "grabowski2014" & parameter == "I", round(mean)]}),
 but the value 1434 was contained ith the 95\% \gls{HPD} interval
-    (X-X).
+    (\Sexpr{d1[dataset == "grabowski2014" & parameter == "I", round(hpd.min.95)]} -
+    \Sexpr{d1[dataset == "grabowski2014" & parameter == "I", round(hpd.max.95)]}).
 The HET/Malawi data represents possibly the oldest population-level \gls{HIV}
@@
 a total of 200 were \gls{HIV}-positive. Although the HET/Malawi data had the
-lowest estimated \gls{I} value among the heterosexual datasets, and 200 was
-contained within the 95\% \gls{HPD} interval
-    (X-X),
-the estimated prevalence was much higher than this value (X). Due to the
-thorough population-level screening, it is most likely that the posterior mean
-represents an overestimate, especially considering the observed upward bias of
-\gls{ABC} estimates of \gls{I} observed on simulated data (\cref{fig:abcptm2}).
-However, it is also possible that the true sexual network extended outside the
-study's geographic area to include a larger number of infected individuals.
-The authors of the HET/Botswana data~\autocite{novitsky2013phylogenetic, 
+lowest estimated \gls{I} value among the heterosexual datasets, 
+the posterior mean
+    (\Sexpr{d1[dataset == "mccormack2002" & parameter == "I", round(mean)]}),
+was much higher than this value. Due to the thorough population-level
+screening, it is most likely that the posterior mean represents an
+overestimate, especially considering the observed upward bias of \gls{ABC}
+estimates of \gls{I} observed on simulated data (\cref{fig:abcptm2}).
+It is also possible that the true sexual network extended outside the study's
+geographic area to include a larger number of infected individuals. The authors
+of the HET/Botswana data~\autocite{novitsky2013phylogenetic,
 novitsky2014impact} estimated that there were 1731 \gls{HIV}-positive
@@
 was much higher 
-    (X), 
-with a 95\% \gls{HPD} interval spanning nearly the entire prior region. Since
-the total population of the study area was approximately 15,000, this value of
-\gls{I} seems unlikely to be accurate. However, the authors noted that the town
-where the study took place was located proximally to the country's capital
-city, and suggested that frequent travel between the two locations may have
-facilitated linking of their sexual networks. Thus, the high estimate of
-\gls{I} we obtained for these data may include a larger network component based
-in the capital city.}
+    (\Sexpr{d1[dataset == "novitsky2014" & parameter == "I", round(mean)]}),
+with a 95\% \gls{HPD} interval spanning nearly the entire prior region. The
+authors noted that the town where the study took place was located proximally
+to the country's capital city, and suggested that frequent travel between the
+two locations may have facilitated linking of their sexual networks. Thus, the
+high estimate of \gls{I} we obtained for these data may include a larger
+network component based in the capital city.}
 
@@
 total of 2120. This is substantially greater than the point estimate of 
-    X 
+    \Sexpr{d1[dataset == "cuevas2009" & parameter == "I", round(mean)]}
 we obtained with \gls{ABC}. This error may in part be due to model
 misspecification -- it is highly unlikely that a mixed risk group population 
-would form a single, highly-connected contact network of the type generated by
+would form a single, well-connected contact network of the type generated by
 the \gls{BA} model.}

### 2016-06-21 doc/thesis/discussion.tex
Almooooooost!

@@
-%<<setup, include=FALSE>>=
-%    source("global.R")
-%@
-
 \label{sec:disc}
@@
 identifiable within their own band. However, when \gls{N} is fixed at 5000, the
-derivatives vary substantially along the \gls{I}-axis, which explains why the
-grid search estimates of \gls{I} were accurate and unbiased
-(\cref{fig:gridptI}).
+derivatives vary substantially along the \gls{I}-axis, which explains why
+the grid search estimates of \gls{I} were accurate and unbiased
+(\cref{fig:gridest,fig:gridptI}). \add{We also note the resemblance of the
+contour surface of \cref{fig:derivs} to the two-dimensional marginal posterior
+distributions on \gls{I} and \gls{N} obtained with simulated data
+(\cref{fig:abclow2,fig:abchigh2}). In addition, both the accuracy and precision
+of the \gls{ABC} estimates of \gls{I} improved substantially when \gls{N} was
+marginalized out (\cref{fig:marginal}).}
 
@@
 
-Both \textcite{de2007preferential} and \textcite{novitsky2014impact} studied
+\add{Both \textcite{de2007preferential} and \textcite{novitsky2014impact} studied
 populations whose primary risk factor for \gls{HIV} infection was heterosexual
@@
     X%\S%expr{d[m_min == 2 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]},
-depending on the prior on \gls{m}; \citeauthor{de2007preferential} estimated
+depending on the prior on \gls{m}; \textcite{de2007preferential} estimated
 0.54, 0.57, and 0.29 for 3-year, 5-year, and lifetime partnership networks
-respectively for the female portion of their sample.
+respectively for the female portion of their sample.}
 
-For both choices of prior on \gls{m}, the datasets derived from \gls{IDU}
+\add{For both choices of prior on \gls{m}, the datasets derived from \gls{IDU}
 populations had a higher estimated \acrlong{PA} power than the other datasets
@@
 community members for advice or supplies, thus increasing the connectivity of
-these high-degree nodes.
+these high-degree nodes.}
 
-One somewhat surprising result was the difference between parameter estimates 
-for the \textcite{li2015hiv} and \textcite{wang2015targeting} datasets. Both
-groups studied cohorts of acutely infected \gls{MSM} in major Chinese cities
-(Shanghai and Beijing respectively); yet, the \citeauthor{li2015hiv} data was 
-estimated to have a lower \acrlong{PA} power and larger infected population
-than the \textcite{wang2015targeting} data\ldots
+\add{Unfortunately, the sub-linear region for \gls{alpha} identified by both
+\textcite{de2007preferential} and \software{netabc} is also the region of
+poorest identifiability (\cref{fig:gridest,fig:abcptm2}). This was reflected in
+the high level of uncertainty in the estimates, with most 95\% \gls{HPD}
+intervals covering the majority of the range [0, 1]. The value $\gls{alpha} =
+0.5$ was contained in the 95\% \gls{HPD} interval for every dataset;
+consequently, it is not possible to say with high confidence that any of the
+\gls{alpha} values are different from each other. In synthetic data, the
+confidence intervals around \gls{alpha} narrowed when other parameters were
+marginalized out (\cref{fig:maginal}). Thus, it is possible that estimates of
+\gls{alpha} could be made more precise by specifying either exact values or
+informative priors on the other \gls{BA} parameters. We argued informally in
+\cref{subsec: whyba} that \gls{I}, \gls{m}, and \gls{N} can sometimes be
+estimated by standard epidemiological methods, although the difficulty may
+depend on the specific epidemic under consideration.}
 
-\subsubsection{Power law exponent}
+\subsubsection{Power law exponent is within reported range, but hard to interpret}
 
-% for discussion
-In order to compare our results to existing literature on networks and
+\add{In order to compare our results to existing literature on networks and
 distributions of partner counts, we have reported estimated values for the
@@
 would be straightforward and seems a natural next step toward improving the
-realism of the \gls{BA} model.
+realism of the \gls{BA} model.}
 
-\textcite{rothenberg2007large} estimated the power law exponents for 15 major
-network studies of sexual and injection drug use networks carried out between
-1981 and 2000. The estimated \gls{gamma} values for ``all contacts'' (that is,
-including both interviewed and non-interviewed individuals) were 
-between 2.08 and 2.87 for all but one of the networks\ldots
+\add{\Cref{tab:litgamma} shows some estimated \gls{gamma} values reported in the
+network literature.  Our estimates are most similar to those of
+\textcite{liljeros2001web, rothenberg2007large}, but interpretation of the
+exact values is challenging. Estimates from the literature mainly fall between
+2 and 4, although substantial variation within this range exists, even for
+networks with apparently similar demographics. For example,
+\textcite{schneeberger2004scale} estimate higher \gls{gamma} values for
+heterosexual females than heterosexual males, while
+\textcite{clemenccon2015statistical} find the opposite. Some of these
+differences are likely due to the differing time intervals considered in each
+study -- some investigate the distribution of lifetime partners, while others
+consider only recent partnerships, such as in the past year. However, the
+effect of the time period on \gls{gamma} is not readily obvious from the
+available estimates.}
 
-%For both priors, the estimated prevalence was extremely high, in fact higher
-%than the estimated HIV prevalence in the sampled region. The authors indicated
-%that the source of the samples was a town in close proximity to the country's
-%capital city, and suggested that there may have been a high degree of migration
-%and partner interchange between the two locations. It is possible that the
-%contact network underlying the subtree we investigated includes a much larger
-%group based in the capital city, which would explain the high estimate of $I$.
-%There is no clear explanation for the discrepancy between the two priors for
-%the \textcite{li2015hiv} data, as the subset we analyzed formed a phylogenetic
-%cluster and therefore was a good candidate for the BA model. However, nearly
-%all the posterior density was assigned to $m = 1$ when this value was allowed,
-%indicating that the network was more likely to have an acyclic tree structure.
+\begin{table}[ht]
+    \centering
+    \input{\tablepath/lit-gamma}
+    \caption[
+        Values of power law exponent \gls{gamma} previously reported in the
+        literature.
+    ]{
+        Values of power law exponent \gls{gamma} previously reported in the
+        literature. In the risk group column, ``mixed sexual'' indicates
+        participants were not selected on the basis of sexual orientation;
+        these values likely reflect primarily heterosexual dynamics.
+        \textcite{brown2011transmission} studied a phylogenetically estimated
+        transmission network, rather than a contact network.
+        The values presented by \textcite{jones2003assessment} were in the
+        context of arguing that the power law distribution is not suitable for
+        use with network data.
+    }
+    \label{tab:litgamma}
+\end{table}
 
-\subsubsection{HIV prevalence}
+\subsubsection{HIV prevalence estimates are often discordant with epidemiological information}
 
-{\color{blue}\uline{The true \gls{HIV} prevalence in a population can be
+\add{The true \gls{HIV} prevalence in a population can be
 difficult to estimate for several reasons. \gls{HIV}-infected individuals may
@@
 (\cref{fig:Icrossv,fig:Ikpca}). In real world studies, the proportion of
-infected individuals sampled is usually low, whic may imede our ability to
-estimate \gls{I}.
+infected individuals sampled is usually low, which may impede our ability to
+estimate \gls{I}. In the following, we discuss any information available about
+the \gls{HIV} prevalence in the analysed study populations, and relate this to
+our estimates. However, due to the aforementioned challenges, this information
+cannot be taken as absolute ``ground truth''. Discrepancies between our results
+and the available data may indicate that the estimates produced by
+\software{netabc} are inaccurate, but they may also indicate that the available
+epidemiological data was not sufficient to estimate the true prevalence. Since
+the estimates of \gls{I} were largely robust to the choice of prior on \gls{m}
+(\cref{fig:abchpd,fig:abchpdm2}), we discuss here only the results obtained
+with the prior $\gls{m} \sim \DiscreteUniform(1, 5)$.}
 
-The estimates of the prevalence \gls{I} were largely robust to the choice of
-prior on \gls{m}} (\cref{fig:abchpd,fig:abchpdm2}).
+\add{The prevalence estimates for two \gls{IDU} epidemics in Eastern Europe,
+IDU/Romania and IDU/Estonia, were 
+    747 and 
+    372.
+The authors of the IDU/Romania study~\autocite{niculescu2015recent} reported
+that 494 new \gls{HIV} infections were diagnosed among the studied population
+(\glspl{IDU} in Bucharest) during the study period, and this value was
+contained in our 95\% \gls{HPD} interval
+    (136 -
+    2379).
+On the other hand, the IDU/Estonia data were collected from several locations
+in the country. Using their reported incidence estimates, there were
+approximately 2000 new infections in the country during the study period, a
+value outside our 95\% \gls{HPD} interval
+    (188 -
+    834).
+The authors indicated that their sample was ``convenience based''; it is
+possible that other, harder-to-reach individuals had no contacts with those
+sampled, reducing the apparent network size. In contrast, the IDU/Canada data
+was collected primarily from \gls{IDU} in Vancouver's downtown east side, a
+population that has been intensively targeted by public health interventions.
+Combined with the routine clinical genotyping performed in British Columbia for
+all new \gls{HIV} infections, it is not unreasonable to suppose that the
+sampled individuals (399) comprise the majority of \gls{HIV}-positive network
+members. This value was the lower bound of our estimated 95\% \gls{HPD}
+interval
+    (399 -
+    2862),
+but it seems likely that the posterior mean of
+    1481
+is an overestimate.}
 
-\textcite{niculescu2015recent} and \textcite{zetterberg2004two} both reported 
-on epidemics in \gls{IDU} populations in Eastern Europe (Romania and Estonia,
-respectively). The point estimates of \gls{I} for these two datasets were
-similar (X and X), although the confidence interval around the
-\textcite{niculescu2015recent} data was wider. \textcite{niculescu2015recent}
-\uline{reported that 494 new \gls{HIV} infections were diagnosed among the
-studied population (\glspl{IDU} in Bucharest) during the study period, and this
-value was contained in our 95\% \gls{HPD} (X-X). On the other hand,
-\textcite{zetterberg2004two} collected samples from several locations
-throughout Estonia; using their reported incidence estimates, there were
-approximately 2000 new infections in the country during the study period. This
-value was outside our 95\% \gls{HPD} (X-X). The authors indicated that their
-sample was ``convenience based''. It is possible that other, harder-to-reach
-individuals had no contacts with those sampled, which would reduce the apparent
-network size. Alternatively, the low estimated prevalence may have been due to
-the low sampling fraction.}
+\add{The prevalence estimates for two \gls{MSM} populations in major Chinese
+cities (MSM/Beijing and MSM/Shanghai) were 
+    676 and
+    1391.
+The individuals sampled for the MSM/Beijing study were recruited by following a
+prospective cohort of size 2000. 179 new infections were identified in the
+cohort during the study period, which is much lower than the estimated
+prevalence obtained with \software{netabc}. However, the authors did not claim
+to have sampled the entire sexual network. In a nationwide survey,
+\textcite{wu2013hiv} estimated that there were 24,198 \gls{MSM} living in
+Beijing, of whom 5.7\%, or 1379, were \gls{HIV}-infected, which is close to the
+upper limit of the 95\% confidence interval of our estimate of \gls{I}
+    (175 -
+    1401).
+The same survey estimated the size of the \gls{MSM} population in Shanghai
+14,511, with an \gls{HIV} prevalence of 6.8\% or approximately 987 individuals.
+This was within the 95\% confidence interval of the estimate obtained with
+\software{netabc} for the MSM/Shanghai data
+    (311 -
+    2822).
+It is worth noting that the authors of the IDU/Shanghai
+study~\autocite{li2015hiv} claimed that there were 80,000 \gls{MSM} living in
+Shanghai, which is a significantly different estimate than that obtained by
+\textcite{wu2013hiv}. The authors of the MSM/Taiwan
+study~\autocite{kao2011surveillance} stated that there were 18,378
+\gls{HIV}-infected individuals in Taiwan, of which approximately 44\% (8086
+individuals) were \gls{MSM}. This value is well outside of our 95\% \gls{HPD} 
+    (268 -
+    772),
+suggesting that the entire \gls{MSM} population does not comprise a single
+network. In fact, our estimate of \gls{I} for these data
+    (524),
+was closer to the number of \gls{MSM} enrolled in the study (301) than to the
+epidemiologically estimated prevalence. The study focused exclusively on newly
+diagnosed individuals. On the other hand, our estimated prevalence for the
+MSM/USA data
+    (482)
+was clearly an underestimate. The study enrolled almost 648 \gls{HIV} infected 
+individuals, which should have been a lower bound on \gls{I}, although the
+value was contained in the 95\% \gls{HPD} interval
+    (180 -
+    1112).}
 
-\uline{The cluster we analysed within the BC data consists primarily of
-\gls{IDU} in Vancouver's downtown east side. This population has been
-intensively targeted by public health interventions. Combined with the routine 
-clinical genotyping performed in BC for all new \gls{HIV} infections, it is not
-unreasonable to suppose that the sampled individuals (399) comprise the
-majority of \gls{HIV}-positive network members.
+\add{The HET/Uganda, HET/Malawi, and HET/Botswana data all originated from
+populations in sub-Saharan Africa, where the dominant mode of transmission is
+believed to be heterosexual. The HET/Uganda study population included
+1434 seropositive individuals. Since the majority of inhabitants of the study
+district were enrolled in the study, it is likely that this number represents
+the majority of \gls{HIV} infected persons. Our estimate was slightly lower
+    (939),
+but the value 1434 was contained ith the 95\% \gls{HPD} interval
+    (226 -
+    2661).
+The HET/Malawi data represents possibly the oldest population-level \gls{HIV}
+sequencing effort, carried out on samples collected for other purposes between
+1981 and 1989. Every individual in the study district was sampled; among these,
+a total of 200 were \gls{HIV}-positive. Although the HET/Malawi data had the
+lowest estimated \gls{I} value among the heterosexual datasets, 
+the posterior mean
+    (919),
+was much higher than this value. Due to the thorough population-level
+screening, it is most likely that the posterior mean represents an
+overestimate, especially considering the observed upward bias of \gls{ABC}
+estimates of \gls{I} observed on simulated data (\cref{fig:abcptm2}).
+It is also possible that the true sexual network extended outside the study's
+geographic area to include a larger number of infected individuals. The authors
+of the HET/Botswana data~\autocite{novitsky2013phylogenetic,
+novitsky2014impact} estimated that there were 1731 \gls{HIV}-positive
+individuals in the study area. \gls{HIV} sequences were obtained from
+approximately 70\% of these individuals. The estimated prevalence we obtained
+was much higher 
+    (5432),
+with a 95\% \gls{HPD} interval spanning nearly the entire prior region. The
+authors noted that the town where the study took place was located proximally
+to the country's capital city, and suggested that frequent travel between the
+two locations may have facilitated linking of their sexual networks. Thus, the
+high estimate of \gls{I} we obtained for these data may include a larger
+network component based in the capital city.}
 
-\textcite{wang2015targeting} and \textcite{li2015hiv} studied \gls{MSM}
-populations in major Chinese cities (Shanghai and Beijing).
+\add{\textcite{cuevas2009hiv} reported that 620 newly infected and 1500
+chronically infected patients had been sampled during the study period, for a
+total of 2120. This is substantially greater than the point estimate of 
+    702
+we obtained with \gls{ABC}. This error may in part be due to model
+misspecification -- it is highly unlikely that a mixed risk group population 
+would form a single, well-connected contact network of the type generated by
+the \gls{BA} model.}
 
-\uline{The individuals sampled by} \textcite{wang2015targeting} \uline{were
-recruited by following a prospective cohort of 2000 \gls{MSM} in Beijing,
-China. Of the 2000, 179 new infections were identified during the study period.
-This number is much lower than the estimated prevalence of X obtained with
-\software{netabc}, however the authors did not claim to have sampled the entire
-sexual network. In a nationwide survey,} \textcite{wu2013hiv} \uline{estimated
-that there were 24,198 \gls{MSM} living in Beijing, of whom 5.7\%, or 1379,
-were \gls{HIV}-infected, which is within the 95\% confidence interval of our
-estimate of \gls{I}.}
-
-\textcite{li2015hiv} \uline{studied an \gls{MSM} population in Shanghai, China.
-Despite the apparent similarity to the context of the
-}\textcite{wang2015targeting} \uline{study, the estimated \gls{HIV} prevalence
-of the } \citeauthor{li2015hiv} \uline{data was higher (X). The aforementioned
-survey~\autocite{wu2013hiv} estimated that there were 14,511 \gls{MSM} living
-in Shanghai, with an \gls{HIV} prevalence of 6.8\% or approximately 987
-individuals. This was within the 95\% confidence interval of the estimate
-obtained with \software{netabc}. It is worth noting that} \citeauthor{li2015hiv}
-\uline{claimed that there were 80,000 \gls{MSM} living in Shanghai, which is a
-significantly different estimate than that obtained by} \textcite{wu2013hiv}.
-
-\textcite{cuevas2009hiv} \uline{reported that 620 newly infected and 1500
-chronically infected patients had been sampled during the study period, for a
-total of 2120. This is substantially greater than the point estimate of X we
-obtained with \gls{ABC}, even without considering the likely possibility that
-there were more infected than sampled individuals. Again, this error may in
-part be due to model misspecification, which highlights the importance of
-selecting a model appropriate to the dataset at hand.}}
-
-\subsubsection{Caveats? TODO}
+\subsubsection{Modelling assumptions}
 

### 2016-06-21 doc/thesis/intro.tex
Started writing intro for BA methods section
Almooooooost!

@@
   \madd{\post(\theta \mid T) = \frac{\lik(T \mid \theta) \prior(\theta)}
-    {\int_\Theta\lik(T \mid \theta) \prior(\theta)}.}
+    {\int_\Theta\lik(T \mid \theta) \prior(\theta) \d\theta}.}
   \label{eq:post}
@@
 
-\add{Here we describe an algorithm called the \gls{SMC} sampler, developed by
-\textcite{del2006sequential}, which forms the basis of the adaptive
-\gls{ABC}-\gls{SMC} algorithm we apply toward the main objective of this work.
-We begin by describing \gls{SIS}, which is a precursor to \gls{SMC} that
-samples from a sequence of distributions defined on spaces of increasing
+\add{In this section, we review an algorithm called the \gls{SMC} sampler,
+developed by \textcite{del2006sequential}, which forms the basis of the
+adaptive \gls{ABC}-\gls{SMC} algorithm we apply toward the main objective of
+this work. We begin by describing \gls{SIS}, which is a precursor to \gls{SMC}
+that samples from a sequence of distributions defined on spaces of increasing
 dimension. We then describe \gls{SMC} itself, which extends \gls{SIS} with a
@@
 mixed population, as most epidemiological models do, network models take the
-more realistic view that human populations are structured. That is, contacts
-which allow for transmission may occur in a nonrandom way, rather than between
-every pair of individuals in the population. A network model parameterizes this
-structure. In particular, our second research objective is to characterize our
-ability to fit the \gls{BA} model, which incorporates preferential attachment
-to generate networks with realistic degree distributions.}
+more realistic view that human populations are structured. For example,
+contacts may tend to occur more often between ``like'' individuals (with
+respect to geometry, socioeconomic status, or other factors) than at random. A
+network model provides a formal representation of this structure. In
+particular, our second research objective is to characterize our ability to fit
+the \gls{BA} model, which incorporates preferential attachment to generate
+networks with realistic degree distributions.}
 
@@
 networks, \acrlong{SMC}, and \acrlong{ABC}. The first two topics together form
-the problem domain. Phylogenetic data is the input to our method, while
+the problem domain. Phylogenetic data are the input to our method, while
 estimates of the parameters of contact network models are the desired output.

### 2016-06-21 doc/thesis/methods_results.Rnw
Almooooooost!

@@
                                 d[true_alpha == 0.5 | true_alpha == 1.5, m_error])
-    #stopifnot(alpha.av["factor(true_m)", "Pr(>F)"] <= 0.05)
-    #stopifnot(alpha.av["factor(true_I)", "Pr(>F)"] > 0.05)
-    #stopifnot(I.I.test$p.value < 1e-5)
-    #stopifnot(I.av["factor(true_m)", "Pr(>F)"] > 0.05)
-    #stopifnot(m.av["factor(true_m)", "Pr(>F)"] > 0.05)
-    #stopifnot(m.av["factor(true_I)", "Pr(>F)"] > 0.05)
-    #stopifnot(m.av["factor(true_alpha)", "Pr(>F)"] > 0.05)
-    #stopifnot(min(N.av[1:3,"Pr(>F)"]) > 0.05)
 
@@
 intervals) of the estimates of \gls{I} were substantially improved when \gls{N}
-and at least one other variable were marginalized out. Neither the accuracy nor
-the precision of any parameter estimates appeared to improve when more
-computing power was used for \gls{SMC} (\cref{fig:moresmc}).}
+was marginalized out. Neither the accuracy nor the precision of any parameter
+estimates appeared to improve when more computing power was used for \gls{SMC}
+(\cref{fig:moresmc}).}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{abc-marginal.pdf}
+    \caption[
+        Marginal posterior mean estimates and 50\%/95\% \gls{HPD} intervals of
+        \gls{BA} model parameters when some parameters were fixed.
+    ]{
+        Marginal posterior mean estimates (points), 50\% \gls{HPD} intervals
+        (notches), and 95\% \gls{HPD} intervals (lines) of \gls{BA} model
+        parameters when some parameters ($x$-axis) were fixed. Parameters were
+        fixed by specifying a Dirac-delta prior at the true value. True values
+        are indicated by horizontal dashed lines.
+    }
+    \label{fig:marginal}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{stock/missing}
+    \caption[
+        Comparsion of \gls{BA} parameter estimates obtained with
+        \software{netabc} on the same dataset with different \gls{SMC}
+        settings.
+    ]{
+        Comparsion of \gls{BA} parameter estimates obtained with
+        \software{netabc} on the same dataset with different \gls{SMC}
+        settings. Points are posterior means, notches are 50\% \gls{HPD}
+        intervals, and lines are 95\% \gls{HPD} intervals.
+    }
+    \label{fig:moresmc}
+\end{figure}
 
@@
 
+  setkey(d1, dataset)
+  setkey(d2, dataset)
+  setkey(d1.env, dataset)
+  setkey(d2.env, dataset)
+
+  pmean1 <- unique(d1)[,.(dataset, ntip)][pmean1]
+  pmean2 <- unique(d2)[,.(dataset, ntip)][pmean2]
+  pmean1.env <- unique(d1.env)[,.(dataset, ntip)][pmean1.env]
+  pmean2.env <- unique(d2.env)[,.(dataset, ntip)][pmean2.env]
+
   stopifnot(pmean1[,dataset[which.max(alpha)]] == "zetterberg2004")
@@
 
-  stopifnot(pmean1[,dataset[which.max(I)]] == "zetterberg2004")
-  stopifnot(pmean1[,dataset[which.min(I)]] == "novitsky2014")
-  stopifnot(pmean1[dataset != "novitsky2014", dataset[which.min(I)]] == "bctree")
-
-  stopifnot(pmean1[,cor.test(ntip, I)$p.value, method="spearman"] > 0.05)
+  stopifnot(pmean1[,dataset[which.min(I)]] == "zetterberg2004")
+  stopifnot(pmean1[,dataset[which.max(I)]] == "novitsky2014")
+  stopifnot(pmean1[dataset != "novitsky2014", dataset[which.max(I)]] == "bctree")
+
+  stopifnot(pmean1[,cor.test(ntip, I, method="spearman")$p.value] > 0.05)
 @
@@
 mass around the mode. When the value $\gls{m} = 1$ was disallowed by the prior, 
-TODO}
+the separation between the \gls{IDU} datasets and the others became more
+striking (\cref{fig:abchpdm2}). All three \gls{IDU} datasets had estimated
+\gls{alpha} values at or above
+    \Sexpr{pmean2[c("zetterberg2004", "niculescu2015", "bctree"), round(min(alpha), 2)]}.
+The estimate for the MSM/Beijing data was slightly lower
+    (\Sexpr{pmean2["wang2015", round(alpha, 2)]}) and
+the estimates for the seven remaining non-\gls{IDU} datasets were bounded above by
+    \Sexpr{pmean2[c("novitsky2014", "mccormack2002", "grabowski2014", "kao2011", "little2014", "li2015", "cuevas2009"), round(max(alpha), 2)]}.}
 
@@
 
+<<genetrees, include=FALSE>>=
+    datasets <- c("grabowski2014", "zetterberg2004", "mccormack2002")
+    paths <- paste0("../../simulations/", datasets, "/subtree/*")
+    paths <- sapply(paths, Sys.glob)
+    trees <- lapply(paths, read.tree)
+    get.ratio <- function (t) {
+        int <- t$edge.length[t$edge[,2] > Ntip(t)]
+        tip <- t$edge.length[t$edge[,2] <= Ntip(t)]
+        mean(int) / mean(tip)
+    }
+    ratios <- data.table(dataset=rep(datasets, each=2),
+                         gene=rep(c("env", "gag"), 3),
+                         ratio=sapply(trees, get.ratio))
+    stopifnot(ratios[gene == "env", mean(ratio)] <
+              ratios[gene == "gag", mean(ratio)])
+@
+
 \add{For all but the HET/Botswana data, the posterior mean estimates for
@@
 prevalence (Spearman correlation, 
-    $p \Sexpr{pp(pmean1[,cor.test(ntip, I)[["p.value"]], method="spearman"], eq=TRUE)}$).
-This indicates that the higher prevalence estimates were not simply due to
+    $p \Sexpr{pp(pmean1[,cor.test(ntip, I, method="spearman")[["p.value"]]], eq=TRUE)}$),
+indicating that the higher prevalence estimates were not simply due to
 increased sampling density. When both \textit{gag} and \textit{env} sequences
 were analyzed, the estimates from the \textit{env} data were substantially
-higher (\cref{fig:genes}). The ratios of terminal to internal branch lengths
-were higher for the \textit{env} trees than the \textit{gag} trees (average X
-for \textit{env} and X for \textit{gag}). As with \gls{alpha}, many of the
-posterior means were outside the 50\% \gls{HPD} intervals, suggesting diffuse
-posterior distributions. The values of \gls{I} were largely robust to
-restricting the prior to $\gls{m} > 1$ (compare
-\cref{fig:abchpd,fig:abchdpm2}), although TODO}
+higher (\cref{fig:genes}). The ratios of internal to terminal branch lengths
+were lower for the \textit{env} trees than the \textit{gag} trees (average 
+    \Sexpr{ratios[gene == "env", round(mean(ratio), 2)]}
+for \textit{env} and 
+    \Sexpr{ratios[gene == "gag", round(mean(ratio), 2)]}
+for \textit{gag}). As with \gls{alpha}, many of the posterior means were
+outside the 50\% \gls{HPD} intervals, suggesting diffuse posterior
+distributions. The values of \gls{I} were fairly robust to restricting the
+prior to $\gls{m} \geq 2$ (compare \cref{fig:abchpd,fig:abchdpm2}), although 
+the 95\% \gls{HPD} intervals were slightly narrower (average width
+    \Sexpr{round(mean(up1[,I] - low1[,I]))}
+for $\gls{m} \geq 1$ and
+    \Sexpr{round(mean(up2[,I] - low2[,I]))}
+for $\gls{m} \geq 2$).}
 
@@
   \Sexpr{replace_number(pmean1[,sum(m == 1)])}
-of the datasets analysed. The widths of the 95\% \gls{HPD} varied from 0 (all
-the mass on the estimated value) to 5 (the entire prior region). When the value
-$\gls{m} = 1$ was disallowed by the prior, TODO.}
+of the datasets analysed. The widths of the 95\% \gls{HPD} intervals varied
+from 0 (all the mass on the estimated value) to 5 (the entire prior region).
+When the value $\gls{m} = 1$ was disallowed by the prior, the posterior means
+for all but the HET/Botswana data took on the value 3, with the \gls{HPD}
+intervals spanning the entire prior region. This is very similar to the results 
+observed for \gls{m} on simulated data (\cref{fig:abcptm2,tab:abchpd}), and
+suggests that \gls{m} is not identifiable from these data with this prior. On
+the other hand, the estimate and 95\% \gls{HPD} interval for the HET/Botswana
+data (4 [2-5]) was the same for both priors.}
 
@@
 posterior mean always fell around the upper two-thirds mark of the region
-allowed by the prior. The results did not change appreciably (TODO) when the
-value $\gls{m} = 1$ was assigned zero prior probability.}
+allowed by the prior. The results did not change appreciably when the value
+$\gls{m} = 1$ was assigned zero prior probability.}
 
@@
 investigated, using the posterior median estimates \del{with the prior allowing
-$\gls{m} = 1$}. Because the estimates of \gls{N} were uninformative, we
+$\gls{m} = 1$}. The results are summarized in \cref{tab:gamma}. All of the
+estimated exponents were in the range $2 \leq \gls{gamma} \leq 2.5$, which is
+on the lower end of the range $2 \leq \gls{gamma} \leq 4$ reported in the
+literature. \add{Because the estimates of \gls{N} were uninformative, we also
 simulated another set of networks with the estimated \gls{I} value, rather than
-\gls{N}, as the number of nodes. The results are summarized in
-\cref{tab:gamma}. All of the estimated exponents were in the range $2 \leq
-\gls{gamma} \leq 2.5$, which is on the lower end of the range $2 \leq
-\gls{gamma} \leq 4$ reported in the literature.
+\gls{N}, as the number of nodes. However, none of the estimated \gls{gamma}
+values were changed by this operation (not shown).}
 
@@
     \caption[
-        Estimated power law exponents for six HIV datasets based on maximum
-        \textit{a priori} estimates of BA model parameters.
+        Estimated power law exponents for six \gls{HIV} datasets based on
+        posterior mean point estimates of \gls{BA} model parameters.
     ]{
-        Estimated power law exponents for six HIV datasets based on maximum
-        \textit{a priori} estimates of BA model parameters. 100 networks were
-        simulated using posterior mean parameter estimates obtained with
-        \software{netabc}. The power law exponent $\gamma$ was estimated for
-        each, and the median of those estimates was used as a point estimate
-        for the corresponding dataset.
+        Estimated power law exponents for six \gls{HIV} datasets based on
+        posterior mean point estimates of \gls{BA} model parameters. 100
+        networks were simulated using posterior mean parameter estimates
+        obtained with \software{netabc}. The power law exponent \gls{gamma}
+        was estimated for each, and the median of those estimates was used as a
+        point estimate for the corresponding dataset.
     }

### 2016-06-21 doc/thesis/methods_results.tex
Almooooooost!

@@
 
-Two additional simulations were performed to address \del{potential sources of error}
-\add{the possible impact of two types of model misspecification}. To consider
-the effect of heterogeneity among nodes, we generated a network where half the
-nodes were attached with power $\gls{alpha} = 0.5$ and the other half with power
-$\gls{alpha} = 1.5$. The other parameters for this network were $\gls{N} = 5000$,
-$\gls{I} = 1000$, and $\gls{m} = 2$. To investigate the effects of potential
-sampling bias~\autocite{karcher2016quantifying}, we simulated a transmission
-tree where the tips were sampled in a peer-driven fashion, rather than at
-random. That is, the probability to sample a node was twice as high if any of
-that node's network peers had already been sampled. The parameters of this
-network were $\gls{N} = 5000$, $\gls{I} = 2000$, $\gls{m} = 2$, and
-$\gls{alpha} = 0.5$.
-
-\end{document}
+Two further simulations were performed to address \del{potential sources of
+error} \add{the possible impact of two types of model misspecification}. To
+consider the effect of heterogeneity among nodes, we generated a network where
+half the nodes were attached with power $\gls{alpha} = 0.5$ and the other half
+with power $\gls{alpha} = 1.5$. The other parameters for this network were
+$\gls{N} = 5000$, $\gls{I} = 1000$, and $\gls{m} = 2$. To investigate the
+effects of potential sampling bias~\autocite{karcher2016quantifying}, we
+simulated a transmission tree where the tips were sampled in a peer-driven
+fashion, rather than at random. That is, the probability to sample a node was
+twice as high if any of that node's network peers had already been sampled. The
+parameters of this network were $\gls{N} = 5000$, $\gls{I} = 2000$, $\gls{m} =
+2$, and $\gls{alpha} = 0.5$.
+
+\add{To assess the impact of the \gls{SMC} settings on \software{netabc}'s
+accuracy, we ran \software{netabc} twice on the same simulated transmission
+tree. For the first run, the \gls{SMC} settings were the same as in the other
+simulations: 1000 particles, 5 simulated transmission trees per particle, and
+\gls{alphaess} = 0.95. The second run was performed with 2000 particles, 10
+simulated transmission trees per particle, and \gls{alphaess} = 0.99. To
+investigate the extent to which errors in the estimated \gls{BA} parameters
+were due to true features of the posterior, rather than an inaccurate \gls{ABC}
+approximation, we performed marginal estimation for one set of parameter
+values. Each combination of 1, 2, or 3 model parameters (14 combinations total)
+was fixed to their known values, and the remaining parameters were estimated
+with \software{netabc}. The parameter values were $\gls{alpha} = 0.0$, $\gls{m}
+= 2$, $\gls{I} = 2000$, and $\gls{N} = 5000$.}
 
@@
 measures tree imbalance, was significantly correlated with all four parameters
-    (for $\alpha$, $I$, $m$, and $N$ respectively: Spearman's rho =
+    (for \gls{alpha}, \gls{I}, \gls{m}, and \gls{N} respectively: Spearman's rho =
     0.85,
@@
   (Spearman's rho
-    \ensuremath{-0.84},
+    \ensuremath{-0.8},
     \ensuremath{-0.69},
-    0.1,
-    0.18;
+    0.09,
+    0.17;
 all $p < 10^{-5}$).
@@
 
-Accuracy of the \gls{kSVR} classifiers varied based on the parameter being
-tested (\cref{fig:rsquared}, left). Classifiers based on two other tree
-statistics, the \gls{nltt} and Sackin's index, generally exhibited worse
-performance than the tree kernel, although the magnitude of the disparity
-varied between the parameters (\cref{fig:rsquared}, centre and right). The
-results were largely robust to variations in the tree kernel meta-parameters
-$\lambda$ and $\sigma$, although accuracy varied between different epidemic and
-sampling scenarios
-(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}).
-
-When classifying $\alpha$, the \gls{kSVR} classifier had an average $R^2$ of 
+The accuracy of each classifier \add{on all \gls{BA} parameters is shown in}
+\del{varied based on the parameter being tested} \cref{fig:rsquared}.
+Classifiers based on \del{two other tree statistics,} the \gls{nltt} and
+Sackin's index generally exhibited worse performance than the tree kernel,
+although the magnitude of the disparity varied between the parameters
+(\cref{fig:rsquared}, centre and right). The results were largely robust to
+variations in the tree kernel meta-parameters \gls{lambda} and \gls{sigma},
+although accuracy varied between different epidemic and sampling scenarios
+(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}). \add{For all
+parameters except \gls{m}, the absolute number of tips in the tree had a much
+greater impact on accuracy than the proportion of infected individuals these
+tips represented. However, \gls{m}, both the number and proportion of sampled
+tips had a strong impact or the accuracy of the \gls{kSVR}
+(\cref{fig:mcrossv}).}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
+    \caption[
+        Cross-validation accuracy of kernel-SVR, nLTT-based SVR, and Sackin's
+        index regression classifiers for BA model parameters.
+    ]{
+        Cross-validation accuracy of kernel-SVR classifier (left), SVR
+        classifier using \gls{nltt} (centre), and linear regression using
+        Sackin's index (right) for \gls{BA} model parameters. Kernel
+        meta-parameters were set to $\gls{lambda} = 0.3$ and $\gls{sigma} = 4$.
+        Each point was calculated based on 300 simulated transmission trees
+        over networks with three different values of the parameter being
+        tested, assuming perfect knowledge of the other parameters. Vertical
+        lines are empirical 95\% confidence intervals based on 1000 two-fold
+        cross-validations. \add{The classifiers for \gls{I} were not evaluated
+        with 1000-tip trees, because one of the tested \gls{I} values was 500,
+        and it is not possible to sample a tree of size 1000 from 500 infected
+        individuals.}
+    }
+    \label{fig:rsquared}
+\end{figure}
+
+The \gls{kSVR} classifier for \gls{alpha} had an average $R^2$ of 
     0.92,
@@
 about the mean for different tree and epidemic sizes. No classifier could
-accurately identify the $m$ parameter in any epidemic scenario, with average
-$R^2$ values of 
+accurately identify \gls{m} in any epidemic scenario, with average $R^2$ values
+of 
   0.12 for \gls{kSVR},
@@
 
-The accuracy of classifiers for $I$ varied significantly with the number of
-tips in the tree. For 100-tip trees, the average $R^2$ values were
+The accuracy of classifiers for $I$ \del{varied significantly
+with}\add{appeared to be strongly influenced by} the number of tips in the
+tree. For 100-tip trees, the average $R^2$ values were
   0.7,
@@
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
-  \caption[Cross-validation accuracy of kernel-SVR, nLTT-based SVR, and
-  Sackin's index regression classifiers for BA model parameters.]{
-      Cross-validation accuracy of kernel-SVR classifier (left), SVR classifier
-      using \gls{nltt} (centre), and linear regression using Sackin's index
-      (right) for \gls{BA} model parameters. Kernel meta-parameters were set to
-      $\lambda = 0.3$ and $\sigma = 4$. Each point was calculated based on 300
-      simulated transmission trees over networks with three different values of
-      the parameter being tested, assuming perfect knowledge of the other
-      parameters. Vertical lines are empirical 95\% confidence intervals based
-      on 1000 two-fold cross-validations. {\color{blue}\uline{ The classifiers
-      for \gls{I} were not evaluated with 1000-tip trees, because one of the
-      tested \gls{I} values was 500, and it is not possible to sample a tree of
-      size 1000 from 500 infected individuals.}}
-  }
-  \label{fig:rsquared}
-\end{figure}
-
 \subsubsection*{Marginal parameter estimates with grid search}
 
-{\color{red}\sout{The accuracy of grid search estimates largely paralleled that
-of the \gls{kSVR} classifiers.}}\Cref{fig:gridest} shows point estimates and
-{\color{blue}\uline{50\%}} and 95\% highest density intervals for each of the
-\gls{BA} parameters, for one replicate experiment with 500-tip trees. Plots
-showing the point estimates for all replicates can be found in
+\del{The accuracy of grid search estimates largely paralleled that of the
+\gls{kSVR} classifiers.} \Cref{fig:gridest} shows point estimates and
+\add{50\%} and 95\% highest density intervals for each of the \gls{BA}
+parameters, for one replicate experiment with 500-tip trees. Plots showing the
+point estimates for all replicates can be found in
 \cref{fig:gridptalpha,fig:gridptI,fig:gridptm,fig:gridptN}. For all parameters
-except $m$, the error of point estimates was negatively correlated with the
+except \gls{m}, the error of point estimates was negatively correlated with the
 number of sampled tips in the tree (for \gls{alpha}, \gls{I}, and \gls{N}
-respectively: Spearman's $\rho$ = \ensuremath{-0.22},
-\ensuremath{-0.51}, \ensuremath{-0.16}; $p$-values
+respectively: Spearman's $\rho$ = 
+    \ensuremath{-0.22},
+    \ensuremath{-0.51}, 
+    \ensuremath{-0.16}; 
+$p$-values
     $4\!\times\!10^{-4}$,
@@
 
-{\color{blue}\uline{Across all replicates, $R^2$ values for the correlations
-between the estimated and true values were
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{gridsearch-example}
+    \caption[
+        Grid search estimates of \gls{BA} model parameters for one replicate
+        experiment with trees of size 500.
+    ]{
+        Grid search estimates of \gls{BA} model parameters for one replicate
+        experiment with trees of size 500. Point estimates (dots), 95\%
+        \glspl{HDI} (lines), and 50\% \glspl{HDI} (notches) for each \gls{BA}
+        model parameter, obtained using grid search. Networks and transmission
+        trees were simulated over a grid of values for each parameter while
+        holding the others fixed \add{to known values}. For a subset of the
+        grid values ($x$-axis), test networks and trees were created and
+        compared to each tree on the grid using the tree kernel. The kernel
+        scores along the grid were normalized to resemble a probability
+        distribution, from which the mode and highest density interval were
+        calculated.
+    } 
+    \label{fig:gridest}
+\end{figure}
+
+\add{Across all replicates, $R^2$ values for the correlations between the
+estimated and true values were
     $0.91$,
@@
     $17$\%
-of the respective grids.}
-
-\uline{Qualitatively, each of the parameters exhibited weak identifiability
-within particular sections of the grid} (\cref{fig:gridest}). \uline{The 50\%
-\glspl{HDI} for \gls{alpha} were nearly identical for the values $\gls{alpha}
-\leq 0.5$ (on average [X-X]) and for $\gls{alpha} \geq 1.5$ ([X-X]).
+of the respective grids. For \gls{I} and \gls{N}, relative errors may be more
+appropriate to consider. An overestimate of 100 individuals would be very
+misleading if the true population was only of size 100, but almost negligable
+in a population of size 5000. The mean relative errors for \gls{I} and \gls{N}
+respectively were
+    $18$\% and 
+    $31$\%.}
+
+\add{Qualitatively, \gls{alpha} and \gls{I} exhibited weak identifiability
+within particular sections of the grid (\cref{fig:gridest}). The 50\%
+\glspl{HDI} for \gls{alpha} were similar for the values $\gls{alpha}
+\leq 0.5$ (on average 
+    0.02 -
+    0.84)
+and for $\gls{alpha} \geq 1.5$ 
+    (1.26 -
+    2).
 For \gls{I}, similar 50\% \gls{HPD} were observed for $\gls{I} \leq 1500$
-(average [X-X]) and for $\gls{I} > 3000$ ([X-X]). For \gls{m}, both the 50\%
-and 95\% \glspl{HDI} were generally identical for $\gls{m} > 2$ (average [X-X]
-for 50\% and [X-X] for 95\% \gls{HDI}). We did not observe a similar pattern
-for \gls{N}, however the boundaries of the \gls{HDI} did not always
-monotonically increase with the actual value.}
-
-\Cref{fig:gridalpha,fig:gridI,fig:gridm,fig:gridN} \uline{show kernel score
+(average 
+    [650 -
+     2846])
+and for $\gls{I} > 3000$ 
+    ([2596 -
+      4802]).
+No similar patterns were observed for \gls{m} or \gls{N} (although the 50\%
+\glspl{HDI} appear to be identical for $\gls{m} > 2$ in \cref{fig:gridexpt},
+this was not consistent across replicates).}
+
+\add{\Cref{fig:gridalpha,fig:gridI,fig:gridm,fig:gridN} show kernel score
 distributions of kernel scores along the grid for each parameter. The
@@
     (Wilcoxon rank-sum test, $p {<}10^{-5}$,
-}\cref{fig:gridptalpha}).
-\sout{These two values exhibited different qualitative behaviour than the other
-values in terms of the distribution of kernel scores along the grid
-}(\cref{fig:gridalpha}). \sout{In particular, there was a pronounced peak in scores
-around the true value, in contrast to the other values where the scores were
-flat around the true value. The effect was most obvious for the value
-\gls{alpha} = 1.25.}}
-
-%The average absolute error of the point estimates for \gls{I} was 
-%    d["I", round(mean(error))] individuals,
-%on a grid of 500 to 5000, and these errors differed between true values of
-%\gls{I}
-%    (one-way \gls{ANOVA}, $p pp(av["I", p], eq=TRUE)$).
-%The errors for $2000 \leq I \leq 3000$ were higher than those for the other
-%values
-%    (average errors
-%     d["I", .SD[true_value >= 2000 & true_value <= 3000, round(mean(error))]]
-%     for $2000 \leq I \leq 3000$ vs.
-%     d["I", .SD[true_value < 2000 | true_value > 3000, round(mean(error))]]
-%     for $I < 2000$ or $I > 3000$),
-%and this difference was significant
-%    (Wilcoxon rank-sum test, $p pp(I.test$p.value, eq=TRUE)$,
-%     \cref{fig:gridptI}).
-%Kernel score distributions for all test values exhibited a similar rounded
-%shape (\cref{fig:gridI}). 
-%
-%The average error for \gls{m} was
-%    d["m", round(mean(error), 2)] edges per vertex,
-%on a grid from 1 to 6. The error varied significantly between the true values
-%of \gls{m} 
-%    (one-way \gls{ANOVA}, $p pp(av["m", p], eq=TRUE)$).
-%Errors for the value \gls{m} = 1 were lower than the other values
-%    (average errors
-%    d["m", .SD[true_value == 1, round(mean(error), 2)]]
-%    for $m = 1$ vs.
-%    d["m", .SD[true_value > 1, round(mean(error), 2)]]
-%    for $m > 1$),
-%and this difference was significant
-%    (Wilcoxon rank-sum test, $p pp(m.test$p.value, eq=TRUE)$,
-%     \cref{fig:gridptm}).
-%The value $m = 1$ causes the network to take on a distinct shape relative to
-%higher \gls{m} values, namely a tree (\ie there are no cycles, see
-%\cref{subsec:treeshape}). The kernel score distribution had a peak at $m = 1$
-%when this was the true value, and a valley at $m = 1$ when the true value of
-%$m$ was greater that 1 (\cref{fig:gridm}).
-%
-%The average error for \gls{N} was 
-%    d[parameter == "N", round(mean(error))] individuals,
-%on a grid from 1000 to 15000, and was varied significantly with the true value
-%of \gls{N}
-%    (one-way \gls{ANOVA}, $p pp(av["N", p], eq=TRUE)$).
-%The errors were lower for $N \leq 3000$
-%    (average errors
-%     d["N", .SD[true_value <= 3000, round(mean(error))]]
-%     for $N \leq 3000$ vs.
-%     d["N", .SD[true_value > 3000, round(mean(error))]]
-%     for $N > 3000$),
-%and this difference was significant
-%    (Wilcoxon rank-sum test, $p pp(N.test$p.value, eq=TRUE)$,
-%     \cref{fig:gridptN}).
-%The kernel score distribution had a peak at $N = 1000$ when this was the true
-%value, and a valley there otherwise (\cref{fig:gridN}). Except for this valley,
-%the distributions were flat for $N > 3000$.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth]{gridsearch-example}
-  \caption[
-    Grid search estimates of \gls{BA} model parameters for one replicate
-    experiment with trees of size 500.
-  ]{
-    Grid search estimates of \gls{BA} model parameters for one replicate
-    experiment with trees of size 500. Point estimates (dots), 95\% highest
-    density intervals (lines), and 50\% highest density intervals (notches) for
-    each \gls{BA} model parameter, obtained using grid search. Networks and
-    transmission trees were simulated over a grid of values for each parameter
-    while holding the others fixed. For a subset of the grid values ($x$-axis),
-    test networks and trees were created and compared to each tree on the grid
-    using the tree kernel. The kernel scores along the grid were normalized to
-    resemble a probability distribution, from which the mode and highest
-    density interval were calculated. 
-  } 
-  \label{fig:gridest}
-\end{figure}
+\cref{fig:gridptalpha}). These two values exhibited different qualitative
+behaviour than the other values in terms of the distribution of kernel scores
+along the grid (\cref{fig:gridalpha}). In particular, there was a pronounced
+peak in scores around the true value, in contrast to the other values where the
+scores were flat around the true value. The effect was most obvious for the
+value \gls{alpha} = 1.25.}
 
@@
     29\%
-of the regions of nonzero prior density. Average \gls{HPD} interval widths were
+of the regions of nonzero prior density. For \gls{I} and \gls{N}, relative
+errors were
+    38\%
+    [20 - 
+    50\%]
+and
+    83\%
+    [73 - 
+    90\%].
+Average \gls{HPD} interval widths were
     0.68,
@@
 it means that the absolute error decreases as the true value of \gls{alpha}
-increases.}}
+increases.}
 
 % latex table generated in R 3.2.3 by xtable 1.8-2 package
-% Thu Jun 16 16:24:45 2016
+% Mon Jun 20 17:18:05 2016
 \begin{table}[ht]
@@
 \end{tabular}
-\caption[Parameters of a fitted GLM relating error in estimated $\alpha$ to true values of BA parameters.]{Parameters of a fitted GLM relating error in estimated $\alpha$ to true values of BA parameters.
-GLM was fitted with a Gaussian distribution and inverse link function.
+\caption[Parameters of a fitted \gls{GLM} relating error in estimated \gls{alpha} to true values of \gls{BA} parameters.]{Parameters of a fitted \gls{GLM} relating error in estimated \gls{alpha} to true values of \gls{BA} parameters.
+\gls{GLM} was fitted with a Gaussian distribution and inverse link function.
 Coefficients are interpretable as additive effects on the inverse of the mean error.}
@@
 % latex table generated in R 3.2.3 by xtable 1.8-2 package
-% Thu Jun 16 16:24:45 2016
+% Mon Jun 20 17:18:05 2016
 \begin{table}[ht]
@@
 \end{tabular}
-\caption[Parameters of a fitted GLM relating error in estimated $I$ to true values of BA parameters.]{Parameters of a fitted GLM relating error in estimated $I$ to true values of BA parameters.
-GLM was fitted with a gaussian distribution and inverse link function.
+\caption[Parameters of a fitted \gls{GLM} relating error in estimated \gls{I} to true values of \gls{BA} parameters.]{Parameters of a fitted \gls{GLM} relating error in estimated \gls{I} to true values of \gls{BA} parameters.
+GLM was fitted with a Gaussian distribution and inverse link function.
 Coefficients are interpretable as additive effects on the inverse of the mean error.}
@@
     ($p =0.01$)
-but was not significatly affected by \gls{I}} (\cref{tab:glmalpha}). 
-\uline{Qualitatively, \gls{alpha} seemed to be only weakly identifiable between
-the values of 0 and 0.5} (\cref{fig:abcptm2}). \uline{The error in the
-estimated prevalence \gls{I} was slightly lower for smaller values of
-\gls{alpha}
+but was not significatly affected by \gls{I} (\cref{tab:glmalpha}).
+Qualitatively, \gls{alpha} seemed to be only weakly identifiable between the
+values of 0 and 0.5 (\cref{fig:abcptm2}). The error in the estimated prevalence
+\gls{I} was slightly lower for smaller values of \gls{alpha}
     ($p {<}10^{-5}$)
@@
     ($p =0.05$),
-but was not significantly affected by the true value of \gls{m}} 
+but was not significantly affected by the true value of \gls{m}
 (\cref{tab:glmI}).}
 
-{\color{red}\sout{
-The accuracy of the parameter estimates obtained with ABC
-paralleled the results from the \gls{kSVR} classifier. Of the four parameters,
-$\alpha$ was the most accurately estimated, with point estimates having a
-median [IQR] absolute error of 
+\del{The accuracy of the parameter estimates obtained with ABC paralleled the
+results from the \gls{kSVR} classifier. Of the four parameters, $\alpha$ was
+the most accurately estimated, with point estimates having a median [IQR]
+absolute error of 
     0.11 
@@
 than those for the other values 
-    (Wilcoxon rank-sum test, $p$ = $0$).
-Errors in estimating $\alpha$ 
-%also varied with the true value of $m$ just at
-%the threshold of statistical significance
-%    (one-way ANOVA, $p 
-%    pp(anova(lm(alpha_error~true_m, d))[1, "Pr(>F)"], eq=TRUE)$),
-did not vary across the true values of $m$ or $I$ (both
-one-way ANOVA).}
-
-\sout{Estimates for $I$ were relatively accurate, with point estimate errors of
+    (Wilcoxon rank-sum test, $p$ = 
+    $0$).
+Errors in estimating $\alpha$ did not vary across the true values of $m$ or $I$
+(both one-way ANOVA).}
+
+\del{Estimates for $I$ were relatively accurate, with point estimate errors of
     492 
@@
 at least 1
-    (Wilcoxon rank-sum test, $p$ = $0$)
+    (Wilcoxon rank-sum test, $p$ = 
+    $0$)
 and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
@@
 
-\sout{The $m$ parameter was estimated correctly in
+\del{The $m$ parameter was estimated correctly in
     37 \%
@@
 
-\sout{Finally, the total number of nodes $N$ was consistently over-estimated by about
+\del{Finally, the total number of nodes $N$ was consistently over-estimated by about
 a factor of two (error 4153
 [3660 - 4489] individuals). No parameters influenced the accuracy of the $N$
-estimates (all one-way ANOVA).}}
+estimates (all one-way ANOVA).}
+
+The dispersion of the \gls{ABC} approximation to the posterior also varied
+between the parameters (\cref{tab:abchpd}). \gls{HPD} intervals around
+\gls{alpha} and \gls{I} were often narrow relative to the region of nonzero
+prior density, whereas the intervals for \gls{m} and \gls{N} were more widely
+dispersed. \Cref{fig:abclow1,fig:abclow2} \del{shows the distributions for one
+simulation.} \add{show one- and two-dimensional marginal distributions for a
+simulation with relatively low error, where the errors in \gls{alpha} and
+\gls{I} were each below the 25\% quantiles. The parameters for this simulation
+were 
+    $\gls{alpha} = 1$,
+    $\gls{I} = 1000$,
+    $\gls{m} = 2$, and
+    $\gls{N} = 5000$.
+\Cref{fig:abchigh1,abchigh2} show the equivalent marginals for a different
+simulation with relatively high error rates above the 75\% quantiles. The
+parameters were
+    $\gls{alpha} = 1$,
+    $\gls{I} = 2000$,
+    $\gls{m} = 3$, and
+    $\gls{N} = 5000$.
+The two-dimensional marginals some dependence between pairs of parameters,
+particularly \gls{I} and \gls{N} which show a diagonally shaped region of high
+posterior density.}
 
@@
     \centering
-  \includegraphics[width=\textwidth]{abc-point-estimate-m2}
-  \vspace{6pt}
-  \caption[
-    Posterior mean point estimates for \gls{BA} model parameters obtained by
-    running \software{netabc} on simulated data, for simulations with \gls{m} =
-    2.
-  ]{
-    Posterior mean point estimates for \gls{BA} model parameters obtained by
-    running \software{netabc} on simulated data, for simulations with \gls{m} =
-    2. Dashed lines indicate true values. (A) Estimates of \gls{alpha} and
-    \gls{I} which were varied in these simulations against known values. (B)
-    Estimates of \gls{m} and \gls{N} which were held fixed in these simulations
-    at the values \gls{m} = 2 and \gls{N} = 5000.
-  }
-  \label{fig:abcptm2}
+    \includegraphics[width=\textwidth]{{abc-posterior/1.0_1000_2_5000_0}.pdf}
+    \caption[
+        One-dimensional marginal posterior distributions of \gls{BA} model
+        parameters estimated with low error by \software{netabc} from a
+        simulated transmission tree.
+    ]{
+        One-dimensional marginal posterior distributions of \gls{BA} model
+        parameters estimated with low error by \software{netabc} from a
+        simulated transmission tree. Dashed lines indicate true values, solid
+        lines indicate posterior means, and shaded areas show 95\%
+        \acrlong{HPD} intervals.
+    }
+    \label{fig:abclow1}
 \end{figure}
 
-\begin{table*}[ht]
-  \centering
-  \input{\tablepath/abc-hpd.tex}
-  \caption[
-      Average posterior mean point estimates and 95\% highest posterior density
-      (HPD) interval widths for BA model parameter estimates obtained with
-      ABC.
-  ]{
-      Average posterior mean point estimates and 95\% highest posterior density
-      (HPD) interval widths for BA model parameter estimates obtained with
-      ABC. Three transmission trees were simulated under each
-      combination of the listed parameter values, and the parameters were
-      estimated with ABC without training.
-  }
-  \label{tab:abchpd}
-\end{table*}
-
-The dispersion of the ABC approximation to the posterior also varied between
-the parameters, with narrower HPD intervals for the parameters with more
-accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows
-the distributions for one simulation. Equivalent plots for one replicate
-simulation with each studied parameter combination can be found in
-\cref{fig:0.0-1000-2-5000-0,fig:0.5-1000-2-5000-0,fig:1.0-1000-2-5000-0,fig:1.5-1000-2-5000-0,fig:0.0-2000-2-5000-0,fig:0.5-2000-2-5000-0,fig:1.0-2000-2-5000-0,fig:1.5-2000-2-5000-0,fig:0.0-1000-3-5000-0,fig:0.5-1000-3-5000-0,fig:1.0-1000-3-5000-0,fig:1.5-1000-3-5000-0,fig:0.0-2000-3-5000-0,fig:0.5-2000-3-5000-0,fig:1.0-2000-3-5000-0,fig:1.5-2000-3-5000-0,fig:0.0-1000-4-5000-0,fig:0.5-1000-4-5000-0,fig:1.0-1000-4-5000-0,fig:1.5-1000-4-5000-0,fig:0.0-2000-4-5000-0,fig:0.5-2000-4-5000-0,fig:1.0-2000-4-5000-0,fig:1.5-2000-4-5000-0}.
-HPD intervals around $\alpha$ and $I$ were often narrow
-relative to the region of nonzero prior density, whereas the intervals for $m$
-and $N$ were more widely dispersed.
-
 \begin{figure}[ht]
     \centering
-    \includegraphics[width=\textwidth]{abc-posterior-example}
-  \vspace{6pt}
-  \caption[
-    Marginal posterior distributions of BA model parameters estimated
-    with ABC for a single simulated transmission tree.
-  ]{
-    Marginal posterior distributions of BA model parameters estimated
-    with ABC for a single simulated transmission tree. Dotted
-    lines indicate true values, solid lines indicate posterior mean point
-    estimates, and shaded areas show 95\% highest posterior density intervals.
-  }
-  \label{fig:abcex}
+    \includegraphics[width=\textwidth]{{abc-posterior-2d/1.0_1000_2_5000_0}.pdf}
+    \caption[
+        Two-dimensional marginal posterior distributions of \gls{BA} model
+        parameters estimated with low error by \software{netabc} from a
+        simulated transmission tree.
+    ]{
+        Two-dimensional marginal posterior distributions of \gls{BA} model
+        parameters estimated with low error by \software{netabc} from a
+        simulated transmission tree. White circles indicate true values,
+        magenta diamonds indicate posterior means.
+    }
+    \label{fig:abclow2}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{{abc-posterior/1.0_2000_3_5000_0}.pdf}
+    \caption[
+        One-dimensional marginal posterior distributions of \gls{BA} model
+        parameters estimated with high error by \software{netabc} from a
+        simulated transmission tree.
+    ]{
+        One-dimensional marginal posterior distributions of \gls{BA} model
+        parameters estimated with high error by \software{netabc} from a
+        simulated transmission tree. Dashed lines indicate true values, solid
+        lines indicate posterior means, and shaded areas show 95\%
+        \acrlong{HPD} intervals.
+    }
+    \label{fig:abchigh1}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{{abc-posterior-2d/1.0_2000_3_5000_0}.pdf}
+    \caption[
+        Two-dimensional marginal posterior distributions of \gls{BA} model
+        parameters estimated with high error by \software{netabc} from a
+        simulated transmission tree.
+    ]{
+        Two-dimensional marginal posterior distributions of \gls{BA} model
+        parameters estimated with high error by \software{netabc} from a
+        simulated transmission tree. White circles indicate true values,
+        magenta diamonds indicate posterior means.
+    }
+    \label{fig:abchigh2}
 \end{figure}
@@
 the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
-other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The posterior mean
-[95\% HPD] estimates for each parameter were: 
-$\alpha$, 
+other half 1.5), with $\gls{m} = 2$, $\gls{N} = 5000$, and $\gls{I} = 1000$.
+The posterior mean [95\% \gls{HPD}] estimates for each parameter were: 
+\gls{alpha},
   1.03 
@@
    1.18];
-$I$,
+\gls{I},
   1474 
@@
    2990];
-$m$,
+\gls{m},
   3 
@@
    5];
-$N$,
+\gls{N},
   9861 
@@
    14977].
-The approximate posterior distributions for this simulation are shown in
-\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
-transmission tree in a peer-driven fashion, where the probability to sample a
-node was twice as high if one of its peers had already been sampled. The
-parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
-$I$ = 2000. The estimated values were
-$\alpha$, 
+The one-dimensional marginal approximate posterior distributions for this
+simulation are shown in \cref{fig:mixed}. To test the effect of sampling bias,
+we sampled one transmission tree in a peer-driven fashion, where the
+probability to sample a node was twice as high if one of its peers had already
+been sampled. The parameters for this experiment were $\gls{N} = 5000$,
+$\gls{m} = 2$, $\gls{alpha} = 0.5$, and $\gls{I} = 2000$. The estimated values
+were: \gls{alpha},
   0.3 
@@
    0.63];
-$I$,
+\gls{I},
   2449 
@@
    3811];
-$m$,
+\gls{m},
   3 
@@
    5];
-$N$,
+\gls{N},
   9132 
@@
    14780].
-The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
-of these results were in line with estimates obtained on other simulated
+The approximate posterior distributions are shown in \cref{fig:peerdriven}.
+Both of these results were in line with estimates obtained on other simulated
 datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
-$\alpha$ was somewhat lower than typical.
+\gls{alpha} was somewhat lower than typical.
+
+\add{\Cref{fig:marginal} shows the effect of performing marginal \gls{ABC}
+estimation of each of the \gls{BA} parameters on the same simulated
+transmission tree. The estimates of \gls{m} are apparently unaffected by
+marginalizing out the other parameters, coroborating the prevous experiments'
+findings that \gls{m} is not an identifiable parameter from scaled tree shapes.
+The estimates of \gls{alpha} and \gls{N} were weakly improved in marginal
+estimation, which is consistent with the result from grid search that all
+\gls{N} values and \gls{alpha} values below 0.5 are weakly identifiable. On the
+other hand, both the accuracy and precision (\ie width of the \gls{HPD}
+intervals) of the estimates of \gls{I} were substantially improved when \gls{N}
+was marginalized out. Neither the accuracy nor the precision of any parameter
+estimates appeared to improve when more computing power was used for \gls{SMC}
+(\cref{fig:moresmc}).}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{abc-marginal.pdf}
+    \caption[
+        Marginal posterior mean estimates and 50\%/95\% \gls{HPD} intervals of
+        \gls{BA} model parameters when some parameters were fixed.
+    ]{
+        Marginal posterior mean estimates (points), 50\% \gls{HPD} intervals
+        (notches), and 95\% \gls{HPD} intervals (lines) of \gls{BA} model
+        parameters when some parameters ($x$-axis) were fixed. Parameters were
+        fixed by specifying a Dirac-delta prior at the true value. True values
+        are indicated by horizontal dashed lines.
+    }
+    \label{fig:marginal}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{stock/missing}
+    \caption[
+        Comparsion of \gls{BA} parameter estimates obtained with
+        \software{netabc} on the same dataset with different \gls{SMC}
+        settings.
+    ]{
+        Comparsion of \gls{BA} parameter estimates obtained with
+        \software{netabc} on the same dataset with different \gls{SMC}
+        settings. Points are posterior means, notches are 50\% \gls{HPD}
+        intervals, and lines are 95\% \gls{HPD} intervals.
+    }
+    \label{fig:moresmc}
+\end{figure}
 
@@
 
-{\color{blue}\uline{
-The synthetic data experiments presented in the previous section demonstrated 
-that the \gls{alpha} and \gls{I} parameters, and to a lesser extent the \gls{N}
-parameter, had measurable impacts on tree shape and were identifiable with
-\gls{ABC}. Therefore, our next step was to use
-\software{netabc} to estimate the parameters of the \gls{BA} model for real
-world \gls{HIV} epidemics. }} Because the \gls{BA} model assumes a single
-connected contact network, it is most appropriate to apply to groups of
-individuals who are epidemiologically related. Therefore, we searched for
-published \gls{HIV} datasets which originated from existing clusters, either
-phylogenetically or geographically defined. In addition, we analysed an
+Because the \gls{BA} model assumes a single connected contact network, it is
+most appropriate to apply to groups of individuals who are epidemiologically
+related. Therefore, we searched for published \gls{HIV} datasets which
+originated from existing clusters, either phylogenetically or geographically
+defined. \add{To identify datasets fitting these criteria, we used the
+\software{Entrez} module in the \software{BioPython}
+library~\autocite{cock2009biopython} to identify all studies which were linked
+to at least 150 sequences of the same \gls{HIV} gene in GenBank (635 at the
+time when the data was collected). We manually curated a subset of these
+articles which, based on their title and abstract, appeared to have sampled one
+sequence per individual in a group likely to be epidemiologically related. For
+example, studies were excluded if they were investigating response to a
+particular drug, pregnant women or pediatric \gls{HIV} patients, intra-host
+evolution, or a multi-region or multi-country cohort. We acknowledge that our
+perusal of the complete set of articles was rather cursory, often limited to
+reading the title, and it is quite possible that we failed to identify other
+studies which would have been suitable to include. Each potential dataset was
+revisited, and those without sampling time annotation in GenBank were excluded.
+We also excluded studies where all sequences were sampled at the same
+timepoint, which was necessary because the method we used to time scale the
+tree requires non-contemporaneous tips.} The datasets are summarized in
+\cref{tab:data}. In addition \add{to the published data}, we analysed an
 in-house dataset sampled from \gls{HIV}-positive individuals in British
-Columbia, Canada (the ``BC data''). The datasets are summarized in
-\cref{tab:data}.
+Columbia, Canada. \add{For clarity, we will refer to each dataset by its risk
+group and location of origin in the text. For example, the
+\textcite{zetterberg2004two} data will be referred to as IDU/Estonia.}
 
@@
 We downloaded all sequences associated with each published study from GenBank.
-{\color{blue}\uline{For the \textcite{niculescu2015recent} data, only sequences
-from \gls{IDU} (whose sequence identifiers included the letters ``DU'') were
-included in the analysis. \textcite{kao2011surveillance} found a very strong
-association in their study population between subtype and risk group - subtype
-B was almost always associated with \gls{MSM}, whereas \gls{IDU} were usually
-infected with a circulating recombinant form. Since there were many more
-subtype B sequences in their data than sequences of other subtypes, we
-restricted our analysis to the subtype B sequences and labelled this dataset as
-\gls{MSM}. Three datasets (\textcite{zetterberg2004two},
-\textcite{grabowski2014role}, and \textcite{mccormack2002early}) included both
-\textit{env} and \textit{gag} sequences. Each gene was analyzed separately to
-assess the robustness of \software{netabc} to the particular \gls{HIV} gene
-sequence used to estimate a transmission tree.}}
-
-{\color{red}\sout{For the \textcite{novitsky2014impact} data,}} Each
-\textit{env} sequence was aligned pairwise to the HXB2 reference sequence
-(GenBank accession number K03455), and the hypervariable regions were clipped
-out with \software{BioPython} version 1.66+~\autocite{cock2009biopython}.
-Sequences were multiply aligned using \software{MUSCLE} version 3.8.31
+\add{For the IDU/Romania data, only sequences from \gls{IDU} (whose sequence
+identifiers included the letters ``DU'') were included in the analysis.
+\textcite{kao2011surveillance} (MSM/Taiwan) found a strong association in their
+study population between subtype and risk group - subtype B was most often
+associated with \gls{MSM}, whereas \gls{IDU} were usually infected with a
+circulating recombinant form. Since there were many more subtype B sequences in
+their data than sequences of other subtypes, we restricted our analysis to the
+subtype B sequences and labelled this dataset as \gls{MSM}. Three datasets
+(IDU/Estonia, HET/Uganda, and HET/Malawi) included both \textit{env} and
+\textit{gag} sequences. Each gene was analyzed separately to assess the
+robustness of \software{netabc} to the particular \gls{HIV} gene sequence used
+to estimate a transmission tree.}
+
+\del{For the \textcite{novitsky2014impact} data,} Each \textit{env} sequence
+was aligned pairwise to the HXB2 reference sequence (GenBank accession number
+K03455), and the hypervariable regions were clipped out with
+\software{BioPython} version 1.66+~\autocite{cock2009biopython}. Sequences were
+multiply aligned using \software{MUSCLE} version 3.8.31
 \autocite{edgar2004muscle}, and alignments were manually inspected with
@@
 
-{\color{red}\sout{Three}} 
-{\color{blue}\uline{Six}} of the datasets 
-\autocite[][and the BC data]{li2015hiv,novitsky2014impact,grabowski2014role,little2014using}
-were initially larger than the others, containing 1265, 1299, 1026/915
-(\textit{env}/\textit{gag}), 648, and 7923 sequences respectively. To ensure
+\del{Three} \add{Six} of the datasets (MSM/Shanghai, HET/Botswana, HET/Uganda, 
+and MSM/USA) were initially larger than the others, containing 1265, 1299,
+1026/915 (\textit{env}/\textit{gag}), and 648 sequences respectively. To ensure
 that the analyses were comparable, we reduced these to a number of sequences
-similar to the smaller datasets. For the \citeauthor{li2015hiv} and BC
-datasets, we detected clusters of size 280 and 399 respectively using a
-patristic distance cutoff of 0.02 as described
-previously~\autocite{poon2015impact}. Only sequences within these clusters were
-carried forward. For the
-\textcite{grabowski2014role,novitsky2014impact,little2014using} datasets, no
+similar to the smaller datasets. For the MSM/Shanghai dataset, we detected a
+cluster of size 280 using a patristic distance cutoff of 0.02 as described
+previously~\autocite{poon2015impact}. Only sequences within this cluster were
+carried forward. For the HET/Uganda, HET/Botswana, and MSM/USA datasets, no
 large clusters were detected using the same cutoff, so we analysed subsets of
-size 225, 180, and 180 respectively. {\color{blue}\uline{The subset of the
-\citeauthor{grabowski2014role} data was chosen by eye such that the individuals
-were monophyletic in both the \textit{gag} and \textit{env} trees. The other
-subsets were arbitrarily chosen subtrees from phylogenies of the complete
-datasets.}}
-
-For all datasets, we used the priors $\alpha$ $\sim$ Uniform(0, 2) and $N$ and
-$I$ jointly uniform on the region \{$n \leq N \leq 10000$, $n \leq I \leq
-10000$, $I \leq N$\}, where $n$ is the number of tips in the tree (see
-\cref{tab:data}). Since the value $m = 1$ produces networks with no cycles,
-which we considered fairly implausible, we ran one analysis with the prior $m
-\sim$ DiscreteUniform(1, 5), and one with the prior $m \sim$ DiscreteUniform(2,
-5). The other parameters to the SMC algorithm were the same as used for the
-simulation experiments, except that we used 10000 particles instead of 1000 to
-increase the accuracy of the estimated posterior. This was computationally
-feasible due to the small number of runs required for this analysis.
-
-Empirical studies of contact networks often report the exponent $\gamma$ of the
-power law degree distribution. To compare our results to the literature, we
-simulated 100 networks each according to the posterior mean parameter estimates
-obtained for each investigated dataset. The power-law exponent $\gamma$ was
-calculated for each network using the \software{fit\_power\_law} function in
-\software{igraph}, with the `R.mle' implementation. The median of the 100
-$\gamma$ values was taken as a point estimate for the associated dataset.
+size 225, 180, and 180 respectively. \add{The subset of the HET/Uganda data was
+chosen by eye such that the individuals were monophyletic in both the
+\textit{gag} and \textit{env} trees. The other subsets were arbitrarily chosen
+subtrees from phylogenies of the complete datasets.}
+
+For all datasets, we used the priors $\gls{alpha} \sim \Uniform(0, 2)$ and
+\gls{N} and \gls{I} jointly uniform on the region $\{n \leq \gls{N} \leq
+10000$, $n \leq \gls{I} \leq 10000$, $\gls{I} \leq \gls{N}\}$, where $n$ is the
+number of tips in the tree (see \cref{tab:data}). Since the value $\gls{m} = 1$
+produces networks with no cycles, which we considered fairly implausible, we
+ran one analysis with the prior $\gls{m} \sim \DiscreteUniform(1, 5)$, and one
+with the prior $\gls{m} \sim \DiscreteUniform(2, 5)$. The other parameters to
+the SMC algorithm were the same as used for the simulation experiments, except
+that we used 10000 particles instead of 1000 to increase the accuracy of the
+estimated posterior. This was computationally feasible due to the small number
+of runs required for this analysis.
+
+Empirical studies of contact networks often report the exponent \gls{gamma} of
+the power law degree distribution~\autocite{colgate1989risk, liljeros2001web,
+schneeberger2004scale, clemenccon2015statistical, rothenberg2007large,
+brown2011transmission}. Although the \gls{BA} model does not produce networks
+with true power law degree distributions except when $\gls{alpha} = 1$, the
+power law still provides a reasonable approximation to the slope when
+$\gls{alpha} \neq 1$ (\cref{fig:powerlaw}).  To compare our results to
+\add{existing network} literature we simulated 100 networks each according to
+the posterior mean parameter estimates obtained for each investigated dataset.
+\add{Although the \gls{BA} model does not produce power law networks except
+when $\gls{alpha} = 1$, simulations show that the power law fit still captures
+the slope of the degree distribution reasonably well (\cref{fig:powerlaw}).}
+\gls{gamma} was calculated for each network using the
+\software{fit\_power\_law} function in \software{igraph}, with the `R.mle'
+implementation. The median of the 100 \gls{gamma} values was taken as a point
+estimate for the associated dataset.
 
@@
 all results discussed henceforth are for the prior $\gls{m} \sim
-\DiscreteUniform(1, 5)$ unless otherwise stated. For clarity, datasets will be
-referred to as their risk group and location. For example, \gls{IDU}/Estonia
-refers to the} \textcite{zetterberg2004two} \uline{dataset.}}
-
-{\color{red}\sout{
-Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had
+\DiscreteUniform(1, 5)$ unless otherwise stated.}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{realdata-hpd-bc}
+    \caption[
+        Posterior means and 50\%/95\% \gls{HPD} intervals for parameters of the
+        \gls{BA} network model, fitted to eleven
+        \gls{HIV} datasets with \software{netabc}.
+    ]{
+        Posterior means (points), 50\% \gls{HPD} intervals (notches), and 95\%
+        (lines) \gls{HPD} intervals for parameters of the \gls{BA} network
+        model, fitted to eleven \gls{HIV} datasets with \software{netabc}.
+        Legend labels indicate risk group and country of origin. Abbreviations:
+        \acrshort{IDU}, \acrlong{IDU}; \acrshort{MSM}, \acrlong{MSM}; HET,
+        heterosexual. Note that posterior means can fall outside of \glspl{HPD} 
+        if the distribution is diffuse.
+    }
+    \label{fig:abchpd}
+\end{figure}
+
+\del{Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had
 estimated $\alpha$ values near unity for the prior allowing $m = 1$ (posterior
@@
    [0 -
-    1.75] for $m \geq 2$). }}
-
-{\color{blue}\uline{Posterior mean point estimates for the \acrlong{PA} power
-\gls{alpha} ranged from 
+    1.75] for $m \geq 2$).}
+
+\add{Posterior mean point estimates for the \acrlong{PA} power \gls{alpha}
+were all sub-linear, ranging from 
     0.9
@@
 where both \textit{gag} and \textit{env} were sequenced, but the estimates for
-\gls{alpha} did not change appreciably between the two genes} (TODO figure).
-\uline{There was a large amount of uncertainty associated with these
-estimates. 95\% \gls{HPD} were very wide for most datasets, often encompassing
-almost the entire range from 0 to 1} (\cref{fig:abchpd}). \uline{For all the
-datasets except HET/Malawi and MSM/Beijing, the posterior mean was either
-outside, or very close to the border of, the 50\% \gls{HPD}. This indicates
-that the posterior distributions were diffuse with heavy tails, rather than
-having most of their mass around the mode.}}
-
-{\color{red}\sout{For all the datasets except \citeauthor{novitsky2014impact},
+\gls{alpha} did not change appreciably between the two genes (\cref{fig:genes}).
+There was a large amount of uncertainty associated with these estimates. 95\%
+\gls{HPD} were very wide for most datasets, often encompassing almost the
+entire range from 0 to 1 (\cref{fig:abchpd}). For all the datasets except
+HET/Malawi and MSM/Beijing, the posterior mean was either outside, or very
+close to the border of, the 50\% \gls{HPD}. This indicates that the posterior
+distributions were diffuse with heavy tails, rather than having most of their
+mass around the mode. When the value $\gls{m} = 1$ was disallowed by the prior, 
+the separation between the \gls{IDU} datasets and the others became more
+striking (\cref{fig:abchpdm2}). All three \gls{IDU} datasets had estimated
+\gls{alpha} values at or above
+    0.9.
+The estimate for the MSM/Beijing data was slightly lower
+    (0.85) and
+the estimates for the seven remaining non-\gls{IDU} datasets were bounded above by
+    0.58.}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{realdata-genes}
+    \caption[
+        Posterior means and 50\%/95\% \gls{HPD} intervals for parameters of the
+        \gls{BA} network model, fitted to three \gls{HIV} datasets where both
+        \textit{gag} and \textit{env} genes were sequenced.
+    ]{
+        Posterior means (points) 50\% \gls{HPD} intervals (notches), and 95\%
+        \gls{HPD} intervals (lines) for parameters of the \gls{BA} network model,
+        fitted to three \gls{HIV} datasets where both \textit{gag} and
+        \textit{env} genes were sequenced.
+    }
+    \label{fig:genes}
+\end{figure}
+
+\del{For all the datasets except \citeauthor{novitsky2014impact},
 estimated values of $I$ were below 2000 when $m = 1$ was allowed, with
@@
 prior on $m$, with slightly narrower HPD intervals (compare
-\cref{fig:abchpd,fig:abchpdm2}).}}
-
-The posterior mean of $m$ was equal to 1 for all but the
+\cref{fig:abchpd,fig:abchpdm2}).}
+
+\add{For all but the HET/Botswana data, the posterior mean estimates for
+the prevalence \gls{I} were between
+    372
+(IDU/Estonia) and
+    1481
+(IDU/Canada). The HET/Botswana data had a much higher estimated \gls{I} value
+    (5432)
+than the other datasets, with a very wide 95\% \gls{HPD} interval covering
+almost the entire prior region (\cref{fig:abchpd}). There was no significant
+correlation between the number of sequences in the tree and the estimated
+prevalence (Spearman correlation, 
+    $p =0.6$),
+indicating that the higher prevalence estimates were not simply due to
+increased sampling density. When both \textit{gag} and \textit{env} sequences
+were analyzed, the estimates from the \textit{env} data were substantially
+higher (\cref{fig:genes}). The ratios of internal to terminal branch lengths
+were lower for the \textit{env} trees than the \textit{gag} trees (average 
+    0.47
+for \textit{env} and 
+    0.76
+for \textit{gag}). As with \gls{alpha}, many of the posterior means were
+outside the 50\% \gls{HPD} intervals, suggesting diffuse posterior
+distributions. The values of \gls{I} were fairly robust to restricting the
+prior to $\gls{m} \geq 2$ (compare \cref{fig:abchpd,fig:abchdpm2}), although 
+the 95\% \gls{HPD} intervals were slightly narrower (average width
+    2183
+for $\gls{m} \geq 1$ and
+    1855
+for $\gls{m} \geq 2$).}
+
+\del{The posterior mean of $m$ was equal to 1 for all but the
 \citeauthor{novitsky2014impact} data, when this value was allowed. However, the
@@
   (5916)
-when $m = 1$ was allowed.
-
-\begin{figure*}[ht]
-  \centering
-  \includegraphics{realdata-hpd-bc}
-  \caption[
-      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
-      parameters of the BA network model, fitted to six HIV datasets
-      with \software{netabc}.]
-  {
-      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
-      parameters of the BA network model, fitted to six HIV datasets with
-      \software{netabc}. Legend labels indicate risk group and country of
-      origin. Abbreviations: IDU, injection drug users; MSM, men who have sex
-      with men; HET, heterosexual.
-  }
-  \label{fig:abchpd}
-\end{figure*}
-
-To make our analyses comparable to the existing network literature, we
-estimated values of the power law exponent $\gamma$ for each of the datasets 
-investigated, using the posterior median estimates with the prior allowing
-\gls{m} = 1. The results are summarized in \cref{tab:gamma}. All of the
-estimated exponents were in the range $2 \leq \gamma \leq 2.5$, which is on the
-lower end of the range $2 \leq \gamma \leq 4$ reported in the literature.
+when $m = 1$ was allowed.}
+
+\add{The posterior means of \gls{m} were equal to one for 
+  zero
+of the datasets analysed. The widths of the 95\% \gls{HPD} intervals varied
+from 0 (all the mass on the estimated value) to 5 (the entire prior region).
+When the value $\gls{m} = 1$ was disallowed by the prior, the posterior means
+for all but the HET/Botswana data took on the value 3, with the \gls{HPD}
+intervals spanning the entire prior region. This is very similar to the results 
+observed for \gls{m} on simulated data (\cref{fig:abcptm2,tab:abchpd}), and
+suggests that \gls{m} is not identifiable from these data with this prior. On
+the other hand, the estimate and 95\% \gls{HPD} interval for the HET/Botswana
+data (4 [2-5]) was the same for both priors.}
+
+\add{Estimates of \gls{N} were  mostly uninformative, with very similar estimates
+for all datasets (mean
+    6234,
+range
+    5881 - 6882).
+This was similar to the pattern observed for the synthetic data, where the
+posterior mean always fell around the upper two-thirds mark of the region
+allowed by the prior. The results did not change appreciably when the value
+$\gls{m} = 1$ was assigned zero prior probability.}
+
+To compare our results to the existing network literature, we estimated
+values of the power law exponent \gls{gamma} for each of the datasets
+investigated, using the posterior median estimates \del{with the prior allowing
+$\gls{m} = 1$}. The results are summarized in \cref{tab:gamma}. All of the
+estimated exponents were in the range $2 \leq \gls{gamma} \leq 2.5$, which is
+on the lower end of the range $2 \leq \gls{gamma} \leq 4$ reported in the
+literature. \add{Because the estimates of \gls{N} were uninformative, we also
+simulated another set of networks with the estimated \gls{I} value, rather than
+\gls{N}, as the number of nodes. However, none of the estimated \gls{gamma}
+values were changed by this operation (not shown).}
 
@@
     \caption[
-        Estimated power law exponents for six HIV datasets based on maximum
-        \textit{a priori} estimates of BA model parameters.
-    ]{
-        Estimated power law exponents for six HIV datasets based on maximum
-        \textit{a priori} estimates of BA model parameters. 100 networks were
-        simulated using posterior mean parameter estimates obtained with
-        \software{netabc}. The power law exponent $\gamma$ was estimated for
-        each, and the median of those estimates was used as a point estimate
-        for the corresponding dataset.
+        Estimated power law exponents for six \gls{HIV} datasets based on
+        posterior mean point estimates of \gls{BA} model parameters.
+    ]{
+        Estimated power law exponents for six \gls{HIV} datasets based on
+        posterior mean point estimates of \gls{BA} model parameters. 100
+        networks were simulated using posterior mean parameter estimates
+        obtained with \software{netabc}. The power law exponent \gls{gamma}
+        was estimated for each, and the median of those estimates was used as a
+        point estimate for the corresponding dataset.
     }

### 2016-06-22 doc/paper/paper.Rnw
Version of paper that was submitted to VE

@@
 \usepackage[T1]{fontenc}
-\usepackage{garamondx}
-\usepackage[garamondx,cmbraces]{newtxmath}
+\usepackage{mathptmx}
 \usepackage[backend=biber]{biblatex}
@@
 \title{Reconstructing contact network parameters from viral phylogenies}
-\author{Rosemary M. McCloskey, Richard H. Liang, and Art F.Y. Poon}
-\abstract{\input{abstract}}
+\author{Rosemary M. McCloskey$^1$, Richard H. Liang$^1$, and Art F.Y. Poon$^{1,2}$ \\
+\small $^1$BC Centre for Excellence in HIV/AIDS, Vancouver, Canada \\ \small $^2$ Department of Medicine, University of British Columbia, Vancouver, Canada}
 
@@
     library(Hmisc)
+
+    pp <- function (p) {
+        if (p < 1e-5) {
+            "{<}10^{-5}"
+        } else {
+            paste0("=", latexSN(round(p, -floor(log10(p)))))
+        }
+    }
 @
@@
 \maketitle
-
-\section{Introduction}
+\abstract{\input{abstract}}
+
+\section*{Introduction}
 
@@
 with a high number of contacts attract new connections at an elevated rate. The
-first contact network model incorporating pcreferential attachment was
-introduced by \textcite{barabasi1999emergence}, and is now creferred to as the
+first contact network model incorporating preferential attachment was
+introduced by \textcite{barabasi1999emergence}, and is now referred to as the
 Barab\'asi-Albert (BA) model. Under this model, networks are formed by
@@
 already led to a new infection. The authors found that a Waring distribution,
-which is produced by a more sophisticated pcreferential attachment model, was a
+which is produced by a more sophisticated preferential attachment model, was a
 good fit to their estimated network.
@@
 that some network parameters can be inferred with reasonable accuracy, while
-others have a minimal detectable impact on tree shape and thecrefore cannot be
+others have a minimal detectable impact on tree shape and therefore cannot be
 estimated accurately. We also find that these parameters can vary considerably
@@
 
-\section{Methods}
-
-\subsection{\textit{Netabc}: phylogenetic inference of contact network
+\section*{Methods}
+
+\subsection*{\textit{Netabc}: phylogenetic inference of contact network
 parameters with kernel-ABC}
@@
 
-\subsection{Kernel classifiers}
+\subsection*{Kernel classifiers}
 
@@
 We constructed a kSVR classifier for $\alpha$ using the \textit{kernlab}
-package~\autocite{karatzoglou2004kernlab}, and evaluated its accuracy with 1000
+package~\autocite{zeileis2004kernlab}, and evaluated its accuracy with 1000
 two-fold cross-validations.
@@
 individuals (\textit{i.e.} the number of tips) and the epidemic size
-(\textit{i.e.} $I$) will be creferred to as an ``epidemic scenario''. When
+(\textit{i.e.} $I$) will be referred to as an ``epidemic scenario''. When
 evaluating the classifier for $I$, we did not consider trees with 1000 tips,
@@
 
-\subsection{ABC simulations}
+\subsection*{ABC simulations}
 
@@
 
-\subsection{Investigation of published data}
+\subsection*{Investigation of published data}
 
@@
 \textcite{novitsky2014impact} data, each sequence was aligned pairwise to the HXB2
-creference sequence (Genbank accession number K03455) and the hypervariable
+reference sequence (Genbank accession number K03455) and the hypervariable
 regions were clipped out with \textit{BioPython} version
@@
 
-\section{Results}
-
-\subsection{Kernel classifiers}
+\section*{Results}
+
+\subsection*{Kernel classifiers}
 
@@
 
-\begin{figure*}[ht]
-  \centering
-  \includegraphics[width=\textwidth]{kernel-rsquared.eps}
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth]{kernel-rsquared}
   \vspace{6pt}
@@
   \label{fig:rsquared}
-\end{figure*}
-
-\subsection{ABC simulations}
-
-%<<point_est, include=FALSE>>=
-%    f <- Sys.glob("../../simulations/abc-pa-free-m/point-estimate/*")
-%    d <- fread(f)
-%    d[,m := floor(m)]
-%    d[,alpha_error := abs(true_alpha - alpha)]
-%    d[,N_error := abs(true_N - N)]
-%    d[,I_error := abs(true_I - I)]
-%    d[,m_error := abs(true_m - m)]
-%
-%    alpha.av <- anova(lm(alpha_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-%    m.av <- anova(lm(m_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-%    N.av <- anova(lm(N_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-%    I.av <- anova(lm(I_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-%
-%    m.tbl <- d[,prop.table(table(m_error))]
-%    zero.alpha.test <- wilcox.test(d[true_alpha == 0, alpha_error], 
-%                                   d[true_alpha != 0, alpha_error])
-%    alpha.I.test <- wilcox.test(d[true_alpha < 1, I_error],
-%                                d[true_alpha >= 1, I_error])
-%    I.I.test <- wilcox.test(d[true_I == 1000, I_error],
-%                            d[true_I == 2000, I_error])
-%    alpha.m.test <- wilcox.test(d[true_alpha == 0 | true_alpha == 1, m_error],
-%                                d[true_alpha == 0.5 | true_alpha == 1.5, m_error])
-%    stopifnot(alpha.av["factor(true_m)", "Pr(>F)"] > 0.05)
-%    stopifnot(alpha.av["factor(true_I)", "Pr(>F)"] > 0.05)
-%    stopifnot(I.I.test$p.value < 1e-5)
-%    stopifnot(I.av["factor(true_m)", "Pr(>F)"] > 0.05)
-%    stopifnot(m.av["factor(true_m)", "Pr(>F)"] > 0.05)
-%    stopifnot(m.av["factor(true_I)", "Pr(>F)"] > 0.05)
-%    stopifnot(min(N.av[1:3,"Pr(>F)"]) > 0.05)
-%@
+\end{figure}
+
+\subsection*{ABC simulations}
+
+<<point_est, include=FALSE>>=
+    f <- Sys.glob("../../simulations/abc-pa-free-m/point-estimate/*")
+    d <- fread(f)
+    d[,m := floor(m)]
+    d[,alpha_error := abs(true_alpha - alpha)]
+    d[,N_error := abs(true_N - N)]
+    d[,I_error := abs(true_I - I)]
+    d[,m_error := abs(true_m - m)]
+
+    alpha.av <- anova(lm(alpha_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
+    m.av <- anova(lm(m_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
+    N.av <- anova(lm(N_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
+    I.av <- anova(lm(I_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
+
+    m.tbl <- d[,prop.table(table(m_error))]
+    zero.alpha.test <- wilcox.test(d[true_alpha == 0, alpha_error], 
+                                   d[true_alpha != 0, alpha_error])
+    alpha.I.test <- wilcox.test(d[true_alpha < 1, I_error],
+                                d[true_alpha >= 1, I_error])
+    I.I.test <- wilcox.test(d[true_I == 1000, I_error],
+                            d[true_I == 2000, I_error])
+    alpha.m.test <- wilcox.test(d[true_alpha == 0 | true_alpha == 1, m_error],
+                                d[true_alpha == 0.5 | true_alpha == 1.5, m_error])
+    stopifnot(alpha.av["factor(true_m)", "Pr(>F)"] <= 0.05)
+    stopifnot(alpha.av["factor(true_I)", "Pr(>F)"] > 0.05)
+    stopifnot(I.I.test$p.value < 1e-5)
+    stopifnot(I.av["factor(true_m)", "Pr(>F)"] > 0.05)
+    stopifnot(m.av["factor(true_m)", "Pr(>F)"] > 0.05)
+    stopifnot(m.av["factor(true_I)", "Pr(>F)"] > 0.05)
+    stopifnot(m.av["factor(true_alpha)", "Pr(>F)"] > 0.05)
+    stopifnot(min(N.av[1:3,"Pr(>F)"]) > 0.05)
+
+    alpha.m.cor <- cor.test(d[,alpha_error], d[,true_m], method="spearman")
+@
 
@@
 median [IQR] absolute error of 
-%    \Sexpr{d[,round(median(alpha_error), 2)]} 
-%    [\Sexpr{d[,round(quantile(alpha_error, 0.25), 2)]} - 
-%    \Sexpr{d[,round(quantile(alpha_error, 0.75), 2)]}].
+    \Sexpr{d[,round(median(alpha_error), 2)]} 
+    [\Sexpr{d[,round(quantile(alpha_error, 0.25), 2)]} - 
+    \Sexpr{d[,round(quantile(alpha_error, 0.75), 2)]}].
 The errors when the true value of $\alpha$ was zero were significantly greater
 than those for the other values 
-%    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(zero.alpha.test$p.value, 4))}$),
-but did not vary across the true values of the other parameters (one-way
-ANOVA). Estimates for $I$ were also relatively accurate, with point estimate
-errors of
-%    \Sexpr{d[,round(median(I_error))]} 
-%    [\Sexpr{d[,round(quantile(I_error, 0.25))]} - 
-%    \Sexpr{d[,round(quantile(I_error, 0.75))]}].
+    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(zero.alpha.test$p.value, 4))}$).
+Errors in estimating $\alpha$ also varied with the true value of $m$ just at
+the threshold of statistical significance
+    ($p \Sexpr{pp(alpha.av["factor(true_m)", "Pr(>F)"])}$),
+but did not vary across the true values of $N$ or $I$ (both one-way ANOVA).
+Estimates for $I$ were relatively accurate, with point estimate errors of
+    \Sexpr{d[,round(median(I_error))]} 
+    [\Sexpr{d[,round(quantile(I_error, 0.25))]} - 
+    \Sexpr{d[,round(quantile(I_error, 0.75))]}] individuals.
 These errors were significantly higher when the true value of $\alpha$ was
 at least 1
-%    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(alpha.I.test$p.value, 4))}$)
+    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(alpha.I.test$p.value, 4))}$)
 and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
@@
 The $m$ parameter was estimated correctly in only
-%    \Sexpr{as.integer(m.tbl[1] * 100)} \%
-of simulations. Oddly, the error in the estimated $m$ was higher for integer
-values of $\alpha$ (\textit{i.e.}~0 and 1) than non-integer values 
-%    (Wilcoxon rank-sum test, $p$ = \Sexpr{round(alpha.m.test$p.value, 3)}).
-The true values of the other parameters did not significantly affect the
-estimates of $m$ (both one-way ANOVA). Finally, the total number of nodes $N$
-was consistently over-estimated by about a factor of two
-%    (error \Sexpr{d[,round(median(N_error))]} 
-%    [\Sexpr{d[,round(quantile(N_error, 0.25))]} - 
-%     \Sexpr{d[,round(quantile(N_error, 0.75))]}]).
+    \Sexpr{as.integer(m.tbl[1] * 100)} \%
+of simulations, barely better than random guessing. The true values of the
+other parameters did not significantly affect the estimates of $m$ (both
+one-way ANOVA). Finally, the total number of nodes $N$ was consistently
+over-estimated by about a factor of two
+    (error \Sexpr{d[,round(median(N_error))]} 
+    [\Sexpr{d[,round(quantile(N_error, 0.25))]} - 
+     \Sexpr{d[,round(quantile(N_error, 0.75))]}] individuals).
 No parameters influenced the accuracy of the $N$ estimates (all one-way ANOVA).
 
-\begin{figure*}[ht]
-  \centering
-  \includegraphics[width=\textwidth]{abc-point-estimate-m2.eps}
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\textwidth]{abc-point-estimate-m2}
   \vspace{6pt}
@@
   \label{fig:abcpt}
-\end{figure*}
+\end{figure}
 
@@
 
+<<supp_labels, echo=FALSE, results="asis">>=
+    N <- 5000
+    replicate <- 0
+    lab <- NULL
+    for (m in c(2, 3, 4)) {
+    for (I in c(1000, 2000)) {
+    for (alpha in c(0, 0.5, 1, 1.5)) {
+        lab <- c(lab, sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate))
+    }
+    }
+    }
+    lab <- paste(lab, collapse = ",")
+@
+
 The dispersion of the ABC approximation to the posterior also varied between
@@
 accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows
-the distributions for for one simulation (equivalent plots for all the
-simulations can be found in Supplementary Figures TODO). HPD intervals around
-$\alpha$ and $I$ were narrow relative to the region of nonzero prior density,
-whereas the intervals for $m$ and $N$ were more widely dispersed.
-
-\begin{figure*}[ht]
-    \centering
-  \includegraphics[width=\textwidth]{abc-posterior-example.eps}
+the distributions for for one simulation. Equivalent plots for one replicate
+simulation with each studied parameter combination can be found in
+\cref{\Sexpr{lab}}. HPD intervals around $\alpha$ and $I$ were often narrow
+relative to the region of nonzero prior density, whereas the intervals for $m$
+and $N$ were more widely dispersed.
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{abc-posterior-example}
   \vspace{6pt}
@@
   \label{fig:abcex}
-\end{figure*}
+\end{figure}
 
@@
 To test the effect of model misspecification, we simulated one network where
-the nodes exhibited heterogeneous pcreferential attachment power (half 0.5, the
+the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
 other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
@@
    \Sexpr{d["N", round(hpd.max)]}].
-To test the effect of sampling bias, we sampled one transmission tree in a
-peer-driven fashion, where the probability to sample a node was twice as high
-if one of its peers had already been sampled. The parameters for this
-experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and $I$ = 2000. The
-estimated values were
+The approximate posterior distributions for this simulation are shown in
+\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
+transmission tree in a peer-driven fashion, where the probability to sample a
+node was twice as high if one of its peers had already been sampled. The
+parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
+$I$ = 2000. The estimated values were
 $\alpha$, 
@@
    \Sexpr{p["N", floor(hpd.max)]}].
-Both of these results were in line with estimates obtained on other simulated
-datasets (\cref{tab:abchpd}).
-
-\subsection{Real data}
+The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
+of these results were in line with estimates obtained on other simulated
+datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
+$\alpha$ was somewhat lower than typical.
+
+\subsection*{Real data}
 
@@
 
-\begin{figure*}[ht]
+\begin{figure}[ht]
   \centering
@@
       parameters of the BA network model, fitted to five published HIV datasets
-      with kernel-ABC.
+      with kernel-ABC. $x$-axes indicate regions of nonzero prior density. In
+      particular, the prior on $m$ was DiscreteUniform(1, 5).
   }
   \label{fig:abchpd}
-\end{figure*}
-
-\section{Discussion}
+\end{figure}
+
+\clearpage
+
+\section*{Discussion}
 
@@
 networks can be simulated. We demonstrated the method on the BA model, which is
-a simple pcreferential attachment model giving rise to the power law degree
+a simple preferential attachment model giving rise to the power law degree
 distributions commonly observed in real-world networks.
@@
 By training a kernel-SVR classifier, we found that the $\alpha$ and $I$
-parameters, representing pcreferential attachment power and number of infected
-nodes, had a strong influence on tree shape. This was creflected in the relative
+parameters, representing preferential attachment power and number of infected
+nodes, had a strong influence on tree shape. This was reflected in the relative
 accuracy of the kernel-ABC estimates of these parameters. The total number of
@@
 parameters may translate to highly informative priors on quantities of
-interest. We observed a non-linear relationship between the pcreferential
+interest. We observed a non-linear relationship between the preferential
 attachment power $\alpha$ and the power law exponent $\gamma$
-(\cref{fig:gamma}). Thecrefore, placing a uniform prior on $\alpha$ between 0
+(\cref{fig:gamma}). Therefore, placing a uniform prior on $\alpha$ between 0
 and 2 is equivalent to placing an informative prior that $\gamma$ is close to
-2. Thecrefore, if we were primarily interested in $\gamma$ rather than
+2. Therefore, if we were primarily interested in $\gamma$ rather than
 $\alpha$, a more sensible choice of prior might have a shape informed by
@@
 most likely misspecified for these data. In particular, the average degree of a
-node in the network is equal to $2m$, and thecrefore is constrained to be a
+node in the network is equal to $2m$, and therefore is constrained to be a
 multiple of 2. Furthermore, we considered the case $m = 1$, where the network
-has no cycles, to be implausible and thecrefore assigned it zero prior
+has no cycles, to be implausible and therefore assigned it zero prior
 probability in one set of analyses. This forced the average degree to be at
 least four, which may be unrealistically high for sexual networks. The fact
-that the estimated values of $\alpha$ differed substatially for three datasets
-depending on whether or not $m = 1$ was allowed by the prior is futher evidence
+that the estimated values of $\alpha$ differed substantially for three datasets
+depending on whether or not $m = 1$ was allowed by the prior is further evidence
 of this potential misspecification. However, we note that for two of the
@@
 modelling assumptions include the network being connected and static, all
-transmission rates being equal, no removal after infection, and identical
-behaviour of all nodes. This last is particularly problematic, as we showed by
-simulating a network where some nodes exhibited a higher attachment power than
-others. The estimated attachment power was simply the average of the two
-values, indicating that, although we could characterize the network in
-aggregate, the estimated parameters could not be said to apply to any
-individual node. Despite these issues, we felt it was best to demonstrate the
-method first on a simple model. It is possible to use this framework to fit
-more complex models which address some of these issues, such as one
-incorporating heterogeneous node behaviour, which may prove a fruitful avenue
+transmission rates being equal, no removal after infection, identical behaviour
+of all nodes, and random sampling. The last two were addressed with small-scale
+experiments. We simulated a network where some nodes exhibited a higher
+attachment power than others, and found that the estimated attachment power was
+simply the average of the two values. This indicated that, although we could
+characterize the network in aggregate, the estimated parameters could not be
+said to apply to any individual node. The effect of biased sampling was
+investigated by analysing a transmission tree which had been sampled in a
+peer-driven fashion. The results were roughly in line with those for random
+sampling, however the estimated value of $\alpha$ was lower than the average
+for randomly-sampled trees. Further experiments would be necessary to fully
+explore the impact of these assumptions on the method's accuracy. However,
+despite these issues, we felt it was best to demonstrate the method first on a
+simple model. It is possible to use this framework to fit more complex models
+which address some of these issues, such as one incorporating heterogeneous
+node behaviour, which may prove a fruitful avenue
 for future investigations.
@@
 
-\section{Acknowledgements}
+\section*{Acknowledgements}
 
@@
 
+\clearpage
 \newpage
@@
 \setcounter{figure}{0}
+\setcounter{page}{1}
 
@@
   \caption{
-      Relationship between pcreferential attachment power parameter $\alpha$
+      Relationship between preferential attachment power parameter $\alpha$
       and power law exponent $\gamma$ for networks simulated under the BA
@@
 \begin{figure}[ht]
+    \centering
+    \includegraphics{mixed-posterior}
+    \caption{
+        Approximate marginal posterior distributions of Barab\'asi-Albert
+        model parameters obtained using kernel-ABC for a network with
+        heterogeneous node behaviour. Half of the nodes were attached with
+        $\alpha = 0.5$, and the other half with $\alpha = 1.5$ (vertical
+        dashed lines, top right). Other parameter values were $m = 2$, $I =
+        1000$, and $N = 5000$ (vertical dashed lines, other than top right).
+        Shaded areas indicate 95\% highest posterior density intervals.
+    }
+    \label{fig:mixed}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{peerdriven-posterior}
+    \caption{Approximate marginal posterior distributions of Barab\'asi-Albert
+        model parameters obtained using kernel-ABC for a network with
+        peer-driven sampling. An epidemic was simulated in the usual fashion,
+        but rather than being sampled at random, infected nodes were sampled
+        with a probability two times higher if they had any sampled neighbours
+        in the contact network. Vertical dashed lines indicate true parameter
+        values, and shaded areas indicate 95\% highest posterior density
+        intervals.
+    }
+    \label{fig:peerdriven}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[ht]
   \includegraphics{realdata-hpd-m2}
@@
       parameters of the BA network model, fitted to five published HIV datasets
-      with kernel-ABC. Regions of nonzero prior density are shown on $x$-axes.
+      with kernel-ABC. $x$-axes indicate regions of nonzero prior density. In
+      particular, the prior on $m$ was DiscreteUniform(2, 5).
   }
@@
 
+<<abc_posterior, echo=FALSE, results="asis">>=
+    N <- 5000
+    replicate <- 0
+    for (m in c(2, 3, 4)) {
+    for (I in c(1000, 2000)) {
+    for (alpha in c(0, 0.5, 1, 1.5)) {
+        fn <- sprintf("{abc-posterior/%.1f_%d_%d_%d_%d}.pdf", alpha, I, m, N, replicate)
+        lab <- sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate)
+        cat("\\begin{figure}[ht]\n")
+        cat("\\centering\n")
+        cat(paste0("\\includegraphics{", fn, "}\n"))
+        cat("\\caption{\n")
+        cat(sprintf("Approximate marginal posterior distributions of Barab\\'asi-Albert model parameters obtained using kernel-ABC for one simulated transmission tree with parameter values $\\alpha = %.1f$, $I = %d$, $m = %d$, and $N = %d$.
+Dashed lines indicate true values, and shaded areas show 95\\%% highest posterior density intervals.\n", alpha, I, m, N))
+        cat("}\n")
+        cat(paste0("\\label{", lab, "}\n"))
+        cat("\\end{figure}\n")
+        cat("\\clearpage\n\n")
+    }
+    }
+    }
+@
+
 \end{document}

### 2016-06-23 doc/thesis/appendix.Rnw
Most thesis changes done

@@
         parameters of the \gls{BA} network model, fitted to eleven \gls{HIV}
-        datasets with \software{netabc} using the prior $\gls{m} \sim
-        \DiscreteUniform(2, 5)$.
+        datasets with \software{netabc} using the prior \gls{m} $\sim$
+        DiscreteUniform(2, 5).
     ]{

### 2016-06-23 doc/thesis/conclusion.tex
Most thesis changes done

@@
 Due to the rapid advancement of nucleotide sequencing technology, viral
-sequence data data have become increasingly feasible to collect on a population
+sequence data have become increasingly feasible to collect on a population
 level. Through phylodynamic methods, these data offer a window into
-epidemiological processes which would otherwise be virtually impossible to
-study on a realistic scale.
+epidemiological processes that would otherwise be virtually impossible to study
+on a realistic scale.
 
@@
 will be of broad interest to the molecular evolution and epidemiology
-community, as it widens the field of epidemiological parameters which may be
+community, as it widens the field of epidemiological parameters that may be
 investigated through viral sequence data. In addition, the characterization of

### 2016-06-23 doc/thesis/discussion.Rnw
Most thesis changes done

@@
 attachment power \gls{alpha} and the fitted power law exponent \gls{gamma}
-(\cref{fig:gamma}, although see the discussion in \cref{TODO}). Therefore,
-placing a uniform prior on \gls{alpha} between 0 and 2 is equivalent to placing an
-informative prior that \gls{gamma} is close to 2. If we were primarily
+(\cref{fig:gamma}, although see the discussion in \cref{subsec:pa}). Therefore,
+placing a uniform prior on \gls{alpha} between 0 and 2 is equivalent to placing
+an informative prior that \gls{gamma} is close to 2. If we were primarily
 interested in \gls{gamma} rather than \gls{alpha}, a more sensible choice of
@@
 \del{With respect to the preferential attachment power $\alpha$, the six
-datasets analysed fell into two categories (\cref{fig:abchpd}). First, we
+datasets analyzed fell into two categories (\cref{fig:abchpd}). First, we
 estimated a preferential attachment power close to 1, indicating linear
@@
 estimate the parameters of a more sophisticated \acrlong{PA} model
-incorporating heterogeneous node behaviour. When used to analyse
+incorporating heterogeneous node behaviour. When used to analyze
 population-level longitudinal partner count data, they found \gls{alpha} values
@@
 confidence intervals around \gls{alpha} narrowed when other parameters were
-marginalized out (\cref{fig:maginal}). Thus, it is possible that estimates of
+marginalized out (\cref{fig:marginal}). Thus, it is possible that estimates of
 \gls{alpha} could be made more precise by specifying either exact values or
 informative priors on the other \gls{BA} parameters. We argued informally in
-\cref{subsec: whyba} that \gls{I}, \gls{m}, and \gls{N} can sometimes be
+\cref{subsec:whyba} that \gls{I}, \gls{m}, and \gls{N} can sometimes be
 estimated by standard epidemiological methods, although the difficulty may
@@
 illegal or stigmatized, which may represent a barrier to testing, treatment,
-and/or disclosure of status. Several of the studies whose data we analysed
+and/or disclosure of status. Several of the studies whose data we analyzed
 report the number of new infections over the study period. We expect these
@@
 estimate \gls{I}. In the following, we discuss any information available about
-the \gls{HIV} prevalence in the analysed study populations, and relate this to
+the \gls{HIV} prevalence in the analyzed study populations, and relate this to
 our estimates. However, due to the aforementioned challenges, this information
@@
     (\Sexpr{d1[dataset == "grabowski2014" & parameter == "I", round(mean)]}),
-but the value 1434 was contained ith the 95\% \gls{HPD} interval
+but the value 1434 was contained within the 95\% \gls{HPD} interval
     (\Sexpr{d1[dataset == "grabowski2014" & parameter == "I", round(hpd.min.95)]} -

### 2016-06-23 doc/thesis/discussion.tex
Most thesis changes done

@@
 estimate the parameters of a more sophisticated \acrlong{PA} model
-incorporating heterogeneous node behaviour. They found \gls{alpha} values
-ranging from 0.26 to 0.62, depending on the gender and time period considered. 
-Our estimates of \gls{alpha} for the \textcite{niculescu2015recent} was above
-this range under both priors, as were the estimates for the
-\textcite{wang2015targeting} data and the BC data when $\gls{m} = 1$ was
-disallowed by the prior. The dataset investigated by
+incorporating heterogeneous node behaviour. When used to analyze
+population-level longitudinal partner count data, they found \gls{alpha} values
+ranging from 0.26 to 0.62 depending on the gender and time period considered.
+Our estimates of \gls{alpha} for the IDU/Romania was above this range under
+both priors, as were the estimates for the MSM/Beijing data and the BC data
+when $\gls{m} = 1$ was disallowed by the prior. The dataset investigated by
 \textcite{de2007preferential} was derived from a survey of a random sample of
@@
 
-\add{Both \textcite{de2007preferential} and \textcite{novitsky2014impact} studied
+\add{Both \textcite{de2007preferential} and the HET/Botswana data studied
 populations whose primary risk factor for \gls{HIV} infection was heterosexual
@@
 \gls{alpha} for these data was 
-    X%\S%expr{d[m_min == 1 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]}
+    0.55
         or 
-    X%\S%expr{d[m_min == 2 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]},
-depending on the prior on \gls{m}; \textcite{de2007preferential} estimated
-0.54, 0.57, and 0.29 for 3-year, 5-year, and lifetime partnership networks
-respectively for the female portion of their sample.}
+    0.53,
+depending on the prior on \gls{m}. Similarly, \textcite{de2007preferential}
+estimated 0.54, 0.57, and 0.29 for 3-year, 5-year, and lifetime partnership
+networks respectively for the female portion of their sample.}
 
@@
     (939),
-but the value 1434 was contained ith the 95\% \gls{HPD} interval
+but the value 1434 was contained within the 95\% \gls{HPD} interval
     (226 -

### 2016-06-23 doc/thesis/intro.tex
Most thesis changes done

@@
 
-%Although phylodynamics is quite new, these phenomena have been studied in
-%evolutionary biology for some time. Viral phylogenies are a specific version of
-%a more general class of trees called \defn{gene trees}, which represent the
-%evolutionary history of a section of genetic material. Transmission trees, on
-%the other hand, are highly analogous to \defn{species trees}, whose tips are
-%species and internal nodes are common ancestors. This analogy derives from the
-%functional similarity between transmission and allopatric speciation. Hence,
-%the potential discordance between transmission trees and viral phylogenies is
-%the similar to that between gene and species trees, which is called
-%\defn{incomplete lineage sorting}.
-
-%What we have just discussed is a two-step procedure for estimating the
-%transmission tree. First, a viral phylogeny is constructed from genetic
-%sequence data, and then it is rooted and time-scaled into a transmission tree.
-%This approach is straightforward, frequently used, and has the advantage of
-%leveraging tried-and-true tools for phylogenetic inference. However, it also
-%has drawbacks, perhaps the most obvious being the multiplication of errors
-%produced by the separate steps. One commonly used alternative method is to
-%directly estimate a time-scaled phylogeny by simultaneously inferring the tree
-%topology, its root and branch lengths, and the parameters of a \defn{molecular
-%clock} model. A molecular clock is a hypothesis about the evolutionary rates
-%along the branches of the tree, such as that they are all equal (a \defn{strict
-%clock}) or that they are \gls{iid} from a common distribution (a \defn{relaxed
-%clock}). This inference is usually done in a Bayesian framework using
-%\gls{MCMC}, so that prior information (including the tip dates) can be included
-%in the analysis, and the so-called nuisance parameters of the molecular clock
-%model can be marginalized out. Software packages for performing these analyses
-%include BEAST~\autocite{bouckaert2014beast} and
-%MrBayes~\autocite{ronquist2012mrbayes}.
-
-%Other methods are tailor-made for inferring transmission trees.
-%\textcite{didelot2014bayesian} develop a Bayesian approach which allows
-%transmissions to occur anywhere along the
-%branches of a transmission tree, rather than being constrained to the branching
-%points in the viral phylogeny. The method requires sampling of every infected
-%individual, although the authors indicate that it could be extended to relax
-%this assumption. \textcite{cottam2008integrating} describe a likelihood-based
-%method which enumerates all transmission trees consistent with an established
-%phylogeny, assigning each a likelihood based on other epidemiological data.
-%This approach is novel in its integration of data from multiple sources,
-%however because it enumerates a large portion of the tree space, it is unlikely
-%to scale to larger epidemics. \textcite{ypma2012unravelling} develop a joint
-%likelihood function integrating temporal, geographic, and genetic observations,
-%and use Bayesian \gls{MCMC} to estimate both the tree and the parameters of the
-%likelihood function. Their approach can handle missing data and produces high
-%resolution transmission trees when multiple types of data are available. A
-%different approach is undertaken by \textcite{jombart2011reconstructing}, who
-%describe a method to build transmission trees directly from sequence data,
-%contingent on the common ancestors also being sampled. This makes the method
-%attractive for slow-evolving pathogens, but less practical for viral outbreaks
-%where samples from common ancestors are unlikely to be available.
-
 \subsection{Tree shapes}
@@
 space. Computing $\varphi(x)$ may be computationally prohibitive due to the
-dimension of $\F$. The utility of a kernel function $K$ is that it is
+dimension of $\F$. The utility of a kernel function is that it is
 constructed in such a way that it can compute the inner product without
@@
 of the population of particles after each sequential step in accordance with
-their importance weights. A similar idea, termed \defn{particle rejuvination},
+their importance weights. A similar idea, termed \defn{particle rejuvenation},
 was proposed by \textcite{liu1995blind}. These approaches cause particles with
@@
 ratio, which must be calculated by computing the individual likelihoods and
-dividing them. If the likeilhood is intractible, this is clearly not a viable
+dividing them. If the likelihood is intractable, this is clearly not a viable
 approach.}

### 2016-06-23 doc/thesis/methods_results.Rnw
Most thesis changes done

@@
 As described in \cref{sec:smc}, \software{netabc} keeps track of a population
-of particles $x^{(k)}$ indexed by an integere $k$, each of which contains
+of particles $x^{(k)}$ indexed by an integer $k$, each of which contains
 particular parameter values $\theta^{(k)}$ for the \add{contact network} model
@@
 contrast to most commonly studied mathematical models, the likelihood
-$\lik(\gls{T} \mid \theta)$ is also likely to be intractible in our case.}
+$\lik(\gls{T} \mid \theta)$ is also likely to be intractable in our case.}
 
@@
 problem of fitting contact network models to phylogenies seems to be of the
-\emph{doubly-intractible} type~\autocite{murray2012mcmc}, which would imply that
+\emph{doubly-intractable} type~\autocite{murray2012mcmc}, which would imply that
 these models are not amenable to neither \gls{ML} nor Bayesian inference
-techniques. Although both methods are able to cope with an intractible
+techniques. Although both methods are able to cope with an intractable
 normalizing constant (for example, by local search for \gls{ML} or Bayesian
@@
 this case $2\gls{m}$, is directly related to \gls{R0}, the basic reproductive
-number~\autocite{britton2002bayesian}. \gls{R0} quantifies the number of
-secondary infections ultimately caused by one infected individual; higher
+number~\autocite{britton2002bayesian}. \gls{R0} quantifies the average number
+of secondary infections ultimately caused by one infected individual; higher
 \gls{R0} generally indicates faster epidemic growth and/or larger eventual
 epidemic size~\autocite{anderson1992infectious}. In a homogeneously mixed
-population, the proportion of the population which must be vaccinated to
-control an epidemic can be expressed in terms of
-\gls{R0}~\autocite{anderson1982directly}. Although optimal vaccination
-strategies may differ in heterogeneous contact
+population, the theoretically expected proportion of the population which must
+be vaccinated to control an epidemic (the \defn{vaccination threshold}) can be
+expressed in terms of \gls{R0}~\autocite{anderson1982directly}. Although
+optimal vaccination strategies may differ in heterogeneous contact
 structures~\autocite{ma2013importance}, it is reasonable to suppose that there
@@
 superspreaders~\autocite{kemper1980identification}, characterize the network
-structure. Superspreaders have been hypothesized to play a greater-than-average
-role in the spread of several diseases~\autocite{stadler2013uncovering,
+structure. Superspreaders have been hypothesized to play a disproportionately
+greater role in the spread of several diseases~\autocite{stadler2013uncovering,
 shen2004superspreading}. If so, network-based
@@
 characteristic, the removal of random nodes (such as through a police
-crackdown) might inadventently make it easier for epidemics to spread. When
+crackdown) might inadvertently make it easier for epidemics to spread. When
 individuals with only one or two connections lose them, they might tend to seek
 out well-known (that is, high-degree) members of the community, thus increasing
-those individuals' connectivity even further.}
-
-\add{All four \gls{BA} parameters can be estimated using more conventional
-approaches, but this estimation may be challenging depending on characteristics
-of the epidemic under consideration. In theory, any network parameter can be
-estimated by explicitly constructing the contact network, although this is
-highly resource intensive and is hampered by misreporting and other
-challenges~\autocite{eames2015six}. All parameters become more difficult to
-estimate when the infected population is ``hidden'' due to illegal or
-stigmatized behaviour, as is sometimes the case with \gls{HIV} outbreaks among
-\gls{IDU}, \gls{MSM}, or sex workers. Of course, phylodynamic methods are
-equally ineffective when the population is completely hidden, since we must at
-least be able to sample viral sequences from its members. If a population is
-accessible, \gls{N} and \gls{I} may be estimable by surveying and/or testing
-each member of the population for the disease. However, this will not tell us
-if there are large compartments of the population that we simply have not
-sampled. Our hope is that estimating \gls{N} and \gls{I} phylogenetically might
-provide this additional information. The average degree of the network,
-$2\gls{m}$, is also estimable by a survey, although individuals may be
-unwilling or unable to disclose how many contacts they have had. The estimation
-of \gls{alpha} is more complex, as this parameter is most strongly reflected in
-the connectivity of very high degree nodes, who are rare in the population.
-Locating them might require contact tracing, or respondent-driven
+those individuals' connectivity even further. Although we do not consider a 
+dynamic network in this work, it is not unreasonable that the network could be
+close to static over short periods of time, and this static structure might be
+informative of future dynamic behaviour.}
+
+\add{Though it is theoretically possible to estimate all four \gls{BA}
+parameters using more conventional approaches, the cost of doing so may be
+high, in addition to other situation-specific challenges. In theory, any
+network parameter can be estimated by explicitly constructing the contact
+network, although this is highly resource intensive and is hampered by
+misreporting and other challenges~\autocite{eames2015six}. All parameters
+become more difficult to estimate when the infected population is ``hidden''
+due to illegal or stigmatized behaviour, as is sometimes the case with
+\gls{HIV} outbreaks among \gls{IDU}, \gls{MSM}, or sex workers. In such cases,
+phylodynamic methods may offer the advantage of providing information about the
+whole population from only a small sample. A survey, for example, will not tell
+us if there are large groups of the population that we simply have not sampled.
+Our hope is that estimating \gls{N} and \gls{I} phylogenetically might provide
+this additional information. The average degree of the network, $2\gls{m}$, is
+also estimable by a survey, although individuals may be unwilling or unable to
+disclose how many contacts they have had. Sequence data also provide an
+advantage in this respect -- they are objective and do not share the same
+biases as self-reported partner counts. The estimation of \gls{alpha} is more
+complex, as this parameter is most strongly reflected in the connectivity of
+very high degree nodes, who are rare in the population. Locating them might
+require contact tracing, or respondent-driven
 sampling~\autocite{heckathorn1997respondent}. Even if the full degree
@@
 experiment with trees located throughout the parameter space allowed us to
-beter quantify the identifiability. Furthermore, doing marginal estimation
+better quantify the identifiability. Furthermore, doing marginal estimation
 (that is, estimating one parameter with all others fixed) can provide insight
@@
 increasing the computing power used for \gls{SMC}. We ran one simulation using
-a larger number of pakticles, more simulated datasets per particle, and a
+a larger number of particles, more simulated datasets per particle, and a
 higher value for \gls{alphaess}. A substantial improvement in accuracy
@@
 \cref{subsec:treeshape}). In contrast with Sackin's index, the \gls{nltt} does
-not explicity consider the trees' topologies, but it does use their normalized
+not explicitly consider the trees' topologies, but it does use their normalized
 branch lengths. While the \gls{nltt} is a newly developed statistic not yet in
@@
 score was taken as the point estimate for the test value.} \add{For all
-parameters except \gls{m}}, 50\% and 95\% highest denstity intervals wer
+parameters except \gls{m}}, 50\% and 95\% highest density intervals were
 obtained using the \software{hpd} function in the \software{TeachingDemos}
@@
 the model, but does not invalidate the use of our method in cases where the
-parameters are more identifiable. Two retrospective experiments were peformed
+parameters are more identifiable. Two retrospective experiments were performed
 to disambiguate some of the observed errors: one where we ran a simulation with
@@
 \add{To evaluate the effects of the true parameter values on the accuracy of
-the posterior mean estimates, we analysed the \gls{alpha} and \gls{I}
+the posterior mean estimates, we analyzed the \gls{alpha} and \gls{I}
 parameters individually using \glspl{GLM}. The response variable was the error
@@
 greater impact on accuracy than the proportion of infected individuals these
-tips represented. However, \gls{m}, both the number and proportion of sampled
-tips had a strong impact or the accuracy of the \gls{kSVR}
+tips represented. However, for \gls{m}, both the number and proportion of
+sampled tips had a strong impact or the accuracy of the \gls{kSVR}
 (\cref{fig:mcrossv}).}
@@
 appropriate to consider. An overestimate of 100 individuals would be very
-misleading if the true population was only of size 100, but almost negligable
+misleading if the true population was only of size 100, but almost negligible
 in a population of size 5000. The mean relative errors for \gls{I} and \gls{N}
@@
     \includegraphics[width=\textwidth]{abc-point-estimate-m2}
-    \vspace{6pt}
     \caption[
@@
 
-\add{To analyse the effects of the true parameter values on the
+\add{To analyze the effects of the true parameter values on the
 accuracy our estimates of \gls{alpha} and \gls{I}, we fitted one \glspl{GLM}
@@
     ($p \Sexpr{pp(alpha.glm[Parameter == "m", p], eq=TRUE)}$)
-but was not significatly affected by \gls{I} (\cref{tab:glmalpha}).
+but was not significantly affected by \gls{I} (\cref{tab:glmalpha}).
 Qualitatively, \gls{alpha} seemed to be only weakly identifiable between the
@@
     #lab <- paste(lab, collapse = ",")
-    worst <- d[alpha_error > quantile(alpha_error, 0.75) & I_error > quantile(I_error, 0.75),]
-    best <- d[alpha_error < quantile(alpha_error, 0.25) & I_error < quantile(I_error, 0.25),]
-    stopifnot(nrow(worst) == 1)
-    stopifnot(nrow(best) == 1)
+    worst.alpha <- d[,mean(alpha_error), by=true_alpha][,true_alpha[which.max(V1)]]
+    best.alpha <- d[,mean(alpha_error), by=true_alpha][,true_alpha[which.min(V1)]]
+    worst.I <- d[,mean(I_error), by=true_I][,true_I[which.max(V1)]]
+    best.I <- d[,mean(I_error), by=true_I][,true_I[which.min(V1)]]
+    worst.m <- d[,mean(m_error), by=true_m][,true_m[which.max(V1)]]
+    best.m <- d[,mean(m_error), by=true_m][,true_m[which.min(V1)]]
+
+    worst <- d[true_alpha == worst.alpha & true_m == worst.m & true_I == worst.I & replicate == 0,]
+    best <- d[true_alpha == best.alpha & true_m == best.m & true_I == best.I & replicate == 0,]
 
@@
 simulation.} \add{show one- and two-dimensional marginal distributions for a
-simulation with relatively low error, where the errors in \gls{alpha} and
-\gls{I} were each below the 25\% quantiles. The parameters for this simulation
-were 
+simulation with relatively low error. Each parameter was chosen based on its
+mean error rate across all simulations. The parameters for this simulation were 
     $\gls{alpha} = \Sexpr{best[,true_alpha]}$,
@@
     $\gls{N} = \Sexpr{best[,true_N]}$.
-\Cref{fig:abchigh1,abchigh2} show the equivalent marginals for a different
-simulation with relatively high error rates above the 75\% quantiles. The
+\Cref{fig:abchigh1,fig:abchigh2} show the equivalent marginals for a different
+simulation with relatively high error rates for each individual parameter. The
 parameters were
@@
 transmission tree. The estimates of \gls{m} are apparently unaffected by
-marginalizing out the other parameters, coroborating the prevous experiments'
+marginalizing out the other parameters, corroborating the previous experiments'
 findings that \gls{m} is not an identifiable parameter from scaled tree shapes.
@@
     \caption[
-        Comparsion of \gls{BA} parameter estimates obtained with
+        Comparison of \gls{BA} parameter estimates obtained with
         \software{netabc} on the same dataset with different \gls{SMC}
@@
     ]{
-        Comparsion of \gls{BA} parameter estimates obtained with
+        Comparison of \gls{BA} parameter estimates obtained with
         \software{netabc} on the same dataset with different \gls{SMC}
@@
 tree requires non-contemporaneous tips.} The datasets are summarized in
-\cref{tab:data}. In addition \add{to the published data}, we analysed an
+\cref{tab:data}. In addition \add{to the published data}, we analyzed an
 in-house dataset sampled from \gls{HIV}-positive individuals in British
@@
 
-\del{Three} \add{Six} of the datasets (MSM/Shanghai, HET/Botswana, HET/Uganda, 
+\del{Three} \add{Four} of the datasets (MSM/Shanghai, HET/Botswana, HET/Uganda, 
 and MSM/USA) were initially larger than the others, containing 1265, 1299,
@@
 similar to the smaller datasets. For the MSM/Shanghai dataset, we detected a
-cluster of size 280 using a patristic distance cutoff of 0.02 as described
+cluster of size 280 using a patriotic distance cutoff of 0.02 as described
 previously~\autocite{poon2015impact}. Only sequences within this cluster were
 carried forward. For the HET/Uganda, HET/Botswana, and MSM/USA datasets, no
-large clusters were detected using the same cutoff, so we analysed subsets of
+large clusters were detected using the same cutoff, so we analyzed subsets of
 size 225, 180, and 180 respectively. \add{The subset of the HET/Uganda data was
-chosen by eye such that the individuals were monophyletic in both the
+chosen by eye such that the individuals were monopolistic in both the
 \textit{gag} and \textit{env} trees. The other subsets were arbitrarily chosen
@@
 distributions. The values of \gls{I} were fairly robust to restricting the
-prior to $\gls{m} \geq 2$ (compare \cref{fig:abchpd,fig:abchdpm2}), although 
+prior to $\gls{m} \geq 2$ (compare \cref{fig:abchpd,fig:abchpdm2}), although 
 the 95\% \gls{HPD} intervals were slightly narrower (average width
@@
   \Sexpr{replace_number(pmean1[,sum(m == 1)])}
-of the datasets analysed. The widths of the 95\% \gls{HPD} intervals varied
+of the datasets analyzed. The widths of the 95\% \gls{HPD} intervals varied
 from 0 (all the mass on the estimated value) to 5 (the entire prior region).

### 2016-06-23 doc/thesis/methods_results.tex
Most thesis changes done

@@
 The \gls{kSVR} classifier for \gls{alpha} had an average $R^2$ of 
-    0.92,
+    0.98,
 compared to 
@@
 % latex table generated in R 3.2.3 by xtable 1.8-2 package
-% Mon Jun 20 17:18:05 2016
+% Thu Jun 23 15:40:35 2016
 \begin{table}[ht]
@@
     ($p =0.01$)
-but was not significatly affected by \gls{I} (\cref{tab:glmalpha}).
+but was not significantly affected by \gls{I} (\cref{tab:glmalpha}).
 Qualitatively, \gls{alpha} seemed to be only weakly identifiable between the
@@
 simulation.} \add{show one- and two-dimensional marginal distributions for a
-simulation with relatively low error, where the errors in \gls{alpha} and
-\gls{I} were each below the 25\% quantiles. The parameters for this simulation
-were 
-    $\gls{alpha} = 1$,
+simulation with relatively low error. Each parameter was chosen based on its
+mean error rate across all simulations. The parameters for this simulation were 
+    $\gls{alpha} = 1.5$,
     $\gls{I} = 1000$,
+    $\gls{m} = 4$, and
+    $\gls{N} = 5000$.
+\Cref{fig:abchigh1,fig:abchigh2} show the equivalent marginals for a different
+simulation with relatively high error rates for each individual parameter. The
+parameters were
+    $\gls{alpha} = 0$,
+    $\gls{I} = 2000$,
     $\gls{m} = 2$, and
-    $\gls{N} = 5000$.
-\Cref{fig:abchigh1,abchigh2} show the equivalent marginals for a different
-simulation with relatively high error rates above the 75\% quantiles. The
-parameters were
-    $\gls{alpha} = 1$,
-    $\gls{I} = 2000$,
-    $\gls{m} = 3$, and
     $\gls{N} = 5000$.
@@
     \centering
-    \includegraphics[width=\textwidth]{{abc-posterior/1.0_1000_2_5000_0}.pdf}
+    \includegraphics[width=\textwidth]{{abc-posterior/1.5_1000_4_5000_0}.pdf}
     \caption[
@@
     \centering
-    \includegraphics[width=\textwidth]{{abc-posterior-2d/1.0_1000_2_5000_0}.pdf}
+    \includegraphics[width=\textwidth]{{abc-posterior-2d/1.5_1000_4_5000_0}.pdf}
     \caption[
@@
     \centering
-    \includegraphics[width=\textwidth]{{abc-posterior/1.0_2000_3_5000_0}.pdf}
+    \includegraphics[width=\textwidth]{{abc-posterior/0.0_2000_2_5000_0}.pdf}
     \caption[
@@
     \centering
-    \includegraphics[width=\textwidth]{{abc-posterior-2d/1.0_2000_3_5000_0}.pdf}
+    \includegraphics[width=\textwidth]{{abc-posterior-2d/0.0_2000_2_5000_0}.pdf}
     \caption[
@@
   zero
-of the datasets analysed. The widths of the 95\% \gls{HPD} intervals varied
+of the datasets analyzed. The widths of the 95\% \gls{HPD} intervals varied
 from 0 (all the mass on the estimated value) to 5 (the entire prior region).

### 2016-06-23 doc/thesis/objective.tex
Most thesis changes done

@@
 marginal probability of $T$ which acts as a normalizing constant on the
-posterior (see \cref{chp:prelim} for a review of mathematical modeling and
+posterior (see \cref{chp:prelim} for a review of mathematical modelling and
 Bayesian inference, including definitions of these concepts). As we shall show

### 2016-06-23 doc/thesis/revision_1_summary.tex
Most thesis changes done

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{mathptmx}
\usepackage{amsmath}
\begin{document}

\section*{Major changes}

\subsection*{Structure}

\begin{itemize}
  \item Took likelihood calculations out of objective section and moved them
    into research chapter (sec. 2.1.4 "Justification for approach").
  \item Instead of displaying many posterior distributions from synthetic data 
    in the appendices, only show two in the main text: one "good" (ie. low
    error rate) and one "bad".
  \item Structure of chapter 2 has been rearranged: one section about method,
    one about synthetic data experiments, one about real data, and then the
    discussion. 
\end{itemize}

\subsection*{Analysis}

\begin{itemize}
  \item It turns out that only networks with alpha = 1 have a power law
    distribution (p. 12). However power laws still capture the salient features
    of the degree distributions (fig. A26) so I continue to report them.
  \item Throughout, maximum a posteriori estimates were replaced with posterior
    means, which are more conventional and give better results.
  \item Removed ad-hoc statistical tests from results on grid search and ABC.
    For grid search, only error distributions were reported numerically.
    Qualitative statements were made about kernel score distributions. For ABC,
    effects of each parameter on error rates was analysed with a p-value
    corrected GLM (p. 52-53).
  \item Two extra analyses to address possible sources of ABC error (a reviewer
    asked for this) described on p. 46, results on p. 62-63.
  \item Show some 2-dimensional marginals (p. 55, 57).
  \item The real data analysis has been significantly expanded to include six
    more datasets.
  \item Analysed three datasets with both env and gag genes and compare results
    (fig. 2.17 on p. 65).
  \item Found several published estimates of power law exponent gamma from
    literature (table 2.9 on p. 74). 
\end{itemize}

\subsection*{Writing}

\begin{itemize}
  \item Added some introduction to SMC section to clarify what the relationship
    to our problem is, ie. we are trying to make a particle approximation to
    the posterior.
  \item Same thing for ABC section (p. 21-23).
  \item Added summary of chapter 1 (p. 26-27).
  \item Added introductory paragraph to chapter 2 (p. 28).
  \item Expanded on why ABC is needed in this setting (p. 33-36), why we should
    study the BA model (p. 36-37).
  \item Wrote high-level summary and motivation of all synthetic data
    experiments (p. 38-39, 40, 44).
  \item Results section for real data experiments has been entirely rewritten.
  \item Discussion of real data results is also entirely new. I relate our
    results to a couple of papers looking at preferential attachment (PA)
    networks, one which found sub-linear PA for real world data (which we also
    found), and one which claimed IDU networks were more PA-ish than sexual
    networks (which we also found too).
\end{itemize}

\section*{Minor changes}

\begin{itemize}
  \item Added text to abstract (UBC limit is 350 words).
  \item Better definitions for "production rule", "decay factor", and
    "ladderize" (p. 8-9).
  \item More conventional definition of importance sampling (p. 16).\
  \item Added subsection under SMC for actual SMC (SIS with resampling), plus
    discussion of particle degeneracy (p. 18).
  \item Throughout, multi-step mathematical derivations have had justifications
    added beside each step (eg. p. 20).
  \item Added a paragraph about the importance of the distance function for ABC
    (p. 23-24).
  \item Changed "$\alpha$" from Del Moral paper to "$\alpha_{\text{ESS}}$" to avoid
    confusion (p. 32).
  \item Fig. 2.7 is now in color.
  \item Added notches for 50\% highest density intervals (similar to IQR) in
    fig. 2.8.
  \item Reported relative errors for I and N in grid search.
  \item Refer to datasets as eg. "IDU/Romania data" instead of "Niculescu et
    al. data" for easier interpretation.
  \item Proper scaling and legends for fig. 2.18 (p. 69).
\end{itemize}

\end{document}

### 2016-06-25 doc/defence/presentation.tex
defences and papers and svm, oh my!

--- 2016-04-25 doc/hiv-dynamics/presentation.tex
@@
 \usetikzlibrary{decorations.pathreplacing}
+\usetikzlibrary{calc}
 
@@
 
-\title{Phylogenetic inference of contact network parameters with kernel approximate Bayesian computation}
-\author[RMM, RHL \& AFYP]{Rosemary M McCloskey$^1$ \and Richard H Liang$^1$ \and Art FY Poon$^{1,2}$}
+\title{Phylogenetic inference of contact network parameters with approximate Bayesian computation}
+\author[RMM \& AFYP]{Rosemary M McCloskey$^1$ \and Bioinformatics Training Program \\ Supervisor: Art FY Poon$^{1,2}$}
 \institute[UBC \& BCCfE]{$^1$BC Centre for Excellence in HIV/AIDS, Vancouver, Canada \\ $^2$Department of Medicine, University of British Columbia, Vancouver, Canada}
-\date{HIV Dynamics \& Evolution, Woods Hole, USA, April 25, 2016}
+\date{M.Sc. Defence, July 26, 2016}
 
@@
 
-\begin{frame}{Most epidemiological models assume homogeneous mixing}
+\begin{frame}{Modeling in epidemiology}
+  \centerline{\includegraphics[width=0.9\textwidth]{compartments}}
+\end{frame}
+
+\begin{frame}{Contact heterogeneity can affect epidemic expansion}
+  \centerline{\includegraphics[width=0.9\textwidth]{sir-trajectories}}
+\end{frame}
+
+\begin{frame}{Contact networks capture contact heterogeneity}
+  \begin{center}
+  \begin{tikzpicture}
+    \node (c) {\includegraphics[width=4cm]{contactnet-empty}};
+
+    \uncover<2->{
+    \node (a) [right=of c.north east, anchor=west] {nodes = individuals};
+    \draw [->, thick, >=stealth] (c.north east) ++(-1.5cm, -0.5cm) -- (a.west);
+    }
+
+    \uncover<3->{
+    \node (b) [below=0.5 of a] {edges = contacts};
+    \draw [->, thick, >=stealth] (c.north east) ++(-0.75cm, -1.25cm) -- (b.west);
+    }
+  \end{tikzpicture}
+  \end{center}
+\end{frame}
+
+\begin{frame}{Network models formally describe network structure}
   \begin{columns}
-    \begin{column}{0.6\textwidth}
-      \centering
-      \includegraphics[scale=0.6]{compartments}
-      \vspace{1cm}
-
-      \uncover<3->{\includegraphics[width=\textwidth]{sir-trajectories}}
-    \end{column}
-    \begin{column}{0.4\textwidth}
-      \vspace{-2cm}
-      \begin{itemize}
-        \setlength{\itemsep}{24pt}
+    \column{0.55\textwidth}
+    \begin{minipage}[c][.6\textheight][c]{\linewidth}
+      \begin{enumerate}
+        \setlength{\itemsep}{12pt}
+        \item Start with \only<1-4>{10}\only<5->{\alert{$n$}} women and 
+          \only<1-4>{10}\only<5->{\alert{$n$}} men.
         \uncover<2->{
-        \item Often provide a reasonable approximation in practice.
+        \item Randomly form \only<2-4>{20}\only<5->{\alert{$p \times n$}} male-female partnerships.
         }
         \uncover<3->{
-        \item Can be inaccurate when substantial contact heterogeneity exists.
+        \item Randomly rewire each edge with \only<3-4>{10}\only<5->{\\\alert{$q$ }}\% probability.
         }
-      \end{itemize}
-    \end{column}
+      \end{enumerate}
+    \end{minipage}
+    \column{0.5\textwidth}
+    \begin{minipage}[c][.6\textheight][c]{\linewidth}
+      \only<1>{\includegraphics[width=5cm, page=1]{homophily}}
+      \only<2>{\includegraphics[width=5cm, page=2]{homophily}}
+      \only<3>{\includegraphics[width=5cm, page=3]{homophily}}
+      \only<4->{\includegraphics[width=5cm, page=4]{homophily}}
+    \end{minipage}
+  \end{columns}
+
+  \vspace{0.5cm}
+  \begin{columns}
+    \column{0.55\textwidth}
+    \uncover<6->{\centerline{\large{network model}}}
+    \column{0.55\textwidth}
+    \uncover<6->{\centerline{\large{contact network}}}
   \end{columns}
@@
 
-\begin{frame}{Network models capture contact heterogeneity}
+\begin{frame}{Barab\'asi-Albert model incorporates preferential attachment}
   \begin{columns}
-    \begin{column}{0.5\textwidth}
-      \includegraphics[width=\textwidth]{contactnet-empty}
-    \end{column}
-    \begin{column}{0.5\textwidth}
-      \begin{itemize}
-        \setlength{\itemsep}{12pt}
-          \pause
-        \item May offer more accurate predictions for highly structured
-          populations.
-          \pause
-        \item Network parameters may be of interest for their own sake,
-          \textit{e.g.} are there superspreaders?
-          \pause
-        \item Extremely difficult to estimate in practice.
-      \end{itemize}
-    \end{column}
+  \column{0.5\textwidth}
+  \begin{tikzpicture}[
+      scale=2,
+      every edge/.style={thick}
+    ]
+    \node (n1) at (1, 0) [circle, fill=black] { };
+    \node (n2) at (0.86, 0.5) [circle, fill=black] { };
+    \node (n3) at (0.5, 0.86) [circle, fill=black] { };
+    \draw (n1) -- (n2) -- (n3) -- (n1);
+
+    \uncover<2->{
+      \node (n4) at (0, 1) [circle, fill=black] { };
+    }
+    \only<2-3>{
+      \node [right=0 of n4.north east, anchor=south west] {$m = 2$};
+    }
+    \only<2-4>{
+      \draw (n4) -- ++ (0.2, -0.2);
+      \draw (n4) -- ++ (-0.2, -0.2);
+    }
+
+    \node [right=0 of n1] {%
+      \only<3-4>{$\Pr \propto 2^\alpha$}%
+      \only<6-7>{$\Pr \propto 3^\alpha$}%
+    };
+    \node [right=0 of n2] {%
+      \only<3-4>{$\Pr \propto 2^\alpha$}%
+      \only<6-7>{$\Pr \propto 3^\alpha$}%
+    };
+    \node [right=0 of n3] {%
+      \only<3-4>{$\Pr \propto 2^\alpha$}%
+      \only<6-7>{$\Pr \propto 2^\alpha$}%
+    };
+
+    \only<4>{
+      \draw (n4) -- ++ (0.2, -0.2) -- (n2);
+      \draw (n4) -- ++ (-0.2, -0.2) -- (n1);
+    }
+
+    \uncover<5->{
+      \draw (n4) -- (n2);
+      \draw (n4) -- (n1);
+    }
+
+    \uncover<6->{
+      \node (n5) at (-0.5, 0.86) [circle, fill=black] { };
+    }
+
+    \only<6-7> {
+      \draw (n5) -- ++ (0.2, -0.2);
+      \draw (n5) -- ++ (-0.2, -0.2);
+    }
+
+    \uncover<6-7>{ \node [right=0 of n4.north east, anchor=south west] { $\Pr \propto 2^\alpha$ }; }
+
+    \only<7>{
+      \draw (n5) -- ++ (0.2, -0.2) -- (n3);
+      \draw (n5) -- ++ (-0.2, -0.2) -- (n2);
+    }
+
+    \uncover<8->{
+      \draw (n5) -- (n3);
+      \draw (n5) -- (n2);
+    }
+
+    \uncover<9->{
+      \node (n5) at (-0.86, 0.5) [circle, fill=black] { };
+      \draw (n5) -- (n2);
+      \draw (n5) -- (n1);
+    }
+
+    \uncover<10->{
+      \node (n6) at (-1, 0) [circle, fill=black] { };
+      \draw (n6) -- (n3);
+      \draw (n6) -- (n4);
+    }
+
+    \uncover<11->{
+      \node (n7) at (-0.86, -0.5) [circle, fill=black] { };
+      \node (n8) at (-0.5, -0.86) [circle, fill=black] { };
+      \node (n9) at (0, -1) [circle, fill=black] { };
+      \node (n10) at (0.5, -0.86) [circle, fill=black] { };
+      \node (n11) at (0.86, -0.5) [circle, fill=black] { };
+
+      \draw (n7) -- (n2);
+      \draw (n7) -- (n4);
+      \draw (n8) -- (n2);
+      \draw (n8) -- (n1);
+      \draw (n9) -- (n2);
+      \draw (n9) -- (n7);
+      \draw (n10) -- (n3);
+      \draw (n10) -- (n8);
+      \draw (n11) -- (n2);
+      \draw (n11) -- (n7);
+    }
+  \end{tikzpicture}
+
+  \column{0.4\textwidth}
+  \uncover<2->{
+    $m$ = number of new edges per vertex \\
+  }
+
+  \uncover<3->{
+    $\alpha$ = preferential attachment power \\
+  }
+
+  \uncover<11->{
+    $N$ = number of nodes
+  }
   \end{columns}
-
 \end{frame}
 
-\begin{frame}{Contact networks shape transmission trees}
-  \includegraphics[width=\textwidth]{contactnet}
-
-  \begin{itemize}
-      \pause
-    \item Transmission trees in turn shape viral phylogenies.
-      \pause
-    \item \textbf{Aim:} estimate contact network parameters from viral
-      phylogenies.
-  \end{itemize}
-\end{frame}
-
-%\begin{frame}{Preferential attachment generates scale-free networks}
-%  \includegraphics[width=\textwidth]{pa-example}
-%  \begin{itemize}
-%    \item Sexual networks are often scale-free: degree distributions appear
-%      follow a power law ($\Pr(\text{degree} = d) \propto d^{-\gamma}$).
-%      \pause
-%    \item Preferential attachment models generate scale-free networks.
-%  \end{itemize}
-%\end{frame}
+\end{document}
 
@@
 \end{frame}
+
+\begin{frame}{Why fit network models?}
+  \begin{columns}
+    \column{0.33\textwidth}
+    \includegraphics[width=0.85\textwidth]{sir-trajectories-vertical}
+
+    \column{0.33\textwidth}
+    \uncover<2->{
+    \includegraphics[width=0.75\textwidth]{nettypes}
+    }
+
+    \column{0.33\textwidth}
+    \uncover<3->{
+    \begin{tikzpicture}
+      \node (pic) { \includegraphics[width=0.75\textwidth]{vaccinate} };
+      \node at (pic.east) { \includegraphics[width=1cm]{stock/syringe} };
+    \end{tikzpicture}
+    }
+  \end{columns}
+
+  \vspace{-0.5cm}
+
+  \begin{columns}
+    \column{0.33\textwidth}
+    \uncover<1->{
+    \centerline{\large predict}
+    }
+
+    \column{0.33\textwidth}
+    \uncover<2->{
+    \centerline{\large characterize}
+    }
+
+    \column{0.33\textwidth}
+    \uncover<3->{
+    \centerline{\large simulate}
+    }
+  \end{columns}
+\end{frame}
+
+\begin{frame}{Phylogenetic methods may help fit network models}
+\end{frame}
+
+\begin{frame}{Contact networks shape transmission trees}
+  \begin{center}
+    \begin{tikzpicture}
+      [every node/.style = {inner sep=0pt},
+       every path/.style = {->, >=stealth, very thick},
+       label/.style = {anchor=south, color=white, inner sep=4pt},
+       time/.style = {anchor=north west, inner sep=2pt}]
+      \def\pht{1cm}
+      \node (a) {\includegraphics[height=\pht]{stock/person}};
+      \node at (a.south) [label] {\Large a};
+      \node [below left=0.75 of a] (b) {\includegraphics[height=\pht]{stock/person}};
+      \node at (b.south) [label] {\Large b};
+      \node [below=0.75 of b] (e) {\includegraphics[height=\pht]{stock/person}};
+      \node at (e.south) [label] {\Large e};
+      \node [below right=0.75 of a] (c) {\includegraphics[height=\pht]{stock/person}};
+      \node at (c.south) [label] {\Large c};
+      \node [below=0.75 of c] (d) {\includegraphics[height=\pht]{stock/person}};
+      \node at (d.south) [label] {\Large d};
+
+      \fill [shade, opacity=0.2] (a.east) -- (c.north) -- (c.north west) -- (a.south east) -- cycle;
+      \uncover<3->{
+      \draw [blue] ($(a.east) + (0, -12pt)$) -- ($(c.north) + (-8pt, 0)$);
+      }
+      \fill [shade, opacity=0.2] (a.west) -- (b.north) -- (b.north east) -- (a.south west) -- cycle;
+      \uncover<2->{
+      \draw [red] ($(a.west) + (0, -12pt)$) -- ($(b.north) + (8pt, 0)$);
+      }
+      \fill [shade, opacity=0.2] (a.south) -- (a.south west) -- (e.north) -- (e.north east) -- cycle;
+      \fill [shade, opacity=0.2] (b.south west) -- (b.south) -- (e.north) -- (e.north west) -- cycle;
+      \fill [shade, opacity=0.2] ($(c.south) + (-8pt, 0)$) -- ++(16pt, 0) -- ($(d.north) + (8pt, 0)$) -- ++(-16pt, 0) -- cycle;
+      \fill [shade, opacity=0.2] (b.east) -- (b.south east) -- (d.west) -- (d.north west) -- cycle;
+      \uncover<4->{
+      \draw [green!80!black] (c) -- (d);
+      }
+
+      \node (abcd) [right=6cm of a.north] {\includegraphics[height=\pht]{stock/person}};
+      \node at (abcd.south) [label] {\Large a};
+      \uncover<2->{
+      \node (b) [below right=1 and 0.75 of abcd] {\includegraphics[height=\pht]{stock/person}};
+      \node at (b.south) [label] {\Large b};
+      \node (acd) [below left=1 and 0.5 of abcd] {\includegraphics[height=\pht]{stock/person}};
+      \node at (acd.south) [label] {\Large a};
+      }
+      \uncover<3->{
+      \node (cd) [below right=0.5 and 0.5 of acd] {\includegraphics[height=\pht]{stock/person}};
+      \node at (cd.south) [label] {\Large c};
+
+      \node (a) [below left=0.25 and 0.25 of acd] {\includegraphics[height=\pht]{stock/person}};
+      \node at (a.south) [label] {\Large a};
+      }
+      \uncover<4->{
+      \node (c) [below left=0.25 and 0.25 of cd] {\includegraphics[height=\pht]{stock/person}};
+      \node at (c.south) [label] {\Large c};
+      \node (d) [below right=0.4 and 0.4 of cd] {\includegraphics[height=\pht]{stock/person}};
+      \node at (d.south) [label] {\Large d};
+      }
+
+      \uncover<2->{
+      \draw [red] (abcd) -- (b);
+      \draw [-, dashed] (abcd) -- (acd);
+      }
+      \uncover<3->{
+      \draw [blue] (acd) -- (cd);
+      \draw [-, dashed] (acd) -- (a);
+      }
+      \uncover<4->{
+      \draw [-, dashed] (cd) -- (c);
+      \draw [green!80!black] (cd) -- (d);
+      }
+    \end{tikzpicture}
+
+  $\qquad$contact network \hfill transmission tree$\qquad$
+  \end{center}
+\end{frame}
+
+\begin{frame}{Transmission trees shape viral phylogenies}
+\end{frame}
+
+\begin{frame}{Can we fit network models with viral phylogenies?}
+\end{frame}
+
+\end{document}
+
+%\begin{frame}{Preferential attachment generates scale-free networks}
+%  \includegraphics[width=\textwidth]{pa-example}
+%  \begin{itemize}
+%    \item Sexual networks are often scale-free: degree distributions appear
+%      follow a power law ($\Pr(\text{degree} = d) \propto d^{-\gamma}$).
+%      \pause
+%    \item Preferential attachment models generate scale-free networks.
+%  \end{itemize}
+%\end{frame}
 

### 2016-06-25 doc/paper/abstract.tex
defences and papers and svm, oh my!

@@
 explicitly representing each contact which could possibly lead to a
-transmission. We developed a method based on kernel approximate Bayesian
-computation (kernel-ABC) for estimating structural parameters of the contact
-network underlying an observed viral phylogeny. The method combines adaptive
-sequential Monte Carlo for ABC, Gillespie simulation for propagating epidemics
-though networks, and a kernel-based tree similarity score. We used the method
-to fit the Barab\'{a}si-Albert network model to simulated transmission trees,
-and also applied it to viral phylogenies estimated from five published HIV
-sequence datasets. On simulated data, we found that the preferential attachment
-power and the number of infected nodes in the network can often be accurately
-estimated. On the other hand, the mean degree of the network, as well as the
-total number of nodes, were not estimable with kernel-ABC. We observed
+transmission. We developed a method based on approximate Bayesian
+computation (ABC) for estimating structural parameters of the contact network
+underlying an observed viral phylogeny. The method combines adaptive sequential
+Monte Carlo for ABC, Gillespie simulation for propagating epidemics though
+networks, and a kernel-based tree similarity score. We used the method to fit
+the Barab\'{a}si-Albert network model to simulated transmission trees, and also
+applied it to viral phylogenies estimated from \del{five}\add{ten} published
+HIV sequence datasets. On simulated data, we found that the preferential
+attachment power and the number of infected nodes in the network can often be
+accurately estimated. On the other hand, the mean degree of the network, as
+well as the total number of nodes, were not estimable with ABC. We observed
 substantial heterogeneity in the parameter estimates on real datasets, with
-point estimates for the preferential attachment power ranging from 0.2 to 1.1.
-These results underscore the importance of considering contact structures when
-performing phylodynamic inference. Our method offers the potential to
-quantitatively investigate the network structure underlying viral epidemics,
-using models appropriate to the context being studied.
+point estimates for the preferential attachment power ranging from 0.06 to
+1.05. These results underscore the importance of considering contact structures
+when performing phylodynamic inference. Our method offers the potential to
+quantitatively investigate the contact network structure underlying viral
+epidemics.

### 2016-06-25 doc/paper/coverletter.tex
defences and papers and svm, oh my!

--- 2016-04-20 doc/paper/mbe_submission.tex
@@
-\documentclass[12pt]{article}
-\usepackage{fullpage}
+\documentclass[12pt]{letter}
+\usepackage[T1]{fontenc}                                                         
 \usepackage{garamondx}                                                           
-\usepackage{parskip}
-\usepackage{setspace}
+\usepackage[garamondx,cmbraces]{newtxmath}
 
-% http://tex.stackexchange.com/questions/44618/dynamically-count-and-return-number-of-words-in-a-section
-\newcommand\wordcount{\immediate\write18{texcount -sub=section \jobname.tex  | grep "Section" | sed -e 's/+.*//' | sed -n \thesection p > '/tmp/count.txt'}\ignorespaces\input{/tmp/count.txt}}
+\signature{Rosemary M. McCloskey}
+\address{BC Centre for Excellence in HIV/AIDS \\ 608-1081 Burrard Street \\ Vancouver, BC, Canada V6Z 1Y6}
+\begin{document}
+\begin{letter}{ }
 
-\title{Phylodynamic inference of contact network parameters through approximate Bayesian computation}
-\author{Rosemary M. McCloskey, Richard H. Liang, and Art F.Y. Poon}
-
-\doublespacing
-\pagestyle{empty}
-\begin{document}
-\maketitle
-
-\section{Abstract}
-
-(\wordcount / 250 words)
-
-Models of the spread of disease in a population often make the simplifying
-assumption that the population is homogeneously mixed, or is divided into
-homogeneously mixed compartments. However, human populations have complex
-structures formed by social contacts, which can have a significant influence on
-the rate of epidemic spread. Contact network models capture this structure by
-explicitly representing each contact which could possibly lead to a
-transmission. We developed a method based on kernel approximate Bayesian
-computation (kernel-ABC) for estimating structural parameters of the contact
-network underlying an observed viral phylogeny. The method combines adaptive
-sequential Monte Carlo for ABC, Gillespie simulation for propagating epidemics
-though networks, and a kernel-based tree similarity score. We used the method
-to fit the Barab\'{a}si-Albert network model to simulated transmission trees,
-and also applied it to viral phylogenies estimated from five published HIV
-sequence datasets. On simulated data, we found that the preferential attachment
-power and the number of infected nodes in the network can often be accurately
-estimated. On the other hand, the mean degree of the network, as well as the
-total number of nodes, were not estimable with kernel-ABC. We observed
-substantial heterogeneity in the parameter estimates on real datasets, with
-point estimates for the preferential attachment power ranging from 0.2 to 1.1.
-These results underscore the importance of considering contact structures when
-performing phylodynamic inference. Our method offers the potential to
-quantitatively investigate the network structure underlying viral epidemics,
-using models appropriate to the context being studied.
-
-\section{Keywords}
-
-Phylogenetics, phylodynamics, approximate Bayesian computation, contact network,
-transmission tree, human immunodeficiency virus.
-
-\section{Cover Letter}
-
-Dear editors of \textit{Molecular Biology and Evolution},
+\opening{Dear editors of \textit{Virus Evolution},}
 
 On behalf of my coauthors and myself, I am submitting a manuscript entitled
-``\textit{Phylodynamic inference of contact network parameters through
-approximate Bayesian computation}'' for your consideration.
+``\textit{Reconstructing contact network parameters from viral phylogenies}''
+for your consideration.
 
@@
 
-All authors fulfill the authorship criteria, and have agreed to this
-submission. \textit{Molecular Biology and Evolution} is our first choice for
-this manuscript, which has not been submitted to any other journal.
-
-Thank you for your time,
-
-Rosemary M. McCloskey
-
-\section{Reviewers}
-
-\begin{enumerate}
-  \item David Welch
-  \item Gabriel Leventhal
-  \item Erik Volz
-  \item Tanja Stadler
-  \item Oliver Ratmann
-  \item Andrew Rambaut
-\end{enumerate}
-
-\section{Editors}
-
-\begin{enumerate}
-  \item Thomas Leitner
-  \item Beth Shapiro
-  \item Kieth Crandall
-  \item John Novembre
-\end{enumerate}
-
-\section{Funding}
-
-\begin{enumerate}
-  \item Bill \& Melinda Gates Foundation (OPP1110049)
-  \item Canadian Institutes for Health Research (HOP-111406)
-\end{enumerate}
-
-\section{Significant discoveries}
-
-(\wordcount / 150 words)
-
-Through a simulation study, we have found that certain parameters of the
-contact network underlying a viral epidemic can be quantitatively estimated
-from phylogenetic data. In particular, we investigated a preferential
-attachment network model and were able to accurately infer the attachment power
-and the number of infected nodes in the network, in most simulation scenarios
-we considered. On the other hand, the average degree and total number of nodes
-in the network had an insufficient effect on tree shape and could not be
-quantitatively estimated. When applied five real-world HIV epidemics, we found
-substantial heterogeneity of these parameters, which are known to have a
-significant impact on the speed and pattern of epidemic spread.
-
-\section{Significant methodological or theoretical advances}
-
-(\wordcount / 150 words)
-
-We have developed a phylodynamic method to quantitatively estimate the
-parameters of contact network models from phylogenetic data. Our method is
-based on approximate Bayesian computation, and combines adaptive sequential
-Monte Carlo, Gillespie simulation, and a kernel-based tree similarity score.
-Due to the use of the tree kernel, the method falls under the heading of
-"kernel-ABC", which has been previously demonstrated to be a robust and
-effective means of estimating epidemiological quantities from observed
-phylogenies. To our knowledge, no other methods have been developed to estimate
-network parameters from this type of data. Therefore, our method expands the
-domain of phylodynamic inference to include network structure, offering
-molecular epidemiologists a new perspective into ongoing epidemics.
-
-\section{Significant new tools or resources}
-
-(\wordcount / 150 words)
-
-We provide a computer program "netabc" implementing our kernel-ABC method. The
-C source code for the tool is freely available under a permissive license at
-github.com/rmcclosk/netabc. This allows users to both apply the method as-is to
-their data, and to incorporate new network models tailored to particular
-research questions. The program is multithreaded, facilitating faster analysis
-when more computing power is available. In addition, we provide two
-supplemental programs to assist in phylodynamic analyses involving contact
-networks: "nettree", which simulates a transmission tree over a contact network
-by Gillespie simulation, and "treekernel", which quickly computes the kernel
-similarity score between two trees. These tools will enable researchers to
-quickly test hypotheses about the effect of network structure on transmission
-tree shape.
-
-\section{Broad impact}
-
-(\wordcount / 150 words)
-
-Phylodynamics and molecular epidemiology are relatively new areas of study, yet
-they have undergone rapid development and have already seen public health
-applications for real epidemics. Recent advances in sequencing technology have
-allowed the collection of viral genetic data on a population scale in diverse
-geographic and demographic settings. These data, together with phylodynamic
-methods, provide an opportunity to gain insight into the real-world dynamics of
-disease spread. Our work expands the domain of phylodynamic inference to
-include structural parameters of the contact network underlying an epidemic.
-Because our method is broadly applicable to any model from which networks can
-be sampled, it will enable researchers to quantitatively investigate population
-structure and contact heterogeneity in a variety of contexts.
-
+\closing{Thank you for your time,}
+\end{letter}
 \end{document}

### 2016-06-25 doc/paper/paper.Rnw
defences and papers and svm, oh my!

@@
 \usepackage{mathptmx}
-\usepackage[backend=biber]{biblatex}
+\usepackage[backend=biber, style=numeric-comp, sorting=none]{biblatex}
 \usepackage{fullpage}
@@
 \usepackage{cleveref}
+\usepackage[normalem]{ulem}
+\usepackage{color}
+\usepackage{amsmath}
 
 \addbibresource{papers.bib}
+
+% beginning of diff stuff
+\usepackage{twoopt}
+
+\let\mtextcite\textcite
+\let\mautocite\autocite
+\let\mciteauthor\citeauthor
+\renewcommandtwoopt{\textcite}[3][][]{\mbox{\mtextcite[#1][#2]{#3}}}
+\renewcommandtwoopt{\autocite}[3][][]{\mbox{\mautocite[#1][#2]{#3}}}
+\renewcommand{\citeauthor}[1]{\mbox{\mciteauthor{#1}}}
+
+\newcommand{\add}[1]{\color{blue} \uline{#1} \color{black}}
+\newcommand{\del}[1]{\color{red} \sout{#1} \color{black}}
+% end of diff stuff
+
+\newcommand{\tablepath}{../tables}
+\graphicspath{{../figures/}}
+
 \begin{document}
 \onehalfspacing
-
-\newcommand{\tablepath}{../tables}
-\graphicspath{{../figures/}}
 
@@
 type of data to collect, requiring extensive epidemiological
-investigation~\autocite{welch2011statistical}.
+investigation~\autocite{welch2011statistical, eames2015six}.
 
@@
 Survey-based studies of sexual networks~\autocite{colgate1989risk,
-liljeros2001web, schneeberger2004scale} have found that these networks tend to
-have a degree distribution which follows a power law \autocite[although there
-has been some disagreement, see][]{handcock2004likelihood}. That is, the number
-of nodes of degree $k$ is proportional to $k^{-\gamma}$ for some constant
-$\gamma$. These networks are also referred to as
+liljeros2001web, schneeberger2004scale, latora2006network, rothenberg2007large,
+clemenccon2015statistical} have found that these networks tend to have a degree
+distribution which follows a power law \autocite[although there has been some
+disagreement, see][]{handcock2004likelihood}. That is, the number of nodes of
+degree $k$ is proportional to $k^{-\gamma}$ for some constant $\gamma$. These
+networks are also referred to as
 ``scale-free''~\autocite{barabasi1999emergence}. One process by which
@@
 some non-negative constant $\alpha$, and we use this extension in this work.
+\add{When $\alpha \neq 1$, the degree distribution is no longer a power law:
+for $\alpha < 1$, the distribution is a stretched exponential, while for
+$\alpha > 1$, it is a ``gelation'' type distribution where one or a few hub
+nodes are connected to nearly every other node in the graph.}
 
@@
 function, $\hat{D}$ is a simulated dataset according to $\theta$, and
-$\varepsilon$ is a small tolerance~\autocite{sunnaaker2013approximate}. In the
-specific case when $\rho$ is a kernel function, the approach is known as
-kernel-ABC~\autocite{nakagome2013kernel, poon2015phylodynamic}.
-
-Here, we develop a method using kernel-ABC to estimate the parameters of
-contact network models from observed phylogenetic data. The distance function
-we use is the tree kernel developed by \textcite{poon2013mapping}, which computes
-a weighted dot product of the trees' representations in the space of all
-possible subset trees. We apply the method to investigate the parameters of the
-BA network model on a variety of simulated and real datasets. Our results show
+$\varepsilon$ is a small tolerance~\autocite{sunnaaker2013approximate}. \del{In
+the specific case when $\rho$ is a kernel function, the approach is known as
+ABC~\autocite{nakagome2013kernel, poon2015phylodynamic}.} \add{Our
+group~\autocite{poon2015phylodynamic} and others~\autocite{park2015k2} have
+demonstrated that taking $\rho$ to be kernel function TODO}
+
+Here, we develop a method using ABC to estimate the parameters of contact
+network models from observed phylogenetic data. The distance function we use is
+the tree kernel developed by \textcite{poon2013mapping}, which computes a
+weighted dot product of the trees' representations in the space of all possible
+subset trees. We apply the method to investigate the parameters of the BA
+network model on a variety of simulated and real datasets. Our results show
 that some network parameters can be inferred with reasonable accuracy, while
-others have a minimal detectable impact on tree shape and therefore cannot be
-estimated accurately. We also find that these parameters can vary considerably
-between real epidemics from different settings.
+others \del{have a minimal detectable impact on tree shape and therefore cannot
+be estimated accurately}\add{are weakly- or non-identifiable with ABC}. We also
+find that these parameters can vary considerably between real epidemics from
+different settings. \add{TODO: something about sub-linear PA and IDU}
 
@@
 \subsection*{\textit{Netabc}: phylogenetic inference of contact network
-parameters with kernel-ABC}
-
-We have developed a kernel-ABC-based method to perform statistical inference of
+parameters with ABC}
+
+We have developed an ABC-based method to perform statistical inference of
 contact network parameters from a transmission tree estimated from an observed
@@
 is assumed to approximate the desired posterior~\autocite{del2012adaptive,
-sunnaaker2013approximate}. A computer program implementing our method is freely
-available at \url{https://github.com/rmcclosk/netabc} (last accessed \today).
+sunnaaker2013approximate}.
+
+\add{In the original formulation of ABC-SMC~\autocite{sisson2007sequential,
+beaumont2009adaptive}, the user is required to specify a decreasing sequence of
+tolerances $\lbrace\varepsilon_i\rbrace$. At iteration $i$, particles with no
+associated simulated datasets within distance $\varepsilon_i$ of the observed
+data are removed from the population. In the adaptive version of
+\textcite{del2012adaptive}, the sequence of tolerances is determined
+automatically by fixing the decay rate of the population's expected sample size
+(ESS) to a user-defined value. \citeauthor{del2012adaptive} call this value
+$\alpha$, but we will refer to it here as $\alpha_{\text{ESS}}$ to avoid
+confusion with the preferential attachment power parameter of the BA model. A
+computer program implementing our method is freely available at
+\url{https://github.com/rmcclosk/netabc} (last accessed \today).}
 
@@
 
-\subsection*{Kernel classifiers}
-
-We used the phylogenetic kernel developed by \textcite{poon2013mapping} to test
-whether the parameters of the BA model had an effect on tree shape. 100
-networks were simulated under each of three different values of $\alpha$: 0.5,
-1.0, and 1.5 (300 networks total). The other parameters were fixed to the
-following values: $N$ = 5000, $I$ = 1000, and $m$ = 2. A transmission tree with
-500 tips was simulated over each network (300 transmission trees total). The
-300 trees were compared pairwise with the tree kernel to form a $300 \times
-300$ kernel matrix. The kernel meta-parameters $\lambda$ (the ``decay
-factor''), and $\sigma$ (the ``radial basis function
-variance'')~\autocite[see][]{poon2013mapping}, were set to 0.3 and 4 respectively.
-We constructed a kSVR classifier for $\alpha$ using the \textit{kernlab}
-package~\autocite{zeileis2004kernlab}, and evaluated its accuracy with 1000
-two-fold cross-validations.
+\subsection*{\del{Kernel classifiers}\add{Classifiers for BA model parameters from tree shapes}}
+
+\add{We considered four parameters related to the BA model, denoted $N$, $m$,
+$alpha$, and $I$. The first three of these parameterize the network structure,
+while $I$ is related to the simulation of transmission trees over the network.
+However, we will refer to all four as BA parameters. $N$ denotes the total
+number of nodes in the network, or equivalently, susceptible individuals in the
+population. $m$ is the number of new undirected edges added for each new
+vertex, or equivalently one-half of the average degree. $\alpha$ is the power
+of preferential attachment -- new nodes are attached to existing nodes of
+degree $d$ with probability proportional to $d^{\alpha} + 1$. Finally, $I$ is
+the number of infected individuals at the time when sampling occurs. The
+$\alpha$ parameter is unitless, while $m$ has units of edges or connections per
+vertex, and $N$ and $I$ both have units of nodes or individuals.}
+
+\add{Before proceeding with a full validation of \textit{netabc} on simulated
+data, we undertook two experiments designed to assess the identifiability of
+the BA parameters. These experiments only investigated one parameter of the
+BA model at a time while holding all others fixed, a strategy commonly used
+when performing sensitivity analyses of mathematical models. This allowed us to
+perform a fast preliminary analysis without dealing with the ``curse of
+dimensionality'' of the full parameter space. We simulated trees under three
+different values of each parameter, and asked how well we could tell the
+different trees apart. The better we are able to distinguish the trees, the
+more identifiability we might expect for the corresponding parameter when we
+attempt to estimate it with ABC.}
+
+\add{This experiment also had the secondary purpose of validating our choice of
+the tree kernel as a distance measure in ABC. To tell the trees apart, we used
+a classifier based on the tree kernel, but we also tested two other tree shape
+statistics. Sackin's index~\autocite{shao1990tree} is a measure of tree
+imbalance which not take branch lengths into account, considering only the
+topology. The normalized lineages-through-time
+\autocite[nLTT,][]{janzen2015approximate} compares two trees based on normalized
+distributions of their branching times, and does not explicitly consider the
+topology. In addition, the tree kernel can be tuned by adjusting the values of
+the meta-parameters $\lambda$ and $\sigma$ (the ``decay factor'' and ``radial
+basis function variance'', see \textcite{poon2013mapping}). The results of this
+experiment were used to select values for these meta-parameters to carry
+forward based on their accuracy in distinguishing the different trees.}
+
+\del{We used the phylogenetic kernel developed by \textcite{poon2013mapping}}
+To test whether the parameters of the BA model had a \add{measurable} effect on
+tree shape, 100 networks were simulated under each of three different values of
+$\alpha$: 0.5, 1.0, and 1.5 (300 networks total). The other parameters were
+fixed to the following values: $N = 5000$, $I = 1000$, and $m = 2$. A
+transmission tree with 500 tips was simulated over each network (300
+transmission trees total). The 300 trees were compared pairwise with the tree
+kernel to form a $300 \times 300$ kernel matrix. The kernel meta-parameters
+$\lambda$ \del{(the ``decay factor''),} and $\sigma$ \del{(the ``radial basis
+function variance'')~\autocite[see][]{poon2013mapping},} were set to 0.3 and 4
+respectively. \add{We also computed a $300 \times 300$ matrix of pairwise nLTT
+values, and a $1 \times 300$ vector of Sackin's index values.} We constructed 
+\add{three classifiers for $\alpha$:} a kernel support vector regression (kSVR)
+\add{from the kernel matrix} with the \textit{kernlab}
+package~\autocite{zeileis2004kernlab}, an ordinary SVR \add{from the nLTT
+matrix} with the e1071 package~\autocite{meyer2015e1071}, and a linear
+regression from the Sackin's index values. The accuracy of each classifier was 
+evaluated with 1000 two-fold cross validations.
 
@@
 tested were fixed at the values $N$ = 5000, $I$ = 1000, $m$ = 2, and $\alpha$ =
-1. Thus, we performed a total of four kSVR cross-validations, one for each of
-the BA model parameters $\alpha$, $I$, $m$, and $N$. We repeated these four
-cross-validations with different values of $\lambda$ (0.2, 0.3, and 0.4) and
-$\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$), as well as on trees with
-differing numbers of tips (100, 500, and 1000) and in epidemics of differing
-size (500, 1000, and 2000). The combination of the number of sampled
-individuals (\textit{i.e.} the number of tips) and the epidemic size
-(\textit{i.e.} $I$) will be referred to as an ``epidemic scenario''. When
-evaluating the classifier for $I$, we did not consider trees with 1000 tips,
-because one of the tested $I$ values was 500, and the number of tips cannot be
-larger than $I$.
-
-For each of the four parameters, we also tested a linear regression against
+1. Thus, we performed a total of four cross-validations \add{for each
+classifier}, one for each of the BA model parameters $\alpha$, $I$, $m$, and
+$N$. We repeated these four cross-validations with different values of
+$\lambda$ (0.2, 0.3, and 0.4) and $\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$),
+as well as on trees with differing numbers of tips (100, 500, and 1000).
+\add{For the structural parameters $\alpha$, $m$, and $N$, the experiments were
+repeated with three different values of $I$ (500, 1000, and 2000).} \del{and in
+epidemics of differing size (500, 1000, and 2000).} The combination of the
+number of sampled individuals (\textit{i.e.} the number of tips) and the
+epidemic size (\textit{i.e.} $I$) will be referred to as an ``epidemic
+scenario''. When evaluating the classifier for $I$, we did not consider trees
+with 1000 tips, because one of the tested $I$ values was 500, and the number of
+tips cannot be larger than $I$.
+
+\del{For each of the four parameters, we also tested a linear regression against
 Sackin's index~\autocite{shao1990tree} and an ordinary SVR based on the normalized
-lineages-through-time (nLTT) statistic~\autocite{janzen2015approximate}.
+lineages-through-time (nLTT) statistic~\autocite{janzen2015approximate}.}
 
@@
 
-We simulated three transmission trees, each with 500 tips, under every element
-of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
-\{1000, 2000\}, $m$ = \{2, 3, 4\}, and $\alpha$ = \{0.0, 0.5, 1,
-1.5\}. This produced a total of 24 parameter combinations $\times$ three trees
-per combination = 72 trees total. The adaptive ABC algorithm was applied to
-each tree with these priors: $m \sim$ DiscreteUniform(1, 5), $\alpha \sim$
-Uniform(0, 2), and $(N, I)$ jointly uniform on the region \{$500 \leq N \leq
-15000$, $500 \leq I \leq 5000$, $I \leq N$\}. Proposals for $\alpha$, $N$, and
-$I$ were Gaussian, while proposals for $m$ were Poisson. Following
-\textcite{del2012adaptive} and \textcite{beaumont2009adaptive}, the variance of
-all proposals was equal to the empirical variance of the particles.
-
-The algorithm was run with 1000 particles, 5 simulated datasets per particle,
-and the ``quality'' parameter controlling the decay rate of the tolerance
-$\varepsilon$ set to 0.95. We used the same stopping criterion as
-\citeauthor{del2012adaptive}, namely when the MCMC acceptance rate dropped
-below 1.5\%. Point estimates for the parameters were obtained by taking the
-highest point of an estimated kernel density on the final set of particles,
-calculated using the \textit{density} function with the default parameters in
-\textit{R}. Highest posterior density (HPD) intervals were calculated with the
+\add{We tested \textit{netabc} by jointly estimating the four parameters of the
+BA model. We used the standard validation approach of simulating transmission
+trees under the model with known parameter values and attempting to recover
+those values with \textit{netabc}. The algorithm was not informed of any of
+the true parameter values for the main set of simulations.} We simulated three
+transmission trees, each with 500 tips, under every element of the Cartesian
+product of these parameter values: $N$ = 5000, $I$ = \{1000, 2000\}, $m$ = \{2,
+3, 4\}, and $\alpha$ = \{0.0, 0.5, 1, 1.5\}. This produced a total of 24
+parameter combinations $\times$ three trees per combination = 72 trees total.
+The adaptive ABC algorithm was applied to each tree with these priors: $m \sim$
+DiscreteUniform(1, 5), $\alpha \sim$ Uniform(0, 2), and $(N, I)$ jointly
+uniform on the region \{$500 \leq N \leq 15000$, $500 \leq I \leq 5000$, $I
+\leq N$\}. Proposals for $\alpha$, $N$, and $I$ were Gaussian, while proposals
+for $m$ were Poisson. Following \textcite{del2012adaptive} and
+\textcite{beaumont2009adaptive}, the variance of all proposals was equal to the
+empirical variance of the particles.
+
+The SMC settings used were 1000 particles, 5 simulated datasets per particle,
+and \del{the ``quality'' parameter controlling the decay rate of the tolerance
+$\varepsilon$ set to}\add{$\alpha_{\text{ESS}} =  0.95$}. We used the same
+stopping criterion as \citeauthor{del2012adaptive}, namely when the MCMC
+acceptance rate dropped below 1.5\%. \del{Point estimates for the parameters
+were obtained by taking the highest point of an estimated kernel density on the
+final set of particles, calculated using the \textit{density} function with the
+default parameters in \textit{R}.} \add{Approximate posterior means for the
+parameters were obtained by taking the weighted average of the final set of
+particles.} Highest posterior density (HPD) intervals were calculated with the
 \textit{HPDinterval} function from the \textit{R} package
@@
 
-Two further analyses were performed to address potential sources of error. To
+\add{To evaluate the effects of the true parameter values on the accuracy of
+the posterior mean estimates, we analyzed the $\alpha$ and $I$ parameters
+individually using generalized linear models (GLMs) The response variable was
+the error of the point estimate, and the predictor variables were the true
+values of $\alpha$, $I$, and $m$. We did not test for differences across true
+values of $N$, because $N$ was not varied in these simulations. The
+distribution family and link function for the GLMs were Gaussian and inverse,
+respectively, chosen by examination of residual plots and Akaike information
+criteria (AIC). The $p$-values of the estimated GLM coefficients were corrected
+using Holm-Bonferroni correction~\autocite{holm1979simple} with $n =
+6$ (two GLMs with three predictors each). Because there was clearly little to
+no identifiability of $N$ and $m$ with ABC (see results in next section), we
+did not construct GLMs for those parameters.}
+
+Two further simulations were performed to address \del{potential sources of
+error} \add{the possible impact of two types of model misspecification}. To
 evaluate the effect of model misspecification in the case of heterogeneity
@@
 that node's network peers had already been sampled. The parameters of this
-network were $N$ = 5000, $I$ = 2000, $m$ = 2, and $\alpha$ = 0.5.
+network were $N= 5000$, $I = 2000$, $m = 2$, and $\alpha = 0.5$.
+
+\add{Despite the fact that the parameter values used to generate the simulated
+transmission trees were known, the true posterior distributions of the BA
+parameters were unknown. Therefore, any apparent errors or biases in the
+estimates could be due to either poor performance of our method, or to real
+features of the posterior distribution. Two retrospective experiments were
+performed to disambiguate some of the observed errors. To assess the impact of
+the SMC settings on \textit{netabc}'s accuracy, we ran \textit{netabc}
+twice on the same simulated transmission tree. For the first run, the SMC
+settings were the same as in the other simulations: 1000 particles, 5 simulated
+transmission trees per particle, and $\alpha_{\text{ESS}} = 0.95$. The second
+run was performed with 2000 particles, 10 simulated transmission trees per
+particle, and $\alpha_{\text{ESS}} = 0.99$. To investigate the extent to which
+errors in the estimated BA parameters were due to true features of the
+posterior, rather than an inaccurate ABC approximation, we performed marginal
+estimation for one set of parameter values. Each combination of 1, 2, or 3
+model parameters (14 combinations total) was fixed to their known values, and
+the remaining parameters were estimated with \textit{netabc}. The parameter
+values were $\alpha = 0.0$, $m = 2$, $I = 2000$, and $N = 5000$.}
 
@@
 
-We applied our kernel-ABC method to several published HIV datasets. Because the
-BA model generates networks with a single connected component, we
-specifically searched for datasets which originated from existing clusters,
-either phylogenetically or geographically defined. Characteristics of the
-datasets we investigated are given in \cref{tab:data}.
+We applied our ABC method to ten published HIV datasets. Because the BA model
+generates networks with a single connected component, we specifically searched
+for datasets which originated from existing clusters, either phylogenetically
+or geographically defined. Characteristics of the datasets we investigated are
+given in \cref{tab:data}. \add{For clarity, we will refer to each dataset by
+its risk group and location of origin in the text. For example, the
+\textcite{zetterberg2004two} data will be referred to as IDU/Estonia.}
 
@@
   \caption{
-    Characteristics of published datasets investigated with kernel-ABC.
+    Characteristics of published datasets investigated with ABC.
     Acronyms: MSM, men who have sex with men; IDU, injection drug users; HET,
-    heterosexual. The \textcite{novitsky2013phylogenetic, novitsky2014impact}
-    data was sampled from a primarily heterosexual risk environment, but did
-    not explicitly exclude other risk groups.
+    heterosexual. The HET data were sampled from a primarily heterosexual risk
+    environment, but did not explicitly exclude other risk factors. The number
+    of sequences column indicates how many sequences were included in our
+    analysis; there may have been additional sequences linked to the study
+    which we excluded for various reasons (see methods).
   }
@@
 
-We downloaded all sequences associated with each study from GenBank. For the
-\textcite{novitsky2014impact} data, each sequence was aligned pairwise to the HXB2
-reference sequence (Genbank accession number K03455) and the hypervariable
-regions were clipped out with \textit{BioPython} version
-1.66+~\autocite{cock2009biopython}. Sequences were multiply aligned using
-\textit{MUSCLE} version 3.8.31 \autocite{edgar2004muscle}, and alignments were
-manually inspected with \textit{Seaview} version 4.4.2
-\autocite{gouy2010seaview}. Phylogenies were constructed from the nucleotide
-alignments by approximate maximum likelihood using \textit{FastTree2} version
-2.1.7 \autocite{price2010fasttree} with the generalized time-reversible model.
-Transmission trees were estimated by rooting and time-scaling the phylogenies
-by root-to-tip regression, using a modified version of Path-O-Gen (distributed
-as part of BEAST~\autocite{drummond2007beast}) as described
+We downloaded all sequences associated with each published study from GenBank.
+\add{For the IDU/Romania data, only sequences from injection drug users (IDU,
+whose sequence identifiers included the letters ``DU'') were included in the
+analysis. \textcite{kao2011surveillance} (MSM/Taiwan) found a strong
+association in their study population between subtype and risk group - subtype
+B was most often associated with men who have sex with men (MSM), whereas IDU
+were usually infected with a circulating recombinant form. Since there were
+many more subtype B sequences in their data than sequences of other subtypes,
+we restricted our analysis to the subtype B sequences and labelled this dataset
+as MSM. Three datasets (IDU/Estonia, HET/Uganda, and HET/Malawi) included both
+\textit{env} and \textit{gag} sequences. Each gene was analyzed separately to
+assess the robustness of \textit{netabc} to the particular HIV gene sequence
+used to estimate a transmission tree.}
+
+\del{For the \textcite{novitsky2014impact} data,} Each \textit{env} sequence
+was aligned pairwise to the HXB2 reference sequence (GenBank accession number
+K03455), and the hypervariable regions were clipped out with
+\textit{BioPython} version 1.66+~\autocite{cock2009biopython}. Sequences were
+multiply aligned using \textit{MUSCLE} version 3.8.31
+\autocite{edgar2004muscle}, and alignments were manually inspected with
+\textit{Seaview} version 4.4.2 \autocite{gouy2010seaview}. Phylogenies were
+constructed from the nucleotide alignments by approximate maximum likelihood
+using \textit{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
+generalized time-reversible (GTR) model~\autocite{tavare1986some}. Transmission
+trees were estimated by rooting and time-scaling the phylogenies by root-to-tip
+regression, using a modified version of Path-O-Gen (distributed as part of
+BEAST~\autocite{drummond2007beast}) as described
 previously~\autocite{poon2015phylodynamic}.
 
-Two of the datasets \autocite{li2015hiv,novitsky2014impact} were initially much
-larger than the others, containing 1265 and 1299 sequences respectively. To
+\del{Two} \add{Four} of the datasets \del{\autocite{li2015hiv,novitsky2014impact}}
+(MSM/Shanghai, HET/Botswana, HET/Uganda, and MSM/USA)
+were initially much larger than the others, containing 1265, 1299,
+\add{1026/915 (\textit{env}/\textit{gag}), and 648} sequences respectively. To
 ensure that the analyses were comparable, we reduced these to a number of
-sequences similar to the smaller datasets. For the \textcite{li2015hiv} data,
-we detected a cluster of size 280 using a patristic distance cutoff of 0.02 as
+sequences similar to the smaller datasets. For the MSM/Shanghai data, we
+detected a cluster of size 280 using a patristic distance cutoff of 0.02 as
 described previously~\autocite{poon2015impact}. Only sequences within this
-cluster were carried forward. For the \textcite{novitsky2014impact} data, no
-large clusters were detected using the same cutoff, so we analysed a subtree of
-size 180 chosen arbitrarily.
+cluster were carried forward. For the HET/Uganda, HET/Botswana, and MSM/USA
+data, no large clusters were detected using the same cutoff, so we analysed
+\del{a subtree} subsets of sizes 255, 180, and 180 respectively. \add{The
+subset of the HET/Uganda data was chosen by eye such that the individuals were
+monopolistic in both the \textit{gag} and \textit{env} trees. The other subsets
+were arbitrarily chosen subtrees from phylogenies of the complete datasets.}
 
@@
 and one with the prior $m \sim$ DiscreteUniform(2, 5). The other parameters to
-the SMC algorithm were the same as used for the simulation experiments.
+the SMC algorithm were the same as used for the simulation experiments,
+\add{except that we used 10000 particles instead of 1000 to increase the
+accuracy of the estimated posterior. This was computationally feasible due to
+the small number of runs required for this analysis.}
 
@@
 
-\subsection*{Kernel classifiers}
+\subsection*{\del{Kernel classifiers}\add{Classifiers for BA model parameters from tree shapes}}
 
@@
 
-We investigated the parameters of the BA network
-model~\autocite{barabasi1999emergence}. In addition to $m$ and $\alpha$ (see
-Introduction), we considered $N$, which denotes the total number of nodes in
-the network, and $I$, which is the number of infected nodes at which to stop
-the simulation and sample the transmission tree. To examine the effect of these
-parameters on tree shape, we simulated transmission trees under different
+We investigated the \add{identifiability of four} parameters of the BA network
+model~\autocite{barabasi1999emergence}: the number of nodes $N$, the
+preferential attachment power $\alpha$, the number of edges added per vertex
+$m$, and the number of infected nodes $I$. \del{In addition to $m$ and $\alpha$
+(see Introduction), we considered $N$, which denotes the total number of nodes
+in the network, and $I$, which is the number of infected nodes at which to stop
+the simulation and sample the transmission tree.} To examine the effect of
+these parameters on tree shape, we simulated transmission trees under different
 parameter values, calculated pairwise tree kernel scores between them, and
@@
 the normalized lineages-through-time (nLTT)
-statistic~\autocite{janzen2015approximate}. Accuracy of the kSVRs varied based
-on the parameter being tested (\cref{fig:rsquared}, left). Classifiers based on
-two other tree statistics, the nLTT and Sackin's index, generally exhibited
-worse performance than the tree kernel, although the magnitude of the disparity
-varied between the parameters (\cref{fig:rsquared}, centre and right). The
-results were largely robust to variations in the tree kernel meta-parameters
-$\lambda$ and $\sigma$ (\cref{fig:alphacrossv,fig:mcrossv,fig:Ncrossv,fig:Icrossv}).
-
-When classifying $\alpha$, the kernel-SVR classifier had an average $R^2$ of 
+statistic~\autocite{janzen2015approximate}. The accuracy of each classifier
+\add{on all BA parameters is shown in} \del{varied based on the parameter
+being tested} \cref{fig:rsquared}. Classifiers based on \del{two other tree
+statistics,} the nLTT and Sackin's index generally exhibited worse performance
+than the tree kernel, although the magnitude of the disparity varied between
+the parameters (\cref{fig:rsquared}, centre and right). The results were
+largely robust to variations in the tree kernel meta-parameters $\lambda$ and
+$\sigma$, \add{although accuracy varied between different epidemic and sampling
+scenarios} (\cref{fig:alphacrossv,fig:mcrossv,fig:Ncrossv,fig:Icrossv}).
+\add{For all parameters except $m$, the absolute number of tips in the tree had
+a much greater impact on accuracy than the proportion of infected individuals
+these tips represented. However, for $m$, both the number and proportion of
+sampled tips had a strong impact or the accuracy of the kSVR
+(\cref{fig:mcrossv}).}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
+    \caption[
+        Cross-validation accuracy of kernel-SVR, nLTT-based SVR, and Sackin's
+        index regression classifiers for BA model parameters.
+    ]{
+        Cross-validation accuracy of kernel-SVR classifier (left), SVR
+        classifier using nLTT (centre), and linear regression using
+        Sackin's index (right) for BA model parameters. Kernel meta-parameters
+        were set to $\lambda = 0.3$ and $\sigma = 4$. Each point was calculated
+        based on 300 simulated transmission trees over networks with three
+        different values of the parameter being tested, assuming perfect
+        knowledge of the other parameters. Vertical lines are empirical 95\%
+        confidence intervals based on 1000 two-fold cross-validations. \add{The
+        classifiers for $I$ were not evaluated with 1000-tip trees, because one
+        of the tested $I$ values was 500, and it is not possible to sample a
+        tree of size 1000 from 500 infected individuals.}
+    }
+    \label{fig:rsquared}
+\end{figure}
+
+\add{The accuracy of each classifier on all BA parameters is shown in
+\cref{fig:rsquared}. Classifiers based on the nLTT and Sackin's index generally
+exhibited worse performance than the tree kernel, although the magnitude of the
+disparity varied between the parameters (\cref{fig:rsquared}, centre and
+right). The results were largely robust to variations in the tree kernel
+meta-parameters $\lambda$ and $\sigma$, although accuracy varied between
+different epidemic and sampling scenarios
+(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}). For all
+parameters except $m$, the absolute number of tips in the tree had a much
+greater impact on accuracy than the proportion of infected individuals these
+tips represented. However, for $m$, both the number and proportion of sampled
+tips had a strong impact or the accuracy of the kSVR (\cref{fig:mcrossv}).}
+
+\end{document}
+
+The \gls{kSVR} classifier for \gls{alpha} had an average $R^2$ of 
     \Sexpr{kernel.alpha[,round(mean(rsquared), 2)]},
 compared to 
-    \Sexpr{nltt.alpha[,round(mean(rsquared), 2)]}
+    %\S%expr{nltt.alpha[,round(mean(rsquared), 2)]}
 for the nLTT-based SVR, and
-    \Sexpr{sackin.alpha[,round(mean(rsquared), 2)]}
+    %\S%expr{sackin.alpha[,round(mean(rsquared), 2)]}
 for the linear regression against Sackin's index. There was little variation
@@
 $R^2$ values of 
-  \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for kSVR,
-  \Sexpr{nltt.m[,round(mean(rsquared), 2)]} for the nLTT, and
-  \Sexpr{sackin.m[,round(mean(rsquared), 2)]}
+  %\S%expr{kernel.m[,round(mean(rsquared), 2)]} for kSVR,
+  %\S%expr{nltt.m[,round(mean(rsquared), 2)]} for the nLTT, and
+  %\S%expr{sackin.m[,round(mean(rsquared), 2)]}
 for Sackin's index. Again, there was little variation in accuracy between
@@
 tips in the tree. For 100-tip trees, the average $R^2$ values were
-  \Sexpr{kernel.I[ntip == 100, round(mean(rsquared), 2)]},
-  \Sexpr{nltt.I[ntip == 100, round(mean(rsquared), 2)]}, and
-  \Sexpr{sackin.I[ntip == 100, round(mean(rsquared), 2)]}
+  %\S%expr{kernel.I[ntip == 100, round(mean(rsquared), 2)]},
+  %\S%expr{nltt.I[ntip == 100, round(mean(rsquared), 2)]}, and
+  %\S%expr{sackin.I[ntip == 100, round(mean(rsquared), 2)]}
 for the tree kernel, nLTT, and Sackin's index respectively. For 500-tip
 trees, the values increased to
-  \Sexpr{kernel.I[ntip == 500, round(mean(rsquared), 2)]},
-  \Sexpr{nltt.I[ntip == 500, round(mean(rsquared), 2)]}, and
-  \Sexpr{sackin.I[ntip == 500, round(mean(rsquared), 2)]}.
+  %\S%expr{kernel.I[ntip == 500, round(mean(rsquared), 2)]},
+  %\S%expr{nltt.I[ntip == 500, round(mean(rsquared), 2)]}, and
+  %\S%expr{sackin.I[ntip == 500, round(mean(rsquared), 2)]}.
 Finally, the performance of classifiers for $N$ depended heavily on the
 epidemic scenario. The $R^2$ of the kSVR classifier ranged from
-  \Sexpr{kernel.N[,round(min(rsquared), 2)]}
+  %\S%expr{kernel.N[,round(min(rsquared), 2)]}
 for the smallest epidemic and smallest sample size, to
-  \Sexpr{kernel.N[,round(max(rsquared), 2)]}
+  %\S%expr{kernel.N[,round(max(rsquared), 2)]}
 for the largest. Likewise, $R^2$ for the nLTT-based SVR ranged from 
-  \Sexpr{nltt.N[,round(min(rsquared), 2)]}
+  %\S%expr{nltt.N[,round(min(rsquared), 2)]}
 to
-  \Sexpr{nltt.N[,round(max(rsquared), 2)]}.
+  %\S%expr{nltt.N[,round(max(rsquared), 2)]}.
 Sackin's index did not accurately classify $N$ in any scenario, with an average
 $R^2$ of
-  \Sexpr{sackin.N[,round(mean(rsquared), 2)]}
+  %\S%expr{sackin.N[,round(mean(rsquared), 2)]}
 and little variation between scenarios.
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth]{kernel-rsquared}
-  \vspace{6pt}
-  \caption{
-      Cross-validation accuracy of kernel-SVR classifier (left), SVR classifier
-      using nLTT (centre), and linear regression using Sackin's index
-      (right) for BA model parameters. Kernel meta-parameters were set to
-      $\lambda = 0.3$ and $\sigma = 4$. Each point was calculated based on 300
-      simulated transmission trees over networks with three different values of
-      the parameter being tested. Vertical lines are empirical 95\% confidence
-      intervals based on 1000 two-fold cross-validations.
-  }
-  \label{fig:rsquared}
-\end{figure}
-
 \subsection*{ABC simulations}
 
-<<point_est, include=FALSE>>=
+<<point_est, include=FALSE, eval=FALSE>>=
     f <- Sys.glob("../../simulations/abc-pa-free-m/point-estimate/*")
@@
 \Cref{fig:abcpt} shows maximum \textit{a posteriori} (MAP) point
-estimates of the BA model parameters obtained with kernel-ABC on simulated
+estimates of the BA model parameters obtained with ABC on simulated
 data. The estimates shown correspond only to the simulations where the $m$
@@
 
-The accuracy of the parameter estimates obtained with kernel-ABC
+The accuracy of the parameter estimates obtained with ABC
 paralleled the results from the kSVR classifier. Of the four parameters,
@@
 median [IQR] absolute error of 
-    \Sexpr{d[,round(median(alpha_error), 2)]} 
-    [\Sexpr{d[,round(quantile(alpha_error, 0.25), 2)]} - 
-    \Sexpr{d[,round(quantile(alpha_error, 0.75), 2)]}].
+    %\S%expr{d[,round(median(alpha_error), 2)]} 
+    [%\S%expr{d[,round(quantile(alpha_error, 0.25), 2)]} - 
+    %\S%expr{d[,round(quantile(alpha_error, 0.75), 2)]}].
 The errors when the true value of $\alpha$ was zero were significantly greater
 than those for the other values 
-    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(zero.alpha.test$p.value, 4))}$).
+    (Wilcoxon rank-sum test, $p$ = %\S%expr{latexSN(round(zero.alpha.test$p.value, 4))}$).
 Errors in estimating $\alpha$ also varied with the true value of $m$ just at
 the threshold of statistical significance
-    ($p \Sexpr{pp(alpha.av["factor(true_m)", "Pr(>F)"])}$),
+    ($p$ %\S%expr{pp(alpha.av["factor(true_m)", "Pr(>F)"])}$),
 but did not vary across the true values of $N$ or $I$ (both one-way ANOVA).
 Estimates for $I$ were relatively accurate, with point estimate errors of
-    \Sexpr{d[,round(median(I_error))]} 
-    [\Sexpr{d[,round(quantile(I_error, 0.25))]} - 
-    \Sexpr{d[,round(quantile(I_error, 0.75))]}] individuals.
+    %\S%expr{d[,round(median(I_error))]} 
+    [%\S%expr{d[,round(quantile(I_error, 0.25))]} - 
+    %\S%expr{d[,round(quantile(I_error, 0.75))]}] individuals.
 These errors were significantly higher when the true value of $\alpha$ was
 at least 1
-    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(alpha.I.test$p.value, 4))}$)
+    (Wilcoxon rank-sum test, $p$ = %\S%expr{latexSN(round(alpha.I.test$p.value, 4))}$)
 and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
@@
 The $m$ parameter was estimated correctly in only
-    \Sexpr{as.integer(m.tbl[1] * 100)} \%
+    %\S%expr{as.integer(m.tbl[1] * 100)} \%
 of simulations, barely better than random guessing. The true values of the
@@
 over-estimated by about a factor of two
-    (error \Sexpr{d[,round(median(N_error))]} 
-    [\Sexpr{d[,round(quantile(N_error, 0.25))]} - 
-     \Sexpr{d[,round(quantile(N_error, 0.75))]}] individuals).
+    (error %\S%expr{d[,round(median(N_error))]} 
+    [%\S%expr{d[,round(quantile(N_error, 0.25))]} - 
+     %\S%expr{d[,round(quantile(N_error, 0.75))]}] individuals).
 No parameters influenced the accuracy of the $N$ estimates (all one-way ANOVA).
@@
   \caption{
-    Point estimates of BA model parameters obtained by running kernel-ABC
+    Point estimates of BA model parameters obtained by running ABC
     on simulated phylogenies without training, for simulations with $m = 2$.
@@
       posterior density (HPD) interval widths for BA model parameter estimates
-      obtained with kernel-ABC. Three transmission trees were simulated under
+      obtained with ABC. Three transmission trees were simulated under
       each combination of the listed parameter values, and the parameters were
-      estimated with kernel-ABC without training.
+      estimated with ABC without training.
   }
@@
 simulation with each studied parameter combination can be found in
-\cref{\Sexpr{lab}}. HPD intervals around $\alpha$ and $I$ were often narrow
+\cref{%\S%expr{lab}}. HPD intervals around $\alpha$ and $I$ were often narrow
 relative to the region of nonzero prior density, whereas the intervals for $m$
@@
     Marginal posterior distributions of BA model parameters estimated
-    with kernel-ABC for a single simulated transmission tree. Dotted
+    with ABC for a single simulated transmission tree. Dotted
     lines and shaded polygon indicate true values.
@@
 
-<<mixed_peerdriven, include=FALSE>>=
+<<mixed_peerdriven, include=FALSE, eval=FALSE>>=
   library(coda)
@@
 $\alpha$, 
-  \Sexpr{round(d["alpha", map], 2)} 
-  [\Sexpr{round(d["alpha", hpd.min], 2)} -
-   \Sexpr{round(d["alpha", hpd.max], 2)}];
+  %\S%expr{round(d["alpha", map], 2)} 
+  [%\S%expr{round(d["alpha", hpd.min], 2)} -
+   %\S%expr{round(d["alpha", hpd.max], 2)}];
 $I$,
-  \Sexpr{d["I", round(map)]} 
-  [\Sexpr{d["I", round(hpd.min)]} -
-   \Sexpr{d["I", round(hpd.max)]}];
+  %\S%expr{d["I", round(map)]} 
+  [%\S%expr{d["I", round(hpd.min)]} -
+   %\S%expr{d["I", round(hpd.max)]}];
 $m$,
-  \Sexpr{d["m", floor(map)]} 
-  [\Sexpr{d["m", floor(hpd.min)]} -
-   \Sexpr{d["m", floor(hpd.max)]}];
+  %\S%expr{d["m", floor(map)]} 
+  [%\S%expr{d["m", floor(hpd.min)]} -
+   %\S%expr{d["m", floor(hpd.max)]}];
 $N$,
-  \Sexpr{d["N", round(map)]} 
-  [\Sexpr{d["N", round(hpd.min)]}-
-   \Sexpr{d["N", round(hpd.max)]}].
+  %\S%expr{d["N", round(map)]} 
+  [%\S%expr{d["N", round(hpd.min)]}-
+   %\S%expr{d["N", round(hpd.max)]}].
 The approximate posterior distributions for this simulation are shown in
@@
 $\alpha$, 
-  \Sexpr{p["alpha", round(map, 2)]} 
-  [\Sexpr{p["alpha", round(hpd.min, 2)]} -
-   \Sexpr{p["alpha", round(hpd.max, 2)]}];
+  %\S%expr{p["alpha", round(map, 2)]} 
+  [%\S%expr{p["alpha", round(hpd.min, 2)]} -
+   %\S%expr{p["alpha", round(hpd.max, 2)]}];
 $I$,
-  \Sexpr{p["I", floor(map)]} 
-  [\Sexpr{p["I", floor(hpd.min)]} -
-   \Sexpr{p["I", floor(hpd.max)]}];
+  %\S%expr{p["I", floor(map)]} 
+  [%\S%expr{p["I", floor(hpd.min)]} -
+   %\S%expr{p["I", floor(hpd.max)]}];
 $m$,
-  \Sexpr{p["m", floor(map)]} 
-  [\Sexpr{p["m", floor(hpd.min)]} -
-   \Sexpr{p["m", floor(hpd.max)]}];
+  %\S%expr{p["m", floor(map)]} 
+  [%\S%expr{p["m", floor(hpd.min)]} -
+   %\S%expr{p["m", floor(hpd.max)]}];
 $N$,
-  \Sexpr{p["N", floor(map)]} 
-  [\Sexpr{p["N", floor(hpd.min)]} -
-   \Sexpr{p["N", floor(hpd.max)]}].
+  %\S%expr{p["N", floor(map)]} 
+  [%\S%expr{p["N", floor(hpd.min)]} -
+   %\S%expr{p["N", floor(hpd.max)]}].
 The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
@@
 
-We applied kernel-ABC to five published HIV datasets (\cref{tab:data}),
+We applied ABC to five published HIV datasets (\cref{tab:data}),
 and found substantial heterogeneity among the parameter estimates
@@
 estimates [95\% HPD] 
-  \Sexpr{map1["niculescu2015", round(alpha, 2)]} 
-  [\Sexpr{low1["niculescu2015", round(alpha, 2)]} - 
-   \Sexpr{up1["niculescu2015", round(alpha, 2)]}]
+  %\S%expr{map1["niculescu2015", round(alpha, 2)]} 
+  [%\S%expr{low1["niculescu2015", round(alpha, 2)]} - 
+   %\S%expr{up1["niculescu2015", round(alpha, 2)]}]
 and
-  \Sexpr{map1["wang2015", round(alpha, 2)]} 
-  [\Sexpr{low1["wang2015", round(alpha, 2)]} -
-   \Sexpr{up1["wang2015", round(alpha, 2)]}] respectively).
+  %\S%expr{map1["wang2015", round(alpha, 2)]} 
+  [%\S%expr{low1["wang2015", round(alpha, 2)]} -
+   %\S%expr{up1["wang2015", round(alpha, 2)]}] respectively).
 The MAP estimates did not change appreciably when $m = 1$ was disallowed by the
@@
 data was narrower
-  (\Sexpr{low1["niculescu2015", round(alpha, 2)]} - 
-   \Sexpr{up1["niculescu2015", round(alpha, 2)]}).
+  (%\S%expr{low1["niculescu2015", round(alpha, 2)]} - 
+   %\S%expr{up1["niculescu2015", round(alpha, 2)]}).
 When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
 low estimated $\alpha$ values
-  (\Sexpr{map1["li2015", round(alpha, 2)]} 
-  [\Sexpr{low1["li2015", round(alpha, 2)]} - 
-  \Sexpr{up1["li2015", round(alpha, 2)]}]
+  (%\S%expr{map1["li2015", round(alpha, 2)]} 
+  [%\S%expr{low1["li2015", round(alpha, 2)]} - 
+  %\S%expr{up1["li2015", round(alpha, 2)]}]
 and
-  \Sexpr{map1["cuevas2009", round(alpha, 2)]} 
-  [\Sexpr{low1["cuevas2009", round(alpha, 2)]} -
-   \Sexpr{up1["cuevas2009", round(alpha, 2)]}]). 
+  %\S%expr{map1["cuevas2009", round(alpha, 2)]} 
+  [%\S%expr{low1["cuevas2009", round(alpha, 2)]} -
+   %\S%expr{up1["cuevas2009", round(alpha, 2)]}]). 
 However, the MAP estimates increased when $m = 1$ was not permitted, although
 the HPD intervals remained roughly the same
-  (\Sexpr{map2["li2015", round(alpha, 2)]} 
-  [\Sexpr{low2["li2015", round(alpha, 2)]} - 
-  \Sexpr{up2["li2015", round(alpha, 2)]}]
+  (%\S%expr{map2["li2015", round(alpha, 2)]} 
+  [%\S%expr{low2["li2015", round(alpha, 2)]} - 
+  %\S%expr{up2["li2015", round(alpha, 2)]}]
 and
-  \Sexpr{map2["cuevas2009", round(alpha, 2)]} 
-  [\Sexpr{low2["cuevas2009", round(alpha, 2)]} -
-   \Sexpr{up2["cuevas2009", round(alpha, 2)]}]).
+  %\S%expr{map2["cuevas2009", round(alpha, 2)]} 
+  [%\S%expr{low2["cuevas2009", round(alpha, 2)]} -
+   %\S%expr{up2["cuevas2009", round(alpha, 2)]}]).
 The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
 for both priors on $m$
-  (\Sexpr{map1["novitsky2014", round(alpha, 2)]} for $m \geq 1$;
-   \Sexpr{map2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).
+  (%\S%expr{map1["novitsky2014", round(alpha, 2)]} for $m \geq 1$;
+   %\S%expr{map2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).
 However, the confidence interval was much wider when $m = 1$ was allowed
-  ([\Sexpr{low1["novitsky2014", round(alpha, 2)]} -
-    \Sexpr{up1["novitsky2014", round(alpha, 2)]}] for $m \geq 1$ vs.
-    \Sexpr{low2["novitsky2014", round(alpha, 2)]} -
-    \Sexpr{up2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).
+  ([%\S%expr{low1["novitsky2014", round(alpha, 2)]} -
+    %\S%expr{up1["novitsky2014", round(alpha, 2)]}] for $m \geq 1$ vs.
+    %\S%expr{low2["novitsky2014", round(alpha, 2)]} -
+    %\S%expr{up2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).
 
@@
 intervals compared to the nonzero prior density region
-  (\citeauthor{cuevas2009hiv}, \Sexpr{map1["cuevas2009", floor(I)]} 
-  [\Sexpr{low1["cuevas2009", floor(I)]} -
-   \Sexpr{up1["cuevas2009", floor(I)]}];
-   \citeauthor{niculescu2015recent}, \Sexpr{map1["niculescu2015", floor(I)]}
-  [\Sexpr{low1["niculescu2015", floor(I)]} - 
-   \Sexpr{up1["niculescu2015", floor(I)]}];
-  \citeauthor{li2015hiv}, \Sexpr{map1["li2015", floor(I)]} 
-  [\Sexpr{low1["li2015", floor(I)]} -
-   \Sexpr{up1["li2015", floor(I)]}];
-   \citeauthor{wang2015targeting}, \Sexpr{map1["wang2015", floor(I)]}
-  [\Sexpr{low1["wang2015", floor(I)]} - 
-   \Sexpr{up1["wang2015", floor(I)]}]).
+  (\citeauthor{cuevas2009hiv}, %\S%expr{map1["cuevas2009", floor(I)]} 
+  [%\S%expr{low1["cuevas2009", floor(I)]} -
+   %\S%expr{up1["cuevas2009", floor(I)]}];
+   \citeauthor{niculescu2015recent}, %\S%expr{map1["niculescu2015", floor(I)]}
+  [%\S%expr{low1["niculescu2015", floor(I)]} - 
+   %\S%expr{up1["niculescu2015", floor(I)]}];
+  \citeauthor{li2015hiv}, %\S%expr{map1["li2015", floor(I)]} 
+  [%\S%expr{low1["li2015", floor(I)]} -
+   %\S%expr{up1["li2015", floor(I)]}];
+   \citeauthor{wang2015targeting}, %\S%expr{map1["wang2015", floor(I)]}
+  [%\S%expr{low1["wang2015", floor(I)]} - 
+   %\S%expr{up1["wang2015", floor(I)]}]).
 The \citeauthor{novitsky2014impact} data was the outlier, with a very high
 estimated $I$, and HPD interval spanning almost the entire prior region
-  (\Sexpr{map1["novitsky2014", floor(I)]} 
-  [\Sexpr{low1["novitsky2014", floor(I)]} -
-   \Sexpr{up1["novitsky2014", floor(I)]}]).
+  (%\S%expr{map1["novitsky2014", floor(I)]} 
+  [%\S%expr{low1["novitsky2014", floor(I)]} -
+   %\S%expr{up1["novitsky2014", floor(I)]}]).
 The $I$ estimates and HPD intervals were generally robust to the choice of
@@
 upper bound of the HPD interval was different for each dataset
-  (\citeauthor{niculescu2015recent}, \Sexpr{up1["niculescu2015", round(m)]};
-   \citeauthor{wang2015targeting}, \Sexpr{up1["wang2015", round(m)]};
-   \citeauthor{li2015hiv}, \Sexpr{up1["li2015", round(m)]};
-   \citeauthor{cuevas2009hiv}, \Sexpr{up1["cuevas2009", round(m)]}).
+  (\citeauthor{niculescu2015recent}, %\S%expr{up1["niculescu2015", round(m)]};
+   \citeauthor{wang2015targeting}, %\S%expr{up1["wang2015", round(m)]};
+   \citeauthor{li2015hiv}, %\S%expr{up1["li2015", round(m)]};
+   \citeauthor{cuevas2009hiv}, %\S%expr{up1["cuevas2009", round(m)]}).
 When $m = 1$ was disallowed, the MAP for all datasets was either 2 or 3, with
@@
 data, for which the MAP estimate was lower 
-  (\Sexpr{map1["li2015", floor(N)]})
+  (%\S%expr{map1["li2015", floor(N)]})
 when $m = 1$ was allowed.
@@
       parameters of the BA network model, fitted to five published HIV datasets
-      with kernel-ABC. $x$-axes indicate regions of nonzero prior density. In
+      with ABC. $x$-axes indicate regions of nonzero prior density. In
       particular, the prior on $m$ was DiscreteUniform(1, 5).
@@
 shape~\autocite{poon2015phylodynamic}. In this work, we developed a
-kernel-ABC-based method to infer the parameters of a contact network model. The
+ABC-based method to infer the parameters of a contact network model. The
 method is general, and could be applied to any model from which contact
@@
 nodes, had a strong influence on tree shape. This was reflected in the relative
-accuracy of the kernel-ABC estimates of these parameters. The total number of
+accuracy of the ABC estimates of these parameters. The total number of
 nodes $N$ had a weak influence on tree shape, which was most prominent when the
@@
 did not produce much variation in tree shape, resulting in in both poorly
-performing classifiers and uninformative kernel-ABC estimates.
-
-$N$ was almost always significantly over-estimated using kernel-ABC. Since the
+performing classifiers and uninformative ABC estimates.
+
+$N$ was almost always significantly over-estimated using ABC. Since the
 prior on $N$ and $I$ is jointly uniform on a non-rectangular region ($I \leq
@@
 work, and previous work by our group~\autocite{poon2015phylodynamic}, has
-demonstrated that kernel-ABC is a broadly applicable and effective framework in
+demonstrated that ABC is a broadly applicable and effective framework in
 which to perform phylodynamic inference.
@@
   \caption{
-    Point estimates of BA model parameters obtained by running kernel-ABC
+    Point estimates of BA model parameters obtained by running ABC
     on simulated phylogenies without training, for simulations with $m = 3$.
@@
   \caption{
-    Point estimates of BA model parameters obtained by running kernel-ABC
+    Point estimates of BA model parameters obtained by running ABC
     on simulated phylogenies without training, for simulations with $m = 4$.
@@
         Approximate marginal posterior distributions of Barab\'asi-Albert
-        model parameters obtained using kernel-ABC for a network with
+        model parameters obtained using ABC for a network with
         heterogeneous node behaviour. Half of the nodes were attached with
@@
     \caption{Approximate marginal posterior distributions of Barab\'asi-Albert
-        model parameters obtained using kernel-ABC for a network with
+        model parameters obtained using ABC for a network with
         peer-driven sampling. An epidemic was simulated in the usual fashion,
@@
       parameters of the BA network model, fitted to five published HIV datasets
-      with kernel-ABC. $x$-axes indicate regions of nonzero prior density. In
+      with ABC. $x$-axes indicate regions of nonzero prior density. In
       particular, the prior on $m$ was DiscreteUniform(2, 5).
@@
 
-<<abc_posterior, echo=FALSE, results="asis">>=
+<<abc_posterior, echo=FALSE, results="asis", eval=FALSE>>=
     N <- 5000
@@
         cat("\\caption{\n")
-        cat(sprintf("Approximate marginal posterior distributions of Barab\\'asi-Albert model parameters obtained using kernel-ABC for one simulated transmission tree with parameter values $\\alpha = %.1f$, $I = %d$, $m = %d$, and $N = %d$.
+        cat(sprintf("Approximate marginal posterior distributions of Barab\\'asi-Albert model parameters obtained using ABC for one simulated transmission tree with parameter values $\\alpha = %.1f$, $I = %d$, $m = %d$, and $N = %d$.
 Dashed lines indicate true values, and shaded areas show 95\\%% highest posterior density intervals.\n", alpha, I, m, N))

### 2016-06-25 programs/netabc/R/cv.R

@@ -107,7 +107,8 @@ ksvm.cv <- function (kmat, y, n.cv=1000,
           stats=stats, show.progress=show.progress, nthread=nthread)
 }
 
-#' Perform replicate 2-fold cross-validations on an SVM classifier.
+#' Perform replicate 2-fold cross-validations on an SVM classifier from a
+#' distance matrix.
 #'
 #' Returns a data.frame with the indicated statistics for each individual
 #' cross-validation. Currently the only supported statistics are "accuracy" for
@@ -131,6 +132,33 @@ dsvm.cv <- function (dmat, y, n.cv=1000,
     fit.model <- function(d, train.y) svm(d, train.y)
     predict <- predict
     do.cv(dmat, y, get.train, get.test, fit.model, predict, n.cv=n.cv,
+          stats=stats, show.progress=show.progress, nthread=nthread)
+}
+
+#' Perform replicate 2-fold cross-validations on an SVM classifier from data.
+#'
+#' Returns a data.frame with the indicated statistics for each individual
+#' cross-validation. Currently the only supported statistics are "accuracy" for
+#' categorical labels or "rsquared" for numeric labels.
+#'
+#' @param fmat data matrix
+#' @param y labels for data
+#' @param stats statistics to calculate
+#' @param show.progress whether to display a progress bar
+#' @param nthread number of threads to use
+#' @return A data.frame with the indicated statistics for each CV
+#' @seealso \url{http://stackoverflow.com/questions/1753299}
+#' @seealso \code{\link{kernlab::ksvm}}
+#' @export
+fsvm.cv <- function (fmat, y, n.cv=1000, 
+                    stats=ifelse(is.factor(y), "accuracy", "rsquared"),
+                    show.progress=TRUE, nthread=1)
+{
+    get.train <- function(d, holdout) d[-holdout,,drop=FALSE]
+    get.test <- function(d, holdout, m) d[holdout,,drop=FALSE]
+    fit.model <- function(d, train.y) svm(d, train.y)
+    predict <- predict
+    do.cv(fmat, y, get.train, get.test, fit.model, predict, n.cv=n.cv,
           stats=stats, show.progress=show.progress, nthread=nthread)
 }
 

### 2016-06-27 doc/defence/presentation.tex
Paper and defence presentation updates

@@
 \usepackage{qrcode}
-\usepackage[backend=biber, style=authoryear, uniquelist=minyear, uniquename=false]{biblatex}
+\usepackage[backend=biber, style=numeric-comp, sorting=none]{biblatex}
 
@@
     \uncover<9->{
-      \node (n5) at (-0.86, 0.5) [circle, fill=black] { };
-      \draw (n5) -- (n2);
-      \draw (n5) -- (n1);
+      \node (n6) at (-0.86, 0.5) [circle, fill=black] { };
+      \draw (n6) -- (n2);
+      \draw (n6) -- (n1);
     }
@@
     \uncover<10->{
-      \node (n6) at (-1, 0) [circle, fill=black] { };
-      \draw (n6) -- (n3);
-      \draw (n6) -- (n4);
+      \node (n7) at (-1, 0) [circle, fill=black] { };
+      \draw (n7) -- (n3);
+      \draw (n7) -- (n4);
     }
@@
     \uncover<11->{
-      \node (n7) at (-0.86, -0.5) [circle, fill=black] { };
-      \node (n8) at (-0.5, -0.86) [circle, fill=black] { };
-      \node (n9) at (0, -1) [circle, fill=black] { };
-      \node (n10) at (0.5, -0.86) [circle, fill=black] { };
-      \node (n11) at (0.86, -0.5) [circle, fill=black] { };
+      \node (n8) at (-0.86, -0.5) [circle, fill=black] { };
+      \node (n9) at (-0.5, -0.86) [circle, fill=black] { };
+      \node (n10) at (0, -1) [circle, fill=black] { };
+      \node (n11) at (0.5, -0.86) [circle, fill=black] { };
+      \node (n12) at (0.86, -0.5) [circle, fill=black] { };
 
-      \draw (n7) -- (n2);
-      \draw (n7) -- (n4);
       \draw (n8) -- (n2);
-      \draw (n8) -- (n1);
+      \draw (n8) -- (n4);
       \draw (n9) -- (n2);
-      \draw (n9) -- (n7);
-      \draw (n10) -- (n3);
+      \draw (n9) -- (n1);
+      \draw (n10) -- (n2);
       \draw (n10) -- (n8);
-      \draw (n11) -- (n2);
-      \draw (n11) -- (n7);
+      \draw (n11) -- (n3);
+      \draw (n11) -- (n9);
+      \draw (n12) -- (n2);
+      \draw (n12) -- (n8);
     }
+
+    \uncover<12->{
+      \node (n1) at (1, 0) [circle, fill=red] { };
+      \node (n2) at (0.86, 0.5) [circle, fill=red] { };
+      \node (n5) at (-0.86, 0.5) [circle, fill=red] { };
+      \node (n7) at (-0.86, -0.5) [circle, fill=red] { };
+      \node (n8) at (-0.5, -0.86) [circle, fill=red] { };
+
+      \draw [ultra thick, red] (n2) -- (n7);
+      \draw [ultra thick, red] (n2) -- (n1);
+      \draw [ultra thick, red] (n8) -- (n1);
+      \draw [ultra thick, red] (n5) -- (n2);
+    }
+
+    \only<11>{\node [below=of n10] {$N = 12$};}
+    \uncover<12->{\node [below=of n10] {$I = 5$};}
   \end{tikzpicture}
@@
     $m$ = number of new edges per vertex \\
+    \hfill\\
   }
@@
     $\alpha$ = preferential attachment power \\
+    \hfill\\
   }
@@
   \uncover<11->{
-    $N$ = number of nodes
+    $N$ = number of nodes \\
+    \hfill\\
   }
-  \end{columns}
-\end{frame}
 
-\end{document}
-
-\begin{frame}{Barab\'asi-Albert model incorporates preferential attachment}
-  \begin{columns}
-    \begin{column}{0.5\textwidth}
-      \begin{minipage}[p][\textheight][t]{\textwidth}
-        \centering
-        \only<1>{\includegraphics[page=1, width=\textwidth]{pa.pdf}}
-        \only<2>{\includegraphics[page=2, width=\textwidth]{pa.pdf}}
-        \only<3>{\includegraphics[page=3, width=\textwidth]{pa.pdf}}
-        \only<4>{\includegraphics[page=4, width=\textwidth]{pa.pdf}}
-        \only<5>{\includegraphics[page=5, width=\textwidth]{pa.pdf}}
-        \only<6>{\includegraphics[page=6, width=\textwidth]{pa.pdf}}
-      \end{minipage}
-    \end{column}
-    \begin{column}{0.5\textwidth}
-      \begin{minipage}[t][\textheight][t]{\textwidth}
-      \begin{itemize}
-        \uncover<1->{ \item Start with a small number of connected nodes. }
-        \uncover<2->{ \item Attach new nodes with $m$ edges. }
-        \uncover<3->{ \item Other endpoints of degree $d$ are chosen with probability $\propto d^{\alpha} + 1$. }
-        \uncover<5->{ \item Continue until network has $N$ nodes. }
-        \uncover<6->{ \item Also consider the prevalence $I$. }
-      \end{itemize}
-      \end{minipage}
-    \end{column}
+  \uncover<12->{
+    $I$ = number of infected nodes (prevalence) 
+  }
   \end{columns}
@@
 \begin{frame}{Phylogenetic methods may help fit network models}
+  \begin{itemize}
+    \item network models are difficult to fit with conventional epidemiological
+      methods (surveys etc.)
+    \item sequence data may be easier to collect and free from some biases
+      (misreporting etc.)
+    \item instead of asking the people, ask the virus! (some cartoon)
+  \end{itemize}
 \end{frame}
@@
 \begin{frame}{Transmission trees shape viral phylogenies}
+  some demo of viruses evolving and being transmitted at the same time
 \end{frame}
@@
 \begin{frame}{Can we fit network models with viral phylogenies?}
+  \begin{center}
+  network $\to$ transmission tree $\to$ viral phylogeny \\
+  \hfill\\
+  network $\stackrel{\alert{?}}{\leftarrow}$ transmission tree $\leftarrow$ viral phylogeny \\
+  \end{center}
 \end{frame}
 
-\end{document}
-
-%\begin{frame}{Preferential attachment generates scale-free networks}
-%  \includegraphics[width=\textwidth]{pa-example}
-%  \begin{itemize}
-%    \item Sexual networks are often scale-free: degree distributions appear
-%      follow a power law ($\Pr(\text{degree} = d) \propto d^{-\gamma}$).
-%      \pause
-%    \item Preferential attachment models generate scale-free networks.
-%  \end{itemize}
-%\end{frame}
-
-\begin{frame}{Do network parameters measurably affect tree shape?}
-  \only<1>{
-    \includegraphics[height=1in, trim=0 2in 4in 0, clip]{kernel-idea}
-  }
-  \only<2>{
-    \includegraphics[height=1in, trim=0 2in 2in 0, clip]{kernel-idea}
-  }
-  \only<3->{
-    \includegraphics[height=1in, trim=0 2in 0 0, clip]{kernel-idea}
-  }
-  \begin{itemize}
-    \setlength{\itemsep}{0pt}
-    \item Generate networks under different parameter values (number of nodes
-      $N$, number of edges per vertex $m$, pereferential attachment power
-      $\alpha$).
-    \uncover<2->{
-    \item Simulate epidemic over each network until $I$ nodes are infected.
-    }
-    \uncover<3->{
-    \item Randomly subsample to form transmission trees.
-    }
-    \uncover<4->{
-    \item Compare trees pairwise using tree kernel\footfullcite{poon2013mapping}.
-    }
-  \end{itemize}
-\end{frame}
-
-\begin{frame}{Preferential attachment power $\alpha$ affects tree shape}
-  \begin{tikzpicture}[remember picture, overlay]
-    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=0.2in 2.7in 3.3in 0, clip]{kernel-kpca.pdf}};
-    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-alpha-tree.pdf}};
-  \end{tikzpicture}
-\end{frame}
-
-\begin{frame}{Prevalence $I$ affects tree shape}
-  \begin{tikzpicture}[remember picture, overlay]
-    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=3.5in 2.7in 0in 0, clip]{kernel-kpca.pdf}};
-    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-I-tree.pdf}};
-  \end{tikzpicture}
-\end{frame}
-
-\begin{frame}{Number of edges per vertex $m$ does not affect tree shape}
-  \begin{tikzpicture}[remember picture, overlay]
-    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=0.2in 0.3in 3.3in 2.2in, clip]{kernel-kpca.pdf}};
-    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-m-tree.pdf}};
-  \end{tikzpicture}
-\end{frame}
-
-\begin{frame}{Number nodes $N$ modestly affects tree shape}
-  \begin{tikzpicture}[remember picture, overlay]
-    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=3.5in 0.3in 0 2.2in, clip]{kernel-kpca.pdf}};
-    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-N-tree.pdf}};
-  \end{tikzpicture}
-\end{frame}
-
-\begin{frame}{Simulated trees can be used to fit a network model}
+\begin{frame}{Are network parameters identifiable from tree shape?}
   \begin{minipage}[p][\textheight][t]{\textwidth}
-    \only<1>{\includegraphics[width=0.9\textwidth, trim=0 3in 0 0, clip]{abc-idea.pdf}}
-    \only<2>{\includegraphics[width=0.9\textwidth, trim=0 1.8in 0 0, clip]{abc-idea.pdf}}
-    \only<3>{\includegraphics[width=0.9\textwidth, trim=0 1in 0 0, clip]{abc-idea.pdf}}
-    \uncover<4->{\includegraphics[width=0.9\textwidth, trim=0 0in 0 0, clip]{abc-idea.pdf}}
-    \uncover<5->{\vspace{0.5cm}\centerline{\textbf{kernel approximate Bayesian computation (kernel-ABC)}}}
+    \only<1>{\includegraphics[width=0.9\textwidth, trim=0 3in 0 0, clip]{kernel-idea.pdf}}
+    \only<2>{\includegraphics[width=0.9\textwidth, trim=0 1.7in 0 0, clip]{kernel-idea.pdf}}
+    \only<3>{\includegraphics[width=0.9\textwidth, trim=0 0.8in 0 0, clip]{kernel-idea.pdf}}
+    \only<4->{\includegraphics[width=0.9\textwidth, trim=0 0in 0 0, clip]{kernel-idea.pdf}}
   \end{minipage}
@@
 
-\begin{frame}{Example simulation shows $\alpha$ and $I$ can be reconstructed}
+\begin{frame}{yes they do}
+  \centerline{\includegraphics[height=0.8\textheight]{kernel-kpca}}
+\end{frame}
+
+\begin{frame}{Simulated trees can be used to fit network models}
+  \centerline{\includegraphics[width=0.9\textwidth]{abc-smc}}
+
+  \centerline{animate this?}
+\end{frame}
+
+\begin{frame}{Posterior mean estimates are more accurate for $\alpha$ and $I$}
+  \begin{minipage}[p][\textheight][t]{\textwidth}
+    \vspace{-0.5cm}
+    preferential attachment power \hfill \uncover<2>{edges per vertex}
+
+    \hspace{0.75cm}\only<1>{\includegraphics[height=0.8\textheight, trim=0 0 3.4in 0, clip]{abc-point-estimate-m3.pdf}}
+    \only<2>{\includegraphics[height=0.8\textheight]{abc-point-estimate-m3.pdf}}
+
+    prevalence \hfill \uncover<2>{total nodes}
+  \end{minipage}
+\end{frame}
+
+\begin{frame}{Example simulation\ldots}
   \begin{minipage}[p][\textheight][t]{\textwidth}
@@
 
-\begin{frame}{Maximum \textit{a posteriori} estimates are accurate for $\alpha$ and $I$}
-  \begin{minipage}[p][\textheight][t]{\textwidth}
-    \vspace{-0.5cm}
-    preferential attachment power \hfill \uncover<2>{edges per vertex}
-
-    \hspace{0.75cm}\only<1>{\includegraphics[height=0.8\textheight, trim=0 0 3.4in 0, clip]{abc-point-estimate-m2.pdf}}
-    \only<2>{\includegraphics[height=0.8\textheight]{abc-point-estimate-m2.pdf}}
-
-    prevalence \hfill \uncover<2>{total nodes}
-  \end{minipage}
+\begin{frame}{Example simulation continued, $N$ and $I$ confounded}
+  %\includegraphics[width=\textwidth]{{abc-posterior-2d/1.0_1000_3_5000_0}.pdf}
 \end{frame}
 
-\begin{frame}{Real world HIV datasets exhibit network heterogeneity}
-  \uncover<2->{
-    \centerline{\includegraphics[height=0.8\textheight]{realdata-hpd-bc}}
-  }
+\begin{frame}{Application to real world HIV datasets}
+  \centerline{\input{\tablepath/realdata-bc-short}}
+\end{frame}
 
-%  \scriptsize
-%    \begin{tabular}{ccccc}
-%      Reference & Sequences ($n$) & Location & Risk group & Gene \\
-%      \hline
-%      \color{magenta} \textcite{novitsky2013phylogenetic} &
-%        \color{magenta} \multirow{2}{*}{180} &
-%        \color{magenta} \multirow{2}{*}{Mochudi, Botswana} & 
-%        \color{magenta} \multirow{2}{*}{HET} &
-%        \color{magenta} \multirow{2}{*}{\textit{env}} \\ 
-%      \color{magenta} \textcite{novitsky2014impact} \\
-%      \color{blue} \textcite{cuevas2009hiv} & 
-%        \color{blue} 287 & 
-%        \color{blue} Basque Country, Spain & 
-%        \color{blue} mixed & 
-%        \color{blue} \textit{pol} \\
-%      \color{cyan} \textcite{li2015hiv} & 
-%        \color{cyan} 280 & 
-%        \color{cyan} Shanghai, China & 
-%        \color{cyan} MSM & 
-%        \color{cyan} \textit{pol} \\
-%      \color{green} unpublished & 
-%        \color{green} 399 & 
-%        \color{green} British Columbia, Canada & 
-%        \color{green} IDU & 
-%        \color{green} \textit{pol} \\
-%      \color{orange} \textcite{wang2015targeting} & 
-%        \color{orange} 173 & 
-%        \color{orange} Beijing, China & 
-%        \color{orange} MSM & 
-%        \color{orange} \textit{pol} \\
-%      \color{red} \textcite{niculescu2015recent} & 
-%        \color{red} 136 & 
-%        \color{red} Romaina & 
-%        \color{red} IDU & 
-%        \color{red} \textit{pol} \\
-%      \hline
-%    \end{tabular}
-%  \normalsize
+\begin{frame}{HIV networks have sub-linear preferential attachment}
+    Also, higher PA for IDU networks.
+
+    \includegraphics[trim=0 3.5in 3.3in 0, clip, width=0.5\textwidth]{realdata-hpd-bc}
+    \includegraphics[trim=3.4in 1.5in 0 2.0in, clip, width=0.5\textwidth]{realdata-hpd-bc}
 \end{frame}
@@
   \begin{itemize}
-    \item Kernel-ABC is the first phylodynamic method to fit contact network
-      models to phylogenetic data.
+    \item We developed a phylodynamic method to fit contact network models to
+      phylogenetic data.
       \pause
@@
       model, which is challenging to estimate by traditional epidemiological
-      methods, can be estimated with kernel-ABC.
+      methods, can be estimated with ABC.
       \pause
@@
       phylodynamic analyses.
-      \pause
   \end{itemize}
-  \centerline{\qrcode{github.com/rmcclosk/netabc} github.com/rmcclosk/netabc}
 \end{frame}
@@
 
-\begin{frame}{Classifiers for network parameters}
-  \vspace{-1cm}
-  \begin{itemize}
-    \item kSVR using tree kernel
-    \item SVR using normalized
-      lineages-through-time\footfullcite{janzen2015approximate}
-    \item linear regression using Sackin's index~\footfullcite{shao1990tree}
-  \end{itemize}
-
-  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
-\end{frame}
-
-\begin{frame}{Sequential Monte Carlo for ABC}
-  \begin{tikzpicture}
-    \only<1-2>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 4in 0, clip]{abc-smc.pdf}}};
-    \only<1>{\fill[white] (0.2,-6.5) rectangle (4,-3.5)};
-    \only<3>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 2in 0, clip]{abc-smc.pdf}}};
-    \only<4>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 0 0, clip]{abc-smc.pdf}}};
-  \end{tikzpicture}
-\end{frame}
-
 \end{document}

### 2016-06-27 doc/figures/transtree.tex
Paper and defence presentation updates

\documentclass[landscape]{article}
\usepackage{tikz}
\usepackage[T1]{fontenc}
\usepackage{mathptmx}
\usepackage{fullpage}

\usetikzlibrary{positioning}
\usetikzlibrary{calc}
\usetikzlibrary{arrows}
\usetikzlibrary{automata}

\graphicspath{{stock/}}

\begin{document}
\pagestyle{empty}

\begin{tikzpicture}
  [every node/.style = {inner sep=0pt},
   every path/.style = {->, >=stealth, very thick},
   label/.style = {anchor=south, color=white, inner sep=6pt},
   time/.style = {anchor=north west, inner sep=2pt}]
  \def\pht{1.5cm}
  \def\pd{0.5cm}
  \def\nd{1cm}

  \node (abc) {\includegraphics[height=\pht]{person}};
  \node at (abc.south) [label] {\huge a};

  \node (ac) [below left=\pd of abc] {\includegraphics[height=\pht]{person}};
  \node at (ac.south) [label] {\huge a};

  \node (a) [below left=\pd of ac] {\includegraphics[height=\pht]{person}};
  \node at (a.south) [label] {\huge a};
  \node (b) [below right=\pd of abc] {\includegraphics[height=\pht]{person}};
  \node at (b.south) [label] {\huge b};
  \node (c) [below right=\pd of ac] {\includegraphics[height=\pht]{person}};
  \node at (c.south) [label] {\huge c};

  \draw [red] (abc) -- (b);
  \draw [-, dashed] (abc) -- (ac);
  \draw [blue] (ac) -- (c);
  \draw [-, dashed] (ac) -- (a);

  %\node (a) [left=7cm of abc] {\includegraphics[height=\pht]{person}};
  %\node at (a.south) [label] {\huge a};
  %\node (b) [below left=\nd of a] {\includegraphics[height=\pht]{person}};
  %\node at (b.south) [label] {\huge b};
  %\node (c) [below right=\nd of a] {\includegraphics[height=\pht]{person}};
  %\node at (c.south) [label] {\huge c};

  %\draw [red] (a) -- (b);
  %\draw [blue] (a) -- (c);

  \node [below=6cm of abc] (abc) {\includegraphics[height=\pht]{person}};
  \node at (abc.south) [label] {\huge b};

  \node (ac) [below left=\pd of abc] {\includegraphics[height=\pht]{person}};
  \node at (ac.south) [label] {\huge c};

  \node (a) [below left=\pd of ac] {\includegraphics[height=\pht]{person}};
  \node at (a.south) [label] {\huge a};
  \node (b) [below right=\pd of abc] {\includegraphics[height=\pht]{person}};
  \node at (b.south) [label] {\huge b};
  \node (c) [below right=\pd of ac] {\includegraphics[height=\pht]{person}};
  \node at (c.south) [label] {\huge c};

  \draw [-, dashed] (abc) -- (b);
  \draw [green!60!black] (abc) -- (ac);
  \draw [-, dashed] (ac) -- (c);
  \draw [cyan] (ac) -- (a);

  %\node (a) [left=7cm of abc] {\includegraphics[height=\pht]{person}};
  %\node at (a.south) [label] {\huge a};
  %\node (b) [below left=\nd of a] {\includegraphics[height=\pht]{person}};
  %\node at (b.south) [label] {\huge b};
  %\node (c) [below right=\nd of a] {\includegraphics[height=\pht]{person}};
  %\node at (c.south) [label] {\huge c};

  %\draw [green!60!black] (b) -- (c);
  %\draw [cyan] (c) -- (a);

  \node [above left=2cm and 8cm of abc] (abc) {\includegraphics[height=\pht]{person}};
  \node (ac) [below left=\pd of abc] {\includegraphics[height=\pht]{person}};

  \node (a) [below left=\pd of ac] {\includegraphics[height=\pht]{person}};
  \node at (a.south) [label] {\huge a};
  \node (b) [below right=\pd of abc] {\includegraphics[height=\pht]{person}};
  \node at (b.south) [label] {\huge b};
  \node (c) [below right=\pd of ac] {\includegraphics[height=\pht]{person}};
  \node at (c.south) [label] {\huge c};

  \draw [-] (abc) -- (b);
  \draw [-] (abc) -- (ac);
  \draw [-] (ac) -- (c);
  \draw [-] (ac) -- (a);

  %\node (a) [left=7cm of abc] {\includegraphics[height=\pht]{person}};
  %\node at (a.south) [label] {\huge a};
  %\node (b) [below left=\nd of a] {\includegraphics[height=\pht]{person}};
  %\node at (b.south) [label] {\huge b};
  %\node (c) [below right=\nd of a] {\includegraphics[height=\pht]{person}};
  %\node at (c.south) [label] {\huge c};

\end{tikzpicture}
\end{document}

### 2016-06-27 doc/paper/paper.Rnw
Paper and defence presentation updates

@@
 
+\let\mref\cref
+\let\Mref\Cref
 \let\mtextcite\textcite
@@
 \let\mciteauthor\citeauthor
+\renewcommand{\cref}[1]{\mbox{\mref{#1}}}
+\renewcommand{\Cref}[1]{\mbox{\Mref{#1}}}
 \renewcommandtwoopt{\textcite}[3][][]{\mbox{\mtextcite[#1][#2]{#3}}}
@@
 
-\end{document}
-
-The \gls{kSVR} classifier for \gls{alpha} had an average $R^2$ of 
+The kSVR classifier for $\alpha$ had an average $R^2$ of 
     \Sexpr{kernel.alpha[,round(mean(rsquared), 2)]},
 compared to 
-    %\S%expr{nltt.alpha[,round(mean(rsquared), 2)]}
+    \Sexpr{nltt.alpha[,round(mean(rsquared), 2)]}
 for the nLTT-based SVR, and
-    %\S%expr{sackin.alpha[,round(mean(rsquared), 2)]}
+    \Sexpr{sackin.alpha[,round(mean(rsquared), 2)]}
 for the linear regression against Sackin's index. There was little variation
@@
 $R^2$ values of 
-  %\S%expr{kernel.m[,round(mean(rsquared), 2)]} for kSVR,
-  %\S%expr{nltt.m[,round(mean(rsquared), 2)]} for the nLTT, and
-  %\S%expr{sackin.m[,round(mean(rsquared), 2)]}
+  \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for kSVR,
+  \Sexpr{nltt.m[,round(mean(rsquared), 2)]} for the nLTT, and
+  \Sexpr{sackin.m[,round(mean(rsquared), 2)]}
 for Sackin's index. Again, there was little variation in accuracy between
@@
 tips in the tree. For 100-tip trees, the average $R^2$ values were
-  %\S%expr{kernel.I[ntip == 100, round(mean(rsquared), 2)]},
-  %\S%expr{nltt.I[ntip == 100, round(mean(rsquared), 2)]}, and
-  %\S%expr{sackin.I[ntip == 100, round(mean(rsquared), 2)]}
+  \Sexpr{kernel.I[ntip == 100, round(mean(rsquared), 2)]},
+  \Sexpr{nltt.I[ntip == 100, round(mean(rsquared), 2)]}, and
+  \Sexpr{sackin.I[ntip == 100, round(mean(rsquared), 2)]}
 for the tree kernel, nLTT, and Sackin's index respectively. For 500-tip
 trees, the values increased to
-  %\S%expr{kernel.I[ntip == 500, round(mean(rsquared), 2)]},
-  %\S%expr{nltt.I[ntip == 500, round(mean(rsquared), 2)]}, and
-  %\S%expr{sackin.I[ntip == 500, round(mean(rsquared), 2)]}.
+  \Sexpr{kernel.I[ntip == 500, round(mean(rsquared), 2)]},
+  \Sexpr{nltt.I[ntip == 500, round(mean(rsquared), 2)]}, and
+  \Sexpr{sackin.I[ntip == 500, round(mean(rsquared), 2)]}.
 Finally, the performance of classifiers for $N$ depended heavily on the
 epidemic scenario. The $R^2$ of the kSVR classifier ranged from
-  %\S%expr{kernel.N[,round(min(rsquared), 2)]}
+  \Sexpr{kernel.N[,round(min(rsquared), 2)]}
 for the smallest epidemic and smallest sample size, to
-  %\S%expr{kernel.N[,round(max(rsquared), 2)]}
+  \Sexpr{kernel.N[,round(max(rsquared), 2)]}
 for the largest. Likewise, $R^2$ for the nLTT-based SVR ranged from 
-  %\S%expr{nltt.N[,round(min(rsquared), 2)]}
+  \Sexpr{nltt.N[,round(min(rsquared), 2)]}
 to
-  %\S%expr{nltt.N[,round(max(rsquared), 2)]}.
+  \Sexpr{nltt.N[,round(max(rsquared), 2)]}.
 Sackin's index did not accurately classify $N$ in any scenario, with an average
 $R^2$ of
-  %\S%expr{sackin.N[,round(mean(rsquared), 2)]}
+  \Sexpr{sackin.N[,round(mean(rsquared), 2)]}
 and little variation between scenarios.
@@
                                 d[true_alpha == 0.5 | true_alpha == 1.5, m_error])
-    stopifnot(alpha.av["factor(true_m)", "Pr(>F)"] <= 0.05)
-    stopifnot(alpha.av["factor(true_I)", "Pr(>F)"] > 0.05)
-    stopifnot(I.I.test$p.value < 1e-5)
-    stopifnot(I.av["factor(true_m)", "Pr(>F)"] > 0.05)
-    stopifnot(m.av["factor(true_m)", "Pr(>F)"] > 0.05)
-    stopifnot(m.av["factor(true_I)", "Pr(>F)"] > 0.05)
-    stopifnot(m.av["factor(true_alpha)", "Pr(>F)"] > 0.05)
-    stopifnot(min(N.av[1:3,"Pr(>F)"]) > 0.05)
+    #stopifnot(alpha.av["factor(true_m)", "Pr(>F)"] <= 0.05)
+    #stopifnot(alpha.av["factor(true_I)", "Pr(>F)"] > 0.05)
+    #stopifnot(I.I.test$p.value < 1e-5)
+    #stopifnot(I.av["factor(true_m)", "Pr(>F)"] > 0.05)
+    #stopifnot(m.av["factor(true_m)", "Pr(>F)"] > 0.05)
+    #stopifnot(m.av["factor(true_I)", "Pr(>F)"] > 0.05)
+    #stopifnot(m.av["factor(true_alpha)", "Pr(>F)"] > 0.05)
+    #stopifnot(min(N.av[1:3,"Pr(>F)"]) > 0.05)
 
@@
 
-\Cref{fig:abcpt} shows maximum \textit{a posteriori} (MAP) point
-estimates of the BA model parameters obtained with ABC on simulated
+\Cref{fig:abcpt} shows \del{maximum \textit{a posteriori} (MAP)} \add{posterior
+mean} point estimates of the BA model parameters obtained with ABC on simulated
 data. The estimates shown correspond only to the simulations where the $m$
@@
 
-The accuracy of the parameter estimates obtained with ABC
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{abc-point-estimate-m2}
+    \caption[
+        Posterior mean point estimates for BA model parameters obtained by
+        running \textit{netabc} on simulated data, for simulations with
+        $m = 2$.
+    ]{
+        Posterior mean point estimates for BA model parameters obtained by
+        running \textit{netabc} on simulated data, for simulations with
+        $m = 2$. Dashed lines indicate true values. (A) Estimates of
+        $\alpha$ and $I$ which were varied in these simulations against known
+        values. (B) Estimates of $m$ and $N$ which were held fixed in these
+        simulations at the values $m = 2$ and $N = 5000$.
+    }
+    \label{fig:abcptm2}
+\end{figure}
+
+\begin{table*}[ht]
+    \centering
+    \input{\tablepath/abc-hpd.tex}
+    \caption[
+        Average posterior mean point estimates and 95\% highest posterior
+        density interval widths for BA model parameter estimates obtained
+        with \textit{netabc} on simulated data.
+    ]{
+        Average posterior mean point estimates and 95\% highest posterior
+        density interval widths for BA model parameter estimates obtained with
+        \textit{netabc} on simulated data. Three transmission trees were
+        simulated under each combination of the listed parameter values, and
+        the parameters were estimated with ABC without training.
+    }
+    \label{tab:abchpd}
+\end{table*}
+
+\del{The accuracy of the parameter estimates obtained with ABC
 paralleled the results from the kSVR classifier. Of the four parameters,
@@
 median [IQR] absolute error of 
-    %\S%expr{d[,round(median(alpha_error), 2)]} 
-    [%\S%expr{d[,round(quantile(alpha_error, 0.25), 2)]} - 
-    %\S%expr{d[,round(quantile(alpha_error, 0.75), 2)]}].
+    \Sexpr{d[,round(median(alpha_error), 2)]} 
+    [\Sexpr{d[,round(quantile(alpha_error, 0.25), 2)]} - 
+    \Sexpr{d[,round(quantile(alpha_error, 0.75), 2)]}].
 The errors when the true value of $\alpha$ was zero were significantly greater
 than those for the other values 
-    (Wilcoxon rank-sum test, $p$ = %\S%expr{latexSN(round(zero.alpha.test$p.value, 4))}$).
+(Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(zero.alpha.test[["p.value"]], 4))}$).
 Errors in estimating $\alpha$ also varied with the true value of $m$ just at
 the threshold of statistical significance
-    ($p$ %\S%expr{pp(alpha.av["factor(true_m)", "Pr(>F)"])}$),
+    $p \Sexpr{pp(alpha.av["factor(true_m)", "Pr(>F)"])}$),
 but did not vary across the true values of $N$ or $I$ (both one-way ANOVA).
 Estimates for $I$ were relatively accurate, with point estimate errors of
-    %\S%expr{d[,round(median(I_error))]} 
-    [%\S%expr{d[,round(quantile(I_error, 0.25))]} - 
-    %\S%expr{d[,round(quantile(I_error, 0.75))]}] individuals.
+    \Sexpr{d[,round(median(I_error))]} 
+    [\Sexpr{d[,round(quantile(I_error, 0.25))]} - 
+    \Sexpr{d[,round(quantile(I_error, 0.75))]}] individuals.
 These errors were significantly higher when the true value of $\alpha$ was
 at least 1
-    (Wilcoxon rank-sum test, $p$ = %\S%expr{latexSN(round(alpha.I.test$p.value, 4))}$)
+    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(alpha.I.test[["p.value"]], 4))}$)
 and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
-did not affect the estimates of $I$ (one-way ANOVA).
-
-The $m$ parameter was estimated correctly in only
-    %\S%expr{as.integer(m.tbl[1] * 100)} \%
+did not affect the estimates of $I$ (one-way ANOVA).}
+
+\del{The $m$ parameter was estimated correctly in only
+    \Sexpr{as.integer(m.tbl[1] * 100)} \%
 of simulations, barely better than random guessing. The true values of the
@@
 over-estimated by about a factor of two
-    (error %\S%expr{d[,round(median(N_error))]} 
-    [%\S%expr{d[,round(quantile(N_error, 0.25))]} - 
-     %\S%expr{d[,round(quantile(N_error, 0.75))]}] individuals).
-No parameters influenced the accuracy of the $N$ estimates (all one-way ANOVA).
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics[width=\textwidth]{abc-point-estimate-m2}
-  \vspace{6pt}
-  \caption{
-    Point estimates of BA model parameters obtained by running ABC
-    on simulated phylogenies without training, for simulations with $m = 2$.
-    Dotted lines indicate true values, and limits of the $y$-axes are regions
-    of uniform prior density. (A) Estimates for $\alpha$ and $I$ against their
-    true values in simulations. (B) Estimates for $m$ and $N$, which were held
-    fixed in these simulations, against true values of $\alpha$.
-  }
-  \label{fig:abcpt}
-\end{figure}
-
-\begin{table*}[ht]
-  \centering
-  \input{\tablepath/abc-hpd.tex}
-  \caption{
-      Average maximum \textit{a posteriori} point estimates and 95\% highest
-      posterior density (HPD) interval widths for BA model parameter estimates
-      obtained with ABC. Three transmission trees were simulated under
-      each combination of the listed parameter values, and the parameters were
-      estimated with ABC without training.
-  }
-  \label{tab:abchpd}
-\end{table*}
-
-<<supp_labels, echo=FALSE, results="asis", eval=FALSE>>=
-    N <- 5000
-    replicate <- 0
-    lab <- NULL
-    for (m in c(2, 3, 4)) {
-    for (I in c(1000, 2000)) {
-    for (alpha in c(0, 0.5, 1, 1.5)) {
-        lab <- c(lab, sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate))
-    }
-    }
-    }
-    lab <- paste(lab, collapse = ",")
+    (error \Sexpr{d[,round(median(N_error))]} 
+    [\Sexpr{d[,round(quantile(N_error, 0.25))]} - 
+     \Sexpr{d[,round(quantile(N_error, 0.75))]}] individuals).
+No parameters influenced the accuracy of the $N$ estimates (all one-way ANOVA).}
+
+\add{Across all simulations, the median [IQR] absolute errors of the parameter 
+estimates obtained with \textit{netabc} were
+    \Sexpr{d[,round(median(alpha_error), 2)]} 
+    [\Sexpr{d[,round(quantile(alpha_error, 0.25), 2)]} - 
+    \Sexpr{d[,round(quantile(alpha_error, 0.75), 2)]}]
+for $\alpha$,
+    \Sexpr{d[,round(median(I_error))]} 
+    [\Sexpr{d[,round(quantile(I_error, 0.25))]} - 
+    \Sexpr{d[,round(quantile(I_error, 0.75))]}]
+for $I$,
+    \Sexpr{d[,round(median(m_error), 2)]} 
+    [\Sexpr{d[,round(quantile(m_error, 0.25), 2)]} - 
+    \Sexpr{d[,round(quantile(m_error, 0.75), 2)]}]
+for $m$, and
+    \Sexpr{d[,round(median(N_error))]} 
+    [\Sexpr{d[,round(quantile(N_error, 0.25))]} - 
+    \Sexpr{d[,round(quantile(N_error, 0.75))]}]
+for $N$. These errors comprised, respectively,
+    \Sexpr{d[,round(median(alpha_error) * 100 / 2)]}\%,
+    \Sexpr{d[,round(median(I_error) * 100 / 4500)]}\%,
+    \Sexpr{d[,round(median(m_error) * 100 / 6)]}\%, and
+    \Sexpr{d[,round(median(N_error) * 100 / 14500)]}\%
+of the regions of nonzero prior density. For $I$ and $N$, relative errors were
+    \Sexpr{d[,round(median(I_error * 100 / true_I))]}\%
+    [\Sexpr{d[,round(quantile(I_error * 100 / true_I, 0.25))]} - 
+    \Sexpr{d[,round(quantile(I_error * 100 / true_I, 0.75))]}\%]
+and
+    \Sexpr{d[,round(median(N_error * 100 / true_N))]}\%
+    [\Sexpr{d[,round(quantile(N_error * 100 / true_N, 0.25))]} - 
+    \Sexpr{d[,round(quantile(N_error * 100 / true_N, 0.75))]}\%].
+Average 95\% HPD interval widths were
+    \Sexpr{d[,round(mean(alpha.upper.95 - alpha.lower.95), 2)]},
+    \Sexpr{d[,round(mean(I.upper.95 - I.lower.95))]},
+    \Sexpr{d[,round(mean(m.upper.95 - m.lower.95), 2)]}, and
+    \Sexpr{d[,format(round(mean(N.upper.95 - N.lower.95)), digits=5)]},
+representing 
+    \Sexpr{d[,round(mean(alpha.upper.95 - alpha.lower.95) * 100 / 2)]}\%,
+    \Sexpr{d[,round(mean(I.upper.95 - I.lower.95) * 100 / 4500)]}\%,
+    \Sexpr{d[,round(mean(m.upper.95 - m.lower.95) * 100 / 6)]}\%, and
+    \Sexpr{d[,round(mean(N.upper.95 - N.lower.95) * 100 / 14500)]}\%
+of the nonzero prior density regions. Point estimates of $I$ were upwardly
+biased: $I$ was overestimated in
+    \Sexpr{d[,sum(I > true_I)]}
+out of 
+    \Sexpr{nrow(d)}
+simulations
+    (\Sexpr{round(d[,sum(I > true_I)] * 100 / nrow(d))}\%).
+The estimates for $m$ and $N$ were similar across all simulations (median [IQR]
+point estimates
+    \Sexpr{d[,round(median(m), 2)]}
+    [\Sexpr{d[,round(quantile(m, 0.25), 2)]} - 
+    \Sexpr{d[,round(quantile(m, 0.75), 2)]}]
+and 
+    \Sexpr{d[,round(median(N))]}
+    [\Sexpr{d[,round(quantile(N, 0.25))]} - 
+    \Sexpr{d[,round(quantile(N, 0.75))]}])
+regardless of the true values of any of the BA parameters.}
+
+\add{To analyze the effects of the true parameter values on the
+accuracy our estimates of $\alpha$ and $I$, we fitted one GLM for each of these
+two parameters, with error rate as the dependent variable and the true
+parameter values as independent variables. Since the estimates of $m$ and
+$N$ were roughly equal across all simulations
+(\cref{fig:abcptm2,fig:abcptm3,fig:abcptm4}), GLMs were not fitted for these
+parameters. The estimated coefficients are shown in \cref{tab:glm}.}
+
+<<abc_glm, include=FALSE>>=
+    source("../thesis/global.R")
+    stopifnot(alpha.glm[Parameter == "alpha",p] < 0.05)
+    stopifnot(alpha.glm[Parameter == "I",p] > 0.05)
+    stopifnot(alpha.glm[Parameter == "m",p] < 0.05)
+
+    stopifnot(I.glm[Parameter == "alpha",p] < 0.05)
+    stopifnot(I.glm[Parameter == "I",p] < 0.05)
+    stopifnot(I.glm[Parameter == "m",p] > 0.05)
 @
 
+<<glm_table, echo=FALSE, warning=FALSE, error=FALSE, results="asis">>=
+    format.glm <- function (df) {
+        df.copy <- copy(df)
+        df.copy[Parameter == "alpha", Parameter := "\\alpha"]
+        df.copy[Parameter != "(Intercept)", Parameter := paste0("$", Parameter, "$")]
+        df.copy[,p := paste0("$", sapply(p, pp, eq=FALSE), "$")]
+        df.copy[,Estimate := paste0("$", sapply(Estimate, pp, eq=FALSE, thres=NA), "$")]
+        df.copy[,`Standard error` := paste0("$", sapply(`Standard error`, pp, eq=FALSE, thres=NA), "$")]
+        setnames(df.copy, "p", "$p$-value")
+        df.copy
+    }
+
+    glm.tbl <- rbind(format.glm(alpha.glm), format.glm(I.glm))
+    glm.tbl <- cbind(`Dependent variable`=c("$\\alpha$", "", "", "", "$I$", "", "", ""),
+                     glm.tbl)
+    setnames(glm.tbl, "Parameter", "Independent variable")
+
+    capt.short <- paste("Parameters of fitted GLMs relating error in estimated",
+                        "$\\alpha$ and $I$ to true values of BA parameters.")
+    capt <- paste(capt.short, "GLMs ere fitted with the Gaussian distribution and",
+                  "inverse link function. Coefficients are interpretable as",
+                  "additive effects on the inverse of the mean error.")
+    print(xtable(glm.tbl, label="tab:glm", caption=c(capt, capt.short)),
+          include.rownames=FALSE, sanitize.text.function=identity,
+          sanitize.colnames.function=identity)
+@
+
+\add{The GLM analysis indicated that the error in estimates of $\alpha$
+decreased with larger true values of $\alpha$
+    ($p \Sexpr{pp(alpha.glm[Parameter == "alpha", p], eq=TRUE)}$)
+and $m$
+    ($p \Sexpr{pp(alpha.glm[Parameter == "m", p], eq=TRUE)}$)
+but was not significantly affected by $I$. Qualitatively, $\alpha$ seemed to be
+only weakly identifiable between the values of 0 and 0.5 (\cref{fig:abcptm2}).
+The error in the estimated prevalence $I$ was slightly lower for smaller values
+of $\alpha$
+    ($p \Sexpr{pp(I.glm[Parameter == "alpha", p], eq=TRUE)}$)
+and $I$
+    ($p \Sexpr{pp(I.glm[Parameter == "I", p], eq=TRUE)}$),
+but was not significantly affected by the true value of $m$.}
+
+<<marginals, include=FALSE>>=
+    q <- 0.42
+    middle <- d[alpha_error > quantile(alpha_error, q) & 
+                alpha_error < quantile(alpha_error, 1-q) & 
+                I_error > quantile(I_error, q) & 
+                I_error < quantile(I_error, 1-q),]
+    middle.plot <- middle[1,sprintf("%.1f_%.0f_%.0f_%.0f_%.0f", true_alpha, true_I, true_m, true_N, replicate)]
+@
+
 The dispersion of the ABC approximation to the posterior also varied between
-the parameters, with narrower HPD intervals for the parameters with more
-accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows
-the distributions for for one simulation. Equivalent plots for one replicate
-simulation with each studied parameter combination can be found in
-\cref{%\S%expr{lab}}. HPD intervals around $\alpha$ and $I$ were often narrow
-relative to the region of nonzero prior density, whereas the intervals for $m$
-and $N$ were more widely dispersed.
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{abc-posterior-example}
-  \vspace{6pt}
-  \caption{
-    Marginal posterior distributions of BA model parameters estimated
-    with ABC for a single simulated transmission tree. Dotted
-    lines and shaded polygon indicate true values.
-  }
-  \label{fig:abcex}
-\end{figure}
-
-<<mixed_peerdriven, include=FALSE, eval=FALSE>>=
+the parameters\del{, with narrower HPD intervals for the parameters with more
+accurate point estimates} (\cref{tab:abchpd}). HPD intervals around $\alpha$
+and $I$ were often narrow relative to the region of nonzero prior
+density, whereas the intervals for $m$ and $N$ were more widely dispersed.
+\Cref{fig:abc1,fig:abc2} \del{shows the distributions for one
+simulation.} \add{show one- and two-dimensional marginal distributions for a
+simulation with $\alpha$ and $I$ errors close to their respective medians. In
+particular, the simulation shown is one of
+\Sexpr{replace_number(nrow(middle))} with errors in both $\alpha$ and $I$
+between the \Sexpr{round(q*100)}nd and \Sexpr{round((1-q)*100)}th percentiles.
+The parameters for this simulation were
+    $\alpha = \Sexpr{middle[1,true_alpha]}$,
+    $I = \Sexpr{middle[1,true_I]}$,
+    $m = \Sexpr{middle[1,true_m]}$, and
+    $N = \Sexpr{middle[1,true_N]}$.
+The two-dimensional marginals indicate some dependence between pairs of
+parameters, particularly $I$ and $N$ which show a diagonally shaped region of
+high posterior density.}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior", middle.plot)}}.pdf}
+    \caption[
+        One-dimensional marginal posterior distributions of BA model
+        parameters estimated by \textit{netabc} from a simulated transmission
+        tree.
+    ]{
+        One-dimensional marginal posterior distributions of BA model parameters
+        estimated by \textit{netabc} from a simulated transmission tree. Dashed
+        lines indicate true values, solid lines indicate posterior means, and
+        shaded areas show 95\% highest posterior density intervals.
+    }
+    \label{fig:abc1}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior-2d", middle.plot)}}.pdf}
+    \caption[
+        Two-dimensional marginal posterior distributions of BA model parameters
+        estimated by \textit{netabc} from a simulated transmission tree.
+    ]{
+        Two-dimensional marginal posterior distributions of BA model parameters
+        estimated by \textit{netabc} from a simulated transmission tree. White
+        circles indicate true values, magenta diamonds
+        indicate posterior means. 
+    }
+    \label{fig:abc2}
+\end{figure}
+
+<<mixed_peerdriven, include=FALSE>>=
   library(coda)
@@
   d <- d[iter == max(iter)]
-  d <- d[sample(1:nrow(d), prob=weight, replace=TRUE)]
   params <- c("alpha", "m", "N", "I")
   d <- melt(d, measure.vars=params, variable.name="parameter")
-  f <- function (x) {
-    dens <- density(x)
-    hpd <- HPDinterval(mcmc(x))
-    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
-         hpd.max=hpd[,"upper"])
-  }
-  d <- d[,f(value), by=parameter]
+  f <- function (x, wt) {
+    hpd <- wtd.hpd(x, wt/sum(wt))
+    list(pmean=wtd.mean(x, wt, normwt=TRUE), 
+         hpd.min=hpd[1], 
+         hpd.max=hpd[2])
+  }
+  d <- d[,f(value, weight), by=parameter]
   setkey(d, parameter)
@@
   p <- p[iter == max(iter)]
-  p <- p[sample(1:nrow(p), prob=weight, replace=TRUE)]
   params <- c("alpha", "m", "N", "I")
   p <- melt(p, measure.vars=params, variable.name="parameter")
-  f <- function (x) {
-    dens <- density(x)
-    hpd <- HPDinterval(mcmc(x))
-    list(map=dens$x[which.max(dens$y)], hpd.min=hpd[,"lower"], 
-         hpd.max=hpd[,"upper"])
-  }
-  p <- p[,f(value), by=parameter]
+  p <- p[,f(value, weight), by=parameter]
   setkey(p, parameter)
@@
 the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
-other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
-HPD] estimates for each parameter were: 
+other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The posterior mean
+[95\% HPD] estimates for each parameter were: 
 $\alpha$, 
-  %\S%expr{round(d["alpha", map], 2)} 
-  [%\S%expr{round(d["alpha", hpd.min], 2)} -
-   %\S%expr{round(d["alpha", hpd.max], 2)}];
+  \Sexpr{round(d["alpha", pmean], 2)} 
+  [\Sexpr{round(d["alpha", hpd.min], 2)} -
+   \Sexpr{round(d["alpha", hpd.max], 2)}];
 $I$,
-  %\S%expr{d["I", round(map)]} 
-  [%\S%expr{d["I", round(hpd.min)]} -
-   %\S%expr{d["I", round(hpd.max)]}];
+  \Sexpr{d["I", round(pmean)]} 
+  [\Sexpr{d["I", round(hpd.min)]} -
+   \Sexpr{d["I", round(hpd.max)]}];
 $m$,
-  %\S%expr{d["m", floor(map)]} 
-  [%\S%expr{d["m", floor(hpd.min)]} -
-   %\S%expr{d["m", floor(hpd.max)]}];
+  \Sexpr{d["m", floor(pmean)]} 
+  [\Sexpr{d["m", floor(hpd.min)]} -
+   \Sexpr{d["m", floor(hpd.max)]}];
 $N$,
-  %\S%expr{d["N", round(map)]} 
-  [%\S%expr{d["N", round(hpd.min)]}-
-   %\S%expr{d["N", round(hpd.max)]}].
-The approximate posterior distributions for this simulation are shown in
-\cref{fig:mixed}. To test the effect of sampling bias, we sampled one
-transmission tree in a peer-driven fashion, where the probability to sample a
-node was twice as high if one of its peers had already been sampled. The
-parameters for this experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and
-$I$ = 2000. The estimated values were
-$\alpha$, 
-  %\S%expr{p["alpha", round(map, 2)]} 
-  [%\S%expr{p["alpha", round(hpd.min, 2)]} -
-   %\S%expr{p["alpha", round(hpd.max, 2)]}];
+  \Sexpr{d["N", round(pmean)]} 
+  [\Sexpr{d["N", round(hpd.min)]}-
+   \Sexpr{d["N", round(hpd.max)]}].
+The approximate one-dimensional marginal posterior distributions for this
+simulation are shown in \cref{fig:mixed}. To test the effect of sampling bias,
+we sampled one transmission tree in a peer-driven fashion, where the
+probability to sample a node was twice as high if one of its peers had already
+been sampled. The parameters for this experiment were $N$ = 5000, $m$ = 2,
+$\alpha$ = 0.5, and $I$ = 2000. The estimated values were $\alpha$, 
+  \Sexpr{p["alpha", round(pmean, 2)]} 
+  [\Sexpr{p["alpha", round(hpd.min, 2)]} -
+   \Sexpr{p["alpha", round(hpd.max, 2)]}];
 $I$,
-  %\S%expr{p["I", floor(map)]} 
-  [%\S%expr{p["I", floor(hpd.min)]} -
-   %\S%expr{p["I", floor(hpd.max)]}];
+  \Sexpr{p["I", floor(pmean)]} 
+  [\Sexpr{p["I", floor(hpd.min)]} -
+   \Sexpr{p["I", floor(hpd.max)]}];
 $m$,
-  %\S%expr{p["m", floor(map)]} 
-  [%\S%expr{p["m", floor(hpd.min)]} -
-   %\S%expr{p["m", floor(hpd.max)]}];
+  \Sexpr{p["m", floor(pmean)]} 
+  [\Sexpr{p["m", floor(hpd.min)]} -
+   \Sexpr{p["m", floor(hpd.max)]}];
 $N$,
-  %\S%expr{p["N", floor(map)]} 
-  [%\S%expr{p["N", floor(hpd.min)]} -
-   %\S%expr{p["N", floor(hpd.max)]}].
-The approximate posterior distributions are shown in \cref{fig:peerdriven}. Both
-of these results were in line with estimates obtained on other simulated
-datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
-$\alpha$ was somewhat lower than typical.
+  \Sexpr{p["N", floor(pmean)]} 
+  [\Sexpr{p["N", floor(hpd.min)]} -
+   \Sexpr{p["N", floor(hpd.max)]}].
+The approximate one-dimensional marginal posterior distributions are shown in
+\cref{fig:peerdriven}. Both of these results were in line with estimates
+obtained on other simulated datasets (\cref{tab:abchpd}), although the estimate
+of peer-driven sampling for $\alpha$ was somewhat lower than typical.
 
@@
 
-<<realdata, include=FALSE, eval=FALSE>>=
+<<realdata, include=FALSE>>=
   m <- collect.metadata("../../simulations/aggregates/hpd/*")
-  d1 <- fread(rownames(m)[m$m_min == 1])
-  d2 <- fread(rownames(m)[m$m_min == 2])
-
-  map1 <- dcast.data.table(d1, dataset~parameter, value.var="map")
-  up1 <- dcast.data.table(d1, dataset~parameter, value.var="hpd.max")
-  low1 <- dcast.data.table(d1, dataset~parameter, value.var="hpd.min")
-  setkey(map1, dataset)
+  d1 <- fread(rownames(m)[m$m_min == 1 & m$agg_gene == "gag"])
+  d2 <- fread(rownames(m)[m$m_min == 2 & m$agg_gene == "gag"])
+
+  d1.env <- fread(rownames(m)[m$m_min == 1 & m$agg_gene == "env"])
+  d2.env <- fread(rownames(m)[m$m_min == 2 & m$agg_gene == "env"])
+
+  pmean1 <- dcast.data.table(d1, dataset~parameter, value.var="mean")
+  up1 <- dcast.data.table(d1, dataset~parameter, value.var="hpd.max.95")
+  low1 <- dcast.data.table(d1, dataset~parameter, value.var="hpd.min.95")
+  setkey(pmean1, dataset)
   setkey(up1, dataset)
@@
 
-  map2 <- dcast.data.table(d2, dataset~parameter, value.var="map")
-  up2 <- dcast.data.table(d2, dataset~parameter, value.var="hpd.max")
-  low2 <- dcast.data.table(d2, dataset~parameter, value.var="hpd.min")
-  setkey(map2, dataset)
+  pmean2 <- dcast.data.table(d2, dataset~parameter, value.var="mean")
+  up2 <- dcast.data.table(d2, dataset~parameter, value.var="hpd.max.95")
+  low2 <- dcast.data.table(d2, dataset~parameter, value.var="hpd.min.95")
+  setkey(pmean2, dataset)
   setkey(up2, dataset)
   setkey(low2, dataset)
+
+  pmean1.env <- dcast.data.table(d1.env, dataset~parameter, value.var="mean")
+  up1.env <- dcast.data.table(d1.env, dataset~parameter, value.var="hpd.max.95")
+  low1.env <- dcast.data.table(d1.env, dataset~parameter, value.var="hpd.min.95")
+  setkey(pmean1.env, dataset)
+  setkey(up1.env, dataset)
+  setkey(low1.env, dataset)
+
+  pmean2.env <- dcast.data.table(d2.env, dataset~parameter, value.var="mean")
+  up2.env <- dcast.data.table(d2.env, dataset~parameter, value.var="hpd.max.95")
+  low2.env <- dcast.data.table(d2.env, dataset~parameter, value.var="hpd.min.95")
+  setkey(pmean2.env, dataset)
+  setkey(up2.env, dataset)
+  setkey(low2.env, dataset)
+
+  setkey(d1, dataset)
+  setkey(d2, dataset)
+  setkey(d1.env, dataset)
+  setkey(d2.env, dataset)
+
+  pmean1 <- unique(d1)[,.(dataset, ntip)][pmean1]
+  pmean2 <- unique(d2)[,.(dataset, ntip)][pmean2]
+  pmean1.env <- unique(d1.env)[,.(dataset, ntip)][pmean1.env]
+  pmean2.env <- unique(d2.env)[,.(dataset, ntip)][pmean2.env]
+
+  stopifnot(pmean1[,dataset[which.max(alpha)]] == "zetterberg2004")
+  stopifnot(pmean1[,dataset[which.min(alpha)]] == "cuevas2009")
+
+  stopifnot(pmean1[,dataset[which.min(I)]] == "zetterberg2004")
+  stopifnot(pmean1[,dataset[which.max(I)]] == "novitsky2014")
+  stopifnot(pmean1[dataset != "novitsky2014", dataset[which.max(I)]] == "bctree")
+
+  stopifnot(pmean1[,cor.test(ntip, I, method="spearman")$p.value] > 0.05)
 @
 
-We applied ABC to five published HIV datasets (\cref{tab:data}),
+\del{We applied ABC to five published HIV datasets (\cref{tab:data}),
 and found substantial heterogeneity among the parameter estimates
-(\cref{fig:abchpd,fig:abchpdm2}). Plots of the marginal posterior distributions
-for each dataset are shown in
-\cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.
-Two of the datasets~\autocite{niculescu2015recent, wang2015targeting} had
+(\cref{fig:abchpd,fig:abchpdm2}).} \add{Posterior mean point estimates and 50\%
+and 95\% HPD intervals for each parameter are shown in \cref{fig:abchpd}.}
+\del{Plots of the marginal posterior distributions for each dataset are shown
+in \cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.}
+\add{\Cref{fig:abchpdm2} shows point estimates and HPD intervals obtained when
+the value $m = 1$ was disallowed by the prior. Since the results indicated that
+$m = 1$ was the most credible value for several datasets, all results discussed
+henceforth are for the prior $m \sim \text{DiscreteUniform}(1, 5)$ unless
+otherwise stated.}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{realdata-hpd}
+    \caption[
+        Posterior means and 50\%/95\% HPD intervals for parameters of the
+        BA network model, fitted to ten HIV datasets with \textit{netabc}.
+    ]{
+        Posterior means (points), 50\% HPD intervals (notches), and 95\%
+        HPD intervals (lines) for parameters of the BA network model, fitted to
+        ten HIV datasets with \textit{netabc}. Legend labels indicate risk
+        group and country of origin. Abbreviations: IDU, injection drug users;
+        MSM, men who have sex with men; HET, heterosexual. Note that posterior
+        means can fall outside of the HPD interval if the distribution is
+        diffuse.
+    }
+    \label{fig:abchpd}
+\end{figure}
+
+\del{Two of the datasets~\autocite{niculescu2015recent, wang2015targeting} had
 estimated $\alpha$ values near unity for the prior allowing $m = 1$ (MAP
 estimates [95\% HPD] 
-  %\S%expr{map1["niculescu2015", round(alpha, 2)]} 
-  [%\S%expr{low1["niculescu2015", round(alpha, 2)]} - 
-   %\S%expr{up1["niculescu2015", round(alpha, 2)]}]
+  \Sexpr{pmean1["niculescu2015", round(alpha, 2)]} 
+  [\Sexpr{low1["niculescu2015", round(alpha, 2)]} - 
+   \Sexpr{up1["niculescu2015", round(alpha, 2)]}]
 and
-  %\S%expr{map1["wang2015", round(alpha, 2)]} 
-  [%\S%expr{low1["wang2015", round(alpha, 2)]} -
-   %\S%expr{up1["wang2015", round(alpha, 2)]}] respectively).
+  \Sexpr{pmean1["wang2015", round(alpha, 2)]} 
+  [\Sexpr{low1["wang2015", round(alpha, 2)]} -
+   \Sexpr{up1["wang2015", round(alpha, 2)]}] respectively).
 The MAP estimates did not change appreciably when $m = 1$ was disallowed by the
@@
 data was narrower
-  (%\S%expr{low1["niculescu2015", round(alpha, 2)]} - 
-   %\S%expr{up1["niculescu2015", round(alpha, 2)]}).
+  (\Sexpr{low1["niculescu2015", round(alpha, 2)]} - 
+   \Sexpr{up1["niculescu2015", round(alpha, 2)]}).
 When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
 low estimated $\alpha$ values
-  (%\S%expr{map1["li2015", round(alpha, 2)]} 
-  [%\S%expr{low1["li2015", round(alpha, 2)]} - 
-  %\S%expr{up1["li2015", round(alpha, 2)]}]
+  (\Sexpr{pmean1["li2015", round(alpha, 2)]} 
+  [\Sexpr{low1["li2015", round(alpha, 2)]} - 
+  \Sexpr{up1["li2015", round(alpha, 2)]}]
 and
-  %\S%expr{map1["cuevas2009", round(alpha, 2)]} 
-  [%\S%expr{low1["cuevas2009", round(alpha, 2)]} -
-   %\S%expr{up1["cuevas2009", round(alpha, 2)]}]). 
+  \Sexpr{pmean1["cuevas2009", round(alpha, 2)]} 
+  [\Sexpr{low1["cuevas2009", round(alpha, 2)]} -
+   \Sexpr{up1["cuevas2009", round(alpha, 2)]}]). 
 However, the MAP estimates increased when $m = 1$ was not permitted, although
 the HPD intervals remained roughly the same
-  (%\S%expr{map2["li2015", round(alpha, 2)]} 
-  [%\S%expr{low2["li2015", round(alpha, 2)]} - 
-  %\S%expr{up2["li2015", round(alpha, 2)]}]
+  (\Sexpr{pmean2["li2015", round(alpha, 2)]} 
+  [\Sexpr{low2["li2015", round(alpha, 2)]} - 
+  \Sexpr{up2["li2015", round(alpha, 2)]}]
 and
-  %\S%expr{map2["cuevas2009", round(alpha, 2)]} 
-  [%\S%expr{low2["cuevas2009", round(alpha, 2)]} -
-   %\S%expr{up2["cuevas2009", round(alpha, 2)]}]).
+  \Sexpr{pmean2["cuevas2009", round(alpha, 2)]} 
+  [\Sexpr{low2["cuevas2009", round(alpha, 2)]} -
+   \Sexpr{up2["cuevas2009", round(alpha, 2)]}]).
 The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
 for both priors on $m$
-  (%\S%expr{map1["novitsky2014", round(alpha, 2)]} for $m \geq 1$;
-   %\S%expr{map2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).
+  (\Sexpr{pmean1["novitsky2014", round(alpha, 2)]} for $m \geq 1$;
+   \Sexpr{pmean2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).
 However, the confidence interval was much wider when $m = 1$ was allowed
-  ([%\S%expr{low1["novitsky2014", round(alpha, 2)]} -
-    %\S%expr{up1["novitsky2014", round(alpha, 2)]}] for $m \geq 1$ vs.
-    %\S%expr{low2["novitsky2014", round(alpha, 2)]} -
-    %\S%expr{up2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).
-
-For all the datasets except \citeauthor{novitsky2014impact}, estimated values
+  ([\Sexpr{low1["novitsky2014", round(alpha, 2)]} -
+    \Sexpr{up1["novitsky2014", round(alpha, 2)]}] for $m \geq 1$ vs.
+    \Sexpr{low2["novitsky2014", round(alpha, 2)]} -
+    \Sexpr{up2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).}
+
+\end{document}
+
+\add{Posterior mean point estimates for the preferential attachment power $\alpha$
+were all sub-linear, ranging from 
+    \Sexpr{pmean1["zetterberg2004", round(alpha, 2)]}
+for the IDU/Estonia data to
+    \Sexpr{pmean1["cuevas2009", round(alpha, 2)]}
+for the mixed/Spain data. When aggregated by risk group, the average estimates
+were
+    \Sexpr{pmean1[c("zetterberg2004", "niculescu2015", "bctree"), round(mean(alpha), 2)]}
+for IDU,
+    \Sexpr{pmean1[c("novitsky2014", "mccormack2002", "grabowski2014"), round(mean(alpha), 2)]}
+for primarily heterosexual risk, and
+    \Sexpr{pmean1[c("wang2015", "kao2011", "little2014", "li2015"), round(mean(alpha), 2)]}
+for MSM. These values were obtained with \textit{gag} for the datasets where
+both \textit{gag} and \textit{env} were sequenced, but the estimates for
+$\alpha$ did not change appreciably between the two genes (\cref{fig:genes}).
+There was a large amount of uncertainty associated with these estimates. 95\%
+HPD intervals were very wide for most datasets, often encompassing almost the
+entire range from 0 to 1 (\cref{fig:abchpd}). For all the datasets except
+HET/Malawi and MSM/Beijing, the posterior mean was either outside, or very
+close to the border of, the 50\% HPD intervals. This indicates that the
+posterior distributions were diffuse with heavy tails, rather than having most
+of their mass around the mode. When the value $m = 1$ was disallowed by the
+prior, the separation between the IDU datasets and the others became more
+striking (\cref{fig:abchpdm2}). All three IDU datasets had estimated $\alpha$
+values at or above
+    \Sexpr{pmean2[c("zetterberg2004", "niculescu2015", "bctree"), round(min(alpha), 2)]}.
+The estimate for the MSM/Beijing data was slightly lower
+    (\Sexpr{pmean2["wang2015", round(alpha, 2)]}) and
+the estimates for the seven remaining non-IDU datasets were bounded above by
+    \Sexpr{pmean2[c("novitsky2014", "mccormack2002", "grabowski2014", "kao2011", "little2014", "li2015", "cuevas2009"), round(max(alpha), 2)]}.}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{realdata-genes}
+    \caption[
+        Posterior means and 50\%/95\% HPD intervals for parameters of the
+        BA network model, fitted to three HIV datasets where both \textit{gag}
+        and \textit{env} genes were sequenced.
+    ]{
+        Posterior means (points), 50\% HPD intervals (notches), and 95\%
+        HPD intervals (lines) for parameters of the BA network model, fitted to
+        three HIV datasets where both \textit{gag} and \textit{env} genes were
+        sequenced.
+    }
+    \label{fig:genes}
+\end{figure}
+
+\del{For all the datasets except \citeauthor{novitsky2014impact}, estimated values
 of $I$ were below 2000 when $m = 1$ was allowed, with relatively narrow HPD
 intervals compared to the nonzero prior density region
-  (\citeauthor{cuevas2009hiv}, %\S%expr{map1["cuevas2009", floor(I)]} 
-  [%\S%expr{low1["cuevas2009", floor(I)]} -
-   %\S%expr{up1["cuevas2009", floor(I)]}];
-   \citeauthor{niculescu2015recent}, %\S%expr{map1["niculescu2015", floor(I)]}
-  [%\S%expr{low1["niculescu2015", floor(I)]} - 
-   %\S%expr{up1["niculescu2015", floor(I)]}];
-  \citeauthor{li2015hiv}, %\S%expr{map1["li2015", floor(I)]} 
-  [%\S%expr{low1["li2015", floor(I)]} -
-   %\S%expr{up1["li2015", floor(I)]}];
-   \citeauthor{wang2015targeting}, %\S%expr{map1["wang2015", floor(I)]}
-  [%\S%expr{low1["wang2015", floor(I)]} - 
-   %\S%expr{up1["wang2015", floor(I)]}]).
+  (\citeauthor{cuevas2009hiv}, \Sexpr{pmean1["cuevas2009", floor(I)]} 
+  [\Sexpr{low1["cuevas2009", floor(I)]} -
+   \Sexpr{up1["cuevas2009", floor(I)]}];
+   \citeauthor{niculescu2015recent}, \Sexpr{pmean1["niculescu2015", floor(I)]}
+  [\Sexpr{low1["niculescu2015", floor(I)]} - 
+   \Sexpr{up1["niculescu2015", floor(I)]}];
+  \citeauthor{li2015hiv}, \Sexpr{pmean1["li2015", floor(I)]} 
+  [\Sexpr{low1["li2015", floor(I)]} -
+   \Sexpr{up1["li2015", floor(I)]}];
+   \citeauthor{wang2015targeting}, \Sexpr{pmean1["wang2015", floor(I)]}
+  [\Sexpr{low1["wang2015", floor(I)]} - 
+   \Sexpr{up1["wang2015", floor(I)]}]).
 The \citeauthor{novitsky2014impact} data was the outlier, with a very high
 estimated $I$, and HPD interval spanning almost the entire prior region
-  (%\S%expr{map1["novitsky2014", floor(I)]} 
-  [%\S%expr{low1["novitsky2014", floor(I)]} -
-   %\S%expr{up1["novitsky2014", floor(I)]}]).
+  (\Sexpr{pmean1["novitsky2014", floor(I)]} 
+  [\Sexpr{low1["novitsky2014", floor(I)]} -
+   \Sexpr{up1["novitsky2014", floor(I)]}]).
 The $I$ estimates and HPD intervals were generally robust to the choice of
 prior on $m$, with slightly narrower HPD intervals (compare
-\cref{fig:abchpd,fig:abchpdm2}).
-
-The MAP estimate of $m$ was equal to 1 for all but the
+\cref{fig:abchpd,fig:abchpdm2}).}
+
+<<genetrees, include=FALSE>>=
+    datasets <- c("grabowski2014", "zetterberg2004", "mccormack2002")
+    paths <- paste0("../../simulations/", datasets, "/subtree/*")
+    paths <- sapply(paths, Sys.glob)
+    trees <- lapply(paths, read.tree)
+    get.ratio <- function (t) {
+        int <- t$edge.length[t$edge[,2] > Ntip(t)]
+        tip <- t$edge.length[t$edge[,2] <= Ntip(t)]
+        mean(int) / mean(tip)
+    }
+    ratios <- data.table(dataset=rep(datasets, each=2),
+                         gene=rep(c("env", "gag"), 3),
+                         ratio=sapply(trees, get.ratio))
+    stopifnot(ratios[gene == "env", mean(ratio)] <
+              ratios[gene == "gag", mean(ratio)])
+@
+
+\add{For all but the HET/Botswana data, the posterior mean estimates for
+the prevalence $I$ were between
+    \Sexpr{pmean1["zetterberg2004", round(I)]}
+(IDU/Estonia) and
+    \Sexpr{pmean1["bctree", round(I)]}
+(IDU/Canada). The HET/Botswana data had a much higher estimated $I$ value
+    (\Sexpr{pmean1["novitsky2014", round(I)]})
+than the other datasets, with a very wide 95\% HPD interval covering almost the
+entire prior region (\cref{fig:abchpd}). There was no significant correlation
+between the number of sequences in the tree and the estimated prevalence
+(Spearman correlation, 
+    $p \Sexpr{pp(pmean1[,cor.test(ntip, I, method="spearman")[["p.value"]]], eq=TRUE)}$),
+indicating that the higher prevalence estimates were not simply due to
+increased sampling density. When both \textit{gag} and \textit{env} sequences
+were analyzed, the estimates from the \textit{env} data were substantially
+higher (\cref{fig:genes}). The ratios of internal to terminal branch lengths
+were lower for the \textit{env} trees than the \textit{gag} trees (average 
+    \Sexpr{ratios[gene == "env", round(mean(ratio), 2)]}
+for \textit{env} and 
+    \Sexpr{ratios[gene == "gag", round(mean(ratio), 2)]}
+for \textit{gag}). As with $\alpha$, many of the posterior means were
+outside the 50\% HPD intervals, suggesting diffuse posterior
+distributions. The values of $I$ were fairly robust to restricting the
+prior to $m \geq 2$ (compare \cref{fig:abchpd,fig:abchpdm2}), although 
+the 95\% HPD intervals were slightly narrower (average width
+    \Sexpr{round(mean(up1[,I] - low1[,I]))}
+for $m \geq 1$ and
+    \Sexpr{round(mean(up2[,I] - low2[,I]))}
+for $m \geq 2$).}
+
+\del{The MAP estimate of $m$ was equal to 1 for all but the
 \citeauthor{novitsky2014impact} data, when this value was allowed. However, the
 upper bound of the HPD interval was different for each dataset
-  (\citeauthor{niculescu2015recent}, %\S%expr{up1["niculescu2015", round(m)]};
-   \citeauthor{wang2015targeting}, %\S%expr{up1["wang2015", round(m)]};
-   \citeauthor{li2015hiv}, %\S%expr{up1["li2015", round(m)]};
-   \citeauthor{cuevas2009hiv}, %\S%expr{up1["cuevas2009", round(m)]}).
+  (\citeauthor{niculescu2015recent}, \Sexpr{up1["niculescu2015", round(m)]};
+   \citeauthor{wang2015targeting}, \Sexpr{up1["wang2015", round(m)]};
+   \citeauthor{li2015hiv}, \Sexpr{up1["li2015", round(m)]};
+   \citeauthor{cuevas2009hiv}, \Sexpr{up1["cuevas2009", round(m)]}).
 When $m = 1$ was disallowed, the MAP for all datasets was either 2 or 3, with
@@
 data, for which the MAP estimate was lower 
-  (%\S%expr{map1["li2015", floor(N)]})
-when $m = 1$ was allowed.
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{realdata-hpd}
-  \vspace{8pt}
-  \caption{
-      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
-      parameters of the BA network model, fitted to five published HIV datasets
-      with ABC. $x$-axes indicate regions of nonzero prior density. In
-      particular, the prior on $m$ was DiscreteUniform(1, 5).
-  }
-  \label{fig:abchpd}
-\end{figure}
-
-\clearpage
+  (\Sexpr{pmean1["li2015", floor(N)]})
+when $m = 1$ was allowed.}
+
+\add{The posterior means of \gls{m} were equal to one for 
+  \Sexpr{replace_number(pmean1[,sum(m == 1)])}
+of the datasets analyzed. The widths of the 95\% \gls{HPD} intervals varied
+from 0 (all the mass on the estimated value) to 5 (the entire prior region).
+When the value $\gls{m} = 1$ was disallowed by the prior, the posterior means
+for all but the HET/Botswana data took on the value 3, with the \gls{HPD}
+intervals spanning the entire prior region. This is very similar to the results 
+observed for \gls{m} on simulated data (\cref{fig:abcptm2,tab:abchpd}), and
+suggests that \gls{m} is not identifiable from these data with this prior. On
+the other hand, the estimate and 95\% \gls{HPD} interval for the HET/Botswana
+data (4 [2-5]) was the same for both priors.}
+
+\add{Estimates of \gls{N} were  mostly uninformative, with very similar estimates
+for all datasets (mean
+    \Sexpr{pmean1[,round(mean(N))]},
+range
+    \Sexpr{pmean1[,round(min(N))]} - \Sexpr{pmean1[,round(max(N))]}).
+This was similar to the pattern observed for the synthetic data, where the
+posterior mean always fell around the upper two-thirds mark of the region
+allowed by the prior. The results did not change appreciably when the value
+$\gls{m} = 1$ was assigned zero prior probability.}
 
@@
 may have been influenced by this prior, which places more mass on low $I$
-values. However, the MAP estimate of $I$ was very high for the
+values. However, the estimate of $I$ was very high for the
 \textcite{novitsky2013phylogenetic,novitsky2014impact} data, suggesting that a
@@
 depending on whether or not $m = 1$ was allowed by the prior is further evidence
-of this potential misspecification. However, we note that for two of the
-datasets, the estimated values of $\alpha$ did not change much between priors,
-and the estimates of $I$ were robust to the choice of prior for all datasets
-studied. More sophisticated models, for example models incorporating
-heterogeneity in node behaviour, are likely to provide a better fit to these
-data.
-
-With respect to the preferential attachment power $\alpha$, the five datasets
-analysed fell into three categories (\cref{fig:abchpd}). First, we
+of this potential misspecification. However, we note that for \del{two}\add{the
+majority} of the datasets, the estimated values of $\alpha$ did not change much
+between priors, and the estimates of $I$ were robust to the choice of prior for
+all datasets studied (compare \cref{fig:abchpd,fig:abchpdm2}). More
+sophisticated models, for example models incorporating heterogeneity in node
+behaviour, are likely to provide a better fit to these data.
+
+<<disc, include=FALSE>>=
+    source("../thesis/global.R")
+    d <- setDT(collect.data("../../simulations/aggregates/hpd/*"))
+    d <- d[agg_gene == "gag"]
+    d1 <- d[m_min == 1]
+    d2 <- d[m_min == 2]
+@
+
+\del{With respect to the preferential attachment power $\alpha$, the five
+datasets analysed fell into three categories (\cref{fig:abchpd}). First, we
 estimated a preferential attachment power close to 1, indicating linear
@@
 connected network, relatively high mean degree, and preferential attachment
-dynamics, to hold.
-
-The remaining three datasets (\textcite{cuevas2009hiv, novitsky2014impact,
+dynamics, to hold.}
+
+\del{The remaining three datasets (\textcite{cuevas2009hiv, novitsky2014impact,
 li2015hiv}) had estimated values of $\alpha$ below 0.5 when $m = 1$ was
@@
 all the posterior density was assigned to $m = 1$ when this value was allowed,
-indicating that the network was more likely to have an acyclic tree structure.
+indicating that the network was more likely to have an acyclic tree structure.}
+
+\subsubsection{Preferential attachment power is sub-linear and higher for \gls{IDU}}
+
+\add{For all datasets we examined, the posterior mean estimates for \gls{alpha}
+were sub-linear, ranging from 
+    \Sexpr{d[m_min == 1 & parameter == "alpha", round(min(mean), 2)]}
+to 
+    \Sexpr{d[m_min == 1 & parameter == "alpha", round(max(mean), 2)]}.
+The sub-linearity is consistent with the results of
+\textcite{de2007preferential}, who developed a statistical inference method to
+estimate the parameters of a more sophisticated \acrlong{PA} model
+incorporating heterogeneous node behaviour. When used to analyze
+population-level longitudinal partner count data, they found \gls{alpha} values
+ranging from 0.26 to 0.62 depending on the gender and time period considered.
+Our estimates of \gls{alpha} for the IDU/Romania was above this range under
+both priors, as were the estimates for the MSM/Beijing data and the BC data
+when $\gls{m} = 1$ was disallowed by the prior. The dataset investigated by
+\textcite{de2007preferential} was derived from a survey of a random sample of
+the Norwegian population, whereas our investigation focused on datasets from
+known phylogenetic or geographic clusters of \gls{HIV} infected persons. It is
+therefore unsurprising that we detected stronger \acrlong{PA} dynamics in some
+cases. For instance, random sampling is much less likely to discover the
+high-degree nodes characterizing the tail of the degree distribution, simply
+because those individuals are rare in the general population. In addition, it
+is plausible that \gls{HIV}-positive individuals are more likely to be highly
+connected in their sexual networks, as the odds of acquiring \gls{HIV} increase
+with the number of unprotected sexual contacts.}
+
+\add{Both \textcite{de2007preferential} and the HET/Botswana data studied
+populations whose primary risk factor for \gls{HIV} infection was heterosexual
+contact. \citeauthor{de2007preferential} explicitly excluded reported
+homosexual contacts; \citeauthor{novitsky2014impact} did not, but noted that
+heterosexual contact is the primary mode of transmission in Botswana where the
+study was done. In the first of the two papers describing the Botswana
+study~\autocite{novitsky2013phylogenetic}, the authors noted that their sample
+was gender-biased, being composed of approximately 75\% women. Our estimate of
+\gls{alpha} for these data was 
+    \Sexpr{d[m_min == 1 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]}
+        or 
+    \Sexpr{d[m_min == 2 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]},
+depending on the prior on \gls{m}. Similarly, \textcite{de2007preferential}
+estimated 0.54, 0.57, and 0.29 for 3-year, 5-year, and lifetime partnership
+networks respectively for the female portion of their sample.}
+
+\add{For both choices of prior on \gls{m}, the datasets derived from \gls{IDU}
+populations had a higher estimated \acrlong{PA} power than the other datasets
+(\cref{fig:abchpd,fig:abchpdm2}). This finding is in line with
+\textcite{dombrowski2013topological}, who reanalyzed a network of \glspl{IDU}
+in Brooklyn, USA, collected between 1991 and
+1993~\autocite{friedman2006social}. They found that the \gls{IDU} network
+resembled a \gls{BA} network much more closely than other social and sexual 
+networks, and offered sociological explanations for the apparent \acrlong{PA}
+dynamics in this population. Importantly, from a public health perspective,
+the authors asserted that the removal of \emph{random} individuals from
+\gls{IDU} networks may have the paradoxical effect of increasing the network's
+epidemic susceptibility. When low-degree nodes are removed, as would occur
+during a police crackdown, their network neighbours may turn to well-known
+community members for advice or supplies, thus increasing the connectivity of
+these high-degree nodes.}
+
+\add{Unfortunately, the sub-linear region for \gls{alpha} identified by both
+\textcite{de2007preferential} and \software{netabc} is also the region of
+poorest identifiability (\cref{fig:gridest,fig:abcptm2}). This was reflected in
+the high level of uncertainty in the estimates, with most 95\% \gls{HPD}
+intervals covering the majority of the range [0, 1]. The value $\gls{alpha} =
+0.5$ was contained in the 95\% \gls{HPD} interval for every dataset;
+consequently, it is not possible to say with high confidence that any of the
+\gls{alpha} values are different from each other. In synthetic data, the
+confidence intervals around \gls{alpha} narrowed when other parameters were
+marginalized out (\cref{fig:marginal}). Thus, it is possible that estimates of
+\gls{alpha} could be made more precise by specifying either exact values or
+informative priors on the other \gls{BA} parameters. We argued informally in
+\cref{subsec:whyba} that \gls{I}, \gls{m}, and \gls{N} can sometimes be
+estimated by standard epidemiological methods, although the difficulty may
+depend on the specific epidemic under consideration.}
+
+\add{I is underestimated in some cases and overestimated in others. Some of the
+datasets have bost of the posteriormass on m = 1. Estimates of N are
+meaningless.}
+
+\subsubsection{Modelling assumptions}
 
@@
 
-This work was supported by grants from the Canadian Institutes of Health
-Research (CIHR, operating grant HOP-111406), and the Bill \& Melinda Gates
-Foundation (award number OPP1110049). R.M.M. was supported by a scholarship
-from the CIHR Strategic Training Program in Bioinformatics. A.F.Y.P. was
-supported by a CIHR New Investigator Award (Canadian HIV Vaccine Initiative,
-Vaccine Discovery and Social Research) and by a Career Investigator Scholar
-Award from the Michael Smith Foundation for Health Research, in partnership
-with the Providence Health Care Research Institute and St. Paul's Hospital
-Foundation.
+\add{We are grateful to Dr. Sally Otto, Dr. Alexandre Bouchard-C\^ot\'e, 
+Dr. Richard Harrigan, and the anonymous reviewers for many helpful
+suggestions.} This work was supported by grants from the Canadian Institutes of
+Health Research (CIHR, operating grant HOP-111406), and the Bill \& Melinda
+Gates Foundation (award number OPP1110049). R.M.M. was supported by a
+scholarship from the CIHR Strategic Training Program in Bioinformatics.
+A.F.Y.P. was supported by a CIHR New Investigator Award (Canadian HIV Vaccine
+Initiative, Vaccine Discovery and Social Research) and by a Career Investigator
+Scholar Award from the Michael Smith Foundation for Health Research, in
+partnership with the Providence Health Care Research Institute and St. Paul's
+Hospital Foundation.
 

### 2016-06-27 doc/tables/realdata-bc-short.tex
Paper and defence presentation updates

\begin{tabular}{ccc}
  Dataset & Sequences ($n$) & Gene \\
  \hline
  IDU/Estonia~\autocite{zetterberg2004two} & 171/188 & \textit{env}/\textit{gag} \\
  IDU/Romania~\autocite{niculescu2015recent} & 136 & \textit{pol} \\
  IDU/Canada & 399 & \textit{pol} \\
  HET/Botswana~\autocite{novitsky2013phylogenetic,novitsky2014impact} & 180 & \textit{env} \\
  HET/Malawi~\autocite{mccormack2002early} & 141/154 & \textit{env}/\textit{gag} \\
  HET/Uganda~\autocite{grabowski2014role} & 225 & \textit{env}/\textit{gag} \\
  MSM/Beijing~\autocite{wang2015targeting} & 173 & \textit{pol} \\
  MSM/Taiwan~\autocite{kao2011surveillance} & 275 & \textit{pol} \\
  MSM/USA~\autocite{little2014using} & 180 & \textit{pol} \\
  MSM/Shanghai~\autocite{li2015hiv} & 280 & \textit{pol} \\
  mixed/Spain~\autocite{cuevas2009hiv} & 287 & \textit{pol} \\
  \hline
\end{tabular}

### 2016-06-27 doc/tables/realdata.tex
Paper and defence presentation updates

@@
   \hline
-  \textcite{wang2015targeting} & 173 & Beijing, China & MSM & \textit{pol} \\
-  \textcite{cuevas2009hiv} & 287 & Basque Country, Spain & mixed & \textit{pol} \\
+  \textcite{zetterberg2004two} & 171/188 & Estonia & IDU & \textit{env}/\textit{gag} \\
+  \textcite{niculescu2015recent} & 136 & Romania & IDU & \textit{pol} \\
   \textcite{novitsky2013phylogenetic} & \multirow{2}{*}{180} &
@@
   \multirow{2}{*}{\textit{env}} \\ \textcite{novitsky2014impact} \\
+  \textcite{mccormack2002early} & 141/154 & Karonga District, Malawi & HET & \textit{env}/\textit{gag} \\
+  \textcite{grabowski2014role} & 225 & Rakai District, Uganda & HET & \textit{env}/\textit{gag} \\
+  \textcite{wang2015targeting} & 173 & Beijing, China & MSM & \textit{pol} \\
+  \textcite{kao2011surveillance} & 275 & Taiwan & MSM & \textit{pol} \\
+  \textcite{little2014using} & 180 & San Fransisco, USA & MSM & \textit{pol} \\
   \textcite{li2015hiv} & 280 & Shanghai, China & MSM & \textit{pol} \\
-  \textcite{niculescu2015recent} & 136 & Romania & IDU & \textit{pol} \\
+  \textcite{cuevas2009hiv} & 287 & Basque Country, Spain & mixed & \textit{pol} \\
   \hline

### 2016-06-27 doc/thesis/appendix.Rnw
Paper and defence presentation updates

@@
 
-\clearpage
-
-\begin{figure}[ht]
-    \includegraphics{bctree-posterior}
-    \caption
-    [
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to IDU/Canada data with \software{netabc}.
-    ]{
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to IDU/Canada data with \software{netabc}. Vertical
-        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
-        intervals. $x$-axis indicates regions of nonzero prior density.
-    }
-    \label{fig:bctree}
-\end{figure}
-
-\begin{figure}[ht]
-    \includegraphics{cuevas2009-posterior}
-    \caption[
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to mixed/Spain data with \software{netabc}.
-    ]{
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to mixed/Spain data with \software{netabc}. Vertical
-        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
-        intervals. $x$-axis indicates regions of nonzero prior density.
-    }
-    \label{fig:cuevas}
-\end{figure}
-
-\begin{figure}[ht]
-    \includegraphics{grabowski2014-posterior}
-    \caption[
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to HET/Uganda data with \software{netabc}.
-    ]{
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to HET/Uganda data with \software{netabc}. Vertical
-        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
-        intervals. $x$-axis indicates regions of nonzero prior density.
-    }
-    \label{fig:grabowski}
-\end{figure}
-
-\begin{figure}[ht]
-    \includegraphics{kao2011-posterior}
-    \caption[
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to MSM/Taiwan data with \software{netabc}.
-    ]{
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to MSM/Taiwan data with \software{netabc}. Vertical
-        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
-        intervals. $x$-axis indicates regions of nonzero prior density.
-    }
-    \label{fig:kao}
-\end{figure}
-
-\begin{figure}[ht]
-    \includegraphics{li2015-posterior}
-    \caption[
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to MSM/Shanghai data with \software{netabc}.
-    ]{
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to MSM/Shanghai data with \software{netabc}. Vertical
-        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
-        intervals. $x$-axis indicates regions of nonzero prior density. 
-    }
-    \label{fig:li}
-\end{figure}
-
-\begin{figure}[ht]
-    \includegraphics{little2014-posterior}
-    \caption[
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to MSM/USA data with \software{netabc}.
-    ]{
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to MSM/USA data with \software{netabc}. Vertical
-        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
-        intervals. $x$-axis indicates regions of nonzero prior density. 
-    }
-    \label{fig:little}
-\end{figure}
-
-\begin{figure}[ht]
-    \includegraphics{mccormack2002-posterior}
-    \caption[
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to HET/Malawi data with \software{netabc}.
-    ]{
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to HET/Malawi data with \software{netabc}. Vertical
-        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
-        intervals. $x$-axis indicates regions of nonzero prior density. 
-    }
-    \label{fig:mccormack}
-\end{figure}
-
-\begin{figure}[ht]
-    \includegraphics{niculescu2015-posterior}
-    \caption[
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to IDU/Romania data with \software{netabc}.
-    ]{
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to IDU/Romania data with \software{netabc}. Vertical
-        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
-        intervals. $x$-axis indicates regions of nonzero prior density.
-    }
-    \label{fig:niculescu}
-\end{figure}
-
-\begin{figure}[ht]
-    \includegraphics{novitsky2014-posterior}
-    \caption[
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to HET/Botswana data with \software{netabc}.
-    ]{
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to HET/Botswana data with \software{netabc}. Vertical
-        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
-        intervals. $x$-axis indicates regions of nonzero prior density.
-    }
-    \label{fig:novitsky}
-\end{figure}
-
-\begin{figure}[ht]
-    \includegraphics{wang2015-posterior}
-    \caption[
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to MSM/Beijing data with \software{netabc}.
-    ]{
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to MSM/Beijing data with \software{netabc}. Vertical
-        lines indicate posterior means, and shaded areas are 95\% highest
-        posterior density intervals. $x$-axis indicates regions of nonzero
-        prior density.
-    }
-    \label{fig:wang}
-\end{figure}
-
-\begin{figure}[ht]
-    \includegraphics{zetterberg2004-posterior}
-    \caption[
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to IDU/Estonia data with \software{netabc}.
-    ]{
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters fitted to IDU/Estonia data with \software{netabc}. Vertical
-        lines indicate posterior means, and shaded areas are 95\% highest
-        posterior density intervals. $x$-axis indicates regions of nonzero
-        prior density.
-    }
-    \label{fig:zetterberg}
-\end{figure}
-
+%\clearpage
+%
+%\begin{figure}[ht]
+%    \includegraphics{bctree-posterior}
+%    \caption
+%    [
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to IDU/Canada data with \software{netabc}.
+%    ]{
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to IDU/Canada data with \software{netabc}. Vertical
+%        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
+%        intervals. $x$-axis indicates regions of nonzero prior density.
+%    }
+%    \label{fig:bctree}
+%\end{figure}
+%
+%\begin{figure}[ht]
+%    \includegraphics{cuevas2009-posterior}
+%    \caption[
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to mixed/Spain data with \software{netabc}.
+%    ]{
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to mixed/Spain data with \software{netabc}. Vertical
+%        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
+%        intervals. $x$-axis indicates regions of nonzero prior density.
+%    }
+%    \label{fig:cuevas}
+%\end{figure}
+%
+%\begin{figure}[ht]
+%    \includegraphics{grabowski2014-posterior}
+%    \caption[
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to HET/Uganda data with \software{netabc}.
+%    ]{
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to HET/Uganda data with \software{netabc}. Vertical
+%        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
+%        intervals. $x$-axis indicates regions of nonzero prior density.
+%    }
+%    \label{fig:grabowski}
+%\end{figure}
+%
+%\begin{figure}[ht]
+%    \includegraphics{kao2011-posterior}
+%    \caption[
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to MSM/Taiwan data with \software{netabc}.
+%    ]{
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to MSM/Taiwan data with \software{netabc}. Vertical
+%        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
+%        intervals. $x$-axis indicates regions of nonzero prior density.
+%    }
+%    \label{fig:kao}
+%\end{figure}
+%
+%\begin{figure}[ht]
+%    \includegraphics{li2015-posterior}
+%    \caption[
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to MSM/Shanghai data with \software{netabc}.
+%    ]{
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to MSM/Shanghai data with \software{netabc}. Vertical
+%        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
+%        intervals. $x$-axis indicates regions of nonzero prior density. 
+%    }
+%    \label{fig:li}
+%\end{figure}
+%
+%\begin{figure}[ht]
+%    \includegraphics{little2014-posterior}
+%    \caption[
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to MSM/USA data with \software{netabc}.
+%    ]{
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to MSM/USA data with \software{netabc}. Vertical
+%        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
+%        intervals. $x$-axis indicates regions of nonzero prior density. 
+%    }
+%    \label{fig:little}
+%\end{figure}
+%
+%\begin{figure}[ht]
+%    \includegraphics{mccormack2002-posterior}
+%    \caption[
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to HET/Malawi data with \software{netabc}.
+%    ]{
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to HET/Malawi data with \software{netabc}. Vertical
+%        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
+%        intervals. $x$-axis indicates regions of nonzero prior density. 
+%    }
+%    \label{fig:mccormack}
+%\end{figure}
+%
+%\begin{figure}[ht]
+%    \includegraphics{niculescu2015-posterior}
+%    \caption[
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to IDU/Romania data with \software{netabc}.
+%    ]{
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to IDU/Romania data with \software{netabc}. Vertical
+%        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
+%        intervals. $x$-axis indicates regions of nonzero prior density.
+%    }
+%    \label{fig:niculescu}
+%\end{figure}
+%
+%\begin{figure}[ht]
+%    \includegraphics{novitsky2014-posterior}
+%    \caption[
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to HET/Botswana data with \software{netabc}.
+%    ]{
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to HET/Botswana data with \software{netabc}. Vertical
+%        lines indicate posterior means, and shaded areas are 95\% \gls{HPD}
+%        intervals. $x$-axis indicates regions of nonzero prior density.
+%    }
+%    \label{fig:novitsky}
+%\end{figure}
+%
+%\begin{figure}[ht]
+%    \includegraphics{wang2015-posterior}
+%    \caption[
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to MSM/Beijing data with \software{netabc}.
+%    ]{
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to MSM/Beijing data with \software{netabc}. Vertical
+%        lines indicate posterior means, and shaded areas are 95\% highest
+%        posterior density intervals. $x$-axis indicates regions of nonzero
+%        prior density.
+%    }
+%    \label{fig:wang}
+%\end{figure}
+%
+%\begin{figure}[ht]
+%    \includegraphics{zetterberg2004-posterior}
+%    \caption[
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to IDU/Estonia data with \software{netabc}.
+%    ]{
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters fitted to IDU/Estonia data with \software{netabc}. Vertical
+%        lines indicate posterior means, and shaded areas are 95\% highest
+%        posterior density intervals. $x$-axis indicates regions of nonzero
+%        prior density.
+%    }
+%    \label{fig:zetterberg}
+%\end{figure}
+%
 %<<abc_posterior, echo=FALSE, results="asis">>=

### 2016-06-27 doc/thesis/methods_results.Rnw
Paper and defence presentation updates

@@
 \subsection{Using synthetic data to investigate identifiability and sources of
-estimation error and bias}
+estimation error}
 
@@
 
-The accuracy of classifiers for $I$ \del{varied significantly
+The accuracy of classifiers for \gls{I} \del{varied significantly
 with}\add{appeared to be strongly influenced by} the number of tips in the
@@
     \caption[
-        Average posterior mean point estimates and 95\% \glspl{HDI} interval
+        Average posterior mean point estimates and 95\% \gls{HPD} interval
         widths for \gls{BA} model parameter estimates obtained with
@@
     ]{
-        Average posterior mean point estimates and 95\% \glspl{HDI} interval
+        Average posterior mean point estimates and 95\% \gls{HPD} interval
         widths for \gls{BA} model parameter estimates obtained with
@@
     \Sexpr{d[,round(quantile(N_error * 100 / true_N, 0.75))]}\%].
-Average \gls{HPD} interval widths were
+Average 95\% \gls{HPD} interval widths were
     \Sexpr{d[,round(mean(alpha.upper.95 - alpha.lower.95), 2)]},
@@
 The estimates for \gls{m} and \gls{N} were similar across all simulations
-(median [IQR] point estimates
+(median [\gls{IQR}] point estimates
     \Sexpr{d[,round(median(m), 2)]}
@@
 \add{To analyze the effects of the true parameter values on the
-accuracy our estimates of \gls{alpha} and \gls{I}, we fitted one \glspl{GLM}
+accuracy our estimates of \gls{alpha} and \gls{I}, we fitted one \gls{GLM}
 for each of these two parameters, with error rate as the dependent variable and
@@
 these parameters. The estimated coefficients are shown in
-\cref{tab:glmalpha,tab:glmI}. For the parameters \gls{alpha}, \gls{I}, and
-\gls{N}, the \gls{GLM} were fitted using the inverse link function. That is, if
-$p$ is the true value of the parameter and $\hat{p}$ is a random variable
-representing our estimate of the parameter, the \gls{GLM} posits a relationship
-of the form}
+\cref{tab:glmalpha,tab:glmI}. The \gls{GLM} were fitted using the inverse link
+function. That is, if $p$ is the true value of the parameter and $\hat{p}$ is a
+random variable representing our estimate of the parameter, the \gls{GLM}
+posits a relationship of the form}
 \[
@@
         df.copy <- copy(df)
-        df.copy[Parameter == "alpha", Parameter := "$\\alpha$"]
+        df.copy[Parameter == "alpha", Parameter := "\\alpha"]
+        df.copy[Parameter != "(Intercept)", Parameter := paste0("$", Parameter, "$")]
         df.copy[,p := paste0("$", sapply(p, pp, eq=FALSE), "$")]
@@
 
-<<supp_labels, echo=FALSE, results="asis">>=
-    #N <- 5000
-    #replicate <- 0
-    #lab <- NULL
-    #for (m in c(2, 3, 4)) {
-    #for (I in c(1000, 2000)) {
-    #for (alpha in c(0, 0.5, 1, 1.5)) {
-    #    lab <- c(lab, sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate))
-    #}
-    #}
-    #}
-    #lab <- paste(lab, collapse = ",")
+<<marginals, include=FALSE>>=
     worst.alpha <- d[,mean(alpha_error), by=true_alpha][,true_alpha[which.max(V1)]]
@@
     $\gls{N} = \Sexpr{worst[,true_N]}$.
-The two-dimensional marginals some dependence between pairs of parameters,
-particularly \gls{I} and \gls{N} which show a diagonally shaped region of high
-posterior density.}
+The two-dimensional marginals indicate some dependence between pairs of
+parameters, particularly \gls{I} and \gls{N} which show a diagonally shaped
+region of high posterior density.}
 
@@
 and found substantial heterogeneity among the parameter
-estimates.} \add{Posterior mean point estimates and 50\% and 95\% \glspl{HPD}
-for each parameter are shown in \cref{fig:abchpd}.} \del{and marginal posterior
-distributions in \cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}}.
-\add{\Cref{fig:abchpdm2} shows point estimates and \glspl{HPD} obtained when
-the value $\gls{m} = 1$ was disallowed by the prior. Since the results
+estimates.} \add{Posterior mean point estimates and 50\% and 95\% \gls{HPD}
+intervals for each parameter are shown in \cref{fig:abchpd}.} \del{and marginal
+posterior distributions in
+\cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}}.
+\add{\Cref{fig:abchpdm2} shows point estimates and \gls{HPD} intervals obtained
+when the value $\gls{m} = 1$ was disallowed by the prior. Since the results
 indicated that $\gls{m} = 1$ was the most credible value for several datasets,
@@
         Posterior means (points), 50\% \gls{HPD} intervals (notches), and 95\%
-        (lines) \gls{HPD} intervals for parameters of the \gls{BA} network
+        \gls{HPD} intervals (lines) for parameters of the \gls{BA} network
         model, fitted to eleven \gls{HIV} datasets with \software{netabc}.
@@
         \acrshort{IDU}, \acrlong{IDU}; \acrshort{MSM}, \acrlong{MSM}; HET,
-        heterosexual. Note that posterior means can fall outside of \glspl{HPD} 
-        if the distribution is diffuse.
+        heterosexual. Note that posterior means can fall outside of \gls{HPD} 
+        intervals if the distribution is diffuse.
     }
@@
 There was a large amount of uncertainty associated with these estimates. 95\%
-\gls{HPD} were very wide for most datasets, often encompassing almost the
-entire range from 0 to 1 (\cref{fig:abchpd}). For all the datasets except
+\gls{HPD} intervals were very wide for most datasets, often encompassing almost
+the entire range from 0 to 1 (\cref{fig:abchpd}). For all the datasets except
 HET/Malawi and MSM/Beijing, the posterior mean was either outside, or very
-close to the border of, the 50\% \gls{HPD}. This indicates that the posterior
-distributions were diffuse with heavy tails, rather than having most of their
-mass around the mode. When the value $\gls{m} = 1$ was disallowed by the prior, 
-the separation between the \gls{IDU} datasets and the others became more
-striking (\cref{fig:abchpdm2}). All three \gls{IDU} datasets had estimated
+close to the border of, the 50\% \gls{HPD} intervals. This indicates that the
+posterior distributions were diffuse with heavy tails, rather than having most
+of their mass around the mode. When the value $\gls{m} = 1$ was disallowed by
+the prior, the separation between the \gls{IDU} datasets and the others became
+more striking (\cref{fig:abchpdm2}). All three \gls{IDU} datasets had estimated
 \gls{alpha} values at or above
@@
     ]{
-        Posterior means (points) 50\% \gls{HPD} intervals (notches), and 95\%
+        Posterior means (points), 50\% \gls{HPD} intervals (notches), and 95\%
         \gls{HPD} intervals (lines) for parameters of the \gls{BA} network model,

### 2016-06-27 doc/thesis/methods_results.tex
Paper and defence presentation updates

@@
 The \gls{kSVR} classifier for \gls{alpha} had an average $R^2$ of 
-    0.98,
+    0.92,
 compared to 
@@
     90\%].
-Average \gls{HPD} interval widths were
+Average 95\% \gls{HPD} interval widths were
     0.68,
@@
 The estimates for \gls{m} and \gls{N} were similar across all simulations
-(median [IQR] point estimates
+(median [\gls{IQR}] point estimates
     3
@@
 % latex table generated in R 3.2.3 by xtable 1.8-2 package
-% Thu Jun 23 15:40:35 2016
+% Mon Jun 27 14:24:46 2016
 \begin{table}[ht]
@@
   $\alpha$ & $10$ & $2$ & ${<}10^{-5}$ \\ 
-  I & $-3\!\times\!10^{-4}$ & $2\!\times\!10^{-4}$ & $0.7$ \\ 
-  m & $0.5$ & $0.2$ & $0.01$ \\ 
+  $I$ & $-3\!\times\!10^{-4}$ & $2\!\times\!10^{-4}$ & $0.7$ \\ 
+  $m$ & $0.5$ & $0.2$ & $0.01$ \\ 
    \hline
@@
   $\alpha$ & $-0.001$ & $2\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
-  I & $-4\!\times\!10^{-7}$ & $2\!\times\!10^{-7}$ & $0.05$ \\ 
-  m & $-7\!\times\!10^{-5}$ & $8\!\times\!10^{-5}$ & $1$ \\ 
+  $I$ & $-4\!\times\!10^{-7}$ & $2\!\times\!10^{-7}$ & $0.05$ \\ 
+  $m$ & $-7\!\times\!10^{-5}$ & $8\!\times\!10^{-5}$ & $1$ \\ 
    \hline
@@
     $\gls{N} = 5000$.
-The two-dimensional marginals some dependence between pairs of parameters,
-particularly \gls{I} and \gls{N} which show a diagonally shaped region of high
-posterior density.}
+The two-dimensional marginals indicate some dependence between pairs of
+parameters, particularly \gls{I} and \gls{N} which show a diagonally shaped
+region of high posterior density.}
 

### 2016-06-27 doc/figures/abc-idea.R

@@ -10,7 +10,7 @@ br <- list(rep(1, t - 1), c(rep(1, t/2), rep(0.1, t/2)), c(1, rep(0.1, t-1)))
 tree <- lapply(mapply(rcoal, list(t), br=br, SIMPLIFY=FALSE), ladderize)
 net <- mapply(sample_pa, list(n), power=list(0, 1, 2), m=2, directed=FALSE, SIMPLIFY=FALSE)
 
-pdf("abc-idea.pdf", width=6, height=4, family="Gillius ADF")
+pdf("abc-idea.pdf", width=6, height=4)
 par(mfrow=c(4, 4), mar=c(0, 0, 0, 0))
 plot.new()
 text(0.5, 0.6, "sample", cex=2)

### 2016-06-27 doc/figures/acc-prec.R

@@ -25,7 +25,7 @@ lines <- data.frame(
     accurate=c("yes", "yes", "no", "no")
 )
 
-pdf("acc-prec.pdf", height=4, family="Gillius ADF")
+pdf("acc-prec.pdf", height=4)
 ggplot(d, aes(x=x, y=y, color=col)) + geom_line() + theme_bw() + xlim(-4, 4) +
     geom_vline(data=lines, aes(xintercept=v), lty=2) +
     facet_grid(precise~accurate, scales="free", labeller="label_both") +

### 2016-06-27 doc/figures/alpha-bounds.R

@@ -13,7 +13,7 @@ g2.large <- sample_pa(5000, m=2, power=2, directed=FALSE)
 summary(degree(g0.large))
 summary(degree(g2.large))
 
-pdf("alpha-bounds.pdf", family="Gillius ADF", width=6, height=6)
+pdf("alpha-bounds.pdf", width=6, height=6)
 par(mfrow=c(2, 2))
 par(mar=c(2, 2, 2, 2) + 0.1)
 

### 2016-06-27 doc/figures/gridsearch-expt-extra.R

@@ -22,7 +22,7 @@ y <- dens$y
 gradient <- colorRampPalette(c("blue", "white"))(length(x))
 true.x <- x[length(x)*0.4]
 true.x.col <- gradient[length(x)*0.4]
-pdf("tinykscore.pdf", family="Gillius ADF", width=3, height=0.8)
+pdf("tinykscore.pdf", width=3, height=0.8)
 par(mar=c(1, 1, 0, 0) + 0.1, mgp=c(0, 1, 0), cex=1.25)
 plot(x, y, type="p", xlim=c(-4, 4), main=NA, pch=16,
      xlab="training value", ylab="score", col=gradient, xaxt="n", yaxt="n")

### 2016-06-27 doc/figures/homophily.R

@@ -1,0 +1,39 @@
+#!/usr/bin/env Rscript
+
+library(igraph)
+
+seed <- 381
+
+n <- 10
+vcol <- rep(c("blue", "magenta"), each=n)
+
+repeat {{
+    print(seed)
+    set.seed(seed)
+    g <- bipartite.random.game(n, n, type="gnm", m=2*n)
+    g2 <- rewire(g, each_edge(p = .1, loops = FALSE))
+    if (min(degree(g2)) > 0) break
+    seed <- seed+1
+}} 
+
+seed <- 390
+repeat {{
+    print(seed)
+    set.seed(seed)
+    g3 <- bipartite.random.game(n, n, type="gnm", m=2*n)
+    g3 <- rewire(g3, each_edge(p = .1, loops = FALSE))
+    if (min(degree(g3)) > 0) break
+    seed <- seed+1
+}} 
+
+layout <-  layout.auto(g2)
+pdf("homophily.pdf")
+plot(g, vertex.label=NA, vertex.color=vcol, vertex.frame.color=vcol,
+     edge.color=NA, edge.width=NA, layout=layout)
+plot(g, vertex.label=NA, vertex.color=vcol, vertex.frame.color=vcol,
+     edge.color="black", edge.width=2, layout=layout)
+plot(g2, vertex.label=NA, vertex.color=vcol, vertex.frame.color=vcol,
+     edge.color="black", edge.width=2, layout=layout)
+plot(g3, vertex.label=NA, vertex.color=vcol, vertex.frame.color=vcol,
+     edge.color="black", edge.width=2)
+dev.off()

### 2016-06-27 doc/figures/kernel-idea.R

@@ -2,73 +2,43 @@
 
 suppressPackageStartupMessages(library(netabc))
 library(RColorBrewer)
-library(grid)
 
-epidemic.mst <- function (g, nI) {
-    t <- mst(g)
-    I <- vcount(g)
-
-    while (I > nI) {
-        v <- which(degree(t) == 1)[1]
-        t <- delete_edges(t, E(t)[inc(v)])
-        I <- I - 1
-    }
-
-    inodes <- which(degree(t) > 0)
-    iedges <- get.edge.ids(g, c(t(as_edgelist(t))))
-    list(nodes=inodes, edges=iedges)
-}
-
+col <- as.list(brewer.pal(3, "Set1"))
 n <- 40
-t <- 12
-alpha <- c(0, 2)
-m <- 2
-
-br <- list(rep(1, t - 1), c(1, rep(0.1, t-1)))
+t <- 8
+br <- list(rep(1, t - 1), c(rep(1, t/2), rep(0.1, t/2)), c(1, rep(0.1, t-1)))
 tree <- lapply(mapply(rcoal, list(t), br=br, SIMPLIFY=FALSE), ladderize)
-net <- mapply(sample_pa, list(n), power=as.list(alpha), m=m, directed=FALSE, SIMPLIFY=FALSE)
+net <- mapply(sample_pa, list(n), power=list(0, 1, 2), m=2, directed=FALSE, SIMPLIFY=FALSE)
 
-epi <- lapply(net, epidemic.mst, t)
-inodes <- lapply(epi, "[[", "nodes")
-iedges <- lapply(epi, "[[", "edges")
-ecol <- mapply(rep, list(NA), lapply(net, ecount), SIMPLIFY=FALSE)
-ecol <- mapply(function (x, i) { x[i] <- "red"; x}, ecol, iedges, SIMPLIFY=FALSE)
-vcol <- mapply(rep, list("black"), lapply(net, ecount), SIMPLIFY=FALSE)
-vcol <- mapply(function (x, i) { x[i] <- "red"; x}, vcol, inodes, SIMPLIFY=FALSE)
+pdf("kernel-idea.pdf", width=6, height=4)
+par(mfrow=c(4, 4), mar=c(0, 0, 0, 0))
+plot.new()
+text(0.5, 0.625, "sample", cex=2)
+text(0.5, 0.4, "parameters", cex=2)
+plot.new()
+text(0.5, 0.5, expression(theta[1]), cex=4, col=col[[1]])
+plot.new()
[…]

### 2016-06-27 doc/figures/nettypes.R

@@ -1,0 +1,18 @@
+#!/usr/bin/env Rscript
+
+library(igraph)
+
+n <- 15
+
+set.seed(1)
+
+g1 <- sample_pa(n, m=2, power=0, directed=FALSE)
+g2 <- sample_pa(n, m=2, power=2, directed=FALSE)
+
+pdf("nettypes.pdf")
+par(mfrow=c(2, 1), mar=c(1, 0, 1, 0))
+plot(g1, vertex.label=NA, vertex.color="black", edge.color="black",
+     edge.width=2)
+plot(g2, vertex.label=NA, vertex.color="black",
+     edge.color="black", edge.width=2)
+dev.off()

### 2016-06-27 doc/figures/nltt.R

@@ -12,7 +12,7 @@ sf2 <- stepfun(c(0, 1/3, 2/3), c(0, 1/3, 2/3, 1))
 af1 <- approxfun(c(0, 0.5, 1), c(0, 0.5, 1))
 af2 <- approxfun(c(0, 1/3, 2/3, 1), c(0, 1/3, 2/3, 1))
 
-pdf("nltt.pdf", family="Gillius ADF")
+pdf("nltt.pdf")
 par(mfrow=c(2, 2), lwd=3)
 old.mar <- par("mar")
 par(mar=c(0, 2, 4, 2) + 0.1)

### 2016-06-27 doc/figures/objects.R

@@ -9,7 +9,7 @@ net <- sample_pa(n, m=2, directed=FALSE)
 trans.tree <- ladderize(rcoal(t))
 phylo <- ladderize(rtree(t))
 
-pdf("objects.pdf", width=7, height=2, family="Gillius ADF")
+pdf("objects.pdf", width=7, height=2)
 par(mfrow=c(1, 3), mar=c(0, 0, 0, 0))
 plot(net, vertex.label=NA, vertex.size=3, vertex.color="black")
 plot(trans.tree, show.tip.label=FALSE, direction="down", edge.width=2)

### 2016-06-27 doc/figures/pa-example.R

@@ -3,13 +3,15 @@
 library(igraph)
 library(extrafont)
 
-set.seed(0)
+set.seed(1)
 g0.small <- sample_pa(50, m=2, power=0, directed=FALSE)
+g1.small <- sample_pa(50, m=2, power=1, directed=FALSE)
 g2.small <- sample_pa(50, m=2, power=2, directed=FALSE)
 
-pdf("pa-example.pdf", family="Gillius ADF", width=6, height=3)
-par(mfrow=c(1, 2))
-par(mar=c(2, 2, 2, 2) + 0.1)
+pdf("pa-example.pdf", width=6, height=3)
+par(mfrow=c(1, 3))
+par(mar=c(2, 1, 2, 1) + 0.1)
 plot(g0.small, vertex.label=NA, vertex.color="black", vertex.size=5)
+plot(g1.small, vertex.label=NA, vertex.color="black", vertex.size=5)
 plot(g2.small, vertex.label=NA, vertex.color="black", vertex.size=5)
 dev.off()

### 2016-06-27 doc/figures/smc-test-bimodal.R

@@ -3,17 +3,17 @@
 library(data.table)
 library(extrafont)
 
-trace.file <- system2(c("locate", "check_smc_trace.tsv"), stdout=TRUE)
+trace.file <- system2(c("locate", "check_smc_bimodal_trace.tsv"), stdout=TRUE)
 trace <- fread(trace.file[1])
 
 trace <- trace[iter == max(iter)]
 trace <- trace[sample(1:nrow(trace), prob=weight, replace=TRUE)]
 dens <- density(trace$theta)
 
-x <- seq(-4, 4, 0.01)
-target <- 0.5 * dnorm(x) + 0.5 * dnorm(x, sd=0.1)
+x <- seq(-10, 10, 0.01)
+target <- 0.5 * dnorm(x, mean=-4) + 0.5 * dnorm(x, mean=4)
 
-pdf("smc-test.pdf", width=5, height=5)
+pdf("smc-test-bimodal.pdf", width=5, height=5)
 plot(x, target, type="n", xlab="theta", ylab="density")
 polygon(dens$x, dens$y, col="gray80")
 lines(x, target)

### 2016-06-27 doc/figures/vaccinate.R

@@ -1,0 +1,48 @@
+#!/usr/bin/env Rscript
+
+library(igraph)
+
+seed <- 111
+n <- 20
+nI <- 10
+nI2 <- 5
+
+repeat {
+    set.seed(seed)
+    g <- sample_gnm(n, n)
+    if (length(subcomponent(g, 1)) == n) break
+    seed <- seed + 1
+    print(seed)
+}
+
+I <- sample(1:vcount(g), 1)
+for (i in 1:nI) {
+    I <- c(I, sample(setdiff(as.list(V(g)[nei(I)]), I), 1))
+    if (i == nI2) {
+        I2 <- unlist(I)
+    }
+}
+I <- unlist(I)
+
+layout <- layout.auto(g)
+
+vcol <- rep("black", vcount(g))
+vcol[I] <- "red"
+
+ecol <- rep("black", ecount(g))
+ecol[E(g)[I %--% I]] <- "red"
+
+pdf("vaccinate.pdf")
+par(mfrow=c(2, 1), mar=c(1, 0, 1, 0))
+plot(g, vertex.label=NA, edge.color=ecol, vertex.color=vcol,
+     vertex.frame.color=vcol, edge.width=2, layout=layout)
+
+vcol <- rep("black", vcount(g))
+vcol[I2] <- "red"
+
+ecol <- rep("black", ecount(g))
+ecol[E(g)[I2 %--% I2]] <- "red"
+
+plot(g, vertex.label=NA, edge.color=ecol, vertex.color=vcol, 
+     vertex.frame.color=vcol, edge.width=2, layout=layout)
+dev.off()

### 2016-06-28 doc/defence/trees.tex
another cartoon

--- 2016-06-27 doc/defence/presentation.tex
@@
 \usepackage{multirow}
-\usepackage{qrcode}
 \usepackage[backend=biber, style=numeric-comp, sorting=none]{biblatex}
@@
 \usetikzlibrary{calc}
+\usetikzlibrary{automata}
 
@@
 
-\maketitle
+\begin{frame}{Transmission trees shape viral phylogenies}
+  \begin{center}
+    \begin{tikzpicture}[
+        every node/.style = {inner sep=0pt},
+        every path/.style = {->, >=stealth, very thick},
+        label/.style = {anchor=south, color=white, inner sep=4pt}
+      ]
+      \def\pht{1cm}
+      \node (abcd) {\includegraphics[height=\pht]{stock/person}};
+      \node at (abcd.south) [label] {\Large a};
 
-\begin{frame}{Modeling in epidemiology}
-  \centerline{\includegraphics[width=0.9\textwidth]{compartments}}
-\end{frame}
-
-\begin{frame}{Contact heterogeneity can affect epidemic expansion}
-  \centerline{\includegraphics[width=0.9\textwidth]{sir-trajectories}}
-\end{frame}
-
-\begin{frame}{Contact networks capture contact heterogeneity}
-  \begin{center}
-  \begin{tikzpicture}
-    \node (c) {\includegraphics[width=4cm]{contactnet-empty}};
-
-    \uncover<2->{
-    \node (a) [right=of c.north east, anchor=west] {nodes = individuals};
-    \draw [->, thick, >=stealth] (c.north east) ++(-1.5cm, -0.5cm) -- (a.west);
+      \uncover<2>{
+      \node (b) at (abcd.east) [anchor=west] {\includegraphics[height=\pht]{stock/person}};
+      \node at (b.south) [label] {\Large b};
+      \draw [red] (abcd.south) to [bend right=75] (b.south);
       }
@@
       \uncover<3->{
-    \node (b) [below=0.5 of a] {edges = contacts};
-    \draw [->, thick, >=stealth] (c.north east) ++(-0.75cm, -1.25cm) -- (b.west);
+      \node (b) [below right=1 and 0.75 of abcd] {\includegraphics[height=\pht]{stock/person}};
+      \node at (b.south) [label] {\Large b};
+      \node (acd) [below left=1 and 0.5 of abcd] {\includegraphics[height=\pht]{stock/person}};
+      \node at (acd.south) [label] {\Large a};
+      \draw [red] (abcd) -- (b);
+      \draw [-, dashed] (abcd) -- (acd);
       }
+
+      \uncover<4>{
+      \node (cd) at (acd.east) [anchor=west] {\includegraphics[height=\pht]{stock/person}};
+      \node at (cd.south) [label] {\Large c};
+      \draw [blue] (acd.south) to [bend right=75] (cd.south);
+      }
+
+      \uncover<5->{
+      \node (cd) [below right=0.5 and 0.5 of acd] {\includegraphics[height=\pht]{stock/person}};
+      \node at (cd.south) [label] {\Large c};
+      \node (a) [below left=0.25 and 0.25 of acd] {\includegraphics[height=\pht]{stock/person}};
+      \node at (a.south) [label] {\Large a};
+      \draw [blue] (acd) -- (cd);
+      \draw [-, dashed] (acd) -- (a);
+      }
+
+      \uncover<6>{
+      \node (d) at (cd.east) [anchor=west] {\includegraphics[height=\pht]{stock/person}};
+      \node at (d.south) [label] {\Large d};
+      \draw [green] (cd.south) to [bend right=75] (d.south);
+      }
+
+      \uncover<7->{
+      \node (c) [below left=0.25 and 0.25 of cd] {\includegraphics[height=\pht]{stock/person}};
+      \node at (c.south) [label] {\Large c};
+      \node (d) [below right=0.4 and 0.4 of cd] {\includegraphics[height=\pht]{stock/person}};
+      \node at (d.south) [label] {\Large d};
+      \draw [-, dashed] (cd) -- (c);
+      \draw [green!80!black] (cd) -- (d);
+      }
+
+      \coordinate [below right=0.5 and 3 of b] (abcd);
+      \coordinate [below right=1 and 1 of abcd] (b);
+      \coordinate [above left=1 and 0.25 of abcd] (acd);
+      \coordinate [above left=1 and 0.25 of acd] (a);
+      \coordinate [right=of acd] (cd);
+      \coordinate [right=1.5 of cd] (c);
+      \coordinate [above right=of cd] (d);
+
+      \uncover<1-2>{
+      \node (x) at (abcd) {\includegraphics[height=\pht]{stock/virus}};
+      \node at (x.center) {\large a};
+      }
+
+      \uncover<2>{
+      \node (x) [below right=0.3 of abcd, anchor=north west] {\includegraphics[height=\pht]{stock/virus}}; 
+      \node at (x.center) {\large b};
+      }
+
+      \uncover<3-4>{
+      \node (x) [above=0 of acd, anchor=south] {\includegraphics[height=\pht]{stock/virus}}; 
+      \node at (x.center) {\large a};
+      }
+
+      \uncover<3->{
+      \node (x) [below right=-0.2 of b, anchor=north west] {\includegraphics[height=\pht]{stock/virus}}; 
+      \node at (x.center) {\large b};
+
+      \draw (abcd) -- (acd);
+      \draw (abcd) -- (b);
+      }
+
+      \uncover<4>{
+      \node (x) [right=0.3 of acd, anchor=west] {\includegraphics[height=\pht]{stock/virus}}; 
+      \node at (x.center) {\large c};
+      }
+
+      \uncover<5-6>{
+      \node (x) at (cd) [anchor=west] {\includegraphics[height=\pht]{stock/virus}};  
+      \node at (x.center) {\large c};
+      }
+
+      \uncover<5->{
+      \node (x) at (a) [anchor=south] {\includegraphics[height=\pht]{stock/virus}};
+      \node at (x.center) {\large a};
+
+      \draw (acd) -- (cd);
+      \draw (acd) -- (a);
+      }
+
+      \uncover<6>{
+      \node (x) [above right=0.4\pht and 0.4\pht of cd] [anchor=south west] {\includegraphics[height=\pht]{stock/virus}};
+      \node at (x.center) {\large d};
+      }
+
+      \uncover<7->{
+      \node (x) at (c) [anchor=west] {\includegraphics[height=\pht]{stock/virus}};
+      \node at (x.center) {\large c};
+
+      \node (x) at (d) [anchor=south west] {\includegraphics[height=\pht]{stock/virus}};
+      \node at (x.center) {\large d};
+
+      \draw (cd) -- (c);
+      \draw (cd) -- (d);
+      }
+
     \end{tikzpicture}
@@
 
-\begin{frame}{Network models formally describe network structure}
-  \begin{columns}
-    \column{0.55\textwidth}
-    \begin{minipage}[c][.6\textheight][c]{\linewidth}
-      \begin{enumerate}
-        \setlength{\itemsep}{12pt}
-        \item Start with \only<1-4>{10}\only<5->{\alert{$n$}} women and 
-          \only<1-4>{10}\only<5->{\alert{$n$}} men.
-        \uncover<2->{
-        \item Randomly form \only<2-4>{20}\only<5->{\alert{$p \times n$}} male-female partnerships.
-        }
-        \uncover<3->{
-        \item Randomly rewire each edge with \only<3-4>{10}\only<5->{\\\alert{$q$ }}\% probability.
-        }
-      \end{enumerate}
-    \end{minipage}
-    \column{0.5\textwidth}
-    \begin{minipage}[c][.6\textheight][c]{\linewidth}
-      \only<1>{\includegraphics[width=5cm, page=1]{homophily}}
-      \only<2>{\includegraphics[width=5cm, page=2]{homophily}}
-      \only<3>{\includegraphics[width=5cm, page=3]{homophily}}
-      \only<4->{\includegraphics[width=5cm, page=4]{homophily}}
-    \end{minipage}
-  \end{columns}
-
-  \vspace{0.5cm}
-  \begin{columns}
-    \column{0.55\textwidth}
-    \uncover<6->{\centerline{\large{network model}}}
-    \column{0.55\textwidth}
-    \uncover<6->{\centerline{\large{contact network}}}
-  \end{columns}
-\end{frame}
-
-\begin{frame}{Barab\'asi-Albert model incorporates preferential attachment}
-  \begin{columns}
-  \column{0.5\textwidth}
-  \begin{tikzpicture}[
-      scale=2,
-      every edge/.style={thick}
-    ]
-    \node (n1) at (1, 0) [circle, fill=black] { };
-    \node (n2) at (0.86, 0.5) [circle, fill=black] { };
-    \node (n3) at (0.5, 0.86) [circle, fill=black] { };
-    \draw (n1) -- (n2) -- (n3) -- (n1);
-
-    \uncover<2->{
-      \node (n4) at (0, 1) [circle, fill=black] { };
-    }
-    \only<2-3>{
-      \node [right=0 of n4.north east, anchor=south west] {$m = 2$};
-    }
-    \only<2-4>{
-      \draw (n4) -- ++ (0.2, -0.2);
-      \draw (n4) -- ++ (-0.2, -0.2);
-    }
-
-    \node [right=0 of n1] {%
-      \only<3-4>{$\Pr \propto 2^\alpha$}%
-      \only<6-7>{$\Pr \propto 3^\alpha$}%
-    };
-    \node [right=0 of n2] {%
-      \only<3-4>{$\Pr \propto 2^\alpha$}%
-      \only<6-7>{$\Pr \propto 3^\alpha$}%
-    };
-    \node [right=0 of n3] {%
-      \only<3-4>{$\Pr \propto 2^\alpha$}%
-      \only<6-7>{$\Pr \propto 2^\alpha$}%
-    };
-
-    \only<4>{
-      \draw (n4) -- ++ (0.2, -0.2) -- (n2);
-      \draw (n4) -- ++ (-0.2, -0.2) -- (n1);
-    }
-
-    \uncover<5->{
-      \draw (n4) -- (n2);
-      \draw (n4) -- (n1);
-    }
-
-    \uncover<6->{
-      \node (n5) at (-0.5, 0.86) [circle, fill=black] { };
-    }
-
-    \only<6-7> {
-      \draw (n5) -- ++ (0.2, -0.2);
-      \draw (n5) -- ++ (-0.2, -0.2);
-    }
-
-    \uncover<6-7>{ \node [right=0 of n4.north east, anchor=south west] { $\Pr \propto 2^\alpha$ }; }
-
-    \only<7>{
-      \draw (n5) -- ++ (0.2, -0.2) -- (n3);
-      \draw (n5) -- ++ (-0.2, -0.2) -- (n2);
-    }
-
-    \uncover<8->{
-      \draw (n5) -- (n3);
-      \draw (n5) -- (n2);
-    }
-
-    \uncover<9->{
-      \node (n6) at (-0.86, 0.5) [circle, fill=black] { };
-      \draw (n6) -- (n2);
-      \draw (n6) -- (n1);
-    }
-
-    \uncover<10->{
-      \node (n7) at (-1, 0) [circle, fill=black] { };
-      \draw (n7) -- (n3);
-      \draw (n7) -- (n4);
-    }
-
-    \uncover<11->{
-      \node (n8) at (-0.86, -0.5) [circle, fill=black] { };
-      \node (n9) at (-0.5, -0.86) [circle, fill=black] { };
-      \node (n10) at (0, -1) [circle, fill=black] { };
-      \node (n11) at (0.5, -0.86) [circle, fill=black] { };
-      \node (n12) at (0.86, -0.5) [circle, fill=black] { };
-
-      \draw (n8) -- (n2);
-      \draw (n8) -- (n4);
-      \draw (n9) -- (n2);
-      \draw (n9) -- (n1);
-      \draw (n10) -- (n2);
-      \draw (n10) -- (n8);
-      \draw (n11) -- (n3);
-      \draw (n11) -- (n9);
-      \draw (n12) -- (n2);
-      \draw (n12) -- (n8);
-    }
-
-    \uncover<12->{
-      \node (n1) at (1, 0) [circle, fill=red] { };
-      \node (n2) at (0.86, 0.5) [circle, fill=red] { };
-      \node (n5) at (-0.86, 0.5) [circle, fill=red] { };
-      \node (n7) at (-0.86, -0.5) [circle, fill=red] { };
-      \node (n8) at (-0.5, -0.86) [circle, fill=red] { };
-
-      \draw [ultra thick, red] (n2) -- (n7);
-      \draw [ultra thick, red] (n2) -- (n1);
-      \draw [ultra thick, red] (n8) -- (n1);
-      \draw [ultra thick, red] (n5) -- (n2);
-    }
-
-    \only<11>{\node [below=of n10] {$N = 12$};}
-    \uncover<12->{\node [below=of n10] {$I = 5$};}
-  \end{tikzpicture}
-
-  \column{0.4\textwidth}
-  \uncover<2->{
-    $m$ = number of new edges per vertex \\
-    \hfill\\
-  }
-
-  \uncover<3->{
-    $\alpha$ = preferential attachment power \\
-    \hfill\\
-  }
-
-  \uncover<11->{
-    $N$ = number of nodes \\
-    \hfill\\
-  }
-
-  \uncover<12->{
-    $I$ = number of infected nodes (prevalence) 
-  }
-  \end{columns}
-\end{frame}
-
-\begin{frame}{Why fit network models?}
-  \begin{columns}
-    \column{0.33\textwidth}
-    \includegraphics[width=0.85\textwidth]{sir-trajectories-vertical}
-
-    \column{0.33\textwidth}
-    \uncover<2->{
-    \includegraphics[width=0.75\textwidth]{nettypes}
-    }
-
-    \column{0.33\textwidth}
-    \uncover<3->{
-    \begin{tikzpicture}
-      \node (pic) { \includegraphics[width=0.75\textwidth]{vaccinate} };
-      \node at (pic.east) { \includegraphics[width=1cm]{stock/syringe} };
-    \end{tikzpicture}
-    }
-  \end{columns}
-
-  \vspace{-0.5cm}
-
-  \begin{columns}
-    \column{0.33\textwidth}
-    \uncover<1->{
-    \centerline{\large predict}
-    }
-
-    \column{0.33\textwidth}
-    \uncover<2->{
-    \centerline{\large characterize}
-    }
-
-    \column{0.33\textwidth}
-    \uncover<3->{
-    \centerline{\large simulate}
-    }
-  \end{columns}
-\end{frame}
-
-\begin{frame}{Phylogenetic methods may help fit network models}
-  \begin{itemize}
-    \item network models are difficult to fit with conventional epidemiological
-      methods (surveys etc.)
-    \item sequence data may be easier to collect and free from some biases
-      (misreporting etc.)
-    \item instead of asking the people, ask the virus! (some cartoon)
-  \end{itemize}
-\end{frame}
-
-\begin{frame}{Contact networks shape transmission trees}
-  \begin{center}
-    \begin{tikzpicture}
-      [every node/.style = {inner sep=0pt},
-       every path/.style = {->, >=stealth, very thick},
-       label/.style = {anchor=south, color=white, inner sep=4pt},
-       time/.style = {anchor=north west, inner sep=2pt}]
-      \def\pht{1cm}
-      \node (a) {\includegraphics[height=\pht]{stock/person}};
-      \node at (a.south) [label] {\Large a};
-      \node [below left=0.75 of a] (b) {\includegraphics[height=\pht]{stock/person}};
-      \node at (b.south) [label] {\Large b};
-      \node [below=0.75 of b] (e) {\includegraphics[height=\pht]{stock/person}};
-      \node at (e.south) [label] {\Large e};
-      \node [below right=0.75 of a] (c) {\includegraphics[height=\pht]{stock/person}};
-      \node at (c.south) [label] {\Large c};
-      \node [below=0.75 of c] (d) {\includegraphics[height=\pht]{stock/person}};
-      \node at (d.south) [label] {\Large d};
-
-      \fill [shade, opacity=0.2] (a.east) -- (c.north) -- (c.north west) -- (a.south east) -- cycle;
-      \uncover<3->{
-      \draw [blue] ($(a.east) + (0, -12pt)$) -- ($(c.north) + (-8pt, 0)$);
-      }
-      \fill [shade, opacity=0.2] (a.west) -- (b.north) -- (b.north east) -- (a.south west) -- cycle;
-      \uncover<2->{
-      \draw [red] ($(a.west) + (0, -12pt)$) -- ($(b.north) + (8pt, 0)$);
-      }
-      \fill [shade, opacity=0.2] (a.south) -- (a.south west) -- (e.north) -- (e.north east) -- cycle;
-      \fill [shade, opacity=0.2] (b.south west) -- (b.south) -- (e.north) -- (e.north west) -- cycle;
-      \fill [shade, opacity=0.2] ($(c.south) + (-8pt, 0)$) -- ++(16pt, 0) -- ($(d.north) + (8pt, 0)$) -- ++(-16pt, 0) -- cycle;
-      \fill [shade, opacity=0.2] (b.east) -- (b.south east) -- (d.west) -- (d.north west) -- cycle;
-      \uncover<4->{
-      \draw [green!80!black] (c) -- (d);
-      }
-
-      \node (abcd) [right=6cm of a.north] {\includegraphics[height=\pht]{stock/person}};
-      \node at (abcd.south) [label] {\Large a};
-      \uncover<2->{
-      \node (b) [below right=1 and 0.75 of abcd] {\includegraphics[height=\pht]{stock/person}};
-      \node at (b.south) [label] {\Large b};
-      \node (acd) [below left=1 and 0.5 of abcd] {\includegraphics[height=\pht]{stock/person}};
-      \node at (acd.south) [label] {\Large a};
-      }
-      \uncover<3->{
-      \node (cd) [below right=0.5 and 0.5 of acd] {\includegraphics[height=\pht]{stock/person}};
-      \node at (cd.south) [label] {\Large c};
-
-      \node (a) [below left=0.25 and 0.25 of acd] {\includegraphics[height=\pht]{stock/person}};
-      \node at (a.south) [label] {\Large a};
-      }
-      \uncover<4->{
-      \node (c) [below left=0.25 and 0.25 of cd] {\includegraphics[height=\pht]{stock/person}};
-      \node at (c.south) [label] {\Large c};
-      \node (d) [below right=0.4 and 0.4 of cd] {\includegraphics[height=\pht]{stock/person}};
-      \node at (d.south) [label] {\Large d};
-      }
-
-      \uncover<2->{
-      \draw [red] (abcd) -- (b);
-      \draw [-, dashed] (abcd) -- (acd);
-      }
-      \uncover<3->{
-      \draw [blue] (acd) -- (cd);
-      \draw [-, dashed] (acd) -- (a);
-      }
-      \uncover<4->{
-      \draw [-, dashed] (cd) -- (c);
-      \draw [green!80!black] (cd) -- (d);
-      }
-    \end{tikzpicture}
-
-  $\qquad$contact network \hfill transmission tree$\qquad$
-  \end{center}
-\end{frame}
-
-\begin{frame}{Transmission trees shape viral phylogenies}
-  some demo of viruses evolving and being transmitted at the same time
-\end{frame}
-
-\begin{frame}{Can we fit network models with viral phylogenies?}
-  \begin{center}
-  network $\to$ transmission tree $\to$ viral phylogeny \\
-  \hfill\\
-  network $\stackrel{\alert{?}}{\leftarrow}$ transmission tree $\leftarrow$ viral phylogeny \\
-  \end{center}
-\end{frame}
-
-\begin{frame}{Are network parameters identifiable from tree shape?}
-  \begin{minipage}[p][\textheight][t]{\textwidth}
-    \only<1>{\includegraphics[width=0.9\textwidth, trim=0 3in 0 0, clip]{kernel-idea.pdf}}
-    \only<2>{\includegraphics[width=0.9\textwidth, trim=0 1.7in 0 0, clip]{kernel-idea.pdf}}
-    \only<3>{\includegraphics[width=0.9\textwidth, trim=0 0.8in 0 0, clip]{kernel-idea.pdf}}
-    \only<4->{\includegraphics[width=0.9\textwidth, trim=0 0in 0 0, clip]{kernel-idea.pdf}}
-  \end{minipage}
-\end{frame}
-
-\begin{frame}{yes they do}
-  \centerline{\includegraphics[height=0.8\textheight]{kernel-kpca}}
-\end{frame}
-
-\begin{frame}{Simulated trees can be used to fit network models}
-  \centerline{\includegraphics[width=0.9\textwidth]{abc-smc}}
-
-  \centerline{animate this?}
-\end{frame}
-
-\begin{frame}{Posterior mean estimates are more accurate for $\alpha$ and $I$}
-  \begin{minipage}[p][\textheight][t]{\textwidth}
-    \vspace{-0.5cm}
-    preferential attachment power \hfill \uncover<2>{edges per vertex}
-
-    \hspace{0.75cm}\only<1>{\includegraphics[height=0.8\textheight, trim=0 0 3.4in 0, clip]{abc-point-estimate-m3.pdf}}
-    \only<2>{\includegraphics[height=0.8\textheight]{abc-point-estimate-m3.pdf}}
-
-    prevalence \hfill \uncover<2>{total nodes}
-  \end{minipage}
-\end{frame}
-
-\begin{frame}{Example simulation\ldots}
-  \begin{minipage}[p][\textheight][t]{\textwidth}
-    \begin{center}
-    preferential attachment power \hfill prevalence
-    \only<1>{
-      \includegraphics[width=0.8\textwidth, trim=0 2.5in 0 0, clip]{abc-posterior-example.pdf}
-    }
-    \only<2>{
-      \includegraphics[width=0.8\textwidth]{abc-posterior-example.pdf}
-
-      \vspace{-0.25cm}
-      edges per vertex \hfill total nodes
-    }
-    \end{center}
-  \end{minipage}
-\end{frame}
-
-\begin{frame}{Example simulation continued, $N$ and $I$ confounded}
-  %\includegraphics[width=\textwidth]{{abc-posterior-2d/1.0_1000_3_5000_0}.pdf}
-\end{frame}
-
-\begin{frame}{Application to real world HIV datasets}
-  \centerline{\input{\tablepath/realdata-bc-short}}
-\end{frame}
-
-\begin{frame}{HIV networks have sub-linear preferential attachment}
-    Also, higher PA for IDU networks.
-
-    \includegraphics[trim=0 3.5in 3.3in 0, clip, width=0.5\textwidth]{realdata-hpd-bc}
-    \includegraphics[trim=3.4in 1.5in 0 2.0in, clip, width=0.5\textwidth]{realdata-hpd-bc}
-\end{frame}
-
-\begin{frame}{Conclusions}
-  \begin{itemize}
-    \item We developed a phylodynamic method to fit contact network models to
-      phylogenetic data.
-      \pause
-    \item The preferential attachment power of the Barab\'asi-Albert network
-      model, which is challenging to estimate by traditional epidemiological
-      methods, can be estimated with ABC.
-      \pause
-    \item The networks underlying real epidemics are heterogeneous,
-      underscoring the importance of considering network structure in
-      phylodynamic analyses.
-  \end{itemize}
-\end{frame}
-
-\begin{frame}{Acknowledgements}
-  \begin{columns}
-    \begin{column}{0.6\textwidth}
-
-      \textbf{BC Centre for Excellence in HIV/AIDS}
-
-      Art Poon
-
-      Jeff Joy
-
-      Richard Liang
-
-      Thuy Nguyen
-
-      P. Richard Harrigan
-
-      \hfill\\
-      \textbf{University of British Columbia}
-
-      Sarah Otto
-
-      Alexandre Bouchard-C\^ot\'e
-
-      \vfill
-      \vspace{0.5cm}
-      $\qquad\qquad$\includegraphics[width=2cm]{logos/genomecanada}
-      $\qquad$
-      \includegraphics[width=2cm]{logos/genomequebec}
-    \end{column}
-    \begin{column}{0.4\textwidth}
-      \centering
-
-      \includegraphics[width=3cm]{logos/cfe}
-      \vspace{0.5cm}
-
-      \includegraphics[width=3cm]{logos/cihr}
-
-      \includegraphics[width=3cm]{logos/genomebc}
-
-      \includegraphics[width=3cm]{logos/bmgf}
-      \vspace{0.5cm}
-
-      \includegraphics[width=3cm]{logos/btp}
-    \end{column}
-  \end{columns}
-\end{frame}
-
 \end{document}

### 2016-06-29 doc/defence/contactnet.tex
working on defence talk

--- 2016-04-25 doc/hiv-dynamics/presentation.tex
@@
-\documentclass{beamer}
-\usepackage{tikz}
-\usepackage{gillius}
-\usepackage{abraces}
-\usepackage{etex}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{qrcode}
-\usepackage[backend=biber, style=authoryear, uniquelist=minyear, uniquename=false]{biblatex}
+\begin{frame}{Contact networks represent possible transmission routes}
+  \begin{center}
+  \begin{tikzpicture}
+    \node (c) {\includegraphics[width=4cm]{contactnet-empty}};
 
-% http://tex.stackexchange.com/questions/12703/how-to-create-fixed-width-table-columns-with-text-raggedright-centered-raggedlef
-\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-
-\usetikzlibrary{external}
-\usetikzlibrary{shapes}
-\usetikzlibrary{arrows}
-\usetikzlibrary{positioning}
-\usetikzlibrary{decorations.pathreplacing}
-
-\usetheme[everytitleformat=regular]{m}
-\setbeamertemplate{navigation symbols}{}
-
-\graphicspath{{../figures/}}
-\newcommand{\tablepath}{../tables}
-
-\title{Phylogenetic inference of contact network parameters with kernel approximate Bayesian computation}
-\author[RMM, RHL \& AFYP]{Rosemary M McCloskey$^1$ \and Richard H Liang$^1$ \and Art FY Poon$^{1,2}$}
-\institute[UBC \& BCCfE]{$^1$BC Centre for Excellence in HIV/AIDS, Vancouver, Canada \\ $^2$Department of Medicine, University of British Columbia, Vancouver, Canada}
-\date{HIV Dynamics \& Evolution, Woods Hole, USA, April 25, 2016}
-
-\newcommand{\dd}[2]{\frac{\text{d}\,#1}{\text{d}\,#2}}
-
-\addbibresource{papers.bib}
-
-\begin{document}
-\setbeamercolor{background canvas}{bg=white}
-\renewcommand{\footnotesize}{\tiny}
-\definecolor{red}{RGB}{228,26,28}
-\definecolor{blue}{RGB}{55,126,184}
-\definecolor{green}{RGB}{77,175,74}
-\definecolor{purple}{RGB}{152,78,163}
-
-\maketitle
-
-\begin{frame}{Most epidemiological models assume homogeneous mixing}
-  \begin{columns}
-    \begin{column}{0.6\textwidth}
-      \centering
-      \includegraphics[scale=0.6]{compartments}
-      \vspace{1cm}
-
-      \uncover<3->{\includegraphics[width=\textwidth]{sir-trajectories}}
-    \end{column}
-    \begin{column}{0.4\textwidth}
-      \vspace{-2cm}
-      \begin{itemize}
-        \setlength{\itemsep}{24pt}
     \uncover<2->{
-        \item Often provide a reasonable approximation in practice.
-        }
-        \uncover<3->{
-        \item Can be inaccurate when substantial contact heterogeneity exists.
-        }
-      \end{itemize}
-    \end{column}
-  \end{columns}
-\end{frame}
-
-\begin{frame}{Network models capture contact heterogeneity}
-  \begin{columns}
-    \begin{column}{0.5\textwidth}
-      \includegraphics[width=\textwidth]{contactnet-empty}
-    \end{column}
-    \begin{column}{0.5\textwidth}
-      \begin{itemize}
-        \setlength{\itemsep}{12pt}
-          \pause
-        \item May offer more accurate predictions for highly structured
-          populations.
-          \pause
-        \item Network parameters may be of interest for their own sake,
-          \textit{e.g.} are there superspreaders?
-          \pause
-        \item Extremely difficult to estimate in practice.
-      \end{itemize}
-    \end{column}
-  \end{columns}
-
-\end{frame}
-
-\begin{frame}{Contact networks shape transmission trees}
-  \includegraphics[width=\textwidth]{contactnet}
-
-  \begin{itemize}
-      \pause
-    \item Transmission trees in turn shape viral phylogenies.
-      \pause
-    \item \textbf{Aim:} estimate contact network parameters from viral
-      phylogenies.
-  \end{itemize}
-\end{frame}
-
-%\begin{frame}{Preferential attachment generates scale-free networks}
-%  \includegraphics[width=\textwidth]{pa-example}
-%  \begin{itemize}
-%    \item Sexual networks are often scale-free: degree distributions appear
-%      follow a power law ($\Pr(\text{degree} = d) \propto d^{-\gamma}$).
-%      \pause
-%    \item Preferential attachment models generate scale-free networks.
-%  \end{itemize}
-%\end{frame}
-
-\begin{frame}{Barab\'asi-Albert model incorporates preferential attachment}
-  \begin{columns}
-    \begin{column}{0.5\textwidth}
-      \begin{minipage}[p][\textheight][t]{\textwidth}
-        \centering
-        \only<1>{\includegraphics[page=1, width=\textwidth]{pa.pdf}}
-        \only<2>{\includegraphics[page=2, width=\textwidth]{pa.pdf}}
-        \only<3>{\includegraphics[page=3, width=\textwidth]{pa.pdf}}
-        \only<4>{\includegraphics[page=4, width=\textwidth]{pa.pdf}}
-        \only<5>{\includegraphics[page=5, width=\textwidth]{pa.pdf}}
-        \only<6>{\includegraphics[page=6, width=\textwidth]{pa.pdf}}
-      \end{minipage}
-    \end{column}
-    \begin{column}{0.5\textwidth}
-      \begin{minipage}[t][\textheight][t]{\textwidth}
-      \begin{itemize}
-        \uncover<1->{ \item Start with a small number of connected nodes. }
-        \uncover<2->{ \item Attach new nodes with $m$ edges. }
-        \uncover<3->{ \item Other endpoints of degree $d$ are chosen with probability $\propto d^{\alpha} + 1$. }
-        \uncover<5->{ \item Continue until network has $N$ nodes. }
-        \uncover<6->{ \item Also consider the prevalence $I$. }
-      \end{itemize}
-      \end{minipage}
-    \end{column}
-  \end{columns}
-\end{frame}
-
-\begin{frame}{Do network parameters measurably affect tree shape?}
-  \only<1>{
-    \includegraphics[height=1in, trim=0 2in 4in 0, clip]{kernel-idea}
-  }
-  \only<2>{
-    \includegraphics[height=1in, trim=0 2in 2in 0, clip]{kernel-idea}
-  }
-  \only<3->{
-    \includegraphics[height=1in, trim=0 2in 0 0, clip]{kernel-idea}
-  }
-  \begin{itemize}
-    \setlength{\itemsep}{0pt}
-    \item Generate networks under different parameter values (number of nodes
-      $N$, number of edges per vertex $m$, pereferential attachment power
-      $\alpha$).
-    \uncover<2->{
-    \item Simulate epidemic over each network until $I$ nodes are infected.
-    }
-    \uncover<3->{
-    \item Randomly subsample to form transmission trees.
-    }
-    \uncover<4->{
-    \item Compare trees pairwise using tree kernel\footfullcite{poon2013mapping}.
-    }
-  \end{itemize}
-\end{frame}
-
-\begin{frame}{Preferential attachment power $\alpha$ affects tree shape}
-  \begin{tikzpicture}[remember picture, overlay]
-    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=0.2in 2.7in 3.3in 0, clip]{kernel-kpca.pdf}};
-    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-alpha-tree.pdf}};
-  \end{tikzpicture}
-\end{frame}
-
-\begin{frame}{Prevalence $I$ affects tree shape}
-  \begin{tikzpicture}[remember picture, overlay]
-    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=3.5in 2.7in 0in 0, clip]{kernel-kpca.pdf}};
-    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-I-tree.pdf}};
-  \end{tikzpicture}
-\end{frame}
-
-\begin{frame}{Number of edges per vertex $m$ does not affect tree shape}
-  \begin{tikzpicture}[remember picture, overlay]
-    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=0.2in 0.3in 3.3in 2.2in, clip]{kernel-kpca.pdf}};
-    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-m-tree.pdf}};
-  \end{tikzpicture}
-\end{frame}
-
-\begin{frame}{Number nodes $N$ modestly affects tree shape}
-  \begin{tikzpicture}[remember picture, overlay]
-    \node at (current page.south west) [anchor=south west] {\includegraphics[width=0.6\textwidth, trim=3.5in 0.3in 0 2.2in, clip]{kernel-kpca.pdf}};
-    \node (t) at (current page.north east) [anchor=north east, inner ysep=1.5cm, inner xsep=0.5cm] {\includegraphics[width=0.7\textwidth]{kernel-N-tree.pdf}};
-  \end{tikzpicture}
-\end{frame}
-
-\begin{frame}{Simulated trees can be used to fit a network model}
-  \begin{minipage}[p][\textheight][t]{\textwidth}
-    \only<1>{\includegraphics[width=0.9\textwidth, trim=0 3in 0 0, clip]{abc-idea.pdf}}
-    \only<2>{\includegraphics[width=0.9\textwidth, trim=0 1.8in 0 0, clip]{abc-idea.pdf}}
-    \only<3>{\includegraphics[width=0.9\textwidth, trim=0 1in 0 0, clip]{abc-idea.pdf}}
-    \uncover<4->{\includegraphics[width=0.9\textwidth, trim=0 0in 0 0, clip]{abc-idea.pdf}}
-    \uncover<5->{\vspace{0.5cm}\centerline{\textbf{kernel approximate Bayesian computation (kernel-ABC)}}}
-  \end{minipage}
-\end{frame}
-
-\begin{frame}{Example simulation shows $\alpha$ and $I$ can be reconstructed}
-  \begin{minipage}[p][\textheight][t]{\textwidth}
-    \begin{center}
-    preferential attachment power \hfill prevalence
-    \only<1>{
-      \includegraphics[width=0.8\textwidth, trim=0 2.5in 0 0, clip]{abc-posterior-example.pdf}
-    }
-    \only<2>{
-      \includegraphics[width=0.8\textwidth]{abc-posterior-example.pdf}
-
-      \vspace{-0.25cm}
-      edges per vertex \hfill total nodes
-    }
-    \end{center}
-  \end{minipage}
-\end{frame}
-
-\begin{frame}{Maximum \textit{a posteriori} estimates are accurate for $\alpha$ and $I$}
-  \begin{minipage}[p][\textheight][t]{\textwidth}
-    \vspace{-0.5cm}
-    preferential attachment power \hfill \uncover<2>{edges per vertex}
-
-    \hspace{0.75cm}\only<1>{\includegraphics[height=0.8\textheight, trim=0 0 3.4in 0, clip]{abc-point-estimate-m2.pdf}}
-    \only<2>{\includegraphics[height=0.8\textheight]{abc-point-estimate-m2.pdf}}
-
-    prevalence \hfill \uncover<2>{total nodes}
-  \end{minipage}
-\end{frame}
-
-\begin{frame}{Real world HIV datasets exhibit network heterogeneity}
-  \uncover<2->{
-    \centerline{\includegraphics[height=0.8\textheight]{realdata-hpd-bc}}
+    \node (a) [right=of c.north east, anchor=west] {nodes = individuals};
+    \draw [->, thick, >=stealth] (c.north east) ++(-1.5cm, -0.5cm) -- (a.west);
     }
 
-%  \scriptsize
-%    \begin{tabular}{ccccc}
-%      Reference & Sequences ($n$) & Location & Risk group & Gene \\
-%      \hline
-%      \color{magenta} \textcite{novitsky2013phylogenetic} &
-%        \color{magenta} \multirow{2}{*}{180} &
-%        \color{magenta} \multirow{2}{*}{Mochudi, Botswana} & 
-%        \color{magenta} \multirow{2}{*}{HET} &
-%        \color{magenta} \multirow{2}{*}{\textit{env}} \\ 
-%      \color{magenta} \textcite{novitsky2014impact} \\
-%      \color{blue} \textcite{cuevas2009hiv} & 
-%        \color{blue} 287 & 
-%        \color{blue} Basque Country, Spain & 
-%        \color{blue} mixed & 
-%        \color{blue} \textit{pol} \\
-%      \color{cyan} \textcite{li2015hiv} & 
-%        \color{cyan} 280 & 
-%        \color{cyan} Shanghai, China & 
-%        \color{cyan} MSM & 
-%        \color{cyan} \textit{pol} \\
-%      \color{green} unpublished & 
-%        \color{green} 399 & 
-%        \color{green} British Columbia, Canada & 
-%        \color{green} IDU & 
-%        \color{green} \textit{pol} \\
-%      \color{orange} \textcite{wang2015targeting} & 
-%        \color{orange} 173 & 
-%        \color{orange} Beijing, China & 
-%        \color{orange} MSM & 
-%        \color{orange} \textit{pol} \\
-%      \color{red} \textcite{niculescu2015recent} & 
-%        \color{red} 136 & 
-%        \color{red} Romaina & 
-%        \color{red} IDU & 
-%        \color{red} \textit{pol} \\
-%      \hline
-%    \end{tabular}
-%  \normalsize
+    \uncover<3->{
+    \node (b) [below=0.5 of a] {edges = contacts};
+    \draw [->, thick, >=stealth] (c.north east) ++(-0.75cm, -1.25cm) -- (b.west);
+    }
+  \end{tikzpicture}
+  \end{center}
 \end{frame}
-
-\begin{frame}{Conclusions}
-  \begin{itemize}
-    \item Kernel-ABC is the first phylodynamic method to fit contact network
-      models to phylogenetic data.
-      \pause
-    \item The preferential attachment power of the Barab\'asi-Albert network
-      model, which is challenging to estimate by traditional epidemiological
-      methods, can be estimated with kernel-ABC.
-      \pause
-    \item The networks underlying real epidemics are heterogeneous,
-      underscoring the importance of considering network structure in
-      phylodynamic analyses.
-      \pause
-  \end{itemize}
-  \centerline{\qrcode{github.com/rmcclosk/netabc} github.com/rmcclosk/netabc}
-\end{frame}
-
-\begin{frame}{Acknowledgements}
-  \begin{columns}
-    \begin{column}{0.6\textwidth}
-
-      \textbf{BC Centre for Excellence in HIV/AIDS}
-
-      Art Poon
-
-      Jeff Joy
-
-      Richard Liang
-
-      Thuy Nguyen
-
-      P. Richard Harrigan
-
-      \hfill\\
-      \textbf{University of British Columbia}
-
-      Sarah Otto
-
-      Alexandre Bouchard-C\^ot\'e
-
-      \vfill
-      \vspace{0.5cm}
-      $\qquad\qquad$\includegraphics[width=2cm]{logos/genomecanada}
-      $\qquad$
-      \includegraphics[width=2cm]{logos/genomequebec}
-    \end{column}
-    \begin{column}{0.4\textwidth}
-      \centering
-
-      \includegraphics[width=3cm]{logos/cfe}
-      \vspace{0.5cm}
-
-      \includegraphics[width=3cm]{logos/cihr}
-
-      \includegraphics[width=3cm]{logos/genomebc}
-
-      \includegraphics[width=3cm]{logos/bmgf}
-      \vspace{0.5cm}
-
-      \includegraphics[width=3cm]{logos/btp}
-    \end{column}
-  \end{columns}
-\end{frame}
-
-\begin{frame}{Classifiers for network parameters}
-  \vspace{-1cm}
-  \begin{itemize}
-    \item kSVR using tree kernel
-    \item SVR using normalized
-      lineages-through-time\footfullcite{janzen2015approximate}
-    \item linear regression using Sackin's index~\footfullcite{shao1990tree}
-  \end{itemize}
-
-  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
-\end{frame}
-
-\begin{frame}{Sequential Monte Carlo for ABC}
-  \begin{tikzpicture}
-    \only<1-2>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 4in 0, clip]{abc-smc.pdf}}};
-    \only<1>{\fill[white] (0.2,-6.5) rectangle (4,-3.5)};
-    \only<3>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 2in 0, clip]{abc-smc.pdf}}};
-    \only<4>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 0 0, clip]{abc-smc.pdf}}};
-  \end{tikzpicture}
-\end{frame}
-
-\end{document}

### 2016-06-29 doc/defence/identifiable.tex
working on defence talk

\begin{frame}{Are network parameters identifiable from tree shape?}
  \begin{minipage}[p][\textheight][t]{\textwidth}
    \only<2>{\includegraphics[width=0.9\textwidth, trim=0 3in 0 0, clip]{kernel-idea.pdf}}
    \only<3>{\includegraphics[width=0.9\textwidth, trim=0 1.7in 0 0, clip]{kernel-idea.pdf}}
    \only<4>{\includegraphics[width=0.9\textwidth, trim=0 0.8in 0 0, clip]{kernel-idea.pdf}}
    \only<5->{\includegraphics[width=0.9\textwidth, trim=0 0in 0 0, clip]{kernel-idea.pdf}}
  \end{minipage}
\end{frame}

### 2016-06-29 doc/defence/inference.tex
working on defence talk

\begin{frame}{Can we fit network models from viral sequence data?}
    \begin{tikzpicture}
      [
        every path/.style={very thick, ->, >=stealth}
      ]
      \uncover<2->{
        \node at (0, 0) [anchor=center] {\includegraphics[width=\textwidth]{objects}};
        \draw (-3.5, 2) to [bend left=75] (-1.5, 2);
      }
      \only<2>{ 
        \fill [white] (0, 2) rectangle (6, -2); 
      }
      \only<3>{
        \fill [white] (2.8, 2) rectangle (6, -2); 
      }
      \uncover<3->{
        \draw (-1, 2) to [bend left=75] (1, 2);
      }

      \uncover<4->{
        \draw (1.5, 2) to [bend left=75] (3.5, 2);
      }

      \uncover<5->{
        \draw [green] (3.5, -2) to [bend left=75] node [auto, green] {\Large \checkmark} (1.5, -2);
      }

      \uncover<6->{
        \draw [green] (1, -2) to [bend left=75] node [auto, green] {\Large ``\checkmark''} (-1, -2);
      }
      \uncover<7->{
        \draw [orange] (-1.5, -2) to [bend left=75] node [auto] {\Large \alert{?}} (-3.5, -2);
      }
    \end{tikzpicture}
\end{frame}

### 2016-06-29 doc/defence/phylonet.tex
working on defence talk

\begin{frame}{Phylogenetic methods may help fit network models}
  \begin{center}
  \begin{tikzpicture}
    [
      label/.style = {anchor=south, color=white, inner sep=6pt},
      every path/.style = {very thick}
    ]
    \def\pht{1cm}
    \uncover<2-5>{
    \node (a) {\includegraphics[height=\pht]{stock/person}};
    \node (b) [below right=3 and 1.5 of a] {\includegraphics[height=\pht]{stock/person}};
    \node (c) [below left=3 and 1.5 of a] {\includegraphics[height=\pht]{stock/person}};
    }

    \uncover<2-5>{
    \draw (a) -- ++ (-1, 0);
    \draw (a) -- ++ (-0.86, -0.5);
    }

    \uncover<3-5>{
    \draw (a) -- ++ (-0.5, -0.86) node [anchor=north east] {\alert{?}};
    \draw (a) -- ++ (0, -1) node [anchor=north] {\alert{?}};
    \draw (a) -- ++ (0.5, -0.86) node [anchor=north west] {\alert{?}};
    \draw (a) -- ++ (0.86, -0.5) node [anchor=west] {\alert{?}};
    \draw (a) -- ++ (1, 0) node [anchor=west] {\alert{?}};
    }

    \uncover<4-5>{
    \draw [->, >=stealth] (b) to [bend right] (c);
    \draw [->, >=stealth] (c) to [bend right] node 
      [-, red, forbidden sign, draw=red, inner sep=6pt] {} (b);
    \node [below=3.3 of a] {\Large \alert{?}};
    }

    \uncover<5>{
    \node (d) [below right=0 and 2 of a] {\includegraphics[height=\pht]{stock/person}};
    \node (e) [below left=0 and 2 of a] {\includegraphics[height=\pht]{stock/person}};
    \node at (d.south) [label] {\Large \alert{?}};
    \node at (e.south) [label] {\Large \alert{?}};
    }

    \uncover<6->{
    \node (a) [opacity=0.3] {\includegraphics[height=\pht]{stock/person}};
    \node (b) [opacity=0.3, below right=3 and 1.5 of a] {\includegraphics[height=\pht]{stock/person}};
    \node (c) [opacity=0.3, below left=3 and 1.5 of a] {\includegraphics[height=\pht]{stock/person}};

    \node (va) at (a) [anchor=north] {\includegraphics[height=\pht]{stock/virus}};
    \node (vb) at (b) [anchor=east] {\includegraphics[height=\pht]{stock/virus}};
    \node (vc) at (c) [anchor=west] {\includegraphics[height=\pht]{stock/virus}};
    }

    \uncover<7->{
    \node (e) [opacity=0.3, below left=0 and 2 of a] {\includegraphics[height=\pht]{stock/person}};
    \coordinate [below=2 of a] (mid);
    \draw (va.south) -- (mid) -- (vb.north west);
    \draw (vc.north) -- (e.east) -- (mid);
    \draw [dashed] (va.south east) -- (vb.north);
    }
  \end{tikzpicture}
  \end{center}
\end{frame}

### 2016-06-29 doc/defence/presentation.tex
working on defence talk

@@
 \usepackage[backend=biber, style=numeric-comp, sorting=none]{biblatex}
+\usepackage{hyperref}
 
@@
 \usetheme[everytitleformat=regular]{m}
-\setbeamertemplate{navigation symbols}{}
+%\setbeamertemplate{navigation symbols}{}
 
@@
 \maketitle
+\input{contactnet}
+\input{netmodel}
+\input{why}
+\input{ba}
+\input{phylonet}
+\input{transtree}
+\input{virusphylo}
+\input{inference}
+\input{identifiable}
 
-\begin{frame}{Modeling in epidemiology}
-  \centerline{\includegraphics[width=0.9\textwidth]{compartments}}
-\end{frame}
-
-\begin{frame}{Contact heterogeneity can affect epidemic expansion}
-  \centerline{\includegraphics[width=0.9\textwidth]{sir-trajectories}}
-\end{frame}
-
-\begin{frame}{Contact networks capture contact heterogeneity}
-  \begin{center}
-  \begin{tikzpicture}
-    \node (c) {\includegraphics[width=4cm]{contactnet-empty}};
-
-    \uncover<2->{
-    \node (a) [right=of c.north east, anchor=west] {nodes = individuals};
-    \draw [->, thick, >=stealth] (c.north east) ++(-1.5cm, -0.5cm) -- (a.west);
-    }
-
-    \uncover<3->{
-    \node (b) [below=0.5 of a] {edges = contacts};
-    \draw [->, thick, >=stealth] (c.north east) ++(-0.75cm, -1.25cm) -- (b.west);
-    }
-  \end{tikzpicture}
-  \end{center}
-\end{frame}
-
-\begin{frame}{Network models formally describe network structure}
-  \begin{columns}
-    \column{0.55\textwidth}
-    \begin{minipage}[c][.6\textheight][c]{\linewidth}
-      \begin{enumerate}
-        \setlength{\itemsep}{12pt}
-        \item Start with \only<1-4>{10}\only<5->{\alert{$n$}} women and 
-          \only<1-4>{10}\only<5->{\alert{$n$}} men.
-        \uncover<2->{
-        \item Randomly form \only<2-4>{20}\only<5->{\alert{$p \times n$}} male-female partnerships.
-        }
-        \uncover<3->{
-        \item Randomly rewire each edge with \only<3-4>{10}\only<5->{\\\alert{$q$ }}\% probability.
-        }
-      \end{enumerate}
-    \end{minipage}
-    \column{0.5\textwidth}
-    \begin{minipage}[c][.6\textheight][c]{\linewidth}
-      \only<1>{\includegraphics[width=5cm, page=1]{homophily}}
-      \only<2>{\includegraphics[width=5cm, page=2]{homophily}}
-      \only<3>{\includegraphics[width=5cm, page=3]{homophily}}
-      \only<4->{\includegraphics[width=5cm, page=4]{homophily}}
-    \end{minipage}
-  \end{columns}
-
-  \vspace{0.5cm}
-  \begin{columns}
-    \column{0.55\textwidth}
-    \uncover<6->{\centerline{\large{network model}}}
-    \column{0.55\textwidth}
-    \uncover<6->{\centerline{\large{contact network}}}
-  \end{columns}
-\end{frame}
-
-\begin{frame}{Barab\'asi-Albert model incorporates preferential attachment}
-  \begin{columns}
-  \column{0.5\textwidth}
-  \begin{tikzpicture}[
-      scale=2,
-      every edge/.style={thick}
-    ]
-    \node (n1) at (1, 0) [circle, fill=black] { };
-    \node (n2) at (0.86, 0.5) [circle, fill=black] { };
-    \node (n3) at (0.5, 0.86) [circle, fill=black] { };
-    \draw (n1) -- (n2) -- (n3) -- (n1);
-
-    \uncover<2->{
-      \node (n4) at (0, 1) [circle, fill=black] { };
-    }
-    \only<2-3>{
-      \node [right=0 of n4.north east, anchor=south west] {$m = 2$};
-    }
-    \only<2-4>{
-      \draw (n4) -- ++ (0.2, -0.2);
-      \draw (n4) -- ++ (-0.2, -0.2);
-    }
-
-    \node [right=0 of n1] {%
-      \only<3-4>{$\Pr \propto 2^\alpha$}%
-      \only<6-7>{$\Pr \propto 3^\alpha$}%
-    };
-    \node [right=0 of n2] {%
-      \only<3-4>{$\Pr \propto 2^\alpha$}%
-      \only<6-7>{$\Pr \propto 3^\alpha$}%
-    };
-    \node [right=0 of n3] {%
-      \only<3-4>{$\Pr \propto 2^\alpha$}%
-      \only<6-7>{$\Pr \propto 2^\alpha$}%
-    };
-
-    \only<4>{
-      \draw (n4) -- ++ (0.2, -0.2) -- (n2);
-      \draw (n4) -- ++ (-0.2, -0.2) -- (n1);
-    }
-
-    \uncover<5->{
-      \draw (n4) -- (n2);
-      \draw (n4) -- (n1);
-    }
-
-    \uncover<6->{
-      \node (n5) at (-0.5, 0.86) [circle, fill=black] { };
-    }
-
-    \only<6-7> {
-      \draw (n5) -- ++ (0.2, -0.2);
-      \draw (n5) -- ++ (-0.2, -0.2);
-    }
-
-    \uncover<6-7>{ \node [right=0 of n4.north east, anchor=south west] { $\Pr \propto 2^\alpha$ }; }
-
-    \only<7>{
-      \draw (n5) -- ++ (0.2, -0.2) -- (n3);
-      \draw (n5) -- ++ (-0.2, -0.2) -- (n2);
-    }
-
-    \uncover<8->{
-      \draw (n5) -- (n3);
-      \draw (n5) -- (n2);
-    }
-
-    \uncover<9->{
-      \node (n6) at (-0.86, 0.5) [circle, fill=black] { };
-      \draw (n6) -- (n2);
-      \draw (n6) -- (n1);
-    }
-
-    \uncover<10->{
-      \node (n7) at (-1, 0) [circle, fill=black] { };
-      \draw (n7) -- (n3);
-      \draw (n7) -- (n4);
-    }
-
-    \uncover<11->{
-      \node (n8) at (-0.86, -0.5) [circle, fill=black] { };
-      \node (n9) at (-0.5, -0.86) [circle, fill=black] { };
-      \node (n10) at (0, -1) [circle, fill=black] { };
-      \node (n11) at (0.5, -0.86) [circle, fill=black] { };
-      \node (n12) at (0.86, -0.5) [circle, fill=black] { };
-
-      \draw (n8) -- (n2);
-      \draw (n8) -- (n4);
-      \draw (n9) -- (n2);
-      \draw (n9) -- (n1);
-      \draw (n10) -- (n2);
-      \draw (n10) -- (n8);
-      \draw (n11) -- (n3);
-      \draw (n11) -- (n9);
-      \draw (n12) -- (n2);
-      \draw (n12) -- (n8);
-    }
-
-    \uncover<12->{
-      \node (n1) at (1, 0) [circle, fill=red] { };
-      \node (n2) at (0.86, 0.5) [circle, fill=red] { };
-      \node (n5) at (-0.86, 0.5) [circle, fill=red] { };
-      \node (n7) at (-0.86, -0.5) [circle, fill=red] { };
-      \node (n8) at (-0.5, -0.86) [circle, fill=red] { };
-
-      \draw [ultra thick, red] (n2) -- (n7);
-      \draw [ultra thick, red] (n2) -- (n1);
-      \draw [ultra thick, red] (n8) -- (n1);
-      \draw [ultra thick, red] (n5) -- (n2);
-    }
-
-    \only<11>{\node [below=of n10] {$N = 12$};}
-    \uncover<12->{\node [below=of n10] {$I = 5$};}
-  \end{tikzpicture}
-
-  \column{0.4\textwidth}
-  \uncover<2->{
-    $m$ = number of new edges per vertex \\
-    \hfill\\
-  }
-
-  \uncover<3->{
-    $\alpha$ = preferential attachment power \\
-    \hfill\\
-  }
-
-  \uncover<11->{
-    $N$ = number of nodes \\
-    \hfill\\
-  }
-
-  \uncover<12->{
-    $I$ = number of infected nodes (prevalence) 
-  }
-  \end{columns}
-\end{frame}
-
-\begin{frame}{Why fit network models?}
-  \begin{columns}
-    \column{0.33\textwidth}
-    \includegraphics[width=0.85\textwidth]{sir-trajectories-vertical}
-
-    \column{0.33\textwidth}
-    \uncover<2->{
-    \includegraphics[width=0.75\textwidth]{nettypes}
-    }
-
-    \column{0.33\textwidth}
-    \uncover<3->{
-    \begin{tikzpicture}
-      \node (pic) { \includegraphics[width=0.75\textwidth]{vaccinate} };
-      \node at (pic.east) { \includegraphics[width=1cm]{stock/syringe} };
-    \end{tikzpicture}
-    }
-  \end{columns}
-
-  \vspace{-0.5cm}
-
-  \begin{columns}
-    \column{0.33\textwidth}
-    \uncover<1->{
-    \centerline{\large predict}
-    }
-
-    \column{0.33\textwidth}
-    \uncover<2->{
-    \centerline{\large characterize}
-    }
-
-    \column{0.33\textwidth}
-    \uncover<3->{
-    \centerline{\large simulate}
-    }
-  \end{columns}
-\end{frame}
-
-\begin{frame}{Phylogenetic methods may help fit network models}
-  \begin{itemize}
-    \item network models are difficult to fit with conventional epidemiological
-      methods (surveys etc.)
-    \item sequence data may be easier to collect and free from some biases
-      (misreporting etc.)
-    \item instead of asking the people, ask the virus! (some cartoon)
-  \end{itemize}
-\end{frame}
-
-\begin{frame}{Contact networks shape transmission trees}
-  \begin{center}
-    \begin{tikzpicture}
-      [every node/.style = {inner sep=0pt},
-       every path/.style = {->, >=stealth, very thick},
-       label/.style = {anchor=south, color=white, inner sep=4pt},
-       time/.style = {anchor=north west, inner sep=2pt}]
-      \def\pht{1cm}
-      \node (a) {\includegraphics[height=\pht]{stock/person}};
-      \node at (a.south) [label] {\Large a};
-      \node [below left=0.75 of a] (b) {\includegraphics[height=\pht]{stock/person}};
-      \node at (b.south) [label] {\Large b};
-      \node [below=0.75 of b] (e) {\includegraphics[height=\pht]{stock/person}};
-      \node at (e.south) [label] {\Large e};
-      \node [below right=0.75 of a] (c) {\includegraphics[height=\pht]{stock/person}};
-      \node at (c.south) [label] {\Large c};
-      \node [below=0.75 of c] (d) {\includegraphics[height=\pht]{stock/person}};
-      \node at (d.south) [label] {\Large d};
-
-      \fill [shade, opacity=0.2] (a.east) -- (c.north) -- (c.north west) -- (a.south east) -- cycle;
-      \uncover<3->{
-      \draw [blue] ($(a.east) + (0, -12pt)$) -- ($(c.north) + (-8pt, 0)$);
-      }
-      \fill [shade, opacity=0.2] (a.west) -- (b.north) -- (b.north east) -- (a.south west) -- cycle;
-      \uncover<2->{
-      \draw [red] ($(a.west) + (0, -12pt)$) -- ($(b.north) + (8pt, 0)$);
-      }
-      \fill [shade, opacity=0.2] (a.south) -- (a.south west) -- (e.north) -- (e.north east) -- cycle;
-      \fill [shade, opacity=0.2] (b.south west) -- (b.south) -- (e.north) -- (e.north west) -- cycle;
-      \fill [shade, opacity=0.2] ($(c.south) + (-8pt, 0)$) -- ++(16pt, 0) -- ($(d.north) + (8pt, 0)$) -- ++(-16pt, 0) -- cycle;
-      \fill [shade, opacity=0.2] (b.east) -- (b.south east) -- (d.west) -- (d.north west) -- cycle;
-      \uncover<4->{
-      \draw [green!80!black] (c) -- (d);
-      }
-
-      \node (abcd) [right=6cm of a.north] {\includegraphics[height=\pht]{stock/person}};
-      \node at (abcd.south) [label] {\Large a};
-      \uncover<2->{
-      \node (b) [below right=1 and 0.75 of abcd] {\includegraphics[height=\pht]{stock/person}};
-      \node at (b.south) [label] {\Large b};
-      \node (acd) [below left=1 and 0.5 of abcd] {\includegraphics[height=\pht]{stock/person}};
-      \node at (acd.south) [label] {\Large a};
-      }
-      \uncover<3->{
-      \node (cd) [below right=0.5 and 0.5 of acd] {\includegraphics[height=\pht]{stock/person}};
-      \node at (cd.south) [label] {\Large c};
-
-      \node (a) [below left=0.25 and 0.25 of acd] {\includegraphics[height=\pht]{stock/person}};
-      \node at (a.south) [label] {\Large a};
-      }
-      \uncover<4->{
-      \node (c) [below left=0.25 and 0.25 of cd] {\includegraphics[height=\pht]{stock/person}};
-      \node at (c.south) [label] {\Large c};
-      \node (d) [below right=0.4 and 0.4 of cd] {\includegraphics[height=\pht]{stock/person}};
-      \node at (d.south) [label] {\Large d};
-      }
-
-      \uncover<2->{
-      \draw [red] (abcd) -- (b);
-      \draw [-, dashed] (abcd) -- (acd);
-      }
-      \uncover<3->{
-      \draw [blue] (acd) -- (cd);
-      \draw [-, dashed] (acd) -- (a);
-      }
-      \uncover<4->{
-      \draw [-, dashed] (cd) -- (c);
-      \draw [green!80!black] (cd) -- (d);
-      }
-    \end{tikzpicture}
-
-  $\qquad$contact network \hfill transmission tree$\qquad$
-  \end{center}
-\end{frame}
-
-\begin{frame}{Transmission trees shape viral phylogenies}
-  some demo of viruses evolving and being transmitted at the same time
-\end{frame}
-
-\begin{frame}{Can we fit network models with viral phylogenies?}
-  \begin{center}
-  network $\to$ transmission tree $\to$ viral phylogeny \\
-  \hfill\\
-  network $\stackrel{\alert{?}}{\leftarrow}$ transmission tree $\leftarrow$ viral phylogeny \\
-  \end{center}
-\end{frame}
-
-\begin{frame}{Are network parameters identifiable from tree shape?}
-  \begin{minipage}[p][\textheight][t]{\textwidth}
-    \only<1>{\includegraphics[width=0.9\textwidth, trim=0 3in 0 0, clip]{kernel-idea.pdf}}
-    \only<2>{\includegraphics[width=0.9\textwidth, trim=0 1.7in 0 0, clip]{kernel-idea.pdf}}
-    \only<3>{\includegraphics[width=0.9\textwidth, trim=0 0.8in 0 0, clip]{kernel-idea.pdf}}
-    \only<4->{\includegraphics[width=0.9\textwidth, trim=0 0in 0 0, clip]{kernel-idea.pdf}}
-  \end{minipage}
-\end{frame}
+\end{document}
 
@@
 \begin{frame}{Example simulation\ldots}
-  \begin{minipage}[p][\textheight][t]{\textwidth}
-    \begin{center}
-    preferential attachment power \hfill prevalence
-    \only<1>{
-      \includegraphics[width=0.8\textwidth, trim=0 2.5in 0 0, clip]{abc-posterior-example.pdf}
-    }
-    \only<2>{
-      \includegraphics[width=0.8\textwidth]{abc-posterior-example.pdf}
-
-      \vspace{-0.25cm}
-      edges per vertex \hfill total nodes
-    }
-    \end{center}
-  \end{minipage}
+  1d posterior goes here
+  \pause
+%  \begin{minipage}[p][\textheight][t]{\textwidth}
+%    \begin{center}
+%    preferential attachment power \hfill prevalence
+%    \only<1>{
+%      \includegraphics[width=0.8\textwidth, trim=0 2.5in 0 0, clip]{abc-posterior-example.pdf}
+%    }
+%    \only<2>{
+%      \includegraphics[width=0.8\textwidth]{abc-posterior-example.pdf}
+%
+%      \vspace{-0.25cm}
+%      edges per vertex \hfill total nodes
+%    }
+%    \end{center}
+%  \end{minipage}
 \end{frame}
@@
 \begin{frame}{Example simulation continued, $N$ and $I$ confounded}
+  2d posterior goes here
   %\includegraphics[width=\textwidth]{{abc-posterior-2d/1.0_1000_3_5000_0}.pdf}

### 2016-06-29 doc/defence/transtree.tex
working on defence talk

--- 2016-06-28 doc/defence/trees.tex
@@
-\documentclass{beamer}
-\usepackage{tikz}
-\usepackage{gillius}
-\usepackage{abraces}
-\usepackage{etex}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage[backend=biber, style=numeric-comp, sorting=none]{biblatex}
+\begin{frame}{Contact networks shape transmission trees}
+  \vspace{-1cm}
+  \begin{center}
+    \begin{tikzpicture}
+      [every node/.style = {inner sep=0pt},
+       every path/.style = {->, >=stealth, very thick},
+       label/.style = {anchor=south, color=white, inner sep=4pt},
+       time/.style = {anchor=north west, inner sep=2pt}]
+      \def\pht{1cm}
+      \node (a) {\includegraphics[height=\pht]{stock/person}};
+      \node at (a.south) [label] {\Large a};
+      \node [below left=0.75 of a] (b) {\includegraphics[height=\pht]{stock/person}};
+      \node at (b.south) [label] {\Large b};
+      \node [below=0.75 of b] (e) {\includegraphics[height=\pht]{stock/person}};
+      \node at (e.south) [label] {\Large e};
+      \node [below right=0.75 of a] (c) {\includegraphics[height=\pht]{stock/person}};
+      \node at (c.south) [label] {\Large c};
+      \node [below=0.75 of c] (d) {\includegraphics[height=\pht]{stock/person}};
+      \node at (d.south) [label] {\Large d};
 
-% http://tex.stackexchange.com/questions/12703/how-to-create-fixed-width-table-columns-with-text-raggedright-centered-raggedlef
-\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+      \fill [shade, opacity=0.2] (a.east) -- (c.north) -- (c.north west) -- (a.south east) -- cycle;
+      \fill [shade, opacity=0.2] (a.west) -- (b.north) -- (b.north east) -- (a.south west) -- cycle;
+      \fill [shade, opacity=0.2] (a.south) -- (a.south west) -- (e.north) -- (e.north east) -- cycle;
+      \fill [shade, opacity=0.2] (b.south west) -- (b.south) -- (e.north) -- (e.north west) -- cycle;
+      \fill [shade, opacity=0.2] ($(c.south) + (-8pt, 0)$) -- ++(16pt, 0) -- ($(d.north) + (8pt, 0)$) -- ++(-16pt, 0) -- cycle;
+      \fill [shade, opacity=0.2] (b.east) -- (b.south east) -- (d.west) -- (d.north west) -- cycle;
+      \uncover<2->{
+      \draw [red] ($(a.west) + (0, -12pt)$) -- ($(b.north) + (8pt, 0)$);
+      }
+      \uncover<4->{
+      \draw [blue] ($(a.east) + (0, -12pt)$) -- ($(c.north) + (-8pt, 0)$);
+      }
+      \uncover<6->{
+      \draw [green!80!black] (c) -- (d);
+      }
 
-\usetikzlibrary{external}
-\usetikzlibrary{shapes}
-\usetikzlibrary{arrows}
-\usetikzlibrary{positioning}
-\usetikzlibrary{decorations.pathreplacing}
-\usetikzlibrary{calc}
-\usetikzlibrary{automata}
-
-\usetheme[everytitleformat=regular]{m}
-\setbeamertemplate{navigation symbols}{}
-
-\graphicspath{{../figures/}}
-\newcommand{\tablepath}{../tables}
-
-\title{Phylogenetic inference of contact network parameters with approximate Bayesian computation}
-\author[RMM \& AFYP]{Rosemary M McCloskey$^1$ \and Bioinformatics Training Program \\ Supervisor: Art FY Poon$^{1,2}$}
-\institute[UBC \& BCCfE]{$^1$BC Centre for Excellence in HIV/AIDS, Vancouver, Canada \\ $^2$Department of Medicine, University of British Columbia, Vancouver, Canada}
-\date{M.Sc. Defence, July 26, 2016}
-
-\newcommand{\dd}[2]{\frac{\text{d}\,#1}{\text{d}\,#2}}
-
-\addbibresource{papers.bib}
-
-\begin{document}
-\setbeamercolor{background canvas}{bg=white}
-\renewcommand{\footnotesize}{\tiny}
-\definecolor{red}{RGB}{228,26,28}
-\definecolor{blue}{RGB}{55,126,184}
-\definecolor{green}{RGB}{77,175,74}
-\definecolor{purple}{RGB}{152,78,163}
-
-\begin{frame}{Transmission trees shape viral phylogenies}
-  \begin{center}
-    \begin{tikzpicture}[
-        every node/.style = {inner sep=0pt},
-        every path/.style = {->, >=stealth, very thick},
-        label/.style = {anchor=south, color=white, inner sep=4pt}
-      ]
-      \def\pht{1cm}
-      \node (abcd) {\includegraphics[height=\pht]{stock/person}};
+      \node (abcd) [above right=0 and 5cm of a] {\includegraphics[height=\pht]{stock/person}};
       \node at (abcd.south) [label] {\Large a};
@@
 
-      \coordinate [below right=0.5 and 3 of b] (abcd);
-      \coordinate [below right=1 and 1 of abcd] (b);
-      \coordinate [above left=1 and 0.25 of abcd] (acd);
-      \coordinate [above left=1 and 0.25 of acd] (a);
-      \coordinate [right=of acd] (cd);
-      \coordinate [right=1.5 of cd] (c);
-      \coordinate [above right=of cd] (d);
-
-      \uncover<1-2>{
-      \node (x) at (abcd) {\includegraphics[height=\pht]{stock/virus}};
-      \node at (x.center) {\large a};
-      }
-
-      \uncover<2>{
-      \node (x) [below right=0.3 of abcd, anchor=north west] {\includegraphics[height=\pht]{stock/virus}}; 
-      \node at (x.center) {\large b};
-      }
-
-      \uncover<3-4>{
-      \node (x) [above=0 of acd, anchor=south] {\includegraphics[height=\pht]{stock/virus}}; 
-      \node at (x.center) {\large a};
-      }
-
-      \uncover<3->{
-      \node (x) [below right=-0.2 of b, anchor=north west] {\includegraphics[height=\pht]{stock/virus}}; 
-      \node at (x.center) {\large b};
-
-      \draw (abcd) -- (acd);
-      \draw (abcd) -- (b);
-      }
-
-      \uncover<4>{
-      \node (x) [right=0.3 of acd, anchor=west] {\includegraphics[height=\pht]{stock/virus}}; 
-      \node at (x.center) {\large c};
-      }
-
-      \uncover<5-6>{
-      \node (x) at (cd) [anchor=west] {\includegraphics[height=\pht]{stock/virus}};  
-      \node at (x.center) {\large c};
-      }
-
-      \uncover<5->{
-      \node (x) at (a) [anchor=south] {\includegraphics[height=\pht]{stock/virus}};
-      \node at (x.center) {\large a};
-
-      \draw (acd) -- (cd);
-      \draw (acd) -- (a);
-      }
-
-      \uncover<6>{
-      \node (x) [above right=0.4\pht and 0.4\pht of cd] [anchor=south west] {\includegraphics[height=\pht]{stock/virus}};
-      \node at (x.center) {\large d};
-      }
-
-      \uncover<7->{
-      \node (x) at (c) [anchor=west] {\includegraphics[height=\pht]{stock/virus}};
-      \node at (x.center) {\large c};
-
-      \node (x) at (d) [anchor=south west] {\includegraphics[height=\pht]{stock/virus}};
-      \node at (x.center) {\large d};
-
-      \draw (cd) -- (c);
-      \draw (cd) -- (d);
-      }
+      \coordinate [above right=0 and 2.5cm of abcd] (top);
+      \draw (top) -- ++ (0, -6cm) node [anchor=west, inner xsep=4pt] {time};
 
     \end{tikzpicture}
+
+  $\qquad$contact network \hfill transmission tree$\qquad$
   \end{center}
 \end{frame}
-
-\end{document}

### 2016-06-29 doc/defence/virusphylo.tex
working on defence talk

--- 2016-06-28 doc/defence/trees.tex
@@
-\documentclass{beamer}
-\usepackage{tikz}
-\usepackage{gillius}
-\usepackage{abraces}
-\usepackage{etex}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage[backend=biber, style=numeric-comp, sorting=none]{biblatex}
-
-% http://tex.stackexchange.com/questions/12703/how-to-create-fixed-width-table-columns-with-text-raggedright-centered-raggedlef
-\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-
-\usetikzlibrary{external}
-\usetikzlibrary{shapes}
-\usetikzlibrary{arrows}
-\usetikzlibrary{positioning}
-\usetikzlibrary{decorations.pathreplacing}
-\usetikzlibrary{calc}
-\usetikzlibrary{automata}
-
-\usetheme[everytitleformat=regular]{m}
-\setbeamertemplate{navigation symbols}{}
-
-\graphicspath{{../figures/}}
-\newcommand{\tablepath}{../tables}
-
-\title{Phylogenetic inference of contact network parameters with approximate Bayesian computation}
-\author[RMM \& AFYP]{Rosemary M McCloskey$^1$ \and Bioinformatics Training Program \\ Supervisor: Art FY Poon$^{1,2}$}
-\institute[UBC \& BCCfE]{$^1$BC Centre for Excellence in HIV/AIDS, Vancouver, Canada \\ $^2$Department of Medicine, University of British Columbia, Vancouver, Canada}
-\date{M.Sc. Defence, July 26, 2016}
-
-\newcommand{\dd}[2]{\frac{\text{d}\,#1}{\text{d}\,#2}}
-
-\addbibresource{papers.bib}
-
-\begin{document}
-\setbeamercolor{background canvas}{bg=white}
-\renewcommand{\footnotesize}{\tiny}
-\definecolor{red}{RGB}{228,26,28}
-\definecolor{blue}{RGB}{55,126,184}
-\definecolor{green}{RGB}{77,175,74}
-\definecolor{purple}{RGB}{152,78,163}
-
 \begin{frame}{Transmission trees shape viral phylogenies}
@@
       }
-
     \end{tikzpicture}
@@
 \end{frame}
-
-\end{document}

### 2016-06-29 doc/ismb/abstract.tex
working on defence talk

In rapidly evolving organisms such as RNA viruses, evolutionary and
epidemiological dynamics occur on the same time scale. Phylodynamics attempt to
detect and quantify the signatures of epidemiological processes in viral
phylogenies. Due to the rapid advancement of nucleotide sequencing technology,
viral sequence data data have become increasingly feasible to collect on a
population level. Through phylodynamic methods, these data offer a window into
epidemiological processes which would otherwise be virtually impossible to
study on a realistic scale. However, the vast majority of phylodynamic methods
assume a homogeneously mixed host population, where every pair of individuals
is equally likely to come into contact. This assumption is clearly unrealistic
for most human communities.

We developed a method based on kernel approximate Bayesian computation to fit
contact network models to viral phylogenies without calculating intractible
likelihoods. Our method can be used to fit any network model from which
simulated networks can be generated. We applied our method to investigate a
preferential attachment model using simulated and real-world HIV sequence
datasets. On simulated data, we found that the preferential attachment power
and the number of infected nodes could often be accurately estimated, while the
edge density and total number of nodes were more challenging to determine. We
found significant heterogeneity in the networks underlying several real-world
datasets, with estimated preferential attachment power ranging from 0.35 to
1.16. These results underscore the importance of taking contact structure into
account when investigating viral epidemics, and suggest kernel-ABC as an
effective tool for such investigations.

### 2016-06-29 doc/ismb/poster.tex
working on defence talk

--- 2016-04-20 doc/big-retreat/poster.tex
@@
 
-\title{Phylogenetic estimation of contact network parameters with kernel-ABC}
-\author{Rosemary M. McCloskey$^1$ and Art F. Y. Poon $^{1,2}$}
+\title{Likelihood-free estimation of contact network parameters from viral phylogenies}
+\author{Rosemary M. McCloskey$^1$, Richard H. Liang$^1$, and Art F. Y. Poon $^{1,2}$}
 \date{March 11, 2015}
@@
 
+\colorlet{blockbodybgcolor}{white}
 \node at (current page.west) 
@@
 \node [text width=6in] at ($ (current page.west) + (-0.42\paperwidth, -1.5in) $)
-  { \LARGE B.I.G. Research Day \\ March 11, 2016 \\ Vancouver, Canada};
+  { \LARGE ISMB \\ July 8-12, 2016 \\ Orlando, USA};
 \node [text width=6in, align=right] at ($ (current page.west) + (0.42\paperwidth, -1.5in) $)
@@
   \block{Barab\'asi-Albert preferential attachment networks}{
-    Nodes are added one-at-a-time until there are $N$ total. Each time a node
+    Nodes are added one at a time until there are $N$ total. Each time a node
     is added, $m$ new edges are connected to it. Each new edge is attached to
@@
     \vspace{1cm}
-    \centerline{\includegraphics[width=0.2\textwidth, trim=0 22in 0 0, clip]{alpha-bounds}}
+    \centerline{\includegraphics[width=0.2\textwidth]{pa-example}}
+    \vspace{1cm}
 
@@
   \block{Accurate estimates for $\alpha$, $I$; not $m$, $N$}{
-    \centerline{\includegraphics[width=0.16\textwidth]{abc-point-estimate}}
+    \centerline{\includegraphics[width=0.16\textwidth]{abc-point-estimate-m2}}
     Dotted lines are true values. $\alpha$ and $I$, which had the most impact
@@
   \block{Transmission trees were estimated from viral sequences}{
-    \centerline{\includegraphics[width=0.2\textwidth]{pipeline}}
+    \vspace{1cm}
+    \centerline{\includegraphics[width=0.25\textwidth]{pipeline}}
+    \vspace{1cm}
 
@@
   \block{Parameter heterogeneity among real datasets}{
-    \centerline{\includegraphics[width=0.2\textwidth]{realdata-hpd}}
+    \vspace{1cm}
+    \centerline{\includegraphics[width=0.25\textwidth]{realdata-hpd}}
+    \vspace{1cm}
   }

### 2016-06-29 doc/figures/objects.R

@@ -1,7 +1,8 @@
 #!/usr/bin/env Rscript
 
-suppressPackageStartupMessages(library(netabc))
-library(RColorBrewer)
+suppressPackageStartupMessages(library(ape))
+suppressPackageStartupMessages(library(igraph))
+library(png)
 
 n <- 40
 t <- 8
@@ -10,8 +11,20 @@ trans.tree <- ladderize(rcoal(t))
 phylo <- ladderize(rtree(t))
 
 pdf("objects.pdf", width=7, height=2)
-par(mfrow=c(1, 3), mar=c(0, 0, 0, 0))
+par(mfrow=c(1, 4), mar=c(0, 0, 0, 0))
 plot(net, vertex.label=NA, vertex.size=3, vertex.color="black")
 plot(trans.tree, show.tip.label=FALSE, direction="down", edge.width=2)
 plot(phylo, type="unrooted", show.tip.label=FALSE, edge.width=2)
+plot.new()
+y <- seq(0.1, 0.9, length.out=5)
+segments(0.25, y, 1, y, lwd=2)
+
+nmut <- 12
+xmut <- sample(seq(0.25, 1, 0.01), nmut, replace=TRUE)
+ymut <- sample(y, nmut, replace=TRUE)
+segments(xmut, ymut-0.03, xmut, ymut+0.03, lwd=2)
+
+virus <- readPNG("stock/virus.png")
+rasterImage(virus, 0, y-0.08, 0.18, y+0.08)
+
 dev.off()

### 2016-06-30 doc/paper/changes.tex
paper edits and response letter

\documentclass[12pt]{letter}
\usepackage[T1]{fontenc}
\usepackage{mathptmx}
\usepackage[garamondx,cmbraces]{newtxmath}

\signature{Rosemary M. McCloskey}
\address{Centre for Excellence in HIV/AIDS \\ 608-1081 Burrard Street \\ Vancouver, BC, Canada V6Z 1Y6}
\begin{document}
\begin{letter}{ }

\opening{Dear editors of \textit{Virus Evolution},}

Thank you for the opportunity to revise our manuscript ``Reconstructing contact
network parameters from viral phylogenies''.

While our submission was pending, we made some additions and improvements to
the work that we would like you and the reviewers to consider. It was pointed
out to us by a colleague that the posterior mean is a more conventionally used
estimator than the maximum \textit{a posteriori} (MAP) value. Therefore, we
have replaced the MAP point estimates by posterior means. Qualitatively, the
results are not much different, except that it is more obvious that $m$ and $N$
are not identifiable with this method. We have replaced some ad hoc statistical
tests in the analysis of the simulation results with a $p$-value corrected
generalized linear model, and rewritten the relevant section of the results
accordingly. For clarity, the real HIV datasets are now referred to as their
risk group and location in the text, rather than by their author (eg.
MSM/Beijing instead of Wang et al.). We have found an analysis of the
Barab\'asi-Albert model in the physics literature which demonstrates that the
degree distribution is only a power law when $\alpha = 1$. We acknowledge this
but do not believe it detracts from the validity of our results, especially in
light of a paper by de Blasio et al. (2007) which found sub-linear preferential
attachment through a population-level survey.

We did some additional literature searching and found five more published HIV
datasets that seemed to be appropriate for our method. The highest $\alpha$
values were associated with IDU networks, coroborating the assertion of
Dombrowski et al. (2013) that IDU networks exhibit the scale-free property more
strongly than sexual networks do. Three of these new datasets had sequenced two
HIV genes, which we analysed separately. We found that estimates of $\alpha$
were very robust to the choice of gene; $I$ was less robust, with \textit{env}
genes tending to lead to higher estimated values.

Our responses to each of the reviewers' comments are in line below. We were
asked by the Editor in Chief to consider how to disambiguate ABC approximation
error from a truly noisy posterior, which this is addressed under the first
point brought up by Reviewer 1.

\textbf{Reviewer 1}

\begin{quote}
  \itshape

  I think the question of how we might learn about contact network structure
  from trees -- transmission or phylogenetic -- is interesting, and it is
  refreshing to see a non-BEAST approach to a phylodynamic question (nice
  though BEAST is, one of several limitations is its unsuitability to large
  datasets, for example). \\

  Overall, though, it seems that the conclusion is that in this simulation
  study, transmission trees from the BA network in these conditions don't carry
  sufficient information about the network parameters to be very informative.
  Alpha is most robustly identified, perhaps because of the dramatic effect of
  alpha on gamma (fig S6) so that alpha < ~1.2 would have a very different
  gamma than alpha > ~1.2. The HIV datasets yielded wide and somewhat varying
  estimates. The authors concluded something about contact heterogeneity, but
  is it robust inference of heterogeneous contact, or just poor posterior
  resolution?
\end{quote}

We are grateful that the reviewer finds our work to be of some interest, and we
acknowledge that the identifiability of the BA parameters from tree shapes is
somewhat limited. TODO

We have included several new analyses that attempt to disambiguate true
features of the posterior distribution from artefacts of a poor ABC
approximation. First, we used ABC to estimate marginal posterior distributions
when some combinations of parameters were held fixed to known values.
Estimation error which is removed during marginal estimation is likely to be
due to confounding between model parameters; error which remains may be due to
either poor resolution or truely diffuse posteriors. TODO

Second, we examined the possible Monte Carlo error by increasing the resolution
of the ABC approximation (more particles, more simulated datasets per particle,
and more iterations). The results were nearly identical, TODO

Finally, three of the newly included HIV datasets included both \textit{gag}
and \textit{env} sequences which we analysed separately. The estimates of
$\alpha$ for the two genes were nearly identical for each of the datasets,
which seems unlikely if the ABC approximations were very poor.

\begin{quote}
  \itshape

  I appreciate the detail and clarity in the Methods section. One concern I had
  was the use of $R^2$. As many quantities can be called $R^2$, and you have
  categorical data, how have you defined it, and why is it reflective of
  goodness of fit? Usually an $R^2$ (linear model?) is not considered the best
  way to test categorical relationships. How much does the linear assumption
  behind $R^2$ impact the quality of the reported results?
\end{quote}

We apologise for the misunderstanding. Our choice of words was poor -- TODO

We have redone this analysis using both SVMs (classifiers) and SVRs
(regressions). Figure 1 has been expanded to show both sets of results --
accuracy for the SVMs and $R^2$ for the SVRs. The results are qualitatively
very similar.

\begin{quote}
  \itshape

  Could you describe how the cross-validation was done (numbers training vs test etc)?
\end{quote}

We have updated the text as follows (addition in italics): ``The accuracy of
each classifier was evaluated with 1000 two-fold cross validations \emph{with
equally-sized folds}.''

\begin{quote}
  \itshape

  Please describe how you extract the transmission trees -- are they binary
  trees, with individual transmission events resolved in real time? Or are they
  not binary trees, eg case 1 infecting cases 2, 3, 4, 5 corresponding to an
  internal node with degree 5 (in-degree 1, out-degree 4)? If the latter, it's
  not really comparable to using HIV phylogenies as these will be binary.
\end{quote}

The method we used to estimate transmission trees simply reroots and adjusts
the branch lengths of an estimated viral phylogeny, which is always binary as
long as there are no duplicated sequences. We have added the following
clarifications in the methods section: (1) ``Duplicated sequences were removed
with BioPython'' (2) ``Due to the removal of duplicated sequences, all
estimated transmission trees were fully binary.''

\begin{quote}
  \itshape

  Is inferring m any easier for fixed alpha, N, I than varying? You note in the
  Discussion that inferring N is optimistic (and this seems particularly true
  if you stop the simulation when either I or the target number of tips is
  reached), and you briefly comment about the relationships between the
  variables. You might be able to get stronger inference if some parameters
  were fixed. Anyway it would be nice to see some joint posteriors with two
  parameters, to see if some of the posterior uncertainty is due to
  correlations.
\end{quote}

We thank the reviewer for this helpful suggestion. We individually fixed each
combination of parameters and performed ABC inference on the remaining free
parameters. The inference of $I$ was dramatically improved when $N$ was fixed.
Fixing each of the parameters contributed slightly to the estimate of $\alpha$,
but none as dramatically. However, the posterior mean estimate of $m$ was
totally unaffected by fixing the other parameters. This indicates that our
inability to estimate $m$ is simply due to lack of identifiability, rather than
to parameter correlations.

As the reviewer suggested, we have also included a set of two-dimensional
posterior distributions in the text (Figure TODO). The correlation between $N$
and $I$ demonstrated in the marginal analysis is clearly observable, but there
are no obvious correlations among the other parameters.

\begin{quote}
  \itshape

  You found that m is hard to infer. Could this be because with pre-determined
  I, even if m increases, only so many edges per host will actually be active
  in transmitting infection? Overall I would have been interested in more
  intuitive-level discussion of why the results were what they were.
\end{quote}

  TODO

\begin{quote}
  \itshape

  You qualitatively describe errors, and show lots of posteriors, but a direct
  visual quantification of how well the mean / mode /etc of the posterior
  reflects the true value would be helpful - for example as a box plot, where
  you could then illustrate all the tests in the same figure, stratified as you
  like.
\end{quote}

We agree with this assessment of our visual presentation of the results. We
have reduced the number of posterior plots shown to one in the text and two in
the supplemental materials. TODO

\begin{quote}
  \itshape

  The Discussion focusses around model mis-specification, which is very
  relevant (and to which the inference isn't that robust, suggesting that it
  might also not be robust to other things like dynamics, clustering,
  age-related mixing etc). But another question is how much of the signal you
  would expect to find in this simulation strategy, and how you would expect to
  see it using these kinds of tree comparisons. I don't have much intuition
  about the kernel but the Sackin imbalance will presumably be lower where
  there are very high-degree nodes, and higher where there are long
  transmission chains. The nLTT will depend on the epidemic curves, presumably.
\end{quote}

The identifiability of the BA parameters is of course a critical prerequisite
to the kind of inference we are trying to perform. Indeed, the objective of the 
analysis with classifiers was to investigate identifiability, and we apologise
for not making this clearer. The reviewer's intuition is correct -- we do
expect imbalance to be sensitive to the degree distribution (related to
$\alpha$) and the nLTT to be sensitive to the epidemic curves (related to $I$
and $N$). This is reflected in the performance of the respective classifiers:
Sackin's index does better than the nLTT at classifying $\alpha$, but the nLTT
is better at classifying $I$ and $N$.

We have included some expository text near the start of the methods section to
emphasize these points. In particular, ``Sackin's index (Shao 1990)
is a measure of tree imbalance which not take branch lengths into account,
considering only the topology. The normalized lineages-through-time (nLTT,
Janzen et al. 2015) compares two trees based on normalized distributions of
their branching times, and does not explicitly consider the topology. Since the
tree kernel incorporates both of these sources of information, we expected it
to outperform the other two statistics.''

\begin{quote}
  \itshape

  ABC is computationally intensive, as you note, but does it perform better
  than classification? Enough better to be worth it?
\end{quote}

Our classification experiments were marginal -- they only estimated one
parameter at a time, when all others were fixed and known. In this situation,
something like k-nearest-neighbours classification or grid search might be just
as effective as ABC, and will certainly be faster. The justification for our
method comes in overcoming the curse of dimensionality when jointly estimating
multiple parameters of a network model. The text added to address the previous
point makes it clearer that the classifier experiments were marginal in nature.
(TODO: We have emphasized this point in the discussion as well: ).

\textbf{Reviewer 2}

\begin{quote}
  \itshape
  The paper describes a study using ABC to estimate the parameters of contact
  networks from transmission trees, using phylogenies as a proxy for the
  latter. The paper is well-written and the method and results are clearly
  presented. While it must be said that the method as presented appears to lack
  accuracy in the reconstruction of many features of an epidemic, the authors
  do not pretend otherwise. I regard this as an important class of models whose
  development should be encouraged, and this is one of the first few such
  examples. As a result, I am happy to recommend publication with some
  revisions. \\

  General comment: This paper was clearly written with results coming before
  methods and wasn't adequately reformatted; various mathematical symbols are
  mentioned in the methods but only defined later. As a result, both methods
  and results sections need another look.
\end{quote}

We are quite embarrassed by this oversight and have revised the text
accordingly. In particular, the first three paragraphs of the methods section
define the BA parameters ($\alpha$, $I$, $m$, and $N$) and introduce the
meta-parameters to the tree kernel.

\begin{quote}
  \itshape

  P2 \\
  L19: This is the effective size of the viral population, so best to specify
  that explicitly given work on the ``effective number of infections''.
\end{quote}

\begin{quote}
  \itshape

  L25-26: ``Straightforward to collect'' is a bit of a strong statement. While
  sequencing is getting easier, representative sample collection may still be
  challenging.
\end{quote}

We have revised the statement as follows: ``Viral sequence data, on the other
hand, has become easier to collect as the cost of sequencing has declined. In
the case of HIV, genotyping has become part of routine clinical care in several
health regions.'' We have also added the following qualification to the end of
the paragraph: ``In the context of networks, sequence data have the advantage
of being objective, in that they are not affected by misreporting. However,
just as with survey data, it is important to collect a representative sample
from the population to perform accurate inference (Novitsky et al. 2014).''

\begin{quote}
  \itshape

  P4\\
  L53-54: The y-axis does not match closely with the Leventhal figure; is this
  an error or genuine? If the latter, please comment.
\end{quote}

We thank the reviewer for their sharp eye. This was indeed an error on our
part. Leventhal et al. constructed Watts-Strogatz (``small world'') networks by
rewiring each edge with probability $p = 0.01$. However, in the implementation
we used, $p$ is the probability of rewiring each \emph{endpoint}, not each
edge. Therefore, when we took $p = 0.01$, the net effect was that each edge had
probability $2p = 0.02$ to be rewired (the sum of the probabilities of each
endpoint). We remade the figure with $p = 0.005$, which should match up with $p
= 0.01$ from Leventhal et al., and the scales are now the same. The lines
corresponding to the Erd\H{o}s-R\'enyi and Barab\'asi-Albert networks are
unchanged.

\begin{quote}
  \itshape

  P5 \\
  L45 How were the tips selected? I'm assuming randomly, but make this explicit.
\end{quote}

The reviewer's assumption is correct. We have clarified this in the section
describing \textit{netabc}, since it is an aspect of the method itself and
applies to all analyses: ``Tips of the simulated transmission tree are randomly
removed until the simulated tree has the same number of tips as the input
tree.''

\begin{quote}
  \itshape

  L45-46: “…and in epidemics of differing size (500, 1000 and 2000)”. Unless
  I’m missing something this is just a restatement of what’s already in this
  paragraph about I, hence repetitive. 
\end{quote}

We apologise for the misunderstanding. Because $I$ is not a structural
parameter of the network, we attempted to classify each of $\alpha$, $N$, and
$m$ under several different fixed values of $I$. We have rephrased this as
follows: ``For the structural parameters $\alpha$, $m$, and $N$, the
experiments were repeated with three different fixed values of $I$ (500, 1000,
and 2000).''

\begin{quote}
  \itshape

  P7 \\
  L20-24: It’s slightly surprising to me that the time trees were constructed
  using “fast” methods when the datasets are not especially large. Were there
  problems in obtaining trees with other algorithms? While I do not insist that
  every analysis be redone using alternative methods, it would be reassuring to
  know that the results were consistent, in at least one example, if
  reconstruction were done with, for example, RAxML+LSD, or by taking the BEAST
  MCC tree. Similarly, are results similar if analysis is performed on several
  bootstraps or posterior samples?
\end{quote}

Although they are very approximate methods, our experience has been that using
approximate maximum likelihood phylogeny reconstruction with root-to-tip
regression does not result in large errors in phylodynamic inference (Poon et
al. 2015). We reanalysed the datasets with the highest and lowest estimated
$\alpha$ (Zetterberg et al. 2004 and Cuevas et al. 2009) using a BEAST MCC tree 
from a run with simple models of evolution and population size. The results
were TODO

\begin{quote}
  \itshape

  P8 \\
  L15: Higher values of sigma do seem to consistently lead to worse estimates
  of alpha, and estimates are better on smaller datasets. Is there an
  explanation for this?
\end{quote}

TODO dunno man

\begin{quote}
  \itshape

  P9 \\
  Figure 1: Could this be in colour rather than shades of grey?
\end{quote}

The figure is now in color.

\begin{quote}
  \itshape

  L32: “Estimated correctly” – i.e. the posterior mode was correct?
\end{quote}

This statement has been removed in our updated version. When we calculated
posterior means rather than modes, as is more conventional, we found that
nearly all of the estimates were very near to the middle of the range. This is
consistent with our observed lack of signal for $m$ in phylogenetic data.

\begin{quote}
  \itshape

  P10-11 \\
  I think I might prefer that a firm decision on whether it is preferable to
  allow m=1 be made and justified and the results of that analysis be presented
  first and foremost, with the alternative results also presented, but in a
  separate paragraph.
\end{quote}

Since several of the datasets had the majority of their posterior mass on $m =
1$, it seemed that this choice of prior was more reasonable, and we therefore
focused our discussion on those results.

\begin{quote}
  \itshape

  P12 \\
  Figure 2: Since there were three simulations done of each type, presumably
  the variable number of points in the top right figure is due to the existence
  of duplicate values. But this needs to be clearer. Also presumably the points
  from the same sort of simulation were jittered so they appear in discrete
  columns, but this makes the regression line perhaps a little misleading.
  Perhaps this figure could also be made a little larger?
\end{quote}

We agree that the sloped lines were misleading and have replated them with
horizontal line segments. TODO

\begin{quote}
  \itshape

  P13 \\
  Figure 3, legend: I’m guessing that the shaded polygon is the HPD region,
  rather than the “true value”?
\end{quote}

This is correct and the caption has been updated as follows: ``shaded areas
show 95\% highest posterior density intervals.''

\begin{quote}
  \itshape

  P15 \\
  L38-40: While I understand the point, it’s surely quite unlikely that the
  network will be identical. Better to say that the number of infected nodes
  with new children is small, and hence there is little difference in the
  outcomes of the simulations.
\end{quote}

We struggled with the phrasing of this particular point and are happy to have
recieved some guidance. TODO

\begin{quote}
  \itshape

  L41-42: “We note also that our accurate estimates of I may have been
  influenced by this prior”. Couldn’t you check?
\end{quote}

TODO

\begin{quote}
  \itshape

  P16 \\
  L9-27: As stated above, I'd prefer your preference for whether to include m=1
  to be stated and justified.
\end{quote}

As mentioned above, we have focused our discussion on $m = 1$ due to the
several datasets for which $m = 1$ contained most of the posterior mass.

\begin{quote}
  \itshape

  P17
  L1: Presumably by estimated prevalence you mean just I/N?
\end{quote}

We had been using the term ``prevalence'' interchangably with $I$, but since
this is ambiguous we have replaced all instances of ``prevalence'' with $I$
except for those which do not refer to $I$ explicitly.

\begin{quote}
  \itshape

  A point I’d like to see at least discussed – presumably if some iterations
  add nodes with one connection and others add them with two, you’d get cycles,
  but not as many as you’d expect with m=2. Is this worth considering in the
  question of why the binary choice to include m=1 is so problematic? Would it
  be better if m was drawn from a discrete distribution at every step?
\end{quote}

This is a valid point, and in fact drawing $m$ from a distribution is an
approach we considered when we were performing this work. However, since this
is our first attempt to fit network models, we thought it best to stick with
a simple, well-known model rather than defining our own parameters. (TODO: We
have added to the discussion the point that this binary choice is sub-optimal
and suggested this enhancement for future work.)

\begin{quote}
  Supplement:
  Figure S1 legend: What is the shaded area?
\end{quote}

The original figure in Leventhal et al. has two shaded regions, one
representing the interquartile range, and the other representing the 95\%
quantile range. We have edited the figure to contain both regions and
identified them in the figure legend.

\closing{Thank you for your consideration,}
\end{letter}
\end{document}

### 2016-06-30 doc/paper/paper.Rnw
paper edits and response letter

@@
 
-Viral sequence data, on the other hand, has become relatively inexpensive and
-straightforward to collect on a population level. Due to the high mutation rate
-of RNA viruses, epidemiological processes impact the course of viral evolution,
-thereby shaping the inter-host viral
+Viral sequence data, on the other hand, has become \del{relatively inexpensive
+and straightforward to collect on a population level.} \add{easier to collect
+as the cost of sequencing has declined. In the case of HIV, genotyping has
+become part of routine clinical care in several health regions.'' Due to the
+high mutation rate of RNA viruses, epidemiological processes impact the course
+of viral evolution, thereby shaping the inter-host viral
 phylogeny~\autocite{drummond2003measurably}. The term ``phylodynamics'' was
@@
 villandre2016assessment}, suggesting that such statistical inference might be
-possible~\autocite{welch2011statistical}.
+possible~\autocite{welch2011statistical}. \add{In the context of networks,
+sequence data have the advantage of being objective, in that they are not
+affected by misreporting. However, just as with survey data, it is important to
+collect a representative sample from the population to perform accurate
+inference~\autocite{novitsky2014impact}.}
 
@@
 several previous studies \autocite[\textit{e.g.}][]{robinson2013dynamics,
-leventhal2012inferring}. The particles are weighted according to the similarity
-between their associated simulated trees and the observed tree. To quantify
-this similarity, we used the tree kernel developed by
-\textcite{poon2013mapping}. Particles are iteratively perturbed by applying a
-Metropolis-Hastings kernel and, if the move is accepted, simulating new
-datasets under the new parameters. When a particle's weight drops to zero,
+leventhal2012inferring}. \add{Tips of the simulated transmission tree are
+randomly removed until the simulated tree has the same number of tips as the
+input tree.} The particles are weighted according to the similarity between
+their associated simulated trees and the observed tree. To quantify this
+similarity, we used the tree kernel developed by \textcite{poon2013mapping}.
+Particles are iteratively perturbed by applying a Metropolis-Hastings kernel
+and, if the move is accepted, simulating new datasets under the new parameters.
+When a particle's weight drops to zero,
 because its simulated trees are too dissimilar to the observed tree, the
@@
 distributions of their branching times, and does not explicitly consider the
-topology. In addition, the tree kernel can be tuned by adjusting the values of
-the meta-parameters $\lambda$ and $\sigma$ (the ``decay factor'' and ``radial
-basis function variance'', see \textcite{poon2013mapping}). The results of this
-experiment were used to select values for these meta-parameters to carry
-forward based on their accuracy in distinguishing the different trees.}
+topology. Since the tree kernel incorporates both of these sources of
+information, we expected it to outperform the other two statistics. Finally,
+the tree kernel can be tuned by adjusting the values of the meta-parameters
+$\lambda$ and $\sigma$ (the ``decay factor'' and ``radial basis function
+variance'', see \textcite{poon2013mapping}). The results of this experiment
+were used to select values for these meta-parameters to carry forward based on
+their accuracy in distinguishing the different trees.}
 
@@
 regression from the Sackin's index values. The accuracy of each classifier was 
-evaluated with 1000 two-fold cross validations.
+evaluated with 1000 two-fold cross validations \add{with equally-sized folds}.
 
@@
 \add{For the structural parameters $\alpha$, $m$, and $N$, the experiments were
-repeated with three different values of $I$ (500, 1000, and 2000).} \del{and in
-epidemics of differing size (500, 1000, and 2000).} The combination of the
-number of sampled individuals (\textit{i.e.} the number of tips) and the
+repeated with three different fixed values of $I$ (500, 1000, and 2000).}
+\del{and in epidemics of differing size (500, 1000, and 2000).} The combination
+of the number of sampled individuals (\textit{i.e.} the number of tips) and the
 epidemic size (\textit{i.e.} $I$) will be referred to as an ``epidemic
@@
 \autocite{edgar2004muscle}, and alignments were manually inspected with
-\textit{Seaview} version 4.4.2 \autocite{gouy2010seaview}. Phylogenies were
-constructed from the nucleotide alignments by approximate maximum likelihood
-using \textit{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
+\textit{Seaview} version 4.4.2 \autocite{gouy2010seaview}. \add{Duplicated
+sequences were removed with BioPython.} Phylogenies were constructed from the
+nucleotide alignments by approximate maximum likelihood using
+\textit{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
 generalized time-reversible (GTR) model~\autocite{tavare1986some}. Transmission
@@
 BEAST~\autocite{drummond2007beast}) as described
-previously~\autocite{poon2015phylodynamic}.
+previously~\autocite{poon2015phylodynamic}. \add{Due to the removal of
+duplicated sequences, all estimated transmission trees were fully binary.}
 
@@
 only weakly identifiable between the values of 0 and 0.5 (\cref{fig:abcptm2}).
-The error in the estimated prevalence $I$ was slightly lower for smaller values
-of $\alpha$
+The error in the estimated $I$ value was slightly lower for smaller values of
+$\alpha$
     ($p \Sexpr{pp(I.glm[Parameter == "alpha", p], eq=TRUE)}$)
@@
 \add{For all but the HET/Botswana data, the posterior mean estimates for
-the prevalence $I$ were between
+$I$ were between
     \Sexpr{pmean1["zetterberg2004", round(I)]}
@@
 entire prior region (\cref{fig:abchpd}). There was no significant correlation
-between the number of sequences in the tree and the estimated prevalence
+between the number of sequences in the tree and the estimated $I$ value
 (Spearman correlation, 
     $p \Sexpr{pp(pmean1[,cor.test(ntip, I, method="spearman")[["p.value"]]], eq=TRUE)}$),
-indicating that the higher prevalence estimates were not simply due to
-increased sampling density. When both \textit{gag} and \textit{env} sequences
-were analyzed, the estimates from the \textit{env} data were substantially
-higher (\cref{fig:genes}). The ratios of internal to terminal branch lengths
-were lower for the \textit{env} trees than the \textit{gag} trees (average 
+indicating that the higher estimates were not simply due to increased sampling
+density. When both \textit{gag} and \textit{env} sequences were analyzed, the
+estimates from the \textit{env} data were substantially higher
+(\cref{fig:genes}). The ratios of internal to terminal branch lengths were
+lower for the \textit{env} trees than the \textit{gag} trees (average 
     \Sexpr{ratios[gene == "env", round(mean(ratio), 2)]}
@@
 1$ was excluded from the prior, the confidence interval shrank significantly.
-For both priors, the estimated prevalence was extremely high, in fact higher
+For both priors, the estimated $I$ value was extremely high, in fact higher
 than the estimated HIV prevalence in the sampled region. The authors indicated
@@
 
-\subsubsection{Preferential attachment power is sub-linear and higher for \gls{IDU}}
-
-\add{For all datasets we examined, the posterior mean estimates for \gls{alpha}
+\subsubsection{Preferential attachment power is sub-linear and higher for IDU networks}
+
+\add{For all datasets we examined, the posterior mean estimates for $\alpha$
 were sub-linear, ranging from 
@@
 \textcite{de2007preferential}, who developed a statistical inference method to
-estimate the parameters of a more sophisticated \acrlong{PA} model
+estimate the parameters of a more sophisticated preferential attachment model
 incorporating heterogeneous node behaviour. When used to analyze
-population-level longitudinal partner count data, they found \gls{alpha} values
+population-level longitudinal partner count data, they found $\alpha$ values
 ranging from 0.26 to 0.62 depending on the gender and time period considered.
-Our estimates of \gls{alpha} for the IDU/Romania was above this range under
-both priors, as were the estimates for the MSM/Beijing data and the BC data
-when $\gls{m} = 1$ was disallowed by the prior. The dataset investigated by
+Our estimates of $\alpha$ for the IDU/Romania was above this range under both
+priors, as were the estimates for the MSM/Beijing data and the BC data when $m
+= 1$ was disallowed by the prior. The dataset investigated by
 \textcite{de2007preferential} was derived from a survey of a random sample of
 the Norwegian population, whereas our investigation focused on datasets from
-known phylogenetic or geographic clusters of \gls{HIV} infected persons. It is
-therefore unsurprising that we detected stronger \acrlong{PA} dynamics in some
-cases. For instance, random sampling is much less likely to discover the
-high-degree nodes characterizing the tail of the degree distribution, simply
-because those individuals are rare in the general population. In addition, it
-is plausible that \gls{HIV}-positive individuals are more likely to be highly
-connected in their sexual networks, as the odds of acquiring \gls{HIV} increase
-with the number of unprotected sexual contacts.}
+known phylogenetic or geographic clusters of HIV-infected persons. It is
+therefore unsurprising that we detected stronger preferential attachment
+dynamics in some cases. For instance, random sampling is much less likely to
+discover the high-degree nodes characterizing the tail of the degree
+distribution, simply because those individuals are rare in the general
+population. In addition, it is plausible that HIV-positive individuals are more
+likely to be highly connected in their sexual networks, as the odds of
+acquiring HIV increase with the number of unprotected sexual contacts.}
 
 \add{Both \textcite{de2007preferential} and the HET/Botswana data studied
-populations whose primary risk factor for \gls{HIV} infection was heterosexual
+populations whose primary risk factor for HIV infection was heterosexual
 contact. \citeauthor{de2007preferential} explicitly excluded reported
@@
 was gender-biased, being composed of approximately 75\% women. Our estimate of
-\gls{alpha} for these data was 
+$\alpha$ for these data was 
     \Sexpr{d[m_min == 1 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]}
@@
     \Sexpr{d[m_min == 2 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]},
-depending on the prior on \gls{m}. Similarly, \textcite{de2007preferential}
+depending on the prior on $m$. Similarly, \textcite{de2007preferential}
 estimated 0.54, 0.57, and 0.29 for 3-year, 5-year, and lifetime partnership
@@
 
-\add{For both choices of prior on \gls{m}, the datasets derived from \gls{IDU}
-populations had a higher estimated \acrlong{PA} power than the other datasets
-(\cref{fig:abchpd,fig:abchpdm2}). This finding is in line with
-\textcite{dombrowski2013topological}, who reanalyzed a network of \glspl{IDU}
-in Brooklyn, USA, collected between 1991 and
-1993~\autocite{friedman2006social}. They found that the \gls{IDU} network
-resembled a \gls{BA} network much more closely than other social and sexual 
-networks, and offered sociological explanations for the apparent \acrlong{PA}
-dynamics in this population. Importantly, from a public health perspective,
-the authors asserted that the removal of \emph{random} individuals from
-\gls{IDU} networks may have the paradoxical effect of increasing the network's
-epidemic susceptibility. When low-degree nodes are removed, as would occur
-during a police crackdown, their network neighbours may turn to well-known
-community members for advice or supplies, thus increasing the connectivity of
-these high-degree nodes.}
-
-\add{Unfortunately, the sub-linear region for \gls{alpha} identified by both
-\textcite{de2007preferential} and \software{netabc} is also the region of
-poorest identifiability (\cref{fig:gridest,fig:abcptm2}). This was reflected in
-the high level of uncertainty in the estimates, with most 95\% \gls{HPD}
-intervals covering the majority of the range [0, 1]. The value $\gls{alpha} =
-0.5$ was contained in the 95\% \gls{HPD} interval for every dataset;
-consequently, it is not possible to say with high confidence that any of the
-\gls{alpha} values are different from each other. In synthetic data, the
-confidence intervals around \gls{alpha} narrowed when other parameters were
-marginalized out (\cref{fig:marginal}). Thus, it is possible that estimates of
-\gls{alpha} could be made more precise by specifying either exact values or
-informative priors on the other \gls{BA} parameters. We argued informally in
-\cref{subsec:whyba} that \gls{I}, \gls{m}, and \gls{N} can sometimes be
-estimated by standard epidemiological methods, although the difficulty may
-depend on the specific epidemic under consideration.}
+\add{For both choices of prior on $m$, the datasets derived from \gls{IDU}
+populations had a higher estimated preferential attachment power than the other
+datasets (\cref{fig:abchpd,fig:abchpdm2}). This finding is in line with
+\textcite{dombrowski2013topological}, who reanalyzed a network of IDUs in
+Brooklyn, USA, collected between 1991 and 1993~\autocite{friedman2006social}.
+They found that the IDU network resembled a BA network much more closely than
+other social and sexual networks, and offered sociological explanations for the
+apparent preferential attachment dynamics in this population. Importantly, from
+a public health perspective, the authors asserted that the removal of
+\emph{random} individuals from IDU networks may have the paradoxical effect of
+increasing the network's epidemic susceptibility. When low-degree nodes are
+removed, as would occur during a police crackdown, their network neighbours may
+turn to well-known community members for advice or supplies, thus increasing
+the connectivity of these high-degree nodes.}
+
+\add{Unfortunately, the sub-linear region for $\alpha$ identified by both
+    \textcite{de2007preferential} and \textit{netabc} is also the region of
+poorest identifiability (\cref{fig:abcptm2,fig:abcptm3,fig:abcptm4}). This was
+reflected in the high level of uncertainty in the estimates, with most 95\%
+HPD intervals covering the majority of the range [0, 1]. The value $\alpha =
+0.5$ was contained in the 95\% HPD interval for every dataset; consequently, it
+is not possible to say with high confidence that any of the $\alpha$ values are
+different from each other. In synthetic data, the confidence intervals around
+$\alpha$ narrowed when other parameters were marginalized out
+(\cref{fig:marginal}). Thus, it is possible that estimates of $\alpha$ could be
+made more precise by specifying either exact values or informative priors on
+the other BA parameters when these are known.}
 
 \add{I is underestimated in some cases and overestimated in others. Some of the
-datasets have bost of the posteriormass on m = 1. Estimates of N are
+datasets have most of the posterior mass on m = 1. Estimates of N are
 meaningless.}
@@
 \add{We are grateful to Dr. Sally Otto, Dr. Alexandre Bouchard-C\^ot\'e, 
-Dr. Richard Harrigan, and the anonymous reviewers for many helpful
+Dr. Richard Harrigan, and the two anonymous reviewers for many helpful
 suggestions.} This work was supported by grants from the Canadian Institutes of
@@
     Transmission trees were simulated over three types of network, with
-    pathogen transmissibility varying from 0 to 1. Sackin's index was
-    calculated for each simulated transmission tree.
+    pathogen transmissibility varying from 0 to 1 (1000 trees per
+    transmissibility value). Sackin's index was calculated for each simulated
+    transmission tree. Lines are means, light shaded areas are 95\% quantile
+    range, and dark shaded areas are interquartile ranges. 
     }

### 2016-07-01 doc/figures/pipeline.tex
GLM thingies
Progress on paper changes

@@
   [every node/.style={rectangle, draw, text width=4cm, align=center, rounded corners},
-   every path/.style={->, >=stealth, thick}]
+   every path/.style={->, >=stealth, thick},
+   node distance=0.5 and 1]
   \node (dl) {download sequences \\ from GenBank};
@@
   \node (tree) [right=of dl] {estimate unrooted phylogeny with FastTree2};
-  \node (cluster) [below=of tree, dashed] {select cluster by patristic distance};
+  \node (cluster) [below=of tree, dashed] {select cluster (patristic distance) or subtree};
   \node (root) [below=of cluster] {root and scale phylogeny with root-to-tip regression};

### 2016-07-01 doc/ismb/poster.tex
Progress on paper changes

@@
 \setlength{\bibhang}{0pt}
+\DeclareBibliographyCategory{ignore}
+\addtocategory{ignore}{zetterberg2004two}
+\addtocategory{ignore}{niculescu2015recent}
+\addtocategory{ignore}{novitsky2013phylogenetic}
+\addtocategory{ignore}{novitsky2014impact}
+\addtocategory{ignore}{mccormack2002early}
+\addtocategory{ignore}{grabowski2014role}
+\addtocategory{ignore}{wang2015targeting}
+\addtocategory{ignore}{kao2011surveillance}
+\addtocategory{ignore}{little2014using}
+\addtocategory{ignore}{li2015hiv}
+\addtocategory{ignore}{cuevas2009hiv}
+
 \tikzposterlatexaffectionproofoff
@@
 \node [text width=6in, align=right] at ($ (current page.west) + (0.42\paperwidth, -1.5in) $)
-  { \LARGE rmccloskey@cfenet.ubc.ca \\ (604) 682-6344 ext. 63266 };
+  { \LARGE rmccloskey@cfenet.ubc.ca \\ (604) 558-8865 };
 
@@
   \block{Contact networks shape viral transmission trees}{
-    \centerline{\includegraphics[width=0.2\textwidth]{contactnet}}
+    \centerline{\includegraphics[width=0.22\textwidth]{contactnet}}
       Network structure can change the speed and pattern of epidemic spread,
@@
     an existing node of degree $d$ with probability proportional to
-    $d^\alpha$.
+    $d^\alpha + 1$.
 
-    \vspace{1cm}
-    \centerline{\includegraphics[width=0.2\textwidth]{pa-example}}
-    \vspace{1cm}
+    %\node at (0, 0) {sub-linear $\alpha$};
+    %\node at (14, 6.5) {linear $\alpha$};
+    %\node at (23, 7.5) {super-linear $\alpha$};
+
+    \begin{center}
+      \begin{tikzpicture}
+        \node (a) {\includegraphics[width=0.22\textwidth]{pa-example}};
+        \node at (a.south) [anchor=north] {sub-linear $\alpha$ $\qquad\qquad\qquad$ linear $\alpha$ $\qquad\qquad\qquad$ super-linear $\alpha$};
+      \end{tikzpicture}
+    \end{center}
 
@@
     \end{center}
-    \node at (1, 10) {low $\alpha$};
-    \node at (14, 10) {high $\alpha$};
   }
@@
   \block{Varying $\alpha$ and $I$ affects tree shape, but not $m$ and $N$}{
-    \centerline{\includegraphics[width=0.18\textwidth]{kernel-kpca}}
+    \centerline{\includegraphics[width=0.16\textwidth]{kernel-kpca}}
 
@@
   \block{Narrow dispersion of $\alpha$, $I$ compared to $m$, $N$}{
-    \centerline{\includegraphics[width=0.16\textwidth]{abc-posterior-example}}
+    \centerline{\includegraphics[width=0.15\textwidth]{abc-posterior-example}}
   }
@@
     \centering
+    \normalsize
     \input{\tablepath/realdata}
@@
 
-  \block{Parameter heterogeneity among real datasets}{
-    \vspace{1cm}
-    \centerline{\includegraphics[width=0.25\textwidth]{realdata-hpd}}
+  \block{Stronger preferential attachment in IDU networks}{
+    \includegraphics[trim=0 3.5in 3.3in 0, clip, width=0.12\textwidth]{realdata-hpd}
+    \includegraphics[trim=3.4in 1.5in 0 2.0in, clip, width=0.12\textwidth]{realdata-hpd}
     \vspace{1cm}
@@
   }
+  \column{0.35}
+  \block{References} {
+    \renewcommand*{\bibfont}{\footnotesize}
+    \printbibliography[notcategory=ignore, heading=none]
+    \small
+    \includegraphics[height=1cm]{logos/westgrid} Analyses were run in the
+    Westgrid high performance computing environment.
+  }
+\end{columns}
 \block{}{
-    \begin{minipage}{0.55\linewidth}
+  \begin{minipage}{0.6\linewidth}
     \vspace{1cm}
-      \includegraphics[width=3in]{logos/btp}
-      \includegraphics[width=3in]{logos/cfe}
-      \includegraphics[width=1.5in]{logos/cihr}
+    \includegraphics[width=4in]{logos/btp}
+    \quad
+    \includegraphics[width=4in]{logos/cfe}
+    \quad
+    \includegraphics[width=2in]{logos/cihr}
+    \quad
     \includegraphics[width=3in]{logos/msfhr}
-      \includegraphics[width=2in]{logos/phc}
-      \includegraphics[width=0.9in]{logos/ubc}
+    \quad
+    \includegraphics[width=3in]{logos/phc}
+    \quad
+    \includegraphics[width=1in]{logos/ubc}
+    \quad
+    \includegraphics[width=1.8in]{logos/genomecanada}
+    \quad
+    \includegraphics[width=2.2in]{logos/genomebc}
+    \quad
+    \includegraphics[width=1.8in]{logos/genomequebec}
     \vspace{0.5cm}
@@
   \begin{adjustbox}{valign=t}
-      \begin{minipage}[t]{0.45\linewidth}
+    \begin{minipage}[t]{0.4\linewidth}
       \small
@@
 
-  \column{0.35}
-  \block{References} {
-    \renewcommand*{\bibfont}{\footnotesize}
-    \printbibliography[heading=none]
-    \small
-    \includegraphics[height=1cm]{logos/westgrid} Analyses were run in the
-    Westgrid high performance computing environment.
-  }
-
-\end{columns}
-
 \end{document}

### 2016-07-01 doc/paper/changes.tex
Progress on paper changes

@@
 Thank you for the opportunity to revise our manuscript ``Reconstructing contact
-network parameters from viral phylogenies''.
+network parameters from viral phylogenies''. While our submission was pending,
+we made some additions and improvements to the work that we would like you and
+the reviewers to consider.
 
-While our submission was pending, we made some additions and improvements to
-the work that we would like you and the reviewers to consider. It was pointed
-out to us by a colleague that the posterior mean is a more conventionally used
-estimator than the maximum \textit{a posteriori} (MAP) value. Therefore, we
-have replaced the MAP point estimates by posterior means. Qualitatively, the
-results are not much different, except that it is more obvious that $m$ and $N$
-are not identifiable with this method. We have replaced some ad hoc statistical
-tests in the analysis of the simulation results with a $p$-value corrected
-generalized linear model, and rewritten the relevant section of the results
-accordingly. For clarity, the real HIV datasets are now referred to as their
-risk group and location in the text, rather than by their author (eg.
-MSM/Beijing instead of Wang et al.). We have found an analysis of the
-Barab\'asi-Albert model in the physics literature which demonstrates that the
-degree distribution is only a power law when $\alpha = 1$. We acknowledge this
-but do not believe it detracts from the validity of our results, especially in
-light of a paper by de Blasio et al. (2007) which found sub-linear preferential
-attachment through a population-level survey.
+\begin{itemize}
 
-We did some additional literature searching and found five more published HIV
-datasets that seemed to be appropriate for our method. The highest $\alpha$
-values were associated with IDU networks, coroborating the assertion of
-Dombrowski et al. (2013) that IDU networks exhibit the scale-free property more
-strongly than sexual networks do. Three of these new datasets had sequenced two
-HIV genes, which we analysed separately. We found that estimates of $\alpha$
-were very robust to the choice of gene; $I$ was less robust, with \textit{env}
-genes tending to lead to higher estimated values.
+  \item It was pointed out to us by a colleague that the posterior mean is a
+    more conventionally used Bayesian estimator than the maximum \textit{a
+    posteriori} (MAP) value. Therefore, we have replaced the MAP point
+    estimates by posterior means. Qualitatively, the results are not much
+    different, except that it is more obvious that $m$ and $N$ are not
+    identifiable with this method. 
+  \item We have replaced some ad hoc statistical tests in the analysis of the
+    simulation results with a $p$-value corrected generalized linear model, and
+    rewritten the relevant section of the results accordingly.
+  \item We have found an analysis of the Barab\'asi-Albert model in the physics
+    literature (Krapivsky et al. 2000) which demonstrates that the degree
+    distribution is only a power law when $\alpha = 1$. We acknowledge this but
+    do not believe it detracts from the validity of our results, especially in
+    light of a paper by de Blasio et al. (2007) which found sub-linear
+    preferential attachment through a population-level survey. We have added to
+    the supplemental materials a series of power law fits to the degree
+    distributions of BA networks with $\alpha \neq 1$ (Figure TODO), which
+    shows that, although the power law is cleary an imperfect fit, the slope
+    is still reasonably representative of the true distribution.
+  \item We did some additional literature searching and found five more
+    published HIV datasets that seemed to be appropriate for our method. For
+    clarity, the real HIV datasets are now referred to by risk group and
+    geographic location, rather than by author (eg. MSM/Beijing instead of Wang
+    et al.). The highest $\alpha$ values were associated with IDU networks,
+    coroborating the assertion of Dombrowski et al. (2013) that IDU networks
+    exhibit the scale-free property more strongly than sexual networks do. 
+  \item Three of these new datasets had sequenced two HIV genes, which we
+    analysed separately. We found that estimates of $\alpha$ were very robust
+    to the choice of gene; $I$ was less robust, with \textit{env} genes tending
+    to lead to higher estimated values.
+  \item Figure 5 (formerly Figure 4) now shows 50\% HPD intervals (similar to
+    interquartile ranges) in addition to 95\% HPD intervals.
+  \item We discovered that our use of the term ``kernel-ABC'' to refer to ABC 
+    with a kernel for the distance function is in conflict with another usage,
+    namely an expression of Bayes' rule in terms of kernel means (Fukumizu et
+    al. ``Kernel Bayes' rule.'' 2011). We therefore simply use the term ``ABC''
+    to refer to our method.
+  \item The discussion has been divided into subsections to make it easier to
+    follow.
+\end{itemize}
 
@@
 asked by the Editor in Chief to consider how to disambiguate ABC approximation
-error from a truly noisy posterior, which this is addressed under the first
-point brought up by Reviewer 1.
+error from a truly noisy posterior, which is addressed under the first point
+brought up by Reviewer 1.
 
@@
 We are grateful that the reviewer finds our work to be of some interest, and we
-acknowledge that the identifiability of the BA parameters from tree shapes is
-somewhat limited. TODO
+acknowledge that our ability to precisely resolve numerical values of the BA
+parameters is limited. We believe that our simulation results show that some
+network parameters (specifically $\alpha$ in our case) can be at least roughly
+estimated from the information contained in tree shapes.
 
@@
 due to confounding between model parameters; error which remains may be due to
-either poor resolution or truely diffuse posteriors. TODO
+either poor resolution or truely diffuse posteriors. The accuracy of the
+estimates was improved about 50\% in marginal estimation.
 
@@
 
-We apologise for the misunderstanding. Our choice of words was poor -- TODO
-
-We have redone this analysis using both SVMs (classifiers) and SVRs
-(regressions). Figure 1 has been expanded to show both sets of results --
-accuracy for the SVMs and $R^2$ for the SVRs. The results are qualitatively
-very similar.
+We apologise for this misunderstanding due to our poor choice of words. An SVR
+is in fact a regressor, not a classifier, so the reported $R^2$ were of true
+values against predictions. However, given that the point of this experiment
+was to demonstrate identifiability by categorizing tree shapes into
+predetermined values, we redid the analysis using SVMs (which are true
+classifiers) instead of SVRs and linear regression. We now report the accuracy
+of the categorical classification, and figures 1 and S3-S6 have been revised
+accordingly. Some additional variation among meta-parameters and epidemic
+scenarios is observed, but the overall results are similar and the best choice
+of meta-parameters is the same.
 
@@
 
-  Could you describe how the cross-validation was done (numbers training vs test etc)?
+  Could you describe how the cross-validation was done (numbers training vs
+  test etc)?
 \end{quote}
@@
 
-We thank the reviewer for this helpful suggestion. We individually fixed each
-combination of parameters and performed ABC inference on the remaining free
-parameters. The inference of $I$ was dramatically improved when $N$ was fixed.
-Fixing each of the parameters contributed slightly to the estimate of $\alpha$,
-but none as dramatically. However, the posterior mean estimate of $m$ was
-totally unaffected by fixing the other parameters. This indicates that our
-inability to estimate $m$ is simply due to lack of identifiability, rather than
-to parameter correlations.
+We thank the reviewer for this helpful suggestion, which we addressed by
+marginal estimation as outlined under the first comment. The inference of $I$
+was dramatically improved when $N$ was fixed. Fixing each of the parameters
+contributed slightly to the estimate of $\alpha$, but none as dramatically.
+However, the posterior mean estimate of $m$ was totally unaffected by fixing
+the other parameters. This indicates that our inability to estimate $m$ is
+simply due to lack of identifiability, rather than to parameter correlations.
 
-As the reviewer suggested, we have also included a set of two-dimensional
-posterior distributions in the text (Figure TODO). The correlation between $N$
-and $I$ demonstrated in the marginal analysis is clearly observable, but there
-are no obvious correlations among the other parameters.
+As suggested, we have also included a set of two-dimensional posterior
+distributions in the text (Figure 4). The correlation between $N$ and $I$
+demonstrated in the marginal analysis is clearly observable, but there are no
+obvious correlations among the other parameters.
 
@@
 have reduced the number of posterior plots shown to one in the text and two in
-the supplemental materials. TODO
+the supplemental materials. Figure 2, which formerly represented each
+simulation by an individual point, has been replaced by a box plot showing the
+$\alpha$ and $I$ estimates stratified by each of the other parameters. A
+similar boxplot for $m$ and $N$, which were not identifiable with ABC, has been
+added to the supplemental materials.
 
@@
 
+We have changed the phrase to ``effective viral population size.''
+
 \begin{quote}
@@
 approximate maximum likelihood phylogeny reconstruction with root-to-tip
-regression does not result in large errors in phylodynamic inference (Poon et
-al. 2015). We reanalysed the datasets with the highest and lowest estimated
+regression does not result in large errors in phylodynamic inference (Poon
+2015). We reanalysed the datasets with the highest and lowest estimated
 $\alpha$ (Zetterberg et al. 2004 and Cuevas et al. 2009) using a BEAST MCC tree
@@
 were TODO
+
+With respect to the suggestion to analyse bootstrap replicates, we believe that
+the three new datasets containing both \textit{gag} and \textit{env} data fill
+the need for within-dataset replication.
 
@@
 
-TODO dunno man
+We think the reviewer is referring to Figure S3, which shows the
+cross-validation accuracy of the SVM classifier for $\alpha$ under several
+values of the tree kernel meta-parameters $\lambda$ and $\sigma$. However,
+higher $\sigma$ values ($x$-axis) lead to better estimates, not worse, and for
+most meta-parameter combinations the estimates get worse, not better, on
+smaller datasets.
+
+In any case, we can offer a few comments on this figure. When $\sigma = 0$,
+subset trees only match when their branch lengths are identical. As $\sigma \to
+\infty$, branch lengths are not taken into account at all, and matches are
+based on topology alone. Thus, it seems that topology is more important for
+classifying $\alpha$ than branch lengths are, which is consistent with our
+observation that $\alpha$ strongly affects tree balance. The decay factor
+$\lambda$ is intended to penalize very large subset tree matches which can
+dominate the kernel score (see Poon et al. 2013). This ``large match'' problem
+becomes worse as the trees become larger. For the largest trees we considered, 
+it seems that $\lambda = 0.4$ was not a strict enough penalty, leading to a
+decrease in classification accuracy.
+
+Part of the new text introducing the classification experiments now contains a
+short description of the two meta-parameters. In addition, a brief comment
+on the effects of $\lambda$ and $\sigma$ on the classifiers has been added to
+the first paragraph of the results section.
 
@@
 
-This statement has been removed in our updated version. When we calculated
-posterior means rather than modes, as is more conventional, we found that
-nearly all of the estimates were very near to the middle of the range. This is
-consistent with our observed lack of signal for $m$ in phylogenetic data.
+The reviewer's interpretation is correct, but this statement has been removed
+in our updated version. When we calculated posterior means rather than modes,
+as is more conventional, we found that nearly all of the estimates were very
+near to the middle of the range. This is consistent with our observed lack of
+signal for $m$ in phylogenetic data.
 
@@
 1$, it seemed that this choice of prior was more reasonable, and we therefore
-focused our discussion on those results.
+focused our discussion on those results. The alternative results are presented
+in the last paragraph of the discussion.
 
@@
 
-We agree that the sloped lines were misleading and have replated them with
-horizontal line segments. TODO
+As per Reviewer 1's suggestion we have replaced Figure 2 with a boxplot showing
+posterior means stratified by the true parameter values. We agree that the
+sloped lines were misleading and have replaced them with horizontal line
+segments.
 
@@
 
-TODO
+This issue was also addressed by performing marginal estimation. If the
+accurate estimates of $I$ were solely due to the prior, they would have gotten
+worse, not better, when $N$ was fixed and the prior on $I$ was truly uniform.
+We make this point in the discussion: ``when $I$ was estimated marginally with
+fixed $N$, the accuarcy of the estimate improved even though there was no
+longer any extra prior mass on low $I$ values.''
 
@@
 
-As mentioned above, we have focused our discussion on $m = 1$ due to the
-several datasets for which $m = 1$ contained most of the posterior mass.
+We have focused our discussion on $m = 1$ due to the several datasets for which
+$m = 1$ contained most of the posterior mass.
 
@@
 is our first attempt to fit network models, we thought it best to stick with
-a simple, well-known model rather than defining our own parameters. (TODO: We
+a simple, well-known model rather than defining our own parameters. We did
+perform a small experiment allowing heterogeneity among nodes, but in $\alpha$
+rather than in $m$, and recovered the average of the two $\alpha$ values.
+
+(TODO: We
 have added to the discussion the point that this binary choice is sub-optimal

### 2016-07-01 doc/paper/paper.Rnw
Progress on paper changes

@@
 and epidemic threshold~\autocite{barthelemy2005dynamical}, in turn affecting
-the estimates of quantities such as effective population
+the estimates of quantities such as effective viral population
 size~\autocite{goodreau2006assessing}.  From a public health perspective,
@@
 as the cost of sequencing has declined. In the case of HIV, genotyping has
-become part of routine clinical care in several health regions.'' Due to the
+become part of routine clinical care in several health regions.} Due to the
 high mutation rate of RNA viruses, epidemiological processes impact the course
@@
 $\alpha > 1$, it is a ``gelation'' type distribution where one or a few hub
-nodes are connected to nearly every other node in the graph.}
+nodes are connected to nearly every other node in the
+graph~\autocite{krapivsky2000connectivity}.}
 
@@
 group~\autocite{poon2015phylodynamic} and others~\autocite{park2015k2} have
-demonstrated that taking $\rho$ to be kernel function TODO}
+demonstrated that taking $\rho$ to be a well-chosen kernel function can produce
+a more accurate ABC approximation than the typical choice of a difference of
+summary statistics.}
 
@@
 $\alpha$, but we will refer to it here as $\alpha_{\text{ESS}}$ to avoid
-confusion with the preferential attachment power parameter of the BA model. A
-computer program implementing our method is freely available at
-\url{https://github.com/rmcclosk/netabc} (last accessed \today).}
+confusion with the preferential attachment power parameter of the BA model.}
 
@@
 graphs. The analyses were run on Westgrid (\url{https://www.westgrid.ca/}) and
-a local computer cluster.
+a local computer cluster. \add{A computer program implementing our method is
+freely available at \url{https://github.com/rmcclosk/netabc} (last accessed
+\today).}
 
@@
 $\lambda$ and $\sigma$ (the ``decay factor'' and ``radial basis function
-variance'', see \textcite{poon2013mapping}). The results of this experiment
-were used to select values for these meta-parameters to carry forward based on
-their accuracy in distinguishing the different trees.}
+variance'', see \textcite{poon2013mapping}). $\lambda$ is used to penalize
+ween large subset trees which tend to dominate the kernel score. When $\lambda
+= 0$, all but the root branches of each subset tree are ignored, while when
+$\lambda = 1$, no penalty is applied. $\sigma$ controls how strictly the notion
+of similarity is applied to branch lengths. When $\sigma = 0$, branch lengths
+must match exactly, while as $\sigma \to \infty$, branch lengths are not
+considered at all.}
 
@@
 values, and a $1 \times 300$ vector of Sackin's index values.} We constructed 
-\add{three classifiers for $\alpha$:} a kernel support vector regression (kSVR)
+\add{three classifiers for $\alpha$:} a kernel support vector regression (kSVM)
 \add{from the kernel matrix} with the \textit{kernlab}
-package~\autocite{zeileis2004kernlab}, an ordinary SVR \add{from the nLTT
-matrix} with the e1071 package~\autocite{meyer2015e1071}, and a linear
-regression from the Sackin's index values. The accuracy of each classifier was 
-evaluated with 1000 two-fold cross validations \add{with equally-sized folds}.
+package~\autocite{zeileis2004kernlab}, and two ordinary SVMs \add{from the nLTT
+matrix and Sackin's index vector} with the e1071
+package~\autocite{meyer2015e1071}.\del{, and a linear regression from the
+Sackin's index values.} The accuracy of each classifier was evaluated with 1000
+two-fold cross validations \add{with equally-sized folds}.
 
@@
 \del{For each of the four parameters, we also tested a linear regression against
-Sackin's index~\autocite{shao1990tree} and an ordinary SVR based on the normalized
+Sackin's index~\autocite{shao1990tree} and an ordinary SVM based on the normalized
 lineages-through-time (nLTT) statistic~\autocite{janzen2015approximate}.}
@@
 run was performed with 2000 particles, 10 simulated transmission trees per
-particle, and $\alpha_{\text{ESS}} = 0.99$. To investigate the extent to which
+particle, and $\alpha_{\text{ESS}} = 0.97$. To investigate the extent to which
 errors in the estimated BA parameters were due to true features of the
@@
 subset of the HET/Uganda data was chosen by eye such that the individuals were
-monopolistic in both the \textit{gag} and \textit{env} trees. The other subsets
-were arbitrarily chosen subtrees from phylogenies of the complete datasets.}
+monophyletic in both the \textit{gag} and \textit{env} trees. The other
+subsets were arbitrarily chosen subtrees from phylogenies of the complete
+datasets.}
 
@@
         d <- setDT(collect.data(rownames(md)))
-        if (param %in% c('alpha', 'm')) {
+        if (param == 'alpha') {
+            d[,list(rsquared=mean(rsquared), accuracy=mean(accuracy)), by=c('nsimnode', 'ntip')]
+        } else if (param == 'm') {
             d[,list(rsquared=mean(rsquared)), by=c('nsimnode', 'ntip')]
         } else if (param == 'I') {
-            d[,list(rsquared=mean(rsquared)), by=ntip]
+            d[,list(rsquared=mean(rsquared), accuracy=mean(accuracy)), by=ntip]
         } else {
@@
 parameter values, calculated pairwise tree kernel scores between them, and
-attempted to classify the trees using a kernel support vector machine (kSVR).
+attempted to classify the trees using a kernel support vector machine (kSVM).
 We also tested classifiers based on Sackin's index~\autocite{shao1990tree} and
 the normalized lineages-through-time (nLTT)
-statistic~\autocite{janzen2015approximate}. The accuracy of each classifier
-\add{on all BA parameters is shown in} \del{varied based on the parameter
-being tested} \cref{fig:rsquared}. Classifiers based on \del{two other tree
+statistic~\autocite{janzen2015approximate}. \add{We report the accuracy of the
+classifiers, which is simply the proportion of trees which were assigned the
+correct parameter value. Since there were three possible values, random
+guessing would produce an accuracy of 0.33.} \del{The accuracy of each
+classifier varied based on the parameter being tested} \add{The results are
+shown in} \cref{fig:rsquared}. Classifiers based on \del{two other tree
 statistics,} the nLTT and Sackin's index generally exhibited worse performance
 than the tree kernel, although the magnitude of the disparity varied between
-the parameters (\cref{fig:rsquared}, centre and right). The results were
+the parameters (\cref{fig:rsquared}, centre and right). \del{The results were
 largely robust to variations in the tree kernel meta-parameters $\lambda$ and
-$\sigma$, \add{although accuracy varied between different epidemic and sampling
-scenarios} (\cref{fig:alphacrossv,fig:mcrossv,fig:Ncrossv,fig:Icrossv}).
-\add{For all parameters except $m$, the absolute number of tips in the tree had
-a much greater impact on accuracy than the proportion of infected individuals
-these tips represented. However, for $m$, both the number and proportion of
-sampled tips had a strong impact or the accuracy of the kSVR
-(\cref{fig:mcrossv}).}
+$\sigma$, although accuracy varied between different epidemic and sampling
+scenarios} \add{Larger datasets were generally classified more accurately
+(\cref{fig:alphacrossv,fig:mcrossv,fig:Ncrossv,fig:Icrossv}), although large
+values of $\lambda$ produced worse estimates on large datasets. Extremely low
+$\sigma$ values, which require nearly-exact matches between branch lengths,
+resulted in low accuracy in some cases (\textit{e.g.} \cref{fig:alphacrossvn},
+center row).}
 
@@
     \caption[
-        Cross-validation accuracy of kernel-SVR, nLTT-based SVR, and Sackin's
+        Cross-validation accuracy of kernel-SVM, nLTT-based SVM, and Sackin's
         index regression classifiers for BA model parameters.
     ]{
-        Cross-validation accuracy of kernel-SVR classifier (left), SVR
-        classifier using nLTT (centre), and linear regression using
-        Sackin's index (right) for BA model parameters. Kernel meta-parameters
-        were set to $\lambda = 0.3$ and $\sigma = 4$. Each point was calculated
-        based on 300 simulated transmission trees over networks with three
-        different values of the parameter being tested, assuming perfect
-        knowledge of the other parameters. Vertical lines are empirical 95\%
-        confidence intervals based on 1000 two-fold cross-validations. \add{The
-        classifiers for $I$ were not evaluated with 1000-tip trees, because one
-        of the tested $I$ values was 500, and it is not possible to sample a
-        tree of size 1000 from 500 infected individuals.}
+        Cross-validation accuracy of kernel-SVM classifier (left), SVM
+        classifiers using nLTT (centre) and Sackin's index (right) for BA model
+        parameters. Kernel meta-parameters were set to $\lambda = 0.3$ and
+        $\sigma = 4$. Each point was calculated based on 300 simulated
+        transmission trees over networks with three different values of the
+        parameter being tested, assuming perfect knowledge of the other
+        parameters. Vertical lines are empirical 95\% confidence intervals
+        based on 1000 two-fold cross-validations. \add{The classifiers for $I$
+        were not evaluated with 1000-tip trees, because one of the tested $I$
+        values was 500, and it is not possible to sample a tree of size 1000
+        from 500 infected individuals.}
     }
@@
 
-\add{The accuracy of each classifier on all BA parameters is shown in
-\cref{fig:rsquared}. Classifiers based on the nLTT and Sackin's index generally
-exhibited worse performance than the tree kernel, although the magnitude of the
-disparity varied between the parameters (\cref{fig:rsquared}, centre and
-right). The results were largely robust to variations in the tree kernel
-meta-parameters $\lambda$ and $\sigma$, although accuracy varied between
-different epidemic and sampling scenarios
-(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}). For all
-parameters except $m$, the absolute number of tips in the tree had a much
-greater impact on accuracy than the proportion of infected individuals these
-tips represented. However, for $m$, both the number and proportion of sampled
-tips had a strong impact or the accuracy of the kSVR (\cref{fig:mcrossv}).}
-
-The kSVR classifier for $\alpha$ had an average $R^2$ of 
-    \Sexpr{kernel.alpha[,round(mean(rsquared), 2)]},
+The kSVM classifier for $\alpha$ had an average accuracy of 
+    \Sexpr{kernel.alpha[,round(mean(accuracy), 2)]},
 compared to 
-    \Sexpr{nltt.alpha[,round(mean(rsquared), 2)]}
-for the nLTT-based SVR, and
-    \Sexpr{sackin.alpha[,round(mean(rsquared), 2)]}
-for the linear regression against Sackin's index. There was little variation
-about the mean for different tree and epidemic sizes. No classifier could
-accurately identify the $m$ parameter in any epidemic scenario, with average
-$R^2$ values of 
-  \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for kSVR,
+    \Sexpr{nltt.alpha[,round(mean(accuracy), 2)]}
+for the nLTT, and
+    \Sexpr{sackin.alpha[,round(mean(accuracy), 2)]}
+for Sackin's index. \del{There was little variation about the mean for
+different tree and epidemic sizes.} No classifier could accurately identify $m$
+in any epidemic scenario, with average $R^2$ values of 
+  \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for kSVM,
   \Sexpr{nltt.m[,round(mean(rsquared), 2)]} for the nLTT, and
   \Sexpr{sackin.m[,round(mean(rsquared), 2)]}
-for Sackin's index. Again, there was little variation in accuracy between
-epidemic scenarios, although the accuracy of the kSVR was slightly higher
-on 1000-tip trees (\cref{fig:rsquared}, left).
+for Sackin's index. There was little variation in accuracy between epidemic
+scenarios, although the accuracy of the kSVM was slightly higher on 1000-tip
+trees (\cref{fig:rsquared}, left).
 
 The accuracy of classifiers for $I$ varied significantly with the number of
-tips in the tree. For 100-tip trees, the average $R^2$ values were
-  \Sexpr{kernel.I[ntip == 100, round(mean(rsquared), 2)]},
-  \Sexpr{nltt.I[ntip == 100, round(mean(rsquared), 2)]}, and
-  \Sexpr{sackin.I[ntip == 100, round(mean(rsquared), 2)]}
-for the tree kernel, nLTT, and Sackin's index respectively. For 500-tip
+tips in the tree. For 100-tip trees, the average accuracy was
+  \Sexpr{kernel.I[ntip == 100, round(mean(accuracy), 2)]},
+  \Sexpr{nltt.I[ntip == 100, round(mean(accuracy), 2)]}, and
+  \Sexpr{sackin.I[ntip == 100, round(mean(accuracy), 2)]}
+for the tree kernel, nLTT, and Sackin's index respectively.
+For 500-tip
 trees, the values increased to
-  \Sexpr{kernel.I[ntip == 500, round(mean(rsquared), 2)]},
-  \Sexpr{nltt.I[ntip == 500, round(mean(rsquared), 2)]}, and
-  \Sexpr{sackin.I[ntip == 500, round(mean(rsquared), 2)]}.
+  \Sexpr{kernel.I[ntip == 500, round(mean(accuracy), 2)]},
+  \Sexpr{nltt.I[ntip == 500, round(mean(accuracy), 2)]}, and
+  \Sexpr{sackin.I[ntip == 500, round(mean(accuracy), 2)]}.
 Finally, the performance of classifiers for $N$ depended heavily on the
-epidemic scenario. The $R^2$ of the kSVR classifier ranged from
+epidemic scenario. The $R^2$ of the kSVM classifier ranged from
   \Sexpr{kernel.N[,round(min(rsquared), 2)]}
@@
   \Sexpr{kernel.N[,round(max(rsquared), 2)]}
-for the largest. Likewise, $R^2$ for the nLTT-based SVR ranged from 
+for the largest. Likewise, $R^2$ for the nLTT-based SVM ranged from 
   \Sexpr{nltt.N[,round(min(rsquared), 2)]}
@@
 
-\Cref{fig:abcpt} shows \del{maximum \textit{a posteriori} (MAP)} \add{posterior
-mean} point estimates of the BA model parameters obtained with ABC on simulated
-data. The estimates shown correspond only to the simulations where the $m$
-parameter was set to 2, however the results for $m = 3$ and $m = 4$ were
-similar (\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% HPD
-intervals are given in \cref{tab:abchpd}.
+\Cref{fig:abcpt} shows \del{maximum \textit{a posteriori} (MAP)}
+\add{stratified posterior mean} point estimates of the BA model parameters
+\add{$\alpha$ and $I$}, obtained with ABC on simulated data. \add{The
+parameters $m$ and $N$ were not identifiable with ABC for any parameter
+combinations (\cref{fig:abcpt2}).} \del{The estimates shown correspond only to
+the simulations where the $m$ parameter was set to 2, however the results for
+$m = 3$ and $m = 4$ were similar.} Average boundaries of 95\% HPD intervals
+\add{for all parameters} are given in \cref{tab:abchpd}.
 
@@
     \centering
-    \includegraphics[width=\textwidth]{abc-point-estimate-m2}
+    \includegraphics[width=\textwidth]{abc-boxplot}
     \caption[
-        Posterior mean point estimates for BA model parameters obtained by
-        running \textit{netabc} on simulated data, for simulations with
-        $m = 2$.
+        Posterior mean point estimates for BA model parameters $\alpha$ and $I$
+        obtained by running \textit{netabc} on simulated data, stratified by
+        true parameter values.
     ]{
-        Posterior mean point estimates for BA model parameters obtained by
-        running \textit{netabc} on simulated data, for simulations with
-        $m = 2$. Dashed lines indicate true values. (A) Estimates of
-        $\alpha$ and $I$ which were varied in these simulations against known
-        values. (B) Estimates of $m$ and $N$ which were held fixed in these
-        simulations at the values $m = 2$ and $N = 5000$.
+        Posterior mean point estimates for BA model parameters $\alpha$ and $I$
+        obtained by running \textit{netabc} on simulated data, stratified by
+        true parameter values. First row of plots contains true versus
+        estimated values of $\alpha$; second row contains true versus estimated
+        values of $I$. Columns are stratified by $\alpha$, $I$, and $m$
+        respectively. Dashed lines indicate true values. 
     }
@@
 
-\begin{table*}[ht]
+\begin{table}[ht]
     \centering
@@
     \label{tab:abchpd}
-\end{table*}
+\end{table}
 
 \del{The accuracy of the parameter estimates obtained with ABC
-paralleled the results from the kSVR classifier. Of the four parameters,
+paralleled the results from the kSVM classifier. Of the four parameters,
 $\alpha$ was the most accurately estimated, with point estimates having a
@@
     \Sexpr{d[,round(quantile(N, 0.75))]}])
-regardless of the true values of any of the BA parameters.}
+regardless of the true values of any of the BA parameters (\cref{fig:abcpt2}).}
 
@@
 parameter values as independent variables. Since the estimates of $m$ and
-$N$ were roughly equal across all simulations
-(\cref{fig:abcptm2,fig:abcptm3,fig:abcptm4}), GLMs were not fitted for these
-parameters. The estimated coefficients are shown in \cref{tab:glm}.}
+$N$ were roughly equal across all simulations (\cref{fig:abcpt2}), GLMs were
+not fitted for these parameters. The estimated coefficients are shown in
+\cref{tab:glm}.}
 
@@
 simulation.} \add{show one- and two-dimensional marginal distributions for a
-simulation with $\alpha$ and $I$ errors close to their respective medians. In
-particular, the simulation shown is one of
-\Sexpr{replace_number(nrow(middle))} with errors in both $\alpha$ and $I$
-between the \Sexpr{round(q*100)}nd and \Sexpr{round((1-q)*100)}th percentiles.
+simulation with $\alpha$ and $I$ errors close to their respective medians. 
+%In particular, the simulation shown is one of
+%\Sexpr{replace_number(nrow(middle))} with errors in both $\alpha$ and $I$
+%between the \Sexpr{round(q*100)}nd and \Sexpr{round((1-q)*100)}th percentiles.
 The parameters for this simulation were
@@
     \centering
-    \includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior-2d", middle.plot)}}.pdf}
+    %\includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior-2d", middle.plot)}}.pdf}
     \caption[
@@
 
+<<marginal, include=FALSE>>=
+    d <- setDT(collect.data("../../simulations/abc-marginal/point-estimate/*", sep="\t"))
+
+    joint.alpha <- d[is.na(fixed), abs(alpha-true_alpha)]
+    marginal.alpha <- d[fixed == "m+N+I", abs(alpha-true_alpha)]
+
+    joint.N <- d[is.na(fixed), abs(N-true_N)]
+    marginal.N <- d[fixed == "I+alpha+m", abs(N-true_N)]
+
+    joint.I <- d[is.na(fixed), abs(I-true_I)]
+    marginal.I <- d[fixed == "N+alpha+m", abs(I-true_I)]
+@
+
+\add{\Cref{fig:marginal} shows the effect of performing marginal ABC estimation
+of each of the BA parameters on the same simulated transmission tree. The
+estimates of $m$ are apparently unaffected by marginalizing out the other
+parameters, corroborating the previous experiments' findings that $m$ is not an
+identifiable parameter from scaled tree shapes. Compared to allowing all
+parameters to vary, estimates of $\alpha$, $I$, and $N$ were improved by
+    \Sexpr{ round((joint.alpha - marginal.alpha) * 100 / joint.alpha) }\%,
+    \Sexpr{ round((joint.I - marginal.I) * 100 / joint.I) }\%, and
+    \Sexpr{ round((joint.N - marginal.N) * 100 / joint.N) }\%
+when all other parameters were fixed. Neither the accuracy nor the precision of
+any parameter estimates appeared to improve when more computing power was used
+for SMC (\cref{fig:moresmc}).}
+
 \subsection*{Real data}
@@
 $m = 1$ was the most credible value for several datasets, all results discussed
-henceforth are for the prior $m \sim \text{DiscreteUniform}(1, 5)$ unless
+henceforth apply to the prior $m \sim \text{DiscreteUniform}(1, 5)$ unless
 otherwise stated.}
@@
 
-\end{document}
-
 \add{Posterior mean point estimates for the preferential attachment power $\alpha$
@@
 HPD intervals were very wide for most datasets, often encompassing almost the
-entire range from 0 to 1 (\cref{fig:abchpd}). For all the datasets except
-HET/Malawi and MSM/Beijing, the posterior mean was either outside, or very
-close to the border of, the 50\% HPD intervals. This indicates that the
-posterior distributions were diffuse with heavy tails, rather than having most
-of their mass around the mode. When the value $m = 1$ was disallowed by the
-prior, the separation between the IDU datasets and the others became more
-striking (\cref{fig:abchpdm2}). All three IDU datasets had estimated $\alpha$
-values at or above
-    \Sexpr{pmean2[c("zetterberg2004", "niculescu2015", "bctree"), round(min(alpha), 2)]}.
-The estimate for the MSM/Beijing data was slightly lower
-    (\Sexpr{pmean2["wang2015", round(alpha, 2)]}) and
-the estimates for the seven remaining non-IDU datasets were bounded above by
-    \Sexpr{pmean2[c("novitsky2014", "mccormack2002", "grabowski2014", "kao2011", "little2014", "li2015", "cuevas2009"), round(max(alpha), 2)]}.}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{realdata-genes}
-    \caption[
-        Posterior means and 50\%/95\% HPD intervals for parameters of the
-        BA network model, fitted to three HIV datasets where both \textit{gag}
-        and \textit{env} genes were sequenced.
-    ]{
-        Posterior means (points), 50\% HPD intervals (notches), and 95\%
-        HPD intervals (lines) for parameters of the BA network model, fitted to
-        three HIV datasets where both \textit{gag} and \textit{env} genes were
-        sequenced.
-    }
-    \label{fig:genes}
-\end{figure}
+entire range from 0 to 1 (\cref{fig:abchpd}).}
 
@@
 estimates from the \textit{env} data were substantially higher
-(\cref{fig:genes}). The ratios of internal to terminal branch lengths were
+    (average \Sexpr{pmean1[datasets, round(mean(I))]} for \textit{gag} vs.
+     \Sexpr{pmean1.env[datasets, round(mean(I))]} for \textit{env},
+\cref{fig:genes}). The ratios of internal to terminal branch lengths were
 lower for the \textit{env} trees than the \textit{gag} trees (average 
+    \Sexpr{ratios[gene == "gag", round(mean(ratio), 2)]}
+for \textit{gag} and 
     \Sexpr{ratios[gene == "env", round(mean(ratio), 2)]}
-for \textit{env} and 
-    \Sexpr{ratios[gene == "gag", round(mean(ratio), 2)]}
-for \textit{gag}). As with $\alpha$, many of the posterior means were
-outside the 50\% HPD intervals, suggesting diffuse posterior
-distributions. The values of $I$ were fairly robust to restricting the
-prior to $m \geq 2$ (compare \cref{fig:abchpd,fig:abchpdm2}), although 
-the 95\% HPD intervals were slightly narrower (average width
-    \Sexpr{round(mean(up1[,I] - low1[,I]))}
-for $m \geq 1$ and
-    \Sexpr{round(mean(up2[,I] - low2[,I]))}
-for $m \geq 2$).}
+for \textit{env}).}
 
@@
 
-\add{The posterior means of \gls{m} were equal to one for 
+\add{The posterior means of $m$ were equal to one for 
   \Sexpr{replace_number(pmean1[,sum(m == 1)])}
-of the datasets analyzed. The widths of the 95\% \gls{HPD} intervals varied
-from 0 (all the mass on the estimated value) to 5 (the entire prior region).
-When the value $\gls{m} = 1$ was disallowed by the prior, the posterior means
-for all but the HET/Botswana data took on the value 3, with the \gls{HPD}
-intervals spanning the entire prior region. This is very similar to the results 
-observed for \gls{m} on simulated data (\cref{fig:abcptm2,tab:abchpd}), and
-suggests that \gls{m} is not identifiable from these data with this prior. On
-the other hand, the estimate and 95\% \gls{HPD} interval for the HET/Botswana
-data (4 [2-5]) was the same for both priors.}
-
-\add{Estimates of \gls{N} were  mostly uninformative, with very similar estimates
-for all datasets (mean
+of the datasets analyzed. The widths of the 95\% HPD intervals varied from 0
+(all the mass on the estimated value) to 5 (the entire prior region). Estimates
+of $N$ were  mostly uninformative, with very similar estimates for all datasets
+(mean
     \Sexpr{pmean1[,round(mean(N))]},
@@
 This was similar to the pattern observed for the synthetic data, where the
-posterior mean always fell around the upper two-thirds mark of the region
-allowed by the prior. The results did not change appreciably when the value
-$\gls{m} = 1$ was assigned zero prior probability.}
+posterior mean always fell around the upper two-thirds mark of the range
+(\cref{fig:abcpt2}).}
+
+\add{When the value $m = 1$ was disallowed by the prior, the separation in
+$\alpha$ between the IDU datasets and the others became more striking
+(\cref{fig:abchpdm2}). All three IDU datasets had estimated $\alpha$ values at
+or above
+    \Sexpr{pmean2[c("zetterberg2004", "niculescu2015", "bctree"), round(min(alpha), 2)]}.
+The estimate for the MSM/Beijing data was slightly lower
+    (\Sexpr{pmean2["wang2015", round(alpha, 2)]}) and
+the estimates for the seven remaining non-IDU datasets were bounded above by
+    \Sexpr{pmean2[c("novitsky2014", "mccormack2002", "grabowski2014", "kao2011", "little2014", "li2015", "cuevas2009"), round(max(alpha), 2)]}. 
+The values of $I$ were fairly robust to the choice of prior (compare
+\cref{fig:abchpd,fig:abchpdm2}), although the 95\% HPD intervals were slightly
+narrower (average width
+    \Sexpr{round(mean(up1[,I] - low1[,I]))}
+for $m \geq 1$ and
+    \Sexpr{round(mean(up2[,I] - low2[,I]))}
+for $m \geq 2$). When $m = 1$ was disallowed, the posterior means of $m$ for
+all but the HET/Botswana data took on the value 3, with the HPD intervals
+spanning the entire prior region. This is very similar to the results observed
+for $m$ on simulated data (\cref{tab:abchpd}), and suggests that $m$ is not
+identifiable from these data with this prior. On the other hand, the estimate
+and 95\% HPD interval for the HET/Botswana data (4 [2-5]) was the same for both
+priors. The results for $N$ did not change appreciably between the two choices
+of prior.}
 
@@
 related work]{groendyke2011bayesian, volz2008sir, brown2011transmission,
-leventhal2012inferring}. Kernel-ABC is a model-agnostic method which can be
-used to investigate any quantity that affects tree
-shape~\autocite{poon2015phylodynamic}. In this work, we developed a
-ABC-based method to infer the parameters of a contact network model. The
-method is general, and could be applied to any model from which contact
-networks can be simulated. We demonstrated the method on the BA model, which is
-a simple preferential attachment model giving rise to the power law degree
-distributions commonly observed in real-world networks.
-
-By training a kernel-SVR classifier, we found that the $\alpha$ and $I$
+leventhal2012inferring}. ABC is a model-agnostic method which can be used to
+investigate any quantity that affects tree
+shape~\autocite{poon2015phylodynamic}. In this work, we developed a ABC-based
+method to infer the parameters of a contact network model. The method is
+general, and could be applied to any model from which contact networks can be
+simulated. We demonstrated the method on the BA model, which is a simple
+preferential attachment model. \add{For some parameter choices,} the BA model
+gives rise to the power law degree distributions commonly observed in
+real-world networks.
+
+\subsection*{Analysis of BA model with synthetic data}
+
+\del{By training a kernel-SVM classifier, we found that the $\alpha$ and $I$
 parameters, representing preferential attachment power and number of infected
@@
 did not produce much variation in tree shape, resulting in in both poorly
-performing classifiers and uninformative ABC estimates.
-
-$N$ was almost always significantly over-estimated using ABC. Since the
+performing classifiers and uninformative ABC estimates.}
+
+\add{The preferential attachment power $\alpha$ had a strong influence on tree
+shape in the range of values we considered. Although the tree kernel was the
+most effective classifier for $alpha$, a Sackin's index of tree imbalance
+performed nearly as well (\cref{fig:rsquared}). High $\alpha$ values produce
+networks with few well-connected ``superspreader'' nodes which are involved in
+a large number of transmissions, resulting in a highly unbalanced ladder-like
+tree structure. There appeared to be weaker identifiability for $\alpha < 1$
+than for $\alpha \geq 1$ (\cref{fig:abcpt,tab:abchpd}), which may be partially
+explained by the relationship between $\alpha$ and the power law exponent
+$\gamma$ (\cref{fig:gamma}). Although the degree distributions do not truly
+follow a power law for $\alpha \neq 1$, the fitted exponent still captures the
+shape of the degree distribution reasonably well (\cref{fig:powerlaw}). The
+$\gamma$ values fitted to $\alpha = 0$ and $\alpha = 0.5$ are nearly identical
+(about 2.28 for $\alpha = 0$ and 2.33 for $\alpha = 0.5$ with $N = 5000$ and $m
+= 2$). In other words, the degree distributions of networks with $\alpha < 1$
+are similar to each other, which may result in similarity of corresponding
+transmission trees as well.}
+
+\add{$I$, representing the number of infected individuals at the time of
+sampling, was also identifiable, albeit slightly over-estimated with ABC for
+both values we considered. Sackin's index was better able to discern $I$ from
+tree shape than the nLTT (\cref{fig:rsquared,fig:Icrossv}), suggesting that
+this parameter impacts the distribution of branching times in the tree more
+than the topology. In a homogeneously-mixed population, branching times can be
+modelled by the coalescent process~\autocite{kingman1982coalescent}, in our
+case under the SI model~\autocite{volz2009phylodynamics}. Although networks are
+not homogeneously mixed, the forces which affect the distributions of branching
+times still apply. In our simulations, all discordant edges shared the same
+transmission rate, so that the waiting time until the next transmission in the
+entire network was always inversely proportional to the number of discordant
+edges. In the initial phase of the epidemic, when $I$ is small, each new
+transmission results in many new discordant edges. Hence, there is an early
+exponential growth phase, producing many short branches near the root of the
+tree. As the epidemic gets closer to saturating the network, the number of
+discordant edges decays, causing longer waiting times.}
+
+% OLD STUFF
+ Since the
 prior on $N$ and $I$ is jointly uniform on a non-rectangular region ($I \leq
-N$), there is more prior mass on high $N$ values. In retrospect, it is
-unreasonable to expect good estimation of $N$, because adding more nodes to a
-BA network does not change the edge density or overall shape. This can be
+N$), there is more prior mass on high $N$ values.
+
+This can be
 illustrated by imagining that we add a small number of nodes to a network after
@@
 simulation again on the new, larger network could produce the exact same
-transmission tree as before. We note also that our accurate estimates of $I$
-may have been influenced by this prior, which places more mass on low $I$
-values. However, the estimate of $I$ was very high for the
-\textcite{novitsky2013phylogenetic,novitsky2014impact} data, suggesting that a
-strong enough signal in the data can overcome the prior.
+transmission tree as before. We note also that \del{our accurate estimates of
+$I$ may have been influenced by} this prior\del{, which} places more mass on
+low $I$ values. However, the estimate of $I$ was very high for the HET/Botswana
+data, suggesting that a strong enough signal in the data can
+overcome the prior. \add{Furthermore, when $I$ was estimated marginally with
+fixed $N$, the accuracy of the estimate improved even though there was no
+longer any extra prior mass on low $I$ values.}
+
+% END OLD STUFF
+
+\add{The number of nodes in the network, $N$, exhibited the most variation in
+terms of its effect on tree shape. There was almost no measurable difference
+between trees simulated under different $N$ values when the number of infected
+nodes $I$ was small (\cref{fig:Ncrossv}).} $N$ was always significantly
+over-estimated using ABC (\cref{fig:abcpt2}).
+
+In retrospect, it is unreasonable to expect good estimation of $N$, because
+adding additional nodes does not change the edge density or overall shape of a
+BA network. This can be illustrated by imagining that we add a small number of
+nodes to a network after the epidemic simulation has already been completed. It
+is possible that none of these new nodes attains a connection to any infected
+node. Thus, running the simulation again on the
+new, larger network could produce the exact same transmission tree as before.
+On the other hand, when $I$ is large relative to $N$, the coalescent dynamics
+discussed above also apply. The waiting times until the next infection
+increase, resulting in longer coalescence times toward the tips. The relative
+accuracy of the nLTT in these situations (\cref{fig:rsquared,fig:Ncrossv})
+corroborates this hypothesis, as the nLTT uses only information about the
+coalescence times. When all BA parameters were simultaneously estimated with
+ABC, $N$ was nearly always over-estimated by approximately a factor of two
+(\cref{fig:abcptm2,tab:abchpd}). One factor which may have contributed to this
+bias was our choice of prior distribution. Since the prior for $I$ and $N$ was
+jointly uniform on a region where $I \leq N$, more prior weight was assigned to
+higher $N$ values.
+
+\add{Another possible contributing factor to the overestimation of $I$ and $N$
+relates to the dynamics of the SI model and the coalescent process. The number
+of infected individuals follows a logistic growth curve under the SI model.
+This kind of growth curve has three qualitative phases: a slow ramp-up, an
+exponential growth phase, and a slow final phase when the susceptible
+population is almost depleted. The waiting times until the next transmission,
+which determine the coalescence times in the tree, are dependent on the growth
+phase of the epidemic. Therefore, we hypothesize that it is the growth phase at
+the time of sampling which most affects tree shape, rather than the specific
+values of $I$ or $N$. To investigate this hypothesis, we simulated transmission
+trees over networks on a grid of $I$ and $I$ values and fit logistic growth
+curves to the proportion of infected individuals over time. We and calculated
+the first and second derivatives of these curves (with respect to time) at the
+time of transmission tree sampling. As shown in \cref{fig:derivs}, there are
+bands along which both derivatives are similar which contain the values we
+tested. These bands span mostly higher values of $N$ and $I$ than the true
+values. Thus, if $N$ and $I$ are free to vary (as is the case in ABC), both
+parameters will tend to be overestimated due to being less identifiable within
+their own band. However, when $N$ is fixed at 5000, the derivatives vary
+substantially along the $I$-axis, which explains why a marginal estimate of $I$
+was more accurate (\cref{fig:marginal}). We also note the resemblance of the
+contour surface of \cref{fig:derivs} to the two-dimensional marginal posterior
+distribution on $I$ and $N$ obtained with simulated data (\cref{fig:abc2}).}
+
+\add{The $m$ parameter, which controls the number of connections added to the
+network per vertex, did not have a measurable impact on tree shape and was not
+identifiable with ABC. However, all the analyses presented here did not take
+the absolute size of the transmission trees into account, as the branch lengths
+were rescaled by their mean. Because higher $m$ values imply higher edge
+density, an epidemic should spread more quickly for higher $m$ than lower $m$
+with the same per-edge transmission probability. Hence, considering the
+absolute height of the trees may improve our method's ability to reconstruct
+$m$. It was also pointed out to us by an anonymous reviewer that for a fixed
+$I$, an infected node may only end up transmitting along a fraction of its
+outgoing edges, which could mask the presence of the extra edges associated
+with higher $m$.}
+
+\end{document}
 
@@
 range of permitted $\gamma$ values may be warranted.
+
+\subsection*{Analysis of real world HIV datasets}
 
@@
 
-\add{For both choices of prior on $m$, the datasets derived from \gls{IDU}
+\add{For both choices of prior on $m$, the datasets derived from IDU
 populations had a higher estimated preferential attachment power than the other
@@
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics{abc-point-estimate-m3.pdf}
-  \caption{
-    Point estimates of BA model parameters obtained by running ABC
-    on simulated phylogenies without training, for simulations with $m = 3$.
-    Dotted lines indicate true values, and limits of the $y$-axes are regions
-    of uniform prior density. (A) Estimates for $\alpha$ and $I$ against their
-    true values in simulations. (B) Estimates for $m$ and $N$, which were held
-    fixed in these simulations, against true values of $\alpha$.
-  }
-  \label{fig:abcptm3}
+\begin{figure}
+    \includegraphics[width=\textwidth]{abc-boxplot-mN}
+    \caption[
+        Posterior mean point estimates for BA model parameters $m$ and $N$
+        obtained by running \textit{netabc} on simulated data, stratified by
+        true parameter values.
+    ]{
+        Posterior mean point estimates for BA model parameters $m$ and $N$
+        obtained by running \textit{netabc} on simulated data, stratified by
+        true parameter values. First row of plots contains true versus
+        estimated values of $m$; second row contains true versus estimated
+        values of $N$. Columns are stratified by $\alpha$, $I$, and $m$
+        respectively. Dashed lines indicate true values. 
+    }
 \end{figure}
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics{abc-point-estimate-m4.pdf}
-  \caption{
-    Point estimates of BA model parameters obtained by running ABC
-    on simulated phylogenies without training, for simulations with $m = 4$.
-    Dotted lines indicate true values, and limits of the $y$-axes are regions
-    of uniform prior density. (A) Estimates for $\alpha$ and $I$ against their
-    true values in simulations. (B) Estimates for $m$ and $N$, which were held
-    fixed in these simulations, against true values of $\alpha$.
-  }
-  \label{fig:abcptm4}
+\begin{figure}
+    \includegraphics[width=\textwidth]{abc-marginal}
+    \caption[
+        Posterior means, 50\% HPD intervals, and 95\% HPD intervals for BA
+        model parameter estimated marginally with ABC.
+    ]{
+        Posterior means (points), 50\% HPD intervals (notches), and 95\% HPD
+        intervals (lines) for BA model parameter estimated marginally with ABC.
+        $x$-axis labels indicate parameters which were fixed to their true
+        values by specifying Dirac-delta priors.
+    }
+    \label{fig:marginal}
 \end{figure}
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{realdata-genes}
+    \caption[
+        Posterior means and 50\%/95\% HPD intervals for parameters of the
+        BA network model, fitted to three HIV datasets where both \textit{gag}
+        and \textit{env} genes were sequenced.
+    ]{
+        Posterior means (points), 50\% HPD intervals (notches), and 95\%
+        HPD intervals (lines) for parameters of the BA network model, fitted to
+        three HIV datasets where both \textit{gag} and \textit{env} genes were
+        sequenced.
+    }
+    \label{fig:genes}
+\end{figure}
+
 \begin{figure}[ht]
-  \includegraphics{cuevas2009-posterior}
-  \caption{
-      Posterior distribution of BA model parameters for
-      \textcite{cuevas2009hiv} data. Vertical lines indicate maximum \textit{a
-      posteriori} estimates, and shaded areas are 95\% highest posterior
-      density intervals. $x$-axis indicates regions of nonzero prior density.
-  }
-  \label{fig:cuevas}
+  \centering
+  \includegraphics{derivatives.pdf}
+  \caption[
+    First and second time derivatives of epidemic growth curves at time of
+    sampling for various values of $I$ and $N$.
+  ]{
+    First and second \add{time} derivatives of epidemic growth curves at time
+    of sampling for various values of \gls{I} and \gls{N}. Networks were
+    simulated under the \gls{BA} model with \gls{alpha} = 1.0, \gls{m} = 2, and
+    \gls{N} varied along the values shown on the $x$-axis. Transmission trees
+    were sampled at the time when \gls{I} nodes were infected ($y$-axis).
+    Logistic growth curves were fit to epidemic trajectories derived from the
+    transmission trees, and their first and second derivatives were calculated
+    at the time of sampling. Contours show first derivatives, while colours
+    indicate second derivatives. Values of \gls{I} and \gls{N} used in
+    simulation experiments with \gls{ABC} are indicated by diamonds.
+  }
+  \label{fig:derivs}
 \end{figure}
 
-\begin{figure}[ht]
-  \includegraphics{li2015-posterior}
-  \caption{
-      Posterior distribution of BA model parameters for \textcite{li2015hiv}
-      data. Vertical lines indicate maximum \textit{a posteriori} estimates,
-      and shaded areas are 95\% highest posterior density intervals. $x$-axis
-      indicates regions of nonzero prior density.
-  }
-  \label{fig:li}
-\end{figure}
-
-\begin{figure}[ht]
-  \includegraphics{niculescu2015-posterior}
-  \caption{
-      Posterior distribution of BA model parameters for
-      \textcite{niculescu2015recent} data. Vertical lines indicate maximum
-      \textit{a posteriori} estimates, and shaded areas are 95\% highest
-      posterior density intervals. $x$-axis indicates regions of nonzero prior
-      density.
-  }
-  \label{fig:niculescu}
-\end{figure}
-
-\begin{figure}[ht]
-  \includegraphics{novitsky2014-posterior}
-  \caption{
-      Posterior distribution of BA model parameters for
-      \textcite{novitsky2013phylogenetic, novitsky2014impact} data. Vertical
-      lines indicate maximum \textit{a posteriori} estimates, and shaded areas
-      are 95\% highest posterior density intervals. $x$-axis indicates regions
-      of nonzero prior density.
-  }
-  \label{fig:novitsky}
-\end{figure}
-
-\begin{figure}[ht]
-  \includegraphics{wang2015-posterior}
-  \caption{
-      Posterior distribution of BA model parameters for
-      \textcite{wang2015targeting} data. Vertical lines indicate maximum
-      \textit{a posteriori} estimates, and shaded areas are 95\% highest
-      posterior density intervals. $x$-axis indicates regions of nonzero prior
-      density.
-  }
-  \label{fig:wang}
-\end{figure}
-
-<<abc_posterior, echo=FALSE, results="asis", eval=FALSE>>=
-    N <- 5000
-    replicate <- 0
-    for (m in c(2, 3, 4)) {
-    for (I in c(1000, 2000)) {
-    for (alpha in c(0, 0.5, 1, 1.5)) {
-        fn <- sprintf("{abc-posterior/%.1f_%d_%d_%d_%d}.pdf", alpha, I, m, N, replicate)
-        lab <- sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate)
-        cat("\\begin{figure}[ht]\n")
-        cat("\\centering\n")
-        cat(paste0("\\includegraphics{", fn, "}\n"))
-        cat("\\caption{\n")
-        cat(sprintf("Approximate marginal posterior distributions of Barab\\'asi-Albert model parameters obtained using ABC for one simulated transmission tree with parameter values $\\alpha = %.1f$, $I = %d$, $m = %d$, and $N = %d$.
+%\begin{figure}[ht]
+%  \includegraphics{cuevas2009-posterior}
+%  \caption{
+%      Posterior distribution of BA model parameters for
+%      \textcite{cuevas2009hiv} data. Vertical lines indicate maximum \textit{a
+%      posteriori} estimates, and shaded areas are 95\% highest posterior
+%      density intervals. $x$-axis indicates regions of nonzero prior density.
+%  }
+%  \label{fig:cuevas}
+%\end{figure}
+%
+%\begin{figure}[ht]
+%  \includegraphics{li2015-posterior}
+%  \caption{
+%      Posterior distribution of BA model parameters for \textcite{li2015hiv}
+%      data. Vertical lines indicate maximum \textit{a posteriori} estimates,
+%      and shaded areas are 95\% highest posterior density intervals. $x$-axis
+%      indicates regions of nonzero prior density.
+%  }
+%  \label{fig:li}
+%\end{figure}
+%
+%\begin{figure}[ht]
+%  \includegraphics{niculescu2015-posterior}
+%  \caption{
+%      Posterior distribution of BA model parameters for
+%      \textcite{niculescu2015recent} data. Vertical lines indicate maximum
+%      \textit{a posteriori} estimates, and shaded areas are 95\% highest
+%      posterior density intervals. $x$-axis indicates regions of nonzero prior
+%      density.
+%  }
+%  \label{fig:niculescu}
+%\end{figure}
+%
+%\begin{figure}[ht]
+%  \includegraphics{novitsky2014-posterior}
+%  \caption{
+%      Posterior distribution of BA model parameters for
+%      \textcite{novitsky2013phylogenetic, novitsky2014impact} data. Vertical
+%      lines indicate maximum \textit{a posteriori} estimates, and shaded areas
+%      are 95\% highest posterior density intervals. $x$-axis indicates regions
+%      of nonzero prior density.
+%  }
+%  \label{fig:novitsky}
+%\end{figure}
+%
+%\begin{figure}[ht]
+%  \includegraphics{wang2015-posterior}
+%  \caption{
+%      Posterior distribution of BA model parameters for
+%      \textcite{wang2015targeting} data. Vertical lines indicate maximum
+%      \textit{a posteriori} estimates, and shaded areas are 95\% highest
+%      posterior density intervals. $x$-axis indicates regions of nonzero prior
+%      density.
+%  }
+%  \label{fig:wang}
+%\end{figure}
+%
+%<<abc_posterior, echo=FALSE, results="asis", eval=FALSE>>=
+%    N <- 5000
+%    replicate <- 0
+%    for (m in c(2, 3, 4)) {
+%    for (I in c(1000, 2000)) {
+%    for (alpha in c(0, 0.5, 1, 1.5)) {
+%        fn <- sprintf("{abc-posterior/%.1f_%d_%d_%d_%d}.pdf", alpha, I, m, N, replicate)
+%        lab <- sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate)
+%        cat("\\begin{figure}[ht]\n")
+%        cat("\\centering\n")
+%        cat(paste0("\\includegraphics{", fn, "}\n"))
+%        cat("\\caption{\n")
+%        cat(sprintf("Approximate marginal posterior distributions of Barab\\'asi-Albert model parameters obtained using ABC for one simulated transmission tree with parameter values $\\alpha = %.1f$, $I = %d$, $m = %d$, and $N = %d$.
 Dashed lines indicate true values, and shaded areas show 95\\%% highest posterior density intervals.\n", alpha, I, m, N))
-        cat("}\n")
-        cat(paste0("\\label{", lab, "}\n"))
-        cat("\\end{figure}\n")
-        cat("\\clearpage\n\n")
-    }
-    }
-    }
-@
+%        cat("}\n")
+%        cat(paste0("\\label{", lab, "}\n"))
+%        cat("\\end{figure}\n")
+%        cat("\\clearpage\n\n")
+%    }
+%    }
+%    }
+%@
 

### 2016-07-01 doc/thesis/discussion.Rnw
Progress on paper changes

@@
 the Norwegian population, whereas our investigation focused on datasets from
-known phylogenetic or geographic clusters of \gls{HIV} infected persons. It is
+known phylogenetic or geographic clusters of \gls{HIV}-infected persons. It is
 therefore unsurprising that we detected stronger \acrlong{PA} dynamics in some
@@
     (\Sexpr{d1[dataset == "little2014" & parameter == "I", round(mean)]})
-was clearly an underestimate. The study enrolled almost 648 \gls{HIV} infected 
+was clearly an underestimate. The study enrolled almost 648 \gls{HIV}-infected 
 individuals, which should have been a lower bound on \gls{I}, although the
@@
 district were enrolled in the study, it is likely that this number represents
-the majority of \gls{HIV} infected persons. Our estimate was slightly lower
+the majority of \gls{HIV}-infected persons. Our estimate was slightly lower
     (\Sexpr{d1[dataset == "grabowski2014" & parameter == "I", round(mean)]}),

### 2016-07-01 doc/thesis/methods_results.tex
Progress on paper changes

@@
 \end{figure}
-
-The accuracy of each classifier \add{on all \gls{BA} parameters is shown in}
-\del{varied based on the parameter being tested} \cref{fig:rsquared}.
-Classifiers based on \del{two other tree statistics,} the \gls{nltt} and
-Sackin's index generally exhibited worse performance than the tree kernel,
-although the magnitude of the disparity varied between the parameters
-(\cref{fig:rsquared}, centre and right). The results were largely robust to
-variations in the tree kernel meta-parameters \gls{lambda} and \gls{sigma},
-although accuracy varied between different epidemic and sampling scenarios
-(\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}). \add{For all
-parameters except \gls{m}, the absolute number of tips in the tree had a much
-greater impact on accuracy than the proportion of infected individuals these
-tips represented. However, for \gls{m}, both the number and proportion of
-sampled tips had a strong impact or the accuracy of the \gls{kSVR}
-(\cref{fig:mcrossv}).}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
-    \caption[
-        Cross-validation accuracy of kernel-SVR, nLTT-based SVR, and Sackin's
-        index regression classifiers for BA model parameters.
-    ]{
-        Cross-validation accuracy of kernel-SVR classifier (left), SVR
-        classifier using \gls{nltt} (centre), and linear regression using
-        Sackin's index (right) for \gls{BA} model parameters. Kernel
-        meta-parameters were set to $\gls{lambda} = 0.3$ and $\gls{sigma} = 4$.
-        Each point was calculated based on 300 simulated transmission trees
-        over networks with three different values of the parameter being
-        tested, assuming perfect knowledge of the other parameters. Vertical
-        lines are empirical 95\% confidence intervals based on 1000 two-fold
-        cross-validations. \add{The classifiers for \gls{I} were not evaluated
-        with 1000-tip trees, because one of the tested \gls{I} values was 500,
-        and it is not possible to sample a tree of size 1000 from 500 infected
-        individuals.}
-    }
-    \label{fig:rsquared}
-\end{figure}
-
-The \gls{kSVR} classifier for \gls{alpha} had an average $R^2$ of 
-    0.92,
-compared to 
-    0.56
-for the \gls{nltt}-based SVR, and
-    0.75
-for the linear regression against Sackin's index. There was little variation
-about the mean for different tree and epidemic sizes. No classifier could
-accurately identify \gls{m} in any epidemic scenario, with average $R^2$ values
-of 
-  0.12 for \gls{kSVR},
-  0.01 for the \gls{nltt}, and
-  0.06
-for Sackin's index. Again, there was little variation in accuracy between
-epidemic scenarios, although the accuracy of the \gls{kSVR} was slightly higher
-on 1000-tip trees 
-    (average $R^2$ 
-     0.01,
-     0.11,
-     0.32
-     for 100, 500, and 1000 tips respectively).
-
-The accuracy of classifiers for \gls{I} \del{varied significantly
-with}\add{appeared to be strongly influenced by} the number of tips in the
-tree. For 100-tip trees, the average $R^2$ values were
-  0.7,
-  0.55, and
-  0.02
-for the tree kernel, \gls{nltt}, and Sackin's index respectively. For 500-tip
-trees, the values increased to
-  0.93,
-  0.83, and
-  0.07.
-Finally, the performance of classifiers for $N$ depended heavily on the
-epidemic scenario. The $R^2$ of the \gls{kSVR} classifier ranged from
-  0.08
-for the smallest epidemic and smallest sample size, to
-  0.82
-for the largest. Likewise, $R^2$ for the \gls{nltt}-based SVR ranged from 
-  0.01
-to
-  0.54.
-Sackin's index did not accurately classify $N$ in any scenario, with an average
-$R^2$ of
-  0.03
-and little variation between scenarios.
-
-\subsubsection*{Marginal parameter estimates with grid search}
-
-\del{The accuracy of grid search estimates largely paralleled that of the
-\gls{kSVR} classifiers.} \Cref{fig:gridest} shows point estimates and
-\add{50\%} and 95\% highest density intervals for each of the \gls{BA}
-parameters, for one replicate experiment with 500-tip trees. Plots showing the
-point estimates for all replicates can be found in
-\cref{fig:gridptalpha,fig:gridptI,fig:gridptm,fig:gridptN}. For all parameters
-except \gls{m}, the error of point estimates was negatively correlated with the
-number of sampled tips in the tree (for \gls{alpha}, \gls{I}, and \gls{N}
-respectively: Spearman's $\rho$ = 
-    \ensuremath{-0.22},
-    \ensuremath{-0.51}, 
-    \ensuremath{-0.16}; 
-$p$-values
-    $4\!\times\!10^{-4}$,
-    ${<}10^{-5}$,
-    $0.01$).
-The 95\% highest density intervals obtained for all parameters were extremely
-wide, occupying $>$75\% of the grid in all cases (\cref{fig:gridest}).
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{gridsearch-example}
-    \caption[
-        Grid search estimates of \gls{BA} model parameters for one replicate
-        experiment with trees of size 500.
-    ]{
-        Grid search estimates of \gls{BA} model parameters for one replicate
-        experiment with trees of size 500. Point estimates (dots), 95\%
-        \glspl{HDI} (lines), and 50\% \glspl{HDI} (notches) for each \gls{BA}
-        model parameter, obtained using grid search. Networks and transmission
-        trees were simulated over a grid of values for each parameter while
-        holding the others fixed \add{to known values}. For a subset of the
-        grid values ($x$-axis), test networks and trees were created and
-        compared to each tree on the grid using the tree kernel. The kernel
-        scores along the grid were normalized to resemble a probability
-        distribution, from which the mode and highest density interval were
-        calculated.
-    } 
-    \label{fig:gridest}
-\end{figure}
-
-\add{Across all replicates, $R^2$ values for the correlations between the
-estimated and true values were
-    $0.91$,
-    $0.91$,
-    $0.25$, and
-    $0.54$
-for \gls{alpha}, \gls{I}, \gls{m}, and \gls{N} respectively. The mean absolute
-errors of the point estimates were
-    $0.14$,
-    $310$,
-    $1.31$, and
-    $2419$,
-representing
-    $7$\%,
-    $8$\%,
-    $26$\%, and
-    $17$\%
-of the respective grids. For \gls{I} and \gls{N}, relative errors may be more
-appropriate to consider. An overestimate of 100 individuals would be very
-misleading if the true population was only of size 100, but almost negligible
-in a population of size 5000. The mean relative errors for \gls{I} and \gls{N}
-respectively were
-    $18$\% and 
-    $31$\%.}
-
-\add{Qualitatively, \gls{alpha} and \gls{I} exhibited weak identifiability
-within particular sections of the grid (\cref{fig:gridest}). The 50\%
-\glspl{HDI} for \gls{alpha} were similar for the values $\gls{alpha}
-\leq 0.5$ (on average 
-    0.02 -
-    0.84)
-and for $\gls{alpha} \geq 1.5$ 
-    (1.26 -
-    2).
-For \gls{I}, similar 50\% \gls{HPD} were observed for $\gls{I} \leq 1500$
-(average 
-    [650 -
-     2846])
-and for $\gls{I} > 3000$ 
-    ([2596 -
-      4802]).
-No similar patterns were observed for \gls{m} or \gls{N} (although the 50\%
-\glspl{HDI} appear to be identical for $\gls{m} > 2$ in \cref{fig:gridexpt},
-this was not consistent across replicates).}
-
-\add{\Cref{fig:gridalpha,fig:gridI,fig:gridm,fig:gridN} show kernel score
-distributions of kernel scores along the grid for each parameter. The
-distributions for some values, such as $\gls{alpha} = 1.25$, $\gls{I} = 500$
-and $4500$, $\gls{m} = 1$, and $\gls{N} = 1000$, exhibited distinct peaks
-around the true value. This indicates that these values produce distinctively
-shaped trees that can be identified with the tree kernel, when the other
-parameter values are fixed and known. However, for the majority of values of
-each parameter, the score distributions were fairly flat around the true value. 
-This means there is a range of values which produce similarly shaped trees, and
-the parameter is less identifiable within that range. The exception was
-\gls{I}, whose score distributions exhibited a more or less rounded shape with
-the highest point near the true value.}
-
-\del{The \gls{alpha} parameter was the most accurately estimated,
-with point estimates having an average deviation of 
-    0.14
-from the true value, on a grid from 0 to 2. The error of point estimates varied
-significantly between true values of \gls{alpha}
-    (one-way \gls{ANOVA}, $p$ ${<}10^{-5}$). In
-particular, errors were lower for the values \gls{alpha} = 1.0 and 1.25 than
-for the other values
-    (average errors 
-    0.03
-    for \gls{alpha} = 1.0 or 1.5 vs.
-    0.17
-    for \gls{alpha} $\neq$ 1.0 or 1.5),
-and this difference was significant
-    (Wilcoxon rank-sum test, $p {<}10^{-5}$,
-\cref{fig:gridptalpha}). These two values exhibited different qualitative
-behaviour than the other values in terms of the distribution of kernel scores
-along the grid (\cref{fig:gridalpha}). In particular, there was a pronounced
-peak in scores around the true value, in contrast to the other values where the
-scores were flat around the true value. The effect was most obvious for the
-value \gls{alpha} = 1.25.}
-
-\subsubsection*{Joint parameter estimates with \software{netabc}}
-
-\Cref{fig:abcptm2} shows \del{MAP} \add{posterior mean} point estimates of the
-\gls{BA} model parameters obtained with \gls{ABC} on simulated data. The
-estimates shown correspond only to the simulations where \gls{m} was set to 2,
-however the results for $\gls{m} = 3$ and $\gls{m} = 4$ were similar
-(\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% \gls{HPD}
-intervals are given in \cref{tab:abchpd}.
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{abc-point-estimate-m2}
-    \caption[
-        Posterior mean point estimates for \gls{BA} model parameters obtained
-        by running \software{netabc} on simulated data, for simulations with
-        $\gls{m} = 2$.
-    ]{
-        Posterior mean point estimates for \gls{BA} model parameters obtained
-        by running \software{netabc} on simulated data, for simulations with
-        $\gls{m} = 2$. Dashed lines indicate true values. (A) Estimates of
-        \gls{alpha} and \gls{I} which were varied in these simulations against
-        known values. (B) Estimates of \gls{m} and \gls{N} which were held
-        fixed in these simulations at the values $\gls{m} = 2$ and 
-        $\gls{N} = 5000$.
-    }
-    \label{fig:abcptm2}
-\end{figure}
-
-\begin{table*}[ht]
-    \centering
-    \input{\tablepath/abc-hpd.tex}
-    \caption[
-        Average posterior mean point estimates and 95\% \gls{HPD} interval
-        widths for \gls{BA} model parameter estimates obtained with
-        \software{netabc} on simulated data.
-    ]{
-        Average posterior mean point estimates and 95\% \gls{HPD} interval
-        widths for \gls{BA} model parameter estimates obtained with
-        \software{netabc} on simulated data. Three transmission trees were
-        simulated under each combination of the listed parameter values, and
-        the parameters were estimated with \gls{ABC} without training.
-    }
-    \label{tab:abchpd}
-\end{table*}
-
-\add{Across all simulations, the median [\gls{IQR}] absolute errors of the parameter 
-estimates obtained with \software{netabc} were
-    0.11 
-    [0.03 - 
-    0.25]
-for \gls{alpha},
-    492 
-    [294 - 
-    782]
-for \gls{I},
-    1 
-    [0 - 
-    1]
-for \gls{m}, and
-    4153 
-    [3660 - 
-    4489]
-for \gls{N}. These errors comprised, respectively,
-    6\%,
-    11\%,
-    17\%, and
-    29\%
-of the regions of nonzero prior density. For \gls{I} and \gls{N}, relative
-errors were
-    38\%
-    [20 - 
-    50\%]
-and
-    83\%
-    [73 - 
-    90\%].
-Average 95\% \gls{HPD} interval widths were
-    0.68,
-    2454,
-    3.01, and
-    12046,
-representing 
-    34\%,
-    55\%,
-    50\%, and
-    83\%
-of the nonzero prior density regions. Point estimates of \gls{I} were upwardly
-biased: \gls{I} was overestimated in
-    69
-out of 
-    72
-simulations
-    (96\%).
-The estimates for \gls{m} and \gls{N} were similar across all simulations
-(median [\gls{IQR}] point estimates
-    3
-    [3 - 
-    3]
-and 
-    9153
-    [8660 - 
-    9489])
-regardless of the true values of any of the \gls{BA} parameters.}
-
-\add{To analyze the effects of the true parameter values on the
-accuracy our estimates of \gls{alpha} and \gls{I}, we fitted one \gls{GLM}
-for each of these two parameters, with error rate as the dependent variable and
-the true parameter values as independent variables. Since the estimates of
-\gls{m} and \gls{N} were roughly equal across all simulations
-(\cref{fig:abcptm2,fig:abcptm3,fig:abcptm4}), \glspl{GLM} were not fitted for
-these parameters. The estimated coefficients are shown in
-\cref{tab:glmalpha,tab:glmI}. The \gls{GLM} were fitted using the inverse link
-function. That is, if $p$ is the true value of the parameter and $\hat{p}$ is a
-random variable representing our estimate of the parameter, the \gls{GLM}
-posits a relationship of the form}
-\[
-    \madd{\E(|p-\hat{p}|) = (\beta_0 + \beta_\alpha \alpha + \beta_I I + \beta_m m)^{-1}},
-\]
-\add{where the $\beta$'s are coefficients to be fitted. If the true value
-\gls{alpha}, say, is increased by one, the \emph{inverse} of the expected
-absolute error will increase by $\beta_\alpha$. If $\beta_\alpha$ is positive,
-it means that the absolute error decreases as the true value of \gls{alpha}
-increases.}
-
-% latex table generated in R 3.2.3 by xtable 1.8-2 package
-% Mon Jun 27 14:24:46 2016
-\begin{table}[ht]
-\centering
-\begin{tabular}{llll}
-  \hline
-Parameter & Estimate & Standard error & $p$-value \\ 
-  \hline
-(Intercept) & $2$ & $0.6$ & $0.01$ \\ 
-  $\alpha$ & $10$ & $2$ & ${<}10^{-5}$ \\ 
-  $I$ & $-3\!\times\!10^{-4}$ & $2\!\times\!10^{-4}$ & $0.7$ \\ 
-  $m$ & $0.5$ & $0.2$ & $0.01$ \\ 
-   \hline
-\end{tabular}
-\caption[Parameters of a fitted \gls{GLM} relating error in estimated \gls{alpha} to true values of \gls{BA} parameters.]{Parameters of a fitted \gls{GLM} relating error in estimated \gls{alpha} to true values of \gls{BA} parameters.
-\gls{GLM} was fitted with a Gaussian distribution and inverse link function.
-Coefficients are interpretable as additive effects on the inverse of the mean error.}
-\label{tab:glmalpha}
-\end{table}
-% latex table generated in R 3.2.3 by xtable 1.8-2 package
-% Mon Jun 27 14:24:46 2016
-\begin{table}[ht]
-\centering
-\begin{tabular}{llll}
-  \hline
-Parameter & Estimate & Standard error & $p$-value \\ 
-  \hline
-(Intercept) & $0.004$ & $5\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
-  $\alpha$ & $-0.001$ & $2\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
-  $I$ & $-4\!\times\!10^{-7}$ & $2\!\times\!10^{-7}$ & $0.05$ \\ 
-  $m$ & $-7\!\times\!10^{-5}$ & $8\!\times\!10^{-5}$ & $1$ \\ 
-   \hline
-\end{tabular}
-\caption[Parameters of a fitted \gls{GLM} relating error in estimated \gls{I} to true values of \gls{BA} parameters.]{Parameters of a fitted \gls{GLM} relating error in estimated \gls{I} to true values of \gls{BA} parameters.
-GLM was fitted with a Gaussian distribution and inverse link function.
-Coefficients are interpretable as additive effects on the inverse of the mean error.}
-\label{tab:glmI}
-\end{table}
-
-\add{The \gls{GLM} analysis indicated that the error in estimates of
-\gls{alpha} decreased with larger true values of \gls{alpha} 
-    ($p {<}10^{-5}$)
-and \gls{m}
-    ($p =0.01$)
-but was not significantly affected by \gls{I} (\cref{tab:glmalpha}).
-Qualitatively, \gls{alpha} seemed to be only weakly identifiable between the
-values of 0 and 0.5 (\cref{fig:abcptm2}). The error in the estimated prevalence
-\gls{I} was slightly lower for smaller values of \gls{alpha}
-    ($p {<}10^{-5}$)
-and \gls{I}
-    ($p =0.05$),
-but was not significantly affected by the true value of \gls{m}
-(\cref{tab:glmI}).}
-
-\del{The accuracy of the parameter estimates obtained with ABC paralleled the
-results from the \gls{kSVR} classifier. Of the four parameters, $\alpha$ was
-the most accurately estimated, with point estimates having a median [IQR]
-absolute error of 
-    0.11 
-    [0.03 - 
-    0.25].
-The errors when the true value of $\alpha$ was zero were significantly greater
-than those for the other values 
-    (Wilcoxon rank-sum test, $p$ = 
-    $0$).
-Errors in estimating $\alpha$ did not vary across the true values of $m$ or $I$
-(both one-way ANOVA).}
-
-\del{Estimates for $I$ were relatively accurate, with point estimate errors of
-    492 
-    [294 - 
-    782] individuals.
-These errors were significantly higher when the true value of $\alpha$ was
-at least 1
-    (Wilcoxon rank-sum test, $p$ = 
-    $0$)
-and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
-did not affect the estimates of $I$ (one-way ANOVA).}
-
-\del{The $m$ parameter was estimated correctly in
-    37 \%
-of simulations barely better than random guessing. The true values of the other
-parameters did not significantly affect the
-estimates of $m$ (both one-way ANOVA).}
-
-\del{Finally, the total number of nodes $N$ was consistently over-estimated by about
-a factor of two (error 4153
-[3660 - 4489] individuals). No parameters influenced the accuracy of the $N$
-estimates (all one-way ANOVA).}
-
-The dispersion of the \gls{ABC} approximation to the posterior also varied
-between the parameters (\cref{tab:abchpd}). \gls{HPD} intervals around
-\gls{alpha} and \gls{I} were often narrow relative to the region of nonzero
-prior density, whereas the intervals for \gls{m} and \gls{N} were more widely
-dispersed. \Cref{fig:abclow1,fig:abclow2} \del{shows the distributions for one
-simulation.} \add{show one- and two-dimensional marginal distributions for a
-simulation with relatively low error. Each parameter was chosen based on its
-mean error rate across all simulations. The parameters for this simulation were 
-    $\gls{alpha} = 1.5$,
-    $\gls{I} = 1000$,
-    $\gls{m} = 4$, and
-    $\gls{N} = 5000$.
-\Cref{fig:abchigh1,fig:abchigh2} show the equivalent marginals for a different
-simulation with relatively high error rates for each individual parameter. The
-parameters were
-    $\gls{alpha} = 0$,
-    $\gls{I} = 2000$,
-    $\gls{m} = 2$, and
-    $\gls{N} = 5000$.
-The two-dimensional marginals indicate some dependence between pairs of
-parameters, particularly \gls{I} and \gls{N} which show a diagonally shaped
-region of high posterior density.}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{{abc-posterior/1.5_1000_4_5000_0}.pdf}
-    \caption[
-        One-dimensional marginal posterior distributions of \gls{BA} model
-        parameters estimated with low error by \software{netabc} from a
-        simulated transmission tree.
-    ]{
-        One-dimensional marginal posterior distributions of \gls{BA} model
-        parameters estimated with low error by \software{netabc} from a
-        simulated transmission tree. Dashed lines indicate true values, solid
-        lines indicate posterior means, and shaded areas show 95\%
-        \acrlong{HPD} intervals.
-    }
-    \label{fig:abclow1}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{{abc-posterior-2d/1.5_1000_4_5000_0}.pdf}
-    \caption[
-        Two-dimensional marginal posterior distributions of \gls{BA} model
-        parameters estimated with low error by \software{netabc} from a
-        simulated transmission tree.
-    ]{
-        Two-dimensional marginal posterior distributions of \gls{BA} model
-        parameters estimated with low error by \software{netabc} from a
-        simulated transmission tree. White circles indicate true values,
-        magenta diamonds indicate posterior means.
-    }
-    \label{fig:abclow2}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{{abc-posterior/0.0_2000_2_5000_0}.pdf}
-    \caption[
-        One-dimensional marginal posterior distributions of \gls{BA} model
-        parameters estimated with high error by \software{netabc} from a
-        simulated transmission tree.
-    ]{
-        One-dimensional marginal posterior distributions of \gls{BA} model
-        parameters estimated with high error by \software{netabc} from a
-        simulated transmission tree. Dashed lines indicate true values, solid
-        lines indicate posterior means, and shaded areas show 95\%
-        \acrlong{HPD} intervals.
-    }
-    \label{fig:abchigh1}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{{abc-posterior-2d/0.0_2000_2_5000_0}.pdf}
-    \caption[
-        Two-dimensional marginal posterior distributions of \gls{BA} model
-        parameters estimated with high error by \software{netabc} from a
-        simulated transmission tree.
-    ]{
-        Two-dimensional marginal posterior distributions of \gls{BA} model
-        parameters estimated with high error by \software{netabc} from a
-        simulated transmission tree. White circles indicate true values,
-        magenta diamonds indicate posterior means.
-    }
-    \label{fig:abchigh2}
-\end{figure}
-
-To test the effect of model misspecification, we simulated one network where
-the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
-other half 1.5), with $\gls{m} = 2$, $\gls{N} = 5000$, and $\gls{I} = 1000$.
-The posterior mean [95\% \gls{HPD}] estimates for each parameter were: 
-\gls{alpha},
-  1.03 
-  [0.67 -
-   1.18];
-\gls{I},
-  1474 
-  [511 -
-   2990];
-\gls{m},
-  3 
-  [1 -
-   5];
-\gls{N},
-  9861 
-  [3710-
-   14977].
-The one-dimensional marginal approximate posterior distributions for this
-simulation are shown in \cref{fig:mixed}. To test the effect of sampling bias,
-we sampled one transmission tree in a peer-driven fashion, where the
-probability to sample a node was twice as high if one of its peers had already
-been sampled. The parameters for this experiment were $\gls{N} = 5000$,
-$\gls{m} = 2$, $\gls{alpha} = 0.5$, and $\gls{I} = 2000$. The estimated values
-were: \gls{alpha},
-  0.3 
-  [0 -
-   0.63];
-\gls{I},
-  2449 
-  [1417 -
-   3811];
-\gls{m},
-  3 
-  [2 -
-   5];
-\gls{N},
-  9132 
-  [2852 -
-   14780].
-The approximate posterior distributions are shown in \cref{fig:peerdriven}.
-Both of these results were in line with estimates obtained on other simulated
-datasets (\cref{tab:abchpd}), although the estimate of peer-driven sampling for
-\gls{alpha} was somewhat lower than typical.
-
-\add{\Cref{fig:marginal} shows the effect of performing marginal \gls{ABC}
-estimation of each of the \gls{BA} parameters on the same simulated
-transmission tree. The estimates of \gls{m} are apparently unaffected by
-marginalizing out the other parameters, corroborating the previous experiments'
-findings that \gls{m} is not an identifiable parameter from scaled tree shapes.
-The estimates of \gls{alpha} and \gls{N} were weakly improved in marginal
-estimation, which is consistent with the result from grid search that all
-\gls{N} values and \gls{alpha} values below 0.5 are weakly identifiable. On the
-other hand, both the accuracy and precision (\ie width of the \gls{HPD}
-intervals) of the estimates of \gls{I} were substantially improved when \gls{N}
-was marginalized out. Neither the accuracy nor the precision of any parameter
-estimates appeared to improve when more computing power was used for \gls{SMC}
-(\cref{fig:moresmc}).}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{abc-marginal.pdf}
-    \caption[
-        Marginal posterior mean estimates and 50\%/95\% \gls{HPD} intervals of
-        \gls{BA} model parameters when some parameters were fixed.
-    ]{
-        Marginal posterior mean estimates (points), 50\% \gls{HPD} intervals
-        (notches), and 95\% \gls{HPD} intervals (lines) of \gls{BA} model
-        parameters when some parameters ($x$-axis) were fixed. Parameters were
-        fixed by specifying a Dirac-delta prior at the true value. True values
-        are indicated by horizontal dashed lines.
-    }
-    \label{fig:marginal}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics{stock/missing}
-    \caption[
-        Comparison of \gls{BA} parameter estimates obtained with
-        \software{netabc} on the same dataset with different \gls{SMC}
-        settings.
-    ]{
-        Comparison of \gls{BA} parameter estimates obtained with
-        \software{netabc} on the same dataset with different \gls{SMC}
-        settings. Points are posterior means, notches are 50\% \gls{HPD}
-        intervals, and lines are 95\% \gls{HPD} intervals.
-    }
-    \label{fig:moresmc}
-\end{figure}
-
-\section{Application to real world HIV data}
-\label{sec:hiv}
-
-\subsection{Methods}
-
-Because the \gls{BA} model assumes a single connected contact network, it is
-most appropriate to apply to groups of individuals who are epidemiologically
-related. Therefore, we searched for published \gls{HIV} datasets which
-originated from existing clusters, either phylogenetically or geographically
-defined. \add{To identify datasets fitting these criteria, we used the
-\software{Entrez} module in the \software{BioPython}
-library~\autocite{cock2009biopython} to identify all studies which were linked
-to at least 150 sequences of the same \gls{HIV} gene in GenBank (635 at the
-time when the data was collected). We manually curated a subset of these
-articles which, based on their title and abstract, appeared to have sampled one
-sequence per individual in a group likely to be epidemiologically related. For
-example, studies were excluded if they were investigating response to a
-particular drug, pregnant women or pediatric \gls{HIV} patients, intra-host
-evolution, or a multi-region or multi-country cohort. We acknowledge that our
-perusal of the complete set of articles was rather cursory, often limited to
-reading the title, and it is quite possible that we failed to identify other
-studies which would have been suitable to include. Each potential dataset was
-revisited, and those without sampling time annotation in GenBank were excluded.
-We also excluded studies where all sequences were sampled at the same
-timepoint, which was necessary because the method we used to time scale the
-tree requires non-contemporaneous tips.} The datasets are summarized in
-\cref{tab:data}. In addition \add{to the published data}, we analyzed an
-in-house dataset sampled from \gls{HIV}-positive individuals in British
-Columbia, Canada. \add{For clarity, we will refer to each dataset by its risk
-group and location of origin in the text. For example, the
-\textcite{zetterberg2004two} data will be referred to as IDU/Estonia.}
-
-\begin{table}[ht]
-    \centering
-    \input{\tablepath/realdata-bc}
-    \caption[
-        Characteristics of published \gls{HIV} datasets analyzed with
-        \software{netabc}.
-    ]{
-        Characteristics of published \gls{HIV} datasets analyzed with
-        \software{netabc}. Abbreviations: \acrshort{MSM}, \acrlong{MSM}; HET,
-        heterosexual; \acrshort{IDU}, \acrlong{IDU}. The HET data were sampled
-        from a primarily heterosexual risk environment but did not explicitly
-        exclude other risk factors. The number of sequences column indicates
-        how many sequences were included in our analysis; there may have been
-        additional sequences linked to the study which we excluded for various
-        reasons (see methods).
-    }
-  \label{tab:data}
-\end{table}
-
-We downloaded all sequences associated with each published study from GenBank.
-\add{For the IDU/Romania data, only sequences from \gls{IDU} (whose sequence
-identifiers included the letters ``DU'') were included in the analysis.
-\textcite{kao2011surveillance} (MSM/Taiwan) found a strong association in their
-study population between subtype and risk group - subtype B was most often
-associated with \gls{MSM}, whereas \gls{IDU} were usually infected with a
-circulating recombinant form. Since there were many more subtype B sequences in
-their data than sequences of other subtypes, we restricted our analysis to the
-subtype B sequences and labelled this dataset as \gls{MSM}. Three datasets
-(IDU/Estonia, HET/Uganda, and HET/Malawi) included both \textit{env} and
-\textit{gag} sequences. Each gene was analyzed separately to assess the
-robustness of \software{netabc} to the particular \gls{HIV} gene sequence used
-to estimate a transmission tree.}
-
-\del{For the \textcite{novitsky2014impact} data,} Each \textit{env} sequence
-was aligned pairwise to the HXB2 reference sequence (GenBank accession number
-K03455), and the hypervariable regions were clipped out with
-\software{BioPython} version 1.66+~\autocite{cock2009biopython}. Sequences were
-multiply aligned using \software{MUSCLE} version 3.8.31
-\autocite{edgar2004muscle}, and alignments were manually inspected with
-\software{Seaview} version 4.4.2 \autocite{gouy2010seaview}. Phylogenies were
-constructed from the nucleotide alignments by approximate maximum likelihood
-using \software{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
-\gls{GTR} model~\autocite{tavare1986some}. Transmission trees were estimated by
-rooting and time-scaling the phylogenies by root-to-tip regression, using a
-modified version of Path-O-Gen (distributed as part of
-BEAST~\autocite{drummond2007beast}) as described
-previously~\autocite{poon2015phylodynamic}.
-
-\del{Three} \add{Four} of the datasets (MSM/Shanghai, HET/Botswana, HET/Uganda, 
-and MSM/USA) were initially larger than the others, containing 1265, 1299,
-1026/915 (\textit{env}/\textit{gag}), and 648 sequences respectively. To ensure
-that the analyses were comparable, we reduced these to a number of sequences
-similar to the smaller datasets. For the MSM/Shanghai dataset, we detected a
-cluster of size 280 using a patriotic distance cutoff of 0.02 as described
-previously~\autocite{poon2015impact}. Only sequences within this cluster were
-carried forward. For the HET/Uganda, HET/Botswana, and MSM/USA datasets, no
-large clusters were detected using the same cutoff, so we analyzed subsets of
-size 225, 180, and 180 respectively. \add{The subset of the HET/Uganda data was
-chosen by eye such that the individuals were monopolistic in both the
-\textit{gag} and \textit{env} trees. The other subsets were arbitrarily chosen
-subtrees from phylogenies of the complete datasets.}
-
-For all datasets, we used the priors $\gls{alpha} \sim \Uniform(0, 2)$ and
-\gls{N} and \gls{I} jointly uniform on the region $\{n \leq \gls{N} \leq
-10000$, $n \leq \gls{I} \leq 10000$, $\gls{I} \leq \gls{N}\}$, where $n$ is the
-number of tips in the tree (see \cref{tab:data}). Since the value $\gls{m} = 1$
-produces networks with no cycles, which we considered fairly implausible, we
-ran one analysis with the prior $\gls{m} \sim \DiscreteUniform(1, 5)$, and one
-with the prior $\gls{m} \sim \DiscreteUniform(2, 5)$. The other parameters to
-the SMC algorithm were the same as used for the simulation experiments, except
-that we used 10000 particles instead of 1000 to increase the accuracy of the
-estimated posterior. This was computationally feasible due to the small number
-of runs required for this analysis.
-
-Empirical studies of contact networks often report the exponent \gls{gamma} of
-the power law degree distribution~\autocite{colgate1989risk, liljeros2001web,
-schneeberger2004scale, clemenccon2015statistical, rothenberg2007large,
-brown2011transmission}. Although the \gls{BA} model does not produce networks
-with true power law degree distributions except when $\gls{alpha} = 1$, the
-power law still provides a reasonable approximation to the slope when
-$\gls{alpha} \neq 1$ (\cref{fig:powerlaw}).  To compare our results to
-\add{existing network} literature we simulated 100 networks each according to
-the posterior mean parameter estimates obtained for each investigated dataset.
-\add{Although the \gls{BA} model does not produce power law networks except
-when $\gls{alpha} = 1$, simulations show that the power law fit still captures
-the slope of the degree distribution reasonably well (\cref{fig:powerlaw}).}
-\gls{gamma} was calculated for each network using the
-\software{fit\_power\_law} function in \software{igraph}, with the `R.mle'
-implementation. The median of the 100 \gls{gamma} values was taken as a point
-estimate for the associated dataset.
-
-\subsection{Results}
-
-\del{We applied \software{netabc} to five published HIV datasets (\cref{tab:data}).
-and found substantial heterogeneity among the parameter
-estimates.} \add{Posterior mean point estimates and 50\% and 95\% \glspl{HPD}
-for each parameter are shown in \cref{fig:abchpd}.} \del{and marginal posterior
-distributions in \cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}}.
-\add{\Cref{fig:abchpdm2} shows point estimates and \glspl{HPD} obtained when
-the value $\gls{m} = 1$ was disallowed by the prior. Since the results
-indicated that $\gls{m} = 1$ was the most credible value for several datasets,
-all results discussed henceforth are for the prior $\gls{m} \sim
-\DiscreteUniform(1, 5)$ unless otherwise stated.}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics{realdata-hpd-bc}
-    \caption[
-        Posterior means and 50\%/95\% \gls{HPD} intervals for parameters of the
-        \gls{BA} network model, fitted to eleven
-        \gls{HIV} datasets with \software{netabc}.
-    ]{
-        Posterior means (points), 50\% \gls{HPD} intervals (notches), and 95\%
-        (lines) \gls{HPD} intervals for parameters of the \gls{BA} network
-        model, fitted to eleven \gls{HIV} datasets with \software{netabc}.
-        Legend labels indicate risk group and country of origin. Abbreviations:
-        \acrshort{IDU}, \acrlong{IDU}; \acrshort{MSM}, \acrlong{MSM}; HET,
-        heterosexual. Note that posterior means can fall outside of \glspl{HPD} 
-        if the distribution is diffuse.
-    }
-    \label{fig:abchpd}
-\end{figure}
-
-\del{Two of the datasets (\textcite{niculescu2015recent, wang2015targeting}) had
-estimated $\alpha$ values near unity for the prior allowing $m = 1$ (posterior
-mean [95\% \gls{HPD}] 
-  0.73 
-  [0.05 - 
-   1.18]
-and
-  0.55 
-  [0.01 -
-   0.99] respectively).
-The estimates did not change appreciably when $m = 1$ was disallowed by the
-prior, although the credible interval of the \textcite{niculescu2015recent}
-data was narrower
-  (0.05 - 
-   1.18).
-When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
-low estimated $\alpha$ values
-  (0.33 
-  [0 - 
-  0.76]
-and
-  0.27 
-  [0 -
-   0.59]). 
-However, the estimates increased when $m = 1$ was not permitted, although the
-HPD intervals remained roughly the same
-  (0.58 
-  [0.06 - 
-  0.99]
-and
-  0.48 
-  [0.02 -
-   0.87]).
-The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
-for both priors on $m$
-  (0.55 for $m \geq 1$;
-   0.53 for $m \geq 2$).
-However, the confidence interval was much wider when $m = 1$ was allowed
-  ([0 -
-    1.75] for $m \geq 1$ vs.
-   [0 -
-    1.75] for $m \geq 2$).}
-
-\add{Posterior mean point estimates for the \acrlong{PA} power \gls{alpha}
-were all sub-linear, ranging from 
-    0.9
-for the \gls{IDU}/Estonia data to
-    0.27
-for the mixed/Spain data. When aggregated by risk group, the average estimates
-were
-    0.73
-for \gls{IDU},
-    0.43
-for primarily heterosexual risk, and
-    0.39
-for \gls{MSM}. These values were obtained with \textit{gag} for the datasets
-where both \textit{gag} and \textit{env} were sequenced, but the estimates for
-\gls{alpha} did not change appreciably between the two genes (\cref{fig:genes}).
-There was a large amount of uncertainty associated with these estimates. 95\%
-\gls{HPD} intervals were very wide for most datasets, often encompassing almost
-the entire range from 0 to 1 (\cref{fig:abchpd}). For all the datasets except
-HET/Malawi and MSM/Beijing, the posterior mean was either outside, or very
-close to the border of, the 50\% \gls{HPD} intervals. This indicates that the
-posterior distributions were diffuse with heavy tails, rather than having most
-of their mass around the mode. When the value $\gls{m} = 1$ was disallowed by
-the prior, the separation between the \gls{IDU} datasets and the others became
-more striking (\cref{fig:abchpdm2}). All three \gls{IDU} datasets had estimated
-\gls{alpha} values at or above
-    0.9.
-The estimate for the MSM/Beijing data was slightly lower
-    (0.85) and
-the estimates for the seven remaining non-\gls{IDU} datasets were bounded above by
-    0.58.}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{realdata-genes}
-    \caption[
-        Posterior means and 50\%/95\% \gls{HPD} intervals for parameters of the
-        \gls{BA} network model, fitted to three \gls{HIV} datasets where both
-        \textit{gag} and \textit{env} genes were sequenced.
-    ]{
-        Posterior means (points), 50\% \gls{HPD} intervals (notches), and 95\%
-        \gls{HPD} intervals (lines) for parameters of the \gls{BA} network model,
-        fitted to three \gls{HIV} datasets where both \textit{gag} and
-        \textit{env} genes were sequenced.
-    }
-    \label{fig:genes}
-\end{figure}
-
-\del{For all the datasets except \citeauthor{novitsky2014impact},
-estimated values of $I$ were below 2000 when $m = 1$ was allowed, with
-relatively narrow HPD intervals compared to the nonzero prior density region
-  (\citeauthor{cuevas2009hiv}, 701 
-  [289 -
-   1279];
-   \citeauthor{niculescu2015recent}, 747
-  [136 - 
-   2378];
-  \citeauthor{li2015hiv}, 1390 
-  [310 -
-   2821];
-   \citeauthor{wang2015targeting}, 675
-  [175 - 
-   1400]).
-The \citeauthor{novitsky2014impact} data was the outlier, with a very high
-estimated $I$, and HPD interval spanning almost the entire prior region
-  (5431 
-  [183 -
-   8739]).
-The $I$ estimates and HPD intervals were generally robust to the choice of
-prior on $m$, with slightly narrower HPD intervals (compare
-\cref{fig:abchpd,fig:abchpdm2}).}
-
-\add{For all but the HET/Botswana data, the posterior mean estimates for
-the prevalence \gls{I} were between
-    372
-(IDU/Estonia) and
-    1481
-(IDU/Canada). The HET/Botswana data had a much higher estimated \gls{I} value
-    (5432)
-than the other datasets, with a very wide 95\% \gls{HPD} interval covering
-almost the entire prior region (\cref{fig:abchpd}). There was no significant
-correlation between the number of sequences in the tree and the estimated
-prevalence (Spearman correlation, 
-    $p =0.6$),
-indicating that the higher prevalence estimates were not simply due to
-increased sampling density. When both \textit{gag} and \textit{env} sequences
-were analyzed, the estimates from the \textit{env} data were substantially
-higher (\cref{fig:genes}). The ratios of internal to terminal branch lengths
-were lower for the \textit{env} trees than the \textit{gag} trees (average 
-    0.47
-for \textit{env} and 
-    0.76
-for \textit{gag}). As with \gls{alpha}, many of the posterior means were
-outside the 50\% \gls{HPD} intervals, suggesting diffuse posterior
-distributions. The values of \gls{I} were fairly robust to restricting the
-prior to $\gls{m} \geq 2$ (compare \cref{fig:abchpd,fig:abchpdm2}), although 
-the 95\% \gls{HPD} intervals were slightly narrower (average width
-    2183
-for $\gls{m} \geq 1$ and
-    1855
-for $\gls{m} \geq 2$).}
-
-\del{The posterior mean of $m$ was equal to 1 for all but the
-\citeauthor{novitsky2014impact} data, when this value was allowed. However, the
-upper bound of the HPD interval was different for each dataset
-  (\citeauthor{niculescu2015recent}, 4;
-   \citeauthor{wang2015targeting}, 1;
-   \citeauthor{li2015hiv}, 1;
-   \citeauthor{cuevas2009hiv}, 1).
-When $m = 1$ was disallowed, the estimate for all datasets was either 2 or 3,
-with HPD intervals spanning the entire prior region. The estimates for the
-total number of nodes $N$ were largely uninformative for all samples, with
-almost all posterior mean estimates greater than 7500 and HPD intervals
-spanning almost the entire nonzero prior density region. The only exception was
-the \citeauthor{li2015hiv} data, for which the estimate was lower 
-  (5916)
-when $m = 1$ was allowed.}
-
-\add{The posterior means of \gls{m} were equal to one for 
-  zero
-of the datasets analyzed. The widths of the 95\% \gls{HPD} intervals varied
-from 0 (all the mass on the estimated value) to 5 (the entire prior region).
-When the value $\gls{m} = 1$ was disallowed by the prior, the posterior means
-for all but the HET/Botswana data took on the value 3, with the \gls{HPD}
-intervals spanning the entire prior region. This is very similar to the results 
-observed for \gls{m} on simulated data (\cref{fig:abcptm2,tab:abchpd}), and
-suggests that \gls{m} is not identifiable from these data with this prior. On
-the other hand, the estimate and 95\% \gls{HPD} interval for the HET/Botswana
-data (4 [2-5]) was the same for both priors.}
-
-\add{Estimates of \gls{N} were  mostly uninformative, with very similar estimates
-for all datasets (mean
-    6234,
-range
-    5881 - 6882).
-This was similar to the pattern observed for the synthetic data, where the
-posterior mean always fell around the upper two-thirds mark of the region
-allowed by the prior. The results did not change appreciably when the value
-$\gls{m} = 1$ was assigned zero prior probability.}
-
-To compare our results to the existing network literature, we estimated
-values of the power law exponent \gls{gamma} for each of the datasets
-investigated, using the posterior median estimates \del{with the prior allowing
-$\gls{m} = 1$}. The results are summarized in \cref{tab:gamma}. All of the
-estimated exponents were in the range $2 \leq \gls{gamma} \leq 2.5$, which is
-on the lower end of the range $2 \leq \gls{gamma} \leq 4$ reported in the
-literature. \add{Because the estimates of \gls{N} were uninformative, we also
-simulated another set of networks with the estimated \gls{I} value, rather than
-\gls{N}, as the number of nodes. However, none of the estimated \gls{gamma}
-values were changed by this operation (not shown).}
-
-\begin{table}
-    \centering
-    \input{\tablepath/realdata-gamma.tex}
-    \caption[
-        Estimated power law exponents for six \gls{HIV} datasets based on
-        posterior mean point estimates of \gls{BA} model parameters.
-    ]{
-        Estimated power law exponents for six \gls{HIV} datasets based on
-        posterior mean point estimates of \gls{BA} model parameters. 100
-        networks were simulated using posterior mean parameter estimates
-        obtained with \software{netabc}. The power law exponent \gls{gamma}
-        was estimated for each, and the median of those estimates was used as a
-        point estimate for the corresponding dataset.
-    }
-    \label{tab:gamma}
-\end{table}

### 2016-07-01 doc/paper/changes.tex
updates

@@
 of the ABC approximation (more particles, more simulated datasets per particle,
-and more iterations). The results were nearly identical, TODO
+and more iterations). The results were similar; surprisingly, increased
+settings made the estimation about 10\% worse. We suspect this may be simply
+due to stochastic effects. In any event, this result demonstrates that Monte
+Carlo error from the SMC algorithm is likely not responsible for the observed
+estimation errors.
 
@@
 
-  TODO
+We have substantially expanded the discussion of the synthetic data results,
+including this point about $m$. The weaker identifiability in lower $\alpha$
+values is related to the degree distributions of the networks, and the upward
+biases in $I$ and $N$ are discussed in relation to the epidemic trajectory of
+the SI model. Our intuition about $m$ is weaker, but we include this point by
+the reviewer and hypothesize that $m$ may be more readily estimable from the
+speed of epidemic expansion rather than the shape of the transmission tree.
 
@@
 We struggled with the phrasing of this particular point and are happy to have
-recieved some guidance. TODO
+recieved some guidance. The statement now reads as follows: ``If $I$ is small
+relative to $N$, very few of the infected nodes will gain any new neighbours.
+Thus, the outcome of a second simulation on the same network will likely be
+very similar.''
 

### 2016-07-01 doc/paper/paper.Rnw
updates

@@
 duplicated sequences, all estimated transmission trees were fully binary.}
+
+\add{To check that our results were robust to the choice of phylogenetic
+method, we constructed phylogenies for two of the datasets (mixed/Spain and
+IDU/Estonia) with BEAST~\autocite{drummond2007beast}. We used the TN93 model of
+sequence evolution~\autocite{tamura1993estimation}, a strict molecular clock,
+and constant population size. The chains were run for $10^7$ steps and sampled
+every $10^3$ steps. Maximum clade credibility (MCC) trees were constructed from
+the sampled trees after discarding the first 10\% as burn-in.}
 
@@
     \centering
-    \includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior", middle.plot)}}.pdf}
+    \includegraphics[width=0.6\textwidth]{{\Sexpr{file.path("abc-posterior", middle.plot)}}.pdf}
     \caption[
@@
     marginal.I <- d[fixed == "N+alpha+m", abs(I-true_I)]
+
+    d <- setDT(collect.data("../../simulations/abc-better-smc/point-estimate/*"))
+
+    base.alpha <- d[nparticle == 1000, abs(alpha - true_alpha)]
+    better.alpha <- d[nparticle == 2000, abs(alpha - true_alpha)]
+
+    base.I <- d[nparticle == 1000, abs(I - true_I)]
+    better.I <- d[nparticle == 2000, abs(I - true_I)]
+
+    base.N <- d[nparticle == 1000, abs(N - true_N)]
+    better.N <- d[nparticle == 2000, abs(N - true_N)]
 @
@@
 of each of the BA parameters on the same simulated transmission tree. The
-estimates of $m$ are apparently unaffected by marginalizing out the other
+estimates of $m$ were apparently unaffected by marginalizing out the other
 parameters, corroborating the previous experiments' findings that $m$ is not an
@@
     \Sexpr{ round((joint.N - marginal.N) * 100 / joint.N) }\%
-when all other parameters were fixed. Neither the accuracy nor the precision of
-any parameter estimates appeared to improve when more computing power was used
-for SMC (\cref{fig:moresmc}).}
+when all other parameters were fixed. \Cref{fig:moresmc} shows the impact of
+increasing the number of particles, simulated datasets, and
+$\alpha_{\text{ESS}}$ parameter on the accuracy of a single simulation. The
+results of the two simulations were similar, but surprisingly, the results with
+higher SMC settings were slightly worse (by
+    \Sexpr{ round((better.alpha - base.alpha) * 100 / better.alpha) }\%,
+    \Sexpr{ round((better.I - base.I) * 100 / better.I) }\%, and
+    \Sexpr{ round((better.N - base.N) * 100 / better.N) }\%
+for $\alpha$, $I$, and $N$ respectively). However, the 50\% HPD interval for
+$I$ was slightly closer to the true value of 2000 with the improved settings
+    (\Sexpr{ d[nparticle == 2000, round(I.lower.50)] } - 
+     \Sexpr{ d[nparticle == 2000, round(I.upper.50)] }, vs.
+     \Sexpr{ d[nparticle == 1000, round(I.lower.50)] } - 
+     \Sexpr{ d[nparticle == 1000, round(I.upper.50)] } with basic settings).
+The estimate of $m$, 3 in both cases, was unaffected by the settings.}
 
@@
 
-% OLD STUFF
- Since the
-prior on $N$ and $I$ is jointly uniform on a non-rectangular region ($I \leq
-N$), there is more prior mass on high $N$ values.
-
-This can be
-illustrated by imagining that we add a small number of nodes to a network after
-the epidemic simulation has already been completed. It is possible that none of
-these new nodes attains a connection to any infected node. Thus, running the
-simulation again on the new, larger network could produce the exact same
-transmission tree as before. We note also that \del{our accurate estimates of
-$I$ may have been influenced by} this prior\del{, which} places more mass on
-low $I$ values. However, the estimate of $I$ was very high for the HET/Botswana
-data, suggesting that a strong enough signal in the data can
-overcome the prior. \add{Furthermore, when $I$ was estimated marginally with
-fixed $N$, the accuracy of the estimate improved even though there was no
-longer any extra prior mass on low $I$ values.}
-
-% END OLD STUFF
-
 \add{The number of nodes in the network, $N$, exhibited the most variation in
@@
 between trees simulated under different $N$ values when the number of infected
-nodes $I$ was small (\cref{fig:Ncrossv}).} $N$ was always significantly
-over-estimated using ABC (\cref{fig:abcpt2}).
-
-In retrospect, it is unreasonable to expect good estimation of $N$, because
-adding additional nodes does not change the edge density or overall shape of a
-BA network. This can be illustrated by imagining that we add a small number of
-nodes to a network after the epidemic simulation has already been completed. It
-is possible that none of these new nodes attains a connection to any infected
-node. Thus, running the simulation again on the
-new, larger network could produce the exact same transmission tree as before.
+nodes $I$ was small (\cref{fig:Ncrossv}).} In retrospect, it is unreasonable to
+expect good estimation of $N$, in many cases, because adding additional nodes
+does not change the edge density or overall shape of a BA network. This can be
+illustrated by imagining that we add a small number of nodes to a network after
+the epidemic simulation has already been completed. \add{If $I$ is small
+relative to $N$, very few of the infected nodes will gain any new neighbours.
+Thus, the outcome of a second simulation on the same network will likely be
+very similar.} \del{It is possible that none of these new nodes attains a
+connection to any infected node. Thus, running the simulation again on the new,
+larger network could produce the exact same transmission tree as before.}
 On the other hand, when $I$ is large relative to $N$, the coalescent dynamics
@@
 ABC, $N$ was nearly always over-estimated by approximately a factor of two
-(\cref{fig:abcptm2,tab:abchpd}). One factor which may have contributed to this
+(\cref{fig:abcpt2,tab:abchpd}). One factor which may have contributed to this
 bias was our choice of prior distribution. Since the prior for $I$ and $N$ was
 jointly uniform on a region where $I \leq N$, more prior weight was assigned to
-higher $N$ values.
+higher $N$ values. We note also that \del{our accurate estimates of $I$ may
+have been influenced by} this prior\del{, which} places more mass on low $I$
+values. However, the estimate of $I$ was very high for the HET/Botswana data,
+suggesting that a strong enough signal in the data can overcome the prior.
+\add{Furthermore, when $I$ was estimated marginally with fixed $N$, the
+accuracy of the estimate improved even though there was no longer any extra
+prior mass on low $I$ values.}
 
@@
 
-\end{document}
-
 As noted by \textcite{lintusaari2016identifiability}, uniform priors on model
@@
 range of permitted $\gamma$ values may be warranted.
+
+\end{document}
 
@@
 behaviour, are likely to provide a better fit to these data.
+
+\end{document}
 

### 2016-07-03 doc/paper/changes.tex
Some additional paper stuff

@@
 point makes it clearer that the classifier experiments were marginal in nature.
-(TODO: We have emphasized this point in the discussion as well: ).
 
@@
 perform a small experiment allowing heterogeneity among nodes, but in $\alpha$
-rather than in $m$, and recovered the average of the two $\alpha$ values.
+rather than in $m$, and recovered the average of the two $\alpha$ values. We
+have suggested the possibility of drawing $m$ from a distribution in the
+discussion (emphasis is new text): ``More sophisticated models, for example
+models incorporating heterogeneity in node behaviour, are likely to provide a
+better fit to these data. \emph{For example, rather than using the same $m$ for
+all nodes, we could specify a distribution from which each individual
+node's $m$ value would be drawn}.''
 
-(TODO: We
-have added to the discussion the point that this binary choice is sub-optimal
-and suggested this enhancement for future work.)
+In addition, we have outlined the difficulties of the binary choice of $m$ in
+the discussion of the real data results: ``\ldots the BA model does not allow
+any nodes with degree 1 when $m \geq 2$. For this reason, the choice of whether
+to allow $m = 1$ in the prior is problematic, as we must choose between an
+unrealistic topology (no cycles) and an unrealistic minimum degree.''
 

### 2016-07-03 doc/paper/paper.Rnw
Some additional paper stuff

@@
 
-\end{document}
-
 \subsection*{Analysis of real world HIV datasets}
@@
 sophisticated models, for example models incorporating heterogeneity in node
-behaviour, are likely to provide a better fit to these data.
-
-\end{document}
+behaviour, are likely to provide a better fit to these data. \add{For example,
+rather than using the same $m$ for all nodes, we could specify a distribution
+from which each individual node's $m$ value would be drawn.}
 
@@
 \textcite{de2007preferential} and \textit{netabc} is also the region of
-poorest identifiability (\cref{fig:abcptm2,fig:abcptm3,fig:abcptm4}). This was
-reflected in the high level of uncertainty in the estimates, with most 95\%
-HPD intervals covering the majority of the range [0, 1]. The value $\alpha =
-0.5$ was contained in the 95\% HPD interval for every dataset; consequently, it
-is not possible to say with high confidence that any of the $\alpha$ values are
-different from each other. In synthetic data, the confidence intervals around
-$\alpha$ narrowed when other parameters were marginalized out
-(\cref{fig:marginal}). Thus, it is possible that estimates of $\alpha$ could be
-made more precise by specifying either exact values or informative priors on
-the other BA parameters when these are known.}
-
-\add{I is underestimated in some cases and overestimated in others. Some of the
-datasets have most of the posterior mass on m = 1. Estimates of N are
-meaningless.}
+poorest identifiability (\cref{fig:abcpt}). This was reflected in the high
+level of uncertainty in the estimates, with most 95\% HPD intervals covering
+the majority of the range [0, 1]. The value $\alpha = 0.5$ was contained in the
+95\% HPD interval for every dataset; consequently, it is not possible to say
+with high confidence that any of the $\alpha$ values are different from each
+other. In synthetic data, the confidence intervals around $\alpha$ narrowed
+when other parameters were marginalized out (\cref{fig:marginal}). Thus, it is
+possible that estimates of $\alpha$ could be made more precise by specifying
+either exact values or informative priors on the other BA parameters when these
+are known.}
+
+\subsubsection{Other BA parameters}
+
+\add{The true HIV prevalence in a population can be difficult to estimate for
+several reasons. HIV-infected individuals may be asymptomatic for months or
+years, possibly delaying their awareness of their status. In many contexts, the
+risk factors for acquisition of HIV are illegal or stigmatized, which may
+represent a barrier to testing, treatment, and/or disclosure of status.
+Our simulation study showed that $I$ is weakly identifiable from tree shapes,
+however, the estimates of $I$ obtained with \textit{netabc} were upwardly
+biased (\cref{fig:abcpt}). In addition, our initial exploratory analysis showed
+that the identifiability of $I$ decreases with the number of sampled tips
+(\cref{fig:Icrossv}); in real world studies, the proportion of infected
+individuals sampled is usually low. The estimated $I$ values for the HIV
+datasets ranged from
+    X (D)
+to
+    X (D).}
+
+\add{We were not able to discern and trends toward over- or underestimation of
+$I$ from available prevalence data. For example, the authors of the
+HET/Botswana data~\autocite{novitsky2013phylogenetic, novitsky2014impact}
+estimated that there were 1731 \gls{HIV}-positive individuals in the study
+area. \gls{HIV} sequences were obtained from approximately 70\% of these
+individuals. The estimated prevalence we obtained was much higher 
+    (\Sexpr{d1[dataset == "novitsky2014" & parameter == "I", round(mean)]}),
+with a 95\% \gls{HPD} interval spanning nearly the entire prior region. On the
+other hand, the study which produced the MSM/USA
+data~\autocite{little2014using} enrolled 648 HIV-positive MSM; thus, our
+estimate of $I$
+    (\Sexpr{d1[dataset == "little2014" & parameter == "I", round(mean)]})
+was clearly an underestimate. Post-hoc explanations can be imagined for both of
+these results. The HET/Botswana data were collected from a town located
+proximally to the country's capital city; the authors suggested that frequent
+travel between the two locations may have facilitated linking of their sexual
+networks. Thus, the high estimate of \gls{I} we obtained for these data may
+include a larger network component based in the capital city. The MSM/USA
+result may indicate that the individuals genotyped for the study are members of
+a smaller subnetwork which does not include the entire local MSM population.
+Unfortunately, none of these hypotheses can be easily tested.}
+
+\add{Over half of the datasets were estimated to have $m = 1$, which produces
+tree-like networks without cycles. Since the average degree of a
+Barab\'asi-Albert network is $2m$, this value may simply reflect the fact that
+most people have a small number of sexual partners, especially when only
+recent partnerships are considered~\autocite{liljeros2001web}. In fact, in one
+survey, the most common number of partnerships in the past twelve months was
+one~\autocite{liljeros2001web}; the BA model does not allow any nodes with
+degree 1 when $m \geq 2$. For this reason, the choice of whether to allow $m =
+1$ in the prior is problematic, as we must choose between an unrealistic
+topology (no cycles) and an unrealistic minimum degree. Extensions to the BA
+model which relax this constraint can be imagined and may offer improved
+parameter resolution. Estimates of $N$ were not informative for any of the
+datasets under either choice of prior, consistent with our simulation results.}
 
@@
 significantly~\autocite{ypma2013relating} due to within-host evolution and the
-sampling process. In addition, the ABC-SMC algorithm is
+sampling process. Progress has very recently been made on this front by
+\textcite{giardina2016inference}, who have concurrently developed an ABC-based
+method incorporating within-host evolution. The ABC-SMC algorithm is
 computationally intensive, taking about a day when run on 20 cores in parallel
@@
 work, and previous work by our group~\autocite{poon2015phylodynamic}, has
-demonstrated that ABC is a broadly applicable and effective framework in
-which to perform phylodynamic inference.
+demonstrated that ABC is a broadly applicable and effective framework in which
+to perform phylodynamic inference.
 

### 2016-07-03 doc/defence/kpca.tex
Updates to presentation

--- 2016-06-29 doc/defence/presentation.tex
@@
-\documentclass{beamer}
-\usepackage{tikz}
-\usepackage{gillius}
-\usepackage{abraces}
-\usepackage{etex}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{qrcode}
-\usepackage[backend=biber, style=numeric-comp, sorting=none]{biblatex}
-\usepackage{hyperref}
-
-% http://tex.stackexchange.com/questions/12703/how-to-create-fixed-width-table-columns-with-text-raggedright-centered-raggedlef
-\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-
-\usetikzlibrary{external}
-\usetikzlibrary{shapes}
-\usetikzlibrary{arrows}
-\usetikzlibrary{positioning}
-\usetikzlibrary{decorations.pathreplacing}
-\usetikzlibrary{calc}
-
-\usetheme[everytitleformat=regular]{m}
-%\setbeamertemplate{navigation symbols}{}
-
-\graphicspath{{../figures/}}
-\newcommand{\tablepath}{../tables}
-
-\title{Phylogenetic inference of contact network parameters with approximate Bayesian computation}
-\author[RMM \& AFYP]{Rosemary M McCloskey$^1$ \and Bioinformatics Training Program \\ Supervisor: Art FY Poon$^{1,2}$}
-\institute[UBC \& BCCfE]{$^1$BC Centre for Excellence in HIV/AIDS, Vancouver, Canada \\ $^2$Department of Medicine, University of British Columbia, Vancouver, Canada}
-\date{M.Sc. Defence, July 26, 2016}
-
-\newcommand{\dd}[2]{\frac{\text{d}\,#1}{\text{d}\,#2}}
-
-\addbibresource{papers.bib}
-
-\begin{document}
-\setbeamercolor{background canvas}{bg=white}
-\renewcommand{\footnotesize}{\tiny}
-\definecolor{red}{RGB}{228,26,28}
-\definecolor{blue}{RGB}{55,126,184}
-\definecolor{green}{RGB}{77,175,74}
-\definecolor{purple}{RGB}{152,78,163}
-
-\maketitle
-\input{contactnet}
-\input{netmodel}
-\input{why}
-\input{ba}
-\input{phylonet}
-\input{transtree}
-\input{virusphylo}
-\input{inference}
-\input{identifiable}
-
-\end{document}
-
-\begin{frame}{yes they do}
+\begin{frame}{yes they are}
   \centerline{\includegraphics[height=0.8\textheight]{kernel-kpca}}
 \end{frame}
-
-\begin{frame}{Simulated trees can be used to fit network models}
-  \centerline{\includegraphics[width=0.9\textwidth]{abc-smc}}
-
-  \centerline{animate this?}
-\end{frame}
-
-\begin{frame}{Posterior mean estimates are more accurate for $\alpha$ and $I$}
-  \begin{minipage}[p][\textheight][t]{\textwidth}
-    \vspace{-0.5cm}
-    preferential attachment power \hfill \uncover<2>{edges per vertex}
-
-    \hspace{0.75cm}\only<1>{\includegraphics[height=0.8\textheight, trim=0 0 3.4in 0, clip]{abc-point-estimate-m3.pdf}}
-    \only<2>{\includegraphics[height=0.8\textheight]{abc-point-estimate-m3.pdf}}
-
-    prevalence \hfill \uncover<2>{total nodes}
-  \end{minipage}
-\end{frame}
-
-\begin{frame}{Example simulation\ldots}
-  1d posterior goes here
-  \pause
-%  \begin{minipage}[p][\textheight][t]{\textwidth}
-%    \begin{center}
-%    preferential attachment power \hfill prevalence
-%    \only<1>{
-%      \includegraphics[width=0.8\textwidth, trim=0 2.5in 0 0, clip]{abc-posterior-example.pdf}
-%    }
-%    \only<2>{
-%      \includegraphics[width=0.8\textwidth]{abc-posterior-example.pdf}
-%
-%      \vspace{-0.25cm}
-%      edges per vertex \hfill total nodes
-%    }
-%    \end{center}
-%  \end{minipage}
-\end{frame}
-
-\begin{frame}{Example simulation continued, $N$ and $I$ confounded}
-  2d posterior goes here
-  %\includegraphics[width=\textwidth]{{abc-posterior-2d/1.0_1000_3_5000_0}.pdf}
-\end{frame}
-
-\begin{frame}{Application to real world HIV datasets}
-  \centerline{\input{\tablepath/realdata-bc-short}}
-\end{frame}
-
-\begin{frame}{HIV networks have sub-linear preferential attachment}
-    Also, higher PA for IDU networks.
-
-    \includegraphics[trim=0 3.5in 3.3in 0, clip, width=0.5\textwidth]{realdata-hpd-bc}
-    \includegraphics[trim=3.4in 1.5in 0 2.0in, clip, width=0.5\textwidth]{realdata-hpd-bc}
-\end{frame}
-
-\begin{frame}{Conclusions}
-  \begin{itemize}
-    \item We developed a phylodynamic method to fit contact network models to
-      phylogenetic data.
-      \pause
-    \item The preferential attachment power of the Barab\'asi-Albert network
-      model, which is challenging to estimate by traditional epidemiological
-      methods, can be estimated with ABC.
-      \pause
-    \item The networks underlying real epidemics are heterogeneous,
-      underscoring the importance of considering network structure in
-      phylodynamic analyses.
-  \end{itemize}
-\end{frame}
-
-\begin{frame}{Acknowledgements}
-  \begin{columns}
-    \begin{column}{0.6\textwidth}
-
-      \textbf{BC Centre for Excellence in HIV/AIDS}
-
-      Art Poon
-
-      Jeff Joy
-
-      Richard Liang
-
-      Thuy Nguyen
-
-      P. Richard Harrigan
-
-      \hfill\\
-      \textbf{University of British Columbia}
-
-      Sarah Otto
-
-      Alexandre Bouchard-C\^ot\'e
-
-      \vfill
-      \vspace{0.5cm}
-      $\qquad\qquad$\includegraphics[width=2cm]{logos/genomecanada}
-      $\qquad$
-      \includegraphics[width=2cm]{logos/genomequebec}
-    \end{column}
-    \begin{column}{0.4\textwidth}
-      \centering
-
-      \includegraphics[width=3cm]{logos/cfe}
-      \vspace{0.5cm}
-
-      \includegraphics[width=3cm]{logos/cihr}
-
-      \includegraphics[width=3cm]{logos/genomebc}
-
-      \includegraphics[width=3cm]{logos/bmgf}
-      \vspace{0.5cm}
-
-      \includegraphics[width=3cm]{logos/btp}
-    \end{column}
-  \end{columns}
-\end{frame}
-
-\end{document}

### 2016-07-05 doc/defence/objects.tex
Presentation updates

\begin{frame}{Transmission trees connect contact networks to viral phylogenies}
    \begin{tikzpicture}
      [
        every path/.style={very thick, ->, >=stealth}
      ]
      \uncover<1-2>{
        \node (a) at (0, 0) [anchor=center] {\includegraphics[width=\textwidth, trim=0 0 5cm 0, clip]{objects}};
        \node [below=0 of a] {contact network \qquad transmission tree \qquad viral phylogeny};
      }
      \only<1>{ 
        \fill [white] (-1.6, 2.1) rectangle (2.4, -2.7); 
      }
    \end{tikzpicture}
\end{frame}

### 2016-07-05 doc/defence/phylo.tex
Presentation updates

\begin{frame}{Viral phylogenies relate individuals' viral genotypes}
  \begin{center}
    \begin{tikzpicture}[
        every node/.style = {inner sep=0pt},
        every path/.style = {-, very thick},
        label/.style = {anchor=south, color=white, inner sep=4pt}
      ]
      \def\pht{1cm}

      \coordinate (abcd);
      \coordinate [below right=1 and 1 of abcd] (b);
      \coordinate [above left=1 and 0.25 of abcd] (acd);
      \coordinate [above left=1 and 0.25 of acd] (a);
      \coordinate [right=of acd] (cd);
      \coordinate [right=1.5 of cd] (c);
      \coordinate [above right=of cd] (d);

      \node (x) [right=0.5 of b, anchor=west, opacity=0.4] {\includegraphics[height=\pht]{stock/person}}; 
      \node (v) at (x.south west) {\includegraphics[height=\pht]{stock/virus}}; 
      \node at (v.center) {\large b};

      \draw (abcd) -- (acd);
      \draw (abcd) -- (b);

      \node (x) [above=0.7 of a, anchor=south west, opacity=0.4] {\includegraphics[height=\pht]{stock/person}};
      \node (va) at (x.south west) {\includegraphics[height=\pht]{stock/virus}};
      \node at (va.center) {\large a};

      \draw (acd) -- (cd);
      \draw (acd) -- (a);

      \node (x) [above right=0 and 0.8 of c, opacity=0.4] {\includegraphics[height=\pht]{stock/person}};
      \node (v) at (x.south west) {\includegraphics[height=\pht]{stock/virus}};
      \node at (v.center) {\large c};

      \node (x) [above right=0.8 of d, opacity=0.4] {\includegraphics[height=\pht]{stock/person}};
      \node (v) at (x.south west) {\includegraphics[height=\pht]{stock/virus}};
      \node at (v.center) {\large d};

      \draw (cd) -- (c);
      \draw (cd) -- (d);

      \uncover<2->{
        \node (t1) [below left=of a, inner ysep=6pt] {viruses within individuals};
        \draw [->, >=stealth] (t1.north) -- (va.south west);
      }

      \uncover<3->{
        \node (t2) [below=of t1, inner ysep=6pt] {evolutionary relationships};
        \coordinate [left=0.5 of abcd] (tmp);
        \draw [->, >=stealth] (t2.north) -- (tmp);
      }

    \end{tikzpicture}
  \end{center}
\end{frame}

### 2016-07-05 doc/defence/pointest.tex
Presentation updates

--- 2016-07-01 doc/paper/paper.Rnw
@@
-\documentclass[12pt]{article}
-\usepackage[T1]{fontenc}
-\usepackage{mathptmx}
-\usepackage[backend=biber, style=numeric-comp, sorting=none]{biblatex}
-\usepackage{fullpage}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{setspace}
-\usepackage{cleveref}
-\usepackage[normalem]{ulem}
-\usepackage{color}
-\usepackage{amsmath}
-
-\addbibresource{papers.bib}
-
-% beginning of diff stuff
-\usepackage{twoopt}
-
-\let\mref\cref
-\let\Mref\Cref
-\let\mtextcite\textcite
-\let\mautocite\autocite
-\let\mciteauthor\citeauthor
-\renewcommand{\cref}[1]{\mbox{\mref{#1}}}
-\renewcommand{\Cref}[1]{\mbox{\Mref{#1}}}
-\renewcommandtwoopt{\textcite}[3][][]{\mbox{\mtextcite[#1][#2]{#3}}}
-\renewcommandtwoopt{\autocite}[3][][]{\mbox{\mautocite[#1][#2]{#3}}}
-\renewcommand{\citeauthor}[1]{\mbox{\mciteauthor{#1}}}
-
-\newcommand{\add}[1]{\color{blue} \uline{#1} \color{black}}
-\newcommand{\del}[1]{\color{red} \sout{#1} \color{black}}
-% end of diff stuff
-
-\newcommand{\tablepath}{../tables}
-\graphicspath{{../figures/}}
-
-\begin{document}
-\onehalfspacing
-
-\title{Reconstructing contact network parameters from viral phylogenies}
-\author{Rosemary M. McCloskey$^1$, Richard H. Liang$^1$, and Art F.Y. Poon$^{1,2}$ \\
-\small $^1$BC Centre for Excellence in HIV/AIDS, Vancouver, Canada \\ \small $^2$ Department of Medicine, University of British Columbia, Vancouver, Canada}
-
-<<setup, include=FALSE>>=
-    library(netabc)
-    library(Hmisc)
-
-    pp <- function (p) {
-        if (p < 1e-5) {
-            "{<}10^{-5}"
-        } else {
-            paste0("=", latexSN(round(p, -floor(log10(p)))))
-        }
-    }
-@
-
-\maketitle
-\abstract{\input{abstract}}
-
-\section*{Introduction}
-
-When an infectious disease spreads through a population, transmissions are
-generally more likely to occur between certain pairs of individuals. Such pairs
-must have a particular mode of contact with one another, which varies with the
-mode of transmission of the disease. For airborne pathogens, physical proximity
-may be sufficient, while for sexually transmitted diseases, sexual or in some
-cases blood-to-blood contact is required. The population together with the set
-of links between individuals along which transmission can occur is called the
-contact network~\autocite{klovdahl1985social, morris1993epidemiology}. The
-structure of the contact network underlying an epidemic can profoundly impact
-the speed and pattern of the epidemic's expansion. Network structure can
-influence the prevalence trajectory~\autocite{o2011contact, ma2013importance}
-and epidemic threshold~\autocite{barthelemy2005dynamical}, in turn affecting
-the estimates of quantities such as effective viral population
-size~\autocite{goodreau2006assessing}.  From a public health perspective,
-contact networks have been explored as tools for curtailing epidemic spread, by
-way of interventions targeted to well-connected
-nodes~\autocite{wang2015targeting}. True contact networks are a challenging
-type of data to collect, requiring extensive epidemiological
-investigation~\autocite{welch2011statistical, eames2015six}.
-
-Viral sequence data, on the other hand, has become \del{relatively inexpensive
-and straightforward to collect on a population level.} \add{easier to collect
-as the cost of sequencing has declined. In the case of HIV, genotyping has
-become part of routine clinical care in several health regions.} Due to the
-high mutation rate of RNA viruses, epidemiological processes impact the course
-of viral evolution, thereby shaping the inter-host viral
-phylogeny~\autocite{drummond2003measurably}. The term ``phylodynamics'' was
-coined to describe this interaction, as well as the growing family of inference
-methods to estimate epidemiological parameters from viral
-phylogenies~\autocite{grenfell2004unifying}. These methods have revealed
-diverse properties of local viral outbreaks, from basic reproductive
-number~\autocite{stadler2012estimating}, to the degree of
-clustering~\autocite{hughes2009molecular}, to the elevated transmission risk
-during acute infection~\autocite{volz2012simple}. On the other hand, although
-sophisticated methods have been developed for fitting complex population
-genetic models to phylogenies~\autocite{volz2012complex,
-rasmussen2014phylodynamic}, inference of structural network parameters has to
-date been limited. However, it has been shown that network structure has a
-tangible impact on phylogeny shape~\autocite{leventhal2012inferring,
-colijn2014phylogenetic, goodreau2006assessing, robinson2013dynamics,
-villandre2016assessment}, suggesting that such statistical inference might be
-possible~\autocite{welch2011statistical}. \add{In the context of networks,
-sequence data have the advantage of being objective, in that they are not
-affected by misreporting. However, just as with survey data, it is important to
-collect a representative sample from the population to perform accurate
-inference~\autocite{novitsky2014impact}.}
-
-Survey-based studies of sexual networks~\autocite{colgate1989risk,
-liljeros2001web, schneeberger2004scale, latora2006network, rothenberg2007large,
-clemenccon2015statistical} have found that these networks tend to have a degree
-distribution which follows a power law \autocite[although there has been some
-disagreement, see][]{handcock2004likelihood}. That is, the number of nodes of
-degree $k$ is proportional to $k^{-\gamma}$ for some constant $\gamma$. These
-networks are also referred to as
-``scale-free''~\autocite{barabasi1999emergence}. One process by which
-scale-free networks can be generated is preferential attachment, where nodes
-with a high number of contacts attract new connections at an elevated rate. The
-first contact network model incorporating preferential attachment was
-introduced by \textcite{barabasi1999emergence}, and is now referred to as the
-Barab\'asi-Albert (BA) model. Under this model, networks are formed by
-iteratively adding nodes with $m$ new edges each. In the most commonly studied
-formulation, these new edges are joined to existing nodes of degree $k$ with
-probability proportional to $k$, so that nodes of high degree tend to attract
-more connections. \citeauthor{barabasi1999emergence} suggested an extension
-where the probability of attaching to a node of degree $k$ is $k^\alpha$ for
-some non-negative constant $\alpha$, and we use this extension in this work.
-\add{When $\alpha \neq 1$, the degree distribution is no longer a power law:
-for $\alpha < 1$, the distribution is a stretched exponential, while for
-$\alpha > 1$, it is a ``gelation'' type distribution where one or a few hub
-nodes are connected to nearly every other node in the
-graph~\autocite{krapivsky2000connectivity}.}
-
-Previous work offers precedent for the possibility of statistical inference of
-structural network parameters. \textcite{britton2002bayesian} develop a Bayesian
-approach to estimate the edge density in an Erd\H{o}s-R\'enyi
-network~\autocite{erdos1960evolution} given observed infection dates, and
-optionally recovery dates. Their approach was later extended by
-\textcite{groendyke2011bayesian} and applied to a much larger data set of 188
-individuals. \textcite{volz2007susceptible, volz2008sir} developed differential
-equations describing the spread of a susceptible-infected (SI) epidemic on
-static and dynamic contact networks with several degree distributions, which
-could in principle be used for inference if observed incidence trajectories
-were available. \textcite{brown2011transmission} analysed the degree distribution
-of an approximate transmission network, estimated based on genetic similarity
-and estimated times of infection, relating 60\% of HIV-infected men who have
-sex with men (MSM) in the United Kingdom. The transmission network is a
-subgraph of the contact network which includes only those edges which have
-already led to a new infection. The authors found that a Waring distribution,
-which is produced by a more sophisticated preferential attachment model, was a
-good fit to their estimated network.
-
-Standard methods of model fitting involve calculation of the likelihood of
-observed data under the model. In maximum likelihood estimation, a quantity
-proportional to the likelihood is optimized, often through a standard
-multi-dimensional numerical optimization procedure. Bayesian methods integrate
-prior information by optimizing the posterior probability instead. To avoid
-calculation of a normalizing constant, Bayesian inference is often performed
-using Markov chain Monte Carlo (MCMC), which uses likelihood \emph{ratios} in
-which the normalizing constants cancel out. Unfortunately, it is generally
-difficult to explicitly calculate the likelihood of an observed transmission
-tree under a contact network model, even up to a normalizing constant. To do
-so, it would be necessary to integrate over all possible networks, and also
-over all possible labellings of the internal nodes of the transmission tree.
-While it is not known (to us) whether such integration is tractable, a simpler
-alternative is offered by likelihood-free methods, namely approximate Bayesian
-computation (ABC)~\autocite{tavare1997inferring, beaumont2002approximate}. ABC
-leverages the fact that, although calculating the likelihood may be
-impractical, generating simulated datasets according to a model is often
-straightforward. If our model fits the data well, the simulated data it
-produces should be similar to the observed data. More formally, if $D$ is the
-observed data, the posterior distribution $f(\theta \mid D)$ on model
-parameters $\theta$ is replaced as the target of statistical inference by
-$f(\theta \mid \rho(\hat{D}, D) < \varepsilon)$, where $\rho$ is a distance
-function, $\hat{D}$ is a simulated dataset according to $\theta$, and
-$\varepsilon$ is a small tolerance~\autocite{sunnaaker2013approximate}. \del{In
-the specific case when $\rho$ is a kernel function, the approach is known as
-ABC~\autocite{nakagome2013kernel, poon2015phylodynamic}.} \add{Our
-group~\autocite{poon2015phylodynamic} and others~\autocite{park2015k2} have
-demonstrated that taking $\rho$ to be a well-chosen kernel function can produce
-a more accurate ABC approximation than the typical choice of a difference of
-summary statistics.}
-
-Here, we develop a method using ABC to estimate the parameters of contact
-network models from observed phylogenetic data. The distance function we use is
-the tree kernel developed by \textcite{poon2013mapping}, which computes a
-weighted dot product of the trees' representations in the space of all possible
-subset trees. We apply the method to investigate the parameters of the BA
-network model on a variety of simulated and real datasets. Our results show
-that some network parameters can be inferred with reasonable accuracy, while
-others \del{have a minimal detectable impact on tree shape and therefore cannot
-be estimated accurately}\add{are weakly- or non-identifiable with ABC}. We also
-find that these parameters can vary considerably between real epidemics from
-different settings. \add{TODO: something about sub-linear PA and IDU}
-
-\section*{Methods}
-
-\subsection*{\textit{Netabc}: phylogenetic inference of contact network
-parameters with ABC}
-
-We have developed an ABC-based method to perform statistical inference of
-contact network parameters from a transmission tree estimated from an observed
-viral phylogeny. We implemented the adaptive sequential Monte Carlo (SMC)
-algorithm for ABC developed by \textcite{del2012adaptive}. The SMC algorithm
-keeps track of a population of parameter ``particles'', which are initially
-sampled from the parameters' joint prior distribution. Several datasets are
-simulated under the model of interest for each of the particles. In this case,
-the datasets are transmission trees, which are generated by a two-step process.
-First, a contact network is simulated according to the network model being fit.
-Second, a transmission tree is simulated over that network with a Gillespie
-simulation algorithm~\autocite{gillespie1976general}, in the same fashion as
-several previous studies \autocite[\textit{e.g.}][]{robinson2013dynamics,
-leventhal2012inferring}. \add{Tips of the simulated transmission tree are
-randomly removed until the simulated tree has the same number of tips as the
-input tree.} The particles are weighted according to the similarity between
-their associated simulated trees and the observed tree. To quantify this
-similarity, we used the tree kernel developed by \textcite{poon2013mapping}.
-Particles are iteratively perturbed by applying a Metropolis-Hastings kernel
-and, if the move is accepted, simulating new datasets under the new parameters.
-When a particle's weight drops to zero,
-because its simulated trees are too dissimilar to the observed tree, the
-particle is dropped from the population, and eventually replaced by a resampled
-particle with a higher weight. As the algorithm progresses, the population
-converges to a Monte Carlo approximation of the ABC target distribution, which
-is assumed to approximate the desired posterior~\autocite{del2012adaptive,
-sunnaaker2013approximate}.
-
-\add{In the original formulation of ABC-SMC~\autocite{sisson2007sequential,
-beaumont2009adaptive}, the user is required to specify a decreasing sequence of
-tolerances $\lbrace\varepsilon_i\rbrace$. At iteration $i$, particles with no
-associated simulated datasets within distance $\varepsilon_i$ of the observed
-data are removed from the population. In the adaptive version of
-\textcite{del2012adaptive}, the sequence of tolerances is determined
-automatically by fixing the decay rate of the population's expected sample size
-(ESS) to a user-defined value. \citeauthor{del2012adaptive} call this value
-$\alpha$, but we will refer to it here as $\alpha_{\text{ESS}}$ to avoid
-confusion with the preferential attachment power parameter of the BA model.}
-
-To check that our implementation of Gillespie simulation was correct, we
-reproduced Figure 1A of \textcite{leventhal2012inferring} (our
-\cref{fig:leventhal}), which plots the unbalancedness of transmission trees
-simulated over four network models at various levels of pathogen
-transmissibility. Our implementation of adaptive ABC-SMC was tested by applying
-it to the same mixture of Gaussians used by \citeauthor{del2012adaptive} to
-demonstrate their method (originally used by~\textcite{sisson2007sequential}).
-We were able to obtain a close approximation to the function (see
-\cref{fig:smctest}), and attained the stopping condition used by the authors in
-a comparable number of steps.
-
-Nodes in our networks followed simple SI dynamics, meaning that they became
-infected at a rate proportional to their number of infected neighbours, and
-never recovered. For all analyses, the transmission trees' branch lengths were
-scaled by dividing by their mean. We used the \textit{igraph} library's
-implementation of the BA model~\autocite{csardi2006igraph} to generate the
-graphs. The analyses were run on Westgrid (\url{https://www.westgrid.ca/}) and
-a local computer cluster. \add{A computer program implementing our method is
-freely available at \url{https://github.com/rmcclosk/netabc} (last accessed
-\today).}
-
-\subsection*{\del{Kernel classifiers}\add{Classifiers for BA model parameters from tree shapes}}
-
-\add{We considered four parameters related to the BA model, denoted $N$, $m$,
-$alpha$, and $I$. The first three of these parameterize the network structure,
-while $I$ is related to the simulation of transmission trees over the network.
-However, we will refer to all four as BA parameters. $N$ denotes the total
-number of nodes in the network, or equivalently, susceptible individuals in the
-population. $m$ is the number of new undirected edges added for each new
-vertex, or equivalently one-half of the average degree. $\alpha$ is the power
-of preferential attachment -- new nodes are attached to existing nodes of
-degree $d$ with probability proportional to $d^{\alpha} + 1$. Finally, $I$ is
-the number of infected individuals at the time when sampling occurs. The
-$\alpha$ parameter is unitless, while $m$ has units of edges or connections per
-vertex, and $N$ and $I$ both have units of nodes or individuals.}
-
-\add{Before proceeding with a full validation of \textit{netabc} on simulated
-data, we undertook two experiments designed to assess the identifiability of
-the BA parameters. These experiments only investigated one parameter of the
-BA model at a time while holding all others fixed, a strategy commonly used
-when performing sensitivity analyses of mathematical models. This allowed us to
-perform a fast preliminary analysis without dealing with the ``curse of
-dimensionality'' of the full parameter space. We simulated trees under three
-different values of each parameter, and asked how well we could tell the
-different trees apart. The better we are able to distinguish the trees, the
-more identifiability we might expect for the corresponding parameter when we
-attempt to estimate it with ABC.}
-
-\add{This experiment also had the secondary purpose of validating our choice of
-the tree kernel as a distance measure in ABC. To tell the trees apart, we used
-a classifier based on the tree kernel, but we also tested two other tree shape
-statistics. Sackin's index~\autocite{shao1990tree} is a measure of tree
-imbalance which not take branch lengths into account, considering only the
-topology. The normalized lineages-through-time
-\autocite[nLTT,][]{janzen2015approximate} compares two trees based on normalized
-distributions of their branching times, and does not explicitly consider the
-topology. Since the tree kernel incorporates both of these sources of
-information, we expected it to outperform the other two statistics. Finally,
-the tree kernel can be tuned by adjusting the values of the meta-parameters
-$\lambda$ and $\sigma$ (the ``decay factor'' and ``radial basis function
-variance'', see \textcite{poon2013mapping}). $\lambda$ is used to penalize
-ween large subset trees which tend to dominate the kernel score. When $\lambda
-= 0$, all but the root branches of each subset tree are ignored, while when
-$\lambda = 1$, no penalty is applied. $\sigma$ controls how strictly the notion
-of similarity is applied to branch lengths. When $\sigma = 0$, branch lengths
-must match exactly, while as $\sigma \to \infty$, branch lengths are not
-considered at all.}
-
-\del{We used the phylogenetic kernel developed by \textcite{poon2013mapping}}
-To test whether the parameters of the BA model had a \add{measurable} effect on
-tree shape, 100 networks were simulated under each of three different values of
-$\alpha$: 0.5, 1.0, and 1.5 (300 networks total). The other parameters were
-fixed to the following values: $N = 5000$, $I = 1000$, and $m = 2$. A
-transmission tree with 500 tips was simulated over each network (300
-transmission trees total). The 300 trees were compared pairwise with the tree
-kernel to form a $300 \times 300$ kernel matrix. The kernel meta-parameters
-$\lambda$ \del{(the ``decay factor''),} and $\sigma$ \del{(the ``radial basis
-function variance'')~\autocite[see][]{poon2013mapping},} were set to 0.3 and 4
-respectively. \add{We also computed a $300 \times 300$ matrix of pairwise nLTT
-values, and a $1 \times 300$ vector of Sackin's index values.} We constructed 
-\add{three classifiers for $\alpha$:} a kernel support vector regression (kSVM)
-\add{from the kernel matrix} with the \textit{kernlab}
-package~\autocite{zeileis2004kernlab}, and two ordinary SVMs \add{from the nLTT
-matrix and Sackin's index vector} with the e1071
-package~\autocite{meyer2015e1071}.\del{, and a linear regression from the
-Sackin's index values.} The accuracy of each classifier was evaluated with 1000
-two-fold cross validations \add{with equally-sized folds}.
-
-Three similar experiments were performed for the other BA model parameters (one
-experiment per parameter). $m$ was varied between 2, 3, and 4; $I$ between 500,
-1000, and 2000; and $N$ between 3000, 5000, and 8000. The parameters not being
-tested were fixed at the values $N$ = 5000, $I$ = 1000, $m$ = 2, and $\alpha$ =
-1. Thus, we performed a total of four cross-validations \add{for each
-classifier}, one for each of the BA model parameters $\alpha$, $I$, $m$, and
-$N$. We repeated these four cross-validations with different values of
-$\lambda$ (0.2, 0.3, and 0.4) and $\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$),
-as well as on trees with differing numbers of tips (100, 500, and 1000).
-\add{For the structural parameters $\alpha$, $m$, and $N$, the experiments were
-repeated with three different fixed values of $I$ (500, 1000, and 2000).}
-\del{and in epidemics of differing size (500, 1000, and 2000).} The combination
-of the number of sampled individuals (\textit{i.e.} the number of tips) and the
-epidemic size (\textit{i.e.} $I$) will be referred to as an ``epidemic
-scenario''. When evaluating the classifier for $I$, we did not consider trees
-with 1000 tips, because one of the tested $I$ values was 500, and the number of
-tips cannot be larger than $I$.
-
-\del{For each of the four parameters, we also tested a linear regression against
-Sackin's index~\autocite{shao1990tree} and an ordinary SVM based on the normalized
-lineages-through-time (nLTT) statistic~\autocite{janzen2015approximate}.}
-
-\subsection*{ABC simulations}
-
-\add{We tested \textit{netabc} by jointly estimating the four parameters of the
-BA model. We used the standard validation approach of simulating transmission
-trees under the model with known parameter values and attempting to recover
-those values with \textit{netabc}. The algorithm was not informed of any of
-the true parameter values for the main set of simulations.} We simulated three
-transmission trees, each with 500 tips, under every element of the Cartesian
-product of these parameter values: $N$ = 5000, $I$ = \{1000, 2000\}, $m$ = \{2,
-3, 4\}, and $\alpha$ = \{0.0, 0.5, 1, 1.5\}. This produced a total of 24
-parameter combinations $\times$ three trees per combination = 72 trees total.
-The adaptive ABC algorithm was applied to each tree with these priors: $m \sim$
-DiscreteUniform(1, 5), $\alpha \sim$ Uniform(0, 2), and $(N, I)$ jointly
-uniform on the region \{$500 \leq N \leq 15000$, $500 \leq I \leq 5000$, $I
-\leq N$\}. Proposals for $\alpha$, $N$, and $I$ were Gaussian, while proposals
-for $m$ were Poisson. Following \textcite{del2012adaptive} and
-\textcite{beaumont2009adaptive}, the variance of all proposals was equal to the
-empirical variance of the particles.
-
-The SMC settings used were 1000 particles, 5 simulated datasets per particle,
-and \del{the ``quality'' parameter controlling the decay rate of the tolerance
-$\varepsilon$ set to}\add{$\alpha_{\text{ESS}} =  0.95$}. We used the same
-stopping criterion as \citeauthor{del2012adaptive}, namely when the MCMC
-acceptance rate dropped below 1.5\%. \del{Point estimates for the parameters
-were obtained by taking the highest point of an estimated kernel density on the
-final set of particles, calculated using the \textit{density} function with the
-default parameters in \textit{R}.} \add{Approximate posterior means for the
-parameters were obtained by taking the weighted average of the final set of
-particles.} Highest posterior density (HPD) intervals were calculated with the
-\textit{HPDinterval} function from the \textit{R} package
-\textit{coda}~\autocite{plummer2006coda}.
-
-\add{To evaluate the effects of the true parameter values on the accuracy of
-the posterior mean estimates, we analyzed the $\alpha$ and $I$ parameters
-individually using generalized linear models (GLMs) The response variable was
-the error of the point estimate, and the predictor variables were the true
-values of $\alpha$, $I$, and $m$. We did not test for differences across true
-values of $N$, because $N$ was not varied in these simulations. The
-distribution family and link function for the GLMs were Gaussian and inverse,
-respectively, chosen by examination of residual plots and Akaike information
-criteria (AIC). The $p$-values of the estimated GLM coefficients were corrected
-using Holm-Bonferroni correction~\autocite{holm1979simple} with $n =
-6$ (two GLMs with three predictors each). Because there was clearly little to
-no identifiability of $N$ and $m$ with ABC (see results in next section), we
-did not construct GLMs for those parameters.}
-
-Two further simulations were performed to address \del{potential sources of
-error} \add{the possible impact of two types of model misspecification}. To
-evaluate the effect of model misspecification in the case of heterogeneity
-among nodes, we generated a network where half the nodes were attached with
-power $\alpha$ = 0.5, and the other half with power $\alpha$ = 1.5. The other
-parameters for this network were $N$ = 5000, $I$ = 1000, and $m$ = 2. To
-investigate the effects of potential sampling bias, we simulated a transmission
-tree where the tips were sampled in a peer-driven fashion, rather than at
-random. That is, the probability to sample a node was twice as high if any of
-that node's network peers had already been sampled. The parameters of this
-network were $N= 5000$, $I = 2000$, $m = 2$, and $\alpha = 0.5$.
-
-\add{Despite the fact that the parameter values used to generate the simulated
-transmission trees were known, the true posterior distributions of the BA
-parameters were unknown. Therefore, any apparent errors or biases in the
-estimates could be due to either poor performance of our method, or to real
-features of the posterior distribution. Two retrospective experiments were
-performed to disambiguate some of the observed errors. To assess the impact of
-the SMC settings on \textit{netabc}'s accuracy, we ran \textit{netabc}
-twice on the same simulated transmission tree. For the first run, the SMC
-settings were the same as in the other simulations: 1000 particles, 5 simulated
-transmission trees per particle, and $\alpha_{\text{ESS}} = 0.95$. The second
-run was performed with 2000 particles, 10 simulated transmission trees per
-particle, and $\alpha_{\text{ESS}} = 0.97$. To investigate the extent to which
-errors in the estimated BA parameters were due to true features of the
-posterior, rather than an inaccurate ABC approximation, we performed marginal
-estimation for one set of parameter values. Each combination of 1, 2, or 3
-model parameters (14 combinations total) was fixed to their known values, and
-the remaining parameters were estimated with \textit{netabc}. The parameter
-values were $\alpha = 0.0$, $m = 2$, $I = 2000$, and $N = 5000$.}
-
-\subsection*{Investigation of published data}
-
-We applied our ABC method to ten published HIV datasets. Because the BA model
-generates networks with a single connected component, we specifically searched
-for datasets which originated from existing clusters, either phylogenetically
-or geographically defined. Characteristics of the datasets we investigated are
-given in \cref{tab:data}. \add{For clarity, we will refer to each dataset by
-its risk group and location of origin in the text. For example, the
-\textcite{zetterberg2004two} data will be referred to as IDU/Estonia.}
-
-\begin{table*}[!t]
-  \centering
-  \input{\tablepath/realdata}
-  \caption{
-    Characteristics of published datasets investigated with ABC.
-    Acronyms: MSM, men who have sex with men; IDU, injection drug users; HET,
-    heterosexual. The HET data were sampled from a primarily heterosexual risk
-    environment, but did not explicitly exclude other risk factors. The number
-    of sequences column indicates how many sequences were included in our
-    analysis; there may have been additional sequences linked to the study
-    which we excluded for various reasons (see methods).
-  }
-  \label{tab:data}
-\end{table*}
-
-We downloaded all sequences associated with each published study from GenBank.
-\add{For the IDU/Romania data, only sequences from injection drug users (IDU,
-whose sequence identifiers included the letters ``DU'') were included in the
-analysis. \textcite{kao2011surveillance} (MSM/Taiwan) found a strong
-association in their study population between subtype and risk group - subtype
-B was most often associated with men who have sex with men (MSM), whereas IDU
-were usually infected with a circulating recombinant form. Since there were
-many more subtype B sequences in their data than sequences of other subtypes,
-we restricted our analysis to the subtype B sequences and labelled this dataset
-as MSM. Three datasets (IDU/Estonia, HET/Uganda, and HET/Malawi) included both
-\textit{env} and \textit{gag} sequences. Each gene was analyzed separately to
-assess the robustness of \textit{netabc} to the particular HIV gene sequence
-used to estimate a transmission tree.}
-
-\del{For the \textcite{novitsky2014impact} data,} Each \textit{env} sequence
-was aligned pairwise to the HXB2 reference sequence (GenBank accession number
-K03455), and the hypervariable regions were clipped out with
-\textit{BioPython} version 1.66+~\autocite{cock2009biopython}. Sequences were
-multiply aligned using \textit{MUSCLE} version 3.8.31
-\autocite{edgar2004muscle}, and alignments were manually inspected with
-\textit{Seaview} version 4.4.2 \autocite{gouy2010seaview}. \add{Duplicated
-sequences were removed with BioPython.} Phylogenies were constructed from the
-nucleotide alignments by approximate maximum likelihood using
-\textit{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
-generalized time-reversible (GTR) model~\autocite{tavare1986some}. Transmission
-trees were estimated by rooting and time-scaling the phylogenies by root-to-tip
-regression, using a modified version of Path-O-Gen (distributed as part of
-BEAST~\autocite{drummond2007beast}) as described
-previously~\autocite{poon2015phylodynamic}. \add{Due to the removal of
-duplicated sequences, all estimated transmission trees were fully binary.}
-
-\del{Two} \add{Four} of the datasets \del{\autocite{li2015hiv,novitsky2014impact}}
-(MSM/Shanghai, HET/Botswana, HET/Uganda, and MSM/USA)
-were initially much larger than the others, containing 1265, 1299,
-\add{1026/915 (\textit{env}/\textit{gag}), and 648} sequences respectively. To
-ensure that the analyses were comparable, we reduced these to a number of
-sequences similar to the smaller datasets. For the MSM/Shanghai data, we
-detected a cluster of size 280 using a patristic distance cutoff of 0.02 as
-described previously~\autocite{poon2015impact}. Only sequences within this
-cluster were carried forward. For the HET/Uganda, HET/Botswana, and MSM/USA
-data, no large clusters were detected using the same cutoff, so we analysed
-\del{a subtree} subsets of sizes 255, 180, and 180 respectively. \add{The
-subset of the HET/Uganda data was chosen by eye such that the individuals were
-monophyletic in both the \textit{gag} and \textit{env} trees. The other
-subsets were arbitrarily chosen subtrees from phylogenies of the complete
-datasets.}
-
-For all datasets, we used the priors $\alpha$ $\sim$ Uniform(0, 2) and $N$ and
-$I$ jointly uniform on the region \{$n \leq N \leq 10000$, $n \leq I \leq
-10000$, $I \leq N$\}, where $n$ is the number of tips in the tree. Since the
-value $m = 1$ produces networks with no cycles, which we considered fairly
-implausible, we ran one analysis with the prior $m \sim$ DiscreteUniform(1, 5),
-and one with the prior $m \sim$ DiscreteUniform(2, 5). The other parameters to
-the SMC algorithm were the same as used for the simulation experiments,
-\add{except that we used 10000 particles instead of 1000 to increase the
-accuracy of the estimated posterior. This was computationally feasible due to
-the small number of runs required for this analysis.}
-
-\section*{Results}
-
-\subsection*{\del{Kernel classifiers}\add{Classifiers for BA model parameters from tree shapes}}
-
-<<classifiers, include=FALSE>>=
-    get.kernel.data <- function (param, step) {
-        md <- collect.metadata(paste0('../../simulations/kernel-', param, '/', step, '/*'))
-        if ('rbf_variance' %in% colnames(md)) {
-          md <- subset(md, rbf_variance == 4 & decay_factor == 0.3)
-        }
-        if (param == 'alpha') {
-            md <- subset(md, m == 2)
-        }
-        d <- setDT(collect.data(rownames(md)))
-        if (param == 'alpha') {
-            d[,list(rsquared=mean(rsquared), accuracy=mean(accuracy)), by=c('nsimnode', 'ntip')]
-        } else if (param == 'm') {
-            d[,list(rsquared=mean(rsquared)), by=c('nsimnode', 'ntip')]
-        } else if (param == 'I') {
-            d[,list(rsquared=mean(rsquared), accuracy=mean(accuracy)), by=ntip]
-        } else {
-            d[,list(rsquared=mean(rsquared)), by=c('I', 'ntip')]
-        }
-    }
-    kernel.alpha <- get.kernel.data('alpha' ,'classifier')
-    kernel.m <- get.kernel.data('m', 'classifier')
-    kernel.I <- get.kernel.data('I', 'classifier')
-    kernel.N <- get.kernel.data('N', 'classifier')
-
-    sackin.alpha <- get.kernel.data('alpha', 'stats-classifier')
-    sackin.m <- get.kernel.data('m', 'stats-classifier')
-    sackin.I <- get.kernel.data('I', 'stats-classifier')
-    sackin.N <- get.kernel.data('N', 'stats-classifier')
-
-    nltt.alpha <- get.kernel.data('alpha', 'nltt-classifier')
-    nltt.m <- get.kernel.data('m', 'nltt-classifier')
-    nltt.I <- get.kernel.data('I', 'nltt-classifier')
-    nltt.N <- get.kernel.data('N', 'nltt-classifier')
-@
-
-We investigated the \add{identifiability of four} parameters of the BA network
-model~\autocite{barabasi1999emergence}: the number of nodes $N$, the
-preferential attachment power $\alpha$, the number of edges added per vertex
-$m$, and the number of infected nodes $I$. \del{In addition to $m$ and $\alpha$
-(see Introduction), we considered $N$, which denotes the total number of nodes
-in the network, and $I$, which is the number of infected nodes at which to stop
-the simulation and sample the transmission tree.} To examine the effect of
-these parameters on tree shape, we simulated transmission trees under different
-parameter values, calculated pairwise tree kernel scores between them, and
-attempted to classify the trees using a kernel support vector machine (kSVM).
-We also tested classifiers based on Sackin's index~\autocite{shao1990tree} and
-the normalized lineages-through-time (nLTT)
-statistic~\autocite{janzen2015approximate}. \add{We report the accuracy of the
-classifiers, which is simply the proportion of trees which were assigned the
-correct parameter value. Since there were three possible values, random
-guessing would produce an accuracy of 0.33.} \del{The accuracy of each
-classifier varied based on the parameter being tested} \add{The results are
-shown in} \cref{fig:rsquared}. Classifiers based on \del{two other tree
-statistics,} the nLTT and Sackin's index generally exhibited worse performance
-than the tree kernel, although the magnitude of the disparity varied between
-the parameters (\cref{fig:rsquared}, centre and right). \del{The results were
-largely robust to variations in the tree kernel meta-parameters $\lambda$ and
-$\sigma$, although accuracy varied between different epidemic and sampling
-scenarios} \add{Larger datasets were generally classified more accurately
-(\cref{fig:alphacrossv,fig:mcrossv,fig:Ncrossv,fig:Icrossv}), although large
-values of $\lambda$ produced worse estimates on large datasets. Extremely low
-$\sigma$ values, which require nearly-exact matches between branch lengths,
-resulted in low accuracy in some cases (\textit{e.g.} \cref{fig:alphacrossvn},
-center row).}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
-    \caption[
-        Cross-validation accuracy of kernel-SVM, nLTT-based SVM, and Sackin's
-        index regression classifiers for BA model parameters.
-    ]{
-        Cross-validation accuracy of kernel-SVM classifier (left), SVM
-        classifiers using nLTT (centre) and Sackin's index (right) for BA model
-        parameters. Kernel meta-parameters were set to $\lambda = 0.3$ and
-        $\sigma = 4$. Each point was calculated based on 300 simulated
-        transmission trees over networks with three different values of the
-        parameter being tested, assuming perfect knowledge of the other
-        parameters. Vertical lines are empirical 95\% confidence intervals
-        based on 1000 two-fold cross-validations. \add{The classifiers for $I$
-        were not evaluated with 1000-tip trees, because one of the tested $I$
-        values was 500, and it is not possible to sample a tree of size 1000
-        from 500 infected individuals.}
-    }
-    \label{fig:rsquared}
-\end{figure}
-
-The kSVM classifier for $\alpha$ had an average accuracy of 
-    \Sexpr{kernel.alpha[,round(mean(accuracy), 2)]},
-compared to 
-    \Sexpr{nltt.alpha[,round(mean(accuracy), 2)]}
-for the nLTT, and
-    \Sexpr{sackin.alpha[,round(mean(accuracy), 2)]}
-for Sackin's index. \del{There was little variation about the mean for
-different tree and epidemic sizes.} No classifier could accurately identify $m$
-in any epidemic scenario, with average $R^2$ values of 
-  \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for kSVM,
-  \Sexpr{nltt.m[,round(mean(rsquared), 2)]} for the nLTT, and
-  \Sexpr{sackin.m[,round(mean(rsquared), 2)]}
-for Sackin's index. There was little variation in accuracy between epidemic
-scenarios, although the accuracy of the kSVM was slightly higher on 1000-tip
-trees (\cref{fig:rsquared}, left).
-
-The accuracy of classifiers for $I$ varied significantly with the number of
-tips in the tree. For 100-tip trees, the average accuracy was
-  \Sexpr{kernel.I[ntip == 100, round(mean(accuracy), 2)]},
-  \Sexpr{nltt.I[ntip == 100, round(mean(accuracy), 2)]}, and
-  \Sexpr{sackin.I[ntip == 100, round(mean(accuracy), 2)]}
-for the tree kernel, nLTT, and Sackin's index respectively.
-For 500-tip
-trees, the values increased to
-  \Sexpr{kernel.I[ntip == 500, round(mean(accuracy), 2)]},
-  \Sexpr{nltt.I[ntip == 500, round(mean(accuracy), 2)]}, and
-  \Sexpr{sackin.I[ntip == 500, round(mean(accuracy), 2)]}.
-Finally, the performance of classifiers for $N$ depended heavily on the
-epidemic scenario. The $R^2$ of the kSVM classifier ranged from
-  \Sexpr{kernel.N[,round(min(rsquared), 2)]}
-for the smallest epidemic and smallest sample size, to
-  \Sexpr{kernel.N[,round(max(rsquared), 2)]}
-for the largest. Likewise, $R^2$ for the nLTT-based SVM ranged from 
-  \Sexpr{nltt.N[,round(min(rsquared), 2)]}
-to
-  \Sexpr{nltt.N[,round(max(rsquared), 2)]}.
-Sackin's index did not accurately classify $N$ in any scenario, with an average
-$R^2$ of
-  \Sexpr{sackin.N[,round(mean(rsquared), 2)]}
-and little variation between scenarios.
-
-\subsection*{ABC simulations}
-
-<<point_est, include=FALSE>>=
-    f <- Sys.glob("../../simulations/abc-pa-free-m/point-estimate/*")
-    d <- fread(f)
-    d[,m := floor(m)]
-    d[,alpha_error := abs(true_alpha - alpha)]
-    d[,N_error := abs(true_N - N)]
-    d[,I_error := abs(true_I - I)]
-    d[,m_error := abs(true_m - m)]
-
-    alpha.av <- anova(lm(alpha_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-    m.av <- anova(lm(m_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-    N.av <- anova(lm(N_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-    I.av <- anova(lm(I_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-
-    m.tbl <- d[,prop.table(table(m_error))]
-    zero.alpha.test <- wilcox.test(d[true_alpha == 0, alpha_error], 
-                                   d[true_alpha != 0, alpha_error])
-    alpha.I.test <- wilcox.test(d[true_alpha < 1, I_error],
-                                d[true_alpha >= 1, I_error])
-    I.I.test <- wilcox.test(d[true_I == 1000, I_error],
-                            d[true_I == 2000, I_error])
-    alpha.m.test <- wilcox.test(d[true_alpha == 0 | true_alpha == 1, m_error],
-                                d[true_alpha == 0.5 | true_alpha == 1.5, m_error])
-    #stopifnot(alpha.av["factor(true_m)", "Pr(>F)"] <= 0.05)
-    #stopifnot(alpha.av["factor(true_I)", "Pr(>F)"] > 0.05)
-    #stopifnot(I.I.test$p.value < 1e-5)
-    #stopifnot(I.av["factor(true_m)", "Pr(>F)"] > 0.05)
-    #stopifnot(m.av["factor(true_m)", "Pr(>F)"] > 0.05)
-    #stopifnot(m.av["factor(true_I)", "Pr(>F)"] > 0.05)
-    #stopifnot(m.av["factor(true_alpha)", "Pr(>F)"] > 0.05)
-    #stopifnot(min(N.av[1:3,"Pr(>F)"]) > 0.05)
-
-    alpha.m.cor <- cor.test(d[,alpha_error], d[,true_m], method="spearman")
-@
-
-\Cref{fig:abcpt} shows \del{maximum \textit{a posteriori} (MAP)}
-\add{stratified posterior mean} point estimates of the BA model parameters
-\add{$\alpha$ and $I$}, obtained with ABC on simulated data. \add{The
-parameters $m$ and $N$ were not identifiable with ABC for any parameter
-combinations (\cref{fig:abcpt2}).} \del{The estimates shown correspond only to
-the simulations where the $m$ parameter was set to 2, however the results for
-$m = 3$ and $m = 4$ were similar.} Average boundaries of 95\% HPD intervals
-\add{for all parameters} are given in \cref{tab:abchpd}.
-
-\begin{figure}[ht]
-    \centering
+\begin{frame}{$\alpha$ and $I$ are identifiable with ABC}
     \includegraphics[width=\textwidth]{abc-boxplot}
-    \caption[
-        Posterior mean point estimates for BA model parameters $\alpha$ and $I$
-        obtained by running \textit{netabc} on simulated data, stratified by
-        true parameter values.
-    ]{
-        Posterior mean point estimates for BA model parameters $\alpha$ and $I$
-        obtained by running \textit{netabc} on simulated data, stratified by
-        true parameter values. First row of plots contains true versus
-        estimated values of $\alpha$; second row contains true versus estimated
-        values of $I$. Columns are stratified by $\alpha$, $I$, and $m$
-        respectively. Dashed lines indicate true values. 
-    }
-    \label{fig:abcptm2}
-\end{figure}
-
-\begin{table}[ht]
-    \centering
-    \input{\tablepath/abc-hpd.tex}
-    \caption[
-        Average posterior mean point estimates and 95\% highest posterior
-        density interval widths for BA model parameter estimates obtained
-        with \textit{netabc} on simulated data.
-    ]{
-        Average posterior mean point estimates and 95\% highest posterior
-        density interval widths for BA model parameter estimates obtained with
-        \textit{netabc} on simulated data. Three transmission trees were
-        simulated under each combination of the listed parameter values, and
-        the parameters were estimated with ABC without training.
-    }
-    \label{tab:abchpd}
-\end{table}
-
-\del{The accuracy of the parameter estimates obtained with ABC
-paralleled the results from the kSVM classifier. Of the four parameters,
-$\alpha$ was the most accurately estimated, with point estimates having a
-median [IQR] absolute error of 
-    \Sexpr{d[,round(median(alpha_error), 2)]} 
-    [\Sexpr{d[,round(quantile(alpha_error, 0.25), 2)]} - 
-    \Sexpr{d[,round(quantile(alpha_error, 0.75), 2)]}].
-The errors when the true value of $\alpha$ was zero were significantly greater
-than those for the other values 
-(Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(zero.alpha.test[["p.value"]], 4))}$).
-Errors in estimating $\alpha$ also varied with the true value of $m$ just at
-the threshold of statistical significance
-    $p \Sexpr{pp(alpha.av["factor(true_m)", "Pr(>F)"])}$),
-but did not vary across the true values of $N$ or $I$ (both one-way ANOVA).
-Estimates for $I$ were relatively accurate, with point estimate errors of
-    \Sexpr{d[,round(median(I_error))]} 
-    [\Sexpr{d[,round(quantile(I_error, 0.25))]} - 
-    \Sexpr{d[,round(quantile(I_error, 0.75))]}] individuals.
-These errors were significantly higher when the true value of $\alpha$ was
-at least 1
-    (Wilcoxon rank-sum test, $p$ = $\Sexpr{latexSN(round(alpha.I.test[["p.value"]], 4))}$)
-and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
-did not affect the estimates of $I$ (one-way ANOVA).}
-
-\del{The $m$ parameter was estimated correctly in only
-    \Sexpr{as.integer(m.tbl[1] * 100)} \%
-of simulations, barely better than random guessing. The true values of the
-other parameters did not significantly affect the estimates of $m$ (both
-one-way ANOVA). Finally, the total number of nodes $N$ was consistently
-over-estimated by about a factor of two
-    (error \Sexpr{d[,round(median(N_error))]} 
-    [\Sexpr{d[,round(quantile(N_error, 0.25))]} - 
-     \Sexpr{d[,round(quantile(N_error, 0.75))]}] individuals).
-No parameters influenced the accuracy of the $N$ estimates (all one-way ANOVA).}
-
-\add{Across all simulations, the median [IQR] absolute errors of the parameter 
-estimates obtained with \textit{netabc} were
-    \Sexpr{d[,round(median(alpha_error), 2)]} 
-    [\Sexpr{d[,round(quantile(alpha_error, 0.25), 2)]} - 
-    \Sexpr{d[,round(quantile(alpha_error, 0.75), 2)]}]
-for $\alpha$,
-    \Sexpr{d[,round(median(I_error))]} 
-    [\Sexpr{d[,round(quantile(I_error, 0.25))]} - 
-    \Sexpr{d[,round(quantile(I_error, 0.75))]}]
-for $I$,
-    \Sexpr{d[,round(median(m_error), 2)]} 
-    [\Sexpr{d[,round(quantile(m_error, 0.25), 2)]} - 
-    \Sexpr{d[,round(quantile(m_error, 0.75), 2)]}]
-for $m$, and
-    \Sexpr{d[,round(median(N_error))]} 
-    [\Sexpr{d[,round(quantile(N_error, 0.25))]} - 
-    \Sexpr{d[,round(quantile(N_error, 0.75))]}]
-for $N$. These errors comprised, respectively,
-    \Sexpr{d[,round(median(alpha_error) * 100 / 2)]}\%,
-    \Sexpr{d[,round(median(I_error) * 100 / 4500)]}\%,
-    \Sexpr{d[,round(median(m_error) * 100 / 6)]}\%, and
-    \Sexpr{d[,round(median(N_error) * 100 / 14500)]}\%
-of the regions of nonzero prior density. For $I$ and $N$, relative errors were
-    \Sexpr{d[,round(median(I_error * 100 / true_I))]}\%
-    [\Sexpr{d[,round(quantile(I_error * 100 / true_I, 0.25))]} - 
-    \Sexpr{d[,round(quantile(I_error * 100 / true_I, 0.75))]}\%]
-and
-    \Sexpr{d[,round(median(N_error * 100 / true_N))]}\%
-    [\Sexpr{d[,round(quantile(N_error * 100 / true_N, 0.25))]} - 
-    \Sexpr{d[,round(quantile(N_error * 100 / true_N, 0.75))]}\%].
-Average 95\% HPD interval widths were
-    \Sexpr{d[,round(mean(alpha.upper.95 - alpha.lower.95), 2)]},
-    \Sexpr{d[,round(mean(I.upper.95 - I.lower.95))]},
-    \Sexpr{d[,round(mean(m.upper.95 - m.lower.95), 2)]}, and
-    \Sexpr{d[,format(round(mean(N.upper.95 - N.lower.95)), digits=5)]},
-representing 
-    \Sexpr{d[,round(mean(alpha.upper.95 - alpha.lower.95) * 100 / 2)]}\%,
-    \Sexpr{d[,round(mean(I.upper.95 - I.lower.95) * 100 / 4500)]}\%,
-    \Sexpr{d[,round(mean(m.upper.95 - m.lower.95) * 100 / 6)]}\%, and
-    \Sexpr{d[,round(mean(N.upper.95 - N.lower.95) * 100 / 14500)]}\%
-of the nonzero prior density regions. Point estimates of $I$ were upwardly
-biased: $I$ was overestimated in
-    \Sexpr{d[,sum(I > true_I)]}
-out of 
-    \Sexpr{nrow(d)}
-simulations
-    (\Sexpr{round(d[,sum(I > true_I)] * 100 / nrow(d))}\%).
-The estimates for $m$ and $N$ were similar across all simulations (median [IQR]
-point estimates
-    \Sexpr{d[,round(median(m), 2)]}
-    [\Sexpr{d[,round(quantile(m, 0.25), 2)]} - 
-    \Sexpr{d[,round(quantile(m, 0.75), 2)]}]
-and 
-    \Sexpr{d[,round(median(N))]}
-    [\Sexpr{d[,round(quantile(N, 0.25))]} - 
-    \Sexpr{d[,round(quantile(N, 0.75))]}])
-regardless of the true values of any of the BA parameters (\cref{fig:abcpt2}).}
-
-\add{To analyze the effects of the true parameter values on the
-accuracy our estimates of $\alpha$ and $I$, we fitted one GLM for each of these
-two parameters, with error rate as the dependent variable and the true
-parameter values as independent variables. Since the estimates of $m$ and
-$N$ were roughly equal across all simulations (\cref{fig:abcpt2}), GLMs were
-not fitted for these parameters. The estimated coefficients are shown in
-\cref{tab:glm}.}
-
-<<abc_glm, include=FALSE>>=
-    source("../thesis/global.R")
-    stopifnot(alpha.glm[Parameter == "alpha",p] < 0.05)
-    stopifnot(alpha.glm[Parameter == "I",p] > 0.05)
-    stopifnot(alpha.glm[Parameter == "m",p] < 0.05)
-
-    stopifnot(I.glm[Parameter == "alpha",p] < 0.05)
-    stopifnot(I.glm[Parameter == "I",p] < 0.05)
-    stopifnot(I.glm[Parameter == "m",p] > 0.05)
-@
-
-<<glm_table, echo=FALSE, warning=FALSE, error=FALSE, results="asis">>=
-    format.glm <- function (df) {
-        df.copy <- copy(df)
-        df.copy[Parameter == "alpha", Parameter := "\\alpha"]
-        df.copy[Parameter != "(Intercept)", Parameter := paste0("$", Parameter, "$")]
-        df.copy[,p := paste0("$", sapply(p, pp, eq=FALSE), "$")]
-        df.copy[,Estimate := paste0("$", sapply(Estimate, pp, eq=FALSE, thres=NA), "$")]
-        df.copy[,`Standard error` := paste0("$", sapply(`Standard error`, pp, eq=FALSE, thres=NA), "$")]
-        setnames(df.copy, "p", "$p$-value")
-        df.copy
-    }
-
-    glm.tbl <- rbind(format.glm(alpha.glm), format.glm(I.glm))
-    glm.tbl <- cbind(`Dependent variable`=c("$\\alpha$", "", "", "", "$I$", "", "", ""),
-                     glm.tbl)
-    setnames(glm.tbl, "Parameter", "Independent variable")
-
-    capt.short <- paste("Parameters of fitted GLMs relating error in estimated",
-                        "$\\alpha$ and $I$ to true values of BA parameters.")
-    capt <- paste(capt.short, "GLMs ere fitted with the Gaussian distribution and",
-                  "inverse link function. Coefficients are interpretable as",
-                  "additive effects on the inverse of the mean error.")
-    print(xtable(glm.tbl, label="tab:glm", caption=c(capt, capt.short)),
-          include.rownames=FALSE, sanitize.text.function=identity,
-          sanitize.colnames.function=identity)
-@
-
-\add{The GLM analysis indicated that the error in estimates of $\alpha$
-decreased with larger true values of $\alpha$
-    ($p \Sexpr{pp(alpha.glm[Parameter == "alpha", p], eq=TRUE)}$)
-and $m$
-    ($p \Sexpr{pp(alpha.glm[Parameter == "m", p], eq=TRUE)}$)
-but was not significantly affected by $I$. Qualitatively, $\alpha$ seemed to be
-only weakly identifiable between the values of 0 and 0.5 (\cref{fig:abcptm2}).
-The error in the estimated $I$ value was slightly lower for smaller values of
-$\alpha$
-    ($p \Sexpr{pp(I.glm[Parameter == "alpha", p], eq=TRUE)}$)
-and $I$
-    ($p \Sexpr{pp(I.glm[Parameter == "I", p], eq=TRUE)}$),
-but was not significantly affected by the true value of $m$.}
-
-<<marginals, include=FALSE>>=
-    q <- 0.42
-    middle <- d[alpha_error > quantile(alpha_error, q) & 
-                alpha_error < quantile(alpha_error, 1-q) & 
-                I_error > quantile(I_error, q) & 
-                I_error < quantile(I_error, 1-q),]
-    middle.plot <- middle[1,sprintf("%.1f_%.0f_%.0f_%.0f_%.0f", true_alpha, true_I, true_m, true_N, replicate)]
-@
-
-The dispersion of the ABC approximation to the posterior also varied between
-the parameters\del{, with narrower HPD intervals for the parameters with more
-accurate point estimates} (\cref{tab:abchpd}). HPD intervals around $\alpha$
-and $I$ were often narrow relative to the region of nonzero prior
-density, whereas the intervals for $m$ and $N$ were more widely dispersed.
-\Cref{fig:abc1,fig:abc2} \del{shows the distributions for one
-simulation.} \add{show one- and two-dimensional marginal distributions for a
-simulation with $\alpha$ and $I$ errors close to their respective medians. 
-%In particular, the simulation shown is one of
-%\Sexpr{replace_number(nrow(middle))} with errors in both $\alpha$ and $I$
-%between the \Sexpr{round(q*100)}nd and \Sexpr{round((1-q)*100)}th percentiles.
-The parameters for this simulation were
-    $\alpha = \Sexpr{middle[1,true_alpha]}$,
-    $I = \Sexpr{middle[1,true_I]}$,
-    $m = \Sexpr{middle[1,true_m]}$, and
-    $N = \Sexpr{middle[1,true_N]}$.
-The two-dimensional marginals indicate some dependence between pairs of
-parameters, particularly $I$ and $N$ which show a diagonally shaped region of
-high posterior density.}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior", middle.plot)}}.pdf}
-    \caption[
-        One-dimensional marginal posterior distributions of BA model
-        parameters estimated by \textit{netabc} from a simulated transmission
-        tree.
-    ]{
-        One-dimensional marginal posterior distributions of BA model parameters
-        estimated by \textit{netabc} from a simulated transmission tree. Dashed
-        lines indicate true values, solid lines indicate posterior means, and
-        shaded areas show 95\% highest posterior density intervals.
-    }
-    \label{fig:abc1}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    %\includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior-2d", middle.plot)}}.pdf}
-    \caption[
-        Two-dimensional marginal posterior distributions of BA model parameters
-        estimated by \textit{netabc} from a simulated transmission tree.
-    ]{
-        Two-dimensional marginal posterior distributions of BA model parameters
-        estimated by \textit{netabc} from a simulated transmission tree. White
-        circles indicate true values, magenta diamonds
-        indicate posterior means. 
-    }
-    \label{fig:abc2}
-\end{figure}
-
-<<mixed_peerdriven, include=FALSE>>=
-  library(coda)
-  d <- fread("bzcat ../../simulations/abc-pa-mixed-alpha/abc/*")
-  d <- d[iter == max(iter)]
-  params <- c("alpha", "m", "N", "I")
-  d <- melt(d, measure.vars=params, variable.name="parameter")
-  f <- function (x, wt) {
-    hpd <- wtd.hpd(x, wt/sum(wt))
-    list(pmean=wtd.mean(x, wt, normwt=TRUE), 
-         hpd.min=hpd[1], 
-         hpd.max=hpd[2])
-  }
-  d <- d[,f(value, weight), by=parameter]
-  setkey(d, parameter)
-
-  m <- collect.metadata("../../simulations/abc-pa-peerdriven/abc/*")
-  p <- fread(paste("bzcat", rownames(m)[m$sample_peer == 1]))
-  p <- p[iter == max(iter)]
-  params <- c("alpha", "m", "N", "I")
-  p <- melt(p, measure.vars=params, variable.name="parameter")
-  p <- p[,f(value, weight), by=parameter]
-  setkey(p, parameter)
-  options(scipen=5)
-@
-
-To test the effect of model misspecification, we simulated one network where
-the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
-other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The posterior mean
-[95\% HPD] estimates for each parameter were: 
-$\alpha$, 
-  \Sexpr{round(d["alpha", pmean], 2)} 
-  [\Sexpr{round(d["alpha", hpd.min], 2)} -
-   \Sexpr{round(d["alpha", hpd.max], 2)}];
-$I$,
-  \Sexpr{d["I", round(pmean)]} 
-  [\Sexpr{d["I", round(hpd.min)]} -
-   \Sexpr{d["I", round(hpd.max)]}];
-$m$,
-  \Sexpr{d["m", floor(pmean)]} 
-  [\Sexpr{d["m", floor(hpd.min)]} -
-   \Sexpr{d["m", floor(hpd.max)]}];
-$N$,
-  \Sexpr{d["N", round(pmean)]} 
-  [\Sexpr{d["N", round(hpd.min)]}-
-   \Sexpr{d["N", round(hpd.max)]}].
-The approximate one-dimensional marginal posterior distributions for this
-simulation are shown in \cref{fig:mixed}. To test the effect of sampling bias,
-we sampled one transmission tree in a peer-driven fashion, where the
-probability to sample a node was twice as high if one of its peers had already
-been sampled. The parameters for this experiment were $N$ = 5000, $m$ = 2,
-$\alpha$ = 0.5, and $I$ = 2000. The estimated values were $\alpha$, 
-  \Sexpr{p["alpha", round(pmean, 2)]} 
-  [\Sexpr{p["alpha", round(hpd.min, 2)]} -
-   \Sexpr{p["alpha", round(hpd.max, 2)]}];
-$I$,
-  \Sexpr{p["I", floor(pmean)]} 
-  [\Sexpr{p["I", floor(hpd.min)]} -
-   \Sexpr{p["I", floor(hpd.max)]}];
-$m$,
-  \Sexpr{p["m", floor(pmean)]} 
-  [\Sexpr{p["m", floor(hpd.min)]} -
-   \Sexpr{p["m", floor(hpd.max)]}];
-$N$,
-  \Sexpr{p["N", floor(pmean)]} 
-  [\Sexpr{p["N", floor(hpd.min)]} -
-   \Sexpr{p["N", floor(hpd.max)]}].
-The approximate one-dimensional marginal posterior distributions are shown in
-\cref{fig:peerdriven}. Both of these results were in line with estimates
-obtained on other simulated datasets (\cref{tab:abchpd}), although the estimate
-of peer-driven sampling for $\alpha$ was somewhat lower than typical.
-
-<<marginal, include=FALSE>>=
-    d <- setDT(collect.data("../../simulations/abc-marginal/point-estimate/*", sep="\t"))
-
-    joint.alpha <- d[is.na(fixed), abs(alpha-true_alpha)]
-    marginal.alpha <- d[fixed == "m+N+I", abs(alpha-true_alpha)]
-
-    joint.N <- d[is.na(fixed), abs(N-true_N)]
-    marginal.N <- d[fixed == "I+alpha+m", abs(N-true_N)]
-
-    joint.I <- d[is.na(fixed), abs(I-true_I)]
-    marginal.I <- d[fixed == "N+alpha+m", abs(I-true_I)]
-@
-
-\add{\Cref{fig:marginal} shows the effect of performing marginal ABC estimation
-of each of the BA parameters on the same simulated transmission tree. The
-estimates of $m$ are apparently unaffected by marginalizing out the other
-parameters, corroborating the previous experiments' findings that $m$ is not an
-identifiable parameter from scaled tree shapes. Compared to allowing all
-parameters to vary, estimates of $\alpha$, $I$, and $N$ were improved by
-    \Sexpr{ round((joint.alpha - marginal.alpha) * 100 / joint.alpha) }\%,
-    \Sexpr{ round((joint.I - marginal.I) * 100 / joint.I) }\%, and
-    \Sexpr{ round((joint.N - marginal.N) * 100 / joint.N) }\%
-when all other parameters were fixed. Neither the accuracy nor the precision of
-any parameter estimates appeared to improve when more computing power was used
-for SMC (\cref{fig:moresmc}).}
-
-\subsection*{Real data}
-
-<<realdata, include=FALSE>>=
-  m <- collect.metadata("../../simulations/aggregates/hpd/*")
-  d1 <- fread(rownames(m)[m$m_min == 1 & m$agg_gene == "gag"])
-  d2 <- fread(rownames(m)[m$m_min == 2 & m$agg_gene == "gag"])
-
-  d1.env <- fread(rownames(m)[m$m_min == 1 & m$agg_gene == "env"])
-  d2.env <- fread(rownames(m)[m$m_min == 2 & m$agg_gene == "env"])
-
-  pmean1 <- dcast.data.table(d1, dataset~parameter, value.var="mean")
-  up1 <- dcast.data.table(d1, dataset~parameter, value.var="hpd.max.95")
-  low1 <- dcast.data.table(d1, dataset~parameter, value.var="hpd.min.95")
-  setkey(pmean1, dataset)
-  setkey(up1, dataset)
-  setkey(low1, dataset)
-
-  pmean2 <- dcast.data.table(d2, dataset~parameter, value.var="mean")
-  up2 <- dcast.data.table(d2, dataset~parameter, value.var="hpd.max.95")
-  low2 <- dcast.data.table(d2, dataset~parameter, value.var="hpd.min.95")
-  setkey(pmean2, dataset)
-  setkey(up2, dataset)
-  setkey(low2, dataset)
-
-  pmean1.env <- dcast.data.table(d1.env, dataset~parameter, value.var="mean")
-  up1.env <- dcast.data.table(d1.env, dataset~parameter, value.var="hpd.max.95")
-  low1.env <- dcast.data.table(d1.env, dataset~parameter, value.var="hpd.min.95")
-  setkey(pmean1.env, dataset)
-  setkey(up1.env, dataset)
-  setkey(low1.env, dataset)
-
-  pmean2.env <- dcast.data.table(d2.env, dataset~parameter, value.var="mean")
-  up2.env <- dcast.data.table(d2.env, dataset~parameter, value.var="hpd.max.95")
-  low2.env <- dcast.data.table(d2.env, dataset~parameter, value.var="hpd.min.95")
-  setkey(pmean2.env, dataset)
-  setkey(up2.env, dataset)
-  setkey(low2.env, dataset)
-
-  setkey(d1, dataset)
-  setkey(d2, dataset)
-  setkey(d1.env, dataset)
-  setkey(d2.env, dataset)
-
-  pmean1 <- unique(d1)[,.(dataset, ntip)][pmean1]
-  pmean2 <- unique(d2)[,.(dataset, ntip)][pmean2]
-  pmean1.env <- unique(d1.env)[,.(dataset, ntip)][pmean1.env]
-  pmean2.env <- unique(d2.env)[,.(dataset, ntip)][pmean2.env]
-
-  stopifnot(pmean1[,dataset[which.max(alpha)]] == "zetterberg2004")
-  stopifnot(pmean1[,dataset[which.min(alpha)]] == "cuevas2009")
-
-  stopifnot(pmean1[,dataset[which.min(I)]] == "zetterberg2004")
-  stopifnot(pmean1[,dataset[which.max(I)]] == "novitsky2014")
-  stopifnot(pmean1[dataset != "novitsky2014", dataset[which.max(I)]] == "bctree")
-
-  stopifnot(pmean1[,cor.test(ntip, I, method="spearman")$p.value] > 0.05)
-@
-
-\del{We applied ABC to five published HIV datasets (\cref{tab:data}),
-and found substantial heterogeneity among the parameter estimates
-(\cref{fig:abchpd,fig:abchpdm2}).} \add{Posterior mean point estimates and 50\%
-and 95\% HPD intervals for each parameter are shown in \cref{fig:abchpd}.}
-\del{Plots of the marginal posterior distributions for each dataset are shown
-in \cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.}
-\add{\Cref{fig:abchpdm2} shows point estimates and HPD intervals obtained when
-the value $m = 1$ was disallowed by the prior. Since the results indicated that
-$m = 1$ was the most credible value for several datasets, all results discussed
-henceforth apply to the prior $m \sim \text{DiscreteUniform}(1, 5)$ unless
-otherwise stated.}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics{realdata-hpd}
-    \caption[
-        Posterior means and 50\%/95\% HPD intervals for parameters of the
-        BA network model, fitted to ten HIV datasets with \textit{netabc}.
-    ]{
-        Posterior means (points), 50\% HPD intervals (notches), and 95\%
-        HPD intervals (lines) for parameters of the BA network model, fitted to
-        ten HIV datasets with \textit{netabc}. Legend labels indicate risk
-        group and country of origin. Abbreviations: IDU, injection drug users;
-        MSM, men who have sex with men; HET, heterosexual. Note that posterior
-        means can fall outside of the HPD interval if the distribution is
-        diffuse.
-    }
-    \label{fig:abchpd}
-\end{figure}
-
-\del{Two of the datasets~\autocite{niculescu2015recent, wang2015targeting} had
-estimated $\alpha$ values near unity for the prior allowing $m = 1$ (MAP
-estimates [95\% HPD] 
-  \Sexpr{pmean1["niculescu2015", round(alpha, 2)]} 
-  [\Sexpr{low1["niculescu2015", round(alpha, 2)]} - 
-   \Sexpr{up1["niculescu2015", round(alpha, 2)]}]
-and
-  \Sexpr{pmean1["wang2015", round(alpha, 2)]} 
-  [\Sexpr{low1["wang2015", round(alpha, 2)]} -
-   \Sexpr{up1["wang2015", round(alpha, 2)]}] respectively).
-The MAP estimates did not change appreciably when $m = 1$ was disallowed by the
-prior, although the credible interval of the \textcite{niculescu2015recent}
-data was narrower
-  (\Sexpr{low1["niculescu2015", round(alpha, 2)]} - 
-   \Sexpr{up1["niculescu2015", round(alpha, 2)]}).
-When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
-low estimated $\alpha$ values
-  (\Sexpr{pmean1["li2015", round(alpha, 2)]} 
-  [\Sexpr{low1["li2015", round(alpha, 2)]} - 
-  \Sexpr{up1["li2015", round(alpha, 2)]}]
-and
-  \Sexpr{pmean1["cuevas2009", round(alpha, 2)]} 
-  [\Sexpr{low1["cuevas2009", round(alpha, 2)]} -
-   \Sexpr{up1["cuevas2009", round(alpha, 2)]}]). 
-However, the MAP estimates increased when $m = 1$ was not permitted, although
-the HPD intervals remained roughly the same
-  (\Sexpr{pmean2["li2015", round(alpha, 2)]} 
-  [\Sexpr{low2["li2015", round(alpha, 2)]} - 
-  \Sexpr{up2["li2015", round(alpha, 2)]}]
-and
-  \Sexpr{pmean2["cuevas2009", round(alpha, 2)]} 
-  [\Sexpr{low2["cuevas2009", round(alpha, 2)]} -
-   \Sexpr{up2["cuevas2009", round(alpha, 2)]}]).
-The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
-for both priors on $m$
-  (\Sexpr{pmean1["novitsky2014", round(alpha, 2)]} for $m \geq 1$;
-   \Sexpr{pmean2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).
-However, the confidence interval was much wider when $m = 1$ was allowed
-  ([\Sexpr{low1["novitsky2014", round(alpha, 2)]} -
-    \Sexpr{up1["novitsky2014", round(alpha, 2)]}] for $m \geq 1$ vs.
-    \Sexpr{low2["novitsky2014", round(alpha, 2)]} -
-    \Sexpr{up2["novitsky2014", round(alpha, 2)]} for $m \geq 2$).}
-
-\add{Posterior mean point estimates for the preferential attachment power $\alpha$
-were all sub-linear, ranging from 
-    \Sexpr{pmean1["zetterberg2004", round(alpha, 2)]}
-for the IDU/Estonia data to
-    \Sexpr{pmean1["cuevas2009", round(alpha, 2)]}
-for the mixed/Spain data. When aggregated by risk group, the average estimates
-were
-    \Sexpr{pmean1[c("zetterberg2004", "niculescu2015", "bctree"), round(mean(alpha), 2)]}
-for IDU,
-    \Sexpr{pmean1[c("novitsky2014", "mccormack2002", "grabowski2014"), round(mean(alpha), 2)]}
-for primarily heterosexual risk, and
-    \Sexpr{pmean1[c("wang2015", "kao2011", "little2014", "li2015"), round(mean(alpha), 2)]}
-for MSM. These values were obtained with \textit{gag} for the datasets where
-both \textit{gag} and \textit{env} were sequenced, but the estimates for
-$\alpha$ did not change appreciably between the two genes (\cref{fig:genes}).
-There was a large amount of uncertainty associated with these estimates. 95\%
-HPD intervals were very wide for most datasets, often encompassing almost the
-entire range from 0 to 1 (\cref{fig:abchpd}).}
-
-\del{For all the datasets except \citeauthor{novitsky2014impact}, estimated values
-of $I$ were below 2000 when $m = 1$ was allowed, with relatively narrow HPD
-intervals compared to the nonzero prior density region
-  (\citeauthor{cuevas2009hiv}, \Sexpr{pmean1["cuevas2009", floor(I)]} 
-  [\Sexpr{low1["cuevas2009", floor(I)]} -
-   \Sexpr{up1["cuevas2009", floor(I)]}];
-   \citeauthor{niculescu2015recent}, \Sexpr{pmean1["niculescu2015", floor(I)]}
-  [\Sexpr{low1["niculescu2015", floor(I)]} - 
-   \Sexpr{up1["niculescu2015", floor(I)]}];
-  \citeauthor{li2015hiv}, \Sexpr{pmean1["li2015", floor(I)]} 
-  [\Sexpr{low1["li2015", floor(I)]} -
-   \Sexpr{up1["li2015", floor(I)]}];
-   \citeauthor{wang2015targeting}, \Sexpr{pmean1["wang2015", floor(I)]}
-  [\Sexpr{low1["wang2015", floor(I)]} - 
-   \Sexpr{up1["wang2015", floor(I)]}]).
-The \citeauthor{novitsky2014impact} data was the outlier, with a very high
-estimated $I$, and HPD interval spanning almost the entire prior region
-  (\Sexpr{pmean1["novitsky2014", floor(I)]} 
-  [\Sexpr{low1["novitsky2014", floor(I)]} -
-   \Sexpr{up1["novitsky2014", floor(I)]}]).
-The $I$ estimates and HPD intervals were generally robust to the choice of
-prior on $m$, with slightly narrower HPD intervals (compare
-\cref{fig:abchpd,fig:abchpdm2}).}
-
-<<genetrees, include=FALSE>>=
-    datasets <- c("grabowski2014", "zetterberg2004", "mccormack2002")
-    paths <- paste0("../../simulations/", datasets, "/subtree/*")
-    paths <- sapply(paths, Sys.glob)
-    trees <- lapply(paths, read.tree)
-    get.ratio <- function (t) {
-        int <- t$edge.length[t$edge[,2] > Ntip(t)]
-        tip <- t$edge.length[t$edge[,2] <= Ntip(t)]
-        mean(int) / mean(tip)
-    }
-    ratios <- data.table(dataset=rep(datasets, each=2),
-                         gene=rep(c("env", "gag"), 3),
-                         ratio=sapply(trees, get.ratio))
-    stopifnot(ratios[gene == "env", mean(ratio)] <
-              ratios[gene == "gag", mean(ratio)])
-@
-
-\add{For all but the HET/Botswana data, the posterior mean estimates for
-$I$ were between
-    \Sexpr{pmean1["zetterberg2004", round(I)]}
-(IDU/Estonia) and
-    \Sexpr{pmean1["bctree", round(I)]}
-(IDU/Canada). The HET/Botswana data had a much higher estimated $I$ value
-    (\Sexpr{pmean1["novitsky2014", round(I)]})
-than the other datasets, with a very wide 95\% HPD interval covering almost the
-entire prior region (\cref{fig:abchpd}). There was no significant correlation
-between the number of sequences in the tree and the estimated $I$ value
-(Spearman correlation, 
-    $p \Sexpr{pp(pmean1[,cor.test(ntip, I, method="spearman")[["p.value"]]], eq=TRUE)}$),
-indicating that the higher estimates were not simply due to increased sampling
-density. When both \textit{gag} and \textit{env} sequences were analyzed, the
-estimates from the \textit{env} data were substantially higher
-    (average \Sexpr{pmean1[datasets, round(mean(I))]} for \textit{gag} vs.
-     \Sexpr{pmean1.env[datasets, round(mean(I))]} for \textit{env},
-\cref{fig:genes}). The ratios of internal to terminal branch lengths were
-lower for the \textit{env} trees than the \textit{gag} trees (average 
-    \Sexpr{ratios[gene == "gag", round(mean(ratio), 2)]}
-for \textit{gag} and 
-    \Sexpr{ratios[gene == "env", round(mean(ratio), 2)]}
-for \textit{env}).}
-
-\del{The MAP estimate of $m$ was equal to 1 for all but the
-\citeauthor{novitsky2014impact} data, when this value was allowed. However, the
-upper bound of the HPD interval was different for each dataset
-  (\citeauthor{niculescu2015recent}, \Sexpr{up1["niculescu2015", round(m)]};
-   \citeauthor{wang2015targeting}, \Sexpr{up1["wang2015", round(m)]};
-   \citeauthor{li2015hiv}, \Sexpr{up1["li2015", round(m)]};
-   \citeauthor{cuevas2009hiv}, \Sexpr{up1["cuevas2009", round(m)]}).
-When $m = 1$ was disallowed, the MAP for all datasets was either 2 or 3, with
-HPD intervals spanning the entire prior region. The estimates for the total
-number of nodes $N$ were largely uninformative for all samples, with almost all
-MAP estimates greater than 7500 and HPD intervals spanning almost the entire
-nonzero prior density region. The only exception was the \citeauthor{li2015hiv}
-data, for which the MAP estimate was lower 
-  (\Sexpr{pmean1["li2015", floor(N)]})
-when $m = 1$ was allowed.}
-
-\add{The posterior means of $m$ were equal to one for 
-  \Sexpr{replace_number(pmean1[,sum(m == 1)])}
-of the datasets analyzed. The widths of the 95\% HPD intervals varied from 0
-(all the mass on the estimated value) to 5 (the entire prior region). Estimates
-of $N$ were  mostly uninformative, with very similar estimates for all datasets
-(mean
-    \Sexpr{pmean1[,round(mean(N))]},
-range
-    \Sexpr{pmean1[,round(min(N))]} - \Sexpr{pmean1[,round(max(N))]}).
-This was similar to the pattern observed for the synthetic data, where the
-posterior mean always fell around the upper two-thirds mark of the range
-(\cref{fig:abcpt2}).}
-
-\add{When the value $m = 1$ was disallowed by the prior, the separation in
-$\alpha$ between the IDU datasets and the others became more striking
-(\cref{fig:abchpdm2}). All three IDU datasets had estimated $\alpha$ values at
-or above
-    \Sexpr{pmean2[c("zetterberg2004", "niculescu2015", "bctree"), round(min(alpha), 2)]}.
-The estimate for the MSM/Beijing data was slightly lower
-    (\Sexpr{pmean2["wang2015", round(alpha, 2)]}) and
-the estimates for the seven remaining non-IDU datasets were bounded above by
-    \Sexpr{pmean2[c("novitsky2014", "mccormack2002", "grabowski2014", "kao2011", "little2014", "li2015", "cuevas2009"), round(max(alpha), 2)]}. 
-The values of $I$ were fairly robust to the choice of prior (compare
-\cref{fig:abchpd,fig:abchpdm2}), although the 95\% HPD intervals were slightly
-narrower (average width
-    \Sexpr{round(mean(up1[,I] - low1[,I]))}
-for $m \geq 1$ and
-    \Sexpr{round(mean(up2[,I] - low2[,I]))}
-for $m \geq 2$). When $m = 1$ was disallowed, the posterior means of $m$ for
-all but the HET/Botswana data took on the value 3, with the HPD intervals
-spanning the entire prior region. This is very similar to the results observed
-for $m$ on simulated data (\cref{tab:abchpd}), and suggests that $m$ is not
-identifiable from these data with this prior. On the other hand, the estimate
-and 95\% HPD interval for the HET/Botswana data (4 [2-5]) was the same for both
-priors. The results for $N$ did not change appreciably between the two choices
-of prior.}
-
-\section*{Discussion}
-
-Contact networks can have a strong influence on epidemic progression, and are
-potentially useful as a public health tool~\autocite{wang2015targeting,
-little2014using}. Despite this, few methods exist for investigating contact
-network parameters in a phylodynamic framework \autocite[although see][for
-related work]{groendyke2011bayesian, volz2008sir, brown2011transmission,
-leventhal2012inferring}. ABC is a model-agnostic method which can be used to
-investigate any quantity that affects tree
-shape~\autocite{poon2015phylodynamic}. In this work, we developed a ABC-based
-method to infer the parameters of a contact network model. The method is
-general, and could be applied to any model from which contact networks can be
-simulated. We demonstrated the method on the BA model, which is a simple
-preferential attachment model. \add{For some parameter choices,} the BA model
-gives rise to the power law degree distributions commonly observed in
-real-world networks.
-
-\subsection*{Analysis of BA model with synthetic data}
-
-\del{By training a kernel-SVM classifier, we found that the $\alpha$ and $I$
-parameters, representing preferential attachment power and number of infected
-nodes, had a strong influence on tree shape. This was reflected in the relative
-accuracy of the ABC estimates of these parameters. The total number of
-nodes $N$ had a weak influence on tree shape, which was most prominent when the
-epidemic size $I$ and number of sampled tips were both large. The $m$
-parameter, representing the number of edges created in the network per vertex,
-did not produce much variation in tree shape, resulting in in both poorly
-performing classifiers and uninformative ABC estimates.}
-
-\add{The preferential attachment power $\alpha$ had a strong influence on tree
-shape in the range of values we considered. Although the tree kernel was the
-most effective classifier for $alpha$, a Sackin's index of tree imbalance
-performed nearly as well (\cref{fig:rsquared}). High $\alpha$ values produce
-networks with few well-connected ``superspreader'' nodes which are involved in
-a large number of transmissions, resulting in a highly unbalanced ladder-like
-tree structure. There appeared to be weaker identifiability for $\alpha < 1$
-than for $\alpha \geq 1$ (\cref{fig:abcpt,tab:abchpd}), which may be partially
-explained by the relationship between $\alpha$ and the power law exponent
-$\gamma$ (\cref{fig:gamma}). Although the degree distributions do not truly
-follow a power law for $\alpha \neq 1$, the fitted exponent still captures the
-shape of the degree distribution reasonably well (\cref{fig:powerlaw}). The
-$\gamma$ values fitted to $\alpha = 0$ and $\alpha = 0.5$ are nearly identical
-(about 2.28 for $\alpha = 0$ and 2.33 for $\alpha = 0.5$ with $N = 5000$ and $m
-= 2$). In other words, the degree distributions of networks with $\alpha < 1$
-are similar to each other, which may result in similarity of corresponding
-transmission trees as well.}
-
-\add{$I$, representing the number of infected individuals at the time of
-sampling, was also identifiable, albeit slightly over-estimated with ABC for
-both values we considered. Sackin's index was better able to discern $I$ from
-tree shape than the nLTT (\cref{fig:rsquared,fig:Icrossv}), suggesting that
-this parameter impacts the distribution of branching times in the tree more
-than the topology. In a homogeneously-mixed population, branching times can be
-modelled by the coalescent process~\autocite{kingman1982coalescent}, in our
-case under the SI model~\autocite{volz2009phylodynamics}. Although networks are
-not homogeneously mixed, the forces which affect the distributions of branching
-times still apply. In our simulations, all discordant edges shared the same
-transmission rate, so that the waiting time until the next transmission in the
-entire network was always inversely proportional to the number of discordant
-edges. In the initial phase of the epidemic, when $I$ is small, each new
-transmission results in many new discordant edges. Hence, there is an early
-exponential growth phase, producing many short branches near the root of the
-tree. As the epidemic gets closer to saturating the network, the number of
-discordant edges decays, causing longer waiting times.}
-
-% OLD STUFF
- Since the
-prior on $N$ and $I$ is jointly uniform on a non-rectangular region ($I \leq
-N$), there is more prior mass on high $N$ values.
-
-This can be
-illustrated by imagining that we add a small number of nodes to a network after
-the epidemic simulation has already been completed. It is possible that none of
-these new nodes attains a connection to any infected node. Thus, running the
-simulation again on the new, larger network could produce the exact same
-transmission tree as before. We note also that \del{our accurate estimates of
-$I$ may have been influenced by} this prior\del{, which} places more mass on
-low $I$ values. However, the estimate of $I$ was very high for the HET/Botswana
-data, suggesting that a strong enough signal in the data can
-overcome the prior. \add{Furthermore, when $I$ was estimated marginally with
-fixed $N$, the accuracy of the estimate improved even though there was no
-longer any extra prior mass on low $I$ values.}
-
-% END OLD STUFF
-
-\add{The number of nodes in the network, $N$, exhibited the most variation in
-terms of its effect on tree shape. There was almost no measurable difference
-between trees simulated under different $N$ values when the number of infected
-nodes $I$ was small (\cref{fig:Ncrossv}).} $N$ was always significantly
-over-estimated using ABC (\cref{fig:abcpt2}).
-
-In retrospect, it is unreasonable to expect good estimation of $N$, because
-adding additional nodes does not change the edge density or overall shape of a
-BA network. This can be illustrated by imagining that we add a small number of
-nodes to a network after the epidemic simulation has already been completed. It
-is possible that none of these new nodes attains a connection to any infected
-node. Thus, running the simulation again on the
-new, larger network could produce the exact same transmission tree as before.
-On the other hand, when $I$ is large relative to $N$, the coalescent dynamics
-discussed above also apply. The waiting times until the next infection
-increase, resulting in longer coalescence times toward the tips. The relative
-accuracy of the nLTT in these situations (\cref{fig:rsquared,fig:Ncrossv})
-corroborates this hypothesis, as the nLTT uses only information about the
-coalescence times. When all BA parameters were simultaneously estimated with
-ABC, $N$ was nearly always over-estimated by approximately a factor of two
-(\cref{fig:abcptm2,tab:abchpd}). One factor which may have contributed to this
-bias was our choice of prior distribution. Since the prior for $I$ and $N$ was
-jointly uniform on a region where $I \leq N$, more prior weight was assigned to
-higher $N$ values.
-
-\add{Another possible contributing factor to the overestimation of $I$ and $N$
-relates to the dynamics of the SI model and the coalescent process. The number
-of infected individuals follows a logistic growth curve under the SI model.
-This kind of growth curve has three qualitative phases: a slow ramp-up, an
-exponential growth phase, and a slow final phase when the susceptible
-population is almost depleted. The waiting times until the next transmission,
-which determine the coalescence times in the tree, are dependent on the growth
-phase of the epidemic. Therefore, we hypothesize that it is the growth phase at
-the time of sampling which most affects tree shape, rather than the specific
-values of $I$ or $N$. To investigate this hypothesis, we simulated transmission
-trees over networks on a grid of $I$ and $I$ values and fit logistic growth
-curves to the proportion of infected individuals over time. We and calculated
-the first and second derivatives of these curves (with respect to time) at the
-time of transmission tree sampling. As shown in \cref{fig:derivs}, there are
-bands along which both derivatives are similar which contain the values we
-tested. These bands span mostly higher values of $N$ and $I$ than the true
-values. Thus, if $N$ and $I$ are free to vary (as is the case in ABC), both
-parameters will tend to be overestimated due to being less identifiable within
-their own band. However, when $N$ is fixed at 5000, the derivatives vary
-substantially along the $I$-axis, which explains why a marginal estimate of $I$
-was more accurate (\cref{fig:marginal}). We also note the resemblance of the
-contour surface of \cref{fig:derivs} to the two-dimensional marginal posterior
-distribution on $I$ and $N$ obtained with simulated data (\cref{fig:abc2}).}
-
-\add{The $m$ parameter, which controls the number of connections added to the
-network per vertex, did not have a measurable impact on tree shape and was not
-identifiable with ABC. However, all the analyses presented here did not take
-the absolute size of the transmission trees into account, as the branch lengths
-were rescaled by their mean. Because higher $m$ values imply higher edge
-density, an epidemic should spread more quickly for higher $m$ than lower $m$
-with the same per-edge transmission probability. Hence, considering the
-absolute height of the trees may improve our method's ability to reconstruct
-$m$. It was also pointed out to us by an anonymous reviewer that for a fixed
-$I$, an infected node may only end up transmitting along a fraction of its
-outgoing edges, which could mask the presence of the extra edges associated
-with higher $m$.}
-
-\end{document}
-
-As noted by \textcite{lintusaari2016identifiability}, uniform priors on model
-parameters may translate to highly informative priors on quantities of
-interest. We observed a non-linear relationship between the preferential
-attachment power $\alpha$ and the power law exponent $\gamma$
-(\cref{fig:gamma}). Therefore, placing a uniform prior on $\alpha$ between 0
-and 2 is equivalent to placing an informative prior that $\gamma$ is close to
-2. Therefore, if we were primarily interested in $\gamma$ rather than
-$\alpha$, a more sensible choice of prior might have a shape informed by
-\cref{fig:gamma} and be bounded above by approximately $\alpha$ = 1.5. This
-would uniformly bound $\gamma$ in the region $2 \leq \gamma \leq 4$ commonly
-reported in the network literature~\autocite{liljeros2001web,
-schneeberger2004scale, colgate1989risk, brown2011transmission}. We note however
-that \textcite{jones2003assessment} estimated $\gamma$ values greater than
-four for some datasets, in one case as high as 17, indicating that a wider
-range of permitted $\gamma$ values may be warranted.
-
-\subsection*{Analysis of real world HIV datasets}
-
-Our investigation of published HIV datasets indicated heterogeneity in the
-contact network structures underlying several distinct local epidemics. When
-interpreting these results, we caution that the BA model is quite simple and
-most likely misspecified for these data. In particular, the average degree of a
-node in the network is equal to $2m$, and therefore is constrained to be a
-multiple of 2. Furthermore, we considered the case $m = 1$, where the network
-has no cycles, to be implausible and therefore assigned it zero prior
-probability in one set of analyses. This forced the average degree to be at
-least four, which may be unrealistically high for sexual networks. The fact
-that the estimated values of $\alpha$ differed substantially for three datasets
-depending on whether or not $m = 1$ was allowed by the prior is further evidence
-of this potential misspecification. However, we note that for \del{two}\add{the
-majority} of the datasets, the estimated values of $\alpha$ did not change much
-between priors, and the estimates of $I$ were robust to the choice of prior for
-all datasets studied (compare \cref{fig:abchpd,fig:abchpdm2}). More
-sophisticated models, for example models incorporating heterogeneity in node
-behaviour, are likely to provide a better fit to these data.
-
-<<disc, include=FALSE>>=
-    source("../thesis/global.R")
-    d <- setDT(collect.data("../../simulations/aggregates/hpd/*"))
-    d <- d[agg_gene == "gag"]
-    d1 <- d[m_min == 1]
-    d2 <- d[m_min == 2]
-@
-
-\del{With respect to the preferential attachment power $\alpha$, the five
-datasets analysed fell into three categories (\cref{fig:abchpd}). First, we
-estimated a preferential attachment power close to 1, indicating linear
-preferential attachment, for the outbreaks studied by
-\textcite{niculescu2015recent} and \textcite{wang2015targeting}. These values
-were robust to specifying different priors for $m$. Both studies were of
-populations in which we would expect a high degree of epidemiological
-relatedness: \textcite{niculescu2015recent} studied a recent outbreak among
-Romanian injection drug users (IDU), while \citeauthor{wang2015targeting}
-sampled acutely infected MSM in Beijing, China. Both these are contexts in
-which we would expect some of the assumptions of the BA model, such as a
-connected network, relatively high mean degree, and preferential attachment
-dynamics, to hold.}
-
-\del{The remaining three datasets (\textcite{cuevas2009hiv, novitsky2014impact,
-li2015hiv}) had estimated values of $\alpha$ below 0.5 when $m = 1$ was
-included in the prior, but these were not robust to changing the prior to
-exclude $m = 1$. For the \citeauthor{cuevas2009hiv} data, model
-misspecification is likely partially responsible. While the authors found that
-a large proportion of the samples were epidemiologically linked, these were
-mainly in small local clusters rather than the single large component
-postulated by the BA model. In addition, the mixed risk groups in the dataset
-would be unlikely to significantly interact, further weakening any global
-preferential attachment dynamics. The dataset studied by
-\textcite{novitsky2014impact} originated from a densely sampled population
-where the predominant risk factor was believed to be heterosexual exposure.
-Although the MAP estimate of $\alpha$ was almost unchanged when the value $m =
-1$ was excluded from the prior, the confidence interval shrank significantly.
-For both priors, the estimated $I$ value was extremely high, in fact higher
-than the estimated HIV prevalence in the sampled region. The authors indicated
-that the source of the samples was a town in close proximity to the country's
-capital city, and suggested that there may have been a high degree of migration
-and partner interchange between the two locations. It is possible that the
-contact network underlying the subtree we investigated includes a much larger
-group based in the capital city, which would explain the high estimate of $I$.
-There is no clear explanation for the discrepancy between the two priors for
-the \textcite{li2015hiv} data, as the subset we analyzed formed a phylogenetic
-cluster and therefore was a good candidate for the BA model. However, nearly
-all the posterior density was assigned to $m = 1$ when this value was allowed,
-indicating that the network was more likely to have an acyclic tree structure.}
-
-\subsubsection{Preferential attachment power is sub-linear and higher for IDU networks}
-
-\add{For all datasets we examined, the posterior mean estimates for $\alpha$
-were sub-linear, ranging from 
-    \Sexpr{d[m_min == 1 & parameter == "alpha", round(min(mean), 2)]}
-to 
-    \Sexpr{d[m_min == 1 & parameter == "alpha", round(max(mean), 2)]}.
-The sub-linearity is consistent with the results of
-\textcite{de2007preferential}, who developed a statistical inference method to
-estimate the parameters of a more sophisticated preferential attachment model
-incorporating heterogeneous node behaviour. When used to analyze
-population-level longitudinal partner count data, they found $\alpha$ values
-ranging from 0.26 to 0.62 depending on the gender and time period considered.
-Our estimates of $\alpha$ for the IDU/Romania was above this range under both
-priors, as were the estimates for the MSM/Beijing data and the BC data when $m
-= 1$ was disallowed by the prior. The dataset investigated by
-\textcite{de2007preferential} was derived from a survey of a random sample of
-the Norwegian population, whereas our investigation focused on datasets from
-known phylogenetic or geographic clusters of HIV-infected persons. It is
-therefore unsurprising that we detected stronger preferential attachment
-dynamics in some cases. For instance, random sampling is much less likely to
-discover the high-degree nodes characterizing the tail of the degree
-distribution, simply because those individuals are rare in the general
-population. In addition, it is plausible that HIV-positive individuals are more
-likely to be highly connected in their sexual networks, as the odds of
-acquiring HIV increase with the number of unprotected sexual contacts.}
-
-\add{Both \textcite{de2007preferential} and the HET/Botswana data studied
-populations whose primary risk factor for HIV infection was heterosexual
-contact. \citeauthor{de2007preferential} explicitly excluded reported
-homosexual contacts; \citeauthor{novitsky2014impact} did not, but noted that
-heterosexual contact is the primary mode of transmission in Botswana where the
-study was done. In the first of the two papers describing the Botswana
-study~\autocite{novitsky2013phylogenetic}, the authors noted that their sample
-was gender-biased, being composed of approximately 75\% women. Our estimate of
-$\alpha$ for these data was 
-    \Sexpr{d[m_min == 1 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]}
-        or 
-    \Sexpr{d[m_min == 2 & dataset == "novitsky2014" & parameter == "alpha", round(mean, 2)]},
-depending on the prior on $m$. Similarly, \textcite{de2007preferential}
-estimated 0.54, 0.57, and 0.29 for 3-year, 5-year, and lifetime partnership
-networks respectively for the female portion of their sample.}
-
-\add{For both choices of prior on $m$, the datasets derived from IDU
-populations had a higher estimated preferential attachment power than the other
-datasets (\cref{fig:abchpd,fig:abchpdm2}). This finding is in line with
-\textcite{dombrowski2013topological}, who reanalyzed a network of IDUs in
-Brooklyn, USA, collected between 1991 and 1993~\autocite{friedman2006social}.
-They found that the IDU network resembled a BA network much more closely than
-other social and sexual networks, and offered sociological explanations for the
-apparent preferential attachment dynamics in this population. Importantly, from
-a public health perspective, the authors asserted that the removal of
-\emph{random} individuals from IDU networks may have the paradoxical effect of
-increasing the network's epidemic susceptibility. When low-degree nodes are
-removed, as would occur during a police crackdown, their network neighbours may
-turn to well-known community members for advice or supplies, thus increasing
-the connectivity of these high-degree nodes.}
-
-\add{Unfortunately, the sub-linear region for $\alpha$ identified by both
-\textcite{de2007preferential} and \textit{netabc} is also the region of
-poorest identifiability (\cref{fig:abcptm2,fig:abcptm3,fig:abcptm4}). This was
-reflected in the high level of uncertainty in the estimates, with most 95\%
-HPD intervals covering the majority of the range [0, 1]. The value $\alpha =
-0.5$ was contained in the 95\% HPD interval for every dataset; consequently, it
-is not possible to say with high confidence that any of the $\alpha$ values are
-different from each other. In synthetic data, the confidence intervals around
-$\alpha$ narrowed when other parameters were marginalized out
-(\cref{fig:marginal}). Thus, it is possible that estimates of $\alpha$ could be
-made more precise by specifying either exact values or informative priors on
-the other BA parameters when these are known.}
-
-\add{I is underestimated in some cases and overestimated in others. Some of the
-datasets have most of the posterior mass on m = 1. Estimates of N are
-meaningless.}
-
-\subsubsection{Modelling assumptions}
-
-In addition to the aforementioned possibility of misspecification, additional
-modelling assumptions include the network being connected and static, all
-transmission rates being equal, no removal after infection, identical behaviour
-of all nodes, and random sampling. The last two were addressed with small-scale
-experiments. We simulated a network where some nodes exhibited a higher
-attachment power than others, and found that the estimated attachment power was
-simply the average of the two values. This indicated that, although we could
-characterize the network in aggregate, the estimated parameters could not be
-said to apply to any individual node. The effect of biased sampling was
-investigated by analysing a transmission tree which had been sampled in a
-peer-driven fashion. The results were roughly in line with those for random
-sampling, however the estimated value of $\alpha$ was lower than the average
-for randomly-sampled trees. Further experiments would be necessary to fully
-explore the impact of these assumptions on the method's accuracy. However,
-despite these issues, we felt it was best to demonstrate the method first on a
-simple model. It is possible to use this framework to fit more complex models
-which address some of these issues, such as one incorporating heterogeneous
-node behaviour, which may prove a fruitful avenue
-for future investigations.
-
-Our method has a number of caveats, perhaps the most significant being that it
-takes a transmission tree as input. In reality, true transmission trees are not
-available and must be approximated, often by way of a viral phylogeny. Although
-this has been demonstrated to be a fair
-approximation~\autocite[e.g.][]{leitner1996accurate}, and is frequently used in
-practice~\autocite[e.g.][]{stadler2013uncovering}, the topologies of a viral
-phylogeny and transmission tree can differ
-significantly~\autocite{ypma2013relating} due to within-host evolution and the
-sampling process. In addition, the ABC-SMC algorithm is
-computationally intensive, taking about a day when run on 20 cores in parallel
-with the settings we described in the methods. Nevertheless, our method is
-potentially useful to epidemiological researchers interested in the general
-characteristics of the network structure underlying disease outbreaks. This
-work, and previous work by our group~\autocite{poon2015phylodynamic}, has
-demonstrated that ABC is a broadly applicable and effective framework in
-which to perform phylodynamic inference.
-
-\section*{Acknowledgements}
-
-\add{We are grateful to Dr. Sally Otto, Dr. Alexandre Bouchard-C\^ot\'e, 
-Dr. Richard Harrigan, and the two anonymous reviewers for many helpful
-suggestions.} This work was supported by grants from the Canadian Institutes of
-Health Research (CIHR, operating grant HOP-111406), and the Bill \& Melinda
-Gates Foundation (award number OPP1110049). R.M.M. was supported by a
-scholarship from the CIHR Strategic Training Program in Bioinformatics.
-A.F.Y.P. was supported by a CIHR New Investigator Award (Canadian HIV Vaccine
-Initiative, Vaccine Discovery and Social Research) and by a Career Investigator
-Scholar Award from the Michael Smith Foundation for Health Research, in
-partnership with the Providence Health Care Research Institute and St. Paul's
-Hospital Foundation.
-
-\printbibliography
-
-\clearpage
-\newpage
-
-\section*{Supplementary Figures}
-
-\renewcommand{\thefigure}{S\arabic{figure}}
-\setcounter{figure}{0}
-\setcounter{page}{1}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{leventhal2012fig1.pdf}
-  \caption{
-    Reproduction of Figure 1A from Leventhal \textit{et al.} (2012) used to
-    check the accuracy of our implementation of Gillespie simulation.
-    Transmission trees were simulated over three types of network, with
-    pathogen transmissibility varying from 0 to 1 (1000 trees per
-    transmissibility value). Sackin's index was calculated for each simulated
-    transmission tree. Lines are means, light shaded areas are 95\% quantile
-    range, and dark shaded areas are interquartile ranges. 
-    }
-  \label{fig:leventhal}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{smc-test.pdf}
-  \caption{Approximation of mixture of Gaussians used by
-    Del Moral \textit{et al.} (2012) and Sisson \textit{et al.} (2009) to test
-    SMC. Solid black line indicates true distribution. Grey shaded area shows
-    SMC approximation obtained with our implementation, using 10000
-    particles with one simulated data point per particle.
-  }
-  \label{fig:smctest}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{kernel-alpha-crossv.pdf}
-  \caption{
-    Cross-validation accuracy of kernel-SVM classifiers for $\alpha$ parameter
-    of BA network model, for various tree kernel meta-parameters and
-    epidemic scenarios. Each point was calculated based on 300 simulated
-    transmission trees over networks with $\alpha$ = 0.5, 1.0, or 1.5. Dotted
-    and and dashed lines indicate, respectively, performance of SVM using the
-    nLTT statistic, and linear regression using Sackin's index. Facets
-    are number of infected nodes before the simulation was stopped ($I$) and
-    number of tips in the sampled transmission tree.
-  }
-  \label{fig:alphacrossv}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{kernel-m-crossv.pdf}
-  \caption{Cross-validation accuracy of kernel-SVM classifiers for $m$
-      parameter of BA network model, for various tree kernel
-      meta-parameters and epidemic scenarios. Each point was calculated based
-      on 300 simulated transmission trees over networks with $m$ = 2, 3, or 4.
-      Dotted and and dashed lines indicate, respectively, performance of SVM
-      using the nLTT statistic, and linear regression using Sackin's
-      index. Facets are number of infected nodes before the simulation was
-      stopped ($I$) and number of tips in the sampled transmission tree.
-  }
-  \label{fig:mcrossv}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{kernel-I-crossv.pdf}
-  \caption{Cross-validation accuracy of kernel-SVM classifiers for number of
-      infected nodes ($I$) under BA network model, for various tree
-      kernel meta-parameters and two tree sizes. Each point was calculated
-      based on 300 simulated transmission trees over networks with $I$ = 500,
-      1000, or 2000. Dotted and and dashed lines indicate, respectively,
-      performance of SVM using the nLTT statistic, and linear regression
-      using Sackin's index. Facets are the number of tips in the sampled
-      transmission tree.
-  }
-  \label{fig:Icrossv}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{kernel-N-crossv.pdf}
-  \caption{Cross-validation accuracy of kernel-SVM classifiers for total number
-      of nodes ($N$) under BA network model, for various tree kernel
-      meta-parameters and epidemic scenarios sizes. Each point was calculated
-      based on 300 simulated transmission trees over networks with $N$ = 3000,
-      5000, or 8000. Dotted and and dashed lines indicate, respectively,
-      performance of SVM using the nLTT statistic, and linear regression
-      using Sackin's index. Facets are the number of tips in the sampled
-      transmission tree.
-  }
-  \label{fig:Ncrossv}
-\end{figure}
-
-\begin{figure}
-    \includegraphics[width=\textwidth]{abc-boxplot-mN}
-    \caption[
-        Posterior mean point estimates for BA model parameters $m$ and $N$
-        obtained by running \textit{netabc} on simulated data, stratified by
-        true parameter values.
-    ]{
-        Posterior mean point estimates for BA model parameters $m$ and $N$
-        obtained by running \textit{netabc} on simulated data, stratified by
-        true parameter values. First row of plots contains true versus
-        estimated values of $m$; second row contains true versus estimated
-        values of $N$. Columns are stratified by $\alpha$, $I$, and $m$
-        respectively. Dashed lines indicate true values. 
-    }
-\end{figure}
-
-\begin{figure}
-    \includegraphics[width=\textwidth]{abc-marginal}
-    \caption[
-        Posterior means, 50\% HPD intervals, and 95\% HPD intervals for BA
-        model parameter estimated marginally with ABC.
-    ]{
-        Posterior means (points), 50\% HPD intervals (notches), and 95\% HPD
-        intervals (lines) for BA model parameter estimated marginally with ABC.
-        $x$-axis labels indicate parameters which were fixed to their true
-        values by specifying Dirac-delta priors.
-    }
-    \label{fig:marginal}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{alpha-gamma.pdf}
-  \caption{
-      Relationship between preferential attachment power parameter $\alpha$
-      and power law exponent $\gamma$ for networks simulated under the BA
-      network model with $N$ = 5000 and $m$ = 2.
-  }
-  \label{fig:gamma}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics{mixed-posterior}
-    \caption{
-        Approximate marginal posterior distributions of Barab\'asi-Albert
-        model parameters obtained using ABC for a network with
-        heterogeneous node behaviour. Half of the nodes were attached with
-        $\alpha = 0.5$, and the other half with $\alpha = 1.5$ (vertical
-        dashed lines, top right). Other parameter values were $m = 2$, $I =
-        1000$, and $N = 5000$ (vertical dashed lines, other than top right).
-        Shaded areas indicate 95\% highest posterior density intervals.
-    }
-    \label{fig:mixed}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics{peerdriven-posterior}
-    \caption{Approximate marginal posterior distributions of Barab\'asi-Albert
-        model parameters obtained using ABC for a network with
-        peer-driven sampling. An epidemic was simulated in the usual fashion,
-        but rather than being sampled at random, infected nodes were sampled
-        with a probability two times higher if they had any sampled neighbours
-        in the contact network. Vertical dashed lines indicate true parameter
-        values, and shaded areas indicate 95\% highest posterior density
-        intervals.
-    }
-    \label{fig:peerdriven}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[ht]
-  \includegraphics{realdata-hpd-m2}
-  \vspace{8pt}
-  \caption{
-      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
-      parameters of the BA network model, fitted to five published HIV datasets
-      with ABC. $x$-axes indicate regions of nonzero prior density. In
-      particular, the prior on $m$ was DiscreteUniform(2, 5).
-  }
-  \label{fig:abchpdm2}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{realdata-genes}
-    \caption[
-        Posterior means and 50\%/95\% HPD intervals for parameters of the
-        BA network model, fitted to three HIV datasets where both \textit{gag}
-        and \textit{env} genes were sequenced.
-    ]{
-        Posterior means (points), 50\% HPD intervals (notches), and 95\%
-        HPD intervals (lines) for parameters of the BA network model, fitted to
-        three HIV datasets where both \textit{gag} and \textit{env} genes were
-        sequenced.
-    }
-    \label{fig:genes}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{derivatives.pdf}
-  \caption[
-    First and second time derivatives of epidemic growth curves at time of
-    sampling for various values of $I$ and $N$.
-  ]{
-    First and second \add{time} derivatives of epidemic growth curves at time
-    of sampling for various values of \gls{I} and \gls{N}. Networks were
-    simulated under the \gls{BA} model with \gls{alpha} = 1.0, \gls{m} = 2, and
-    \gls{N} varied along the values shown on the $x$-axis. Transmission trees
-    were sampled at the time when \gls{I} nodes were infected ($y$-axis).
-    Logistic growth curves were fit to epidemic trajectories derived from the
-    transmission trees, and their first and second derivatives were calculated
-    at the time of sampling. Contours show first derivatives, while colours
-    indicate second derivatives. Values of \gls{I} and \gls{N} used in
-    simulation experiments with \gls{ABC} are indicated by diamonds.
-  }
-  \label{fig:derivs}
-\end{figure}
-
-%\begin{figure}[ht]
-%  \includegraphics{cuevas2009-posterior}
-%  \caption{
-%      Posterior distribution of BA model parameters for
-%      \textcite{cuevas2009hiv} data. Vertical lines indicate maximum \textit{a
-%      posteriori} estimates, and shaded areas are 95\% highest posterior
-%      density intervals. $x$-axis indicates regions of nonzero prior density.
-%  }
-%  \label{fig:cuevas}
-%\end{figure}
-%
-%\begin{figure}[ht]
-%  \includegraphics{li2015-posterior}
-%  \caption{
-%      Posterior distribution of BA model parameters for \textcite{li2015hiv}
-%      data. Vertical lines indicate maximum \textit{a posteriori} estimates,
-%      and shaded areas are 95\% highest posterior density intervals. $x$-axis
-%      indicates regions of nonzero prior density.
-%  }
-%  \label{fig:li}
-%\end{figure}
-%
-%\begin{figure}[ht]
-%  \includegraphics{niculescu2015-posterior}
-%  \caption{
-%      Posterior distribution of BA model parameters for
-%      \textcite{niculescu2015recent} data. Vertical lines indicate maximum
-%      \textit{a posteriori} estimates, and shaded areas are 95\% highest
-%      posterior density intervals. $x$-axis indicates regions of nonzero prior
-%      density.
-%  }
-%  \label{fig:niculescu}
-%\end{figure}
-%
-%\begin{figure}[ht]
-%  \includegraphics{novitsky2014-posterior}
-%  \caption{
-%      Posterior distribution of BA model parameters for
-%      \textcite{novitsky2013phylogenetic, novitsky2014impact} data. Vertical
-%      lines indicate maximum \textit{a posteriori} estimates, and shaded areas
-%      are 95\% highest posterior density intervals. $x$-axis indicates regions
-%      of nonzero prior density.
-%  }
-%  \label{fig:novitsky}
-%\end{figure}
-%
-%\begin{figure}[ht]
-%  \includegraphics{wang2015-posterior}
-%  \caption{
-%      Posterior distribution of BA model parameters for
-%      \textcite{wang2015targeting} data. Vertical lines indicate maximum
-%      \textit{a posteriori} estimates, and shaded areas are 95\% highest
-%      posterior density intervals. $x$-axis indicates regions of nonzero prior
-%      density.
-%  }
-%  \label{fig:wang}
-%\end{figure}
-%
-%<<abc_posterior, echo=FALSE, results="asis", eval=FALSE>>=
-%    N <- 5000
-%    replicate <- 0
-%    for (m in c(2, 3, 4)) {
-%    for (I in c(1000, 2000)) {
-%    for (alpha in c(0, 0.5, 1, 1.5)) {
-%        fn <- sprintf("{abc-posterior/%.1f_%d_%d_%d_%d}.pdf", alpha, I, m, N, replicate)
-%        lab <- sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate)
-%        cat("\\begin{figure}[ht]\n")
-%        cat("\\centering\n")
-%        cat(paste0("\\includegraphics{", fn, "}\n"))
-%        cat("\\caption{\n")
-%        cat(sprintf("Approximate marginal posterior distributions of Barab\\'asi-Albert model parameters obtained using ABC for one simulated transmission tree with parameter values $\\alpha = %.1f$, $I = %d$, $m = %d$, and $N = %d$.
-Dashed lines indicate true values, and shaded areas show 95\\%% highest posterior density intervals.\n", alpha, I, m, N))
-%        cat("}\n")
-%        cat(paste0("\\label{", lab, "}\n"))
-%        cat("\\end{figure}\n")
-%        cat("\\clearpage\n\n")
-%    }
-%    }
-%    }
-%@
-
-\end{document}
+\end{frame}

### 2016-07-05 doc/defence/posterior1.tex
Presentation updates

--- 2016-04-15 doc/hiv-dynamics/presentation.tex
@@
-\documentclass{beamer}
-\usepackage{tikz}
-\usepackage{tikz-qtree}
-\usepackage{gillius}
-\usepackage{abraces}
-\usepackage{etex}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{qrcode}
-\usepackage[backend=biber]{biblatex}
+\begin{frame}{Marginal posterior distributions for $\alpha$ show least spread}
+  \begin{minipage}[p][\textheight][t]{\textwidth}
+    \begin{center}
+    preferential attachment power \hfill prevalence
+      \includegraphics[width=0.8\textwidth]{abc-posterior-example.pdf}
 
-% http://tex.stackexchange.com/questions/12703/how-to-create-fixed-width-table-columns-with-text-raggedright-centered-raggedlef
-\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-
-\usetikzlibrary{external}
-\usetikzlibrary{shapes}
-\usetikzlibrary{arrows}
-\usetikzlibrary{positioning}
-\usetikzlibrary{decorations.pathreplacing}
-
-\usetheme[everytitleformat=regular]{m}
-\setbeamertemplate{navigation symbols}{}
-
-\graphicspath{{../figures/}}
-\newcommand{\tablepath}{../tables}
-
-\title{Phylogenetic inference of contact network parameters with kernel approximate Bayesian computation}
-\author[RMM, RHL \& AFYP]{Rosemary M McCloskey$^1$ \and Richard H Liang$^1$ \and Art FY Poon$^{1,2}$}
-\institute[UBC \& BCCfE]{$^1$BC Centre for Excellence in HIV/AIDS, Vancouver, Canada \\ $^2$Department of Medicine, University of British Columbia, Vancouver, Canada}
-\date{HIV Dynamics \& Evolution, Woods Hole, USA, April 25, 2016}
-
-\newcommand{\dd}[2]{\frac{\text{d}\,#1}{\text{d}\,#2}}
-
-\addbibresource{papers.bib}
-
-\begin{document}
-\setbeamercolor{background canvas}{bg=white}
-\renewcommand{\footnotesize}{\tiny}
-\definecolor{red}{RGB}{228,26,28}
-\definecolor{blue}{RGB}{55,126,184}
-\definecolor{green}{RGB}{77,175,74}
-\definecolor{purple}{RGB}{152,78,163}
-
-\maketitle
-
-\begin{frame}{Most epidemiological models assume homogeneous mixing}
-  \begin{columns}
-    \begin{column}{0.6\textwidth}
-      \centering
-      \includegraphics{compartments}
-      \vspace{1cm}
-
-      \uncover<2->{\includegraphics[width=\textwidth]{sir-trajectories}}
-    \end{column}
-    \begin{column}{0.4\textwidth}
-      \vspace{-2cm}
-      \begin{itemize}
-        \setlength{\itemsep}{24pt}
-        \item Often provides reasonable approximation in practice.
-        \uncover<2->{
-        \item Can be inaccurate when substantial contact heterogeneity exists.
-        }
-      \end{itemize}
-    \end{column}
-  \end{columns}
-\end{frame}
-
-\begin{frame}{Network models capture contact heterogeneity}
-  \begin{columns}
-    \begin{column}{0.5\textwidth}
-      \includegraphics[width=\textwidth]{contactnet-empty}
-    \end{column}
-    \begin{column}{0.5\textwidth}
-      \begin{itemize}
-        \setlength{\itemsep}{12pt}
-          \pause
-        \item May offer more accurate predictions for highly structured
-          populations.
-          \pause
-        \item Network parameters may be of interest for their own sake,
-          \textit{e.g.} are there superspreaders?
-          \pause
-        \item Extremely difficult to estimate in practice.
-      \end{itemize}
-    \end{column}
-  \end{columns}
-
-\end{frame}
-
-\begin{frame}{Contact networks shape transmission trees}
-  \includegraphics[width=\textwidth]{contactnet}
-
-  \begin{itemize}
-      \pause
-    \item Transmission trees in turn shape viral phylogenies.
-      \pause
-    \item \textbf{Aim:} estimate contact network parameters from viral
-      phylogenies.
-  \end{itemize}
-\end{frame}
-
-\begin{frame}{Scale-free networks and preferential attachment}
-  \includegraphics[width=\textwidth]{pa-example}
-  \begin{itemize}
-    \item Sexual networks are often scale-free: degree distributions appear
-      follow a power law ($\Pr(\text{degree} = d) \propto d^{\gamma}$).
-      \pause
-    \item Preferential attachment models generate scale-free networks.
-  \end{itemize}
-\end{frame}
-
-\begin{frame}{Barab\'asi-Albert (BA) network model}
-  \begin{columns}
-    \begin{column}{0.5\textwidth}
-      \begin{minipage}[p][\textheight][t]{\textwidth}
-        \centering
-        \only<1>{\includegraphics[page=1, width=\textwidth]{pa.pdf}}
-        \only<2>{\includegraphics[page=2, width=\textwidth]{pa.pdf}}
-        \only<3>{\includegraphics[page=3, width=\textwidth]{pa.pdf}}
-        \only<4>{\includegraphics[page=4, width=\textwidth]{pa.pdf}}
-        \only<5>{\includegraphics[page=5, width=\textwidth]{pa.pdf}}
-        \only<6>{\includegraphics[page=6, width=\textwidth]{pa.pdf}}
-      \end{minipage}
-    \end{column}
-    \begin{column}{0.5\textwidth}
-      \begin{minipage}[t][\textheight][t]{\textwidth}
-      \begin{itemize}
-        \uncover<1->{ \item Start with a small number of connected nodes. }
-        \uncover<2->{ \item Attach new nodes with $m$ edges. }
-        \uncover<3->{ \item Other endpoints of degree $d$ are chosen with probability $\propto d^{\alpha} + 1$. }
-        \uncover<5->{ \item Continue until network has $N$ nodes. }
-        \uncover<6->{ \item Also consider the prevalence $I$ when the transmission
-          tree is sampled.}
-      \end{itemize}
-      \end{minipage}
-    \end{column}
-  \end{columns}
-\end{frame}
-
-\begin{frame}{Do network parameters measurably affect tree shape?}
-  \only<1>{
-    \includegraphics[height=1in, trim=0 2in 4in 0, clip]{kernel-idea}
-  }
-  \only<2>{
-    \includegraphics[height=1in, trim=0 2in 2in 0, clip]{kernel-idea}
-  }
-  \only<3->{
-    \includegraphics[height=1in, trim=0 2in 0 0, clip]{kernel-idea}
-  }
-  \begin{itemize}
-    \item Generate networks under different parameter values ($N$, $m$, $\alpha$).
-    \uncover<2->{
-    \item Simulate epidemic over each network with Gillespie algorithm ($I$).
-    }
-    \uncover<3->{
-    \item Randomly subsample to form transmission trees.
-    }
-    \uncover<4->{
-    \item Compare pairwise using tree kernel\footfullcite{poon2013mapping}.
-    }
-  \end{itemize}
-\end{frame}
-
-\begin{frame}{$\alpha$ and $I$ separate well in feature space}
-  \begin{tikzpicture}
-    \node {\includegraphics[height=0.8\textheight]{kernel-kpca.pdf}};
-    \fill<1>[white] (-4.6,-3.2) rectangle (5,0.2);
-  \end{tikzpicture}
-  \pause
-\end{frame}
-
-\begin{frame}{Classifiers for network parameters}
-  \vspace{-1cm}
-  \begin{itemize}
-    \item kSVM using tree kernel
-    \item SVM using normalized
-      lineages-through-time\footfullcite{janzen2015approximate}
-    \item linear regression using Sackin's index~\footfullcite{shao1990tree}
-  \end{itemize}
-
-  \pause
-  \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
-\end{frame}
-
-\begin{frame}{Approximate Bayesian computation (ABC) for networks}
-  \begin{minipage}[p][\textheight][t]{\textwidth}
-    \only<1>{\includegraphics[width=\textwidth, trim=0 3in 0 0, clip]{abc-idea.pdf}}
-    \only<2>{\includegraphics[width=\textwidth, trim=0 1.8in 0 0, clip]{abc-idea.pdf}}
-    \only<3>{\includegraphics[width=\textwidth, trim=0 1in 0 0, clip]{abc-idea.pdf}}
-    \only<4>{\includegraphics[width=\textwidth, trim=0 0in 0 0, clip]{abc-idea.pdf}}
+      \vspace{-0.25cm}
+      edges per vertex \hfill total nodes
+    \end{center}
   \end{minipage}
 \end{frame}
-
-\begin{frame}{Sequential Monte Carlo for ABC}
-  \begin{tikzpicture}
-    \only<1-2>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 4in 0, clip]{abc-smc.pdf}}};
-    \only<1>{\fill[white] (0.2,-6.5) rectangle (4,-3.5)};
-    \only<3>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 2in 0, clip]{abc-smc.pdf}}};
-    \only<4>{\node at (0, 0) [anchor=north west] {\includegraphics[height=0.7\textheight, trim=0 0in 0 0, clip]{abc-smc.pdf}}};
-  \end{tikzpicture}
-\end{frame}
-
-\begin{frame}{Marginal posterior distributions for one simulated tree}
-  \begin{minipage}[p][\textheight][t]{\textwidth}
-    \only<1>{\centerline{\includegraphics[width=0.9\textwidth, trim=0 2.5in 0 0, clip]{abc-posterior-example.pdf}}}
-    \only<2>{\centerline{\includegraphics[width=0.9\textwidth]{abc-posterior-example.pdf}}}
-  \end{minipage}
-\end{frame}
-
-\begin{frame}{Accuracy of maximum \textit{a posteriori} estimates for $m = 2$}
-  \begin{minipage}[p][\textheight][t]{\textwidth}
-    \vspace{-0.5cm}
-    \only<1>{\includegraphics[height=0.9\textheight, trim=0 0 3.4in 0, clip]{abc-point-estimate-m2.pdf}}
-    \only<2>{\includegraphics[height=0.9\textheight]{abc-point-estimate-m2.pdf}}
-  \end{minipage}
-\end{frame}
-
-\begin{frame}{Application to published HIV datasets}
-  \uncover<2->{
-  \includegraphics[width=\textwidth]{realdata-hpd-bc-horiz}
-  }
-  \vspace{0.5cm}
-
-  \scriptsize
-  \input{\tablepath/realdata-bc}
-  \normalsize
-\end{frame}
-
-\begin{frame}{Conclusions}
-  \begin{itemize}
-    \item Kernel-ABC is a viable method for fitting contact network models to
-      phylogenetic data.
-      \pause
-    \item Some parameters of the Barab\'asi-Albert preferential attachment data
-      have a measurable impact on tree shape and can be re-estimated using
-      kernel-ABC.
-      \pause
-    \item The networks underlying real epidemics may be heterogeneous,
-      underscoring the importance of considering network structure in
-      phylodynamic analyses.
-      \pause
-  \end{itemize}
-  \centerline{\qrcode{github.com/rmcclosk/netabc} github.com/rmcclosk/netabc}
-\end{frame}
-
-\begin{frame}{Acknowledgements}
-  \begin{columns}
-    \begin{column}{0.6\textwidth}
-
-      \textbf{BC Centre for Excellence in HIV/AIDS}
-
-      Art Poon
-
-      Jeff Joy
-
-      Richard Liang
-
-      Thuy Nguyen
-
-      P. Richard Harrigan
-
-      \hfill\\
-      \textbf{University of British Columbia}
-
-      Sarah Otto
-
-      Alexandre Bouchard-C\^ot\'e
-
-      \vfill
-    \end{column}
-    \begin{column}{0.4\textwidth}
-      \centering
-
-      \includegraphics[width=3cm]{logos/cihr}
-      \vspace{0.5cm}
-
-      \includegraphics[width=3cm]{logos/bmgf}
-      \vspace{0.5cm}
-
-      \includegraphics[width=1cm]{logos/ubc}
-      \vspace{0.5cm}
-
-      \includegraphics[width=3cm]{logos/btp}
-    \end{column}
-  \end{columns}
-\end{frame}
-\end{document}

### 2016-07-05 doc/defence/presentation.tex
Updates to presentation
Presentation updates

@@
 \maketitle
-\input{contactnet}
-\input{netmodel}
-\input{why}
-\input{ba}
-\input{phylonet}
-\input{transtree}
-\input{virusphylo}
-\input{inference}
-\input{identifiable}
+
+%\begin{frame}{Outline}
+%    \tableofcontents
+%\end{frame}
+
+%\section{Background and Motivation}
+%\input{contactnet}
+%\input{phylo}
+%\input{objects}
+%\input{transtree}
+%\input{virusphylo}
+%\input{inference}
+%%\input{netmodel}
+%\input{why}
+
+%\section{The Barab\'asi-Albert (BA) Contact Network Model}
+%\input{ba}
+%\input{identifiable}
+%\input{kpca}
+
+%\section{\textit{Netabc}: Phylogenetic Reconstruction of Contact Network Parameters with Approximate Bayesian Computation}
+%\input{smc}
+%\input{posterior1}
+%\input{posterior2}
+%\input{pointest}
+%\input{pointest2}
+
+\section{Application to real HIV epidemics}
+\input{data}
+
+\section{Conclusions}
 
 \end{document}
-
-\begin{frame}{yes they do}
-  \centerline{\includegraphics[height=0.8\textheight]{kernel-kpca}}
-\end{frame}
-
-\begin{frame}{Simulated trees can be used to fit network models}
-  \centerline{\includegraphics[width=0.9\textwidth]{abc-smc}}
-
-  \centerline{animate this?}
-\end{frame}
-
-\begin{frame}{Posterior mean estimates are more accurate for $\alpha$ and $I$}
-  \begin{minipage}[p][\textheight][t]{\textwidth}
-    \vspace{-0.5cm}
-    preferential attachment power \hfill \uncover<2>{edges per vertex}
-
-    \hspace{0.75cm}\only<1>{\includegraphics[height=0.8\textheight, trim=0 0 3.4in 0, clip]{abc-point-estimate-m3.pdf}}
-    \only<2>{\includegraphics[height=0.8\textheight]{abc-point-estimate-m3.pdf}}
-
-    prevalence \hfill \uncover<2>{total nodes}
-  \end{minipage}
-\end{frame}
-
-\begin{frame}{Example simulation\ldots}
-  1d posterior goes here
-  \pause
-%  \begin{minipage}[p][\textheight][t]{\textwidth}
-%    \begin{center}
-%    preferential attachment power \hfill prevalence
-%    \only<1>{
-%      \includegraphics[width=0.8\textwidth, trim=0 2.5in 0 0, clip]{abc-posterior-example.pdf}
-%    }
-%    \only<2>{
-%      \includegraphics[width=0.8\textwidth]{abc-posterior-example.pdf}
-%
-%      \vspace{-0.25cm}
-%      edges per vertex \hfill total nodes
-%    }
-%    \end{center}
-%  \end{minipage}
-\end{frame}
-
-\begin{frame}{Example simulation continued, $N$ and $I$ confounded}
-  2d posterior goes here
-  %\includegraphics[width=\textwidth]{{abc-posterior-2d/1.0_1000_3_5000_0}.pdf}
-\end{frame}
-
-\begin{frame}{Application to real world HIV datasets}
-  \centerline{\input{\tablepath/realdata-bc-short}}
-\end{frame}
 

### 2016-07-05 doc/defence/test.tex
Presentation updates

--- 2016-07-05 doc/defence/presentation.tex
@@
 
-\maketitle
-
-%\begin{frame}{Outline}
-%    \tableofcontents
-%\end{frame}
-
-%\section{Background and Motivation}
-%\input{contactnet}
-%\input{phylo}
-%\input{objects}
-%\input{transtree}
-%\input{virusphylo}
-%\input{inference}
-%%\input{netmodel}
-%\input{why}
-
-%\section{The Barab\'asi-Albert (BA) Contact Network Model}
-%\input{ba}
-%\input{identifiable}
-%\input{kpca}
-
-%\section{\textit{Netabc}: Phylogenetic Reconstruction of Contact Network Parameters with Approximate Bayesian Computation}
-%\input{smc}
-%\input{posterior1}
-%\input{posterior2}
-%\input{pointest}
-%\input{pointest2}
-
-\section{Application to real HIV epidemics}
-\input{data}
-
-\section{Conclusions}
+\input{objects2}
 
 \end{document}
-
-\begin{frame}{HIV networks have sub-linear preferential attachment}
-    Also, higher PA for IDU networks.
-
-    \includegraphics[trim=0 3.5in 3.3in 0, clip, width=0.5\textwidth]{realdata-hpd-bc}
-    \includegraphics[trim=3.4in 1.5in 0 2.0in, clip, width=0.5\textwidth]{realdata-hpd-bc}
-\end{frame}
-
-\begin{frame}{Conclusions}
-  \begin{itemize}
-    \item We developed a phylodynamic method to fit contact network models to
-      phylogenetic data.
-      \pause
-    \item The preferential attachment power of the Barab\'asi-Albert network
-      model, which is challenging to estimate by traditional epidemiological
-      methods, can be estimated with ABC.
-      \pause
-    \item The networks underlying real epidemics are heterogeneous,
-      underscoring the importance of considering network structure in
-      phylodynamic analyses.
-  \end{itemize}
-\end{frame}
-
-\begin{frame}{Acknowledgements}
-  \begin{columns}
-    \begin{column}{0.6\textwidth}
-
-      \textbf{BC Centre for Excellence in HIV/AIDS}
-
-      Art Poon
-
-      Jeff Joy
-
-      Richard Liang
-
-      Thuy Nguyen
-
-      P. Richard Harrigan
-
-      \hfill\\
-      \textbf{University of British Columbia}
-
-      Sarah Otto
-
-      Alexandre Bouchard-C\^ot\'e
-
-      \vfill
-      \vspace{0.5cm}
-      $\qquad\qquad$\includegraphics[width=2cm]{logos/genomecanada}
-      $\qquad$
-      \includegraphics[width=2cm]{logos/genomequebec}
-    \end{column}
-    \begin{column}{0.4\textwidth}
-      \centering
-
-      \includegraphics[width=3cm]{logos/cfe}
-      \vspace{0.5cm}
-
-      \includegraphics[width=3cm]{logos/cihr}
-
-      \includegraphics[width=3cm]{logos/genomebc}
-
-      \includegraphics[width=3cm]{logos/bmgf}
-      \vspace{0.5cm}
-
-      \includegraphics[width=3cm]{logos/btp}
-    \end{column}
-  \end{columns}
-\end{frame}
-
-\end{document}

### 2016-07-06 doc/paper/abstract.tex
Add paper and tex

@@
 accurately estimated. On the other hand, the mean degree of the network, as
-well as the total number of nodes, were not estimable with ABC. We observed
-substantial heterogeneity in the parameter estimates on real datasets, with
-point estimates for the preferential attachment power ranging from 0.06 to
-1.05. These results underscore the importance of considering contact structures
-when performing phylodynamic inference. Our method offers the potential to
+well as the total number of nodes, were not estimable with ABC. \del{We
+observed substantial heterogeneity in the parameter estimates on real datasets,
+with point estimates for the preferential attachment power ranging from 0.06 to
+1.05.} \add{We observed sub-linear preferential attachment (PA) power in all
+datasets, as well as higher PA power in networks of injection drug users.}
+These results underscore the importance of considering contact structures when
+performing phylodynamic inference. Our method offers the potential to
 quantitatively investigate the contact network structure underlying viral

### 2016-07-06 doc/paper/changes.tex
Add paper and tex

--- 2016-07-01 doc/paper/changes.tex
@@
 \signature{Rosemary M. McCloskey}
-\address{Centre for Excellence in HIV/AIDS \\ 608-1081 Burrard Street \\ Vancouver, BC, Canada V6Z 1Y6}
+\address{BC Centre for Excellence in HIV/AIDS \\ 608-1081 Burrard Street \\ Vancouver, BC, Canada V6Z 1Y6}
 \begin{document}
@@
     estimates by posterior means. Qualitatively, the results are not much
-    different, except that it is more obvious that $m$ and $N$ are not
-    identifiable with this method. 
+    different.
   \item We have replaced some ad hoc statistical tests in the analysis of the
-    simulation results with a $p$-value corrected generalized linear model, and
-    rewritten the relevant section of the results accordingly.
+    simulation results with a multiple test corrected generalized linear model,
+    and revised the relevant section of the results accordingly.
   \item We have found an analysis of the Barab\'asi-Albert model in the physics
     literature (Krapivsky et al. 2000) which demonstrates that the degree
-    distribution is only a power law when $\alpha = 1$. We acknowledge this but
-    do not believe it detracts from the validity of our results, especially in
-    light of a paper by de Blasio et al. (2007) which found sub-linear
-    preferential attachment through a population-level survey. We have added to
-    the supplemental materials a series of power law fits to the degree
-    distributions of BA networks with $\alpha \neq 1$ (Figure TODO), which
-    shows that, although the power law is cleary an imperfect fit, the slope
-    is still reasonably representative of the true distribution.
+    distribution is only a power law when $\alpha = 1$. We acknowledge this in
+    the revised manuscript but do not believe it detracts from the validity of
+    our results, especially in light of a paper by de Blasio et al. (2007)
+    which found sub-linear preferential attachment through a population-level
+    survey. We have added to the supplemental materials a series of power law
+    fits to the degree distributions of BA networks with $\alpha \neq 1$
+    (Figure S11), which shows that, although the power law is clearly an
+    imperfect fit, the slope is still reasonably representative of the true
+    distribution.
   \item We did some additional literature searching and found five more
@@
     et al.). The highest $\alpha$ values were associated with IDU networks,
-    coroborating the assertion of Dombrowski et al. (2013) that IDU networks
+    corroborating the assertion of Dombrowski et al. (2013) that IDU networks
     exhibit the scale-free property more strongly than sexual networks do. 
-  \item Three of these new datasets had sequenced two HIV genes, which we
-    analysed separately. We found that estimates of $\alpha$ were very robust
-    to the choice of gene; $I$ was less robust, with \textit{env} genes tending
-    to lead to higher estimated values.
-  \item Figure 5 (formerly Figure 4) now shows 50\% HPD intervals (similar to
-    interquartile ranges) in addition to 95\% HPD intervals.
+  \item Two of these new datasets had sequenced two different HIV genes
+    (\textit{gag} and \textit{env}), which we analysed separately. We found
+    that estimates of $\alpha$ were very robust to the choice of gene. $I$ was
+    less robust -- in one of the two datasets, the \textit{env} sequences led
+    to a noticably higher estimated value than the \textit{gag} sequences.
   \item We discovered that our use of the term ``kernel-ABC'' to refer to ABC 
@@
     to refer to our method.
+  \item Figure 5 (formerly Figure 4) now shows 50\% HPD intervals (similar to
+    interquartile ranges) in addition to 95\% HPD intervals.
   \item The discussion has been divided into subsections to make it easier to
@@
 
-We are grateful that the reviewer finds our work to be of some interest, and we
-acknowledge that our ability to precisely resolve numerical values of the BA
-parameters is limited. We believe that our simulation results show that some
-network parameters (specifically $\alpha$ in our case) can be at least roughly
-estimated from the information contained in tree shapes.
+We are grateful that the reviewer finds our work to be of some interest. Even
+in the relatively simple scenario we studied (susceptible-infected epidemics
+over Barab\'asi-Albert networks), the process leading from a network model to a
+transmission tree is complex. To calculate the likelihood of an observed
+transmission tree exactly, it would be necessary to sum over all possible
+contact networks the model could generate \emph{and} all possible epidemics
+over each contact network. It is therefore unsurprising that the posterior
+distributions are diffuse. In light of this, the level of parameter
+identifiability we observed is better than one might expect \textit{a priori}.
+We are confident that our parameter estimates, especially of $\alpha$, are
+truly representative of the posterior distributions and not simply artifacts of
+a poor ABC approximation. Several new analyses have been included to
+demonstrate this.
 
-We have included several new analyses that attempt to disambiguate true
-features of the posterior distribution from artefacts of a poor ABC
-approximation. First, we used ABC to estimate marginal posterior distributions
-when some combinations of parameters were held fixed to known values.
-Estimation error which is removed during marginal estimation is likely to be
-due to confounding between model parameters; error which remains may be due to
-either poor resolution or truely diffuse posteriors. The accuracy of the
-estimates was improved about 50\% in marginal estimation.
+First, we used ABC to estimate marginal posterior distributions when some
+combinations of parameters were held fixed to known values. Estimation error
+which is removed during marginal estimation is likely to be due to confounding
+between model parameters. The accuracy of the estimates was improved about 50\%
+in marginal estimation (Figure S8), implying that at least half of the observed
+inaccuracy stems from parameter correlations in the posterior. Although we have
+no way to prove it, it seems highly likely that even the marginal posterior
+distributions are diffuse given the complexity of the simulation process noted
+above. If that is the case, even more of the error may be attributed to true
+posterior features.
 
 Second, we examined the possible Monte Carlo error by increasing the resolution
-of the ABC approximation (more particles, more simulated datasets per particle,
-and more iterations). The results were nearly identical, TODO
+of the ABC approximation (twice as many particles, twice as many simulated
+datasets per particle, and roughly 1.5 times as many iterations). The results
+were comparable (Figure S9), which demonstrates that Monte Carlo error from the
+SMC algorithm is likely not responsible for observed estimation errors.
 
-Finally, three of the newly included HIV datasets included both \textit{gag}
+Finally, two of the newly included HIV datasets included both \textit{gag}
 and \textit{env} sequences which we analysed separately. The estimates of
-$\alpha$ for the two genes were nearly identical for each of the datasets,
-which seems unlikely if the ABC approximations were very poor.
+$\alpha$ for the two genes were very similar for each of the datasets (Figure
+S12), which seems unlikely if the ABC approximations were very poor. We also
+reanalyzed two of the datasets with a different phylogenetic reconstruction
+method, and again found very similar estimates of $\alpha$ (Figure S13). The
+estimates of $I$ were less robust to both of these variables, but remained in
+same sub-2000 region of the range in both cases. In the revised discussio, we
+acknowledge that $I$ was overestimated in simulations and offer the possible
+explanation that $I$ is only identifiable up to the epidemic phase it
+represents (\textit{i.e.} exponential growth vs. linear expansion vs. slow
+growth, Figure S15).
 
@@
 
-We apologise for this misunderstanding due to our poor choice of words. An SVR
+We apologize for this misunderstanding due to our poor choice of words. An SVR
 is in fact a regressor, not a classifier, so the reported $R^2$ were of true
@@
 accordingly. Some additional variation among meta-parameters and epidemic
-scenarios is observed, but the overall results are similar and the best choice
-of meta-parameters is the same.
+scenarios is observed, but the overall results are very similar.
 
@@
 clarifications in the methods section: (1) ``Duplicated sequences were removed
-with BioPython'' (2) ``Due to the removal of duplicated sequences, all
+with BioPython''; and (2) ``Due to the removal of duplicated sequences, all
 estimated transmission trees were fully binary.''
@@
 
-  TODO
+We have substantially expanded the discussion of the synthetic data results,
+including this point about $m$. The weaker identifiability in lower $\alpha$
+values may be related to the degree distributions of the networks, and the
+upward biases in $I$ and $N$ are discussed in relation to the epidemic
+trajectory of the SI model. Our intuition about $m$ is weaker, but we include
+this point by the reviewer and hypothesize that, as reviewer 2 suggests, a
+model which allows heterogeneity in $m$ may offer better identifiability.
 
@@
 We agree with this assessment of our visual presentation of the results. We
-have reduced the number of posterior plots shown to one in the text and two in
-the supplemental materials. Figure 2, which formerly represented each
-simulation by an individual point, has been replaced by a box plot showing the
-$\alpha$ and $I$ estimates stratified by each of the other parameters. A
-similar boxplot for $m$ and $N$, which were not identifiable with ABC, has been
-added to the supplemental materials.
+now only show a single set of marginal posterior distributions in the text.
+Figure 2, which formerly represented each simulation by an individual point,
+has been replaced by a box plot showing the $\alpha$ and $I$ estimates
+stratified by each of the other parameters. A similar boxplot for $m$ and $N$,
+which were non- and weakly-identifiable respectively with ABC, has been added
+to the supplemental materials (Figure S7).
 
@@
 
-  The Discussion focusses around model mis-specification, which is very
+  The Discussion focuses around model mis-specification, which is very
   relevant (and to which the inference isn't that robust, suggesting that it
@@
 to the kind of inference we are trying to perform. Indeed, the objective of the 
-analysis with classifiers was to investigate identifiability, and we apologise
+analysis with classifiers was to investigate identifiability, and we apologize
 for not making this clearer. The reviewer's intuition is correct -- we do
@@
 
-  L45-46: “…and in epidemics of differing size (500, 1000 and 2000)”. Unless
-  I’m missing something this is just a restatement of what’s already in this
-  paragraph about I, hence repetitive. 
+  L45-46: ``\ldots and in epidemics of differing size (500, 1000 and 2000)''.
+  Unless I'm missing something this is just a restatement of what's already in
+  this paragraph about I, hence repetitive. 
 \end{quote}
 
-We apologise for the misunderstanding. Because $I$ is not a structural
+We apologize for the misunderstanding. Because $I$ is not a structural
 parameter of the network, we attempted to classify each of $\alpha$, $N$, and
@@
   P7 \\
-  L20-24: It’s slightly surprising to me that the time trees were constructed
-  using “fast” methods when the datasets are not especially large. Were there
+  L20-24: It's slightly surprising to me that the time trees were constructed
+  using ``fast'' methods when the datasets are not especially large. Were there
   problems in obtaining trees with other algorithms? While I do not insist that
@@
 regression does not result in large errors in phylodynamic inference (Poon
-2015). We reanalysed the datasets with the highest and lowest estimated
-$\alpha$ (Zetterberg et al. 2004 and Cuevas et al. 2009) using a BEAST MCC tree
-from a run with simple models of evolution and population size. The results
-were TODO
+2015). We reanalysed two datasets (Niculescu et al. 2015 and Cuevas et al.
+2009) using RAxML and LSD. The results for $\alpha$ were nearly identical, but
+$I$ was not robust to the reconstruction method -- RAxML+LSD produced higher
+$I$ estimates for both datasets (Figure S13). This result has been added to the
+discussion of both parameters.
 
-With respect to the suggestion to analyse bootstrap replicates, we believe that
-the three new datasets containing both \textit{gag} and \textit{env} data fill
-the need for within-dataset replication.
+Rather than analyze bootstrap replicates, we used the two new datasets
+containing both \textit{gag} and \textit{env} sequences for within-dataset
+replication.
 
@@
 
-  L32: “Estimated correctly” – i.e. the posterior mode was correct?
+  L32: ``Estimated correctly'' -- i.e. the posterior mode was correct?
 \end{quote}
@@
   P13 \\
-  Figure 3, legend: I’m guessing that the shaded polygon is the HPD region,
-  rather than the “true value”?
+  Figure 3, legend: I'm guessing that the shaded polygon is the HPD region,
+  rather than the ``true value''?
 \end{quote}
@@
   P15 \\
-  L38-40: While I understand the point, it’s surely quite unlikely that the
+  L38-40: While I understand the point, it's surely quite unlikely that the
   network will be identical. Better to say that the number of infected nodes
@@
 We struggled with the phrasing of this particular point and are happy to have
-recieved some guidance. TODO
+received some guidance. The statement now reads as follows: ``If $I$ is small
+relative to $N$, very few of the infected nodes will gain any new neighbours.
+Thus, the outcome of a second simulation on the same network will likely be
+very similar.''
 
@@
 
-  L41-42: “We note also that our accurate estimates of I may have been
-  influenced by this prior”. Couldn’t you check?
+  L41-42: ``We note also that our accurate estimates of I may have been
+  influenced by this prior''. Couldn't you check?
 \end{quote}
 
-This issue was also addressed by performing marginal estimation. If the
-accurate estimates of $I$ were solely due to the prior, they would have gotten
-worse, not better, when $N$ was fixed and the prior on $I$ was truly uniform.
-We make this point in the discussion: ``when $I$ was estimated marginally with
-fixed $N$, the accuarcy of the estimate improved even though there was no
-longer any extra prior mass on low $I$ values.''
+This issue was addressed by performing marginal estimation. If the accurate
+estimates of $I$ were solely due to the prior, they would have gotten worse,
+not better, when $N$ was fixed and the prior on $I$ was truly uniform. We make
+this point in the discussion: ``when $I$ was estimated marginally with fixed
+$N$, the accuracy of the estimate improved even though there was no longer any
+extra prior mass on low $I$ values.''
 
@@
 
-We had been using the term ``prevalence'' interchangably with $I$, but since
+We had been using the term ``prevalence'' interchangeably with $I$, but since
 this is ambiguous we have replaced all instances of ``prevalence'' with $I$
@@
 
-  A point I’d like to see at least discussed – presumably if some iterations
-  add nodes with one connection and others add them with two, you’d get cycles,
-  but not as many as you’d expect with m=2. Is this worth considering in the
+  A point I'd like to see at least discussed -- presumably if some iterations
+  add nodes with one connection and others add them with two, you'd get cycles,
+  but not as many as you'd expect with m=2. Is this worth considering in the
   question of why the binary choice to include m=1 is so problematic? Would it
@@
 perform a small experiment allowing heterogeneity among nodes, but in $\alpha$
-rather than in $m$, and recovered the average of the two $\alpha$ values.
-
-(TODO: We
-have added to the discussion the point that this binary choice is sub-optimal
-and suggested this enhancement for future work.)
+rather than in $m$, and recovered the average of the two $\alpha$ values. We
+have outlined the difficulties of the binary choice of $m$ in the discussion of
+the real data results: ``\ldots the BA model does not allow any nodes with
+degree 1 when $m \geq 2$. For this reason, the choice of whether to allow $m =
+1$ in the prior is problematic, as we must choose between an unrealistic
+topology (no cycles) and an unrealistic minimum degree.''
 
 \begin{quote}
-  Supplement:
+  \itshape
+
+  Supplement: \\
   Figure S1 legend: What is the shaded area?

### 2016-07-06 doc/paper/paper.Rnw
Add paper and tex

@@
 could in principle be used for inference if observed incidence trajectories
-were available. \textcite{brown2011transmission} analysed the degree distribution
+were available. \textcite{brown2011transmission} analyzed the degree distribution
 of an approximate transmission network, estimated based on genetic similarity
@@
 find that these parameters can vary considerably between real epidemics from
-different settings. \add{TODO: something about sub-linear PA and IDU}
+different settings.
 
@@
 \add{Before proceeding with a full validation of \textit{netabc} on simulated
-data, we undertook two experiments designed to assess the identifiability of
-the BA parameters. These experiments only investigated one parameter of the
-BA model at a time while holding all others fixed, a strategy commonly used
-when performing sensitivity analyses of mathematical models. This allowed us to
-perform a fast preliminary analysis without dealing with the ``curse of
-dimensionality'' of the full parameter space. We simulated trees under three
-different values of each parameter, and asked how well we could tell the
-different trees apart. The better we are able to distinguish the trees, the
-more identifiability we might expect for the corresponding parameter when we
-attempt to estimate it with ABC.}
+data, we undertook an experiment designed to assess the identifiability of the
+BA parameters. One parameter of the BA model was investigated at a time while
+holding all others fixed, a strategy commonly used when performing sensitivity
+analyses of mathematical models. This allowed us to perform a fast preliminary
+analysis without dealing with the ``curse of dimensionality'' of the full
+parameter space. We simulated trees under three different values of each
+parameter, and asked how well we could tell the different trees apart. The
+better we are able to distinguish the trees, the more identifiability we might
+expect for the corresponding parameter when we attempt to estimate it with
+ABC.}
 
@@
 we restricted our analysis to the subtype B sequences and labelled this dataset
-as MSM. Three datasets (IDU/Estonia, HET/Uganda, and HET/Malawi) included both
-\textit{env} and \textit{gag} sequences. Each gene was analyzed separately to
-assess the robustness of \textit{netabc} to the particular HIV gene sequence
-used to estimate a transmission tree.}
+as MSM. Two datasets (HET/Uganda and HET/Malawi) included both \textit{env}
+and \textit{gag} sequences. Each gene was analyzed separately to assess the
+robustness of \textit{netabc} to the particular HIV gene sequence used to
+estimate a transmission tree. The IDU/Estonia data also sequenced both genes, 
+but the highly variable coverage and high homology of the \textit{gag}
+sequences made it impossible to obtain a sufficiently large block of
+non-identical sequences to analyze. Therefore, we analyzed only \textit{env}
+for this dataset.}
 
@@
 \textit{Seaview} version 4.4.2 \autocite{gouy2010seaview}. \add{Duplicated
-sequences were removed with BioPython.} Phylogenies were constructed from the
-nucleotide alignments by approximate maximum likelihood using
+sequences were removed with \textit{BioPython}.} Phylogenies were constructed
+from the nucleotide alignments by approximate maximum likelihood using
 \textit{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
@@
 trees were estimated by rooting and time-scaling the phylogenies by root-to-tip
-regression, using a modified version of Path-O-Gen (distributed as part of
-BEAST~\autocite{drummond2007beast}) as described
+regression, using a modified version of \textit{Path-O-Gen} (distributed as
+part of \textit{BEAST}~\autocite{drummond2007beast}) as described
 previously~\autocite{poon2015phylodynamic}. \add{Due to the removal of
@@
 
-\add{To check that our results were robust to the choice of phylogenetic
-method, we constructed phylogenies for two of the datasets (mixed/Spain and
-IDU/Estonia) with BEAST~\autocite{drummond2007beast}. We used the TN93 model of
-sequence evolution~\autocite{tamura1993estimation}, a strict molecular clock,
-and constant population size. The chains were run for $10^7$ steps and sampled
-every $10^3$ steps. Maximum clade credibility (MCC) trees were constructed from
-the sampled trees after discarding the first 10\% as burn-in.}
+\add{To check if our results were robust to the choice of phylogenetic
+reconstruction method, we built and reanalyzed phylogenies for two of the
+datasets (mixed/Spain and IDU/Romania) with
+\textit{RAxML}~\autocite{stamatakis2014raxml} with the GTR+$\Gamma$ model of
+sequence evolution and rate heterogeneity. The trees were rooted and
+time-scaled with \textit{Least Square
+Dating}~\autocite[\textit{LSD},][]{to2016fast}.}
 
@@
 cluster were carried forward. For the HET/Uganda, HET/Botswana, and MSM/USA
-data, no large clusters were detected using the same cutoff, so we analysed
+data, no large clusters were detected using the same cutoff, so we analyzed
 \del{a subtree} subsets of sizes 255, 180, and 180 respectively. \add{The
@@
         } else if (param == 'm') {
-            d[,list(rsquared=mean(rsquared)), by=c('nsimnode', 'ntip')]
+            d[,list(rsquared=mean(rsquared), accuracy=mean(accuracy)), by=c('nsimnode', 'ntip')]
         } else if (param == 'I') {
@@
         } else {
-            d[,list(rsquared=mean(rsquared)), by=c('I', 'ntip')]
+            d[,list(rsquared=mean(rsquared), accuracy=mean(accuracy)), by=c('I', 'ntip')]
         }
@@
     kernel.I <- get.kernel.data('I', 'classifier')
-    kernel.N <- get.kernel.data('N', 'classifier')
+    kernel.N <- get.kernel.data('N.bk.2', 'classifier')
 
@@
     sackin.I <- get.kernel.data('I', 'stats-classifier')
-    sackin.N <- get.kernel.data('N', 'stats-classifier')
+    sackin.N <- get.kernel.data('N.bk.2', 'stats-classifier')
 
@@
     nltt.I <- get.kernel.data('I', 'nltt-classifier')
-    nltt.N <- get.kernel.data('N', 'nltt-classifier')
+    nltt.N <- get.kernel.data('N.bk.2', 'nltt-classifier')
 @
@@
 $\sigma$ values, which require nearly-exact matches between branch lengths,
-resulted in low accuracy in some cases (\textit{e.g.} \cref{fig:alphacrossvn},
+resulted in low accuracy in some cases (\textit{e.g.} \cref{fig:alphacrossv},
 center row).}
@@
         Cross-validation accuracy of kernel-SVM, nLTT-based SVM, and Sackin's
-        index regression classifiers for BA model parameters.
+        index SVM classifiers for BA model parameters.
     ]{
@@
 different tree and epidemic sizes.} No classifier could accurately identify $m$
-in any epidemic scenario, with average $R^2$ values of 
-  \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for kSVM,
-  \Sexpr{nltt.m[,round(mean(rsquared), 2)]} for the nLTT, and
-  \Sexpr{sackin.m[,round(mean(rsquared), 2)]}
+in any epidemic scenario, with average accuracy values of 
+  \Sexpr{kernel.m[,round(mean(accuracy), 2)]} for kSVM,
+  \Sexpr{nltt.m[,round(mean(accuracy), 2)]} for the nLTT, and
+  \Sexpr{sackin.m[,round(mean(accuracy), 2)]}
 for Sackin's index. There was little variation in accuracy between epidemic
@@
 Finally, the performance of classifiers for $N$ depended heavily on the
-epidemic scenario. The $R^2$ of the kSVM classifier ranged from
-  \Sexpr{kernel.N[,round(min(rsquared), 2)]}
+epidemic scenario. The accuracy of the kSVM classifier ranged from
+  \Sexpr{kernel.N[,round(min(accuracy), 2)]}
 for the smallest epidemic and smallest sample size, to
-  \Sexpr{kernel.N[,round(max(rsquared), 2)]}
-for the largest. Likewise, $R^2$ for the nLTT-based SVM ranged from 
-  \Sexpr{nltt.N[,round(min(rsquared), 2)]}
+  \Sexpr{kernel.N[,round(max(accuracy), 2)]}
+for the largest. Accuracy for the nLTT ranged from 
+  \Sexpr{nltt.N[,round(min(accuracy), 2)]}
 to
-  \Sexpr{nltt.N[,round(max(rsquared), 2)]}.
+  \Sexpr{nltt.N[,round(max(accuracy), 2)]}.
 Sackin's index did not accurately classify $N$ in any scenario, with an average
-$R^2$ of
-  \Sexpr{sackin.N[,round(mean(rsquared), 2)]}
+accuracy of
+  \Sexpr{sackin.N[,round(mean(accuracy), 2)]}
 and little variation between scenarios.
@@
     }
-    \label{fig:abcptm2}
+    \label{fig:abcpt}
 \end{figure}
@@
 
-\add{To analyze the effects of the true parameter values on the
-accuracy our estimates of $\alpha$ and $I$, we fitted one GLM for each of these
-two parameters, with error rate as the dependent variable and the true
-parameter values as independent variables. Since the estimates of $m$ and
-$N$ were roughly equal across all simulations (\cref{fig:abcpt2}), GLMs were
-not fitted for these parameters. The estimated coefficients are shown in
-\cref{tab:glm}.}
-
 <<abc_glm, include=FALSE>>=
@@
 
-\add{The GLM analysis indicated that the error in estimates of $\alpha$
-decreased with larger true values of $\alpha$
+\add{To analyze the effects of the true parameter values on the
+accuracy our estimates of $\alpha$ and $I$, we fitted one GLM for each of these
+two parameters, with error rate as the dependent variable and the true
+parameter values as independent variables. Since the estimates of $m$ and
+$N$ were roughly equal across all simulations (\cref{fig:abcpt2}), GLMs were
+not fitted for these parameters. The estimated coefficients are shown in
+\cref{tab:glm}. The GLM analysis indicated that the error in estimates of
+$\alpha$ decreased with larger true values of $\alpha$
     ($p \Sexpr{pp(alpha.glm[Parameter == "alpha", p], eq=TRUE)}$)
@@
 but was not significantly affected by $I$. Qualitatively, $\alpha$ seemed to be
-only weakly identifiable between the values of 0 and 0.5 (\cref{fig:abcptm2}).
+only weakly identifiable between the values of 0 and 0.5 (\cref{fig:abcpt}).
 The error in the estimated $I$ value was slightly lower for smaller values of
@@
     \centering
-    \includegraphics[width=0.6\textwidth]{{\Sexpr{file.path("abc-posterior", middle.plot)}}.pdf}
+    \includegraphics[width=0.7\textwidth]{{\Sexpr{file.path("abc-posterior", middle.plot)}}.pdf}
     \caption[
@@
     \centering
-    %\includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior-2d", middle.plot)}}.pdf}
+    \includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior-2d", middle.plot)}}.pdf}
     \caption[
@@
    \Sexpr{d["N", round(hpd.max)]}].
-The approximate one-dimensional marginal posterior distributions for this
-simulation are shown in \cref{fig:mixed}. To test the effect of sampling bias,
+\del{The approximate one-dimensional marginal posterior distributions for this
+simulation are shown in \cref{fig:mixed}.} To test the effect of sampling bias,
 we sampled one transmission tree in a peer-driven fashion, where the
@@
    \Sexpr{p["N", floor(hpd.max)]}].
-The approximate one-dimensional marginal posterior distributions are shown in
-\cref{fig:peerdriven}. Both of these results were in line with estimates
+\del{The approximate one-dimensional marginal posterior distributions are shown
+in \cref{fig:peerdriven}.} Both of these results were in line with estimates
 obtained on other simulated datasets (\cref{tab:abchpd}), although the estimate
-of peer-driven sampling for $\alpha$ was somewhat lower than typical.
+of $\alpha$ for peer-driven sampling was somewhat lower than typical.
 
@@
     \Sexpr{ round((joint.N - marginal.N) * 100 / joint.N) }\%
-when all other parameters were fixed. \Cref{fig:moresmc} shows the impact of
+when all other parameters were fixed. \Cref{fig:better} shows the impact of
 increasing the number of particles, simulated datasets, and
 $\alpha_{\text{ESS}}$ parameter on the accuracy of a single simulation. The
-results of the two simulations were similar, but surprisingly, the results with
-higher SMC settings were slightly worse (by
-    \Sexpr{ round((better.alpha - base.alpha) * 100 / better.alpha) }\%,
-    \Sexpr{ round((better.I - base.I) * 100 / better.I) }\%, and
-    \Sexpr{ round((better.N - base.N) * 100 / better.N) }\%
-for $\alpha$, $I$, and $N$ respectively). However, the 50\% HPD interval for
-$I$ was slightly closer to the true value of 2000 with the improved settings
-    (\Sexpr{ d[nparticle == 2000, round(I.lower.50)] } - 
-     \Sexpr{ d[nparticle == 2000, round(I.upper.50)] }, vs.
-     \Sexpr{ d[nparticle == 1000, round(I.lower.50)] } - 
-     \Sexpr{ d[nparticle == 1000, round(I.upper.50)] } with basic settings).
-The estimate of $m$, 3 in both cases, was unaffected by the settings.}
-
-\subsection*{Real data}
+number of iterations until the stopping condition was reached was 81 with the
+basic settings and 124 with the higher settings. The results of the two
+simulations were very similar for all parameters.}
+
+\subsection*{Published HIV data}
 
@@
   m <- collect.metadata("../../simulations/aggregates/hpd/*")
-  d1 <- fread(rownames(m)[m$m_min == 1 & m$agg_gene == "gag"])
-  d2 <- fread(rownames(m)[m$m_min == 2 & m$agg_gene == "gag"])
-
-  d1.env <- fread(rownames(m)[m$m_min == 1 & m$agg_gene == "env"])
-  d2.env <- fread(rownames(m)[m$m_min == 2 & m$agg_gene == "env"])
+  d1 <- fread(rownames(m)[m$m_min == 1 & m$agg_gene == "gag"])[dataset != "bctree"]
+  d2 <- fread(rownames(m)[m$m_min == 2 & m$agg_gene == "gag"])[dataset != "bctree"]
+
+  d1.env <- fread(rownames(m)[m$m_min == 1 & m$agg_gene == "env"])[dataset != "bctree"]
+  d2.env <- fread(rownames(m)[m$m_min == 2 & m$agg_gene == "env"])[dataset != "bctree"]
 
@@
 
-  stopifnot(pmean1[,dataset[which.min(I)]] == "zetterberg2004")
+  stopifnot(pmean1[,dataset[which.min(I)]] == "kao2011")
   stopifnot(pmean1[,dataset[which.max(I)]] == "novitsky2014")
-  stopifnot(pmean1[dataset != "novitsky2014", dataset[which.max(I)]] == "bctree")
+  stopifnot(pmean1[dataset != "novitsky2014", dataset[which.max(I)]] == "li2015")
 
@@
 
-\del{We applied ABC to five published HIV datasets (\cref{tab:data}),
+We applied ABC to five published HIV datasets (\cref{tab:data}),
 and found substantial heterogeneity among the parameter estimates
-(\cref{fig:abchpd,fig:abchpdm2}).} \add{Posterior mean point estimates and 50\%
+(\cref{fig:abchpd,fig:abchpdm2}). \add{Posterior mean point estimates and 50\%
 and 95\% HPD intervals for each parameter are shown in \cref{fig:abchpd}.}
@@
 were all sub-linear, ranging from 
+    \Sexpr{pmean1["cuevas2009", round(alpha, 2)]}
+(mixed/Spain) to
     \Sexpr{pmean1["zetterberg2004", round(alpha, 2)]}
-for the IDU/Estonia data to
-    \Sexpr{pmean1["cuevas2009", round(alpha, 2)]}
-for the mixed/Spain data. When aggregated by risk group, the average estimates
-were
-    \Sexpr{pmean1[c("zetterberg2004", "niculescu2015", "bctree"), round(mean(alpha), 2)]}
+(IDU/Estonia). When aggregated by risk group, the average estimates were
+    \Sexpr{pmean1[c("zetterberg2004", "niculescu2015"), round(mean(alpha), 2)]}
 for IDU,
@@
     \Sexpr{pmean1[c("wang2015", "kao2011", "little2014", "li2015"), round(mean(alpha), 2)]}
-for MSM. These values were obtained with \textit{gag} for the datasets where
-both \textit{gag} and \textit{env} were sequenced, but the estimates for
-$\alpha$ did not change appreciably between the two genes (\cref{fig:genes}).
-There was a large amount of uncertainty associated with these estimates. 95\%
-HPD intervals were very wide for most datasets, often encompassing almost the
-entire range from 0 to 1 (\cref{fig:abchpd}).}
+for MSM. 95\% HPD intervals were very wide for most datasets, often
+encompassing nearly the entire range from 0 to 1 (\cref{fig:abchpd}).}
 
@@
 <<genetrees, include=FALSE>>=
-    datasets <- c("grabowski2014", "zetterberg2004", "mccormack2002")
+    datasets <- c("grabowski2014", "mccormack2002")
     paths <- paste0("../../simulations/", datasets, "/subtree/*")
@@
     ratios <- data.table(dataset=rep(datasets, each=2),
-                         gene=rep(c("env", "gag"), 3),
+                         gene=rep(c("env", "gag"), 2),
                          ratio=sapply(trees, get.ratio))
-    stopifnot(ratios[gene == "env", mean(ratio)] <
-              ratios[gene == "gag", mean(ratio)])
+    setkey(ratios, dataset)
+    stopifnot(pmean1[datasets, round(mean(I))] < pmean1.env[datasets, round(mean(I))])
 @
@@
 $I$ were between
-    \Sexpr{pmean1["zetterberg2004", round(I)]}
-(IDU/Estonia) and
-    \Sexpr{pmean1["bctree", round(I)]}
-(IDU/Canada). The HET/Botswana data had a much higher estimated $I$ value
+    \Sexpr{pmean1["kao2011", round(I)]}
+(MSM/Taiwan) and
+    \Sexpr{pmean1["li2015", round(I)]}
+(MSM/Shanghai). The HET/Botswana data had a much higher estimated $I$ value
     (\Sexpr{pmean1["novitsky2014", round(I)]})
@@
 density. When both \textit{gag} and \textit{env} sequences were analyzed, the
-estimates from the \textit{env} data were substantially higher
-    (average \Sexpr{pmean1[datasets, round(mean(I))]} for \textit{gag} vs.
-     \Sexpr{pmean1.env[datasets, round(mean(I))]} for \textit{env},
-\cref{fig:genes}). The ratios of internal to terminal branch lengths were
-lower for the \textit{env} trees than the \textit{gag} trees (average 
-    \Sexpr{ratios[gene == "gag", round(mean(ratio), 2)]}
-for \textit{gag} and 
-    \Sexpr{ratios[gene == "env", round(mean(ratio), 2)]}
-for \textit{env}).}
+estimates from the \textit{env} data were higher (HET/Uganda,
+    \Sexpr{pmean1["grabowski2014", round(I)]} for \textit{gag} vs.
+    \Sexpr{pmean1.env["grabowski2014", round(I)]} for \textit{env};
+HET/Malawi,
+    \Sexpr{pmean1["mccormack2002", round(I)]} for \textit{gag} vs.
+    \Sexpr{pmean1.env["mccormack2002", round(I)]} for \textit{env}).}
 
@@
 or above
-    \Sexpr{pmean2[c("zetterberg2004", "niculescu2015", "bctree"), round(min(alpha), 2)]}.
+    \Sexpr{pmean2[c("zetterberg2004", "niculescu2015"), round(min(alpha), 2)]}.
 The estimate for the MSM/Beijing data was slightly lower
@@
     \Sexpr{round(mean(up2[,I] - low2[,I]))}
-for $m \geq 2$). When $m = 1$ was disallowed, the posterior means of $m$ for
-all but the HET/Botswana data took on the value 3, with the HPD intervals
-spanning the entire prior region. This is very similar to the results observed
-for $m$ on simulated data (\cref{tab:abchpd}), and suggests that $m$ is not
-identifiable from these data with this prior. On the other hand, the estimate
-and 95\% HPD interval for the HET/Botswana data (4 [2-5]) was the same for both
-priors. The results for $N$ did not change appreciably between the two choices
-of prior.}
+for $m \geq 2$). The posterior means of $m$ for all but the HET/Botswana data
+took on the value 3 with this prior, with the HPD intervals spanning the entire
+prior region. This is very similar to the results observed for $m$ on simulated
+data (\cref{tab:abchpd}), and suggests that $m$ is not identifiable from these
+data with this prior. The results for $N$ did not change appreciably between
+the two choices of prior.}
 
@@
 shape in the range of values we considered. Although the tree kernel was the
-most effective classifier for $alpha$, a Sackin's index of tree imbalance
+most effective classifier for $\alpha$, a Sackin's index of tree imbalance
 performed nearly as well (\cref{fig:rsquared}). High $\alpha$ values produce
@@
 \add{$I$, representing the number of infected individuals at the time of
-sampling, was also identifiable, albeit slightly over-estimated with ABC for
-both values we considered. Sackin's index was better able to discern $I$ from
-tree shape than the nLTT (\cref{fig:rsquared,fig:Icrossv}), suggesting that
-this parameter impacts the distribution of branching times in the tree more
-than the topology. In a homogeneously-mixed population, branching times can be
-modelled by the coalescent process~\autocite{kingman1982coalescent}, in our
-case under the SI model~\autocite{volz2009phylodynamics}. Although networks are
-not homogeneously mixed, the forces which affect the distributions of branching
-times still apply. In our simulations, all discordant edges shared the same
+sampling, was also identifiable, albeit over-estimated with ABC for both values
+we considered. Sackin's index was better able to discern $I$ from tree shape
+than the nLTT (\cref{fig:rsquared,fig:Icrossv}), suggesting that this parameter
+impacts the distribution of branching times in the tree more than the topology.
+In a homogeneously-mixed population, branching times can be modelled by the
+coalescent process~\autocite{kingman1982coalescent}, in our case under the SI
+model~\autocite{volz2009phylodynamics}. Although networks are not homogeneously
+mixed, the forces which affect the distribution of branching times still apply.
+In our simulations, all discordant edges shared the same
 transmission rate, so that the waiting time until the next transmission in the
@@
 network per vertex, did not have a measurable impact on tree shape and was not
-identifiable with ABC. However, all the analyses presented here did not take
-the absolute size of the transmission trees into account, as the branch lengths
-were rescaled by their mean. Because higher $m$ values imply higher edge
-density, an epidemic should spread more quickly for higher $m$ than lower $m$
-with the same per-edge transmission probability. Hence, considering the
-absolute height of the trees may improve our method's ability to reconstruct
-$m$. It was also pointed out to us by an anonymous reviewer that for a fixed
-$I$, an infected node may only end up transmitting along a fraction of its
-outgoing edges, which could mask the presence of the extra edges associated
-with higher $m$.}
+identifiable with ABC. It was pointed out to us by an anonymous reviewer that
+for a fixed $I$, an infected node may only end up transmitting along a fraction
+of its outgoing edges, which could mask the presence of the extra edges
+associated with higher $m$. If $m$ were a continuous variable, it is possible
+that we would observe stronger identifiability for lower values (say between 0
+and 2), where extra edges are more likely to be involved in the epidemic and
+have an impact on the transmission tree. One way to achieve this would be to
+draw $m$ separately for each node from a distribution parameterized by a
+continuous variable.}
 
@@
 depending on whether or not $m = 1$ was allowed by the prior is further evidence
-of this potential misspecification. However, we note that for \del{two}\add{the
-majority} of the datasets, the estimated values of $\alpha$ did not change much
-between priors, and the estimates of $I$ were robust to the choice of prior for
-all datasets studied (compare \cref{fig:abchpd,fig:abchpdm2}). More
+of this potential misspecification. However, we note that \add{the ordering of
+the datasets with respect to $\alpha$ was similar between the two priors,}
+\del{for two of the datasets, the estimated values of $\alpha$ did not change
+much between priors,} and the estimates of $I$ were robust to the choice of
+prior for all datasets studied (compare \cref{fig:abchpd,fig:abchpdm2}). More
 sophisticated models, for example models incorporating heterogeneity in node
-behaviour, are likely to provide a better fit to these data. \add{For example,
-rather than using the same $m$ for all nodes, we could specify a distribution
-from which each individual node's $m$ value would be drawn.}
+behaviour, are likely to provide a better fit to these data.
 
@@
 \del{With respect to the preferential attachment power $\alpha$, the five
-datasets analysed fell into three categories (\cref{fig:abchpd}). First, we
+datasets analyzed fell into three categories (\cref{fig:abchpd}). First, we
 estimated a preferential attachment power close to 1, indicating linear
@@
 
-\subsubsection{Preferential attachment power is sub-linear and higher for IDU networks}
+\subsubsection*{Preferential attachment power is sub-linear and higher for IDU networks}
 
@@
 population-level longitudinal partner count data, they found $\alpha$ values
-ranging from 0.26 to 0.62 depending on the gender and time period considered.
-Our estimates of $\alpha$ for the IDU/Romania was above this range under both
-priors, as were the estimates for the MSM/Beijing data and the BC data when $m
-= 1$ was disallowed by the prior. The dataset investigated by
-\textcite{de2007preferential} was derived from a survey of a random sample of
-the Norwegian population, whereas our investigation focused on datasets from
-known phylogenetic or geographic clusters of HIV-infected persons. It is
-therefore unsurprising that we detected stronger preferential attachment
-dynamics in some cases. For instance, random sampling is much less likely to
-discover the high-degree nodes characterizing the tail of the degree
-distribution, simply because those individuals are rare in the general
-population. In addition, it is plausible that HIV-positive individuals are more
-likely to be highly connected in their sexual networks, as the odds of
-acquiring HIV increase with the number of unprotected sexual contacts.}
+ranging from 0.26 to 0.62 depending on the gender and time period considered.}
 
@@
 
-\add{For both choices of prior on $m$, the datasets derived from IDU
-populations had a higher estimated preferential attachment power than the other
-datasets (\cref{fig:abchpd,fig:abchpdm2}). This finding is in line with
+\add{The datasets derived from IDU populations had a higher estimated
+preferential attachment power than the other datasets
+(\cref{fig:abchpd,fig:abchpdm2}). This finding is in line with
 \textcite{dombrowski2013topological}, who reanalyzed a network of IDUs in
@@
 
-\subsubsection{Other BA parameters}
+\add{The results we have just discussed pertain to the \textit{gag} sequences
+for the two datasets with both \textit{gag} and \textit{env} data. As shown in
+\cref{fig:genes}, the estimates of $\alpha$ were quite robust to the gene
+analyzed. For the two datasets we reanalyzed using
+RAxML~\autocite{stamatakis2014raxml} and LSD~\autocite{to2016fast}, $\alpha$
+was robust to the choice of method (\cref{fig:methods}).}
+
+\subsubsection*{Other BA parameters}
+
+<<realdata_I, include=FALSE>>=
+    #stopifnot(d1[parameter == "I", dataset[which.min(mean)]] == "zetterberg2004")
+    stopifnot(d1[parameter == "I", dataset[which.max(mean)]] == "novitsky2014")
+@
 
@@
 datasets ranged from
-    X (D)
+    \Sexpr{d1[parameter == "I", round(min(mean))]}
+    (IDU/Estonia)
 to
-    X (D).}
+    \Sexpr{d1[parameter == "I", round(max(mean))]}
+    (HET/Botswana).
+The estimates of $I$ were less robust to the HIV gene analyzed
+(\cref{fig:genes}) and the phylogenetic reconstruction method used
+(\cref{fig:methods}).}
 
@@
 HET/Botswana data~\autocite{novitsky2013phylogenetic, novitsky2014impact}
-estimated that there were 1731 \gls{HIV}-positive individuals in the study
-area. \gls{HIV} sequences were obtained from approximately 70\% of these
-individuals. The estimated prevalence we obtained was much higher 
+estimated that there were 1731 HIV-positive individuals in the study
+area. HIV sequences were obtained from approximately 70\% of these individuals.
+The estimated prevalence we obtained was much higher 
     (\Sexpr{d1[dataset == "novitsky2014" & parameter == "I", round(mean)]}),
-with a 95\% \gls{HPD} interval spanning nearly the entire prior region. On the
-other hand, the study which produced the MSM/USA
-data~\autocite{little2014using} enrolled 648 HIV-positive MSM; thus, our
-estimate of $I$
+with a 95\% HPD interval spanning nearly the entire prior region. On the other
+hand, the study which produced the MSM/USA data~\autocite{little2014using}
+enrolled 648 HIV-positive MSM; thus, our estimate of $I$
     (\Sexpr{d1[dataset == "little2014" & parameter == "I", round(mean)]})
@@
 travel between the two locations may have facilitated linking of their sexual
-networks. Thus, the high estimate of \gls{I} we obtained for these data may
+networks. Thus, the high estimate of $I$ we obtained for these data may
 include a larger network component based in the capital city. The MSM/USA
@@
 
-\subsubsection{Modelling assumptions}
+\subsubsection*{Modelling assumptions}
 
@@
 said to apply to any individual node. The effect of biased sampling was
-investigated by analysing a transmission tree which had been sampled in a
+investigated by analyzing a transmission tree which had been sampled in a
 peer-driven fashion. The results were roughly in line with those for random
@@
 significantly~\autocite{ypma2013relating} due to within-host evolution and the
-sampling process. Progress has very recently been made on this front by
+sampling process. \add{Progress has very recently been made on this front by
 \textcite{giardina2016inference}, who have concurrently developed an ABC-based
-method incorporating within-host evolution. The ABC-SMC algorithm is
+method incorporating within-host evolution.} The ABC-SMC algorithm is
 computationally intensive, taking about a day when run on 20 cores in parallel
@@
 
-\printbibliography
-
 \clearpage
 \newpage
+\printbibliography
 
@@
     transmission trees over networks with $\alpha$ = 0.5, 1.0, or 1.5. Dotted
-    and and dashed lines indicate, respectively, performance of SVM using the
-    nLTT statistic, and linear regression using Sackin's index. Facets
-    are number of infected nodes before the simulation was stopped ($I$) and
-    number of tips in the sampled transmission tree.
+    and and dashed lines indicate, respectively, performance of SVMs using the
+    nLTT statistic and Sackin's index. Facets are number of infected nodes
+    before the simulation was stopped ($I$) and number of tips in the sampled
+    transmission tree.
   }
@@
       on 300 simulated transmission trees over networks with $m$ = 2, 3, or 4.
-      Dotted and and dashed lines indicate, respectively, performance of SVM
-      using the nLTT statistic, and linear regression using Sackin's
-      index. Facets are number of infected nodes before the simulation was
-      stopped ($I$) and number of tips in the sampled transmission tree.
+      Dotted and and dashed lines indicate, respectively, performance of SVMs
+      using the nLTT statistic and Sackin's index. Facets are number of
+      infected nodes before the simulation was stopped ($I$) and number of tips
+      in the sampled transmission tree.
   }
@@
       1000, or 2000. Dotted and and dashed lines indicate, respectively,
-      performance of SVM using the nLTT statistic, and linear regression
-      using Sackin's index. Facets are the number of tips in the sampled
-      transmission tree.
+      performance of SVMs using the nLTT statistic and Sackin's index. Facets
+      are the number of tips in the sampled transmission tree.
   }
@@
       5000, or 8000. Dotted and and dashed lines indicate, respectively,
-      performance of SVM using the nLTT statistic, and linear regression
-      using Sackin's index. Facets are the number of tips in the sampled
-      transmission tree.
+      performance of SVMs using the nLTT statistic and Sackin's index. Facets
+      are the number of tips in the sampled transmission tree.
   }
@@
     }
+    \label{fig:abcpt2}
 \end{figure}
@@
         Posterior means, 50\% HPD intervals, and 95\% HPD intervals for BA
-        model parameter estimated marginally with ABC.
+        model parameters estimated marginally with ABC.
     ]{
         Posterior means (points), 50\% HPD intervals (notches), and 95\% HPD
-        intervals (lines) for BA model parameter estimated marginally with ABC.
-        $x$-axis labels indicate parameters which were fixed to their true
+        intervals (lines) for BA model parameters estimated marginally with
+        ABC. $x$-axis labels indicate parameters which were fixed to their true
         values by specifying Dirac-delta priors.
@@
     \label{fig:marginal}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.7\textwidth]{better-smc}
+    \caption[
+        Posterior means, 50\% HPD intervals, and 95\% HPD intervals for BA
+        model parameters estimated with ABC using two sets of SMC settings.
+    ]{
+        Posterior means (points), 50\% HPD intervals (notches), and 95\% HPD
+        intervals (lines) for BA model parameter estimated with ABC using two
+        sets of SMC settings. ``Low'' settings are 1000 particles, 5 simulated
+        datasets per particle, and $\alpha_{\text{ESS}} = 0.95$. ``High''
+        settings are 2000 particles, 10 simulated datasets per particle, and
+        $\alpha_{\text{ESS}} = 0.97.$
+    }
+    \label{fig:better}
 \end{figure}
@@
     \centering
-    \includegraphics{mixed-posterior}
+    \includegraphics{powerlaw.pdf}
     \caption{
-        Approximate marginal posterior distributions of Barab\'asi-Albert
-        model parameters obtained using ABC for a network with
-        heterogeneous node behaviour. Half of the nodes were attached with
-        $\alpha = 0.5$, and the other half with $\alpha = 1.5$ (vertical
-        dashed lines, top right). Other parameter values were $m = 2$, $I =
-        1000$, and $N = 5000$ (vertical dashed lines, other than top right).
-        Shaded areas indicate 95\% highest posterior density intervals.
-    }
-    \label{fig:mixed}
+        Best fit power law and stretched exponential curves for degree
+        distributions of simulated BA networks for several values of $\alpha$
+        and $m$.
+    }
+    \label{fig:powerlaw}
 \end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics{peerdriven-posterior}
-    \caption{Approximate marginal posterior distributions of Barab\'asi-Albert
-        model parameters obtained using ABC for a network with
-        peer-driven sampling. An epidemic was simulated in the usual fashion,
-        but rather than being sampled at random, infected nodes were sampled
-        with a probability two times higher if they had any sampled neighbours
-        in the contact network. Vertical dashed lines indicate true parameter
-        values, and shaded areas indicate 95\% highest posterior density
-        intervals.
-    }
-    \label{fig:peerdriven}
-\end{figure}
-
-\clearpage
 
@@
         Posterior means and 50\%/95\% HPD intervals for parameters of the
-        BA network model, fitted to three HIV datasets where both \textit{gag}
+        BA network model, fitted to two HIV datasets where both \textit{gag}
         and \textit{env} genes were sequenced.
@@
         HPD intervals (lines) for parameters of the BA network model, fitted to
-        three HIV datasets where both \textit{gag} and \textit{env} genes were
+        two HIV datasets where both \textit{gag} and \textit{env} genes were
         sequenced.
@@
     \label{fig:genes}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{realdata-methods}
+    \caption[
+        Posterior means and 50\%/95\% HPD intervals for parameters of the
+        BA network model, fitted to two HIV datasets using two phylogenetic
+        reconstruction methods.
+    ]{
+        Posterior means (points), 50\% HPD intervals (notches), and 95\%
+        HPD intervals (lines) for parameters of the BA network model, fitted to
+        two HIV datasets using two phylogenetic reconstruction methods.
+    }
+    \label{fig:methods}
 \end{figure}
@@
     First and second \add{time} derivatives of epidemic growth curves at time
-    of sampling for various values of \gls{I} and \gls{N}. Networks were
-    simulated under the \gls{BA} model with \gls{alpha} = 1.0, \gls{m} = 2, and
-    \gls{N} varied along the values shown on the $x$-axis. Transmission trees
-    were sampled at the time when \gls{I} nodes were infected ($y$-axis).
-    Logistic growth curves were fit to epidemic trajectories derived from the
-    transmission trees, and their first and second derivatives were calculated
-    at the time of sampling. Contours show first derivatives, while colours
-    indicate second derivatives. Values of \gls{I} and \gls{N} used in
-    simulation experiments with \gls{ABC} are indicated by diamonds.
+    of sampling for various values of $I$ and $I$. Networks were simulated
+    under the BA model with $\alpha = 1.0$, $m = 2$, and $N$ varied along the
+    values shown on the $x$-axis. Transmission trees were sampled at the time
+    when $I$ nodes were infected ($y$-axis). Logistic growth curves were fit to
+    epidemic trajectories derived from the transmission trees, and their first
+    and second derivatives were calculated at the time of sampling. Contours
+    show first derivatives, while colours indicate second derivatives. Values
+    of $I$ and $N$ used in simulation experiments with ABC are indicated by
+    diamonds.
   }
@@
 
-%\begin{figure}[ht]
-%  \includegraphics{cuevas2009-posterior}
-%  \caption{
-%      Posterior distribution of BA model parameters for
-%      \textcite{cuevas2009hiv} data. Vertical lines indicate maximum \textit{a
-%      posteriori} estimates, and shaded areas are 95\% highest posterior
-%      density intervals. $x$-axis indicates regions of nonzero prior density.
-%  }
-%  \label{fig:cuevas}
-%\end{figure}
-%
-%\begin{figure}[ht]
-%  \includegraphics{li2015-posterior}
-%  \caption{
-%      Posterior distribution of BA model parameters for \textcite{li2015hiv}
-%      data. Vertical lines indicate maximum \textit{a posteriori} estimates,
-%      and shaded areas are 95\% highest posterior density intervals. $x$-axis
-%      indicates regions of nonzero prior density.
-%  }
-%  \label{fig:li}
-%\end{figure}
-%
-%\begin{figure}[ht]
-%  \includegraphics{niculescu2015-posterior}
-%  \caption{
-%      Posterior distribution of BA model parameters for
-%      \textcite{niculescu2015recent} data. Vertical lines indicate maximum
-%      \textit{a posteriori} estimates, and shaded areas are 95\% highest
-%      posterior density intervals. $x$-axis indicates regions of nonzero prior
-%      density.
-%  }
-%  \label{fig:niculescu}
-%\end{figure}
-%
-%\begin{figure}[ht]
-%  \includegraphics{novitsky2014-posterior}
-%  \caption{
-%      Posterior distribution of BA model parameters for
-%      \textcite{novitsky2013phylogenetic, novitsky2014impact} data. Vertical
-%      lines indicate maximum \textit{a posteriori} estimates, and shaded areas
-%      are 95\% highest posterior density intervals. $x$-axis indicates regions
-%      of nonzero prior density.
-%  }
-%  \label{fig:novitsky}
-%\end{figure}
-%
-%\begin{figure}[ht]
-%  \includegraphics{wang2015-posterior}
-%  \caption{
-%      Posterior distribution of BA model parameters for
-%      \textcite{wang2015targeting} data. Vertical lines indicate maximum
-%      \textit{a posteriori} estimates, and shaded areas are 95\% highest
-%      posterior density intervals. $x$-axis indicates regions of nonzero prior
-%      density.
-%  }
-%  \label{fig:wang}
-%\end{figure}
-%
-%<<abc_posterior, echo=FALSE, results="asis", eval=FALSE>>=
-%    N <- 5000
-%    replicate <- 0
-%    for (m in c(2, 3, 4)) {
-%    for (I in c(1000, 2000)) {
-%    for (alpha in c(0, 0.5, 1, 1.5)) {
-%        fn <- sprintf("{abc-posterior/%.1f_%d_%d_%d_%d}.pdf", alpha, I, m, N, replicate)
-%        lab <- sprintf("fig:%.1f-%d-%d-%d-%d", alpha, I, m, N, replicate)
-%        cat("\\begin{figure}[ht]\n")
-%        cat("\\centering\n")
-%        cat(paste0("\\includegraphics{", fn, "}\n"))
-%        cat("\\caption{\n")
-%        cat(sprintf("Approximate marginal posterior distributions of Barab\\'asi-Albert model parameters obtained using ABC for one simulated transmission tree with parameter values $\\alpha = %.1f$, $I = %d$, $m = %d$, and $N = %d$.
-Dashed lines indicate true values, and shaded areas show 95\\%% highest posterior density intervals.\n", alpha, I, m, N))
-%        cat("}\n")
-%        cat(paste0("\\label{", lab, "}\n"))
-%        cat("\\end{figure}\n")
-%        cat("\\clearpage\n\n")
-%    }
-%    }
-%    }
-%@
-
 \end{document}

### 2016-07-06 doc/paper/paper.tex
Add paper and tex

@@
 \usepackage[T1]{fontenc}
-\usepackage{garamondx}
-\usepackage[garamondx,cmbraces]{newtxmath}
-\usepackage[backend=biber]{biblatex}
+\usepackage{mathptmx}
+\usepackage[backend=biber, style=numeric-comp, sorting=none]{biblatex}
 \usepackage{fullpage}
@@
 \usepackage{cleveref}
+\usepackage[normalem]{ulem}
+\usepackage{color}
+\usepackage{amsmath}
 
 \addbibresource{papers.bib}
+
+% beginning of diff stuff
+\usepackage{twoopt}
+
+\let\mref\cref
+\let\Mref\Cref
+\let\mtextcite\textcite
+\let\mautocite\autocite
+\let\mciteauthor\citeauthor
+\renewcommand{\cref}[1]{\mbox{\mref{#1}}}
+\renewcommand{\Cref}[1]{\mbox{\Mref{#1}}}
+\renewcommandtwoopt{\textcite}[3][][]{\mbox{\mtextcite[#1][#2]{#3}}}
+\renewcommandtwoopt{\autocite}[3][][]{\mbox{\mautocite[#1][#2]{#3}}}
+\renewcommand{\citeauthor}[1]{\mbox{\mciteauthor{#1}}}
+
+\newcommand{\add}[1]{\color{blue} \uline{#1} \color{black}}
+\newcommand{\del}[1]{\color{red} \sout{#1} \color{black}}
+% end of diff stuff
+
+\newcommand{\tablepath}{../tables}
+\graphicspath{{../figures/}}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 
-\newcommand{\tablepath}{../tables}
-\graphicspath{{../figures/}}
-
 \title{Reconstructing contact network parameters from viral phylogenies}
-\author{Rosemary M. McCloskey, Richard H. Liang, and Art F.Y. Poon}
+\author{Rosemary M. McCloskey$^1$, Richard H. Liang$^1$, and Art F.Y. Poon$^{1,2}$ \\
+\small $^1$BC Centre for Excellence in HIV/AIDS, Vancouver, Canada \\ \small $^2$ Department of Medicine, University of British Columbia, Vancouver, Canada}
+
+\maketitle
 \abstract{\input{abstract}}
 
-\maketitle
-
-\section{Introduction}
+\section*{Introduction}
 
@@
 the speed and pattern of the epidemic's expansion. Network structure can
-influence the prevalence trajectory~\autocite{o2011contact} and epidemic
-threshold~\autocite{barthelemy2005dynamical}, in turn affecting the estimates of
-quantities such as effective population size~\autocite{goodreau2006assessing}.
-From a public health perspective, contact networks have been explored as tools
-for curtailing epidemic spread, by way of interventions targeted to
-well-connected nodes~\autocite{wang2015targeting}. True contact networks are a
-challenging type of data to collect, requiring extensive epidemiological
-investigation~\autocite{welch2011statistical}.
-
-Viral sequence data, on the other hand, has become relatively inexpensive and
-straightforward to collect on a population level. Due to the high mutation rate
-of RNA viruses, epidemiological processes impact the course of viral evolution,
-thereby shaping the inter-host viral
+influence the prevalence trajectory~\autocite{o2011contact, ma2013importance}
+and epidemic threshold~\autocite{barthelemy2005dynamical}, in turn affecting
+the estimates of quantities such as effective viral population
+size~\autocite{goodreau2006assessing}.  From a public health perspective,
+contact networks have been explored as tools for curtailing epidemic spread, by
+way of interventions targeted to well-connected
+nodes~\autocite{wang2015targeting}. True contact networks are a challenging
+type of data to collect, requiring extensive epidemiological
+investigation~\autocite{welch2011statistical, eames2015six}.
+
+Viral sequence data, on the other hand, has become \del{relatively inexpensive
+and straightforward to collect on a population level.} \add{easier to collect
+as the cost of sequencing has declined. In the case of HIV, genotyping has
+become part of routine clinical care in several health regions.} Due to the
+high mutation rate of RNA viruses, epidemiological processes impact the course
+of viral evolution, thereby shaping the inter-host viral
 phylogeny~\autocite{drummond2003measurably}. The term ``phylodynamics'' was
@@
 sophisticated methods have been developed for fitting complex population
-genetic models to phylogenies~\autocite{rasmussen2014phylodynamic}, inference
-of structural network parameters has to date been limited. However, it has been
-shown that network structure has a tangible impact on phylogeny
-shape~\autocite{leventhal2012inferring, colijn2014phylogenetic,
-goodreau2006assessing, robinson2013dynamics, villandre2016assessment},
-suggesting that such statistical inference might be
-possible~\autocite{welch2011statistical}.
-
-Survey-based studies of sexual networks~\autocite{liljeros2001web,
-schneeberger2004scale, colgate1989risk} have found that these networks tend to
-have a degree distribution which follows a power law \autocite[although there has
-been some disagreement, see][]{handcock2004likelihood}. That is, the number of
-nodes of degree $k$ is proportional to $k^{-\gamma}$ for some constant
-$\gamma$. These networks are also creferred to as ``scale-free''. One process by
-which scale-free networks can be generated is pcreferential attachment, where
-nodes with a high number of contacts attract new connections at an elevated
-rate. The first contact network model incorporating pcreferential attachment was
-introduced by \textcite{barabasi1999emergence}, and is now creferred to as the
+genetic models to phylogenies~\autocite{volz2012complex,
+rasmussen2014phylodynamic}, inference of structural network parameters has to
+date been limited. However, it has been shown that network structure has a
+tangible impact on phylogeny shape~\autocite{leventhal2012inferring,
+colijn2014phylogenetic, goodreau2006assessing, robinson2013dynamics,
+villandre2016assessment}, suggesting that such statistical inference might be
+possible~\autocite{welch2011statistical}. \add{In the context of networks,
+sequence data have the advantage of being objective, in that they are not
+affected by misreporting. However, just as with survey data, it is important to
+collect a representative sample from the population to perform accurate
+inference~\autocite{novitsky2014impact}.}
+
+Survey-based studies of sexual networks~\autocite{colgate1989risk,
+liljeros2001web, schneeberger2004scale, latora2006network, rothenberg2007large,
+clemenccon2015statistical} have found that these networks tend to have a degree
+distribution which follows a power law \autocite[although there has been some
+disagreement, see][]{handcock2004likelihood}. That is, the number of nodes of
+degree $k$ is proportional to $k^{-\gamma}$ for some constant $\gamma$. These
+networks are also referred to as
+``scale-free''~\autocite{barabasi1999emergence}. One process by which
+scale-free networks can be generated is preferential attachment, where nodes
+with a high number of contacts attract new connections at an elevated rate. The
+first contact network model incorporating preferential attachment was
+introduced by \textcite{barabasi1999emergence}, and is now referred to as the
 Barab\'asi-Albert (BA) model. Under this model, networks are formed by
@@
 where the probability of attaching to a node of degree $k$ is $k^\alpha$ for
-some non-negative constant $\alpha$, and we use this extension here.
+some non-negative constant $\alpha$, and we use this extension in this work.
+\add{When $\alpha \neq 1$, the degree distribution is no longer a power law:
+for $\alpha < 1$, the distribution is a stretched exponential, while for
+$\alpha > 1$, it is a ``gelation'' type distribution where one or a few hub
+nodes are connected to nearly every other node in the
+graph~\autocite{krapivsky2000connectivity}.}
 
@@
 function, $\hat{D}$ is a simulated dataset according to $\theta$, and
-$\varepsilon$ is a small tolerance~\autocite{sunnaaker2013approximate}. In the
-specific case when $\rho$ is a kernel function, the approach is known as
-kernel-ABC~\autocite{nakagome2013kernel, poon2015phylodynamic}.
-
-Here, we develop a method using kernel-ABC to estimate the parameters of
-contact network models from observed phylogenetic data. The distance function
-we use is the tree kernel developed by \textcite{poon2013mapping}, which computes
-a weighted dot product of the trees' representations in the space of all
-possible subset trees. We apply the method to investigate the parameters of the
-BA network model on a variety of simulated and real datasets. Our results show
+$\varepsilon$ is a small tolerance~\autocite{sunnaaker2013approximate}. \del{In
+the specific case when $\rho$ is a kernel function, the approach is known as
+ABC~\autocite{nakagome2013kernel, poon2015phylodynamic}.} \add{Our
+group~\autocite{poon2015phylodynamic} and others~\autocite{park2015k2} have
+demonstrated that taking $\rho$ to be a well-chosen kernel function can produce
+a more accurate ABC approximation than the typical choice of a difference of
+summary statistics.}
+
+Here, we develop a method using ABC to estimate the parameters of contact
+network models from observed phylogenetic data. The distance function we use is
+the tree kernel developed by \textcite{poon2013mapping}, which computes a
+weighted dot product of the trees' representations in the space of all possible
+subset trees. We apply the method to investigate the parameters of the BA
+network model on a variety of simulated and real datasets. Our results show
 that some network parameters can be inferred with reasonable accuracy, while
-others have a minimal detectable impact on tree shape and thecrefore cannot be
-estimated accurately. We also find that these parameters can vary considerably
-between real epidemics from different settings.
-
-\section{Methods}
-
-\subsection{\textit{Netabc}: phylogenetic inference of contact network
-parameters with kernel-ABC}
-
-We have developed a kernel-ABC-based method to perform statistical inference of
+others \del{have a minimal detectable impact on tree shape and therefore cannot
+be estimated accurately}\add{are weakly- or non-identifiable with ABC}. We also
+find that these parameters can vary considerably between real epidemics from
+different settings.
+
+\section*{Methods}
+
+\subsection*{\textit{Netabc}: phylogenetic inference of contact network
+parameters with ABC}
+
+We have developed an ABC-based method to perform statistical inference of
 contact network parameters from a transmission tree estimated from an observed
@@
 is assumed to approximate the desired posterior~\autocite{del2012adaptive,
-sunnaaker2013approximate}. A computer program implementing our method is freely
-available at \url{https://github.com/rmcclosk/netabc} (last accessed April 3,
-2016).
+sunnaaker2013approximate}.
+
+\add{In the original formulation of ABC-SMC~\autocite{sisson2007sequential,
+beaumont2009adaptive}, the user is required to specify a decreasing sequence of
+tolerances $\lbrace\varepsilon_i\rbrace$. At iteration $i$, particles with no
+associated simulated datasets within distance $\varepsilon_i$ of the observed
+data are removed from the population. In the adaptive version of
+\textcite{del2012adaptive}, the sequence of tolerances is determined
+automatically by fixing the decay rate of the population's expected sample size
+(ESS) to a user-defined value. \citeauthor{del2012adaptive} call this value
+$\alpha$, but we will refer to it here as $\alpha_{\text{ESS}}$ to avoid
+confusion with the preferential attachment power parameter of the BA model.}
 
@@
 graphs. The analyses were run on Westgrid (\url{https://www.westgrid.ca/}) and
-a local computer cluster.
-
-\subsection{Kernel classifiers}
-
-We used the phylogenetic kernel developed by \textcite{poon2013mapping} to test
-whether the parameters of the BA model had an effect on tree shape. 100
-networks were simulated under each of three different values of $\alpha$: 0.5,
-1.0, and 1.5 (300 networks total). The other parameters were fixed to the
-following values: $N$ = 5000, $I$ = 1000, and $m$ = 2. A transmission tree with
-500 tips was simulated over each network (300 transmission trees total). The
-300 trees were compared pairwise with the tree kernel to form a $300 \times
-300$ kernel matrix. The kernel meta-parameters $\lambda$ (the ``decay
-factor''), and $\sigma$ (the ``radial basis function
-variance'')~\autocite[see][]{poon2013mapping}, were set to 0.3 and 4 respectively.
-We constructed a kSVR classifier for $\alpha$ using the \textit{kernlab}
-package~\autocite{zeileis2004kernlab}, and evaluated its accuracy with 1000
-two-fold cross-validations.
+a local computer cluster. \add{A computer program implementing our method is
+freely available at \url{https://github.com/rmcclosk/netabc} (last accessed
+\today).}
+
+\subsection*{\del{Kernel classifiers}\add{Classifiers for BA model parameters from tree shapes}}
+
+\add{We considered four parameters related to the BA model, denoted $N$, $m$,
+$alpha$, and $I$. The first three of these parameterize the network structure,
+while $I$ is related to the simulation of transmission trees over the network.
+However, we will refer to all four as BA parameters. $N$ denotes the total
+number of nodes in the network, or equivalently, susceptible individuals in the
+population. $m$ is the number of new undirected edges added for each new
+vertex, or equivalently one-half of the average degree. $\alpha$ is the power
+of preferential attachment -- new nodes are attached to existing nodes of
+degree $d$ with probability proportional to $d^{\alpha} + 1$. Finally, $I$ is
+the number of infected individuals at the time when sampling occurs. The
+$\alpha$ parameter is unitless, while $m$ has units of edges or connections per
+vertex, and $N$ and $I$ both have units of nodes or individuals.}
+
+\add{Before proceeding with a full validation of \textit{netabc} on simulated
+data, we undertook an experiment designed to assess the identifiability of the
+BA parameters. One parameter of the BA model was investigated at a time while
+holding all others fixed, a strategy commonly used when performing sensitivity
+analyses of mathematical models. This allowed us to perform a fast preliminary
+analysis without dealing with the ``curse of dimensionality'' of the full
+parameter space. We simulated trees under three different values of each
+parameter, and asked how well we could tell the different trees apart. The
+better we are able to distinguish the trees, the more identifiability we might
+expect for the corresponding parameter when we attempt to estimate it with
+ABC.}
+
+\add{This experiment also had the secondary purpose of validating our choice of
+the tree kernel as a distance measure in ABC. To tell the trees apart, we used
+a classifier based on the tree kernel, but we also tested two other tree shape
+statistics. Sackin's index~\autocite{shao1990tree} is a measure of tree
+imbalance which not take branch lengths into account, considering only the
+topology. The normalized lineages-through-time
+\autocite[nLTT,][]{janzen2015approximate} compares two trees based on normalized
+distributions of their branching times, and does not explicitly consider the
+topology. Since the tree kernel incorporates both of these sources of
+information, we expected it to outperform the other two statistics. Finally,
+the tree kernel can be tuned by adjusting the values of the meta-parameters
+$\lambda$ and $\sigma$ (the ``decay factor'' and ``radial basis function
+variance'', see \textcite{poon2013mapping}). $\lambda$ is used to penalize
+ween large subset trees which tend to dominate the kernel score. When $\lambda
+= 0$, all but the root branches of each subset tree are ignored, while when
+$\lambda = 1$, no penalty is applied. $\sigma$ controls how strictly the notion
+of similarity is applied to branch lengths. When $\sigma = 0$, branch lengths
+must match exactly, while as $\sigma \to \infty$, branch lengths are not
+considered at all.}
+
+\del{We used the phylogenetic kernel developed by \textcite{poon2013mapping}}
+To test whether the parameters of the BA model had a \add{measurable} effect on
+tree shape, 100 networks were simulated under each of three different values of
+$\alpha$: 0.5, 1.0, and 1.5 (300 networks total). The other parameters were
+fixed to the following values: $N = 5000$, $I = 1000$, and $m = 2$. A
+transmission tree with 500 tips was simulated over each network (300
+transmission trees total). The 300 trees were compared pairwise with the tree
+kernel to form a $300 \times 300$ kernel matrix. The kernel meta-parameters
+$\lambda$ \del{(the ``decay factor''),} and $\sigma$ \del{(the ``radial basis
+function variance'')~\autocite[see][]{poon2013mapping},} were set to 0.3 and 4
+respectively. \add{We also computed a $300 \times 300$ matrix of pairwise nLTT
+values, and a $1 \times 300$ vector of Sackin's index values.} We constructed 
+\add{three classifiers for $\alpha$:} a kernel support vector regression (kSVM)
+\add{from the kernel matrix} with the \textit{kernlab}
+package~\autocite{zeileis2004kernlab}, and two ordinary SVMs \add{from the nLTT
+matrix and Sackin's index vector} with the e1071
+package~\autocite{meyer2015e1071}.\del{, and a linear regression from the
+Sackin's index values.} The accuracy of each classifier was evaluated with 1000
+two-fold cross validations \add{with equally-sized folds}.
 
@@
 tested were fixed at the values $N$ = 5000, $I$ = 1000, $m$ = 2, and $\alpha$ =
-1. Thus, we performed a total of four kSVR cross-validations, one for each of
-the BA model parameters $\alpha$, $I$, $m$, and $N$. We repeated these four
-cross-validations with different values of $\lambda$ (0.2, 0.3, and 0.4) and
-$\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$), as well as on trees with
-differing numbers of tips (100, 500, and 1000) and in epidemics of differing
-size (500, 1000, and 2000). The combination of the number of sampled
-individuals (\textit{i.e.} the number of tips) and the epidemic size
-(\textit{i.e.} $I$) will be creferred to as an ``epidemic scenario''. When
-evaluating the classifier for $I$, we did not consider trees with 1000 tips,
-because one of the tested $I$ values was 500, and the number of tips cannot be
-larger than $I$.
-
-For each of the four parameters, we also tested a linear regression against
-Sackin's index~\autocite{shao1990tree} and an ordinary SVR based on the normalized
-lineages-through-time (nLTT) statistic~\autocite{janzen2015approximate}.
-
-\subsection{ABC simulations}
-
-We simulated three transmission trees, each with 500 tips, under every element
-of the Cartesian product of these parameter values: $N$ = 5000, $I$ =
-\{1000, 2000\}, $m$ = \{2, 3, 4\}, and $\alpha$ = \{0.0, 0.5, 1,
-1.5\}. This produced a total of 24 parameter combinations $\times$ three trees
-per combination = 72 trees total. The adaptive ABC algorithm was applied to
-each tree with these priors: $m \sim$ DiscreteUniform(1, 5), $\alpha \sim$
-Uniform(0, 2), and $(N, I)$ jointly uniform on the region \{$500 \leq N \leq
-15000$, $500 \leq I \leq 5000$, $I \leq N$\}. Proposals for $\alpha$, $N$, and
-$I$ were Gaussian, while proposals for $m$ were Poisson. Following
-\textcite{del2012adaptive} and \textcite{beaumont2009adaptive}, the variance of
-all proposals was equal to the empirical variance of the particles.
-
-The algorithm was run with 1000 particles, 5 simulated datasets per particle,
-and the ``quality'' parameter controlling the decay rate of the tolerance
-$\varepsilon$ set to 0.95. We used the same stopping criterion as
-\citeauthor{del2012adaptive}, namely when the MCMC acceptance rate dropped
-below 1.5\%. Point estimates for the parameters were obtained by taking the
-highest point of an estimated kernel density on the final set of particles,
-calculated using the \textit{density} function with the default parameters in
-\textit{R}. Highest posterior density (HPD) intervals were calculated with the
+1. Thus, we performed a total of four cross-validations \add{for each
+classifier}, one for each of the BA model parameters $\alpha$, $I$, $m$, and
+$N$. We repeated these four cross-validations with different values of
+$\lambda$ (0.2, 0.3, and 0.4) and $\sigma$ ($2^{-3}$, $2^{-2}$, \ldots, $2^3$),
+as well as on trees with differing numbers of tips (100, 500, and 1000).
+\add{For the structural parameters $\alpha$, $m$, and $N$, the experiments were
+repeated with three different fixed values of $I$ (500, 1000, and 2000).}
+\del{and in epidemics of differing size (500, 1000, and 2000).} The combination
+of the number of sampled individuals (\textit{i.e.} the number of tips) and the
+epidemic size (\textit{i.e.} $I$) will be referred to as an ``epidemic
+scenario''. When evaluating the classifier for $I$, we did not consider trees
+with 1000 tips, because one of the tested $I$ values was 500, and the number of
+tips cannot be larger than $I$.
+
+\del{For each of the four parameters, we also tested a linear regression against
+Sackin's index~\autocite{shao1990tree} and an ordinary SVM based on the normalized
+lineages-through-time (nLTT) statistic~\autocite{janzen2015approximate}.}
+
+\subsection*{ABC simulations}
+
+\add{We tested \textit{netabc} by jointly estimating the four parameters of the
+BA model. We used the standard validation approach of simulating transmission
+trees under the model with known parameter values and attempting to recover
+those values with \textit{netabc}. The algorithm was not informed of any of
+the true parameter values for the main set of simulations.} We simulated three
+transmission trees, each with 500 tips, under every element of the Cartesian
+product of these parameter values: $N$ = 5000, $I$ = \{1000, 2000\}, $m$ = \{2,
+3, 4\}, and $\alpha$ = \{0.0, 0.5, 1, 1.5\}. This produced a total of 24
+parameter combinations $\times$ three trees per combination = 72 trees total.
+The adaptive ABC algorithm was applied to each tree with these priors: $m \sim$
+DiscreteUniform(1, 5), $\alpha \sim$ Uniform(0, 2), and $(N, I)$ jointly
+uniform on the region \{$500 \leq N \leq 15000$, $500 \leq I \leq 5000$, $I
+\leq N$\}. Proposals for $\alpha$, $N$, and $I$ were Gaussian, while proposals
+for $m$ were Poisson. Following \textcite{del2012adaptive} and
+\textcite{beaumont2009adaptive}, the variance of all proposals was equal to the
+empirical variance of the particles.
+
+The SMC settings used were 1000 particles, 5 simulated datasets per particle,
+and \del{the ``quality'' parameter controlling the decay rate of the tolerance
+$\varepsilon$ set to}\add{$\alpha_{\text{ESS}} =  0.95$}. We used the same
+stopping criterion as \citeauthor{del2012adaptive}, namely when the MCMC
+acceptance rate dropped below 1.5\%. \del{Point estimates for the parameters
+were obtained by taking the highest point of an estimated kernel density on the
+final set of particles, calculated using the \textit{density} function with the
+default parameters in \textit{R}.} \add{Approximate posterior means for the
+parameters were obtained by taking the weighted average of the final set of
+particles.} Highest posterior density (HPD) intervals were calculated with the
 \textit{HPDinterval} function from the \textit{R} package
@@
 that node's network peers had already been sampled. The parameters of this
-network were $N$ = 5000, $I$ = 2000, $m$ = 2, and $\alpha$ = 0.5.
-
-\subsection{Investigation of published data}
-
-We applied our kernel-ABC method to several published HIV datasets. Because the
-BA model generates networks with a single connected component, we
-specifically searched for datasets which originated from existing clusters,
-either phylogenetically or geographically defined. Characteristics of the
-datasets we investigated are given in \cref{tab:data}.
+network were $N= 5000$, $I = 2000$, $m = 2$, and $\alpha = 0.5$.
+
+\add{Despite the fact that the parameter values used to generate the simulated
+transmission trees were known, the true posterior distributions of the BA
+parameters were unknown. Therefore, any apparent errors or biases in the
+estimates could be due to either poor performance of our method, or to real
+features of the posterior distribution. Two retrospective experiments were
+performed to disambiguate some of the observed errors. To assess the impact of
+the SMC settings on \textit{netabc}'s accuracy, we ran \textit{netabc}
+twice on the same simulated transmission tree. For the first run, the SMC
+settings were the same as in the other simulations: 1000 particles, 5 simulated
+transmission trees per particle, and $\alpha_{\text{ESS}} = 0.95$. The second
+run was performed with 2000 particles, 10 simulated transmission trees per
+particle, and $\alpha_{\text{ESS}} = 0.97$. To investigate the extent to which
+errors in the estimated BA parameters were due to true features of the
+posterior, rather than an inaccurate ABC approximation, we performed marginal
+estimation for one set of parameter values. Each combination of 1, 2, or 3
+model parameters (14 combinations total) was fixed to their known values, and
+the remaining parameters were estimated with \textit{netabc}. The parameter
+values were $\alpha = 0.0$, $m = 2$, $I = 2000$, and $N = 5000$.}
+
+\subsection*{Investigation of published data}
+
+We applied our ABC method to ten published HIV datasets. Because the BA model
+generates networks with a single connected component, we specifically searched
+for datasets which originated from existing clusters, either phylogenetically
+or geographically defined. Characteristics of the datasets we investigated are
+given in \cref{tab:data}. \add{For clarity, we will refer to each dataset by
+its risk group and location of origin in the text. For example, the
+\textcite{zetterberg2004two} data will be referred to as IDU/Estonia.}
 
@@
   \input{\tablepath/realdata}
-  \caption{Characteristics of published datasets investigated with kernel-ABC.
-  Acronyms: MSM, men who have sex with men; IDU, injection drug users.}
+  \caption{
+    Characteristics of published datasets investigated with ABC.
+    Acronyms: MSM, men who have sex with men; IDU, injection drug users; HET,
+    heterosexual. The HET data were sampled from a primarily heterosexual risk
+    environment, but did not explicitly exclude other risk factors. The number
+    of sequences column indicates how many sequences were included in our
+    analysis; there may have been additional sequences linked to the study
+    which we excluded for various reasons (see methods).
+  }
   \label{tab:data}
@@
 
-We downloaded all sequences associated with each study from GenBank. For the
-\textcite{novitsky2014impact} data, each sequence was aligned pairwise to the HXB2
-creference sequence (Genbank accession number K03455) and the hypervariable
-regions were clipped out with \textit{BioPython} version
-1.66+~\autocite{cock2009biopython}. Sequences were multiply aligned using
-\textit{MUSCLE} version 3.8.31 \autocite{edgar2004muscle}, and alignments were
-manually inspected with \textit{Seaview} version 4.4.2
-\autocite{gouy2010seaview}. Phylogenies were constructed from the nucleotide
-alignments by approximate maximum likelihood using \textit{FastTree2} version
-2.1.7 with the generalized time-reversible model. Transmission trees were
-estimated by rooting and time-scaling the phylogenies by root-to-tip
-regression, using a modified version of Path-O-Gen (distributed as part of
-BEAST~\autocite{drummond2007beast}) as described
-previously~\autocite{poon2015phylodynamic}.
-
-Two of the datasets \autocite{li2015hiv,novitsky2014impact} were initially much
-larger than the others, containing 1265 and 1299 sequences respectively. To
+We downloaded all sequences associated with each published study from GenBank.
+\add{For the IDU/Romania data, only sequences from injection drug users (IDU,
+whose sequence identifiers included the letters ``DU'') were included in the
+analysis. \textcite{kao2011surveillance} (MSM/Taiwan) found a strong
+association in their study population between subtype and risk group - subtype
+B was most often associated with men who have sex with men (MSM), whereas IDU
+were usually infected with a circulating recombinant form. Since there were
+many more subtype B sequences in their data than sequences of other subtypes,
+we restricted our analysis to the subtype B sequences and labelled this dataset
+as MSM. Two datasets (HET/Uganda and HET/Malawi) included both \textit{env}
+and \textit{gag} sequences. Each gene was analyzed separately to assess the
+robustness of \textit{netabc} to the particular HIV gene sequence used to
+estimate a transmission tree. The IDU/Estonia data also sequenced both genes, 
+but the highly variable coverage and high homology of the \textit{gag}
+sequences made it impossible to obtain a sufficiently large block of
+non-identical sequences to analyze. Therefore, we analyzed only \textit{env}
+for this dataset.}
+
+\del{For the \textcite{novitsky2014impact} data,} Each \textit{env} sequence
+was aligned pairwise to the HXB2 reference sequence (GenBank accession number
+K03455), and the hypervariable regions were clipped out with
+\textit{BioPython} version 1.66+~\autocite{cock2009biopython}. Sequences were
+multiply aligned using \textit{MUSCLE} version 3.8.31
+\autocite{edgar2004muscle}, and alignments were manually inspected with
+\textit{Seaview} version 4.4.2 \autocite{gouy2010seaview}. \add{Duplicated
+sequences were removed with \textit{BioPython}.} Phylogenies were constructed
+from the nucleotide alignments by approximate maximum likelihood using
+\textit{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
+generalized time-reversible (GTR) model~\autocite{tavare1986some}. Transmission
+trees were estimated by rooting and time-scaling the phylogenies by root-to-tip
+regression, using a modified version of \textit{Path-O-Gen} (distributed as
+part of \textit{BEAST}~\autocite{drummond2007beast}) as described
+previously~\autocite{poon2015phylodynamic}. \add{Due to the removal of
+duplicated sequences, all estimated transmission trees were fully binary.}
+
+\add{To check if our results were robust to the choice of phylogenetic
+reconstruction method, we built and reanalyzed phylogenies for two of the
+datasets (mixed/Spain and IDU/Romania) with
+\textit{RAxML}~\autocite{stamatakis2014raxml} with the GTR+$\Gamma$ model of
+sequence evolution and rate heterogeneity. The trees were rooted and
+time-scaled with \textit{Least Square
+Dating}~\autocite[\textit{LSD},][]{to2016fast}.}
+
+\del{Two} \add{Four} of the datasets \del{\autocite{li2015hiv,novitsky2014impact}}
+(MSM/Shanghai, HET/Botswana, HET/Uganda, and MSM/USA)
+were initially much larger than the others, containing 1265, 1299,
+\add{1026/915 (\textit{env}/\textit{gag}), and 648} sequences respectively. To
 ensure that the analyses were comparable, we reduced these to a number of
-sequences similar to the smaller datasets. For the \textcite{li2015hiv} data,
-we detected a cluster of size 280 using a patristic distance cutoff of 0.02 as
+sequences similar to the smaller datasets. For the MSM/Shanghai data, we
+detected a cluster of size 280 using a patristic distance cutoff of 0.02 as
 described previously~\autocite{poon2015impact}. Only sequences within this
-cluster were carried forward. For the \textcite{novitsky2014impact} data, no
-large clusters were detected using the same cutoff, so we analysed a subtree of
-size 180 chosen arbitrarily.
+cluster were carried forward. For the HET/Uganda, HET/Botswana, and MSM/USA
+data, no large clusters were detected using the same cutoff, so we analyzed
+\del{a subtree} subsets of sizes 255, 180, and 180 respectively. \add{The
+subset of the HET/Uganda data was chosen by eye such that the individuals were
+monophyletic in both the \textit{gag} and \textit{env} trees. The other
+subsets were arbitrarily chosen subtrees from phylogenies of the complete
+datasets.}
 
@@
 and one with the prior $m \sim$ DiscreteUniform(2, 5). The other parameters to
-the SMC algorithm were the same as used for the simulation experiments.
-
-\section{Results}
-
-\subsection{Kernel classifiers}
-
-We investigated the parameters of the BA network
-model~\autocite{barabasi1999emergence}. In addition to $m$ and $\alpha$ (see
-Introduction), we considered $N$, which denotes the total number of nodes in
-the network, and $I$, which is the number of infected nodes at which to stop
-the simulation and sample the transmission tree. To examine the effect of these
-parameters on tree shape, we simulated transmission trees under different
+the SMC algorithm were the same as used for the simulation experiments,
+\add{except that we used 10000 particles instead of 1000 to increase the
+accuracy of the estimated posterior. This was computationally feasible due to
+the small number of runs required for this analysis.}
+
+\section*{Results}
+
+\subsection*{\del{Kernel classifiers}\add{Classifiers for BA model parameters from tree shapes}}
+
+We investigated the \add{identifiability of four} parameters of the BA network
+model~\autocite{barabasi1999emergence}: the number of nodes $N$, the
+preferential attachment power $\alpha$, the number of edges added per vertex
+$m$, and the number of infected nodes $I$. \del{In addition to $m$ and $\alpha$
+(see Introduction), we considered $N$, which denotes the total number of nodes
+in the network, and $I$, which is the number of infected nodes at which to stop
+the simulation and sample the transmission tree.} To examine the effect of
+these parameters on tree shape, we simulated transmission trees under different
 parameter values, calculated pairwise tree kernel scores between them, and
-attempted to classify the trees using a kernel support vector machine (kSVR).
+attempted to classify the trees using a kernel support vector machine (kSVM).
 We also tested classifiers based on Sackin's index~\autocite{shao1990tree} and
 the normalized lineages-through-time (nLTT)
-statistic~\autocite{janzen2015approximate}. Accuracy of the kSVRs varied based
-on the parameter being tested (\cref{fig:rsquared}, left). Classifiers based on
-two other tree statistics, the nLTT and Sackin's index, generally exhibited
-worse performance than the tree kernel, although the magnitude of the disparity
-varied between the parameters (\cref{fig:rsquared}, centre and right). The
-results were largely robust to variations in the tree kernel meta-parameters
-$\lambda$ and $\sigma$ (\cref{fig:alphacrossv,fig:mcrossv,fig:Ncrossv,fig:Icrossv}).
-
-When classifying $\alpha$, the kernel-SVR classifier had an average $R^2$ of 
+statistic~\autocite{janzen2015approximate}. \add{We report the accuracy of the
+classifiers, which is simply the proportion of trees which were assigned the
+correct parameter value. Since there were three possible values, random
+guessing would produce an accuracy of 0.33.} \del{The accuracy of each
+classifier varied based on the parameter being tested} \add{The results are
+shown in} \cref{fig:rsquared}. Classifiers based on \del{two other tree
+statistics,} the nLTT and Sackin's index generally exhibited worse performance
+than the tree kernel, although the magnitude of the disparity varied between
+the parameters (\cref{fig:rsquared}, centre and right). \del{The results were
+largely robust to variations in the tree kernel meta-parameters $\lambda$ and
+$\sigma$, although accuracy varied between different epidemic and sampling
+scenarios} \add{Larger datasets were generally classified more accurately
+(\cref{fig:alphacrossv,fig:mcrossv,fig:Ncrossv,fig:Icrossv}), although large
+values of $\lambda$ produced worse estimates on large datasets. Extremely low
+$\sigma$ values, which require nearly-exact matches between branch lengths,
+resulted in low accuracy in some cases (\textit{e.g.} \cref{fig:alphacrossv},
+center row).}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{kernel-rsquared.pdf}
+    \caption[
+        Cross-validation accuracy of kernel-SVM, nLTT-based SVM, and Sackin's
+        index SVM classifiers for BA model parameters.
+    ]{
+        Cross-validation accuracy of kernel-SVM classifier (left), SVM
+        classifiers using nLTT (centre) and Sackin's index (right) for BA model
+        parameters. Kernel meta-parameters were set to $\lambda = 0.3$ and
+        $\sigma = 4$. Each point was calculated based on 300 simulated
+        transmission trees over networks with three different values of the
+        parameter being tested, assuming perfect knowledge of the other
+        parameters. Vertical lines are empirical 95\% confidence intervals
+        based on 1000 two-fold cross-validations. \add{The classifiers for $I$
+        were not evaluated with 1000-tip trees, because one of the tested $I$
+        values was 500, and it is not possible to sample a tree of size 1000
+        from 500 infected individuals.}
+    }
+    \label{fig:rsquared}
+\end{figure}
+
+The kSVM classifier for $\alpha$ had an average accuracy of 
     0.92,
 compared to 
-    0.56
-for the nLTT-based SVR, and
-    0.75
-for the linear regression against Sackin's index. There was little variation
-about the mean for different tree and epidemic sizes. No classifier could
-accurately identify the $m$ parameter in any epidemic scenario, with average
-$R^2$ values of 
-  0.12 for kSVR,
-  0.01 for the nLTT, and
-  0.06
-for Sackin's index. Again, there was little variation in accuracy between
-epidemic scenarios, although the accuracy of the kSVR was slightly higher
-on 1000-tip trees (\cref{fig:rsquared}, left).
+    0.6
+for the nLTT, and
+    0.77
+for Sackin's index. \del{There was little variation about the mean for
+different tree and epidemic sizes.} No classifier could accurately identify $m$
+in any epidemic scenario, with average accuracy values of 
+  0.35 for kSVM,
+  0.32 for the nLTT, and
+  0.38
+for Sackin's index. There was little variation in accuracy between epidemic
+scenarios, although the accuracy of the kSVM was slightly higher on 1000-tip
+trees (\cref{fig:rsquared}, left).
 
 The accuracy of classifiers for $I$ varied significantly with the number of
-tips in the tree. For 100-tip trees, the average $R^2$ values were
-  0.7,
-  0.55, and
-  0.02
-for the tree kernel, nLTT, and Sackin's index respectively. For 500-tip
+tips in the tree. For 100-tip trees, the average accuracy was
+  0.59,
+  0.58, and
+  0.34
+for the tree kernel, nLTT, and Sackin's index respectively.
+For 500-tip
 trees, the values increased to
-  0.93,
-  0.83, and
-  0.07.
+  0.99,
+  0.76, and
+  0.37.
 Finally, the performance of classifiers for $N$ depended heavily on the
-epidemic scenario. The $R^2$ of the kSVR classifier ranged from
-  0.08
+epidemic scenario. The accuracy of the kSVM classifier ranged from
+  0.36
 for the smallest epidemic and smallest sample size, to
-  0.82
-for the largest. Likewise, $R^2$ for the nLTT-based SVR ranged from 
-  0.01
+  0.81
+for the largest. Accuracy for the nLTT ranged from 
+  0.33
 to
-  0.54.
+  0.63.
 Sackin's index did not accurately classify $N$ in any scenario, with an average
-$R^2$ of
-  0.03
+accuracy of
+  0.35
 and little variation between scenarios.
 
-\begin{figure*}[ht]
-  \centering
-  \includegraphics[width=\textwidth]{kernel-rsquared.eps}
-  \vspace{6pt}
-  \caption{
-      Cross-validation accuracy of kernel-SVR classifier (left), SVR classifier
-      using nLTT (centre), and linear regression using Sackin's index
-      (right) for BA model parameters. Kernel meta-parameters were set to
-      $\lambda = 0.3$ and $\sigma = 4$. Each point was calculated based on 300
-      simulated transmission trees over networks with three different values of
-      the parameter being tested. Vertical lines are empirical 95\% confidence
-      intervals based on 1000 two-fold cross-validations.
-  }
-  \label{fig:rsquared}
-\end{figure*}
-
-\subsection{ABC simulations}
-
-%<<point_est, include=FALSE>>=
-%    f <- Sys.glob("../../simulations/abc-pa-free-m/point-estimate/*")
-%    d <- fread(f)
-%    d[,m := floor(m)]
-%    d[,alpha_error := abs(true_alpha - alpha)]
-%    d[,N_error := abs(true_N - N)]
-%    d[,I_error := abs(true_I - I)]
-%    d[,m_error := abs(true_m - m)]
-%
-%    alpha.av <- anova(lm(alpha_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-%    m.av <- anova(lm(m_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-%    N.av <- anova(lm(N_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-%    I.av <- anova(lm(I_error ~ factor(true_alpha) + factor(true_m) + factor(true_I), d))
-%
-%    m.tbl <- d[,prop.table(table(m_error))]
-%    zero.alpha.test <- wilcox.test(d[true_alpha == 0, alpha_error], 
-%                                   d[true_alpha != 0, alpha_error])
-%    alpha.I.test <- wilcox.test(d[true_alpha < 1, I_error],
-%                                d[true_alpha >= 1, I_error])
-%    I.I.test <- wilcox.test(d[true_I == 1000, I_error],
-%                            d[true_I == 2000, I_error])
-%    alpha.m.test <- wilcox.test(d[true_alpha == 0 | true_alpha == 1, m_error],
-%                                d[true_alpha == 0.5 | true_alpha == 1.5, m_error])
-%    stopifnot(alpha.av["factor(true_m)", "Pr(>F)"] > 0.05)
-%    stopifnot(alpha.av["factor(true_I)", "Pr(>F)"] > 0.05)
-%    stopifnot(I.I.test$p.value < 1e-5)
-%    stopifnot(I.av["factor(true_m)", "Pr(>F)"] > 0.05)
-%    stopifnot(m.av["factor(true_m)", "Pr(>F)"] > 0.05)
-%    stopifnot(m.av["factor(true_I)", "Pr(>F)"] > 0.05)
-%    stopifnot(min(N.av[1:3,"Pr(>F)"]) > 0.05)
-%@
-
-\Cref{fig:abcpt} shows maximum \textit{a posteriori} (MAP) point
-estimates of the BA model parameters obtained with kernel-ABC on simulated
-data. The estimates shown correspond only to the simulations where the $m$
-parameter was set to 2, however the results for $m = 3$ and $m = 4$ were
-similar (\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% HPD
-intervals are given in \cref{tab:abchpd}.
-
-The accuracy of the parameter estimates obtained with kernel-ABC
-paralleled the results from the kSVR classifier. Of the four parameters,
+\subsection*{ABC simulations}
+
+\Cref{fig:abcpt} shows \del{maximum \textit{a posteriori} (MAP)}
+\add{stratified posterior mean} point estimates of the BA model parameters
+\add{$\alpha$ and $I$}, obtained with ABC on simulated data. \add{The
+parameters $m$ and $N$ were not identifiable with ABC for any parameter
+combinations (\cref{fig:abcpt2}).} \del{The estimates shown correspond only to
+the simulations where the $m$ parameter was set to 2, however the results for
+$m = 3$ and $m = 4$ were similar.} Average boundaries of 95\% HPD intervals
+\add{for all parameters} are given in \cref{tab:abchpd}.
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{abc-boxplot}
+    \caption[
+        Posterior mean point estimates for BA model parameters $\alpha$ and $I$
+        obtained by running \textit{netabc} on simulated data, stratified by
+        true parameter values.
+    ]{
+        Posterior mean point estimates for BA model parameters $\alpha$ and $I$
+        obtained by running \textit{netabc} on simulated data, stratified by
+        true parameter values. First row of plots contains true versus
+        estimated values of $\alpha$; second row contains true versus estimated
+        values of $I$. Columns are stratified by $\alpha$, $I$, and $m$
+        respectively. Dashed lines indicate true values. 
+    }
+    \label{fig:abcpt}
+\end{figure}
+
+\begin{table}[ht]
+    \centering
+    \input{\tablepath/abc-hpd.tex}
+    \caption[
+        Average posterior mean point estimates and 95\% highest posterior
+        density interval widths for BA model parameter estimates obtained
+        with \textit{netabc} on simulated data.
+    ]{
+        Average posterior mean point estimates and 95\% highest posterior
+        density interval widths for BA model parameter estimates obtained with
+        \textit{netabc} on simulated data. Three transmission trees were
+        simulated under each combination of the listed parameter values, and
+        the parameters were estimated with ABC without training.
+    }
+    \label{tab:abchpd}
+\end{table}
+
+\del{The accuracy of the parameter estimates obtained with ABC
+paralleled the results from the kSVM classifier. Of the four parameters,
 $\alpha$ was the most accurately estimated, with point estimates having a
 median [IQR] absolute error of 
-%    d[,round(median(alpha_error), 2)] 
-%    [d[,round(quantile(alpha_error, 0.25), 2)] - 
-%    d[,round(quantile(alpha_error, 0.75), 2)]].
+    0.11 
+    [0.03 - 
+    0.25].
 The errors when the true value of $\alpha$ was zero were significantly greater
 than those for the other values 
-%    (Wilcoxon rank-sum test, $p$ = $latexSN(round(zero.alpha.test$p.value, 4))$),
-but did not vary across the true values of the other parameters (one-way
-ANOVA). Estimates for $I$ were also relatively accurate, with point estimate
-errors of
-%    d[,round(median(I_error))] 
-%    [d[,round(quantile(I_error, 0.25))] - 
-%    d[,round(quantile(I_error, 0.75))]].
+(Wilcoxon rank-sum test, $p$ = $0$).
+Errors in estimating $\alpha$ also varied with the true value of $m$ just at
+the threshold of statistical significance
+    $p =0.5$),
+but did not vary across the true values of $N$ or $I$ (both one-way ANOVA).
+Estimates for $I$ were relatively accurate, with point estimate errors of
+    492 
+    [294 - 
+    782] individuals.
 These errors were significantly higher when the true value of $\alpha$ was
 at least 1
-%    (Wilcoxon rank-sum test, $p$ = $latexSN(round(alpha.I.test$p.value, 4))$)
+    (Wilcoxon rank-sum test, $p$ = $0$)
 and when the true value of $I$ was 2000 ($p < 10^{-5}$). The true value of $m$
-did not affect the estimates of $I$ (one-way ANOVA).
-
-The $m$ parameter was estimated correctly in only
-%    as.integer(m.tbl[1] * 100) \%
-of simulations. Oddly, the error in the estimated $m$ was higher for integer
-values of $\alpha$ (\textit{i.e.}~0 and 1) than non-integer values 
-%    (Wilcoxon rank-sum test, $p$ = round(alpha.m.test$p.value, 3)).
-The true values of the other parameters did not significantly affect the
-estimates of $m$ (both one-way ANOVA). Finally, the total number of nodes $N$
-was consistently over-estimated by about a factor of two
-%    (error d[,round(median(N_error))] 
-%    [d[,round(quantile(N_error, 0.25))] - 
-%     d[,round(quantile(N_error, 0.75))]]).
-No parameters influenced the accuracy of the $N$ estimates (all one-way ANOVA).
-
-\begin{figure*}[ht]
-  \centering
-  \includegraphics[width=\textwidth]{abc-point-estimate-m2.eps}
-  \vspace{6pt}
-  \caption{
-    Point estimates of BA model parameters obtained by running kernel-ABC
-    on simulated phylogenies without training, for simulations with $m = 2$.
-    Dotted lines indicate true values, and limits of the $y$-axes are regions
-    of uniform prior density. (A) Estimates for $\alpha$ and $I$ against their
-    true values in simulations. (B) Estimates for $m$ and $N$, which were held
-    fixed in these simulations, against true values of $\alpha$.
-  }
-  \label{fig:abcpt}
-\end{figure*}
-
-\begin{table*}[ht]
-  \centering
-  \input{\tablepath/abc-hpd.tex}
-  \caption{
-      Average maximum \textit{a posteriori} point estimates and 95\% highest
-      posterior density (HPD) interval widths for BA model parameter estimates
-      obtained with kernel-ABC. Three transmission trees were simulated under
-      each combination of the listed parameter values, and the parameters were
-      estimated with kernel-ABC without training.
-  }
-  \label{tab:abchpd}
-\end{table*}
+did not affect the estimates of $I$ (one-way ANOVA).}
+
+\del{The $m$ parameter was estimated correctly in only
+    37 \%
+of simulations, barely better than random guessing. The true values of the
+other parameters did not significantly affect the estimates of $m$ (both
+one-way ANOVA). Finally, the total number of nodes $N$ was consistently
+over-estimated by about a factor of two
+    (error 4153 
+    [3660 - 
+     4489] individuals).
+No parameters influenced the accuracy of the $N$ estimates (all one-way ANOVA).}
+
+\add{Across all simulations, the median [IQR] absolute errors of the parameter 
+estimates obtained with \textit{netabc} were
+    0.11 
+    [0.03 - 
+    0.25]
+for $\alpha$,
+    492 
+    [294 - 
+    782]
+for $I$,
+    1 
+    [0 - 
+    1]
+for $m$, and
+    4153 
+    [3660 - 
+    4489]
+for $N$. These errors comprised, respectively,
+    6\%,
+    11\%,
+    17\%, and
+    29\%
+of the regions of nonzero prior density. For $I$ and $N$, relative errors were
+    38\%
+    [20 - 
+    50\%]
+and
+    83\%
+    [73 - 
+    90\%].
+Average 95\% HPD interval widths were
+    0.68,
+    2454,
+    3.01, and
+    12046,
+representing 
+    34\%,
+    55\%,
+    50\%, and
+    83\%
+of the nonzero prior density regions. Point estimates of $I$ were upwardly
+biased: $I$ was overestimated in
+    69
+out of 
+    72
+simulations
+    (96\%).
+The estimates for $m$ and $N$ were similar across all simulations (median [IQR]
+point estimates
+    3
+    [3 - 
+    3]
+and 
+    9153
+    [8660 - 
+    9489])
+regardless of the true values of any of the BA parameters (\cref{fig:abcpt2}).}
+
+% latex table generated in R 3.2.3 by xtable 1.8-2 package
+% Wed Jul  6 16:31:52 2016
+\begin{table}[ht]
+\centering
+\begin{tabular}{lllll}
+  \hline
+Dependent variable & Independent variable & Estimate & Standard error & $p$-value \\ 
+  \hline
+$\alpha$ & (Intercept) & $2$ & $0.6$ & $0.01$ \\ 
+   & $\alpha$ & $10$ & $2$ & ${<}10^{-5}$ \\ 
+   & $I$ & $-3\!\times\!10^{-4}$ & $2\!\times\!10^{-4}$ & $0.7$ \\ 
+   & $m$ & $0.5$ & $0.2$ & $0.01$ \\ 
+  $I$ & (Intercept) & $0.004$ & $5\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
+   & $\alpha$ & $-0.001$ & $2\!\times\!10^{-4}$ & ${<}10^{-5}$ \\ 
+   & $I$ & $-4\!\times\!10^{-7}$ & $2\!\times\!10^{-7}$ & $0.05$ \\ 
+   & $m$ & $-7\!\times\!10^{-5}$ & $8\!\times\!10^{-5}$ & $1$ \\ 
+   \hline
+\end{tabular}
+\caption[Parameters of fitted GLMs relating error in estimated $\alpha$ and $I$ to true values of BA parameters.]{Parameters of fitted GLMs relating error in estimated $\alpha$ and $I$ to true values of BA parameters.
+GLMs ere fitted with the Gaussian distribution and inverse link function.
+Coefficients are interpretable as additive effects on the inverse of the mean error.}
+\label{tab:glm}
+\end{table}
+
+\add{To analyze the effects of the true parameter values on the
+accuracy our estimates of $\alpha$ and $I$, we fitted one GLM for each of these
+two parameters, with error rate as the dependent variable and the true
+parameter values as independent variables. Since the estimates of $m$ and
+$N$ were roughly equal across all simulations (\cref{fig:abcpt2}), GLMs were
+not fitted for these parameters. The estimated coefficients are shown in
+\cref{tab:glm}. The GLM analysis indicated that the error in estimates of
+$\alpha$ decreased with larger true values of $\alpha$
+    ($p {<}10^{-5}$)
+and $m$
+    ($p =0.01$)
+but was not significantly affected by $I$. Qualitatively, $\alpha$ seemed to be
+only weakly identifiable between the values of 0 and 0.5 (\cref{fig:abcpt}).
+The error in the estimated $I$ value was slightly lower for smaller values of
+$\alpha$
+    ($p {<}10^{-5}$)
+and $I$
+    ($p =0.05$),
+but was not significantly affected by the true value of $m$.}
 
 The dispersion of the ABC approximation to the posterior also varied between
-the parameters, with narrower HPD intervals for the parameters with more
-accurate point estimates (\cref{tab:abchpd}). \Cref{fig:abcex} shows
-the distributions for for one simulation (equivalent plots for all the
-simulations can be found in Supplementary Figures TODO). HPD intervals around
-$\alpha$ and $I$ were narrow relative to the region of nonzero prior density,
-whereas the intervals for $m$ and $N$ were more widely dispersed.
-
-\begin{figure*}[ht]
-    \centering
-  \includegraphics[width=\textwidth]{abc-posterior-example.eps}
-  \vspace{6pt}
-  \caption{
-    Marginal posterior distributions of BA model parameters estimated
-    with kernel-ABC for a single simulated transmission tree. Dotted
-    lines and shaded polygon indicate true values.
-  }
-  \label{fig:abcex}
-\end{figure*}
+the parameters\del{, with narrower HPD intervals for the parameters with more
+accurate point estimates} (\cref{tab:abchpd}). HPD intervals around $\alpha$
+and $I$ were often narrow relative to the region of nonzero prior
+density, whereas the intervals for $m$ and $N$ were more widely dispersed.
+\Cref{fig:abc1,fig:abc2} \del{shows the distributions for one
+simulation.} \add{show one- and two-dimensional marginal distributions for a
+simulation with $\alpha$ and $I$ errors close to their respective medians. 
+%In particular, the simulation shown is one of
+%replace_number(nrow(middle)) with errors in both $\alpha$ and $I$
+%between the round(q*100)nd and round((1-q)*100)th percentiles.
+The parameters for this simulation were
+    $\alpha = 1$,
+    $I = 1000$,
+    $m = 3$, and
+    $N = 5000$.
+The two-dimensional marginals indicate some dependence between pairs of
+parameters, particularly $I$ and $N$ which show a diagonally shaped region of
+high posterior density.}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=0.7\textwidth]{{abc-posterior/1.0_1000_3_5000_0}.pdf}
+    \caption[
+        One-dimensional marginal posterior distributions of BA model
+        parameters estimated by \textit{netabc} from a simulated transmission
+        tree.
+    ]{
+        One-dimensional marginal posterior distributions of BA model parameters
+        estimated by \textit{netabc} from a simulated transmission tree. Dashed
+        lines indicate true values, solid lines indicate posterior means, and
+        shaded areas show 95\% highest posterior density intervals.
+    }
+    \label{fig:abc1}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics[width=\textwidth]{{abc-posterior-2d/1.0_1000_3_5000_0}.pdf}
+    \caption[
+        Two-dimensional marginal posterior distributions of BA model parameters
+        estimated by \textit{netabc} from a simulated transmission tree.
+    ]{
+        Two-dimensional marginal posterior distributions of BA model parameters
+        estimated by \textit{netabc} from a simulated transmission tree. White
+        circles indicate true values, magenta diamonds
+        indicate posterior means. 
+    }
+    \label{fig:abc2}
+\end{figure}
 
 To test the effect of model misspecification, we simulated one network where
-the nodes exhibited heterogeneous pcreferential attachment power (half 0.5, the
-other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The MAP [95\%
-HPD] estimates for each parameter were: 
+the nodes exhibited heterogeneous preferential attachment power (half 0.5, the
+other half 1.5), with $m$ = 2, $N$ = 5000, and $I$ = 1000. The posterior mean
+[95\% HPD] estimates for each parameter were: 
 $\alpha$, 
-  1.1 
-  [0.6 -
-   1.15];
+  1.03 
+  [0.67 -
+   1.18];
 $I$,
-  1141 
-  [662 -
-   4455];
+  1474 
+  [511 -
+   2990];
+$m$,
+  3 
+  [1 -
+   5];
+$N$,
+  9861 
+  [3710-
+   14977].
+\del{The approximate one-dimensional marginal posterior distributions for this
+simulation are shown in \cref{fig:mixed}.} To test the effect of sampling bias,
+we sampled one transmission tree in a peer-driven fashion, where the
+probability to sample a node was twice as high if one of its peers had already
+been sampled. The parameters for this experiment were $N$ = 5000, $m$ = 2,
+$\alpha$ = 0.5, and $I$ = 2000. The estimated values were $\alpha$, 
+  0.3 
+  [0 -
+   0.63];
+$I$,
+  2449 
+  [1417 -
+   3811];
 $m$,
@@
 $N$,
-  12689 
-  [3733-
-   14977].
-To test the effect of sampling bias, we sampled one transmission tree in a
-peer-driven fashion, where the probability to sample a node was twice as high
-if one of its peers had already been sampled. The parameters for this
-experiment were $N$ = 5000, $m$ = 2, $\alpha$ = 0.5, and $I$ = 2000. The
-estimated values were
-$\alpha$, 
-  0.36 
+  9132 
+  [2852 -
+   14780].
+\del{The approximate one-dimensional marginal posterior distributions are shown
+in \cref{fig:peerdriven}.} Both of these results were in line with estimates
+obtained on other simulated datasets (\cref{tab:abchpd}), although the estimate
+of $\alpha$ for peer-driven sampling was somewhat lower than typical.
+
+\add{\Cref{fig:marginal} shows the effect of performing marginal ABC estimation
+of each of the BA parameters on the same simulated transmission tree. The
+estimates of $m$ were apparently unaffected by marginalizing out the other
+parameters, corroborating the previous experiments' findings that $m$ is not an
+identifiable parameter from scaled tree shapes. Compared to allowing all
+parameters to vary, estimates of $\alpha$, $I$, and $N$ were improved by
+    41\%,
+    59\%, and
+    46\%
+when all other parameters were fixed. \Cref{fig:better} shows the impact of
+increasing the number of particles, simulated datasets, and
+$\alpha_{\text{ESS}}$ parameter on the accuracy of a single simulation. The
+number of iterations until the stopping condition was reached was 81 with the
+basic settings and 124 with the higher settings. The results of the two
+simulations were very similar for all parameters.}
+
+\subsection*{Published HIV data}
+
+We applied ABC to five published HIV datasets (\cref{tab:data}),
+and found substantial heterogeneity among the parameter estimates
+(\cref{fig:abchpd,fig:abchpdm2}). \add{Posterior mean point estimates and 50\%
+and 95\% HPD intervals for each parameter are shown in \cref{fig:abchpd}.}
+\del{Plots of the marginal posterior distributions for each dataset are shown
+in \cref{fig:cuevas,fig:li,fig:niculescu,fig:novitsky,fig:wang}.}
+\add{\Cref{fig:abchpdm2} shows point estimates and HPD intervals obtained when
+the value $m = 1$ was disallowed by the prior. Since the results indicated that
+$m = 1$ was the most credible value for several datasets, all results discussed
+henceforth apply to the prior $m \sim \text{DiscreteUniform}(1, 5)$ unless
+otherwise stated.}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{realdata-hpd}
+    \caption[
+        Posterior means and 50\%/95\% HPD intervals for parameters of the
+        BA network model, fitted to ten HIV datasets with \textit{netabc}.
+    ]{
+        Posterior means (points), 50\% HPD intervals (notches), and 95\%
+        HPD intervals (lines) for parameters of the BA network model, fitted to
+        ten HIV datasets with \textit{netabc}. Legend labels indicate risk
+        group and country of origin. Abbreviations: IDU, injection drug users;
+        MSM, men who have sex with men; HET, heterosexual. Note that posterior
+        means can fall outside of the HPD interval if the distribution is
+        diffuse.
+    }
+    \label{fig:abchpd}
+\end{figure}
+
+\del{Two of the datasets~\autocite{niculescu2015recent, wang2015targeting} had
+estimated $\alpha$ values near unity for the prior allowing $m = 1$ (MAP
+estimates [95\% HPD] 
+  0.73 
+  [0.05 - 
+   1.18]
+and
+  0.55 
   [0.01 -
-   0.63];
-$I$,
-  2392 
-  [1419 -
-   3767];
-$m$,
-  3 
-  [2 -
-   5];
-$N$,
-  8760 
-  [2881 -
-   14788].
-Both of these results were in line with estimates obtained on other simulated
-datasets (\cref{tab:abchpd}).
-
-\subsection{Real data}
-
-We applied kernel-ABC to five published HIV datasets (\cref{tab:data}),
-and found substantial heterogeneity among the parameter estimates
-(\cref{fig:abchpd,fig:abchpdm2}). Two of the
-datasets~\autocite{niculescu2015recent, wang2015targeting} had estimated
-$\alpha$ values near unity for the prior allowing $m = 1$ (MAP estimates
-[95\% HPD] 
-  1.05 
-  [0.04 - 
-   1.27]
-and
-  0.84 
-  [0.01 -
-   1.02] respectively).
+   0.99] respectively).
 The MAP estimates did not change appreciably when $m = 1$ was disallowed by the
@@
 data was narrower
-  (0.04 - 
-   1.27).
+  (0.05 - 
+   1.18).
 When $m = 1$ was permitted, the \textcite{li2015hiv, cuevas2009hiv} both had
 low estimated $\alpha$ values
-  (0.06 
-  [0.01 - 
-  0.73]
+  (0.33 
+  [0 - 
+  0.76]
 and
-  0.19 
-  [0.01 -
-   0.8]). 
+  0.27 
+  [0 -
+   0.59]). 
 However, the MAP estimates increased when $m = 1$ was not permitted, although
 the HPD intervals remained roughly the same
-  (0.78 
+  (0.58 
+  [0.06 - 
+  0.99]
+and
+  0.48 
   [0.02 -
-  0.94]
-and
-  0.59 
-  [0.07 -
-   0.95]).
+   0.87]).
 The \textcite{novitsky2014impact} data had a fairly low estimated $\alpha$
 for both priors on $m$
-  (0.32 for $m \geq 1$;
-   0.39 for $m \geq 2$).
+  (0.55 for $m \geq 1$;
+   0.53 for $m \geq 2$).
 However, the confidence interval was much wider when $m = 1$ was allowed
-  ([0.04 -
-    1.62] for $m \geq 1$ vs.
+  ([0 -
+    1.75] for $m \geq 1$ vs.
     0 -
-    0.73 for $m \geq 2$).
-
-For all the datasets except \citeauthor{novitsky2014impact}, estimated values
+    1.75 for $m \geq 2$).}
+
+\add{Posterior mean point estimates for the preferential attachment power $\alpha$
+were all sub-linear, ranging from 
+    0.27
+(mixed/Spain) to
+    0.81
+(IDU/Estonia). When aggregated by risk group, the average estimates were
+    0.77
+for IDU,
+    0.41
+for primarily heterosexual risk, and
+    0.37
+for MSM. 95\% HPD intervals were very wide for most datasets, often
+encompassing nearly the entire range from 0 to 1 (\cref{fig:abchpd}).}
+
+\del{For all the datasets except \citeauthor{novitsky2014impact}, estimated values
 of $I$ were below 2000 when $m = 1$ was allowed, with relatively narrow HPD
 intervals compared to the nonzero prior density region
-  (\citeauthor{cuevas2009hiv}, 482 
-  [293 -
-   2111];
-   \citeauthor{niculescu2015recent}, 307
+  (\citeauthor{cuevas2009hiv}, 701 
+  [289 -
+   1279];
+   \citeauthor{niculescu2015recent}, 747
   [136 - 
-   2822];
-  (\citeauthor{li2015hiv}, 1183 
-  [413 -
-   2897];
-   \citeauthor{wang2015targeting}, 719
-  [176 - 
-   2114]).
+   2378];
+  \citeauthor{li2015hiv}, 1390 
+  [310 -
+   2821];
+   \citeauthor{wang2015targeting}, 675
+  [175 - 
+   1400]).
 The \citeauthor{novitsky2014impact} data was the outlier, with a very high
 estimated $I$, and HPD interval spanning almost the entire prior region
-  (7409 
-  [187 -
-   8819]).
+  (5431 
+  [183 -
+   8739]).
 The $I$ estimates and HPD intervals were generally robust to the choice of
 prior on $m$, with slightly narrower HPD intervals (compare
-\cref{fig:abchpd,fig:abchpdm2}).
-
-The MAP estimate of $m$ was equal to 1 for all but the
+\cref{fig:abchpd,fig:abchpdm2}).}
+
+\add{For all but the HET/Botswana data, the posterior mean estimates for
+$I$ were between
+    410
+(MSM/Taiwan) and
+    1391
+(MSM/Shanghai). The HET/Botswana data had a much higher estimated $I$ value
+    (5432)
+than the other datasets, with a very wide 95\% HPD interval covering almost the
+entire prior region (\cref{fig:abchpd}). There was no significant correlation
+between the number of sequences in the tree and the estimated $I$ value
+(Spearman correlation, 
+    $p =0.8$),
+indicating that the higher estimates were not simply due to increased sampling
+density. When both \textit{gag} and \textit{env} sequences were analyzed, the
+estimates from the \textit{env} data were higher (HET/Uganda,
+    939 for \textit{gag} vs.
+    1615 for \textit{env};
+HET/Malawi,
+    724 for \textit{gag} vs.
+    845 for \textit{env}).}
+
+\del{The MAP estimate of $m$ was equal to 1 for all but the
 \citeauthor{novitsky2014impact} data, when this value was allowed. However, the
 upper bound of the HPD interval was different for each dataset
-  (\citeauthor{niculescu2015recent}, 5;
-   \citeauthor{wang2015targeting}, 4;
+  (\citeauthor{niculescu2015recent}, 4;
+   \citeauthor{wang2015targeting}, 1;
    \citeauthor{li2015hiv}, 1;
-   \citeauthor{cuevas2009hiv}, 2).
+   \citeauthor{cuevas2009hiv}, 1).
 When $m = 1$ was disallowed, the MAP for all datasets was either 2 or 3, with
@@
 number of nodes $N$ were largely uninformative for all samples, with almost all
-MAP estimates greater than 5000 and HPD intervals spanning almost the entire
+MAP estimates greater than 7500 and HPD intervals spanning almost the entire
 nonzero prior density region. The only exception was the \citeauthor{li2015hiv}
 data, for which the MAP estimate was lower 
-  (6973)
-when $m = 1$ was allowed.
-
-\begin{figure*}[ht]
-  \centering
-  \includegraphics{realdata-hpd}
-  \vspace{8pt}
-  \caption{
-      Maximum \textit{a posteriori} point estimates and 95\% HPD intervals for
-      parameters of the BA network model, fitted to five published HIV datasets
-      with kernel-ABC.
-  }
-  \label{fig:abchpd}
-\end{figure*}
-
-\section{Discussion}
+  (5916)
+when $m = 1$ was allowed.}
+
+\add{The posterior means of $m$ were equal to one for 
+  zero
+of the datasets analyzed. The widths of the 95\% HPD intervals varied from 0
+(all the mass on the estimated value) to 5 (the entire prior region). Estimates
+of $N$ were  mostly uninformative, with very similar estimates for all datasets
+(mean
+    6188,
+range
+    5881 - 6882).
+This was similar to the pattern observed for the synthetic data, where the
+posterior mean always fell around the upper two-thirds mark of the range
+(\cref{fig:abcpt2}).}
+
+\add{When the value $m = 1$ was disallowed by the prior, the separation in
+$\alpha$ between the IDU datasets and the others became more striking
+(\cref{fig:abchpdm2}). All three IDU datasets had estimated $\alpha$ values at
+or above
+    1.
+The estimate for the MSM/Beijing data was slightly lower
+    (0.85) and
+the estimates for the seven remaining non-IDU datasets were bounded above by
+    0.58. 
+The values of $I$ were fairly robust to the choice of prior (compare
+\cref{fig:abchpd,fig:abchpdm2}), although the 95\% HPD intervals were slightly
+narrower (average width
+    2192
+for $m \geq 1$ and
+    1874
+for $m \geq 2$). The posterior means of $m$ for all but the HET/Botswana data
+took on the value 3 with this prior, with the HPD intervals spanning the entire
+prior region. This is very similar to the results observed for $m$ on simulated
+data (\cref{tab:abchpd}), and suggests that $m$ is not identifiable from these
+data with this prior. The results for $N$ did not change appreciably between
+the two choices of prior.}
+
+\section*{Discussion}
 
@@
 little2014using}. Despite this, few methods exist for investigating contact
-network parameters in a phylodynamic framework~\autocite{groendyke2011bayesian,
-volz2008sir, brown2011transmission, leventhal2012inferring}. Kernel-ABC is a
-model-agnostic method which can be used to investigate any quantity that
-affects tree shape~\autocite{poon2015phylodynamic}. In this work, we developed a
-kernel-ABC-based method to infer the parameters of a contact network model. The
-method is general, and could be applied to any model from which contact
-networks can be simulated. We demonstrated the method on the BA model,
-which is a simple pcreferential attachment model giving rise to the power law
-degree distributions commonly observed in real-world networks.
-
-By training a kernel-SVR classifier, we found that the $\alpha$ and $I$
-parameters, representing pcreferential attachment power and number of infected
-nodes, had a strong influence on tree shape. This was creflected in the relative
-accuracy of the kernel-ABC estimates of these parameters. The total number of
+network parameters in a phylodynamic framework \autocite[although see][for
+related work]{groendyke2011bayesian, volz2008sir, brown2011transmission,
+leventhal2012inferring}. ABC is a model-agnostic method which can be used to
+investigate any quantity that affects tree
+shape~\autocite{poon2015phylodynamic}. In this work, we developed a ABC-based
+method to infer the parameters of a contact network model. The method is
+general, and could be applied to any model from which contact networks can be
+simulated. We demonstrated the method on the BA model, which is a simple
+preferential attachment model. \add{For some parameter choices,} the BA model
+gives rise to the power law degree distributions commonly observed in
+real-world networks.
+
+\subsection*{Analysis of BA model with synthetic data}
+
+\del{By training a kernel-SVM classifier, we found that the $\alpha$ and $I$
+parameters, representing preferential attachment power and number of infected
+nodes, had a strong influence on tree shape. This was reflected in the relative
+accuracy of the ABC estimates of these parameters. The total number of
 nodes $N$ had a weak influence on tree shape, which was most prominent when the
@@
 did not produce much variation in tree shape, resulting in in both poorly
-performing classifiers and uninformative kernel-ABC estimates.
-
-$N$ was almost always significantly over-estimated using kernel-ABC. Since the
-prior on $N$ and $I$ is jointly uniform on a non-rectangular region ($I \leq
-N$), there is more prior mass on high $N$ values. In retrospect, it is
-unreasonable to expect good estimation of $N$, because adding more nodes to a
-BA network does not change the edge density or overall shape. This can be
+performing classifiers and uninformative ABC estimates.}
+
+\add{The preferential attachment power $\alpha$ had a strong influence on tree
+shape in the range of values we considered. Although the tree kernel was the
+most effective classifier for $\alpha$, a Sackin's index of tree imbalance
+performed nearly as well (\cref{fig:rsquared}). High $\alpha$ values produce
+networks with few well-connected ``superspreader'' nodes which are involved in
+a large number of transmissions, resulting in a highly unbalanced ladder-like
+tree structure. There appeared to be weaker identifiability for $\alpha < 1$
+than for $\alpha \geq 1$ (\cref{fig:abcpt,tab:abchpd}), which may be partially
+explained by the relationship between $\alpha$ and the power law exponent
+$\gamma$ (\cref{fig:gamma}). Although the degree distributions do not truly
+follow a power law for $\alpha \neq 1$, the fitted exponent still captures the
+shape of the degree distribution reasonably well (\cref{fig:powerlaw}). The
+$\gamma$ values fitted to $\alpha = 0$ and $\alpha = 0.5$ are nearly identical
+(about 2.28 for $\alpha = 0$ and 2.33 for $\alpha = 0.5$ with $N = 5000$ and $m
+= 2$). In other words, the degree distributions of networks with $\alpha < 1$
+are similar to each other, which may result in similarity of corresponding
+transmission trees as well.}
+
+\add{$I$, representing the number of infected individuals at the time of
+sampling, was also identifiable, albeit over-estimated with ABC for both values
+we considered. Sackin's index was better able to discern $I$ from tree shape
+than the nLTT (\cref{fig:rsquared,fig:Icrossv}), suggesting that this parameter
+impacts the distribution of branching times in the tree more than the topology.
+In a homogeneously-mixed population, branching times can be modelled by the
+coalescent process~\autocite{kingman1982coalescent}, in our case under the SI
+model~\autocite{volz2009phylodynamics}. Although networks are not homogeneously
+mixed, the forces which affect the distribution of branching times still apply.
+In our simulations, all discordant edges shared the same
+transmission rate, so that the waiting time until the next transmission in the
+entire network was always inversely proportional to the number of discordant
+edges. In the initial phase of the epidemic, when $I$ is small, each new
+transmission results in many new discordant edges. Hence, there is an early
+exponential growth phase, producing many short branches near the root of the
+tree. As the epidemic gets closer to saturating the network, the number of
+discordant edges decays, causing longer waiting times.}
+
+\add{The number of nodes in the network, $N$, exhibited the most variation in
+terms of its effect on tree shape. There was almost no measurable difference
+between trees simulated under different $N$ values when the number of infected
+nodes $I$ was small (\cref{fig:Ncrossv}).} In retrospect, it is unreasonable to
+expect good estimation of $N$, in many cases, because adding additional nodes
+does not change the edge density or overall shape of a BA network. This can be
 illustrated by imagining that we add a small number of nodes to a network after
-the epidemic simulation has already been completed. It is possible that none of
-these new nodes attains a connection to any infected node. Thus, running the
-simulation again on the new, larger network could produce the exact same
-transmission tree as before. We note also that our accurate estimates of $I$
-may have been influenced by this prior, which places more mass on low $I$
-values. However, the MAP estimate of $I$ was very high for the
-\textcite{novitsky2013phylogenetic,novitsky2014impact} data, suggesting that a
-strong enough signal in the data can overcome the prior.
+the epidemic simulation has already been completed. \add{If $I$ is small
+relative to $N$, very few of the infected nodes will gain any new neighbours.
+Thus, the outcome of a second simulation on the same network will likely be
+very similar.} \del{It is possible that none of these new nodes attains a
+connection to any infected node. Thus, running the simulation again on the new,
+larger network could produce the exact same transmission tree as before.}
+On the other hand, when $I$ is large relative to $N$, the coalescent dynamics
+discussed above also apply. The waiting times until the next infection
+increase, resulting in longer coalescence times toward the tips. The relative
+accuracy of the nLTT in these situations (\cref{fig:rsquared,fig:Ncrossv})
+corroborates this hypothesis, as the nLTT uses only information about the
+coalescence times. When all BA parameters were simultaneously estimated with
+ABC, $N$ was nearly always over-estimated by approximately a factor of two
+(\cref{fig:abcpt2,tab:abchpd}). One factor which may have contributed to this
+bias was our choice of prior distribution. Since the prior for $I$ and $N$ was
+jointly uniform on a region where $I \leq N$, more prior weight was assigned to
+higher $N$ values. We note also that \del{our accurate estimates of $I$ may
+have been influenced by} this prior\del{, which} places more mass on low $I$
+values. However, the estimate of $I$ was very high for the HET/Botswana data,
+suggesting that a strong enough signal in the data can overcome the prior.
+\add{Furthermore, when $I$ was estimated marginally with fixed $N$, the
+accuracy of the estimate improved even though there was no longer any extra
+prior mass on low $I$ values.}
+
+\add{Another possible contributing factor to the overestimation of $I$ and $N$
+relates to the dynamics of the SI model and the coalescent process. The number
+of infected individuals follows a logistic growth curve under the SI model.
+This kind of growth curve has three qualitative phases: a slow ramp-up, an
+exponential growth phase, and a slow final phase when the susceptible
+population is almost depleted. The waiting times until the next transmission,
+which determine the coalescence times in the tree, are dependent on the growth
+phase of the epidemic. Therefore, we hypothesize that it is the growth phase at
+the time of sampling which most affects tree shape, rather than the specific
+values of $I$ or $N$. To investigate this hypothesis, we simulated transmission
+trees over networks on a grid of $I$ and $I$ values and fit logistic growth
+curves to the proportion of infected individuals over time. We and calculated
+the first and second derivatives of these curves (with respect to time) at the
+time of transmission tree sampling. As shown in \cref{fig:derivs}, there are
+bands along which both derivatives are similar which contain the values we
+tested. These bands span mostly higher values of $N$ and $I$ than the true
+values. Thus, if $N$ and $I$ are free to vary (as is the case in ABC), both
+parameters will tend to be overestimated due to being less identifiable within
+their own band. However, when $N$ is fixed at 5000, the derivatives vary
+substantially along the $I$-axis, which explains why a marginal estimate of $I$
+was more accurate (\cref{fig:marginal}). We also note the resemblance of the
+contour surface of \cref{fig:derivs} to the two-dimensional marginal posterior
+distribution on $I$ and $N$ obtained with simulated data (\cref{fig:abc2}).}
+
+\add{The $m$ parameter, which controls the number of connections added to the
+network per vertex, did not have a measurable impact on tree shape and was not
+identifiable with ABC. It was pointed out to us by an anonymous reviewer that
+for a fixed $I$, an infected node may only end up transmitting along a fraction
+of its outgoing edges, which could mask the presence of the extra edges
+associated with higher $m$. If $m$ were a continuous variable, it is possible
+that we would observe stronger identifiability for lower values (say between 0
+and 2), where extra edges are more likely to be involved in the epidemic and
+have an impact on the transmission tree. One way to achieve this would be to
+draw $m$ separately for each node from a distribution parameterized by a
+continuous variable.}
 
@@
 parameters may translate to highly informative priors on quantities of
-interest. We observed a non-linear relationship between the pcreferential
+interest. We observed a non-linear relationship between the preferential
 attachment power $\alpha$ and the power law exponent $\gamma$
-(\cref{fig:gamma}). Thecrefore, placing a uniform prior on $\alpha$ between 0
+(\cref{fig:gamma}). Therefore, placing a uniform prior on $\alpha$ between 0
 and 2 is equivalent to placing an informative prior that $\gamma$ is close to
-2. Thecrefore, if we were primarily interested in $\gamma$ rather than
-$\alpha$, a more sensible choice of prior might have a shape similar to the
-inverse of \cref{fig:gamma} and be bounded above by approximately $\alpha$ =
-1.5. This would uniformly bound $\gamma$ in the region $2 \leq \gamma \leq 4$
-commonly reported in the network literature~\autocite{liljeros2001web,
+2. Therefore, if we were primarily interested in $\gamma$ rather than
+$\alpha$, a more sensible choice of prior might have a shape informed by
+\cref{fig:gamma} and be bounded above by approximately $\alpha$ = 1.5. This
+would uniformly bound $\gamma$ in the region $2 \leq \gamma \leq 4$ commonly
+reported in the network literature~\autocite{liljeros2001web,
 schneeberger2004scale, colgate1989risk, brown2011transmission}. We note however
 that \textcite{jones2003assessment} estimated $\gamma$ values greater than
-four, in one case as high as 17, for some datasets, indicating that a wider
+four for some datasets, in one case as high as 17, indicating that a wider
 range of permitted $\gamma$ values may be warranted.
+
+\subsection*{Analysis of real world HIV datasets}
 
@@
 most likely misspecified for these data. In particular, the average degree of a
-node in the network is equal to $2m$, and thecrefore is constrained to be a
+node in the network is equal to $2m$, and therefore is constrained to be a
 multiple of 2. Furthermore, we considered the case $m = 1$, where the network
-has no cycles, to be implausible and thecrefore assigned it zero prior
+has no cycles, to be implausible and therefore assigned it zero prior
 probability in one set of analyses. This forced the average degree to be at
 least four, which may be unrealistically high for sexual networks. The fact
-that the estimated values of $\alpha$ differed substatially for three datasets
-depending on whether or not $m = 1$ was allowed by the prior is futher evidence
-of this potential misspecification. However, we note that for two of the
-datasets, the estimated values of $\alpha$ did not change much between priors,
-and the estimates of $I$ were robust to the choice of prior for all datasets
-studied. More sophisticated models, for example models incorporating
-heterogeneity in node behaviour, are likely to provide a better fit to these
-data.
-
-With respect to the pcreferential attachment power $\alpha$, the five datasets
-analysed fell into three categories (\cref{fig:abchpd}). First, we
-estimated a pcreferential attachment power close to 1, indicating linear
-pcreferential attachment, for the outbreaks studied by
+that the estimated values of $\alpha$ differed substantially for three datasets
+depending on whether or not $m = 1$ was allowed by the prior is further evidence
+of this potential misspecification. However, we note that \add{the ordering of
+the datasets with respect to $\alpha$ was similar between the two priors,}
+\del{for two of the datasets, the estimated values of $\alpha$ did not change
+much between priors,} and the estimates of $I$ were robust to the choice of
+prior for all datasets studied (compare \cref{fig:abchpd,fig:abchpdm2}). More
+sophisticated models, for example models incorporating heterogeneity in node
+behaviour, are likely to provide a better fit to these data.
+
+\del{With respect to the preferential attachment power $\alpha$, the five
+datasets analyzed fell into three categories (\cref{fig:abchpd}). First, we
+estimated a preferential attachment power close to 1, indicating linear
+preferential attachment, for the outbreaks studied by
 \textcite{niculescu2015recent} and \textcite{wang2015targeting}. These values
@@
 which we would expect some of the assumptions of the BA model, such as a
-connected network, relatively high mean degree, and pcreferential attachment
-dynamics, to hold.
-
-The remaining three datasets (\textcite{cuevas2009hiv, novitsky2014impact,
+connected network, relatively high mean degree, and preferential attachment
+dynamics, to hold.}
+
+\del{The remaining three datasets (\textcite{cuevas2009hiv, novitsky2014impact,
 li2015hiv}) had estimated values of $\alpha$ below 0.5 when $m = 1$ was
@@
 the \textcite{li2015hiv} data, as the subset we analyzed formed a phylogenetic
-cluster and thecrefore was a good candidate for the BA model. However, nearly
+cluster and therefore was a good candidate for the BA model. However, nearly
 all the posterior density was assigned to $m = 1$ when this value was allowed,
-indicating that the network was more likely to have an acyclic tree structure.
+indicating that the network was more likely to have an acyclic tree structure.}
+
+\subsubsection*{Preferential attachment power is sub-linear and higher for IDU networks}
+
+\add{For all datasets we examined, the posterior mean estimates for $\alpha$
+were sub-linear, ranging from 
+    0.27
+to 
+    0.81.
+The sub-linearity is consistent with the results of
+\textcite{de2007preferential}, who developed a statistical inference method to
+estimate the parameters of a more sophisticated preferential attachment model
+incorporating heterogeneous node behaviour. When used to analyze
+population-level longitudinal partner count data, they found $\alpha$ values
+ranging from 0.26 to 0.62 depending on the gender and time period considered.}
+
+\add{Both \textcite{de2007preferential} and the HET/Botswana data studied
+populations whose primary risk factor for HIV infection was heterosexual
+contact. \citeauthor{de2007preferential} explicitly excluded reported
+homosexual contacts; \citeauthor{novitsky2014impact} did not, but noted that
+heterosexual contact is the primary mode of transmission in Botswana where the
+study was done. In the first of the two papers describing the Botswana
+study~\autocite{novitsky2013phylogenetic}, the authors noted that their sample
+was gender-biased, being composed of approximately 75\% women. Our estimate of
+$\alpha$ for these data was 
+    0.55
+        or 
+    0.53,
+depending on the prior on $m$. Similarly, \textcite{de2007preferential}
+estimated 0.54, 0.57, and 0.29 for 3-year, 5-year, and lifetime partnership
+networks respectively for the female portion of their sample.}
+
+\add{The datasets derived from IDU populations had a higher estimated
+preferential attachment power than the other datasets
+(\cref{fig:abchpd,fig:abchpdm2}). This finding is in line with
+\textcite{dombrowski2013topological}, who reanalyzed a network of IDUs in
+Brooklyn, USA, collected between 1991 and 1993~\autocite{friedman2006social}.
+They found that the IDU network resembled a BA network much more closely than
+other social and sexual networks, and offered sociological explanations for the
+apparent preferential attachment dynamics in this population. Importantly, from
+a public health perspective, the authors asserted that the removal of
+\emph{random} individuals from IDU networks may have the paradoxical effect of
+increasing the network's epidemic susceptibility. When low-degree nodes are
+removed, as would occur during a police crackdown, their network neighbours may
+turn to well-known community members for advice or supplies, thus increasing
+the connectivity of these high-degree nodes.}
+
+\add{Unfortunately, the sub-linear region for $\alpha$ identified by both
+\textcite{de2007preferential} and \textit{netabc} is also the region of
+poorest identifiability (\cref{fig:abcpt}). This was reflected in the high
+level of uncertainty in the estimates, with most 95\% HPD intervals covering
+the majority of the range [0, 1]. The value $\alpha = 0.5$ was contained in the
+95\% HPD interval for every dataset; consequently, it is not possible to say
+with high confidence that any of the $\alpha$ values are different from each
+other. In synthetic data, the confidence intervals around $\alpha$ narrowed
+when other parameters were marginalized out (\cref{fig:marginal}). Thus, it is
+possible that estimates of $\alpha$ could be made more precise by specifying
+either exact values or informative priors on the other BA parameters when these
+are known.}
+
+\add{The results we have just discussed pertain to the \textit{gag} sequences
+for the two datasets with both \textit{gag} and \textit{env} data. As shown in
+\cref{fig:genes}, the estimates of $\alpha$ were quite robust to the gene
+analyzed. For the two datasets we reanalyzed using
+RAxML~\autocite{stamatakis2014raxml} and LSD~\autocite{to2016fast}, $\alpha$
+was robust to the choice of method (\cref{fig:methods}).}
+
+\subsubsection*{Other BA parameters}
+
+\add{The true HIV prevalence in a population can be difficult to estimate for
+several reasons. HIV-infected individuals may be asymptomatic for months or
+years, possibly delaying their awareness of their status. In many contexts, the
+risk factors for acquisition of HIV are illegal or stigmatized, which may
+represent a barrier to testing, treatment, and/or disclosure of status.
+Our simulation study showed that $I$ is weakly identifiable from tree shapes,
+however, the estimates of $I$ obtained with \textit{netabc} were upwardly
+biased (\cref{fig:abcpt}). In addition, our initial exploratory analysis showed
+that the identifiability of $I$ decreases with the number of sampled tips
+(\cref{fig:Icrossv}); in real world studies, the proportion of infected
+individuals sampled is usually low. The estimated $I$ values for the HIV
+datasets ranged from
+    410
+    (IDU/Estonia)
+to
+    5432
+    (HET/Botswana).
+The estimates of $I$ were less robust to the HIV gene analyzed
+(\cref{fig:genes}) and the phylogenetic reconstruction method used
+(\cref{fig:methods}).}
+
+\add{We were not able to discern and trends toward over- or underestimation of
+$I$ from available prevalence data. For example, the authors of the
+HET/Botswana data~\autocite{novitsky2013phylogenetic, novitsky2014impact}
+estimated that there were 1731 HIV-positive individuals in the study
+area. HIV sequences were obtained from approximately 70\% of these individuals.
+The estimated prevalence we obtained was much higher 
+    (5432),
+with a 95\% HPD interval spanning nearly the entire prior region. On the other
+hand, the study which produced the MSM/USA data~\autocite{little2014using}
+enrolled 648 HIV-positive MSM; thus, our estimate of $I$
+    (482)
+was clearly an underestimate. Post-hoc explanations can be imagined for both of
+these results. The HET/Botswana data were collected from a town located
+proximally to the country's capital city; the authors suggested that frequent
+travel between the two locations may have facilitated linking of their sexual
+networks. Thus, the high estimate of $I$ we obtained for these data may
+include a larger network component based in the capital city. The MSM/USA
+result may indicate that the individuals genotyped for the study are members of
+a smaller subnetwork which does not include the entire local MSM population.
+Unfortunately, none of these hypotheses can be easily tested.}
+
+\add{Over half of the datasets were estimated to have $m = 1$, which produces
+tree-like networks without cycles. Since the average degree of a
+Barab\'asi-Albert network is $2m$, this value may simply reflect the fact that
+most people have a small number of sexual partners, especially when only
+recent partnerships are considered~\autocite{liljeros2001web}. In fact, in one
+survey, the most common number of partnerships in the past twelve months was
+one~\autocite{liljeros2001web}; the BA model does not allow any nodes with
+degree 1 when $m \geq 2$. For this reason, the choice of whether to allow $m =
+1$ in the prior is problematic, as we must choose between an unrealistic
+topology (no cycles) and an unrealistic minimum degree. Extensions to the BA
+model which relax this constraint can be imagined and may offer improved
+parameter resolution. Estimates of $N$ were not informative for any of the
+datasets under either choice of prior, consistent with our simulation results.}
+
+\subsubsection*{Modelling assumptions}
 
@@
 modelling assumptions include the network being connected and static, all
-transmission rates being equal, no removal after infection, and identical
-behaviour of all nodes. This last is particularly problematic, as we showed by
-simulating a network where some nodes exhibited a higher attachment power than
-others. The estimated attachment power was simply the average of the two
-values, indicating that, although we could characterize the network in
-aggregate, the estimated parameters could not be said to apply to any
-individual node. Despite these issues, we felt it was best to demonstrate the
-method first on a simple model. It is possible to use this framework to fit
-more complex models which address some of these issues, such as one
-incorporating heterogeneous node behaviour, which may prove a fruitful avenue
+transmission rates being equal, no removal after infection, identical behaviour
+of all nodes, and random sampling. The last two were addressed with small-scale
+experiments. We simulated a network where some nodes exhibited a higher
+attachment power than others, and found that the estimated attachment power was
+simply the average of the two values. This indicated that, although we could
+characterize the network in aggregate, the estimated parameters could not be
+said to apply to any individual node. The effect of biased sampling was
+investigated by analyzing a transmission tree which had been sampled in a
+peer-driven fashion. The results were roughly in line with those for random
+sampling, however the estimated value of $\alpha$ was lower than the average
+for randomly-sampled trees. Further experiments would be necessary to fully
+explore the impact of these assumptions on the method's accuracy. However,
+despite these issues, we felt it was best to demonstrate the method first on a
+simple model. It is possible to use this framework to fit more complex models
+which address some of these issues, such as one incorporating heterogeneous
+node behaviour, which may prove a fruitful avenue
 for future investigations.
@@
 significantly~\autocite{ypma2013relating} due to within-host evolution and the
-sampling process. In addition, the ABC-SMC algorithm is
+sampling process. \add{Progress has very recently been made on this front by
+\textcite{giardina2016inference}, who have concurrently developed an ABC-based
+method incorporating within-host evolution.} The ABC-SMC algorithm is
 computationally intensive, taking about a day when run on 20 cores in parallel
@@
 work, and previous work by our group~\autocite{poon2015phylodynamic}, has
-demonstrated that kernel-ABC is a broadly applicable and effective framework in
-which to perform phylodynamic inference.
-
-\section{Acknowledgements}
-
-This work was supported by grants from the Canadian Institutes of Health
-Research (CIHR, operating grant HOP-111406), and the Bill \& Melinda Gates
-Foundation (award number OPP1110049). R.M.M. was supported by a scholarship
-from the CIHR Strategic Training Program in Bioinformatics. A.F.Y.P. was
-supported by a CIHR New Investigator Award (Canadian HIV Vaccine Initiative,
-Vaccine Discovery and Social Research) and by a Career Investigator Scholar
-Award from the Michael Smith Foundation for Health Research, in partnership
-with the Providence Health Care Research Institute and St. Paul's Hospital
-Foundation.
-
+demonstrated that ABC is a broadly applicable and effective framework in which
+to perform phylodynamic inference.
+
+\section*{Acknowledgements}
+
+\add{We are grateful to Dr. Sally Otto, Dr. Alexandre Bouchard-C\^ot\'e, 
+Dr. Richard Harrigan, and the two anonymous reviewers for many helpful
+suggestions.} This work was supported by grants from the Canadian Institutes of
+Health Research (CIHR, operating grant HOP-111406), and the Bill \& Melinda
+Gates Foundation (award number OPP1110049). R.M.M. was supported by a
+scholarship from the CIHR Strategic Training Program in Bioinformatics.
+A.F.Y.P. was supported by a CIHR New Investigator Award (Canadian HIV Vaccine
+Initiative, Vaccine Discovery and Social Research) and by a Career Investigator
+Scholar Award from the Michael Smith Foundation for Health Research, in
+partnership with the Providence Health Care Research Institute and St. Paul's
+Hospital Foundation.
+
+\clearpage
+\newpage
 \printbibliography
-
-\newpage
 
@@
 
-\begin{figure}[ht]
-  \centering
-  \includegraphics{abc-point-estimate-m3.pdf}
-  \caption{
-    Point estimates of BA model parameters obtained by running kernel-ABC
-    on simulated phylogenies without training, for simulations with $m = 3$.
-    Dotted lines indicate true values, and limits of the $y$-axes are regions
-    of uniform prior density. (A) Estimates for $\alpha$ and $I$ against their
-    true values in simulations. (B) Estimates for $m$ and $N$, which were held
-    fixed in these simulations, against true values of $\alpha$.
-  }
-  \label{fig:abcptm3}
-\end{figure}
-
-\begin{figure}[ht]
-  \centering
-  \includegraphics{abc-point-estimate-m4.pdf}
-  \caption{
-    Point estimates of BA model parameters obtained by running kernel-ABC
-    on simulated phylogenies without training, for simulations with $m = 4$.
-    Dotted lines indicate true values, and limits of the $y$-axes are regions
-    of uniform prior density. (A) Estimates for $\alpha$ and $I$ against their
-    true values in simulations. (B) Estimates for $m$ and $N$, which were held
-    fixed in these simulations, against true values of $\alpha$.
-  }
-  \label{fig:abcptm4}
+\begin{figure}
+    \includegraphics[width=\textwidth]{abc-boxplot-mN}
+    \caption[
+        Posterior mean point estimates for BA model parameters $m$ and $N$
+        obtained by running \textit{netabc} on simulated data, stratified by
+        true parameter values.
+    ]{
+        Posterior mean point estimates for BA model parameters $m$ and $N$
+        obtained by running \textit{netabc} on simulated data, stratified by
+        true parameter values. First row of plots contains true versus
+        estimated values of $m$; second row contains true versus estimated
+        values of $N$. Columns are stratified by $\alpha$, $I$, and $m$
+        respectively. Dashed lines indicate true values. 
+    }
+    \label{fig:abcpt2}
+\end{figure}
+
+\begin{figure}
+    \includegraphics[width=\textwidth]{abc-marginal}
+    \caption[
+        Posterior means, 50\% HPD intervals, and 95\% HPD intervals for BA
+        model parameters estimated marginally with ABC.
+    ]{
+        Posterior means (points), 50\% HPD intervals (notches), and 95\% HPD
+        intervals (lines) for BA model parameters estimated marginally with
+        ABC. $x$-axis labels indicate parameters which were fixed to their true
+        values by specifying Dirac-delta priors.
+    }
+    \label{fig:marginal}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.7\textwidth]{better-smc}
+    \caption[
+        Posterior means, 50\% HPD intervals, and 95\% HPD intervals for BA
+        model parameters estimated with ABC using two sets of SMC settings.
+    ]{
+        Posterior means (points), 50\% HPD intervals (notches), and 95\% HPD
+        intervals (lines) for BA model parameter estimated with ABC using two
+        sets of SMC settings. ``Low'' settings are 1000 particles, 5 simulated
+        datasets per particle, and $\alpha_{\text{ESS}} = 0.95$. ``High''
+        settings are 2000 particles, 10 simulated datasets per particle, and
+        $\alpha_{\text{ESS}} = 0.97.$
+    }
+    \label{fig:better}
 \end{figure}
@@
   \label{fig:gamma}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \includegraphics{powerlaw.pdf}
+    \caption{
+        Best fit power law and stretched exponential curves for degree
+        distributions of simulated BA networks for several values of $\alpha$
+        and $m$.
+    }
+    \label{fig:powerlaw}
 \end{figure}
@@
       parameters of the BA network model, fitted to five published HIV datasets
-      with kernel-ABC. Regions of nonzero prior density are shown on $x$-axes.
+      with ABC. $x$-axes indicate regions of nonzero prior density. In
+      particular, the prior on $m$ was DiscreteUniform(2, 5).
   }
@@
 
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{realdata-genes}
+    \caption[
+        Posterior means and 50\%/95\% HPD intervals for parameters of the
+        BA network model, fitted to two HIV datasets where both \textit{gag}
+        and \textit{env} genes were sequenced.
+    ]{
+        Posterior means (points), 50\% HPD intervals (notches), and 95\%
+        HPD intervals (lines) for parameters of the BA network model, fitted to
+        two HIV datasets where both \textit{gag} and \textit{env} genes were
+        sequenced.
+    }
+    \label{fig:genes}
+\end{figure}
+
 \begin{figure}[ht]
-  \includegraphics{cuevas2009-posterior}
-  \caption{
-      Posterior distribution of BA model parameters for
-      \textcite{cuevas2009hiv} data. Vertical lines indicate maximum \textit{a
-      posteriori} estimates, and shaded areas are 95\% highest posterior
-      density intervals. $x$-axis indicates regions of nonzero prior density.
-  }
-  \label{fig:cuevas}
+    \centering
+    \includegraphics[width=\textwidth]{realdata-methods}
+    \caption[
+        Posterior means and 50\%/95\% HPD intervals for parameters of the
+        BA network model, fitted to two HIV datasets using two phylogenetic
+        reconstruction methods.
+    ]{
+        Posterior means (points), 50\% HPD intervals (notches), and 95\%
+        HPD intervals (lines) for parameters of the BA network model, fitted to
+        two HIV datasets using two phylogenetic reconstruction methods.
+    }
+    \label{fig:methods}
 \end{figure}
@@
 \begin{figure}[ht]
-  \includegraphics{li2015-posterior}
-  \caption{
-      Posterior distribution of BA model parameters for \textcite{li2015hiv}
-      data. Vertical lines indicate maximum \textit{a posteriori} estimates,
-      and shaded areas are 95\% highest posterior density intervals. $x$-axis
-      indicates regions of nonzero prior density.
-  }
-  \label{fig:li}
-\end{figure}
-
-\begin{figure}[ht]
-  \includegraphics{niculescu2015-posterior}
-  \caption{
-      Posterior distribution of BA model parameters for
-      \textcite{niculescu2015recent} data. Vertical lines indicate maximum
-      \textit{a posteriori} estimates, and shaded areas are 95\% highest
-      posterior density intervals. $x$-axis indicates regions of nonzero prior
-      density.
-  }
-  \label{fig:niculescu}
-\end{figure}
-
-\begin{figure}[ht]
-  \includegraphics{novitsky2014-posterior}
-  \caption{
-      Posterior distribution of BA model parameters for
-      \textcite{novitsky2013phylogenetic, novitsky2014impact} data. Vertical
-      lines indicate maximum \textit{a posteriori} estimates, and shaded areas
-      are 95\% highest posterior density intervals. $x$-axis indicates regions
-      of nonzero prior density.
-  }
-  \label{fig:novitsky}
-\end{figure}
-
-\begin{figure}[ht]
-  \includegraphics{wang2015-posterior}
-  \caption{
-      Posterior distribution of BA model parameters for
-      \textcite{wang2015targeting} data. Vertical lines indicate maximum
-      \textit{a posteriori} estimates, and shaded areas are 95\% highest
-      posterior density intervals. $x$-axis indicates regions of nonzero prior
-      density.
-  }
-  \label{fig:wang}
+  \centering
+  \includegraphics{derivatives.pdf}
+  \caption[
+    First and second time derivatives of epidemic growth curves at time of
+    sampling for various values of $I$ and $N$.
+  ]{
+    First and second \add{time} derivatives of epidemic growth curves at time
+    of sampling for various values of $I$ and $I$. Networks were simulated
+    under the BA model with $\alpha = 1.0$, $m = 2$, and $N$ varied along the
+    values shown on the $x$-axis. Transmission trees were sampled at the time
+    when $I$ nodes were infected ($y$-axis). Logistic growth curves were fit to
+    epidemic trajectories derived from the transmission trees, and their first
+    and second derivatives were calculated at the time of sampling. Contours
+    show first derivatives, while colours indicate second derivatives. Values
+    of $I$ and $N$ used in simulation experiments with ABC are indicated by
+    diamonds.
+  }
+  \label{fig:derivs}
 \end{figure}

### 2016-07-14 doc/paper/changes.tex
Finished reviewer things finally ugh

@@
 regression does not result in large errors in phylodynamic inference (Poon
-2015). We reanalysed two datasets (Niculescu et al. 2015 and Cuevas et al.
-2009) using RAxML and LSD. The results for $\alpha$ were nearly identical, but
-$I$ was not robust to the reconstruction method -- RAxML+LSD produced higher
-$I$ estimates for both datasets (Figure S13). This result has been added to the
-discussion of both parameters.
+2015). We reanalysed the two datasets with the highest and lowest estimated
+$\alpha$ values (Zetterberg et al. 2004 and Cuevas et al. 2009) using RAxML and
+LSD. The results for $\alpha$ were quite similar, but $I$ was less robust to
+the reconstruction method -- RAxML+LSD produced higher $I$ estimates for both
+datasets (Figure S13). This result has been added to the discussion of both
+parameters.
 
-Rather than analyze bootstrap replicates, we used the two new datasets
-containing both \textit{gag} and \textit{env} sequences for within-dataset
-replication.
+We also analyzed five bootstrap replicate alignments for each of these
+datasets (Figure S15). The posterior means for the replicates were in the same
+region as the posterior mean for the original alignment. For $\alpha$, the
+means spanned a region of width approximately 0.3 (about 15\% of the prior
+region); for $I$, the range was only about 5\% of the prior region. Two new
+datasets containing both \textit{gag} and \textit{env} sequences provide
+additional within-dataset replication.
 

### 2016-07-14 doc/paper/paper.Rnw
Finished reviewer things finally ugh

@@
 \add{To check if our results were robust to the choice of phylogenetic
-reconstruction method, we built and reanalyzed phylogenies for two of the
-datasets (mixed/Spain and IDU/Romania) with
-\textit{RAxML}~\autocite{stamatakis2014raxml} with the GTR+$\Gamma$ model of
-sequence evolution and rate heterogeneity. The trees were rooted and
-time-scaled with \textit{Least Square
-Dating}~\autocite[\textit{LSD},][]{to2016fast}.}
+reconstruction method, we built and reanalyzed phylogenies for the datasets
+with the lowest and highest estimated $\alpha$ values (mixed/Spain and
+IDU/Estonia) with \textit{RAxML}~\autocite{stamatakis2014raxml} with the
+GTR+$\Gamma$ model of sequence evolution and rate heterogeneity. The trees were
+rooted and time-scaled with \textit{Least Square
+Dating}~\autocite[\textit{LSD},][]{to2016fast}. For expediency, the analysis
+was run with the prior $m \sim \text{DiscreteUniform}(2, 5)$, which defines a
+smaller total search space than the prior allowing $m = 1$. For both of these
+datasets, we also analyzed five bootstrap replicate alignments generated by
+resampling alignment columns with replacement.}
 
@@
 \add{except that we used 10000 particles instead of 1000 to increase the
-accuracy of the estimated posterior. This was computationally feasible due to
-the small number of runs required for this analysis.}
+accuracy of the estimated posterior for all analyses except the bootstrap
+replicates. This was computationally feasible due to the small number of runs
+required for this analysis.}
 
@@
 basic settings and 124 with the higher settings. The results of the two
-simulations were very similar for all parameters.}
+simulations were similar, but surprisingly, the results with higher SMC
+settings were slightly worse (by
+    \Sexpr{ round((better.alpha - base.alpha) * 100 / better.alpha) }\%,
+    \Sexpr{ round((better.I - base.I) * 100 / better.I) }\%, and
+    \Sexpr{ round((better.N - base.N) * 100 / better.N) }\%
+for $\alpha$, $I$, and $N$ respectively). However, the 50\% HPD interval for
+$I$ was closer to the true value of 2000 with the improved settings
+    (\Sexpr{ d[nparticle == 2000, round(I.lower.50)] } - 
+     \Sexpr{ d[nparticle == 2000, round(I.upper.50)] }, vs.
+     \Sexpr{ d[nparticle == 1000, round(I.lower.50)] } - 
+     \Sexpr{ d[nparticle == 1000, round(I.upper.50)] } with basic settings).
+The estimate of $m$, 3 in both cases, was unaffected by the settings.}
 
@@
 
-  stopifnot(pmean1[,dataset[which.min(I)]] == "kao2011")
+  stopifnot(pmean1[,dataset[which.min(I)]] == "zetterberg2004")
   stopifnot(pmean1[,dataset[which.max(I)]] == "novitsky2014")
@@
 
+<<raxml, include=FALSE>>=
+    d.raxml <- fread(Sys.glob("../../simulations/aggregates/raxml-hpd/*"))
+@
+
 \add{Posterior mean point estimates for the preferential attachment power $\alpha$
@@
 for MSM. 95\% HPD intervals were very wide for most datasets, often
-encompassing nearly the entire range from 0 to 1 (\cref{fig:abchpd}).}
+encompassing nearly the entire range from 0 to 1 (\cref{fig:abchpd}). As shown
+in \cref{fig:genes}, the estimates of $\alpha$ were quite robust to the gene
+analyzed.}
 
@@
 $I$ were between
-    \Sexpr{pmean1["kao2011", round(I)]}
-(MSM/Taiwan) and
+    \Sexpr{pmean1["zetterberg2004", round(I)]}
+(IDU/Estonia) and
     \Sexpr{pmean1["li2015", round(I)]}
@@
 $\alpha$ between the IDU datasets and the others became more striking
-(\cref{fig:abchpdm2}). All three IDU datasets had estimated $\alpha$ values at
-or above
+(\cref{fig:abchpdm2}). Both IDU datasets had estimated $\alpha$ values at or
+above
     \Sexpr{pmean2[c("zetterberg2004", "niculescu2015"), round(min(alpha), 2)]}.
@@
 the two choices of prior.}
+
+<<bootstrap, include=FALSE>>=
+    d.boot <- fread(Sys.glob("../../simulations/aggregates/bootstrap-hpd/*"))
+@
+
+\add{For the two datasets we reanalyzed using
+RAxML~\autocite{stamatakis2014raxml} and LSD~\autocite{to2016fast}, $\alpha$
+was relatively robust to the choice of method (\cref{fig:methods}, posterior
+means 
+    \Sexpr{pmean2["cuevas2009", round(alpha, 2)]}
+vs. 
+    \Sexpr{d.raxml[dataset == "cuevas2009" & parameter == "alpha", round(mean, 2)]}
+for mixed/Spain and 
+    \Sexpr{pmean2["zetterberg2004", round(alpha, 2)]}
+vs.
+    \Sexpr{d.raxml[dataset == "zetterberg2004" & parameter == "alpha", round(mean, 2)]}
+for IDU/Estonia). However, the estimates of $I$ were about twice as high when
+RAxML was used instead of FastTree to reconstruct the trees 
+    (\Sexpr{pmean2["zetterberg2004", round(I)]} 
+     vs. \Sexpr{d.raxml[dataset == "zetterberg2004" & parameter == "I", round(mean)]} for IDU/Estonia, 
+    \Sexpr{pmean2["cuevas2009", round(I)]}
+    vs. \Sexpr{d.raxml[dataset == "cuevas2009" & parameter == "I", round(mean)]} for mixed/Spain). 
+\Cref{fig:boot} shows estimates obtained for five bootstrap replicate
+alignments for each of these two datasets. For the mixed/Spain data, the
+estimated posterior mean [range of bootstrap posterior means] was
+    \Sexpr{pmean2["cuevas2009", round(alpha, 2)]}
+    [\Sexpr{d.boot[dataset == "cuevas2009" & parameter == "alpha" & replicate < 5, round(min(mean), 2)]} -
+     \Sexpr{d.boot[dataset == "cuevas2009" & parameter == "alpha" & replicate < 5, round(max(mean), 2)]}]
+for $\alpha$, 
+    \Sexpr{pmean2["cuevas2009", round(I)]}
+    [\Sexpr{d.boot[dataset == "cuevas2009" & parameter == "I" & replicate < 5, round(min(mean))]} -
+     \Sexpr{d.boot[dataset == "cuevas2009" & parameter == "I" & replicate < 5, round(max(mean))]}] for $I$,
+    \Sexpr{pmean2["cuevas2009", round(m, 2)]}
+    [\Sexpr{d.boot[dataset == "cuevas2009" & parameter == "m" & replicate < 5, round(min(mean), 2)]} -
+     \Sexpr{d.boot[dataset == "cuevas2009" & parameter == "m" & replicate < 5, round(max(mean), 2)]}] for $m$, and
+    \Sexpr{pmean2["cuevas2009", round(N)]}
+    [\Sexpr{d.boot[dataset == "cuevas2009" & parameter == "N" & replicate < 5, round(min(mean))]} -
+     \Sexpr{d.boot[dataset == "cuevas2009" & parameter == "N" & replicate < 5, round(max(mean))]}] for $N$. 
+For the IDU/Estonia data, values were
+    \Sexpr{pmean2["zetterberg2004", round(alpha, 2)]}
+    [\Sexpr{d.boot[dataset == "zetterberg2004" & parameter == "alpha" & replicate < 5, round(min(mean), 2)]} -
+     \Sexpr{d.boot[dataset == "zetterberg2004" & parameter == "alpha" & replicate < 5, round(max(mean), 2)]}]
+for $\alpha$, 
+    \Sexpr{pmean2["zetterberg2004", round(I)]}
+    [\Sexpr{d.boot[dataset == "zetterberg2004" & parameter == "I" & replicate < 5, round(min(mean))]} -
+     \Sexpr{d.boot[dataset == "zetterberg2004" & parameter == "I" & replicate < 5, round(max(mean))]}] for $I$,
+    \Sexpr{pmean2["zetterberg2004", round(m, 2)]}
+    [\Sexpr{d.boot[dataset == "zetterberg2004" & parameter == "m" & replicate < 5, round(min(mean), 2)]} -
+     \Sexpr{d.boot[dataset == "zetterberg2004" & parameter == "m" & replicate < 5, round(max(mean), 2)]}] for $m$, and
+    \Sexpr{pmean2["zetterberg2004", round(N)]}
+    [\Sexpr{d.boot[dataset == "zetterberg2004" & parameter == "N" & replicate < 5, round(min(mean))]} -
+    \Sexpr{d.boot[dataset == "zetterberg2004" & parameter == "N" & replicate < 5, round(max(mean))]}] for $N$.}
 
@@
 least four, which may be unrealistically high for sexual networks. The fact
-that the estimated values of $\alpha$ differed substantially for three datasets
+that the estimated values of $\alpha$ differed substantially for several datasets
 depending on whether or not $m = 1$ was allowed by the prior is further evidence
@@
 
-\add{The results we have just discussed pertain to the \textit{gag} sequences
-for the two datasets with both \textit{gag} and \textit{env} data. As shown in
-\cref{fig:genes}, the estimates of $\alpha$ were quite robust to the gene
-analyzed. For the two datasets we reanalyzed using
-RAxML~\autocite{stamatakis2014raxml} and LSD~\autocite{to2016fast}, $\alpha$
-was robust to the choice of method (\cref{fig:methods}).}
-
 \subsubsection*{Other BA parameters}
@@
 <<realdata_I, include=FALSE>>=
-    #stopifnot(d1[parameter == "I", dataset[which.min(mean)]] == "zetterberg2004")
+    stopifnot(d1[parameter == "I", dataset[which.min(mean)]] == "zetterberg2004")
     stopifnot(d1[parameter == "I", dataset[which.max(mean)]] == "novitsky2014")
@@
     \Sexpr{d1[parameter == "I", round(max(mean))]}
-    (HET/Botswana).
-The estimates of $I$ were less robust to the HIV gene analyzed
-(\cref{fig:genes}) and the phylogenetic reconstruction method used
-(\cref{fig:methods}).}
+(HET/Botswana).}
 
@@
 significantly~\autocite{ypma2013relating} due to within-host evolution and the
-sampling process. \add{Progress has very recently been made on this front by
-\textcite{giardina2016inference}, who have concurrently developed an ABC-based
-method incorporating within-host evolution.} The ABC-SMC algorithm is
+sampling process~\autocite{giardina2016inference}. The ABC-SMC algorithm is
 computationally intensive, taking about a day when run on 20 cores in parallel
@@
 \clearpage
+\printbibliography
 \newpage
-\printbibliography
 
@@
     \centering
+    \includegraphics[width=\textwidth]{realdata-bootstrap}
+    \caption[
+        Posterior means and 50\%/95\% HPD intervals for parameters of the
+        BA network model, fitted to original and bootstrap replicate alignments
+        of two HIV datasets.
+    ]{
+        Posterior means (points), 50\% HPD intervals (notches), and 95\%
+        HPD intervals (lines) for parameters of the BA network model, fitted to
+        original and bootstrap replicate alignments of two HIV datasets.
+    }
+    \label{fig:boot}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
   \includegraphics{derivatives.pdf}
@@
   ]{
-    First and second \add{time} derivatives of epidemic growth curves at time
+    First and second time derivatives of epidemic growth curves at time
     of sampling for various values of $I$ and $I$. Networks were simulated

### 2016-07-14 doc/paper/paper.tex
Finished reviewer things finally ugh

@@
 % latex table generated in R 3.2.3 by xtable 1.8-2 package
-% Wed Jul  6 16:31:52 2016
+% Thu Jul 14 15:34:47 2016
 \begin{table}[ht]
@@
 basic settings and 124 with the higher settings. The results of the two
-simulations were very similar for all parameters.}
+simulations were similar, but surprisingly, the results with higher SMC
+settings were slightly worse (by
+    10\%,
+    8\%, and
+    11\%
+for $\alpha$, $I$, and $N$ respectively). However, the 50\% HPD interval for
+$I$ was closer to the true value of 2000 with the improved settings
+    (2338 - 
+     3423, vs.
+     2810 - 
+     3767 with basic settings).
+The estimate of $m$, 3 in both cases, was unaffected by the settings.}
 
@@
 (mixed/Spain) to
-    0.81
+    0.83
 (IDU/Estonia). When aggregated by risk group, the average estimates were
-    0.77
+    0.78
 for IDU,
@@
 $I$ were between
-    410
-(MSM/Taiwan) and
+    373
+(IDU/Estonia) and
     1391
@@
 (Spearman correlation, 
-    $p =0.8$),
+    $p =0.9$),
 indicating that the higher estimates were not simply due to increased sampling
@@
 (mean
-    6188,
+    6202,
 range
@@
 $\alpha$ between the IDU datasets and the others became more striking
-(\cref{fig:abchpdm2}). All three IDU datasets had estimated $\alpha$ values at
-or above
+(\cref{fig:abchpdm2}). Both IDU datasets had estimated $\alpha$ values at or
+above
     1.
@@
 narrower (average width
-    2192
+    2159
 for $m \geq 1$ and
@@
 the two choices of prior.}
+
+\add{For the two datasets we reanalyzed using
+RAxML~\autocite{stamatakis2014raxml} and LSD~\autocite{to2016fast}, $\alpha$
+was relatively robust to the choice of method (\cref{fig:methods}, posterior
+means 
+    0.48
+vs. 
+    0.48
+for mixed/Spain and 
+    1.02
+vs.
+    1.12
+for IDU/Estonia). However, the estimates of $I$ were about twice as high when
+RAxML was used instead of FastTree to reconstruct the trees 
+    (228 
+     vs. 437 for IDU/Estonia, 
+    816
+    vs. 1949 for mixed/Spain). 
+\Cref{fig:boot} shows estimates obtained for five bootstrap replicate
+alignments for each of these two datasets. For the mixed/Spain data, the
+estimated posterior mean [range of bootstrap posterior means] was
+    0.48
+    [0.54 -
+     0.67]
+for $\alpha$, 
+    816
+    [403 -
+     886] for $I$,
+    2.76
+    [2.61 -
+     3.24] for $m$, and
+    6639
+    [6652 -
+     7245] for $N$. 
+For the IDU/Estonia data, values were
+    1.02
+    [0.78 -
+     1.07]
+for $\alpha$, 
+    228
+    [313 -
+     741] for $I$,
+    3.11
+    [3.41 -
+     3.46] for $m$, and
+    6803
+    [5941 -
+    6913] for $N$.}
 
@@
 to 
-    0.81.
+    0.83.
 The sub-linearity is consistent with the results of
@@
 datasets ranged from
-    410
+    373
     (IDU/Estonia)
@@
     5432
-    (HET/Botswana).
-The estimates of $I$ were less robust to the HIV gene analyzed
-(\cref{fig:genes}) and the phylogenetic reconstruction method used
-(\cref{fig:methods}).}
+(HET/Botswana).}
 

### 2016-07-14 doc/defence/abc.tex
Presentation

\begin{frame}{Simulated trees can be used to fit network models}
  \begin{minipage}[p][0.8\textheight][t]{\textwidth}
    \only<2>{\includegraphics[width=0.9\textwidth, trim=0 3in 0 0, clip]{abc-idea.pdf}}
    \only<3>{\includegraphics[width=0.9\textwidth, trim=0 1.7in 0 0, clip]{abc-idea.pdf}}
    \only<4>{\includegraphics[width=0.9\textwidth, trim=0 0.8in 0 0, clip]{abc-idea.pdf}}
    \only<5->{\includegraphics[width=0.9\textwidth, trim=0 0in 0 0, clip]{abc-idea.pdf}}

    \vspace{0.5cm}
    \only<6>{\centerline{\Large approximate Bayesian computation}}
  \end{minipage}
\end{frame}

### 2016-07-14 doc/defence/conclusions.tex
Presentation

\begin{frame}
  \begin{itemize}
    \setlength{\itemsep}{12pt}
    \item \textit{Netabc} uses approximate Bayesian computation with sequential
      Monte Carlo, epidemic simulation, and the tree kernel to fit contact
      network models to phylogenetic data.
    \pause
    \item In simulated data, viral phylogenies were informative about the
      $\alpha$, $I$, and $N$ parameters of the Barab\'asi-Albert contact
      network model.
    \pause
    \item The networks underlying real-world HIV epidemics exhibit sub-linear
      preferential attachment, and PA values may be higher for IDU networks.
  \end{itemize}
\end{frame}

### 2016-07-14 doc/defence/posterior1.tex
Presentation

@@
 \begin{frame}{Marginal posterior distributions for $\alpha$ show least spread}
-  \begin{minipage}[p][\textheight][t]{\textwidth}
-    \begin{center}
-    preferential attachment power \hfill prevalence
-      \includegraphics[width=0.8\textwidth]{abc-posterior-example.pdf}
+  \vspace{-0.5cm}
+  \begin{tikzpicture}
+    \node at (0, 0) [anchor=center] {\includegraphics[trim=1cm 0 0 0, clip, width=0.8\textwidth]{{abc-posterior/1.0_1000_3_5000_0}.pdf}};
+    \node at (-4, 3) [anchor=center, text width=2cm, fill=white, inner sep=2pt] {preferential attachment};
+    \node at (4.5, 3) [anchor=center, text width=2cm, fill=white, inner sep=2pt] {number infected};
+    \node at (-4.2, -1) [anchor=center, text width=2cm, align=center, fill=white, inner ysep=2pt, inner xsep=0pt] {edges per vertex};
+    \node at (4.2, -0.7) [anchor=center, text width=1.5cm, align=center, fill=white, inner sep=2pt] {total nodes};
 
-      \vspace{-0.25cm}
-      edges per vertex \hfill total nodes
-    \end{center}
-  \end{minipage}
+    \only<1>{ \fill [white] (0, 0) rectangle (4.8, 3.8);}
+    \only<1-2>{ \fill [white] (0, 0) rectangle (-5, -3.8);}
+    \only<1-3>{ \fill [white] (0, 0) rectangle (4.8, -3.8);}
+    \only<4>{ }
+  \end{tikzpicture}
 \end{frame}

### 2016-07-14 doc/defence/presentation.tex
Presentation

@@
 \usetheme[everytitleformat=regular]{m}
-%\setbeamertemplate{navigation symbols}{}
+\setbeamertemplate{navigation symbols}{}
 
@@
 
-%\begin{frame}{Outline}
-%    \tableofcontents
-%\end{frame}
+\begin{frame}{Outline}
+    \tableofcontents
+\end{frame}
 
-%\section{Background and Motivation}
-%\input{contactnet}
-%\input{phylo}
-%\input{objects}
-%\input{transtree}
-%\input{virusphylo}
-%\input{inference}
-%%\input{netmodel}
-%\input{why}
+\section{Background and motivation}
+\input{contactnet}
+\input{phylo}
+\input{objects}
+\input{transtree}
+\input{virusphylo}
+\input{inference}
+%\input{netmodel}
+\input{why}
 
-%\section{The Barab\'asi-Albert (BA) Contact Network Model}
-%\input{ba}
-%\input{identifiable}
-%\input{kpca}
+\section{The Barab\'asi-Albert (BA) contact network model}
+\input{ba}
+\input{identifiable}
+\input{kpca}
 
-%\section{\textit{Netabc}: Phylogenetic Reconstruction of Contact Network Parameters with Approximate Bayesian Computation}
-%\input{smc}
-%\input{posterior1}
-%\input{posterior2}
-%\input{pointest}
-%\input{pointest2}
+\section{Our inference method: \itshape{netabc}}
+\input{abc}
+\input{smc}
+\input{posterior1}
+\input{posterior2}
+\input{pointest}
+\input{pointest2}
 
 \section{Application to real HIV epidemics}
-\input{data}
+\input{map}
+\input{realdata}
 
 \section{Conclusions}
+\input{conclusions}
 
+\input{acknowledgements}
+
+\input{supp}
 \end{document}
-
-\begin{frame}{HIV networks have sub-linear preferential attachment}
-    Also, higher PA for IDU networks.
-
-    \includegraphics[trim=0 3.5in 3.3in 0, clip, width=0.5\textwidth]{realdata-hpd-bc}
-    \includegraphics[trim=3.4in 1.5in 0 2.0in, clip, width=0.5\textwidth]{realdata-hpd-bc}
-\end{frame}
-
-\begin{frame}{Conclusions}
-  \begin{itemize}
-    \item We developed a phylodynamic method to fit contact network models to
-      phylogenetic data.
-      \pause
-    \item The preferential attachment power of the Barab\'asi-Albert network
-      model, which is challenging to estimate by traditional epidemiological
-      methods, can be estimated with ABC.
-      \pause
-    \item The networks underlying real epidemics are heterogeneous,
-      underscoring the importance of considering network structure in
-      phylodynamic analyses.
-  \end{itemize}
-\end{frame}
-
-\begin{frame}{Acknowledgements}
-  \begin{columns}
-    \begin{column}{0.6\textwidth}
-
-      \textbf{BC Centre for Excellence in HIV/AIDS}
-
-      Art Poon
-
-      Jeff Joy
-
-      Richard Liang
-
-      Thuy Nguyen
-
-      P. Richard Harrigan
-
-      \hfill\\
-      \textbf{University of British Columbia}
-
-      Sarah Otto
-
-      Alexandre Bouchard-C\^ot\'e
-
-      \vfill
-      \vspace{0.5cm}
-      $\qquad\qquad$\includegraphics[width=2cm]{logos/genomecanada}
-      $\qquad$
-      \includegraphics[width=2cm]{logos/genomequebec}
-    \end{column}
-    \begin{column}{0.4\textwidth}
-      \centering
-
-      \includegraphics[width=3cm]{logos/cfe}
-      \vspace{0.5cm}
-
-      \includegraphics[width=3cm]{logos/cihr}
-
-      \includegraphics[width=3cm]{logos/genomebc}
-
-      \includegraphics[width=3cm]{logos/bmgf}
-      \vspace{0.5cm}
-
-      \includegraphics[width=3cm]{logos/btp}
-    \end{column}
-  \end{columns}
-\end{frame}
-
-\end{document}

### 2016-07-14 doc/defence/supp.tex
Presentation

\begin{frame}{Marginalization improves estimates of $I$, $m$, and $N$}
  \centerline{\includegraphics[height=0.8\textheight]{abc-marginal}}
\end{frame}

\begin{frame}{Higher SMC settings do not improve parameter estimates}
  \centerline{\includegraphics[height=0.8\textheight]{better-smc}}
\end{frame}

\begin{frame}{$I$ and $N$ may be identifiable only up to epidemic phase}
  \centerline{\includegraphics[height=0.8\textheight]{derivatives}}
\end{frame}

\begin{frame}{Grid search estimates are accurate with wide intervals}
  \centerline{\includegraphics[height=0.8\textheight]{gridsearch-example}}
\end{frame}

\begin{frame}{Higher $\alpha$ causes unbalanced trees}
  \includegraphics[width=\textwidth]{kernel-alpha-tree}
\end{frame}

\begin{frame}{Higher $I$ causes longer terminal branch lengths}
  \includegraphics[width=\textwidth]{kernel-I-tree}
\end{frame}

\begin{frame}{Tree shape not visibly affected by $m$}
  \includegraphics[width=\textwidth]{kernel-m-tree}
\end{frame}

\begin{frame}{Higher $N$ causes longer internal branch lengths}
  \includegraphics[width=\textwidth]{kernel-N-tree}
\end{frame}

\begin{frame}{Tree kernel and other statistics for parameter classification}
  \includegraphics[width=\textwidth]{kernel-rsquared}
\end{frame}

\begin{frame}{Estimates of $\alpha$ are robust to HIV gene}
  \centerline{\includegraphics[height=0.8\textheight]{realdata-genes}}
\end{frame}

\begin{frame}{$\alpha$ estimates are robust to phylogeny reconstruction method}
  \centerline{\includegraphics[height=0.8\textheight]{realdata-methods}}
\end{frame}

\begin{frame}{Power law fit captures slope of BA degree distributions}
  \centerline{\includegraphics[height=0.8\textheight]{powerlaw}}
\end{frame}

\begin{frame}{SMC was validated on two mixture distributions}
  \begin{columns}
    \column{0.5\textwidth}
    \includegraphics[width=\textwidth]{smc-test}
    \column{0.5\textwidth}
    \includegraphics[width=\textwidth]{smc-test-bimodal}
  \end{columns}
\end{frame}

### 2016-07-25 doc/defence/transtree2.tex
Final changes to defence presentation

--- 2016-06-29 doc/defence/transtree.tex
@@
-\begin{frame}{Contact networks shape transmission trees}
+\begin{frame}{Transmission trees record the progress of an epidemic}
   \vspace{-1cm}
@@
       \def\pht{1cm}
-      \node (a) {\includegraphics[height=\pht]{stock/person}};
-      \node at (a.south) [label] {\Large a};
-      \node [below left=0.75 of a] (b) {\includegraphics[height=\pht]{stock/person}};
-      \node at (b.south) [label] {\Large b};
-      \node [below=0.75 of b] (e) {\includegraphics[height=\pht]{stock/person}};
-      \node at (e.south) [label] {\Large e};
-      \node [below right=0.75 of a] (c) {\includegraphics[height=\pht]{stock/person}};
-      \node at (c.south) [label] {\Large c};
-      \node [below=0.75 of c] (d) {\includegraphics[height=\pht]{stock/person}};
-      \node at (d.south) [label] {\Large d};
-
-      \fill [shade, opacity=0.2] (a.east) -- (c.north) -- (c.north west) -- (a.south east) -- cycle;
-      \fill [shade, opacity=0.2] (a.west) -- (b.north) -- (b.north east) -- (a.south west) -- cycle;
-      \fill [shade, opacity=0.2] (a.south) -- (a.south west) -- (e.north) -- (e.north east) -- cycle;
-      \fill [shade, opacity=0.2] (b.south west) -- (b.south) -- (e.north) -- (e.north west) -- cycle;
-      \fill [shade, opacity=0.2] ($(c.south) + (-8pt, 0)$) -- ++(16pt, 0) -- ($(d.north) + (8pt, 0)$) -- ++(-16pt, 0) -- cycle;
-      \fill [shade, opacity=0.2] (b.east) -- (b.south east) -- (d.west) -- (d.north west) -- cycle;
-      \uncover<2->{
-      \draw [red] ($(a.west) + (0, -12pt)$) -- ($(b.north) + (8pt, 0)$);
-      }
-      \uncover<4->{
-      \draw [blue] ($(a.east) + (0, -12pt)$) -- ($(c.north) + (-8pt, 0)$);
-      }
-      \uncover<6->{
-      \draw [green!80!black] (c) -- (d);
-      }
-
-      \node (abcd) [above right=0 and 5cm of a] {\includegraphics[height=\pht]{stock/person}};
+      \node (abcd) {\includegraphics[height=\pht]{stock/person}};
       \node at (abcd.south) [label] {\Large a};
 
-      \uncover<2>{
-      \node (b) at (abcd.east) [anchor=west] {\includegraphics[height=\pht]{stock/person}};
-      \node at (b.south) [label] {\Large b};
-      \draw [red] (abcd.south) to [bend right=75] (b.south);
-      }
-
-      \uncover<3->{
       \node (b) [below right=1 and 0.75 of abcd] {\includegraphics[height=\pht]{stock/person}};
@@
       \draw [-, dashed] (abcd) -- (acd);
-      }
 
-      \uncover<4>{
-      \node (cd) at (acd.east) [anchor=west] {\includegraphics[height=\pht]{stock/person}};
-      \node at (cd.south) [label] {\Large c};
-      \draw [blue] (acd.south) to [bend right=75] (cd.south);
-      }
-
-      \uncover<5->{
       \node (cd) [below right=0.5 and 0.5 of acd] {\includegraphics[height=\pht]{stock/person}};
@@
       \draw [-, dashed] (acd) -- (a);
-      }
 
-      \uncover<6>{
-      \node (d) at (cd.east) [anchor=west] {\includegraphics[height=\pht]{stock/person}};
-      \node at (d.south) [label] {\Large d};
-      \draw [green] (cd.south) to [bend right=75] (d.south);
-      }
-
-      \uncover<7->{
       \node (c) [below left=0.25 and 0.25 of cd] {\includegraphics[height=\pht]{stock/person}};
@@
       \draw [green!80!black] (cd) -- (d);
-      }
 
@@
 
+      \uncover<2->{
+      \node (t) [above left=0 and 2 of acd, inner xsep=4pt] {transmission};
+      \draw [thin] (t.east) -- (acd);
+      }
+
+      \uncover<3->{
+      \node (t) [below left=0 and 1 of a, inner xsep=4pt] {sampling/removal};
+      \draw [thin, shorten >=4pt] (t.east) -- (a);
+      }
+
     \end{tikzpicture}
-
-  $\qquad$contact network \hfill transmission tree$\qquad$
   \end{center}

### 2016-07-27 doc/thesis/acknowledgements.tex
Defence version of thesis with changes

@@
-First and foremost I am deeply indebted to my supervisor, Dr. Art Poon, 
-for mentoring me through the majority of my academic career.
+First and foremost I am deeply indebted to my supervisor, Dr. Art Poon, for
+mentoring me through the majority of my academic career, and somehow always
+managing to make time for me. My time in his lab has been challenging and
+rewarding, and I feel well prepared for opportunities to come. I also thank my
+labmates, especially Jeff, Richard, and Thuy. I'll miss our Thursday lunches.
 
-I also thank my labmates, especially Jeff, Richard, and Thuy. I'll miss our
-Thursday lunches.
-
-I am grateful for the mentorship of my committee members Dr. Alexandre
+I am grateful for the guidance of my committee members Dr. Alexandre
 Bouchard-C\^ot\'e, Dr. Richard Harrigan, and Dr. Sarah Otto. Thanks also to my
@@
 I owe thanks to my fellow graduate students of the bioinformatics training
-program, especially to Ogan.
+program, especially Ogan, who helped keep me on an even keel when times got
+tough.
 

### 2016-07-27 doc/thesis/appendix.Rnw
Defence version of thesis with changes

@@
 
-\begin{figure}[ht]
-    \centering
-    \includegraphics{abc-point-estimate-m3.pdf}
+\begin{figure}
+    \includegraphics[width=\textwidth]{abc-boxplot-mN}
     \caption[
-        Posterior mean point estimates for \gls{BA} model parameters obtained
-        by running \software{netabc} on simulated data, for simulations with
-        \gls{m} = 3. 
+        Posterior mean point estimates for BA model parameters $m$ and $N$
+        obtained by running \textit{netabc} on simulated data, stratified by
+        true parameter values.
     ]{
-        Posterior mean point estimates for \gls{BA} model parameters obtained
-        by running \software{netabc} on simulated data, for simulations with
-        \gls{m} = 3. Dashed lines indicate true values. (A) Estimates of
-        \gls{alpha} and \gls{I} which were varied in these simulations against
-        known values. (B) Estimates of \gls{m} and \gls{N} which were held
-        fixed in these simulations at the values \gls{m} = 3 and \gls{N} =
-        5000.
+        Posterior mean point estimates for BA model parameters $m$ and $N$
+        obtained by running \textit{netabc} on simulated data, stratified by
+        true parameter values. First row of plots contains true versus
+        estimated values of $m$; second row contains true versus estimated
+        values of $N$. Columns are stratified by $\alpha$, $I$, and $m$
+        respectively. Dashed lines indicate true values. 
     }
-    \label{fig:abcptm3}
+    \label{fig:abcpt2}
 \end{figure}
 
-\begin{figure}[ht]
-    \centering
-    \includegraphics{abc-point-estimate-m4.pdf}
-    \caption[
-        Posterior mean point estimates for \gls{BA} model parameters obtained
-        by running \software{netabc} on simulated data, for simulations with
-        \gls{m} = 3.
-    ]{
-        Posterior mean point estimates for \gls{BA} model parameters obtained
-        by running \software{netabc} on simulated data, for simulations with
-        \gls{m} = 4. Dashed lines indicate true values. (A) Estimates of
-        \gls{alpha} and \gls{I} which were varied in these simulations against
-        known values. (B) Estimates of \gls{m} and \gls{N} which were held
-        fixed in these simulations at the values \gls{m} = 4 and \gls{N} =
-        5000.
-    }        
-    \label{fig:abcptm4}
-\end{figure}
+%\begin{figure}[ht]
+%    \centering
+%    \includegraphics{abc-point-estimate-m3.pdf}
+%    \caption[
+%        Posterior mean point estimates for \gls{BA} model parameters obtained
+%        by running \software{netabc} on simulated data, for simulations with
+%        \gls{m} = 3. 
+%    ]{
+%        Posterior mean point estimates for \gls{BA} model parameters obtained
+%        by running \software{netabc} on simulated data, for simulations with
+%        \gls{m} = 3. Dashed lines indicate true values. (A) Estimates of
+%        \gls{alpha} and \gls{I} which were varied in these simulations against
+%        known values. (B) Estimates of \gls{m} and \gls{N} which were held
+%        fixed in these simulations at the values \gls{m} = 3 and \gls{N} =
+%        5000.
+%    }        
+%    \label{fig:abcptm3}
+%\end{figure}
+%
+%\begin{figure}[ht]
+%    \centering
+%    \includegraphics{abc-point-estimate-m4.pdf}
+%    \caption[
+%        Posterior mean point estimates for \gls{BA} model parameters obtained
+%        by running \software{netabc} on simulated data, for simulations with
+%        \gls{m} = 3.
+%    ]{
+%        Posterior mean point estimates for \gls{BA} model parameters obtained
+%        by running \software{netabc} on simulated data, for simulations with
+%        \gls{m} = 4. Dashed lines indicate true values. (A) Estimates of
+%        \gls{alpha} and \gls{I} which were varied in these simulations against
+%        known values. (B) Estimates of \gls{m} and \gls{N} which were held
+%        fixed in these simulations at the values \gls{m} = 4 and \gls{N} =
+%        5000.
+%    }        
+%    \label{fig:abcptm4}
+%\end{figure}
 
@@
 
-\begin{figure}[ht]
-    \centering
-    \includegraphics{mixed-posterior}
-    \caption[
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters obtained with \software{netabc} for a network with
-        heterogeneous node behaviour.
-    ]{
-        Approximate marginal posterior distributions of \gls{BA} model
-        parameters obtained with \software{netabc} for a network with
-        heterogeneous node behaviour. Half of the nodes were attached with
-        $\gls{alpha} = 0.5$, and the other half with $\gls{alpha} = 1.5$
-        (vertical dashed lines, top left). Other parameter values were $m = 2$,
-        $\gls{I} = 1000$, and $\gls{N} = 5000$ (vertical dashed lines, other
-        than top left). Shaded areas indicate 95\% highest posterior density
-        intervals.
-    }
-    \label{fig:mixed}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \includegraphics{peerdriven-posterior}
-    \caption[
-        Approximate marginal posterior distributions of \gls{BA}
-        model parameters obtained with \software{netabc} for a network with
-        peer-driven sampling.
-    ]{
-        Approximate marginal posterior distributions of \gls{BA}
-        model parameters obtained with \software{netabc} for a network with
-        peer-driven sampling. An epidemic was simulated in the usual fashion,
-        but rather than being sampled at random, infected nodes were sampled
-        with a probability two times higher if they had any sampled neighbours
-        in the contact network. Vertical dashed lines indicate true parameter
-        values, and shaded areas indicate 95\% highest posterior density
-        intervals.
-    }
-    \label{fig:peerdriven}
-\end{figure}
-
+%\begin{figure}[ht]
+%    \centering
+%    \includegraphics{mixed-posterior}
+%    \caption[
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters obtained with \software{netabc} for a network with
+%        heterogeneous node behaviour.
+%    ]{
+%        Approximate marginal posterior distributions of \gls{BA} model
+%        parameters obtained with \software{netabc} for a network with
+%        heterogeneous node behaviour. Half of the nodes were attached with
+%        $\gls{alpha} = 0.5$, and the other half with $\gls{alpha} = 1.5$
+%        (vertical dashed lines, top left). Other parameter values were $m = 2$,
+%        $\gls{I} = 1000$, and $\gls{N} = 5000$ (vertical dashed lines, other
+%        than top left). Shaded areas indicate 95\% highest posterior density
+%        intervals.
+%    }
+%    \label{fig:mixed}
+%\end{figure}
+%
+%%\begin{figure}[ht]
+%%    \centering
+%%    \includegraphics{peerdriven-posterior}
+%%    \caption[
+%%        Approximate marginal posterior distributions of \gls{BA}
+%%        model parameters obtained with \software{netabc} for a network with
+%%        peer-driven sampling.
+%%    ]{
+%%        Approximate marginal posterior distributions of \gls{BA}
+%%        model parameters obtained with \software{netabc} for a network with
+%%        peer-driven sampling. An epidemic was simulated in the usual fashion,
+%%        but rather than being sampled at random, infected nodes were sampled
+%%        with a probability two times higher if they had any sampled neighbours
+%%        in the contact network. Vertical dashed lines indicate true parameter
+%%        values, and shaded areas indicate 95\% highest posterior density
+%%        intervals.
+%%    }
+%%    \label{fig:peerdriven}
+%%\end{figure}
+%
 \begin{figure}[ht]

### 2016-07-27 doc/thesis/changes.tex
Defence version of thesis with changes

\documentclass[12pt]{article}
\usepackage[T1]{fontenc}
\usepackage{mathptmx}
\usepackage{fullpage}
\usepackage{amsmath,amssymb}
\usepackage[garamondx,cmbraces]{newtxmath}
\usepackage{setspace}
\usepackage{parskip}
\begin{document}

\onehalfspacing

In the time since circulating the draft thesis, and submitting the paper, I
have undertaken some additional work and restructuring. \\

\begin{itemize}
  \item I have done some more literature searching and found five additional
    published datasets to which my method could be applied. Out of these
    emerged an interesting trend: IDU networks seem to have higher $\alpha$
    than others. This is in line with the findings of a paper of Drombowski et
    al. 2013, who found that IDU networks are much to typical BA networks than
    sexual networks are.
  \item Two of these new datasets had sequenced two HIV genes, which we
    analysed separately. I found that estimates of $\alpha$ were very robust
    to the choice of gene. $I$ was less robust, with \textit{env} genes tending
    leading to higher estimated values.
  \item There is another study (de Blasio et al. 2007) which estimates $\alpha$
    from partner count data and I compare my estimates to theirs. They are
    similar.
  \item Prompted by one of Alex's comments, I found that the posterior mean is
    a more conventional estimator than the maximum \textit{a posteriori} (MAP)
    value, and have replaced MAP estimates with posterior means throughout the
    thesis.
  \item I searched the literature for estimated power law exponent ($\gamma$)
    values from real datasets (Table 2.9). Unfortunately, no pattern could be
    observed in the published estimates with respect to risk groups, genders,
    or time durations. 
  \item There are only two example posterior distributions now, instead of
    many in the text.
  \item The ``basic'' Bayesian statistics material (ie. what a posterior
    distribution is, what a parametric model is) have been moved into Appendix
    A instead of the introduction section on ABC.
  \item I found an analysis of the Barab\'asi-Albert model in the physics
    literature (Krapivsky et al. 2000) which demonstrates that the degree
    distribution is only a power law when $\alpha = 1$. I acknowledge this but
    do not believe it detracts from the validity of teh results, especially in
    light of the above mentioned de Blasio et al. paper. I added to the
    supplemental materials a series of power law fits to the degree
    distributions of BA networks with $\alpha \neq 1$ (Figure A.25), which
    shows that, although the power law is cleary an imperfect fit, the slope
    is still reasonably representative of the true distribution.
\end{itemize}

The following changes have been made at the request of the anonymous reviewers 
of the paper based on this work. \\

\begin{itemize}
  \item I repeated two of the real data analyses with trees created with RAxML
    and LSD rather than FastTree and root-to-tip regression. The results for
    $\alpha$ are very similar, but the estimates of $I$ are not very robust to
    the change of methods. This indicates weaker identifiability of $I$ than of
    $\alpha$.
  \item Figure 2.9 (formerly 2.8), which previously showed each simulation as
    an individual point, has been replaced by a set of stratified box plots
    showing the point estimates against true values.
  \item I had evaluated the ``classifiers'' for different parameter values
    using $R^2$, which doisn't make sense for classifiers. In fact, an SVR is a
    regressor, not a classifier, so the $R^2$ were for predicted against actual
    values. In any event, I have redone the analyses using SVMs instead of SVRs
    and regressions. The results are qualitatively very similar.
  \item I show two-dimensinal marginal posterior distributions in addition to
    the 1-dimensional ones, which provide evidence of parameter correlations
    (specifically, between $I$ and $N$).
  \item The reviewers pointed out that there are two sources of error: bad ABC
    approximation, and a noisy posterior. To disambiguate these, I performed
    marginal ABC estimation by fixing some parameters to known values. The
    estimates of all but $m$ improved by about 50\%.
  \item I also tried increasing the SMC settings to get better posterior
    resolution. It didn't make much difference (Figure 2.15). In fact it made
    the estimates slightly worse.
\end{itemize}

Changes that I have made in response to your specific comments are outlined
below. Minor grammatical and phrasing changes have been omitted. \\

\subsubsection*{Sally}

\begin{quote}
  \itshape
  In general (and partly suggested by Alex), I would highly recommend that you
  make minor revisions throughout to sign post to the reader why you are about
  to launch into, as you introduce a new topic.  You are strong on the
  technical details, but I often found myself wondering what the broader
  context was.  I think maybe you credit your reader for being super smart and
  not wanting to spell out the motivation for each section, but we aren’t that
  smart!
\end{quote}

Significantly more exposition has been added throughout the thesis. In
particular, each section of Chapter 2 now has some introductory material.
Sections 2.2.1 and 2.2.2 serve as motivation and a road map, respectively, for
the synthetic data experiments.

\begin{quote}
  \itshape
  For example:

  * How does sequential Monte Carlo (section 1.4) an ABC (section 1.5) fit into
  your overall thesis goals?  How do the ``particles'' introduced on page 13
  relate to the network structure and transmission trees just discussed?
  Basically, link the previous sections to sections 1.4 and 1.5.
\end{quote}

Several additional explanatory paragraphs have been prepended to both of these
sections. In particular, for SMC, ``In this work, the distribution of interest
is the posterior distribution 1.3. The particles are particular values of the
parameters $\theta$ of the contact network model being studied.'' For ABC,
``the observed data y is an estimated transmission tree for a viral epidemic
under investigation. The model in question is a contact network model with
parameters $\theta$ . The simulated dataset $z$ is a transmission tree,
obtained by first generating a contact network under the model, and then
simulating the spread of an epidemic over that network.''

\begin{quote}
  \itshape

  * Chapter 2 ends abruptly and needs a connecting paragraph.  How does this overview prepare you (and the reader) to develop the methods in Chapter 3?
\end{quote}

I think you are referring to Chapter 1 (Introduction) not Chapter 2. In any
case, I have added both an introductory paragraph (the last in the Objective
section) and a concluding paragraph to this chapter.

\begin{quote}
  \itshape

  * Similarly, Chapter 3 needs an introductory paragraph or two.  Late in your
  thesis, you note that netabc is the first attempt to use phylogenetic data
  (-> transmission trees) to infer network properties.  Remind the reader of
  this here and give them an overview of how you will proceed with the
  inference -- at a very broad-scale level.
\end{quote}

An introductory paragraph has been added which broadly describes the contents
of the chapter.

\begin{quote}
  \itshape

  * I really needed a road map for section 2.1.2.  Tell the reader why you are
  using different approaches for inference and what you expect their strengths
  and weaknesses to be.  I had to dig to tell what the differences were between
  the Classifiers subsection and the Grid search subsection.  Did the first
  co-estimate all parameters while the second focused on one at a time?  Were
  different methods use (yes - several versus only kernel)?  [Say so]
\end{quote}

Section 2.2.2 now motivates and describes, at a high level, each of the
synthetic data experiments and why they were done. In addition, the sections
describing each of the experiments (headings under 2.2.3) now contain one or
two introductory paragraphs each.

\begin{quote}
  \itshape

  * I found the discussion of the HIV datasets could also have used some
  greater interpretation.  For these datasets, what has been estimated as the
  likely (or at least known) number of infections?  How does this relate to the
  inferred I values?
\end{quote}

I have carefully read each of the studies for any available information about
the HIV prevalence in the study populations. In the case of the Chinese
datasets, I also consulted an article which reported the results of a
nationwide prevalence survey by city. Unfortunately, no general conclusions
could be drawn about how our estimates compare to values estimated with other
means. For some datasets (eg. IDU/Canada), $I$ was apparently overestimated,
while for some (MSM/USA) it was most likely underestimated.

\begin{quote}
  \itshape

  In addition, other more major comments are:

  * In the grid search, you held all of the other unknown parameters to their
  true values (if I understand correctly from Table 2.2), right?  You really
  have to make this explicit in the methods and discuss both there and in the
  results that this will give an overly confident view of the parameter values,
  compared with performing a grid search but allowing the other parameters to
  be inferred.  [Unless the latter is what you actually did.]
\end{quote}

This has been made much clearer both in section 2.2.2, and in the subsections
of both the methods and results where the grid search estimates are reported.
Figure captions emphasize this point as well.

\begin{quote}
  \itshape

  * I would consider adding box plots with interquartile ranges in places like
  Figure 2.7.  The 95\% CI are simply too broad to be able to compare the
  accuracy of the inferences.
\end{quote}

I felt that box plots would be confusing, since we do not report the median and
quantiles, but rather the posterior mean and highest density intervals.
However, I have added notches each plot of this type to indicate the 50\%
highest density intervals, which are similar to the interquartile range.

\begin{quote}
  \itshape

  * When you say that alpha is the most accurately estimated parameter (p. 37
  and elsewhere), I think that this is unfair because it is based on absolute
  differences and alpha, by definition, is a smaller number than I and N, say.
  Visually inspecting Figure 2.7, ``I'' is much better estimated (hugging the
  line more closely).  I think it would be better to discuss the proportional
  deviation (or $R^2$)  rather than absolute errors in this section.  Similarly,
  you compared some data points as being easier to estimate than others (e.g.,
  alpha = 1 and 1.5) but I do not see a trend that justifies picking out the
  best fitting two points and singling them out as ``special''.  In other words,
  your p value needs to be corrected for the fact that you might take any
  combination of the alpha values that seemed, by eye, to hug the regression
  line most closely.  Similarly, on p. 39 you note that the errors for I were
  higher when alpha was at least 1, but this is not at all visible and I doubt
  it is true (based on the bottom left panel in Figure 2.8), even before
  correcting for the fact that you’ve arbitrarily chosen one of several
  possible comparisons to make.
\end{quote}

Figure 2.7 shows only one out of ten replicate simulations for each parameter
value, so it may be somewhat visually misleading with respect to which
parameter is estimated best overall. I elected to represent the data this way,
instead of using box plots with all the simulations, in order to show the 95\%
and 50\% highest density intervals. I have noted this in the text.

In any event, $I$ and $\alpha$ were in fact estimated with comparable accuracy. 
I now simply report the accuracy rates for each parameter and make some
qualitative observations about the distributions without claiming any
statistical significance.

With respect to $p$-value correction, I have also removed the ad-hoc
statistical tests in the analysis of the ABC experiments and replaced them with
two $p$-value corrected GLMs (Tables 2.5 and 2.6). \\

\subsubsection*{Alex}

\begin{quote}
  \itshape

  High-level comments:

  - Terminology: 'kernel ABC' in the way you use it ('ABC with epsilon
  threshold where the distance happens to be computed using the kernel trick')
  might be confusing to the ABC folks, since there is already some work on
  'kernel ABC' used for a different meaning ('ABC where there is not a hard
  threshold on the distance, instead, a kernel is used to weigh the generated
  hypotheses', see for example http://arxiv.org/abs/1009.5736). 
\end{quote}

I have replaced ``kernel-ABC'' with either ``ABC'' or \textit{netabc}
throughout the thesis.

\begin{quote}
  \itshape

  - The material in section 2 is all there and good, but I feel it could be
  reorganized a little bit to make it easier to follow. I would suggest making
  one section with only the description of the method. Then another section for
  the experiments on synthetic data. And then one more for real data analysis.
  The section on synthetic data would also benefit from a short introductory
  paragraph explaining that you first present exploratory data analysis
  (without SMC-ABC) to assess the strength of the signal, and then do the
  SMC-ABC analysis (and subdivide the section into corresponding subsection). 
\end{quote}

The chapter is now divided into four sections: description of the method,
sythetic data analyses, real data analyses, and discussion.

\begin{quote}
  \itshape
  Detailed comments:

  ``were not estimable with kernel-ABC'' -> ''appeared to be weakly or
  non-identifiable with kernel-ABC''. Estimable is typically reserved for the
  more specialized cases of linear models.
\end{quote}

I have replaced all instances of ``estimable'' with either ``identifiable'' or
another term appropriate to the context (ie. parameters were estimable $\to$
classifiers were accurate, etc.).

\begin{quote}
  \itshape

  Argument at bottom of page 14/132 and top of page 15/132: I didn't fully
  understand why you needed labelling---I suspect it will become clearer later
  in the manuscript but in terms of exposition it is a bit hard to follow.
  Perhaps move this section to later, i.e. after having introduced more
  notation? Also, note that showing that ML/Bayes requires a search over an
  exponential space does not necessarily rules out ML/Bayes. For example
  searching over tree is NP hard yet ML/Bayes are still good options in
  practice. A more convincing argument would be that the natural formulation of
  the problem falls in what is called a 'doubly-intractable problem', i.e. a
  search problem over an intractable space where evaluation of the likelihood
  of individual candidates is also computationally hard.
\end{quote}

I have moved that argument down into Chapter 2 (section 2.1.4), and instead
focus on a higher level discussion for the objective. Section 2.1.4 also
contains some explanation of the idea of doubly-intractable problems, and a new
figure showing that a transmission tree topology has several possible internal
node labellings.

\begin{quote}
  \itshape

  ``The tips of a phylogeny, that is, the nodes without any descendants,
  correspond to extant, or observed, taxa.'' -> in a phylodynamics context,
  this is not so accurate of a characterization of the tips.
\end{quote}

I have changed the sentance to read as follows: ``The \emph{internal nodes}            
correspond to their common ancestors, usually extinct (although occasionally
the internal nodes may be observed as well, eg. Jombart et al. 2011).''

\begin{quote}
  \itshape
  Same page: "large matches " -> what does that mean?

  A question, perhaps stupid, on the tree kernel, more specifically on the
  parts of the definition comparing branch lengths: won't that be sensitive to
  the order in which the branch lengths are enumerated? Since the children in a
  tree are unordered, do you need to sort them or something like that?
\end{quote}

Two paragraphs have been added to section 1.2.4 describing each of the terms in
more detail. The paragraphs also clarify that we usually ladderize the tree
before applying the tree kernel.

\begin{quote}
  \itshape

  ``Sequential importance sampling (SIS) [113] is one type of SMC method,'' ->
  it is rather a precursor of, technically.
\end{quote}

This is mentioned both in the section describing SIS and in the introductory
paragraphs at the beginning of the section.

\begin{quote}
  \itshape

  Equation on top of page 27/132, although technically correct, is not quite
  what is the basis for IS; note that the form used would have weights at the
  top and bottom (top for weighting, bottom for self-normalizing).
\end{quote}

When trying to understand IS, I found the formulation I had previously written
to be more intuitive. However, I have replaced it with the expression
\[
  \int f(x) \pi(x) dx = \frac{1}{Z} \int w(x) \eta(x) f(x) dx,
\]
which I think is more conventional. Here, $\pi$ is the distribution of interest,
$Z$ is its (unknown) normalizing constant, $f$ is an integrable function whose
expectation we want to calculate under $\pi$, $w$ is the importance weight, and
$\eta$ is the importance distribution.

\begin{quote}
  \itshape

  ``then $\eta(x)$ can be approximated by a Monte Carlo estimate.'' : in IS we
  do not approximate the importance distribution, rather we pick a simple one
  and sample from it exactly.
\end{quote}

This was a poor choice of words on my part. I have edited the statement as
follows: ``Since $\eta$ can be sampled from exactly, and $\gamma$ and $f$ can
both be evaluated pointwise, the integral $\int w(x) \eta(x) f(x) \d x$ can be
approximated by a Monte Carlo estimate.''

\begin{quote}
  \itshape

  ``The recursive definition eq. (1.3) suggests an algorithm for obtaining a
  sample from $\pi$ (algorithm 1).'' -> note that this algorithm does not
  create an exact sample from $\pi$, but rater an approximate sample, an
  important distinction to make.
\end{quote}

The sentence now reads as follows: ``The recursive definition 1.3 suggests an
algorithm for obtaining a sample from $\eta$ and using it to obtain an
approximate sample from $\pi$ by IS.''

\begin{quote}
  \itshape

  In SIS, you only need to normalize the weights in the final step, and ideally
  you shouldn't do resampling in the last iteration. (similarly, in SMC, you
  should not do resampling in the final iteration. Technically, you still get a
  consistent algorithm if you do resampling in the last step, but for
  efficiency reasons it should be avoided if possible, this is motivated by a
  theoretical analysis of the variance of the weights, described for example in
  the book by Del Moral, 'Feynman-Kac Formulae, Genealogical and Interacting
  Particle Systems with Applications')
\end{quote}

I'm glad you brought this to my attention, as I was making this mistake in all
the experiments. Fortunately it only affected post-processing, so I didn't have
to rerun everything. I have removed the final resampling step from all the
algorithms in this section, and from my analyses as well. When discussing this
point, you mentioned that the posterior expectation is a commonly used
estimator, which prompted me to replace my MAP estimates with posterior means
throughout.

\begin{quote}
  \itshape

  Your background on SMC is currently organized as follows:
  - IS
  - SIS
  - SMC samplers
  Note that a perhaps more natural organization of these ideas (in the sense
  that each in this progression introduces one fundamental new thing at the
  time) would be:
  - IS
  - SIS (which adds online capability)
  - SMC (which introduces the resampling operator to fight degeneracy)
  - 'SMC samplers' (which introduces the capacity to approach spaces that
  cannot be expressed as product spaces)
  Note that confusingly, 'SMC' and 'SMC samplers' are used for different ideas
  (the latter builds on the former). 
\end{quote}

I did not initially realize that the resampling was introduced as part of SMC
and is not part of SIS. I have moved resampling into a separate subsection
about only SMC, so the section now has the suggested structure.

\begin{quote}
  \itshape

  Related comment: one important idea missing in the background is the notion
  of particle degeneracy, which is the fundamental reason why resampling is
  theoretically justified.
\end{quote}

This point has been added to the section describing SMC: ``\ldots the
cumulative errors at each sequential step tend to push many of the weights to
very low values. This results in a poor approximation to $\pi$, since only a
few particles retain high importance weights after all $d$ sequential steps, a
problem known as \emph{particle degeneracy}.''

\begin{quote}
  \itshape

  ``data, x and outcomes y'' -> `outcome' seems a weird terminology name here.
  Perhaps covariate would be better.
\end{quote}

I had actually intended $y$ to be the data and $x$ to be the covariates: I was
thinking of covariates as ``inputs'' and data as ``outputs''. This text has
been moved into Appendix A and changed to read ``observed data $y$ and
covariates $x$.''

\begin{quote}
  \itshape

  ``suitably chosen distance $\rho$'' -> It might be good to include a short
  paragraph expending on this, since these conditions are not only for
  theoretical analysis but also have a big impact in practice. 
\end{quote}

There is now a subheading ``Distance functions and summary statistics in ABC''
which describes sufficient statistics and references some methods to choose
optimal summary statistics.

\begin{quote}
  \itshape

  ``We use the variable x'' -> point out that this definition is unrelated to
  the definition of x used earlier in this section
\end{quote}

Rather than change the wording, the previous usage has been moved into Appendix
A.

\begin{quote}
  \itshape

  ``Classifiers for BA model parameters based on tree shape'' -> specify
  explicitly here that you are doing synthetic data experiments. 
\end{quote}

This is now emphasized in both the new section 2.2.2 and the introductory
paragraphs below the relevant subheading in section 2.2.3.

\end{document}

### 2016-07-27 doc/thesis/discussion.Rnw
Defence version of thesis with changes

@@
 weaker identifiability for \gls{alpha} $<$ 1 than for \gls{alpha} $\geq$ 1}
-(\cref{fig:abcptm2,tab:abchpd}), which may be partially explained by the
+(\cref{fig:abcpt,tab:abchpd}), which may be partially explained by the
 relationship between \gls{alpha} and the power law exponent \gls{gamma}
@@
 always over-estimated by approximately a factor of two
-(\cref{fig:abcptm2,tab:abchpd}). One factor which may have contributed to this
+(\cref{fig:abcpt2,tab:abchpd}). One factor which may have contributed to this
 bias was our choice of prior distribution. Since the prior for \gls{I} and
@@
 confidence intervals were wide for $\gls{alpha} < 1$
-(\cref{tab:abchpd,fig:abcptm2,fig:abcptm3,fig:abcptm4}). The estimates of
-\gls{N} and \gls{I} were both upwardly biased, and the estimates of \gls{m}
-were largely uninformative. Despite these limitations, a major result of our
-investigation is that some contact network parameters have a measurable impact
-on tree shape which can be used to perform statistical inference. Further
-refinements to \software{netabc}, as well as the use of more sophisticated
-network models, may improve the accuracy and precision of these estimates.
+(\cref{fig:abcpt,tab:abchpd}). The estimates of \gls{N} and \gls{I} were both
+upwardly biased, and the estimates of \gls{m} were largely uninformative.
+Despite these limitations, a major result of our investigation is that some
+contact network parameters have a measurable impact on tree shape which can be
+used to perform statistical inference. Further refinements to
+\software{netabc}, as well as the use of more sophisticated network models, may
+improve the accuracy and precision of these estimates.
 
@@
 \textcite{de2007preferential} and \software{netabc} is also the region of
-poorest identifiability (\cref{fig:gridest,fig:abcptm2}). This was reflected in
+poorest identifiability (\cref{fig:gridest,fig:abcpt}). This was reflected in
 the high level of uncertainty in the estimates, with most 95\% \gls{HPD}
@@
 screening, it is most likely that the posterior mean represents an
-overestimate, especially considering the observed upward bias of \gls{ABC}
-estimates of \gls{I} observed on simulated data (\cref{fig:abcptm2}).
-It is also possible that the true sexual network extended outside the study's
-geographic area to include a larger number of infected individuals. The authors
-of the HET/Botswana data~\autocite{novitsky2013phylogenetic,
-novitsky2014impact} estimated that there were 1731 \gls{HIV}-positive
-individuals in the study area. \gls{HIV} sequences were obtained from
-approximately 70\% of these individuals. The estimated prevalence we obtained
-was much higher 
+overestimate, especially considering the upward bias of \gls{ABC} estimates of
+\gls{I} observed on simulated data (\cref{fig:abcpt2}). It is also possible
+that the true sexual network extended outside the study's geographic area to
+include a larger number of infected individuals. The authors of the
+HET/Botswana data~\autocite{novitsky2013phylogenetic, novitsky2014impact}
+estimated that there were 1731 \gls{HIV}-positive individuals in the study
+area. \gls{HIV} sequences were obtained from approximately 70\% of these
+individuals. The estimated prevalence we obtained was much higher 
     (\Sexpr{d1[dataset == "novitsky2014" & parameter == "I", round(mean)]}),

### 2016-07-27 doc/thesis/discussion.tex
Defence version of thesis with changes

@@
 to 
-    0.9.
+    0.81.
 The sub-linearity is consistent with the results of
@@
     747 and 
-    372.
+    465.
 The authors of the IDU/Romania study~\autocite{niculescu2015recent} reported
@@
 value outside our 95\% \gls{HPD} interval
-    (188 -
-    834).
+    (171 -
+    1343).
 The authors indicated that their sample was ``convenience based''; it is
@@
 interval
-    (399 -
-    2862),
+    (384 -
+    3484),
 but it seems likely that the posterior mean of
-    1481
+    1356
 is an overestimate.}
@@
     (268 -
-    772),
+    607),
 suggesting that the entire \gls{MSM} population does not comprise a single
 network. In fact, our estimate of \gls{I} for these data
-    (524),
+    (410),
 was closer to the number of \gls{MSM} enrolled in the study (301) than to the
@@
     (482)
-was clearly an underestimate. The study enrolled almost 648 \gls{HIV} infected 
+was clearly an underestimate. The study enrolled almost 648 \gls{HIV}-infected 
 individuals, which should have been a lower bound on \gls{I}, although the
@@
 district were enrolled in the study, it is likely that this number represents
-the majority of \gls{HIV} infected persons. Our estimate was slightly lower
+the majority of \gls{HIV}-infected persons. Our estimate was slightly lower
     (939),
@@
 the posterior mean
-    (919),
+    (724),
 was much higher than this value. Due to the thorough population-level
 screening, it is most likely that the posterior mean represents an
-overestimate, especially considering the observed upward bias of \gls{ABC}
-estimates of \gls{I} observed on simulated data (\cref{fig:abcptm2}).
-It is also possible that the true sexual network extended outside the study's
-geographic area to include a larger number of infected individuals. The authors
-of the HET/Botswana data~\autocite{novitsky2013phylogenetic,
-novitsky2014impact} estimated that there were 1731 \gls{HIV}-positive
-individuals in the study area. \gls{HIV} sequences were obtained from
-approximately 70\% of these individuals. The estimated prevalence we obtained
-was much higher 
+overestimate, especially considering the upward bias of \gls{ABC} estimates of
+\gls{I} observed on simulated data (\cref{fig:abcpt2}). It is also possible
+that the true sexual network extended outside the study's geographic area to
+include a larger number of infected individuals. The authors of the
+HET/Botswana data~\autocite{novitsky2013phylogenetic, novitsky2014impact}
+estimated that there were 1731 \gls{HIV}-positive individuals in the study
+area. \gls{HIV} sequences were obtained from approximately 70\% of these
+individuals. The estimated prevalence we obtained was much higher 
     (5432),

### 2016-07-27 doc/thesis/methods_results.Rnw
Defence version of thesis with changes

@@
 using the \software{ape} package~\autocite{paradis2004ape}. \add{Both the tree
-kernel and the \gls{nltt} are pairwise statistics, and the \glspl{SVR}
+kernel and the \gls{nltt} are pairwise statistics, and the \glspl{SVM}
 classifiers we used to investigate them operate on pairwise distance matrices.}
@@
 constructed classifiers for \gls{alpha} based on the three tree shape
-statistics discussed above. First, we used the \software{kernlab}
-package~\autocite{zeileis2004kernlab} to create a \gls{kSVR} classifier using
-the computed kernel matrix. Second, we used the \software{e1071}
-package~\autocite{meyer2015e1071} to create an ordinary \gls{SVR} classifier
-using the pairwise \gls{nltt} matrix. Finally, we performed an ordinary linear
-regression of \gls{alpha} against Sackin's index. Each of these classifiers was
-evaluated with 1000 two-fold cross-validations. We also performed a \gls{kPCA}
-projection of the kernel matrix, and used it to visualize the separation of the
-different \gls{alpha} values in the tree kernel's feature space. A schematic of
-this experiment is presented in \cref{fig:kernelexpt}.
+statistics discussed above. The \software{kernlab}
+package~\autocite{zeileis2004kernlab} was used to create a \gls{kSVM}
+classifier using the computed kernel matrix, and the \software{e1071}
+package~\autocite{meyer2015e1071} was used to create ordinary \gls{SVM}
+classifiers using the pairwise \gls{nltt} matrix and Sackin's index values.
+\del{Finally, we performed an ordinary linear regression of \gls{alpha} against
+Sackin's index.} Each of these classifiers was evaluated with 1000 two-fold
+cross-validations \add{with equally-sized folds}. We also performed a
+\gls{kPCA} projection of the kernel matrix, and used it to visualize the
+separation of the different \gls{alpha} values in the tree kernel's feature
+space. A schematic of this experiment is presented in \cref{fig:kernelexpt}.
 
@@
 \gls{alphaess} = 0.95. The second run was performed with 2000 particles, 10
-simulated transmission trees per particle, and \gls{alphaess} = 0.99. To
+simulated transmission trees per particle, and \gls{alphaess} = 0.97. To
 investigate the extent to which errors in the estimated \gls{BA} parameters
@@
         rownames(m) <- sub("../../simulations/", "", rownames(m))
+        rownames(m) <- sub(".bk.2", "", rownames(m))
         d <- merge(d, m, by=0, suffixes=c("", ".1"))
@@
     m.stats <- get.treestats("m")
-    N.stats <- get.treestats("N")
+    N.stats <- get.treestats("N.bk.2")
 
@@
         if (param %in% c('alpha', 'm')) {
-            d[,list(rsquared=mean(rsquared)), by=c('nsimnode', 'ntip')]
+            d[,list(rsquared=mean(rsquared), accuracy=mean(accuracy)), by=c('nsimnode', 'ntip')]
         } else if (param == 'I') {
-            d[,list(rsquared=mean(rsquared)), by=ntip]
+            d[,list(rsquared=mean(rsquared), accuracy=mean(accuracy)), by=ntip]
         } else {
-            d[,list(rsquared=mean(rsquared)), by=c('I', 'ntip')]
+            d[,list(rsquared=mean(rsquared), accuracy=mean(accuracy)), by=c('I', 'ntip')]
         }
@@
     nltt.I <- get.kernel.data('I', 'nltt-classifier')
-    nltt.N <- get.kernel.data('N', 'nltt-classifier')
+    nltt.N <- get.kernel.data('N.bk.2', 'nltt-classifier')
 @
 
-The accuracy of each classifier \add{on all \gls{BA} parameters is shown in}
-\del{varied based on the parameter being tested} \cref{fig:rsquared}.
-Classifiers based on \del{two other tree statistics,} the \gls{nltt} and
-Sackin's index generally exhibited worse performance than the tree kernel,
-although the magnitude of the disparity varied between the parameters
-(\cref{fig:rsquared}, centre and right). The results were largely robust to
-variations in the tree kernel meta-parameters \gls{lambda} and \gls{sigma},
-although accuracy varied between different epidemic and sampling scenarios
+The accuracy of each classifier, which is the proportion of trees assigned the
+correct parameter value, \add{is shown in} \del{varied based on the parameter
+being tested} \cref{fig:rsquared} \add{Since there were three possible values,
+random guessing would produce an accuracy of 0.33.} Classifiers based on
+\del{two other tree statistics,} the \gls{nltt} and Sackin's index generally
+exhibited worse performance than the tree kernel, although the magnitude of the
+disparity varied between the parameters (\cref{fig:rsquared}, centre and
+right). The results were largely robust to variations in the tree kernel
+meta-parameters \gls{lambda} and \gls{sigma}, although accuracy varied between
+different epidemic and sampling scenarios
 (\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}). \add{For all
@@
 tips represented. However, for \gls{m}, both the number and proportion of
-sampled tips had a strong impact or the accuracy of the \gls{kSVR}
+sampled tips had a strong impact or the accuracy of the \gls{kSVM}
 (\cref{fig:mcrossv}).}
@@
     \caption[
-        Cross-validation accuracy of kernel-SVR, nLTT-based SVR, and Sackin's
-        index regression classifiers for BA model parameters.
+        Cross-validation accuracy of kernel-SVM, nLTT-based SVM, and Sackin's
+        SVM classifiers for BA model parameters.
     ]{
-        Cross-validation accuracy of kernel-SVR classifier (left), SVR
-        classifier using \gls{nltt} (centre), and linear regression using
-        Sackin's index (right) for \gls{BA} model parameters. Kernel
-        meta-parameters were set to $\gls{lambda} = 0.3$ and $\gls{sigma} = 4$.
-        Each point was calculated based on 300 simulated transmission trees
-        over networks with three different values of the parameter being
-        tested, assuming perfect knowledge of the other parameters. Vertical
-        lines are empirical 95\% confidence intervals based on 1000 two-fold
-        cross-validations. \add{The classifiers for \gls{I} were not evaluated
-        with 1000-tip trees, because one of the tested \gls{I} values was 500,
-        and it is not possible to sample a tree of size 1000 from 500 infected
-        individuals.}
+        Cross-validation accuracy of kernel-SVM classifier (left), SVM
+        classifier using \gls{nltt} (centre), and SVM using Sackin's index
+        (right) for \gls{BA} model parameters. Kernel meta-parameters were set
+        to $\gls{lambda} = 0.3$ and $\gls{sigma} = 4$. Each point was
+        calculated based on 300 simulated transmission trees over networks with
+        three different values of the parameter being tested, assuming perfect
+        knowledge of the other parameters. Vertical lines are empirical 95\%
+        confidence intervals based on 1000 two-fold cross-validations. \add{The
+        classifiers for \gls{I} were not evaluated with 1000-tip trees, because
+        one of the tested \gls{I} values was 500, and it is not possible to
+        sample a tree of size 1000 from 500 infected individuals.}
     }
@@
 
-The \gls{kSVR} classifier for \gls{alpha} had an average $R^2$ of 
-    \Sexpr{kernel.alpha[,round(mean(rsquared), 2)]},
+The \gls{kSVM} classifier for \gls{alpha} had an average accuracy of 
+    \Sexpr{kernel.alpha[,round(mean(accuracy), 2)]},
 compared to 
-    \Sexpr{nltt.alpha[,round(mean(rsquared), 2)]}
-for the \gls{nltt}-based SVR, and
-    \Sexpr{sackin.alpha[,round(mean(rsquared), 2)]}
-for the linear regression against Sackin's index. There was little variation
-about the mean for different tree and epidemic sizes. No classifier could
-accurately identify \gls{m} in any epidemic scenario, with average $R^2$ values
-of 
-  \Sexpr{kernel.m[,round(mean(rsquared), 2)]} for \gls{kSVR},
-  \Sexpr{nltt.m[,round(mean(rsquared), 2)]} for the \gls{nltt}, and
-  \Sexpr{sackin.m[,round(mean(rsquared), 2)]}
+    \Sexpr{nltt.alpha[,round(mean(accuracy), 2)]}
+for the \gls{nltt}-based SVM, and
+    \Sexpr{sackin.alpha[,round(mean(accuracy), 2)]}
+for Sackin's index. There was little variation about the mean for different
+tree and epidemic sizes. No classifier could accurately identify \gls{m} in any
+epidemic scenario, with average accuracy values of 
+  \Sexpr{kernel.m[,round(mean(accuracy), 2)]} for \gls{kSVM},
+  \Sexpr{nltt.m[,round(mean(accuracy), 2)]} for the \gls{nltt}, and
+  \Sexpr{sackin.m[,round(mean(accuracy), 2)]}
 for Sackin's index. Again, there was little variation in accuracy between
-epidemic scenarios, although the accuracy of the \gls{kSVR} was slightly higher
+epidemic scenarios, although the accuracy of the \gls{kSVM} was slightly higher
 on 1000-tip trees 
-    (average $R^2$ 
-     \Sexpr{kernel.m[ntip == 100,round(mean(rsquared), 2)]},
-     \Sexpr{kernel.m[ntip == 500,round(mean(rsquared), 2)]},
-     \Sexpr{kernel.m[ntip == 1000,round(mean(rsquared), 2)]}
+    (average 
+     \Sexpr{kernel.m[ntip == 100,round(mean(accuracy), 2)]},
+     \Sexpr{kernel.m[ntip == 500,round(mean(accuracy), 2)]},
+     \Sexpr{kernel.m[ntip == 1000,round(mean(accuracy), 2)]}
      for 100, 500, and 1000 tips respectively).
@@
 with}\add{appeared to be strongly influenced by} the number of tips in the
-tree. For 100-tip trees, the average $R^2$ values were
-  \Sexpr{kernel.I[ntip == 100, round(mean(rsquared), 2)]},
-  \Sexpr{nltt.I[ntip == 100, round(mean(rsquared), 2)]}, and
-  \Sexpr{sackin.I[ntip == 100, round(mean(rsquared), 2)]}
+tree. For 100-tip trees, the average accuracy values were
+  \Sexpr{kernel.I[ntip == 100, round(mean(accuracy), 2)]},
+  \Sexpr{nltt.I[ntip == 100, round(mean(accuracy), 2)]}, and
+  \Sexpr{sackin.I[ntip == 100, round(mean(accuracy), 2)]}
 for the tree kernel, \gls{nltt}, and Sackin's index respectively. For 500-tip
 trees, the values increased to
-  \Sexpr{kernel.I[ntip == 500, round(mean(rsquared), 2)]},
-  \Sexpr{nltt.I[ntip == 500, round(mean(rsquared), 2)]}, and
-  \Sexpr{sackin.I[ntip == 500, round(mean(rsquared), 2)]}.
+  \Sexpr{kernel.I[ntip == 500, round(mean(accuracy), 2)]},
+  \Sexpr{nltt.I[ntip == 500, round(mean(accuracy), 2)]}, and
+  \Sexpr{sackin.I[ntip == 500, round(mean(accuracy), 2)]}.
 Finally, the performance of classifiers for $N$ depended heavily on the
-epidemic scenario. The $R^2$ of the \gls{kSVR} classifier ranged from
-  \Sexpr{kernel.N[,round(min(rsquared), 2)]}
+epidemic scenario. The accuracy of the \gls{kSVM} classifier ranged from
+  \Sexpr{kernel.N[,round(min(accuracy), 2)]}
 for the smallest epidemic and smallest sample size, to
-  \Sexpr{kernel.N[,round(max(rsquared), 2)]}
-for the largest. Likewise, $R^2$ for the \gls{nltt}-based SVR ranged from 
-  \Sexpr{nltt.N[,round(min(rsquared), 2)]}
+  \Sexpr{kernel.N[,round(max(accuracy), 2)]}
+for the largest. Likewise, accuracy for the \gls{nltt}-based SVM ranged from 
+  \Sexpr{nltt.N[,round(min(accuracy), 2)]}
 to
-  \Sexpr{nltt.N[,round(max(rsquared), 2)]}.
+  \Sexpr{nltt.N[,round(max(accuracy), 2)]}.
 Sackin's index did not accurately classify $N$ in any scenario, with an average
-$R^2$ of
-  \Sexpr{sackin.N[,round(mean(rsquared), 2)]}
+accuracy of
+  \Sexpr{sackin.N[,round(mean(accuracy), 2)]}
 and little variation between scenarios.
@@
 \del{The accuracy of grid search estimates largely paralleled that of the
-\gls{kSVR} classifiers.} \Cref{fig:gridest} shows point estimates and
+\gls{kSVM} classifiers.} \Cref{fig:gridest} shows point estimates and
 \add{50\%} and 95\% highest density intervals for each of the \gls{BA}
@@
     $\Sexpr{tests["N", round(grid.err)]}$\%
-of the respective grids. For \gls{I} and \gls{N}, relative errors may be more
-appropriate to consider. An overestimate of 100 individuals would be very
-misleading if the true population was only of size 100, but almost negligible
-in a population of size 5000. The mean relative errors for \gls{I} and \gls{N}
-respectively were
+of the respective grids. Note that \cref{fig:gridest} contains only one
+replicate data point per parameter value, out of ten total. For \gls{I} and
+\gls{N}, relative errors may be more appropriate to consider. An overestimate
+of 100 individuals would be very misleading if the true population was only of
+size 100, but almost negligible in a population of size 5000. The mean relative
+errors for \gls{I} and \gls{N} respectively were
     $\Sexpr{tests["I", round(rel.err)]}$\% and 
@@
 
-\Cref{fig:abcptm2} shows \del{MAP} \add{posterior mean} point estimates of the
-\gls{BA} model parameters obtained with \gls{ABC} on simulated data. The
-estimates shown correspond only to the simulations where \gls{m} was set to 2,
-however the results for $\gls{m} = 3$ and $\gls{m} = 4$ were similar
-(\cref{fig:abcptm3,fig:abcptm4}). Average boundaries of 95\% \gls{HPD}
-intervals are given in \cref{tab:abchpd}.
+\Cref{fig:abcpt} shows \del{MAP} \add{stratified posterior mean} point
+estimates of the \gls{BA} model parameters \add{\gls{alpha} and \gls{I}}
+obtained with \gls{ABC} on simulated data. \add{The parameters $m$ and $N$ were
+not identifiable with ABC for any parameter combinations (\cref{fig:abcpt2}).}
+\del{The estimates shown correspond only to the simulations where \gls{m} was
+set to 2, however the results for $\gls{m} = 3$ and $\gls{m} = 4$ were
+similar.} Average boundaries of 95\% \gls{HPD} intervals are given in
+\cref{tab:abchpd}.
 
@@
     \centering
-    \includegraphics[width=\textwidth]{abc-point-estimate-m2}
+    \includegraphics[width=\textwidth]{abc-boxplot}
     \caption[
-        Posterior mean point estimates for \gls{BA} model parameters obtained
-        by running \software{netabc} on simulated data, for simulations with
-        $\gls{m} = 2$.
+        Posterior mean point estimates for BA model parameters $\alpha$ and $I$
+        obtained by running \textit{netabc} on simulated data, stratified by
+        true parameter values.
     ]{
-        Posterior mean point estimates for \gls{BA} model parameters obtained
-        by running \software{netabc} on simulated data, for simulations with
-        $\gls{m} = 2$. Dashed lines indicate true values. (A) Estimates of
-        \gls{alpha} and \gls{I} which were varied in these simulations against
-        known values. (B) Estimates of \gls{m} and \gls{N} which were held
-        fixed in these simulations at the values $\gls{m} = 2$ and 
-        $\gls{N} = 5000$.
-    }
-    \label{fig:abcptm2}
+        Posterior mean point estimates for BA model parameters $\alpha$ and $I$
+        obtained by running \textit{netabc} on simulated data, stratified by
+        true parameter values. First row of plots contains true versus
+        estimated values of $\alpha$; second row contains true versus estimated
+        values of $I$. Columns are stratified by $\alpha$, $I$, and $m$
+        respectively. Dashed lines indicate true values. 
+    }
+    \label{fig:abcpt}
 \end{figure}
 
-\begin{table*}[ht]
+\begin{table}[ht]
     \centering
@@
     \label{tab:abchpd}
-\end{table*}
+\end{table}
 
@@
 \gls{m} and \gls{N} were roughly equal across all simulations
-(\cref{fig:abcptm2,fig:abcptm3,fig:abcptm4}), \glspl{GLM} were not fitted for
-these parameters. The estimated coefficients are shown in
-\cref{tab:glmalpha,tab:glmI}. The \gls{GLM} were fitted using the inverse link
-function. That is, if $p$ is the true value of the parameter and $\hat{p}$ is a
-random variable representing our estimate of the parameter, the \gls{GLM}
-posits a relationship of the form}
+(\cref{fig:abcpt2}), \glspl{GLM} were not fitted for these parameters. The
+estimated coefficients are shown in \cref{tab:glmalpha,tab:glmI}. The \gls{GLM}
+were fitted using the inverse link function. That is, if $p$ is the true value
+of the parameter and $\hat{p}$ is a random variable representing our estimate
+of the parameter, the \gls{GLM} posits a relationship of the form}
 \[
@@
 Qualitatively, \gls{alpha} seemed to be only weakly identifiable between the
-values of 0 and 0.5 (\cref{fig:abcptm2}). The error in the estimated prevalence
+values of 0 and 0.5 (\cref{fig:abcpt}). The error in the estimated prevalence
 \gls{I} was slightly lower for smaller values of \gls{alpha}
@@
 \del{The accuracy of the parameter estimates obtained with ABC paralleled the
-results from the \gls{kSVR} classifier. Of the four parameters, $\alpha$ was
+results from the \gls{kSVM} classifier. Of the four parameters, $\alpha$ was
 the most accurately estimated, with point estimates having a median [IQR]
@@
     \centering
-    \includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior", best.plot)}}.pdf}
+    \includegraphics[width=0.8\textwidth]{{\Sexpr{file.path("abc-posterior", best.plot)}}.pdf}
     \caption[
@@
     \centering
-    \includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior-2d", best.plot)}}.pdf}
+    \includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior-2d", best.plot)}}.png}
     \caption[
@@
     \centering
-    \includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior", worst.plot)}}.pdf}
+    \includegraphics[width=0.8\textwidth]{{\Sexpr{file.path("abc-posterior", worst.plot)}}.pdf}
     \caption[
@@
     \centering
-    \includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior-2d", worst.plot)}}.pdf}
+    \includegraphics[width=\textwidth]{{\Sexpr{file.path("abc-posterior-2d", worst.plot)}}.png}
     \caption[
@@
    \Sexpr{d["N", round(hpd.max)]}].
-The one-dimensional marginal approximate posterior distributions for this
-simulation are shown in \cref{fig:mixed}. To test the effect of sampling bias,
+\del{The one-dimensional marginal approximate posterior distributions for this
+simulation are shown in \cref{fig:mixed}.} To test the effect of sampling bias,
 we sampled one transmission tree in a peer-driven fashion, where the
@@
    \Sexpr{p["N", floor(hpd.max)]}].
-The approximate posterior distributions are shown in \cref{fig:peerdriven}.
+\del{The approximate posterior distributions are shown in \cref{fig:peerdriven}.}
 Both of these results were in line with estimates obtained on other simulated
@@
 
-\add{\Cref{fig:marginal} shows the effect of performing marginal \gls{ABC}
-estimation of each of the \gls{BA} parameters on the same simulated
-transmission tree. The estimates of \gls{m} are apparently unaffected by
-marginalizing out the other parameters, corroborating the previous experiments'
-findings that \gls{m} is not an identifiable parameter from scaled tree shapes.
-The estimates of \gls{alpha} and \gls{N} were weakly improved in marginal
-estimation, which is consistent with the result from grid search that all
-\gls{N} values and \gls{alpha} values below 0.5 are weakly identifiable. On the
-other hand, both the accuracy and precision (\ie width of the \gls{HPD}
-intervals) of the estimates of \gls{I} were substantially improved when \gls{N}
-was marginalized out. Neither the accuracy nor the precision of any parameter
-estimates appeared to improve when more computing power was used for \gls{SMC}
-(\cref{fig:moresmc}).}
+<<marginal, include=FALSE>>=
+    d <- setDT(collect.data("../../simulations/abc-marginal/point-estimate/*", sep="\t"))
+
+    joint.alpha <- d[is.na(fixed), abs(alpha-true_alpha)]
+    marginal.alpha <- d[fixed == "m+N+I", abs(alpha-true_alpha)]
+
+    joint.N <- d[is.na(fixed), abs(N-true_N)]
+    marginal.N <- d[fixed == "I+alpha+m", abs(N-true_N)]
+
+    joint.I <- d[is.na(fixed), abs(I-true_I)]
+    marginal.I <- d[fixed == "N+alpha+m", abs(I-true_I)]
+
+    d <- setDT(collect.data("../../simulations/abc-better-smc/point-estimate/*"))
+
+    base.alpha <- d[nparticle == 1000, abs(alpha - true_alpha)]
+    better.alpha <- d[nparticle == 2000, abs(alpha - true_alpha)]
+
+    base.I <- d[nparticle == 1000, abs(I - true_I)]
+    better.I <- d[nparticle == 2000, abs(I - true_I)]
+
+    base.N <- d[nparticle == 1000, abs(N - true_N)]
+    better.N <- d[nparticle == 2000, abs(N - true_N)]
+@
+
+\add{\Cref{fig:marginal} shows the effect of performing marginal ABC estimation
+of each of the BA parameters on the same simulated transmission tree. The
+estimates of $m$ were apparently unaffected by marginalizing out the other
+parameters, corroborating the previous experiments' findings that $m$ is not an
+identifiable parameter from scaled tree shapes. Compared to allowing all
+parameters to vary, estimates of $\alpha$, $I$, and $N$ were improved by
+    \Sexpr{ round((joint.alpha - marginal.alpha) * 100 / joint.alpha) }\%,
+    \Sexpr{ round((joint.I - marginal.I) * 100 / joint.I) }\%, and
+    \Sexpr{ round((joint.N - marginal.N) * 100 / joint.N) }\%
+when all other parameters were fixed. \Cref{fig:better} shows the impact of
+increasing the number of particles, simulated datasets, and
+$\alpha_{\text{ESS}}$ parameter on the accuracy of a single simulation. The
+results of the two simulations were similar, but surprisingly, the results with
+higher SMC settings were slightly worse (by
+    \Sexpr{ round((better.alpha - base.alpha) * 100 / better.alpha) }\%,
+    \Sexpr{ round((better.I - base.I) * 100 / better.I) }\%, and
+    \Sexpr{ round((better.N - base.N) * 100 / better.N) }\%
+for $\alpha$, $I$, and $N$ respectively). However, the 50\% HPD interval for
+$I$ was closer to the true value of 2000 with the improved settings
+    (\Sexpr{ d[nparticle == 2000, round(I.lower.50)] } - 
+     \Sexpr{ d[nparticle == 2000, round(I.upper.50)] }, vs.
+     \Sexpr{ d[nparticle == 1000, round(I.lower.50)] } - 
+     \Sexpr{ d[nparticle == 1000, round(I.upper.50)] } with basic settings).
+The estimate of $m$, 3 in both cases, was unaffected by the settings.}
 
@@
     \centering
-    \includegraphics{stock/missing}
+    \includegraphics[width=0.7\textwidth]{better-smc}
     \caption[
@@
     }
-    \label{fig:moresmc}
+    \label{fig:better}
 \end{figure}
@@
 their data than sequences of other subtypes, we restricted our analysis to the
-subtype B sequences and labelled this dataset as \gls{MSM}. Three datasets
-(IDU/Estonia, HET/Uganda, and HET/Malawi) included both \textit{env} and
-\textit{gag} sequences. Each gene was analyzed separately to assess the
-robustness of \software{netabc} to the particular \gls{HIV} gene sequence used
-to estimate a transmission tree.}
+subtype B sequences and labelled this dataset as \gls{MSM}. Two datasets
+(HET/Uganda and HET/Malawi) included both \textit{env} and \textit{gag}
+sequences. Each gene was analyzed separately to assess the robustness of
+\software{netabc} to the particular \gls{HIV} gene sequence used to estimate a
+transmission tree. The IDU/Estonia data also sequenced both genes, but the
+highly variable coverage and high homology of the \textit{gag} sequences made
+it impossible to obtain a sufficiently large block of non-identical sequences
+to analyze. Therefore, we analyzed only \textit{env} for this dataset.}
 
@@
 \autocite{edgar2004muscle}, and alignments were manually inspected with
-\software{Seaview} version 4.4.2 \autocite{gouy2010seaview}. Phylogenies were
-constructed from the nucleotide alignments by approximate maximum likelihood
-using \software{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
+\software{Seaview} version 4.4.2 \autocite{gouy2010seaview}. \add{Duplicated
+sequences were removed with \textit{BioPython}.} Phylogenies were constructed
+from the nucleotide alignments by approximate maximum likelihood using
+\software{FastTree2} version 2.1.7 \autocite{price2010fasttree} with the
 \gls{GTR} model~\autocite{tavare1986some}. Transmission trees were estimated by
@@
 BEAST~\autocite{drummond2007beast}) as described
-previously~\autocite{poon2015phylodynamic}.
+previously~\autocite{poon2015phylodynamic}. \add{Due to the removal of
+duplicated sequences, all estimated transmission trees were fully binary.}
+
+\add{To check if our results were robust to the choice of phylogenetic
+reconstruction method, we built and reanalyzed phylogenies for the datasets
+with the lowest and highest estimated $\alpha$ values (mixed/Spain and
+IDU/Estonia) with \textit{RAxML}~\autocite{stamatakis2014raxml} with the
+GTR+$\Gamma$ model of sequence evolution and rate heterogeneity. The trees were
+rooted and time-scaled with \textit{Least Square
+Dating}~\autocite[\textit{LSD},][]{to2016fast}. For expediency, the analysis
+was run with the prior $m \sim \text{DiscreteUniform}(2, 5)$, which defines a
+smaller total search space than the prior allowing $m = 1$.}
 
@@
   stopifnot(pmean1[,dataset[which.max(I)]] == "novitsky2014")
-  stopifnot(pmean1[dataset != "novitsky2014", dataset[which.max(I)]] == "bctree")
+  stopifnot(pmean1[dataset != "novitsky2014", dataset[which.max(I)]] == "li2015")
 
@@
 posterior distributions were diffuse with heavy tails, rather than having most
-of their mass around the mode. When the value $\gls{m} = 1$ was disallowed by
-the prior, the separation between the \gls{IDU} datasets and the others became
-more striking (\cref{fig:abchpdm2}). All three \gls{IDU} datasets had estimated
-\gls{alpha} values at or above
-    \Sexpr{pmean2[c("zetterberg2004", "niculescu2015", "bctree"), round(min(alpha), 2)]}.
-The estimate for the MSM/Beijing data was slightly lower
-    (\Sexpr{pmean2["wang2015", round(alpha, 2)]}) and
-the estimates for the seven remaining non-\gls{IDU} datasets were bounded above by
-    \Sexpr{pmean2[c("novitsky2014", "mccormack2002", "grabowski2014", "kao2011", "little2014", "li2015", "cuevas2009"), round(max(alpha), 2)]}.}
+of their mass around the mode.}
 
@@
         Posterior means and 50\%/95\% \gls{HPD} intervals for parameters of the
-        \gls{BA} network model, fitted to three \gls{HIV} datasets where both
+        \gls{BA} network model, fitted to two \gls{HIV} datasets where both
         \textit{gag} and \textit{env} genes were sequenced.
@@
         \gls{HPD} intervals (lines) for parameters of the \gls{BA} network model,
-        fitted to three \gls{HIV} datasets where both \textit{gag} and
+        fitted to two \gls{HIV} datasets where both \textit{gag} and
         \textit{env} genes were sequenced.
@@
     ratios <- data.table(dataset=rep(datasets, each=2),
-                         gene=rep(c("env", "gag"), 3),
+                         gene=rep(c("env", "gag"), 2),
                          ratio=sapply(trees, get.ratio))
-    stopifnot(ratios[gene == "env", mean(ratio)] <
-              ratios[gene == "gag", mean(ratio)])
 @
@@
 (IDU/Estonia) and
-    \Sexpr{pmean1["bctree", round(I)]}
-(IDU/Canada). The HET/Botswana data had a much higher estimated \gls{I} value
+    \Sexpr{pmean1["li2015", round(I)]}
+(MSM/Shanghai). The HET/Botswana data had a much higher estimated \gls{I} value
     (\Sexpr{pmean1["novitsky2014", round(I)]})
@@
 increased sampling density. When both \textit{gag} and \textit{env} sequences
-were analyzed, the estimates from the \textit{env} data were substantially
-higher (\cref{fig:genes}). The ratios of internal to terminal branch lengths
-were lower for the \textit{env} trees than the \textit{gag} trees (average 
-    \Sexpr{ratios[gene == "env", round(mean(ratio), 2)]}
-for \textit{env} and 
-    \Sexpr{ratios[gene == "gag", round(mean(ratio), 2)]}
-for \textit{gag}). As with \gls{alpha}, many of the posterior means were
-outside the 50\% \gls{HPD} intervals, suggesting diffuse posterior
-distributions. The values of \gls{I} were fairly robust to restricting the
-prior to $\gls{m} \geq 2$ (compare \cref{fig:abchpd,fig:abchpdm2}), although 
-the 95\% \gls{HPD} intervals were slightly narrower (average width
-    \Sexpr{round(mean(up1[,I] - low1[,I]))}
-for $\gls{m} \geq 1$ and
-    \Sexpr{round(mean(up2[,I] - low2[,I]))}
-for $\gls{m} \geq 2$).}
+were analyzed, the estimates from the \textit{env} data were higher
+(HET/Uganda,
+    \Sexpr{pmean1["grabowski2014", round(I)]} for \textit{gag} vs.
+    \Sexpr{pmean1.env["grabowski2014", round(I)]} for \textit{env};
+HET/Malawi,
+    \Sexpr{pmean1["mccormack2002", round(I)]} for \textit{gag} vs.
+    \Sexpr{pmean1.env["mccormack2002", round(I)]} for \textit{env}).
+As with \gls{alpha}, many of the posterior means were outside the 50\%
+\gls{HPD} intervals, suggesting diffuse posterior distributions.}
 
@@
 from 0 (all the mass on the estimated value) to 5 (the entire prior region).
-When the value $\gls{m} = 1$ was disallowed by the prior, the posterior means
-for all but the HET/Botswana data took on the value 3, with the \gls{HPD}
-intervals spanning the entire prior region. This is very similar to the results 
-observed for \gls{m} on simulated data (\cref{fig:abcptm2,tab:abchpd}), and
-suggests that \gls{m} is not identifiable from these data with this prior. On
-the other hand, the estimate and 95\% \gls{HPD} interval for the HET/Botswana
-data (4 [2-5]) was the same for both priors.}
-
-\add{Estimates of \gls{N} were  mostly uninformative, with very similar estimates
+Estimates of \gls{N} were  mostly uninformative, with very similar estimates
 for all datasets (mean
@@
 posterior mean always fell around the upper two-thirds mark of the region
-allowed by the prior. The results did not change appreciably when the value
-$\gls{m} = 1$ was assigned zero prior probability.}
+allowed by the prior (\cref{fig:abcpt2}).}
+
+\add{When the value $m = 1$ was disallowed by the prior, the separation in
+$\alpha$ between the IDU datasets and the others became more striking
+(\cref{fig:abchpdm2}). All three IDU datasets had estimated $\alpha$ values at
+or above
+    \Sexpr{pmean2[c("zetterberg2004", "niculescu2015"), round(min(alpha), 2)]}.
+The estimate for the MSM/Beijing data was slightly lower
+    (\Sexpr{pmean2["wang2015", round(alpha, 2)]}) and
+the estimates for the seven remaining non-IDU datasets were bounded above by
+    \Sexpr{pmean2[c("novitsky2014", "mccormack2002", "grabowski2014", "kao2011", "little2014", "li2015", "cuevas2009"), round(max(alpha), 2)]}. 
+The values of $I$ were fairly robust to the choice of prior (compare
+\cref{fig:abchpd,fig:abchpdm2}), although the 95\% HPD intervals were slightly
+narrower (average width
+    \Sexpr{round(mean(up1[,I] - low1[,I]))}
+for $m \geq 1$ and
+    \Sexpr{round(mean(up2[,I] - low2[,I]))}
+for $m \geq 2$). The posterior means of $m$ for all but the HET/Botswana data
+took on the value 3 with this prior, with the HPD intervals spanning the entire
+prior region. This is very similar to the results observed for $m$ on simulated
+data (\cref{tab:abchpd}), and suggests that $m$ is not identifiable from these
+data with this prior. The results for $N$ did not change appreciably between
+the two choices of prior.}
+
+<<raxml, include=FALSE>>=
+    d.raxml <- fread(Sys.glob("../../simulations/aggregates/raxml-hpd/*"))
+@
+
+\add{For the two datasets we reanalyzed using
+RAxML~\autocite{stamatakis2014raxml} and LSD~\autocite{to2016fast}, $\alpha$
+was relatively robust to the choice of method (\cref{fig:methods}, posterior
+means 
+    \Sexpr{pmean2["cuevas2009", round(alpha, 2)]}
+vs. 
+    \Sexpr{d.raxml[dataset == "cuevas2009" & parameter == "alpha", round(mean, 2)]}
+for mixed/Spain and 
+    \Sexpr{pmean2["zetterberg2004", round(alpha, 2)]}
+vs.
+    \Sexpr{d.raxml[dataset == "zetterberg2004" & parameter == "alpha", round(mean, 2)]}
+for IDU/Estonia). However, the estimates of $I$ were about twice as high when
+RAxML was used instead of FastTree to reconstruct the trees 
+    (\Sexpr{pmean2["zetterberg2004", round(I)]} 
+     vs. \Sexpr{d.raxml[dataset == "zetterberg2004" & parameter == "I", round(mean)]} for IDU/Estonia, 
+    \Sexpr{pmean2["cuevas2009", round(I)]}
+    vs. \Sexpr{d.raxml[dataset == "cuevas2009" & parameter == "I", round(mean)]} for mixed/Spain). }
 

### 2016-07-27 doc/thesis/methods_results.tex
Defence version of thesis with changes

--- 2016-06-27 doc/thesis/methods_results.tex
@@
 
-The accuracy of each classifier \add{on all \gls{BA} parameters is shown in}
-\del{varied based on the parameter being tested} \cref{fig:rsquared}.
-Classifiers based on \del{two other tree statistics,} the \gls{nltt} and
-Sackin's index generally exhibited worse performance than the tree kernel,
-although the magnitude of the disparity varied between the parameters
-(\cref{fig:rsquared}, centre and right). The results were largely robust to
-variations in the tree kernel meta-parameters \gls{lambda} and \gls{sigma},
-although accuracy varied between different epidemic and sampling scenarios
+The accuracy of each classifier, which is the proportion of trees assigned the
+correct parameter value, \add{is shown in} \del{varied based on the parameter
+being tested} \cref{fig:rsquared} \add{Since there were three possible values,
+random guessing would produce an accuracy of 0.33.} Classifiers based on
+\del{two other tree statistics,} the \gls{nltt} and Sackin's index generally
+exhibited worse performance than the tree kernel, although the magnitude of the
+disparity varied between the parameters (\cref{fig:rsquared}, centre and
+right). The results were largely robust to variations in the tree kernel
+meta-parameters \gls{lambda} and \gls{sigma}, although accuracy varied between
+different epidemic and sampling scenarios
 (\cref{fig:alphacrossv,fig:mcrossv,fig:Icrossv,fig:Ncrossv}). \add{For all
@@
     \caption[
-        Cross-validation accuracy of kernel-SVR, nLTT-based SVR, and Sackin's
-        index regression classifiers for BA model parameters.
-    ]{
-        Cross-validation accuracy of kernel-SVR classifier (left), SVR
-        classifier using \gls{nltt} (centre), and linear regression using
-        Sackin's index (right) for \gls{BA} model parameters. Kernel
-        meta-parameters were set to $\gls{lambda} = 0.3$ and $\gls{sigma} = 4$.
-        Each point was calculated based on 300 simulated transmission trees
-        over networks with three different values of the parameter being
-        tested, assuming perfect knowledge of the other parameters. Vertical
-        lines are empirical 95\% confidence intervals based on 1000 two-fold
-        cross-validations. \add{The classifiers for \gls{I} were not evaluated
-        with 1000-tip trees, because one of the tested \gls{I} values was 500,
-        and it is not possible to sample a tree of size 1000 from 500 infected
-        individuals.}
+        Cross-validation accuracy of kernel-SVM, nLTT-based SVM, and Sackin's
+        SVM classifiers for BA model parameters.
+    ]{
+        Cross-validation accuracy of kernel-SVM classifier (left), SVM
+        classifier using \gls{nltt} (centre), and SVM using Sackin's index
+        (right) for \gls{BA} model parameters. Kernel meta-parameters were set
+        to $\gls{lambda} = 0.3$ and $\gls{sigma} = 4$. Each point was
+        calculated based on 300 simulated transmission trees over networks with
+        three different values of the parameter being tested, assuming perfect
+        knowledge of the other parameters. Vertical lines are empirical 95\%
+        confidence intervals based on 1000 two-fold cross-validations. \add{The
+        classifiers for \gls{I} were not evaluated with 1000-tip trees, because
+        one of the tested \gls{I} values was 500, and it is not possible to
+        sample a tree of size 1000 from 500 infected individuals.}
     }
@@
 
-The \gls{kSVR} classifier for \gls{alpha} had an average $R^2$ of 
+The \gls{kSVM} classifier for \gls{alpha} had an average accuracy of 
     0.92,
 compared to 
-    0.56
-for the \gls{nltt}-based SVR, and
-    0.75
-for the linear regression against Sackin's index. There was little variation
-about the mean for different tree and epidemic sizes. No classifier could
-accurately identify \gls{m} in any epidemic scenario, with average $R^2$ values
-of 
-  0.12 for \gls{kSVR},
-  0.01 for the \gls{nltt}, and
-  0.06
+    0.6
+for the \gls{nltt}-based SVM, and
+    0.77
+for Sackin's index. There was little variation about the mean for different
+tree and epidemic sizes. No classifier could accurately identify \gls{m} in any
+epidemic scenario, with average accuracy values of 
+  0.35 for \gls{kSVM},
+  0.32 for the \gls{nltt}, and
+  0.38
 for Sackin's index. Again, there was little variation in accuracy between
-epidemic scenarios, although the accuracy of the \gls{kSVR} was slightly higher
+epidemic scenarios, although the accuracy of the \gls{kSVM} was slightly higher
 on 1000-tip trees 
-    (average $R^2$ 
-     0.01,
-     0.11,
-     0.32
+    (average 
+     0.33,
+     0.35,
+     0.38
      for 100, 500, and 1000 tips respectively).
@@
 with}\add{appeared to be strongly influenced by} the number of tips in the
-tree. For 100-tip trees, the average $R^2$ values were
-  0.7,
-  0.55, and
-  0.02
+tree. For 100-tip trees, the average accuracy values were
+  0.59,
+  0.58, and
+  0.34
 for the tree kernel, \gls{nltt}, and Sackin's index respectively. For 500-tip
 trees, the values increased to
-  0.93,
-  0.83, and
-  0.07.
+  0.99,
+  0.76, and
+  0.37.
 Finally, the performance of classifiers for $N$ depended heavily on the
-epidemic scenario. The $R^2$ of the \gls{kSVR} classifier ranged from
-  0.08
+epidemic scenario. The accuracy of the \gls{kSVM} classifier ranged from
+  0.36
 for the smallest epidemic and smallest sample size, to
-  0.82
-for the largest. Likewise, $R^2$ for the \gls{nltt}-based SVR ranged from 
-  0.01
+  0.81
+for the largest. Likewise, accuracy for the \gls{nltt}-based SVM ranged from 
+  0.33
 to
-  0.54.
+  0.63.
 Sackin's index did not accurately classify $N$ in any scenario, with an average
-$R^2$ of
-  0.03
+accuracy of
+  0.35
 and little variation between scenarios.
@@
     $17$\%
-of the respective grids. For \gls{I} and \gls{N}, relative errors may be more
-appropriate to consider. An overestimate of 100 individuals would be very
-misleading if the true population was only of size 100, but almost negligible
-in a population of size 5000. The mean relative errors for \gls{I} and \gls{N}
-respectively were
+of the respective grids. Note that \cref{fig:gridest} contains only one
+replicate data point per parameter value, out of ten total. For \gls{I} and
+\gls{N}, relative errors may be more appropriate to consider. An overestimate
+of 100 individuals would be very misleading if the true population was only of
+size 100, but almost negligible in a population of size 5000. The mean relative
+errors for \gls{I} and \gls{N} respectively were
     $18$\% and 
@@
     \centering
-    \includegraphics[width=\textwidth]{abc-point-estimate-m2}
-    \caption[
-        Posterior mean point estimates for \gls{BA} model parameters obtained
-        by running \software{netabc} on simulated data, for simulations with
-        $\gls{m} = 2$.
-    ]{
-        Posterior mean point estimates for \gls{BA} model parameters obtained
-        by running \software{netabc} on simulated data, for simulations with
-        $\gls{m} = 2$. Dashed lines indicate true values. (A) Estimates of
-        \gls{alpha} and \gls{I} which were varied in these simulations against
-        known values. (B) Estimates of \gls{m} and \gls{N} which were held
-        fixed in these simulations at the values $\gls{m} = 2$ and 
-        $\gls{N} = 5000$.
-    }
-    \label{fig:abcptm2}
+    \includegraphics[width=\textwidth]{abc-boxplot}
+    \caption[
+        Posterior mean point estimates for BA model parameters $\alpha$ and $I$
+        obtained by running \textit{netabc} on simulated data, stratified by
+        true parameter values.
+    ]{
+        Posterior mean point estimates for BA model parameters $\alpha$ and $I$
+        obtained by running \textit{netabc} on simulated data, stratified by
+        true parameter values. First row of plots contains true versus
+        estimated values of $\alpha$; second row contains true versus estimated
+        values of $I$. Columns are stratified by $\alpha$, $I$, and $m$
+        respectively. Dashed lines indicate true values. 
+    }
+    \label{fig:abcpt}
 \end{figure}
 
-\begin{table*}[ht]
+\begin{table}[ht]
     \centering
@@
 % latex table generated in R 3.2.3 by xtable 1.8-2 package
-% Mon Jun 27 14:24:46 2016
+% Mon Jul 18 11:53:42 2016
 \begin{table}[ht]
@@
     \centering
-    \includegraphics[width=\textwidth]{{abc-posterior/1.5_1000_4_5000_0}.pdf}
+    \includegraphics[width=0.8\textwidth]{{abc-posterior/1.5_1000_4_5000_0}.pdf}
     \caption[
@@
     \centering
-    \includegraphics[width=\textwidth]{{abc-posterior-2d/1.5_1000_4_5000_0}.pdf}
+    \includegraphics[width=\textwidth]{{abc-posterior-2d/1.5_1000_4_5000_0}.png}
     \caption[
@@
     \centering
-    \includegraphics[width=\textwidth]{{abc-posterior/0.0_2000_2_5000_0}.pdf}
+    \includegraphics[width=0.8\textwidth]{{abc-posterior/0.0_2000_2_5000_0}.pdf}
     \caption[
@@
     \centering
-    \includegraphics[width=\textwidth]{{abc-posterior-2d/0.0_2000_2_5000_0}.pdf}
+    \includegraphics[width=\textwidth]{{abc-posterior-2d/0.0_2000_2_5000_0}.png}
     \caption[
@@
    14977].
-The one-dimensional marginal approximate posterior distributions for this
-simulation are shown in \cref{fig:mixed}. To test the effect of sampling bias,
+\del{The one-dimensional marginal approximate posterior distributions for this
+simulation are shown in \cref{fig:mixed}.} To test the effect of sampling bias,
 we sampled one transmission tree in a peer-driven fashion, where the
@@
    14780].
-The approximate posterior distributions are shown in \cref{fig:peerdriven}.
+\del{The approximate posterior distributions are shown in \cref{fig:peerdriven}.}
 Both of these results were in line with estimates obtained on other simulated
@@
 
-\add{\Cref{fig:marginal} shows the effect of performing marginal \gls{ABC}
-estimation of each of the \gls{BA} parameters on the same simulated
-transmission tree. The estimates of \gls{m} are apparently unaffected by
-marginalizing out the other parameters, corroborating the previous experiments'
-findings that \gls{m} is not an identifiable parameter from scaled tree shapes.
-The estimates of \gls{alpha} and \gls{N} were weakly improved in marginal
-estimation, which is consistent with the result from grid search that all
-\gls{N} values and \gls{alpha} values below 0.5 are weakly identifiable. On the
-other hand, both the accuracy and precision (\ie width of the \gls{HPD}
-intervals) of the estimates of \gls{I} were substantially improved when \gls{N}
-was marginalized out. Neither the accuracy nor the precision of any parameter
-estimates appeared to improve when more computing power was used for \gls{SMC}
-(\cref{fig:moresmc}).}
+\add{\Cref{fig:marginal} shows the effect of performing marginal ABC estimation
+of each of the BA parameters on the same simulated transmission tree. The
+estimates of $m$ were apparently unaffected by marginalizing out the other
+parameters, corroborating the previous experiments' findings that $m$ is not an
+identifiable parameter from scaled tree shapes. Compared to allowing all
+parameters to vary, estimates of $\alpha$, $I$, and $N$ were improved by
+    41\%,
+    59\%, and
+    46\%
+when all other parameters were fixed. \Cref{fig:better} shows the impact of
+increasing the number of particles, simulated datasets, and
+$\alpha_{\text{ESS}}$ parameter on the accuracy of a single simulation. The
+results of the two simulations were similar, but surprisingly, the results with
+higher SMC settings were slightly worse (by
+    10\%,
+    8\%, and
+    11\%
+for $\alpha$, $I$, and $N$ respectively). However, the 50\% HPD interval for
+$I$ was closer to the true value of 2000 with the improved settings
+    (2338 - 
+     3423, vs.
+     2810 - 
+     3767 with basic settings).
+The estimate of $m$, 3 in both cases, was unaffected by the settings.}
 
@@
 were all sub-linear, ranging from 
-    0.9
+    0.83
 for the \gls{IDU}/Estonia data to
@@
 for \gls{IDU},
-    0.43
+    0.41
 for primarily heterosexual risk, and
-    0.39
+    0.37
 for \gls{MSM}. These values were obtained with \textit{gag} for the datasets
@@
 posterior distributions were diffuse with heavy tails, rather than having most
-of their mass around the mode. When the value $\gls{m} = 1$ was disallowed by
-the prior, the separation between the \gls{IDU} datasets and the others became
-more striking (\cref{fig:abchpdm2}). All three \gls{IDU} datasets had estimated
-\gls{alpha} values at or above
-    0.9.
-The estimate for the MSM/Beijing data was slightly lower
-    (0.85) and
-the estimates for the seven remaining non-\gls{IDU} datasets were bounded above by
-    0.58.}
+of their mass around the mode.}
 
@@
 the prevalence \gls{I} were between
-    372
+    373
 (IDU/Estonia) and
-    1481
-(IDU/Canada). The HET/Botswana data had a much higher estimated \gls{I} value
+    1391
+(MSM/Shanghai). The HET/Botswana data had a much higher estimated \gls{I} value
     (5432)
@@
 prevalence (Spearman correlation, 
-    $p =0.6$),
+    $p =0.7$),
 indicating that the higher prevalence estimates were not simply due to
 increased sampling density. When both \textit{gag} and \textit{env} sequences
-were analyzed, the estimates from the \textit{env} data were substantially
-higher (\cref{fig:genes}). The ratios of internal to terminal branch lengths
-were lower for the \textit{env} trees than the \textit{gag} trees (average 
-    0.47
-for \textit{env} and 
-    0.76
-for \textit{gag}). As with \gls{alpha}, many of the posterior means were
-outside the 50\% \gls{HPD} intervals, suggesting diffuse posterior
-distributions. The values of \gls{I} were fairly robust to restricting the
-prior to $\gls{m} \geq 2$ (compare \cref{fig:abchpd,fig:abchpdm2}), although 
-the 95\% \gls{HPD} intervals were slightly narrower (average width
-    2183
-for $\gls{m} \geq 1$ and
-    1855
-for $\gls{m} \geq 2$).}
+were analyzed, the estimates from the \textit{env} data were higher
+(HET/Uganda,
+    939 for \textit{gag} vs.
+    1615 for \textit{env};
+HET/Malawi,
+    724 for \textit{gag} vs.
+    845 for \textit{env}).
+As with \gls{alpha}, many of the posterior means were outside the 50\%
+\gls{HPD} intervals, suggesting diffuse posterior distributions.}
 
@@
 from 0 (all the mass on the estimated value) to 5 (the entire prior region).
-When the value $\gls{m} = 1$ was disallowed by the prior, the posterior means
-for all but the HET/Botswana data took on the value 3, with the \gls{HPD}
-intervals spanning the entire prior region. This is very similar to the results 
-observed for \gls{m} on simulated data (\cref{fig:abcptm2,tab:abchpd}), and
-suggests that \gls{m} is not identifiable from these data with this prior. On
-the other hand, the estimate and 95\% \gls{HPD} interval for the HET/Botswana
-data (4 [2-5]) was the same for both priors.}
-
-\add{Estimates of \gls{N} were  mostly uninformative, with very similar estimates
+Estimates of \gls{N} were  mostly uninformative, with very similar estimates
 for all datasets (mean
-    6234,
+    6212,
 range
@@
 posterior mean always fell around the upper two-thirds mark of the region
-allowed by the prior. The results did not change appreciably when the value
-$\gls{m} = 1$ was assigned zero prior probability.}
+allowed by the prior (\cref{fig:abcpt2}).}
+
+\add{When the value $m = 1$ was disallowed by the prior, the separation in
+$\alpha$ between the IDU datasets and the others became more striking
+(\cref{fig:abchpdm2}). All three IDU datasets had estimated $\alpha$ values at
+or above
+    1.
+The estimate for the MSM/Beijing data was slightly lower
+    (0.85) and
+the estimates for the seven remaining non-IDU datasets were bounded above by
+    0.58. 
+The values of $I$ were fairly robust to the choice of prior (compare
+\cref{fig:abchpd,fig:abchpdm2}), although the 95\% HPD intervals were slightly
+narrower (average width
+    2244
+for $m \geq 1$ and
+    1848
+for $m \geq 2$). The posterior means of $m$ for all but the HET/Botswana data
+took on the value 3 with this prior, with the HPD intervals spanning the entire
+prior region. This is very similar to the results observed for $m$ on simulated
+data (\cref{tab:abchpd}), and suggests that $m$ is not identifiable from these
+data with this prior. The results for $N$ did not change appreciably between
+the two choices of prior.}
+
+\add{For the two datasets we reanalyzed using
+RAxML~\autocite{stamatakis2014raxml} and LSD~\autocite{to2016fast}, $\alpha$
+was relatively robust to the choice of method (\cref{fig:methods}, posterior
+means 
+    0.48
+vs. 
+    0.48
+for mixed/Spain and 
+    1.02
+vs.
+    1.12
+for IDU/Estonia). However, the estimates of $I$ were about twice as high when
+RAxML was used instead of FastTree to reconstruct the trees 
+    (228 
+     vs. 437 for IDU/Estonia, 
+    816
+    vs. 1949 for mixed/Spain). }
 

### 2016-07-27 doc/thesis/preface.tex
Defence version of thesis with changes

@@
 ``Likelihood-free estimation of contact network parameters from viral
-phylogenies'' is scheduled for presentation at the Intelligent Systems for
-Molecular Biology meeting on July 8, 2016, in Orlando, Florida, USA.
+phylogenies'' was presented at the Intelligent Systems for Molecular Biology
+meeting on July 8, 2016, in Orlando, Florida, USA.
 
@@
 reviewed and approved by the UBC/Providence Health Care Research Ethics Board
-(H07-02559). Rosemary M. McCloskey completed the Tri-Council Policy Statement:
-Ethical Conduct for Research Involving Humans Course on Research Ethics (TCPS
-2: CORE) tutorial on June 8, 2016.
+(H07-02559). 
+%Rosemary M. McCloskey has completed the Tri-Council Policy Statement: Ethical
+%Conduct for Research Involving Humans Course on Research Ethics (TCPS 2: CORE)
+%tutorial.
 
